### 2019-10-19 repro/papers/mps_theory/response.tex
repro docs dir added

\title{Responses to reviewer's comments on ``Discretization of Multipole Sources in a Finite Difference Setting for Wave Propagation Problems''}
\date{}
\address{
        \footnotemark[1]Institute for Computational and Experimental Research in Mathematics,\\ Brown University,\\ Providence, RI 02912 USA\\
        \footnotemark[2]The Rice Inversion Project,\\ Rice University,\\ Houston, TX
        77005-1892 USA
}
\author{Mario J. Bencomo\footnotemark[1] and William Symes\footnotemark[2]}

\righthead{Responses}

\maketitle
\parskip 12pt
\section{Introduction}
We thank the reviewers for insightful and constructive comments. Their insights led to many changes in this manuscript. We trust that the revised manuscript may answer the questions posed, and generally provide a much improved overview of our work.

\section{Reviewer 1}

Comment: The article presents an extension to the theory on moment conditions for finite differences to multipole point sources. The moment conditions are first derived in a continuous setting followed by the derivation of a discrete set of moment conditions. Further, a convergence theory for staggered grid finite-difference schemes on a family of first order systems of hyperbolic PDEs with multipole source terms is presented. The authors conclude the article by testing the convergence by numerical experiments on the acoustic wave equation.

In the introduction the authors claim that the presented work is an extension of the theory on point source discretization in finite differences. The moment conditions in the first section appears as a valuable extension to current theory, but the convergence analysis is restricted to a specific formulation of hyperbolic PDEs solved with staggered grid finite differences. My interpretation is that these results are not general for all stable finite difference discretisations. If the theory is more general than my interpretation, this must be stated more clearly. Otherwise, the reason for the restriction to staggered grid-finite differences should be explained.

Response: The introduction has been rewritten to emphasize the focus on staggered grid discretizations for symmetric hyperbolic systems of ``div-grad'' form.  The latter does not seem a severe constraint to us, as several physically important wave theories can be cast in that form, and others (e.g. several forms of viscoelasticity) can be viewed lower-order perturbations of systems of this type. The focus on staggered grid schemes is a more serious limitation. It is conceivable that results like those proven here could be established for other stable finite difference schemes, or for finite element schemes of various types, however such generalization is beyond the scope of this paper. The key ingredient in our arguments is the discrete energy estimate proved in the theory section. However as is well-known such an estimate is only a sufficient condition for stability, hence convergence.

%I think the for the most part, the paper does a good job in specifying that the analysis and results are limited to staggered-grid FD schemes. Hopefully this is more apparent now that the theory section has been cleaned up. I added some extra sentences in the conclusion section to emphasize the integral components of the theory. The motivation for using staggered-grid methods was briefly mentioned in the intro (page 4, bottom paragraph). Not sure if it is worth mentioning again. Would it be a good idea to “staggered-grid” somewhere in the title?

Comment: In segments of the article, the line of argument is difficult to follow. For example, the subsection "Examples" on p. 23 appears out of context and the subsection is not well written. In the introduction of this subsection numerical convergence tests in the "next section" are mentioned but the next section does not contain such tests. Figure references are also missing in this section. This subsection must be rewritten and perhaps moved to the section "numerical tests and results".

Response: This section has been cleaned up, and figure references added.

Comment: The structure of the convergence theory is difficult to follow since the more general system which is actually analyzed (21) is mixed with an example (23). I think the example causes confusion about what is actually proven. I believe that the example would fit better in the section "numerical tests and results”.

Response: The convergence theory section has been extensively reworked. All references to the acoustic case have been moved from the theory section to a subsection titled “Acoustic Case” at the end of the respective section.

Comment: The way of presenting the convergence result is non-standard. I suggest that you perform a convergence study with several grid refinements presented similar to figure 7 in Petterson et al. 2016. If if this is not added, a motivation of the current form of convergence test is needed.

Response: Convergence rates are presented in this manner to demonstrate their spatial dependence, in particular, relative to source location in 2D space. For example, this presentation highlights the degradation in convergence rate far from the source location when using source approximations with the wrong approximation order. The convergence rates reported here are equivalent to the slopes of typical convergence figures in log-log plots.

Comment: In figure 6 (a) a green line which corresponds to third order convergence appears vertically in the figure. In figure 7 (a) one can also detect some green areas surrounding the source. Can you comment on this locally lower convergence rate?

Response: I think the author mean figures 6(c) and 7(c). The green line in fig 6(c) coincides with the nodal plane of the dipole.

\section{Reviewer 3:}

Comment: Your analysis is restricted to the Cauchy problem (periodic case), but that is not stated anywhere (please state).  Many applications require treatment of point sources on the boundary. In this case, the moment conditions need to be modified to account for the quadrature rule used by the difference scheme, as shown in Petersson 2016.

Response: Quite correct, and intentional. We do not limit the analysis to the periodic case, but to the full-space case, assuming always that solutions have compact support in space for each time.  It is certainly possible to extend the results to some combinations of boundary conditions and interior and boundary schemes, so long as the latter admit energy estimates. We know of no such staggered grid boundary schemes for the elastic free surface of order higher than two, for example (though we are aware of widely used approximations, such as Robertsson's vacuum scheme). Also we do not study sources supported on the boundary. All of these other questions are interesting, but beyond the scope of the paper. We include explicit statements to that effect in the rewritten introduction.

%Boundary conditions are not required in the analysis, except implicitly in satisfying the skew-adjoint relations in the continuum and discrete systems. In the numerical examples, we use PML boundaries to simulate an unbounded domain, though we implement reflective boundary conditions at the computational domain, consistent with the theory. Source discretization near boundaries is not considered here. Not sure what could be added to the paper.

Comment: Before (16) on page 19, invervals $\rightarrow$ intervals

Response: fixed

Comment: On page 33 "corollary above", what corollary? do you mean theorem?

Response: fixed

Comment: Page 36, did you mean $\tilde{u}$ and not $\tilde{p}$?

Response: fixed.

Comment: Page 37, does $\pm$ indicate addition and subtraction?

Response: Yes.

Comment:  You rely on $P_h = -P_h^T$ for your proofs, but this property is never stated (only for the continuous problem).  It is now mentioned explicitly in the presentation of the general continuum problem and its discretization by the skew-adjoint relations (23) and (25) respectively.

Response: Actually this statement (that $P_h$ must be skew adjoint) was an error on our part, and has been removed entirely from the rewrite - the acoustic example, for instance, does not have this property. The matrix operator $\left(\begin{array}{cc}0 & P^T\\-P & 0\end{array}\right)$ is skew adjoint, for any P of the type considered here (continuous or discrete).

Comment: Page 38, How can you bound ${\cal E}_2$ using $L^2$-error estimates of finite difference solutions for smooth problems, when the solution to the problem is not smooth?

Response: $(\tilde{u}^{n-1/2},\tilde{v}^n)$ solves a FD approximation to the adjoint problem with smooth data $\tilde{f},\tilde{g}$, hence satisfies an energy estimate. The point of using the adjoint formulation is that it replaces inner products of  singular solutions and smooth test functions with inner products of singular sources and smooth solutions with the test functions serving as sources.

Comment: At no point in your proof do you make use of the fact that you have a staggered finite difference discretization.

Response: This is an astonishing statement, in view of the formulation of the abstract staggered grid scheme (equation 24) and its use througout the rest of the paper. We are at a loss as to what the reviewer means.

Comment: However, if the discretization is replaced by the collocated grid, central finite discretizations (e.g., -1/2h, 0,1/2h), then smoothness conditions are necessary for convergence. To prove convergence without imposing smoothness conditions, I believe, at some point you have to invoke the property that the Nyquist mode $(-1)^j$ does not lie in the nullspace of the staggered grid difference operators.

Response: if the Nyquist mode lay in the nullspace of the finite difference propagator, then an energy estimate of the type we prove would be impossible. Conversely, the energy estimate shows that no such phenomenon can occur, provided that the CFL criterion is satisfied.

%We can apply L_2-error estimates to \mathbcal{E}_2 since \delta{\tilde u} and \delta{\tilde v} refer to the finite difference error for a problem with smooth sources (\tilde f, and \tilde g), i.e., see equation (35).  We did most of the heavy lifting for the proof of theorem 8 (weak convergence) a bit ahead of the actual theorem.  May have contributed to some confusion for the reviewer?

Comment: Page 42, how do you define the minimum number of grid points per wavelength? Usually, the frequency at 5\% of
peak amplitude in the spectral domain is defined as $f_{max}$ and then the minimum wavelength is estimated by $ \lambda_{min} = c_{min}/f_{max}$, where $c_{min}$ is the minimum wavespeed. For Ricker wavelets, one obtains $f_{max} = 2.5 f_p$ where $f_p$ is the peak frequency.

%Dr. Symes, not sure how to handle this? Would it be sufficient to reference a paper stating the rule of thumb of 10 grid points per wavelength for 2-2 and 5 grid points per wavelength for 2-4? If so, could you remind me of the reference?

Response: We have removed this paragraph. In our study of convergence rates, gridpoints per wavelength is not a relevant concept, though it might be used to encapsulate the results of such a study.

Comment: Page 43, You state "... our narrow source approximations seem twice as wide and therefore smoother, from the point of view of the difference operator $P_h$. For this reason, smoothness conditions are not necessary for staggeredgrid finite difference schemes". I do not follow this argument. Why would stencil width imply smoothness? As previously mentioned, smoothness conditions are not necessary for convergence when the Nyquist mode does not lie in the nullspace of the difference operator.

%Response: The stencil width and smoothness comment was referring to smoothness in Fourier space. For example, the Fourier transform of a wider the hat function is a narrower $sinc^2$ function, thus the contribution from spurious modes is diminished. Incidentally, source discretizations presented by Petersson 2016 that satisfy m-moment and ssmoothness conditions (with m=s) coincide with approximations that satisfy m-moment conditions and have two times the stencil width relative to the finite difference operator. These are the same source approximations used here.

Response: we have removed this discussion, as it is redundant. We have shown in the theory section that for the class of schemes considered here, the moment conditions imply convergence for multipole rhs (in the weak sense, of course), without need to appeal to any additional conditions such as the smoothness condition of  Petersson et al.

Comment: Page 44, Instead of saying convergence theory "applicable to a large set of wave propagation problems", why not
simply say "wave equations"? In other places you say "symmetric hyperbolic systems, which also can change to
"wave equations" to be more specific.

%Dr. Symes, any thoughts on this?

Response: We are not sure why the referee makes these suggestions. Symmetric hyperbolic system are first order systems of PDEs with wave-like solutions. They do not encompass all wave equations (or even the ``classic'' wave equation), but most wave propagation dynamics can be (re)formulated in terms of such systems. So as we say we are mystified, and with apologies for being so obtuse, have left the prose in question intact.

### 2019-10-19 repro/caam641/basic/paper.tex
repro docs dir added

\title{Basic Imaging of a 2D Marine Survey}
\maketitle
\label{ch:basic}
\inputdir{project}

\section{Introduction}
The purpose of this paper is to explain how to use simple scripts and public-domain software to create images of the subsurface from reasonably tame 2D seismic reflection data. By ``reasonably tame'', I mean having few amplitude anomalies and mostly primary reflection energy. These conditions pretty much imply high quality shallow water marine data from regions with low structural relief and well-defined dip trends, shot along dip lines, with successful multiple removal applied.

The Viking Graben (or Mobil AVO) data placed in the public domain by Mobil Research in 1994 appears to be reasonably tame, in the sense just described.``Viking Graben'' refers to a 25 km long marine line from the North Sea Norwegian sector. The data released by Mobil included both raw field tape ({\tt tape1093}) and parabolic Radon demultiple ({\tt paracdp}) version of the data, along with two logs of $v_p$, $v_s$, and $\rho$, and a far-field wavelet recording. Data are in SEGY format (IBM 4-byte floats, plus text and binary reel headers).

Robert Keys and Douglas Foster were the organizers of the SEG post-convention workshop at which the initial results of various groups' efforts to invert this data were presented. Besides the workshop report volume \cite[]{FosterKeys:98}, Keys and Foster have recently published a web page on this data:
\begin{verbatim}
s3.amazonaws.com/open.source.geoscience/open_data/
Mobil_Avo_Viking_Graben_Line_12/mobil_avo.html
\end{verbatim}

This section of the CAAM 641 course notes discusses so-called standard processing of the Viking Graben data, based on hyperbolic normal moveout (NMO). The connection between inversion and standard processing is the subject of the first part of this course. An earlier treatment can be found in my MSRI lecture notes from 2013, available on the TRIP web page, downloadable materials, short course and summer school materials, especially part 2.4. I also show how to transform the by-products of standard processing into suitable input for prestack imaging via RTM.

My workflow is recorded in the {\tt project/SConstruct} file. The reader should refer to this file for the detailed form of the commands used. The following sections outline the rationale for the choice of commands and their sequence. To reproduce my results, the reader will need to install SU, Madagascar, and TRIP packages. SU should be version 44 or later. The prestack part of the workflow requires parallel installation of TRIP, and at least 10 threads of execution, preferably more, to finish in reasonable time.

\section{Preliminary Steps}
Extract the data using {\tt Fetch}, from the TRIP data repository. The result is {\tt paracdp.segy}, a SEGY-format file. The SU command {\tt segyread} converts this file to SU format (no reel headers, little-endian 4-byte floats, same number of traces). Besides parabolic radon demultiple, this data was delivered with a rather strict mute applied, with no signal before 1 s. No further mute is applied in the processing sequence described here.

Evidently there are some bad traces in some of the low-number CDPs, with floating point garbage in some of the samples. On average, random binary digits form numbers many orders of magnitude larger than the $O(10^8)$ samples encountered in ``good'' CDPs. So an adequate strategy is to use {\tt sugain} to reject samples in the 99\%ile, and this command is included in the conversion to SU format.

\section{Extract Analysis CDPs}
I chose 4 CDPs (Figure \ref{fig:cdp200}, Figure \ref{fig:cdp700}, Figure \ref{fig:cdp1300}, Figure \ref{fig:cdp2000}) roughly evenly spaced, including the end zones, separated by about 8 km. This is a preliminary guess as to the necessary lateral resolution for velocity analysis. It turned out to be reasonable.

Also, I truncated the traces at 3 s, as the logged zone indicated that only reflection arriving before that time would likely image the exploration targets. This is of course an ex post facto decision, but in fact it becomes clear quickly that most of the energy after perhaps 4 s must be other than primary reflection, either multiples or sideswipe.

\multiplot{2}{cdp200,cdp700}{width=0.45\textwidth}{Left: CDP 200; Right: CDP 700}
\multiplot{2}{cdp1300,cdp2000}{width=0.45\textwidth}{Left: CDP 1300; Right: CDP 2000}

\section{Velocity Analysis}
Create velan panels, one for each CDP: Figure \ref{fig:velan200}, Figure \ref{fig:velan700}, Figure \ref{fig:velan1300}, and Figure \ref{fig:velan2000}. Note that in each case there is a trend of strong bullets or streaks at lower velocities than some nearby higher velocity peaks. The higher velocity peaks are weaker, but pick them anyway. The stronger lower velocity features represent multiple reflections - the suppression of multiple energy notwithstanding. (If you carry out this exercise vvwith the field tapes, you will see nothing but water and near-water velocity features).

\multiplot{2}{velan200,velan700}{width=0.45\textwidth}{Left: Velocity spectrum for CDP 200; Right: Velocity spectrum for CDP 700}
\multiplot{2}{velan1300,velan2000}{width=0.45\textwidth}{Left: Velocity spectrum for CDP 1300; Right: Velocity spectrum for CDP 2000}

To QC your picks, plot the NMO correction of the corresponding CDPs as Figure \ref{fig:nmo200}, Figure \ref{fig:nmo700}, Figure \ref{fig:nmo1300}, and Figure \ref{fig:nmo2000}.

\multiplot{2}{nmo200,nmo700}{width=0.45\textwidth}{Left: NMO corrected CDP 200; Right: NMO corrected CDP 700}
\multiplot{2}{nmo1300,nmo2000}{width=0.45\textwidth}{Left: NMO corrected CDP 1300; Right: NMO corrected CDP 2000}

These CDPs are pretty easy to pick from the velocity spectra; however it is instructive to see what might go wrong. For CDP 2000, suppose you were to pick velocities of 1800 m/s at t=1.2, 1.6. and 1.8  s, and 2200 m/s at t=2.1 s - each of these coordinates with a feature on the velocity panel - instead of 1950, 2050, and 2250 m/s at 1.25, 1.6, and 2.25 s, as was done to create Figure \ref{fig:nmo2000}. Then you would generate Figure \ref{fig:nmo2000lo} instead.

\plot{nmo2000lo}{width=0.45\textwidth}{NMO corrected CDP 2000 with tnmo=0.0,1.2,1.5,1.8,2.1,2.6,2.75 vnmo=1500,1800,1800,1800,2200,2400,2700}

\section{Stack and Post-Stack Migration}

The velocity-denominated quantity determined by flattening NMO-corrected CDP gathers is the so-called RMS velocity, or $v_{\rm rms}$, the root mean square of a local wave velocity expressed as a function of vertical travel time $t_0$:
\begin{equation}
\label{eqn:vrms}
v_{\rm rms}^2(t_0)= \frac{2}{t_0}\int_0^{t_0} d\tau \tilde{v}^2(\tau).
\end{equation}
Vertical traveltime is related to depth by 
\begin{equation}
\label{eqn:zt}
z=\int_0^{t_0}d\tau\tilde{v}(\tau)
\end{equation}
in which $\tilde{v}$ is the local (or {\em interval} wave velocity as a function of $t_0$. Velocity as a function of depth is then given by composing $\tilde{v}$ with the inverse of the tranformation (\ref{eqn:zt}).

This construction can be carried out for every midpoint. If the subsurface structure is sufficiently laterally homogeneous, then to good approximation the actual wave velocity is well-approximated by the interval velocity in the ``well'' below each midpoint. The interval velocity as function of $t_0$ or $z$ can then be used to perform migration of zero-offset data, to good approximation. Also, to good approximation zero-offset data is similar to the stack of NMO-corrected CDPs.

This process requires a velocity at every CDP. So far we have only determined four such RMS velocity profiles. However the command {\tt sunmo} will interpolate between 
midpoints, and extrapolate towards the ends of the line as necessary, and even output the interpolated $v_{\rm rms}$ as a function of $t_0$ and midpoint (keyword {\tt voutfile}). This output is critical, as it can be converted later into other forms of velocity suitable for time and depth migration. Of course, the command also produces a stacked section (Figure \ref{fig:parastack}).

\plot{parastack}{width=0.9\textwidth}{Stack with linearly interpolated, constant extrapolated $v_{\rm rms}$}

Because the deeper events are much less energetically imaged than the shallow, automatic gain control (amplitude equalization) produces a much more informative Figure \ref{fig:parastackagc}. This AGC'd stack will be input to all poststack processes.

\plot{parastackagc}{width=0.9\textwidth}{Automatic Gain Control (AGC) applied to data of Figure \ref{fig:parastack}. Note pronouced diffraction hyperbolae in the deeper part of the secion.}

You will note the precence of many {\em diffraction tails} especially below 2 s. These may be partially collapsed either by poststack time migration using Gazdag's algorithm (Figure \ref{fig:paragazmig}) or, more effectively, by poststack depth migration using one of several methods. Gazdag poststack time migration assumes layered velocity structure, hence uses only the first column of the interval velocity as function of time. Nonetheless Figure \ref{fig:paragazmig} shows considerably less interference from diffraction tails; the graben structures that give this prospect its name are now clearly in evidence.

\plot{paragazmig}{width=0.9\textwidth}{Gazdag's phase-shift time migration applied to the stack (Figure \ref{fig:parastack}), followed by AGC. Diffraction artifacts are largely collapsed to the diffracting points that caused them, and the graben structures that give this prospect its name begin to be clearly visible. This migration assumes horzontal layering, which is not correct on the scale of 25 km.}

Poststack depth migration (a process that will be identified later as an approximation to the adjoint linearized zero-offset modeling operator) does even better - Figure \ref{fig:parapspi} displays the output of Phase-Shift-Plus-Interpolation poststack depth migration, using the velocity $v(z,x)$ depicted in Figure \ref{fig:paravintz} (obtained via use of the SU utility {\tt velconv}, see {\tt project/SConstruct} for details). The image gives reasonable depths for the various horizons. Also, this velocity model is the beginning of the next, prestack phase of Viking Graben processing.

\plot{parapspi}{width=0.9\textwidth}{Gazdag's phase-shift-plus-interpolation post-stack depth migration applied to the stack (Figure \ref{fig:parastack}), followed by AGC. Accounts for lateral velocity variation - uses interval velocity as function of midpoint and depth - more accurately than does the time migration of Figure \ref{fig:paragazmig}. The geology is even more clearly delineated, and depths should approximate those to be obtained by more sophisticated imaging.}

\plot{paravintz}{width=0.9\textwidth}{Interval velocity as function of depth, derived from NMO velocity analysis. Probably not to be taken seriously below 3 km - for initial MVA estimate should extend by contant from 3 km and smooth.}

With an eye to the prestack processing to be discussed later in the course, I include analogous results for downfiltered data. Jie Hou suggested a 5-10-30-40 Hz bandpass filter. The resulting AGC'd stack (Figure \ref{fig:parastackagcf}) shows the expected decrease in resolution over the original data, which has significant energy over 50 Hz.

\plot{parastackagcf}{width=0.9\textwidth}{NMO stack of (5,10,30,40) Hz bandpass filtered data, after AGC. Compare to Figure \ref{fig:parastack}.}

[A question: the process, as you will see from the SConstruct, consists in filtering the data then applying NMO. Since the stack is a collection of time traces, you might think that you could get the same result by applying NMO first, then filtering. Is this true?]

The PSPI post-stack depth migration (Figure \ref{fig:parapspif}) reveals the same structure as does the stack, with the same decreased resolution over Figure \ref{fig:parapspi}, but with diffraction tails suppressed and stretched to depth.

\plot{parapspif}{width=0.9\textwidth}{PSPI post stack migrated image from stack in Figure \ref{fig:parastackagcf}.}

\section{Conclusion}
Use of simplified physics (NMO, stack, poststack migration) based on layered modeling produces plausible subsurface structure images from the Viking Graben data. Public domain software - in this exercise, mostly SU - provides enough functionality to carry out this imaging task, provided that some basic pre-processing is performed (multiple suppression, mute). Mobil provided a suitably preprocessed version of the data in 1994.

The sequence of processing steps takes us far enough from the basic physics of wave propagation to raise questions about whether the images produced here are actually images of anything real. Later in the course, we turn to prestack processing based more directly on the acoustic wave equation - only a step further, but meaningful nonetheless. To telegraph part of the punch: prestack processing supports the validity of the images obtained in this paper by poststack processes. It seems that in the case of this ``tame'' data, classic seismic data processing does not steer you wrong.

\section{Suggested Projects}
\begin{enumerate}
\item The velocity analysis presented above seems quite minimal - only four analysis CMPs selected, more or less evenly spaced. Is there any advantage in choosing more analysis points? Does the stack become cleaner, the interval velocity any more believable?

\item Find another field data set and carry out a similar analysis.

\item Taking another tack, find a synthetic model with more lateral variation in velocity - Marmousi is an obvious choice - and carry out NMO based velocity analysis. Does the interval velocity so obtained bear any relation to the actual velocity (which of course you know, since the example is synthetic)? What about the PSPI migrated image and the actual model?

\item AGC is a crude tool for equalizing amplitudes, that completely destroys the relation between event and reflector dynamics. Is it possible to use a simpler scaling, such as by a power of $t$ or $z$, to  achieve the same enhancement of important reflectors at depth while maintaining a simple relation between reflector and reflection amplitudes?
\end{enumerate}

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/caam641/born/paper.tex
repro docs dir added

\title{Born Approximation}
\maketitle
\label{ch:born}

\inputdir{project}

\section{Introduction}
Several features of the seismic wave field are obvious from field recordings:
\begin{itemize}
\item it consists largely of waves, that is, coherent space-time structures with definite apparent velocities;
\item the velocity of seismic waves varies with position in the earth;
\item some of the waves appear to be reflections, that is, the result of interaction with the earth's structure that changes (or even reverses) the direction of wave motion.
\end{itemize}
The simplest model of wave motion that predicts these features is {\em constant density acoustics},  connecting the acoustic pressure field $p(\bx,t)$, the compressional wave velocity $v(\bx)$, and a source wavefield $f(\bx,t)$ through the partial differential equation
\begin{eqnarray}
\label{eqn:cdawe}
\frac{\partial^2 p}{\partial t^2} -v^2 \nabla^2 p& = & f\\
p &= & 0, \,t<<0 \nonumber 
\end{eqnarray} 
The second condition guarantees that the field is causal, that is, vanishes before the onset of nonzero source wavefield values.

This section takes a first look at the relation between $p$ and $v$ implicit in the system \ref{eqn:cdawe}, and introduces the linearization of this relation, often called (somewhat imprecisely) the Born approximation in the seismic literature. The Born approximation is both a powerful analytical tool and the foundation for much of seismic data processing. It is particularly accurate when the reference model is smooth, or slowly varying, on the wavelength scale, and the perturbation is oscillatory, that is, contains all wavelength-scale features. Under these circumstances, Born data represents the part of the wavefield that has interacted once with the wavelength-scale features in the model, that is, the singly-scattered field, and provides a rough definition of the primary reflection field. These features are illustrated via a model constructed from the NMO-based velocity analysis and PSPI depth image derived from the Mobil Viking Graben data in Chapter \ref{ch:basic}.%\cite{basic-caam641:17}.

\section{Refraction}
Note that the wave equation \ref{eqn:cdawe} does not necessarily predict wave motion: if the velocity field $v$ is sufficiently complex, not only may waves not be observable, but energy may not even flow as one would expect in a wave setting. So called Anderson localization, in which a complex material with highly spatially variable index of refraction ($1/v$) causes energy to remain for arbitrarily long periods in the same location, or to diffuse slowly rather than propagate, has been predicted theoretically and even observed in the lab \cite[]{huetal:08,WrightWeaver:10}. Clear evidence of waves at seismic frequencies already tells us something about the structure of the earth's interior.

The classical radiation solution for the wave equation \cite[]{CourHil:62} is the pressure field $p(\bx,t)$ satisfying \ref{eqn:cdawe} with constant (space-independent) $v$ and $f(\bx,t) = w(t)\delta(\bx)$. As is well-known, the solution in 3D is
\begin{equation}
\label{eqn:cdawe-green}
p(\bx,t) = \frac{w(t-r/v)}{4\pi r}, \,r=|\bx| = \sqrt{\bx^T\bx}.
\end{equation}
The 2D case has a more complex expression:
\begin{equation}\label{eqn:acdsol2dalt}
p(\bx,t)=\frac{1}{\pi v^2} \int_0^{\sqrt{t-r/v}}d\sigma \frac{w(t-\sigma^2-r/v)}{\sqrt{\sigma^2 + 2r/c}}. 
\end{equation}
The constant $v$ radiation solution predicts waves moving at speed $v$, but (of course) their velocity does not vary with position. Also, this model does not generate anything that looks like reflected waves. So, reflection (and even more so, variable wave velocity) implies spatially variable wave velocity.

Next examine wave propagation via numerical approximation. Throughout these notes, I use finite difference schemes to model wave propagation. A good overview of finite difference methods appears in \cite[]{Moczoetal:06}. These methods approximate the actual solution with an asymptotic rate called the scheme order. Actually only some solutions are approximated with this order, in general - see \cite[]{SymesTerentyevVdovina:08} or \cite[]{SymesVdovina:09} for examples in which the formal order does not predict the actual convergence behaviour. In general, for smooth solutions and coefficients (like $v$), the formal order actually predicts the rate of convergence: if the scheme is of second order, the error will be roughly proportional to $\Delta t^2$ for small enough $\Delta t$, for instance. For the examples shown in this section of the notes, we use the classic centered difference scheme, with second order centered difference in time replacing the time derivatives in \ref{eqn:cdawe-green}, and $2k$th order centered differences replacing the spatial derivatives (usually $k=2$, so the scheme is fourth order in space). Singular right hand sides are represented by numerical delta functions.

A slightly more realistic propagating medium from the constant case (equation \ref{eqn:cdawe-green})  is  the  interval velocity predicted by NMO velocity analysis from the Mobil AVO data \cite[]{basic-caam641:17}, which is slowly varying on the wavelength scale. [Average velocity = 2500 m/s, wavelength at 20 Hz = 125 m.] Figure \ref{fig:paracsq} shows the square of this interval velocity. Since we have no density information in this model, this field may be regarded as proportional to bulk modulus, with constant density.

Up to this point boundary conditions have not been mentioned: the radiation solution does not interact with any boundaries. The problem solved numerically in the following examples takes place in a bounded rectangle in ${\bf R}^3$, namely the one shown in Figure \ref{fig:paracsq}. Every boundary is subject to the pressure-free condition $p=0$. The domain is so chosen that for the point sources used in the numerical experiments, the interaction of the wavefield with the side and bottom boundaries does not affect the simulated trace data - as will be evident shortly. The free surface at the top of the model does interact with these solutions, however.

\plot{paracsq}{width=0.9\textwidth}{Interval velocity as function of depth, derived from NMO velocity analysis. Probably not to be taken seriously below 3 km.}

Figures \ref{fig:movereflooverlay-frame05}, \ref{fig:movereflooverlay-frame10} and \ref{fig:movereflooverlay-frame15} show the computed pressure wavefield at $t=0.5s, 1.0s$ and $1.5s$, for a shot at $x_s=10012.5 m$ (in the coordinate system of the trace headers).
The wavelet is a 5-10-20-25 Hz zero-phase bandpass filter.

\plot{movereflooverlay-frame05}{width=0.9\textwidth}{Pressure field at $t=0.5s$ for a point source at $x=10012$ m. Wavelet is  5-10-20-25 Hz zero-phase bandpass filter. Velocity field derived from NMO velocity analysis, square depicted in Figure \ref{fig:paracsq}.}

\plot{movereflooverlay-frame10}{width=0.9\textwidth}{Pressure field at $t=1.0s$ for a point source at $x=10012$ m. Wavelet is  5-10-20-25 Hz zero-phase bandpass filter. Velocity field derived from NMO velocity analysis, square depicted in Figure \ref{fig:paracsq}.}

\plot{movereflooverlay-frame15}{width=0.9\textwidth}{Pressure field at $t=1.5s$ for a point source at $x=10012$ m. Wavelet is  5-10-20-25 Hz zero-phase bandpass filter. Velocity field derived from NMO velocity analysis, square depicted in Figure \ref{fig:paracsq}.}

There is little evidence of reflection occuring in this example. The shot gather (all acquisition parameters as in the field data, reported in \cite[]{basic-caam641:17}) shows the direct wave and a diving wave arrival - at longer offsets, the latter will overtake the former.

\plot{data10012}{width=0.9\textwidth}{Pressure shot gather for shot at $x=10012$ m. Wavelet is  5-10-20-25 Hz zero-phase bandpass filter. Velocity field derived from NMO velocity analysis, square depicted in Figure \ref{fig:paracsq}.}

\section{Reflectors and Reflection}
From the homogeneous and slowly varying examples, one could infer that only transmitted (direct and diving wave) arrivals occur in models for which the wave velocity (and eventually other parameters) change slowly (are smooth) on the scale of a wavelength. This supposition is correct, and will be fully justified in the next section \cite[]{raytheory-caam641:17}. Therefore, to make reflections appear in model-derived synthetics requires a model having substantial variation on the wavelength scale.

We create such a model by adding a scaled version of the PSPI depth-migrated image from \cite[]{basic-caam641:17}, shown in Figure \ref{fig:parapspimdlpert}. Keeping the velocity between 1.0 and 5.0 km/s by using a scale factor of 0.15, we obtain the model shown in Figure \ref{fig:parapspimdl}. This model has wavelength scale features, and we would expect it to generate reflections. In fact it does, as can be seen in the snapshots at 0.5, 1.0, and 1.5 s (Figures \ref{fig:moviereflopoverlay-frame05}, \ref{fig:moviereflopoverlay-frame10} and \ref{fig:moviereflopoverlay-frame15}). The shot gather, Figure \ref{fig:data10012p} appears to be rich in reflected events.

\plot{parapspimdlpert}{width=0.9\textwidth}{PSPI depth migration of Mobil AVO data using square velocity of Figure \ref{fig:paracsq}.}

\plot{parapspimdl}{width=0.9\textwidth}{Interval squared velocity as function of depth, combines background model of Figure \ref{fig:paracsq} with PSPI poststack image, scaled by 0.15.}

\plot{moviereflopoverlay-frame05}{width=0.9\textwidth}{Pressure field at $t=0.5s$ for a point source at $x=10012$ m. Wavelet is  5-10-20-25 Hz zero-phase bandpass filter. Velocity field depicted in Figure \ref{fig:parapspimdl}.}

\plot{moviereflopoverlay-frame10}{width=0.9\textwidth}{Pressure field at $t=1.0s$ for a point source at $x=10012$ m. Wavelet is  5-10-20-25 Hz zero-phase bandpass filter. Velocity field depicted in Figure \ref{fig:parapspimdl}.}

\plot{moviereflopoverlay-frame15}{width=0.9\textwidth}{Pressure field at $t=1.5s$ for a point source at $x=10012$ m. Wavelet is  5-10-20-25 Hz zero-phase bandpass filter. Velocity field depicted in Figure \ref{fig:parapspimdl}.}

\plot{data10012p}{width=0.9\textwidth}{Pressure shot gather for shot at $x=10012$ m. Wavelet is  5-10-20-25 Hz zero-phase bandpass filter. Velocity field depicted in Figure \ref{fig:parapspimdl}.}

Recall that the reference data (Figure \ref{fig:data10012}) created from the slowly varying model (Figure \ref{fig:paracsq}) contains only direct and refracted waves. These are also visible in the data (Figure \ref{fig:data10012p}) created from the model with reflectors (Figure \ref{fig:parapspimdl}), and the wavefield snapshots also suggest that the transmitted waves persist. It is natural to speculate that the difference between these two shot gathers might contain only reflections: we plot this difference in Figure \ref{fig:data10012res}.

\plot{data10012res}{width=0.9\textwidth}{Difference between shot gathers for $v^2$ and $v^2 + \epsilon \delta v^2$, with $\epsilon=0.15$ (shown in Figures \ref{fig:data10012} and \ref{fig:data10012p}).}

\section{Linearization}
Of course the relation between $v$ and $p$ implied by \ref{eqn:cdawe} is nonlinear, so the change between the shot gathers of Figures \ref{fig:data10012} and \ref{fig:data10012p} cannot be linear in the perturbation (Figure \ref{fig:parapspimdlpert}). On the other hand, the RMS (or $L^2$ norm) of the perturbation is approximately 2\% of the $L^2$ norm of the square velocity, so in some sense a small perturbation, and it is natural to think that a linear approximation to the model-data relation would be predictive of the data change. Linearization, or (as it is somewhat imprecisely known in this business) the Born approximation. amounts to replacing both $p$ and $v$ by perturbed quantities $p + \delta p$, $v+\delta v$, substituting these in equation \ref{eqn:cdawe}, throwing away all terms with two or more $\delta$s in them, and using \ref{eqn:cdawe} itself to eliminate a few more terms. The upshot is a relation between the reference fields $p, v$ (presumed to solve \ref{eqn:cdawe}) and perturbation fields $\delta p, \delta v$:
\begin{eqnarray}
\label{eqn:pcdawe}
\frac{\partial^2 \delta p}{\partial t^2} -v^2 \nabla^2 \delta p& = & \delta v^2 \nabla^2 p\\
\delta p &= & 0, \,t<<0 \nonumber 
\end{eqnarray} 
This equation may be approximated numerically by the same kind of finite difference method explained above for the reference system. The perturbation field  is sampled in exactly the same way as the other shot gathers shown here to produce Figure \ref{fig:data10012born}. If linearization is an accurate approximation, then this field should approximate the difference between the shot gathers produced from the model with reflectors (Figure \ref{fig:data10012p} and the one without (Figure \ref{fig:data10012}), plotted as Figure \ref{fig:data10012res}. Indeed the two look quite similar: the difference between the data in Figure \ref{fig:data10012res} and FIgure \ref{fig:data10012born} is displayed in Figure \ref{fig:data10012bornres}, all three figures plotted on the same grey scale.

\plot{data10012born}{width=0.9\textwidth}{Born data: shot gather at shot position $x=10012 m$ with the reference model ($v^2$) of Figure \ref{fig:paracsq} and the perturbation ($\delta (v^2)$) of Figure \ref{fig:parapspimdlpert}, scaled by $\epsilon=0.15$ .}

\plot{data10012bornres}{width=0.9\textwidth}{Difference between Born shot gather (Figure \ref{fig:data10012born}) and residual shot gather (Figure \ref{fig:data10012res}) for perturbation scaled by $\epsilon = 0.15$, all three plotted on same grey scale.}

The events appearing in Figure \ref{fig:data10012born} are found in Figure \ref{fig:data10012res}, which however contains other events as well: careful examination shows that these other events tend to be slower than those retained in Figure \ref{fig:data10012born}. These slower events are multiple reflections, bouncing between the free surface at the top of the model (a perfect reflector) and the various reflectors evident in Figure \ref{fig:parapspimdl}, especially those near the surface.

To understand better the relation between the full model (equation \ref{eqn:cdawe}) and the Born model (equation \ref{eqn:pcdawe}), and set the stage for what follows, introduce the {\em forward map} or {\em modeling operator} $\cF$, mapping the square velocity $v^2$ to the trace data:
\begin{equation}
\label{eqn:mdlop}
\cF[v^2] = \{p(\bx_r,t;\bx_s)\}
\end{equation}
in which $\bx_r,\bx_s$ are presumed to run over the combinations of receiver and source locations present in the data. Similarly the linearized forward map, or Born modeling operator, $D\cF$ is defined by
\begin{equation}
\label{eqn:pmdlop}
D\cF[v^2]\delta v^2 = \{\delta p(\bx_r,t;\bx_s)\}.
\end{equation}
The Born modeling operator is actually the derivative of the modeling operator, as the notation suggests, in a suitable sense: this is proven in the theoretical part of this course. In fact, the modeling operator is twice differentiable, so long as the pulse is bandlimited as are the sources used here. One would therefore expect that the remainder of the first order Taylor series should decrease like the second power of the perturbation length:
\begin{equation}
\label{eqn:taylor1}
\cF[v^2 + \epsilon \delta v^2] - \cF[v^2] - \epsilon D\cF[v^2]\delta v^2 = O(\epsilon^2).
\end{equation}
Using the common notation $\|\cdot\|$ for the $L^2$ norm (RMS scaled by cell volume), this would imply
\begin{equation}
\label{eqn:taylor1l2}
\|\cF[v^2 + \epsilon \delta v^2] - \cF[v^2] - D\cF[v^2]\delta v^2 \| \approx K \epsilon^2
\end{equation}
for a constant $K$ depending on everything in sight. We can see this relation by measuring the $L^2$ norms of the Taylor series remainder, which you have already seen, namely as Figure \ref{fig:data10012bornres}, for the choices of $v^2$ (Figure \ref{fig:paracsq}) and $\delta v^2$ (Figure \ref{fig:parapspimdlpert}) introduced above, and $\epsilon = 0.15$. If you replace $\epsilon$ by $0.05$, then the norm of the first order Taylor series residual should drop by roughly a factor of $9$, according to the formula \ref{eqn:taylor1l2}. In fact the norms are respectively 0.0027 ($\epsilon=0.15$) and 0.00029 ($\epsilon = 0.05$), and differ almost exactly a factor of 9, as predicted. The residual and Born approximation (linear term in the Taylor series) appear as Figures \ref{fig:data10012res05} and \ref{fig:data10012born05} respectively, plotted on the same scale as the remainder in the first order Taylor series in \ref{fig:data10012bornres05}.

\plot{data10012res05}{width=0.9\textwidth}{Difference between shot gathers for $v^2$ and $v^2 + \epsilon \delta v^2$, with $\epsilon=0.05$.}

\plot{data10012born05}{width=0.9\textwidth}{Born data: shot gather at shot position $x=10012 m$ with the reference model ($v^2$) of Figure \ref{fig:paracsq} and the perturbation ($\delta (v^2)$) of Figure \ref{fig:parapspimdlpert}, scaled by $\epsilon=0.05$ .}

\plot{data10012bornres05}{width=0.9\textwidth}{Difference between Born shot gather (Figure \ref{fig:data10012born05}) and residual shot gather (Figure \ref{fig:data10012res05}) for perturbation scaled by $\epsilon=0.05$, all three plotted on same grey scale.}

Another important fact to learn about the Born approximation is that its accuracy reflects {\em scale separation}, and the first order Taylor series remainder is much smaller when the perturbation contains little energy on the scale of a wavelength. This freedom from long-scale components is a feature of the PSPI image \ref{fig:parapspimdlpert} used as $\delta v^2$ in the preceding examples, which accounts for the relative accuracy of the Born approximation. If instead one uses the perturbed model (Figure \ref{fig:parapspimdl}) as reference ($v^2$) and perturbs by a multiple of the smooth background model (Figure \ref{fig:paracsq}) ($\delta v^2$), a much larger first order Taylor remainder results. For example, with $\epsilon = 0.02$ to give a perturbation of the same $L^2$ size as the PSPI image, but with $v^2$ and $\delta v^2$ as just suggested, the residual shown in Figure \ref{fig:data10012reslf} results, which may be compared to the Born approximation (linearization) shown in Figure \ref{fig:data10012bornlf}. The larger difference is evident to the eye: the norm of the difference (first order Taylor series remainder) is 0.0089, actually larger than the residual norm of 0.0077.  In contrast, the corresponding numbers in the preceding example were 0.0027  and 0.0048. This phenomonon - the more rapid variation of $\cF$ in smooth directions - is key to understanding many aspects of the seismic inverse problem. The next section of this course will give some inkling about why it occurs, but the full mathematical story has yet to be told.

\plot{data10012reslf}{width=0.9\textwidth}{Difference between shot gathers for $v^2$ and $v^2 + \epsilon \delta v^2$, with $v^2=$ the perturbed model (Figure \ref{fig:parapspimdl}), $\delta v^2 = $ the smooth background model (Figure \ref{fig:paracsq}), and  $\epsilon=0.02$, chosen to create a perturbation of the same $L^2$ size as the PSPI image (Figure \ref{fig:parapspimdlpert}).}

\plot{data10012bornlf}{width=0.9\textwidth}{Born data: shot gather at shot position $x=10012 m$ with the reference, perturbation, and scale factor $\epsilon$ as in Figure \ref{fig:data10012reslf}, plotted on same scale.}

\plot{data10012bornreslf}{width=0.9\textwidth}{Difference between Born shot gather (Figure \ref{fig:data10012bornlf}) and residual shot gather (Figure \ref{fig:data10012reslf}), plotted on same scale.}

\section{Project Notes}
The computational results for this section depend on those of the previous section \cite[]{basic-caam641:17}, and the corresponding project must be built first.

Along with the displays in the paper, the project build ({\tt scons}) creates two propagating wave movies, from which the snapshots in the paper were taken. To view these, use the {\tt xtpen} command in the project directory:
\begin{verbatim}
xtpen < Fig/moviereflooverlay.vpl 
\end{verbatim}
for the radiation solution from a point source in the smooth square velocity model (Figure \ref{fig:paracsq}), and
\begin{verbatim}
xtpen < Fig/moviereflopoverlay.vpl 
\end{verbatim}
for the radiation solution from a point source in the square velocity model with reflectors (Figure \ref{fig:parapspimdl}).

The $L^2$ norms quoted in the text were obtained using {\tt sfattr} - flows are provided in {\tt project/SConstruct} that write the relevant line out to a text file.

\section{Suggested Projects}
\begin{enumerate}
\item Just how separated are the scales between the ``background'' model (Figure \ref{fig:paracsq}) and reflectivity or perturbation (Figure \ref{fig:parapspimdlpert})? This question can be explored using the 2D fft implemented in Madagascar.
\item create examples with well-separated scales and investigate the linearization error for perturbations in smooth background vs. rough reflectivity.
\end{enumerate}

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/caam641/conv/paper.tex
repro docs dir added

\title{The Convolutional Model}
\maketitle
\label{ch:conv}
\inputdir{project}

\section{Introduction}
The model of the title relates a seismic data trace $d(t)$, a {\em reflectivity} $r(t)$, a wavelet $w(t)$, and a noise function $n(t)$ by
\begin{equation}
\label{eqn:convmod}
d = w * r + n
\end{equation}
An enormous literature addresses various properties of this model - see \cite{RobTrei:80} and many works cited there. A large part of this literature is deveoted to suppressing or otherwise managing the influence of the noise term $n$ on the recovery of $r$ from $d$ given $w$ and just some properties of $n$.

Equation \ref{eqn:convmod} is understood to hold for every data trace, after NMO correction as carried out as in Chapter \ref{ch:basic}. Two questions suggest themselves: 
\begin{itemize}
\item what is NMO correction, and 
\item why does it lead to the very simple relation \ref{eqn:convmod}, presumed to hold for every source-receiver combination?
\end{itemize}
The aim of this chapter is to answer these two questions, {\em starting with the wave equation}. The approximations developed in the preceding two chapters play central roles in the answer.

\section{Integral Representation of the Singly Scatterered Field}
At first, one might be tempted to say, well {\em of course} the convolution relation \ref{eqn:convmod} holds. Assuming that the pressure field $p(\bx,t;\bx_s)$ is a solution of the wave equation \ref{eqn:raycdawe}, with isotropic point radiator source $f(\bx,t) = w(t)\delta(\bx-\bx_s)$, then the representaion of the solution by the Green's function (equation \ref{eqn:green}) implies that
\begin{equation}
\label{eqn:greenpt}
p(\bx_r,t;\bx_s) = \int\, dt_r\, w(t-t_r) G(\bx_r,t_r;\bx_s).
\end{equation}
If we model the data as traces of the pressure field, then this would appear to give us the convolutional model \ref{eqn:convmod}, though without the noise term on the right-hand side. Apparently $r(t)$ is idenfied as the same trace of the Green's function. So Q. E. D.

There are two big things wrong with this point of view. First, the Green's function is a pressure field - a particular solution of the wave equation. It is not a property of the material. However, the reflectivity $r(t)$ is meant to correspond {\em directly and locally}  to properties of earth material - and there is no obvious reason why the Green's function should have any relation to a local measurement of these properties (density, stiffness, etc.). In fact, $r(t)$ must somehow encode properties at depth, whereas the Green's function is evaluated at the source and receiver points in equation \ref{eqn:greenpt}, and at best encodes all of the properties that the field encounters in propagating over the time interval $[0,t_r]$. Second, nominallly $G$ is a pressure, and even if it were a local earth property (which it is not), it is not clear what combination of the basic parameters of the acoustic model it represents, or why.

We can make some progress towards a resolution of this conundrum by assuming a {\em known background square velocity $v^2$} and regarding the data as traces of the perturbational field due to a perterbation $\delta v^2$. The perturbational pressure field $\delta p$ satisfies the inhomogeneous wave equation \ref{eqn:pcdawe}, reproduced for convenience here:
\begin{eqnarray}
\label{eqn:pcdaweconv}
\frac{\partial^2 \delta p}{\partial t^2} -v^2 \nabla^2 \delta p& = & \delta v^2 \nabla^2 p\\
\delta p &= & 0, \,t<<0 \nonumber 
\end{eqnarray} 
The traces of $\delta p$ make up the image of the linearized map $D\cF[v^2]$. Recalling that the solution of an inhomogenous wave equation may be expressed in terms of the Green's function, obtain
\[
D\cF[v^2]\delta v^2(\bx_r,t;\bx_s) = \int \, dt_r \, \int \,dx\, G(\bx_r,t_r;\bx) \delta v^2(\bx) \nabla^2 p(\bx,t-t_r;\bx_s).
\]
Using equation \ref{eqn:greenpt} to replace $p$ results in
\begin{eqnarray}
\label{eqn:intrep1}
= w *_t \int \,dt_r\, \int \, dx \, G(\bx_r,t_r;\bx)\delta v^2(\bx) \nabla^2 G(\bx,t-t_r;\bx_s).
\end{eqnarray}
It is convenient to use the wave equation for $G$ to replace the space derivatives in the integrand \ref{eqn:intrep1} with time derivatives. This is possible under the assumption:
\begin{equation}
\label{eqn:deltasupp}
\delta v^2(\bx) = 0 \mbox{ for } \bx \mbox{ near } \bx_s.
\end{equation}
Granted assumption \ref{eqn:deltasupp}, 
\begin{equation}
\label{eqn:intrep}
D\cF[v^2]\delta v^2(\bx_r,t;\bx_s) = w *_t \frac{\partial^2}{\partial t^2}\int \,dt_r\, \int \, dx \, G(\bx_r,t_r;\bx)\frac{\delta v^2}{v^2}(\bx) G(\bx,t-t_r;\bx_s).
\end{equation}
This relation has the great advantage over equation \ref{eqn:greenpt} that a physical parameter, namely the perturbation in square velocity, appears directly. The influence of this parameter, however still appears to be non-local. To see that it is in fact local, one more level of approximation is needed.

\section{Generalized Radon Transform}
Add to the accumulating hypotheses the assumption that the background square velocity $v^2$ is slowly varying on the wavelength scale, and $\delta v^2$ is rapidly varying (oscillatory) with small mean over regions a wavelength in diameter. Also assume that no caustics develop in the ray fields centered at either source or receiver, at least in the region where $\delta v^2$ is non-zero. Then the result of Chapter \ref{ch:ray} suggests that we might write $G$ in terms of its progressing wave expansion \ref{eqn:progwave}, with $S = \delta$ (3D):
\begin{equation}
\label{eqn:greenprog}
G(\bx,t;\bx_s) = a(\bx,\bx_s)\delta(t-\tau(\bx,\bx_s)) + R(\bx,t;\bx_s)
\end{equation}
Since the first term will be most significant for high frequency waves, keep it and drop the second, insert in \ref{eqn:intrep} to obtain
\[
D\cF[v^2]\delta v^2(\bx_r,t;\bx_s) \approx 
\]
\[
w*_t \frac{\partial^2}{\partial t^2}\int \,dt_r\, \int \, dx \, a(\bx_r,\bx) \delta(t_r-\tau(\bx_r,\bx) \frac{\delta v^2}{v^2}(\bx) a(\bx,\bx_s)\delta(t-t_r-\tau(\bx,\bx_s))
\]
(since $\tau$ and $a$ are symmetric functions of their arguments)
\begin{equation}
\label{eqn:grt1}
= w*_t \frac{\partial^2}{\partial t^2}\int \, dx \, a(\bx_r,\bx)a(\bx_s,\bx)\delta(t-\tau(\bx_r,\bx)-\tau(\bx_s,\bx)) \frac{\delta v^2}{v^2}(\bx)
\end{equation}
That is, apart from the convolution with the wavelet $w$, $D\cF$ acts (approximately) by integrating the relative  perturbation $\delta v^2/v^2$ over the {\em isochron surface of equal two-way traveltime} $\{\bx: t=\tau(\bx_r,\bx)+\tau(\bx_s,\bx)\}$.

One additional assumption needs to be made for this reasoning to be correct: the last step in the above argument was this identity:
\begin{equation}
\label{eqn:noscatoverpi}
\int dt_r \delta(t_r-\tau(\bx,\bx_r))\delta(t-t_r-\tau(\bx,\bx_s)) = \delta(t - \tau(\bx_r,\bx)-\tau(\bx_s,\bx)).
\end{equation}
This identity does not hold in general. To take an extreme example, if $t-\tau(\bx,\bx_s) = \tau(\bx,\bx_r)$ over a small region of $\bx$, then in that region the integrand is the square of a delta function, which is not a legal distribution. This is only a hint of what goes wrong, which I am not going to discuss in detail. Suffice it to quote a result from \cite[]{GelShil:58}: the left-hand side of equation \ref{eqn:noscatoverpi} is well-defined, and equal to the right-hand side, when 
\begin{equation}
\label{eqn:noraymeet}
\nabla_{\bx}\tau(\bx,\bx_r) \ne - \nabla_{\bx}\tau(\bx,\bx_s)
\end{equation}
for all $\bx_r,\bx_s$ in the survey geometry and $\bx$ in the support of $\delta v^2$, that is, the region where $\delta v^2 \ne 0$.

Condition \ref{eqn:noraymeet} has a significant physical interpretation: according to the ray equation \ref{eqn:xeqn}, the gradient of $\tau$ is proportional to the ray velocity vector. Thus condition \ref{eqn:noraymeet} means that the ray from $\bx_r$ to $\bx$ should arriva at $\bx$ with a velocity vector unequal to the {\em negative} of the velocity vector of the ray from $\bx_s$ to $\bx$, which is the velocity vector of the ray from $\bx$ to $\bx_s$. If this condition is violated, then the curve got by concatenating the rays from $\bx_s$ to $\bx$ and from $\bx$ to $\bx_r$ is itself a ray (thanks to the uniqueness theory for ordinary differential equations). Thus in that case there is a ray from $\bx_s$ to $\bx_r$, that is, a direct or diving ray connecting source and receiver. This circumstance must be ruled out if equation \ref{eqn:noscatoverpi} is to hold: specifically, {\em no diving ray may intersect the support of $\delta v^2$.}

Add this no-diving-rays hypothesis to the standing assumptions. Two remarks are worth making:
\begin{itemize}
\item A more refined analysis shows that it's OK for diving rays to pass over the support of $\delta v^2$, so long as directions in which $\delta v^2$ oscillates strongly are not perpindicular to the rays. The ideas necessary to properly formulate and justify this assertion will appear later in the book.
\item It is ironic that contemporary FWI relies almost entirely on diving waves, that is, waves associated to the very rays outlawed by condition \ref{eqn:noraymeet}. This point will also get some attention later in the book.
\end{itemize}

The formula \ref{eqn:grt1} is surprisingly accurate under the conditions assumed, as I showed by explicit example in my old notes \cite[]{SymesNotes}. With numerical methods for computation of traveltimes and amplitudes, \ref{eqn:grt1} turns into a so-called Kirchhoff integral operator approximation to $D\cF$. Such numerical integral operator representations are very useful, and widely used.

\section{Layered Models}
The convolutional model is the result of all of the assumptions and approximations made so far, plus a few more. The first of these, in its strictest form, is that {\em the earth is layered}: that is that the phsyical properties of rock affecting seismic wave propagation (meaning $v^2$, in the constant density acoustic model) depend only on $x_3=z$. Also, the sources and receivers are arrayed on horizontal surfaces, that is, $z = z_s$ and $z=z_r$ respectively.

In fact, the earth is not layered (even locallly) - if it were, then the results of seismic experiments would be translation-invariant, and they are not. For example, Figure \ref{fig:cdp-every50th} shows every 50th {\em Common Midpoint} gather. Common midpoint (``CMP'') gathers are the collections of traces sharing a common {\em midpoint} $x_m=(x_r+x_s)/2, y_m=(y_r+y_s)/2$, more on that later).
The half-offset $h=|(x_r-x_s,y_r-y_s)|/2$ indexes traces in a CMP gather. The scalar half offset defined here suffices for present purposes, since the layered hypothesis implies rotational symmetry of the traces in a CMP about the midpoint.

The CMP gathers in Figure \ref{fig:cdp-every50th} change quite a bit one to another, but they are 625 m apart. Figure \ref{fig:cdp1002-1040} on the other hand shows the same number of CMP gathers, but chosen contiguously, with spacing of 12.5 m. Over this range of CMPs, the data changes quite slowly, so perhaps it is not so dreadfully inaccurate to model the earth as layered on this scale.
\plot{cdp-every50th}{width=0.9\textwidth}{Every 50th Common Midpoint Gather from the Viking Graben line introduced in Chapter \ref{ch:basic}. Midpoint spacing is 625 m.}
\plot{cdp1002-1040}{width=0.9\textwidth}{Common Midpoint Gathers 1002-1040 from the Viking Graben line introduced in Chapter \ref{ch:basic}. Midpoint spacing is 12.5 m.}

Due to the rotational symmetry already noted,
\begin{equation}
\label{eqn:cmpdf}
D\cF[v^2]\delta v^2(\bx_r,t;\bx_s) = D\cF[v^2]\delta v^2(\bx_m + (h,0,0),t; \bx_m-(h,0,0))
\end{equation}
This identity suggests a change of coordinates: replace
\begin{equation}
\label{eqn:cmpdfc}
D\cF[v^2]\delta v^2(x_m,y_mh,t) \leftarrow D\cF[v^2]\delta v^2((x_m + h,y_m,z_r)t; (x_m-h,y_m,z_s)).
\end{equation}
These {\em midpoint-offset} coordinates are in common use. To honor the actual non-layered nature of the earth, a vector half-offset $h_x = (x_r-x_s)/2, h_y=(y_r-y_s)/2$ may be used instead.

\section{Stationary Phase}
Analysis of the layered case uses yet another approximation. Write the integral representation \ref{eqn:grt1} in terms of the Fourier transform in time: 
\[
D\cF[v^2]\delta v^2(\bx_r,t;\bx_s) \approx -\frac{1}{2\pi}\int\, d\omega \,\omega^2\hat{w}(\omega)\int \,dz \,\left( \int\,dx\int\,dy \, \right.
\]
\begin{equation}
\label{eqn:grt2}
\left. a(\bx_r,(x,y,z))a(\bx_s,(x,y,z))e^{i\omega(t-\tau(\bx_r,(x,y,z))-\tau(\bx_s,(x,y,z)))}\right)\frac{\delta v^2}{v^2}(z)
\end{equation}

Since $w$ is assumed highly oscillatory on the wavelength scale, its Fourier transform is concentrated a ``large'' frequencies. The principle of stationary
phase \cite[]{GuiSte:79,BleisteinCohenStockwell:01} gives an asymptotic evaluation of integrals such as the $dx$ integral in \ref{eqn:grt2} for large frequency.
To state the general principle: the integral of a rapidly fluctuating function
$g(\mathbf{y})e^{i\omega\psi(\mathbf{y})}$ for large $\omega$ is approximated to every order in $\omega$ by a
sum of terms, one for each stationary phase point $\mathbf{y^*}$ (that is $\nabla\psi(\mathbf{y^*})=0$): 
\begin{equation}
\label{eqn:statphase}
\begin{aligned} 
&\int_{\mathbf{R}^m}d\mathbf{y} g(\mathbf{y})e^{i\omega \psi(\mathbf{y})}\\
&\approx \sum_{\nabla
  \psi(\mathbf{y^*})=0}\left(\frac{2\pi}{\omega}\right)^{\frac{m}{2}}e^{\frac{\pi
    i}{4}\mathrm{sgn}\ \mathrm{Hess} \ \psi(\mathbf{y}^*)}
|\mathrm{det}\ \mathrm{Hess}\psi(\mathbf{y}^*)|^{-\frac{1}{2}}g(\mathbf{y}^*)e^{i\omega \psi(\mathbf{y}^*)}
\end{aligned}
\end{equation}
The main condition for validity of this approximation is that the Hessian determinant not vanish at the stationary points (that is, that the stationary points are {\em nondegenerate}).

\noindent {\bf Note:} The symbol $\mathrm{sgn}$ in \ref{eqn:statphase} means the {\em signature} of the symmetric matrix $\mathrm{Hess}  \psi(\mathbf{y}^*$, that is, the number of positive eigenvalues minus the number of negative eigenvalues.

The stationary phase principle \ref{eqn:statphase} may be applied to approximate the integral in \ref{eqn:grt2} over the horizontal variables $x_1=x,x_2=y$, identifying $m=2$, $\psi(\bx) = t - \tau(\bx,\bx_r) - \tau(\bx,\bx_s)$ and $g(\bx) = -(a(\bx,\bx_r)a(\bx,\bx_s)/2\pi)(\delta v^2(\bx)/v^2(\bx))$. This requires (a) finding the stationary points, and (b) verifying that the Hessian is nonsingular there.

The rotational symmetry implicit in the layered assumption strictly constrains the locatino of stationary points. To see how, note that rotational symmetry implies that $\tau(\bx,\bx_s)$ depends only $|(x-x_s,y-y_s)| = r_s$:
\begin{equation}
\label{eqn:bartaudef}
\tau(\bx,\bx_s) = \bar{\tau}(r_s,z,z_s),
\end{equation}
and similarly $\tau(\bx,\bx_r) = \bar{\tau}(r_r,z,z_r)$. Note that $\bar{\tau}$ also solves the eikonal equation, in 2D:
\begin{equation}
\label{eqn:eikr}
\left(\frac{\partial \bar{\tau}}{\partial r}\right)^2(r,z,z_s) +\left(\frac{\partial \bar{\tau}}{\partial z}\right)^2(r,z,z_s) = \frac{1}{v^2(z)}
\end{equation}
The locations of the stationary points of $\psi$, and their non-degeneracy, follows from several properties of $\bar{\tau}$:
\begin{itemize}
\item There exists a decreasing positive function $z_s < z \mapsto r_{\rm max}(z)$ so that a unique solution  $\bar{\tau}$ of \ref{eqn:eikr} exists on $\{(r,z): z_s< z, |r| < r_{\rm max}\}$ satisfying 
\begin{equation}
\label{eqn:initeikr}
\bar{\tau}(r,z,z_s) \sim \frac{\sqrt{(z-z_s)^2+r^2}}{v(z_s)} \mbox{ as } z \rightarrow z_s, \, r \rightarrow 0.
\end{equation}
and similarly for $z_r$;
\item in $\{(r,z): z_s< z, |r| < r_{\rm max}\}$,
\begin{itemize}
\item \begin{equation}
\label{eqn:bartauincr}
r\frac{\partial \bar{\tau}}{\partial r} >0 \mbox{ if }|r|>0
\end{equation}
\item \begin{equation}
\label{eqn:bartauconv}
\frac{\partial^2 \bar{\tau}}{\partial r^2} (r,z,z_s) > 0,
\end{equation}
\end{itemize}
and similarly for $z_r$.
\end{itemize}
Granted these results, note that 
\[
\nabla_{x,y} \psi(\bx) = -\nabla_{x,y}\tau(\bx,\bx_r) - \nabla_{x,y}\tau(\bx,\bx_s)
\]
\begin{equation}
\label{eqn:statcondr}
=-\frac{(x-x_r,y-y_r)}{|(x-x_r,y-y_r)|}\frac{\partial \bar{\tau}}{\partial r}(r_r,z,z_r)
-\frac{(x-x_s,y-y_s)}{|(x-x_s,y-y_s)|}\frac{\partial \bar{\tau}}{\partial r}(r_s,z,z_s)
\end{equation}
For $\bx$ to be a stationary point of $\psi$, it follows from \ref{eqn:statcondr} that the two unit vectors appearing there must be colinear, that is, $(x,y)$ must be on the line joining $(x_r,y_r)$ and $(x_s,y_s)$. Let 
\[
r_{sr} = |(x_r-x_s,y_r-y_s)|, \,\, (e_x,e_y) = \frac{(x_r-x_s,y_r-y_s)}{r_{sr}}
\]
Then for a suitable $r$,
\[
(x-x_s,y-y_s) = r(e_x,e_y), \,\,(x-x_r, y-y_r) = (r-r_{sr})(e_x,e_y).
\]
 and the stationary phase condition \ref{eqn:statcondr} hold if and  only if
\[
{\rm sgn}(r)\frac{\partial \bar{\tau}}{\partial r}(r_{sr}-r_s,z,z_r) + {\rm sgn}(r-r_{sr})\frac{\partial \bar{\tau}}{\partial r}(r_s,z,z_s) = 0
\] 
Either the two signs are the same, or they are different. In fact, it follows from the condition \ref{eqn:bartauincr} that they cannot be the same (exercise!). Thus stationary phase \ref{eqn:statcondr} implies that 
\begin{equation}
\label{eqn:midptcond}
\frac{\partial \bar{\tau}}{\partial r}(r_{sr}-r,z,z_r) = \frac{\partial \bar{\tau}}{\partial r}(r,z,z_s).
\end{equation}
The meaning of condition \ref{eqn:midptcond} is easiest to understand when $z_s=z_r$: in that case, the two sides of \ref{eqn:midptcond} are the same function with reversed argument. Since this function of $r$ is increasing (result \ref{eqn:bartauincr}), the unique solution is $r=r_{sr}/2$, from which it follows that 
\begin{equation}
\label{eqn:midpt}
x = x_m = \frac{x_r-x_s}{2}, \, y=y_m=\frac{y_r-y_s}{2}.
\end{equation}
That is, the stationary point $\bx$ lies beneath the {\em midpoint} $(x_m,y_m)$ of the line segment between source and receiver. The horizontal distance from either is $r=r_{rs}/2 = h$, the half-offset, so write from now on $h$ instead of $r$.

It is not hard to see that the for typical $z_r,z_s$ in towed streamer marine surveys - both a few meters - and typical target depth $z$ - kilometers - the error caused by assuming $z_s=z_r$ in this calculation is negligible. Therefore \ref{eqn:midpt} is assumed to hold in the following, and I will also assume $z_s=z_r$ to simplify various expressions.

Having found the only stationary points, the other issue that must be settled in order to apply the stationary phase approximation \ref{eqn:statphase} is the nondegeneracy of the phase Hessian. From the definition \ref{eqn:bartaudef} of $\bar{tau}$ and its eikonal \ref{eqn:eikr} it follows that at the midpoint \ref{eqn:midpt}, 
\[
\left(
\begin{array}{cc}
\frac{\partial^2 \psi}{\partial x^2} & \frac{\partial^2 \psi}{\partial x \partial y}\\
\frac{\partial^2 \psi}{\partial x \partial y} & \frac{\partial^2 \psi}{\partial y^2} 
\end{array}
\right) (\bx^*)
=
\]
\begin{equation}
\label{eqn:hess}
-\left(
\begin{array}{cc} 
e_y^2\frac{2}{r}\frac{\partial \bar{\tau}}{\partial r} + e_x^2\frac{\partial^2 \bar{\tau}}{\partial r^2} & 
-e_xe_y\left( \frac{2}{r}\frac{\partial \bar{\tau}}{\partial r} + \frac{\partial^2 \bar{\tau}}{\partial r^2}\right) \\
-e_xe_y\left( \frac{2}{r}\frac{\partial \bar{\tau}}{\partial r} + \frac{\partial^2 \bar{\tau}}{\partial r^2} \right) &
e_x^2\frac{2}{r}\frac{\partial \bar{\tau}}{\partial r} + e_y^2\frac{\partial^2 \bar{\tau}}{\partial r^2}
\end{array}
\right) (h,z)
\end{equation}
In view of the inequalities \ref{eqn:bartauincr} and \ref{eqn:bartauconv}, the determinant of this matrix is nonzero (in fact, positive). Thus the stationary point of $\psi$ (for each $z,t$) is nondegenerate.

The identity \ref{eqn:hess} implies that the Hessian has two negative eigenvalues. The easiest way to see this is to check the case in which the offset vector is parallel to the $x$-axis, so that $|e_x|=1, e_y=0$. Then the Hessian is diagonal, and the diagonal entries are exactly twice the negatives of the quantities estimated in \ref{eqn:bartauincr} and \ref{eqn:bartauconv}. Since the Hessian for other orientations of the offset is related to the $x$-axis case by rotational similarity, the eigenvalues are always these quantities. Conclude that
\begin{equation}
\label{eqn:signat}
{\rm sgn} {\rm Hess}\psi(\bx^*) = -2 \Rightarrow e^{\frac{i \pi}{4} {\rm sgn} {\rm Hess}\psi(\bx^*)} = -i
\end{equation}

Since the stationary point is below the midpoint, rotational symmetry implies that this determinant is a function of $h$ and $z$: abuse notation by writing
\begin{equation}
\label{eqn:hessdef}
{\rm Hess}(h,z) = \left(
\begin{array}{cc}
\frac{\partial^2 \psi}{\partial x^2} & \frac{\partial^2 \psi}{\partial x \partial y}\\
\frac{\partial^2 \psi}{\partial x \partial y} & \frac{\partial^2 \psi}{\partial y^2} 
\end{array}
\right) (\bx^*)
\end{equation}
where $h=r_{rs}/2 = |(x_r-x_s,y_r-y_s)|/2$ and $\bx^*=(x_m,y_m,z)$.

The preceding paragraphs have supplied all of the ingredients necessary to apply the stationary phase approximation \ref{eqn:statphase} to the frequency-domain expression of the Generalized Radon Transform approximation \ref{eqn:grt2}. As the reader will verify, the result is 
\begin{equation}
\label{eqn:grt2sp}
D\cF[v^2]\delta v^2(x_m,y_m,h,t) \approx 
\approx \int\,d\omega\,i\omega \hat{w}(\omega)\int\,dz\,e^{i\omega(t-2\bar{\tau}(h,z,z_s))} a(\bx_s,\bx)a(\bx_r,\bx) |Hess(h,z)|^{-1/2}\delta v^2(z)
\end{equation}
(including a minus sign from the signature of the Hessian, which is 2).

Rotational symmetry also applies to the transport equation \ref{eqn:transp}, and shows that $a(\bx_s,\bx_m) = a(\bx_r,\bx_m)=\bar{a}(h,z)$ (defining $\bar{a}$). Insert this identity and use the Fourier representation of the delta function to arrive at 
\begin{equation}
\label{eqn:convz}
D\cF[v^2]\delta v^2(x_m,y_m,h,t) \approx w *_t \frac{\partial}{\partial t}\int\,dz\,
\delta(t-2\bar{\tau}(h,z,z_s)) \bar{a}^2(h,z)|\det {\rm Hess}(h,z)|^{-1/2}\frac{\delta v^2}{v^2}(z)
\end{equation}
This approximation is very close to the desired model - the remaining steps will be taken in the next section.

The key inequalities \ref{eqn:bartauincr} and \ref{eqn:bartauconv} still require justification. The key observation in this argument is that $\bar{\tau}$ can also be constructed via rays, as shown in Chapter \ref{ch:ray}. The assocated Hamilton's equations \ref{eqn:ham} for the phase space vector $(r,z,p,q)$ read
\begin{eqnarray}
\label{eqn:hamr}
\frac{dr}{dt} & = & v^2(z)p\\
\frac{dz}{dt} & = & v^2(z)q\\
\frac{dp}{dt} & = & 0\\
\frac{dq}{dt} & = & \frac{d \log v}{dz}(z)
\end{eqnarray}
In particular, $p$ is constant along rays. Divding the first equation by the second and using the eikonal equation $v^2(z) (p^2 + q^2) = 1$, obtain 
\begin{equation}
\label{eqn:drdz}
\frac{dr}{dz} = \frac{v(z)p}{\sqrt{1-v(z)^2 p^2}}
\end{equation} 
which gives the evolution of $r$ as a function of $z$ along the ray so long as $q > 0$, that is, $v(z)|p|<1$. The latter condition characterizes rays for which $dz/dt > 0$, implying also that along such rays, $\partial\tau/\partial z > 0$ .Rays violate this condition by turning horizontal. It is not too hard to see that a horizontal ray velocity vector implies the presence of a caustic, violating the conditions that ensure that the approximations \ref{eqn:grt1} and \ref{eqn:grt2} are accurate. Therefore, we assume that over the depth range of interest, and over the rays that are important in this construction, $v(z)|p| < 1$. This assumption will be justified ex post facto.

Integrating \ref{eqn:drdz}, obtain
\begin{equation}
\label{eqn:intrz}
r(z,p)=\int_{z_s}^z \,dz' \, v(z')p (1-v(z')^2 p^2)^{-\frac{1}{2}}
\end{equation}
Then 
\begin{equation}
\label{eqn:drdp}
\frac{\partial r}{\partial p}(z,p) = \int_{z_s}^z \,dz' \, v(z') (1-v(z')^2 p^2)^{-\frac{3}{2}} >0
\end{equation}
under the assumptions made above. That is, define $p_{\rm max}(z) = \min \{1/v(z'): z_s \le z' \le z\}$. Then the function
$p \mapsto r(z,p)$ is strictly increasing on $(-p_{\rm max}(z),p_{\rm max}(z))$ hence invertible: write $p(r,z)$ for the inverse, which is well-defined on $(-r_{\rm max}(z),r_{\rm max}(z))$ with $r_{\rm max}(z) = r(p_{\rm max}(z),z)$.

Of course, in equation \ref{eqn:drdz} and following, $p$ is the $x$-component of the slowness vector $\bP$ in Hamilton's equations (that is, the system \ref{eqn:hamr}). So from \ref{eqn:pdef} and the strict increase of $r \mapsto p(r,z)$ follows inequality \ref{eqn:bartauconv}. Symmetry implies that 
\[
\frac{\partial \bar{\tau}}{\partial r}(0,z) = 0;
\]
together with the convexity result, that's enough to establish inequality \ref{eqn:bartauincr}.

\section{Hyperbolic Normal Moveout}
Transform the integral in equation \ref{eqn:convz} by the change of variable $z \rightarrow T(h,z,z_s) = 2\bar{\tau}(h,z,z_s)$. $T$ is the {\em two-way time} from $(0,z_s)$ to $(h,z)$. Because of the standing hypothesis that $dz/dt>0$ along all rays from source or receiver to $(h,z)$  in the region of defined by the inequalities $h<r_{\rm max}(z,p)$, this change of variables is invertible, and defines depth as a function of two-way time: $Z(h,T(h,z,z_s),z_s) = z$. Thus equation \ref{eqn:convz} can be re-written as 
\[
D\cF[v^2]\delta v^2(x_m,y_m,h,t) \approx 
\]
\[
w *_t \frac{\partial}{\partial t}\int\,dT\,
\approx \int\,dz\,\delta(t-T) \left(\frac{\partial Z}{\partial T}\bar{a}^2|\det {\rm Hess}|^{-1/2}\right)(h,Z(h,T,,z_s)\frac{\delta v^2}{v^2}(Z(h,T,z_s))
\]
\begin{equation}
\label{eqn:convt1}
= w*_t\frac{d}{dt}A(h,t,z_s) R(h,t,z_s)
\end{equation}
in which
\[
A(h,t,z_s) = \left(\frac{\partial Z}{\partial t}\bar{a}^2|\det {\rm Hess}|^{-1/2}\right)(h,Z(h,t,,z_s),\,\, R(h,t,z_s)  = \frac{\delta v^2}{v^2}(Z(h,t,z_s))
\]

Now rely on the separation of scales assumed to justify the use of geometric optics. Since $v^2$ is much smoother than $\delta v^2$, which is oscillatory on the scale of a wavelength, derivatives of $A$ (depending on traveltime and amplitude, which are as smooth as $v^2$ or smoother) are much smaller than corresponding derivatives of $R$. Make another approximation of the same quality as those that have been used to arrive at \ref{eqn:convt1} to arrive at
\begin{equation}
\label{eqn:convt}
D\cF[v^2]\delta v^2(x_m,y_m,h,t) \approx  w*_t A(h,t,z_s) \frac{dR}{dt}(h,t,z_s)
\end{equation}

The hyperbolic normal moveout approximation takes two further steps:
\begin{itemize}
\item replacing the depth coordinate $z$ by the zero-offset two-way time
$T_0(z,z_s) = T(0,z,z_s)$ with inverse $Z_0(t_0,z_s)$ (so that $Z_0(T_0(z,z_s),z_s) = z$), and 
\item replacing the non-zero offset time $T(h,z,z_s)$ with the lowest order nontrivial Taylor series in $h$, which may be expressed as
\begin{equation}
\label{eqn:hnmo}
\bar{T}(h,t_0) = \sqrt{t_0^2 + \frac{h^2}{v_{\rm RMS}^2(t_0)}} = T(h,Z_0(t_0,z_s),z_s)+ O(h^4).
\end{equation}
\end{itemize}
To prove \ref{eqn:hnmo}, begin by noting that
\begin{equation}
\label{eqn:eik2w}
\left(\frac{\partial T}{\partial z}\right)^2+
\left(\frac{\partial T}{\partial h}\right)^2 =\frac{4}{v^2}
\end{equation}
Differentiate this twice with respect to $h$ and use the vanishing of
odd-order $h$ derivatives at $h=0$ (implied by symmetry) to conclude
that 
\begin{equation}
\label{eqn:eiktwh0}
\partial T(0,z,z_s)/\partial z = \frac{2}{v(z)}
\end{equation}
so the second $h$ derivative
\[
q(z)=\frac{\partial^2 T}{\partial h^2}(0,z)
\]
satisfies
\[
\frac{2}{v}\frac{dq}{dz}+q^2=0
\]
Introduce temporarily a new depth coordinate
\[
\sigma(z)=\frac{1}{2}\int_{z_s}^z \, v
\]
Then in terms of $\sigma$, $q$ satisfies the Ricatti equation
\[
\frac{dq}{d\sigma}+q^2=0
\]
The solution which is singular at $\sigma=0$, i.e. $z=z_s$, is
\[
q(\sigma)=\frac{1}{\sigma}=\frac{2}{\int_{z_s}^z \,dz\,v}
\]
From equation \ref{eqn:eiktwh0}, you can also write this as
\[
\frac{\partial ^2\bar{T}}{\partial h^2}(0,t_0) = \frac{\partial^2 T}{\partial h^2}(0,Z_0(t_0,z_s),z_s)  q(\sigma(Z_0(t_0,z_s))=\frac{1}{\int_0^{t_0}\,dt_0\,\bar{v}^2}
\]
where $\bar{v}(t_0) = v(Z_0(t_0,z_s))$.

\noindent {\bf Note:} In the seismic literature, $v(z)$ is called the interval velocity as a function of $z$, $\bar{v}(t_0)$ is called the interval velocity as a function of time - see self-doc for the SU utility {\tt velconv}).

Thus 
\[
\bar{T}(h,t_0)=t_0 + \frac{h^2}{2}\frac{\partial^2 \bar{T}}{\partial h^2}(h,t_0) +...
\]
\[
= t_0 + \frac{h^2}{2\int_0^{t_0}\,dt_0\,\bar{v}^2}+...
\]
Since
\[
\frac{\partial^2}{\partial h^2}(\bar{T}(h,t_0))^2_{h=0}=2 t_0\frac{\partial^2 \bar{T}}
{\partial h^2}(0,t_0)
\]
the above can be rewritten as
\[
\bar{T}(h,t_0)^2 = t_0^2 + \frac{h^2}{\frac{1}{t_0}\int_0^{t_0}\,dt_0\,\bar{v}^2}+...
\]
\begin{equation}
\label{eqn:hnmo2}
=t_0^2+\frac{h^2}{v_{\rm RMS}^2(t_0)} + ...
\end{equation}
where 
\begin{equation}
\label{eqn:vrmsdef}
v_{\rm RMS}(t_0) = \sqrt{\frac{1}{t_0}\int_0^{t_0}\,dt_0\,\bar{v}^2}
\end{equation}
is the Root-Mean Square (RMS) velocity as function of $t_0$. Evidently equation \ref{eqn:hnmo2} is equivalent to \ref{eqn:hnmo}.

\section{NMO Correction}
The NMO correction is a change of variable in the data, namely 
\begin{equation}
\label{eqn:nmo}
d(x_m,y_m,h,t) \rightarrow d(x_m,y_m,h,\bar{T}(h,t_0)).
\end{equation}

This operation has a simple interpretation under the assumption that an effective source signature deconvolution has been performed in the convolutional model \ref{eqn:convt}, that is, that an inverse to the convolution with the source wavelet $w$ has been performed. This would remove $w*_t$ from the right-hand side of \ref{eqn:convt}. Actually exact source signature deconvolution is impossible, as realistic source wavelets are band-limited. Therefore there must be a residual source signature built into the data even after any sort of deconvolution is applied, consisting (at best!!) of a bandpass filter. I will comment further on this matter at the end of the chapter, but for now presume that deconvolution has removed $w*_t$ from the right-hand side of \ref{eqn:convt}, to yield the {\em deconvolved convolutional model} of the NMO-corrected data:
\begin{equation}
\label{eqn:convtdecon}
d(x_m,y_m,h,\bar{T}(h,t_0))  \approx  A(h,t,z_s) \frac{dR}{dt}(h,t,z_s)
\end{equation}
Recall that
\[
R(h,t,z_s) = \frac{\delta v^2}{v^2}(Z(h,t,z_s)) ,
\]
so 
\[
R(h,\bar{T}(h,t_0),z_s) = \frac{\delta v^2}{v^2}(Z(h,\bar{T}(h,t_0),z_s))
=\frac{\delta v^2}{v^2}(Z(h,T(h,Z_0(,t_0,z_s),z_s))) = \frac{\delta v^2}{v^2}(Z_0(t_0,z_s))
\]
since $T$ and $Z$ are an inverse pair. Note that $R(h,\bar{T}(h,t_0),z_s)$ is {\em independent of $t$}. Define \begin{equation}
\label{eqn:refldef}
\bar{R}(t_0) = \frac{\delta v^2}{v^2}(Z_0(t_0,z_s)),
\end{equation}
\begin{equation}
\label{eqn:ampdef}
 bar{A}(h,t_0) = A(n,\bar{T}(h,t_0),z_s)\left(\frac{\partial \bar{T}}{\partial t_0}(h,t_0)\right)^{-1}
\end{equation}
In terms of of $\bar{R}$ and $\bar{A}$, the deconvolved convolutional model \ref{eqn:convtdecon} of a CMP gather is
fantastically simple:
\begin{equation}
\label{eqn:convt0}
d(x_m,y_m,h,\bar{T}(h,t_0)) \approx \bar{A}(h,t_0) \frac{d\bar{R}}{dt_0}(t_0) 
\end{equation}
The second factor is the {\em reflectivity} - simply the derivative of the relative perturbation in $v^2$, parametrized by vertical two-way time $t_0$. The first is a combination of various amplitude effects, including the geometric amplitudes from source/receiver at offset $h$, the reciprolcal square root of the phase Hessian (measuring wavefront curvature), and the rate of change of two-way time at offset $h$ with respect to two-way time at offset zero.

In the case of constant $v$, $\bar{A}$ can certainly be computed explicitly. However, its most important feature is that it is {\em smooth}, that is slowly varying on the wavelength scale, as it combines various geometric optics quantities that are smooth under the standing assumptions on $v$.  The reflectivity on the other hand is not, so a grey-scale plot of NMO-corrected will have {\em flat wavefronts} - rapidly oscillating vertically, slowly varying horizontally. That is what you see in Figures \ref{fig:nmo200}, \ref{fig:nmo700}, \ref{fig:nmo1300}, and \ref{fig:nmo2000}. This feature is an indication that the data $d$ and the velocity $v$ are compatible, that is, that $v$ adequately explains the kinematic features of the data.

On the other hand, if the NMO-corrected CMP gathers are not flat, then the velocity used to compute the NMO correction is not compatible with the data. Figure \ref{fig:nmo2000lo} showed an example of this.

Careful inspection of the relation \ref{eqn:convt0} reveals the reason. To clarify the role of velocity, include it as an argument of $\bar{T}$: that is, $\bar{T}[v](t_0,h)$ 
\begin{equation}
\label{eqn:hnmov}
\bar{T}[v](h,t_0) = \sqrt{t_0^2 + \frac{h^2}{v_{\rm RMS}^2(t_0)}} 
\end{equation}
Denote by $\bar{T}_0[v](h,t)$ its inverse function, that is, 
\[
\bar{T}[v](h,\bar{T}_0[v](h,t)) = t
\]
Suppose that the data are compatible with $v$, that is that relation \ref{eqn:convt0} holds. Choose another velocity $v'$, and construct the NMO-corrected CMP for this velocity: 
\begin{equation}
\label{eqn:convt0}
d(x_m,y_m,h,\bar{T}[v'](h,t_0)) \approx \bar{A}(h,\bar{T}_0[v](h,\bar{T}[v'](h,t_0))) \frac{d\bar{R}}{dt_0}(\bar{T}_0[v](h,\bar{T}[v'](h,t_0)))
\end{equation}
The surfaces of equal phase for the last factor - these delineate the visible wavefronts in the NMO corrected gather, since the other factor is smooth - are the curves defined by 
\[
h \mapsto \bar{T}_0[v](h,\bar{T}[v'](h,t_0)).
\]
To describe these curves conveniently, introduce the {\em two way time discrepancy} between $v$ and $v'$, denoted $\Delta T[v,v'](h,t_0)$, by
\begin{equation}
\label{eqn:tdiscrep}
\sqrt{t_0^2 + \Delta T[v,v'](h,t_0)^2} = \bar{T}_0[v](h,\bar{T}[v'](h,t_0))
\end{equation}
Since $\bar{T}$ and $\bar{T}_0$ are an inverse pair, evaluating $\bar{T}[v](h,\cdot)$ at the right-hand-side of equation \ref{eqn:tdiscrep} gives
\[
t_0^2 + \Delta T[v,v'](h,t_0)^2 +\frac{h^2}{v_{\rm RMS}^2(\bar{T}_0[v](h,\bar{T}[v'](h,t_0)))} = t_0^2 + \frac{h^2}{(v')_{\rm RMS}^2(t_0)}
\]
so 
\begin{equation}
\label{eqn:tdiscrep1}
\Delta T[v,v'](h,t_0)^2 =  h^2\left(\frac{1}{(v')_{\rm RMS}^2(t_0)}-\frac{1}{v_{\rm RMS}^2(\bar{T}_0[v](h,\bar{T}[v'](h,t_0)))} \right)
\end{equation}

It is worth taking a moment to understand the geometric meaning of equation \ref{eqn:tdiscrep1}. Equal phase surfaces $t=\bar{T}[v](h,t_0)$ are called {\em moveout curves} (or surfaces, in 3D). The inverse function $\bar{T}_0[v](h,t)$ gives the zero-offset two-way time for the moveout curve passing through $(h,t)$, for the velocity $v$. So the {\em residual moveout} $\bar{T}_0[v](h,\bar{T}[v'](h,t_0))$ is the zero-offset intercept of the moveout curve for $v$, passing through the point $(h,\bar{T}[v'](h,t_0))$ on the moveout curve for $v'$. $\Delta T[v,v'](h,t_0)$ is the difference between $t_0^2$ and this two-way time (squared) obtained by starting at $(0,t_0)$ on the moveout curve for $v'$, traveling out to offset $h$, and then back to $h=0$ on the moveout curve for $v$. Residual moveout is a direct measure of the difference in the NMO-corrected phase. Equation \ref{eqn:tdiscrep1} relates residual moveout directly to the values of the two velocities $v$ and $v'$.

Since the residual moveout formula samples the two velocities at two different places, one of those depending on the velocities themselves, it is not entirely obvious exactly what equation \ref{eqn:tdiscrep1} tells us about the relation with the velocity difference. The easy case to understand is when both $v$ and $v'$ are constant. Then $v_{\rm RMS} = v$, $v'_{\rm RMS} = v'$, and  \ref{eqn:tdiscrep1} becomes
\begin{equation}
\label{eqn:tdiscrepcv}
\Delta T[v,v'](h,t_0)^2 =  h^2\left(\frac{1}{(v')^2}-\frac{1}{v^2} \right)
\end{equation}
This tells you that if you NMO-correct with a wrong constant velocity, the gathers will definitely not be flat, in fact residual moveout curvature is proportional to the error between $1/v^2$ and $1/(v')^2$. This in this case at least gather flatness directly measures error in velocity.

The non-constant velocity case is harder, but not impossible, to analyse, and the conclusion is the same: velocity errors large enough to cause significant traveltime errors on the scale of a wavelength show up in non-flat NMO corrected CMP gathers. See \cite[]{Symes:99a} for details.

\section{NMO Stretch}
As mentioned earlier, the ``deconvolved convolutional model'' is necessarily a myth, as ideal source signature deconvolution is impossible. Even granted strict validity of the acoustic constant density Born approximation with the isotropic point radiator source model, the optimal result still bandpass filters the Born impulse response (equation \ref{intrep}). This fact has a striking consequence for NMO correction.

Compose both sides of the deconvolved convolutional model equation \ref{eqn:convt0} with the {\em inverse NMO} function $\bar{T}_0$ to obtain tthe perfectly deconvolved trace in terms of the amplitude $\bar{A}$ and the reflectivity $\bar{R}$, both parametrized by offset $h$ and vertical two-way time $t_0$:
\begin{equation}
\label{eqn:convt}
d_{\rm imp}(x_m,y_m,h,t) \approx \bar{A}(h,t_0) \frac{d\bar{R}}{dt_0}(\bar{T}_0(h,t)).
\end{equation}
I've renamed the left-hand side as $d_{\rm imp}$, for the impulsive (perfectly deconvolved) trace. Then admit that there is indeed a wavelet $w$ in the mix:
\[
d(x_m,y_m,h,t) = w *_t d_{\rm imp}(x_m,y_m,h,t)
\]
\[
\approx \int\,dt'\,w(t-t') \bar{A}(h,\bar{T}_0(h,t')) \frac{d\bar{R}}{dt_0}(\bar{T}_0(h,t'))
\]
So the NMO-corrected trace is NOT the convolution of $w$ with its perfectly-deconvolved cousin:
\begin{equation}
\label{eqn:stretch}
d(x_m,y_m,h,\bar{T}(h,t_0)) \approx \int\,dt_0'\,w(\bar{T}(h,t_0)-\bar{T}(h,t_0')) \frac{\partial \bar{T}}{\partial t_0}(h,t_0')\bar{A}(h,t_0') \frac{d\bar{R}}{dt_0}(h,t_0')
\end{equation}
Note that the argument of $w$ inside the integral is (generally) nonlinear in $t$ and $t'$: this is not a convolution. The ratio
\[
\frac{\bar{T}(h,t_0)-\bar{T}(h,t_0')}{t_0-t_0'}
\]
measures the distortion of convolution: it is roughly the Jacobian factor inside the integral, namely the rate of change of time with respect to zero-offset time. This factor is known as {\em NMO stretch}, in honor of its graphical effect.

To see how this works, start with the idealized convolutional model seismogram depicted in Figure \ref{fig:event}. This is a ``pure event'', that is, $\bar{R}$ is a Heaviside function so that its $t_0$ derivative is a delta function, and the amplitude $\bar{A}$ is set to 1. The SU command {\tt suaddevent} is very convenient for creating such idealized reflections with hyperbolic moveout. The RMS velocity is chosen as 3000 m/s (the default), and the zero-offset intercept (of the moveout curve) as $t_0$ = 0.8 s - these two choices completely determine the moveout curve.  The source wavelet is a [2,5,25,35] Hz trapezoidal bandpass filter.

Figure \ref{fig:nmoevent} shows the NMO correction. Notice that the wavelet seems to spread as offset increases: that's why the non-convolution of equation \ref{eqn:stretch} is called ``NMO stretch''.  If you were to add more events to the data, you would find that the degree of stretch varies with $t_0$.

\multiplot{2}{event,nmoevent}{width=0.45\textwidth}{Left: idealized event created with {\tt suaddevent}, with [2,5,25,35] Hz bandpass wavelet; Right: result of NMO correction}

Note that Viking Graben NMO-corrected CMP gathers (Figures \ref{fig:nmo200}, \ref{fig:nmo700}, \ref{fig:nmo1300}, and \ref{fig:nmo2000}) inherit a {\em mute} from the CMP gathers themselves, and this mute is chosen so that the NMO-corrected events are cut off before the offset becomes large enough to exhibit severe NMO stretch as in Figure \ref{fig:nmoevent}. However careful inspection will reveal that even the Viking Graben events are stretched a bit.

The implication of NMO stretch is that NMO correction does not actually reproduce a material parameter, the reflectivity, even accounting for amplitude. Instead, it produces a nonlinearly filtered version of a material parameter, with the degree of filtering varying with offset and vertical travel time. This feature is a fundamental defect of the convolutional model and NMO correction, with which the community has struggled for decades. See for example \cite{Harlan:14} for a clever use of optimization to choose a {\em time-varying wavelet} so model, and eventually remove, NMO stretch and similar effects, also many references to older work.

\section{Suggested Projects}
\bibliographystyle{seg}  % style file is seg.bst 
\bibliography{../../bib/masterref}

### 2019-10-19 repro/caam641/fd/paper.tex
repro docs dir added

\title{Notes on Convergence of FD Methods}
\maketitle

\section{Energy Estimates}
Abstract setup: 
\begin{itemize} 
\item Hilbert space $H=H_1 \oplus H_2$ (``continuum fields'')
\item Family of Hilbert spaces $G(h) =U(h) \oplus V(h), h>0$ (``grids'')
\item bounded self-adjoint positive definite operators 
\[
A(h): U(h) \rightarrow U(h), B(h): V(h) \rightarrow V(h)
\]
upper and lower bounds uniform in $h$ (``coefficients''):
\[
A_*\|u\|^2 \le \langle Au,u \rangle \le A^*\|u\|^2, \, B_*\|v\|^2 \le
\langle B v, v\rangle \le B^*\|v\|.
\]
\item bounded operators $D(h,r), D_1(h):U(h) \rightarrow V(h)$,
\[
\|D_1(h)\| \le D^* \mbox{ for all }h>0, D(h,r) = rD_1(h)
\]
(``stress divergence'').
\item Bounded operator $S(h):H \rightarrow G(h)$ (``sampling'')
\end{itemize}

Staggered grid dynamics: $(u^n(h),v^{n+1/2}(h)) \in G(h)$ satisfy 
\begin{eqnarray}
\label{eqn:sg}
A(u^{n+1} - u^n) & = & -D^T v^{n+1/2} \nonumber\\
B(v^{n+1/2}-v^{n-1/2}) & = & D u^n
\end{eqnarray}

Energy form:
\[
E = \frac{1}{2}(\langle Au,u\rangle + \langle Bv,v\rangle + \langle D^Tv,u\rangle ) 
\]
\[
= \frac{1}{2} \left(
\begin{array}{c}
u\\
v 
\end{array}
\right)^T
\left(
\begin{array}{cc}
A & \frac{1}{2}D^T\\
\frac{1}{2}D & B
\end{array}
\right)
\left(
\begin{array}{c}
u\\
v 
\end{array}
\right)
\]

\noindent Claim: for a solution $\{u^n,v^{n-1/2}$ of the system \ref{eqn:sg},
\[
E^n = E(u^n,v^{n-1/2})
\]
is independent of $n$.

Proof: 
\[
E^{n+1} - E^n =\langle A(u^{n+1}-u^n), (u^{n+1}+u^n)\rangle 
+ \langle B(v^{n+1/2}-v^{n-1/2}), (v^{n+1/2}+v^{n-1/2})\rangle +
\]
\[
\langle D^Tv^{n+1/2},u^{n+1}\rangle ) - \langle D^Tv^{n-1/2},u^n\rangle )
\]
\[
= \langle -D^Tv^{n+1/2},(u^{n+1}+u^n) \rangle 
+ \langle Du^n. (v^{n+1/2}+v^{n-1/2}) \rangle
+\langle D^Tv^{n+1/2},u^{n+1}\rangle ) - \langle D^Tv^{n-1/2},u^n\rangle )
\]
\[ 
=0.
\]
Q.E.D.

\noindent Claim: for sufficiently small $r$, there exist $0 < C_* \le C^*$ such that 
\begin{equation}
\label{eqn:pd}
C_*\|u,v\|^2 \le E(u,v) \le C^*\|u,v\|^2
\end{equation}

Proof: upper bound is clear, from $h$-uniform bounds on $A$, $B$,
$D$. 
\[
E(u,v) \ge A_*\|u\|^2 + B_*\|v\|^2 - rD^*\|u\|\|v\|
\]
which is a positive definite form in $\|u\|,\|v\|$ when $rD^* <
\sqrt{A^*B^*}$.
Q.E.D.

\section{$L^2$ Convergence of Smooth Solutions}
Inhomogeneous system:

\begin{eqnarray}
\label{eqn:in}
A(u^{n+1} - u^n) & = & -D^T v^{n+1/2} + hf^{n+1/2}\nonumber \\
B(v^{n+1/2}-v^{n-1/2}) & = & D u^n + h g^n
\end{eqnarray}

\noindent Claim: for $r$ sufficiently small that $E$ is positive definite, there
exists $K \ge 0$, $\lambda > 1$ so that so that
\[
E^n = E(u^n,v^{n-1/2}) \le \lambda^nE(u^0,v^{-1/2}) + Kh\sum_{m=0}^{n-1}\lambda^{n-m+1}(\|f^{m-1/2}\|^2 + \|g^m\|^2)
\]
Proof: The same arithmetic as used in the homogeneous case leads to
\[
E^{n+1}-E^n = h\langle f^{n+1/2}, u^{n+1}+u^n\rangle + h\langle g^n,
v^{n+1/2}+v^{n-1/2}\rangle
\]
\[
\le \frac{h}{2\alpha^2} (\|f^{n+1/2}\|^2 + \|g^n\|^2) +
\frac{\alpha^2h}{C_*}(E^{n+1}+E^n)
\]
for any $\alpha \in (0,1)$. Choose $\alpha$ so that $\alpha^2 h < C_*$
(that is, can make fixed choice of $\alpha$ for $h$ small enough), and
set 
\[
\lambda = \left(1+\frac{\alpha^2h}{C_*}\right) \left(1-\frac{\alpha^2h}{C_*}\right)^{-1}
\]
and 
\[
K = \frac{\frac{1}{2\alpha^2}}{ \left(1+\frac{\alpha^2h}{C_*}\right) }
\]
Then
\[
\lambda^{-1}E^{n+1} - E^n = Kh (\|f^{n+1/2}\|^2 + \|g^n\|^2)  
\] 
so 
\[
\lambda^{-(n+1)}E^{n+1} - \lambda^{-n}E^n = \lambda^{-n}Kh (\|f^{n+1/2}\|^2 + \|g^n\|^2) 
\]
whence
\[
\lambda^{-n}E^n = E^0 +
Kh\sum_{m=0}^{n-1}\lambda^{-m+1}(\|f^{m-1/2}\|^2 + \|g^m\|^2)
\]
Q.E.D.

Suppose that $(U^n,V^{n-1/2}) \in H$. Set
\begin{eqnarray}
\label{eqn:tr}
f^{n+1/2}(h)& =& h^{-1}(A(h)(S(h)U^{n+1} - S(h)U^n) + D(h,r)^T S(h)V^{n+1/2}))\\
g^{n}(h)& =& h^{-1}(B(h)(S(h)V^{n+1/2}-S(h)V^{n-1/2}) - D(h,r) S(h)U^n)
\end{eqnarray}

Suppose $(u^n,v^{n-1/2})$ solves \ref{eqn:sg} with 
\[
(u^0, v^{-1/2})=S(h)(U^0,V^{-1/2})
\]

\noindent Claim: Define $K, \lambda$ as above, then
\[
E((u^n,v^{n-1/2})-S(h)(U^n,V^{n-1/2})) \le
Kh\sum_{m=0}^{n-1}\lambda^{n-m+1}(\|f^{m-1/2}\|^2 + \|g^m\|^2)
\]
Proof: follows immediately from previous claim.

\noindent Corollary: Suppose $T>0$ given independent of $h$. If
$\|f^{m-1/2}\|^2 + \|g^m\|^2 \le L^2 h^{2p}$ for $0\le m \le N$, $Nh=T$, then 
\begin{equation}
\label{eqn:ee}
E((u^n,v^{n-1/2})-S(h)(U^n,V^{n-1/2})) \le M^2L^2h^{2p}
\end{equation}
Proof: from the preceding estimate,
\[
E((u^n,v^{n-1/2})-S(h)(U^n,V^{n-1/2})) \le KL^2
h^{2p+1}\frac{\lambda^N-1}{\lambda-1}
\]
Note that if $h$ is sufficiently small, then $\lambda < 1 +
2\frac{\alpha^2h}{C_*}$ so $\lambda^N \le
exp(2\frac{\alpha^2}{C_*}T)$, where $T=Nh$. If $h$ is perhaps smaller
yet, then $\lambda > 1+\frac{\alpha^2h}{2C_*}$. Putting this all
together, there is $M^2$ depending on $T$ and all of the other constants
in the setup, so that if $nh <T$ so that \ref{eqn:ee} holds.
Q. E. D.

We apply these results to 3D acoustics and elasticity. In either case,
$H_1 = L^2(\bR^3,\bR^N)$,  $H_2 = L^2(\bR^3,\bR^3)$. Members of $H_1$
are stresses ($N=6$) or pressures ($N=1$), members of $H_2$ are velocities.

$U(h) = $ square integrable $\bR^N$-valued functions on rectangular
grid of mesh $h$, $V(h) = $ square integrable $\bR^3$-valued functions
on half-cell shifted grids ($V_i$ shifted along $i$th axis) (discrete velocities)

$A$ = smooth symmetric $N \times N$ matrix-valued function on $\bR^3$,
uniformly bounded with univormly bounded inverse

$B$ = smooth symmetric $3 \times 3$ matrix-valued function on $\bR^3$,
uniformly bounded with univormly bounded inverse

View $(U,V)$ as $\bR^{N+3}$ valued, then 
\[
S(h)(U,V)_{i,j,k} = \left(
\begin{array}{c}
U(ih,jh,kh)\\
V_1((i+1/2)h,jh,kh)\\
V_2((ih,(j+1/2)h,kh)\\
V_3((ih,jh,(k+1/2)h)
\end{array}\right)
\]

Corresponding discretizations of $A$ and $B$: in the acoustic case, $A$ is scalar,
$(A(h) u)_{i,j,k} = \frac{1}{\kappa(ih,jh,kh)}u_{i,j,k}$. For
elasticity, it is a more complex evaluation of the Hooke tensor at
various grid points, which I will not give explicitly. In all cases,
$B(h)$ is a diagonal sampling of the density: 
\begin{eqnarray*}
(B(h)_{i,j,k})_{1,1} & = & \rho((i+1/2)h,jh,kh),\\
(B(h)_{i,j,k})_{2,2} & = & \rho(ih,(j+1/2)h,kh),\\
(B(h)_{i,j,k})_{3,3} & = & \rho(ih,jh,(k+1/2)h),\\
(B(h)_{i,j,k})_{l,m} & = & 0,\,l \ne m
\end{eqnarray*}

$D$ is a finite difference approximation to the strain (or gradient,
for acoustics) of order $2p$. For example, for $p=1$, one possibility is 
\[
D u =
\{(u_{i+1,j,k}-u_{i,j,k},u_{i,j+1,k}-u_{i,j,k},u_{i,j,k+1}-u_{i,j,k}\}
\]
This discrete gradient operator is a second-order approximation to the
partial derivatives of $u$ at half-cell positions along each
coordinate axis, corresponding to velocity sample points as described
above.

$D^T$ is discrete divergence, $r = \Delta t/h$.

Standard truncation error calculation: if $(U(t), V(t))$ is smooth solution of
acoustic (pressure-velocity) system with smooth right-had sides $F(t),G(t)$, 
\begin{eqnarray}
\label{vdawe.fd}
\frac{1}{\kappa}\frac{\partial U}{\partial t} + \nabla \cdot V & = &
                                                                     F,\\
\rho \frac{\partial V}{\partial t} + \nabla U & = & G
\end{eqnarray}
each time slice of compact
support, $(U^n(h),V^{n-1/2}(h)) = S(h)(U(n\Delta t),V((n-1/2)\Delta t))$, then
\begin{eqnarray}
\label{eqn:tr2}
f^{n+1/2}(h)& =& h^{-1}(A(h)(U^{n+1}(h) - S(h)U^n(h)) + D(h,r)^T
                 V^{n+1/2}(h))) = O(h^{2p})\\
g^{n}(h)& =& h^{-1}(B(h)(V^{n+1/2}(h)-V^{n-1/2}(h)) - D(h,r)
             U^n(h)) = O(h^{2p}+\Delta t^2)
\end{eqnarray}
pointwise, uniformly in $i,j,k$. Since the support is uniformly bounded if
$0\le n\Delta t = nh/r \le T$, the $O$ statements apply  to the $L^2$
norms as well.

Let $u$ denote the solution of \ref{eqn:sg} with
$(u^0,v^{-1/2})=S(h)(U^0,V^{-1/2})$. Then $(u,v)-S(h)(U,V)$ solves
\ref{eqn:in} with $f,g$ as just given.  Therefore the error estimate
\ref{eqn:ee} implies that 
\[
\|(u,v)-S(h)(U,V)\| = O(h^{2p} + \Delta t^2),
\]
that is, $(2,2p)$-th order convergence in the $L^2$ sense.

\section{Weak Convergence of Singular Solutions}

### 2019-10-19 repro/caam641/intro.tex
repro docs dir added

\title{Introduction}

\section{Goals}
This book will explore some of the core concepts and methods of industrial seismic imaging and inversion from a mathematical point of view. The content arose from my efforts (and those of many others) to explain {\em precisely} how many clever and effective processing algorithms originate in the physics of waves. Evidently, either such explanation is possible, or our understanding of seismic waves is lacking in some important respect. To some extent this project has been successful: wave theory - linear elastodynamics -  actually does explain {\em precisely} why some important seismic imaging and inversion methods work. Seismic processing is a huge field, and the account given here is not comprehensive by any means. However, in the pages to follow you will find a few complete accounts, starting in basic linear elastodynamics and arriving at some practical and important imaging and inversion algorithms.

These notes owe a great deal, in structure, content, and spirit, to Jon Claerbout. In preparing the text, I re-read the introduction to Jon's book {\em Image Estimation by Example}. In Jon's terms, you might say that this course is about ``the theory of the similarity between theory and practice'' (an emphasis that would clearly not appeal to Jon!). In particular, it is about both theory and practice, with practice represented by computational examples illustrating the theory, using open source software packages and public domain data, both synthetic and field. These examples take the form of {\em reproducible computational research}, another pioneering concept of Claerbout's. The reader can rebuild the results from raw data and code by executing a few simple commands. Most importantly, the reader can also modify the examples to explore beyond the content of the text.

\section{Syllabus}
This book's current incarnation is a set of notes for a course, CAAM 641, Spring 2017 edition. The structure of this course it a bit unusual, and requires some explanation. It will divide into three segments, each consisting of several in-person lectures and work sessions, followed by an internet-based project period. Each segment consists of two parts, one oriented towards data processing, the other concerned with mathematical foundations. The two parts actually support each other but develop to some extent independently, with the final story emerging over the course. These notes mostly support the processing part. The mathematical part will read through some recent (and not-so-recent) papers, or notes in those cases where satisfactory references are not available.

THe three segments are:

\begin{itemize}
\item {\bf Basic Imaging} (Chapters 2 - 6) - starting with an imaging exercise, using 2D marine survey data: RMS velocity estimation, NMO correction of CMP gathers, stack, and post-stack time and depth migration. The object of this part of the course is to identify the origin in wave theory of each process used in this exercise. Two approximations to solution of the wave equation are central to this story: linearization, aka single scattering, aka the Born approximation, and geometric optics, aka high frequency asymptotics, aka ray theory. Applied to ``locally layered'' modeling, these lead to the convolutional model, basic time processing of seismic reflection surveys (deconvolution, statics, NMO, stack), and NMO based velocity analysis. Applied to models with mild lateral variation, it leads to poststack imaging. NMO based velocity analysis provides a starting point for prestack imaging (the topic of the second part of the course) and migration velocity analysis, and serves as an interesting and accessible test bed for inversion algorithms.

The mathematical focus of this segment is the basic theory of the wave equation, following the lead of J.-L. Lions. We will work through existence of finite energy solutions, regularity as a function of the right hand side and of the coefficients, and the examples of acoustics and elasticity.

At the end we will have an understanding at various levels of the Born approximation (when it is justified!), and of its collaboration with high frequency asymptotics to produce (some) post-stack imaging algorithms.

\item {\bf Linear Inversion:} prestack depth imaging via diffraction sums (Kirchhoff migration), Generalized Radon Transform inversion for acoustics and elasticity, Reverse Time Migration, Least Squares Migration (iterative and true amplitude migration). This time we start with the theory, as the computational examples will use Reverse Time and True Amplitude migration, but the justification for these is essentially the theory of the Generalized Radon Transform. We end by applying these techniques to the 2D marine survey of segment 1.

The mathematical part of this segment has two goals: first, a rapid survey of the convergence theory of finite difference methods, used throughout the course (and seismology) to generate examples and process data; second, filling in some of the mathematical holes in high-frequency asymptotics as developed in the other part of the course, in particular extension of the asymptotic description beyond caustics.

This section will complete the analysis of asymptotics and its role in post-stack imaging, and extend that understanding to pre-stacck imaging.

\item {\bf Nonlinear Inversion:} Theory of reflection and transmission tomography, relation to migration velocity analysis in extended time and image domains. Inversion beyond single scattering - full waveform inversion.  Application to the marine survey of segment 1.

Mathematical topics: extended modeling and inversion, source and medium extensions, analysis of convergence for inversion via iterative solution of inner problems. 
\end{itemize}

Examples presented in course materials use Seismic Unix, Madagascar, and TRIP software packages to realize the concepts introduced in class. The course notes take the form of a directory tree of Madagascar papers, each one with any examples and resulting figures produced from Python scripts in a project subdirectory. The participants can rebuild every one of these papers from scratch, and the scripts can (and should!) serve as departure points for additional projects.

I suggest questions at the end of each chapter that can serve as nuclei of projects, and welcome suggestions for others.

\section{Schedule}
The class schedule divides into three periods, each with two weeks of class meetings and three weeks of internet-based project work. Classes will meet on Tuesdays and Thursdays from 1600 to 1830 in Duncan Hall room 3076 on the Rice campus. A tentative schedule for class meetings is: January 10, 12, 17, and 19; February 14, 16, 21, and 23; and March 21, 23, 28, and 30. I will communicate with participants during the intervening project periods by email and, when appropriate, by skype.

\section{Expectations}
In view of the advanced graduate nature of this course, no general performance metrics are appropriate: each participant will determine their level of involvement and expectations for the course.

### 2019-10-19 repro/caam641/intro/paper.tex
repro docs dir added

\title{0.0: Introduction}
\author{William W. Symes}
\address{The Rice Inversion Project,
Rice University,
Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}

\maketitle
\parskip 12pt

\lefthead{Symes}
\righthead{CAAM 641 Section 0.0}

\begin{abstract}
For Spring semester 2017, CAAM 641 will be devoted to the foundations and algorithms of seismic imaging and inversion.
\end{abstract}

\section{Goals}
This course will explore some of the core concepts and methods of seismic imaging and inversion from a mathematical point of view. The course will

\begin{itemize}
\item emphasize the close correspondence between mathematical foundations and practical algorithms, while allowing the participants to concentrate their attention on one or the other (or both) of these aspects, and
\item offer structured opportunities to exercise and explore these concepts computationally, using open source software packages and public domain data, both synthetic and field.
\end{itemize}

\section{Syllabus}
This course will divide into three segments, each consisting of several in-person lectures and work sessions, followed by internet-based project periods. Each segment consists of two parts, one oriented towards data processing, the other concerned with mathematical foundations. The two parts have points of contact but are not necessarily parallel. I will provide notes as needed, largely for the processing part. The mathematical part will read through some recent (and not-so-recent) papers, or notes in those cases where satisfactory references are not available.

\begin{itemize}
\item {\bf Basic Imaging:} We will begin with basic processing exercise, using 2D marine survey data: NMO correction of CMP gathers, stack, and post-stack time and depth migration. The object of this part of the course is to identify the origin in wave theory of each process used in this exercise. Two approximations to solution of the wave equation are central to this project: linearization, aka single scattering, aka the Born approximation, and geometric optics, aka high frequency asymptotics, aka ray theory. Applied to ``locally layered'' modeling, these lead to the convolutional model, basic time processing of seismic reflection surveys (deconvolution, statics, NMO, stack), and NMO based velocity analysis. Applied to models with mild lateral variation, it leads to poststack imaging. NMO based velocity analysis provides a starting point for prestack imaging (the topic of the second part of the course) and migration velocity analysis, and serves as an interesting and accessible test bed for inversion algorithms.

The mathematical focus of this segment is the basic theory of the wave equation, following the lead of J.-L. Lions. The main reference is \cite{BlazekStolkSymes:13}, which is made available on the web site. We will work through existence of finite energy solutions, regularity as a function of the right hand side and of the coefficients, and the examples of acoustics and elasticity.

\item {\bf Linear Inversion:} prestack depth imaging via diffraction sums (Kirchhoff migration), Generalized Radon Transform inversion for acoustics and elasticity, Reverse Time Migration, Least Squares Migration (iterative and true amplitude migration). This time we start with the theory, as the main examples will use Reverse Time and True Amplitude migration, but the justification for these is essentially the theory of the Generalized Radon Transform. We end by applying these techniques to the 2D marine survey of segment 1.

The mathematical part of this segment has two goals: first, a rapid survey of the convergence theory of finite difference methods, used throughout the course (and seismology) to generate examples and process data; second, to extend high-frequency asymptotics beyond caustics.

\item {\bf Nonlinear Inversion:} Theory of reflection and transmission tomography, relation to migration velocity analysis in extended time and image domains. Inversion beyond single scattering - full waveform inversion.  Application to the marine survey of segment 1.

Mathematical topics: extended modeling and inversion, source and medium extensions, analysis of convergence for inversion via iterative solution of inner problems. 
\end{itemize}

Examples presented in course materials will use Seismic Unix, Madagascar, and TRIP software packages to realize the concepts introduced in class. In fact, the course notes are organized in the form of directory tree of Madagascar papers, each one with any examples and resulting figures produced from Python scripts in a project subdirectory. The participants can rebuild every one of these papers from scratch, and the scripts should serve as departure points for additional projects.

I suggest questions in each paper that can serve as nuclei of projects, and welcome suggestions for others.

\section{Schedule}
The class schedule divides into three periods, each with two weeks of class meetings and three weeks of internet-based project work. Classes will meet on Tuesdays and Thursdays from 1600 to 1830 in Duncan Hall room 3076 on the Rice campus. A tentative schedule for class meetings is: January 10, 12, 17, and 19; February 14, 16, 21, and 23; and March 21, 23, 28, and 30. I will communicate with participants during the intervening project periods by email and, when appropriate, by skype.

\section{Expectations}
In view of the advanced graduate nature of this course, no general performance metrics are appropriate: each participant will determine their level of involvement and expectations for the course.

\section{Resources}
A number of good sources exist for the topics of this course, some with links to codes and computational exercises, and I will refer participants to them whenever appropriate. Probably the closest in ``Courant and Hilbert'' spirit is Laurent Demanet's 18.325 class notes \cite[]{Demanet:325notes}. These were written for a graduate course in mathematics at MIT, and it shows: topics such as NMO do not show up, but the foundations of imaging theory and the adjoint state method, amongst other things, receive an elegant and complete treatment. The now-classic book by Norm Bleistein, Jack Cohen, and John Stockwell \cite[]{BleisteinCohenStockwell:01} gives a very thorough treatment of high-frequency asymptotics and ``Kirchhoff'' imaging.  Gerry Schuster's book on seismic imaging basics \cite[]{Schuster:10} is an excellent overview of the fundamental geophysical and data processing concepts, with Matlab exercises. Gary Margrave's book \cite[]{Margrave:book} gives a comprehensive account of seismic data processing, with links to the extensive CREWES Matlab codes. Finally, several conventional texts, for example those by \cite{Yil:01} and \cite{DobSav:88}, are excellent resources.

\bibliographystyle{seg}  % style file is seg.bst 
\bibliography{../../bib/masterref}

### 2019-10-19 repro/caam641/madfigs/paper.tex
repro docs dir added

\title{Resources: Reading and Software}
\maketitle 
\label{ch:madfigs}

\section{Introduction}
This book rests on a foundation of science and software provided by many talented researchers. The following paragraphs list some important background reading, and discusses the book's software environment and its use.

\section{Reading}
Lots of good background reading exists for the topics of this course, some with links to codes and computational exercises, and I will refer participants to them whenever appropriate. Probably the closest in ``Courant and Hilbert'' spirit is Laurent Demanet's 18.325 class notes \cite[]{Demanet:325notes}. These were written for a graduate course in mathematics at MIT, and it shows: topics such as NMO do not appear, but the foundations of imaging theory and the adjoint state method, amongst other things, receive an elegant and complete treatment. The now-classic book by Norm Bleistein, Jack Cohen, and John Stockwell \cite[]{BleisteinCohenStockwell:01} gives a very thorough treatment of high-frequency asymptotics and ``Kirchhoff'' imaging.  Jerry Schuster's book on seismic imaging basics \cite[]{Schuster:10} is an excellent overview of fundamental geophysical and data processing concepts, with Matlab exercises. Gary Margrave's book \cite[]{Margrave:book} gives a comprehensive account of seismic data processing, with links to the extensive CREWES Matlab codes.

Full waveform inversion (FWI) is a new enough (as a topic of widespread interest) that comprehensive references to complement the classic 1987 Tarantola book (republished by SIAM as \cite[]{Tarantola:05}) are only recently beginning to appear. See for example \cite{Fichtner:10}. FWI is well-known to be only locallly convergent (a phenomenon that this course will illustrate, and to some extent explain). In some sense you have to ``know the answer before you ask the question''.  Amongst many approaches that have been suggested for globalizing FWI convergence, I will dwell on extended inversion, an idea borrowed from velocity analysis. The literature on extended inversion has become fairly large. The reveiw paper \cite{geoprosp:2008} includes an extensive bibliography up to about 2007. Course reading for this segment will include this review paper and several papers appearing in the last 12 months.

Excellent general texts on geophysical prospecting, for example those by \cite{Yil:01} and \cite{DobSav:88}, set the topics discussed here in a wider context.

For the fundamental properties of wave equations and their solutions, the main reference is \cite{BlazekStolkSymes:13}, an open access paper available through the journal's web site, {\tt http://iopscience.iop.org/journal/0266-5611}. For finite difference methods, the classic is \cite{RichtMort}, which (so far as I know) remains the only mathematically complete account of convergence in book form, half a century after its publication. I will provides notes on a special case using energy estimates, that suffices for for the FD methods used in this course (and most of those used in computational seismology). Excellent books on other aspects of finite difference methods include \cite{Cohen:01} and \cite{Leveque:07}. For the section on asymptotic inversion, I will refer to my old notes \cite[]{SymesNotes}, and to the papers \cite[]{HouSymes:15}, \cite[]{HouSymes:17}, along with many papers cited there. The mathematics of extended inversion exists only in fragmentary form. I will present an account in the final part of these notes.

\section{Software} 
A project like this book would simply be impossible without a high quality open source foundation. In particular, I make extensive use of Seismic Unix (SU) and Madagascar, and of the TRIP package develped by my group. These three packages are the results of multi-decade efforts by dozens of talented people, to whom I am deeply grateful. Both SU and Madagascar are community projects, with many contributors. Both have chief contributor/maintainers: for SU, John Stockwell, and for Madagascar, Sergey Fomel (who is responsible for the concept, design, and a great deal of the code).

I presume that the active reader will work in a Un*x-like operating environment, either a flavor of Linux or Mac OSX. The packages on which the book's examples depend will need to be built from C/C++ source code, so you will need C/C++ compilers. Several parts of the code take advantage of relatively recent language innovations: you compilers will need to support key features of the C99 and C++11 standards. Gnu ({\tt gcc, g++}) version 4.4 and later, and Intel ({\tt icc, icpc}) version 13 and later, are adequate and widely available. You will also need Python (version 2.7 or later) and SConstruct (any recent version). Recent Linux distributions generally include installed Python, and so do at least the latest Mac OSX releases. Otherwise you will need to install Python from the {\tt www.python.org}. SConstruct will be installed automatically as the first step in installing Madagascar, if you haven't already installed it, and is needed for installation of TRIP. Seismic Unix (SU) uses Gnu make as its build system. All of the environments I have mentioned come with a suitable version of Gnu make installed.

Given the prerequisites outlined in the preceding paragraph, you should be able to download and install SU, Madagascar, and TRIP from their web sites:
\begin{verbatim}
reproducibility.org

cwp.mines.edu/cwpcodes

www.trip.caam.rice.edu/software
\end{verbatim}
In each source tree you will find installation instructions at the top level. Madagascar and SU have well-established release procedures and should install with minimal headaches. TRIP is newer, and has received less systematic testing; the release from the web site has installed correctly on Linux and Mac OSX under recent versions of Gnu and Intel suites. TRIP defines several modes of parallel execution via MPI. To enable these, it will be necessary to supply also a compatibly installed MPI.

That much will get you through building intermediate data from raw data and the figure files (in Madagascar {\tt .vpl} format) from intermediate data. Inclusion in finished papers, or in this book, requires \LaTeX. The Tutorial on the Madagascar web site explains how to obtain and install \LaTeX, if you haven't already, and how to make it available to the Madagascar framework.

\section{Seismic Unix}
Installation of SU is straightforward - just remember to comment out {\tt XDRFLAG} in Makefile.config, otherwise you will have to set XDR flags in TRIP software as well and various other complications will ensue. The world is little endian now, and XDR solves a problem that went away.

There is lots of tutorial material about SU on the web, beginning with
\begin{verbatim}
http://www.cwp.mines.edu/sututor/sututor.html
\end{verbatim}

\section{Madagascar}
Madagascar is both a processing system, with hundreds of core and user-contributed utilities, and a framework for relatively straightforward publication of reproducible research. While the processing system is very useful (and I use it a lot in the examples to follow), the reproducible research framework is unique: to my knowledge, no other available framework is as well-adapted to merging large-scale computation with text describing its significance. However, the system documentation tends to leave the novice user in the dark about a number of important issues. Most current documentation consists of oversimplified or incomplete examples, with little guidance about incorporating high-performance computing environments. This section gives some additional information on the steps necessary to produce a paper via Madagascar, {\em beyond the information provided on the Madagascar web site}. I strongly recommend that you
\begin{itemize}
\item read {\tt http://reproducibility.org/wiki/Tutorial} and try some of the examples
\item read {\tt http://reproducibility.org/wiki/Reproducible\_}{\tt computational\_experiments\_using\_SCons}
\end{itemize}
I will only add a few remarks about aspects glossed over or not treated in the Madagascar web site tutorial materials. To illustrate these points, I'll use the Chapter 2 project directory {\tt caam641/basic/project}. The  {\tt SConstruct} file found there implements simple processing workflow.

There are five major steps, starting with raw data and ending with a finished book chapter.

\subsection{Import Raw Data: {\tt Fetch}} 
{\tt Fetch} is well-described in the ``Reproducibility with Scons'' page on the Madagascar site. Its syntax is
\begin{verbatim}
Fetch(file, dir, server=URL)
\end{verbatim}
Here {\tt file} is the name of the file to be fetched from the directory {\tt dir}. The key point is that {\tt dir} must be a subdirectory of {\tt htdocs/data}, where {\tt htdocs} is the usual doc directory under the root accessed by the URL. For example, in {\tt caam641/basic/project/SConstruct} you will find the URL {\tt http://www.trip.caam.rice.edu}. The sysadmin at our site has aliased the root directory (under {\tt /}) accessed by this URL as {\tt /www.trip}. So the file accessed is {\tt /www.trip/htdocs/data/dir/file}.

With this understood, you can now set up your own {\tt Fetch} data sources.

One common plaint heard from Madagascar users is ``where did my data go?''. The answer is usually ``to the {\tt DATAPATH}''. Madagascar puts binary data in a directory signified by the {\tt DATAPATH} environment variable, and the output of {\tt Fetch} is assumed to be a binary data file. The default value of {\tt DATAPATH} is {\tt /var/tmp}, and that's what you'll get unless you set it yourself. Madagascar expects all working data process directories to be {\tt project} subdirectories of a paper directory (say, {\tt paper}, and each of those to be a subdirectory of a book directory (either a real book, like this one, in which case the papers are actually chapters, or an anthology or multi-author report, in which case the papers are stand-alone documents. In any case, the canonical directory tree {\tt book/paper/project} is how Madagascar organizes practically everything. For {\tt Fetch} executed in {\tt book/paper/project}, Madagascar creates a subdirectory of {\tt DATAPATH} named {\tt book/paper/project} and places the actual file retrieved by {\tt Fetch} there. A link is placed in the working directory, that is, {\tt book/paper/project}.

\subsection{Generate Intermediate Data: {\tt Flow}}
{\tt Flow} is well-explained in the Madagascar documentation. The only point I wish to add is that it's entirely possible to use commands other than those provided by Madagascar in implementing {\tt Flow}, and it's even possible to override the filter design of {\tt Flow}. For example, in {\tt basic/project/SConstruct}  I use the SU command {\tt segyread} to convert the SEGY data file (with standard SEGY structure and IBM binary word order) to an SU file (stripped of ebcidic and reel header, trace data converted to native four-byte floats). {\tt segyread} is not a filter: it expects the input file to be specified as the vallue of {\tt tape=}. {\tt segyread} does dump its output to {\tt stdout}, but I wanted to clean up the binary and real headers after returning from {\tt segyread}, so the command I chose to embed in the flow is not even half a filter. To signify that data is {\em not} read from {\tt stdin}, add {\tt, stdin=0} after the command string. To signify that data is {\tt not} written to {\tt stdout}, add {\tt, stdout=-1}. In the {\tt segyread} example, I use both. You will see lots of similar constructions in the examples to come.

The command in a flow is a Python string, with any literal parts enclosed in single quotes. If you have defined non-Madagascar commands as string variables, as I recommend, these must be outside the quoted string literals: for example,
\begin{verbatim}
.. segyread + ' tape=${SOURCE}...'
\end{verbatim}
Here {\tt segyread} has been assigned the value {\tt<path to CWP bin dir>/segyread} at the top of the SConstruct file: in the above fragment, it is a Python string variable, concatenated (Python addition of strings) with a literal that defines its arguments. Notice that the source file (second argument of {\tt Flow}) is referenced as {\tt \$\{SOURCE\}}, a convention that makes it easy to generate a number of similar commands with minimal typing. See the Tutorial for more on this.

\subsection{Generate Figures: {\tt Result}}
Chapter 2 \LaTeX source resides in the subdirectory {\tt basic} of the {\tt caam641} book directory.
As for every other such chapter or paper organized according to the Madagascar framework, each figure to be included in the paper (or chapter) is recorded as a {\tt .vpl} file in the {\tt project/Fig} subdirectory, by evaluation of the {\tt Result} function.

{\tt Result} takes three arguments: the name of a target figure file, the name of a source data file, and a command for producing the former from the latter. 
A typical example from {\tt basic/project/SConstruct}: 
\begin{verbatim}
Result('parastack','parastack.su','suread endian=0  read=data | put label1=Time label2=Trace unit1=s | grey clip=1.e+7 xinch=10 yinch=5')
\end{verbatim}
Note that the graphics file is in Madagascar graphics format ({\tt .vpl}), created in {\tt project/Fig}, but the source data file is SU (SEGY) format. Even though the rest of the workflow is implemented using SU commands, the graphics commands in our examples use Madagascar utilities only, in order to conform to the reproducible research design. {\tt sfsuread} converts the SEGY traces in {\tt parastack.su} into RSF format, {\tt sfput} adds header words for labels and units, and {\tt sfgrey} converts the RSF data to graphics format. As usual, the prefix {\tt sf} that is part of the name of every Madagascar command may be left off within the command argument for {\tt Result} or {\tt Flow}. Also, the figure file suffix {\tt .vpl} is understood, and the name of the target is simply the filename root. If the input data file ({\tt parastack.su} in this case) were an {\tt .rsf} file, then the figure file root name and the data file root name can be presumed to be the same, and the souce data file name (second argument of {\tt Result}) may be left out.

For example, if we were to convert {\tt parastack.su} to RSF format first, in a separate {\tt Flow}, then the prduction of this figure could read
\begin{verbatim}
Flow('parastack','parastack.su','suread endian=0  read=data | 'put label1=Time label2=Trace unit1=s')
Result('parastack', 'grey clip=1.e+7 xinch=10 yinch=5')
\end{verbatim}
The final result ({\tt basic/project/Fig/parastack.vpl}) is exactly the same, but now there is an extra intermediate file, {\tt parastack.rsf} in {\tt basic/project} (and its corresponding data file, {\tt parastack.rsf@}, residing in the data path). Note that the suffice {\tt .rsf} is also presumed in any file not containing a period {\tt .} in its name, and may also be left off, as {\tt .vpl} was left off in the specification of the {\tt Result} target..

SU commands (and other non-Madagascar commands, such as compiled-in-place IWAVE) must be defined, most reliably by full pathnames. See the {\tt SConstruct} file in {\tt basic/project} for a good way to do this.

Otherwise, only the standard RSF project framework is required ({\tt from rsf.proj import *} at the top of the {\tt SConstruct}) for this part of the process.

You can do this at home: in your copy of {\tt caam641/basic/project}, run {\tt scons -c} to clean everything up, then run {\tt scons} to regenerate everything.

\subsection{Archive the Figures}
Madagascar defines a sort of figure repository - actually, just a directory with a specific subdirectory tree structure in which copies of figure files are stored for future use. You control the location of this directory with the environment variable {\tt RSFFIGS}.  This is a local (platform-specific) definition - for example, on one of my laptops, {\tt .bashrc} contains the line
\begin{verbatim}
export RSFFIGS=/Users/symes/Documents/tripbooks/figs
\end{verbatim}
and you would do something similar if you use {\tt tcsh}.

The figures get into the RSFFIGS figure ``repository'' by the {\tt scons lock} command, run in the {\tt project} subdirectory.

For example, if you have successfully run {\tt scons} in {\tt basic/project}, then execute {\tt scons lock} to ``lock'' the figure files, i.e. copy them to the RSFFIGS directory.

After you do this, inspect {\tt RSFFIGS}: you will see a {\tt caam641} subdirectory, i.e. one with the same name as the parent directory of {\tt basic} - this would be created, if it did not exist before. In {\tt \$RSFFIGS/caam641}, you will see a {\tt basic} subdir, and within that a {\tt project} subdir, containing copies of all the figure files from {\tt caam641/basic/project/Fig}. This is Fomel's device for keeping the figures for each project in a unique place: create a subdirectory path in {\tt RSFFIGS} mimicing the directory path under the ``book'' ({\tt scratch}, in this case) and put the figures there. The construction follows the standard {\tt book/paper/project} structure of Madagascar reproducible research.

\subsection{Build the Paper}
You can now build the paper in the {\tt caam641/basic}, by typing {\tt scons}. You will see the {\tt .vpl} files from the {\tt RSFFIGS} subdir converted to {\tt .pdf} and incorporated in the paper {\tt .pdf} file. The rigid choice of directory structure and environment variables makes all this work: you have to follow {\tt book/paper/project}!

If you visit all of the {\tt caam641/chapter/project} directories and execute {\tt scons lock} successfully in each, then you can execute {\tt scons} in {\tt caam641} and reproduce the book that you are now reading!

Once again, only the standard RSF reproducible research framework is needed to make all of this work.

A bit reflection may suggest that actually the procedure outlined here and on the Madagascar web site really only makes sense when all of the work takes place in a single address space. At the moment, Madagascar does not provide adequate tools for incorporating the results of remote computations, for example at supercomputer centers. Later in the book, I'll outline a workaround.

\section{TRIP}
TRIP contains two subpackages:
\begin{itemize}
\item Rice Vector Library: a middleware package for connecting complex simulations to linear algebra and optimization algorithms, along with a number of linear algebra and optimization algorithms implemented in terms of this middleware;
\item IWAVE, a framework for regular grid finite difference simulation, supplied with RVL interfaces, with several acoustic and elastic simulators implements in terms of the framework. Includes commands invoking the simulators, and their associated maps (Born approximation, adjoint, etc.).
\end{itemize}
TRIP's user interface slavishly imitates Madagascar and SU (and SEPLib, their immediate ancestor), insofar as possible: commands self-doc, and parameters are passed by keyword=value pairs. The chief difference is that TRIP commands are not filters (reading from stdin, writing to stdout). That unavoidable, as they represent interactions amongst data stored in many files: there are usually multiple inputs, and often multiple outputs. Note that SU commands are also not all structured as filters, for the same reasons, and one uses the same devices to include TRIP commands in Madagascar {\tt Flow}s. The first chapter to include examples of this is Chapter 3, Born Approximation, and there are many examples in following chapters.

In the current release, commands read/write trace data from/to SU files, and regularly gridded spatial data from/to RSF file pairs.

\bibliographystyle{seg}  % style file is seg.bst 
\bibliography{../../bib/masterref}

### 2019-10-19 repro/caam641/ray/paper.tex
repro docs dir added

\title{Geometric Optics}
\maketitle

\label{ch:ray}

\inputdir{project}

\section{Introduction}
The ``geometric optics'' of the section title refers to a high frequency asymptotic approximation to wave motion. In combination with the Born approximation explained in the last section, this approximation leads to most of the intuition and many of the methods used in practical seismic data processing. Recall that the Born approximation is particularly accurate when the background is smooth and the perturbation oscillatory on the wavelength scale. This is exactly the setting in which geometric optics is effective in approximating the Born approximation.

\section{Progressing Wave Expansion}

%To understand the perturbational field $\delta p$, it is evident from 
%\ref{pert}
%(2)
%that we must first understand the background field $p$.  As might be guessed
%from the point-source assumption, this field is singular --- in fact, 
Begin once again with the wave equation \ref{eqn:cdawe}, which we
repeat here for convenience:
\begin{eqnarray}
\label{eqn:raycdawe}
\frac{\partial^2 p}{\partial t^2} -v^2 \nabla^2 p& = & f\\
p &= & 0, \,t<<0 \nonumber 
\end{eqnarray} 
Denote by $G(\bx,t;\bx_s)$ the Green's function, that is, the solution of equation \ref{eqn:raycdawe} for 
$f(\bx,t) = \delta(t)\delta(\bx-\bx_s)$. In view of the linearity of the wave equation and the time-independence of the coefficients, the solution of \ref{eqn:raycdawe} may be expressed as 
\begin{equation}
\label{eqn:green}
p(\bx,t ) = \int dt_s \int dx_s \,f(\bx_s,t-t_s)G(\bx,t_s;\bx_s)
\end{equation}
As noted in the last section, if $v$ is constant, then in 3D 
\begin{equation}
\label{eqn:green3d}
G(\bx,t;\bx_s) = \frac{\delta(t-r/v)}{4\pi r}, \,r=|\bx-\bx_s| = \sqrt{(\bx-\bx_s)^T(\bx-\bx_s)}.
\end{equation}
whereas in 2D,
\begin{equation}\label{eqn:green2d}
G(\bx,t;\bx_s)=\frac{1}{\pi v^2} \int_0^{\sqrt{t-r/v}}d\sigma \frac{\delta(t-\sigma^2-r/v)}{\sqrt{\sigma^2 + 2r/c}}. 
\end{equation}
While it is not possible to write such explicit expressions for the
fundamental solution in , it is possible
to describe the leading singularity of $G$, assuming smoothness of the
wave velocity $v$ (physically, separation of scales).  This is accomplished {\em via} the
{\em progressing wave expansion} (~\cite{CourHil:62}, Ch. VI).  Each
of the formulas for $G$ above is of the form $a(\bx,\bx_s)S(t-\tau(\bx,\bx_s))$
where $a$ and the travel time function $\tau$ are smooth except possibly at
$\bx=\bx_s$, and $S(t)$ is singular at $t=0$.  The progressing wave expansion
allows the extension of this expression away from $\bx=\bx_s$, up to a limit
signaled by a fundamental change in the nature of the wavefield, and with an
error which is {\em smoother} than $S$: it takes the form
\begin{equation}
\label{eqn:progwave}
G(\bx,t;\bx_s) = a(\bx,\bx_s)S(t-\tau(\bx,\bx_s)) + R(\bx,t;\bx_s)
\end{equation}
where $R$ is in some sense to be smoother than $S$, and $a$ and $\tau$ are
assumed to be smooth in some as-yet unspecified region. The first term
clearly expresses wave motion: if it is possible to show that it is
truly identifiable as separate (more singular) than the remainder $R$,
then the progressing wave expansion justifies the name of the wave equation.

Applying the wave
operator to the right-hand side of equation \ref{eqn:progwave}, we obtain
\begin{eqnarray}
\label{eqn:progwaveexp}
\lefteqn{\left(\frac{\partial^2}{\partial t^2} - v^2\nabla^2 \right)
  G} \nonumber \\[1ex]
& = & a( 1 - v^2|\nabla \tau|^2)S''(t-\tau) \nonumber\\[1ex]
  && - \;  v^2( 2\nabla a \cdot \nabla \tau +a \nabla^2 \tau)
     S'(t-\tau) \nonumber \nonumber \\[1ex]
  && - v^2\nabla^2 a S(t-\tau) 
   + \frac {\partial ^2 R}{\partial t^2} -  v^2 \nabla^2 R \nonumber \\[1ex]
&=&  \delta(\bx-\bx_s)\delta(t) \;.
\end{eqnarray}
Formally, the terms written in the above order have decreasing orders of
singularity, so that if $G$ is to solve the wave equation for $\bx \ne
\bx_s$, each of the coefficients above ought to vanish.  Certainly, if
\begin{equation} 
  \label{eqn:eik}
|\nabla \tau|^2 = \frac{1}{v^2} 
\end{equation}
\begin{equation}
   \label{eqn:transp} 
2\nabla a \cdot \nabla \tau + a\nabla
^2\tau = 0  
\end{equation}
then the first two terms vanish. The last term vanishes if
\begin{equation}
\label{eqn:rem0}
\delta(\bx-\bx_s)\delta(t) + v^2 \nabla^2 a S(t-\tau) = \frac {\partial ^2 R}{\partial t^2} -  v^2
\nabla^2 R,
\end{equation}
an inhomogeneous wave equation for the remainder $R$.

Equation \ref{eqn:eik} is the {\em eikonal equation} of geometric optics (of which the
progressing wave expansion is a variant). Inspecting the Green's functions \ref{eqn:green2d} and \ref{eqn:green3d}, evidently $\tau(\bx,\bx_s) = |\bx-\bx_s|/v$ for constant $v$, and indeed it is easy to verify that this function satisfies the eikonal equation \ref{eqn:eik}. The correct ``initial'' condition at $\bx=\bx_s$ forces $\tau$ to be asymptotic to the constant $v$ solution there:
%(3) 
\begin{equation}
\label{eqn:eik_init}
\tau(\bx,\bx_s)/|\bx-\bx_s| = v(\bx_s) \mbox{ as }  |\bx-\bx_s| \rightarrow 0.
\end{equation}

The second condition, the {\em transport equation} \ref{eqn:transp},
may be rewritten as
\begin{equation}
\label{eqn:transpalt}
\nabla\cdot (a^2 \nabla \tau) = 0
\end{equation}
Reference to the constant-coefficient Green's functions suggests that
solutions will need a specified singularity at $\bx=\bx_s$. This topic
will be addressed in the next section.

As stated above, the progressing wave expansion is not much of an
expansion. In fact, the decompositionn $G = aS(t-\tau)+R$ is the first
term of a series, 
\begin{equation}
\label{eqn:progwaveN}
G(\bx,t;\bx_s) = \sum_{n=0}^N a_n(\bx,\bx_s) S_n(t-\tau(\bx,\bx_s)  + R_N(\bx,t;\bx_s)
\end{equation}
in which $S_k$ is the $k$th primitive of $S$, that is,
\[
S_0 = S;\,\, a_0=a;\,\,\frac{dS_k}{dt} = S_{k-1}, \,k=1,2,...
\]
That is, \ref{eqn:progwaveN} decomposes $G$ into successively less
singular terms. The case $N=1$ will be important in understanding the
singularity of $G$:
\begin{equation}
\label{eqn:progwave1}
G(\bx,t;\bx_s) = a_0(\bx,\bx_s) S_0(t-\tau(bx,\bx_s) +
a_1)\bx,\bx_s)S_1(t-\tau(\bx,\bx_s) + R_1(\bx,t;\bx_s).
\end{equation}
Applying the wave operator, one sees that $G$ solves the wave equation
for $\bx \ne \bx_s$ provided that
\begin{eqnarray}
\label{eqn:transp1}
2\nabla a_1 \cdot \nabla \tau + a_1 \nabla^2 \tau &=& \nabla^2 a_0 \nonumber \\
v^2 \nabla^2 a_1 & = & \frac {\partial ^2 R_1}{\partial t^2} -  v^2 \nabla^2 R_1
\end{eqnarray}
\section{Ray Theory}

The {\em method of characteristics} extrapolates $\tau$ away from $\bx_s$. To see how this works,
suppose first that $\tau$ solves the eikonal equation, and let $\bX(t)$
be a solution of the system of ordinary differential equations
\begin{equation}
\label{eqn:xeqn}
\frac{d\bX}{dt} = v^2(\bX) \nabla\tau(\bX).
\end{equation}
Then
\begin{eqnarray}
\frac{d}{dt} \tau (\bX(t)) & = & \nabla\tau(\bX(t))\cdot
\frac{d\bX}{dt}(t)\\
& = & v^2(\bX(t)) |\nabla\tau(\bX(t))|^2 = 1 \;\;.
\end{eqnarray}
Therefore we can identify $\tau$ with $t$: 
if the segment $\{\bX(t'):t_0 \le t' \le t\}$
lies entirely in a domain in which $\tau$ is defined, then
\begin{equation}
\label{eqn:tausig}
\tau(\bX(t)) = \tau(\bX(t_0)) + t - t_0
\end{equation}
Thus from knowledge of the characteristic curves (rays) $\bX(t)$, we can
construct $\tau$.

Somewhat more surprisingly, it is possible
to construct the rays directly, which furnishes a construction of $\tau$ as
well. Define the vector 
\begin{equation}
\label{eqn:pdef}
\bP(t) = \nabla\tau(\bX(t))
\end{equation}
with units of time/length, or {\em slowness}. Note that you can write \ref{eqn:xeqn} in the form
\begin{equation}
\label{eqn:xeqn1}
\frac{d\bX}{dt} = v^2(\bX) \nabla\tau(\bX) = v^2(\bX) \bP.
\end{equation}
$\bP$ also solves a differential equation in terms of $\tau$:
\begin{eqnarray}
\label{eqn:pderiv}
\frac{d \bP}{dt}(t) & = & (\nabla\nabla\tau)(\bX(t)) \cdot \dot{\bX}(t)\\
& = & v^2(\bX(t))\nabla\nabla\tau(\bX(t)) \cdot \nabla\tau(\bX(t))\\
& = & \frac{1}{2} v^2(\bX(t))\nabla | (\nabla\tau)(\bX(t))|^2\\
& = & \frac{1}{2} v^2(x)(\nabla v^{-2})(\bX(t))\\
& = & - \frac{1}{2} |\nabla\tau(x)|^2(\nabla v^{2})(\bX(t)).
\end{eqnarray}
Define the {\em Hamiltonian}
\begin{equation}
\label{eqn:hamdef}
H(\bx,\bP) = \frac{1}{2}v^{2}(\bx)|\bP|^2
\end{equation}
then the equation \ref{eqn:xeqn1} for $\bX$ and \ref{pderiv} for $\bP$ can be written in the form
\begin{eqnarray}
\label{eqn:ham}
\dot{\bX} & = & \nabla_{\bP} H(\bX,\bP)\\
\dot{\bP} & = & -\nabla_{\bx} H(\bX,\bP)
\end{eqnarray}
These are  {\em Hamilton's equations} of classical mechanics, a system
of $2n$ autonomous ordinary differential equations. If $(\bX,\bP)$ is
a solution of the system \ref{eqn:ham}, both
\[
t \mapsto \bX(t)
\]
and
\[
t \mapsto (\bX(t),\bP(t))
\]
are called rays; in the mathematical literature, the latter
trajectory, with values in phase space, is known as a {\em
  bicharacteristic strip} or just a {\em bicharacteristic}.

It's an easy exercise to see that the equations \ref{eqn:ham} implies that $H$ is constant along rays, that is,
\begin{equation}
\label{eqn:constH}
\frac{d}{dt}H(\bX(t),\bP(t)) = 0
\end{equation}
if $(\bX,\bP)$ is a ray. The rays that figure in construction of the traveltime $\tau$ 
are related to $\tau$ through equations \ref{eqn:xeqn} and \ref{eqn:pdef}, so because of the iekonal equation satisfy $v(\bX)|\bP| = 1$ - so along such rays,
$H(\bX,\bP)=1/2$. Since $H-1/2$ generates the same Hamiltonian system,
these rays are also called {\em null bicharacterisitics}.

The rays that play a role in the construction of the Green's function
for a source at $\bx=\bx_s$ are those for which 
\begin{equation}
\label{eqn:initH}
\bX(0)=\bx_s, \, \bP(0) = \bP_s, \, v(\bX_s)|\bP_s|=1.
\end{equation}
Solving the system \ref{eqn:ham} and initial conditions
\ref{eqn:initH} generates a mapping
\begin{equation}
\label{eqn:polar}
(t,\bP_s) \mapsto (\bX(t,\bP_s), \bP(t,\bP_s)) 
\end{equation}
called the ray (polar) coordinate system centered at $\bx_s$. For
sufficiently small $t>0$, the Jacobian of the map \ref{eqn:polar} is
nonsingular for any $\bP_s$ with $v(\bx_s)|\bP_s|=1$ (exercise!). It
follows that ray polar coordinates define an invertible differentiable
coordinate system on an open set $\Omega(\bx_s)$ containing
$\bx_s$. For constant $v$, of course, $\Omega(\bx_s)$ consists of all
of Euclidean space, and
\begin{eqnarray}
\label{eqn:polar0}
\bX(t,\bP_s) &=& \bx_s + t v(\bx_s)^2\bP_s,\\
\bP(t,\bP_s)) &=& \bP_s. 
\end{eqnarray}

Given a ray coordinate system centered at $\bx_s$ and valid in the set
$\Omega(\bx_s)$, define
\begin{equation}
\label{eqn:taudef}
\tau(\bx, \bx_s) = t \mbox{ for } \bx=\bX(t,\bP_s).
\end{equation}
That is, $\tau(\bx,\bx_s)$ is the ``radius'' in the polar coordinate
system.

The first step in showing that $\tau$, so defined, satisfies the eikonal equation in
$\Omega(\bx_s)$ is the proof of the

\noindent {\bf Claim} The ray tangent vector $d\bX/dt$ at
$\bX(t,\bP_s)$ is perpindicular to the surface (isochron) $\{\bx=\bX(t,\bP_s):\mbox{ all } \bP_s \mbox{ such that } v(\bx_s)|\bP_s|=1\}$.

That is, for any vector $\delta \bP_s \perp \bP_s$, 
\begin{equation}
\label{eqn:perp}
(\nabla_{\bP_s} \bX(t,\bP_s) \delta \bP_s) \cdot d\bX(t,\bP_s)/dt = 0.
\end{equation}
since all of the tangent vectors to the isochron take the form of the
vector on the left. To see this, differentiate the left hand side of
\ref{eqn:perp}:
\[
\frac{d}{dt} \left((\nabla_{\bP_s}\bX \delta \bP_s) \cdot \frac{d\bX}{dt}\right) 
\]
\begin{equation}
\label{eqn:dperp}
= \left(\nabla_{\bP_s}\frac{d \bX}{dt} \delta \bP_s\right) \cdot \frac{d \bX}{dt}  + (\nabla_{\bP_s}\bX \delta \bP_s) \cdot \frac{d^2\bX}{dt^2}
\end{equation}
Differentiation of the first equation in \ref{eqn:ham} and using both
equations and the constant-$H$ condition \ref{eqn:constH} yields the
second-order ray equation, useful in its own right:
\begin{equation}
\label{eqn:rayde}
\frac{d^2\bX}{dt^2} = 2 \left(\frac{\nabla v}{v} \cdot
  \frac{d\bX}{dt}\right)\frac{d\bX}{dt} - v \nabla v
\end{equation}
The first term in \ref{eqn:perp} may be rewritten as 
\[
\frac{1}{2}\left(\nabla_{\bP_s} \left|\frac{d \bX}{dt}\right|^2\right)\delta \bP_s =
\frac{1}{2}\nabla_{\bP_s}v^2 \delta \bP_s
\]
\begin{equation}
\label{eqn:perp1}
=v \nabla v \cdot \nabla_{\bP_s}\bX \delta \bP_s.
\end{equation}
Re-write equation \ref{eqn:dperp} using the second-order ODE
\ref{eqn:rayde} in the first term and the identigy \ref{eqn:perp1} in
the second: net result is that the second term cancels, leaving
\begin{equation}
\label{eqn:perp2}
\frac{d}{dt} \left(\nabla_{\bP_s}\bX \delta \bP_s\right) \cdot \frac{d\bX}{dt} = 2 \left(\frac{\nabla v}{v} \cdot
  \frac{d\bX}{dt}\right)\frac{d\bX}{dt} \cdot (\nabla_{\bP_s}\bX \delta \bP_s) 
\end{equation}
This is a homogeneous linear ordinary differential equation for the
quantity
\begin{equation}
\label{eqn:perp3}
\frac{d\bX}{dt} \cdot (\nabla_{\bP_s}\bX \delta \bP_s).
\end{equation}
Rays are asymptotic to the frozen-coefficient trajectories specified
in \ref{eqn:polar0} as $t \rightarrow 0$, and for these, the dot
product in \ref{eqn:perp3} is zero identically - at the origin of the polar coordinates, the ray direction is perpindicular to the directions of constant $|\bP_s|$. Therefore the
effective initial condition for the ODE \ref{eqn:perp2} is zero, and
so the inner produce in \ref{eqn:perp3} vanishes identically,
establishing the Claim.

Choose any basis for the $d-1$-dimensional subspace of $\bR^d$
perpindicular to $\bP_s$ (reminder - $d$ = space dimension). Think of
$\bX(t,\bP_s$ as a function of $t$ and the $d-1$ coefficients of this
basis, which generate a vector perpindicular to $\bP_s$. For a
coefficient vector near the origin, this vector added to $\bP_s$
projects uniquely onto the sphere of radius $1/v(\bx_s)$. A
restatement of the Claim is that the first column ($d\bX/dt$) of the
Jacobian of $(t,\bP_s) \rightarrow \bX$ is orthogonal to to the other
columns. Therefore the first row of the inverse, which by the chain
rule is $\nabla \tau(\bx,\bx_s)$, has length reciprocal to that of
$d\bX/dt$. Since the latter has length $v$, the eikonal equation
follows.

The identification $\tau(\bx,\bx_s) = t$ when $\bx=\bX(t,\bP_s)$
shows that $\tau$ is the time of arrival of the ray from
$\bx_s$ to $\bx$. So $\tau$ is often called the travel- or arrival-time.

Note that this ``ray-tracing'' construction of $\tau$ works only in
regions $\Omega(\bx_s)$ within which ray polar coordinates are
valid. The loss of validity comes when rays with two different values
of takeoff slowness $\bP_s$ intersect - that is, when two or more rays
pass over both $\bx_s$ and $\bx$. If such {\em conjugate points}
exist, then it is not hard to see that envelopes of ray families
(defined by ranges of $\bP_s$) must exist; these envelopes are known
as {\em caustics}. Analysis of progressing waves near such caustics is
somewhat harder than for the non-caustic case. The developments of
this and the next few sections assumes that all of the important
energy in the wavefield may be explained with rays that have not yet
touched a caustic in their travel from the source point $\bx_s$, that
is, the analysis will be confined to the region $\Omega(\bx_s)$ of
{\em simple ray geometry}.

The Hamiltonian system \ref{eqn:ham} is the basis for the most common
approach to computing $\tau$, by solving the initial value problem for
this system of ODEs. The main difficulty to be overcome in this
approach is that the rays do not generally pass over grid points, so a
gridded version of $\tau$ is not an immediate by-product. Many clever
workarounds have been proposed \cite[]{Chapman:85,Vinje:93a}.

It is also possible to treat $\tau$ as a solution of the eikonal
equation \ref{eqn:eik}, as an initial value problem for a partial
differential equation. However $\tau$ appears to be no longer a
function if one uses the ray construction above, outside of the region
$\Omega(\bx_s)$ in which each point is connected to $\bx_s$ by a
unique ray. One approach constructs a function that is equal to $\tau$
in $\Omega(\bx_s)$ and elsewhere is the least value of $t$ along all
of the rays connecting to $\bx_s$. This first-arrival-time
construction may be approximated by solving special gridded finite difference or
finite element discretizations of the eikonal equation (~\cite{Vid:88}, 
~\cite{VantrSy:90}). For recent viewpoint and many references see
\cite[]{LuoQianBurridge:14}. It has also been possible to extend this
``Eulerian'' point of view, via appropriately posed PDE problems, to
computing global (multi-valued) travel time fields
\cite[]{FomelSethian:02,QianChengOsher:03}.

A simple 2D example exhibiting many of the features just described
appears in Figure \ref{fig:lensmedium}, showing a velocity field ($v(\bx)$)
with a slow (low value) anomaly below the source point $\bx_s$ at the
top of the figure. Figure \ref{fig:lensrayswavefronts} shows both rays
$t \mapsto \bX(t,\bP_s)$, and wavefronts $\tau=$ constant. The rays
and wavefronts are perpindicular: this is precisely equation
\ref{eqn:perp}, established above. The rays bend towards the lower
velocity region in the center, as follows directly from the 2nd order
equation \ref{eqn:rayde}. The envelope of rays (caustic) are clearly
visible. In the region beyond the caustics, three rays pass over each point.

\plot{lensmedium}{width=0.9\textwidth}{Low velocity lens in constant
  velocity (2 km/s) background. Dark bar at depth = 2 km is a
  reflector, used in later examples but not here. Borrowed from \cite{StolkSymes:04}.}
\plot{lensrayswavefronts}{width=0.9\textwidth}{Rays and wavefronts
  emanating from a source point above the center of the lens. Note
  that rays are perpindicular to wavefronts, consistent with the
  theory developed in the text, and that the bending of rays towards
  low velocity leads to the development of a caustic (ray envelope),
  which itself has a cusp singularity just below the lens. This
  ``butterfly'' caustic joins two ``fold'' caustic branches at the
  cusp. Below the caustic, each point is joined to the source by three
  rays: one (with greater traveltime) passing through the slow lens,
  and two passing around the lens through the faster background
  medium.  Borrowed from \cite{StolkSymes:04}.}

\section{Amplitudes}
Having computed $\tau$, it is easy to compute $a$.  Indeed, the em
transport equation \ref{eqn:transp}
%(4) 
may be re-written as an ordinary differential equation along rays
\[
\frac{d}{dt} a (\bX(t)) - b(\bX(t))a(\bX(t)) = 0
\]
where
\[ 
b = \frac{1}{2} \nabla^2\tau.
\]
Thus $a$ may be computed by quadrature along the ray family associated with
$\tau$.  Initial values (for small $t$) for $a(\bX(t))$ cannot be read
off directly for $t=0$, since the amplitude is singular at
$\bx=\bx_s$, but must be inferred from the $t \rightarrow 0$ asymptotics.

Rather than go through that exercise, note that the solution of the
transport equation has a nice geometric interpretation.
Construct a ``ray tube'' $R$ (note - this is not the remainder term R, defined above and discussed in the next section!), whose lateral boundary surface $B$ is swept out
by rays, and whose
ends are surfaces $l$ and $L$ on which $\tau =$ constant:
$\partial R
= B \cup l \cup L$. A 2D schematic appears in Figure \ref{fig:amp}.

\plot{amp}{width=0.9\textwidth}{Sketch of ``ray tube'' construction. The segment ``dx'' is a projection onto the horizontal surface - it is useful in understanding prestack amplitudes, discussed in a later section. Borrowed from \cite{HouSymes:15}.}

Now integrate both sides of the re-written transport equation
\ref{eqn:traspalt} over $R$. The divergence theorem implies that 
\begin{equation}
\label{eqn:raytubeident}
\int_{R} \nabla\cdot(a^2\nabla\tau)=\int_{l\cup L}a^2\nabla\tau\cdot \mathbf{n}dA =0
\end{equation}
where $\mathbf{n}$ is the outward normal vector on $\partial \Gamma$.
The integral over the lateral boundary $B$ vanishes as the normal is
perpindicular to the rays that form it, hence to $\nabla \tau$. A
second consequence of the orthogonality \ref{eqn:perp} is that 
$\nabla \tau = v^{-1}{\bf n}$ on $L$, $=-v^{-1}{\bf n} $ on $l$ so the identity
\ref{eqn:raytubeident} implies that
\begin{equation}
\label{eqn:transpsoln}
\int_{l} a^2 v^{-1} =\int_{L} a^2 v^{-1}. 
\end{equation}
Shrinking the ray tube to a single central ray, one obtains in the limit
\begin{equation}
\label{eqn:transpsolnpt}
\frac{a^2(\bX(t_0))}{v(\bX(t_0))} = J(t_0,t_1)
\frac{a^2(\bX(t_1))}{v(\bX(t_1))}
\end{equation}
in which $J$ is the transverse Jacobian determinant, that is, the Jacobian determinant of the
map from $l$ to $L$ at the central ray intersection, defined by
tracing rays from $l$ to $L$. This determinant is the limit of the
volume ratio $|L|/|l|$ as the tube shrinks to the central ray.

The Jacobian and its determinant satisfy evolution 
equations along rays - solving these equations is {\em dynamic ray 
  tracing}, see \cite{cer85}. Equation \ref{eqn:transpsolnpt} implies that this construction fails
when the Jacobian becomes singular, which happens precisely at the
intersection of the central ray with a caustic (envelope of rays), and
at the source point itself (a sort of degenerate caustic). In principle, the
amplitude grows without bound near such points.

The actual behaviour
of the wavefield near caustics can't be inferred from the progressing wave expansion
in the form presented so far: extensions are required, discussed in a
later section. However, The analysis of the remainder term in
the next section requires an explicit description of the behaviour of
$a(\bx,\bx_s)$ as $\bx \rightarrow \bx_s$.

The 3D case is slightly
simpler. Introduce polar coordinates for the ray field $\bX$ centered at $\bx_s$:
\[
\bX = \bx_s + 
\left(\begin{array}{c}
r \cos \theta \cos \phi\\
r \sin \theta \cos\phi\\
r \sin \phi 
\end{array}\right),
\,\,r > 0, -\pi \le \theta \le \pi, -\frac{\pi}{2} \le \phi \le 
\frac{\pi}{2}
\]
and parametrize the initial slowness $\bP_s$ in terms of initial
azimuth and polar angles $\theta_s,\phi_s$:
\[
\bP_s(\theta_s,\phi_s) = \frac{1}{v(\bx_s)}
\left(\begin{array}{c}
\cos \theta_s \cos \phi_s\\
\sin \theta_s \cos\phi_s\\
\sin \phi_s 
\end{array}\right).
\]
Denote by $\bP^0_s = \bP_s(\theta^0_s,\phi^0_s)$ the initial slowness
for the central ray of the ray tube construction, and define the tube
to have distorted square cross section, of radius $\delta > 0$:
\[
R = \{\bX(t,\theta_s,\phi_s): 0 < t_0 \le t \le t_1,
|\theta_s-\theta_s^0| \le \delta, |\phi_s-\phi_s^0|<\delta\}.
\]
Then
\begin{equation}
\label{eqn:transpsoln}
\int_{l} a^2 v^{-1} =\int_{L} a^2 v^{-1}. 
\end{equation}
\begin{eqnarray}
\label{eqn:tubeends}
\int_{l} a^2 v^{-1}  & = & \int_{\theta^0_s-\delta}^{\theta^0_s+\delta}\,d\theta_s \,
\int_{\phi^0_s-\delta}^{\phi^0_s+\delta}\,d\phi_s 
          \,\mu(t_0,\theta_s,\phi_s)
                           (a^2v^{-1})(\bX(t_0,\theta_s,\phi_s))d\theta_s
                           d\phi_s \nonumber \\
\int_{L} a^2 v^{-1} & = & \int_{\theta^0_s-\delta}^{\theta^0_s+\delta}\,d\theta_s \,
\int_{\phi^0_s-\delta}^{\phi^0_s+\delta}\,d\phi_s 
          \,\mu(t_1,\theta_s,\phi_s)
          (a^2v^{-1})(\bX(t_0,\theta_s,\phi_s)) d\theta_s d\phi_s
\end{eqnarray}
The surface area density $\mu(t,\theta_s,\phi_s)$ satisfies
\begin{equation}
\label{eqn:areaelt}
\mu(t,\theta_s,\phi_s)d\theta_s \wedge d\phi_s =
\bX(t,\theta_s,\phi_s)^*dS_t
\end{equation}
in which $dS_t$ is the area element on the isochron surface
$S_t = \{\bx:\tau(\bx,\bx_s) = t\}$. By definition, $dS_t$ satisfies 
\begin{equation}
\label{eqn:areaeltprop}
dS_t({\bf V}_1,{\bf V}_2) = 1
\end{equation}
for any vector fields ${\bf V}_1,{\bf V_2}$ on  $S_t$ that form an
orthonormal oriented basis at every point. The orientation of $S_t$ is
given by choosing $v^{-1}\nabla \tau$ to be the unit normal vector.

Note that 
\begin{equation}
\label{eqn:drdt}
\frac{\partial \bX}{\partial t} = 
\frac{\partial r}{\partial t} e_r
+
r \cos \phi\frac{\partial \theta}{\partial t} e_{\theta}
+
r\frac{\partial \phi}{\partial t} e_{\phi}
\end{equation}
where
\[
e_r = 
\left(\begin{array}{c}
\cos \theta \cos \phi\\
\sin \theta \cos\phi\\
\sin \phi 
\end{array}\right),\,\,
e_{\theta}=\left(\begin{array}{c}
-\sin \theta \\
\cos \theta \\
0 
\end{array}\right),\,\,
e_{\phi}=\left(\begin{array}{c}
-\cos \theta \sin \phi\\
-\sin \theta \sin\phi\\
\cos \phi 
\end{array}\right),
\]
form an orthonormal basis of $\bR^3$. Similarly,
\begin{equation}
\label{eqn:drdtheta}
\frac{\partial \bX}{\partial \theta_s} = 
\frac{\partial r}{\partial \theta_s} e_r
+
r \cos \phi\frac{\partial \theta}{\partial \theta_s} e_{\theta}
+
r\frac{\partial \phi}{\partial \theta_s} e_{\phi}
\end{equation}
\begin{equation}
\label{eqn:drdphi}
\frac{\partial \bX}{\partial \phi_s} = 
\frac{\partial r}{\partial \phi_s} e_r
+
r \cos \phi\frac{\partial \theta}{\partial \phi_s} e_{\theta}
+
r\frac{\partial \phi}{\partial \phi_s} e_{\phi}
\end{equation}
Suppose ${\bf W_1},{\bf W_2}$ are arbitrary vector fields on $S_t$:
orthogonalizing them and using the algebraic properties of the 2-form
$dS_t$, it follows from the volume element property
\ref{eqn:areaeltprop} that 
\[
dS_t({\bf W}_1,{\bf W}_2) = \sqrt{|{\bf W}_{1}|^2|{\bf W}_2|^2 -
  \langle {\bf W}_1,{\bf W}_2\rangle^2}
\]
So 
\[
\mu_t = (\mu_t d\theta_s\wedge d\phi_s) \left(\frac{\partial}{\partial 
    \theta_s},\frac{\partial}{\partial \phi_s}\right)
\]
\[
= \bX^* dS_t \left(\frac{\partial}{\partial 
    \theta_s},\frac{\partial}{\partial \phi_s}\right) = dS_t
\left(\frac{\partial \bX}{\partial 
    \theta_s},\frac{\partial \bX}{\partial \phi_s}\right)
\]
\[
= \left(\left|\frac{\partial \bX}{\partial 
    \theta_s}\right|^2\left|\frac{\partial \bX}{\partial 
    \phi_s}\right|^2 - \left\langle \frac{\partial \bX}{\partial 
    \theta_s}, \frac{\partial \bX}{\partial \phi_s}\right\rangle^2
\right)^{\frac{1}{2}}
\]
Using the orthonormality of $e_r, e_{\theta}$, and $e_{\phi}$ and
the chain rules \ref{eqn:drdtheta} and \ref{eqn:drdphi}, see that this is
\[
= \left\{
\left[
\left(\frac{\partial r}{\partial \theta_s}\right)^2 
+
r^2\left(\cos^2 \phi\left(\frac{\partial \theta}{\partial \theta_s}\right)^2 
+
\left(\frac{\partial \phi}{\partial \theta_s}\right)^2\right)\right]
\left[
\left(\frac{\partial r}{\partial \phi_s}\right)^2 
+
r^2\left(\cos^2 \phi\left(\frac{\partial \theta}{\partial \phi_s}\right)^2 
+
\left(\frac{\partial \phi}{\partial \phi_s}\right)^2\right)\right]\right.
\]
\begin{equation}
\label{eqn:mucalc}
\left.-
\left[
\left(\frac{\partial r}{\partial \theta_s}\right) 
\left(\frac{\partial r}{\partial \phi_s}\right)
+
r^2\left(\cos^2 \phi
\left(\frac{\partial \theta}{\partial \theta_s}\right)
\left(\frac{\partial \theta}{\partial \phi_s}\right) 
+
\left(\frac{\partial \phi}{\partial \theta_s}\right) 
\left(\frac{\partial \phi}{\partial \phi_s}\right)
\right)
\right]^2
\right\}^{\frac{1}{2}}
\end{equation}

So asymptotic behaviour of the density $\mu$ as $t
\rightarrow 0$ follow from the asymptotics of $r,\theta,$ and $\phi$
and their angular derivatives.
I have implicitly assumed that the velocity field $v$ has as many
derivatives as are convenient; for the following argument, three
continuous derivatives are certainly sufficient. Then the solutions of
Hamilton's equations \ref{eqn:ham} are several times differentiable as
functions of $t$ and of the initial conditions, that is, of the
azimuth and polar angles, so
\ref{eqn:initH} imply that
\[
\bX(t,\theta_s,\phi_s) = \bx_s + t v(\bx_s)^2\bP_s(\theta_s,\phi_s) + t^2{\bf
  Y}(t,\theta_s,\phi_s)
\]
in which ${\bf Y}$ is smooth in $t \ge 0, \theta_s,\phi_s$. Therefore
\begin{equation}
\label{eqn:rasympt}
r(t,\theta_s,\phi_s) = v(x_s)t + t^2 y(t,\theta_s,\phi_s),
\end{equation}
where $y$ is several times differentiable for $t\ge 0$ and
real-valued. This equation in turn implies that for a non-vanishing
range of times $0 \le t < t^*$, there are constants $0 < v_* \le v^*$
so that
\begin{equation}
\label{eqn:rasgoodast}
v_*t \le r(t,\theta_s,\phi_s) \le v^* t,
\end{equation}
uniformly in $\theta_s,\phi_s$.

Under these conditions, the map $(t_0,\theta_s,\phi_s) \mapsto
(r,\theta,\phi)$ is a $C^k$ diffeomorphism on $[0,t^*)\times
(-\pi,\pi) \times (-\pi/2,\pi/2)$, into the slit $\bR^3$. Of course
part of $\bR^3$ is left out, but by using several versions of polar
coordinates with different polar axes, it is possible to cover all of
$\bR^3$ except the origin. For all of these rotated coordinates, the
coordinate change is of class $C^k$ up to $t_0=0$, so derivatives are
bounded uniformly on compact subsets of the domain. I will assume that
derivatives of order at least two are continuous.

Equation \ref{eqn:rasympt} also implies that
\begin{eqnarray}
\label{eqn:partr}
\frac{\partial r}{\partial \theta_s} & =& O(t^2)\nonumber \\
\frac{\partial r}{\partial \phi_s}    & =& O(t^2) 
\end{eqnarray}
Together with equation \ref{eqn:rasgoodast}, this shows that
\begin{eqnarray}
\label{eqn:quadr}
\frac{1}{r}\frac{\partial r}{\partial \theta_s} & =& O(t)\nonumber \\
\frac{1}{r}\frac{\partial r}{\partial \phi_s}    & =& O(t) 
\end{eqnarray}
A similar examination of the Hamilton equation for the slowness yields
the conclusion that
\begin{eqnarray}
\label{eqn:angles}
\theta &=& \theta_s + O(t) \nonumber\\
\phi    &=& \phi_s + O(t) \nonumber\\
\frac{\partial \theta}{\partial \theta_s} &=& 1 + O(t) \nonumber\\
\frac{\partial \theta}{\partial \phi_s} &=& O(t) \nonumber\\
\frac{\partial \phi}{\partial \theta_s} &=& O(t) \nonumber\\
\frac{\partial \phi}{\partial \phi_s} &=& 1 + O(t) 
\end{eqnarray}
In all cases, the remainders hiding inside the big $O$s are several 
times differentiable in $t \ge 0,\theta_s,\phi_s$.

Putting together equations \ref{eqn:mucalc}, \ref{eqn:quadr}, and
\ref{eqn:angles}, conclude that
\begin{equation}
\label{eqn:muasympt}
\mu((t,\theta_s,\phi_s) = r^2 \cos \phi_s(1 +t^2\nu(t,\theta_s,\phi_s)), 
\end{equation}
with $\nu$ of class $C^k([0,t^*) \times (-\pi,\pi) \times
(-\pi/2,\pi/2))$, $k \ge 2$.

That is,  $\mu$ is to leading order the same as the Euclidean area
element on the sphere of radius $r$. So the relations
\ref{eqn:tubeends} imply
\[
\int_{\theta^0_s-\delta}^{\theta^0_s+\delta}\,d\theta_s \,
\int_{\phi^0_s-\delta}^{\phi^0_s+\delta}\,d\phi_s 
          \,r(t_0,\theta_s,\phi_s)^2 \cos \phi_s (1 +t_0\nu(t_0,\theta_s,\phi_s))(a^2v^{-1})(\bX(t_0,\theta_s,\phi_s)) 
\]
\begin{equation}
\label{eqn:tubeends1}
= \int_{\theta^0_s-\delta}^{\theta^0_s+\delta}\,d\theta_s \,
\int_{\phi^0_s-\delta}^{\phi^0_s+\delta}\,d\phi_s 
          \,r(t_1,\theta_s,\phi_s)^2 \cos \phi_s (1 +t_1\nu(t_1,\theta_s,\phi_s))(a^2v^{-1})(\bX(t_1,\theta_s,\phi_s))
\end{equation}
Dividing by $4\delta^2$ and letting $\delta \rightarrow 0$, obtain
\begin{equation}
\label{eqn:tubeends2}
r(t_0,\theta_s,\phi_s)^2 \cos \phi_s (1
+t^2_0\nu(t_0,\theta_s,\phi_s))(a^2v^{-1})(\bX(t_0,\theta_s,\phi_s))=
r(t_1,\theta_s,\phi_s)^2 \cos \phi_s (1
+t^2_1\nu(t_1,\theta_s,\phi_s))(a^2v^{-1})(\bX(t_1,\theta_s,\phi_s))
\end{equation}
Changing variables to $r,\theta,\phi$ and discarding the factor $\cos
\phi_s$,
\begin{equation}
\label{eqn:tubeends3}
r_0^2 (1 +
\tau(r_0,\theta,\phi)^2\nu(r_0,\theta,\phi))(a^2v^{-1})(r_0,\theta,\phi) =
r_1^2 (1 +
\tau(r_1,\theta,\phi)^2\nu(r_1,\theta,\phi))(a^2v^{-1})(r_1,\theta,\phi)
\end{equation}
Divide both sides of \ref{eqn:tubeends1} by $v^{-1}(r_0,\theta,\phi)1 +\tau(r_0,\theta,\phi)^2\nu(r_0,\theta,\phi) $
and denote by $\gamma(\theta,\phi)$ the limit of the modiffiedLHS as  $r_0
\rightarrow 0$:
\begin{equation}
\label{eqn:tubeend4}
\gamma(\theta,\phi) = r_1^2 (1 +
\tau(r_1,\theta,\phi)^2\nu(r_1,\theta,\phi))(a^2v(\bx_s)v^{-1})(r_1,\theta,\phi)
\end{equation}
Since the RHS is $>0$, so is $\gamma$. Since the RHS is actually the
same as the LHS of \ref{eqn:tubeends3} with $r_0 \leftrightarrow r_1$,
there is $r^* >0$ so that
\[
a(r,\theta,\phi)\sqrt{\frac{v(\bx_s)}{v(r,\theta,\phi)}} = 
\frac{\sqrt{\gamma(\theta,\phi)}}{r} + r \tilde{b}(r,\theta,\phi), \,\tilde{b}\in C^k([0,r^*] \times
(-\pi,\pi) \times (-\pi/2,\pi/2)) 
\]
To put this relation in the form that will be useful in analyzing the remainder
term in the progressing wave expansion, Taylor-expand
$(v(\bx)v(\bx_s))^{1/2}$ to second order and amalgamate the
second-order remainder with $\tilde{b}$ above to obtain
\begin{equation}
\label{eqn:ampfinal}
a(\bx,\bx_s) = \frac{\sqrt{\gamma(\theta,\phi)}}{r} +
\frac{\sqrt{\gamma(\theta,\phi)}}{r} {\bf g}\cdot (\bx-\bx_s) + r b(r,\theta,\phi), \,b\in C^k([0,r^*] \times
(-\pi,\pi) \times (-\pi/2,\pi/2)) 
\end{equation}
where ${\bf g} = (v(\bx_s)^{-1/2}(\nabla v^{1/2})(\bx_s)$.

\section{Remainder}

The entire construction is justified by the final step, showing that
the remainder
$R$ in equation \ref{eqn:progwave} is actually less singular than the
leading term $aS(t-\tau)$, which justifies separating the two. The
remainder satisfies equation 
\ref{eqn:rem0}, which I repeat for convenience:
\begin{equation}
\label{eqn:rem}
(v^2\nabla^2 a) S(t-\tau) + \delta(\bx-\bx_s)\delta(t)
   = \frac {\partial ^2 R}{\partial t^2} -  v^2 \nabla^2 R).
\end{equation}

Introduce the {\em energy form}: for (any) function $p(\bx,t)$,
\begin{equation}
\label{eqn:energy2}
E(t) = \frac{1}{2}\int\, dx\, \left[\frac{1}{v^2(\bx)}\left(\frac{\partial p}{\partial
    t}(\bx,t)\right)^2 + |\nabla p(\bx,t)|^2\right].
\end{equation}
For solutions of the wave equation \ref{eqn:raycdawe}, the {\em energy
  identity} holds:
\begin{equation}
\label{eqn:energy2id}
E(t_1)-E(t_0) = \int_{t_0}^{t_1}\,dt\,\int \,dx\, \frac{1}{v^2(\bx)}\frac{\partial p}{\partial
    t}(\bx,t)f(\bx,t).
\end{equation}
To see this, differntiate the definition \ref{eqn:energy2} with respect to time. Passing the time derivative under the integral sign, obtain
\[
\frac{dE}{dt}(t) = \int\, dx\, \left[\frac{1}{v^2(\bx)}\left(\frac{\partial p}{\partial
    t}\frac{\partial^2 p}{\partial
    t^2}(\bx,t)\right) + \nabla p(\bx,t) \cdot \nabla \frac{\partial p}{\partial
    t}\right]
\]
\[
=\int\, dx\, \frac{\partial p}{\partial
    t}(\bx,t)\left(\frac{1}{v^2(\bx)}\frac{\partial^2 p}{\partial
    t^2}(\bx,t) - \nabla^2 p(\bx,t) \right)
\]
after integrating by parts to shift $\nabla$ (this step assumes that $p=0$ for large $|\bx|$).
Use the wave equation satisfied by $p$ to substitute $f$ for the factor in parantheses, and so obtain 
\begin{equation}
\label{eqn:energy2idt}
= \int \,dx\, \frac{1}{v^2(\bx)}\frac{\partial p}{\partial
    t}(\bx,t)f(\bx,t).
\end{equation}
Integrate \ref{eqn:energy2idt} in time from $t_0$ to $t_1$ to obtain \ref{eqn:energy2id}.

The Cauchy-Schwarz inequatity and the observation that the square of
$\partial p/\partial t$ is part of the energy form \ref{eqn:energy2} leads to the
inequality
\begin{equation}
\label{eqn:energy2ineq}
E(t_1)-E(t_0) \le C
\left(\int_{t_0}^{t_1}\,dt\,E(t)\right)^{\frac{1}{2}} \|f\|_{[t_0,t_1]}
\end{equation}
in which $\|\cdot\|$ denotes the space-time $L^2$ norm:
\begin{equation}
\label{eqn:L2}
\|f\|_{[t_0,t_1]} = \left(\int_{t_0}^{t_1}\,dt\,\int \,dx\, f(\bx,t)^2
  \right)^{\frac{1}{2}},
\end{equation}
and $C>0$ depends on the minimum of $v$. Using the famous inequality
$ab \le \frac{\alpha}{2}a^2 + \frac{1}{2\alpha}b^2$ with proper choice
of $\alpha>0$, inequality \ref{eqn:energy2ineq} implies in turn that
\begin{equation}
\label{eqn:energy2gron}
E(t_1)-E(t_0) \le 
\int_{t_0}^{t_1}\,dt\,E(t) + C \|f\|_{[t_0,t_1]}^2
\end{equation}
This integral inequality can be solved (a fact that goes under the
name ``Gronwall's inequality''): the result is a bound on the growth
of $E$ in terms of the 
\begin{equation}
\label{eqn:energy2final}
E(t_1)-E(t_0) \le Ce^{t_1-t_0}((t_1-t_0)E(t_0) + \|f\|_{[t_0,t_1]}).
\end{equation}

Apply this bound to $R_1$. From equation \ref{eqn:rem1}, for this case
$f = (v^2\nabla^2 a) H(t-\tau)$, which is square-integrable (but see
remark below), take $t_0 \le 0$ so that $E(t_0) =0$. Note that
$\partial R_1/\partial t_0 = R$ is part of $E$, so the specialization
of inequality \ref{eqn:energy2final} to this case leads to
\begin{equation}
\label{eqn:rem2}
\|R\|_{[0,t]} \le Ce^t
\end{equation}
in which $C$ depends on the transport coefficient $a$ and the
traveltime $\tau$ (which determines the size of the set $\{(\bx,t):
\tau(\bx,\bx_s) \le t\}$).

Continuing as in the last section, I will discuss explicitly the 3D
case, $S = \delta$ per \ref{eqn:green3d}. In order to apply the energy
estimates just explained to $R$, the main requirement that the
the inhomogeneous term in \ref{eqn:rem} be square-integrable. That is
obviously not the case, no matter what the properties of $a$, because
$\delta(t-\tau)$ is not. So the first step is to modify \ref{eqn:rem} to
correct this deficiency.

Recall that the Green's function defined here is the causal one,
therefore zero for t<0. Integrating both sides of equation
\ref{eqn:rem} in $t$ from $-\infty$ to $t$, obtain for 
\begin{equation}
\label{eqn:R1def}
R_{1}(\bx,t) = \int_{-\infty}^t dt' R(\bx,t')
\end{equation}
\begin{equation}
\label{eqn:rem1}
-(v^2\nabla^2 a) H(t-\tau) + \delta(\bx-\bx_S)H(t-\tau)
   = \frac {\partial ^2 R_1}{\partial t^2} -  v^2 \nabla^2 R_1) \;. 
\end{equation}
with $H$ being the Heaviside function. Square-integrability of the RHS
clearly depends on the properties of $a$ near $\bx=\bx_s$, given by
equaiton \ref{eqn:ampfinal}.

From \ref{eqn:ampfinal}, $a$ is the sum of three terms. The
Laplacian of  the first term is

\[
\sqrt{\gamma(\theta,\phi)}  \frac{\partial^2}{\partial r^2} + \frac{2}{r}\frac{\partial}{\partial 
  r} \frac{1}{r} + \frac{1}{r^3} \Delta_S\sqrt{\gamma(\theta,\phi)} 
\]
in which 
\[
\Delta_S = \frac{1}{\cos^2 \phi}\frac{\partial^2}{\partial \theta^2} +
\frac{1}{\cos \phi} \frac{\partial}{\partial \phi}\cos \phi 
\frac{\partial}{\partial \phi}
\]
is the spherical Laplacian. In order for the second term to define a
disribution of order zero (like $\delta(\bx-\bx_s)$), the coefficient
of $1/r^3$ must vanish, implying that $\gamma$ is constant, and the
second term drops out. The first
term is the same as 
\[
\sqrt{\gamma} \nabla^2 {1}{r} = \gamma 4 \pi \delta(\bx-\bx_s)
\]
So if we chose 
\[
\gamma = \frac{1}{4 \pi v^2(\bx_s)}
\]
then the Laplacian of the first term in \ref{eqn:ampfinal} cancels the
space-time delta in \ref{eqn:rem}.

Assume for the moment that the second term in \ref{eqn:ampfinal}
vanishes, that is, $\nabla v$ vanishes at $\bx=\bx_s$. The Laplacian
of the third term is
\[
\left(\frac{\partial^2}{\partial r^2} + \frac{2}{r}\frac{\partial}{\partial 
  r} + \frac{1}{r^2 \cos^2 \phi}\frac{\partial^2}{\partial \theta^2} +
\frac{1}{r^2 \cos \phi} \frac{\partial}{\partial \phi}\cos \phi 
\frac{\partial}{\partial \phi}\right) r b(r,\theta,\phi)
\]
\[
\frac{1}{r}h(r,\theta,\phi), \, h \in C^0([0,r^*] \times (-\pi,\pi)
\times (-\pi/2,\pi/2))
\]
The singularity is square-integrable in 3D, and remains so in 4D after
multiplication by $H(t-\tau)$. That is, equation \ref{eqn:rem1}
reduces to 
\begin{equation}
\label{eqn:rem1r}
\frac{1}{r}h(r,\theta,\phi) H(t-\tau)
   = \frac {\partial ^2 R_1}{\partial t^2} -  v^2 \nabla^2 R_1) \;. 
\end{equation}
with a square-integrable left-hand side. The energy estimate
\ref{eqn:energy2final} applies: the first derivatives of $R_1$ are
square-integrable. In particular, $R = \partial R_1/\partial t$ is
square integrable, justifying the progressing wave expansion in this
case.

For the general case in which $\nabla v$ does not vanish at
$\bx=\bx_s$, the left-hand side of \ref{eqn:rem} has the additional summand
\begin{equation}
\label{eqn:term2}
-v^2(\bx)\nabla^2\frac{1}{4\pi v^2(\bx_s) r} {\bf g}\cdot (\bx-\bx_s) 
= v^2(\bx)\frac{1}{4\pi v^2(\bx_s)}\frac{{\bf g}\cdot(\bx-\bx_s)}{r^3} = O(r^-2).
\end{equation}
This function is not square-integrable, so the energy argument does
not apply directly.

\section{Suggested Projects}

\begin{enumerate}
\item In the ray theory section, I used the idea that the traveltime
 should be asymptotic to the constant velocity
  specializations as $\bx \rightarrow \bx_s$. I made a similar
  implicit argument for the amplitude in the following
  section. However both traveltime and amplitude are singular at
  $\bx=\bx_s$. Supply the missing justification.

\item Singularity is measured by the rate of growth/decay of the
  Fourier transform at infinite frequency. Recast the progressing wave
  expansion in terms of frequency:
\[
u(\bx,\omega) = a(\bx)e^{i\omega (t-\tau(\bx))}  +
\frac{1}{i\omega}R(\bx,\omega)
\] 
with $R$ bounded as $\omega \rightarrow \infty$. A classic reference
from this point of view is \cite{Lax:57}.

\item What really happens at a caustic? The actual behaviour of the
  wavefield was first understood in the 1960's. Major contributors,
  whose papers are well worth a read, are ~\cite{Lud:66} and
~\cite{Krav:68}, and \cite{StickAhl:81} provide a very readable
account.

\item Having worked that out, create an example like the low-velocity
  lens (Figure \ref{fig:lensmedium}) and compute the wavefield with a
  finite difference code, for example IWave. What actually happens to
  the wavefield in the vicinity of the caustic shown in Figure
  \ref{fig:lensrayswavefronts}? Why? How is this consistent with
  theory?

\item Work out all the missing details in the argument presented in
  the last section, and in particular answer the question posed in the
  last paragraph.
\end{enumerate}

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/caam641/stack/paper.tex
repro docs dir added

\title{Post-stack Migration}
\maketitle
\label{ch:post}
\inputdir{project}

\section{Introduction}
The term ``migration'', as used in exploration geophysics, has a complicated origin. The reader should consult some of the standard texts on the subject, such as \cite{Dob:88}, or \cite{SheriffGeldart:1995}, for an account of this history. For these notes, the term ``migration'' is a synonym for ``imaging'' or ``image formation'', that is, a process of creating a subsurface image from data.

This chapter gives an idiosyncratic overview of post-stack migration, that is to say, creation of an image from the stack. The reader has already seen examples in Chapter \ref{ch:basic}. The term ``post-stack migration'' refers to techniques that convert zero-offset data to subsurface images. Since zero-offset data is seldom acquired, it is important that the stack and zero-offset data are roughly equivalent. I will explain this equivalence, and derive the {\em exploding reflector} model of zero-offset data. In terms of this model, I will pose a simple linear inverse problem for recovery of the square velocity (or bulk modulus) perturbation from the stacked data, by minimizing the mean square misfit between the stack and its exploding reflector prediction. From the various approximations and replacements necessary to get to this point, it will be clear that the field extracted from the data by solving this inverse problem cannot be expected to represent the actual values of a physical parameter - but perhaps can be expected to show the same geometric structure, that is, to be an {\em image} of the subsurface. Like all remotely reasonable formulations of the seismic inverse problem, this one is so large computationally that only iterative solution methods are remotely feasible. I introduce the {\em Conjugate Gradient} algorithm for iterative solution of linear systems, and apply it to the exploding reflector inverse problem. The first iterate, a multiple of the gradient of the mean squares error, is already a reasonably good image. I will explain why this is to be expected. Computation of the gradient is essentially equivalent to applying the {\em adjoint} or transpose of the simulation operator. I introduce the {\em adjoint state method} for computation of this adjoint, which is a verion of {\em reverse time migaration}, introduced into seismology in the early 80's and a key ingredient in many advanced processing methods.

\section{Zero-Offset Modeling}
Recall that in case of simple ray geometry (unique rays connecting sources and receivers to scattering points), geometric optics approximates the Born approximation, equation \ref{eqn:grt1}, as:
\[
D\cF[v^2]\delta v^2(\bx_r,t;\bx_s) \approx 
\]
\begin{equation}
\label{eqn:grt1r}
= w*_t \frac{\partial^2}{\partial t^2}\int \, dx \, a(\bx_r,\bx)a(\bx_s,\bx)\delta(t-\tau(\bx_r,\bx)-\tau(\bx_s,\bx)) \frac{\delta v^2}{v^2}(\bx)
\end{equation}
``Zero offset'' means that $\bx_s = \bx_r=\bx_m$. So the zero-offset trace at receiver location $\bx_r$ is
\[
D\cF_{\rm ZO}[v^2]\delta v^2(\bx_r,t) \approx 
\]
\begin{equation}
\label{eqn:grt1r1}
= w*_t \frac{\partial^2}{\partial t^2}\int \, dx \, a^2(\bx_r,\bx)\delta(t-2\tau(\bx_r,\bx)) \frac{\delta v^2}{v^2}(\bx) 
\end{equation}
Recall that $\tau$ is a solution of the eikonal equation \ref{eqn:eik}, written concisely as $v(\bx)|\nabla \tau(\bx,\bx_r)|=1$. Obviously $2\tau(\bx,\bx_r)$ solves the same eikonal equation with the velocity $ v/2$: ``if you go half as fast, it takes twice as long to get there''.

I claim that in fact the quantity $a \delta(t-2\tau)$ is the most singular (high frequency asymptotic) part of the Green's function $G_{v/2}(\bx,t;\bx_r)$ for the velocity $v/2$, with ``source'' at $\bx_r$, up to a factor of 2. Specifically, the relationship is
\begin{equation}
\label{eqn:2green}
G_{v/2}(\bx,t;\bx_r) = \frac{1}{2}G_v\left(\bx,\frac{t}{2};\bx_r\right).
\end{equation}
To see this, substitute the right-hand side of \ref{eqn:2green} in the wave equation for velocity field $v/2$:
\[
\left(\frac{1}{\left(\frac{v}{2}(\bx)\right)^2} \frac{\partial^2}{\partial t^2} - \nabla^2\right) \frac{1}{2}G_v\left(\bx,\frac{t}{2};\bx_r\right)
\]
\[
=\left(\frac{4}{v(\bx)^2}\frac{\partial^2}{\partial t^2} - \nabla^2\right)\frac{1}{2}G_v\left(\bx,\frac{t}{2};\bx_r\right)
\]
\[
= \frac{1}{2}\left(\frac{1}{v(\bx)^2}\frac{\partial^2G_v}{\partial t^2} - \nabla^2G_v\right)\left(\bx,\frac{t}{2};\bx_r\right)
\]
\begin{equation}
\label{eqn:2greenderiv}
=\frac{1}{2}\delta(\bx-\bx_r)\delta\left(\frac{t}{2}\right) = \delta(\bx-\bx_r)\delta(t).
\end{equation}
The last step in the sequence of equations \ref{eqn:2greenderiv} is the identity $\delta(t) = c\delta(ct)$, which follows from the definition of the delta function and the substitution rule for integrals.

Equation \ref{eqn:2greenderiv} shows that the right-hand side of \ref{eqn:2green} solves the defining wave equation of the Green's function $G_{v/2}$. Since both sides are causal, \ref{eqn:2green} is established.

The geometric optics computation from Chapter 5 shows that the leading singularity of $G_{v/2}$ takes the form
\begin{equation}
\label{eqn:gohalf1}
G_{v/2}(\bx,t;\bx_r) \approx a_{v/2}(\bx,\bx_r)\delta(t-2\tau(\bx,\bx_r)) = \frac{1}{2}a_{v/2}(\bx,\bx_r)\delta\left(\frac{t}{2}-\tau(\bx,\bx_r)\right)
\end{equation}
However the same geometric optics computation shows that
\begin{equation}
\label{eqn:gohalf2}
\frac{1}{2}G_v\left(\bx,\frac{t}{2};\bx_r\right) \approx \frac{1}{2}a_{v}(\bx,\bx_r)\delta\left(\frac{t}{2}-\tau(\bx,\bx_r)\right)
 =a_v(\bx,\bx_r)\delta(t-2\tau(\bx,\bx_r)) 
\end{equation}
Combining equations \ref{eqn:2green}, \ref{eqn:gohalf1}, and \ref{eqn:gohalf2}, conclude that $a_{v/2} = a_v$, and that the integrand in the RHS of equation \ref{eqn:grt1r1} contains exactly the leading order approximation to the Green's function $G_{v/2}$:
\[
D\cF_{\rm ZO}[v^2]\delta v^2(\bx_r,t) \approx  w*_t \frac{\partial^2}{\partial t^2}\int \, dx \,a_v(\bx,\bx_r) G_{v/2}(\bx,t;\bx_r) \frac{\delta v^2}{v^2}(\bx) 
\]
\[
= w*_t \frac{\partial^2}{\partial t^2}\int\, dt'\,\int \, dx \, a(\bx,\bx_r)G_{v/2}(\bx_r,t-t';\bx) \frac{\delta v^2}{v^2}(\bx) 
\delta(t')
\]
\begin{equation}
\label{eqn:grt1g}
= \int\, dt'\,\int \, dx \, a(\bx,\bx_r)G_{v/2}(\bx_r,t-t';\bx) \frac{\delta v^2}{v^2}(\bx)\frac{d^2w}{dt^2}(t').
\end{equation}

Equation \ref{eqn:grt1g} unfortunately does {\em not} expresses the zero-offset data gather as sampling the solution of an acoustic wave equation, because of the extra amplitude factor under the integral sign. 
The very creative step from zero-offset to exploding reflector modeling consists dropping $a$ and the factor $1/v^2\partial^2/\partial t^2$. Both are $a$ and $1/v^2$ positive and (under standing assumptions) slowly varying on the wavelength scale, so that the overall effect ought to be scaling oscillatory reflection events without modifying their local relative amplitudes or locations. The second time derivative modifies the phase also without modifying locations and orientations of events, and combines with the other factors to yield an overall dimensionless change. A complete mathematical justification requires tools developed later in these notes.

Accepting these changes, re-interpret zero-offset data as represented by the {\em exploding reflector model}
\begin{equation}
\label{eqn:grt2g}
D\cF_{\rm ZO}[v^2]\delta v^2(\bx_r,t) \approx  \int\, dt'\,\int \, dx \, G_{v/2}(\bx_r,t-t';\bx) \delta v^2(\bx)w(t').
\end{equation}
The Green's function $G_{v/2}$ is the solution operator of the initial value problem
\begin{eqnarray}
\label{eqn:cdawe7}
\frac{\partial^2 p}{\partial t^2} - \frac{v^2}{4} \nabla^2 p & = & f \nonumber\\
p & = & 0, \,t<<0
\end{eqnarray}
$D\cF_{\rm ZO}$ applies this solution operator to the right-hand side 
\begin{equation}
\label{eqn:explref}
f(\bx,t) = \frac{\delta v^2}(\bx) w(t) 
\end{equation}
This observation explains the term {\em exploding reflector model} for the representation \ref{eqn:grt1g}: the reflectors ($\delta v^2$) ``explode'' with time dependence $w(t)$.

\section{Exploding Reflectors in Pressure-Velocity Acoustics}
The pressure-velocity form of acoustics relates pressure $p$ and the particle velocity vector field $\bv$ through a pair of time-dependent PDEs: 
\begin{eqnarray}
\label{eqn:vdawe}
\frac{\partial p}{\partial t}&=& -\kappa (\nabla \cdot \bv + g) \nonumber \\
\frac{\partial \bv}{\partial t} &=& -\beta \nabla p
\end{eqnarray}
In this system, $\kappa = \rho v^2$ where $\rho$ is the material density, and $\beta = 1/\rho$ is the {\em buoyancy}.

Each equation in \ref{eqn:vdawe} has clear physical meaning. The first is the constituitve law of linear acoustics: the rate of change of pressure is proportional to the rat of change of volume (the divergence of the velocity field), with the constant of proportionality being the bulk modulus. The the energy source is represented as a defect ($g$) in this constitutive law: more specifically, a defect in the infinitesimal volume rate field. The second equation is Newton's law.

To make this system equivalent to the constant density second order equation \ref{eqn:cdawe7}, simply assume $\beta=\rho=1$ and set 
\[
\bv = \nabla \int_{-\infty}^t p,\,\,\kappa \frac{\partial g}{\partial t} = f.
\]

In terms of the pressure-velocity form of acoustics, the exploding reflector model from the previous section is expressed as
\[
D\cF_{\rm ZO}[v^2]\delta v^2(\bx_r,t) \approx p(x_r,t),
\]
where $(p,\bv)$ solve the system \ref{eqn:vdawe} with $\kappa= v^2/4$, that is, the physical bulk modulus is divided by 4, and the right-hand-side
\begin{equation}
\label{eqn:explrefderiv}
g(\bx,t) = \frac{\delta v^2(\bx)}{v^2(\bx)}w_1(t). 
\end{equation}
where $w_1$ is the indefinite integral of $w$.

At this point, I will take possibly unfair advantage of a point I have not raised until now - we have actually no knowledge of the wavelet $w(t)$. Of course, whatever it is (if indeed the isotropic point radiator is even a fair model of the seismic source) it can be changed by filtering. Therefore I will from now on assume that $w$ is the derivative of an approximate delta, filtered appropriately and with appropriate units, and therefore that $w_1$ is a filtered delta function. For convenience, I will even ignore the filter. Therefore the model of the right-hand side in the exploding reflector model becomes
\begin{equation}
\label{eqn:explrefdelta}
g(\bx,t) = \frac{\delta v^2(\bx)}{v^2(\bx)}\delta(t). 
\end{equation}

\section{An Example}

In Chapter \ref{ch:born}, I used the stack as input to various forms of migration to produce images in both time and depth, without explaining why the stack is appropriate input (or, indeed, what the migration algorithms did). This chapter will clear up some of those mysteries. However, as I have done in earlier chapters, I will begin by building an example based on migration of a stack.

I construct an example similar to those in Chapter \ref{ch:born}, by using the $v^2$ model estimated by flattening NMO-corrected gathers for the Viking Graben data (Figure \ref{fig:paracsqer}), and for the perturbation $\delta v^2$ the PSPI post-stack depth migrated image Figure \ref{fig:parapspifixagcfilt}. The input data for the PSPI image is shown in Figure \ref{fig:parastackfixagcfilt}: it is the result of applying AGC to the stack of the NMO-corrected data from Chapter \ref{ch:born}, then applying a (2,5,20,25) Hz bandpass filter. The additinoal low-pass filter allows accurate finite difference calculations with a coarser grid than is possible with the original data, and is used only to make the computations a bit less time-consuming.

\plot{paracsqer}{width=0.9\textwidth}{Scaled bulk modulus $v^2/4$ ($\rho=1$) for exploding reflector simulation: obtained from Viking Graben data by flattening NMO gathers, as in Chapter \ref{ch:basic}.}

\plot{parastackfixagcfilt}{width=0.9\textwidth}{AGC'd and filtered stack from Chapter \ref{ch:born}. Filter is (2, 5, 20, 25) Hz bandpass.}

\plot{parapspifixagcfilt}{width=0.9\textwidth}{Bulk modulus perturbation $\delta v^2$ for exploding reflector simulation, obtained from Viking Graben data by Gazdag PSPI depth migration of filtered stack (Figure \ref{fig:parastackfixagcfilt}) using velocity model of Figure \ref{fig:paracsqer}.}

The Viking Graben survey does not contain zero-offset data. The closest approximation is the near offset, at $h=-262$ m. With AGC to make the events more visible, this gather appears as Figure \ref{fig:paracdpnearoffagc}. Comparison with the result of the exploding reflector simulation, Figure \ref{fig:expldatanmo}, shows general agreement of positioning and orientation of events. I have plotted both field data and simulated data after AGC, to emphasize the similarity in apparent structure.

\plot{paracdpnearoffagc}{width=0.9\textwidth}{Near offset section, after AGC.}
\plot{expldata}{width=0.9\textwidth}{Exploding reflector simulated zero offset data based on PSPI depth image (Figure \ref{fig:parastackfixagcfilt}) as input reflectivity (right-hand side in equation \ref{eqn:explrefdelta}, and scaled NMO-derived velocity model (Figure \ref{fig:paracsqer}).}

\section{The Stack As a Stand-In for ZO Data}
Recall from the last chapter that the convolutional model predicts a simple relation between perfectly deconvolved CMP gathers $d(x_m,y_m,h,t)$ after NMO correction using event traveltime $T(t_0,h)$  as function of zero offset time $t_0$ and (half) offset $h$, and a reflectivity function $R(t_0)$:
\begin{equation}
\label{eqn:convt0rep}
d(x_m,y_m,h,\bar{T}(h,t_0)) \approx \bar{A}(h,t_0) \frac{d\bar{R}}{dt_0}(x_m,y_m,t_0) 
\end{equation}
SInce $\bar{T}(0,t_0)=t_0$, this relation implies that for zero offset,
\begin{equation}
\label{eqn:convt0h0}
d(x_m,y_m,0,t_0) \approx \bar{A}(0,t_0) \frac{d\bar{R}}{dt_0}(x_m,y_m,t_0). 
\end{equation}
It also yields an approximation to the stack:
\begin{equation}
\label{eqn:convstack}
d_{\rm stack}(x_m,y_m,t_0) = \int \, dh \,d(x_m,y_m,h,\bar{T}(h,t_0)) \approx \bar{A}_{\rm stack}(t_0) \frac{d\bar{R}}{dt_0}(x_m,y_m,t_0). 
\end{equation}
where the stacked amplitude is defined by
\begin{equation}
\label{eqn:stackamp}
\bar{A}_{\rm stack}(t_0) = \int \, dh \,\bar{A}(h,t_0) .
\end{equation}
Comparison of \ref{eqn:convt0h0} and \ref{eqn:convstack} shows that the stack is proportional to zero-offset data by the ratio of the zero-offset and stacked amplitudes, both of which are positive slow-varying functions of $t_0$. Thus if the deconvolved convolutional model \ref{eqn:convt0rep} is to be believed, the stack has precisely the same {\em locations in time} of high-frequency events as zero-offset data, so that as input to imaging for structure, the two are interchangeable.

Of course, the deconvolved convolutional model is not entirely correct, as discussed in the last chapter, for (at least) three reasons. First, since perfect deconvolution is impossible, NMO stretch will decrease the frequency content of far offsets, hence lower decrease the energy content of the stack at high frequencies, hence its apparent resolution. Second, imperfectly defined NMO velocity will result in some averaging of the reflectivity, again depressing frequency content. Both of these effects are observable in the stacked data, shown in Figure \ref{fig:parastackfixagc} (also with AGC, but without low-pass filter) for comparison with the near-offset section (Figure \ref{fig:paracdpnearoffagc2}, repeated here for the reader's convenience).

\plot{paracdpnearoffagc2}{width=0.9\textwidth}{Near offset section, after AGC.}
\plot{parastackfixagc}{width=0.9\textwidth}{Stack of NMO-corrected Viking Graben data, AGC applied.}

Finally, of course the convolutional model in any form is the result of several layers of approximation - linearization, high-frequency asymptotics, the ``locally layered'' assumption. Nontheless, in fact the stack is even more similar to the modeled exploding reflector data (Figure \ref{fig:expldataagc}) than is the zero offset section.

\section{A Linear Inverse Problem}
Seismic imaging is the construction of an image of the subsurface from seismic data - a statement which begs the question, ``what is an image?''. [One could give the Justice Potter Stewart answer, of course - ``I know it when I see it'' - but it's possible to do better.] To start with, I will pose the imaging problem in the simplest possible way: given the data, recover parameters that predict it. For the constant density acoustic model, ``parameters'' means the squared velocity field $v^2(\bx)$, and in principle the source model as well. Admitting the Born approximation, ``parameters'' means the source and the pair of squared velocity $v^2(\bx)$ and its perturbation $\delta v^2(\bx)$, with the former slowly varying on the wavelength scale hence transparent thanks to geometric optics, and the latter oscillatory and accounting for reflection dynamics. Recovery of these parameter fields must surely encompass producing their image, and therefore an image of earth structure, to the extent that the modeling assumptions underlying the formulation of the problem are valid.

I will specialize this parameter recovery or {\em inverse} problem to the zero-offset (or stacked) data case with the exploding reflector approximation replacing the acoustic model. As it turns out, this combination of data and model does not permit all parameters to be determined. If the source model is fixed as in equation \ref{eqn:explrefdelta}, and the squared velocity is regarded as data, rather than as a parameter field to be determined, it turns out to be possible to estimate the remaining parameter, the square velocity perturbation $\delta v^2(\bx)$, so that the data is quite well predicted. The predicted exploding reflector data is linear in $\delta v^2(\bx)$, as the latter is essentially the right-hand side of a wave equation. Therefore the recovery of $\delta v^2(\bx)$ from $v^(\bx)$ and $d_{\rm stack}(x_m,y_m,t_0)$ is a {\em linear inverse problem}. Formally, the problem amounts to the (approximate) solution of the linear system
\begin{equation}
\label{eqn:psinv}
D\cF_{\rm ZO}[v^2]\delta v^2(x_m,y_m,t_0) \approx  d_{\rm stack}(x_m,y_m,t_0). 
\end{equation}

Because of the very many approximations and unrealistic physical assumptions made in arriving at the system \ref{eqn:psinv}, actually solving it (even in the approximate sense that I will develop in this chapter) is overkill - asking too much of the data and modeling assumptions. The solution can't be confused with the ``real'' short-scale perturbations in compressional wave velocity occuring in the subsurface beneath the Norwegian North Sea. Nonetheless, one might hope first that some evidence of subsurface structure would emerge, and perhaps to see how the calculations might be short-circuited to provide just such image information without actually bothering to fit the data. Both hopes turn out to be realistic.

It is very likely that equation \ref{eqn:psinv} cannot be solved by any choice of $\delta v^2$, that is, the data cannot be exactly fit, for example because the domain within which $\delta v^2$ is allowed to vary does not predict the full range of midpoints or times present in the stack. An example of exactly this type of behaviour appears below. Therefore it is natural to converte \ref{eqn:psinv} to a {\em best fit} problem. The mean square error is a misfit measure with pleasant mathematical and computational properties. It also enjoys some physical justification, see \cite{SantosaSymes:00}. Thus seek $\delta v^2$ to minimize (approximately)
\begin{equation}
\label{eqn:psinvls}
J[\delta v^2] =\frac{1}{2} \sum_{x_m,y_m}\int \,dt_0\,|(D\cF_{\rm ZO}[v^2]\delta v^2)(x_m,y_m,t_0) -  d_{\rm stack}(x_m,y_m,t_0)|^2. 
\end{equation}
Since $J$ is a positive semidefinite quadratic form, all of its stationary points are global minimizers. A stationary point is a zero of the gradient:
\begin{equation}
\label{eqn:psinvgrad}
\nabla J[\delta v^2] = D\cF_{\rm ZO}[v^2]^T( D\cF_{\rm ZO}[v^2]\delta v^2 -  d_{\rm stack})(x_m,y_m,t_0)) = 0
\end{equation}
or alternatively in the form of a linear system, the so-called {\em normal equation}:
\begin{equation}
\label{eqn:psinvnorm}
(D\cF_{\rm ZO}[v^2]^T D\cF_{\rm ZO}[v^2]\delta v^2) (x,y,z) = D\cF_{\rm ZO}[v^2]^T d_{\rm stack})(x,y,z).
\end{equation}
The transpose, or adjoint, operator $D\cF_{\rm ZO}[v^2]^T$ is the unique operator making the {\em dot product test} true:
\[
\sum_{x_m,y_m}\int \,dt_0\,(D\cF_{\rm ZO}[v^2]\delta v^2)(x_m,y_m,t_0)d_{\rm stack}(x_m,y_m,t_0) 
\]
\begin{equation}
\label{eqn:dotprodtest}
=\int\int\int \,dx\,dy\,dz\,(D\cF_{\rm ZO}[v^2]^T d_{\rm stack})(x,y,z)\delta v^2(x,y,z).
\end{equation}
Note that the dot product test should hold for {\em any} choice of square velocity perturbation $\delta v^2$ and stacked data section $d_{\rm stack}$, not just the ones that you might be currently interested in.

After discretization, say by finite differences as are used in the IWAVE package, the fields $\delta v^2$ and $d_{\rm stack}$ become vectors, and the linear operators $D\cF_{\rm ZO}[v^2]$, $D\cF_{\rm ZO}[v^2]^T$, and $D\cF_{\rm ZO}[v^2]^TD\cF_{\rm ZO}[v^2]$ become matrices. 
The matrix of $D\cF_{\rm ZO}[v^2]^T$ is the transpose of the matrix of $D\cF_{\rm ZO}[v^2]$, and the matrix of the {\em normal operator} $D\cF_{\rm ZO}[v^2]^TD\cF_{\rm ZO}[v^2]$ is symmetric positive (semi-)definite.

The integrals in the equation \ref{eqn:dotprodtest} become scaled versions of the dot product, scaled by the cell volume on each side. On the left, I have continued to write a sum over midpoints, rather than integrals over midpoint coordinates, because midpoint coordinates may not be uniformly sampled and devising a quadrature method is difficult. Instead, I presume that the midpoint sampling is sufficiently uniform that I can treat it as perfectly uniform and ignore the midpoint area factor that would go into an approximation to an integral. So the only cell volume on the left hand side is the time step. On the other hand, the spatial fields - $\delta v$ and the output of the transpose operator - are sampled on a regular finite difference grid, so the integral is well-approximated by the ordinary dot product multiplied by the volume of a grid cell. The consequence is the the results are stable against resampling in $t$ and in $x,y,z$.

The inversion task boils down to solving (approximately, at least) the normal system \ref{eqn:psinvnorm}. A first impediment to computational implementation is the size of this system. For the example created in this book from the Viking Graben survey, $d_{\rm stack}$ has 1.6 $\times 10^6$ samples, and $\delta v^2$ about 0.5 $\times 10^6$. That is, there are more equations than unknowns, by about a factor of 3: the system \ref{eqn:psinv} is overdetermined, and is very unlikely to have a literal solution, justifying the least squares approach. Storing the matrix of $D\cF_{\rm ZO}[v^2]$ requires $O(10^{12})$ words of memory - not inconceivable for vintage 2017 high-end computing equipment, but a LOT of storage nonetheless. The normal marix is smaller, $O(10^{11})$ words, especially if one stores only the upper or lower triangle, taking advantage of symmetry. It is still an unpleasantly large amount of data, and the vast number of loads and stores necessary to manipulate it make efficient implementation difficult. Finally, Gaussian elimination requires $O(10^{18})$ floating point operations, or more than 100 days at 100 Gflops - and this for a rather academic 2D problem.

\section{Conjugate Gradient Iteration}
Remarkably, it is possible to solve the least squares problem described in the last section (effectively, the normal equations \ref{eqn:psinvnorm}) to an acceptable level of approximation, by means of an efficient {\em iterative} method, in a tiny fraction of the time estimated above for Gaussian elimination, and using very modest computing resources - for the examples to be shown in the next section, a few minutes on a vintage 2015 Apple laptop with a single thread of execution.

The efficient iterative method is Conjugate Gradient (CG) iteration \cite[]{NocedalWright,Golub:2012}. For completeness, I describe the general form of this computation, in the context of the least squares problem defined in the last section. To make the notation tractable, I introduce the two abbreviations:
\begin{eqnarray}
\label{eqn:Fdef}
F & \leftarrow & D\cF_{\rm ZO}[v^2]\nonumber \\
r & \leftarrow & \frac{\delta v^2}{v^2}
\end{eqnarray}
That is, $v^2$ is fixed for the course of the discussion, so suppress it from the notation, and abbreviate the notation for derivative. Also, the relative perturbation in $v^2$ is the source of reflection signal in the exploding reflector model: accordingly, abbreviate it as the reflectivity, denoted $r$, and regard the predicted signal as being a function of the relative perturbation rather than the perturbation itself - since $v^2$ is regarded as fixed, the two points of view are equivalent. Thus $r \mapsto Fr$ is a linear map, which would be represented after discretization by a (huge!) matrix if you were to be so foolish as to compute all of its entries. Instead, a perfectly good approach to computing the matrix-vector product $F r$ is available: use a finite difference method to approximately solve the exploding reflector problem \ref{eqn:vdawe}, \ref{eqn:explrefderiv}, which in the notation just introduced take the form
\begin{eqnarray}
\label{eqn:vdawer}
\frac{\partial p}{\partial t}&=& -\kappa (\nabla \cdot \bv - r \delta(t)) \nonumber \\
\frac{\partial \bv}{\partial t} &=& -\beta \nabla p\nonumber \\
p,\bv & = & 0, \,t<0 
\end{eqnarray}

The CG algorithm needs only the matrix-vector product, not the matrix entries themselves:
\begin{itemize}
\item Initialize:
\begin{itemize}
 \item $r_0 = 0$
  \item $e_0 = d$
  \item $s_0 = F^Td$
\item $g_0 = s_0$
  \item $k = 0$
\end{itemize}
\item Repeat:
\begin{itemize}
  \item $\alpha_k = \frac{\langle s_k,s_k \rangle}{\langle Fs_k,Fs_k\rangle}$
  \item $r_{k+1} = r_k + \alpha_k s_k$
 \item $e_{k+1} = e_k - \alpha_kFs_k$
\item $g_{k+1} = g_k - \alpha_k F^TFs_k$
  \item $\beta_{k+1} = \frac{\langle g_{k+1},g_{k+1}\rangle}{\langle g_k,g_k\rangle}$
  \item $s_{k+1}= g_{k+1}+\beta_{k+1}s_k$
  \item $k = k+1$
\end{itemize}
\item Until $\|e_k\|$ sufficiently small, or max iteration count exceeded 
\end{itemize}

These relations follow from the statement of the algorithm:
\begin{eqnarray}
\label{eqn:cgcons}
e_k & = & d - Fr_k\nonumber \\
g_k&=& F^T(d-Fr_k)
\end{eqnarray}
That is, $e_k$ is the data error (or residual) at the $k$th iteration, and $g_k$ is the gradient (also the normal residual). $s_k$ is the search direction, $\alpha_k$ is the correct multiple of $s_k$ to attain the minimum of the quadratic objective on the line through $r_k$ in the direction $s_k$. The justification for the direction update parameter ($\beta_k$) is more difficult to explain - see the cited references.

The corner brackets are the scaled dot products mentioned in the previous section: for instance,
 for data sections $d_1$ and $d_2$,
\begin{equation}
\label{eqn:dataip}
\langle d_1, d_2 \rangle = \sum_{x_m,y_m}  \sum_{t_0}\Delta t_0\,d_1(x_m,y_m,t_0) d_2(x_m,y_m,t_0) 
\end{equation}
and for depth fields $g_1(x,y,z), g_2(x,y,z)$
\begin{equation}
\label{eqn:modelip}
\langle g_1, g_2 \rangle = \sum_{x,y,z}  \Delta x \Delta y \Delta z g_1(x,y,z) g_2(x,y,z) 
\end{equation}
For any inner product, the corresponding norm is the square root of the inner product ssquare:
$\|g\|=\sqrt{\langle g,g\rangle}$.

Conjugate Gradient iteration has many fascinating properties, explained in the books cited earlier. I will mention just one: in its application to least squares problems as in the algorithm explained above, the residual norm $\|e_k\|$ decreases monotonically with $k$, whereas the gradient norm $\|g_k\|$ does not, necessarily.

\section{Inversion of the Viking Graben Stack}
I used CG iteration to solve the inverse problem set out in this chapter, assuming that the exploding reflector $r$ is zero outside of the subrectangle $\{3000 m \le x \le 27000m, 400 m \le z \le 2800 m\}$. The modeling is done in 2D, which means that some of the considerations developed above that depend on the nature of the leading singularity really ought to be changed. I ignored these conflicts, and simply used the system \ref{eqn:vdawer} without modification.

In order to make sure that all inputs were correct, I first calculated a single iteration of the CG algorithm described in  the last section. The RVL implementation of CG outputs key information, as displayed in Table 
%\ref{tab:cg1it}
1.
\begin{table}
\label{tab:cg1it}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Iteration   &  Residual Norm &  Gradient Norm\\
\hline 
         1  &  2.18368799e+03    & 3.05936885e+00\\
\hline 
         2  &  9.36336853e+02    & 1.13921440e+00\\
\hline
\end{tabular}
\end{center}
\caption{Conjugate Gradient first iteration.}
\end{table} 
Note that the RVL CG algorithm counts iterations Fortran-style, that is, starting from 1 instead of from 0, C style, as the algorithm pseudo-code in the last section would have it. I wonder why that is.

The first iteration has dropped the data residual norm to 75\% of its initial value - not so impressive. The first iterate ($r_1$) is more interesting:
\plot{explnmo1it}{width=0.9\textwidth}{Viking Graben ZO inversion: CG iteration 1} 
Compare it with the PSPI migration in Figure \ref{fig:parapspifixagcfilt}).
The resemblance is remarkable: in fact, the first iterate is a reasonably accurate {\em image}, in that it shows the same geometrical structures picked out in grey scale. Note from the CG listing that the first iterate is a multiple of the gradient of the least squares function at $r=0$. So this observation poses a question:

{\em Why is the gradient an image?}

The answer comes in the section after next.

Note that the obvious diffraction hyperbolae in the stack (Figure \ref{fig:parastackfixagc}) are absent in any of the depth images - collapsing of diffraction hyperbolae is a salient feature of migration (and indeed the first CG iterate is a migration).

Since the setup seemed to be working I went ahead with 6 steps. The residual record appears in Table %\ref{tab:cg6it}. 
2.
Note that the next-to-last iteration actually increases the gradient norm, as is possible during CG iteration, but the data residual norm monotonically decreases as it should.
\begin{table}
\label{tab:cg6it}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Iteration   &  Residual Norm &  Gradient Norm\\
\hline 
         1  &  2.18368799e+03    & 3.05936885e+00\\
\hline 
         2  &  9.36336853e+02    & 1.13921440e+00\\
\hline 
         3  &  5.54742188e+02  & 3.24299669e+00\\
\hline 
         4  &  5.08972382e+02  & 8.39157641e-01\\
\hline 
         5  &  4.04111511e+02  & 4.64890152e-01\\
\hline 
         6  &  3.84719269e+02  & 1.62544572e+00\\
\hline 
         7  &  3.25789581e+02  & 5.23918450e-01\\
\hline
\end{tabular}
\end{center}
\caption{Residual and Gradient norms, six Conjugate Gradient iterations.}
\end{table}
The recovered reflectivity $r$ after 6 iterations appears in Figure \ref{fig:explnmo6it}, and is somewhat less noisy than the 1 iteration result (Figure \ref{fig:explnmo1it}); both are very similar to the PSPI image in Figure \ref{fig:parapspifixagcfilt}.
\plot{explnmo6it}{width=0.9\textwidth}{Viking Graben ZO inversion: CG iteration 6}

The data predicted from this model (Figure \ref{fig:expldatanmo6it}) appears to match fairly well with the target data in Figure \ref{fig:parastackfixagcfilt} - the two figures are plotted on the same grey scale. According to the table above, the iteration has reduced the data misfit to less than 15\% of its initial size (RMS). Plotting the data residual on the same scale (Figure \ref{fig:resdatanmo6it}) shows how small the remaining error actually is.

\plot{expldatanmo6it}{width=0.9\textwidth}{Simulated data from reflectivity produced by 6 CG iterations (Figure \ref{fig:explnmo6it}).}

\plot{parastackfixagcfilt2}{width=0.9\textwidth}{Target data for inversion: duplicates Figure \ref{fig:parastackfixagcfilt}, filtered AGC'd NMO-corrected stack of Viking Graben line.}

\plot{resdatanmo6it}{width=0.9\textwidth}{Data residual, Viking Graben ZO inversion: CG iteration 6}

The alert reader will have possibly formulated a critical question. I have explained that manipulation of $F$ as a matrix is out of the question, even for this rather tame problem. If $F$ were represented as a matrix, then the meaning of $F^T$ would be clear, as would be its computation (swap the indices!).  However, $F$ is not represented as a matrix, but rather its action on a vector (representing an exploding reflecetor field) results from numerical solution of a wave equation. Accordingly,

{\em How can $F^T$ be computed without accessing its matrix?}

\section{Computation of the Adjoint: Reverse Time Migration}
The transpose or adjoint operator is {\em defined} by the dot product test: in the notation introduced in the last two sections, 
\begin{equation}
\label{eqn:dotprodtest1}
\sum_{x_m,y_m}\int \,dt_0\,(Fr)(x_m,y_m,t_0)d(x_m,y_m,t_0) 
=\int\int\int \,dx\,dy\,dz\,r(x,y,z)(F^T d)(x,y,z).
\end{equation}
From the definition of F and the delta function, the left-hand side of equation \ref{eqn:dotprodtest1} is the same as
\[
=\sum_{x_m,y_m}\int \,dt_0\,p(x_m,y_m,,z_m,t)d(x_m,y_m,t) 
\]
\begin{equation}
\label{eqn:dotprodrecast}
=\int \int \int \int dx\,dy\,dz\,dt\,p(x,y,z,t)\left(\sum_{x_m,y_m}d(x_m,y_m,t) \delta(x-x_m)\delta(y-y_m)\delta(z-z_m)\right)
\end{equation}
In this last version of the dot product test, the data has been converted to a ``data source'', a collection of point radiators with the data traces as their time dependence. Note that data is only recorded for a finite time (a few seconds, for exploration data) so make the very important presumption that $d$ vanishes for $t$ larger than a maximum time of recording $t_{\rm max}$,

If \ref{eqn:dotprodtest1} can be re-written so that the integrand displays an explicit factor of $r$, then the $F^Td$ can be read off by inspection - whatever multiplies $r$ must then be $F^Td$. From the system \ref{eqn:vdawer}, $r$ is the right-hand side of an equation satisfied by {\em derivatives} of $p, \bv$. As it currently stands, $p$ appears {\em without} derivatives in \ref{eqn:dotprodtest1}. However, derivatives can be ``borrowed'' from one factor in an integral to another via integration by parts. So if we can replace the ``data source'' in \ref{eqn:dotprodtest1} by a suitable combination of derivatives of another field, then we could presumably ``borrow'' those derivatives to set up the left hand side of the first equation in \ref{eqn:vdawer} hence introduce $r$.

Suppose that the factor multiplying $p$ in equation \ref{eqn:dotprodrecast} were itself the right-hand side of a wave equation, one of a pair similar to \ref{eqn:vdawer}:
\begin{eqnarray}
\label{eqn:vdaweadj}
\frac{\partial q}{\partial t}& = &-\kappa\left(\nabla \cdot \bw - \left(\sum_{x_m,y_m}d(x_m,y_m,t) \delta(x-x_m)\delta(y-y_m)\delta(z-z_m)\right)\right)\nonumber \\
\frac{\partial \bw}{\partial t} & = & -\beta \nabla q,\nonumber \\
q, \bw & = & 0, \, t > t_{\rm max}. 
\end{eqnarray}
It will become clear shortly why the fields $q, \bw$ should vanish for large positive time, instead of for large negative time as is the case of $p, \bv$. Then the right-hand side of equation \ref{eqn:dotprodrecast} becomes
\[
= \int \int \int \int dx\,dy\,dz\,dt\, p\left(\frac{\frac{1}{\kappa}\partial q}{\partial t}+\nabla \cdot \bw  \right)(x,y,z,t) 
\]
Integrate by parts in all space variables and time to move the derivatives onto $p$:
\[
= -\int \int \int \int dx\,dy\,dz\,dt\, \left((\frac{1}{\kappa}\frac{\partial p}{\partial t}q+\nabla p \cdot \bw  \right)x,y,z,t) 
\]
There are no boundary terms from integraion by parts in $t$ because $p$ vanishes for large negative time, and $q$ vanishes for large positive time. Since the midpoint (or receiver) locations occupy only a finite volume, and the data vanishes for $t>t_{\rm max}$ it follows from the finite speed of propagation of acoustic waves that at any fixed time, the fields $q, \bw$ vanish outside of a large ball, so there all the boundary terms in space also vanish. Now you know why the $q,\bw$ were required to vanish for large positive rather than negative time!

Use the equations \ref{eqn:vdawer} to replace the partial derivatives of $p$ with those of $\bv$:
\[
= -\int \int \int \int dx\,dy\,dz\,dt\, \left((-\nabla \cdot \bv + r\delta(t))q+
\frac{1}{\beta}\frac{\partial \bv}{\partial t} \cdot \bw  \right) (x,y,z,t) 
\]
then segregate the term involving $r$ and integrate by parts again:
\[
= - \int \int \int \int dx\,dy\,dz\,dt\, (r\delta(t) q) (x,y,z,t) 
\]
\[
+\int \int \int \int dx\,dy\,dz\,dt\, \bv \cdot \left(\nabla q + \frac{1}{\beta}\frac{\partial \bw}{\partial t}\right)
\]
The second summand vanishes because of the second equation in the system \ref{eqn:vdaweadj}, so the end result is
\begin{equation}
\label{eqn:asm}
= - \int \int \int \int dx\,dy\,dz\,dt\, r(x,y,z) q(x,y,z,0)
\end{equation}
Comparing equations \ref{eqn:dotprodtest1} and \ref{eqn:asm}, conclude that
\begin{equation}
\label{eqn:adjer}
(F^Td)(x,y,z)=-q(x,y,z,0).
\end{equation}
in which $q$ is the first component of the solution of system \ref{eqn:vdaweadj}.

This remarkable formula is a simple instance of the {\em adjoint state method}: it computes the adjoint of the modeling operator via solution of another system of wave equations. A discrete version also holds, and that is what is implemented in the IWAVE exploding reflector adjoint calculation.

\section{Why the Gradient is an Image}
Here is a slightly more ambitious statement about the meaning of ``image'': a function of 2 (or 3) variables is an image of another, if the two share the location and orientation of rapid oscillations. For example, this statement describes the relations between various subsurface images derived from the Viking Graben line in this text.

Location and rapid oscillation are to some extent contradictory properties, according to Heisenberg's uncertainty principle: it is not possible to localize a function in both space and frequency simultaneously  with arbitrary precision. However it is possible to construct functions localized to some extent in space and spatial frequency, and use them as exploding reflector sources to see which parts of the data they influence. Figure \ref{fig:pulse45} shows a localized rapid oscillation, oriented 45 degrees to the vertical. The Fourier transform is quite localized in the spatial frequency plane, and the function itself is nonzero onliy over an ellipse of horizontal diameter 2 km and vertical diameter 0.5 km. The center of the ellipse is $z=2000, x=10000$. Exploding reflector modeling with this data and constant velocity $v/2 = 1 km/s$ produces the data shown in \ref{fig:pulsedata45}. You will note that the approximate center position of the enlongated blob of energy is $x=12000$: it appears that the signal moved about 2 km laterally and 2 km vertically, as if it has traveled along a ray oriented at 45 degrees to the vertical, that is, perpindicularly to the equal phase surfaces of the oscillation.

In fact that is exactly what has happened, and it is possible to explain why with a variant of the geometric optics construction of Chapter \ref{ch:ray}.

A movie of the exploding reflector pressure field can be viewed with
\begin{verbatim}
scons pulsemovie45.rsf
sfgrey < pulsemovie45.rsf | xtpen
\end{verbatim}
for example - the reader should stretch the window horizontally so that the aspect ratio is closer to 1:1, to begin with. The pulse clearly moves up and to the right, in the direction normal to the equal phase surfaces.
\plot{pulse45}{width=0.9\textwidth}{Exploding reflector oscillating pulse data.}
\plot{pulsedata45}{width=0.9\textwidth}{Data with same geometry as Viking Graben stack, from data of Figure \ref{fig:pulse45} with constant (half-)velocity of 1 km/s, and density of 1 g/cm$^3$.}

The exploding reflector source has the form
\begin{equation}
\label{eqn:explrefrhs}
f(x,z,t) = cos(k(x \sin \theta + z \cos \theta))\delta(t).
\end{equation}
Since $\bv(x,z,t) = 0$ for $t<0$, it follows from the first equation of system \ref{eqn:vdawer} that
\begin{equation}
\label{eqn:explrefinit}
\lim_{t \rightarrow 0^+} p(z,x,t) = \kappa(x,z)cos(k(x \sin \theta + z \cos \theta))
\end{equation}
(in this case $\kappa = 1$). The data plot \ref{fig:pulsedata45} and the movie mboth suggest that the solution takes the form
\begin{eqnarray}
\label{eqn:go}
p(x,z,t) & \approx & cos(k \phi(x,z,t))p_0(x,z,t)\nonumber \\
\bv(x,z,t) & \approx & cos(k \phi(x,z,t))\bv_0(x,z,t) 
\end{eqnarray}
with $\phi(x,z,t)$ roughly linear in $x,z$ and shifting with some velocity as $t$ changes. Also $p_0$ and $\bv_0$ represent the observed envelopes of the solution (observed for $p_0$ - if you were to plot the components of $\bv$ you would see the same), which appear to move along the direction normal to the oscillations at velocity $v/2 =1$ km/s.

Plug the functional forms \ref{eqn:go} into the acoustic system \ref{eqn:vdawer}: for $t > 0$, obtain
\begin{eqnarray}
\label{eqn:go1}
-k \sin k\phi \frac{\partial \phi}{\partial t} p_0 + \cos k\phi \frac{\partial p_0}{\partial t}
& \approx & - \kappa (-k \sin k\phi \nabla \phi \cdot \bv_0 + \cos k\phi \nabla\cdot \bv_0)\nonumber \\
-k \sin k\phi \frac{\partial \phi}{\partial t} \bv_0 + \cos k\phi \frac{\partial \bv_0}{\partial t} 
& \approx & - \beta (-k \sin k\phi \nabla \phi p_0 + \cos k\phi \nabla p_0)
\end{eqnarray}
The terms with factors of $k$ are presumably dominant when $k >> 0$, so isolate them:
\begin{eqnarray}
\label{eqn:go1a}
-k \sin k\phi \frac{\partial \phi}{\partial t} p_0 & \approx & - \kappa (-k \sin k\phi \nabla \phi \cdot \bv_0 )\nonumber \\
-k \sin k\phi \frac{\partial \phi}{\partial t} \bv_0 & \approx & - \beta (-k \sin k\phi \nabla \phi p_0)
\end{eqnarray}
You can view this as a matrix system for the vector $u=(p_0,v_x,v_y,v_z)^T$, 
\begin{equation}
\label{eqn:go1b}
k \sin k\phi Mu =0
\end{equation}
with matrix
\begin{equation}
\label{eqn:go1c}
M =\left(\begin{array}{cccc}
 \frac{\partial \phi}{\partial t} & \kappa\frac{\partial \phi}{\partial x} & \kappa\frac{\partial \phi}{\partial y} & \kappa\frac{\partial \phi}{\partial z} \nonumber \\
\beta\frac{\partial \phi}{\partial x} &   \frac{\partial \phi}{\partial t} & 0 & 0\nonumber \\
\beta\frac{\partial \phi}{\partial y} & 0 &  \frac{\partial \phi}{\partial t} & 0 \nonumber \\
\beta\frac{\partial \phi}{\partial z} & 0 & 0 &  \frac{\partial \phi}{\partial t}
\end{array}\right)
\end{equation}
Equation \ref{eqn:go1b} implies that (assuming that $p_0, \bv_0$ are not all zero) the matrix $M$ is singular. Its determinant is
\begin{equation}
\label{eqn:go1d}
\left(\frac{\partial \phi}{\partial t}\right)^2 \left[\left(\frac{\partial \phi}{\partial t}\right)^2 -\kappa \beta \|\nabla \phi\|^2\right]
\end{equation}
Since $\kappa \beta = v^2$ and the first factor should be non-zero (why?), conclude that
\begin{equation}
\label{eqn:teik}
\frac{1}{v^2} \left(\frac{\partial \phi}{\partial t}\right)^2 = \|\nabla \phi\|^2.
\end{equation}
This is the time-dependent version of the eikonal equation \ref{eqn:eik}. Just how close the relation is follows from the ray construction of the phase $\phi$. As before, start with the solutions of the physical space part of the ray equations, posed in terms of the phase and a dimensional parameter $\sigma$:
\begin{eqnarray}
\label{eqn:go2a}
\frac{dX}{d\sigma} & = & -\nabla \phi(X,T) \nonumber \nonumber \\
 \frac{dT}{d\sigma} & = & \frac{1}{v^2(X)}\frac{\partial \phi}{\partial t}(X,T)
\end{eqnarray}
Since $k\phi$ (the exponent, or argument of $\cos$) must be dimensionless, and $k$ is a spatial frequency, $\phi$ has units of length. $T$ is a time, Therefore from the second equation in \ref{eqn:go2a}, $\sigma$ must have dimensions of length. The first equation is then consistent, as both sides are dimensionless. In fact, it will turn out that $\sigma$ is arc length.

The form of equations \ref{eqn:go2a} suggest the definition of a Hamiltonian:
\[
H(X,T,\xi,\omega) = \frac{1}{2}\left(\frac{\omega^2}{v^2(X)} - \|\xi\|^2\right)
\]
Then the system \ref{eqn:go2a} are the Hamilton equations for the evolution of $X$ and $T$:
\begin{equation}
\label{eqn:he1}
\frac{dX}{d\sigma} = \nabla_{\xi}H(X,T,\xi,\omega) = -\xi=-\nabla \phi(X,T)
\end{equation}
and 
\begin{equation}
\label{eqn:he2}
\frac{dT}{d\sigma} = \frac{\partial}{\partial \omega} H(X,T,\xi,\omega) = \frac{\omega}{v(X)^2} = \frac{1}{v^2}\frac{\partial \phi}{\partial t},
\end{equation}
 if you identify
\begin{equation}
\label{eqn:xiomega}
\xi = \nabla \phi, \omega = \frac{\partial \phi}{\partial t}
\end{equation}
Then
\[
\frac{d\omega}{d\sigma} = \frac{d}{d\sigma} \frac{\partial \phi}{\partial t}(X,T) 
\]
\[
= \nabla\frac{\partial \phi}{\partial t}\frac{dX}{d\sigma} + \frac{\partial^2 \phi}{\partial t^2}\frac{dT}{d\sigma}
= -\nabla\frac{\partial \phi}{\partial t}\nabla\phi + 
\frac{1}{v^2}\frac{\partial^2 \phi}{\partial t^2}\frac{\partial \phi}{\partial t}
\]
\[
=\frac{1}{2}\frac{d}{dt} \left(-|\nabla \phi|^2+\frac{1}{v^2}\left(\frac{\partial \phi}{\partial t}\right)^2\right)
\]
\begin{equation}
\label{eqn:he4}
= 0 = -\frac{d}{dT}H(X,T,\xi,\omega).
\end{equation}
Similarly,
\[
\frac{d\xi}{d\sigma} = \frac{d}{d\sigma} \nabla \phi(X,T)
\]
\[
=\nabla \nabla \phi(X,T)\frac{dX}{d\sigma} + \nabla \frac{\partial \phi}{\partial T} \frac{dT}{d\sigma}
= -\nabla\nabla \phi{X,T}\nabla \phi(X,T) + \nabla \frac{\partial \phi}{\partial T} \frac{1}{v^2}\frac{\partial \phi}{\partial T}
\]
\[
=\frac{1}{2}\nabla\left(-\|\nabla \phi\|^2 + \frac{1}{v^2}\left(\frac{\partial \phi}{\partial t}\right)^2\right)
-\frac{1}{2}\nabla\frac{1}{v(X)^2} \left(\frac{\partial \phi}{\partial t}\right)^2
\]
\begin{equation}
\label{eqn:he3}
= \frac{1}{2}\nabla\frac{\omega}{v(X)^2} = -\nabla_X H(X,T,\xi,\omega)
\end{equation}
The equations \ref{eqn:he1}, \ref{eqn:he2}. \ref{eqn:he3}, and \ref{eqn:he4} form the Hamiltonian system for the Hamiltonian $H$. The foregoing calculations have shown that if $\phi$ is a solution of the time-dependent eikonal equation \ref{eqn:teik}, then the trajectories $X,T,\xi,\omega$ obtained by solving 
\ref{eqn:go2a} and defining $\xi,\omega$ by \ref{eqn:xiomega} are solutions of Hamilton's equation. In Chapter \ref{ch:ray} I establihsed a similar set of identities for the time-independent eikonal \ref{eqn:eik}, then showed that one could start with the solution of the Hamilton system and construct a solution of the eikonal equation from it - that is, of course, what is commonly termed ray-tracing. Similarly, it is possible start with family of solutions of the time-dependent Hamiltonian system \ref{eqn:he1}, \ref{eqn:he2}. \ref{eqn:he3}, and \ref{eqn:he4}, and construct from them a solution of the time-dependent eikonal \ref{eqn:teik}. The apprriate initial conditions are (back to the 2D case!)
\[
X(\bx) = \bx, \,T(\bx) = 0,\, \xi(\bx) = -(\cos \theta, \sin \theta)^T,\, \omega(\bx)=\frac{1}{v(\bx)}
\]
Since $\frac{\partial T}{\partial \sigma} > 0$ (why?), you can parametrize the Hamiltonian trajectories by $T$, and 
As was the case with time-independent geometric optics in Chapter \ref{ch:ray}, the amplitudes $p_0$, $\bv_0$ satisfy transport equations: they evolve along the rays, so that $p_0(X(\sigma),T(\sigma))/p_0(X(0),0)$ is a positive slowly varying (smooth) function of $X(0)$.

For the constant velocity case illustrated in the figures, the Hamilton equations can be solved by inspection:
\[
X = \bx + v \xi \sigma, T = \frac{\sigma}{v}
\]
with $\xi$ and $\omega$ both constant along each ray. Note that $\sigma = vT$ and $dX/dT = v$, so that $\sigma$ is arc length as stated earlier. The trajectories (``rays'') $X(T)$ take off perpindicular to the equal phase surfaces of the exploding reflector source \ref{eqn:go} and the amplitudes advect along the straight rays.

This construction completely explains the appearance of the examples, assuming that the geometric optics approximation \ref{eqn:go} is accurate. In fact, you can show that the error is $O(1/k)$, by an argument similar to that explained in Chapter \ref{ch:ray}.

This reasoning does not quite explain why the first step of CG iteration produces an image. Instead of going further in this direction to construction such an explanation (as is certainly possible), I will take a different approach, that will prove useful in later chapters. This approach starts by noting that the first step of CG is proportional to the gradient ($F^Td$, in the notation of the last section) of the mean square error function $J$.

The goal is an analysis of the oscillatory behaviour of $F^Td$, in the case that the data $d$ is consistent with the approximation being used, that is, $d=Fr$ for a suitable $r$. That is, assume that $d$ is actually exploding reflector data for an exploding reflector source $r$. Then $F^Td = F^TFr$, so the question becomes:

To what extent is $F^TFr$ an ``image'' of $r$ - that is, to what extent are its rapid oscillation present in the same locations and orientations as those of $r$?

The key to the answer lies in the integral representation of the exploding reflector modeling operator, derived in section 2 as equation \ref{eqn:grt2g} (2D version), reproduced here for convenience: 
\begin{equation}
\label{eqn:erg0}
D\cF_{\rm ZO}[v^2]\delta v^2(x_r,,t) \approx  \int\, dt'\,\int \, dx\,dz\, \, G_{v/2}(x_r,t-t';x,z) \delta v^2(x,z) w(t'). 
\end{equation}
I have several times made changes in the time wavelet $w(t)$, not to mention amplitudes, and the justification for accepting these various changes will emerge finallly from the analysis to follow. So I will go ahead with another such change. Also I will suppress $z_r$ from $\bx_r=(x_r,z_r)$, using the simplifying assumption about acquisition geometry that all receivers are located at the same fixed depth.
In the notation introduced in the last section, and with the choice $w(t) = \delta(t)$, \ref{eqn:erg0} is equivalent to 
\begin{equation}
\label{eqn:erg}
Fr(x_r,t) \approx \int \, dx \, G_{v/2}(x_r,t;x,z) r(x,z). 
\end{equation}
This expression the analogue of a matrix representation of $F$, with $G_{v/2}$ as the matrix. Note that $G_{v/2}$ is not computationally accessible: the motivation for the construction of the last section is very much still active. However it is certainly possible to represent the adjoint $F^T$ in terms of the ``matrix'' $G_{v/2}$, simply by reversing the roles of the arguments (``indices''):
\begin{equation}
\label{eqn:erg}
F^Td(x,z) \approx  \int\, dt\,\int \, dx_r \, G_{v/2}(x_r,t;x,z) d(\bx_r,t). 
\end{equation}
Hence 
\begin{equation}
\label{eqn:norm0}
F^TFr(x,z) \approx \int\, dt\,\int \, dx_r \, \int \, dx'\,\int\,dz' \, G_{v/2}(x_r,t;x,z) G_{v/2}(x_r,t;x',z') r(x',z')
\end{equation}
Introduce the geometric optics approximation \ref{eqn:gohalf2} for $G_{v/2}$ in \ref{eqn:norm0}:
\[
F^TFr(\bx) \approx 
\]
\begin{equation}
\label{eqn:normgo}
\int\, dt\,\int \, dx_r \, \int \, dx'\,\int\,dz'\, a(x_r,x,z)a(x_r,x',z')\delta(t-2\tau(x_r,x,z))\delta(t-2\tau(x_r,x',z'))r(x',z')
\end{equation}
You can justify the evaluation of the delta functions by an argument similar to that mentioned in Chapter \ref{eqn:ch2}. The upshot is
\begin{equation}
\label{eqn:normgo1}
F^TFr(x,z) \approx \int \, dx_r \, \int \, dx'\,\int\,dz'\, a(x_r,x,z)a(x_r,x',z')\delta(2(\tau(x_r,x,z)-\tau(x_r,x',z')))r(x',z') 
\end{equation}
This integral is a distribution representation of a surface integral, over the family of surfaces 
$\{(x',z'): \tau(x_r,x',z')=\tau(x_r,x,z)\}$ parametrized by $x_r,x,z$. For constant velocity, these surfaces are simply spheres of radii $v\tau(x_r,x,z)$. For velocities not too far from constant, and $x',z'$ not too far from $x,z$, it is possible to solve the equation $t = \tau(x_r,x',z')$ for $z'$ in terms of the horizontal variables $x',x_r$. [Note that the 3D case is quite similar, just messier!] Express the solution of $\tau(x_r,x',z')=t$ as $z' = \zeta(x_r,x',t)$. Change variables from $z'$ to $\tau(x_r,x',z')$ in the integral \ref{eqn:normgo1} requires the Jacobian factor
\[
\frac{dz'}{dt}(x_r,x',t) = \left(\frac{\partial \tau}{\partial  z'}(x_r,x',\zeta(x_r,x',t))\right)^{-1}.
\]
Making the change of variable and integrating out the delta function gives 
\begin{equation}
\label{eqn:normgo2}
F^TFr(x,z) \approx \int \, dx_r \, \int \, dx'\, b(x_r,x,z,x')r(x',\zeta(x_r,x',\tau(x_r,x,z)) )
\end{equation}
in which I have abbreviated
\begin{equation}
\label{eqn:normgoamp}
b(x_r,x,z,x')=a(x_r,x,z)a(x_r,x',\zeta(x_r,x',\tau(x_r,x,z))) \left(\frac{\partial \tau}{\partial z'}(x_r,x',\zeta(x_r,x',t))\right)^{-1}.
\end{equation}

Remember that the goal is to understand the relation between the high spatial frequency components of $r$ and $F^TFr$. Introduce the frequency components of $r$ into the relation \ref{eqn:normgo2} via the Fourier transform of $r$:
\begin{equation}
\label{eqn:normgoft}
F^TFr(x,z) \approx \frac{1}{(2\pi)^2}\int\,dk_x \,\int \, dk_z \int \, dx_r \, \int \, dx'\, b(x_r,x,z,x')e^{i(k_xx'+k_z\zeta(x_r,x',\tau(x_r,x,z)) )}\hat{r}(k_x,k_z) 
\end{equation}
The goal is to understand the behaviour of this oscillatory integral for large $k_x,k_z$. For this purpose, use the {\em Stationary Phase Principle}. It approximates an oscillatory integral, with {\em amplitude} $g$, smooth and vanishing outside of a ball in  Euclidean n-space $\bR^n$, and a real-valued {\em phase} $\psi$ defined on the same ball and having the

\noindent {\bf Nondegenerate Stationary Point} property: if $\nabla \psi(\by^*) = 0$, then $\det \nabla \nabla \psi(\by^*) \ne 0$

The approximation is
\begin{equation}
\label{eqn:statphase}
\begin{aligned} 
&\int_{\bR^n}dy g(\by)e^{i\omega \psi(\by)}\\
&\approx \sum_{\nabla
  \psi(\by^*)=0}\left(\frac{2\pi}{\omega}\right)^{\frac{n}{2}}e^{\frac{\pi
    i}{4}\mathrm{sgn}\  \nabla\nabla\psi(\by^*)}
|\det \nabla \nabla \psi(\by^*)|^{-\frac{1}{2}}g(\by^*)e^{i\omega \psi(\by^*)} + O(\omega^{-n-1}.
\end{aligned}
\end{equation}

In the application to \ref{eqn:normgoft}, $n=2$, $x,z$ are static parameters, and $x_r,x$ are the integration variables. For the large parameter ($\omega$ in \ref{eqn:statphase}), take $k=\sqrt{k_x^2+k_z^2}$, and write  
\begin{equation}
\label{eqn:normgophase}
\psi(x_r,x';x,z,\theta) = \sin \theta x' + \cos \theta \zeta(x_r,x',\tau(x_r,x,z))
\end{equation}
in which 
\[
\sin \theta = \frac{k_x}{k},\,\cos\theta=\frac{k_z}{k}
\]
The amplitude is $b$ as defined in \ref{eqn:normgoamp}, divided by $(2\pi)^2$.

The stationary phase condition is
\begin{eqnarray}
\label{eqn:normgostat}
\frac{\partial \psi}{\partial x'} = \sin \theta + \cos \theta \frac{\partial \zeta}{\partial x'} & = & 0\nonumber \\
\frac{\partial \psi}{\partial x_r} = \cos \theta\left(\frac{\partial \zeta}{\partial x_r} + \frac{\partial \zeta}{\partial t}\frac{\partial \tau}{\partial x_r}\right)& =& 0
\end{eqnarray}
From the defining equation of $\zeta$ and implicit differentiation, obtain
\[
\frac{\partial \zeta}{\partial t} = \left(\frac{\partial \tau}{\partial z}\right)^{-1}; \,
\frac{\partial \zeta}{\partial x'} = -\frac{\partial \tau}{\partial x'}\left(\frac{\partial \tau}{\partial z}\right)^{-1}; \,\frac{\partial \zeta}{\partial x_r} = -\frac{\partial \tau}{\partial x_r}\left(\frac{\partial \tau}{\partial z}\right)^{-1},
\]
with appropriate arguments prescribed. Hence the first stationary phase equation implies that
\[
\frac{\partial \tau}{\partial z'} \sin \theta - \frac{\partial \tau}{\partial x'}\cos \theta = 0,
\]
that is $\nabla \tau(x_r,x',z')$ is perpindicular to $(-\cos \theta, \sin \theta)$, hence parallel to 
$(\sin \theta, \cos \theta)$ and therefore to $(k_x,k_z)$ at a stationary point $k_r,x'$, with $z'=\zeta(x_r,x',\tau(x_r,x,z))$ whence $\tau(x_r,x',z') = \tau(x_r,x,z)$. That is: $(x,z)$ and $(x',z')$ lie on the same isochron, and $(k_z,k_z)$ is parallel to $\nabla \tau$ at $(x',z')$.

The second stationary phase condition is equivalent to 
\begin{equation}
\label{normgorays}
\frac{\partial \tau}{\partial x_r}(x_r,x',z') = \frac{\partial \tau}{\partial x_r}(x_r,x,z).
\end{equation}
From the discussion in Chapter \ref{ch:rays}, the left hand side is the horizontal component of the ray slowness at $x=x_r, z=z_r$ for the ray connecting $(x,z)$ with $(x_r,z_r)$, and similarly for the right-hand side. A hidden hypothesis, implicit in our use of the convolutional model to justify replacing the zero-offset modeling with exploding reflector modeling, is that rays do not turn horizontal, so that both slownesses must have negative z-components at the receiver point. Therefore they are the same, by the eikonal equation. Since the two rays arrive at the same point, and with the same slownessess, all coordinates of the two rays are the same at the receiver point - therefore they are the same ray!!! Since both $(x,z)$ and $(x',z')$ lie on the same ray, and since they lie on the same isochron, as mentioned in the last paragraph, they are the same point. That is, we conclude that $x=x', z=z'$ and $(k_x,k_z)$ is parallel to $\nabla \tau(x_r,x,z)$.

Having identified the stationary point (there is only one), it remains to compute the Hessian $\nabla \nabla \psi$ and its signature. In fact, the signature is $0$, and I leave the Hessian as an exercise. Conclude that the oscillatory integral over $x_r,x'$ in \ref{eqn:normgoft} is 
\begin{equation}
\label{eqn:normgosymb}
\int \, dx_r \, \int \, dx'\, b(x_r,x,z,x')e^{i(k_xx'+k_z\zeta(x_r,x',\tau(x_r,x,z)) )}\hat{r}(k_x,k_z)
= \frac{2\pi}{k} B(x,z,k_x,k_z) e^{k_x x+k_z z}\hat{r}(k_x,k_z)
\end{equation}
to leading order in $k$. The quantity $B(x,z,k_x,k_z)$ is the product of the amplitude $b(x_r,x,z,x')$ and the Hessian determinant factor from the stationary phase principle, evaluated at $x=x',z=x'$ and $x_r$ equal to the receiver horizontal location at which the ray taking off from $(x,z)$ with initial slowness $(k_x,k_z)$ reaches the receiver depth $z=z_r$. Note that $(k_x,k_z)$ enters into the construction of $B$ only through the direction vector $(\sin \theta, \cos \theta)$, so that $B(x,z,k_x,k_z)$ is homogeneous in $(k_x,k_z)$ of  order 0.

Putting all of this together, obtain an approximation for highly oscillatory $r$ (that is, large $k$):
\begin{equation}
\label{eqn:normgopsido}
F^TF(x,z) \approx \frac{1}{2\pi}\int \,dk_x\,dk_z\,\frac{1}{k} B(x,z,k_x,k_z) e^{k_x x+k_z z}\hat{r}(k_x,k_z)
\end{equation}
Oscillatory integral operators of this type have come to be called {\em pseudodifferential}, and will be discussed in some detail in the next chapter. The quantity multiplying the exponential, excluding the Fourier transform $\hat{r}$ is called the {\em symbol} of the operator, and in this case is homogeneous in $(k_x,k_z) $ of order -1.

\noindent {\bf Note, 2/17 draft:} The careful reader will see that I have committed two errors in the preceding derivation: (1) I used the leading order asymptotic form of the 3D Green's function in a 2D calculation, and (2) the exploding reflector source is in effect used as the RHS in the 2nd order wave equation, not the first order system. These errors affect the details but not the overall conclusion of the argument. The operator $F^TF$ is correctly identified as a pseudodifferential operator, but the order and the detailed calculation of the symbol is wrong. The effect of the first error is to make the Green's function 1/2 order more singular than it should be. The correct $F^TF$ should be of one order less, that is, of order -2 rather than -1. The second error has the opposite effect: if the definition of $F$ rests on the system \ref{eqn:vdawer}, then the exploding reflector source must be differentiated in time to be used as the RHS in the 2nd order wave equation. The net result is that $F^TF$ defined by solving \ref{eqn:vdawer} is 0th order, and this conclusion is independent of space dimension.

The chief properties of pseudodifferential operators important at this point are these:
\begin{itemize}
\item Preservation of high frequency asymptotics: if $r$ does {\em not} oscillate rapidly in the direction $(\sin \theta,\cos\theta)$ near $(x,z)$, then neither does $F^TF$. If the symbol is positive on at $(x,z,\sin\theta,\cos\theta)$, then the converse is also true.
\item Insensitivity to amplitude and frequency factors: the property just enunciated is preserved if the symbol is scaled by additional factors of spatial frequency or positive smooth amplitude.
\item Changes of amplitude or spatial or temporal frequency in the representation of $F$ result in the sort of changes to the symbol mentioned in the previous bullet.
\end{itemize}

Taken together, these properties and equation \ref{eqn:normgopsido} validate the assertions made throughout this chapter about the exploding reflector model. In particular, the first step of CG iteration, that is, the application of the adjoint modeling operator to the data, results in an image - assuming that the data is itself consistent with the exploding reflector model. In other words, exploding reflector migration creates an output with the same localized high frequency oscillations as the exploding reflector source used to create the data.

The Viking Graben example was consistent with this conclusion, however the data was ``real'', that is, extracted from field data tapes. To verify this assertion in vacuo, synthetic examples are useful. Figure \ref{fig:pulse15} shows a oscillating pulse exploding reflector source, with the phase normal 15 degrees from vertical. Figure \ref{fig:pulsedata15} shows that exploding reflector data computed from it, and Figure \ref{fig:pulsemig15} the exploding reflector migration of, that is, application of the adjoint operator to, this data. The recovery of the location and orientation of the oscillations is virtually perfect - evidently the spatial frequency used in this example is high enough that the error in stationary phase approximation is unimportant.
\plot{pulse15}{width=0.9\textwidth}{Exploding reflector oscillating pulse data - normal to phase surfaces 15 deg from vertical.}
\plot{pulsedata15}{width=0.9\textwidth}{Data with same geometry as Viking Graben stack, from exploding reflector source of Figure \ref{fig:pulse15} with constant (half-)velocity of 1 km/s, and density of 1 g/cm$^3$.}
\plot{pulsemig15}{width=0.9\textwidth}{Exploding reflector migration of data in Figure \ref{fig:pulsedata15}.}

\section{A One-Step Approximate Inversion}
A re-write of the linear inverse problem introduced earlier in this chapter (equation \ref{eqn:psinv}) in terms of the operator $F$ would be: given trace data $d$, find an exploding reflector source $r$ so that
\begin{equation}
\label{eqn:psinvr}
Fr \approx d.
\end{equation} 
The analysis in the preceding section illuminates a feature of this problem that was not obvious before: in general, it does not have unique solutions. For purely dimensional reasons, \ref{eqn:psinvr} does not always have a solution, as indicated earlier: it is an overdetermined problem, at least for the data layout and discretization choices made in this chapter.

It is also an underdetermined problem. For example, the exploding reflector  depicted in Figure \ref{fig:pulse90}, with horizontal equal-phase normals, produces the data depicted in Figure \ref{fig:pulsedata90}. This figure is not a misprint - the data is effectively zero. Closer examination reveals that it is not exactly zero, but contains a region of apparent noise with amplitude $O(10^{-6})$ smaller than the data in Figure \ref{fig:pulsedata15}, for instance. These nonzero samples are a combination of finite difference and round-off error. 
\plot{pulse90}{width=0.9\textwidth}{Exploding reflector oscillating pulse data - normal to phase surfaces 90 deg from vertical.}
\plot{pulsedata90}{width=0.9\textwidth}{Data with same geometry as Viking Graben stack, from exploding reflector source of Figure \ref{fig:pulse90} with constant (half-)velocity of 1 km/s, and density of 1 g/cm$^3$. Grey scale at same level as that of Figure \ref{fig:pulsedata15}.}
Since $d=0$, a perfectly good solution to equation \ref{eqn:psinvr} is $r=0$. The exploding reflector source in Figure \ref{fig:pulse90} is a null vector for $F$. It must also be a null vector for $F^TF$, so solutions of the least-squares version of the inverse problem, or the normal equation \ref{eqn:psinvnorm}, are also not unique.

We can however single out a solution of the least-squares problem, namely the solution with least length. In effect, this {\em pseudoinverse} to the normal operator produces the part of the exploding reflector source that fits as much of the data as possible. A standard approximation is the solution of the {\em regularized least-squares problem}: choose $r$ to minimize
\begin{equation}
\label{eqn:psinvreg}
J_{\lambda}[r] = \|Fr -d\|^2 + \lambda^2 \|r\|^2
\end{equation}
The minimizer is a solution of the regularized normal equation
\begin{equation}
\label{eqn:psinvregnorm}
(F^TF + \lambda^2 I)r = F^Td.
\end{equation}
Since $F^TF + \lambda^2I$ is positive definite for $\lambda \ne 0$, this problem has unique solutions. For small $\lambda$, the solution approximates the least squares solution of least length.

The regularized least squares problem can also be solved by conjugate gradient iteration, as before. [So why did we get away with solving the unregularized problem with CG? There is a good answer!]

I will next describe an approximation to this solution that only involves a single application of $F^T$. To begin with, observe that any $r$ can be assembled to arbitrarily good approximation out of oscialling pulses similar to the last few examples: namely, choose a convenience partition of unity $\{\chi_i\}$ of $\bR^2$ (we will stay with the 2D problem in this section, but 3D follows exactly the same pattern). Then
\[
r(x,z) = \sum_i \frac{1}{(2\pi)^2}\int\,dx\,\int\,dz\,\hat{r}(k_x,k_z) \chi_i(x,z)\cos(k_x x + k_z z)
\]
and the right-hand side consists of sums and integrals of oscillating pulses. The data for a typical summand 
\begin{equation}
\label{eqn:pulsesource}
r(x,z) = \chi(x,z)\cos(k_x x + k_z z) 
\end{equation}
can be read off from the time-dependent geometric optics approximation developed in the last section. The pressure component of the solution looks like
\begin{equation}
\label{eqn:pulsepressure}
p(x,z,t) =  \cos(k \phi(x,z,t)) p_0(x,z,t) + O\left(\frac{1}{k}\right),\,\,k=\sqrt{k_x^2+k_z^2}.
\end{equation}
so the modeled data is
\begin{equation}
\label{eqn:pulsedata}
Fr(x_r,t) = \cos(k \phi(x_r,z_r,t)) p_0(x_r,z_r,t) + O\left(\frac{1}{k}\right).
\end{equation}
The amplitude $p_0$ is transported along the ray field that begins at time $t=0$. If all of the rays starting in the support of $\chi$ arrive at the receiver depth within the recording time interval, then it appears that there is enough information to recover $r$ from the data traces: the amplitude $p_0$ can be reconstructed by solving the transport equation backwards in time, starting at points on the recording surface $z=z_r$ - the same goes for the phase, and then at $t=0$ one recovers the initial values of $\phi$ and $p_0$ and therefore of $r$, to good approximation for large $k$.

This is correct, with the two important caveats. First, those who have viewed the propagating-pulse movies will understand that the soltion $(p,\bv)$ of the pressure-velocity system \ref{eqn:vdawer} contains not one, but two propagating pulses, that move in opposite directsions $\pm (k_x,k_z)$ from a monochromatic pulse source like that defined in \ref{eqn:pulsesource}. Correspondingly, two geometric optics solutions of the form \ref{eqn:pulsepressure} add together to approximate the pressure field. It is easy to see that the two amplitudes must be the same for $t=0$, therefore each is half of the initial pressure data:
\[
p_0(x,z,0) = \frac{1}{2}\chi(x,z).
\]
In the half-space geometry of the synthetic survey, if one of the two traveling pulses generated by $r$ travels upward toward the surface at an angle close enough to vertical that the entire signal (except for the $O(1/k)$ error) arrives at the receivers in the recorded interval, then the other pulse travels in the positive $z$ direction and is not recorded. [The 45 degree pulse movie shows this behaviour clearly.] So the recovered initial value of the amplitude must be doubled.

The other caveat is that separating the recorded data into pulses and picking apart the amplitude and phase factors is a nontrivial task, best avoided if at all possible.

In fact, it turns out that it can be avoided. The basic idea is simple: equation \ref{eqn:pulsedata} simply reads off the trace of the pressure field on the surface $z = z_r$. So you could regard the acoustic field $(p,\bv)$ as the solution of the boundary value problem
\begin{eqnarray}
\label{eqn:vdawerdir0}
\frac{\partial p}{\partial t}&=& -\kappa (\nabla \cdot \bv - r \delta(t)) \nonumber \\
\frac{\partial \bv}{\partial t} &=& -\beta \nabla p\nonumber \\
p(x,z_r,t) &= &Fr(x,z_r,t) \nonumber \\
p,\bv & = & 0, \,t<0 
\end{eqnarray}
The additional third equation makes sense if the receivers are dense enough to be regarded as continuous - for the example used here, the receiver sampling is the same as the $x-$ increment, so the discretization is compatible with that assumption.

The problem \label{eqn:vdawerdir0} is however posed in a circular way. I have assumed that the acoustic system is posed in all of $\bR^2$, and more than that, the region $z<z_r$ ``above the surface'' is filled with homogenous material that does not generate reflections. In effect, an {\em absorbing boundary condition} is posed at $z=z_r$ - in fact, in the computational implementation, that is exactlyl the setup. So it is not possible to specify another boundary condition at $z=z_r$ - any other boundary condition is redundant, and implies a compatibility condition with the absorbing condition that $(p,\bv)$ is already expected to satisfy.

The resolution of this paradox comes in two steps. First, observe that the initial conditions (last equations in \ref{eqn:vdawerdir1}) imply  that $\lim_{t\rightarrow 0^+} \bv(x,z,t) = 0$, but the form of the exploding reflector source implies that $\lim_{t\rightarrow 0^+} p(x,z,t) = \kappa(x,z)r(x,z)$. So you could reformulate the exploding reflector source as an initial condition:
\begin{eqnarray}
\label{eqn:vdawerdir1}
\frac{\partial p}{\partial t}&=& -\kappa \nabla \cdot \bv  \nonumber \\
\frac{\partial \bv}{\partial t} &=& -\beta \nabla p\nonumber \\
p(x,z,0) & = & \kappa(x,z)r(x,z)\nonumber \\
\bv(x,z,0) & = & 0. 
\end{eqnarray}
If you know the value of $p$ at $z=z_r$, that is, $d=Fr$, then you can drop the absorbing boundary condition and substitute a Dirichlet condition:
\begin{eqnarray}
\label{eqn:vdawerdir2}
\frac{\partial p}{\partial t}&=& -\kappa \nabla \cdot \bv , \, z>z_r \nonumber \\
\frac{\partial \bv}{\partial t} &=& -\beta \nabla p, \, z>z_r \nonumber \\
p(x,z_r,t) & = & d(x,t) \nonumber \\
p(x,z,0) & = & \kappa(x,z)r(x,z), z > z_r \nonumber \\
\bv(x,z,0) & = & 0, z > z_r. 
\end{eqnarray}
This problem has a unique solution, and it is the same as the solution of \ref{eqn:vdawerdir1} if indeed $d$ is the value at $z=z_r$ of $p$.

However the absorbing boundary condition has disappeared from the formulation of \ref{eqn:vdawerdir2}, so the relation between $r$ and $d$ is broken: it is possible to solve this system for {\em any} choice of $r$ and $d$. The connection between the two lies in $(p,\bv)$ actually being the solution of (say) \ref{eqn:vdawerdir1}, with absorbing boundary at $z=z_r$, restricted to $z>z_r$. This condition is difficult to check directly, however one of its consequences is obvious. If $r$ is the data of a pulse exploding reflector source of the form \ref{eqn:pulsesource} with large $k$ and wavenumber oriented sufficiently vertically that the part of the solution traveling in the negative $z$ direction passes $z=z_r$ in the negative $z$ direction, in the recording interval $0 \le t \le t_{\rm max}$, the the field $(p,\bv)$ at $t=t_{\rm max}$ in $z > z_r$ consists only of the other part of the pulse, propagating in the positive $z$ direction. That part of the pulse propagates independently of the part propagating upward ( = negative $z$ direction), so you can {\em reproduce the upward propagating part of the pulse by ignoring the downward propagating part}: that is, you can simply zero out the field at $t=t_{\rm max}$. In so doing, you have committed two errors: an error of leaving out the entire downward propagating pulse, and an $O(1/k)$ error stemming from the use of the geometric optics approximation. Setting the field at this later time is inconsistent with initial conditions at time $t=0$, so drop them:
\begin{eqnarray}
\label{eqn:vdawertr}
\frac{\partial p}{\partial t}&=& -\kappa \nabla \cdot \bv , \, z>z_r \nonumber \\
\frac{\partial \bv}{\partial t} &=& -\beta \nabla p, \, z>z_r \nonumber \\
p(x,z_r,t) & = & d(x,t) \nonumber \\
p(x,z,t_{\rm max}) & = & 0, z > z_r \nonumber \\
\bv(x,z,t_{\rm max}) & = & 0, z > z_r. 
\end{eqnarray}
Under the hypotheses presented above, the solution $(p,\bv)$ of this system should be an $O(1/k)$ approximation the the upward propagating pulse part of the geometric optics approximation.

The relation to $r$ is recovered by remembering that the two parts of the pulse, upward and downward propagating, are each equal to half of the exploding reflector source at time $t=0$. Therefore 
\[
\frac{1}{2}\kappa(x,z)r(x,z) \approx p(x,z,0)
\]
for the solution $(p,\bv)$ of \ref{eqn:vdawertr}, under these assumptions. For this solution, define
\begin{equation}
\label{eqn:psinvpsinv}
F^{\dagger}d(x,z) = \frac{2}{\kappa(x,z)} p(x,z,0).
\end{equation}
Then the principle of superposition (linearity of the wave equation) implies:

\noindent {\em For exploding reflector sources $r$ that can be decomposed into oscillating pulses with wavenumbers sufficiently close to vertical, }
\[
Fr = d \, \Rightarrow\, r \approx F^{\dagger}d
\]

Otherwise put, for exploding reflector sources $r$ that generate ``big'' data $d$, $F^{\dagger}F r \approx r$. Note that if $r$ generates near-zero data, that is, is in the approximate null space of $F$, then  $F^{\dagger}F r \approx 0$. The properties indicate the $F^{\dagger}$ is an approximate pseudoinverse of $F$.

Because the initial data for \ref{eqn:vdawertr} is posed at the maximum simulation time $t=t_{\rm max}$, the system is actually solved backwards in time. Such a backwards-in-time system has appeared earlier in this chapter, namely the adjoint state system for computation of $F^T$, which I reproduce here for convenience, in its 2D version:
\begin{eqnarray}
\label{eqn:vdaweadj1}
\frac{\partial q}{\partial t}& = &-\kappa\left(\nabla \cdot \bw - \sum_{x_m,y_m}d(x_t,t) \delta(x-x_r)\delta(z-z_r)\right)\nonumber \\
\frac{\partial \bw}{\partial t} & = & -\beta \nabla q,\nonumber \\
q, \bw & = & 0, \, t > t_{\rm max}. 
\end{eqnarray}
The difference between this system and the system \ref{eqn:vdawertr} is that \ref{eqn:vdaweadj1} is posed in all of $\bR^3\times \bR$, and its inhomogeneous term is a right hand side, whereas \ref{eqn:vdawertr} is posed only in the half-space $\{z > z_r\}$ and its inhomogenous term is a boundary condition.

It is possible to convert \ref{eqn:vdawertr} to a system in the whole space with inhomogeneous term, and thereby establish a relation between $F^{\dagger}$ and $F^T$. The convention in this book has been that the system \ref{eqn:vdawe} is defined in all of $\bR^4$, in particular that $\kappa$ and $\beta$ are defined in all of $\bR^3$. So it is possible replace $z>z_r$ in \ref{eqn:vdawertr} with $z<z_r$ and so define a second system in the complementary half-space. Note that the boundary condition for $p$ is the same in both cases, so if you regard $p$ as being defined in $\bR^4$, then it is {\em continuous} across $z=z_r$. It will turn out that $\bv$ is not generally continuous, but still $(p,\bv)$ is a vector-valued function and solves the homogeneous wave equation at any point in $\bR^4$ {\em except} on $z=z_r$.

Next observe that this extended version of $(p,\bv)$ solves a wave equation, but one with a non-zero right-hand side in the $p$ equation, supported (non-zero) on $z=z_r$. To figure out what this right-hand side is, write the condition for a weak solution of the acoustic system \ref{eqn:vdawe} (the only kind of solution $(p,\bv)$ can be if $\bv$ is discontinuous): for $(\phi,{\bf \psi}) \in C^{\infty}_0(\bR^3,\bR \times \bR^3)$, 
\[
\int_{\bR^3} \left(\frac{1}{\kappa}\frac{\partial \phi}{\partial t}+\nabla \cdot {\bf \psi}\right)p +\left(\frac{1}{\beta}\frac{\partial {\bf \psi}}{\partial t} + \nabla \phi\right)\cdot \bv
\]
\begin{equation}
\label{eqn:vdaweweak}
= \int \phi g
\end{equation}
To identify the right choice of $g$ to make \ref{eqn:vdawe} with initial condition at $t = t_{\rm max}$ (rather than $t=0$) equivalent to \ref{eqn:vdawertr}, divide the integral on the left into two integrals, one from $z=-{\infty}$ to $z=z_r$, the other from $z=z_r$ to $z=\infty$. Assume that $(p,\bv)$ are regular, and that integration by parts is legitimate, except at $z=z_r$. Then the left hand side above is
\[
= -\left[\int_{z\le z_r} dz dx dy dt + \int_{z\ge z_r} dz dx dy dt\right]\left(\phi\frac{1}{\kappa}\frac{\partial p}{\partial t}+{\bf \psi} \cdot\nabla p\right.
\]
\[
\left. + {\bf \psi} \cdot\frac{1}{\beta}\frac{\partial \bv} {\partial t} + \phi \nabla \cdot \bv\right)
\]
\begin{equation}
\label{eqn:vdawertrweak}
- \int_{z=z_r}dx dy dt (\psi_z [p] + \phi [v_z])(x,y,z_r,t)
\end{equation}
in which the notation $[f]$ signifies the jump of a function $f$ across $z=z_r$:
\[
[f](x,y,z_r,t) = lim_{z\rightarrow z_r^+}f(x,y,z,t) - lim_{z\rightarrow z_r^-}f(x,y,z,t).
\]
As mentioned before, $[p]=0$ as its limit from both sides is the (same) Dirichlet condition. Comparing \ref{eqn:vdaweweak} and \ref{eqn:vdawertrweak} shows that in $z>z_r$, the solution $(p,\bv)$ of \ref{eqn:vdawertr} is the same as the weak (distribution) solution of 
\begin{eqnarray}
\label{eqn:vdawertrhs}
\frac{\partial p}{\partial t}&=& -\kappa (\nabla \cdot \bv -  [v_z]\delta(z-z_r)) , \, \nonumber \\
\frac{\partial \bv}{\partial t} &=& -\beta \nabla p, \, \nonumber \\
p(x,z_r,t) & = & d(x,t) \nonumber \\
p(x,z,t_{\rm max}) & = & 0,  \nonumber \\
\bv(x,z,t_{\rm max}) & = & 0.
\end{eqnarray}
Comparing the systems \ref{eqn:vdawertrhs} and \ref{eqn:vdaweadj1}, the solution $(q,\bw)$ of the latter must be equal to the solution $(p,\bv)$ if $d$ is replaced by $[v_z]$. Since $F^Td = -q_{t=0}$, 
\begin{equation}
\label{eqn:pseudoadj}
F^{\dagger}d = -\frac{2}{\kappa}F^T[v_z]
\end{equation}

\section{Dirichlet-to-Neumann map}

I have not said how $\kappa$ and $\beta$ are extended to the exterior domain $z<z_r$. A simple choice is evenly: that is, impose the condition
\[
\kappa(x,y,z)=\kappa(x,y,2z_r-z), \,\beta(x,y,z)=\beta(x,y,2z_r-z).
\]
This condition may render $\kappa$ and $\beta$ only continuous, not differentiable, at $z=z_r$, but the concept of weak solution is still applicable, and weak solutions may be constructed out of ordinary solutions in the two half-spaces that and right-hand sides supported on $z=z_r$.

Suppose $(p,\bv)$ is an ordinary (smooth) solution of \ref{eqn:vdawertr} in $z\ge z_r$. Then extending $p, v_x, v_y$ to be even and $v_z$ to be odd generates a solution of the same system in $z<z_r$, as is easily verified. Viewed as functions on all of $\bR^4$, the evenly-extended $p,v_x,v_y$ are continuous, whereas, $v_z$ generally acquires a discontinuity. Since the extension is odd, $[v_z](x,y,z_r,t) =2 \lim_{z \rightarrow z_r^+}v_z(x,y,z,t)$.

Newton's law (second equation in \ref{eqn:vdawe} implies - for anti-causal solutions such as specified in \ref{eqn:vdawertr} - that
\[
v_z(\bx,t) = \int^{\infty}_t \,ds\,\frac{\partial p}{\partial z}(\bx,s)
\]
so knowledge of the limiting value of $v_z$ at $z=z_r$ is equivalent to knowledge of $\partial p/\partial t$. Since the boundary value of $p$ (that is, $d$ in the system \ref{eqn:vdawertr}) determines the anti-causal solution in $z_r$ hence the boundary value of $\partial p/\partial z$ or equivalently $v_z$, in fact \ref{eqn:vdawertr} defines a (clearly linear) map 
\begin{equation}
\label{eqn:dtondef}
\Lambda d =-2 \lim_{z \rightarrow z_r^+}v_z.
\end{equation} 
Since the boundary value of $p$ is conventionally termed {\em Dirichlet data}, and that of $\partial p/\partial z$ {\em Neumann data}, on $z=z_r$, the map $\Lambda$ is a version of the {\em Dirchlet-to-Neumann map}, which has figured in many works on inverse problems for partial differential equations (for example \cite{Uhl18}).

In terms of the Dirichlet-to-Neumann map, the relations \ref{eqn:psinvpsinv} and \ref{eqn:pseudoadj} imply that
\begin{equation}
\label{eqn:pseudodton}
F^{\dagger} = \frac{2}{\kappa} F^T \Lambda .
\end{equation}

The Dirichlet-to-Neumann map has two very important properties: applied to higly oscillatory data,
\begin{itemize}
\item it is positive definite, at least when restricted to data that can be decomposed into upcoming pulses;
\item it depends only on the properties of the material parameters $\kappa$ and $\rho$ in an arbitrarily thin layer $z_r\le z <z_{\rm max}$.
\end{itemize}

As a consequence of the first property, \ref{eqn:pseudodton} can be interpreted as meaning that $F^{\dagger}$ is the adjoint of $F$ in the weighted inner product
\[
\langle d_1, d_2 \rangle_{\rm data} = \langle d_1,\Lambda d_2 \rangle_{L^2}
\]
SInce $F^{\rm dagger}$ is an approximate inverse of $F$, \ref{eqn:pseudodton} implies tht $F$ is approimately unitary with resepct to the inner product $\langle \cdot,\cdot \rangle_{\rm data}$. That in turn implies that the CG algorithm, for instance, will converge rapidly if reformulated in terms of $\langle \cdot,\cdot \rangle_{\rm data}$.

The second property implies that, insofar as highly osicllatory data is concerned, the 
choice of extension of the density and bulk modulus does not matter: only the behaviour of these coefficients inside the half-space $z>z_r$, and in fact near the boundary $z=z_r$, are significant.

Both properties follow from the high-frequency approximation for pulse data introduced earier. Recall the  form of the the time-dependent geometric optics approximate solution \ref{eqn:go}, which in 3D reads
\begin{eqnarray}
\label{eqn:go1}
p_{go}(x,y,z,t) & \approx & cos(k \phi(x,y,z,t))p_0(x,y,z,t)\nonumber \\
\bv_{go}(x,y,z,t) & \approx & cos(k \phi(x,y,z,t))\bv_0(x,y,z,t) 
\end{eqnarray}
If the phase $\phi$ satisfies the time-dependent eikonal equation, and the amplitudes $p_0$, $\bv_0$ the corresponding transport equations, then $(p_{go},\bv_{go})$ differs from a solution of the wave equation by a field that is $O(1/k)$ in any compact subset of space-time, as shown in Chapter \ref{ch:ray}. 
It is possible to construct an approximate solution for which the phase is a given linear phase, when restricted to $z=z_r$:
\[
\phi(x,y,z_r,t) = t + \hat{k}_x x + \hat{k}_y y
\]
Then the eikonal equation determines the normal derivative of $\phi$ at the boundary:
\begin{equation}
\label{eqn:dphidz}
\frac{\partial \phi}{\partial z}(x,y,z_r,t) = \pm \sqrt{\frac{1}{v(x,y,z_r)^2}-\hat{k}_x^2-\hat{k}_y^2}
\end{equation}
Evidently, for this prescription to make sense, a necessary condition is that
\begin{equation}
\label{eqn:timelike}
1 > v(x,y,z_r)^2(\hat{k}_x^2+\hat{k}_y^2)
\end{equation}
over the support of $p_0,\bv_0$. We will come back to the significance of this condition below.

Assuming that condition \ref{eqn:timelike} is satisfied, construct an upcoming pulse, i.e. one associated with rays along which $z$ decreases as $t$ increases by solving the Hamilton equations \ref{eqn:he1} and \ref{eqn:he2}. These equations imply that the correct sign above is $+$.

These relations imply that for an approximate solution of \ref{eqn:vdawertrhs} of the form \ref{eqn:go1}, representing an upcoming pulse,
\[
v_z(x,y,z_r,t) = \int_t^{\infty}\,ds\,\beta(x,y,z_r)\frac{\partial p}{\partial z}(x,y,z_r,t)
\]
\[
=\int_t^{\infty}\,ds\,\beta(x,y,z_r)\left(\frac{\partial p_0}{\partial z} \cos(k\phi) - p_0(x,y,z_r,t)\sin(k\phi)k\frac{\partial \phi}{\partial z}\right) (x,y,z_r,t)
\]
\[
=\int_t^{\infty}\,ds\,\beta(x,y,z_r)\left(\frac{\partial p_0}{\partial z} \cos(k\phi) + p_0(x,y,z_r,t)\frac{\partial}{\partial t}\cos(k\phi)\frac{\partial \phi}{\partial z}\left(\frac{\partial \phi}{\partial t}\right)^{-1}\right) (x,y,z_r,t)
\]
\[
=-p_0(x,y,z_r,t)\cos(k\phi(x,y,z_r,t)) \left(\frac{\partial \phi}{\partial z}\left(\frac{\partial \phi}{\partial t}\right)^{-1} \right)(x,y,z_r,t)+ O\left(\frac{1}{k}\right)
\]
after integrating by parts, which due to \ref{eqn:dphidz} becomes
\begin{equation}
\label{eqn:dton1}
=-p_0(x,y,z_r,t)\cos(k(t+\hat{k}_x x + \hat{k}_y y)) \sqrt{\frac{1}{v(x,y,z_r)^2}-\hat{k}_x^2-\hat{k}_y^2} + O\left(\frac{1}{k}\right)
\end{equation}
With this result, it is possible to write a high-frequency approximation to the Dirichlet-to-Neumann map. First construct a local approximation: presume that $d$ is supported near a single point on $z=z_r$, and let $p_0 \in C^{\infty}_0$ be $=1$ on the support of $d$. Then 
\[
d(x,y,z_r,t) = p_0(x,y,z_r,t)d(x,y,z_r,t) 
\]
\[
= \frac{1}{(2\pi)^3} \int\,dk\,dk_{x}\,dk_y p_0(x,y,z_r,t)\hat{d}(k,k_{x},k_{y})cos(k(t+\hat{k}_x x +\hat{k}_y y))
\]
where $\hat{k}_x=k_x/k,\hat{k}_y=k_y/k$. From \ref{eqn:dton1}, the corresponding $v_z$ is 
\[
v_z(x,y,z_r,t) = -\frac{1}{(2\pi)^3} \int\,dk\,dk_{x}\,dk_y
\]
\[
\times  p_0(x,y,z_r,t)\hat{d}(k,k_{x},k_{y})cos(k(t+\hat{k}_x x +\hat{k}_y y)) \sqrt{\frac{1}{v(x,y,z_r)^2}-\hat{k}_x^2-\hat{k}_y^2} +...
\]
where the elided terms have lower frequency content than the leading term displayed, as they derive from the $O(1/k)$ remainder in \ref{eqn:dton1}. Recalling the definition \ref{eqn:dtondef}, and the fact that $p_0$ is $=1$ on the support of $d$, we obtain the approximation
\[
\Lambda d (x,y,z_r,t) \approx 
\]
\begin{equation}
\label{eqn:dtonpseudo}
\frac{1}{(2\pi)^3} \int\,dk\,dk_{x}\,dk_y \hat{d}(k,k_{x},k_{y})cos(k(t+k_x x +k_y y)) \sqrt{\frac{1}{v(x,y,z_r)^2}-\left(\frac{k_x}{k}\right)^2-\left(\frac{k_y}{k}\right)^2} +...
\end{equation}
exhibiting $\Lambda$ in the form of a pseudodifferential operator on the boundary $z=z_r$.

This derivation is incomplete in several ways. To begin with, I have dispensed with the constraint \ref{eqn:timelike}. It must be reintroduced, via a factor under the integral sign in \ref{eqn:dtonpseudo} that enforces it. The identity with the reconstructed pressure field at $t=0$ is then only correct if the data $d$ really does consist (at least in high-frequency limit) of Fourier components satisfying \ref{eqn:timelike}. That and other gaps in the reasoning above may be filled by invoking the theory of pseudodifferential operators more completely than I have done here.

Note that the multiplier inside the integral in \ref{eqn:dtonpseudo} is positive (when it is well-defined). This means that $\Lambda$, as defined here, is a positive operator, modulo low frequency (relatively smoothing) error. Also, operators of this form are symmetric, again modulo low frequency error. Therefore $\Lambda$ may be used to define a norm in the data space, as noted above.

\section{Computational Notes}

\section{Suggested Projects}

\bibliographystyle{seg}  % style file is seg.bst 
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/IPrev20181210/paper.tex
repro docs dir added

\title{Review of  "Relaxation algorithms for matrix completion, with applications to seismic travel-time data interpolation", by Baraldi et al.}
\date{}
\author{}

\righthead{}

\maketitle
\parskip 12pt

This paper presents a new algorithm for data completion (rather than interpolation, which I shall argue is inappropriately used in the title) via matrix completion. Rank minimization to fill in missing matrix entries is a well-established technique. The algorithm pressented is novel, so far as I know, and the application is interesting and important. Missing data infill is in a sense the original inverse problem, so the paper is well-suited to Inverse Problems and should find plenty of interested readers.

That said, there are several defects in the presentation that render the paper quite difficult to read. Probably the main defect is that the two major methodological sections are not well-connected. A related issue is that several details, critically important to any attempt at reproducing the authors' results, are omitted or insufficiently described. I will give a quick overview of the paper's main sections, identifying as I go the matters that seem to me to need further attention.

Section 2 gives an overview of matrix completion, introducing a few ideas particular to the intended application. After introducing nuclear norm regularization, the authors assert that a smoothness constraint is also appropriate for seismic travel-time data, leading to a ``Morozov'' regularization with an objecctive that linearly combines the nuclear norm and the L2 norm of the Laplacian. Then the nuclear norm, uncomputable for large problems, is replaced with a computable bound via matrix factorization, with low-rank factors. Nothing is said at this point about how this rank ($k$ in their notation) is selected, though that choice appears likely to be a critical factor in the performance of the algorithm to follow. One finds out only at the beginning of Section 5 that $k$ has been set $=40$, without guidance given for how this choice is made.

Section 3 introduces a slack variable $W$ and relaxes the formulation further (equation (9)), introducing yet another penalty weight $\eta$. A block coordinate descent algorithm is introduced to minimize this objective. The critical step 6 is solved via a root-finding algorithm, indeed a variant of the familiar root-finding approach to solving the secular equation occuring in the trust region subproblem. This step involves a SPD matrix solve, implemented via Cholesky decomposition

So far, so good. However the very next sentence brings up the first major disconnect (p. 6, just before section 3.1): ``This system only changes when $\eta$ is updated.''. This is a critical point - the expense of step 6 appears to be the major cost of Algorithm 1, in which $\eta$ is merely a parameter. The listing includes no rule for updating $\eta$, despite the critical role played by this parameter. Eventually, in the discussion around the first set of experiments in Section 5,  p. 11, a rule for updating $\eta$ is mentioned, without justification or complete description (what is the ``pre-interpolated matrix''? why every 30 iterations, rather than every 10 or 10$^{30}$?).

Section 3 concludes by pointing out that both FISTA and LBFGS can be used to solve variants of the minimization formulated there. Algorithm 1 will be benchmarked against these standard methods. FISTA gets an algorithm listing, whereas LBFGS does not (approriately!).

Section 4 abruptly changes gear, to discussing the particulars of the iMUSH project. Some of this discussion is simply out of place - it really belongs in the introductory Section 1, particularly the paragraph beginning ``Our goal here is to...''. The reader should not have to wade through eight pages to find the goal of the paper.

A quite disturbing statement follows: ``Approximations or projections of this raw data onto the grid have yielded poor results for all of the aforementioned algorithms.'' I understand this statement to mean that the algorithms described in section 3 have been attempted with field data, involving necessarily some sort of interpolation from gridded (synthetic) receiver locations to actual field coordinates, and the results found wanting. This is surprising, for two reasons. First, it calls into question the interest of the entire enterprise: if low rank and smoothness cannot be used to estimate gridded times that interpolate onto field stations with something like the believed accuracy of the field picks, then why bother? Second, it appears to contradict the statement about smoothness at the beginning of subsection 2.3: if the time fields are indeed smooth as functions of receiver coordinates, then accurate interpolation should be possible.

I really don't think the authors should leave this tension unresolved. The full-blown problem with field data need not be tackled, but the interest of the synthetic problem constructed here is minimal without a roadmap to dealing with the qualities of the actual data.

Finally, the word ``interpolation'' in the title should really be replaced, perhaps by ``completion'', since the problem solved does not include the interpolation to arbitrary sensor locations.

There follows a description of the time residuals from a 1D model - these residuals form the actual data to be in-filled - and the ``matricization'' (unfolding of the 3D survey receiver grid into a matrix) of this data. The description is rendered unnecessarily confusing by the consistent use of source coordinate pairs $(S_{x_i},S_{y_i}): i=1,...,n_s$, which actually play no role whatsoever in the formulation of the algorithm - the sources could as well have been teleseismic, in the form of approximate plane waves. In any case their geometry is irrelevant, but it takes a bit of reading to see this. A revised description would be welcome - just get rid of the S's and keep the R's and keep the indices.

What does play a role is the matricization of the sources, and the receiver set for each source - natural if, as in this synthetic setup, they each form a rectangular grid. The first matricization is what might be called organization into receiver gathers, in another branch of seismology; the second simply blows up each source indes {\em pair} into a $20 \times 20$ submatrix. Both of these have a natural description if you parametrize the time picks as 4-tensors $T_{ijkl}$, where $i,j$ are the receiver indices and $k,l$ are the source indices.

Another puzzling assertiion appears at the end of p. 10: ``The ideal situation is for the subsampled data to have high rank ...''. Why? In what sense is this ideal? The following assertion, ``Then it is feasible...'' is no help: That the full (i.e. completed) data should be ``low rank'' is the core hypothesis of this approach to travel-time data in-fill, as was revealed in the introduction. Why it should be important that the subsampled data havae full rank is not obvious at all, at least to this reviewer, and no reason is given.

Even more puzzling is the organization of sources by energy, mentioned (but not explained) on p. 11. The data here is travel-time picks, not seismic traces. What can possibly be means by source energy in this connection? The concept disappears without a trace after the first half of p. 11, but it should either be left out or explained.

An overall critique of Sections 2-4 is that they are notationally incoherent, and this incoherence is responsible for the much of the difficulty this reader experienced. The matrix $X$ is described verbally in its two formulations, on p. 10 - but why not state the relation explicitly, namely
\[
X_{20i+j,20k+l} = T_{ijkl}
\]
in the parametrization I have suggested for formulation 1,
\[
X_{i+20k,j+20l}= T_{ijkl}
\]
for formulation 2. Such an explicit mathematical relation must be part of an actual implemention - why leave it out, and thereby leave the reader uncertain?

Again, the only thing we find out about the subsampling operator ${\cal A}$ is that it outputs a (randomly selected?) 15\% of the time picks. A bit more precision would be appreciated, and re-use of the notation from section 2 would be helpful in connecting the discussion there with the applicaiton.

Section 5 presents the results of several numerical experiments, testing the ability of Algorithm 1 to recover gridded data from a sub-sampling, the relative importance of smoothness and low rank constraints, and the relative performance of Algorithm 1, Algorithm 2 (FISTA), and LBFGS. The presentation seems clear and readable to me.

The only comment I have on the results section is on the relative importance of the low rank constraint. Its use is central to the entire approach, but it is not obvious to me that mere Tihonov regularization with the Laplacian constraint would not reconsitute the missing data equally well - the results presented seem to support that notion.

In summary, I find the methodology of the paper to be interesting, novel, and likely useful to a substantial number of IP readers. However the presentation could use quite a bit of modification, to ease the reader's burden and to completly specify the algorithm which is the main content of the paper.

### 2019-10-19 repro/papers/Jon20181209/paper.tex
repro docs dir added

\title{Perturbations by a Pool}
\date{}
\address{
        \footnotemark[1]{West Sound, Orcas Island}
}
\author{William Symes\footnotemark[1]}

\righthead{Perturbations}

\maketitle
\parskip 12pt

It may be that my take on this misses the point, but here is a try at an answer.

In this horizontally layered business, every trace depends on the same reflectivity $r(t_0)$.

The trace $d(t,x)$ at offset $x$ has an amplitude $a(t,x)$ and a moveout function $\tau_0(t,x) = t_0$ corresponding to arrival time $t$ at offset $x$ - ignore the noise series and the wavelet.

Officially, 
\[d(t,x) = a(t,x) r(\tau_0(t,x))\]

I think of the last factor as a non-stationary convolution - the moveout function stretches $t$ to $t_0$.

introduce velocity perturbation $\delta v$ and reflectivity perturbation $\delta r$. $\delta v$ results in perturbations $\delta a$ in amplitude and $\delta \tau_0$ in moveout function.

by Taylor, Leibnitz, chain rule, etc., the resulting data perturbation $\delta d$  is to first order
\[
\delta d(t,x) = \delta a(t,x) r(\tau_0(t,x)) + a(t,x) \delta \tau_0(t,x) \frac{dr}{dt_0} (\tau_0(t,x)) + a(t,x) \delta r(\tau_0(t,x))
\]
So, yes, the effect of small velo and refl changes is a sum of three nonstationary convolutions.

However note that $\delta v$ appears implicitly - notably through $\delta \tau_0$ - rather than as the explicit target of a non-stationary filter. It is the reflectivity and its perturbation that are subject to non-stationary filter.

The middle term is "blue", that is, preferentially emphasizes short wavelengths in the reflectivity. The other two are "white".

But maybe I have gone completely off the rails, so will stop here.

### 2019-10-19 repro/papers/Li/paper.tex
repro docs dir added

\title{Report on ``Subsurface seismic imaging based on inversion velocity analysis in both image and data domains'', by Yubing Li, submitted as Th\`{e}se de Doctorat}

\author{
William Symes\thanks{Department of Computational and Applied Mathematics, Rice University,
Houston, TX, 77005, USA,
{\tt symes@rice.edu}}
}

\lefthead{Symes}

\righthead{Li Thesis}

\maketitle

\begin{abstract}
This thesis presents several versions of surfacec oriented IVA and comparisons with similar MVA algorithms. It develops a full account of the algorithms and presents a number of 2D examples.
\end{abstract}

\section{Report}
Li's thesis gives a thorough account of a surface-oriented approach to Inversion Velocity Analysis (IVA), based on the shot-record extension of constant density acoustic modeling. This Differential Semblance approach to velocity model building goes back to early work of the author of this report in the 1980's, or even further: the essential idea is the basis for NMO-based velocity analysis, a technique so ubiquitous that many geophysics students no longer even learn it.

IVA has several major variants. The one discussed in this thesis relies on asymptotic least-squares inversion of the extended Born modeling operator. Other implementations of shot-record DSO using wave equation solvers (mostly from my group \cite[]{KerSy:94,HuangSymes:SEG15,YinHuang:16} but also consult Ali Almomin's Stanford thesis and recent work by Robert Soubaras) have used iterative inversion to solve the ``inner'' problem. Li does a very good job of explaining how this asymptotic inversion is used, and of justifying its use (as opposed to Reverse Time Migration) to construct Common Image Gathers which drive the velocity update scheme. He also argues correctly that shot-record extended DSO based on asymptotic inversion is much less computationally expensive than the similar algorithm based on subsurface offset extension \cite[]{HouSymes:SEG16a,Herve2017}, costing no more per iteration than standard FWI, hence feasible in the near future for 3D field scale application.

The thesis appears to contain two draft papers. The first, already submitted to {\em Geophysics}, explains the common-shot IVA algorithm, with details of optimization and gradient calculation relegated to appendices, then presents several examples. The second compares the ``image domain'' algorithm of Chapter 3 to the Differential Waveform Inversion algorithm of \cite{ChaurisPlessix:EAGE13}, to an inversion-driven variant of DWI which the author calls ``data domain IVA''.

 I have specific comments on two points that I think may be worth Mr. Li's attention:

\subsection{Multipathing}

The second example presented in Chapter 3 uses a low-velocity anomaly to create data containing caustics and multiple arrivals. This example is included because a number of past studies indicated that even at the correct velocity, such data may give rise to non-flat CIGs. Li concludes that ``the impact of the triplication is not that severe for the common-shot IVA approach'' (p. 85, similar statements are found in Chapter 5).

I find this conclusion questionable. To begin with, one consequence of multiple arrivals is that shot record extended Born modeling operator is either literallly not invertible, or has relatively tiny singular values - effectively, it is not invertible. So it is not clear what is meant by ``inversion''.  This conclusion is implicit in some of the cited references \cite[]{NolanSymes:96,StolkSymes:04}: kinematically, a reflection horizon in a shot record can correspond to two (or more) reflecting horizons in the subsurface - that is the origin of the non-flat CIGs. The dynamics (amplitudes) are easily filled out to make each reflecting horizon fit the shot record data separately. So one has constructed a null space for the extended modeling operator, or alternatively two completely different extended reflectivities that reproduce the same data. One of these is ``physical'', i.e. has flat CIGs, whereas the other does not. However both are isolated - that is, without changing the velocity, there are no reflectivities near these two that fit the data. Possibly the two models are connected through an intermediate suite of velocities, along which the DSO objective changes monotonically - but I don't see any particular reason to believe that. In any case, the change in velocity would at least for part of the trajectory introduce non-flat CIGs for the ``correct'' reflectivity branch.

The examples shown in this chapter do not address this question - as noted in the discussion of the 2nd and 3rd examples, the inverted velocity does not create the caustics of the ``true'' model, so in some sense is not close enough to it to engage this problem. Because a couple of examples do not exhibit a plausible pathology is no reason to think that it does not exist - one must remember the absolutely reliable Murphy's Law, that is: anything that can go wrong, will go wrong.

Therefore I think that the discussion in this thesis of the the influence of triplication on the global convergence properties of surface-oriented IVA is incomplete. To be fair, this matter has not really been discussed thoroughly by anyone - there are a couple of examples in the literature of failure to converge in zones with complex ray geometry (one is in Almomin's thesis, another in Huang's thesis \cite[]{HuangSymes:SEG15,YinHuang:16}), but failure to converge can occur for several reasons, eg. not enough iterations. It remains an open question. The examples given here are provocative, and should stimulate more work.

The great advantage of surface-oriented IVA and related DSO-like algorithms is that their computation cost per iteration is comparable to that of FWI, especially with a ``direct'' inversion used either as the one-step inner loop or as a preconditioner.  The community should try out these algorithms, but be aware of the potential pitfall.

\subsection{Single-shot Inversion}
Formulas like (3.9) first appeared (to the best of my knowledge) in t'Root's thesis 
\cite[]{RootStolkHoop:SEG09,RootStolkHoop:12}. Apparently the CGG group rediscovered this result independently - hardly surprising, happens all the time in this business - but t'Root has priority and should be cited.

\subsection{Conclusion}
Overall, the conclusions of this work are clearly explained and well-supported, and the examples are well-chosen. This is a thorough exploration of a class of algorithms that have potential for gaining widespread use. In my opinion, this thesis amply justifies the awarding of a PhD in computational geophysics to Mr. Yubing Li.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/asg/paper1_madplot.tex
repro docs dir added

\inputdir{project}
\plot{frame13}{width=15cm}{Point source field, homogeneous medium with
$v_p=1.5$ km/s, at 1.2 s}
\plot{frame40-1}{width=15cm}{Point source field at 4.0 s, after interaction with reflecting (zero-pressure) boundaries}
\plot{frame40-2}{width=15cm}{Point source field at 4.0 s, after interaction
with 250 m PML boundary zones on bottom and sides ($\eta_0=1.0$) -
same grey scale as Figure \ref{fig:frame40-1}. Longest wavelength
carrying significant energy is roughly 500 m.}
\plot{frame40-3}{width=15cm}{Point source field at 4.0 s, after interaction
with 100 m PML boundary zones on bottom and sides ($\eta_0=1.0$) -
same grey scale as Figure \ref{fig:frame40-1}.}
\plot{bm1}{width=15cm}{Dome bulk modulus}
\plot{by1}{width=15cm}{Dome buoyancy}
\plot{data1}{width=15cm}{2D shot record, (2,4) staggered grid scheme, $\Delta x = \Delta z =$ 5 m,
  appropriate $\Delta t$, 301 traces: shot x = 3300 m, shot z = 40 m, receiver x =
 100 - 6100 m, receiver z = 20 m, number of time samples = 1501, time
sample interval = 2 ms. Source pulse = zero phase trapezoidal [0.0,
2.4, 15.0, 20.0] Hz bandpass filter.}

\plot{trace}{width=15cm}{Trace 100 (receiver x = 2100 m) for $\Delta x = \Delta z = $
 20 m (black), 10 m (blue), 5 m (green), and 2.5 m (red). Note
 arrival time discrepancy after 1 s: this is the interface error
 discussed in \cite[]{SymesVdovina:09}. Except for the 20 m result,
 grid dispersion error is minimal.}

\plot{wtrace}{width=15cm}{Trace 100 detail, 1.8-2.5 s, showing more clearly the
 first-order interface error: the time shift between computed events
 and the truth (the 2.5 m result, more or less) is proportional to
 $\Delta t$, or equivalently to $\Delta z$.}

\plot{data8k1}{width=15cm}{2D shot record, (2,8) scheme, other
  parameters as in Figure \ref{fig:data1}.}

\plot{trace8k}{width=15cm}{Trace 100 computed with the (2,8) scheme,
  other parameters as described in the captions of Figures
  \ref{fig:data1} and \ref{fig:trace}.}

\plot{wtrace8k}{width=15cm}{Trace 100 detail, 1.8-2.5 s, (2,8) scheme..
Comparing to Figure \ref{fig:wtrace}, notice that the dispersion error for
the 20 m grid is considerably smaller, but the results for finer grids
are nearly identical to those produced by the (2,4) grids - almost all
of the remaining error is due to the presence of discontinuities in
the model.}

### 2019-10-19 repro/papers/asg/paper1_main.tex
repro docs dir added

\begin{abstract}
  IWAVE is a framework for time-domain regular grid finite difference
  and finite element methods. The IWAVE package includes
  source code for infrastructure component, and 
  implementations of several wave physics modeling categories. This
  paper presents two sets of examples using IWAVE acoustic staggered grid
  modeling. The first set illustrates the effectiveness of a simple
  version of Perfectly Matched Layer absorbing
  boundary conditions. The second set reproduce illustrations from a  recent paper on error propagation for
  heterogeneous medium simulation using finite differences, and
  demostrate the interface error effect
  which renders all FD methods effectively first-order accurate.  
  The source code for these examples is packaged with the paper
  source, and supports the user in duplicating the results presented
  here and using IWAVE in other settings.
\end{abstract}

\section{Introduction}
Domain-specific simulation such as seismic modeling begs for
software re-use via modular design. All applications of this type have
the same structure: static fields are initialized, dynamic fields
updated, output extracted. A modular approach to code architecture is
implicit in this structure, and further specialization leads to even more opportunity
for code re-use via modular design. 
%The general topic of this paper, time-domain
%simulation on regular rectangular grids, presents specific openings
%for code re-use via abstraction: time steps may have internal
%structure which is repeated in a defined pattern, parallelism via
%domain decomposition may be computed from stencil shape information
%rather than hard-wired into the code, sampling rules for output
%buffering are simple to formulate and universal, and external data
%formats may be hidden behind uniform interfaces.

IWAVE is open source
software for finite difference or finite element time-domain
simulation on regular rectangular grids, written exclusively in the
C/C++. IWAVE is built around a core framework: that is,
a collection of separate software packages which together provide
essential services upon which applications may be built. These
service components completely define the interfaces to which
additional code must be written to formulate a complete
application. The core framework defines
\begin{itemize}
\item parameter-driven job control;
\item grid generation and memory allocation in 1D, 2D, and 3D space;
\item serial, loop-parallel, and task-parallel execution models,
scaling to thousands of threads; 
\item arbitrary source and receiver locations, and flexible
source specification including simultaneous source modeling (random,
plane-wave,...)
\item standard input and output data formats (SEGY, RSF)
\item predefined support for linearized (Born) modeling and adjoint
  linearized (RTM) modeling, both first and second order;
\item uniform interface to optimization and linear algebra for
  creation of inversion applications via the
  Rice Vector Library (``RVL'') \cite[]{RVLTOMS,GeoPros:11}.
\end{itemize}
\cite{trip14:struct} describes the desgn principles
underlying the IWAVE core framework, and illustrates the construction
of a complete acoustic modeling application using centered finite
differences for the second order acoustic constant denstity wave equation.

The primary purpose of this short paper is to illustrate 
synthetic seismogram generation using another finite
difference scheme implemented in IWAVE, the staggered grid approximation to
variable-density velocity-pressure acoustodynamics \cite[]{Vir:84}. 
Exactly the same framework supports this application as was described
in \cite{trip14:struct}; as explained there, only two data structures
and six principal functions need be defined to implement this (or any) finite
difference method in IWAVE.

The examples illustrate two aspects of finite difference modeling.
The IWAVE staggered grid implementation includes a version of PML
absorbing boundary conditions \cite[]{Habashy:07}, permitting accurate
finite grid approximation of wave propagation in a full- or
half-space. The first set of examples demonstrates the effectiveness
of these very simple PML conditions.
The second set reproduce the examples presented in 
\cite{SymesVdovina:09}, and illustrate a fundamental limitation in the
use of straightforward finite-difference methods for modeling waves in
heterogeneous media.

IWAVE was used in a quality control role in the SEAM Phase I project -
see \cite{FehlerKeliher:11} for an account, including discussion of
the many difficulties of large scale numerical simulation of seismograms.

The internal details of IWAVE are not discussed here, except insofar
as is necessary to explain the use of the main
commands. As mentioned above, \cite{trip14:struct} overviews the design
of IWAVE and the main features of its internal
structure, and defines the elements necessary to compile a new IWAVE application. \cite{GeoPros:11} briefly describe the IWAVE/RVL
mechanisms for coupling modeling with optimization packages
to produce inversion applications.

The paper begins with a brief review of the system of partial
differential equations solved (approximately) by IWAVE's acoustic
application, and the choice of finite difference method. The next
section evaluates the effectiveness of the PML absorbing boundary
conditions included in the IWAVE staggered grid acoustic application.
The following section presents the examples of \cite{SymesVdovina:09},
along with some additional examples based on the same distribution of
mechanical parameters which shed light on the impact of finite
difference order on solution accuracy. Instructions follow for recreating
these examples, and for using them as starting points for further
modeling exercises. The paper ends with a brief discussion of the
prospects for improvements in performance and accuracy in FD
technology, and the evolutionary advantages flowing from the modular,
or object, orientation of IWAVE. Two appendices describe the job
parameters used in the examples, and download and install
instructions.

\section{Acoustodynamics}
The pressure-velocity form of
acoustodynamics consists of two coupled first-order partial
differential equations:
\begin{eqnarray}
\label{awe}
\rho \frac{\partial {\bf v}}{\partial t} &=& - \nabla p \\
\frac{1}{\kappa}\frac{\partial p}{\partial t} &=& -\nabla \cdot {\bf v} + g
\end{eqnarray}
In these equations, $p({\bf x},t)$ is the pressure (excess, relative to an
ambient equilibrium pressure), ${\bf v}({\bf x},t)$ is the particle velocity,
$\rho({\bf x})$ and $\kappa({\bf x})$ are the density and bulk modulus
respectively. Bold-faced symbols denote vectors; the above formulation
applies in 1, 2, or 3D.

The inhomogeneous term $g$ represents externally supplied energy (a
``source''), via a defect in the acoustic constitutive relation. A
typical example is the {\em isotropic point source}
\[
g({\bf x},t) = w(t) \delta({\bf x}-{\bf x}_s)
\]
at source location ${\bf x}_s$.

\cite{Vir:84} introduced finite difference methods based on this
formulation of acoustodynamics to the active source seismic
community. \cite{Vir:86} extended the technique to elastodynamics, and
\cite{Lev:88} demonstrated the use of higher (than second) order
difference formulas and the consequent improvement in dispersion
error. Many further developments are described in the review paper
\cite{moczoetal:06}. IWAVE's acoustic application uses the principles introduced by
these authors to offer a suite of finite difference schemes, all
second order in time and of various orders of accuracy in space.

The bulk modulus and buoyancy (reciprocal density) are the natural
parameters whose grid samplings appear in the difference
formulae. These are the parameters displayed in the figures below,
rather than, say, velocity and density, which might seem more natural.

\section{PML Effectiveness}

The IWAVE acoustic staggered grid scheme implements the Perfectly
Matched Layer (PML) approach to absorbing boundary conditions, in one of the
simpler of its many guises (a split field approach -
\cite[]{Habashy:07}). After some manipulation, the acoustic PML system
for the physical velocity $\bv$ and an artificial vector pressure
$\bp$ takes the form
\begin{eqnarray}
\label{eqn:pml}
\rho \left(\frac{\partial v_k}{\partial t} + \eta_k(x_k)v_k\right) &=&
- \frac{\partial p_k}{\partial x_k}, \nonumber \\
\frac{1}{\kappa}\left(\frac{\partial p_k}{\partial t} + \eta_k(x_k)p_k\right) &=& -\nabla \cdot {\bf v} + g
\end{eqnarray}
in which the $k$th component of the attenuation profile vector ${\bf \eta}$ depends
only on $x_k$, and can be stored as a collection of 1D objects. Ordinary acoustic
wave propagation takes place where ${\bf \eta}={\bf 0}$, and if the
components of the vector pressure $\bp$ are all the same in this zone,
then they remain the same there, and any one of them may be regarded
as the same as the physical pressure field. Outside of the physical
domain, where waves are to be attenuated, ${\bf {\eta}}$ should
ibe positive; at the boundary of the physical domain, it should vanish to positve order. We
elected to make ${\bf {\eta}}$ cubic in distance to the boundary: for
a PML layer of width $L_{k,r}$, beginning at $x_k=x_{k,r}$ along the $k$th
coordinate axes,
\[
\eta(x_k) = \eta_0 \left(\frac{x_k-x_{k,r}}{L_{k,r}}\right)^3
\]
Thus there are four PML boundary layer thicknesses in 2D, six in 3D, one
for each side of the simulation cube. The IWAVE convention imposes
pressure-free boundary conditions on the exterior boundary of the PML
domain. Thus $L=0$ signifies a free surface boundary face. Any face of
the boundary may be assigned a zero-pressure condition ($L=0$) or a
PML zone of any width ($L>0$).

Many implementations of PML, especially for elasticity, confine the
extra PML fields (in this case, the extra pressure variables) to
explicitly constructed zones around the boundary, and use the standard
physical system in the domain interior. We judged that for acoustics little would be
lost in either memory or efficiency, and much code bloat avoided, if
we were to solve the system (\ref{eqn:pml}) in the entire domain.

Considerable experience and some theory
\cite[]{Habashy:07,moczoetal:06} suggest that the system \ref{eqn:pml} will
effectively absorb waves that impinge on the boundary, emulating free
space in the exterior of the domain, if the PML zones outside the
physical domain in which ${\bf {\eta}}$ are roughly a half-wavelength
wide, and $\eta_0=0$.

A simple 2D example illustrates the performance of this type of
PML. The physical 
domain is a 1.8 x 7.6 km; the same domain is used in the experiments
reported in the next section. A point source is placed at $z$=40 m,
$x=3.3$ km, with a Gaussian derivative time dependence with peak
amplitude at about 5 Hz, and signifcant energy at 3 Hz but little below. The acoustic velocity is 1.5 km/s throughout
the domain, so the effective maximum wavelength is roughly 500 m. The
density is also constant, at 1 g/${\rm cm}^3$. A
snapshot of the wavefield at 1.2 s after source onsiet
(Figure \ref{fig:frame13}), before the wave has reached the boundary of the
domain, shows the expected circular wavefront. At 4.0 s, a simulation
with zero-pressure boundary conditions on all sides of the physical
domain produces the expected reflections, Figure \ref{fig:frame40-1}. With
PML zones of 250 m on the bottom and sides of the domain, so that only
the top is a zero-pressure surface, and $\eta_0=1$, the wave and its
free-surfacec ghost both appear to leave the domain
(Figure \ref{fig:frame40-2}, plotted on the same grey scale). The
maximum amplitude visible in Figure \ref{fig:frame40-1} is roughly
$7.1 \times 10^{-2}$, whereas the maximum amplitude in Figure
\ref{fig:frame40-2} is $7.0 \times 10^{-5}$. The actual reflection
coefficient is likely less than $10^{-3}$, as the 2D free space field
does not have a lacuna behind the wavefront, but decays smoothly, so
the low end of the wavelet spectrum remains.

It is not possible to decrease the PML layer thickness much beyond the
nominal longest half-wavelength and enjoy such small
reflections. Figure \ref{fig:frame40-3} shows the field at 4.0 s with
PML zones of width 100 m on bottom and sides, and an apparently
optimal choice of $\eta_0$. The maximum amplitude is $2.3 \times
10^{-4}$, and a reflected wave is clearly visible at the same grey scale.

\section{All FD Schemes are First Order in Heterogeneous Media}

The bulk modulus and buoyancy models
depicted in Figures \ref{fig:bm1} and \ref{fig:by1} embed an anticline or dome in an otherwise
undisturbed package of layers. These
figures display sampled versions of the models with $\Delta x = \Delta
z = $ 5 m; the model fields are actually given analytically, and can
be sampled at any spatial rate. The IWAVE utility $sfstandardmodel$ (in the
Madagascar {\tt bin} directory) builds this example and a number of
others that can be sampled arbitrarily for grid refinement
studies. See its self-doc for usage instructions.

\cite{SymesVdovina:09} use the model depicted in Figures \ref{fig:bm1} and \ref{fig:by1} to illustrate the {\em
  interface error} phenomenon: the tendency, first reported by
\cite{Brown:84}, of all finite difference schemes for wave
propagation to exhibit first order error, regardless of formal order,
for models with material parameter discontinuities. 
Figure \ref{fig:data1} exhibits a shot gather, computed with a (2,4) (= 2nd order in time,
4th order in space) staggered grid scheme, $\Delta
x = \Delta z = $ 5 m (more than 20 gridpoints per wavelength at the
wavelength corresponding to the highest frequency, 12 Hz, with
significant energy, and the smallest $v_p=1.5$ km/s)  and an appropriate near-optimal time step, acquisition geometry as described in
caption. The same gather computed at different spatial sample rates
seem identical, at first glance, however in fact the sample rate has a considerable effect. Figures
\ref{fig:trace} and \ref{fig:wtrace} compare traces computed from models sampled
at four different spatial rates (20 m to 2.5 m), with proportional
time steps. The scheme used is formally 2nd order
convergent like the original 2nd order scheme suggested by
\cite{Vir:84}, but has better dispersion suppression due to the use of
4th order spatial derivative approximation. Nonetheless,
the figures clearly show the first order error, in the form of a
grid-dependent time shift, predicted by \cite{Brown:84}.

Generally, even higher order approximation of spatial derivatives
yields less dispersive propagation error, which dominates the finite
difference error for smoothly varying material models. For
discontinuous models, the dispersive component of error is still
improved by use of a higher order spatial derivative approximation,
but the first order interface error eventually dominates as the grids
are refined. Figure \ref{fig:data8k1} shows the same shot gather as
displayed earlier, with the same spatial and temporal sampling and
acquisition geometry, but computed via the (2,8) (8th order in space)
scheme. The two gather figures are difficult to disinguish. The trace
details (Figures \ref{fig:trace8k}, \ref{fig:wtrace8k}) show clearly
that while the coarse grid simulation is more accurate than the (2,4)
result, but the convergence rate stalls out to 1st order as the grid
is refined, and for fine grids the (2,4) and (2,8) schemes produce
very similar results: dispersion error has been suppressed, and what
remains is due to the presence of model discontinuities.

See
\cite{SymesVdovina:09} for more examples, analysis, and discussion,
also \cite{FehlerKeliher:11} for an account of consequences for quality control in
large-scale simulation.

Note that the finest (2.5 m) grid consists of roughly 10 million
gridpoints. Consequently the modeling runs collectively take a
considerable time, from a minutes to a substantial fraction of an hour
depending on platform,
on a single thread. This example is computationally large enough that
parallelism via domain decomposition is worthwhile. IWAVE is designed
from the ground up to support parallel computation; a companion report
will demonstrate parallel use of IWAVE.

\section{Creating the examples - running IWAVE applications}
IWAVE builds with SConstruct ({\tt http://www.scons.org}), either as an
independent package or as part of Madagascar
\cite[]{Madagascar}. See the Madagascar web site 
\begin{verbatim}  
http://www.ahay.org/wiki/Main_Page
\end{verbatim}
for download and install
instructions. Source for IWAVE and other TRIP software reside in the {\tt
  trip} subdirectory of the top-level Madagascar source directory. A
{\tt README} file describes how to install TRIP software independently
of the rest of Madagascar, which is useful to configuring TRIP
differently from other parts of the package (for example, with MPI support).

The IWAVE acoustic staggered grid modeling command is
{\tt sfasg} for the Madagascar build, stored in the Madagascar {\tt
  \$RSFROOT/bin} directory, or 
\begin{verbatim}
$RSFSRC/trip/iwave/asg/main/asg.x
\end{verbatim}
for the standalone build. All
IWAVE commands self-document: entering the command path prints a usage
statement to the terminal, including descriptions of all parameters.

The paper you are currently reading follows the reproducible research pattern described on
the Madagascar web site, using Madagascar reproducible research tools. You can find the LaTeX source in the
subdirectory {\tt book/trip/asg} of the Madagascar source directory,
and the script for building the data in 
\begin{verbatim}
$RSFSRC/book/trip/asg/project/SConstruct
\end{verbatim} 
This script, together with the
self-doc for the acoustic staggered grid command and the remarks in
the remainder of this section, should enable you to
build your own examples after the pattern used in this project.

IWAVE applications currently expect model data files in the RSF
format of Madagascar \cite[]{Madagascar}. Data
from other sources will need conversion to this format. An RSF data
set consists of two files, an ascii header (grid metadata) file and a
flat binary data file. The data set is referenced by the header file
name; one of the parameters listed in the header file is the pathname
of the binary data file, with key {\tt in}. The header file is small
and easily created by hand with an editor, if necessary. Madagascar
commands add processing history information to header files, and
modify their parameters. By convention, the last value of a parameter
({\tt key=value} pair) appearing in the file is the current value. Many archival
data formats make the grid sample values available as a flat binary
file - this is true for instance of the gridded models output by GOCAD
({\tt http://www.gocad.org}), for which the {\tt vo} files contain virtually
the same information as (so may easily be translated to) RSF header
files in ascii form, and the {\tt vodat} files are flat binary files
which may be used unaltered as RSF binary files.

IWAVE uses two extensions of the Madagascar RSF standard. The first is
the optional inclusion of the {\tt dim} and {\tt gdim} keywords. These
permit IWAVE applications to treat an RSF file image as defining a {\tt
  gdim} dimensional data hypercube divided into {\tt dim} dimensional
slices. The second is the axis identification keyword set, {\tt id1}, {\tt id2},
etc.: these supply information on the physical meaning of various axes. For an IWAVE {\tt dim} space-dimensional modeling problem, axes labeled {\tt
  id1},...,{\tt id[dim-1]} are the spatial grid axes. If {\tt gdim} >
{\tt dim}, then {\tt id[dim]} labels the time axis, and {\tt id[n]},
{\tt n > dim}, axes other than those of space-time. The IWAVE
structure paper \cite[]{trip14:struct} explains the use of the
additional keywords in more detail.

An example of this construction appears in the script that builds the
PML examples above, which are actually frames of movies. The output of
the 2D simulations are 3D RSF files ({\tt gdim=3, dim=2}) with {\tt id3=2},
that is, the third axis is treated as time. Madagascar applications
ignore these keywords: in particular, you can view the 3D RSF simulation
output as a movie using
{\tt sfgrey} and {\tt xtpen} as usual. The presence of the additional keywords is necessary in order
for IWAVE to correctly interpret the data geometry.

This example illustrates another important feature of IWAVE
applications: any output data files must exist prior to execution -
their data samples are overwritten. The {\tt SConstruct} for this
project uses {\tt sfmakevel} to create the movie output files and {\tt
  sfput} to add the IWAVE-specific keywords to the headers, before
invoking the IWAVE command.

By IWAVE convention, the dimension of the problem is that of the primary
model grid. In the acoustic staggered grid application, the primary
model grid is that associated with the bulk modulus data. This grid is also the primary grid of the
simulation: that is, the space steps used in the finite difference
method are precisely those of the bulk modulus data.
Thus the choice of simulation grid is made externally to IWAVE.

The IWAVE acoustic application uses specific internal scales - m/ms
for velocity, g/cm$^3$ for density, and corresponding units for other
parameters. To ensure that data in other (metric) units are properly
scaled during i/o, the RSF header file may specify a value for the {\tt scale}
key, equal to the power of 10 by which the data should be multiplied
on being read into the application, to convert to the internal
scale. For example, if velocities are given in m/s, the header file
should include the line {\tt scale = -3}. In forthcoming releases,
this device will be deprecated in favor of explicit unit
specifications.

The current release is configured to use Seismic Unix (``SU'') (SEGY without reel header)
format for trace data i/o. Units of
length and time are m and ms respectively, consistent with other
internal unit choices. Two peculiarities of which the user should be
aware: (i) {\em receiver}
coordinates ({\tt gx}, {\tt gy}, and {\tt gelev} keywords) {\em always} specify trace location , that is, the
location at which values are sampled in space-time,
and (ii) on input, traces are regarded as point sources, so that each
trace multiplies a discrete spatial delta (hence values are scaled by the
reciprocal grid cell volume). Both of these design choices stem from the
migration (adjoint modeling) and inversion uses of IWAVE, discussed
for example in \cite[]{GeoPros:11,trip14:struct}.

Source traces must be modified to conform to this rubric. The {\tt
  sftowed\_array} application relieves the user of the necessity to
manually adjust the headers of an SU file containing source traces. It
accepts three arguments: (i) an input source source file containing {\tt gx}, {\tt gy}, and {\tt gelev}
values representing source trace location {\em relative} to a source
center location - the 
source coordinates of source traces are ignored; (ii)
a data file whose {\tt sx}, {\tt sy}, and {\tt selev} values are the
source center locations to be used - its receiver coordinates are ignored, and (iii) an output file (name), 
to which output source traces will be written, each with {\em source}
coordinates equal to those of a data trace, and {\em receiver}
coordinates equal to the sums of the source trace {\em receiver}
coordinates and the data trace {\em source} coordinates. The result is
a collection of source coordinate gathers with the same source
coordinates as the data file, but within each gather the same receiver
coordinates {\em relative} to the source coordinates as the source
file. Thus the source array is translated to each of the source
centers specified in the data file. Because the source file may
contain arbitrarily many traces with arbitrary relative locations, any
source radiation pattern may be approximated \cite[]{SantosaSymes:00}.

The example scripts in the {\tt project}
subdirectory use Madagascar
commands to create these prototype trace files.

One of IWAVE's design criteria is that acquisition geometry parameters
should have no {\em a priori} relation to the computational grid geometry:
source and receiver locations may be specified anywhere in Euclidean
space.

\section{Discussion and Conclusion}
The rather large and only slowly disappearing error revealed by the
examples from \cite{SymesVdovina:09} suggests strong limits for the
accuracy of regular grid finite difference methods. Finite element
methods suffer from the same limitations: accurate solution of
acoustodynamic or elastodynamic problems appears to demand
interface-fitted meshed \cite[]{Cohen:01}, with the attendant increase
in code and computational complexity.

The situation may not be so bleak, however. For one special case,
namely constant density acoustics, 
\cite{Terentyev:09} show that a regular grid finite difference method,
derived from a regular grid Galerkin finite element method, has
accuracy properties one would expect in homogeneous media (second
order convergence, reduction of grid dispersion through higher order
space differencing) even for discontinuous models: the interface error
effect is attenuated. This type of result actually goes quite far back
in computational geophysics (see for example \cite{Muiretal:92}),
though theoretical support has been slower in coming.

Pure regular grid methods cannot take advantage of changes in average
velocity across the model, and concommitant changes in
wavelength. Coupling of local regular grids is possible, however, and
can yield substantial computational efficiency through grid coarsening
in higher velocity zones - see \cite{moczoetal:06}. IWAVE already
accommodates multiple grids (in domain decomposition parallelism), and
extension to incommensurable multiple grids would be a significant
change, but in principle
straightforward. The use of logically rectangular but
geometrically irregular (``stretched'') grids is completely
straightforward, on the other hand.

These and other extensions, both past and future, are eased by the
reusability designed into the IWAVE core framework. This design has produced
reasonably well-performing and easy-to-use applications, and has proven
extensible to new models and schemes. Moreover, as explained by
\cite{GeoPros:11}, the object-oriented design of IWAVE dovetails with
similarly designed optimization software to support the construction
of waveform inversion software. The inversion applications resulting
from this marriage inherit the features of IWAVE - parallel execution,
high-order stencils, efficient boundary conditions, simple job control
- without requiring that these aspects be reworked in the code extensions.

\section{Acknowledgements}
IWAVE has been a team effort: the original design of the core
framework is due to Igor Terentyev, and Tanya Vdovina, Dong Sun, Marco
Enriquez, Xin Wang, Yin Huang, Mario Bencomo, and Muhong Zhou have each made major contributions.
Development of IWAVE was supported by the SEG Advanced Modeling (SEAM)
project, by the National Science Foundation under awards 0620821 and
0714193, and by the sponsors of The Rice Inversion Project. The IWAVE
project owes a great deal to several open source seismic software
packages (Seismic Un*x, SEPlib, Madagascar), debts which we gratefully
acknowledge. The author wishes to record his special gratitude to
Sergey Fomel, the architect of Madagascar, for his inspiring ideas and
his generous and crucial help in the integration of IWAVE into Madagascar.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/dso1d/paper.tex
repro docs dir added

\title{Why it works}
\author{William. W. Symes \thanks{The Rice Inversion Project,
Department of Computational and Applied Mathematics, Rice University,
Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}

\lefthead{Symes}

\righthead{Why it works}

\maketitle
\begin{abstract}
An extremely simple example shows how extended modeling plus variable projection plus a suitable annihilator lead to constructive velocity updates.
\end{abstract}

\section{Introduction}
Many of the extension concepts introduced to combat cycle-skipping in FWI sound good, and appear to work at least to some extent as one might hope from their heuristic justifications. However very few of these approaches have been assured of success via mathematical argument: in essence, they are mostly justified only ``in the rear-view mirror'', with no assurance that failure is not just around the corner, at the next example. On top of that, some of these approaches, for example those based on the computationally attractive Variable Projection Method (``VPM'') of \cite{GolubPereyra:03} have attributes such as gradient formulae cast in such form that the reasons for success are not readily apparent.

This note shows exactly how VPM leads to successful velocity updates for a particular extension of a very simple inverse problem, which asks that a homogeneous velocity model be deduced from one trace at known offset. I put this computation forward not because there are not simpler ways of answering the question it poses - there certainly are - but because the formal ingredients of waveform-based velocity estimation in this very simple setting are common to many similar extended inversion algorithms, and because in this case every computation can be done analytically, nearly to completion. In particular, it becomes clear why the VPM gradient formula produces a constructive update, despite not incorporating explicitly the extension measure (``annihilator'') central to the approach.

\section{Theory}
Suppose that a single trace is recorded, at distance $r$ from a source, and both source and receiver locations far enough from physical boundaries that the recorded wave has effectively propagated in free space.

Assume acoustic propagation, constant density, and isotropic point source and receiver. With a single trace, at best the average velocity in a region between source and receiver can be estimated, so assume that the velocity is constant. Then the modeling operator depends on $v$ and maps the source pulse $f$ to the predicted trace $S[v]f$ by
\begin{equation}
\label{eqn:mod}
S[v]f (t)= \frac{1}{4\pi r}f\left(t-\frac{r}{v}\right).
\end{equation}
Ignoring amplitude, this map implements a $v$-dependent time shift, which is everyone's favorite example to show why FWI cycle-skips and get stuck.

An obvious extended model for this example is $\{v,f\}$, that is, pairs of velocity and source wavelet, with the extended modeling operator already defined by equation \ref{eqn:mod}. With an annihilator $A$, to be chosen below, the penalty form of extended inversion is: minimize over $\{v,f\}$
\begin{equation}
\label{eqn:obj}
J[v,f] = \frac{1}{2}(\|S[v]f-d\|^2 + \alpha^2 \|Af\|^2).
\end{equation}
VPM \cite[]{GolubPereyra:03} takes advantage of $J$ being quadratic in $f$ to solve for $f$ given $v$, thus producing a reducted objective of $v$ alone:
\begin{equation}
\label{eqn:red}
J_{\rm red}[v] = \min_f J[v,f] = J[v,f[v]],
\end{equation}
where $f[v]$ is the minimizer of $J$ over $f$ for given $v$.
For this example, $J_{\rm red}$ is simple to compute. First observe that apart from amplitude, $S[v]$ is unitary:
\begin{equation}
\label{eqn:tran}
S[v]^T g (t) = \frac{1}{4\pi r}g\left(t+\frac{r}{v}\right)
\end{equation}
so
\begin{equation}
\label{eqn:unit}
S[v]^TS[v] f (t) = \frac{1}{(4\pi r)^2} f(t).
\end{equation}
Therefore the normal equation for the minimizer on the RHS of equation \ref{eqn:red} is
\begin{equation}
\label{eqn:norm1}
\left(\frac{1}{(4\pi r)^2} I + \alpha^2 A^TA\right)f = S[v]^Td.
\end{equation}

At this point I have to come clean about the actual choice of $A$. As usual with source extensions, the proper annihilator depends on modeling assumptions, not fundamental physics. Assume that signature decon has been applied and the target source $f_*$ is a zero-phase bandpass filter, reasonably well focused at time $t=0$. Since there is no spatial structure to exploit in this hyper-simple problem, we construct $A$ to penalize energy away from $t=0$:
\begin{equation}
\label{eqn:ann}
Af(t) = tf(t).
\end{equation}
Then the normal operator on the LHS of \ref{eqn:norm1} is simply multiplication by a positive function of time, and can be inverted by inspection. Using the identity \ref{eqn:tran} for the adjoint operator, obtain
\begin{equation}
\label{eqn:soln} 
f[v](t) = \left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2\right)^{-1}\frac{1}{4\pi r}d\left(t+\frac{r}{v}\right)  
\end{equation} 
Suppose that the data is noise-free: 
\begin{equation}
\label{eqn:nonoise}
d(t) = S[v_*]f_*(t) = \frac{1}{4\pi r}f_*\left(t-\frac{r}{v_*}\right)
\end{equation}
Then
\begin{equation}
\label{eqn:solnnon}
f[v](t) = \left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2\right)^{-1}\frac{1}{(4\pi r)^2}f_*\left(t-\frac{r}{v_*}+\frac{r}{v}\right) 
\end{equation}
so the residual is, after a little algebra,
\begin{equation}
\label{eqn:resid}
(S[v]f[v]-d)(t) =\frac{1}{4\pi r}[ (1+(4\pi r)^2 \alpha^2 (t-r/v)^2)^{-1} - 1]f_*(t-r/v_*)
\end{equation}
The gradient of $J_{\rm red}$ is given by the famous VPM formula (well-known to those who know it well):
\begin{equation}
\label{eqn:grad}
\nabla J[v] = (DS[v]f[v])^*(S[v]f[v]-d)
\end{equation}.
In this formula, $DS[v]$ is the derivative with respect to $v$ - since we can make explicit use of the Green's function, easy to see that
\begin{equation}
\label{eqn:deriv}
(DS[v]\delta v)f (t) = \frac{\delta v}{4\pi v^2}\frac{df}{dt}(t-r/v) = S[v](Q[v]\delta v)f (t),
\end{equation}
where 
\begin{equation}
\label{eqn:defq}
(Q[v]\delta v)f = \frac{r \delta v}{v^2} \frac{df}{dt}.
\end{equation}
That is, $Q[v]\delta v$ is a skew-adjoint operator depending linearly on $\delta v$ - more on this below.
The adjoint $(DS[v]f)^*$ is the adjoint of the map from model space to data space:
\[
(DS[v]f): \delta v \mapsto (DS[v]\delta v)f
\]
and is NOT the same as the adjoint denoted by $S[v]^T$, which is the data-space-to-data-space adjoint (got that?).
That is, this adjoint make this relation true, for all $\delta v,f,$ and $d$:
\begin{equation}
\label{eqn:vadj}
\delta v \cdot (DS[v]f)^*d = \int \,dt\,[(DS[v]\delta v)f] (t) d(t) 
\end{equation}
On the left side is the dot product in model space - since model space is just 1D in this example, that's just the numerical product. On the right is the dot product in data space, in the idealized continuum limit. Using equations \ref{eqn:deriv}, \ref{eqn:defq} and \ref{eqn:vadj},
\[
\delta v \cdot \nabla J[v] =  \int \,dt\,[(DS[v]\delta v)f[v]] (t) (S[v]f[v]-d)(t)
\]
\[
= \int \,dt\, [S[v](Q[v]\delta v)f[v](t)] (S[v]f[v]-d)(t) = \int \,dt\,(Q[v]\delta v)f[v](t) S[v]^T(S[v]f[v]-d)(t)
\]
Now invoke the normal equation \ref{eqn:norm1}: and replace the last factor:
\begin{equation}
\label{eqn:comm1}
= -\alpha^2\int \,dt\, (Q[v]\delta v)f[v](t)A^TAf[v](t)
\end{equation}
Since $Q[v]\delta v$ is skew-symmetric, shift it onto the other factor in this $L^2$ inner product (why not):
\[
= \alpha^2\int \,dt\, f[v](t)(Q[v]\delta v)A^TAf[v](t) 
\]
\begin{equation}
\label{eqn:comm2}
= \alpha^2\int \,dt\, (A^TAf[v](t)(Q[v]\delta v)f[v](t) + f[v](t)[(Q[v]\delta v),A^TA]f[v](t))
\end{equation}
where I swapped $Q$ and $A^TA$ at the cost of introducing a term involving the commutator $[Q,A^TA] = QA^TA-A^TAQ$, and rearranged the first term using the symmetry of $A^TA$. Now notice that this first term is exactly the same as the RHS of equation \ref{eqn:comm1}, except for the minus sign - so subtracting the RHS of \ref{eqn:comm2}
from \ref{eqn:comm1} and rearranging, get
\[
-\alpha^2\int \,dt\, (Q[v]\delta v)f[v](t)A^TAf[v](t) = \frac{1}{2}\alpha^2\int \,dt\,f[v](t)[(Q[v]\delta v),A^TA]f[v](t)
\]
hence 
\begin{equation}
\label{eqn:gradcomm}
\delta v \cdot \nabla J[v] = \frac{1}{2}\alpha^2\int \,dt\,f[v](t)[(Q[v]\delta v),A^TA]f[v](t)
\end{equation}
Remember that $A^TA$ amounts to multiplying by $t^2$, and $Q$ is the scaled time derivative (equation \ref{eqn:defq}), so
\begin{equation}
\label{eqn:comm3}
[(Q[v]\delta v),A^TA]=\frac{2r\delta v}{v^2}t
\end{equation}
Insert this identity into equation \ref{eqn:gradcomm} to obtain
\begin{equation}
\label{eqn:comm4}
\delta v \cdot \nabla J_{\rm red}[v] =  \delta v\frac{r \alpha^2}{v^2}\int \,dt\,tf[v]^2(t)
\end{equation}
Combine this identity with the formula \ref{eqn:solnnon} for the solution of the inner problem, and divide out the common factor $\delta v$, to obtain
\[
\nabla J_{\rm red}[v] = \frac{r\alpha^2}{v^2}\int \,dt\,t\left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2\right)^{-1}\frac{1}{(4\pi r)^2}f_*\left(t-\frac{r}{v_*}+\frac{r}{v}\right)^2
\]
\begin{equation}
\label{eqn:gradfinal}
= \frac{r\alpha^2}{v^2}\int \,dt\,\left(t+\frac{r}{v_*}-\frac{r}{v}\right)\left(1 + (4\pi r)^2\alpha^2 \left(t+\frac{r}{v_*}-\frac{r}{v}\right)^2\right)^{-1}f_*(t)^2
\end{equation}

Recall that the definition of $A$ (multiplication by $t$) was justified by a modeling assumption, that the actual source is (possibly after designature) highly localized at $t=0$ - ideally a Dirac delta, but because of bandlimitation not quite. Still, a high degree of localization, as implicit in this choice of annihilator, means that the integral above is approximately proportional to the integrand at $t=0$, that is, 
\[
\approx \frac{r\alpha^2}{v^2}\left(\frac{r}{v_*}-\frac{r}{v}\right)\left(1 + (4\pi r)^2\alpha^2 \left(\frac{r}{v_*}-\frac{r}{v}\right)^2\right)^{-1}
\]
Note that to the extent that this approximation is accurate,
\begin{itemize}
\item if $v > v_*$, then $\nabla J_{\rm red}[v] > 0$, and
\item if $v < v_*$, then $\nabla J_{\rm red}[v] < 0$, and
\end{itemize}
That is, within the domain of validity of this approximation (setting $t=0$ in equation \ref{eqn:gradfinal}), {\em there are no local mins:} the gradient has the correct sign and velocity updates computed from it will be constructive.

On the other hand, careful examination of equation \ref{eqn:gradfinal} (which involves no approximations) shows that if $\alpha$ is too large, then the finite-frequency spread of $f_*$ is guaranteed to result in non-convex behaviour, since the side-lobes will dominate. This message is consistent with the import of many other similar computations.

\section{Discussion}
There remain two important points to be made.

First, the formal computations centering on the operator $Q$ (equations \ref{eqn:deriv} through \ref{eqn:gradcomm}) depend only on the relation \ref{eqn:deriv} and the skew-symmetry of $Q$, which hold with minor modifications for other more complex waveform inversion problems. These other more complex problems do not submit to such a simple treatment as is shown in the equations following \ref{eqn:gradcomm}, but for example it is possible in some cases to use the relation \ref{eqn:deriv} to extract a relation between extended waveform inversion and traveltime tomography, via analysis of the Hessian at a zero-residual global minimizer. See for instance \cite{tenKroode:IPTA14,Symes:IPTA14,Symes:Madrid}. Up to that point the reasoning is quite general, and central to the understanding gained so far of extended inversion methods.

The cited papers are set in the context of subsurface-offset extended Born inversion, and demand a single scattering approximation and high-frequency asymptotics - that is, microlocal anallysis. Somewhat surprisingly, it has recently proven possible to establish the existence of a variant of the operator $Q$ for acoustic problems with transmission configuration but without the assumed smoothness of the material parameter fields essential for microlocal analysis. That is, at least in some circumstances $DS=SQ$ with approximately skew-symmetric $Q$, even when reflections (even multiple) occur in the material model. It remains to be seen whether enough can be established about these more general versions of $Q$ (which are for example in general not pseudodifferential0 to gain any hold over the corresponding inverse problems.

The second point to be made has to do with the necessity of nested optimization, of which VPM is a variant. The aim is to find a minimizer of the bivariate objective $J[v,f]$ (and ultimately of data fitting, that is, FWI), but the bivariate problem is computationally intractable. The framework developed here can be used to explain why this is so: optimization in $v$ and $f$ simultaneously is very ill-conditioned (in the continuum limit, infinitely so), whereas reduced objetives (of which VPM produces one possibility) are much easier to optimize. Yin Huang's thesis \cite[]{YinHuang:16} shows this contrast between global and nested optimization very clearly, using a different extension.

\section{Conclusion}
Desipite its simplicity, the single-trace inversion transmission problem proves typical of many more complex waveform inversion problems. The structure of the derivative is similar in many of these problems, and for the particularly simple one explained here, can be analysed on paper to the point of showing explicitly why this approach to waveform inversino works.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/dton/paper.tex
repro docs dir added

\title{Waveform Inversion via Blended Source Extension}
\author{William. W. Symes \thanks{The Rice Inversion Project,
Department of Computational and Applied Mathematics, Rice University,
Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}

\lefthead{Symes}

\righthead{Blended Source Approximate Inverse}

\maketitle
\begin{abstract}
Define the blended source extension, introduce annihilator, waveform inversion based on it
and variable projection.
Explain how approximate inverse via energy decay  provides a
substitute for the $\Psi$DO property of the normal operator for smooth
medium transmission problems.
\end{abstract}
\section{Introduction}
\section{Acoustics - sketch}
Acoustic system with pressure source:
\begin{eqnarray}
\label{eqn:awe}
\gamma \frac{\partial p}{\partial t} & = & - \nabla \cdot \bv +
F, \nonumber \\
\rho \frac{\partial \bv}{\partial t} & = & - \nabla p + {\bf G},\nonumber \\
p & =& p_0 \mbox{ for } t= t_0\nonumber\\ 
\bv & = & \bv_0 \mbox{ for } t= t_0 
\end{eqnarray}

$\Omega$  is a smoothly bounded open set in $\bR^d$. Denote by
$T^{\rm int}_{\Omega}$ and $T^{\rm ext}_{\Omega}$ the traces on $\partial
\Omega$ from interior and exterior of $\Omega$, and use the same notation
for functions of time. These maps are a priori defined with domains
$C^{\infty}_0(\Omega \times \bR)$ and $C^{\infty}_0((\bR^d\setminus
\Omega) \times \bR)$ and ranges $C^{\infty}_0(\partial \Omega \times \bR)$.

For the moment, assume that $p_0=0, \bv_0=0$. This restriction will be
removed later. The initial system can then be re-stated as
\begin{eqnarray}
\label{eqn:ichom}
\gamma \frac{\partial p}{\partial t} & = & - \nabla \cdot \bv +
F, \nonumber \\
\rho \frac{\partial \bv}{\partial t} & = & - \nabla p + {\bf G},\nonumber \\
p & = & 0, t < 0\\
\bv & = & 0, t < 0 
\end{eqnarray}

The system \ref{eqn:ichom} takes the form described in
\cite{BlazekStolkSymes:13}, with $P(\nabla) (p,\bv) = (-\nabla p,
-\nabla \cdot \bv)$ (use $P(\nabla)$ here rather than $p(\nabla)$ as
in \cite{BlazekStolkSymes:13} to avoid conflict with the notation for
the pressure field. As shown in Appendix A of
\cite{BlazekStolkSymes:13}, $P(\nabla)$ is densely defined and skew adjoint  on $L^2(\Omega)^{d+1}$ with domain
$H^1_0(\Omega) \times H^1_{\rm div}(\Omega)$. Therefore
Theorem 1 of  \cite{BlazekStolkSymes:13} applies to the system
\ref{eqn:ichom}: if $(F, {\bf G}) \in H^1_{\rm loc}(\bR,(L^2(\Omega))^{d+1})$
and $= 0$ for $t<0$, then there exists a unique strong solution
$(p,\bv) \in C^1(\bR,(L^2(\Omega))^{d+1}) \cap C^0(\bR,  H^1_0(\Omega) 
\times H^1_{\rm div}(\Omega))$.

The same reasoning applies to the exterior domain $\bR \times (\bR^d
\setminus \Omega)$, to yield a unique strong solution with similar conditions.

The trace operators $\Ti$ and $\Te$ extend to domains $C^k(\bR,
H^1(\Omega))$ and $C^k(\bR,H^1_{\rm loc}(\bR,\bR^d \setminus
\Omega))$
and range $C^k(\bR,L^2(\partial \Omega))$, $k \ge 0$, per the standard trace
theorem for Sobolev spaces. Define the interior (exterior) Dirichlet
problem for $d \in C^0(\bR,L^2(\partial \Omega))$ by

\begin{thm}
  Suppose that
  $d \in H^1([0,T], H^1(\partial \Omega)) \cap H^2([0,T],L^2(\partial
  \Omega))$. Then there exist unique 
$(\ip,\ibv) \in C^1(\bR,(L^2(\Omega))^{d+1}) \cap C^0(\bR,  H^1(\Omega) 
\times H^1_{\rm div}(\Omega))$.
for which
\begin{eqnarray}
\label{eqn:iawebv}
\gamma \frac{\partial \ip}{\partial t} & = & - \nabla \cdot \ibv, \nonumber \\
\rho \frac{\partial \ibv}{\partial t} & = & - \nabla \ip,\nonumber \\
\Ti \ip & = & d \\
\ip & = & 0, t < 0\\
\ibv & = & 0, t < 0.
\end{eqnarray}
Also, there extst unique 
$(\ep,\ebv) \in C^1(\bR,(L^2(\bR^d \setminus \Omega))^{d+1}) \cap C^0(\bR,  H^1(\bR^d \setminus \Omega) 
\times H^1_{\rm div}(\bR^d \setminus \Omega))$.
for which
\begin{eqnarray}
\label{eqn:eawebv}
\gamma \frac{\partial \ep}{\partial t} & = & - \nabla \cdot \ebv, \nonumber \\
\rho \frac{\partial \ebv}{\partial t} & = & - \nabla \ep,\nonumber \\
\Te \ep & = & d \\
\ep & = & 0, t < 0\\
\ebv & = & 0, t < 0.
\end{eqnarray}
\end{thm}
\begin{proof}
Since $\Omega$ is precompact, it is possible to construct a collar
neigborhood of $\partial \Omega$ in which a finite number of local
coordinate systems may be defined, covering $\partial \Omega$. Using
these coordinate systems, it is possible to construct an extension of 
$d \in H^2_{\rm loc}(\bR,L^2_0(\bR^d)) \cap
H^1_{\rm loc}(\bR,H^1_0(\bR^d))$. For this extension,
\[
\gamma\frac{\partial d}{\partial t} \in H^1_{\rm
  loc}(\bR,L^2_0(\bR^d)), \,
\nabla d \in H^1_{\rm
  loc}(\bR,L^2_0(\bR^d))^d
\]
Define $\ip_1, \ibv_1$ to be the strong solutions of the system
\ref{eqn:ichom} in $\bR \times \Omega$, respectively $\bR \times \bR^d
\setminus \Omega$, with 
\[
F = -\gamma\frac{\partial d}{\partial t}, {\bf G} = -\nabla d.
\]
Then $\ip = \ip_1 + d, \ibv = \ibv_1$ is the unique strong solution of
system \ref{eqn:iawebv}, $\ep = \ep_1 + d, \ebv = \ebv_1$ the unique
strong solution of system \ref{eqn:eawebv}.
\end{proof}

Note that
\begin{equation}
\label{eqn:trace}
T^{\rm  int}\ip = T^{\rm ext}\ep = d.
\end{equation}

Denote by $\bn$ the outward unit normal on $\partial \Omega$. note
that $\bn \cdot \Ti$ extends continuously to  $C^0(\bR,H_{\rm
  div}(\Omega))$. Similarly,
$\bn \cdot \Te$ extends continuously to $C^0(\bR, H_{\rm div}(\bR^d \setminus \Omega))$.
Therefore Theorem 2 implies that if $d \in H^1([0,T], H^1(\partial \Omega)) \cap H^2([0,T],L^2(\partial
  \Omega))$
\begin{equation}
\label{eqn:dton}
D[\kappa,\rho]d =[v_n] = \Te \ebv - \Ti \ibv)
\end{equation}
is well defined.

\begin{thm}
Suppose that $d \in H^1([0,T], H^1(\partial \Omega)) \cap H^2([0,T],L^2(\partial
  \Omega))$, and denote by $(\ip,\ibv)$ and $(\ep,\ebv)$ solutions of
  \ref{eqn:iawebv} and \ref{eqn:eawebv} respectively. Define 
\begin{equation}
\label{eqn:both}
(p,\bv)= \left\{
\begin{array}{c}
(\ip,\ibv) \mbox{ in } \Omega \times [0,T]\\
(\ep,\ebv) \mbox{ in } (\bR^d \setminus \Omega) \times [0,T] 
\end{array}
\right.
\end{equation}
Then $(p,\bv)$ is a weak solution of 
\begin{eqnarray}
\label{eqn:awe1}
\gamma \frac{\partial p}{\partial t} & = & - \nabla \cdot \bv +
D[\kappa,\rho]d \delta_{\partial \Omega}, \nonumber \\
\rho \frac{\partial \bv}{\partial t} & = & - \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t= 0\nonumber\\ 
\bv & = & 0 \mbox{ for } t= 0
\end{eqnarray}
\end{thm}

\begin{proof}: (sketch) choose smooth test functions $\phi, {\bf \psi}$ vanishing for $t=T$. Then
\[
\int_{\Omega \times [0,T]}\left (\gamma \frac{\partial \phi}{\partial t} +
\nabla \cdot {\bf \psi}\right)p + \left(\rho\frac{\partial {\bf
    \psi}}{\partial t}
  + \nabla \phi\right) \cdot \bv
\]
\[
= - \int_{\Omega \times [0,T]}\left (\gamma \frac{\partial p}{\partial t} +
\nabla \cdot \bv\right) \phi + \left(\rho\frac{\partial \bv}{\partial t}
  + \nabla p\right) \cdot {\bf \psi}
\]
\begin{equation}
\label{eqn:rhs1}
+ \int_{\partial \Omega \times [0,T]} \bn \cdot {\bf \psi} \ip + \bn
  \cdot \ibv \phi
\end{equation}
Similarly,
\[
\int_{(\bR^d\setminus \Omega) \times [0,T]}\left (\gamma \frac{\partial \phi}{\partial t} +
\nabla \cdot {\bf \psi}\right)p + \left(\rho\frac{\partial {\bf
    \psi}}{\partial t}
  + \nabla \phi\right) \cdot \bv
\]
\[
= - \int_{(\bR^d \setminus\Omega) \times [0,T]}\left (\gamma \frac{\partial p}{\partial t} +
\nabla \cdot \bv\right) \phi + \left(\rho\frac{\partial \bv}{\partial t}
  + \nabla p\right) \cdot {\bf \psi}
\]
\begin{equation}
\label{eqn:rhs2}
- \int_{\partial \Omega \times [0,T]} \bn \cdot {\bf \psi} \ep + \bn
  \cdot \ebv \phi
\end{equation}
Adding equations \ref{eqn:rhs1} and \ref{eqn:rhs2} and taking into
account that $\ip$ and $\ep$ coincide (by construction) on $\partial
\Omega \times [0,T]$, obtain
\[
\int_{(\bR^d \times [0,T]}\left (\gamma \frac{\partial \phi}{\partial t} +
\nabla \cdot {\bf \psi}\right)p + \left(\rho\frac{\partial {\bf
    \psi}}{\partial t}
  + \nabla \phi\right) \cdot \bv
\]
\[
= \int_{\partial \Omega \times [0,T]} (\bn \cdot \ibv -  \bn
  \cdot \ebv) \phi
\]
$\phi$ and ${\bf \psi}$ being arbitrary, other than vanishing at $t=T$,
\end{proof}

Conversely,
\begin{thm}: Suppose $f \in  H^2([0,T],L^2(\partial \Omega)) \cap
  H^1([0,T],H^1(\partial \Omega))$, and $(p,\bv)$ is a weak solution of
\begin{eqnarray}
\label{eqn:awe2}
\gamma \frac{\partial p}{\partial t} & = & - \nabla \cdot \bv +
f \delta_{\partial \Omega}, \nonumber \\
\rho \frac{\partial \bv}{\partial t} & = & - \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t= 0\nonumber\\ 
\bv & = & 0 \mbox{ for } t= 0
\end{eqnarray}
Then $T^{\rm ext}_{\Omega} p = T^{\rm int}_{\Omega}p$.
% and $f =
%D[\kappa,\rho]T^{\rm int}_{\Omega}p$.
\end{thm}

\begin{proof}: (sketch) Choosing test functions $\phi,\psi$ as before, the
system \ref{eqn:awe2} means that
\[
\int_{(\bR^d \times [0,T]}\left (\gamma \frac{\partial \phi}{\partial t} +
\nabla \cdot {\bf \psi}\right)p + \left(\rho\frac{\partial {\bf
    \psi}}{\partial t}
  + \nabla \phi\right) \cdot \bv
\]
\[
= \int_{\partial \Omega \times [0,T]} \phi f
\]
Extend $\bn$ to a collar neighborhood of $\partial \Omega$ and choose
smooth $\chi$ supported in that neighborhood and $=1$ on $\partial
\Omega$. Also extend $f$ so that it remains in the same class. Set
\[
\bw = \chi f \bn \left({\bf 1}_{\Omega} - \frac{1}{2}\right)
\]
\[
\int (\nabla \cdot \bw)  \phi = \int_{\Omega \times [0,T]}(...) + \int_{(\bR^d
  \setminus \Omega) \times [0,T]}(...) 
\]
\[
= \int_{\partial \Omega \times [0,T]} \bn \cdot (\chi f \bn) \phi
\]
\[
= \int_{\partial \Omega \times [0,T]} f  \phi
\]
That is,
\[
 \nabla \cdot \bw = f \delta_{\partial \Omega \times [0,T]}
\]
Then $(p,\bv-\bw))$ is a weak solution of 
\begin{eqnarray}
\label{eqn:awe12}
\gamma \frac{\partial p}{\partial t} & = & - \nabla \cdot \bv \nonumber \\
\rho \frac{\partial (\bv-\bw)}{\partial t} & = & -\rho \frac{\partial\bw}{\partial t} - \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t= 0\nonumber\\ 
\bv & = & 0 \mbox{ for } t= 0
\end{eqnarray}
The RHS of the second equation in the system \ref{eqn:awe12} is of class $H^1([0,T],L^2(\bR^d))$ so Theorem 1 in
\cite{BlazekStolkSymes:13} implies that $(p, \bv-\bw)$ is a strong
solution of system \ref{eqn:awe12}, in particular that $p \in C^0([0,T],H^1(\bR^d))$,
whence the assertion about continuity at the boundary follows from the
trace theorem.

The remaining statement now follows from the previous theorem.
\end{proof}

\section{Ins and Outs}
Assume that $\kappa, \rho$ are of class $C^{\infty}$ in a neighborhood
of $\partial \Omega$.

Denote by $P(\bx,\omega,\bk)$ symbol of the acoustic system,
\[
P(\bx,\Omega,\bk) = 
i\left(
\begin{array}{cc}
\gamma(\bx)i\omega & \bk^T\\
\bk &  \rho(\bx)\omega I_{3\times3}
\end{array}
\right)
\]
Determinant is zero iff $\omega=0$ or $\omega^2 = c^2|\bk|^2$.
At a point $\bx$ on $\partial \Omega$, write $\bk = \bk^{\perp} +
k_n\bn$. The characteristic eqn has two solutions for $k_n$, namely
\[
k_n = \pm \sqrt{\frac{\omega^2}{c^2}-|\bk^{\perp}|^2}
\]
and correspondingly, the normal component of $\bv$ satisfies
\[
\rho v_n = \bn \cdot \bv = \int \bn \cdot \nabla p
\]
or in the Fourier domain 
\[
c \rho v_n = \frac{k_n}{\omega} p = \pm
\sqrt{1-\frac{c^2|\bk^{\perp}|^2}{\omega^2}}p
\]
Define $V$ to be the $\Psi$DO on $\partial \Omega \times \bR$ with symbol
\[
V(\bx,\bk^{\perp})=\frac{1}{c\rho}\sqrt{1-\frac{c^2|\bk^{\perp}|^2}{\omega^2}}
\]
Then if $(p,\bv)$ is incoming/outgoing, $T^{\rm int}v_n = \pm VT^{\rm int}p$ 
The incoming/outgoing subspace corresponds to $\pm$.
Note that if $(p, \bv^{\perp} + v_n\bn)$ is incoming, then $(p,
\bv^{\perp} - v_n\bn)$ is outgoing, and vis-versa.

Projections onto Incoming/outgoing data: $P_{\pm}$. Data is outgoing,
source incoming. So to extract data from overlapping signal with 
source, apply $P_-$. Then solution backwards in time has source field 
as its incoming component.

Decomposition of $(p,v_n)$: 
\[
(p,v_n) = (p_+, Vp_+) + (p_-,-Vp_-) = (p_++p_-, V(p_+-p_-)) \Rightarrow
\]
\[
(p,V^{-1}v_n) = (p_++p_-,p_+-p_-)
\]
\[
p_+ = \frac{1}{2}(p + V^{-1}v_n), p_-=\frac{1}{2}(p-V^{-1}v_n)
\]
\[
P_{\pm}(p,v_n) = \left( \frac{1}{2}(p \pm V^{-1}v_n),  \frac{1}{2}(Vp \pm
  v_n)\right)
\]

Ray theory assumption about exterior: all rays leaving $\partial
\Omega$ keep going, never to return. Then the outer solution
$(\ep,\ebv)$ defined in Theorem 1 is purely outgoing, i.e. 
\[
P_-T^{\rm ext}(\ep,\ebv) = T^{\rm ext}(\ep,\ebv), \,P_+T^{\rm ext}(\ep,\ebv) = 0
\]
hence
\[
v_n^{\rm ext} = -Vp 
\]
and
\begin{equation}
\label{eqn:did}
D[\kappa,\rho]p = v_n^{\rm int}-v_n^{\rm ext} = v_n^{\rm int}+Vp =
2 P_+(p,v^{\rm int}_n)_2
\end{equation}
that is, $D[\kappa,\rho]p$ is the normal v-component of the incoming projection of the
solution of system \ref{eqn:awe2}.
[Since $\ip=\ep$ on boundary, and initial conditions are same, follows
that the tangential components of $\ibv, \ebv$ are the same.]

Equation \ref{eqn:did} relates $D[\kappa,\rho]$ to the $\Psi$DO
$V$. Note however that in general it is not a $\Psi$DO itself, unless
$v_n=Vp$, that is, unless the entire field $(p,\bv)$ is incoming at
the boundary. This occurs under certain circumstances, but in general
both components must be computed, and the relation between $p$ and
$D[\kappa,\rho]p$ is global.

1D example.

\section{Linear Inverse Problem}
%Problem: Given uniformly positive, bounded, and measureable $\kappa,\rho$ and
%$d \in H^2([0,T],L^2(\partial \Omega)) \cap  H^1([0,T],H^1(\partial
%\Omega))$, find $f \in H^2([0,T],L^2(\partial \Omega)) \cap
% H^1([0,T],H^1(\partial \Omega))$ so that the solution $(p,\bv)$ of 
%the system \ref{eqn:awe2} satisfies $d = T^{\rm int} p$.

Suppose that $(p,\bv)$ is the solution of \ref{eqn:awe2} as described
in Theorem 2, and define 
\[
S[\kappa,\rho] f = T^{\rm int}p^{\rm int}.
\]
By standard uniqueness thm, if you record final data at
$t=T$, say $p_T(\bx)=p(\bx,T), \bv_T(\bx)=\bv(\bx,T)$ for $\bx \in
\Omega$, and $d = T^{\rm int}p$, then $(p,\bv)=(p_r,\bv_r)$, the
solution of the time-reversed problem
\begin{eqnarray}
\label{eqn:awer}
\gamma \frac{\partial p_r}{\partial t} & = & - \nabla \cdot \bv_r \nonumber \\
\rho \frac{\partial \bv_r}{\partial t} & = & - \nabla p_r \nonumber \\
p_r & =& p_T \mbox{ for } t= T\nonumber\\ 
\bv_r & = & \bv_T \mbox{ for } t= T \\
p_r & = & d \mbox{ on } \partial \Omega \times [0,T]
\end{eqnarray}
In general, suppose $d$ is given on $\partial \Omega \times [0,T]$,
and define
\[
S[\kappa,\rho]^{\dagger}d = v_{r,n} + Vp_r.
\]
If specifically $d = S[\kappa,\rho]f$, then the last section together
with Theorem 2 implies that
\begin{equation}
\label{eqn:idealinv}
S[\kappa,\rho]^{\dagger}S[\kappa,\rho] = I
\end{equation}

Implicitly, $S[\kappa,\rho]^{\dagger}$ depends on the final data
$(p_T,\bv_T)$. Suppose that local energy decays:

\noindent {\bf Local Decay Assumption:} There exist $C, \alpha > 0$ so
that the solution of system \ref{eqn:awe} with $F = {\bf G} = 0$
satisfies
\[
E(T) \le Ce^{-\alpha T} E(0),
\]
where $E$ is the acoustic strain energy:in $\Omega$:
\[
E(t) = \frac{1}{2}\int_{\Omega} dx \left(\rho |\bv(\bx,t)|^2 +
  \frac{1}{\kappa} p(\bx,t)^2\right).
\]
In that case, for any $\epsilon > 0$ there is $T>0$ so that $E(T) <
\epsilon \sup_{0 \le t \le T}E(t)$. Define $(p_a,\bv_a)$ to be the
solution of
\begin{eqnarray}
\label{eqn:awer}
\gamma \frac{\partial p_a}{\partial t} & = & - \nabla \cdot \bv_a \nonumber \\
\rho \frac{\partial \bv_a}{\partial t} & = & - \nabla p_a \nonumber \\
p_a & =& 0 \mbox{ for } t= T\nonumber\\ 
\bv_a & = & 0 \mbox{ for } t= T \\
p_a & = & d \mbox{ on } \partial \Omega \times [0,T]
\end{eqnarray}
$(p_a-p_r,\bv_a-\bv_r)$ satisfies a similar system with zero Dirichlet
data. Since the energy in the final data is small, it follows that the
predicted Neumann data is small. NOTE this requires a collar
neighborhood with smooth coefficients and microlocal restrictions on
the sources, so that restriction to the boundary acts like order zero op.

(then relation with adjoint: $S^{\dagger} = S^TW$)

Slightly less trivially, suppose that the boundary $\partial \Omega
\times [0,T]$ is divided into two subdomains, $\Gamma$ and its
complement $\Gamma'$, and $d$ is given only on $\Gamma$.

\section{Gradient}
\[
J[\kappa] = \frac{1}{2}\|AS[\kappa]^{\dagger}d\|^2=\frac{1}{2}\|AS[\kappa]^TWd\|^2
\]
Note that if $\kappa$ is not varied near wells then W is independent
of constant.
\[
DJ[\kappa]\delta \kappa= = \langle(DS[\kappa]\delta \kappa)^T Wd, A^TA 
S[\kappa]^TWd\rangle
\]
\begin{equation}
\label{eqn:deriv}
=\langle( Wd, DS[\kappa]\delta \kappa) A^TA S[\kappa]^TWd\rangle
\end{equation}
\begin{equation}
\label{eqn:grad}
=\langle DS[\kappa]^*(Wd,A^TAS[\kappa]^TWd\rangle_{\rm model}
\end{equation}
Here $DS[\kappa]^*(Wd,A^TAS[\kappa]^TWd)$ is the reverse time
migration of $Wd$ with source field $A^TA S[\kappa]^TWd$. The model
inner produce may incorporate a weight (usually should, eg. power of
the Laplacian, in which case its inverse postmultiplies the Gradient.

\section{Hessian}
Differentiating equation \ref{eqn:deriv} again,
\[
D^2J[\kappa](\delta \kappa_1,\delta \kappa_2)
= \langle( Wd, D^2S[\kappa](\delta \kappa_1,\delta \kappa_2) A^TA
S[\kappa]^TWd\rangle
\]
\[
+ \langle( Wd, DS[\kappa](\delta \kappa_1) A^TA DS[\kappa](\delta 
\kappa_2)^TWd\rangle. 
\]
Now suppose that $d = S[\kappa]f$ with physical $f$, that is
$Af=0$. Thus $AS[\kappa]^TWd \approx 0$ and the Hessian is
\begin{equation}
\label{eqn:hess1}
+ \langle A(DS[\kappa](\delta \kappa_1))^T Wd, A (DS[\kappa](\delta 
\kappa_2))^TWd\rangle. 
\end{equation}

 Also suppose that the factorization lemma holds:
\begin{equation}
\label{eqn:fact}
DS[\kappa]\delta \kappa \approx S[\kappa]Q[\kappa,\delta \kappa]
\end{equation}
with $Q$ an essentiall skew-adjoint $\Psi$DO of order 1.

Remark. It may not be necessary to assume this much: setting 
$Q = S^{\dagger}DS$ may do. That even works when geometric optics is
not valid. Need to better understand structure.

Then as noted elsewhere
\begin{equation}
\label{eqn:hess2}
D^2J[\kappa](\delta \kappa_1,\delta \kappa_2) 
\approx \langle [A,Q[\kappa,\delta \kappa_1]]S[\kappa]^TWd,
[A,Q[\kappa,\delta \kappa_2]]S[\kappa]^TWd\rangle
\end{equation}

\section{Asymptotics}
Assume 3D, single arrival. Then with source surface = $\{z=z_s\}$ and
similarly for receivers,
\[
S[\kappa]f(\bx_r,t) = \int dx_sdy_s a(\bx_r,\bx_s)
f(t-T(\bx_s,\bx_r),\bx_s)
\]
\[
(DS[\kappa]\delta \kappa)f(\bx_r,t) = -\int dx_sdy_s f'(t-T(\bx_s,\bx_r),\bx_s)
(DT[\kappa]\delta \kappa)(\bx_s,\bx_r)
\]
\[
S[\kappa]^T(DS[\kappa]\delta \kappa)f(\bx_s,t) = -\int dx_r dy_r \int
dx'_s dy'_s a(\bx_r,\bx_s) a(\bx_r,\bx'_s) f'(t+T(\bx_s,\bx_r) -
T(\bx_r,\bx'_s)) (DT[\kappa]\delta \kappa)(\bx_s',\bx_r)
\]
\[
=\int dx_r dy_r dx'_s dy'_s d\omega  a(\bx_r,\bx_s) a(\bx_r,\bx'_s) \hat{f}(\omega)e^{i\omega(t +T(\bx_s,\bx_r) -T(\bx_r,\bx'_s))}
\]
Inner ($\bx_r$) integral, stationary phase (for 3D, so integral is
over $x_r,y_r$:
\[
\approx
\sum
\frac{2\pi}{\omega}|\det\left(\nabla_{\bx_r}\nabla_{\bx_r}(T(\bx_s,\bx_r)-T(\bx_r,\bx_s'))\right)|^{-\frac{1}{2}}
  e^{i\omega(t +T(\bx_s,\bx_r) -T(\bx_r,\bx'_s))} e^{i\pi\sigma/4}(DT[\kappa]\delta \kappa)(\bx_s',\bx_r)
\]
$\sigma$ is the signature of the Hessian - take it to be zero. The sum is over $\bx_r$ satisfying the stationary phase condition
\[
\nabla_{\bx_r} T(\bx_s,\bx_r) = \nabla_{\bx_r}T(\bx'_s,\bx_r)
\]
which may be interpreted as stating that the ray from $\bx_s$ to
$\bx_r$ arrives at $\bx_r$ with the same tangential slowness vector as the ray
from $\bx_s'$ to $\bx_r$. Assuming that all such rays cross the
surface transversally, and using the eikonal equation to determine the
$z$ component, conclude that the slowness vectors are the same. Thus
these must be the same rays, hence $\bx_s$ and $\bx'_s$ must lie on
the same ray. Assuming that rays cross the source surface only once,
this implies that $\bx_s=\bx_s'$. 
\section{General Case}
Begin with an abstract hyperbolic system of the sort decribed by
\cite{BlazekStolkSymes:13}, equation (1):
\begin{equation}
\label{symmhyp}
a \frac{\partial u}{\partial t} + p(\nabla) u  = f\mbox{ in } \bR^d \times \bR; \, u = u_0 \mbox{ for } t = 0,
\end{equation}
without the lower-order terms considered in that paper. Here $a$ is a
bounded measurable positive definite symmetric matrix-valued time-independent function, and $p(\nabla)$ is
a skew-symmetric first order PDO.

Assume that 
$f$ is supported in an open domain $\Omega \in \bR^d$, and that $a$ is 
smooth in a neighborhood of  $\bR^d \setminus \Omega$.

The key assumption on $a$ is that it has the {\em local energy decay
  property}: there exist $C, \alpha >0$ so that for any $f \in
L^2_{\rm comp}(\Omega \times (-\infty,0)), u_0 \in L^2_{\rm comp}(\Omega)$,
\begin{equation}
\label{eqn:decay}
\|u(t)\|_{L^2(\Omega)} \le Ce^{-\alpha t} (\|u_0\|_{L^2(\Omega)} +
\|f\|_{L^2(\Omega \times \bR)}).
\end{equation}

D-to-N map: suppose $\Gamma \subset \bR^d$ is a smooth submanifold of
dimension $d-1$, and define the trace map $S[m]$ as in the discussion
around Theorem 2 in \cite{BlazekStolkSymes:13}. Let $V$ be the
subspace as defined there, and assume that $S[m]$ extends to $V$.

\section{Adjoint}
\begin{equation}
\label{eqn:dfwd1d}
\frac{p^{n+1}-p^n}{\kappa\Delta t} = D^T \bv^{n+1/2} + \frac{f^n}{\kappa} \delta_{z_s},\,\frac{\bv^{n+1/2}-\bv^{n-1/2}}{\beta \Delta t} = - Dp^n,\, p^0=0, \bv^{1/2}=\bf{0}
\end{equation}
\begin{equation}
\label{eqn:dadj1d}
\frac{q^{n+1}-q^n}{\kappa \Delta t} = D^T \bw^{n+1/2} + \frac{g^{n+1}}{\kappa} \delta_{z_r},\,\frac{\bw^{n+1/2}-\bw^{n-1/2}}{\beta \Delta t} = - Dq^n, q^N=0, \bw^{N-1/2}=\bf{0}
\end{equation}
\[
\sum^N_{n=1} \frac{\Delta t}{\kappa} p^n g^n|_{z=z_r} = \sum_{\bx,n} dV \frac{\Delta t}{\kappa} p^n g^n \delta_{z_r}
\]
\[
= \sum_{\bx,n} dV \Delta t p^n\left(\frac{q^{n}-q^{n-1}}{\kappa\Delta t} -D^T \bw^{n-1/2}\right)
\]
\[
= \sum_{\bx,n} dV \Delta t \left(\frac{p^{n}-p^{n+1}}{\kappa\Delta t}\right)q^n - D p^n \cdot \bw^{n-1/2}
\]
\[
= \sum_{\bx,n} dV \Delta t \left(-D^T\bv^{n+1/2} - \frac{f^n}{\kappa}\delta_{z_s}\right)q^n + \frac{\bv^{n+1/2}-\bv^{n-1/2}}{\beta \Delta t}\bw^{n-1/2}
\]
\[
=-\sum_{\bx,n} dV \frac{\Delta t}{\kappa} f^nq^n \delta_{z_s} 
-\bv^{n+1/2} \cdot \left(D q^n - \frac{\bw^{n-1/2}-\bw^{n+1/2}}{\beta \Delta t} \right) 
\]
\[
=-\sum_{\bx,n} dV \frac{\Delta t}{\kappa} f^nq^n \delta_{z_s} 
-\bv^{n+1/2}\cdot  \left(D q^n + \frac{\bw^{n+1/2}-\bw^{n-1/2}}{\beta \Delta t} \right) 
\]
\begin{equation}
\label{eqn:adjreln}
=-\sum^{N-1}_{n=0} \frac{\Delta t}{\kappa} f^nq^n|_{z=z_s}
\end{equation}
\begin{equation}
\label{eqn:opdef}
S[\kappa,\beta]f = p|_{z=z_r}, \,S[\kappa,\beta]^*g = -q|_{z=z_s}
\end{equation}
\begin{equation}
\label{eqn:ipdef}
\langle u, v \rangle_{\kappa,z_0} = \sum_{\bx',n} \frac{dV' \Delta t}{\kappa(\bx',z_0)} u(\bx',z_0,n\Delta t) v(\bx',z_0,n\Delta t)
\end{equation}
\begin{equation}
\langle S[\kappa,\beta]f, g \rangle_{\kappa,z_s} = \langle f, S[\kappa,\beta]^*g \rangle_{\kappa,z_r}
\end{equation}

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/geom/paper.tex
repro docs dir added

\title{Geometry of Extended Reflection}
\author{
\ William Symes\thanks{Department of Computational and Applied Mathematics, Rice University,
Houston, TX, 77005, USA,
{\tt symes@rice.edu}}
}

\lefthead{Symes}

\righthead{Extended Reflection Geometry}

\maketitle
\parskip 12pt

\begin{abstract}
Subsurface offset extended modeling implies relations between reflection properties (reflector dip, scattering angle) and ray geometry. The relation is similar to that of non-extended modeling, but differs in that a simple relation between scattering angle and image wavenumber holds only for zero offset...
\end{abstract}

\section{Introduction}
\cite{SavaFomel:03} introduced the construction of angle domain extended images by Radon transform of subsurface offset extended images. Notably, the ``scattering angle'' was identified by \cite{SavaFomel:03} as linked to the image wavenumber in offset.

Generally, the physical meaning of this ``angle domain'' has been misunderstood. 
For example, the ``scattering angle'' computed via Radon transform is the opening angle between incident and reflected rays only when these rays meet at the scattering point - that is, for zero subsurface offset. All of the energy is focused at zero offset when the velocity if correct (in the infinite frequency limit, of course), but not otherwise.

The ray geometry of the subsurface offset extension has been explored in a number of papers. I use here the account of \cite{HouSymes:15}, which completely and rigorously develops the geometry of extended reflection. Note that essentially the same relations appear in \cite{tenKroode:12}, and possibly elsewhere; these relations are implicit in the asymptotic analysis of extended modeling and migration.

\section{Theory}
In constant-density acoustic Born modeling extended by horizontal subsurface offset, the model perturbation $\delta v$ is a function of horizontal spatial coordinates $\bx$, horizontal subsurface offset $\bh$, and depth $z$. For 2D modeling, $\bx $ and $\bh $ are 1D vectors, whereas for 3D modeling they are 2D: we will use the same notation in either case.

The dual wavevectors to these coordinates are $\bk_x, \bk_h,$, and $k_z$.Two distinguished wavevector subspaces play a key role, namely $\bk_h=0$ with coordinates $\bk_x,k_z$ and $\bk_x=0$ with coordinates $\bk_h,k_z$. We will write 
\begin{equation}
\label{eqn:hyp}
k_{xz}=\sqrt{|\bk_x|^2 + k_z^2},\,\,k_{hz}=\sqrt{|\bk_h|^2+k_z^2}.
\end{equation} 
For 2D, the angle subtended with the distinguished axis, which for horizontal subsurface offset is of course the $k_z$ axis, are denoted $\nu $ for the $\bk_x,k_z$ subspace, and $\gamma $ for the $\bk_h,k_z$ subspace:
\begin{equation}
\label{eqn:cos}
\cos \nu = \frac{k_z}{k_{xz}},\,\, \cos \gamma = \frac{k_z}{k_{xz}}.
\end{equation}
For 3D, these angles are denoted $\nu_1$ and $\gamma_1$, and
the corresponding azimuths are $\nu_2 $ and $\gamma_2 $ respectively. Equations (\ref{eqn:cos}) are still correct, with $\nu$ replaced by $\nu_1$ and $\gamma$ by $\gamma_1$.

In the high-frequency asymptotics of reflection, the wavevector $(\bk_x,\bk_h,k_z)$ is an extended reflector normal. For physical reflectors, the reflectivity Fourier transform is independent of $\bk_h$, and the subvector with coordinates $(\bk_x,k_z)$ is thus the physical reflector normal. Accordingly, $nu$ (2D) or $\nu_1$ (3D) is the dip (polar) angle and $\nu_2 $ is the dip azimuth. The physical significance of the angle $\gamma$, or $\gamma_1$ and $\gamma_2$ in the 3D case, is not as obvious, as it is related to the ray geometry of reflection.

For the purposes of this paper, assume that all traveltimes between source or receiver and reflecting point are single-valued, that is, that no caustics occur in the scattering domain. For a particular pair $\bx_r,\bx_s$ of source and receiver coordinates (assuming that these lie on horizontal planes), the two-way traveltime $T$ is defined in terms of the one-way traveltime $\tau$ by
\begin{equation}
\label{eqn:twoway}
T(\bx,\bh,z;\bx_r,\bx_s) = \tau(\bx+\bh,z,\bx_r) + \tau(\bx-\bh,z,\bx_s).
\end{equation}
$T$ is closely related to the phase function in the expression of the scattering normal operator - see \cite{HouSymes:15}, Appendix A, equation (A-1). Note that in this reference $T$ is denoted $\phi$, and the one-way times are denoted as $T_r$ and $T_s$ respectively. The computations leading up to equation (A-6) are essentially the same, with $x$ replaced by $\bx$ and $k_x,k_h$ by $\bk_x,\bk_h$ and so on to accommodate the 3D case. Note that in 3D, two time integrations are required for each copy of $\bar{F}$ on the left-hand side of equation (A-1), in order that the resulting expression be equivalent to integration over a surface. The upshot is essentially the same stationary phase conditions as expressed in (A-6). In particular,
\begin{eqnarray}
\label{eqn:statph-x}
(\bk_x,k_z) &\mbox{ is parallel to }& \nabla_{\bx,z} T,\\
\label{eqn:statph-h}
(\bk_h,k_z) &\mbox{ is parallel to }& \nabla_{\bh,z} T,
\end{eqnarray}
from which it follows that
\begin{eqnarray}
\label{eqn:statphquot-x}
\frac{k_z}{k_{xz}} & = & \frac{\nabla_z T}{|\nabla_{\bx,z}T|},\\
\label{eqn:statphquot-h}
\frac{k_z}{k_{hz}} & = & \frac{\nabla_z T}{|\nabla_{\bh,z}T|},
\end{eqnarray}
and
\begin{eqnarray}
\label{eqn:statphquot-x-tan}
\tan \nu & = & \frac {|\nabla_{\bx}T|}{\nabla_z T},\\
\label{eqn:statphquot-h-tan}
\tan \gamma & = & \frac {|\nabla_{\bh}T|}{\nabla_z T}.
\end{eqnarray}

The $\bx,z$ gradient of $T$ is the sum of the $\bx,z$ gradients of $\tau$ for the arguments indicated in equation (\ref{eqn:twoway}). These gradients are in turn the ray velocity vectors from source and receiver. 
For $\bh=0$, the eikonal equation states that these gradients have the same length, so that their sum is their bisector. Thus in this case equation (\ref{eqn:statph-x}) states Snell's law of reflection, and the bisector of the angle subtended by incident and reflected rays is the dip (reflector normal) vector. For $\bh \ne 0$, the values of velocity at the points of evaluation $(\bx \pm \bh,z)$ are not necessarily the same, so that while the reflector normal is still parallel to the isochron normal, it is not necessarily the bisector of the incident and reflected ray pair.

The scattering (or opening) (half-) angle $\gamma_{\rm scat}$ is one-half the angle subtended by the incident and reflected rays. Denoting $s=v^{-1}(\bx,z), s_{\pm} = v^{-1}(\bx \pm \bh,z)$, and using the eikonal equation several times,
\[
|\nabla_{\bx,z}T(\bx,\bh,z;\bx_r,\bx_s)|^2 = s_+^2 + s_-^2 + 2s_+s_-\cos 2 \gamma_{\rm scat}
\]
\begin{equation}
\label{eqn:len}
= 4s_+s_-\left(\frac{s_+}{2s_-}+\frac{s_-}{2s_+}  -1 + \cos^2 \gamma_{\rm scat}\right). 
\end{equation}
This equation is difficult to interpret in general, however in the special case $\bh=0$ it takes on a familiar meaning:
\begin{equation}
\label{eqn:lenh0}
|\nabla_{\bx,z}T(\bx,0,z;\bx_r,\bx_s)|^2 = 4 s^2 \cos^2 \gamma_{\rm scat}
\end{equation}
On the other hand,
\begin{equation}
\label{eqn:lenh0-1}
|\nabla_{\bx,z}T|^2 = |\nabla_{\bx}T|^2 + |\nabla_zT|^2 
=\sec^2\nu |\nabla_z T|^2.
\end{equation}
The computation of $|\nabla_zT|^2$ is accomplished in equations (A-21) through (A-26) and accompanying discussion in Appendix A of \cite{HouSymes:15}, and needs no change other than the lexicographic ones already noted to accommodate 3D: the starting point is the stationary phase (A-6), or its rewrite (\ref{eqn:statph-x}), (\ref{eqn:statph-h}) above. The net result is (for general $\bh$, rewritten from equations (A-24), (A-25) and (A-26))
\begin{equation}
\label{eqn:tz}
|\nabla_z T|^2 = \frac{-b + \sqrt{b^2-4ac}}{2a}
\end{equation}
in which
\begin{equation}
\label{eqn:tzabc}
a = \frac{k_{xz}^2k_{hz}^2}{k_z^4},\,b=-2\left[(s_+^2=s_-^2)\frac{\bk_x\cdot \bk_h}{k_z^2}+(s_+^2+s_-^2)\right],\, c=(s_+^2-s_-^2)^2.
\end{equation}
For $\bh=0$, these expressions collapse to
\begin{equation}
\label{eqn:tzh0}
|\nabla_z T|^2 = 4s^2 \frac{k_z^4}{k_{xz}^2k_{hz}^2}
\end{equation}
so from (\ref{eqn:lenh0}), (\ref{eqn:lenh0-1}, and (\ref{eqn:cos})
\[
4 s^2 \cos^2 \gamma_{\rm scat} = 4 s^2 \sec^2\nu  \frac{k_z^4}{k_{xz}^2k_{hz}^2}
\]
\begin{equation}
\label{scatid}
=4 s^2 \cos^2 \gamma
\end{equation}
so the angles $\gamma$ and $\gamma_{\rm scat}$ are identical for $\bh=0$. Inspection of (\ref{eqn:tz}), (\ref{eqn:tzabc}) suggests that otherwise, when $s_+ \ne s_-$, these two angles are not necessarily equal.

For the 2D case with $\bh=0$, reference to (\ref{eqn:cos}), (\ref{eqn:statph-x}), (\ref{eqn:statph-h}) and (\ref{eqn:tzh0}) yield identities previously stated by \cite{SavaFomel:03}:
\begin{eqnarray}
\label{savafomel}
\frac{\partial T}{\partial x} &=& 2 s \sin\nu \cos\gamma, \nonumber \\
\frac{\partial T}{\partial h} &=& 2 s \cos\nu \sin\gamma, \nonumber \\
\frac{\partial T}{\partial z} &=& 2 s \cos\nu \cos\gamma,
\end{eqnarray}
and
\begin{eqnarray}
-\frac{\partial z}{\partial x} = \tan \nu &=& -\frac{k_x}{k_z} \nonumber \\
-\label{savafomeltan}
-\frac{\partial z}{\partial h} = \tan \gamma &=& -\frac{k_h}{k_z}
\end{eqnarray}
which are simply the equations (\ref{eqn:statphquot-x-tan}) and (\ref{eqn:statphquot-h-tan}) with sign normalization. On the left-hand side, $z$ is regarded implicitly as a function of $x,h$ describing the plane tangent to the extended reflector at the reflection point.
Note that the identities (\ref{savafomel}) are correct in general {\em only} for $\bh=0$, whereas from the point of view developed in this paper, equations (\ref{savafomeltan}) are valid in general.

For completeness, we state the 3D generalizations of these relations, in coordinates, using $h_x$ and $h_y$ for the $x-$ and $y-$ components of subsurface offset:
\begin{eqnarray}
\label{savafomel3D}
\frac{\partial T}{\partial x} &=& 2 s \sin\nu_1 cos\nu_2 \cos\gamma_1, \nonumber \\
\frac{\partial T}{\partial y} &=& 2 s \sin\nu_1 sin\nu_2 \cos\gamma_1, \nonumber \\
\frac{\partial T}{\partial h_x} &=& 2 s \cos\nu_1 \sin\gamma_2\cos\gamma_2, \nonumber \\
\frac{\partial T}{\partial h_y} &=& 2 s \cos\nu_1\sin\gamma_1\sin\gamma_2, \nonumber \\
\frac{\partial T}{\partial z} &=& 2 s \cos\nu_1\cos\gamma_1,
\end{eqnarray}
and 
\begin{eqnarray}
\label{savafomeltan3D}
-\frac{\partial z}{\partial x} = \tan \nu_1\cos\nu_2 &=& -\frac{k_x}{k_z},\nonumber\\
-\frac{\partial z}{\partial y} = \tan \nu_1\sin\nu_2 &=& 
-\frac{k_y}{k_z}, \nonumber\\
-\frac{\partial z}{\partial h_x} = \tan \gamma_1\cos\gamma_2  &=& -\frac{k_{h_x}}{k_z}, \nonumber\\
-\frac{\partial z}{\partial h_y} = \tan \gamma_1\sin\gamma_2  &=& -\frac{k_{h_y}}{k_z}.
\end{eqnarray}
Again, equations (\ref{savafomel3D}) relate rays to reflectors and hold only for $h=0$, whereas equations (\ref{savafomeltan3D}) simply identify the angles in relation to the reflector wavenumbers and hold in general.
\section{Conclusion}
The geometrical reasoning developed by \cite{HouSymes:15} encapsulates the relations between ray angles and reflector normals. These relations take the usual form proposed by \cite{SavaFomel:03} only for physical reflection, that is, zero subsurface offset.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/iva/junk.tex
repro docs dir added

\section{Quantifying Microlocal Asymptotics}

\[
\Lambda(\xi) \le 2 \Lambda(\eta) \Lambda(\xi + \eta)
\]
pf: if $||\xi|-|\eta|| \le \frac{1}{2}|\xi|$, then $|\eta| \ge \frac{1}{2}|\xi|$ so
\[
\Lambda(\eta)^2 \Lambda(\xi + \eta)^2 = (1+|\eta|^2)(1+ ...) \ge \frac{1}{4}(1+|\xi|^2)
\]
Otherwise, if $||\xi|-|\eta|| \ge \frac{1}{2}|\xi|$, then reverse triangle inequality implies 
$|\xi + \eta| \ge \frac{1}{2}|\xi|$ so same story.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The theory to be developed here applies for example to acoustic
scattering, governed by the acoustic wave equation in $R^3$,
\bea
\label{awe}
\frac{\partial p}{\partial t} & = & -\kappa \nabla \cdot \bv + f\\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p.
\eea
The components of this system are the excess pressure $p$, particle
velocity $\bv$, bulk modulus $\kappa$, buoyancy (reciprocal density)
$\beta$, and source term $f$. The material parameters $\kappa$ and
$\beta$ are functions on $\bR^3$ and must be positive in the region of
interest. The fields $f$, $p$ and $\bv$ are functions on space-time $\bR^3
\times \bR$ and of source position, which ranges over a set $Y_s \subset
\bR^3$.

\cite{BlazekStolkSymes:13} show that this system has solutions
depending ``nicely'' on the coefficients $\kappa$ and
$\beta$. Specifically, suppose that $\Omega$ is a cube, 
\begin{itemize}
\item $V = H^1_{0}(\Omega) \times (H^1(\Omega))^3$,
\item $H = L^2(\Omega) \times (L^2(\Omega))^3$
\item $S^k = \{ f\in C^k(Y_s \times \bR, H): f(\bx_s,t)=0, \bx_s \in
  Y_s, t < 0\}$
\item $(\log \kappa, \log \beta) \in L^{\infty}(\Omega) \equiv M_0$.
\end{itemize}
Then  there is a unique solution $(p,\bv) \in C^k(Y_s \times \bR, H)$ of a suitable weak version of
\eqref{awe}. Also, if $k \ge n \ge 0$, then $\cal{U}: M_0 \ C^{k-n}(Y_s \times
\bR, H)$ is $n$ times G\^ateaux differentiable. We denote the
directional derivative operator by $D^n\cal{U}$. For example,
\[
D\cal{U}(\kappa,\beta) (\delta \kappa, \delta \beta) = (\delta p,
\delta \bv)
\]
for the (weak) solution $(\delta p, \delta \bv)$ of 
\bea
\label{dawe}
\frac{\partial \delta p}{\partial t} & = & -\kappa \nabla \cdot \delta
\bv - \delta \kappa \nabla \cdot \bv \\
\frac{\partial \delta \bv}{\partial t} & = & - \beta \nabla \delta p - \delta
\beta \nabla p.
\eea

\cite{BlazekStolkSymes:13} also show that weak solutions of
\eqref{awe} propagate with finite speed given as usual by $c =
\sqrt{\kappa \beta}$. Therefore if $T > 0$ is fixed and $\Omega$ is chosen sufficiently
large relative to 
\[
\cup \{\mbox{supp } f(\bx_s,t): \bx_s \in Y_s, t \in [0,T]\}
\]
then $(p,\bv)(t) = 0$ near $\partial \Omega$ for $0\le t \le T$. In
effect, the solution of \eqref{awe} propagates in $\bR^3$ for $ t \le
T$.

If more regularity is assumed of $\kappa, \beta$, then additional
constraints follow for the solution. Assume from now on that $Y_s
\subset \Omega \cap \{\bx:x_3=0 \}$

In order to avoid spurious difficulties with behaviour at infinity,
abbreviate $H^s = H^s(T^3)$ with $T^3$ = a torus of large enough size
that

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Stuff}

\noindent {\bf Theorem 2:} Suppose that  $\de \in H^1_0(Y)$ for all $\lambda \in (0,\lambda_0]$. Then $\bre[m] \in H^1_0(X)$, and there exist $C,\lambda_0>0$ as in Theorem 1 so that for all $m \in U$, $\lambda \in (0,\lambda_0]$,
\[
\|\bre[m]\|_1 \le C \|\de\|_1.
\]
\noindent{\bf Proof:}
As $\Ne[m] \in OPS^0$ is elliptic, it has a (global, not microlocal) parametrix $\Ne[m]^{\dagger} \in OPS^0$, so that $\Ne[m]^{\dagger}\Ne[m]-I \in OPS^{-\infty}$. Accordingly,
\[
\bre[m] = \Ne[m]^{\dagger}\Ne[m]\bre[m] + (I-\Ne[m]^{\dagger}\Ne[m])\bre[m]
\]
\[
= \Ne[m]^{\dagger}\oF[m]^*\de + (I-\Ne[m]^{\dagger}\Ne[m])\bre[m].
\]
so standard $L^2$ estimates imply the first part of the conclusion. As for the second part,
\[
\Ne[m]\Lambda\bre[m] = [\Ne[m],\Lambda]\bre[m]+\Lambda\oF[m]^*\de
\]
Write $\Lambda \bre[m]=u+w$, where
\[
\Ne[m]u = [\Ne[m],\Lambda]\bre[m]
\]
and
\[
\Ne[m]w=\Lambda \oF[m]^*\de.
\]
From Lemma 2 with $s=1, k=1$, the hypotheses of Corollary A6 are satisfied for the second of these equations, so
\[
\|w\|_0 \le C_0 \|\Lambda \oF[m]^*\de\|_{0}.
\]
The basic bound on $\Ne[m]^{-1}$ implies that
\[
\|u\|_0 \le C_1 \lambda^{-1}\|[\Ne[m],\Lambda]\bre[m]\|_0
\]
\[
\le C_2 \lambda^{-1}\|\bre[m]\|_0,
\]
since $[\Ne[m],\Lambda] \in OPS^{0}$.  From Theorem 1, this is
\[
\le C_3 \lambda^{-1}\|\oF[m]^*\de\|_{0}.
\]
Invoking Lemma 2 once again and adding the bounds on $u$ and $w$, the result follows.

Q. E. D.

\noindent{\bf Theorem 3:} Denote by $\Ne[m]^{\dagger}$ the microlocal parametrix for $\Ne[m]$ described in Lemma A2. Then there exist $L_1, K_1 \ge 0$ so that
for $0 \le \lambda \le \lambda_0$, $m \in U$, $s=0,1$, 
\be
\label{theorem3:res1}
\|(I-\Pi[m])\Ne[m]^{\dagger}\oF[m]^*\de\|_s \le L_1\|\Ne[m]^{\dagger}\oF[m]^*\de\|_{s-1}
\ee
and
\be
\label{theorem3:res2}
\|\Ne[m]^{\dagger}\oF[m]^*\de\|_{s-1} \le K_1 \lambda \|\Ne[m]^{\dagger}\oF[m]^*\de\|_{s}.
\ee

\noindent{\bf Proof:} Lemma 2 shows that $\{q_{\lambda}=\oF[m]^*\de: \lambda \in (0,\lambda_0]\}$ satisfies inequalities of the form (\ref{microlocal_1}) and (\ref{hfc_1}) with $s=0,1$ and $k=0,1$. 
The inequalities (\ref{theorem3:res1}) and (\ref{theorem3:res2}) are precisely the conclusion Lemma A5.

Q. E. D.

\noindent{\bf Theorem 4:} In Assumptions 2 and 3, require that $k_0 \ge 2$. Then there exist $L_2,K_2 \ge 0$ so that
for $0 \le \lambda \le \lambda_0$, $m \in U$, $s=0,1$, 
\be
\label{theorem4:res1}
\|(\Pi[m]-I)\bre[m]\|_s \le L_2\|\bre[m]\|_{s-1}
\ee
\be
\label{theorem4:res2}
\|\bre[m]\|_{s-1}\le K_2\lambda \|\bre[m]\|_s
\ee
\noindent{\bf Proof:} For conciseness, write $p=\Ne[m]^{\dagger}\oF[m]^*\de$.
\[
\|(\Pi[m]-I)\bre[m]\|_s \le \|(\Pi[m]-I)p\|_s + \|(\Pi[m]-I)(\bre[m]-p)\|_s  
\]
\[
\le L_1\|p\|_{s-1} + C\|\bre[m]-p\|_s
\]
Take first the case $s=0$. Note that
\[
\Ne[m] (\bre[m]-p) = (I-\Ne[m]\Ne[m]^{\dagger})\oF[m]^*\de
\]
\[
=(I-\Ne[m]\Ne[m]^{\dagger})\Pi[m]\oF[m]^*\de + (I-\Ne[m]\Ne[m]^{\dagger})(I-\Pi[m])\oF[m]^*\de.
\]
According to Lemma A2,  $(I-\Ne[m]\Ne[m]^{\dagger})\Pi[m] \in OPS^{-k}$ for any $k \ge 0$, whereas from Lemma 2 with $s=0, k=2$, $\|(I-\Pi[m])\oF[m]^*\de\|_0 \le L_1\|\oF[m]^*\de\|_{-2}$. Since $I-\Ne[m]\Ne[m]^{\dagger}$ is bounded on $L^2(X)$, uniformly in $\lambda$, the RHS is bounded in $L^2(X)$ by a multiple of $\|\oF[m]^*\de\|_{-2}$. Since $\Ne[m] \ge \lambda I$, 
\[
\|\bre[m]-p\|_0 \le C_2\lambda^{-1}\|\oF[m]^*\de\|_{-2}
\]
From Assumption 2 with $s=-1, k=1$ it follows that this is
\be
\label{theorem4:eqn1}
\le C_3 \|\oF[m]^*\de\|_{-1} \le C_3\|\Ne[m]\|_{-1,-1}\|\bre[m]\|_{-1}.
\ee
For future use, note that alternatively 
\be
\label{theorem4:eqn2}
\|\bre[m]-p\|_0 \le C_2\lambda \|\oF[m]^*\de\|_{0} \le C_3\lambda \|\Ne[m]\|_{0,0}\|\bre[m]\|_{0}.
\ee
Similarly,
\[
\|p\|_{-1} \le \|\bre[m]-p\|_{-1} + \|\bre[m]\|_{-1}
\]
\[
\le \|\bre[m]-p\|_0 +  \|\bre[m]\|_{-1}
\]
\[
\le C_4 \|\bre[m]\|_{-1}
\]
via another use of (\ref{theorem4:eqn1}). So
\[
\|(\Pi[m]-I)\bre[m]\|_0 \le \|(\Pi[m]-I)p\|_0 + \|(\Pi[m]-I)(\bre[m]-p)\|_0  
\]
\[
\le C_5 \|p\|_{-1} + C_4\|\bre[m]\|_{-1} \le C_6\|\bre[m]\|_{-1}.
\]
which establishes (\ref{theorem4:res1}) for $s=0$.

From Theorem 3, 
\[
\|\bre[m]\|_{-1} \le \|p\|_{-1} + \|\bre[m]-p\|_{-1} \le \lambda K_1\|p\|_0 + \|\bre[m]-p\|_{0}.
\]
\[
\le \lambda K_1\|\bre[m]\|_0 + (\lambda K_1+1) \|\bre[m]-p\|_0;
\]
estimate the second term from (\ref{theorem4:eqn2}) to obtain
\[
\le \lambda ((2+\lambda_{\rm max})C_3\lambda \|\Ne[m]\|_{0,0})\|\bre[m]\|_0,
\]
which estimates the constant in (\ref{theorem4:res2}) for $s=0$.

For $s=1$, follow the bootstrap pattern introduced in the proof of Theorem 2. Note that $\Lambda^{-1}\Ne[m]\Lambda$ is self-adjoint, bounded, and $\ge \lambda$ on $H^1(X)$, and
\[
\Lambda^{-1}\Ne[m]\Lambda(\bre[m]-p) = \Lambda^{-1}[\Ne[m],\Lambda](\bre[m]-p) + (I-\Ne[m]\Ne[m]^{\dagger})\oF[m]^*\de.
\]
As in the proof of Theorem 2, write $\bre[m]-p = u + w$, where
\[
\Lambda^{-1}\Ne[m]\Lambda u = \Lambda^{-1}[\Ne[m],\Lambda](\bre[m]-p)
\]
and
\[
\Lambda^{-1}\Ne[m]\Lambda w = (I-\Ne[m]\Ne[m]^{\dagger})\oF[m]^*\de.
\]
Since $\Lambda^{-1}\Ne[m]\Lambda w \ge \lambda I$,
\[
\|u\|_1 \le \lambda^{-1}\|\Lambda^{-1}[\Ne[m],\Lambda](\bre[m]-p)\|_1
\]
\be
\label{theorem4:eqn3}
\le \lambda^{-1}C_5 \|\bre[m]-p\|_0 \le \lambda^{-1}C_6\|\bre[m]\|_{-1} \le C_7\|\bre[m]\|_0, 
\ee
from (\ref{theorem4:eqn1}) and the case $s=0$. 
To bound $w$, write $w=w_1 + w_2$, where
\[
\Lambda^{-1}\Ne[m]\Lambda w_1 = (I-\Ne[m]\Ne[m]^{\dagger})\Pi[m]\oF[m]^*\de,
\]
\[
\Lambda^{-1}\Ne[m]\Lambda w_2 = (I-\Ne[m]\Ne[m]^{\dagger})(I-\Pi[m])\oF[m]^*\de.
\]
According to Lemma A2,  $(I-\Ne[m]\Ne[m]^{\dagger})\Pi[m] \in OPS^{-k}$ for any $k\ge 0$, whereas from Lemma 2 $\|(I-\Pi[m])\oF[m]^*\de\|_1 \le L_1\|\oF[m]^*\de\|_0$. Since $I-\Ne[m]\Ne[m]^{\dagger}$ is bounded on $H^1_0(X)$, uniformly in $\lambda$, conclude that
\[
\|w_1\|_1 \le \lambda^{-1}\|(I-\Ne[m]\Ne[m]^{\dagger})\Pi[m]\oF[m]^*\de\|_1
\]
\[
\le C_8 \lambda^{-1}\|\oF[m]^*\de\|_{-1}\le C_8 K_1\|\oF[m]^*\de\|_{0}
\]
\be
\label{theorem4:eqn4}
\le C_7 K_1\|\Ne[m]\|_{0,0}\|\bre[m]\|_0.
\ee
\be
\label{theorem4:eqn5}
\|w_2\|_1 \le C_8\|\oF[m]^*\de\|_0 \le C_8\|\Ne[m]\|{0,0}\|\bre[m]\|_{0}.
\ee
Adding up (\ref{theorem4:eqn3}), (\ref{theorem4:eqn4}), and (\ref{theorem4:eqn5}), 
conclude that
\be
\label{theorem4:eqn6}
\|\bre[m]-p\|_1 \le C_9 \|\bre[m]\|_0.
\ee
\[
\|p\|_{0} \le \|\bre[m]-p\|_{0} + \|\bre[m]\|_{0}
\]
which from (\ref{theorem4:eqn2}) is 
\[
\le C_{10}\|\bre[m]\|_{0}
\]
So
\[
\|(\Pi[m]-I)\bre[m]\|_1 \le \|(\Pi[m]-I)p\|_1 + \|(\Pi[m]-I)(\bre[m]-p)\|_1  
\]
\[
\le C_11 \|p\|_{0} + C_12\|\bre[m]\|_{0} \le C_13\|\bre[m]\|_{0}.
\]<
Finally, From Theorem 3, 
\[
\|\bre[m]\|_0 \le \|p\|_0 + \|\bre[m]-p\|_0 \le \lambda K_1\|p\|_1 + \|\bre[m]-p\|_{0}.
\]
\[
\le \lambda (C_5 |\Ne[m]\|_{0,0} + K_1\|\|\Ne[m]^{\dagger}\|_{1,1}\|\Ne[m]\|_{1,1})\|\bre[m]\|_1
\]
which provides the claimed inequality of form  (\ref{theorem4:res2}) for $s=1$.

Q. E. D.

### 2019-10-19 repro/papers/iva/old.tex
repro docs dir added

\title{Inversion Velocity Analysis}
\date{}
\author{William W. Symes, The Rice Inversion Project}

\maketitle
\parskip 12pt
\begin{abstract}
Minimization of semblance subject to regularized linearized inversion with fixed penalty parameter(s) results in computable gradient approximation with controllable error. This is an intermediate result for the constrained formulation.
\end{abstract}

\section{Introduction}

\section{Mapping properties of separable extended modeling operators}
This is a Hilbert space story. For any Hilbert space $H$, the inner
product will be denoted $\langle \cdot, \cdot \rangle_H$, the norm
denoted $\| \cdot \|_H$, with the subscript omitted whenever the
choice of space is clear from context.

The natural objects in this study are {\em Hilbert space scales}, that
is sequences of Hilbert spaces $H = \{ H^s, s \in \bZ\}$, decreasing in
the sense that
$H^{s+1} \subset H^s, s \in \bZ$. Define the sets
\[
H^{\infty} = \cap_{s \in \bZ} H^s, \,H^{-\infty} = \cup_{s \in
  \bZ}H^s.
\]
An {\em operator} $L$  {\em of order $k \in \bZ$} from
a scale $H_1$ to a scale $H_2$ is linear function from $H_1^{-\infty}$ to
$H_2^{-\infty}$ for which $L|_{H_1^s} \in \sB(H_1^s,H_2^{s-k})$ for
all $ s \in
\bZ$. Denote by $OP^k(H_1,H_2)$ the set of all operators of order $k$.

The motivating example of this structure are scales of $L^2$ Sobolev
spaces, and (pseudo)differential operators on them.

The domain of a separable extended model consists of two
components. The first is a (background) model Hilbert space $M_b$ and
an open set of admissible models
$U \subset M_b$. In all examples, $M_b$ is a space of (possibly)
vector-valued) smooth ($C^{\infty}$, ``low frequency'') functions on a suitable physical
domain representing the physical parameters of a wave dynamics model, and members of $U$ obey additional constraints (such as
bounds) required to make the dynamical laws they parametrize
well-posed.

The other component of the model domain is a Hilbert space scale of extended
models $\oM =\{\oM_k, k\in \bZ\}$. 
A distinguished subscale $M = \{M_k, k\in \bZ\}, M_k \subset \oM_k$,
represents the non-extended or physical models. 
The  {\em annihilator} $A \in OP^0(\oM,N)$ has $M$ as its kernel:
\begin{equation}
\label{eqn:ann}
\om \in \oM,\,\,A\om=0 \Leftrightarrow \om \in M.
\end{equation}.
The {\em extension map} $E \in OP^0(M,\oM)$ is injective.

The model range or data space is another Hilbert scale $D$.

The extended modeling operator, or forward map, is a function
\[
\oF: U \rightarrow OP^0(\oM,D)
\]
It is {\em smooth with loss of regularity}: $m_b \mapsto \oF[m_b]|_{\oM^k}$ is a 
of class $C^p$ as a map from $U \rightarrow \sB(\oM^k,D^{k-p})$.

The physical modeling operator is a similar map
\[
F:U \rightarrow OP^0(M,D)
\]
with the same smoothness with loss of regularity.

The extended and physical modeling operators are related through the
extension map: for each $m_b \in U$,
\[
F[m_b] = \oF[m_b] \circ E.
\]

\section{Idealized extended inversion}
The main alleged advantage of extended modeling over standard modeling
is that {\em data can always be fit}, thus avoiding any possibility of
cycle-skipping. For the moment, then, idealize the extended modeling
operator $\oF[m_b]$ as bijective, for each admissible
background model $m_b \in U$: that is, assume that for any admissible
background model, it is always possible to perfectly fit data (certainly untrue
of standard modeling) with proper choice of extended model, and in
precisely one way. In some simple cases, this property actually holds,
while in others it holds in an approximate sense; in the latter case,
regularization is required, as will be explained in the next section.

In terms of the framework introduced in the last section, the inverse
problem is: given data $d \in D^0$, find a model $m_b \in U, m \in M^0$ so that
$F[m_b]m \approx d$. 
Call $d \in D$ {\em model-consistent} if there
exists $m_b^* \in U, m^* \in M$ so that $F[m_b^*]m^*=d$. 
In view of the extension property, and the assumed
bijectivity of $\oF[m_b]$, for any $m_b \in U$ there exists a unique
$\om[m_b,d] \in \oM$ so that
\be
\label{eqn:perfect}
\oF[m_b]\om[m_b,d] = d
\ee

{\em Inversion velocity analysis} (IVA) is the minimization of the objective
\be
\label{eqn:iva}
J[m_b] = \frac{1}{2}\|A\om[m_b,d]\|_0^2
\ee
over $m_b \in U$. If the data $d$ is model-consistent, then the global
minimum value is zero, since $M$ is the kernel of $A$. Therefore in
this case any global minimizer of $J$ is the first component of a
solution of the inverse problem, or equivalently a global minimizer of
the mean-square error $\|F[m_b]m-d\|_0$.

\ref{eqn:perfect}), is necessarily the background component of a
solution to the inverse problem. Note that this solution is not necessarily the
same as $m_b^*,m^*$, since $\oF$ (as opposed to $\oF[m_b], m_b \in U$)
has not been assumed injective.
If the data is not model-consistent, 
then $J[m_b]$ measures the distance to the feasible set ($M$) amongst
all extended models that fit the data, so a projection (left inverse
to the injection $M \subset \oM$) from $m_b,\om[m_b,d]$ onto $M$ will produce an
approximate solution.

Evidently (\ref{eqn:iva}) could be viewed as a constrained problem,
with constraint (\ref{eqn:perfect}). On the other hand, FWI could be
posed as minimzation of the extended residual (error in
(\ref{eqn:perfect})) subject to the constraint $A\om=0$. Thus IVA
may be viewed as FWI with the roles of objective and
constraint switched. While the two problems have the same solutions
for model-consistent data, the solutions are different for
model-inconsistent data. Also, all nontrivial examples of output least
squares inversion tend to generate a 
multiplicity of uninformative stationary points arbitrarily close to a
consistent-data global minimizer (the ``local min''
problem). In contrast, it can sometimes be shown that IVA
has only global minimizers, or at least that the IVA objective
is convex in a ball about the global solution.

Local minimization via Newton-like optimization requires computation
of the gradient. According to the smoothness assumptions made in the
last section,  $m_b \mapsto \om[m_b,d]$ is differentiable; by implicit
differentiation of (\ref{eqn:perfect}),
\begin{equation}
\label{eqn:dperfect}
(D\oF[m_b]\delta m_b)\om[m_b,d] + \oF[m_b]D\om[m_b,d]\delta m_b = 0,
\end{equation}
so
\[
DJ[m_b]\delta m_b = \langle D\om[m_b]\delta m_b, A^*A\om[m_b]\rangle
\]
\begin{equation}
\label{eqn:naive_jderiv}
= -\langle \oF[m_b]^{-1}(D\oF[m_b]\delta m_b)\om[m_b],A^*A\om[m_b]\rangle.
\end{equation}
For $m_b \in M_b$ and $\om \in \oM$, denote by $D\oF[m_b]\om$ the
linear map: $M_b \rightarrow D$ defined by
\[
\delta m_b \mapsto D(\oF[m_b]\om)\delta m_b
\]
and by $(D\oF[m_b]\om)^T$ its adjoint (note the distinction with the
adjoint $(D\oF[m_b]\delta m_b)^*$). Then 
whence the formal gradient of $J$ is
\be
\label{eqn:naive0_g}
\nabla J[m_b] = - (D\oF[m_b]\om[m_b])^T\bq[m_b]
\ee
in which $\bq \in D$ is the solution of
\be
\label{eqn:naive0_q}
\oF[m_b]^*\bq = A^*A\om[m_b].
\ee

Thus computation of the gradient appears to involve solution of two
linear systems ((\ref{eqn:perfect}) and (\ref{eqn:naive0_q})) and
application of the  {\em tomographic operator}
$(D\oF[m_b]\om[m_b,d])^T$. As explained elsewhere
\cite[]{KerSy:94,BiondiSava:04}. the tomographic operator is quite
computable via a variant of the {\em adjoint state method}, so modulo
the solution of the linear systems, (\ref{eqn:naive0_g}) appears to be
a computational {\em entr\'{e}} into IVA via quasi-Newton or conjugate
gradient methods. However several obstacles arise to practical use of
these ideas:
\begin{itemize}
\item for all but the simplest examples of extended seismic modeling, the linearized extended forward map $\oF$ is not actually surjective, therefore the equality constraint (\ref{eqn:perfect}) must be replaced by a best-fit criterion, such as least squares.
\item for all but the simplest examples of extended seismic modeling, the linearized extended forward map is not coercive, that is, the ratio $\|\oF[m_b]\om\|/\|\om\|$ attains arbitrarily small values (this is the case even if $\oF[m_b]$ is injective, which hardly matters). Therefore the least squares problem replacing (\ref{eqn:perfect}) must be regularized. On the other hand, for a natural choice of norm, the cone defined by $\|\oF[m_b]\om\| \ge \alpha \|\om\|$ for suitable $\alpha > 0$ contains a subspace whose dimension grows without bound as the frequency content of the modeled data grows without bound - that is an expression of the high information content of carried by propagating waves.
\item for all but the simplest examples of extended seismic modeling,
  the optimization problem replacing (\ref{eqn:perfect}) must be
  solved by an iterative process, simply because of computational size - therefore only an estimate of $\om[m_b,d]$ is available, and its error propagates into error in (regularized versions of) $J$ and $\nabla J$.
\item if the model parameters include wave velocities (that is to say,
almost always), the tomographic operator $(D\oF[m_b]\om[m_b,d])^T$ is not
continuous in the choice of norms for which the $\oF[m_b]$ is bounded
and has the ``partly coercive'' property mentioned in the second
bullet above. Otherwise put, in a choice of norms for which
$(D\oF[m_b]\om[m_b,d])^T$ is continuous, $\oF[m_b]$ is compact. If on the other hand
the iterative estimate oof $\om[m_b,d]$ relies on a choice of norms
for which $\oF[m_b]$ has a ``large'' well-conditioned subspace, then
no bound is possible for the effect of $(D\oF[m_b]\om[m_b,d])^T$ on
these error, hence no error control of the gradient is possible. In
practice, after discretization, this means that even a minimally accurate gradient requires
an excessively precise, hence computationally costly, estimation of $\om[m_b,d]$.
\end{itemize}

\section{Regularized extended inversion}
Overcoming the obstacles outlined at the end of the last section
requires introduction of additional abstract structure mirroring the
properties of important exaamples.

First, the extended domain space $\oM=\oM_0$ is member of a scale of Hilbert
spaces $\{\oM_s: s \in \bf{Z}\}$, each compactly embedded in the next: $\oM_s \subset \subset \oM_{s+1}$. Denote by $\langle \cdot ,\cdot \rangle_{s}$ the inner product
on $\oM_s$. Typically, $\{\oM_s\}$ is some version of the Sobolev
scale. Motivated by that example, assume that the inner product on
$\oM_0$ extends to a pairing: $\oM_s \times \oM_{-s} \rightarrow
\bf{R}$, realizing an isomorphism of $\oM_{-s}$ with the dual of
$\oM_s$, all $s \in \bf{Z}$.

We will assume that the data space $D = D_0$ is also a member of a
Hibert scale $\{D_s: s \in \bf{Z}\}$.

A linear map $L: \cup_{s \in \bf{Z}} X_s \rightarrow \cup_{s \in
  \bf{Z}}Y_s$ between scales $X, Y$ is {\em of order} $k$ iff $L: \in
{\cal L}(X_s,Y_{s-k})$ for every $s \in \bf{Z}$.

\noindent {\bf Assumption 1:} Assume that the operators $\oF[m_b]$ and $A$ extend to
the  {\em order 0} operators on the domain and range scales: $\oF[m_b]
\in {\cal L}(\oM_s,D_s)$ for every $s \in \bf{Z}$, similarly for $A$. The
notation $*$ is reserved for the adjoint of
$\oF[m_b]: \oM_0 \rightarrow D_0$.

\noindent {\bf Assumption 2:}  The {\em normal operator} $ m_b
\mapsto \oF[m_b]^*\oF[m_b]$ defines a smooth
${\cal L}(\oM_s,\oM_s)$-valued function on $U$ for every $s\in
\bf{Z}$, mapping $U$ to a bounded set. Similarly, the (``abnormal''?) operator $ m_b
\mapsto \oF[m_b]\oF[m_b]^*$ defines a smooth
${\cal L}(D_s,D_s)$-valued function on $U$ with bounded image, for every  $s \in \bf{Z}$.

\noindent {\bf Remark 1:} $\oF$ itself is {\em not} assumed to define a differentiable
map $\oM_s \rightarrow D_s$: that is because it does not, when wave
velocities are amongst the parameters in $m_b$. The (Gateaux) derivative does
exist, but only as an operator of order 1 - more on this below.

\noindent {\bf Remark 2:} Except in trivial instances, extended
modeling is {\em necessary} in order that $\oF\oF^*$ has the
smoothness property indicated.

Denote by $\bre[m_b,d]$ the solution of the $L^2$ least-squares problem,
\be
\label{eqn:reg}
\mbox{min}_{\br \in \oM_0}\, \frac{1}{2}(\|\oF[m_b]\br -\de\|_0^2 + \lambda \|\br\|_0^2).
\ee
This condition is of course a regularized version of (\ref{eqn:perfect}).
The first order condition for this problem is
\be
\label{l2normal}
\Ne[m_b]\br = \oF[m_b]^*\de
\ee
in which 
\[
\Ne[m_b] = \oF[m_b]^*\oF[m_b] + \lambda I
\]
Note that $\Ne$ is positive definite and self-adjoint on $M_0$, so the
Lax-Milgram Theorem asserts the existence of a unique $\bre[m_b,d] \in
L^2(X)$ satisfying (\ref{l2normal}).

In general the best possible estimate for the solution of (\label{l2normal})
is
\be
\label{eqn:genlest}
\|\om[m_b,d]\|_0  = O(\lambda^{-1}\|d\|_0)
\ee
However (\ref{eqn:genlest}) does not reflect the information content
of wave data, as remarked at the end of the last section.  To capture
this additional control possible in certain instances, posit the
existance of self-adjoint approximate projectors
\be
\label{eqn:dataproj}
\Pi_D,\bar{\Pi}_D \in \cap_{s \in \bf{Z}}{\cal L}(D_s, D_s), \,\Pi_D^*
= \Pi_D, \,\bar{\Pi}_D^* = \bar{\Pi}_D,
\ee
so that $\bar{\Pi}_D$ acts as an approximate identity on the range of
  $\Pi_D$, in the sense that 
\be
\label{eqn:projsmoothing}
(I-\bar{\Pi}_D)\Pi_D \in {\cal L}(D_s, D_{s+k}) \mbox{ for every } s
\in \bf{Z}, k \in \bf{Z}^+.
\ee

\noindent {\bf Assumption 3:} There exist $\gamma >0$, and for every $s \in \bf{Z}$, $k \in
\bf{Z}^+$ $C_{s,k} \ge 0$, so that
\be
\label{eqn:microell}
\|\oF[m_b]\oF[m_b]^*\bar{\Pi}_Dd\|_s \ge \gamma \|\bar{\Pi}_Dd\|_s -
\|d\|_{s-k}
\ee

\noindent {\bf Remark 3:} It follows from (\ref{eqn:projsmoothing})
that a similar inequality holds for $\Pi_D$.

\noindent {\bf Remark 4:} The inequality (\ref{eqn:microell})
abstracts a key feature of several important examples, a microlocal version of G\aa rding's inequality
\cite[]{Tay:81}; it is established by appeal to microlocal ellipticity
of the extended modeling operator.

The final two assumptions concern properties of the data, expressed in
terms of homogeneous inequalities defining cones in $D_s$. The results
to follow are {\em aymptotic}, so concern, not a single datum but a family
$\{\de\}_{\lambda \in (0,\lambda_{\rm max}}$ of data. The index
parameter $\lambda$ will turn out to signify (roughly) wavelength.

The first property expresses concentration of energy in ``imageable''
parts of data space.

\noindent {\bf Assumption 4:} For every $s \in \bf{Z}$, there exists
$L_s \ge 0$ so that for all $k \in \bf{Z}^+$, $\lambda \in
(0, \lambda_{\rm max}]$,
\be
\label{micro}
\|(I- \Pi_D)\de\|_s \le L_s \|\de\|_{s-k}.
\ee

The second property makes the connection with wavelength, and expresses
the assumption that the data is concentrated at high frequency:

\noindent {\bf Assumption 5 -  High Frequency Cone Condition:} 
For each $s \in \bf{Z}$ there exists $K_s
\ge 0$ so that for  $\lambda \in
(0, \lambda_{\rm max}]$,
\be
\label{hfcd}
\|\de\|_{s} \le K_s \lambda\|\de\|_{s+1}
\ee

\noindent{\bf Theorem 1:} 
Under assumptions 1 - 5 above, there exist $C,\lambda_0>0$ so that $\lambda_0 \le \lambda_{\rm max}$ and
\[
\|\bre[m_b,d]\|_0 \le C\|\de\|_0
\]
uniformly in $\lambda \in (0,\lambda_0]$, $m \in U$.

\noindent{\bf Proof:} See following section. % Take $M=\oF[m_b]^*\oF[m_b]$, $P=I$,
%$q_{\lambda}=\oF[m_b]^*\de$, $s=0$ and $k=1$ in the statement of Corollary A6. Lemma 2 verifies the required cond%itions on $M$, $P$, and $q_{\lambda}$. Q. E. D.

The semblance optimization functional [REF TO DEF IN INTRO] may be written as
\be
\label{tgt}
\Je[m_b] = \frac{1}{2}\|A\bre[m_b,d]\|^2
\ee
\[
=\frac{1}{2}\langle \de, \oF[m_b]\Ne[m_b]^{-1} A^*A \Ne[m_b]^{-1}\oF[m_b]^*\de \rangle
\]
%From Egorov's Theorem and Assumption 1 follows that operator in this expression is $\Psi$DO with symbol depending smoothly on $m$. Therefore $\Je$ is smooth in $m$.

\noindent {\bf Proposition 1:} $\Ne \in C^1(U,P_0)$, and for $\br \in \oM_1$, $\delta m_b \in M$,
\[
(D\Ne[m,\delta m]\br = (D\oF[m,\delta m]\oF[m_b]+\oF[m_b]D\oF[m,\delta m])\br
\]
\noindent {\bf Proof:} ...

For $\delta m \in M$,
\[
D\Je[m_b]\delta m = \langle D\bre[m_b,d]\delta m, A^*A\bre[m_b,d]\rangle.
\]
Thanks to Proposition 1,
\[
D\bre[m_b,d]\delta m = 
D\Ne[m_b]^{-1}(\delta m,\oF^*[m_b]\de) + \Ne[m_b]^{-1}D\oF[m_b]^*(\delta m, \de))
\]
\[
= -\Ne[m_b]^{-1}D\Ne[m_b](\delta m,\Ne[m_b]^{-1}\oF^*[m_b]\de) + \Ne[m_b]^{-1}D\oF[m_b]^*(\delta m, \de)
\]
\[
=-\Ne[m_b]^{-1}D\oF[m_b]^*(\delta m,\oF[m_b]\bre[m_b,d]) - \Ne[m_b]^{-1}\oF[m_b]^*D\oF[m_b](\delta m,\bre[m_b,d])
\]
\be
\label{prerderiv}
+ \Ne[m_b]^{-1}D\oF[m_b]^*(\delta m, \de)
\ee
\be
\label{rderiv}
=\Ne^{-1}(D\oF[m_b]^*(\delta m, \de - \oF[m_b]\bre[m_b,d]) - \oF[m_b]^*D\oF[m_b](\delta m, \bre[m_b,d])).
\ee
Define
\be
\label{qdef}
\bqe[m_b]=\Ne[m_b]^{-1}A^*A\bre[m_b,d].
\ee
Then
\be
\label{Jderiv}
D\Je[m_b]\delta m =\langle D\oF[m_b]^*(\delta m, \de - \oF[m_b]\bre[m_b,d]) - \oF[m_b]^*D\oF[m_b](\delta m, \bre[m_b,d]), \bqe[m_b]\rangle 
\ee
from which it follows that the formal ($L^2$) gradient of $\Je$ is 
\be
\label{grad}
\nabla \Je[m_b] = D\oF[m_b]^*(\bqe[m_b],\de - \oF[m_b]\bre[m_b,d])- D\oF[m_b]^*(\bre[m_b,d],\oF[m_b]\bqe[m_b]),
\ee

\noindent {\bf Remark:} That is, two linear least squares solves are required to compute the gradient of $\Je[m_b]$: (\ref{l2normal}) for $\bre[m_b,d]$ and (\ref{qdef}) for $\bqe[m_b]$.

\noindent {\bf Factorization Lemma:} $D\oF[m_b](\delta m,\br) = \oF[m_b]Q[m,\delta m]\br$, in which $Q[m,\delta m]$ is a smooth $OPS^1$-valued function of $(m,\delta m) \in U \times M$, linear in $\delta m$, whose values are essentially skew-adjoint:
\[
Q[m,\delta m] + Q[m,\delta m]^* = R[m,\delta m] \in OPS^0.
\]
For $u \in H^1$, 
\be
\label{Qnorm}
\|Q[m,\delta m]u\|_0 \le C\|\delta m\|_M \|u\|_1,
\ee
\be
\label{Rnorm}
\|R[m,\delta m]u\|_0 \le C\|\delta m\|_M \|u\|_0,
\ee
with $C$ uniform for $m\in U$.

It will be convenient to view $Q[m,\cdot]\br, m\in U, \br \in \oM$, as a continuous linear map $:M \rightarrow \oM$, and denote its adjoint by $Q^*[m,\cdot]\br$. That is,
\[
\langle Q^*[m,\br_2]\br_1,\delta m\rangle_M = \langle Q[m,\delta m]\br_1,\br_2\rangle.
\]
In the statement (\ref{Qnorm}), I have used $Q[m,\delta m]^*$ to denote the ``other'' adjoint, defined by 
\[
\langle Q[m,\delta m]^*\br_1,\br_2\rangle = \langle \br_1,Q[m,\delta m]\br_2\rangle
\]

\noindent {\bf Theorem 5:} 
\be
\label{qrderiv}
D\bre[m_b,d]\delta m = -Q[m,\delta m]\bre[m_b,d] + \lambda \Ne^{-1}R[m,\delta m]\bre[m_b,d].
\ee
\be
\label{eqn:gradq}
\nabla \Je[m_b] = (-Q^*[m,A^*A\bre[m_b,d]] + \lambda R^*[m,\bqe[m_b]])\bre[m_b,d].
\ee
For $C \ge 0$ uniform in small $\lambda$, $m \in U$,
\be
\label{eqn:firstgradest}
\|\nabla \Je[m_b] - -Q^*[m,A^*A\bre[m_b,d]]\bre[m_b,d] \|_M \le C\lambda \|\de\|_0.
\ee
Also, the $M$-valued quadratic form on $H^1$, defined by
\be
\label{qquad}
\br \mapsto -Q^*[m,A^*A\br]\br
\ee
extends continuously to $H^0$.

\noindent {\bf Remark:} The importance of the last statement stems from the effect of approximating $\bre[m_b,d]$, of course, as will be described below.

\noindent {\bf Proof:} From (\ref{prerderiv}) above and the definition of $Q$,
\[
D\bre[m_b,d]\delta m = 
-\Ne[m_b]^{-1}(Q[m,\delta m]^*\oF[m_b]^*\oF[m_b]\bre[m_b,d] + \oF[m_b]^*\oF[m_b]Q[m,\delta m])\bre[m_b,d]
\]
\[
 + \Ne[m_b]^{-1}Q[m,\delta m]^*\oF[m_b]^*\de)
\]
\[
= 
-\Ne[m_b]^{-1}([Q[m,\delta m]^*,\oF[m_b]^*\oF[m_b]] + \oF[m_b]^*\oF[m_b] (Q^*[m,\delta m]+Q[m,\delta m]))\bre[m_b,d]
\]
\[
 + \Ne[m_b]^{-1}Q[m,\delta m]^*\oF[m_b]^*\de)
\]
Using (\ref{l2normal}) and the definitions of $Q,R$, this is
\[
(-\Ne[m_b]^{-1}[Q[m,\delta m]^*,\oF[m_b]^*\oF[m_b]]-(I-\lambda \Ne[m_b]^{-1})R[m,\delta m])\bre[m_b,d]
\]
\[
+ \Ne[m_b]^{-1}([Q[m,\delta m]^*,\oF[m_b]^*\oF[m_b]] + \Ne[m_b]Q[m,\delta m]^*)\bre[m_b,d]
\]
\[
=(Q[m,\delta m]^*-(I-\lambda \Ne[m_b]^{-1})R[m,\delta m])\bre[m_b,d]
\]
\[
= -Q[m,\delta m]\bre[m_b,d] + \lambda \Ne^{-1}R[m,\delta m]\bre.
\]
as claimed.

The stated expression for $\nabla \Je[m_b]$ follows. The estimate amounts to a bound on the second term on the RHS. To see this, note that according to Theorem 4, $\bre[m_b,d]$ satisfies the hypotheses of Lemma A-4 with $s=0, k=1, M \leftarrow \oF[m_b]^*\oF[m_b], q \leftarrow \bre[m_b,d]$, and $\Pi \leftarrow \Pi[m_b]$. Accordingly from (\ref{a4:res1}) follows
\[
\|\bqe[m_b]\|_0 \le C\|\bre[m_b,d]\|_0.
\]
Since $R[m,\delta m]$ is continuous $OPS^0$-valued, the estimate follows from Theorem 1.

Finally, note that
\[
|\langle Q[m,\delta m]\br,A^*A \br\rangle| = |\langle ([A,Q[m,\delta m]]+Q[m,\delta m]A)\br,A\br\rangle|
\]
\[
=|\langle[A,Q[m,\delta m]]\br+\frac{1}{2}R[m,\delta m]A\br,A\br\rangle| \le C\|\br\|_0^2,
\]
since the operators appearing in the last expression are $\in OPS^0$.
Q. E. D.

The estimate (\ref{eqn:firstgradest}) suggests the use of the first
term in (\ref{eqn:gradq}) as an approximation to the gradient: in
essence, the use of the Factorization Lemma has isolated the result
$\bqe$ of the otherwise apparently necessary second system solve in a
term which is uniformly $O(\lambda)$. As $\lambda$ will in
applications be proportional to a wavelength, the achievable
resolution of the inversion, it seems reasonable to neglect this
term. Thus $\bqe$ need not be computed.

However the key player in the first term in (\ref{eqn:gradq}), namely
the operator $Q$, is not directly computable. Fortunatly, key examples
add yet another property to the extended modeling operator, formalized
in

\noindent {\bf Assumption 6:} For $m_b \in U$, $\oF[m_b]^{\dagger}$
denotes a {\em partial parametrix} for $\oF[m_ b]$, in the sense that
for every $s \in \bf{Z}$, $k \in \bf{Z}^*$,
\be
\label{eqn:paradef}
\bar{\Pi}_D (I-\oF[m_b]\oF[m_b]^{\dagger}) \in {\cal L}(D_s,D_{s-k}).
\ee

\noindent {\bf Remark 7}: In fact, the inequality (\ref{eqn:paradef}) will be required only for 
a finite range of $k$, actually for $k=2$, and can even be concocted
from an operator satisfying (\ref{eqn:paradef}) for $k=1$. This fact
is important, as common algorithms for constructing such partial
parametrices conveniently produce operators satisfying
(\ref{eqn:paradef}) only for $k=1$. %In fact write $\bar{\Pi}_D(I-

\section{Linearized Extended Inversion}
The first order of business is to describe the key model
properties. An example possessing these properties will be described
later.

\begin{itemize}
\item The space of space of p-parameter reference models $M \subset
  (H_{\rm loc}^m(\bR^3))^p,\,m>>3/2$. Such models must have a number of continuous
  derivatives, so $m$ must be quite large.
\item The feasible reference model set $U \subset M$ is defined by
  pointwise bounds.
\item Perturbational extended model spaces $\oM_s \subset (H^s_{\rm
    loc}(\bR^3 \times \bR^2))^p$, $s=0, \pm 1, \pm 2,...$, Hilbert norms $\|\cdot\|_{s} = \|\Lambda^s \cdot\|_0$, where $\Lambda = \sqrt{I-l^2\nabla^2}$ is the usual Sobolev weight operator, extended to act diagonally on vector-valued functions, with nondimensionalizing weight $l$ (units of length). 
\item Corresponding data spaces $D_s \subset \subset (H^s_{\rm
    loc}(\bR^3 \times \bR^2))^d$, $s=0, \pm 1, \pm 2,...$, 
\item Linearized extended forward map = $\oF_{\rm raw} \cap_{s \in
    {\rm Z}} C^1(U , {\cal L}(\oM_s,D_s)$ 
\item Semblance operator (``annihilator'') = $A \in \cap_{s \in
    {\rm Z}} {\cal L}(\oM_s,\oM_s)$;
\item Precompact model and data supports $\Omega_M, \Omega_D$. Pre-
  and post-cutoff versions of $\oF, A$ so that $\oM_s$ can be
  replaced by $(H^s_0(\Omega_M))^p$, $D_s$ by
  $(H^s_0(\Omega_D))^d$. $\Omega_M$ is precompact in $\bR^5$,
  $\Omega_D$ is a precompact smooth manifold of dimension 5 in $R^6$.
\end{itemize}

These abstract the structure of a typical wave inverse problem setting for seismology, with a number of critical assumptions, detailed in the appendix, along with an important example. The examples have a number of crucial additional features:

\noindent {\bf Assumption 1:} For $m \in U$, both $\oF[m_b]^*\oF[m_b]$ and $\oF[m_b]\oF[m_b]^*$ are uniformly bounded on each $\oM_s$ respectively $D_s$ for every $s \in {\bf Z}$, uniformly in $m \in U$.

\noindent {\bf Assumption 2:} $\oF[m_b]^*\oF[m_b] \in OPS^0(\Omega_M)$,
$\oF[m_b]\oF[m_b]^* \in OPS^0(\Omega_D)$. Here $OPS^k$ denotes $p \times
p$, resp. $d \times d$, matrix pseudodifferential operators of order
k. Officially, reference models must be $\in C^{\infty}$ for this to
literally true, however the exponent $m$ in the first bullet above
should be chosen large enough that it makes no
difference. Alternative, restrict $M$ to a finite-dim'l subspace of
smooth models.

\noindent {\bf Remark:} Extended modeling is {\em necessary} in order that $\oF\oF^*$ is $\Psi$DO;

\noindent {\bf Assumption 3 - Uniform Microlocal Ellipticity:}  $\Gamma_D \subset T^*(Y)$ is conic so that $\oF[m_b]\oF[m_b]^*$ is elliptic in $\Gamma_D$ uniformly for $m \in U$.

\noindent {\bf Remark:} Denote by ${\cal C}[m_b]$ the canonical relation of $\oF[m_b]$. According to Egorov's Theorem, $\oF[m_b] P \oF[m_b]^*$ is elliptic in $\Gamma_D$ if $P \in OPS^0$ is elliptic in ${\cal C}[m_b]^{-1}\Gamma_D \equiv \Gamma[m_b]$. Also $\oF[m_b]^*\oF[m_b]$ is elliptic in $\Gamma[m_b]$.

The final two assumptions concern properties of the data, expressed in
terms of homogeneous inequalities defining cones in $D_s$. The results
to follow are {\em aymptotic}, so concern, not a single datum but a family
$\{\de\}_{\lambda \in (0,\lambda_{\rm max}}$ of data. The index
parameter $\lambda$ will turn out to signify (roughly) wavelength.

The first property expresses microlocal concentration of energy.
Pseudodifferential ``projection'' operators $\Pi_D, \bar{\Pi}_D \in
OPS^0$ will appear throughout the following results: these are chosen
to satisfy 
\[
\mbox{ess supp } \bar{\Pi}_D \subset \Gamma_D,
\]
\[
\sigma_0(\bar{\Pi}_D) = 1 \mbox{ on ess supp } \Pi_D
\]

In the following two Assumptions, $k_0>0$ is 
independent of $m \in U$ and $\lambda \in (0,\lambda_{\rm max}]$.

\noindent {\bf Assumption 4 - Data Microlocalization:} There exists $L
\ge 0$ so that for $0 \le s,  k \le k_0, \lambda \in
(0, \lambda_{\rm max}]$,
\be
\label{micro}
\|(I- \Pi_D)\de\|_s \le L_s \|\de\|_{s-k}.
\ee

The last property makes the connection with wavelength, and expresses
the assumption that the data is concentrated at high frequency:

\noindent {\bf Assumption 5 -  High Frequency Cone Condition:} 
For each $s \ge -k_0$ there exists $K_s
\ge 0$ so that for  $\lambda \in
(0, \lambda_{\rm max}]$,
\be
\label{hfcd}
\|\de\|_{s} \le K_s \lambda\|\de\|_{s+1}
\ee

The estimates (\ref{micro}) and (\ref{hfcd}) involve constants independent of the parameter $\lambda \in (0,\lambda_0]$, so it is not immediately obvious that these assumptions define non-empty cones. To build a family $\{\de: \lambda \in (0,\lambda_0]\}$ satisfying estimates of the forms (\ref{micro}) and (\ref{hfcd}). choose $u \in C_0^{\infty}(\bR^3)$ and define $\Phi_{\lambda}(\bx)=(x_1/\lambda,x_2,x_3)$, $u_{\lambda} = \Phi_{\lambda}^*u, \de= \Lambda^{k_0}u_{\lambda}$. Recall the simple anisotropic Poincar\'{e} inequality:
\[
\int u^2 = \int_0^1dx_2 \int_0^1dx_3 \int_0^1dx_1\left(\int_0^{x_1} D_1u\right) ^2
\]
\[
\le \int_0^1dx_2 \int_0^1dx_3 \int_0^1dx_1\left(x_1\int_0^{x_1} (D_1u)^2\right)
\le \int (D_1u)^2
\]
whence
\[ 
\int u_\lambda^2 = \int (\Phi_{\lambda}^{-1})^*u_{\lambda} |det D(\Phi_{\lambda}^{-1})|
 = \lambda \int u^2 \le \lambda \int (D_1u)^2 = \lambda^3 \int (\Phi_{\lambda}^{-1})^*(D_1 u_{\lambda})^2 
\]
\[
= \lambda^2 \int (D_1 u_{\lambda})^2 \le \lambda^2 (u^2 + l^2|\nabla u|^2), 
\]
so
\[
\|u_{\lambda}\|_0 \le \lambda \|D_1 u_{\lambda}\| \le \lambda \|u_{\lambda}\|_1.
\]
Since $D^{\alpha}u_{\lambda}$ is of the same type as $u_{\lambda}$, the same inequality holds, hence for $s \ge 0$  
\[
\|u_{\lambda}\|_s \le C_s \lambda \|u_{\lambda}\|_{s+1}.
\]
Then
\[
\|\de\|_s = \|u_{\lambda}\|_{s+k_0} \le C_{s+k_0}\lambda\|u_{\lambda}\|_{s+k_0+1} = C_{s+k_0}\lambda\|\de|_{s+1}
\]
for $s \ge -k_0$, that is, $\{\de:\lambda > 0\}$ satistfies (\ref{hfcd}) with $K_s = C_{s+k_0}$.

Choose a pseudodifferential partition of unity $\{\Pi_D,I-\Pi_D\}$ with
the scalar matrix symbol of $\Pi$ being
\begin{itemize}
\item independent of $\bx$,
\item homogeneous of order 0 outside a neighborhood of the zero, 
  section, with  values between 0 and 1;
\item $\equiv 1$ in a conic neighborhood of $\{(\bx,{\bf
    \xi}):\xi_2=\xi_3=0\}$, say  $\{(\bx,{\bf \xi}):|\xi_1|\le \beta
  |\xi'|\}$ ($\xi'=(\xi_2,\xi_3)$); 
\item $\equiv 0$ in a conic neighborhood of  $\{(\bx,{\bf
    \xi}):\xi_1=0\}$. 
\end{itemize}
Then for $0\le k \le k_0, s\ge-k_0+k$, 
\[
\|(I-\Pi)\de\|_s^2  = \int d\xi |\hat{u_\lambda}(\xi)|^2\Pi(\xi)^2\Lambda(\xi)^{2(s+k_0)} 
\]
\[
\le \int_{|\xi_1| \le \beta|\xi'|} d\xi |\hat{u_\lambda}(\xi)|^2\Lambda(\xi)^{2(s+k_0)} 
\]
\[
= \lambda^2 \int_{|\xi_1| \le \beta|\xi'|} d\xi
\left|\hat{u}\left(\lambda \xi_1,\xi'\right)\right|^2\Lambda(\xi)^{2(s+k_0)} .
\]
Since $\Lambda(\xi) \le (1+\beta)\Lambda(\xi')$ provided that $|\xi_1|
\le \beta |\xi'|$, this is
\[
\le (1+\beta)^{2k} \lambda^2 \int_{|\xi_1| \le \beta|\xi'|} d\xi
\left|\hat{u}\left(\lambda
    \xi_1,\xi'\right)\right|^2\Lambda(\xi')^{2k}\Lambda(\xi)^{2(s+k_0-k)} 
\]
\[
=(1+\beta)^{2k} \lambda^2 \int_{|\xi_1| \le \beta|\xi'|} d\xi
\left|\widehat{\Lambda(D')^ku}\left(\lambda
    \xi_1,\xi'\right)\right|^2\Lambda(\xi)^{2(s+k_0-k)}  \le (1+\beta^{2k})\|\Lambda(D')^ku_{\lambda}\|_{s+k_0-k}
\]
Since $s+k_0-k \ge 0$, the alternate expression for the norm is
available:
\[
\le C(1+\beta)^2 \sum_{ |\alpha| \le s+k_0-k}
\|D^{\alpha}\Lambda(D')^ku_{\lambda}\|_0^2 
\]
\[
\le \le C(1+\beta)^2 \sum_{
  0 \le \alpha_1 \le s+k_0-k, |\alpha'| \le s+k_0-k-\alpha_1}
\lambda^{1-2\alpha_1}\|D_1^{\alpha_1}(D')^{\alpha'}\Lambda(D')^ku\|_0^2
\]
\[
 \le C_1C(1+\beta)^2 \sum_{
  0 \le \alpha_1 \le s+k_0-k, |\alpha'| \le s+k_0-k-\alpha_1}
\lambda^{1-2\alpha_1}\|D_1^{\alpha_1}(D')^{\alpha'}u\|_0^2
\]
\[
 \le C_1C(1+\beta)^2 \sum_{
  0 \le \alpha_1 \le s+k_0-k, |\alpha'| \le s+k_0-k-\alpha_1}
\|D_1^{\alpha_1}(D')^{\alpha'}u_{\lambda}\|_0^2
\]
\[
\le C_1C(1+\beta)^2 \sum_{ |\alpha| \le s+k_0-k}
\|D^{\alpha}u_{\lambda}\|_0^2 \le C_2 C_1 C (1+\beta)^2
\|u_\lambda\|_{s+k_0-k} \le L \|\de\|_{s-k}
\]
in which 
\[
C_1 = \max_{|\alpha| \le
  s+k_0-k}\frac{\|D^{\alpha}\Lambda(D')^ku\|_0^2}
{\|D^{\alpha}u\|_0^2}
\]

\noindent{\bf Lemma 1:} There exist $C_*,\lambda_0>0$, $\lambda_0 \le \lambda_{\rm max}$ depending on the same quantities as $K,L$ so that for $\lambda \in (0,\lambda_0]$, $m \in U$, $s=-k_0,...,t$, 
\be
\label{lemma1:res}
\|\de\|_s \le C_*\|\oF[m_b]^*\de\|_s.
\ee

\noindent{\bf Proof:} 
The principal symbol of $\oF[m_b]\oF[m_b]^*$ is nonnegative, and positive in $\Gamma_D$:
\[
\sigma_0(\oF[m_b]\oF[m_b]^*) \ge \gamma^2 \mbox{ in } \Gamma_D,
\]
so by a standard construction there exists a microlocal square root $B[m_b] \in OPS^0$ with real nonnegative principal symbol, for which 
\[
\sigma_0(B[m_b]) \ge \gamma \mbox{ in } \Gamma_D,
\]
\[
(\oF[m_b]\oF[m_b]^*-B[m_b]^*B[m_b]) \Pi_D \in OPS^{-2},
\]
whence if $\lambda \in (0,\lambda_{\rm max}]$,
\[
\|\oF[m_b]^*\de\|^2_s = \langle \de, \oF[m_b]\oF[m_b]^* \de \rangle_s =
\|B[m_b]\de\|_s+ \langle \de,(\oF[m_b]\oF[m_b]^*-B[m_b]^*B[m_b]) \Pi_D\de
\rangle_s
\]
\[
+ \langle \de,(\oF[m_b]\oF[m_b]^*-B[m_b]^*B[m_b]) (I-\Pi_D)\de \rangle_s \ge
\|B[m_b]\Pi_D\de\|^2_s  -\|B[m_b](I-\Pi_D)\de\|^2_s  - C\|\de\|^2_{s-1}
\]
\[
\ge \gamma \|\de\|^2_s - C\|\de\|^2_{s-1} \ge (\gamma - C_1\lambda^2)\|\de\|^2_s,
\]
using Lemma A1 (microlocal G\aa rding inequality), $L^2$ estimates, and Assumptions 4 and 5 (with $k=1$). Take $\lambda_0 = \mbox{min }(\lambda_{\rm max}, (\gamma/2C_1))^{1/2})$ (for instance): then for $0 < \lambda \le \lambda_0$,
\be
\label{lemma1:eq5}
\|\oF[m_b]^*\de\|_s \ge \frac{\gamma}{2} \|\de\|_s
\ee

Q. E. D.

From her on replace $\lambda_{\rm max}$ by $\lambda_0$.

\noindent{\bf Lemma 2:} There exist $K_0, L_0 \ge 0$ so that for $0
\le s \le t, 1 \le k \le k_0, \lambda \in (0, \lambda_{\rm max}]$,
\be
\label{lemma2:res1}
\|(I-\Pi[m_b])\oF[m_b]^*\de\|_s \le L_0 \|\oF[m_b]^*\de\|_{s-k}
\ee
\be
\label{lemma2:res2}
\|\oF[m_b]^*\de\|_{s-1} \le K_0\lambda\|\oF[m_b]^*\de\|_{s}
\ee

\noindent{\bf Proof:} 
Egorov's Theorem (REF) implies that $\oF[m_b]^*\Pi_D = \Pi[m_b]\oF[m_b]^*+R[m_b]$, in which $\Pi[m_b] \in OPS^0$ has principal symbol $=1$ in $\Gamma[m_b]$, and $R[m_b] \in OPS^{-\infty}$.  
\[
\|(\Pi[m_b]-I)\oF[m_b]^*\de\|_s = \|\oF[m_b]^*(\Pi_D-I)\de\|_s + \|R[m_b]\|_{s-k,s}\|\de\|_{s-k}  
\]
\[
\le (\|\oF[m_b]^*\|_{s}L_k+\|R[m_b]\|_{s-k,s})\|\de\|_{s-k} \le L_{0,k}\|\oF[m_b]^*\de\|_{s-k},
\]
from (\ref{lemma1:res}), with $L_{0,k} = (\|\oF[m_b]^*\|_{s}L_k+\|R[m_b]\|_{s-k,s})C_*$.

Similarly,
\[
\|\oF[m_b]^*\de\|_{s-1} \le \lambda\|\oF[m_b]^*\|_{s-1}\|\de\|_{s} \le \lambda K_0\|\oF[m_b]^*\de\|_s,
\]
with $K_0 = \|\oF[m_b]^*\|_{s-1}C_*$.

Q. E. D.

\noindent {\bf Lemma A1: Standard Microlocal G\aa rding Estimate:} Suppose that $\Gamma \subset T^*\bR^n$ is conic open, and let $\Pi$, $\bar{\Pi} \in OPS^0$ be such that 
\begin{itemize}
\item[1. ]$\mbox{ess supp} \,\bar{\Pi} \subset \Gamma$;
\item[2. ]$(\bar{\Pi}-I)\Pi \in OPS^{-\infty}$
\end{itemize}
Suppose that $M \in OPS^0$ is microlocally elliptic in $\Gamma$: for certain $\gamma, R>0$,  $\sigma_0(M)(\bx,\xi)  \ge \gamma > 0$ for $(\bx,\xi) \in \Gamma$, $|\xi|>R$.   Then for $u \in H^s(\bR^n)$, any $k \ge 0$, there exists $C_{s,k} \ge 0$ depending on $\Gamma$ and on the symbol semi-norms of  $\bar{\Pi}, \Pi$ and $M$ so that
\be
\label{garding}
\|M\Pi u\|_s \ge \gamma \|\Pi u\|_s - C_{s,k} \|u\|_{s-k}.
\ee

\noindent Proof: Define
\[
\tilde{M} = M \bar{\Pi} + \gamma (I-\bar{\Pi}).
\]
Then $\sigma_0(\tilde{M})(\bx,\xi) \ge \gamma$ so by standard G\aa rding, 
\[
\|\tilde{M}(\Pi u)\|_s \ge \gamma \|\Pi u\|_s - C \|u\|_{s-k},
\]
with $C \ge 0$ as described. Since $(I-\bar{\Pi})\Pi \in OPS^{-\infty}$, the conclusion follows with a different $C$ of same type. Q. E. D.

\noindent {\bf Lemma A2: Standard Microlocal Parametrix Construction:} With the same hypotheses, there exist $M^{\dagger} \in OPS^0$ so that for $s \in {\bf Z}, k \in {\bf Z}_+$, there are $K_s, C_{s,k} \ge 0$ so that for $u \in H^s(\bR^n)$,
\[
\|\Pi (MM^{\dagger}u - u)\|_s \le C_{s,k} \|u\|_{s-k},
\]
\[
\|\Pi (M^{\dagger}Mu - u)\|_s \le C_{s,k} \|u\|_{s-k},
\]
\[
\|M^{\dagger}u\|_s \le \frac{1}{\gamma}\|u\|_s + K_s \|u\|_{s-1}.
\]
The constants may be taken uniform over any bounded subset of $M \in OPS^0$ that are uniformly elliptic in $\Gamma$.

\noindent {\bf Lemma A3:} Same hypotheses as Lemma A1, and $P \in OPS^0$. Suppose $t \ge 0$ and $q \in H^t(\bR^n)$ satisfies
\be
\label{microlocal}
\|\Pi q - q \|_s \le L\|q\|_{s-k}
\ee
for $0 \le s \le t, 1 \le k \le k_0 > 0$. Then there exists $p\in H^t(\bR^n)$ so that
\[
\|Mp - Pq\|_s \le C_{s,k}\|q\|_{s-k} 
\]
for $0 \le s \le t, 1 \le k \le k_0 > 0$, with $C$ depending on the constant in the $\Gamma$-ellipticity estimate of $M$,  semi-norms of $\Pi, \bar{\Pi}, M$, and $P$, and on $L$.

\noindent Proof: Let $M^{\dagger}$ be a $\Gamma$-microlocal parametrix as in Lemma A2, set
\[
p = M^{\dagger}Pq.
\]
Then for $0 \le s \le t$, 
\[
\|Mp-Pq\|_s \le \|\bar{\Pi}(MM^{\dagger}-I)Pq\|_s + \|(I-\bar{\Pi})(MM^{\dagger}-I)P\Pi q\|_s 
\]
\[
+ \|(I-\bar{\Pi})(MM^{\dagger}-I)P(I-\Pi) q\|_s
\]
For $k=1,...,k_0$, the first term is bounded by a multiple of $\|q\|_{s-k}$ as a consequence of the parametrix bound. The operator in the second term is $\in OPS^{-\infty}$, and the third term is also bounded by a multiple of $\|q\|_{s-k}$ due to the microlocal bound on $q$. All of the constants are uniform as indicated. Q. E. D.

\noindent {\bf Lemma A4:} Suppose that $M \in OPS^0$ is positive semidefinite on $H^0$. Then $M_{\lambda} \equiv M + \lambda I$ is invertible on $H^s$ for $s \in Z^+$, and for $C_{s} \ge 0$ independent of $\lambda \in (0,\infty)$,
\be
\label{a4:res0}
\|M_{\lambda}^{-1}q \|_s \le C_s\sum_{j=0}^s \lambda^{-(j+1)} \|q\|_{s-j},\,\,q \in H^s.
\ee
\noindent {\bf Proof:} $M_{\lambda}$ is invertible on $H^0$, as a consequence of the bounds
\[
C \|u\|^2  \ge \langle u, M_{\lambda}u\rangle_0 \ge \lambda \|u\|_0^2
\]
and the Lax-Milgram Theorem. In particular $\|M_{\lambda}^{-1}q\|_0\le \lambda^{-1}\|q\|_0$, that is, the case $s=0$ is established.

For $q \in H^s$, $M_{\lambda}^{-1}q \in H^s$ as a consequence of the existence of a parametrix $\in  OPS^0$. Note that for $u \in H^s$, 
\[
\frac{\lambda}{4}\|u\|_s^2 + \frac{1}{\lambda}\|M_{\lambda}u\|_s^2 \ge \|u\|_s\|M_{\lambda}u\|_s \ge \langle \Lambda^s u, \Lambda^s M_{\lambda} u\rangle_0  
\]
\[
= \langle \Lambda^s u, M_{\lambda} \Lambda^s u\rangle_0 + \langle \Lambda^s u, [\Lambda^s, M_{\lambda}]u\rangle_0 
\ge \lambda \|u\|_s^2 - C\|u\|_s \|u\|_{s-1} \ge \frac{\lambda}{2}\|u\|_s^2-\frac{C^2}{2\lambda}\|u\|_{s-1}^2,
\]
whence
\[
\|M_{\lambda}u\|_s + \frac{C^2}{2}\|u\|_{s-1} \ge \sqrt{2} (\|M_{\lambda}u\|_s^2 + \frac{C^2}{2}\|u\|_{s-1}^2)^{\frac{1}{2}} \ge \frac{\lambda}{\sqrt{2}}\|u\|_s
\]
Take $u=M_{\lambda}^{-1}q$, and suppose that the conclusion holds for $s-1$. Then
\[
\frac{\lambda}{\sqrt{2}}\|M_{\lambda}^{-1}q\|_s \le \|q\|_s + \frac{C^2}{2}\|M_{\lambda}^{-1}q\|_{s-1} 
\]
\[
\le \|q\|_s + \frac{C^2}{2}C_{s-1}\sum_{j=0}^{s-1}\lambda^{-(j+1)}\|q\|_{s-1-j}.
\]
Rearranging and renaming constants completes the induction step. Q. E. D.

\noindent {\bf Lemma A5:} Suppose that $M \in OPS^0$ is microlocally elliptic in $\Gamma$, self-adjoint and positive semidefinite on $H^0$. Suppose $t \ge 0$ and $q \in H^t(\bR^n)$ satisfies (\ref{microlocal}) for $0 \le s \le t, k=1,...,k_0 > 0$. 
Then there exist $C_{s,k} \ge 0$ and $\lambda_0>0$ so that for $\lambda \in (0,\lambda_0]$,
 \be
 \label{a4:res1}
\| M_{\lambda}^{-1} P q \|_s \le C_{s,k} (\|Pq\|_s + \sum_{j=0}^s \lambda^{-(j+1)} \|q\|_{s-j-k}).
\ee
Denote by $M_{\lambda}^{\dagger}$ the $\Gamma$-microlocal parametrix for $M_{\lambda}$ constructed in Lemma A-2, and note that $M_{\lambda}^{\dagger}$ is bounded on each $H^s$ with bound independent of $\lambda$. Then
\be
\label{a4:res2}
\|M_{\lambda}^{-1} P q - M_{\lambda}^{\dagger} Pq\|_s \le C_{s,k} \sum_{j=0}^s \lambda^{-(j+1)} \|q\|_{s-j-k}.
\ee
Finally,
\be
\label{a4:res3}
\|(I-\bar{\Pi})M_{\lambda}^{-1}Pq\|_s \le C_{s,k} \sum_{j=0}^s \lambda^{-(j+1)} \|q\|_{s-j-k}.
\ee
for $0 \le s \le t, k=1,...,k_0 > 0$. 
The constants $C_{s,k}$ depend on $s,k$ and the symbol semi-norms of $M, P, \Pi, \bar{\Pi}$ and on $L$, but are independent of $\lambda \in (0,\lambda_0]$.

\noindent {\bf Proof:} Apply Lemma A3 to $M_{\lambda}$ to construct $p \in H^t(\bR^n)$ so that for $0 \le s \le t$, $0 \le k \le k_0$,
\[
\|M_{\lambda}p-Pq\|_s \le C\|q\|_{s-k}.
\]
Note that $\{M_{\lambda}: \lambda \in (0,\lambda_0]\}$ is bounded in $OPS^0$ and uniformly $\Gamma$-elliptic, hence the constants $C$ from Lemma A3 are also independent of $\lambda$. 
Then
\[
\|M_{\lambda}^{-1}Pq\|_s \le \|p\|_s + \|M_{\lambda}^{-1}(Pq-M_{\lambda}p)\|_s 
\]
\[
\le \|p\|_s + C_s\sum_{j=0}^s \lambda^{-(j+1)}\|Pq-M_{\lambda}p\|_{s-j} 
\]
\[
\le \|M_{\lambda}^{\dagger}Pq\|_s + C_s\sum_{j=0}^s \lambda^{-(j+1)} C_{s-j,k}\|q\|_{s-j-k}
\]
\[
\le \|M^{\dagger}\|\|Pq\|_s + C_{s,k} \sum_{j=0}^s \lambda^{-(j+1)} \|q\|_{s-j-k}
\]
after suitable redefinition of constants.
Also
\[
\|M_{\lambda}(M_{\lambda}^{-1} P q - M_{\lambda}^{\dagger} Pq)\|_s = \|Pq-M_{\lambda}p\|_s
\]
\[
\le C\|q\|_{s-k},
\]
from Lemma A3, whence (\ref{a4:res2}) follows from (\ref{a4:res0}). Finallly, (\ref{a4:res2}) implies
\[
\|(I-\bar{\Pi})M_{\lambda}^{-1}Pq\|_s \le \|M_{\lambda}^{-1} P q - M_{\lambda}^{\dagger} Pq\|_s + \|(I-\bar{\Pi})M_{\lambda}^{\dagger}Pq\|_s
\]
\[
\le C_{s,k}\sum_{j=0}^{s}\lambda^{-(j+1)}\|q\|_{s-j-k} + \|(I-\bar{\Pi})M_{\lambda}^{\dagger}P \Pi q\|_s
\]
\[
+\|(I-\bar{\Pi})M_{\lambda}^{\dagger}P(I-\Pi)q\|_s
\]
The operator inside the second term on the RHS is smoothing, and the third term is $O(\|q\|_{s-k})$ for any $k=1,...,k_0$ by hypothesis, hence bounded by a sum of the given form, in fact by its $j=0$ term. Once again redefining the $C_{s,k}$ yields (\ref{a4:res3}).

Q. E. D.

\noindent {\bf Corollary A6:} Suppose that in addition to the assumptions of Lemma A6, $\{q_{\lambda}: 0<\lambda<\lambda_{\rm max}\} \subset H^t$ satisfies 
\be
\label{hfc}
\|q_{\lambda}\|_{k} \le K \lambda \|q_{\lambda}\|_{k+1}
\ee
for $-k_0 \le k < t$, $K$ independent of $\lambda$, $k_0 \ge 2$. Then there exists
$C >0$ so that for $0 \le s
\le t$, 
\be
\label{main_est}
\|M_{\lambda}^{-1}Pq_{\lambda}\|_s \le C \|q_{\lambda}\|_s.
\ee
and for $1 \le k \le k_0$,
\be
\label{microlocal_est}
\|(I-\bar{\Pi})M_{\lambda}^{-1}Pq_{\lambda}\|_s \le C \lambda\|q\|_{s-k+2}.
\ee
If additionally $P=I$ and $k_0 \ge 2$, then $L_1, K_1>0$ exist so that for $0 \le s \le t, 1 \le k \le \min(k_0,s+1)$,
\be
\label{hfc_sol}
\|M_{\lambda}^{-1}q_{\lambda}\|_{s-1} \le K_1\lambda\|M_{\lambda}^{-1}q\|_{s}
\ee
and
\be
\label{microlocal_sol}
\|(I-\bar{\Pi})M_{\lambda}^{-1}q_{\lambda}\|_s \le L_1\lambda\|M_{\lambda}^{-1}q_\lambda\|_{s}.
\ee
for $s=0,...t$.

\noindent{\bf Proof:} 
The inequalities (\ref{hfc}) imply that 
\[
\|q_{\lambda}\|_{s-j-k} \le K \lambda \|q_{\lambda}\|_{s-(j-1)-k} \le ...\le (K\lambda)^{j} \|q_{\lambda}\|_{s-k}.
\]
So all terms in the sum on the RHS of (\ref{a4:res0}) are bounded by multiples of $\lambda^{-1}\|q_{\lambda}\|_{s-k}$, and add up to
\[
\|M_{\lambda}^{-1}Pq_{\lambda}\|_s \le C (\|Pq_{\lambda}\|_s + \lambda^{-1} \|q_{\lambda}\|_{s-k} )
\]
Since $k \ge 1$, (\ref{hfc}) implies (\ref{main_est}), with the constant $C$ inheriting its properties from standard bounds on $P$, $K$, and the $C_{s,k}$.

Apply the same trick to (\ref{a4:res3}) to obtain 
\[
\|(I-\bar{\Pi})M_{\lambda}^{-1}Pq_{\lambda}\|_s \le C
\lambda^{-1}\|q\|_{s-k}.
\]
Since $s \ge 0$, $s-k \ge -k_0$, so you can apply (\ref{hfc}) twice to obtain (\ref{microlocal_est}).

To show (\ref{hfc_sol}), write
\[
\|M_{\lambda}^{-1} q \|_{s-1} \le \|M_{\lambda}^{-1} q_{\lambda} - M_{\lambda}^{\dagger} q_{\lambda}\|_s + \|M_{\lambda}^{\dagger} q_{\lambda}\|_{s-1}
\]
For $s =0, ..., t$, (\ref{a4:res2}) together with the trick applied twice already using (\ref{hfc}) estimates the first term as
\[
\le  C \lambda^{-1}\|q_{\lambda}\|_{s-k} \le C \lambda \|q_{\lambda}\|_{s}
\]
taking $k=2$, as we may thanks to $k_0 \ge 2$. Lemma A-2 provides a uniform (in $\lambda$) estimate for $\|M_{\lambda}^{\dagger}\|_{s-1,s-1}$ for any $s \in {\bf Z}$: 
\[
\|M_{\lambda}^{\dagger} q\|_{s-1} \le C \|q\|_{s-1} \le C\lambda \|q\|_{s}
\]
Combining these estimates, obtain
\[
\|M_{\lambda}^{-1} q \|_{s-1} \le C \lambda \|q\|_{s} \le C\lambda \|M_{\lambda}^{-1} q \|_{s}.
\]

Finally, for $P=I$ and $k=2$, (\ref{microlocal_est}) takes the form
\[
\|(I-\bar{\Pi})M_{\lambda}^{-1}q_{\lambda}\|_s \le C \lambda\|q\|_{s}
\]
\[
\le C\lambda\|M_{\lambda}\|_{s,s}\|M_{\lambda}^{-1}q_{\lambda}\|_{s}
\]
which gives (\ref{microlocal_sol}) with proper choice of $L_1$.

Q. E. D.

\bibliographystyle{seg}
\bibliography{../bib/masterref}

### 2019-10-19 repro/papers/iva/paper.tex
repro docs dir added

\title{Gradients of Reduced Objective Functions for Separable Wave
  Inverse Problems}
\date{}
\author{William W. Symes, The Rice Inversion Project}
\lefthead{Symes}
\righthead{Accurate Reduced Gradients}
\maketitle
\parskip 12pt
\begin{abstract}
The derivative of the forward (modeling) map for wave inverse problems
is unbounded in the natural metric of the model and data spaces. This
unboundedness complicates the accurate computation of the reduced
gradient in separable wave inverse problems: iterative estimation of
the linear variables may not lead to convergent computation of the
gradient. If the linear part of the modeling operator possesses a computable
inverse up to a smoothing error, however, and if the data is
high-frequency in a precise sense, then the gradient may be decomposed
into a computable non-iterative part and a convergent remainder. 
The resulting form of the
gradient error control is well-suited to integration with inexact trust-region
optimization methods.

\end{abstract}

\section{Introduction}
Separable (partly linear) modeling operators for wave inverse problems
arise either via Born approximation
\cite[]{KerSy:94,ChaurisNoble:01,MuldertenKroode:01,
  ShenSymes:08,WeibullArntsen:13,BiondiAlmomin:14} or by including
source (right-hand side) parameters as model degrees of freedom
\cite[]{SongSymes:94b,Plessix:00,PrattSymes:02,LuoSava:11,Warner:14,LeeuwenHerrmannWRI:13}.
Extended model spaces add yet more degrees of freedom to the linear
parameters of a separable modeling operator, to permit data fit for a
large set of nonlinear parameter (``velocity model'') values. The
non-extended (``physical'') subspace is the kernel of an
``annihilator''. Minimizing the norm of the annihilator output over
all data-fitting models picks out an approximation to the solution of
the physical least squares inverse problem. A natural approach to this
minimization is via elimination of the linear variables to create a
reduced function of the nonlinear variables
\cite[]{KerSy:94,vanLeeuwenMulder:09,LiRickettAbubakar:13}. If the
nonlinear parameters include velocities, indices of refraction, or
equivalent, then the modeling operator is differentiable with loss of
regularity: a convergent sequence of approximate solutions of the
data-fit problem does not (necessarily) lead to a convergent sequence
of derivative values. Consequently, the computable approximate
solutions of the data-fit problem do not lead to convergent gradient
approximations for the reduced objective, which may reduce the
effectiveness of iterative optimization strategies
\cite[]{HuangSymes:SEG15}.

In the paragraphs to follow, I show that the gradient computation may
be modified to become convergent if three conditions are satisfied:
\begin{itemize}
\item the linear part of the forward modeling operator possesses a
  computable parametrix (inverse modulo compacts, or ``asymptotic
  inverse'') in the form of its adjoint with respect to suitable
  weighted norms, and
\item the inverse problem is embedded in a family of problems indexed
  by a data frequency parameter.
\item the (inevitable) Tihonov regularization parameter is coupled to
  data frequency.
\end{itemize}
Then the reduced gradient 
computation may be modified to become convergent as well.

Reliance on parametrices might seem surprising. However, parametrices
for linearized inverse wave problems have a long history, mostly as a
subtopic of computational geometric optics
\cite[]{CohBlei:77,Beyl:85,BeylBurr:90,VirJinMadLam:92,ArayaSymes:96,Operto:00}. Recently
several authors have described direct (``wave equation'') parametrix
constructions without any ray-geometric computations
\cite[]{ZhangBleistein:05,ZhangSunGray:07,XuZhangTang:11,tenKroode:12,HouSymes:15}. For
inversion of source parameters, even the usual ``exact'' inversion
by Fourier division is actually a parametrix, due to the inevitable
presence of time cut-off.

The existence of the parametrix permits the decomposition of the
gradient into two summands: the first does not involve the solution of
the inner (linear) problem at all, while the second does but is
manifestly continuous in the linear variables. Hence approximation of
the linear part of the solution by a convergent iteration results
in a convergent gradient approximation.

While convergence is assured by the three conditions above, properly
interpreted, control over the actual error in the gradient is
indirect, depending on quantities that are difficult or impossible ot
estimate accurately. However I also show that gradient error control
of the type established here is compatible with so-called inexact
trust-region optimization algorithms \cite[]{HeinkenVicente:01b}, with
(in principle) assured convergence to a stationary point. That is,
control over gradient error must be linked to convergence metrics of
the optimization algorithm.

Some uses of model extension for inverse wave problems have been
criticized for so-called ``gradient artifacts'', that is, computed
gradients that appear to give rise to slow convergence or undesired
model features \cite[]{FeiWilliamson:10,VyasTang:10}. However it now
appears that nothing was wrong with the gradients - instead, the
reduced objective functions of separable least squares problems were
defined with insufficiently precise solution of the linear (inner)
inverse problem \cite[]{KerSy:94,LiuSymesLi:14,ChaurisGP:14}. The present paper
could be viewed as an elaboration upon the latter work.

\section{Operators}
This is a Hilbert space story. 
In fact, the natural objects in this study are {\em Hilbert space scales}, that
is sequences of Hilbert spaces $H = \{ H^s, s \in \bZ\}$, decreasing in
the sense that
$H^{s+1} \subset H^s, s \in \bZ$. Define the sets
\[
H^{\infty} = \cap_{s \in \bZ} H^s, \,H^{-\infty} = \cup_{s \in
  \bZ}H^s.
\]
The inner 
product in $H^s$ will be denoted $\langle \cdot, \cdot \rangle_s$, and
the corresponding norm is
$\| \cdot \|_s$. Usually several scales will be involved in most of
our assertions; the scale to which the inner product or norm belongs
should be clear from context.

An {\em operator} $L$  {\em of order $k \in \bZ$} from
a scale $H_1$ to a scale $H_2$ is linear function from $H_1^{-\infty}$ to
$H_2^{-\infty}$ for which $L|_{H_1^s} \in \sB(H_1^s,H_2^{s-k})$ for
all $ s \in
\bZ$. Denote by $Op^k(H_1,H_2)$ a set of operators of order $k$,
satisfying the additional order properties: for $L_i \in
Op^{k_i}_1(H_1,H_2), i=1,2$,
\begin{equation}
\label{eqn:comm}
L_1 L_2 \in Op^{k_1+k_2}(H_1,H_2),\,[L_1,L_2] \in
Op^{k_1+k_2-1}(H_1,H_2).
\end{equation}
$Inv^k(H_1,H_2) \subset Op^k(H_1,H_2)$ is the subset of invertible
operators of order $k$.

The motivating example of this structure are scales of $L^2$ Sobolev
spaces, and scalar (pseudo-)\-differential operators on them.

The domain of a separable extended model consists of two
components. The first is a (background) model Hilbert space $M_b$ and
an open set of admissible models
$U \subset M_b$. In all examples, $M_b$ is a space of (possibly)
vector-valued) smooth ($C^{\infty}$, ``low frequency'') functions on a suitable physical
domain representing the physical parameters of a wave dynamics model, and members of $U$ obey additional constraints (such as
bounds) required to make the dynamical laws they parametrize
well-posed.

The other component of the model domain is a Hilbert space scale of extended
models $\oM =\{\oM_k, k\in \bZ\}$. 
A distinguished subscale $M = \{M_k, k\in \bZ\}, M_k \subset \oM_k$,
represents the non-extended or physical models. 
The  {\em annihilator} $A \in Op^0(\oM,N)$ has $M$ as its kernel:
\begin{equation}
\label{eqn:ann}
\om \in \oM,\,\,A\om=0 \Leftrightarrow \om \in M.
\end{equation}.
The {\em extension map} $E \in Op^0(M,\oM)$ is injective.

The model range or data space is another Hilbert scale $D$.

The extended modeling operator, or forward map, is a continuous function
\[
\oF: U \rightarrow Op^0(\oM,D)
\]
It is {\em smooth with loss of regularity}: 
\begin{equation}
\label{eqn:nonsmooth1} 
m_b \mapsto \oF[m_b]|_{\oM^k} \in C^p(U,\sB(\oM^k,D^{k-p})),
\end{equation}
but
\begin{equation}
\label{eqn:nonsmooth2} 
m_b \mapsto \oF[m_b]|_{\oM^k} \notin C^p(U,\sB(\oM^k,D^{k-p+1})),
\end{equation}
and in particular
\begin{equation}
\label{eqn:nonsmooth3} 
m_b \mapsto \oF[m_b]|_{\oM^0} \notin C^p(U,\sB(\oM^0,D^0)). 
\end{equation}
However, the normal operator is smooth 
\begin{equation}
\label{eqn:smoothnormal}
F^TF \in C^{\infty}(U,Op^0(\oM,\oM),
\end{equation}
as follows from the
{\em factorization lemma}:
\begin{equation}
\label{eqn:factor}
D\oF[m_b] \delta m_b = F[m_b](Q_1[m_b]\delta m_b)
\end{equation}
in which $Q_1 \in C^{\infty}(U\times M_b,Op^1(\oM,\oM))$ is {\em essentially
  skew-symmetric}:
\begin{equation}
\label{eqn:skew}
Q_1 + Q_1^T = Q_0 \in C^{\infty}(U\times M_b,Op^0(\oM,\oM)).
\end{equation}
Finally, assume that $\oF$ has the {\em Egorov property}: if $L_1 \in
C^{\infty}(U,Op^k(\oM,\oM))$, then there exists $L_2 \in
C^{\infty}(U,Op^k(D,D))$ so that for every $m_b \in U$,
\begin{equation}
\label{eqn:egorov}
\oF[m_b] L_1[m_b] = L_2[m_b] \oF[m_b].
\end{equation}
and vis-versa.

\noindent {\bf Remark:} In the examples, $\oF[m_b]$ is a Fourier
Integral Operator whose canonical relation is a local canonical graph,
and whose symbol and canonical relation depends smoothly on $m_b$.
Both the regularity with loss of a derivative
\cite[]{BlazekStolkSymes:13} and normal regularity of $\oF^T\oF$
follow from these facts: in this case, $\oF^T\oF$ is a
pseudodifferential operator. For examples of the factorization lemma,
see \cite{tenKroode:IPTA14,Symes:IPTA14}. The Egorov
property follows from the usual Egorov's theorem (see for example \cite{Tay:81}).

The physical modeling operator is a similar map
\[
F:U \rightarrow Op^0(M,D)
\]
with the same smoothness with loss of regularity.

The extended and physical modeling operators are related through the
extension map: for each $m_b \in U$,
\[
F[m_b] = \oF[m_b] \circ E.
\]

\section{Objective}
In the notation just introduced, the separable inverse problem of this
paper may be crudely stated as:
\begin{equation}
\label{eqn:basicip}
\mbox{Given }d\in D^0,\mbox{ find }m\in M_b,m_l \in M^0 \mbox{ so 
  that }\\
F[m]m_l \approx d 
\end{equation}
In view of the extension structure, this problem is equivalent to
\begin{equation}
\label{eqn:ip}
\mbox{Given }d\in D^0,\mbox{ find }m\in M_b,\om \in \oM^0 \mbox{ so 
  that }\\
\oF[m]\om \approx d, \,\,A\om \approx 0
\end{equation}
For consistent data (that is, for which ``$\approx$'' can be replaced
with ``$=$'' in either of the above statements, a solution is also a
global minimizer of both
\begin{equation}
\label{eqn:residuals}
\|\oF[m]\om - d\|^2 \mbox{ and } \|A\om\|^2
\end{equation}
The fundamental difficulty of this type of problem lies in the
mapping properties of $m \mapsto F[m] \mbox{ or } \oF[m]$: because of
these maps are differentiable only with loss of regularity (statements
\ref{eqn:nonsmooth1}, \ref{eqn:nonsmooth2}, and \ref{eqn:nonsmooth3}),
the first of the two functions in display \ref{eqn:residuals} (the
``data residual'') is not smooth as a map $M_b \times \oM \rightarrow
\bR$. This lack of smoothness is the ultimate reason that solution of the basic
``full waveform inversion'' problem \ref{eqn:basicip} is still a topic
of active research, rather than a routine tool in exploration and
academic seismology, more than thirty years after it was introduced.

Two objectives such as \ref{eqn:residuals} can be combined in various
ways to produce constrained and unconstrained optimization
problems. Both are quadratic in $\om$, so minimization of either in
this variable seems a tractable problem. Minimizing the second
(``semblance'') function so that the data residual becomes a function
of $m$ only (a so called reduced objective) does not improve matters:
the resulting unconstrained optimization is equivalent to
\ref{eqn:basicip} and has a non-smooth objective.

Remarkably, eliminating $\om$ by minimizing the data residual,  yields a smooth reduced semblance
objective: this conclusion will be justified below.
Generally it cannot
be assumed that $\oF$ is coercive, even if it is
injective. Therefore some form of regularization must be added to the
data residual to guarantee existence of a stable minimizer. It is also
turns out to be essential to
permit more freedom in the metric structure of
domain and range spaces also.

Introduce a data weighting operator $W_d\in C^{\infty}(U,Inv^0(D,D))$.
When restricted to $D^0$, $W_d$ is self-adjoint and positive definite,
hence induces a norm on the order $0$ data space:
\begin{equation}
\label{eqn:wdata}
\langle d_1,d_2 \rangle_d = \langle d_1, W_d d_2\rangle_0 \mbox{ for }
d_1,d_2 \in D^0 
\end{equation}
Similarly, introduce a model weight operator
$W_m\in C^{\infty}(U, Inv^0(\oM,\oM))$ on the model space, self-adjoint and positive definite when restricted to
$\oM^0$, with corresponding norm and inner product
\begin{equation}
\label{eqn:wmodel}
\langle \om_1,\om_2 \rangle_m = \langle \om_1, W_d \om_2\rangle_0 \mbox{ for }
\om_1,\om_2 \in \oM^0 
\end{equation}
Reformulate data residual minimization for the linear 
variables $\om$ as
\begin{equation}
\label{eqn:wredn}
\mbox{min}_{\om}\|\oF[m]\om-d\|^2_d + \lambda^2\|R\om\|_m^2 
\end{equation}
in which $R$ is a bounded and coercive regularization operator, whose
range is some other Hilbert space. The normal equation is
\begin{equation}
\label{eqn:normal}
N_{\lambda}[m]\om = \odF[m]d,\,\,N_{\lambda}[m]
=\odF[m]\oF[m]+\lambda^2 \dR R,
\end{equation}
in which $\odF=W_m^{-1}\oF[m]^TW_d[m]$ is the adjoint of $\oF$ with
respect to the weighted norms and $\oF^T$ is the adjoint with respect
to the scale 0-norms. Similarly $\dR$ is the adjoint of $R$ with respect to
the weighted model-space norm.

Denote by $\bre \in C^{\infty}(U \times D,\oM^0)$ the solution of
\ref{eqn:normal}, and define the reduced objective function 
\begin{equation}
\label{eqn:aredn}
J_{\lambda}[m,d]= \frac{1}{2}\|A\bre[m,d]\|_0^2
\end{equation}
If $J_{\lambda}[m,d]$ attains its obvious lower bound, then $\bre[m,d] \in \oM^0$ and
$(m,\bre[m,d])$ solves the extended inverse problem \ref{eqn:ip}.

\section{Gradient}

{\bf Calculation}
 $W = W_m$, $W_d=I$
\[
\langle u,v\rangle_m = \langle u, W[m]^{-1} v \rangle_0
\]
\[
W[m]S[m]^T S[m] = I + R[m], \, R \in Op^{-1}
\]
also 
\[ 
SWS^T \approx I
\]
\[
S[m]^{\dagger} = W[m]S[m]^T
\]
\[
J[m] = \frac{1}{2}\|AS[m]^{\dagger}d\|_m^2 =
\frac{1}{2}\|AW[m]S[m]^Td\|_m^2
\]
\[
= \frac{1}{2}\langle AW[m]S[m]^Td, W[m]^{-1}AW[m]S^T[m]d\langle_0
\]
\[
DJ[m] = \langle A D(W[m]S[m]^T)d, W[m]^{-1}AW[m]S^T[m]d \rangle_0
- \frac{1}{2}\langle AW[m]S[m]^Td,
W[m]^{-1}DW[m]W[m]^{-1}AW[m]S[m]^Td\rangle_0
\]
\[
=\langle A (DWW^{-1})(WS^T)d, W^{-1}AWS^Td
\rangle_0
+\langle A WDS^Td, W^{-1}AWS^Td \rangle_0
\]
\[
- \frac{1}{2}\langle AWS^Td,W^{-1}DWW^{-1}AWS^Td\rangle_0
\]
Definition of $Q$:
\[
Q \approx WS^TDS, DS = SQ, DS^T = Q^TS^T
\]
so 
\[
DJ[m] =
\langle A (DWW^{-1})(WS^T)d, W^{-1}AWS^Td
\rangle_0
+\langle A WQ^TS^Td, W^{-1}AWS^Td \rangle_0
\]
\[
- \frac{1}{2}\langle AW[m]S[m]^Td,
W[m]^{-1}DW[m]W[m]^{-1}AW[m]S[m]^Td\rangle_0 
\]
\[
=
\langle A (DWW^{-1})(WS^T)d, W^{-1}AWS^Td
\rangle_0
+\langle A WQ^TW^{-1}WS^Td, W^{-1}AWS^Td \rangle_0
\]
\[
- \frac{1}{2}\langle AWS[m]^Td,
W[m]^{-1}DW[m]W[m]^{-1}AW[m]S[m]^Td\rangle_0 
\]
Symmetry of $Q$:
\[
Q + WQ^TW^{-1} = WS^TDS + W(DS^TSW)W^{-1} = W(S^TDS + DS^TS)
\]
\[
= WD(S^TS) = D(WS^TS) - DW W^{-1}WS^TS \approx -DW W^{-1}
\]
\[
QW + WQ^T \approx -DW
\]
So
\[
\langle A WQ^TW^{-1}WS^Td, W^{-1}AWS^Td \rangle_0 =
\langle A WQ^TS^Td, W^{-1}AWS^Td \rangle_0
\]
\[
=-\langle A QWS^Td, W^{-1}AWS^Td \rangle_0 -\langle ADWW^{-1}(WS^Td),
W^{-1}AWS^Td\rangle_0 
\]
\[ 
=-\langle W^{-1}AWA QWS^Td, W^{-1}(WS^Td) \rangle_0 -\langle ADWW^{-1}(WS^Td),W^{-1}AWS^Td\rangle_0
\]
On the other hand,
\[
\langle A WQ^TW^{-1}WS^Td, W^{-1}AWS^Td \rangle_0 =
\langle Q^TS^Td, WA^TW^{-1}AWS^Td \rangle_0
\]
\[
=\langle WS^Td, W^{-1}Q WA^TW^{-1}AWS^Td \rangle_0
\]
Adding left and right hand sides of last two equations and dividing by
2, get
\[
\langle A WQ^TW^{-1}WS^Td, W^{-1}AWS^Td \rangle_0 
\]
\[
= \frac{1}{2}\left[\langle WS^Td, W^{-1}[Q,WA^TW^{-1}A]WS^Td\rangle_0
 - \langle ADWW^{-1}(WS^Td),W^{-1}AWS^Td\rangle_0\right]
\]
Combine with previous calc to get
\[
DJ =
\langle A (DWW^{-1})(WS^Td), W^{-1}AWS^Td \rangle_0
+\frac{1}{2}\left[\langle WS^Td, W^{-1}[Q,WA^TW^{-1}A]WS^Td\rangle_0
 - \langle A(DWW^{-1})(WS^Td),W^{-1}AWS^Td\rangle_0\right]
\]
\[
- \frac{1}{2}\langle W^{-1} AWS^Td,(DWW^{-1})A(WS^Td)\rangle_0 
\]
\[
=\frac{1}{2}\langle WS^Td, W^{-1}[Q,WA^TW^{-1}A]WS^Td\rangle_0
- \frac{1}{2}\langle W^{-1} AWS^Td,[(DWW^{-1}),A](WS^Td)\rangle_0 
\]
Since $[(DWW^{-1}),A] \in Op^{-1}$, the last term is negligible, and
we get
\[
DJ \approx \frac{1}{2}\langle WS^Td,
  W^{-1}[Q,WA^TW^{-1}A]WS^Td\rangle_0
\]
\[
= \frac{1}{2}\langle S^{\dagger}d, [Q,A^{\dagger}A]S^{\dagger}
d\rangle_m
\]

Minimization of $J_{\lambda}$ 
by any variant of Newton's method requires computation of the gradient  
of $J_{\lambda}$. Formally,
\[
DJ_{\lambda}[m,d]\delta m = \langle D\bre[m,d]\delta m,A^TA\bre[m,d]\rangle_0
\]
\[
= \langle -\Ne[m]^{-1}(D\Ne[m]\delta m) \bre[m,d]
+ \Ne[m]^{-1}(D\odF[m]\delta m)d ,A^TA\bre[m,d]\rangle_0
\]
Inserting $W_mW_m^{-1}$, referring to the definition of $\langle
\cdot,\cdot \rangle_m$, and noting that $W_m^{-1}A^T = A^{\dagger}$
and that $\Ne$ is self-adjoint in the model norm, obtain
\begin{equation}
\label{eqn:deriv}
= \langle -(D\Ne[m]\delta m)\bre[m,d] +(D\odF[m]\delta m)d, \bqe[m,d] \rangle_m
\end{equation}
where $\bqe[m,d]$ is the solution of
\begin{equation}
\label{eqn:q}
\Ne[m] \bqe[m,d] = A^{\dagger}A\bre[m].
\end{equation}

The first term in right-hand side of equation \ref{eqn:deriv}  defines
a continuous function of $\bre[m,d]$. 
Since $\Ne = W_m^{-1}\oF^TW_d\oF + \lambda^2 \dR R$, the rule \ref{eqn:egorov} implies
the existence of $P \in C^{\infty}(U,Inv^0(\oM,\oM))$ so that $\Ne =
W_m^{-1}\oF^T\oF P + \lambda^2 \dR R$. This shows in turn that $\Ne \in
C^{\infty}(U,Op^0(\oM,\oM))$, thanks to the assumptions on $W_d, W_m$,
\ref{eqn:egorov} once again, and the \ref{eqn:smoothnormal}. In
particular,
$D\Ne[m]\delta m \in Op^0(\oM,\oM)$, so the bilinear form 
\begin{equation}
\label{eqn:norcont}
\bq, \br \mapsto \langle (D\Ne[m]\delta m)\br,\bq \rangle_m
\end{equation}
is continuous in $\oM^0 \times \oM^0$. Also,
\[
\langle (D\Ne[m]\delta m)\br,\bq\rangle_m=
\langle (D(\odF[m])\delta m)\oF[m]\br,\bq \rangle_m
\]
\[
+\langle(\odF[m] (D\oF[m]\delta m)\br,\bq \rangle _m
= -\langle W_m[m]^{-1}(DW_m[m]\delta m)\odF[m]\oF[m]\br,\bq\rangle_m
\]
\[
+ \langle \odF[m] W_d[m]^{-1}DW_d[m]\oF[m]\br,\bq\rangle_m
+ \langle (D\oF[m]\delta m)^{\dagger}\oF[m]\br,\bq \rangle_m
\]
\[
+ \langle \odF[m](D\oF[m]\delta m)\br,\bq \rangle_m
\]
\[
= -\langle (DW_m[m]\delta m)\odF[m]\oF[m]\br,\bq\rangle_0
+ \langle (DW_d[m]\delta m)\oF[m]\br,\oF[m]q \rangle_0
\]
\begin{equation}
\label{eqn:firstterm}
+ \langle \oF[m]\br,(D\oF[m]\delta m) \bq \rangle_d
+ \langle (D\oF[m]\delta m)\br,\oF[m]\bq \rangle_d
\end{equation}

Concerning the second term in \ref{eqn:deriv}, note that
\[
(D\odF[m]\delta m) = D[(W_m[m])^{-1}\oF[m]^TW_d[m]d](\delta m)
\]
\begin{equation}
\label{eqn:1}
=-W_m[m]^{-1}(DW_m[m]\delta m)\odF[m]d + \odF[m]W_d[m]^{-1}DW_d[m]d 
+ (D\oF[m]\delta m)^{\dagger}d
\end{equation}
The first two terms in equation \ref{eqn:1} involve only bounded
operators on $\oM^0$. The contribution to the second
term in the right hand side of equation \ref{eqn:deriv} is
\[
\langle (D\odF[m]\delta m)d,\bq\rangle_m =
\]
\[
- \langle W_m[m]^{-1}(DW_m[m]\delta m)\odF[m]d, \bq\rangle_m + 
\langle \odF[m]W_d[m]^{-1}DW_d[m]d,\bq\rangle_m
\]
\[
+ \langle (D\oF[m]\delta m)^{\dagger}d,\bq\rangle_m
\]
\[
= -\langle (DW_m[m]\delta m)\delta m)\odF[m]d,\bq\rangle_0 +
+ \langle (DW_d[m]\delta m)d,\oF[m]\bq\rangle_0
\]
\begin{equation}
\label{eqn:secondterm} 
+ \langle d, (D\oF[m]\delta m)\bq \rangle_d
\end{equation}
Now the difficulty is plain to see: if $\bq=\bqe[m,d]$ in the last
term on the right hand side of equation \ref{eqn:secondterm} is replaced with an
approximation in the sense of the $0$ norm, as would be the result of
an iterative process for solving \ref{eqn:q} (or equation
\ref{eqn:normal}), no bound on the resulting error in the inner
product above, hence in the resulting approximation to $DJ_{\lambda}[m,d]\delta
m$ or the gradient of $J_{\lambda}$, can be asserted, since $D\oF[m]\delta m$ is not continuous in the sense of $\oM^0$.

Ignoring for the moment the apparent instability, combine equations
\ref{eqn:firstterm} and \ref{eqn:secondterm} to obtain an expression for the
derivative of $J_{\lambda}[m,d]$:
\[
DJ_{\lambda}[m,d]\delta m = \langle (DW_m[m]\delta m)
\odF[m](\oF[m]\bre[m,d]-d),\bqe[m,d]\rangle_0 
\]
\[
+ \langle (DW_d[m]\delta
m)(\oF[m]\bre[m,d]-d),\oF[m]\bqe[m,d]\rangle_0
\]
\begin{equation}
\label{deriv1}
- \langle F[m]\bre[m,d]-d,(D\oF[m]\delta m)\bqe[m,d]\rangle_d
- \langle (D\oF[m]\delta m)\bre[m,d],F[m]\bqe[m,d]\rangle_d
\end{equation}
There follows an expression for the gradient, using the partial
duals $D\oF[m]^*$, a continuous quadratic form $: D^0 \times \oM^1
\rightarrow M_b$ defined by
\[
\langle \delta m, D\oF[m]^*(d,\br)\rangle_{M_b} = \langle d,
(D\oF[m]\delta m)\br\rangle_d = \langle (D\oF[m]\delta
m)^{\dagger}d,\br\rangle_m
\]
\begin{equation}
\label{eqn:dual}
= \langle W_d[m]^{-1}d,(D\oF[m]\delta m)\br\rangle_0
\end{equation}
and $DW_m[m]^t$, a continuous bilinear map $\oM^0 \times \oM^0
\rightarrow M_b$ defined by 
\[
\langle \delta m, DW_m[m]^t(\br,\bq) \rangle_{M_b} = 
\langle (DW_m[m]\delta m)\br,\bq\rangle_0
\]
$DW_d[m]^t$ is defined similarly.
Then the $m$-gradient of $J_{\lambda}$ is given by
\[
\nabla_m J_{\lambda}[m,d] = DW_m[m]^t(\odF[m](\oF[m]\bre[m,d]-d),\bqe[m,d])
\]
\[
+ DW_d[m]^t(\oF[m]\bre[m,d]-d),\oF[m]\bqe[m,d]
\]
\[
- D\oF[m]^*(\oF[m]\bre[m,d],\bqe[m,d]) -
D\oF[m]^*(\bre[m,d],\oF[m]\bqe[m,d])
\]
\begin{equation}
\label{eqn:gradient}
+D\oF[m]^*(d,\bqe[m,d])
\end{equation}
Every part of this expression is manifestly {\em stable}, that is,  substitutions of
approximations $\brea,\bqea$ in their computation will entail 
$O(\|\bre - \brea\|_0, \|\bqe-\bqea\|_0)$ error, {\em except} for the
last term

\section{Stablility}
In this section, I will show that the last term in \ref{eqn:gradient} can 
expressed as the value at $\br=\bre[m,d]$ and $\bq=\bqe[m,d]$ of 
of continuous function of $d, \br,\bq$, thus giving a stable
computation of the gradient in the sense explained at the end of the
last section.

The construction depends on the availability of a {\em parametrix} or
approximate inverse modulo lower order error, of a particular
form. The utility of the construction depends on the computability of
the parametrix, that is, on its being a straightforward modification of the
transpose. \cite[]{HouSymes:15} demonstrated the existence of such special
parametrices for a particular separable inverse problem (extended
linearized constant density acoustic inversion with horizontal
subsurface offset extension) and parametrices with similar properties
exist for other extended modeling operators as well.

In fact, most parametrices of the type I've mentioned are only
microlocal, and moreover can only be computed approximately. These
limitations will be addressed in coming sections.

Assume that the data and model weight operators $W_d, W_m$ can be
chosen so that $\oF$ is approximately unitary with respect to the
norms $\|\cdot\|_d, \|\cdot\|_m$:
satisfies 
\begin{equation}
\label{eqn:parametrix}
\odF\oF - I \equiv S \in C^{\infty}( U, Op^{-1}(\oM,\oM)). 
\end{equation}
Thus $\Ne = S +
I +\lambda^2\dR R)$. Thus equations \ref{eqn:normal}, \ref{eqn:q} are equivalent to 
\begin{eqnarray}
\label{eqn:magic}
\bre &=& (I+\lambda^2\dR R)^{-1}(\odF d - S\bre)\\
\bqe &=&(I+\lambda^2\dR R)^{-1}(A^{\dagger}A\bre - S\bqe)\\
   & = &(I+\lambda^2\dR R)^{-1}(A^{\dagger}A (I+\lambda^2 \dR R)^{-1}(\odF d -
             S \bre)-(I+\lambda^2\dR R)^{-1}S\bqe
\end{eqnarray}
Replacing $\bqe$ in \ref{eqn:gradient} with the right hand side in the last equality of
\ref{eqn:magic}, obtain
\[
D\oF[m]^*(d,\bqe[m,d])=
D\oF[m]^*(d,(I+\lambda^2\dR R)^{-1}A^{\dagger}A (I+\lambda^2 \dR R)^{-1}\odF[m]d 
\]
\[
- (I+\lambda^2\dR R)^{-1}A^{\dagger}A (I+\lambda^2 \dR R)^{-1}
S[m]\bre[m,d] 
\]
\begin{equation}
\label{eqn:protogradient}
- (I+\lambda^2\dR R)^{-1}S[m]\bqe[m,d]).
\end{equation}
The operators appearing in the last two terms of
\ref{eqn:protogradient} are of
order $-1$, mapping $\oM^0$ continuously to $\oM^1$. So the continuity
property of $D\oF[m]^*$ noted above implies that the last two terms
above are stable, that is, substitution of approximations $\brea$ and
$\bqea$ for $\bre$ and $\bqe$ in the right-hand side of equation
\ref{eqn:protogradient} will result in an
$O(\|\brea-\bre\|_0,\|\bqea-\bqe\|_0)$ error.

It remains to be seen that the first term on the right hand side of
equation \ref{eqn:protogradient} is stable: {\em a priori}, it only
makes sense for $d\in D^1$ (hence $\bre \in \oM^1$). A continuous
extension to $d \in D^0$ follows however from the factorization
property \ref{eqn:factor}. For convenience, define 
\[
B = (1+\lambda^2\dR R)^{-1}A^{\dagger}A (1+\lambda^2 \dR R)^{-1}.
\]
Then the first term on the right-hand side of \ref{eqn:protogradient} is
\[
\langle \delta m, D\oF[m]^*(d,B\odF[m]d) \rangle_{M_b}
\]
\[
= \langle d, (D\oF[m]\delta m)B\odF[m]d \rangle_d 
\]
\[
= \langle d, \oF[m](Q_1[m]\delta m)B\odF[m]d\rangle_d
\]
\begin{equation}
\label{eqn:2}
=\langle \odF[m]d, (Q_1[m]\delta m)B\odF[m]d\rangle_m 
\end{equation}
The essentially-skew property of $Q$ (equation \ref{eqn:skew}) holds
also for the weighted inner product $\langle \cdot, \cdot \rangle_m$,
since the weight operator $W_m$ is invertible and of order 0:
specifically,
\begin{equation}
\label{eqn:skew-m}
(Q_1[m]\delta m)^{\dagger} = -(Q_1[m]\delta m) + (Q_{0,m}[m]\delta m),
\end{equation}
with $Q_{0,m} = Q_0 + W_m^{-1}[Q_1,W_m] \in C^{\infty}(U, Op^0(\oM,\oM))$.
 So the
right-hand side of equation \ref{eqn:2} is
\[
=\langle \odF[m]d, ([B,(Q_1[m]\delta m)] + B(Q_1[m]\delta m))
\odF[m]d\rangle_m 
\]
\begin{equation}
\label{eqn:3}
+\langle \odF[m]d, ([B,(Q_1[m]\delta m)] + ((Q_{0,m}[m]- Q_1[m])\delta m)B)\odF[m]d\rangle_m.
\end{equation}
Add the right hand sides of equations \ref{eqn:2} and \ref{eqn:3} and
divide by 2 to obtain
\[
\langle \delta m, D\oF[m]^*(d,B\odF[m]d) \rangle_{M_b}
\]
\begin{equation}
\label{eqn:4}
=\frac{1}{2}\langle \odF[m]d,([B,(Q_1[m]\delta m)] +
(Q_{0,m}[m]\delta m))\odF[m]d\rangle_m.
\end{equation}
The right-hand side of equation \ref{eqn:4} is a quadratic form in
$\odF[m]d$, defined by a self-adjoint operator (the sum of the various
commutators and products in the above expression) of order
zero. Therefore 
\[
D\oF[m]^*(d,B\odF[m]d)
\]
is a $\|\cdot\|_0$-continuous $M_b$-valued quadratic form in $d$,
hence extends continuously to $d \in D^0$, whence $\nabla_m J_{\lambda}[m,d]$ is
$\|\cdot\|_0$-continuous in $d$.

For Hilbert spaces $H_1, H_2$, denote by ${\cal Q}(H_1,H_2)$ the
Banach space of continuous $H_2$-valued quadratic forms on $H_1$.
Then the result of the foregoing calculations is summarized in

\begin{thm} Define $G_{\lambda} \in C^{\infty}(U, {\cal
  Q}(D^0 \times \oM^0 \times \oM^0, M_b)$ by
\begin{equation}
\label{eqn:grad_approx}
\begin{split}
G_{\lambda}[m](d,\br,\bq) = 
& D\oF[m]^*(d,(1+\lambda^2\dR R)^{-1}A^{\dagger}A (1+\lambda^2 \dR R)^{-1}\odF[m]d \\
& - (1+\lambda^2\dR R)^{-1}A^{\dagger}A (1+\lambda^2 \dR R)^{-1} S[m]\br\\
& - (1+\lambda^2\dR R)^{-1}S[m]\bq]) \\
& + DW_m[m]^t(\odF[m](\oF[m]\br-d),\bq)\\
&+ DW_d[m]^t(\oF[m]\br-d,\oF[m]\bq)\\
&- D\oF[m]^*(\oF[m]\br,\bq) \\
& -D\oF[m]^*(\br,\oF[m]\bq)
\end{split}
\end{equation}
Then
\begin{equation}
\label{eqn:thm1}
\nabla_m J_{\lambda}[m,d] = G_{\lambda}[m](d,\bre[m,d],\bqe[m,d]),
\end{equation}
in which $\bre[m,d]$ and $\bqe[m,d]$ are solutions of equations \ref{eqn:normal} and
\ref{eqn:q} respectively.
\end{thm}

\begin{proof} The content of the definition \ref{eqn:grad_approx} of
  $G_{\lambda}$ is that it is a continuous quadratic-form-valued
  function. This fact has been established for each of the summands:
  for the first term by equation \ref{eqn:4}, for the second and third
  by equation \ref{eqn:protogradient} and following discussion, the
  fourth and fifth by equation \ref{eqn:1}, and the last two by
  equations \ref{eqn:norcont} and \ref{eqn:firstterm} and surrounding
  discussion. These calculations also established equation
  \ref{eqn:thm1}.
\end{proof}

\begin{rem} For computational purposes, it is more convenient to work
directly with the defining Hilbert space structure of the background, domain and
range spaces, rather than with the background, model and data norms, and to
formulate the background model space norm also as a weighted norm. For example,
\[
\langle \delta m, D\oF[m]^*(d,A^{\dagger}A\odF[m]d) \rangle_{M_b}
\]
\[
= \langle d, (D\oF[m]\delta m)A^{\dagger}A\odF[m]d \rangle_d 
\]
\[
= \langle W_d d, (D\oF[m]\delta m)W_m^{-1}A^TA\odF[m]d \rangle_0.
\]
So
\begin{equation}
\label{eqn:thm1b}
\nabla_m J_{\lambda}[m,d] = W_b^{-1}D\oF[m]^t(W_d[m]d,W_m[m]^{-1}A^TA\odF[m]d) +
K[m](\bre,\bqe),
\end{equation}
where $W_b$ is the background model space weight operator defining the
norm $\|\cdot\|_{M_b}$, and $D\oF[m]^t$ is the transpose with respect
to the 0-norms (usually, $L^2$, or in the discrete case Euclidean
length) rather than the $M_b$ and $\|\cdot\|_d$ norms:
\begin{equation}
\label{eqn:dual0}
\langle \delta m, D\oF[m]^t(d,\br)\rangle_0 = \langle d,
(D\oF[m]\delta m)\br\rangle_0.
\end{equation}
Assuming that $\oF$ is implemented by a time-stepping finite
difference or finite element method, this notion of transpose can be
computed by a variant of the adjoint state method. For the case that
$\oF$ is a linearization (that is, Born approximation), so
that $D\oF$ is actually a second derivative, this
application of the adjoint state method was introduced by
\cite{SymSant:88}, and employed by \cite{KerSy:94} in computations
similar to those presented here.
\end{rem}

\section{Controllability}
The stability result of the last two sections is not in itself
sufficient foundation for a convergent optimization
algorithm, for two reasons: 
\begin{enumerate}
\item The solution errors $\brea-\bre$ and $\bqea-\bqe$ are not
directly observable, since neither $\bre$ nor $\bqe$ are known in
practice;
\item typical iterative solution algorithms for
equations \ref{eqn:normal} and \ref{eqn:q} do not necessarily reduce
their solution errors.
\end{enumerate}
Observable quantities actually reduced by iterative algorithms include the
data residual $\|\oF[m]\brea - d\|_d$, the residual for each of the
equations \ref{eqn:normal} and \ref{eqn:q}, and the norms of the
approximate solutions $\brea$ and $\bqea$.

It might be objected that the solution errors and residuals for
equations \ref{eqn:normal} and \ref{eqn:q} are related, the former
being at most $\lambda^{-2}$ times the latter ($\lambda^2$ being a
lower bound for the normal operator $\Ne$). Of course such bounds are
not uniform in $\lambda$, hence are not useful in formulation of an
effective algorithm.

As will be explained in the next section, the necessary form for the
gradient error estimate is
\begin{equation}
\label{eqn:graderr}
\|g - \nabla J_{\lambda}[m,d]\|_{M_b} \le K \epsilon
\end{equation}
in which $K \ge 0$ and $\epsilon$ is a parameter of the estimation process
that produces $\brea,\bqea$, in effect that $\brea,\bqea$
are functions of $m \in U, d\in D^0, e >0$. Only minimal assumptions
will be made about the estimation process - as will be seen below,
only requirements on the residuals produced in equations
\ref{eqn:normal} and \ref{eqn:q}.

However, such estimates appear to
require explicit use of scale separation, abstracted in part by introducing a
{\em family} $d=\{\de: \lambda > 0\} \subset D^0$ of data, and
corresponding separable least squares problems. The {\em high
  frequency cone condition}
\begin{equation}
\label{eqn:hfc}
\|\de\|_k \le C_k\lambda^2 \|\de\|_{k+1},\, k \ge k_0
\end{equation}
gives a quantitative expression of scale separation. The lower index
bound $k_0$ is presumed to be at most $-1$.

\begin{rem} For the Sobolev scale, this condition identifies the
  regularization parameter $\lambda^2$ with wavelength.
\end{rem}

From
here on, $\bre[m,d]$ and $\bqe[m,d]$ denote the solutions of the
parametrized systems
\begin{equation}
\label{eqn:normalbis}
N_{\lambda}[m]\bre[m,d] = \odF[m]\de  
\end{equation} 
and
\begin{equation}
\label{eqn:qbis}
N_{\lambda}[m]\bqe[m,d] = A^{\dagger}A\bre[m,d]. 
\end{equation}

Linking the size of the residuals
\begin{eqnarray}
\label{eqn:mres}
e_{m,\lambda}[m,d] & = & \Ne[m]\brea[m,d]-\odF[m]\de,\\
\label{eqn:qres}
e_{q,\lambda}[m,d] & = & \Ne[m]\bqea[m,d]-A^{\dagger}A\brea[m,d,e] 
\end{eqnarray}
to the solution errors $\brea-\bre$ and $\bqea-\bqe$ requires some
assumption about the approximate solutions $\brea$ and $\bqea$. Note
that any Krylov method for solution of the systems
$\ref{eqn:normalbis}$ and $\ref{eqn:qbis}$, preconditioned by $(I+\lambda^2\dR
R)^{-1}$, yields first iterates
\begin{eqnarray}
\label{eqn:krylov1}
\breone[m,d]& =& (I+\lambda^2\dR R)^{-1}\odF[m]\de\\
\bqeone[m,d]&=&(I+\lambda^2 \dR R)^{-1}A^{\dagger}A\breone[m,d]
\end{eqnarray}
for which the residuals are
\begin{eqnarray}
\label{eqn:mres1}
e_{m,\lambda,1} & = & S[m](I+\lambda^2\dR R)^{-1}\odF[m]\de,\\
e_{q,\lambda,1} & = & S[m](I+\lambda^2\dR R)^{-1}A^{\dagger}A\breone[m,d],
\end{eqnarray}

\begin{thm} Suppose that for $\epsilon > 0$, $\brea[m,d,\epsilon]$ and
  $\bqea[m,d,\epsilon]$ are approximate solutions of \ref{eqn:normalbis} and
  \ref{eqn:qbis} for which the residuals $e_{m,\lambda,\epsilon}$ and
  $e_{q,\lambda,\epsilon}$ satisfy
\begin{eqnarray}
\label{eqn:mresn}
\|e_{m,\lambda,\epsilon}\|_m & \le & \epsilon \|e_{m,\lambda,1}\|_m\\
\|e_{q,\lambda,\epsilon}\|_m & \le & \epsilon \|e_{q,\lambda,1}\|_m 
\end{eqnarray}
Then there exists $K>0$ and $\lambda_0 \ge 0$ so that for $\lambda \le
\lambda_0$,
\begin{eqnarray}
\label{eqn:postbd1}
\|\brea[m,d,\epsilon] -\bre[m,d]\|_m & \le & K \epsilon\|\de\|_d\\
\label{eqn:postbd2}
\|\bqea[m,d,\epsilon] -\bqe[m,d]\|_m & \le & K \epsilon \|\de\|_d
\end{eqnarray}
uniformly in $m \in U$, for fixed choice of $\{C_k\}$ in
\ref{eqn:hfc}.
\end{thm}

\begin{proof}  Since 
\[
\Ne[m] (\brea[m,d,\epsilon] -\bre[m,d]) = e_{m,\lambda,\epsilon},
\]
\[
\|\brea[m,d,\epsilon] -\bre[m,d]\|_0 \le \frac{K}{\lambda^2}\| e_{m,\lambda,\epsilon}\|_0
\le \frac{K\epsilon}{\lambda^2}\| e_{m,\lambda,1}\|_0
\]
\[
\le \frac{K\epsilon}{\lambda^2}\|\de\|_{-1} \le K \epsilon \|\de\|_d
\]
The next-to-the last inequality follows from the definition of
$e_{m,\lambda,1}$ and the uniform $Op^{-1}$ bound on $S[m]$, the last
from \ref{eqn:hfc} for $k=-1$. Thus \ref{eqn:postbd1} is established. Since $e_{q,\lambda,1}$
takes the same form, a similar estimate holds for it.
\end{proof}
\begin{cor}
Under the assumptions of Theorem 2, there exists $K \ge 0$ for which
\begin{eqnarray}
\label{eqn:graderr1}
\left|\frac{1}{2}\|A\brea[m,d,\epsilon]\|^2 - J_{\lambda}[m,d]\right |
  \le K \epsilon\\
\|G_{\lambda}[m](d,\brea[m,d,\epsilon],\bqea[m,d,\epsilon]) - \nabla J_{\lambda}[m,d]\|_{M_b} \le K \epsilon
\end{eqnarray}
 for $m \in U$ and sufficiently small $\lambda$.
\end{cor}

\begin{proof}
Follows from continuity of $G_{\lambda}$ as described in Theorem 1,
and bounds on the solution errors established in Theorem 2. 
\end{proof}

\section{Optimization}

In view of the error inherent in the 
calculations explained in the previous sections, standard variants of
Newton's method for local minimization cannot be guaranteed to
converge to a local minimizer of $J_{\lambda}$. Minimization of the
reduced objective requires use of iterations that converge in the
presence of inexact gradient and objective function evaluations. Such
an algorithm would necessarily need to couple error and step size control.

Since error-contaminated function and gradient computations are hardly
rare, it is unsurprising that this topic has a fairly large
literature
\cite[]{Dembo:82,Carter:91,Deuflhard:91,Carter:93,EisenstatWalker:94}. All
of these works and many more recent ones share a major drawback: they
mandate {\em absolute} control of function and/or gradient error,
implying in particular that the computed gradient be a descent
direction. While of course this condition must eventually hold, it is
nearly impossible to check in practice: many sources of error are like
those considered in the preceding sections, in that they give no
direct measure of gradient or function error, but only allow indirect
control.

One exception to this pattern is the work of \cite{HeinkenVicente:01b,DPKouri_MHeinkenschloss_DRidzal_BGvanBloemenWaanders_2013a,DPKouri_MHeinkenschloss_DRidzal_BGvanBloemenWaanders_2014a}
on error control in the context of {\em trust region} methods for
sequential qudratic programming, an approach to constrained
optimization. Trust region globalization is a general concept that
applies also to unconstrained formulations
\cite[]{ConnGouldToint:00,NocedalWright}. The condition that
\cite{HeinkenVicente:01b} place on gradient error takes the form:
error $\le K \epsilon$, where $\epsilon$ is a control parameter but
$K$ is unknown or at least poorly controlled. Thus nothing can be said
{\em a priori} about the size of the gradient error, or even whether
the computed gradient is a descent direction, except in an asymptotic
sense. Reference to Corollary 1 shows that the theory developed above
provides exactly this sort of condition.

The trust region approach to minimization of a $C^2$ function $f$
defined on a Hilbert space $H$ bases its $k$th step on a quadratic model
approximating $f$ near $x_k$.
\[
m_k(s) = f_k + \langle s,g_k \rangle + \frac{1}{2}\langle s,H_k s
\rangle.
\]
Here $f_k$ is the computed value and $g_k$ the computed gradient at
$x_k$. $H_k$ is an approximation, possibly quite crude, to the Hessian
of $f$ at $x_k$. The basic trust region algorithm seeks the step $s=x_{k+1}-x_k$
as the optimum of the constrained problem
\begin{equation}
\label{eqn:trustregion}
\mbox{minimize }m_k(s) \mbox{ subject to } \|s\| \le \Delta_k,
\end{equation}
The {\em trust radius} $\Delta_k$ is also subject to update as the iteration
proceeds, so as to satisfy a {\em sufficient decrease} criterion
leading to assured global convergence to a local minimizer, under
various conditions (smoothness, Hessian definiteness, etc.) some of
which will be mentioned below.

I will describe a simple variant of the trust radius step, depending
on two computed quantities: actual reduction,
\[
actred = f(x_k) - f(x_k +s)
\]
and predicted reduction, 
\[
predred = m_k(0) - m_k(s) = -(\langle s,g_k \rangle +
\frac{1}{2}\langle s, H_k s \rangle)
\],
and on four magic numbers, $0 < \eta_1 <\eta_2 < 1$ and 
$0 < \gamma_1 <1 < \gamma_2 $.

Choose a {\em search direction} $p$ by
minimizing the model as an unconstrained problem: that is,
\[
p = -H_k^{-1}g_k.
\]
Note that $H_k$ is only an approximation to the Hessian of $f$, and
that in a rather loose sense: it may be computed by building up a
Krylov or quasi-Newton approximation to the Hessian or its
Gauss-Newton simplification, for example. Thus $p$ may be only a crude
approximation to a Newton or Gauss-Newton step: this algorithm
description encompasses so-called truncated Newton-Krylov methods, for
example, and the step selection may interact with the computation of
$H_k$.

Next enter the step update loop:
\begin{enumerate}
\item set $s=\Delta_k p$
\item compute $actred, predred$.
\begin{enumerate}

\item if $actred \le \eta_1 predred$, $\Delta_k \leftarrow \gamma_1 
  \Delta_k$, go to 1.
\item if $actred \ge \eta_2 predred$, $\Delta_k \leftarrow \min(1,\gamma_2 
  \Delta_k)$, 
\end{enumerate}
\item $x_{k+1}=x_k + s$, $\Delta_{k+1} = \Delta_k$, exit.
\end{enumerate}

To see how this algorithm accommodates function and gradient error,
first presume that there is none: that is, $f_k=f(x_k), g_k = \nabla
f(x_k)$, and $H_k \ge \beta > 0$ for all $k$.  The theory presumes that the
function $f$ is twice continuously differentiable, bounded below, and
has uniform bounds on the gradient and condition number of the Hessian
over the sub-level sets of the objective. It is easy to see that the
condition for a successful step implies that
\begin{equation}
\label{eqn:trstep}
f_k - f_{k+1} \ge C \|g_k\|\Delta_k
\end{equation}
references, these assumptions imply a natural floor
under the trust radius $\Delta_k$. The trust radius decreases only when the actual
function 
reduction is not at least the lower proportion $\eta_1$ of the
predicted reduction. However as the trust radius, hence the step, gets
smaller, the actual reduction and the predicted reduction become
close, since the model becomes close to the function to first
order. Once the trust radius is small enough, the conditions for
further decrease are never met, and the trust radius never decreases
beyond a fixed positive thresshold. Since the function values are
decreasing from step to step and bounded below, they must
converge. Therefore the actual reduction converges to zero, hence the
gradient must converge to zero thanks to inequality
\ref{eqn:trstep}. Note that this reasoning does not establish that
$\{x_k\}$ converges - that requires a bit more reasoning. However, if
the sequence of iterates does converge, it must converge to a
stationary point.

In fact the expected behaviour of this algorithm is that it eventually
settles down at $\Delta_k = 1$, takes the full step, and converges at
whatever rate the underlying approximation to Newton's method yields.

Convergence with inexact function and gradient evaluations follows
from an extension of the same reasoning. Assume for the moment that
the function values are exact. Then the primary condition to be imposed on the
gradient, according to \cite{HeinkenVicente:01b}, is of the form
\begin{equation}
\label{eqn:hv}
\|g_k - \nabla f(x_k)\| \le \xi \min(\|g_k\|,\Delta_k).
\end{equation}
in which it is merely required that the constant $\xi \ge 0$ be
uniform over the sublevel set of $f$ in which the iterates lie. Then
once again reduction of $\Delta_k$ during a step eventually stops
short of a uniform positive lower bound, since the inequality
\ref{eqn:hv} forces $g_k$, hence the step $s$, towards the steepest
descent step. For short enough steepest descent steps the
actual reduction exceeds $\eta_1$ times the predicted reduction, which
leads both to the step being taken and in the computed gradient $g_k$
being reduced in length.

I embed this mechanism in an algorithm for minimization of
$J_{\lambda}$ following the pattern of \cite{HeinkenVicente:01b},
section 5. Add $\xi >0$ to the list of magic numbers required by the
trust region algorithm; in theory at least $\xi=1$ is adequate.

Define $(f,g,H) = JET(m,\epsilon)$ via the algorithm

  \begin{enumerate}
    \item compute initial approximate solutions $\br_1,\bq_1$
      \begin{eqnarray}
      \label{eqn:krylov1alg}
      \br_1& =& (I+\lambda^2\dR R)^{-1}\odF[m]\de\\
      \bq_1&=&(I+\lambda^2 \dR R)^{-1}A^{\dagger}A\br_1
      \end{eqnarray}
       with residuals $e_{m,1}, e_{q,1}$;
    \item compute approximate solutions $\br,\bq$ of the system
      \ref{eqn:normalbis},\ref{eqn:qbis} satisfying with residuals
      $e=(e_{m}, e_{q})$ satisfying
      \begin{eqnarray}
      \label{eqn:mresnalg}
      \|e_{m}\|_m & \le & \epsilon \|e_{m,1}\|_m\\
      \|e_{q}\|_m & \le & \epsilon \|e_{q,1}\|_m 
      \end{eqnarray}
    \item compute approximate function value, gradient
      \begin{eqnarray}
      \label{eqn:fgalg}
      f& =& \frac{1}{2}\|A\br\|^2 \\
      g& = & G_{\lambda}[m](d,\br,\bq)
      \end{eqnarray}
      and approximate Hessian $H$
  \end{enumerate}

Formulation of inexact Trust Region algorithm:

\begin{enumerate}
  \item $(f_k,g_k,H_k) =JET(m_k,\epsilon_k)$
  \item while $\epsilon_k > \xi \min(\|g_k\|_m,\Delta_k)$ 
  \begin{enumerate}
    \item $\epsilon_k \leftarrow \xi \min(\|g_k\|_m,\Delta_k)$
    \item $(f_k,g_k,H_k) =JET(m_k,\epsilon_k)$
  \end{enumerate}
  \item $s = -\Delta_k H_k^{-1}g_k, m_{k+1}=m_k+s, \epsilon_{k+1} = \epsilon_k,
    (f_{k+1},g_{k+1},H_{k+1}) = JET(m_{k+1},\epsilon_{k+1})$
    \begin{eqnarray}
    \label{eqn:actpred}
    actred &=&f_k-f_{k+1}\\
    predred &=& \left(1-\frac{\Delta_k}{2}\right)
                \Delta_k\langle g_k,H_k^{-1}g_k\rangle_m
    \end{eqnarray}
  \item if $actred \le \eta_1 predred$, $\Delta_k \leftarrow \gamma_1 
      \Delta_k$, go to 3.
  \item if $actred \ge \eta_2 predred$, $\Delta_k \leftarrow \min(1,\gamma_2 
      \Delta_k)$, 
  \item next k
\end{enumerate}

\begin{cor} The sequence $\{m_k\}$ produced by the
  inexact Trust Region algorithm just described converges to a
  stationary point of $J_{\lambda}$.
\end{cor}

\section{Microlocalization}

With a few exceptions, no examples of forward maps $\oF$ defined by
separable inverse wave problems have actual parametrices, that is,
operators $\odF$ satisfying the definition
\ref{eqn:parametrix}. Instead, they have {\em microlocal}
parametrices. Abstractly, the microlocal property is captured in an
approximate projector $\Pi \in Op^0(\oM,\oM)$, self-adjoint with
respect to $\|\cdot\|_m$ and close to idempotent in
the sense that 
\begin{equation}
\label{eqn:approxidem}
\|\Pi - \Pi^2\|_m \le \frac{1}{4},\,\,0 \le \Pi \le 1.
\end{equation}
A microlocal parametrix $\odF$ satisfies
\begin{equation}
\label{eqn:micropar}
\odF[m]\oF[m] - \Pi = S[m] \in Op^{-1}(\oM,\oM).
\end{equation}
In this section, the approximate projector $\Pi$ is locally constant,
that is, the relation \ref{eqn:micropar} holds in a neighborhood of
$m_0 \in M$.

Augment the linear least squares problem \ref{eqn:wredn} by adding
$\langle \om, (I-\Pi)\om\rangle_m$ to objective. The corresponding
effect on the normal operator is:
\[
\Ne[m] = \odF[m]\oF[m] + (I-\Pi) + \lambda^2 \dR R = I + S[m] +
\lambda^2\dR R.
\]
Define as before $\bre[m,d]$ to be the solution of $\Ne[m]
\bre[m,d] = \odF[m]d$, and the reduced objective $J_{\lambda}$ by
equation \ref{eqn:aredn}.

Computation of $J_{\lambda}$ and $\nabla J_{\lambda}$ goes exactly as
before, and Theorems 1 and 2 and Corollary 1 hold verbatim. The only
change comes in the relation between the minimizer of $J_{\lambda}$
and the solution of the inverse problem $m,\om$: the data misfit is
small only if one assumes that $d$ can be fit with $\om$ nearly
annihilated by $I-\Pi$. That is an {\em a priori} assumption about the
solution, which in examples entails a similar assumption about data. 
% explore meaning in an example

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/iwavecal/paper.tex
repro docs dir added

\title{Calibrating IWAVE}
\date{}
\author{William W. Symes, The Rice Inversion Project}

\lefthead{Symes}
\righthead{Calibrating IWAVE}

\maketitle
\parskip 12pt
\begin{abstract}
IWAVE simulation of wavefields involves many individual computations combined into a single command. While each computation can be tested for accuracy on its own, it is also important to ensure that the final result accurately approximates the solution of the target continuum problem by direct comparison with solutions of known accuracy. This paper presents comparison of acoustic simulations on staggered and non-staggered grids with numerical implementations of analytic solutions in homogeneous media. The comparisons suggest that IWAVE correctly solves these problems with reasonable accuracy.
\end{abstract}

\section{Theory}
Linear acoustics describes small amplitude motions of a compressible fluid in terms of the evolution of (excess) pressure $p(\bx,t)$ and particle velocity $\bv(\bx,t)$: 
\begin{eqnarray}
\label{eqn:aws}
\frac{\partial p}{\partial t}(\bx,t) & = &\kappa(\bx) (-\nabla \cdot \bv(\bx,t) + f(\bx,t))\nonumber\\
\frac{\partial \bv}{\partial t}(\bx,t) & = & -\beta(\bx) \nabla p(\bx,t)\nonumber\\
p, \bv &=& 0,\, t << 0.
\end{eqnarray}
(\cite[]{gur81}, section 19, or \cite[]{Frie:58}, Chapter 1). The linear acoustic constitutive law makes relative change of volume and pressure proportional by the the bulk modulus $\kappa$, which therefore has units of pressure = force/area. The first of the equations in \ref{eqn:aws} therefore represents input of energy into the system via a defect in the acoustic consitutive relation: the forcing (or source) term $f$ has units of 1/time, that is, relative rate of change. The second equation in \ref{eqn:aws} expresses Newton's law relating acceleration and force for the acoustic model in which stress is scalar and its divergence (by definition) equal to the negative pressure gradient. The buoyancy $\beta$ is the reciprocal of density, $\beta=1/\rho$. The final condition in \ref{eqn:aws} specifies causal solutions.

The system \ref{eqn:aws} is equivalent to a single second order PDE for pressure: differentiate the first equation with respect to time, then use the second equation to eliminate $\bv$ and obtain
\begin{eqnarray}
\label{eqn:awe}
\frac{\partial^2 p}{\partial t^2}(\bx,t) & = &\kappa(\bx)\left(\nabla \cdot \beta(x) \nabla p(\bx,t) + \frac{\partial f}{\partial t}(\bx,t)\right) \nonumber\\
p & = & 0, \,t<<0
\end{eqnarray}

The systems \ref{eqn:aws} and \ref{eqn:awe} describe linear acoustodynamics in any dimension, in particular in 2D or 3D, and far from physical boundaries, or in Euclidean space without boundaries. Appropriate boundary conditions must be supplied to complete the description of acoustics in  domains with boundaries.

In this section we will develop analytic expressions for the pressure field $p$ in \ref{eqn:aws} or \ref{eqn:awe} for 2D and 3D, assuming that $\kappa$ and $\beta$ are constant and $f$ has the form of an {\em isotropic point radiator} located at the origin:
\begin{equation}
\label{eqn:ipr}
f(\bx,t) = w(t) \delta(\bx).
\end{equation}
The source pulse, or wavelet, $w$ is smooth and of compact support.
Since $f$ has units of rate (1/time), $w$ must have units of volume rate.

Note that in this case \ref{eqn:awe} simplifies to 
\begin{eqnarray}
\label{eqn:aweccipr}
\frac{1}{c^2}\frac{\partial^2 p}{\partial t^2}(\bx,t) & = &\nabla^2 p(\bx,t) + \rho\frac{dw}{dt}(t)\delta(\bx) \nonumber\\
p & = & 0, \,t<<0 
\end{eqnarray}
using the wave velocity $c=\sqrt{\kappa\beta}$. Note that the forcing term has units of density rate, or force/distance. The same for would be obtained starting with a version of \ref{eqn:aws} with a body force, that is, inhomogenous term in Newton's law, in which case the source would be the divergence of a force.

\subsection{3D}
The solution of the point radiator problem is well-known, see for example \cite{CouHil:62}, Chapter 6 or \cite{GuiSte:79}, Chapter 1. I repeat it here for completeness. Since the isotropic point radiator \ref{eqn:ipr} is rotationally symmetric, it is natural to expect the same of the pressure field, i.e. that $p(\bx,t) = P(r,t)$ with $r=|\bx|$. The simple identity
\begin{equation}
\label{eqn:calc}
\nabla^2 \frac{F(r)}{r} = \left(\frac{\partial^2}{\partial r^2} + \frac{2}{r}\frac{\partial}{\partial r}\right) \frac{F(r)}{r} = \frac{1}{r}\frac{\partial^2 F}{\partial r}^2(r)
\end{equation}
shows that $rP(r,t)$ obeys the homogeneous 1D wave equation for $r>0$. To achieve causality, necessarily $rP(r,t)$ is a function of $t-r/c$. Set
\[
P(r,t) = \frac{g(t-r/c)}{r}
\]
and assume $g$ to be smooth and of compact support. Then $P$ has an integrable singularity at $r=0$, and is a distribution solution of \ref{eqn:aweccipr} exactly if
\[
g(t) = \frac{\rho\frac{dw}{dt}(t)}{4\pi}
\]
as one sees by writing the equation in weak form, expressing the integration against at test function image under the wave operator as the limit of integrals over the exterior of the ball of radius $\epsilon >0$, integrating by parts, and taking the limit $\epsilon \rightarrow 0$. Therefore we obtain for $p=p_{3D}$ the analytic expression
\begin{equation}
\label{eqn:awsol3D}
p_{3D}(\bx,t) = \frac{\rho\frac{dw}{dt}(t-r/c)}{4\pi r}
\end{equation}

As stated in the introduction, the IWAVE discretizations of systems \ref{eqn:aws} and \ref{eqn:awe} normalize the source as a black-box right-hand side. The IWAVE acoustic constant density (ACD) solver (code in {\tt iwave/acd}) approximates the solution of 
\begin{eqnarray}
\label{eqn:acd}
\frac{\partial^2 p_{ACD}}{\partial t^2}(\bx,t) & = &c^2\nabla^2 p_{ACD}(\bx,t) + f_{ACD}(\bx,t)\nonumber\\
p_{ACD} & = & 0, \,t<<0 
\end{eqnarray}
The point source version of this problem has
\begin{equation}
\label{eqn:acdipr}
f_{ACD} = w_{ACD}(t)\delta(\bx)
\end{equation}
Comparing equations \ref{eqn:aweccipr}, \ref{eqn:acd} and \ref{eqn:acd}, one sees that $p_{ACD} = p$ if 
\[
w_{ACD}(t) = \rho c^2 \frac{dw}{dt}(t) 
\]
so for 3D one obtains the solution of \ref{eqn:acd}, \ref{eqn:acdipr}:
\begin{equation}
\label{eqn:acdsol3D}
p_{ACD3D}(\bx,t) = \frac{ w_{ACD}(t-r/c)}{4\pi c^2 r}.
\end{equation}

The IWAVE acoustic staggered grid (ASG) solver (code in {\tt iwave/asg}) approximates the solution of
\begin{eqnarray}
\label{eqn:asg}
\frac{\partial p_{ASG}}{\partial t}(\bx,t) & = &-\kappa(\bx) \nabla \cdot \bv_{ASG}(\bx,t) + f_{ASG}(\bx,t)\nonumber\\
\frac{\partial \bv_{ASG}}{\partial t}(\bx,t) & = & -\beta(\bx) \nabla p_{ASG}(\bx,t)\nonumber\\
p_{ASG}, \bv_{ASG} &=& 0,\, t << 0.
\end{eqnarray}
This system differs from \ref{eqn:aws} only by having the right-hand side vector scaled by $1/\kappa$ - that is, $p_{ASG} = p$ if $f_{ASG} = \kappa f$. Thus for the 3D point source problem, that is, the solution of \ref{eqn:asg} with 
\[
f_{ASG}(\bx,t) = w_{ASG}(t) \delta(\bx),
\]
the source pulse $w_{ASG}=\kappa w$ has units of pressure $\cdot$ volume / time, and equation \ref{eqn:awsol3D} implies
\[
p_{ASG3D}(\bx,t) = \frac{\rho\frac{dw_{ASG}}{dt}(t-r/c)}{4\pi \kappa r}
\]
\begin{equation}
\label{eqn:asgsol3d}
= \frac{\frac{dw_{ASG}}{dt}(t-r/c)}{4\pi c^2 r}.
\end{equation}

An interesting consequence of that for any space dimension, isotropic point source wavelets used in \ref{eqn:asg} and \ref{eqn:acd} will produce the same pressure field if 
\begin{equation}
\label{eqn:acdasg}
w_{ACD} = \frac{\partial w_{ASG}}{\partial t}.
\end{equation}
Equation \ref{eqn:acdasg} is consistent with $w_{ASG}$ having units of force $\cdot$ velocity, and $w_{ACD}$ force $\cdot$ acceleration, as follows from equations \ref{eqn:asg} and \ref{eqn:acd} respectively.

\subsection{2D}
I will regard 2D solutions of equations \ref{eqn:aws} and \ref{eqn:awe} as line source solutions of the 3D equations. In this subsection, I use the notation $\bx=(x,z)$ for the 2D position vector in the ``inline'' plane perpindicular to the $y$ axis, and $\bx_{3D} = (x,y,z)$. The isotropic line radiator appropriate for the ASG version of the acoustic system \ref{eqn:asg} is 
\begin{equation}
\label{eqn:ilr}
f(\bx_{3D},t) = w_{ASG2D}(t)\delta(\bx).
\end{equation}
For constant $\kappa$ and $\beta$, the weak solution $(p_{ASG2D}, \bv_{ASG2D})$ of system \ref{eqn:asg} is also independent of $y$. Since $\delta(\bx)$ has units of 1/area, $w_{ASG2D}$ must have units of pressure $\cdot$ area / time.

To find an analytic form for the pressure field of an isotropic line radiator, use D'Alembert's method of descent (\cite[]{CouHil:62}, Ch. 6). This approach begins with the observation that
\[
\delta(\bx) = \int dy \delta(\bx_{3D})
\]
(with the right-hand side interpreted as a pull-back). Therefore the solution of system \ref{eqn:asg} with $f$ given by equation \ref{eqn:ilr} is  
\begin{equation}
\label{eqn:descent}
p_{ASG2D}(\bx,t) = \int dy \tilde{p}_{ASG3D}(\bx_{3D},t).
\end{equation}
where $\tilde{p}_{3D}$ is the solution of the isotropic point radiator problem given by equation \ref{eqn:asgsol3d}, but with $w_{ASG}$ replaced by $w_{ASG2D}$, so that $\tilde{p}_{ASG3D}$ has units of pressure/length and $p_{ASG2D}$ is a pressure.

From equation \ref{eqn:asgsol3d},
\[
p_{ASG2D}(\bx,t) = \int dy \tilde{p}_{ASG3D}(\bx_{3D},t).
\]
\[
= \int dy \frac{\frac{dw_{ASG2D}}{dt}(t-r/c)}{4\pi c^2 r}
\]
\[
= \int_{-\infty}^{\infty} dy \frac{\frac{dw_{ASG2D}}{dt}(t-\sqrt{|\bx|^2 + y^2}/c)}{4\pi c^2 \sqrt{|\bx|^2 + y^2}}
\]
\[
= 2 \int_0^{\infty} dy \frac{\frac{dw_{ASG2D}}{dt}(t-\sqrt{|\bx|^2 + y^2}/c)}{4\pi c^2 \sqrt{|\bx|^2 + y^2}}
\]
\begin{equation}
\label{eqn:green2dpart1}
= 2 \lim_{\epsilon \rightarrow 0}\int_{\epsilon}^{\infty} dy \frac{\frac{dw_{ASG2D}}{dt}(t-\sqrt{|\bx|^2 + y^2}/c)}{4\pi c^2\sqrt{|\bx|^2 + y^2}}
\end{equation}
since the integrand is continuous, of compact support, and even in $y$. Set $\tau = \sqrt{|\bx|^2 + y^2}/c$. Then $y \rightarrow \tau$ is a nonsingular diffeomorphism $[\epsilon,\infty) \rightarrow [\sqrt{|\bx|^2+\epsilon^2}/c,\infty)$. Changing variables in the \ref{eqn:green2dpart1} gives
\[
= 2 \lim_{\epsilon \rightarrow 0} \int_{\sqrt{|\bx|^2+\epsilon^2}/c}^{\infty} \frac{c^2 \tau d\tau}{\sqrt{c^2 \tau - |\bx|^2}} \frac{\frac{dw_{ASG2D}}{dt}(t-\tau)}{4\pi c^3 \tau} 
\]
\[
=\lim_{\epsilon \rightarrow 0} \frac{1}{2\pi c}\int_{\sqrt{|\bx|^2+\epsilon^2}/c}^{\infty} d\tau \frac{\frac{dw_{ASG2D}}{dt}(t-\tau)}{\sqrt{c^2 \tau^2 -|\bx|^2}}
\]
so taking the limit $\epsilon \rightarrow 0$ and introducing the Heavyside function to express the integral as a convolution, obtain
\begin{equation}
\label{eqn:asgsol2d}
p_{ASG2D}(\bx,t) = \frac{1}{2\pi c}\int d\tau \frac{dw_{ASG2D}}{dt}(t-\tau)\frac{H(c\tau-|\bx|)}{\sqrt{c^2 \tau^2 -|\bx|^2}}
\end{equation}

The integral \ref{eqn:asgsol2d} is singular, and inconvenient for computation. The singular factor in the denominator is $\sigma = \sqrt{\tau -|\bx|/c}$, which if used as an alternate integration variable desingularizes the integral. Note that
$\tau = \sigma^2 + |\bx|/c$, $d\tau = 2 \sigma d\sigma$, that the range of $\sigma$ is $[0,\sqrt{t-|\bx|/c}]$, and that
\[
\sqrt{c^2\tau^2-|\bx|^2} = c\sigma \sqrt{\sigma^2 + 2|\bx|/c}.
\]
So
\begin{equation}\label{eqn:asgsol2dalt}
p_{ASG2D}(\bx,t)=\frac{1}{\pi c^2} \int_0^{\sqrt{t-|\bx|/c}}d\sigma \frac{\frac{dw_{ASG2D}}{dt}(t-\sigma^2-|\bx|/c)}{\sqrt{\sigma^2 + 2|\bx|/c}}. 
\end{equation}
Discrete evaluation for this expression requires interpolation of $f$, but the integral is proper so ordinary numerical quadrature works well.

From the relation \ref{eqn:asgacd}, 
\begin{equation}\label{eqn:acdsol2dalt}
p_{ACD2D}(\bx,t)=\frac{1}{\pi c^2} \int_0^{\sqrt{t-|\bx|/c}}d\sigma \frac{w_{ACD2D}(t-\sigma^2-|\bx|/c)}{\sqrt{\sigma^2 + 2|\bx|/c}}. 
\end{equation}
Here $p_{ACD2D}$ solves the 2D version of the system \ref{eqn:acd} with line source pulse $w_{ACD2D}$.

It is easily checked that the right-hand sides of equations \ref{eqn:asgsol2d}, \ref{eqn:asgsol2dalt}, and \ref{eqn:acdsol2dalt} have units of pressure.

### 2019-10-19 repro/papers/iwavepar/paper.tex
repro docs dir added

\title{Parallel IWAVE}
\date{}
\author{William W. Symes, The Rice Inversion Project}

\lefthead{Symes}
\righthead{Parallel IWAVE}

\maketitle
\parskip 12pt
\begin{abstract}
IWAVE supports loop-level, block-level, and task-level parallelism. OpenMP parallelization of loops is available, as is domain decomposition via MPI. MPI also enables parallelism over shot records. This brief report desribes the mechanics of IWAVE parallel execution modes, and includes examples of typical use cases. 
\end{abstract}

\section{Introduction}
IWAVE was designed from the outset for parallel execution. The first release (version 1.0, fall 2009) used domain decomposition via MPI, and optionally loop distribution via OpenMP, for simulation (forward map) only. Later releases introduced task-level parallelism, that is, concurrent execution of shots optionally in combination with domain decomposition, and extended all of these modes of parallelism to derivative and adjoint derivative computations.

This paper explains how to invoke IWAVE in various modes of parallel execution. I apply a selection of these modes to a small 2D example, and so illustrate that the same results are obtained for serial execution as for any mix of parallel options. I will discuss parallelization over shots and domain decomposition here: loop distribution via OpenMP will be discussed elsewhere. The {\tt SConstruct} file in the {\tt project} subdirectory contains examples of all of the constructions mentioned here, and should be regarded as part of the paper.

\section{The IWAVE approach to parallelism}

Domain decomposition is {\em computed} from minimal user input, namely the number of subdomains along each grid axis. The global grid is partitioned into subgrids of equal, or near-equal, size, and each subgrid is assigned to an MPI process. The subgrids contain ghost points required to execute the scheme stencil. Exchange of data to re-initialize ghost points uses the {\tt MPI\_Vectortype} structure to realize virtual subgrids, thus avoiding redundant memory allocation and data motion. OpenMP-based threaded execution co-exists with domain decomposition. See \cite{Terentyev:09a} for details on the initial design of IWAVE, \cite{FehlerKeliher:11} for information on its use in the QC of the SEAM Phase I project. and \cite{GeoPros:11} for a description of a later release.

IWAVE uses minimal user data, namely the number of shots to execute in parallel, to generate a vector of communicators, one per parallel shot. Each communicator is assigned a subset of shots, the subsets being chosen as close to equal size as possible, and a subset of processes that participate in computing each shot assigned to that communicator - one for pure task-level parallelism, more than one if domain decomposition is also being used.

The same principles apply to the forward map and to its derivatives and adjoint derivatives: all are parallelized in the same ways.

IWAVE currently implements disk-to-disk operations. Common input data (eg. model fields) is read and broadcast, shot-dependent data (such as shot gathers) are read or written as needed or generated. The current i/o design is very MPI 1, that is, all i/o takes place on rank 0 of each communicator (note that there may possibly be multiple communicators active in an IWAVE run). Reduction data (non-extended adjoint output) is presumed to be summed (stacked). In task-parallel domain decomposition, each communicator assembles a partial stack over the shots assigned to it, then these results are summed in a final reduction phase.

\section{Building Parallel IWAVE}
IWAVE can be built MPI-enabled, but not as part of the overall Madagascar build. To enable MPI execution, run SConstruct in {\tt RSFSRC/trip}, with a {\tt config.py} file specifying MPI compilation and linking. You will need to put the path to the MPI root directory in your environment as {\tt MPIROOT}, or use one of the module loading systems common at supercomputer sites to provide equivalent information to your shell, then follow the model (or use) one of the config files in {\tt RSFSRC/trip/admin}, for example {\tt RSFSRC/trip/admin/linux.mpi.py} - to use, simply copy to {\tt RSFSRC/trip/config.py}, then {\tt scons}.

This process creates the MPI-enabled executables in the {\tt main} subdirectories, not in the {\tt RSFROOT/bin} directory where other Madagascar commands are found - for example, in this exercise, {\tt RSFSRC/trip/iwave/acd/main/acd.x}. You will need to instrument your {\tt SConstruct} files to find these commands, since the default paths from Madagascar will not work. See {\tt project/SConstruct} for an example.

\noindent {\bf NOTE:} at this writing (2015.09.10) the TRIP directory tree head revision does NOT reside in {\tt RSFSRC} - it is independent of Madagascar, for purely accidental reasons flowing from a catastrophic failure at SourceForge earlier in summer 2015, and lives in the CAAM TRIP repository as {\tt trip2.1}. I anticipate that sometime soon the TRIP tree will migrate back to the new {\tt RSFSRC} GitHub repository. For the time being, where you see {\tt RSFSRC/trip} in this discussion, substitute {\tt trip2.1}. Do NOT use the version of the TRIP tree found in {\tt RSFSRC}, until further notice, as it is increasingly out-of-date.

\section{Job Control}
The TRIP software stack includes a Python module ({\tt RSFSRC/trip/admin/newbatch.py}) defining functions that read standardized job information for serial, command line parallel, or batch execution of IWAVE commands and builds appropriate Madagascar {\tt Flow}s for each case. These data structures specify each essential item of information necessary for job definition with minimal dependencies on other items.

An {\tt SConstruct} file using the IWAVE batch module should include standard boilerplate at the top:
\begin{verbatim}
from rsf.proj import *
from newbatch import tripExec
\end{verbatim}
The {\tt SConstruct} should also define
\begin{itemize}
\item an array of {\em job dictionaries};
\item one or more {\em parallel environment} dictionaries;
\item one or more {\em batch environment} dictionaries.
\end{itemize}
and include a call to {\tt tripExec} following these definitions.
This section describes these data structures and their relations and uses.

The central data structure is the {\em job}, a Python dictionary with these fields:
\begin{itemize}
\item {\tt job}: job name (string),
\item {\tt pre}: preliminary command, to be executed in serial mode. This is mainly provided to include creation of output data files by copy or by another command (from Madagascar or SU, for example) - these need to exist, with correct metadata, before acting as target in an IWAVE command, however they cannot be built in a separate {\tt Flow} since no object can serve as target in two different {\tt Flow}s. Generally these are (low-intensity floating point) commands which one wishes to execute in serial, hence defined separately from the main command below, which may be executed in parallel.
\item {\tt src}: list of source files, including all those used in as sources in the IWAVE command and in the preliminary command. 
\item {\tt tgt}: list of target files.
\item {\tt cmd}: main command, to be executed as serial command, under mpirun, or as part of a batch script.
\item {\tt exe}: execution environment dictionary - defines type of parallel execution (serial, command-line mpi, or batch) with necessary parameters for each.
\end{itemize}
Since each project typically specifies several jobs, the project {\tt SConstruct} file should organize its job dictionaries into an array, eg.
\begin{verbatim}
jobs = [ {job1: {...}}, {job2: {...}},...]
\end{verbatim}

Three types of parallel execution environment are recognized, corresponding to three dictionary structures:
\begin{itemize}
\item serial: an empty dictionary
\item command line mpi: a dictionary with two items,
\begin{itemize}
\item {\tt platf: mpi} - specifies command line mpi
\item {\tt ppn: ...} - number of MPI processes
\end{itemize}
\item batch:
\begin{itemize}
\item {\tt platf}: batch platform name - serves as index into batch environment dictionary, see below
\item {\tt nodes}: number of nodes
\item {\tt ppn}: processes/cores per node
\item {\tt wall}: wallclock time limit, in form xx:xx:xx
\end{itemize}
\end{itemize}

Batch execution requires some additional job-independent characteristics of each environment, such as scripting language, launcher name or path, project name and other accounting information, etc. These need to be listed in a {\em batch environment} dictionary. The information provided in the batch environment dictionary is peculiar to each site and/or machine operated by the site, and may be expected to change as supercomputing sites evolve their software stacks and environments. The dictionary has a standard structure: it uses platform names (from the batch instances in the parallel environment dictionary) as keys; the corresponding values are themselves dictionaries, defining values for the standard keys:
\begin{itemize}
\item {\tt batch}: batch system name, eg. pbs, sge, slurm,...
\item {\tt queue}: name of execution queue.
\item {\tt acct}: account name, to which jobs are to be charged.
\item {\tt mail}: email address to which notification of job begin and end should be sent
\item {\tt launcher}: name of MPI launcher used, eg. mpiexec, ibrun,...
\end{itemize}

Having defined the array of job dictionaries, the parallel environment dictionary (or dictionaries, as different jobs can use different parallel environments), and the batch environment dictionary (which may be empty, if your project has no need of batch job submission), include the line
\begin{verbatim}
tripExec(<name of jobs array>, <name of batch env dict>)
\end{verbatim}
in the project {\tt SConstruct} somewhere below the definitions. The {\tt tripExec} function parses the information contained in its arguments into Madagascar {\tt Flow}s that execute the various jobs with the chosen modes of parallelism. Each {\tt Flow} executes in a subdirectory of the {\tt project} directory, named {\tt jobname.work}, in which {\tt jobname} is the job name assigned in the job dictionary (containing the item {\tt 'job': jobname}). This subdirectory captures the diagnostic output of IWAVE and MPI (for instance, the {\tt cout....txt} files generated by IWAVE, one for each MPI process). The subdirectory is a {\tt Flow} target, so {\tt scons -c} in the project directory gets rid of all of these execution directories and their contents.

\noindent Notes:

\noindent 1. The {\tt Flow}s generated by {\tt tripExec} do not do standard i/o, so this tool is really suitable only for applications like IWAVE. For SU or Madagascar commands structured as filters (i.e. {\tt < inp cmd >outp}, write ordinary Madagascar {\tt Flow}s.

\noindent 2. The implementation of {\tt tripExec} uses {\tt sfbatch}, the Madagascar batch utility, enabling use of the Madagascar {\tt SOURCES} and {\tt TARGETS} macros in the command definition, and keeping each data file name to a unique location (in the {\tt src} or {\tt tgt} list). {\tt sfbatch} incorporates current standard choices for SLURM and PBS at many sites, but given the variety and continued evolution of supercomputer batch environments it's natural to suspect that it may need updating - so beware.

\noindent 3. The various modes of execution may be mixed in a Madagascar project {\tt SConstruct}. In particular, dependence of results on other results produced by batch commands is respected: a {\tt Flow} dependent on the result of a batch computation executes when the batch computation is complete, rather than when the batch submission command is executed.

\noindent 4. For all but serial execution models, the number of processes, and for domain decomposition their geometric layout, must be described. This information can be captured in various ways, but probably the easiest is via a set of Python integer variables, as in the example described in the next section. For example, a script fragment describing a 2D 2 $\times$ 3 domain decomposition of 6 shots, computed 3 at a time, would include in the definition of the IWAVE driver command something like
\begin{verbatim}
NP1=2
NP2=3
NPT=3
...
cmd='... mpi_np1=' + str(NP1) + ' mpi_np2=' + str(NP2) +
    ' partask=' + str(NPT)
\end{verbatim}
For command-line MPI execution, one would naturally use exactly the required number of threads, so
\begin{verbatim}
jobs=[..., {...,'exe': {'platf': 'mpi', 'ppn': str(NP1*NP2*NPT)} },...]
\end{verbatim}
whereas queue management algorithms for batch sites argue for an independent definition of the number of threads, like
\begin{verbatim}
jobs=[..., {...,'exe': {'platf': 'euclid',
                        'nodes': str(NODES),
                        'ppn': str(PPN),
                        'wall': '04:00:00'} },...]
\end{verbatim}
The IWAVE MPI environment function checks that sufficient resources are available: thus IWAVE will abort if {\tt NP1*NP2*NPT > NODES*PPN}.

\noindent 5. {\bf WARNING:} In many cluster environments, partial use of a node (using fewer cores than available on the number of nodes allocated) can result in jobs hanging. Therefore, until we see some portable and transparent way around this limitation, the number of tasks defined by IWAVE ({\tt NP1*NP2*NPT} in the preceding example) should be the same as the number of cores allocated ({\tt NODES*PPN} in the example).

\section{Examples}
\inputdir{project}

I have used the simple 4-layer OBC example from \cite[]{trip14:struct} to illustrate the various parallel IWAVE modes for serial and command line mpi execution. The command used in these examples is {\tt iwave/acd/main/acd.x}, which implements the forward map, its first and second derivatives, and their adjoints. This application is described in some detail in \cite{trip14:struct}.

The paper \cite[]{SymesIWEXT:15} describes a similar example. The reader should examing the {\tt project/SConstruct} files for both papers to see ``live'' examples of the framework. In particular, the examples attached to this paper are configured for command-line MPI, rather than batch. To see an explicit example of the mechanics of batch submission, see the examples attached to the companion paper \cite[]{SymesIWEXT:15}.

The velocity model for the 4-layer OBS example appears as Figure \ref{fig:csq4layer}. Other details of the simulation are as described in \cite{trip14:struct}, and in the figure captions.
Figures \ref{fig:shot11}, \ref{fig:shot12}, \ref{fig:shot21}, and \ref{fig:shot22} show the results of modeling an OBC gather (effectively, a shot on the seafloor and near-surface receivers) with no domain decomposition (or 1 $\times$ 1), versus 1 $\times$ 2, 2 $\times$ 1, and $2 \times 2$ decompositions, all plotted on the same scale. The differences are at round-off level. The parameters {\tt mpi\_np1} and {\tt mpi\_np2} describe the number of domains along the first and second axes, and were set at 1 and 2 respectively. Thus, 1, 2, 2, and 4 MPI processes were involved in these simulations respectively. It is an error to provide too few processes to assign each domain a process. However providing more processes is not an error; the unused processes are simply idled.

Figures \ref{fig:born11}, \ref{fig:born12}, \ref{fig:born21}, and \ref{fig:born22} describe the analogous results for the Born approximation. Once again, the differences are at round-off level. Finally, Figures \ref{fig:rtm11}, \ref{fig:rtm12}, \ref{fig:rtm21}, and \ref{fig:rtm22} describe the reverse time migration (adjoint linearized modeling) of the corresponding Born data.

All of these results should be compared the analogous plots in \cite{trip14:struct}.

Finally, to illustrate shot parallelization and its combination with domain decomposition, I simulated 9 shots, with 3 computed concurrently ({\tt partask=3}). I obtained the result displayed in Figure \ref{fig:shot8-12km}. This runs under mpi command line control with {\tt -np 3} (resulting from the parallel environment dictionary entry {\tt 'ppn': '3'}. Initially, shots 0, 3, and 6 are distributed to processes 1, 2, and 3. Then shots 1, 4, and 7 are distributed in the same way, and finally 2,5, and 8.

Note that the communicator design described above causes the algorithm to behave like an asynchronous queue - for example process 2 could be working on shot 4 when process 1 is still on shot 0. In general, IWAVE takes an optimistic point of view regarding load balancing: it simply relies on similar-sized jobs taking similar lengths of time to complete. This approach has definite limits, but it is perhaps reasonable as a first cut for a solver based on uniform rectangular meshes.

Figure \ref{fig:shot8-12kmdd} shows the same job processed with 3 concurrent shots and a 1 $\times$ 2 domain decomposition. In this case, processes 1 and 2 start on shot 0, processes 3 and 4 on shot3, and processes 5 and 6 on shot 6, and so on. As was the case in other examples, the differences between results obtained with different levels of parallelism are within round-off of each other.

\plot{csq4layer}{width=0.8\textwidth}{4-layer velocity-squared model. First layer has $H_2O$ properties: model sea bottom is at depth = 1875 m.}
\multiplot{4}{shot11,shot12,shot21,shot22}{width=0.45\textwidth}{Shot over model of Figure \ref{fig:csq4layer} at OBC position $x=$ 8 km, $z=$ 1.875 km, 400 receivers in 5 km $\le x \le $ 15 km, $\Delta x$ = 25 m. Isotropic point source as constitutive law anomaly, Gaussian derivative wavelet, 5 Hz peak frequency. Domain decomposition parameters ({\tt mpi\_np1,mpi\_np2}) = (1,1) (no domain decomp), (1,2), (2,1) and (2,2).}

\plot{csq4layersm}{width=0.8\textwidth}{Background model: smoothing of 4-layer velocity-squared.}
\plot{dcsq4layer}{width=0.8\textwidth}{Reflectivity model: difference between velocity-squared (Figure \ref{fig:csq4layer}) and a less aggressive smoothing than that shown in Figure \ref{fig:csq4layersm}.}

\multiplot{4}{born11,born12,born21,born22}{width=0.45\textwidth}{Linearized (``Born'') shot over model of Figure \ref{fig:csq4layer} at OBC position $x=$ 8 km, $z=$ 1.875 km, 400 receivers in 5 km $\le x \le $ 15 km, $\Delta x$ = 25 m. Isotropic point source as constitutive law anomaly, Gaussian derivative wavelet, 5 Hz peak frequency. Velocity-squared as in Figure \ref{fig:csq4layersm}, reflectivity (velocity-squared perturbation) as in Figure \ref{fig:dcsq4layer}. Domain decomposition parameters ({\tt mpi\_np1,mpi\_np2}) = (1,1) (no domain decomp), (1,2), (2,1) and (2,2).}

\multiplot{4}{rtm11,rtm12,rtm21,rtm22}{width=0.45\textwidth}{Adjoint linearized map (RTM) applied to data of Figures \ref{fig:born11}, \ref{fig:born12}, \ref{fig:born21}, and \ref{fig:born22}, with same domain decomposition parameters.}

\multiplot{2}{shot8-12km,shot8-12kmdd}{width=0.45\textwidth}{9 shots from $x_s$ = 8 km to $x_s$ = 12 km, 3 shots computed at once ({\tt partask=3}); no domain decomposition on left, 1 $\times$ 2 domain decomposition on right.}

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/iwavext/paper.tex
repro docs dir added

\title{Extended IWAVE}
\date{}
\author{Raanan Dafni and William W. Symes, The Rice Inversion Project}

\lefthead{Dafni and Symes}
\righthead{Extended IWAVE}

\maketitle
\parskip 12pt
\begin{abstract}
IWAVE accommodates several extended modeling modes. This paper explains how to implement  shot record extension (independent simulation of shots with one model per shot), subsurface offset extension (mechanical parameters as non-diagonal operators), and source extension (independent source parameters for each trace, or each shot, or both), for constant-density acoustic modeling. The modeling operators and their first and second derivatives inherit all features of IWAVE simulation - dataflow design, shot and multidomain parallelism, computed memory allocation, command line job control, etc.  
\end{abstract}

\section{Introduction}

\section{Mathematical Setting}
The central design principle of IWAVE/RVL can be stated as follows: {\em the mathematics is the API}. So the first task is to describe the mathematics of extended modeling.

To make the illustrations in this paper as simple as possible, we choose constant density acoustic wave propagation as the underlying physics. The sole material parameter field is the velocity squared, $m=v^2$, proportional to bulk modulus by the (constant) density. The second order wave equation for causal acoustic dynamics is
\be
\label{awe}
\frac{\partial^2 p}{\partial t^2} - m\nabla^2 p = f,\,\,p = 0, t<<0.
\ee
in which $p=p(\bx,t,\bx_s)$ is the pressure field, $\bx_s$ is the source parameter (which may be location, or slowness, or ...), $f = f(\bx,t,\bx_s)$ is the source field. The modeled seismogram is the trace, or sampling, of $p$ at (possibly $\bx_s$-dependent) receiver locations. For the moment, we will regard the source as known and fixed, and write $\cF(m)$ for the collection of traces, so determined, a point in the data space $D$, the range of the forward map $\cF$ whose  domain is the model space $M$.

An extended model space $\oM$ is simply larger: $M$ ``$\subset$'' $\oM$ - as we will see, the inclusion need not be literally true. Typically extended modeling means parameters added somehow to the model, beyond those specified by the basic physics, but the fundamental requirement is that there be an extended modeling operator $\ocF: \oM \rightarrow D$ and an extension operator $E: M \rightarrow \oM$ so that 
\be
\label{ext}
\ocF(E(m)) = \cF(m), \,\,m \in M.
\ee

\subsection{Shot Record Extension}

\subsection{Subsurface Offset Extension}

\subsection{Dataflow Control}
IWAVE controls the selection of extension (or no extension) by {\em input and/or output data format}. This idea is described in \cite[]{trip14:struct}. Briefly,
\begin{itemize}
\item if the grid of the input fields (or output fields, in the case of the adjoint linearized map) contains one or more extended axes, then IWAVE computes an extended map. As explained in \cite[]{trip14:struct}, IWAVE uses an enhanced version of the Madagascar data structure to differentiate extended from non-extended axes: added keywords are
\begin{itemize}
\item {\tt dim} = spatial dimension, should be same as reference grid
\item {\tt gdim} = global dimension, including extended axes - for typical 2D extended modeling, this is {\tt dim} + 1.
\item {\tt idxxx} for {\tt xxx}=1,...{\tt dim}-1 are the id's of the spatial axes, {\tt xxx}={\tt dim} signifies the time axis, {\tt xxx}={\tt dim} + 1,...,99 are available for external extended axes such as shot number, and {\tt xxx}=100,... signify internal extended axes such as subsurface offset.  ``External'' means that only a single value participates in the simulation of a single shot; ``internal''  means that all (or many) points on the axis particpate in single shot simulation, as is the case for spatial axes.
\item for linearized or adjoint linearized maps, 
\begin{itemize}
\item if the grid of the input reference fields is the same as the grid of the perturbation inputs or outputs ({\tt csq\_d1} or {\tt csq\_b1} for {\tt acd} first derivative and adjoint first derivative respectively), then IWAVE computes the linearization (or adjoint linearization) of the (extended or non-extended) forward map.
\item if the grid of the input reference fields is not extended (has no extended axes), but one or more perturbations is extended, then IWAVE computes the extended linearization (or its adjoint) about the implicitly extended reference field. This case is typical for migration velocity analysis.
\end{itemize}

\end{itemize}
In all cases, the identification keywords of extended axes tell IWAVE what sort of extension to compute.
\end{itemize}

\inputdir{project}

\section{Example: Coarse Grid Marmousi}
This example uses the Marmousi model subsampled to $dx = dz = $ 24 m. The Born simulation  uses a well-smoothed background model (Figure \ref{fig:csq24big}) and a reflectivity derived from the original model with a less agressive smoothing removed (Figure \ref{fig:dcsq24}). Velocity-squared and its perturbation are plotted, as these are the quantities used in the simulation. The acquisition geometry is the original, subsampled by a factor of 4 to give a 60-shot simulated towed streamer geometry. Time of recording is cut back to 2 s.

\plot{csq24big}{width=0.8\textwidth}{Background model: square of smoothing of Marmousi velocity.}
\plot{dcsq24}{width=0.8\textwidth}{Reflectivity model: difference of squares of Marmousi velocity and a smoothing.
}

All of the computations reported here were performed at the Texas Advanced Computing Center (TACC), University of Texas-Austin, using TACC's Stampede Linux cluster. The source directory for this paper contains the {\tt project/SConstruct} script file describing every command, and illustrates the configuration of reproducible computational experiments in a batch environment (SLURM). The (small) jobs described below ran on 32 threads (2 nodes) and complete in well under a minute. See the companion paper \cite{}{SymesIWPAR:15} for a description of the software tools used in these examples. Note that the number of tasks (keyword {\tt partask} = number of shots to execute in parallel) was also 32, for reasons explained in \cite[]{SymesIWPAR:15}. Since this example involves 60 shots, IWAVE idles some threads near the end of the run.

\subsection{Shot Record Extension}
All output (and, of course, input) data objects must exist for IWAVE to function properly. To build a prototype shot record extended model that can be used as an output object for shot record extended migration, use {\tt sfspray} to duplicate the perturbational model (or any other RSF data object with the reference grid) 60 times. [For implementations of this and all other computations described here, see the {\tt SConstruct} file in the {\tt project} subdirectory.]

In addition to duplicating the reference grid, your command must decorate the rsf header file as described above: use {\tt sfput} to add {\tt id1=0, id2=1, id3=3, dim=2,} and {\tt gdim=3} to the header file. Note that {\tt id3=3} sets the first axis numbered above the time axis (hence extended) to {\tt dim} + 1, which is the index of the shot record axis in the simulation grid (see \cite[]{trip14:struct} for explanation). These choices will cause IWAVE to overwrite a shot record extended image on the output (Figure \ref{fig:migsr}). As expected, the common image gathers reflect the correct velocity in being as flat as possible subject to the limited aperture of shot record migration and presence of edge effects (Figure \ref{fig:migsr}, right panel).

\plot{migsr}{width=0.8\textwidth}{Shot record extended migration cube: front face is image}

\subsection{Subsurface Offset Extension}
In this case, use {\tt sfpad} to add a subsurface offset axis to the perturbational model to create a prototype for migration output. I chose to use 20 grid cells to the left and right of offset 0, with offset increment the same as the other spatial axis increments (24 m), so a maximum subsurface offset of 480 m. Header corrections for this case are {\tt id1=1, id2=0, id3=100, d3=24, o3=-480, dim=2,} and {\tt gdim=3}. Note all of the spatial fields are transposed in this case: while the model field fetched from the TRIP web site has $z$ as the fast axis, {\em horizontal} subsurface offset extension is {\em required} to use $x$ as fast axis. There are two reasons for this choice: (1) the organization of loops in the numerical kernals can be maximally vectorized with this choice, and (2) domain decomposition is not permitted on the fast axis, since the effective stencil width is very large in this dimension, and the restriction to $x$ allows a simple implementation of this restriction. Since the physical significance of the axes is part of the data structure, IWAVE can take axis ordering into account so that it is transparent to the simulation. The data setup must include transposing of the velocity squared field and all perturbations, of course, and to display the output of the migration in the normal way it must be transposed back to fast-$z$ order. The choice {\tt id3=100} toggles internal extension: at the level of IWAVE itself, this choice causes appropriate memory allocation to occur. The interpretation of the extended axis as horizontal subsurface offset is an attribute of the numerical kernels that go into {\tt acd} (not of the core IWAVE code).

The resulting migration is reasonably well-focused at offset zero (Figure \ref{fig:migso}, right panel), and the zero-offset section is identical to non-extended migration (Figure \ref{fig:migso}, front panel).

\plot{migso}{width=0.8\textwidth}{Subsurface offset extended migration cube: front face is image}

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/lsmabs18/paper.tex
repro docs dir added

\title{Accelerated Acoustic Least Squares Migration}
\renewcommand{\thefootnote}{\fnsymbol{footnote}} \author{Raanan
  Dafni, William W. Symes\footnotemark[1], The Rice Inversion Project,
  Rice University}
\captionsetup{font=scriptsize,labelfont=scriptsize}
\righthead{Accelerated LSM}
\lefthead{Dafni and Symes}
\maketitle

\begin{abstract}
Least squares migration cost depends on the number of iterations
required to produce a satisfactory image volume and/or fit to data. The
cost can be reduced considerably through effective
preconditioning. One such preconditioner derives from true-amplitude
or asymptotic approximate inversion. We describe implementation details of a
true-amplitude preconditioned least squares migration algorithm, and illustrate
its performance using a synthetic 2D example.
\end{abstract}

\section{Introduction}
Least squares migration (LSM) based on two-way wave equation modeling
produces optimized image volumes - in fact inversions - that correct
many of the basic defects of reverse time migration images
\cite[]{Dutta:SEG14,DaiBS:2010,DaiSchuster2013,Nemeth:99,ChaventPlessix:98}. LSM
accommodates any wave physics that can be modeled in the linearized
(``Born'') sense \cite[]{duan2017elastic,sun:SEG15,DuttaSchuster:14},
and can produce information about the earth's structure well beyond a
mere image. However LSM is necessarily an iterative process, as the
linear least squares problem it defines is far to large to admit
matrix solution methods at any interesting scale. Each iteration
requires a combination of modeling (demigration) and migration. The
computational expense can therefore be considerable.

\cite{HouSymes:EAGE16} have described the use of asymptotic approximate
  inverse formulae (``true amplitude migration'') of an appropriate form, derived by
  \cite{HouSymes:15}, to accelerate the
  convergence of conjugate gradient LSM. Here we base the construction
  of a preconditioner instead on an alternative construction of approximate inversion 
  \cite[]{HouSymes:17,tenKroode:12}, leading to a simpler and equally
  effective algorithm. We explain several crucial 
  implementation details which have not been described elsewhere but
  which are essential for successful convergence
  acceleration, and illustrates the speedup obtainable in the acoustic
  case via a synthetic example.

\section{Method}
In this section we explain our approach to acoustic modeling and RTM, the
structure of the approximate inverse and its key role in successful
preconditioning of the LSM iteration, and several computational
constraints and short-cuts that lead to a practical and efficient
algorithm.

\subsection{Modeling and Migration}
Born acoustic modeling is the solution of the system
\ref{eqn:awe}-\ref{eqn:icwe} of partial differential equations for the
excess pressure and particle velocity fields $p, \bv$ and their
perturbations $\delta p, \delta \bv$ in response to perturbations
$\delta \kappa, \delta \beta$ in the bulk modulus and
buoyancy (reciprocal density) fields $\kappa$, $\beta$:
\begin{equation}
\label{eqn:awe}
\frac{\partial p}{\partial t} = - \kappa \nabla \cdot \bv + f;
\,\frac{\partial \bv}{\partial t} = - \beta \nabla p
\end{equation} 
\begin{equation}
\label{eqn:bwe}
\frac{\partial \delta p}{\partial t} = -\delta  \kappa \nabla \cdot
\bv - \kappa \nabla \cdot \delta \bv ;
\,\frac{\partial \delta \bv}{\partial t} = - \delta \beta \nabla p -
\beta \nabla \delta p
\end{equation} 
\begin{equation}
\label{eqn:icwe}
p, \delta p, \bv, \delta \bv, f =0,\, t<<0
\end{equation}
In addition to the initial conditions \ref{eqn:icwe}, boundary
conditions are also required to make the solution unique. In all of
the examples presented here, we solve the system
\ref{eqn:awe}-\ref{eqn:icwe} in a rectangular subdomain of the 2D
half-space $z \ge 0$, use PML absorbing boundary conditions \cite[]{Habashy:07} on the
lateral sides and bottom, and impose either the pressure-release
condition $p=0$ or a PML condition (both are needed, as will be seen)
on the top surface $z=0$

The source fields $f$ represent isotropic point sources
$f(\bx,t)=w(t)\delta(\bx-\bx_s)$ for a line of source positions
$\bx_s$. The linearized, or Born, traces $\delta d(\bx_s,\bx_r,t)$ are
the samples of $\delta p$ at source-receiver pairs
$\bx_s,\bx_r$. Since these depend on both the background parameters
$\kappa$, $\beta$ and the perturbations $\delta \kappa$, $\delta
\beta$, and are linear in the latter, this relation defines the
absorbing and free surface Born modeling operators $F$ and $F_0$ respectively:
\begin{eqnarray}
\label{eqn:afdef}
\{\delta p(\bx_r,t;\bx_s)\} &=& F[\kappa,\beta][\delta \kappa, \delta \beta]\mbox{
  (abs. surface)}\\
\label{eqn:ffdef}
\{\delta p(\bx_r,t;\bx_s)\} &=& F_0[\kappa,\beta][\delta \kappa, \delta \beta]\mbox{ (free. surface)}
\end{eqnarray}
Since in LSM the background fields $\kappa, \beta$ are not updated, we
drop them from the notation. Also, we will set $\delta \beta = 0$
throughout, for simplicity, thus writing $F \delta \kappa$ for the
output of the Born modeling operator.

 We use the well-known staggered grid
finite difference method, of order 2 in time and 2$k$ in space
($k=$ 4 in the example below) to approximate solutions of the system
\ref{eqn:awe}-\ref{eqn:icwe}.

Reverse-time migration is taken to mean application of the {\em
  transpose} $F^T$ or $F^T_0$ of the Born modeling operator. We use the
well-known adjoint state method to compute the action of these operators
\cite[]{Plessix:06}. We
do not specify the adjoint system of differential
equations  here, as
we do not directly discretize it, but rather create the transpose
mapping of the discrete Born modeling operator as produced by the
finite difference method. This approach guarantees that the computed
$F^T$ (resp. $F_0^T$) is actually the transpose of the discrete $F$
(resp. $F_0$), that is, passes the ``dot
product test'' to machine precision. This discrete adjointness
relation is very important in ensuring that the conjugate gradient
method exhibits its theoretical convergence properties.

Our method for computing $F^T$, $F_0^T$ with these properties relies on the
automatic differentiationn software {\em TAPENADE}
\cite[]{TapenadeRef13} to create finite
difference kernels for the Born fields (equation \ref{eqn:bwe}) and
their corresponding adjoint kernels, and iterating these in an
abstract loop framework adaptable to any wave physics \cite[]{GeoPros:11}. The adjoint state method requires backwards-in-time
reconstruction of the source wavefields $p,\bv$: we use optimal
checkpointing \cite[]{Griewank:00,Symes:06a-pub} to avoid introducing any error
beyond round-off in this step.

We also note that in the theoretical derivation of the adjointness
relation, and the asymptotics calculations in the next section,
$F^T$ and $F_0^T$ involve an {\em integral} over source and
receiver positions. Thus the sum over source and receiver positions
found in naive descriptions of RTM must be replaced by a
quadrature. If the source and receiver spacings are uniform, then a
simple cell-size factor is sufficient - we have taken advantage of
that observation in the synthetic example below.

\subsection{Asymptotic Inversion}
Beginning with \cite{CohBlei:77}, high frequency asymptotics has been
used to construct approximate inverses to Born modeling. All such
constructions require normalization of the energy source. In terms of
the system \ref{eqn:awe}-\ref{eqn:icwe}, the relevant normalization is
the step/Heaviside point source, $f(\bx,t) =
H(t)\delta(\bx-\bx_s)$. In practice this idealized source must be
modified by a bandpass filter. For data supplied with a source pulse
estimate, a shaping filter should be constructed outputting a
bandlimited Heaviside pulse, and applied to the data.

We focus first on the absorbing surface operator
$F$. The results derived in \cite{HouSymes:17} imply that with the source normalized as just
described,
\begin{equation}
\label{eqn:alt1}
F^{\dagger} = V_m^{-1}F^TV_d,\, V_m^{-1}= 32\beta \kappa^3 \partial_z,\, 
V_d = I_t (I_t\partial_{z_s})(I_t\partial_{z_r}) 
\end{equation}
defines an asymptotic (in frequency) inverse $F^{\dagger}$ for $F$,
that is,
\begin{equation}
\label{eqn:appinv}
F^{\dagger}F \approx I.
\end{equation}
In equation
\ref{eqn:alt1}, $I_t$ represents the indefinite $t$ integral, and
$\partial$ signifies the partial derivative with respect to the
variable in the subscript.

Actually, the result of \cite{HouSymes:17} pertains to the {\em
  subsurface offset} extension of Born modeling, not to the operators
discussed here which do not require extended modeling calculations. In
fact equation \ref{eqn:alt1}, with the meaning of the various symbols
as defined here, expresses the {\em zero (subsurface) offset} section
of the inversion defined by \cite{HouSymes:17}, possible because the
operators $V_m^{-1}$ and $V_d$ operate trace-by-trace so can be
restricted to the zero subsurface offset section. In the extended
modeling context, the analogs of $F^{\dagger},F$ are approximately
inverse to each other, as asserted in equation \ref{eqn:appinv}. The
zero-offset section of the extended approximate inverse, namely
$F^{\dagger}$ as defined here, cannot be exactly inverse to $F$, as
the energy in the extended inverse is necessarily spread over at least
a wavelength in the offset direction. Recovery of the bulk modulus
perturbation from the extended image volume requires a {\em stack},
approximating the integral over subsurface offset, discretely
the sum multiplied by the offset step (which must be the same as the
horizontal space step $\Delta x$). Thus the zero offset section gives
an image smaller by roughly the number of gridpoints per wavelength,
multiplied by $\Delta x$, than would be the case for an approximate
inverse. We introduce a factor of $\Delta x$ in the approximate
inverse output, so that the remaining scale difference is roughly the
number of gridpoints per wavelength - a less well-defined quantity. As
shall be shown below this residual scale error is absorbed in the LSM
process.

Two amendments are necessary to render $F^{\dagger}$ as defined above
a useful accelerator for LSM.
$F^{\dagger}$ would be the adjoint of $F$ with respect to weighted inner
products with Gram operators $V_m$ and $V_d$, if it were the case that these have
the necessary properties of Gram operators, namely symmetry and
positivity. Neither $V_m$ (or $V_m^{-1}$) nor $V_d$, as defined above,
have either property. We can however alter the definitions to recover
these properties. Note that $\partial_t = H |\partial_t|$, in which
$H$ is the Hilbert transform and $|\partial_t|$ is represented in the
Fourier domain by multiplication with
the absolute value of frequency. A well-known result from analysis
(Egorov's Theorem, \cite{Tay:81}) allows us to pass $H$ from one
side of $F^T$ to the other in equation \ref{eqn:alt1}, where it is
``absorbed'' by its inverse $-H$ coming from the first $I_t$. The
upshot is an alternate expression 
\begin{eqnarray}
\label{eqn:alt2}
F^{\dagger} & = & W_m^{-1}F^TW_d, \nonumber \\
 W_m^{-1}&=&32\beta^{1/2} \kappa^{3/2} K_z \beta^{1/2}
             \kappa^{3/2},\nonumber \\
W_d &=& -K_t^{-3} \partial_{z_s}\partial_{z_r}.
\end{eqnarray}
Here $K_z$ is defined in the frequency domain by multiplication with
$|k_z|$, supplemented with an appropriate bandpass filter, and $K_t$
is defined similarly. $W_m^{-1}$ is manifestly symmetric and positive.

Concerning $W_d$, \cite{HouSymes:15} demonstrated
that $-\partial_{z_s}\partial_{z_r}$ is symmetric and
positive if the source wavefield is downgoing near the surface, and
the receiver wavefield is upcoming. In general guaranteeing this
property requires decomposition of the fields into up- and down-going
components. However, when sources and receivers are embedded in a
reflector-free layer (simulating the ocean) then these properties
necessarilly hold. Therefore it is only necessary in that setting to
apply a taper operator on the input side of $F$ (and the output side
of $F^T$) to make sure that no artificial reflectors
appear in the water column. Note that if this step is left
out, reflectors inevitably develop in the water column, $W_d$ loses
its positivity, and the algorithm defined in the next section fails to
converge.

We have idenfitied several methods to compute the $z_s, z_r$ derivatives in the definition of $W_d$. If the $z$ component of motion is measured at the receiver (eg. OBN or
Geostreamer$^{TM}$ data), then the second equation of motion
\ref{eqn:bwe} supplies $\partial_{z}\delta p|_{\bx=\bx_r}
= \partial_{z_r}F$. In general, if the source and receiver wavefields
are separated into up- and down-going components, then each obeys a
one-way wave equation, which specifies the $z$ derivative in terms of
the $t$ derivative and other spatial derivatives.

In the example reported below, we have used yet another method,
mentioned briefly by \cite{HouSymes:15}. We
presume that the source and receiver depths for the absorbing surface
fields are both $=0$. The corresponding free surface ($z=0$) acquisition geometry satisfies the {\em shallow tow depth}
assumption if
\begin{equation}
\label{eqn:std}
\partial_{z_s}\partial_{z_r} \delta p_{z_s = z_r=0} \approx
\frac{1}{4z_sz_r}Z\delta p_0
\end{equation}
In equation \ref{eqn:std}, $\delta p_0$ is the free surface Born field at prescribed $z_s, z_r$, and $Z$ sets the source and receiver depths in
each free surface trace to zero and leaves all other header words and
data samples unchanged. The level of error implicit in the $\approx$
symbol should be on the order of the discretization error in the finite
difference scheme, which is the case if $z_s$ and $z_r$ are comparable
to the grid step in $z$. If the condition \ref{eqn:std} holds, then 
\begin{equation}
\label{eqn:stdop}
\partial_{z_s}\partial_{z_r} F \approx \frac{1}{4z_sz_r}Z F_0.
\end{equation}
Therefore the definition \ref{eqn:alt2} translates to
\begin{equation}
\label{eqn:appinvab}
F^{\dagger} \approx - W_m^{-1} \frac{1}{4z_sz_r}F_0^TZ^TK_t^{-3}.
\end{equation}
That is, we have used the ghosted traces to emulate the $z_s,z_r$
derivatives. Such emulation is only valid up to the first ghost notch,
however that is precisely the frequency range for which the criterion
mentioned above assures accuracy of the finite difference simulation.

The examples below employ the free surface condition. Similar logic
leads to the approximate inverse formula
\begin{equation}
\label{eqn:appinvfr}
F_0^{\dagger} \approx - W_m^{-1} \frac{1}{4z_sz_r}F^TK_t^{-3}Z.
\end{equation}

Note that the absorbing surface RTM operator appears in the expression
for the free surface true amplitude migration, and vis-versa.

\subsection{Preconditioned Conjugate Gradient Iteration}
Use of the inner products with weight operators $W_m$ and $W_d$ in
model and data spaces and the transpose $F^{\dagger}$ with respect to
these inner products in conjugate gradient iteration produces a rapidly convergent algorithm, since
with respect to the weighted inner products $F$ is approximately
unitary (up to a scale factor). The computation of the weighted inner products is
tedious. Fortunately, the preconditioned conjugate gradient method
\cite[]{Golub:2012} is
mathematically equivalent, and uses only the Euclidean inner
products. Define $F_0^*$ to be the transpose of $F_0$ with respect to the
Euclidean inner product in model space, and the $W_d$-weighted inner
product in data space:
\begin{equation}
\label{eqn:uwadj}
F_0^* = F_0^T W_d = - \frac{1}{4z_sz_r}F^TK_t^{-3}Z
\end{equation}
Use $F_0$, $F_0^*$ to define the least squares problem and and
$W_m^{-1}$ as the preconditioner. The iteration converges to the minimizer of the
data-weighted mean square error
\begin{equation}
\label{eqn:obj}
J[\delta \kappa] = \frac{1}{2}(F_0\delta \kappa -\delta d)^T W_d
(F_0\delta \kappa -\delta d).
\end{equation}

\section{Numerical Example}
We present a 2D synthetic numerical illustration based on the Marmousi
blind test model \cite[]{BoLaVe:91}. We take $\beta = 1, \delta \beta = 0$
throughout. The smoothed background bulk modulus $\kappa$ appears in
Figure \ref{fig:bulk}, and the target bulk modulus perturbation in
Figure \ref{fig:dbulk}.  We discretized the model on a 12 m $\times$
12 m grid, and used a [2.5, 5, 20, 25] Hz zero-phase bandpass filter, integrated
once to yield a bandlimited Heaviside function, as the source in both
modeling and migration. 180 source positions at $z_s=12 m$ and 360
receiver positions at $z_r=12 m$ were spread across the top of the
model (free surface, full aperture, fixed spread). 
%A typical shot gather appears as
%Figure \ref{fig:shot151}.

LSM via 20 iterations of (unpreconditioned) conjugate gradient
iteration produces the estimated bulk modulus perturbation in Figure
\ref{fig:lsm}, plotted on the same color scale as Figure
\ref{fig:dbulk}. While relative amplitudes above 2500 m depth are
reasonably well recovered, the deeper part of the model is not, and
the data residual is reduced only to 40\% of its initial value. 
Application of the approximate inverse yields the estimate in Figure
\ref{fig:appinv}, with relative amplitudes that are actually more
accurate (for the price of 1 RTM) than was achieved with 20 LSM
iterations. The preconditioned LSM achieves the result displayed in
Figure \ref{fig:plsm} after 20 iterations. The recovered approximation
of the bulk modulus perturbation is roughly as accurate as the data
bandwidth allows, and the data residual is under 10\%. The convergence
curves in Figure \ref{fig:conv} show clearly the cost reduction
achieved through true amplitude preconditioning.

\section{Discussion}

%\cite{HouSymes:SEG16b} show that the same technique used here to
%accelerate LSM may be used to accelerate Full Waveform
%Inversion (FWI). Acceleration does not mean avoidance of cycle-skipping: if
%basic FWI does not converge to a useful result, the accelerated
%algorithm simply achieves the useless result faster. However the
%improvement in convergence is as dramatic as shown here for LSM, and
%could conceivably be combined with other techniques that render FWI
%more robust.

The preconditioning approach presented here is not the
only one suggested for acceleration of LSM iteration
\cite[]{Aoki:09,Berkhout:09d,HuangNammourSymes:16,RochaSavaGuitton:17,Guitton:17}. A
very limited comparison suggests that true-amplitude preconditioning
may be more effective than some of the other techniques discussed in
these references, when it applies. However, as described here, it is
limited to recovery of a single parameter (we have used bulk
modulus). True amplitude recovery of multiple parameters is possible,
but in principle requires the use of extended modeling and migration
\cite[]{YuZhang:14}. On the other hand, analogous results should be
possible for various forms of
pseudo-acoustic anisotropy.

\section{Conclusion}
We have presented details of a true-amplitude acceleration of LSM. The
accelerated iteration has essentially the same cost per step as
ordinary LSM, but converges substantially more rapidly. We have
presented the 2D version here, corresponding to our example, but
similar acceleration formulae are available for 3D, at the same low cost.

\section{ACKNOWLEDGMENTS}
We are grateful to the sponsors of The Rice Inversion Project (TRIP)
for their long-term support, to Shell International Exploration
and Production Inc for its support of Raanan Dafni's postdoctoral work
and  Jie Hou's PhD studies in TRIP, and to Fons ten Kroode for inspiring our
work on this topic. We acknowledge the Texas Advanced Computing
Center (TACC) and the Rice University Research Computing Support Group
(RCSG) for providing essential HPC resources.

\plot{bulk}{width=0.45\textwidth}{Marmousi bulk modulus sampled to 12
  m $\times$ 12 m grid, smoothed with 4 iterations of a 3 $\times$ 3
  rectangular moving average.}

\plot{dbulk}{width=0.45\textwidth}{Marmousi bulk modulus perturbation
  = difference between original bulk modulus and slightly milder
  smoothing than that shown in Figure \ref{fig:bulk}.}

\plot{lsm}{width=0.45\textwidth}{LSM from data described in text, via
  20 conjugate gradient iterations. Note that color scale
  is same as in Figure \ref{fig:dbulk}.}

\plot{appinv}{width=0.45\textwidth}{True amplitude migration of
  Marmousi data described in text. Note that relative amplitudes are
  quite similar to target bulk modulus perturbation (Figure
  \ref{fig:dbulk}), however overall amplitude is smaller by a nearly
  position independent factor.}

\plot{plsm}{width=0.45\textwidth}{LSM from data described in text, via
  20 preconditioned conjugate gradient iterations. Note that color
  scale is same as that in Figure \ref{fig:dbulk}.}

\plot{conv}{width=0.45\textwidth}{Mean square data error as function
  of iteration, with (blue) and without (red) true amplitude preconditioning.}
\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/lsmabs18/slides.tex
repro docs dir added

\documentclass[xcolor=dvipsnames,12pt,aspectratio=169]{beamer}
\usecolortheme[named=violet]{structure}

%\usetheme{umbc2} 
\usepackage{pgf}
%\usepackage[pdftex]{graphicx}
\usepackage{color}
\usepackage{amssymb,amsmath}
\usepackage[english]{babel}
\mode<presentation>

 \usepackage{multimedia}
\usefonttheme{structurebold}

%\useinnertheme{umbcboxes}

%\setbeamercolor{umbcboxes}{bg=violet!15,fg=black}

\setbeamercovered{transparent}
\setbeamertemplate{navigation symbols}{}
\input{./math-macros}
\setlength{\parskip}{2ex}   % place a blank line between paragraphs
\def\Logo{\@ifnextchar(\beamer@Logo\logo}
\def\beamer@Logo(#1,#2){\logo}

\AtBeginSection[]{
{\setbeamertemplate{logo}{}
 \begin{frame}<beamer>
    \frametitle{Agenda}
    \tableofcontents[currentsection]
  \end{frame}
}
}
% Images
% Macros
\newcommand{\peq}{\,+\hspace{-0.15cm}=}
\newcommand{\bm}{\bar{m}}
\newcommand{\dm}{\delta m}
\newcommand{\dom}{\delta \bar{m}}
\newcommand{\doma}{\delta \bar{m}_{\alpha}}
\newcommand{\om}{\bar{m}}
\newcommand{\bH}{{\bf H}}
\newcommand{\bR}{{\bf R}}
\newcommand{\bcF}{{\bar{\cal F}}}
\newcommand{\cF}{{\cal F}}
\newcommand{\oF}{\bar{F}}
\newcommand{\odF}{\bar{F}^{\dagger}}
\newcommand{\oS}{\bar{S}}
\newcommand{\oM}{\bar{M}}
\newcommand{\Ja}{J_{\alpha}}
\newcommand{\Na}{N_{\alpha}}
\newcommand{\dd}{\delta d}
\newcommand{\ddl}{\delta d_{\alpha}}
\title[]{Accelerating Acoustic Least Squares Migration}
\author[]{Raanan Dafni \& William W. Symes}
\institute[]{The Rice Inversion Project\\Computational and Applied Mathematics\\Rice University}
\date{SEG Anaheim 2018}

\begin{document}
\frame{
\titlepage
}

\begin{frame}\frametitle{Overview}

Takeaway:  {\color{blue}T}rue {\color{blue}A}mplitude {\color{blue}M}igration = transpose of Born modeling w.r.t. {\em weighted dot products} $\Rightarrow$ rapidly convergent {\color{blue}P}reconditioned {\color{blue}C}onjugate {\color{blue}G}radient iteration for {\color{blue}L}east {\color{blue}S}quares {\color{blue}M}igration

Agenda:
\begin{itemize}
\item Feasible TAM computation 
\item How to formulate accelerated LSM via PCG
\item Necessary conditions for successful PCG
\end{itemize}
\end{frame}

\section{True Amplitude Migration}

\begin{frame}
Varieties of TAM:

\begin{itemize}
\item Generalized Radon Transform / ray-Born / Kirchhoff inversion: Beylkin 85, Bleistein 87,....
\item Scaling methods: Claerbout \& Nichols 94, Nammour \& S. 09, ...
\item {\color{blue}Wave Equation TAM}: Zhang, Bleistein et al. 03, 05,,..., ten Kroode 12, Hou \& S. 15, 17
\end{itemize}

Aim of Wave Equation TAM: achieve inversion similar to GRT with {\em no ray tracing} - only wave equation solves
\end{frame}

\begin{frame}
Acoustic wave system, isotropic point source - bulk modulus = $\kappa(\bx)$, density = $\rho(\bx)$, $w(t)$ = source pulse, $\bx_s$ = source position: pressure = $p(\bx,t;\bx_s)$, particle velocity = $\bv(\bx,t;\bx_s)$ (``source'' fields) satisfy
\[
\rho \frac{\partial \bv}{\partial t} = - \nabla p,\,\,
\frac{1}{\kappa}\frac{\partial p}{\partial t} = - \nabla \cdot \bv + w(t)\delta(\bx-\bx_s)
\]
\[
p, \bv = 0, t<<0
\]

Born system for perturbations $\delta \kappa$, {\color{blue} $\delta \rho = 0$}, $\delta p$, $\delta \bv$:
\[
\rho \frac{\partial \delta \bv}{\partial t} = - \nabla \delta p,\,\,
\frac{1}{\kappa}\frac{\partial \delta p}{\partial t} = - \nabla \cdot \delta \bv + \frac{\delta \kappa}{\kappa^2} \frac{\partial p}{\partial t} 
\]
\[
\delta p, \delta \bv = 0, t << 0
\]

Born modeling op: $F_w \delta \kappa (\bx_r,t; \bx_s) = \delta p(\bx_r,t;\bx_s) $ 
\end{frame}

\begin{frame}
\vspace{-0.5cm}
\begin{center}
%\includegraphics[height=5cm]{Fig/coarse1bulk}\includegraphics[height=5cm]{Fig/equals}
%\includegraphics[height=5cm]{Fig/fine1bulk}\includegraphics[height=5cm]{Fig/equals}
\includegraphics[height=5.5cm]{Fig/fine1bulk}\\
$\kappa_0$ = Marmousi bulk modulus = $\kappa + \delta \kappa$
\end{center}
\vspace{-0.5cm}
\end{frame}

\begin{frame}
\vspace{-0.5cm}
\begin{center}
%\includegraphics[height=5cm]{Fig/coarse1bulkbig}\includegraphics[height=5cm]{Fig/plus}
%\includegraphics[height=5cm]{Fig/fine1bulkbig}\includegraphics[height=5cm]{Fig/plus}
\includegraphics[height=5.5cm]{Fig/fine1bulkbig}\\
$\kappa$ = background = spatial moving average 
\end{center}
\vspace{-0.5cm}
\end{frame}

\begin{frame}
\vspace{-0.5 cm}
\begin{center}
%\includegraphics[height=5.5cm]{Fig/coarse1dbulk}\\
\includegraphics[height=5.5cm]{Fig/fine1dbulk}\\
$\delta \kappa$ = model perturbation = $\kappa_0-\kappa$
\end{center}
\end{frame}

\begin{frame}
%\vspace{-1.0cm}
\begin{center}
%\includegraphics[height=6cm]{Fig/coarse1bornr45}
\includegraphics[height=6cm]{Fig/fine1bornr90}\\
$\delta d = F_w\delta \kappa$ = First order predicted (linearized,
``Born'') free surface shot 90 \\(180 shots @ 48 m, 361 recvrs @ 24 m), $z_s=z_r=12$ m, $w = H * $ [2.5, 5, 25, 30] Hz trapezoidal bandpass
\end{center}
\end{frame}

\begin{frame}
Reverse Time Migration = transpose of Born modeling

RTM of perturbation data $\delta d(\bx_r,t;\bx_s)$:

- compute adjoint (``receiver'') fields $q(\bx,t;\bx_s), \bw(\bx,t;\bx_s)$
\[
\rho \frac{\partial \bw}{\partial t} = - \nabla q,\,\,
\frac{1}{\kappa}\frac{\partial q}{\partial t} = - \nabla \cdot \bw + \sum_{\bx_r} d(\bx_r,t;\bx_s) \delta(\bx-\bx_r)
\]
\[
q, \bw = 0, t >>0
\]

- cross-correlate with zero lag, sum over sources:
\[
F_w^T\delta d(\bx) = -\frac{1}{\kappa(\bx)^2}\sum_{\bx_s} \int\,dt\,\frac{\partial p}{\partial t}(\bx,t;\bx_s)q(\bx,t:\bx_s) 
\]
\end{frame}

\begin{frame}
\vspace{-1.0cm}
\begin{center}
%\includegraphics[height=7cm]{Fig/coarse1mbulk}\\
\includegraphics[height=7cm]{Fig/fine1mbulk}\\
$F_w^T\delta d$ = RTM image of $\delta \kappa$  
\end{center}
\end{frame}

\begin{frame}
\vspace{-0.7cm}
\begin{center}
%\includegraphics[height=5.1cm]{Fig/coarse1dbulk}\\
\includegraphics[height=5.8cm]{Fig/fine1dbulk}\\
$\delta \kappa$ = model perturbation
\end{center}
\end{frame}

\begin{frame}
Observations:

\begin{itemize}
\item +: reflector positions correct (correct velocity)
\item -: low spatial frequency noise (diving waves in source fields)
\item -: {\color{blue} wrong reflector amplitudes - relative and absolute errors}
\end{itemize}

\end{frame}

\begin{frame}
Asymptotic analysis
(Beylkin 85, Bleistein 87, Rakesh 88, Beylkin \& Burridge 89,...) - $w(t)=H(t)$ (practically: integrated bandpass filter), drop from notation:

\[
(I_tF)^TI_tF\delta \kappa(\bx) \approx \int d{\bf k} e^{i {\bf k}\cdot \bx} \frac{(...)}{\cos \theta_r \cos \theta_s} \delta \hat{\kappa}({\bf k})
\]
\begin{itemize}
\item $I_t$ = indef t-integral (Heaviside filter)
\item $\cos \theta_{s,r}$ = wave/ray angle of incidence at source, receiver
\item ``(...)'' = explicit filters,  functions of $\kappa_0(\bx),\rho(\bx)$
\end{itemize}
\end{frame}

\begin{frame}
Get rid of $\cos \theta_{s,r}$ by applying $D_{z_{s,r}}$

Get rid of filters, functions by applying inverse filters, reciprocals

$\Rightarrow$  true/preserved amplitude migration operator $(I_tF)^{\dagger}$ (Zhang \& Bleistein 03, 05,...,ten Kroode 12,...) Hou \& S. 15, 17: factorization
\[
(I_tF)^{\dagger} = W_m^{-1} (I_tF)^T W_d, 
\]
$W_m, W_d$ simple explicit filters: $W_d =  -|f|^{-3} D_{z_s} D_{z_r},\, W_m^{-1} = 32 \rho^{-1}\kappa^3 |k_z|$
\[
(I_tF)^{\dagger}(I_tF) \kappa \propto P \kappa
\]
$P$ = phase space projection onto migration aperture, $f$ = temporal frequency, 
\end{frame}

\begin{frame}
\vspace{-0.62cm}
\begin{center}
%\includegraphics[height=7cm]{Fig/coarse1bulkinvp1it0}\\
\includegraphics[height=6.8cm]{Fig/fine1bulkinvp1it0}\\
$(I_tF)^{\dagger}(I_tF)\delta \kappa$ = TAM
\end{center}
\end{frame}

\begin{frame}
\vspace{-0.5cm}
\begin{center}
%\includegraphics[height=7cm]{Fig/coarse1dbulk1}\\
\includegraphics[height=7cm]{Fig/fine1dbulk}\\
$\delta \kappa$ = model perturbation
\end{center}
\end{frame}

\begin{frame}
Observations:
\begin{itemize}
\item reflector positions still correct
\item amplitudes, spectrum much better balanced 
\item low frequency noise absent
\item {\color{blue} not an approximate inversion - absolute amplitudes too small}
\end{itemize}

Constraints:
\begin{itemize}
\item absorbing surface data: $z_s,z_r=0$
\item DSR hypothesis: rays involved in imaging migration aperture {\em do not turn}
\end{itemize}
\end{frame}

\section{Computing TAM}

\begin{frame}
\begin{itemize}
\item Accessing $z_s,z_r$ derivatives
\item Migrating free surface data
\end{itemize}
\end{frame}

\begin{frame}
Access to $D_{z_s,r}$ via {\em shallow tow depth} hypothesis (Hou \& S. 15): $F^{\rm free}$ = free surface data with $z_s, z_r > 0$,
\[
D_{z_s}D_{z_r}F \approx \frac{1}{4 z_r z_s}F^{\rm free}
\]
Error = $O(z_s^2, z_r^2)$ = same order as FD RTM error if $z_{s,r} = O(\Delta z)$

$D_{z_s}D_{z_r}$ {\em symmetric} $\Rightarrow$
\[
(I_tF)^{\dagger} = -W_m^{-1}(I_tF)^T|f|^{-3}D_{z_s}D_{z_r} = -W_m^{-1}[D_{z_s}D_{z_r}(I_tF)]^T|f|^{-3}
\]
\[
=-\frac{1}{4 z_r z_s}W_m^{-1}(I_tF^{\rm free})^T|f|^{-3}
\]
\end{frame}

\begin{frame}
More work - similar for free surface TAM
\[
(I_tF^{\rm free})^{\dagger} =-\frac{1}{4 z_r z_s}W_m^{-1}(I_tF)^T|f|^{-3}
\]

{\bf NB:} {\em all} examples shown here actually free surface, $F \rightarrow F^{\rm free}$

TAM workflow, free surface data:
\begin{itemize}
\item filter by $|f|^{-3}$
\item absorbing surface RTM with Heaviside source pulse ($I_t$)
\item apply $W_m^{-1}$ - filter by $|k_z|$, scale
\end{itemize}

All steps except RTM - trace-by-trace filters, scaling ops $\Rightarrow$ {\color{blue} negligible expense beyond RTM}
\end{frame}

\begin{frame}
Observations:

\begin{itemize}
\item ghost construction of $(I_tF^{\rm free})^{\dagger}$ only works for tow depth $<<$ wavelength - vulnerable to depth variations, swell noise,...
\item input $D_{z_s}D_{z_r} d$ to RTM $\Leftrightarrow$ use data, source traces as Dirichlet conditions on $z=0$ (Zhang et al. 09, 11, 14,...)
\item other acquisition geometries: 
\begin{itemize}
\item 4C OBCs, shallow tow depth sources: $D_{z_r}p \propto v_z$ 
\item Geostreamer$^{\rm TM}$ - measure $v_z$
\item ...
\end{itemize}
\end{itemize}
\end{frame}

\section{TAM as Preconditioner}

\begin{frame}
Form of approx scaled inverse $(I_tF)^{\dagger} = W_m^{-1}(I_tF)^TW_d$ $\Rightarrow$ can apply
PCG (eg. Golub \& van Loan 2012) - minimizes

\[
J[\delta \kappa] = (I_t (d-F\delta \kappa))^T W_d (I_t(d-F\delta \kappa))
\]

{\bf NB:} $W_d$ plays no explicit role in step!

\begin{itemize}
\item application of {\em model-unweighted} adjoint $(I_tF)^TW_d = -\frac{1}{4 z_r z_s}(I_tF^{\rm free})^T|f|^{-3}$
\item application of $W_m^{-1}$
\item dot products, vector arithmetic
\item similar for $F^{\rm free}$
\end{itemize}
\end{frame}

\begin{frame}
\vspace{-0.62cm}
\begin{center}
\includegraphics[height=6.8cm]{Fig/fine1bulkinvp1it20}\\
$\delta \kappa$ estimate, 20 iterations TAM-preconditioned CG
\end{center}
\end{frame}

\begin{frame}
\vspace{-0.5cm}
\begin{center}
\includegraphics[height=7cm]{Fig/fine1dbulk}\\
$\delta \kappa$ = target model perturbation
\end{center}
\end{frame}

\begin{frame}
\vspace{-0.5cm}
\begin{center}
\includegraphics[height=7cm]{Fig/fine1comp}\\
Mean square error (unweighted!) vs. iteration: Red = CG, Blue = PCG
\end{center}
\end{frame}

\begin{frame}
Observations:
\begin{itemize}
\item $(I_tF)^{\dagger}$, $(I_tF^{\rm free})^{\dagger}$ not approx inverses - differs by slowly varying scale factor $< 1$
\item reason: = zero-offset section of band-limited subsurface offset approximate inverse (ten Kroode 12, Hou \& Symes 15, 17, Chauris \& Cocher 17) 
\item like $t=0$ value of band-limited impulse
\item doesn't matter - preconditioning still compresses spectrum 
\item major issue: must ensure that $W_d$ positive
\end{itemize}
\end{frame}

\begin{frame}
Positivity of $W_d$:  source, receiver slownesses have same sign,
\[
-|f|^{-2}D_{z_r}D_{z_s} = -\frac{ik_{z_s}}{f}\frac{ik_{z_r}}{f} > 0
\]

Sufficient condition (absorbing surface): sources, receivers in non-reflecting near-surface layer (``ocean'') - {\em cannot permit artificial reflectors to develop in layer during iteration}

For sources, receivers in water layer: depth-taper, $\delta \kappa = 0$ near surface in every iteration
\end{frame}

\begin{frame}
\vspace{-0.5cm}
\begin{center}
\includegraphics[height=6cm]{Fig/fine1compnt}\\
Residual curves without depth-taper on $\delta \kappa$ - Red = CG, Blue = PCG. Horizontally propagating noise $\Rightarrow$ artificial reflectors imaged in water $\Rightarrow$ non-positive data weight, leads to divergence of PCG
\end{center}
\end{frame}

\section{Summary, Prospects}

\begin{frame}
``True Amplitude'' RTM can serve as effective preconditioner for least squares migration

Reason: can be written as {\em transpose of $F$ using weighted dot products}

``Ghost trick'' for free surface TARTM $\Rightarrow$ $W_d$ implicit, uses absorbing surface RTM

Essential: $W_d > 0$: sufficient to ensure downgoing source, upcoming receiver wave fields

TARTM differs from approximate inverse by slowly varying amplitude factor - still works as preconditioner!
\end{frame}

\begin{frame}\frametitle{Acknowledgements}
\begin{itemize}
\item A. ten Kroode, J. Hou
\item Shell 
\item sponsors of The Rice Inversion Project
\item Texas Advanced Computing Center, U. of Texas, Austin
\end{itemize}
\end{frame}

\end{document}

### 2019-10-19 repro/papers/madfigs/paper.tex
repro docs dir added

\title{Reproducible Research with Local and Remote Computation - building Madagascar Papers the TRIP way}
\date{}
\author{William W. Symes, The Rice Inversion Project}

\maketitle 
\parskip 12pt 
\begin{abstract}
Madagascar makes publishing reproducible research results relatively straightforward. To my knowledge, no other available framework is as well-adapted to merging large-scale computation with text describing its significance. However, the system is poorly documented: most current documentation consists of oversimplified or incomplete examples, with little guidance about incorporating high-performance computing environments. This paper gives a complete description of the steps necessary to produce a paper via Madagascar that combines the results of local computations on the writer's laptop or desktop computer with those of other computations performed at remote supercomputer centers.
\end{abstract}

\section{Introduction}
\section{Local Computation}
The Viking Graben, or Mobil AVO, data is available on the TRIP data web page. 
I've written a draft paper in {\tt tripbooks/scratch/viking} describing very simple standard processing for this data, using the Radon-demultiple version as a stand-in for primaries-only reflection data. See this paper for a description of the workflow, implemented mostly with SU. Here, I want to emphasize the method for generating and including figures as part of the reproducible research process. The Viking Graben paper and its project directory are a good example: you can build the project from scratch in a few minutes, and carry out the entire procedure described here to obtain a readable paper.

There are four major steps.

\subsection{Generating Figure Files}

Each figure is recorded as a {\tt .vpl} file in the {\tt project/Fig} subdirectory, by evaluation of the {\tt Result} function. A typical example from {\tt viking/project/SConstruct}: 
\begin{verbatim}
Result('parastack','parastack.su','suread endian=0  read=data | put label1=Time label2=Trace unit1=s | grey clip=1.e+7 xinch=10 yinch=5')
\end{verbatim}
Note that the graphics file is in Madagascar graphics format ({\tt .vpl}), created in {\tt project/Fig}. Even though the rest of the workflow is implemented using SU commands, the graphics are Madagascar. {\tt sfsuread} converts the SEGY traces in {\tt parastack.su} into RSF format, {\tt sfput} adds header words for labels and units, and {\tt sfgrey} converts the RSF data to graphics format. As usual, the prefix {\tt sf} may be left off within the command argument for {\tt Result} or {\tt Flow}.

SU commands (and other non-Madagascar commands, such as compiled-in-place IWAVE) must be defined, most reliably by full pathnames. See the {\tt SConstruct} file in {\tt viking/project} for a good way to do this.

Otherwise, only the standard RSF project framework is required ({\tt from rsf.proj import *} at the top of the {\tt SConstruct}) for this part of the process.

You can do this at home: update your copy of {\tt tripbooks/scratch} (and of course Madagascar and SU if you haven't updated them recently) and run {\tt scons} in {\tt viking/project}

\subsection{Archiving the Figures}
Madagascar defines a sort of figure repository - actually, just a directory with a specific subdirectory tree structure in which copies of figure files are stored for future use. You control the location of this directory with the environment variable {\tt RSFFIGS}.

For TRIP projects (papers in subdirs of {\tt tripbooks}, the correct choice is the fully qualified path to {\tt tripbooks/figs}. This is a local (platform-specific) definition - for example, on one of my laptops, {\tt .bashrc} contains the line
\begin{verbatim}
export RSFFIGS=/Users/symes/Documents/tripbooks/figs
\end{verbatim}
and you would do something similar if you use {\tt tcsh}. The directory at this path is a by-product of installing {\tt tripbooks} by {\tt svn import}.

The figures get into the figure ``repository'' by the {\tt scons lock} command, run in the {\tt project} subdirectory.

For example, if you have successfully run {\tt scons} in {\tt viking/project}, then execute {\tt scons lock} to ``lock'' the figure files, i.e. copy them to the RSFFIGS directory.

After you do this, inspect {\tt RSFFIGS}: you will see a {\tt scratch} subdirectory - this would be created, if it did not exist before. In {\tt figs/scratch}, you will see a {\tt viking} subdir, and within that a {\tt project} subdir, containing copies of all the figure files from {\tt tripbooks/scratch/viking/project/Figs}. This is Fomel's device for keeping the figures for each project in a unique place: create a subdirectory path in {\tt RSFFIGS} mimicing the directory path under the ``book'' ({\tt scratch}, in this case) and put the figures there. The construction follows the standard {\tt book/paper/project} structure of Madagascar reproducible research.

You can now build the paper in the {\tt scratch/viking}. You will see the {\tt .vpl} files from the {\tt RSFFIGS} subdir converted to {\tt .pdf} and incorporated in the paper {\tt .pdf} file. The rigid choice of directory structure and environment variables makes all this work: you have to follow {\tt book/paper/project}!

Once again, only the standard RSF reproducible research framework is needed to make all of this work.

\subsection{Archiving the Archive}
The process described so far is documented on the Madagascar web page (though not all in one place, or in as much detail). It works fine if you do all your work in one environment (like your laptop). However, if you use a number of platforms, not so much. Even more of an issue, other people won't be able to build your paper without running all of your code and rebuilding the figures from scratch - not a problem if your project builds in a few seconds, but definitely an obstacle if it runs several hours.

You can get around this problem by using {\tt svn} to archive figure directories, thus breaking the chain of reproduction. The figures are still reproducible - you can rebuild the project - but it's not necessary to rebuild them to make the paper readable.

The basic idea is to commit the subdirectory of {\tt RSFFIGS} containing the figures. There are some subtleties. For example, you must commit just the path that you have added - for example, if {\tt RSFFIGS/scratch} exists (the ``book''), and you are adding {\tt mypaper}, then you should go to {\tt RSFFIGS/scratch}, which contains the not-yet-committed subdirectory {\tt mypaper} created by your use of {\tt scons lock}, {\tt svn add} it, then {\tt svn commit}. It gets more complicated if you are adding or removing figures from an earlier version of a project. Then you need to go to the {\tt RSFFIGS/book/paper/project} directory that you have previously created, and use {\tt svn add} and {\tt svn rm} to make the needed changes (followed by {\tt svn commit} of course!). The easiest way to do this is by a command-line script: after you've locked your new figures, go to the {\tt RSFFIGS/book/paper/project} and 
\begin{verbatim}
foreach i (`ls -1 *.vpl`)
svn add $i
end
\end{verbatim}
(in {\tt tcsh}, something similar in {\tt bash}). You will get complaints about existing figure files, but who cares. Old and unneeded files won't be removed, but again, who cares, they're not very big.

Another approach, perhaps simpler, is to trash your old directory in its entirety by {\tt svn rm} and {\tt svn commit}, generate the new directory using {\tt scons lock}, and archive the whole thing using {\tt svn add} and {\tt svn commit}.

When a project is truly finished, it is civilized behaviour to groom your figure archive.

If you do archive your figures in this way, you'll be able to portably build your paper, and others will be able to build your paper without rebuilding your figures from scratch.

If you want to check the reproducibility of someone else's paper that has been organized as I suggest, then simply {\tt rm -r} or move aside the paper subdirectory in {\tt tripbooks/figs}, and re-run {\tt scons lock} in the project directory. If you want to completely rebuild {\em your} figures, then use {\tt svn rm} and {\tt svn commit} instead, and don't forget to add and commit your new path afterwards.

\subsection{Including Figures in the Paper}
The procedure for including figures in a paper is well-described on the Madagascar web site. As long as the preconditions are defined:
\begin{itemize}
\item {\tt RSFFIGS} defined in environment, points to local {\tt .../tripbooks/figs},
\item project organization according to {\tt book/paper/project},
\item figures locked, and
\item paper LaTeX file built according to Madagascar principles - that is, all macros and extra packages defined in the SConstruct rather than paper, graphics included via {\tt plot} and {\tt multiplot},
\end{itemize}
the paper will build.

One final point: make sure the book builds also before you include your paper in the master copy of the book-level {\tt SConstruct}.

\section{Remote Computation}
Remote computation poses a particular problem: how can you retain the ``build portability'' of the paper compiling process in Madagascar, without making local copies of large binary files, computed remotely, from which the figures are built? Or, alternatively, keeping the figures in the paper directory in pdf form, as many members of our group have done recently - that breaks the reproducibility link between figure and data/code, since a pdf could a priori come from anywhere.

The solution is simple, and uses the same procedure that works locally. Only one step need be changed in the process just outlined to make use of remotely built figures in a reproducible research project. Obviously the paper that results will not be reproducible in the same sense as those relying entirely on local computation, which was really the conceptual model for the Madagascar framework. However it is possible to attain essentially the same result with a bit of care.

There are several equally effective workflows, but the simplest and most in-synch with the Madagascar process is this:

\begin{itemize}
\item log in to your remote platform, {\tt cd} to the disk where you will do the work. Define {\tt DATAPATH} suitably so that large binary files will go to the right place.
\item check out the appropriate book from {\tt \$SVNURL/trip/trunk/tripbooks}. Example:
\begin{verbatim}
svn co $SVNURL/trip/trunk/tripbooks/trip15 trip15
\end{verbatim}
If all of the papers in the book are built according to Madagascar standard, the book consists of a few subdirectories, LaTeX files and scripts, so there's no harm in copying the whole thing. If it's got more stuff in it than that (the current state of affairs with our books), then it will take longer - you can get rid of the papers you're not going to work with, and the pdfs or whatever that the authors cluttered them up with.
\item Now do the same with the TRIP figures directory: go somewhere convenient and
\begin{verbatim}
svn co $SVNURL/trip/trunk/tripbooks/figs $RSFFIGS
\end{verbatim}
Make sure that {\tt RSFFIGS} is defined but does not point to an existing directory before you do this.
\item Now go ahead as you did before, building all of the intermediate results and figures and locking the figures.
\item MOST IMPORTANT: then archive the archive, as above - use {\tt svn} to add and commit the subdirectory of {\tt RSFFIGS}.
\item Now go home and write your paper in the paper directory. When you are ready to compile it, go to your local version of {\tt RSFFIGS} and {\tt svn update} the subdirectory you've committed in the previous step. 
\item When (inevitably) you don't like your first set of figures, or want to add more, just go back to the remote platform, update and add figures in your {\tt project} directory, and when you've got what you want (tested by viewing the {\tt .vpl} files in {\tt project/Fig} remotely, if possible), then lock'em again, and update the archive. If you add figures, then you'll have to add their copies in {\tt RSFFIGS} to the archive - as mentioned before, probably the easiest way to do that is to {\tt svn rm} the paper subdirectory of {\tt RSFFIGS/book}, {\tt svn commit} that change, go back to your project directory, {\tt scons lock} again (which will regenerate the {\tt paper/project} subdirectory of {\tt RSFFIGS/book}, then go back to {\tt RSFFIGS/book}, {\tt svn add} the {\tt paper} directory, then {\tt svn commit}. It sounds complicated, but the idea is straightforward: to ensure that everything's added, trash the repo directory containing your figures, rebuild it locally with {\tt scons lock}, and add it back to the repo.
\end{itemize}

This is essentially what I've done to build the figures for {\tt trip15/iwavext}. Go ahead and try it - go to your copy of {\tt tripbooks/figs}, {\tt svn update}, then go to the nice clean directory {\tt trip15/iwavext}, nothing there but a {\tt tex} file and a few scripts, and build the paper.

Note what has been accomplished here: the figures are saved, but as {\tt .vpl} files in the standard place where Madagascar expects to find figures stored in this format, each a copy via {\tt scons lock} of a local {\tt .vpl} file built as part of the project, via a {\tt Result} command in the project {\tt SConstruct}. So even though the figures do not have to be rebuilt from scratch to compile the paper, they CAN be rebuilt from scratch if you like - all of the code is there to do so. Contrast this with the practice of saving pdfs, which are intrinsically decoupled from the command that builds the input data to the pdf conversion.

### 2019-10-19 repro/papers/mps_framework/paper.tex
repro docs dir added

\title{A Framework for the Representation and Estimation of Seismic Sources via Multipoles}
\date{}
%\lefthead{Huang \& Nammour \& Symes}
\address{
        \footnotemark[1]The Rice Inversion Project,\\ Rice University,\\ Houston, TX
        77005-1892 USA\\
        %\footnotemark[2]Total E\&P R\&T USA
}
\author{Mario J. Bencomo\footnotemark[1] and William Symes\footnotemark[1]}

\righthead{MPS Framework}

\maketitle
\parskip 12pt

%%%%%%%%%
\begin{abstract}
%%%%%%%%%

%Seismic sources are commonly idealized as concentrated at a source point, due
%to their small spatial extent relative to seismic wavelengths.
%The acoustic isotropic point radiator is inadequate as a model of seismic wave
%generation and propagation for seismic sources that are known to exhibit directivity.
%Therefore accurate modeling of seismic wavefields must include energy 
%source (right-hand side) representations generating anisotropic radiation patterns.
%We present here a mathematical and computational framework for representating 
%arbitrarily anisotropic seismic sources as linear combinations of base {\em multipole point-sources}.
%Our framework couples source representation with generic finite-difference
%solvers via multipole discretizations over uniform grids based on discrete moment matching criteria
%as developed by \cite{Walden:1999} and \cite{TorEng:04}. 
%We give a particular instance of our multipole source construction using the C++
%packages \emph{IWave}, framework for finite difference solvers over uniform grids, 
%and \emph{Rice Vector Library} (RVL), library containing classes for expression of
%gradient-based optimization algorithms over Hilbert spaces;\citep{GeoPros:11,RVL_TOMS}.
%Numerical tests in 2D and 3D acoustics demonstrate the accuracy of discretized multipole
%sources coupled with finite-difference solvers for arbitrary source location of varying order. 
%In particular, optimal convergence rates as suggested by spatial approximation order of numerical 
%schemes are obtained when using appropriate source discretizations.

Seismic sources are commonly idealized as concentrated at a source point, though they may exhibit anisotropy in their radiation pattern. 
We present here a mathematical and computational framework for representing arbitrarily anisotropic seismic sources as linear combinations of base multipole point-sources. 
Our framework couples source representation with generic finite difference solvers via multipole discretizations over uniform grids based on discrete moment matching criteria developed in the literature for the delta function. 
We give a particular instance of our multipole source construction using the C++ packages IWave, framework for finite difference solvers over uniform grids, and Rice Vector Library (RVL), library containing classes for expression of gradient-based optimization algorithms over Hilbert spaces.
As proof of concept we present some synthetic inversion examples, in particular a multipole source inversion and a joint source-medium inversion for acoustic wave modeling.

\end{abstract}

%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%

Seismic sources are typically idealized as concentrated at a point, i.e., of point support, due to their small spatial extent relative to the wavelengths of generated seismic waves.
A seismic source is mathematically represented as a forcing term $f$ in some modeling equations, in particular $f$ is most commonly taken to be of the form
\begin{equation}
\label{eq:delta-src}
	f(\mathbf x,t) = w(t)\delta(\mathbf x-\mathbf x^*)
\end{equation}
where $\delta(\mathbf x-\mathbf x^*)$ is the Dirac delta function shifted by $\mathbf x^*$, the spatial coordinates of where the source is localized.
Using the acoustic wave equation,
\begin{equation}
\label{eqn:isorad}
\begin{split}
        \frac{\partial^2}{\partial t^2} p(\mathbf x,t) - c^2\nabla^2
  	p(\mathbf x,t) =& f(\mathbf x,t)\\
	p(\mathbf x,t) = & 0,  \quad t<<0,
\end{split}
\end{equation}
to model the propagation of seismic waves, one observes that source term $f$ as given by \ref{eq:delta-src} results in a pressure field $p$ that is spherically symmetric\footnote{Assuming wave speed $c$ is constant, and in an unbounded medium.}; the solution for the isotropic point radiator problem \ref{eqn:isorad} in Euclidean 3-space \cite[]{CourHil:62} is given by
\begin{equation}
\label{eq:green3d}
	p(\mathbf x,t) = \frac{w\left(t-\frac{r}{c}\right)}{4\pi c^2 r},
	\quad r = \sqrt{\mathbf x^T \mathbf x}.
\end{equation}
In this context, $f$ given by \ref{eq:delta-src} is referred to as an {\em isotropic source}, that is, a source with isotropic response. 
Both active and earthquake seismic sources, however, generate spatially asymmetric wavefields (see for example \cite{Shearer:2009}, \cite{Yil:01}), and thus an isotropic source representation is inadequate.

Generic acoustic sources can be approximated by a truncated series of {\em multipole point-sources}, or simply multipoles, \cite[]{SantosaSymes:00},
\begin{equation}
\label{eq:MPSappx}
	f(\mathbf x,t) \approx \sum_{|\mathbf s|\le N} w_{\mathbf s}(t) D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)
\end{equation}
in which we have introduced {\em multi-index notation}: for spatial dimension $d$ and multi-index (integer $d$-tuple) ${\bf s} = (s_1,...,s_d)$, the $\mathbf s$-mixed partial derivative operator denoted $D^{\bf s}$
and its (total) order $|\mathbf s|$ are
\begin{equation}\label{eq:PDO}
        D^{\mathbf s} = \prod_{k=1}^d \left( \frac{\partial}{\partial x_k}\right)^{s_k}, \quad |\mathbf s| = \sum_{k=1}^{d} s_k.
\end{equation}
Approximation \ref{eq:MPSappx} is motivated by Peetre's Theorem \cite[]{Horm:69}, where any function concentrated entirely at a point in space (of {\em point support}) is a multipole of finite order; multipole \ref{eq:MPSappx} is said to be of order $N$.
The number of terms required in the series \ref{eq:MPSappx} is dependent on how valid the point-source assumption is, source anisotropy, and the given error tolerance.

%The coefficient time functions $w_{\mathbf s}(t)$ may be scalar-, vector-, or
%tensor-valued, according to the nature of quantity being updated
%(pressure, velocity, or stress) in the equation in which
%$f(\mathbf x,t)$ appears as right-hand side.  
%Multipoles may approximate arbitrary sources highly localized on the wavelength 
%scale, in the sense of generating approximately the same field away from the
%source location, and for this reason have enjoyed widespread use in
%seismic source representation \cite[]{Shearer:2009}.

This paper describes an implementation of the multipole representation \ref{eq:MPSappx} with
\begin{itemize}
	\item natural input, namely the time functions $\{w_{\mathbf s}\}$, 
	\item systematic discrete approximations to singular sources of the form \ref{eq:MPSappx} that preserve the convergence order of finite difference solutions to wave equations, resulting in
	\item natural output, namely time series attached to points on a regular grid that can be fed as source functions to a finite-difference wave equation solver, and
	\item a flexible object-oriented implementation that follows closely the structure of the underlying mathematics.
\end{itemize}
To achieve the second goal, we use the approach of {\em matching discrete moments} pioneered by \cite{Walden:1999} to attain such singular source approximations over uniform grids, regardless of the actual location $\mathbf x^*$ (on- or off-grid).

%%FIX THIS
%We refer to our previous paper, \cite{MPS_theory}, for a more detailed account of the singular source approximation as implemented here.
%multipole source discretizations where
%we give an overview of the theory and present numerical results demonstrating
%point-wise convergence at optimal rates away from the source.

We have realized the multipole source construction explained below using the C++ packages \emph{IWave} and \emph{Rice Vector Library} (RVL); \citep{GeoPros:11,RVL_TOMS}.  
The IWave package is a framework for finite difference solvers over uniform grids while the RVL package provides a system of classes for expression of gradient-based optimization algorithms over Hilbert spaces.  
IWave and RVL come together to form a modeling engine for seismic inversion and migration. 
Implementation of multipole sources as RVL objects enables straightforward composition with IWave solvers and inclusion in inversion algorithms powered by RVL optimization code. 
Any other wave equation solver wrapped in the appropriate RVL interfaces could be coupled to the multipole source objects in the same way.

We describe the multipole source representation in detail in the first section, and review the mechanism of high-order convergence for the approximation/discretization of singular source-terms.  
The second section summarizes IWave and RVL, and describes the implementation of the multipole source framework based on these packages and on the Madagascar and Seismic Unix open source libraries.  
A number of reproducible tests accompanying this paper are discussed in the third section, demonstrating the structure of the implementation and the use of its principal components.
In particular, we give explicit examples of some simple forward modeling and inversion code with multipole sources.

In this paper we do not deal with several important issues, for example the number of multipole terms required to represent a source with a given degree of anisotropy, and the degree of anisotropy required in successful inversion of field data.  
These topics are mostly outside the scope of our paper, but we mention them in the discussion section.
We also avoid some of the mathematical challenges associated with the ill-posedness of the multipole source inverse problem, though we discuss the possibility of preconditioning the source-to-data map for more rapid convergence of iterative algorithms, a matter we will treat fully in another place.

%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory}
%%%%%%%%%%%%%%%%%%%%%%%%%

The localized sources modeled in this paper occur as right-hand sides in the first order forms of time domain linear acoustics and elasticity:\\
{\em (velocity-pressure form of acoustic equations)}
\begin{equation}
\label{eq:acous}
\begin{split}
		\frac{\partial}{\partial t} p(\mathbf x,t) + \kappa(\mathbf x)\frac{\partial}{\partial x_j}v_j(\mathbf x,t) &= f(\mathbf x,t),\\
                \frac{\partial}{\partial t} v_i(\mathbf x,t) + \beta(\mathbf x)\frac{\partial}{\partial x_i}p(\mathbf x,t) &= f_i(\mathbf x,t),
\end{split}
\end{equation}
\emph{(velocity-stress form of linear elasticity equations)}
\begin{equation}\label{eq:elas}
\begin{split}
		\frac{\partial}{\partial t} v_i(\mathbf x,t) - \beta(\mathbf x) \frac{\partial}{\partial x_j} \sigma_{ij}(\mathbf x,t) &= f_i(\mathbf x,t), \\
		\frac{\partial}{\partial t} \sigma_{ij}(\mathbf x,t) - c_{ijmn}(\mathbf x) \frac{\partial}{\partial x_n} v_m(\mathbf x,t) &= f_{ij}(\mathbf x,t),
\end{split}
\end{equation}
as well as the second order (constant density) acoustic wave equation:
\begin{equation}
\label{eq:wave_acous}
		\frac{\partial^2}{\partial t^2} p(\mathbf x,t) - c(\mathbf x)^2 \nabla^2 p(\mathbf x,t) = f(\mathbf x,t),
\end{equation}
where
\begin{singlespace}
\begin{itemize}
	\item $\beta =$ buoyancy (reciprocal of density), 
	\item $\kappa =$ bulk modulus,
	\item $c = \sqrt{\kappa\beta} = $ speed of sound, also referred to as the (wave) velocity,
	\item $c_{ijmn} = $ $ijmn$-component of fourth-order Hooke's tensor $\bf C$,
	\item $p = $ pressure field, 
	\item $v_i =$ $i^{th}$-component of particle-velocity vector $\bf v$,
	\item $\sigma_{ij} = $ $ij$-component of second-order stress tensor,
	\item $f =$ scalar pressure source = defect in acoustic
          constitutive law,
	\item $f_i = $ $i^{th}$-component of velocity
          source vector $\bf f=$ body force,
	\item $f_{ij} = $ $ij$-component of second-order
          tensor stress source $\bf F=$ defect in elastic constitutive law.
\end{itemize}
\end{singlespace}
Indices $i,j,m,n$ run from $1,...,d$, where $d=$ space dimension. Einstein
summation convention is assumed, that is, repeated indices are summed.
Scalar, vector, and second-order tensor right-hand sides occur in
these equations, representing energy sources of several types.
There is some redundancy amongst these possibilities: for example a
spatially localized
constitutive law defect (pressure or stress source) may be exchanged
for an equivalent body force that generates the same acoustic or
elastic field away from the source location. Because of these
trade-off possibilities, we will tacitly assume that only one of the
three types of source field - scalar, vector, or tensor - is present at a time.

%%%%%%
\subsection{Multipole Representation: MPS Spaces}

The scalar multipole of order $N$ concentrated at $\mathbf x^*\in\mathbb R^d$ has already been introduced in equation \ref{eq:MPSappx}.
Expressions for vector and tensor multipoles of order $N$ are similar:
\begin{equation}\label{eq:MPSappxVec}
	f_i(\mathbf x,t) = \sum_{|\mathbf s| \le N} w_{i;\mathbf s}(t) \; D^{\mathbf s} \delta(\mathbf x- \mathbf x^*),
\end{equation}
\begin{equation}\label{eq:MPSappxTen}
	f_{ij}(\mathbf x,t) = \sum_{|\mathbf s| \le N} w_{ij;\mathbf s}(t) \; D^{\mathbf s} \delta(\mathbf x- \mathbf x^*).
\end{equation}

Evidently, multipoles of each type and order form vector spaces. Subspaces of
these spaces are determined by linear constraints (for example, tensor
symmetry). To describe vector spaces of multipoles conveniently, we introduce the concept of
{\em multipole bases} $\{b_i\}$, $\{\mathbf b_i\}$, and $\{\mathbf
B_i\}$. These are simply linearly independent sets of multipoles, whose
linear combinations with scalar time function coefficients $\{w_i\}$, called the
{\em multipole coefficients}, span various vector spaces of multipoles:
\begin{equation}\label{eq:MPS}
\begin{split}
        f(\mathbf x,t) &= \sum_{i} w_i(t) b_i(\mathbf x),\\
        \mathbf f(\mathbf x,t) &= \sum_{i} w_i(t) \mathbf b_i(\mathbf x),\\
        \mathbf F(\mathbf x,t) &= \sum_{i} w_i(t) \mathbf B_i(\mathbf x).
\end{split}
\end{equation}
Each basis element combines a scalar, vector, or tensor
coefficient $c_{ij},{\bf c}_{ij}$ or ${\bf C}_{ij}$, and mixed partial
derivatives of the spatial delta function, parametrized by a multi-indices ${\bf
  s}_{ij}$:
\begin{equation}\label{eq:realdefn}
\begin{split}
	b_i(\bf x) = & \sum_j c_{ij} D^{{\bf s}_{ij}}\delta({\bf x}-{\bf x^*}),  \\
	{\bf b}_i(\bf x) =& \sum_j {\bf c}_{ij} D^{{\bf s}_{ij}}\delta({\bf x}-{\bf x^*}), \\
	{\bf B}_i(\bf x) =& \sum_j {\bf C}_{ij} D^{{\bf s}_{ij}}\delta({\bf x}-{\bf x^*}). 
\end{split}
\end{equation}
Many choices of combinations $c_{ij},{\bf s}_{ij}$ (and so on) are
possible - we show some of these possible choices below. We will assume in the sequel
that such a choice has been made.

For example, the scalar multipole basis of order zero centered 
at $\mathbf x^*$ consists of one element, $b_1$, given by
\[
        b_1(\mathbf x) = \delta(\mathbf x-\mathbf x^*).
\]
A maximal choice of scalar multipole basis of order one (in 3-D) centered at $\mathbf x^*$ is
\[
        b_1(\mathbf x) = \delta(\mathbf x - \mathbf x^*), \quad 
        b_2(\mathbf x) = \frac{\partial}{\partial x_1}\delta(\mathbf x-\mathbf x^*), \quad
        b_3(\mathbf x) = \frac{\partial}{\partial x_2}\delta(\mathbf x-\mathbf x^*), \quad
        b_4(\mathbf x) = \frac{\partial}{\partial x_3}\delta(\mathbf x-\mathbf x^*).
\]
A vector multipole basis for the space of order one (in 3-D) centered at $\mathbf x^*$ is
\[
        \mathbf b_1(\mathbf x) = \delta(\mathbf x-\mathbf x^*) \vet{1\\ 0\\0}, \quad 
        \mathbf b_2(\mathbf x) = \delta(\mathbf x-\mathbf x^*) \vet{0\\ 1\\0}, \quad 
        \mathbf b_3(\mathbf x) = \delta(\mathbf x-\mathbf x^*) \vet{0\\ 0\\1}.
\]
A maximal tensor multipole basis of order zero (in 3D)  centered at $\mathbf x^*$ is
\[
        \mathbf B_1(\mathbf x) = \delta(\mathbf x-\mathbf x^*) \vet{1,0,0\\ 0,0,0\\0,0,0}, \quad 
        \mathbf B_2(\mathbf x) = \delta(\mathbf x-\mathbf x^*) \vet{0,1,0\\ 0,0,0\\0,0,0}, \quad 
        \mathbf B_3(\mathbf x) = \delta(\mathbf x-\mathbf x^*) \vet{0,0,1\\ 0,0,0\\0,0,0}.
\]
\[
        \mathbf B_4(\mathbf x) = \delta(\mathbf x-\mathbf x^*) \vet{0,0,0\\ 1,0,0\\0,0,0}, \quad 
        \mathbf B_5(\mathbf x) = \delta(\mathbf x-\mathbf x^*) \vet{0,0,0\\ 0,1,0\\0,0,0}, \quad 
        \mathbf B_6(\mathbf x) = \delta(\mathbf x-\mathbf x^*) \vet{0,0,0\\ 0,0,1\\0,0,0}.
\]
\[
        \mathbf B_7(\mathbf x) = \delta(\mathbf x-\mathbf x^*) \vet{0,0,0\\ 0,0,0\\1,0,0}, \quad 
        \mathbf B_8(\mathbf x) = \delta(\mathbf x-\mathbf x^*) \vet{0,0,0\\ 0,0,0\\0,1,0}, \quad 
        \mathbf B_9(\mathbf x) = \delta(\mathbf x-\mathbf x^*) \vet{0,0,0\\ 0,0,0\\0,0,1}.
\]

Some applications may require constraints on the source representation for physical reasons or assumptions on the radiation structure of the source. 
For example, in the study of earthquake source mechanisms, sources are commonly represented by the seismic moment tensor, which in the terminology introduced here is precisely a tensor multipole source of order $N=0$.  
Earthquake sources must preserve angular momentum, so the moment tensor source must be symmetric: $f_{ij} = f_{ji},$ for all $i,j=1,2,3$.
This symmetry constraint suggests another tensor multipole basis for this corresponding multipole subspace,
\[
        \mathbf B_1(\mathbf x) = \delta(\mathbf x -\mathbf x^*) \mat{1,0,0\\0,0,0\\0,0,0}, \quad
        \mathbf B_2(\mathbf x) = \delta(\mathbf x -\mathbf x^*) \mat{0,0,0\\0,1,0\\0,0,0}, \quad
        \mathbf B_3(\mathbf x) = \delta(\mathbf x -\mathbf x^*) \mat{0,0,0\\0,0,0\\0,0,1},
\]
\[
        \mathbf B_4(\mathbf x) = \delta(\mathbf x -\mathbf x^*) \mat{0,1,0\\1,0,0\\0,0,0}, \quad
        \mathbf B_5(\mathbf x) = \delta(\mathbf x -\mathbf x^*) \mat{0,0,1\\0,0,0\\1,0,0}, \quad
        \mathbf B_6(\mathbf x) = \delta(\mathbf x -\mathbf x^*) \mat{0,0,0\\0,0,1\\0,1,0}.
\]

The multipole expansions \ref{eq:MPS} are determined by 
\begin{itemize}
	\item the source location ${\bf x}^*$,
	\item the choices of multipole basis functions $\{b_i\}$ 
		(for the scalar case, similar for vector and tensor - in this discussion, we give results explicitly for the scalar case), and
	\item the coefficient time functions $\{w_i\}$.
\end{itemize}
We collect the multipole coefficient time series together as the {\em multipole coefficient vectors} $\mathbf w = \{w_i: i=1,2,...\}$, and endow them with the obvious definition of linear combination: 
for multipole coefficient vectors $\mathbf w^{1}$ and $\mathbf w^{2}$ and scalars $a_1, a_2\in\mathbb R$, 
\begin{equation} \label{eq:linstruct}
	a_1 \mathbf w^{1}(t) + a_2 \mathbf w^{2}(t) 
	= \{a_1 w_{i}^{1}(t) + a_2  w_{i}^{2}(t) : i=1,2,...\}.
\end{equation}
With definition \ref{eq:linstruct}, the set of multipole coefficient vectors becomes a vector space, a {\em multipole source (MPS) space}. 
We regard the other two identifying items listed above (source location, choice of multipole basis) as attributes of an MPS space as well.

\noindent {\bf Remark:} While not important for the our current purposes, we should mention that norms and inner (``dot'') products can also be implemented for MPS spaces. 
For example, the obvious (Euclidean or $L^2$) inner product is
\begin{equation} \label{eq:euclid}
	\langle \mathbf w^{1}, \mathbf w^{2}\rangle = \sum_i \int w^{1}_i(t)\; w^{2}_i(t) \;dt,
\end{equation}
however, many other inner products are possible. 
As we shall show elsewhere, there is a better choice of inner product than that given in equation \ref{eq:euclid}, in that it improves the condition number of the linear map from multipole coefficient functions to data traces, hence accelerates the convergence of iterative inversion algorithms.

Equation \ref{eq:MPS} defines a linear mapping $\mathbf L$ from corresponding MPS space to the distributions in physical space: in the scalar case,
\begin{equation} \label{eq:mpstorhsdefn}
	(\mathbf L\mathbf w)({\bf x},t)= \sum_{i} w_i(t) b_i(\mathbf x).
\end{equation}
We call $\mathbf L$ the {\em MPS-to-RHS} map, since $\mathbf L\mathbf w$ may be used as a right-hand side source-term ($f$ or ${\bf f}$ in the system \ref{eq:acous}, ${\bf f}$ or ${\bf F}$ in the system \ref{eq:elas}).

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Discretization of Multipoles}

Let the collection of points associated with a Cartesian grid centered at $\mathbf x^0\in\mathbb R^d$ and with grid size $h>0$ be denoted by
\[
	\mathcal G(\mathbf x^0,h) = \{ \mathbf x^{\mathbf n}\in\mathbb R^d : 
		\mathbf x^{\mathbf n}=(x_{1}^{n_1},...,x_{d}^{n_d}), \; \mathbf n\in\mathbb Z^d\}
\] 
where
\[
	x_{k}^{n_k} = x_{k}^0 + hn_k, \quad \forall k=1,...,d.
\]
For simplicity of presentation we continue to work with Cartesian grids hereon, though generalizations to regular grids are possible.
We will denote grid functions and other grid-dependent quantities with an $h$ subscript.

We obtain a discrete analog, the {\em discrete multipole representation}, by replacing $D^{{\bf s}_{ij}}\delta({\bf x} - {\bf x}^*)$ in equations \ref{eq:realdefn} with functions $\eta_h^{ij}(\mathbf x;\mathbf x^*)$ defined over a given grid, i.e., 
\[
	\eta_{h}^{ij}(\cdot;\mathbf x^*) : \mathcal G(\mathbf x^0,h) \to \mathbb R,
\]
for each $i,j$:
\begin{equation}
\label{eq:MPSdiscrete}
\begin{split}
        f_h(\mathbf x,t) &= \sum_{i} w_i(t) b_{h,i}(\mathbf x),\\
        \mathbf f_h(\mathbf x,t) &= \sum_{i} w_i(t) \mathbf b_{h,i}(\mathbf x)\\
        \mathbf F_h(\mathbf x,t) &= \sum_{i} w_i(t) \mathbf B_{h,i}(\mathbf x),
\end{split}
\end{equation}
with
\begin{equation}
\label{eq:realdefndiscrete}
\begin{split}
	b_{h,i}(\mathbf x) & = \sum_j c_{ij} \eta_{h}^{ij}(\mathbf x;\mathbf x^*)  \\
	\mathbf b_{h,i}(\mathbf x) & = \sum_j {\bf c}_{ij} \eta_{h}^{ij}(\mathbf x;\mathbf x^*) \\
	\mathbf B_{h,i}(\mathbf x) & = \sum_j {\bf C}_{ij} \eta_{h}^{ij}(\mathbf x;\mathbf x^*).
\end{split}
\end{equation}
Note that we parameterize grid functions $\eta_h^{ij}$ by $\mathbf x^*\in\mathbb R^d$ given their dependence on source location as discussed below.

Grid functions $\eta_h^{ij}(\mathbf x;\mathbf x^*)$ correspond to discretizations of singular sources terms $D^{\mathbf s_{ij}}\delta(\mathbf x-\mathbf x^*)$ via the {\em discrete moment conditions} pioneered by \cite{Walden:1999}.
In our previous paper, \cite{MPS_theory}, we give a detailed account of the theory and our implementation of the singular source approximation.
Our main contribution in that paper is a novel error analysis for the discretization of multipole sources in wave propagation problems.
In particular, we prove a weak error estimate that demonstrates convergence of numerical solutions.
Strong convergence (point-wise in space) at optimal rates is however observed, away from the support of source approximation, in the reported numerical experiments.
We briefly discuss some aspects of the singular source approximation here.
Our primary focus will be to present the construction of singular source approximations, that is the $\eta_{h}^{ij}$ grid functions.

In the simpler 1-D case, 
we say that $\eta_h(x;x^*)$ is $q$-order approximation of $\tfrac{d^s}{dx^s}\delta(x-x^*)$ if the grid function satisfies \begin{equation}
\label{eq:momcond}
       h \sum_{x\in\mathcal G(x^0,h)} \eta_h(x;x^*) (x-x^*)^{\alpha} = 
       \int \tfrac{d^s}{dx^s}\delta(x-x^*)(x-x^*)^{\alpha} \; dx,
       \quad\quad \forall \alpha =0,1,... q+{\bf s}-1,
\end{equation}
also known as the {\em discrete $(q,s)$-moment conditions at $x^*$}.
In practice we construct grid functions $\eta_{h}(x;x^*)$ of compact support proportional to grid size $h$.
In particular, we assume the grid function is defined to be zero outside the interval $[x^*-\tfrac{N}{2},x^*+\tfrac{N}{2})$ for $N=q+s$.
We denote the $N$ grid points contained within the interval $[x^*-\tfrac{N}{2},x^*+\tfrac{N}{2})$ by $\{\tilde x_\ell\}_{\ell=1}^N$, assuming $\tilde x_1 < \tilde x_2< \cdots < \tilde x_N$.
The discrete $(q,s)$-moment conditions thus result in a $(q+s)\times(q+s)$ system of equations for $\eta_h(\tilde x_\ell;x^*)$,
\[
	\mathbf A \mathbf d = \mathbf b
\]
with
\[
	\{A\}_{kl} = (\tilde x_\ell - x^*)^{k-1}, \quad
	\{d\}_\ell = \eta_{h}(\tilde x_\ell;x^*), \quad
	\{b\}_k = \frac{s!(-1)^{s}}{h} \delta_{s,k-1}
\]
for $\ell,k=1,2,...,q+s$.
Note that $\mathbf A$ is a Vandermonde matrix of full rank and is guaranteed to have a unique solution.
In fact, using the Vandermonde matrix inverse formula, we have
\begin{equation}\label{eq:eta}
	 \eta_h(\tilde x_\ell;x^*) = \left\{ \begin{array}{cl}
		\displaystyle
		s! (-1)^{N-1} \; 
		\frac{
			\displaystyle
			\sum_{\underset{m_1,...,m_{q-1}\neq \ell}{1\le m_1<\cdots<m_{q-1}\le N}}
			 ( \tilde x_{m_1} - x^*) \cdots ( \tilde x_{m_{q-1}} -x^*)
		}
		{
			\displaystyle
			h^{N}\prod_{\underset{m\neq \ell}{1\le m\le N}} ( \ell - m)
		}
		,& \text{for} \; q>1  \vspace{10pt} \\ 
		\displaystyle
		\frac{
			s! (-1)^{s}
		}
		{
			\displaystyle
			h^{N} \prod_{\underset{m\neq \ell}{1\le m\le N}} (\ell - m)
		}
		,& \text{for}\; q=1.
	\end{array}\right.
\end{equation}

Higher dimensional approximations are generated by taking tensor products of 1-D approximations. 
Namely $\eta_{h}:\mathcal G(\mathbf x^0,h) \to \mathbb R$, an approximation to $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$, is given by
\[
	\eta_h(\mathbf x;\mathbf x^*) = \prod_{k=1}^d \eta^k_h(x_k;x^*_k)
\]
where $\eta^k_h(x_k;x^*_k)$ satisfies the discrete $(q,s_k)$-moment conditions at $x^*_k$.

Discretization of multipole sources produce gridded approximations \ref{eq:MPSdiscrete}, where the formula may be viewed as defining a discrete version of the MPS-to-RHS map (equation \ref{eq:mpstorhsdefn}):
\begin{equation} \label{eq:mpstorhsdefndiscrete}
	(\mathbf L_h \mathbf w)(\mathbf x,t) = \sum_{i} w_i(t) b_{h,i}(\mathbf x),
\end{equation}
over grid points $\mathbf x\in\mathcal G(\mathbf x_0,h)$ as before.
We call $\mathbf L_h$ the {\em discrete MPS-to-RHS} map. 
Its output (vector in range) may be used as the right-hand side in a finite difference approximation to acoustodynamics (equations \ref{eq:acous}, \ref{eq:wave_acous}) or elastodynamics (equations \ref{eq:elas}). 
The domain of this map is a discrete version of the MPS space described in the preceding section.

\noindent {\bf Remark:} Throughout the mathematical development, we will continue to regard time as continuous, and write $w_i(t)$ for the multipole coefficients. 
Of course, in any implementation time will be discretely sampled as well.

The range of $\mathbf L_h$ consists of the discrete functions of space and time, limited in space to those grid points that actually appear in the discrete multipole basis functions (equation \ref{eq:realdefndiscrete}). 
Thus the information that characterizes the range of the discrete MPS-to-RHS map is
(i) the locations of the grid points required to express the moment-consistent discrete multipoles listed in equation \ref{eq:realdefndiscrete}, these locations depending on the choice of discrete multipole basis as in \ref{eq:MPSdiscrete}, 
and (ii) the time sampling information (time grid) which for sake of simplicity we assume is the same for the right-hand side and for all of the multipole coefficients.

%
%A multipole representation of seismic sources poses an interesting mathematical challenge for the numerical solution of hyperbolic PDEs such as \ref{eq:acous}, \ref{eq:elas}, and \ref{eq:wave_acous}.
%Namely, how should one discretize such singular source terms? 
%Subsequently, how does the discretization affect the accuracy of numerical PDE methods used to solve such problems?
%For numerical schemes based on variational forms, that is Galerkin methods, the discretization of $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$ is rather straightforward, one simply applies the distribution definition of such singularities;
%\begin{equation}
%\label{eq:distribution}
%	\int D^{\mathbf s}\delta(\mathbf x-\mathbf x^*) \phi(\mathbf x)\; d\mathbf x = (-1)^{|\mathbf s|} D^{\mathbf s}\phi(\mathbf x)|_{\mathbf x=\mathbf x^*}
%\end{equation}
%for any smooth test function $\phi$.
%In this work, however, we focus on regular (rectangular) {\em staggered-grid} finite difference methods, which are widely used for basin- and exploration-scale modeling (see \cite{moczoetal:06} for an excellent overview and many older references). 
%Such methods pose an immediate problem for singular source models such as multipole sources: finite difference algorithms ``know'' only gridded fields, so a source located at an arbitrary point $\mathbf x^*$ in space must be represented somehow by virtual sources at nearby grid points. 
%This task is complicated by the nature of the field, as is evident for instance from inspection of the 3-D analytical solution \ref{eq:green3d}.
%Similarly, solutions of the wave equations \ref{eq:acous}, \ref{eq:elas}, and \ref{eq:wave_acous} with singular right hand sides are generally themselves singular at the source point, so the Taylor-series based analysis of finite difference accuracy does not apply.
%%Imitating finite-element singular source representation, for example
%%by using adjoint interpolation, may reduce the accuracy of the modeled
%%field, in the sense of convergence order: visually, large errors can
%%pollute the field near the source, and propagate throughout the
%%simulation domain. 
%
%\cite{Walden:1999} developed approximations to singular source terms in a finite difference and finite element setting for the 1D Helmholtz equation for which his analysis and numerical examples demonstrated point-wise convergence of numerical solutions with optimal convergence rates.
%The key idea was for approximations to satisfy a set of {\em discrete moment conditions} depending on the derivative and desired approximation order, a generalization of previous work for discretizations of the Dirac delta function in the context of the immersed boundary method \citep{Pes:02}.
%
%%\cite{TorEng:04} refined these ideas and provided further error analysis, 
%%primarily for finite difference solutions to simple ODEs, and parabolic and elliptic PDEs.
%%Recently, \cite{hoss:16} presented a unified theory for the approximation, or 
%%regularizations, of the delta function independent of numerical scheme and their
%%application to prototypical elliptic and hyperbolic PDEs.
%%Work by \cite{Petersson:2016} presented even more relevant analysis, in particular  
%%singular source discretizations as related to central difference approximations of the 
%%1-D advection equation.
%%They demonstrated that the discrete moment conditions given by \cite{Walden:1999},
%%along with smoothness conditions, were sufficient for convergence of finite difference
%%solutions, though their analysis proved suboptimal convergence results.
%
%Other works have provided further error analysis and theory to understand the coupling between source approximations and PDE discretizations; see for example \cite{TorEng:04}, \cite{hoss:16}, \cite{Petersson:2016}.
%However, the singular source approximation theory presented in the current literature does not provide the relevant error analysis for staggered-grid finite difference solutions to linear acoustics or elasticity (first order form) with multipole sources.
%In our previous paper, \cite{MPS_theory}, we present a more relevant error analysis: we prove a weak error estimate that demonstrates convergence of numerical solutions.
%Strong convergence (point-wise in space) at optimal rates is however observed, away from the support of source approximation, in the reported numerical experiments.
%
%%FIX THIS
%In the following we give a short overview of the theory developed by \cite{Walden:1999} and \cite{MPS_theory}. 
%Our primary focus will be to present the construction of singular source approximations for multipoles with arbitrary source location $\mathbf x^*$.
%
%Recall one way to view truncation order of a finite difference scheme is via the action of the difference operator on polynomials: the (truncation) order is the largest integer $p$ for which the difference operator, applied to a polynomial of degree $p$, gives the same result at grid points as the differential operator it is intended to approximate. 
%\cite{Walden:1999} applies a version of this idea for constructing singular source approximations. 
%We first define the (continuum) {\em moment} of a function: for the general $d$-dimensional case and $d-$tuple multi-index $\bs \alpha =(\alpha_1,...,\alpha_d)$, the {\em $\bs \alpha$-moment centered at $\mathbf x^*$} of some $L^1$ function with compact support, $\eta$, is defined by
%\begin{equation}
%\label{eq:cont_moment}
%	M^{\bs \alpha}(\eta,\mathbf x^*):= \int_{\mathbb R^d} \eta(\mathbf x)\;(\mathbf x-\mathbf x^*)^{\bs \alpha} \; d\mathbf x,
%\end{equation}
%where the monomial term $({\bf x}-{\bf x}^*)^{\bs \alpha}$ is interpreted as the a product of monomials in each dimension, i.e.,
%\[
%        ({\bf x}-{\bf x}^*)^{\bs \alpha} = \prod_{k=1}^d (x_k-x_k^*)^{\alpha_k}.
%\]
%Using \ref{eq:distribution}, we can interpret the $\bs \alpha$-moment of $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$ as
%\[
%	M^{\bs \alpha}(D^{\mathbf s}\delta(\cdot-\mathbf x^*),\mathbf x^*) = \left\{\begin{array}{rl}
%								0, & \bs \alpha\neq \mathbf s, \\
%								(-1)^{|\mathbf s|}{\mathbf s !}, & \bs \alpha=\mathbf s.
%								\end{array}\right.
%\]
%
%In a similar fashion, we define the {\em discrete moment} by replacing the integration in \ref{eq:cont_moment} with summation.
%First, let the collection of points associated with a Cartesian grid centered at $\mathbf x^0\in\mathbb R^d$ and with grid size $h>0$ be denoted by
%\[
%	\mathcal G(\mathbf x^0,h) = \{ \mathbf x^{\mathbf n}\in\mathbb R^d : 
%		\mathbf x^{\mathbf n}=(x_{1}^{n_1},...,x_{d}^{n_d}), \; \mathbf n\in\mathbb Z^d\}
%\] 
%where
%\[
%	x_{k}^{n_k} = x_{k}^0 + hn_k, \quad \forall k=1,...,d.
%\]
%For simplicity of presentation we continue to work with Cartesian grids hereon, though generalizations to regular grids are possible.
%We will denote grid functions and other grid-dependent quantities with an $h$ subscript.
%
%The {\em discrete $\bs \alpha$-moment centered at $\mathbf x^*$} ($\mathbf x^*$ need not coincide with a grid point) of a grid function $\eta_h:\mathcal G(\mathbf x_0,h)\to\mathbb R$ is given by
%\[
%	M^{\bs \alpha}_h(\eta_h,\mathbf x^*) := h^d \sum_{\mathbf x \in\mathcal G(\mathbf x_0,h)} \eta_h(\mathbf x) \; (\mathbf x-\mathbf x^*)^{\bs \alpha}.
%\] 
%Given integer $q>0$ and multi-index $\mathbf s$, grid function $\eta_h$ is said to satisfy the {\em discrete $(q,\mathbf s)$-moment conditions at $\mathbf x^*$} if
%\begin{equation}
%\label{eq:momcond}
%        M_h^{\bs \alpha}(\eta_h,\mathbf x^*) = M^{\bs \alpha}(D^{\bf s}\delta( \cdot-\mathbf x^*),\mathbf x^*), \quad \forall |\bs \alpha|\le q+|{\bf s}|-1.
%\end{equation}
%
%In practice we construct approximations to $D^{\mathbf s}\delta(\mathbf x)$ with grid functions of compact support proportional to grid size $h$. 
%The following theorem \citep{MPS_theory} gives error estimates for such approximations:
%
%\begin{theorem}\label{thm:discweakconv}
%Let nonnegative integer $q$, multi-index $\mathbf s$, and $\mathbf x^*\in\mathbb R^d$ be given.
%Suppose $\{\eta_h^\epsilon(\cdot;\mathbf x^*): \epsilon>0\}$ is a sequence of grid functions $\eta_h^\epsilon(\cdot;\mathbf x^*):\mathcal G(\mathbf x_0,h)\to\mathbb R$.
%Furthermore, assume that the support of $\eta_h^\epsilon(\cdot;\mathbf x^*)$ is contained in $B(\mathbf x^*,\epsilon)$ with $\epsilon=O(h)$, and that there exists constant $K>0$ independent of $\epsilon$ such that
%\begin{equation*}\label{eq:boundh}
%	h^d \sum_{\mathbf x\in\mathcal G(\mathbf x_0,h)} |\eta_h^\epsilon(\mathbf x;\mathbf x^*)|\; |(\mathbf x-\mathbf x^*)^{\bs \alpha}| \le K, \quad \forall |\bs \alpha| = |\mathbf s|.
%\end{equation*}
%If $\{\eta_h^\epsilon(\cdot;\mathbf x^*): \epsilon>0\}$ satisfy the discrete $(q,\mathbf s)$-moment conditions at $\mathbf x^*$ and $\phi$ is of class $C^{q+|\mathbf s|}$ over $B(\mathbf x^*,\epsilon)$, then 
%\begin{equation}
%\label{eq:delta_error}
%	\left| \int_{\mathbb R^d} D^{\mathbf s}\delta(\mathbf x- \mathbf x^*) \;\phi(\mathbf x)\; d\mathbf x - h^d \sum_{\mathbf x\in\mathcal G(\mathbf x_0,h)}\eta_h^\epsilon(\mathbf x;\mathbf x^*) \; \phi(\mathbf x)\right| = O(h^q).
%\end{equation}
%\end{theorem}
%
%Theorem \ref{thm:discweakconv} is a discrete extension of what is presented in \cite{hoss:16}, for $|{\bf s}|\neq0$.
%Given equation \ref{eq:delta_error}, we refer to $q$ as the \emph{singular source approximation order} and $\eta^\epsilon_h(\mathbf x;\mathbf x^*)$ as being a $q$-order approximation of $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$.
%
%To construct approximations $\eta_h^\epsilon(\mathbf x;\mathbf x^*)$ for $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$ we first focus on the 1-D case, where we assume that the support of our approximation contains exactly $N$ grid points, that is, $2\epsilon=N$.
%For the case $N=q+s$, it can be shown that $\eta_h^\epsilon$ is uniquely determined:
%for a given grid point $x\in\mathcal G(x_0,h)$,
%\[
%	\eta_h^\epsilon(x; x^*) = \eta^\epsilon(x-x^*)
%\]
%where the continuum function $\eta^\epsilon$ is a piecewise polynomial,
% \begin{equation}\label{eq:eta}
% 	\eta^\epsilon(x) = \left\{ \begin{array}{rl}
%		P_\ell(x),& \quad x\in[a_\ell,a_{\ell+1}), \; \text{for}\; \ell=1,...,N\\
%		0,& \quad \text{otherwise}
%	\end{array}\right.
% \end{equation}
%with $P_\ell$ a polynomial over the considered interval, and $a_{\ell}=-\epsilon+ (\ell-1)h$ for $\ell=1,...,N+1$,
%\begin{equation}\label{eq:poly}
%	 P_\ell(x) = \left\{ \begin{array}{cl}
%		\displaystyle
%		s! (-1)^{N-1} \; 
%		\frac{
%			\displaystyle
%			\sum_{\underset{m_1,...,m_{q-1}\neq \ell}{1\le m_1<\cdots<m_{q-1}\le N}}
%			 ( h(m_1- \ell) + x) \cdots ( h(m_{q-1} - \ell) + x)
%		}
%		{
%			\displaystyle
%			h^{N}\prod_{\underset{m\neq \ell}{1\le m\le N}} ( \ell - m)
%		}
%		,& \text{for} \; q>1  \vspace{10pt} \\ 
%		\displaystyle
%		\frac{
%			s! (-1)^{s}
%		}
%		{
%			\displaystyle
%			h^{N} \prod_{\underset{m\neq \ell}{1\le m\le N}} (\ell - m)
%		}
%		,& \text{for}\; q=1.
%	\end{array}\right.
%\end{equation}
%Inspection of equation \ref{eq:poly} reveals that $\eta^\epsilon$ is a piecewise polynomial of degree $q-1$.
%The explicit form for $\eta^\epsilon$ (equations \ref{eq:eta} and \ref{eq:poly}) can be derived from imposing the discrete $(q,s)$-moment conditions over all $x^*\in\mathbb R$.
%
%Higher dimensional approximations are generated by taking tensor products of 1-D approximations. Namely $\eta^\epsilon:\mathbb R^d\to \mathbb R$ is given by
%\[
%	\eta^\epsilon(\mathbf x) = \prod_{k=1}^d \eta^{\epsilon_k}(x_k)
%\]
%where $\eta^{\epsilon_k}:\mathbb R\to\mathbb R$ is the continuum function that satisfies the discrete $(q,s_k)$-moment conditions for all $x^*_k\in \mathbb R$.
%
%
%A simple example in 1-D is given by taking $q=2$ and $s=0$, that is a second-order approximation to the delta function;
%\begin{equation}
%\label{eq:hat}
%\begin{split}
%       \eta^\epsilon(x) =& \left\{ \begin{array}{rc}
%                             \frac{x+h}{h^2}, & x\in[-h,0)\\
%                             \frac{h-x}{h^2}, & x\in[0,h)\\
%                             0, & \text{otherwise}
%                              \end{array}\right. \\
%\end{split}
%\end{equation}
%which coincides with the familiar {\em hat function}.
%It follows that the hat function is, for arbitrary $x^*$, only a second order approximation.
%
%
%%%%%%%
%\subsection{Discrete MPS-to-RHS Map}
%
%Recall our multipole source representation as given by equations \ref{eq:MPS} and \ref{eq:realdefn}. 
%We obtain a discrete analog, the {\em discrete multipole representation}, by replacing $D^{{\bf s}_{ij}}\delta({\bf x} - {\bf x}^*)$ in equations \ref{eq:realdefn} with grid functions $\eta^\epsilon_{h,ij}(\mathbf x;\mathbf x^*)$ for each $i,j$, according to the singular source approximation discussed above:
%\begin{equation}
%\label{eq:MPSdiscrete}
%\begin{split}
%        f_h(\mathbf x,t) &= \sum_{i} w_i(t) b_{h,i}(\mathbf x),\\
%        \mathbf f_h(\mathbf x,t) &= \sum_{i} w_i(t) \mathbf b_{h,i}(\mathbf x)\\
%        \mathbf F_h(\mathbf x,t) &= \sum_{i} w_i(t) \mathbf B_{h,i}(\mathbf x),
%\end{split}
%\end{equation}
%with
%\begin{equation}
%\label{eq:realdefndiscrete}
%\begin{split}
%	b_{h,i}(\mathbf x) & = \sum_j c_{ij} \eta^\epsilon_{h,ij}(\mathbf x;\mathbf x^*)  \\
%	\mathbf b_{h,i}(\mathbf x) & = \sum_j {\bf c}_{ij} \eta^\epsilon_{h,ij}(\mathbf x;\mathbf x^*) \\
%	\mathbf B_{h,i}(\mathbf x) & = \sum_j {\bf C}_{ij} \eta^\epsilon_{h,ij}(\mathbf x;\mathbf x^*).
%\end{split}
%\end{equation}
%According to our analysis and numerical experiments in our previous paper \citep{MPS_theory}, we take $q$ to be equal to the spatial order of the finite difference scheme.%, this will preserve numerical convergence rates.
%
%\noindent {\bf Remark:} Throughout the mathematical development, we will continue to regard time as continuous, and write $w_i(t)$ for the multipole coefficients. 
%Of course, in any implementation time will be discretely sampled as well.
%
%Discretization of multipole sources produces gridded approximations \ref{eq:MPSdiscrete}, where the formula may be viewed as defining a discrete version of the MPS-to-RHS map (equation \ref{eq:mpstorhsdefn}):
%\begin{equation} \label{eq:mpstorhsdefndiscrete}
%	(\mathbf L_h \mathbf w)(\mathbf x,t) = \sum_{i} w_i(t) b_{h,i}(\mathbf x),
%\end{equation}
%over grid points $\mathbf x\in\mathcal G(\mathbf x_0,h)$ as before.
%We call $\mathbf L_h$ the {\em discrete MPS-to-RHS} map. 
%Its output (vector in range) may be used as the right-hand side in a finite difference approximation to acoustodynamics (equations \ref{eq:acous}, \ref{eq:wave_acous}) or elastodynamics (equations \ref{eq:elas}). 
%The domain of this map is a discrete version of the MPS space described in the preceding section. 
%
%The range of $\mathbf L_h$ consists of the discrete functions of space and time, limited in space to those grid points that actually appear in the discrete multipole basis functions (equation \ref{eq:realdefndiscrete}). 
%Thus the information that characterizes the range of the discrete MPS-to-RHS map is
%(i) the locations of the grid points required to express the moment-consistent discrete multipoles listed in equation \ref{eq:realdefndiscrete}, these locations depending on the choice of discrete multipole basis as in \ref{eq:MPSdiscrete}, 
%and (ii) the time sampling information (time grid) which for sake of simplicity we assume is the same for the right-hand side and for all of the multipole coefficients.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

As mentioned in the introduction, the overarching software design
principle of this project is to organize the computation in a fashion
as parallel as possible with the underlying mathematics. The
mathematical venue of the computations described in the preceding
section is vector calculus and linear algebra. However, the key
concepts of vector calculus - vectors, linear operators, vector-valued
functions, derivatives, and so on - have no immediate reflection in
the basic types of any high level programming language. The programmer
of numerical algorithms such as those described here must define the
missing types and provide them with the proper behavior. While
straightforward programming with native arrays, in the fashion
traditional to computational science, can express vector calculus
algorithms, such expression does not encompass (for example) data
stored on disk or in a network, and the additional code
required to deal with such extensions inevitably intrudes into the
expression of mathematical algorithms to which they are irrelevant.

Programming with abstract types - that is, types defined by their
behavior rather than their implementation or data structure details -
provides a way out of this dilemma, allowing expression of abstract
algorithms that apply transparently, regardless of data structure and
low-level computational details. Code centered around
programmer-defined types has come to be called {\em object oriented}. 
While object oriented design can be implemented in almost
any language environment, some programming languages (C++, Python,
Java,...) offer explicit support for type definition and use. We chose
to program in C++, representing abstract numerical types as C++ classes.

%%%%%%
\subsection{The Rice Vector Library}
The Rice Vector Library (RVL) \cite[]{RVLTOMS} is an attempt to mimic
the structure of vector calculus in a system of C++ classes, so that common linear
algebra and optimization algorithms may be expressed in natural and
portable form.

RVL provides the basic interfaces for definition of abstract vector
calculus types in C++ and coding of important numerical algorithms in
terms of these types. Other packages offer a similar application
programmer interface, but tend to imprint the needs of particular
application areas in the core design. Recent examples include
Trilinos Rapid Optimization Library for simulation-driven optimization
\cite[]{ROL:15} and Operator Discretization Library for biomedical
tomography \cite[]{ODL:16}. Amongst these, RVL alone is built on those
interfaces, and only those interfaces, native to the mathematics
underlying continuous optimization and linear algebra.

The class {\tt RVL::Space} expresses the central abstraction of
calculus in Hilbert space, namely the Hilbert space. It acts as an
abstract factory for {\tt RVL::Vector}s, realizing in code the
mathematical commonplace ``let $x$ be a member of $V$''. 
%Most scientific vector objects have Cartesian product structure (traces in a
%gather, gathers in a line...) so most RVL vector spaces are instances
%of the subclass {\tt RVL::ProductSpace}. 
{\tt RVL::LinearOp} subtypes implement matrix-vector and transpose (adjoint)
matrix-vector product, without requiring explicit storage of a matrix.
{\tt RVL::Operator}s are nonlinear vector-valued
functions, with derivatives and second derivatives and their
transposes. Just as a linear operator is a special type of operator,
{\tt RVL::LinearOp} is a subclass of {\tt RVL::Operator}. Operators
act as vectors - they can be added and multiplied by scalars - but
there is no natural inner product on the vector space of
operators. Since {\tt RVL::Space} realizes the structure of Hilbert
space, {\tt RVL::LinearOp} cannot be regarded as a subclass of {\tt
  RVL::Space}.  However linear-operator valued functions are useful
and occur naturally. So we supply a separate {\tt RVL::LinOpValOp}
class to express such behavior, as an appropriate subclass of {\tt
  RVL::Operator}. We give an example of its use below.

For much more detail, design considerations, and examples, see
\cite{RVLTOMS}.

%%%%%%
\subsection{Out of core classes}

We have built the framework described here on top two open source seismic data processing packages, Seismic Unix
(SU, \cite[]{SU44}) and Madagascar \cite[]{Madagascar}. Both packages offer extensive file-to-file
utilities based on two different file data formats. The SU variant of SEGY tape format consists of
SEGY traces without reel headers. We refer to this as SU file format,
and use it to represent time series data throughout the project. The
Madagascar file format, Regularly Sampled Function (RSF) is ideal
for rectangularly gridded data. Since we will couple the MPS
representation to finite difference codes on regular grids, RSF format
is natural to represent grids and spatial functions on  them.

We have built {\tt RVL::Space} classes around these file formats, with
appropriate in-core data structures mirroring the information content
of the file data. These will be described in detail below.

%%%%%%
\subsection{IWave}

IWave is a wave equation simulator code, implementing several wave propagation models. 
The ones used in our examples are staggered-grid finite difference for acoustodynamic equations \ref{eq:acous} \citep{Vir:84} and centered differences for wave equation \ref{eq:wave_acous}. 
Each of these models defines an operator which we abstractly refer to as the {\em forward map} $\mathcal F$, mapping the pair of model parameters $(\mathbf m,\mathbf f)$ to data traces $\mathbf d$:
\begin{equation} \label{eq:fwdmodel}
	\mathcal F(\mathbf m,\mathbf f) = \mathbf d,
\end{equation}
where
\begin{itemize}
	\item $\mathbf m$ is the medium parameters, i.e., mechanical parameters in the modeling PDE,
	\item $\mathbf f$ is the source-term, which in this context abstractly refers to any of the right-hand side source terms.
\end{itemize}	
In the discrete setting, $(\mathbf m,\mathbf f)$ will correspond to the space-time discretizations of medium and source functions.
Similarly, $\mathbf d$ is typically space-time sampling of state variables, such as pressure field.

It can be shown that $\mathcal F$ is in fact linear in the $\mathbf f$ argument, which will be of particular importance when formulating the source and joint source-medium inverse problems.
In other words, $\mathcal F$ defines a linear operator (on $\mathbf f$) valued function of $\mathbf m$. 
RVL supplies the class {\tt RVL::LinOpValOp} for abstracting the behavior of such objects, and we wrap the IWave simulation code in this interface under the operator subclass {\tt TSOpt::IWaveLOVOp}.
The domain of these IWave operators are the Cartesian product of {\tt RVL:Space}'s encapsulating out-of-core RSF data for $\mathbf m$ and {\tt RVL::Space}'s based on SU trace data for $\mathbf f$.
The range is also an {\tt RVL::Space} (or Cartesian product of same) based on SU trace data. 
The action of the operator is file-to-file, very much like most SU and Madagascar utilities.

Along with the modeling relation \ref{eq:fwdmodel}, IWave also
provides access to derivatives and adjoints as is appropriate for
assembling inversion code. See \cite{FehlerKeliher:11} for the use of
an earlier version in QC of phase I of the SEAM project,
\cite[]{GeoPros:11} discussion of the design and implementation of
the IWave inversion interface, and \cite[]{trip14:struct} for a detailed
description of the overall code design. \cite[]{trip14:struct} is also a
reproducible paper, and can be found in the {\tt repro} subdirectory
of the code package accompanying this paper.

%%%%%%%%%%%
\subsection{MPS Framework Implementation}

The three fundamental constructs of the multipole point-source MPS framework are MPS spaces, the
MPS-to-RHS operator, and the composition of the latter with a wave
equation simulator to create an MPS-to-data modeling operator.
The MPS framework is built out of RVL types, which are explicitly
adapted to out-of-core data such as is stored in the SU and RSF data
formats.

%%%%%%%%%%%
\subsubsection{MPS Space}

The subclass {\tt MPS\_Space} of {\tt RVL::Space}
implements the MPS space abstraction described in the last section. 
As MPS spaces are parametrized by their multipole bases, {\tt MPS\_Space} is an
abstract or purely virtual class with three purely virtual member functions:
\begin{itemize}
	\item {\tt get\_MPS\_basis()}, returning a list of multipole basis vectors in the form of a \\
		{\tt std::vector<MPS\_base>};
	\item {\tt get\_type()} returns an identifying {\tt std::string} for use in error messages and the like;
	\item {\tt clone()}, a virtual copy constructor mandated by the base class {\tt RVL::Space}.
\end{itemize}
The {\tt MPS\_base} struct stores a vector of derivative index vectors, for example $(0,0,0)$ for no derivative (monopole), $(1,0,0)$ for dipole in the first direction, $(1,1,0)$ for a double couple in the $x_1$-$x_2$ coordinate plane, $(0,0,2)$ for a second order multipole in the $x_3$-direction, etc. 
Examples of {\tt get\_MPS\_basis()} for various MPS space types may be found in the directory {\tt iwave/mps/lib/MPS\_Space\_Examples.cc}. 
These examples also show how to implement the virtual copy constructor {\tt clone()} of the {\tt RVL::Space} class, by calling operator {\tt new} using the copy constructor of the class. 
Once {\tt get\_MPS\_basis()}, {\tt get\_type()}, and {\tt clone()} are defined in a subclass of {\tt MPS\_Space}, that subclass is fully defined and may be instantiated.

The main constructor of {\tt MPS\_Space} is inherited by every subspace and may be used in lieu of a subspace constructor. 
Its arguments are an {\tt MPS\_KEYS} struct, and a {\tt PARARRAY}.
{\tt PARARRAY} is an associative array class, part of the {\tt iwave/base} package. 
It stores {\tt key=value} pairs and offers a number of convenient manipulation routines. 
A {\tt PARARRAY} can be read from the command line or from a file recording key-value pairs.
The file format constraints are quite non-restrictive compared to other similar packages appearing in similar applications. 
For example, white space is inconsequential, and key-value pairs are extracted by finding two strings without white space, separated by an ``='' sign surrounded by any amount of white space. 
Any other string pattern is treated as a comment, which makes files defining {\tt PARARRAYS} easy to annotate.

{\tt MPS\_KEYS} is a kitchen-sink struct listing all key strings needed by any of the MPS classes, spaces and operators, to extract values from {\tt  PARARRAY}s.
For {\tt MPS\_Space}, the core members of {\tt MPS\_KEYS} that must be initialized are
\begin{itemize}
	\item {\tt grid\_file} - the key used to identify an RSF-formatted gridded data-file pair defining the simulation grid, usually the name of one of the spatial coefficient arrays such as bulk modulus or density, and
	\item {\tt MPS\_file} - key used to identify an SU file containing traces with the correct source geometry.
\end{itemize}
Only the source geometry and time sampling info (SEGY keywords {\tt sx, sy, selev, ns, dt, delrt}) are extracted from the {\tt MPS\_file} file data. 
The main purpose in passing the name of an SU file with these attributes, rather than the attributes
themselves, is to provide a prototype file that can be simply copied to create new instances of {\tt RVL::Vector}'s in the {\tt MPS\_Space}. 
SU provides excellent file initialization utilities, and the {\tt MPS\_file} file is likely to play other roles beyond specifying geometry, such as being the source for multipole coefficient data, so this is a convenient way to transfer the admittedly small amount of information needed to initialized a space.

To illustrate these concepts with concrete, compilable code, we have created the {\tt MPS\_howto.cc} example, a driver defined in  the source directory {\tt iwave/mps/main}, 
%FIX THIS
which we run as part of the {\tt project/SConstruct} accompanying this paper. 
This example shows how all of the fundamental
constructs in our framework are built and used.

Construction of {\tt MPS\_Space} requires that an {\tt MPS\_KEYS}
struct be initialized with at least the two items listed above. A
fragment from {\tt MPS\_howto.cc} shows how this is done:
{\small
\begin{verbatim}
    MPS_KEYS mk_sp;
    mk_sp.MPS_file  = "MPS_file"; 
    mk_sp.grid_file = "bulkmod";  
    ExScal_MPS_Space MPS_SP(mk_sp,*pars);
\end{verbatim}
}
The {\tt ExScal\_MPS\_Space} is a subclass of {\tt MPS\_Space}, defined in header file {\tt MPS\_howto.hh} in the MPS package {\tt include} directory. 
This particular space is an example of an MPS space for scalar multipoles with basis
\[
	b_1(\mathbf x) = \delta(\mathbf x-\mathbf x^*),\quad b_2(\mathbf x) = \left(\frac{\partial}{\partial x_1} + \frac{\partial}{\partial x_2}\right)\delta(\mathbf x-\mathbf x^*).
\]
Moreover, the {\tt MPS\_howto.cc} driver specializes with scalar source terms for the pressure equations of the acoustic system \ref{eq:acous} in either 2-D or 3-D.

Having constructed an {\tt MPS\_Space} instance, it is simple to
create an object corresponding to a vector in the space:
{\small
\begin{verbatim}
    RVL::Vector<float> mps(MPS_SP);
    RVL::AssignFilename 
      af_mps(RVL::valparse<string>(*pars,"MPS_file"));
    mps.eval(af_mps);
\end{verbatim}
}
Line 1 of this fragment constructs {\tt RVL::Vector} workspace {\tt
  mps}, representing a vector in the space {\tt MPS\_SP}.
Note that a floating point representation of real numbers (in this
case, {\tt float}) parametrizes {\tt RVL::Vector} via the C++ template
mechanism. In fact, {\tt MPS\_Space} is a subclass of {\tt
  RVL::Space<float>}. RVL accommodates other scalar fields, notably
{\tt double} and the {\tt std::complex} specializations, but MPS is
entirely a float (single precision real) package.

In line 3, {\tt mps} is connected to file storage through use of
function object {\tt RVL::AssignFilename}, constructed in line 2. 
Evaluation ({\tt eval(...)}) with this particular function object must
be the first operation on the {\tt RVL::Vector} after its creation,
else RVL will open a scratch file to hold the data array associated
with the vector. Such scratch files are deleted upon return from the
process in which they are created. For archival storage, the progammer
must make explicit use of {\tt RVL::AssignFilename}.  In line 2, the
{\tt valparse} function extracts a string from the {\tt PARARRAY}
pointed to by {\tt pars}, using the key string {\tt
  mks.MPS\_file}, and returns it for use as the archival filename,
passed to the constructor of {\tt RVL::AssignFilename}.

Lines 2 and 3 illustrate the encapsulation of data enforced throughout
RVL: with few exceptions, the only interactions with the data
underlying a vector - i/o, arithmetic, or otherwise - take place
through evaluation of function objects such as {\tt
  RVL::AssignFilename}. The white paper \cite[]{RVLTOMS} discusses at
length the justification for this strict encapsulation.

%Several MPS space subclasses are also defined in the file
%\texttt{MPS\_Space\_Examples.cc}, in {\tt iwave/mps/lib}. 
%One subclass used throughout our
%numerical examples is the \emph{scalar MPS space}
%(\texttt{Scal\_MPS\_Space} class), spanned by a single
%multipole basis $b_1$,
%\begin{equation}\label{eq:ScalMPS}
%        b_1({\bf x}) = D^{\bf s}\delta({\bf x}-{\bf x^*}),
%\end{equation}
%for a given multi-index ${\bf s}$.
%%MJB edit
%%vector $b^h_{\bf n}$ ($bf n = (n_1,n_2,n_3)$)  chosen to match
%%sufficiently many moments with a specified mixed
%%partial derivative of spatial delta.

%%%
\subsubsection{MPS-to-RHS map}

Our framework realizes the discrete MPS-to-RHS map $\mathbf L_h$ (equation
\ref{eq:mpstorhsdefndiscrete}) as a subclass of {\tt
  RVL::LinearOp}, an abstract base class encoding the behaviour of a linear map or
operator. Amongst other things, linear operators have domain and range
(vector) spaces. Just so, two of {\tt RVL::LinearOp}'s public member
functions are {\tt getDomain()} and {\tt getRange()}, which return
{\tt const} references to the {\tt RVL::Space}s representing domain
and range respectively. Two other important functions are {\tt
  applyOp} and {\tt applyAdjOp}, implementing the matrix-vector
product and transpose matrix-vector product respectively.

The class {\tt MPS\_to\_RHS} implements the MPS-to-RHS map. The domain
space of an instance is an {\tt MPS\_Space}, which is an argument to
the constructor, along with {\tt MPS\_KEYS} and {\tt PARARRAY} objects
as was the case for {\tt MPS\_Space}:
{\small
\begin{verbatim}
    MPS_KEYS mk_m2r;
    mk_m2r.MPS_file  = "MPS_file"; 
    mk_m2r.grid_file = "bulkmod";  
    mk_m2r.appx_ord  = "appx_ord"; 
    mk_m2r.RHS_files.push_back("source_p");
    MPS_to_RHS M2R(mk_m2r,*pars,MPS_SP);
\end{verbatim}
}
The keys {\tt bulkmod} and {\tt source\_p} point to the names of files
containing bulk modulus (a spatial field, therefore RSF format) and
source traces for the pressure equation (first equation in system
\ref{eq:acous}).

Another code fragment from {\tt MPS\_howto.cc} illustrates a typical usage
of the RVL version of the MPS-to-RHS map:
{\small
\begin{verbatim}
    RVL::Vector<float> rhs(M2R.getRange());
    RVL::AssignFilename
      af_rhs(RVL::valparse<string>(*pars,"source_p"));
    rhs.eval(af_rhs);
    M2R.applyOp(mps,rhs);
\end{verbatim}
}
Lines 1-3 show creation of a vector, this time in the range space of
the {\tt MPS\_to\_RHS} operator {\tt M2R}, and link to archival file storage.
In line 4, the {\tt MPS\_to\_RHS} operator {\tt M2R} is applied to the
multipole coefficient vector {\tt mps}, created a few lines before, and the
output written to {\tt rhs}.

In our {\tt MPS\_howto} example, we provide the MPS-to-RHS map with 
input SU file {\tt howto\_mps.su}, containing multipole coefficient vector $\mathbf w=(w_1,w_2)$ as shown in figure \ref{fig:howto-mps}.
The output is a single SU file {\tt howto\_RHS.su}, plotted in figure \ref{fig:howto-RHS} 
in the form of a collection of time traces plotted in grey scale.
Each trace in {\tt howto\_RHS.su} has a spatial location corresponding to spatial grid points associated
with the output of the discrete MPS-to-RHS map. 
The number of traces in the output of an {\tt MPS\_to\_RHS} operator, 
i.e., the number of grid points within the support of our singular source approximation,
is given by
\[
	\prod_{k=1}^d (s_{k,max} +q)
\]
where $s_{k,max}$ is the maximum number of derivatives in the spatial $k$-axis for the MPS space in consideration.
Note that there are a total of $100$ traces in {\tt howto\_RHS.su}, which is 
to be expected for this example in 3-D with source approximation $q=4$,
\[
	(1+4)\times(1+4)\times 4 = 100.
\]
%The MPS-to-RHS map is implemented as the \texttt{MPS\_to\_RHS} class,
%a derived class of \texttt{RVL::LinearOp}, templated by its
%\texttt{MPS\_Space} domain.  The \texttt{MPS\_to\_RHS} class draws
%from its domain {\tt MPS\_Space} all of the necessary information
%required to implement the singular source approximation as a linear
%operator (see \ref{eq:L}) to output IWave compatible RHS sources.

%The \texttt{MPS\_to\_RHS} operator will also have ``make'' flags for both its domain and range, similar to the ``make'' flag for the \texttt{MPS\_Space} class.
%The essential forward and adjoint actions of the \texttt{MPS\_to\_RHS} operator will be carried out by kernels, mainly \texttt{MPS\_to\_RHS\_kern} and \texttt{RHS\_to\_MPS\_kern} respectively, through low level function calls to standard SU I/O methods such as \texttt{fgettr(s)} and \texttt{fputtr()}.
%See appendix A for more discussion on the forward action of \texttt{MPS\_to\_RHS} operator and a derivation of its adjoint.

%Consider the forward modeling problem using the acoustic equations from \ref{eq:acous}. 
%Let $F$ abstractly denote the operator (forward map) from both medium/model and source parameters to data,
%\[
%        F(\mathbf m,\mathbf f) = \mathbf d
%\]  
%where $\mathbf m=\{\beta,\kappa\}$ and $\mathbf f$ corresponds to IWave RHS source parameters.
%For this example, and in our numerical tests, we take data vector $\mathbf d$ to be time traces of the pressure field at reciever locations.
%Forward map $F$ has a special structure which will be exploited in
%the context of joint medium-source inversion, namely $F$ is linear
%with respect to a subset of parameters (in this case $\mathbf f$)
%thus resulting in a \emph{separable} nonlinear least squares problem.

%%%%
\subsubsection{Composition of MPS and IWave}

%The range space of the discrete MPS-to-RHS operator is a space-time
%grid function, a discrete representation of a linear combination of
%delta functions. This is in principle exactly what is required as a
%right-hand side for the finite difference discretization of the wave
%equations \ref{eq:acous} and \ref{eq:elas}.

IWave implements the forward modeling operator (equation \ref{eq:fwdmodel}) as an {\tt TSOpt::IWaveLOVOp}, a subtype of {\tt RVL::LinOpValOp}. 
This operator has a product space domain, consisting of pairs (medium parameter fields, source fields). 
To compute the data traces generated by a multipole represented by a multipole coefficient vector, the IWave operator must be chained or {\em composed} with the MPS-to-RHS map (equation \ref{eq:mpstorhsdefndiscrete}). 
As just noted, the latter operator is represented as a {\tt RVL::LinearOp} in our framework.
RVL provides a number of classes for  operator composition. 
These are of the type sometimes known as ``decorator'' or ``wrapper'' classes: they combine instances of various types to obtain the behavior of yet another type. 
In this instance, we use an {\tt RVL::LinCompLOVOp}, which implements the composition
\begin{equation}
\label{eq:iwavemps}
	(\mathbf m,\mathbf w) \mapsto \mathcal F(\mathbf m,\mathbf L \mathbf w).
\end{equation}

The code fragment from {\tt MPS\_howto.cc} illustrates instantiation of a {\tt TSOpt::IWaveLOVOp} and its composition with the {\tt RVL::LinearOp} subtype {\tt MPS\_to\_RHS}:
{\small
\begin{verbatim}
    TSOpt::IWaveLOVOp IWOP(M2R.get_pars(),stream);
    RVL::LinCompLOVOp<float> F(M2R,IWOP);
\end{verbatim}
}
At this point in the code, the {\tt MPS\_to\_RHS} instance {\tt M2R}
has already been built. Its internal {\tt PARARRAY} object contains
all parameters needed by {\tt TSOpt::IWaveLOVOp} as well. In
particular, the file used by {\tt MPS\_to\_RHS} is listed under the
key words used to generate the linear part of the domain for {\tt
  TSOpt::IWaveLOVOp}, so that the domain/range sanity checks built
into the {\tt RVL::LinCompLOVOp} are satisfied.  {\tt RVL::LinCompLOVOp} 
builds a functional {\tt RVL:LinOpValOp} out of
the two arguments to its constructor, implementing equation
\ref{eq:iwavemps}.
Figure \ref{fig:howto-data} plots the resulting output, a single trace of the pressure field for our running {\tt MPS\_howto} example; see {\tt hdr\_3D.su} for more details on source-receiver geometry.

The singular source approximation, as discussed in the theory section, is encoded in the discrete MPS-to-RHS map and streamlined into IWave forward modeling via operator composition as mentioned above.
With the resulting composition, it is possible to compute the data directly given the multipole coefficient vector and the acoustic or elastic medium parameter fields (density, bulk modulus,...). 
Since the operator class built this way implements derivatives up to order two and their adjoints, it is possible to couple the results to RVL optimization code and so construct inversion applications. 
In the next section we offer some examples of inversion code that take advantage of our multipole source framework.

%The MPS and IWave framework come together through operator composition in the linear parameters. 
%Let $\mathbb L$ denote the \texttt{MPS\_to\_RHS} operator and $\mathbf f_{\text{MPS}}$ an element %of some derived class of \texttt{MPS\_Space}.
%Then we can define a forward map $F_{\text{MPS}}$ as an operator of $\mathbf f_{\text{MPS}}$ by 
%\[
%        F_{\text{MPS}}(\mathbf m,\mathbf f_{\text{MPS}}) := F(\mathbf m, \mathbb L\mathbf f_{\text{MPS%}}).
%\]

%It is worth mentioning that $F_{\text{MPS}}$ is more than just a linear operator with respect to $\mathbf f_{\text{MPS}}$, it can be shown to be a multichannel convolution between MPS coefficients and proper Green's functions.
%In fact, the \texttt{MPS\_IWaveLOVOp} implements $F_{\text{MPS}}$ as a multichannel convolution, and its adjoint as corresponding cross-correlation, by computing and storing appropriate Green's functions for a given set of model parameters through the use of \texttt{IWaveLOVOp}, \texttt{MPS\_to\_RHS} and other IWave operators.

%%%%%%%%%%%%%%%%
\section{Numerical Examples}
%%%%%%%%%%%%%%%%

In the numerical examples to come, we make use of our multipole source parametrization and computational framework to implement inversion code for source estimation problems.
Numerical examples will pertain to inversion problems associated with forward modeling via the acoustic wave equation \ref{eq:wave_acous} in 2-D.
Moreover, we consider scalar multipole sources spanned by the single base
\begin{equation}\label{eq:singleMPS}
	b_1(\mathbf x) = D^{\mathbf s}\delta(\mathbf x-\mathbf x^*),
\end{equation}
particularly for $\mathbf s=(0,0)$.

%%%%%
\subsection{Forward Modeling}

As alluded to in our discussion of IWave and wave propagation modeling, we can define a modified forward map $\tilde{\mathcal F}$ as the abstract operator mapping model parameters $(\mathbf m,\mathbf w)$ to data $\mathbf d$ through operator composition with the MPS-to-RHS map $\mathbf L$ and the IWave forward map $\mathcal F$:
\begin{equation}\label{eq:Fmap}
	\tilde{\mathcal F}(\mathbf m,\mathbf w) := \mathcal F(\mathbf m,\mathbf L\mathbf w) = \mathbf d
\end{equation}
Note that $\tilde{\mathcal F}$ is also linear with respect to its second argument, and thus we use the following notation for convenience at times,
\begin{equation}\label{eq:F_lin}
	\tilde{\mathcal F}(\mathbf m,\mathbf w) = \tilde{\mathcal F}(\mathbf m)\mathbf w.
\end{equation}

The medium parameter $\mathbf m$ for acoustic wave modeling corresponds to the square of the wave velocity, i.e., $\mathbf m = c(\mathbf x)^2$.
Also, given our choice of MPS space (scalar multipoles with basis \ref{eq:singleMPS}), we have $\mathbf w=w_1(t)$.
Lastly, in our numerical examples, $\mathbf d$ consists of the pressure field $p$ sampled at receiver locations $\{\mathbf x_{i_r} : i_r=1,2,...\}$, that is,
\[
	\mathbf d = \{ p(\mathbf x_i,t) : i_r=1,2,...\}.
\]

Driver {\tt acd\_scal.cc} is used to generate ``observed'' data, $\mathbf d_{\rm obs}$, for our inversion examples.
The driver implements forward map \ref{eq:Fmap}, as done in {\tt MPS\_howto.cc}, through operator composition of {\tt TSOpt::IWaveLOVOp} and {\tt MPS\_to\_RHS} operators.
We discuss some of the input key-value pairs required for running {\tt acd\_scal}, see {\tt SConstruct} located in the {\tt project} directory of this paper:
\begin{itemize}
	\item {\tt csq=csq24.rsf}, input RSF file for medium parameter;
	\item {\tt source=empty}, input SU file related to right-hand side source terms, it is set to {\tt empty} as a place holder since it is necessary for IWave code;
	\item {\tt data=data.su}, output SU file containing pressure field time traces with specified source-receiver geometry;
	\item {\tt MPS\_file=mps\_towed.su}, input SU file containing multipole coefficients and source locations;
	\item {\tt MPS\_ord\_0=0 MPS\_ord\_1=0}, equivalent to multi-index $\mathbf s=(0,0)$ for multipole source \ref{eq:singleMPS};
	\item {\tt appx\_ord=4}, singular source approximation order, $q=4$;
	\item {\tt order=2}, spatial half-order of FD scheme, thus a fourth-order in space scheme here;
	\item {\tt cfl=0.75 cmin=1.5 cmax=4.0}, CFL factor and bounds on wave velocity;
\end{itemize}

Input medium parameter file {\tt csq24.rsf} is plotted in figures \ref{fig:csq24}, it consists of a fraction of the Marmousi model with an added water layer to simulate a marine environment.
The acoustic wave equation is discretized using a standard second-order in time and fourth-order in space centered finite difference scheme.
The spatial grid size $h$ is taken from the medium parameter file which contains information of the underlying spatial grid; $h=24m$ in {\tt csq24.rsf}.
Time step size $\Delta t$ is given by CFL relation
\[
	\Delta t = \alpha \frac{h}{c_{\rm max}},
\] 
which can be tweaked through user input CFL factor $\alpha$ and $c_{\rm max}$, corresponding to keys {\tt cfl} and {\tt  cmax} respectively.

Source-receiver geometry for the generation of data file {\tt data.su} is based on marine seismic surveying, where the source is activated multiple times while being towed across the ocean by a boat, along with a string of trailing receivers.
Each activation of the source is referred to as a {\em shot}, and each shot results in data measurements that are independent from each other.
File $\tt data.su$ contains a concatenation of measured data for 16 shots.
The first shot occurs at location $\mathbf x^* = (8,2500)$, where the coordinates correspond to depth and horizontal displacement, in meters, respectively.
The first line of receivers are located at depth $12m$ spanning a horizontal displacement from $500$ to $2300m$, with $25m$ spacing between receivers.  
The remainder of the shots consist of towing the initial source-receiver configuration to the right of the domain $100m$ at a time.
Thus, the last shot is located at $4000m$ horizontal displacement.
Input SU file {\tt mps\_towed.su} contains multipole coefficient $\mathbf w$ for each shot, consisting of a shifted Ricker wavelet of $5Hz$ peak frequency; see figure \ref{fig:mps-towed}.

%%%%%
\subsection{Multipole Source Inversion}

We assume domain and range spaces of $\tilde{\mathcal F}$ have inner-product structure for which it makes sense to talk about norms and adjoints of linear operators; an obvious Euclidian inner-product for MPS spaces is given by \ref{eq:euclid}.
Multipole source inversion is formulated as a (linear) least squares problem: given $\mathbf m$, find $\mathbf w^*$ such that it minimizes the misfit between predicted and observed data in a least squares sense, i.e.,
\begin{equation}\label{eq:LLS}
	\mathbf w^* = {\rm argmin}_{\mathbf w} \; 
	\| \tilde{\mathcal F}(\mathbf m)\mathbf w - \mathbf d_{\rm obs}\|^2.
\end{equation}
Making use of our linear operator notation \ref{eq:F_lin}, it follows that least squares estimate $\mathbf w^*$ is also the solution to the normal equations.
\begin{equation}\label{eq:normal}
	\tilde{\mathcal F}(\mathbf m)^T (\tilde{\mathcal F}(\mathbf m)\mathbf w^* - \mathbf d_{\rm obs}) =0.
\end{equation}

Driver {\tt acd\_scal\_inv\_towed.cc} solves the multipole source inverse problem iteratively by solving normal equations \ref{eq:normal} via conjugate gradient.
Recall that multi-shot data {\tt data.su} is generated from a single multipole source that is being towed horizontally.
Therefore, we should only estimate a single multipole coefficient, rather than a multipole coefficient for each shot.
To this end we introduce the {\tt MPS\_spread} operator, a subtype of the {\tt RVL::LinearOp} class, which implements the linear operation of taking a multipole source and copying (or spreading) it over many source locations specified by the range MPS space.
The {\tt MPS\_spread} constructor requires only the domain and range {\tt MPS\_Space}'s to initialize such operators.
The following piece of code from {\tt acd\_scal\_inv\_towed.cc} shows how to initialize an {\tt MPS\_spread} operator:
{\small
\begin{verbatim}
    MPS_KEYS mk_sp;
    mk_sp.MPS_file  = "MPS_file_inv";
    mk_sp.grid_file = "csq";
    Scal_MPS_Space MPS_SP(mk_sp,*pars);
    MPS_KEYS mk_tw;
    mk_tw.MPS_file  = "MPS_towed";
    mk_tw.grid_file = "csq"
    Scal_MPS_Space MPS_TW(mk_tw,*pars);
    MPS_spread MPS_SPR(MPS_SP,MPS_TW);
\end{verbatim}
}
Making use of operator composition we can create the target forward map, an operator from medium parameters and the single multipole source (rather than the towed source) to data traces:
{\small
\begin{verbatim}
    MPS_KEYS mk;
    mk.MPS_file   = "MPS_towed";
    mk.grid_file  = "csq";
    mk.appx_ord   = "appx_ord";
    mk.delta_file = "MPS_delta";
    mk.RHS_files.push_back("source");
    mk.hdr_files.push_back("data");
    mk.G_files.push_back("data_g");
    MPS_IWaveLOVOp<Scal_MPS_Space> MPS_LOVO(mk,*pars,stream);
    RVL::LinCompLOVOp<float> F(MPS_SPR,MPS_LOVO);
\end{verbatim}
}
The {\tt MPS\_IWaveLOVOp} class is a subtype of {\tt RVL::LinOpValOp} and is templated by {\tt MPS\_Space}.
This version of {\tt IWaveLOVOp} implements forward map $\tilde{\mathcal F}$ (equation \ref{eq:Fmap}) as a convolution between multipole coefficients and a kernel related to a Green's function.
The convolution kernel is computed once and stored for a given medium parameter $\mathbf m$, this involves one PDE solve.
Computing forward (and adjoint) actions of linear operator $\tilde{\mathcal F}(\mathbf m)$ during source estimation thus comes at a cost of a convolution, rather than a more costly PDE solve.
%{\tt MPS\_KEY.delta\_file} refers to a compatible SU file containing a filtered $\delta(t)$ as multipole coefficients as input for computing the Green's function, stored in another SU file referred by {\tt MP\_KEY.G\_files}.
{\tt MPS\_KEY}'s, {\tt "MPS\_delta"} and {\tt "data\_g"}, refer to SU files required for the computation of the convolution kernel, while the other keys are needed for an operator composition between {\tt MPS\_to\_RHS} and {\tt IWaveLOVOp} internal to {\tt MPS\_IWaveLOVOp}.

Before we can apply a conjugate gradient algorithm to solve the normal equations, we must first extract the linear operator $\tilde{\mathcal F}(\mathbf m)$ from our linear operator valued operator $\tilde{\mathcal F}$.
This is done through the {\tt RVL::LinearRestrictOp} class, a subtype of {\tt RVL::LinearOp}, which upon initialization takes an {\tt RVL::LinOpValOp} and an {\tt RVL::Vector} from the nonlinear domain of said operator and yields the desired linear operator;
in our example code:
{\small
\begin{verbatim}
    RVL::Vector<float> m(F.getProductDomain()[0]);
    RVL::AssignFilename af_csq(valparse<string>(*pars,"csq"));
    m.eval(af_csq);
    RVL::LinearRestrictOp<float> F_lin(F,m);
\end{verbatim}
}

Lastly, in the next lines of code we initialize and apply a {\em conjugate gradient for normal equations} (CGNE) algorithm \citep{Paige:82}, implemented by the {\tt RVLUmin::CGNEAlg} abstract class.
Without going into too much detail, initializing an {\tt RVLUmin::CGNEAlg} object requires {\tt const} references to {\tt RVL::Vector}'s (estimated multipole coefficient vector {\tt w}, and observed data vector {\tt d}) and an {\tt RVL::LinearOp} (linear forward map {\tt F\_lin}), among other things:
{\small
\begin{verbatim}
    Vector<float> w(F.getProductDomain()[1]);
    Vector<float> d(F.getRange());
    AssignFilename af_mps(valparse<string>(*pars,"MPS_file_inv"));
    AssignFilename af_data(valparse<string>(*pars,"data"));
    w.eval(af_mps);
    d.eval(af_data);

    RVLUmin::CGNEAlg<float> cgalg (w,F_lin,d,
        rnorm,nrnorm,rtol,gtol,maxiter,trustrad,sstream);
    cgalg.run();

\end{verbatim}
}
The {\tt run()} method runs the CGNE algorithm, overwriting {\tt RVL::Vector w} with the computed solution.
{\tt RVLUmin::CGNEAlg} is part of the unconstrained minimization library of RVL; codes can be found in the {\tt trip/rvl/umin} subdirectory.
The estimated $\mathbf w^*$ for this example is plotted in figure \ref{fig:test2-mps} (dashed red curve), superimposed with the true solution (solid blue curve). 
We remark that a constant zero function is used as the initial guess for this result.

%%%%%
\subsection{Joint Source-Medium Inversion}

The joint source-medium inversion problem consists of estimating both source and medium parameters simultaneous in a similar minimum misfit least squares setting:
\begin{equation}\label{eq:NLS}
	(\mathbf m^*,\mathbf w^*) = {\rm argmin}_{\mathbf (\mathbf m,\mathbf w)} \;
		\| \mathcal R(\mathbf m,\mathbf w)\|^2,
\end{equation}
where we now introduce the {\em residual function} $\mathcal R$,
\[
	\mathcal R(\mathbf m,\mathbf w) := \tilde{\mathcal F}(\mathbf m)\mathbf w - \mathbf d_{\rm obs}.
\]
Note that \ref{eq:NLS} is now a nonlinear least squares problem if $\mathcal R$ is nonlinear in $\mathbf m$, which typically is the case. 
Nonetheless, we can capitalize on the particular structure of the resulting {\em separable nonlinear least squares} problem \ref{eq:NLS} by a reformulation known as the {\em variable projection method} accredited to \cite{GolubPereyra:73}.
Let $\mathbf w_{\rm VP}(\mathbf m)$ denote the solution to the linear least squares problems \ref{eq:LLS} which is, as suggested by notation, dependent on the given medium parameter $\mathbf m$, and thus can be viewed as a function of $\mathbf m$.
Replacing $\mathbf w$ with $\mathbf w_{\rm VP}(\mathbf m)$ in the residual function yields a reduced (still nonlinear) least squares problem, reduced in the sense that $\mathbf w$ is effectively eliminated from the inversion:
\begin{equation}\label{eq:VPM}
	\mathbf m^* = {\rm argmin}_{\mathbf m} \; \| \mathcal R_{\rm VP}(\mathbf m)\|^2
\end{equation}
where $\mathcal R_{\rm VP}$ is the {\em reduced residual function},
\begin{equation}\label{eq:VPMfun}
	\mathcal R_{\rm VP}(\mathbf m) := \mathcal R(\mathbf m,\mathbf w_{\rm VP}(\mathbf m)).
\end{equation}

Solving the reduced least squares problem \ref{eq:VPM} amounts to a nested optimization scheme, where linear parameter $\mathbf w$ and nonlinear parameter $\mathbf m$ are updated in the inner and outer optimization routines respectively.  
In practice, variable projection coupled with Gauss-Newton type algorithms for solving nonlinear least squares has been proven to outperform other multiparameter estimation algorithms in some cases \citep{Ruhe:1980}.
\cite{Rickett:SEG12} demonstrate the effectiveness of variable projection over simultaneous descent and alternating direction in the context of joint source-medium inversion.
Future work regarding joint inversions will incorporate variable projection as means to mitigate the difficulty of multiparameter inversion.

Driver {\tt acd\_scal\_Jinv\_towed.cc} implements the joint source-medium inversion code.
The reduced residual operator is actualized in code by the {\tt RVLUmin::ReduceOp} class, part of RVL's unconstrained optimization library.
Given the {\tt RVL::LinOpValOp} forward map, data {\tt RVL::Vector}, and a CGNE algorithm (for the inner linear least squares solves), initializes an {\tt RVLUmin::ReduceOp} object:
{\small
\begin{verbatim}
    ReduceOp<float,CGNEPolicy<float>,CGNEPolicyData<float> > 
        R_VP(F,d,cgne,sstream);
\end{verbatim}
}

We solve nonlinear least squares problem \ref{eq:VPM} by a variant of the Steihaug-Toint trust region globalization of the Gauss-Newton (TRGN) algorithm, with flexible specification of the least-squares substep, implemented by {\tt RVLUmin::TRGNAlg}.
See for instance \cite{Steihaug:83}, \cite{ConnGouldToint:00}, and \cite{NocedalWright}.

%The {\tt RVLUmin::ReduceOp} class, applying the variable projection formulation, and the various linear and nonlinear least square iterative solvers (CGNE and TRGN respectively) used in the numerical examples here are defined in terms of abstract RVL classes, that is, with the defined behavior of operators (derivatives and adjoints thereof) and vectors over inner-product spaces.
%These abstract algorithms can be found in the subdirectory {\tt trip/rvl/umin}.

Figure \ref{fig:test4-csq-inv} plots the recovered medium parameter after three TRGN steps, with a maximum of 30 CGNE iterations for the inner optimization loop for multipole coefficients.
Given that TRGN is a local optimization method, we started with an initial guess not too far from the true medium parameter, in fact a smooth version of {\tt csq24.rsf} given by {\tt csq24sm.rsf} and plotted in \ref{fig:csq24sm}.
Figure \ref{fig:trace-compare} plots a 1-D slice of the recovered, starting, and true medium parameter as a function of depth at $3500m$ of horizontal displacement.
Lastly, figure \ref{fig:test4-mps} plots a comparison of the true and estimated $w_1(t)$ function for the joint inversion.

%We use the code described in the last section to generate traces of pressure fields 
%induced by multipole sources.
%In particular, we focus on 3-D acoustics with unbounded homogenous medium.
%%where we compare our numerical solutions to that of analytical solutions derived in the appendix.
%We use the IWave implementation of staggered grid finite difference schemes for 
%the acoustic system \ref{eq:acous} \cite[]{Vir:84}, of order 2 in time and order 4 in 
%space - we refer to this as the 2-4 order scheme. 
%In these experiments we use only scalar (pressure) sources and pressure trace data. 
%Similar results are obtained with other choices. 
%Boundary conditions are of PML type, as described by \cite{Habashy:07}, as to
%simulate unbounded domains.
%
%The code used in these numerical examples incorporates sources of scalar type, 
%namely, sources with multipole bases $\{b_1,b_2\}$ for the {\tt ExScal\_MPS\_Space} (from {\tt MPS\_howto.cc})
%\[
%	b_1(\mathbf x) = \delta(\mathbf x-\mathbf x^*), \quad b_2(\mathbf x) = \left( \frac{\partial}{\partial x_1} + \frac{\partial}{\partial x_2}\right)\delta(\mathbf x-\mathbf x^*),
%\]
%and a single base term $\{b_1\}$ for the {\tt Scal\_MPS\_Space} (from {\tt asg\_scal\_p.cc}),
%\[
%	b_1(\mathbf x) = D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)
%\]
%given multi-index ${\bf s}$. 
%Discretization of multipole terms $b_{i}$ is chosen to match the spatial order of the finite difference method, namely $q=4$ for the 2-4 scheme.
%
%Multipole coefficients $w_1, w_2$ for the {\tt MPS\_howto} example are chosen to be the derivative of a Gaussian and a Ricker wavelet respectively, see figures 3 and 4.
%Multipole coefficient $w_1($ for the simulated data when running $asg\_scal\_p.cc$ code is chosen to be a Gaussian wavelet of the form
%\[
%	w_1(t) = e^{-\pi^2f^2(t-t_c)^2}
%\]
%with $f=10Hz$ and $t_c = 0.4 s$, see figure \ref{fig:wlt-gauss}.
%
%Coordinates are oriented such that $x_1$ is the depth axis.
%See figure \ref{fig:../../model-geo} for a diagram of the source-receiver geometry.
%The following are some other specifications that apply to all experiments:
%\begin{itemize}
%	\item constant density $\rho=1kg/m^3$ and bulk modulus $\kappa = 9 Gpa$, i.e., constant velocity $c=3km/s$;
%        \item total recording time is $1.5s$;
%        \item spatial grid sizes $h=20m$;
%        \item time step $\Delta t=0.002s$;
%        \item source location $\mathbf x^* = (-298m,602m,303m)$;
%        \item receiver location $\mathbf x_r$ = (-300m,2400m,300m).
%\end{itemize}
%Note that the source location is chosen as to not coincide with a grid point.
%Moreover, the grid cell size is $20m$, which implies $5$ {\em grid points per wavelength} at $30Hz$ (see figure \ref{fig:wlt-gauss-fft}), thus minimizing the effects of grid dispersion.
%
%Figure \ref{fig:MPS-howto} plots the input MPS coefficient vector $w=\{w_1,w_2\}$, again a collection of waveforms associated with certain multipole bases, in particular two terms $\{b_1,b_2\}$ given our example.
%The actual source that gets injected as a right-hand side term in the finite difference
%scheme is given by figure \ref{fig:howto-RHS}; each trace corresponds to a point source on a grid point in the vicinity of the source location $\mathbf x^*$.
%These effective sources are, as discussed in the methods section, the output of the 
%discrete MPS-to-RHS operator that applies the singular source approximation given a multipole representation (i.e., an MPS space).
%We have effectively gone form two point sources, i.e., multipoles $w_1b_1$ and $w_2b_2$, to a spread of 100 point sources, where the 100 can be accounted for by 
%analyzing the support of our singular source approximation.
%Let $s_{i,max}$ denote the maximum derivative order in the $x_i$-coordinate over all
%MPS basis, thus we can expect the RHS source file to have a total number of traces
%given by
%\[
%	\prod_{i=1}^d (s_{i,max} +q).
%\]
%Indeed, for the {\tt ExScal\_MPS\_Space}, we have
%\[
%	(1+4) \times (1+4) \times (0+4) = 100.
%\]
%
%
%As a way to validate the correctness of our MPS framework and implementation, 
%we compare numerical and analytical solutions for the simplest scenario for which
%we can derive such expressions, namely, in unbounded homogeneous media for single multipole sources.
%We refer to the appendix for the derivation of analytical formulas used here.
%Figures \ref{fig:test3D-s000}, \ref{fig:test3D-s010}, \ref{fig:test3D-s020} plot the analytical and numerical solutions, as well as the error, for different multipole sources; $\mathbf s=(0,0,0),(0,1,0),(0,2,0)$ respectively.
%Results demonstrate that the numerical solutions, that is the finite difference and source discretizations, yield accurate approximations to the analytical solutions.
%

\newpage
%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%

The RVL-based structure of our multipole source framework does get a solid workout in this study. 
Inspection of the code and scripts shows that the framework is a convenient environment for implementing the numerical examples presented in the last section.
For example, singular source approximation discussed in the theory section is implemented as an abstract linear operator and streamlined into IWave forward modeling via operator composition, e.g., {\tt MPS\_howto.cc} and {\tt acd\_scal.cc}.
Moreover, this allowed for the seamless application of our multipole source parametrization into inversion application: source inversion ({\tt acd\_scal\_inv\_towed.cc}) and joint source-medium inversion ({\tt acd\_scal\_Jinv\_towed.cc}).

Thus far we have been keen in avoiding some of the difficulties associated with multipole source inversion by presenting results for the simplest multipole possible, i.e., sources of the form $f(\mathbf x,t) = w_1(t)\delta(\mathbf x-\mathbf x^*)$.
Going beyond the Dirac delta function results in a very ill-conditioned forward map stemming in part from the effect of the spatial delta derivatives on the output pressure field related to temporal (possibly fractional) derivatives of multipole coefficients.
In other words, the linear forward map $\tilde{\mathcal F}(\mathbf m)$ resembles a differential operator defined over an $L^2$-normed domain resulting in an unbounded operator.
In a future paper we plan on defining a weighted inner-product to better bound the forward map.
Equivalently, we introduce a preconditioner in order to dramatically accelerate the convergence of source recovery by Krylov subspace iteration.
Acceleration of the inner linear least squares problem will be vital for the overall computational cost of the joint source-medium inverse problem.

Other sources of ill-conditioning of the forward map (i.e., poor data coverage, non-empty null space) can be addressed through other more standard means such as Tikhonov regularization.
This of course leads to the proverbial question of how to choose the Tikhonov regularization parameter.
The matter of the null space will be more important for higher order multipoles, given that these source representations can lead to the phenomena of null-radiating point-sources, that is, non-zero point-sources that produce null pressure responses outside a ball of finite radius.

We have not addressed two other important issues.
The first is the order of multipole necessary to represent a given degree of source anisotropy. 
Point support is an idealization: active source regions may be small or comparable to a wavelength in spatial extent. 
\cite{SantosaSymes:00} showed that multipole approximation (equation \ref{eq:MPSappx}) of acoustic sources vanishing outside a small region of space exhibits a threshold effect, in that the error in the resulting acoustic fields drops abruptly as the length of the series ($N$ in the expression \ref{eq:MPSappx}) is increased past a critical value.
This threshold in the number of terms necessary for accurate approximation depends on the size of the active region and a measure of energy output relative to energy input (related to the degree of anisotropy).
The analysis in \cite[]{SantosaSymes:00} pertains to acoustics, however we expect similar results to hold for linear elastodynamics.
Earthquake seismology has long used multipole source approximations in elastic media to describe earthquake mechanisms. 
The seismic moment tensor represents earthquake source mechanisms
 by a combinations of point-dipole body forces, that is, a first-order vector-valued multipole \cite[]{Backus:1976a,Shearer:2009}.
Moment tensors of higher order (that is, higher order multipole series) have been shown to be important in cases where finiteness of the source is of issue, that is the fault size is comparable to propagating wavelengths \cite[]{Stump:1982}.

A second natural question is: how much source anisotropy is really necessary to fit field data well?
\cite{SymMink:97} show that inverting marine reflection data for multipole source parameters together with a layered viscoelastic model, resulted in fitting 25\% more of the data than was possible with any isotropic source, and allowed 90\% data fit to the target portion of the data. 
Moreover, recovered p-wave and s-wave impedance parameters matched closely the expected seismic-lithologic signature of the gas sand only when viscoelastic model and anisotropic source parameters were simultaneously estimated.
The order of multipole required to achieve this degree of data fit and
well log tie was $N=6$. 
Obviously the necessary order depends on many factors, and varies from survey to survey. 
The framework we have provided here offers a platform in which to develop an approach to source anisotropy estimation as part of the overall inversion process.

% For example, we have shown in
%another place REFERENCE??? that a choice of norm in an MPS space, that
%reflects the effect of the spatial delta derivatives on temporal
%frequency, can dramatically accelerate the convergence of
%source recovery by Krylov subspace iteration. For this, it was only
%necessary to pass the {\tt RVL::LinOpValOp} representing the MPS-to-data 
%trace map, and an {\tt RVL:LinearOp}
%representing the Gram operator of the norm,  to an RVL implementation of the
%preconditioned conjugate gradient method, used completely without
%alteration. It will also be possible to incorporate the MPS based
%operator into the RVL implementation of variable projection method
%\cite[]{LiRickettAbubakar:13} for joint source-medium inversion
%without any change either to the MPS code or to the RVL variable
%projection algorithm.
%
%%We would be remised if we did not point out some of the attractive features present in this singular source approximation.
%%For one, incorporating source approximations is rather straight-forward and minimaly invasive to the finite difference implementation, a matter of cooking up the correct collection of RHS sources and injecting them at proper grid points.
%%Moreover, the relationship between our source parameterization via MPS coefficients and RHS source-terms for finite difference solvers is a linear one, aptly implemented by our \texttt{MPS\_to\_RHS} operator.
%%Ultimately the singular source approximation implemented in our framework preserves linearity of the MPS-to-data map in a natural way that will be crucial for source inversion.
%
%We have not addressed two important issues in the work reported here.
%The first is the order of multipole necessary to represent a given
%degree of source anisotropy. Point support is an idealization: active source regions may
%be small or comparable to a wavelength in spatial
%extent. \cite{SantosaSymes:00} showed that multipole approximation
%(equation \ref{eq:MPSappx}) of acoustic sources vanishing outside
%a small region of space
%exhibits a threshold effect, in that the error in the
%resulting acoustic fields drops abruptly as the length of the series
%($N$ in the expression \ref{eq:MPSappx})
%is increased past a critical value.
%This threshold in the number of terms necessary for accurate
%approximation depends on the size of the active region and a measure of energy
%output relative to energy input (related to the degree of
%anisotropy).
%The analysis in \cite[]{SantosaSymes:00} pertains to acoustics,
%however we expect similar results to hold for linear elastodynamics.
%Earthquake seismology has 
%long used multipole source approximations in elastic media 
%to describe earthquake mechanisms. 
%The seismic moment tensor represents earthquake source mechanisms
% by a combinations of point-dipole body forces, that is, a
%first-order vector-valued multipole \cite[]{Backus:1976a,Shearer:2009}.
%Moment tensors of higher order (that is, higher order multipole series) have been shown to be important in cases where finiteness of the source is of issue, that is the fault size is comparable to propagating wavelengths \cite[]{Stump:1982}.
%
%A second natural question is: how much source anisotropy is really
%necessary to fit field data well?
%\cite{SymMink:97} show that inverting marine reflection data for multipole source
%parameters together with a layered viscoelastic model, resulted in fitting 25\% 
%more of the data than was possible with any
%isotropic source, and allowed
%90\% data fit to the target portion of the data. Moreover, recovered
%p-wave and s-wave impedance parameters matched closely the expected
%seismic-lithologic signature of the gas sand only when viscoelastic model and anisotropic source parameters were simultaneously estimated.
%The order of multipole required to achieve this degree of data fit and
%well log tie was $N=6$. Obviously the necessary order depends on many
%factors, and varies from survey to survey. The framework we have
%provided here offers a platform in which to develop an approach to
%source anisotropy estimation as part of the overall inversion process.

%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
%%%%%%%%%%%%%%%%%%%%%

In this paper we have presented a framework for representing 
multipole sources as linear combinations of multipole bases.
Sources are thus uniquely specified by their multipole coefficients 
given an MPS space.
We also briefly discussed how to deal with the singularity of multipole 
source-terms in the context of finite difference solvers. 
In particular, we implemented the singular source approximation 
developed by \cite{Walden:1999}, \cite{TorEng:04}, and \cite{MPS_theory}
as a linear map from MPS-to-RHS source terms.
Mathematical constructs of the MPS space and the MPS-to-RHS 
map were realized using RVL and pipelined into finite difference solvers via IWave.
%Numerical results validate the correctness of our implementation by 
%showing an agreement between numerical and analytical solutions, particularly for
%3-D acoustics in first order form with an unbounded homogeneous medium.
Numerical experiments demonstrate the applicability of our computational framework for forward modeling and inversion of multipole sources, as well as source-medium joint inversion.

%\newpage
%%%%%%%%%%%
%\section{Appendix}
%
%Consider the system \ref{eq:acous} in an unbounded homogeneous medium with source term
%\[
%	f(\mathbf x,t) = w(t) D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)
%\]
%for some given multi-index $\mathbf s$.
%Under constant medium parameters, we can derive the second order wave equation
%from the first order system by taking an extra time derivative in the pressure equation
%and use the velocity equation to substitute the derivative in time of the velocity field, 
%thus yielding
%\begin{equation}
%\label{eq:acous_hom}
%	\frac{\partial^2}{\partial t^2} p - c^2 \nabla^2 p = w'(t) D^{\mathbf s}\delta(\mathbf x-\mathbf x^*),
%\end{equation}
%where $c^2 = \kappa \rho$.
%Note the extra time derivative on the source wavelet, due to the fact that the source
%term $f(\mathbf x,t)$ originated from the first order system.
%
%Let $p_{\mathbf s}$ denote the analytical solution to \ref{eq:acous_hom} with given multi-index $\mathbf s$. In particular, for $\mathbf s=\mathbf 0$ we have
%\[
%	p_{\mathbf 0}(\mathbf x,t) =  \frac{w'(\tau)}{4\pi c^2 r},
%\]
%where
%\[
%	\tau = t - \frac{r}{c}, \quad \text{and} \quad r = \sqrt{\mathbf x^T\mathbf x}.
%\]
%For $\mathbf s\neq \mathbf 0$, it follows that
%\[
%	p_{\mathbf s} = D^{\mathbf s}p_{\mathbf 0},
%\]
%which can be shown by leveraging the smoothness of $p_{\mathbf 0}$ away from its singularity at $\mathbf x^*$.
%The following table summarizes the analytical expressions for $p_{\mathbf s}$ used in the numerical experiments, where 
%\[
%	\gamma_i = \frac{x_i}{r}
%\]
%and $\delta_{ij}$ is the Kronocker delta;
%
%\begin{table}[!h]
%	\centering
%	\begin{tabular}{c | l }
%		  Source Term	& Analytical Solution  \\  \hline 
%		$w(t)\delta(\mathbf x)$ & $\displaystyle \frac{1}{4\pi c^2 r} w'(\tau)$ \\
%		$w(t)\frac{\partial}{\partial x_i}\delta(\mathbf x)$ 
%			& $ 	\displaystyle -\frac{1}{4\pi c^2} \left\{ 
%					\frac{1}{c}\frac{\gamma_i}{r} w''(\tau) +
%					\frac{\gamma_i}{r^2} w'(\tau) 
%				\right\} $ \\
%		$w(t)\frac{\partial^2}{\partial x_j\partial x_i}\delta(\mathbf x)$ 
%			& $ 	\displaystyle \frac{1}{4\pi c^2} \left\{ 
%					 \frac{\gamma_i\gamma_j}{c^2r} w'''(\tau) -
%					\frac{1}{cr^2} \left( \delta_{ij}-3\gamma_i\gamma_j \right) w''(\tau) -
%					\frac{1}{r^3}\left( \delta_{ij}-3\gamma_i\gamma_j \right) w'(\tau)					 
%				\right\} $
%	\end{tabular}
%	\caption{Analytical solutions to first-order acoustic equations in 3-D, with unbounded homogeneous medium and a single multipole source term.}
%	\label{tab:ansol}
%\end{table}

%%%%%%%%%%%%%%%%%%%%%%%
%\section{Appendix: Deriving Adjoint Action of \texttt{MPS\_to\_RHS}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Deriving the adjoint of the MPS-to-RHS map $\mathbb L$ is rather straight forward using the tensor notation developed in the theory section.
%First we must define the $L^2$-inner products of the input and output of $\mathbb L$\footnote{These are the inner product defined for the \texttt{SEGYSpace}s and tensor products of such spaces, and currently in use for domain and range of \texttt{MPS\_to\_RHS} operators.}:
%\[
%	\langle \mathbf F^{\text{MPS}}_1,\mathbf F^{\text{MPS}}_2 \rangle = 
%        \Delta t \;\sum_{i_t,i_m} \mathbf F^{\text{MPS}}_1[i_t,i_m] \times \mathbf F^{\text{MPS}}_2[i_t,i_m],
%\]
%\[
%	\langle \mathbf F^{\text{RHS}}_1, \mathbf F^{\text{RHS}}_2 \rangle = 
%        \Delta t \; \sum_{i_t,i_c,i_s} \mathbf F^{RHS}_1[i_t,i_c,i_s] \times \mathbf F^{RHS}_2[i_t,i_c,i_s].
%\]
%Using \ref{eq:L},
%\[
%\begin{array}{rcl}
%	\langle \mathbb L\mathbf F^{\text{MPS}}, \mathbf F^{\text{RHS}} \rangle 
%		&=& \Delta t\; \sum_{i_t,i_c,i_s} \left( \sum_{i_m} \mathbf W[i_m,i_c,i_s] \times \mathbf F^{\text{MPS}}[i_t,i_m] \right) \times \mathbf F^{\text{RHS}}[i_t,i_c,i_s] \\
%		&=& \Delta t\; \sum_{i_t,i_m} \mathbf F^{\text{MPS}}[i_t,i_m] \times \left( \sum_{i_c,i_s} \mathbf W[i_m,i_c,i_s] \times \mathbf F^{\text{RHS}}[i_t,i_c,i_s]\right)\\
%		&=& \langle \mathbf F^{\text{MPS}}, \mathbb L^\dagger \mathbf F^{\text{RHS}} \rangle,
%\end{array}
%\]
%from which we define adjoint of $\mathbb L$ by,
%\[
%	(\mathbb L^\dagger \mathbf F^{\text{RHS}})[i_t,i_m] = \sum_{i_c,i_s} \mathbf W[i_m,i_c,i_s] \times \mathbf F^{\text{RHS}}[i_t,i_c,i_s].
%\]
%The equation for the action of $L^\dagger$ states that the adjoint action of a \texttt{MPS\_to\_RHS} operators consists of linear combinations of time traces for different RHS components (index $i_r$) and stencil points (index $i_n$), point wise in time (index $i_t$) for each source (index $i_s$).

%%%%%%%%%
% PLOTS %
%%%%%%%%%

\inputdir{project}

%\plot{../../model-geo}{width=0.8\textwidth}{Model and source-receiver geometry specifications; source position $\mathbf x^*=(-298m,602m,303m)$, receiver position $\mathbf x_r = (-300m,2400m,300m)$.}
%
%\multiplot{2}{wlt-gauss,wlt-gauss-fft}{width=0.45\textwidth}{Gaussian wavelet (a) and its power spectrum (b).}
%
%\multiplot{2}{wlt-dgauss,wlt-dgauss-fft}{width=0.45\textwidth}{Derivative of Gaussian wavelet (a) and its power spectrum (b).}
%
%\multiplot{2}{wlt-ricker,wlt-ricker-fft}{width=0.45\textwidth}{Ricker wavelet (a) and its power spectrum (b).}

\multiplot{3}{howto-mps,howto-RHS,howto-data}{width=0.45\textwidth}{Plots from the how-to example: (a) input multipole coefficients, (b) output RHS or gridded source traces, (c) output pressure data.}

\plot{csq24}{width=\textwidth}{True medium parameter $\mathbf m=c(\mathbf x)^2$ (velocity squared), fraction of Marmousi model with water layer.}

\plot{data}{width=\textwidth}{Observed multi-shot data, in the form of pressure field time traces, for inversions.}

\plot{mps-towed}{width=\textwidth}{Multipole coefficients for generating the observed multi-shot data.}

\plot{test2-mps}{width=\textwidth}{Multipole source inversion results. Estimated $w_1(t)$ (dashed red curve) plotted against true solution (solid blue curve).}

\multiplot{3}{csq24sm,test4-csq-inv,trace-compare}{width=0.5\textwidth}{Joint source-medium inversion results. (a) Initial guess and (b) estimated wave velocity squared. (c) Single trace comparison of medium parameter at $3500m$ horizontal displacement.}

\plot{test4-mps}{width=\textwidth}{Joint source-medium inversion results. Estimated $w_1(t)$ (dashed red curve) plotted against true solution (solid blue curve).}

%\multiplot{3}{test3D-s000,test3D-s010,test3D-s020}{width=0.45\textwidth}{Pressure field at a single receiver location $\mathbf x_r$ induced by a single multipole source $f(\mathbf x,t) = w(t)D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$.
%Numerical solution given by dotted magenta curve.
%Analytical solution given by solid blue curve.
%Error given by dash-dot red curve.
%${\bf s} = (0,0,0)$ in (a), ${\bf s}= (0,1,0)$ in (b), and ${\bf s}=(0,2,0)$ in (c).}

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/mps_precond/paper.tex
repro docs dir added

\newcommand{\Langle}{\bigg\langle} 
\newcommand{\Rangle}{\bigg\rangle} 
\newcommand{\Lparan}{\bigg(} 
\newcommand{\Rparan}{\bigg)} 
\newcommand{\WE}{\Lparan\frac{\partial^2}{\partial t^2} - c^2\nabla^2 \Rparan} 
\def\mathbi#1{\textbf{\em{#1}}}
\def\bhat#1{\hat{\mathbf{#1}}}

\title{Preconditioner for Estimation of Multipole Sources via Full Waveform Inversion}
\date{}

%\lefthead{Huang \& Nammour \& Symes}

%\address{
%\footnotemark[1]The Rice Inversion Project,\\ Rice University,\\ Houston, TX
%77005-1892 USA\\
%%\footnotemark[2]Total E\&P R\&T USA
%}
%\author{Mario J. Bencomo\footnotemark[1] and William Symes\footnotemark[1]}
\author{Mario J. Bencomo and William W. Symes}

\righthead{multipole source estimation, preconditioner }

\maketitle
\parskip 12pt

%%%%%%%%%
\begin{abstract}

%Abstract needs to be updated ...

Accurate representation and estimation of seismic sources is crucial to the joint medium-source full waveform inversion problem. 
In the work presented here we focus on the source estimation subproblem for 2-D acoustics, where source terms are represented as truncated series of multipoles. 
Our main contribution lies in developing a preconditioner for the normal equations related to the source-parameter linear least squares problem.
The proposed preconditioner consists of fractional time derivative/integral operators based on the analytical solutions for unbounded, homogeneous medium with multipole sources.  
Numerical results demonstrate an acceleration of conjugate gradient iterates when our preconditioning scheme is used.

%%%%%%%%%
\end{abstract}

%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%

%Reworked intro, as of 05/22/19

Accurate representation and estimation of seismic sources is essential for the recovery of medium parameters in the seismic inverse problem.
Joint determination of medium and source parameters via \emph{full waveform inversion} (FWI) has been shown to be successful in many applications as a means of addressing the impact of source parameters on the estimated medium. 
For example, \cite{Wang:09} demonstrated the feasibility of joint medium-source parameter estimation for 2-D, constant density, acoustics with multiples in the data. 
Furthermore, their synthetic tests show how robust the source inversion is due to inaccuracies in the estimated medium by imposing constraints on the spectrum and time duration on the source time-dependence.
\cite{SymMink:97} further motivate the need for joint medium-source inversion and highlight the importance of taking into account the source anisotropy in their study of marine reflection data with plane-wave viscoelastic modeling.

In this paper, we model general anisotropic sources as a \emph{multipole} series involving derivatives of the spatial delta function,
\begin{equation}\label{eq:MPS}
	f(\mathbf x,t) = \sum_{|\mathbf s|=0}^N w_{\mathbf s}(t) D^{\mathbf s} \delta(\mathbf x-\mathbf x^*),
\end{equation}
for a source centered at $\mathbf x^*$, where $f(\mathbf x,t)$ plays the role of the source term (or inhomogeneous term) for a given wave equation; we present our modeling equations in full detail in the theory section.
Equation \ref{eq:MPS} makes use of multi-index notation:
given multi-index (integer $d$-tuple) $\mathbf s=(s_1,...,s_d)$, define the $\mathbf s$-mixed partial derivative operator $D^{\mathbf s}$ and its order $|\mathbf s|$ as follows,
\[
	D^{\mathbf s} = \prod_{k=1}^d \left(\frac{\partial}{\partial x_k}\right)^{s_k},
         \quad |\mathbf s| = \sum_{k=0}^d s_k,
\]
with $d\in\{1,2,3\}$ denoting the spatial dimension. 
We note that equation \ref{eq:MPS} is in fact the time-space analogue of the source-model considered by \cite{SymMink:97}.
Moreover, equation \ref{eq:MPS} is a generalization of the commonly used \emph{moment tensor} representation of anisotropic point-sources in earthquake seismology; see \cite{AkiRich:80}, \cite{Backus:1976a}, \cite{Shearer:2009} for a detailed discussion on seismic moment tensors.
Moment tensors of higher degree (related to multipole terms of higher order) have been shown to be important in cases where finiteness of the source is of issue, that is, the fault size is comparable to propagating wavelengths \citep{Stump:1982}.

Though the ultimate goal is joint medium-source inversion, we focus on the source estimation subproblem and its challenges in this paper:
We formulate the source inversion as FWI in the time domain, assuming medium parameters are known.
The novelty of our work is in the representation, and thus estimation, of seismic sources as multipoles.
We discuss and provide means to address the main challenge of ill-conditioning in the resulting inverse problem.
In particular, FWI yields a linear \emph{least squares} (LS) problem which we solve iteratively by \emph{conjugate gradient} (CG).
Our main contribution is to propose a preconditioner, based on (potentially fractional) derivative/integral operators, as a means to better condition the source inversion problem.

We first present a review of prior work on seismic source inversion, as well as joint medium-source estimation, in the context of FWI to motivate our methodology. 
In the following section we formulate the multipole source FWI problem and discuss the root causes of its ill-conditioning.
Lastly, we motivate and derive our proposed preconditioner and demonstrate its effectiveness in accelerating convergence of CG on some synthetic examples.

%%%%%%%
\subsection{Literature Review}

%Reworked lit review, as of 05/22/19

%Motivating source estimation in seismic inversion
Seismic data contains information about source and receiver responses as well as properties of the earth's medium, the latter being of priority in exploration seismology. 
Conventional methodology initially focused on removing the source response from data as a preprocessing step in seismic imaging by estimating the source signature through statistical methods like predictive deconvolution \citep{Rob:57} and homomorphic deconvolution \citep{Ulr:71}. 
\cite{Zio:91} has criticized these statistical methods for imposing unrealistic and at times theoretically unjustifiable constraints on both the source and medium, yielding results vulnerable to subjectivity. 
Alternatively, source signatures for the \emph{Vibroseis$^{TM}$} and airgun arrays have been estimated using near-field measurements with some success; see \cite{Zio:91} and \cite{Landro:92}.
It should be noted that source estimation with near-field measurements still depends on how the source is modeled and the data's dependency on the medium which can be unknown in applications where the source-receiver path is partially submerged in unknown medium or the direct arrival cannot be clearly discerned from the data.

%Joint medium-source inversion
Starting in the late 1980's, efforts in decoupling medium-source interactions shifted from preprocessing the source signature out of seismic traces to the joint inversion of source and medium parameters. 
Early attempts focused on proving theoretically the co-determinability of source time-dependent parameters and medium parameters from reflectivity data under simplifying assumptions: acoustic layered medium, quasi-impulsive and non-impulsive sources, primaries only data, inverting for only one medium parameter (reflectivity),  \citep{Ramm:85, Lew:89, BubeLaiSacksSanSy:88, MinkSym:95}. 
Later works have implemented and tested the feasibility of joint medium-source inversion beyond the theory with some success in a variety of synthetic and field examples; \citep{SymMink:97, Wang:09, Zhou:97}.

%VPM
In most cases, an FWI formulation of joint medium-source estimation results in a \emph{separable LS} problem in which the source parameters are quadratic with respect to the FWI objective function; we discuss this in the theory section with more detail.
Some recent works have employed the \emph{variable projection} (VP) method, first developed by \cite{GolubPereyra:73} (see \cite{GolubPereyra:03} for a more current discussion on the topic), to exploit the separable structure;
source parameters are effectively eliminated by an orthogonal projection when solving a linear LS subproblem.
In practice, VP, coupled with Gauss-Newton type algorithms for solving the primary nonlinear LS problem, has been proven to outperform other multiparameter estimation algorithms, \citep{Ruhe:1980}.
For example, \cite{Rickett:SEG12} and \cite{LiRickettAbubakar:13} demonstrate the effectiveness of VP over simultaneous descent and alternating direction in the context of joint medium-source inversion via FWI in the time and frequency domains respectively.
%Formulating the multipole source subproblem via FWI will allow us to couple our methodology to VP algorithms for joint medium-source estimation in future work as a means to mitigate the difficulty of multiparameter inversion.
%Future work regarding inversion of medium-source parameters will incorporate VP as a means to mitigate the difficulty of multiparameter inversion.

%Motivating anisotropic source modeling
All of the works mentioned above, with the exception of \cite{SymMink:97}, idealize the source contribution to that of an isotropic source localized at a single point in space, i.e., a point-source. 
The point-source assumption for seismic sources is justified by the fact that the spatial dimensions of sources considered in exploration seismology are considerably smaller (typically by an order of magnitude) than the propagating wavelengths of seismic waves. 
The isotropy assumption, however, is questionable, especially when the source is known or potentially expected to exhibit directivity or anisotropy in its radiation pattern. 
A perfect example is the airgun-array used in marine seismic surveying where the output pressure field of the source is known to exhibit vertical directivity. 
Work by \cite{SymMink:97} has been the prime motivator for this paper.
Their results highlight the necessity of modeling source anisotropy, as well as estimating source parameters in conjunction with the medium for an accurate inversion.
Anisotropy of their seismic source, which consisted of an airguns-arary, was modeled in the plane-wave domain by representing the source term as a series of Legendre polynomials in the slowness variable $p$, that is
\begin{equation}\label{eq:MPSradon}
	f(p,t) = \sum_{n=0}^N w_n(t)L_n(p),
\end{equation}
where $L_n$ is the $n$-th Legendre polynomial.
Their results show that inverting for anisotropic source parameters $\{w_n(t)\}$, as oppose to using a given isotropic source or even inverting for an isotropic source, allowed them to account for 25\% more of the data and were even able to achieve a dramatic 90\% data fit up to a gas-sand target.
Moreover, recovered reflectivity parameters matched closely expected lithology, but only when jointly estimating for medium and anisotropic source parameters.

%Prior work on anisotropic source inversion
Active work in anisotropic seismic source inversion primarily has focused on the determination of earthquake mechanisms \citep{Sipkin:82,Koch:91}, and most recently in the characterization of microseismic events (small earthquakes) resulting from hydraulic fracturing \citep{Song:11,Eaton:10}.
%These applications use the seismic moment tensor representation, which are related to the multipole source representation given in equation \ref{eq:MPS}.

Time-domain methods, based on an FWI approach, result in having to solve a block Toeplitz system of equations, which can be done efficiently with Levinson recursion algorithms \citep{Robinson:83,Trench:64}.
The equivalent problem in the frequency domain is no longer a simple Fourier division.
%, however, since we are dealing with \emph{multichannel deconvolution} (i.e., source being estimated has multiple components).
Instead, one has to solve a system of $M$ equations for $M$ unknowns point-wise in frequency, where $M$ is the number of source components we are inverting for.%; we discuss this in more detail later.
Some of the references cited above consider an alternate frequency-domain method, the so-called \emph{appraisal deconvolution} method by \cite{Oldenburg:82}.
Appraisal deconvolution consists of estimating, in the frequency domain, a set of optimal inverse filters, rather than source parameters.
What sets this method apart from a standard LS approach in the frequency-domain is the choice of objective function that defines the optimal inverse filters. 
In the scalar deconvolution case, one can show equivalency between appraisal deconvolution and a standard LS approach in time and frequency domain (also equivalent to Wiener deconvolution); this is no longer true for the general multichannel case.
Given that an inverse filter is computed, the appraisal deconvolution method readily offers access to qualitative measures of the inversion, primarily the sensitivity matrix.
On the other hand, this method requires the careful calibration of several hyper-parameters, and can be computationally intractable for moderate to large scale problems.

%\cite{Sipkin:82} estimates earthquake source parameters, mainly seismic moment (rate) tensor components, via FWI using two methods: a time-domain method equivalent to a standard LS formulation, and a frequency-domain method referred to as \emph{appraisal deconvolution}.
%The time-domain approach results in having to solve a block Toeplitz system which can be done efficiently with recursive solvers \citep{Robinson:83}.
%The appraisal deconvolution \citep{Oldenburg:82} consists of estimating for optimal filters in a LS sense in the frequency domain. 
%This method, however, requires one to tune several hyper-parameters, though it does offer easy access to qualitative measures of the inversion (e.g., the sensitivity matrix).

%Prior work on dealing with ill-conditioning of src problem
A particular challenge of the multipole source inverse problem, as we discuss in the next section, is ill-conditioning.
Some attempts at regularization have aimed to improve data coverage, as well as constraining or oversimplifying the source representation.
For example, \cite{Eaton:10} and \cite{Song:11} invert for time-independent moment tensor components and \cite{Eaton:10} further discusses ill-conditioning of the inverse problem with respect to angle subtended by the receiver array.
\cite{Koch:91} applies a more standard approach via Tikhonov regularization.
Their numerical results indeed demonstrate the benefits of this method, especially in cases where data acquisition geometry results in known ill-conditioning. 
Furthermore, \cite{Koch:91} illustrated robustness of their results against errors in time lag, model parameters (i.e., velocity model), and extended sources (non-point sources). 
It is worth highlighting that their regularization parameter is chosen in a trial by error manner given information about the singular value decomposition of the forward map, which could be impractical for larger problems.
It should also be noted that time-domain algorithms, based on Levinson recursion, are weakly stable \citep{Bunch:85}, making them ill-suited for ill-conditioned problems unless adequate regularization is introduced.
The appraisal deconvolution method offers a form of regularization (equivalent to Tikhonov regularization) through a ``trade-off'' parameter, though there is no efficient manner in setting said parameter.

%Motivating our approach
Formulating the multipole source subproblem via FWI allows us to couple our methodology to VP algorithms for joint medium-source inversion in future work, giving us a means of mitigating the difficulty of multiparameter inversion.
We decide to solve the multipole source inversion problem iteratively via CG in the time-domain, as done by \cite{SymMink:97} and \cite{Wang:09}.

%An iterative approach offers an efficient method of solving the source 
%regularizing via CG early stop -> more control in accuracy of src estimation subproblem (relevant for joint inversion and VPM)
%%  - can determine Tikhonov reg param iteratively
%A lot of the methodology from the previously cited works fail to address this challenge in a manner that is efficient to larger-scale problems.

%%Prior work on anisotropic source inversion and multichannel deconvolution
%As alluded to earlier, the source estimation can be framed as a deconvolution problem; we formalize this in the following section.
%In fact, an FWI formulation of the source estimation subproblem can be shown to be equivalent to deconvolution under simplifying assumption in the data noise.
%%FWI with Tikhonov regularization is equivalent to Wiener deconvolution.
%In the case where the source term consists of a single component, we have a scalar deconvolution problem which consists of a solving Toeplitz systems in the time domain or applying a regularized form Fourier division in the frequency domain.
%The multipole source case, however, results in a \emph{multichannel deconvolution} problem which is no longer trivial to solve in the frequency domain.
%In the time-domain, one is required to solve a block Toeplitz system.

%time vs frequency domain
%iterative vs direct solvers
%issues with ill-conditioning/posedness and how to alleviate it

%Benefits of our method (time-domain iterative solver via CG for source estimation subproblem):
%  - easy to time window the source function
%  - regularizing via CG early stop -> more control in accuracy of src estimation subproblem (relevant for joint inversion and VPM)
%  - can determine Tikhonov reg param iteratively
%  - consistent with our time-domain approach

%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory}
%%%%%%%%%%%%%%%%%%%%%%%%%

We model seismic wave phenomena via the acoustic equations, particularly in the first-order form,
\begin{equation}\label{eq:acous1}
\begin{split}
		\frac{\partial}{\partial t} \mathbf v(\mathbf x,t) +  
		\beta(\mathbf x) \nabla p(\mathbf x,t) &= 0, \\
		\frac{\partial}{\partial t} p(\mathbf x,t) + 
		\kappa(\mathbf x)\nabla \cdot \mathbf v (\mathbf x,t) &= f(\mathbf x,t),
\end{split}
\end{equation}
with
\begin{itemize}
	\item $\beta = $ buoyancy (reciprocal of density), 
	\item $\kappa = $ bulk modulus,
	\item $p = $ pressure scalar field, 
	\item $\mathbf v = $ particle velocity vector field,
	\item $f = $ (pressure) source term,
\end{itemize}
for $\mathbf x\in \Omega\subset\mathbb R^d$ and $t\in[0,T]$, with homogeneous boundary and initial conditions, when appropriate.
Our choice of PDE for modeling seismic waves is predicated on its simplicity and applicability. 
Furthermore, our proposed preconditioning strategy, related to this first-order system, will have a natural extension to elasticity with related velocity-stress equations and will be explored in future work.

Note that the source term $f$ is introduced as a scalar in the pressure equation only, though we could have just as easily introduced a vector force in the velocity equations.
Indeed, the relationship among corresponding velocity-pressure (or stress in linear elasticity) source terms will be of interest when modeling seismic sources, however this paper will abstain from such discussions given its intended scope. 
Moreover, though we focus on pressure sources for brevity, our approach can easily be applied to velocity sources.

We represent source term $f$ as a slightly different but equivalent multipole series, after some reordering for ease of enumeration, 
\begin{equation}\label{eq:MPSrep}
	f(\mathbf x,t) = \sum_{j=1}^M w_j(t) D^{\mathbf s_j} \delta(\mathbf x-\mathbf x^*).
\end{equation}
Note that multipole source $f$ as stated above, given multi-indexes $\mathbf s_j$ and source location $\mathbf x^*$, is uniquely determined by its time-dependent functions $w_j(t)$ which we refer to as {\em multipole coefficients}.
%In particular, multipole sources are linear combinations of multipoles $D^{\mathbf s_j}\delta(\mathbf x-\mathbf x^*)$ with time varying coefficients $w_j(t)$, which we refer to as {\em multipole coefficients}.

In our discussion of analytical solutions, and the derivation of our preconditioning scheme, we will assume we are dealing with a homogenous and unbounded medium, i.e., $\beta$ and $\kappa$ are constants and $\Omega=\mathbb R^d$.
Moreover, in our analysis we consider the second order form of the wave equation \begin{equation}\label{eq:WE}
	\WE p(\mathbf x,t) = \tilde f(\mathbf x,t),
\end{equation}
where $c=\sqrt{\kappa \beta}$ is the speed of wave propagation and $\tfrac{\partial}{\partial t} f(\mathbf x,t) = \tilde f(\mathbf x,t)$;
the pressure field from the first order system \ref{eq:acous1} is also a solution to the second order wave equation as described here.

%%%%%
\subsection{Analytical Solutions to the Wave Equation with Multipole Sources}

The following theorem states that the solution of wave equation \ref{eq:WE}, with multipole source \ref{eq:MPSrep}, can be written as a multichannel convolution of multipole coefficients $\tilde w_j(t) = \tfrac{d}{dt}w_j(t)$ with some kernels $g_j(\mathbf x,t)$.
Without loss of generality, we assume $\mathbf x^*=0$.

\begin{theorem}\label{thm:MPSsol}
	The solution $p$ of wave equation \ref{eq:WE}, with multipole source \ref{eq:MPSrep}, is given by
	\begin{equation}\label{eq:multiconv}
		p(\mathbf x,t) = \sum_{j=1}^M g_j(\mathbf x,t) * \tilde w_j(t),
	\end{equation}
	where `$*$' denotes convolution in time, and $g_j(\mathbf x,t)$ satisfy the following wave equation,
	\[
		\WE g_j(\mathbf x,t) = \delta(t) D^{\mathbf s_j}\delta(\mathbf x).
	\]
	Furthermore, 
	\[
		g_j(\mathbf x,t) = D^{\mathbf s_j} g(\mathbf x,t)
	\]
	where $g$ is the solution to the wave equation with an impulsive point source, i.e.,
	\[
		\WE g(\mathbf x,t) = \delta(t) \delta(\mathbf x).
	\]
\end{theorem}
\begin{proof}
	With some forethought, we clarify that partial derivatives of convolution kernels $g_j$ and Green's function $g$ are taken in a weak (distribution) sense given that these functions may be singular, in fact distributions.
	Moreover, satisfaction of the wave equation is also taken in a weak sense since we are dealing with singular sources, even when the resulting pressure field $p$ may be smooth.

	Assume that $p$ is given by equation \ref{eq:multiconv}.
	We apply the wave equation operator, multiply with a test function $\phi$ that is smooth and has compact support in space and time (denoted by $\phi\in C_0^\infty$), and integrate:
	\[
		\Langle \WE p(\mathbf x,t),\phi(\mathbf x,t) \Rangle 
		= \sum_{j=1}^M \Langle  \left(\frac{\partial^2}{\partial t^2}g_j - c^2\nabla^2g_j\right)(\mathbf x,t)*\tilde w_j(t), \phi(\mathbf x,t) \Rangle,
	\]
	where $\langle \cdot\rangle$ denotes the space-time inner product.
	Given that $g_j$ satisfy the wave equation with an particular multipole source, we have
\begin{equation*}
\begin{split}		
	\Langle\WE p(\mathbf x,t),\phi(\mathbf x,t) \Rangle
		&= \sum_{j=1}^M \Langle D^{\mathbf s_j}\delta(\mathbf x)\delta(t)*\tilde w_j(t), \phi(\mathbf x,t) \Rangle \\
		&= \Langle \sum_{j=1}^M \tilde w_j(t) D^{\mathbf s_j}\delta(\mathbf x), \phi(\mathbf x,t) \Rangle
\end{split}
\end{equation*}
for all $\phi \in C_0^\infty$, i.e., the desired result.

To obtain the second part of the theorem we use the definition of derivatives for distributions, the fact that the wave equation operator is a differential operator of even order, and that differential operators commute on smooth test functions:
\begin{equation*}
\begin{split}		
	\Langle\WE D^{\mathbf s_j} g(\mathbf x,t),\phi(\mathbf x,t) \Rangle
		&= \Langle g(\mathbf x,t), (-1)^{|\mathbf s_j|} D^{\mathbf s_j} \WE \phi(\mathbf x,t) \Rangle \\
		&= \Langle g(\mathbf x,t), \WE(-1)^{|\mathbf s_j|} D^{\mathbf s_j} \phi(\mathbf x,t)\Rangle \\
		&= \Langle D^{\mathbf s_j}\WE g(\mathbf x,t), \phi(\mathbf x,t) \Rangle \\
		&= \Langle \delta(t) D^{\mathbf s_j} \delta(\mathbf x), \phi(\mathbf x,t)\Rangle,
\end{split}
\end{equation*}
for all $\phi\in C_0^\infty$.
\end{proof}

The Green's function $g(\mathbf x,t)$, in 1D, 2D and 3D space, is given in table \ref{tab:Gfun}.
Note that none of these Green's functions are differentiable in the classical sense, and thus weak derivatives must be carefully applied when deriving expressions for $g_j(\mathbf x,t)$.
Alternatively, we can leverage the smoothness of the resulting pressure field from the monopole case, i.e., $f(\mathbf x,t) = w(t)\delta(\mathbf x)$, to compute the solution of the wave equation with multipole sources. 
In particular, it follows from theorem \ref{thm:MPSsol} that we can express $p(\mathbf x,t)$ as the linear combination of spatial derivatives of other pressure fields $p_j(\mathbf x,t)$:
\[
	p(\mathbf x,t) = \sum_{j=1}^M D^{\mathbf s_j} p_j(\mathbf x,t)
\]
where
\[
	p_j(\mathbf x,t) = g(\mathbf x,t) * \tilde w_j(t).
\]

\begin{table}[!h]
	\centering
	\begin{tabular}{c | c }
		  	& $g(\mathbf x,t)$ \\  \hline 
		1-D & $\displaystyle \frac{1}{2c} H(t-\tfrac{|x|}{c})$ \vspace{5pt}\\ \hline \vspace{5pt}
		2-D & $\displaystyle \frac{1}{2\pi c} \frac{H(t-\frac{|\mathbf x|}{c})}{\sqrt{c^2t^2 - |\mathbf x|^2}}$ \\ \hline
		3-D & $\displaystyle \frac{1}{4\pi c^2} \frac{\delta(t-\tfrac{|\mathbf x|}{c})}{|\mathbf x|} $
	\end{tabular}
	\caption{Green's functions for wave equation.}
	\label{tab:Gfun}
\end{table}

We summarize in table \ref{tab:AnSol} the analytical solution to the wave equation, though related to the first-order system (thus the occurrence of $w_j(t)$ instead of $\tilde w_j(t)$), with a single multipole source term, where
\[
	\tau = t - \frac{|\mathbf x|}{c},
	\quad
	\gamma_k = \frac{x_k}{|\mathbf x|},
	\quad
	\Omega(\mathbf x,\sigma) = \left( \sigma^2 + 2\frac{|\mathbf x|}{c} \right)^{-1/2},
\]
\[
	{\rm sgn}(x) = \left\{ \begin{array}{rl}
		-1,& x<0\\
		1,& x>0
	\end{array}\right.,
	\quad 
	\delta_{kl} = \left\{ \begin{array}{rl}
		1,& k=l\\
		0,& k\neq l 
	\end{array}\right..
\]
The 1D solutions, interestingly enough, show that the pressure field is only smooth in the monopole case, with discontinuous and singular pressure responses for dipole and quadrapole sources respectively. 
The 2D\footnote{Given the complexity of the Green's function in 2D, we provide a more detailed derivation of analytical solutions in the appendix.} and 3D solutions, however,  are indeed smooth\footnote{Assuming $w(t)$ is sufficiently smooth.}.
In all cases, the pressure field due to a monopole is indeed isotropic, that is $p(\mathbf x,t)$ is a function of $|\mathbf x|$ and time only.
Furthermore, pressure field response of a dipole point source is evidently anisotropic given the $\rm{sgn}(x)$ factor in 1D and the $\gamma_k$ factor 2D and 3D.

\begin{table}[!h]
\renewcommand{\arraystretch}{2}
	\centering
	\small
	\begin{tabular}{c | c | l }
	& \multicolumn{1}{c|}{Source Term $f$} & \multicolumn{1}{l}{Analytical Solution $p$}\\ \hline\hline
	\parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{1-D}}} 
		&$w(t)\delta(x)$ 				& $ \displaystyle \frac{1}{2c} w(\tau)$ \\
		&$w(t)\frac{d}{dx}\delta(x)$ 		& $ \displaystyle -\frac{1}{2c^2} w'(\tau) \; \text{sgn}(x)$ \\
		&$w(t)\frac{d^2}{dx^2}\delta(x)$	& $ \displaystyle \frac{1}{2c^3} w''(\tau) - \frac{1}{c^2} w'(\tau) \delta(x)$\\ 
	\hline\hline
	\parbox[t]{2mm}{\multirow{2}{*}{\rotatebox[origin=c]{90}{2-D}}} 
		&$w(t)\delta(\mathbf x)$ & $ \displaystyle \frac{1}{\pi c^2} \int_0^{\sqrt{\tau}} w'(\tau-\sigma^2)\Omega(\mathbf x, \sigma) \; d\sigma$\\
		&$w(t) \frac{\partial}{\partial x_k} \delta(\mathbf x)$ 
		& $\displaystyle-\frac{\gamma_k}{\pi c^3}  \int_{0}^{\sqrt{\tau}} 
		\Big\{ w''(\tau-\sigma^2) \Omega(\mathbf x,\sigma)
			+ w'(\tau-\sigma^2)\Omega^3(\mathbf x,\sigma) \Big\}\; d\sigma $ \\
		\hline\hline
	\parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{3-D\;\;}}} 
		&$w(t) \delta(\mathbf x)$ & $\displaystyle \frac{1}{4\pi c^2 |\mathbf x|} w'(\tau)$ \\
		&$w(t) \frac{\partial}{\partial x_k}\delta(\mathbf x)$ 
			& $ 	\displaystyle -\frac{\gamma_k}{4\pi c^2} \left\{ 
					\frac{1}{c|\mathbf x|} w''(\tau) +
					\frac{1}{|\mathbf x|^2} w'(\tau) 
				\right\} $ \\
		&$w(t)\frac{\partial^2}{\partial x_l\partial x_k}\delta(\mathbf x)$ 
			& $ 	\displaystyle \frac{1}{4\pi c^2} \left\{ 
					 \frac{\gamma_k\gamma_l}{c^2|\mathbf x|} w'''(\tau) -
					\frac{\delta_{kl}-3\gamma_k\gamma_l}{c|\mathbf x|^2} w''(\tau) -
					\frac{\delta_{kl}-3\gamma_k\gamma_l}{|\mathbf x|^3} w'(\tau)					 
				\right\} $
	\end{tabular}
    \caption{Analytical solutions to 1D, 2D and 3D acoustic equations in first-order form with multipole sources in homogeneous unbounded medium.}
    \label{tab:AnSol}
\end{table}

For 1D and 3D it is quite straight forward to make a connection between multipole coefficient $w(t)$ and the analytical pressure field $p(\mathbf x,t)$, if we focus on the far-field terms, that is, terms that dominate as $|\mathbf x|\to\infty$.
Essentially, the pressure field will consist of the some derivatives of $w$ evaluated at $\tau$, and scaled by $1/c$.
The higher the multipole order, the more derivatives and factors of $1/c$ are tacked onto multipole coefficients $w(t)$.
In 2D, we still have derivatives of $w$ and scaling by $1/c$ but with the added complexity of an integral.
It turns out that the differ-integral form for the 2D analytical solution is quiet similar to the (Caputo) fractional derivative; the fractional derivative of order $m-1/2$ for $m\in\mathbb N$ is defined as
\[
        \mathcal{D}^{m-1/2}_C w(t) := \frac{1}{\Gamma(\tfrac{1}{2})} \int_{0}^t \frac{w^{(m)}(t')}{ \sqrt{t-t'}} \; dt'.
\]
As qualitative proof, we plot the output pressure field $p$ and fractional time derivatives of input multipole coefficient $w$ in figure \ref{fig:test0-data-0,test0-D-0,test0-data-1,test0-D-1,test0-data-2,test0-D-2}, related to multipole source $f(\mathbf x, t) =  w(t) D^{\mathbf s}\delta(\mathbf x - \mathbf x^*)$ in a 2D homogeneous unbounded acoustic medium for $\mathbf s = (0,0), (0,1), (0,2)$.
Multipole coefficient $w(t)$ was a 10Hz Ricker wavelet, see figure \ref{fig:test0-MPS,test0-MPS-spec}. 
Source location is given by $\mathbf x^* = (-1km,1km)$, and pressure fields are measured at $\mathbf x = (-0.8km,0.8km)$. 
Plots show almost identical waveforms when comparing pressure fields to fractional derivatives of multipole coefficient $w$, modulo a scaling factor.

%%%%%
\subsection{FWI Formulation}

We now state multipole source inversion as a linear LS problem, and introduce some basic notation.
We first group multipole coefficients to form the {\em multipole coefficient vector} $\mathbf w$,
\[
	\mathbf w(t) = \{w_j(t) : j=1,...,M\},
\]
which has a natural vector space structure if we endow such vectors with the following obvious definition of linear combinations: for two multipole coefficient vectors $\mathbf w^1$ and $\mathbf w^2$, and scalars $a_1,a_2$,
\[
	a_1\mathbf w^1(t) + a_2 \mathbf w^2(t) = \{a_1w^{1}_j(t) + a_2 w^{2}_j(t) : j=1,...,M\}.
\]
The space of multipole coefficient vectors, $\mathfrak W$, is referred to as the {\em multipole source (MPS) space}.
Note that we regard source position $\mathbf x^*$ and multipole basis (i.e., multi-indexes $\mathbf s_j$) as attributes of an MPS space.

Similarly, we group pressure field traces sampled at receiver points $\{\mathbf x_i : i=1,2,...,N\}$ to form the {\em data vector} $\mathbf d$,
\[
	\mathbf d = \{ p(\mathbf x_i,t) : t\in[0,T], i=1,...,N\}.
\]
We let $\mathfrak D$ denote the space of data vectors, or simply the {\em data space}.
Furthermore, we define a natural $L^2$ inner-product for MPS and data spaces as follows,
\begin{equation}\label{eq:Wprod}
	\langle \mathbf w^1,\mathbf w^2 \rangle_{\mathfrak W} := \sum_{j=1}^M \int_{0}^T w^1_j(t) w^2_j(t) \;dt,
\end{equation}
\begin{equation}\label{eq:Dprod}
	\langle \mathbf d^1,\mathbf d^2 \rangle_{\mathfrak D} := \sum_{i=1}^N \int_{0}^T p^1(\mathbf x_i,t) p^2(\mathbf x_i,t) \; dt.
\end{equation}

Define the \emph{forward map} as an operator $F:\mathfrak M\times \mathfrak W \to \mathfrak D$ from medium-source parameter space $\mathfrak M\times \mathfrak W$ to data space $\mathfrak D$.
In other words,
\[
	F(\mathbf m,\mathbf w) = \mathbf d
\]
where $\mathbf m\in\mathfrak M$ denotes medium parameters (possibly $(\beta, \kappa)$ or other parameterizations).
It was shown in the previous subsection that the forward map is linear with respect to source parameters.
Linearity of $F$ with respect to source parameters will be emphasized for convenience by the following notation,
\[
	F(\mathbf m,\mathbf w) \equiv \mathbf F \mathbf w,
\]
where $\mathbf F:\mathfrak W\to\mathfrak D$ has a tacit dependence on $\mathbf m$.
We can go a step further and give a more explicit form of $\mathbf F$ using the multichannel representation in theorem \ref{thm:MPSsol}.
In particular, we can view $\mathbf F$ as an $N\times M$ matrix of linear operators,
\[
	\mathbf F = \mat{  F_{11} & \cdots &  F_{1M}\\
				   \vdots & \ddots & \vdots \\
				    F_{N1} & \cdots &  F_{NM} }
\]
 where 
\[
	\mathbf (F_{ij} w_j)(t) = g_j(x_i,t)*w_j(t).
\]

The multipole source FWI problem consists of finding $\mathbf w^*$ that minimizes the error between observed and predicted data in a LS sense:
\begin{equation}\label{eq:FWI}
	\mathbf w^* = \underset{\mathbf w\in\mathfrak W}{\text{argmin}} \; \| \mathbf F \mathbf w - \mathbf d_{obs}\|^2_{\mathfrak D},
\end{equation}
where $\|\cdot\|_{\mathfrak D}$ is the norm induced by the data space inner product defined in equation \ref{eq:Dprod}.
Solution to minimization problem \ref{eq:FWI} is given by the solution to the normal equations
\begin{equation}\label{eq:NEQ}
	\mathbf F^T \mathbf F \mathbf w^* = \mathbf F^T \mathbf d_{obs},
\end{equation}
$\mathbf F^T$ denoting the adjoint of $\mathbf F$ with respect to source parameter and data spaces and their respective inner products.

We also consider the regularized linear inverse problem, as done in \cite{Koch:91}:
\begin{equation}\label{eq:FWIreg}
	\mathbf w^*_{\lambda} = \underset{\mathbf w\in\mathfrak W}{\text{argmin}} \; \| \mathbf F \mathbf w - \mathbf d_{\rm obs}\|^2_{\mathfrak D} + \lambda \| \mathbf w\|^2_{\mathfrak W},
\end{equation}
equivalent to solving the modifyed the normal equations to
\begin{equation}\label{eq:RegNEQ}
	\Big( \mathbf F^T \mathbf F + \lambda\mathbf I\Big) \mathbf w^*_{\lambda} = \mathbf F^T \mathbf d_{\rm obs},
\end{equation}
where $\lambda$ is the damping constant, also referred to as the regularization parameter.

%%%%
\subsection{1D Analytical Case}

Consider the 1D case, with multipole source
\begin{equation}\label{eq:1Ddipole}
	f(x,t) = w_1(t)\delta(x) + w_2(t) \frac{d}{dx}\delta(x).
\end{equation}
To simplify the analysis, we assume the medium is homogenous and unbounded and pose the inversion in the frequency domain.
Applying the Plancherel theorem results in the following equivalent LS problem: find $\bhat{w}^*$ such that
\[
	\bhat{w}^* = \underset{\bhat{w}}{\rm argmin} \| \bhat{F}\bhat{w} - \bhat{d}_{\rm obs}\|^2_{\mathfrak D}
\]
where $\hat w_j(\omega)$ denotes the Fourier transform of $w_j(t)$,
\[
	\hat w_j(\omega) = \int_{\mathbb R} w_j(t) e^{-2\pi i t \omega} \; dt
\]
and similarly for other functions.
It follows from the convolution theorem that the forward map in the frequency domain, $\bhat{F}$, consists of simply matrix multiplication, point-wise in frequency.
In particular,
\[
	\bhat F = \mat{ \hat g_1(x_1,\omega) & \cdots & \hat g_M(x_1,\omega) \\
			        \vdots & \ddots & \vdots \\
			        \hat g_1(x_N,\omega) & \cdots & \hat g_M(x_N,\omega) }.
\]
One still has to solve a system of normal equations, analogous to that of \ref{eq:NEQ}, though point-wise in frequency:
\[
	\bhat{F}^H\bhat{F} \bhat w^* = \bhat{F}^H \bhat{d}_{\rm obs}
\]
where $\bhat{F}^H$ denotes the conjugate transpose of $\bhat{F}$.

For dipole source \ref{eq:1Ddipole}, we have $M=2$ and
\[
	g_1(x,t) = \frac{1}{2c} \delta(t-\tfrac{|x|}{c}),
\]
\[
	g_2(x,t) = \frac{1}{2c^2}\delta'(t-\tfrac{|x|}{c})\rm{sgn}(x),
\]
in accordance to theorem \ref{thm:MPSsol}.
Thus,
\[
	\bhat{F} = \frac{1}{2c} \mat{ {\phi}(\omega), &
						     -\frac{2\pi i}{c} \omega \psi(\omega)}
\]
with
\[
	\phi(\omega) = \mat{{\rm exp}(-2\pi i \tfrac{|x_1|}{c}\omega)\\
					\vdots \\
					{\rm exp}(-2\pi i \tfrac{|x_N|}{c}\omega)}
\]
and 
\[
	\psi(\omega) = \mat{ {\rm exp}(-2\pi i \tfrac{|x_1|}{c}\omega) \; {\rm sgn}(x_1) \\
					\vdots\\
					{\rm exp}(-2\pi i \tfrac{|x_N|}{c}\omega) \; {\rm sgn}(x_N)}
\]
Moreover, the normal operator $\bhat{F}^H\bhat{F}$ is given by
\begin{equation*}
\begin{split}
	\bhat{F}^H\bhat{F} 
	&= 
		\frac{1}{4c^2}\mat{ \phi(\omega)^H \phi(\omega) & 
		\frac{2\pi i}{c}w \phi(\omega)^T \psi(\omega) \\
		-\frac{2\pi i}{c} \omega \psi(\omega)^H \phi(\omega) &
		\frac{4\pi^2}{c^2} \omega^2 \psi(\omega)^H\psi(\omega) }\\
	&=
		\frac{1}{4c^2}\mat{N & \frac{2\pi i}{c}\omega \sum_{i=1}^N {\rm sgn}(x_i) \\
		-\frac{2\pi i}{c}\omega \sum_{i=1}^N {\rm sgn}(x_i) & \frac{4\pi^2}{c^2} w^2 N}.
\end{split}
\end{equation*}

Consider the case where all of the receiver points $\{x_i\}_{i=1}^N$ lie either to the right or left of the source, resulting in $\sum_{i=1}^N {\rm sgn}(x_i) = N$ or  $-N$ respectively.
For this case, we have that $\bhat{F}^H\bhat{F}(\omega) \in \mathbb C^{2\times 2}$ is rank one and thus singular.
In other words, such data coverage fails to capture any of the anisotropy, making the monopole term (isotropic source term) indistinguishable from the dipole term.
This placement of receiver points $\{x_i\}_{i=1}^N$ constitutes the worst case scenario resulting in ill-posed inverse problem.
The optimal data coverage scenario consists of having half of receiver points on the left and right of the source.
Assuming we have an even number of receiver points, this leads to $\sum_{i=1}^N {\rm sgn}(x_i) = 0$, and
\[
	\bhat{F}^H\bhat{F} = \frac{1}{4c^2}\mat{ N & 0 \\
		0 & \frac{4\pi^2}{c^2} \omega^2 N}.
\]
Normal operator $\bhat{F}^H\bhat{F}$ is clearly invertible in the case of optimal data coverage, though with a condition number that is quadratic in wavelength, i.e., 
\begin{equation}\label{eq:cond1}
	\kappa(\bhat{F}^H\bhat{F}) = \mathcal O(\tfrac{\omega^2}{c^2}) = \mathcal O(\lambda^{-2}).
\end{equation}

Given our discussion of analytical solutions of the wave equation with multipole sources, we can extrapolate result \ref{eq:cond1} for source with higher order multipoles. 
Suppose source term $f$ contains a monopole term and that the highest order multipole is of order $M^*$, then we have that $\kappa(\bhat{F}^H\bhat{F}) = \mathcal O(\lambda^{-2M^*})$.
This result follows from observing that every extra derivative on the spatial delta function in the multipole source incurs a factor of $\omega/c$ in the Fourier transform of the kernel $\hat g_j$.
Moreover, given that the Fourier transform is a unitary operator, we can conclude that the original forward map in the time domain, $\mathbf F$, has condition number $\mathcal O(\lambda_{\rm min}^{M^*})$, where $\lambda_{\rm min}$ is the smallest wavelength, assuming that observed data is bounded in the frequency domain.

Clearly, data coverage has a direct impact on the well-posedness and conditioning of the inverse problem.
However, the analysis above reveals that multipole source inversion is inherently, and potentially, highly ill-conditioned.
This ill-conditioning stems from the singular nature of the multipole source terms, which translate into a pressure field response as derivatives of the input waveform $\mathbf w(t)$.

\newpage

%%%%%
\subsection{Preconditioner}

%Solution to normal equations \ref{eq:NEQ}, in the context of moment tensor inversion, has been shown to suffer from numerical instability stemming from $\mathbf F$ having a large condition number. 
%Most attempts at regularization have aimed to improve the condition number by optimizing data coverage, as well as constraining or oversimplifying the source representation; see \cite{Eaton:10} and \cite{Song:11} for examples of moment tensor inversions from microseismic data.
%\cite{Koch:91} applies a more general approach to improving the condition number, namely by introducing a damping term in the underlying linear LS problem (i.e., Tikhonov regularization), thus modifying the normal equations to
%\begin{equation}\label{eq:RegNEQ}
%	\Big( \mathbf F^T \mathbf F + \lambda\mathbf I\Big) \mathbf w = \mathbf F^T \mathbf d_{\rm obs},
%\end{equation}
%where $\lambda$ is the damping constant also referred to as the regularization parameter.
%Their numerical results indeed demonstrate the benefits of this method, especially in cases where data acquisition geometry results in known ill-conditioning.
%Furthermore, \cite{Koch:91} illustrated robustness of their results against errors in time lag, model parameters (i.e., velocity model), and extended sources (non-point sources).
%It is worth highlighting that $\lambda$ is chosen in a trial by error manner given information about the singular value decomposition of $\mathbf F$ which could be impractical for larger problems.

The approach developed here seeks to better condition the multipole source inversion by redefining the domain space $\mathfrak W$ to yield a better bounded operator $\mathbf F$, hence improving the condition number associated with solving the normal equations.
We redefine $\mathfrak W$ by replacing its native inner-product $\langle\cdot,\cdot\rangle_{\mathfrak W}$ by a weighted one denoted by $\langle\cdot,\cdot\rangle_{\mathfrak W'}$,
\[
	\langle \mathbf w^1, \mathbf w^2 \rangle_{\mathfrak W'} := \langle \mathbf Q\mathbf w^1, \mathbf Q \mathbf w^2 \rangle_{\mathfrak W} \equiv \langle \mathbf w^1, \mathbf Q^T\mathbf Q \mathbf w^2 \rangle_{\mathfrak W}.
\] 
Weight operator $\mathbf Q$ will consists of a diagonal operator with respect to multipole coefficients, i.e.,
\[
	\mathbf Q \mathbf w = \text{diag}(Q_1 w_1, \cdots, Q_M w_M)
\]
for some linear operators $Q_j$.
Note that we can express the adjoint of $\mathbf F$ with respect to our weighted inner-product $\langle\cdot,\cdot\rangle_{\mathfrak W'}$ in terms of $\mathbf F^T$, namely
\[
	\mathbf F^T \to (\mathbf Q^T\mathbf Q)^{-1} \mathbf F^T.
\]
The LS problem with respect to our new inner-product results in the following modified normal equations,
\begin{equation}\label{eq:PNEQ}
	\mathbf M\mathbf F^T\mathbf F \mathbf w = \mathbf M\mathbf F^T \mathbf d_{\rm obs},
\end{equation}
where we have effectively (left) preconditioned by $\mathbf M = (\mathbf Q^T\mathbf Q)^{-1}$.
Implementation wise, we take the preconditioning approach instead of actually weighting the MPS space inner product.

The effectiveness of our preconditioner hinges on a proper choice of $Q_j$: a linear operator that is numerically stable, has an inverse we can readily implement, and most importantly better-conditions our normal equations.
Based on our discussion of analytical solutions to the wave equation we choose $Q_j$ in general to be a temporal integro-differential operator, more precisely a fractional derivative/integral, scaled by wave speed $c=\sqrt{\beta\kappa}$,
\[
	Q_j = \frac{1}{c^{|\mathbf s_j|}} \left( \frac{d}{d t} \right)^{\alpha_j}, \quad \text{with} \quad \alpha_j = \gamma + |\mathbf s_j| \in\mathbb R.
\]
The value of $\gamma$ is potentially non-integer and dependent on the spatial dimension and the choice of PDE modeling our wave phenomenology.

As an example, for multipole source term of the form
\[
	f(\mathbf x,t) = w_1(t) \delta(\mathbf x-\mathbf x^*) + 
			w_2(t) \frac{\partial}{\partial x_1}\delta(\mathbf x-\mathbf x^*) +
			w_3(t) \frac{\partial}{\partial x_2}\delta(\mathbf x-\mathbf x^*).
\]
in the first-order acoustic equations \ref{eq:acous1}, then we take $\gamma=1/2$ and
\[
        Q_1 = \left(\frac{d}{dt}\right)^{1/2}, \quad 
        Q_2 = Q_3 = \frac{1}{c} \left(\frac{d}{d t}\right)^{3/2}.       
\]
The inverse of fractional derivatives, which will be needed for our proposed preconditioner, will coincide with fractional integrals.
Our numerical approach and implementation of fractional derivative/integral operators is discussed in the appendix.

%%%%
\section{Numerical Experiments}

In this section we present some source inversion results illustrating the better-conditioning of normal equations \ref{eq:NEQ} when using our preconditioning scheme.
We solve system \ref{eq:NEQ} iteratively via conjugate gradient for normal equations (CGNE/CGLS), both without and with preconditioner 
\[
	\mathbf M = (\mathbf Q^T\mathbf Q)^{-1}.
\]
See \cite{Paige:82} for a description of the CGLS algorithm and \cite{NocedalWright} for the preconditioned variant which we refer to as PCGLS.
Previous work on moment tensor inversion exclusively use direct solvers given the manageable size of the underlying linear systems.
Direct solvers are indeed the tools to use when the problem is posed in the frequency domain or when simplifying assumptions on the source lead to tractable linear systems.
The size of system \ref{eq:NEQ} we expect to solve for problems related to exploration seismology, as oppose to earthquake seismology, is quite large due to the number of receivers typical in these applications, thus we are motivated in our use of iterative solvers.

The multipole source representation and inversion framework has been implemented as part of the \emph{IWave} and the \emph{Rice Vector Library} (RVL) packages; see \cite{GeoPros:11} and \cite{RVL_TOMS} for more information on the related software.%, and in particular \cite{MPSframework} detailing our implementation.
The underlying PDE solver for acoustic system \ref{eq:acous1} used here, via IWave, is a staggered-grid \emph{finite difference} (FD) method of second-order in time and fourth-order in space, also referred to as the 2-4 scheme, based on \cite{mad:dyn} for elasticity.

The action of linear operator $\mathbf F$, as alluded to before, is implemented as a multichannel convolution of multipole coefficients $\{w_j(t):j=1,2,...\}$ with appropriate Green's functions, thus avoiding further PDE solves during CG iterations.
It is worth mentioning that special care is taken when computing Green's functions since one would have to solve acoustic system \ref{eq:acous1} with impulsive multipole source terms, i.e., 
\[
 	f(\mathbf x,t) = \delta(t) D^{\mathbf s}\delta(\mathbf x-\mathbf x^*).
\]
We take the approach developed by \cite{Walden:1999} for discretizing singular terms $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$, namely replacing such singularities with regular functions of compact support that satisfy vanishing discrete moment conditions.
Such discretizations of singular source terms have been shown to be accurate by preserving convergence rates of FD solvers; for example see \cite{Walden:1999} and \cite{Petersson:2010} for numerical results related to 1D Helmholtz and 2D elastic wave equation respectively.
We reference some of our previous work on singular source approximations \cite{MPStheory} for more details on the methodology applied here.

%%%%%%%%%%
\subsection{Test 1}

The first three tests we present involve estimating some simple multipole sources in a homogeneous medium with ``unbounded'' domain simulated using \emph{perfectly matched layers} (PMLs), as described by \cite{Habashy:07}.
Figures \ref{fig:bmod-hom} and \ref{fig:buoy-hom} plot medium parameters, bulk modulus and buoyancy respectively, as functions of space discretized over a uniform grid with $20m\times 20m$ spacing.
A single multipole source is positioned at $\mathbf x^*=(1.003km,1.003km)$, where components $x_1$ and $x_2$ refer to depth and horizontal axes respectively.
Receivers points $\mathbf x_{i_r}$ are placed at depth $0.9km$, spanning the horizontal axis from $0km$ to $2km$ with a spacing of $20m$.

In test 1 we estimate multipole coefficient $\mathbf w = w_1(t)$ for monopole source
\[
	f(\mathbf x,t) = w_1(t) \delta(\mathbf x-\mathbf x^*).
\]
Observed data $\mathbf d_{\rm obs}^{\epsilon}$ is shown in figure \ref{fig:test1-data-p} with added Gaussian noise $\epsilon$, that is
\[
	\mathbf d_{\rm obs}^{\epsilon} = \mathbf d_{\rm obs} + \epsilon
\]
such that 
\[
	\frac{{\rm RMS}(\epsilon)}{{\rm RMS}(\mathbf d_{\rm obs})} = 20\%,
\]
as is be done in all other test cases.

Figures \ref{fig:test1-CG-rnrm} and \ref{fig:test1-CG-gnrm} plot the log of the $L^2$-norm of the residual $\mathbf r^{(k)} = \mathbf F \mathbf w^{(k)} -\mathbf d_{\rm obs}^{\epsilon}$ and the normal residual $\mathbf g^{(k)} = \mathbf F^T \mathbf r^{(k)}$ versus CG iteration index $k$ respectively.
For PCGLS the normal residual is actually given by $\mathbf g^{(k)} = \mathbf M \mathbf F^T\mathbf r^{(k)}$.
The following CG stoping criterion is applied to all test cases: stop if either of the following three conditions are satisfied,
\begin{itemize} 
	\item reached maximum number of iterations, $k=150$;
	\item sufficient reduction in residual, $\tfrac{\|\mathbf r^{(k)}\|_{\mathfrak D}}{\|\mathbf r^{(0)}\|_{\mathfrak D}} \le 1\times 10^{-4}$;
	\item sufficient reduction in normal residual, $\tfrac{\|\mathbf g^{(k)}\|_{\mathfrak W}}{\|\mathbf g^{(0)}\|_{\mathfrak W}} \le 1\times 10^{-4}$.
\end{itemize}
Note that both CGLS and PCGLS terminate before the 150 iterations for test 1 since they both achieved a sufficient reduction in the normal residual. 
Moreover, applying preconditioning accelerated convergence by almost 100 iterations.

In figure \ref{fig:test1-err} we plot the log of the normalized inversion error, i.e.,
\[
	\log \frac{\|\mathbf w^{(k)}-\mathbf w_{\rm true}\|}{\|\mathbf w_{\rm true}\|}
\]
given that we have access to the true solution $\mathbf w_{\rm true}$.
We report errors using the unweighted norm $\|\cdot\|_{\mathfrak W}$, and for PCGLS estimates we also display the weighted norm $\|\cdot\|_{\mathfrak W'}$ since it is accessible within the PCGLS algorithm.
Inversion error plots demonstrate the semi-convergence behavior of CG when applied to ill-conditioned inverse problems with noisy data, namely that errors begin to increase as one keeps iterating.
For the preconditioned case, we observe that the minimal error and the onslaught of increasing error occurs at earlier iterations.
Stopping criteria that capitalize on the semi-convergence of Krylov space methods are thus crucial in the context of large-scale ill-conditioned inverse problems and are subject of current research.
Given the intended focus of the paper, we postpone dealing with the issue of optimal stopping criteria and simply analyze inversion error plots such as figure \ref{fig:test1-err}.
In the discussion section we will offer some suggestions based on standard techniques in the current literature.

Looking at error plot \ref{fig:test1-err} we see that minimal error estimates are obtained at iterates 28 and 5 for the CGLS and PCGLS respectively.\footnote{PCGLS obtains its minimal error estimate at the same iterate in both the unweighted and weighted norms.}
Figure \ref{fig:test1-MPS} plots the true multipole coefficient $\mathbf w_{\rm true}$ and the minimal error estimates for CGLS and PCGLS, both are qualitatively similar.

%%%%%%%%%
\subsection{Test 2}

In the second test case we invert for a dipole source,
\[
	f(\mathbf x,t) = w_1(t) \frac{\partial}{\partial x_2}\delta(\mathbf x-\mathbf x^*).
\]
Observed data $\mathbf d_{\rm obs}^{\epsilon}$ is shown in figure \ref{fig:test1-data-p} and the log of the $L^2$-norm of the residual and the normal residual plotted in figure \ref{fig:test2-CG-rnrm,test2-CG-gnrm}.
For this case, only PCGLS converges before 150 iterations, again by sufficient normal residual reduction.
Looking at the inversion error plots, figure \ref{fig:test2-err}, we see that PCGLS obtains a larger minimal error than CGLS, though after 5 iterations as oppose to 68 for CGLS.
True multipole coefficient is plotted along with the minimal errors in figure \ref{fig:test2-MPS}.

%%%%%%%%%%
\subsection{Test 3}

In test three we invert for a first-order multipole series of the form
\begin{equation}\label{eq:test3MPS}
	f(\mathbf x,t) = w_1(t) \delta(\mathbf x-\mathbf x^*) 
		+ w_2(t) \frac{\partial}{\partial x_1}\delta(\mathbf x-\mathbf x^*)
		+ w_3(t) \frac{\partial}{\partial x_2}\delta(\mathbf x-\mathbf x^*).
\end{equation}
Though neither CGLS or PCGLS converge in under 150 iteration we do observe that PCGLS achieves a greater reduction in both the residual and the normal residual, see figure \ref{fig:test3-CG-rnrm,test3-CG-gnrm}.
PCGLS also produces marginally more accurate inversion compared to CGLS though under less iterations, i.e., 36 versus 150 iterations in reference to inversion error plots \ref{fig:test3-err}.
Nonetheless, neither CGLS or PCGLS yield satisfactory multipole coefficients as shown by the minimal error plots in figure \ref{fig:test3-MPS-0,test3-MPS-1,test3-MPS-2}.

In hopes of improving our inversion results we apply Tikhonov regularization, or equivalently introduce a damping term in the linear LS problem, that is system \ref{eq:RegNEQ}.
We used a variant of CGLS for solving the regularized normal equations, discussed in algorithm 3 of \cite{Frommer:99}, which we refer to as CGLS+Reg.
Similarly, we modify PCGLS to incorporate regularization yielding PCGLS+Reg.
We avoid the difficulty associated with choosing regularization parameter $\lambda$ by simply choosing it in a trial and error manner as to yield the best of result, specifically smaller solution errors. 
Again, in the discussion section we offer some recommendations and insight as to which current methodologies to apply for choosing $\lambda$.
In this test case, $\lambda= 1e-10$ for both CGLS+Reg and PCGLS+Reg.

Results for both CGLS+Reg and PCGLS+Reg are shown in figures \ref{fig:test3-CG-rnrm-reg,test3-CG-gnrm-reg}, \ref{fig:test3-err-reg}, and \ref{fig:test3-MPS-reg-0,test3-MPS-reg-1,test3-MPS-reg-2}.
Inversion error plots demonstrate a significant improvement for PCGLS+Reg, where it is able to achieve a relative error below $10\%$.
CGLS+Reg however demonstrated little reduction in its solution error.
In fact, it was observed that CGLS+Reg was quite insensitive to regularization, producing similar results for $\lambda$ of varying magnitudes.
This might be perhaps due to the fact that convergence of CGLS with or without regularization is quite sluggish.

%FIX

%%%%%%%%%%
\subsection{Test 4}

A fourth test was conducted for a more complex data set, namely reflectivity data resulting from a dome model; see \ref{fig:bmod-dome} and \ref{fig:buoy-dome} for plots of medium parameters.
The 2-4 staggered grid FD method was used with PML boundaries added at the $x_2=0km$, $x_2=4km$, and $x_1=2km$ edges of the medium while using the free surface condition at depth $x_1=0km$ to simulate marine seismic data.
The source was located at $\mathbf x^*=(0.103km,1.003km)$, and a line of receivers was placed at a depth of $8m$ spanning the horizontal axis from $1.4km$ to $3.4km$ with a $20m$ separation between receivers.
The same multipole source term as in test 3, equation \ref{eq:test3MPS}, is estimated in this final test.

Inversion error plots \ref{fig:test4-err} show a slow but steady decrease in error for CGLS.
Conversely the behavior of PCGLS is rather erratic with inversion error significantly increasing after 20 iterations, despite PCGLS producing smaller residuals (figure \ref{fig:test4-CG-rnrm}).
Interestingly enough, this is the first case where the behavior of the error in the unweighted and weighted norms vary drastically for PCGLS.
Similar to test 3, Tikhonov regularization comes to the rescue demonstrating a significant improvement particularly for PCGLS+Reg.
Optimal regularization parameter was found to be $\lambda=1e-10$ for both CGLS and PCGLS.

%Table \ref{tab:summary} provides a quantitative summary of the estimated solutions, with minimal error, for all test cases.

%%%
\section{Discussions}

Numerical results overall show an acceleration in the semi-convergence of CG iterates when our preconditioning scheme is used.
As the complexity of the source representation increases, such as in the latter test cases, we observe increasingly slow convergence of CGLS and worse yet erratic behavior for PCGLS when analyzing inversion errors.
The goal of preconditioning is to accelerate convergence of CG.
According to standard CG convergence theory, acceleration is achieved if singular values of the preconditioned normal operator $\mathbf M\mathbf F^T\mathbf F$ are more clustered.
An unintended consequence with regards to the semi-convergence properties of CG is that this spectral clustering in the preconditioned system may introduce components related to the ill-conditioned part of the spectrum at earlier iterates.
This is our theory for explaining why PCGLS produces poor results for tests 3 and 4.
Adding, however, Tikhonov regularization seems to alleviated these spectral issues as demonstrated by PCGLS+Reg solutions.
%insensitivity of CGLS to regularization?

Of course, in order to take advantage of CG semi-convergence and Tikhonov regularization requires one to have optimal stopping criteria and a method for choosing the regularization parameter.
Consider first the CGLS algorithm.
It is well known that the residual monotonically decreases with iteration index while the norm of the solution monotonically increases\footnote{If a zero initial guess is used.} under exact arithmetic.
In other words, an early termination of CGLS effectively provides Tikhonov regularization where the terminating iteration count plays the role of regularization parameter.
The L-curve method consists of plotting the residual and solution norm at varying regularization parameter values yielding a characteristic L shaped curve, and selecting the regularization parameter closes to the corner of the L. 
The idea behind the L-curve criterion is that the corner point produces a regularized solution that is a balance between overfitting (smaller residual and larger solution norm) and over-regularizing (smaller solution norm and larger residual).
Given that the residual and solution norm are rather accessible during CGLS solution update makes the L-curve stopping criterion an attractive choice;
we refer to \cite{Hansen:05}, chapter 6 and 7 for a full discussion of the L-curve and its application as a stopping criteria for CGLS.

In the case of PCGLS, with preconditioner of the form $\mathbf M = (\mathbf Q^T\mathbf Q)^{-1}$, it turns out we can relate this algorithm to CGLS via the following standard form transformation, $\tilde{\mathbf w}\leftarrow \mathbf Q \mathbf w$.
Thus we have
\[
	\min_{\mathbf w} \|\mathbf F \mathbf w -\mathbf d \| 
	= \min_{\tilde{\mathbf w}}\| \tilde{\mathbf F} \tilde{\mathbf w} -\mathbf d\|
\]
where $\tilde {\mathbf F} = \mathbf F \mathbf Q^{-1}$.
CGLS iterates of the transformed system can be written as polynomials involving the modified linear operator $\tilde{\mathbf F}$; in particular
\begin{align*}
	\tilde {\mathbf w}^{(k)} 
		&= \mathcal P_k(\tilde{\mathbf F}^T \tilde{\mathbf F})\tilde{\mathbf F}^T \mathbf d \\
		&= \mathcal P_k( (\mathbf Q^{-1})^T \mathbf F^T \mathbf F \mathbf Q^{-1} ) (\mathbf Q^{-1})^T \mathbf F^T \mathbf d
\end{align*}
where $\mathcal P_k(\cdot)$ is a polynomial of degree $k-1$ of its input argument.
Furthermore,
\[
	\mathbf w^{(k)} = \mathcal P_k( (\mathbf Q^T\mathbf Q)^{-1} \mathbf F^T \mathbf F) (\mathbf Q^T\mathbf Q)^{-1} \mathbf F^T \mathbf d
\]
since $\mathbf Q^{-1} \tilde {\mathbf w}^{(k)} = \mathbf w^{(k)}$.
Hence, PCGLS with preconditioner $\mathbf M = (\mathbf Q^T\mathbf Q)^{-1}$ is equivalent to CGLS under the transformation $\tilde{\mathbf w}\leftarrow \mathbf Q\mathbf w$ and $\tilde{\mathbf F}\leftarrow \mathbf F\mathbf Q^{-1}$.
It follows that PCGLS produces iterates with monotonically increasing $\|\mathbf Q\mathbf w^{(k)}\|$ which implies that one should monitor the weighted norm of the solution when applying the L-curve stopping criterion.
Note that this would be a form of generalized Tikhonov regularization with Tikhonov operator $\mathbf Q$.

It is clear from our results that regularization is pivotal, especially for PCGLS in the context of complex sources.
Choosing an optimal regularization parameter $\lambda$ for the regularized normal equations \ref{eq:RegNEQ}, particularly in a manner that is computationally tractable for iterative methods, is also of fundamental importance.% for both the theory of linear ill-posed inverse problems and our particular application. 
A naive implementation of the L-curve criterion for determining an optimal $\lambda$ would entail fully solving system \ref{eq:RegNEQ} for several values of $\lambda$, usually to the order of $O(10)$ times.
Alternatively, there have been variants of the L-curve method for determining optimal $\lambda$ that have been tailored to CG to reduce computational cost.
One such method is the {\em envelope guided} CG algorithm developed by \cite{Kaufman:96} initially for PET image reconstruction.
The method is based on theoretical properties of the L-curve, mainly that its envelope contains all points $(\|\mathbf r\|^2,\|\mathbf w\|^2)$ for all $\mathbf w$.
This is used to guide CG iterates towards the L-curve corner as $\lambda$ is refined according to some updating scheme.
A drawback of this method is that it requires one to store several iterates as potential solutions, similar to the naive implementation of the L-curve.

\cite{Frommer:99} offer two other algorithms for determining regularization parameter $\lambda$ in the context of CG.
Their first method is referred to as a {\em truncated CG} algorithm that, for a current value of $\lambda$, terminates early from CGLS if their theoretical lower bound based on the discrepancy principle is violated.
Regularization parameter is refined in this process until the discrepancy principle is satisfied.
The second method is dubbed the {\em shifted CG} method and utilizes a Lanczos process to compute CG iterates for several $\lambda$ values simultaneously.
This results in less $\mathbf F\mathbf w$ and $\mathbf F^T\mathbf y$ computations overall but requires one to store $O(k_{\rm max})$ more vectors, where $k_{\rm max}$ is the number of different $\lambda$ values being tested.
It should be mentioned that the shifted CG method also relies on the discrepancy principle, like the truncated CG algorithm, and thus require knowledge of the noise in the data.

%using L-curve criterion for stopping criteria of CGLS, 
%monitor ||x|| and ||r|| when there is no preconditioning
%monitor ||Qx|| and ||r|| when there is preconditioning
%Properties of CG, monotonically increasing ||x|| and ||Qx|| resp
%Equivalency of preconditioning and Tikhonov reg standard form transformation
%Preconditioning CGLS is more sensitive to null space issues, early CG termination is not sufficient to reg problem

%choosing regularization parameter  for damped LS:

%extension of preconditioner to elasticity

%We have demonstrated better-conditioning of multipole inversions when preconditioning, in the context of CGLS as our iterative solver for the source estimation problem.
%Our preconditioning approach is based on fundamentally improving condition number of normal equations by redefining the domain inner-product of linear operator $\mathbf F$, thus resulting in a preconditioned normal equations; see equation \ref{eq:PNEQ}.
%We thus argue that our methodology for better-conditioning the multipole estimation, or equivalently the moment tensor, problem via FWI can be applied in general regardless of the choice of solver for the normal equation.
%Moreover, regularization techniques such as optimized data coverage and dampening of the normal equations as suggested by \cite{Koch:91} can and should be applied conjunctly to further improve condition number of system \ref{eq:PNEQ}.
%

%%%
\section{Conclusion}
In this paper we have presented a framework for inverting seismic sources as series of multipoles.
The source estimation problem reduces to solving normal equations for multipole coefficients, reminiscent of moment tensor components in earthquake seismology.
Our main contribution has been to introduce a preconditioner when solving the aforementioned normal equations via conjugate gradient.
This preconditioner is based on fractional derivative/integral operators whose order is based semi-heuristically on the analytical solutions of the acoustic wave equation in unbounded media with multipole source terms.
Numerical experiments conducted show the acceleration of CG iterates and accuracy of estimated sources when using preconditioning and in conjunction with Tikhonov regularization for more complicated inversions.
%Our approach for better-conditioning the source inversion problem for seismic sources represented as multipoles is fundamental  and necessary, especially for complex sources, and should be applied along with other regularization techniques.

\newpage

%%%%%%%%%%%%%%%%%%%
\section{Appendix}
%%%%%%%%%%%%%%%%%%%

%%%%%%
\subsection{Analytical Solutions to the Wave Equation with Multipole Sources}

We first focus our attention to the second order wave equation, in particular
\begin{equation}\label{eq:2DWE}
        \left( \frac{\partial}{\partial t^2} - c^2 \nabla^2 \right) p(\mathbf x,t) = \tilde f(\mathbf x,t),
\end{equation}
with multipole source spatially centered at the origin,
\[
	\tilde f(\mathbf x,t) = \sum_{i} \tilde w_i D^{\mathbf s_i} \delta(\mathbf x).
\]

we relate equations \ref{eq:2Dansol0} and \ref{eq:2Dansol1} to analytical solutions for the acoustic equations in first order form by noting that $\tilde f = \tfrac{\partial}{\partial t}f$.

\subsection{Deriving Analytical Solutions for 2D}

We focus our attention to the 2D case, since the 1D and 3D cases lead to simpler derivations.
Consider the wave equation in an unbounded homogenous medium for pressure field $p$, with homogenous initial conditions and causal source term $\tilde f$ spatially centered at the origin,
\begin{equation}\label{eq:2DWE}
        \left( \frac{\partial}{\partial t^2} - c^2 \nabla^2 \right) p(\mathbf x,t) = \tilde f(\mathbf x,t),
\end{equation}
where $c$ is the speed of sound in the medium.
An analytical solution to equation \ref{eq:2DWE} is given by convolving source term $\tilde f$ in time and space with the 2D Green's function $G$,
\[
        G(\mathbf x,t) = \frac{1}{2\pi c} \frac{H(t-\tfrac{|\mathbf x|}{c})}{\sqrt{c^2t^2 - |\mathbf x|^2}},
\] 
where $H$ denoting the Heaviside function.
In the special case where $\tilde f$ is a point source, e.g., $\tilde f(\mathbf x,t) = \tilde w(t) \delta(\mathbf x)$, we obtain an expression for the pressure field in terms of a temporal deconvolution,
\[
        p(\mathbf x,t) = \int_{\mathbb R} g(\mathbf x,s) \tilde w(t-s) \; ds
\]
with Green's function
\[
        g(\mathbf x,t) = \frac{1}{2\pi c} \frac{H(t-\tfrac{|\mathbf x|}{c})}{\sqrt{c^2t^2-|\mathbf x|^2}} 
\]
Applying a change of variables simplifies the expression above and eliminates the singularity in the resulting integral for the pressure field;
take $\sigma = \sqrt{s-\tfrac{|\mathbf x|}{c}}$, then the resulting integral takes the form
\begin{equation}\label{eq:2Dansol0}
        p(\mathbf x,t) = \frac{1}{\pi c^2} \int_{0}^{\sqrt{\tau}} \tilde w(\tau-\sigma^2) \Omega(\mathbf x,\sigma) \; d\sigma
\end{equation}
where we have introduced the variable $\tau$ and function $\Omega(\mathbf x,\sigma)$ in order to clean up our expressions;
\[
        \tau = t - \frac{|\mathbf x|}{c}, \quad \text{and} \quad \Omega(\mathbf x,\sigma) = \frac{1}{\sqrt{\sigma^2 + 2 \frac{|\mathbf x|}{c}}}.
\]

Analytical solutions to the wave equation with multipole sources can be derived by appropriately differentiating the pressure field from equation \ref{eq:2Dansol0}.
For example, if $p_{\mathbf s}$ denotes the solution to the wave equation with $\tilde f(\mathbf x,t) = \tilde w(t) D^{\mathbf s} \delta(\mathbf x)$ for some multi-index $\mathbf s$, then it can be shown that $D^{\mathbf s}_{\mathbf x} p(\mathbf x,t) = p_{\mathbf s}(\mathbf x,t)$, where $D^{\mathbf s}_{\mathbf x}$ is the $\mathbf s$-mixed partial derivative with respect to $\mathbf x$.
We give expressions for the case $|\mathbf s|=1$ here, i.e., $D^{\mathbf s} = \tfrac{\partial}{\partial x_i}$.

In order to differentiate equation \ref{eq:2Dansol0} with respect to $x_i$ we must invoke Leibniz rule: Let 
\[ 
   \phi(\mathbf x,t,\sigma) = \tilde w(\tau -\sigma^2)\Omega(\mathbf x,\sigma).
\]
and note that $\phi$ has continuous partial derivatives in $\sigma$ for $|\mathbf x|\neq 0$.
Then
\[
        \frac{\partial}{\partial x_i} \left( \int_{0}^{b(\mathbf x,t)} \phi(\mathbf x,t,\sigma) \; d\sigma \right) = 
        \phi(\mathbf x,t,b(\mathbf x,t)) \cdot \frac{\partial b}{\partial x_i}(\mathbf x,t) +
        \int_{0}^{b(\mathbf x,t)} \frac{\partial}{\partial x_i} \phi(\mathbf x,t,\sigma) \; d\sigma,
\]
where we have taken $b(\mathbf x,t) = \sqrt{\tau}$.
Computing partial derivatives,
\begin{align*}
        \frac{\partial}{\partial x_i} b(\mathbf x,t) = -\frac{1}{2c}\frac{\gamma_i}{\sqrt{\tau}} 
        \quad \text{with} \quad \gamma_i=\frac{\partial}{\partial x_i}|\mathbf x| = \tfrac{x_i}{|\mathbf x|}. 
\end{align*}
Thus,
\begin{align*}
        \phi(\mathbf x,t,b(\mathbf x,t)) \cdot \frac{\partial b}{\partial x_i}(\mathbf x,t) 
                     &= \frac{\tilde w(\tau-\tau)}{ \sqrt{\tau + 2\tfrac{|\mathbf x|}{c}} } \cdot  
                        \left( -\frac{1}{2c} \frac{\gamma_i}{\sqrt{\tau}} \right) \\
                               &= -\frac{\gamma_i}{2c} \frac{\tilde w(0)}{\sqrt{ t^2 - \tfrac{|\mathbf x|^2}{c^2} }}\\
                                  &= 0,
\end{align*}
assuming $\tilde w$ is causal, that is $\tilde w(t)=0$ for $t\le 0$.
Furthermore, 
\begin{align*}
        \frac{\partial}{\partial x_i} \phi(\mathbf x,t,\sigma) &= 
                                 \tilde w'(\tau-\sigma^2) \left( -\frac{\gamma_i}{c}\right) \Omega(\mathbf x,\sigma)
                                                   + \tilde w(\tau-\sigma^2) \left(-\frac{1}{2}\right)  \Omega^3(\mathbf x,\sigma) \left( \frac{2\gamma_i}{c} \right) \\
                                                     &= -\frac{\gamma_i}{c}  
                                                        \Big\{ \tilde w'(\tau-\sigma^2) \Omega(\mathbf x,\sigma) + \tilde w(\tau-\sigma^2)\Omega^3(\mathbf x,\sigma) \Big\}.
\end{align*}
Putting it all together we have the pressure field produced by a point dipole pressure source, for $|\mathbf s| = 1$,
\begin{align}\label{eq:2Dansol1}
        \frac{\partial }{\partial x_i} p(\mathbf x,t) = p_{\mathbf s}(\mathbf x,t) = -\frac{\gamma_i}{\pi c^3}  \int_{0}^{\sqrt{\tau}} \; 
                       \Big\{ \tilde w'(\tau-\sigma^2) \Omega(\mathbf x,\sigma)
                                                + \tilde w(\tau-\sigma^2)\Omega^3(\mathbf x,\sigma) \Big\} \; d\sigma.
\end{align}

Lastly, we relate equations \ref{eq:2Dansol0} and \ref{eq:2Dansol1} to analytical solutions for the acoustic equations in first order form by noting that $\tilde f = \tfrac{\partial}{\partial t}f$.
The table below summarizes these analytical solutions, including the cases for 1D and 3D. 
The sign function is denoted by ${\rm sgn}(x)$.

%%%%%
\subsection{Fractional Derivative/Integral Operators}

We will follow the Gr\"unwald-Letnikov definition of the left fractional derivative as a basis for a numerical implementation; see \cite{Li:15} for a detailed account of numerical methods for fractional calculus.
The left fractional derivative of order $\alpha>0$, of a given function $w(t)$, $t\in(0,T)$, is defined as
\begin{equation}\label{eq:DGL}
	\mathcal{D}^{\alpha}_{GL} w(t) := \lim_{\underset{N\Delta t=t}{\Delta t\to0}} \Delta t^{-\alpha} \sum_{j=0}^{N} (-1)^{j} {\alpha \choose j} w(t-j\Delta t),
\end{equation}
where ${\alpha \choose j}$ is interpreted to be the generalized binomial coefficient for $\alpha\in\mathbb R$, i.e.,
\[
	{\alpha \choose j} = \frac{\Gamma(\alpha+1)}{\Gamma(\alpha-j+1) j!}.
\]
The discrete operator, which we will simply denote by $\mathcal D^\alpha$, is given by not taking the limit in equation \ref{eq:DGL}, thus yielding a finite-difference-like approach to computing fractional derivatives. 
Suppose we discretize the time interval $(0,T)$ uniformly into $N_t$ segments, where $\Delta t = \tfrac{T}{N_t}$.
The discrete fractional derivative of order $\alpha>0$, for $w$ at time $t=n\Delta t$, $\forall n=0,1,...,N_t$, is defined as
\begin{equation}\label{eq:Derv}
	\mathcal D^{\alpha} w(t) :=  \Delta t^{-\alpha} \sum_{j=0}^{N^*} (-1)^{j}  \frac{\Gamma(\alpha+1)}{\Gamma(\alpha-j+1) j!} w((n-j)\Delta t).
\end{equation}
where
\[
	N^* = \left\{\begin{array}{cl}
		\text{min}(n,\alpha), &\alpha\in\mathbb N, \\
		n, & \text{otherwise}.
	\end{array}\right.
\]
%For $\alpha\in\mathbb N$, it follows that $D^{\alpha}$ is equivalent the discrete matrix $\mathbf D$ to the $\alpha$ power;
%\[
%	\mathbf D = \frac{1}{\Delta t} \mat{1 \\
%				 -1 & 1 \\
%				     & -1 & 1\\
%				     &     & \ddots & \ddots \\
%				     &     &      &      & -1 & 1}.
%\]

The discrete fractional integral of order $\alpha>0$ is derived by taking $\alpha \to -\alpha$ in equation \ref{eq:DGL}, and making sense out of the binomial coefficients.
First lets expand the binomial coefficient for $\alpha\in\mathbb N$,
\[
	{\alpha \choose j} = \frac{\alpha(\alpha-1)\cdots(\alpha-j+1)}{j!}.
\]
Replacing $\alpha$ with $-\alpha$ gives
\begin{subequations}
\begin{align*}
	{-\alpha \choose j} 
		&= \frac{-\alpha(-\alpha-1)\cdots(-\alpha-j+1)}{j!}, \\
		&= (-1)^{j} \frac{\alpha(\alpha+1)\cdots(\alpha+j-1)}{j!}, \\
		&= (-1)^{j} \frac{(\alpha+j-1)!}{(\alpha-1)! j!}.
\end{align*}
\end{subequations}
The discrete fractional inetgral of order $\alpha>0$, for $f$ at time $t=n\Delta t$, $\forall n=0,1,...,N_t$, is defined as
\begin{equation}\label{eq:Inte} 
	\mathcal D^{-\alpha} w(t) :=  \Delta t^{\alpha} \sum_{j=0}^{n}   \frac{\Gamma(\alpha+j)}{\Gamma(\alpha) j!} w((n-j)\Delta t).
\end{equation}

%PLOTTING FIGS

\inputdir{project}

\multiplot{2}{test0-MPS,test0-MPS-spec}{width=0.45\textwidth}{10Hz Ricker wavelet time series (a) and its frequency spectrum (b).}

\multiplot{6}{test0-data-0,test0-D-0,test0-data-1,test0-D-1,test0-data-2,test0-D-2}{width=0.45\textwidth}{Output pressure field (left column) and fractional time derivatives of input waveform (right column). First, second, and third rows of plots correspond to $\mathbf s=(0,0)$, $\mathbf s=(0,1)$, and $\mathbf s=(0,2)$ cases respectively. Plots (b), (d), and (f) correspond to fractional derivatives of order 1/2, 3/2, and 5/2 respectively.}

\multiplot{4}{bmod-hom,buoy-hom,bmod-dome,buoy-dome}{width=0.45\textwidth}{Medium parameters: constant bulk modulus (a) and buoyancy (b) for tests 1-3; dome bulk modulus (c) and buoyancy (d) for test 4.}

\multiplot{4}{test1-data-p,test2-data-p,test3-data-p,test4-data-p}{width=0.45\textwidth}{Time traces of pressure field at receiver locations: (a) test 1, (b) test 2, (c) test 3, (d) test 4.}

%TEST 1

\multiplot{2}{test1-CG-rnrm,test1-CG-gnrm}{width=0.6\textwidth}{Convergence plots of residual {\color{red}(a)} and normal residual {\color{blue}(b)} for test 1, CGLS (dashed) and PCGLS (solid).}

\plot{test1-err}{width=1\textwidth}{Relative error of estimated multipole coefficient for test 1, CGLS (dashed) and PCGLS (solid). Error measured in the $L^2$-norm $\|\cdot\|_{\mathfrak W}$ {\color{red}(red)} and the weighted norm $\|\cdot\|_{\mathfrak W'}$ {\color{blue}(blue)}.}

\plot{test1-MPS}{width=1\textwidth}{True (solid) and minimal errors of multipole coefficient for test 1; CGLS {\color{red}(dashed)} after 28 and PCGLS {\color{blue}(dotted)} after 5 iterations.}

%\multiplot{2}{test1-Lcurve,test1-Lcurve-PC}{width=0.8\textwidth}{L-curves}

%TEST 2

\multiplot{2}{test2-CG-rnrm,test2-CG-gnrm}{width=0.6\textwidth}{Convergence plots of residual {\color{red}(a)} and normal residual {\color{blue}(b)} for test 2, CGLS (dashed) and PCGLS (solid).}

\plot{test2-err}{width=1\textwidth}{Relative error of estimated multipole coefficient for test 2, CGLS (dashed) and PCGLS (solid). Error measured in the $L^2$-norm $\|\cdot\|_{\mathfrak W}$ {\color{red}(red)} and the weighted norm $\|\cdot\|_{\mathfrak W'}$ {\color{blue}(blue)}.}

\plot{test2-MPS}{width=1\textwidth}{True (solid) and minimal errors of multipole coefficient for test 2; CGLS {\color{red}(dashed)} after 68 and PCGLS {\color{blue}(dotted)} after 5 iterations.}

%TEST 3

\multiplot{2}{test3-CG-rnrm,test3-CG-gnrm}{width=0.6\textwidth}{Convergence plots of residual {\color{red}(a)} and normal residual {\color{blue}(b)} for test 3, CGLS (dashed) and PCGLS (solid).}

\plot{test3-err}{width=1\textwidth}{Relative error of estimated multipole coefficient for test 3, CGLS (dashed) and PCGLS (solid). Error measured in the $L^2$-norm $\|\cdot\|_{\mathfrak W}$ {\color{red}(red)} and the weighted norm $\|\cdot\|_{\mathfrak W'}$ {\color{blue}(blue)}.}

\multiplot{3}{test3-MPS-0,test3-MPS-1,test3-MPS-2}{width=0.45\textwidth}{True (solid) and minimal errors of multipole coefficients for test 3; CGLS {\color{red}(dashed)} after 150 and PCGLS {\color{blue}(dotted)} after 36 iterations. Multipole component 1 (a), component 2 (b), and component (3).}

\multiplot{2}{test3-CG-rnrm-reg,test3-CG-gnrm-reg}{width=0.6\textwidth}{Convergence plots of residual {\color{red}(a)} and normal residual {\color{blue}(b)} for test 3 with regularization, CGLS (dashed) and PCGLS (solid).}

\plot{test3-err-reg}{width=1\textwidth}{Relative error of estimated multipole coefficient for test 3 with regularization, CGLS (dashed) and PCGLS (solid). Error measured in the $L^2$-norm $\|\cdot\|_{\mathfrak W}$ {\color{red}(red)} and the weighted norm $\|\cdot\|_{\mathfrak W'}$ {\color{blue}(blue)}.}

\multiplot{3}{test3-MPS-reg-0,test3-MPS-reg-1,test3-MPS-reg-2}{width=0.45\textwidth}{True (solid) and minimal errors of multipole coefficients for test 3 with regularization; CGLS {\color{red}(dashed)} after 150 and PCGLS {\color{blue}(dotted)} after 79 iterations. Multipole component 1 (a), component 2 (b), and component (3).}

%No PC
%mu=0       , min_err = 0.37466
%mu=1e-15, min_err = 0.37543
%mu=1e-14, min_err = 0.37436
%mu=1e-13, min_err = 0.37528
%mu=1e-12, min_err = 0.37446
%mu=1e-11, min_err = 0.37452
%mu=1e-10, min_err = 0.37581 ***
%mu=1e-09, min_err = 0.38843
%mu=1e-08, min_err = 0.51442

%with PC
%mu=0       , min_err = 0.32676  k=36
%mu=1e-13, min_err = 0.30634  k=41
%mu=1e-12, min_err = 0.20584  k=39
%mu=1e-11, min_err = 0.10887  k=78
%mu=1e-10, min_err = 0.07561  k=79 ***
%mu=1e-09, min_err = 0.29535  k=125

%TEST 4

\multiplot{2}{test4-CG-rnrm,test4-CG-gnrm}{width=0.6\textwidth}{Convergence plots of residual {\color{red}(a)} and normal residual {\color{blue}(b)} for test 4, CGLS (dashed) and PCGLS (solid).}

\plot{test4-err}{width=1\textwidth}{Relative error of estimated multipole coefficient for test 4, CGLS (dashed) and PCGLS (solid). Error measured in the $L^2$-norm $\|\cdot\|_{\mathfrak W}$ {\color{red}(red)} and the weighted norm $\|\cdot\|_{\mathfrak W'}$ {\color{blue}(blue)}.}

\multiplot{3}{test4-MPS-0,test4-MPS-1,test4-MPS-2}{width=0.45\textwidth}{True (black, solid) and errors of ``best'' estimated multipole coefficient for test 4; without {\color{red}(dashed)} and with {\color{blue}(dotted)} preconditioning after 150 and 150 CGLS iterations respectively.}

\multiplot{2}{test4-CG-rnrm-reg,test4-CG-gnrm-reg}{width=0.6\textwidth}{Convergence plots of residual {\color{red}(a)} and normal residual {\color{blue}(b)} for test 4 with regularization, CGLS (dashed) and PCGLS (solid).}

\plot{test4-err-reg}{width=1\textwidth}{Relative error of estimated multipole coefficient for test 4 with regularization, CGLS (dashed) and PCGLS (solid). Error measured in the $L^2$-norm $\|\cdot\|_{\mathfrak W}$ {\color{red}(red)} and the weighted norm $\|\cdot\|_{\mathfrak W'}$ {\color{blue}(blue)}.}

\multiplot{3}{test4-MPS-reg-0,test4-MPS-reg-1,test4-MPS-reg-2}{width=0.45\textwidth}{True (solid) and minimal errors of multipole coefficients for test 4 with regularization; CGLS {\color{red}(dashed)} after 150 and PCGLS {\color{blue}(dotted)} after 150 iterations. Multipole component 1 (a), component 2 (b), and component (3).}

%No PC
%mu=0       , min_err = 0.47596
%mu=1e-14, min_err = 0.47581
%mu=1e-13, min_err = 0.47610
%mu=1e-12, min_err = 0.47618
%mu=1e-11, min_err = 0.47626 
%mu=1e-10, min_err = 0.47749 ***
%mu=1e-09, min_err = 0.48270
%mu=1e-08, min_err = 0.53256

%with PC
%mu=0       , min_err = 0.73065  k=5
%mu=1e-12, min_err = 0.50234  k=150
%mu=1e-11, min_err = 0.30488  k=150
%mu=5e-11, min_err = 0.19310  k=150
%mu=8e-11, min_err = 0.15954  k=150
%mu=1e-10, min_err = 0.14998  k=150 ***
%mu=5e-10, min_err = 0.15316  k=149
%mu=1e-09, min_err = 0.19811  k=150
%mu=1e-08, min_err = 0.47450  k=150

%TABLE
%\begin{table}
%\renewcommand{\arraystretch}{2}
%	\footnotesize
%	\centering
%	\begin{tabular}{ c |c | c | c | c | c | c }
%		test \# & 
%		CG alg. &
%		$\frac{\|\mathbf r^{(i)}\|}{\|\mathbf r^{(0)}\|}$ & 
%		$\frac{\|\mathbf g^{(i)}\|}{\|\mathbf g^{(0)}\|}$ & 
%		$\frac{\|\mathbf w^{(i)}-\mathbf w_{\rm true}\|_{\mathfrak W}}{\|\mathbf w_{\rm true}\|_{\mathfrak W}}$ & 
%		$\frac{\|\mathbf w^{(i)}-\mathbf w_{\rm true}\|_{\mathfrak W'}}{\|\mathbf w_{\rm true}\|_{\mathfrak W'}}$ & 
%		\# iter.\\ 
%		\hline
%		\multirow{2}{*}{1} &
%			CGLS &	  
%		  	$1.84\times 10^{-4}$ &
%		  	$9.63\times 10^{-5}$ & 
%			$9.77\times 10^{-4}$ &
%			- &
%			$33$ 
%		\\ \cline{2-7}&
%			\cellcolor{Gray} PCGLS &
%		  	\cellcolor{Gray} $8.53\times 10^{-5}$ &
%		  	\cellcolor{Gray} $1.03\times 10^{-4}$ &
%			\cellcolor{Gray} $1.73\times 10^{-4}$ &
%			\cellcolor{Gray} $8.22\times 10^{-5}$ &
%			\cellcolor{Gray} $4$
%		\\ \hline\hline
%		\multirow{2}{*}{2} &
%			CGLS &	  
%		  	$1.84\times 10^{-4}$ &
%		  	$9.63\times 10^{-5}$ & 
%			$9.77\times 10^{-4}$ &
%			- &
%			$33$ 
%		\\ \cline{2-7}&
%			\cellcolor{Gray} PCGLS &
%		  	\cellcolor{Gray} $8.53\times 10^{-5}$ &
%		  	\cellcolor{Gray} $1.03\times 10^{-4}$ &
%			\cellcolor{Gray} $1.73\times 10^{-4}$ &
%			\cellcolor{Gray} $8.22\times 10^{-5}$ &
%			\cellcolor{Gray} $4$
%		\\ \hline\hline
%		\multirow{4}{*}{3} &
%			CGLS &	  
%		  	$2.20e-01$ &
%		  	$1.25e-02$ & 
%			$9.77\times 10^{-4}$ &
%			- &
%			$33$ 
%		\\ \cline{2-7}&
%			\cellcolor{Gray} PCGLS &
%		  	\cellcolor{Gray} $8.53\times 10^{-5}$ &
%		  	\cellcolor{Gray} $1.03\times 10^{-4}$ &
%			\cellcolor{Gray} $1.73\times 10^{-4}$ &
%			\cellcolor{Gray} $8.22\times 10^{-5}$ &
%			\cellcolor{Gray} $4$
%		\\ \cline{2-7}&
%			CGLS+Reg &
%		  	$8.53\times 10^{-5}$ &
%		  	$1.03\times 10^{-4}$ &
%			$1.73\times 10^{-4}$ &
%			- &
%			$4$
%		\\ \cline{2-7}&
%			\cellcolor{Gray} PCGLS+Reg &
%		  	\cellcolor{Gray} $8.53\times 10^{-5}$ &
%		  	\cellcolor{Gray} $1.03\times 10^{-4}$ &
%			\cellcolor{Gray} $1.73\times 10^{-4}$ &
%			\cellcolor{Gray} $8.22\times 10^{-5}$ &
%			\cellcolor{Gray} $150$
%		\\ \hline\hline
%		\multirow{4}{*}{4} &
%			CGLS &	  
%		  	$2.20e-01$ &
%		  	$1.25e-02$ & 
%			$9.77\times 10^{-4}$ &
%			- &
%			$150$ 
%		\\ \cline{2-7}&
%			\cellcolor{Gray} PCGLS &
%		  	\cellcolor{Gray} $8.53\times 10^{-5}$ &
%		  	\cellcolor{Gray} $1.03\times 10^{-4}$ &
%			\cellcolor{Gray} $1.73\times 10^{-4}$ &
%			\cellcolor{Gray} $8.22\times 10^{-5}$ &
%			\cellcolor{Gray} $150$
%		\\ \cline{2-7}&
%			CGLS+Reg &
%		  	$8.53\times 10^{-5}$ &
%		  	$1.03\times 10^{-4}$ &
%			$1.73\times 10^{-4}$ &
%			- &
%			$150$
%		\\ \cline{2-7}&
%			\cellcolor{Gray} PCGLS+Reg &
%		  	\cellcolor{Gray} $8.53\times 10^{-5}$ &
%		  	\cellcolor{Gray} $1.03\times 10^{-4}$ &
%			\cellcolor{Gray} $1.73\times 10^{-4}$ &
%			\cellcolor{Gray} $8.22\times 10^{-5}$ &
%			\cellcolor{Gray} $150$
%		\\ \hline
%	\end{tabular}
%	\caption{Summary of results for single scalar MPS inversions with and without preconditioning, containing data residual reduction, gradient reduction, and errors in MPS coefficients with respect to $\|\cdot\|_{\mathfrak W}$ and $\|\cdot\|_{\mathfrak W'}$ norms.}
%	\label{tab:summary}
%\end{table}

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/mps_theory/paper.tex
repro docs dir added

%\title{Discretization of Multipole Sources in a Finite Difference Setting for Wave Propagation Problems}
%\date{}
%\address{
%        \footnotemark[1]Institute for Computational and Experimental Research in Mathematics,\\ Brown University,\\ Providence, RI 02912 USA\\
%        \footnotemark[2]The Rice Inversion Project,\\ Rice University,\\ Houston, TX
%        77005-1892 USA
%}
%\author{Mario J. Bencomo\footnotemark[1] and William Symes\footnotemark[2]}

%\righthead{Multipole Source Discretizations}

%\maketitle
%\parskip 12pt

%%%%%%%%%
\begin{abstract}
%%%%%%%%%

Seismic sources are commonly idealized as point-sources due
to their small spatial extent relative to seismic wavelengths.
The acoustic isotropic point-radiator is inadequate as a model 
of seismic wave generation for seismic sources that are known to exhibit directivity. 
Therefore, accurate modeling of seismic wavefields must include source 
representations generating anisotropic radiation patterns. 
Such seismic sources can be modeled as 
{\em multipoles}, that is, time-dependent linear combination of
spatial derivatives of the spatial delta function. Since the solutions of linear hyperbolic
systems with point-source right hand sides are necessarily singular, standard results on
convergence of grid-based numerical methods (finite difference or
finite element) do not imply convergence of numerical solutions.
We present a method for discretizing multipole sources 
in a finite difference setting, an extension of the moment matching 
conditions developed for the Dirac delta function in other applications.
along with numerical evidence demonstrating the accuracy of these approximations.
Using this analysis, we develop a weak convergence theory for the discretization of 
a family of symmetric hyperbolic systems of first-order partial
differential equations, with singular source terms, solved via
staggered-grid finite difference methods: we show that
grid-independent space-time averages of the numerical solutions
converge to the same averages of the continuum solution, and provide
an estimate for the error in terms of moment matching and truncation
error conditions.
Numerical experiments confirm this result, but also suggest a stronger
one: optimal convergence rates appear to be achieved
achieved point-wise in space away from the source.

\end{abstract}

%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%

%Seismic sources are commonly idealized as concentrated at a source point due
%to their small spatial extent relative to seismic wavelengths.
Linear wave propagation problems modeling physical systems with
concentrated (in space) energy sources arise in a number of
applications, such as seismology, nondestructive material evaluation,
and radar. A familiar example of a wave propagation model with
spatially concentrated 
%energy
energy
source is the isotropic point-radiator problem
for the acoustic wave equation in Euclidean 3-space \cite[]{CourHil:62}:
\begin{equation}\label{eq:isorad}
\begin{split}
       \frac{1}{c^2} \frac{\partial^2}{\partial t^2} p(\mathbf x,t) - \nabla^2
  p(\mathbf x,t) & = f(\mathbf x,t) \equiv w(t)\delta(\mathbf x),\\ 
p(\mathbf x,t) & =  0,  \quad t<<0,
\end{split}
\end{equation}
$\delta(\mathbf x)$ being the Dirac delta function.
The solution $p$ is spherically symmetric:
\begin{equation}
\label{eq:green3d}
	p(\mathbf x,t) = \frac{w\left(t-\frac{r}{c}\right)}{4\pi r}, \quad 
	r= \sqrt{\mathbf x^T \mathbf x}.
\end{equation}

% here is a heaven-sent opportunity to tell the reader what the paper
% is about

The form \ref{eq:green3d} of the isotropic point radiator solution
suggests an obstacle to successful  approximation by gridded (finite
difference, finite element,...) numerical methods: the fundamental
rationale for the accuracy of such methods, namely truncation error
analysis based on Taylor expansion, is invalid in any neighborhood of
the source point $\mathbf x = \mathbf 0$, as the solution is singular
there. Moreover, due to the wave nature of both continuum and gridded
approximate solutions for problems like \ref{eq:isorad}, errors
arising near the source point propagate throughout the region in which
the solution is computed. Examples presented later in this paper
illustrate this generation and propagation of error due to source singularity.

The main result of this paper is that certain special approximations
to wave dynamics problems such as \ref{eq:isorad} with (singular)
right-hand sides of point support lead to finite difference solutions
that converge globally in a weak sense - that is, averages over fixed
regions converge as the grid cell size is reduced.

In the remainder of this introduction, we describe the class of wave
dynamics problems to which our arguments apply, the representation and approximation
of right-hand sides,  the type of numerical
approximation used in our work, and the weak convergence of approximate solutions.

We focus our attention on a symmetric hyperbolic systems of the form
\begin{equation}\label{eq:hyper_fam}
\begin{split}
	A \frac{\partial}{\partial t}u + P^T v &= f \\
	B \frac{\partial}{\partial t}v  - P u & = g
\end{split}
\end{equation}
connecting dynamical fields $u,v$, energy source fields $f,g$, and symmetric
positive definite matrix fields $A,B$. The operator $P$ is a
constant-coefficient differential operator of order 1; $P^T$ denotes its
formal adjoint or transpose.
Many important physical systems supporting wave propagation can be
cast in the form \ref{eq:hyper_fam}. These include linear
elastodynamics and Maxwell's equations. In all such examples, the
matrices $A$ and $B$ encode time-independent physical parameters characteristic of the
continuum mechanical model. For example, elastodynamics may be cast in
the form \ref{eq:hyper_fam}, with $A$ being the compliance tensor,
$B$ being the material density multiplied by the $3 \times 3$
identity, and $P$ is the strain operator. As is well-known, the acoustic
wave equation \ref{eq:isorad} is also equivalent to such a system, in the
sense that its solution $p$ can be identified with $u$ above, with $v$
chosen as the indefinite time integral of $-\nabla p$. In this case,
$A$ is the $1 \times 1$ matrix with entry $c^2$, $B$ is the $3
\times 3$ identity matrix, and $P$ is the gradient operator. The resulting system expresses
acoustodynamics with material density normalized to 1, and $c^2$
identified with bulk modulus of the fluid material. The dynamic fields
$p$ and $v$ represent pressure and particle velocity respectively.

Sources of small support relative to typical wavelengths are reasonable models
for some seismic energy generation mechanisms. Such sources may be
approximated by point support right hand sides such as occur in the
isotropic radiator problem \ref{eq:isorad}. Of course, ``functions''
of space and time having point support in space are not really
functions at all, but (Schwarz) distributions. 
Peetre's Theorem \cite[]{Horm:69} implies that any distribution in
${\bf R}^{d+1}$ 
supported on $\{\mathbf x^* \} \times {\bf R}$, ${\bf x^*} \in {\bf R}^d$, takes the form of a finite series
\begin{equation}\label{eq:MPSappx}
        f(\mathbf x,t) = \sum_{|\mathbf s|=0}^N w_{\mathbf s}(t) \; D^{\mathbf s}\delta(\mathbf x-\mathbf x^*),
\end{equation}
in which we have introduced {\em multi-index notation}: for spatial
dimension $d$ and multi-index (integer d-tuple) ${\bf s} =
(s_1,...,s_d)$, the $\mathbf s$-mixed partial derivative operator, denoted $D^{\bf s}$,
and its (total) order $|{\bf s}|$, are defined as
\begin{equation}\label{eq:PDO}
        D^{\mathbf s} = \prod_{i=1}^d \left( \frac{\partial}{\partial x_i}\right)^{s_i}, \quad |\mathbf s| = \sum_{i=1}^{d} s_i.
\end{equation}
In the context of energy source representation, we shall refer to
distributions of point support, necessarily taking the form
\ref{eq:MPSappx}, as {\em multipole sources}, or simply {\em multipoles}.
Such sources combine
localization of energy and (possibly) anisotropic radiation pattern.
%distribution in space and time. ?? MJB
The coefficient distributions $w_{\mathbf s}(t)$ may be scalar-, vector-, or
tensor-valued, according to the nature of the quantity (pressure, velocity, or stress) 
in the equation in which
$f(\mathbf x,t)$ appears as right-hand side.  Multipoles may
approximate arbitrary sources highly localized on the wavelength scale, in
the sense of generating approximately the same field away from the
source location, and for this reason have enjoyed widespread use in
the representation of seismic sources \cite[]{Shearer:2009}.

The isotropic acoustic point radiator problem \ref{eq:isorad} is
(equivalent to) a
simple special case of the symmetric hyperbolic system
\ref{eq:hyper_fam} with multipole
right hand sides $f,g$. Therefore gridded approximation of the more general
dynamics modeled by \ref{eq:hyper_fam} suffers from the difficulty
pointed out earlier for \ref{eq:isorad}.

Our results pertain to a specific class of finite difference
approximations for systems of the form \ref{eq:hyper_fam},
the so called {\em staggered grid schemes}, defined in detail in below
(equation \ref{eq:sg} and surrounding discussion). These schemes have the
virtue of obeying {\em energy estimates}, which are the chief
analytical tool in our work, and are widely used for basin-
and exploration-scale seismic modeling, acoustics computations,
structural dynamics, and many other purposes; see \cite{moczoetal:06} for an
excellent overview and many older references. Other classes of finite
difference approximations are also used in many of the same applications
\cite[]{Cohen:01,moczoetal:06,Petersson:2010,Petersson:2016}, as are finite element 
(spectral element, continuous and discontinous Galerkin) methods 
\cite[]{KomTromp:00,Cohen:01,Ghattas:IP25}. Results similar to those
reported here may hold for some of these approaches, however we do not
address this question here.

Staggered grid finite difference approximations $(u^n_h,v^{n+1/2}_h)$
to a solution  $(u,v)$ of
\ref{eq:hyper_fam} depend on the spatial grid cell size $h>0$ and the
time step $\Delta t>0$: $u^n_h$ approximates $u(\cdot,n\Delta t)$,
$v^{n+1/2}_h$ approximates $u(\cdot,(n+1/2)\Delta t)$. A sufficient
condition for consistent staggered grid schemes to produce convergent
approximations to 
smooth solutions of systems \ref{eq:hyper_fam} takes
the form of a {\em Courant-Friedrichs-Levy stability condition}, that is, a bound on
$\Delta t/h$, depending on the coefficient matrices $A,B$. In order to
express the error asymptotics of these schemes in a convenient way, we
will assume that $\Delta t$ is a function of $h$, obeying the
stability condition.

As pointed out earlier, point-wise convergence to singular solutions
cannot be expected, as the truncation error analysis on which it is
based makes no sense in the singular case. Instead, we turn to {\em
  weak} approximation: the approximate
solution $(u^n_h,v^{n+1/2}_h)$ converges weakly to continuum fields $(u,v)$ iff
arbitrary positively weighted averages of $(u^n_h,v^{n+1/2}_h)$
converge to the same weighted averages of
samplings of $(u,v)$ on the space-time $(h,\Delta t)$ grid, as $h,
\Delta t \rightarrow 0$ subject to the stability condition. 
The averaging weights are smooth functions of position (or position and time)
independent of grid, sampled at grid-points to create discrete weight functions.
Precise definitions are supplied below.

Our main result is that staggered grid approximations
$(u^n_h,v^{n+1/2}_h)$ 
converges weakly to a solution of \ref{eq:hyper_fam} with multipole right-hand side $(f,g)$, provided that the
finite-difference representation of the right-hand side
satisfies certain {\em discrete moment conditions}
\cite[]{Walden:1999,TorEng:04}, expressing accuracy of the 
right-hand side
approximation in a weak sense. ``Singular'' is used in
this paper as a synonym for {\em non-smooth distribution}, that is,
continuous linear functional on various topological vector spaces of
smooth functions, not representable by integration against a smooth kernel. For example, the right-hand side of the isotropic
radiator problem \ref{eq:isorad} defines a continuous linear
functional on the space of continuous functions with the sup norm. The
moment conditions used in this work simply specify that the numerical
approximations (linear combinations of grid values, with coefficients
depending on the grid diameter $h$) give the {\em same} result as the distribution being
approximated on polynomials with degree less than a bound depending on
the dimension $d$, the nature of the distribution (right-hand
side). This degree bound must be at least the {\em multipole order}
$N$, the highest derivative appearing in the definition
\ref{eq:MPSappx}. Express the degree bound as $N+q$, $q \ge 0$: we
call $q$ the {\em approximation order} of the discrete multipole representation.

The weak error asymptotics of a staggered grid scheme applied to a
problem with multipole right-hand side, using a discrete multipole
representation satisfying a discrete moment condition, depend on the
quantities introduced in the preceding paragraph, and on the error
asymptotics of the staggered grid scheme for smooth solutions. The
schemes we shall investigate produce approximations with error $O(\Delta t(h)^2 + h^p)$
(abbreviated as order $(2,p)$) to smooth solutions of systems \ref{eq:hyper_fam}.
%We give explicit rates of convergence, depending on on singular source
%approximation order $q$, spatial order $p$ for a staggered-grid
%finite difference scheme of with truncation error $O(\Delta t(h)^2,h^{p})$, the maximum multipole
%order between $f$ and $g$ denoted by $N^*$, and spatial dimension $d$.
%include explicitly the error bounds here
For spatial dimension $d$, multipole right-hand sides $(f,g)$  in
\ref{eq:hyper_fam}  of order $N$, discrete multipole
representation of approximation order $q$, and staggered grid scheme
of order $(2,p)$, we show that the error of averages with smooth averaging
functions $\tilde f, \tilde g$ satisfies the estimate
\begin{equation}
\label{eq:intro_weak_est}
\begin{split}
	\left| \int_0^T  \Big\{ \langle u,\tilde f\rangle + \langle v,\tilde g\rangle \Big\} dt -
		\Delta t \sum_{n=0}^N \Big\{ \langle u_h^{n}, \tilde f_h^{n} \rangle +
						 	   \langle v_h^{n+1/2}, \tilde g_h^{n+1/2}  \rangle \Big\} \right|\\
	= O(\Delta t(h)^2 + h^q + (\Delta t(h)^2+h^{p})h^{-N-d/2}).
\end{split}
\end{equation}
(Theorem \ref{thm:conv}). As discussed in the subsection on weak
convergence, staggered grid schemes typically involve multiple grids
with diameter $h$. In the estimate \ref{eq:intro_weak_est},
$\tilde{f}_h^{n}$ represents the samples on the grid(s) of diameter
$h$ of the smooth function $\bx \mapsto \tilde{f}(\bx,n\Delta t)$, and
similarly for $g$. Thus left-hand side of \ref{eq:intro_weak_est}
compares continuum ($L^2$) and discrete (weighted $l^2$) inner
products, the latter being the result of quadrature rules applied to
the former.

Note the implications of this estimate: to assure an error tending to
zero like $O(h^m)$, it is sufficient to require that $\Delta t(h) = O(h^{p/2})$, $q
\ge m$, $p \ge m +N+d/2$. These appear to be very stringent
requirements, compared to those for point-wise convergence to smooth solutions: for example for $N=0$ (that is, isotropic point source),
to achieve $m=2$ it is required that $q \ge 2$, $p \ge 2+d/2$ and $\Delta t(h) =
O(h^{1+d/4})$. In remarks after the proof of Theorem \ref{thm:conv},
we point out that if staggered grid approximations to {\em smooth} solutions can be shown to obey certain
$L^{\infty}$ estimates, as opposed to the $L^2$ estimates used here,
then the $h^{d/2}$ factor drops out of the right-hand side of the
estimate \ref{eq:intro_weak_est}, and the error asymptotics would
become optimal, that is, the same as those for smooth solutions. We
conjecture that such sup norm estimates for convergence of averages
hold, as our numerical results suggest, but their proof is beyond the
scope of this paper.

We assume throughout that the time-independent
coefficient matrices $A,B$ in the system \ref{eq:hyper_fam} are
smooth ($C^{\infty}$) in their dependence on the space variables. It
is certainly possible to extend some of these results to larger
classes of coefficients defined by weaker regularity
conditions - indeed, the weak solutions for systems of
the form \ref{eq:hyper_fam} exist if $A, B$ and their inverses are
uniformly bounded and measurable functions of the space variables
\cite[]{BlazekStolkSymes:13}. However the extent to which (and sense
in which) convergence
of finite difference methods might be established with such weak
regularity assumptions is an open question, beyond the scope of this work.

We analyze the approximation of waves in unbounded material models:
the spatial domain of wave propagation is the entire space $\bR^d$. We
consider only initial and right-hand side data of compact support in
space, so that the solutions of the wave propagation systems also have
compact support in space, for each time. Extension of our results to
settings in which waves have nontrivial interaction with boundaries is
possible in some cases, notably when the boundary conditions are
conservative or energy-dissipative. However we do not explicitly
formulate such generalizations, nor do we have anything to say about
sources located on physical boundaries. An effective extension to free
surface modeling in elasticity, for instance, is constrained by the
lack of conservative boundary conditions of order higher than
two. These are topics for further research.

We should note that numerical tests presented in the last section of
the paper suggest that, in the case of smooth coefficients, the actual 
quality of finite difference approximate solutions is better than our 
convergence result (or even the hypothetical ``optimal'') result
mentioned above) indicates: in some cases, away from the source 
point, we observe point-wise (strong) convergence with optimal 
order. We will mention this phenomenon again in the discussion 
section. The convergence behaviour suggested by the numerical examples 
is local, in regions bounded away from the source point. The 
convergence result proved here, on the other hand, is global.

We begin this paper with a review of the (rather sparse) prior work on
finite-difference approximation of linear partial differential 
equations with singular right-hand sides, emphasizing results for 
hyperbolic problems. Then we give a detailed discussion of the moment conditions which form the {\em a priori} conditions for accuracy of 
source approximations. We then explain the convergence theory, based 
on energy estimates, for hyperbolic systems with singular right-hand 
sides. We introduce linear acoustics as an example. We have performed 
a number of numerical experiments that illustrate the results of the 
theory - we present these next. We conclude with a discussion of
matters not addressed in the body of this paper, and a summary of the 
results.

%Numerical results presented here, consistent with other similar works (e.g., \cite{Petersson:2010}), appear to %indicate, however, stronger convergence results: optimal convergence point-wise away from source.
%In particular, we report second and fourth order rates when studying the spatial convergence of 2-2 and 2-4 %finite difference methods respectively, where the source approximation order matched the spatial order of th%e numerical scheme.

%Though numerical results, both reported in other works (e.g., \cite{Petersson:2010}) and in this paper, demonstrate stronger results, mainly optimal convergence point-wise away from source, a weak convergence error estimate is an appropriate first step towards a complete theory.

%There are, of course, many ways to discretize/regularize singularities.
%However, we are primarily concerned with direct discretization of the Dirac delta function and its derivatives through the use of discrete moment conditions, as done by \cite{Walden:1999} in a finite difference and finite element setting for source terms in the 1-D Helmholtz equation.
%His analysis and numerical examples, though limited to his particular problem, demonstrated point-wise convergence of numerical solutions with optimal convergence rates (as suggested by the numerical scheme) away from the source location when appropriately discretizing the singular source term.

\subsection{Literature Review}

The analysis and numerical examples presented by \cite{Walden:1999}, though limited to the 1-D Helmholtz equation, demonstrated point-wise convergence of numerical solutions with optimal convergence rates (as suggested by the numerical scheme) away from the source location when appropriately discretizing the singular source term.
The theory of singular source approximations has been further extended to a range of applications, most notably for the Dirac delta function in the context of the {\em immersed boundary method} \citep{Pes:02}. 
Several authors have addressed questions regarding the convergence of source approximations and subsequently their effect on solutions to more complicated differential equations.
Consider the following abstract problem: find $u$ such that
\begin{equation}\label{eq:pde}
	\mathcal L u = f,
\end{equation}
in which $\mathcal L$ is a differential operator and the right-hand
side $f$ may be singular.
Define the regularization of problem \ref{eq:pde} by replacing $f$ with some regular (at least piecewise continuous) function $f^\epsilon$ parameterized by regularization parameter $\epsilon>0$, that is, find $u^\epsilon$ such that
\begin{equation}\label{eq:pde_reg}
	\mathcal L u^\epsilon = f^{\epsilon}.
\end{equation}
Regularized source term $f^\epsilon$ is said to approximate $f$ in that $f^\epsilon \to f$ as $\epsilon \to 0$ in some sense.
The end goal is of course to have $u^\epsilon$ approximate the continuum solution $u$, i.e.,
\[
	\lim_{\epsilon\to 0}\|u^\epsilon-u\|_X = 0
\]
under some suitable norm $\|\cdot\|_X$.

\cite{TorEng:03} have studied the regularization error $\|u^\epsilon - u\|_X$, point-wise away from the source location.
Their analysis is based on a simple ODE case where they prove convergence of regularized solutions $u^\epsilon$ if $f^{\epsilon}$ satisfy what we call the {\em continuum moment conditions}. 
We use the qualifier ``continuum'' to differentiate at times between the discrete moment conditions.
Recent work by \cite{hoss:16} addresses the mode of convergence of $f^\epsilon\to f$ subject to regularized source terms satisfying the continuum moment conditions, mainly convergence in a weak-$*$ topology (distribution sense) and in a weighted Sobelev norm.
Both \cite{TorEng:03} and \cite{hoss:16} argue that $f^{\epsilon}\to f$ implies $u^\epsilon\to u$ as $\epsilon \to 0$, point-wise away from the source location, in particular for elliptic operators $\mathcal L$. 
This argument hinges on the integral representation of elliptic operators and the smoothness of their kernel (i.e., Green's functions) away from source location.

Suppose that the regularized problem \ref{eq:pde_reg} is discretized, with mesh or cell size $h>0$;
\begin{equation}\label{eq:pde_disc}
	\mathcal L_h u^{\epsilon}_h = f^{\epsilon}_h.
\end{equation}
In the context of finite difference methods $\mathcal L_h$ is the finite difference approximation of differential operator $\mathcal L$.
The discrete source term $f^\epsilon_h$ can be interpreted as the discretization 
(e.g., sampling over grid points) of the regularized source term $f^\epsilon$  generated by the continuum moment conditions, or as the direct discretization of $f$ through the discrete moment conditions.
In practice the regularization parameter $\epsilon$ is related to the discretization parameter $h$, that is, $\epsilon = \epsilon(h)$ such that
\[
	\lim_{h\to 0} \epsilon(h) = 0.
\]
\cite{TorEng:04} provided insight into the convergence of $f^\epsilon_h$ as $h\to 0$ for direct discretizations of $f$ via the discrete moment conditions, in particular for discretizations whose support is proportional to $\epsilon$ and $\epsilon(h) = O(h)$.
Consistent with results by \cite{Walden:1999}, \cite{TorEng:04} demonstrated the convergence of numerical solutions $u^\epsilon_h$ for problem \ref{eq:pde_disc}, in particular
\[
	\| u^\epsilon_h - u\|_X = O(h^p)
\]
where $p$ is the convergence rate of the numerical scheme and $f^{\epsilon}_h$ satisfies a sufficient number of moment conditions. 
The norm $\|\cdot\|_X$ in this case coincides with the sup-norm with a deleted neighborhood containing the source location.
Theory presented by \cite{TorEng:04} is based on analysis of the 1-D Poisson equation discretized by second- and fourth-order finite difference approximations.%, in particular, studying the numerical Green's functions.

Green's functions for hyperbolic problems are singular at the wavefront they propagate as well as at the source location, thus the analysis of \cite{TorEng:04} does not apply.
\cite{Petersson:2016} studied a finite difference approximations to various
advection systems, and to the linear acoustodynamics, with singular
source terms. based on centered difference approximations and accurate
Runge-Kutta time differentiation. An plane wave analysis revealed that
both moment conditions of the type discussed here, and additional {\em smoothness}
constraints on the source representation
They show that the discrete moment conditions are necessary but not sufficient for the convergence of numerical solutions at optimal rates away from the source location.
They demonstrated that {\emph smoothness constraints} on the numerical
solution are also required to achieve convergence, due to the presence
of spurious modes injected by the singular source approximation. 
%The main result of \cite{Petersson:2016} is however suboptimal, in that
%\[
%	\|u^\epsilon_h - u\|_2 \le Ch^{p-\frac{1}{2m}}
%\]
%where $p$ is the order of the finite difference scheme and the number of moment and smoothness conditions satisfied by the source discretization, and integer $m$ is related to the regularity of the time-dependent component of the singular source term.

%%%
%Our main contribution is a weak convergence theory of
% $u^\epsilon_h\to u$ that is, unlike current convergence analysis
% that is limited by its particular application, applicable to a large
% set of symmetric hyperbolic problems solved via staggered grid
% finite difference methods.
%%Though numerical results, both reported in other works and in this
%% paper, demonstrate stronger results, mainly point-wise convergence
%% away from source, we believe a weak convergence error estimate is
%% an appropriate first step towards a complete theory.
%The remainder of the paper is organized as follows: In the theory
% section we begin by presenting an overview of the singular source
% approximation method via continuum and discrete moment conditions.
% perhaps include regularity error estimates for 3-D acoustics? We
% focus primarily on source approximations of narrow support, as
% discussed in \cite{TorEng:04}, though we provide explicit formulas
% for approximations of arbitrary order. Moreover, we show that the
% discrete moment conditions in fact define a sequence of continuum
% functions that converge to target distributions in a weak sense, a
% new result. The theory section also covers known convergence results
% of staggered-grid finite difference methods via energy estimates,
% applied to our set of differential equations under smooth
% coefficients and smooth source terms. At this point we present our
% weak convergence theory. The last section covers numerical results,
% mainly for staggered-grid finite difference solutions to the 2-D
% acoustic equations in first-order form with multipole
% sources. Consistent with numerical results presented in the
% literature (see for example \cite{Petersson:2010}), we observe
% optimal convergence rates of our numerical solutions when the source
% discretization satisfied the proper number of moment conditions.

\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory}
%%%%%%%%%%%%%%%%%%%%%%%%%

The weak convergence theory of finite difference solutions to hyperbolic
systems with multipole sources rests on two foundations: (i) the
analysis of weak convergence conditions for gridded approximations to
point support distributions, and (ii) the convergence theory for
smooth solutions with smooth right-hand sides. The next two
subsections develop these foundations.

%%%%
\subsection{Singular Source Approximation}
%%%%

%%%%%%%
\subsubsection{(Continuum) Moment Conditions}
%%%%%%%

%

Let $\mathcal D$ denote the {\em space of test functions} over $\mathbf R^d$, that is, the space of $C^\infty_0(\mathbf R^d)$ endowed with the standard topology of test functions.
The set of \emph{distributions} is the topological dual of the space
of test functions, and is commonly denoted by $\mathcal D'$.
It is conventional to represent the application of a distribution on a function by the integral of the product, even when the distribution is not actually a function that can be integrated in the usual sense.
For example, given multi-index $\mathbf s=(s_1,...,s_d)$, the $\mathbf s$-mixed partial derivative of the Dirac delta function, shifted by $\mathbf x^*\in\mathbf R^d$, is defined by
\[
	\int_{\mathbf R^d} D^{\mathbf s} \delta(\mathbf x-\mathbf x^*) \; \psi(\mathbf x)\; d\mathbf x =
	(-1)^{|\mathbf s|} D^{\mathbf s} \psi(\mathbf x^*), \quad \forall \psi \in\mathcal D.
\]
%The distribution $D^{\mathbf s}\delta(\cdot;\mathbf 0)$ is simply denoted by $D^{\mathbf s}\delta$.
%In the remainder of the paper we focus on deriving the approximation theory for $\mathbf x^*=\mathbf 0$ (assuming $\mathbf 0\in\Omega$) since the general case follows by simply shifting the approximation accordingly.

%
The key idea for constructing approximations to $D^{\bf s}\delta(\mathbf x-\mathbf x^*)$ is based on mimicking the behavior of the target distribution on polynomials, reminiscent of finite difference approximations for differential operators.
Consider $\psi(\mathbf x) = (\mathbf x - \mathbf x^*)^{\alpha}$, with multi-index $\alpha=(\alpha_1,...,\alpha_d)$, where multi-indexed monomials are interpreted as the product of monomials in each dimension,
\[
       {\bf x}^{\alpha} = \prod_{k=1}^d x_k^{\alpha_k}.
\] 
It can be shown that
\[
	\int_{\mathbf R^d} D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)\; \psi(\mathbf x) \; d\mathbf x = \mathbf s! (-1)^{|\mathbf s|} \delta_{\mathbf s \alpha}
\]
where $\delta_{\mathbf s \alpha}$ is the Kronecker delta, defined as follows for multi-indexes,
\[
	\delta_{\mathbf s \alpha} := \prod_{k=1}^d \delta_{s_k \alpha_k}.
\]

%
Given $\eta\in L^1_0(\mathbf R^d)$ (i.e., integrable function of compact support) and multi-index $\alpha$, the {\em $\alpha$-moment} of $\eta$ centered at $\mathbf x^*\in\mathbf R^d$, denoted by $M^\alpha(\cdot,\mathbf x^*)$, is defined as
%definition of moment
\begin{equation}\label{eq:def_mom}
	M^\alpha(\eta,\mathbf x^*) := \int_{\mathbf R^d} \eta(\mathbf x) \; (\mathbf x-\mathbf x^*)^{\alpha} \; d\mathbf x.
\end{equation}
%Note that integration in equation \ref{eq:def_mom} is translation invariant hence $M_{\alpha}(\eta,\mathbf x^*)$ is constant with respect to $\mathbf x^*$, which I denote by $M_{\alpha}(\eta)$.
For given nonnegative integer $q$ and multi-index $\mathbf s$, the function $\eta$ is said to satisfy the {\em continuum $(q,\mathbf s)$-moment conditions} at $\mathbf x^*\in \mathbf R^d$ if
%definition of moment conditions
\begin{equation}\label{eq:momcond}
	M^\alpha(\eta,\mathbf x^*) = \mathbf s! (-1)^{|\mathbf s|} \delta_{\mathbf s \alpha}, \quad \forall |\alpha|=0,...,q+|\mathbf s|-1.
\end{equation}

If $\eta$ satisfies the $(q,\mathbf s)$-moment conditions at $\mathbf x^*$, then its associated distribution, that is
\[
	\int_{\mathbf R^d} \eta(\mathbf x) \; \psi(\mathbf x)\; d\mathbf x, \quad \forall \psi\in\mathcal D,
\]	
is an approximation to $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$ in that it is exact on polynomials of order $q+|\mathbf s|-1$.

The following theorem states that a sequence of (regular) distributions of compact support, satisfying the $(q,\mathbf s)$-moment conditions, will converge in the weak-$*$ topology at a rate $q$ to the target distribution as the width of the supports approach zero. 
Let $B(\mathbf x^*,\epsilon)$ denote the $d$-dimensional ball of radius $\epsilon$ centered at $\mathbf x^*$.

%theorem weak convergence of approximations to distribution
\begin{theorem}\label{thm:weakconv}
	Let nonnegative integer $q$, multi-index $\mathbf s$, and $\mathbf x^*\in\mathbf R^d$ be given.
	Suppose $\{\eta^\epsilon\}\subset L^1_{0}(\mathbf R^d)$ is a sequence of functions as $\epsilon\to 0$, where $supp(\eta^\epsilon)\subset B(\mathbf x^*,\epsilon)$.
	Furthermore, suppose that there exists a constant $K>0$ independent of $\epsilon$ such that 
	\begin{equation}\label{eq:boundH}
		\int_{\mathbf R^d} |\eta^\epsilon(\mathbf x)|\; |(\mathbf x-\mathbf x^*)^{\alpha}| \; d\mathbf x \le K, \quad \forall |\alpha| = |\mathbf s|.
	\end{equation}
	If $\{\eta^\epsilon\}$ satisfy the $(q,\mathbf s)$-moment conditions at $\mathbf x^*$, equation \ref{eq:momcond}, then the sequence of distribution they generate converges to $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$ in the weak-$*$ topology as $\epsilon\to 0$.
	In particular, if $\psi$ is of class $C^{q+|\mathbf s|}$ over $B(\mathbf x^*,\epsilon)$, then
	\begin{equation}\label{eq:delta_error}
		E:= \left| \int_{\mathbf R^d} D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)\;\psi(\mathbf x) \; d\mathbf x\; - 
		\int_{\mathbf R^d} \eta^\epsilon(\mathbf x) \; \psi(\mathbf x)\; d\mathbf x \right| = O(\epsilon^q).
	\end{equation}
\end{theorem}

%PROOF
\begin{proof}
We first apply multi-variate Taylor's theorem to $\psi$, centered at $\mathbf x^*$ and truncated after $N=q+|\mathbf s|-1$ terms \cite[]{konigsberger2013}, assuming $\psi$ is $C^{q+|\mathbf s|}$ over $B(\mathbf x^*,\epsilon)$,
\begin{align*} 
	\int_{\mathbf R^d} \eta^\epsilon (\mathbf x) \; \psi(\mathbf x) \; d\mathbf x
	&= \int_{\mathbf R^d} \eta^\epsilon(\mathbf x) \left( \sum_{|\alpha|=0}^{N} \frac{D^{\alpha} \psi(\mathbf x^*)}{\alpha!} \; (\mathbf x - \mathbf x^*)^\alpha 
	+ \sum_{|\beta|=N+1}R_{\beta}(\mathbf x) \; (\mathbf x - \mathbf x^*)^\beta \right)
	d\mathbf x \\
	&= \sum_{|\alpha|=0}^N \frac{D^{\alpha}\psi(\mathbf x^*)}{\alpha!}  \left( \int_{\mathbf R^d}  \eta^\epsilon(\mathbf x) \; (\mathbf x -\mathbf x^*)^\alpha \;d\mathbf x\right) 
	+  \sum_{|\beta|=N+1} \int_{\mathbf R^d} \eta^{\epsilon}(\mathbf x) R_{\beta}(\mathbf x) \; (\mathbf x-\mathbf x^*)^\beta \; d\mathbf x,
\end{align*}
where $R_{\beta}$ is the remainder term,
\[
	R_\beta(\mathbf x) = \frac{|\beta|}{\beta!} \int_{0}^1 (1-t)^{|\beta|-1} D^\beta \psi(\mathbf x^* + t(\mathbf x-\mathbf x^*)) \; dt.
\]
Note that the term in the parenthesis in the bottom equation corresponds to the $\alpha$-moment centered at $\mathbf x^*$ with $|\alpha|\le q+|\mathbf s|-1$, hence the ($q,\mathbf s$)-moment conditions apply;
\begin{align*}
	\int_{\mathbf R^d} \eta^\epsilon(\mathbf x) \; \psi(\mathbf x)\;d\mathbf x
	&= \sum_{|\alpha|=0}^{N} \frac{1}{\alpha!} D^{\alpha}\psi(\mathbf x^*) \Big( \mathbf s! (-1)^{|\mathbf s|}\delta_{\mathbf s \alpha} \Big) 
	+  \sum_{|\beta|=N+1} \int_{\mathbf R^d} \eta^\epsilon(\mathbf x) R_{\beta}(\mathbf x)\; (\mathbf x-\mathbf x^*)^\beta \; d\mathbf x\\
	&= (-1)^{|\mathbf s|}D^{\mathbf s} \psi(\mathbf x^*) 
	+  \sum_{|\beta|=N+1} \int_{\mathbf R^d} \eta^\epsilon (\mathbf x) R_{\beta}(\mathbf x) \; (\mathbf x-\mathbf x^*)^\beta \; d\mathbf x.
\end{align*}

The remainder term is bounded uniformly over $B(\mathbf x^*,\epsilon)$,
\[
	\underset{\mathbf x\in B(\mathbf x^*,\epsilon)}{\sup} |R_\beta(\mathbf x)| \le C(\beta,\psi) := \frac{1}{\beta!} \max_{|\alpha|=|\beta|} \; \max_{\mathbf y\in B(\mathbf x^*,\epsilon)} |D^{\alpha}\psi(\mathbf y)|,
\]
using the fact that $\psi\in C^{N+1}$ in $B(\mathbf x^*,\epsilon)$.
This gives the following error estimate,
\begin{align*}
	 E
	 & \le \sum_{|\beta|=N+1} C(\beta,\psi) \int_{B(\mathbf x^*,\epsilon)} 
	  |\eta^\epsilon(\mathbf x)| \; |(\mathbf x-\mathbf x^*)^\beta| \; d\mathbf x.
\end{align*}
Let $\gamma$ be a multi-index such that $|\gamma|=q$, thus $|\beta-\gamma| = |\mathbf s|$.
This yields,
\begin{align*}
	E
	&\le \sum_{|\beta|=N+1} C(\beta,\psi) \left( \sup_{\mathbf x\in B(\mathbf x^*,\epsilon)} \left|(\mathbf x-\mathbf x^*)^{\gamma}\right|\right) 
	\int_{B(\mathbf x^*,\epsilon)} |\eta^\epsilon(\mathbf x)| \; |(\mathbf x-\mathbf x^*)^{\beta-\gamma}| \; d\mathbf x \\
	&\le \sum_{|\beta|=N+1} C(\beta,\psi) \left( \sup_{\mathbf x\in B(\mathbf x^*,\epsilon)} \left|(\mathbf x-\mathbf x^*)^{\gamma}\right|\right)  K\\
	&= O(\epsilon^q)
\end{align*}
\end{proof}
%END PROOF

Theorem \ref{thm:weakconv} and moment conditions given in equation \ref{eq:momcond} are extensions of what is presented in \cite{hoss:16} for $|{\bf s}|\neq0$.
Given equation \ref{eq:delta_error}, we refer to $q$ as the \emph{singular source approximation order} and $\eta^\epsilon$ as being a $q$-order approximation of $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$.

%%%%%%%
\subsubsection{Discrete Moment Conditions}
%%%%%%%

Define the regular grid family $\mathcal G_h(\mathbf x_0,\mathbf l)$
with mesh or scale $h$, offset $\bx_0 \in
\bR^d$ and cell given by ${\bf l} \in \bR_+^d$, namely $\{\bx \in
\bR^d: 0 \le x_i\le l_k\}$:
\[
	\mathcal G_h(\mathbf x_0,\mathbf l) = \{\bx_{\bf n} =
        h((x_0)_1+n_1l_1,...,(x_0)_d+n_dl_d), {\bf n} \in {\bf Z}^d\}
\]
We will sometimes refer to the scaled cell side lengths $h_k=hl_k,k=1,...,d$.
%The dependence of real-valued grid functions with respect to a given regular grid $\mathcal G_h(\mathbf x_0,\mathbf h)$, say $\eta^h: \mathcal G(\mathbf x_0,\mathbf h)\to \mathbf R$, is made implicit through the use of multi-indexing notation.
%For example, given $\mathbf x_{\mathbf n}\in \mathcal G(\mathbf x_0,\mathbf h)$ for some multi-index $\mathbf n=(n_1,...,n_d)$,
%\[
%	\eta^h_{\mathbf n}= \eta^h(\mathbf x_{\mathbf n}).
%\]
Note that there is no reason to assume that the grid cell is cubical: it may have different lengths along different axes.

The obvious definition of the \emph{discrete $\alpha$-moment}, centered at $\mathbf x^*\in\mathbf R^d$ (note that $\mathbf x^*$ need not coincide with a grid point in $\mathcal G_h(\mathbf x_0,\mathbf l)$), of a grid function $\eta_h:\mathcal G_h(\mathbf x_0,\mathbf l)\to \mathbf R$ is given as follows: 
\begin{equation*}\label{eq:def_discmom}
        M_h^\alpha(\eta_h,\mathbf x^*) := \left(\prod_{k=1}^{d} h_k \right) \sum_{\mathbf x\in\mathcal G_h(\mathbf x_0,\mathbf l)}  \eta_h(\mathbf x)\; ({\bf x}-{\bf x}^*)^{\alpha}.
\end{equation*}
It is worth pointing out that the discrete moment defined above is dependent on choice of grid, in particular dependent on the source location $\mathbf x^*$ relative to the grid.
Similar to the continuum moment conditions (equation \ref{eq:momcond}), grid function $\eta_h$ is said to satisfy the \emph{discrete $(q,\mathbf s)$-moment conditions} at $\mathbf x^*\in\mathbf R^d$ if
\begin{equation}\label{eq:discmomcond}
	M_h^\alpha(\eta_h,\mathbf x^*) = \mathbf s! (-1)^{|\mathbf s|} \delta_{\mathbf s \alpha}, \quad \forall |\alpha|=0,...,q+|\mathbf s|-1.
\end{equation}
The following theorem is a discrete analogue of theorem \ref{thm:weakconv}.
%the discrete $(q,\mathbf s)$-moment conditions imply convergence of gridded functions $\{S^h\}$ as the characteristic grid size $h$ is refined, assuming the support of $S^h$ is proportional to $h$, in a ``discrete'' weak-$*$ topology.
%theorem discrete weak convergence for approximations to distribution

\begin{theorem}\label{thm:discweakconv}
Let nonnegative integer $q$, multi-index $\mathbf s$, and $\mathbf x^*\in\mathbf R^d$ be given.
Suppose $\{\eta_h^\epsilon\}$ is a sequence of grid functions $\eta_h^\epsilon:\mathcal G_h(\mathbf x_0,\mathbf l)\to\mathbf R$ as $\epsilon\to 0$.
Furthermore, assume that the support of $\eta_h^\epsilon$ is contained in $B(\mathbf x^*,\epsilon)$ with $\epsilon=O(h)$, and that there exists constant $K>0$ independent of $\epsilon$ such that
\begin{equation*}\label{eq:boundh}
	\left(\prod_{k=1}^d h_k \right) \sum_{\mathbf x\in\mathcal G_h(\mathbf x_0,\mathbf l)} |\eta_h^\epsilon(\mathbf x)|\; |(\mathbf x-\mathbf x^*)^{\alpha}| \le K, \quad \forall |\alpha| = |\mathbf s|.
\end{equation*}
If $\{\eta_h^\epsilon\}$ satisfy the discrete $(q,\mathbf s)$-moment conditions at $\mathbf x^*$ (equation \ref{eq:discmomcond}) and $\psi$ is of class $C^{q+|\mathbf s|}$ over $B(\mathbf x^*,\epsilon)$, then 
\[
	\left| \int_{\mathbf R^d} D^{\mathbf s}\delta(\mathbf x- \mathbf x^*) \;\psi(\mathbf x)\; d\mathbf x - \left(\prod_{k=1}^d h_k\right) \sum_{\mathbf x\in\mathcal G_h(\mathbf x_0,\mathbf l)}\eta_h^\epsilon (\mathbf x) \; \psi(\mathbf x)\right| = O(h^q).
\]
\end{theorem}

%%
\begin{proof}
The proof of this theorem is omitted since it is nearly identical to that of the continuum case (theorem \ref{thm:weakconv}), replacing integrals with summations over grid points.
The jump from $O(\epsilon^q)$ to $O(h^q)$ follows from $\epsilon=O(h)$.
\end{proof}

Theorem \ref{thm:discweakconv} and discrete moment conditions \ref{eq:discmomcond} are generalizations of work by \cite{TorEng:04} for $|\mathbf s|\neq 0$. 
The discrete moment conditions are also an extension of \cite{Walden:1999} for dimensions higher than one.
We now discuss with more detail how to construct said sequences of gridded and continuum functions starting in 1-D.

%%%%
\subsubsection{1-D Constructions}
%%%%%

We make the choice of having 1-D gridded approximations be centered around source location $x^*\in\mathbf R$, and define them to be zero outside the interval $[-\epsilon+x^*,\epsilon+x^*)$, with $2\epsilon=Nh$ for some positive integer $N$.
In other words, there exists $N$ grid points, denoted by $\{\tilde x_\ell\}_{\ell=1}^{N}$, such that they are contained in the interval $[-\epsilon+x^*,\epsilon+x^*)$ for a given grid $\mathcal G_h(x_0,1)$\footnote{We assume $l=1$ for simplicity, since any scaling effects $l$ may have can be incorporated into $h$ in 1-D}.
These grid points $\{\tilde x_\ell\}_{\ell=1}^N$ are referred to as the \emph{stencil points} of the approximation.
We assume that the stencil points are ordered, i.e., $\tilde x_1<\tilde x_2<\cdots<\tilde x_{N}$.
The discrete $(q,s)$-moment conditions thus results in a $N\times(q+s)$ system of equations for the grid function $\eta_h^\epsilon$ evaluated at stencil points,
\[
	\mathbf A \mathbf d = \mathbf b
\]
with 
\[
	\{\mathbf A\}_{k\ell} = (\tilde x_\ell-x^*)^{k-1}, \quad \{\mathbf d\}_{\ell}=\eta_h^\epsilon(\tilde x_\ell), \quad \{\mathbf b\}_{k}= \frac{s! (-1)^s}{h} \delta_{s,k-1},
\]
for $\ell=1,...,N$ and $k=1,...,q+s$.
Note that $\mathbf A$ is a Vandermonde matrix of full rank and is guaranteed a solution if $N\ge q+s$ and no solution for $N<q+s$ under general $x^*\in\mathbf R$.
%It will be of benefit to pick grid functions of minimal support, that is $N=q+s$; this will be more apparent when coupling source approximations with finite difference schemes.

We choose the case where $N=q+s$, which we refer to as grid functions of narrow support.
The system above will result in a unique solution for a given $x^*\in\mathbf R$.
In fact the inverse matrix for $\mathbf A$ can be written explicitly using the following Vandermonde matrix inverse formula:
%VANDERMONDE INVERSE
\begin{equation*}\label{eq:vandinv}
	\{\mathbf A^{-1} \}_{\ell k} = \left\{ \begin{array}{cl}
		\displaystyle
		(-1)^{N-k} \; 
		\frac{
			\displaystyle
			\sum_{\underset{m_1,...,m_{N-k}\neq \ell}{1\le m_1<\cdots<m_{N-k}\le N}}
			 (\tilde x_{m_1}-x^*) \cdots (\tilde x_{m_{N-k}}-x^*)
		}
		{
			\displaystyle
			\prod_{\underset{m\neq \ell}{1\le m\le N}} (\tilde x_\ell - \tilde x_m)
		}
		,& \text{for} \; 1\le k\le N  \vspace{10pt} \\ 
		\displaystyle
		\frac{
			1
		}
		{
			\displaystyle
			\prod_{\underset{m\neq \ell}{1\le m\le N}} (\tilde x_\ell - \tilde x_m)
		}
		,& \text{for}\; k=N.
	\end{array}\right.
\end{equation*}
Given the particular form of right-hand side vector $\mathbf b$, it follows that $\mathbf d$ is simply the scaled $(s+1)$-column of $\mathbf A^{-1}$, whence
%GRID SOLUTION
\begin{equation}\label{eq:gridfunsol}
	\eta_h^\epsilon(\tilde x_\ell) = \left\{ \begin{array}{cl}
		\displaystyle
		s! (-1)^{N-1} \; 
		\frac{
			\displaystyle
			\sum_{\underset{m_1,...,m_{q-1}\neq \ell}{1\le m_1<\cdots<m_{q-1}\le N}}
			 (\tilde x_{m_1}-x^*) \cdots (\tilde x_{m_{q-1}}-x^*)
		}
		{
			\displaystyle
			h^{N} \prod_{\underset{m\neq \ell}{1\le m\le N}} (\ell - m)
		}
		,& \text{for} \; q>1  \vspace{10pt} \\ 
		\displaystyle
		\frac{
			s!(-1)^{s}
		}
		{
			\displaystyle
			h^N \prod_{\underset{m\neq \ell}{1\le m\le N}} (\ell - m)
		}
		,& \text{for}\; q=1.
	\end{array}\right.
\end{equation}

Note that the equation above above is dependent on $x^*$.
To be more precise, approximation $\eta_h^\epsilon$ is actually dependent on the relative position of $x^*$ with respect to the stencil points $\{\tilde x_\ell\}_{\ell=1}^N$, or equivalently dependent on $x_0$ and $h$.
For example, shifting the source location by $h$ would result in the same grid function $\eta_h^\epsilon$ though shifted by a grid point.
However, arbitrary shifts in $x^*$ yield approximations that may vary by more than a simple translation.
%Intuitively enough, the same cannot be said of the continuum case. 
%In other words, if $\eta(x)$ is a $q$-order approximation of $D^{s}\delta(x)$ then it follows that $\eta(x-x^*)$ is also a $q$-order approximation of $D^{s}\delta(x-x^*)$, for all $x^*\in\mathbf R$.
Given that equation \ref{eq:gridfunsol} is unique for a particular source location, we show that imposing the discrete moment conditions over all $x^*\in\mathbf R$ indeed defines a function over the reals.
Moreover, we show that these continuum functions satisfy the continuum moment conditions and thus define a sequence of distributions that converge to $D^{s}\delta$ in the weak-$*$ topology, a result previously not known.

%Equation \ref{eq:gridfunsol} is used to generate gridded functions $\eta^{h_k}$ satisfying the discrete $(q,s_k)$-moment conditions at $x^*_k\in\mathbf R$ on each dimension $k=1,...,d$.
%Approximations on higher dimension are then constructed by taking the tensor product of the 1-D approximations.
%Note, however, that in order for theorem \ref{thm:discweakconv} to apply it is required that the multi-variate grid function, that is $\eta^h(\mathbf x) = \eta^{h_1}(x_1) \cdots \eta^{h_d}(x_d)$, satisfy the discrete $(q,\mathbf s)$-moment conditions.
%Though it may not appear obvious, the proposition below shows that my tensor construction indeed satisfies the appropriate discrete moment conditions.
%
%%theorem proving tensor prod appx satisfy discrete moment conditions
%\begin{theorem}\label{thm:prodmomcond}
%	Let $q\in\mathbf N$, multi-index $\mathbf s$, and $\mathbf x^*\in\mathbf R^d$ be given. 
%Suppose $\eta^h:\mathcal G(\mathbf x_0,\mathbf h)\to \mathbf R$ is a multi-variate grid function given by the tensor product of 1-D grid functions $\eta^{h_k}:\mathcal G(x_{0,k},h_k)\to\mathbf R$.
%If $\eta^{h_k}$ satisfy the discrete $(q,s_k)$-moment conditions at $x^*_k$ for all $k=1,...,d$, then it follows that $\eta^h$ satisfies the discrete $(q,\mathbf s)$-moment conditions at $\mathbf x^*$.
%\end{theorem}
%
%%%
%\begin{proof}
%Suppose for each $k=1,...,d$ that $\eta^{h_k}$ satisfies the discrete $(q,s_k)$-moment conditions at $x^*_k$. 
%Let $\alpha$ be some multi-index with $0\le |\alpha|\le q+|\mathbf s|-1$. 
%Note that,
%\[
%	M^{h}_{\alpha}(S^{h},\mathbf x^*) = \prod_{k=1}^d M^{h_k}_{\alpha_k}(\eta^{h_k},x^*_k).
%\]
%Clearly, if $\alpha_k\le q+s_k-1$ for all $k=1,...,d$, then the result follows from the supposition. 
%Same applies for $\alpha=\mathbf s$.
%Suppose then, that there exists index $\ell$ such that $\alpha_\ell>q+s_\ell-1$, that is $a_\ell=q+s_\ell -1 +i$ for some $i\in\mathbf N$. Thus,
%\begin{align*}
%	& |\alpha| = \sum_{k\neq \ell} \alpha_k + \alpha_\ell = \sum_{k\neq \ell} \alpha_k + q+s_\ell-1+i \le q + |\mathbf s| -1\\
%\Longrightarrow	&  \sum_{k\neq\ell} \alpha_k + i \le \sum_{k\neq \ell} s_k.
%\end{align*}
%which implies that $\alpha_k<s_k$ for at least one $k\neq\ell$;
%for this particular $k$, it follows that 
%\[
%	M^{h_k}_{\alpha_k}(\eta^{h_k},x^*_k) = s_k!(-1)^{s_k} \delta_{s_k \alpha_k} = 0
%\]
%since it has been established that $\alpha_k\neq s_k$, i.e., the product over $k$ is zero if $\mathbf s\neq \alpha$.
%\end{proof}
%
%It is important to note that grid functions $\eta^h$ as computed by equation \ref{eq:gridfunsol}, will be dependent on three things:
%\begin{description}
%	\item{1.} source location $\mathbf x^*$,
%	\item{2.} choice of grid $\mathcal G(\mathbf x_0,\mathbf h)$,
%	\item{3.} and choice of support of $\eta^h$.
%\end{description}
%The last point refers to the fact that the center of the support, that is $a^*$, was never specified above; different choices of $a^*$ will result in different grid functions.
%In practice, however, I will consider only grid functions that are centered at source location $x^*$, as shown in section \ref{sc:examples}.

%%%%%%%%
%\subsection{Relating Discrete and Continuous Moments Conditions}
%%%%%%%%

%The discrete $(q,\mathbf s)$-moment conditions, for a given $\mathbf x^*$, yield gridded approximations from an explicit formula (equation \ref{eq:gridfunsol}) assuming that the diameter of the support of the grid function is equal to $(q+s_k)h_k$ in each dimension.

%%%%%%
\subsubsection{Connection between Continuum and Discrete Moment Conditions}

We first focus on the $x^*=0$ case and define our continuum approximation $\eta^\epsilon$ to be zero outside $[-\epsilon,\epsilon)$, with $2\epsilon=Nh$ for $N=q+s$. 
Furthermore, we define $\eta^\epsilon$ to be piecewise polynomial over $N$ \hl{intervals}: 
%ETA
 \begin{equation}\label{eq:eta}
 	\eta^\epsilon(x) = \left\{ \begin{array}{rl}
		P_\ell(x),& \quad x\in[a_\ell,a_{\ell+1}), \; \text{for}\; \ell=1,...,N\\
		0,& \quad \text{otherwise}
	\end{array}\right.
 \end{equation}
where $P_\ell$ is some polynomial over the considered interval, and $a_{\ell}=-\epsilon+ (\ell-1)h$ for $\ell=1,...,N+1$.
Given the support of our approximation, and a regular grid $\mathcal G_h(x_0,1)$, it follows that there are $N$ grid points contained in the interval $[-\epsilon,\epsilon)$, again denoted by $\{\tilde x_\ell\}_{\ell=1}^N$.
In fact,
\[
	\tilde x_\ell \in [a_\ell,a_{\ell+1}), \quad \forall \ell=1,...,N.
\]
Let $\ell^*$ be the index such that $0\in[a_{\ell^*},a_{\ell^*+1})$ and define $\zeta\in(0,h]$ by $\zeta=a_{\ell^*+1}-\tilde x_{\ell^*}$.
Thus, if we vary $x^*$ within the interval $(-\zeta,h-\zeta]$ if follows that
\[
	\tilde x_\ell - x^* \in [a_\ell,a_{\ell+1}), \quad \ell=1,...,N.
\]

Let $\eta_h^\epsilon(\cdot;x^*)$ denote the grid function that satisfies the discrete $(q,s)$-moment conditions for a given $x^*\in(-\zeta,h-\zeta]$.
Then $\eta^\epsilon(\tilde x_\ell - x^*) := \eta_h^\epsilon(\tilde x_\ell; x^*)$  defines $\eta^\epsilon$ over $[a_\ell,a_{\ell+1})$ by allowing $x^*$ to vary over the prescribed interval. 
Moreover, slightly modifying equation \ref{eq:gridfunsol} as a function of $x=\tilde x_\ell - x^*$ defines the $P_\ell(x)$ polynomials of $\eta^\epsilon$,
%POLY P_L
\begin{equation}\label{eq:poly}
	 P_\ell(x) = \left\{ \begin{array}{cl}
		\displaystyle
		s! (-1)^{N-1} \; 
		\frac{
			\displaystyle
			\sum_{\underset{m_1,...,m_{q-1}\neq \ell}{1\le m_1<\cdots<m_{q-1}\le N}}
			 ( h(m_1- \ell) + x) \cdots ( h(m_{q-1} - \ell) + x)
		}
		{
			\displaystyle
			h^{N}\prod_{\underset{m\neq \ell}{1\le m\le N}} ( \ell - m)
		}
		,& \text{for} \; q>1  \vspace{10pt} \\ 
		\displaystyle
		\frac{
			s! (-1)^{s}
		}
		{
			\displaystyle
			h^{N} \prod_{\underset{m\neq \ell}{1\le m\le N}} (\ell - m)
		}
		,& \text{for}\; q=1.
	\end{array}\right.
\end{equation}
Inspection of equation \ref{eq:poly} reveals that $P_\ell$ is a polynomial of degree $q-1$.

%By construction of our approximations, it follows that $\eta_H(x-x^*)$ satisfies the discrete $(q,s)$-moment conditions at $x^*$ in a manner that is independent of how the grid $\Omega^h$ is centered, or grid origin in other words.

%%%
\begin{theorem}\label{thm:DCmomcond}
	Let nonnegative integer $q$, positive integer $s$, and $x^*\in \mathbf R$ be given. 
	Suppose $\eta^\epsilon$ is constructed according to equations \ref{eq:eta} and \ref{eq:poly} for a given $h>0$.
	Then it follows that $\eta^\epsilon(x)$ is a $q$-order approximation of $D^s\delta(x)$. 
\end{theorem}

%%%
\begin{proof}
	In order to apply theorem \ref{thm:weakconv} we need to verify that $\eta^\epsilon$ indeed satisfies the (continuum) $(q,s)$-moment conditions at $0$ as well as the estimate given by equation \ref{eq:boundH}.
	We first evaluate the $\alpha$-moment of $\eta^\epsilon$ for $\alpha=0,...,q+s-1$;
	\begin{subequations}
	\begin{align*}
		M^\alpha(\eta^\epsilon,0) &= \int_{\mathbf R} \eta^\epsilon(x) \;x^\alpha\; dx \\
			&= \sum_{\ell=1}^N \int_{a_\ell}^{a_{\ell+1}} P_\ell(x) \;x^\alpha \; dx.
	\end{align*}
	\end{subequations}
	Applying the following change of variables, $x= a_\ell + \xi$ with $\xi\in[0,h)$, over each interval yields
	\begin{subequations}
	\begin{align*}
		M^\alpha(\eta^\epsilon,0) &= \sum_{\ell=1}^N \int_{0}^h P_\ell(a_\ell+\xi) (a_\ell+\xi)^\alpha \; d\xi \\
			&= \int_{0}^h \frac{1}{h} \left[ h \sum_{\ell=1}^N P_\ell(a_\ell+\xi) (a_\ell+\xi)^\alpha \right]\; d\xi.
	\end{align*}
	\end{subequations}
	Note that the term in the bracket coincides with the discrete $\alpha$-moment of $\eta^\epsilon$ with respect to a uniform grid $\mathcal G_h(a_1,1)$ (containing stencil points $\{a_\ell\}_{\ell=1}^{N+1}$) for a source located at $x^*=-\xi$.
	In other words,
	\[
		h \sum_{\ell=1}^N P_\ell(a_\ell+\xi) (a_\ell+\xi)^\alpha = M_h^\alpha(\eta_h^\epsilon,-\xi)
	\]
	where $\eta_h^\epsilon:\mathcal G_h(a_1,1)\to \mathbf R$ defined by $\eta_h^\epsilon(x) = \eta^\epsilon(x+\xi)$ satisfies the discrete $(q,s)$-moment conditions at $-\xi$ by construction.
	We can conclude
	\begin{align*}
		M^\alpha(\eta^\epsilon,0) &= \int_{0}^h  \frac{1}{h} \left[ s! (-1)^{s} \delta_{s\alpha}\right]  \; d\xi \\
		&= s! (-1)^{s} \delta_{s\alpha}.
	\end{align*}
	Lastly, since $\eta^\epsilon$ consist of piecewise polynomials of order $q-1$ divided by a factor of $h^N$, where $N=q+s$, and $\epsilon=O(h)$, we have that
\[
	\sup_{x\in B(0,\epsilon)} |\eta^\epsilon(x)| = O(\epsilon^{-s-1}).
\] 
Thus
\[
	\int_{B(0,\epsilon)} dx\; |\eta^\epsilon(x)| \; |x^s| \le \sup_{B(0,\epsilon)} |\eta^\epsilon(x)|  \; \int_{B(0,\epsilon)} dx\; |x^s| = O(1),
\]
as required for estimate \ref{eq:boundH}

\end{proof}

%%%%%
\subsubsection{Constructions in Higher Dimensions via Tensor Products}
%%%%%

We construct approximations in general $d$-dimension by taking tensor products of 1-D approximations, similar to work by \cite{TorEng:04} for $\mathbf s=\mathbf 0$.
Namely, given approximation order $q$ and multi-index $\mathbf s$, multivariate continuum approximation $\eta:\mathbf R^d\to \mathbf R$ is given by
\begin{equation}\label{eq:tensor_eta}
	\eta(\mathbf x) = \prod_{k=1}^{d} \eta^k(x_k) 
\end{equation}
where $\eta^k:\mathbf R\to \mathbf R$ is a continuum function over the $k$-th axis as given by equations \ref{eq:eta} and \ref{eq:poly}, satisfying the $(q,s_k)$-moment conditions for each $k=1,...,d$.
It follows that these tensor approximations are indeed approximations to multipoles in the higher spatial dimension.

\begin{theorem}\label{thm:prodmomcond}
	Let nonnegative integer $q$, multi-index $\mathbf s$, and $\mathbf x^*\in\mathbf R^d$ be given. 
Suppose $\eta:\mathbf R^d\to\mathbf R$ is a multi-variate grid function given by the tensor product of 1-D approximations $\eta^{k}:\mathbf R\to\mathbf R$.
If $\eta^{k}$ satisfy the discrete $(q,s_k)$-moment conditions at $x^*_k$ for each $k=1,...,d$, then it follows that $\eta$ satisfies the discrete $(q,\mathbf s)$-moment conditions at $\mathbf x^*$.
\end{theorem}

%%
\begin{proof}
Suppose for each $k=1,...,d$ that $\eta^{k}$ satisfies the $(q,s_k)$-moment conditions at $x^*_k$. 
Let $\alpha$ be some multi-index with $|\alpha|\le q+|\mathbf s|-1$. 
Note that,
\[
	M^\alpha(\eta,\mathbf x^*) = \prod_{k=1}^d M^{\alpha_k}(\eta^{k},x^*_k).
\]
Clearly, if $\alpha_k\le q+s_k-1$ for all $k=1,...,d$, then the result follows from the supposition. 
Same applies for $\alpha=\mathbf s$.
Suppose then, that there exists index $\ell$ such that $\alpha_\ell>q+s_\ell-1$, that is $a_\ell=q+s_\ell -1 +i$ for some $i\in\mathbf N$. Thus,
\begin{align*}
	& |\alpha| = \sum_{k\neq \ell} \alpha_k + \alpha_\ell = \sum_{k\neq \ell} \alpha_k + q+s_\ell-1+i \le q + |\mathbf s| -1\\
\Longrightarrow	&  \sum_{k\neq\ell} \alpha_k + i \le \sum_{k\neq \ell} s_k.
\end{align*}
which implies that $\alpha_k<s_k$ for at least one $k\neq\ell$;
for this particular $k$, it follows that 
\[
	M^{\alpha_k}(\eta^{k},x^*_k) = s_k!(-1)^{s_k} \delta_{s_k \alpha_k} = 0
\]
since it has been established that $\alpha_k\neq s_k$, i.e., the product over $k$ is zero if $\mathbf s\neq \alpha$.
\end{proof}

%%%%%%%%
\subsubsection{1-D and 2-D Examples}
%%%%%%%%

%MOD
%Numerical convergence rate tests in the following section will employ the singular source approximation discussed here, replacing multipole terms with grid functions to be used in finite difference schemes.
\hl{Numerical examples in this paper will employ the singular source approximation discussed here, replacing multipole terms with grid functions to be used in finite difference schemes.}
Equation \ref{eq:gridfunsol} gives an explicit formula for such grid functions, depending on the source location $\mathbf x^*$, multi-index $\mathbf s$, approximation order $q$, and of course the underlying finite difference grid.
Alternatively, equations \ref{eq:eta} and \ref{eq:poly} define the continuum form of the approximations, where grid functions are obtained by shifting and sampling over the grid.
The following figures plot 1-D and 2-D continuum approximations $\eta^\epsilon$,
in particular, we plot second- and fourth-order approximations of 
%MOD
$D^s\delta(\mathbf x)$ for $s=0,1,2$ in 1-D \hl{(figure 1)} and $\mathbf s=(0,0),(0,1),(0,2)$ in 2-D \hl{(figure 2)}, with $h=1$.
Two dimensional approximations are constructed via tensor product of 1-D approximations as discussed.
In particular, the $q=2$ approximation for the 1-D Dirac delta function $(s=0)$ is none other than the well known hat/triangular function of unit mass.

%%%%%%%%%
\subsection{Convergence of finite difference approximations to smooth solutions}
%%%%%%%%%

%We prove weak convergence of finite difference solutions with discrete multipole
%sources satisfying the discrete moment conditions discussed above. 
%Convergence theory for smooth solutions is in some sense standard. 
%However 
%The easily available results (those based on Lax's Equivalence Theorem) 
%pertain to $L^2$ convergence for smooth solutions and smooth source terms,
%which we begin with.
%whereas weak convergence of multipole solutions depends on uniform 
%($L^{\infty}$) error estimates for smooth solutions. 
%Therefore we begin with a full account of the convergence of smooth solutions.

With smooth Cauchy data and
right-hand sides, symmetric hyperbolic systems have unique smooth solutions depending
continuously in the sense of ${\cal D}$ on the data
\cite[]{CourHil:62,Lax:PDENotes}. Moreover, restriction of the
solution to constant-time hyperplanes is well-defined, and if the data has compact
support, then the restrictionts have compact support as well - this is
the finite propagation speed property.

Finite difference approximation to smooth solutions of hyperbolic PDE has been
extensively studied \cite[]{RichMor:67,Cohen:01,Leveque:07}. Rather
than quote results from the literature, however, we prove suitable
convergence results for staggered grid approximation to systems of the
form \ref{eq:hyper_fam} directly, by introducing a type of energy estimate
appropriate to such approximations. This technique of proof will be
central to our approach to weak convergence for singular solutions,
developed in the next subsection.

%Theoretical results presented here pertain to the family of symmetric hyperbolic
%systems of  first-order partial differential equations for vector-valued fields $(u,v)$,
%defined over a domain $\Omega\subset \mathbf R^d$ in $d=\{1,2,3\}$ dimension
%space and some interval in time, given by \ref{eq:hyper_fam} which we restate 
%with greater detail.

\subsubsection{Smooth and Distribution Solutions}

From here on, we will denote by $k_1, k_2 \in {\bf N}$ the dimensions of
the two subsystems in \ref{eq:hyper_fam}.  The references cited above
establish (result which imply) the following theorem.%something missing here?? MJB

\begin{theorem}
\label{thm:well-posed}
Suppose
\begin{itemize}  
\item $f \in C_0^{\infty}(\bR^{d+1},\bR^{k_1})$, $g \in
  C_0^{\infty}(\bR^{d+1},\bR^{k_2})$;
\item $u^0 \in C_0^{\infty}(\bR^d,\bR^{k_1}), v^0 \in C_0^{\infty}(\bR^d,\bR^{k_1})$;
\item $A \in C^{\infty}(\bR^d, \bR^{k_1 \times k_1})$, $B \in
  C^{\infty}(\bR^d, \bR^{k_2 \times k_2})$ are symmetric for all $\mathbf x\in\Omega$ 
  and uniformly bounded and positive definite , i.e., there exist constants 
  $0< A_*\le A^*$ and $0< B_*\le B^*$ for which
  \[
    A_*I \le A(\bx) \le A^*I,\,\,   B_*I \le B(\bx) \le B^*I, \quad \forall \mathbf
    x\in \bR^d;
  \]

\item $P$ is a constant coefficient, 
  first-order differential operator of the form
  \[
    Pu = \sum_{j=1}^d P_j \frac{\partial u}{\partial x_j}, \quad P_j\in \bR^{k_2\times k_1},
  \]
  for $u\in C^{\infty}(\bR^d,\bR^{k_1})$.
\end{itemize}
Then there exist unique $u \in C^{\infty}(\bR^{d+1},\bR^{k_1})$, $v \in
  C^{\infty}(\bR^{d+1},\bR^{k_2})$ satisfying the following system,
  identically in  $t \in \bR, \bx \in \bR^d$:
\begin{equation}\label{eq:cont_system}
\begin{split}
	A(\mathbf x)\frac{\partial}{\partial t}u(\mathbf x,t) + P^T v(\mathbf x,t) 
		&= f(\mathbf x,t), \\
	B(\mathbf x)\frac{\partial}{\partial t}v(\mathbf x,t) - Pu(\mathbf x,t) 
		&= g(\mathbf x,t), \\
        u(\bx,0) & = u^0(\bx),\\
        v(\bx,0) & = v^0(\bx), 
\end{split}
\end{equation}
Moreover, for each $t \in \bR$, the restrictions $u(\cdot,t), v(\cdot,t)$ have compact support.
\end{theorem}

Solution of the system \ref{eq:cont_system} defines a linear map from from
$u^0,v^0,f,g$ to $u,v$, continuous in the $C_0^{\infty}$ respectively
$C^{\infty} $ topologies. The restriction of the formal adjoint of
this map to smooth input will play an important role in the sequel.

It will be sufficient to restrict $u^0=v^0=0$, so that the restricted map
has inputs $f,g$. A function $f \in C^{\infty}(\bR^{d+1},\bR^k)$ is {\em causal} if there 
is $t_0 \in \bR$ so that $f(\bx,t)=0$ for $t<t_0, \bx \in
\bR^d$. Similarly, $f$ is {\em anticausal} if there is $t_0 \in \bR$
so that $f(\bx,t)=0$ for $t>t_0$.

%should \tilde f and \tilde g be anticausal ?? MJB
\begin{theorem}\label{thm:adjointstate}
In addition to the hypotheses of Theorem \ref{thm:well-posed},
suppose that (i) $u^0=v^0=0$, and (ii)
$\tilde{f} \in C_0^{\infty}(\bR^{d+1},\bR^{k_1})$, $\tilde{g} \in
C_0^{\infty}(\bR^{d+1},\bR^{k_2})$. Let $u,v$ be the unique causal
solution of the system \ref{eq:cont_system}, and $\tilde{u},\tilde{v}$
the unique anticausal solution of the system
\begin{equation}\label{eq:adj_system}
\begin{split}
	A(\mathbf x)\frac{\partial}{\partial t}\tilde{u}(\mathbf x,t) - P^T \tilde{v}(\mathbf x,t) 
		&= -\tilde{f}(\mathbf x,t), \\
	B(\mathbf x)\frac{\partial}{\partial t}\tilde{v}(\mathbf x,t) + P\tilde{u}(\mathbf x,t) 
		&= -\tilde{g}(\mathbf x,t).
\end{split}
\end{equation}
Then
\begin{equation}
\label{eq:contadj}
\int (\langle u,\tilde{f} \rangle + \langle v, \tilde{g} \rangle) \, dt
= \int (\langle\tilde{u}, f \rangle + \langle \tilde{v},g \rangle) \, dt.
\end{equation}
\end{theorem}

%The integrals appearing in the identity \ref{eq:contadj} are
%well-defined, as the integrands have uniformly bounded support in
%$\bx$ for each $t$, and vanish outside of a bounded interval in $t$
%due to the presence of both causal and anticausal factors in each term.

%First time t=0 and t=T appear in integrals, should we mention 
%perhaps that causality begins at t0=0 and anticausality at t0=T ?? MJB
\begin{proof}
\[
\int (\langle u,\tilde{f} \rangle + \langle v, \tilde{g} \rangle) \, dt
=
\int \left(\left\langle u,-A\frac{\partial \tilde{u}}{\partial t} -
    P^T \tilde{v}\right\rangle\right.
+
\left. \left\langle v,-B\frac{\partial \tilde{v}}{\partial t} + P
    \tilde{u}\right\rangle\right) \, dt
\]
\[
=
\int \left(\left\langle A\frac{\partial u}{\partial
      t},\tilde{u}\right\rangle
-\langle P u,\tilde{v}\rangle + \left\langle B\frac{\partial
    v}{\partial t},\tilde{v}\right\rangle + \langle
P^Tv,\tilde{u}\rangle
\right) \, dt
\]
\[
=
\int \left(\left\langle A\frac{\partial u}{\partial
      t} + P^Tv,\tilde{u}\right\rangle + \left\langle B\frac{\partial
    v}{\partial t}-Pu,\tilde{v}\right\rangle\right) \, dt
\]
\[
= \int (\langle\tilde{u}, f \rangle + \langle \tilde{v},g
\rangle) \, dt.
\]
\end{proof}

The definition of distribution solution is related to the definition
of the adjoint system \ref{eq:adj_system}. A pair $(u,v)$ of
distributions is a {\em weak} or {\em distribution} solution of
%what is mathcal E ?? MJB
%may be missing a mathcal ?? MJB
\ref{eq:cont_system} with right-hand sides $(f,g) \in \mathcal
D'(\bR^{d+1},\bR^{k_1}) \times \mathcal D'(\bR^{d+1},\bR^{k_2})$ if for any pair $(\phi,\psi) \in
C_0^{\infty}(\bR^{d+1},\bR^{k_1}) \times C_0^{\infty}(\bR^{d+1},\bR^{k_2})$,
\[
\int \left( \left\langle u, -A\frac{\partial \phi}{\partial 
    t} + P^T \psi\right\rangle + \left\langle v, -B\frac{\partial\psi}{\partial t} + P\phi\right\rangle \right) \, dt
\]
\begin{equation}
\label{eq:defdistsol}
= \langle f,\phi \rangle + \langle g,\psi \rangle
\end{equation}

The references cited above show that unique causal (or anticausal)
distribution solutions exist for distribution right hand sides of
compact support. Moreover, it is implicit in the definition that these
distribution solutions satisfy the relation \ref{eq:contadj}, with the
$\langle \cdot, \cdot \rangle$ interpreted as the pairing of
distribution and test function rather than as the $L^2$ inner product
(the former is in a sense a generalization of the latter, of course).

% WWS added 2018/01/03
In particular, unique causal distribution solutions of the system
\ref{eq:cont_system} exist for multipole right-hand sides $(f,g)$ of
the form \ref{eq:MPSappx}, provided that the multipole coefficients
$w_{\mathbf s}$ are distributions of compact support.
  %\item we assume solution $(u,v)$ and source terms $(f,g)$ are causal, i.e., for $t < 0$
%		\[
%			u(\mathbf x,t) = f(\mathbf x,t) = 0, \quad 
%			v(\mathbf x,t) = g(\mathbf x,t) = 0.
%		\]
%\end{itemize}
%See \cite{BlazekStolkSymes:13} for proof of existence of solutions, stability, and other mathematical properties for a larger class of partial (integro-)differential systems for wave modeling, applicable to system \ref{eq:cont_system}.

%As an example, consider the acoustic equations in first-order (pressure-velocity) form:
%\begin{equation}\label{eq:acoustics}
%\begin{split}
%	\frac{1}{\kappa(\mathbf x)} \frac{\partial}{\partial t} p(\mathbf x,t) + \nabla\cdot \mathbf v(\mathbf x,t) 
%		&= f(\mathbf x,t)\\
%	\rho(\mathbf x) \frac{\partial}{\partial t} \mathbf v(\mathbf x,t) + \nabla p(\mathbf x,t) 
%		&= \mathbf g(\mathbf x,t)
%\end{split}
%\end{equation}
%where $u=p$ is the scalar pressure field and $v=\mathbf v\in\mathbf R^d$ is the vector particle velocity field; we take $\mathcal H_1=H^1_0(\Omega)$ and $\mathcal H_2=H^1(\Omega)^d$. %MOD
%Coefficient operator $A(\mathbf x)=1/\kappa(\mathbf x)$ and $B(\mathbf x)=\rho(\mathbf x) I$, with $\kappa$ denoting bulk-modulus and $\rho$ density of the medium; here $I\in\mathbf R^{d\times d}$ is the identity matrix.
%Lastly, the differential operator $P$ coincides with the gradient and its adjoint with the negative of the divergence,
%\[
%	P = \mat{\frac{\partial}{\partial x_1}\\ \vdots \\ \frac{\partial}{\partial x_d}}, \quad 
%	P^T = - \mat{\frac{\partial}{\partial x_1},...,\frac{\partial}{\partial x_d}}.
%\]
%%ADDED
%Given homogenous boundary conditions for the pressure field, it can be shown that $P$ given as the gradient operator satisfies the skew-adjoint relation \ref{eq:skew}.

%%%%%%%%%%%
\subsubsection{Staggered grid discretization}
Staggered grid discretizations of hyperbolic systems can be quite
complex in structure, as examples presented by \cite{moczoetal:06}
show. Fortunately the essential properties of these discrete dynamical
systems can be captured in a fairly simple abstraction.

%We don't make use of <,>_h or || ||_h notation throughout ?? MJB
We presume a family $\{H^1_h, H^2_h: h \in \bR_+\}$ of inner product
space pairs, and surjections $S^i_h:
C_0^{\infty}(\bR^d,\bR^{k_i}) \rightarrow H^i_h$, $i=1,2$. We will
denote the inner product in $H^i_h$ by $\langle \cdot,\cdot\rangle$, 
with the index $i = 1,2$ and inferred from context. 
Moreover, distinguishing inner products between elements in
$H_h^i$ or continuum functions (e.g., $C^\infty$ functions) is discerned by context.
The corresponding induced norm is denoted $\|\cdot\|$.  
In the examples that motivate this work, $H^i_h$ consists of vector valued
functions of compact support on uniform rectangular grids in $\bR^d$ of mesh $h$, and $S^i_h$
samples the smooth input on grid points. Several
grids may figure in the definition of each of these spaces, and
with some components defined on each grid - see \cite{moczoetal:06}
for numerous examples.

The next ingredient is a uniformly bounded family $\{Q_h \in {\cal
  B}(H^1_n,H^2_h)\}$ of linear maps. This family defines an 
approximation to the differential operator $P$, in a sense to be made
precise below.

For future use, set
\[
Q^* = \sup_{h \in \bR_+} \|Q_h\|.
\]

Taking advantage of a bit of hindsight, we define an abstract staggered grid
dynamical system to be the recursion for a sequence of vector pairs $(u^n_h,v^{n-1/2}_h)
\in H^1_h \times H^2_h: n \in {\bf Z}$:
\begin{equation}\label{eq:sg}
\begin{split}
	A_h(u^{n+1}_h - u^n_h)  + P_h(r)^T v^{n+1/2}_h &= rh\,   f_h^{n+1/2} \\
	B_h(v^{n+1/2}_h-v^{n-1/2}_h) -  P_h(r) u^n_h &=  rh\, g_h^{n} 
\end{split}
\end{equation}
in which
\begin{itemize}

	\item $A_h, \, B_h$ are bounded self-adjoint positive-definite operators 
\[
	A_h: H_h^1 \rightarrow H_h^1,\quad B_h: H_h^2 \rightarrow H_h^2
\]
with upper and lower bounds uniform in $h$: there exists constants $0<A_*\le A^*$ and $0<B_*\le B^*$ such that
\begin{equation*}
\begin{split}
	A_*\|u_h\|^2 \le \langle A_h u_h,u_h \rangle \le A^*\|u_h\|^2, 
	\quad \forall u_h\in H_h^1, \\
	B_*\|v_h\|^2 \le \langle B_h v_h,v_h \rangle \le B^*\|v_h\|^2,
	\quad \forall v_h\in H_h^2; \\
\end{split}
\end{equation*}
\item $f_h^{n+1/2} \in H^1_h,\,\, g_h^{n} \in H^2_h$,
\item $P_h(r) = rQ_h, \,\ r > 0$.
\end{itemize}

%Consider again the acoustic case (related to problem \ref{eq:acoustics}).
%For simplicity we assume we are dealing with rectangular grids, in 2-D for this example, of the form
%\[
%	\mathcal G(\mathbf 0,\mathbf h) = \{ \mathbf x_{i,j} = (ih,jh) \; : \; i,j\in\mathbf Z \}.
%\]
%The simplest staggered-grid finite difference scheme, second-order in time and space, is given by
%\begin{equation}\label{eq:acoustics_sgfd}
%\begin{split}
%	\frac{1}{\kappa(ih,jh)} \frac{1}{\Delta t} \Big[  (p_h)^{n+1}_{i,j} - (p_h)^{n}_{i,j} \Big] + \hspace{7cm}\\
%	\frac{1}{h} \Big[ (v_{1,h})_{i+1/2,j}^{n+1/2} - (v_{1,h})_{i-1/2,j}^{n+1/2} + 
%				(v_{2,h})_{i,j+1/2}^{n+1/2} - (v_{2,h})_{i,j-1/2}^{n+1/2} \Big] 
%		&= (f_h)^{n+1/2}_{i,j}\\
%	\rho((i+1/2)h,jh) \frac{1}{\Delta t} \Big[ (v_{1,h})^{n+1/2}_{i+1/2,j} - (v_{1,h})^{n-1/2}_{i+1/2,j} \Big] +
%	\frac{1}{h} \Big[ (p_h)_{i+1,j}^{n} - (p_h)_{i,j}^{n} \Big] 
%		&= (g_{1,h})_{i+1/2,j}^{n}\\
%	\rho(ih,(j+1/2)h) \frac{1}{\Delta t} \Big[ (v_{2,h})^{n+1/2}_{i,j+1/2} - (v_{2,h})^{n-1/2}_{i,j+1/2} \Big] +
%	\frac{1}{h} \Big[ (p_h)_{i,j+1}^{n} - (p_h)_{i,j}^{n} \Big] 
%		&= (g_{2,h})_{i,j+1/2}^{n}
%\end{split}
%\end{equation}
%where finite difference solution $(p_h,v_{1,h},v_{2,h})$ approximates the continuum fields $(p,v_1,v_2)$,% for system \ref{eq:cont_system}, mainly,
%\begin{equation*}
%\begin{split}
%	(p_h)_{i,j}^{n} &\approx p(ih,jh,n\Delta t), \\
%	(v_{1,h})_{i+1/2,j}^{n+1/2} &\approx v_1((i+1/2)h,jh,(n+1/2)\Delta t),\\
%	(v_{2,h})_{i,j+1/2}^{n+1/2} &\approx v_2(ih,(j+1/2)h,(n+1/2)\Delta t).
%\end{split}
%\end{equation*}
%We emphasize that the velocity fields for finite difference scheme \ref{eq:acoustics_sgfd} are staggered with respect to spatial grids, more specifically each component is shifted by half a cell size in its respective axis.
%
%Relating system \ref{eq:acoustics_sgfd} with \ref{eq:sg}, we see
%\[
%	(u_h)_{i,j} = (p_h)_{i,j}, \quad 
%	(v_h)_{i,j} = \mat{(v_{1,h})_{i+1/2,j}\\ 
%				  (v_{2,h})_{i,j+1/2} }.
%\]
%The space $H_{1,h}$ corresponds to the set of square summable scalar-valued
%%MOD
%functions on rectangular grids that satisfy the homogenous boundary condition, and $H_{2,h}$ is the set of square summable $\mathbf R^2$-valued functions on half-cell shifted grids, both equipped with a discrete $L^2$ inner-products,  
%\begin{equation*}
%\begin{split}
%	\langle u_h,\tilde u_h\rangle = h^2 \sum_{i,j} (u_h)_{i,j} (\tilde u_h)_{i,j} &\quad 	\text{for } u_h,\tilde u_h \in H_{1,h}\\
%	\langle v_h, \tilde v_h \rangle = h^2 \sum_{i,j} (v_h)_{i,j} \cdot (\tilde v_h)_{i,j} &\quad
%	\text{for } v_h,\tilde v_h \in H_{2,h}.
%\end{split}
%\end{equation*}
%Inner-products and norms of the spaces $\mathcal H_1, \mathcal H_2, H_{1,h}, H_{2,h}$ are all denoted by $\langle\cdot,\cdot\rangle$ and $\|\cdot\|$ and interpreted given the context, unless otherwise specified.
%
%Sampling operators $S_{1,h}$ and $S_{2,h}$ coincide with evaluating continuum functions on grid or shifted grid points accordingly,
%\[
%	(S_{1,h} u)_{i,j} = u(ih,jh), \quad 
%	(S_{2,h} v)_{i,j} = \mat{v_{1}((i+1/2)h,jh)\\ 
%				  v_{2}(ih,(j+1/2)h) }.
%\]
%Discretized coefficient operators $A_h$ and $B_h$ are given by,
%\[
%	(A_h u_h)_{i,j} = \frac{1}{\kappa(ih,jh)} (u_h)_{ij}, \quad
%	(B_h v_h)_{i,j} = \mat{ \rho((i+1/2)h,jh) (v_{1,h})_{i+1/2,j} \\
%					 \rho(ih,(j+1/2)h) (v_{2,h})_{i,j+1/2} }.
%\]
%Lastly, $P_h$ and $P_h^T$ correspond to second-order approximations of the gradient and the negative of the divergence respectively;
%\begin{equation*}
%\begin{split}
%	(P_{h}(r) u_h)_{i,j} &= r \mat{ (u_h)_{i+1,j} - (u_h)_{i,j} \\
%					     (u_h)_{i,j+1} - (u_h)_{i,j} }, \\
%	(P_{h}(r)^T v_h)_{i,j} &= -r \Big[ (v_{1,h})_{i+1/2,j} - (v_{1,h})_{i-1/2,j} + (v_{2,h})_{i,j+1/2} - (v_{2,h})_{i,j-1/2} \Big]
%\end{split}
%\end{equation*}
%again, with $r =  \Delta t/h$.
%In general, the differential operator $P$ can be approximated by a family of central difference operators $P_h$ of even order $2p$ for $p\in\mathbf N$, resulting in the 2-$2p$ staggered-grid finite difference schemes.
%%ADDED
%Similar to the continuum case, it can be shown that the $P_h$ indeed satisfy the skew-adjoint relation \ref{eq:skew_h}.

%%%%%%%
\subsubsection{Energy Estimates}

%We aren't consistent between E or E_h ?? MJB
Inspired by the work of Gustafsson and colleagues on staggered grid
schemes for acoustics \cite[]{GusWahl:04a,GusWahl:04b,GusMoss:04},
define $E: H^1_h \times H^2_h \rightarrow \bR_+$ by
\begin{equation*}
\begin{split}
	E(u_h,v_h) &:= \frac{1}{2} \Big( \langle A_h u_h,u_h\rangle + 
				       \langle B_hv_h,v_h\rangle +
				       \langle P^T_h v_h,u_h\rangle \Big).
%	&= \frac{1}{2} \mat{ u_h^n \\ v_h^{n-1/2}}^T
%			     \mat{ A_h & \frac{1}{2}P^T_h\\
%			     	     \frac{1}{2}P_h & B_h }
%			     \mat{ u_h^n\\ v_h^{n-1/2} }.
\end{split}
\end{equation*}

\begin{theorem}
Denote by $SPD_k$ the set of symmetric positive definite $k\times k$
matrices. Then there exist $R, C_*, C^*: SPD_{k_1} \times SPD_{k_2} \times
\bR^{k_2\times k_1} \rightarrow \bR_+$ so that 
\begin{itemize}
\item if
\begin{equation}
\label{eq:cfl}
r < R(A_h,B_h,Q_h)
\end{equation}
then
\begin{equation}
\label{eqn:pd}
	C_*(A_h,B_h,Q_h) \Big( \|u_h\|^2 + \|v_h\|^2 \Big) \le E(u_h,v_h) \le C^*(A_h,B_h,Q_h)\Big( \|u_h\|^2 +\|v_h\|^2 \Big).
\end{equation}
\item $R(A_h,B_h,Q_h) \ge \sqrt{A_*B_*}/Q^*$.
\end{itemize}
\end{theorem}

\begin{proof}
The upper bound in \ref{eqn:pd} is clear from $h$-uniform bounds $A^*$on $A_h$, $B_h$, 
$Q_h$. 
The lower bound follows from the inequality
\[
	E(u_h,v_h) \ge A_*\|u_h\|^2 + B_*\|v_h\|^2 - rQ^*\|u_h\|\|v_h\|
\]
and the RHS is positive definite in $\|u_h\|,\|v_h\|$ when $rQ^* <\sqrt{A_*B_*}$. 
\end{proof}

That is,  the quadratic form $E$ defines a norm equivalent to the
product norm $\|\cdot\|$ on $H^1_h \times H^2_h$, provided that $r$
is sufficiently small relative to spectral bounds for the coefficient
matrices. This condition is in fact the Courant-Friedrichs-Levy
condition for staggered grid schemes.

\begin{theorem}
Suppose that $\{(u_h^n,v^{n-1/2}_h): n \in {\bf Z}\}$ is a solution of
system \ref{eq:sg} with $(f^{n+1/2}_h, g^n_h)\equiv 0$ for all n. 
Then $E^n=E(u_h^n,v^{n-1/2}_h)$ is independent of time index $n$.
\end{theorem}

\begin{proof}
\begin{equation*}
\begin{split}
	E^{n+1} - E^n =\langle A_h( u^{n+1}_h-u^n_h ), (u^{n+1}_h+u^n_h)\rangle 
		&+ \langle B_h( v^{n+1/2}_h-v^{n-1/2}_h ), (v^{n+1/2}_h+v^{n-1/2}_h)\rangle \\
		&+ \langle P_h^T v^{n+1/2}_h, u^{n+1}_h\rangle - \langle P_h^T v^{n-1/2}_h, u^n_h\rangle
\end{split}
\end{equation*} 
\begin{equation*}
\begin{split}
	= \langle -P_h^Tv^{n+1/2}_h,(u^{n+1}_h+u^n_h) \rangle 
		&+ \langle P_hu^n_h, (v^{n+1/2}_h+v^{n-1/2}_h) \rangle\\
		&+\langle P_h^Tv^{n+1/2}_h,u^{n+1}_h\rangle - \langle P^T_h v^{n-1/2}_h,u^n_h\rangle
	=0.
\end{split}
\end{equation*}
\end{proof}

%%%%%%%%

%Inhomogeneous system:
%\begin{equation}\label{eqn:in}
%\begin{split}
%A(u^{n+1}_h - u^n_h) & =  -D^T v^{n+1/2}_h + hf^{n+1/2}_h \\
%B(v^{n+1/2}_h-v^{n-1/2}_h) & =  D u^n_h + h g^n_h 
%\end{split}
%\end{equation}

\begin{theorem}
\label{thm:inhom}
Suppose that $\{(u_h^n,v^{n-1/2}_h): n \in {\bf Z}\}$ is a solution of
system \ref{eq:sg}. Assume that $r$ sufficiently small that $E$ is
positive definite, recall the lower bound $C_*$ in the 
equivalence relation \ref{eqn:pd}. Set $a = 2/C_*$. For $n \ge 0$ and $h < C_*/r$,
\begin{equation}
\label{eq:inhom}
	E^n \le e^{anrh} \left(E^0 + 
	C_* \sup_{0 \le m \le n-1}(\|f^{m+1/2}_h\|^2 + \|g^m_h\|^2)\right)
\end{equation}
\end{theorem}

\begin{proof}
The same arithmetic as used in the homogeneous case leads to
\begin{equation*}
\begin{split}
	E^{n+1}-E^n 
	&= rh\langle f^{n+1/2}_h, u^{n+1}_h+u^n_h\rangle + rh\langle g^n_h,
v^{n+1/2}_h+v^{n-1/2}_h\rangle\\
	&\le \frac{rh}{2\alpha^2} (\|f^{n+1/2}_h\|^2 + \|g^n_h\|^2) +
\frac{\alpha^2rh}{C_*}(E^{n+1}+E^n)
\end{split}
\end{equation*}
for any $\alpha \in (0,1)$. Choose $\alpha^2=1/2$. From the assumption $rh \le C_* <2C_*$,
$\alpha^2 rh < C_*$. Set 
\begin{equation}\label{eq:K}
	\lambda = \left(1+\frac{rh}{2C_*}\right) \left(1-\frac{rh}{2C_*}\right)^{-1}
	\quad
	\text{and}
	\quad
	K = \frac{1}{ \left(1+\frac{rh}{2C_*}\right) }.
\end{equation}
Then $\lambda > 1$ and 
\begin{equation*}
\begin{split}
	\lambda^{-1}E^{n} - E^{n-1} &\le Krh (\|f^{n-1/2}_h\|^2 + \|g^{n-1}_h\|^2)  \\
	\Longrightarrow
	\lambda^{-n}E^{n} - \lambda^{-(n-1)}E^{n-1} &\le \lambda^{-(n-1)}Krh (\|f^{n-1/2}_h\|^2 + \|g^{n-1}_h\|^2) 
\end{split}
\end{equation*}
whence
\begin{equation}
\label{eq:est1}
\lambda^{-n}E^n \le E^0 +
Krh\sum_{m=0}^{n-1}\lambda^{-m}(\|f^{m+1/2}_h\|^2 + \|g^m_h\|^2). 
\end{equation}

Set $M_n=\sup_{0 \le m \le n-1}(\|f^{m+1/2}_h\|^2 +
\|g^m_h\|^2)$. Then the inequality \ref{eq:est1} implies

\[
\lambda^{-n}E^n \le E^0 +
Krh \frac{1-\lambda^{-n}}{1-\lambda^{-1}} M_n
\]
\[
\le E^0 + Krh\frac{1}{1-\lambda^{-1}} M_n
\]
\[
\le E^0 + Krh \frac{1 + \frac{rh}{2C_*}}{2\left(
    \frac{rh}{2C_*}\right)}M_n
\]
From the definition \ref{eq:K} of $K$, this is equivalent to 
\begin{equation}
\label{eq:est2}
\lambda^{-n}E^n \le E^0 + C_* M_n.
\end{equation}
The simple inequality 
\[
\frac{1+x}{1-x} \le e^{4x},\, 0\le x \le \frac{1}{2}
\]
together with the hypothesis that $rh \le C_*$ so that $rh/2C_* < 1/2$, and the definition
\ref{eq:K} of $\lambda$ implies that 
\[
\lambda \le e^{2\frac{rh}{C_*}}
\]
Thus the estimate \ref{eq:est2} implies the conclusion \ref{eq:inhom}.
\end{proof}

In view of the relation between the continuum adjoint and the system
\ref{eq:adj_system}, it is not surprising that a similar relation
exists between the adjoint of the map defined by the system
\ref{eq:sg} and a related discrete adjoint state system.

A sequence $\{f^n\}$ (or $\{f^{n+1/2}\}$) vector-valued functions is {\em causal} if there is
$n_{\rm min} \in {\bf Z}$ so that $f^n=0$ (or $f^{n+1/2}=0$) for
$n<n_{\rm min}$. Likewise, such a function is {\em anticausal} if
there is $n_{\rm max}$ for which  $f^n=0$ (or $f^{n+1/2}=0$) for
$n>n_{\rm max}$.

\begin{theorem}
Suppose that $\{(f_h^{n+1/2},g_h^{n})\}$ is a causal sequence,
$\{(\tilde{f}_h^{n},\tilde{g}_h^{n+1/2})\}$ an anticausal
sequence in $H^1_h \times H^2_h$.
Let $\{ (u_h^n,v_h^{n+1/2})\}$ be the unique causal solution of 
the system \ref{eq:sg}, and
 $\{(\tilde u_h^{n+1/2},\tilde v_h^{n})\}$ the unique anticausal
solution of
\begin{equation}\label{eqn:inadj}
\begin{split}
	A_h( \tilde u_h^{n-1/2}-\tilde u_h^{n+1/2} ) - P_h^T \tilde v^n & =  rh\, \tilde f_h^{n} \\
	B_h( \tilde v^{n} - \tilde v^{n+1} ) + P_h \tilde u^{n+1/2} &=  rh \, \tilde g_h^{n+1/2}. 
\end{split}
\end{equation}
Then 
\begin{equation}
\label{eqn:weak}
	 rh\sum_{n} \Big( \langle \tilde u_h^{n+1/2}, f_h^{n+1/2} \rangle +
	 			 	  \langle \tilde v_h^n, g_h^n \rangle \Big) 
	= rh\sum_{n} \Big( \langle \tilde f_h^n, u_h^n \rangle + 
					     \langle \tilde g_h^{n+1/2}, v_h^{n+1/2}\rangle \Big). 
\end{equation}
\end{theorem}

Note that both sums in the identity \ref{eqn:weak} are finite, as all
terms have both causal and anticausal factors.

\begin{proof}
\[
\sum_{n} \Big( \langle \tilde u_h^{n+1/2}, f^{n+1/2}\rangle +
					   \langle \tilde v_h^n, g_h^n \rangle \Big)
\]
\begin{equation*}
\begin{split}
	= \sum_n \Big( \langle \tilde u_h^{n+1/2}, 
					A_h( u_h^{n+1} - u_h^n) + P_h^T v_h^{n+1/2} \rangle 
			  +  \langle \tilde v_h^n, 
					B_h ( v_h^{n+1/2}-v_h^{n-1/2} ) - P_h u_h^n \rangle \Big)
\end{split}
\end{equation*}
\begin{equation*}
\begin{split}
	= \sum_n \Big( \langle A_h ( \tilde u_h^{n-1/2}-\tilde u_h^{n+1/2} ), u_h^n \rangle 
			       -\langle P_h \tilde u_h^{n+1/2}, v_h^{n+1/2}\rangle 
	 		       +\langle B_h ( \tilde v_h^{n} - \tilde v_h^{n+1} ), v_h^{n+1/2} \rangle
			       +\langle P_h^T \tilde v_h^n, u_h^n\rangle \Big)
\end{split}
\end{equation*}
\begin{equation}\label{eqn:sbp}
\begin{split}
	= \sum_n \Big( \langle A_h (\tilde u_h^{n-1/2} - \tilde u_h^{n+1/2}) - P_h^T \tilde v_h^n, 
				u_h^n\rangle 
	                      +	\langle B_h (\tilde v_h^{n} - \tilde v_h^{n+1}) + P_h \tilde u_h^{n+1/2},
	                      	v_h^{n+1/2}\rangle
\end{split}
\end{equation}
In view of the system \ref{eqn:inadj}, the last line is equal to the
right hand side of \ref{eqn:weak}.

\end{proof}

%%%%%%%%%
\subsubsection{Norm Convergence to Smooth Solutions}

We will suppose from now on that $Q_h$ approximates $P$ to order $p\in {\bf N}$, in the sense that
for all $u_i \in C_0^{\infty}(\bR^d,\bR^{k_i}),\, i=1,2$,
\begin{equation}
\label{eq:fdord}
\begin{split}
\left\| S^2_hPu_1 - \frac{1}{h}Q_h S^1_h u_1\right\| &= O(|h|^p),\\
\left\| S^1_hP^Tu_2 - \frac{1}{h}Q_h^T S^2_h u_2\right\| &= O(|h|^p).
\end{split}
\end{equation}
For the grid-based examples described earlier, this
condition defines precisely the usual truncation-based order of
accuracy. In the second equation, the transpose $P^T$ is the usual
formal adjoint, whereas $Q_h^T$ is the adjoint with respect to the
assumed inner product structure on $H^i_h, \,i=1,2$.

Since we have assumed nothing {\em a priori} about the relation between distribution
pairing in $\bR^d$ and the inner products on $H^i_h, \,i=1,2$, 
the second estimate \ref{eq:fdord} does not follow from the first, and must be
asserted explicitly. In many examples, on the other hand, such as the
acoustic staggered grid approximation to be described later, the 
transpose automatically has the same order of accuracy.

Note that the implicit constants hiding behind the ``big O'' symbols
depend on the functions $u_1, u_2$, .

\begin{theorem}
\label{thm:smconvg}
Suppose that $Q_h$ approximates $P$ to order $p$, $T>0$, and that $\Delta t:
\bR_+ \rightarrow \bR_+$ satisfies $\Delta t(h) <
hR(A,B,Q)$ as in condition \ref{eq:cfl}. Suppose also that $u \in C^{\infty}(\bR^{d+1},\bR^{k_1})$, $v \in
  C^{\infty}(\bR^{d+1},\bR^{k_2})$ solve the hyperbolic system
  \ref{eq:cont_system} with coefficients $A,B$, right hand sides
  $f,g$, and Cauchy data $u^0,v^0$ satisfying the conditions of
  Theorem \ref{thm:well-posed}. Define
  $A_h,B_h,f^{n+1/2}_h,g^{n}_h,$ and $u^0_h,v^{-1/2}_h$ by
\begin{itemize}
\item $A_h S^1_hw_1=S^1_h Aw_1, \,\,B_hS^2_hw_2 = S^2_hBw_2$ for any $w_i
  \in C_0^{\infty}(\bR^d,\bR^{k_i})$; 
\item $f^{n+1/2}_h = S^1_hf(\cdot,(n+1/2)\Delta t(h)),\,\, g^n_h=S^2_h
  g(\cdot,n\Delta t(h))$
\item $u^0_h=S^1_hu(\cdot,0),\,\, v^{-1/2}_h=S^2_hv(\cdot,-1/2 \Delta
  t(h))$
\end{itemize}
Let $(u^n_h,v^{n-1/2}_h)$ denote the solution of the staggered grid
scheme \ref{eq:sg} with data $A_h,B_h,f^{n+1/2}_h,g^{n}_h,$ and
$u^0_h,v^{-1/2}_h$, $N(h)=T/\Delta t(h)$, and $r(h) = \Delta
t(h)/h$. Then for $h>0$ and $0 \le n \le N(h)$, 
\begin{equation}
\label{eq:errest}
\begin{split}
\|S^1_hu(\cdot,n \Delta t(h)) - u^n_h\| &= O(\Delta t(h)^2+h^p)\\
\|S^2_hv(\cdot,(n-1/2)\Delta t(h)) - v^{n-1/2}_h\| & = O(\Delta
t(h)^2 +h^p).
\end{split}
\end{equation}
\end{theorem}

\begin{proof}
Recall that $\Delta t = rh$ and that $r=r(h)$, hence $\Delta t=\Delta
t(h)$, is a (possibly linear) function of $h$, satisfying the
stability condition \ref{eq:cfl} for all $h>0$. For the remainder of
the proof, we suppress the dependence on $h$ from the notation.
Set
\begin{equation}\label{eqn:tr}
\begin{split}
  \delta u^n_h &  = S^1_hu(\cdot,n \Delta t) - u^n_h\\
  \delta v^{n-1/2}_h & = S^2_hv(\cdot,(n-1/2)\Delta t) - v^{n-1/2}_h\\
  \delta f^{n+1/2}_h & = \frac{1}{rh} \Big( A_h( S^1_h u(\cdot,(n+1)\Delta t) - S^1_h u(\cdot,n\Delta t)) + 
  P_h(r(h))^T S^2_h v(\cdot,(n+1/2)\Delta t) \Big) + f_h^{n+1/2}\\
  \delta g^{n}_h  &= \frac{1}{rh} \Big( B_h (S^2_h v(\cdot,(n+1/2)\Delta t) - S^2_h v(\cdot,(n-1/2)\Delta t) ) - 
  P_h(r(h)) S^1_h u(\cdot,n\Delta t) \Big) + g_h^{n}.
\end{split}
\end{equation}
Then the truncation error vector $(\delta u^n_h, \delta v^{n-1/2}_h)$ solves the staggered grid
system
\begin{equation}\label{eq:truncsg}
\begin{split}
  A_h(\delta u^{n+1}_h - \delta u^n_h)  + P_h(r)^T \delta
  v^{n+1/2}_h &= rh\,   \delta f_h^{n+1/2} \\
  B_h(\delta v^{n+1/2}_h-\delta v^{n-1/2}_h) -  P_h(r) \delta
  u^n_h &=  rh\, \delta g_h^{n} 
\end{split}
\end{equation}
which is precisely of the form \ref{eq:sg}. Note that $\delta u^0_h = 0,
\delta v^{-1/2}_h=0$. Therefore from Theorem
\ref{thm:inhom},
\begin{equation}
\label{eq:truncest1}
E(\delta u^n_h, \delta v^{n-1/2}_h) \le e^{anr(h)h}C_*\sup_{0 \le m
  \le n-1}(\|\delta f^{m+1/2}_h\|^2 + \|\delta g^m_h\|^2)
\end{equation}

From the definitions of $A_h$, $f^{n+1/2}_h$, 
\[
rh \delta f_h^{n+1/2} =  A_h(\delta u^{n+1}_h - \delta u^n_h)  + P_h(r)^T \delta
  v^{n+1/2}_h 
\]
\[
= A_h(S^1_h(u(\cdot,(n+1)\Delta t)-u(\cdot,n\Delta t)) - A_h(u^{n+1}_h-u^n_h) +
P_h(r(h))^T (S^2_h(v(\cdot,(n+1/2)\Delta t)) - v^{n+1/2}_h)
\]
\[
= S^1_h(A(u(\cdot,(n+1)\Delta t)-u(\cdot,n\Delta t))-P_h(r(h))^T
S^2_h(v(\cdot,(n+1/2)\Delta t))
-rh f^{n+1/2}_h
\]
where we have used the first of the two equations in \ref{eq:sg} to
eliminate the discrete fields $u^n_h,v^{n-1/2}_h$ in favor of $f^{n+1/2}_h$.
Because of the smoothness of $u$, 
\begin{equation}
\label{eq:truncdt}
u(\cdot,(n+1)\Delta t)-u(\cdot,n\Delta t) = 2\Delta t \frac{\partial u}{\partial 
  t}(\cdot,(n+1/2)\Delta t) +
O(\Delta t^3).
\end{equation}
From the order assumption \ref{eq:fdord} on $P_h = r Q_h$, 
\begin{equation}
\label{eq:truncdx}
P_h(r)^T S^2_h(v(\cdot,(n+1/2)\Delta t) = \Delta t S^1_h (P^T v)(\cdot,(n+1/2)\Delta t) + O(\Delta
t h^p)
\end{equation}
We conclude that 
\[
\Delta t \delta f_h^{n+1/2} = \Delta t  S^1_h\left(A \frac{\partial u}{\partial 
  t}(\cdot,(n+1/2)\Delta t) + (P^T v)(\cdot,(n+1/2)\Delta t) -  f(\cdot,(n+1/2)\Delta
t)\right)
\]
\[
 + O(\Delta t^3) + O(\Delta t h^p)
\]
\[
= O(\Delta t^3) + O(\Delta t h^p)
\]
since $(u,v)$ are smooth solutions of \ref{eq:cont_system}.
Similarly,
\[
\Delta t \delta g_h^n =  O(\Delta t^3) + O(\Delta t  h^p).
\]
The constants hiding inside the big ``O''s are uniform over compact
sets. Since $0 \le n\Delta t \le T$ for $0 \le n \le N(h)$, and the
data of the problem \ref{eq:cont_system} has compact support, the
domain in which the errors are evaluated has compact support, so these
inequalities imply that
\[
\|\delta f^{n+1/2}_h\|^2 + \|\delta g^n_h\|^2 = (O(\Delta t^2) + O(
h^p))^2
\]
uniformly for $0 \le n \le N(h)$. Combined with the inhomogeneous
energy estimate \ref{eq:inhom}, this observation implies the
conclusion of the theorem.
\end{proof}

%MOD
%If $(u,v)$ is a smooth solution of the acoustic (pressure-velocity) system 
%\ref{eq:acoustics} with smooth right-hand sides $(f,g)$,
%and $(u_h,v_h)$ is the staggered-grid finite difference approximation of 
%order $2$-$2p$, then, from standard truncation error calculations, for 
%each time slice of compact support we have,
%\hl{If $(u,v)$ is a smooth solution of the continuum problem} \ref{eq:cont_system}
%\hl{with smooth right-hand sides $(f,g)$, and $(u_h,v_h)$ is the finite difference 
%solution where $P_h$ is of order $2p$,} then, from standard truncation error 
%calculations, for each time slice of compact support we have,
%\begin{equation*}
%\begin{split}
%	\delta f_h^{n+1/2} &= O( \Delta t^2+h^{2p})\\
%	\delta g_h^{n} &= O( \Delta t^2+h^{2p})
%\end{split} 
%\end{equation*}
%point-wise and uniformly in $i,j$. 
%Since the support is uniformly bounded if $0\le n\Delta t \le T$, 
%the $O$ statements apply to the $L^2$ norms as well.
%Thus, by equivalency of the energy form with the $L^2$-norm, and the \hl{theorem} above, 
%we can conclude
%\[
%	\|u_h-S_{1,h}u\| + \|v_h-S_{2,h}v\| = O(\Delta t^2+h^{2p}),
%\]
%that is, $2$-$2p$ order convergence in the $L^2$ sense.

%Let $u$ denote the solution of \ref{eqn:sg} with
%$(u^0,v^{-1/2})=S(h)(U^0,V^{-1/2})$. Then $(u,v)-S(h)(U,V)$ solves
%\ref{eqn:in} with $f,g$ as just given.  Therefore the error estimate
%\ref{eqn:ee} implies that 
%\[
%\|(u,v)-S(h)(U,V)\| = O(h^{2p} + \Delta t^2),
%\]
%that is, $(2,2p)$-th order convergence in the $L^2$ sense.

%As mentioned at the beginning, to establish weak convergence of
%multipole solutions requires $L^{\infty}$, rather than $L^2$,
%convergence of smooth solutions. We obtain the stronger convergence
%theory by imitating the proof of the Sobolev Embedding Theorem.
%
%Sketch:
%
%apply difference operators $D_j$ to difference equations, commute with
%coefficients, note that same error estimates inherited by $D_j(u,v)$
%(with respect to $\partial_j (U,V)$) assuming that data is smooth enough
%
%iterate, addemup to show that discrete Laplacian of $(u,v)$ converges
%at same rate, that is, discrete Laplacian of $(u,v)-S(h)(U,V)$ is
%$O(h^{2p} + \Delta t^2)$ also (in $L^2$ sense)
%
%use discrete FT, Plancherel identity, and C-S inequality to bound
%$(u,v)-S(h)(U,V)$ pointwise. This requires only energy estimate on
%Laplacian for dimensions 1 and 2; for dimension 3, requires square of
%Laplacian for dimension 3.

%%%%%%%
\subsection{Weak Convergence to Singular Solutions}

In order to combine the moment-based analysis of gridded multipole
approximations with the convergence theory developed in the last
section, it is necessary to explicitly identify the Hilbert space
families $H^1_h,H^2_h$ with vector spaces of grid functions. In order
to accommodate the structure of commonly used staggered grid schemes
\cite[]{moczoetal:06}, this definition must allow for participation of
multiple grid types in each of these spaces.

Recall the definition of a regular grid family $\mathcal G_h(\mathbf
x_0,\mathbf l)$ with mesh $h \in \bR_+$, offset $\bx_0 \in
\bR^d$ and cell given by ${\bf l} \in \bR_+^d$, namely $\{\bx \in
\bR^d: 0 \le x_i\le l_k\}$:
\[
	\mathcal G_h(\mathbf x_0,\mathbf l) = \{\bx_{\bf n} =
        h((x_0)_1+n_1l_1,...,(x_0)_d+n_dl_d), {\bf n} \in {\bf Z}^d\}.
\]
We define the {\em grid Hilbert space} $\mathcal H_h(\bx_0, {\bf l})$ to be the real-valued
functions $f_h$ on $\mathcal G_h(\mathbf x_0,\mathbf l)$ for which
\[
\|f_h\|^2 = h^d \left(\Pi_{i=1}^dl_i\right) \sum_{\bx \in \mathcal G_h(\mathbf x_0,\mathbf l)}\,
|f_h(\bx)|^2
\]
is finite. That is, $\| \cdot \|$ is the standard $l_2$ norm, scaled
by the scaled cell volume.

The sampling operator defined by a grid $\mathcal G_h(\mathbf
x_0,\mathbf l)$ is the map
\[
\mathcal S_h(\bx_0,{\bf l}): C^{\infty}_0(\bR^d) \rightarrow \mathcal
H_h(\bx_0, {\bf l})
\]
defined by evaluating $f_h$ at gridpoints:
\[
(\mathcal S_h(\bx_0,{\bf l})f_h)(\bx) = f_h(\bx).
\]

Define the spaces $H^1_h,H^2_h$ to be direct sums of grid Hilbert
spaces with common mesh $h$. That is, given offsets $\bx_0^{j,i}$ and
cell side vectors ${\bf l}^{j,i}, \,1 \le j \le k_i, i=1,2$, set
\begin{equation}
\label{eq:def-hilb}
H^i_h = \oplus_{j=1}^{k_i}\mathcal H_h(\bx_0^{j,i},{\bf l}^{j,i})
\end{equation}
with the usual product Hilbert structure. The sampling operators
\[
S^i_h: C^{\infty}_0(\bR^d,\bR^{k_i}) \rightarrow H^i_h
\]
are diagonal:
\begin{equation}
\label{eq:def-samp}
(S^i_h f)_j = \mathcal S_h(\bx_0^{j,i},{\bf l}^{j,i})(f_j),
\,j=1,...,k_i
\end{equation}

With these preliminaries, we can state the central result of this paper:

\begin{theorem}\label{thm:conv}
Choose $q, N^*\ge 0$, $p>0$. Assume that $Q_h$ approximates $P$ to
order $p$. Let $(u,v)$ be the weak causal solution to continuum
problem \ref{eq:cont_system} with singular source terms $(f,g)$
given by order $N^*$ multipoles,
\begin{equation*}
\begin{split}
	f(\mathbf x,t) &=  \sum_{|{\bf s}_1| \le N^*}w_{{\bf s}_1}(t) D^{{\bf s}_1}\delta(\mathbf x-\mathbf x^*),  
        \; w_{{\bf s}_1} \in C_0^{\infty}(\bR,\bR^{k_1});\\
	g(\mathbf x,t) &= \sum_{|{\bf s}_2| \le N^*} z_{{\bf s}_2}(t) D^{{\bf s}_2}\delta(\mathbf x-\mathbf x^*),  
	\;z_{{\bf s}_2} \in C_0^{\infty}(\bR,\bR^{k_2}). 
\end{split}
\end{equation*}
Let $A_h,B_h$ be defined as in the statement of Theorem
\ref{thm:smconvg}. Suppose that $(f^{n+1/2}_h,g^n_h)$ to be discrete
multipoles
\begin{equation}
\label{eq:mprhs}
\begin{split}
	f^{n+1/2}_h &=  \sum_{|{\bf s}_1| \le N^*}w_{{\bf
            s}_1}((n+1/2)\Delta t) \eta_h^{\bf s_1},  \\
	g^n_h &= \sum_{|{\bf s}_2| \le N^*} z_{{\bf
            s}_2}(n\Delta t) \eta_h^{\bf s_2}
\end{split}
\end{equation}
in which $\eta_h^{\bf s_i} \in H^i_h, i=1,2$ satisfies the discrete $(q,{\bf s_{\bf i}})$-moment condition
\ref{eq:discmomcond} {\em in each of its components}. Let
$(u^n_h,v^{n-1/2}_h)$ be the causal solution of \ref{eq:sg}
with right-hand sides given by \ref{eq:mprhs}.

Suppose that $(\tilde f,\tilde g) \in
C^{\infty}_0(\bR^{d+1},\bR^{k_1}) \times
C^{\infty}_0(\bR^{d+1},\bR^{k_2})$, and set
\begin{equation}
\label{eq:defadjrhs}
\begin{split}
\tilde{f}^n_h &= S^1_h\tilde{f}(\cdot,n\Delta t)\\
\tilde{g}^{n+1/2}_h &= S^2_h\tilde{g}(\cdot,(n+1/2)\Delta t).
\end{split}
\end{equation}

Then
\begin{equation}\label{eq:weak_conv}
\begin{split}
	\mathcal E &:= \left| \int_0^T  \Big\{ \langle u,\tilde f\rangle + \langle v,\tilde g\rangle \Big\} dt -
		\Delta t \sum_{n=0}^N \Big\{ \langle u_h^{n}, \tilde f_h^{n} \rangle +
						 	   \langle v_h^{n+1/2}, \tilde g_h^{n+1/2}  \rangle \Big\} \right|\\
	&= O(\Delta t^2 + h^q + (\Delta t^2+h^{p})h^{-N^*-d/2}),
\end{split}
\end{equation}
\end{theorem}

\begin{proof}
Denote by $(\tilde u, \tilde v)$ the unique anticausal solution of the adjoint system \ref{eq:adj_system}
with right-hand sides $(\tilde f,\tilde g)$. Likewise, denote by $(\tilde{u}^{n-1/2}_h,\tilde{v}^n_h)$ the unique
anticausal solution of the discrete adjoint system \ref{eqn:inadj}
with right-hand sides $(\tilde{f}^{n}_h,\tilde{g}^{n+1/2}_h) $ as in
equation \ref{eq:defadjrhs}.

Using \ref{eqn:weak}, and its continuum version \ref{eq:adj_system}, error $\mathcal E$ is rewritten in terms of inner products with the singular source terms, both in the continuum and discrete sense, i.e.,
\[
	\mathcal = \left| \int_0^T  \Big\{ \langle f,\tilde u \rangle + \langle g,\tilde v\rangle \Big\} dt -
		\Delta t \sum_{n=0}^N \Big\{ \langle f_h^{n+1/2}, \tilde u_h^{n+1/2} \rangle +
						 	   \langle g_h^{n}, \tilde v_h^{n}  \rangle \Big\} \right|.
\]
%MOD Note $(\tilde p,\tilde v)$
Note \hl{$(\tilde u,\tilde v)$} are solutions to problem \ref{eq:cont_system} with smooth source terms 
$(-\tilde f,-\tilde g)$, and $(\tilde u_h,\tilde v_h)$ is the respective staggered grid finite difference solution with sources $(-\tilde f_h,-\tilde g_h)$.
Furthermore,
\begin{equation*}
\begin{split}
\mathcal E &= \left| \int_0^T  \Big\{ \langle f,\tilde u \rangle + \langle g,\tilde v\rangle \Big\} dt -
		\Delta t \sum_{n=0}^N \Big\{ \langle f_h^{n+1/2}, \tilde u_h^{n+1/2} \pm S_{h}^1\tilde u(\cdot,(n+1/2)\Delta t) \rangle +
						 	   \langle g_h^{n}, \tilde v_h^{n} \pm S_{h}^2 \tilde v(\cdot,n\Delta t) \rangle \Big\} \right|\\
	& \le \mathcal E_1 + \mathcal E_2.
\end{split}
\end{equation*}
with
\[
\mathcal E_1 := \left|  \int_0^T  \Big\{ \langle f,\tilde u \rangle + \langle g,\tilde v\rangle \Big\} dt
			-\Delta t \sum_{n=0}^{N} \Big\{  \langle f_h^{n+1/2}, S_{h}^1\tilde u(\cdot,(n+1/2)\Delta t) \rangle +
			\langle g_h^n, S_{h}^2 \tilde v(\cdot,n\Delta t) \rangle \Big\} \right|
\]
\[
\mathcal E_2 := \left| \Delta t \sum_{n=0}^N \Big\{ \langle f_h^{n+1/2}, 
				\tilde u_h^{n+1/2} - S_h^1 \tilde u(\cdot,(n+1/2)\Delta t) \rangle 
			+ \langle g_h^n, \tilde v_h^n - S_h^2 \tilde v(\cdot,n\Delta t) \rangle \Big\} \right|.
\]

For $\mathcal E_1$, we first focus on the terms involving $f$ and $\tilde u$;
\begin{equation*}
\begin{split}
	\left| \int_0^T \langle f, \tilde u\rangle \;  dt -
		\Delta t \sum_{n=0}^N \langle f_h^{n+1/2}, S_{h}^1\tilde u(\cdot,(n+1/2)\Delta t)\rangle \right| \\
	\le
	\left| \int_0^T \langle f,\tilde u \rangle \; dt - \Delta t \sum_{n=0}^N \langle f(\cdot,(n+1/2)\Delta t),\tilde u(\cdot,(n+1/2)\Delta t)\rangle \right| \\
	+ \left| \Delta t \sum_{n=0}^N \Big\{ \langle f_h^{n+1/2}, S_{h}^1\tilde u(\cdot,(n+1/2)\Delta t)\rangle - 
							     \langle f(\cdot,(n+1/2)\Delta t), \tilde u(\cdot,(n+1/2)\Delta t) \rangle \Big\} \right |,
\end{split}
\end{equation*}
which is nothing more than quadrature error for the time integration and the singular source discretization error.
In particular,
\[
	\left| \int_0^T \langle f,\tilde u \rangle \; dt - \Delta t \sum_{n=0}^N \langle f(\cdot,(n+1/2)\Delta t),\tilde u(\cdot,(n+1/2)\Delta t)\rangle \right| = O(\Delta t^2)
\]
from standard error estimates of the midpoint rule, and
\[
	\left| \Delta t \sum_{n=0}^N \Big\{ \langle f_h^{n+1/2}, S_{h}^1\tilde u(\cdot,(n+1/2)\Delta t)\rangle - 
							     \langle f(\cdot,(n+1/2)\Delta t), \tilde u(\cdot,(n+1/2)\Delta t) \rangle \Big\} \right |
	= O(h^q)
\]
from the singular source approximation theory (theorem 2).
Similar error estimates can be derived for the terms involving $g$ and $\tilde v$.\footnote{$O(\Delta t^2)$ quadrature error estimates follow from noting that the summation term coincides with the trapezoidal rule, assuming that $g(t)$ and $\tilde v(t)$ satisfy homogenous initial and final time conditions respectively.}

We bound $\mathcal E_2$ using $L^2$-error estimates of finite difference solutions for smooth problems (i.e., theorem \ref{thm:smconvg}), and using $L^2$-bounds of the discrete singular source approximations; one can show
\[
 	\| \eta_h^{\bf s_{\bf i}}\| = O(h^{-|{\bf s_{\bf i}}|-d/2}), \; i=1,2.
\]
%The abounds above need a summation over  |s_i| <= N^*
%Maybe consider simpler source case of a single multipole ?? MJB
Whence,
\begin{equation*}
\begin{split}
	\mathcal E_2 &\le \Delta t \sum_{n=0}^N \Big\{ \|f_h^{n+1/2}\| \; 
				\| \tilde u_h^{n+1/2} - S_h^1 \tilde u(\cdot,(n+1/2)\Delta t) \|
			+ \|g_h^{n}\|  \; \| \tilde v_h^n - S_h^2 \tilde v(\cdot,n\Delta t) \| \Big\}\\
					&= O((\Delta t^2+h^p)h^{-N^*-d/2}).
\end{split}
\end{equation*}
Estimate \ref{eq:weak_conv} thus follows.
\end{proof}

Consider the simpler case where only multipoles of order zero are present, that is $N^*=0$.
According to estimate \ref{eq:weak_conv}, we have weak convergence at rate $2-d/2$ if we choose $q=p$ and $\Delta t = O(h)$.
In other words, convergence (in the weak sense) is guaranteed though at a suboptimal rate.
For multipoles of order $N^*>0$, to retrieve the smooth solution behavior, we would require $p > N^*+d/2$ and $\Delta t\to 0$ like a positive power of $h$.

The (weak) error estimate we present here can be improved by making the following observation: the $h^{-d/2}$ factor originates from $L^2$-bounds of multipole source approximations.
In particular, we can remove this factor by using $L^1$-bounds instead, mainly
\[
	\| \eta_h^{\mathbf s_{\bf i}}\|_1 = O(h^{-|\mathbf s_{\bf i}|}), \; i=1,2.
\] 
Applying H\"older's inequality to $\mathcal E_2$ yields
\begin{equation}
\label{eq:weak_conv_infty}
\begin{split}
	\mathcal E_2 &\le \Delta t \sum_{n=0}^N \Big\{ \|f_h^{n+1/2}\|_1 \; 
			\| \tilde u_h^{n+1/2} - S_h^1 \tilde u(\cdot,(n+1/2)\Delta t) \|_{\infty}
			+ \|g_h^{n}\|_1  \; \| \tilde v_h^n - S_h^2 \tilde v(\cdot,n\Delta t) \|_{\infty} \Big\}\\
					&= O((\Delta t^2+h^{p})h^{-N^*}).
\end{split}
\end{equation}
We conjecture point-wise bounds hold for smooth $\tilde u$ and $\tilde v_h$, whence
\begin{equation}\label{eq:conjecture}
	\mathcal E = O(\Delta t^2 +h^q + (\Delta t^2+h^{p})h^{-N^*}),
\end{equation}
yielding optimal rates for the $N^*=0$ case, again if $q\ge p$.
Note that \ref{eq:weak_conv_infty} follows through if $L^\infty$ error terms $\| \tilde u_h - S_h^1 u \|_\infty$ and $\| \tilde v_h - S_h^2 v \|_\infty$ have optimal rates.
The authors are not aware of any such $L^\infty$ error
estimates for the types of hyperbolic systems considered here, see
however \cite{Brenner:06} and \cite{Layton:82} for some work in this
direction.

%see \cite{Brenner:06}, an overview of the theory for $L^p$ estimates and stability for initial value problems, this theory is based on Fourier multipliers and Besov spaces.
%Alternatively, there is a way to produce $L^\infty$ estimates requiring only $L^2$ stability (this is also discussed in \cite{Brenner:06}).
%In particular, we highlight work by \cite{Layton:82}, for his $L^\infty$ estimates (with $L^2$ stability) based on simple $L^2$-techniques demonstrating optimal convergence rates, albeit for a simple hyperbolic initial value problem.
%Again, we reiterate that the current theory on $L^\infty$ error estimate finite difference approximations is not immediately applicable to the class of hyperbolic systems considered here, and thus the improved convergence rates remain as a conjecture.

%Any discrete multipole of order zero (not a lot of variety there) is
%uniformly bounded in $L^1$. Therefore this last estimate shows that
%the multipole solution is weakly convergent of the same order as a
%smooth solution, mainly $2$-$2p$ convergence if source approximation order $q\ge 2p$.
%For a multipole of order $N^*$, to retrieve the smooth
%solution behaviour must have $2p > N^*$ and
%$\Delta t$ must go to zero like a positive power of $h$.

%\newpage
%So it follows from
%\ref{eqn:weak} that
%\[
%\Delta t \sum_{n} \langle r^n,u^n \rangle + \langle
%z^{n+1/2},v^{n+1/2}\rangle
%\]
%\[
%=\Delta t \sum_{n} \langle S(h)(W((n+1/2)\Delta t),Q(n\Delta t)),
%(f^{n+1/2},g^n)\rangle + O(h^{2p}+\Delta t^2) \times \|(f^{n+1/2},g^n)\|_{L^1}
%\]
%using the $L^{\infty}$ error estimate for the smooth solution $(w,q)$
%proved at the end of the last subsection. 
%
%Any discrete multipole of order zero (not a lot of variety there) is
%uniformly bounded in $L^1$. Therefore this last estimate shows that
%the multipole solution is weakly convergent of the same order as a
%smooth solution. For a multipole of order $k$, to retrieve the smooth
%solution behaviour must have $2p > k$ and
%$\Delta t$ must go to zero like a positive power of $h$.

%ADDED
\subsubsection{Acoustic Case}

As an example, consider the acoustic equations in first-order (pressure-velocity) form:
\begin{equation}\label{eq:acoustics}
\begin{split}
	\frac{1}{\kappa(\mathbf x)} \frac{\partial}{\partial t} p(\mathbf x,t) + \nabla\cdot \mathbf v(\mathbf x,t) 
		&= f(\mathbf x,t)\\
	\rho(\mathbf x) \frac{\partial}{\partial t} \mathbf v(\mathbf x,t) + \nabla p(\mathbf x,t) 
		&= \mathbf g(\mathbf x,t)
\end{split}
\end{equation}
where $u=p$ is the scalar pressure field and $v=\mathbf v\in\mathbf R^d$ is the vector particle velocity field.
Coefficient operator $A(\mathbf x)=1/\kappa(\mathbf x)I$ and $B(\mathbf x)=\rho(\mathbf x) I$, with $\kappa$ denoting bulk-modulus and $\rho$ density of the medium; here $I\in\mathbf R^{d\times d}$ is the identity matrix.
Lastly, the differential operator $P$ coincides with the gradient and its adjoint with the negative of the divergence,
\[
	P = \mat{\frac{\partial}{\partial x_1}\\ \vdots \\ \frac{\partial}{\partial x_d}}, \quad 
	P^T = - \mat{\frac{\partial}{\partial x_1},...,\frac{\partial}{\partial x_d}}.
\]

For the remainder of this section, we assume that the spatial
dimension $d=2$. 
The simplest staggered-grid finite difference scheme, second-order in time and space, is given by
\begin{equation}\label{eq:acoustics_sgfd}
\begin{split}
	\frac{1}{\kappa(ih,jh)} \frac{1}{\Delta t} \Big[  (p_h)^{n+1}_{i,j} - (p_h)^{n}_{i,j} \Big] + \hspace{7cm}\\
	\frac{1}{h} \Big[ (v_{1,h})_{i+1/2,j}^{n+1/2} - (v_{1,h})_{i-1/2,j}^{n+1/2} + 
				(v_{2,h})_{i,j+1/2}^{n+1/2} - (v_{2,h})_{i,j-1/2}^{n+1/2} \Big] 
		&= (f_h)^{n+1/2}_{i,j}\\
	\rho((i+1/2)h,jh) \frac{1}{\Delta t} \Big[ (v_{1,h})^{n+1/2}_{i+1/2,j} - (v_{1,h})^{n-1/2}_{i+1/2,j} \Big] +
	\frac{1}{h} \Big[ (p_h)_{i+1,j}^{n} - (p_h)_{i,j}^{n} \Big] 
		&= (g_{1,h})_{i+1/2,j}^{n}\\
	\rho(ih,(j+1/2)h) \frac{1}{\Delta t} \Big[ (v_{2,h})^{n+1/2}_{i,j+1/2} - (v_{2,h})^{n-1/2}_{i,j+1/2} \Big] +
	\frac{1}{h} \Big[ (p_h)_{i,j+1}^{n} - (p_h)_{i,j}^{n} \Big] 
		&= (g_{2,h})_{i,j+1/2}^{n}
\end{split}
\end{equation}
where grid functions $\{p_h,v_{1,h},v_{2,h}\}$ approximate the continuum fields $\{p,v_1,v_2\}$ respectively, % for system \ref{eq:cont_system}, mainly,
\begin{equation*}
\begin{split}
	(p_h)_{i,j}^{n} &\approx p(ih,jh,n\Delta t), \\
	(v_{1,h})_{i+1/2,j}^{n+1/2} &\approx v_1((i+1/2)h,jh,(n+1/2)\Delta t),\\
	(v_{2,h})_{i,j+1/2}^{n+1/2} &\approx v_2(ih,(j+1/2)h,(n+1/2)\Delta t).
\end{split}
\end{equation*}
We emphasize that the velocity fields for finite difference scheme \ref{eq:acoustics_sgfd} are staggered with respect to spatial grids, more specifically each component is shifted by half a cell size in its respective axis.
Relating system \ref{eq:acoustics_sgfd} with \ref{eq:sg}, we see
\[
	(u_h)_{i,j} = (p_h)_{i,j}, \quad 
	(v_h)_{i,j} = \mat{(v_{1,h})_{i+1/2,j}\\ 
				  (v_{2,h})_{i,j+1/2} }.
\]
The space $H_h^1$ in this example corresponds to the set of square summable 
scalar-valued functions on rectangular grid
\[
	\mathcal G((0,0),(1,1)) = \{ \mathbf x_{i,j} = (ih,jh) \; : \; i,j\in\mathbf Z \}.
\]
In other words, $H_h^1 = \mathcal H_h((0,0),(1,1))$.
$H_h^2$ is the set of square summable $\mathbf R^2$-valued functions on half-cell shifted grids, that is, 
\[
	H_h^2 = \mathcal H_h((1/2,0),(1,1)) \times \mathcal H_h((0,1/2),(1,1)).
\]

%\begin{equation*}
%\begin{split}
%	\langle u_h,\tilde u_h\rangle = h^2 \sum_{i,j} (u_h)_{i,j} (\tilde u_h)_{i,j} &\quad 	\text{for } u_h,\tilde u_h \in H_h^1\\
%	\langle v_h, \tilde v_h \rangle = h^2 \sum_{i,j} (v_h)_{i,j} \cdot (\tilde v_h)_{i,j} &\quad
%	\text{for } v_h,\tilde v_h \in H_h^2.
%\end{split}
%\end{equation*}
%Inner-products and norms of the spaces $\mathcal H_1, \mathcal H_2, H_{1,h}, H_{2,h}$ are all denoted by $\langle\cdot,\cdot\rangle$ and $\|\cdot\|$ and interpreted given the context, unless otherwise specified.

Sampling operators $S_h^1$ and $S_h^2$ coincide with evaluating continuum functions on grid or shifted grid points accordingly,
\[
	(S_h^1 u)_{i,j} = u(ih,jh), \quad 
	(S_h^2 v)_{i,j} = \mat{v_{1}((i+1/2)h,jh)\\ 
				  v_{2}(ih,(j+1/2)h) }.
\]
Discretized coefficient operators $A_h$ and $B_h$ are given by,
\[
	(A_h u_h)_{i,j} = \frac{1}{\kappa(ih,jh)} (u_h)_{ij}, \quad
	(B_h v_h)_{i,j} = \mat{ \rho((i+1/2)h,jh) (v_{1,h})_{i+1/2,j} \\
					 \rho(ih,(j+1/2)h) (v_{2,h})_{i,j+1/2} }.
\]
Lastly, $P_h$ and $P_h^T$ correspond to second-order approximations of the gradient and the negative of the divergence respectively;
\begin{equation*}
\begin{split}
	(P_{h}(r) u_h)_{i,j} &= r \mat{ (u_h)_{i+1,j} - (u_h)_{i,j} \\
					     (u_h)_{i,j+1} - (u_h)_{i,j} }, \\
	(P_{h}(r)^T v_h)_{i,j} &= -r \Big[ (v_{1,h})_{i+1/2,j} - (v_{1,h})_{i-1/2,j} + (v_{2,h})_{i,j+1/2} - (v_{2,h})_{i,j-1/2} \Big]
\end{split}
\end{equation*}
with $r =  \Delta t/h$.
In general, the differential operator $P$ can be approximated by a family of central difference operators $P_h$ of even order $p=2p^*$ for $p^*\in\mathbf N$, resulting in the $(2,2p^*)$ staggered-grid finite difference schemes.

%\hl{It follows that our weak convergence analysis applies to this example.}

\newpage
%%%%%%%%%%%%%%%%
\section{Numerical Tests}
%%%%%%%%%%%%%%%%

We have implemented the singular source approximation as discussed in the theory section
using the C++ packages \emph{IWave} and \emph{Rice Vector Library} (RVL);
\citep{GeoPros:11,RVL_TOMS}.  The IWave package %serves as
is a framework for finite difference solvers over uniform grids while
the RVL package provides a system of classes for expression of
gradient-based optimization algorithms over Hilbert spaces.  IWave and
RVL come together to form a modeling engine for seismic inversion and
migration. Implementation of multipole sources as RVL objects enables
straightforward composition with IWave solvers and inclusion in
inversion algorithms powered by RVL optimization code. Any other wave
equation solver wrapped in the appropriate RVL interfaces could be
coupled to the multipole source objects in the same way.

A convergence rate study is performed to corroborate theoretical results
pertaining to the accuracy of moment-consistent approximations to multipole sources.
In particular, our numerical experiments explore the semi-discrete error of staggered-grid finite difference solutions (time discretization errors are 
minimized by taking sufficiently small time steps).
We used the IWave implementation of staggered-grid finite difference schemes 
for the acoustic system \ref{eq:acoustics} \cite[]{Vir:84}, of order 2 in time and orders 2 and 4 in space - we
refer to these as the $(2,2)$ and $(2,4)$ order schemes respectively. In these
experiments we use only scalar (pressure) sources and pressure trace
data. Similar results are obtained with other choices. Boundary
conditions are of PML type, as described by \cite{Habashy:07},

The numerical experiments carried out concern multipoles in 2-D of the form
\[
	f(\mathbf x,t) = w(t) D^{\mathbf s} \delta(\mathbf x-\mathbf x^*),
\]
for $\mathbf s =(0,0),(0,1),(0,2)$.
The discretizations of $D^{\mathbf s}\delta$ are chosen as to achieve a target order of
convergence for the difference schemes, in most cases the nominal
spatial order ($q=2$ for the $(2,2)$ scheme, $q=4$ for the $(2,4)$ scheme). 
Again, the time step is fixed small enough that the time discretization error plays
essentially no role in the global error - it reflects truncation error
of the spatial derivative and the source approximation only.
Time-dependent function $w(t)$ is chosen to be a Ricker wavelet with peak 
frequency of $5Hz$, see figure \ref{fig:wlt-ricker}.

We approximate the convergence rate $R(\mathbf x)$ at a given location $\mathbf x$ via Richardson extrapolation,
\[
        R(\mathbf x) = \text{log}_2 \left( \frac{\| p_{h}(\mathbf x,\cdot)-p_{h/2}(\mathbf x,\cdot)\| }{\|p_{h/2}(\mathbf x,\cdot)-p_{h/4}(\mathbf x,\cdot)\| } \right)
\] 
where $p_h$ denotes the computed pressure field via finite difference using a grid size $h$ respectively.
The norm $\|\cdot\|$ is chosen to be 
\[
        \|p(\mathbf x,\cdot) \| := \sqrt{ \Delta t \sum_{k} |p(\mathbf x,t_k)|^2},
\]

Coordinates are aligned such that $\mathbf x = (z,x)$, 
where $z$ and $x$ refer to depth and horizontal distance respectively.  
%Similarly for 3-D, $\mathbf x=(z,x,y)$ where $y$ is
%measured along the remaining perpendicular direction.  
The computational domain consists of a $4km$ by $4km$ square medium
with a constant bulk modulus of $9 GPa$ and buoyancy of $1 cm^3/kg$,
thus a constant velocity of $3 km/s$.
%Receivers are placed in a 2-D grid covering the entire domain at every $40 m$ in each direction.
The source is placed slightly off of the center, $(z^*,x^*)=(-2003m,2003m)$, 
as to not coincide with a grid point for any of the computational uniform grids.
The following are some other specifications that apply to all tests carried out here:
\begin{itemize}
        \item total recording time is $1.5s$;
        \item spatial grid sizes $h=40m$, $h/2=20m$, $h/4=10m$;
        \item time step $\Delta t=0.5ms$;
\end{itemize}
%Note that the coarsest grid cell size is $40m$ which implies that at least $15$ 
%grid {\em points per wavelength (gpw)} at peak frequency of $5Hz$ or $5$ gpw at $15Hz$ 
%(see figure \ref{fig:wlt-ricker-fft}), in all numerical experiments, 
%thus minimizing the effects of grid dispersion on approximated convergence rates.

%Figures \ref{fig:model-geo-2D} and \ref{fig:model-geo-3D} describe the model 
%and source-receiver geometry for convergence rate tests in 2-D and 3-D respectively.
%The following are some other specifications that apply to all tests carried out here:
%\begin{itemize}
%        \item total recording time is $1.5s$;
%        \item spatial grid sizes $h=40m$, $h/2=20m$, $h/4=10m$;
%        \item time step $\Delta t=0.5ms$;
%\end{itemize}
%Note that the source location is chosen as to not coincide with a grid point for any of the computational uniform grids.
%Moreover, the coarsest grid cell size is $40m$, which implies at least $15$ grid {\em points per wavelength (gpw)}
%at peak frequency of $5Hz$ or $5$ gpw at $15Hz$ (see figure \ref{fig:wlt-ricker-fft}), in all numerical experiments, 
%thus minimizing the effects of grid dispersion on approximated convergence rates.

%%%%%%%%%%%%%%%%
\subsection{Results}

Figures \ref{fig:sq-test-s00q2-fd2-plotsnap} -- \ref{fig:sq-test-s02q2-fd2-plotsnap} 
plot a snapshot of the computed pressure field for different multipole sources.
Consistent with analytical formulas for the homogeneous unbounded medium case, 
the observed pressure field exhibited a polarity reversal (or lack of), a symmetry about the $x$-axis (i.e., $z=2003$), and an overall decrease in amplitude dependent on the multipole.

%Figures \ref{fig:test-s00q2-fd2-data10} -- \ref{fig:test-s02q2-fd2-data10} 
%and \ref{fig:test-s000q2-fd2-data10} -- \ref{fig:test-s020q2-fd2-data10} 
%plot pressure field traces versus horizontal location for different multipole sources in 2-D and 3-D respectively.
%Consistent with analytical formulas for the homogeneous unbounded medium case, 
%traces exhibited a polarity reversal (or lack of) and an overall decrease in amplitude dependent on 
%the multipole.

Convergence rates are plotted for the $\mathbf s=(0,0)$ case in figure 5, 
over the entire domain and at a particular depth of $z=2000$ in the left and right column plots respectively.
The first row of graphs shows results when using the $(2,2)$ finite difference scheme and the second-order source approximation.
We observe that second order convergence is indeed achieved away from the source.
Similarly, fourth order convergence results are observed when using the $(2,4)$ finite difference scheme and the fourth-order source approximation, i.e., second row in figure 5.
Lastly, the third row of plots demonstrates the negative effects of using a lower order source approximation relative to the spatial finite difference order, namely a second-order source approximation with a fourth-order method in space.
Clearly, fourth order rates are not achieved and moreover, there seems to be regions 
where rates dip below or above the expected second order.
Convergence results for $\mathbf s=(0,1)$ and $\mathbf s=(0,2)$ are given in figures 6 and 7 respectively with similar results.
The green line in figure \ref{fig:sq-test-s01q4-fd4-plots-wind} coincides with the nodal plane of the dipole source, that is, a line in the $zx$-plane where the source produces a null response.

%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%

Numerical results presented here validate the accuracy of moment-consistent singular source discretizations for controlling the propagation of finite difference truncation error for multipole sources.  
In particular, optimal spatial convergence rates for the $(2,2)$ and $(2,4)$ staggered grid finite difference methods are achieved, point-wise in space away from source location, when the source approximation order is equal to that of the spatial finite difference approximation order.

The onset distance of optimal order convergence is consistent with the support of the source approximation for the coarsest grid used in the convergence study, that is, $h(q+|{\bf s}|)$ with $h=40m$.
The erratic behavior of numerical convergence rates within the support of source approximations can be explained in part by how convergence rates were approximated.
Namely, numerical solutions over consecutively refined grids are essentially compared, though over the coarsest grid.
It follows that the support of $f_{h}$, the source approximation with respect to grid with cell size $h$, will contain the support of $f_{h/2}$ and a region in the $h$-grid outside the direct influence of $f_{h/2}$, resulting in irregular convergence rates.

As mentioned in the Literature Review, \cite{Petersson:2016}
demonstrated that optimal convergence rates can be achieved if the
source discretization satisfied moment as well as smoothness
conditions. \cite{Petersson:2016} used central difference
approximations to the space derivatives and Runge-Kutta time
integration, for various advection problems and for acoustics. The
staggered grid schemes used in our work, however, do not require
additional constraints beyond the moment conditions for the multipole
source approximations.

%As our numerical tests demonstrate, smoothness conditions were not necessary here, which is attributed to the fact that we use staggered-grid finite difference schemes. 
%In particular, the central finite difference operator used in \cite{Petersson:2016} have an associated grid spacing of $2h$ with respect to an $h$-sized grid, coupled with a source approximation of narrow support, i.e., $2\epsilon=(q+|\mathbf s|)h$, results in the triggering of spurious modes unresolved by the numerical solution.
%The finite difference operators used here however have a grid spacing of $h$ given that they approximate derivatives over staggered grids.
%Thus, in essence our narrow source approximations ``seem'' twice as wide, and therefore smoother, from the point of view of the difference operator $P_h$.
%For this reason, smoothness conditions are not necessary for staggered-grid finite difference schemes.

%
%%MJB edit
%As mentioned in our discussion of singular source approximation theory,
%current literature fails to provide analysis directly applicable to the solution of 
%acoustodynamics (equations \ref{eq:acous}) or elastodynamics (equations \ref{eq:elas}) with multipole sources.
%The most complete account, as far as we are aware, of the analysis of approximations 
%to PDEs with singular sources is given by \cite{hoss:16} in the context
%of regularizations of the delta distribution.
%\citeauthor{hoss:16} developed a unified framework for analyzing approximation errors 
%that can be readily applied to any numerical scheme.
%Their work in particular addressed two fundamental questions: 
%(i) What form of convergence should be used to examine $f^h\to f$? 
%That is, how does our source approximations converge to multipoles as distributions?
%(ii) What form of convergence should be used to examine $u^h\to u$, 
%convergence of numerical solution $u^h$ with source approximation $f^h$ to true
%solution $u$ with source $f$?
%Authors showed that convergence of $f^h\to f$ in the distribution sense (weak-* topology),
%and in some weighted Sobolev norm, can be achieved at a desired rate if a set of 
%{\em continuous moment conditions} are satisfied by the approximation $f^h$, 
%analogous to the discrete moment conditions discussed here.
%Furthermore, they discussed answers to question (ii) and the interplay of approximation $f^h$ with
%the convergence of $u^h\to u$ for protopical elliptic and hyperbolic PDEs.
%
%%MJB edit
%Numerical results presented here, and by \cite{Petersson:2010}, 
%help motivate the use of approximations to multipoles as suggested by
%\cite{Walden:1999} and \cite{TorEng:04}.
%Moreover, we believe it is possible to provide a formal error analysis of our conjecture 
%by building upon the theoretical framework laid out by \cite{hoss:16}.
%%This would entail answering the following questions:
%%Can the theory be extended to distributions of the form $D^{\bf s}\delta$?
%%If so, do source approximations that satisfy discrete moments (equation \ref{eq:qmom})
%%also satisfy some version of the continuous moments as given by \cite{hoss:16}?
%%Question (ii) and the interplay of approximation $f^h$ with
%This would entail analyzing the interplay of approximations $f^h$ with
%the convergence of $u^h\to u$ for our particular PDEs and choice of 
%numerical scheme.
%
%The RVL-based structure of our MPS framework does not really get a
%solid workout in this study. Inspection of the code and scripts
%shows that the framework is a convenient environment for 
%implementing the numerical examples presented in the last
%section. However the capabilities of this mode of code organization
%emerge much more clearly in inversion applications. For example, we have shown in
%another place REFERENCE??? that a choice of norm in an MPS space, that
%reflects the effect of the spatial delta derivatives on temporal
%frequency, can dramatically accelerate the convergence of
%source recovery by Krylov subspace iteration. For this, it was only
%necessary to pass the {\tt RVL::LinOpValOp} representing the MPS-to-data 
%trace map, and an {\tt RVL:LinearOp}
%representing the Gram operator of the norm,  to an RVL implementation of the
%preconditioned conjugate gradient method, used completely without
%alteration. It will also be possible to incorporate the MPS based
%operator into the RVL implementation of variable projection method
%\cite[]{LiRickettAbubakar:13} for joint source-medium inversion
%without any change either to the MPS code or to the RVL variable
%projection algorithm.
%
%%We would be remised if we did not point out some of the attractive features present in this singular source approximation.
%%For one, incorporating source approximations is rather straight-forward and minimaly invasive to the finite difference implementation, a matter of cooking up the correct collection of RHS sources and injecting them at proper grid points.
%%Moreover, the relationship between our source parameterization via MPS coefficients and RHS source-terms for finite difference solvers is a linear one, aptly implemented by our \texttt{MPS\_to\_RHS} operator.
%%Ultimately the singular source approximation implemented in our framework preserves linearity of the MPS-to-data map in a natural way that will be crucial for source inversion.
%
%We have not addressed two important issues in the work reported here.
%The first is the order of multipole necessary to represent a given
%degree of source anisotropy. Point support is an idealization: active source regions may
%be small or comparable to a wavelength in spatial
%extent. \cite{SantosaSymes:00} showed that multipole approximation
%(equation \ref{eq:MPSappx}) of acoustic sources vanishing outside
%a small region of space
%exhibits a threshhold effect, in that the error in the
%resulting acoustic fields drops abruptly as the length of the series
%($N$ in the expression \ref{eq:MPSappx})
%is increased past a critical value.
%This threshhold in the number of terms necessary for accurate
%approximation depends on the size of the active region and a measure of energy
%output relative to energy input (related to the degree of
%anisotropy).
%The analysis in \cite[]{SantosaSymes:00} pertains to acoustics,
%% the leap to linear elasticity is rather justifiable. (REALLY?)
%however we expect similar results to hold for linear elastodynamics.
%Earthquake seismology has 
%%indeed benefitted from the applicability of
%long used multipole source approximations in elastic media 
%%particularly for the determination of
%to describe earthquake mechanisms. 
%The %basic earthquake representation is  
%%what is referred to as 
%%the 
%seismic moment tensor
%%which is 
%represents earthquake source mechanisms by a combinations of point-dipole body forces, that is, a
%first-order vector-valued multipole \cite[]{Backus:1976a,Shearer:2009}.
%Moment tensors of higher order (that is, higher order multipole series) have been shown to be important in cases where finiteness of the source is of issue, that is the fault size is comparable to propagating wavelengths \cite[]{Stump:1982}.
%
%%WWSAccurate source representation and estimation is crucial to the seismic inversion problem, whos focus is the recovery of medium parameters.
%%The moderate success of
%%Joint determination of medium and source parameters
%% in 
%%for improving seismic inversion 
%%has motivated 
%%motivates our work on MPS representation of seismic sources.
%A second natural question is: how much source anisotropy is really
%necessary to fit field data well?
%%Since source radation pattern anisotropy has a first-order effect on
%%seismic amplitudes, it is natural to expect that 
%%anisotropic source representation, such as multipole series, would be
%%an important ingredient in seismic inversion.
%\cite{SymMink:97} 
%%motivate the need for joint medium-source parameter inversion and a source representation that accounts for anisotropy in the context of plane-wave viscoelastic modeling of marine reflection data.
%%Their results 
%show that inverting marine reflection data for multipole source
%parameters together with a layered viscoelastic model,% as oppose to
%%using a given isotropic modeled source or even inverting for an
%%isotropic source, %allowed them to account 
%resulted in fitting 25\% more of the data than was possible with any
%isotropic source, and allowed
%%were even able to achieve a dramatic 
%90\% data fit to the target portion of the data. Moreover, recovered
%p-wave and s-wave impedance parameters matched closely the expected
%seismic-lithologic signature of the gas sand only when %jointly
%                                %estimating for 
%viscoelastic model and anisotropic source parameters were simultaneously esitmated.
%The order of multipole required to achieve this degree of data fit and
%well log tie was $N=6$. Obviously the necessary order depends on many
%factors, and varies from survey to survey. The framework we have
%provided here offers a platform in which to develop an approach to
%source anisotropy estimation as part of the overall inversion process.

%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
%%%%%%%%%%%%%%%%%%%%%

In this paper we have covered the singular source approximation theory
based on moment conditions, which essentially has approximations mimic 
the behavior of the target distribution $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$ on polynomials.
Moreover, we give explicit forms of source approximations with narrow support,
based on the discrete moment conditions in the context of finite difference solvers:
diameter $2\epsilon = (q+|\mathbf s|) h$ for a multipole of order $|\mathbf s|$
over a grid of size $h$.
As a new result, we connect the discrete and continuum singular source approximation
theory by proving that continuum functions generated from the discrete moment conditions 
indeed satisfy the continuum moment conditions.

Our main contribution was the development of a weak convergence theory that is applicable to a large set of wave propagation problems (including acoustics and elasticity in first-order form) solved via a family of staggered-grid finite difference schemes.%MOD, larger than what is reported in the current literature.
Posing the convergence mode of numerical solutions in terms of weak convergence was indeed a natural choice given that source terms are derivatives of the Dirac delta function, that is, distributions.
%ADDED
\hl{
The weak convergence theory relied on the structure of the the general continuum problem} \ref{eq:cont_system} \hl{and the ability for its discretization to preserve this structure, i.e, coefficient operators ($A,B,A_h,$ and $B_h$) are symmetric and bounded.% and the differential operators $P$ and $P_h$ satisfy a skew-adjoint relation.
The general staggered finite difference scheme} \ref{eq:sg} \hl{is also assumed to be staggered in time, in particular a central difference approximation of the time derivative was applied.
Given the aforementioned assumptions on the continuum problem and its discretization, energy estimates along with the singular source approximation theory (theorem }\ref{thm:discweakconv} \hl{) are the key ingredients that gives us our main weak convergence result, theorem} \ref{thm:conv}.
%END ADDED
Numerical results, however, give evidence of stronger convergence, namely 
optimal convergence rates given by numerical scheme under smooth conditions,
point-wise away from source location for appropriate source discretizations and
in particular for multipoles of order $|\mathbf s |=0,1$ and $2$.

%%%%%%%%%
\section{Acknowledgements}

We are grateful to the sponsors of The Rice Inversion Project for their long-term support,
and to the Rice Graduate Education for Minorities (RGEM) and XSEDE scholarship programs for their support of M. Bencomo's Ph.D. research.
This material is also based upon work supported by the National Science Foundation under Grant No. DMS-1439786 while the author was in residence at the Institute for Computational and Experimental Research in Mathematics in Providence, RI, during the Fall 2017 semester.

%%%%%%%%%
% PLOTS 
%%%%%%%%%
\inputdir{project}

\multiplot{6}{ssappx-s0q2,ssappx-s0q4,ssappx-s1q2,ssappx-s1q4,ssappx-s2q2,ssappx-s2q4}{width=0.33\textwidth}{Plots of $\eta^\epsilon(x)$, 1-D approximations to $D^s\delta(x)$ with $h=1$. (a) $s=0$ and $q=2$, (b) $s=0$ and $q=4$, (c) $s=1$ and $q=2$, (d) $s=1$ and $q=4$, (e) $s=2$ and $q=2$, (f) $s=2$ and $q=4$.}

\multiplot{6}{ssappx-s00q2,ssappx-s00q4,ssappx-s01q2,ssappx-s01q4,ssappx-s02q2,ssappx-s02q4}{width=0.33\textwidth}{Plots of $\eta^\epsilon(\mathbf x)$, 2-D approximations to $D^{\mathbf s}\delta(\mathbf x)$ with $\mathbf l=(1,1)$ and $h=1$. (a) $\mathbf s=(0,0)$ and $q=2$, (b) $\mathbf s=(0,0)$ and $q=4$, (c) $\mathbf s=(0,1)$ and $q=2$, (d) $\mathbf s=(0,1)$ and $q=4$, (e) $\mathbf s=(0,2)$ and $q=2$, (f) $\mathbf s=(0,2)$ and $q=4$.}

\multiplot{2}{wlt-ricker,wlt-ricker-fft}{width=0.6\textwidth}{Ricker wavelet with peak frequency $5Hz$: (a) time plot, (b) power spectrum plot.}

\multiplot{3}{sq-test-s00q2-fd2-plotsnap,sq-test-s01q2-fd2-plotsnap,sq-test-s02q2-fd2-plotsnap}{width=0.48\textwidth}{Snapshot of pressure field ($t=0.75$ sec) computed using $(2,2)$ finite difference scheme ($h=10m, \Delta t=0.5$ ms) and second-order approximation of multipole source $D^{\bf s}\delta({\bf x}-{\bf x^*})$: (a) ${\bf s}=(0,0)$, (b) ${\bf s}=(0,1)$, and (c) ${\bf s}=(0,2)$.}

\multiplot{9}{sq-test-s00q2-fd2-plots,sq-test-s00q2-fd2-plots-wind,sq-test-s00q2-fd2-plots-wind2,sq-test-s00q4-fd4-plots,sq-test-s00q4-fd4-plots-wind,sq-test-s00q4-fd4-plots-wind2,sq-test-s00q2-fd4-plots,sq-test-s00q2-fd4-plots-wind,sq-test-s00q2-fd4-plots-wind2}
{width=0.25\textwidth}{Convergence rates results for ${\mathbf s}=(0,0)$. Using (a) $(2,2)$ FD scheme and second-order source approximation, (d) $(2,4)$ FD scheme and fourth-order source approximation, (g) $(2,4)$ FD scheme and second-order source approximation. Plots (b), (e), (h) correspond to convergence rates at $z=2000$, related to the first column from the left. Similarly, plots (c), (f), (i) show rates at $z=-1600$.}

\multiplot{9}{sq-test-s01q2-fd2-plots,sq-test-s01q2-fd2-plots-wind,sq-test-s01q2-fd2-plots-wind2,sq-test-s01q4-fd4-plots,sq-test-s01q4-fd4-plots-wind,sq-test-s01q4-fd4-plots-wind2,sq-test-s01q2-fd4-plots,sq-test-s01q2-fd4-plots-wind,sq-test-s01q2-fd4-plots-wind2}
{width=0.25\textwidth}{Convergence rate results for ${\mathbf s}=(0,1)$. Using (a) $(2,2)$ FD scheme and second-order source approximation, (d) $(2,4)$ FD scheme and fourth-order source approximation, (g) $(2,4)$ FD scheme and second-order source approximation. Plots (b), (e), (h) correspond to convergence rates at $z=2000$, related to the first column from the left. Similarly, plots (c), (f), (i) show rates at $z=-1600$.}

\multiplot{9}{sq-test-s02q2-fd2-plots,sq-test-s02q2-fd2-plots-wind,sq-test-s02q2-fd2-plots-wind2,sq-test-s02q4-fd4-plots,sq-test-s02q4-fd4-plots-wind,sq-test-s02q4-fd4-plots-wind2,sq-test-s02q2-fd4-plots,sq-test-s02q2-fd4-plots-wind,sq-test-s02q2-fd4-plots-wind2}
{width=0.25\textwidth}{Convergence rate results for ${\mathbf s}=(0,2)$. Using (a) $(2,2)$ FD scheme and second-order source approximation, (d) $(2,4)$ FD scheme and fourth-order source approximation, (g) $(2,4)$ FD scheme and second-order source approximation. Plots (b), (e), (h) correspond to convergence rates at $z=2000$, related to the first column from the left. Similarly, plots (c), (f), (i) show rates at $z=-1600$.}

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/mps_theory/paper_orig.tex
repro docs dir added

--- 2019-10-19 repro/papers/mps_theory/paper.tex
@@
-%\title{Discretization of Multipole Sources in a Finite Difference Setting for Wave Propagation Problems}
-%\date{}
-%\address{
-%        \footnotemark[1]Institute for Computational and Experimental Research in Mathematics,\\ Brown University,\\ Providence, RI 02912 USA\\
-%        \footnotemark[2]The Rice Inversion Project,\\ Rice University,\\ Houston, TX
-%        77005-1892 USA
-%}
-%\author{Mario J. Bencomo\footnotemark[1] and William Symes\footnotemark[2]}
-
-%\righthead{Multipole Source Discretizations}
-
-%\maketitle
-%\parskip 12pt
+\title{Discretization of Multipole Sources in a Finite Difference Setting for Wave Propagation Problems}
+\date{}
+\address{
+        \footnotemark[1]Institute for Computational and Experimental Research in Mathematics,\\ Brown University,\\ Providence, RI 02912 USA\\
+        \footnotemark[2]The Rice Inversion Project,\\ Rice University,\\ Houston, TX
+        77005-1892 USA
+}
+\author{Mario J. Bencomo\footnotemark[1] and William Symes\footnotemark[2]}
+
+\righthead{Multipole Source Discretizations}
+
+\maketitle
+\parskip 12pt
 
@@
 representations generating anisotropic radiation patterns. 
-Such seismic sources can be modeled as 
-{\em multipoles}, that is, time-dependent linear combination of
-spatial derivatives of the spatial delta function. Since the solutions of linear hyperbolic
-systems with point-source right hand sides are necessarily singular, standard results on
-convergence of grid-based numerical methods (finite difference or
-finite element) do not imply convergence of numerical solutions.
-We present a method for discretizing multipole sources 
+Such seismic sources can be modeled as linear combinations of 
+{\em multipole point-sources}.
+In this paper we present a method for discretizing multipole sources 
 in a finite difference setting, an extension of the moment matching 
 conditions developed for the Dirac delta function in other applications.
-along with numerical evidence demonstrating the accuracy of these approximations.
-Using this analysis, we develop a weak convergence theory for the discretization of 
-a family of symmetric hyperbolic systems of first-order partial
-differential equations, with singular source terms, solved via
-staggered-grid finite difference methods: we show that
-grid-independent space-time averages of the numerical solutions
-converge to the same averages of the continuum solution, and provide
-an estimate for the error in terms of moment matching and truncation
-error conditions.
-Numerical experiments confirm this result, but also suggest a stronger
-one: optimal convergence rates appear to be achieved
-achieved point-wise in space away from the source.
+We also provide the necessary analysis and numerical evidence 
+to demonstrate the accuracy of our singular source approximations.
+In particular, we develop a weak convergence theory for the discretization of 
+a family of symmetric hyperbolic systems of first-order partial differential equations, with singular source terms, solved via staggered-grid finite difference methods.
+Numerical experiments demonstrate a stronger result than 
+what is presented in our convergence theory, namely, optimal convergence rates 
+of numerical solutions are achieved point-wise in space away from the source if an 
+appropriate source discretization is used.
 
@@
 
-%Seismic sources are commonly idealized as concentrated at a source point due
-%to their small spatial extent relative to seismic wavelengths.
-Linear wave propagation problems modeling physical systems with
-concentrated (in space) energy sources arise in a number of
-applications, such as seismology, nondestructive material evaluation,
-and radar. A familiar example of a wave propagation model with
-spatially concentrated 
-%energy
-energy
-source is the isotropic point-radiator problem
+Seismic sources are commonly idealized as concentrated at a source point due
+to their small spatial extent relative to seismic wavelengths.
+A simple and familiar example of a wave propagation model with
+spatially concentrated source is the isotropic point-radiator problem
 for the acoustic wave equation in Euclidean 3-space \cite[]{CourHil:62}:
@@
 \begin{split}
-       \frac{1}{c^2} \frac{\partial^2}{\partial t^2} p(\mathbf x,t) - \nabla^2
+        \frac{\partial^2}{\partial t^2} p(\mathbf x,t) - c^2\nabla^2
   p(\mathbf x,t) & = f(\mathbf x,t) \equiv w(t)\delta(\mathbf x),\\ 
@@
 \label{eq:green3d}
-	p(\mathbf x,t) = \frac{w\left(t-\frac{r}{c}\right)}{4\pi r}, \quad 
+	p(\mathbf x,t) = \frac{w\left(t-\frac{r}{c}\right)}{4\pi c^2 r}, \quad 
 	r= \sqrt{\mathbf x^T \mathbf x}.
 \end{equation}
-
-% here is a heaven-sent opportunity to tell the reader what the paper
-% is about
-
-The form \ref{eq:green3d} of the isotropic point radiator solution
-suggests an obstacle to successful  approximation by gridded (finite
-difference, finite element,...) numerical methods: the fundamental
-rationale for the accuracy of such methods, namely truncation error
-analysis based on Taylor expansion, is invalid in any neighborhood of
-the source point $\mathbf x = \mathbf 0$, as the solution is singular
-there. Moreover, due to the wave nature of both continuum and gridded
-approximate solutions for problems like \ref{eq:isorad}, errors
-arising near the source point propagate throughout the region in which
-the solution is computed. Examples presented later in this paper
-illustrate this generation and propagation of error due to source singularity.
-
-The main result of this paper is that certain special approximations
-to wave dynamics problems such as \ref{eq:isorad} with (singular)
-right-hand sides of point support lead to finite difference solutions
-that converge globally in a weak sense - that is, averages over fixed
-regions converge as the grid cell size is reduced.
-
-In the remainder of this introduction, we describe the class of wave
-dynamics problems to which our arguments apply, the representation and approximation
-of right-hand sides,  the type of numerical
-approximation used in our work, and the weak convergence of approximate solutions.
-
-We focus our attention on a symmetric hyperbolic systems of the form
-\begin{equation}\label{eq:hyper_fam}
-\begin{split}
-	A \frac{\partial}{\partial t}u + P^T v &= f \\
-	B \frac{\partial}{\partial t}v  - P u & = g
-\end{split}
-\end{equation}
-connecting dynamical fields $u,v$, energy source fields $f,g$, and symmetric
-positive definite matrix fields $A,B$. The operator $P$ is a
-constant-coefficient differential operator of order 1; $P^T$ denotes its
-formal adjoint or transpose.
-Many important physical systems supporting wave propagation can be
-cast in the form \ref{eq:hyper_fam}. These include linear
-elastodynamics and Maxwell's equations. In all such examples, the
-matrices $A$ and $B$ encode time-independent physical parameters characteristic of the
-continuum mechanical model. For example, elastodynamics may be cast in
-the form \ref{eq:hyper_fam}, with $A$ being the compliance tensor,
-$B$ being the material density multiplied by the $3 \times 3$
-identity, and $P$ is the strain operator. As is well-known, the acoustic
-wave equation \ref{eq:isorad} is also equivalent to such a system, in the
-sense that its solution $p$ can be identified with $u$ above, with $v$
-chosen as the indefinite time integral of $-\nabla p$. In this case,
-$A$ is the $1 \times 1$ matrix with entry $c^2$, $B$ is the $3
-\times 3$ identity matrix, and $P$ is the gradient operator. The resulting system expresses
-acoustodynamics with material density normalized to 1, and $c^2$
-identified with bulk modulus of the fluid material. The dynamic fields
-$p$ and $v$ represent pressure and particle velocity respectively.
-
-Sources of small support relative to typical wavelengths are reasonable models
-for some seismic energy generation mechanisms. Such sources may be
-approximated by point support right hand sides such as occur in the
-isotropic radiator problem \ref{eq:isorad}. Of course, ``functions''
-of space and time having point support in space are not really
-functions at all, but (Schwarz) distributions. 
-Peetre's Theorem \cite[]{Horm:69} implies that any distribution in
-${\bf R}^{d+1}$ 
-supported on $\{\mathbf x^* \} \times {\bf R}$, ${\bf x^*} \in {\bf R}^d$, takes the form of a finite series
+Both active and earthquake seismic sources, however, generate
+spatially asymmetric wavefields (see for example \cite{Shearer:2009}, \cite{Yil:01}). 
+The acoustic isotropic point-radiator 
+is therefore inadequate as a model of seismic wave generation
+and propagation because of its prediction of spatial symmetry.
+The symmetry of the solution arises in part from the spherical symmetry of
+the right-hand side in equation \ref{eq:isorad}. Therefore, accurate
+modeling of seismic wavefields must include energy source (right-hand
+side) representations generating anisotropic radiation patterns.
+
+A {\em multipole}, or {\em multipole source} in the 
+context of source modeling, is a finite linear combination of partial 
+derivatives of the spatial Dirac delta function. Such sources combine
+localization of energy and anisotropic radiation pattern. In fact,
+Peetre's Theorem \cite[]{Horm:69} implies that any function of space
+and time $f(\mathbf x,t)$ concentrated entirely at a point in space
+(of {\em point support}) is a multipole of finite order $N \ge 0$,
 \begin{equation}\label{eq:MPSappx}
@@
 in which we have introduced {\em multi-index notation}: for spatial
-dimension $d$ and multi-index (integer d-tuple) ${\bf s} =
+dimension $d=\{1,2,3\}$ and multi-index (integer d-tuple) ${\bf s} =
 (s_1,...,s_d)$, the $\mathbf s$-mixed partial derivative operator, denoted $D^{\bf s}$,
@@
 \end{equation}
-In the context of energy source representation, we shall refer to
-distributions of point support, necessarily taking the form
-\ref{eq:MPSappx}, as {\em multipole sources}, or simply {\em multipoles}.
-Such sources combine
-localization of energy and (possibly) anisotropic radiation pattern.
-%distribution in space and time. ?? MJB
-The coefficient distributions $w_{\mathbf s}(t)$ may be scalar-, vector-, or
+The coefficient time functions $w_{\mathbf s}(t)$ may be scalar-, vector-, or
 tensor-valued, according to the nature of the quantity (pressure, velocity, or stress) 
@@
 
-The isotropic acoustic point radiator problem \ref{eq:isorad} is
-(equivalent to) a
-simple special case of the symmetric hyperbolic system
-\ref{eq:hyper_fam} with multipole
-right hand sides $f,g$. Therefore gridded approximation of the more general
-dynamics modeled by \ref{eq:hyper_fam} suffers from the difficulty
-pointed out earlier for \ref{eq:isorad}.
-
-Our results pertain to a specific class of finite difference
-approximations for systems of the form \ref{eq:hyper_fam},
-the so called {\em staggered grid schemes}, defined in detail in below
-(equation \ref{eq:sg} and surrounding discussion). These schemes have the
-virtue of obeying {\em energy estimates}, which are the chief
-analytical tool in our work, and are widely used for basin-
-and exploration-scale seismic modeling, acoustics computations,
-structural dynamics, and many other purposes; see \cite{moczoetal:06} for an
-excellent overview and many older references. Other classes of finite
-difference approximations are also used in many of the same applications
-\cite[]{Cohen:01,moczoetal:06,Petersson:2010,Petersson:2016}, as are finite element 
-(spectral element, continuous and discontinous Galerkin) methods 
-\cite[]{KomTromp:00,Cohen:01,Ghattas:IP25}. Results similar to those
-reported here may hold for some of these approaches, however we do not
-address this question here.
-
-Staggered grid finite difference approximations $(u^n_h,v^{n+1/2}_h)$
-to a solution  $(u,v)$ of
-\ref{eq:hyper_fam} depend on the spatial grid cell size $h>0$ and the
-time step $\Delta t>0$: $u^n_h$ approximates $u(\cdot,n\Delta t)$,
-$v^{n+1/2}_h$ approximates $u(\cdot,(n+1/2)\Delta t)$. A sufficient
-condition for consistent staggered grid schemes to produce convergent
-approximations to 
-smooth solutions of systems \ref{eq:hyper_fam} takes
-the form of a {\em Courant-Friedrichs-Levy stability condition}, that is, a bound on
-$\Delta t/h$, depending on the coefficient matrices $A,B$. In order to
-express the error asymptotics of these schemes in a convenient way, we
-will assume that $\Delta t$ is a function of $h$, obeying the
-stability condition.
-
-As pointed out earlier, point-wise convergence to singular solutions
-cannot be expected, as the truncation error analysis on which it is
-based makes no sense in the singular case. Instead, we turn to {\em
-  weak} approximation: the approximate
-solution $(u^n_h,v^{n+1/2}_h)$ converges weakly to continuum fields $(u,v)$ iff
-arbitrary positively weighted averages of $(u^n_h,v^{n+1/2}_h)$
-converge to the same weighted averages of
-samplings of $(u,v)$ on the space-time $(h,\Delta t)$ grid, as $h,
-\Delta t \rightarrow 0$ subject to the stability condition. 
-The averaging weights are smooth functions of position (or position and time)
-independent of grid, sampled at grid-points to create discrete weight functions.
-Precise definitions are supplied below.
-
-Our main result is that staggered grid approximations
-$(u^n_h,v^{n+1/2}_h)$ 
-converges weakly to a solution of \ref{eq:hyper_fam} with multipole right-hand side $(f,g)$, provided that the
-finite-difference representation of the right-hand side
-satisfies certain {\em discrete moment conditions}
-\cite[]{Walden:1999,TorEng:04}, expressing accuracy of the 
-right-hand side
-approximation in a weak sense. ``Singular'' is used in
-this paper as a synonym for {\em non-smooth distribution}, that is,
-continuous linear functional on various topological vector spaces of
-smooth functions, not representable by integration against a smooth kernel. For example, the right-hand side of the isotropic
-radiator problem \ref{eq:isorad} defines a continuous linear
-functional on the space of continuous functions with the sup norm. The
-moment conditions used in this work simply specify that the numerical
-approximations (linear combinations of grid values, with coefficients
-depending on the grid diameter $h$) give the {\em same} result as the distribution being
-approximated on polynomials with degree less than a bound depending on
-the dimension $d$, the nature of the distribution (right-hand
-side). This degree bound must be at least the {\em multipole order}
-$N$, the highest derivative appearing in the definition
-\ref{eq:MPSappx}. Express the degree bound as $N+q$, $q \ge 0$: we
-call $q$ the {\em approximation order} of the discrete multipole representation.
-
-The weak error asymptotics of a staggered grid scheme applied to a
-problem with multipole right-hand side, using a discrete multipole
-representation satisfying a discrete moment condition, depend on the
-quantities introduced in the preceding paragraph, and on the error
-asymptotics of the staggered grid scheme for smooth solutions. The
-schemes we shall investigate produce approximations with error $O(\Delta t(h)^2 + h^p)$
-(abbreviated as order $(2,p)$) to smooth solutions of systems \ref{eq:hyper_fam}.
-%We give explicit rates of convergence, depending on on singular source
-%approximation order $q$, spatial order $p$ for a staggered-grid
-%finite difference scheme of with truncation error $O(\Delta t(h)^2,h^{p})$, the maximum multipole
-%order between $f$ and $g$ denoted by $N^*$, and spatial dimension $d$.
-%include explicitly the error bounds here
-For spatial dimension $d$, multipole right-hand sides $(f,g)$  in
-\ref{eq:hyper_fam}  of order $N$, discrete multipole
-representation of approximation order $q$, and staggered grid scheme
-of order $(2,p)$, we show that the error of averages with smooth averaging
-functions $\tilde f, \tilde g$ satisfies the estimate
-\begin{equation}
-\label{eq:intro_weak_est}
-\begin{split}
-	\left| \int_0^T  \Big\{ \langle u,\tilde f\rangle + \langle v,\tilde g\rangle \Big\} dt -
-		\Delta t \sum_{n=0}^N \Big\{ \langle u_h^{n}, \tilde f_h^{n} \rangle +
-						 	   \langle v_h^{n+1/2}, \tilde g_h^{n+1/2}  \rangle \Big\} \right|\\
-	= O(\Delta t(h)^2 + h^q + (\Delta t(h)^2+h^{p})h^{-N-d/2}).
-\end{split}
-\end{equation}
-(Theorem \ref{thm:conv}). As discussed in the subsection on weak
-convergence, staggered grid schemes typically involve multiple grids
-with diameter $h$. In the estimate \ref{eq:intro_weak_est},
-$\tilde{f}_h^{n}$ represents the samples on the grid(s) of diameter
-$h$ of the smooth function $\bx \mapsto \tilde{f}(\bx,n\Delta t)$, and
-similarly for $g$. Thus left-hand side of \ref{eq:intro_weak_est}
-compares continuum ($L^2$) and discrete (weighted $l^2$) inner
-products, the latter being the result of quadrature rules applied to
-the former.
-
-Note the implications of this estimate: to assure an error tending to
-zero like $O(h^m)$, it is sufficient to require that $\Delta t(h) = O(h^{p/2})$, $q
-\ge m$, $p \ge m +N+d/2$. These appear to be very stringent
-requirements, compared to those for point-wise convergence to smooth solutions: for example for $N=0$ (that is, isotropic point source),
-to achieve $m=2$ it is required that $q \ge 2$, $p \ge 2+d/2$ and $\Delta t(h) =
-O(h^{1+d/4})$. In remarks after the proof of Theorem \ref{thm:conv},
-we point out that if staggered grid approximations to {\em smooth} solutions can be shown to obey certain
-$L^{\infty}$ estimates, as opposed to the $L^2$ estimates used here,
-then the $h^{d/2}$ factor drops out of the right-hand side of the
-estimate \ref{eq:intro_weak_est}, and the error asymptotics would
-become optimal, that is, the same as those for smooth solutions. We
-conjecture that such sup norm estimates for convergence of averages
-hold, as our numerical results suggest, but their proof is beyond the
-scope of this paper.
-
-We assume throughout that the time-independent
-coefficient matrices $A,B$ in the system \ref{eq:hyper_fam} are
-smooth ($C^{\infty}$) in their dependence on the space variables. It
-is certainly possible to extend some of these results to larger
-classes of coefficients defined by weaker regularity
-conditions - indeed, the weak solutions for systems of
-the form \ref{eq:hyper_fam} exist if $A, B$ and their inverses are
-uniformly bounded and measurable functions of the space variables
-\cite[]{BlazekStolkSymes:13}. However the extent to which (and sense
-in which) convergence
-of finite difference methods might be established with such weak
-regularity assumptions is an open question, beyond the scope of this work.
-
-We analyze the approximation of waves in unbounded material models:
-the spatial domain of wave propagation is the entire space $\bR^d$. We
-consider only initial and right-hand side data of compact support in
-space, so that the solutions of the wave propagation systems also have
-compact support in space, for each time. Extension of our results to
-settings in which waves have nontrivial interaction with boundaries is
-possible in some cases, notably when the boundary conditions are
-conservative or energy-dissipative. However we do not explicitly
-formulate such generalizations, nor do we have anything to say about
-sources located on physical boundaries. An effective extension to free
-surface modeling in elasticity, for instance, is constrained by the
-lack of conservative boundary conditions of order higher than
-two. These are topics for further research.
-
-We should note that numerical tests presented in the last section of
-the paper suggest that, in the case of smooth coefficients, the actual 
-quality of finite difference approximate solutions is better than our 
-convergence result (or even the hypothetical ``optimal'') result
-mentioned above) indicates: in some cases, away from the source 
-point, we observe point-wise (strong) convergence with optimal 
-order. We will mention this phenomenon again in the discussion 
-section. The convergence behaviour suggested by the numerical examples 
-is local, in regions bounded away from the source point. The 
-convergence result proved here, on the other hand, is global.
-
-We begin this paper with a review of the (rather sparse) prior work on
-finite-difference approximation of linear partial differential 
-equations with singular right-hand sides, emphasizing results for 
-hyperbolic problems. Then we give a detailed discussion of the moment conditions which form the {\em a priori} conditions for accuracy of 
-source approximations. We then explain the convergence theory, based 
-on energy estimates, for hyperbolic systems with singular right-hand 
-sides. We introduce linear acoustics as an example. We have performed 
-a number of numerical experiments that illustrate the results of the 
-theory - we present these next. We conclude with a discussion of
-matters not addressed in the body of this paper, and a summary of the 
-results.
-
-%Numerical results presented here, consistent with other similar works (e.g., \cite{Petersson:2010}), appear to %indicate, however, stronger convergence results: optimal convergence point-wise away from source.
-%In particular, we report second and fourth order rates when studying the spatial convergence of 2-2 and 2-4 %finite difference methods respectively, where the source approximation order matched the spatial order of th%e numerical scheme.
+While finite element methods are also used in earthquake seismic modeling and inversion \cite[]{KomTromp:00,Cohen:01,Ghattas:IP25}, this paper focuses on regular (rectangular) grid finite difference methods, in particular the {\em staggered-grid} variety, which are widely used for basin- and exploration-scale modeling; see \cite{moczoetal:06} for an excellent overview and many older references.
+Such methods pose an immediate problem for singular source models such as multipoles: finite
+difference algorithms ``know'' only gridded fields, so a source located at an arbitrary point $\mathbf x^*$ in space must be represented somehow by virtual sources at nearby grid points. 
+This task is complicated by the nature of the field, as is evident for instance from inspection of the 3-D analytical function \ref{eq:green3d}: solutions of the acoustic equations with singular right-hand sides are generally themselves singular at the source point, so the Taylor-series based analysis of finite difference accuracy does not apply.
+%Imitating finite-element singular source representation, for example by using adjoint interpolation, may reduce the accuracy of the modeled field, in the sense of convergence order: visually, large errors can pollute the field near the source, and propagate throughout the simulation domain.
+
+The focus of this paper is to study singular source discretizations, particularly in the context of staggered-grid finite difference solvers for a family of wave propagation differential equations (e.g., linear acoustics and elasticity in first-order form) with multipole sources.
+Specifically, we are interested in the following class of symmetric hyperbolic systems of first-order partial differential equations: find vector-valued fields $(u,v)$ such that
+\begin{equation}\label{eq:hyper_fam}
+\begin{split}
+	A \frac{\partial}{\partial t}u + P^T v &= f \\
+	B \frac{\partial}{\partial t}v  - P u & = g
+\end{split}
+\end{equation}
+for given source terms $(f,g)$.
+Operators $A$ and $B$ are bounded symmetric matrix-valued and represent medium parameters.
+$P$ is a first-order spatial differential operator with constant coefficients, e.g., $P$ is the gradient in the acoustic equations \ref{eq:acoustics} with $P^T$ being the negative of the divergence.
+We directly discretize multipole sources using {\em discrete moment conditions} as developed by \cite{Walden:1999}, and elaborated by \cite{TorEng:04} for multi-dimensional approximations.
+Let $(u_h,v_h)$ denote the finite difference solution to \ref{eq:hyper_fam} via staggered-grid finite differences, with discretized source terms $(f_h,g_h)$, where $h>0$ denotes the grid size.
+The class of staggered-grid finite difference methods we deal with are second-order in time and $2p$-order in space accurate, and we refer to such a scheme as being of order $2$-$2p$.
+
+Our main contribution is a weak convergence theory, a novel approach to the singular source approximation problem, that is applicable to \ref{eq:hyper_fam} in that $(u_h,v_h)$ converge to $(u,v)$ given that source approximations $(f_h,g_h)$ converge to $(f,g)$ as $h\to 0$, both in a weak sense which we make clear in the theory section.
+In particular, we show
+\begin{equation}\label{eq:result}
+	\langle (u,v),(\tilde f,\tilde g) \rangle - \langle (u_h,v_h), (\tilde f,\tilde g)\rangle = O(h^q + \Delta t^2 + (h^{2p}+\Delta t)h^{-N^*-d/2})
+\end{equation}
+given smooth test functions $(\tilde f,\tilde g)$. 
+Inner products $\langle\cdot,\cdot\rangle$ are interpreted as standard $L^2$ space-time inner products in a continuum or discrete sense depending on their arguments.
+Estimate \ref{eq:result} thus depends on singular source approximation order $q$, spatial half-order $p$ for a staggered-grid finite difference scheme of order $2$-$2p$, the maximum multipole order between $f$ and $g$ denoted by $N^*$, and spatial dimension $d$.
+Numerical results presented here, consistent with other similar works (e.g., \cite{Petersson:2010}), appear to indicate, however, stronger convergence results: optimal convergence point-wise away from source.
+In particular, we report second and fourth order rates when studying the spatial convergence of 2-2 and 2-4 finite difference methods respectively, where the source approximation order matched the spatial order of the numerical scheme.
 
@@
 
-\subsection{Literature Review}
+\subsection{Literature Review} %ADDED
 
@@
 \end{equation}
-in which $\mathcal L$ is a differential operator and the right-hand
-side $f$ may be singular.
+for some differential operator $\mathcal L$ and singular source term $f$.
 Define the regularization of problem \ref{eq:pde} by replacing $f$ with some regular (at least piecewise continuous) function $f^\epsilon$ parameterized by regularization parameter $\epsilon>0$, that is, find $u^\epsilon$ such that
@@
 Regularized source term $f^\epsilon$ is said to approximate $f$ in that $f^\epsilon \to f$ as $\epsilon \to 0$ in some sense.
-The end goal is of course to have $u^\epsilon$ approximate the continuum solution $u$, i.e.,
+The end goal is of course to have $u^\epsilon$ approximate true solution $u$, i.e.,
 \[
@@
 
-Suppose that the regularized problem \ref{eq:pde_reg} is discretized, with mesh or cell size $h>0$;
+Suppose that the regularized problem \ref{eq:pde_reg} is discretized with mesh or cell size $h>0$;
 \begin{equation}\label{eq:pde_disc}
@@
 Green's functions for hyperbolic problems are singular at the wavefront they propagate as well as at the source location, thus the analysis of \cite{TorEng:04} does not apply.
-\cite{Petersson:2016} studied a finite difference approximations to various
-advection systems, and to the linear acoustodynamics, with singular
-source terms. based on centered difference approximations and accurate
-Runge-Kutta time differentiation. An plane wave analysis revealed that
-both moment conditions of the type discussed here, and additional {\em smoothness}
-constraints on the source representation
+Recent work by \cite{Petersson:2016} presented some of the most relevant analysis based on centered difference approximations to the 1-D advection equation with singular source term.
 They show that the discrete moment conditions are necessary but not sufficient for the convergence of numerical solutions at optimal rates away from the source location.
-They demonstrated that {\emph smoothness constraints} on the numerical
-solution are also required to achieve convergence, due to the presence
-of spurious modes injected by the singular source approximation. 
+They demonstrated that {\emph smoothness conditions} are also required to achieve convergence; these conditions are based on Fourier analysis and the ability of the finite difference operators at play to resolve spurious modes injected by the singular source approximation.
 %The main result of \cite{Petersson:2016} is however suboptimal, in that
@@
 %%%
-%Our main contribution is a weak convergence theory of
-% $u^\epsilon_h\to u$ that is, unlike current convergence analysis
-% that is limited by its particular application, applicable to a large
-% set of symmetric hyperbolic problems solved via staggered grid
-% finite difference methods.
-%%Though numerical results, both reported in other works and in this
-%% paper, demonstrate stronger results, mainly point-wise convergence
-%% away from source, we believe a weak convergence error estimate is
-%% an appropriate first step towards a complete theory.
-%The remainder of the paper is organized as follows: In the theory
-% section we begin by presenting an overview of the singular source
-% approximation method via continuum and discrete moment conditions.
-% perhaps include regularity error estimates for 3-D acoustics? We
-% focus primarily on source approximations of narrow support, as
-% discussed in \cite{TorEng:04}, though we provide explicit formulas
-% for approximations of arbitrary order. Moreover, we show that the
-% discrete moment conditions in fact define a sequence of continuum
-% functions that converge to target distributions in a weak sense, a
-% new result. The theory section also covers known convergence results
-% of staggered-grid finite difference methods via energy estimates,
-% applied to our set of differential equations under smooth
-% coefficients and smooth source terms. At this point we present our
-% weak convergence theory. The last section covers numerical results,
-% mainly for staggered-grid finite difference solutions to the 2-D
-% acoustic equations in first-order form with multipole
-% sources. Consistent with numerical results presented in the
-% literature (see for example \cite{Petersson:2010}), we observe
-% optimal convergence rates of our numerical solutions when the source
-% discretization satisfied the proper number of moment conditions.
+%Our main contribution is a weak convergence theory of $u^\epsilon_h\to u$ that is, unlike current convergence analysis that is limited by its particular application, applicable to a large set of symmetric hyperbolic problems solved via staggered grid finite difference methods.
+%%Though numerical results, both reported in other works and in this paper, demonstrate stronger results, mainly point-wise convergence away from source, we believe a weak convergence error estimate is an appropriate first step towards a complete theory.
+The remainder of the paper is organized as follows:
+In the theory section we begin by presenting an overview of the singular source approximation method via continuum and discrete moment conditions.
+%perhaps include regularity error estimates for 3-D acoustics?
+We focus primarily on source approximations of narrow support, as discussed in \cite{TorEng:04}, though we provide explicit formulas for approximations of arbitrary order.
+Moreover, we show that the discrete moment conditions in fact define a sequence of continuum functions that converge to target distributions in a weak sense, a new result.
+The theory section also covers known convergence results of staggered-grid finite difference methods via energy estimates, applied to our set of differential equations under smooth coefficients and smooth source terms.
+At this point we present our weak convergence theory.
+The last section covers numerical results, mainly for staggered-grid finite difference solutions to the 2-D acoustic equations in first-order form with multipole sources.
+Consistent with numerical results presented in the literature (see for example \cite{Petersson:2010}), we observe optimal convergence rates of our numerical solutions when the source discretization satisfied the proper number of moment conditions.
 
@@
 
-The weak convergence theory of finite difference solutions to hyperbolic
-systems with multipole sources rests on two foundations: (i) the
-analysis of weak convergence conditions for gridded approximations to
-point support distributions, and (ii) the convergence theory for
-smooth solutions with smooth right-hand sides. The next two
-subsections develop these foundations.
-
 %%%%
@@
 %
-
-Let $\mathcal D$ denote the {\em space of test functions} over $\mathbf R^d$, that is, the space of $C^\infty_0(\mathbf R^d)$ endowed with the standard topology of test functions.
-The set of \emph{distributions} is the topological dual of the space
-of test functions, and is commonly denoted by $\mathcal D'$.
+We begin by noting that the delta function and its derivatives (and thus multipoles) are not actually functions but rather so-called distributions, functionals that return a real number when applied to a test function.
+Let $\mathcal D$ denote the {\em space of test functions} over $\mathbb R^d$, that is, the space of $C^\infty_0(\mathbb R^d)$ endowed with the standard topology of test functions.
+The set of \emph{distributions} is given by the dual of the space of test functions, denoted by $\mathcal D'$.
 It is conventional to represent the application of a distribution on a function by the integral of the product, even when the distribution is not actually a function that can be integrated in the usual sense.
-For example, given multi-index $\mathbf s=(s_1,...,s_d)$, the $\mathbf s$-mixed partial derivative of the Dirac delta function, shifted by $\mathbf x^*\in\mathbf R^d$, is defined by
-\[
-	\int_{\mathbf R^d} D^{\mathbf s} \delta(\mathbf x-\mathbf x^*) \; \psi(\mathbf x)\; d\mathbf x =
+For example, given multi-index $\mathbf s=(s_1,...,s_d)$, the $\mathbf s$-mixed partial derivative of the Dirac delta function, shifted by $\mathbf x^*\in\mathbb R^d$, is defined by
+\[
+	\int_{\mathbb R^d} D^{\mathbf s} \delta(\mathbf x-\mathbf x^*) \; \psi(\mathbf x)\; d\mathbf x =
 	(-1)^{|\mathbf s|} D^{\mathbf s} \psi(\mathbf x^*), \quad \forall \psi \in\mathcal D.
@@
 \[
-	\int_{\mathbf R^d} D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)\; \psi(\mathbf x) \; d\mathbf x = \mathbf s! (-1)^{|\mathbf s|} \delta_{\mathbf s \alpha}
+	\int_{\mathbb R^d} D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)\; \psi(\mathbf x) \; d\mathbf x = \mathbf s! (-1)^{|\mathbf s|} \delta_{\mathbf s \alpha}
 \]
@@
 %
-Given $\eta\in L^1_0(\mathbf R^d)$ (i.e., integrable function of compact support) and multi-index $\alpha$, the {\em $\alpha$-moment} of $\eta$ centered at $\mathbf x^*\in\mathbf R^d$, denoted by $M^\alpha(\cdot,\mathbf x^*)$, is defined as
+Given $\eta\in L^1_0(\mathbb R^d)$ (i.e., integrable function of compact support) and multi-index $\alpha$, the {\em $\alpha$-moment} of $\eta$ centered at $\mathbf x^*\in\mathbb R^d$, denoted by $M^\alpha(\cdot,\mathbf x^*)$, is defined as
 %definition of moment
 \begin{equation}\label{eq:def_mom}
-	M^\alpha(\eta,\mathbf x^*) := \int_{\mathbf R^d} \eta(\mathbf x) \; (\mathbf x-\mathbf x^*)^{\alpha} \; d\mathbf x.
+	M^\alpha(\eta,\mathbf x^*) := \int_{\mathbb R^d} \eta(\mathbf x) \; (\mathbf x-\mathbf x^*)^{\alpha} \; d\mathbf x.
 \end{equation}
 %Note that integration in equation \ref{eq:def_mom} is translation invariant hence $M_{\alpha}(\eta,\mathbf x^*)$ is constant with respect to $\mathbf x^*$, which I denote by $M_{\alpha}(\eta)$.
-For given nonnegative integer $q$ and multi-index $\mathbf s$, the function $\eta$ is said to satisfy the {\em continuum $(q,\mathbf s)$-moment conditions} at $\mathbf x^*\in \mathbf R^d$ if
+For given nonnegative integer $q$ and multi-index $\mathbf s$, the function $\eta$ is said to satisfy the {\em continuum $(q,\mathbf s)$-moment conditions} at $\mathbf x^*\in \mathbb R^d$ if
 %definition of moment conditions
@@
 \[
-	\int_{\mathbf R^d} \eta(\mathbf x) \; \psi(\mathbf x)\; d\mathbf x, \quad \forall \psi\in\mathcal D,
+	\int_{\mathbb R^d} \eta(\mathbf x) \; \psi(\mathbf x)\; d\mathbf x, \quad \forall \psi\in\mathcal D,
 \]	
 is an approximation to $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$ in that it is exact on polynomials of order $q+|\mathbf s|-1$.
-
 The following theorem states that a sequence of (regular) distributions of compact support, satisfying the $(q,\mathbf s)$-moment conditions, will converge in the weak-$*$ topology at a rate $q$ to the target distribution as the width of the supports approach zero.
@@
 \begin{theorem}\label{thm:weakconv}
-	Let nonnegative integer $q$, multi-index $\mathbf s$, and $\mathbf x^*\in\mathbf R^d$ be given.
-	Suppose $\{\eta^\epsilon\}\subset L^1_{0}(\mathbf R^d)$ is a sequence of functions as $\epsilon\to 0$, where $supp(\eta^\epsilon)\subset B(\mathbf x^*,\epsilon)$.
+	Let nonnegative integer $q$, multi-index $\mathbf s$, and $\mathbf x^*\in\mathbb R^d$ be given.
+	Suppose $\{\eta^\epsilon\}\subset L^1_{0}(\mathbb R^d)$ is a sequence of functions as $\epsilon\to 0$, where $supp(\eta^\epsilon)\subset B(\mathbf x^*,\epsilon)$.
 	Furthermore, suppose that there exists a constant $K>0$ independent of $\epsilon$ such that 
 	\begin{equation}\label{eq:boundH}
-		\int_{\mathbf R^d} |\eta^\epsilon(\mathbf x)|\; |(\mathbf x-\mathbf x^*)^{\alpha}| \; d\mathbf x \le K, \quad \forall |\alpha| = |\mathbf s|.
+		\int_{\mathbb R^d} |\eta^\epsilon(\mathbf x)|\; |(\mathbf x-\mathbf x^*)^{\alpha}| \; d\mathbf x \le K, \quad \forall |\alpha| = |\mathbf s|.
 	\end{equation}
@@
 	\begin{equation}\label{eq:delta_error}
-		E:= \left| \int_{\mathbf R^d} D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)\;\psi(\mathbf x) \; d\mathbf x\; - 
-		\int_{\mathbf R^d} \eta^\epsilon(\mathbf x) \; \psi(\mathbf x)\; d\mathbf x \right| = O(\epsilon^q).
+		E:= \left| \int_{\mathbb R^d} D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)\;\psi(\mathbf x) \; d\mathbf x\; - 
+		\int_{\mathbb R^d} \eta^\epsilon(\mathbf x) \; \psi(\mathbf x)\; d\mathbf x \right| = O(\epsilon^q).
 	\end{equation}
@@
 \begin{align*} 
-	\int_{\mathbf R^d} \eta^\epsilon (\mathbf x) \; \psi(\mathbf x) \; d\mathbf x
-	&= \int_{\mathbf R^d} \eta^\epsilon(\mathbf x) \left( \sum_{|\alpha|=0}^{N} \frac{D^{\alpha} \psi(\mathbf x^*)}{\alpha!} \; (\mathbf x - \mathbf x^*)^\alpha 
+	\int_{\mathbb R^d} \eta^\epsilon (\mathbf x) \; \psi(\mathbf x) \; d\mathbf x
+	&= \int_{\mathbb R^d} \eta^\epsilon(\mathbf x) \left( \sum_{|\alpha|=0}^{N} \frac{D^{\alpha} \psi(\mathbf x^*)}{\alpha!} \; (\mathbf x - \mathbf x^*)^\alpha 
 	+ \sum_{|\beta|=N+1}R_{\beta}(\mathbf x) \; (\mathbf x - \mathbf x^*)^\beta \right)
 	d\mathbf x \\
-	&= \sum_{|\alpha|=0}^N \frac{D^{\alpha}\psi(\mathbf x^*)}{\alpha!}  \left( \int_{\mathbf R^d}  \eta^\epsilon(\mathbf x) \; (\mathbf x -\mathbf x^*)^\alpha \;d\mathbf x\right) 
-	+  \sum_{|\beta|=N+1} \int_{\mathbf R^d} \eta^{\epsilon}(\mathbf x) R_{\beta}(\mathbf x) \; (\mathbf x-\mathbf x^*)^\beta \; d\mathbf x,
+	&= \sum_{|\alpha|=0}^N \frac{D^{\alpha}\psi(\mathbf x^*)}{\alpha!}  \left( \int_{\mathbb R^d}  \eta^\epsilon(\mathbf x) \; (\mathbf x -\mathbf x^*)^\alpha \;d\mathbf x\right) 
+	+  \sum_{|\beta|=N+1} \int_{\mathbb R^d} \eta^{\epsilon}(\mathbf x) R_{\beta}(\mathbf x) \; (\mathbf x-\mathbf x^*)^\beta \; d\mathbf x,
 \end{align*}
@@
 \begin{align*}
-	\int_{\mathbf R^d} \eta^\epsilon(\mathbf x) \; \psi(\mathbf x)\;d\mathbf x
+	\int_{\mathbb R^d} \eta^\epsilon(\mathbf x) \; \psi(\mathbf x)\;d\mathbf x
 	&= \sum_{|\alpha|=0}^{N} \frac{1}{\alpha!} D^{\alpha}\psi(\mathbf x^*) \Big( \mathbf s! (-1)^{|\mathbf s|}\delta_{\mathbf s \alpha} \Big) 
-	+  \sum_{|\beta|=N+1} \int_{\mathbf R^d} \eta^\epsilon(\mathbf x) R_{\beta}(\mathbf x)\; (\mathbf x-\mathbf x^*)^\beta \; d\mathbf x\\
+	+  \sum_{|\beta|=N+1} \int_{\mathbb R^d} \eta^\epsilon(\mathbf x) R_{\beta}(\mathbf x)\; (\mathbf x-\mathbf x^*)^\beta \; d\mathbf x\\
 	&= (-1)^{|\mathbf s|}D^{\mathbf s} \psi(\mathbf x^*) 
-	+  \sum_{|\beta|=N+1} \int_{\mathbf R^d} \eta^\epsilon (\mathbf x) R_{\beta}(\mathbf x) \; (\mathbf x-\mathbf x^*)^\beta \; d\mathbf x.
+	+  \sum_{|\beta|=N+1} \int_{\mathbb R^d} \eta^\epsilon (\mathbf x) R_{\beta}(\mathbf x) \; (\mathbf x-\mathbf x^*)^\beta \; d\mathbf x.
 \end{align*}
@@
 
-Define the regular grid family $\mathcal G_h(\mathbf x_0,\mathbf l)$
-with mesh or scale $h$, offset $\bx_0 \in
-\bR^d$ and cell given by ${\bf l} \in \bR_+^d$, namely $\{\bx \in
-\bR^d: 0 \le x_i\le l_k\}$:
-\[
-	\mathcal G_h(\mathbf x_0,\mathbf l) = \{\bx_{\bf n} =
-        h((x_0)_1+n_1l_1,...,(x_0)_d+n_dl_d), {\bf n} \in {\bf Z}^d\}
-\]
-We will sometimes refer to the scaled cell side lengths $h_k=hl_k,k=1,...,d$.
-%The dependence of real-valued grid functions with respect to a given regular grid $\mathcal G_h(\mathbf x_0,\mathbf h)$, say $\eta^h: \mathcal G(\mathbf x_0,\mathbf h)\to \mathbf R$, is made implicit through the use of multi-indexing notation.
+We define the regular grid centered at $\mathbf x_0\in\mathbb R^d$ with cell size $\mathbf h = (h_1,...,h_d)>\mathbf 0$, as the collection of points denoted by $\mathcal G(\mathbf x_0,\mathbf h)$,
+\[
+	\mathcal G(\mathbf x_0,\mathbf h) = \{ \mathbf x_n = (x_{1,n_1},...,x_{d,n_d}), \; \mathbf n \in\mathbb Z^d\},
+\]
+where
+\[
+	x_{k,n_k} = x_{0,k} + h_kn_k, \quad \forall k=1,...,d.
+\]
+%The dependence of real-valued grid functions with respect to a given regular grid $\mathcal G(\mathbf x_0,\mathbf h)$, say $\eta^h: \mathcal G(\mathbf x_0,\mathbf h)\to \mathbb R$, is made implicit through the use of multi-indexing notation.
 %For example, given $\mathbf x_{\mathbf n}\in \mathcal G(\mathbf x_0,\mathbf h)$ for some multi-index $\mathbf n=(n_1,...,n_d)$,
@@
 %\]
-Note that there is no reason to assume that the grid cell is cubical: it may have different lengths along different axes.
-
-The obvious definition of the \emph{discrete $\alpha$-moment}, centered at $\mathbf x^*\in\mathbf R^d$ (note that $\mathbf x^*$ need not coincide with a grid point in $\mathcal G_h(\mathbf x_0,\mathbf l)$), of a grid function $\eta_h:\mathcal G_h(\mathbf x_0,\mathbf l)\to \mathbf R$ is given as follows:
+Note that there is no reason to assume that the grid cell is cubical, it may have different lengths along different axes.
+However, we assume that the grid is refined by scaling a characteristic grid cell size $h$ (e.g., $h=\max_{k} h_k$) and which we will use to denote grid functions and other grid-dependent quantities with subscript $h$ even in dimensions higher than one.
+
+The obvious definition of the \emph{discrete $\alpha$-moment}, centered at $\mathbf x^*\in\mathbb R^d$ (note that $\mathbf x^*$ need not coincide with a grid point in $\mathcal G(\mathbf x_0,\mathbf h)$), of a grid function $\eta_h:\mathcal G(\mathbf x_0,\mathbf h)\to \mathbb R$ is given as follows:
 \begin{equation*}\label{eq:def_discmom}
-        M_h^\alpha(\eta_h,\mathbf x^*) := \left(\prod_{k=1}^{d} h_k \right) \sum_{\mathbf x\in\mathcal G_h(\mathbf x_0,\mathbf l)}  \eta_h(\mathbf x)\; ({\bf x}-{\bf x}^*)^{\alpha}.
+        M_h^\alpha(\eta_h,\mathbf x^*) := \left(\prod_{k=1}^{d} h_k \right) \sum_{\mathbf x\in\mathcal G(\mathbf x_0,\mathbf h)}  \eta_h(\mathbf x)\; ({\bf x}-{\bf x}^*)^{\alpha}.
 \end{equation*}
 It is worth pointing out that the discrete moment defined above is dependent on choice of grid, in particular dependent on the source location $\mathbf x^*$ relative to the grid.
-Similar to the continuum moment conditions (equation \ref{eq:momcond}), grid function $\eta_h$ is said to satisfy the \emph{discrete $(q,\mathbf s)$-moment conditions} at $\mathbf x^*\in\mathbf R^d$ if
+Similar to the continuum moment conditions (equation \ref{eq:momcond}), grid function $\eta_h$ is said to satisfy the \emph{discrete $(q,\mathbf s)$-moment conditions} at $\mathbf x^*\in\mathbb R^d$ if
 \begin{equation}\label{eq:discmomcond}
@@
 \begin{theorem}\label{thm:discweakconv}
-Let nonnegative integer $q$, multi-index $\mathbf s$, and $\mathbf x^*\in\mathbf R^d$ be given.
-Suppose $\{\eta_h^\epsilon\}$ is a sequence of grid functions $\eta_h^\epsilon:\mathcal G_h(\mathbf x_0,\mathbf l)\to\mathbf R$ as $\epsilon\to 0$.
+Let nonnegative integer $q$, multi-index $\mathbf s$, and $\mathbf x^*\in\mathbb R^d$ be given.
+Suppose $\{\eta_h^\epsilon\}$ is a sequence of grid functions $\eta_h^\epsilon:\mathcal G(\mathbf x_0,\mathbf h)\to\mathbb R$ as $\epsilon\to 0$.
 Furthermore, assume that the support of $\eta_h^\epsilon$ is contained in $B(\mathbf x^*,\epsilon)$ with $\epsilon=O(h)$, and that there exists constant $K>0$ independent of $\epsilon$ such that
 \begin{equation*}\label{eq:boundh}
-	\left(\prod_{k=1}^d h_k \right) \sum_{\mathbf x\in\mathcal G_h(\mathbf x_0,\mathbf l)} |\eta_h^\epsilon(\mathbf x)|\; |(\mathbf x-\mathbf x^*)^{\alpha}| \le K, \quad \forall |\alpha| = |\mathbf s|.
+	\left(\prod_{k=1}^d h_k \right) \sum_{\mathbf x\in\mathcal G(\mathbf x_0,\mathbf h)} |\eta_h^\epsilon(\mathbf x)|\; |(\mathbf x-\mathbf x^*)^{\alpha}| \le K, \quad \forall |\alpha| = |\mathbf s|.
 \end{equation*}
@@
 \[
-	\left| \int_{\mathbf R^d} D^{\mathbf s}\delta(\mathbf x- \mathbf x^*) \;\psi(\mathbf x)\; d\mathbf x - \left(\prod_{k=1}^d h_k\right) \sum_{\mathbf x\in\mathcal G_h(\mathbf x_0,\mathbf l)}\eta_h^\epsilon (\mathbf x) \; \psi(\mathbf x)\right| = O(h^q).
+	\left| \int_{\mathbb R^d} D^{\mathbf s}\delta(\mathbf x- \mathbf x^*) \;\psi(\mathbf x)\; d\mathbf x - \left(\prod_{k=1}^d h_k\right) \sum_{\mathbf x\in\mathcal G(\mathbf x_0,\mathbf h)}\eta_h^\epsilon (\mathbf x) \; \psi(\mathbf x)\right| = O(h^q).
 \]
@@
 
-We make the choice of having 1-D gridded approximations be centered around source location $x^*\in\mathbf R$, and define them to be zero outside the interval $[-\epsilon+x^*,\epsilon+x^*)$, with $2\epsilon=Nh$ for some positive integer $N$.
-In other words, there exists $N$ grid points, denoted by $\{\tilde x_\ell\}_{\ell=1}^{N}$, such that they are contained in the interval $[-\epsilon+x^*,\epsilon+x^*)$ for a given grid $\mathcal G_h(x_0,1)$\footnote{We assume $l=1$ for simplicity, since any scaling effects $l$ may have can be incorporated into $h$ in 1-D}.
+We make the choice of having 1-D gridded approximations be centered at source location $x^*\in\mathbb R^d$, and define them to be zero outside the interval $[-\epsilon+x^*,\epsilon+x^*)$, with $2\epsilon=Nh$ for some positive integer $N$.
+In other words, there exists $N$ grid points, denoted by $\{\tilde x_\ell\}_{\ell=1}^{N}$, such that they are contained in the interval $[-\epsilon+x^*,\epsilon+x^*)$ for a given grid $\mathcal G(x_0,h)$.
 These grid points $\{\tilde x_\ell\}_{\ell=1}^N$ are referred to as the \emph{stencil points} of the approximation.
@@
 for $\ell=1,...,N$ and $k=1,...,q+s$.
-Note that $\mathbf A$ is a Vandermonde matrix of full rank and is guaranteed a solution if $N\ge q+s$ and no solution for $N<q+s$ under general $x^*\in\mathbf R$.
+Note that $\mathbf A$ is a Vandermonde matrix of full rank and is guaranteed a solution if $N\ge q+s$ and no solution for $N<q+s$ under general $x^*\in\mathbb R$.
 %It will be of benefit to pick grid functions of minimal support, that is $N=q+s$; this will be more apparent when coupling source approximations with finite difference schemes.
@@
 We choose the case where $N=q+s$, which we refer to as grid functions of narrow support.
-The system above will result in a unique solution for a given $x^*\in\mathbf R$.
+The system above will result in a unique solution for a given $x^*\in\mathbb R$.
 In fact the inverse matrix for $\mathbf A$ can be written explicitly using the following Vandermonde matrix inverse formula:
@@
 %Intuitively enough, the same cannot be said of the continuum case. 
-%In other words, if $\eta(x)$ is a $q$-order approximation of $D^{s}\delta(x)$ then it follows that $\eta(x-x^*)$ is also a $q$-order approximation of $D^{s}\delta(x-x^*)$, for all $x^*\in\mathbf R$.
-Given that equation \ref{eq:gridfunsol} is unique for a particular source location, we show that imposing the discrete moment conditions over all $x^*\in\mathbf R$ indeed defines a function over the reals.
+%In other words, if $\eta(x)$ is a $q$-order approximation of $D^{s}\delta(x)$ then it follows that $\eta(x-x^*)$ is also a $q$-order approximation of $D^{s}\delta(x-x^*)$, for all $x^*\in\mathbb R$.
+Given that equation \ref{eq:gridfunsol} is unique for a particular source location, we show that imposing the discrete moment conditions over all $x^*\in\mathbb R$ indeed defines a function over the reals.
 Moreover, we show that these continuum functions satisfy the continuum moment conditions and thus define a sequence of distributions that converge to $D^{s}\delta$ in the weak-$*$ topology, a result previously not known.
 
-%Equation \ref{eq:gridfunsol} is used to generate gridded functions $\eta^{h_k}$ satisfying the discrete $(q,s_k)$-moment conditions at $x^*_k\in\mathbf R$ on each dimension $k=1,...,d$.
+%Equation \ref{eq:gridfunsol} is used to generate gridded functions $\eta^{h_k}$ satisfying the discrete $(q,s_k)$-moment conditions at $x^*_k\in\mathbb R$ on each dimension $k=1,...,d$.
 %Approximations on higher dimension are then constructed by taking the tensor product of the 1-D approximations.
@@
 %\begin{theorem}\label{thm:prodmomcond}
-%	Let $q\in\mathbf N$, multi-index $\mathbf s$, and $\mathbf x^*\in\mathbf R^d$ be given. 
-%Suppose $\eta^h:\mathcal G(\mathbf x_0,\mathbf h)\to \mathbf R$ is a multi-variate grid function given by the tensor product of 1-D grid functions $\eta^{h_k}:\mathcal G(x_{0,k},h_k)\to\mathbf R$.
+%	Let $q\in\mathbb N$, multi-index $\mathbf s$, and $\mathbf x^*\in\mathbb R^d$ be given. 
+%Suppose $\eta^h:\mathcal G(\mathbf x_0,\mathbf h)\to \mathbb R$ is a multi-variate grid function given by the tensor product of 1-D grid functions $\eta^{h_k}:\mathcal G(x_{0,k},h_k)\to\mathbb R$.
 %If $\eta^{h_k}$ satisfy the discrete $(q,s_k)$-moment conditions at $x^*_k$ for all $k=1,...,d$, then it follows that $\eta^h$ satisfies the discrete $(q,\mathbf s)$-moment conditions at $\mathbf x^*$.
@@
 %Same applies for $\alpha=\mathbf s$.
-%Suppose then, that there exists index $\ell$ such that $\alpha_\ell>q+s_\ell-1$, that is $a_\ell=q+s_\ell -1 +i$ for some $i\in\mathbf N$. Thus,
+%Suppose then, that there exists index $\ell$ such that $\alpha_\ell>q+s_\ell-1$, that is $a_\ell=q+s_\ell -1 +i$ for some $i\in\mathbb N$. Thus,
 %\begin{align*}
@@
 We first focus on the $x^*=0$ case and define our continuum approximation $\eta^\epsilon$ to be zero outside $[-\epsilon,\epsilon)$, with $2\epsilon=Nh$ for $N=q+s$. 
-Furthermore, we define $\eta^\epsilon$ to be piecewise polynomial over $N$ \hl{intervals}: 
+Furthermore, we define $\eta^\epsilon$ to be piecewise polynomial over $N$ \hl{intervals}: %MOD
 %ETA
@@
 where $P_\ell$ is some polynomial over the considered interval, and $a_{\ell}=-\epsilon+ (\ell-1)h$ for $\ell=1,...,N+1$.
-Given the support of our approximation, and a regular grid $\mathcal G_h(x_0,1)$, it follows that there are $N$ grid points contained in the interval $[-\epsilon,\epsilon)$, again denoted by $\{\tilde x_\ell\}_{\ell=1}^N$.
+Given the support of our approximation, and a regular grid $\mathcal G(x_0,h)$, it follows that there are $N$ grid points contained in the interval $[-\epsilon,\epsilon)$, again denoted by $\{\tilde x_\ell\}_{\ell=1}^N$.
 In fact,
@@
 \begin{theorem}\label{thm:DCmomcond}
-	Let nonnegative integer $q$, positive integer $s$, and $x^*\in \mathbf R$ be given. 
+	Let nonnegative integer $q$, positive integer $s$, and $x^*\in \mathbb R$ be given. 
 	Suppose $\eta^\epsilon$ is constructed according to equations \ref{eq:eta} and \ref{eq:poly} for a given $h>0$.
@@
 	\begin{align*}
-		M^\alpha(\eta^\epsilon,0) &= \int_{\mathbf R} \eta^\epsilon(x) \;x^\alpha\; dx \\
+		M^\alpha(\eta^\epsilon,0) &= \int_{\mathbb R} \eta^\epsilon(x) \;x^\alpha\; dx \\
 			&= \sum_{\ell=1}^N \int_{a_\ell}^{a_{\ell+1}} P_\ell(x) \;x^\alpha \; dx.
@@
 	\end{subequations}
-	Note that the term in the bracket coincides with the discrete $\alpha$-moment of $\eta^\epsilon$ with respect to a uniform grid $\mathcal G_h(a_1,1)$ (containing stencil points $\{a_\ell\}_{\ell=1}^{N+1}$) for a source located at $x^*=-\xi$.
+	Note that the term in the bracket coincides with the discrete $\alpha$-moment of $\eta^\epsilon$ with respect to a uniform grid $\mathcal G(a_1,h)$ (containing stencil points $\{a_\ell\}_{\ell=1}^{N+1}$) for a source located at $x^*=-\xi$.
 	In other words,
@@
 	\]
-	where $\eta_h^\epsilon:\mathcal G_h(a_1,1)\to \mathbf R$ defined by $\eta_h^\epsilon(x) = \eta^\epsilon(x+\xi)$ satisfies the discrete $(q,s)$-moment conditions at $-\xi$ by construction.
+	where $\eta_h^\epsilon:\mathcal G(a_1,h)\to \mathbb R$ defined by $\eta_h^\epsilon(x) = \eta^\epsilon(x+\xi)$ satisfies the discrete $(q,s)$-moment conditions at $-\xi$ by construction.
 	We can conclude
@@
 We construct approximations in general $d$-dimension by taking tensor products of 1-D approximations, similar to work by \cite{TorEng:04} for $\mathbf s=\mathbf 0$.
-Namely, given approximation order $q$ and multi-index $\mathbf s$, multivariate continuum approximation $\eta:\mathbf R^d\to \mathbf R$ is given by
+Namely, given approximation order $q$ and multi-index $\mathbf s$, multivariate continuum approximation $\eta:\mathbb R^d\to \mathbb R$ is given by
 \begin{equation}\label{eq:tensor_eta}
@@
 \end{equation}
-where $\eta^k:\mathbf R\to \mathbf R$ is a continuum function over the $k$-th axis as given by equations \ref{eq:eta} and \ref{eq:poly}, satisfying the $(q,s_k)$-moment conditions for each $k=1,...,d$.
+where $\eta^k:\mathbb R\to \mathbb R$ is a continuum function over the $k$-th axis as given by equations \ref{eq:eta} and \ref{eq:poly}, satisfying the $(q,s_k)$-moment conditions for each $k=1,...,d$.
 It follows that these tensor approximations are indeed approximations to multipoles in the higher spatial dimension.
@@
 \begin{theorem}\label{thm:prodmomcond}
-	Let nonnegative integer $q$, multi-index $\mathbf s$, and $\mathbf x^*\in\mathbf R^d$ be given. 
-Suppose $\eta:\mathbf R^d\to\mathbf R$ is a multi-variate grid function given by the tensor product of 1-D approximations $\eta^{k}:\mathbf R\to\mathbf R$.
+	Let nonnegative integer $q$, multi-index $\mathbf s$, and $\mathbf x^*\in\mathbb R^d$ be given. 
+Suppose $\eta:\mathbb R^d\to\mathbb R$ is a multi-variate grid function given by the tensor product of 1-D approximations $\eta^{k}:\mathbb R\to\mathbb R$.
 If $\eta^{k}$ satisfy the discrete $(q,s_k)$-moment conditions at $x^*_k$ for each $k=1,...,d$, then it follows that $\eta$ satisfies the discrete $(q,\mathbf s)$-moment conditions at $\mathbf x^*$.
@@
 Same applies for $\alpha=\mathbf s$.
-Suppose then, that there exists index $\ell$ such that $\alpha_\ell>q+s_\ell-1$, that is $a_\ell=q+s_\ell -1 +i$ for some $i\in\mathbf N$. Thus,
+Suppose then, that there exists index $\ell$ such that $\alpha_\ell>q+s_\ell-1$, that is $a_\ell=q+s_\ell -1 +i$ for some $i\in\mathbb N$. Thus,
 \begin{align*}
@@
 %%%%%%%%%
-\subsection{Convergence of finite difference approximations to smooth solutions}
+\subsection{Convergence Theory of Wave Equation Solutions with
+  Multipole Sources}
 %%%%%%%%%
 
-%We prove weak convergence of finite difference solutions with discrete multipole
-%sources satisfying the discrete moment conditions discussed above. 
-%Convergence theory for smooth solutions is in some sense standard. 
+We prove weak convergence of finite difference solutions with discrete multipole
+sources satisfying the discrete moment conditions discussed above. 
+Convergence theory for smooth solutions is in some sense standard. 
 %However 
-%The easily available results (those based on Lax's Equivalence Theorem) 
-%pertain to $L^2$ convergence for smooth solutions and smooth source terms,
-%which we begin with.
+The easily available results (those based on Lax's Equivalence Theorem) 
+pertain to $L^2$ convergence for smooth solutions and smooth source terms,
+which we begin with.
 %whereas weak convergence of multipole solutions depends on uniform 
@@
 
-With smooth Cauchy data and
-right-hand sides, symmetric hyperbolic systems have unique smooth solutions depending
-continuously in the sense of ${\cal D}$ on the data
-\cite[]{CourHil:62,Lax:PDENotes}. Moreover, restriction of the
-solution to constant-time hyperplanes is well-defined, and if the data has compact
-support, then the restrictionts have compact support as well - this is
-the finite propagation speed property.
-
-Finite difference approximation to smooth solutions of hyperbolic PDE has been
-extensively studied \cite[]{RichMor:67,Cohen:01,Leveque:07}. Rather
-than quote results from the literature, however, we prove suitable
-convergence results for staggered grid approximation to systems of the
-form \ref{eq:hyper_fam} directly, by introducing a type of energy estimate
-appropriate to such approximations. This technique of proof will be
-central to our approach to weak convergence for singular solutions,
-developed in the next subsection.
-
-%Theoretical results presented here pertain to the family of symmetric hyperbolic
-%systems of  first-order partial differential equations for vector-valued fields $(u,v)$,
-%defined over a domain $\Omega\subset \mathbf R^d$ in $d=\{1,2,3\}$ dimension
-%space and some interval in time, given by \ref{eq:hyper_fam} which we restate 
-%with greater detail.
-
-\subsubsection{Smooth and Distribution Solutions}
-
-From here on, we will denote by $k_1, k_2 \in {\bf N}$ the dimensions of
-the two subsystems in \ref{eq:hyper_fam}.  The references cited above
-establish (result which imply) the following theorem.%something missing here?? MJB
-
-\begin{theorem}
-\label{thm:well-posed}
-Suppose
+Theoretical results presented here pertain to the family of symmetric hyperbolic
+systems of  first-order partial differential equations for vector-valued fields $(u,v)$,
+defined over a domain $\Omega\subset \mathbb R^d$ in $d=\{1,2,3\}$ dimension
+space and some interval in time, given by \ref{eq:hyper_fam} which we restate 
+with greater detail.
+Let $k_1\in\mathbb N$ and $k_2\in\mathbb N$ denote the dimension of 
+vector-fields $u$ and $v$ respectively. 
+The systems we are interested in are of the following form:
+\begin{equation}\label{eq:cont_system}
+\begin{split}
+	A(\mathbf x)\frac{\partial}{\partial t}u(\mathbf x,t) + P^T v(\mathbf x,t) 
+		&= f(\mathbf x,t) \\
+	B(\mathbf x)\frac{\partial}{\partial t}v(\mathbf x,t) - Pu(\mathbf x,t) 
+		&= g(\mathbf x,t)
+\end{split}
+\end{equation}
+where
 \begin{itemize}
-\item $f \in C_0^{\infty}(\bR^{d+1},\bR^{k_1})$, $g \in
-  C_0^{\infty}(\bR^{d+1},\bR^{k_2})$;
-\item $u^0 \in C_0^{\infty}(\bR^d,\bR^{k_1}), v^0 \in C_0^{\infty}(\bR^d,\bR^{k_1})$;
-\item $A \in C^{\infty}(\bR^d, \bR^{k_1 \times k_1})$, $B \in
-  C^{\infty}(\bR^d, \bR^{k_2 \times k_2})$ are symmetric for all $\mathbf x\in\Omega$ 
-  and uniformly bounded and positive definite , i.e., there exist constants 
-  $0< A_*\le A^*$ and $0< B_*\le B^*$ for which
-  \[
-    A_*I \le A(\bx) \le A^*I,\,\,   B_*I \le B(\bx) \le B^*I, \quad \forall \mathbf
-    x\in \bR^d;
-  \]
-
-\item $P$ is a constant coefficient, 
+	\item $u\in L^2_{loc}(\mathbb R,\mathcal H_1)$ and 
+		$v\in L^2_{loc}(\mathbb R,\mathcal H_2)$ with Hilbert spaces 
+		$\mathcal H_1, \mathcal H_2$;
+	\item ``coefficient'' operators $A$ and $B$ are $k_1\times k_1$ 
+		and $k_2\times k_2$ matrix-valued functions respectively; 
+		we assume they are symmetric for all $\mathbf x\in\Omega$ 
+		and uniformly-positive definite , i.e., there exists constants 
+		$0< A_*\le A^*$ and $0< B_*\le B^*$ such that 
+	\begin{equation}
+	\begin{split}
+		A_*I \le A(\mathbf x) \le A^*I, \quad \forall \mathbf x\in\Omega,\\
+		B_*I \le B(\mathbf x) \le B^*I, \quad \forall \mathbf x\in\Omega;
+	\end{split}
+	\end{equation}
+
+	\item $P:\mathcal H_1 \to \mathcal H_2$ is a constant coefficient, 
 		first-order differential operator of the form
 		\[
-    Pu = \sum_{j=1}^d P_j \frac{\partial u}{\partial x_j}, \quad P_j\in \bR^{k_2\times k_1},
-  \]
-  for $u\in C^{\infty}(\bR^d,\bR^{k_1})$.
+			Pu = \sum_{j=1}^d P_j \frac{\partial u}{\partial x_j}, \quad P_j\in \mathbb R^{k_2\times k_1},
+		\]
+		%ADDED
+		\hl{satisfying the following skew-adjoint relation}
+		\begin{equation}\label{eq:skew}
+			\langle P u, v \rangle = - \langle u, P^T v\rangle
+		\end{equation}
+		for all $u\in \mathcal H_1$ and $v\in\mathcal H_2$;
+	\item source terms $(f,g)$ are for the time being assumed to be smooth, i.e., 
+		 $f\in L^2_{loc}(\mathbb R,\mathcal H_1)$ 
+		and $g\in L^2_{loc}(\mathbb R,\mathcal H_2)$ ;
+	\item we assume solution $(u,v)$ and source terms $(f,g)$ are causal, i.e., for $t < 0$
+		\[
+			u(\mathbf x,t) = f(\mathbf x,t) = 0, \quad 
+			v(\mathbf x,t) = g(\mathbf x,t) = 0.
+		\]
 \end{itemize}
-Then there exist unique $u \in C^{\infty}(\bR^{d+1},\bR^{k_1})$, $v \in
-  C^{\infty}(\bR^{d+1},\bR^{k_2})$ satisfying the following system,
-  identically in  $t \in \bR, \bx \in \bR^d$:
-\begin{equation}\label{eq:cont_system}
-\begin{split}
-	A(\mathbf x)\frac{\partial}{\partial t}u(\mathbf x,t) + P^T v(\mathbf x,t) 
-		&= f(\mathbf x,t), \\
-	B(\mathbf x)\frac{\partial}{\partial t}v(\mathbf x,t) - Pu(\mathbf x,t) 
-		&= g(\mathbf x,t), \\
-        u(\bx,0) & = u^0(\bx),\\
-        v(\bx,0) & = v^0(\bx), 
-\end{split}
-\end{equation}
-Moreover, for each $t \in \bR$, the restrictions $u(\cdot,t), v(\cdot,t)$ have compact support.
-\end{theorem}
-
-Solution of the system \ref{eq:cont_system} defines a linear map from from
-$u^0,v^0,f,g$ to $u,v$, continuous in the $C_0^{\infty}$ respectively
-$C^{\infty} $ topologies. The restriction of the formal adjoint of
-this map to smooth input will play an important role in the sequel.
-
-It will be sufficient to restrict $u^0=v^0=0$, so that the restricted map
-has inputs $f,g$. A function $f \in C^{\infty}(\bR^{d+1},\bR^k)$ is {\em causal} if there 
-is $t_0 \in \bR$ so that $f(\bx,t)=0$ for $t<t_0, \bx \in
-\bR^d$. Similarly, $f$ is {\em anticausal} if there is $t_0 \in \bR$
-so that $f(\bx,t)=0$ for $t>t_0$.
-
-%should \tilde f and \tilde g be anticausal ?? MJB
-\begin{theorem}\label{thm:adjointstate}
-In addition to the hypotheses of Theorem \ref{thm:well-posed},
-suppose that (i) $u^0=v^0=0$, and (ii)
-$\tilde{f} \in C_0^{\infty}(\bR^{d+1},\bR^{k_1})$, $\tilde{g} \in
-C_0^{\infty}(\bR^{d+1},\bR^{k_2})$. Let $u,v$ be the unique causal
-solution of the system \ref{eq:cont_system}, and $\tilde{u},\tilde{v}$
-the unique anticausal solution of the system
-\begin{equation}\label{eq:adj_system}
-\begin{split}
-	A(\mathbf x)\frac{\partial}{\partial t}\tilde{u}(\mathbf x,t) - P^T \tilde{v}(\mathbf x,t) 
-		&= -\tilde{f}(\mathbf x,t), \\
-	B(\mathbf x)\frac{\partial}{\partial t}\tilde{v}(\mathbf x,t) + P\tilde{u}(\mathbf x,t) 
-		&= -\tilde{g}(\mathbf x,t).
-\end{split}
-\end{equation}
-Then
-\begin{equation}
-\label{eq:contadj}
-\int (\langle u,\tilde{f} \rangle + \langle v, \tilde{g} \rangle) \, dt
-= \int (\langle\tilde{u}, f \rangle + \langle \tilde{v},g \rangle) \, dt.
-\end{equation}
-\end{theorem}
-
-%The integrals appearing in the identity \ref{eq:contadj} are
-%well-defined, as the integrands have uniformly bounded support in
-%$\bx$ for each $t$, and vanish outside of a bounded interval in $t$
-%due to the presence of both causal and anticausal factors in each term.
-
-%First time t=0 and t=T appear in integrals, should we mention 
-%perhaps that causality begins at t0=0 and anticausality at t0=T ?? MJB
-\begin{proof}
-\[
-\int (\langle u,\tilde{f} \rangle + \langle v, \tilde{g} \rangle) \, dt
-=
-\int \left(\left\langle u,-A\frac{\partial \tilde{u}}{\partial t} -
-    P^T \tilde{v}\right\rangle\right.
-+
-\left. \left\langle v,-B\frac{\partial \tilde{v}}{\partial t} + P
-    \tilde{u}\right\rangle\right) \, dt
-\]
-\[
-=
-\int \left(\left\langle A\frac{\partial u}{\partial
-      t},\tilde{u}\right\rangle
--\langle P u,\tilde{v}\rangle + \left\langle B\frac{\partial
-    v}{\partial t},\tilde{v}\right\rangle + \langle
-P^Tv,\tilde{u}\rangle
-\right) \, dt
-\]
-\[
-=
-\int \left(\left\langle A\frac{\partial u}{\partial
-      t} + P^Tv,\tilde{u}\right\rangle + \left\langle B\frac{\partial
-    v}{\partial t}-Pu,\tilde{v}\right\rangle\right) \, dt
-\]
-\[
-= \int (\langle\tilde{u}, f \rangle + \langle \tilde{v},g
-\rangle) \, dt.
-\]
-\end{proof}
-
-The definition of distribution solution is related to the definition
-of the adjoint system \ref{eq:adj_system}. A pair $(u,v)$ of
-distributions is a {\em weak} or {\em distribution} solution of
-%what is mathcal E ?? MJB
-%may be missing a mathcal ?? MJB
-\ref{eq:cont_system} with right-hand sides $(f,g) \in \mathcal
-D'(\bR^{d+1},\bR^{k_1}) \times \mathcal D'(\bR^{d+1},\bR^{k_2})$ if for any pair $(\phi,\psi) \in
-C_0^{\infty}(\bR^{d+1},\bR^{k_1}) \times C_0^{\infty}(\bR^{d+1},\bR^{k_2})$,
-\[
-\int \left( \left\langle u, -A\frac{\partial \phi}{\partial 
-    t} + P^T \psi\right\rangle + \left\langle v, -B\frac{\partial\psi}{\partial t} + P\phi\right\rangle \right) \, dt
-\]
-\begin{equation}
-\label{eq:defdistsol}
-= \langle f,\phi \rangle + \langle g,\psi \rangle
-\end{equation}
-
-The references cited above show that unique causal (or anticausal)
-distribution solutions exist for distribution right hand sides of
-compact support. Moreover, it is implicit in the definition that these
-distribution solutions satisfy the relation \ref{eq:contadj}, with the
-$\langle \cdot, \cdot \rangle$ interpreted as the pairing of
-distribution and test function rather than as the $L^2$ inner product
-(the former is in a sense a generalization of the latter, of course).
-
-% WWS added 2018/01/03
-In particular, unique causal distribution solutions of the system
-\ref{eq:cont_system} exist for multipole right-hand sides $(f,g)$ of
-the form \ref{eq:MPSappx}, provided that the multipole coefficients
-$w_{\mathbf s}$ are distributions of compact support.
-  %\item we assume solution $(u,v)$ and source terms $(f,g)$ are causal, i.e., for $t < 0$
-%		\[
-%			u(\mathbf x,t) = f(\mathbf x,t) = 0, \quad 
-%			v(\mathbf x,t) = g(\mathbf x,t) = 0.
-%		\]
-%\end{itemize}
-%See \cite{BlazekStolkSymes:13} for proof of existence of solutions, stability, and other mathematical properties for a larger class of partial (integro-)differential systems for wave modeling, applicable to system \ref{eq:cont_system}.
+See \cite{BlazekStolkSymes:13} for proof of existence of solutions, stability, and other mathematical properties for a larger class of partial (integro-)differential systems for wave modeling, applicable to system \ref{eq:cont_system}.
 
@@
 %\end{equation}
-%where $u=p$ is the scalar pressure field and $v=\mathbf v\in\mathbf R^d$ is the vector particle velocity field; we take $\mathcal H_1=H^1_0(\Omega)$ and $\mathcal H_2=H^1(\Omega)^d$. %MOD
-%Coefficient operator $A(\mathbf x)=1/\kappa(\mathbf x)$ and $B(\mathbf x)=\rho(\mathbf x) I$, with $\kappa$ denoting bulk-modulus and $\rho$ density of the medium; here $I\in\mathbf R^{d\times d}$ is the identity matrix.
+%where $u=p$ is the scalar pressure field and $v=\mathbf v\in\mathbb R^d$ is the vector particle velocity field; we take $\mathcal H_1=H^1_0(\Omega)$ and $\mathcal H_2=H^1(\Omega)^d$. %MOD
+%Coefficient operator $A(\mathbf x)=1/\kappa(\mathbf x)$ and $B(\mathbf x)=\rho(\mathbf x) I$, with $\kappa$ denoting bulk-modulus and $\rho$ density of the medium; here $I\in\mathbb R^{d\times d}$ is the identity matrix.
 %Lastly, the differential operator $P$ coincides with the gradient and its adjoint with the negative of the divergence,
@@
 
-%%%%%%%%%%%
-\subsubsection{Staggered grid discretization}
-Staggered grid discretizations of hyperbolic systems can be quite
-complex in structure, as examples presented by \cite{moczoetal:06}
-show. Fortunately the essential properties of these discrete dynamical
-systems can be captured in a fairly simple abstraction.
-
-%We don't make use of <,>_h or || ||_h notation throughout ?? MJB
-We presume a family $\{H^1_h, H^2_h: h \in \bR_+\}$ of inner product
-space pairs, and surjections $S^i_h:
-C_0^{\infty}(\bR^d,\bR^{k_i}) \rightarrow H^i_h$, $i=1,2$. We will
-denote the inner product in $H^i_h$ by $\langle \cdot,\cdot\rangle$, 
-with the index $i = 1,2$ and inferred from context. 
-Moreover, distinguishing inner products between elements in
-$H_h^i$ or continuum functions (e.g., $C^\infty$ functions) is discerned by context.
-The corresponding induced norm is denoted $\|\cdot\|$.  
-In the examples that motivate this work, $H^i_h$ consists of vector valued
-functions of compact support on uniform rectangular grids in $\bR^d$ of mesh $h$, and $S^i_h$
-samples the smooth input on grid points. Several
-grids may figure in the definition of each of these spaces, and
-with some components defined on each grid - see \cite{moczoetal:06}
-for numerous examples.
-
-The next ingredient is a uniformly bounded family $\{Q_h \in {\cal
-  B}(H^1_n,H^2_h)\}$ of linear maps. This family defines an 
-approximation to the differential operator $P$, in a sense to be made
-precise below.
-
-For future use, set
-\[
-Q^* = \sup_{h \in \bR_+} \|Q_h\|.
-\]
-
-Taking advantage of a bit of hindsight, we define an abstract staggered grid
-dynamical system to be the recursion for a sequence of vector pairs $(u^n_h,v^{n-1/2}_h)
-\in H^1_h \times H^2_h: n \in {\bf Z}$:
+The discretization of continuum problem \ref{eq:cont_system} is given by the following abstract staggered grid dynamical system:
 \begin{equation}\label{eq:sg}
@@
 \end{equation}
-in which
+where
 \begin{itemize} 
-
-	\item $A_h, \, B_h$ are bounded self-adjoint positive-definite operators 
-\[
-	A_h: H_h^1 \rightarrow H_h^1,\quad B_h: H_h^2 \rightarrow H_h^2
+	\item $u_h^n\in H_{1,h}$ and $v_h^{n+1/2}\in H_{2,h}$
+	for a 	family of Hilbert spaces $H_{1,h}, H_{2,h}$ (``space of spatial-grid functions''), with $h>0$ ("cell size").
+Superscript indexes in $u_h^{n}$ and $v_h^{n+1/2}$ refer to the discretized time axis at times $n\Delta t$ and $(n+1/2)\Delta t$ respectively, with time step size $\Delta t = rh$; $r$ will play the role of the CFL constant, crucial for stability of the discretization.
+Note that the fields $u_h$ and $v_h$ are staggered in time;
+
+	\item bounded self-adjoint positive-definite operators (discretizations of $A$ and $B$)
+\[
+	A_h: H_{1,h} \rightarrow H_{1,h},\quad B_h: H_{2,h} \rightarrow H_{2,h}
 \]
@@
 	A_*\|u_h\|^2 \le \langle A_h u_h,u_h \rangle \le A^*\|u_h\|^2, 
-	\quad \forall u_h\in H_h^1, \\
+	\quad \forall u_h\in H_{1,h}, \\
 	B_*\|v_h\|^2 \le \langle B_h v_h,v_h \rangle \le B^*\|v_h\|^2,
-	\quad \forall v_h\in H_h^2; \\
+	\quad \forall v_h\in H_{2,h}; \\
 \end{split}
 \end{equation*}
-\item $f_h^{n+1/2} \in H^1_h,\,\, g_h^{n} \in H^2_h$,
-\item $P_h(r) = rQ_h, \,\ r > 0$.
+\item bounded finite difference operators $P_h(r), Q_h:H_{1,h} \rightarrow H_{2,h}$ such that there is a constant $Q^*>0$ where
+\[
+	\|Q_h\| \le Q^* \mbox{ for all } h>0,  \quad P_h(r) = rQ_h
+\]
+%ADDED
+\hl{and $P_h(r)$ satisfies the following skew-adjoint relation}
+\begin{equation}\label{eq:skew_h}
+	\langle P_h(r) u_h, v_h \rangle = - \langle u_h, P_h(r)^T v_h\rangle
+\end{equation}
+for all $u_h\in H_{1,h}$ and $v_h\in H_{2,h}$;
+\item discrete source terms $f_h^{n+1/2}\in H_{1,h}$ and $g_h^n\in H_{2,h}$; if $(f,g)$ are smooth source terms, then
+\[
+	f_h^{n+1/2} = S_{1,h} f(\cdot,(n+1/2)\Delta t), \quad g_h^n = S_{2,h} g(\cdot,n\Delta t),
+\]
+where we define bounded operator $S_{1,h}:\mathcal H_1\to H_{1,h}$, and $S_{2,h}:\mathcal H_2\to H_{2,h}$ for ``sampling'' continuum fields onto the spatial grids.
 \end{itemize}
@@
 %\[
-%	\mathcal G(\mathbf 0,\mathbf h) = \{ \mathbf x_{i,j} = (ih,jh) \; : \; i,j\in\mathbf Z \}.
+%	\mathcal G(\mathbf 0,\mathbf h) = \{ \mathbf x_{i,j} = (ih,jh) \; : \; i,j\in\mathbb Z \}.
 %\]
@@
 %%MOD
-%functions on rectangular grids that satisfy the homogenous boundary condition, and $H_{2,h}$ is the set of square summable $\mathbf R^2$-valued functions on half-cell shifted grids, both equipped with a discrete $L^2$ inner-products,
+%functions on rectangular grids that satisfy the homogenous boundary condition, and $H_{2,h}$ is the set of square summable $\mathbb R^2$-valued functions on half-cell shifted grids, both equipped with a discrete $L^2$ inner-products,
 %\begin{equation*}
@@
 %again, with $r =  \Delta t/h$.
-%In general, the differential operator $P$ can be approximated by a family of central difference operators $P_h$ of even order $2p$ for $p\in\mathbf N$, resulting in the 2-$2p$ staggered-grid finite difference schemes.
+%In general, the differential operator $P$ can be approximated by a family of central difference operators $P_h$ of even order $2p$ for $p\in\mathbb N$, resulting in the 2-$2p$ staggered-grid finite difference schemes.
 %%ADDED
@@
 
-%We aren't consistent between E or E_h ?? MJB
-Inspired by the work of Gustafsson and colleagues on staggered grid
-schemes for acoustics \cite[]{GusWahl:04a,GusWahl:04b,GusMoss:04},
-define $E: H^1_h \times H^2_h \rightarrow \bR_+$ by
+Define the following energy form:
 \begin{equation*}
 \begin{split}
-	E(u_h,v_h) &:= \frac{1}{2} \Big( \langle A_h u_h,u_h\rangle + 
-				       \langle B_hv_h,v_h\rangle +
-				       \langle P^T_h v_h,u_h\rangle \Big).
+	E^n := E(u_h^n,v_h^{n-1/2}) &:= \frac{1}{2} \Big( \langle A_h u_h^n,u_h^n\rangle + 
+				       \langle B_hv_h^{n-1/2},v_h^{n-1/2}\rangle +
+				       \langle P^T_h v_h^{n-1/2},u_h^n\rangle \Big).
 %	&= \frac{1}{2} \mat{ u_h^n \\ v_h^{n-1/2}}^T
@@
 \begin{theorem}
-Denote by $SPD_k$ the set of symmetric positive definite $k\times k$
-matrices. Then there exist $R, C_*, C^*: SPD_{k_1} \times SPD_{k_2} \times
-\bR^{k_2\times k_1} \rightarrow \bR_+$ so that 
-\begin{itemize}
-\item if
-\begin{equation}
-\label{eq:cfl}
-r < R(A_h,B_h,Q_h)
-\end{equation}
-then
+For a solution $(u_h,v_h)$ of system \ref{eq:sg} with $(f_h, g_h)\equiv 0$, it follows that $E^n$ is independent of time index $n$.
+Moreover, for sufficiently small $r$, there exist $0 < C_* \le C^*$ such that 
 \begin{equation}
 \label{eqn:pd}
-	C_*(A_h,B_h,Q_h) \Big( \|u_h\|^2 + \|v_h\|^2 \Big) \le E(u_h,v_h) \le C^*(A_h,B_h,Q_h)\Big( \|u_h\|^2 +\|v_h\|^2 \Big).
-\end{equation}
-\item $R(A_h,B_h,Q_h) \ge \sqrt{A_*B_*}/Q^*$.
-\end{itemize}
+	C_* \Big( \|u_h\|^2 + \|v_h\|^2 \Big) \le E(u_h,v_h) \le C^*\Big( \|u_h\|^2 +\|v_h\|^2 \Big).
+\end{equation}
 \end{theorem}
@@
 \begin{proof}
-The upper bound in \ref{eqn:pd} is clear from $h$-uniform bounds $A^*$on $A_h$, $B_h$, 
-$Q_h$. 
-The lower bound follows from the inequality
-\[
-	E(u_h,v_h) \ge A_*\|u_h\|^2 + B_*\|v_h\|^2 - rQ^*\|u_h\|\|v_h\|
-\]
-and the RHS is positive definite in $\|u_h\|,\|v_h\|$ when $rQ^* <\sqrt{A_*B_*}$. 
-\end{proof}
-
-That is,  the quadratic form $E$ defines a norm equivalent to the
-product norm $\|\cdot\|$ on $H^1_h \times H^2_h$, provided that $r$
-is sufficiently small relative to spectral bounds for the coefficient
-matrices. This condition is in fact the Courant-Friedrichs-Levy
-condition for staggered grid schemes.
-
-\begin{theorem}
-Suppose that $\{(u_h^n,v^{n-1/2}_h): n \in {\bf Z}\}$ is a solution of
-system \ref{eq:sg} with $(f^{n+1/2}_h, g^n_h)\equiv 0$ for all n. 
-Then $E^n=E(u_h^n,v^{n-1/2}_h)$ is independent of time index $n$.
-\end{theorem}
-
-\begin{proof}
+First we show the energy conservation property:
 \begin{equation*}
@@
 		&+ \langle B_h( v^{n+1/2}_h-v^{n-1/2}_h ), (v^{n+1/2}_h+v^{n-1/2}_h)\rangle \\
-		&+ \langle P_h^T v^{n+1/2}_h, u^{n+1}_h\rangle - \langle P_h^T v^{n-1/2}_h, u^n_h\rangle
+		&+ \langle P_h^T v^{n+1/2}_h, u^{n+1}_h\rangle ) - \langle P_h^T v^{n-1/2}_h, u^n_h\rangle )
 \end{split}
@@
 		&+ \langle P_hu^n_h, (v^{n+1/2}_h+v^{n-1/2}_h) \rangle\\
-		&+\langle P_h^Tv^{n+1/2}_h,u^{n+1}_h\rangle - \langle P^T_h v^{n-1/2}_h,u^n_h\rangle
+		&+\langle P_h^Tv^{n+1/2}_h,u^{n+1}_h\rangle ) - \langle P^T_h v^{n-1/2}_h,u^n_h\rangle )
 	=0.
@@
 \end{equation*}
+The upper bound in \ref{eqn:pd} is clear from $h$-uniform bounds on $A_h$, $B_h$,
+$P_h$. 
+The lower bound follows from
+\[
+	E(u_h,v_h) \ge A_*\|u_h\|^2 + B_*\|v_h\|^2 - rQ^*\|u_h\|\|v_h\|
+\]
+which is a positive definite form in $\|u_h\|,\|v_h\|$ when $rQ^* <\sqrt{A^*B^*}$.
+In other words, we have established positive definiteness of $E(\cdot,\cdot)$, and more importantly, that the energy form is equivalent to a norm.
 \end{proof}
@@
 %%%%%%%%
+\subsubsection{$L^2$ Convergence of Smooth Solutions}
 
@@
 
+Given theorem 5, that is, equivalency of between the $L^2$ and energy norms, we focus on proving stability estimates with respect to the energy norm in order to imply $L^2$ convergence.
+
 \begin{theorem}
-\label{thm:inhom}
-Suppose that $\{(u_h^n,v^{n-1/2}_h): n \in {\bf Z}\}$ is a solution of
-system \ref{eq:sg}. Assume that $r$ sufficiently small that $E$ is
-positive definite, recall the lower bound $C_*$ in the 
-equivalence relation \ref{eqn:pd}. Set $a = 2/C_*$. For $n \ge 0$ and $h < C_*/r$,
-\begin{equation}
-\label{eq:inhom}
-	E^n \le e^{anrh} \left(E^0 + 
-	C_* \sup_{0 \le m \le n-1}(\|f^{m+1/2}_h\|^2 + \|g^m_h\|^2)\right)
-\end{equation}
+For $r$ sufficiently small that $E$ is positive definite, stability follows, i.e., there exists $K \ge 0$, $\lambda > 1$ so that
+\[
+	E^n \le \lambda^n E^0 + 
+	Kh\sum_{m=0}^{n-1}\lambda^{n-m+1}(\|f^{m-1/2}_h\|^2 + \|g^m_h\|^2)
+\]
+where $(u_h,v_h)$ satisfy the inhomogeneous system \ref{eq:sg} with source terms $(f_h,g_h)$.
 \end{theorem}
@@
 \end{equation*}
-for any $\alpha \in (0,1)$. Choose $\alpha^2=1/2$. From the assumption $rh \le C_* <2C_*$,
-$\alpha^2 rh < C_*$. Set 
+for any $\alpha \in (0,1)$. Choose $\alpha$ so that $\alpha^2 rh < C_*$
+(that is, can make fixed choice of $\alpha$ for $rh$ small enough), and
+set 
 \begin{equation}\label{eq:K}
-	\lambda = \left(1+\frac{rh}{2C_*}\right) \left(1-\frac{rh}{2C_*}\right)^{-1}
+	\lambda = \left(1+\frac{\alpha^2rh}{C_*}\right) \left(1-\frac{\alpha^2rh}{C_*}\right)^{-1}
 	\quad
@@
 	\quad
-	K = \frac{1}{ \left(1+\frac{rh}{2C_*}\right) }.
-\end{equation}
-Then $\lambda > 1$ and 
+	K = \frac{\frac{r}{2\alpha^2}}{ \left(1+\frac{\alpha^2rh}{C_*}\right) }.
+\end{equation}
+Then
 \begin{equation*}
 \begin{split}
-	\lambda^{-1}E^{n} - E^{n-1} &\le Krh (\|f^{n-1/2}_h\|^2 + \|g^{n-1}_h\|^2)  \\
+	\lambda^{-1}E^{n+1} - E^n &= Kh (\|f^{n+1/2}_h\|^2 + \|g^n_h\|^2)  \\
 	\Longrightarrow
-	\lambda^{-n}E^{n} - \lambda^{-(n-1)}E^{n-1} &\le \lambda^{-(n-1)}Krh (\|f^{n-1/2}_h\|^2 + \|g^{n-1}_h\|^2) 
+	\lambda^{-(n+1)}E^{n+1} - \lambda^{-n}E^n &= \lambda^{-n}Kh (\|f^{n+1/2}_h\|^2 + \|g^n_h\|^2) 
 \end{split}
@@
 whence
-\begin{equation}
-\label{eq:est1}
-\lambda^{-n}E^n \le E^0 +
-Krh\sum_{m=0}^{n-1}\lambda^{-m}(\|f^{m+1/2}_h\|^2 + \|g^m_h\|^2). 
-\end{equation}
-
-Set $M_n=\sup_{0 \le m \le n-1}(\|f^{m+1/2}_h\|^2 +
-\|g^m_h\|^2)$. Then the inequality \ref{eq:est1} implies
-
-\[
-\lambda^{-n}E^n \le E^0 +
-Krh \frac{1-\lambda^{-n}}{1-\lambda^{-1}} M_n
-\]
-\[
-\le E^0 + Krh\frac{1}{1-\lambda^{-1}} M_n
-\]
-\[
-\le E^0 + Krh \frac{1 + \frac{rh}{2C_*}}{2\left(
-    \frac{rh}{2C_*}\right)}M_n
-\]
-From the definition \ref{eq:K} of $K$, this is equivalent to 
-\begin{equation}
-\label{eq:est2}
-\lambda^{-n}E^n \le E^0 + C_* M_n.
-\end{equation}
-The simple inequality 
-\[
-\frac{1+x}{1-x} \le e^{4x},\, 0\le x \le \frac{1}{2}
-\]
-together with the hypothesis that $rh \le C_*$ so that $rh/2C_* < 1/2$, and the definition
-\ref{eq:K} of $\lambda$ implies that 
-\[
-\lambda \le e^{2\frac{rh}{C_*}}
-\]
-Thus the estimate \ref{eq:est2} implies the conclusion \ref{eq:inhom}.
+\[
+\lambda^{-n}E^n = E^0 +
+Kh\sum_{m=0}^{n-1}\lambda^{-m+1}(\|f^{m-1/2}_h\|^2 + \|g^m_h\|^2).
+\]
 \end{proof}
 
-In view of the relation between the continuum adjoint and the system
-\ref{eq:adj_system}, it is not surprising that a similar relation
-exists between the adjoint of the map defined by the system
-\ref{eq:sg} and a related discrete adjoint state system.
-
-A sequence $\{f^n\}$ (or $\{f^{n+1/2}\}$) vector-valued functions is {\em causal} if there is
-$n_{\rm min} \in {\bf Z}$ so that $f^n=0$ (or $f^{n+1/2}=0$) for
-$n<n_{\rm min}$. Likewise, such a function is {\em anticausal} if
-there is $n_{\rm max}$ for which  $f^n=0$ (or $f^{n+1/2}=0$) for
-$n>n_{\rm max}$.
-
 \begin{theorem}
-Suppose that $\{(f_h^{n+1/2},g_h^{n})\}$ is a causal sequence,
-$\{(\tilde{f}_h^{n},\tilde{g}_h^{n+1/2})\}$ an anticausal
-sequence in $H^1_h \times H^2_h$.
-Let $\{ (u_h^n,v_h^{n+1/2})\}$ be the unique causal solution of 
-the system \ref{eq:sg}, and
- $\{(\tilde u_h^{n+1/2},\tilde v_h^{n})\}$ the unique anticausal
-solution of
-\begin{equation}\label{eqn:inadj}
-\begin{split}
-	A_h( \tilde u_h^{n-1/2}-\tilde u_h^{n+1/2} ) - P_h^T \tilde v^n & =  rh\, \tilde f_h^{n} \\
-	B_h( \tilde v^{n} - \tilde v^{n+1} ) + P_h \tilde u^{n+1/2} &=  rh \, \tilde g_h^{n+1/2}. 
-\end{split}
-\end{equation}
-Then 
-\begin{equation}
-\label{eqn:weak}
-	 rh\sum_{n} \Big( \langle \tilde u_h^{n+1/2}, f_h^{n+1/2} \rangle +
-	 			 	  \langle \tilde v_h^n, g_h^n \rangle \Big) 
-	= rh\sum_{n} \Big( \langle \tilde f_h^n, u_h^n \rangle + 
-					     \langle \tilde g_h^{n+1/2}, v_h^{n+1/2}\rangle \Big). 
-\end{equation}
-\end{theorem}
-
-Note that both sums in the identity \ref{eqn:weak} are finite, as all
-terms have both causal and anticausal factors.
-
-\begin{proof}
-\[
-\sum_{n} \Big( \langle \tilde u_h^{n+1/2}, f^{n+1/2}\rangle +
-					   \langle \tilde v_h^n, g_h^n \rangle \Big)
-\]
-\begin{equation*}
-\begin{split}
-	= \sum_n \Big( \langle \tilde u_h^{n+1/2}, 
-					A_h( u_h^{n+1} - u_h^n) + P_h^T v_h^{n+1/2} \rangle 
-			  +  \langle \tilde v_h^n, 
-					B_h ( v_h^{n+1/2}-v_h^{n-1/2} ) - P_h u_h^n \rangle \Big)
-\end{split}
-\end{equation*}
-\begin{equation*}
-\begin{split}
-	= \sum_n \Big( \langle A_h ( \tilde u_h^{n-1/2}-\tilde u_h^{n+1/2} ), u_h^n \rangle 
-			       -\langle P_h \tilde u_h^{n+1/2}, v_h^{n+1/2}\rangle 
-	 		       +\langle B_h ( \tilde v_h^{n} - \tilde v_h^{n+1} ), v_h^{n+1/2} \rangle
-			       +\langle P_h^T \tilde v_h^n, u_h^n\rangle \Big)
-\end{split}
-\end{equation*}
-\begin{equation}\label{eqn:sbp}
-\begin{split}
-	= \sum_n \Big( \langle A_h (\tilde u_h^{n-1/2} - \tilde u_h^{n+1/2}) - P_h^T \tilde v_h^n, 
-				u_h^n\rangle 
-	                      +	\langle B_h (\tilde v_h^{n} - \tilde v_h^{n+1}) + P_h \tilde u_h^{n+1/2},
-	                      	v_h^{n+1/2}\rangle
-\end{split}
-\end{equation}
-In view of the system \ref{eqn:inadj}, the last line is equal to the
-right hand side of \ref{eqn:weak}.
-
-\end{proof}
-
-%%%%%%%%%
-\subsubsection{Norm Convergence to Smooth Solutions}
-
-We will suppose from now on that $Q_h$ approximates $P$ to order $p\in {\bf N}$, in the sense that
-for all $u_i \in C_0^{\infty}(\bR^d,\bR^{k_i}),\, i=1,2$,
-\begin{equation}
-\label{eq:fdord}
-\begin{split}
-\left\| S^2_hPu_1 - \frac{1}{h}Q_h S^1_h u_1\right\| &= O(|h|^p),\\
-\left\| S^1_hP^Tu_2 - \frac{1}{h}Q_h^T S^2_h u_2\right\| &= O(|h|^p).
-\end{split}
-\end{equation}
-For the grid-based examples described earlier, this
-condition defines precisely the usual truncation-based order of
-accuracy. In the second equation, the transpose $P^T$ is the usual
-formal adjoint, whereas $Q_h^T$ is the adjoint with respect to the
-assumed inner product structure on $H^i_h, \,i=1,2$.
-
-Since we have assumed nothing {\em a priori} about the relation between distribution
-pairing in $\bR^d$ and the inner products on $H^i_h, \,i=1,2$, 
-the second estimate \ref{eq:fdord} does not follow from the first, and must be
-asserted explicitly. In many examples, on the other hand, such as the
-acoustic staggered grid approximation to be described later, the 
-transpose automatically has the same order of accuracy.
-
-Note that the implicit constants hiding behind the ``big O'' symbols
-depend on the functions $u_1, u_2$, .
-
-\begin{theorem}
-\label{thm:smconvg}
-Suppose that $Q_h$ approximates $P$ to order $p$, $T>0$, and that $\Delta t:
-\bR_+ \rightarrow \bR_+$ satisfies $\Delta t(h) <
-hR(A,B,Q)$ as in condition \ref{eq:cfl}. Suppose also that $u \in C^{\infty}(\bR^{d+1},\bR^{k_1})$, $v \in
-  C^{\infty}(\bR^{d+1},\bR^{k_2})$ solve the hyperbolic system
-  \ref{eq:cont_system} with coefficients $A,B$, right hand sides
-  $f,g$, and Cauchy data $u^0,v^0$ satisfying the conditions of
-  Theorem \ref{thm:well-posed}. Define
-  $A_h,B_h,f^{n+1/2}_h,g^{n}_h,$ and $u^0_h,v^{-1/2}_h$ by
-\begin{itemize}
-\item $A_h S^1_hw_1=S^1_h Aw_1, \,\,B_hS^2_hw_2 = S^2_hBw_2$ for any $w_i
-  \in C_0^{\infty}(\bR^d,\bR^{k_i})$; 
-\item $f^{n+1/2}_h = S^1_hf(\cdot,(n+1/2)\Delta t(h)),\,\, g^n_h=S^2_h
-  g(\cdot,n\Delta t(h))$
-\item $u^0_h=S^1_hu(\cdot,0),\,\, v^{-1/2}_h=S^2_hv(\cdot,-1/2 \Delta
-  t(h))$
-\end{itemize}
-Let $(u^n_h,v^{n-1/2}_h)$ denote the solution of the staggered grid
-scheme \ref{eq:sg} with data $A_h,B_h,f^{n+1/2}_h,g^{n}_h,$ and
-$u^0_h,v^{-1/2}_h$, $N(h)=T/\Delta t(h)$, and $r(h) = \Delta
-t(h)/h$. Then for $h>0$ and $0 \le n \le N(h)$, 
-\begin{equation}
-\label{eq:errest}
-\begin{split}
-\|S^1_hu(\cdot,n \Delta t(h)) - u^n_h\| &= O(\Delta t(h)^2+h^p)\\
-\|S^2_hv(\cdot,(n-1/2)\Delta t(h)) - v^{n-1/2}_h\| & = O(\Delta
-t(h)^2 +h^p).
-\end{split}
-\end{equation}
-\end{theorem}
-
-\begin{proof}
-Recall that $\Delta t = rh$ and that $r=r(h)$, hence $\Delta t=\Delta
-t(h)$, is a (possibly linear) function of $h$, satisfying the
-stability condition \ref{eq:cfl} for all $h>0$. For the remainder of
-the proof, we suppress the dependence on $h$ from the notation.
+Suppose that $(u^n,v^{n-1/2})\in \mathcal H_1 \times \mathcal H_2$,
+and $(u^n_h,v^{n-1/2}_h)$ solves the discretized system \ref{eq:sg} with initial data
+\[
+	u^0_h = S_{1,h} u^0, \quad v^{-1/2}_h = S_{2,h} v^{-1/2}.
+\]
 Set
@@
 \begin{split}
-  \delta u^n_h &  = S^1_hu(\cdot,n \Delta t) - u^n_h\\
-  \delta v^{n-1/2}_h & = S^2_hv(\cdot,(n-1/2)\Delta t) - v^{n-1/2}_h\\
-  \delta f^{n+1/2}_h & = \frac{1}{rh} \Big( A_h( S^1_h u(\cdot,(n+1)\Delta t) - S^1_h u(\cdot,n\Delta t)) + 
-  P_h(r(h))^T S^2_h v(\cdot,(n+1/2)\Delta t) \Big) + f_h^{n+1/2}\\
-  \delta g^{n}_h  &= \frac{1}{rh} \Big( B_h (S^2_h v(\cdot,(n+1/2)\Delta t) - S^2_h v(\cdot,(n-1/2)\Delta t) ) - 
-  P_h(r(h)) S^1_h u(\cdot,n\Delta t) \Big) + g_h^{n}.
-\end{split}
-\end{equation}
-Then the truncation error vector $(\delta u^n_h, \delta v^{n-1/2}_h)$ solves the staggered grid
-system
-\begin{equation}\label{eq:truncsg}
-\begin{split}
-  A_h(\delta u^{n+1}_h - \delta u^n_h)  + P_h(r)^T \delta
-  v^{n+1/2}_h &= rh\,   \delta f_h^{n+1/2} \\
-  B_h(\delta v^{n+1/2}_h-\delta v^{n-1/2}_h) -  P_h(r) \delta
-  u^n_h &=  rh\, \delta g_h^{n} 
-\end{split}
-\end{equation}
-which is precisely of the form \ref{eq:sg}. Note that $\delta u^0_h = 0,
-\delta v^{-1/2}_h=0$. Therefore from Theorem
-\ref{thm:inhom},
+	\delta f^{n+1/2}_h & = \frac{1}{rh} \Big( A_h( S_{1,h} u^{n+1} - S_{1,h}u^{n}) + 
+							   P_h^T S_{2,h} v^{n+1/2} \Big) + f_h^{n+1/2}\\
+	\delta g^{n}_h  &= \frac{1}{rh} \Big( B_h (S_{2,h} v^{n+1/2} - S_{2,h} v^{n-1/2}) - 
+							P_h S_{1,h}u^n \Big) + g_h^{n}.
+\end{split}
+\end{equation}
+Moreover, let $T>0$ be given, independent of $h$. If $(\delta f_h,\delta g_h)$, as defined above, satisfy the estimate
+\begin{equation}\label{eq:estimate}
+	\|\delta f_h^{m-1/2}\|^2 + \|\delta g_h^m\|^2 \le L^2 h^{2p}
+\end{equation}
+	for $0\le m \le N$, $N\Delta t=T$, then 
 \begin{equation}
-\label{eq:truncest1}
-E(\delta u^n_h, \delta v^{n-1/2}_h) \le e^{anr(h)h}C_*\sup_{0 \le m
-  \le n-1}(\|\delta f^{m+1/2}_h\|^2 + \|\delta g^m_h\|^2)
-\end{equation}
-
-From the definitions of $A_h$, $f^{n+1/2}_h$, 
-\[
-rh \delta f_h^{n+1/2} =  A_h(\delta u^{n+1}_h - \delta u^n_h)  + P_h(r)^T \delta
-  v^{n+1/2}_h 
-\]
-\[
-= A_h(S^1_h(u(\cdot,(n+1)\Delta t)-u(\cdot,n\Delta t)) - A_h(u^{n+1}_h-u^n_h) +
-P_h(r(h))^T (S^2_h(v(\cdot,(n+1/2)\Delta t)) - v^{n+1/2}_h)
-\]
-\[
-= S^1_h(A(u(\cdot,(n+1)\Delta t)-u(\cdot,n\Delta t))-P_h(r(h))^T
-S^2_h(v(\cdot,(n+1/2)\Delta t))
--rh f^{n+1/2}_h
-\]
-where we have used the first of the two equations in \ref{eq:sg} to
-eliminate the discrete fields $u^n_h,v^{n-1/2}_h$ in favor of $f^{n+1/2}_h$.
-Because of the smoothness of $u$, 
-\begin{equation}
-\label{eq:truncdt}
-u(\cdot,(n+1)\Delta t)-u(\cdot,n\Delta t) = 2\Delta t \frac{\partial u}{\partial 
-  t}(\cdot,(n+1/2)\Delta t) +
-O(\Delta t^3).
-\end{equation}
-From the order assumption \ref{eq:fdord} on $P_h = r Q_h$, 
-\begin{equation}
-\label{eq:truncdx}
-P_h(r)^T S^2_h(v(\cdot,(n+1/2)\Delta t) = \Delta t S^1_h (P^T v)(\cdot,(n+1/2)\Delta t) + O(\Delta
-t h^p)
-\end{equation}
-We conclude that 
-\[
-\Delta t \delta f_h^{n+1/2} = \Delta t  S^1_h\left(A \frac{\partial u}{\partial 
-  t}(\cdot,(n+1/2)\Delta t) + (P^T v)(\cdot,(n+1/2)\Delta t) -  f(\cdot,(n+1/2)\Delta
-t)\right)
-\]
-\[
- + O(\Delta t^3) + O(\Delta t h^p)
-\]
-\[
-= O(\Delta t^3) + O(\Delta t h^p)
-\]
-since $(u,v)$ are smooth solutions of \ref{eq:cont_system}.
-Similarly,
-\[
-\Delta t \delta g_h^n =  O(\Delta t^3) + O(\Delta t  h^p).
-\]
-The constants hiding inside the big ``O''s are uniform over compact
-sets. Since $0 \le n\Delta t \le T$ for $0 \le n \le N(h)$, and the
-data of the problem \ref{eq:cont_system} has compact support, the
-domain in which the errors are evaluated has compact support, so these
-inequalities imply that
-\[
-\|\delta f^{n+1/2}_h\|^2 + \|\delta g^n_h\|^2 = (O(\Delta t^2) + O(
-h^p))^2
-\]
-uniformly for $0 \le n \le N(h)$. Combined with the inhomogeneous
-energy estimate \ref{eq:inhom}, this observation implies the
-conclusion of the theorem.
+\label{eqn:ee}
+	E(u_h^n - S_{1,h}u^n, v_h^{n-1/2} - S_{2,h}v^{n-1/2} ) \le M^2L^2h^{2p}.
+\end{equation}
+\end{theorem}
+
+\begin{proof}
+Define $K, \lambda$ as in \ref{eq:K}, then it follows from previous theorem that
+\[
+	E(u^n_h - S_{1,h}u^n, v^{n-1/2}_h - S_{2,h} v^{n-1/2}) \le
+	Kh\sum_{m=0}^{n-1}\lambda^{n-m+1}(\| \delta f_h^{m-1/2}\|^2 + \|\delta g_h^m\|^2).
+\]
+which implies
+\[
+	E(u_h^n - S_{1,h}u^n, v_h^{n-1/2} - S_{2,h}v^{n-1/2} ) \le 
+	KL^2 h^{2p+1}\frac{\lambda^N-1}{\lambda-1}
+\]
+using estimate \ref{eq:estimate}.
+Note that if $h$ is sufficiently small, then $\lambda < 1 +
+2\frac{\alpha^2rh}{C_*}$ so $\lambda^N \le
+exp(2\frac{\alpha^2}{C_*}T)$, where $T=Nrh$. If $h$ is perhaps smaller
+yet, then $\lambda > 1+\frac{\alpha^2rh}{2C_*}$. Putting this all
+together, there is $M^2$ depending on $T$ and all of the other constants
+in the setup, so that if $nrh <T$ so that \ref{eqn:ee} holds.
 \end{proof}
@@
 %each time slice of compact support we have,
-%\hl{If $(u,v)$ is a smooth solution of the continuum problem} \ref{eq:cont_system}
-%\hl{with smooth right-hand sides $(f,g)$, and $(u_h,v_h)$ is the finite difference 
-%solution where $P_h$ is of order $2p$,} then, from standard truncation error 
-%calculations, for each time slice of compact support we have,
-%\begin{equation*}
-%\begin{split}
-%	\delta f_h^{n+1/2} &= O( \Delta t^2+h^{2p})\\
-%	\delta g_h^{n} &= O( \Delta t^2+h^{2p})
-%\end{split} 
-%\end{equation*}
-%point-wise and uniformly in $i,j$. 
-%Since the support is uniformly bounded if $0\le n\Delta t \le T$, 
-%the $O$ statements apply to the $L^2$ norms as well.
-%Thus, by equivalency of the energy form with the $L^2$-norm, and the \hl{theorem} above, 
-%we can conclude
-%\[
-%	\|u_h-S_{1,h}u\| + \|v_h-S_{2,h}v\| = O(\Delta t^2+h^{2p}),
-%\]
-%that is, $2$-$2p$ order convergence in the $L^2$ sense.
+\hl{If $(u,v)$ is a smooth solution of the continuum problem} \ref{eq:cont_system}
+\hl{with smooth right-hand sides $(f,g)$, and $(u_h,v_h)$ is the finite difference 
+solution where $P_h$ is of order $2p$,} then, from standard truncation error 
+calculations, for each time slice of compact support we have,
+\begin{equation*}
+\begin{split}
+	\delta f_h^{n+1/2} &= O( \Delta t^2+h^{2p})\\
+	\delta g_h^{n} &= O( \Delta t^2+h^{2p})
+\end{split} 
+\end{equation*}
+point-wise and uniformly in $i,j$. 
+Since the support is uniformly bounded if $0\le n\Delta t \le T$, 
+the $O$ statements apply to the $L^2$ norms as well.
+Thus, by equivalency of the energy form with the $L^2$-norm, and the \hl{theorem} above, 
+we can conclude
+\[
+	\|u_h-S_{1,h}u\| + \|v_h-S_{2,h}v\| = O(\Delta t^2+h^{2p}),
+\]
+that is, $2$-$2p$ order convergence in the $L^2$ sense.
 
@@
 %%%%%%%
-\subsection{Weak Convergence to Singular Solutions}
-
-In order to combine the moment-based analysis of gridded multipole
-approximations with the convergence theory developed in the last
-section, it is necessary to explicitly identify the Hilbert space
-families $H^1_h,H^2_h$ with vector spaces of grid functions. In order
-to accommodate the structure of commonly used staggered grid schemes
-\cite[]{moczoetal:06}, this definition must allow for participation of
-multiple grid types in each of these spaces.
-
-Recall the definition of a regular grid family $\mathcal G_h(\mathbf
-x_0,\mathbf l)$ with mesh $h \in \bR_+$, offset $\bx_0 \in
-\bR^d$ and cell given by ${\bf l} \in \bR_+^d$, namely $\{\bx \in
-\bR^d: 0 \le x_i\le l_k\}$:
-\[
-	\mathcal G_h(\mathbf x_0,\mathbf l) = \{\bx_{\bf n} =
-        h((x_0)_1+n_1l_1,...,(x_0)_d+n_dl_d), {\bf n} \in {\bf Z}^d\}.
-\]
-We define the {\em grid Hilbert space} $\mathcal H_h(\bx_0, {\bf l})$ to be the real-valued
-functions $f_h$ on $\mathcal G_h(\mathbf x_0,\mathbf l)$ for which
-\[
-\|f_h\|^2 = h^d \left(\Pi_{i=1}^dl_i\right) \sum_{\bx \in \mathcal G_h(\mathbf x_0,\mathbf l)}\,
-|f_h(\bx)|^2
-\]
-is finite. That is, $\| \cdot \|$ is the standard $l_2$ norm, scaled
-by the scaled cell volume.
-
-The sampling operator defined by a grid $\mathcal G_h(\mathbf
-x_0,\mathbf l)$ is the map
-\[
-\mathcal S_h(\bx_0,{\bf l}): C^{\infty}_0(\bR^d) \rightarrow \mathcal
-H_h(\bx_0, {\bf l})
-\]
-defined by evaluating $f_h$ at gridpoints:
-\[
-(\mathcal S_h(\bx_0,{\bf l})f_h)(\bx) = f_h(\bx).
-\]
-
-Define the spaces $H^1_h,H^2_h$ to be direct sums of grid Hilbert
-spaces with common mesh $h$. That is, given offsets $\bx_0^{j,i}$ and
-cell side vectors ${\bf l}^{j,i}, \,1 \le j \le k_i, i=1,2$, set
+\subsubsection{Weak Convergence of Singular Solutions}
+
+Let $n_{\rm min}, n_{\rm max}$ be integers.
+Suppose that the grid-function sequence $(u_h^n,v_h^{n+1/2})$, 
+satisfies \ref{eq:sg} with source term $(f_h^{n+1/2},g_h^{n})$
+and vanish identically for $n<n_{\rm min}$.
+Let $(\tilde u_h^{n+1/2},\tilde v_h^{n})$ be another sequence of 
+grid functions that vanish for $n>n_{\rm max}$. 
+Then,
+%$(q^n(h),w^{n+1/2}(h))$ be yet another $G(h)$-valued sequence that
+%vanished for $n > n_{\rm max}$. Then (suppressing $h$-dependence for
+%the moment)
+\[
+	\Delta t \sum_{n} \Big( \langle \tilde u_h^{n+1/2}, f^{n+1/2}\rangle +
+					   \langle \tilde v_h^n, g_h^n \rangle \Big)
+\]
+\begin{equation*}
+\begin{split}
+	= \sum_n \Big( \langle \tilde u_h^{n+1/2}, 
+					A_h( u_h^{n+1} - u_h^n) + P_h^T v_h^{n+1/2} \rangle 
+			  +  \langle \tilde v_h^n, 
+					B_h ( v_h^{n+1/2}-v_h^{n-1/2} ) - P_h u_h^n \rangle \Big)
+\end{split}
+\end{equation*}
+\begin{equation*}
+\begin{split}
+	= \sum_n \Big( \langle A_h ( \tilde u_h^{n-1/2}-\tilde u_h^{n+1/2} ), u_h^n \rangle 
+			       -\langle P_h \tilde u_h^{n+1/2}, v_h^{n+1/2}\rangle 
+	 		       +\langle B_h ( \tilde v_h^{n} - \tilde v_h^{n+1} ), v_h^{n+1/2} \rangle
+			       +\langle P_h^T \tilde v_h^n, u_h^n\rangle \Big)
+\end{split}
+\end{equation*}
+\begin{equation}\label{eqn:sbp}
+\begin{split}
+	= \sum_n \Big( \langle A_h (\tilde u_h^{n-1/2} - \tilde u_h^{n+1/2}) - P_h^T \tilde v_h^n, 
+				u_h^n\rangle 
+	                      +	\langle B_h (\tilde v_h^{n} - \tilde v_h^{n+1}) + P_h \tilde u_h^{n+1/2},
+	                      	v_h^{n+1/2}\rangle
+\end{split}
+\end{equation}
+
+Define $(\tilde f_h^{n},\tilde g_h^{n+1/2})\in H_{1,h}\times H_{2,h}$ by
+\begin{equation}\label{eqn:inadj}
+\begin{split}
+	A_h( \tilde u_h^{n-1/2}-\tilde u_h^{n+1/2} ) - P_h^T \tilde v^n & =  rh\, \tilde f_h^{n} \\
+	B_h( \tilde v^{n} - \tilde v^{n+1} ) + P_h \tilde u^{n+1/2} &=  rh \, \tilde g_h^{n+1/2}.
+\end{split}
+\end{equation}
+Then, identity \ref{eqn:sbp} can be re-written as
 \begin{equation}
-\label{eq:def-hilb}
-H^i_h = \oplus_{j=1}^{k_i}\mathcal H_h(\bx_0^{j,i},{\bf l}^{j,i})
-\end{equation}
-with the usual product Hilbert structure. The sampling operators
-\[
-S^i_h: C^{\infty}_0(\bR^d,\bR^{k_i}) \rightarrow H^i_h
-\]
-are diagonal:
+\label{eqn:weak}
+	\Delta t \sum_{n} \Big( \langle \tilde u_h^{n+1/2}, f_h^{n+1/2} \rangle +
+	 			 	  \langle \tilde v_h^n, g_h^n \rangle \Big)
+	= \Delta t \sum_{n} \Big( \langle \tilde f_h^n, u_h^n \rangle + 
+					     \langle \tilde g_h^{n+1/2}, v_h^{n+1/2}\rangle \Big).
+\end{equation}
+
+%MOD
+%Now specialize to the acoustic case - the elastic case is
+%similar. In this case $g$ (and $\tilde g$) is vector-valued, and $f$
+ %(and $\tilde f$) is scalar-valued. Once again, $P_h$ is a discrete gradient;
+%assume that it is accurate of order $2p$, as before. $P_h^T$ is a discrete
+%(negative) divergence, also accurate of order $2p$.  
+%%\hl{Assuming $P_h$ is accurate of order $2p$, it follows that $P_h^T$ is a $2p$-order discretization of $P^T$.}
+The adjoint system
+\ref{eqn:inadj} is recognized as another discretization of the
+original system \ref{eq:cont_system}, with two differences: 
+the time index is dual to the one used in \ref{eq:sg}, that is, 
+shifted by $\Delta t/2$, and the inhomogenous terms get a negative sign. Evidently, the truncation error analysis is exactly the same.
+
+Suppose that $(\tilde f,\tilde g)$ are smooth in both temporal and spatial variables, and that 
+\begin{equation}\label{eq:tilde_f}
+	\tilde f_h^n = S_{1,h}\tilde f^{n}, \quad
+	\tilde g_h^{n+1/2} = S_{2,h} \tilde g^{n+1/2}.
+\end{equation}
+Denote by $(\tilde u,\tilde v)$ the (weak) solution of the acoustic system \ref{eq:acoustics} with
+right-hand sides $(-\tilde f,-\tilde g)$ that vanishes for $t>T$. 
+The standard theory for hyperbolic systems shows that $(\tilde u,\tilde v)$ is also smooth. 
+Let $(\tilde u_h^{n+1/2},\tilde v_h^n)$ be the solution of the discrete adjoint system
+\ref{eqn:inadj} with right-hand sides $(\tilde f_h^n, \tilde g_h^{n+1/2})$. 
+Then (as before assuming that $P_h$ is accurate of order $2p$)
+\[
+	\delta \tilde u_h^{n+1/2} = \tilde u_h^{n+1/2} - S_{1,h} \tilde u^{n+1/2},\quad
+	\delta \tilde v_h^{n} = \tilde v_h^{n} - S_{2,h} \tilde v^{n}
+\]
+solve \ref{eqn:inadj} with right-hand side that is
+$O(\Delta t^2+h^{2p})$, so according to the error estimate \ref{eqn:ee}
+(which applies ipso facto to the system \ref{eqn:inadj}),
 \begin{equation}
-\label{eq:def-samp}
-(S^i_h f)_j = \mathcal S_h(\bx_0^{j,i},{\bf l}^{j,i})(f_j),
-\,j=1,...,k_i
-\end{equation}
-
-With these preliminaries, we can state the central result of this paper:
-
+\label{eqn:truncadj}
+	\| \delta \tilde u_h^{n+1/2}\|^2 + \|\delta \tilde v_h^{n}\|^2 = O(\Delta t^2+h^{2p})
+\end{equation}
+over any finite range of $n\Delta t$.
+
+We now have all of the ingredients to prove weak convergence, which we now state:
 \begin{theorem}\label{thm:conv}
-Choose $q, N^*\ge 0$, $p>0$. Assume that $Q_h$ approximates $P$ to
-order $p$. Let $(u,v)$ be the weak causal solution to continuum
-problem \ref{eq:cont_system} with singular source terms $(f,g)$
-given by order $N^*$ multipoles,
+Let $(u,v)$ be a solution to continuum problem \ref{eq:cont_system} with singular source terms $(f,g)$ of the form
 \begin{equation*}
 \begin{split}
-	f(\mathbf x,t) &=  \sum_{|{\bf s}_1| \le N^*}w_{{\bf s}_1}(t) D^{{\bf s}_1}\delta(\mathbf x-\mathbf x^*),  
-        \; w_{{\bf s}_1} \in C_0^{\infty}(\bR,\bR^{k_1});\\
-	g(\mathbf x,t) &= \sum_{|{\bf s}_2| \le N^*} z_{{\bf s}_2}(t) D^{{\bf s}_2}\delta(\mathbf x-\mathbf x^*),  
-	\;z_{{\bf s}_2} \in C_0^{\infty}(\bR,\bR^{k_2}). 
+	f(\mathbf x,t) &=  w(t) D^{\mathbf s_1}\delta(\mathbf x-\mathbf x^*),\\
+	g(\mathbf x,t) &= z(t) D^{\mathbf s_2}\delta(\mathbf x-\mathbf x^*),
 \end{split}
 \end{equation*}
-Let $A_h,B_h$ be defined as in the statement of Theorem
-\ref{thm:smconvg}. Suppose that $(f^{n+1/2}_h,g^n_h)$ to be discrete
-multipoles
-\begin{equation}
-\label{eq:mprhs}
-\begin{split}
-	f^{n+1/2}_h &=  \sum_{|{\bf s}_1| \le N^*}w_{{\bf
-            s}_1}((n+1/2)\Delta t) \eta_h^{\bf s_1},  \\
-	g^n_h &= \sum_{|{\bf s}_2| \le N^*} z_{{\bf
-            s}_2}(n\Delta t) \eta_h^{\bf s_2}
-\end{split}
-\end{equation}
-in which $\eta_h^{\bf s_i} \in H^i_h, i=1,2$ satisfies the discrete $(q,{\bf s_{\bf i}})$-moment condition
-\ref{eq:discmomcond} {\em in each of its components}. Let
-$(u^n_h,v^{n-1/2}_h)$ be the causal solution of \ref{eq:sg}
-with right-hand sides given by \ref{eq:mprhs}.
-
-Suppose that $(\tilde f,\tilde g) \in
-C^{\infty}_0(\bR^{d+1},\bR^{k_1}) \times
-C^{\infty}_0(\bR^{d+1},\bR^{k_2})$, and set
-\begin{equation}
-\label{eq:defadjrhs}
-\begin{split}
-\tilde{f}^n_h &= S^1_h\tilde{f}(\cdot,n\Delta t)\\
-\tilde{g}^{n+1/2}_h &= S^2_h\tilde{g}(\cdot,(n+1/2)\Delta t).
-\end{split}
-\end{equation}
-
-Then
+where $w(t)\in\mathbb R^{k_1}$ and $z(t)\in\mathbb R^{k_2}$ are smooth vector-valued functions in time.
+Also, let $(u_h,v_h)$ denote the corresponding finite difference solution with singular source approximates $(f_h,g_h)$ of order $q$.
+Then, for any smooth test functions $(\tilde f,\tilde g)$, we have the following error estimate
 \begin{equation}\label{eq:weak_conv}
@@
 						 	   \langle v_h^{n+1/2}, \tilde g_h^{n+1/2}  \rangle \Big\} \right|\\
-	&= O(\Delta t^2 + h^q + (\Delta t^2+h^{p})h^{-N^*-d/2}),
-\end{split}
-\end{equation}
+	&= O(\Delta t^2 + h^q + (\Delta t^2+h^{2p})h^{-N^*-d/2}),
+\end{split}
+\end{equation}
+with $N^* = \max\{|\mathbf s_1|,|\mathbf s_2|\}$. 
+Again, $(\tilde f_h,\tilde g_h)$ as given by \ref{eq:tilde_f}. 
 \end{theorem}
@@
 \begin{proof}
-Denote by $(\tilde u, \tilde v)$ the unique anticausal solution of the adjoint system \ref{eq:adj_system}
-with right-hand sides $(\tilde f,\tilde g)$. Likewise, denote by $(\tilde{u}^{n-1/2}_h,\tilde{v}^n_h)$ the unique
-anticausal solution of the discrete adjoint system \ref{eqn:inadj}
-with right-hand sides $(\tilde{f}^{n}_h,\tilde{g}^{n+1/2}_h) $ as in
-equation \ref{eq:defadjrhs}.
-
-Using \ref{eqn:weak}, and its continuum version \ref{eq:adj_system}, error $\mathcal E$ is rewritten in terms of inner products with the singular source terms, both in the continuum and discrete sense, i.e.,
+Using \ref{eqn:weak}, and its continuum version, error $\mathcal E$ is rewritten in terms of inner products with the singular source terms, both in the continuum and discrete sense, i.e.,
 \[
@@
 Note \hl{$(\tilde u,\tilde v)$} are solutions to problem \ref{eq:cont_system} with smooth source terms 
-$(-\tilde f,-\tilde g)$, and $(\tilde u_h,\tilde v_h)$ is the respective staggered grid finite difference solution with sources $(-\tilde f_h,-\tilde g_h)$.
+$(-\tilde f,-\tilde g)$, and $(\tilde u_h,\tilde v_h)$ is the respective staggered grid finite difference solutions with sources $(-\tilde f_h,-\tilde g_h)$.
 Furthermore,
@@
 \mathcal E &= \left| \int_0^T  \Big\{ \langle f,\tilde u \rangle + \langle g,\tilde v\rangle \Big\} dt -
-		\Delta t \sum_{n=0}^N \Big\{ \langle f_h^{n+1/2}, \tilde u_h^{n+1/2} \pm S_{h}^1\tilde u(\cdot,(n+1/2)\Delta t) \rangle +
-						 	   \langle g_h^{n}, \tilde v_h^{n} \pm S_{h}^2 \tilde v(\cdot,n\Delta t) \rangle \Big\} \right|\\
+		\Delta t \sum_{n=0}^N \Big\{ \langle f_h^{n+1/2}, \tilde u_h^{n+1/2} \pm S_{1,h}\tilde u^{n+1/2} \rangle +
+						 	   \langle g_h^{n}, \tilde v_h^{n} \pm S_{2,h}\tilde v^{n} \rangle \Big\} \right|\\
 	& \le \mathcal E_1 + \mathcal E_2.
@@
 \mathcal E_1 := \left|  \int_0^T  \Big\{ \langle f,\tilde u \rangle + \langle g,\tilde v\rangle \Big\} dt
-			-\Delta t \sum_{n=0}^{N} \Big\{  \langle f_h^{n+1/2}, S_{h}^1\tilde u(\cdot,(n+1/2)\Delta t) \rangle +
-			\langle g_h^n, S_{h}^2 \tilde v(\cdot,n\Delta t) \rangle \Big\} \right|
-\]
-\[
-\mathcal E_2 := \left| \Delta t \sum_{n=0}^N \Big\{ \langle f_h^{n+1/2}, 
-				\tilde u_h^{n+1/2} - S_h^1 \tilde u(\cdot,(n+1/2)\Delta t) \rangle 
-			+ \langle g_h^n, \tilde v_h^n - S_h^2 \tilde v(\cdot,n\Delta t) \rangle \Big\} \right|.
+			-\Delta t \sum_{n=0}^{N} \Big\{  \langle f_h^{n+1/2}, S_{1,h}\tilde u^{n+1/2} \rangle +
+			\langle g_h^n, S_{2,h} \tilde v^{n} \rangle \Big\} \right|
+\]
+\[
+\mathcal E_2 := \left| \Delta t \sum_{n=0}^N \Big\{ \langle f_h^{n+1/2}, \delta \tilde u_h^{n+1/2} \rangle +
+			\langle g_h^n, \delta \tilde v_h^n \rangle \Big\} \right|.
 \]
@@
 	\left| \int_0^T \langle f, \tilde u\rangle \;  dt -
-		\Delta t \sum_{n=0}^N \langle f_h^{n+1/2}, S_{h}^1\tilde u(\cdot,(n+1/2)\Delta t)\rangle \right| \\
-	\le
-	\left| \int_0^T \langle f,\tilde u \rangle \; dt - \Delta t \sum_{n=0}^N \langle f(\cdot,(n+1/2)\Delta t),\tilde u(\cdot,(n+1/2)\Delta t)\rangle \right| \\
-	+ \left| \Delta t \sum_{n=0}^N \Big\{ \langle f_h^{n+1/2}, S_{h}^1\tilde u(\cdot,(n+1/2)\Delta t)\rangle - 
-							     \langle f(\cdot,(n+1/2)\Delta t), \tilde u(\cdot,(n+1/2)\Delta t) \rangle \Big\} \right |,
+		\Delta t \sum_{n=0}^N \langle f_h^{n+1/2}, S_{1,h}\tilde u^{n+1/2}\rangle \right|
+	\le \\
+	\left| \int_0^T \langle f,\tilde u \rangle \; dt - \Delta t \sum_{n=0}^N \langle f^{n+1/2},\tilde u^{n+1/2}\rangle \right|
+	+ \left| \Delta t \sum_{n=0}^N \Big\{ \langle f_h^{n+1/2}, S_{1,h}\tilde u^{n+1/2}\rangle - 
+							     \langle f^{n+1/2}, \tilde u^{n+1/2} \rangle \Big\} \right |,
 \end{split}
@@
 \[
-	\left| \int_0^T \langle f,\tilde u \rangle \; dt - \Delta t \sum_{n=0}^N \langle f(\cdot,(n+1/2)\Delta t),\tilde u(\cdot,(n+1/2)\Delta t)\rangle \right| = O(\Delta t^2)
+	\left| \int_0^T \langle f,\tilde u \rangle \; dt - \Delta t \sum_{n=0}^N \langle f^{n+1/2},\tilde u^{n+1/2}\rangle \right| = O(\Delta t^2)
 \]
@@
 \[
-	\left| \Delta t \sum_{n=0}^N \Big\{ \langle f_h^{n+1/2}, S_{h}^1\tilde u(\cdot,(n+1/2)\Delta t)\rangle - 
-							     \langle f(\cdot,(n+1/2)\Delta t), \tilde u(\cdot,(n+1/2)\Delta t) \rangle \Big\} \right |
+	\left| \Delta t \sum_{n=0}^N \Big\{ \langle f_h^{n+1/2}, S_{1,h}\tilde u^{n+1/2}\rangle - 
+							     \langle f^{n+1/2}, \tilde u^{n+1/2} \rangle \Big\} \right |
 	= O(h^q)
@@
 
-We bound $\mathcal E_2$ using $L^2$-error estimates of finite difference solutions for smooth problems (i.e., theorem \ref{thm:smconvg}), and using $L^2$-bounds of the discrete singular source approximations; one can show
-\[
- 	\| \eta_h^{\bf s_{\bf i}}\| = O(h^{-|{\bf s_{\bf i}}|-d/2}), \; i=1,2.
-\]
-%The abounds above need a summation over  |s_i| <= N^*
-%Maybe consider simpler source case of a single multipole ?? MJB
+We bound $\mathcal E_2$ using $L^2$-error estimates of finite difference solutions for smooth problems \hl{(i.e., equation }\ref{eqn:truncadj} \hl{)}, and using $L^2$-bounds of the discrete singular source approximations; one can show
+\[
+ 	\| f_h\|_2 = O(h^{-|\mathbf s_1|-d/2}) \quad {\rm and} \quad 
+	\| g_h\|_2 = O(h^{-|\mathbf s_2|-d/2}).
+\]
 Whence,
-\begin{equation*}
-\begin{split}
-	\mathcal E_2 &\le \Delta t \sum_{n=0}^N \Big\{ \|f_h^{n+1/2}\| \; 
-				\| \tilde u_h^{n+1/2} - S_h^1 \tilde u(\cdot,(n+1/2)\Delta t) \|
-			+ \|g_h^{n}\|  \; \| \tilde v_h^n - S_h^2 \tilde v(\cdot,n\Delta t) \| \Big\}\\
-					&= O((\Delta t^2+h^p)h^{-N^*-d/2}).
-\end{split}
-\end{equation*}
+\[
+	\mathcal E_2 \le \Delta t \sum_{n=0}^N \Big\{ \|f_h^{n+1/2}\|_2 \; \|\delta \tilde u_h^{n+1/2}\|_{2}
+								+ \|g_h^{n}\|_2  \; \| \delta \tilde v_h^n\|_2 \Big\}
+					= O((\Delta t^2+h^{2p})h^{-N^*-d/2}).
+\]
 Estimate \ref{eq:weak_conv} thus follows.
@@
 Consider the simpler case where only multipoles of order zero are present, that is $N^*=0$.
-According to estimate \ref{eq:weak_conv}, we have weak convergence at rate $2-d/2$ if we choose $q=p$ and $\Delta t = O(h)$.
+According to estimate \ref{eq:weak_conv}, we have weak convergence at rate $2-d/2$ if we choose $q=2p$ and $\Delta t = O(h)$.
 In other words, convergence (in the weak sense) is guaranteed though at a suboptimal rate.
-For multipoles of order $N^*>0$, to retrieve the smooth solution behavior, we would require $p > N^*+d/2$ and $\Delta t\to 0$ like a positive power of $h$.
+For multipoles of order $N^*>0$, to retrieve the smooth solution behavior, we would require $2p > N^*+d/2$ and $\Delta t\to 0$ like a positive power of $h$.
 
@@
 \[
-	\| \eta_h^{\mathbf s_{\bf i}}\|_1 = O(h^{-|\mathbf s_{\bf i}|}), \; i=1,2.
+	\| f_h\|_1 = O(h^{-|\mathbf s_1|}) \quad {\rm and} \quad \|g_h\|_1 = O(h^{-|\mathbf s_2|}).
 \] 
@@
 \label{eq:weak_conv_infty}
-\begin{split}
-	\mathcal E_2 &\le \Delta t \sum_{n=0}^N \Big\{ \|f_h^{n+1/2}\|_1 \; 
-			\| \tilde u_h^{n+1/2} - S_h^1 \tilde u(\cdot,(n+1/2)\Delta t) \|_{\infty}
-			+ \|g_h^{n}\|_1  \; \| \tilde v_h^n - S_h^2 \tilde v(\cdot,n\Delta t) \|_{\infty} \Big\}\\
-					&= O((\Delta t^2+h^{p})h^{-N^*}).
-\end{split}
-\end{equation}
-We conjecture point-wise bounds hold for smooth $\tilde u$ and $\tilde v_h$, whence
+	\mathcal E_2 \le \Delta t \sum_{n=0}^N \Big\{ \|f_h^{n+1/2}\|_1 \; \|\delta \tilde u_h^{n+1/2}\|_{\infty}
+								+ \|g_h^{n}\|_1  \; \| \delta \tilde v_h^n\|_{\infty} \Big\}
+					= O((\Delta t^2+h^{2p})h^{-N^*}).
+\end{equation}
+We conjecture point-wise bounds hold for smooth $\delta \tilde u$ and $\delta \tilde v$, whence
 \begin{equation}\label{eq:conjecture}
-	\mathcal E = O(\Delta t^2 +h^q + (\Delta t^2+h^{p})h^{-N^*}),
-\end{equation}
-yielding optimal rates for the $N^*=0$ case, again if $q\ge p$.
-Note that \ref{eq:weak_conv_infty} follows through if $L^\infty$ error terms $\| \tilde u_h - S_h^1 u \|_\infty$ and $\| \tilde v_h - S_h^2 v \|_\infty$ have optimal rates.
-The authors are not aware of any such $L^\infty$ error
-estimates for the types of hyperbolic systems considered here, see
-however \cite{Brenner:06} and \cite{Layton:82} for some work in this
-direction.
+	\mathcal E = O(\Delta t^2 +h^q + (\Delta t^2+h^{2p})h^{-N^*}),
+\end{equation}
+yielding optimal rates for the $N^*=0$ case, again if $q\ge2p$.
+Note that \ref{eq:weak_conv_infty} follows through if $L^\infty$ error terms $\|\delta \tilde u_h\|_\infty$ and $\|\delta \tilde v_h\|_\infty$ have optimal rates.
+The authors are not, however, aware of any such $L^\infty$ error estimates for the types of hyperbolic systems considered here. 
+One could potentially attempt to prove stability in $L^\infty$ in order to apply the Lax equivalence theorem and imply $L^\infty$ estimates (see \cite{Brenner:06}) or alternatively  
+proof $L^\infty$ estimates requiring only $L^2$ stability (see \cite{Layton:82}).
 
@@
 \end{equation}
-where $u=p$ is the scalar pressure field and $v=\mathbf v\in\mathbf R^d$ is the vector particle velocity field.
-Coefficient operator $A(\mathbf x)=1/\kappa(\mathbf x)I$ and $B(\mathbf x)=\rho(\mathbf x) I$, with $\kappa$ denoting bulk-modulus and $\rho$ density of the medium; here $I\in\mathbf R^{d\times d}$ is the identity matrix.
+where $u=p$ is the scalar pressure field and $v=\mathbf v\in\mathbb R^d$ is the vector particle velocity field; we take \hl{$\mathcal H_1=H^1_0(\Omega)$ and $\mathcal H_2=H^1(\Omega)^d$}. %MOD
+Coefficient operator $A(\mathbf x)=1/\kappa(\mathbf x)$ and $B(\mathbf x)=\rho(\mathbf x) I$, with $\kappa$ denoting bulk-modulus and $\rho$ density of the medium; here $I\in\mathbb R^{d\times d}$ is the identity matrix.
 Lastly, the differential operator $P$ coincides with the gradient and its adjoint with the negative of the divergence,
@@
 \]
-
-For the remainder of this section, we assume that the spatial
-dimension $d=2$. 
+%ADDED
+\hl{
+Given homogenous boundary conditions for the pressure field, it can be shown that the gradient operator $P$ satisfies skew-adjoint relation} \ref{eq:skew}.
+
+For simplicity we assume we are dealing with rectangular grids, in 2-D for this example, of the form
+\[
+	\mathcal G(\mathbf 0,\mathbf h) = \{ \mathbf x_{i,j} = (ih,jh) \; : \; i,j\in\mathbb Z \}.
+\]
 The simplest staggered-grid finite difference scheme, second-order in time and space, is given by
@@
 We emphasize that the velocity fields for finite difference scheme \ref{eq:acoustics_sgfd} are staggered with respect to spatial grids, more specifically each component is shifted by half a cell size in its respective axis.
+
 Relating system \ref{eq:acoustics_sgfd} with \ref{eq:sg}, we see
@@
 \]
-The space $H_h^1$ in this example corresponds to the set of square summable 
-scalar-valued functions on rectangular grid
-\[
-	\mathcal G((0,0),(1,1)) = \{ \mathbf x_{i,j} = (ih,jh) \; : \; i,j\in\mathbf Z \}.
-\]
-In other words, $H_h^1 = \mathcal H_h((0,0),(1,1))$.
-$H_h^2$ is the set of square summable $\mathbf R^2$-valued functions on half-cell shifted grids, that is, 
-\[
-	H_h^2 = \mathcal H_h((1/2,0),(1,1)) \times \mathcal H_h((0,1/2),(1,1)).
-\]
-
-%\begin{equation*}
-%\begin{split}
-%	\langle u_h,\tilde u_h\rangle = h^2 \sum_{i,j} (u_h)_{i,j} (\tilde u_h)_{i,j} &\quad 	\text{for } u_h,\tilde u_h \in H_h^1\\
-%	\langle v_h, \tilde v_h \rangle = h^2 \sum_{i,j} (v_h)_{i,j} \cdot (\tilde v_h)_{i,j} &\quad
-%	\text{for } v_h,\tilde v_h \in H_h^2.
-%\end{split}
-%\end{equation*}
-%Inner-products and norms of the spaces $\mathcal H_1, \mathcal H_2, H_{1,h}, H_{2,h}$ are all denoted by $\langle\cdot,\cdot\rangle$ and $\|\cdot\|$ and interpreted given the context, unless otherwise specified.
-
-Sampling operators $S_h^1$ and $S_h^2$ coincide with evaluating continuum functions on grid or shifted grid points accordingly,
-\[
-	(S_h^1 u)_{i,j} = u(ih,jh), \quad 
-	(S_h^2 v)_{i,j} = \mat{v_{1}((i+1/2)h,jh)\\ 
+The space $H_{1,h}$ corresponds to the set of square summable scalar-valued
+%MOD
+functions on rectangular grids \hl{that satisfy the homogenous boundary condition}, and $H_{2,h}$ is the set of square summable $\mathbb R^2$-valued functions on half-cell shifted grids, both equipped with a discrete $L^2$ inner-products,
+\begin{equation*}
+\begin{split}
+	\langle u_h,\tilde u_h\rangle = h^2 \sum_{i,j} (u_h)_{i,j} (\tilde u_h)_{i,j} &\quad 	\text{for } u_h,\tilde u_h \in H_{1,h}\\
+	\langle v_h, \tilde v_h \rangle = h^2 \sum_{i,j} (v_h)_{i,j} \cdot (\tilde v_h)_{i,j} &\quad
+	\text{for } v_h,\tilde v_h \in H_{2,h}.
+\end{split}
+\end{equation*}
+Inner-products and norms of the spaces $\mathcal H_1, \mathcal H_2, H_{1,h}, H_{2,h}$ are all denoted by $\langle\cdot,\cdot\rangle$ and $\|\cdot\|$ and interpreted given the context, unless otherwise specified.
+
+Sampling operators $S_{1,h}$ and $S_{2,h}$ coincide with evaluating continuum functions on grid or shifted grid points accordingly,
+\[
+	(S_{1,h} u)_{i,j} = u(ih,jh), \quad 
+	(S_{2,h} v)_{i,j} = \mat{v_{1}((i+1/2)h,jh)\\ 
 				  v_{2}(ih,(j+1/2)h) }.
@@
 with $r =  \Delta t/h$.
-In general, the differential operator $P$ can be approximated by a family of central difference operators $P_h$ of even order $p=2p^*$ for $p^*\in\mathbf N$, resulting in the $(2,2p^*)$ staggered-grid finite difference schemes.
-
-%\hl{It follows that our weak convergence analysis applies to this example.}
+In general, the differential operator $P$ can be approximated by a family of central difference operators $P_h$ of even order $2p$ for $p\in\mathbb N$, resulting in the $2$-$2p$ staggered-grid finite difference schemes.
+%ADDED
+\hl{Similar to the continuum case, it can be shown that the $P_h$ for these type of staggered grid schemes indeed satisfy skew-adjoint relation} \ref{eq:skew_h}.
+\hl{It follows that our weak convergence analysis applies to this example.}
 
@@
 for the acoustic system \ref{eq:acoustics} \cite[]{Vir:84}, of order 2 in time and orders 2 and 4 in space - we
-refer to these as the $(2,2)$ and $(2,4)$ order schemes respectively. In these
+refer to these as the 2-2 and 2-4 order schemes respectively. In these
 experiments we use only scalar (pressure) sources and pressure trace
@@
 convergence for the difference schemes, in most cases the nominal
-spatial order ($q=2$ for the $(2,2)$ scheme, $q=4$ for the $(2,4)$ scheme). 
+spatial order ($q=2$ for the 2-2 scheme, $q=4$ for the 2-4 scheme). 
 Again, the time step is fixed small enough that the time discretization error plays
@@
 \end{itemize}
-%Note that the coarsest grid cell size is $40m$ which implies that at least $15$ 
-%grid {\em points per wavelength (gpw)} at peak frequency of $5Hz$ or $5$ gpw at $15Hz$ 
-%(see figure \ref{fig:wlt-ricker-fft}), in all numerical experiments, 
-%thus minimizing the effects of grid dispersion on approximated convergence rates.
+Note that the coarsest grid cell size is $40m$ which implies that at least $15$ 
+grid {\em points per wavelength (gpw)} at peak frequency of $5Hz$ or $5$ gpw at $15Hz$ 
+(see figure \ref{fig:wlt-ricker-fft}), in all numerical experiments, 
+thus minimizing the effects of grid dispersion on approximated convergence rates.
 
@@
 over the entire domain and at a particular depth of $z=2000$ in the left and right column plots respectively.
-The first row of graphs shows results when using the $(2,2)$ finite difference scheme and the second-order source approximation.
+The first row of graphs shows results when using the 2-2 finite difference scheme and the second-order source approximation.
 We observe that second order convergence is indeed achieved away from the source.
-Similarly, fourth order convergence results are observed when using the $(2,4)$ finite difference scheme and the fourth-order source approximation, i.e., second row in figure 5.
+Similarly, fourth order convergence results are observed when using the 2-4 finite difference scheme and the fourth-order source approximation, i.e., second row in figure 5.
 Lastly, the third row of plots demonstrates the negative effects of using a lower order source approximation relative to the spatial finite difference order, namely a second-order source approximation with a fourth-order method in space.
@@
 
+%Convergence rates versus receiver horizontal location for 2-D and 3-D cases, 
+%when using the 2-2 finite difference scheme and an
+%$O(h^2)$ approximation for the various multipoles, are plotted in figures  
+%\ref{fig:test-s00q2-fd2-plots} -- \ref{fig:test-s02q2-fd2-plots} and 
+%\ref{fig:test-s000q2-fd2-plots} -- \ref{fig:test-s020q2-fd2-plots}.
+%Second-order convergence is observed for all cases in both the $\|\cdot\|_2$ (solid blue curve) and $\|\cdot\|_\infty$ (dashed red curve) norms away from the source location.
+%Analogous results for the 2-4 finite difference scheme in conjunction with an $O(h^4)$ singular source approximation 
+%are shown in figures 
+%\ref{fig:test-s00q4-fd4-plots} -- \ref{fig:test-s02q4-fd4-plots} and 
+%\ref{fig:test-s000q4-fd4-plots} -- \ref{fig:test-s020q4-fd4-plots},
+%where the target quartic convergence rate is again achieved away from the source.
+
+%Lastly, the last set of tests consider the case where a lower singular source approximation order
+%is used, relative to the spatial finite difference order.
+%In particular, figures 
+%\ref{fig:test-s00q2-fd4-plots} -- \ref{fig:test-s02q2-fd4-plots} and 
+%\ref{fig:test-s000q2-fd4-plots} -- \ref{fig:test-s020q2-fd4-plots}
+%plot the convergence rates for the case when using the 2-4 
+%finite difference scheme with an $O(h^2)$ source approximation.
+
+%The onset distance of optimal order convergence is consistent with the
+%theoretical prediction, that is, roughly $h(q+|{\bf s}|)$ for the coarses $h$ used in
+%the convergence study ($h=40m$), which is the same as the trace spacing.
+
+%Test 6 demonstates the negative effects of using the wrong multipole
+%approximation order, namely $O(h^2)$ (i.e., tensor product of shifted hat functions)
+%for a fourth order finite difference method in 2-D. 
+%The theory makes no prediction in this case; apparently the
+%convergence rate can be lower than would be the case with a difference
+%scheme of lower order (second, in this case), even at a substantial
+%distance from the source. We interpret this as meaning that 
+%higher-order difference operators create larger truncation error when
+%applied to discrete multipole sources of lower approximation
+%order. This is not what one would expect of a discretely sampled smooth 
+%right hand side, and is a matter worth further study.
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 Numerical results presented here validate the accuracy of moment-consistent singular source discretizations for controlling the propagation of finite difference truncation error for multipole sources.
-In particular, optimal spatial convergence rates for the $(2,2)$ and $(2,4)$ staggered grid finite difference methods are achieved, point-wise in space away from source location, when the source approximation order is equal to that of the spatial finite difference approximation order.
+In particular, optimal spatial convergence rates for the 2-2 and 2-4 staggered grid finite difference methods are achieved, point-wise in space away from source location, when the source approximation order is equal to that of the spatial finite difference approximation order.
 
@@
 
-As mentioned in the Literature Review, \cite{Petersson:2016}
-demonstrated that optimal convergence rates can be achieved if the
-source discretization satisfied moment as well as smoothness
-conditions. \cite{Petersson:2016} used central difference
-approximations to the space derivatives and Runge-Kutta time
-integration, for various advection problems and for acoustics. The
-staggered grid schemes used in our work, however, do not require
-additional constraints beyond the moment conditions for the multipole
-source approximations.
-
-%As our numerical tests demonstrate, smoothness conditions were not necessary here, which is attributed to the fact that we use staggered-grid finite difference schemes. 
-%In particular, the central finite difference operator used in \cite{Petersson:2016} have an associated grid spacing of $2h$ with respect to an $h$-sized grid, coupled with a source approximation of narrow support, i.e., $2\epsilon=(q+|\mathbf s|)h$, results in the triggering of spurious modes unresolved by the numerical solution.
-%The finite difference operators used here however have a grid spacing of $h$ given that they approximate derivatives over staggered grids.
-%Thus, in essence our narrow source approximations ``seem'' twice as wide, and therefore smoother, from the point of view of the difference operator $P_h$.
-%For this reason, smoothness conditions are not necessary for staggered-grid finite difference schemes.
-
+Work by \cite{Petersson:2016} demonstrated that optimal convergence rates can be achieved if the source discretization satisfied moment as well as smoothness conditions.
+Again, their work was centered around central difference approximations of the 1-D advection equation.
+As our numerical tests demonstrate, smoothness conditions were not necessary here, which is attributed to the fact that we use staggered-grid finite difference schemes. 
+In particular, the central finite difference operator used in \cite{Petersson:2016} have an associated grid spacing of $2h$ with respect to an $h$-sized grid, coupled with a source approximation of narrow support, i.e., $2\epsilon=(q+|\mathbf s|)h$, results in the triggering of spurious modes unresolved by the numerical solution.
+The finite difference operators used here however have a grid spacing of $h$ given that they approximate derivatives over staggered grids.
+Thus, in essence our narrow source approximations ``seem'' twice as wide, and therefore smoother, from the point of view of the difference operator $P_h$.
+For this reason, smoothness conditions are not necessary for staggered-grid finite difference schemes.
+
+%, which is the same as the trace spacing.
+%For example, consider test 12 (figure \ref{fig:test12-plots}) 
+%where $D^{\bf s}\delta({\bf x}-{\bf x}^*)$ with $|{\bf s}|=2$ is replaced by
+%a fourth order approximation in 3-D; the source stencil had a diameter of about
+%$q+|{\bf s}|=6$ traces, equivalent to 6 grid points in the $x_2$-direction for the coarsest
+%($40m$) mesh.
+
+%the observed region of optimal order convergence is also roughly as predicted.
+%We note that accuracy claims made above are in relation to receiver positions away from the source stencil.
+%Source approximations with large source stencils are problematic since they produce large regions where the numerical solution is not guaranteed to be as accurate.
+
+%%In particular, we observe good agreement between finite difference and
+%%analytical solutions, see figures \ref{fig:comp2D-00} --
+%%\ref{fig:comp2D-01} and figures \ref{fig:comp3D-000} --
+%%\ref{fig:comp3D-020}; this is to be expected given since our choice of
+%%grid cell size $h=40m$, or equivalently $15$ grid points per wavelength at peak frequency of $5Hz$, 
+%%and time step size $\Delta t=0.5ms$.
+%
 %
@@
 \hl{
-The weak convergence theory relied on the structure of the the general continuum problem} \ref{eq:cont_system} \hl{and the ability for its discretization to preserve this structure, i.e, coefficient operators ($A,B,A_h,$ and $B_h$) are symmetric and bounded.% and the differential operators $P$ and $P_h$ satisfy a skew-adjoint relation.
+The weak convergence theory relied on the structure of the the general continuum problem} \ref{eq:cont_system} \hl{and the ability for its discretization to preserve this structure, i.e, coefficient operators ($A,B,A_h,$ and $B_h$) are symmetric and bounded and the differential operators $P$ and $P_h$ satisfy a skew-adjoint relation.
 The general staggered finite difference scheme} \ref{eq:sg} \hl{is also assumed to be staggered in time, in particular a central difference approximation of the time derivative was applied.
@@
 
-\multiplot{6}{ssappx-s00q2,ssappx-s00q4,ssappx-s01q2,ssappx-s01q4,ssappx-s02q2,ssappx-s02q4}{width=0.33\textwidth}{Plots of $\eta^\epsilon(\mathbf x)$, 2-D approximations to $D^{\mathbf s}\delta(\mathbf x)$ with $\mathbf l=(1,1)$ and $h=1$.
+\multiplot{6}{ssappx-s00q2,ssappx-s00q4,ssappx-s01q2,ssappx-s01q4,ssappx-s02q2,ssappx-s02q4}{width=0.33\textwidth}{Plots of $\eta^\epsilon(\mathbf x)$, 2-D approximations to $D^{\mathbf s}\delta(\mathbf x)$ with $\mathbf h=(1,1)$.
 (a) $\mathbf s=(0,0)$ and $q=2$, (b) $\mathbf s=(0,0)$ and $q=4$, (c) $\mathbf s=(0,1)$ and $q=2$, (d) $\mathbf s=(0,1)$ and $q=4$, (e) $\mathbf s=(0,2)$ and $q=2$, (f) $\mathbf s=(0,2)$ and $q=4$.}
@@
 
-\multiplot{3}{sq-test-s00q2-fd2-plotsnap,sq-test-s01q2-fd2-plotsnap,sq-test-s02q2-fd2-plotsnap}{width=0.48\textwidth}{Snapshot of pressure field ($t=0.75$ sec) computed using $(2,2)$ finite difference scheme ($h=10m, \Delta t=0.5$ ms) and second-order approximation of multipole source $D^{\bf s}\delta({\bf x}-{\bf x^*})$: (a) ${\bf s}=(0,0)$, (b) ${\bf s}=(0,1)$, and (c) ${\bf s}=(0,2)$.}
-
-\multiplot{9}{sq-test-s00q2-fd2-plots,sq-test-s00q2-fd2-plots-wind,sq-test-s00q2-fd2-plots-wind2,sq-test-s00q4-fd4-plots,sq-test-s00q4-fd4-plots-wind,sq-test-s00q4-fd4-plots-wind2,sq-test-s00q2-fd4-plots,sq-test-s00q2-fd4-plots-wind,sq-test-s00q2-fd4-plots-wind2}
-{width=0.25\textwidth}{Convergence rates results for ${\mathbf s}=(0,0)$.
-Using (a) $(2,2)$ FD scheme and second-order source approximation, (d) $(2,4)$ FD scheme and fourth-order source approximation, (g) $(2,4)$ FD scheme and second-order source approximation.
-Plots (b), (e), (h) correspond to convergence rates at $z=2000$, related to the first column from the left.
-Similarly, plots (c), (f), (i) show rates at $z=-1600$.}
-
-\multiplot{9}{sq-test-s01q2-fd2-plots,sq-test-s01q2-fd2-plots-wind,sq-test-s01q2-fd2-plots-wind2,sq-test-s01q4-fd4-plots,sq-test-s01q4-fd4-plots-wind,sq-test-s01q4-fd4-plots-wind2,sq-test-s01q2-fd4-plots,sq-test-s01q2-fd4-plots-wind,sq-test-s01q2-fd4-plots-wind2}
-{width=0.25\textwidth}{Convergence rate results for ${\mathbf s}=(0,1)$.
-Using (a) $(2,2)$ FD scheme and second-order source approximation, (d) $(2,4)$ FD scheme and fourth-order source approximation, (g) $(2,4)$ FD scheme and second-order source approximation.
-Plots (b), (e), (h) correspond to convergence rates at $z=2000$, related to the first column from the left.
-Similarly, plots (c), (f), (i) show rates at $z=-1600$.}
-
-\multiplot{9}{sq-test-s02q2-fd2-plots,sq-test-s02q2-fd2-plots-wind,sq-test-s02q2-fd2-plots-wind2,sq-test-s02q4-fd4-plots,sq-test-s02q4-fd4-plots-wind,sq-test-s02q4-fd4-plots-wind2,sq-test-s02q2-fd4-plots,sq-test-s02q2-fd4-plots-wind,sq-test-s02q2-fd4-plots-wind2}
-{width=0.25\textwidth}{Convergence rate results for ${\mathbf s}=(0,2)$.
-Using (a) $(2,2)$ FD scheme and second-order source approximation, (d) $(2,4)$ FD scheme and fourth-order source approximation, (g) $(2,4)$ FD scheme and second-order source approximation.
-Plots (b), (e), (h) correspond to convergence rates at $z=2000$, related to the first column from the left.
-Similarly, plots (c), (f), (i) show rates at $z=-1600$.}
+%\plot{model-geo-2D}{width=0.8\textwidth}{Model and source-receiver geometry specifications for 2-D convergence rate tests; constant density ($1kg/m^3$) and velocity ($3km/s$), receiver positions $(z_r,x_r)=(-200m,0m:40m:6000m)$, source position $(z^*,x^*)=(-203m,3003m)$.}
+
+%\plot{model-geo-3D}{width=0.8\textwidth}{Model and source-receiver geometry specifications for 3-D convergence rate tests; constant density ($\rho=1kg/m^3$) and velocity ($c=3km/s$), receiver positions $(z_r,x_r,y_r)=(-200m,0m:40m:6000m,100m)$, source position $(z^*,x^*,y^*)=(-203m,3003m,103m)$.}
+
+%%%%%%%
+%% 2-D %
+%%%%%%%
+
+\multiplot{3}{sq-test-s00q2-fd2-plotsnap,sq-test-s01q2-fd2-plotsnap,sq-test-s02q2-fd2-plotsnap}{width=0.48\textwidth}{Snapshot of pressure field ($t=0.748$ sec) computed using 2-2 finite difference scheme ($h=10m, \Delta t=0.5$ ms) and second-order approximation of multipole source $D^{\bf s}\delta({\bf x}-{\bf x^*})$: (a) ${\bf s}=(0,0)$, (b) ${\bf s}=(0,1)$, and (c) ${\bf s}=(0,2)$.}
+
+\multiplot{6}{sq-test-s00q2-fd2-plots,sq-test-s00q2-fd2-plots-wind,sq-test-s00q4-fd4-plots,sq-test-s00q4-fd4-plots-wind,sq-test-s00q2-fd4-plots,sq-test-s00q2-fd4-plots-wind}
+{width=0.33\textwidth}{Convergence rate results for ${\mathbf s}=(0,0)$.
+Using 2-2 FD scheme and second-order source approximation (a),  2-4 FD scheme and fourth-order source approximation (c), 2-4 FD scheme and second-order source approximation (e).
+Plots on right column correspond to convergence rates at $z=2000$, related to the left column.}
+
+\multiplot{6}{sq-test-s01q2-fd2-plots,sq-test-s01q2-fd2-plots-wind,sq-test-s01q4-fd4-plots,sq-test-s01q4-fd4-plots-wind,sq-test-s01q2-fd4-plots,sq-test-s01q2-fd4-plots-wind}
+{width=0.33\textwidth}{Convergence rate results for ${\mathbf s}=(0,1)$.
+Using 2-2 FD scheme and second-order source approximation (a),  2-4 FD scheme and fourth-order source approximation (c), 2-4 FD scheme and second-order source approximation (e).
+Plots on right column correspond to convergence rates at $z=2000$, related to the left column.}
+
+\multiplot{6}{sq-test-s02q2-fd2-plots,sq-test-s02q2-fd2-plots-wind,sq-test-s02q4-fd4-plots,sq-test-s02q4-fd4-plots-wind,sq-test-s02q2-fd4-plots,sq-test-s02q2-fd4-plots-wind}
+{width=0.33\textwidth}{Convergence rate results for ${\mathbf s}=(0,2)$.
+Using 2-2 FD scheme and second-order source approximation (a),  2-4 FD scheme and fourth-order source approximation (c), 2-4 FD scheme and second-order source approximation (e).
+Plots on right column correspond to convergence rates at $z=2000$, related to the left column.}
+
+%\multiplot{3}{test-s00q2-fd2-data10,test-s01q2-fd2-data10,test-s02q2-fd2-data10}{width=0.48\textwidth}{2-D pressure field data using 2-2 finite difference scheme ($h=10m$) and $O(h^2)$ approximation of multipole source $D^{\bf s}\delta({\bf x}-{\bf x^*})$: ${\bf s}=(0,0)$ in (a), ${\bf s}=(0,1)$ in (b), and ${\bf s}=(0,2)$ in (c).}
+%
+%\multiplot{3}{test-s00q2-fd2-plots,test-s01q2-fd2-plots,test-s02q2-fd2-plots}{width=0.48\textwidth}{2-D convergence rate results using 2-2 finite difference scheme and $O(h^2)$ approximation for multipole source $D^{\bf s}\delta({\bf x}-{\bf x^*})$.
+$L^2$- (solid blue) and $L^{\infty}$- (dashed red) norms.
+(a) ${\bf s} = (0,0)$, (b) ${\bf s}= (0,1)$, and (c) ${\bf s}= (0,2)$.}
+%
+%\multiplot{3}{test-s00q4-fd4-plots,test-s01q4-fd4-plots,test-s02q4-fd4-plots}{width=0.48\textwidth}{2-D convergence rate results using 2-4 finite difference scheme and $O(h^4)$ approximation for multipole source $D^{\bf s}\delta({\bf x}-{\bf x^*})$.
+$L^2$- (solid blue) and $L^{\infty}$- (dashed red) norms.
+(a) ${\bf s} = (0,0)$, (b) ${\bf s}= (0,1)$, and (c) ${\bf s}= (0,2)$.}
+%
+%\multiplot{3}{test-s00q2-fd4-plots,test-s01q2-fd4-plots,test-s02q2-fd4-plots}{width=0.48\textwidth}{2-D convergence rate results using 2-4 finite difference scheme and $O(h^2)$ approximation for multipole source $D^{\bf s}\delta({\bf x}-{\bf x^*})$.
+$L^2$- (solid blue) and $L^{\infty}$- (dashed red) norms.
+(a) ${\bf s} = (0,0)$, (b) ${\bf s}= (0,1)$, and (c) ${\bf s}= (0,2)$.}
+
+%%%%%%
+% 3-D %
+%%%%%%
+
+%\multiplot{3}{test-s000q2-fd2-data10,test-s010q2-fd2-data10,test-s020q2-fd2-data10}{width=0.48\textwidth}{3-D pressure field data using 2-2 finite difference scheme ($h=10m$) and $O(h^2)$ approximation of multipole sources $D^{\bf s}\delta({\bf x}-{\bf x^*})$: ${\bf s} = (0,0,0)$ in (a), ${\bf s}= (0,1,0)$ in (b), and ${\bf s}=(0,2,0)$ in (c).}
+%
+%\multiplot{3}{test-s000q2-fd2-plots,test-s010q2-fd2-plots,test-s020q2-fd2-plots}{width=0.48\textwidth}{3-D convergence rate results using 2-2 finite difference scheme and $O(h^2)$ approximation for multipole source $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$.
+$L^2$- (solid blue) and $L^{\infty}$- (dashed red) norms.
+(a) ${\bf s}=(0,0,0)$, (b) ${\bf s}=(0,1,0)$, and (c) ${\bf s}=(0,2,0)$.}
+%
+%\multiplot{3}{test-s000q4-fd4-plots,test-s010q4-fd4-plots,test-s020q4-fd4-plots}{width=0.48\textwidth}{3-D convergence rate results using 2-4 finite difference scheme and $O(h^4)$ approximation for multipole source $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$.
+$L^2$- (solid blue) and $L^{\infty}$- (dashed red) norms.
+(a) ${\bf s}=(0,0,0)$, (b) ${\bf s}=(0,1,0)$, and (c) ${\bf s}=(0,2,0)$.}
+%
+%\multiplot{3}{test-s000q2-fd4-plots,test-s010q2-fd4-plots,test-s020q2-fd4-plots}{width=0.48\textwidth}{3-D convergence rate results using 2-4 finite difference scheme and $O(h^2)$ approximation for multipole source $D^{\mathbf s}\delta(\mathbf x-\mathbf x^*)$.
+$L^2$- (solid blue) and $L^{\infty}$- (dashed red) norms.
+(a) ${\bf s}=(0,0,0)$, (b) ${\bf s}=(0,1,0)$, and (c) ${\bf s}=(0,2,0)$.}
 

### 2019-10-19 repro/papers/ms/f.tex
repro docs dir added

\title{Figures}
\author{William W. Symes}

\begin{abstract}
abstract goes here
\end{abstract}

\inputdir{project}

\plot{res0}{width=\textwidth}{res0}
\plot{res1}{width=\textwidth}{res1}
\plot{res2}{width=\textwidth}{res2}
\plot{res3}{width=\textwidth}{res3}
\plot{res4}{width=\textwidth}{res4}
\plot{res5}{width=\textwidth}{res5}
\plot{res6}{width=\textwidth}{res6}
\plot{res7}{width=\textwidth}{res7}
\plot{res8}{width=\textwidth}{res8}
\plot{res9}{width=\textwidth}{res9}
\plot{res10}{width=\textwidth}{res10}

\plot{shot6kmbigdecon0tscale}{width=\textwidth}{shot6kmbigdecon0tscale}
\plot{shot6kmbigdecon1tscale}{width=\textwidth}{shot6kmbigdecon1tscale}
\plot{shot6kmbigdecon2tscale}{width=\textwidth}{shot6kmbigdecon2tscale}
\plot{shot6kmbigdecon3tscale}{width=\textwidth}{shot6kmbigdecon3tscale}
\plot{shot6kmbigdecon4tscale}{width=\textwidth}{shot6kmbigdecon4tscale}
\plot{shot6kmbigdecon5tscale}{width=\textwidth}{shot6kmbigdecon5tscale}
\plot{shot6kmbigdecon6tscale}{width=\textwidth}{shot6kmbigdecon6tscale}
\plot{shot6kmbigdecon7tscale}{width=\textwidth}{shot6kmbigdecon7tscale}
\plot{shot6kmbigdecon8tscale}{width=\textwidth}{shot6kmbigdecon8tscale}
\plot{shot6kmbigdecon9tscale}{width=\textwidth}{shot6kmbigdecon9tscale}
\plot{shot6kmbigdecon10tscale}{width=\textwidth}{shot6kmbigdecon10tscale}

\plot{shot6kmbig}{width=\textwidth}{shot6kmbig}
\plot{shot6kmhom}{width=\textwidth}{shot6kmhom}
\plot{fwiscanlo}{width=\textwidth}{fwiscanlo}
\plot{ivascan}{width=\textwidth}{ivascan}
\plot{waveletbase}{width=\textwidth}{waveletbase}

### 2019-10-19 repro/papers/ms/paper.tex
repro docs dir added

\title{Solution of an Inverse Problem for the Wave Equation by means of the Source-Receiver Extension}
%\author{Guanghui Huang and William. W. Symes}

%\address{The Rice Inversion Project,
%Rice University,
%Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}

\author{
Guanghui Huang\thanks{Department of Computational and Applied Mathematics, Rice University,
Houston, TX, 77005, USA,
{\tt ghhuang@rice.edu}} \ and \ William Symes\thanks{Department of Computational and Applied Mathematics, Rice University,
Houston, TX, 77005, USA,
{\tt symes@rice.edu}}
}

\lefthead{Huang and Symes}

\righthead{Source-Receiver Extension}

\maketitle
\parskip 12pt

\begin{abstract}
  Iterative gradient-based minimization of mean-square misfit often
  results in suboptimal estimation of sound speed from remote
  recordings of waves. Model extension, that
  is, enlargement of the model space (domain of the simulation
  operator), coupled with a penalty for deviation from the original
  domain, permits descent methods to converge to a global minimizer of
  mean-square misfit. This paper applies the model extension concept
  to a simple transmission inverse problem for the wave equation modeled on crosswell
  tomography. The model extension studied here permits the energy
  source to depend source and receiver coordinates, whereas the
  modeled physics dictates that all energy sources should be the same
  (and given a priori). Minimization of a suitable penalty for
  deviation from the prescribed source model drives the sound velocity
  towards a global solution of the least squares data fit problem. An
  analysis based on geometric optics explains why the straightforward
  least-squares formulation fails to yield a useful algorithm, and why the
  extended formulation succeeds. The analysis requires the absence of
  caustics; we show that a substantial modification of the algorithm
  is required if these are present, and suggest such a modification
  for which numerical experiments provide some evidence of
  effectiveness.
\end{abstract}

\section{Introduction}
Estimation of sound speed in the wave equation from by least-squares
fitting of time-like seismic trace data (``full waveform inversion''
or FWI, in the seismic literature) has demonstrated potential to
produce highly detailed maps of the earth's sound speed
distribution. As an optimization problem, however, FWI is notoriously
difficult. Many realistic examples, using both synthetic and field
data, appear to show that iterative descent methods tend to stagnate
quite far from any global minimizer, at points in parameter space
having little to do with any accurate rendition of earth
structure. This commonly-observed stagnation has led many authors to
infer the existence of local minimizers of the least-squares
objective, caused by serendipitous partial fits of oscillatory data,
better than those produced by any nearby model, and preventing
improvement of fit by local optimization. A combination of data accuracy
at sufficiently low frequency and sufficiently accurate initial
velocity estimates will overcome this obstacle, but such combinations
are not always available, and in any case difficult to verify.
% This ``cycle
%skipping'' phenomenon has been conclusively demonstrated in only
%trivial cases, but is widely believed to be generic.  In any case,
%premature stagnation of iterative optimization methods is the biggest
%impediment to routine use of least-squares data fitting. It may be
%overcome by use of either sufficiently low-frequency data that the
%correct cycles in the simulated data overlap those in the measured
%data for reasonable initial estimate of sound velocity, or
%sufficiently accurate initial estimate that the correct cycles overlap
%for the measured frequency range. However frequency content of seismic
%data is limited by experimental design, and a priori knowledge of
%earth structure on the required scales is limited. It is not
%necessarily possible to achieve the right combination of frequency
%content and initial sound velocity accuracy, or even to know whether
%it has been achieved. 
See \cite{VirieuxOperto:09} for a good overview
%of FWI as currently construed by industrial and academic
seismologists.

This paper explores another solution of the FWI stagnation problem,
via extending the modeling operator to a larger  {\em infeasible} domain, so that the
data is easy to fit (in principle perfectly) by local optimization. We discuss the structure of FWI, and of its
{\em source-receiver} extension, for a simple acoustic transmission
problem, idealizing crosswell waveform tomography. Extended
(infeasible) solutions must be driven towards feasibility somehow.
We introduce measures of infeasibility, in the form of {\em
  annihilation} operators, the kernels of which are the original (feasible, physical)
domain of the modeling operator. We use these operators to define
auxiliary constrained minimization problems, and show that these
problems are convex over much larger domains than is the feasible
least squares problem (FWI).

The data of the idealized crosswell tomography problem
are functions of time, representing the acoustic
fields recorded at receivers position on a line in the plane. Source
processes generating these fields lie on another line in the plane,
and the sound velocity of the region between the lines is to  be
recovered from the data. It is assumed a priori that source processes
differ only by position on the line, and that this unique source
process is known. This assumption is an idealization: the effective
source process in actual crosswell surveys may well change from one
source position to another, even if the energy source apparatus
remains the same, due to coupling with the rock formation and other
near-field effects. Also, this source is seldom measured directly in
any usable way. For the purpose of this paper, however, we set aside
these issues.

The {\em source-receiver} extension discussed here models each
function of time, recording the acoustic field for each source and
receiver posiiton, as resulting from an independent source
process. That is, the source processes are permitted to depend on both
source and receiver coordinates, whereas physically they should be
independent of these coordinates. The larger model space consists of
the sound velocity field together with a set of source/receiver
dependent filters, one for each time function (``trace'') in the
data), which modify the output of the simulation operator at each
source/receiver combination. Unsurprisingly, the addition of so many
parameters to the domain of the modeling operator makes data fitting
considerably easier - in fact, for essentially any distribution of
sound velocity, it is possible to adjust the filters provided by the
extension to fit the data. Of the myriad possible measures of
non-physicality, we single out two for careful study - the mean square
of the image under multiplication by $t$, and the mean square of the derivative with respect to source and
receiver coordinates. Forcing either
quantity toward zero, while maintaining data fit, implicitly recovers a minimizer of the original
mean-square misfit function, for model-consistent data, provided that
none of the velocity fields encountered along the way produce
conjugate points, or caustics. Our main result in this case is that
the objective functions described above closely approximate objectives
for the traveltime inverse problem (``tomography''), hence inherit
their good properties, including in some cases global convergence.

In the presence of caustics, the close relation between tomography and
the source-receiver extended source inverse problems breaks down. We describe the
mechanism of this failure: it results from implicit attempts to match
different branches of traveltime. We show how to overcome this
problem, at least to some extent, via exponential weighting of data, a
technique introduced by Pratt (REF) to force emphasize fitting of
first-arriving waveforms in FWI.

\section{A model problem based on crosswell tomography'}

Crosswell seismic tomography  aims to  estimate acoustic structure between
two wells from seismic signals (``traces'') recorded in one of the wells due to seismic
energy sources in the other. The simplest useful realization of this
inverse problem uses linear acoustics to describe the generation and
propagation of seismic waves in the earth between the source and
receiver wells. In this paper we adopt the further simplifying
assumption that material density is constant and can be absorbed in
other quantities. The remaining parameter of linear acoustics is the
sound velocity, which varies with position, to reflect the
spatial heterogeniety of sedimentary rocks. With these assumptions,
crosswell tomography seeks to recover the sound velocity, or an
equivalent quantity, in the region
between the wells from measurements of acoustic fields at the wells.

Linear acoustics is a gross approximation to actual seismic wave
physics. We will make quite a few other approximations and
idealizations. The discussion section will offer some thoughts about
how important or limiting these assumptions are, and what prospects
exist for transcending them.

To describe the problem setting more precisely, we introduce
notation that will be used throughout the discussion. Coordinates in
$\bR^3$ are $\bx=(x,y,z)$, $z$ representing depth. We assume that the
source and receiver wells are 
perfectly straight and vertical line segments in $\bR^3$(certainly an
idealization!), lying in the plane $y=0$: the
source ``well'' $W_s$and receiver ``well'' $W_r$ are defined by the
limit depths $z_s^{\rm min}$,$z_s^{\rm max}$,$z_r^{\rm min}$,$z_r^{\rm
  max}$: 
\begin{eqnarray}
\label{eqn:wells} 
W_s & = & \{\bx_s = (x_s,0,z_s): z_s^{\rm min} \le z_s \le z_s^{\rm 
          max}\} \nonumber \\
W_r & = & \{\bx_r = (x_r,0,z_r): z_r^{\rm min} \le z_r \le z_r^{\rm 
          max}\} 
\end{eqnarray}
As we will frequently need to refer to the depth ranges in the wells,
introduce
\begin{eqnarray}
\label{eqn:depths} 
Z_s & = & Z_s\} \nonumber \\
Z_r & = & Z_r\} 
\end{eqnarray}
We idealize energy sources as isotropic point radiators. A radiator at
$\bx=\bx_s$ appears as the right-hand side in the acoustic wave
equation, of the form $\delta(\bx-\bx_s)f(t)$. 
The function $f(t)$ is causal (vanishes on a negative half-axis, that
is,  for $t<<0$), and has units of
mass/time/time. The idealized model posits that every
source is has the same mechanism, modeled by the
isotropic point radiator with the same time dependence $f$, differing only in
depth $z=z_s$ in the source well.

Slowness (reciprocal velocity) turns out to be a more convenient
choice than velocity to represent the spatially varying acoustic
properties of the interwell material. We denote the slowness by $m$.
For physical reasons, $m$ should be positive and bounded uniformly above and 
below, say by $0 < m_{\rm min} < m_{\rm max}$. We assume that the
material is {\em acoustically transparent}, that is, generates no
reflected waves of significant energy in the seismic band. The geometric
optics approximation, to be summarized below, shows that a convenient
mathematical proxy for transparency is the assumption that $m$is
smooth, with derivatives up to
order 2 bounded by another constant $\mu > 0$. These conditions are
summarized in the definition of the feasible set of slowness models $M$:
\begin{equation}
\label{eqn:vfeas}
M=\{m \in C^{\infty}(\bR^3): m_{\rm min} < m(\bx) < m_{\rm max},
|D^{\alpha}m(\bx)| \le \mu,
\bx \in \bR^3, |\alpha| \le 2\}
\end{equation}

The pressure field 
$p(\bx,t;\bx_s)$ due to the isotropic point radiator source at 
$\bx_s=(x_s,0,z_s)$, with time dependence $f(t)$, is the solution of the initial value problem 
\begin{eqnarray}
\left(m^2 \frac{\pa^2 p}{\pa t^2} - \nabla^2 p\right)(\bx,t) &=&
     \delta(\bx-\bx_s) f(t), \nonumber \\
 p&=&0, \, t<<0 \label{eqn:pressure}
\end{eqnarray}
We will establish the existance, uniqueness, and essential properties
of solutions $p$ in the next section. Proceeding formally for the moment,
define the modeling operator $S[m]$ to relate the slowness $m$ and
source function $f$ to the pressure field at the receiver locations
$\bx_r$ and source locations $\bx_s$, for a finite time interval $0
\le t \le t^{\rm max}$. Since the horizontal coordinates of sources and
receivers are fixed as part of the idealized cross-well acquisition
geometry, we regard the output of $S[m]$ as a function of $z_s,z_r,$
and $t$, with domain
\begin{equation}
\label{eqn:omdef}
\Omega = Z_r\times Z_s \times \bR
\end{equation}
 A smooth cutoff factor provides
convenient way to incorporate the limited time duration of measured
signals, and to smoothly taper the measured data to zero near the
boundary of the measurement domain, with favorable signal-processing
consequences. We denote this factor by
\begin{equation}
\label{eqn:wdef}
w \in C_0^{\infty}( \Omega): \,{\rm supp}\,w \subset 
Z_r\times Z_s \times (0,t^{\rm max}).
\end{equation}
Proceeding formally, define
\begin{equation}
\label{eqn:sdef}
S[m]f(z_r,t;z_s) = w(z_r,t;z_s) p(\bx_r,t;\bx_s), 
\end{equation}
The crosswell waveform inversion problem, in the highly idealized form 
studied here, is: given data $d$, a function of on $\Omega$,
find slowness $m \in M$ and source function $f$ such that 
$S[m]f \approx wd$.

\section{Properties of the Modeling Operator}

The problem formulation just stated begs several questions. For
example, what sort of functions $d$, $f$, and $m$ should be admitted?
Under what circumstances is the restriction of the pressure field $p$
to the submanifold $W_s \times W_r \times \bR$ well-defined?

For numerical convenience, and on physical grounds
\cite[]{SantosaSymes:00}, $f$ and $d$ should be square-integrable and
the natural (and conventional) measure for the data misfit $S[m]f-d$
is $L^2$. Indeed, iff $m \in M$ is constant, $f \in C_0^{\infty}(\bR)$,
$\bx \ne \bx_s$, then
\be\label{eqn:cv} 
p(\bx,t;\bx_s) = 
\frac{f(t-mr)}{4\pi r}, \,\,r=|\bx-\bx_s|.  
\ee 
The distribution on the right-hand side of equation \ref{eqn:cv} has a
well-defined restriction to $\bx=\bx_r$, for reasons to be reviewed
shortly. so that $S[m]$ given by
equation \ref{eqn:sdef} is well-defined, and extends continuously to a
map $f \in L^2(\bR) \mapsto S[m]f \in L^2(\bar{\Omega})$.

More generally,

\begin{theorem}
\label{thm:trace}
For any $m \in M$, $f \in {\cal E}'(\bR)$, there exists a unique solution $p \in {\cal D}'(\bR^4)$ of problem
\ref{eqn:pressure}. Define $\chi_{\bx_r}: \bR \rightarrow \bR^4$ by
$\chi_{\bx_r}(t) = (\bx_r,t)$. For any $\bx_r \in \bR^3\setminus
\{\bx_s\}$ , the map $f\mapsto \chi_{\bx_r}^*p$ is continuous: ${\cal E}'(\bR) \rightarrow
{\cal D}'(\bR)$. Denote by $G(\cdot,\cdot;\bx_s) \in {\cal D}'(\bR^4)$
the solution of the initial value problem
\ref{eqn:pressure} with $f = \delta$. Then $\chi_{\bx_r}^*p =
(\chi_{\bx_r}^*G(\cdot,\cdot;\bx_s))*f =
\chi_{\bx_r}(G(\cdot,\cdot;\bx_s)*(\delta_{\bR^3}\otimes f))$. 
\end{theorem}
\begin{proof}
  A proof of the existence and uniqueness of causal distribution
  solutions to \ref{eqn:pressure} may be found in \cite{Lax:PDENotes},
  Chapter 6. Alternatively, since $\delta(\cdot-\bx_s)\otimes f \in
  H^s(\bR^4)$ for some $s < -2$, \cite{Tay:81} 2.1-2.2 shows
  how to construct local solutions, which can be patched together
  taking advantage of the uniformly bounded speed of propagation
  implicit in $m \in M$. From the definition (\cite{Tay:81}, Chapter VI), the wave front set
  $WF(p)$ is contained in the set
\[
C \cup \Pi^{-1}({\{\bx_s\} \times \bR})
\]
in which $C$ is the characteristic variety of the wave operator,
\begin{equation}
\label{eqn:charvar}
C[m] = \{(\bx,t,\xi,\omega) \in T^*(\bR^4): m^2(\bx)\omega^2 =
|\xi|^2\}
\end{equation}
and  $\Pi: T^*(\bR^4) \rightarrow \bR^4$ is the fiber projection.

The normal bundle of $\{\bx_r\} \times \bR$ consists of covectors with
coordinates $(\bx_r,t,\xi,0)$, hence has trivial intersection with
$C[m]$. Therefore the restriction of $p$ to $\{\bx_r\} \times \bR$ is
well-defined (\cite{Dui:95}, Proposition 1.3.3), as is the pull-back
by $\chi_{\bx_r}$. Convolution in time
with $f$ is by definition convolution in space-time with
$\delta_{\bR^3} \otimes f \in {\cal E}'(\bR^4)$, a continuous map on ${\cal
  D'}(\bR^4)$,  and this latter convolution commutes with
the wave operator in \ref{eqn:pressure}. This convolution also maps
the right-hand side for $G$ to the right-hand-side for $p$.
\end{proof}

\begin{rem} $G$ is of course the causal Green's function, or distribution
  kernel of the retarded fundamental solution \cite[]{CourHil:62}.
\end{rem}

An immediate consequence of this result is a proper definition of the
map $S[m]$.

\begin{theorem}
\label{thm:gather}
Denote by $\Delta = \{(\bx,\bx): \bx \in \bR^3\}$ the diagonal. With
notation as in Theorem \ref{thm:trace}, the map $(\bR^3 \times \bR^3)
\setminus \Delta \times {\cal E}'(\bR) \rightarrow {\cal D}'(\bR)$ defined by $(\bx_r,\bx_s,f)
\mapsto \chi_{\bx_r}^*p(\cdot,\cdot;\bx_s)$ is continuous. 
\end{theorem}
\begin{proof}
The right-hand side of \ref{eqn:pressure} is a continuous
${\cal E}'(\bR^4)$- valued function of $\bx_s$. Denote by $N$ the
normal bundle of the fibration $\{\{\bx\} \times \bR: \bx \in \bR^3\}$.
Denote by $\Gamma$ a closed conic submanifold of
$T^*(\bR^4)$ containing $C$ in its interior. chosen so that $\Gamma
\cap N=\{0\}$. The solution of the wave equation is a continuous
${\cal D}'(\bR^4)$-valued function of the ${\cal E}'(\bR^4)$- valued
right-hand side. Therefore solution of \ref{eqn:pressure} is a continuous
${\cal D}'(\bR^4)$-valued function of $\bx_s\in \bR^3$ and $f\in {\cal
  E}'(\bR)$, and also continous in the sense of 
${\cal D}_{\Gamma}'(\bR^4\setminus \{\bx_s\})$ as a consequence of the
propagation of singularities theorem. Since $N \cap \Pi^{-1}(\bx_r) =\{\bx_r\} \times \bR$,
$\chi_{\bx_r}^*$ is continuous: $\bR^3 \times {\cal D_{\Gamma}}'(\bR^4) \rightarrow
{\cal D}'(\bR)$. Putting all of this together, obtain the 
statement of the theorem. 
\end{proof}

Theorem \ref{thm:gather} permits a precise rewrite of the
formal definition \ref{eqn:sdef}: define $S[m]: {\cal E}'(\bR)
\rightarrow C(Z_r \times [z_x^{\rm min},z_s^{\rm max}], {\cal E}'(\bR))$ by 
\[
S[m]f(z_r,\cdot;z_s) = w(z_r,\cdot;z_s) \chi_{\bx_r}^*p(\cdot,\cdot;\bx_s)
\]

To formulate the inverse problem as envisioned in the last section, an
analogous result is needed with $L^2$ replacing ${\cal E}'$. This step
requires the introduction of geometric optics.

Recall that the rays of geometric optics are trajectories
$(X(t),P(t))$ in $\bR^3 \times (\bR^3 \setminus \{0\})$ satisfy 
\begin{eqnarray}
\label{eqn:ham}
\frac{dX}{dt} &=&m(X)^{-2}P,\nonumber \\
\frac{dP}{dt} &=&\nabla \log m(X), \nonumber\\
|P| &=& m(X).
\end{eqnarray}
The condition $m \in M$ implies that trajectories exist that solve the
first two conditions in \ref{eqn:ham} for all $t>0$, for any initial
values of $X$ and $P$. The third condition is compatible with the
other two: if the initial $X$, $P$ satisfy it, then it is satisfied
along the entire ray.

Define the {\em geodesic exponential map}
\[
\exp(t,P_s; \bx_s) = X(t)
\]
where $(X(t),P(t))$ solves \ref{eqn:ham} with $X(0)=\bx_s$,
$P(0)=P_s, |P_s|=m(\bx_s)$: $(t,P_s)$ are {\em
  geodesic normal coordinates} centered at $\bx_s$. Denote by
$S_m(\bR^3)$ the sphere bundle defined by the Riemannian metric $m (dx
\otimes dx + dy \otimes dy + dz \otimes dz)$ defined by the slowness
$m \in M$. It is a standard
result that $t^*:S_m(\bR^3) \rightarrow \bR_+$ exists for which
$\exp$ is a diffeomorphism from 
\begin{equation}
\label{eqn:prenorm}
\tilde{{\cal N}} = \{(\bx_s,t,P_s) \in
\bR^3 \times \bR_+ \times m(\bx_s)S^2: 0 < t < t^*(\bx,P_s)\}
\end{equation}
 onto a punctured
neighborhood ${\cal N}(\bx_s) \setminus \{\bx_s\}$ of $\bx_s$. A {\em normal neighborhood}
${\cal N}$ is an open subset of $\bR^3$ for which ${\cal N} \subset
{\cal N}(\bx)$ for every $\bx \in {\cal N}$. It is also a standard
result that every point $\bx \in \bR^3$ has a normal neighborhood (\cite{Friedlander:75}, Chapter 2).

In a normal neighborhood of $\bx_s$, the inverse of the exponential
map expresses $t$ as a function of $\bx$: this function is the
{\em travel time} $\tau(\bx,\bx_s)$.
The travel time is differentiable and satisifies the eikonal equation in the punctured normal
neighborhood (and in a suitable generalized sense in the entire normal
neighborhood):
\begin{eqnarray}
\label{eqn:eik}
|\nabla_{\bx} \tau(\bx,\bx_s)|&=&m(\bx), \bx \in {\cal N}, \nonumber \\
\tau(\bx,\bx_s) & \sim & m(\bx_s)|\bx-\bx_s| \mbox{ as } \bx \rightarrow 
\bx_s. 
\end{eqnarray}

A slowness field $m \in M$ satisfies the {\em simple ray geometry}
hypothesis iff for every $\bx_s \in W_s$, the
the receiver well $W_r$ is contained in a
normal neighborhood ${\cal N}(\bx_s)$ centered at $\bx_s$. In this case, there is a common
neighborhood ${\cal N}$, normal for all points $\bx_s$ in the source
well and containing the closure of the receiver well.

A slowness fails to satisfy simple ray geometry hypothesis if the
exponential map at some source point $\bx_s$ fails to be a
diffeomorphism in a neighborhood of the receiver well, by failing to
be injective, that is, more than one ray (``path'') connects $\bx_s$
with another point,

In the simple ray geometry case, the Green's function has a useful
decomposition.

\begin{theorem}
\label{thm:simplegreen}
Suppose that $m$ satisfies the simple ray geometry assumption. Then there exists $t^{\rm max} >
\max\{\tau(\bx_r,\bx_s): \bx_r \in W_r, \bx_s \in W_s\}$, and 
functions $a \in C^{\infty}(W_r\times W_s), b\in
C^{\infty}(W_r,W_s,(0,t^{\rm max})$ so that for
$\bx_r \in W_r, \bx_s \in W_s, t <t^{\rm max}$,
\begin{equation}
\label{eqn:simplegreen}
G(\bx_r,t;\bx_s) = a(\bx_r,\bx_s)\delta(t-\tau(\bx_r,\bx_s)) + b(\bx_r,t;\bx_s)H(t-\tau(\bx_r,\bx_s)).
\end{equation}
\end{theorem}
\begin{proof}
The starting point is the parametrix contruction due to Hadamard, as described
by \cite{Friedlander:75}, Theorem 4.3.2. In the
notation used here, for $\bx,\bx_s \in {\cal N}$ (a normal
neighborhood), there exist $U, V \in C^{\infty}({\cal N}\times{\cal N}
\times \bR)$ so that
\begin{equation} 
\label{eqn:parametrix}
\tilde{G}(\bx,t;\bx_s) = U(\bx,t;\bx_s)\delta^+(t^2-\tau^2(\bx,\bx_s))
+ V(\bx,t;\bx_s)H^+(t^2-\tau(\bx,\bx_s))
\end{equation}
Here $\delta^+(t^2-\tau^2)$ is defined as the limit
\[
\delta^+(t^2-\tau^2) = \lim_{\epsilon \rightarrow
  0^+}H(t-\tau)\delta(t^2-\tau^2-\epsilon)
\]
The product on the right is well-defined, as the singular supports of
the factors are disjoint, and picks out the positive-time sheet of the
distorted hyperboloid $t^2 = \tau^2$. The limit exists in the sense of
${\cal D}'$.

$\tilde{G}$ is a $C^{\infty}$ parametrix, in the sense that 
\[
\left(m\frac{\partial^2}{\partial
    t^2}-\nabla^2\right)\tilde{G}(\bx,t;\bx_s) =
\delta(\bx-\bx_s) + R(\bx,t;\bx_s), \,\,R \in C^{\infty}({\cal
  N}\times{\cal N}\times \bR),\,R=0 \mbox{ for } t<\tau.
\]
Define $\tilde{R}$ to be the solution of the initial value problem
\[
\left(m\frac{\partial^2}{\partial
    t^2}-\nabla^2\right)\tilde{R} = -R; \,\,\tilde{R}=0, t<0
\]
This problem has a unique $C^{\infty}$ solution. The
domain-of-dependence property implies that $\tilde{R}=0$ for $t <
\tau$ also.

Any singularity of $G$ in ${\cal N}\times{\cal N}\times \bR$ lies
either on the light cone $t=\tau$ or must occur in $t > t^{\rm
  max}$. So on ${\cal N}\times{\cal N}\times (0,t^{\rm max})$, 
\[
G = \tilde{G} + \tilde{R}.
\]

Finally, since 
\[
\delta^+(t^2-\tau^2) = \frac{\delta(t-\tau)}{2\tau}
\]
for $t>0$, or equivalently, $\tau>0$, set $a = U/2\tau, b = V +
\tilde{R}$ to obtain \ref{eqn:simplegreen}.
\end{proof}

\begin{rem}
In the notation used here, 
\[
U(\bx,\bx_s) = \frac{m(\bx_s)}{2\pi} + O(|\bx-\bx_s|)
\]
so for constant $m$,
\[
a(\bx_r,\bx_s) = \frac{m}{4\pi\tau(\bx,\bx_s)} +
O(1)=\frac{1}{4\pi|\bx-\bx_s|} + O(1)
\]
which recovers \ref{eqn:cv} approximately for $\bx_r$ close to $\bx_s$
- of course, the error term actually vanishes in this case.
\end{rem}

\begin{cor}
\label{thm:gosimple}
Under the assumptions of Theorem \ref{thm:simplegreen}, 
for $\bx_r \in W_r, \bx_s \in W_s$, and $t < t^{\rm max}$, set $A(z_r,z_s) = a(\bx_r,\bx_s),
T(z_r,z_s)=\tau(\bx_r,\bx_s), B(z_r,t;z_s) = b(\bx_r,t;\bx_s)$. Then
$A,T \in C^{\infty}(Z_r \times 
Z_s)$,
$B \in \in C^{\infty}(Z_r \times Z_s\times (-\infty,t^{\rm max}))$, $T>0$. For $f \in
L^2(\bR)$,
\begin{equation}
\label{eqn:gosimple} 
S[m]f (z_r,t;z_s) = w(z_r,t;z_s)\left(A(z_r,z_s)f(t-T(z_r,z_s)) + \int_{T(z_r,z_s)}^{\infty}
\,ds\,B(z_r,s;z_s)f(t-s)\right). 
\end{equation}
\end{cor}

For highly oscillatory $f$, the first term in (\ref{eqn:gosimple}) is much
larger than the second, and can be taken as a high frequency
asymptotic approximation of $S[m]f$.

Analysis of the inverse problem requires some information on 
the smooth dependence of $S$ on $m$.

\begin{theorem}
\label{thm:smdep}
Denote by $M^k$ be the interior of the closure of $M$ in $C^k({\cal
  N})$; $M^k$ is defined by the same conditions as $M$ but with
$C^{\infty}$ replaces by $C^k$.  Then
\begin{itemize}

\item[1. ] If $N$ is a normal neighborhood for $m_0 \in M^k$,
  $k \ge 2$, and $\tilde{\cal N}$ is its pre-image under the
  exponential map for $m_0$ as defined in \ref{eqn:prenorm}, then there is an open neighborhood ${\cal U}$ of $m_0$ in
  $M^k$ so that ${\cal N}$ is a normal neighborhood for every
  $m \in {\cal U}$.
\item[2. ] For ${\cal N}, \tilde{\cal N}$, and ${\cal U}$ as defined
  in item 1, 
\begin{equation}
\label{eqn:expreg}
\exp \in C^0({\cal U}, C^k(\tilde{\cal N},{\cal N})) \cap C^1({\cal
  U}, C^{k-1}(\tilde{\cal N},{\cal N}))
\end{equation}
\item[3. ] Suppose that the simple ray geometry assumption holds for
an open set ${\cal U} \subset M^k$ (that is, for every $m \in {\cal
  U}$ and that ${\cal N}$ is a normal neighborhood for every $m \in
{\cal U}$ containing $W_r$ and $W_s$. Then $T, A,$ and
$B$ extend to functions on ${\cal U}$ for $k \ge 2$ so that
\begin{itemize}
\item[3.1 ]$T \in C^1({\cal U},C^{k-2}({\cal N}\times{\cal N}))$ 
\item[3.2 ]$A \in C^1({\cal U},C^{k-2}({\cal N}\times{\cal N}))$ 
\item[3.3 ]$B \in C^1({\cal U},,C^{k-2}({\cal N}\times{\cal N}\times(-\infty,t^{\rm
    max})))$
\item[3.4 ]$S \in C^0({\cal U},{\cal L}_s(L^2(\bR),
C^0(Z_r\times Z_s))),
  L^2(\bR))))$
\item[3.5 ]$S \in G^1({\cal U},{\cal L}_s(H^1_0(\bR),
C^0(Z_r\times Z_s))),
  L^2(\bR))))$
\end{itemize} 
\end{itemize}
\end{theorem}
\begin{proof}
The items 1., 2., 3.1, 3.2, and 3.3 follow from the standard theorem
  on continuous dependence of solutions of ordinary differential
  equations on parameters applied to the Hamiltonian system
  \ref{eqn:ham}, and to the transport equations defining the functions
  $U$ and $V$ appearing in the Hadamard construction, and the inverse
  function theorem applied to the exponential map. Note that $B$ is
  implicitly extended smoothly to $t < \tau(\bx,\bx_s)$, possible
  because the intersection of the light cone $t=\tau(\bx,\bx_s)$ with
  $W_r$ is a smooth curve. In items 3.4 and
  3.5, ${\cal L}_s({\cal B}_1,{\cal B}_2)$ denotes the topological
  vector space of continuous linear maps from Banach space
  ${\cal B}_1$ to Banach space $ {\cal B}_2$, endowed with the strong
  operator topology, that is, the topology of pointwise
  convergence. That is, if $m_j \rightarrow m$ in
  ${\cal U} \subset M^k$, then for each $f \in L^2(\bR)$,
  $S[m_j]f \rightarrow S[m]f$ in the sense indicated in item
  3.4. In item 3.5, $G^1$ indicates the class of
  G\^{a}teaux-differentiable functions: that is, for each $m \in {\cal
    U}$, there exists a linear map 
\[
DS[m]: M^k \rightarrow {\cal L}_s(H^1_0(\bR),
C^0(Z_r\times Z_s,L^2(\bR))),
\]
so that for each $\delta m \in M^k$, $f \in L^2(\bR)$,
\[
\|(S[m + h\delta m]-S[m]-hDS[m]\delta m) f \| = o(h) \mbox{ as }
h\rightarrow 0.
\]
Items 3.4 and 3.5 follow directly from the the other
  conclusions and Corollary \ref{thm:gosimple}.
\end{proof}

\begin{rem}
\label{rem:nounif}
  It is not possible to replace ${\cal L}_s$ with ${\cal L}$, the same
  vector space of continuous linear maps with the (Banach) topology of
  uniform convergence, in the statements of items 3.4 and 3.5. To see
  this, suppose that $m_j \rightarrow m$ in $M^k$. Assume that $m$
  satisfies the simple ray geometry condition with normal neighborhood
  ${\cal N}$ containing $W_r$ and $W_s$, and that the sequence
  $\{m_j\}$ lies inside an open ball ${\cal U}$ sufficiently small
  that ${\cal N}$ is a normal neighborhood for every member of this
  sequence. Finally suppose that
\[
\inf\{\tau[m_j](z_s,z_r) -\tau[m](z_s,z_r)\}=\epsilon_j > 0.
\]
This condition can be arranged, for example, by taking
$m_j=(1+c2^{-j})m$ for sufficiently small $c>0$. For each $j$, choose
$f_j \in L^2(\bR)$ so that 
\[
\|f_j\|=1,\,\,{\rm supp}\,f_j \subset [0,\epsilon_j/2].
\]
Then ${\rm supp}\,f_j(\cdot-\tau[m_j])$ and ${\rm supp}\,f_j(\cdot-\tau[m])$ are
disjoint. It is easy to see that the integrated term in the expansion
\ref{eqn:gosimple} tends to zero,  and item 2 above implies that
$A[m_j] \rightarrow A[m]$, so 
\[
\|S[m_j]f_j-S[m]f_j\|^2\rightarrow \|A[m_j]f_j(\cdot-\tau[m_j])\|^2 + \|A[m]f_j(\cdot-\tau[m])\|^2 
\]
\[
\rightarrow 2 \int_{z_r^{\rm min}}^{z_r^{\rm
    max}}\,dz_r\,\int_{z_s^{\rm min}}^{z_s^{\rm
    max}}\,dz_s\,|A[m](z_r,z_s)|^2 > 0
\]
Thus $S[m_j]$ does not tend to $S[m]$ in operator norm.

This observation captures a very important characteristic of all
inverse problems in wave propagation in which wave velocity is an
unknown: the data prediction map is not uniformly continuous in
material parameters and source parameters jointly. This fact explains
the character of the full waveform inversion problem, to be analyzed
in the next section.
\end{rem}

\begin{rem}
\label{rem:derivloss}
Note that for $m, \delta m \in M^k$, $DS[m]\delta m$ is continuous
on a smaller domain than is the case for $S[m]$. In fact, it follows
from Corollary \ref{thm:gosimple} that for $f \in H^1_0(\bR)$,
\[
D(S[m]f)\delta m  = w\left(-A[m](DT[m]\delta m)\frac{df}{dt}(t-T[m]) +
((DA[m]\delta m)\right.
\]
\begin{equation}
\label{eqn:preq}
\left. -B[m](DT[m]\delta m))f(t-T[m]) + \int_{T(z_r,z_s)}^{\infty} \,ds\,
(DB[m]\delta m)(s)f(t-s)\right)
\end{equation}
This observation shows that the statement of item 5 is sharp.
\end{rem}

\begin{rem}
\label{rem:multipath}
In general, geometric optics predicts more than one traveltime between
each source and receiver (the multipath case), hence a principal term
involving multiple shifted copies f. It is possible to show that in a
subset of $W_r \times W_s$ of second Baire category, smooth travel times
$T_i$, amplitudes $a_i$, and remainders $B_i$ exist so that
\begin{equation}
\label{eqn:gomultipath} 
S[m]f (z_r,t;z_s) = w(z_r,t;z_s)\left(\sum_{i=0}^NA_i(z_r,z_s)f(t-T_i(z_r,z_s)) + \int_{T_i(z_r,z_s)}^{\infty} 
\,ds\,B_i(z_r,s;z_s)f(t-s)\right). 
\end{equation}
We will not give a detailed justification of this expansion, but will
note one of its consequences later.
\end{rem}

\section{Full Waveform Inversion (FWI)}

In view of the mapping property described in the previous section, a
natural explicit formulation of the inverse problem is via least
squares: given $d\in L^2(\Omega)$ and $f\in L^2(\bR)$, find $m \in M$ to minimize 
\be\label{eqn:ls}
 J_{\rm LS}[m;f,d] =
\frac{1}{2}\|(S[m]f-d)\|^2.  
\ee

The first consequence to draw from the expansion \ref{eqn:gosimple}
is a very strong limitation on the possible convexity of $J_{\rm
  LS}$. To state this fact conveniently, introduce a family of source
functions based on a ``mother'' function $f_1 \in
C^{\infty}_0(\bR)$, for which
\begin{equation}
\label{eqn:mutha}
\|f_1\|_{L^2(\bR)} = 1.
\end{equation}
For $\epsilon>0$, define
\be\label{eqn:srcfam}
f_{\epsilon} =
\frac{1}{\sqrt{\epsilon}}f_1\left(\frac{t}{\epsilon}\right).
\ee

\begin{theorem}\label{thm:FWIconv}
Suppose that $k \ge 2$, ${\cal U} \subset M^k$ is open and convex and that the
simple ray geometry hypothesis (Definition \ref{def:srg}) is satisfied
at every $m \in {\cal U}$.  Suppose that $m_*, m_0 \in {\cal U}$, and
for some $\rho < 1$,
\be\label{eqn:ttgap}
T[m_*](z_s,z_r) \le \rho T[m_0](z_s,z_r).
\ee
 for every $\bx_s \in W_s, \bx_r \in W_r$,  Define $d_{\epsilon} = S[m_*]f_{\epsilon}$. Then there exists
$\epsilon_*>0$ so that for $0 <\epsilon \le \epsilon_*$, $J_{\rm LS}(\cdot,d_{\epsilon},f_{\epsilon})$
is {\em not} convex on ${\cal U}$.
\end{theorem}
\begin{rem}
\label{rem:scale}
The relation \ref{eqn:ttgap} is easy to achieve. If $m_0 \in {\cal U}$ and
$1-\rho$ is sufficiently small that $m_* = \rho m_0 \in {\cal U}$, then
\ref{eqn:ttgap} holds.
\end{rem} 
\begin{proof}
It sufficies to show that the function $j:[0,1] \rightarrow \bR$ defined by 
\[
j(\sigma) = J_{\rm LS}[\sigma m_* + (1-\sigma)m_0;d_{\epsilon},f_{\epsilon}]
\]
is not convex.

The crux of the matter is the representation \ref{eqn:gosimple}. To signify the dependence of the quantities $A, T,$ and $B$ from
equation \ref{eqn:gosimple} on the slowness field, throughout this
argument and subsequent discussion we write $A=A[m]$ and so on.

Theorem \ref{thm:smdep} that implies that given $\alpha>0$, there
exists $\sigma_0 \in [0,1)$ for which 
\[
(1+\alpha) \int_{z_r^{\rm min}}^{z_r^{\rm max}}\,dz_r\, \int_{z_s^{\rm
    min}}^{z_s^{\rm max}}\,dz_s\,|A[m_*](z_r,z_s)|^2.
\]
\[
\ge \int_{z_r^{\rm min}}^{z_r^{\rm max}}\,dz_r\, \int_{z_s^{\rm
    min}}^{z_s^{\rm max}}\,dz_s\,|A[\sigma m_* +
(1-\sigma)m_0](z_r,z_s)|^2 
\]
\[
\ge (1-\alpha) \int_{z_r^{\rm min}}^{z_r^{\rm max}}\,dz_r\, \int_{z_s^{\rm
    min}}^{z_s^{\rm max}}\,dz_s\,|A[m_*](z_r,z_s)|^2.
\]
for $\sigma \in [\sigma_0,1]$.

Define $S_0$ to be the first term in the expansion \ref{eqn:gosimple},
\[
S_0[m]f(z_r,t;z_s) = A[m](z_r,z_s)f(t-T(z_r,z_s)).
\]
Condition \ref{eqn:ttgap} implies that there exists $\epsilon_*>0$,
depending on $\alpha$, so that if $0 <\epsilon\le \epsilon_*$, then the supports of $S_0[\sigma m_* + (1-\sigma)m_0]f_{\epsilon}$ and
$S_0[m_*]f_{\epsilon}$ are disjoint if $\sigma_0 \le \sigma \le \frac{1}{2}(1+\sigma_0)$, so
\[
\|S_0[\sigma m_* + (1-\sigma)m_0]f_{\epsilon}-S_0[m_*]f_{\epsilon}\|^2 =
\|S_0[\sigma m_* + (1-\sigma)m_0]f_{\epsilon}\|^2 + \|S_0[m_*]f_{\epsilon}\|^2 
\]
\[
= \int_{z_r^{\rm min}}^{z_r^{\rm max}}\,dz_r\, \int_{z_s^{\rm
    min}}^{z_s^{\rm max}}\,dz_s\,|A[\sigma m_* + (1-\sigma)m_0](z_r,z_s)|^2 +
|A[m_*](z_r,z_s)|^2 
\]
\begin{equation}
\label{eqn:orthominus}
\ge (2-\alpha)\left(\int_{z_r^{\rm min}}^{z_r^{\rm max}}\,dz_r\, \int_{z_s^{\rm
    min}}^{z_s^{\rm max}}\,dz_s\,|A[m_*](z_r,z_s)|^2\right) 
\end{equation}
Similarly, for $\sigma \in [\sigma_0, \frac{1}{2}(1+\sigma_0)]$,
\begin{equation}
\label{eqn:orthoplus}
\|S_0[\sigma m_* + (1-\sigma)m_0]f_{\epsilon}-S_0[m_*]f_{\epsilon}\|^2 
\le (2+\alpha)\left(\int_{z_r^{\rm min}}^{z_r^{\rm max}}\,dz_r\, \int_{z_s^{\rm
    min}}^{z_s^{\rm max}}\,dz_s\,|A[m_*](z_r,z_s)|^2\right) 
\end{equation}

Since $ B[\sigma m_* + (1-\sigma)m_0]$ is uniformly bounded in
$C^{k-2}$,
\[
\sup_{t \in (0,t^{\rm max})} \left|\int_{T(z_r,z_s)}^{\infty}
  \,ds\,B[\sigma m_* + (1-\sigma)m_0](z_r,s;z_s)f_{\epsilon}(t-s)\right| = O(\sqrt{\epsilon})
\]
If $\epsilon_*>0$ is chosen possibly smaller yet, again depending on
$\alpha$, then for $0 <\epsilon\le\epsilon_*$,
\[
\|S[\sigma m_* + (1-\sigma)m_0]f_{\rm epsilon}-S_0[\sigma m_* +
(1-\sigma)m_0]f_{\rm epsilon}\|^2 
\]
\begin{equation}
\label{eqn:lot}
\le \alpha \int_{z_r^{\rm min}}^{z_r^{\rm max}}\,dz_r\,
  \int_{z_s^{\rm min}}^{z_s^{\rm max}}\,dz_s\,|A[m_*](z_r,z_s)|^2
\end{equation}
Combining \ref{eqn:orthoplus} and \ref{eqn:lot} for $\sigma=\sigma_0$, obtain
\begin{equation}
\label{eqn:ub}
\|S[\sigma_0 m_* + (1-\sigma_0)m_0]f_{\epsilon}-S[m_*]f_{\epsilon}\|^2 
\le (\sqrt{2+\alpha}+2\sqrt{\alpha})^2 \int_{z_r^{\rm min}}^{z_r^{\rm max}}\,dz_r\,
  \int_{z_s^{\rm min}}^{z_s^{\rm max}}\,dz_s\,|A[m_*](z_r,z_s)|^2 
\end{equation}
Similarly, combining \ref{eqn:orthominus} and \ref{eqn:lot} for
$\sigma=\frac{1}{2}(1+\sigma_0)$, 
\[
\|S[\frac{1}{2}(1+\sigma_0) m_* + (1-\frac{1}{2}(1+\sigma_0))m_0]f_{\epsilon}-S[m_*]f_{\epsilon}\|^2 
\]
\begin{equation}
\label{eqn:lb}
\ge (\sqrt{2-\alpha}-2\sqrt{\alpha})^2 \int_{z_r^{\rm min}}^{z_r^{\rm max}}\,dz_r\,
  \int_{z_s^{\rm min}}^{z_s^{\rm max}}\,dz_s\,|A[m_*](z_r,z_s)|^2 
\end{equation}
Since 
\[
\frac{\sqrt{2-\alpha}-2\sqrt{\alpha}}{\sqrt{2+\alpha}+2\sqrt{\alpha}}
\rightarrow 1 \mbox{ as } \alpha \rightarrow 0,
\]
inequalities \ref{eqn:ub} and \ref{eqn:lb} imply that for $\alpha$
small enough (and consequenty $\epsilon_*$ small enough,
\[
\frac{1}{2}j(\sigma_0) = \frac{1}{2}(j(\sigma_0) + j(1)) <
j\left(\frac{1}{2}(1+\sigma_0)\right),
\]
\end{proof}

\begin{rem} Since 
\[
\frac{\int d\omega \omega^2 |\hat{f}_{\epsilon}(\omega)|^2}{\int
  d\omega |\hat{f}_{\epsilon}(\omega)|^2}
= \frac{\int dt \left(\frac{df_{\epsilon}}{dt}(t)\right)^2}{\int dt
  \left(f_{\epsilon}(t)\right)^2}
\]
\[
=\epsilon^{-2}\frac{\int dt \left(\frac{df_{1}}{dt}(t)\right)^2}{\int dt
  \left(f_{1}(t)\right)^2}
\]
and the left-hand side can be interpreted as mean square frequency in
$f_{\epsilon}$, $\epsilon$ can be interpreted as relative wavelength
(relative to the root mean square wavelength in $f_1$). Thus the maxiumum
diameter of an inscribed sphere in the domain of convexity of $J_{\rm
  LS}$ is proportional to a wavelength. This fact is often expressed
in the literature by stating that to ensure success of FWI, the
initial estimate of slowness must be sufficient to predict traveltimes
within a half-wavelength. Of course, Theorem \ref{thm:FWIconv} does not show that
local optimization would not succeed, that is, that stationary points
other than the global minimizer exist, merely that the domain of
convexity has diameter on the order of a mean wavelength.

Note that the conclusion of Theorem \ref{thm:FWIconv} holds regardless of whether
$f_1$ is oscillatory or mean-zero, so the common description of this
phenomenon as ``cycle-skipping'' is somewhat misleading.
\end{rem}

\section{A Synthetic Example: FWI}
A simple 2D synthetic example illustrates the impact of Theorem
\ref{thm:FWIconv}. We will also use this example throughout the paper
to illustrate alternative solution methods for the crosswell inverse
problem.

The target slowness $m_*$ is the reciprocal of a piecewise
linear velocity (Figure
\ref{fig:velbig}). This corresponding velocity has a constant $2$ km/s layer near $x=0$ and
increases linearly for $x\ge 0.5$ km. Boundary conditions are Dirichlet
for pressure on all parts of the boundary, although the wavefields
occuring in the experiments described here interact only with the $x=0$
surface so that the effective geometry is that of a half-space. Data
for a single source at depth $x_s=$ 10 m, $z_s=$ 5100 m (coordinates as
depicted in Figure \ref{fig:velbig}. A line of receivers is positioned
at depth $x_r=$ 3000 m, with 4000 m $\le z_r \le$ 7000 m, spaced
$\Delta z_r$=30 m apart.

The source pulse ($f$ in equation (\ref{eqn:pressure})) is a [2.5, 5, 25, 30]
Hz trapezoidal bandpass filter produces the synthetic data
depicted in Figure \ref{fig:shot6kmbig}. Note that the finite
difference scheme used to produce this data approximates 2D wave
propagation rather than the 3D propagation mentioned in the discussion
so far, so that the data waveform is asymmetric, unlike the
source. Also the presence of the pressure-free surface near the source
contributes effectively another derivative. Nonetheless the data
resemble the prediction of the constant velocity calculation (equation
(\ref{eqn:cv})).

For a very smooth $m$ such as that depicted in Figure
\ref{fig:velbig}, the curve $z_r \mapsto \tau[m](z_s,z_r)$ resembles a hyperbola, as one sees from Figure
\ref{fig:shot6kmbig}. The onset time of the signal in this figure
depends on the slowness used in the simulation. For
$m_0 \equiv 0.5$ s/km, the predicted data $S[m_0]f$ are depicted in
Figure \ref{fig:shot6kmhom}. The general appearance is similar to   Figure
\ref{fig:shot6kmbig}, but the onset time of the wave is later, due to
the slower wave velocity.

Now consider the inverse problem (\ref{eqn:ls}) with $d=S[m_*]f$. Since
$m_*=m_0$ near the source point, it is not unreasonable to take $m_0$
as an initial guess for the wave velocity. Conventional FWI uses
iterative optimization algorithm to minimize $J_{\rm LS}$. Descent
algorithms related to Newton's method have been favored for FWI, due
to the computational cost of function evaluation
\cite[]{VirieuxOperto:09}. Of course, any such algorithm finds a
stationary point of the objective, assuming convergence, and not
necessarily a global minimizer.

A plot (Figure \ref{fig:fwicurve}) 
of $J_{\rm LS}$ along the line
segment from $m_0$ to $m_*$ immediately shows the difficulty faced by
a local optimization approach: as one
would predict from Theorem \ref{thm:FWIconv} and its proof, the predicted data change
little in size along this segment and have support nearly disjoint
from that of the target data $d$, hence are orthogonal to it, until the trial slowness becomes
quite close to the target $m_*$, so that the local behaviour of $J_{\rm LS}$ in a small
neighborhood of $m_0$ gives little or
no hint about a constructive update direction. One would not expect a
local optimization algorithm to be effective under such circumstances.

Most descent methods require computed gradient, as input to the
computation of a descent step. Assuming that $f \in H^1_0(\bR)$, the gradient of $J_{\rm LS}$ is 
\be\label{eqn:gradls}
\nabla J_{\rm LS}[m] = D(S[m]f)^T(w^2(S[m]f-d)).
\ee
In the expression \ref{eqn:gradls}, note that $m \rightarrow S[m]f$ is
a map from $M$ to $L^2(\Omega)$, the derivative of which is given in
equation  \ref{eqn:preq}. The symbol $D(S[m]f)^T$ denotes the
transpose of this derivative with respect to a Hilbert space norm
chosen for $M$, about which we have said nothing up to this point. We
will mention some sensible possibilities in the Discussion section;
for now, we simply identify $M$ with $L^2(\bR^2)$ for the purpose of
computing the transpose, hence the gradient of $J_{\rm LS}$. Thus the
gradient as computed here is the image of the formal or $L^2$ adjoint
of $D(S[m]f)$, applied
to the time-cutoff residual $w^2(S[m]f-d)$. 
%There is no reason to suppose that this
%image would be smooth, given the weak constraints on $f$ and $d$, so
%in general it will not define an acceptable update direction for
%velocity. One way out of this predicament is to extend the domain of
%$S$ to a subset of a larger Hilbert space containing
%$C^{\infty}(\bR^2)$, defined by control of finitely many derivatives,
%such as a Sobolev space, and define the adjoint with respect to this
%inner product structure, obtained by concatenating a smoothing
%operator with the formal adjoint. The reader can consult for instance
%\cite{KerSy:94} for more on this idea. For the purposes of this
%discussion we will display the formal gradient (\ref{eqn:gradls})..

Figure \ref{fig:fwigrad6kmbighom} shows the result of the computation
(\ref{eqn:gradls}) for the example given previously, $m=m_0$, $d=S[m_*]f$. The adjoint $D(S[m]f)^T$ is
computed by the adjoint state method
\cite[]{Chavent:74,GauTarVir:86,Plessix:06}. Geometric optics explains
the apparent interrogation of only the triangle spanning the source
and the receiver segment - perturbations of velocity outside of it do
not affect the geometric optics travel times (rays are straight lines,
in constant velocity) connecting source and receivers. However there
is very little indication of a productive update: the edge effects one
sees in this figure (``rabbit ears'') average to very nearly zero when
weighted by the difference between $m_0$ and $m_*$ - that is, the
gradient is essentially orthogonal to the best search direction.

If we use the gradient as a search direction, as one would for the
steepest descent algorithm, for instance, then one obtains the values
of $J_{\rm LS}$ displayed in Figure \ref{fig:fwigrad} - clearly not a
productive search direction.

\section{Source-Receiver Extension}

As illustrated in the previous section, mean-square data misfit tends
to be large, and instensitive to small
velocity changes, except in the immediate vicinity of its global
minimizer. This misfit saturation is the root cause of the poor
performance of local optimization methods for $J_{\rm LS}$..

Several cures for this pathology have been explored. The state of the
art relies on small enough error in the initial velocity estimate,
relative to the longest wavelengths reliably present in the data. The non-overlapping
signal supports of the example given in the last section are
impossible if enough low-frequency signal exists in the
data. Unfortunately frequency content is limited by acquisition
technology. Sometimes it is possible to supply a sufficiently
accurate initial model in view of spectral limitations, and sometimes
it is not, or even to know
whether one has been successfully supplied

Another class of alternatives extends the simulation map
$(m,f) \mapsto S[m]f$ to a larger domain, so chosen that the ``relaxed''
inverse problem set in this larger domain always has a small-residual
solution. Some alternative objective must then be supplied, for which
a minimizer (at least for model-consistent data) is a model of the
original type, which by construction also fits the data.
This section introduces and analyzes such a data-fitting {\em
  extension} for the idealized crosswell tomography problem.

Before we introduce the extended simulation map, we redefine $S$ so
that $S[m], m\in M$ is injective, and the linear least squares problem
for $f$ coercive.

Suppose that ${\cal U} \subset M$ consists of slowness fields
satisfying the simple ray geometry hypothesis, having a common normal
neighborhood ${\cal N}$ containing $W_s$ and $W_r$. Suppose also that
${\cal U}$ contains at least one constant.
For typical $\bx_s \in W_s, \bx_r \in W_r$, denote by $r$ and $r_0$
the ray generated by the exponential map connecting $\bx_s,\bx_r$, respectively
the straight line segment between $\bx_s$ and $\bx_r$. Note that there
may be other connecting rays, but any such rays must exit ${\cal N}$ and be associated to larger
travel times than $\tau[m](\bx_s,\bx_r)$, which is the travel time
along $r$. Also, Fermat's principle holds: the travel time along $r$
is minimum amongst travel time along all paths from $\bx_s$ to
$\bx_r$, lying entirely inside of ${\cal N}$. Write
\[
\tau[m](\bx_s,\bx_r) = \int_r \,dl\, m,
\]
with $dl$ denoting the Euclidean arc length element. it follows that
\[
\tau[m](\bx_s,\bx_r) \le \int_{r_0} \,dl, m \le m_{\rm max}|\bx_r -\bx_s|
\bx_s|.
\]
Also, 
\[
\tau[m](\bx_s,\bx_r) \ge \int_r \,dl m_{\rm min} \ge m_{\rm min}
|\bx_r-\bx_s|
\]
since $r_0$ is geodesic for the Euclidean metric. Since we have
presumed that ${\cal U}$ contains at least one constant slowness
field, both bounds can be approximated arbitrarily well.
Set
\begin{eqnarray}
\label{eqn:rdef}
r_{\rm max} &=& \sup \{|\bx_r-\bx_s|: \bx_s \in W_s, \bx_r \in
                W_r\},\nonumber\\
r_{\rm min} &=& \inf \{|\bx_r-\bx_s|: \bx_s \in W_s, \bx_r \in W_r\}.
\end{eqnarray}
Putting these simple results together, it follows that
\begin{eqnarray}
\label{eqn:tauglob}
m_{\rm max}r_{\rm max} & = & \sup \{\tau[m](\bx_s,\bx_r): \bx_s \in W_s, \bx_r \in 
  W_r, m \in {\cal U}\} \nonumber\\
m_{\rm min}r_{\rm min} |& = & \inf \{\tau[m](\bx_s,\bx_r): \bx_s \in W_s, \bx_r \in 
  W_r, m \in {\cal U}\}
\end{eqnarray}
As these quantities will recur throughout the arguments to come,
define
\begin{eqnarray}
\label{eqn:tauextr}
\tau_{\rm max}& = & m_{\rm max}r_{\rm max} \nonumber\\
\tau_{\rm min}& = & m_{\rm min}r_{\rm min}
\end{eqnarray}

We now add the only hypothesis on ${\cal U}$ pertaining to the
behaviour of slowness fields outside of the common normal
neighborhood:

The maximum measurement time $t^{\rm max}$ satisfies the {\em
  no-return hypothesis}:
\begin{itemize}
\item
\begin{equation}
\label{eqn:tmaxbd}
t^{\rm max} >  \tau_{\rm max}-\tau_{\rm min}
\end{equation}
\item for any $m \in {\cal U}$ no ray with initial $X=\bx_s \in W_s$ passes over $\bx_r \in W_r$
at any time in the interval $(\tau[m](\bx_s,\bx_r), t^{\rm max}+\tau_{\rm max})$.
\end{itemize}
Note that the second condition means that $t^{\rm max}$ as specified 
here will may be used in the statement of Theorems \ref{thm:gosimple}
and \ref{thm:smdep}. In fact, the larger limit $t^{\rm max} +\tau_{\rm max})$ may be so used. The extra margin will be needed to
make some later estimates uniform over large sets of
slowness fields.

Since rays passing over $W_r$ a second time necessarily leave ${\cal
  N}$, this hypothesis actually concerns the global behaviour of the
slowness field. It is satisfied for any constant slowness in ${\cal
  U}$, and is clearly open. Therefore beginning with an open set of
slownesses in $M^k$, $k \ge 2$, satisfying the simple ray geometry
hypothesis and containing a constant slowness, there must be a subset
satisfying also the no-return hypothesis.

One might naturally ask how a slowness field may be
extended outside the image domain ${\cal N}$ of the exponential map to
satisfy the no-return hypothesis with the largest possible $t^{\rm
  max}$, however we shall not address this question here.

Choose $\Delta > 0$ and define
\begin{equation}
\label{eqn:fwddomdef}
D_{\Delta} = \{f \in L^2(\bR): {\rm supp}\,f \subset (-\tau_{\rm min} +
\Delta, t^{\rm max} -\Delta - \tau_{\rm max})\}
\end{equation}

\begin{theorem}
\label{thm:fwdinj}
Suppose that ${\cal U} \subset M$, $t^{\rm max}$ satisfy the simple 
ray geometry and no-return hypotheses, $\Delta >0$, and $w=1$ on 
$[\Delta, t^{\rm max}-\Delta]$. Then for $m \in {\cal U}$,
$S[m]:D_{\Delta} \rightarrow L^2(\Omega)$ is coercive, and 
\[
f \in D_{\Delta} \,\Rightarrow \, S[m]f (z_r,t;z_s) =
A(z_r,z_s)f(t-T(z_r,z_s)) 
\]
\begin{equation}
\label{eqn:gosimplemod} 
+ w(z_r,t;z_s)\int_{T(z_r,z_s)}^{\infty}
\,ds\,B(z_r,s;z_s)f(t-s). 
\end{equation}
\end{theorem}
\begin{proof}
Inequalities \ref{eqn:tauglob} and \ref{eqn:tmaxbd} imply that if
$t \in (-\tau_{\rm min} +
\Delta, t^{\rm max} -\Delta - \tau_{\rm max})$,  then $t+\tau[m](\bx_s,\bx_r) \in (\Delta ,t^{\rm
  max}-\Delta)$. Thus if $f \in D_{\Delta}$, then $w=1$ on ${\rm
  supp}\,f(\cdot-\tau[m](\bx_s,\bx_r))$, establishing
\ref{eqn:gosimplemod}. The injectivity of $S[m]$ follows from Picard
iteration, applied to \ref{eqn:gosimplemod}.
\end{proof}

\begin{definition} (Source-receiver extension)
Suppose that ${\cal U} \subset M$, $t^{\rm max}$ satisfy the simple
ray geometry and no-return hypotheses. Define $\bar{w}_0$ to be the
characteristic function of $\{t<t^{\rm max}\}$ in $\Omega$.

Define the
{\em source-receiver} extended source domain
\[
\oD_0 = \{\of \in L^2(\Omega): {\rm supp}\,f \subset 
Z_r\times Z_s 
\]
\begin{equation}
\label{eqn:extdomdef}
\times (-\tau_{\rm max}, t^{\rm max} - \tau_{\rm  min})\}
\end{equation}
and extended modeling operator 
\be\label{eqn:extdom}
\oS[m]: \oD_{0} \rightarrow L^2(\Omega) 
\ee
by
\be\label{eqn:extmap}
\oS[m]\of (z_r,t;z_s) = \bar{w}_0(z_r,t;z_s)\bar{p}(\bx_r,t;\bx_s).
\ee
\end{definition}
in which $\bar{p}$ is the solution of the wave equation
\ref{eqn:pressure} with $f(t) = \of(z_r,t;z_s)$.  That is, $\oS$ acts
by computing each data trace $\bar{p}(\bx_r,t;\bx_s)$ as if it were
generated by an indepedent source function $\of(z_r,\cdot;z_s)$ for
each source and receiver location.

Note that the time interval in the definition of $\oD_0$ strictly
contains that in the definition of $D_{\Delta}$.
Define the extension map $E: D_{\Delta} \rightarrow \oD_0$ by 
$(Ef)(z_t,t;z_s) = f(t)$, with $f$ on the RHS being regarded as
a constant function of $z_sz_r$. Then $\oS$ is an
extension of $S$, in the sense that
\be\label{eqn:extprop}
\oS[m](Ef) = S[m]f,\,f \in D_{\Delta}.
\ee

At first blush, it would appear that computing $\oS$ would be a very
expensive proposition, even after discretization in $z_s$ and $z_r$,
as each trace would have to be simulated independently. However this
is not the case, as \be\label{eqn:extgreen} \oS[m]\of(z_r,t;z_s) =
\bar{w}_0(z_r,t;z_s(G[m](\bx_r,\cdot;\bx_s) *_t \of(,z_r,\cdot;z_s))(t), \ee
with $G[m]$ being the Green's function of the problem
(\ref{eqn:pressure}), defined above. $G[m]$ can be approximated
numerically, one numerical simulation for each value of $z_s$. Once
this is done, the evaluation of $\oS[m]\of$ is a relatively
inexpensive convolution for each trace.

The main advantage of the extended model is that any data can be fit -
for any slowness model (within the class specified in the last
section)! More precisely, define 
\begin{equation}
\label{eqn:extrange}
\oF_0 = \{d \in L^2(\Omega): d=0 \mbox{ if } t<0 \mbox{ or } t>t^{\rm 
  max} \}. 
\end{equation}
\begin{theorem}
\label{thm:extsur}
Suppose that ${\cal U} \subset M$, $t^{\rm max}$ satisfy the simple 
ray geometry and no-return hypotheses
Then for $m \in {\cal U}$,
\begin{equation}
\label{eqn:extsur}
\oF_0 \subset \oS[m]\oD_0.
\end{equation}
Define $\oS_0[m]: L^2(\Omega) \rightarrow L^2(\Omega)$ and its inverse by
\begin{eqnarray}
\label{eqn:approxrinv}
\oS_0[m]\of(z_r,t;z_s) & = &
A[m](z_r,z_s)\of(z_r,t-T[m](z_r,z_s);z_s),\nonumber\\
\oS_0^{-1}[m]d(z_r,t;z_s) &=&
A[m]^{-1}(z_r,z_s)d(z_r,t+T[m](z_r,z_s);z_s),
\end{eqnarray}
and $\bar{R}_0[m]: L^2_{\rm comp}(\Omega) \rightarrow L^2_{\rm comp}$
by
\[
\bar{R}_0[m]d(z_r,t;z_s) 
\]
\begin{equation}
\label{eqn:approxres}
= \frac{\bar{w}_0(z_r,t+T[m](z_r,z_s);z_s)}{A(z_r,z_s)}
\int_0^t\,ds\,B[m](z_r,t+T[m](z_r,z_s)-s;z_s)d(z_r,s;z_s)
\end{equation}
Then 
\begin{equation}
\label{eqn:rinvdef}
\oS^R[m] = \oS_0^{-1}
\sum_{k=0}^{\infty}(-\bar{R}_0[m])^k \bar{w}_0
\end{equation}
converges in norm to a bounded operator $\oS^R[m]: \oF_0
\rightarrow \oD_0$, and $\oS^R[m]$ is a right inverse
for $\oS[m]$ on $\oF_0$, that is, for $d \in \oF_0$,
\begin{equation}
\label{eqn:invreln}
\oS[m]\oS[m]^Rd = d.
\end{equation}
\end{theorem}

\begin{proof} 
Observe that  the definition \ref{eqn:extmap} can be re-written using
the Green's function components $A, T,$ and $B$ and the cutoff function
$\bar{w}_0$ in the form
\[
\oS[m]\of(z_r,t;z_s) = \bar{w}_0(t) \left(
A[m](z_r,z_s)\of(z_r.t-T[m](z_r,z_s);z_s) \right.
\]
\begin{equation}
\label{eqn:extmapexp0}
\left.+ \int_{T[m](z_r,z_s)}^{\infty}\,ds\,B[m](z_r,s;z_s)\of(z_r,t-s;z_s)\right)
\end{equation}
According to the no-return hypothesis, $B[m]$ is well-defined and
smooth for $t \le t^{\rm max} + \tau_{\rm max}$. For $\of \in
\oD_0$, the upper limit of integration may be replaced by $t + \tau_{\rm max}$. Since only
$t\le t^{\rm max}$ is significant on the right-hand side of
\ref{eqn:extmapexp0}, the integrand in the last summand is
well-defined, and the integral defines an
operator $\bar{B}[m]: \oD_0  \rightarrow L^2(\Omega)$,
\begin{equation}
\label{eqn:barb}
\bar{B}[m]\of(z_r,t;z_s) = 
\int_{T[m](z_r,z_s)}^{t^{\rm max}+\tau_{\rm max}}\,ds\,B[m](z_r,s;z_s)\of(z_r,t-s;z_s)
\end{equation}
Then equation \ref{eqn:extmapexp0} can be abbreviated as
\begin{equation}
\label{eqn:extmapexpabbrev}
\oS[m] = \bar{w}_0 (\oS_0[m] + \bar{B}[m]).
\end{equation}
From the definitions of $\oS_0,\bar{R}_0$,
\[
\bar{w}_0\bar{B}[m] = \bar{R}_0[m]\oS_0[m]
\]
whence
\begin{equation}
\label{eqn:extmapexp}
\oS[m]\of = \bar{w}_0(I + \bar{R}_0[m]) \oS_0[m]\of.
\end{equation}

Note that $\bar{R}_0$ preserves supports of the form $t_0 \le t \le
t^{\rm max}$, in particular maps $\oF_0$ into itself.
Straightforward estimates using a weighted $L^2$ norm with decreasing
exponential weight in $t$ shows that as an operator on $\oF_0$, $I + \bar{R}_0[m]$ is invertible,
and its inverse is given by the Neumann series appearing in the
definition \ref{eqn:rinvdef} of $\oS^R[m]$.

For $d \in \oF_0$, then, $\bar{w}_0 d = d $, and
$\sum_{k=0}^{\infty}(-\bar{R}_0[m])^kd  \in \oF_0$. The definition of
$\oD_0$ is concocted to guarantee that if $d \in \oF_0$, then
$\oS[m]^{-1}d \in \oD_0$ for any $m \in {\cal U}$. Thus equation
\ref{eqn:rinvdef} defines a map from $\oF_0$ to $\oD_0$. While in
general $\oS_0[m]\oD_0$ is not a subset of $\oF_0$,
$\oS_0[m]\oS_0[m]^{-1} = I $ so membership in $\oF_0$ is preserved:
\[
\oS_0[m]\oS_0[m]^{-1}\sum_{k=0}^{\infty}(-\bar{R}_0[m])^kd  \in \oF_0.
\]
Tacking on the other factors in the expansion \ref{eqn:extmapexp} of
$\oS[m]$,  the conclusion \ref{eqn:invreln} follows, which in turn
implies the inclusion \ref{eqn:extsur}.
\end{proof}

\section{Extended Full Waveform Inversion}
Theorem \ref{thm:extsur} shows that data fitting with $\oS$, rather
than $S$, cannot suffer from the difficulty identified in Theorem
\ref{thm:FWIconv}, that is, failure to fit the data until the slowness
field predicts the right kinematics within a wavelength or so: in
fact, data can be fit via $\oS$ with essentiallly any slowness, however
kinematically inaccurate.. However
this pleasant property does not in itself yield a solution of the
inverse problem, because it is achieved at the price of adding a large
number of parameters to the models space. This enlargement of the
model space has two negative consequecnes:
\begin{itemize}
\item[1. ] The extended modeling map $\oS$ is not
injective - it is not possible to define the domain and range so that
this map is bijective for an open set ${\cal U}$ of slowness
models. 
\item[2. ] More fundamentally, since data
can be fit with any slowness model at all, data fitting no longer
constrains the slowness model.
\end{itemize}
Item 2 is a
consequence of the unphysical nature of the extended source, with each
data trace modeled by an independent source trace. The large number of
parameters added in extending the model must somehow be controlled, and 
forced to assume trivial values in order to recover  a slowness $m$
and physical source $f$ that together explain the data.

Compensation for the null space of $\oS[m]$ is straightforward via
Tihonov regularization. That is, choose a regularization parameter
$\alpha >0$ and define the inverted extended source $\ofa[m,d], m \in
M, d \in L^2(\Omega)$, by
\begin{equation}
\label{eqn:regextinv}
\ofa[m,d] = \mbox{ argmin}_{\of}\,\frac{1}{2}(\|\oS[m]\of - d\|^2 + \alpha^2\|\of\|^2).
\end{equation}
If ${\cal U}$ and $t^{\rm max}$ are chosen as in the preceding
discussion, that is, to satisfy the simple ray geometry and no-return
hypotheses, and $m \in {\cal U}, d \in \oF_0$, then
as $\alpha \rightarrow 0$, $\ofa[m,d] \rightarrow \oS[m]^Rd$.

A hint about how the second issue might be addressed comes from
further examination of the synthetic example defined earlier. Recall
that $m_*$ is the target slowness 
(corresponding to the velocity depicted in Figure \ref{fig:velbig})
used to generate the data $d=S[m_*]f$ (Figure \ref{fig:shot6kmbig}).
Figures \ref{fig:shot6kmbig_bigdecon} and \ref{fig:shot6kmbig_decon} display
$\ofa[m,d]$ for $d=S[m_*]f$ and $m=m_*$ and $m=m_0 = 0.5 s/km$
respectively. In both cases, the regularization parameter $\alpha$ is
set = $10^{-3}$. The non-extended or
physical source $f$ is localized near $t=0$. The inverted extended
source for correct slowness (Figure \ref{fig:shot6kmbig_bigdecon})
yields a result very close to the extension $Ef$, and has the same
quality (energy focused at $t=0$, independent of $z_r$). Inversion with the incorrect slowness
shifts the extended source into $t \ne 0$. Also, the level surfaces of
$\ofa[m_0,d]$ (Figure \ref{fig:shot6kmbig_decon}) are not
constant in $z_r$, but vary significantly. If this were not true, then
any $z_r=const.$ trace could be taken as the source function $f$, and
the inverse problem would be solved.

These observations suggest two alternative methods for measuring the
correctness of $m$ from the properties of $\ofa[m,d]$. First,
concentration near $t=0$ may be measurred by multplying $\ofa[m,d]$ by $t$ and taking the $L^2$ norm of the result. If the velocity
is correct, {\em and} the target source function actually is supported
near $t=0$, this norm will be small, as the support of $\ofa[m,d]$
will be close to $t=0$ also (see Figure \ref{fig:shot6kmbig_bigdecon}). Second, the variance with respect to $z_r$ may
be measured by taking the norm of the $z_r$ derivative, which will be
zero (not just small) for an extended model in the range of the
extension operator. That is, the range of the extension operator (the
``physical'' extended models is the null space of the $z_r$ derivative.
Either case may be captured by small values of the objective function
\be\label{eqn:dsdef}
J_{\rm DS}[m,d] = \frac{1}{2}\|A\ofa[m,d]\|_{L^2(\Omega)}^2,
\ee
in which
\begin{itemize}
\item $A\of(z_s,z_r,t) = t\of(z_s,z_r,t)$, or
\item $A\of(z_s,z_r,t) = \nabla_{z_s,z_r}\of(z_s,z_r,t)$.
\end{itemize}
Operators $A$ which indirectly detect the kinematic correctness of a
wave propagation model via qualities of auxiliary data generated with
that model (in this case, the extended source estimate $\ofa[m,d]$)
have come to be known as {\em annihilators}, since in many cases the
auxiliary data corresponding to correct models is in (or near) the
null space. Model estimation via minimizations of functions such as
$J_{\rm DS}$ is sometimes called {\em differential semblance}
optimization, after a name introduced in some of the early papers on
this concept \cite[]{SymesCar:91,KerSy:94} to reflect the use of
differential operators as annihilators. A differential annihilator
(the second option above) was used in the first papers known to us on the
source-receiver extension \cite[]{SongSymes:94b,Symes:94c}. A brief
survey of the literature on this topic appeared above, in the
Introduction.

Both choices have drawbacks. The first choice (multiplication by $t$)
defines a bounded operator on $D_0$, however without a null space;
without some further compactness-inducing constraint on $\of$, a
minimizer of $J_{\rm DS}$ may not exist. For computational purposes,
this defect appears to be minor, as discretization effectively forces
the search for $\ofa$ into a compact subspace of
$L^2(\Omega)$. However this effect has not been analyzed; the applied
literature uses it without mathematical justification (but with
considerable success - see for example \cite{Warner:16}). The second
choice has a complementary problem: the derivative with respect to
$z_r,z_s$ is of course unbounded, rendering $J_{\rm DS} $ only densely
defined. The obvious remedy is to use a subspace of $H^1(\Omega)$ as
the domain of $\oS$ - see \cite{KerSy:94} for more discussion of this
option, and implementation details.

We will accept the drawbacks of the first option above (multiplication
by $t$) in the remainder of our discussion, simply because the
implementation is simpler and is the approach taken in the
computational examples accompanying this paper.

\section{Gradient and Hessian}

Throughout this section we will assume without further comment that
${\cal U}$ and $t^{\rm max}$ are chosen to satisfy the simple ray
geometry and no-return hypotheses, as explained earlier.

Recall from Theorem \ref{thm:smdep}, and surrounding discussion,
especially Remark \ref{rem:derivloss}, that $(m,f) \mapsto S[m]f$ is
only differentiable in $m$ if $f$ has one square integrable derivative
or better. The same is true fore $\oS$. The local
behavioiur of $J_{\rm DS}$ follows from unpacking the directional derivative of
$\oS$ at $m \in {\cal U}$ in a direction $\delta m \in
C^{\infty}(\bR^3)$ into a leading term with explicit dependence on $\partial
\of/ \partial t$, plus an $L^2$-continuous remainder.
The leading term factors as $\oS[m]Q[m]\delta m$, in which $Q$ is a
first order differential operator:
\begin{equation}
\label{eqn:qdef}
Q[m]\delta m =\frac{1}{A[m](z_r,z_s)}\left(-(DT[m]\delta
m) (z_r,z_s)\frac{\partial}{\partial t} + (D A[m]\delta m)(z_r,z_s)\right)
\end{equation}
A suitable domain for $Q[m]\delta m$ is
\begin{equation}
\label{eqn:d1def}
\oD_1 = H^1_0(Z_r \times Z_s \times [-\tau_{\rm max},t^{\rm
  max}-\tau_{\rm min}])
\end{equation}
Note that $Q[m]\delta m$ is {\em essentially skew-symmetric}, that is
\begin{equation}
\label{eqn:qess}
(Q[m]\delta m) + (Q[m]\delta m)^T =2 DA[m]\delta m.
\end{equation}
The right-hand side of equation \ref{eqn:qess} is a differential
operator of order zero, in particular bounded on $L^2(\Omega)$,
whereas $Q$ is first order.
\begin{theorem}
\label{thm:q}
There exists continuous map $P: {\cal U} \times C^{\infty}(\bR^3)
\rightarrow {\cal L}(\oD_0,\oD_0)$, linear in its second argument, 
so that for $\of \in \oD_1$,
\begin{equation}
\label{eqn:paramderiv}
D(\oS[m]\of)\delta m = \oS[m]((Q[m]\delta m) + (P[m]\delta
m))\of
\end{equation}
$P$ is a Volterra convolution operator of the first kind. In
particular, the operators $P$ and $Q$ commute.
\end{theorem}
\begin{proof}
Suppressing $z_r,z_s$ for the moment, recall the definition of $\oS_0[m]$
\[
(\oS_0[m]\of)(t) = A[m]\of(t-T[m]).
\]
Also, from the defintion \ref{eqn:barb}, 
\[
(\bar{B}[m]\of)(t) = 
\int_{T[m]}^{\infty}\,ds\,B[m](s)\of(t-s) 
\]
\[
=\int_{-\infty}^{t-T[m]}\,ds\,B[m](s)(t-s)\of(s)
\]
\begin{equation}
\label{eqn:barbalt}
= (\oS_0[m]\tilde{B}[m])\of(t),
\end{equation}
where
\begin{equation}
\label{eqn:tildeb}
(\tilde{B}[m])\of(t) = \int_{-\infty}^{t}\,ds\,B[m](t+T[m]-s)\of(s) 
\end{equation}
In order to view $\tilde{B}[m]$ as a Volterra integral operator on $D_0$ with
smooth kernel, it is necessary to extend $B[m]$ smoothly to $t >
t^{\rm max} + \tau_{\rm max}$. The extended values do not play
any role in the result of the computation \ref{eqn:barbalt} for $t >
t^{\rm max}$.

These manipulations combine to yield a re-write of the expression
\ref{eqn:extmapexpabbrev} for $\oS[m]$:
\begin{equation}
\label{eqn:extmapexpabbrevalt}
\oS[m] = \bar{w}_0 \oS_0[m](I + \tilde{B}[m]).
\end{equation}
Its directional derivative at $m \in {\cal U}$ in the direction
$\delta m \in C^{\infty}(\bR^3)$ is
\[
D(\oS[m]\of)\delta m = \bar{w}_0( (D(\oS_0[m])\delta m) (I +
\tilde{B}[m])\of+\oS_0[m]D(\tilde{B}[m]\of)\delta m)
\]
Note that 
\[
(D(\oS_0[m]\of)\delta m)(t) = (DA[m]\delta m)\of(t-T[m]) -A[m](DT[m]\delta
m)\frac{\partial \of}{\partial t}(t-T[m])
\]
\[
=\oS_0[m]\left(-\frac{(DT[m]\delta m)}{A[m]}
  \frac{\partial \of}{\partial t} + (D\log A[m]\delta m)\of\right)(t)
\]
so 
\[
\bar{w}_0(D(\oS_0[m]\of)\delta m) =  \bar{w}_0\oS_0[m]Q[m]\delta m
\]
whence
\[
D(\oS[m]\of)\delta m = \bar{w}_0\oS_0[m] (Q[m]\delta m)(I+\tilde{B}[m])\of 
\]
\[
+ \bar{w}_0\oS_0[m]D(\tilde{B}[m]\of)\delta m 
\]
Since $\tilde{B}$ is a convolution operator in $t$, it commutes with
multiplication by functions of $z_s,z_r$, and with $\partial/\partial
t$, so this is actually equation \ref{eqn:paramderiv}, provided that
$P$ is defined as 
\begin{equation}
\label{eqn:pdef}
P[m]\delta m = (I + \tilde{B}[m])^{-1}D\tilde{B}[m] \delta m
\end{equation}
Since both factors on the right-hand side of equation \ref{eqn:pdef}
are Volterra convolution operators (of second and first types
respectively), so their product is a Volterra convolution operator of
the first kind, and $P$ commutes with $Q$ as claimed.
\end{proof}

In the remainder of this section we will use the equivalence between $D(\oS[m]\of)\delta
m$, the directional derivative of the vector-valued function $m
\mapsto \oS[m]\of$,  and $(D\oS[m]\delta m)\of$, the directional
derivative of the operator-valued function $m \mapsto \oS[m]$,
evaluated at the vector $\of$, as is valid if $\of \in \oD_1$.

Just as existence of a derivative with respect to $m$ of $\oS[m]\of$
requires that $\of$ have a square-integrable time derivative, so
existence of a derivative with respecto $m$ of $\oS[m]^Td$ requires
that $d$ have a time derivative. Accordingly, define
\[
\oF_1 = \oF_0 \cap H^1(\Omega)
\]
and metrize $\oF_1$ with the Sobolev 1-norm.

\begin{theorem}
\label{thm:qcomm}
For $d \in \oF_1$, $m \in {\cal U}$,
\[
DJ_{\rm DS}[m,d]\delta m =\alpha^2 \langle ((P[m]\delta m)^T +(P[m]\delta
m))\ofa[m,d],\oga[m,d]\rangle 
\]
\begin{equation}
\label{eqn:qderiv}
-\langle (P[m]\delta
m)\ofa[m,d],A^TA\ofa[m,d]\rangle + \langle [Q[m]\delta m,A^TA]\ofa[m,d],\ofa[m,d]\rangle.
\end{equation}
For $m \in {\cal U}$, $\delta m \in C^{\infty}(\bR^3)$, the linear
functional on $\oF_1$ defined by $d \mapsto DJ_{\rm DS}[m,d]\delta m$
extends by continuity to $\oF_0$.
\end{theorem}
\begin{proof}
The normal equation equivalent to \ref{eqn:regextinv} are:
\begin{eqnarray}
\label{eqn:normal}
N_{\alpha}[m]\of &=& \oS[m]^Td\nonumber\\
N_{\alpha}[m] & = & \oS[m]^T\oS[m] + \alpha^2 I
\end{eqnarray}
So
\begin{equation}
\label{eqn:ofanorm}
\ofa[m,d] = N_{\alpha}[m]^{-1}\oS[m]^Td
\end{equation}
and 
\[
J_{\rm DS}[m,d] = \frac{1}{2}\|A N_{\alpha}[m]^{-1}\oS[m]^Td\|^2
\]
For $d \in \oF_1$,
\[
DJ_{\rm DS}[m,d] \delta m = \langle A (DN_{\alpha}[m]^{-1}\delta m)
\oS[m]^Td + AN_{\alpha}[m]^{-1}D(\oS[m]^Td)\delta m,A\ofa[m,d]\rangle
\]
\[
=\langle - N_{\alpha}[m]^{-1}(DN_{\alpha}[m]\delta m)\ofa[m,d]
  +N_{\alpha}[m]^{-1}D(\oS[m]^Td)\delta m,A^TA\ofa[m,d]\rangle
\]
Introduce $\oga[m,d]$, defined by
\begin{equation}
\label{eqn:gdef}
\oga[m,d] = N_{\alpha}[m]^{-1}A^TA\ofa[m,d].
\end{equation}
Then
\[
DJ_{\rm DS}[m,d] \delta m = \langle - (DN_{\alpha}[m]\delta m)\ofa[m,d] +
  D(S[m]^Td)\delta m, \oga[m,d]\rangle
\]
\begin{equation}
\label{eqn:pregrad1}
= \langle (DS[m]^T\delta m)(d-S[m]\ofa[m,d]) - S[m]^T (DS[m]\delta m)\ofa[m,d],\oga[m,d]\rangle
\end{equation}
From Theorem \ref{thm:q}, this is
\[
=\langle (Q[m]\delta m + P[m]\delta m)^TS[m]^T(d-S[m]
\ofa[m,d]) - S[m]^TS[m]
(Q[m]\delta m + P[m]\delta m)\ofa[m,d],\oga[m,d]\rangle
\]
which because of the normal equation \ref{eqn:normal} is
\[
= \langle (Q[m]\delta m + P[m]\delta m)^T\alpha^2\ofa[m,d],\oga[m,d]\rangle 
-\langle (Q[m]\delta m + P[m]\delta m)\ofa[m,d], (N_{\alpha} -
\alpha^2I)\oga[m,d]\rangle
\]
\[
= \alpha^2 \langle (Q[m]\delta m + P[m]\delta
m)^T\ofa[m,d],\oga[m,d]\rangle 
+ \alpha^2 \langle (Q[m]\delta m + P[m]\delta
m)\ofa[m,d],\oga[m,d]\rangle
\]
\[
-\langle (Q[m]\delta m + P[m]\delta
m)\ofa[m,d],A^TA\ofa[m,d]\rangle
\]
\[
=\alpha^2 \langle ((P[m]\delta m)^T +(P[m]\delta
m))\ofa[m,d],\oga[m,d]\rangle 
\]
\[
-\langle (Q[m]\delta m + P[m]\delta
m)\ofa[m,d],A^TA\ofa[m,d]\rangle
\]
by virtue of the defintion \ref{eqn:gdef}. Finally, the essential skew-symmetry of
$Q[m]\delta m$ (equation \ref{eqn:qess}) implies that
\[
-\langle (Q[m]\delta m)\ofa[m,d],A^TA\ofa[m,d]\rangle
\]
\[
=-\langle [A^TA,Q[m]\delta m]\ofa[m,d],\ofa[m,d]\rangle- \langle A^TA
\ofa[m,d],(Q[m]\delta m)^T\ofa[m,d]\rangle
\]
\[
=-\langle [A^TA,Q[m]\delta m]\ofa[m,d],\ofa[m,d]\rangle +\langle
(Q[m]\delta m - 2DA[m]\delta m)\ofa[m,d],A^TA\ofa[m,d]\rangle
\]
whence
\begin{equation}
\label{eqn:qcomm}
-\langle (Q[m]\delta m)\ofa[m,d],A^TA\ofa[m,d]\rangle =
\langle\left(\frac{1}{2} [Q[m]\delta m,A^TA] + A^TA DA[m]\delta m\right)\ofa[m,d],\ofa[m,d]\rangle,
\end{equation}
and equation \ref{eqn:qderiv} is established.

Recall that $Q[m]\delta m$ is a $t$-independent multiple of
$\partial/\partial t$, and $A = $ multiplication by $t$. Therefore
\begin{equation}
\label{eqn:qcommdt}
[Q[m]\delta m,A^TA] = 2t\frac{DT[m]\delta m}{A[m]}
\end{equation}
defines a bounded multiplier on $\oD_0$. Since all of the other operators
appearing in equation \ref{eqn:qderiv} are bounded on $\oD_0$, and
$\ofa[m,d]$ and $\oga[m,d]$ are continuous in $d \in \oF_0$, 
continuous extension of $DJ_{\rm DS}[m,d]\delta m$ to $d \in \oF_0$ follows.
\end{proof}

To understand how the term involving the commutator $[Q,A^TA]$ may
dominate the derivative of $J_{\rm DS}$, we introduce a measure of
oscillation in time:

\begin{definition} For $u \in L^2(\Omega)$, $a < b$,
\begin{equation}
\label{eqn:lamdef}
\lambda_{[a,b]}(u) =
\left(\frac{\int_{Z_r}\,dz_r\,\int_{Z_s}\,dz_s\,\int_a^b\,dt,\,\left(\int_a^t\,ds\,u(z_r,s;z_s)\right)^2}{\int_{Z_r}\,dz_r\,\int_{Z_s}\,dz_s\,\int_a^b\,dt\,u^2}\right)^{\frac{1}{2}}.
\end{equation}
\end{definition}
Essentially, $\lambda$ is the ratio of the partial Sobolev $W^{-1,2}$ norm to
the $L^2$ norm, and measures oscillation in $t$. If $u(t) = \sin \pi \omega
t$, then $\lambda_{[a,b]}(u) = \frac{C}{ \omega} + O(\omega^{-2})$, so
$\lambda$ is more or less wavelength, hence the notation. Its
importance for the analysis of the gradient lies in a simple lemma,
the proof of which we leave to the reader:

\begin{lem}
\label{lem:lamlem}
Suppose that $k \in C^1(\Omega)$, $a < b$, and define for $u \in L^2(\Omega)$
\[
Ku(z_r,z_s,t) = \int_a^t\,ds\,k(z_r,t-s;z_s) u(z_r,s;z_s).
\]
Then 
\[
\|Ku\|_{L^2(Z_r\times Z_s \times [a,b])} \le
(2+(b-a)^2)^{\frac{1}{2}}\|k\|_{C^1(\Omega)}\lambda_{[a,b]}(u)\|u\|.
\]
Also, if ${\rm supp} \,u \subset [a,b]$ and $\int_a^bu = 0$, then $a' \le a < b \le b'$
implies that
\[
\lambda_{[a',b']}(u) = \lambda_{[a,b]}(u)
\]
and
\[
\|K^Tu\|_{L^2(Z_r\times Z_s \times [a,b])} \le
(2+(b-a)^2)^{\frac{1}{2}}\|k\|_{C^1(\Omega)}\lambda_{[a,b]}(u)\|u\|.
\]
\end{lem}

\begin{theorem}
\label{thm:qerr}
Under the hypotheses of Theorem \ref{thm:qcomm}, for $d \in \oF_0$ for
which
\[
\int_0^{t^{\rm max}}\,d(z_r,t;z_s) = 0, z_r \in Z_r, z_s \in Z_s,
\]
there is $C >0$
depending on ${\cal U}$ and $t^{\rm max}$ so that
\[
|DJ_{\rm DS}[m,d]\delta m - \langle [Q[m]\delta
m,A^TA]\ofa[m,d],\ofa[m,d]\rangle| 
\]
\[
\le C \alpha^{-2}\lambda_{[0,t^{\rm max}]}(d)(1+\alpha^{-2}\lambda_{[0,t^{\rm max}]}(d))\|d\|^2
\]
\end{theorem}
\begin{proof}
Every operator appearing in the following expressions depends on $m
\in {\cal U}$, and $\ofa$ and $\oga$ depend on $d$ as well. Since these
dependencies are universal, we suppress them from the notation.

Using the expansion \ref{eqn:extmapexpabbrevalt}, rewrite the normal
equation \ref{eqn:normal} as
\[
(\alpha^2 I + (I+\tilde{B})^TS_0^t \bar{w}_0^2 S_0 (I+\tilde{B})) \ofa
= (I+\tilde{B})^TS_0^T \bar{w}_0d
\]
\[
=S_0^T d +\tilde{B}^TS_0^T d
\] 
from the definition of $\oF_0$. The left-hand side is
\[
= (\alpha^2 + S_0^T\bar{w}_0^2S_0) \ofa  + K \ofa,
\]
where
\[
K = \tilde{B}^T S_0^T\bar{w}_0^2S_0 + S_0^T\bar{w}_0^2S_0\tilde{B}
+ \tilde{B}^T S_0^T\bar{w}_0^2S_0\tilde{B}.
\]
Decompose $\ofa = \ofa^0 + \ofa^1$, where
\[
\ofa^0 = (\alpha^2 + A^2)^{-1}S_0^T d.
\]
Tracking supports, you see that
\[
S_0^T\bar{w}_0^2S_0 \ofa^0 = \S_0^TS_0\ofa^0 = A^2 \ofa^0,
\]
so 
\[
(\alpha^2 + S_0^T\bar{w}_0^2S_0) \ofa^0 =S_0^T d.
\]
Therefore, 
\[
N_{\alpha}\ofa^1 = \tilde{B}^TS_0^T \bar{w}_0d - K\ofa^0
\]
Since $\ofa^0$ is a multiple of a translate of $d$, from Lemma \ref{lem:lamlem} 
\[
\lambda_{[-\tau_{\rm max},t^{\rm max}-\tau_{\rm min}]}(\ofa^0) =
  \lambda_{[0,t^{\rm max}]}(d).
\]
Every summand in $K$ has a factor of $\tilde{B}$ or $\tilde{B}^T$ (or
both). Lemma \ref{lem:lamlem} implies that
\[
\|N_{\alpha}\ofa^1\| \le C \lambda_{[0,t^{\rm max}]}\|d\|
\] 
for suitable $C$ depending on $A, B$ hence on ${\cal U}$, $t^{\rm
  max}$. Since $N_{\alpha}-\alpha^2I \ge 0$ and $\|\ofa^0\|\le C\|d\|$,
obtain
\begin{eqnarray}
\label{eqn:f1barbd}
\|\ofa^1\| & \le & C \alpha^{-2}\lambda_{[0,t^{\rm max}]}(d))\|d\|\\
\label{eqn:fbarbd}
\|\ofa\| & \le & C(1 + \alpha^{-2}\lambda_{[0,t^{\rm max}]}(d))\|d\|
\end{eqnarray}
hence
\[
\alpha^2\|\oga\| \le C(1 + \alpha^{-2}\lambda_{[0,t^{\rm max}]})\|d\|.
\]
Finally, estimate $P \ofa = P \ofa^0  + P \ofa^1$ by applying Lemma
\ref{lem:lamlem} again to the first summand on the right, and using
the already-established bound for $\|\ofa^1\|$ to bound the second
term. Similar reasoning applied to $P^T \ofa$ and appeal to Theorem
\ref{thm:qcomm} establishes the result.
\end{proof}

\section{Transmission Caustics}

\section{Discussion}

\section{Conclusion}

\newpage

\inputdir{project}
%\plot{csqlens10}{width=\textwidth}{Lens: diameter = 0.5 km, thickness = 1.0 km, background velocity = 2 km/s, velocity at center = 1.4142  km/s.}
%\plot{shot6kmlens10conv}{width=\textwidth}{Shot gather - source at x=5100 m, z=10 m. Receivers between 4 km and 6km at depth = 3.0 km.}
%\plot{awicurve}{width=\textwidth}{RMS amplitudes of inversions for
%trace-dependent source (matched source), scaled by t (second moment
%objective). Convex combinations of homogeneous (2 km/s) and lens models.}
\plot{velbig_dec_exp}{width=\textwidth}{Example: $v$ = 2 km/s for $z \le
  0.5 km$, $=2.0 + (z-0.5)*0.5$ km/s for $z>0.5$ km. Source location
  $x=5100$, $z=10$ m and receiver locations $x=3000 - 6000$ m,
$z=3000$ m indicated.}
\plot{shot6kmhom}{width=\textwidth}{Shot gather for geometry indicated
  in Figure \ref{fig:velbig_dec_exp}, generated with homogeneous velocity:
  $v_0=2$ km/s}
\plot{shot6kmbig}{width=\textwidth}{Shot gather for geometry and
  $v$ as in Figure \ref{fig:velbig_dec_exp}.}
\plot{shot6kmbig_decon}{width=\textwidth}{Inverse image of data in 
  Figure (\ref{shot6kmbig}) under extended modeling operator at $v_0$
  = 2 km/s}
\plot{shot6kmbig_bigdecon}{width=\textwidth}{Inverse image of data in 
  Figure \ref{shot6kmbig} under extended modeling operator at target
  $v$ (Figure \ref{fig:velbig_dec_exp}).}
\plot{fwigrad6kmbighom}{width=\textwidth}{FWI gradient for shot gather
defined in Figure \ref{fig:velbig_dec_exp}, at $v=v_0$ = 2 km/s. Scaled by
10$^6$ for plotting. Note very
small values, and sign is positive, indicating that FWI objective {\em
increases} in the direction of the target velocity.}
\plot{fwiscan}{width=\textwidth}{Plot of values of the FWI objective
  along the line segment between $v_0$ and $v$, for the shot gather
  described in Figure \ref{fig:velbig_dec_exp}. For abscissa $h$, value
  plotted is FWI objective at $h v_0 + (1-h) v$. Note that objective
  does in fact initially increase from its value at $v_0$, consistent
  with the positive average value of the gradient as shown in Figure \ref{fig:fwigrad6kmbighom}. }
\plot{ivagrad6kmbighom}{width=\textwidth}{Source-receiver Extended FWI
  gradient, same geometry, data, and velocity model as in Figure
  \ref{fig:fwigrad6kmbighom}. Note that overall sign is negative, indicating
  correct conclusion that $v_0$ is slow.}
\plot{fwistack}{width=\textwidth}{FWI gradient at $v=v_0$ for line of shots 
  extending from 2100 m to 8000 m. Scaled by
10$^6$ for plotting. Receiver locations are windowed so 
  that the range of offsets is the same for all shots. In effect, the 
  gradient is a stack of translates of the gradient in Figure
  \ref{fig:fwigrad6kmbighom}.}
\plot{ivastack}{width=\textwidth}{Source-receiver Extended FWI gradient at $v=v_0$ for line of shots 
  extending from 2100 m to 8000 m. Receiver locations are windowed so 
  that the range of offsets is the same for all shots. In effect, the 
  gradient is a stack of translates of the gradient in Figure
  \ref{fig:ivagrad6kmbighom}.}

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/ms/q.tex
repro docs dir added

\title{Computing Q}
\author{William W. Symes}

\begin{abstract}
abstract goes here
\end{abstract}

Basic physics: acoustics
\be
\label{awe}
\frac{\partial^2 u}{\partial t^2} - m \nabla^2 u = S
\ee

Forward map: $F(m) = Pu$, $P$ = trace operator on receiver surface $\Sigma$ (sampling).

Source extension: $\oF(m,\oS) = Pu$ with $S$ in (\ref{awe}) replaced by $\oS$.

Derivative: $D\oF(m)\delta m = P \delta u$ where
\be
\label{dawe}
\frac{\partial^2 \delta u}{\partial t^2} - m \nabla^2 \delta u = \delta m \nabla^2 u
\ee

Duhamel's principle:
\[
\delta u(x,z,t) = \int_0^t ds v(x,z,t,s)
\]
where
 \begin{eqnarray}
\label{vawe}
\frac{\partial^2 v}{\partial t^2} - m \nabla^2 v &=& 0,\,\,t \ge s; \nonumber\\
v(x,z,s,s)&=&0; \nonumber\\
\frac{\partial v}{\partial t}(x,z,s,s)&=& (\delta m \nabla^2 u)(x,z,s). 
\ee

\noindent {\bf Assumption 1:} 
\be
\label{supp}
\mbox{supp }\delta m \cap \mbox{supp }\oS = \emptyset
\ee

Then can rewrite 3rd condition in (\ref{vawe}) as 
\be
\label{vawe3}
\frac{\partial v}{\partial t}(x,z,s,s)&=& \left(\frac{\delta m}{m} \frac{\partial^2 u}{\partial t^2}\right)(x,z,s) 
\ee

Set 
\[
w(x,z,t,s) = \frac{1}{2}\left( \frac{\partial u}{\partial t}(x,z,t) - \frac{\partial u}{\partial t}(x,z,2s-t)\right)
\]

Then in complement of $\mbox{supp }\oS$, $w$ solves (\ref{vawe}), provided $\delta m/m \equiv 1$.

\noindent {\bf Assumption 2:} $u$ is supported near $t \mapsto (X(t),Z(t))$, and there exists exactly one $t_r$ so that $(X(t_r),Z(t_r)) \in \Sigma$.

\noindent {\bf Assumption 3:} for each $t$, the support of $u$ is small enough that $\delta m$ may be approximated by a constant on $\mbox{supp }u(\cdot,\cdot,t)$.

Claim: 
\[
v(x,z,t,s) \approx  \frac{\delta m}{m}(X(s),Z(s)) w(x,z,t,s)
\]
so for $t near t_r$,
\[
\delta u(x,z,t) = \frac{1}{2}\frac{\partial u}{\partial t}(x,z,t)\int_0^t ds  \frac{\delta m}{m}(X(s),Z(s)) 
\]
\be
\label{a1}
+ \frac{1}{2} \int_0^t ds  \frac{\delta m}{m}(X(s),Z(s))\frac{\partial u}{\partial t}(x,z,2s-t)
\ee
The second term integrates $u$ {\em backwards} along the trajectory $t \mapsto (X(t),Z(t))$, so if the trajectory intersects $\Sigma$ at $t=t_r$, this involves only times $<< t_r$, hence does not contribute to the trace. Upshot:
under assumptions 1 - 3,
\be
\label{a2}
\oF(m,\oS) \approx \frac{1}{2}\frac{\partial }{\partial t}\oF(m,\oS)(x,z,t)\int_0^t ds  \frac{\delta m}{m}(X(s),Z(s)) 
\ee

Application: transmission configuration, construct $\oS$ as integral over localized oscillatory data generating solutions remaining close to rays. Read off symbol of Q.

Note that this will be correct even when caustics are present, by Huyghens construction.

### 2019-10-19 repro/papers/ms/r.tex
repro docs dir added

\title{Volume/Surface Source Equivalence}
\author{William W. Symes}

\begin{abstract}
abstract goes here
\end{abstract}

\section{Transmission Problem}

Model problem: $u: \bR^{d+1} \rightarrow \bR$
\bea
\label{awe}
\frac{\partial^2 u}{\partial t^2} - m \nabla^2 u &=& f(t)
\delta(\bx-\bx_s)\\
u&=&0, t<0
\eea

$\Omega \subset \bR^d$, $\bx_s \in \partial \Omega$,
\[
\partial \Omega = \Gamma_R \cup \Gamma_D
\]
Forward map: $F(m)f = Pu$, $P$ = trace operator on receiver
surface $\Gamma_R \times [0,T]$

Volume source extension:
\[
\mbox{supp }\bar{f} \subset \bar\Omega \times [0,\infty)
\]
\bea
\label{extawe}
\frac{\partial^2 \bar{u}}{\partial t^2} - m \nabla^2 \bar{u} & = &
\bar{f}(\bx_s,\bx,t)\\
\bar{u}&=&0, t<0
\eea
Extended forward map: $\bar{F}(m)\bar{f} = P\bar{u}$, $P$ = trace operator on receiver
surface $\Gamma_R \times [0,T]$

Write $\bar{u} = u_0 + u_1,$, with $u_0 = 0$ in $(\bR^d \setminus
\bar{\Omega})\times \bR$, 
\bea
\label{awe0}
\frac{\partial^2 u_0}{\partial t^2} - m \nabla^2 u_0 &=& \bar{f}
\mbox{ in }\Omega
\times \bR,\\
u_0 & = & 0 \mbox{ on } \partial \Omega\\
u_0 & = & 0, t < 0
\eea
and 
\bea
\label{awe1}
\frac{\partial^2 u_1}{\partial t^2} - m \nabla^2 u_1 &=& 0 \mbox{ in }\Omega
\times \bR,\\
u_1 & = & \bar{u} \mbox{ on } \partial \Omega\\
u_1 & = & 0, t < 0
\eea
Note that $\bar{u}=u_1$ in $(\bR^d \setminus
\bar{\Omega})\times \bR$ . Standard uniqueness for WE $\Rightarrow \bar{u}=u_0+u_1 $ in $\bR^d \times \bR$.

SInce $u_0$ satisfies homogeneous Dirichlet condition on all of
$\partial \Omega$, $\bar{F}(m)\bar{f} = Pu_1. $

From the definition of  that in a sufficiently small neighborhood of any point of
$\partial \Omega \times \bR$, 
\bea
\label{awe2}
 \frac{\partial^2 u_1}{\partial t^2} - m \nabla^2 u_1 &=&[\partial_n
 u_1] \delta_{\partial \Omega \times \bR}
\eea
in which $[\cdot]$ is the trace of the jump on $\partial \Omega$ and
$\delta_{\partial \Omega \times \bR}$ is the delta distribution on the
space-time boundary. In view
of (\ref{awe1}) and $\bar{u}=u_1$ outside of $\Omega \times \bR$, in
fact $u_1$ solves (\ref{awe2}) globally, and $u_1=0$ for $t<0$, which
determines $u_1$ uniquely in terms of $G[m,\bar{f}] = [\partial_n
u_1]|_{\partial \Omega \times \bR}$.

For $g \in H^{-1/2}(\partial \Omega \times \bR), g=0$ for $t<0$,
define the boundary trace extension $\tilde{F}(m)g = Pw$ for the
solution $w$ of 
\bea
\label{awe3}
 \frac{\partial^2 w}{\partial t^2} - m \nabla^2 w &=& g
 \delta_{\partial \Omega \times \bR},\\
w&=&0,t<0
\eea

The foregoing shows that
\[
\tilde{F}[m]G[m,\bar{f}] = \bar{F}[m]\bar{f}.
\]
That is, the boundary trace extension is equivalent to the volume
source extension, in the sense that the two have the same
range. Since $\bx_s \in \partial \Omega$, a suitable annihilator for
the boundary trace extended IP is still $A =$ multiply by $|\bx-\bx_s|$.

### 2019-10-19 repro/papers/pw/paper.tex
repro docs dir added

\title{Planewave Modeling and Migration with IWAVE}
\author{William. W. Symes}

\address{The Rice Inversion Project,
Rice University,
Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}

\lefthead{Symes}

\righthead{Planewaves with IWAVE}

\maketitle
\parskip 12pt

\begin{abstract}
Since IWAVE accepts any forcing term, or source, modeling plane waves simply involves creating a plane wave source. This paper describes the {\tt planewave} utility: it creates both plane wave source files and prototype output (header) files for plane wave data. The examples show both how to perform ordinary plane wave modeling and migration, and also how to create an extended plane wave image volume for use in velocity analysis and AVO.
\end{abstract}

\inputdir{project}

\section{Introduction}
This paper describes the IWAVE implementation of plane wave modeling and migration. The examples use constant density acoustic modeling, but the same principles apply to plane wave modeling based on any wave physics implemented in IWAVE.

IWAVE is a framework for solving time-dependent partial differential equations by Finite Element or Finite Difference methods. The current implementation focuses exclusively on uniform grid FD methods. A simple example of the target problem class is constant density acoustics, governed by the acoustic wave equation
\mybe
\label{awe}
\left(\frac{\partial^2 u}{\partial t^2} - c^{2}\nabla u\right)(s;t,\bx) = f(s;t\bx).
\myee
together with appropriate initial and boundary conditions, defining a family of fields depending on a source parameter $s$. Plane wave modeling results from the choice
\mybea
\label{pw3}
s & = & (p_x,p_y),\\
f(s;t,z,x,y) & = &\delta(z-z_s)w(t-p_x x - p_y y)
\myeea
or, for 2D,
\mybea
\label{pw2}
s & = & (p_x),\\
f(s;t,z,x) & = & \delta(z-z_s)w(t-p_x x)
\myeea
The 2D plane wave source (\ref{pw2}), for example, produces a planar wave at a source point $(z_s,x)$ propagating at an angle of $\theta = \arcsin c(z_s,x)p_x$ to the vertical. Plane waves propagate only in the region in which $|cp_x|<1$. Thus a plane wave will transit the entire region only if $|c_{\rm max}p_x| <1$. This criterion places an intrinsic limit on the image region for migration.

The basic acoustic constant density driver {\tt acd} computes approximate solutions to (\ref{awe}) for arbitrary right-hand side families $f(s;t,\bx)$, so plane wave simulation requires another utility to produce appropriate right hand side families. As the notation in (\ref{awe}) suggests, the field also depends on the parameter $s$, so storage of the simulation output requires a data structure including parametrization by $s$. The utility {\tt planewave} does both tasks, producing a compatible pair of SEGY files for input to {\tt acd} (and, with appropriate modification, to other IWAVE modeling tools).

Any solution of the acoustic or elastic wave equations in a homogeneous material model may be decomposed into propagating plane waves: this observation is fundamental to a basic understanding of these equations \cite[]{CourHil:62,Ach:73}. Plane wave modeling has a long history in seismology: classic references include \cite{AkiRich:80}, \cite{DieSt:81}, \cite{sto81}, \cite{TreiGutWag:82}, and \cite{CarriKuoStof:84}. Plane waves are a persistent waveform for wave propagation in stratified or layered media, and accordingly there is a large literature on analysis and inversion for layered models based on this observation. More recently, Dong Sun used plane wave modeling in his pathbreaking investigation of nonlinear image domain inversion \cite[]{Sun:09a,SunSymes:SEG12}. For imaging (or inversion), plane wave sources have the advantage over point sources of illuminating a substantial part of the model, rather than a relatively narrow beam: as the examples to be presented in this paper will show, plane wave images consist largely of image, with relatively small volume occupied by edge diffraction, unlike point source images. The chief disadvantage of plane wave sources is the necessary extension of the simulation time axis: some parts of the plane wave are activated earlier than others (see (\ref{pw2})!), whereas time-stepping methods must initiate when any of their dynamic fields become nonzero. Possibilities for ameliorating this added simulation expense exist, for instance broken and time-shifted plane waves or spatial sinusoid modulation \cite[]{Soubaras:07}.

The {\tt SConstruct} file in the {\tt project} subdirectory contains
complete annotated scripts for producing the examples shown here. The
reader should consult this script, along with the self-docs for {\tt
acd} and {\tt planewave} and the basic IWAVE white
paper \cite[]{trip14:struct} to fill in the usage details not
mentioned here. The examples build in a few minutes on any modern
workstation or laptop via {\tt scons} in {\tt project} - the reader
will need to do this, for example, to view the movie files {\tt
movie1p.rsf} and {\tt movie1pm01.rsf}. The examples require that the
Madagascar revision number at least 13459.

\section{Planewave}
Since IWAVE permits source input at any point in the spatial modeling domain, there is no need to restrict the source points composing a plane wave emitter to simulation grid points. The {\tt planewave} utility assumes uniformly spaced source and receiver points, arranged in horizontal arrays. This is OK for synthetic experiments. On the TO-DO list for {\tt planewave'} is addition of a facility to take arbitrary source and receiver locations from SEGY data files.

The salient characteristic of each plane wave is slope, or slowness (the vectors $(p_x)$ and $(p_x,p_y)$ in (\ref{pw2}), (\ref{pw3})). Since planewave sources are synthetic in any case, there seems little harm in limiting the sampling in slowness to uniform in some sense. Current implementation samples uniformly in slowness. Analysis of layered medium kinmatics suggests that uniform sampling of slowness-squared may be more useful - on the TO-DO list.

Due to a bug in the SU utility {\tt suplane}, to which {\tt planewave} delegates plane wave construction, all  plane waves must pivot around the midpoint of the array - that is, the time=0 point in each plane wave will occur at the source array midpoint.

Upshot: {\tt planewave} requires parameters to determine several axes:

\begin{itemize}
\item receiver time axis - just {\tt nt} and {\tt ot}, as {\tt dt} is taken from the source pulse header
\item receiver horizontal axis ({\tt nx}, {\tt ox}, {\tt dx})
\item receiver depth ({\tt zr})
\item source horizontal axis ({\tt nxs}, {\tt oxs}, {\tt dxs})
\item source depth ({\tt zs})
\item slowness axis ({\tt np}, {\tt op}, {\tt dp})
\end{itemize}

Because the {\tt Flow} does not automatically inherit the ambient environment, it is also necessary to pass the path to the SU root directory, {\tt CWPROOT}, via a parameter of the same name.

The source time axis is computed - the user does not specify it. The computation uses the extreme slownesses and the extreme $x$ coordinates of the source array, to compute an interval containing the support of every trace in the entire impulse (spike) plane wave for every slowness. Then {\tt suconv} adds this to the time axis of the pulse to create a time axis containing the support of every trace in every plane wave. As described in \cite[]{trip14:struct}, IWAVE creates a simulation time axis containing the convex hull of the source time axis and the output data time axis.

\section{Examples}
This section presents several examples of plane wave modeling and migration. I'll describe the key parameter selections; the reader can consult the annotated {\tt SConstruct} file in the project directory for precise details.
I use the acoustic constant density driver {\tt acd} in all of the examples below - for usage direction, see the self-doc, or for more extensive discussion \cite[]{trip14:struct}.

The examples in this section use a subsampled version of the Marmousi model, Figure \ref{fig:csq24}, with the water depth increased to 400 m and a horizontal extension on the left side of the model. The extended geometry has 444 points in the x direction, and extends from $x=0$ m (the left side) to $x=10632$ m (the right side). The depth range is $[0,3000]$ m, so to accommodate the deeper water layer some of the original model is dropped off the bottom. The spatial sample rate is 24 m in each direction. Horizontal locations are measured from the left edge, from 0 to 10632 m.

\plot{csq24}{width=\textwidth}{Marmousi model, sampled at 24 m in z and x, layered extension on right to 10632 m.}

The choice of modeling algorithm, the venerable (2,4) scheme, requires at least 5 gridpoints per wavelength for reasonable accuracy over a few seconds' propagation time, so the maximum frequency adequately modeled in the water layer is 12.5 Hz. The source wavelet for these simulations is a $[1,3.5,10,12.5]$ Hz trapezoidal bandpass filter (Figure \ref{fig:wavelet}). Free surface boundary conditions are imposed on all boundaries. The recording time interval is 2 s, short enough that few boundary reflections will be observed, and those only for the larger slownesses.

\plot{wavelet}{width=\textwidth}{Source pulse: [1.0, 3.5, 10, 12.5] Hz trapezoidal bandpass filter.}

Source points are placed at all 444 horizontal grid locations (multiples of $dx=24$ m) at a depth of 12 m. [IWAVE evaluates fields at non-grid points by multilinear interpolation, and inserts sources by multilinear adjoint interpolation.]

The receiver array occupies 241 contiguous horizontal locations spaced at 24 m and beginning at 3000 m, at a depth of 12 m.

\subsection{Single plane wave at normal incidence}
The parameter choice for this example is {\tt op=0, np=1, dp=} whatever.

A movie of the wavefield shows the plane wave starting near the surface (the ghost occurs at a short enough time lag that the effect is visible only as a change of wavelet shape). The sea floor creates a sizeable reflection, of which the first seafloor multiple is within the time range of the simulation - it is however of such low amplitude that it is very difficult to distinguish from other, primary reflections. Figure \ref{fig:movie1p} 
shows the final frame. View the entire movie as follows: 
\begin{verbatim}
scons movie.rsf
sfgrey clip=2 < movie.rsf|xtpen
\end{verbatim}
\plot{movie1p}{width=\textwidth}{Pressure field simulation at t = 2 s, normal incidence plane wave source.}

Trace data for this example appears as Figure \ref{fig:shot1p}.

\plot{shot1p}{width=\textwidth}{Plane wave gather, normal incidence.}

Born (linearized) simulation requires definition of a background model and a reflectivity (perturbational model). Figure \ref{fig:csq24big} shows a smoothed version of the Marmousi model, which serves as a macromodel for this example. The difference of the base model and a smoothing on a shorter length scale gives a suitable, scale-separated reflectivity model (Figure \ref{fig:dcsq24}).

\plot{csq24big}{width=\textwidth}{Smoothed Marmousi c-squared model: moving average on 240 m window, iterated 10 times.}
\plot{dcsq24}{width=\textwidth}{Reflectivity: difference of base model (Figure \ref{fig:csq24}) and moving average on 120 m window, iterated twice. Plot window chosen to match image zone for migration.}

Trace data for this example appears as Figure \ref{fig:born1p}.

\plot{born1p}{width=\textwidth}{Born plane wave gather, normal incidence.}

Migration of Born data amounts to application of the adjoint linearized modeling operator, a kinematic inverse in the high frequency limit provided that the background model is transparent, as it is in this well-scale-separated example. Compare the migrated image, Figure \ref{fig:mig1p}, to the reflectivity model, Figure \ref{fig:dcsq24}.

\plot{mig1p}{width=\textwidth}{Migration of normal incidence Born data using correct background model.}

The FWI gradient is equivalent to a (reverse time) migration of the simulation residual data, that is, the difference of the modeled and observed data. To illustrate this construction, take for the ``observed'' data the Marmousi normal incidence gather (Figure \ref{fig:shot1p}), and for the current FWI iterate a homogeneous model with $c = 1.5$ m/ms at all locations. The difference is plotted in Figure \ref{fig:diff}, and the migration = FWI gradient in Figure \ref{fig:migshot1p}.

\plot{diff}{width=\textwidth}{Difference of modeled data (Figure \ref{fig:shot1p}) and data from a homogeneous model, which contains only the incident wave. Like Born data (Figure \ref{fig:born1p}), the incident wave is missing (because the difference cancels it); unlike Born data, the difference contains all nonlinear effects (eg. multiple reflections) present in the modeled data.}

\plot{migshot1p}{width=\textwidth}{Migration of the difference data of Figure \ref{fig:diff}, in homogenous background model (c=1.5 m/ms). Note kinematic distortion compared to migration of Born data in consistent background (Figure \ref{fig:mig1p}).}

\subsection{Multiple plane waves, oblique incidence}

As explained  in \cite[]{trip14:struct}, IWAVE implicitly loops over axes labeled {\tt dim}+1 and greater, in which {\tt dim} is the dimension of the simulation domain ( = 2, in the examples presented here). By rigid convention, axes 0,...{\tt dim}-1 are the spatial grid axes, {\tt dim} is the time axis, and all axes with labes {\tt dim} + 1 and greater represent additional axes over which the simulation should loop.

For SEGY, the shot axis is implicitly axis {\tt dim} +1. A change in either {\tt sx} or {\tt sy} increments this axis. Therefore to simulate a line of plane wave ``shots'', the user merely need construct trace headers containing the requred number of traces for each slowness, and encode the slowness in {\tt sx}. Similarly, the plane wave source should contain one gather for each slowness, with precisely the same slownesses and number of slownesses as the occur in the trace headers.

The {\tt planewave} utility does this job for the user, creating a matching set of plane wave headers and source gathers suitable for IWAVE input. The example plane wave source displayed in Figure \ref{fig:wav11p} shows the plane wave source gathers for 11 evenly spaced slownesses between $p=-0.1$ and $0.1$, corresponding to propagation angles of approximately $-6$ to $6$ degrees. Figure \ref{fig:movie1pm01} shows the computed pressure field at 0.5 s for p=-0.1. It is plotted at approximately 1:1 aspect ratio; clearly the propagation angle in the water later is indeed about $6^{\circ}$.  Figure \ref{fig:shot11p} shows the 11 plane wave gathers created by the source array depicted in Figure \ref{fig:wav11p}, with the same receiver array as in Figure \ref{fig:shot1p}.

\plot{wav11p}{width=\textwidth}{Multiple slowness source gathers: $p\in [-0.1,0.1]$, $\Delta p = 0.02$ ms/m.}

\plot{movie1pm01}{width=\textwidth}{Pressure field response to plane wave source at $p=-0.1$ ms/m, $t = 500$ ms.}

\plot{shot11p}{width=\textwidth}{Simulation for 11 plane wave sources in Figure \ref{fig:wav11p}.}

Born modeling and migration work the same way. Figure \ref{fig:born1p} shows the result of linearized modeling with the background of Figure \ref{fig:csq24big} and reflectivitly of Figure \ref{fig:dcsq24}, with the source gathers of Figure \ref{fig:wav11p}. Figure \ref{fig:mig11p} show the result of migrating these 11 gathers. This image is a bit crisper than the normal incidence migration of Figure \ref{fig:mig1p}, with migration swings suppressed somewhat by stacking and some of the more steeply dipping features towards the bottom better imaged.

\plot{born11p}{width=\textwidth}{Linearized simulation for 11 plane wave sources, same model parameters as Figure \ref{fig:born1p}.}

\plot{mig11p}{width=\textwidth}{Migration of data in Figure \ref{fig:born11p}.}

\subsection{Shot-record extended modeling}

Stacking of individual shot images is responsible for artifact suppression in line images. To understand how this happens, exposure of the shot images is useful. The full volume of shot images is also underlies one approach to migration velocity analysis. IWAVE can output this full volume.

In fact, IWAVE interprets the shot record volume as the output of the adjoint linearized map of an {\em extended} model, in which each shot uses an independent copy of the coefficient fields ($c^2$, for the constant density acoustics system). The extended modeling concept underlies much recent work on ``image domain'' inversion. For an extensive discussion of this idea, see \cite[]{geoprosp:2008}.

IWAVE requires the definition of an extended model space, with one ``panel'' or {\tt dim}-dimensional model for each shot (in this form of extended modeling - there are others!). Thus as a first step, one must create data files exhibiting the extended structure. The {\tt SConstruct} file accomplishes this task via use of {\tt sfspray} - see the rule for creating {\tt csq24big\_ext}. Note that an important part of this construction is the addition of nonstandard keywords {\tt dim}, {\tt gdim}, and {\tt idx}, {\tt x} = 0, 1, and 3. The functioning of these keywords is explained in  \cite[]{trip14:struct}. In brief, {\tt x} = 0, 1 signify the two spatial dimensions, and 3 = {\tt dim} + 1 tells IWAVE to treat the third axis in {\tt csq24big\_ext} as the same as the shot axis implied by the structure of the SU files in the simulation (see discussion above).

Use these extended files as input under the keywords for the background model fields (active, input fields - see the discussion of the {\tt fields} struct in \cite[]{trip14:struct}). For constant density acoustic IWAVE, there is only one such field, with keyword {\tt csq}. All related fields, eg. the model perturbation inputs for the first derivative (keyword {\tt csq\_d1} in constant density acoustic IWAVE) or the adjoint first derivative (RTM) outputs (keyword {\tt csq\_b1}), must have the same structure including the same additional keywords. The {\tt SConstruct} uses {\tt sfadd} to copy the c-squared file onto the migration output file to initialize it, including all of its header info.

Models such as the shot-record extension are {\em extended} because the modeling operator or forward map is an extension of the ordinary one. That means that the data output by ordinary modeling is the same as the data output by a ``physical'' extended model, one in the range of the extension map. In the case of the shot record extension, the extension map simply duplicates (or ``sprays'', hence the use of {\tt sfspray}) an ordinary model field as many times as required, and identifies the axis introduced in this way. The adjoint of this spray operation is the stack, which explains the relation between extended migration (adjoint linearized modeling) and ordinary migration: output of the latter is the stack of the output of the former.

An example appears as Figure \ref{fig:mig11pext}. Panel (shot record migration) 5, or p=-0.02, is displayed in the large front section; this image is comparable to Figure \ref{fig:mig1p}. The right panel is an example of a {\em common image gather}: the horizontal axis is shot index, and the flatness (or lack of same) indicates the kinematic correctness of the background model - that is, the 11 trace. Since this example is an inverse crime, of course the image gather is as flat as possible. Note that it is not perfectly flat or uniform in amplitude, as it would be if the image volume were replaced by the physical extended model (result of spray) which would have generated the data - migration is only the adjoint of linearized modeling, not its inverse.

\plot{mig11pext}{width=\textwidth}{Shot-record extended migration of data in Figure \ref{fig:born11p} with correct background model, i.e. that used to generate the data.}

Replacing the correct c-squared model (Figure \ref{fig:csq24big}) by a convex combination with 20\% homogeneous water c-squared produces a comparable image volume, shown in Figure \ref{fig:mig11p80pctext}. Comparison with Figure \ref{fig:mig11pext} reveals two obvious differences: the image (front, large rectangle) is misplaced (too shallow), and the image gather is not quite flat. Correct location of the image is not known {\em a priori}, but it is known that the images for the various values of $p$ should all be the same, at least in the location of events. So that latter provides a means of judging the correctness of the background model. This flatness criterion, and related criteria for other extended models. are the basis of velocity analysis, or as it has come to be known recently, image domain inversion.

\plot{mig11p80pctext}{width=\textwidth}{Shot-record extended migration of data in Figure \ref{fig:born11p} with 80\% correct background model, 20\% water c-squared. Note non-zero slope of bottom events in image gather (right-hand panel).}

\section{Conclusion}

IWAVE is built to solve equation \ref{awe} and similar systems, so the particular choice of right-hand side implicit in plane wave modeling must be the result of an external (to an IWAVE modeling driver) utility. The {\tt planewave} utility generates a matching pair of SEGY trace files: (i) a trace header file, serves as a prototype for the output data of a simulation or the input data of a migration, and (ii) ah source gather file, also SEGY traces, encoding the plane wave source (or sources, for a multiple plane wave simulation). Coupled with IWAVE driver code (such as {\tt acd}), the {\tt planewave} utility creates a basic tool for plane wave simulation and migration.

\section{Acknowledgements}
The IWAVE
project owes a great deal to several open source seismic software
packages (Seismic Un*x, SEPlib, Madagascar), debts which we gratefully
acknowledge. This work was supported in part by the sponsors of The Rice Inversion Project.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/pwinv/paper.tex
repro docs dir added

\title{Planewave Inversion with IWAVE and RVL}
\author{William. W. Symes}

\address{The Rice Inversion Project,
Rice University,
Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}

\maketitle
\parskip 12pt

\begin{abstract}
IWAVE implements forward modeling operators, and their derivatives and adjoint derivatives up to second order. RVL supplies optimization algorithms and a vector calculus interface, including windowing, muting, scaling, and other operators useful in constructing inversion algorithms. This paper shows how to use these tools to invert plane wave data for the long spatial wavelengths of velocity models.
\end{abstract}

\inputdir{project}

\section{Normal Incidence Plane Wave}
Model = Marmousi, 6x subsampled to 24 m grid, supports roughly 12 Hz max with acceptable dispersion over 2 s.

Modifications: extended to 12000 m by layered addition on right. 384 m of ``water'' added on top, then model is extended by 33 m of ``water'' in negative depth. See (\ref{fig:deepcsq24}).

FD method = (2,8) scheme, cmin=1.0 m/ms, cmax=6.0 m/ms, cfl=0.5.

Trust region method implemented in RVL:

{\tt rvl/umin/include/trgnalg.hh}.

 Includes standard trust region adjustment rules plus ``internal doubling'' logic to extend step in case of high ratio of actual to predicted decrease, linear solver passed by policy. Use conjugate gradient method for normal equation in this set:

{\tt rvl/umin/include/cgne.hh}

Attempt to fit plane wave data by modifying velocity model in window $[384,3384] \times [0, 12000]$ m. Initial velocity is 1.5 m/ms throughout $[-3384,3384] \times [0,12000]$ m. Velocity (squared) represented as initial model plus update in window, injected into global grid.

Implementation: composed IWAVE modeling operator, instance of {\tt IWaveOp}, defined in

{\tt iwave/core/include/iwop.hh}

with affine windowing operator, instance of {\tt GridWindowOp}, defined in

{\tt iwave/grid/include/gridops.hh}

via RVL operator composition {\tt OpComp}, defined in

{\tt rvl/rvl/include/op.hh}.

\plot{deepcsq24}{width=0.8\textwidth}{Modified Marmousi Model, windowed.}

\plot{deepshot1ph}{width=0.8\textwidth}{Response of Marmousi model to vertical incidence (p=0.0) Heaviside plane wave. Heaviside pulse filtered by 10 Hz (half power) Gaussian wavelet. Sources and receivers at z=0. Sources extend over entire surface z=0, at every grid point. Receivers from x=3000 to x=9000, at every grid point. Max = 8.0e+03 GPa.}

\plot{deepshot1pg}{width=0.8\textwidth}{Response of Marmousi model to vertical incidence (p=0.0) zero phase Gaussian 10 Hz plane wave. Same source-receiver geometry as Figure \ref{fig:deepshot1ph}.}

\plot{deepshot1p}{width=0.8\textwidth}{Response of Marmousi model to vertical incidence (p=0.0) zero phase trapezoidal [1, 2.5, 10, 12.5] Hz bandpass plane wave. Same source-receiver geometry as Figure \ref{fig:deepshot1ph}.}

\plot{heavifinal}{width=0.8\textwidth}{Three iteration of Gauss-Newton Trust Region algorithm, with initial model = constant 1.5 km/s, data = Figure \ref{fig:deepshot1ph}. (Display is velocity-squared.) Three steps each with 10 inner CGNE iterations. Objective reduction by 0.005, gradient norm by 0.014.}

\plot{gauss10Hzfinal}{width=0.8\textwidth}{Three iteration of Gauss-Newton Trust Region algorithm, with initial model = Figure \ref{fig:heavifinal}, data = Figure \ref{fig:deepshot1pg}. (Display is velocity-squared.) Three steps each with 10 inner CGNE iterations. Objective reduction by 0.005, gradient norm by 0.014.}

\plot{deepbandfinal}{width=0.8\textwidth}{Three iteration of Gauss-Newton Trust Region algorithm, with initial model = Figure \ref{fig:gauss10Hzfinal}, data = Figure \ref{fig:deepshot1p}. (Display is velocity-squared.) Direct wave muted in data and simulation. Three steps each with 10 inner CGNE iterations. Objective reduction by 0.05, gradient by 0.14.}

\plot{deepbandres}{width=0.8\textwidth}{Final residual for model displayed in Figure \ref{fig:deepbandfinal}. Most eerror in last 0.2 s; error reduction (re Figure \ref{fig:deepshot1p}) in time interval 0-1.6 s is 0.19.}

### 2019-10-19 repro/papers/sseabs19/paper.tex
repro docs dir added

%% This file is automatically generated. Do not edit!
%\documentclass[11pt]{segabs}

%\usepackage{amsmath}
%\usepackage{setspace}
%\usepackage{wasysym}
%\usepackage{graphicx}

%\newcommand{\mb}{\mathbf}
%\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
%\newcommand{\bv}{\mathbf{v}}
%\newcommand{\bx}{\mathbf{x}}
%\newtheorem{exmp}{Example}

%\setfigdir{Fig}

%\begin{document}
\title{Waveform Inversion via Source Extension}
\author{Guanghui Huang, Michigan State University, Rami Nammour, Total $E\&P$ $R\&T$ USA, William W. Symes, Rice University, Mohamed Dolliazal, Total $E\&P$ $R\&T$ USA}

%\extrafloats{100}

\lefthead{Huang, Nammour, and Symes}

\righthead{Source Extension}

\maketitle
\begin{abstract}
Extended modeling is one of a number of modifications suggested to enhance the reliability of iterative Full Waveform Inversion (FWI), by making it less prone to stagnation away from useful model estimates (``cycle skipping''). All extended modeling methods add parameters to wave modeling beyond those suggested by basic wave physics, in some cases entailing substantial computational expense. Source extension methods add parameters to the description of the physical energy source; many of these cost little more computationally than standard FWI. The first purpose of this paper is to present a (necessarily incomplete) taxonomy of proposed source extensions and related inversion methods. The second purpose is to introduce a particular such method, the Surface Source Extension, and give a couple of examples of its use. We will also explain the theoretical justification for some of these methods: for pure transmission (including diving wave) data, a close link to traveltime tomography explains the absence of cycle skipping observed in numerical experiments. No such theoretical link is known to exist in the case of pure reflection data, but some positive numerical results suggest that some source extension methods may ameliorate cycle-skipping in that case as well .
\end{abstract}

\section{Overview}
FWI can be described in terms of a wave operator $L[c]$ depending on an array of space-dependent coefficient fields $c$, a trace sampling operator $P$, and a dynamic wavefield $u$. The components of coefficient fields $c$ describe material parameters (density, elasticity tensor components, attenuation coefficients,...) and depend on spatial position $\bx$. The dynamical field $u$ may be vector-valued and depends on $\bx$ and on time $t$. The trace sampling operator converts $u$ to a a vector of trace gathers indexed by source position $\bx_s$ representing sensor outputs, or ``data'' for short. The basic FWI problem is: given data $d$,  find coefficients $c$ and source space-time fields $f$ (also indexed by $\bx_s$) so that
\begin{equation}
\label{eqn:fwi}
Pu \approx d \mbox{ and } L[c]u = f.
\end{equation}
Sometimes $f$ is also regarded as given, sometimes only some aspects are given (for example, localization at a source point) and others are to be determined as part of the solution. A basic optimization formulation of FWI is the nonlinear least squares problem:
\[
\mbox{choose } c, f \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2
\]
often with some sort of regularization thrown in.

As is well-known, local optimization methods are the only feasible approach given the dimensions of a typical instance of \ref{eqn:fwi}, and those have a tendency to stall at physically uninformative models due to ``cycle-skipping''.

Many modifications of FWI have been proposed to avoid cycle-skipping. One category of such modifications are the extension methods, in which the FWI problem is ``relaxed'' by adding degrees of freedom to the model. Proposed extension methods fall roughly into two categories: medium extensions, in which parameters are added to the coefficients in the wave equation, and source extension, in which parameters are added to the energy source representation. We shall focus on source extension methods here; see \cite{geoprosp:2008} for an overview of medium extensions, also \cite{Herve2017} for more recent developments. Source extensions effectively impose the wave equation as a soft, as opposed to hard, constraint. For sake of argument (and because almost all work so far has assumed it), adopt the isotropic point source as the physical source model: for the data gather at source position $\bx_s$, the source takes the form
\begin{equation}
\label{eqn:ptsrc}
f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s).
\end{equation}
Every method explained below can be altered to accommodate more complex physical source models.

Source extension replaces $f(\bx,t;\bx_s)$ with an artificial source $g(\bx,t; \bx_s)$ with more degrees of freedom than merely a single common source wavelet $w(t)$ as in definition \ref{eqn:ptsrc}. Assuming that \ref{eqn:ptsrc} is actually correct physics, these extra degrees of freedom have to be suppressed in the optimal solution, via an {\em annihilator} $A$, an operator acting on extended sources $g$, whose null space is precisely the physical sources (of the form \ref{eqn:ptsrc}). Extended source inversion replaces the FWI objective with a penalty function
\begin{equation}
\label{eqn:esi}
\mbox{choose } c, g \mbox{ to minimize } \|PL[c]^{-1}g -d \|^2 + \alpha^2 \|A[g]\|^2
\end{equation}

The way in which degrees of freedom are introduced, and the choice of annihilator, distinguish the various source extension methods proposed in the literature. In the next few paragraphs we describe some examples.

\subsection{Wavefield Reconstruction Inversion}
\cite[]{LeeuwenHerrmannWRI:13,LeeuwenHerrmann:16,WangYingst:SEG16,Aghamiry:19} Extended source space consists of arbitrary functions of space-time and source position $g(\bx,t;\bx_s)$. Annihilator is difference with actual (known) source: $A[g]=g-f$.
With these choices, WRI takes the form introduced by \cite{WangYingst:SEG16}. The original approach is equivalent: \cite{LeeuwenHerrmann:16} use the dynamic wavefield $u$ as the additional degrees of freedom, and minimize
\[
\|Pu-d\|^2 + \alpha^2 \|L[c]u-f\|^2
\]
Since the wave equation (any reasonable version!) has a unique solution for each right-hand side, replacing $L[c]u$ with $g$ (and $u$ with $L[c]^{-1}g$) turns this original form of WRI into the form \ref{eqn:esi}.
WRI requires storage of, and access to, the full space-time (or space-frequency) field $g$ (or $u$) at each iteration, potentially a 6D field.

\subsection{Space-Time Extension}
\cite[]{HuangSymes:SEG16b,HuangSymes:Geo18a} Differs from WRI by including $f$ in the unknowns, and in choice of annihilator:
\begin{equation}
\label{eqn:msste}
A[g](\bx,t;\bx_s) = |\bx-\bx_s|g(\bx,t;\bx_s).
\end{equation}
This annihilator merely enforces the position constraint on the source, without relying on {\em a priori} knowledge of the source wavelet. In fact, the source wavelet is a by-product of the minimization \ref{eqn:esi}. Also requires storage of the full (potentially 6D) field.

\subsection{Volume Extension}
\cite[]{HuangSymes:SEG16a,HuangSymes:Geo18b} Assumes that source signature deconvolution has been performed, so that the actual wavelet is (a bandlimited version of) $\delta(t)$. The extended source is an {\em exploding reflector}:
\begin{equation}
\label{eqn:msvol}
g(\bx,t;\bx_s) = h(\bx,\bx_s)\delta(t).
\end{equation}
Uses the same annihilator as the Space-Time Extension, but loses the advantage of solving for the source as by-product - the source must be known, so that decon can be carried out prior to inversion.
However, only requires a (3D) spatial storage volume per source position, essentially equivalent to a second (5D) data volume.

%A disadvantage of this approach, for inversion of reflections, is that unlike post-stack migration the actual physical wave velocity is used in modeling, so that the the region in which $h$ must be defined may considerably larger than the ordinary modeling domain - reflection two-way time is effectively modeled by one-way time.

%The volume extension concept is similar to an idea discussed by \cite{}.

\subsection{Surface Extension}
So far as we know, this extension is described for the first time in this paper (however \cite{ZhangGao:08} describe a related idea). Similar to the volume extension, but the source is spread over the source space-time hypersurface $z=z_s$, rather than over space at $t=0$. In the simplest case, where the sources all lie on $z=z_s$, the extended source space consists of functions (distributions) of the form ($\bx=(x,y,z)$)
\begin{equation}
\label{eqn:mssur}
g(\bx,t;\bx_s) = h(x,y,t;x_s,y_s)\delta(z-z_s).
\end{equation}
As for the previous two cases, the annihilator is given by the localization penalty (equation \ref{eqn:msste}). Therefore the source wavelet is a by-product of the inversion, and does not need to be specified {\em a priori}. Like the volume extension, the additional data volume required is on the same order of size as the data. Unlike the volume extension, it does not require a larger-than-physical simulation volume.

\subsection{Source-Receiver Extension}
\cite{HuangSymes:SEG15a} describe the {\em source-receiver extension}: each trace regarded as separate gather, with separate point source waveform. In effect, the source is permitted to depend on receiver position as well:
\begin{equation}
\label{eqn:awi-alt}
g(\bx,t;\bx_r,\bx_s) = w(t;\bx_r,\bx_s)\delta(\bx-\bx_s).
\end{equation}
Since the corresponding dynamical field $u$ (also dependent on $\bx_r$) satisfies
\begin{equation}
\label{eqn:awi-green}
u(\bx,t;\bx_r,\bx_s)=w(t;\bx_r,\bx_s) *_{t} G(t;\bx,\bx_s)
\end{equation}
it is not necessary to construct $u$ by solving a $\bx_r,\bx_s$-dependent wave equation, only to construct the $\bx_s$-dependent Green's function $G(\cdot; \cdot,\bx_s)$. Thus the modeling cost per iteration is the same as that of FWI.

This extension seems to have appeared first in unpublished work from the third author's group \cite[]{Song:94c,SoSy:92,SongSymes:94a,SongSymes:94b,Symes:94c}. Various annihilators have been suggested, see also \cite[]{Plessix:00a,Plessix:00, PrattSymes:02,LuoSava:11, Warner:14,Warner:16}. Since the source-receiver dependent wavelet $w(t;\bx_r,\bx_s)$ parametrizes the source (equation \ref{eqn:awi-alt}), the annihilator may be viewed as acting on $w$. One popular choice is including multiplication by $t$, which annihilates $w=\delta(t)$:
\begin{equation}
\label{eqn:awi-ann}
A[w](\bx,t;\bx_r,\bx_s) = t w(t;\bx_r,\bx_s)
\end{equation}
This makes sense if the actual source-receiver independent point source wavelet is known and deconvolved from the data traces, to produce an approximate delta source at zero time lag when the coefficients $c$ include kinematically correct velocities.

\subsection{Adaptive Waveform Inversion}
\cite{Warner:14,Warner:16} presented an extended FWI based on source-receiver extension (``Adaptive Waveform Inversion'', AWI) using the scale-by-$t$ annihilator (equation \ref{eqn:awi-ann}). Rather than use the penalty method \ref{eqn:esi}, Warner and Guatsch effectively force the data fit term to zero, by solving the deconvolution problem  $G(t;\bx_r,\bx_s)*_tw(t;\bx_r,\bx_s) = d(t;\bx_r,\bx_s)$, thus making the estimated source wavelet $w(t;\bx_r,\bx_s)$ a function of the coefficients $c$, and using the annihilator term as the objective (several other works cited above have also made use of this $\alpha \rightarrow 0$ limiting case). Further, Warner and Guatsch normalize the objective with the estimated wavelet $L^2$ norm: choose $c$ to minimize
\begin{equation}
\label{eqn:awi-obj}
\sum_{\bx_r,\bx_s}\frac{\int dt\,|t|^2 |w(t;\bx_r,\bx_s)|^2}{\int dt\,|w(t;\bx_r,\bx_s)|^2}
\end{equation}
Note the order of summation and normalization: each trace is normalized independently. This normalization avoids the familiar source/reflector amplitude ambiguity, and has other constructive effects as well.

\section{Theory}
For several of the source extension algorithms, applied to transmission data, the objective is known to be closely linked to traveltime tomography in one form or another, and to inherit its convergence properties. This connection seems first to have been established for the source-receiver extension by \cite{Song:94c} in the case of a so-called differential semblance annihilator, and in \cite[]{HuangSymes:SEG15a,HuangSymes:Geo17} for the time-delay annihilator \ref{eqn:awi-ann}. The latter references study source-receiver extension of constant density acoustic modeling, so that the coefficients $c$ consist of the compressional wave velocity only. As in previous paragraph, make $w$ depend on $c$ by solving the deconvolution problem $G*_tw = d$, then update $c$ to minimize the objective
\begin{equation}
\label{eqn:sr-obj}
 J_0[c,d]= \sum_{\bx_r,\bx_s}\int dt\,|t|^2 |w(t;\bx_r,\bx_s)|^2
\end{equation}
(note that this objective differs from the AWI objective \ref{eqn:awi-obj} by absence of normalization). Assume that the candidate velocity fields are smooth (that is, have no embedded reflectors) and that ray paths connecting sources and receivers are unique. Then for each such velocity $c$, the corresponding Green's function has the asymptotic expansion 
\begin{equation}\label{eqn:green}
G[c](t;\bx_s,\bx_r) = a[c](\bx_s,\bx_r)\delta(t-\tau[c](\bx_r,\bx_s))+R[c](t;\bx_r,\bx_s),
\end{equation}
in which $\tau(\bx_r,\bx_s)$ is the (unique) ray travel time from $\bx_s$ to $\bx_r$ and $a$ is the corresponding geometric amplitude. The remainder $R[c]$ has a Heaviside-type singularity on the light cone $t=\tau(\bx_s,\bx_r) $ and is otherwise smooth \cite[]{Friedlander:75} Choose a target velocity field $c_*$ and create a parametrized family of noise-free data $d_{\lambda}=w_{\lambda}*G[c_*]$ defined by a family of  pulses $w_{\lambda}(t) =\lambda^{-1/2} w_1(t/\lambda) $: the parameter $\lambda$ is effectively the dominant wavelength in the pulse. \cite{HuangSymes:SEG15a} show that under these circumstances,
\begin{equation}
\label{eqn:sr-asympt}
J_0[c,d] \propto \sum_{\bx_r,\bx_s} |\tau[c](\bx_r,\bx_s)-\tau[c_*](\bx_r,\bx_s)|^2 +O(\lambda).
\end{equation}
That is, the objective $J_0$ is proportional to the standard objective function of traveltime tomography, up to an error on the order of dominant wavelength. The same is true of its gradient and Hessian, so the any minimizer of $J_0$ differs from a traveltime tomography solution by an error proportional to wavelength. This fact accounts for the apparent success of source-receiver extension FWI, reported in the cited references.

Unfortunately, the relation between source-receiver extension FWI and traveltime tomography collapses when the rays connecting source and receiver fail to be unique, as occurs generically when velocity field are sufficiently heterogeneous \cite[]{Whi:82}. \cite{Symes:94c} explained this failure and illustrated it numerically. \cite{HuangSymes:Geo17} give a different perspective on the theory, and an explicit illustration in a diving wave setting: if the diving ray field is triplicated, source-receiver extension FWI is just as prone to ``cycle skipping'' as is ordinary FWI.

The volume and surface source extensions in transmission mode are also closely related to traveltime tomography, in the specific guise of {\em slope tomography} or {\em stereotomography}, which has been developed mostly for migration velocity analysis (see \cite{Prieux:13} and references cited there). In contrast with the source-receiver extension, this relationship continues to hold in the presence of multiple ray paths connecting sources and receivers, and explains the success of surface- and volume-extended FWI. Numerical illustrations of this feature can be found in \cite[]{HuangSymes:Geo18b} and in the numerical results section below.

Insofar as the authors are aware, almost nothing is known about why either WRI or the space-time extended FWI converge, or when. Nothing appears to be known theoretically about the behaviour of any of these methods for reflection data, and numerical evidence is ambiguous (see second example below).

\section{Numerical Examples}
We provide two 2D numerical examples of surface-source extended FWI. The first is based on pure transmission data, and illustrates the capability of this extended inversion method to resolve velocities in the presence of multiple ray paths connecting sources and receivers. The second is based on the Marmousi model \cite[]{BoLaVe:91}, modified to ensure that almost all of the recorded energy consists of reflections. Its success underlines a gap in the current theory.

In both examples, modeling and inversion computations are carried out in the frequency domain with standard finite difference stencils and direct matrix methods. The penalty parameter $\alpha$ is set so that the initial data residual in the surface source extended inversion is roughly half the initial data residual for FWI, and held constant throughout the inversion. This resulted in the choices $\alpha=0.001$ for the first (lens) example, $\alpha=1.0$ for the second (Marmousi) example. We also added a regularization term, a multiple of the mean square gradient of the velocity field, weighted to be initially 10\% of the sum of the data and penalty terms, as in our work on the volume source extension \cite[]{HuangSymes:Geo18b}. We used the LBFGS algorithm \cite[]{NocedalWright} to perform all inversions.

{\exmp{(Low velocity zone)}} This example closely resembles one used by \cite{HuangSymes:SEG15a,HuangSymes:Geo18a} to demonstrate the failure of source-receiver extended inversion in the presence of multiple ray paths. The velocity model consists of a Gaussian low velocity anomaly embedded in a constant background model $v_0=2$ km/s (Figure \ref{fig:GaussVt}):
\begin{equation}
    v(x,z) = 2 - 1.4e^{-\frac{(x-1)^2}{0.25^2} -\frac{(z-1)^2}{0.5^2}} km/s.
\end{equation}
Source depth is $z_s=1.99$ km, receiver depth $z_r=0.01$ km.
A shot gather generated with a 10 Hz Ricker wavelet at $x_s=1.5$ km appears as Figure \ref{fig:GvData1500m}. Triplicated arrivals due to the low velocity anomaly are clearly evident.

\inputdir{./Fig/gv}

\begin{figure}[h]
 \centering
     \subfloat[\label{fig:GaussVt}]{\includegraphics[width=0.5\columnwidth]{Fig/gv/GaussVt}}
     \subfloat[\label{fig:GvData1500m}]{\includegraphics[width=0.245\columnwidth]{Fig/gv/GvData1500m}}
     \caption{Transmission problem: (a) Gaussian low velocity anomaly model and (b) recorded data. Receivers (blue star) and sources (green star) are placed on the top and bottom to mimic the transmission configuration.}
\end{figure}

%\multiplot{2}{,GvData1500mV0}{width=0.3\columnwidth}{Transmission problem: Shot gather of (a) recorded data and (b) simulated data by constant velocity $v_0$ for shot at $x_s=1.5$ km. Due to the low velocity anomaly, complicated triplications are present in the recorded data, and arrive later than the event by constant velocity.}

%\multiplot{3}{GaussV0pt8,GaussV0pt8,GaussVt}{width=0.3\textwidth}{Transmission problem: (a) constant velocity ($v_0$), (b) testing velocity $v = 0.8v_t+0.2v_0$ and (c) true velocty $v_t$.}

%\multiplot{3}{GvBackpro1500m,GvExtSrc1500m,GvExtData1500m,GvV0pt8Backpro1500m,GvV0pt8ExtSrc1500m,GvV0pt8ExtData1500m,GvVtBackpro1500m,GvVtExtSrc1500m,GvVtExtData1500m}{width=0.27\columnwidth}{Transmission problem: (a) backpropagation wavefield of recorded data at $x_s=1.5$ km, (b) extended source model and (c) extended data by constant velocity $v_0$; (d)-(f) are similar plots as the top row but computed by testing velocity $v=0.8v_t+0.2v_0$; (g)-(i) are similar plots as the top row but computed by true velocity $v_t$;  }

\begin{figure}[h]
 \centering
     \subfloat[\label{fig:GaussV31its}]{\includegraphics[width=0.5\columnwidth]{Fig/gv/GaussV31its}}
\subfloat[\label{fig:GvExtSrc1500m}]{\includegraphics[width=0.245\columnwidth]{Fig/gv/GvExtSrc1500m}}
     \subfloat[\label{fig:GvV31its10HzExtSrc1500m}]{\includegraphics[width=0.245\columnwidth]{Fig/gv/GvV31its10HzExtSrc1500m}}
     \caption{Transmission problem: (a) inverted velocity by surface source extended FWI; (b) extended source estimate using initial constant velocity; (c) extended source estimate using the final inverted velocity. Both (b) and (c) for shot at $x_s=1.5$ km.}
\end{figure}
After 31 LBFGS iterations, the surface source extended FWI algorithm obtains the velocity estimate depicted in Figure \ref{fig:GaussV31its}.  Source energy is much more focused in time and space after inversion (Figures \ref{fig:GvExtSrc1500m}, \ref{fig:GvV31its10HzExtSrc1500m}).

{\exmp{(Marmousi)}}
We modified the Marmousi model by adding a $2.2$ km water layer and truncating the depth range at $4$ km, also adding a box-like high velocity anomaly at $z=$2.2 km, $x=$2.8 km in distance (Figure \ref{fig:Marm2kmVt}). Grid increments were $\Delta x = \Delta z = 0.02$ km. The 2.2 km deep water effectively gates out refraction (diving) waves, so that the data energy is almost purely reflection. We generated 8 s of data for a fixed receiver spread using an isotropic point source with 8 Hz Ricker wavelet, receiver coordinates $z_r=0.04$, $x_r=$ 0.05 km to 9.15 km, $\Delta x_r=$ 0.05 km, source coordinates $z_s=0.08$ km, $x_s=0.6$ km to $x_s=8.6$, $\Delta x_s$ = 0.1 km. Figures \ref{fig:Marm2kmShot1}-\ref{fig:Marm2kmShot81} show shot gathers at $x_s=0.6$ km and $x_s=8.6$ km.

The initial velocity (Figure \ref{fig:Marm2kmV0}) is linearly increasing with depth below the water layer. Data for both FWI and surface source extended FWI, method are sampled from the Fourier-transformed time domain traces with frequency band [5,8] Hz. The correct source amplitudes were input for FWI, whereas the surface source extended FWI generates these as by-product.

\inputdir{./Fig/marm2km}

\plot{Marm2kmVt}{width=0.8\columnwidth}{Marmousi example: the modified Marmousi model added with 2.2 km  water layer on the top and a high velocity salt box around the left part of the model. The water layer of 2.2 km in depth is used to avoid the refraction energy present in the recorded data.}

\multiplot{1}{Marm2kmV0,Marm2kmFWI,Marm2kmSurf}{width=0.8\columnwidth}{Marmousi example: (a) 1D initial velocity; inverted results by (b) FWI  with target source wavelet and (c) by surface source based MSWI without given source wavelet.}

\multiplot{2}{Marm2kmShot1,Marm2kmShot81}{width=0.4\columnwidth}{Marmousi example: shot gather of recorded data for shot at (a) $x_s=0.6$ km and (b) $x_s=8.6$ km.}

\multiplot{2}{Marm2kmShot1Inv,Marm2kmShot81Inv}{width=0.4\columnwidth}{Marmousi example: shot gather of  simulated data generated using the inverted velocity by surface source based MSWI for shot at (a) $x_s=0.6$ km and (b) $x_s=8.6$ km..}

The 200th FWI iterate is shown in Figure \ref{fig:Marm2kmFWI}, the 150th surface source extended FWI iterate in Figure \ref{fig:Marm2kmSurf}. All features of the target model, at both large and small scales to the resolution afforded by the data, are much better imaged in the latter.  Figures \ref{fig:Marm2kmShot1Inv}-\ref{fig:Marm2kmShot81Inv} show the simulated  data produced by the final inverted velocity, which compare very well with the target data (Figures \ref{fig:Marm2kmShot1}, \ref{fig:Marm2kmShot81}).

\section{Conclusion}
We have described the source extension modification of FWI, detailed several methods of this type found in the literature, and used the surface source extension to illustrate the type of result obtainable using this class of FWI modifications. Theoretical understanding is currently limited to transmission data inversion: several of the extended source FWI algorithms are closely related to variants of travel time tomography, and produce results of similar quality. No theoretical results are available for reflection data, nor for several source extensions described in the literature, notably WRI. Also, several promising algorithms are numerically challenging, requiring data storage and/or computation far in excess of the norm for conventional FWI. Both theory and algorithm development are promising avenues for further research on extended source FWI.

\paragraph{Acknowledgement:} GH and WS thank The Rice Inversion Project and Total E\&P USA for support. We thank Total E\&P USA for permission to publish these results.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}
%\bibliography{masterref}
%\end{document}

### 2019-10-19 repro/papers/sseprecond/paper.tex
repro docs dir added

\title{Efficient Computation of Extended Sources}
\author{William. W. Symes \thanks{The Rice Inversion Project,
Department of Computational and Applied Mathematics, Rice University,
Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}

\lefthead{Symes}

\righthead{Approximate Source Inversion}

\maketitle
\begin{abstract}
Source extension is a reformulation of inverse problems in wave propagation, that at least in some cases leads to computationally tractable iterative solution methods. The core subproblem in all source extension methods is the solution of a linear inverse problem for a source (right hand side in a system of wave equations) through minimization of data error in the least squares sense with soft imposition of physical constraints on the source via an additive quadratic penalty. A variant of the time reversal method from photoacoustic tomography provides an approximate solution that can be used to precondition Krylov space iteration for rapid convergence to the solution of this subproblem. An acoustic 2D example for sources supported on a surface, with a soft contraint enforcing point support, illustrates the effectiveness of this preconditioner.
\end{abstract}

\section{Introduction}
Full Waveform Inversion (FWI) can be described in terms of 
\begin{enumerate}
\item a linear wave operator $L[c]$, depending on an array of space-dependent coefficients $c$ and acting on wavefields $u$ vanishing in negative time:
\begin{equation}
\label{eqn:init}
u \equiv 0, t \ll 0; 
\end{equation}
\item a trace sampling operator $P$ acting on wavefields and producing data traces;
\item and a source function (of space and time) $f$ representing energy input to the system. 
\end{enumerate}
The basic FWI problem is: given data $d$, find $c$ so that 
\begin{equation}
\label{eqn:fwi}
Pu \approx d \mbox{ and } L[c]u = f.
\end{equation}

The energy source $f$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time. In fact, additional source degrees of freedom, beyond those needed to describe physically realized sources, may be useful in rendering the FWI problem \ref{eqn:fwi} more amenable to numerical solution, via so-called extended modeling (see \cite{geoprosp:2008}, \cite{LeeuwenHerrmannWRI:13}, \cite{HuangNammourSymesDollizal:SEG19}, and many references cited there). Therefore it is natural to view $f$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
%A typical nonlinear least squares formulation is:
%\begin{equation}
%\label{eqn:ols}
%\mbox{choose } c \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2.
%\end{equation}

%As is well-known, local optimization methods are the only feasible approach given the dimensions of a typical instance of \ref{eqn:fwi}, and those have a tendency to stall due to ``cycle-skipping''. Source extension is one approach to avoiding this problem. It consists in imposing the wave equation as a soft as opposed to hard constraint, by allowing the source to have more degrees of freedom than is permitted by a faithful  model model of the seismic experiment, and constraining these additional degrees of freedom by means of an additive penalty modifying the probem \ref{eqn:ols}:
\begin{equation}
\label{eqn:esi}
\mbox{choose } c, f \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2 + \alpha^2 \|Af\|^2 
\end{equation}
The operator $A$ penalizes deviation from known (or assumed) characteristics of the source function - its null space consists of feasible source models.

This paper presents a numerically efficient approach to solving the {\em source subproblem}
\begin{equation}
\label{eqn:esis}
\mbox{given } c, \mbox{ choose } f \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2 + \alpha \|Af\|^2 
\end{equation}

The penalty operator $A$ is assumed linear in this work, so the source subproblem is a linear least squares problem. Under some additional assumptions to be described below, I shall show how to construct an accurate approximate solution operator for problem \ref{eqn:esis}. This approximate solution operator may be used to accelerate Krylov space methods for the solution of the source subproblem \ref{eqn:esis}. Numerical examples suggest the effectiveness of this acceleration.

The source subproblem is a necessary step in solving the overall FWI problem \ref{eqn:fwi}. In particular, the nonlinear least squares realization \ref{eqn:esi} of source-extended FWI is naturally approached via the {\em variable projection method} \cite[]{GolubPereyra:73,GolubPereyra:03,vanLeeuwenMulder:09,Rickett:SEG12}, in which the minimum over $f$ in \ref{eqn:esis} is treated as a function of $c$, which is in turn minimized. Variable projection thus treats the source subproblem as an inner problem, which must be solved for an estimate of $f$ at every step of an iterative method for the outer problem of minimization over $c$. Efficiency in solving the source subproblem \ref{eqn:esis} is critical to efficient implementation of variable projection for solving the regularized FWI problem \ref{eqn:esi}.

This paper restricts the choice of wave physics embodied in $L[c]$ to linear acoustics.  The principal characteristics assumed of sources are localization and isotropy. That is, physical sources will be assumed to be isotropic point radiators, described by the functional form
\begin{equation}
\label{eqn:ptsrc}
f_{\rm pt}(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s). 
\end{equation}
Source locations $\bx_s$ are assumed known (as is natural for active source methods), and the source {\em wavelet} $w(t)$ is to be found as part of the solution of \ref{eqn:esi}. A convenient choice of penalty operator $A$ is then multiplication by the {\em offset} $|\bx-\bx_s|$: its null space consists precisely of distributions of the form \ref{eqn:ptsrc}.

More complex radiation characteristics may be accommodated with similar but more complex choice for the operator $A$. Since the core analysis is similar, this paper deals only with the simplest case \ref{eqn:ptsrc}.

The wavefield $u$ in this work is presumed to exist in all of Euclicean space-time: in other words, boundary effects have been removed from $u$, by surface-related multiple elimination or other techniques. Predicted data traces are simply the samples of one or more components of $u$ at receiver points $\bx_r$.

%The domain of wave propagation is denoted $\Omega$, and two subsets of its boundary are identified: $\Gamma_s$ contains the support (non-zeroes) of the (extended) sources and in particular the locations $\bx_s$ of physical sources. Data trace locations $\bx_r$ lie in another subset $\Gamma_r$. The data projection operator $P$ samples the acoustic field $u$ on $\Gamma_r$, and includes a finite aperture mute.

Achieving a small residual in the source subproblem \ref{eqn:esis} for arbitrary $c$ is a critical part of the convergence mechanism for the extended source approach to FWI \cite[]{HuangNammourSymesDollizal:SEG19}. However, for arbitrary data $d$, coefficient array $c$, and source location $\bx_s$, there does not exist an isotropic point source $f_{\rm pt}$ of the form \ref{eqn:ptsrc} for which $PL[c]^{-1}f_{\rm pt} \approx d$. This is so even if there does exist a coefficient array $c^*$ and point source $f_{\rm pt}^*$ located at $\bx_s$ with wavelet $w^*(t)$ for which $d = PL[c^*]^{-1}f_{\rm pt}^*$, if $c$ differs substantially from $c^*$. It is this fact that underlies the effectiveness of the extended source approach to FWI:  driving the residual in the problem \ref{eqn:esi} towards zero necessarily requires modifying $c$ to resemble $c^*$, assuming the latter exists, as well as pushing the optimal $f$ towards the null space of $A$. Conversely, the source subproblem \ref{eqn:esis} does not have a small residual solution of point source form \ref{eqn:ptsrc}, in general, for a given choice of $c$, even if that were the case for a different choice ($c^*$). Therefore a more general class of source models than that specified by the condition \ref{eqn:ptsrc} must be admitted to the feasible set for the source subproblem \ref{eqn:esis} if it is to serve as the inner problem in a variable projection formulation of extended source FWI.

The class of non-point sources investigated here are those confined a hypersurface assumed to contain the locations of physical (point) sources, but not a priori required to have point support. These {\em surface extended sources} are able to yield a relatively small residual in the source subproblem \ref{eqn:esis} for more or less arbitrary coefficient array $c$. This observation is a by-product of the approximate solution for \ref{eqn:esis} constructed below. The construction of the approximate solution is closely related to the {\em time reversal} method of photoacoustic tomography (see \cite{StefanovUhlmannIP:09} and references cited there).

The next section gives a precise description of an acoustic version of problem \ref{eqn:esis}, and points out two interesting special cases, the {\em crosswell} and {\em diving wave} configurations. The assumptions detailed here imply that the data $d$ of the inverse problems \ref{eqn:esi}, \ref{eqn:esis} have the physical character of {\em transmitted waves}. There are no mathematical results at present concerning reflected wave inverse problems treated by source extension methods, though there are tantalizing numerical clues \cite[]{LeeuwenHerrmannWRI:13,Warner:14,Warner:16,LeeuwenHerrmann:16,HuangSymes:Geo17,HuangSymes:Geo18a,HuangSymes:Geo18b}.

The following section describes the construction of the approximate inverse and explains the conditions under which accuracy should be expected. I observe that the approximate inverse is approximately the adjoint of the modeling operator $PL[c]^{-1}$ with respect to weighted norms in its domain and range (see \cite{HouSymes:15} for a similar observation in a different context). This fact immediately suggests use of the approximate inverse as a preconditioner to accelerate convergence of Krylov space methods such as conjugate gradient iteration applied to the least squares problem \ref{eqn:esis}. However, its straightforward application in the preconditioned conjugate gradient (PCG) method \cite[]{Golub:2012} involves explicitly a version the so called Dirichlet-to-Neumann (D2N) operator, which is computationally awkward. I show how to reorganize the PCG iteration so that only solutions of $L[c]u=f$ are required. The penultimate section presents some 2D examples in the crosswell configuration, illustrating the accuracy of the approximate inversion and the accelerated convergence of PCG. The examples are chosen to emphasize that surface extended sources can be constructed to fit more or less arbitrary data, unlike point sources - as mentioned before, this approximate invertibility property of the modeling operator is of critical importance in the application to extended FWI. The paper ends with a discussion of several important matters not addressed here, and a restatement of the conclusions.

The intent of the main body of this paper is to present the formal structure of the extended source subproblem and the way in which this structure leads to accelerated numerical solution. This formal structure is however supported by a rigorous foundation. In order to avoid disrupting the formal account, this foundation material is relegated to an appendix.

%The penalty operator $A$ is chosen to penalize non-physical aspects of the trial source $g$: physical sources are supposed to constitute the null space of $A$.

%In some cases, the resulting  relaxed data fit problem has been observed to be more amenable to solution by local optimization. The main goal of this paper is explain why this should be so,  in a particular instance of source-extended FWI. I will show that for a specific choice of {\em annihilator} $A$, the optimization problem \ref{eqn:esi} possesses a {\em qualitatively} larger domain of convexity for consistent data ($PL[c]^{-1}g = d, Ag=0$) than does the FWI problem \ref{eqn:ols}, while possessing the same solution.

%For sake of argument (and because almost all work so far has assumed it), adopt the isotropic point source as the model: for the data gather at source positon $\bx_s$, the source takes the form
%\begin{equation}
%\label{eqn:ptsrc}
%f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s). 
%\end{equation}
%in which the {\em wavelet} $w(t)$ has whatever structure is necessary for $f$ to lie in the range of the wave operator $L[c]$. FWI can also accommodate more complex source models reflecting the nontrivial radiation patterns of actual sources. 
%The data projection operator $P$ involves the restriction of one or more components of the field $u$ to a hypersurface $X_r$. The simplest interesting choice of $X_r$ is a hyperplane, say $z=z_r$ (with the convention that $\bx=(x,y,z)$). To emulate the limited aperture of any experimental data system, include multiplication by a smooth cut-off or taper function $\phi$. Component selection is a special case of multiplication by a matrix $P_c \in \bR^{n_d \times n_s}$, in which $n_d$ is the number of data components and $n_s$ is the dimension of the state vector $u$.
%\begin{equation}
%\label{eqn:data}
%Pu(x,y,t) = \phi(x,y,z_r)P_cu(x,y,z_r,t)
%\end{equation}
%in which $P_c$ is a

%Source extension replaces $f(\bx,t;\bx_s)$ with an artificial source $g(\bx,t; \bx_s)$ with more degrees of freedom than merely a single common source wavelet $w(t)$ as in definition \ref{eqn:ptsrc}. Assuming that \ref{eqn:ptsrc} is actually correct physics, these extra degrees of freedom have to be suppressed in the optimal solution, via an {\em annihilator} $A$, an operator acting on extended sources $g$, whose null space is precisely the physical sources (of the form \ref{eqn:ptsrc}). Extended source inversion replaces the FWI objective with a penalty function

%The way in which degrees of freedom are introduced, and the choice of annihilator, distinguish the various source extension methods, of which there are many - see \cite{HuangNammourSymesDollizal:SEG19}, who introduce the {\em surface source} extension. It is this surface source extension that is the topic of this paper. It is similar  to the volume extension \cite[]{HuangSymes:Geo18b}, but the source is spread over the source space-time surface, rather than over space at $t=0$.
%Similar to the space-time and volume extensions \cite[]{HuangSymes:Geo18a,HuangSymes:Geo18b}, the annihilator is given by the {\em localization penalty}:
%\begin{equation}
%\label{eqn:msste}
%A[g](\bx,t;\bx_s) = |\bx-\bx_s|g(\bx,t;\bx_s). 
%\end{equation} 
%Therefore an estimate of the the source wavelet is a by-product of the inversion, and does not need to be specified {\em a priori}. Like the volume extension (and unlike, for example, Wavefield Reconstruction Inversion \cite[]{LeeuwenHerrmannWRI:13,LeeuwenHerrmann:16}), the additional data volume required is on the same order of size as the data.

%The volume and surface extensions have a well-defined relation to tomographic principles (slope or traveltime, depending on type of annihilator). This relation holds regardless of ray geometry (with or without triplication) in transmission configuration.

%Again, nothing is known theoretically about the behaviour of any of these methods for reflection data, and numerical evidence is ambiguous.

\section{Extended Source Modeling}
For acoustic wave physics, the coefficient vector $c$ is the pair $c=(\kappa,\beta)$ of bulk modulus $\kappa$ and buoyancy (reciprocal density) $\beta$, and the state vector $u=(p,\bv)$ consists of pressure $p$ (a scalar space-time field) and particle velocity $\bv$ (a vector space-time field). The wave operator $L[c]$ is given by the linear acoustics system:
\begin{equation}
\label{eqn:awe}
L[c]u = 
\left(
\begin{array}{c}
\partial_t p  + \kappa \nabla \cdot \bv \\
\partial_t \bv + \beta \nabla p
\end{array}
\right) 
\end{equation}

The space dimension will be denoted by $d$. Most of what follows is valid for any $d >0$. The choice $d=2$ is computationally convenient, if non-physical.

As mentioned earlier, in this paper the coefficients $c$ is regarded as defined throughout space $\bR^d$, the state vector $u$ throughout space-time $\bR^{d+1}$.

Real-world sampling is discrete and direction-sensitive. This paper ignores such issues and presumes that the sampling operator $P$ simply restricts its argument to a cylindrical space-time hypersurface $\Gamma_r \times \bR$.  Here $\Gamma_r \subset \bR^d$ is an open subset of a smooth codimension 1 oriented hypersurface. To distinguish this output or receiver sampling operator from another similar operator, add the subscript $r$ to the notation, that is, write $P_r$ rather than $P$. For technical reasons (and to emulate actual acquisition practice) include multiplication by a cutoff or mute $\phi_r \in C_0^{\infty}(\bR^{d+1})$, and extract only samples of the pressure field:
\begin{equation}
\label{eqn:recp}
(P_ru)(\bx,t) = \phi_r(\bx,t)p(\bx,t),\,\bx \in \Gamma_r, t \in \bR. 
\end{equation}
A similar sampling operator for the velocity field will also be required below. Denote by ${\bf n}_r$ the outward unit normal field on $\Gamma_r$, and define
\begin{equation}
\label{eqn:recv}
(V_ru)(\bx,t) = \phi_r(\bx,t){\bf n}_r(\bx)\cdot \bv(\bx,t),\,\bx \in \Gamma_r, t \in \bR.
\end{equation}

Physical sources are modeled as an isotropic constitutive law defects of point support. As mentioned earlier, it is simple to accommodate more complex source models. Assume that all source lie in an open subset $\Gamma_s \subset \bR^d$ of a smooth orientable codimension 1 hypersurface in $\bR^d$, possibly different from the hypersurface containing $\Gamma_r$. Denote by $P_s, V_s$ sampling operators on $\Gamma_s \times \bR$ defined precisely analogously to $P_r$ and $V_r$ (equations \ref{eqn:recp}, \ref{eqn:recv}):
\begin{equation}
\label{eqn:recp}
(P_su)(\bx,t) = \phi_s(\bx,t)p(\bx,t),\,\bx \in \Gamma_s, t \in \bR. 
\end{equation}
and similarly for $V_s$. Denote by $\delta_{\bx_s}$ the delta distribution on $\Gamma_s$ at $\bx_s \in \Gamma_s$, and by $w \in C_0^{\infty}(\bR)$ a choice of source wavelet. The corresponding physical source is then the right-hand side in the acoustic system
\begin{equation}
\label{eqn:awe-src}
L[c]u = P_s^T(w\delta_{\bx_s}) 
\end{equation}
With dummy space and time variables, the right hand side above is exactly the form $w(t)e_1\delta(\bx-\bx_s)$ given in the point source definition \ref{eqn:ptsrc}, where , ${\bf e}_1$ the standard basis vector $e_1=(1,0,...,0) \in \bR^{d+1}$.

As explained in the introduction, unless the data $d$ and the coefficient array $c$ are compatible, the source subproblem \ref{eqn:esis} has no low-residual solution of the form $f = P_s^T(w\delta_{\bx_s}) $. A larger class of source models is required to fit data. A precise description of the class described in the introduction is simply the range of $P_s^T$: that is, for $h \in C^{\infty}(\Gamma_s \times \bR)$, define
\begin{equation}
S[c]h = P_rL[c]^{-1}P_s^T h.
\end{equation}
$S[c]$ is the surface source extension modeling operator. For this class of sources, the source subproblem becomes 
\begin{equation}
\label{eqn:esis-sse}
\mbox{given } c, \mbox{ choose } h  \mbox{ to minimize } \|S[c]h -d \|^2 + \alpha \|Ah\|^2 
\end{equation}
To complete the definition of this problem, a choice of {\em annihilator} $A$ is required. A simple choice is
\begin{equation}
\label{eqn:msste}
Ah(\bx,t) = |\bx-\bx_s|h(\bx,t). 
\end{equation}

Formally, the solution of the source subproblem is also the solution of the normal equation
\begin{equation}
\label{eqn:normal}
(S[c]^TS[c] + \alpha^2 A^TA) h = S[c]^T d 
\end{equation}
This relation holds, and a solution $h$ is guaranteed to exist, if the $\alpha=0$ normal operator $S[c]^TS[c]$ is well-defined and continuous in $L^2(\bar{\Gamma}_s \times \bR)$, and there exists $C>0$ so that 
\begin{equation}
\label{eqn:coercive}
\|S[c]h\|^2 \le C|h|^2,\,\,\|S[c]h\|^2 + \alpha^2 \|Ah\|^2 \ge C^{-1}\|h\|^2
\end{equation}
for all $h \in L^2(\bar{\Gamma}_s)$. These estimates hold under various conditions. One particular setting in which the least squares problem \ref{eqn:esis} is well-posed is given by

\noindent {\bf Assumption A:} There exists $\theta_*>0$ so that for every ray connecting a point in $\mbox{supp }\phi_s$ with a point in $\mbox{supp }\phi_r$, the angles $\theta_s$ and $\theta_r$ subtended by the ray and the tangent planes of $\Gamma_s$ and $\Gamma_r$ respectively are both greater than $\theta_*$ in absolute value.

The first appendix provides a proof of the following statement:

\begin{theorem} 
\label{thm:one}
Suppose that $\log \kappa, \log \beta \in C^{\infty}(\bR^d) \cap L^{\infty}(\bR^d)$, and that Assumption A holds. Then the estimates \ref{eqn:coercive} hold for $h \in C^{\infty}_0(\Gamma_s \times \bR)$, hence $S[c]$ extends to $L^2(\bar{\Gamma}_s \times \bR)$ with the same estimates.
\end{theorem}

The assumptions made in the hypotheses of \ref{thm:one} imply that the inverse wave problems under consideration in this paper are of the transmission type, that is, do not involve reflected waves. The reflected wave concept is well-defined only in terms of high-frequency asymptotics: because the coefficients are assumed smooth, there are no material singularities to generate reflections.  Two configurations of sources and receivers that naturally fit this framework, at least in an ideal sense, are
\begin{itemize}
\item the {\em crosswell} configuration, in which $\Gamma_s$ and $\Gamma_r$ are parallel hyperplanes. In reality, wells are approximately 1D structures embedded in 3D ambient media, so this configuration does not really model them well. However it has become commonplace to use the 2D version of this configuration as a model, assuming that the waves propagate in-plane between the wells  \cite[]{BregChapBai:86,Harr:93,Zhouetal:93,Pratt:99b,Plessix:00}.
\item the {\em diving wave} configuration, $\Gamma_s$ and $\Gamma_r$ are both subsets of a horizontal plane at the source, respectively receiver, depth (assumed well-defined). In marine towed streamer and ocean bottom seismometer surveys, a set of refracted waves arrives earlier than the direct wave in the water column, at sufficient distance from the source. The mutes $\phi_s, \phi_r$ isolate these refacted (``diving wave'') arrivals. This data is the focus of the ``classic'' period of FWI \cite[]{SirgPratt02,BrendersPratt:06a,BrendersPratt:06b,VirieuxOperto:09,Vigh:10,Vigh13}.
\end{itemize}
In both cases, it is possible (though not necessary) that Assumption A is satisfied. Note that in the diving wave configuration, the mutes $\phi_S,\phi_r$ are responsible for eliminating direct wave energy, hence enforcing the acute incidence of the waves connecting the admitted source and receiver points. The numerical examples presented below will fall into the crosswell category.

Assumption A is not by any means the most general hypothesis under which the source subproblem has a well-behaved solution. However weaker hypotheses typically involve multidimensional filters applied to data (and extended sources) to eliminate waves propagating tangent to the source and receiver surfaces, so require more involved implementations.

%Denote by $J_{\alpha}[g,c;d]$ the objective function defined in \ref{eqn:esi}:
%\begin{equation}
%\label{eqn:jdef}
%J_{\alpha}[g,c;d] = \frac{1}{2}( \|PL[c]^{-1}g -d \|^2 + \alpha \|A[g]\|^2) 
%\end{equation}
%To compute the objective 
%\begin{itemize}
%\item solve the acoustic wave system \ref{eqn:awe} with $g$ as in \ref{eqn:mssur}
%\item sample at receiver locations and compute mean-square difference with pressure data traces, take mean square
%\item add scaled mean square of annihilator applied to the source
%\end{itemize}

%The variable projection method \cite[]{GolubPereyra:03} applied to the objective \ref{eqn:esi} approaches its minimization via a nested algorithm: first minimize \ref{eqn:esi} over the source $g$ for given coefficients $c$, thus making $g = g[c]$ a function of $c$ (inner minimization); then substitute $g[c]$ in $J$ and minimize the resulting function of $c$ (outer minimization). The optimal $g[c]$ solves the first order necessary condition for minimization of $J$ over $g$: $\nabla_g J_{\alpha}[g,c;d] = 0$. For linear annihilators such as that defined by \ref{eqn:mssur}, $J$ is quadratic in $g$, so the first order necessary condition is also sufficient, and is equivalent to the linear {\em normal equation}:

%where I have written $S[c]=PL[c]^{-1}$ for the operator that maps the (extended) source to the data traces. Substitution of $g[c]$ in $J$ results in the {\em reduced objective}
%\begin{equation}
%\label{eqn:jred}
%\tilde{J}_{\alpha}[c;d] = J_{\alpha}[g[c],c;d].
%\end{equation}
%The gradient would appear to involve the derivative of the inner solution $g[c]$ with respect to $c$, computation of which implies solution of additional linear systems. However by fortunate accident, the chain rule and the normal equation \ref{eqn:normal}
%conspire to eliminate all terms involving derivatives of $g[c]$. Consequently the direction derivative of $\tilde{J}_{\alpha}$ in the direction $\delta c$ is
%\begin{equation}
%\label{eqn:tildejderiv}
%D_c\tilde{J}_{\alpha}[c;d]\delta c = \langle D(S[c]g)\delta c,S[c]g-d \rangle|_{g=g[c]}
%\end{equation}
%whence
%\begin{equation}
%\label{eqn:l2grad}
%\nabla \tilde{J}_{\alpha}[c;d] = D_c(S[c]g)^T|_{g=g[c]}(S[c]g[c]-d)
%\end{equation}
%Let us field-strip the expression on the right. The rightmost factor is the data residual, which is a by-product of many methods for the solution of the normal equation \ref{eqn:normal}. $D_cS$ is the linearization or Born approximation of $S$, or perturbation with respect to $c$, with $g$ fixed:
%\[
%S[c+\delta c]g \approx S[c]g + D_c(S[c]g)\delta c
%\]
%That is, $D_c(S[c]g)$ is the object commonly called the Born modeling operator, with source wavefield $g$. Its adjoint or transpose is $D_c(S[c]g)^T$ - as is well known, this is the RTM operator (or rather a version of it), computed via the adjoint state method in the time domain. Note that the Born and RTM operators need to match, i.e. both be derived from the same underlying full waveform modeling operator (in this case acoustodynamics, but the same would apply for any other wave theory).

%The upshot is that the computation of the gradient goes like this:
%\begin{itemize}
%\item solve the normal equation \ref{eqn:normal} for $g[c]$
%\item compute (or retain) the data residual $S[c]g[c]-d$
%\item reverse-time migrate the data residual, using $g[c]$ to generate the source wavefield in the adjoint state method - the output is the gradient of $\tilde{J}_{\alpha}$.
%\end{itemize}

%To make this algorithm practical and efficient, effective preconditioning needs to be applied to both inner and outer problems.

\subsection{Time Reversal}
%In contrast to the outer problem, a preconditioner for the inner problem is available that works regardless of medium smoothness, so for both transmission and reflection (the latter with some caveats). 
%First the transmission case - a sort of rotated crosswell configuration. Idealize the receiver set as a continuum plane $z=z_r$, with $z_r > z_s$ - not essential but simpler.

%NB: the definition of the forward map (and the data) should include a taper $\phi(\bx_r,\bx_s)$ masking out the shot gather $d_r(\cdot,\cdot;\bx_s)$ for each source position $\bx_s$. However I will not explicitly mention this operation.

Change notation slightly, and write $S[c]h_s$ rather than $S[c]g$ with $g = h_s \delta(z-z_s)$, as $z_s$ will be presumed ot be the same for all shots. The pressure and velocity fields occuring in the computation of $d_r=S[c]h_s$ form the solution of
\begin{eqnarray}
\label{eqn:awedata}
\frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv +
h \delta(z-z_s), \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t<0\nonumber\\ 
\bv & = & 0 \mbox{ for } t<0 \nonumber\\
 p|_{z=z_r}& = & d_r
\end{eqnarray}
Suppose that the recording time interval $[0,T]$ is long enough that $ d_r \approx 0$ for $t > T$. Note that $p,\bv$ are {\em incoming} (downgoing) in $z>z_s$. Therefore in $z<z_r$ {\em the same fields} approximately solve
\begin{eqnarray}
\label{eqn:awetr}
\frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t>T\nonumber\\ 
\bv & = & 0 \mbox{ for } t>T \nonumber\\
p|_{z=z_r} &=&  d_r
\end{eqnarray} 
The last equation defines $p$ on the entire plane $z=z_r$ for $t<T$, and the previous two define final conditions at $t=T$. So the {\em time-reversed} system \ref{eqn:awetr} has unique solutions in $z<z_r, t<T$ and $z>z_r,t<T$. The solution in $z<z_r$ is incoming (downgoing) at $z=z_s$, and is {\em the same vector of fields} (approximately) as the solution of \ref{eqn:awedata} in the region $z_s<z<z_r$ between the source and receiver surfaces.

Define $d_s =p|_{z=z_s}$ for the solution of \ref{eqn:awedata}. Then $p,\bv$ is also the solution of
\begin{eqnarray}
\label{eqn:awedir}
\frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
p &=&0, t< 0,\nonumber\\
\bv &=&0,t<0,\nonumber\\
p|_{z=z_s }& = &d_s
\end{eqnarray}
in both $z>z_s$ and $z<z_s$, since the pressure field for the solution of \ref{eqn:awedata} is continuous across $z=z_s$. Since the same fields solve the two different systems, there must be a relation between the data. In fact, the normal component of velocity $v_z$ is discontinuous, and (recalling the $g(\bx,t;\bx_s) = h_s(x,y)\delta(z-z_s)$). A calculation using integration by parts shows that $h_s$ is precisely the jump in $v_z$ across $z=z_s$:
\begin{equation}
\label{eqn:vjump}
h_s = [v_z]|_{z=z_s}.
\end{equation}
A further simplification occurs if $\kappa, \rho$ are constant near $z=z_s$, then by symmetry
\begin{equation}
\label{eqn:vsimp}
[v_z]|_{z=z_s}=2 \lim_{z \rightarrow z_s^+} v_z
\end{equation}

With these preparations, define 
\begin{equation}
\label{eqn:psinv}
S[c]^{\dagger}d_r= [v_z]|_{z=z_s}.
\end{equation}
where the jump in $v_z$ at $z=z_s$ is computed by
\begin{itemize}
\item solving \ref{eqn:awetr} backwards in time,
\item reading off $d_s=p$ at $z=z_s$,
\item solving \ref{eqn:awedir} and computing $[v_z]|_{z=z_s}$, or if appropriate simply using \ref{eqn:vsimp}.
\end{itemize}
 The relations developed in the last few paragraphs may be summarized as follows:
\begin{equation}
\label{eqn:appinv}
d_r = S[c]h_s \,\,\Leftrightarrow \,\, h_s \approx S[c]^{\dagger}d_r
\end{equation}
that is, $S[c]^{\dagger}$ is an approximate inverse of $S[c]$.

As I will show in the example section, the approximation may be accurate enough without any further improvement. However, a closer examination shows that in fact it is possible to express the relation between $S[c]$ and $S[c]^{\dagger}$ in terms of a change of norm, that is, in terms of a preconditioner, and this relation can be used to considerably accelerate an iterative solution of the normal equation \ref{eqn:normal}.

To see this, first compute the adjoint $S[c]^T$, by a variant of the adjoint state method, in this case a by-product of the conservation of energy. Suppose that 
$\tilde{p},\tilde{\bv}$ solve
\begin{eqnarray}
\label{eqn:aweadj}
\frac{\partial \tilde{p}}{\partial t} & = & \kappa (-\nabla \cdot \tilde{\bv} + 
 d_r \delta(z-z_r))\nonumber \\
\frac{\partial \tilde{\bv}}{\partial t} & = & - \beta \nabla \tilde{p},\nonumber \\
\tilde{p} & =& 0 \mbox{ for } t>T\nonumber\\ 
\tilde{\bv} & = & 0 \mbox{ for } t>T 
\end{eqnarray} 
Then
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  
\frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=T}
-
\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=0}
\]
\[
= 
\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)
\]
\[
= 
\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \tilde {p} + \kappa p \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} \right.
\]
\[
+
\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \tilde{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \tilde{\bv}}{\partial t} \right)
\]
\[
= 
\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
 \frac{h_s}{\kappa} \delta(z-z_s)\right) \tilde{p} + p \left(- \nabla \cdot \tilde{\bv} + 
 d_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  (- \nabla p) \cdot \tilde{\bv} + \bv \cdot (-\nabla \tilde{p}) \right)
\]
\[
= 
\int_{0}^{T}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
 \frac{h_s}{\kappa} \delta(z-z_s)\right) \tilde{p} + p \left(- \nabla \cdot \tilde{\bv} + 
 d_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  p (\nabla \cdot \tilde{\bv}) + (\nabla \cdot \bv) \tilde{p} \right)
\]
after integration by parts in the last two terms. Most of what is left cancels, leaving 
\[
0 = \int\,dt\,dx\,dy\, h_s \left(\frac{\tilde{p}}{\kappa}\right)_{z=z_s} + (p  d_r)_{z=z_r}
\]
whence
\begin{equation}
\label{eqn:sadj}
 S[c]^T d_r = -\left(\frac{\tilde{p}}{\kappa}\right)|_{z=z_s}
\end{equation}
Both $S[c]^T$ and $S[c]^{\dagger}$ are defined by time-reversed solutions of the same acoustic system, except with different data. System \ref{eqn:aweadj} places $d_r$ as a pressure source on $z=z_r$, with an additional factor of $\kappa$, and its trace on $z=z_s$ must be scaled by $1/\kappa$. System \ref{eqn:awetr} employs $d_r$ as a Dirichlet boundary condition, and the trace of its pressure component must be employed as a Dirichlet boundary condition for the forward-in-time problem \ref{eqn:awedir}, after which the jump in the normal velocity is extracted.

The relation between a Dirichlet boundary condition and a source has already been explained, in the context of justifying the approximation property of $S[c]^{\dagger}$:
\begin{quote}
The surface source producing the same field as a Dirichlet boundary condition (on either side of the boundary surface) is the jump in the normal velocity component of the Dirichlet field. 
\end{quote}
Call the operator that produces the jump in normal velocity from Dirichlet pressure data $\Lambda[c]_s$ for the source surface $z=z_s$, $\Lambda[c]_r$ for the receiver surface $z=z_r$ (in the math literature, this is the ``Dirichlet-to-Neumann operator'', more or less). The discussion so far can be summarized as follows:
\begin{equation}
\label{eqn:wadj}
S[c]^{\dagger} = W[c]_m^{-1}S[c]^TW[c]_d 
\end{equation}
in which 
\begin{equation}
\label{eqn:weights}
W[c]_m = \kappa^{-1},\,\, W[c]_d = \kappa^{-1}\Lambda[c]_r
\end{equation}
in which ``$\kappa$'' means the operator ``multiply by $\kappa$''. This means that $S[c]^{\dagger}$ {\em is} the adjoint of $S[c]$ with respect to weighted norms 
\begin{itemize}
\item on the source space ($h_s$): weight $W[c]_m$
\item on the data space ($d_r$): weight $W[c]_d$
\end{itemize}
Since $S[c]^{\dagger}S[c] \approx I$ as shown above, $S[c]$ is approximately unitary in the Hilbert spaces of source and data traces with norms defined by the weighting operators $W_m$ and $W_d$ respectively. Therefore a Krylov space method employing these norms will converge rapidly.

\cite{HouSymes:EAGE16} demonstrated a very similar preconditioner for Least Squares Migration, also for its subsurface offset extension \cite[]{HouSymes:16}, motivated by \cite{tenKroode:12}. These constructions all involve the Dirichlet-to-Neumann operator. This concept also turns up in hidden form in the work of Yu Zhang and collaborators on true amplitude migration \cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,ZhangYuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.

\section{Prototype Numerical Examples}
I present a collection of simple examples that illustrate the features of the surface source extension claimed in preceding sections.

I used the IWAVE acoustic staggered grid package to carry out these calculations. This package implements (2,2k) schemes for k=1,2,..., and outputs traces (of either velocity or pressure at any point in space via multilinear interpolation. The discretized modeling operator is thus of second order accuracy, though as usual I have used higher order in space to reduce grid dispersion.

Source injection is implemented as the adjoint of trace sampling, resulting in another second-order error [REFERENCES].

The data is a single shot gather, with a source at coordinates $x_s=z_s=3000$ m (units of length are meters in all cases). The receiver line occupies $1500 \le x_r \le 5500$ m, with receiver depth $z_r=1000$ m.  Extended sources occupy $1500 \le x_s \le 5500$ m, with the same depth $z_s=3000$ m as the ``physical'' source used to generate the data. This region turns out to be adequate to represent the extended sources that approximately invert the data, for the cases examined below. An algorithm to automatically identify an appropriate region can be based on the ideas developed by \cite{Fu:Geo17}.

I have used absorbing boundary conditions (split-field PML) on all four sides of the 4000 m (vertical) $\times$ 8000 m (horizontal) simulation domain. Evidently inclusion of a free surface is important to the application of the ideas explained here to diving wave marine data, and I have not addressed the necessary modifications here.

The Dirichlet-to-Neumann operator $\Lambda$ is an essential part of the inner problem preconditioner just presented. Implementation can be accomplished in several ways:
\begin{itemize}
\item \cite{tenKroode:12} suggests using a one-way operator;
\item if both pressure and normal partical velocity are measured (or simulated), then the two are related by $\Lambda$ and the velocity component can simply be used as the output;
\item presence of a free surface implies all of the usual problems, such as the need for removal of receiver-side ghosts. On the other hand, if the free surface is within a quarter-wavelength throughout the useful bandwidth of the data, then the ghosted data differs from $\Lambda p$ by a time integration and a scale factor, a fact used to good effect by \cite{HouSymes:15}.
\end{itemize}

In the examples, I have used the second observation. With a finite difference implementation of the pressure-velocity system \ref{eqn:awe}, velocity components are available ``for free'', short-circuiting explicit computation of $\Lambda$.

The  {\tt project/SConstruct} script is set up to carry out the necessary computations on grids with spacings $\Delta x = \Delta z = $ 20, 10, and 5 m, with a jump of roughly 8 in computation time resulting from each refinement. For present purposes, the coarsest (20 m) grid seems to be sufficient, and that is the grid used in the examples presented below. The source pulses are chosen so that the computation is reasonably accurate. For the 20 m grid case, I use a zero-phase trapezoidal bandpass filter source with corner frequencies of 1.0, 2.0, 7.5, and 12.5 Hz.

The IWAVE asg driver has been set up to recognize the case {\tt deriv=0} as defining the map from source (right-hand side in the pressure equation) to data (pressure) traces. The adjoint to this map, as explained above, is reverse-time propagation of the data traces as pressure sources, followed by scaling (formula \ref{eqn:sadj}). The approximate inverse is computed by application of the Dirichlet-to-Neumann operator to the pressure traces to produce corresponding velocity traces, followed by injection as pressure sources and reverse time propagation, followed by another application of the Dirichlet-to-Neumann map and scaling (formulas \ref{eqn:appinv}, \ref{eqn:adj}).

The script implements these operatations step-by-step via calls to IWAVE, Madagascar, and SU commands. A peculiarity of the {\tt asg.x} driver needs to be mentioned: it is based on an un-scaled version of the constitutive law defect source representation, that is, {\tt asg.x} approximately computes the solution of the system \ref{eqn:awedata} with the first equation replaced by
\begin{equation}
\label{eqn:asgdata}
\frac{\partial p}{\partial t}  =  - \kappa \nabla \cdot \bv +
h \delta(z-z_s).
\end{equation}
Denote by $S_{\rm asg}[c]$ the forward map produced by {\tt asg.x}. Then comparison of \ref{eqn:awedata} and \ref{eqn:asgdata} reveals that
\begin{equation}
\label{eqn:sreln}
S[c]=S_{\rm asg}[c]\kappa
\end{equation}
where $\kappa$ is shorthand for the operator of multiplication by $\kappa$. Accordingly, and approximate inverse for $S_{\rm asg}[c]$ is
\[
I \approx S_{\rm asg}[c]^{\dagger}S_{\rm asg}[c] = S_{\rm asg}[c]^{\dagger}S[c]\kappa^{-1}
\]
Since $S[c]S[c]^{\dagger} \approx I \approx S[c]^{\dagger}S[c]$, it follows that
\[
S_{\rm asg}[c]^{\dagger}=\kappa S[c]^{\dagger} = \kappa\Lambda[c]_s S[c]^T \Lambda[c]_r 
\]
\begin{equation}
\label{eqn:asginv}
= \kappa \Lambda[c]_s \kappa S_{\rm asg}[c]^T\Lambda[c]_r
\end{equation}
This is the approximate inverse computed in the examples. Note that only the values of $\kappa$ near the source datum $z=z_s$ play a role in the relation \ref{eqn:sreln} or in the definition \ref{eqn:asginv} of the IWAVE ASG approximate inverse.
\bibliographystyle{seg}
\bibliography{../../bib/masterref}

\append{Proof of Theorem 1}
Following the notation of \cite{BaoSy:91b}, denote by $\gamma_s$ ($\gamma_r$)the closed conic subset of $T^*(\Gamma_s\times \bR)|_{\mbox{supp } \phi_s}$ ($T^*(\Gamma_r\times \bR)|_{\mbox{supp } \phi_r}$) consisting of tangential components of null bicharacteristics passing over points of $\mbox{supp }\phi_s$ and $\mbox{supp }\phi_r$. That is, the ray with initial condition $(\bx,t,\xi,\omega) \in T^*(\Gamma_s\times \bR)|_{\mbox{supp } \phi_s}$ passes over $\mbox{supp }\phi_r$ if and only if $(\bx,t,\xi,\omega) \in \gamma_s$. Since each such ray is paired with the ray with the same spatial projection traversed backwards, it is also the case that $(\bx,t,\xi,\omega) \in T^*(\Gamma_r\times \bR)|_{\mbox{supp } \phi_r}$ passes over $\mbox{supp }\phi_s$ if and only if $(\bx,t,\xi,\omega) \in \gamma_r$.

Define $s=\sqrt{\kappa/\rho}$ to be the wave slowness (reciprocal velocity), and
\[
c(\bx,t,\xi,\omega) = s(\bx)^2\omega^2-|\xi|^2
\]
to be the characteristic polynomial of the acoustic wave system, and $C=c^{-1}(0)$ the characteristic variety.  The angle condition in Assumption A implies the existence of a conic neighborhood $\tilde{\gamma}_s$ of $\gamma_s$, so that for  $(\bx,t,\xi,\omega) \in \tilde{\gamma}_s$, the polynomial $\tau \mapsto c(\bx,t,\xi + \tau{\bf n}_s(\bx),\omega)$ has two distinct roots, $\pm \tau_s$. Let $\gamma_{1,s}$ be a conic neighborhood of the complement $\tilde{\gamma}_s^C$, chosen so that $\gamma_{1,s} \cap \gamma_s = \emptyset$. Let $\gamma_s' = \Pi_s^*(\gamma_{1,s})$, where $\Pi_s^*$ is the pullback map: $T^*(\Gamma_s\times \bR) \rightarrow T^*(\bR^{d+1})$. Similarly, construct $\gamma_r'$ and $\gamma_{1,r}$.

The null bicharacteristic through $(\bx,\xi,\omega,\xi) \in \gamma_r' \cap C$ does not pass over $\mbox{supp } \phi_s$ by construction, hence the RHS of the system $L[c]u=P_r^Th$ is microlocally of class $H^1$ (in fact, smooth) near this bicharacteristic,  thanks to the microlocal propagation of singularities theorem \cite[]{BealsReed:82}: that is, $p \in H^1_{ml}(\gamma_r')$.

Need to know $p \in L^2$. Restart: consider $F_{r,s}$ mapping Cauchy data at $t=T$ by reverse time to $P_{r,s} p$, $\mbox{supp }\phi_s \cup \mbox{supp }\phi_r \subset \{t<T\}$. Only relevant rays are continuations of those in $\gamma_r$. Represent $S$ as composition: $S=F_rF_s^T$. $F_s$ $L^2$-bounded $\Rightarrow$ for $u, v \in C_0^{\infty}$,
\[
|\langle F_s u, v \rangle| = \le C\|u\|\|v\|=|\langle u, F_s^T v\rangle|
\]
Indep arg establishes that $F_s^Tv$

%for which $(\bx,t,\xi\pm \tau_s{\bf n}_s(\bx),\omega)$ is the initial condition on $\Gamma_s \times \bR$ of a ray passing over $\mbox{supp }\phi_r$. Choose a conic neighborhood $\tilde{\gamma}_s$ of $\gamma_s$

Denote by

%,  $T^*(\Gamma_r\times \bR)|_{\mbox{supp \phi_r}}$ for which Note that the angle condition mentioned in Theorem \ref{thm:one} implies that if $(\bx,t,\xi,\omega) \in \gamma' $ and $\bx \in \Gamma_s$ or $\Gamma_r$, then $\omega$ is a simple root of are distinct, and even a lower bound on the difference proportional to $\epsilon$. It also implies that $\gamma'$ is disjoint from a neighborhood of the normal bundles of $\Gamma_s \times \bR$, $\Gamma_r \times \bR$.

\append{Transpose of Full Acoustic Modeling Operator}

In this section I compute the adjoint operator form modeling with both pressure and velocity source inputs and pressure and velocity trace outputs. The forward modeling operator is defined by solving
\begin{eqnarray}
\label{eqn:fawedata}
\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
h_s \delta(z-z_s)), \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta ( \nabla p + l_s {\bf e}_z\delta(z-z_s)),\nonumber \\
p & =& 0 \mbox{ for } t<0\nonumber\\ 
\bv & = & 0 \mbox{ for } t<0 \nonumber\\
\end{eqnarray}
The new elements here are the point load time function $l_s(\bx,t;\bx_s)$, and the vertical unit vector ${\bf e}_z=(0,1)^T$ for 2D, $=(0,0,1)^T$ for 3D. The forward map ${\bf F}$ is defined as implicit in the discussion of the Intrinsically Preconditioned CG algorithm: let ${\bf h}_s=(h_s,l_s)^T$, then 
\begin{equation}
\label{eqn:ffwd}
{\bf F}{\bf h}_s =\left\{ \left[
\begin{array}{c}
p(\bx,t;\bx_s)\\
v_z(\bx,t;\bx_s)
\end{array}
\right]_{z=z_r}\right\}
\end{equation}

Let $\tilde{\bf h}_r=(\tilde{h}_r,\tilde{l}_r)^T$ be a function on $z=z_r$, 
and 
$\tilde{p},\tilde{\bv}$ solve the backwards-in-time boundary value problem:
\begin{eqnarray}
\label{eqn:faweadj}
\frac{\partial \tilde{p}}{\partial t} & = & \kappa (-\nabla \cdot \tilde{\bv} + 
 \tilde{h}_r \delta(z-z_r))\nonumber \\
\frac{\partial \tilde{\bv}}{\partial t} & = & \beta (\nabla \tilde{p} + \tilde{l}_r{\bf e}_z\delta(z-z_r),\nonumber \\
\tilde{p} & =& 0 \mbox{ for } t>T\nonumber\\ 
\tilde{\bv} & = & 0 \mbox{ for } t>T 
\end{eqnarray} 
Then
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  
\frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=T}
-
\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=0}
\]
\[
= 
\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)
\]
\[
= 
\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \tilde {p} +  p \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} \right.
\]
\[
+
\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \tilde{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \tilde{\bv}}{\partial t} \right)
\]
\[
= 
\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
 h_s \delta(z-z_s)\right) \tilde{p} + p \left(- \nabla \cdot \tilde{\bv} + 
 \tilde{h}_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  (- \nabla p + l_s{\bf e}_z\delta(z-z_s)) \cdot \tilde{\bv} + \bv \cdot (-\nabla \tilde{p} + \tilde{l}_r{\bf e}_z\delta(z-z_r)) \right)
\]
\[
= 
\int_{0}^{T}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv \tilde{p} + 
 h_s\tilde{p}|_{z=z_s} - p\nabla \cdot \tilde{\bv} + 
 \tilde{h}_r p|_{z=z_r}\right) \right.
\]
\[
+
\left.  p (\nabla \cdot \tilde{\bv}) + l_s \tilde{v}_z|_{z=z_r}+ (\nabla \cdot \bv) \tilde{p}  + l_r  v_z|_{z=z_r}\right)
\]
after integration by parts in the last two terms. Most of what is left cancels, leaving 
\[
0 = \int\,dt\,dx\,dy\, (h_s \tilde{p})_{z=z_s} + (p  \tilde{h}_r)_{z=z_r} + (l_s\tilde{v}_z)_{z=z_s} + (v_z \tilde{l}_r)_{z=z_r},
\]
whence
\begin{equation}
\label{eqn:fadj}
 {\bf F}^T {\bf h}_r = -\left[
\begin{array}{c}
\tilde{p}(\bx,t;\bx_s)\\
\tilde{\bf v}(\bx,t;\bx_s)
\end{array}
\right]_{z=z_s}
\end{equation}

\append{Transpose of Cauchy Data to Trace operator}

For $u_0=(p_0,\bv_0) \in (C_0^{\infty}(\bR^d))^{d+1}$, define ${\bf F}u_0 = (P_r p, P_r {\bf n}_r\cdot \bv)$ where $u=(p,\bv)$ solves
\begin{eqnarray}
\label{eqn:fawedata}
\frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv, \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p ,\nonumber \\
p(\cdot,T) & =& p_0 \\ 
\bv(\cdot,T) & = & \bv_0\\
\end{eqnarray}
(Note: assumed that $\mbox{supp }\phi_r \subset \bR^d \times (0,T)$.)

Let $\tilde{\bf h}_r=(\tilde{p}_r,\tilde{v_{nr}})^T \in (C_0^{\infty}(\Gamma_r \times (0,T)))^2$ 
and 
$\tilde{u}=(\tilde{p},\tilde{\bv})$ solve 
\begin{eqnarray}
\label{eqn:faweadj}
\frac{\partial \tilde{p}}{\partial t} & = & \kappa (-\nabla \cdot \tilde{\bv} + 
 \tilde{p}_r \phi_r \delta_{\Gamma_r \times \bR})\nonumber \\
\frac{\partial \tilde{\bv}}{\partial t} & = & \beta (-\nabla \tilde{p} + \tilde{v}_{nr}{\bf n}_r \phi_r \delta_{\Gamma_r \times \bR},)\nonumber \\
\tilde{p} & =& 0 \mbox{ for } t<0\nonumber\\ 
\tilde{\bv} & = & 0 \mbox{ for } t<0 
\end{eqnarray} 
Then
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p_0 \tilde{p}}{\kappa} +  
\frac{\bv_0 \cdot \tilde{\bv}}{\beta} \right)|_{t=T}
\]
\[
= 
\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)
\]
\[
= 
\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \tilde {p} +  p \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} \right.
\]
\[
+
\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \tilde{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \tilde{\bv}}{\partial t} \right)
\]
\[
= 
\int_{0}^{T} \,dt\, \int\, dx\,dy\,dz\, \left((- \nabla \cdot \bv) \tilde{p} + p (- \nabla \cdot \tilde{\bv})\right.
\]
\[
\left.- \nabla p  \cdot \tilde{\bv} + \bv \cdot( -\nabla \tilde{p})\right)
\]
\[
+ \int_0^T \,dt \, \int_{\Gamma_r}\, dS_{\Gamma_r}( p \tilde{p}_r + \bv \cdot {\bf n}_r \tilde{v}_{nr})
\]
The first term vanishes after integration by parts. Accordingly,
\begin{equation}
\label{eqn:fadj}
 {\bf F}^T {\bf h}_r= -\tilde{u}(\cdot,T).
\end{equation}

\append{$L^2$ estimates}
This appendix lays out the proof of

\begin{theorem} 
\label{thm:one}
Suppose that $\log \kappa, \log \beta \in C^{\infty}(\bR^d) \cap L^{\infty}(\bR^d)$, and that Assumption A holds. Then the estimates \ref{eqn:coercive} hold for $h \in C^{\infty}_0(\Gamma_s \times \bR)$, hence $S[c]$ extends to $L^2(\bar{\Gamma}_s \times \bR)$ with the same estimates.
\end{theorem}

For convenience, I restate

\noindent {\bf Assumption A:} There exists $\theta_*>0$ so that for every ray connecting a point in $\mbox{supp }\phi_s$ with a point in $\mbox{supp }\phi_r$, the angles $\theta_s$ and $\theta_r$ subtended by the ray and the tangent planes of $\Gamma_s$ and $\Gamma_r$ respectively are both greater than $\theta_*$ in absolute value.

Recall that $\Gamma_r \subset \bR^d$ is a smooth embedded surface containing the receiver locations. Denote by 
${\cal I}_r$ the injection of $\Gamma_r \times \bR$ into $\bR^{d+1}$. Define the angle $\theta(\xi)$ made by a covector $\xi \in T^*(\bR^d)|_{\Gamma_r}$ with $\Gamma_r$ by 
\begin{equation}
\label{eqn:angledef}
\cos \theta(\xi) = \frac{\|  (D{\cal I}_r)^T\xi\|}{\|\xi\|}.
\end{equation}
in which $\| \cdot \|$ denotes the induced Euclidean norm.

For a covector $(\xi,\omega) \in  T^*(\bR^{d+1})$, define the angle with $\Gamma_r \times \bR$ to be the angle made by $\xi$ with $\Gamma_r$.

Choose $\theta^*>0$, and denote by $\lambda_r$ the open conic subset of $T^*(\bR^{d+1}) |_{\Gamma_r \times \bR}$ of covectors making angle $> \theta*$ with $\Gamma_r \times \bR$. The projection $\gamma_r = (D{\cal I}_r)^T\lambda_r$ consists of tangential components of these covectors; it is easy to see that $\gamma_r$ is also open.

Let $P$ denote the symbol of the acoustic system
\begin{equation}
P(\bx,\omega,\bk) = 
i\left(
\begin{array}{cc}
\frac{\omega}{\kappa(\bx)} & \bk^T\\
\bk &  \rho(\bx)\omega I_{3\times3}
\end{array}
\right)
\end{equation}
For a normal covector $\nu \in {\cal N}(\Gamma_r \times \bR)  = ((D{\cal I}_r)^T)^{-1}(0)$ and $\bk \in \gamma_r$, the polynomial
\[
\omega \mapsto \det P(\bx,\omega,\bk)
\]
has two distinct nonzero roots $\pm \omega(\bx,\bk)$.

Let $\beta_r$ be an open conic neighborhood of $\gamma_r^C$, and define $\alpha_r = ((D{\cal I}_r)^T)^{-1}\beta_r \subset T^*(\bR^{d+1}) |_{\Gamma_r \times \bR}$.

Let $H \in OPS^0(\bR^{d+1})$ be elliptic with symbol asymptotic to 1 near the characteristic variety 
\begin{equation}
\label{eqn:chardef}
C = \{(\bx,t,\bk,\omega): \det P(\bx,\bk,\omega)=0\}
\end{equation}. That is,
 $ES(H)$ is a small nbhd of characteristic variety. Note that
\[ ES(H) \cap {\cal N}(\Gamma_r \times \bR) = \emptyset\]
From the second of these facts and the definition of $\beta_r$, the Theorem
in \cite{BaoSy:91b} implies that for $\Phi \in C_0^{\infty}(\bR^{d+1})$,
\begin{equation}
\label{eqn:baosymes}
Pu=0 \mbox{ and } u \in L^2(\bR^{d+1}) \cap H^1_{ml}(ES(H) \cap \beta_r) \,\Rightarrow\, {\cal I_r}^*(\Phi u) \in L^2(\Gamma_r \times \bR)
\end{equation}

Unpack this: $u \in H^1_{ml}(ES(H) \cap \beta_r),\, s \ge 0$ means that there exists $A \in OPS^1(\bR^{d+1})$, elliptic in $ES(H) \cap \beta_r$, so that $Au \in L^2_{\rm loc}(\bR^{d+1})$. Choose $\Phi \in C_0^{\infty}(\bR^{d+1})$, that is, $\Phi Au \in L^2_{\rm loc}(\bR^{d+1})$, that is, $u \in {\rm Dom}(\Phi A)$. Denote by $\| \cdot \|_{\Phi A}$ the graph norm of $\Phi A$, and by $H^s_{\Phi A}$ the completion of $C_0^{\infty}(\bR^{d+1})$ with respect to $\| \cdot \|_{\Phi A}$, with its natural injection into $L^2(\bR^{d+1})$. Then  
\[
u \in H^s_{ml}(ES(H) \cap \beta_r) \,\Leftrightarrow \, u \in H^s_{\Phi A}
\]
for any $\Phi$, suitable choice of $A$. Further, the proof of the Theorem in \cite{BaoSy:91b} actually implies that for $\Phi$, $A$ as above with $s=1$,
\[
\|{\cal I}_r^*(\Phi u)\| \le C \|u\|_{\Phi A},
\]
That is, the localized restriction map is continuous on $H^1_{\Phi A}$.

 $\lambda_T \subset T^*(\bR^{d+1})_{t=T}$ = intersection with flow-out of $\lambda_r$, ${\cal I}_t =$ injection $\bR^d \rightarrow \bR^d \times \{t\} \subset \bR^{d+1}$, $\gamma_T = (D{\cal I})^T\lambda_T$. $\gamma_T$ = projection of flow-out to $\{t=T\}$ of lift of $\gamma_r$.

$R_T \in OPS^1(\bR^d)$ elliptic of order 1 in nbhd of $\bar{\gamma}_T^c$ so $ES(R_T)^c \subset \gamma_T$. $R \in C^{\infty}(\bR, OPS^1(\bR^d))$ flow-out of $R_T$. Write (lifted essential support) 
\[
LES(R) = \cup_{t \in \bR} ((D({\cal I}_t)^T)^{-1}ES(R_t). 
\]
\begin{lemma}
\label{thm:lem1} $(D{\cal I}_r)^T (LES(R) \cap T^*(\bR^{d+1})_{\Gamma_r \times \bR})$ contains a neighborhood of $\bar{\gamma}_r^c$.
\end{lemma}
\begin{proof}
$(\bx,t,\xi,\omega) \in T^*(\bR^{d+1})_{\Gamma_r \times \bR}, \notin LES(R)$ means that $(D({\cal I}_r)^T(\bx,t,\xi,\omega) = (\bx,\xi) \notin ES(R_t)$. Let ${\bf H}$ denote the flow. Then ${\bf H}_{T-t}(\bx,\xi) \notin ES(R_T)$ so ${\bf H}_{T-t}(\bx,\xi) = (\bx_T,\xi_T) \in \gamma_T$ so there are $\omega_T, \omega_r$ for which $(\bx_T,T,\xi_T,\omega_T) \in \lambda_T \cap \mbox{ Char }$ is on ray through $(\bx,t,\xi,\omega_r) \in \lambda_r$, whence $(D{\cal I}_r)^T (\bx,t,\xi,\omega_r) \in \gamma_r$.
\end{proof}

Then Prop. 2 in BaoSymes implies that for $ \Phi \in C_0^{\infty}(\bR^{d+1})$, $R\Phi H \in OPS^1(\bR^{d+1})$ and $ \lambda_r^C \subset ES(R\Phi H) \subset LES(R) \cap ES(H)$. Standard energy estimates imply if $R_Tu_T \in L^2(\bR^d)$ then $Ru \in L^2(\bR^{d+1})$. Also $(I-H)\Psi u \in H^m_{\rm loc}(\bR^{d+1})$ for any $m>0$ whence $R\Phi(I-H)\Psi u \in H^m_{\rm loc}(\bR^{d+1})$ for any $m \ge 0$ assume $\Psi \Phi = \Phi $. So $R\Phi Hu \in L^2(\bR^{d+1})$.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/sseprecond/plots.tex
repro docs dir added

These examples are based on two different material models. The first is a homogeneous model, with $\kappa=4.0$ GPa and $\rho=1.0$ g/cm$^{3}$. Figure \ref{fig:ptpwindh0} shows the shot gather computed in the configuration descritbed above for this choice of propagation medium. As mentioned in the preceding section, the velocity traces can be recorded in the same IWAVE workflow: these are displayed in 
\ref{fig:ptvzwindh0}.

The first group of plots show pressure traces for source position $z_s=3$ km, $x_s=3$ km, 201 receivers at $z_r=1$ km, $2 \le x_r \le 6$ km.

\plot{ptpwindh0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km. Homogeneous material model, $\kappa=4$ GPa, $\rho$=1 g/cm$^{3}$. Coarse grid: $dx=dz=20$ m, $dt=$ 8 ms, (2,4) staggered grid scheme. }
\plot{ptvzwindhh0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindh0}.}
\plot{srcvzghh0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindhh0} by time reversal (approximate inverse) algorithm explained in text.}
\plot{reptpwindhh0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzghh0}.} 
\plot{trcftrhh0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindh0}, \ref{fig:reptpwindhh0}.}.

\plot{bml0}{width=15cm}{Inhomogenous bulk modulus field, with low velocity lens.}
\plot{ptpwindl0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km. bulk modulus with low velocity lens as in Figure \ref{fig:bml0}, $\rho$=1 g/cm$^{3}$. Coarse grid: $dx=dz=20$ m, $dt=$ 8 ms, (2,4) staggered grid scheme. }
\plot{ptvzwindll0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindl0}.}

\plot{srcvzglh0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindll0} by time reversal (approximate inverse) algorithm explained in text, propagation in homogeneous material model (that is, ``wrong velocity'').}
\plot{reptpwindlh0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzglh0}, both approximate inversion and re-simulation in  homogeneous material model (``wrong velocity'').} 
\plot{trcftrlh0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindl0}, \ref{fig:reptpwindlh0}.}

\plot{srcvzgll0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindll0} by time reversal (approximate inverse) algorithm explained in text, propagation in bulk modulus of Figure \ref{fig:bml0} (that is, ``correct velocity'').}
\plot{reptpwindll0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzgll0}, both approximate inversion and re-simulation in bulk modulus of Figure \ref{fig:bml0} (``correct velocity'').} 
\plot{trcftrll0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindl0}, \ref{fig:reptpwindll0}.}

### 2019-10-19 repro/papers/sserefl/paper.tex
repro docs dir added

\title{Surface Source Extension, Reflection Case}
\author{William. W. Symes \thanks{The Rice Inversion Project,
Department of Computational and Applied Mathematics, Rice University,
Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}

\lefthead{Symes}

\righthead{Reflection SSE}

\maketitle
\begin{abstract}
FWI via surface source extension (SSE) applies equally well to reflection or transmission data. In the transmission case, it is asymptotically equivalent to a form of travel time tomography, fitting event slowness rather than arrival time. Use of this method for reflection data simply involves permitting reflectors in the velocity and other subsurface mechanical parameters. It is possible to analyze a simple layered medium special case in the plane wave domain. The analysis suggests that in fact reflection SSE cannot constructively update velocity in absence of correctly positioned reflectors, so more closely resembles so-called Reflection FWI than transmission SSE in its velocity updating capabilities. This theoretical result is supported by numerical experiments with 2D reflection data.
\end{abstract}

\section{Introduction}
FWI based on source extension has been around since the early 1990's,
and some versions are provably robust - converge from substantially
incorrect initial velocity models. However the theoretical robustness
results all pertain to purely transmission configurations, such as
crosswell or diving wave geometry. It has been an open question
whether source extended FWI remains robust if all kinematic
information in the data is in the form of reflection
moveout. Some numerical evidence for such robustness has appeared, but
in each case it is not obvious that post-critical events did not play
a role in the inversion.

The goal of this paper is to produce a proof that in one precisely
defined case and in a precisely defined sense, one particular variant
of source-extended FWI is {\em not} robust for inversion of reflection
data. This counterexample negates any general claim of robustness for
such methods of reflection inversion, of course.

Surface source extension, defined below, is the method investigated
here. The objective of the reasoning laid out in the following sections is
to explicitly calculate the gradient of a particular surface source
extended FWI objective. A context that permits such explicit
calculation is plane-wave propagation in a layered medium. An explicit
computation of the gradient at a homogeneous background medium with
data containing a single reflector shows that the gradient has
vanishing low-frequency components as the central data frequency goes
to infinity: in fact, the mean square in any fixed frequency band goes
to zero. Therefore a step in the negative gradient direction cannot
affect traveltimes significantly and therefore this technique cannot
produce a productive velocity update in this case.

It is important to note the contrast with other extension methods. The
surface source extended FWI objective in the closest analogue for
transmission, namely the estimation of a homogeneous velocity from one
or more transmission traces, has been shown to be locally convex
independent of data frequency content. Also, the subsurface offset
extended FWI objective for surface reflection data has been shown to
have similar properties.

In the following sections, I will first give a more precise statement
of the main result. Then I will review plane-wave modeling, and construct
an asymptotic approximation to single-reflector plane wave data. In
the final section I will present the surface source extended FWI
objective and compute its gradient at a homogeneous background with
single reflector data. The main result will follow immediately from
this calculation.

I include several appendices, on 1D wave propagation and plane wave
reflection modeling, as this material is difficult to find in
sufficiently precise form.

Whenever convenient, I abbreviate ``surface source extended FWI'' as SSE-FWI.

\section{Main Result}

We adopt the usual pressure-velocity acoustics system as the
underlying physics of waves, see beginning of next section. The
surface source extension allows the source to extend over the surface,
in the form $w(x,t)\delta(z)$ for a nominal source position of
$(0,0)$. After Radon transform, this source takes the form
$\tw(\xi,t)\delta(z)$, where $\xi$ is the plane wave slowness.

The underlying hypothesis is that the data comes from a point
source. Extension methods encode such information in an annihilator,
an operator that vanishes for models that satisfy the hypothesis. For
SSE-FWI, a natural annilator is a variant on the operator introduced
for this purpose by \cite{HuangSymes:Geo18a,HuangSymes:Geo18b} for two
similar source extension methods, namely multiplication by offset,
also reminiscent of a now-standard annihilator for subsurface offset
introduced by \cite{stolk2001c}. In the plane wave domain,
multiplication by offset becomes 
\[
\tA = D_{\xi}I_t,
\]
that is, differentiation in $\xi$ and integration in $t$, in either
order.

The forward map $\tS$ for plane-wave transformed inverse problem maps
extended plane wave source wavelets $\tw(\xi,t)$ to plane wave data
traces (sampling of the plane-wave field at $z=0$, assuming this to be the
depth of sources and receivers, as I do here). I will regard this operator as a
function of the bulk modulus $\kappa(z)$, by assuming the density
$\rho$ to be fixed and homogeneous.

$\tS[\kappa]$ is 
actually invertible, at least for $\kappa$ close to homogeneous. This conclusion follows for instance
from explicit expressions in two special cases: for homogeneous models
$\kappa=\kappa_0$, 
\[
\tS[\kappa_0]\tw(\xi,t) = a_0(\xi)\tw(\xi,t),
\]
and for a smooth model $\kappa_1$ with an embedded reflector at $z=z_1$,
\[
\tS[\kappa_1](\xi,t) = a_0(\xi)\wl(t) +
R(\xi)\wl\left(t-2\tau(\xi,z_1)\right) + O(\lambda)
\]
Here the source is non-extended: $\tw(\xi,t)=\wl(t)$, and 
\[
\wl(t)=\frac{1}{\sqrt{\lambda}}w_1\left(\frac{t}{\lambda}\right)
\]
is a family of wavelets depending on central wavelength $\lambda$. The
vertical travel time $\tau(\xi,z_1)$ is given by
\[
\tau(\xi,z_1) = \int_0^{z_1}\sqrt{\frac{1}{c^2}-\xi^2},\,\,c=\sqrt{\kappa/rho}.
\]

Since $\tS$ is invertible, it is possible to use the
simplified objective presented by
\cite{Symes:EAGE15}:
\begin{equation}
\label{eqn:dsobj}
J_0[\kappa]=\frac{1}{2}\|\tA \tS[\kappa]^{-1}\td\|^2
\end{equation}
A derivation of this limit case from the standard penalty formulation
of extended FWI 
appears in the last section. Inversion based on this type of objective
has been called ``inversion velocity analysis'' in the subsurface
offset context \cite[]{Herve2017},\cite[]{HouSymes:Geo18}, and is also the
basis of Adaptive Waveform Inversion
\cite{Warner:14,Warner:16}.

The main result pertains to the gradient at homogeneous
$\kappa=\kappa_0$ for data resulting from a single flat reflector
embedded in a smooth background:
\[
\tdl = \tS[\kappa_1]\wl
\]
I calculate the gradient of $J_0$ in this configuration.
The result is given in equation \ref{eqn:jgrad} in the last
section. Converted to spatial frequency via the Fourier transform, it
becomes
\begin{equation}
\label{eqn:jgradft-intro}
\widehat{\nabla J[\kappa_0]}(k) = -\frac{c_0}{16}\int\, d\xi\,a_1(\xi) 
\frac{1}{ik}\hat{u}_{\lambda}(\xi,k)D_{\xi}^2\overline{\hat{u}_{\lambda}(\xi,k)}
\end{equation}
where
\[
\hat{u}_{\lambda}(\xi,k) = \left(1 +
  \frac{R(\xi)}{a_0(\xi)}e^{4ik\tau(\xi,z_1)/\tc(\xi)}\right)\hat{\wl}\left(\frac{2k}{\tc(\xi)}\right) 
\]
and 
\[
\tc(\xi) = \left(\frac{1}{c_0}^2-\xi^2\right)^{-1/2}
\]
is the vertical plane wave velocity for homogenous wave velocity
$c_0 = \sqrt{\kappa_0/rho}$.

As is reasonable, even mandatory, for simulation of seismic data,
assume that the ``mother wavelet'' $w_1$ has at least two vanishing
moments. Then for any $k_{\rm max} >0$, $\xi_{\rm max} > 0$,
\[
\max \{\left|\hat{w}_{\lambda}\left(2k/\tc{\xi}\right)\right|:
|k|\le k_{\rm max}, |\xi|\le \xi_{\rm max} \} \rightarrow 0\mbox{ as }
\lambda \rightarrow 0.
\]
It follows immediately from equation \ref{eqn:jgradft-intro} that the
same is true of the gradient:
\begin{equation}
\label{eqn:bigdeal}
\max \{\left|\widehat{\nabla J[\kappa_0]}(k)\right|:
|k|\le k_{\rm max}, |\xi|\le \xi_{\rm max} \} \rightarrow 0\mbox{ as }
\lambda \rightarrow 0.
\end{equation}

Therefore the rate at which traveltime changes for any plane wave
component in the specified range, as the bulk modulus is changed along
the direction of the negative gradient, is vanishingly small as
$\lambda \rightarrow 0$.

Otherwise put, for high-frequency reflection data, this method cannot
update the velocity away from a homogenous initial guess in a way that
is kinematically constructive.

It may even be possible to show that the update may be destructive, by
more careful examination of the gradient approximation. However, who
cares, really: it is established that SSE-FWI has similar behaviour to
FWI for at least one simple instance of reflection data.

\section{Plane Wave Modeling}
Start with 2D acoustics in velocity-pressure form: $p$ is excess
pressure, $\bv=(v_x,v_z)$ is particle velocity about equilibrium, $f$
is a (possibly) extended source model, all linked by the acoustic wave system
\begin{eqnarray}
\label{eqn:awe}
\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
                                                                f(\bx,t;\bx_s) \nonumber\\
\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
p,{\bf v}&=&0, t \ll 0
\end{eqnarray}

As in other work on this topic, I will assume $\rho$ to be fixed, and
for sake of simplicity constant, and focus the discussion on
$\kappa$. Thus the modeled pressure traces at receiver locations
depend on $\kappa$ and the source function $f$. For simplicity again, assume that
receivers are located at every point of the hyperplane $z=z_r$. Then
the modeling operator is defined as
\begin{equation}
\label{eqn:fwd}
S[\kappa]f = p|_{z=z_r}
\end{equation}

Assume that sources all lie on the surface $z=z_s$, so that surface
extended sources take the form $f(\bx,t;\bx_s) =
w(x,t)\delta(z-z_s)$. By abuse of notation, regard $S[\kappa]$ as
acting on $w$.

Assume that $f$ is square-integrable, $\kappa = \kappa(z)$,
$\rho=\rho(z)$ (layered medium). Then the modeled shots differ from
each other by a horizontal translation, so that there is in effect
only one shot, for which I take the location to be
$\bx_s=(0,0,z_s)$ and suppress $\bx_s$ from the notation.

To make things even simpler, set $z_s=z_r=0$. Permitting arbitrary
values just makes the arithmetic more complicated, without affecting
the essential results.

 If $\log \kappa$ and $\log \rho$ are globally
bounded, there is $\xi_{\rm max} > 0$ so that for $\|\xi\|<\xi_{\rm
  max}$, the integrals defining slant stack fields
$\tilde{p},\tilde{\bv}$, namely
\begin{eqnarray}
\label{eqn:slant}
\tp(\xi,z,t)& = &\int dx p(x,z,t+\xi x),\nonumber\\
\tv(\xi,z,t)&=&\int dx \bv(x,z,t+\xi x),\nonumber\\
\tf(\xi,z,t)&=&\int dx f(x,z,t+\xi x),
\end{eqnarray}
are absolutely convergent, and $\tp, \tv$ satisfy 
\begin{eqnarray}
\label{eqn:awepw}
(1-c^2\xi^2)\frac{\partial \tp}{\partial t} + \kappa\frac{\partial
  \tv}{\partial z} &=& \tf \nonumber\\
\rho \frac{\partial \tv}{\partial t} + \frac{\partial \tp}{\partial
  z}&=&0\nonumber\\
\tp,\tv&=&0, t \ll 0.
\end{eqnarray}
Here $c=\sqrt{\kappa/\rho}$ is the sound velocity. I use $\xi$ instead
of the more usual $p$ for plane wave slowness since $p$ already has
been assigned to mean pressure. Note that if
$f(\bx,t;\bx_s)=w(t)\delta(\bx)$, then
$\tf(\xi,z,t)=w(t)\delta(z)$. Likewise for a surface extended source,
$f(\bx,t;\bx_s) = w(x,t)\delta(z)$, $\tf(\xi,z,t) = \tw(\xi,t)\delta(z)$.

From now on assume a surface extended source, so in plane wave domain
$\tf(\xi,z,t) = \tw(\xi,t)\delta(z)$. The plane wave domain modeling
operator samples $\tp$ on $z=z_r=0$:
\begin{equation}
\label{eqn:modpw}
\tS[\kappa]\tw = \tp|_{z=0}
\end{equation}

For homogeneous $\kappa=\kappa_0, \rho=\rho_0, c=c_0$, a computation
recapitulated in the Appendix C, relying on computations for the 1D
acoustics system detailed in Appendices A and B, yields
\begin{equation}
\label{eqn:modhom}
\tS[\kappa_0]\tw(\xi,t) = \frac{1}{2(1-c_0^2\xi^2)}\tw(\xi,t).
\end{equation}
That is, at a homogeneous model, the plane wave modeling operator is
simply scaling by a $\xi$-dependent factor.

The derivative of the plane wave modeling operator $\tS$ with respect
to $\kappa$ is the $z=0$ trace of the pressure perturbation field
$\delta \tp$, a solution component of the perturbational plane wave
system
\begin{eqnarray}
\label{eqn:dawepw}
(1-c^2\xi^2)\frac{\partial \delta \tp}{\partial t} -
  \delta \kappa \frac{\xi^2}{\rho}\frac{\partial \tp}{\partial t}
  \nonumber\\
+\kappa\frac{\partial
  \delta \tv}{\partial z} + \delta \kappa\frac{\partial
  \tv}{\partial z} &=& 0 \nonumber\\
\rho \frac{\partial \delta \tv}{\partial t} + \frac{\partial \delta \tp}{\partial
  z}&=&0\nonumber\\
\delta \tp,\delta \tv&=&0, t \ll 0.
\end{eqnarray}
The derivative of the modeling operator $\tS$ is simply the trace at
$z=0$ of the plane wave pressure field $\tp$:
\begin{equation}
\label{eqn:dmodpw}
D(\tS[\kappa]\tw)\delta \kappa = \delta \tp|_{z=0}
\end{equation}
For homogeneous background model $\kappa=\kappa_0$, the appendices
explain how to develop an explicit expression for this operator: one
obtains
\begin{equation}
\label{eqn:dmodhom}
D(\tS[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
\frac{c_0}{16\kappa_0 (1-c_0^2\xi^2)^{5/2}}\int \,ds\,\tw(\xi,t-s)(\delta \kappa)'  \left(\frac{c_0
      s}{2\sqrt{1-c_0^2\xi^2}}\right)
\end{equation}
To compute the gradient of the SSE objective function, the transpose
of this operator (as a linear map on $\delta \kappa$) is required:
that is
\begin{equation}
\label{eqn:dmodhomadj}
D(\tS[\kappa]\tw)_{\kappa=\kappa_0}^T\td(z) =
- 
\frac{1}{8} \int \,d\xi\,\frac{1}{\kappa_0 (1-c_0^2\xi^2)^2}\frac{\partial}{\partial z} \int \,dt\,
  w\left(\xi,t-\frac{2z\sqrt{1-c_0^2\xi^2}}{c_0}\right)d(t).
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Simple Reflection Data}

Inclusion of reflections simply means permitting $\kappa$ to be
non-smooth. The simplest example is the layer-over-half-space. The
data to be described in this section belongs to a model having the
same properties at the measurement surface $z=0$ as the homogeneous
reference model: that is, $\kappa(0)=\kappa_0$. So to introduce a
reflector with different moveout than that implied by the reference
model, it is necessary to allow $\kappa$ to be non-constant between
the surface an the reflector depth, say $z_1 > 0$.

Suppose that $\kappa_1$ is smooth in the intervals $[0,z_1)$ and
$(z_1,\infty)$, but undergoes a nonzero jump at $z=z_1$, 
\[
[\kappa_1]_{z=z_1} \ne 0.
\]
The proper notion of solution for wave equations such as equation
\ref{eqn:awe} or \ref{eqn:awepw} with discontinuous coefficients such
as $\kappa_1$ is that of {\em weak solution}, see
\cite{BlazekStolkSymes:13}. For the point source case, $\tw(\xi,t) =
w(t)$, with smooth $w$, a weak solution can be constructed that is
smooth in $z \ne z_1$: this construction is the familiar combination
of reflected and transmitted waves. Since causal weak solutions are
unique, this is {\em the} solution. The smooth fields on either side
of the reflector at $z=z_1$ can approximated via geometric optics,
leading to an asymptotic
approximation to the solution. To express the asymptotic nature of
this approximation, consider a family of source wavelets, rather than
a single wavelet
\[
\wl = \frac{1}{\sqrt{\lambda}}w_1\left(\frac{t}{\lambda}\right)
\]
where $w_1\in C_0^{\infty}(\bR)$ has zero mean. Then [REFERENCE - WHERE?]

The pressure field at the receiver 
location $z_r=0$ takes the familiar form  
\begin{equation}
\label{eqn:modlohs}
\tS[\kappa_1]\wl(\xi,t) = \frac{1}{2c_0\sqrt{1-c_0^2\xi^2}}\wl(t) +
R(\xi)\wl\left(t-2\tau(\xi,z_1)\right) + O(\lambda)
\end{equation}
where $\tau$ is the vertical plane wave travel time:
\begin{equation}
\label{eqn:pwtt}
\tau(\xi,z) = \int_0^{z} \sqrt{\frac{1}{c(z)^2}-\xi^2}.
\end{equation}
The reflection coefficient $R$ depends on $\xi$ and on
the values of $\kappa_1$ and $\rho$ near $z=z_1$: the precise form is not
important for the argument to follow.

The remainder in \ref{eqn:modlohs} is $O(\lambda)$ in the sense of
$L^2$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{SSE Objective and Gradient: vanishing low frequencies}
In this section I will formulate the SSE objective, and calculate its
gradient at homogeneous background for single reflector data as
described in the last section. I will show that this gradient has
low-frequency content as the data wavelength $\lambda$, introduced in
the last section and in the introduction, tends to zero.

FWI via SSE requires specification of an {\em annihilator}, the null
space of which consists of non-extended models, as explained in the
introduction.  Construct a suitable plane-wave domain annihilator
$\tA$ for
SSE by intertwining the Radon transform $R$ with the space-domain
annihilator $A = $ multiplication by offset. In order to do this
properly, it is necessary to introduce a cutoff in offset: not a
hardship, since infinite offsets are unavailable. Denote by $\chi$ a
offset mute. Define
\[
Rw(\chi,t) = \int \,dx\,\chi(x) w(x,t+\xi x)
\]
\[
Aw(x,t)=x w(x,t)
\]
\[
\tilde{A} Rw (\xi,t) = R Aw (\xi,t)
\]
\[
=\int dx x \chi(x) w(x,t+\xi x) = \int dx \chi(x)\frac{\partial}{\partial \xi}
I_tw(x,t+\xi x)
\]
\[
= D_{\xi} I_t R  w \equiv \tilde{A} Rw (\xi,t) 
\]

Since $\tS[\kappa]$ is continuous in $\kappa$
\cite[]{BlazekStolkSymes:13}, and $\tS[\kappa_0]$ is a multiple of the
identity, $\tS[\kappa]$ is invertible in sup-norm neighborhood of
$\kappa$. Introduce the the scaled reduced penalty function objective
\[
\kappa \mapsto \min_{\tw}(\frac{1}{\alpha^2}\|\tS[\kappa]\tw-\td\|^2 + \|\tA
\tw\|^2)
\]
\[
=\frac{1}{\alpha^2}\|\tS[\kappa]\tw_{\alpha}[\kappa]-\td\|^2 + \|\tA
\tw_{\alpha}[\kappa]\|^2
\]
where $\tw_{\alpha}[\kappa]$ solves the normal equation:
\[
\tS[\kappa]^T\tS[\kappa]\tw  + \alpha^2\tA^T\tA\tw = \tS[\kappa]^T\td.
\]
Since $\tS[\kappa]$ is invertible, $\tw_{\alpha}[\kappa]$ is uniformly
bounded as $\alpha \rightarrow 0$. Let $\tw_0[\kappa] =
\tS[\kappa]^{-1}\td$. Then
\[
\tS[\kappa]^T\tS[\kappa](\tw_{\alpha}[\kappa]-\tw_0[\kappa])+\alpha^2\tA^T\tA\tw_{\alpha}[\kappa]
= 0,
\]
so $\tw_{\alpha}[\kappa] \rightarrow \tw_0[\kappa]$ as $\alpha
\rightarrow 0$. Consequently the scaled objective above becomes
\[
\frac{1}{\alpha^2}\|\tS[\kappa]\tw_{\alpha}[\kappa]-\td\|^2 + \|\tA 
\tw_{\alpha}[\kappa]\|^2
=\frac{1}{\alpha^2}\|\tS[\kappa](\tw_{\alpha}[\kappa]-\tw_0[\kappa])\|^2 + \|\tA 
\tw_{\alpha}[\kappa]\|^2
\]
\[
=\frac{1}{\alpha^2}\langle (\tw_{\alpha}[\kappa]-\tw_0[\kappa]),
\tS[\kappa]^T\tS[\kappa](\tw_{\alpha}[\kappa]-\tw_0[\kappa])\rangle+ \|\tA 
\tw_{\alpha}[\kappa]\|^2
\]
\[
=\frac{1}{\alpha^2}\langle (\tw_{\alpha}[\kappa]-\tw_0[\kappa]),
-\alpha^2\tA^T\tA \tw_{\alpha}[\kappa] \rangle + \|\tA\tw_{\alpha}[\kappa]\|^2
=\langle (\tw_{\alpha}[\kappa]-\tw_0[\kappa]),
-\tA^T\tA \tw_{\alpha}[\kappa] \rangle + \|\tA\tw_{\alpha}[\kappa]\|^2
\]
\begin{equation}
\label{eqn:j0def}
\rightarrow \|\tA\tw_0[\kappa]\|^2 \equiv 2 J_0[\kappa,\td].
\end{equation}
This $\alpha \rightarrow 0$ limit is the simplified objective used by
\cite{Symes:EAGE15} and mentioned in the introduction.

%\[
%DJ[\kappa_0]\delta \kappa = -\langle D(\tS[\kappa_0]\tw)\delta \kappa,
%\tA^T\tA S[\kappa_0]^{-1}\td \rangle , \,\tw=\tS[\kappa_0]^{-1}\td
%\]
Now assume that $\td$ is data computed from a simple reflector as
described in the previous section, depending on the wavelength
parameter $\lambda$:
\[
\td_{\lambda}(\xi,t) = a_0(\xi)\wl(t) +
R(\xi)\wl\left(t-2\tau(\xi,z_1)\right) + O(\lambda)
\]
Then
\begin{equation}
\label{eqn:wtildedef}
\tS[\kappa_0]^{-1}\td_{\lambda}(\xi,t) = \wl(t) +
\frac{R(\xi)}{a_0(\xi)}\wl\left(t-2\tau(\xi,z_1)\right) + O(\lambda)
\equiv \tw(\xi,t) + O(\lambda)
\end{equation}
Write 
\[
a_1(\xi)=\frac{1}{\kappa_0 (1-c_0^2\xi^2)^2}, \tc_0(\xi) =
\frac{c_0}{\sqrt{1-c_0^2\xi^2}}
\]
%Then
%\[
%DJ[\kappa_0]\delta \kappa =
%\left\langle \frac{c_0}{16\kappa_0 (1-c_0^2\xi^2)^{5/2}}\int
%\,ds\,\tw(\xi,t-s)(\delta \kappa)'
%\left(\frac{c_0s}{2\sqrt{1-c_0^2\xi^2}}\right), D_{\xi}^2I_t^2
%\tw(\xi,t) \right\rangle
%\]
%Per the derivation of simple reflection, presume that (1) $w_1(t) = 0$
%for $t>1$, and $w_1$ (hence $\wl$) has two vanishing moments, and (2) $\delta \kappa = 0 $ in an interval of positive
%length about $z=0$, and (3) $\lambda$ is small enough that $\wl(t)=0$
%if $\delta(\kappa)(t\tc/2) \ne 0$ for all $\xi$ for which $\chi(\xi)
%\ne 0$ (I.e. the range of the $\xi$ integral).

%Then the convolution of $\delta \kappa(t\tc/2)$ and $\wl$ vanishes for
%small $t$, hence the product with $\wl(t)$

%and

From the definition \ref{eqn:j0def} of $J_0$ and a simple calculation,
\[
\nabla J[\kappa_0,\td] = (D\tS[\kappa_0]\tw)^T(\tS[\kappa_0]^T)^{-1}\tA^T\tA
\tS[\kappa_0]^{-1}\td,  \,\tw=\tS[\kappa_0]^{-1}\td
\]
Since $D_{\xi}$ and $I_t$ are both skew-symmetric and commute, $\tA$
is symmetric. $\tS$ is $\xi-$ dependent scaling. For simplicity
abbreviated the relation \ref{eqn:modhom} as
\[
\tS[\kappa_0]\tw(\xi,t) = a_0(\xi)\tw(\xi,t)
\]
Then
\[
\nabla J[\kappa_0](z) =
- 
\frac{1}{8} \int \,d\xi\,a_1(\xi)\frac{\partial}{\partial z} \int \,dt\, 
  \tw\left(\xi,t-\frac{2z}{\tc_0}\right)\left(D_{\xi}^2I_t^2 \tw(\xi,t)\right)
\]

\begin{equation}
\label{eqn:jgrad}
=-\frac{c_0}{16} \int \,d\xi\,a_1(\xi)\frac{\partial}{\partial z} \int \,dz'\, 
  \tw\left(\xi,\frac{2(z'-z)}{\tc_0}\right)\left(D_{\xi}^2I_z^2
    \tw\left(\xi,\frac{2z'}{\tc_0}\right)\right)
\end{equation} 
Set $u(\xi,z) = \tw(\xi,2z/\tc)$. Denote by $\hat{f}$ the Fourier
transform of $f$.  
\begin{equation}
\label{eqn:jgradft}
\widehat{\nabla J[\kappa_0]}(k) = -\frac{c_0}{16}\int\, d\xi\,a_1(\xi) 
\frac{1}{ik}\hat{u}(\xi,k)D_{\xi}^2\overline{\hat{u}(\xi,k)}
\end{equation}
From the definition \ref{eqn:wtildedef} of $\tw$,
\[
\hat{\tw}(\xi,\omega) = \left(1 + \frac{R(\xi)}{a_0(\xi)}e^{2i\omega\tau(\xi,z_1)}\right)\hat{\wl}(\omega)
\]
whence
\[
\hat{u}(\xi,k) = \left(1 +
  \frac{R(\xi)}{a_0(\xi)}e^{4ik\tau(\xi,z_1)/\tc(\xi)}\right)\hat{\wl}\left(\frac{2k}{\tc(\xi)}\right) 
\]
Together with the expression \ref{eqn:jgradft} for the Fourier
transform of the gradient, this observation establishes the assertion
stated in the introduction.

\append{1D Wave Motion}
Begin with the 1D acoustics point source system. 
\begin{eqnarray}
\label{eqn:awe1d}
\frac{\partial p}{\partial t} +\kappa\frac{\partial 
  v}{\partial z} &=& w(t)\delta(z-z_s) \nonumber\\
\rho \frac{\partial v}{\partial t} + \frac{\partial p}{\partial 
  z}&=&0\nonumber\\
 p,v&=&0, t \ll 0. 
\end{eqnarray}
Since the right hand side is singular, so is the solution, so it must
be a solution in the weak sense; since it is causal, it is uniquely
determined by the system above. It follows from the weak solution
conditions that the pressure is continuous at $p=0$, whence $v$ must
have a discontinuity.

Now assume that $\kappa, \rho$ are constant. 
In $z \ne 0$, the right hand side 
vanishes, so the solution must be locally a combination of plane
waves; causality implies that
\[
p(z,t)=a\left(t \mp \frac{z}{c}\right), \, v(z,t)=\pm b\left(t \mp
  \frac{z}{c}\right)
\]
From the second dynamical equation (Newton's law) it follows that $b =
a/(\rho c)$. The singularity on the LHS of the first dynamical
equation (constitutive law) is
\[
\kappa [v]_{z=0}\delta(z) =
2\frac{\kappa}{c}b\delta(z) = 2a\delta(z).
\] 
This must in turn equal the RHS of the constitutive law, whence
$a=w/2$. Thus
\begin{eqnarray}
\label{eqn:sol1d}
p(z,t) &=& \frac{1}{2}w\left(t \mp \frac{z}{c}\right) \nonumber \\
v(z,t) &=& \pm\frac{1}{2\rho c}w\left(t \mp \frac{z}{c}\right)
           \nonumber \\
\pm z &>& 0.
\end{eqnarray}

\append{1D Born Approximation}

The perturbational system for \ref{eqn:awe1d} is:
\begin{eqnarray}
\label{eqn:dawe1d}
\frac{\partial \delta p}{\partial t} +\kappa\frac{\partial
  \delta v}{\partial z} + \delta \kappa\frac{\partial
  v}{\partial z} &=& 0 \nonumber\\
\rho \frac{\partial \delta v}{\partial t} + \frac{\partial \delta p}{\partial
  z}&=&0\nonumber\\
\delta p,\delta v&=&0, t \ll 0.
\end{eqnarray}

If $\kappa, \rho $ are constant, it is possible to develop explicit
expressions for the solution of this system. Assuming that $\delta
\rho=0$ and $\delta \kappa$ is nonzero only in $z>0$, the reflected
(perturbational) wavefield should be upcoming (i.e. a function of
$z+ct$, away from the wavefront), as the wave scatters
only once, at the wavefront implicit in the expressions
\ref{eqn:sol1d}. On the other hand, the field must vanish before the
first arrival.

The key is to understand the perturbational impulse response. That is,
set $w(t)=H(t)$, and adopt adopt the anzatz
\begin{eqnarray}
\label{eqn:ansatz1d}
\delta p(z,t) &=& a_0\left(t + \frac{z}{c}\right) \delta\left(t -
  \frac{z}{c}\right) + a_1\left(t + \frac{z}{c}\right) H\left(t -
  \frac{z}{c}\right),\nonumber\\
\delta v(z,t) &=& b_0\left(t + \frac{z}{c}\right) \delta\left(t -
  \frac{z}{c}\right) + b_1\left(t + \frac{z}{c}\right) H\left(t -
  \frac{z}{c}\right),
\end{eqnarray}
Assume that $w$ is smooth and
of compact support, with zero mean. Then all derivatives of $w$ are
necessarily linearly independent, as any dependence would amount to a
differential equation with constant coefficients, the solution of
which would be entire analytic hence not of compact support.

In the following calculations, the arguments of the coefficients
$a_0,a_1,b_0,$ and $b_1$ are $\left(t + \frac{z}{c}\right)$ in all
cases, and are suppressed for the sake of brevity, until the final result.

Inserting the ansatz \ref{eqn:ansatz1d} in the second equation in \ref{eqn:dawe1d}, obtain
\[
0=\rho \frac{\partial \delta v}{\partial t} + \frac{\partial \delta p}{\partial
  z} = 
\]
\[
\left(\rho b_0 - \frac{a_0}{c}\right) \delta'\left(t -
  \frac{z}{c}\right)
\]
\[
+\left(\rho b_0' + \frac{a_0'}{c}+\rho b_1 - \frac{a_1}{c} \right)\left(t + \frac{z}{c}\right) \delta\left(t -
  \frac{z}{c}\right)
\]
\[
 + \left(\rho b_1' + \frac{a_1'}{c}\right) \left(t + \frac{z}{c}\right) H\left(t -
  \frac{z}{c}\right)
\]
Since $\delta', \delta$ and $H$ are linearly independent, this equation is
equivalent to the system
\begin{eqnarray}
\label{eqn:asnbs}
\rho b_0 - \frac{a_0}{c} &=& 0\nonumber\\
\rho b_0' + \frac{a_0'}{c}+\rho b_1 - \frac{a_1}{c} &=&0 \nonumber\\
\rho b_1' + \frac{a_1'}{c} &=&0.
\end{eqnarray}
The first and third equations imply that 
\begin{eqnarray}
\label{eqn:handy}
\rho c b_0&=&a_0,\nonumber\\
\rho c b_1&=&-a_1,
\end{eqnarray}
 so the second equation becomes
\begin{equation}
\label{eqn:as}
a_0' = a_1
\end{equation}
Now insert the asatz \ref{eqn:ansatz1d} into the first of the
dynamical equations \ref{eqn:dawe1d}. Note that \ref{eqn:sol1d}
implies
\[
\delta \kappa \frac{\partial v}{\partial z} = - \frac{\delta
  \kappa}{2\kappa}\delta\left(t - \frac{z}{c}\right)
\]
so that
\[
0 = \left(\frac{\partial \delta p}{\partial t} +\kappa\frac{\partial
  \delta v}{\partial z} + \delta \kappa\frac{\partial
  v}{\partial z}\right)(z,t)=
\]
\[
\left(a_0 - b_0\frac{\kappa}{c}\right)\delta'\left(t - \frac{z}{c}\right) +
\]
\[
\left(a_0' + b_0'\frac{\kappa}{c} + a_1-b_1\frac{\kappa}{c}\right)\delta\left(t -
  \frac{z}{c}\right)+
\]
\[
\left(a_1' + b_1'\frac{\kappa}{c}\right) H\left(t - \frac{z}{c}\right)
\]
\[
- \frac{\delta \kappa}{2\kappa}\delta\left(t - \frac{z}{c}\right)
\]
The first and last equations from \ref{eqn:handy} imply that the first
and third terms in the sum above vanish, and that the second and fourth
amount to
\[
=\left(2 (a_0'+a_1) -\frac{\delta \kappa}{2\kappa}\right) \delta\left(t -
  \frac{z}{c}\right)
\]
Taking the identity \ref{eqn:as} into account and re-inserting the
argument of $a_1$, the upshot is
\[
4a_1\left(t + \frac{z}{c}\right) \delta\left(t -
  \frac{z}{c}\right) = \frac{\delta \kappa(z)}{2\kappa} \delta\left(t
  -  \frac{z}{c}\right)
\]
In view of the delta function factor, $z=ct$ in the arguments of $a_1$
and $\delta \kappa $, so the above equation is equivalent to
\begin{equation}
\label{eqn:asol}
a_1(t)=\frac{\delta \kappa\left(\frac{ct}{2}\right)}{8\kappa}
\end{equation}
It remains to determine the integration constant for $a_0$. In view of
the condition, mentioned above, that $\delta \kappa$ vanish in $z \le
0$, it follows from \ref{eqn:dawe1d} that $\delta p(0,0)=0$, so
$a_0(0)=0$. Consequently the ansatz \ref{eqn:ansatz1d} is validated,
and 
\begin{equation}
\label{eqn:imp1d}
D(S[\kappa]H)\delta \kappa (t) = \delta p(0,t)=  \frac{\delta 
  \kappa\left(\frac{ct}{2}\right)}{8\kappa} 
\end{equation}
whence for arbitrary causal wavelet $w$,
\begin{equation}
\label{eqn:fwd1d}
D(S[\kappa]w)\delta \kappa (t) = \int_0^t\,ds\,w'(t-s)\frac{\delta 
  \kappa\left(\frac{cs}{2}\right)}{8\kappa} 
\end{equation}

\append{Plane Wave Modeling of Reflection Traces}
Re-write the plane wave system \ref{eqn:awepw} as
\begin{eqnarray}
\label{eqn:awepw1}
\frac{\partial \tp}{\partial t} + \tk\frac{\partial
  \tv}{\partial z} &=& (1-c^2\xi^2)^{-1}\tf \nonumber\\
\rho \frac{\partial \tv}{\partial t} + \frac{\partial \tp}{\partial
  z}&=&0\nonumber\\
\tp,\tv&=&0, t \ll 0,
\end{eqnarray}
in which $\tk(\xi,z)=(1-c(z)^2\xi^2)^{-1}\kappa(z)$ is a scaled version of
bulk modulus. Similarly define the vertical plane wave velocity as
\[
\tc(\xi,z)=\frac{c(z)}{1-c(z)^2\xi^2} =
\left(\frac{1}{c(z)^2}-\xi^2\right)^{-1/2}.
\]
If $\tf(\xi,z,t)=\tw(\xi,t)\delta(z)$ is a surface
extended source, plane wave version, this is equivalent to replacing
$\tw(\xi,t)$ by $(1-c(0)^2\xi^2)^{-1}\tw(\xi,t)$.

Since the system \ref{eqn:awepw1} has the form of the 1D system
\ref{eqn:awe1d} for each $\xi $, both $\tS[\kappa_0] $ and
$D\tS[\kappa_0]$ can be computed from the corresponding 1D
results. From \ref{eqn:sol1d},
\[
\tS[\kappa_0]\tw(\xi,t)=\frac{1}{2}(1-c_0^2\xi^2)^{-1}\tw(\xi,t)
\]
which is equation \ref{eqn:modhom}.

From \ref{eqn:fwd1d},
\[
D(\tS[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa = \int_0^t\,ds\,(1-c_0^2\xi^2)\tw'(\xi,t-s)\frac{\delta 
  \tk \left(\frac{\tc s}{2}\right)}{8\tk} 
\]
Note that 
\[
\delta \tk = \delta \left(\frac{\kappa}{1-\kappa \xi^2/\rho}\right) = 
\]
\[
\delta \kappa (1-c^2\xi^2)^{-2}
\]
Consequently 
\[
D(\tS[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
\int_0^t\,ds\,\tw'(\xi,t-s)\frac{\delta \kappa  \left(\frac{\tc 
      s}{2}\right)}{8\kappa_0 (1-c_0^2\xi^2)^2} 
\]
\[
= \frac{c_0}{2}\int_0^t\,ds\,\tw(\xi,t-s)\frac{(\delta \kappa)'  \left(\frac{\tc 
      s}{2}\right)}{8\kappa_0 (1-c_0^2\xi^2)^{5/2}} 
\]
which is equation \ref{eqn:dmodhom}.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/ssevelan/paper.tex
repro docs dir added

\title{Velocity Inversion by Surface Source Extension}
\author{William. W. Symes \thanks{The Rice Inversion Project,
Department of Computational and Applied Mathematics, Rice University,
Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}

\lefthead{Symes}

\righthead{Velocity Inversion by Surface Source Extension}

\maketitle
\begin{abstract}
Extension methods reformulate inverse problems in wave propagation by adding degrees of freedom to the domain of an objective function, the minimizer of which is the solution of the inverse problem. In favorable cases the modified objective function has a larger domain of convexity than does the original, leading to effective solution of the inverse problem via numerical optimization. Source extension adds degrees of freedom to the energy source representation. When the extended sources are confined to a hypersurface, they are uniquely determined by velocity and other material parameters, even when those parameters are inconsistent with the data of the in verse problem. This ability of extended surface sources to fit data regardless of the choice of other parameters is one of the crucial links in the relation of a natural extended objective function with travel time tomography.

\end{abstract}

\inputdir{../trans/project}

\section{Full Waveform Inversion}

A simple example: acoustic transmission inversion - isotropic point radiator (``source'') located at  $\bx=\bx_s$
\begin{equation}
\label{eqn:awept}
\frac{1}{\kappa}\frac{\partial p}{\partial t} = -\nabla \cdot {\bf v} + f(t)\delta(\bx-\bx_s), \,\rho \frac{\partial {\bf v}}{\partial t} = - \nabla p; \,p,{\bf v}=0, t \ll 0
\end{equation}
Bulk modulus is depicted in Figure \ref{fig:bml0}.

\plot{bml0}{width=\textwidth}{Lens example: target bulk modulus $\kappa$}

$x_s=3500$ m, $z_s=3000$ m, receiver positions $x_r=2000-6000$ m, $z_r=1000$ m

Modeling operator $(S[\kappa]f)(t,x_r) = \phi(x_r) p(t,x_r,z_r)$ - linear in $f$, nonlinear in $\kappa$

%\includegraphics[height=1.5in]{Fig/pulse00.pdf} %\includegraphics[height=1.5in]{Fig/ptpwindl0.pdf}\\
\plot{pulse00}{width=0.5\textwidth}{Bandpass filter pulse $f$}
\plot{ptpwindl0}{width=\textwidth}{Receiver line  $S(\kappa)$  at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km. bulk modulus $\kappa$ with low velocity lens as in Figure \ref{fig:bml0}, $\rho$=1 g/cm$^{3}$. Coarse grid: $dx=dz=20$ m, $dt=$ 8 ms, (2,4) staggered grid scheme. }
% at receiver position $1000 m$ for source at $x=3500$ m, $z=3000$ m}

{\bf F}ull {\bf W}aveform {\bf I}nversion = output least squares: minimize over $\kappa$ (for example)
\[
J_{\rm FWI}[\kappa,d] = \frac{1}{2}\|S[\kappa]f-d\|^2
\]
Variants of Newton's method use preconditioned gradient
\[
\nabla J_{\rm FWI}[\kappa,d] = W^{-1}DS[\kappa]^T(S[\kappa]f-d)
\]
$W^{-1}$ = preconditioner $\Leftrightarrow$ $W$ = Gram op for inner product on model space

{\color{blue} this talk: constrain bulk modulus $\kappa$ to be {\em smooth} ($\Rightarrow W^{-1}$ smoothing), density $\rho$ constant \& fixed}

What goes wrong: ``cycle skipping'', ``trapped in local minima'',...

Assume local geometric optics: $\tau(x_s,x_r) =$ ray travel time from $(x_s,z_s)$ to $(x_r,z_r)$, $a(x_s,x_r)$ = geometric amplitude
\[
S[\kappa]f(\bx_r,t) \approx a(x_s,x_r)f(t-\tau(x_s,x_r))
\]
then $S[\kappa]f=d$ $\Rightarrow$
\[
D^2J_{\rm FWI}[\kappa,d](\delta \kappa,\delta \kappa) \approx \|(...)D\tau[\kappa]\delta \kappa\|^2\left\|\frac{df}{dt}\right\|^2
\]
High frequency asymptotics: $f_{\lambda}(t) = \frac{1}{\sqrt{\lambda}} f\left(\frac{t}{\lambda}\right), d_{\lambda} = S[\kappa]f_{\lambda}$

$
\|D\tau[\kappa]\delta \kappa\|=O(1) \Rightarrow D^2J_{\rm FWI}[\kappa,d_{\lambda}](\delta \kappa,\delta \kappa)=O(\lambda^{-2})$ but $ J_{\rm FWI}[\kappa,d_{\lambda}]=O(1)
$
$\Rightarrow$ {\color{blue} $J_{\rm FWI}$ convex in region of diameter $O(\lambda)$}

Extension of $S$ = map $\oS$ on larger domain, with embedding $E$ 
\[
S = \oS \circ E
\]
Annihilator $A$: $\mbox{ker }A = \mbox{range E}$.

Example - {\bf S}urface {\bf S}ource {\bf E}xtension

Domain of $\oS = \{\mbox{(smooth bulk moduli)} \times {\cal E}'\{z=z_s, x_s-r < x < x_s+r\} \}$

Compute $\oS$: solve wave equation with point source {\color{blue} $f(t)\delta(x-x_s)$}$\delta(z-z_s)$ replaced by {\color{blue}$\of(t,x)$}$\delta(z-z_s) \in {\cal E}\{z=z_s, x_s-r < x < x_s+r\}$

\begin{equation}
\label{eqn:extann}
Ef(t,x) = f(t)\delta(x-x_s),\,A\of(t,x)=(x-x_s)\of(t,x)
\end{equation}

Big deal: $\oS[\kappa]$ has a microlocal approximate inverse $\oS[\kappa]^{\dagger}$ [$S$ has approximate left inverse {\em but not right!}]:
\begin{eqnarray}
\oS[\kappa]^{\dagger}\oS[\kappa] &=& \Pi_s[\kappa]\\
\oS[\kappa]\oS[\kappa]^{\dagger} &=& \Pi_r[\kappa]
\end{eqnarray}
The approximate projectors $\Pi_s,\Pi_r \in OPS^0$ have symbols $\equiv 1$ in conic $\gamma_{s,r} \subset T^*(\Gamma_{s,r})$

%\includegraphics[height=1.2in]{Fig/ptpwindl0.pdf}\includegraphics[height=1.2in]{Fig/cgwhestsourceplh0.pdf}\\
\plot{cgwhestsourceplh0}{width=\textwidth}{$\of=\oS(\kappa_0)^{\dagger}d$ for $\kappa_0 \equiv 4.0$ GPa, approx. via CG - $r=2000 m$}

At solution source should be supported at $x=x_s$ - in $\mbox{rng }E = \mbox{ker }A$

One way to incorporate this: penalty function
\begin{equation}
\label{eqn:Jnaive}
\tilde{J}_{\alpha}[\kappa,\of,d] = \frac{1}{2}(\|\oS[\kappa]\of -d\|^2 + \alpha^2 \|A\of\|^2)
\end{equation}

Reduced objective:
\[
J_{\alpha}[\kappa,d] = \mbox{min}_{\of} \tilde{J}_{\alpha}[\kappa,\of,d] = \tilde{J}_{\alpha}[\kappa,\of_{\alpha}[\kappa,d],d]
\]
$\of_{\alpha}[\kappa,d]$ = solution of {\em inner problem}
\[
N_{\alpha}[\kappa]\of = \oS[\kappa]^Td,\,\,N_{\alpha}[\kappa] = \oS[\kappa]^T\oS[\kappa] + \alpha^2 A^TA
\]

Solving the outer problem (``variable projection method'', Golub \& Pereyra 73, 03):
\[
\nabla J_{\alpha}[\kappa,d] = W^{-1} D\oS[\kappa]^T(\of_{\alpha}[\kappa,d],\oS[\kappa]\of_{\alpha}[\kappa,d]-d)
\]
closely resembles FWI gradient, efficiently computable via adjoint state method

Algorithm:
\begin{itemize}
\item solve inner problem using CG
\item adjust $\alpha$ using Discrepancy Principle (keep residual $\in (e_-, e_+)$) - $\oS^{\dagger} \Rightarrow$ can start with $\alpha=0$ (Fu \& S. 17)
\item outer update using steepest descent, line search
\end{itemize}

\section{Modeling}

First, a formal definition of $\oS$: choose cutoff functions $\chi_a \in C_0^{\infty}(\Gamma_a \times \bR)$ with $a \in \{s,r\}$. For $\of \in C_0^{\infty}(\Gamma_s \times \bR)$, the initial value problem
\begin{equation}
\label{eqn:awept}
\frac{1}{\kappa}\frac{\partial p}{\partial t} = -\nabla \cdot {\bf v} + \chi_s\of, \,\rho \frac{\partial {\bf v}}{\partial t} = - \nabla p; \,p,{\bf v}=0, t \ll 0
\end{equation}
has a unique solution $(p,\bv)$, smooth except on $\Gamma_s \times \bR$. 
Define
\begin{equation}
\label{eqn:os1def}
\oS_1[\kappa]\of = \chi_r p|_{\Gamma_r \times \bR}
\end{equation}

First Key Observation: adopt

\noindent {\bf Assumption A:} There exists $\theta_*>0$ so that for every ray connecting a point in $\mbox{supp }\chi_s$ with a point in $\mbox{supp }\chi_r$, the angles $\theta_s$ and $\theta_r$ subtended by the ray and the tangent planes of $\Gamma_s$ and $\Gamma_r$ respectively are both greater than $\theta_*$ in absolute value.

Denote by $\gamma_{s,r}$ the open set of rays $\{(\bx,r \xi): r>0\}$ through projections $(\bx,\xi)$ onto $T^*(\Gamma_{s,r} \times \bR)$ of initial/final conditions for rays connecting points in the interiors of $\mbox{supp }\chi_{s,r}$: these are open conic sets. Note that $\gamma_{s,r}$ is $\kappa$-dependent. If $C[\kappa]$ denotes the twisted canonical relation for $\oS$, $\gamma_r = C[\kappa]\gamma_s$.

Then 
$\oS_1[\kappa]$ has a microlocal approximate inverse $(\oS[\kappa]^1)^{\dagger}$, defined by 
\begin{equation}
\label{eqn:appinv}
\oS_1^{\dagger} = W_s^{-1}\oS_1^TW_r
\end{equation}
in which $W_s, W_r$ are properly supported $\Psi$DOs on $\Gamma_s \times \bR,\Gamma_r \times \bR$ respectively [from here on, suppress the arguments whenever these are obvious: $\oS_1 = \oS_1[\kappa]$ and so on]. The sense in which $\oS_1^{\dagger}$ is an inverse is that
\begin{eqnarray}
\label{eqn:proj}
\oS_1^{\dagger}\oS &=& \Pi^1_s\\
\oS \oS_1^{\dagger} &=& \Pi^1_r
\end{eqnarray}
in which $\Pi^1_{s,r} \in OPS^0(\Gamma_{s,r}\times \bR)$ have principal symbols $\sim 1$ in the conic sets $\gamma_{s,r} \subset T^*(\Gamma_{s,r}\times \bR)$.
%, and $\ge 1/2$ in the larger conic sets $\tilde{\gamma}^1_{s,r} \supset \gamma^1_{s,r}$.

It follows immediately from \ref{eqn:proj} that
\begin{equation}
\label{eqn:picomm}
\oS_1 \Pi^1_s = \Pi^1_r\oS_1
\end{equation}

The (G\^{a}teaux) derivative $D\oS_1 = D\oS_1[\kappa]\delta \kappa$ (that is, a perturbation $\delta \kappa \in C_0^{\infty}(\bR^d)$ is implicit in the notation) is a Fourier Integral Operator operator of order 1, with the same canonical relation: a proof of this statement is given in appendix 1. It follows that
\begin{eqnarray}
\label{eqn:qdef}
Q_s &\equiv& \oS_1^{\dagger}D\oS_1 \nonumber \\
Q_r &\equiv& D\oS_1 \oS_1^{\dagger} 
\end{eqnarray}
are microlocally eilliptic $\Psi$DOs of order 1, and satisfy
\begin{eqnarray}
\label{eqn:qrel}
\oS_1 Q_s &=& \Pi^1_r D\oS_1 \nonumber \\
Q_r \oS_1 &=& D\oS_1 \Pi^1_s \nonumber \\
Q_s\oS_1^{\dagger} &=& \oS_1^{\dagger} Q_r
\end{eqnarray}
Since $D\oS_1$ is linear in $\delta \kappa$, so is $Q_{s,r} = Q_{s,r}[\kappa]\delta \kappa$.

\begin{lemma}\label{thm:qskew} $Q_s$ is essentially skew-symmetric, that is, $B \equiv Q_s+Q_s^T \in OPS^0(\Gamma_s \times \bR)$.
\end{lemma}

\begin{proof}: In the definition $\oS_1^{\dagger}=W_s^{-1}\oS_1^TW_r$, the weights $W_s, W_r$ are symmetric invertible members of $OPS^{0}(\Gamma_{s,r} \times \bR)$.
\[
D(\oS_1^{\dagger}\oS_1) = D\Pi^1_s
\]
\[
=D(W_m^{-1})\oS_1^TW_d \oS_1 + W_m^{-1}D\oS_1^TW_d \oS_1 + W_m^{-1}\oS_1^TDW_d \oS_1 + W_m^{-1}\oS_1^TW_dD\oS_1
\]
Also 
\[
Q_s^T = D\oS_1^T W_d ^T\oS_1 W_m^{-1}
\]
so
\[
W_m^{-1}Q_s^TW_m = W_m^{-1}D\oS_1^TW_d \oS_1
\]
is identical to the second term above, and by the $\Psi$DO calculus differs from $Q_s^T$ by a member of  $OPS^0(\Gamma_S \times \bR)$. The first and the third define members of $OPS^0(\Gamma_S \times \bR)$, by the rules for composition of FIO, and the fourth term is $Q_s$. \end{proof}

Select $\kappa^0 \in C^{\infty}(\bR^d)$, and choose a neighborhood $U \subset C^{\infty}(\bR^d)$ of $\kappa_0$ so that there exist non-void open conic subsets $\gamma^0_s, \tilde{\gamma}^0_s$ of $ T^*(\Gamma_{s,r} \times \bR)$ with $\overline{\tilde{\gamma}^0_s} \subset \gamma^0_s$ so that for $\kappa \in U$, 
\begin{itemize}
\item[A1: ] $\gamma^0_s \subset \gamma_s[\kappa]$
\item[A2: ] $C[\kappa]\gamma^0_s \subset  \gamma_r[\kappa]$
\end{itemize}
(recall that $C[\kappa]$ is the twisted canonical relation of $S$).

Let $\Pi_s$ denote an order zero $\Psi$DO with principal symbol $\sim 1$ in $\tilde{\gamma}^0_s$, $\mbox{ess supp }\Pi_s \subset \gamma^0_s$. From relation \ref{eqn:picomm}
\begin{eqnarray}
\label{eqn:pistuff3}
\Pi^1_r \oS_1 \Pi_s = \oS_1 \Pi^1_s \Pi_s  =\oS_1(\Pi_s^1-I)\Pi_s + \oS_1 \Pi_s
\end{eqnarray}
Since the symbol of $\Pi^1_s$ is $ \sim 1$ on $\mbox{ess supp}(\Pi_s)$, $(\Pi_s^1-I)\Pi_s \in OPS^{-\infty}(\Gamma_s \times \bR)$. Thus the first term on the RHS in equation aboveis (infinitely) smoothing,

In the sequel, I will write $\approx$ for ``differs by smoothing operator''. Thus
\begin{eqnarray}
\label{eqn:pistuff3}
\Pi^1_r \oS_1 \Pi_s \approx \oS_1 \Pi_s
\end{eqnarray}
Define 
\begin{equation}
\label{osdef}
\oS = \oS_1 \Pi_s
\end{equation}
Since $\Pi_s$ is independent of $\kappa \in U$,
\[
D\oS = D\oS_1 \Pi_s \approx (\Pi^1_r D\oS_1-D\Pi \oS_1 \Pi_s)
\]
from \ref{eqn:pistuff3}. The second term on the RHS is smoothing because of the null intersection of essential supports, so
\[
\approx \oS_1 Q_s \Pi_s = \oS Q_s + \oS_1 [Q_s,\Pi_s]+ = \oS(Q_s+[Q_s,\Pi_s]) + \oS_1(I-\Pi_s)[Q_s,\Pi_s]
\]
Thus
\begin{equation}
\label{eqn:qrel0}
D\oS \approx \oS Q + \oS_1R
\end{equation}
in which $Q=Q_s + [Q_s,\Pi_s] \in OPS^1(\Gamma_s \times \bR)$ and $R = (I-\Pi_s)[Q_s,\Pi_s] \in OPS^0(\Gamma_s \times \bR)$. Note that the principal symbols of $Q$ and $Q_s$ are identical, and 
\begin{equation}
\label{eqn:esssuppr}
\mbox{ess supp}(R) \subset \gamma_s \setminus \tilde{\gamma}_s.
\end{equation}
That is, $R$ is essentially supported near the boundary of $\gamma_s$.

\section{Regularization}
The naive version of the least squares problem given in definition \ref{eqn:Jnaive} is unsatisfactory for several reasons. First, the intended null space of the annihilator (equation \ref{eqn:extann}) does not lie in its domain ($L^2$): in fact, the null space is actually the $\{0\}$. Moreover, the multiplier defining $A$ has a singularity, so $A$ is not pseudodifferential. Second, the normal operator of the least squares problem of minimizing the function \ref{eqn:Jnaive} is not coercive, so there is no guarantee that the reduced form is well-defined. Therefore the variable projection method actually makes no sense for this problem.

Two amendments remedy both problems. To give the annihilator a nonempty null space approximating the desired one (i.e. supported near $\bx_s$), and make it a pseudodifferential operator, cut it off smoothly near $\bx_s$. Specifically, choose $\chi_1 \in C_0^{\infty}(\bR^d)$ so that $\chi_1(x)=0$ for $|x|<1$, $\chi_1(x)=1$ for $|x| >2$. For $r>0$, define $\chi_r \in C_0^{\infty}(\bR^d)$ by
\[
\chi_r(x) = \chi_1(x/r).
\]
Then set 
\[
A_r = A(1-\chi_{2r}), \, Au(\bx,t)=|\bx-\bx_s|u(\bx,t)|
\]
$A_r$ is a regularized version of $A$, the null space of which consists of square-integrable functions supported in the cylinder $|\bx-\bx_s| \le r$. In practice, $r$ might be chosen to be a couple of wavelengths (or even ignored altogether - this seems to work, though it is not altogether clear why). Since the multiplier defining $A_r$ is smooth, $A_r$ is also pseudodifferential of order $0$.

The second problem is solved simply by Tihonov regularization. Introduce an additional penalty proportional to $\|\of\|^2$:
\begin{equation}
\label{eqn:jdef}
\Ja[\kappa,\of]=\frac{1}{2}(\|\oS \of - d\|^2 + \alpha^2 \|A_r\of\|^2 + \lambda^2\|\of\|^2)
\end{equation}
The normal operator 
\begin{equation}
\label{eqn:normop}
\oNa = \oS^T\oS + \alpha^2 A_r^TA_r + \lambda^2 I
\end{equation}
is coercive for $\lambda>0$, so the minimizer $\ofa=\ofa[\kappa,d]$ of $\Ja[\kappa,\cdot]$ is well-defined as the unique solution of the normal equation
\begin{equation}
\label{eqn:normal}
\oNa \ofa = \oS^Td.
\end{equation}
The reduced objective
\begin{equation}
\label{eqn:normred}
\Jared[\kappa,d] = \Ja[\kappa,\ofa[\kappa,d],d]
\end{equation}
is well-defined and smooth in its arguments.

The parametrices introduced above for $\oS_1$ lead to a useful approximation of $\ofa$. To simplify notation in the following calculations, define 
\[
M_1=\oS_1^T\oS_1,\, M=\oS^T\oS = \Pi_s^TM_1\Pi_s
\]
and
\[
M_1^{\dagger} = \oS_1^{\dagger}(\oS_1^{\dagger})^T
\]
From the definitions \ref{eqn:proj} and the relation \ref{eqn:picomm},
\[
M_1^{\dagger}M_1=\oS_1^{\dagger}(\oS_1^{\dagger})^T \oS_1^T\oS_1
=\oS_1^{\dagger}(\oS^1\oS_1^{\dagger})^T\oS_1
\]
\[
=\oS_1^{\dagger}\Pi^1_r\oS_1=\oS_1^{\dagger}\oS_1\Pi^1_s = (\Pi^1_s)^2
\]
and
\[
M_1 M_1^{\dagger}=\oS_1^T\oS_1 \oS_1^{\dagger}(\oS_1^{\dagger})^T 
=\oS_1^T(\Pi^1_r)^T(\oS_1^{\dagger})^T 
\]
\[
=(\oS_1^{\dagger}\Pi^1_r\oS_1)^T = ((\Pi^1_s)^2)^T \approx (\Pi^1_s)^2
\]
where the final equality modulo smoothing error follows from the observation the $\Pi_s$ is a scalar $\Psi$DO with real symbol, so that $\Pi_s$ and $\Pi_s^T$ have the same principal symbol.

That is, $M_1^{\dagger}$ is a two-sided microlocal parametrix for $M_1$. It turns out to be a parametrix for $M$ as well:
\[
M_1^{\dagger} M = M_1^{\dagger} \Pi_s M_1 \Pi_s =  M_1^{\dagger}M_1\Pi_s^2 +  M_1^{\dagger}[\Pi_s,M_1]\Pi_s
\]
The last term is in $OPS^{-1}$, so this is
\[
\approx (\Pi^1_s)^2\Pi_s^2 \approx \Pi_s^2
\]
since $\Pi^1_s\Pi_s \approx \Pi_s$ etc. Similarly,
\[
M M_1^{\dagger} \approx \Pi_s^2.
\]
That is, $M_1^{\dagger}$ is also a two-sided microlocal parametrix for $M$.

To understand the behaviour of $\ofa$ as $\lambda \rightarrow 0$ it is also useful to introduce an approximate inverse for $\oS^T\oS + \lambda^2 I = \Ml$. Define
\[
\Mld = M_1^{\dagger} (I + \lambda^2 M_1^{\dagger})^{-1}.
\]
The inverse is well-defined for $\lambda$ sufficiently small. Then
\[
(\Ml ) \Mld \approx (\Pi_s^2 + \lambda^2 M_1^{\dagger}) (I + \lambda^2 M_1^{\dagger})^{-1}
\]
\[
=I - (I-(\Pi_s)^2) (I + \lambda^2 M_1^{\dagger})^{-1}
\]
Similarly,
\[
\Mld (\Ml) \approx I - (I + \lambda^2 M_1^{\dagger})^{-1} (I-(\Pi_s)^2) 
\]
Thus $\Mld$ is a two-sided parametrix for $\Ml$ in the sense that 
\[
\Mld (\Ml) P \approx P, \, (\Ml) \Mld P \approx P
\]
for and $P \in OPS^0$ essentially supported in $\tilde{\gamma}_s$.

\section{Consistency}

I will call the triple $(\kappa,\of,d)$ {\em consistent} if $\oS[\kappa]\of = d$. A distributed source $\of\in {\cal D}(\Gamma_s \times \bR)$ is {\em $r$-approximately point} ($r>0$) if 
\begin{equation}
\label{eqn:apt}
\of = \tilde{P} f \delta_{\Gamma \times \bR},
\end{equation}
for $f \in C^{\infty}_0(\Gamma_s \times \bR)$, $\mbox{supp }f \subset B_r(\bx_s) \times \bR$ and $\tilde{P} \in OPS^0(\Gamma \times \bR)$ with essential support $\subset \tilde{\gamma}_s$.

Will call $f$ in \ref{eqn:apt} the {\em wavelet}, and $\tilde{P}$ the {\em aperture operator}. Note that if $\of$ is $r$-approximately point, then its wavelet satisfies $f=\chi_r f$.

\begin{lemma}
\label{thm:asm}
Suppose that $\of = \tilde{P}f$ is $r$-approximately point. Then for any $s \in \bR, B \in OPS^s(\Gamma_s \times \bR)$, 
\[
A B \of = E f
\]
in which $E \in OPS^{-\infty}(\Gamma_s \times \bR)$.
\end{lemma}
\begin{proof}
True because of pseudolocality and the null intersection of the supports of $A_r=A(1-\chi_{2r})$ and $\chi_r$.
\end{proof}

Suppose that $\of$ is approximately point, with wavelet $f$ and aperture operator $\tilde{P}$. Then
\[
(\Ml + \alpha^2 A_r^TA_r)(\ofa - \Mld M\tilde{P}f) = M\tilde{P}f 
\]
\[
- (I + (I + \lambda^2 M_1^{\dagger})^{-1} (I-(\Pi_s)^2)) M\tilde{P}\chi_rf -\alpha^2 A_r^TA_r \Mld M\tilde{P}\chi_rf
\]
\begin{equation}
\label{eqn:old}
= E f
\end{equation}
in which 
\[
E = (I + \lambda^2 M_1^{\dagger})^{-1} (I-(\Pi_s)^2) M\tilde{P} - 
-\alpha^2 A_r^TA_r \Mld M\tilde{P}\chi_r
\]
is a smoothing operator: the first term on the right is smoothing because the essential supports of $\tilde{P}$ and $(I-(\Pi_s)^2)$ have null intersection, and the second term is smoothing per Lemma \ref{thm:asm}.

I shall use $E$ as a symbol denoting various smoothing operators that occur in the calculations to come.

Now introduce a family $\{f_{\lambda}: \lambda>0\}$ of wavelets satisfying
\begin{itemize}
\item $\|f_{\lambda}\|=1$ for all $\lambda$;
\item for all $k \ge 0$, there is $c_k>0$ so that $\|(I - \Delta)^{-k/2}f_{\lambda}\| \le c_k \lambda^{-2k}$
\end{itemize}
In the last condition, $\Delta$ is the Laplace-Beltrami operator in the induced metric on $\Gamma_s \times \bR$. The condition can be interpreted as stating that $f_{\lambda}$ oscillates with wavelength $\lambda^2$. Accordingly I will call such a family {\em high-frequency}.

\begin{lemma}\label{thm:ofaerr}
Suppose that $\of$ is approximately point, with wavelet $f$ and aperture operator $\tilde{P}$, and $d=S\tilde{P}$. Then
for $s, k \ge 0$, there exist $c_{s,k}\ge 0$ so that
\begin{equation}
\label{eqn:ofaerr}
\|\ofa - \Mld M\tilde{P}f_{\lambda}\|_s \le c_{s,k}\lambda^{2k}, \,k\ge s+1
\end{equation}
\end{lemma}
\begin{proof}
\[
(\Ml + \alpha^2 A_r^TA_r)(\ofa - \Mld M\tilde{P}f) = M\tilde{P}f 
\]
\[
- (I + (I + \lambda^2 M_1^{\dagger})^{-1} (I-(\Pi_s)^2)) M\tilde{P}\chi_rf -\alpha^2 A_r^TA_r \Mld M\tilde{P}\chi_rf
\]
\begin{equation}
\label{eqn:old}
= E f
\end{equation}
in which 
\[
E = (I + \lambda^2 M_1^{\dagger})^{-1} (I-(\Pi_s)^2) M\tilde{P} - 
-\alpha^2 A_r^TA_r \Mld M\tilde{P}\chi_r
\]
is a smoothing operator: the first term on the right is smoothing because the essential supports of $\tilde{P}$ and $(I-(\Pi_s)^2)$ have null intersection, and the second term is smoothing per Lemma \ref{thm:asm}. Note that 
\[
\Ml + \alpha^2 A_r^TA_r \ge \lambda^2 I
\]
whence invertible, thanks to Lax-Milgram, with inverse bounded by $\lambda^{-2}$. So invoking the defining properties of $f_{\lambda}$,
\[
\|\ofa - \Mld M\tilde{P}f\| \le \lambda^{-2}\|Ef\| = \lambda^{-2}\|E(I-\Delta)^{k/2}(I-\Delta)^{-k/2}f_{\lambda}\|
\]
$E(I-\Delta)^{k/2}$ is smoothing hence bounded, so the second defining property of $f_{\lambda}$ gives 
\[
\le c_{0,k}\|f_{\lambda}\|\lambda^{2(k-1)}
\]
which is the $s=0$ case of the bound \ref{eqn:ofaerr}.

Assume that the bound has been established for $s-1$. Note that
\[
(I-\Delta)^{s/2}(\Ml + \alpha^2 A_r^TA_r)(\ofa - \Mld M\tilde{P}f) 
\]
\[
=(\Ml + \alpha^2 A_r^TA_r)(I-\Delta)^{s/2}(\ofa - \Mld M\tilde{P}f) 
+[(I-\Delta)^{s/2},(\Ml + \alpha^2 A_r^TA_r)](\ofa - \Mld M\tilde{P}f) 
\]
\[
=(I-\Delta)^{s/2}Ef
\]
The operator in the second term is of order $s-1$, so that contribution is bounded in $L^2$ by 
\[
\mbox{const. } \times \|\ofa - \Mld M\tilde{P}f_{\lambda}\|_{s-1} \le c_{s-1,k} \lambda^{k}
\]
for $k \ge 1$. Repeating the manipulations of the $k=0$ case, arrive at the conclusion.
\end{proof},

The next observations will permit $r$ approximately point wavelets to be treated as if they generated point sources in the null space of the penalty operator $A_r$.
\begin{lemma}
\label{thm:aptsm}
Suppose that $\{f_{\lambda}\}$ is an $r$-approximate point high frequency wavelet family, and $B \in OPS^{s}(\Gamma \times \bR)$. Then for any $s \in \bR$, $k \in {\bf N}$
\[
\|AB f_{\lambda}\|_s = O(\lambda^{2k}).
\]
\end{lemma}
\begin{proof}
Follows from Lemma \ref{thm:asm} and the defining properties of high-frequency wavelet families.
\end{proof}

\begin{lemma}
\label{thm:ofasm}
Suppose that $d=S\tilde{\Pi}f_{\lambda}$ is $r$-approximately point and $\{f_{\lambda}\}$ is a high-frequency wavelet family. Then for any $s \in \bR$, $k \in {\bf N}$, there is $C>0$ so that for sufficiently small $\lambda>0$,
\[
\|A_r \ofa\|_s \le C\lambda^{2k}
\]
\end{lemma}
\begin{proof}
From the bound \ref{eqn:ofaerr}, Lemma \ref{thm:aptsm} and the definitions above,
\[
\|A_r \ofa \|_s \le \|A_r(\ofa - \Mld M\tilde{P}f_{\lambda})\|_s + \|A_r \Mld M \tilde{P}f_{\lambda}\|_s \le c_{s,k}\lambda^{2k}
\]
for suitable $c_{s,k}>0$.
\end{proof}

\begin{lemma}
\label{thm:dataerr}
Under the assumptions of Lemma \ref{thm:ofaerr}, in particular $d_{\lambda} = \oS\tilde{P}f_{\lambda}$, for all sufficiently small $\lambda$, any $s \in \bR$, there exists $C_s>0$ so that,
\[
\|\oS \ofa - d_{\lambda}\|_s \le C_s \lambda^2.
\]
\end{lemma}
\begin{proof}
\[
\oS\ofa - d_{\lambda} = \oS(\ofa - \tilde{P}f_{\lambda}) 
\]
\begin{equation}
\label{eqn:derr1}
= \oS(\ofa - \Mld M\tilde{P}f) + \oS((\Mld M-I)\tilde{P}f)
\end{equation}
\[
\Mld M = \Mld(\Ml) - \lambda^2 \Mld \approx I - (I+\lambda^2M_1^{\dagger})^{-1}(I-\Pi_s^2) - \lambda^2 M_1^{\dagger}(I+\lambda^2M_1^{\dagger})^{-1}
\]
\[
= I - (I+\lambda^2M_1^{\dagger})^{-1}(I-\Pi_s^2) - (I+\lambda^2 M_1^{\dagger})(I+\lambda^2M_1^{\dagger})^{-1} + (I+\lambda^2M_1^{\dagger})^{-1}
\]
\[
-(I+\lambda^2M_1^{\dagger})^{-1}(I-\Pi_s^2)+ (I+\lambda^2M_1^{\dagger})^{-1}
\]
\[
= (I+\lambda^2M_1^{\dagger})^{-1}\Pi_s^2
\]
So
\[
\Mld M - I =(I+\lambda^2M_1^{\dagger})^{-1}\Pi_s^2 - (I+\lambda^2M_1^{\dagger})^{-1}(I+\lambda^2M_1^{\dagger})
\]
\[
=
(I+\lambda^2M_1^{\dagger})^{-1}(\Pi_s^2-I -\lambda^2M_1^{\dagger})
\]
whence there exists $C>0$ so that for any $k>0$, sufficiently small $\lambda$,
\[
\|\oS\ofa - d\|_s  \le C (\|\ofa - \Mld M\tilde{P}f_{\lambda}\|_s 
\]
\[
+ \|(I+\lambda^2M_1^{\dagger})^{-1}(\Pi_s^2-I -\lambda^2M_1^{\dagger})\tilde{P}f_{\lambda}\|_s
\]
\[
\le C(\lambda^{2k} + \lambda^2\|(I+\lambda^2M_1^{\dagger})^{-1}M_1^{\dagger}\tilde{P}f_{\lambda}\|_s)
\]
\[
\le C_s\lambda^2
\]
for suitable choice of $C_s > 0$.
\end{proof}

\section{Derivatives}
The variable projection method amounts to minimization of the reduced objective
\[
\Jared[\kappa] = \Ja[\kappa,\ofa[\kappa]].
\]
It follows from the normal equation \ref{eqn:normal} that the derivative of $\Jared$ is given by
VPM derivative formula:
\[
D\Jared[\kappa]\delta \kappa = \langle D(\oS[\kappa] \ofa[\kappa])\delta[\kappa],\oS[\kappa] \ofa[\kappa] -d\rangle
\]
To simplify notation, in the following I will write suppress the arguments of the derivatives in this formula, in particular the bulk modulus perturbation: $D\oS = D\oS[kappa]\delta \kappa$ and so on.

The relation \ref{eqn:qrel0} implies that
\[
D\Jared \approx \langle (\oS Q + \oS_1 R + E) \ofa,\oS\ofa -d\rangle
\]
As mentioned above, $E$ is a smoothing operator, the identity of which may change from equation to equation.
\[
= \langle Q\ofa,\oS^T(\oS\ofa -d)\rangle
+ \langle \oS_1R \ofa, \oS_1^T\oS\ofa-d \rangle
+\langle E\ofa,\oS\ofa-d\rangle
\]
\[
 = \langle Q \ofa, -(\alpha^2 A_r^TA_r + \lambda^2I)\ofa \rangle 
+ \langle R\ofa,   \oS_1^T\oS\ofa -d \rangle 
+\langle E\ofa,\oS\ofa-d\rangle
\]
\[
 = \langle Q \ofa, -(\alpha^2 A_r^TA_r + \lambda^2I)\ofa \rangle 
+ \langle R\ofa,   \oS_1^T(\oS\ofa -d) \rangle 
+\langle E\ofa,\oS\ofa-d\rangle
\]
using the normal equation \ref{eqn:normal}. From Lemma \ref{thm:qskew} and the observation that $Q_s$ and $Q$ have the same principal symbol,
\[
 = -\alpha^2\langle [Q,A_r] \ofa, A_r\ofa \rangle -\frac{\alpha^2}{2} \langle (Q+Q^T)A_r\ofa,A_r \ofa \rangle - \frac{\lambda^2}{2} \langle (Q + Q^T)\ofa,\ofa \rangle 
\]
\[
+ \langle R\ofa,   \oS_1^T(\oS\ofa -d) \rangle 
+\langle E\ofa,\oS\ofa-d\rangle 
\]
\[
 = -\alpha^2\langle [Q,A_r] \ofa, A_r\ofa \rangle -\frac{\alpha^2}{2} \langle BA_r\ofa,A_r \ofa \rangle - \frac{\lambda^2}{2} \langle B\ofa,\ofa \rangle 
\]
\begin{equation}
\label{eqn:vpderiv}
+ \langle R\ofa,   \oS_1^T(\oS\ofa -d) \rangle 
+\langle E\ofa,\oS\ofa-d\rangle 
\end{equation}

Computation of the 2nd derivative requires differentiation of $\ofa$.
\begin{lemma}
\label{thm:dofa}
The derivative $d\ofa$ may be expressed as
\begin{equation}
\label{eqn:dofa}
= G \oS_1^T d + Q_1 \ofa 
\end{equation}
in which $Q_1 \in OPS^1(\Gamma_s \times \bR)$ is essentially skew and in fact has the same principal symbol as $Q$, and $G \in OPS^0(\Gamma_s \times \bR)$.
\end{lemma}
\begin{proof}
\[
D\ofa = D(\oNa^{-1}\oS d) = -\oNa^{-1}(D(\oS^T)\oS + \oS^TD\oS)\oNa^{-1}\oS d
+ \oNa^{-1}D\oS d
\]
\[
=-\oNa^{-1}((\oS Q + \oS_1 R)^T\oS +\oS^T(\oS Q + \oS_1 R))\oNa^{-1}\oS^Td
+\oNa^{-1}(\oS Q + \oS_1 R)^Td
\]
\[
=-\oNa^{-1}(Q^T \oS^T\oS + R^T\oS_1^T\oS +\oS^T\oS Q + \oS^T\oS_1 R))\oNa^{-1}\oS^Td +\oNa^{-1}(\oS Q + \oS_1 R)^Td
\]
\[
=-\oNa^{-1}(B\oS^T\oS + [\oS^T\oS,Q] + R^T\oS_1^T\oS + \oS^T\oS_1 R))\ofa +([\oNa^{-1},Q]\oS^T+ R^T\oS_1^T)d + Q\ofa
\]
\[
= G \oS_1^T d + Q_1 \ofa 
\]
in which $G$ and $Q_1$ are as described.
\end{proof}

The remainder of this section is devoted to computing the second derivative at a consistent model with an $r$-approximate point source and a high-frequency wavelet. Equation \ref{eqn:vpderiv} decomposes the first derivative into five terms, which I will treat one-by-one.

\subsection{Term 1}
\[
D\langle [Q,A_r] \ofa, A_r\ofa \rangle = \langle D[Q,A_r]\ofa,A_r\ofa\rangle 
\]
\[+
\langle [Q,A_r] D\ofa, A_r \ofa \rangle + \langle [Q,A_r]\ofa, A_r D\ofa \rangle
\]
Since $D[Q,A_r] \in OPS^0$, the first summand is bounded by a multiple of $\|A_r\ofa\|$ hence by a multiple of $\lambda^{2k}$ for any $k$, thanks to Lemma \ref{thm:aptsm}.

From Lemma \ref{thm:dofa}, the second summand is 
\[
= \langle [Q,A_r](GS_1^Td+Q_1\ofa,A_r\ofa \rangle = \langle [Q,A_r]GS_1^TS \tilde{\Pi}f_{\lambda},A_r\ofa \rangle +\langle [[Q,A_r],Q_1]\ofa,A_r\ofa \rangle + \langle [Q,A_r]\ofa,Q_1^TA_r\ofa\rangle
\]
All terms here are $O(\lambda^{2k})$ because of Lemmas\ref{thm:asm} and \ref{thm:aptsm} with $s=0$ or $1$.

The third term can be rewritten using Lemma \ref{thm:dofa} as 
\[
\langle [Q,A_r] \ofa, A(G\oS_1^T\oS\tilde{\Pi}f_{\lambda} + Q_1\ofa)\rangle.
\]
Since $Q_1$ and $Q$ have the same principal symbol, $Q_1=Q+B_1$ with $B_1 \in OPS^0$, so this becomes
\[
 \langle [Q,A_r] \ofa, [A_r,G\oS_1^T\oS\tilde{\Pi}]f_{\lambda} +G\oS_1^T\oS\tilde{\Pi} A_r f_{\lambda}\rangle 
\]
\[
+ \langle[Q,A_r] \ofa, ([A_r, Q] + [A_r,B] + Q_1A_r)\ofa\rangle
\]
The last term is $O(\lambda^{2k})$ thanks to Lemma \ref{thm:ofasm} with $s=1$. To bound the second, write 
\[
[A_r,B] \ofa = [A_r,B] (\ofa - \Mld M\tilde{P}f_{\lambda}) + [A_r,B] \Mld M\tilde{P}f_{\lambda}
\]
The first summand is $O(\lambda^{2k})$ thanks to Lemma \ref{thm:ofaerr}. Since $[A_r,B]  \in OPS^{-1}$, the defining property of the high frequency wavelet family yields
a bound $O(\lambda^2)$ fo the second summand.

The upshot can be stated:
\begin{equation}
\label{eqn:term1}
D\langle [Q,A_r] \ofa, A_r\ofa \rangle = -\langle [Q,A_r]\ofa,[Q,A_r]\ofa \rangle + O(\lambda^2)
\end{equation}

\subsection{Term 2}
Note that $B$ might as well be regarded as symmetric.
\[
D\langle BA_r\ofa,A_r\ofa \rangle = \langle DB A_r \ofa, A_r \ofa \rangle + \langle B A_r D \ofa, A_r \ofa \rangle
\]
As $DB$ is bounded and $\|A_r \ofa\| = O(\lambda^{2k})$, the first term is $O(\lambda^{2k})$. The second term is
\[
= \langle B A_r (G \oS_1^T \oS\tilde{P}f_{\lambda} + Q_1\ofa),A_r \ofa \rangle
\]
\[
= \langle BA_r G \oS_1^T \oS\tilde{P}f_{\lambda}, A_r \ofa \rangle
+ \langle BA_r Q_1 \ofa, A_r\ofa \rangle.
\]
The first summand in this expression is $O(\lambda^{2k})$, any $k$, thanks to Lemma \ref{thm:ofasm}. The second term is an $L^2$-bounded quadratic form:
\[
\langle BA_r Q_1 \ofa, A_r\ofa \rangle = \langle [BA_r, Q_1]\ofa, A_r\ofa\rangle + \langle Q_1 BA_r \ofa, A_r \ofa
\]
Lemma \ref{thm:ofasm} again applies to both the first term ($s=0$) and the second ($s=1$).

Putting all of this together,
\[
D\langle BA_r\ofa,A_r\ofa \rangle = O(\lambda^{2k})
\]

\subsection{Term 3}
\[
D\langle \ofa,\ofa \rangle = 2\langle G\oS_1^T \oS \tilde{P}f_{\lambda} + Q_1\ofa,\ofa\rangle 
\]
\[
= 2\langle G\oS_1^T \oS \tilde{P}f_{\lambda},\ofa\rangle + \langle (Q_1+Q_1^T)\ofa,\ofa \rangle
\]
Since $Q_1$ has the same principal symbol as $Q$, it is essentially skew-adjoint, so the second term here is also $L^2$-bounded.

Since the coefficient of this term is $O(\lambda^2)$, that is its bound.

\subsection{Term 4}
\[
D \langle R\ofa,   \oS_1^T(\oS\ofa -d) \rangle = \langle DR \ofa, \oS_1^T(\oS\ofa -d) \rangle
\]
\[
+ \langle R D\ofa, \oS_1^T(\oS\ofa -d) \rangle + \langle R\ofa, (D\oS_1)^T(\oS\ofa -d) \rangle
\]
\[
+\langle R\ofa,\oS_1^T D\oS\ofa \rangle + \langle R\ofa, \oS_1^T \oS D\ofa \rangle
\]
The five summands are all $O(\lambda^2)$, but for different reasons. The first is the simplest: $DR$ is also bounded, so Lemma \ref{thm:dataerr} shows that this term is $O(\lambda^2)$. The second can be re-written
\[
\langle R(G \oS_1^T \oS\tilde{P}f_{\lambda} + Q_1\ofa, \oS_1^T(\oS\ofa -d) \rangle
\]
In this expression, the first summand is $O(\lambda^{2k})$, any $k$, since $\mbox{ess supp }R \subset \gamma_s \setminus \tilde{\gamma}_s$ and $\mbox{ess supp }\tilde{P} \subset \tilde{\gamma}$.

The third is also straightforward: it is
\[
= \langle R\ofa,Q^T(\Pi^1_s)^T\oS_1^T (\oS\ofa -d)
\]
Lemma \ref{thm:dataerr} applies again: since $\oS_1^T$ is bounded from $H^s(\Gamma_r \times \bR)$ to $H^s(\Gamma_s \times \bR)$ for any $s \in \bR$, in particular for $s=1$, it is clear that this quantity is $O(\lambda^2)$.

The fourth and fifth terms can be handled together:
\[
\langle R\ofa,\oS_1^T D\oS\ofa \rangle + \langle R\ofa, \oS_1^T \oS D\ofa \rangle
\]
\[
=\langle R\ofa,\oS_1^T (\oS Q + \oS_1 R)\ofa \rangle + \langle R\ofa, \oS_1^T \oS (G \oS_1^T\oS \tilde{P}f_{\lambda} + Q_1\ofa) \rangle
\]
From Lemma \ref{thm:ofaerr}, it is possible to replace $\ofa$ with $\Mld M \tilde{P}f_{\lambda}$ at the cost of an $O(\lambda^{2k})$ error in $H^s$, any $k$ and $s$. Having done so, the non-intersection of the essential supports of $R$ and $\tilde{P}$ together with the defining properties of the family $\{f_{\lambda}\}$ implies that the whole mess is $O(\lambda^{2k})$, hence Term 4 is altogether $O(\lambda^2)$.

\subsection{Term 5}
This one is $O(\lambda^{2k})$ any $k$ thanks to the presence of the smoothing operator $E$, which remains smoothing after differentiation, and reasoning of the sort that has become tiresomely familiar at this point.

Taking these results together, we see that the only contribution to $D^2\Jared$ that does not vanish like $\lambda^2$ resides in Term 1:

\begin{theorem}
\label{d2j}
Suppose that $d=S\tilde{\Pi}f_{\lambda}$ is $r$-approximately point and $\{f_{\lambda}\}$ is a high-frequency wavelet family. Then
\[
D^2\Jared = \|[Q,A_r]\ofa\|^2 + O(\lambda^2)
\]
\end{theorem}

\begin{remark}
As mentioned in the introduction, the key assumption underlying this result is the tying together of regularization and frequency content.
\end{remark}

\section{Symbols}

$Q$ is a pseudodifferential operator (``$\Psi$DO'') 
\[
Q\of(t,x) = \int dk d\omega e^{i(kx+\omega t)}q(x,t,k.\omega) \hat{\of}(k,\omega)
\]
{\em Symbol} $q(x,t,k.\omega)$ grows like poly as $|k|,|\omega| \rightarrow \infty$

{\em Principal symbol} = leading term in frequency of $q$ = 
\[
q_1(x,t,k,\omega) = i\omega \left((D\tau[\kappa]\delta \kappa)(x,x_r)\right)_{x_r=x_r(x,k/\omega)}
\]
$x_r(x,k/\omega)$ = point of arrival of ray on $\{z=z_r\}$ with horizontal takeoff slowness $k/\omega$ on $\{z=z_s\}$

principal symbol of $[A,Q]= -i\{x-x_s,q_1\} $
\[
=\frac{\partial}{\partial \hat{k}}\left((D\tau[\kappa]\delta \kappa)(x,x_r)\right)_{x_r=x_r(x,\hat{k}=k/\omega)}
\]
(= linearized modeling operator of {\em slope tomography})

$\Psi$DO calculus $\Rightarrow$ as $\lambda \rightarrow 0$,
\begin{itemize}
\item $\|[A,Q]\ofa\|=O(1), \|[A^TA,Q]\ofa\|=O(\lambda)$ 
\item $D^2J_{\alpha} \approx \alpha^2 \|[A,Q]\ofa\|^2 = O(1)$
\end{itemize}

... $\Rightarrow$  for $C>0$,

$D^2J_{\alpha}[\kappa](\delta \kappa,\delta \kappa) \ge (C+ O(\lambda)) \times \left\|\frac{\partial}{\partial \hat{k}}(D\tau[\kappa]\delta \kappa)\right\|^2 $

{\color{blue} To 2nd order at consistent (model, data), SSE inversion $\approx$ traveltime tomography - $J_{\alpha}$ convex in ball of radius $O(1)$ in ``tomographically detectable'' subspace - {\em much} larger region of convexity than FWI}

### 2019-10-19 repro/papers/struct/paper.tex
repro docs dir added

\title{IWAVE Structure and Basic Use Cases}
\author{William. W. Symes}
\address{The Rice Inversion Project, Rice University,
Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}

\maketitle
\parskip 12pt

\lefthead{Symes}

\righthead{IWAVE Structure}

\begin{abstract}
  The IWAVE control structure facilitates construction of wave
  simulators with flexible specification of input and output. This
  document describes synthesis of seismograms and wavefield movies
  from initial data and from single and multiple sources (right-hand
  sides), and linearized (``Born'') and linearized adjoint
  (reverse time migration) modeling. The choice of physical model and
  simulation method - constant density acoustics with Dirichlet
  boundary conditions and $(2,2k)$ finite difference schemes - is the
  simplest possible, but the framework accommodates any regularly
  gridded stencil-based discretization of arbitrary wave physics in the same way.
\end{abstract}

\inputdir{project}

\section{Introduction}
IWAVE combines minimal physics- and scheme-specific code with
a common base of memory allocation, SPMD communication and other forms
of parallelism, i/o, and
job control code to produce complete wave modeling applications. This
paper explains what code must be written to implement a wave modeling
application in IWAVE, and how IWAVE makes various modeling options
available via a set of examples. It ends with a discussion of various
other extensions and capabilities, implemented or planned, including
various forms of inversion.

The next section describes the general organization of the IWAVE core
code and the additional code that must be added to form an
application. Following this generic structural description, I explain
how to implement it in the case of the simplest useful example,
constant density acoustics.

Some design aspects of IWAVE arise from the intended use cases not
discussed here: parallel processing of shots, subdomains, and loops,
and interface with an object oriented optimization package, the Rice
Vector Library, to facilitate inversion applications. These use cases
and their design implications will form the subjects of other reports.

\section{Structures and Interfaces}
Three components and their relations define a time-stepping simulator
based on regular gridding and stencil field updates:
\begin{itemize}
\item a collection of (discrete) dynamic and static (coefficient) fields;
\item model-specific functions implementing time step stencils, 
  determining space and time grids and stencil shape, and
  sanity-checking coefficient fields;
\item a list of i/o procedures to populate the fields at initial and
  intermediate times, and to extract results at final and intermediate
  times.
\end{itemize}
Definitions and interfaces for these three components form the core of
IWAVE. The critical functions and data structures are static members of the {\tt IWaveInfo} class;
definining them completely defines the application. Declarations
appear in \\{\tt iwave/core/include/iwaveinfo.hh}, which should be
consulted for function signatures etc. Applications must supply
definitions, each unique (as is requlred for static global data), as
described in the final paragraphs of this section.

\subsection{Fields}
The top level data structure characterizing an application implemented
in IWAVE is a list of field keywords and other attributes. Each field
gets a keyword, acting as a mnemonic index for internal and external
reference. Fields are dynamic or static, and may be primal or dual in
each coordinate axis: ``dual'' here refers to the grid, so dual fields
represent values assigned to edges, faces, or volumes, rather than
grid vertices (nodes). IWAVE encodes these boolean attributes as 0's
or 1's. Finally, a bit of information about the scheme intrudes: each
dynamic field is updated in one of the (sub)steps of a (possibly)
multistep method, and for that substep, and that substep only, needs
ghost cell data exchanged.

The {\tt FIELD} struct containing this information is organized in the
order 
\begin{itemize}
\item (string) keyword;
\item (int) dynamic flag;
\item (int) substep number;
\item (int array) duality flag for each axis.
\end{itemize}
For instance, a $(z,x)$ stress component field for a staggered grid elastic
finite difference method is a dynamic fielld, representing a grid
sampled centered in the $(z,x)$ faces of the grid cells, and 
updated in second substep of each time step \cite{moczoetal:06}. The
{\tt FIELD} struct
\begin{verbatim}
{"szx", 1, 1, {1, 1, 0}}
\end{verbatim}
captures this information (here coordinate order is $(z,x,y)$).. Note that the only arbitrary
choice here is the keyword string. The choice of keyword for each field must be
consistent throughout the code.

The distinction between static and dynamic fields is more than a
convenience. Static fields, representing the coefficients of the model
differential equations, must exist in the simulation environment prior
to simulation, in some form of persistent store. Simulation input and output
fields, defined below, must also exist as persistent store prior to
simulation. The current
implementation of IWAVE presumes that ``persistent store'' is a
synonym for ``disk file'', however the logic is simply that the data
for these fields should exist outside of the simulation scope. Future
versions of IWAVE may accommodate distributed data as persistent
store, for instance. Dynamic fields exist only within the scope of the
simulation: IWAVE creates and destroys them in the course of a run.

By convention, the first field listed in the application {\tt FIELD}
array is the source of the primary simulation grid, to which all other
grids are referenced. Since this
information will need to enter the compuation via i/o, this first
field should be static (i.e. a coefficient), which will exist in the
simulation environment prior to execution of the application.

The static array {\tt iwave\_fields} data member of the {\tt
  IWaveInfo} class lists the static
and dynamic fields of an IWAVE
application and their top-level attributes using the {\tt FIELD} struct
explained above. Being static, {\tt iwave\_fields} must be
initialized once, and only once, somewhere in global namespace. That
is, the application author must supply a definition
\begin{verbatim}
iwave_fields = { ... };
\end{verbatim}
somewhere - the preferred location for this definition is in the
model definition header file containing the other required definition to
be reviewed below. This model definition file should be included {\em
  only} in main program source files, as its static declarations must
appear only once in program text.

\subsection{Field allocation}
IWAVE stores all fields as {\em
  rarrays} (or {\tt RARR}s). {\tt RARR} is an intelligent array type
that defines gridpoints by reference to global coordinates in ${\bf
  Z}^d$ for $d$-dimensional simulations, provides both one- and
multi-dimensional access to array data, and includes many utiliities
for data manipulation. The time step function implementing the dynamic
field updates will be written in terms of {\tt RARR} utilities and
data members. The {\tt RARRs} corresponding to the application's
fields together form an {\rm rdomain} ({\tt RDOM}), which is simply an
array of {\tt RARR}, indexed in the order that they appear in the {\tt
  iwave\_fields} array.

Derivatives (and adjoint derivatives) of the simulation map involve
multiple copies of the basic application fields: the first derivative
needs two copies, a set of reference fields and a corresponding set of
perturbation fields. Since the $k$th derivative is the derivative of
the $k-1$st derivative, the $k$th derivative
requires $2^k$ sets of fields in its definition. The {\tt IWaveTree}
data structure of order $k$ is a vector of $2^k$ {\tt RDOM}s suitable
for defining the $k$th derivative; the first $2^{k-1}$ {\tt RDOM}s
comprise the reference {\tt IWaveTree}, the second form the
perturbation.

\subsection{Functions}
The implementation of an IWAVE application depends on a half-dozen
functions, which the application author must supply. The most obvious
of these is the time step, or dynamic field update, function, and it
is described here. All six essential functions are documented in 
\begin{verbatim}
RSFSRC/trip/iwave/core/include/iwinfo.hh
\end{verbatim}

The time step interface accommodates (in principle) the update
functions for all orders of derivative, and adjoint (reverse mode) as
well as forward time stepping. Its signature is encapsulated in a {\tt
  typedef}:
\begin{verbatim}
typedef void (*FD_TIMESTEP)(std::vector<RDOM *> dom, 
                            bool fwd, 
                            int iv, 
                            void* fdpars);
\end{verbatim}
The first argument is the {\tt RDOM} array extracted from an {\tt
  IWaveTree}. The order of derivative to be computed is the base 2 log
of {\tt dom.size()}. The implementation should take the form of a case
list, one major case for each order of derivative implemented. Each
case (except the simulation itself, or order 0) should be divided into
two subcases, one each for forward and adjoint modes, switched by the
second {\tt bool} argument {\tt fwd}. Each subcase is further divided
into sub-sub-cases according to the {\em substep index} {\tt iv}. This
third refinement permits implementation of multistep methods in this
framework. To take a prominent example, staggered grid methods for
elastodynamics, in their basic form, use leapfrog time stepping, 
a two-step methods: velocity fields are updated from stress fields
in the first substep, stress fields from velocity fields in the
second. This subdivision of time steps in this fashion allows for the
simplest coding and reduces the amount of data in ghost cell exchange
for domain decomposition, an advantage for low-latency systems. The
final {\tt void *} pointer is the usual dodge for faking private class
data members in ``object oriented C'': the opaque object passed by
address should be of a type defined for the given application and
encapsulating all information needed to compute the time step, such as
Courant numbers, auxiliary damping arrays for absorbing layers, and so
on. One of the other five functions initializes this object.

\section{I/O}
The {\tt FIELD} array member of the {\tt IWaveInfo} class defines the
physical fields of a modeling application. The function data members
describe the memory allocation and updating of the arrays representing
these fields. It remains to describe initialization and finalization,
that is, how external data is read to and
written from these arrays.

The basic principle of IWAVE i/o is that {\em the structure of a data item
  determines the manner in which it is read/written}. Absent
self-describing i/o formats, this principle implies that the
relevant part of the code implements a case list. Persistent data is
presumed to exist in disk files, or in any case be identified by
single strings. Accordingly the case switch is the {\em filename suffix}. At
this writing, two file data structures are realized: SEGY minus reel
header, or Seismic Unix (SU) format, suffix {\tt su}, and Regularly
Sampled Function format, suffix {\tt rsf}. [Other file or network data
structures may be added to IWAVE later.]

Since the details of i/o are implicit in the choice of data structure,
it remains only to connect external data units with the data arrays
with which they communicate. Since the data arrays are indexed by
keyword, and since several external data objects may communicate with
the same data array, it follows that a keyword must be introduced to
stand as a proxy for
each external data object in the program text. At runtime, the keyword
appears as an index into an associative array describing program
inputs and ouputs; the value associated to each key is a path.

Accordingly, the final data member of the
{\tt IWaveInfo} class is the {\tt IOKEY} array {\tt
  iwave\_iokeys}. {\tt IOKEY} is a typedef for a struct consisting of
\begin{itemize}
\item an external data item keyword
\item a field index
\item a flag to indicate input (1) or output (0)
\item an activity flag, which is key to defining the RVL Operator
  interface (see discussion below) - active arrays (1) define
  components of the operator domain (input) or range (output), whereas
  non-active arrays (0) represent auxiliary inputs/outputs.
\end{itemize}
For example, an input to the pressure Cauchy data for
pressure-velocity acoustics, regarded as auxiliary data, might be
represented by the {\tt IOKEY}
\begin{verbatim}
{"init_p", 1, 1, 0}
\end{verbatim}
(assuming that the pressure array has index 1) whereas the output of the same array data, regarded as defining part
of the range of the simulation operator, might correspond to
\begin{verbatim}
{"movie_p",1, 0, 1}
\end{verbatim}
Note that any sampling in space and/or time may be represented in this
way, and is determined at the time of interaction with the external
data unit. Some details of file structure controlling IWAVE i/o are
discussed in later in this paper.

The indirection afforded by the external data unit keyword allows the
the same IWAVE-based command to be used for many different
combinations of inputs and outputs. Parameter
pairs {\tt keyword = filename } passed from a driver serve to connect
specific disk files or other data repositories to external data
keywords, hence to read/write operations.

The IOKEYS array defines i/o info for the basic simulation. Keywords
for perturbation fields, used in derivatives and adjoint derivatives, are generated
automatically. For example, if the keyword for density is {\tt rho},
then the keyword for the first perturbation density (input to the
first derivative) is {\tt rho\_d1}, The adjoint output density
is {\tt rho\_b1}. The choices of suffix correspond to
those generated by the {\em Tapenade} automatic differentiation
package \cite{TapenadeRef13}. These keywords would be used in as parameter keys in pairs
passed to driver routines, for example, {\tt rho\_d1 = my\_density\_pert.rsf}.

\subsection{IWaveInfo}
Definition of an IWAVE application amounts to initialization, in
global namespace of the static public data members of the {\tt
  IWaveInfo} class:
\begin{verbatim}
  static FIELD iwave_fields[]; 
  static IOKEY iwave_iokeys[];
  static FD_MODELINIT minit;
  static FD_MODELDEST mdest;
  static FD_TIMESTEP timestep;
  static FD_TIMEGRID timegrid;
  static FD_STENCIL createstencil;
  static FD_CHECK check;
\end{verbatim}
These initializations must occur precisely once for each application. The recommended
procedure is to assign all members except {\tt iwave\_iokeys} in
a {\em model definition}  header file, which includes both the declarations of the
static data types (via {\tt \#include "iwinfo.hh"}) and of the
function data members (via {\tt \#include} of the main model header
file).  The {\tt iwave\_iokeys} array, on the other hand, should be
assigned at the top of each command file, after {\tt \#include} of the
model definition file.

This organization allows the model application library (object files
defining various functions) to be linked
without alteration to a variety of
drivers (main program files) implementing different input/output 
choices. I have identified two main use cases:
\begin{itemize}
\item standalone command: a standalone driver such as {\tt acd} (next
  section) must have keys provided for all inputs and outputs
  contemplated. I/O keys for which key=value pairs don't appear
  amongst the parameters passed to the application at runtime are simply
  ignored. So this command can be used for all possibilities covered
  by the chosen I/O keys without any alteration. See the next section
  for illustration.
\item RVL Operator interface, for use in inversion applications: the constructor builds its domain and
  range spaces using the supplied I/O keys - this is the reason for
  the ``active'' flag.  I/O keys flagged as active define components
  of domain and range (product) spaces, with geomtry metadata taken
  from the files which the keys point to via the parameter list. Data
  sources corresponding to these keys must be present in the
  environment and correspond to parameter values. This use case will
  be described in more detail in another report.
\end{itemize}

A negative consequence of IWAVE's reliance on static global
definitions is that only one IWAVE modeling application can be active
in a given process. This disadvantage is offset by two considerations:
(1) in the use cases for which IWAVE was designed, only one modeling
application per process is required; (2) the creation of a new IWAVE
application is as simple as one can imagine, requiring only the
definition of a couple of arrays and half-a-dozen functions with
specified signatures, and their assignment to static data of {\tt
  IWaveInfo}. If in the future interesting use cases arise in which
truly distinct modeling packages must be combined in the same process,
then simple modificiations of the IWAVE overall structure will answer,
with slightly increased burden on the programmer.

\section{Case: Constant-density Acoustics}
A simple example illustrating the framework described above is the
IWAVE implementation of
the constant-density acoustic wave equation with Dirichlet
(pressure-free) boundary conditions, connecting the acoustic
potential field $u({\bf x},t)$ and a right-hand side $f({\bf x},t)$
representing a source of mechanical energy, defined in a spatial
domain $\Omega$ over a suitable time interval,
\[
\frac{\partial^2 u}{\partial t^2} - c^2 \nabla^2 u = f,
\]
\[
u(\bx,0) = u_0(\bx),\,\,\frac{\partial u}{\partial t}(\bx,0) =
v(\bx)\,\, \bx \in \Omega,
\]
\[
u(\bx,t) = 0, \,\bx \in \partial \Omega.
\]
The examples to be discussed use the centered
difference approximation \cite[]{kel76}
\[
u^{n+1} = 2 u^{n} - u^{n-1} + \Delta t^2 c^2 Lu^n + \Delta t^2 f^n
\]
in which $L$ is a regular grid difference approximation to the Laplacian, and $u^n$
represents the array of acoustic potential samples for time $n\Delta t$. The choice of $L$
used below is a sum of centered coordinate second difference
operators of order $2k$, $k=1, 2, 4,...$ resulting in a scheme of
formal order $2$ in time and $2k$ in space. Lax-Wendroff extension to
higher order time approximation fits this pattern also.

Since each array element in $u^{n-1}$ appears exactly once in a loop
through the array, it is possible to store only the two arrays for
time indices $n-1$ and $n$, represented by {\tt RARR}s {\tt up} and
{\tt uc} respectively, and store $c^2$ in the {\tt RARR} {\tt csq}.
The the three-level scheme above becomes
\begin{verbatim}
up = 2 * uc - up + dt2 * csq .* L uc + dt2 * f^n
[swap up, uc]
\end{verbatim}
With the type of discrete finite difference Laplacian described above,
the grids for {\tt uc}, {\tt up}, and {\tt csq} are all primal, and
commensurable. Since {\tt csq} must exist along with all of its
metadata (its grid information, basically) in the scope of the
simulation, it is natural to read the primal grid geometry from it.
Thus an appropriate {\tt iwave\_fields} array for acoustic constant
density modeling is
\begin{verbatim}
FIELD IWaveInfo::iwave_fields[]
= {
  {"csq",    0,    0,  {0, 0, 0}},
  {"uc",     1,    0,  {0, 0, 0}},
  {"up",     1,    0,  {0, 0, 0}},
  {"",       0,    0,  {0, 0, 0}}
};
\end{verbatim}
The last line functions the same way as the traling {\tt NUL} for C
strings, that is, to signal the end of the structure.

Inspection of the pseudo-code above reveals that {\tt up} and {\tt
  csq} need to be available at precisely the same gridpoints, whereas
{\tt uc} must store additional gridpoints around the boundary
(``halo'' or ``ghost'' points) in order that the Laplacian can be
built at the {\tt csq} gridpoints (the ``physical'' grid). So the
memory allocations for {\tt uc} and {\tt up} differ, and the ``swap''
mentioned in the algorithm exchanges only values of the two fields at
physical grid points. The algorithm must be completed with a boundary
loop which updates the non-physical gridpoints of {\tt uc}, for
example with odd reflection implementing a Dirichlet condition.

The design described in the preceding paragraphs is realized in the
IWAVE acoustic constant density package, {\tt
  RSFSRC/trip/iwave/acd}. The standalone executable implementing the
various options provided by IWAVE is also called {\tt acd}. It can be
built as part of a Madagascar top-down build, in which case it shows
up as {\tt RSFROOT/bin/sfacd} and can be referenced as {\tt acd} in
Madagascar {\tt Flow}s, or standalone via invocation of {\tt scons} in
either {\tt RSFSRC/trip} or {\tt RSFSRC/trip/iwave}. In the latter
case, the dependency {\tt RSFSRC/trip/rvl} must be built first. The
standalone build has the virtue of permitting local control of build
environment. The {\tt RSFSRC/trip/admin} directory includes a number
of example configuration scripts for build options - to use, copy one
of these, or create a similar file, as
{\tt config.py} in the root build directory.

The standalone build results in the executable command 
\begin{verbatim}
  RSFSRC/trip/iwave/acd/main/acd.x. 
\end{verbatim}
In the remainder of this paper, I
will refer to this command as {\tt acd.x}. The {\tt SConstruct} file
in the {\tt project} subdirectory of the paper directory is also
configured to use this standalone-built command.

The IWAVE acoustic constant density implementation includes code for
the acoustic simulator and its derivatives (with respect to
velocity-squared) of orders 1 and 2, and
their adjoints, built from a simple numerical kernel (or set
of kernels) of truncation orders in space 2, 4, and 8, and truncation
order 2 in time. My research group has used the Tapenade
\cite[]{TapenadeRef13} automatic differentiation package to produce the
code for derivatives and adjoints. For example, the signature of an
implementation of the $(2, 2k)$ scheme for 3D is
\begin{verbatim}
acd_3d_[2k](float *** uc3, 
            float *** up3, 
            float *** csq3, 
            int * s, 
            int * e,
            ...,
            int * lbc,
            int * rbc);
\end{verbatim}
in which {\tt uc3} is the 3-dimensional array view of the array {\tt
  uc}, and so on; {\tt s} and {\tt e} are the vectors of start and end
indices for the loop over gridpoints; and {\tt ...} stands in for a
list of difference formula coefficients, the the number and value of
which depend on the order ($2k$). The integer arrays {\tt lbc} and
{\tt rbc} flag whether the left and right boundaries of the
computational domain, delimited by {\tt s} and {\tt e}, are external
(physical) boundaries or internal boundaries. In the former case,
phsical boundary conditions must be applied; these are also part of
the code.

Tapenade produces similar code for the first derivative of
this stencil (with respect to the {\tt uc}, {\tt up}, and {\tt csq}
arguments, with signature
\begin{verbatim}
void acd_3d_[2k]_d(float *** uc, 
                   float *** ucd, 
                   float *** up, 
                   float *** upd, 
                   float *** csq,
                   float *** csqd, 
                   int * s, 
                   int * e, 
                   ...,
                   int *lbc,
                   int *rbc);
\end{verbatim}
in which {\tt ucd}, {\tt upd}, and {\tt csqd} are the perturbations of
the arrays without the {\tt d}'s.

These kernels can be folded into an obvious case list, switched by the
inputs to the timestep interface described above.

Tapenade output is not entirely suitable for immediate use: some minor
cleanup is necessary, and any serious optimizations (vectorization,
for instance) will need to be applied in a tuning phase. However the
code as it comes from the package is correct and reasonably readable,
and can serve as a baseline with which to verify tuned versions.

Definition of a command based on the fields and functions described
above requires one more piece of information: the connection of
fields to external data sources and sinks, intermediated by i/o
keywords. Many choices are possible; one reasonable choice for the 
standalone command option and constant density acoustics ({\tt acd.x})
is:
\begin{verbatim}
IOKEY IWaveInfo::iwave_iokeys[]
= {
  {"csq",    0, true,  true },
  {"data",   1, false, true },
  {"source", 1, true,  false},
  {"movie",  1, false, false},
  {"initc",  1, true,  false},
  {"initp",  2, true,  false},
  {"",       0, false, false}
};
\end{verbatim}
Clearly the velocity (or rather velocity-squared) must be made
available. Two outputs from {\tt uc} are identified, {\tt "data"}
and {\tt "movie"}: while nothing about the specs demands this usage,
the first is intended for trace output, the second for time slices,
as the keywords choices are intended to suggest. Since the precise
mechanism of I/O is inherent in the data unit (file structure, for
instance) rather than the directed by the code, in fact these mnemonic
suggestions could be ignored, and {\tt "data"} used to store a
movie, for example. However it is an intended use case that movies
might be generated at a byproduct of trace generation, so two output
slots are provided. Similarly, several input keywords 
suggest a right-hand side input (time dependent force divergence
traces) ({\tt "source"}) and Cauchy data ({\tt "initc"}, {\tt "initp"}).

Note that the discrete Cauchy data represent pressure at two successive time levels,
whereas the natural Cauchy data for the wave equation would provide
presure and its time derivative. An application accepting this natural
Cauchy data would need to pre-process it into discrete Cauchy data as
indicated above. I have elected to ``un-bundle'' this type of
pre-processing, that is, it is not included in the IWAVE code
itself. Similarly, the natural SEGY representation of the RHS source
traces needs to be pre-processed to code the source positions as
receiver coordinates, as reviewed below.

\subsection{Single Shot Examples}
The examples presented here are built in the {\tt project} subdirectory of the
paper directory, and organized in standard Madagascar
fashion with SConstruct \cite[]{Madagascar}. Details of the build are best
appreciated by reading the {\tt SConstruct} script in {\tt
  ./project}. All use the standalone build of the acoustic constant
density command {\tt acd.x} (in {\tt
  RSFSRC/trip/iwave/acd/main}). This command self-docs: all of the
parameters described below, and others not used in these examples, are
described in the self-doc, which the reader should consult in
conjunction with this discussion.

All of these examples use the simple layered velocity
(-squared) model depicted in Figure \ref{fig:csq-4layer}. 
This data derives from a corresponding velocity model, fetched from
and squared to create a velocity-squared field.

\plot{csq-4layer}{width=15cm}{Four layer deep-water cartoon - quantity
  plotted is velocity squared.}

\noindent \paragraph{Movie from Initial Data.} 
Figure \ref{fig:gauss}
displays a Gaussian initial field, input as the initial potential
field {\tt uc}.  Since  {\tt up} is not input, it remains at its
default value of zero. The resulting IWAVE solution is not convergent
to the solution of any particular Cauchy problem with full convergence
order (2), but produces a reasonable-looking movie. Another (minor)
application would produce a correction to be added to the {\tt up}
field to restore 2nd order convergence.

\plot{gauss}{width=15cm}{Gaussian initial datum}

The data depicted in Figures \ref{fig:csq-4layer} and \ref{fig:gauss}
is input to the simulation, so clearly must exist prior to
simulation. However the output must also exist: IWAVE I/O, both reads
and writes, is driven by the target data structure. Therefore the
movie output file must be constructed before the simulation fills it
with data. The {\tt SConstruct} script contains an invocation of the
{\tt sfmakevel} command which creates a 3D {\tt rsf} file {\tt
  movieinit.rsf}. On completion of the command, this file holds the
movie output.

Perusal of this command reveals some customization of the {\tt rsf}
file format, as compared to its standard use (\cite{Madagascar}). The
duration of the movie determines the duration of the simulation: the
initial simulation time is the time of the initial movie frame, and
similarly for the final time. Thus IWAVE must be able to determine
which axis specified in {\tt movieinit.rsf} is the time axis. Three
additional header word categories, beyond those of the {\tt rsf}
standard, make this feat possible:
\begin{itemize}
\item {\tt dim} = spatial dimension - should be same as spatial
  dimension of the primary static file (keyword {\tt csq} in this
  case).
\item {\tt gdim} = global dimension - of the data array, at least {\tt
    dim}
\item {\tt id1, id2, id3,...}: identification tags of axes 1, 2, 3,...
\end{itemize}
By IWAVE convention, axis with tag = {\tt dim} is the time axis, and axes 0,
1,..., {\tt dim}-1 are the spatial axes. More than {\tt dim} axes are
permitted - they are used to represent other acquisition or modeling
parameters, see section on multi-shot simulation, below.

In this example, the space dimension is 2, so {\tt id3=2} indicates
that the 3rd axis is the time axis.

After propagating 5.12 s and interacting with both the
reflecting (Dirichlet) boundaries and the interfaces in the model, the
potential field becomes that depicted in Figure
\ref{fig:frameinit}.

\plot{frameinit}{width=15cm}{Acoustic potential field at
  5.12 s, resulting from Gaussian initial data}

Parameters passed to the command {\tt acd.x} included 
\begin{verbatim}
           csq = ../csq_4layer.rsf
         initc = ../init.rsf
         movie = ../movieinit.rsf
\end{verbatim}
Keywords {\tt data} and {\tt source} were ignored. Many other
paramters were required; a brief description of these is included in
the self-doc of the command {\tt acd.x}.

Note that the pathnames refer to the directory level above the working
directory. IWAVE produces various diagnostic output at runtime,
switched by various flags passed as parameters. These outputs, and
possibly other auxiliary outputs of commands built upon IWAVE (eg. the
data residual in an inversion) vary with application and data, so are
inconvenient to specify individually as cleanup targets. Instead, the
SConstruct script creates a working subdirectory and executes (and
dumps its auxiliary output) there. The entire directory is cleaned up
by {\tt scons -c}. So the correct parameter specification for archival
input and output files is one directory level up.

\noindent \paragraph{Movie from Point Source.} 
The pulse in Figure \ref{fig:wavelet} the derivative of a 5 Hz
Gaussian; it is embedded in a space-time field via the {\tt
  iwave/trace/main/towed\_array.x} utility, and used as the RHS in the
wave equation. The resulting field (starting from homogeneous Cauchy
data) at 5.12 s is depicted in Figure \ref{fig:framesrc}.

\plot{wavelet}{width=15cm}{Derivative of 5 Hz Gaussian}

\plot{framesrc}{width=15cm}{Acoustic potential field at
  5.12 s, resulting from Gaussian derivative point source pulse
  located at x=12 km, z = 1.875 km.}

In this case, the parameters passed to {\tt acd.x} (in addition to the
other required numerical parameters) are 
\begin{verbatim}
           csq = ../csq_4layer.rsf
        source = ../wavelet12000.su
         movie = ../movieinit.rsf
\end{verbatim}
The point source data are presented as SEGY traces, rather than as an
RSF data file; the latter would in principle also be possible, but at
the time of this writing RSF trace I/O is not enabled.

\noindent \paragraph{Trace Data from Point Source}
The same point source as in the previous example produces sthe traces
at 6 m depth depicted in Figure \ref{fig:shot12000}. The sampling is
also pure point, that is, does not filter the potential field in any
way. So this trace data could be interpreted as the 2D pressure
traces of a point source with 5 Hz Gaussian pulse, for example.
In this case, the required parameters are
\begin{verbatim}
           csq = ../csq_4layer.rsf
        source = ../wavelet12000.su
          data = ../shot12000.su
\end{verbatim}
As noted before, there is nothing sacred about the string {\tt data} -
any keyword identified in the I/O key struct as output from {\tt uc}
would do just as well. This key is provided for mnemonic convenience.
Note that the file data structure ({\tt shot12000.su}) entirely
determines the  {\em way} in which the data is written - sampling,
source and receiver geometry are all taken from this file.

\plot{shot12000}{width=15cm}{Response of seafloor pressure sensor at
  1.875 km depth, to 400 shots spaced 25 m apart at 6 m depth, source is isotropic point
  radiator, source pulse is Gaussian
  derivative with peak frequency 5 Hz,}

\noindent \paragraph{Born Approximation}
The linearization of the acoustic field with respect to the wave
velocity is the solution $\delta u$ of
\begin{eqnarray}
\frac{\partial^2 u}{\partial t^2} - c^2 \nabla^2 u & = & f,\\
\frac{\partial^2 \delta u}{\partial t^2} - c^2 \nabla^2 \delta u & = &
\frac{2\delta c}{c^3}\frac{\partial^2 u}{\partial t^2},
\end{eqnarray}
\[
u(\bx,0) = u_0(\bx),\,\,\frac{\partial u}{\partial t}(\bx,0) =
v_0(\bx)\,\, \bx \in \Omega,
\]
\[
\delta u(\bx,0) = 0,\,\,\frac{\partial \delta u}{\partial t}(\bx,0) =
0\,\, \bx \in \Omega,
\]
\[
u(\bx,t) = \delta u(\bx,t) = 0, \,\bx \in \partial \Omega.
\]
As explained above, the IWAVE framework provides approximations for
computing the linearization (widely called the ``Born approximation'',
even though this is strictly a misnomer), along with its adjoint and
higher derivatives.

The parameter key {\tt deriv} flags the computation, or not, of
derivatives. The value assigned is the order of derivative, with
default 0. Each input perturbation (representing quantities such as
$\delta c$ in the linearized system of PDEs, above) is assigned a key
equal to the key for the unperturbed quantity with {\tt \_d1} appended
(for the first derivative - higher derivatives require multiple input
perturbations, keys for which have {\tt \_d2},
{\tt \_d3},... appended). Output keys remain the same as for the reference
computation.

For the acoustic constant density application, Born approximation
requires reference and perturbation square velocity fields. Figures
\ref{fig:dcsq-4layer} and \ref{fig:csq-4layersm} show perturbation and
reference square velocity fields, respectively, that will generate
Born data roughly corresponding to the preceding example. The required
parameters are
\begin{verbatim}
         deriv = 1
           csq = ../csq_4layer.rsf
        csq_d1 = ../dcsq_4layer.rsf
        source = ../wavelet12000.su
          data = ../born12000.su
\end{verbatim}
The linearized response (Born modeling) corresponding to Figure 
\ref{fig:shot12000} appears as Figure \ref{fig:born12000}.

\plot{dcsq-4layer}{width=15cm}{Velocity-squared perturbation -
  localized oscillations at layer boundaries}
\plot{csq-4layersm}{width=15cm}{Smooth velocity-squared obtained from
  velocity of Figure \ref{fig:csq-4layer} by filtering with a cubic spline window.}
\plot{born12000}{width=15cm}{Linearized response (``Born modeling'') of seafloor pressure
  sensor, due to perturbation (Figure \ref{fig:dcsq-4layer}) about
  smooth background (Figure \ref{fig:csq-4layersm}); other parameters
  as in Figure \ref{fig:shot12000}.}

\noindent \paragraph{Reverse Time Migration}
One version of Reverse Time Migration (RTM) is simply the adjoint of
Born modeling. IWAVE provides adjoint computations for every
derivative mapping (first, second,...) using the {\em optimal
  checkpointing} method of reverse time propagation
\cite[]{Griewank:book,BlanVerSy:95,Plessix:06,Symes:06a-pub}. Other approaches to time reversal can be more efficient in
special cases, especially when the interior dynamics are conservative
(acoustics, elasticity) \cite[]{Dussaud:08,Clapp:09}. However
none are more effective in general, in particular when energy
attenuation is significant part of wave dynamics, as is the case for
all realistic models of seismic wave motion.

Figure \ref{fig:migr12000} displays the migration of the single
Born ``shot'' gather (really, OBS receiver gather) located at $x_s = 12000$ m from
the left edge of the model. No effort has been made to remove the
low-frequency noise caused by the sea bottom reflection.

\plot{migr12000}{width=15cm}{Reverse-time migration of Born data
  from Figure \ref{fig:born12000}.}

The parameters required for this job are
\begin{verbatim}
         deriv = 1
       adjoint = 1
        nsnaps = 10
           csq = ../csq_4layer.rsf
        csq_b1 = ../migr12000.rsf
        source = ../wavelet12000.su
          data = ../born12000.su
\end{verbatim}

The {\tt adjoint} key flags the adjoint computation. The checkpointing
algorithm requires allocation of workspace for checkpoints (copies of
wavefield Cauchy data, consisting of all dynamic arrays).  The number
of checkpoints allocated is the value for key {\tt nsnaps}. The
appropriate number of checkpoints depends on the number of time
steps. Reasonable numbers to achieve a cost ratio of adjoint to
forward computations of around 5 are
\begin{itemize}
\item up to 1000 time steps: 5 checkpoints
\item up to 5000 time steps: 10 checkpoints
\item up to 10000 time steps: 20 checkpoints
\item up to 20000 time steps: 30 checkpoints
\end{itemize}

\subsection{Interlude: The Internal Grid}

The basic control structure of IWAVE is its {\em internal grid}, which
describes the global state space of IWAVE simulations. This
rectangular grid is logically equivalent to a tuple of axes; each axis
is regularly sampled, hence described by the usual $(n,d,o)$ triple
familiar from RSF. [In common with conventional usage for this
subject, ``axis'' in this discussion will mean ``finite regularly
sampled interval on an axis''.]

Like RSF, the restriction to regular sampling implies some workarounds
for cases in which data sampling is actually not regular - the notable
instance being trace acquisition geometry. Irregular
sampling parameters find a natural home in the attributes of {\tt IWaveSampler}
i/o objects, which encapsulate all of the information required for i/o
of a given type or format. A reference to the internal grid is an key data member of
every {\tt IWaveSampler} instance.

The spatial axes are obvious members of the internal grid, as is the
time axis. The {\tt IWaveSim} constructor creates a version of the
time axis that combines the various time axes of associated data
objects that have them, and resets the time step to one appropriate
for the simulation (returned by the {\tt TIMESTEP} member function of
{\tt IWaveInfo}). Thus for example the source time axis may even
end earlier than the output trace time axis begins; the internal time
axis contains the convex hull the two, resampled to the simulation time
step.

Other axes represent other aspects of simulation - most important,
other sampled parameters such as source positions that parametrize
individual shots within a simulation. The internal grid allocates
additional axes to record these parameters. For instance, SEGY data is
organized shot coordinates, on a {\em shot axis}. All coordinates are
recorded in the {\tt IWaveSampler} for SEGY data, but only one
additional axis with stride $1$ is introduced to describe the various
shots.

Additional axes may represent internal degrees of freedom available to
the simulation, for instance space- or time-shift simulation \cite[]{geoprosp:2008}. The use of these internal degrees of freedom will be described
in another report.

With many axes interacting, it is essential to provide some method to
determine their roles. Other data handling systems either carry out
this task in a similar way to that described here (eg. DDS,...) or
ignore it (not an option for IWAVE!). The additional information is
arranged as additional keywords in RSF data files, mentioned earlier
in the discussion of movie output:
\begin{itemize}
\item {\tt dim} = spatial dimension of the simulation
\item {\tt gdim} = global dimension = dimension of internal grid
\item {\tt id1, id2, id3,...} = axis identifiers, coded as follows:
\begin{itemize}
\item {\tt 0,...,dim-1}: spatial dimensions corresponding to these
  choices in SEGY coordinates: 0=z, 1=x, 2=y;
\item{\tt dim}: time
\item{\tt dim+1,...,99}: additional axes for various simulation
  parameters (eg. shot index)
\item{\tt 100,...,dim+99}: internal degrees of freedom (extended
  modeling)
\end{itemize}
\end{itemize}
The labeling of spatial axes avoids external data transpose: if the
data for coefficient arrays is organized with $x$ as the fast
variable, for instance ($x$ meaning the coordinate with that name in
the SEGY data standard, for example), then setting {\tt id1=1}
etc. enables correct sampling of traces. The default settings are {\tt
  id1=0, id1=2, id2=3}. [Note that axis keywords are numbered
Fortran-style, whereas axis values are C-style (beginning with
0). This is done to maintain compatibility with RSF header keywords,
which appear to be inherited from SEPlib.]

\subsection{Multi-shot (survey) simulation}

{\tt IWaveSim} loops over any axes beyond the time axis (signified by {\tt
  idxxx=dim} through {\tt idxxx=99}) that {\tt IWaveSampler} objects 
add to the internal grid. In particular, SEGY data file identified as output or
input adds a simulation axis with {\tt idxxx=dim+1} to the internal
grid. The loop over this axis increments when a the keywords {\tt sx},
{\tt sy}, or {\tt sz} change from one trace to the next, signifying a
new shot.

From the user point of view, this means that multi-shot simulation is
automatic: if multiple shots are part of the output data structure,
then all shots will be simulated.

We show several examples that illustrate this data-flow feature of
IWAVE. The examples are identical to those reviewed above for single
shots, except that the additional computation load of multi-shot
simulation suggests the use of parallelism. The parallel features of
IWAVE (parallel shots, parallel subdomains, parallel loops) will be
the subject of a subsequent report. For now, we note that the {\tt
  partask} keyword indicates the number of shots to run in
parallel. The {\tt SConstruct} files for the several multishot
simulations include a line (near the top) to set the variable {\tt
  NP}. If {\tt NP=1}, then the simulations described below are run in
serial  mode. If {\tt NP} is set to a value larger than one, then this value
indicates the number of shots to process in parallel, via a collection
of MPI communicators. Running several shots in parallel requires that
IWAVE be installed with MPI enabled {{\tt IWAVE\_USE\_MPI} defined as a
compiler parameter, see the {\tt README\_INSTALL} file in the top-level
directory. The number of MPI processes assigned (via {\tt mpirun -np}
can be fewer than the number of shots to be simulated - in that case
the simulations run in batches until all shots are completed. Any
unnecessary processes at the terminal stage of the simulation are simply left
idle, so there is no necessary relation between the number of MPI
processes and the number of shots. The {\tt SConstruct} script in the
project directory for this paper uses {\tt mpirun -np NP} to initiate
MPI and assign the number of processes to be used. The follow-on
report will describe the use of IWAVE in a batch environment, for both
parallelization over shots and via domain decomposition.

The next few examples are large enough that completion single-threaded
execution requires perhaps half an hour on a typical (circa 2014) desktop
CPU.  The data displayed were obtained on a typical multicore desktop
machine, using MPI with {\tt NP=6}. These results are precisely the
same as those that are obtained with a single process, but required
less than 4 minutes walltime.

\noindent \paragraph{Simulation}
Recall that output files for IWAVE applications must exist prior to
execution. Therefore, to simulate 12 shots at intervals of 400 m over
the model depicted in Figure \ref{fig:csq-4layer}, the first step is
to create a suitable SEGY data set to hold the output. Therefore the {\tt
  SConstruct} script creates the output file of zero SEGY traces ({\tt
  line8-12km.su}), subsequently filled with samples by IWAVE, all as
part of the same {\tt Flow} command. As note above, the source must
be properly embedded in a space-time field (collection of SEGY traces)
for each shot location. The utility {\tt trace/main/towed\_array.x}
provides a convenient method for combining the position data of a
prototype SEGY trace file with a choice of source array to produce
translations of a source array to correct shot positions. Other
parameters are as in the single-shot case. The result is displayed in
Figure \ref{fig:line8-12km}.

\plot{line8-12km}{width=15cm}{Response of 13 seafloor pressure sensors
  spaced 400 m apart at
  1.875 km depth, to 400 shots spaced 25 m apart at 6 m depth, sources
  are isotropic point
  radiators, source pulse is Gaussian
  derivative with peak frequency 5 Hz. Velocity-squared model depicted
in Figure \ref{fig:csq-4layer}.}

Perusal of the {\tt project/SConstruct} script shows that in the
invocation of {\tt acd.x}, only the filenames have changed. The
dataflow design of IWAVE, in which data objects determine the way in
which they are read/written, implies that the information about
changed source/receiver geometry and data volume does not need to be
explicitly passed to the simulation command.

\noindent \paragraph{Linearized Simulation}
Linearized simulation (Born modeling) works the same way. The data
predicted from the perturbation in Figure \ref{fig:dcsq-4layer} of the smooth
background velocity (squared) in Figure \ref{fig:csq-4layer}, same
source-receiver geometry as in the preceding example, is displayed in
Figure \ref{fig:born8-12km}.

\plot{born8-12km}{width=15cm}{Linearized response of 13 seafloor pressure sensors
  space 400 m apart at
  1.875 km depth, to 400 shots spaced 25 m apart at 6 m depth, sources
  are isotropic point
  radiators, source pulse is Gaussian
  derivative with peak frequency 5 Hz, Background or reference model
  as in Figure \ref{fig:csq-4layersm}, perturbation as in
  Figure \ref{fig:dcsq-4layer}.}

\noindent \paragraph{Adjoint Linearized Simulation}
Adjoint linearized simulation (reverse time migration) of the 12 shot
linearized data (Figure \ref{fig:born8-12km}) produces the image
displayed in Figure \ref{fig:migr8-12km}.

\plot{migr8-12km}{width=15cm}{Reverse-time migration of Born data
  from Figure \ref{fig:born8-12km}. Reflectors are correctly
  positioned and clearly visible, along with migration artifacts
  characteristic of raw OBS migration, for example a free-surface
  multiple image slightly above 5000 m depth.}

\section{Conclusion}
This paper has described the overall structure of the IWAVE framework,
and illustrated a few of its capabilities.

The release of IWAVE described in this paper is 2.0. It implements many use
cases not described here:  
\begin{itemize}
\item multipole source representation - essential to simulate the
  directional nature of field source and receiver arrays;
\item plane wave and other extended sources, including randomized;
\item PML absorbing boundary conditions for acoustic constant density
  simulation, following \cite{GrotePML:10};
\item shot record extension of the acoustic constant density model.
  \cite[]{KerSy:94}
\end{itemize}
All of these simulation modes are as accesssible as easily as those
illustrated in the preceding seciton.

Earlier releases (1.x) of IWAVE included implementations of staggered
grid schemes for acoustics and isotropic linear elasticity
\cite[]{moczoetal:06}. See  \cite{FehlerKeliher:11} for an account of
the role played by the IWAVE acoustic staggered grid code as a QC tool
in the SEAM Phase I project. Revival 
of these applications in the IWAVE 2.0 infrastruction is intended
for a future release.

The QC role in the SEAM project heavily influenced the design of 
IWAVE, particularly the requirement that its performance scale well to
very large simulations and large numbers of threads. All versions of
IWAVE have offered parallelism via domain decomposition: SEAM
simulations typically split simulations involving eight fields and
3000 $\times$ 3000 $\times$ 1500 grid points amongst 1000-4000
processes. The user interface specifies the subdivision of the domain
with a few parameters. Another report will detail domain
decomposition, parallelization over simulations (mentioned above), and
multithreaded parallel loop execution aspects of IWAVE, as well as
performance enhancements implemented in the finite difference kernels.

Another ground-up design goal is to provide a simple interface to
inversion software. \cite{Geopros:11} describe the concept behind
the data interchange mechanism: it is file-based, and motivates the
dataflow design mentioned several times already. Since data objects,
represented as files or collections of files, determine their own i/o
modalities, inversion software can simply communicate pathnames to the
IWAVE interface, rather than some more complex data structure. The
current release couples to (and in fact depends on)
the Rice Vector Library (``RVL'') \cite[]{RVLTOMS}, a
framework for linear algebra and optimization.
IWAVE 2.0 includes full waveform inversion, linearized inversion
(``least squares migration''), and shot record extended linearized inversion or
differential semblance optimization. These and other inversion
applications of IWAVE will be discussed in other reports.

\section{Acknowledgements}
IWAVE has been a team effort: the original design of the core
framework is due to Igor Terentyev, and Tanya Vdovina, Dong Sun, Marco
Enriquez, Xin Wang, Yin Huang, Mario Bencomo, Lei Fu, and Muhong Zhou
have each made major contributions.
Development of IWAVE was supported by the SEG Advanced Modeling (SEAM)
project, by the National Science Foundation under awards 0620821 and
0714193, and by the sponsors of The Rice Inversion Project. The IWAVE
project owes a great deal to several open source seismic software
packages (Seismic Un*x, SEPlib, Madagascar), debts which we gratefully
acknowledge. The author wishes to record his special gratitude to
Sergey Fomel, the architect of Madagascar, for his inspiring ideas and
his generous and crucial help in the integration of IWAVE into Madagascar.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/trans/paper.tex
repro docs dir added

\title{Waveform Inversion via Source Extension: Major Questions and Research Plan}
\author{William. W. Symes \thanks{The Rice Inversion Project,
Department of Computational and Applied Mathematics, Rice University,
Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}

\extrafloats{100}

\lefthead{Symes}

\righthead{Source Extension}

\maketitle
\begin{abstract}
FWI via source extension is a workaround for cycle skipping. For pure transmission (including diving wave) data, a close link to traveltime tomography explains the good results observed in experiments. No such link is known to exist in the case of pure reflection data, and numerical results are mixed. This proposal outlines several numerical experiments that should complement theoretical investigations and decide the effectiveness of these algorithms.
\end{abstract}

\section{Overview}
FWI can be described in terms of a wave operator $L[c]$, depending on an array of space-dependent coefficients $c$, a trace sampling operator $P$, a wavefield $u$, and a source function (of space and time) $f$. The basic FWI problem is: given $d$, find $c$ so that 
\begin{equation}
\label{eqn:fwi}
Pu \approx d \mbox{ and } L[c]u = f.
\end{equation}
Sometimes $f$ is regarded as given, sometimes only some aspects are given (for example, localization at a source point) and others are to be determined as part of the solution. A typical formulation is via nonlinear least squares:
\[
\mbox{choose } c \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2 
\]
usually with some sort of regularization thrown in.

Note that $f$ and $d$ are typically vectors of trace gathers, one for each source position in the survey (or possibly some other parameter, produced by processing - will stick with source position $\bx_s$).

As is well-known, local optimization methods are the only feasible approach given the dimensions of a typical instance of \ref{eqn:fwi}, and those have a tendency to stall due to ``cycle-skipping''. Source extension is one approach to avoiding this problem. It consists in imposing the wave equation as a soft as opposed to hard constraint, by allowing thw source to have more degrees of freedom than is permitted by the basic model. For sake of argument (and because almost all work so far has assumed it), adopt the isotropic point source as the model: for the data gather at source positon $\bx_s$, the source takes the form
\begin{equation}
\label{eqn:ptsrc}
f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s).
\end{equation}
Every method explained below can be altered to accommodate more complex source models.

Source extension replaces $f(\bx,t;\bx_s)$ with an artificial source $g(\bx,t; \bx_s)$ with more degrees of freedom than merely a single common source wavelet $w(t)$ as in definition \ref{eqn:ptsrc}. Assuming that \ref{eqn:ptsrc} is actually correct physics, these extra degrees of freedom have to be suppressed in the optimal solution, via an {\em annihilator} $A$, an operator acting on extended sources $g$, whose null space is precisely the physical sources (of the form \ref{eqn:ptsrc}). Extended source inversion replaces the FWI objective with a penalty function
\begin{equation}
\label{eqn:esi}
\mbox{choose } c, g \mbox{ to minimize } \|PL[c]^{-1}g -d \|^2 + \alpha \|A[g]\|^2 
\end{equation}

The way in which degrees of freedom are introduced, and the choice of annihilator, distinguish the various source extension methods.

\subsection{Wavefield Reconstruction Inversion}
Introduced by \cite{LeeuwenHerrmannWRI:13} and significantly modified by \cite{WangYingst:SEG16}. Extended source space: arbitrary functions of space-time $g(\bx,t;\bx_s)$, one for each source position. Annihilator: 
\begin{equation}
\label{eqn:wri}
A[g] = g-f
\end{equation}
With these choices, WRI takes the form introduced by \cite{WangYingst:SEG16}. The original approach is equivalent: \cite{LeeuwenHerrmannWRI:13} use the dyamic wavefield $u$ as the additional degrees of freedom, and minimize
\[
\|Pu-d\|^2 + \alpha \|L[c]u-f\|^2
\]
Since the wave equation (any reasonable version!) has a unique solution for each right-hand side, replacing $L[c]u$ turns this original form of WRI into the form \ref{eqn:esi}.

Note that in princple at least, the actual source $f$ (positions and wavelet) must be known {\em a priori}.

The Wang-Yingst variant leads to a huge reduction in the number of degrees of freedom: either the original or the modified versions, as originally presented, appear to require storage of an entire space-time source, which is prohibitive for 3D and/or time domain formulation. I will explain in an appendix how the Wang-Yingst approach drastically shrinks the storage requirements of this method, and makes it feasible for time-domain 3D application, which Wang and Yingst and their co-authors have demonstrated.

\subsection{Space-Time Extension}
\cite{HuangSymes:SEG16b}. Differs from WRI by choice of annihilator:
\begin{equation}
\label{eqn:msste}
A[g](\bx,t;\bx_s) = |\bx-\bx_s|g(\bx,t;\bx_s).
\end{equation}
This annihilator merely enforces the position constraint on the source, without relying on {\em a priori} knowledge of the source wavelet. In fact, the source wavelet is a by-product of the minimization \ref{eqn:esi}.

Appears to require storage of full space-time source volume - the trick explained in the appendix for WRI does not work, due the form of the annihilator.

\subsection{Volume Extension}
\cite{HuangSymes:SEG16a}. Assumes that source signature deconvolution has been performed, so that the actual wavelet is (a bandlimited version of) $\delta(t)$. The extended source is an {\em exploding reflector}:
\begin{equation}
\label{eqn:msvol}
g(\bx,t;\bx_s) = h(\bx;\bx_s)\delta(t).
\end{equation}
Uses the same annihilator as the Space-Time Extension, but loses the advantage of solving for the source as by-product - the source must be known, so that decon can be carried out prior to inversion.

However, only requires a spatial storage volume per source position, essentially equivalent to a second data volume, therefore practical for 3D (though I don't believe it has been tried except in 2D).

The variable projection method applied to this extension has essentially exploding reflector inversion (as opposed to exploding reflector migration) as its inner problem, i.e. estimation of $h$ in \ref{eqn:msvol}. The same problem arises in so-called photoacoustic tomography, and has received a lot of attention in the mathematical inverse problems and biomedical imaging literature. In fact there is a quite accurate approximate inverse available for the inner problem, consisting of converting pressure traces to normal velocity traces and back-projecting (this is known as ``time reversal'') in the IP literature).

A disadvantage of this approach, if you want to handle reflections, is that unlike post-stack migration the actual physical wave velocity is used in modeling, so that the the region in which $h$ must be defined may considerably larger than the ordinary modeling domain - reflection two-way time is effectively modeled by one-way time.

The volume extension concept is similar to an idea discussed by \cite{ZhangGao:08}.

\subsection{Surface Extension}
No publication yet - subject of my workshop talk at the 2016 SEG, and GH has described it in talks at Total. Similar to the volume extension, but the source is spread over the source space-time surface, rather than over space at $t=0$. In the simplest case, where the sources all lie on $z=z_s$, the extended source space consists of functions (distributions) of the form ($\bx=(x,y,z)$)
\begin{equation}
\label{eqn:mssur}
g(\bx,t;\bx_s) = h(x,y,t)\delta(z-z_s).
\end{equation}
As for the previous two cases, the annihilator is given by the localization penalty \ref{eqn:msste}. Therefore the source wavelet is a by-product of the inversion, and does not need to be specified {\em a priori}. Like the volume extension, the additional data volume required is on the same order of size as the data.

\subsection{Adaptive Waveform Inversion = Source-Receiver Extension}
 Popularized by \cite{Warner:14}, \cite{Warner:16}, has many antecedents, listed in \cite[]{HuangSymes:17}. Each trace regarded as separate gather, with separate point source waveform. In effect,
\begin{equation}
\label{eqn:awi}
g(\bx,t;\bx_s) = w(t;\bx_s)\delta(\bx-\bx_s) 
\end{equation}
with repeated occurances of $\bx_s$ regarded as distinct. Usually written by making the wavelet a function of the receiver position also, to distinguish redundant source positions:
\begin{equation}
\label{eqn:awi-alt}
g(\bx,t;\bx_r,\bx_s) = w(t;\bx_r,\bx_s)\delta(\bx-\bx_s) 
\end{equation}
hence the alternate name.

Various annihilators have been suggested - see references. Warner proposes a variant (``AWI'')  that (in principle) requires the source wavelet to be known, and deconvolved from the data traces. Then uses multiplication by $t$, which annihilates $\delta(t)$:
\begin{equation}
\label{eqn:awi-ann}
A[g](\bx,t;\bx_r,\bx_s) = t g (\bx,t;\bx_r,\bx_s)
\end{equation}
AWI distinguished from earlier variants by normalizing the objective with the estimated wavelet $L^2$ norm:
\begin{equation}
\label{eqn:awi-obj}
\mbox{choose } c, g \mbox{ to minimize } \sum_{\bx_r,\bx_s}\frac{\int dt\,|t|^2 |w(t;\bx_r,\bx_s)|^2}{\int dt\,|w(t;\bx_r,\bx_s)|^2}
\end{equation}
Note the order of summation and normalization: each trace is normalized independently (this is not clear from the notation in \cite[]{Warner:16}).

The justification for normalization given by \cite{Warner:16} is that it avoids a familar source/reflector ambiguity. Actually it is more important than that, and plays an important role when no reflectors are present. It is possible to recast the objective \ref{eqn:awi-obj} as equivalent to \ref{eqn:esi} with a weighted norm used in the latter. [I am writing this up - take my word for it!] The weighted norm makes the linear map from source to data {\em unitary} (at least to leading order in frequency), and that is an essential property in relating AWI to traveltime tomography.

\section{What is known}
For several of the source extension algorithms, the objective is known to be closely linked to traveltime tomography in one form or another, and to inherit its convergence properties. These connections are known {\em only} for pure transmission, and not always then. Nothing theoretical is known about any of these methdods for reflection data.

The connection to traveltime tomography was first (so far as I know) established for the source-receiver extension in several TRIP technical reports related to Hua Song's PhD thesis \cite[]{Song:94c,SoSy:92,SongSymes:94a,SongSymes:94b,Symes:94c}. This work established a form of differential semblance velocity estimation for crosswell tomography, and the related tomographic objective was a form of slope, or stereo-, tomography. The time moment objective \ref{eqn:awi-obj} (without normalization) first appears in \cite[]{Plessix:00a,Plessix:00}. A similar idea appears in \cite[]{LuoSava:11}. Neither of these refferences made an explicit connection to tomography - that was done in \cite[]{HuangSymes2015SEG,HuangSymes:17}.

The defect, in the relation between AWI/source-receiver extention and traveltime tomography, is that it collapses when the ray field (must be able to speak of rays, to invoke traveltime tomography) develops caustics. This was already demonstrated both theoretically and by numerical example in \cite[]{Symes:94c}, though the fact does not seem to have percolated very far. \cite{HuangSymes:17} give an explicit illustration of this failure in a diving wave setting. As soon as the diving wave field is triplicated, AWI is just as bad as FWI.

On the other end of the list, almost nothing is known about why WRI converges, or when it does. It may be possible to repeat a version of the analysis mentioned next for the volume and surface extensions, based on geometric optics and the Wang-Yingst variant as described in the appendix, though that has not been tried yet.

The volume and surface extensions have a well-defined relation to tomographic principles (slope or traveltime, depending on type of annihilator). This relation holds regardless of ray geometry (with or without triplication) in transmission configuration.

Again, nothing is known theoretically about the behaviour of any of these methods for reflection data, and numerical evidence is ambiguous.

\section{Proposed Project}.

Amongst the source extension options described in the previous section, only the surface extension has all of these properties:
\begin{itemize}
\item requires only additional storage roughly equivalent to the data volume
\item does not require prior knowledge of the source wavelet, in fact recovers an estimate as a by-product
\item inner problem (in variable projection - estimation of the extended source) has a cheap approximate solution, via back-propagation
\item relation with traveltime inversion persists, regardless of ray geometry
\end{itemize}

Therefoire I propose that we begin with a time-domain implementation of the surface source extension.

\section{Detailed Description}
The simplest version of the surface source extension assumes acoustic wave physics, and that the physical source locations lie on the horizontal surface $z=z_s$. The coefficient vector $c$ is the pair $(\kappa,\beta)$ of bulk modulus and buoyancy, and the wave operator $L[c]$ is given by the linear acoustics system:
\begin{eqnarray}
\label{eqn:awe}
\frac{\partial p}{\partial t} & = & - \kappa (\nabla \cdot \bv +
w(t)\delta(\bx-\bx_s)), \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t<0\nonumber\\ 
\bv & = & 0 \mbox{ for } t<0 
\end{eqnarray}
The physical source is modeled as an isotropic point constitutive law defect. As mentioned earlier, it is simple to accommodate more complex source models.

The dynamical field is the 4-tuple $u=(p,\bv)$. Either the pressure $p$ or the particle velocity $\bv$ could be sampled (or both) - presume that the pressure is sampled, that is $Pu = \{p(\bx_r,t;\bx_s)\}$, where $\{\bx_r\}$ are the receiver positions.

Extended sources are spread over $z=z_s$ (\ref{eqn:mssur}). The annihilator is the source localization penalty, equation \ref{eqn:msste}.

Denote by $J_{\alpha}[g,c;d]$ the objective function defined in \ref{eqn:esi}:
\begin{equation}
\label{eqn:jdef}
J_{\alpha}[g,c;d] = \frac{1}{2}( \|PL[c]^{-1}g -d \|^2 + \alpha \|A[g]\|^2) 
\end{equation}
To compute the objective 
\begin{itemize}
\item solve the acoustic wave system \ref{eqn:awe} with $g$ as in \ref{eqn:mssur}
\item sample at receiver locations and compute mean-square difference with pressure data traces, take mean square
\item add scaled mean square of annihilator applied to the source
\end{itemize}

The variable projection method \cite[]{GolubPereyra:03} applied to the objective \ref{eqn:esi} approaches its minimization via a nested algorithm: first minimize \ref{eqn:esi} over the source $g$ for given coefficients $c$, thus making $g = g[c]$ a function of $c$ (inner minimization); then substitute $g[c]$ in $J$ and minimize the resulting function of $c$ (outer minimization). The optimal $g[c]$ solves the first order necessary condition for minimization of $J$ over $g$: $\nabla_g J_{\alpha}[g,c;d] = 0$. For linear annihilators such as that defined by \ref{eqn:mssur}, $J$ is quadratic in $g$, so the first order necessary condition is also sufficient, and is equivalent to the linear {\em normal equation}:
\begin{equation}
\label{eqn:normal}
(S[c]^TS[c] + \alpha A^TA) g = S[c]^T d
\end{equation}
where I have written $S[c]=PL[c]^{-1}$ for the operator that maps the (extended) source to the data traces. Substitution of $g[c]$ in $J$ results in the {\em reduced objective}
\begin{equation}
\label{eqn:jred}
\tilde{J}_{\alpha}[c;d] = J_{\alpha}[g[c],c;d].
\end{equation}
The gradient would appear to involve the derivative of the inner solution $g[c]$ with respect to $c$, computation of which implies solution of additional linear systems. However by fortunate accident, the chain rule and the normal equation \ref{eqn:normal}
conspire to eliminate all terms involving derivatives of $g[c]$. Consequently the direction derivative of $\tilde{J}_{\alpha}$ in the direction $\delta c$ is
\begin{equation}
\label{eqn:tildejderiv}
D_c\tilde{J}_{\alpha}[c;d]\delta c = \langle D(S[c]g)\delta c,S[c]g-d \rangle|_{g=g[c]}
\end{equation}
whence
\begin{equation}
\label{eqn:l2grad}
\nabla \tilde{J}_{\alpha}[c;d] = D_c(S[c]g)^T|_{g=g[c]}(S[c]g[c]-d)
\end{equation}
Let us field-strip the expression on the right. The rightmost factor is the data residual, which is a by-product of many methods for the solution of the normal equation \ref{eqn:normal}. $D_cS$ is the linearization or Born approximation of $S$, or perturbation with respect to $c$, with $g$ fixed:
\[
S[c+\delta c]g \approx S[c]g + D_c(S[c]g)\delta c
\]
That is, $D_c(S[c]g)$ is the object commonly called the Born modeling operator, with source wavefield $g$. Its adjoint or transpose is $D_c(S[c]g)^T$ - as is well known, this is the RTM operator (or rather a version of it), computed via the adjoint state method in the time domain. Note that the Born and RTM operators need to match, i.e. both be derived from the same underlying full waveform modeling operator (in this case acoustodynamics, but the same would apply for any other wave theory).

The upshot is that the computation of the gradient goes like this:
\begin{itemize}
\item solve the normal equation \ref{eqn:normal} for $g[c]$
\item compute (or retain) the data residual $S[c]g[c]-d$
\item reverse-time migrate the data residual, using $g[c]$ to generate the source wavefield in the adjoint state method - the output is the gradient of $\tilde{J}_{\alpha}$.
\end{itemize}

To make this algorithm practical and efficient, effective preconditioning needs to be applied to both inner and outer problems.

\subsection{Preconditioning the Outer Problem: Gradient Smoothing}
Conditioning the outer (velocity etc.) update has different meaning for transmission versus reflection problems. One interpretation of ``transmission'' is that no reflection occurs on any ray path connecting source and receiver, whereever these may be located. Is is practically implicit in the use of ``ray path'' in this statement that the material model is at least mostly smooth, or essentially localized in low spatial frequencies. Smoothness is also one way to interpret ``lack of reflections''. The classic work explaining the need for model smoothness in tomography is \cite{DelpratLailly:92}. Since the surface source extension method gets its ability to identify kinematically correct velocity from its relation to tomography, it also requires model smoothness, at least for transmission.

A natural method for enforcing smoothness is use of a norm in model space that penalizes lack of smoothness. This method is also properly called preconditioning, which consists in altering the choice of norm to enhance convergence. For pure transmission, the choice recommended by \cite[]{DelpratLailly:92} is a norm from the Sobolev family. These are weighted $L^2$ (RMS) norms with weight operators indexed by a real parameter $s$:
\begin{equation}
\label{eqn:sob}
\|u\|^2_s = \langle u, W_s u \rangle_{L^2},\,\,W_s = (I - \nabla^2)^s
\end{equation}
Choosing $s>0$ penalizes lack of smoothness. According to the Sobolev Embedding Theorem and the basic well-posedness theorem for ordinary differential equations (which governs the behaviour of the ray system), it is neccessary to choose $s > 2 + d/2$ in dimension $d$, in order that metrizing the velocity (amongst other things) with the $s$-norm will guarantee sufficient smoothness that ray theory is well-behaved, and the link with tomography honored. For both $d=2$ and $d=3$, $s=4$ would work.

The gradient of a function is defined with respect to the domain norm. Using the $s$-norm, the gradient is
\begin{equation}
\label{eqn:grads}
\nabla_s \tilde{J}_{\alpha}[c,d] = W_s^{-1}\nabla \tilde{J}_{\alpha}[c,d]
\end{equation}
If $s>0$ then $W_s^{-1}$ is a smoothing operator. It is best implemented in the Fourier domain.

The most natural way view the reflection problem is to allow non-smoothness in $c$. It is not clear at this point what would be the optimal choice of weight operator in this case, or even whether there is a choice of Hilbert space structure that would offer an easily computable gradient and development of short-scale structures to generate reflections. GH used very low frequency data (in frequency domain) and Tihonov regularization. His results seemed to depend heavily on the regularization weight. Of course we also do not know how to connect the reflection case to a tomographic principle, so have no guidance for how a model space norm should be selected. These are the main target questions of this project.

\subsection{Preconditioning the Inner Problem: Time Reversal}
%In contrast to the outer problem, a preconditioner for the inner problem is available that works regardless of medium smoothness, so for both transmission and reflection (the latter with some caveats). 
In contrast to the outer problem, the inner problem admits a conventional preconditioner, that is, a change of variables that in the spaces of source traces and data traces respectively that moves the singular values of the modeling operator closer to 1. This preconditioner also creates an approximate solution operator for the inner problem, of the form
\begin{equation}
\label{eqn:appinv}
S[c]^{\dagger} = W_m[c]^{-1}S[c]^TW_d[c],
\end{equation}
in which $W_m$ and $W_c$ are $c$-dependent positive definite symmetric operators that define weighted inner products in domain and range of $S[c]$, and their associated weighted norms:
\begin{eqnarray}
\label{eqn:norms}
\langle f,g \rangle_m & = & \langle f, W_mg\rangle; \|f\|_m^2 = \langle f,f\rangle_m \nonumber \\
\langle f,g \rangle_d & = & \langle f, W_d g\rangle; \|f\|_d^2 = \langle f,f \rangle_d
\end{eqnarray}
We have implicitly used the notations $\langle \cdot,\cdot \rangle, \|\cdot \|$ (without subscripts) to denote the (unweighted) $L^2$ norm, in whatever form is appropriate for the arguments. That is what these symbols mean in display \ref{eqn:norms}, and will continue to mean.

The approximate inverse $S[c]^{\dagger}$ defined by the relation \ref{eqn:appinv} is the transpose of $S[c]$ with respect to the inner products $\langle \cdot,\cdot \rangle_m,, \langle \cdot,\cdot \rangle_d$, so $S[c]$ is approximately unitary in the same sense. Consequently the singular spectrum (which is also dependent on choice of inner product) is close to that of the identity - or most of it is, as we shall explain - and the convergence of  a Krylov subspace method like Conjugate Gradient (CG) Iteration converge relatively rapidly.

The system that needs solving, of course, is the normal equation \ref{eqn:normal},  so it is the convergence of CG or the like applied to that system that is the important issue. The normal equation is the first order necessary condition for minimization of the objective
\begin{equation}
\label{eqn:jdefs}
J_{\alpha}[g,c;d] = \frac{1}{2}( \|S[c]g -d \|^2 + \alpha \|A[g]\|^2).
\end{equation}
over the choice of surface source $g$. The choice of norms appearing in \ref{eqn:jdefs} is actually arbitrary, and the norms appearing in the first and second terms on the right-hand side do not even have to be the same.

%First the transmission case - a sort of rotated crosswell configuration. Idealize the receiver set as a continuum plane $z=z_r$, with $z_r > z_s$ - not essential but simpler.

%Two key facts and a geometric assumption underly the construction of the time-reversal preconditioner: (i) a local relation between the surface source and the normal particle velocity, and (ii) the uniqueness of solutions of the Dirichlet boundary value problem for the wave equation. Together these facts allow us to reconstruct the pressure field, or the significant part of it, from the transmission data, and to approximate a distributed source responsible for a given set of data traces.

%The geometric assumption is that the pressure field is {\em outgoing} at the recording datum: that is, that it is the restriction to $z>z_s$ of a pressure field that is negligible in $z<z_r$ for large enough time. ``Negligible'', rather than vanishing, as the latter is strictly speaking impossible: this property is asymptotic in frequency, and assured by an assumption about the ray field associated to the acoustic system.

%NB: the definition of the forward map (and the data) should include a taper $\phi(\bx_r,\bx_s)$ masking out the shot gather $d_r(\cdot,\cdot;\bx_s)$ for each source position $\bx_s$. However I will not explicitly mention this operation. I will also ignore it in the numerical examples, but there will be a price to pay.

To begin with, change notation slightly, and write $S[c]h_s$ rather than $S[c]g$ with $g = h_s \delta(z-z_s)$, as $z_s$ will be presumed ot be the same for all shots. 
%The addition of the subscript indicates that the sources lie on the plane $z=z_s$ and the receivers are on the plane $z=z_r$. As will be apparent shortly, a similar operator with source and receiver planes swapped will also be important, hence the need for the additional information in the notation.

The pressure and velocity fields occuring in the computation of $d_r=p|_{z=z_r}=S[c]_{s,r}h_s$ form the solution of
\begin{eqnarray}
\label{eqn:awedata}
\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
h \delta(z-z_s)), \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t \ll 0 \nonumber\\ 
\bv & = & 0 \mbox{ for } t \ll 0 \nonumber\\
\end{eqnarray}
Actually $p,\bv$ form a {\em weak solution} of the system \ref{eqn:awedata}, defined by integration by parts. A more general setting for this type of system provides an advantageous framework within which to explain the properties of these weak solutions.

Let $\omega \subset \bR^d$ be a smooth orientable surface. Then the distribution $\delta_{\omega}$ is well-defined: for $u \in C_0^{\infty}(\bR^d)$, 
\[
\langle \delta_{\omega}, u \rangle = \int_{\omega} u dS
\]
The system \ref{eqn:awedata} is a special case of
\begin{eqnarray}
\label{eqn:aweomega}
\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
h \delta_{\omega}, \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t \ll 0 \nonumber\\ 
\bv & = & 0 \mbox{ for } t \ll 0 \nonumber\\
\end{eqnarray}
in which $h \in C_0^{\infty}(\omega)$.

\begin{thm}
\label{thm:bvs}
Suppose that $\kappa,\beta$ are smooth. Then the system \ref{eqn:aweomega} has a unique weak solution $(p,\bv)$. Denote by $[v_n] \in C^{\infty}(\omega)$ the jump of the normal component of $\bv$ across $\omega$:
\[
[v_n](\bx) = \lim_{\epsilon \rightarrow 0} (\bv(\bx + \epsilon {\bf n}(\bx))-\bv(\bx - \epsilon {\bf n}(\bx)))\cdot {\bf n}(\bx),\,\bx \in \omega.
\]
Then $h(\bx) = [v_n](\bx), \bx \in \omega$.

%Let $p,\bv$ be the weak solution of the system \ref{eqn:awedata}, and assume that $\kappa, \beta$ are smooth. Denote by $[v_z]|_{z=z_s}$ the jump of the z-component of particle velocity in the z-direction:
%\[
%[v_z](x,z) = v_z^+(x,z)-v_z^-(x,z),  v_z^{\pm}(x,z)=\lim_{z'\rightarrow z^{\pm}} v_z(x,z')
%\]
%Then $[v_z](x,z)=0$ unless $z=z_s$, and
%\[
%h(x) = [v_z](x,z_s).
%\]
\end{thm}
For the proof, see  Appendix B.

For the rest of this section, $\omega$ will be a coordinate hyperplane, defined by $z=$ constant, with its usual orientation, and I will write $[v_z] = [v_b]$.

To begin with, change notation slightly, and write $S[c]h_s$ rather than $S[c]g$ with $g = h_s \delta(z-z_s)$, as $z_s$ will be presumed ot be the same for all shots. 
%The addition of the subscript indicates that the sources lie on the plane $z=z_s$ and the receivers are on the plane $z=z_r$. As will be apparent shortly, a similar operator with source and receiver planes swapped will also be important, hence the need for the additional information in the notation.

The pressure and velocity fields occuring in the computation of $d_r=p|_{z=z_r}=S[c]_{s,r}h_s$ form the solution of
\begin{eqnarray}
\label{eqn:awedata1}
\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
h \delta(z-z_s)), \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t \ll 0 \nonumber\\ 
\bv & = & 0 \mbox{ for } t \ll 0 \nonumber\\
\end{eqnarray}
Actually $p,\bv$ form a {\em weak solution} of the system \ref{eqn:awedata1}, defined by integration by parts.
\begin{thm}
\label{thm:vz2}
If $\kappa,\beta$ are constant near the $z=z_s$, then $v_z^-=-v_z^+$ so $[v_z] = 2 v_n^+$. 
\end{thm}
(for proof see Appendix 2). The formulation is for 2D propagation, but a precisely similar fact is true for 3D - in that case, $x$ is replaced everywhere by the horizontal coordinates.

The import of Theorem \ref{thm:bvs} is that we can read off the source if we can infer the (particle) velocity field near the source surface, from knowledge of the data traces (values of the pressure at $z=z_r$). The key observation that enables this feat is a well-posedness theorem for the wave equation:

\begin{thm} \label{thm:bvp}
Suppose that $\omega \subset \bR^d$ is a smooth orientable surface, $\omega = \partial  \Omega^- = \partial \Omega^+$, where $\Omega^{\pm}$ are the open exterior/interior regions defined by the outward unit normal field of $\omega$, and let $d \in C^{\infty}_0(\omega \times \bR)$. Then there exist unique smooth solutions $p^{\pm},\bv^{\pm}$ of the initial/boundary value problems 
\begin{eqnarray}
\label{eqn:awebdry}
\frac{\partial p^{\pm}}{\partial t} & = & - \kappa \nabla \cdot \bv^{\pm} \mbox{ in } \Omega^{\pm}, \nonumber \\
\frac{\partial \bv^{\pm}}{\partial t} & = & - \beta \nabla p^{\pm} \mbox{ in } \Omega^{\pm},\nonumber \\
p^{\pm} & =& 0 \mbox{ for } t \ll 0\nonumber\\ 
\bv^{\pm} & = & 0 \mbox{ for } t \ll 0 \nonumber\\
\lim_{\bx \rightarrow \omega^{\pm}}p & = & d 
\end{eqnarray}
\end{thm}

Putting the two theorems together, obtain
\begin{cor}\label{thm:bvps}
Suppose that $d_s$ is a sufficiently regular function, = 0 for $t<0$, and denote by $p^{\pm},\bv^{\pm}$ the half-space solutions provided by Theorem \ref{thm:bvp}. Define the vector fields $p,\bv$ in $\bR^4$ by
\begin{eqnarray}
\label{eqn:putemtogether} 
p(x,z) &=& p^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
p(x,z_s)&=&d_s(x)\\
\bv(x,z) &=& \bv^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
\end{eqnarray}
Then $p,\bv$ is a weak solution of the system \ref{eqn:awedata1} with $h=[v]_n$.
\end{cor}

Now invoke the {\em outgoing assumption}: $p,\bv$ is the restriction to $z>z_s$ of a solution vanishing in $z<z_r$ for large $t$. This condition needs to be interpreted asymptotically, that is, $p,\bv$ is relatively smooth (lower order in frequency) for $t \gg 0, z<z_r$. It holds if the rays carrying significant energy in $p,\bv$ intersect $z=z_r$ transversally, and do not return to $z<z_r$ in forward time. The examples below will satisfy this condition, because the wave velocity is constant in $z > z_r$.

Note that isotropic point sources do not produce fields outgoing in a half space, in the sense just described - simply because they radiate in all directions. However, for any specified finite time of propagation, it is possible to replace a point source by an anisotropic source that fits the point source data to within an arbitrarily small error, and does produce outgoing fields. In fact, the numerical examples below will construct such sources.

This assumption implies that for $t$ sufficiently large, $d_r = p|_{z=z_r}$ is negligible for $t \gg 0$, so $p,\bv$ differs in $z_s<z < z_r$ by a negligible error from the solution of $p^-,\bv^-$, part of the solution of the  backwards-in-time  initial value problem, similar to the system \ref{eqn:awebdry}:
\begin{eqnarray}
\label{eqn:awetr}
\frac{\partial \tilde{p}^{\pm}}{\partial t} & = & - \kappa \nabla \cdot \tilde{\bv}^{\pm}, \pm(z-z_r)>0 \nonumber \\
\frac{\partial \tilde{\bv}^{\pm}}{\partial t} & = & - \beta \nabla \tilde{p}^{\pm},\nonumber \\
\tilde{p}^{\pm} & =& 0 \mbox{ for } t \gg 0,\nonumber\\ 
\tilde{\bv}{\pm} & = & 0 \mbox{ for } t \gg 0,\nonumber\\
\lim_{z \rightarrow z_r^{\pm}}\tilde{p}^{\pm} &=&  d_r, \nonumber\\,
\tilde{p} &=& \tilde{p}^{\pm}, \pm(z-z_r)>0, \nonumber\\, 
\tilde{\bv} &=& \tilde{\bv}^{\pm}, \pm(z-z_r)>0.
\end{eqnarray}

This must be so, as $p,\bv$ satisfies the wave equation system (first two conditions), and the rest as well except for a negligible error. The last condition must be interpreted consistently, that is $\tilde{p}=0$ for $t$ large enough that $d_r$ is negligible. (In practice consistency will be obtained by replacing $t \gg 0$ respectively $t \ll \infty$ by $t > t_{\rm max}$, $t < t_{\rm max}$ for a suitably large $t_{\rm max}$). Then the standard energy estimate shows that $p, \tilde{p}$ differ also by a negligible error: in particular, $p|_{z=z_s} \approx \tilde{p}_{z=z_s}$.

According to Corollary \ref{thm:bvps}, recovering $p_{z=z_s}$ entails recovering $h_s=[v_z]|_{z=z_s}$. So a procedure for computing an approximate inverse $S[c]_{s,r}^{\dagger}$ for $S[c]_{s,r}$ is 
\begin{equation}
\label{eqn:psinv}
S[c]_{s,r}^{\dagger}d_r\equiv [v_z]_{z=z_s}.
\end{equation}
where the jump in $v_z$ at $z=z_s$ is computed from knowledge of $d_r=p|_{z=z_r}$ by
\begin{itemize}
\item solving \ref{eqn:awetr} backwards in time,
\item reading off $d_s=\tilde{p}$ at $z=z_s$,
\item solving \ref{eqn:awebdry} forwards in time, and computing $[v_z]|_{z=z_s}$, or if appropriate ($\kappa$ and $\beta$ constant near $z=z_s$) simply using $\bar{v}_{z=z_s^+} = v_z^{+}=-v_z^{-}$ at $z=z_s$ (Theorem \ref{thm:bvs}).
\end{itemize}

The relations developed in the last few paragraphs may be summarized as follows:
\begin{equation}
\label{eqn:appinv}
d_r = S[c]_{s,r}h_s \,\,\Leftrightarrow \,\, h_s \approx S[c]_{s,r}^{\dagger}d_r
\end{equation}
that is, $S[c]_{s,r}^{\dagger}$ is an approximate inverse of $S[c]_{s,r}$.

Inspection of systems \ref{eqn:awebdry} and \ref{eqn:awetr} reveals that they are actually the same, except for the direction of time, and that observation leads to another useful relation. Denote by $R$ the time-reversal operator:
\[
Ru(\bx,t)=u(\bx,-t)
\]
and by {\cal R} the time reversal operator for acoustic fields:
\[
{\cal R}\left(\begin{array}{c}
p\\
\bv
\end{array}
\right) = 
\left(\begin{array}{c}
Rp\\
-R\bv
\end{array}
\right) 
\]
The sign in the definition of ${\cal R}$ is chosen so that if $(\tilde{p},\tilde{\bv})^T$ solves the system \ref{eqn:awetr}, then ${\cal R}(\tilde{p},\tilde{\bv})^T=(\hat{p},\hat{\bv})^T$ is the solution of 
\begin{eqnarray}
\label{eqn:awetrs}
\frac{\partial \hat{p}^{\pm}}{\partial t} & = & - \kappa \nabla \cdot \hat{\bv}^{\pm}, \pm(z-z_r) > 0 \nonumber \\
\frac{\partial \hat{\bv}^{\pm}}{\partial t} & = & - \beta \nabla \hat{p}^{\pm},\nonumber \\
\hat{p}^{\pm} & =& 0 \mbox{ for } t \ll 0,\nonumber\\ 
\hat{\bv}^{\pm} & = & 0 \mbox{ for } t \ll 0,\nonumber\\
\lim _{z=z_r^{\pm}}\hat{p}^{\pm} &=&  Rd_r,\nonumber\\
\hat{p} &=& \hat{p}^{\pm}, \pm(z-z_r)>0, \nonumber\\, 
\hat{\bv} &=& \hat{\bv}^{\pm}, \pm(z-z_r)>0.
\end{eqnarray}

%For a smooth orientable surface $\omega$, denote by $\Lambda[c,\omega]$ the operator that maps Dirichlet pressure data $d$ on $\omega \times \bR$ to the jump in normal velocity $[v_n]$ across $\omega \times \bR$ via solution of the system \ref{eqn:awebdry}.

 From Corollary \ref{thm:bvps} applied to the system \ref{eqn:awetrs}, $\hat{p}|_{z=z_s} = S[c]_{r,s}\Lambda[c]_rRd_r$. Denote by $\Lambda[c]_r$ the same operator with $z_s$ replaced by $z_r$. Then $\hat{p}|_{z=z_s} = S[c]_{r,s}\Lambda[c]_rRd_r=R\tilde{p}|_{z=z_s}$. Thus the definition \ref{eqn:psinv} implies
\begin{equation}
\label{eqn:psinvrev}
S[c]_{s,r}^{\dagger}= -\Lambda[c]_sRS[c]_{r,s}R\Lambda[c]_r
\end{equation}
since $\Lambda[c]_rR=-R\Lambda[c]_r$ from the definition of ${\cal R}$.
Define $T[c]_{s,r}d_s = p|_{z=z_r}$ for the solution of \ref{eqn:awebdry}. Since ${\cal R}^2=I$, Then $\hat{p}|_{z=z_s} = T[c]_{r,s}Rd_r$. Since ${\cal R}^2=I$, ${\cal R}(\hat{p},\hat{\bv})^T=(\tilde{p},\tilde{\bv})^T$, whence $\tilde{p}|_{z=z_s}=RT[c]_{r,s}Rd_r$.

Denote by $\Lambda[c]_s$ the operator that produces the jump in normal velocity $[v_z]=\lim_{z \rightarrow z_s^+}v_z^+ - \lim_{z\rightarrow z_s^-} v_z^-$ from Dirichlet pressure data $d_s$ and causal solution of the system \ref{eqn:awebdry}. From Corollary \ref{thm:bvps}, $T[c]_{s,r} = S[c]_{s,r}\Lambda[c]_s$. Denote by $\Lambda[c]_r$ the same operator with $z_s$ replaced by $z_r$. Then $\hat{p}|_{z=z_s} = S[c]_{r,s}\Lambda[c]_rRd_r=R\tilde{p}|_{z=z_s}$. Thus the definition \ref{eqn:psinv} implies
\begin{equation}
\label{eqn:psinvrev}
S[c]_{s,r}^{\dagger}= -\Lambda[c]_sRS[c]_{r,s}R\Lambda[c]_r
\end{equation}
since $\Lambda[c]_rR=-R\Lambda[c]_r$ from the definition of ${\cal R}$.

As I will show in the example section, the approximation may be accurate enough without any further improvement. However, a closer examination shows that in fact it is possible to express the relation between $S[c]_{s,r}$ and $S[c]_{s,r}^{\dagger}$ in terms of a change of norm, that is, in terms of a preconditioner, and this relation can be used to considerably accelerate an iterative solution of the normal equation \ref{eqn:normal}.

To see this, compute the formal or $L^2$ adjoint $S[c]_{s,r}^T$, by a variant of the adjoint state method, in this case a by-product of the conservation of energy. Suppose that $\bar{h}_r$ is a function on $z=z_r$ (data traces), and 
$\bar{p},\bar{\bv}$ solve the backwards-in-time boundary value problem:
\begin{eqnarray}
\label{eqn:aweadj}
\frac{\partial \bar{p}}{\partial t} & = & \kappa (-\nabla \cdot \bar{\bv} + 
 \bar{h}_r \delta(z-z_r))\nonumber \\
\frac{\partial \bar{\bv}}{\partial t} & = & - \beta \nabla \bar{p},\nonumber \\
\bar{p} & =& 0 \mbox{ for } t \gg 0\nonumber\\ 
\bar{\bv} & = & 0 \mbox{ for } t \gg 0 
\end{eqnarray} 
Then
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p \bar{p}}{\kappa} +  
\frac{\bv \cdot \bar{\bv}}{\beta} \right)|_{t=T}
-
\left(\int\, dx\,dy\,dz\, \frac{p \bar{p}}{\kappa} +  \frac{\bv \cdot \bar{\bv}}{\beta} \right)|_{t=0}
\]
\[
= 
\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \bar{p}}{\kappa} +  \frac{\bv \cdot \bar{\bv}}{\beta} \right)
\]
\[
= 
\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \bar {p} +  p \frac{1}{\kappa}\frac{\partial \bar{p}}{\partial t} \right.
\]
\[
+
\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \bar{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \bar{\bv}}{\partial t} \right)
\]
\[
= 
\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
 h_s \delta(z-z_s)\right) \bar{p} + p \left(- \nabla \cdot \bar{\bv} + 
 d_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  (- \nabla p) \cdot \bar{\bv} + \bv \cdot (-\nabla \bar{p}) \right)
\]
\[
= 
\int_{0}^{T}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
 h_s \delta(z-z_s)\right) \bar{p} + p \left(- \nabla \cdot \bar{\bv} + 
 d_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  p (\nabla \cdot \bar{\bv}) + (\nabla \cdot \bv) \bar{p} \right)
\]
after integration by parts in the last two terms. Most of what is left cancels, leaving 
\[
0 = \int\,dt\,dx\,dy\, (h_s \bar{p})_{z=z_s} + (p  \bar{h}_r)_{z=z_r}
\]
whence
\begin{equation}
\label{eqn:sadj}
 S[c]_{s,r}^T \bar{h}_r = -\bar{p}|_{z=z_s}
\end{equation}
Compare the systems \ref{eqn:awetr} and \ref{eqn:aweadj}: they differ in that 
\begin{itemize}
\item in \ref{eqn:awetr}, $d_r$ appears as a Dirichlet condition on $z=z_r$,
\item in \ref{eqn:aweadj}, $\bar{h}_r$ appears as a distributed source on $z=z_r$.
\end{itemize}
According to Corollary \ref{thm:bvps}, $\bar{p}=\bar{p}, \bar{\bv}=\bar{\bv}$ if $\bar{h}_r = [\bar{v}_z]|_{z=z_r}$, i.e. the source is that which generates the same solution as the Dirichlet condition..

Simlarly denote by $\tilde{\Lambda}[c]_r$ the operator that produces the jump in normal velocity at $z=z_r$ from pressure data $d_r$ via {\em anticausal} solution of the system 
\begin{eqnarray}
\label{eqn:awebdrytr}
\frac{\partial {\tilde p}^{\pm}}{\partial t} & = & - \kappa \nabla \cdot {\tilde \bv}^{\pm} \mbox{ in } \pm(z-z_r) > 0, \nonumber \\
\frac{\partial {\tilde \bv}^{\pm}}{\partial t} & = & - \beta \nabla {\tilde p}^{\pm} \mbox{ in } \pm(z-z_r) > 0,\nonumber \\
{\tilde p}^{\pm} & =& 0 \mbox{ for } t \gg 0\nonumber\\ 
{\tilde \bv}^{\pm} & = & 0 \mbox{ for } t \gg 0 \nonumber\\
\lim_{z\rightarrow z_r^{\pm}}{\tilde p} & = & {\tilde d}_r 
\end{eqnarray}
%\end{thm}
 for the receiver surface $z=z_r$ (in the math literature, this is the ``Dirichlet-to-Neumann operator'', more or less). The discussion so far can be summarized as follows:
\begin{equation}
\label{eqn:wadj}
\bar{p}|_{z=z_s} = S[c]_{s,r}^T\Lambda[c]_r \bar{p}|_{z=z_r} 
\end{equation}
According to the discussion above, $\bar{p}|_{z=z_s} \approx p|_{z=z_s}$, and $h_s = \Lambda[c]_s p|_{z=z_s} = S[c]_{s,r}^{\dagger}d_r$. That is,
\begin{equation}
\label{eqn:appinv}
S[c]_{s,r}^{\dagger} = \Lambda[c]_s S[c]_{s,r}^T \Lambda[c]_r
\end{equation}
This relation can be re-written in the useful form
\begin{equation}
\label{eqn:adj}
S[c]_{s,r}^{\dagger} = W_m[c]^{-1}S[c]_{s,r}^TW_d[c]
\end{equation}
in which 
\begin{equation}
\label{eqn:weights}
W_m[c] = \Lambda[c]_s^{-1}\,\, W[c]_d = \Lambda[c]_r
\end{equation}
Since $\Lambda[c]$ is positive (semi-)definite, equation \ref{eqn:adj} exhibits $S[c]_{s,r}^{\dagger}$ as the adjoint of $S[c]_{s,r}$ with respect to weighted norms 
\begin{itemize}
\item on the source space ($h_s$): weight $W[c]_m$
\item on the data space ($d_r$): weight $W[c]_d$
\end{itemize}
Since $S[c]_{s,r}^{\dagger}S[c]_{s,r} \approx I$ as shown above, $S[c]_{s,r}$ is approximately unitary in the Hilbert spaces of source and data traces with norms defined by the weighting operators $W_m$ and $W_d$ respectively. Therefore a Krylov space method employing these norms will converge rapidly.

\cite{HouSymes:EAGE16} demonstrated a very similar preconditioner for Least Squares Migration, also for its subsurface offset extension \cite[]{HouSymes:16}, motivated by \cite{tenKroode:12}. These constructions all involve the Dirichlet-to-Neumann operator. This concept also turns up in hidden form in the work of Yu Zhang and collaborators on true amplitude migration \cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,ZhangYuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.

\subsection{Implementation Considerations}
The Dirichlet-to-Neumann operator $\Lambda$ is an essential part of the inner problem preconditioner just presented. Implementation can be accomplished in several ways:
\begin{itemize}
\item \cite{tenKroode:12} suggests using a one-way operator;
\item if both pressure and normal partical velocity are measured (or simulated), then the two are related by $\Lambda$ and the velocity component can simply be used as the output;
\item presence of a free surface implies all of the usual problems, such as the need for removal of receiver-side ghosts. On the other hand, if the free surface is within a quarter-wavelength throughout the useful bandwidth of the data, then the ghosted data differs from $\Lambda p$ by a time integration and a scale factor, a fact used to good effect by \cite{HouSymes:15}.
\end{itemize}

In the examples, I have used the second observation. With a finite difference implementation of the pressure-velocity system \ref{eqn:awe}, velocity components are available ``for free'', short-circuiting explicit computation of $\Lambda$.

The variable projection gradient formula \ref{eqn:l2grad} has an Achilles heel, not mentioned so far: it is correct only when the normal equation \ref{eqn:normal} is satisfied {\em exactly}. Any iterative method (or the asymptotic inverse developed in the last subsection) solves \ref{eqn:normal} only approximately, so a nonzero error term is neglected in using \ref{eqn:l2grad}. This error term can be large, because it amplifies the high frequencies in the solution error, in fact depends on derivatives of the solution. For pure transmission through transparent (smooth) material models, the approximate inverse $S[c]_{s,r}^{\dagger}$ is asymptotically exact (at least in principle), and that fact can be used to devise a modification of \ref{eqn:l2grad} that converges with a convergent iteration for \ref{eqn:normal}. [LAST YEAR TRIP REPORT, MAY PUT SOMETHING IN APPENDIX]

Choice of the penalty weight $\alpha$, and of region in which the extended source $g$ is permitted to be nonzero, must be managed. These two choices are linked. Lei Fu's thesis \cite[]{Fu:Geo17,Fu:Geo17b} explains and tests algorithms for similar choices that occur in subsurface offset WEMVA. All of these ideas apply {\em ipso facto} to the surface source extension.

\subsection{Practical Preconditioning}

The formula \ref{eqn:adj} allows direct use of the conjugate gradient (or other Krylov space iterative) algorithm with the inner products described above. This is however computationally inconvenient, as it requires explicit computation of the weight operators, in this case the Dirichlet-to-Neumann operator. There are several ways to carry out this computation, as mentioned somewhere, but it is in any case a  pain. On the other hand, it is possible to rearrange the calculation with weighted norms into the form of the {\em preconditioned conjugate gradient (PCG) algorithm}. For the current problem, identify $A=S[c]_{s,r}$, $A^* = S[c]_{s,r}^T W_d$. PCG takes the form
\begin{algorithm}[H]
\caption{Preconditioned Conjugate Gradient Algorithm}
\begin{algorithmic}[1]
\State Choose $x_0$ somehow
  \State $r_0 \gets A^*(b-Ax_0)$
  \State $p_0 \gets W_m^{-1}r_0$
  \State $g_0 \gets p_0$
  \State $q_0 \gets A^*Ap_0$
  \State $k \gets 0$
  \Repeat
  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle p_k,q_k\rangle}$
  \State $x_{k+1} \gets x_k + \alpha_k p_k$
  \State $r_{k+1} \gets r_k - \alpha_kq_k$
  \State $g_{k+1} \gets W_m^{-1}r_{k+1}$
  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
  \State $q_{k+1} \gets A^*Ap_{k+1}$
  \State $k \gets k+1$
  \Until{Error is sufficiently small, or max iteration count exceeded} 
\end{algorithmic}
\end{algorithm}
Of course, this variant of CG merely hides the data sides weight operator, and requires application on the model side. This is still inconvenient.

The simplest solution is to use the coexistence of the $d_r=p$ and $\Lambda[c]_rd_r =[v_z]=2v_z$ traces (assuming a horizontal datum), and similarly for the source datum. That is, the computation of $\Lambda[c]_r$ is intrinsic: when $S[c]_{s,r}h_s$ has been computed, so has $\Lambda[c]_rS[c]h_s$ - it is simply the collection of corresponding vertical velocity traces. Similarly,  $\Lambda[c]_sS[c]^T$ is the collection of vertical velocity traces at $z=z_s$ arising from the time-reversal computation input equal to sources traces on $z=z_r$. If we can rewrite the PCG loop to require only these computations, then the acoustic simulations in forward and reverse time will suffice to compute the step.

In the loop body above, introduce the vector sequences $t_k=Ap_k,v_k=W_d t_k, s_k=W_m^{-1}q_k$, and re-write the iteration by unfolding $W_m^{-1}S[c]^T W_d S[c]$ and noting that $\langle p_k, q_k \rangle =\langle p_k,A^TW_dAp_k \rangle =\langle Ap_k,W_dAp_k \rangle =\langle t_k,v_k\rangle$: 
\begin{algorithm}[H]
\caption{Modified Preconditioned Conjugate Gradient Algorithm}
\begin{algorithmic}[1]
\State Choose $x_0$ somehow
  \State $r_0 \gets A^*(b-Ax_0)$
  \State $p_0 \gets W_m^{-1}r_0$
  \State $g_0 \gets p_0$
  \State $k \gets 0$
  \Repeat
  \State $t_k \gets Ap_k$
  \State $v_k \gets W_dt_k$
  \State $q_k \gets A^Tv_k$
  \State $s_k \gets W_m^{-1}q_k$
  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle t_k,v_k\rangle}$
  \State $x_{k+1} \gets x_k + \alpha_k p_k$
  \State $r_{k+1} \gets r_k - \alpha_kq_k$
  \State $g_{k+1} \gets g_k - \alpha_k s_k$
  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
  \State $k \gets k+1$
  \Until{Error is sufficiently small, or max iteration count exceeded} 
\end{algorithmic}
\end{algorithm}
This rewrite does not look like an improvement - four operator applications appear to be required:
\begin{itemize}
\item $p_k \rightarrow S[c] p_k = t_k$
\item $p_k \rightarrow  \Lambda[c]_r S[c] p_k = v_k$
\item $v_k \rightarrow S[c]^Tv_k = q_k$
\item $v_k \rightarrow \Lambda[c]_s S[c]^T v_k = s_k$
\end{itemize}
However this rearrangement reveals that actually only two wave equation solves are needed. For $q_k$ and $s_k$, both by-products of the adjoint (time-reversed) simulation: $q_k$ stores the pressure traces on $z=z_s$, $s_k$ the velocity traces. The same is true of $t_k$ and $v_k$, with the latter being the input for the $q_k,s_k$ computation.

Denote by $F_{pp}$ the acoustic simulation operator with a $p$ source (constitutive law defect, as in the system \ref{eqn:awedata}), and $p$ output traces. Similarly, denote by $F_{pv}$ the simulation operator with $p$ source and $v_z$ output traces, likewise $F_{vp}$ and $F_{vv}$, and by ${\bf F}$ the matrix operator
\begin{equation}
\label{eqn:iwop}
{\bf F} = \left(
\begin{array}{cc}
F_{pp} & F_{vp} \\
F_{pv} & F_{vv}
\end{array}
\right)
\end{equation}
The {\tt IWaveLOVOp} built in the asg package implements this matrix operator if the keywords {\tt source\_p, source\_v0, data\_p, data\_v0} are defined in its parameter list: these keywords point to SEGY trace files that define the spaces in which the $p$ and $v_z$ sources respectively data traces are treated as vectors. That is, IWAVE provides an easy access to ${\bf F}$. In terms of the operators used in the statement of the inverse problem, $F_{pp}=S[c]$, $F_{pv}=\Lambda[c]_rS[c]$, $F_{vp} = S[c]\Lambda[c]_s$, $F_{pp}^T = S[c]^T$, and $F_{vp}^T = \Lambda[c]_sS[c]^T$. So if we define 
\begin{equation}
\label{eqn:vecvec}
{\bf p}_k = \left[
\begin{array}{c}
p_k\\
0
\end{array}
\right],\,\,
{\bf r}_k = \left[
\begin{array}{c}
r_k\\
g_k
\end{array}
\right],\,\,
{\bf t}_k = \left[
\begin{array}{c}
t_k\\
v_k
\end{array}
\right],\,\,
{\bf v}_k = \left[
\begin{array}{c}
v_k\\
0
\end{array}
\right],\,\,
{\bf q}_k = \left[
\begin{array}{c}
q_k\\
s_k
\end{array}
\right],
\end{equation}
and
\begin{equation}
\label{eqn:proj}
{\bf P} = \left(
\begin{array}{cc}
0 & I \\
0 & 0
\end{array}
\right),
\end{equation}
then the algorithm listing above can be rewritten as
\begin{algorithm}[H]
\caption{Further modified Preconditioned Conjugate Gradient Algorithm}
\begin{algorithmic}[1]
\State Choose $x_0$ somehow
  \State $r_0 \gets A^TW_d(b-Ax_0)$
  \State $p_0 \gets W_mr_0$
  \State $g_0 \gets p_0$
  \State $k \gets 0$
  \Repeat
  \State ${\bf t}_k \gets {\bf F}{\bf p}_k$
  \State ${\bf v}_k \gets {\bf P}{\bf t}_k$
  \State ${\bf q}_k \gets {\bf F}^T{\bf v}_k$
  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle t_k,v_k\rangle}$
  \State $x_{k+1} \gets x_k + \alpha_k p_k$
  \State ${\bf r}_{k+1} \gets {\bf r}_k - \alpha_k{\bf q}_k$
  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
  \State ${\bf p}_{k+1}\gets {\bf P}{\bf r}_{k+1}+\beta_{k+1}{\bf p}_k$
  \State $k \gets k+1$
  \Until{Error is sufficiently small, or max iteration count exceeded} 
\end{algorithmic}
\end{algorithm}

Note that
\begin{equation}
\label{eqn:note}
{\bf r}_0 = \left[
\begin{array}{c}
r_0\\
g_0 
\end{array}
\right] = 
\left[
\begin{array}{c}
r_0\\
p_0 
\end{array}
\right] = 
\left[
\begin{array}{c}
A^TW_d(b-Ax_0)\\
W_m^{-1}A^TW_d(b-Ax_0)
\end{array}
\right]
\end{equation}
Define 
\begin{equation}
\label{eqn:rhsdef}
{\bf b}
= 
\left[
\begin{array}{c}
W_d(b-Ax_0)\\
0
\end{array}
\right].
\end{equation}
Then identity \ref{eqn:note} implies that
\begin{equation}
\label{eqn:init}
{\bf r}_0= {\bf F}^T{\bf b},\,\, {\bf p}_0 = {\bf P}{\bf r}_0.
\end{equation}
This observation allows a convenient ``hoist'' of the $k=0$ operator applications out of the iteration loop. Also, it is convenient to introduce the indefinite quadratic form that returns the inner product of the two components of a vector in the product space introduced above:
\begin{equation}
\label{eqn:dotfunc}
Q({\bf x})=\langle x_0,x_1\rangle=\left\langle \bx,
\left(
\begin{array}{cc}
0 & I \\
I & 0 
\end{array}
\right) \bx \right\rangle
\end{equation}
and the scalar sequences
\begin{eqnarray}
\label{eqn:scalars}
\gamma_k &=& \langle r_k,g_k \rangle = Q({\bf r})\\
\delta_k &=& \langle t_k,v_k \rangle = Q({\bf t})
\end{eqnarray}

With these modifications, we obtain 
\begin{algorithm}[H]
\caption{Intrinsically Preconditioned Conjugate Gradient Algorithm}
\begin{algorithmic}[1]
\State Choose $x_0$ somehow
\State ${\bf b} \gets \left[
\begin{array}{c}
W_d(b-Ax_0)\\
0
\end{array}
\right]$
  \State ${\bf r}_0 \gets {\bf F}^T{\bf b}$
  \State ${\bf p}_0 \gets {\bf P}{\bf r}_0$
  \State $\gamma_0 \gets Q({\bf r}_0)$
  \State $k \gets 0$
  \Repeat
  \State ${\bf t}_k \gets {\bf F}{\bf p}_k$
  \State $\delta_k \gets Q({\bf t})$
  \State ${\bf v}_k \gets {\bf P}{\bf t}_k$
  \State ${\bf q}_k \gets {\bf F}^T{\bf v}_k$
  \State $\alpha_k \gets \frac{\gamma_k}{\delta_k}$
  \State $x_{k+1} \gets x_k + \alpha_k p_k$
  \State ${\bf r}_{k+1} \gets {\bf r}_k - \alpha_k{\bf q}_k$
  \State $\gamma_{k+1} \gets Q({\bf r}_{k+1})$
  \State $\beta_{k+1} \gets \frac{\gamma_{k+1}}{\gamma_k}$
  \State ${\bf p}_{k+1}\gets {\bf P}{\bf r}_{k+1}+\beta_{k+1}{\bf p}_k$
  \State $k \gets k+1$
  \Until{Error is sufficiently small, or max iteration count exceeded} 
\end{algorithmic}
\end{algorithm}

\section{Prototype Numerical Examples}
\inputdir{project}

I present a collection of simple examples that illustrate the features of the surface source extension claimed in preceding sections.

I used the IWAVE acoustic staggered grid package to carry out these calculations. This package implements (2,2k) schemes for k=1,2,..., and outputs traces (of either velocity or pressure at any point in space via multilinear interpolation. The discretized modeling operator is thus of second order accuracy, though as usual I have used higher order in space to reduce grid dispersion.

Source injection is implemented as the adjoint of trace sampling, resulting in another second-order error [REFERENCES].

The data is a single shot gather, with a source at coordinates $x_s=z_s=3000$ m (units of length are meters in all cases). The receiver line occupies $1500 \le x_r \le 5500$ m, with receiver depth $z_r=1000$ m.  Extended sources occupy $1500 \le x_s \le 5500$ m, with the same depth $z_s=3000$ m as the ``physical'' source used to generate the data. This region turns out to be adequate to represent the extended sources that approximately invert the data, for the cases examined below. An algorithm to automatically identify an appropriate region can be based on the ideas developed by \cite{Fu:Geo17}.

I have used absorbing boundary conditions (split-field PML) on all four sides of the 4000 m (vertical) $\times$ 8000 m (horizontal) simulation domain. Evidently includion of a free surface is important to the application of the ideas explained here to diving wave marine data, and I have not addressed the necessary modifications here.

The script is set up to carry out the necessary computations on grids with spacings $\Delta x = \Delta z = $ 20, 10, and 5 m, with a jump of roughly 8 in computation time resulting from each refinement. For present purposes, the coarsest (20 m) grid seems to be sufficient, and that is the grid used in the examples presented below. The source pulses are chosen so that the computation is reasonably accurate. For the 20 m grid case, I use a zero-phase trapezoidal bandpass filter source with corner frequencies of 1.0, 2.0, 7.5, and 12.5 Hz.

The IWAVE asg driver has been set up to recognize the case {\tt deriv=0} as defining the map from source (right-hand side in the pressure equation) to data (pressure) traces. The adjoint to this map, as explained above, is reverse-time propagation of the data traces as pressure sources, followed by scaling (formula \ref{eqn:sadj}). The approximate inverse is computed by application of the Dirichlet-to-Neumann operator to the pressure traces to produce corresponding velocity traces, followed by injection as pressure sources and reverse time propagation, followed by another application of the Dirichlet-to-Neumann map and scaling (formulas \ref{eqn:appinv}, \ref{eqn:adj}).

The script for the first set of examples implements these operatations step-by-step via calls to IWAVE, Madagascar, and SU commands. A peculiarity of the {\tt asg.x} driver needs to be mentioned: it is based on an un-scaled version of the constitutive law defect source representation, that is, {\tt asg.x} approximately computes the solution of the system \ref{eqn:awedata} with the first equation replaced by
\begin{equation}
\label{eqn:asgdata}
\frac{\partial p}{\partial t}  =  - \kappa \nabla \cdot \bv +
h \delta(z-z_s).
\end{equation}
Denote by $S_{\rm asg}[c]$ the forward map produced by {\tt asg.x}. Then comparison of \ref{eqn:awedata} and \ref{eqn:asgdata} reveals that
\begin{equation}
\label{eqn:sreln}
S[c]=S_{\rm asg}[c]\kappa
\end{equation}
where $\kappa$ is shorthand for the operator of multiplication by $\kappa$. Accordingly, and approximate inverse for $S_{\rm asg}[c]$ is
\[
I \approx S_{\rm asg}[c]^{\dagger}S_{\rm asg}[c] = S_{\rm asg}[c]^{\dagger}S[c]\kappa^{-1}
\]
Since $S[c]S[c]^{\dagger} \approx I \approx S[c]^{\dagger}S[c]$, it follows that
\[
S_{\rm asg}[c]^{\dagger}=\kappa S[c]^{\dagger} = \kappa\Lambda[c]_s S[c]^T \Lambda[c]_r 
\]
\begin{equation}
\label{eqn:asginv}
= \kappa \Lambda[c]_s \kappa S_{\rm asg}[c]^T\Lambda[c]_r
\end{equation}
This is the approximate inverse computed in the examples. Note that only the values of $\kappa$ near the source datum $z=z_s$ play a role in the relation \ref{eqn:sreln} or in the definition \ref{eqn:asginv} of the IWAVE ASG approximate inverse.

These examples are based on two different material models. The first is a homogeneous model, with $\kappa=4.0$ GPa and $\rho=1.0$ g/cm$^{3}$. Figure \ref{fig:ptpwindh0} shows the shot gather computed in the configuration descritbed above for this choice of propagation medium. As mentioned in the preceding section, the velocity traces can be recorded in the same IWAVE workflow: these are displayed in 
\ref{fig:ptvzwindh0}.

The first group of plots show pressure traces for source position $z_s=3$ km, $x_s=3$ km, 201 receivers at $z_r=1$ km, $2 \le x_r \le 6$ km.

\plot{ptpwindh0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km. Homogeneous material model, $\kappa=4$ GPa, $\rho$=1 g/cm$^{3}$. Coarse grid: $dx=dz=20$ m, $dt=$ 8 ms, (2,4) staggered grid scheme. }
\plot{ptvzwindhh0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindh0}.}
\plot{srcvzghh0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindhh0} by time reversal (approximate inverse) algorithm explained in text.}
\plot{reptpwindhh0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzghh0}.} 
\plot{trcftrhh0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindh0}, \ref{fig:reptpwindhh0}.}.

\plot{bml0}{width=15cm}{Inhomogenous bulk modulus field, with low velocity lens.}
\plot{ptpwindl0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km. bulk modulus with low velocity lens as in Figure \ref{fig:bml0}, $\rho$=1 g/cm$^{3}$. Coarse grid: $dx=dz=20$ m, $dt=$ 8 ms, (2,4) staggered grid scheme. }
\plot{ptvzwindll0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindl0}.}

\plot{srcvzglh0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindll0} by time reversal (approximate inverse) algorithm explained in text, propagation in homogeneous material model (that is, ``wrong velocity'').}
\plot{reptpwindlh0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzglh0}, both approximate inversion and re-simulation in  homogeneous material model (``wrong velocity'').} 
\plot{trcftrlh0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindl0}, \ref{fig:reptpwindlh0}.}

\plot{srcvzgll0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindll0} by time reversal (approximate inverse) algorithm explained in text, propagation in bulk modulus of Figure \ref{fig:bml0} (that is, ``correct velocity'').}
\plot{reptpwindll0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzgll0}, both approximate inversion and re-simulation in bulk modulus of Figure \ref{fig:bml0} (``correct velocity'').} 
\plot{trcftrll0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindl0}, \ref{fig:reptpwindll0}.}

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

\append{RWI (=Wang-Yingst WRI) - a Linear Algebra Viewpoint}

Define $S[c] = PL[c]^{-1}$ = operator mapping extended source to data traces.

The Wang-Yingst version of the WRI objective function may be written
\[
\|S[c]g-d\|^2 + \alpha \|g-f\|^2
\]
``Most'' extended sources result in zero data traces - those are non-radiating sources, the null space of $S[c]$.

Rank Nullity Theorem: any $g$ may be written in a unique way as $g=S[c]^Th + n$, where $h$ is a collection of data traces, $S[c]^T$ is back propagation, and $n$ is in the null space of $S[c]$. Furthermore, the decomposition is orthogonal. Thus
\[
\|g-f\|^2 = \|S[c]^Th + n -f\|^2 = \|S[c]^Th - f_p\|^2 + \|n-f_n\|^2
\]
in which $f_p$ and $f_n$ are the orthoprojections on the orthocomplement of the null space, and the null space, of $S[c]$ respectively.

There is no use keeping $f_n$ around, as it radiates nothing - in fact, I suspect it is possible to see that point sources such as represented in \ref{eqn:ptsrc} are perpindicular to the null space already. In any case, simply choose $n=f_n$ and eliminate the second term, without affecting the data fit (by definition!).

Then WRI objective function can be re-written as
\begin{equation}
\label{eqn:wriobj}
\|S[c]S[c]^Th-d\|^2 + \alpha \|S[c]^Th -f_p\|^2
\end{equation}
and the extended model has been reduced from a full-space time volume ($g$) to a data gather ($h$).

{\bf Note added 2019.03.14:} Chao Wang's talk at GS 19 revealed that that they actually store the entire source ($g$ above). That is not so crazy, since the operator $S[c]S[c]^T$ poses the same complexity issue as time-domain RTM.

To see this, need an explicit description of $S[c]^T$. Regard $L[c]$ as the 2nd ord wave operator with velocity $c$: $L[c]=\frac{1}{c^2}\frac{\partial}{\partial t^2} - \nabla^2$. $L[c]^{-1}$ is ambiguous: define 
\[
L[c]^{-1_+}g = u, \mbox{ for } L[c]u=g \mbox{ and } u=0 \mbox{ for } t<0
\]
that is, $L[c]^{-1}_+$ is the causal inverse of $L[c]$. Solutions defined on time interval $[0,T]$. Let $R$ be time-reversal: $Ru(t)=u(T-t)$. Then the anticausal inverse $L[c]^{-1}_-$ of $L[c]$, solution $=0$ for $t>T$, can be written as $L[c]^{-1}_- = RL[c]^{-1}_+R$.

Denote by $\langle \cdot,\cdot \rangle$ the $L^2$ inner product in space.. Suppose $u$ is a causal solution of $L[c]u=f$, $w$ an anti-causal solution of $L[c]w=g$. Then 
\[
\int_0^T \langle L[c]^{-1}_+f, g \rangle = int_0^T \langle u, g \rangle = \int_0^T \langle u, \left( \frac{1}{c^2}\frac{\partial}{\partial t^2} - \nabla^2\right)w\rangle
\]
\[
= \int_0^T \langle \left( \frac{1}{c^2}\frac{\partial}{\partial t^2} - \nabla^2\right)u, w \rangle = \int_0^T \langle f, w \rangle = \int_0^T \langle f, L[c]^{-1}_-g \rangle
\]
since the causal resp. anti-causal solutions have zero Cauchy data at $t=0$ resp. $t=T$, and since the integration is over all of $\bR^3$ and I am assuming everything vanishing at $\infty$. Conclude that
\[
(L[c]^{-1}_+)^T = L[c]^{-1}_-
\]
so
\[
S[c]^T = (PL[c]^{-1}_+)^T = L[c]^{-1}_-P^T
\]

For convenience, ignore sampling and aperture, and define $P$ to be the restriction operator on $z=0$. Then $(P^T h)(x,y,z,t) = h(x,y,t)\delta(z)$.

The troublesome operator in the definition \ref{eqn:wriobj} is 
\[
S[c]S[c]^T = PL[c]^{-1}_+L[c]^{-1}_-P^T
\]
The first solution operator involves evolution backwards in time, and provides the data for the second solution operator. This leadss to the same computational complexity problem as encountered in RTM. It could be handled by storing the output of $L[c]^{-1}_-$, as the IONGeo folks apparently do, or storing boundary data (for acoustics - as usual, this would not work if the approach is extended to a lossy model such as viscoacoustics), or using a checkpointing scheme.

Another approach: write $w = L[c]^{-1}_-P^T h$. The object is to compute $u=L[c]^{-1}_+w$, that is, the solution of 
\[
(c^{-2}\partial_t^2 - \nabla^2)u(t) = w(t), 0\le t \le T; \,u(0) = \partial_t u(0) = 0
\]
Duhamel's principle would provide a solution in the form of integrals over time of solutions of homogeneous initial value problems:
\[
u(t) = \int_0^t d\tau v(t,\tau), \, (c^{-2}\partial_t^2 - \nabla^2)v(t,\tau) = 0, 0\le t \le T; \,v(t,\tau)_{t=\tau} = 0, \partial_t v(t,\tau)_{t=\tau} = w(\tau)
\]
Note that $w$ solves the homogenous wave equation except at $z=0$. That makes it tempting to use $w$ directly to define $v$. In fact, set 
\[
W(t) = -int_t^T ds w(s),\,v_0(t,\tau) = \frac{1}{2}(W(t) - W(2\tau-t)) 
\]
Then $v_0$ satisfies the Duhamel initial conditions. Note that because $w$ has zero Cauchy data at $t=T$,
\[
\partial_t W(t)= -\int_t^T ds \partial_t w(s),\,\partial_t^2 W(t)=-\int_t^T ds \partial^2_t w(s)
\]
so 
\[
(c^{-2}\partial_t^2 - \nabla^2)v_0(t,\tau) = -\int_t^T(c^{-2}\partial^2_t -\nabla^2)w + \int_{2\tau-t}^T(c^{-2}\partial^2_t -\nabla^2)
\]
For $t>\tau$, $t > 2\tau-t$, so 
\[
= -\int_{2\tau-t}^t(c^{-2}\partial^2_t -\nabla^2)w = -\int_{2\tau-t}^t P^Th
\]
and 
\[ 
v_0(t,\tau) = \int_{2\tau-t}^t w  
\] 
Consequently $v=v_0 + v_1$, in which $v_1$ is defined by 
\[
(c^{-2}\partial_t^2 - \nabla^2)v_1(t,\tau) = \int_{2\tau-t}^t P^Th, \,  v_1(t,\tau)_{t=\tau} = 0, \,  \partial_t v_1(t,\tau)_{t=\tau} = 0
\]

\append{Transpose of Full Acoustic Modeling Operator}
In this section I compute the adjoint operator form modeling with both pressure and velocity source inputs and pressure and velocity trace outputs. The forward modeling operator is defined by solving
\begin{eqnarray}
\label{eqn:fawedata}
\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
h_s \delta(z-z_s)), \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta ( \nabla p + l_s {\bf e}_z\delta(z-z_s)),\nonumber \\
p & =& 0 \mbox{ for } t<0\nonumber\\ 
\bv & = & 0 \mbox{ for } t<0 \nonumber\\
\end{eqnarray}
The new elements here are the point load time function $l_s(\bx,t;\bx_s)$, and the vertical unit vector ${\bf e}_z=(0,1)^T$ for 2D, $=(0,0,1)^T$ for 3D. The forward map ${\bf F}$ is defined as implicit in the discussion of the Intrinsically Preconditioned CG algorithm: let ${\bf h}_s=(h_s,l_s)^T$, then 
\begin{equation}
\label{eqn:ffwd}
{\bf F}{\bf h}_s =\left\{ \left[
\begin{array}{c}
p(\bx,t;\bx_s)\\
v_z(\bx,t;\bx_s)
\end{array}
\right]_{z=z_r}\right\}
\end{equation}

Let $\tilde{\bf h}_r=(\tilde{h}_r,\tilde{l}_r)^T$ be a function on $z=z_r$, 
and 
$\tilde{p},\tilde{\bv}$ solve the backwards-in-time boundary value problem:
\begin{eqnarray}
\label{eqn:faweadj}
\frac{\partial \tilde{p}}{\partial t} & = & \kappa (-\nabla \cdot \tilde{\bv} + 
 \tilde{h}_r \delta(z-z_r))\nonumber \\
\frac{\partial \tilde{\bv}}{\partial t} & = & \beta (\nabla \tilde{p} + \tilde{l}_r{\bf e}_z\delta(z-z_r),\nonumber \\
\tilde{p} & =& 0 \mbox{ for } t>T\nonumber\\ 
\tilde{\bv} & = & 0 \mbox{ for } t>T 
\end{eqnarray} 
Then
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  
\frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=T}
-
\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=0}
\]
\[
= 
\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)
\]
\[
= 
\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \tilde {p} +  p \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} \right.
\]
\[
+
\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \tilde{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \tilde{\bv}}{\partial t} \right)
\]
\[
= 
\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
 h_s \delta(z-z_s)\right) \tilde{p} + p \left(- \nabla \cdot \tilde{\bv} + 
 \tilde{h}_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  (- \nabla p + l_s{\bf e}_z\delta(z-z_s)) \cdot \tilde{\bv} + \bv \cdot (-\nabla \tilde{p} + \tilde{l}_r{\bf e}_z\delta(z-z_r)) \right)
\]
\[
= 
\int_{0}^{T}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv \tilde{p} + 
 h_s\tilde{p}|_{z=z_s} - p\nabla \cdot \tilde{\bv} + 
 \tilde{h}_r p|_{z=z_r}\right) \right.
\]
\[
+
\left.  p (\nabla \cdot \tilde{\bv}) + l_s \tilde{v}_z|_{z=z_r}+ (\nabla \cdot \bv) \tilde{p}  + l_r  v_z|_{z=z_r}\right)
\]
after integration by parts in the last two terms. Most of what is left cancels, leaving 
\[
0 = \int\,dt\,dx\,dy\, (h_s \tilde{p})_{z=z_s} + (p  \tilde{h}_r)_{z=z_r} + (l_s\tilde{v}_z)_{z=z_s} + (v_z \tilde{l}_r)_{z=z_r},
\]
whence
\begin{equation}
\label{eqn:fadj}
 {\bf F}^T {\bf h}_r = -\left[
\begin{array}{c}
\tilde{p}(\bx,t;\bx_s)\\
\tilde{\bf v}(\bx,t;\bx_s)
\end{array}
\right]_{z=z_s}
\end{equation}

%\begin{cor}\label{thm:bvps}
%Suppose that $d_s$ is a sufficiently regular function, = 0 for $t<0$, and denote by $p^{\pm},\bv^{\pm}$ the half-space solutions provided by Theorem \ref{thm:bvp}. Define the vector fields $p,\bv$ in $\bR^4$ by
%\begin{eqnarray}
%\label{eqn:putemtogether} 
%p(x,z) &=& p^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
%p(x,z_s)&=&d_s(x)\\
%\bv(x,z) &=& \bv^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
%\end{eqnarray}
%Then $p,\bv$ is a weak solution of the system \ref{eqn:awedata} with $h=[v]_n$.
%\end{cor}

Proof of Theorem \ref{thm:bvp}:

\begin{proof}
Since $\omega \cap \mbox{ supp }d$ is compact, it is possible to construct an extension of 
$d \in C^{\infty}_{0}(\bR^{d+1})$. Define $p_1^{\pm}, \bv_1^{\pm}$ to be strong solutions of the system
\begin{eqnarray}
\label{eqn:ichom}
\frac{\partial p^{\pm}_1}{\partial t} & = & - \kappa \nabla \cdot \bv^{\pm}_1 + F\mbox{ in } \Omega^{\pm}, \nonumber \\
\frac{\partial \bv^{\pm}_1}{\partial t} & = & - \beta \nabla p^{\pm}_1 + {\bf G}\mbox{ in } \Omega^{\pm},\nonumber \\
p^{\pm}_1 & = & 0 \mbox{ on } \omega \times \bR,\\
p^{\pm}_1 & = & 0, t \ll 0\\
\bv^{\pm}_1 & = & 0, t \ll 0 
\end{eqnarray}
with 
\[
F = -\gamma\frac{\partial d}{\partial t}, {\bf G} = -\nabla d.
\]
Existence, uniqueness, and regularity of these solutions are assured by a minor extension of results of \cite{BlazekStolkSymes:13}.
Set $p^{\pm} = p_1^{\pm} + d, \bv^{\pm} = \bv_1^{\pm}$; then $p^{\pm},\bv^{\pm}$ are the unique strong solutions of the boundary value problems \ref{eqn:awebdry}, and are smooth up to the boundary as indicated. 
\end{proof}

Proof of Corollary \ref{thm:bvps}:

\begin{proof}: Choose smooth functions $\phi \in C^{\infty}_0(\bR^{d+1}), {\bf \psi} \in  (C^{\infty}_0(\bR^{d+1}))^d$. Then
\[
\int_{\Omega^{\pm} \times \bR}\left (\gamma \frac{\partial \phi}{\partial t} +
\nabla \cdot {\bf \psi}\right)p^{\pm} + \left(\rho\frac{\partial {\bf
    \psi}}{\partial t}
  + \nabla \phi\right) \cdot \bv^{\pm}
\]
\[
= - \int_{\Omega \times \bR}\left (\gamma \frac{\partial p^{\pm}}{\partial t} +
\nabla \cdot \bv^{\pm}\right) \phi + \left(\rho\frac{\partial \bv^{\pm}}{\partial t}
  + \nabla p^{\pm}\right) \cdot {\bf \psi}
\]
\begin{equation}
\label{eqn:rhs1}
\pm \int_{\omega \times \bR} \bn \cdot {\bf \psi} p^{\pm} + \bn
  \cdot \bv^{\pm} \phi
\end{equation}
Adding equations \ref{eqn:rhs1} for the two choices of sign and taking into
account that $p^{\pm}$ coincide (by construction) on $\omega \times \bR$, obtain
\[
\int_{(\bR^d \times \bR}\left (\gamma \frac{\partial \phi}{\partial t} +
\nabla \cdot {\bf \psi}\right)p + \left(\rho\frac{\partial {\bf
    \psi}}{\partial t}
  + \nabla \phi\right) \cdot \bv
\]
\[
= \int_{\partial \Omega \times [0,T]} (\bn \cdot \bv^{+} -  \bn
  \cdot \bv^{-}) \phi
\]
Since $\phi$ and ${\bf \psi}$ are arbitrary, other than being smooth of compact support, this observation finishes the proof.
\end{proof}

\append{Variable Projection}

\[
J[c]=\min_h\frac{1}{2}(\|S[c]h-d\|^2 + \alpha^2\|Ah\|^2)
\]
\[
= \min_h \frac{1}{2}\|F[c]h-e\|^2
\]
\[
F[c]=\left(
\begin{array}{c}
S[c]\\
\alpha A
\end{array}
\right),\,\,
e=\left(
\begin{array}{c}
d\\
0
\end{array}
\right)
\]

\[
J[c]=\frac{1}{2}\|(F[c]N[c]^{-1}F[c]^T-I)e\|^2,\,\,N[c]=F[c]^TF[c]
\]
\[
DJ[c]\delta c = \langle (F[c]N[c]^{-1}F[c]^T-I)e, 
(DF[c]\delta c)N[c]^{-1}F[c]^Te +
F[c](-N[c]^{-1}(DN[c]\delta c)N[c]^{-1})F[c]^Te +
F[c]N[c]^{-1}(DF[c]\delta c)^Te \rangle 
\]

\[
= \langle (F[c]N[c]^{-1}F[c]^T-I)e, 
(DF[c]\delta c)N[c]^{-1}F[c]^Te -
F[c]N[c]^{-1}(DF[c]\delta c)^T F[c]N[c]^{-1})F[c]^Te -
F[c]N[c]^{-1}F[c]^T (DF[c]\delta c]N[c]^{-1})F[c]^Te +
F[c]N[c]^{-1}(DF[c]\delta c)^Te \rangle 
\]
\[
= \langle (F[c]N[c]^{-1}F[c]^T-I)e, 
(I-F[c]N[c]^{-1}F[c]^T) DF[c]\delta c)N[c]^{-1}F[c]^Te
F[c]N[c]^{-1}(DF[c]\delta c)^T(I-F[c]N[c]^{-1}F[c]^T) \rangle
\]
\[
=-\langle (I-F[c]N[c]^{-1}F[c]^T)^2e, DF[c]\delta c)N[c]^{-1}F[c]^Te\rangle
-\langle e,(I-F[c]N[c]^{-1}F[c]^T)F[c]N[c]^{-1}(DF[c]\delta c)^T (I-F[c]N[c]^{-1}F[c]^T)e
\]
\[
F[c]N[c]^{-1}F[c]^TF[c]N[c]^{-1}=F[c]N[c]^{-1}
\]
so 
\[
(I-F[c]N[c]^{-1}F[c]^T)F[c]N[c]^{-1} = 0
\]
and
\[
(I-F[c]N[c]^{-1}F[c]^T)^2 = (I-F[c]N[c]^{-1}F[c]^T)
\]
Whence
\[
DJ[c]\delta c =
-\langle (I-F[c]N[c]^{-1}F[c]^T)e, (DF[c]\delta c)N[c]^{-1}F[c]^Te\rangle
\]
Define $h[c]=N[c]^{-1}F[c]^Te$ to be the solution of the inner problem as function of $c$. Then
\[
DJ[c]\delta c = \langle F[c]h[c]-e,(DF[c]\delta c)h[c] \rangle
\]
which is the usual VPM gradient formula.

Gauss-Newton: define $G[c]=F[c]N[c]^{-1}F[c]^Te$. Need $DG[c], DG[c]^T$.
\[
DG[c]\delta c =( DF[c]\delta c)N[c]^{-1}F[c]^Te 
\]
\[
- F[c]N[c]^{-1}(DF[c]\delta c)^TF[c]N[c]^{-1}F[c]^Te + F[c]N[c]^{-1} F[c]^T(DF[c]\delta c))N[c]^{-1}F[c]^Te 
\]
\[
+ F[c]N[c]^{-1}(DF[c]\delta c)^Te
\]
\[
=(I-F[c]N[c]^{-1} F[c]^T)(DF[c]\delta c))N[c]^{-1}F[c]^Te + F[c]N[c]^{-1}(DF[c]\delta c)^T(I-F[c]N[c]^{-1} F[c]^T) e
\]
Reinterpretation: set
\begin{eqnarray}
h[c] &=&N[c]^{-1}F[c]^Te \nonumber\\
r[c] &= &F[c]h[c]-e \nonumber\\
E_1[c]\delta c & = & (DF[c]\delta c)h[c] \nonumber\\
H_1[c]\delta c &=&N[c]^{-1}F[c]^T E_1[c]\delta c \nonumber\\
R_1[c]\delta c &=& (F[c]H_1[c]-E_1[c])\delta c \nonumber\\
H_2[c]\delta c &=&N[c]^{-1}(DF[c]\delta c)^Tr[c] \nonumber\\
R_2[c]\delta c &=&F[c]H_2[c]\delta c
\end{eqnarray}
Then
\[
DG[c]\delta c = -(R_1[c]+R_2[c])\delta c
\]
\[
= (I-F[c]N[c]^{-1}F[c]^T)DF[c]\delta c)h[c] -F[c]N[c]^{-1}(DF[c]\delta c)^Tr[c]
\]
Adjoints: write $(DF[c]\delta c)h = DF[c](\delta c,h)$, 
\begin{eqnarray}
\langle F[c]h,\phi \rangle &=& \langle h,F[c]^T\phi\rangle \nonumber\\
\langle DF[c](\delta c,h),\phi \rangle &=& \langle \delta c, DF[c]^*(\phi,h) \rangle
\end{eqnarray}
\[
\langle DF[c](\delta c,h),\phi \rangle = \langle \delta c,DF[c]^*(\phi,h)\rangle
\]
\[
\langle E_1[c]\delta c,\phi \rangle = \langle DF[c](\delta c,h[c]),\phi \rangle
\]
\[
= \langle \delta c, DF[c]^*(\phi,h[c]) \rangle
\]
so $E_1[c]^* = DF[c]^*(\cdot,h[c])$, and
\[
\langle F[c]H_1[c]\delta c, \phi \rangle = \langle F[c] N[c]^{-1}F[c]^TE_1[c]\delta c,\phi \rangle
\]
\[
=\langle E_1[c]\delta c,F[c]N[c]^{-1}F[c]^T\phi\rangle
\]
\[
=\langle \delta c, DF[c]^*(F[c]N[c]^{-1}F[c]^T\phi,h[c])\rangle
\]
\[
\langle R_1[c]\delta c,\phi \rangle = \langle \delta c, DF[c]^*((F[c]N[c]^{-1}F[c]^T-I)\phi,h[c])
\]
That is,
\[
R_1[c]^*\phi = DF[c]^*((F[c]N[c]^{-1}F[c]^T-I)\phi,h[c])
\]
\[
\langle R_2[c]\delta c,\phi \rangle = \langle F[c]N[c]^{-1}(DF[c]\delta c)^Tr[c],\phi \rangle
\]
\[
\langle r[c],DF[c](\delta c,N[c]^{-1}F[c]^T\phi)\rangle = \langle \delta c, DF[c]^*(r[c], N[c]^{-1}F[c]^T\phi) \rangle
\]
That is,
\[
R_2[c]^*\phi = DF[c]^*(r[c], N[c]^{-1}F[c]^T\phi)
\]

\end{document}
\section{Figures for talk}
\plot{pulse00}{width=\textwidth}{Bandpass filter pulse}
\plot{fwitrbulkmodinvlh0}{width=\textwidth}{Inversion for bulk modulus of data displayed in Figure \ref{fig:ptpwindl0}. Initial model: constant bulk modulus = 4.0 GPa. Algorithm: 20 iterations of trust region Krylov-Gauss-Newton algorithm with preconditioning by box smoother of radius 5 points, repeated 10 times.  Each KGN step computed by up to 20 Conjugate Gradient iterations.} 
\plot{fwitrdatapreslh0}{width=\textwidth}{Data residual produced by 20 itersions of KGN FWI for bulk modulus, starting at constant bulk modulus. Residual RMS approximately 26\% of data RMS.}
\plot{cgwhestsourceplh0}{width=\textwidth}{Extended inversion for source on 1500 m $\le x \le$ 5500 m, $z$=3000 m, data from Figure \ref{fig:ptpwindl0}, homogeneous bulk modulus = 4.0 GPA.}
\plot{cgwhresdataplh0}{width=\textwidth}{Residual data from extended source inversion shown in Figure \ref{fig:cgwhestsourceplh0}. RMS is 3\% of data RMS.}
\plot{bmh0s}{width=\textwidth}{Iteration 1}
\plot{cgw1bulkupdlh0}{width=\textwidth}{Iteration 1 SSE bulkmod estimate}
\plot{cgw2bulkupdlh0}{width=\textwidth}{Iteration 2 SSE bulkmod estimate}
\plot{cgw3bulkupdlh0}{width=\textwidth}{Iteration 3 SSE bulkmod estimate}
\plot{cgw4bulkupdlh0}{width=\textwidth}{Iteration 4 SSE bulkmod estimate}
\plot{cgw5bulkupdlh0}{width=\textwidth}{Iteration 5 SSE bulkmod estimate}
\plot{cgw6bulkupdlh0}{width=\textwidth}{Iteration 6 SSE bulkmod estimate}
\plot{cgw7bulkupdlh0}{width=\textwidth}{Iteration 7 SSE bulkmod estimate}
\plot{cgw8bulkupdlh0}{width=\textwidth}{Iteration 8 SSE bulkmod estimate}
\plot{cgw9bulkupdlh0}{width=\textwidth}{Iteration 9 SSE bulkmod estimate}
\plot{cgw10bulkupdlh0}{width=\textwidth}{Iteration 10 SSE bulkmod estimate}
\plot{cgw11bulkupdlh0}{width=\textwidth}{Iteration 11 SSE bulkmod estimate}
\plot{cgw12bulkupdlh0}{width=\textwidth}{Iteration 12 SSE bulkmod estimate}
\plot{cgw13bulkupdlh0}{width=\textwidth}{Iteration 13 SSE bulkmod estimate}
\plot{cgw14bulkupdlh0}{width=\textwidth}{Iteration 14 SSE bulkmod estimate}
\plot{cgw15bulkupdlh0}{width=\textwidth}{Iteration 15 SSE bulkmod estimate}
\plot{cgw22bulkupdlh0}{width=\textwidth}{Iteration 22 SSE bulkmod estimate}
\plot{cgw30bulkupdlh0}{width=\textwidth}{Iteration 30 SSE bulkmod estimate}
\plot{fwitrsse1bulkmodinvlh0}{width=\textwidth}{Inversion for bulk modulus of data displayed in Figure \ref{fig:ptpwindl0}. Initial model: result of 15 steepest descent steps of SSE, Figure \ref{fig:cgw15bulkupdlh0}. Algorithm: 20 iterations of trust region Krylov-Gauss-Newton algorithm with preconditioning by box smoother of radius 5 points, repeated 10 times.  Each KGN step computed by up to 20 Conjugate Gradient iterations.} 
\plot{fwitrsse1datapreslh0}{width=\textwidth}{Data residual produced by 20 itersions of KGN FWI for bulk modulus, starting at output of 15 steepest descent steps of SSE, Figure \ref{fig:cgw15bulkupdlh0}. Residual RMS approximately 1.2\% of data RMS.}
\plot{cgw0estsourceplh0}{width=\textwidth}{Iteration 1 SSE source estimate alpha=4.36 $\times 10^{-6}$}
\plot{cgw1estsourceplh0}{width=\textwidth}{Iteration 2 SSE source estimate alpha=5.93 $\times 10^{-6}$}
\plot{cgw2estsourceplh0}{width=\textwidth}{Iteration 3 SSE source estimate alpha=6.82 $\times 10^{-6}$}
\plot{cgw3estsourceplh0}{width=\textwidth}{Iteration 4 SSE source estimat alpha=7.93 $\times 10^{-6}$e}
\plot{cgw4estsourceplh0}{width=\textwidth}{Iteration 5 SSE source estimate alpha=9.29 $\times 10^{-6}$}
\plot{cgw5estsourceplh0}{width=\textwidth}{Iteration 6 SSE source estimate alpha=1.1 $\times 10^{-5}$}
\plot{cgw6estsourceplh0}{width=\textwidth}{Iteration 7 SSE source estimate alpha=1.32 $\times 10^{-5}$}
\plot{cgw7estsourceplh0}{width=\textwidth}{Iteration 8 SSE source estimate alpha=1.32 $\times 10^{-5}$}
\plot{cgw8estsourceplh0}{width=\textwidth}{Iteration 9 SSE source estimate alpha=1.72 $\times 10^{-5}$}
\plot{cgw9estsourceplh0}{width=\textwidth}{Iteration 10 SSE source estimate alpha=2.2 $\times 10^{-5}$}
\plot{cgw10estsourceplh0}{width=\textwidth}{Iteration 11 SSE source estimate alpha=2.79 $\times 10^{-5}$}
\plot{cgw11estsourceplh0}{width=\textwidth}{Iteration 12 SSE source estimate alpha=2.79 $\times 10^{-5}$}
\plot{cgw12estsourceplh0}{width=\textwidth}{Iteration 13 SSE source estimate alpha=2.79 $\times 10^{-5}$}
\plot{cgw13estsourceplh0}{width=\textwidth}{Iteration 14 SSE source estimate alpha=2.79 $\times 10^{-5}$}
\plot{cgw14estsourceplh0}{width=\textwidth}{Iteration 15 SSE source estimate alpha=2.79 $\times 10^{-5}$}
\plot{cgw21estsourceplh0}{width=\textwidth}{Iteration 22 SSE source estimate alpha=5.73 $\times 10^{-5}$}
\plot{cgw29estsourceplh0}{width=\textwidth}{Iteration 30 SSE source estimate alpha=9.1 $\times 10^{-5}$}
\plot{cgw0resdataplh0}{width=\textwidth}{Iteration 1 SSE data residual}
\plot{cgw1resdataplh0}{width=\textwidth}{Iteration 2 SSE data residual}
\plot{cgw2resdataplh0}{width=\textwidth}{Iteration 3 SSE data residual}
\plot{cgw3resdataplh0}{width=\textwidth}{Iteration 4 SSE data residual}
\plot{cgw4resdataplh0}{width=\textwidth}{Iteration 5 SSE data residual}
\plot{cgw5resdataplh0}{width=\textwidth}{Iteration 6 SSE data residual}
\plot{cgw6resdataplh0}{width=\textwidth}{Iteration 7 SSE data residual}
\plot{cgw7resdataplh0}{width=\textwidth}{Iteration 8 SSE data residual}
\plot{cgw8resdataplh0}{width=\textwidth}{Iteration 9 SSE data residual}
\plot{cgw9resdataplh0}{width=\textwidth}{Iteration 10 SSE data residual}
\plot{cgw10resdataplh0}{width=\textwidth}{Iteration 11 SSE data residual}
\plot{cgw11resdataplh0}{width=\textwidth}{Iteration 12 SSE data residual}
\plot{cgw12resdataplh0}{width=\textwidth}{Iteration 13 SSE data residual}
\plot{cgw13resdataplh0}{width=\textwidth}{Iteration 14 SSE data residual}
\plot{cgw14resdataplh0}{width=\textwidth}{Iteration 15 SSE data residual}

### 2019-10-19 repro/papers/vardentheory/paper.tex
repro docs dir added

\title{Asymptotic Inversion of the Variable Density Acoustic
  Model\\Part 1: Theory}

\author{
Raanan Dafni\thanks{Department of Computational and Applied Mathematics, Rice University, 
Houston, TX, 77005, USA, 
{\tt symes@rice.edu}},
William Symes\thanks{Department of Computational and Applied Mathematics, Rice University, 
Houston, TX, 77005, USA, 
{\tt symes@rice.edu}}
}

\lefthead{Dafni Symes}

\righthead{Asymptotic Acoustic Inversion}

\maketitle
\parskip 12pt

\begin{abstract}
Techniques developed for asymptotic wave equation based inversion of
subsurface offset extended constant density acoustics also apply to variable density
acoustics. The data may be matched by proper choice of an extended
relative bulk modulus perturbation, with zero density
perturbation. This bulk modulus perturbation may be treated as an
offset-dependent reflectivity volume, and AVA-like techniques applied
to extract asymptotically correct estimates of velocity and density
perturbations, assuming that the background velocity is kinematically
consistent with the data. This first part of the paper presents a 
theoretical justification of the inversion algorithm. The second part
describes implementation details and shows numerical examples.
\end{abstract}
\setlength{\parindent}{0cm}
\section{Extended Variable Density Acoustics}
The dynamical laws of linear acoustics may be written
\begin{eqnarray}
\label{eqn:awe}
\frac{1}{\kappa}\frac{\partial p}{\partial t}& =& -\nabla \cdot \bv +
g,\nonumber \\
\rho \frac{\partial \bv}{\partial t} & = & -\nabla p.
\end{eqnarray}
In this system, $p$ is pressure, $\bv$ is particle velocity, $\kappa$
is bulk modulus, $\rho$ is material density, and $g$ is the defect
representation of an energy source. The first equation is the
constitutive law relating stress (pressure) rate to strain rate (actually to its
trace, which is proportional to the velocity divergence). The second
expresses momentum balance. When completed by suitable initial and
boundary conditions, the system (\ref{eqn:awe}) has a unique finite
energy solution $(p,\bv)$ under very mild constraints on $\kappa,
\rho$ \cite[]{BlazekStolkSymes:13}.

Define the acoustic forward map as
\begin{equation}
\label{eqn:afm}
{\cal F}[\kappa,\rho] = {\cal S}[p,\bv],
\end{equation}
in which ${\cal S}$ is a (linear) sampling operator, extracting traces of pressure
and/or velocity at receiver points $\bx_r$ for $t \in [0,T]$. Note that
the source function $g$ also depends on an experimental
parameter. While other possibilities are interesting, for present
purposes assume that $g$ depends on a location $\bx_s$, so the value of
$\cal{F}$ does as well.

The subsurface offset extension of this system generalizes $\kappa$
and $\rho$ to be operators. That is, the relations between stress
and strain and between force and acceleration may be non-local.
Of course one
can think of the ordinary stress-strain relation as being mediated by
an operator, namely  the operator of multiplication by
$\kappa$. Multiplication by $1/\kappa$ is just the inverse of this
operator. Extended model replaces multiplication by $\kappa$ by the
action of a (possibly non-local) operator $\bar{\kappa}$, and multiplication by
$\rho$ by the action of a (possibly non-local) operator $\bar{\rho}$. Just as $\kappa$ is
positive and multiplication by (real) $\kappa$ is a symmetric
operator, require $\bar{\kappa}$ to be symmetric positive
definite and bounded on $L^2$, and similar for $\bar{\rho}$.

As it happens, the extension of both parameters $\kappa$ and $\rho$,
as opposed to just $\kappa$, is not
terribly useful. Also the quasi-physical justification for extending just
$\kappa$ is clearer. However it is no more difficult to work through the
consequences of extending both, so we shall do so, and specialize at
the end to draw conclusions for $\kappa$-only extended inversion, and
for ordinary non-extended inversion for
$\kappa$ and $\rho$.

The extended acoustics system for extended pressure and velocity
fields $\bar{p}$ and $\bar{\bv}$ is
\begin{eqnarray}
\label{eqn:eawe}
\bar{\kappa}^{-1}\frac{\partial \bar{p}}{\partial t}& =& -\nabla \cdot
\bar{\bv} + g,\nonumber \\
\bar{\rho} \frac{\partial \bar{\bv}}{\partial t} & = & -\nabla \bar{p}.
\end{eqnarray}
\cite{BlazekStolkSymes:13} show that $\bar{p}$ and $\bar{\bv}$ are
well-defined, as is the extended forward map, defined as 
\begin{equation}
\label{eqn:eafm}
\bar{{\cal F}}[\bar{\kappa},\bar{\rho}] = {\cal S}(\bar{p},\bar{\bv}).
\end{equation}

\section{Linearization}
Using the symbol $\delta$ as shorthand for perturbation, obtain for the
perturbation $\delta \bar{p}, \delta \bar{\bv}$ in the solution of
(\ref{eqn:eawe}) resulting from $\delta \bar{\kappa}, \delta{\rho}$ the
system
\begin{eqnarray}
\label{eqn:leawe}
\bar{\kappa}^{-1}\frac{\partial \delta \bar{p}}{\partial t}& =& -\nabla \cdot
\delta \bar{\bv} +
\bar{\kappa}^{-1}\delta\bar{\kappa}\bar{\kappa}^{-1}\frac{\partial \bar{p}}{\partial t},\nonumber\\
\bar{\rho} \frac{\partial \delta \bar{\bv}}{\partial t} & = & -\nabla \delta
\bar{p} - \delta \bar{\rho} \frac{\partial \bar{\bv}}{\partial t}.
\end{eqnarray}
The linearized forward map is defined in terms of the solution of
(\ref{eqn:leawe}) as 
\begin{equation}
\label{eqn:leafm}
D\bar{{\cal F}}[\bar{\kappa},\bar{\rho}][\delta \bar{\kappa},\delta \bar{\rho}] =
{\cal S}(\delta \bar{p},\delta \bar{\bv}),
\end{equation}
thanks to the linearity of ${\cal S}$.

Physically, of course, the extended system is nonsense, in that it
violates the continuum mechanical axiom of
no-action-at-a-distance.

\section{Expression in terms of Green's Functions}
From here on, assume that the reference fields are physical.

Differentiate the first equation in (\ref{eqn:leawe}) and
(\ref{eqn:awe}) with respect to $t$,
and use the second equation to eliminate $\delta \bar{\bv}$ and $\bv$ to arrive at
the second order wave equations for $p,\delta \bar{p}$:
\begin{eqnarray}
\label{eqn:awe2}
\frac{1}{\kappa}\frac{\partial^2 p}{\partial t^2} &=&
\nabla \cdot \frac{1}{\rho} \nabla p + \frac{\partial g}{\partial t}\\
\label{eqn:leawe2}
\frac{1}{\kappa}\frac{\partial^2 \delta \bar{p}}{\partial t^2} &=&
\nabla \cdot \frac{1}{\rho} \nabla \delta \bar{p} + \frac{1}{\kappa}
\delta \bar{\kappa} \frac{1}{\kappa}\frac{\partial^2 p}{\partial t^2} -
  \nabla \cdot \frac{1}{\rho}\delta \bar{\rho}\frac{1}{\rho} \nabla p. 
\end{eqnarray}
Note that the two copies of $1/\kappa$ must be kept separate, on
either side of $\delta \bar{\kappa}$, as the latter is an operator and
does not commute with multiplication by $1/\kappa$ unless $\delta
\bar{\kappa}$ is itself physical, that is, multiplication by a
function. The same applies to $1/\rho$.

To express the relation between $\delta \bar{\kappa}$, $\delta \rho$
and $D\bar{{\cal F}}[\bar{\kappa},\rho][\delta \bar{\kappa},\delta
\rho]$ in integral form, assume that (at least formally) $\delta
\bar{\kappa}$, $\delta \bar{\rho}$ are integral operators:
\begin{eqnarray}
\label{eqn:dkappa}
(\delta \bar{\kappa} u)(\bx) &=& \int d\by \delta 
  \bar{\kappa}(\bx,\by)u(\by),\\
(\delta \bar{\rho} u)(\bx) &=& \int d\by \delta 
  \bar{\rho}(\bx,\by)u(\by),\\
\end{eqnarray}
and introduce the Green's function $G(\bx,t;\by)$, the causal solution
of
\begin{equation}
\label{eqn:green}
\left(\frac{1}{\kappa}\frac{\partial^2 G}{\partial t^2} - \nabla_{\bx} \cdot
  \frac{1}{\rho} \cdot \nabla_{\bx}G\right)(\bx,t;\by) =
\delta(\bx-\by)\delta(t), \,\, G=0 \mbox{ for } t < 0.
\end{equation}
Then, by superposition,
\[
\delta \bar{p}(\bx,t)  = \int d\by \int d\tau G(\bx,t-\tau;\by)
\frac{1}{\kappa(\by)}  \int d\bz \delta \bar{\kappa}(\by,\bz) \frac{1}{\kappa(\bz)}
 \frac{\partial^2 p}{\partial t^2}(\bz,\tau) 
\]
\begin{equation}
\label{eqn:intrep1}
- \int d\by \int d\tau G(\bx,t-\tau;\by)\nabla_{\by} \cdot \left(
\frac{1}{\rho(\by)}\int d\bz \delta \bar{\rho}(\by,\bz)\frac{1}{\rho(\bz)}) \nabla_{\bz} p(\bz,\tau)\right). 
\end{equation}
It is conventional to introduce alternate integration variables
\begin{eqnarray}
\label{eqn:jon}
\bx &=& \frac{1}{2}(\by + \bz) \nonumber\\
\bh &=& \frac{1}{2}(\by - \bz).
\end{eqnarray}
These play the roles of Claerbout's sunken midpoint and half-offset
vectors (REF). Re-parametrize $\delta\bar{\kappa}(\by,\bz) \rightarrow
\delta\bar{\kappa}(\bx,\bh)$, and similarly for $\delta \bar{\rho}$. Then (\ref{eqn:intrep1}) becomes
\[
\delta \bar{p}(\bx,t)  = \int d\bx\int d\bh \int d\tau
\left( G(\bx_r,t-\tau;\bx+\bh) 
\frac{1}{\kappa(\bx+\bh)}  \delta \bar{\kappa}(\bx,\bh) \frac{1}{\kappa(\bx-\bh)}
 \frac{\partial^2 p}{\partial t^2}(\bx-\bh,\tau)
\right.
\]
\begin{equation}
\label{eqn:intrep2}
-\left.  G(\bx_r,t-\tau;\bx+\bh) \nabla \cdot \left[
\frac{1}{\rho(\bx+\bh)}  \delta \bar{\rho}(\bx,\bh) \frac{1}{\rho(\bx-\bh)}\nabla
 p(\bx-\bh,\tau)\right]\right)
\end{equation}

\section{The Impulse Response}
We make two restrictions on the source term $g$ in the systems
\ref{eqn:awe} and \ref{eqn:eawe}:
\begin{itemize}
\item For each source position $\bx_s$, $g$ represents an isotropic
  point radiator located at $\bx_s$;
\item The source spectrum is flat, at least for frequencies in the
  passband of significant S/N.
\end{itemize}
Neither assumption reflects the reality of sources recorded in the
field. While the localization assumption may be acceptable, many field
sources have significant radiation patterns, and certainly non-flat
spectra. We are in effect ignoring directional variation in radiation,
and assuming that (in principle perfect) source signature
deconvolution has been performed.

Introduction of a multipole source
expansion would allow us to carry out an analysis similar to that to
follow, while accounting for significantly directional
sources, however we reserve that opportunity for another
work. Similarly, while source estimation could be folded into the
estimation of the material parameters, but that is for another time.

Since the time dependence is equivalent to an impulse over the data
bandwidth, we go ahead and treat it as an impulse with infinite
bandwidth. This assumption is standard in work on asymptotics since
its outset \cite[]{CohBlei:77}, and has some negative consequences, for example
inaccuracy at passband edges. However we will stick with it for
convenience.

The upshot is 
\begin{equation}
\label{eqn:isopt}
g(\bx,t) = \delta(t)\delta(\bx-\bx_s).
\end{equation}
Then $p(\bx,t) = \frac{\partial G}{\partial t}(\bx,\cdot;\bx_s)(t)$ depends on $\bx_s$ as
does its perturbation.

Further assume that the sampling operator simply sampling pressure at receiver
points $\bx_r$. Then
\[
D\bar{{\cal F}}[\kappa,\rho][\delta 
\bar{\kappa},\delta\bar{\rho}](\bx_r,t;\bx_s) 
\]
\[
= \int d\bx\int d\bh \int d\tau \left(G(\bx_r,t-\tau;\bx+\bh) 
\frac{1}{\kappa(\bx+\bh)}  \delta \bar{\kappa}(\bx,\bh) \frac{1}{\kappa(\bx-\bh)}
 \frac{\partial^3 G}{\partial t^3}(\bx-\bh,\tau;\bx_s) 
\right. 
\]
\[
-\left.  G(\bx_r,t-\tau;\bx+\bh) \nabla \cdot \left[
\frac{1}{\rho(\bx+\bh)}  \delta \bar{\rho}(\bx,\bh) \frac{1}{\rho(\bx-\bh)}\nabla 
 \frac{\partial G}{\partial t}(\bx-\bh,\tau;\bx_s)\right]\right)
\]
\[
= \int d\bx\int d\bh \int d\tau \left(G(\bx_r,t-\tau;\bx+\bh) 
\frac{1}{\kappa(\bx+\bh)}  \delta \bar{\kappa}(\bx,\bh) \frac{1}{\kappa(\bx-\bh)}
 \frac{\partial^3 G}{\partial t^3}(\bx-\bh,\tau;\bx_s) 
\right.
\]
\begin{equation}
\label{eqn:intrep3}
+\left. \nabla G(\bx_r,t-\tau;\bx+\bh) \cdot \left[
\frac{1}{\rho(\bx+\bh)}  \delta \bar{\rho}(\bx,\bh) \frac{1}{\rho(\bx-\bh)}\nabla 
\frac{ \partial G}{\partial t}(\bx-\bh,\tau;\bx_s)\right]\right)
\end{equation}
after integration by parts.

For future reference, we include the integral expression for the
adjoint of $D\bar{{\cal F}}[\kappa,\rho]$:
\[
D\bar{{\cal F}}[\kappa,\rho]^Td(\bx,\bh) =
\]
\begin{equation}
\label{intrep3adj}
\left[
\begin{array}{c}
\frac{1}{\kappa(\bx+\bh)}\frac{1}{\kappa(\bx-\bh)}\int d\bx_r d\bx_s dt d\tau G(\bx_r,t-\tau;\bx+\bh) 
 \frac{\partial^3 G}{\partial t^3}(\bx-\bh,\tau;\bx_s)
  d(\bx_r,t;\bx_s)  \\
\frac{1}{\rho(\bx+\bh)}\frac{1}{\rho(\bx-\bh)}\int d\bx_r d\bx_s dt
  d\tau \nabla G(\bx_r,t-\tau;\bx+\bh) \cdot \nabla \frac{ \partial
  G}{\partial t}(\bx-\bh,\tau;\bx_s) d(\bx_r,t;\bx_s)
\end{array}
\right].
\end{equation}

It is convenient for analysis (and for QC!) to make the perturbations
$\delta \bar{\kappa}$ and $\delta \bar{\rho}$ relative to the
reference quantities. Define the {\em extended reflectivities} $\bar{r}_{\kappa}$
and $\bar{r}_{\rho}$ as 
\begin{eqnarray}
\label{eqn:relfs}
\bar{r}_{\kappa}(\bx,\bh)& = & \frac{\delta
                               \bar{\kappa}(\bx,\bh)}{\sqrt{\kappa(\bx+\bh)\kappa(\bx-\bh)}}
                               \nonumber \\
\bar{r}_{\rho}(\bx,\bh)& = & \frac{\delta \bar{\rho}(\bx,\bh)}{\sqrt{\rho(\bx+\bh)\rho(\bx-\bh)}}
\end{eqnarray}
Define the relative linearized forward operator $\bar{F}[\kappa,\rho]$ by
\begin{equation}
\label{eqn:nondim}
\bar{F}[\kappa,\rho][\bar{r}_{\kappa},\bar{r}_{\rho}] =
D\bar{{\cal F}}[\kappa,\rho][\delta 
\bar{\kappa},\delta\bar{\rho}](\bx_r,t;\bx_s) 
\end{equation}
where $\bar{r}_{\kappa},\bar{r}_{\rho}$ are related to $\delta 
\bar{\kappa},\delta\bar{\rho}$ by equation \ref{eqn:relfs}. Then
\[
\bar{F}[\kappa,\rho][\bar{r}_{\kappa},\bar{r}_{\rho}](\bx_r,t;\bx_s) 
=\int d\bx\int d\bh \int d\tau 
\]
\[
\left(G(\bx_r,t-\tau;\bx+\bh) 
\frac{1}{\sqrt{\kappa(\bx+\bh)}}  \bar{r}_{\kappa}(\bx,\bh) \frac{1}{\sqrt{\kappa(\bx-\bh)}}
 \frac{\partial^3 G}{\partial t^3}(\bx-\bh,\tau;\bx_s) 
\right.
\]
\begin{equation}
\label{eqn:intrep3nd}
+\left. \nabla G(\bx_r,t-\tau;\bx+\bh) \cdot \left[
\frac{1}{\sqrt{\rho(\bx+\bh)}} \bar{r}_{\rho}(\bx,\bh) \frac{1}{\sqrt{\rho(\bx-\bh)}}\nabla 
 \frac{\partial G}{\partial t}(\bx-\bh,\tau;\bx_s)\right]\right)
\end{equation}

\section{Simple Ray Geometry}

Suppose that unique rays connect sources and receivers to scattering
points, as is the case for smooth $\kappa, \rho$ when sources,
receivers, and scatterers are sufficiently close together. Then in 2D,
\begin{equation}
\label{eqn:simp2d}
G(\bx,t;\by) = a(\bx,\by)t_+^{-1/2}(t-T(\bx,\by)) 
\end{equation}
whereas in 3D,
\begin{equation}
\label{eqn:simp3d}
G(\bx,t;\by) = a(\bx,\by)\delta(t-T(\bx,\by)) 
\end{equation}
in which $T$ is traveltime and $a$ is geometric amplitude.

For simplicity assume dimension = 3 for the remainder of this
section. The following section will present analgous computations for 2D.

Up to a lower order error,
\[
\nabla G \approx -a \delta'(t-T)\nabla T.
\]
Substituting this approximation and using $\delta'*\delta''=\delta'''$, \ref{eqn:intrep3} becomes
\[
\bar{F}[\kappa,\rho][\bar{r}_{\kappa},\bar{r}_{\rho}](\bx_r,t;\bx_s) 
\]
\[
\approx \int d\bx\int d\bh a(\bx_r,\bx+\bh) a(\bx_s,\bx-\bh) 
\delta'''(t-(T(\bx_r,\bx+\bh)+T(\bx_s,\bx-\bh))) 
\]
\[
\times \left(
\frac{1}{\sqrt{\kappa(\bx+\bh)}}  \bar{r}_{\kappa}(\bx,\bh) \frac{1}{\sqrt{\kappa(\bx-\bh)}}
%\frac{1}{\kappa(\bx+\bh)}  \delta \bar{\kappa}(\bx,\bh) \frac{1}{\kappa(\bx-\bh)}
\right.
\]
\begin{equation}
\label{eqn:intrep4}
+\left. 
   \nabla  T(\bx_r,\bx+\bh) \cdot \nabla T(\bx_s,\bx-\bh) 
\frac{1}{\sqrt{\rho(\bx+\bh)}} \bar{r}_{\rho}(\bx,\bh) \frac{1}{\sqrt{\rho(\bx-\bh)}}
%\frac{1}{\rho(\bx+\bh)}  \delta \bar{\rho}(\bx,\bh) \frac{1}{\rho(\bx-\bh)}
\right)
\end{equation}
We remove the third 
time derivative by applying the causal time antiderivative by $I_t$,
\[
I_t^3 \bar{F}[\kappa,\rho][\bar{r}_{\kappa},\bar{r}_{\rho}](\bx_r,t;\bx_s) 
\]
\[
= \int d\bx\int d\bh a(\bx_r,\bx+\bh) a(\bx_s,\bx-\bh) 
\delta(t-(T(\bx_r,\bx+\bh)+T(\bx_s,\bx-\bh)) ) 
\]
\[
\times \left(
\frac{1}{\sqrt{\kappa(\bx+\bh)}}  \bar{r}_{\kappa}(\bx,\bh) \frac{1}{\sqrt{\kappa(\bx-\bh)}}
%\frac{1}{\kappa(\bx+\bh)}  \delta \bar{\kappa}(\bx,\bh) \frac{1}{\kappa(\bx-\bh)}
\right. 
\]
\begin{equation}
\label{eqn:intrep5}
+\left. 
   \nabla  T(\bx_r,\bx+\bh) \cdot \nabla T(\bx_s,\bx-\bh) 
%\frac{1}{\rho(\bx+\bh)}  \delta \bar{\rho}(\bx,\bh)
%\frac{1}{\rho(\bx-\bh)}
\frac{1}{\sqrt{\rho(\bx+\bh)}} \bar{r}_{\rho}(\bx,\bh) \frac{1}{\sqrt{\rho(\bx-\bh)}}
\right)
\end{equation}
%The normal operator of $I_t^2D{\cal F}$ is a operator matrix of the
%form
%\begin{equation}
%\label{eqn:normalop}
%(I_t^2D{\cal F}[\kappa,\rho])^*(I_t^2D{\cal F}[\kappa,\rho])
%=\left(\begin{array}{cc}
%M_{\kappa \kappa} & M_{\kappa \rho}\\
%M_{\kappa \rho}^* & M_{\rho \rho}
%\end{array}
%\right)
%\end{equation}
%in which $M_{i,j}, i,j \in \{\kappa, \rho\}$ take the form of the GRT
%operator defined by a 3D form of \cite{HouSymes:17}, equation 4, 
%\begin{equation}
%\begin{aligned}
%\label{eqn:grt3d}
%M u(\bx,\bh) =& \int d\bx_r d\bx_s d\bx' d\bh' A(\bx_r,\bx_s,\bx,\bh,\bx',\bh')\\
%&\times \delta(\phi(\bx_r,\bx_s,\bx,\bh) - \phi(\bx_r,\bx_s,\bx',\bh')) u(\bx',\bh'),
%\end{aligned}
%\end{equation}

\section{Two Dimensions and Horizontal Offset}
Since the 2D case is a bit simpler, we now turn to it, and also
introduce the horizontal offset restriction. Instead of $\bx$ we shall
write $(x,z)$, also $x_r,x_s$ rather than $\bx_r,\bx_s$, suppressing
the $z$ coordinates of the sources and receiver, idealized to
constants. [So the following derivation would have to be modified to
model so-called broadband acquisition with variable streamer depth.] 
Since offset is now one-dimensional (and horizontal), we
will write $(x,z,h)$ instead of $(\bx,\bh)$, and $(x\pm h,z)$ instead
of $\bx\pm \bh$.

For 2D, the $t^{-1/2}_+$ singularity of the Green's function removes
one of the time derivatives appearing in \ref{eqn:intrep5}, due to the
identity
\begin{equation}
\label{eqn:convsq}
t_+^{-1/2}*t_+^{-1/2} = \pi H(t)
\end{equation}
(see \cite{GelShil:58}, p. 116, formula (3')), so
\[
I_t^2\bar{F}[\kappa,\rho][\bar{r}_{\kappa},\bar{r}_{\rho}]
%\bar{{\cal F}}[\kappa,\rho][\delta 
%\bar{\kappa},\delta\bar{\rho}](x_r,t;x_s) 
\]
\[
= \pi\int dx \int dz \int dh a(x_r,x+h,z) a(x_s,x-h,z) 
\delta(t-(T(x_r,x+h,z)+T(x_s,x-h,z))) 
\]
\[
\times \left(\frac{1}{\sqrt{\kappa(x+h,z)}} \bar{r}_{\kappa}(x,z,h) \frac{1}{\sqrt{\kappa(x-h,z)}}
\right. 
\]
\begin{equation}
\label{eqn:intrep2d}
+\left. \left. 
   \nabla_{x,z}  T(x_r,x+h,z) \cdot \nabla_{x,z} T(x_s,x-h,z) 
\frac{1}{\sqrt{\rho(x+h,z)}}  \bar{r}_{\rho}(x,z,h) \frac{1}{\sqrt{\rho(x-h,z)}}
\right)\right\}
\end{equation}

Note the following relation btween
partial derivatives of $T$ and the ray takeoff angles {\em at the
  source and receiver surfaces}
$\theta_r,\theta_s$ of the rays from $(x_r,z_r)$ respectively
$(x_s,z_s)$ to the scattering points $(x+h,z)$ respectively $(x-h,z)$:
\begin{eqnarray}
\frac{\partial T}{\partial z_r}(x_r,x+h,z)  &=& -\frac{\cos 
  \theta_r}{v(x_r,z_r)} \nonumber \\
\label{eqn:dtdxr}
\frac{\partial T}{\partial z_s}(x_s,x-h,z)  &=& -\frac{\cos 
  \theta_s}{v(x_s,z_s)}\\
\end{eqnarray}
Thus
\[
I_t^2\frac{\partial^2}{\partial z_s \partial z_r}
I_t^2\bar{F}[\kappa,\rho][\bar{r}_{\kappa},\bar{r}_{\rho}]
\]
\[
=\pi\int dx \int dz \int dh \frac{\cos \theta_r(x_r,x+h,z)}{v(x_r,z_r)}
\frac{\cos \theta_s(x_s,x-h,z)}{v(x_s,z_s)}
\]
\[
\times a(x_r,x+h,z) a(x_s,x-h,z) 
\delta(t-(T(x_r,x+h,z)+T(x_s,x-h,z)))  
\]
\[
\times \left(\frac{1}{\sqrt{\kappa(x+h,z)}} \bar{r}_{\kappa}(x,z,h) \frac{1}{\sqrt{\kappa(x-h,z)}}
\right. 
\]
\begin{equation}
\label{eqn:intrep2dsr}
+\left. \left. 
   \nabla_{x,z}  T(x_r,x+h,z) \cdot \nabla_{x,z} T(x_s,x-h,z) 
\frac{1}{\sqrt{\rho(x+h,z)}}  \bar{r}_{\rho}(x,z,h) \frac{1}{\sqrt{\rho(x-h,z)}}
\right)\right\}
\end{equation}
The operator matrix 
\begin{equation}
\label{eqn:dwnop}
{\bf M} = (I_t^2\bar{F}[\kappa,\rho])^T
I_t^2\frac{\partial^2}{\partial z_s \partial z_r}I_t^2\bar{F}[\kappa,\rho]
\end{equation}
thus has components of the form given by
\cite{HouSymes:17}, equation 4,
% which we repeat 
%here for the readers convenience:
\begin{equation}
\begin{aligned}
\label{eqn:grt2d}
M u(x,z,h) =& \int dx_r dx_s dx' dz' dh' \tilde{A}(x_r,x_s,x,z,h,x',z',h')\\
&\times \delta(\phi(x_r,x_s,x,z,h) - \phi(x_r,x_s,x',z',h')) u(x',z',h'),
\end{aligned}
\end{equation}
in which $\phi(x_r,x_s,x,z,h) = T(x_r,x+h,z) + T(x_s,x-h,z)$. The
coefficients $\tilde{A} = \tilde{A}_{i,j}, i,j \in \{\kappa, \rho\}$ are
\[
\tilde{A}_0(x_r,x_s,x,z,h,x',z',h')=
\]
\[
\pi^2 a(x_r,x+h,z) 
a(x_s,x-h,z) \frac{\cos \theta_r(x_r,x'+h',z')}{v(x_r,z_r)}
\]
\begin{equation}
\label{eqn:grt2dcoeff1}
\times \, \frac{\cos \theta_s(x_s,x'-h',z')}{v(x_s,z_s)} a(x_r,x'+h',z')a(x_s,x'-h',z')
\end{equation}
\begin{equation}
\label{eqn:grt2dcoeff2}
\tilde{A}_{\kappa,\kappa}(x_r,x_s,x,z,h,x',z',h') = 
\frac{\tilde{A}_0(x_r,x_s,x,z,h,x',z',h')}{\sqrt{\kappa(x+h,z)}\sqrt{\kappa(x-h,z)}\sqrt{\kappa(x'+h',z')}\sqrt{\kappa(x'-h',z')}}
\end{equation}
\[
\tilde{A}_{\rho,\kappa}(x_r,x_s,x,z,h,x',z',h') =
\]
\begin{equation}
\label{eqn:grt2dcoeff3}
\frac{\tilde{A}_0(x_r,x_s,x,z,h,x',z',h')\nabla T(x_r,x+h,z)
\cdot\nabla T(x_s,x-h,z)}
{\sqrt{\rho(x+h,z)}\sqrt{\rho(x-h,z)}\sqrt{\kappa(x'+h',z')}\sqrt{\kappa(x'-h',z')}}
\end{equation}
\begin{equation}
\label{eqn:grt2dcoeff4}
\tilde{A}_{\kappa,\rho}(x_r,x_s,x,z,h,x',z',h') =
\tilde{A}_{\rho,\kappa}(x_r,x_s,x',z',h',x,z,h)
\end{equation}
\[
\tilde{A}_{\rho,\rho}(x_r,x_s,x,z,h,x',z',h') =\tilde{A}_0(x_r,x_s,x,z,h,x',z',h') \times
\]
\begin{equation}
\label{eqn:grt2dcoeff5}
\frac{ \nabla_{x,z}  T(x_r,x+h,z)\cdot\nabla_{x,z}  T(x_s,x-h,z)
\nabla_{x',z'} T(x_r,x'+h',z')\cdot \nabla_{x',z'}T(x_s,x'-h',z')}
{\sqrt{\rho(x+h,z)}\sqrt{\rho(x-h,z)}\sqrt{\rho(x'+h',z')}\sqrt{\rho(x'-h',z')}}
\end{equation}

Asymptotic evaluation of the operator matrix $\bf{M}$ leads to one of
our two major results. The other follows by asymptotic evaluation of a
related integral. Note that

\[
\frac{\partial}{\partial t}I_t^2\frac{\partial^2}{\partial z_s \partial z_r}
I_t^2\bar{F}[\kappa,\rho][\bar{r}_{\kappa},\bar{r}_{\rho}]
\]
\[
=\pi\int dx \int dz \int dh \frac{\cos \theta_r(x_r,x+h,z)}{v(x_r,z_r)}
\frac{\cos \theta_s(x_s,x-h,z)}{v(x_s,z_s)}
\]
\[
\times a(x_r,x+h,z) a(x_s,x-h,z) 
\delta'(t-(T(x_r,x+h,z)+T(x_s,x-h,z))) 
\]
\[
\times \left(\frac{1}{\sqrt{\kappa(x+h,z)}} \bar{r}_{\kappa}(x,z,h) \frac{1}{\sqrt{\kappa(x-h,z)}}
\right. 
\]
\[
+\left. \left. 
   \nabla_{x,z}  T(x_r,x+h,z) \cdot \nabla_{x,z} T(x_s,x-h,z) 
\frac{1}{\sqrt{\rho(x+h,z)}}  \bar{r}_{\rho}(x,z,h) \frac{1}{\sqrt{\rho(x-h,z)}}
\right)\right\}
\]
\[
=\pi\int dx \int dz \int dh \frac{\cos \theta_r(x_r,x+h,z)}{v(x_r,z_r)}
\frac{\cos \theta_s(x_s,x-h,z)}{v(x_s,z_s)}
\]
\[
\times a(x_r,x+h,z) a(x_s,x-h,z) 
\frac{\partial}{\partial z}\delta(t-(T(x_r,x+h,z)+T(x_s,x-h,z))) 
\]
\[
  \times \left[-\left(\frac{\partial T}{\partial 
  z}(x_r,x+h,z) + \frac{\partial T}{\partial z}(x_s,x-h,z)\right)^{-1}\right] 
\]
\[
\times \left(\frac{1}{\sqrt{\kappa(x+h,z)}} \bar{r}_{\kappa}(x,z,h) \frac{1}{\sqrt{\kappa(x-h,z)}}
\right. 
\]
\[
+\left. \left. 
   \nabla_{x,z}  T(x_r,x+h,z) \cdot \nabla_{x,z} T(x_s,x-h,z) 
\frac{1}{\sqrt{\rho(x+h,z)}}  \bar{r}_{\rho}(x,z,h) \frac{1}{\sqrt{\rho(x-h,z)}}
\right)\right\}
\]
\[
\approx \pi\int dx \int dz \int dh \frac{\cos \theta_r(x_r,x+h,z)}{v(x_r,z_r)}
\frac{\cos \theta_s(x_s,x-h,z)}{v(x_s,z_s)}
\]
\[
\times a(x_r,x+h,z) a(x_s,x-h,z) 
\delta(t-(T(x_r,x+h,z)+T(x_s,x-h,z))) 
\]
\[
  \times \left(\frac{\partial T}{\partial 
  z}(x_r,x+h,z) + \frac{\partial T}{\partial z}(x_s,x-h,z)\right)^{-1}
\]
\[
\times \left(\frac{1}{\sqrt{\kappa(x+h,z)}} \frac{\partial}{\partial z}\bar{r}_{\kappa}(x,z,h) \frac{1}{\sqrt{\kappa(x-h,z)}}
\right. 
\]
\begin{equation}
\label{eqn:intrep2dsrbis}
+\left. \left. 
   \nabla_{x,z}  T(x_r,x+h,z) \cdot \nabla_{x,z} T(x_s,x-h,z) 
\frac{1}{\sqrt{\rho(x+h,z)}}  \frac{\partial}{\partial z} \bar{r}_{\rho}(x,z,h) \frac{1}{\sqrt{\rho(x-h,z)}}
\right)\right\}
\end{equation}
after integration by parts in $z$. The approximation in the last step
consists of leaving out all terms except those involving the $z$
derivative of $\bar{r}_{\kappa}, \bar{r}_{\rho}$, as is consistent
with the asymptotic evaluation to come: the reflectivities will be
presumed oscillatory, while all other quantities appearing in the
integral \ref{eqn:intrep2dsrbis} are smooth. In the limit of infinite
spatial frequency components dominating $\bar{r}_{\kappa},
\bar{r}_{\rho}$, that is, for singular reflectivities, the terms
retained on the RHS of \ref{eqn:intrep2dsrbis} dominate.

Thus the operator matrix 
\[
{\bf M_1} = (I_t^2\bar{F}[\kappa,\rho])^{*}
\frac{\partial}{\partial t}I_t^2\frac{\partial^2}{\partial z_s \partial z_r}I_t^2\bar{F}[\kappa,\rho]
\]
\begin{equation}
\label{eqn:dwnopbis}
 = (I_t^2\bar{F}[\kappa,\rho])^{*}
I_t\frac{\partial^2}{\partial z_s \partial z_r}I_t^2\bar{F}[\kappa,\rho]
\end{equation}
also has components of the integral form \ref{eqn:grt2d}  
(\cite{HouSymes:17}, equation 4), in fact precisely the same as in
\ref{eqn:grt2dcoeff1}-\ref{eqn:grt2dcoeff5}, except that in the integrand,
$\partial \bar{r}_{\kappa}/\partial z$ and $\partial
\bar{r}_{\rho}/\partial z$ replace $\bar{r}_{\kappa},\bar{r}_{\rho}$ on the RHS and there is an extra factor of $(\partial
\phi/\partial z)^{-1}$.

\cite{HouSymes:17} develop an asymptotic evaluation
of the operator $M$ in definition \ref{eqn:grt2d}. 
This result involves the
rate of change of the takeoff angles {\em at the scattering points}
$\alpha_r,\alpha_s$ with respect to the source and receiver
coordinates, and also the mapping from the extended phase space point
$(x,z,h,k_x,k_z,k_h)$ to the source and receiver coordinates
$x_r,x_s$. \cite{tenKroode:12} intoduced this map and explained its
central importance: it defines the scattering (``canonical'') relation of
$D{\cal F}$. \cite{HouSymes:15} present an alternate description,
which we will follow here: the extended phase space
point determines the ray pair, by determining the traveltime gradients
which are also the ray takeoff vectors at the extended scattering
point $(x,z,h)$. The relation is homogeneous in the wavevector
$(k_x,k_h,k_z)$, so can be expressed in terms of projective
coordinates $(k_x/k_z,k_h/k_z)$ under the assumption (implicit in the
use of the horizontal offset extension) that $k_z > 0$. Accordingly we
will write $x_{r,s} = x_{r,s}(x,z,h,k_x/k_z,k_h/k_z)$ for the
emergence points of the receiver, respectively source, rays. In terms
of these quantities, the asymptotic evaluation of $M$ is given by
equation A-4 in \cite[]{HouSymes:17}:
\[
Mu(x,z,h) \approx \int 
dk_{x}dk_{z}dk_{h}\frac{e^{i(xk_{x}+zk_{z}+hk_{h})}\widehat{u}(k_{x},k_{z},k_{h})}{2\pi 
  k^2_z}
\]
\[
\times \, A(x_r(x,z,h,k_x/k_z,k_h/k_z),x_s(x,z,h,k_x/k_z,k_h/k_z),x,z,h) 
\]
\[
\times \left(\frac{\partial T}{\partial 
    z}(x_r(x,z,h,k_x/k_z,k_h/k_z),x+h,z) +
\frac{\partial T}{\partial 
    z}(x_s(x,z,h,k_x/k_z,k_h/k_z),x-h,z) \right)^{-1}
\]
\begin{equation}
\label{eqn:grt2deval}
\times |\det \mbox{
  Hess}(x_r(x,z,h,k_x/k_z,k_h/k_z),x_s(x,z,h,k_x/k_z,k_h/k_z),x,z,h)|^{-1/2} . 
\end{equation}
The coefficient $A$ is the diagonal value of the coefficient
$\tilde{A}$ in equation \ref{eqn:grt2d}, for $x'=x,z'=z,h'=h$:
\begin{equation}
\label{eqn:diag}
A(x_r,x_s,x,z,h) = \tilde{A}(x_r,x_s,x,z,h,x,z,h).
\end{equation}
The restriction $x'=x,z'=z,h'=h$ is one of the stationary phase
conditions that results from 
asymptotic evaluation of the integral in \ref{eqn:grt2d}. This
calculation also leads to the
conclusion that the vectors $(k_x,k_z)$ and $(k_h,k_z)$ are parallel
to $\nabla_{x,z}T(x_r,x+h,z)+\nabla_{x,z}T(x_s,x-h,z)$ and
$\nabla_{h,z}T(x_r,x+h,z)+\nabla_{h,z}T(x_s,x-h,z)$ respectively. This
condition in turn leads to the determination of
$\nabla_{x,z}T(x_r,x+h,z)$ and $\nabla_{x,z}T(x_s,x-h,z)$
individually, hence the rays connecting $(x\pm h,z)$ with $(x_r,z_s)$
and $(x_s,z_s)$ respectively, from the phase space vector
$(x,z,h,k_x,k_z,k_h)$, as
explained by \cite{HouSymes:15}. The determination of the rays in turn
determines $x_r,x_s$ as functions of  $(x,z,h,k_x,k_z,k_h)$, as
mentioned above.

The Hessian factor $|\det \mbox{ Hess}|^{-1/2}$ is most
conveniently expressed in terms of the ray slownesses, velocity $v
= \sqrt{\kappa/\rho}$, and slowness $s=1/v$, for which we
introduce abbreviations, regarding the arguments as understood and
dropping them from the notation:
\begin{eqnarray}
p_r & = & \frac{\partial T}{\partial x}(x_r,x+h,z) \nonumber \\  
p_s & = & \frac{\partial T}{\partial x}(x_s,x-h,z) \nonumber \\  
q_r & = & \frac{\partial T}{\partial z}(x_r,x+h,z) \nonumber \\
q_s & = & \frac{\partial T}{\partial z}(x_s,x-h,z) \nonumber \\
v_{\pm} & = & v(x\pm h,z) \nonumber \\
\label{eqn:rayslow}   
s_{\pm} & = & 1/v_{\pm}
\end{eqnarray}
Then
\begin{equation}
\label{eqn:hess}
|\det \mbox{ Hess}|^{-1/2} =\frac{1}{2} (q_r+q_s)^4\left(\frac{\partial 
    \alpha_r}{\partial x_r}\frac{\partial \alpha_s}{\partial 
    x_s}\right)^{-1}[s_-^2q_r^2 + s_+^2q_s^2 + (s_+^2+s_-^2)q_rq_s]^{-1}
\end{equation}
%Note that the right-hand sides in the system \ref{eqn:rayslow} are all
%functions of $(x,z,h,k_x,k_z,k_h)$, because of the stationary phase
%conditions, as already explained. Thus $Q$ is also a function of
%$(x,z,h,k_x,k_z,k_h)$. Since $Q$ is homogeneous of degree zero in
%$(k_x,k_z,k_h)$, it can be expressed as a function of $(k_x/k_z,
%k_h,k_z)$, as indicated in equation \ref{eqn:grt2deval}. 
(\cite{HouSymes:15}, equation A-19).

Because of the stationarity conditions $x'=x, z'=z, h'=h$, the
coefficients appearing in the asymptotic evalution of the integral
defining the matrix operator $\bf M$ simplify. It is convenient to
introduce the generalized scattering angle $\theta =
\theta(x,z,h,k_x/k_z,k_h/k_z)$, usually defined as half the angle
between source and receiver rays at the scattering point:
\[
\nabla_{x,z}T(x_r(x,z,h,k_x/k_z,k_h/k_z),x+h,z)\cdot 
\nabla_{x,z}T(x_s(x,z,h,k_x/k_z,k_h/k_z),x-h,z) 
\]
\begin{equation}
\label{eqn:thetadef}
= \frac{\cos 2\theta(x,z,h,k_x/k_z,k_h/k_z)}{v(x+h,z) v(x-h,z)} =
\frac{(1-2\sin^2\theta(x,z,h,k_x/k_z,k_h/k_z))}{v(x+h,z) v(x-h,z)}
\end{equation}
Then the 
coefficients $A = A_{i,j}, i,j \in \{\kappa, \rho\}$ appearing in the
asymptotic evaluation \ref{eqn:grt2deval} of $\bf M$ are 
\[
A_0(x_r,x_s,x,z,h)=
\]
\[
\pi^2 a(x_r,x+h,z)^2  
a(x_s,x-h,z)^2 
\]
\begin{equation}
\label{eqn:grt2dcoeff1red}
\times \frac{\cos \theta_r(x_r,x+h,z)}{v(x_r,z_r)} \frac{\cos\theta_s(x_s,x-h,z)}{v(x_s,z_s)} 
\end{equation}
\begin{equation}
\label{eqn:grt2dcoeff2red}
A_{\kappa,\kappa}(x_r,x_s,x,z,h) = 
\frac{A_0(x_r,x_s,x,z,h)}{\kappa(x+h,z)\kappa(x-h,z)}
\end{equation}
\[
A_{\rho,\kappa}(x_r,x_s,x,z,h) 
\]
\[
=\frac{A_0(x_r,x_s,x,z,h) \cos 2 \theta(x,z,h,k_x/k_z,k_h/k_z)}
{v(x+h,z)
  v(x-h,z)\sqrt{\rho(x+h,z)}\sqrt{\rho(x-h,z)}\sqrt{\kappa(x+h,z)}\sqrt{\kappa(x-h,z)}}
\]
\begin{equation}
\label{eqn:grt2dcoeff3red}
=\frac{A_0(x_r,x_s,x,z,h) \cos 2 \theta(x,z,h,k_x/k_z,k_h/k_z)}
{\kappa(x+h,z)\kappa(x-h,z)}
\end{equation}
\begin{equation}
\label{eqn:grt2dcoeff4red}
A_{\kappa,\rho}(x_r,x_s,x,z,h) =
A_{\rho,\kappa}(x_r,x_s,x,z,h)
\end{equation}
\[
A_{\rho,\rho}(x_r,x_s,x,z,h) =
\]
\[
=\frac{A_0(x_r,x_s,x,z,h) \cos^2 2 \theta(x,z,h,k_x/k_z,k_h/k_z)}
{v(x+h,z)^2v(x-h,z)^2\rho(x+h,z)\rho(x-h,z)}
\]
\begin{equation}
\label{eqn:grt2dcoeff5red}
=\frac{A_0(x_r,x_s,x,z,h) \cos^2 2 \theta(x,z,h,k_x/k_z,k_h/k_z)}
{\kappa(x+h,z)\kappa(x-h,z)}
\end{equation}
whence
\[
{\bf M} 
\left(\begin{array}{c}
\bar{r}_{\kappa} \\
\bar{r}_{\rho}
\end{array}
\right) = 
\]
\[
= \int dk_x dk_z dk_h \frac{A_0 e^{i(xk_{x}+zk_{z}+hk_{h})}}{2\pi 
  k^2_z(q_r + q_s)\kappa_+\kappa_-} |\mbox{det Hess}|^{-1/2} 
\]
\begin{equation}
\label{eqn:rk2}
\times 
\left(
\begin{array}{c}
\hat{\bar{r}}_{\kappa} + \hat{\bar{r}}_{\rho} \cos 2 \theta \\ 
(...) 
\end{array}
\right)
\end{equation}
Following the remark after equation \ref{eqn:dwnopbis}, $\bf M_1$ has
a similar expression:
\[
{\bf M_1} 
\left(\begin{array}{c}
\bar{r}_{\kappa} \\
\bar{r}_{\rho}
\end{array}
\right) = 
\]
\[
= \int dk_x dk_z dk_h \frac{A_0 e^{i(xk_{x}+zk_{z}+hk_{h})}}{2\pi 
  (-ik_z)(q_r + q_s)^2\kappa_+\kappa_-} |\mbox{det Hess}|^{-1/2} 
\]
\begin{equation}
\label{eqn:rk2bis}
\times 
\left(
\begin{array}{c}
\hat{\bar{r}}_{\kappa} + \hat{\bar{r}}_{\rho} \cos 2 \theta \\ 
(...) 
\end{array}
\right)
\end{equation}

%Define
%\begin{equation}
%\label{eqn:rsig}
%\bar{r}_{\sigma} \equiv \frac{1}{2}(\bar{r}_{\kappa} + \bar{r}_{\rho})
%\end{equation}
%Note that for $h=0$, or alternatively for physical perturbations
%$\delta \bar{\kappa}(x,z,h) = \delta \kappa(x,z)\delta(h),
%\delta \bar{\rho}(x,z,h) = \delta \rho(x,z)\delta(h)$, 
%\[
%\bar{r}_{\kappa} = \frac{\delta \kappa}{\kappa}\delta(h),
%\bar{r}_{\rho} = \frac{\delta \rho}{\rho}\delta(h),
%\]
%whence
%\[
%\bar{r}_{\sigma} = \frac{\delta \sigma}{\sigma} \delta(h), \,\sigma =
%\rho v = \sqrt{\kappa \rho}.
%\]
%That is, in the physical case, $\bar{r}_{\sigma}$ is the relative
%perturbation in acoustic impedance. In general, $\bar{r}_{\sigma}$
%does not have a simple physical meaning.

%In terms of $\bar{r}_{\sigma}$, the $\kappa$ component in equation
%\ref{eqn:rk2} may be expressed as
%\[
%{\bf M} 
%\left(\begin{array}{c}
%2\bar{r}_{\sigma}-\bar{r}_{\rho} \\
%\bar{r}_{\rho}
%\end{array}
%\right) = 
%\]
%\[
%= \int dk_x dk_z dk_h \frac{A_0 e^{i(xk_{x}+zk_{z}+hk_{h})}}{2\pi 
%  k^2_z(q_r + q_s)\kappa_+\kappa_-} |\mbox{det Hess}|^{-1/2} 
%\]
%\begin{equation}
%\label{eqn:rk2sigma}
%\times 
%\left(
%\begin{array}{c}
%2(\hat{\bar{r}}_{\sigma} - \hat{\bar{r}}_{\rho} \sin^2 \theta) \\ 
%(...) 
%\end{array}
%\right)
%\end{equation}

Equations \ref{eqn:hess} and
\ref{eqn:grt2dcoeff1red} combine to show that the common scalar factor in
the integral \ref{eqn:rk2} takes the form
\[
\frac{A_0 e^{i(xk_{x}+zk_{z}+hk_{h})}}{2\pi 
  k^2_z(q_r + q_s)\kappa_+\kappa_-} |\mbox{det Hess}|^{-1/2} 
\]
\[
=\pi^2 a(x_r,x+h,z)^2  a(x_s,x-h,z)^2 
 \frac{\cos \theta_r(x_r,x+h,z)}{v(x_r,z_r)}
\frac{\cos\theta_s(x_s,x-h,z)}{v(x_s,z_s)} 
\]
\[
\times \frac{1}{2} (q_r+q_s)^4\left(\frac{\partial 
    \alpha_r}{\partial x_r}\frac{\partial \alpha_s}{\partial 
    x_s}\right)^{-1}[s_-^2q_r^2 + s_+^2q_s^2 +
(s_+^2+s_-^2)q_rq_s]^{-1}
\]
\begin{equation}
\label{eqn:rk3}
\times \frac{e^{i(xk_{x}+zk_{z}+hk_{h})}}{2\pi 
  k^2_z(q_r + q_s)\kappa_+\kappa_-} 
\end{equation}
Similarly,  the common scalar factor in
the integral \ref{eqn:rk2bis} takes the form
\[
\frac{A_0 e^{i(xk_{x}+zk_{z}+hk_{h})}}{2\pi 
  (-ik_z)(q_r + q_s)^2\kappa_+\kappa_-} |\mbox{det Hess}|^{-1/2} 
\]
\[
=\pi^2 a(x_r,x+h,z)^2  a(x_s,x-h,z)^2 
 \frac{\cos \theta_r(x_r,x+h,z)}{v(x_r,z_r)}
\frac{\cos\theta_s(x_s,x-h,z)}{v(x_s,z_s)} 
\]
\[
\times \frac{1}{2} (q_r+q_s)^4\left(\frac{\partial 
    \alpha_r}{\partial x_r}\frac{\partial \alpha_s}{\partial 
    x_s}\right)^{-1}[s_-^2q_r^2 + s_+^2q_s^2 +
(s_+^2+s_-^2)q_rq_s]^{-1}
\]
\begin{equation}
\label{eqn:rk3bis}
\times \frac{e^{i(xk_{x}+zk_{z}+hk_{h})}}{2\pi 
  (-ik_z)(q_r + q_s)^2\kappa_+\kappa_-} 
\end{equation}

Invoke the crucial observation of Norm Bleistein, Yu
Zhang, and others, that the geometric amplitude $a(...)$ may be
expressed in terms of the rate of change of arrival time and takeoff
angles of rays reaching the source and receiver surfaces with respect
to the source or receiver coordinate
\cite[]{ZhangBleistein:03,ZhangBleistein:05,ZhangSunGray:07,ZhangYuSun:08}.
Specifically,
\[
\frac{a(x_r,x+h,z)^2}{\rho(x+h,z)}\frac{\cos
  \theta_r(x_r,x+h,z)}{v_r(x_r,z_r)}\left(\frac{\partial
    \alpha_r}{\partial x_r}(x_r,x+h,z)\right)^{-1} 
\]
\[
 = \frac{a(x_s,x-h,z)^2}{\rho(x-h,z)}\frac{\cos
   \theta_s(x_s,x-h,z)}{v_s(x_s,z_s)}\left(\frac{\partial
     \alpha_s}{\partial x_s}(x_s,x-h,z)\right)^{-1}
\]
\begin{equation}
\label{eqn:raytube}
= \frac{1}{8\pi^2}. 
\end{equation}
(\cite[]{HouSymes:15}, equations B-6, B-7 - note that for variable
density acoustics, the transport equation reads $\nabla \cdot
(a^2/\rho) \nabla T=0$). Thus writing $\rho(x\pm h,z)=\rho_{\pm}$, the
right-hand side of equation \ref{eqn:rk3} becomes
\[
\frac{A_0 e^{i(xk_{x}+zk_{z}+hk_{h})}\rho_+\rho_-}{2\pi 
  k^2_z(q_r + q_s)\kappa_+\kappa_-} |\mbox{det Hess}|^{-1/2} 
\]
\[
=\pi^2 \left(\frac{1}{8\pi^2}\right)^2 
 \frac{1}{2} (q_r+q_s)^4[s_-^2q_r^2 + s_+^2q_s^2 +
(s_+^2+s_-^2)q_rq_s]^{-1}
\]
\begin{equation}
\label{eqn:rk4}
\times \frac{e^{i(xk_{x}+zk_{z}+hk_{h})}}{2\pi 
  k^2_zv_+^2v_-^2(q_s+q_r)} 
\end{equation}
and the right-hand side of equation \ref{eqn:rk3bis} becomes
\[
\frac{A_0 e^{i(xk_{x}+zk_{z}+hk_{h})}\rho_+\rho_-}{2\pi 
  (-ik_z)(q_r + q_s)^2\kappa_+\kappa_-} |\mbox{det Hess}|^{-1/2} 
\]
\[
=\pi^2 \left(\frac{1}{8\pi^2}\right)^2 
 \frac{1}{2} (q_r+q_s)^4[s_-^2q_r^2 + s_+^2q_s^2 +
(s_+^2+s_-^2)q_rq_s]^{-1}
\]
\begin{equation}
\label{eqn:rk4bis}
\times \frac{e^{i(xk_{x}+zk_{z}+hk_{h})}}{2\pi 
  (-ik_z)v_+^2v_-^2(q_s+q_r)^2} 
\end{equation}

Combining equations \ref{eqn:rk4}, respectively \ref{eqn:rk4bis}, and \ref{eqn:grt2dcoeff1red}-\ref{eqn:grt2dcoeff5red}, obtain
\[
{\bf M} 
\left(\begin{array}{c}
\bar{r}_{\kappa} \\
\bar{r}_{\rho}
\end{array}
\right) = 
\]
\[
= \int dk_x dk_z dk_h \frac{e^{i(xk_{x}+zk_{z}+hk_{h})}(q_r+q_s)^3}{256\pi^3 
  k^2_z(s_-^2q_r^2 + s_+^2q_s^2 +
(s_+^2+s_-^2)q_rq_s)v^2_+v^2_-} 
\]
\begin{equation}
\label{eqn:rk3sigma}
\times 
\left(
\begin{array}{c}
\hat{\bar{r}}_{\kappa} + \hat{\bar{r}}_{\rho} \cos 2 \theta) \\ 
(...) 
\end{array}
\right)
\end{equation}
and
\[
{\bf M_1} 
\left(\begin{array}{c}
\bar{r}_{\kappa} \\
\bar{r}_{\rho}
\end{array}
\right) = 
\]
\[
= \int dk_x dk_z dk_h \frac{e^{i(xk_{x}+zk_{z}+hk_{h})}(q_r+q_s)^2}{256\pi^3 
  (-ik_z)(s_-^2q_r^2 + s_+^2q_s^2 +
(s_+^2+s_-^2)q_rq_s)v^2_+v^2_-} 
\]
\begin{equation}
\label{eqn:rk3sigmabis}
\times 
\left(
\begin{array}{c}
\hat{\bar{r}}_{\kappa} + \hat{\bar{r}}_{\rho} \cos 2 \theta) \\ 
(...) 
\end{array}
\right)
\end{equation}

The computations done so far are formal in nature, that is,
restrictions necessary to ensure that rays with appropriate slowness
vectors actually exist have not been imposed, nor have the
reflectivities $\bar{r}$ or the data $d$ been constrained so that the
expressions \ref{eqn:rk3sigma},  \ref{eqn:rk3sigmabis} and related equations are correct to
leading order in frequency. To make such constraints manifest and
recover an asymptotically correct result from \ref{eqn:rk3sigma}, \ref{eqn:rk3sigmabis}
requires a closer study of the ray geometry of extended acoustics.

\section{The Canonical Relation}
To make sense of the integrand in equation \ref{eqn:rk3sigma}, it is
necessary to
understand how to compute the generalized scattering angle $\theta$ in
terms of the extended phase space coordinates $(x,z,h,k_x/k_z,
k_h/k_z)$. This in turn requires computation of the ray slowness
vectors $(p_r,q_r)$ and $(p_s,q_s)$ in terms of the same
quantities.

The critical hypothesis for horizontal subsurface offset extended
scattering is that rays carrying significant energy do not turn
horizontal, that is, $q_r, q_s > 0$ at all points on such
rays. \cite{deHoopStolk:05,deHoopStolk:06} call this assumption the
``DSR'' (double square root) hypothesis, as it is also the assumption
that permits to the wave equation to be replaced by an asymptotically
equivalent evolution equation in depth known as the double square root equation.

The stationary phase conditions include 
\begin{eqnarray}
p_r & = & \frac{1}{2}\frac{k_x+k_h}{k_z}(q_r+q_s) \nonumber \\
\label{eqn:stat}
p_s & = & \frac{1}{2}\frac{k_x-k_h}{k_z}(q_r+q_s) 
\end{eqnarray}
(\cite[]{HouSymes:15}, equation A-20). Introduce projective
coordinates $p_x=k_x/k_z,
p_h=k_h/k_z$. Then an equivalent system is
\begin{eqnarray}
p_r & = & \frac{1}{2}(p_x+p_h)(q_r+q_s) \nonumber \\
\label{eqn:stat1}
p_s & = & \frac{1}{2}(p_x-p_h)(q_r+q_s) 
\end{eqnarray}
Rearranging,
\begin{eqnarray}
p_x & = &\frac{p_r+p_s}{q_r+q_s} \nonumber \\
\label{eqn:stat2}
p_h & = &\frac{p_r-p_s}{q_r+q_s} 
\end{eqnarray}
Since $(p_r,q_r)$ and $(p_s,q_s)$ traverse the product of hemispheres
$q_r,q_s > 0$ of
radius $s_+$ and $s_-$ respectively, $(p_x,p_h)$ must lie in the image
of this product of hemispheres under the map defined in equation
\ref{eqn:stat2}. This set is not so easy to describe explicitly in
general, however for the case $h=0$ (or, more precisely, $s_+=s_-=s$),
the description is simple: in that case, 
\[
p_xp_h = \frac{p_r^2-p_s^2}{(q_r+q_s)^2} = - \frac{q_r-q_s}{q_r+q_s}
\]
\begin{equation}
\label{eqn:stat3}
=-\frac{\frac{q_r}{q_s}-1}{\frac{q_r}{q_s}+1}
\end{equation}
The map $q_r/q_s \mapsto p_xp_h$ so defined is a bijection from
$\bR^+$ to $(-1,1)$. That is, for $h=0$ necessarily
\begin{equation}
\label{eqn:stat4}
|p_xp_h| < 1.
\end{equation}
See \cite[]{tenKroode:12}, definition of the set $\Omega_X$ for a related
construction. As in that reference, we shall see that for $(p_x,p_h)$
satisfying equation \ref{eqn:stat4}, there is a unique DSR ray pair
with slowness vectors $(p_r,q_r)$ and $(p_s,q_s)$ satisfying equation 
\ref{eqn:stat1}.

To begin the construction of the inverse map from
$p_x=k_x/k_z,p_h=k_h/k_z$ to $(q_s,q_r)$ and hence the rays connecting
scattering points with sources and receivers, note that the eikonal
equation together with equation \ref{eqn:stat} implies that
\begin{eqnarray}
q_r & = & \left(s_+^2 -
          \frac{1}{4}\frac{(k_x+k_h)^2}{k_z^2}(q_s+q_r)^2\right)^{1/2} \nonumber \\
\label{eqn:eik}
q_s & = & \left(s_-^2 -
          \frac{1}{4}\frac{(k_x-k_h)^2}{k_z^2}(q_s+q_r)^2\right)^{1/2}
\end{eqnarray}
whence
\[
s_+s_-\cos 2 \theta = p_rp_s + q_r q_s 
= \frac{1}{4}\frac{k_x^2-k_h^2}{k_z^2}(q_r+q_s)^2
\]
\begin{equation}
\label{eqn:theta1}
+  \left(s_+^2 -
          \frac{1}{4}\frac{(k_x+k_h)^2}{k_z^2}(q_s+q_r)^2\right)^{1/2}
\left(s_-^2 -
          \frac{1}{4}\frac{(k_x-k_h)^2}{k_z^2}(q_s+q_r)^2\right)^{1/2}
\end{equation}

Thus the vertical ray slownesses $q_s$ and $q_r$ (equation
\ref{eqn:eik}), the phase Hessian factor in the evaluation $\bf M$
(equation \ref{eqn:hess}), and the scattering angle (equation
\ref{eqn:theta1}) may all be expressed in terms of the vertical
gradient of the two-way time at the stationary point, $q_r+q_s$. This
quantity is in turn a function of the generalized phase
coordinates: it follows from equations \ref{eqn:stat} and
\ref{eqn:eik} that its square solves a quadratic with coefficients
that are algebraic functions of $s_+,s_-$ and $k_x/k_z, k_h/k_z$
(\cite[]{tenKroode:12}, also \cite[]{HouSymes:15}, equations A-20
through A-28). Explicitly, 
\begin{equation}
\label{eqn:dphidz}
(q_r+q_s)^2 = \frac{-b+\sqrt{b^2-4ac}}{2a}
\end{equation}
where
\begin{eqnarray}
a & = & \frac{(k_x^2+k_z^2)(k_h^2+k_z^2)}{k_z^4}\nonumber \\
b & = & -2 \left[(s_+^2-s_-^2)\frac{k_xk_h}{k_z^2} +
        (s_+^2+s_-^2)\right] \nonumber \\
\label{eqn:abc}
c & = & (s_+^2-s_-^2)^2.
\end{eqnarray}

While explicit, the expression \ref{eqn:dphidz} is somewhat
opaque. Fortunately it simplifies considerably in the physical case,
that is, $h=0$: then $s_+=s_-$, $b=-4 s^2$, $c=0$, and 
\begin{equation}
\label{eqn:dphidzh0}
(q_r+q_s)^2 = 4\frac{k_z^4s^2}{(k_x^2+k_z^2)(k_h^2+k_z^2)}=\frac{4s^2}{(p_x^2+1)(p_h^2+1)}
\end{equation}
whence
\begin{eqnarray}
q_r & = & s\left(1 -
          \frac{(p_x+p_h)^2}{(p_x^2+1)(p_h^2+1)}\right)^{1/2} \nonumber \\
\label{eqn:eikh0}
q_s & = & s\left( 1 -
          \frac{(p_x-p_h)^2}{(p_x^2+1)(p_h^2+1)}\right)^{1/2}.
\end{eqnarray}
Since $s_+=s_-=s$ when $h=0$, all factors of $s$ in equation
\ref{eqn:theta1} are common and can be eliminated, leading to
\[
\cos 2 \theta
= \frac{(p_x^2-p_h^2)}{(p_x^2+1)(p_h^2+1)}
\]
\begin{equation}
\label{eqn:theta2}
+  \left(1 -
          \frac{(p_x+p_h)^2}{(p_x^2+1)(p_h^2+1)}\right)^{1/2}
\left(1 -
          \frac{(p_x-p_h)^2}{(p_x^2+1)(p_h^2+1)}\right)^{1/2}
\end{equation}
A bit of algebra, which we leave
to the reader, togerther with the condition that (equation
\ref{eqn:stat4}) $|p_xp_h|<1$ for $h=0$  yields
\begin{equation}
\label{eqn:theta3}
\cos 2 \theta = 
\frac{1-p_h^2}{1+p_h^2}, \, \sin \theta = \frac{p_h}{\sqrt{1+p_h^2}},
  \, \tan \theta = p_h.
\end{equation}
(Again, these identities hold only for $h=0$). It is easy to check
that $q_r,q_s$ given by \ref{eqn:eikh0} are positive, and finally set
$p_r,p_s$ by \ref{eqn:stat1}. That is, we have described a homogeneous
bijective map from $k_x,k_z,k_h$ to rays satisfying the DSR condition
in the special case of coincident scattering points $h=0$. The
explicit description of the domain of this map description for
$h \ne 0$ is less straightforward than condition \ref{eqn:stat4}.

Following \cite{tenKroode:12}, we denote by $\Omega_X$ the set of
points $(x,z,h,k_x,k_z,k_z)$ in phase space, satsifying the scattering
(stationary phase) conditions for a pair of downgoing rays. In view of
equation \ref{eqn:stat4}, 
\begin{equation}
\label{eqn:omegaxh0}
(x,z,0,k_x,k_z,k_z) \in \Omega_X \Leftrightarrow
\left|\frac{k_xk_h}{k_z^2}\right| < 1
\end{equation}
Let $B$ be
the symbol of a pseudodifferential operator essentially supported on this (conic)
set, a smoothing of the characteristic function of $\Omega_X$. When
convenient we will treat $B$ as if it were the characteristic function
of $\Omega_X$. Under the assumption that only downgoing rays carry
significant energy (technically, singularities), $D{\cal F}$ produces
asymptotically (up to a smooth error)
the same data from $(\delta \kappa,\delta \rho)$ and from 
$(Op(B)\delta \kappa, Op(B)\delta \rho)$. Thus $B$ may be regarded as
being part of the symbol of $\bf M$, without loss of generality.

\section{Near-focus Approximation}
The asymptotic evaluation in equation \ref{eqn:rk3sigma} implies that
\[
({\bf M}(\bar{r}_{\kappa},\bar{r}_{\rho})^T)_{\kappa} = \left((I_t^2\bar{F}[\kappa,\rho])^{*}
I_t^2\frac{\partial^2}{\partial z_s \partial z_r}I_t^2\,
\bar{F}[\kappa,\rho][\bar{r}_{\kappa},\bar{r}_{\rho}]\right)_{\kappa}
\]
\[
 \approx \int dk_x dk_z dk_h \frac{e^{i(xk_{x}+zk_{z}+hk_{h})}(q_r+q_s)^3}{256\pi^3 
  k^2_z(s_-^2q_r^2 + s_+^2q_s^2 +
(s_+^2+s_-^2)q_rq_s)v^2_+v^2_-} 
\]
\begin{equation}
\label{eqn:rk3mig}
\times 
(\hat{\bar{r}}_{\kappa} + \hat{\bar{r}}_{\rho} \cos 2 \theta)
\end{equation}
Similarly, \ref{eqn:rk3sigmabis} implies that
\[
({\bf M_1}(\bar{r}_{\kappa},\bar{r}_{\rho})^T)_{\kappa} = \left((I_t^2\bar{F}[\kappa,\rho])^{*}
I_t\frac{\partial^2}{\partial z_s \partial z_r}I_t^2\,
\bar{F}[\kappa,\rho][\bar{r}_{\kappa},\bar{r}_{\rho}]\right)_{\kappa}
\]
\[
 \approx \int dk_x dk_z dk_h \frac{e^{i(xk_{x}+zk_{z}+hk_{h})}(q_r+q_s)^2}{256\pi^3 
  (-ik_z)(s_-^2q_r^2 + s_+^2q_s^2 +
(s_+^2+s_-^2)q_rq_s)v^2_+v^2_-} 
\]
\begin{equation}
\label{eqn:rk3migbis}
\times 
(\hat{\bar{r}}_{\kappa} + \hat{\bar{r}}_{\rho} \cos 2 \theta)
\end{equation}
These relation
significantly simplify using the evaluation of slownesses $q_s,q_r$ for $h=0$,
explained in the last section. The error resulting from these
approximations amounts to inserting a 
pseudodifferential operator $P$ of order zero, whose symbol is $\equiv
1$ for $h=0$.

To begin with, \ref{eqn:dphidzh0} implies that
\[
\frac{(q_r+q_s)^3}{128\pi^3 
  k^2_z(s_-^2q_r^2 + s_+^2q_s^2 +
(s_+^2+s_-^2)q_rq_s)v^2_+v^2_-} 
\]
\[
\approx  
\frac{2sk_z^2}{(k_x^2+k_z^2)^{1/2}(k_h^2+k_z^2)^{1/2}}\frac{1}{256  
  \pi^3 s^2k_z^2v^4}
\] 
\begin{equation}
=\frac{1}{128\pi^3 v^3 \sqrt{k_x^2+k_z^2}\sqrt{k_h^2+k_z^2}}
\label{eqn:kksymbh0}
\end{equation} 
and 
\[
\frac{(q_r+q_s)^2}{256\pi^3 
  (-ik_z)(s_-^2q_r^2 + s_+^2q_s^2 +
(s_+^2+s_-^2)q_rq_s)v^2_+v^2_-} 
\]
\begin{equation}
\label{eqn:kksymbh0bis}
\approx 
\frac{1}{256 \pi^3 (-ik_z)v^2}
\end{equation}

``Approximately'' here means ``up to multiplication by a symbol that
is $\equiv 1$ for $h=0$''. So 
\[
({\bf M}(\bar{r}_{\kappa},\bar{r}_{\rho})^T)_{\kappa} 
\approx \int dk_x dk_z dk_h \frac{e^{i(xk_{x}+zk_{z}+hk_{h})} }{128 \pi^3 v^3 \sqrt{k_x^2+k_z^2}\sqrt{k_h^2+k_z^2}}PB 
\]
\begin{equation}
\label{eqn:rk4sigma}
\times 
(\hat{\bar{r}}_{\kappa} + \hat{\bar{r}}_{\rho} \cos 2 \theta ) 
\end{equation}
in which $P=P(x,z,h,k_x,k_z,k_h)$ is a symbol of order 0 satisfying
$P=1$ for $h=0$, and $B$ is the symbol mentioned at the end of the
last section. We claim that these symbols are actually of class
$S^0_{1,0}$, that is, classical symbols of order zero, and that the
RHS of equation \ref{eqn:rk4sigma} defines a classical
pseudodifferential operator of order zero. We defer the proof.

Similarly,
\[
({\bf M_1}(\bar{r}_{\kappa},\bar{r}_{\rho})^T)_{\kappa} 
= \int dk_x dk_z dk_h \frac{e^{i(xk_{x}+zk_{z}+hk_{h})} }{256 \pi^3 (-ik_z)v^2}P_1B 
\]
\begin{equation}
\label{eqn:rk4sigmabis}
\times 
(\hat{\bar{r}}_{\kappa} + \hat{\bar{r}}_{\rho} \cos 2 \theta ) 
\end{equation}
in which $P_1$ has the same properties as $P$, that is, it is also a
classical symbol of order 0.

\begin{theorem}
Define
\begin{equation}
\label{eqn:defL}
L=(-\nabla^2_{x,z})^{1/2}(-\nabla^2_{h,z})^{1/2},
\end{equation}
\begin{eqnarray}
\label{eqn:defWm}
W_m & = & \frac{1}{16}v^{-\frac{3}{2}}L^{-1}v^{-\frac{3}{2}},\\
\label{eqn:defWminv}
W_m^{-1} & = & 16 v^{\frac{3}{2}}Lv^{\frac{3}{2}},\\
\label{eqn:defWd}
W_d &=& I_t^2 I_t\frac{\partial}{\partial z_s}I_t\frac{\partial}{\partial z_r}I_t^2
\end{eqnarray}
and 
\begin{equation}
\label{eqn:rk5sigma} 
\bar{F}[\kappa,\rho]^{\dagger}  = W_m^{-1} (\bar{F}^T W_d)_{\kappa} 
\end{equation}
Then
\begin{equation}
\label{eqn:rk6sigma}
\bar{F}[\kappa,\rho]^{\dagger}\bar{F}[\kappa,\rho][\bar{r}_{\kappa},\bar{r}_{\rho}]
\approx  
\frac{1}{8 \pi^3}\int dk_x dk_z dk_h e^{i(xk_{x}+zk_{z}+hk_{h})}B P  
 (\hat{\bar{r}}_{\kappa} + \hat{\bar{r}}_{\rho} \cos 2 \theta )  
\end{equation} 
\end{theorem}
\begin{proof}
The concluion \ref{eqn:rk6sigma} follows from \ref{eqn:rk4sigma} and the calculus of 
pseudodifferential operators.
\end{proof}

The definition \ref{eqn:rk5sigma} is equivalent to equation 21 in \cite[]{HouSymes:15}, or
equation 10 in \cite[]{HouSymes:17}, once the translation from
$\kappa$ to $v$ (with $\rho=1$) is accounted for, for the focused case
where most energy in $R$ is concentrated at $h=0$. This equivalence
will be explained in detail in the next section.

Recall from \cite[]{HouSymes:15} that $W_d$ (defined in
\ref{eqn:defWd}) is acts as a pseudodifferential operator with a
positive real symbol of order -2, when restricted to a conic set of
covectors corresponding to downgoing source and upcoming receiver
wavefields. Therefore, for data satisfying the downgoing/upcoming
constraint, $W_d$ defines a weighted norm equivalent to the Sobolev
$-2$ norm. Since $\bar{F}$ and $\bar{F}^T$ are of order zero, in the
sense that they preserve the order of Sobolev classes, and $W_m$ is
plainly of order 2, $F^{\dagger}$ is also of order zero. Furthermore,
since $W_m$ is also positive (semi-)definite symmetric, it also
defines a norm on the a subspace of relative model perturbations equivalent
to the Sobolev 2-norm. We conclude that when restricted to relative
perturbations of the form $[\bar{r}_{\kappa},0]$,  $\bar{F}^{\dagger}$ is the
adjoint of $\bar{F}$ with respect to the norms in model and data
spaces defined by $W_m$ and $W_d$. This observation was used to
accelerate the solution of the least squares inverse problem in
\cite[]{HouSymes:16}.

A similar result is possible based on the second approximationt
developed above. Define 
\begin{eqnarray}
\label{eqn:defVminv}
V_m & = & 32 v^2 \frac{\partial}{\partial z} \\
\label{eqn:defVd}
V_d &=& -\frac{\partial}{\partial t}I_t^2 I_t\frac{\partial}{\partial z_s}I_t\frac{\partial}{\partial z_r}I_t^2
\end{eqnarray}
and 
\begin{equation}
\label{eqn:rk5sigmabis} 
\bar{F}[\kappa,\rho]^{\ddagger}  = V_m (\bar{F}^T V_d)_{\kappa} 
\end{equation}
Using the calculus again, together with equation \ref{eqn:rk4sigmabis}, obtain
\begin{equation}
\label{eqn:rk6sigmabis}
\bar{F}[\kappa,\rho]^{\ddagger}\bar{F}[\kappa,\rho][\bar{r}_{\kappa},\bar{r}_{\rho}]
\approx  
\frac{1}{8 \pi^3}\int dk_x dk_z dk_h e^{i(xk_{x}+zk_{z}+hk_{h})}B P_1
 (\hat{\bar{r}}_{\kappa} + \hat{\bar{r}}_{\rho} \cos 2 \theta ) 
\end{equation}
This result is equivalent to conclusion of \cite{HouSymes:17}, as will
be explained below.

One key difference between the representations \ref{eqn:rk5sigma} and
\ref{eqn:rk5sigmabis} is that the operators figuring in the latter,
$V_m$ and $V_d$, are not even symmetric, let alone positive definite,
so do not define alternate norms. Thus \ref{eqn:rk5sigmabis} does not
appear to exhibit $\bar{F}^{\dagger}$ as an adjoint. This appearance
is however deceptive. Note that $ik_z = |k_z|i{\rm sgn}\,k_z$. The
first factor is positive real, and the rest (the
Fourier multiplier defining the Hilbert transform)  is skew symmetric;
both can be made into symbols of order 1 and 0 respectively by
modification around $k_z=0$.  Define
\begin{eqnarray}
\label{eqn:kzkt1} 
K_zu(x,z,h) &=& \int dk_z e^{i(k_xx+k_z z+k_hh)}
                (1-\phi(k_z))|k_z|\hat{u}(k_x,k_z,k_h)\\
\label{eqn:kzkt2} 
H_zu(x,z,h) &=& \int dk_z e^{i(k_xx+k_z z+k_hh)}
                (1-\phi(k_z))i\rm{sgn}\,k_z\,\hat{u}(k_x,k_z,k_h)\\
\label{eqn:kzkt3} 
K_tu(x_r,t;x_s) &=& \int d\omega e^{i(k_{x_r} x_r+k_{x_s} x_s+\omega
                    t)}
                    (1-\phi(\omega))|\omega|\hat{u}(k_{x_r},\omega;k_{x_s}) \\
\label{eqn:kzkt4} 
H_tu(x_r,t;x_s) &=& \int d\omega e^{i(k_{x_r} x_r+k_{x_s} x_s+\omega
                    t)}
                    (1-\phi(\omega))i\rm{sgn}\,\omega\,\hat{u}(k_{x_r},\omega;k_{x_s}) 
\end{eqnarray}
in which $\phi \in C^{\infty}_0(\bR)$ is $=1$ in a neigborhood of
$0$. 
\begin{theorem} The operators defined in
  \ref{eqn:kzkt1}-\ref{eqn:kzkt4}  satisfy
\begin{eqnarray}
\frac{\partial}{\partial z} & \approx& K_z H_z = H_z K_z, \nonumber \\
\frac{\partial}{\partial t} &\approx& K_t H_t = H_t K_t, \nonumber\\
H_z \bar{F}^T &\approx& \bar{F}^T H_t.
\label{eqn:kzkt5}
\end{eqnarray}
Define 
\begin{eqnarray}
\label{eqn:defbarWm}
\bar{W}_m^{-1} &=& 32 v^2 K_z \\
\label{eqn:defbarWd}
\bar{W}_d & = & K_t I_t^2 I_t\frac{\partial}{\partial 
  z_s}I_t\frac{\partial}{\partial z_r}I_t^2\\
\label{eqn:defdd}
\bar{F}^{\ddagger} &=& \bar{W}_m^{-1}
(\bar{F}^T\bar{W}_d)_{\kappa}.
\end{eqnarray}
Then
\begin{equation}
\label{eqn:rk6sigmaterce}
\bar{F}[\kappa,\rho]^{\ddagger}\bar{F}[\kappa,\rho][\bar{r}_{\kappa},\bar{r}_{\rho}]
\approx  
\frac{1}{8 \pi^3}\int dk_x dk_z dk_h e^{i(xk_{x}+zk_{z}+hk_{h})}B P_1
 (\hat{\bar{r}}_{\kappa} + \hat{\bar{r}}_{\rho} \cos 2 \theta ) .
\end{equation}
\end{theorem}
\begin{proof}
The first two approximations in \ref{eqn:kzkt5} follow directly from
the definitions, the third from Egorov's Theorem \cite[]{Tay:81}.
Since $H_t^2 \approx -I$,  
\[ 
V_m \bar{F}^T V_d =- 32 v^2 \frac{\partial}{\partial z} \bar{F}^T  
\frac{\partial}{\partial t}I_t^2 I_t\frac{\partial}{\partial 
  z_s}I_t\frac{\partial}{\partial z_r}I_t^2  
\]
\begin{equation}
\label{eqn:getspd}
=32 v^2 K_z \bar{F}^T K_t I_t^2 I_t\frac{\partial}{\partial 
  z_s}I_t\frac{\partial}{\partial z_r}I_t^2  
\end{equation} 
and  equations \ref{eqn:getspd} and \ref{eqn:rk5sigmabis} combine to 
yield the results \ref{eqn:rk6sigmaterce}.
\end{proof}

Since $\bar{W}_m$ and $\bar{W}_d$ are positive (semi-)definite
symmetric under the same circumstances as $W_m$ and $W_d$,
$\bar{F}^{\ddagger}$ can actually be viewed as an adjoint, hence
the representation \ref{eqn:rk6sigmaterce} gives another way of seeing
$\bar{F}$ as almost-unitary.

\section{Focused Inversion}
In this section, assume that
$\bar{r}_{\kappa}$ and $\bar{r}_\rho$ are physical,
  that is,
\[
\bar{r}_{\kappa} = r_{\kappa}\delta(h), r_{\kappa} =\frac{\delta
  \kappa}{\kappa}; \,\,
\bar{r}_{\rho} = r_{\rho}\delta(h), r_{\rho}=\frac{\delta \rho}{\rho} 
\]
Also, as noted before,
\[
r_{\kappa} = r_{\rho} + 2 r_{v}, \,\,r_{\sigma} = r_{\rho}+r_{v} =
\frac{1}{2}(r_{\kappa}+r_{\rho})
\]

The first task of this section is to establish the equivalence of the
inversion formulae \ref{eqn:rk6sigma} and \ref{eqn:rk6sigmabis} with the
results presented by \cite{HouSymes:15,HouSymes:17}. The remainder
will be devoted to estimation of $r_{\kappa}$ and $r_{\rho}$ (or the
equivalent) separately.

The papers \cite[]{HouSymes:15,HouSymes:17} analyzed the mapping
$\delta v \rightarrow \delta d$, assuming that $\rho = 1$, $\delta \rho =
0$. Denote this map by $F_v$, and its subsurface offset extension by
$\bar{F}_v$. Since $\delta \rho = 0$ implies that $r_{\kappa} = 2 r_v
= 2 \delta v / v$, $F_v$ is related to $F$ by 
\[
Fr_{\kappa} = F_v \delta v = F_v (v r_{\kappa}/2
\]
whence 
\[ 
F = \frac{1}{2}F_v v
\]
For nearly focused extended models $\bar{r}_{\kappa}$, supported near
$h=0$, this implies that 
\[
\bar{F} \approx \bar{F}_v \frac{v}{2}
\]
so in that case
\begin{equation}
\label{eqn:vkap}
\bar{F}^{\dagger} \bar{F} = W_m^{-1} \bar{F}^T W_d \bar{F}  \approx
W_m^{-1} \frac{v}{2} \bar{F}_v^T W_d \bar{F}_v \frac{v}{2} \approx
W_m^{-1}\frac{v^2}{4} \bar{F}_v^T W_d \bar{F}_v
\end{equation}
From equations \ref{eqn:rk6sigma} and \ref{eqn:defWminv}, it follows
that for focused extended models, 
\begin{equation}
\label{eqn:hou1}
\bar{F}_v^{\dagger} \bar{F}_v \approx P_v
\end{equation}
where $P_v$ is a pseudodifferential operator that acts as an
approximate identity on focused extended models, and 
\begin{eqnarray}
\label{eqn:hou2}
\bar{F}_v^{\dagger} &=& W_v^{-1}\bar{F}_v^TW_d\\
\label{eqn:hou3}
W_v^{-1} &=& 4 v^5 L.
\end{eqnarray}
Equations \ref{eqn:hou1}, \ref{eqn:hou2}, and \ref{eqn:hou3} precisely
recapitulate the main result of \cite{HouSymes:15}.

The factor $B$ did not appear in these earlier analyses - for the constant density
case, it plays little role, since the method for extracting an
estimate of $r_v$ is simply stacking over $h$. The stack weight in
principle can be any function of $h$ that $=1$ at $h=0$. The choice
made in \cite[]{HouSymes:15} is to use the full range (in principle,
infinite) of $h$, which is equivalent to sampling at $k_h=0$, or $p_h=0$.
As noted in the last section, for $h=0$, $B=1$ for $|p_xp_h| < 1$, and
therefore for the purposes of extracting $r_v$ the factor $B$ can also
be ignored.

However to extract both $r_{\sigma}$ and $r_{\rho}$, it is no longer
possible to sample at $k_h=0$, since that also implies $\theta=0$,
i.e. normal incidence reflection, at which $r_{\rho}$ has no influence.
According to the identities \ref{eqn:theta3}, valid at $h=0$, the
non-oscillatory part of the integrand in \ref{eqn:rk5sigma} is
\[
B (\hat{r}_{\sigma} - \hat{r}_{\rho} \sin^2 \theta )
 = B \left(\hat{r}_{\sigma} - \hat{r}_{\rho} \frac{k_h^2}{k_z^2+k_h^2}\right)
\]
To isolate $r_{\rho}$, note that
\[
\left.-\frac{1}{2}\frac{\partial^2}{\partial k_h^2}
  k_z^2\left(\hat{r}_{\sigma} -
  \frac{k_h^2}{k_z^2+k_h^2}\hat{r}_{\rho}(k_x,k_z)\right)\right|_{k_h=0} = \hat{r}_{\rho}(k_x,k_z)
\]
The second $k_h$ derivative corresponds to multiplication by $-h^2$,
and multiplication by $k_z^2$ to application of
$-(\partial_z^2+\partial_h^2)$. Finally, restriction to $k_h=0$ is
equivalent to (unweighted) stacking over $h$. Note once again that
stacking in effect makes $B=1$, to leading order in frequency.

The upshot is that in the focused case, the relative impedance and
density perturbations  can be extracted from the modified migrated
image volume $\bar{R}$:
\begin{eqnarray}
r_{\sigma}(x,z) &=& \int \, dh \, \bar{R}(x,z,h) \nonumber \\
r_{\rho}(x,z) &=& -\frac{1}{2}\int \, dh \,h^2
                  \frac{\partial^2}{\partial z^2}\bar{R}(x,z,h) 
\label{eqn:vdinv}
\end{eqnarray}

Note that if $\bar{R}$ is defined by \ref{eqn:rk5sigma}, then $d \approx
\bar{F}[\kappa,\rho](\bar{R},0)$. That is, the $\kappa$ component
defines an asymptotic right inverse. A precise statement of this type requires microlocal
constraints on $d$, following ten Kroode. Modulo these constraints,
this observation may be paraphrased, ``the extended
model fits everything'', in which by ``extended model'' is understood
the offset-dependent relative bulk modulus perturbation constructed
above, constructed with possibly variable background density
but zero density perturbation.

The estimates \ref{eqn:vdinv} are asymptotically correct hence can be
used to precondition the least squares estimation of the acoustic parameters.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-19 repro/papers/mps_precond/project/parseCG.py

@@ -1,0 +1,185 @@
+import os
+import sys
+
+class InputError(Exception):
+    """Exception raised for errors in the input.
+    Attributes:
+        inarg -- missing input argument key
+        message -- explanation of the error
+    """
+
+    def __init__(self, inarg):
+        self.inarg = inarg
+        self.message = 'Missing input argument, '+str(inarg)+'\n'
+
+
+if __name__ == '__main__':
+    """ 
+    This method parses output from CG algorithm and spits out an ASCII file 
+    with residual and gradient norms.
+
+    Arguments:
+        infile -- name of input CG file
+        out_nres -- file name of output ASCII file with norm of residuals
+        out_grad -- file name of output ASCII file with norm of gradient (normal residuals)
+        out_xnrm -- file name of output ASCII file with norm of solution
+        out_Lxnrm --
+        out_xerr --
+        out_Lxerr --
+        Niter -- size of output vector containing residual and gradient norms.
+                 If Niter > # iterations reported by CG file then the ASCII 
+                 file will be padded by NaN.
+    """
+
+    try:
+        #reading input args
+        infile = []
+        out_nres = []
+        out_grad = []
+        out_xnrm = []
+        out_Lxnrm = []
+        out_xerr = []
+        out_Lxerr = []
+        Niter = []
+        
+        for tmp in sys.argv:
+            if tmp.find("infile=")==0:
+                infile = tmp[7:]
+            if tmp.find("out_nres=")==0:
+                out_nres = tmp[9:]
+            if tmp.find("out_grad=")==0:
+                out_grad = tmp[9:]
+            if tmp.find("out_xnrm=")==0:
+                out_xnrm = tmp[9:]
+            if tmp.find("out_Lxnrm=")==0:
+                out_Lxnrm = tmp[10:]
+            if tmp.find("out_xerr=")==0:
+                out_xerr = tmp[9:]
+            if tmp.find("out_Lxerr=")==0:
+                out_Lxerr = tmp[10:]
+            if tmp.find("Niter=")==0:
+                Niter = int(tmp[6:])
+                
+        #checking input args
+        if infile==[]:
+            raise InputError("infile")
+        if out_nres==[]:
+            raise InputError("out_nres")
+        if out_grad==[]:
+            raise InputError("out_grad")
+        if Niter==[]:
+            raise InputError("Niter")
+
+        #reading in CG file
+        iter=[]
+        nres=[]
[…]

### 2019-10-19 repro/papers/mps_theory/project/appxrates.py

@@ -1,0 +1,148 @@
+import os
+import sys
+import array
+import math
+
+class InputError(Exception):
+    """Exception raised for errors in the input.
+    Attributes:
+        inarg -- missing input argument key
+        message -- explanation of the error
+    """
+
+    def __init__(self, inarg):
+        self.inarg = inarg
+        self.message = 'Missing input argument, '+str(inarg)+'\n'
+
+
+if __name__ == '__main__':
+    """ 
+    This method reads in SU files and to compute approximate convergence rates
+    via Richardson extrapolation.
+
+    Arguments:
+        in_h -- input SU file, solution with h spatial discretization
+        in_2h -- input SU file, solution with 2h spatial discretization
+        in_4h -- input SU file, solution with 4h spatial discretization
+        nt -- size of time traces
+        ntr -- number of traces
+        outfile -- name of output ASCII file, containing convergence rates 
+                   computed using L2 and sup norms in time.
+    """
+
+    try:
+        #reading input args
+        in_h  = []
+        in_2h = []
+        in_4h = []
+        nt = []
+        ntr = []
+        outfile = []
+        
+        for tmp in sys.argv:
+            if tmp.find("in_h=")==0:
+                in_h = tmp[5:]
+            if tmp.find("in_2h=")==0:
+                in_2h = tmp[6:]
+            if tmp.find("in_4h=")==0:
+                in_4h = tmp[6:]
+            if tmp.find("nt=")==0:
+                nt = int(tmp[3:])
+            if tmp.find("ntr=")==0:
+                ntr = int(tmp[4:])
+            if tmp.find("outfile=")==0:
+                outfile = tmp[8:]
+
+        #checking input args
+        if in_h==[]:
+            raise InputError("in_h")
+        if in_2h==[]:
+            raise InputError("in_2h")
+        if in_4h==[]:
+            raise InputError("in_4h")
+        if nt==[]:
+            raise InputError("nt")
+        if ntr==[]:
+            raise InputError("ntr")
+        if outfile==[]:
+            raise InputError("outfile")
+
+        data_h = array.array('f')
+        f_h = open(in_h,'rb')
+        data_h.fromfile(f_h,nt*ntr)
+        f_h.close()
+
+        data_2h = array.array('f')
+        f_2h = open(in_2h,'rb')
+        data_2h.fromfile(f_2h,nt*ntr)
+        f_2h.close()
[…]

### 2019-10-19 repro/papers/mps_theory/project/appxrates2.py

@@ -1,0 +1,154 @@
+import os
+import sys
+import array
+import math
+
+class InputError(Exception):
+    """Exception raised for errors in the input.
+    Attributes:
+        inarg -- missing input argument key
+        message -- explanation of the error
+    """
+
+    def __init__(self, inarg):
+        self.inarg = inarg
+        self.message = 'Missing input argument, '+str(inarg)+'\n'
+
+
+if __name__ == '__main__':
+    """ 
+    This method reads in SU files and to compute approximate convergence rates
+    via Richardson extrapolation.
+
+    Arguments:
+        in_h -- input SU file, solution with h spatial discretization
+        in_2h -- input SU file, solution with 2h spatial discretization
+        in_4h -- input SU file, solution with 4h spatial discretization
+        nt -- size of time traces
+        ntr -- number of traces
+        outfile -- name of output ASCII file, containing convergence rates 
+                   computed using L2 or sup norms in time.
+    """
+
+    try:
+        #reading input args
+        in_h  = []
+        in_2h = []
+        in_4h = []
+        nt = []
+        ntr = []
+        outfile = []
+        norm = []
+        
+        for tmp in sys.argv:
+            if tmp.find("in_h=")==0:
+                in_h = tmp[5:]
+            if tmp.find("in_2h=")==0:
+                in_2h = tmp[6:]
+            if tmp.find("in_4h=")==0:
+                in_4h = tmp[6:]
+            if tmp.find("nt=")==0:
+                nt = int(tmp[3:])
+            if tmp.find("ntr=")==0:
+                ntr = int(tmp[4:])
+            if tmp.find("outfile=")==0:
+                outfile = tmp[8:]
+            if tmp.find("norm=")==0:
+                norm = tmp[5:]
+
+        #checking input args
+        if in_h==[]:
+            raise InputError("in_h")
+        if in_2h==[]:
+            raise InputError("in_2h")
+        if in_4h==[]:
+            raise InputError("in_4h")
+        if nt==[]:
+            raise InputError("nt")
+        if ntr==[]:
+            raise InputError("ntr")
+        if outfile==[]:
+            raise InputError("outfile")
+        if norm==[]:
+            raise InputError("norm")
+
+        data_h = array.array('f')
+        f_h = open(in_h,'rb')
+        data_h.fromfile(f_h,nt*ntr)
+        f_h.close()
+
+        data_2h = array.array('f')
[…]

### 2019-10-19 repro/papers/mps_theory/project/snapshot.py

@@ -1,0 +1,82 @@
+import os
+import sys
+import array
+import math
+
+class InputError(Exception):
+    """Exception raised for errors in the input.
+    Attributes:
+        inarg -- missing input argument key
+        message -- explanation of the error
+    """
+
+    def __init__(self, inarg):
+        self.inarg = inarg
+        self.message = 'Missing input argument, '+str(inarg)+'\n'
+
+
+if __name__ == '__main__':
+    """ 
+    This method reads in SU files and to compute approximate convergence rates
+    via Richardson extrapolation.
+
+    Arguments:
+        infile -- input binary file
+        it -- input time index for snapshot
+        nt -- size of time traces
+        ntr -- number of traces
+        outfile -- name of output ASCII file, containing convergence rates 
+                   computed using L2 and sup norms in time.
+    """
+
+    try:
+        #reading input args
+        it = []
+        infile = []
+        nt = []
+        ntr = []
+        outfile = []
+        
+        for tmp in sys.argv:
+            if tmp.find("it=")==0:
+                it = int(tmp[3:])
+            if tmp.find("infile=")==0:
+                infile = tmp[7:]
+            if tmp.find("nt=")==0:
+                nt = int(tmp[3:])
+            if tmp.find("ntr=")==0:
+                ntr = int(tmp[4:])
+            if tmp.find("outfile=")==0:
+                outfile = tmp[8:]
+
+        #checking input args
+        if it==[]:
+            raise InputError("it")
+        if infile==[]:
+            raise InputError("infile")
+        if nt==[]:
+            raise InputError("nt")
+        if ntr==[]:
+            raise InputError("ntr")
+        if outfile==[]:
+            raise InputError("outfile")
+
+        data = array.array('f')
+        fp = open(infile,'rb')
+        data.fromfile(fp,nt*ntr)
+        fp.close()
+
+        buff = []
+        f = open(outfile,'w') 
+
+        print('after having read file\n');
+
+        for itr in range(ntr):
+            buff = data[it+itr*nt]
+            f.write(str(buff)+' ')
+        f.write('\n')
+
+        f.close()
+    except InputError as err:
+        print 'ERROR from snapshot.py :',err.message
+        raise

### 2019-10-19 repro/papers/ms/code/main/msfcnl.cc

@@ -1,0 +1,358 @@
+#include "gridpp_top.hh"
+#include "op.hh"
+
+using TSOpt::GridSpace;
+using namespace RVL;
+
+class mvaop: public Operator<float> {
+private:
+
+  // wavelet file input name
+  string wname;
+
+  // data file input name
+  string dname;
+
+  // model space : bvel
+  GridSpace<float> const & dom;
+
+  // data space : image
+  GridSpace<float> const & rng;
+
+  // elevation of sources and receivers
+  float r_z;
+  float s_z;
+    
+  // default construction disabled
+  mvaop();
+    
+protected:
+    
+  void apply(const Vector<float> & x, 
+	     Vector<float> & y) const {
+    try {
+
+      // extract const Grid reference from range (data) space -
+      Grid<float> const & g = rng.getGrid();
+
+      // the rsf/grid data is in the various axes of
+      // the Grid g; assign to variables named in your
+      // usual way
+      int h_x_n = g.getAxis(2).n;
+      float h_x_d = g.getAxis(2).d;
+      float h_x_o = g.getAxis(2).o;
+
+      // note that r_z, s_z are not properties of the
+      // rsf grid of the data - so they must be stored in 
+      // data of the operator class. This is wrong, as these
+      // numbers are really properties of the data. To include
+      // them, we will need to start using the tfile construction.
+
+      // next, get the filenames associated with the input and 
+      // output vectors. The AssignParams function does this:
+      // create it, then evaluate it on the input and output
+      // vectors. The effect of AssignParam evaluation is to 
+      // copy the data_type and filename properties of a GridSpace
+      // vector into a "data_type = filename" pair in the PARARRAY
+      // referenced by the AssignParams object.
+ 
+      // create empty param lists
+      PARARRAY * xpar = ps_new();
+      PARARRAY * ypar = ps_new();
+
+      // create two AssignParams functions, make them refer
+      // to the two PARARRAYs.
+      AssignParams xap(*xpar,stdout);
+      AssignParams yap(*ypar,stdout);
+
+      // transfer filenames from input and output vectors to 
+      // PARARRAYS. If x is a vector in a ProductSpace, i.e. 
+      // has more than one component, the AssignParam function
[…]

### 2019-10-20 repro/papers/varden/paper.tex
added repro/papers/varden

\title{Asymptotic Inversion of the Variable Density Acoustic
  Model\\Part 2: Implementation and Examples}

\author{
Raanan Dafni\thanks{Department of Computational and Applied Mathematics, Rice University, 
Houston, TX, 77005, USA, 
{\tt symes@rice.edu}},
William Symes\thanks{Department of Computational and Applied Mathematics, Rice University, 
Houston, TX, 77005, USA, 
{\tt symes@rice.edu}}
}

\lefthead{Dafni Symes}

\righthead{Asymptotic Acoustic Inversion}

\maketitle
\parskip 12pt

\begin{abstract}
Techniques developed for asymptotic wave equation based linearized inversion of
subsurface offset extended constant density acoustics also apply to variable density
acoustics. The data may be matched by proper choice of an extended
relative bulk modulus perturbation, with zero density
perturbation. This bulk modulus perturbation may be treated as an
offset-dependent reflectivity volume, and AVA-like techniques applied
to extract asymptotically correct estimates of velocity and density
perturbations, assuming that the background velocity is kinematically
consistent with the data.

The first part of this paper presents a 
theoretical analysis of the linearized inversion, leading to the
asymptotic wave equation algorithm. This second part
describes implementation details, explains how the asymptotic inverse
may be used to accelerate iterative inversion algorithms (``least
squares migration''), and  presents numerical examples of both
asymptotic and iterative inversion.
\end{abstract}
\setlength{\parindent}{0cm}

\section{Preliminaries}

Position vectors are $\bx = (x,z)$; $x$ is the horizontal position
vector, 1D or 2D for 2D or 3D modeling respectively.

The techniques explained here apply to acoustic data, or data that can
be satisfactorily treated as acoustic, recorded in a water layer at
the top of the half-space $z>0$. It's conceivable that similar
techniques may be developed in the future for other wave physics and
other acquisition geometries, but so far as I know such extensions
have not yet been worked out.

The pair $\kappa,\beta$ of bulk modulus and buoyancy (reciprocal
density) respectively is a convenient parametrization of the acoustic
model for present purposes. The acoustic fields $p,\bv$ solve the
acoustic wave equations
\begin{eqnarray}
\label{eqn:awe}
\frac{\partial p}{\partial t}& =& -\kappa (\nabla \cdot \bv +
w(t)\delta(\bx-\bx_s)),\nonumber \\
\frac{\partial \bv}{\partial t} & = & -\beta \nabla p. 
\end{eqnarray}  
The perturbational acoustic fields $\delta p, \delta \bv$ due to
perturbations $\delta \kappa, \delta \beta$ solve the
related equations 
\begin{eqnarray}
\label{eqn:pawe}
\frac{\partial \delta p}{\partial t}& =& -\kappa \nabla \cdot \delta \bv -
\delta \kappa (\nabla \cdot \bv + w(t)\delta(\bx-\bx_S),\nonumber \\
\frac{\partial \delta \bv}{\partial t} & = & -\beta \nabla \delta p - \delta
                                      \beta \nabla p. 
\end{eqnarray}  
in which $p,\bv$ are solutions of equation \ref{eqn:awe}.
The source characteristics (wavelet $w(t)$, locations $\bx_s$) are
assumed known.

$\dF$ denotes the acoustic Born modeling operator in the half-space $z>0$
adjoined to an infinite homogeneous half-space in $z<0$, that is, an
infinitely deep ocean. An implementable approximation of this
configuration uses an
absorbing boundary condition (PML) at the $z=0$ surface of a
rectangular domain, and also use absorbing boundary conditions at the
other boundary surfaces of the rectangle. Assume that the recorded
data are simply the samples of pressure at receiver
locations, {\em for both sources and receivers located on the
  bounndary $z=0$}. Explicitly,
\[
\dF[\kappa,\beta][\delta \kappa,\delta \beta](x_r,t;x_s) = \{\delta p((x_r,0),t;(x_s,0))\}
\]
with $x_r, x_s$ ranging over survey source-receiver horizontal
locations.

$\dF_0$ denotes the free surface (pressure-release) Born modeling operator in
the same half-space. The free surface acoustic field $p_0,\bv_0$ solves
the acoustic system \ref{eqn:awe} with the additional
Dirichlet boundary condition $p_0=0$ on $z=0$. The perturbational
field $\delta p_0,\delta \bv_0$ is defined similarly. Sources and receivers
must be positioned at positive depths $z_r>0$, $z_s>0$ (else the
output is boring) 
and the same horizontal locations as the hypothetical
absorbing-surface survey.

Define ``shallow tow depth'' to mean that for the data
spectrum of interest, the finite difference approximation 
\begin{equation}
\label{eqn:dipole}
\left(\frac{\partial}{\partial z_r}\frac{\partial}{\partial
  z_s}\delta p\right)((x_r,0),t; (x_s,0)) \approx
\frac{1}{4z_rz_s}\delta p_0((x_r,z_r),t;(x_s,z_s))
\end{equation}
is acceptably accurate.  
Assuming that $\delta \kappa, \delta \rho$ are non-zero only in $z>0$,
the absorbing surface perturbational field $\delta p,\delta \bv$ is
upcoming at $z=0$, that is, its high frequency energy is carried along
rays for which $z$ decreases as $t$ increases. Consequently, the depth
derivatives in equation \ref{eqn:dipole} can be computed
asymptotically to any order in frequency from the surface data $\dF$
alone, since asymptotically the fields satisfy one-way wave
equations. Let the operators on the surface data corresponding to the
partial derivatives in $z_s$, $z_r$ be denoted $D_{z_s}$ and $D_{z_r}$
respectively. \cite{HouSymes:15} show that $-D_{z_s}D_{z_r}$ is a
positive semi-definite operator of order 2. The definition of
``shallow tow depth'' \ref{eqn:dipole}
can be re-written
\begin{equation}
\label{eqn:opdipole}
D_{z_s}D_{z_r}\dF[\kappa,\rho][\delta \kappa,\delta \beta](x_r,t;x_s) \approx 
\frac{1}{4z_rz_s} Z \dF_0[\kappa,\beta][\delta \kappa,\delta \beta](x_r,t;x_s) 
\end{equation}
The operator $Z$ assigns traces with source and receiver depths $z_s,
z_r$ to traces with source and receiver depths $=0$, with all other
header and sample data the same.

The subsurface offset flavor of extended modeling makes the mechanical
parameters $\kappa,\beta$ into operators. In the Born case, only the
perturbational parameters $\delta \kappa,\delta \beta$ are regarded as
operators. In fact, it turns out that there is really no point in
making $\delta \beta$ into an operator, as it plays no necessary role at
all in subsequent developments so set
\begin{equation}
\label{eqn:zerodbeta}
\delta \beta = 0
\end{equation}
and drop $\delta \beta$ from the argument list of the linearized operators.

The extended bulk modulus
perturbation (operator kernel) $\delta \bar{\kappa}$ has a convenient
parametrization in terms of sunken midpoint $(m,z)$ and (horizontal)
half-offset $h$. The extended perturbation fields
$\delta \bar{p}, \delta \bar{\bv}$ are solutions of the extended
version of the system \ref{eqn:pawe}:
\begin{eqnarray}
\label{eqn:pawe}
\frac{\partial \delta \bar{p}}{\partial t}& =& -\kappa \nabla \cdot \delta \bar{\bv} -
\int\,dh\,\delta \bar{\kappa}(x-h,z,h) \nabla \cdot \bv(x-2h,z,t),\nonumber \\
\frac{\partial \delta \bar{\bv}}{\partial t} & = & -\beta \nabla \delta \bar{p} 
\end{eqnarray}  
Here is how to interpret the somewhat peculiar form of the
$h$-integral: $x$ is the horizontal sunken receiver position, $m=x-h$ the
sunken midpoint, and $x-2h$ the sunken source.

With this
interpretation, the absorbing surface extended modeling operator defined by 
\begin{equation}
\label{eqn:extop}
\doF[\kappa,\beta][\delta \bar{\kappa}] =
\{\bar{p}((x_r,0),t;(x_s,0))\}
\end{equation}
is the adjoint of Claerbout's survey sinking migration operator,
interpreted as a reverse-time construction.

The free surface extended modeling operator $\doF_0$ has a similar
definition. The exended absorbing surface fields $\delta
\bar{p},\delta \bar{\bv}$ are also upcoming at $z=0$, and ``shallow
tow depth'' can also be taken to mean
\begin{equation}
\label{eqn:opdipole}
D_{z_s}D_{z_r}\doF[\kappa,\rho][\delta \bar{\kappa}]\approx 
\frac{1}{4z_rz_s} Z \doF_0[\kappa,\beta][\delta \bar{\kappa}]
\end{equation}
to acceptable accuracy.

\section{Asymptotic Unitarity}
The first part of this paper analyzes unitarity relations for a
normalized version of $\doF$, acting on {\em extended reflectivities} $\bar{r}_{\kappa}$
and $\bar{r}_{\rho}$ as 
\begin{eqnarray}
\label{eqn:relfs}
\bar{r}_{\kappa}(\bx,\bh)& = & \frac{\delta \bar{\kappa}(\bx,\bh)}{\sqrt{\kappa(\bx+\bh)\kappa(\bx-\bh)}} \nonumber \\
\bar{r}_{\rho}(\bx,\bh)& = & \frac{\delta \bar{\rho}(\bx,\bh)}{\sqrt{\rho(\bx+\bh)\rho(\bx-\bh)}}
\end{eqnarray}
The relative linearized forward operator $\bar{F}[\kappa,\rho]$ is
\begin{equation}
\label{eqn:nondim}
\bar{F}[\kappa,\rho][\bar{r}_{\kappa},\bar{r}_{\rho}] =
D\bar{{\cal F}}[\kappa,\rho][\delta 
\bar{\kappa},\delta\bar{\rho}]
\end{equation}
where $\bar{r}_{\kappa},\bar{r}_{\rho}$ are related to $\delta 
\bar{\kappa},\delta\bar{\rho}$ by equation \ref{eqn:relfs}. That is, 
\begin{equation}
\label{eqn:nondimbis}
\bar{F}[\kappa,\rho][\frac{\delta \bar{\kappa}}{\sqrt{\kappa_+\kappa_-}}, \frac{\delta \bar{\rho}}{\sqrt{\rho_+\rho_-}}] =
D\bar{{\cal F}}[\kappa,\rho][\delta 
\bar{\kappa},\delta\bar{\rho}]
\end{equation}
using the notation $\kappa_{\pm}(\bx,\bh) = \kappa(\bx \pm \bh)$ and
so on, introduced in part 1.

Recall from the last section that $\delta \bar{\rho} = 0$ in the
sequel, and that $D\bar{\cal F}$ is regarded here as acting only on
$\delta \bar{\kappa}$. Therefore the same is assumed for the
normalized operator $\bar{F}$. Also, since $\kappa_{\pm}$ differs from
$\kappa$ by a (multiplication) $\Psi$DO whose symbol is $=1$ at $h=0$,
substituting $\kappa$ for $\kappa_{\pm}$ and $\rho$ for $\rho_{\pm}$
commits an error of the type we will ignore in the approximations to
follow, so we make this substitution without comment. Combining these
two observations, we write
\begin{equation}
\label{eqn:nondimterce}
\bar{F}[\kappa,\rho]\left[\frac{\delta \bar{\kappa}}{\kappa}\right] \approx  
D\bar{\cal F}[\kappa,\rho]\delta \bar{\kappa} 
\end{equation} 
or more concisely 
\begin{equation}
\label{eqn:nondimquatro}
\bar{F} \approx\doF \kappa
\end{equation}
in which $\kappa$ is shorthand for the operator of multiplication by $\kappa$.

We will use only the second of the unitarity results for $\bar{F}$,
Theorem 2, from the first part of the paper. We repeat this result for
convenience. First, define asymptotic approximations of the Hilbert transforms of the $z$ and $t$
derivatives:
\begin{eqnarray}
\label{eqn:kzkt1} 
K_zu(x,z,h) &=& \int dk_z e^{i(k_xx+k_z z+k_hh)}
                (1-\phi(k_z))|k_z|\hat{u}(k_x,k_z,k_h)\\
\label{eqn:kzkt3} 
K_tu(x_r,t;x_s) &=& \int d\omega e^{i(k_{x_r} x_r+k_{x_s} x_s+\omega
                    t)}
                    (1-\phi(\omega))|\omega|\hat{u}(k_{x_r},\omega;k_{x_s}) \\
\end{eqnarray}
in which $\phi \in C^{\infty}_0(\bR)$ is $=1$ in a neigborhood of
$0$. 
Define the model and data space weight operators by
\begin{eqnarray}
\label{eqn:defbarWm}
\bar{W}_m^{-1} &=& 32 v^2 K_z \\
\label{eqn:defbarWd}
\bar{W}_d & = & K_t I_t^2 I_t
D_{z_s}I_t
D_{z_r}I_t^2
\end{eqnarray}
Then
\begin{equation}
\label{eqn:defdd}
\bar{F}^{\ddagger} = \bar{W}_m^{-1}(\bar{F}^T\bar{W}_d)_{\kappa}.
\end{equation}
satisfies
\begin{equation}
\label{eqn:rk6sigmaterce}
\bar{F}[\kappa,\rho]^{\ddagger}\bar{F}[\kappa,\rho][\bar{r}_{\kappa}]
\approx  
\frac{1}{8 \pi^3}\int dk_x dk_z dk_h e^{i(xk_{x}+zk_{z}+hk_{h})}B P
 (\hat{\bar{r}}_{\kappa}) .
\end{equation}

Even though no ray-based computations are required to evaluate $\bar{F}^{\ddagger}$,
the derivation of its properties is
ray-theoretic, and depends on the assumptions
that the background fields $\kappa, \beta$ are smooth (slowly varying
on wavelenghth scale) and satisfy the ``DSR hypothesis'': that all
rays from source or receiver to scattering point maintain a nonzero
vertical velocity throughout their extent. A consequence of the DSR
hypothesis is that the energetically significant part of the source
wavefield should be downgoing, and the receiver wavefield should be
upcoming, at least near the sources and receivers.

The DSR assumption can be
stated: turning rays carry negligible energy. This hypothesis is
sometime true, sometimes not, and constitutes a real limitation on the
approximate inverse construction.

As discussed in part 1, for physical $\bar{r}_{\kappa} =
r_{\kappa}\delta(h)$, the operators $B$ and $P$ act as the identity, 
whence equation \ref{eqn:rk6sigmaterce} exhibits approximate unitarity 
of $\doF$ with respect to the weighted norms 
\begin{equation}
\label{eqn:wnorms}
\langle m_1,m_2\rangle_m = \langle m_1,W_m m_2 \rangle,\,\, 
\langle d_1,d_2\rangle_d = \langle d_1,W_d d_2 \rangle. 
\end{equation}

A unitarity result for $\doF$ follows immediately from that for
$\bar{F}$. Indeed, from \ref{eqn:nondimquatro} and
\ref{eqn:rk6sigmaterce} obtain
\[
PB \approx \bar{F}^{\ddagger} \bar{F} = \bar{W}_m^{-1} \bar{F}^T
\bar{W}_d  \bar{F} 
\]
\[
\approx \bar{W}_m^{-1} \kappa \doF^T \bar{W}_d \doF \kappa
\]
\begin{equation}
\label{eqn:rk6sigmaquatro}
\approx W_m^{-1}\doF^T \bar{W}_d \doF 
\end{equation}
in which
\begin{eqnarray}
\label{eqn:wmdefk}
W_m^{-1} = \kappa^2 \bar{W_m} &=& 32 \kappa^2 v^2 K_z = 32 \kappa^3 \beta
                             K_z
\end{eqnarray}

A rearrangement of $\bar{W}_d$ implicitly expresses the Green's
function perturbation, and corresponds to the implementation discussed
below. From the first sections of part 1, the pressure perturbation
corresponding to a reference field with a space-time delta source on
the RHS of the pressure equation is actually the time derivative of
the Green's function perturbation. Its trace is $\doF \delta \bar{\kappa}$, so
$I_t \doF$ is the trace of the Green's function perturbation. This
trace is already implicit in the data space weight operator:
\[
\doF^T \bar{W}_d \doF = \doF^T K_t I_t^2 I_t 
D_{z_s}I_t
D_{z_r}I_t^2 \doF
\]
\[
= -(I_t \doF)^T K_t I_t^2 
D_{z_s}
D_{z_r}I_t^2 (I_t \doF)
\]
\begin{equation}
\label{eqn:wdrearr}
=-(I_t \doF)^T K_t^{-3} 
D_{z_s}
D_{z_r} (I_t \doF)
\end{equation}
since $I_t^2 \approx -K_t^{-2}$. Define 
\begin{equation}
\label{eqn:wdredef}
W_d = -K_t^{-3}D_{z_s}D_{z_r}
\end{equation}
Then the approximation \ref{eqn:rk6sigmaquatro} is equivalent to the
unitarity relation for $I_t\doF$:
\begin{equation}
\label{eqn:rk6sigmaquinze}
PB \approx W_m^{-1} (I_t\doF)^T W_d (I_t \doF), 
\end{equation}
so that 
\begin{equation}
\label{eqn:dddef}
\iddoF = W_m^{-1} (I_t\doF)^T W_d
\end{equation}
is both the adjoint of $(I_t \doF)$ with respect to the weighted norms
defined by $W_m$ and $W_d$, and its approximate inverse for
near-physical extended $\delta \bar{\kappa}$.
[Note that we are re-purposing the symbols $W_m^{-1}$ and $W_d$ used 
in the statement of Theorem 1, part 1.]

Computation of $I_t \doF$ and its transpose simply requires use of the
$\delta(\bx-\bx_s)H(t)$, or a bandlimited approximation, in the RHS of
the reference pressure equation, that is, a bandpass filtered version of $w=H$ in \ref{eqn:awe}.

\section{Asymptotic Unitarity for Free Surface modeling}
An approximate inversion of
the free-surface modeling operator $\doF_0$ can be constructed on the
same principles. Begin with the dipole relation \ref{eqn:opdipole},
which we abbreviate as 
\begin{equation}
\label{eqn:opdipolebis}
D_{z_s}D_{z_r}\doF \approx \frac{1}{4z_rz_s} Z \doF_0
\end{equation}
The ``ghosting'' operator $D_{z_s}D_{z_r}$ (without time integrations)
is microlocally negative
definite symmetric as a consequence of the DSR hypothesis
\cite[]{HouSymes:15}, and in particular microlocally
invertible. Abusing notation a bit, write
\begin{equation}
\label{eqn:opdipoleterce}
\doF = (4z_rz_s D_{z_s}D_{z_r})^{-1} Z \doF_0
\end{equation}
and substitute \ref{eqn:opdipoleterce} in \ref{eqn:rk6sigmaquinze} to
obtain
\[
PB \approx W_m^{-1} (ZI_t\doF_0)^T (4z_rz_s D_{z_s}D_{z_r})^{-1} W_d
(4z_rz_s D_{z_s}D_{z_r})^{-1} (ZI_t \doF_0) 
\]
\begin{equation}
\label{eqn:rk6sigmads}
= W_m^{-1} (ZI_t\doF_0)^T (4z_rz_s)^{-2} K_t^{-3} D_{z_s}D_{z_r})^{-1} (ZI_t \doF_0).
\end{equation}
The relation \ref{eqn:rk6sigmads} may be re-written
\begin{equation}
\label{eqn:rk6sigmadsbis}
PB = W_m^{-1} (I_t\doF_0)^T W_d^0 (I_t\doF_0) 
\end{equation}
with 
\begin{equation}
\label{eqn:wddeffs}
W_d^0 = (4z_sz_r)^{-2} Z^T K_t^{-3} (D_{z_s}D_{z_r})^{-1} Z
\end{equation}
thus exhibiting the weighted adjoint
\begin{equation}
\label{eqn:appinvfs}
\iddoFf= W_m^{-1} (I_t \doF_0)^T W_d^0
\end{equation}
as an approximate inverse for near-physical arguments.

Construction of $\iddoFf$ according to \ref{eqn:appinvfs} appears to
require application of the ``de-ghosting'' operator
$(D_{z_s}D_{z_r})^{-1}$, per the definition \ref{eqn:wddeffs} of
$W_d^0$. As we will point out below, the shallow tow-depth hypothesis
proides an approximation to $\iddoFf$ that does not require
de-ghosting amplitude and phase adjustment.

\section{Preconditioned Conjugate Gradient Iteration}
The availability of an approximate inverse for $I_t \doF$ (and $I_t
\doF_0$) suggests that the least squares migration problem should be
reformulated as
\begin{equation}
\label{eqn:elsm}
\mbox{min}_{\delta \bar{\kappa}} \|I_t\doF[\kappa,\beta][\delta 
\bar{\kappa}] -I_t\delta d\|^2_d. 
\end{equation}
In this section, we will discuss the Conjugate Gradient (CG) algorithm
for generic problems of this form. Because the algorithms have general
applicability, and for brevity, we will use the notation $A$ for a
linear operator such as $I_t \doF$, $W_m$ and $W_d$ for positive
definite symmetric weight operators defining weighted inner products
in the domain and range of $A$ respectively, $b$ for a data vector in
the range ($I_t \delta d$ in our application), and $x$ for a model
vector in the domain ($=\delta \bar{\kappa}$ here). 
The CG algorithm is defined in terms of the weigted inner products
domain space $\langle x_1, x_2 \rangle_m = x_1^TW_m x_2$ and similarly
in the range. Denote by $A^{\dagger}$ the adjoint of $A$ with respect to the
weighted inner products.
Near-unitarity of $A$, that is, $A^{\dagger}A \approx I$, implies
rapid convergence of the CG iteration. 
\begin{algorithm}[H]
\caption{Conjugate Gradient Algorithm}
\begin{algorithmic}[1]
  \State Choose $x_0$ somehow
  \State $g_0 \gets A^{\dagger}(b-Ax_0)$
  \State $p_0 \gets g_0$ 
  \State $w_0 \gets A^{\dagger}A p_0$
  \State $k \gets 0$
  \Repeat
  \State $\alpha_k \gets \frac{\langle g_k,g_k \rangle_m}{\langle p_k,w_k\rangle_m}$
  \State $x_{k+1} \gets x_k + \alpha_k p_k$
\State $g_{k+1} \gets g_k - \alpha_kw_k$
  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},g_{k+1}\rangle_m}{\langle g_k,g_k\rangle_m}$
  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
  \State $w_{k+1} \gets A^{\dagger}Ap_{k+1}$
  \State $k \gets k+1$
  \Until{Error is sufficiently small, or max iteration count exceeded} 
\end{algorithmic}
\end{algorithm}
Several things to note:
\begin{itemize}
\item Each step requires on application of the normal operator $A^{\dagger}A$. Except for the first step, only the action of the normal  operator is required.
\item The initial step requires an application of $A^{\dagger}$ to the
  data. A common choice is $x_0 = 0$.
\item In the form presented here, the algorithm requires explicitly only the model space inner product (of course, the data space inner
  product is implicit in the definition \ref{eqn:wadj} of
  $A^{\dagger}$). 
\end{itemize}

The algorithm described above is equivalent to the well-known
Preconditioned Conjugate Gradient algorithm \cite[]{gvl}. Define
$A^*$ to be the adjoint of $\doF$ with respect to the (unaltered)
$L^2$ norm in model space and the weighted norm defined above in data
space. That is,
\begin{eqnarray}
A^* &=& A^T W_d \nonumber\\
A^{\dagger} &=& W_m^{-1}\doF^*
\label{eqn:dwadj}
\end{eqnarray}
 Also, using the notation of the
algorithm above, set 
\begin{eqnarray}
r_k &=& A^*(b - Ax_k) = W_m g_k\nonumber \\
q_k &=& W_m w_k
\label{eqn:pcgvars}
\end{eqnarray}
Then for example
\[
\langle g_k, g_k\rangle_m = \langle g_k, W_m g_k \rangle = \langle
g_k,r_k \rangle
\]
and so on, so the Conjugate Gradient Algorithm is equivalent to
\begin{algorithm}[H]
\caption{Preconditioned Conjugate Gradient Algorithm}
\begin{algorithmic}[1]
\State Choose $x_0$ somehow
  \State $r_0 \gets A^*(b-Ax_0)$
  \State $p_0 \gets W_m^{-1}r_0$
  \State $g_0 \gets p_0$
  \State $q_0 \gets A^*Ap_0$
  \State $k \gets 0$
  \Repeat
  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle p_k,q_k\rangle}$
  \State $x_{k+1} \gets x_k + \alpha_k p_k$
  \State $r_{k+1} \gets r_k - \alpha_kq_k$
  \State $g_{k+1} \gets W_m^{-1}r_{k+1}$
  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle_m}{\langle g_k,r_k\rangle_m}$
  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
  \State $q_{k+1} \gets A^*Ap_{k+1}$
  \State $k \gets k+1$
  \Until{Error is sufficiently small, or max iteration count exceeded} 
\end{algorithmic}
\end{algorithm}
Note that the Preconditioned Conjugate Gradient algorithm involves
only one application of {\em model-unweighted} normal operator
$A^*A$ and one applicatino of the model weight operator $W_m^{-1}$
per step, and only the unweighted (Euclidean) inner product in the
model space - that is, (weighted) the model space inner product has
disappeared as well. This algorithm is considerably simpler to
implement than the Conjugate Gradient Algorithm in weighted spaces, so
we will use it in all experiments reported below.

\section{Practical Computation of Adjoint and Normal Operators}
For absorbing boundary (deghosted) data, 
\[
\iddoF = -32 \kappa^3 \beta K_z (I_t\doF)^T  K_t^{-3} D_{z_s}D_{z_r} = -32 \kappa^3 \beta K_z (D_{z_s}D_{z_r}I_t\doF)^T  K_t^{-3} 
\]
\begin{equation}
\label{eqn:abadj}
\approx -32 \kappa^3 \beta K_z \frac{1}{4z_sz_r}(I_t\doF_0)^T Z^T K_t^{-3}
\end{equation}
using the dipole relation \ref{eqn:opdipolebis} and the symmetry of
the ``ghosting'' operator.

As noted earlier, the $\Psi$DOs $K_z$ and $K_t$ can be computed via 1D
Fourier transforms using tapered high- and low-pass cutoffs to
regulate frequency content to avoid aliasing, and $(I_t\doF_0)^T $ is
computed via the adjoint state method.  The normal operator is simply
the composition of the operator expressed in \ref{eqn:abadj} with
$I_t\doF_0$.

The upshot: equation \ref{eqn:abadj} provides a computation of the
approximate inverse of $I_t \doF$ that does not require explicit
computation of the ``ghosting'' operator $D_{z_s}D_{z_r}$.

Similarly, \ref{eqn:appinvfs} states that
\[
\iddoFf = -32 \kappa^3 \beta K_z (ZI_t \doF_0)^T K_t^{-3}
(4z_sz_r)^{-2} (D_{z_s}D_{z_r})^{-1}Z
\]
 \begin{eqnarray}
\label{eqn:fsadj}
\approx -32 \kappa^3 \beta K_z \frac{1}{4z_sz_r}(I_t\doF)^T K_t^{-3} Z
\end{eqnarray}
which also provides a route to computation of $\iddoFf$ via the adjoint
state method and simple filters, without requiring explicit
computation of the ``de-ghosting'' operator $(D_{z_s}D_{z_r})^{-1}$,
as expression \ref{eqn:appinvfs} appeared to require.

Note that the expressions \ref{eqn:abadj} and \ref{eqn:fsadj} depend
on the shallow tow depth assumption,  appropriate for marine towed
streamer acquisition only, and then only for a limited range of
frequencies (up to the first ghost notch).
For some other configurations, similar constructions should be
possible. For example, OBS node arrays record 4C data, including
vertical velocity data. Vertical velocity is precisely $I_tD_{z_r}$
applied to pressure data. With shallow tow depth sources as typically
used in node surveys, the $I_tD_{z_s}$ factor can be handled by free
surface modeling and the method of images as is done here. For
streamers at deeper tow depth, or to go beyond the notch, some
additional data is necessary. If source tow depth is shallow enough
that the method of images is adequate to approximate $I_t D_{z_s}$,
then the receiver side vertical velocity data could be supplied by
Geostreamer$^{TM}$ for instance without regard to tow depth . These
possibilities remain to be investigated.

It is important to note that the adjoint computations are
inexact, and that the weights as presented above are not quite SPD
operators, so need modification if they are to precisely define inner
products. Besides the microlocal ellipticity of the weights, the model
weight $W_m^{-1}$ is clearly not symmetric, but can easily be 
symmetrized:
\[
W_m^{-1} \leftarrow 32\kappa^{3/2}\beta^{1/2} K_z \beta^{1/2}\kappa^{3/2}
\]

In the special case of a constant-property layer (ideal ocean) near
$z=0$, the data weights $W_d$ and $W_d^0$ are prima facie symmetric, as their
factors are all convolutions hence commute. Thus the inner products
are actually inner products, except for being semidefinite, and act as
inner products for high-frequency microlocalized data.

The adjoint formulae \ref{eqn:abadj} and \ref{eqn:fsadj} are only as
good as the ``shallow tow depth'' assumption. That is, there is an
error of size $O(k_z^2z_rz_s)^2$ between computed and actual adjoint,
in which $k_z$ is the largest frequency carrying significant
energy. Since the error likely contributes to a nonsymmetric error in
the normal operator, CG iteration is likely to stall eventually,
requiring restart. Note that direct assessment of the error in the
adjoint is difficult, as the data side inner product is not easily
computable. However the asymmetry of the computed normal operator can
be assessed by random vector computations.

A final note on implementation: the (microlocal) definite property of
the ghosting operator is dependent on the DSR assumption, as explained
by \cite{HouSymes:15}: the source wavefields must be downgoing,
receiver wavefields upcoming, at least insofar as these carry
significan energy in the computations explained above. If these
requirements are violated, then the partial weighted adjoint $A^*A
(A=I_t \doF)$ may acquire negative eigenvalues, causing the conjugate
gradient iteration to diverge.

The downgoing/upcoming property
could be enforced by dip filtering of source and receiver
data. However for the simplified marine configuarions used in the
experiments reported below, it was simpler to separate the scattering
region from the source-receiver manifold by a horizontally homogeneous
zone (water layer).  The separation is accomplished by a depth
tapering operator, which tapers the extended bulk modulus perturbation
$\delta \bar{\kappa}$ to zero near the surface. A simple ray analysis
show that the source and receiver wavefields then must have the
required properties, and consequently we observe convergence in the
conjugate gradient iteration. We emphasize that without this
precaution, horizontally propagating energy inevitably develops during
the iteration and leads to divergence.

%%%%%%%%%%%%%%%% numerics %%%%%%%%%%%%%%

\inputdir{project}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage

\section{Example 0: Three Reflectors, Homogeneous
  Background}
This example is adapted from the first example in
\cite[]{HouSymes:15}, and reproduces it almost exactly.

The background parameters are $\kappa$=2.5 GPa, $\rho$ = 1.0 g/cc. The
density perturbation vanishes, the bulk modulus perturbation is
plotted in \ref{fig:boxdbulk}.

\plot{boxdbulk}{width=\textwidth}{Example 0: Bulk modulus perturbation for three
  layer example. Max excursion = 0.4 GPa.}

\plot{boxborn37}{width=\textwidth}{Example 0: Born simulated free surface traces,
  center midpoing 37 of 76, 
  (integrated) [2.5,5.0, 35, 40] Hz bandpass filter source.}

\plot{boxfreeinvp1it0import}{width=\textwidth}{Example 0: Asymptotic extended inversion of
  free surface data.}

\plot{boxfreerebornp1it0shot37}{width=\textwidth}{Example 0: Resimulation of data
  in Figure \ref{fig:boxborn37}.}

\plot{boxfreerebornp1it0diff37}{width=\textwidth}{Example 0: Data residual for shot
  depicted in Figure \ref{fig:boxborn37}.}

\plot{boxfreeinvp1it0stack}{width=\textwidth}{Example 0: Asymptotic extended
  inversion of free surface data, stacked over offset - best estimate
  of bulk modulus perturbation.}

\plot{boxcomp150}{width=\textwidth}{Example 0: comparison of central
  depth trace, blue (target) vs. red (asymptotic inversion).}

\newpage

\section{Example 1: Marmousi LSM, Coarse Grid}

The aim in this example and in examples 2 and 4 is to use the
zero-offset restriction of the approximate inverse \ref{eqn:fsadj} to
accelerate (ordinary, non-extended) least squares migration (LSM). The
example is based on the Marmousi model \cite[]{BoLaVe:91} treated as a
constant density acoustic medium, $\rho = 1$. The bulk modulus is shown
in Figure \ref{fig:coarse1bulk}.

\plot{coarse1bulk}{width=\textwidth}{Example 1: Bulk modulus derived from 
  Marmousi model, with $rho = 1$ - that is, velocity 
  squared. Subsampled by factor of 6 in both directions, so that $dx =
  dz = 24$ m.}

The scale division used in this experiment is displayed in Figures
\ref{fig:coarse1bulkbig} and \ref{fig:coarse1dbulk}. The smoothed bulk
modulus displayed in Figure \ref{fig:coarse1bulkbig} is the result of
4 repetitions of 6 $\times $ 6 box averaging of the subsampled
Marmousi model (Figure \ref{fig:coarse1bulk}) on the 24 m grid. The
  bulk modulus perturbation (Figure \ref{fig:coarse1dbulk}) is simply
  the difference.

\plot{coarse1bulkbig}{width=\textwidth}{Example 1: Smooth background bulk modulus
  resulting from 6 $\times$ 6 box averaging of the field in Figure
  \ref{fig:coarse1bulk}, repeated 4 times.}

\plot{coarse1dbulk}{width=\textwidth}{Example 1: Difference between unsmoothed
  and smoothed bulk moduli, used as bulk modulus perturbation to
  generate Born data for this experiment series.}

This model is used to simulate 90 shots spaced 96 m apart, with 190
receivers in a fixed (shot-independent) array. Shot and receiver
depths are 24 m. The acquisition geometry is summarized in the output
of {\tt surange}:
\begin{verbatim}
17100 traces:
tracl    1 17100 (1 - 17100)
tracr    1 17100 (1 - 17100)
fldr     0 89 (0 - 89)
tracf    1 190 (1 - 190)
offset   -8760 8856 (-216 - 312)
gelev    -24
selev    -24
sx       240 8784 (240 - 8784)
gx       24 9096 (24 - 9096)
ns       501
dt       8000

Shot coordinate limits:
        North(240,0) South(240,0) East(8784,0) West(240,0)

Receiver coordinate limits:
        North(24,0) South(24,0) East(9096,0) West(24,0)

Midpoint coordinate limits:
        North(132,0) South(132,0) East(8940,0) West(132,0)
\end{verbatim}

Sources and receivers are point and isotropic. The source wavelet is
is a once-integrated bandpass filter with corner frequencies 2.5, 5,
12.5, and 15 Hz. The pressure (source) field is a
bandpass-filtered Green's function.

The source wavelet is scaled by the value of the bulk modulus near the
source positions, that is, in the water column, to compensate for the
lack of such scaling in the IWAVE acoustic staggered grid
implementation: otherwise, the computed fields would not correspond to
solutions of equation \ref{eqn:awe}.

The surface $z=0$ is pressure-free, that is, the boundary condition is
$p=0$ on this surface.

We use a very conventional staggered-grid finite difference method of
second order in time and 8th order in space to simulate all wavefields
contributing the results reported here \cite[]{Vir:84,Vir:86}.

Shot gather 45, at source position 4560 m, is displayed in Figure
\ref{fig:coarse1bornr45}.

\plot{coarse1bornr45}{width=\textwidth}{Example 1: Shot gather 45, $x_s$ = 4560
  m. }

Migration (Reverse Time Migration = RTM) of this data with yields the
image displayed in Figure \ref{fig:coarse1mbulk}. Our implementation
of RTM has a couple of features worthy of note:
\begin{itemize}
\item it realizes the {\em adjoint} or {\em transpose} of Born
  modeling. This choice differs from some more conventional choices of
  RTM in filtering and scaling. We verify the adjointness relation
  between RTM and Born modeling operators by the well-known {\em dot
    product test} \cite[]{claer:PVI} which passes at a low multiple of
  machine (single) precision.
\item For reconstruction of the source wavefield, we use the optimal
  checkpointing method
  \cite[]{Griewank:92,Griewank:00,Griewank:book,Symes:06a-pub}, which
  has the virtue of achieving the adjoint relation to machine
  precision for virtually any time-domain simulation method, but the
  drawback of somewhat more computational complexity than would be
  required for some specialized techniques suited for special systems
  such as our acoustic problem. Generally the cost of RTM via optimal
  checkpointing is roughly 4 -5 times the cost of forward Born
  modeling for problems with $O(10000)$ time steps, such as those
  solved here.
\end{itemize}

Several aspects are immediately apparent from comparison with the of
the RTM image \ref{fig:coarse1mbulk} with the target image, that is,
the actual bulk modulus perturbation used to generate the data, see
Figure \ref{fig:coarse1dbulk}:
\begin{itemize}
\item locations and orientation of events in the image
  \ref{fig:coarse1mbulk} are correct, up to limits imposed by the bandlimited
  nature of the image, which follows from the bandlimited nature of
  the data.
\item the overal amplitude is grossly incorrect, and the relation
  between reflector strengths in various parts of the image is also
  much different: in particular, the deeper part of the image is much
  fainter than the shallow part. This feature follows from the uneven
  illumination of the subsurface: both the fall-off of source
  wavefield and receiver wavefield amplitude with distance from the
  surface, and the focusing and defocusing of these wavefields as they
  traverse the heterogeneous refractive velocity model, contribute to
  this effect.
\item some evidence of transmitted wavefield is present: the
  low-frequency swings visible in the image are due to refracted
  (diving) wave components in source and receiver field contributing
  to image amplitude along the entire diving wave ray path, not just
  at a reflector location. These signal components are actually quite
  important, and are the principal contribution to FWI. However in
  reflector imaging they are often regarded as noise.
\end{itemize}

\plot{coarse1mbulk}{width=\textwidth}{Example 1: Reverse time migration of Born
  data. Notice the uneven illumination, and the faint but visible
  low spatial frequency diving wave artifacts.}

LSM is the name acquired in recent years by the data fitting problem:
given data (regarded as) resulting from a Born or perturbational
model, estimate the corresponding model perturbation by minimization
the mean-square misfit between predicted and given Born data. This
task defines a linear inverse problem, and an equally good (and older)
synonym is ``linearized inversion''. It has long been recognized that
reducing or minimizing data misfit tends to resolve the illumination
and amplitude defects of migrated images, noted above
\cite[]{BourJiaLai:89}.  Data misfit minimization is commonly
implemented via conjugate gradient iteration or an equivalent
\cite[]{Nemeth:99}. Each iteration requires one Born forward modeling
operation and one (variant of) Reverse Time Migration.

The 2017 SEG Annual International Meeting and Exposition featured
several sessions on LSM, including elastic LSM for estimation of shear
properties, acoustic LSM as is demonstrated here, viscoelastic LSM for
correct estimation of amplitudes in the presence of attenuation and/or
attentuation parameters, and a variety of case studies.

We use Born data (one shot of which was just displayed) as target data
in our LSM exercise. Twenty iterations of conjugate gradient iteration
produces the estimate of the bulk modulus perturbation displayed in
FIgure \ref{fig:coarse1bulkinvp0it20}.

\plot{coarse1bulkinvp0it20}{width=\textwidth}{Example 1: Estimate of bulk modulus
  perturbation resulting from least squres migration, 20 conjugate
  gradient iterations. Plotted on same color scale as target bulk
  modulus perturbation (Figure \ref{fig:coarse1dbulk}).}

LSM has indeed balanced the amplitudes throughout the image: the
relative amplitudes between various parts are much closer to those of
the target image \ref{fig:coarse1dbulk}. The overall amplitude is also
much closer to correct, so one to some extent regard the image point
values as local and band-limited estimates of physical
parameter (perturbation in bulk modulus). The ampltudes are generally
low, partly because of the spatial averaging effect of bandlimitation,
partly because further iterations would decrease data residual and
recover model parameters more accurately.

LSM achieves its aim by presuming that achieving data fit will result
in a more accurate image of the subsurface. Figure
\ref{fig:coarse1rebornp0it20r45} shows the predicted data for
shot 45 based on the bulk modulus perturbation in Figure
\ref{fig:coarse1bulkinvp0it20}. These traces indeed closely resemble
the target data \ref{fig:coarse1bornr45}; the difference (residual
shot record) appears as \ref{fig:coarse1residrebornp0it20r45}. The RMS
fit level achieved is part of the information displayed in the
following table:

\begin{verbatim}
========================== BEGIN CGNE =========================
Iteration   |  Residual Norm |  Gradient Norm
         1    2.50883907e-01   2.06906028e-04
         2    2.11527959e-01   1.23586302e-04
         3    1.93715423e-01   1.00169396e-04
         4    1.79163083e-01   9.69018802e-05
         5    1.64154008e-01   9.32183611e-05
         6    1.47441313e-01   8.68227726e-05
         7    1.32597432e-01   7.44872232e-05
         8    1.20940886e-01   6.83080798e-05
         9    1.11524783e-01   5.57000349e-05
        10    1.03391834e-01   5.03123520e-05
        11    9.70416740e-02   4.44237339e-05
        12    9.14672911e-02   4.11237088e-05
        13    8.63379091e-02   3.77963879e-05
        14    8.15358013e-02   3.53451178e-05
        15    7.70760477e-02   3.34034703e-05
        16    7.26942420e-02   3.14817880e-05
        17    6.85755685e-02   3.00511110e-05
        18    6.44731000e-02   2.88680549e-05
        19    6.06474578e-02   2.66172028e-05
        20    5.71314394e-02   2.49373679e-05
-----------------------------------------------
        21    5.40441014e-02   2.32873826e-05
=========================== END CGNE ==========================

 ****************** CGNE summary *****************  
initial residual norm      = 2.50883907e-01
residual norm              = 5.40441014e-02
residual redn              = 2.15414777e-01
initial gradient norm      = 2.06906028e-04
gradient norm              = 2.32873826e-05
gradient redn              = 1.12550527e-01
\end{verbatim}
Evidently convergence is slow but sure. The left-hand column lists the
data residual, which is (for mathematically consistent implementations
of the conjugate gradient algorithm) guaranteed to decrease at each
step. The right-hand column is the gradient length, or normal
residual, whcih also converges to zero but not (necessarily)
monotnically.

The reduction in data misfit achieved over 20 iterations is
approximately 21.5\%.

\plot{coarse1rebornp0it20r45}{width=\textwidth}{Example 1: Predicted data from 
  model of Figure \ref{fig:coarse1bulkinvp0it20}.}

\plot{coarse1residrebornp0it20r45}{width=\textwidth}{Example 1: Data residual from 
  model of Figure \ref{fig:coarse1bulkinvp0it20}. RMS error is
  approximately 21.5\%.}

We can reach two tentative conclusions from this example:
\begin{itemize}
\item data fitting, that is, LSM, does indeed tend to extract image 
  amplitudes so that they resemble those in the ``earth'', hence
  incidentally correct uneven illumination;
\item convergence can be slow, and the premium in computational cost
  over RTM is a large factor (20X, in this example).
\end{itemize}

The cost of LSM is a major impediment to its use. To speed it up,
either each iteration can be made less expensive, or the number of
necessary iterations can be reduced. We use the approximate
inverse constructed in the preceeding sections to accelerate the
convergence of the LSM loop. Note that in this section we are dealing
with a physical or non-extended model. The second form of the
approximate inverse  
involves a model weight operator that acts diagonally on the various
offset sections of the extended model, and in particular on the
zero-offset section, which is the result of conventional RTM such as
the image displayed above. There is no reason to think particularly
that the zero-offset section of the approximate inverse output is an
approximate inverse of the zero-offset (physical) problem, however we
observe that indeed the amplitudes are considerably
improved over basic RTM, so we may attempt to use it as a
preconditioner.

We explain the link between the zero-offset weighted adjoint and a
{\em bona fide} approximate inverse in the discussion of Example 3.

The preconditioned conjugate gradient iteration described earlier is
applied to estimate $\delta \kappa$. The algorithm approximates a
minimizer of the least-squares objective
\begin{equation}
\label{eqn:elsmex1}
\|I_t\dF \delta \kappa -I_t\delta d\|^2_d = \langle (I_t \doF \delta \kappa \delta(h) -
I_d \delta d) W_d^0 (I_t \doF \delta \kappa \delta(h)- I_d \delta d)\rangle). 
\end{equation}
The expression on the RHS uses the extension relation $\dF \delta
\kappa = \doF \delta \kappa \delta(h)$. The code implements $\dF$
through this relation, with care to avoid inefficiency. The data
weight operator $W_d^0$ is defined in \ref{eqn:wddeffs}. As pointed
out above, the PCG algorithm does not require explicit computation of
$W_d^0$, a good thing as it would be quite expensive. THe expression
\ref{eqn:elsmex1} emphasizes that the data residual is minimized in
the sense of the weighted norm, not in the sense of the unweighted
($L^2$) norm. Since we do not compute the weighted norm explicitly,
the implemented PCG algorithm displays records the $L^2$ norm, rather
than the weighted norm, and the former is not guaranteed to decrease
monotonically.

Application of the asymptotic inverse \ref{eqn:appinvfs}, in the
convenient form \ref{eqn:fsadj} provided by the shallow tow depth hypothesis,  yields the image displayed in Figure
\ref{fig:coarse1bulkinvp1it0}. This image has the same overall
amplitude balance as the 20-iteration LSM image
\ref{fig:coarse1bulkinvp0it20}, but results from a single migration
pass, is less noisy, and has even more
correct relative amplitudes. On the other hand the overall amplitude
is roughly 4X too small, and the data residual is still large - around 80\% in
fact. However if you scale the image by a factor of 4, the difference
drops to less than 50\%, indicating that the approximate inverse
zero section \ref{fig:coarse1bulkinvp1it0} may be a good search
direction. We will explain why this is so in the discussion of Example 3.

\plot{coarse1bulkinvp1it0}{width=\textwidth}{Example 1: Approximate inversion or
  true amplitude image = zero offset section of subsurface offset
  approximate inversion. Plotted on same color scale as target bulk
  modulus perturbation in Figure \ref{fig:coarse1dbulk}.}

Ten iterations of PCG produces the approximate inversion in Figure
\ref{fig:coarse1bulkinvp1it10}. The amplitudes are roughly correct,
allowing for the spatial averaging influence of band limitation, and the
resolution improved somewhat over that displayed in Figure
\ref{fig:coarse1bulkinvp1it0}. More than that, this is a good
inversion: the RMS error is less than 10\%. The output of the RVL CGNE
tool describes the convergence:

\begin{verbatim}
========================== BEGIN PCGNE =========================
Iteration   |  Residual Norm |  Gradient Norm
         1    2.50883907e-01   6.89441109e+00
         2    1.18942171e-01   2.66392207e+00
         3    7.03184232e-02   1.10846794e+00
         4    5.15148342e-02   6.83876038e-01
         5    4.13431711e-02   4.61543351e-01
         6    3.50607708e-02   3.48029107e-01
         7    3.10531463e-02   2.71396458e-01
         8    2.83174179e-02   2.21747950e-01
         9    2.63601393e-02   1.83933064e-01
        10    2.49932427e-02   1.60866663e-01
-----------------------------------------------
        11    2.39612553e-02   1.46033272e-01
=========================== END PCGNE ==========================

 ****************** CGNE summary *****************  
initial residual norm      = 2.50883907e-01
residual norm              = 2.39612553e-02
residual redn              = 9.55073461e-02
initial gradient norm      = 6.89441109e+00
gradient norm              = 1.46033272e-01
gradient redn              = 2.11813990e-02
\end{verbatim}

The predicted data shot 45 is displayed in Figure
\ref{fig:coarse1rebornp1it10r45}, plotted on the same scale as Figure
\ref{fig:coarse1bornr45} - it is hard to see a difference between the
two. The residual practically disappears (Figure \ref{fig:coarse1residrebornp1it10r45}).

\plot{coarse1bulkinvp1it10}{width=\textwidth}{Example 1: Estimated bulk modulus
  perturbation from ten precondiitioned CG iterations. Compare to
  Figure \ref{fig:coarse1dbulk}.}

\plot{coarse1rebornp1it10r45}{width=\textwidth}{Example 1: Predicted shot record 45 from
  inversion displayed in Figure \ref{fig:coarse1bulkinvp1it10} -
  compare with Figure \ref{fig:coarse1bornr45}, which shows target
  data plotted on same grey scale, or with next figure, which
  replicates Figure \ref{fig:coarse1bornr45} for the reader's convenience.}

\plot{coarse1bornr45bis}{width=\textwidth}{Example 1: Replication of Figure \ref{fig:coarse1bornr45} for
convenient comparison with previous figure. Note that the only readily
visible difference is the appearance of water column energy in the
resimulated data (Figure \ref{fig:coarse1rebornp1it10r45}), absent in
the target data (Figure \ref{fig:coarse1bornr45}).}

\plot{coarse1residrebornp1it10r45}{width=\textwidth}{Example 1: Residual =
  difference of data displayed in previous two figures, plotted on
  same grey scale.}

These results suggest the following conclusions: 
\begin{itemize}
\item The zero-offset restriction of the asymptotic 
  inversion operator \ref{eqn:appinvfs} (implemented via \ref{eqn:fsadj}) is an effective preconditioner for LSM: it gives a initial
  search direction and continues to accelerate CG in later
  iterations. Its deficient overall amplitude is implicitly corrected
  by the CG step length calculation.
Convergence is much faster with than without preconditioning: the
preconditioned algorithm takes 4 iterations to reach (or better) the
residual obtained by the unpreconditioned algorithm in 20 iterations.
\item Good RMS convergence does not guarantee convergence of the
  estimated bulk modulus (or other parameter) to the target model {\em
    pointwise} -
  only to within the resolution of the data bandwidth, which affects
  estimated amplitudes via spatial averaging. 
\end{itemize}

\newpage
\section{Example 2: Marmousi LSM, Fine Grid}

\inputdir{project}

This section replicates the results of the preceding section with
higher frequency data and finer spatial grid. The Marmousi bulk
modulus on this 12 m $\times$ 12 m grid is shown 
in Figure \ref{fig:fine1bulk}.

\plot{fine1bulk}{width=\textwidth}{Example 2: Bulk modulus derived from 
  Marmousi model, with $rho = 1$ - that is, velocity 
  squared. Subsampled by factor of 6 in both directions, so that $dx =
  dz = 12$ m.}

The scale division used in this experiment is displayed in Figures
\ref{fig:fine1bulkbig} and \ref{fig:fine1dbulk}. The smoothed bulk
modulus displayed in Figure \ref{fig:fine1bulkbig} is the result of
4 repetitions of 6 $\times $ 6 box moving average of the subsampled
Marmousi model (Figure \ref{fig:fine1bulk}) on the 12 m grid. The
  bulk modulus perturbation (Figure \ref{fig:fine1dbulk}) is the
  difference between bulk moduli corresponding to the Marmousi model
  and the result of 2 repititions of 3 $\times$ 3 box moving average.

\plot{fine1bulkbig}{width=\textwidth}{Example 2: Smooth background bulk modulus
  resulting from 6 $\times$ 6 box averaging of the velocity
  corresponding to the field in Figure
  \ref{fig:fine1bulk}, repeated 4 times.}

\plot{fine1dbulk}{width=\textwidth}{Example 2: Difference between base bulk modulus (Fig
  \ref{fig:fine1bulk} and the bulk modulus obtained by a milder
  smothing (3 $\times$ 3 moving average repeated twice) of the
  Marmousi velocity field. Used as bulk modulus perturbation to
  generate Born data for this experiment series.}

This model is used to simulate 180 shots spaced 48 m apart, with 380
receivers in a fixed (shot-independent) array. Shot and receiver
depths are 12 m. The acquisition geometry is summarized in the output
of {\tt surange}:
\begin{verbatim}
68760 traces:
tracl    1 68760 (1 - 68760)
tracr    1 68760 (1 - 68760)
fldr     0 179 (0 - 179)
tracf    1 382 (1 - 382)
offset   -8820 8916 (-228 - 324)
gelev    -12
selev    -12
sx       240 8832 (240 - 8832)
gx       12 9156 (12 - 9156)
ns       1001
dt       4000

Shot coordinate limits:
        North(240,0) South(240,0) East(8832,0) West(240,0)

Receiver coordinate limits:
        North(12,0) South(12,0) East(9156,0) West(12,0)

Midpoint coordinate limits:
        North(126,0) South(126,0) East(8994,0) West(126,0)
\end{verbatim}

Sources and receivers are point and isotropic. The source wavelet is
is a once-integrated bandpass filter with corner frequencies 2.5, 5,
25, and 30 Hz. The pressure (source) field is a
bandpass-filtered Green's function.

The surface $z=0$ is pressure-free, that is, the boundary condition is
$p=0$ on this surface.

Shot gather 90, at source position 4560 m, is displayed in Figure
\ref{fig:fine1bornr90}.

\plot{fine1bornr90}{width=\textwidth}{Example 2: Shot gather 90, $x_s$ = 4560
  m. }

RTM once again produces an image with major reflectors correctly
positioned, but with badly unbalanced amplitudes and low-frequency
diving wave artifacts (``rabbit ears'' - in FWI applications not
useful, but in fact completely useless here, as the velocity model is
already precisely correct).

\plot{fine1mbulk}{width=\textwidth}{Example 2: Reverse time migration of Born
  data. Notice the uneven illumination, and the clearly visible
  low spatial frequency diving wave artifacts.}

The additional bandwidth appears to make this problem more difficult
(more ill-conditioned) than the problem discussed in the last section.
LSM via twenty iterations of conjugate gradient iteration
produces the estimate of the bulk modulus perturbation displayed in
FIgure \ref{fig:fine1bulkinvp0it20}.

\plot{fine1bulkinvp0it20}{width=\textwidth}{Example 2: Estimate of bulk modulus
  perturbation resulting from least squres migration, 20 conjugate
  gradient iterations. Plotted on same color scale as target bulk
  modulus perturbation (Figure \ref{fig:fine1dbulk}).}

LSM has partly balanced the amplitudes throughout the image: the
relative amplitudes between various parts are much closer to those of
the target image \ref{fig:fine1dbulk}, but not as close as the
analogous result from the narrow-band low frequency roblem, Figure
\ref{fig:coarse1bulkinvp0it20}. Figure
\ref{fig:fine1rebornp0it20r90} shows the predicted data for
shot 90 based on the bulk modulus perturbation in Figure
\ref{fig:fine1bulkinvp0it20}. These traces resemble
the target data \ref{fig:fine1bornr90}, but not as closely as the
analogous comparison of the lower-frequency data. The difference (residual
shot record) appears as \ref{fig:fine1residrebornp0it20r90}. The RMS
fit level achieved is part of the information displayed in the
following table:

\begin{verbatim}
========================== BEGIN CGNE =========================
Iteration   |  Residual Norm |  Gradient Norm
         1    1.32567376e-01   9.79450924e-05
         2    1.20665237e-01   1.74745495e-04
         3    1.10067330e-01   1.08682645e-04
         4    1.05383545e-01   7.54519788e-05
         5    1.00764051e-01   6.28481284e-05
         6    9.69513506e-02   6.03487533e-05
         7    9.27119926e-02   4.58884570e-05
         8    8.90481845e-02   4.39115865e-05
         9    8.59621167e-02   4.09871645e-05
        10    8.23227838e-02   4.35815928e-05
        11    7.92605504e-02   4.27014675e-05
        12    7.76133314e-02   5.97550788e-05
        13    7.50623047e-02   4.16568146e-05
        14    7.14619383e-02   3.48134454e-05
        15    6.85999393e-02   3.68092878e-05
        16    6.53088316e-02   3.76127427e-05
        17    6.37752339e-02   3.79335834e-05
        18    6.12660609e-02   3.35733057e-05
        19    5.85174002e-02   2.77393119e-05
        20    5.61746396e-02   2.67642426e-05
-----------------------------------------------
        21    5.39411120e-02   2.79296019e-05
=========================== END CGNE ==========================

 ****************** CGNE summary *****************  
initial residual norm      = 1.32567376e-01
residual norm              = 5.39411120e-02
residual redn              = 4.06895816e-01
initial gradient norm      = 9.79450924e-05
gradient norm              = 2.79296019e-05
gradient redn              = 2.85155714e-01

\end{verbatim}
Evidently convergence is somewhat slower than before: the final data
misfit achieved after 20 iterations is approximately 40\%.

\plot{fine1rebornp0it20r90}{width=\textwidth}{Example 2: Predicted data from 
  model of Figure \ref{fig:fine1bulkinvp0it20}.}

\plot{fine1residrebornp0it20r90}{width=\textwidth}{Example 2: Data residual from 
  model of Figure \ref{fig:fine1bulkinvp0it20}. RMS error is
  approximately 40.7\%.}

\plot{fine1comp}{width=\textwidth}{Example 2: Relative mean square 
  error vs. iteration for CG (red curve) and PCG (blue curve).}

\plot{fine1compnt}{width=\textwidth}{Example 2: Relative mean square 
  error vs. iteration for CG (red curve) and PCG (blue curve), without
  depth tapering in water layer.}

Application of the asymptotic inverse (zero offset section of
\ref{eqn:appinvfs}) yields the image displayed in Figure
\ref{fig:fine1bulkinvp1it0}. This image better relative amplitudes than the 20-iteration LSM image
\ref{fig:fine1bulkinvp0it20}, and is less noisy, but the overall
amplitude is quite small compared to the target bulk modulus
perturbation (Figure \ref{fig:fine1dbulk}). As was true for the lower
frequency example, the good relative amplitudes and reflector
waveforms  indicates that the approximate inverse
zero section \ref{fig:fine1bulkinvp1it0} may be a good search
direction.

\plot{fine1bulkinvp1it0}{width=\textwidth}{Example 2: Approximate inversion or
  true amplitude image = zero offset section of subsurface offset
  approximate inversion. Plotted on same color scale as target bulk
  modulus perturbation in Figure \ref{fig:fine1dbulk}.}

Twenty iterations of preconditioned
CG produces the approximate inversion in Figure
\ref{fig:fine1bulkinvp1it20}. Comparison with Figure
\ref{fig:fine1dbulk} suggests that the amplitudes and waveforms are
roughly correct,
allowing for the spatial averaging influence of band limitation, and the
resolution improved somewhat over that displayed in Figure
\ref{fig:fine1bulkinvp1it0}. More than that, this is a good
inversion: the RMS error is less than 10\%. The output of the RVL CGNE
tool describes the convergence:

\begin{verbatim}
========================== BEGIN PCGNE =========================
Iteration   |  Residual Norm |  Gradient Norm
         1    1.32567376e-01   2.19551253e+00
         2    8.64119902e-02   8.03773761e-01
         3    6.29559085e-02   4.04524982e-01
         4    5.03380559e-02   2.41569892e-01
         5    4.18944918e-02   1.65786415e-01
         6    3.59481946e-02   1.23230457e-01
         7    3.13160010e-02   9.72592682e-02
         8    2.77100820e-02   7.86418393e-02
         9    2.49173287e-02   6.50660247e-02
        10    2.26768292e-02   5.65260053e-02
        11    2.08535902e-02   4.91942577e-02
        12    1.93884857e-02   4.33817171e-02
        13    1.81311555e-02   3.99240106e-02
        14    1.70452520e-02   3.66419256e-02
        15    1.61075015e-02   3.42206433e-02
        16    1.53041314e-02   3.18240337e-02
        17    1.46190701e-02   3.01125068e-02
        18    1.40266838e-02   2.82831118e-02
        19    1.35276383e-02   2.65362691e-02
        20    1.30786821e-02   2.57860348e-02
-----------------------------------------------
        21    1.26670329e-02   2.50269361e-02
=========================== END PCGNE ==========================

 ****************** CGNE summary *****************  
initial residual norm      = 1.32567376e-01
residual norm              = 1.26670329e-02
residual redn              = 9.55516621e-02
initial gradient norm      = 2.19551253e+00
gradient norm              = 2.50269361e-02
gradient redn              = 1.13991313e-02

\end{verbatim}

The predicted data shot 90 is displayed in Figure
\ref{fig:fine1rebornp1it20r90}, plotted on the same scale as Figure
\ref{fig:fine1bornr90} - it is hard to see a difference between the
two. The residual practically disappears (Figure \ref{fig:fine1residrebornp1it20r90}).

\plot{fine1bulkinvp1it20}{width=\textwidth}{Example 2: Estimated bulk modulus
  perturbation from 20 precondiitioned CG iterations. Compare to
  Figure \ref{fig:fine1dbulk}.}

\plot{fine1rebornp1it20r90}{width=\textwidth}{Example 2: Predicted shot record 90 from
  inversion displayed in Figure \ref{fig:fine1bulkinvp1it20} -
  compare with Figure \ref{fig:fine1bornr90}, which shows target
  data plotted on same grey scale, or with next figure, which
  replicates Figure \ref{fig:fine1bornr90} for the reader's convenience.}

\plot{fine1bornr90bis}{width=\textwidth}{Example 2: Replication of Figure \ref{fig:fine1bornr90} for
convenient comparison with previous figure. Note that the only readily
visible difference is the appearance of water column energy in the
resimulated data (Figure \ref{fig:fine1rebornp1it20r90}), absent in
the target data (Figure \ref{fig:fine1bornr90}).}

\plot{fine1residrebornp1it20r90}{width=\textwidth}{Example 2: Residual =
  difference of data displayed in previous two figures, plotted on
  same grey scale.}

These results support the same conclusions as in the preceding
section. The approximate inverse provides an effective
preconditioner, in fact the data residual of the preconditioned loop is
smaller at 4 iterations than the data residual of the
un-preconditioned loop after 20.

\newpage
\section{Example 3: Marmousi Extended LSM}
The setting and parameters for this example are exactly as in the
last,  except that the bulk modulus perturbation is extended in a
subsurface offset interval of [-125 m, 125 m], or 10 grid points on
either side of zero offset. The free surface data is precisely the
same as in the previous example (shot 90 shown in Figure \ref{fig:fine1bornr90bis}).

The asymptotic inverse operator $\iddoFf$ produces the $\delta
\bar{\kappa}$ shown in Figure \ref{fig:fine10bulkinvp1it0}. Recall
from Part 1, Theorem 3, that the a left inverse for the physical
modeling operator may be extracted from $\delta \bar{\kappa} $ by
stacking over subsurface offset. The stack in Figure
\ref{fig:fine10bulkinvp1it0stack} is plotted on the same color scale
as the bulk modulus (Figure \ref{fig:fine1dbulk}.

\plot{fine10bulkinvp1it0}{width=\textwidth}{Example 3: Extended bulk
  modulus estimate, application of asymptotic inverse operator to data
  of Example 2.}

\plot{fine10bulkinvp1it0stack}{width=\textwidth}{Example 3: Stack of extended
  bulk modulus inversion - approximate left inverse of physical
  forward modeling operator.}

\plot{fine10bulkinvp1it0stack0}{width=\textwidth}{Example 3: Zero
  offset section of the extended asymptotic inversion. Note that color
scale differs by factor of 4 from that in Figure
\ref{fig:fine10bulkinvp1it0stack}.}

\plot{fine10bulkinvp1it0g6000}{width=\textwidth}{Example 3: Subsurface
  offset gather at position $x_s=$ 6000 m indicated in Figure
  \ref{fig:fine10bulkinvp1it0}, plotted on same grey scale. Note apparent concentration of energy
within 50 m of $h=0$.}

\plot{fine10bulkinvp1it0g6000rat}{width=\textwidth}{Example 3: Subsurface
  offset gather at position $x_s=$ 6000 m, depth range 2000 m $<z<$
  3200 m, plotted to 1:1 aspect ratio
  and padded with zeroes to left and right to obtain square
  region. Plotted on same grey scale as previous figure. Note that extend of energy spread is approximately one
  apparent wavelength.}

Note that the similarity of FIgures \ref{fig:fine10bulkinvp1it0stack}
and \ref{fig:fine1bulkinvp1it0} - the two differ mostly by scale, with
the latter being an order of magnitude smaller. In fact, the
zero-offset asymptotic inverse shown in Figure \ref{fig:fine1bulkinvp1it0} is
exactly the same as the zero offset section of the extended inversion
(Figure \ref{fig:fine10bulkinvp1it0stack0}. An explanation for the deficient
amplitude of the zero-offset asymptotic inversion is now evident: the
stack has the proper amplitude, but it is clear from Figure
\ref{fig:fine10bulkinvp1it0} that the energy in the image is spread
over a number of offsets, exactly how many depending on depth.

The extent of the spread depends on the ray geometry (and on the surface
offset range, as will become evident by comparison with Example
5). Figure \ref{fig:fine10bulkinvp1it0g6000} shows the subsurface
offset image gather at $x=$ 6000 m: the image energy appears to
concentrate within a few grid cells of $h=$0. To visualize the extent
of this concentraation more clearly, Figure
\ref{fig:fine10bulkinvp1it0g6000rat} displays the part of the same
data between $z=$ 2000 and 3200 m, with enough zero padding on the
sides that the plot has 1:1 aspect ratio. Evidently the energy in this
gather is maximally focused, in view of Heisenberg's principle: the
width of the corridor containing most of the energy is about a
wavelength. It has been speculated that image gather energy would
focus in a Fresnel zone, but that is clearly not correct, at least for
a full aperture example such as this one.

In any case the amplitude of each offset section near zero offset is
more or less the same as any other as one sees from Figure
\ref{fig:fine10bulkinvp1it0g6000}, so the stack multiplies the
amplitude, and by the theory developed in part 1 produces the
``proper'' amplitude. Therefore the amplitudes of the indivitual
offset sections, in particular zero offset, must be a fraction of this
``proper'' amplitude. In the middle part of the gather, the two grid
cells on either side of zero offset contain the bulk of the energy, so
the multiplier would be roughly 5 (in view of the maximal focusing,
this is the same as the number of spatial grid points per wavelength
at peak frequency, which indeed is roughly 5 in this example). The
extent of the energy spread is not constant, and therefore the
amplitude of the zero-offset inversion is not precisely a constant
multiple of the target amplitude, but it varies slowly with depth. So
the asymptotic zero-offset inverse acts as a slowly varying scaling of
the target bulk modulus perturbation hence is usable as as
preconditioner, even though it is not itself an accurate approximate
inversion - as we have seen.

On the other hand, we have shown that \ref{eqn:appinvfs} is an
asymptotic inverse for the extended forward map $\iddoFf$, provided
that most of the energy in the image volume is focused near $h=0$, as
it appears to be in this example. Therefore PCG should converge
rapidly, to a data-fitting extended model. Figure
\ref{fig:fine10bulkinvp1it20} displays the result of 20 PCG
iterations. As was the case for the zero-offset verson of this
example, the data residual is roughly 7.5\%, and the stack (Figure
\ref{fig:fine10bulkinvp1it20stack} closely approximates the target
bulk modulus perturbation (Figure \ref{fig:fine1dbulk}), allowing for
resolution limitation due to limited bandwidth. In fact, it differs
little from the zero-offset PCG result \ref{fig:fine1bulkinvp1it20}
and therefore also generates a data misfit via physical modeling of
less than 10%.

\plot{fine10bulkinvp1it20}{width=\textwidth}{Example 3: Extended bulk
  modulus estimate, estimated from 20 PCG iteratoins from data
  of Example 2.}

\plot{fine10bulkinvp1it20stack}{width=\textwidth}{Example 3: Stack of extended
  bulk modulus inversion from 20 PCG iterations - near-exact left inverse of physical
  forward modeling operator.}

Inversion, rather than asymptotic inversion, also refines the focusing
of energy in the image volume. Figure
\ref{fig:fine10bulkinvp1it20g6000} shows the gather at $x=$ 6000 m
from the volume depicted in Figure \ref{fig:fine10bulkinvp1it20}. The
additional focusing is even more evident in the 1:1 aspect ratio plot
(Figure \ref{fig:fine10bulkinvp1it20g6000rat}), constructed in the same
way as Figure \ref{fig:fine10bulkinvp1it20g6000rat}.

\plot{fine10bulkinvp1it20g6000}{width=\textwidth}{Example 3: Subsurface
  offset gather of 20 PCG iteration inversion at position $x_s=$ 6000 m indicated in Figure
  \ref{fig:fine10bulkinvp1it20}, plotted on same grey scale. Note apparent concentration of energy
within 50 m of $h=0$.}

\plot{fine10bulkinvp1it20g6000rat}{width=\textwidth}{Example 3:
  Subsurface offset gather of 20 PCG iteration inversion at position
  $x_s=$ 6000 m, depth range 2000 m $<z<$ 3200 m, plotted to 1:1
  aspect ratio and padded with zeroes to left and right to obtain
  square region. Plotted on same grey scale as previous figure. Note
  that extent of energy spread is very close to one apparent
  wavelength.}

\newpage
\section{Example 4: Marmousi LSM, Towed Streamer Geometry}
This example differs from the preceding three in that it uses towed
streamer, rather than full aperture, acquisition geometry. The
parameters are as
specified by IFP in the original 1989 data distribution
\cite[]{BoLaVe:91}, with only one change: the shot points are subsampled by a factor of 4, leaving 60
shots with 100 m interval. The output of surange gives the details:
\begin{verbatim}
5760 traces:
tracl    1 22752 (1 - 22752)
tracr    658 23409 (658 - 23409)
fldr     3000 8900 (3000 - 8900)
tracf    1 96 (1 - 96)
cdp      1 568 (1 - 568)
cdpt     1 48 (1 - 1)
trid     1
duse     1
offset   -2575 -200 (-2575 - -200)
gelev    -8
selev    -12
sx       3000 8900 (3000 - 8900)
gx       425 8700 (425 - 8700)
ns       751
dt       4000

Shot coordinate limits:
        North(3000,0) South(3000,0) East(8900,0) West(3000,0)
Receiver coordinate limits:
        North(425,0) South(425,0) East(8700,0) West(425,0)
Midpoint coordinate limits:
        North(1712.5,0) South(1712.5,0) East(8800,0) West(1712.5,0)
\end{verbatim}

The spatial grid is subsampled by a factor of 5 from IFP
model, so that $dz = dx = $ 20 m. 
The density in this example is once again set = 1.0 g/cc. The velocity
is smoothed by by 4 passes of a 6-point rectangular average and
squared to create the background velocity model 
(Figure \ref{fig:marm1bulkbig}). The bulk modulus perturbation is
computed by creating a slightly less smoothed bulk modulus (velocity
smoothed by 2 passes of a 3 point rectangular average, squared) then
subtracting that from the unsmoothed bulk modulus. The perturbation is
displayed in Figure \ref{fig:marm1dbulk}.

\plot{marm1bulkbig}{width=\textwidth}{Example 4: background bulk
  modulus, 20 m grid.}

\plot{marm1dbulk}{width=\textwidth}{Example 4: bulk modulus perturbation.}

The source pulse ($w(t)$ in the
system \ref{eqn:awe}) is a [2.5, 5.0, 15, 20] Hz bandpass filter. The
time integral in $I_t \doF$ is accommodated by trapezoidal rule
integration of this pulse.

Figure \ref{fig:marm1r6000} shows the synthetic at shot position 6000
m.

\plot{marm1r6000}{width=\textwidth}{Example 4: linearized shot gather at $x_s$ =
  6000 m.}

The behaviour of the asymptotic inverse for this non-extended
(zero-offset) example is no different than in previous similar
experiments: it gives a good relative amplitude map, but with small
overall amplitude. It is not in itself a usable inverse, but functions
as a quite successful preconditioner. Ten PCG iterations generate a quite accurate
inversion displayed in Figure \ref{fig:marm1invp1it10}.

The iteration record illustrates the successful convergence
acceleration - we do not show the record for unpreconditioned CG, but
it is considerably slower, as in previous examples.

\begin{verbatim}
========================== BEGIN PCGNE =========================
Iteration   |  Residual Norm |  Gradient Norm
         1    4.29975577e-02   2.53198719e+00
         2    2.50279326e-02   1.15877616e+00
         3    1.54467793e-02   6.48228705e-01
         4    1.08529218e-02   3.65152717e-01
         5    8.42937827e-03   2.28351489e-01
         6    6.79261191e-03   1.60824731e-01
         7    5.62188262e-03   1.16707794e-01
         8    4.78196330e-03   9.19547901e-02
         9    4.19104146e-03   7.66246468e-02
        10    3.71039868e-03   6.24982305e-02
-----------------------------------------------
        11    3.33797187e-03   5.21011576e-02
=========================== END PCGNE ==========================

 ****************** CGNE summary *****************  
initial residual norm      = 4.29975577e-02
residual norm              = 3.33797187e-03
residual redn              = 7.76316598e-02
initial gradient norm      = 2.53198719e+00
gradient norm              = 5.21011576e-02
gradient redn              = 2.05771811e-02

\end{verbatim}

The resimulation at shot location 6000
m is plotted in Figure \ref{fig:marm1rebornp1it10r6000}, and the
residual on the same grey scale in Figure \ref{fig:marm1residp1it10r6000}.

\plot{marm1invp1it10}{width=\textwidth}{Example 4: inverted bulk
  modulus perturbation, 10 PCG iterations.}
\plot{marm1rebornp1it10r6000}{width=\textwidth}{Example 4: resimulation of shot gather at $x_s$ =
  6000 m from inverted bulk modulus perturbation (Figure
  \ref{fig:marm1invp1it10}).}
\plot{marm1residp1it10r6000}{width=\textwidth}{Example 4: data
  residual at $x_s$= 6000 m.}

Evidently the original rather short cable (at least, rather short by
modern standards) produces data more than sufficient for an effective
inversion, provided that the background model is kinematically
consistent with the data.

\newpage

\section{Example 5: Marmousi Extended LSM, Towed Streamer Geometry}
This example bears the same relation to Example 4 that Example 3 did
to Example 2. All parameters are the same, except that the bulk
modulus perturbation is extended, -200 m $< h < $ 200 m, so a total of
21 grid points in $h$ (for computational efficiency, $\Delta h$ =
$\Delta x$).

For this example, we show only the result of PCG inversion with 10
iterations. Figure \ref{fig:marm10freeinvp1it10} shows the extended
inversion for $\delta \bar{\kappa}$, and Figure
\ref{fig:marm10freeinvp1it10stack} the stack (optimal estimate of $\delta
\kappa$). Comparison with Figure \ref{fig:marm1invp1it10} reveals that
the dipping layers on the left of the model are truncated several
hundred meters further to the right in Figure
\ref{fig:marm10freeinvp1it10stack}. This truncation is due to the lack of
energy in image gathers for $h<0$ in the left hand part of the model,
which in turn results from the off-end cable geometry and the slope of
the layer package. Thus image energy available at offset zero in Example
4 is diluted by missing energy in $h<0$ in Example 5, in the left
side of the model.

\plot{marm10freeinvp1it10}{width=\textwidth}{Example 5: PCG extended inversion, 10
  iterations, Marmousi with original streamer geometry, 20 m spatial
  grid, 2.5-5-15-20 Hz bandpass source, free surface, subsurface
  offset axis $-200 < h < 200$ m.}

\plot{marm10freeinvp1it10stack}{width=\textwidth}{Example 5: Stack of data in
  Figure \ref{fig:marm10freeinvp1it10}.}

Figure
\ref{fig:marm10freeinvp1it10g6000} shows the gather at $x=$ 6000 m
from the volume depicted in Figure \ref{fig:marm10freeinvp1it10}. The
 1:1 aspect ratio plot
(Figure \ref{fig:marm10freeinvp1it10g6000rat}), constructed in the same
way as Figure \ref{fig:fine10bulkinvp1it0g6000rat}, indicates that
once again the image energy is mostly confined to a corridor one
wavelength wide, although the wavelength is longer in this example,
corresponding to the lower source freequency band ([2.5,5,15,20] Hz,
as opposed to [2.5,5,25,30] Hz for Example 3).

\plot{marm10freeinvp1it10g6000}{width=\textwidth}{Example 5: Subsurface
  offset gather of 10 PCG iteration inversion at position $x_s=$ 6000 m indicated in Figure
  \ref{fig:marm10freeinvp1it10}, plotted on same grey scale. Note apparent concentration of energy
within 100 m of $h=0$.}

\plot{marm10freeinvp1it10g6000rat}{width=\textwidth}{Example 5: Subsurface
  offset gather of 10 PCG iteration inversion at position $x_s=$ 6000 m, depth range 2000 m $<z<$
  3200 m, plotted to 1:1 aspect ratio
  and padded with zeroes to left and right to obtain square
  region. Plotted on same grey scale as Figure \ref{fig:marm10freeinvp1it10g6000}. Note that extend of energy spread is very close to one
  apparent wavelength.}

%\newpage
%\plot{vg1invp0it0agc}{width=\textwidth}{Example6: RTM image after AGC, 
%Viking Graben data.}

%\plot{vg1invp1it0}{width=\textwidth}{Example6: Asymptotic inverse, 
%Viking Graben data.}

\section{Plots for DL Lecture}

\plot{coarse1unbornr45}{width=\textwidth}{Simulation of shot 45
  (nonlinear)}
\plot{coarse1bornr45c}{width=\textwidth}{Born simulation of shot 45}
%\plot{fine1diffr90}{width=\textwidth}{Difference between target,
%  background simulations of shot 90 (nonlinear)}
%\plot{fine1linerrr90}{width=\textwidth}{Linearization error}

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2019-10-20 repro/papers/varden/stuff.tex
added repro/papers/varden

\section{Plots for DL Lecture}

\plot{fine1unbornr90}{width=\textwidth}{Simulation of shot 90
  (nonlinear)}
\plot{fine1bgunbornr90}{width=\textwidth}{Simulation of shot 90 in
  background model (nonlinear)}
\plot{fine1diffr90}{width=\textwidth}{Difference between target,
  background simulations of shot 90 (nonlinear)}
\plot{fine1linerrr90}{width=\textwidth}{Linearization error}

Regularization by DSO annihilator: minimization of 
\begin{equation}
\label{eqn:regobj}
J_{\alpha}[\bar{\kappa}] =\frac{1}{2}( \|\doF \delta \bar{\kappa}
-\delta d\|_d^2 +\alpha^2
\|A\delta \bar{\kappa}\|^2_m)
\end{equation}

\begin{equation}
\label{eqn:regnormal}
W_m^{-1}(\doF^T W_d \doF + \alpha^2 A^TA)\delta \bar{\kappa} =
W_m^{-1}\doF^T W_d \delta d
\end{equation}

Define the augmented modeling operator
\begin{equation}
\label{eqn:columnop}
L = \left(\begin{array}{c}
\doF\\
A
\end{array}
\right) 
\end{equation}
View the range of this operator as the cartesian product of data
space and model space, with

### 2019-10-20 repro/papers/vikingbasic/paper.tex
added viking graben projects

--- 2019-10-19 repro/caam641/basic/paper.tex
@@
-\title{Basic Imaging of a 2D Marine Survey}
+\title{Basic Imaging: Viking Graben}
+\date{}
+\author{William W. Symes, The Rice Inversion Project}
+
 \maketitle 
-\label{ch:basic}
+\parskip 12pt 
+\begin{abstract}
+Presents very simple processing sequence resulting in post-stack and prestack migrated images of Viking Graben data.
+\end{abstract}
+
 \inputdir{project}
@@
 
-This section of the CAAM 641 course notes discusses so-called standard processing of the Viking Graben data, based on hyperbolic normal moveout (NMO).
-The connection between inversion and standard processing is the subject of the first part of this course.
-An earlier treatment can be found in my MSRI lecture notes from 2013, available on the TRIP web page, downloadable materials, short course and summer school materials, especially part 2.4.
+This paper discusses so-called standard processing of the Viking Graben data, based on hyperbolic normal moveout (NMO).
+The connection between inversion and standard processing is covered in detail in my MSRI lecture notes from 2013, available on the TRIP web page, downloadable materials, short course and summer school materials, especially part 2.4.
 I also show how to transform the by-products of standard processing into suitable input for prestack imaging via RTM.
@@
 The following sections outline the rationale for the choice of commands and their sequence.
-To reproduce my results, the reader will need to install SU, Madagascar, and TRIP packages.
+To reproduce my results, the reader will need to install SU, Madagascar, and trip2.1 packages.
 SU should be version 44 or later.
-The prestack part of the workflow requires parallel installation of TRIP, and at least 10 threads of execution, preferably more, to finish in reasonable time.
+The prestack part of the workflow requires parallel installation of trip2.1, and at least 10 threads of execution, preferably more, to finish in reasonable time.
 
@@
 
-\multiplot{2}{cdp200,cdp700}{width=0.45\textwidth}{Left: CDP 200; Right: CDP 700}
-\multiplot{2}{cdp1300,cdp2000}{width=0.45\textwidth}{Left: CDP 1300; Right: CDP 2000}
-
 \section{Velocity Analysis}
@@
 
-\multiplot{2}{velan200,velan700}{width=0.45\textwidth}{Left: Velocity spectrum for CDP 200; Right: Velocity spectrum for CDP 700}
-\multiplot{2}{velan1300,velan2000}{width=0.45\textwidth}{Left: Velocity spectrum for CDP 1300; Right: Velocity spectrum for CDP 2000}
-
 To QC your picks, plot the NMO correction of the corresponding CDPs as Figure \ref{fig:nmo200}, Figure \ref{fig:nmo700}, Figure \ref{fig:nmo1300}, and Figure \ref{fig:nmo2000}.
-
-\multiplot{2}{nmo200,nmo700}{width=0.45\textwidth}{Left: NMO corrected CDP 200; Right: NMO corrected CDP 700}
-\multiplot{2}{nmo1300,nmo2000}{width=0.45\textwidth}{Left: NMO corrected CDP 1300; Right: NMO corrected CDP 2000}
 
@@
 
-\plot{nmo2000lo}{width=0.45\textwidth}{NMO corrected CDP 2000 with tnmo=0.0,1.2,1.5,1.8,2.1,2.6,2.75 vnmo=1500,1800,1800,1800,2200,2400,2700}
-
 \section{Stack and Post-Stack Migration}
 
-The velocity-denominated quantity determined by flattening NMO-corrected CDP gathers is the so-called RMS velocity, or $v_{\rm rms}$, the root mean square of a local wave velocity expressed as a function of vertical travel time $t_0$:
+As explained in the short course lectures, the velocity-denominated quantity determined by flattening NMO-corrected CDP gathers is the so-called RMS velocity, or $v_{\rm rms}$, the root mean square of a local wave velocity expressed as a function of vertical travel time $t_0$:
 \begin{equation}
@@
 
-\plot{parastack}{width=0.9\textwidth}{Stack with linearly interpolated, constant extrapolated $v_{\rm rms}$}
-
 Because the deeper events are much less energetically imaged than the shallow, automatic gain control (amplitude equalization) produces a much more informative Figure \ref{fig:parastackagc}.
@@
 
-\plot{parastackagc}{width=0.9\textwidth}{Automatic Gain Control (AGC) applied to data of Figure \ref{fig:parastack}. Note pronouced diffraction hyperbolae in the deeper part of the secion.}
-
 You will note the precence of many {\em diffraction tails} especially below 2 s.
 These may be partially collapsed either by poststack time migration using Gazdag's algorithm (Figure \ref{fig:paragazmig}) or, more effectively, by poststack depth migration using one of several methods.
-Gazdag poststack time migration assumes layered velocity structure, hence uses only the first column of the interval velocity as function of time.
+Gazdag poststack time migration assumes layered velocity structure, hence uses only the first column of the interval velocity as function of tim.
 Nonetheless Figure \ref{fig:paragazmig} shows considerably less interference from diffraction tails; the graben structures that give this prospect its name are now clearly in evidence.
-
-\plot{paragazmig}{width=0.9\textwidth}{Gazdag's phase-shift time migration applied to the stack (Figure \ref{fig:parastack}), followed by AGC.
-Diffraction artifacts are largely collapsed to the diffracting points that caused them, and the graben structures that give this prospect its name begin to be clearly visible.
-This migration assumes horzontal layering, which is not correct on the scale of 25 km.}
-
-Poststack depth migration (a process that will be identified later as an approximation to the adjoint linearized zero-offset modeling operator) does even better - Figure \ref{fig:parapspi} displays the output of Phase-Shift-Plus-Interpolation poststack depth migration, using the velocity $v(z,x)$ depicted in Figure \ref{fig:paravintz} (obtained via use of the SU utility {\tt velconv}, see {\tt project/SConstruct} for details).
+Poststack depth migration (that is, what one would naturally mean - an approximation to the adjoint linearized zero-offset modeling operator) does even better - Figure \ref{fig:parapspi} displays the output of Phase-Shift-Plus-Interpolation poststack depth migration, using the velocity $v(z,x)$ depicted in Figure \ref{fig:paravintz} (obtained via use of the SU utility {\tt velconv}, see {\tt project/SConstruct} for details).
 The image gives reasonable depths for the various horizons.
@@
 
-\plot{parapspi}{width=0.9\textwidth}{Gazdag's phase-shift-plus-interpolation post-stack depth migration applied to the stack (Figure \ref{fig:parastack}), followed by AGC.
-Accounts for lateral velocity variation - uses interval velocity as function of midpoint and depth - more accurately than does the time migration of Figure \ref{fig:paragazmig}.
-The geology is even more clearly delineated, and depths should approximate those to be obtained by more sophisticated imaging.}
-
-\plot{paravintz}{width=0.9\textwidth}{Interval velocity as function of depth, derived from NMO velocity analysis.
-Probably not to be taken seriously below 3 km - for initial MVA estimate should extend by contant from 3 km and smooth.}
-
-With an eye to the prestack processing to be discussed later in the course, I include analogous results for downfiltered data.
+With an eye to the prestack processing to be discussed in another paper, I include analogous results for downfiltered data.
 Jie Hou suggested a 5-10-30-40 Hz bandpass filter.
 The resulting AGC'd stack (Figure \ref{fig:parastackagcf}) shows the expected decrease in resolution over the original data, which has significant energy over 50 Hz.
-
-\plot{parastackagcf}{width=0.9\textwidth}{NMO stack of (5,10,30,40) Hz bandpass filtered data, after AGC. Compare to Figure \ref{fig:parastack}.}
 
@@
 
-\plot{parapspif}{width=0.9\textwidth}{PSPI post stack migrated image from stack in Figure \ref{fig:parastackagcf}.}
-
 \section{Conclusion}
@@
 The sequence of processing steps takes us far enough from the basic physics of wave propagation to raise questions about whether the images produced here are actually images of anything real.
-Later in the course, we turn to prestack processing based more directly on the acoustic wave equation - only a step further, but meaningful nonetheless.
+In a subsequent paper, we turn to prestack processing based on the acoustic wave equation - only a step further, but meaningful nonetheless.
 To telegraph part of the punch: prestack processing supports the validity of the images obtained in this paper by poststack processes.
@@
 
-\section{Suggested Projects}
-\begin{enumerate}
-\item The velocity analysis presented above seems quite minimal - only four analysis CMPs selected, more or less evenly spaced.
-Is there any advantage in choosing more analysis points?
-Does the stack become cleaner, the interval velocity any more believable?
-
-\item Find another field data set and carry out a similar analysis.
-
-\item Taking another tack, find a synthetic model with more lateral variation in velocity - Marmousi is an obvious choice - and carry out NMO based velocity analysis.
-Does the interval velocity so obtained bear any relation to the actual velocity (which of course you know, since the example is synthetic)?
-What about the PSPI migrated image and the actual model?
-
-\item AGC is a crude tool for equalizing amplitudes, that completely destroys the relation between event and reflector dynamics.
-Is it possible to use a simpler scaling, such as by a power of $t$ or $z$, to  achieve the same enhancement of important reflectors at depth while maintaining a simple relation between reflector and reflection amplitudes?
-\end{enumerate}
-
 \bibliographystyle{seg}
 \bibliography{../../bib/masterref}
+
+\multiplot{2}{cdp200,cdp700}{width=0.45\textwidth}{Left: CDP 200; Right: CDP 700}
+\multiplot{2}{cdp1300,cdp2000}{width=0.45\textwidth}{Left: CDP 1300; Right: CDP 2000}
+
+\multiplot{2}{velan200,velan700}{width=0.45\textwidth}{Left: Velocity spectrum for CDP 200; Right: Velocity spectrum for CDP 700}
+\multiplot{2}{velan1300,velan2000}{width=0.45\textwidth}{Left: Velocity spectrum for CDP 1300; Right: Velocity spectrum for CDP 2000}
+
+\multiplot{2}{nmo200,nmo700}{width=0.45\textwidth}{Left: NMO corrected CDP 200; Right: NMO corrected CDP 700}
+\multiplot{2}{nmo1300,nmo2000}{width=0.45\textwidth}{Left: NMO corrected CDP 1300; Right: NMO corrected CDP 2000}
+
+\plot{nmo2000lo}{width=0.45\textwidth}{NMO corrected CDP 2000 with tnmo=0.0,1.2,1.5,1.8,2.1,2.6,2.75 vnmo=1500,1800,1800,1800,2200,2400,2700}
+
+\plot{parastack}{width=0.9\textwidth}{Stack with linearly interpolated, constant extrapolated $v_{\rm rms}$}
+
+\plot{parastackagc}{width=0.9\textwidth}{Automatic Gain Control (AGC) applied to data of Figure \ref{fig:parastack}. Note pronouced diffraction hyperbolae in the deeper part of the secion.}
+
+\plot{paragazmig}{width=0.9\textwidth}{Gazdag's phase-shift time migration applied to the stack (Figure \ref{fig:parastack}), followed by AGC.
+Diffraction artifacts are largely collapsed to the diffracting points that caused them, and the graben structures that give this prospect its name begin to be clearly visible.
+This migration assumes horzontal layering, which is not correct on the scale of 25 km.}
+
+\plot{parapspi}{width=0.9\textwidth}{Gazdag's phase-shift-plus-interpolation post-stack depth migration applied to the stack (Figure \ref{fig:parastack}), followed by AGC.
+Accounts for lateral velocity variation - uses interval velocity as function of midpoint and depth - more accurately than does the time migration of Figure \ref{fig:paragazmig}.
+The geology is even more clearly delineated, and depths should approximate those to be obtained by more sophisticated imaging.}
+
+\plot{parastackagcf}{width=0.9\textwidth}{NMO stack of (5,10,30,40) Hz bandpass filtered data, after AGC. Compare to Figure \ref{fig:parastack}.}
+
+\plot{parapspif}{width=0.9\textwidth}{PSPI post stack migrated image from stack in Figure \ref{fig:parastackagcf}.}
+
+\plot{paravintz}{width=0.9\textwidth}{Interval velocity as function of depth, derived from NMO velocity analysis.
+Probably not to be taken seriously below 3 km - for initial MVA estimate should extend by contant from 3 km and smooth.}

### 2019-10-20 repro/papers/vikingmig/paper.tex
added viking graben projects

\title{Prestack Imaging: Viking Graben}
\date{}
\author{William W. Symes, The Rice Inversion Project}

\maketitle 
\parskip 12pt 
\begin{abstract}
Uses the output of NMO-based processing as the starting point for prestack imaging of the Viking Graben data using the IWAVE package.
\end{abstract}

\inputdir{project}

\section{Introduction}
This paper describes how to use IWAVE to apply prestack RTM to the Viking Graben data. The data and a simple time processing workflow are described in the companion article \cite[]{trip15:vikingbasic}. A copy of the source for this article is included in this book, and should be built first, as this project uses the intermediate data for that project: from the paper directory housing the LaTeX source for this paper ({\tt vikingmig}), execute 
\begin{itemize}
\item {\tt cd ../vikingbasis/project}
\item {\tt scons}
\end{itemize}

This project uses IWAVE's {\tt acd} command to produce 
\begin{itemize}
\item postmigration stacked image,
\item image volume with shot axis (shot record extension), and 
\item image volume with subsurface offset (space shift) axis (subsurface offset extension).
\end{itemize}
The method is described in the \cite[]{SymesIWEXT:15}.

\section{Prestack RTM with IWAVE}
IWAVE modeling, migration, and inversion commands requires as input:
\begin{itemize}
\item data, in SU native format (XDR is still allowed, but because of SU's design must be compiled in, and I don't recommend it)
\item model, as specific physical fields and in specific units. For constant density acoustics, the primary physical field is velocity-squared, and the unit is m/ms (squared). 
\item source, in the form of the right-hand side of the wave equation. 
\end{itemize}
The self-doc for IWAVE's constant density acoustics ({\tt acd}) modeling and migration command includes these points and also reviews the other required and optional parameters.

The data requires no special preparation, other than what has already been done for the post-stack processing. The downfiltered (5-10-30-40 Hz bandpass) data is used, in order to keep numerical dispersion under control: 30 Hz at 1500 m/s (the lowest modeled velocity) is equivalent to 50 m spatial wavelength, or 5 grid cells at 10 m sampling. Thus the spectral roll-off is at 5 gridpoints/wavelength, at least vertically. In order to keep this demonstration example small, and avoid having to extend the velocity grid to eliminate boundary reflections, {\tt acd} currently having no usable absorbing boundary conditions, I chose to migrate only the middle 200 shots ({\tt fldr} = 401-601).

The velocity-squared requires several preparatory steps. The starting point is the interval velocity depth function prepared from NMO velocity analysis, as described earlier (Figure \ref{fig:paravintz}). The SU commands produce a flat binary file, sampling this velocity field at 10 m in depth (that was a somewhat arbitrary choice) starting at and at every CDP, that is, every 12.5 m starting at the first CDP location, 1617.5 m from the horizontal coordinate origin as determined from the headers {\tt sx} and {\tt gx} for cdp 1. While the SU self-docs do not specify the unit of velocity (a frequent failing of self-docs), from the values one can infer m/s. Use the 2D grid parameters to build an RSF header file pointing to the flat binary data produced by SU. Use {\tt sfadd} to scale the velocity $10^{-3}$ thus changing units to m/ms, and {\tt sfadd} again to create a velocity-squared file for input to IWAVE.

The source data starts with creation of a SU file containing the 5-10-30-40 Hz bandpass filter, converted via the IWAVE {\tt towed\_array} command to a file containing right-hand sides for each of the selected shots.

The IWAVE command itself is embedded in a python dictionary set up to drive parallel execution on various platforms, as described in \cite{SymesIWPAR:15}.

\noindent {\bf Remark:} Former versions of IWAVE incorporated several of these preparatory steps - conversion between various choices of physical parameters, unit conversions, conversion of a source wavelet to a RHS. The latest version (trip2.1) has these things unbundled: I have learned that the resulting flexibility is worth having. It makes sense to create programs (like {\tt towed\_array} and scripts to carry out conventional preparation steps conveniently, and more of these are needed. The sole exception, I believe, is unit conversion. IWAVE needs to be modified to deal with units automatically and consistently - it's fine, indeed on some level unavoidable, to rigidly choose units internally, however that choice should be transparent to the user. It's really obnoxious to have to remember to convert velocities to m/ms. The C {\tt units} package can serve as a portable basis for automatic conversion during reads and writes - definitely on the TO-DO list!

The result of RTM of shots 401 - 601 is displayed in Figure \ref{fig:mig401-601}. Obviously the migration aperture is limited by the shot window. To make comparison easier, I windowed the depth range to [0,3000] m, and the inline range to [14000, 18000] m. The windowed image is plotted in Figure \ref{fig:mig401-601wind}, alongside the same window of the filtered PSPI image. The locations and characters of the events are very similar, as one would expect. I have AGC'd the RTM result - it is merely a migration, not an inversion, and the amplitudes are in themselves meaningless and highly unequal, with deeper events being quite faint. In lieu of inversion, AGC makes the structural information content manifest.

\section{Prestack Velocity QC: shot record extended migration}

The IWAVE driver {\tt acde} does modeling and migration for shot record and horizontal subsurface offset extensions. Figure \ref{fig:mig401-601extsrwind1} shows the faces of the shot record extended image cube for the data described in the previous section. The right hand panel is the shot record image gather at $x=15750$.

The appearance of this image volume is typical for the shot record extension. The aperture of the individual shot images is quite small, and especially at shallow depths the image is dominated by non-specular energy. In order for the plot energy not to be dominated by this ``noise'' (from the specular reflection point of view), I plot only depths greater than 800 m. The gather is mostly flat, indicating that the velocity is reasonably consistent with the data, though there are hints of possible improvement available, and some mildly sloping events.

[Another question: we know from the NMO exercise that the {\tt paracdp} data contains visible amounts of residual multiple energy. What is the shot record image gather signature of a multiple, assuming that the migration velocity is consistent with primaries?]

To see more clearly the degree to which this migration has flattened the shot record gathers, I have clipped out and AGC'd signal windows around the four gathers starting at $x=15000$ m, at 500 m intervals. These appear as Figures \ref{fig:cig401-601extsr150}-\ref{fig:cig401-601extsr165} .

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

\plot{paravintz}{width=0.9\textwidth}{Interval velocity as function of depth, derived from NMO velocity analysis. Probably not to be taken seriously below 3 km - for initial MVA estimate should extend by contant from 3 km and smooth.}

\plot{mig401-601}{width=0.9\textwidth}{RTM image from shots 401-601, using interval velocity displayed in Figure \ref{fig:paravintz}. Compare to poststack migrations displayed in \cite[]{trip15:vikingbasic}. Note limited migration aperture due to limited shot range.}

\plot{mig401-601extsrwind1}{width=0.9\textwidth}{RTM shot record extended output from shots 401-601, using interval velocity displayed in Figure \ref{fig:paravintz}. Right panel is common image gather for $x=15750$ m.}

%\multiplot{4}{cig401-601extsr150,cig401-601extsr155,cig401-601extsr160,cig401-601extsr165}{width=0.24\textwidth}{Shot-record CIGs at $x=$ 15000, 15500, 16000 and 16500 m. With AGC.}

\plot{mig401-601extso}{width=0.9\textwidth}{RTM subsurface offset extended output from shots 401-601, using interval velocity displayed in Figure \ref{fig:paravintz}. Right panel is common image gather for $x=16500$ m.}

### 2019-10-20 admin/linux.mpi.py
removed openmp from generic mpi config

@@ -1,6 +1,5 @@
 CC = 'mpicc' 
-CCFLAGS = '-O3 -fopenmp -ffast-math -pedantic -Wunused -Wno-long-long -D_FILE_OFFSET_BITS=64 -D_LARGEFILE_SOURCE -D_GNU_SOURCE -DIWAVE_USE_MPI'
+CCFLAGS = '-O3 -ffast-math -pedantic -Wunused -Wno-long-long -D_FILE_OFFSET_BITS=64 -D_LARGEFILE_SOURCE -D_GNU_SOURCE -DIWAVE_USE_MPI'
 CFLAGS = '-std=c99 -Wimplicit'
 CXX = 'mpicxx'
 CXXFLAGS = '-std=c++11'
-LINKFLAGS = '-fopenmp'

### 2019-10-23 repro/papers/trans/append1.tex
update

--- 2019-10-19 repro/papers/trans/paper.tex
@@
-\title{Waveform Inversion via Source Extension: Major Questions and Research Plan}
-\author{William. W. Symes \thanks{The Rice Inversion Project,
-Department of Computational and Applied Mathematics, Rice University,
-Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}
-
-\extrafloats{100}
-
-\lefthead{Symes}
-
-\righthead{Source Extension}
-
-\maketitle
-\begin{abstract}
-FWI via source extension is a workaround for cycle skipping.
-For pure transmission (including diving wave) data, a close link to traveltime tomography explains the good results observed in experiments.
-No such link is known to exist in the case of pure reflection data, and numerical results are mixed.
-This proposal outlines several numerical experiments that should complement theoretical investigations and decide the effectiveness of these algorithms.
-\end{abstract}
-
-\section{Overview}
-FWI can be described in terms of a wave operator $L[c]$, depending on an array of space-dependent coefficients $c$, a trace sampling operator $P$, a wavefield $u$, and a source function (of space and time) $f$.
-The basic FWI problem is: given $d$, find $c$ so that
-\begin{equation}
-\label{eqn:fwi}
-Pu \approx d \mbox{ and } L[c]u = f.
-\end{equation}
-Sometimes $f$ is regarded as given, sometimes only some aspects are given (for example, localization at a source point) and others are to be determined as part of the solution.
-A typical formulation is via nonlinear least squares:
-\[
-\mbox{choose } c \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2 
-\]
-usually with some sort of regularization thrown in.
-
-Note that $f$ and $d$ are typically vectors of trace gathers, one for each source position in the survey (or possibly some other parameter, produced by processing - will stick with source position $\bx_s$).
-
-As is well-known, local optimization methods are the only feasible approach given the dimensions of a typical instance of \ref{eqn:fwi}, and those have a tendency to stall due to ``cycle-skipping''.
-Source extension is one approach to avoiding this problem.
-It consists in imposing the wave equation as a soft as opposed to hard constraint, by allowing thw source to have more degrees of freedom than is permitted by the basic model.
-For sake of argument (and because almost all work so far has assumed it), adopt the isotropic point source as the model: for the data gather at source positon $\bx_s$, the source takes the form
-\begin{equation}
-\label{eqn:ptsrc}
-f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s).
-\end{equation}
-Every method explained below can be altered to accommodate more complex source models.
-
-Source extension replaces $f(\bx,t;\bx_s)$ with an artificial source $g(\bx,t; \bx_s)$ with more degrees of freedom than merely a single common source wavelet $w(t)$ as in definition \ref{eqn:ptsrc}.
-Assuming that \ref{eqn:ptsrc} is actually correct physics, these extra degrees of freedom have to be suppressed in the optimal solution, via an {\em annihilator} $A$, an operator acting on extended sources $g$, whose null space is precisely the physical sources (of the form \ref{eqn:ptsrc}).
-Extended source inversion replaces the FWI objective with a penalty function
-\begin{equation}
-\label{eqn:esi}
-\mbox{choose } c, g \mbox{ to minimize } \|PL[c]^{-1}g -d \|^2 + \alpha \|A[g]\|^2 
-\end{equation}
-
-The way in which degrees of freedom are introduced, and the choice of annihilator, distinguish the various source extension methods.
-
-\subsection{Wavefield Reconstruction Inversion}
-Introduced by \cite{LeeuwenHerrmannWRI:13} and significantly modified by \cite{WangYingst:SEG16}.
-Extended source space: arbitrary functions of space-time $g(\bx,t;\bx_s)$, one for each source position.
-Annihilator:
-\begin{equation}
-\label{eqn:wri}
-A[g] = g-f
-\end{equation}
-With these choices, WRI takes the form introduced by \cite{WangYingst:SEG16}.
-The original approach is equivalent: \cite{LeeuwenHerrmannWRI:13} use the dyamic wavefield $u$ as the additional degrees of freedom, and minimize
-\[
-\|Pu-d\|^2 + \alpha \|L[c]u-f\|^2
-\]
-Since the wave equation (any reasonable version!) has a unique solution for each right-hand side, replacing $L[c]u$ turns this original form of WRI into the form \ref{eqn:esi}.
-
-Note that in princple at least, the actual source $f$ (positions and wavelet) must be known {\em a priori}.
-
-The Wang-Yingst variant leads to a huge reduction in the number of degrees of freedom: either the original or the modified versions, as originally presented, appear to require storage of an entire space-time source, which is prohibitive for 3D and/or time domain formulation.
-I will explain in an appendix how the Wang-Yingst approach drastically shrinks the storage requirements of this method, and makes it feasible for time-domain 3D application, which Wang and Yingst and their co-authors have demonstrated.
-
-\subsection{Space-Time Extension}
-\cite{HuangSymes:SEG16b}. Differs from WRI by choice of annihilator:
-\begin{equation}
-\label{eqn:msste}
-A[g](\bx,t;\bx_s) = |\bx-\bx_s|g(\bx,t;\bx_s).
-\end{equation}
-This annihilator merely enforces the position constraint on the source, without relying on {\em a priori} knowledge of the source wavelet.
-In fact, the source wavelet is a by-product of the minimization \ref{eqn:esi}.
-
-Appears to require storage of full space-time source volume - the trick explained in the appendix for WRI does not work, due the form of the annihilator.
-
-\subsection{Volume Extension}
-\cite{HuangSymes:SEG16a}.
-Assumes that source signature deconvolution has been performed, so that the actual wavelet is (a bandlimited version of) $\delta(t)$.
-The extended source is an {\em exploding reflector}:
-\begin{equation}
-\label{eqn:msvol}
-g(\bx,t;\bx_s) = h(\bx;\bx_s)\delta(t).
-\end{equation}
-Uses the same annihilator as the Space-Time Extension, but loses the advantage of solving for the source as by-product - the source must be known, so that decon can be carried out prior to inversion.
-
-However, only requires a spatial storage volume per source position, essentially equivalent to a second data volume, therefore practical for 3D (though I don't believe it has been tried except in 2D).
-
-The variable projection method applied to this extension has essentially exploding reflector inversion (as opposed to exploding reflector migration) as its inner problem, i.e.
-estimation of $h$ in \ref{eqn:msvol}.
-The same problem arises in so-called photoacoustic tomography, and has received a lot of attention in the mathematical inverse problems and biomedical imaging literature.
-In fact there is a quite accurate approximate inverse available for the inner problem, consisting of converting pressure traces to normal velocity traces and back-projecting (this is known as ``time reversal'') in the IP literature).
-
-A disadvantage of this approach, if you want to handle reflections, is that unlike post-stack migration the actual physical wave velocity is used in modeling, so that the the region in which $h$ must be defined may considerably larger than the ordinary modeling domain - reflection two-way time is effectively modeled by one-way time.
-
-The volume extension concept is similar to an idea discussed by \cite{ZhangGao:08}.
-
-\subsection{Surface Extension}
-No publication yet - subject of my workshop talk at the 2016 SEG, and GH has described it in talks at Total.
-Similar to the volume extension, but the source is spread over the source space-time surface, rather than over space at $t=0$.
-In the simplest case, where the sources all lie on $z=z_s$, the extended source space consists of functions (distributions) of the form ($\bx=(x,y,z)$)
-\begin{equation}
-\label{eqn:mssur}
-g(\bx,t;\bx_s) = h(x,y,t)\delta(z-z_s).
-\end{equation}
-As for the previous two cases, the annihilator is given by the localization penalty \ref{eqn:msste}.
-Therefore the source wavelet is a by-product of the inversion, and does not need to be specified {\em a priori}.
-Like the volume extension, the additional data volume required is on the same order of size as the data.
-
-\subsection{Adaptive Waveform Inversion = Source-Receiver Extension}
- Popularized by \cite{Warner:14}, \cite{Warner:16}, has many antecedents, listed in \cite[]{HuangSymes:17}. Each trace regarded as separate gather, with separate point source waveform. In effect,
-\begin{equation}
-\label{eqn:awi}
-g(\bx,t;\bx_s) = w(t;\bx_s)\delta(\bx-\bx_s) 
-\end{equation}
-with repeated occurances of $\bx_s$ regarded as distinct. Usually written by making the wavelet a function of the receiver position also, to distinguish redundant source positions:
-\begin{equation}
-\label{eqn:awi-alt}
-g(\bx,t;\bx_r,\bx_s) = w(t;\bx_r,\bx_s)\delta(\bx-\bx_s) 
-\end{equation}
-hence the alternate name.
-
-Various annihilators have been suggested - see references.
-Warner proposes a variant (``AWI'')  that (in principle) requires the source wavelet to be known, and deconvolved from the data traces.
-Then uses multiplication by $t$, which annihilates $\delta(t)$:
-\begin{equation}
-\label{eqn:awi-ann}
-A[g](\bx,t;\bx_r,\bx_s) = t g (\bx,t;\bx_r,\bx_s)
-\end{equation}
-AWI distinguished from earlier variants by normalizing the objective with the estimated wavelet $L^2$ norm:
-\begin{equation}
-\label{eqn:awi-obj}
-\mbox{choose } c, g \mbox{ to minimize } \sum_{\bx_r,\bx_s}\frac{\int dt\,|t|^2 |w(t;\bx_r,\bx_s)|^2}{\int dt\,|w(t;\bx_r,\bx_s)|^2}
-\end{equation}
-Note the order of summation and normalization: each trace is normalized independently (this is not clear from the notation in \cite[]{Warner:16}).
-
-The justification for normalization given by \cite{Warner:16} is that it avoids a familar source/reflector ambiguity.
-Actually it is more important than that, and plays an important role when no reflectors are present.
-It is possible to recast the objective \ref{eqn:awi-obj} as equivalent to \ref{eqn:esi} with a weighted norm used in the latter.
-[I am writing this up - take my word for it!] The weighted norm makes the linear map from source to data {\em unitary} (at least to leading order in frequency), and that is an essential property in relating AWI to traveltime tomography.
-
-\section{What is known}
-For several of the source extension algorithms, the objective is known to be closely linked to traveltime tomography in one form or another, and to inherit its convergence properties.
-These connections are known {\em only} for pure transmission, and not always then.
-Nothing theoretical is known about any of these methdods for reflection data.
-
-The connection to traveltime tomography was first (so far as I know) established for the source-receiver extension in several TRIP technical reports related to Hua Song's PhD thesis \cite[]{Song:94c,SoSy:92,SongSymes:94a,SongSymes:94b,Symes:94c}.
-This work established a form of differential semblance velocity estimation for crosswell tomography, and the related tomographic objective was a form of slope, or stereo-, tomography.
-The time moment objective \ref{eqn:awi-obj} (without normalization) first appears in \cite[]{Plessix:00a,Plessix:00}.
-A similar idea appears in \cite[]{LuoSava:11}.
-Neither of these refferences made an explicit connection to tomography - that was done in \cite[]{HuangSymes2015SEG,HuangSymes:17}.
-
-The defect, in the relation between AWI/source-receiver extention and traveltime tomography, is that it collapses when the ray field (must be able to speak of rays, to invoke traveltime tomography) develops caustics.
-This was already demonstrated both theoretically and by numerical example in \cite[]{Symes:94c}, though the fact does not seem to have percolated very far.
-\cite{HuangSymes:17} give an explicit illustration of this failure in a diving wave setting.
-As soon as the diving wave field is triplicated, AWI is just as bad as FWI.
-
-On the other end of the list, almost nothing is known about why WRI converges, or when it does.
-It may be possible to repeat a version of the analysis mentioned next for the volume and surface extensions, based on geometric optics and the Wang-Yingst variant as described in the appendix, though that has not been tried yet.
-
-The volume and surface extensions have a well-defined relation to tomographic principles (slope or traveltime, depending on type of annihilator).
-This relation holds regardless of ray geometry (with or without triplication) in transmission configuration.
-
-Again, nothing is known theoretically about the behaviour of any of these methods for reflection data, and numerical evidence is ambiguous.
-
-\section{Proposed Project}.
-
-Amongst the source extension options described in the previous section, only the surface extension has all of these properties:
-\begin{itemize}
-\item requires only additional storage roughly equivalent to the data volume
-\item does not require prior knowledge of the source wavelet, in fact recovers an estimate as a by-product
-\item inner problem (in variable projection - estimation of the extended source) has a cheap approximate solution, via back-propagation
-\item relation with traveltime inversion persists, regardless of ray geometry
-\end{itemize}
-
-Therefoire I propose that we begin with a time-domain implementation of the surface source extension.
-
-\section{Detailed Description}
-The simplest version of the surface source extension assumes acoustic wave physics, and that the physical source locations lie on the horizontal surface $z=z_s$.
-The coefficient vector $c$ is the pair $(\kappa,\beta)$ of bulk modulus and buoyancy, and the wave operator $L[c]$ is given by the linear acoustics system:
-\begin{eqnarray}
-\label{eqn:awe}
-\frac{\partial p}{\partial t} & = & - \kappa (\nabla \cdot \bv +
-w(t)\delta(\bx-\bx_s)), \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
-p & =& 0 \mbox{ for } t<0\nonumber\\ 
-\bv & = & 0 \mbox{ for } t<0 
-\end{eqnarray}
-The physical source is modeled as an isotropic point constitutive law defect. As mentioned earlier, it is simple to accommodate more complex source models.
-
-The dynamical field is the 4-tuple $u=(p,\bv)$.
-Either the pressure $p$ or the particle velocity $\bv$ could be sampled (or both) - presume that the pressure is sampled, that is $Pu = \{p(\bx_r,t;\bx_s)\}$, where $\{\bx_r\}$ are the receiver positions.
-
-Extended sources are spread over $z=z_s$ (\ref{eqn:mssur}). The annihilator is the source localization penalty, equation \ref{eqn:msste}.
-
-Denote by $J_{\alpha}[g,c;d]$ the objective function defined in \ref{eqn:esi}:
-\begin{equation}
-\label{eqn:jdef}
-J_{\alpha}[g,c;d] = \frac{1}{2}( \|PL[c]^{-1}g -d \|^2 + \alpha \|A[g]\|^2) 
-\end{equation}
-To compute the objective 
-\begin{itemize}
-\item solve the acoustic wave system \ref{eqn:awe} with $g$ as in \ref{eqn:mssur}
-\item sample at receiver locations and compute mean-square difference with pressure data traces, take mean square
-\item add scaled mean square of annihilator applied to the source
-\end{itemize}
-
-The variable projection method \cite[]{GolubPereyra:03} applied to the objective \ref{eqn:esi} approaches its minimization via a nested algorithm: first minimize \ref{eqn:esi} over the source $g$ for given coefficients $c$, thus making $g = g[c]$ a function of $c$ (inner minimization); then substitute $g[c]$ in $J$ and minimize the resulting function of $c$ (outer minimization).
-The optimal $g[c]$ solves the first order necessary condition for minimization of $J$ over $g$: $\nabla_g J_{\alpha}[g,c;d] = 0$.
-For linear annihilators such as that defined by \ref{eqn:mssur}, $J$ is quadratic in $g$, so the first order necessary condition is also sufficient, and is equivalent to the linear {\em normal equation}:
-\begin{equation}
-\label{eqn:normal}
-(S[c]^TS[c] + \alpha A^TA) g = S[c]^T d
-\end{equation}
-where I have written $S[c]=PL[c]^{-1}$ for the operator that maps the (extended) source to the data traces. Substitution of $g[c]$ in $J$ results in the {\em reduced objective}
-\begin{equation}
-\label{eqn:jred}
-\tilde{J}_{\alpha}[c;d] = J_{\alpha}[g[c],c;d].
-\end{equation}
-The gradient would appear to involve the derivative of the inner solution $g[c]$ with respect to $c$, computation of which implies solution of additional linear systems.
-However by fortunate accident, the chain rule and the normal equation \ref{eqn:normal}
-conspire to eliminate all terms involving derivatives of $g[c]$. Consequently the direction derivative of $\tilde{J}_{\alpha}$ in the direction $\delta c$ is
-\begin{equation}
-\label{eqn:tildejderiv}
-D_c\tilde{J}_{\alpha}[c;d]\delta c = \langle D(S[c]g)\delta c,S[c]g-d \rangle|_{g=g[c]}
-\end{equation}
-whence
-\begin{equation}
-\label{eqn:l2grad}
-\nabla \tilde{J}_{\alpha}[c;d] = D_c(S[c]g)^T|_{g=g[c]}(S[c]g[c]-d)
-\end{equation}
-Let us field-strip the expression on the right.
-The rightmost factor is the data residual, which is a by-product of many methods for the solution of the normal equation \ref{eqn:normal}.
-$D_cS$ is the linearization or Born approximation of $S$, or perturbation with respect to $c$, with $g$ fixed:
-\[
-S[c+\delta c]g \approx S[c]g + D_c(S[c]g)\delta c
-\]
-That is, $D_c(S[c]g)$ is the object commonly called the Born modeling operator, with source wavefield $g$.
-Its adjoint or transpose is $D_c(S[c]g)^T$ - as is well known, this is the RTM operator (or rather a version of it), computed via the adjoint state method in the time domain.
-Note that the Born and RTM operators need to match, i.e.
-both be derived from the same underlying full waveform modeling operator (in this case acoustodynamics, but the same would apply for any other wave theory).
-
-The upshot is that the computation of the gradient goes like this:
-\begin{itemize}
-\item solve the normal equation \ref{eqn:normal} for $g[c]$
-\item compute (or retain) the data residual $S[c]g[c]-d$
-\item reverse-time migrate the data residual, using $g[c]$ to generate the source wavefield in the adjoint state method - the output is the gradient of $\tilde{J}_{\alpha}$.
-\end{itemize}
-
-To make this algorithm practical and efficient, effective preconditioning needs to be applied to both inner and outer problems.
-
-\subsection{Preconditioning the Outer Problem: Gradient Smoothing}
-Conditioning the outer (velocity etc.) update has different meaning for transmission versus reflection problems.
-One interpretation of ``transmission'' is that no reflection occurs on any ray path connecting source and receiver, whereever these may be located.
-Is is practically implicit in the use of ``ray path'' in this statement that the material model is at least mostly smooth, or essentially localized in low spatial frequencies.
-Smoothness is also one way to interpret ``lack of reflections''.
-The classic work explaining the need for model smoothness in tomography is \cite{DelpratLailly:92}.
-Since the surface source extension method gets its ability to identify kinematically correct velocity from its relation to tomography, it also requires model smoothness, at least for transmission.
-
-A natural method for enforcing smoothness is use of a norm in model space that penalizes lack of smoothness.
-This method is also properly called preconditioning, which consists in altering the choice of norm to enhance convergence.
-For pure transmission, the choice recommended by \cite[]{DelpratLailly:92} is a norm from the Sobolev family.
-These are weighted $L^2$ (RMS) norms with weight operators indexed by a real parameter $s$:
-\begin{equation}
-\label{eqn:sob}
-\|u\|^2_s = \langle u, W_s u \rangle_{L^2},\,\,W_s = (I - \nabla^2)^s
-\end{equation}
-Choosing $s>0$ penalizes lack of smoothness.
-According to the Sobolev Embedding Theorem and the basic well-posedness theorem for ordinary differential equations (which governs the behaviour of the ray system), it is neccessary to choose $s > 2 + d/2$ in dimension $d$, in order that metrizing the velocity (amongst other things) with the $s$-norm will guarantee sufficient smoothness that ray theory is well-behaved, and the link with tomography honored.
-For both $d=2$ and $d=3$, $s=4$ would work.
-
-The gradient of a function is defined with respect to the domain norm. Using the $s$-norm, the gradient is
-\begin{equation}
-\label{eqn:grads}
-\nabla_s \tilde{J}_{\alpha}[c,d] = W_s^{-1}\nabla \tilde{J}_{\alpha}[c,d]
-\end{equation}
-If $s>0$ then $W_s^{-1}$ is a smoothing operator. It is best implemented in the Fourier domain.
-
-The most natural way view the reflection problem is to allow non-smoothness in $c$.
-It is not clear at this point what would be the optimal choice of weight operator in this case, or even whether there is a choice of Hilbert space structure that would offer an easily computable gradient and development of short-scale structures to generate reflections.
-GH used very low frequency data (in frequency domain) and Tihonov regularization.
-His results seemed to depend heavily on the regularization weight.
-Of course we also do not know how to connect the reflection case to a tomographic principle, so have no guidance for how a model space norm should be selected.
-These are the main target questions of this project.
-
-\subsection{Preconditioning the Inner Problem: Time Reversal}
-%In contrast to the outer problem, a preconditioner for the inner problem is available that works regardless of medium smoothness, so for both transmission and reflection (the latter with some caveats).
-In contrast to the outer problem, the inner problem admits a conventional preconditioner, that is, a change of variables that in the spaces of source traces and data traces respectively that moves the singular values of the modeling operator closer to 1.
-This preconditioner also creates an approximate solution operator for the inner problem, of the form
-\begin{equation}
-\label{eqn:appinv}
-S[c]^{\dagger} = W_m[c]^{-1}S[c]^TW_d[c],
-\end{equation}
-in which $W_m$ and $W_c$ are $c$-dependent positive definite symmetric operators that define weighted inner products in domain and range of $S[c]$, and their associated weighted norms:
-\begin{eqnarray}
-\label{eqn:norms}
-\langle f,g \rangle_m & = & \langle f, W_mg\rangle; \|f\|_m^2 = \langle f,f\rangle_m \nonumber \\
-\langle f,g \rangle_d & = & \langle f, W_d g\rangle; \|f\|_d^2 = \langle f,f \rangle_d
-\end{eqnarray}
-We have implicitly used the notations $\langle \cdot,\cdot \rangle, \|\cdot \|$ (without subscripts) to denote the (unweighted) $L^2$ norm, in whatever form is appropriate for the arguments.
-That is what these symbols mean in display \ref{eqn:norms}, and will continue to mean.
-
-The approximate inverse $S[c]^{\dagger}$ defined by the relation \ref{eqn:appinv} is the transpose of $S[c]$ with respect to the inner products $\langle \cdot,\cdot \rangle_m,, \langle \cdot,\cdot \rangle_d$, so $S[c]$ is approximately unitary in the same sense.
-Consequently the singular spectrum (which is also dependent on choice of inner product) is close to that of the identity - or most of it is, as we shall explain - and the convergence of  a Krylov subspace method like Conjugate Gradient (CG) Iteration converge relatively rapidly.
-
-The system that needs solving, of course, is the normal equation \ref{eqn:normal},  so it is the convergence of CG or the like applied to that system that is the important issue.
-The normal equation is the first order necessary condition for minimization of the objective
-\begin{equation}
-\label{eqn:jdefs}
-J_{\alpha}[g,c;d] = \frac{1}{2}( \|S[c]g -d \|^2 + \alpha \|A[g]\|^2).
-\end{equation}
-over the choice of surface source $g$.
-The choice of norms appearing in \ref{eqn:jdefs} is actually arbitrary, and the norms appearing in the first and second terms on the right-hand side do not even have to be the same.
-
-%First the transmission case - a sort of rotated crosswell configuration. Idealize the receiver set as a continuum plane $z=z_r$, with $z_r > z_s$ - not essential but simpler.
-
-%Two key facts and a geometric assumption underly the construction of the time-reversal preconditioner: (i) a local relation between the surface source and the normal particle velocity, and (ii) the uniqueness of solutions of the Dirichlet boundary value problem for the wave equation.
-Together these facts allow us to reconstruct the pressure field, or the significant part of it, from the transmission data, and to approximate a distributed source responsible for a given set of data traces.
-
-%The geometric assumption is that the pressure field is {\em outgoing} at the recording datum: that is, that it is the restriction to $z>z_s$ of a pressure field that is negligible in $z<z_r$ for large enough time.
-``Negligible'', rather than vanishing, as the latter is strictly speaking impossible: this property is asymptotic in frequency, and assured by an assumption about the ray field associated to the acoustic system.
-
-%NB: the definition of the forward map (and the data) should include a taper $\phi(\bx_r,\bx_s)$ masking out the shot gather $d_r(\cdot,\cdot;\bx_s)$ for each source position $\bx_s$.
-However I will not explicitly mention this operation.
-I will also ignore it in the numerical examples, but there will be a price to pay.
-
-To begin with, change notation slightly, and write $S[c]h_s$ rather than $S[c]g$ with $g = h_s \delta(z-z_s)$, as $z_s$ will be presumed ot be the same for all shots. 
-%The addition of the subscript indicates that the sources lie on the plane $z=z_s$ and the receivers are on the plane $z=z_r$.
-As will be apparent shortly, a similar operator with source and receiver planes swapped will also be important, hence the need for the additional information in the notation.
-
-The pressure and velocity fields occuring in the computation of $d_r=p|_{z=z_r}=S[c]_{s,r}h_s$ form the solution of
-\begin{eqnarray}
-\label{eqn:awedata}
-\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
-h \delta(z-z_s)), \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
-p & =& 0 \mbox{ for } t \ll 0 \nonumber\\ 
-\bv & = & 0 \mbox{ for } t \ll 0 \nonumber\\
-\end{eqnarray}
-Actually $p,\bv$ form a {\em weak solution} of the system \ref{eqn:awedata}, defined by integration by parts.
-A more general setting for this type of system provides an advantageous framework within which to explain the properties of these weak solutions.
-
-Let $\omega \subset \bR^d$ be a smooth orientable surface. Then the distribution $\delta_{\omega}$ is well-defined: for $u \in C_0^{\infty}(\bR^d)$, 
-\[
-\langle \delta_{\omega}, u \rangle = \int_{\omega} u dS
-\]
-The system \ref{eqn:awedata} is a special case of
-\begin{eqnarray}
-\label{eqn:aweomega}
-\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
-h \delta_{\omega}, \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
-p & =& 0 \mbox{ for } t \ll 0 \nonumber\\ 
-\bv & = & 0 \mbox{ for } t \ll 0 \nonumber\\
-\end{eqnarray}
-in which $h \in C_0^{\infty}(\omega)$.
-
-\begin{thm}
-\label{thm:bvs}
-Suppose that $\kappa,\beta$ are smooth.
-Then the system \ref{eqn:aweomega} has a unique weak solution $(p,\bv)$.
-Denote by $[v_n] \in C^{\infty}(\omega)$ the jump of the normal component of $\bv$ across $\omega$:
-\[
-[v_n](\bx) = \lim_{\epsilon \rightarrow 0} (\bv(\bx + \epsilon {\bf n}(\bx))-\bv(\bx - \epsilon {\bf n}(\bx)))\cdot {\bf n}(\bx),\,\bx \in \omega.
-\]
-Then $h(\bx) = [v_n](\bx), \bx \in \omega$.
-
-%Let $p,\bv$ be the weak solution of the system \ref{eqn:awedata}, and assume that $\kappa, \beta$ are smooth.
-Denote by $[v_z]|_{z=z_s}$ the jump of the z-component of particle velocity in the z-direction:
-%\[
-%[v_z](x,z) = v_z^+(x,z)-v_z^-(x,z),  v_z^{\pm}(x,z)=\lim_{z'\rightarrow z^{\pm}} v_z(x,z')
-%\]
-%Then $[v_z](x,z)=0$ unless $z=z_s$, and
-%\[
-%h(x) = [v_z](x,z_s).
-%\]
-\end{thm}
-For the proof, see  Appendix B.
-
-For the rest of this section, $\omega$ will be a coordinate hyperplane, defined by $z=$ constant, with its usual orientation, and I will write $[v_z] = [v_b]$.
-
-To begin with, change notation slightly, and write $S[c]h_s$ rather than $S[c]g$ with $g = h_s \delta(z-z_s)$, as $z_s$ will be presumed ot be the same for all shots. 
-%The addition of the subscript indicates that the sources lie on the plane $z=z_s$ and the receivers are on the plane $z=z_r$.
-As will be apparent shortly, a similar operator with source and receiver planes swapped will also be important, hence the need for the additional information in the notation.
-
-The pressure and velocity fields occuring in the computation of $d_r=p|_{z=z_r}=S[c]_{s,r}h_s$ form the solution of
-\begin{eqnarray}
-\label{eqn:awedata1}
-\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
-h \delta(z-z_s)), \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
-p & =& 0 \mbox{ for } t \ll 0 \nonumber\\ 
-\bv & = & 0 \mbox{ for } t \ll 0 \nonumber\\
-\end{eqnarray}
-Actually $p,\bv$ form a {\em weak solution} of the system \ref{eqn:awedata1}, defined by integration by parts.
-\begin{thm}
-\label{thm:vz2}
-If $\kappa,\beta$ are constant near the $z=z_s$, then $v_z^-=-v_z^+$ so $[v_z] = 2 v_n^+$. 
-\end{thm}
-(for proof see Appendix 2). The formulation is for 2D propagation, but a precisely similar fact is true for 3D - in that case, $x$ is replaced everywhere by the horizontal coordinates.
-
-The import of Theorem \ref{thm:bvs} is that we can read off the source if we can infer the (particle) velocity field near the source surface, from knowledge of the data traces (values of the pressure at $z=z_r$).
-The key observation that enables this feat is a well-posedness theorem for the wave equation:
-
-\begin{thm} \label{thm:bvp}
-Suppose that $\omega \subset \bR^d$ is a smooth orientable surface, $\omega = \partial  \Omega^- = \partial \Omega^+$, where $\Omega^{\pm}$ are the open exterior/interior regions defined by the outward unit normal field of $\omega$, and let $d \in C^{\infty}_0(\omega \times \bR)$.
-Then there exist unique smooth solutions $p^{\pm},\bv^{\pm}$ of the initial/boundary value problems
-\begin{eqnarray}
-\label{eqn:awebdry}
-\frac{\partial p^{\pm}}{\partial t} & = & - \kappa \nabla \cdot \bv^{\pm} \mbox{ in } \Omega^{\pm}, \nonumber \\
-\frac{\partial \bv^{\pm}}{\partial t} & = & - \beta \nabla p^{\pm} \mbox{ in } \Omega^{\pm},\nonumber \\
-p^{\pm} & =& 0 \mbox{ for } t \ll 0\nonumber\\ 
-\bv^{\pm} & = & 0 \mbox{ for } t \ll 0 \nonumber\\
-\lim_{\bx \rightarrow \omega^{\pm}}p & = & d 
-\end{eqnarray}
-\end{thm}
-
-Putting the two theorems together, obtain
-\begin{cor}\label{thm:bvps}
-Suppose that $d_s$ is a sufficiently regular function, = 0 for $t<0$, and denote by $p^{\pm},\bv^{\pm}$ the half-space solutions provided by Theorem \ref{thm:bvp}.
-Define the vector fields $p,\bv$ in $\bR^4$ by
-\begin{eqnarray}
-\label{eqn:putemtogether} 
-p(x,z) &=& p^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
-p(x,z_s)&=&d_s(x)\\
-\bv(x,z) &=& \bv^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
-\end{eqnarray}
-Then $p,\bv$ is a weak solution of the system \ref{eqn:awedata1} with $h=[v]_n$.
-\end{cor}
-
-Now invoke the {\em outgoing assumption}: $p,\bv$ is the restriction to $z>z_s$ of a solution vanishing in $z<z_r$ for large $t$.
-This condition needs to be interpreted asymptotically, that is, $p,\bv$ is relatively smooth (lower order in frequency) for $t \gg 0, z<z_r$.
-It holds if the rays carrying significant energy in $p,\bv$ intersect $z=z_r$ transversally, and do not return to $z<z_r$ in forward time.
-The examples below will satisfy this condition, because the wave velocity is constant in $z > z_r$.
-
-Note that isotropic point sources do not produce fields outgoing in a half space, in the sense just described - simply because they radiate in all directions.
-However, for any specified finite time of propagation, it is possible to replace a point source by an anisotropic source that fits the point source data to within an arbitrarily small error, and does produce outgoing fields.
-In fact, the numerical examples below will construct such sources.
-
-This assumption implies that for $t$ sufficiently large, $d_r = p|_{z=z_r}$ is negligible for $t \gg 0$, so $p,\bv$ differs in $z_s<z < z_r$ by a negligible error from the solution of $p^-,\bv^-$, part of the solution of the  backwards-in-time  initial value problem, similar to the system \ref{eqn:awebdry}:
-\begin{eqnarray}
-\label{eqn:awetr}
-\frac{\partial \tilde{p}^{\pm}}{\partial t} & = & - \kappa \nabla \cdot \tilde{\bv}^{\pm}, \pm(z-z_r)>0 \nonumber \\
-\frac{\partial \tilde{\bv}^{\pm}}{\partial t} & = & - \beta \nabla \tilde{p}^{\pm},\nonumber \\
-\tilde{p}^{\pm} & =& 0 \mbox{ for } t \gg 0,\nonumber\\ 
-\tilde{\bv}{\pm} & = & 0 \mbox{ for } t \gg 0,\nonumber\\
-\lim_{z \rightarrow z_r^{\pm}}\tilde{p}^{\pm} &=&  d_r, \nonumber\\,
-\tilde{p} &=& \tilde{p}^{\pm}, \pm(z-z_r)>0, \nonumber\\, 
-\tilde{\bv} &=& \tilde{\bv}^{\pm}, \pm(z-z_r)>0.
-\end{eqnarray}
-
-This must be so, as $p,\bv$ satisfies the wave equation system (first two conditions), and the rest as well except for a negligible error.
-The last condition must be interpreted consistently, that is $\tilde{p}=0$ for $t$ large enough that $d_r$ is negligible.
-(In practice consistency will be obtained by replacing $t \gg 0$ respectively $t \ll \infty$ by $t > t_{\rm max}$, $t < t_{\rm max}$ for a suitably large $t_{\rm max}$).
-Then the standard energy estimate shows that $p, \tilde{p}$ differ also by a negligible error: in particular, $p|_{z=z_s} \approx \tilde{p}_{z=z_s}$.
-
-According to Corollary \ref{thm:bvps}, recovering $p_{z=z_s}$ entails recovering $h_s=[v_z]|_{z=z_s}$. So a procedure for computing an approximate inverse $S[c]_{s,r}^{\dagger}$ for $S[c]_{s,r}$ is 
-\begin{equation}
-\label{eqn:psinv}
-S[c]_{s,r}^{\dagger}d_r\equiv [v_z]_{z=z_s}.
-\end{equation}
-where the jump in $v_z$ at $z=z_s$ is computed from knowledge of $d_r=p|_{z=z_r}$ by
-\begin{itemize}
-\item solving \ref{eqn:awetr} backwards in time,
-\item reading off $d_s=\tilde{p}$ at $z=z_s$,
-\item solving \ref{eqn:awebdry} forwards in time, and computing $[v_z]|_{z=z_s}$, or if appropriate ($\kappa$ and $\beta$ constant near $z=z_s$) simply using $\bar{v}_{z=z_s^+} = v_z^{+}=-v_z^{-}$ at $z=z_s$ (Theorem \ref{thm:bvs}).
-\end{itemize}
-
-The relations developed in the last few paragraphs may be summarized as follows:
-\begin{equation}
-\label{eqn:appinv}
-d_r = S[c]_{s,r}h_s \,\,\Leftrightarrow \,\, h_s \approx S[c]_{s,r}^{\dagger}d_r
-\end{equation}
-that is, $S[c]_{s,r}^{\dagger}$ is an approximate inverse of $S[c]_{s,r}$.
-
-Inspection of systems \ref{eqn:awebdry} and \ref{eqn:awetr} reveals that they are actually the same, except for the direction of time, and that observation leads to another useful relation.
-Denote by $R$ the time-reversal operator:
-\[
-Ru(\bx,t)=u(\bx,-t)
-\]
-and by {\cal R} the time reversal operator for acoustic fields:
-\[
-{\cal R}\left(\begin{array}{c}
-p\\
-\bv
-\end{array}
-\right) = 
-\left(\begin{array}{c}
-Rp\\
--R\bv
-\end{array}
-\right) 
-\]
-The sign in the definition of ${\cal R}$ is chosen so that if $(\tilde{p},\tilde{\bv})^T$ solves the system \ref{eqn:awetr}, then ${\cal R}(\tilde{p},\tilde{\bv})^T=(\hat{p},\hat{\bv})^T$ is the solution of
-\begin{eqnarray}
-\label{eqn:awetrs}
-\frac{\partial \hat{p}^{\pm}}{\partial t} & = & - \kappa \nabla \cdot \hat{\bv}^{\pm}, \pm(z-z_r) > 0 \nonumber \\
-\frac{\partial \hat{\bv}^{\pm}}{\partial t} & = & - \beta \nabla \hat{p}^{\pm},\nonumber \\
-\hat{p}^{\pm} & =& 0 \mbox{ for } t \ll 0,\nonumber\\ 
-\hat{\bv}^{\pm} & = & 0 \mbox{ for } t \ll 0,\nonumber\\
-\lim _{z=z_r^{\pm}}\hat{p}^{\pm} &=&  Rd_r,\nonumber\\
-\hat{p} &=& \hat{p}^{\pm}, \pm(z-z_r)>0, \nonumber\\, 
-\hat{\bv} &=& \hat{\bv}^{\pm}, \pm(z-z_r)>0.
-\end{eqnarray}
-
-%For a smooth orientable surface $\omega$, denote by $\Lambda[c,\omega]$ the operator that maps Dirichlet pressure data $d$ on $\omega \times \bR$ to the jump in normal velocity $[v_n]$ across $\omega \times \bR$ via solution of the system \ref{eqn:awebdry}.
-
- From Corollary \ref{thm:bvps} applied to the system \ref{eqn:awetrs}, $\hat{p}|_{z=z_s} = S[c]_{r,s}\Lambda[c]_rRd_r$.
-Denote by $\Lambda[c]_r$ the same operator with $z_s$ replaced by $z_r$.
-Then $\hat{p}|_{z=z_s} = S[c]_{r,s}\Lambda[c]_rRd_r=R\tilde{p}|_{z=z_s}$.
-Thus the definition \ref{eqn:psinv} implies
-\begin{equation}
-\label{eqn:psinvrev}
-S[c]_{s,r}^{\dagger}= -\Lambda[c]_sRS[c]_{r,s}R\Lambda[c]_r
-\end{equation}
-since $\Lambda[c]_rR=-R\Lambda[c]_r$ from the definition of ${\cal R}$.
-Define $T[c]_{s,r}d_s = p|_{z=z_r}$ for the solution of \ref{eqn:awebdry}.
-Since ${\cal R}^2=I$, Then $\hat{p}|_{z=z_s} = T[c]_{r,s}Rd_r$.
-Since ${\cal R}^2=I$, ${\cal R}(\hat{p},\hat{\bv})^T=(\tilde{p},\tilde{\bv})^T$, whence $\tilde{p}|_{z=z_s}=RT[c]_{r,s}Rd_r$.
-
-Denote by $\Lambda[c]_s$ the operator that produces the jump in normal velocity $[v_z]=\lim_{z \rightarrow z_s^+}v_z^+ - \lim_{z\rightarrow z_s^-} v_z^-$ from Dirichlet pressure data $d_s$ and causal solution of the system \ref{eqn:awebdry}.
-From Corollary \ref{thm:bvps}, $T[c]_{s,r} = S[c]_{s,r}\Lambda[c]_s$.
-Denote by $\Lambda[c]_r$ the same operator with $z_s$ replaced by $z_r$.
-Then $\hat{p}|_{z=z_s} = S[c]_{r,s}\Lambda[c]_rRd_r=R\tilde{p}|_{z=z_s}$.
-Thus the definition \ref{eqn:psinv} implies
-\begin{equation}
-\label{eqn:psinvrev}
-S[c]_{s,r}^{\dagger}= -\Lambda[c]_sRS[c]_{r,s}R\Lambda[c]_r
-\end{equation}
-since $\Lambda[c]_rR=-R\Lambda[c]_r$ from the definition of ${\cal R}$.
-
-As I will show in the example section, the approximation may be accurate enough without any further improvement.
-However, a closer examination shows that in fact it is possible to express the relation between $S[c]_{s,r}$ and $S[c]_{s,r}^{\dagger}$ in terms of a change of norm, that is, in terms of a preconditioner, and this relation can be used to considerably accelerate an iterative solution of the normal equation \ref{eqn:normal}.
-
-To see this, compute the formal or $L^2$ adjoint $S[c]_{s,r}^T$, by a variant of the adjoint state method, in this case a by-product of the conservation of energy.
-Suppose that $\bar{h}_r$ is a function on $z=z_r$ (data traces), and
-$\bar{p},\bar{\bv}$ solve the backwards-in-time boundary value problem:
-\begin{eqnarray}
-\label{eqn:aweadj}
-\frac{\partial \bar{p}}{\partial t} & = & \kappa (-\nabla \cdot \bar{\bv} + 
- \bar{h}_r \delta(z-z_r))\nonumber \\
-\frac{\partial \bar{\bv}}{\partial t} & = & - \beta \nabla \bar{p},\nonumber \\
-\bar{p} & =& 0 \mbox{ for } t \gg 0\nonumber\\ 
-\bar{\bv} & = & 0 \mbox{ for } t \gg 0 
-\end{eqnarray} 
-Then
-\[
-0 = 
-\left(\int\, dx\,dy\,dz\, \frac{p \bar{p}}{\kappa} +  
-\frac{\bv \cdot \bar{\bv}}{\beta} \right)|_{t=T}
--
-\left(\int\, dx\,dy\,dz\, \frac{p \bar{p}}{\kappa} +  \frac{\bv \cdot \bar{\bv}}{\beta} \right)|_{t=0}
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \bar{p}}{\kappa} +  \frac{\bv \cdot \bar{\bv}}{\beta} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \bar {p} +  p \frac{1}{\kappa}\frac{\partial \bar{p}}{\partial t} \right.
-\]
-\[
-+
-\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \bar{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \bar{\bv}}{\partial t} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
- h_s \delta(z-z_s)\right) \bar{p} + p \left(- \nabla \cdot \bar{\bv} + 
- d_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  (- \nabla p) \cdot \bar{\bv} + \bv \cdot (-\nabla \bar{p}) \right)
-\]
-\[
-= 
-\int_{0}^{T}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
- h_s \delta(z-z_s)\right) \bar{p} + p \left(- \nabla \cdot \bar{\bv} + 
- d_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  p (\nabla \cdot \bar{\bv}) + (\nabla \cdot \bv) \bar{p} \right)
-\]
-after integration by parts in the last two terms. Most of what is left cancels, leaving 
-\[
-0 = \int\,dt\,dx\,dy\, (h_s \bar{p})_{z=z_s} + (p  \bar{h}_r)_{z=z_r}
-\]
-whence
-\begin{equation}
-\label{eqn:sadj}
- S[c]_{s,r}^T \bar{h}_r = -\bar{p}|_{z=z_s}
-\end{equation}
-Compare the systems \ref{eqn:awetr} and \ref{eqn:aweadj}: they differ in that 
-\begin{itemize}
-\item in \ref{eqn:awetr}, $d_r$ appears as a Dirichlet condition on $z=z_r$,
-\item in \ref{eqn:aweadj}, $\bar{h}_r$ appears as a distributed source on $z=z_r$.
-\end{itemize}
-According to Corollary \ref{thm:bvps}, $\bar{p}=\bar{p}, \bar{\bv}=\bar{\bv}$ if $\bar{h}_r = [\bar{v}_z]|_{z=z_r}$, i.e.
-the source is that which generates the same solution as the Dirichlet condition..
-
-Simlarly denote by $\tilde{\Lambda}[c]_r$ the operator that produces the jump in normal velocity at $z=z_r$ from pressure data $d_r$ via {\em anticausal} solution of the system 
-\begin{eqnarray}
-\label{eqn:awebdrytr}
-\frac{\partial {\tilde p}^{\pm}}{\partial t} & = & - \kappa \nabla \cdot {\tilde \bv}^{\pm} \mbox{ in } \pm(z-z_r) > 0, \nonumber \\
-\frac{\partial {\tilde \bv}^{\pm}}{\partial t} & = & - \beta \nabla {\tilde p}^{\pm} \mbox{ in } \pm(z-z_r) > 0,\nonumber \\
-{\tilde p}^{\pm} & =& 0 \mbox{ for } t \gg 0\nonumber\\ 
-{\tilde \bv}^{\pm} & = & 0 \mbox{ for } t \gg 0 \nonumber\\
-\lim_{z\rightarrow z_r^{\pm}}{\tilde p} & = & {\tilde d}_r 
-\end{eqnarray}
-%\end{thm}
- for the receiver surface $z=z_r$ (in the math literature, this is the ``Dirichlet-to-Neumann operator'', more or less). The discussion so far can be summarized as follows:
-\begin{equation}
-\label{eqn:wadj}
-\bar{p}|_{z=z_s} = S[c]_{s,r}^T\Lambda[c]_r \bar{p}|_{z=z_r} 
-\end{equation}
-According to the discussion above, $\bar{p}|_{z=z_s} \approx p|_{z=z_s}$, and $h_s = \Lambda[c]_s p|_{z=z_s} = S[c]_{s,r}^{\dagger}d_r$. That is,
-\begin{equation}
-\label{eqn:appinv}
-S[c]_{s,r}^{\dagger} = \Lambda[c]_s S[c]_{s,r}^T \Lambda[c]_r
-\end{equation}
-This relation can be re-written in the useful form
-\begin{equation}
-\label{eqn:adj}
-S[c]_{s,r}^{\dagger} = W_m[c]^{-1}S[c]_{s,r}^TW_d[c]
-\end{equation}
-in which 
-\begin{equation}
-\label{eqn:weights}
-W_m[c] = \Lambda[c]_s^{-1}\,\, W[c]_d = \Lambda[c]_r
-\end{equation}
-Since $\Lambda[c]$ is positive (semi-)definite, equation \ref{eqn:adj} exhibits $S[c]_{s,r}^{\dagger}$ as the adjoint of $S[c]_{s,r}$ with respect to weighted norms 
-\begin{itemize}
-\item on the source space ($h_s$): weight $W[c]_m$
-\item on the data space ($d_r$): weight $W[c]_d$
-\end{itemize}
-Since $S[c]_{s,r}^{\dagger}S[c]_{s,r} \approx I$ as shown above, $S[c]_{s,r}$ is approximately unitary in the Hilbert spaces of source and data traces with norms defined by the weighting operators $W_m$ and $W_d$ respectively.
-Therefore a Krylov space method employing these norms will converge rapidly.
-
-\cite{HouSymes:EAGE16} demonstrated a very similar preconditioner for Least Squares Migration, also for its subsurface offset extension \cite[]{HouSymes:16}, motivated by \cite{tenKroode:12}.
-These constructions all involve the Dirichlet-to-Neumann operator.
-This concept also turns up in hidden form in the work of Yu Zhang and collaborators on true amplitude migration \cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,ZhangYuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.
-
-\subsection{Implementation Considerations}
-The Dirichlet-to-Neumann operator $\Lambda$ is an essential part of the inner problem preconditioner just presented. Implementation can be accomplished in several ways:
-\begin{itemize}
-\item \cite{tenKroode:12} suggests using a one-way operator;
-\item if both pressure and normal partical velocity are measured (or simulated), then the two are related by $\Lambda$ and the velocity component can simply be used as the output;
-\item presence of a free surface implies all of the usual problems, such as the need for removal of receiver-side ghosts.
-On the other hand, if the free surface is within a quarter-wavelength throughout the useful bandwidth of the data, then the ghosted data differs from $\Lambda p$ by a time integration and a scale factor, a fact used to good effect by \cite{HouSymes:15}.
-\end{itemize}
-
-In the examples, I have used the second observation.
-With a finite difference implementation of the pressure-velocity system \ref{eqn:awe}, velocity components are available ``for free'', short-circuiting explicit computation of $\Lambda$.
-
-The variable projection gradient formula \ref{eqn:l2grad} has an Achilles heel, not mentioned so far: it is correct only when the normal equation \ref{eqn:normal} is satisfied {\em exactly}.
-Any iterative method (or the asymptotic inverse developed in the last subsection) solves \ref{eqn:normal} only approximately, so a nonzero error term is neglected in using \ref{eqn:l2grad}.
-This error term can be large, because it amplifies the high frequencies in the solution error, in fact depends on derivatives of the solution.
-For pure transmission through transparent (smooth) material models, the approximate inverse $S[c]_{s,r}^{\dagger}$ is asymptotically exact (at least in principle), and that fact can be used to devise a modification of \ref{eqn:l2grad} that converges with a convergent iteration for \ref{eqn:normal}.
-[LAST YEAR TRIP REPORT, MAY PUT SOMETHING IN APPENDIX]
-
-Choice of the penalty weight $\alpha$, and of region in which the extended source $g$ is permitted to be nonzero, must be managed.
-These two choices are linked.
-Lei Fu's thesis \cite[]{Fu:Geo17,Fu:Geo17b} explains and tests algorithms for similar choices that occur in subsurface offset WEMVA.
-All of these ideas apply {\em ipso facto} to the surface source extension.
-
-\subsection{Practical Preconditioning}
-
-The formula \ref{eqn:adj} allows direct use of the conjugate gradient (or other Krylov space iterative) algorithm with the inner products described above.
-This is however computationally inconvenient, as it requires explicit computation of the weight operators, in this case the Dirichlet-to-Neumann operator.
-There are several ways to carry out this computation, as mentioned somewhere, but it is in any case a  pain.
-On the other hand, it is possible to rearrange the calculation with weighted norms into the form of the {\em preconditioned conjugate gradient (PCG) algorithm}.
-For the current problem, identify $A=S[c]_{s,r}$, $A^* = S[c]_{s,r}^T W_d$.
-PCG takes the form
-\begin{algorithm}[H]
-\caption{Preconditioned Conjugate Gradient Algorithm}
-\begin{algorithmic}[1]
-\State Choose $x_0$ somehow
-  \State $r_0 \gets A^*(b-Ax_0)$
-  \State $p_0 \gets W_m^{-1}r_0$
-  \State $g_0 \gets p_0$
-  \State $q_0 \gets A^*Ap_0$
-  \State $k \gets 0$
-  \Repeat
-  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle p_k,q_k\rangle}$
-  \State $x_{k+1} \gets x_k + \alpha_k p_k$
-  \State $r_{k+1} \gets r_k - \alpha_kq_k$
-  \State $g_{k+1} \gets W_m^{-1}r_{k+1}$
-  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
-  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
-  \State $q_{k+1} \gets A^*Ap_{k+1}$
-  \State $k \gets k+1$
-  \Until{Error is sufficiently small, or max iteration count exceeded} 
-\end{algorithmic}
-\end{algorithm}
-Of course, this variant of CG merely hides the data sides weight operator, and requires application on the model side. This is still inconvenient.
-
-The simplest solution is to use the coexistence of the $d_r=p$ and $\Lambda[c]_rd_r =[v_z]=2v_z$ traces (assuming a horizontal datum), and similarly for the source datum.
-That is, the computation of $\Lambda[c]_r$ is intrinsic: when $S[c]_{s,r}h_s$ has been computed, so has $\Lambda[c]_rS[c]h_s$ - it is simply the collection of corresponding vertical velocity traces.
-Similarly,  $\Lambda[c]_sS[c]^T$ is the collection of vertical velocity traces at $z=z_s$ arising from the time-reversal computation input equal to sources traces on $z=z_r$.
-If we can rewrite the PCG loop to require only these computations, then the acoustic simulations in forward and reverse time will suffice to compute the step.
-
-In the loop body above, introduce the vector sequences $t_k=Ap_k,v_k=W_d t_k, s_k=W_m^{-1}q_k$, and re-write the iteration by unfolding $W_m^{-1}S[c]^T W_d S[c]$ and noting that $\langle p_k, q_k \rangle =\langle p_k,A^TW_dAp_k \rangle =\langle Ap_k,W_dAp_k \rangle =\langle t_k,v_k\rangle$:
-\begin{algorithm}[H]
-\caption{Modified Preconditioned Conjugate Gradient Algorithm}
-\begin{algorithmic}[1]
-\State Choose $x_0$ somehow
-  \State $r_0 \gets A^*(b-Ax_0)$
-  \State $p_0 \gets W_m^{-1}r_0$
-  \State $g_0 \gets p_0$
-  \State $k \gets 0$
-  \Repeat
-  \State $t_k \gets Ap_k$
-  \State $v_k \gets W_dt_k$
-  \State $q_k \gets A^Tv_k$
-  \State $s_k \gets W_m^{-1}q_k$
-  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle t_k,v_k\rangle}$
-  \State $x_{k+1} \gets x_k + \alpha_k p_k$
-  \State $r_{k+1} \gets r_k - \alpha_kq_k$
-  \State $g_{k+1} \gets g_k - \alpha_k s_k$
-  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
-  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
-  \State $k \gets k+1$
-  \Until{Error is sufficiently small, or max iteration count exceeded} 
-\end{algorithmic}
-\end{algorithm}
-This rewrite does not look like an improvement - four operator applications appear to be required:
-\begin{itemize}
-\item $p_k \rightarrow S[c] p_k = t_k$
-\item $p_k \rightarrow  \Lambda[c]_r S[c] p_k = v_k$
-\item $v_k \rightarrow S[c]^Tv_k = q_k$
-\item $v_k \rightarrow \Lambda[c]_s S[c]^T v_k = s_k$
-\end{itemize}
-However this rearrangement reveals that actually only two wave equation solves are needed.
-For $q_k$ and $s_k$, both by-products of the adjoint (time-reversed) simulation: $q_k$ stores the pressure traces on $z=z_s$, $s_k$ the velocity traces.
-The same is true of $t_k$ and $v_k$, with the latter being the input for the $q_k,s_k$ computation.
-
-Denote by $F_{pp}$ the acoustic simulation operator with a $p$ source (constitutive law defect, as in the system \ref{eqn:awedata}), and $p$ output traces.
-Similarly, denote by $F_{pv}$ the simulation operator with $p$ source and $v_z$ output traces, likewise $F_{vp}$ and $F_{vv}$, and by ${\bf F}$ the matrix operator
-\begin{equation}
-\label{eqn:iwop}
-{\bf F} = \left(
-\begin{array}{cc}
-F_{pp} & F_{vp} \\
-F_{pv} & F_{vv}
-\end{array}
-\right)
-\end{equation}
-The {\tt IWaveLOVOp} built in the asg package implements this matrix operator if the keywords {\tt source\_p, source\_v0, data\_p, data\_v0} are defined in its parameter list: these keywords point to SEGY trace files that define the spaces in which the $p$ and $v_z$ sources respectively data traces are treated as vectors.
-That is, IWAVE provides an easy access to ${\bf F}$.
-In terms of the operators used in the statement of the inverse problem, $F_{pp}=S[c]$, $F_{pv}=\Lambda[c]_rS[c]$, $F_{vp} = S[c]\Lambda[c]_s$, $F_{pp}^T = S[c]^T$, and $F_{vp}^T = \Lambda[c]_sS[c]^T$.
-So if we define
-\begin{equation}
-\label{eqn:vecvec}
-{\bf p}_k = \left[
-\begin{array}{c}
-p_k\\
-0
-\end{array}
-\right],\,\,
-{\bf r}_k = \left[
-\begin{array}{c}
-r_k\\
-g_k
-\end{array}
-\right],\,\,
-{\bf t}_k = \left[
-\begin{array}{c}
-t_k\\
-v_k
-\end{array}
-\right],\,\,
-{\bf v}_k = \left[
-\begin{array}{c}
-v_k\\
-0
-\end{array}
-\right],\,\,
-{\bf q}_k = \left[
-\begin{array}{c}
-q_k\\
-s_k
-\end{array}
-\right],
-\end{equation}
-and
-\begin{equation}
-\label{eqn:proj}
-{\bf P} = \left(
-\begin{array}{cc}
-0 & I \\
-0 & 0
-\end{array}
-\right),
-\end{equation}
-then the algorithm listing above can be rewritten as
-\begin{algorithm}[H]
-\caption{Further modified Preconditioned Conjugate Gradient Algorithm}
-\begin{algorithmic}[1]
-\State Choose $x_0$ somehow
-  \State $r_0 \gets A^TW_d(b-Ax_0)$
-  \State $p_0 \gets W_mr_0$
-  \State $g_0 \gets p_0$
-  \State $k \gets 0$
-  \Repeat
-  \State ${\bf t}_k \gets {\bf F}{\bf p}_k$
-  \State ${\bf v}_k \gets {\bf P}{\bf t}_k$
-  \State ${\bf q}_k \gets {\bf F}^T{\bf v}_k$
-  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle t_k,v_k\rangle}$
-  \State $x_{k+1} \gets x_k + \alpha_k p_k$
-  \State ${\bf r}_{k+1} \gets {\bf r}_k - \alpha_k{\bf q}_k$
-  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
-  \State ${\bf p}_{k+1}\gets {\bf P}{\bf r}_{k+1}+\beta_{k+1}{\bf p}_k$
-  \State $k \gets k+1$
-  \Until{Error is sufficiently small, or max iteration count exceeded} 
-\end{algorithmic}
-\end{algorithm}
-
-Note that
-\begin{equation}
-\label{eqn:note}
-{\bf r}_0 = \left[
-\begin{array}{c}
-r_0\\
-g_0 
-\end{array}
-\right] = 
-\left[
-\begin{array}{c}
-r_0\\
-p_0 
-\end{array}
-\right] = 
-\left[
-\begin{array}{c}
-A^TW_d(b-Ax_0)\\
-W_m^{-1}A^TW_d(b-Ax_0)
-\end{array}
-\right]
-\end{equation}
-Define 
-\begin{equation}
-\label{eqn:rhsdef}
-{\bf b}
-= 
-\left[
-\begin{array}{c}
-W_d(b-Ax_0)\\
-0
-\end{array}
-\right].
-\end{equation}
-Then identity \ref{eqn:note} implies that
-\begin{equation}
-\label{eqn:init}
-{\bf r}_0= {\bf F}^T{\bf b},\,\, {\bf p}_0 = {\bf P}{\bf r}_0.
-\end{equation}
-This observation allows a convenient ``hoist'' of the $k=0$ operator applications out of the iteration loop.
-Also, it is convenient to introduce the indefinite quadratic form that returns the inner product of the two components of a vector in the product space introduced above:
-\begin{equation}
-\label{eqn:dotfunc}
-Q({\bf x})=\langle x_0,x_1\rangle=\left\langle \bx,
-\left(
-\begin{array}{cc}
-0 & I \\
-I & 0 
-\end{array}
-\right) \bx \right\rangle
-\end{equation}
-and the scalar sequences
-\begin{eqnarray}
-\label{eqn:scalars}
-\gamma_k &=& \langle r_k,g_k \rangle = Q({\bf r})\\
-\delta_k &=& \langle t_k,v_k \rangle = Q({\bf t})
-\end{eqnarray}
-
-With these modifications, we obtain 
-\begin{algorithm}[H]
-\caption{Intrinsically Preconditioned Conjugate Gradient Algorithm}
-\begin{algorithmic}[1]
-\State Choose $x_0$ somehow
-\State ${\bf b} \gets \left[
-\begin{array}{c}
-W_d(b-Ax_0)\\
-0
-\end{array}
-\right]$
-  \State ${\bf r}_0 \gets {\bf F}^T{\bf b}$
-  \State ${\bf p}_0 \gets {\bf P}{\bf r}_0$
-  \State $\gamma_0 \gets Q({\bf r}_0)$
-  \State $k \gets 0$
-  \Repeat
-  \State ${\bf t}_k \gets {\bf F}{\bf p}_k$
-  \State $\delta_k \gets Q({\bf t})$
-  \State ${\bf v}_k \gets {\bf P}{\bf t}_k$
-  \State ${\bf q}_k \gets {\bf F}^T{\bf v}_k$
-  \State $\alpha_k \gets \frac{\gamma_k}{\delta_k}$
-  \State $x_{k+1} \gets x_k + \alpha_k p_k$
-  \State ${\bf r}_{k+1} \gets {\bf r}_k - \alpha_k{\bf q}_k$
-  \State $\gamma_{k+1} \gets Q({\bf r}_{k+1})$
-  \State $\beta_{k+1} \gets \frac{\gamma_{k+1}}{\gamma_k}$
-  \State ${\bf p}_{k+1}\gets {\bf P}{\bf r}_{k+1}+\beta_{k+1}{\bf p}_k$
-  \State $k \gets k+1$
-  \Until{Error is sufficiently small, or max iteration count exceeded} 
-\end{algorithmic}
-\end{algorithm}
-
-\section{Prototype Numerical Examples}
-\inputdir{project}
-
-I present a collection of simple examples that illustrate the features of the surface source extension claimed in preceding sections.
-
-I used the IWAVE acoustic staggered grid package to carry out these calculations.
-This package implements (2,2k) schemes for k=1,2,..., and outputs traces (of either velocity or pressure at any point in space via multilinear interpolation.
-The discretized modeling operator is thus of second order accuracy, though as usual I have used higher order in space to reduce grid dispersion.
-
-Source injection is implemented as the adjoint of trace sampling, resulting in another second-order error [REFERENCES].
-
-The data is a single shot gather, with a source at coordinates $x_s=z_s=3000$ m (units of length are meters in all cases).
-The receiver line occupies $1500 \le x_r \le 5500$ m, with receiver depth $z_r=1000$ m.
-Extended sources occupy $1500 \le x_s \le 5500$ m, with the same depth $z_s=3000$ m as the ``physical'' source used to generate the data.
-This region turns out to be adequate to represent the extended sources that approximately invert the data, for the cases examined below.
-An algorithm to automatically identify an appropriate region can be based on the ideas developed by \cite{Fu:Geo17}.
-
-I have used absorbing boundary conditions (split-field PML) on all four sides of the 4000 m (vertical) $\times$ 8000 m (horizontal) simulation domain.
-Evidently includion of a free surface is important to the application of the ideas explained here to diving wave marine data, and I have not addressed the necessary modifications here.
-
-The script is set up to carry out the necessary computations on grids with spacings $\Delta x = \Delta z = $ 20, 10, and 5 m, with a jump of roughly 8 in computation time resulting from each refinement.
-For present purposes, the coarsest (20 m) grid seems to be sufficient, and that is the grid used in the examples presented below.
-The source pulses are chosen so that the computation is reasonably accurate.
-For the 20 m grid case, I use a zero-phase trapezoidal bandpass filter source with corner frequencies of 1.0, 2.0, 7.5, and 12.5 Hz.
-
-The IWAVE asg driver has been set up to recognize the case {\tt deriv=0} as defining the map from source (right-hand side in the pressure equation) to data (pressure) traces.
-The adjoint to this map, as explained above, is reverse-time propagation of the data traces as pressure sources, followed by scaling (formula \ref{eqn:sadj}).
-The approximate inverse is computed by application of the Dirichlet-to-Neumann operator to the pressure traces to produce corresponding velocity traces, followed by injection as pressure sources and reverse time propagation, followed by another application of the Dirichlet-to-Neumann map and scaling (formulas \ref{eqn:appinv}, \ref{eqn:adj}).
-
-The script for the first set of examples implements these operatations step-by-step via calls to IWAVE, Madagascar, and SU commands.
-A peculiarity of the {\tt asg.x} driver needs to be mentioned: it is based on an un-scaled version of the constitutive law defect source representation, that is, {\tt asg.x} approximately computes the solution of the system \ref{eqn:awedata} with the first equation replaced by
-\begin{equation}
-\label{eqn:asgdata}
-\frac{\partial p}{\partial t}  =  - \kappa \nabla \cdot \bv +
-h \delta(z-z_s).
-\end{equation}
-Denote by $S_{\rm asg}[c]$ the forward map produced by {\tt asg.x}. Then comparison of \ref{eqn:awedata} and \ref{eqn:asgdata} reveals that
-\begin{equation}
-\label{eqn:sreln}
-S[c]=S_{\rm asg}[c]\kappa
-\end{equation}
-where $\kappa$ is shorthand for the operator of multiplication by $\kappa$. Accordingly, and approximate inverse for $S_{\rm asg}[c]$ is
-\[
-I \approx S_{\rm asg}[c]^{\dagger}S_{\rm asg}[c] = S_{\rm asg}[c]^{\dagger}S[c]\kappa^{-1}
-\]
-Since $S[c]S[c]^{\dagger} \approx I \approx S[c]^{\dagger}S[c]$, it follows that
-\[
-S_{\rm asg}[c]^{\dagger}=\kappa S[c]^{\dagger} = \kappa\Lambda[c]_s S[c]^T \Lambda[c]_r 
-\]
-\begin{equation}
-\label{eqn:asginv}
-= \kappa \Lambda[c]_s \kappa S_{\rm asg}[c]^T\Lambda[c]_r
-\end{equation}
-This is the approximate inverse computed in the examples.
-Note that only the values of $\kappa$ near the source datum $z=z_s$ play a role in the relation \ref{eqn:sreln} or in the definition \ref{eqn:asginv} of the IWAVE ASG approximate inverse.
-
-These examples are based on two different material models.
-The first is a homogeneous model, with $\kappa=4.0$ GPa and $\rho=1.0$ g/cm$^{3}$.
-Figure \ref{fig:ptpwindh0} shows the shot gather computed in the configuration descritbed above for this choice of propagation medium.
-As mentioned in the preceding section, the velocity traces can be recorded in the same IWAVE workflow: these are displayed in
-\ref{fig:ptvzwindh0}.
-
-The first group of plots show pressure traces for source position $z_s=3$ km, $x_s=3$ km, 201 receivers at $z_r=1$ km, $2 \le x_r \le 6$ km.
-
-\plot{ptpwindh0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km.
-Homogeneous material model, $\kappa=4$ GPa, $\rho$=1 g/cm$^{3}$.
-Coarse grid: $dx=dz=20$ m, $dt=$ 8 ms, (2,4) staggered grid scheme.
-}
-\plot{ptvzwindhh0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindh0}.}
-\plot{srcvzghh0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindhh0} by time reversal (approximate inverse) algorithm explained in text.}
-\plot{reptpwindhh0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzghh0}.} 
-\plot{trcftrhh0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindh0}, \ref{fig:reptpwindhh0}.}.
-
-\plot{bml0}{width=15cm}{Inhomogenous bulk modulus field, with low velocity lens.}
-\plot{ptpwindl0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km.
-bulk modulus with low velocity lens as in Figure \ref{fig:bml0}, $\rho$=1 g/cm$^{3}$.
-Coarse grid: $dx=dz=20$ m, $dt=$ 8 ms, (2,4) staggered grid scheme.
-}
-\plot{ptvzwindll0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindl0}.}
-
-\plot{srcvzglh0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindll0} by time reversal (approximate inverse) algorithm explained in text, propagation in homogeneous material model (that is, ``wrong velocity'').}
-\plot{reptpwindlh0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzglh0}, both approximate inversion and re-simulation in  homogeneous material model (``wrong velocity'').}
-\plot{trcftrlh0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindl0}, \ref{fig:reptpwindlh0}.}
-
-\plot{srcvzgll0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindll0} by time reversal (approximate inverse) algorithm explained in text, propagation in bulk modulus of Figure \ref{fig:bml0} (that is, ``correct velocity'').}
-\plot{reptpwindll0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzgll0}, both approximate inversion and re-simulation in bulk modulus of Figure \ref{fig:bml0} (``correct velocity'').}
-\plot{trcftrll0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindl0}, \ref{fig:reptpwindll0}.}
-
-\bibliographystyle{seg}
-\bibliography{../../bib/masterref}
-
 \append{RWI (=Wang-Yingst WRI) - a Linear Algebra Viewpoint}
@@
 It could be handled by storing the output of $L[c]^{-1}_-$, as the IONGeo folks apparently do, or storing boundary data (for acoustics - as usual, this would not work if the approach is extended to a lossy model such as viscoacoustics), or using a checkpointing scheme.
-
-Another approach: write $w = L[c]^{-1}_-P^T h$. The object is to compute $u=L[c]^{-1}_+w$, that is, the solution of 
-\[
-(c^{-2}\partial_t^2 - \nabla^2)u(t) = w(t), 0\le t \le T; \,u(0) = \partial_t u(0) = 0
-\]
-Duhamel's principle would provide a solution in the form of integrals over time of solutions of homogeneous initial value problems:
-\[
-u(t) = \int_0^t d\tau v(t,\tau), \, (c^{-2}\partial_t^2 - \nabla^2)v(t,\tau) = 0, 0\le t \le T; \,v(t,\tau)_{t=\tau} = 0, \partial_t v(t,\tau)_{t=\tau} = w(\tau)
-\]
-Note that $w$ solves the homogenous wave equation except at $z=0$. That makes it tempting to use $w$ directly to define $v$. In fact, set 
-\[
-W(t) = -int_t^T ds w(s),\,v_0(t,\tau) = \frac{1}{2}(W(t) - W(2\tau-t)) 
-\]
-Then $v_0$ satisfies the Duhamel initial conditions. Note that because $w$ has zero Cauchy data at $t=T$,
-\[
-\partial_t W(t)= -\int_t^T ds \partial_t w(s),\,\partial_t^2 W(t)=-\int_t^T ds \partial^2_t w(s)
-\]
-so 
-\[
-(c^{-2}\partial_t^2 - \nabla^2)v_0(t,\tau) = -\int_t^T(c^{-2}\partial^2_t -\nabla^2)w + \int_{2\tau-t}^T(c^{-2}\partial^2_t -\nabla^2)
-\]
-For $t>\tau$, $t > 2\tau-t$, so 
-\[
-= -\int_{2\tau-t}^t(c^{-2}\partial^2_t -\nabla^2)w = -\int_{2\tau-t}^t P^Th
-\]
-and 
-\[ 
-v_0(t,\tau) = \int_{2\tau-t}^t w  
-\] 
-Consequently $v=v_0 + v_1$, in which $v_1$ is defined by 
-\[
-(c^{-2}\partial_t^2 - \nabla^2)v_1(t,\tau) = \int_{2\tau-t}^t P^Th, \,  v_1(t,\tau)_{t=\tau} = 0, \,  \partial_t v_1(t,\tau)_{t=\tau} = 0
-\]
-
-\append{Transpose of Full Acoustic Modeling Operator}
-In this section I compute the adjoint operator form modeling with both pressure and velocity source inputs and pressure and velocity trace outputs. The forward modeling operator is defined by solving
-\begin{eqnarray}
-\label{eqn:fawedata}
-\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
-h_s \delta(z-z_s)), \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta ( \nabla p + l_s {\bf e}_z\delta(z-z_s)),\nonumber \\
-p & =& 0 \mbox{ for } t<0\nonumber\\ 
-\bv & = & 0 \mbox{ for } t<0 \nonumber\\
-\end{eqnarray}
-The new elements here are the point load time function $l_s(\bx,t;\bx_s)$, and the vertical unit vector ${\bf e}_z=(0,1)^T$ for 2D, $=(0,0,1)^T$ for 3D.
-The forward map ${\bf F}$ is defined as implicit in the discussion of the Intrinsically Preconditioned CG algorithm: let ${\bf h}_s=(h_s,l_s)^T$, then
-\begin{equation}
-\label{eqn:ffwd}
-{\bf F}{\bf h}_s =\left\{ \left[
-\begin{array}{c}
-p(\bx,t;\bx_s)\\
-v_z(\bx,t;\bx_s)
-\end{array}
-\right]_{z=z_r}\right\}
-\end{equation}
-
-Let $\tilde{\bf h}_r=(\tilde{h}_r,\tilde{l}_r)^T$ be a function on $z=z_r$, 
-and 
-$\tilde{p},\tilde{\bv}$ solve the backwards-in-time boundary value problem:
-\begin{eqnarray}
-\label{eqn:faweadj}
-\frac{\partial \tilde{p}}{\partial t} & = & \kappa (-\nabla \cdot \tilde{\bv} + 
- \tilde{h}_r \delta(z-z_r))\nonumber \\
-\frac{\partial \tilde{\bv}}{\partial t} & = & \beta (\nabla \tilde{p} + \tilde{l}_r{\bf e}_z\delta(z-z_r),\nonumber \\
-\tilde{p} & =& 0 \mbox{ for } t>T\nonumber\\ 
-\tilde{\bv} & = & 0 \mbox{ for } t>T 
-\end{eqnarray} 
-Then
-\[
-0 = 
-\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  
-\frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=T}
--
-\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=0}
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \tilde {p} +  p \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} \right.
-\]
-\[
-+
-\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \tilde{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \tilde{\bv}}{\partial t} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
- h_s \delta(z-z_s)\right) \tilde{p} + p \left(- \nabla \cdot \tilde{\bv} + 
- \tilde{h}_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  (- \nabla p + l_s{\bf e}_z\delta(z-z_s)) \cdot \tilde{\bv} + \bv \cdot (-\nabla \tilde{p} + \tilde{l}_r{\bf e}_z\delta(z-z_r)) \right)
-\]
-\[
-= 
-\int_{0}^{T}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv \tilde{p} + 
- h_s\tilde{p}|_{z=z_s} - p\nabla \cdot \tilde{\bv} + 
- \tilde{h}_r p|_{z=z_r}\right) \right.
-\]
-\[
-+
-\left.  p (\nabla \cdot \tilde{\bv}) + l_s \tilde{v}_z|_{z=z_r}+ (\nabla \cdot \bv) \tilde{p}  + l_r  v_z|_{z=z_r}\right)
-\]
-after integration by parts in the last two terms. Most of what is left cancels, leaving 
-\[
-0 = \int\,dt\,dx\,dy\, (h_s \tilde{p})_{z=z_s} + (p  \tilde{h}_r)_{z=z_r} + (l_s\tilde{v}_z)_{z=z_s} + (v_z \tilde{l}_r)_{z=z_r},
-\]
-whence
-\begin{equation}
-\label{eqn:fadj}
- {\bf F}^T {\bf h}_r = -\left[
-\begin{array}{c}
-\tilde{p}(\bx,t;\bx_s)\\
-\tilde{\bf v}(\bx,t;\bx_s)
-\end{array}
-\right]_{z=z_s}
-\end{equation}
-
-%\begin{cor}\label{thm:bvps}
-%Suppose that $d_s$ is a sufficiently regular function, = 0 for $t<0$, and denote by $p^{\pm},\bv^{\pm}$ the half-space solutions provided by Theorem \ref{thm:bvp}.
-Define the vector fields $p,\bv$ in $\bR^4$ by
-%\begin{eqnarray}
-%\label{eqn:putemtogether} 
-%p(x,z) &=& p^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
-%p(x,z_s)&=&d_s(x)\\
-%\bv(x,z) &=& \bv^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
-%\end{eqnarray}
-%Then $p,\bv$ is a weak solution of the system \ref{eqn:awedata} with $h=[v]_n$.
-%\end{cor}
-
-Proof of Theorem \ref{thm:bvp}:
-
-\begin{proof}
-Since $\omega \cap \mbox{ supp }d$ is compact, it is possible to construct an extension of 
-$d \in C^{\infty}_{0}(\bR^{d+1})$. Define $p_1^{\pm}, \bv_1^{\pm}$ to be strong solutions of the system
-\begin{eqnarray}
-\label{eqn:ichom}
-\frac{\partial p^{\pm}_1}{\partial t} & = & - \kappa \nabla \cdot \bv^{\pm}_1 + F\mbox{ in } \Omega^{\pm}, \nonumber \\
-\frac{\partial \bv^{\pm}_1}{\partial t} & = & - \beta \nabla p^{\pm}_1 + {\bf G}\mbox{ in } \Omega^{\pm},\nonumber \\
-p^{\pm}_1 & = & 0 \mbox{ on } \omega \times \bR,\\
-p^{\pm}_1 & = & 0, t \ll 0\\
-\bv^{\pm}_1 & = & 0, t \ll 0 
-\end{eqnarray}
-with 
-\[
-F = -\gamma\frac{\partial d}{\partial t}, {\bf G} = -\nabla d.
-\]
-Existence, uniqueness, and regularity of these solutions are assured by a minor extension of results of \cite{BlazekStolkSymes:13}.
-Set $p^{\pm} = p_1^{\pm} + d, \bv^{\pm} = \bv_1^{\pm}$; then $p^{\pm},\bv^{\pm}$ are the unique strong solutions of the boundary value problems \ref{eqn:awebdry}, and are smooth up to the boundary as indicated.
-\end{proof}
-
-Proof of Corollary \ref{thm:bvps}:
-
-\begin{proof}: Choose smooth functions $\phi \in C^{\infty}_0(\bR^{d+1}), {\bf \psi} \in  (C^{\infty}_0(\bR^{d+1}))^d$. Then
-\[
-\int_{\Omega^{\pm} \times \bR}\left (\gamma \frac{\partial \phi}{\partial t} +
-\nabla \cdot {\bf \psi}\right)p^{\pm} + \left(\rho\frac{\partial {\bf
-    \psi}}{\partial t}
-  + \nabla \phi\right) \cdot \bv^{\pm}
-\]
-\[
-= - \int_{\Omega \times \bR}\left (\gamma \frac{\partial p^{\pm}}{\partial t} +
-\nabla \cdot \bv^{\pm}\right) \phi + \left(\rho\frac{\partial \bv^{\pm}}{\partial t}
-  + \nabla p^{\pm}\right) \cdot {\bf \psi}
-\]
-\begin{equation}
-\label{eqn:rhs1}
-\pm \int_{\omega \times \bR} \bn \cdot {\bf \psi} p^{\pm} + \bn
-  \cdot \bv^{\pm} \phi
-\end{equation}
-Adding equations \ref{eqn:rhs1} for the two choices of sign and taking into
-account that $p^{\pm}$ coincide (by construction) on $\omega \times \bR$, obtain
-\[
-\int_{(\bR^d \times \bR}\left (\gamma \frac{\partial \phi}{\partial t} +
-\nabla \cdot {\bf \psi}\right)p + \left(\rho\frac{\partial {\bf
-    \psi}}{\partial t}
-  + \nabla \phi\right) \cdot \bv
-\]
-\[
-= \int_{\partial \Omega \times [0,T]} (\bn \cdot \bv^{+} -  \bn
-  \cdot \bv^{-}) \phi
-\]
-Since $\phi$ and ${\bf \psi}$ are arbitrary, other than being smooth of compact support, this observation finishes the proof.
-\end{proof}
-
-\append{Variable Projection}
-
-\[
-J[c]=\min_h\frac{1}{2}(\|S[c]h-d\|^2 + \alpha^2\|Ah\|^2)
-\]
-\[
-= \min_h \frac{1}{2}\|F[c]h-e\|^2
-\]
-\[
-F[c]=\left(
-\begin{array}{c}
-S[c]\\
-\alpha A
-\end{array}
-\right),\,\,
-e=\left(
-\begin{array}{c}
-d\\
-0
-\end{array}
-\right)
-\]
-
-\[
-J[c]=\frac{1}{2}\|(F[c]N[c]^{-1}F[c]^T-I)e\|^2,\,\,N[c]=F[c]^TF[c]
-\]
-\[
-DJ[c]\delta c = \langle (F[c]N[c]^{-1}F[c]^T-I)e, 
-(DF[c]\delta c)N[c]^{-1}F[c]^Te +
-F[c](-N[c]^{-1}(DN[c]\delta c)N[c]^{-1})F[c]^Te +
-F[c]N[c]^{-1}(DF[c]\delta c)^Te \rangle 
-\]
-
-\[
-= \langle (F[c]N[c]^{-1}F[c]^T-I)e, 
-(DF[c]\delta c)N[c]^{-1}F[c]^Te -
-F[c]N[c]^{-1}(DF[c]\delta c)^T F[c]N[c]^{-1})F[c]^Te -
-F[c]N[c]^{-1}F[c]^T (DF[c]\delta c]N[c]^{-1})F[c]^Te +
-F[c]N[c]^{-1}(DF[c]\delta c)^Te \rangle 
-\]
-\[
-= \langle (F[c]N[c]^{-1}F[c]^T-I)e, 
-(I-F[c]N[c]^{-1}F[c]^T) DF[c]\delta c)N[c]^{-1}F[c]^Te
-F[c]N[c]^{-1}(DF[c]\delta c)^T(I-F[c]N[c]^{-1}F[c]^T) \rangle
-\]
-\[
-=-\langle (I-F[c]N[c]^{-1}F[c]^T)^2e, DF[c]\delta c)N[c]^{-1}F[c]^Te\rangle
--\langle e,(I-F[c]N[c]^{-1}F[c]^T)F[c]N[c]^{-1}(DF[c]\delta c)^T (I-F[c]N[c]^{-1}F[c]^T)e
-\]
-\[
-F[c]N[c]^{-1}F[c]^TF[c]N[c]^{-1}=F[c]N[c]^{-1}
-\]
-so 
-\[
-(I-F[c]N[c]^{-1}F[c]^T)F[c]N[c]^{-1} = 0
-\]
-and
-\[
-(I-F[c]N[c]^{-1}F[c]^T)^2 = (I-F[c]N[c]^{-1}F[c]^T)
-\]
-Whence
-\[
-DJ[c]\delta c =
--\langle (I-F[c]N[c]^{-1}F[c]^T)e, (DF[c]\delta c)N[c]^{-1}F[c]^Te\rangle
-\]
-Define $h[c]=N[c]^{-1}F[c]^Te$ to be the solution of the inner problem as function of $c$. Then
-\[
-DJ[c]\delta c = \langle F[c]h[c]-e,(DF[c]\delta c)h[c] \rangle
-\]
-which is the usual VPM gradient formula.
-
-Gauss-Newton: define $G[c]=F[c]N[c]^{-1}F[c]^Te$. Need $DG[c], DG[c]^T$.
-\[
-DG[c]\delta c =( DF[c]\delta c)N[c]^{-1}F[c]^Te 
-\]
-\[
-- F[c]N[c]^{-1}(DF[c]\delta c)^TF[c]N[c]^{-1}F[c]^Te + F[c]N[c]^{-1} F[c]^T(DF[c]\delta c))N[c]^{-1}F[c]^Te 
-\]
-\[
-+ F[c]N[c]^{-1}(DF[c]\delta c)^Te
-\]
-\[
-=(I-F[c]N[c]^{-1} F[c]^T)(DF[c]\delta c))N[c]^{-1}F[c]^Te + F[c]N[c]^{-1}(DF[c]\delta c)^T(I-F[c]N[c]^{-1} F[c]^T) e
-\]
-Reinterpretation: set
-\begin{eqnarray}
-h[c] &=&N[c]^{-1}F[c]^Te \nonumber\\
-r[c] &= &F[c]h[c]-e \nonumber\\
-E_1[c]\delta c & = & (DF[c]\delta c)h[c] \nonumber\\
-H_1[c]\delta c &=&N[c]^{-1}F[c]^T E_1[c]\delta c \nonumber\\
-R_1[c]\delta c &=& (F[c]H_1[c]-E_1[c])\delta c \nonumber\\
-H_2[c]\delta c &=&N[c]^{-1}(DF[c]\delta c)^Tr[c] \nonumber\\
-R_2[c]\delta c &=&F[c]H_2[c]\delta c
-\end{eqnarray}
-Then
-\[
-DG[c]\delta c = -(R_1[c]+R_2[c])\delta c
-\]
-\[
-= (I-F[c]N[c]^{-1}F[c]^T)DF[c]\delta c)h[c] -F[c]N[c]^{-1}(DF[c]\delta c)^Tr[c]
-\]
-Adjoints: write $(DF[c]\delta c)h = DF[c](\delta c,h)$, 
-\begin{eqnarray}
-\langle F[c]h,\phi \rangle &=& \langle h,F[c]^T\phi\rangle \nonumber\\
-\langle DF[c](\delta c,h),\phi \rangle &=& \langle \delta c, DF[c]^*(\phi,h) \rangle
-\end{eqnarray}
-\[
-\langle DF[c](\delta c,h),\phi \rangle = \langle \delta c,DF[c]^*(\phi,h)\rangle
-\]
-\[
-\langle E_1[c]\delta c,\phi \rangle = \langle DF[c](\delta c,h[c]),\phi \rangle
-\]
-\[
-= \langle \delta c, DF[c]^*(\phi,h[c]) \rangle
-\]
-so $E_1[c]^* = DF[c]^*(\cdot,h[c])$, and
-\[
-\langle F[c]H_1[c]\delta c, \phi \rangle = \langle F[c] N[c]^{-1}F[c]^TE_1[c]\delta c,\phi \rangle
-\]
-\[
-=\langle E_1[c]\delta c,F[c]N[c]^{-1}F[c]^T\phi\rangle
-\]
-\[
-=\langle \delta c, DF[c]^*(F[c]N[c]^{-1}F[c]^T\phi,h[c])\rangle
-\]
-\[
-\langle R_1[c]\delta c,\phi \rangle = \langle \delta c, DF[c]^*((F[c]N[c]^{-1}F[c]^T-I)\phi,h[c])
-\]
-That is,
-\[
-R_1[c]^*\phi = DF[c]^*((F[c]N[c]^{-1}F[c]^T-I)\phi,h[c])
-\]
-\[
-\langle R_2[c]\delta c,\phi \rangle = \langle F[c]N[c]^{-1}(DF[c]\delta c)^Tr[c],\phi \rangle
-\]
-\[
-\langle r[c],DF[c](\delta c,N[c]^{-1}F[c]^T\phi)\rangle = \langle \delta c, DF[c]^*(r[c], N[c]^{-1}F[c]^T\phi) \rangle
-\]
-That is,
-\[
-R_2[c]^*\phi = DF[c]^*(r[c], N[c]^{-1}F[c]^T\phi)
-\]
-
-\end{document}
-\section{Figures for talk}
-\plot{pulse00}{width=\textwidth}{Bandpass filter pulse}
-\plot{fwitrbulkmodinvlh0}{width=\textwidth}{Inversion for bulk modulus of data displayed in Figure \ref{fig:ptpwindl0}.
-Initial model: constant bulk modulus = 4.0 GPa.
-Algorithm: 20 iterations of trust region Krylov-Gauss-Newton algorithm with preconditioning by box smoother of radius 5 points, repeated 10 times.
-Each KGN step computed by up to 20 Conjugate Gradient iterations.}
-\plot{fwitrdatapreslh0}{width=\textwidth}{Data residual produced by 20 itersions of KGN FWI for bulk modulus, starting at constant bulk modulus. Residual RMS approximately 26\% of data RMS.}
-\plot{cgwhestsourceplh0}{width=\textwidth}{Extended inversion for source on 1500 m $\le x \le$ 5500 m, $z$=3000 m, data from Figure \ref{fig:ptpwindl0}, homogeneous bulk modulus = 4.0 GPA.}
-\plot{cgwhresdataplh0}{width=\textwidth}{Residual data from extended source inversion shown in Figure \ref{fig:cgwhestsourceplh0}. RMS is 3\% of data RMS.}
-\plot{bmh0s}{width=\textwidth}{Iteration 1}
-\plot{cgw1bulkupdlh0}{width=\textwidth}{Iteration 1 SSE bulkmod estimate}
-\plot{cgw2bulkupdlh0}{width=\textwidth}{Iteration 2 SSE bulkmod estimate}
-\plot{cgw3bulkupdlh0}{width=\textwidth}{Iteration 3 SSE bulkmod estimate}
-\plot{cgw4bulkupdlh0}{width=\textwidth}{Iteration 4 SSE bulkmod estimate}
-\plot{cgw5bulkupdlh0}{width=\textwidth}{Iteration 5 SSE bulkmod estimate}
-\plot{cgw6bulkupdlh0}{width=\textwidth}{Iteration 6 SSE bulkmod estimate}
-\plot{cgw7bulkupdlh0}{width=\textwidth}{Iteration 7 SSE bulkmod estimate}
-\plot{cgw8bulkupdlh0}{width=\textwidth}{Iteration 8 SSE bulkmod estimate}
-\plot{cgw9bulkupdlh0}{width=\textwidth}{Iteration 9 SSE bulkmod estimate}
-\plot{cgw10bulkupdlh0}{width=\textwidth}{Iteration 10 SSE bulkmod estimate}
-\plot{cgw11bulkupdlh0}{width=\textwidth}{Iteration 11 SSE bulkmod estimate}
-\plot{cgw12bulkupdlh0}{width=\textwidth}{Iteration 12 SSE bulkmod estimate}
-\plot{cgw13bulkupdlh0}{width=\textwidth}{Iteration 13 SSE bulkmod estimate}
-\plot{cgw14bulkupdlh0}{width=\textwidth}{Iteration 14 SSE bulkmod estimate}
-\plot{cgw15bulkupdlh0}{width=\textwidth}{Iteration 15 SSE bulkmod estimate}
-\plot{cgw22bulkupdlh0}{width=\textwidth}{Iteration 22 SSE bulkmod estimate}
-\plot{cgw30bulkupdlh0}{width=\textwidth}{Iteration 30 SSE bulkmod estimate}
-\plot{fwitrsse1bulkmodinvlh0}{width=\textwidth}{Inversion for bulk modulus of data displayed in Figure \ref{fig:ptpwindl0}.
-Initial model: result of 15 steepest descent steps of SSE, Figure \ref{fig:cgw15bulkupdlh0}.
-Algorithm: 20 iterations of trust region Krylov-Gauss-Newton algorithm with preconditioning by box smoother of radius 5 points, repeated 10 times.
-Each KGN step computed by up to 20 Conjugate Gradient iterations.}
-\plot{fwitrsse1datapreslh0}{width=\textwidth}{Data residual produced by 20 itersions of KGN FWI for bulk modulus, starting at output of 15 steepest descent steps of SSE, Figure \ref{fig:cgw15bulkupdlh0}.
-Residual RMS approximately 1.2\% of data RMS.}
-\plot{cgw0estsourceplh0}{width=\textwidth}{Iteration 1 SSE source estimate alpha=4.36 $\times 10^{-6}$}
-\plot{cgw1estsourceplh0}{width=\textwidth}{Iteration 2 SSE source estimate alpha=5.93 $\times 10^{-6}$}
-\plot{cgw2estsourceplh0}{width=\textwidth}{Iteration 3 SSE source estimate alpha=6.82 $\times 10^{-6}$}
-\plot{cgw3estsourceplh0}{width=\textwidth}{Iteration 4 SSE source estimat alpha=7.93 $\times 10^{-6}$e}
-\plot{cgw4estsourceplh0}{width=\textwidth}{Iteration 5 SSE source estimate alpha=9.29 $\times 10^{-6}$}
-\plot{cgw5estsourceplh0}{width=\textwidth}{Iteration 6 SSE source estimate alpha=1.1 $\times 10^{-5}$}
-\plot{cgw6estsourceplh0}{width=\textwidth}{Iteration 7 SSE source estimate alpha=1.32 $\times 10^{-5}$}
-\plot{cgw7estsourceplh0}{width=\textwidth}{Iteration 8 SSE source estimate alpha=1.32 $\times 10^{-5}$}
-\plot{cgw8estsourceplh0}{width=\textwidth}{Iteration 9 SSE source estimate alpha=1.72 $\times 10^{-5}$}
-\plot{cgw9estsourceplh0}{width=\textwidth}{Iteration 10 SSE source estimate alpha=2.2 $\times 10^{-5}$}
-\plot{cgw10estsourceplh0}{width=\textwidth}{Iteration 11 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw11estsourceplh0}{width=\textwidth}{Iteration 12 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw12estsourceplh0}{width=\textwidth}{Iteration 13 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw13estsourceplh0}{width=\textwidth}{Iteration 14 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw14estsourceplh0}{width=\textwidth}{Iteration 15 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw21estsourceplh0}{width=\textwidth}{Iteration 22 SSE source estimate alpha=5.73 $\times 10^{-5}$}
-\plot{cgw29estsourceplh0}{width=\textwidth}{Iteration 30 SSE source estimate alpha=9.1 $\times 10^{-5}$}
-\plot{cgw0resdataplh0}{width=\textwidth}{Iteration 1 SSE data residual}
-\plot{cgw1resdataplh0}{width=\textwidth}{Iteration 2 SSE data residual}
-\plot{cgw2resdataplh0}{width=\textwidth}{Iteration 3 SSE data residual}
-\plot{cgw3resdataplh0}{width=\textwidth}{Iteration 4 SSE data residual}
-\plot{cgw4resdataplh0}{width=\textwidth}{Iteration 5 SSE data residual}
-\plot{cgw5resdataplh0}{width=\textwidth}{Iteration 6 SSE data residual}
-\plot{cgw6resdataplh0}{width=\textwidth}{Iteration 7 SSE data residual}
-\plot{cgw7resdataplh0}{width=\textwidth}{Iteration 8 SSE data residual}
-\plot{cgw8resdataplh0}{width=\textwidth}{Iteration 9 SSE data residual}
-\plot{cgw9resdataplh0}{width=\textwidth}{Iteration 10 SSE data residual}
-\plot{cgw10resdataplh0}{width=\textwidth}{Iteration 11 SSE data residual}
-\plot{cgw11resdataplh0}{width=\textwidth}{Iteration 12 SSE data residual}
-\plot{cgw12resdataplh0}{width=\textwidth}{Iteration 13 SSE data residual}
-\plot{cgw13resdataplh0}{width=\textwidth}{Iteration 14 SSE data residual}
-\plot{cgw14resdataplh0}{width=\textwidth}{Iteration 15 SSE data residual}

### 2019-10-23 repro/papers/trans/prop.tex
update

--- 2019-10-19 repro/papers/trans/paper.tex
@@
-\title{Waveform Inversion via Source Extension: Major Questions and Research Plan}
-\author{William. W. Symes \thanks{The Rice Inversion Project,
-Department of Computational and Applied Mathematics, Rice University,
-Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}
-
-\extrafloats{100}
-
-\lefthead{Symes}
-
-\righthead{Source Extension}
-
-\maketitle
-\begin{abstract}
-FWI via source extension is a workaround for cycle skipping.
-For pure transmission (including diving wave) data, a close link to traveltime tomography explains the good results observed in experiments.
-No such link is known to exist in the case of pure reflection data, and numerical results are mixed.
-This proposal outlines several numerical experiments that should complement theoretical investigations and decide the effectiveness of these algorithms.
-\end{abstract}
-
-\section{Overview}
-FWI can be described in terms of a wave operator $L[c]$, depending on an array of space-dependent coefficients $c$, a trace sampling operator $P$, a wavefield $u$, and a source function (of space and time) $f$.
-The basic FWI problem is: given $d$, find $c$ so that
-\begin{equation}
-\label{eqn:fwi}
-Pu \approx d \mbox{ and } L[c]u = f.
-\end{equation}
-Sometimes $f$ is regarded as given, sometimes only some aspects are given (for example, localization at a source point) and others are to be determined as part of the solution.
-A typical formulation is via nonlinear least squares:
-\[
-\mbox{choose } c \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2 
-\]
-usually with some sort of regularization thrown in.
-
-Note that $f$ and $d$ are typically vectors of trace gathers, one for each source position in the survey (or possibly some other parameter, produced by processing - will stick with source position $\bx_s$).
-
-As is well-known, local optimization methods are the only feasible approach given the dimensions of a typical instance of \ref{eqn:fwi}, and those have a tendency to stall due to ``cycle-skipping''.
-Source extension is one approach to avoiding this problem.
-It consists in imposing the wave equation as a soft as opposed to hard constraint, by allowing thw source to have more degrees of freedom than is permitted by the basic model.
-For sake of argument (and because almost all work so far has assumed it), adopt the isotropic point source as the model: for the data gather at source positon $\bx_s$, the source takes the form
-\begin{equation}
-\label{eqn:ptsrc}
-f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s).
-\end{equation}
-Every method explained below can be altered to accommodate more complex source models.
-
-Source extension replaces $f(\bx,t;\bx_s)$ with an artificial source $g(\bx,t; \bx_s)$ with more degrees of freedom than merely a single common source wavelet $w(t)$ as in definition \ref{eqn:ptsrc}.
-Assuming that \ref{eqn:ptsrc} is actually correct physics, these extra degrees of freedom have to be suppressed in the optimal solution, via an {\em annihilator} $A$, an operator acting on extended sources $g$, whose null space is precisely the physical sources (of the form \ref{eqn:ptsrc}).
-Extended source inversion replaces the FWI objective with a penalty function
-\begin{equation}
-\label{eqn:esi}
-\mbox{choose } c, g \mbox{ to minimize } \|PL[c]^{-1}g -d \|^2 + \alpha \|A[g]\|^2 
-\end{equation}
-
-The way in which degrees of freedom are introduced, and the choice of annihilator, distinguish the various source extension methods.
-
-\subsection{Wavefield Reconstruction Inversion}
-Introduced by \cite{LeeuwenHerrmannWRI:13} and significantly modified by \cite{WangYingst:SEG16}.
-Extended source space: arbitrary functions of space-time $g(\bx,t;\bx_s)$, one for each source position.
-Annihilator:
-\begin{equation}
-\label{eqn:wri}
-A[g] = g-f
-\end{equation}
-With these choices, WRI takes the form introduced by \cite{WangYingst:SEG16}.
-The original approach is equivalent: \cite{LeeuwenHerrmannWRI:13} use the dyamic wavefield $u$ as the additional degrees of freedom, and minimize
-\[
-\|Pu-d\|^2 + \alpha \|L[c]u-f\|^2
-\]
-Since the wave equation (any reasonable version!) has a unique solution for each right-hand side, replacing $L[c]u$ turns this original form of WRI into the form \ref{eqn:esi}.
-
-Note that in princple at least, the actual source $f$ (positions and wavelet) must be known {\em a priori}.
-
-The Wang-Yingst variant leads to a huge reduction in the number of degrees of freedom: either the original or the modified versions, as originally presented, appear to require storage of an entire space-time source, which is prohibitive for 3D and/or time domain formulation.
-I will explain in an appendix how the Wang-Yingst approach drastically shrinks the storage requirements of this method, and makes it feasible for time-domain 3D application, which Wang and Yingst and their co-authors have demonstrated.
-
-\subsection{Space-Time Extension}
-\cite{HuangSymes:SEG16b}. Differs from WRI by choice of annihilator:
-\begin{equation}
-\label{eqn:msste}
-A[g](\bx,t;\bx_s) = |\bx-\bx_s|g(\bx,t;\bx_s).
-\end{equation}
-This annihilator merely enforces the position constraint on the source, without relying on {\em a priori} knowledge of the source wavelet.
-In fact, the source wavelet is a by-product of the minimization \ref{eqn:esi}.
-
-Appears to require storage of full space-time source volume - the trick explained in the appendix for WRI does not work, due the form of the annihilator.
-
-\subsection{Volume Extension}
-\cite{HuangSymes:SEG16a}.
-Assumes that source signature deconvolution has been performed, so that the actual wavelet is (a bandlimited version of) $\delta(t)$.
-The extended source is an {\em exploding reflector}:
-\begin{equation}
-\label{eqn:msvol}
-g(\bx,t;\bx_s) = h(\bx;\bx_s)\delta(t).
-\end{equation}
-Uses the same annihilator as the Space-Time Extension, but loses the advantage of solving for the source as by-product - the source must be known, so that decon can be carried out prior to inversion.
-
-However, only requires a spatial storage volume per source position, essentially equivalent to a second data volume, therefore practical for 3D (though I don't believe it has been tried except in 2D).
-
-The variable projection method applied to this extension has essentially exploding reflector inversion (as opposed to exploding reflector migration) as its inner problem, i.e.
-estimation of $h$ in \ref{eqn:msvol}.
-The same problem arises in so-called photoacoustic tomography, and has received a lot of attention in the mathematical inverse problems and biomedical imaging literature.
-In fact there is a quite accurate approximate inverse available for the inner problem, consisting of converting pressure traces to normal velocity traces and back-projecting (this is known as ``time reversal'') in the IP literature).
-
-A disadvantage of this approach, if you want to handle reflections, is that unlike post-stack migration the actual physical wave velocity is used in modeling, so that the the region in which $h$ must be defined may considerably larger than the ordinary modeling domain - reflection two-way time is effectively modeled by one-way time.
-
-The volume extension concept is similar to an idea discussed by \cite{ZhangGao:08}.
-
-\subsection{Surface Extension}
-No publication yet - subject of my workshop talk at the 2016 SEG, and GH has described it in talks at Total.
-Similar to the volume extension, but the source is spread over the source space-time surface, rather than over space at $t=0$.
-In the simplest case, where the sources all lie on $z=z_s$, the extended source space consists of functions (distributions) of the form ($\bx=(x,y,z)$)
-\begin{equation}
-\label{eqn:mssur}
-g(\bx,t;\bx_s) = h(x,y,t)\delta(z-z_s).
-\end{equation}
-As for the previous two cases, the annihilator is given by the localization penalty \ref{eqn:msste}.
-Therefore the source wavelet is a by-product of the inversion, and does not need to be specified {\em a priori}.
-Like the volume extension, the additional data volume required is on the same order of size as the data.
-
-\subsection{Adaptive Waveform Inversion = Source-Receiver Extension}
- Popularized by \cite{Warner:14}, \cite{Warner:16}, has many antecedents, listed in \cite[]{HuangSymes:17}. Each trace regarded as separate gather, with separate point source waveform. In effect,
-\begin{equation}
-\label{eqn:awi}
-g(\bx,t;\bx_s) = w(t;\bx_s)\delta(\bx-\bx_s) 
-\end{equation}
-with repeated occurances of $\bx_s$ regarded as distinct. Usually written by making the wavelet a function of the receiver position also, to distinguish redundant source positions:
-\begin{equation}
-\label{eqn:awi-alt}
-g(\bx,t;\bx_r,\bx_s) = w(t;\bx_r,\bx_s)\delta(\bx-\bx_s) 
-\end{equation}
-hence the alternate name.
-
-Various annihilators have been suggested - see references.
-Warner proposes a variant (``AWI'')  that (in principle) requires the source wavelet to be known, and deconvolved from the data traces.
-Then uses multiplication by $t$, which annihilates $\delta(t)$:
-\begin{equation}
-\label{eqn:awi-ann}
-A[g](\bx,t;\bx_r,\bx_s) = t g (\bx,t;\bx_r,\bx_s)
-\end{equation}
-AWI distinguished from earlier variants by normalizing the objective with the estimated wavelet $L^2$ norm:
-\begin{equation}
-\label{eqn:awi-obj}
-\mbox{choose } c, g \mbox{ to minimize } \sum_{\bx_r,\bx_s}\frac{\int dt\,|t|^2 |w(t;\bx_r,\bx_s)|^2}{\int dt\,|w(t;\bx_r,\bx_s)|^2}
-\end{equation}
-Note the order of summation and normalization: each trace is normalized independently (this is not clear from the notation in \cite[]{Warner:16}).
-
-The justification for normalization given by \cite{Warner:16} is that it avoids a familar source/reflector ambiguity.
-Actually it is more important than that, and plays an important role when no reflectors are present.
-It is possible to recast the objective \ref{eqn:awi-obj} as equivalent to \ref{eqn:esi} with a weighted norm used in the latter.
-[I am writing this up - take my word for it!] The weighted norm makes the linear map from source to data {\em unitary} (at least to leading order in frequency), and that is an essential property in relating AWI to traveltime tomography.
-
-\section{What is known}
-For several of the source extension algorithms, the objective is known to be closely linked to traveltime tomography in one form or another, and to inherit its convergence properties.
-These connections are known {\em only} for pure transmission, and not always then.
-Nothing theoretical is known about any of these methdods for reflection data.
-
-The connection to traveltime tomography was first (so far as I know) established for the source-receiver extension in several TRIP technical reports related to Hua Song's PhD thesis \cite[]{Song:94c,SoSy:92,SongSymes:94a,SongSymes:94b,Symes:94c}.
-This work established a form of differential semblance velocity estimation for crosswell tomography, and the related tomographic objective was a form of slope, or stereo-, tomography.
-The time moment objective \ref{eqn:awi-obj} (without normalization) first appears in \cite[]{Plessix:00a,Plessix:00}.
-A similar idea appears in \cite[]{LuoSava:11}.
-Neither of these refferences made an explicit connection to tomography - that was done in \cite[]{HuangSymes2015SEG,HuangSymes:17}.
-
-The defect, in the relation between AWI/source-receiver extention and traveltime tomography, is that it collapses when the ray field (must be able to speak of rays, to invoke traveltime tomography) develops caustics.
-This was already demonstrated both theoretically and by numerical example in \cite[]{Symes:94c}, though the fact does not seem to have percolated very far.
-\cite{HuangSymes:17} give an explicit illustration of this failure in a diving wave setting.
-As soon as the diving wave field is triplicated, AWI is just as bad as FWI.
-
-On the other end of the list, almost nothing is known about why WRI converges, or when it does.
-It may be possible to repeat a version of the analysis mentioned next for the volume and surface extensions, based on geometric optics and the Wang-Yingst variant as described in the appendix, though that has not been tried yet.
-
-The volume and surface extensions have a well-defined relation to tomographic principles (slope or traveltime, depending on type of annihilator).
-This relation holds regardless of ray geometry (with or without triplication) in transmission configuration.
-
-Again, nothing is known theoretically about the behaviour of any of these methods for reflection data, and numerical evidence is ambiguous.
-
 \section{Proposed Project}.
@@
 \plot{trcftrll0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindl0}, \ref{fig:reptpwindll0}.}
-
-\bibliographystyle{seg}
-\bibliography{../../bib/masterref}
-
-\append{RWI (=Wang-Yingst WRI) - a Linear Algebra Viewpoint}
-
-Define $S[c] = PL[c]^{-1}$ = operator mapping extended source to data traces.
-
-The Wang-Yingst version of the WRI objective function may be written
-\[
-\|S[c]g-d\|^2 + \alpha \|g-f\|^2
-\]
-``Most'' extended sources result in zero data traces - those are non-radiating sources, the null space of $S[c]$.
-
-Rank Nullity Theorem: any $g$ may be written in a unique way as $g=S[c]^Th + n$, where $h$ is a collection of data traces, $S[c]^T$ is back propagation, and $n$ is in the null space of $S[c]$.
-Furthermore, the decomposition is orthogonal.
-Thus
-\[
-\|g-f\|^2 = \|S[c]^Th + n -f\|^2 = \|S[c]^Th - f_p\|^2 + \|n-f_n\|^2
-\]
-in which $f_p$ and $f_n$ are the orthoprojections on the orthocomplement of the null space, and the null space, of $S[c]$ respectively.
-
-There is no use keeping $f_n$ around, as it radiates nothing - in fact, I suspect it is possible to see that point sources such as represented in \ref{eqn:ptsrc} are perpindicular to the null space already.
-In any case, simply choose $n=f_n$ and eliminate the second term, without affecting the data fit (by definition!).
-
-Then WRI objective function can be re-written as
-\begin{equation}
-\label{eqn:wriobj}
-\|S[c]S[c]^Th-d\|^2 + \alpha \|S[c]^Th -f_p\|^2
-\end{equation}
-and the extended model has been reduced from a full-space time volume ($g$) to a data gather ($h$).
-
-{\bf Note added 2019.03.14:} Chao Wang's talk at GS 19 revealed that that they actually store the entire source ($g$ above).
-That is not so crazy, since the operator $S[c]S[c]^T$ poses the same complexity issue as time-domain RTM.
-
-To see this, need an explicit description of $S[c]^T$.
-Regard $L[c]$ as the 2nd ord wave operator with velocity $c$: $L[c]=\frac{1}{c^2}\frac{\partial}{\partial t^2} - \nabla^2$.
-$L[c]^{-1}$ is ambiguous: define
-\[
-L[c]^{-1_+}g = u, \mbox{ for } L[c]u=g \mbox{ and } u=0 \mbox{ for } t<0
-\]
-that is, $L[c]^{-1}_+$ is the causal inverse of $L[c]$.
-Solutions defined on time interval $[0,T]$.
-Let $R$ be time-reversal: $Ru(t)=u(T-t)$.
-Then the anticausal inverse $L[c]^{-1}_-$ of $L[c]$, solution $=0$ for $t>T$, can be written as $L[c]^{-1}_- = RL[c]^{-1}_+R$.
-
-Denote by $\langle \cdot,\cdot \rangle$ the $L^2$ inner product in space.. Suppose $u$ is a causal solution of $L[c]u=f$, $w$ an anti-causal solution of $L[c]w=g$. Then 
-\[
-\int_0^T \langle L[c]^{-1}_+f, g \rangle = int_0^T \langle u, g \rangle = \int_0^T \langle u, \left( \frac{1}{c^2}\frac{\partial}{\partial t^2} - \nabla^2\right)w\rangle
-\]
-\[
-= \int_0^T \langle \left( \frac{1}{c^2}\frac{\partial}{\partial t^2} - \nabla^2\right)u, w \rangle = \int_0^T \langle f, w \rangle = \int_0^T \langle f, L[c]^{-1}_-g \rangle
-\]
-since the causal resp.
-anti-causal solutions have zero Cauchy data at $t=0$ resp.
-$t=T$, and since the integration is over all of $\bR^3$ and I am assuming everything vanishing at $\infty$.
-Conclude that
-\[
-(L[c]^{-1}_+)^T = L[c]^{-1}_-
-\]
-so
-\[
-S[c]^T = (PL[c]^{-1}_+)^T = L[c]^{-1}_-P^T
-\]
-
-For convenience, ignore sampling and aperture, and define $P$ to be the restriction operator on $z=0$. Then $(P^T h)(x,y,z,t) = h(x,y,t)\delta(z)$.
-
-The troublesome operator in the definition \ref{eqn:wriobj} is 
-\[
-S[c]S[c]^T = PL[c]^{-1}_+L[c]^{-1}_-P^T
-\]
-The first solution operator involves evolution backwards in time, and provides the data for the second solution operator.
-This leadss to the same computational complexity problem as encountered in RTM.
-It could be handled by storing the output of $L[c]^{-1}_-$, as the IONGeo folks apparently do, or storing boundary data (for acoustics - as usual, this would not work if the approach is extended to a lossy model such as viscoacoustics), or using a checkpointing scheme.
-
-Another approach: write $w = L[c]^{-1}_-P^T h$. The object is to compute $u=L[c]^{-1}_+w$, that is, the solution of 
-\[
-(c^{-2}\partial_t^2 - \nabla^2)u(t) = w(t), 0\le t \le T; \,u(0) = \partial_t u(0) = 0
-\]
-Duhamel's principle would provide a solution in the form of integrals over time of solutions of homogeneous initial value problems:
-\[
-u(t) = \int_0^t d\tau v(t,\tau), \, (c^{-2}\partial_t^2 - \nabla^2)v(t,\tau) = 0, 0\le t \le T; \,v(t,\tau)_{t=\tau} = 0, \partial_t v(t,\tau)_{t=\tau} = w(\tau)
-\]
-Note that $w$ solves the homogenous wave equation except at $z=0$. That makes it tempting to use $w$ directly to define $v$. In fact, set 
-\[
-W(t) = -int_t^T ds w(s),\,v_0(t,\tau) = \frac{1}{2}(W(t) - W(2\tau-t)) 
-\]
-Then $v_0$ satisfies the Duhamel initial conditions. Note that because $w$ has zero Cauchy data at $t=T$,
-\[
-\partial_t W(t)= -\int_t^T ds \partial_t w(s),\,\partial_t^2 W(t)=-\int_t^T ds \partial^2_t w(s)
-\]
-so 
-\[
-(c^{-2}\partial_t^2 - \nabla^2)v_0(t,\tau) = -\int_t^T(c^{-2}\partial^2_t -\nabla^2)w + \int_{2\tau-t}^T(c^{-2}\partial^2_t -\nabla^2)
-\]
-For $t>\tau$, $t > 2\tau-t$, so 
-\[
-= -\int_{2\tau-t}^t(c^{-2}\partial^2_t -\nabla^2)w = -\int_{2\tau-t}^t P^Th
-\]
-and 
-\[ 
-v_0(t,\tau) = \int_{2\tau-t}^t w  
-\] 
-Consequently $v=v_0 + v_1$, in which $v_1$ is defined by 
-\[
-(c^{-2}\partial_t^2 - \nabla^2)v_1(t,\tau) = \int_{2\tau-t}^t P^Th, \,  v_1(t,\tau)_{t=\tau} = 0, \,  \partial_t v_1(t,\tau)_{t=\tau} = 0
-\]
-
-\append{Transpose of Full Acoustic Modeling Operator}
-In this section I compute the adjoint operator form modeling with both pressure and velocity source inputs and pressure and velocity trace outputs. The forward modeling operator is defined by solving
-\begin{eqnarray}
-\label{eqn:fawedata}
-\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
-h_s \delta(z-z_s)), \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta ( \nabla p + l_s {\bf e}_z\delta(z-z_s)),\nonumber \\
-p & =& 0 \mbox{ for } t<0\nonumber\\ 
-\bv & = & 0 \mbox{ for } t<0 \nonumber\\
-\end{eqnarray}
-The new elements here are the point load time function $l_s(\bx,t;\bx_s)$, and the vertical unit vector ${\bf e}_z=(0,1)^T$ for 2D, $=(0,0,1)^T$ for 3D.
-The forward map ${\bf F}$ is defined as implicit in the discussion of the Intrinsically Preconditioned CG algorithm: let ${\bf h}_s=(h_s,l_s)^T$, then
-\begin{equation}
-\label{eqn:ffwd}
-{\bf F}{\bf h}_s =\left\{ \left[
-\begin{array}{c}
-p(\bx,t;\bx_s)\\
-v_z(\bx,t;\bx_s)
-\end{array}
-\right]_{z=z_r}\right\}
-\end{equation}
-
-Let $\tilde{\bf h}_r=(\tilde{h}_r,\tilde{l}_r)^T$ be a function on $z=z_r$, 
-and 
-$\tilde{p},\tilde{\bv}$ solve the backwards-in-time boundary value problem:
-\begin{eqnarray}
-\label{eqn:faweadj}
-\frac{\partial \tilde{p}}{\partial t} & = & \kappa (-\nabla \cdot \tilde{\bv} + 
- \tilde{h}_r \delta(z-z_r))\nonumber \\
-\frac{\partial \tilde{\bv}}{\partial t} & = & \beta (\nabla \tilde{p} + \tilde{l}_r{\bf e}_z\delta(z-z_r),\nonumber \\
-\tilde{p} & =& 0 \mbox{ for } t>T\nonumber\\ 
-\tilde{\bv} & = & 0 \mbox{ for } t>T 
-\end{eqnarray} 
-Then
-\[
-0 = 
-\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  
-\frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=T}
--
-\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=0}
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \tilde {p} +  p \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} \right.
-\]
-\[
-+
-\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \tilde{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \tilde{\bv}}{\partial t} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
- h_s \delta(z-z_s)\right) \tilde{p} + p \left(- \nabla \cdot \tilde{\bv} + 
- \tilde{h}_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  (- \nabla p + l_s{\bf e}_z\delta(z-z_s)) \cdot \tilde{\bv} + \bv \cdot (-\nabla \tilde{p} + \tilde{l}_r{\bf e}_z\delta(z-z_r)) \right)
-\]
-\[
-= 
-\int_{0}^{T}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv \tilde{p} + 
- h_s\tilde{p}|_{z=z_s} - p\nabla \cdot \tilde{\bv} + 
- \tilde{h}_r p|_{z=z_r}\right) \right.
-\]
-\[
-+
-\left.  p (\nabla \cdot \tilde{\bv}) + l_s \tilde{v}_z|_{z=z_r}+ (\nabla \cdot \bv) \tilde{p}  + l_r  v_z|_{z=z_r}\right)
-\]
-after integration by parts in the last two terms. Most of what is left cancels, leaving 
-\[
-0 = \int\,dt\,dx\,dy\, (h_s \tilde{p})_{z=z_s} + (p  \tilde{h}_r)_{z=z_r} + (l_s\tilde{v}_z)_{z=z_s} + (v_z \tilde{l}_r)_{z=z_r},
-\]
-whence
-\begin{equation}
-\label{eqn:fadj}
- {\bf F}^T {\bf h}_r = -\left[
-\begin{array}{c}
-\tilde{p}(\bx,t;\bx_s)\\
-\tilde{\bf v}(\bx,t;\bx_s)
-\end{array}
-\right]_{z=z_s}
-\end{equation}
-
-%\begin{cor}\label{thm:bvps}
-%Suppose that $d_s$ is a sufficiently regular function, = 0 for $t<0$, and denote by $p^{\pm},\bv^{\pm}$ the half-space solutions provided by Theorem \ref{thm:bvp}.
-Define the vector fields $p,\bv$ in $\bR^4$ by
-%\begin{eqnarray}
-%\label{eqn:putemtogether} 
-%p(x,z) &=& p^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
-%p(x,z_s)&=&d_s(x)\\
-%\bv(x,z) &=& \bv^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
-%\end{eqnarray}
-%Then $p,\bv$ is a weak solution of the system \ref{eqn:awedata} with $h=[v]_n$.
-%\end{cor}
-
-Proof of Theorem \ref{thm:bvp}:
-
-\begin{proof}
-Since $\omega \cap \mbox{ supp }d$ is compact, it is possible to construct an extension of 
-$d \in C^{\infty}_{0}(\bR^{d+1})$. Define $p_1^{\pm}, \bv_1^{\pm}$ to be strong solutions of the system
-\begin{eqnarray}
-\label{eqn:ichom}
-\frac{\partial p^{\pm}_1}{\partial t} & = & - \kappa \nabla \cdot \bv^{\pm}_1 + F\mbox{ in } \Omega^{\pm}, \nonumber \\
-\frac{\partial \bv^{\pm}_1}{\partial t} & = & - \beta \nabla p^{\pm}_1 + {\bf G}\mbox{ in } \Omega^{\pm},\nonumber \\
-p^{\pm}_1 & = & 0 \mbox{ on } \omega \times \bR,\\
-p^{\pm}_1 & = & 0, t \ll 0\\
-\bv^{\pm}_1 & = & 0, t \ll 0 
-\end{eqnarray}
-with 
-\[
-F = -\gamma\frac{\partial d}{\partial t}, {\bf G} = -\nabla d.
-\]
-Existence, uniqueness, and regularity of these solutions are assured by a minor extension of results of \cite{BlazekStolkSymes:13}.
-Set $p^{\pm} = p_1^{\pm} + d, \bv^{\pm} = \bv_1^{\pm}$; then $p^{\pm},\bv^{\pm}$ are the unique strong solutions of the boundary value problems \ref{eqn:awebdry}, and are smooth up to the boundary as indicated.
-\end{proof}
-
-Proof of Corollary \ref{thm:bvps}:
-
-\begin{proof}: Choose smooth functions $\phi \in C^{\infty}_0(\bR^{d+1}), {\bf \psi} \in  (C^{\infty}_0(\bR^{d+1}))^d$. Then
-\[
-\int_{\Omega^{\pm} \times \bR}\left (\gamma \frac{\partial \phi}{\partial t} +
-\nabla \cdot {\bf \psi}\right)p^{\pm} + \left(\rho\frac{\partial {\bf
-    \psi}}{\partial t}
-  + \nabla \phi\right) \cdot \bv^{\pm}
-\]
-\[
-= - \int_{\Omega \times \bR}\left (\gamma \frac{\partial p^{\pm}}{\partial t} +
-\nabla \cdot \bv^{\pm}\right) \phi + \left(\rho\frac{\partial \bv^{\pm}}{\partial t}
-  + \nabla p^{\pm}\right) \cdot {\bf \psi}
-\]
-\begin{equation}
-\label{eqn:rhs1}
-\pm \int_{\omega \times \bR} \bn \cdot {\bf \psi} p^{\pm} + \bn
-  \cdot \bv^{\pm} \phi
-\end{equation}
-Adding equations \ref{eqn:rhs1} for the two choices of sign and taking into
-account that $p^{\pm}$ coincide (by construction) on $\omega \times \bR$, obtain
-\[
-\int_{(\bR^d \times \bR}\left (\gamma \frac{\partial \phi}{\partial t} +
-\nabla \cdot {\bf \psi}\right)p + \left(\rho\frac{\partial {\bf
-    \psi}}{\partial t}
-  + \nabla \phi\right) \cdot \bv
-\]
-\[
-= \int_{\partial \Omega \times [0,T]} (\bn \cdot \bv^{+} -  \bn
-  \cdot \bv^{-}) \phi
-\]
-Since $\phi$ and ${\bf \psi}$ are arbitrary, other than being smooth of compact support, this observation finishes the proof.
-\end{proof}
-
-\append{Variable Projection}
-
-\[
-J[c]=\min_h\frac{1}{2}(\|S[c]h-d\|^2 + \alpha^2\|Ah\|^2)
-\]
-\[
-= \min_h \frac{1}{2}\|F[c]h-e\|^2
-\]
-\[
-F[c]=\left(
-\begin{array}{c}
-S[c]\\
-\alpha A
-\end{array}
-\right),\,\,
-e=\left(
-\begin{array}{c}
-d\\
-0
-\end{array}
-\right)
-\]
-
-\[
-J[c]=\frac{1}{2}\|(F[c]N[c]^{-1}F[c]^T-I)e\|^2,\,\,N[c]=F[c]^TF[c]
-\]
-\[
-DJ[c]\delta c = \langle (F[c]N[c]^{-1}F[c]^T-I)e, 
-(DF[c]\delta c)N[c]^{-1}F[c]^Te +
-F[c](-N[c]^{-1}(DN[c]\delta c)N[c]^{-1})F[c]^Te +
-F[c]N[c]^{-1}(DF[c]\delta c)^Te \rangle 
-\]
-
-\[
-= \langle (F[c]N[c]^{-1}F[c]^T-I)e, 
-(DF[c]\delta c)N[c]^{-1}F[c]^Te -
-F[c]N[c]^{-1}(DF[c]\delta c)^T F[c]N[c]^{-1})F[c]^Te -
-F[c]N[c]^{-1}F[c]^T (DF[c]\delta c]N[c]^{-1})F[c]^Te +
-F[c]N[c]^{-1}(DF[c]\delta c)^Te \rangle 
-\]
-\[
-= \langle (F[c]N[c]^{-1}F[c]^T-I)e, 
-(I-F[c]N[c]^{-1}F[c]^T) DF[c]\delta c)N[c]^{-1}F[c]^Te
-F[c]N[c]^{-1}(DF[c]\delta c)^T(I-F[c]N[c]^{-1}F[c]^T) \rangle
-\]
-\[
-=-\langle (I-F[c]N[c]^{-1}F[c]^T)^2e, DF[c]\delta c)N[c]^{-1}F[c]^Te\rangle
--\langle e,(I-F[c]N[c]^{-1}F[c]^T)F[c]N[c]^{-1}(DF[c]\delta c)^T (I-F[c]N[c]^{-1}F[c]^T)e
-\]
-\[
-F[c]N[c]^{-1}F[c]^TF[c]N[c]^{-1}=F[c]N[c]^{-1}
-\]
-so 
-\[
-(I-F[c]N[c]^{-1}F[c]^T)F[c]N[c]^{-1} = 0
-\]
-and
-\[
-(I-F[c]N[c]^{-1}F[c]^T)^2 = (I-F[c]N[c]^{-1}F[c]^T)
-\]
-Whence
-\[
-DJ[c]\delta c =
--\langle (I-F[c]N[c]^{-1}F[c]^T)e, (DF[c]\delta c)N[c]^{-1}F[c]^Te\rangle
-\]
-Define $h[c]=N[c]^{-1}F[c]^Te$ to be the solution of the inner problem as function of $c$. Then
-\[
-DJ[c]\delta c = \langle F[c]h[c]-e,(DF[c]\delta c)h[c] \rangle
-\]
-which is the usual VPM gradient formula.
-
-Gauss-Newton: define $G[c]=F[c]N[c]^{-1}F[c]^Te$. Need $DG[c], DG[c]^T$.
-\[
-DG[c]\delta c =( DF[c]\delta c)N[c]^{-1}F[c]^Te 
-\]
-\[
-- F[c]N[c]^{-1}(DF[c]\delta c)^TF[c]N[c]^{-1}F[c]^Te + F[c]N[c]^{-1} F[c]^T(DF[c]\delta c))N[c]^{-1}F[c]^Te 
-\]
-\[
-+ F[c]N[c]^{-1}(DF[c]\delta c)^Te
-\]
-\[
-=(I-F[c]N[c]^{-1} F[c]^T)(DF[c]\delta c))N[c]^{-1}F[c]^Te + F[c]N[c]^{-1}(DF[c]\delta c)^T(I-F[c]N[c]^{-1} F[c]^T) e
-\]
-Reinterpretation: set
-\begin{eqnarray}
-h[c] &=&N[c]^{-1}F[c]^Te \nonumber\\
-r[c] &= &F[c]h[c]-e \nonumber\\
-E_1[c]\delta c & = & (DF[c]\delta c)h[c] \nonumber\\
-H_1[c]\delta c &=&N[c]^{-1}F[c]^T E_1[c]\delta c \nonumber\\
-R_1[c]\delta c &=& (F[c]H_1[c]-E_1[c])\delta c \nonumber\\
-H_2[c]\delta c &=&N[c]^{-1}(DF[c]\delta c)^Tr[c] \nonumber\\
-R_2[c]\delta c &=&F[c]H_2[c]\delta c
-\end{eqnarray}
-Then
-\[
-DG[c]\delta c = -(R_1[c]+R_2[c])\delta c
-\]
-\[
-= (I-F[c]N[c]^{-1}F[c]^T)DF[c]\delta c)h[c] -F[c]N[c]^{-1}(DF[c]\delta c)^Tr[c]
-\]
-Adjoints: write $(DF[c]\delta c)h = DF[c](\delta c,h)$, 
-\begin{eqnarray}
-\langle F[c]h,\phi \rangle &=& \langle h,F[c]^T\phi\rangle \nonumber\\
-\langle DF[c](\delta c,h),\phi \rangle &=& \langle \delta c, DF[c]^*(\phi,h) \rangle
-\end{eqnarray}
-\[
-\langle DF[c](\delta c,h),\phi \rangle = \langle \delta c,DF[c]^*(\phi,h)\rangle
-\]
-\[
-\langle E_1[c]\delta c,\phi \rangle = \langle DF[c](\delta c,h[c]),\phi \rangle
-\]
-\[
-= \langle \delta c, DF[c]^*(\phi,h[c]) \rangle
-\]
-so $E_1[c]^* = DF[c]^*(\cdot,h[c])$, and
-\[
-\langle F[c]H_1[c]\delta c, \phi \rangle = \langle F[c] N[c]^{-1}F[c]^TE_1[c]\delta c,\phi \rangle
-\]
-\[
-=\langle E_1[c]\delta c,F[c]N[c]^{-1}F[c]^T\phi\rangle
-\]
-\[
-=\langle \delta c, DF[c]^*(F[c]N[c]^{-1}F[c]^T\phi,h[c])\rangle
-\]
-\[
-\langle R_1[c]\delta c,\phi \rangle = \langle \delta c, DF[c]^*((F[c]N[c]^{-1}F[c]^T-I)\phi,h[c])
-\]
-That is,
-\[
-R_1[c]^*\phi = DF[c]^*((F[c]N[c]^{-1}F[c]^T-I)\phi,h[c])
-\]
-\[
-\langle R_2[c]\delta c,\phi \rangle = \langle F[c]N[c]^{-1}(DF[c]\delta c)^Tr[c],\phi \rangle
-\]
-\[
-\langle r[c],DF[c](\delta c,N[c]^{-1}F[c]^T\phi)\rangle = \langle \delta c, DF[c]^*(r[c], N[c]^{-1}F[c]^T\phi) \rangle
-\]
-That is,
-\[
-R_2[c]^*\phi = DF[c]^*(r[c], N[c]^{-1}F[c]^T\phi)
-\]
-
-\end{document}
-\section{Figures for talk}
-\plot{pulse00}{width=\textwidth}{Bandpass filter pulse}
-\plot{fwitrbulkmodinvlh0}{width=\textwidth}{Inversion for bulk modulus of data displayed in Figure \ref{fig:ptpwindl0}.
-Initial model: constant bulk modulus = 4.0 GPa.
-Algorithm: 20 iterations of trust region Krylov-Gauss-Newton algorithm with preconditioning by box smoother of radius 5 points, repeated 10 times.
-Each KGN step computed by up to 20 Conjugate Gradient iterations.}
-\plot{fwitrdatapreslh0}{width=\textwidth}{Data residual produced by 20 itersions of KGN FWI for bulk modulus, starting at constant bulk modulus. Residual RMS approximately 26\% of data RMS.}
-\plot{cgwhestsourceplh0}{width=\textwidth}{Extended inversion for source on 1500 m $\le x \le$ 5500 m, $z$=3000 m, data from Figure \ref{fig:ptpwindl0}, homogeneous bulk modulus = 4.0 GPA.}
-\plot{cgwhresdataplh0}{width=\textwidth}{Residual data from extended source inversion shown in Figure \ref{fig:cgwhestsourceplh0}. RMS is 3\% of data RMS.}
-\plot{bmh0s}{width=\textwidth}{Iteration 1}
-\plot{cgw1bulkupdlh0}{width=\textwidth}{Iteration 1 SSE bulkmod estimate}
-\plot{cgw2bulkupdlh0}{width=\textwidth}{Iteration 2 SSE bulkmod estimate}
-\plot{cgw3bulkupdlh0}{width=\textwidth}{Iteration 3 SSE bulkmod estimate}
-\plot{cgw4bulkupdlh0}{width=\textwidth}{Iteration 4 SSE bulkmod estimate}
-\plot{cgw5bulkupdlh0}{width=\textwidth}{Iteration 5 SSE bulkmod estimate}
-\plot{cgw6bulkupdlh0}{width=\textwidth}{Iteration 6 SSE bulkmod estimate}
-\plot{cgw7bulkupdlh0}{width=\textwidth}{Iteration 7 SSE bulkmod estimate}
-\plot{cgw8bulkupdlh0}{width=\textwidth}{Iteration 8 SSE bulkmod estimate}
-\plot{cgw9bulkupdlh0}{width=\textwidth}{Iteration 9 SSE bulkmod estimate}
-\plot{cgw10bulkupdlh0}{width=\textwidth}{Iteration 10 SSE bulkmod estimate}
-\plot{cgw11bulkupdlh0}{width=\textwidth}{Iteration 11 SSE bulkmod estimate}
-\plot{cgw12bulkupdlh0}{width=\textwidth}{Iteration 12 SSE bulkmod estimate}
-\plot{cgw13bulkupdlh0}{width=\textwidth}{Iteration 13 SSE bulkmod estimate}
-\plot{cgw14bulkupdlh0}{width=\textwidth}{Iteration 14 SSE bulkmod estimate}
-\plot{cgw15bulkupdlh0}{width=\textwidth}{Iteration 15 SSE bulkmod estimate}
-\plot{cgw22bulkupdlh0}{width=\textwidth}{Iteration 22 SSE bulkmod estimate}
-\plot{cgw30bulkupdlh0}{width=\textwidth}{Iteration 30 SSE bulkmod estimate}
-\plot{fwitrsse1bulkmodinvlh0}{width=\textwidth}{Inversion for bulk modulus of data displayed in Figure \ref{fig:ptpwindl0}.
-Initial model: result of 15 steepest descent steps of SSE, Figure \ref{fig:cgw15bulkupdlh0}.
-Algorithm: 20 iterations of trust region Krylov-Gauss-Newton algorithm with preconditioning by box smoother of radius 5 points, repeated 10 times.
-Each KGN step computed by up to 20 Conjugate Gradient iterations.}
-\plot{fwitrsse1datapreslh0}{width=\textwidth}{Data residual produced by 20 itersions of KGN FWI for bulk modulus, starting at output of 15 steepest descent steps of SSE, Figure \ref{fig:cgw15bulkupdlh0}.
-Residual RMS approximately 1.2\% of data RMS.}
-\plot{cgw0estsourceplh0}{width=\textwidth}{Iteration 1 SSE source estimate alpha=4.36 $\times 10^{-6}$}
-\plot{cgw1estsourceplh0}{width=\textwidth}{Iteration 2 SSE source estimate alpha=5.93 $\times 10^{-6}$}
-\plot{cgw2estsourceplh0}{width=\textwidth}{Iteration 3 SSE source estimate alpha=6.82 $\times 10^{-6}$}
-\plot{cgw3estsourceplh0}{width=\textwidth}{Iteration 4 SSE source estimat alpha=7.93 $\times 10^{-6}$e}
-\plot{cgw4estsourceplh0}{width=\textwidth}{Iteration 5 SSE source estimate alpha=9.29 $\times 10^{-6}$}
-\plot{cgw5estsourceplh0}{width=\textwidth}{Iteration 6 SSE source estimate alpha=1.1 $\times 10^{-5}$}
-\plot{cgw6estsourceplh0}{width=\textwidth}{Iteration 7 SSE source estimate alpha=1.32 $\times 10^{-5}$}
-\plot{cgw7estsourceplh0}{width=\textwidth}{Iteration 8 SSE source estimate alpha=1.32 $\times 10^{-5}$}
-\plot{cgw8estsourceplh0}{width=\textwidth}{Iteration 9 SSE source estimate alpha=1.72 $\times 10^{-5}$}
-\plot{cgw9estsourceplh0}{width=\textwidth}{Iteration 10 SSE source estimate alpha=2.2 $\times 10^{-5}$}
-\plot{cgw10estsourceplh0}{width=\textwidth}{Iteration 11 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw11estsourceplh0}{width=\textwidth}{Iteration 12 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw12estsourceplh0}{width=\textwidth}{Iteration 13 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw13estsourceplh0}{width=\textwidth}{Iteration 14 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw14estsourceplh0}{width=\textwidth}{Iteration 15 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw21estsourceplh0}{width=\textwidth}{Iteration 22 SSE source estimate alpha=5.73 $\times 10^{-5}$}
-\plot{cgw29estsourceplh0}{width=\textwidth}{Iteration 30 SSE source estimate alpha=9.1 $\times 10^{-5}$}
-\plot{cgw0resdataplh0}{width=\textwidth}{Iteration 1 SSE data residual}
-\plot{cgw1resdataplh0}{width=\textwidth}{Iteration 2 SSE data residual}
-\plot{cgw2resdataplh0}{width=\textwidth}{Iteration 3 SSE data residual}
-\plot{cgw3resdataplh0}{width=\textwidth}{Iteration 4 SSE data residual}
-\plot{cgw4resdataplh0}{width=\textwidth}{Iteration 5 SSE data residual}
-\plot{cgw5resdataplh0}{width=\textwidth}{Iteration 6 SSE data residual}
-\plot{cgw6resdataplh0}{width=\textwidth}{Iteration 7 SSE data residual}
-\plot{cgw7resdataplh0}{width=\textwidth}{Iteration 8 SSE data residual}
-\plot{cgw8resdataplh0}{width=\textwidth}{Iteration 9 SSE data residual}
-\plot{cgw9resdataplh0}{width=\textwidth}{Iteration 10 SSE data residual}
-\plot{cgw10resdataplh0}{width=\textwidth}{Iteration 11 SSE data residual}
-\plot{cgw11resdataplh0}{width=\textwidth}{Iteration 12 SSE data residual}
-\plot{cgw12resdataplh0}{width=\textwidth}{Iteration 13 SSE data residual}
-\plot{cgw13resdataplh0}{width=\textwidth}{Iteration 14 SSE data residual}
-\plot{cgw14resdataplh0}{width=\textwidth}{Iteration 15 SSE data residual}

### 2019-10-23 repro/papers/trans/paper.tex
update

@@
-\title{Waveform Inversion via Source Extension: Major Questions and Research Plan}
+\title{Waveform Inversion via Source Extension: Overview as of Q1 2019}
 \author{William. W. Symes \thanks{The Rice Inversion Project,
@@
 
-\section{Proposed Project}.
-
-Amongst the source extension options described in the previous section, only the surface extension has all of these properties:
-\begin{itemize}
-\item requires only additional storage roughly equivalent to the data volume
-\item does not require prior knowledge of the source wavelet, in fact recovers an estimate as a by-product
-\item inner problem (in variable projection - estimation of the extended source) has a cheap approximate solution, via back-propagation
-\item relation with traveltime inversion persists, regardless of ray geometry
-\end{itemize}
-
-Therefoire I propose that we begin with a time-domain implementation of the surface source extension.
-
-\section{Detailed Description}
-The simplest version of the surface source extension assumes acoustic wave physics, and that the physical source locations lie on the horizontal surface $z=z_s$.
-The coefficient vector $c$ is the pair $(\kappa,\beta)$ of bulk modulus and buoyancy, and the wave operator $L[c]$ is given by the linear acoustics system:
-\begin{eqnarray}
-\label{eqn:awe}
-\frac{\partial p}{\partial t} & = & - \kappa (\nabla \cdot \bv +
-w(t)\delta(\bx-\bx_s)), \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
-p & =& 0 \mbox{ for } t<0\nonumber\\ 
-\bv & = & 0 \mbox{ for } t<0 
-\end{eqnarray}
-The physical source is modeled as an isotropic point constitutive law defect. As mentioned earlier, it is simple to accommodate more complex source models.
-
-The dynamical field is the 4-tuple $u=(p,\bv)$.
-Either the pressure $p$ or the particle velocity $\bv$ could be sampled (or both) - presume that the pressure is sampled, that is $Pu = \{p(\bx_r,t;\bx_s)\}$, where $\{\bx_r\}$ are the receiver positions.
-
-Extended sources are spread over $z=z_s$ (\ref{eqn:mssur}). The annihilator is the source localization penalty, equation \ref{eqn:msste}.
-
-Denote by $J_{\alpha}[g,c;d]$ the objective function defined in \ref{eqn:esi}:
-\begin{equation}
-\label{eqn:jdef}
-J_{\alpha}[g,c;d] = \frac{1}{2}( \|PL[c]^{-1}g -d \|^2 + \alpha \|A[g]\|^2) 
-\end{equation}
-To compute the objective 
-\begin{itemize}
-\item solve the acoustic wave system \ref{eqn:awe} with $g$ as in \ref{eqn:mssur}
-\item sample at receiver locations and compute mean-square difference with pressure data traces, take mean square
-\item add scaled mean square of annihilator applied to the source
-\end{itemize}
-
-The variable projection method \cite[]{GolubPereyra:03} applied to the objective \ref{eqn:esi} approaches its minimization via a nested algorithm: first minimize \ref{eqn:esi} over the source $g$ for given coefficients $c$, thus making $g = g[c]$ a function of $c$ (inner minimization); then substitute $g[c]$ in $J$ and minimize the resulting function of $c$ (outer minimization).
-The optimal $g[c]$ solves the first order necessary condition for minimization of $J$ over $g$: $\nabla_g J_{\alpha}[g,c;d] = 0$.
-For linear annihilators such as that defined by \ref{eqn:mssur}, $J$ is quadratic in $g$, so the first order necessary condition is also sufficient, and is equivalent to the linear {\em normal equation}:
-\begin{equation}
-\label{eqn:normal}
-(S[c]^TS[c] + \alpha A^TA) g = S[c]^T d
-\end{equation}
-where I have written $S[c]=PL[c]^{-1}$ for the operator that maps the (extended) source to the data traces. Substitution of $g[c]$ in $J$ results in the {\em reduced objective}
-\begin{equation}
-\label{eqn:jred}
-\tilde{J}_{\alpha}[c;d] = J_{\alpha}[g[c],c;d].
-\end{equation}
-The gradient would appear to involve the derivative of the inner solution $g[c]$ with respect to $c$, computation of which implies solution of additional linear systems.
-However by fortunate accident, the chain rule and the normal equation \ref{eqn:normal}
-conspire to eliminate all terms involving derivatives of $g[c]$. Consequently the direction derivative of $\tilde{J}_{\alpha}$ in the direction $\delta c$ is
-\begin{equation}
-\label{eqn:tildejderiv}
-D_c\tilde{J}_{\alpha}[c;d]\delta c = \langle D(S[c]g)\delta c,S[c]g-d \rangle|_{g=g[c]}
-\end{equation}
-whence
-\begin{equation}
-\label{eqn:l2grad}
-\nabla \tilde{J}_{\alpha}[c;d] = D_c(S[c]g)^T|_{g=g[c]}(S[c]g[c]-d)
-\end{equation}
-Let us field-strip the expression on the right.
-The rightmost factor is the data residual, which is a by-product of many methods for the solution of the normal equation \ref{eqn:normal}.
-$D_cS$ is the linearization or Born approximation of $S$, or perturbation with respect to $c$, with $g$ fixed:
-\[
-S[c+\delta c]g \approx S[c]g + D_c(S[c]g)\delta c
-\]
-That is, $D_c(S[c]g)$ is the object commonly called the Born modeling operator, with source wavefield $g$.
-Its adjoint or transpose is $D_c(S[c]g)^T$ - as is well known, this is the RTM operator (or rather a version of it), computed via the adjoint state method in the time domain.
-Note that the Born and RTM operators need to match, i.e.
-both be derived from the same underlying full waveform modeling operator (in this case acoustodynamics, but the same would apply for any other wave theory).
-
-The upshot is that the computation of the gradient goes like this:
-\begin{itemize}
-\item solve the normal equation \ref{eqn:normal} for $g[c]$
-\item compute (or retain) the data residual $S[c]g[c]-d$
-\item reverse-time migrate the data residual, using $g[c]$ to generate the source wavefield in the adjoint state method - the output is the gradient of $\tilde{J}_{\alpha}$.
-\end{itemize}
-
-To make this algorithm practical and efficient, effective preconditioning needs to be applied to both inner and outer problems.
-
-\subsection{Preconditioning the Outer Problem: Gradient Smoothing}
-Conditioning the outer (velocity etc.) update has different meaning for transmission versus reflection problems.
-One interpretation of ``transmission'' is that no reflection occurs on any ray path connecting source and receiver, whereever these may be located.
-Is is practically implicit in the use of ``ray path'' in this statement that the material model is at least mostly smooth, or essentially localized in low spatial frequencies.
-Smoothness is also one way to interpret ``lack of reflections''.
-The classic work explaining the need for model smoothness in tomography is \cite{DelpratLailly:92}.
-Since the surface source extension method gets its ability to identify kinematically correct velocity from its relation to tomography, it also requires model smoothness, at least for transmission.
-
-A natural method for enforcing smoothness is use of a norm in model space that penalizes lack of smoothness.
-This method is also properly called preconditioning, which consists in altering the choice of norm to enhance convergence.
-For pure transmission, the choice recommended by \cite[]{DelpratLailly:92} is a norm from the Sobolev family.
-These are weighted $L^2$ (RMS) norms with weight operators indexed by a real parameter $s$:
-\begin{equation}
-\label{eqn:sob}
-\|u\|^2_s = \langle u, W_s u \rangle_{L^2},\,\,W_s = (I - \nabla^2)^s
-\end{equation}
-Choosing $s>0$ penalizes lack of smoothness.
-According to the Sobolev Embedding Theorem and the basic well-posedness theorem for ordinary differential equations (which governs the behaviour of the ray system), it is neccessary to choose $s > 2 + d/2$ in dimension $d$, in order that metrizing the velocity (amongst other things) with the $s$-norm will guarantee sufficient smoothness that ray theory is well-behaved, and the link with tomography honored.
-For both $d=2$ and $d=3$, $s=4$ would work.
-
-The gradient of a function is defined with respect to the domain norm. Using the $s$-norm, the gradient is
-\begin{equation}
-\label{eqn:grads}
-\nabla_s \tilde{J}_{\alpha}[c,d] = W_s^{-1}\nabla \tilde{J}_{\alpha}[c,d]
-\end{equation}
-If $s>0$ then $W_s^{-1}$ is a smoothing operator. It is best implemented in the Fourier domain.
-
-The most natural way view the reflection problem is to allow non-smoothness in $c$.
-It is not clear at this point what would be the optimal choice of weight operator in this case, or even whether there is a choice of Hilbert space structure that would offer an easily computable gradient and development of short-scale structures to generate reflections.
-GH used very low frequency data (in frequency domain) and Tihonov regularization.
-His results seemed to depend heavily on the regularization weight.
-Of course we also do not know how to connect the reflection case to a tomographic principle, so have no guidance for how a model space norm should be selected.
-These are the main target questions of this project.
-
-\subsection{Preconditioning the Inner Problem: Time Reversal}
-%In contrast to the outer problem, a preconditioner for the inner problem is available that works regardless of medium smoothness, so for both transmission and reflection (the latter with some caveats).
-In contrast to the outer problem, the inner problem admits a conventional preconditioner, that is, a change of variables that in the spaces of source traces and data traces respectively that moves the singular values of the modeling operator closer to 1.
-This preconditioner also creates an approximate solution operator for the inner problem, of the form
-\begin{equation}
-\label{eqn:appinv}
-S[c]^{\dagger} = W_m[c]^{-1}S[c]^TW_d[c],
-\end{equation}
-in which $W_m$ and $W_c$ are $c$-dependent positive definite symmetric operators that define weighted inner products in domain and range of $S[c]$, and their associated weighted norms:
-\begin{eqnarray}
-\label{eqn:norms}
-\langle f,g \rangle_m & = & \langle f, W_mg\rangle; \|f\|_m^2 = \langle f,f\rangle_m \nonumber \\
-\langle f,g \rangle_d & = & \langle f, W_d g\rangle; \|f\|_d^2 = \langle f,f \rangle_d
-\end{eqnarray}
-We have implicitly used the notations $\langle \cdot,\cdot \rangle, \|\cdot \|$ (without subscripts) to denote the (unweighted) $L^2$ norm, in whatever form is appropriate for the arguments.
-That is what these symbols mean in display \ref{eqn:norms}, and will continue to mean.
-
-The approximate inverse $S[c]^{\dagger}$ defined by the relation \ref{eqn:appinv} is the transpose of $S[c]$ with respect to the inner products $\langle \cdot,\cdot \rangle_m,, \langle \cdot,\cdot \rangle_d$, so $S[c]$ is approximately unitary in the same sense.
-Consequently the singular spectrum (which is also dependent on choice of inner product) is close to that of the identity - or most of it is, as we shall explain - and the convergence of  a Krylov subspace method like Conjugate Gradient (CG) Iteration converge relatively rapidly.
-
-The system that needs solving, of course, is the normal equation \ref{eqn:normal},  so it is the convergence of CG or the like applied to that system that is the important issue.
-The normal equation is the first order necessary condition for minimization of the objective
-\begin{equation}
-\label{eqn:jdefs}
-J_{\alpha}[g,c;d] = \frac{1}{2}( \|S[c]g -d \|^2 + \alpha \|A[g]\|^2).
-\end{equation}
-over the choice of surface source $g$.
-The choice of norms appearing in \ref{eqn:jdefs} is actually arbitrary, and the norms appearing in the first and second terms on the right-hand side do not even have to be the same.
-
-%First the transmission case - a sort of rotated crosswell configuration. Idealize the receiver set as a continuum plane $z=z_r$, with $z_r > z_s$ - not essential but simpler.
-
-%Two key facts and a geometric assumption underly the construction of the time-reversal preconditioner: (i) a local relation between the surface source and the normal particle velocity, and (ii) the uniqueness of solutions of the Dirichlet boundary value problem for the wave equation.
-Together these facts allow us to reconstruct the pressure field, or the significant part of it, from the transmission data, and to approximate a distributed source responsible for a given set of data traces.
-
-%The geometric assumption is that the pressure field is {\em outgoing} at the recording datum: that is, that it is the restriction to $z>z_s$ of a pressure field that is negligible in $z<z_r$ for large enough time.
-``Negligible'', rather than vanishing, as the latter is strictly speaking impossible: this property is asymptotic in frequency, and assured by an assumption about the ray field associated to the acoustic system.
-
-%NB: the definition of the forward map (and the data) should include a taper $\phi(\bx_r,\bx_s)$ masking out the shot gather $d_r(\cdot,\cdot;\bx_s)$ for each source position $\bx_s$.
-However I will not explicitly mention this operation.
-I will also ignore it in the numerical examples, but there will be a price to pay.
-
-To begin with, change notation slightly, and write $S[c]h_s$ rather than $S[c]g$ with $g = h_s \delta(z-z_s)$, as $z_s$ will be presumed ot be the same for all shots. 
-%The addition of the subscript indicates that the sources lie on the plane $z=z_s$ and the receivers are on the plane $z=z_r$.
-As will be apparent shortly, a similar operator with source and receiver planes swapped will also be important, hence the need for the additional information in the notation.
-
-The pressure and velocity fields occuring in the computation of $d_r=p|_{z=z_r}=S[c]_{s,r}h_s$ form the solution of
-\begin{eqnarray}
-\label{eqn:awedata}
-\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
-h \delta(z-z_s)), \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
-p & =& 0 \mbox{ for } t \ll 0 \nonumber\\ 
-\bv & = & 0 \mbox{ for } t \ll 0 \nonumber\\
-\end{eqnarray}
-Actually $p,\bv$ form a {\em weak solution} of the system \ref{eqn:awedata}, defined by integration by parts.
-A more general setting for this type of system provides an advantageous framework within which to explain the properties of these weak solutions.
-
-Let $\omega \subset \bR^d$ be a smooth orientable surface. Then the distribution $\delta_{\omega}$ is well-defined: for $u \in C_0^{\infty}(\bR^d)$, 
-\[
-\langle \delta_{\omega}, u \rangle = \int_{\omega} u dS
-\]
-The system \ref{eqn:awedata} is a special case of
-\begin{eqnarray}
-\label{eqn:aweomega}
-\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
-h \delta_{\omega}, \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
-p & =& 0 \mbox{ for } t \ll 0 \nonumber\\ 
-\bv & = & 0 \mbox{ for } t \ll 0 \nonumber\\
-\end{eqnarray}
-in which $h \in C_0^{\infty}(\omega)$.
-
-\begin{thm}
-\label{thm:bvs}
-Suppose that $\kappa,\beta$ are smooth.
-Then the system \ref{eqn:aweomega} has a unique weak solution $(p,\bv)$.
-Denote by $[v_n] \in C^{\infty}(\omega)$ the jump of the normal component of $\bv$ across $\omega$:
-\[
-[v_n](\bx) = \lim_{\epsilon \rightarrow 0} (\bv(\bx + \epsilon {\bf n}(\bx))-\bv(\bx - \epsilon {\bf n}(\bx)))\cdot {\bf n}(\bx),\,\bx \in \omega.
-\]
-Then $h(\bx) = [v_n](\bx), \bx \in \omega$.
-
-%Let $p,\bv$ be the weak solution of the system \ref{eqn:awedata}, and assume that $\kappa, \beta$ are smooth.
-Denote by $[v_z]|_{z=z_s}$ the jump of the z-component of particle velocity in the z-direction:
-%\[
-%[v_z](x,z) = v_z^+(x,z)-v_z^-(x,z),  v_z^{\pm}(x,z)=\lim_{z'\rightarrow z^{\pm}} v_z(x,z')
-%\]
-%Then $[v_z](x,z)=0$ unless $z=z_s$, and
-%\[
-%h(x) = [v_z](x,z_s).
-%\]
-\end{thm}
-For the proof, see  Appendix B.
-
-For the rest of this section, $\omega$ will be a coordinate hyperplane, defined by $z=$ constant, with its usual orientation, and I will write $[v_z] = [v_b]$.
-
-To begin with, change notation slightly, and write $S[c]h_s$ rather than $S[c]g$ with $g = h_s \delta(z-z_s)$, as $z_s$ will be presumed ot be the same for all shots. 
-%The addition of the subscript indicates that the sources lie on the plane $z=z_s$ and the receivers are on the plane $z=z_r$.
-As will be apparent shortly, a similar operator with source and receiver planes swapped will also be important, hence the need for the additional information in the notation.
-
-The pressure and velocity fields occuring in the computation of $d_r=p|_{z=z_r}=S[c]_{s,r}h_s$ form the solution of
-\begin{eqnarray}
-\label{eqn:awedata1}
-\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
-h \delta(z-z_s)), \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
-p & =& 0 \mbox{ for } t \ll 0 \nonumber\\ 
-\bv & = & 0 \mbox{ for } t \ll 0 \nonumber\\
-\end{eqnarray}
-Actually $p,\bv$ form a {\em weak solution} of the system \ref{eqn:awedata1}, defined by integration by parts.
-\begin{thm}
-\label{thm:vz2}
-If $\kappa,\beta$ are constant near the $z=z_s$, then $v_z^-=-v_z^+$ so $[v_z] = 2 v_n^+$. 
-\end{thm}
-(for proof see Appendix 2). The formulation is for 2D propagation, but a precisely similar fact is true for 3D - in that case, $x$ is replaced everywhere by the horizontal coordinates.
-
-The import of Theorem \ref{thm:bvs} is that we can read off the source if we can infer the (particle) velocity field near the source surface, from knowledge of the data traces (values of the pressure at $z=z_r$).
-The key observation that enables this feat is a well-posedness theorem for the wave equation:
-
-\begin{thm} \label{thm:bvp}
-Suppose that $\omega \subset \bR^d$ is a smooth orientable surface, $\omega = \partial  \Omega^- = \partial \Omega^+$, where $\Omega^{\pm}$ are the open exterior/interior regions defined by the outward unit normal field of $\omega$, and let $d \in C^{\infty}_0(\omega \times \bR)$.
-Then there exist unique smooth solutions $p^{\pm},\bv^{\pm}$ of the initial/boundary value problems
-\begin{eqnarray}
-\label{eqn:awebdry}
-\frac{\partial p^{\pm}}{\partial t} & = & - \kappa \nabla \cdot \bv^{\pm} \mbox{ in } \Omega^{\pm}, \nonumber \\
-\frac{\partial \bv^{\pm}}{\partial t} & = & - \beta \nabla p^{\pm} \mbox{ in } \Omega^{\pm},\nonumber \\
-p^{\pm} & =& 0 \mbox{ for } t \ll 0\nonumber\\ 
-\bv^{\pm} & = & 0 \mbox{ for } t \ll 0 \nonumber\\
-\lim_{\bx \rightarrow \omega^{\pm}}p & = & d 
-\end{eqnarray}
-\end{thm}
-
-Putting the two theorems together, obtain
-\begin{cor}\label{thm:bvps}
-Suppose that $d_s$ is a sufficiently regular function, = 0 for $t<0$, and denote by $p^{\pm},\bv^{\pm}$ the half-space solutions provided by Theorem \ref{thm:bvp}.
-Define the vector fields $p,\bv$ in $\bR^4$ by
-\begin{eqnarray}
-\label{eqn:putemtogether} 
-p(x,z) &=& p^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
-p(x,z_s)&=&d_s(x)\\
-\bv(x,z) &=& \bv^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
-\end{eqnarray}
-Then $p,\bv$ is a weak solution of the system \ref{eqn:awedata1} with $h=[v]_n$.
-\end{cor}
-
-Now invoke the {\em outgoing assumption}: $p,\bv$ is the restriction to $z>z_s$ of a solution vanishing in $z<z_r$ for large $t$.
-This condition needs to be interpreted asymptotically, that is, $p,\bv$ is relatively smooth (lower order in frequency) for $t \gg 0, z<z_r$.
-It holds if the rays carrying significant energy in $p,\bv$ intersect $z=z_r$ transversally, and do not return to $z<z_r$ in forward time.
-The examples below will satisfy this condition, because the wave velocity is constant in $z > z_r$.
-
-Note that isotropic point sources do not produce fields outgoing in a half space, in the sense just described - simply because they radiate in all directions.
-However, for any specified finite time of propagation, it is possible to replace a point source by an anisotropic source that fits the point source data to within an arbitrarily small error, and does produce outgoing fields.
-In fact, the numerical examples below will construct such sources.
-
-This assumption implies that for $t$ sufficiently large, $d_r = p|_{z=z_r}$ is negligible for $t \gg 0$, so $p,\bv$ differs in $z_s<z < z_r$ by a negligible error from the solution of $p^-,\bv^-$, part of the solution of the  backwards-in-time  initial value problem, similar to the system \ref{eqn:awebdry}:
-\begin{eqnarray}
-\label{eqn:awetr}
-\frac{\partial \tilde{p}^{\pm}}{\partial t} & = & - \kappa \nabla \cdot \tilde{\bv}^{\pm}, \pm(z-z_r)>0 \nonumber \\
-\frac{\partial \tilde{\bv}^{\pm}}{\partial t} & = & - \beta \nabla \tilde{p}^{\pm},\nonumber \\
-\tilde{p}^{\pm} & =& 0 \mbox{ for } t \gg 0,\nonumber\\ 
-\tilde{\bv}{\pm} & = & 0 \mbox{ for } t \gg 0,\nonumber\\
-\lim_{z \rightarrow z_r^{\pm}}\tilde{p}^{\pm} &=&  d_r, \nonumber\\,
-\tilde{p} &=& \tilde{p}^{\pm}, \pm(z-z_r)>0, \nonumber\\, 
-\tilde{\bv} &=& \tilde{\bv}^{\pm}, \pm(z-z_r)>0.
-\end{eqnarray}
-
-This must be so, as $p,\bv$ satisfies the wave equation system (first two conditions), and the rest as well except for a negligible error.
-The last condition must be interpreted consistently, that is $\tilde{p}=0$ for $t$ large enough that $d_r$ is negligible.
-(In practice consistency will be obtained by replacing $t \gg 0$ respectively $t \ll \infty$ by $t > t_{\rm max}$, $t < t_{\rm max}$ for a suitably large $t_{\rm max}$).
-Then the standard energy estimate shows that $p, \tilde{p}$ differ also by a negligible error: in particular, $p|_{z=z_s} \approx \tilde{p}_{z=z_s}$.
-
-According to Corollary \ref{thm:bvps}, recovering $p_{z=z_s}$ entails recovering $h_s=[v_z]|_{z=z_s}$. So a procedure for computing an approximate inverse $S[c]_{s,r}^{\dagger}$ for $S[c]_{s,r}$ is 
-\begin{equation}
-\label{eqn:psinv}
-S[c]_{s,r}^{\dagger}d_r\equiv [v_z]_{z=z_s}.
-\end{equation}
-where the jump in $v_z$ at $z=z_s$ is computed from knowledge of $d_r=p|_{z=z_r}$ by
-\begin{itemize}
-\item solving \ref{eqn:awetr} backwards in time,
-\item reading off $d_s=\tilde{p}$ at $z=z_s$,
-\item solving \ref{eqn:awebdry} forwards in time, and computing $[v_z]|_{z=z_s}$, or if appropriate ($\kappa$ and $\beta$ constant near $z=z_s$) simply using $\bar{v}_{z=z_s^+} = v_z^{+}=-v_z^{-}$ at $z=z_s$ (Theorem \ref{thm:bvs}).
-\end{itemize}
-
-The relations developed in the last few paragraphs may be summarized as follows:
-\begin{equation}
-\label{eqn:appinv}
-d_r = S[c]_{s,r}h_s \,\,\Leftrightarrow \,\, h_s \approx S[c]_{s,r}^{\dagger}d_r
-\end{equation}
-that is, $S[c]_{s,r}^{\dagger}$ is an approximate inverse of $S[c]_{s,r}$.
-
-Inspection of systems \ref{eqn:awebdry} and \ref{eqn:awetr} reveals that they are actually the same, except for the direction of time, and that observation leads to another useful relation.
-Denote by $R$ the time-reversal operator:
-\[
-Ru(\bx,t)=u(\bx,-t)
-\]
-and by {\cal R} the time reversal operator for acoustic fields:
-\[
-{\cal R}\left(\begin{array}{c}
-p\\
-\bv
-\end{array}
-\right) = 
-\left(\begin{array}{c}
-Rp\\
--R\bv
-\end{array}
-\right) 
-\]
-The sign in the definition of ${\cal R}$ is chosen so that if $(\tilde{p},\tilde{\bv})^T$ solves the system \ref{eqn:awetr}, then ${\cal R}(\tilde{p},\tilde{\bv})^T=(\hat{p},\hat{\bv})^T$ is the solution of
-\begin{eqnarray}
-\label{eqn:awetrs}
-\frac{\partial \hat{p}^{\pm}}{\partial t} & = & - \kappa \nabla \cdot \hat{\bv}^{\pm}, \pm(z-z_r) > 0 \nonumber \\
-\frac{\partial \hat{\bv}^{\pm}}{\partial t} & = & - \beta \nabla \hat{p}^{\pm},\nonumber \\
-\hat{p}^{\pm} & =& 0 \mbox{ for } t \ll 0,\nonumber\\ 
-\hat{\bv}^{\pm} & = & 0 \mbox{ for } t \ll 0,\nonumber\\
-\lim _{z=z_r^{\pm}}\hat{p}^{\pm} &=&  Rd_r,\nonumber\\
-\hat{p} &=& \hat{p}^{\pm}, \pm(z-z_r)>0, \nonumber\\, 
-\hat{\bv} &=& \hat{\bv}^{\pm}, \pm(z-z_r)>0.
-\end{eqnarray}
-
-%For a smooth orientable surface $\omega$, denote by $\Lambda[c,\omega]$ the operator that maps Dirichlet pressure data $d$ on $\omega \times \bR$ to the jump in normal velocity $[v_n]$ across $\omega \times \bR$ via solution of the system \ref{eqn:awebdry}.
-
- From Corollary \ref{thm:bvps} applied to the system \ref{eqn:awetrs}, $\hat{p}|_{z=z_s} = S[c]_{r,s}\Lambda[c]_rRd_r$.
-Denote by $\Lambda[c]_r$ the same operator with $z_s$ replaced by $z_r$.
-Then $\hat{p}|_{z=z_s} = S[c]_{r,s}\Lambda[c]_rRd_r=R\tilde{p}|_{z=z_s}$.
-Thus the definition \ref{eqn:psinv} implies
-\begin{equation}
-\label{eqn:psinvrev}
-S[c]_{s,r}^{\dagger}= -\Lambda[c]_sRS[c]_{r,s}R\Lambda[c]_r
-\end{equation}
-since $\Lambda[c]_rR=-R\Lambda[c]_r$ from the definition of ${\cal R}$.
-Define $T[c]_{s,r}d_s = p|_{z=z_r}$ for the solution of \ref{eqn:awebdry}.
-Since ${\cal R}^2=I$, Then $\hat{p}|_{z=z_s} = T[c]_{r,s}Rd_r$.
-Since ${\cal R}^2=I$, ${\cal R}(\hat{p},\hat{\bv})^T=(\tilde{p},\tilde{\bv})^T$, whence $\tilde{p}|_{z=z_s}=RT[c]_{r,s}Rd_r$.
-
-Denote by $\Lambda[c]_s$ the operator that produces the jump in normal velocity $[v_z]=\lim_{z \rightarrow z_s^+}v_z^+ - \lim_{z\rightarrow z_s^-} v_z^-$ from Dirichlet pressure data $d_s$ and causal solution of the system \ref{eqn:awebdry}.
-From Corollary \ref{thm:bvps}, $T[c]_{s,r} = S[c]_{s,r}\Lambda[c]_s$.
-Denote by $\Lambda[c]_r$ the same operator with $z_s$ replaced by $z_r$.
-Then $\hat{p}|_{z=z_s} = S[c]_{r,s}\Lambda[c]_rRd_r=R\tilde{p}|_{z=z_s}$.
-Thus the definition \ref{eqn:psinv} implies
-\begin{equation}
-\label{eqn:psinvrev}
-S[c]_{s,r}^{\dagger}= -\Lambda[c]_sRS[c]_{r,s}R\Lambda[c]_r
-\end{equation}
-since $\Lambda[c]_rR=-R\Lambda[c]_r$ from the definition of ${\cal R}$.
-
-As I will show in the example section, the approximation may be accurate enough without any further improvement.
-However, a closer examination shows that in fact it is possible to express the relation between $S[c]_{s,r}$ and $S[c]_{s,r}^{\dagger}$ in terms of a change of norm, that is, in terms of a preconditioner, and this relation can be used to considerably accelerate an iterative solution of the normal equation \ref{eqn:normal}.
-
-To see this, compute the formal or $L^2$ adjoint $S[c]_{s,r}^T$, by a variant of the adjoint state method, in this case a by-product of the conservation of energy.
-Suppose that $\bar{h}_r$ is a function on $z=z_r$ (data traces), and
-$\bar{p},\bar{\bv}$ solve the backwards-in-time boundary value problem:
-\begin{eqnarray}
-\label{eqn:aweadj}
-\frac{\partial \bar{p}}{\partial t} & = & \kappa (-\nabla \cdot \bar{\bv} + 
- \bar{h}_r \delta(z-z_r))\nonumber \\
-\frac{\partial \bar{\bv}}{\partial t} & = & - \beta \nabla \bar{p},\nonumber \\
-\bar{p} & =& 0 \mbox{ for } t \gg 0\nonumber\\ 
-\bar{\bv} & = & 0 \mbox{ for } t \gg 0 
-\end{eqnarray} 
-Then
-\[
-0 = 
-\left(\int\, dx\,dy\,dz\, \frac{p \bar{p}}{\kappa} +  
-\frac{\bv \cdot \bar{\bv}}{\beta} \right)|_{t=T}
--
-\left(\int\, dx\,dy\,dz\, \frac{p \bar{p}}{\kappa} +  \frac{\bv \cdot \bar{\bv}}{\beta} \right)|_{t=0}
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \bar{p}}{\kappa} +  \frac{\bv \cdot \bar{\bv}}{\beta} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \bar {p} +  p \frac{1}{\kappa}\frac{\partial \bar{p}}{\partial t} \right.
-\]
-\[
-+
-\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \bar{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \bar{\bv}}{\partial t} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
- h_s \delta(z-z_s)\right) \bar{p} + p \left(- \nabla \cdot \bar{\bv} + 
- d_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  (- \nabla p) \cdot \bar{\bv} + \bv \cdot (-\nabla \bar{p}) \right)
-\]
-\[
-= 
-\int_{0}^{T}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
- h_s \delta(z-z_s)\right) \bar{p} + p \left(- \nabla \cdot \bar{\bv} + 
- d_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  p (\nabla \cdot \bar{\bv}) + (\nabla \cdot \bv) \bar{p} \right)
-\]
-after integration by parts in the last two terms. Most of what is left cancels, leaving 
-\[
-0 = \int\,dt\,dx\,dy\, (h_s \bar{p})_{z=z_s} + (p  \bar{h}_r)_{z=z_r}
-\]
-whence
-\begin{equation}
-\label{eqn:sadj}
- S[c]_{s,r}^T \bar{h}_r = -\bar{p}|_{z=z_s}
-\end{equation}
-Compare the systems \ref{eqn:awetr} and \ref{eqn:aweadj}: they differ in that 
-\begin{itemize}
-\item in \ref{eqn:awetr}, $d_r$ appears as a Dirichlet condition on $z=z_r$,
-\item in \ref{eqn:aweadj}, $\bar{h}_r$ appears as a distributed source on $z=z_r$.
-\end{itemize}
-According to Corollary \ref{thm:bvps}, $\bar{p}=\bar{p}, \bar{\bv}=\bar{\bv}$ if $\bar{h}_r = [\bar{v}_z]|_{z=z_r}$, i.e.
-the source is that which generates the same solution as the Dirichlet condition..
-
-Simlarly denote by $\tilde{\Lambda}[c]_r$ the operator that produces the jump in normal velocity at $z=z_r$ from pressure data $d_r$ via {\em anticausal} solution of the system 
-\begin{eqnarray}
-\label{eqn:awebdrytr}
-\frac{\partial {\tilde p}^{\pm}}{\partial t} & = & - \kappa \nabla \cdot {\tilde \bv}^{\pm} \mbox{ in } \pm(z-z_r) > 0, \nonumber \\
-\frac{\partial {\tilde \bv}^{\pm}}{\partial t} & = & - \beta \nabla {\tilde p}^{\pm} \mbox{ in } \pm(z-z_r) > 0,\nonumber \\
-{\tilde p}^{\pm} & =& 0 \mbox{ for } t \gg 0\nonumber\\ 
-{\tilde \bv}^{\pm} & = & 0 \mbox{ for } t \gg 0 \nonumber\\
-\lim_{z\rightarrow z_r^{\pm}}{\tilde p} & = & {\tilde d}_r 
-\end{eqnarray}
-%\end{thm}
- for the receiver surface $z=z_r$ (in the math literature, this is the ``Dirichlet-to-Neumann operator'', more or less). The discussion so far can be summarized as follows:
-\begin{equation}
-\label{eqn:wadj}
-\bar{p}|_{z=z_s} = S[c]_{s,r}^T\Lambda[c]_r \bar{p}|_{z=z_r} 
-\end{equation}
-According to the discussion above, $\bar{p}|_{z=z_s} \approx p|_{z=z_s}$, and $h_s = \Lambda[c]_s p|_{z=z_s} = S[c]_{s,r}^{\dagger}d_r$. That is,
-\begin{equation}
-\label{eqn:appinv}
-S[c]_{s,r}^{\dagger} = \Lambda[c]_s S[c]_{s,r}^T \Lambda[c]_r
-\end{equation}
-This relation can be re-written in the useful form
-\begin{equation}
-\label{eqn:adj}
-S[c]_{s,r}^{\dagger} = W_m[c]^{-1}S[c]_{s,r}^TW_d[c]
-\end{equation}
-in which 
-\begin{equation}
-\label{eqn:weights}
-W_m[c] = \Lambda[c]_s^{-1}\,\, W[c]_d = \Lambda[c]_r
-\end{equation}
-Since $\Lambda[c]$ is positive (semi-)definite, equation \ref{eqn:adj} exhibits $S[c]_{s,r}^{\dagger}$ as the adjoint of $S[c]_{s,r}$ with respect to weighted norms 
-\begin{itemize}
-\item on the source space ($h_s$): weight $W[c]_m$
-\item on the data space ($d_r$): weight $W[c]_d$
-\end{itemize}
-Since $S[c]_{s,r}^{\dagger}S[c]_{s,r} \approx I$ as shown above, $S[c]_{s,r}$ is approximately unitary in the Hilbert spaces of source and data traces with norms defined by the weighting operators $W_m$ and $W_d$ respectively.
-Therefore a Krylov space method employing these norms will converge rapidly.
-
-\cite{HouSymes:EAGE16} demonstrated a very similar preconditioner for Least Squares Migration, also for its subsurface offset extension \cite[]{HouSymes:16}, motivated by \cite{tenKroode:12}.
-These constructions all involve the Dirichlet-to-Neumann operator.
-This concept also turns up in hidden form in the work of Yu Zhang and collaborators on true amplitude migration \cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,ZhangYuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.
-
-\subsection{Implementation Considerations}
-The Dirichlet-to-Neumann operator $\Lambda$ is an essential part of the inner problem preconditioner just presented. Implementation can be accomplished in several ways:
-\begin{itemize}
-\item \cite{tenKroode:12} suggests using a one-way operator;
-\item if both pressure and normal partical velocity are measured (or simulated), then the two are related by $\Lambda$ and the velocity component can simply be used as the output;
-\item presence of a free surface implies all of the usual problems, such as the need for removal of receiver-side ghosts.
-On the other hand, if the free surface is within a quarter-wavelength throughout the useful bandwidth of the data, then the ghosted data differs from $\Lambda p$ by a time integration and a scale factor, a fact used to good effect by \cite{HouSymes:15}.
-\end{itemize}
-
-In the examples, I have used the second observation.
-With a finite difference implementation of the pressure-velocity system \ref{eqn:awe}, velocity components are available ``for free'', short-circuiting explicit computation of $\Lambda$.
-
-The variable projection gradient formula \ref{eqn:l2grad} has an Achilles heel, not mentioned so far: it is correct only when the normal equation \ref{eqn:normal} is satisfied {\em exactly}.
-Any iterative method (or the asymptotic inverse developed in the last subsection) solves \ref{eqn:normal} only approximately, so a nonzero error term is neglected in using \ref{eqn:l2grad}.
-This error term can be large, because it amplifies the high frequencies in the solution error, in fact depends on derivatives of the solution.
-For pure transmission through transparent (smooth) material models, the approximate inverse $S[c]_{s,r}^{\dagger}$ is asymptotically exact (at least in principle), and that fact can be used to devise a modification of \ref{eqn:l2grad} that converges with a convergent iteration for \ref{eqn:normal}.
-[LAST YEAR TRIP REPORT, MAY PUT SOMETHING IN APPENDIX]
-
-Choice of the penalty weight $\alpha$, and of region in which the extended source $g$ is permitted to be nonzero, must be managed.
-These two choices are linked.
-Lei Fu's thesis \cite[]{Fu:Geo17,Fu:Geo17b} explains and tests algorithms for similar choices that occur in subsurface offset WEMVA.
-All of these ideas apply {\em ipso facto} to the surface source extension.
-
-\subsection{Practical Preconditioning}
-
-The formula \ref{eqn:adj} allows direct use of the conjugate gradient (or other Krylov space iterative) algorithm with the inner products described above.
-This is however computationally inconvenient, as it requires explicit computation of the weight operators, in this case the Dirichlet-to-Neumann operator.
-There are several ways to carry out this computation, as mentioned somewhere, but it is in any case a  pain.
-On the other hand, it is possible to rearrange the calculation with weighted norms into the form of the {\em preconditioned conjugate gradient (PCG) algorithm}.
-For the current problem, identify $A=S[c]_{s,r}$, $A^* = S[c]_{s,r}^T W_d$.
-PCG takes the form
-\begin{algorithm}[H]
-\caption{Preconditioned Conjugate Gradient Algorithm}
-\begin{algorithmic}[1]
-\State Choose $x_0$ somehow
-  \State $r_0 \gets A^*(b-Ax_0)$
-  \State $p_0 \gets W_m^{-1}r_0$
-  \State $g_0 \gets p_0$
-  \State $q_0 \gets A^*Ap_0$
-  \State $k \gets 0$
-  \Repeat
-  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle p_k,q_k\rangle}$
-  \State $x_{k+1} \gets x_k + \alpha_k p_k$
-  \State $r_{k+1} \gets r_k - \alpha_kq_k$
-  \State $g_{k+1} \gets W_m^{-1}r_{k+1}$
-  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
-  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
-  \State $q_{k+1} \gets A^*Ap_{k+1}$
-  \State $k \gets k+1$
-  \Until{Error is sufficiently small, or max iteration count exceeded} 
-\end{algorithmic}
-\end{algorithm}
-Of course, this variant of CG merely hides the data sides weight operator, and requires application on the model side. This is still inconvenient.
-
-The simplest solution is to use the coexistence of the $d_r=p$ and $\Lambda[c]_rd_r =[v_z]=2v_z$ traces (assuming a horizontal datum), and similarly for the source datum.
-That is, the computation of $\Lambda[c]_r$ is intrinsic: when $S[c]_{s,r}h_s$ has been computed, so has $\Lambda[c]_rS[c]h_s$ - it is simply the collection of corresponding vertical velocity traces.
-Similarly,  $\Lambda[c]_sS[c]^T$ is the collection of vertical velocity traces at $z=z_s$ arising from the time-reversal computation input equal to sources traces on $z=z_r$.
-If we can rewrite the PCG loop to require only these computations, then the acoustic simulations in forward and reverse time will suffice to compute the step.
-
-In the loop body above, introduce the vector sequences $t_k=Ap_k,v_k=W_d t_k, s_k=W_m^{-1}q_k$, and re-write the iteration by unfolding $W_m^{-1}S[c]^T W_d S[c]$ and noting that $\langle p_k, q_k \rangle =\langle p_k,A^TW_dAp_k \rangle =\langle Ap_k,W_dAp_k \rangle =\langle t_k,v_k\rangle$:
-\begin{algorithm}[H]
-\caption{Modified Preconditioned Conjugate Gradient Algorithm}
-\begin{algorithmic}[1]
-\State Choose $x_0$ somehow
-  \State $r_0 \gets A^*(b-Ax_0)$
-  \State $p_0 \gets W_m^{-1}r_0$
-  \State $g_0 \gets p_0$
-  \State $k \gets 0$
-  \Repeat
-  \State $t_k \gets Ap_k$
-  \State $v_k \gets W_dt_k$
-  \State $q_k \gets A^Tv_k$
-  \State $s_k \gets W_m^{-1}q_k$
-  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle t_k,v_k\rangle}$
-  \State $x_{k+1} \gets x_k + \alpha_k p_k$
-  \State $r_{k+1} \gets r_k - \alpha_kq_k$
-  \State $g_{k+1} \gets g_k - \alpha_k s_k$
-  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
-  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
-  \State $k \gets k+1$
-  \Until{Error is sufficiently small, or max iteration count exceeded} 
-\end{algorithmic}
-\end{algorithm}
-This rewrite does not look like an improvement - four operator applications appear to be required:
-\begin{itemize}
-\item $p_k \rightarrow S[c] p_k = t_k$
-\item $p_k \rightarrow  \Lambda[c]_r S[c] p_k = v_k$
-\item $v_k \rightarrow S[c]^Tv_k = q_k$
-\item $v_k \rightarrow \Lambda[c]_s S[c]^T v_k = s_k$
-\end{itemize}
-However this rearrangement reveals that actually only two wave equation solves are needed.
-For $q_k$ and $s_k$, both by-products of the adjoint (time-reversed) simulation: $q_k$ stores the pressure traces on $z=z_s$, $s_k$ the velocity traces.
-The same is true of $t_k$ and $v_k$, with the latter being the input for the $q_k,s_k$ computation.
-
-Denote by $F_{pp}$ the acoustic simulation operator with a $p$ source (constitutive law defect, as in the system \ref{eqn:awedata}), and $p$ output traces.
-Similarly, denote by $F_{pv}$ the simulation operator with $p$ source and $v_z$ output traces, likewise $F_{vp}$ and $F_{vv}$, and by ${\bf F}$ the matrix operator
-\begin{equation}
-\label{eqn:iwop}
-{\bf F} = \left(
-\begin{array}{cc}
-F_{pp} & F_{vp} \\
-F_{pv} & F_{vv}
-\end{array}
-\right)
-\end{equation}
-The {\tt IWaveLOVOp} built in the asg package implements this matrix operator if the keywords {\tt source\_p, source\_v0, data\_p, data\_v0} are defined in its parameter list: these keywords point to SEGY trace files that define the spaces in which the $p$ and $v_z$ sources respectively data traces are treated as vectors.
-That is, IWAVE provides an easy access to ${\bf F}$.
-In terms of the operators used in the statement of the inverse problem, $F_{pp}=S[c]$, $F_{pv}=\Lambda[c]_rS[c]$, $F_{vp} = S[c]\Lambda[c]_s$, $F_{pp}^T = S[c]^T$, and $F_{vp}^T = \Lambda[c]_sS[c]^T$.
-So if we define
-\begin{equation}
-\label{eqn:vecvec}
-{\bf p}_k = \left[
-\begin{array}{c}
-p_k\\
-0
-\end{array}
-\right],\,\,
-{\bf r}_k = \left[
-\begin{array}{c}
-r_k\\
-g_k
-\end{array}
-\right],\,\,
-{\bf t}_k = \left[
-\begin{array}{c}
-t_k\\
-v_k
-\end{array}
-\right],\,\,
-{\bf v}_k = \left[
-\begin{array}{c}
-v_k\\
-0
-\end{array}
-\right],\,\,
-{\bf q}_k = \left[
-\begin{array}{c}
-q_k\\
-s_k
-\end{array}
-\right],
-\end{equation}
-and
-\begin{equation}
-\label{eqn:proj}
-{\bf P} = \left(
-\begin{array}{cc}
-0 & I \\
-0 & 0
-\end{array}
-\right),
-\end{equation}
-then the algorithm listing above can be rewritten as
-\begin{algorithm}[H]
-\caption{Further modified Preconditioned Conjugate Gradient Algorithm}
-\begin{algorithmic}[1]
-\State Choose $x_0$ somehow
-  \State $r_0 \gets A^TW_d(b-Ax_0)$
-  \State $p_0 \gets W_mr_0$
-  \State $g_0 \gets p_0$
-  \State $k \gets 0$
-  \Repeat
-  \State ${\bf t}_k \gets {\bf F}{\bf p}_k$
-  \State ${\bf v}_k \gets {\bf P}{\bf t}_k$
-  \State ${\bf q}_k \gets {\bf F}^T{\bf v}_k$
-  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle t_k,v_k\rangle}$
-  \State $x_{k+1} \gets x_k + \alpha_k p_k$
-  \State ${\bf r}_{k+1} \gets {\bf r}_k - \alpha_k{\bf q}_k$
-  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
-  \State ${\bf p}_{k+1}\gets {\bf P}{\bf r}_{k+1}+\beta_{k+1}{\bf p}_k$
-  \State $k \gets k+1$
-  \Until{Error is sufficiently small, or max iteration count exceeded} 
-\end{algorithmic}
-\end{algorithm}
-
-Note that
-\begin{equation}
-\label{eqn:note}
-{\bf r}_0 = \left[
-\begin{array}{c}
-r_0\\
-g_0 
-\end{array}
-\right] = 
-\left[
-\begin{array}{c}
-r_0\\
-p_0 
-\end{array}
-\right] = 
-\left[
-\begin{array}{c}
-A^TW_d(b-Ax_0)\\
-W_m^{-1}A^TW_d(b-Ax_0)
-\end{array}
-\right]
-\end{equation}
-Define 
-\begin{equation}
-\label{eqn:rhsdef}
-{\bf b}
-= 
-\left[
-\begin{array}{c}
-W_d(b-Ax_0)\\
-0
-\end{array}
-\right].
-\end{equation}
-Then identity \ref{eqn:note} implies that
-\begin{equation}
-\label{eqn:init}
-{\bf r}_0= {\bf F}^T{\bf b},\,\, {\bf p}_0 = {\bf P}{\bf r}_0.
-\end{equation}
-This observation allows a convenient ``hoist'' of the $k=0$ operator applications out of the iteration loop.
-Also, it is convenient to introduce the indefinite quadratic form that returns the inner product of the two components of a vector in the product space introduced above:
-\begin{equation}
-\label{eqn:dotfunc}
-Q({\bf x})=\langle x_0,x_1\rangle=\left\langle \bx,
-\left(
-\begin{array}{cc}
-0 & I \\
-I & 0 
-\end{array}
-\right) \bx \right\rangle
-\end{equation}
-and the scalar sequences
-\begin{eqnarray}
-\label{eqn:scalars}
-\gamma_k &=& \langle r_k,g_k \rangle = Q({\bf r})\\
-\delta_k &=& \langle t_k,v_k \rangle = Q({\bf t})
-\end{eqnarray}
-
-With these modifications, we obtain 
-\begin{algorithm}[H]
-\caption{Intrinsically Preconditioned Conjugate Gradient Algorithm}
-\begin{algorithmic}[1]
-\State Choose $x_0$ somehow
-\State ${\bf b} \gets \left[
-\begin{array}{c}
-W_d(b-Ax_0)\\
-0
-\end{array}
-\right]$
-  \State ${\bf r}_0 \gets {\bf F}^T{\bf b}$
-  \State ${\bf p}_0 \gets {\bf P}{\bf r}_0$
-  \State $\gamma_0 \gets Q({\bf r}_0)$
-  \State $k \gets 0$
-  \Repeat
-  \State ${\bf t}_k \gets {\bf F}{\bf p}_k$
-  \State $\delta_k \gets Q({\bf t})$
-  \State ${\bf v}_k \gets {\bf P}{\bf t}_k$
-  \State ${\bf q}_k \gets {\bf F}^T{\bf v}_k$
-  \State $\alpha_k \gets \frac{\gamma_k}{\delta_k}$
-  \State $x_{k+1} \gets x_k + \alpha_k p_k$
-  \State ${\bf r}_{k+1} \gets {\bf r}_k - \alpha_k{\bf q}_k$
-  \State $\gamma_{k+1} \gets Q({\bf r}_{k+1})$
-  \State $\beta_{k+1} \gets \frac{\gamma_{k+1}}{\gamma_k}$
-  \State ${\bf p}_{k+1}\gets {\bf P}{\bf r}_{k+1}+\beta_{k+1}{\bf p}_k$
-  \State $k \gets k+1$
-  \Until{Error is sufficiently small, or max iteration count exceeded} 
-\end{algorithmic}
-\end{algorithm}
-
-\section{Prototype Numerical Examples}
-\inputdir{project}
-
-I present a collection of simple examples that illustrate the features of the surface source extension claimed in preceding sections.
-
-I used the IWAVE acoustic staggered grid package to carry out these calculations.
-This package implements (2,2k) schemes for k=1,2,..., and outputs traces (of either velocity or pressure at any point in space via multilinear interpolation.
-The discretized modeling operator is thus of second order accuracy, though as usual I have used higher order in space to reduce grid dispersion.
-
-Source injection is implemented as the adjoint of trace sampling, resulting in another second-order error [REFERENCES].
-
-The data is a single shot gather, with a source at coordinates $x_s=z_s=3000$ m (units of length are meters in all cases).
-The receiver line occupies $1500 \le x_r \le 5500$ m, with receiver depth $z_r=1000$ m.
-Extended sources occupy $1500 \le x_s \le 5500$ m, with the same depth $z_s=3000$ m as the ``physical'' source used to generate the data.
-This region turns out to be adequate to represent the extended sources that approximately invert the data, for the cases examined below.
-An algorithm to automatically identify an appropriate region can be based on the ideas developed by \cite{Fu:Geo17}.
-
-I have used absorbing boundary conditions (split-field PML) on all four sides of the 4000 m (vertical) $\times$ 8000 m (horizontal) simulation domain.
-Evidently includion of a free surface is important to the application of the ideas explained here to diving wave marine data, and I have not addressed the necessary modifications here.
-
-The script is set up to carry out the necessary computations on grids with spacings $\Delta x = \Delta z = $ 20, 10, and 5 m, with a jump of roughly 8 in computation time resulting from each refinement.
-For present purposes, the coarsest (20 m) grid seems to be sufficient, and that is the grid used in the examples presented below.
-The source pulses are chosen so that the computation is reasonably accurate.
-For the 20 m grid case, I use a zero-phase trapezoidal bandpass filter source with corner frequencies of 1.0, 2.0, 7.5, and 12.5 Hz.
-
-The IWAVE asg driver has been set up to recognize the case {\tt deriv=0} as defining the map from source (right-hand side in the pressure equation) to data (pressure) traces.
-The adjoint to this map, as explained above, is reverse-time propagation of the data traces as pressure sources, followed by scaling (formula \ref{eqn:sadj}).
-The approximate inverse is computed by application of the Dirichlet-to-Neumann operator to the pressure traces to produce corresponding velocity traces, followed by injection as pressure sources and reverse time propagation, followed by another application of the Dirichlet-to-Neumann map and scaling (formulas \ref{eqn:appinv}, \ref{eqn:adj}).
-
-The script for the first set of examples implements these operatations step-by-step via calls to IWAVE, Madagascar, and SU commands.
-A peculiarity of the {\tt asg.x} driver needs to be mentioned: it is based on an un-scaled version of the constitutive law defect source representation, that is, {\tt asg.x} approximately computes the solution of the system \ref{eqn:awedata} with the first equation replaced by
-\begin{equation}
-\label{eqn:asgdata}
-\frac{\partial p}{\partial t}  =  - \kappa \nabla \cdot \bv +
-h \delta(z-z_s).
-\end{equation}
-Denote by $S_{\rm asg}[c]$ the forward map produced by {\tt asg.x}. Then comparison of \ref{eqn:awedata} and \ref{eqn:asgdata} reveals that
-\begin{equation}
-\label{eqn:sreln}
-S[c]=S_{\rm asg}[c]\kappa
-\end{equation}
-where $\kappa$ is shorthand for the operator of multiplication by $\kappa$. Accordingly, and approximate inverse for $S_{\rm asg}[c]$ is
-\[
-I \approx S_{\rm asg}[c]^{\dagger}S_{\rm asg}[c] = S_{\rm asg}[c]^{\dagger}S[c]\kappa^{-1}
-\]
-Since $S[c]S[c]^{\dagger} \approx I \approx S[c]^{\dagger}S[c]$, it follows that
-\[
-S_{\rm asg}[c]^{\dagger}=\kappa S[c]^{\dagger} = \kappa\Lambda[c]_s S[c]^T \Lambda[c]_r 
-\]
-\begin{equation}
-\label{eqn:asginv}
-= \kappa \Lambda[c]_s \kappa S_{\rm asg}[c]^T\Lambda[c]_r
-\end{equation}
-This is the approximate inverse computed in the examples.
-Note that only the values of $\kappa$ near the source datum $z=z_s$ play a role in the relation \ref{eqn:sreln} or in the definition \ref{eqn:asginv} of the IWAVE ASG approximate inverse.
-
-These examples are based on two different material models.
-The first is a homogeneous model, with $\kappa=4.0$ GPa and $\rho=1.0$ g/cm$^{3}$.
-Figure \ref{fig:ptpwindh0} shows the shot gather computed in the configuration descritbed above for this choice of propagation medium.
-As mentioned in the preceding section, the velocity traces can be recorded in the same IWAVE workflow: these are displayed in
-\ref{fig:ptvzwindh0}.
-
-The first group of plots show pressure traces for source position $z_s=3$ km, $x_s=3$ km, 201 receivers at $z_r=1$ km, $2 \le x_r \le 6$ km.
-
-\plot{ptpwindh0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km.
-Homogeneous material model, $\kappa=4$ GPa, $\rho$=1 g/cm$^{3}$.
-Coarse grid: $dx=dz=20$ m, $dt=$ 8 ms, (2,4) staggered grid scheme.
-}
-\plot{ptvzwindhh0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindh0}.}
-\plot{srcvzghh0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindhh0} by time reversal (approximate inverse) algorithm explained in text.}
-\plot{reptpwindhh0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzghh0}.} 
-\plot{trcftrhh0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindh0}, \ref{fig:reptpwindhh0}.}.
-
-\plot{bml0}{width=15cm}{Inhomogenous bulk modulus field, with low velocity lens.}
-\plot{ptpwindl0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km.
-bulk modulus with low velocity lens as in Figure \ref{fig:bml0}, $\rho$=1 g/cm$^{3}$.
-Coarse grid: $dx=dz=20$ m, $dt=$ 8 ms, (2,4) staggered grid scheme.
-}
-\plot{ptvzwindll0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindl0}.}
-
-\plot{srcvzglh0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindll0} by time reversal (approximate inverse) algorithm explained in text, propagation in homogeneous material model (that is, ``wrong velocity'').}
-\plot{reptpwindlh0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzglh0}, both approximate inversion and re-simulation in  homogeneous material model (``wrong velocity'').}
-\plot{trcftrlh0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindl0}, \ref{fig:reptpwindlh0}.}
-
-\plot{srcvzgll0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindll0} by time reversal (approximate inverse) algorithm explained in text, propagation in bulk modulus of Figure \ref{fig:bml0} (that is, ``correct velocity'').}
-\plot{reptpwindll0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzgll0}, both approximate inversion and re-simulation in bulk modulus of Figure \ref{fig:bml0} (``correct velocity'').}
-\plot{trcftrll0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindl0}, \ref{fig:reptpwindll0}.}
+%\input{prop.tex}
 
@@
 
-\append{RWI (=Wang-Yingst WRI) - a Linear Algebra Viewpoint}
-
-Define $S[c] = PL[c]^{-1}$ = operator mapping extended source to data traces.
-
-The Wang-Yingst version of the WRI objective function may be written
-\[
-\|S[c]g-d\|^2 + \alpha \|g-f\|^2
-\]
-``Most'' extended sources result in zero data traces - those are non-radiating sources, the null space of $S[c]$.
-
-Rank Nullity Theorem: any $g$ may be written in a unique way as $g=S[c]^Th + n$, where $h$ is a collection of data traces, $S[c]^T$ is back propagation, and $n$ is in the null space of $S[c]$.
-Furthermore, the decomposition is orthogonal.
-Thus
-\[
-\|g-f\|^2 = \|S[c]^Th + n -f\|^2 = \|S[c]^Th - f_p\|^2 + \|n-f_n\|^2
-\]
-in which $f_p$ and $f_n$ are the orthoprojections on the orthocomplement of the null space, and the null space, of $S[c]$ respectively.
-
-There is no use keeping $f_n$ around, as it radiates nothing - in fact, I suspect it is possible to see that point sources such as represented in \ref{eqn:ptsrc} are perpindicular to the null space already.
-In any case, simply choose $n=f_n$ and eliminate the second term, without affecting the data fit (by definition!).
-
-Then WRI objective function can be re-written as
-\begin{equation}
-\label{eqn:wriobj}
-\|S[c]S[c]^Th-d\|^2 + \alpha \|S[c]^Th -f_p\|^2
-\end{equation}
-and the extended model has been reduced from a full-space time volume ($g$) to a data gather ($h$).
-
-{\bf Note added 2019.03.14:} Chao Wang's talk at GS 19 revealed that that they actually store the entire source ($g$ above).
-That is not so crazy, since the operator $S[c]S[c]^T$ poses the same complexity issue as time-domain RTM.
-
-To see this, need an explicit description of $S[c]^T$.
-Regard $L[c]$ as the 2nd ord wave operator with velocity $c$: $L[c]=\frac{1}{c^2}\frac{\partial}{\partial t^2} - \nabla^2$.
-$L[c]^{-1}$ is ambiguous: define
-\[
-L[c]^{-1_+}g = u, \mbox{ for } L[c]u=g \mbox{ and } u=0 \mbox{ for } t<0
-\]
-that is, $L[c]^{-1}_+$ is the causal inverse of $L[c]$.
-Solutions defined on time interval $[0,T]$.
-Let $R$ be time-reversal: $Ru(t)=u(T-t)$.
-Then the anticausal inverse $L[c]^{-1}_-$ of $L[c]$, solution $=0$ for $t>T$, can be written as $L[c]^{-1}_- = RL[c]^{-1}_+R$.
-
-Denote by $\langle \cdot,\cdot \rangle$ the $L^2$ inner product in space.. Suppose $u$ is a causal solution of $L[c]u=f$, $w$ an anti-causal solution of $L[c]w=g$. Then 
-\[
-\int_0^T \langle L[c]^{-1}_+f, g \rangle = int_0^T \langle u, g \rangle = \int_0^T \langle u, \left( \frac{1}{c^2}\frac{\partial}{\partial t^2} - \nabla^2\right)w\rangle
-\]
-\[
-= \int_0^T \langle \left( \frac{1}{c^2}\frac{\partial}{\partial t^2} - \nabla^2\right)u, w \rangle = \int_0^T \langle f, w \rangle = \int_0^T \langle f, L[c]^{-1}_-g \rangle
-\]
-since the causal resp.
-anti-causal solutions have zero Cauchy data at $t=0$ resp.
-$t=T$, and since the integration is over all of $\bR^3$ and I am assuming everything vanishing at $\infty$.
-Conclude that
-\[
-(L[c]^{-1}_+)^T = L[c]^{-1}_-
-\]
-so
-\[
-S[c]^T = (PL[c]^{-1}_+)^T = L[c]^{-1}_-P^T
-\]
-
-For convenience, ignore sampling and aperture, and define $P$ to be the restriction operator on $z=0$. Then $(P^T h)(x,y,z,t) = h(x,y,t)\delta(z)$.
-
-The troublesome operator in the definition \ref{eqn:wriobj} is 
-\[
-S[c]S[c]^T = PL[c]^{-1}_+L[c]^{-1}_-P^T
-\]
-The first solution operator involves evolution backwards in time, and provides the data for the second solution operator.
-This leadss to the same computational complexity problem as encountered in RTM.
-It could be handled by storing the output of $L[c]^{-1}_-$, as the IONGeo folks apparently do, or storing boundary data (for acoustics - as usual, this would not work if the approach is extended to a lossy model such as viscoacoustics), or using a checkpointing scheme.
-
-Another approach: write $w = L[c]^{-1}_-P^T h$. The object is to compute $u=L[c]^{-1}_+w$, that is, the solution of 
-\[
-(c^{-2}\partial_t^2 - \nabla^2)u(t) = w(t), 0\le t \le T; \,u(0) = \partial_t u(0) = 0
-\]
-Duhamel's principle would provide a solution in the form of integrals over time of solutions of homogeneous initial value problems:
-\[
-u(t) = \int_0^t d\tau v(t,\tau), \, (c^{-2}\partial_t^2 - \nabla^2)v(t,\tau) = 0, 0\le t \le T; \,v(t,\tau)_{t=\tau} = 0, \partial_t v(t,\tau)_{t=\tau} = w(\tau)
-\]
-Note that $w$ solves the homogenous wave equation except at $z=0$. That makes it tempting to use $w$ directly to define $v$. In fact, set 
-\[
-W(t) = -int_t^T ds w(s),\,v_0(t,\tau) = \frac{1}{2}(W(t) - W(2\tau-t)) 
-\]
-Then $v_0$ satisfies the Duhamel initial conditions. Note that because $w$ has zero Cauchy data at $t=T$,
-\[
-\partial_t W(t)= -\int_t^T ds \partial_t w(s),\,\partial_t^2 W(t)=-\int_t^T ds \partial^2_t w(s)
-\]
-so 
-\[
-(c^{-2}\partial_t^2 - \nabla^2)v_0(t,\tau) = -\int_t^T(c^{-2}\partial^2_t -\nabla^2)w + \int_{2\tau-t}^T(c^{-2}\partial^2_t -\nabla^2)
-\]
-For $t>\tau$, $t > 2\tau-t$, so 
-\[
-= -\int_{2\tau-t}^t(c^{-2}\partial^2_t -\nabla^2)w = -\int_{2\tau-t}^t P^Th
-\]
-and 
-\[ 
-v_0(t,\tau) = \int_{2\tau-t}^t w  
-\] 
-Consequently $v=v_0 + v_1$, in which $v_1$ is defined by 
-\[
-(c^{-2}\partial_t^2 - \nabla^2)v_1(t,\tau) = \int_{2\tau-t}^t P^Th, \,  v_1(t,\tau)_{t=\tau} = 0, \,  \partial_t v_1(t,\tau)_{t=\tau} = 0
-\]
-
-\append{Transpose of Full Acoustic Modeling Operator}
-In this section I compute the adjoint operator form modeling with both pressure and velocity source inputs and pressure and velocity trace outputs. The forward modeling operator is defined by solving
-\begin{eqnarray}
-\label{eqn:fawedata}
-\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
-h_s \delta(z-z_s)), \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta ( \nabla p + l_s {\bf e}_z\delta(z-z_s)),\nonumber \\
-p & =& 0 \mbox{ for } t<0\nonumber\\ 
-\bv & = & 0 \mbox{ for } t<0 \nonumber\\
-\end{eqnarray}
-The new elements here are the point load time function $l_s(\bx,t;\bx_s)$, and the vertical unit vector ${\bf e}_z=(0,1)^T$ for 2D, $=(0,0,1)^T$ for 3D.
-The forward map ${\bf F}$ is defined as implicit in the discussion of the Intrinsically Preconditioned CG algorithm: let ${\bf h}_s=(h_s,l_s)^T$, then
-\begin{equation}
-\label{eqn:ffwd}
-{\bf F}{\bf h}_s =\left\{ \left[
-\begin{array}{c}
-p(\bx,t;\bx_s)\\
-v_z(\bx,t;\bx_s)
-\end{array}
-\right]_{z=z_r}\right\}
-\end{equation}
-
-Let $\tilde{\bf h}_r=(\tilde{h}_r,\tilde{l}_r)^T$ be a function on $z=z_r$, 
-and 
-$\tilde{p},\tilde{\bv}$ solve the backwards-in-time boundary value problem:
-\begin{eqnarray}
-\label{eqn:faweadj}
-\frac{\partial \tilde{p}}{\partial t} & = & \kappa (-\nabla \cdot \tilde{\bv} + 
- \tilde{h}_r \delta(z-z_r))\nonumber \\
-\frac{\partial \tilde{\bv}}{\partial t} & = & \beta (\nabla \tilde{p} + \tilde{l}_r{\bf e}_z\delta(z-z_r),\nonumber \\
-\tilde{p} & =& 0 \mbox{ for } t>T\nonumber\\ 
-\tilde{\bv} & = & 0 \mbox{ for } t>T 
-\end{eqnarray} 
-Then
-\[
-0 = 
-\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  
-\frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=T}
--
-\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=0}
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \tilde {p} +  p \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} \right.
-\]
-\[
-+
-\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \tilde{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \tilde{\bv}}{\partial t} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
- h_s \delta(z-z_s)\right) \tilde{p} + p \left(- \nabla \cdot \tilde{\bv} + 
- \tilde{h}_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  (- \nabla p + l_s{\bf e}_z\delta(z-z_s)) \cdot \tilde{\bv} + \bv \cdot (-\nabla \tilde{p} + \tilde{l}_r{\bf e}_z\delta(z-z_r)) \right)
-\]
-\[
-= 
-\int_{0}^{T}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv \tilde{p} + 
- h_s\tilde{p}|_{z=z_s} - p\nabla \cdot \tilde{\bv} + 
- \tilde{h}_r p|_{z=z_r}\right) \right.
-\]
-\[
-+
-\left.  p (\nabla \cdot \tilde{\bv}) + l_s \tilde{v}_z|_{z=z_r}+ (\nabla \cdot \bv) \tilde{p}  + l_r  v_z|_{z=z_r}\right)
-\]
-after integration by parts in the last two terms. Most of what is left cancels, leaving 
-\[
-0 = \int\,dt\,dx\,dy\, (h_s \tilde{p})_{z=z_s} + (p  \tilde{h}_r)_{z=z_r} + (l_s\tilde{v}_z)_{z=z_s} + (v_z \tilde{l}_r)_{z=z_r},
-\]
-whence
-\begin{equation}
-\label{eqn:fadj}
- {\bf F}^T {\bf h}_r = -\left[
-\begin{array}{c}
-\tilde{p}(\bx,t;\bx_s)\\
-\tilde{\bf v}(\bx,t;\bx_s)
-\end{array}
-\right]_{z=z_s}
-\end{equation}
-
-%\begin{cor}\label{thm:bvps}
-%Suppose that $d_s$ is a sufficiently regular function, = 0 for $t<0$, and denote by $p^{\pm},\bv^{\pm}$ the half-space solutions provided by Theorem \ref{thm:bvp}.
-Define the vector fields $p,\bv$ in $\bR^4$ by
-%\begin{eqnarray}
-%\label{eqn:putemtogether} 
-%p(x,z) &=& p^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
-%p(x,z_s)&=&d_s(x)\\
-%\bv(x,z) &=& \bv^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
-%\end{eqnarray}
-%Then $p,\bv$ is a weak solution of the system \ref{eqn:awedata} with $h=[v]_n$.
-%\end{cor}
-
-Proof of Theorem \ref{thm:bvp}:
-
-\begin{proof}
-Since $\omega \cap \mbox{ supp }d$ is compact, it is possible to construct an extension of 
-$d \in C^{\infty}_{0}(\bR^{d+1})$. Define $p_1^{\pm}, \bv_1^{\pm}$ to be strong solutions of the system
-\begin{eqnarray}
-\label{eqn:ichom}
-\frac{\partial p^{\pm}_1}{\partial t} & = & - \kappa \nabla \cdot \bv^{\pm}_1 + F\mbox{ in } \Omega^{\pm}, \nonumber \\
-\frac{\partial \bv^{\pm}_1}{\partial t} & = & - \beta \nabla p^{\pm}_1 + {\bf G}\mbox{ in } \Omega^{\pm},\nonumber \\
-p^{\pm}_1 & = & 0 \mbox{ on } \omega \times \bR,\\
-p^{\pm}_1 & = & 0, t \ll 0\\
-\bv^{\pm}_1 & = & 0, t \ll 0 
-\end{eqnarray}
-with 
-\[
-F = -\gamma\frac{\partial d}{\partial t}, {\bf G} = -\nabla d.
-\]
-Existence, uniqueness, and regularity of these solutions are assured by a minor extension of results of \cite{BlazekStolkSymes:13}.
-Set $p^{\pm} = p_1^{\pm} + d, \bv^{\pm} = \bv_1^{\pm}$; then $p^{\pm},\bv^{\pm}$ are the unique strong solutions of the boundary value problems \ref{eqn:awebdry}, and are smooth up to the boundary as indicated.
-\end{proof}
-
-Proof of Corollary \ref{thm:bvps}:
-
-\begin{proof}: Choose smooth functions $\phi \in C^{\infty}_0(\bR^{d+1}), {\bf \psi} \in  (C^{\infty}_0(\bR^{d+1}))^d$. Then
-\[
-\int_{\Omega^{\pm} \times \bR}\left (\gamma \frac{\partial \phi}{\partial t} +
-\nabla \cdot {\bf \psi}\right)p^{\pm} + \left(\rho\frac{\partial {\bf
-    \psi}}{\partial t}
-  + \nabla \phi\right) \cdot \bv^{\pm}
-\]
-\[
-= - \int_{\Omega \times \bR}\left (\gamma \frac{\partial p^{\pm}}{\partial t} +
-\nabla \cdot \bv^{\pm}\right) \phi + \left(\rho\frac{\partial \bv^{\pm}}{\partial t}
-  + \nabla p^{\pm}\right) \cdot {\bf \psi}
-\]
-\begin{equation}
-\label{eqn:rhs1}
-\pm \int_{\omega \times \bR} \bn \cdot {\bf \psi} p^{\pm} + \bn
-  \cdot \bv^{\pm} \phi
-\end{equation}
-Adding equations \ref{eqn:rhs1} for the two choices of sign and taking into
-account that $p^{\pm}$ coincide (by construction) on $\omega \times \bR$, obtain
-\[
-\int_{(\bR^d \times \bR}\left (\gamma \frac{\partial \phi}{\partial t} +
-\nabla \cdot {\bf \psi}\right)p + \left(\rho\frac{\partial {\bf
-    \psi}}{\partial t}
-  + \nabla \phi\right) \cdot \bv
-\]
-\[
-= \int_{\partial \Omega \times [0,T]} (\bn \cdot \bv^{+} -  \bn
-  \cdot \bv^{-}) \phi
-\]
-Since $\phi$ and ${\bf \psi}$ are arbitrary, other than being smooth of compact support, this observation finishes the proof.
-\end{proof}
-
-\append{Variable Projection}
-
-\[
-J[c]=\min_h\frac{1}{2}(\|S[c]h-d\|^2 + \alpha^2\|Ah\|^2)
-\]
-\[
-= \min_h \frac{1}{2}\|F[c]h-e\|^2
-\]
-\[
-F[c]=\left(
-\begin{array}{c}
-S[c]\\
-\alpha A
-\end{array}
-\right),\,\,
-e=\left(
-\begin{array}{c}
-d\\
-0
-\end{array}
-\right)
-\]
-
-\[
-J[c]=\frac{1}{2}\|(F[c]N[c]^{-1}F[c]^T-I)e\|^2,\,\,N[c]=F[c]^TF[c]
-\]
-\[
-DJ[c]\delta c = \langle (F[c]N[c]^{-1}F[c]^T-I)e, 
-(DF[c]\delta c)N[c]^{-1}F[c]^Te +
-F[c](-N[c]^{-1}(DN[c]\delta c)N[c]^{-1})F[c]^Te +
-F[c]N[c]^{-1}(DF[c]\delta c)^Te \rangle 
-\]
-
-\[
-= \langle (F[c]N[c]^{-1}F[c]^T-I)e, 
-(DF[c]\delta c)N[c]^{-1}F[c]^Te -
-F[c]N[c]^{-1}(DF[c]\delta c)^T F[c]N[c]^{-1})F[c]^Te -
-F[c]N[c]^{-1}F[c]^T (DF[c]\delta c]N[c]^{-1})F[c]^Te +
-F[c]N[c]^{-1}(DF[c]\delta c)^Te \rangle 
-\]
-\[
-= \langle (F[c]N[c]^{-1}F[c]^T-I)e, 
-(I-F[c]N[c]^{-1}F[c]^T) DF[c]\delta c)N[c]^{-1}F[c]^Te
-F[c]N[c]^{-1}(DF[c]\delta c)^T(I-F[c]N[c]^{-1}F[c]^T) \rangle
-\]
-\[
-=-\langle (I-F[c]N[c]^{-1}F[c]^T)^2e, DF[c]\delta c)N[c]^{-1}F[c]^Te\rangle
--\langle e,(I-F[c]N[c]^{-1}F[c]^T)F[c]N[c]^{-1}(DF[c]\delta c)^T (I-F[c]N[c]^{-1}F[c]^T)e
-\]
-\[
-F[c]N[c]^{-1}F[c]^TF[c]N[c]^{-1}=F[c]N[c]^{-1}
-\]
-so 
-\[
-(I-F[c]N[c]^{-1}F[c]^T)F[c]N[c]^{-1} = 0
-\]
-and
-\[
-(I-F[c]N[c]^{-1}F[c]^T)^2 = (I-F[c]N[c]^{-1}F[c]^T)
-\]
-Whence
-\[
-DJ[c]\delta c =
--\langle (I-F[c]N[c]^{-1}F[c]^T)e, (DF[c]\delta c)N[c]^{-1}F[c]^Te\rangle
-\]
-Define $h[c]=N[c]^{-1}F[c]^Te$ to be the solution of the inner problem as function of $c$. Then
-\[
-DJ[c]\delta c = \langle F[c]h[c]-e,(DF[c]\delta c)h[c] \rangle
-\]
-which is the usual VPM gradient formula.
-
-Gauss-Newton: define $G[c]=F[c]N[c]^{-1}F[c]^Te$. Need $DG[c], DG[c]^T$.
-\[
-DG[c]\delta c =( DF[c]\delta c)N[c]^{-1}F[c]^Te 
-\]
-\[
-- F[c]N[c]^{-1}(DF[c]\delta c)^TF[c]N[c]^{-1}F[c]^Te + F[c]N[c]^{-1} F[c]^T(DF[c]\delta c))N[c]^{-1}F[c]^Te 
-\]
-\[
-+ F[c]N[c]^{-1}(DF[c]\delta c)^Te
-\]
-\[
-=(I-F[c]N[c]^{-1} F[c]^T)(DF[c]\delta c))N[c]^{-1}F[c]^Te + F[c]N[c]^{-1}(DF[c]\delta c)^T(I-F[c]N[c]^{-1} F[c]^T) e
-\]
-Reinterpretation: set
-\begin{eqnarray}
-h[c] &=&N[c]^{-1}F[c]^Te \nonumber\\
-r[c] &= &F[c]h[c]-e \nonumber\\
-E_1[c]\delta c & = & (DF[c]\delta c)h[c] \nonumber\\
-H_1[c]\delta c &=&N[c]^{-1}F[c]^T E_1[c]\delta c \nonumber\\
-R_1[c]\delta c &=& (F[c]H_1[c]-E_1[c])\delta c \nonumber\\
-H_2[c]\delta c &=&N[c]^{-1}(DF[c]\delta c)^Tr[c] \nonumber\\
-R_2[c]\delta c &=&F[c]H_2[c]\delta c
-\end{eqnarray}
-Then
-\[
-DG[c]\delta c = -(R_1[c]+R_2[c])\delta c
-\]
-\[
-= (I-F[c]N[c]^{-1}F[c]^T)DF[c]\delta c)h[c] -F[c]N[c]^{-1}(DF[c]\delta c)^Tr[c]
-\]
-Adjoints: write $(DF[c]\delta c)h = DF[c](\delta c,h)$, 
-\begin{eqnarray}
-\langle F[c]h,\phi \rangle &=& \langle h,F[c]^T\phi\rangle \nonumber\\
-\langle DF[c](\delta c,h),\phi \rangle &=& \langle \delta c, DF[c]^*(\phi,h) \rangle
-\end{eqnarray}
-\[
-\langle DF[c](\delta c,h),\phi \rangle = \langle \delta c,DF[c]^*(\phi,h)\rangle
-\]
-\[
-\langle E_1[c]\delta c,\phi \rangle = \langle DF[c](\delta c,h[c]),\phi \rangle
-\]
-\[
-= \langle \delta c, DF[c]^*(\phi,h[c]) \rangle
-\]
-so $E_1[c]^* = DF[c]^*(\cdot,h[c])$, and
-\[
-\langle F[c]H_1[c]\delta c, \phi \rangle = \langle F[c] N[c]^{-1}F[c]^TE_1[c]\delta c,\phi \rangle
-\]
-\[
-=\langle E_1[c]\delta c,F[c]N[c]^{-1}F[c]^T\phi\rangle
-\]
-\[
-=\langle \delta c, DF[c]^*(F[c]N[c]^{-1}F[c]^T\phi,h[c])\rangle
-\]
-\[
-\langle R_1[c]\delta c,\phi \rangle = \langle \delta c, DF[c]^*((F[c]N[c]^{-1}F[c]^T-I)\phi,h[c])
-\]
-That is,
-\[
-R_1[c]^*\phi = DF[c]^*((F[c]N[c]^{-1}F[c]^T-I)\phi,h[c])
-\]
-\[
-\langle R_2[c]\delta c,\phi \rangle = \langle F[c]N[c]^{-1}(DF[c]\delta c)^Tr[c],\phi \rangle
-\]
-\[
-\langle r[c],DF[c](\delta c,N[c]^{-1}F[c]^T\phi)\rangle = \langle \delta c, DF[c]^*(r[c], N[c]^{-1}F[c]^T\phi) \rangle
-\]
-That is,
-\[
-R_2[c]^*\phi = DF[c]^*(r[c], N[c]^{-1}F[c]^T\phi)
-\]
-
+\input{append1.tex}
+%\input{append2.tex}
+%\input{append3.tex}
+%\input{talkfigs.tex}
 \end{document}
-\section{Figures for talk}
-\plot{pulse00}{width=\textwidth}{Bandpass filter pulse}
-\plot{fwitrbulkmodinvlh0}{width=\textwidth}{Inversion for bulk modulus of data displayed in Figure \ref{fig:ptpwindl0}.
-Initial model: constant bulk modulus = 4.0 GPa.
-Algorithm: 20 iterations of trust region Krylov-Gauss-Newton algorithm with preconditioning by box smoother of radius 5 points, repeated 10 times.
-Each KGN step computed by up to 20 Conjugate Gradient iterations.}
-\plot{fwitrdatapreslh0}{width=\textwidth}{Data residual produced by 20 itersions of KGN FWI for bulk modulus, starting at constant bulk modulus. Residual RMS approximately 26\% of data RMS.}
-\plot{cgwhestsourceplh0}{width=\textwidth}{Extended inversion for source on 1500 m $\le x \le$ 5500 m, $z$=3000 m, data from Figure \ref{fig:ptpwindl0}, homogeneous bulk modulus = 4.0 GPA.}
-\plot{cgwhresdataplh0}{width=\textwidth}{Residual data from extended source inversion shown in Figure \ref{fig:cgwhestsourceplh0}. RMS is 3\% of data RMS.}
-\plot{bmh0s}{width=\textwidth}{Iteration 1}
-\plot{cgw1bulkupdlh0}{width=\textwidth}{Iteration 1 SSE bulkmod estimate}
-\plot{cgw2bulkupdlh0}{width=\textwidth}{Iteration 2 SSE bulkmod estimate}
-\plot{cgw3bulkupdlh0}{width=\textwidth}{Iteration 3 SSE bulkmod estimate}
-\plot{cgw4bulkupdlh0}{width=\textwidth}{Iteration 4 SSE bulkmod estimate}
-\plot{cgw5bulkupdlh0}{width=\textwidth}{Iteration 5 SSE bulkmod estimate}
-\plot{cgw6bulkupdlh0}{width=\textwidth}{Iteration 6 SSE bulkmod estimate}
-\plot{cgw7bulkupdlh0}{width=\textwidth}{Iteration 7 SSE bulkmod estimate}
-\plot{cgw8bulkupdlh0}{width=\textwidth}{Iteration 8 SSE bulkmod estimate}
-\plot{cgw9bulkupdlh0}{width=\textwidth}{Iteration 9 SSE bulkmod estimate}
-\plot{cgw10bulkupdlh0}{width=\textwidth}{Iteration 10 SSE bulkmod estimate}
-\plot{cgw11bulkupdlh0}{width=\textwidth}{Iteration 11 SSE bulkmod estimate}
-\plot{cgw12bulkupdlh0}{width=\textwidth}{Iteration 12 SSE bulkmod estimate}
-\plot{cgw13bulkupdlh0}{width=\textwidth}{Iteration 13 SSE bulkmod estimate}
-\plot{cgw14bulkupdlh0}{width=\textwidth}{Iteration 14 SSE bulkmod estimate}
-\plot{cgw15bulkupdlh0}{width=\textwidth}{Iteration 15 SSE bulkmod estimate}
-\plot{cgw22bulkupdlh0}{width=\textwidth}{Iteration 22 SSE bulkmod estimate}
-\plot{cgw30bulkupdlh0}{width=\textwidth}{Iteration 30 SSE bulkmod estimate}
-\plot{fwitrsse1bulkmodinvlh0}{width=\textwidth}{Inversion for bulk modulus of data displayed in Figure \ref{fig:ptpwindl0}.
-Initial model: result of 15 steepest descent steps of SSE, Figure \ref{fig:cgw15bulkupdlh0}.
-Algorithm: 20 iterations of trust region Krylov-Gauss-Newton algorithm with preconditioning by box smoother of radius 5 points, repeated 10 times.
-Each KGN step computed by up to 20 Conjugate Gradient iterations.}
-\plot{fwitrsse1datapreslh0}{width=\textwidth}{Data residual produced by 20 itersions of KGN FWI for bulk modulus, starting at output of 15 steepest descent steps of SSE, Figure \ref{fig:cgw15bulkupdlh0}.
-Residual RMS approximately 1.2\% of data RMS.}
-\plot{cgw0estsourceplh0}{width=\textwidth}{Iteration 1 SSE source estimate alpha=4.36 $\times 10^{-6}$}
-\plot{cgw1estsourceplh0}{width=\textwidth}{Iteration 2 SSE source estimate alpha=5.93 $\times 10^{-6}$}
-\plot{cgw2estsourceplh0}{width=\textwidth}{Iteration 3 SSE source estimate alpha=6.82 $\times 10^{-6}$}
-\plot{cgw3estsourceplh0}{width=\textwidth}{Iteration 4 SSE source estimat alpha=7.93 $\times 10^{-6}$e}
-\plot{cgw4estsourceplh0}{width=\textwidth}{Iteration 5 SSE source estimate alpha=9.29 $\times 10^{-6}$}
-\plot{cgw5estsourceplh0}{width=\textwidth}{Iteration 6 SSE source estimate alpha=1.1 $\times 10^{-5}$}
-\plot{cgw6estsourceplh0}{width=\textwidth}{Iteration 7 SSE source estimate alpha=1.32 $\times 10^{-5}$}
-\plot{cgw7estsourceplh0}{width=\textwidth}{Iteration 8 SSE source estimate alpha=1.32 $\times 10^{-5}$}
-\plot{cgw8estsourceplh0}{width=\textwidth}{Iteration 9 SSE source estimate alpha=1.72 $\times 10^{-5}$}
-\plot{cgw9estsourceplh0}{width=\textwidth}{Iteration 10 SSE source estimate alpha=2.2 $\times 10^{-5}$}
-\plot{cgw10estsourceplh0}{width=\textwidth}{Iteration 11 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw11estsourceplh0}{width=\textwidth}{Iteration 12 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw12estsourceplh0}{width=\textwidth}{Iteration 13 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw13estsourceplh0}{width=\textwidth}{Iteration 14 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw14estsourceplh0}{width=\textwidth}{Iteration 15 SSE source estimate alpha=2.79 $\times 10^{-5}$}
-\plot{cgw21estsourceplh0}{width=\textwidth}{Iteration 22 SSE source estimate alpha=5.73 $\times 10^{-5}$}
-\plot{cgw29estsourceplh0}{width=\textwidth}{Iteration 30 SSE source estimate alpha=9.1 $\times 10^{-5}$}
-\plot{cgw0resdataplh0}{width=\textwidth}{Iteration 1 SSE data residual}
-\plot{cgw1resdataplh0}{width=\textwidth}{Iteration 2 SSE data residual}
-\plot{cgw2resdataplh0}{width=\textwidth}{Iteration 3 SSE data residual}
-\plot{cgw3resdataplh0}{width=\textwidth}{Iteration 4 SSE data residual}
-\plot{cgw4resdataplh0}{width=\textwidth}{Iteration 5 SSE data residual}
-\plot{cgw5resdataplh0}{width=\textwidth}{Iteration 6 SSE data residual}
-\plot{cgw6resdataplh0}{width=\textwidth}{Iteration 7 SSE data residual}
-\plot{cgw7resdataplh0}{width=\textwidth}{Iteration 8 SSE data residual}
-\plot{cgw8resdataplh0}{width=\textwidth}{Iteration 9 SSE data residual}
-\plot{cgw9resdataplh0}{width=\textwidth}{Iteration 10 SSE data residual}
-\plot{cgw10resdataplh0}{width=\textwidth}{Iteration 11 SSE data residual}
-\plot{cgw11resdataplh0}{width=\textwidth}{Iteration 12 SSE data residual}
-\plot{cgw12resdataplh0}{width=\textwidth}{Iteration 13 SSE data residual}
-\plot{cgw13resdataplh0}{width=\textwidth}{Iteration 14 SSE data residual}
-\plot{cgw14resdataplh0}{width=\textwidth}{Iteration 15 SSE data residual}

### 2019-12-12 repro/papers/sserefl/paper.tex
many edits

@@
 
-In the following sections, I will first give a more precise statement
-of the main result. Then I will review plane-wave modeling, and construct
+The first of the following sections sets out the framework of this
+study. It does {\em not} include mathematically complete definitions
+or fully qualified statements of results - these may be found in their
+entirety in later sections. Instead, it is intended to introduce the
+concepts that figure in the main results, which are stated succinctly
+in the third section. The  I will review plane-wave modeling, and construct
 an asymptotic approximation to single-reflector plane wave data. In
@@
 
-\section{Main Result}
+\section{Preliminaries}
 
-We adopt the usual pressure-velocity acoustics system as the
-underlying physics of waves, see beginning of next section. The
-surface source extension allows the source to extend over the surface,
-in the form $w(x,t)\delta(z)$ for a nominal source position of
-$(0,0)$. After Radon transform, this source takes the form
+This paper will use linear acoustics in velocity-pressure form as its
+model for wave motion. In this model, the space-time fields $p$ is excess
+pressure, $\bv$ is particle velocity about equilibrium, $f$
+represents an energy source as a defect in the acoustic constitutive
+law. The bulk modulus $\kappa$ and density $\rho$ are positive functions of
+spatial position. The acoustic wave system couples these quantities:
+\begin{eqnarray}
+\label{eqn:awe}
+\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
+                                                                f \nonumber\\
+\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
+p,{\bf v}&=&0, t \ll 0
+\end{eqnarray}
+In terms of the solution of the system \ref{eqn:awe}, the modeling
+operator is defined by
+\begin{equation}
+\label{eqn:fwd}
+S[\kappa]f = p|_{z=z_r}
+\end{equation}
+Two special types of right-hand side are important in the discussion
+to follow: isotropic point sources, of the form
+\begin{equation}
+  \label{eqn:ipr}
+  f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s).  
+\end{equation}
+and surface sources, of the form
+\begin{equation}
+  \label{eqn:ss}
+  f(\bx,t;\bx_s) = u(x,y,t;x_s,y_s)\delta(z-z_s). 
+\end{equation}
+Point sources are a special case of surface sources. Introduce appropriately
+parametrized versions of the modeling operator for these two cases:
+\begin{eqnarray}
+ f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s) & \Rightarrow & S[\kappa]f = S_p
+  [\kappa]w\\ \nonumber
+  \label{eqn:specsrc}
+f(\bx,t;\bx_s) = u(x,y,t;x_s,y_s)\delta(z-z_s) & \Rightarrow & S[\kappa]f =
+  S_s[\kappa]u
+\end{eqnarray}
+
+The standing assumption in this work is that observed data to be fit
+is to good approximation the result of a point source. That is, there
+are bulk modulus $\kappa_*(\bx)$ and source wavelet or pulse $w_*(t)$
+so that
+\begin{equation}
+  \label{eqn:datadef}
+  d = S_p[\kappa_*]w_*
+\end{equation}
+In equation \ref{eqn:datadef}, $d$ is a function of receiver position
+$\bx_r$, source position $\bx_s$, and time $t$.
+
+Full waveform inversion (FWI), as construed in this paper, means:
+\begin{equation}
+  \label{eqn:deffwi}
+  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }w \mbox{ so that }d
+  \approx S_p[\kappa]w.
+\end{equation}
+In the following paragraphs, I will discuss four approaches to FWI:
+
+\subsection{``Straight'' FWI}
+
+The standard approach to FWI is to formulate the task \ref{eqn:deffwi}
+as a nonlinear least squares problem:
+\begin{equation}
+  \label{eqn:defols}
+  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }w \mbox{ to
+    minimize }
+  \| d -  S_p[\kappa]w\|^2.
+\end{equation}
+Here the symbol $\|\cdot\|^2$ signifies {\em mean-square}, that is,
+sum the squares of every value of the field between the vertical bars.
+
+This concept was introduced in the 1980's
+(\cite{BamChavLai:79,Tara:84a,KolbColLai:86,Crasetal:90},
+and many other papers since then). Usually some form of regularization
+is applied, to compensate for poorly determined aspects of $\kappa$
+and/or $w$, as is explained in Tarantola's influential book
+\cite[]{Tarantola:05}. The approach was of only academic interest
+until roughly fifteen years ago, when 3D modeling and related computations
+first became feasible at industrial scale, and evidence began to emerge 
+that FWI offered potentially decisive improvements in imaging accuracy
+over other inversion (imaging) methods, especially for construction of
+large-scale earth model features
+\cite[]{Plessix:10,Vigh:10}. In the last ten years, FWI has become
+both a production service offered by the geophysical industry, and a
+leading topic of both industry-oriented and academic research.
+
+As was understood very early in the evolution of FWI, the basic
+least-squares formulation \ref{eqn:defols} suffers from a major
+defect: the parameter spaces tend to be so large that only iterative
+optimization (that is, a variant
+of Newton's method) may feasibly be applied, due to computational cost
+- this remains true today and for the foreseeable future - but these
+method tend not to produce usable results unless the initial iterate
+is in some sense ``good'' (predicts travel times to within a
+half-wavelength) \cite[]{GauTarVir:86,Plessix:10}.``Good'' initial models are not always
+available. Accordingly, a number of alternative approaches have been
+proposed, many of them variations on the least squares theme, amongst
+which are the three approaches described below.
+
+\subsection{Source Extensions}
+
+Careful analysis of the defect mentioned in the last paragraph shows
+that the failure of iterative optimization applied to \ref{eqn:defols}
+to converge (or at least to converge usefully quickly) appears to
+arise from failure to fit the data: some parts of the oscillatory time
+series $d(\cdot,t)$ overlap oscillations in $S_p[\kappa]w$, while
+others do not. It seems reasonable to guess that this ``cycle skipping'' phenomenon 
+could be avoided if the data could be fit throughout the inversion
+process. That requirement in turn implies a modeling operator with
+more degrees of freedom, that is, a bigger domain, than the basic
+modeling operator $S_p$, which should produce the same results as
+$S_p$ when applied to pairs of bulk moduli and isotropic point sources \ref{eqn:ipr}.
+
+There is
+well-established mathematical terminology for this relationship:
+operator A is an {\em extension} of operator B if (i) the domain of $A$
+contains the domain of $B$, and whenever $v$ is in the domain of $B$,
+$Av = Bv$. Two extensions of $S_p$ already appear
+above: the isotropic point radiator \ref{eqn:ipr} is a special case of
+both the surface source \ref{eqn:ss}, which is in turn a special case
+of the generic right-hand side in the system \ref{eqn:awe}. So the
+domain of $S_s$ contains the domain of $S_p$, as does the domain of
+$S$. From the relations \ref{eqn:specsrc}, $S$ and $S_s$ produce the
+same result as $S_p$ when applied to sources in their common domain,
+so these operators are both extensions of $S_p$.
+
+Since these extensions of $S_p$ involve additional degrees of freedom
+for the energy source in the system \ref{eqn:awe}, they have come to
+be called {\em source extensions}. These are not the only source
+extensions - see \cite{HuangNammourSymesDollizal:SEG19} for an
+overview of source extensions and their properties. Extensions may
+also be constructed by adding degrees of freedom to the coefficients
+in the wave system (bulk modulus, in the present case), 
+rather than the right hand side - see \cite{geoprosp:2008} for a
+description of some of these {\em medium extensions}.
+
+The purpose of any extension is to add options for data fitting,
+thereby reducing the chances of cycle-skipping. As will be seen below,
+both extension $S$ and $S_s$ permit arbitrarily accurate fitting of
+essentially any data with essentially any bulk modulus field $\kappa$,
+with proper choice of $f$ or $w$ respectively. However, the standing
+assumption on the data $d$ is that it derives from a point source, per
+\ref{eqn:datadef}. So the added degrees of freedom represented by the
+extensions $S$ and $S_s$ must be suppressed somehow. As explained by
+\cite{geoprosp:2008}, a convenient mechanism for suppression of the
+extension degrees of freedom is a penalty term added to the objective
+of the extended
+version of the least squares problem \ref{eqn:defols}. Choice of penalty
+term, along with choice of extension, characterizes each of the extended
+inversion methods detailed in the next few paragraphs.
+
+\subsubsection{Wavefield Reconstruction Inversion}
+This concept was introduced by \cite{LeeuwenHerrmannWRI:13}, and further
+developed by \cite{LeeuwenHerrmann:16,WangYingst:SEG16} and other
+authors. It is based on the extension $S$ (equation \ref{eqn:fwd}), with full
+space-time volume source $f$, and presumes that the correct point
+source wavelet $w_*$ is known. 
+\begin{equation}
+  \label{eqn:defwri}
+  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }f \mbox{ to
+    minimize }
+  \| d -  S[\kappa]f\|^2+\lambda^2\|f-w_*\delta(\cdot-\bx_s)\|^2.
+\end{equation}
+Evidently the difference $f-w_*\delta(\cdot-\bx_s)$ should be such
+that the scaled sum-of-squares makes sense - this is always the case
+for discretizations, but for the continuum limit entails an actual
+assumption.
+
+\cite{LeeuwenHerrmannWRI:13} posed this problem slightly
+differently, in terms of the second order wave equation for the
+pressure wavefield $p$. From this viewpoint, the 
+second terms becomes a scaled mean square of the residual in the
+wave equation. The formulation presented here is equivalent, and was
+introduced by \cite{WangYingst:SEG16}. \cite{LeeuwenHerrmann:16} used
+the variable projection method \cite[]{GolubPereyra:03}, eliminating the source $f$ in the
+inner step and updating the bulk modulus (or an equivalent quantity)
+in the outer step. In any penalty method, control of the penalty
+parameter has a large influence on the speed of
+convergence. \cite{Aghamiry:19} use an augmented Lagrangian algorithm
+to minimize the influence of the penalty weight choice.
+
+The Achilles' heel of Wavefield Reconstruction Inversion (WRI), in any of
+the formulations described so far, is the necessity of storing a full
+space time field, namely $f$ (or $p$). Especially for time-domain
+formulation and high frequency waves, this storage requirement can be
+formidable. In fact, ``most'' source fields $f$ are non-radiating,
+that is, $S[\kappa]f=0$. The orthocomplement of this subspace of
+non-radiating sources is the range of the adjoint operator
+$S[\kappa]^T$. Therefore the sources may be re-parametrized as images of
+$S[\kappa]^T$. Write $f = g + w_*\delta(\cdot-\bx_s)$: then the second term in
+the objective displayed in \ref{eqn:defwri} is simply $\|g\|^2$.
+Decompose $g = S[\kappa]^Te + n$, in
+which $e$ is the same type of object as $d$ and $S[\kappa]n=0$ (that
+is, $n$ is a non-radiating
+source), and note that the decomposition is orthogonal. Then the WRI problem \ref{eqn:defwri} may be reformulated as
+\begin{equation}
+  \label{eqn:defwrireform}
+  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }e \mbox{ to
+    minimize }
+  \| d -  S[\kappa]w_*\delta(\cdot-\bx_s) - S[\kappa]S[\kappa]^Te\|^2+\lambda^2\|S[\kappa]^Te\|^2.
+\end{equation}
+Note that $n$ has disappeared: {\em a priori} it does not contribute to the first
+term, and $n \ne 0$ makes the second term larger. So minimization
+necessarily entails the choice $n=0$.
+
+This reformulation does {\em not} require storage of $d+1$-dimensional
+fields, as does the original WRI formulation \ref{eqn:defwri}. The
+field $e$ has taken the place of the space-time source $f$, and its
+size is precisely the same as that of the data field $d$. Therefore
+the extra memory cost over modeling is the same as another copy of
+data. A close examination reveals that in computing the composition
+$SS^T$ via time-stepping methods, the same simultaneity issue arises
+as in all applications of the adjoint state method. However good
+solutions are known \cite[]{Griewank:book}, and since precisely the
+same issue arises in time-domain FWI, the cost is the same. Therefore,
+it is this reformulation of WRI that will be analyzed below.
+
+\subsubsection{Space-Time Extension}
+\cite{HuangSymes:Geo18a} introduced a variant of WRI, also using the space-time extension, but not
+requiring knowledge of the target point source wavelet $w_*$: in fact,
+an estimate of this wavelet is a by-product of the algorithm. This
+approach relies on an {\em annihilator}, a linear operator whose null
+space is precisely the desired class of source models, in this case
+isotropic point radiators. Such an operator can be used to generate a
+quadratic form whose zero set is also the desired class, hence can be
+used as a penalty in a nonlinear least squares objective, designed to
+drive the source towards to the desired type. See \cite{geoprosp:2008}
+for a discussion of annihilators in extended inversion.
+
+The annihilator proposed by \cite{HuangSymes:Geo18a} is multiplication
+by $|\bx-\bx_s|$, that is,
+\begin{equation}
+  \label{eqn:defann}
+  Af(\bx,t;\bx_s) = |\bx-\bx_s|f(\bx,t;\bx_s).
+\end{equation}
+The space-time extension approach consists of solving the optimization
+problem:
+\begin{equation}
+  \label{eqn:defste}
+  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }f \mbox{ to
+    minimize }
+  \| d -  S[\kappa]f\|^2+\lambda^2\|Af\|^2.
+\end{equation}
+This problem can be reformulated in the same way as the WRI problem. Write
+$f = w\delta(\cdot-\bx_s) + S[\kappa]^Te + n$, where $n$ is
+non-radiating as before: then the problem \ref{eqn:defste} is
+equivalent to
+\[
+   \mbox{Given } d, \mbox{ find }\kappa, w, e \mbox{ and }n \mbox{ to
+     minimize }
+ \]
+\begin{equation}
+  \label{eqn:defstereform}
+  \| d -  S[\kappa]w_*\delta(\cdot-\bx_s) - S[\kappa]S[\kappa]^Te\|^2+\lambda^2\|A(S[\kappa]^Te+n)\|^2.
+\end{equation}
+Note that the non-radiating component $n$ must remain in the second
+term of the objective, because there is no reason for $n$ to be
+$A^TA$-orthogonal to the range of $S[\kappa]^T$. Also, the isotropic
+point radiator component $w\delta(\cdot-\bx_s)$ must also remain as
+part of the unknowns - simply because it is unknown! It does not
+appear in the second term because $A$ annihilates it.
+
+\subsubsection{Surface Source Extension}
+Beside the reformulation of WRI described above, it is also possible
+to avoid the storage problem of WRI by limiting the support of the
+extended source $f$ to lower dimensional sets. Of several
+possibilities discussed in the literature, I will discuss the 
+  surface source extension approach introduced by
+\cite{HuangNammourSymesDollizal:SEG19}. The extension in question is
+defined in equation \ref{eqn:defss}, and denoted $S_s$. The same
+annihilator \ref{eqn:defann} is used in this approach, this time
+interpreted as acting in the plane $z=z_s$, that is,
+\[
+  Au(x,y,t;x_s,y_s) = \sqrt{(x-x_s)^2+(y-y_s)^2}u(x,y,t;x_s,y_s).
+\]
+The problem formulation is
+\begin{equation}
+  \label{eqn:defsse}
+  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }u \mbox{ to
+    minimize }
+  \| d -  S_s[\kappa]u\|^2+\lambda^2\|Au\|^2.
+\end{equation}
+Evidently the extended source has as many parameters as the data, so
+the memory complexity of this problem is comparable to that of FWI or
+reformulated WRI. The computational complexity compares favorably to
+that of reformulated WRI.
+
+\subsection{Layered Media, Plane Waves}
+As explained in the introduction, this study explores the properties
+of the various FWI variants explained above for the special case of
+layered media: $\kappa = \kappa(z)$,
+$\rho=\rho(z)$. Then the modeled shots differ from
+each other by a horizontal translation, so that there is in effect
+only one shot, for which I take the location to be
+$\bx_s=(0,0,0)$ and suppress $\bx_s$ from the notation.
+
+If $\log \kappa$ and $\log \rho$ are globally bounded and $f$ has
+compact support (and some other properties that will be made precise
+later), there is $\xi_{\rm max} > 0$ so that for
+$\|\xi\|<\xi_{\rm max}$, the integrals defining slant stack fields
+$\tilde{p},\tilde{\bv}$, namely
+\begin{eqnarray}
+\label{eqn:slant}
+\tp(\xi,z,t)& = &\int dx\, dy\, p(x,y,z,t+\xi x),\nonumber\\
+\tv(\xi,z,t)&=&\int dx\, dy\, v_z(x,y,z,t+\xi x),\nonumber\\
+\tf(\xi,z,t)&=&\int dx\,dy\, f(x,y,z,t+\xi x),
+\end{eqnarray}
+are absolutely convergent, and $\tp, \tv$ satisfy the plane wave
+version of the acoustic system \ref{eqn:awe}:
+\begin{eqnarray}
+\label{eqn:awepw}
+(1-c^2\xi^2)\frac{\partial \tp}{\partial t} + \kappa\frac{\partial
+  \tv}{\partial z} &=& \tf \nonumber\\
+\rho \frac{\partial \tv}{\partial t} + \frac{\partial \tp}{\partial
+  z}&=&0\nonumber\\
+\tp,\tv&=&0, t \ll 0.
+\end{eqnarray}
+Here $c=\sqrt{\kappa/\rho}$ is the sound velocity. I use $\xi$ instead
+of the more usual $p$ for plane wave slowness since $p$ already has
+been assigned to mean pressure. Note that for a isotropic point
+source, respectively a surface source,
+\begin{eqnarray}
+  f(\bx,t;\bx_s)=w(t)\delta(\bx) & \Rightarrow &
+                                               \tf(\xi,z,t)=w(t)\delta(z)\\ \nonumber 
+  f(\bx,t;\bx_s) = u(x,y,t)\delta(z) & \Rightarrow &
+                                                     \tf(\xi,z,t) =
+                                                     \tu(\xi,t)\delta(z)\\
+  \nonumber
+  \tu(\xi,t) &=& \int dx\,dy\, u(x,y,t+\xi x)
+\end{eqnarray}
+The plane wave domain modeling
+operator for space-time extended sources samples $\tp$ on $z=z_r$:
+\begin{equation}
+\label{eqn:modpw}
+\tS[\kappa]\tf = \tp|_{z=z_r}
+\end{equation}
+Similarly, the plane wave domain versions of $S_p$ and $S_s$ are
+\begin{eqnarray}
+ \tf(\xi,z,t) = w(t)\delta(z) & \Rightarrow & \tS[\kappa]\tf = \tS_p
+  [\kappa]w\\ \nonumber
+  \label{eqn:specsrc}
+\tf(\xi,z,t) = \tu(\xi,t)\delta(z-z_s) & \Rightarrow & \tS[\kappa]\tf =
+  \tS_s[\kappa]\tu
+\end{eqnarray}
+
+For homogeneous $\kappa=\kappa_0, \rho=\rho_0, c=c_0$, a computation
+recapitulated in the Appendix C, relying on computations for the 1D
+acoustics system detailed in Appendices A and B, yields
+\begin{equation}
+\label{eqn:modhom}
+\tS_s[\kappa_0]\tu(\xi,t) = \frac{1}{2(1-c_0^2\xi^2)}\tu
+\left(\xi,t-z_r\sqrt{\frac{1}{c_0^2}-\xi^2}\right).
+\end{equation}
+That is, at a homogeneous model, the plane wave modeling operator is
+simply scaling by a $\xi$-dependent factor.
+
+\section{Main Results}
+
+The space dimension is not terribly important in the results to
+follow. Generally I will assume it to take the physical value $d=3$,
+though the numerical examples will use $d=2$ for reasons of
+computational cost. For spatial position, write $\bx = (x,y,z)$ (or $\bx = (x,z)$ for
+2D).
+
+\cite{BlazekStolkSymes:13} give conditions under which the system
+\ref{eqn:awe} has a strong solution, that is, in which all of the
+terms in equations above are interpretable as functions of space-time. Assume that
+
+\indent Assumption A: $\log \kappa, \log \rho$ are bounded and
+measureable. \\
+
+\indent Assumption B: $f$ is square-integrable, has a square-integrable time
+  derivative, and \\has compact support.\\
+
+Then Theorem 1 in \cite{BlazekStolkSymes:13} implies that there exists
+a unique causal (vanishing for $t \ll 0$) strong solution
+$(p,\bv)$ of the system \ref{eqn:awe}, that is, $p$ and $\bv$ are
+$C^1$ functions of time with square-integrable (in space) values,
+$\nabla p$ and $\nabla \cdot \bv$ are $C^0$ functions of time with
+square-integrable values,  and the equations \ref{eqn:awe} are
+satisfied pointwise a. e.
+
+The central object in this study is the relation between the
+coefficients $\kappa$ and $\rho$, on the one hand, and the fields $p$
+and $\bv$ on the other. The influence of the wave velocity $c$ is the
+most important, so from here on $\rho$ will be fixed and constant:
+$\rho(\bx) = \rho_0 \in \bR^+$. Thus view $p$ and $\bv$ as functions of
+$\kappa$ and $f$.
+
+In seismic and ultrasonic experiments, the acoustic
+fields are recorded at or near a hypersurface (the surface of the
+earth, or the skin, or...). Theorem 2 in \cite[]{BlazekStolkSymes:13}
+assures that the trace of $p$ on a cylindrical hypersurface $\Gamma
+\times [0,T]$ is well-defined, as is the trace of the normal component
+of $\bv$. These traces will serve as idealized data in the discussion
+to follow, ignoring issues of sampling and other details of actual
+measurement processes.
+
+This discussion further idealizes acquisition by setting $\Gamma =
+\{\bx: z=z_r\}$. This is not a bad approximation for some field
+settings, and is convenient for the eventual use of the Radon
+transform.
+
+Active (artificial) sources (right-hand sides in the wave equations)
+are frequently localized in space, relative to other active length scales. The
+simplest such localized source is the {\em isotropic radiator},
+\begin{equation}
+  \label{eqn:ipr}
+  f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s).  
+\end{equation} 
+in which $\bx_s$ is the location of the radiator. The radiator is a
+special case of the {\em surface source}, in which the support of the
+right-hand side in \ref{eqn:awe} is confined to a hypersurface. For
+present purposes, it suffices to consider sources localized on the
+hyperplane $z=z_s$:
+\begin{equation}
+  \label{eqn:ss}
+  f(\bx,t;\bx_s) = w(x,y,t)\delta(z-z_s). 
+\end{equation}
+Neither distribution defined by \ref{eqn:ipr}, \ref{eqn:ss} is 
+square-integrable, so the theory in \cite[]{BlazekStolkSymes:13} does
+not apply directly to the system \ref{eqn:awe} with $f$ defined by
+equation \ref{eqn:ipr}. However a simple trick produces a unique
+(necessarily weak) solution in both cases, provided an additional
+constraint is placed on $\kappa$:
+
+\indent Assumption C: There exists $\epsilon >0$ so that $\kappa =
+\kappa_0 \in \bR^+$ in $\{\bx: |z-z_s|<\epsilon\}$.
+
+\noindent That is, $\kappa$ is constant near the plane $z=z_s$. See
+Appendix \appSingular for a proof of this statement.
+
+Define the coefficient-to-data map (``forward map'', ``modeling
+operator'') $S$ by
+\begin{equation}
+\label{eqn:fwd}
+S[\kappa]f = p|_{z=z_r}
+\end{equation}
+The notation reminds the reader that $S$ is linear in $f$ but
+nonlinear in $\kappa$.
+
+Theorem 2 in \cite[]{BlazekStolkSymes:13} assures that $S[\kappa]$ is
+well-defined as a map from right-hand sides $f$ having
+square-integrable time derivatives to square-integrable traces. With
+more regularity in time, the map is also differentiable in its
+dependence on $\kappa$. In Appendix \ref{appSingular}, this result is
+extended to the singular sources \ref{eqn:ipr}, \ref{eqn:ss}.
+
+In fact, stronger results hold under Assumption C. The special cases
+considered below have stronger continuity properties.
+
+After Radon transform, this source takes the form
 $\tw(\xi,t)\delta(z)$, where $\xi$ is the plane wave slowness.
@@
 \section{Plane Wave Modeling}
-Start with 2D acoustics in velocity-pressure form: $p$ is excess
-pressure, $\bv=(v_x,v_z)$ is particle velocity about equilibrium, $f$
-is a (possibly) extended source model, all linked by the acoustic wave system
-\begin{eqnarray}
-\label{eqn:awe}
-\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
-                                                                f(\bx,t;\bx_s) \nonumber\\
-\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
-p,{\bf v}&=&0, t \ll 0
-\end{eqnarray}
-
-As in other work on this topic, I will assume $\rho$ to be fixed, and
-for sake of simplicity constant, and focus the discussion on
-$\kappa$. Thus the modeled pressure traces at receiver locations
-depend on $\kappa$ and the source function $f$. For simplicity again, assume that
-receivers are located at every point of the hyperplane $z=z_r$. Then
-the modeling operator is defined as
-\begin{equation}
-\label{eqn:fwd}
-S[\kappa]f = p|_{z=z_r}
-\end{equation}
-
-Assume that sources all lie on the surface $z=z_s$, so that surface
-extended sources take the form $f(\bx,t;\bx_s) =
-w(x,t)\delta(z-z_s)$. By abuse of notation, regard $S[\kappa]$ as
-acting on $w$.
 
@@
 
-\append{1D Wave Motion}
+\append{Singular RIght-hand Sides and Weak Solutions}
+Theorem 1 in \cite[]{BlazekStolkSymes:13} provides a strong solution of the system \ref{eqn:awe}: $p$, respectively $\bv$, is continuous
+in time with values in the Sobolev space $H^1(\bR^d)$, respectively
+$H^1_{\rm div}(\bR^d)$ and
+also continuously differentiable with values in $L^2(\bR^d)$,
+respectively $(L^2(\bR^d))^d$. Therefore all of the terms in
+\ref{eqn:awe} are square-integrable functions.
+
+Theorem 4 in \cite[]{BlazekStolkSymes:13} implies that for each $t$, $p(\cdot,t)$ and $\bv(\cdot,t)$
+have compact support, and that the support expands in $t$ at the rate
+bounded by the sup of the velocity $c=\sqrt{\kappa/\rho}$, as one
+would expect. Therefore, for any fixed max time $T$, it is possible to
+treat the problem for $0\le t \le T$ as a Dirichlet problem in a
+domain $\Omega$ of sufficiently large diameter. Without loss of
+generality, assume that $\mbox{supp }f \subset \Omega \times [0,T]$.
+
+The isotropic point radiator \ref{eqn:ipr} is not
+square-integrable, so the theory in \cite[]{BlazekStolkSymes:13} does
+not apply directly to the system
+\begin{eqnarray}
+\label{eqn:aweptsrc}
+\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
+                                                                w(t)\delta(\bx-\bx_s) \nonumber\\
+\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
+p,{\bf v}&=&0, t \ll 0
+\end{eqnarray}
+However there is a simple trick that makes this
+theory applicable, provided one assumes that $\kappa$ is constant
+($=\kappa_0 \in \bR^+$) near
+$\bx=\bx_s$ (Assumption C). With this constraint, the problem
+\ref{eqn:aweptsrc} may be shown to have a unique causal weak solution.,
+
+Note that the existence of weak solutions for problems of this type
+has been 
+well-known for decades, with the additional assumption that the coefficients are
+smooth, see for example \cite{Lax:PDENotes}. Using the progressing
+wave expansion of the fundamental solution for smooth coefficient
+problems (for example, \cite{Friedlander:75}) in place of the analytic
+expression for the constant-coefficient point radiator solution given
+below, it is possible to obtain a similar result under the weaker
+\begin{itemize}
+\item[Assumption C']: $\kappa$ is smooth near $\bx_s$.
+\end{itemize}
+
+Denote by $(p_0,\bv_0)$ the (weak) solution of the
+system \ref{eqn:awe} with $\kappa$ replaced by $\kappa_0$. For $d=3$,
+\[
+ p_0(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-\frac{ |\bx-\bx_s|}{c_0}\right)
+\]
+in which $c_0=\sqrt{\kappa_0/\rho_0}$. There is a similar expression
+for $\bv_0$, and for other space dimensions $d$. Because $w$ is smooth, both $p$ and $\bv$ are smooth
+except at $\bx=\bx_s$. Pick $\phi \in C_0^{\infty}(\bR^d)$ so that
+$\phi = 1$ in a neighborhood of $\bx_s$ and $\kappa=\kappa_0$ on
+$\mbox{supp }\phi$. Then
+\begin{eqnarray}
+\label{eqn:awecut}
+\frac{\partial (\phi p_0)}{\partial t} + \kappa \nabla \cdot (\phi {\bf v}_0) &=&
+                           w(t)\delta(\bx-\bx_s) + f_0  \nonumber\\
+\rho \frac{\partial (\phi {\bf v}_0)}{\partial t} + \nabla (\phi p_0) = {\bf g}_0 \nonumber\\
+\end{eqnarray}
+in which $f_0$ and ${\bf g}_0$ are smooth of compact support. Denote by
+$(p_1,\bv_1)$ the strong solution of
+\begin{eqnarray}
+\label{eqn:awesupp}
+\frac{\partial p_1}{\partial t} + \kappa \nabla \cdot {\bf v}_1 &=&
+                                                                -f_0 \nonumber\\
+\rho \frac{\partial {\bf v}_1}{\partial t} + \nabla p_1 = -{\bf g}_0 \nonumber\\
+p_1,{\bf v}_1&=&0, t \ll 0
+\end{eqnarray}
+as provided by Theorem 1 of \cite{BlazekStolkSymes:13}. Then
+\begin{eqnarray*}
+  p & = & \phi p_0 + p_1\\
+  \bv & = & \phi \bv_0 + \bv_1
+\end{eqnarray*}
+defines the unique weak solution of the system \ref{eqn:aweptsrc}.
+
+\cite{SantosaSymes:00} use an analogous construction for the
+second-order wave equation to give $L^2$ bounds on the point radiator
+solution.
+
+As mentioned in the Preliminaries section, a similar argument
+establishes the existence of weak solutions for the surface source problem
+\begin{eqnarray}
+\label{eqn:awess}
+\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
+                                                                w(x,y,t)\delta(z-z_s) \nonumber\\
+\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
+p,{\bf v}&=&0, t \ll 0
+\end{eqnarray}
+under Assumption D, that $\kappa=\kappa_0$ near $z=z_s$. An analytic
+solution for $\kappa \equiv \kappa_0$ follows from that for the point
+radiator problem, by superposition. Then a similar cutoff construction
+and appeal to Theorem 1 of \cite{BlazekStolkSymes:13} proveds a weak
+solution to \ref{eqn:awess}.
+
+This construction also works via recourse to the progressing wave
+expansion for the weaker
+\begin{itemize}
+\item[Assumption D']: $\epsilon > 0$ exists for which $\kappa$ is
+smooth in the collar neigborhood $\{\bx: |z-z_s|<\epsilon\}$.
+\end{itemize}
+
+\append{1D Radiation Problem}
 Begin with the 1D acoustics point source system. 
@@
 Now assume that $\kappa, \rho$ are constant. 
-In $z \ne 0$, the right hand side 
+In $z \ne z_s$, the right hand side 
 vanishes, so the solution must be locally a combination of plane
@@
 \[
-p(z,t)=a\left(t \mp \frac{z}{c}\right), \, v(z,t)=\pm b\left(t \mp
-  \frac{z}{c}\right)
+p(z,t)=a\left(t \mp \frac{z-z_s}{c}\right), \, v(z,t)=\pm b\left(t \mp
+  \frac{z_s}{c}\right)
 \]
@@
 \[
-\kappa [v]_{z=0}\delta(z) =
-2\frac{\kappa}{c}b\delta(z) = 2a\delta(z).
+\kappa [v]_{z=z_s}\delta(z-z_s) =
+2\frac{\kappa}{c}b\delta(z-z_s) = 2a\delta(z-z_s).
 \] 
@@
 \label{eqn:sol1d}
-p(z,t) &=& \frac{1}{2}w\left(t \mp \frac{z}{c}\right) \nonumber \\
-v(z,t) &=& \pm\frac{1}{2\rho c}w\left(t \mp \frac{z}{c}\right)
+p(z,t) &=& \frac{1}{2}w\left(t \mp \frac{z-z_s}{c}\right) \nonumber \\
+v(z,t) &=& \pm\frac{1}{2\rho c}w\left(t \mp \frac{z-z_s}{c}\right)
            \nonumber \\
-\pm z &>& 0.
+\pm z-z_s &>& 0.
 \end{eqnarray}
@@
 If $\kappa, \rho $ are constant, it is possible to develop explicit
-expressions for the solution of this system. Assuming that $\delta
-\rho=0$ and $\delta \kappa$ is nonzero only in $z>0$, the reflected
-(perturbational) wavefield should be upcoming (i.e. a function of
-$z+ct$, away from the wavefront), as the wave scatters
-only once, at the wavefront implicit in the expressions
-\ref{eqn:sol1d}. On the other hand, the field must vanish before the
-first arrival.
+expressions for the solution of this system. Only $\kappa$ will be
+subject to perturbation in any of the problems considered in this
+report, so assume that that $\delta\rho=0$.
 
 The key is to understand the perturbational impulse response. That is,
-set $w(t)=H(t)$, and adopt adopt the anzatz
+set $w(t)=H(t)$. Then the identity \ref{eqn:sol1d} implies that the
+inhomogeneous term in the first equation of the system \ref{eqn:dawe1d}
+is 
+\begin{equation}
+  \label{eqn:pertrhs}
+-\delta \kappa(z) \frac{\partial v}{\partial z}(z,t) = \frac{\delta
+  \kappa(z_s)}{\rho c}H(t)\delta(z-z_s) +
+\frac{\delta\kappa(z)}{2\kappa}\delta\left(t - \frac{|z-z_s|}{c}\right)
+\end{equation}
+Correspondingly, the solution $(\delta p, \delta v)$ is the sum of
+three components:
 \begin{eqnarray}
-\label{eqn:ansatz1d}
-\delta p(z,t) &=& a_0\left(t + \frac{z}{c}\right) \delta\left(t -
-  \frac{z}{c}\right) + a_1\left(t + \frac{z}{c}\right) H\left(t -
-  \frac{z}{c}\right),\nonumber\\
-\delta v(z,t) &=& b_0\left(t + \frac{z}{c}\right) \delta\left(t -
-  \frac{z}{c}\right) + b_1\left(t + \frac{z}{c}\right) H\left(t -
-  \frac{z}{c}\right),
+  \label{eqn:pertrhsterms}
+  \delta p & = & \delta p_s + \delta p_d \nonumber\\
+  \delta v & = & \delta v_s + \delta v_d 
 \end{eqnarray}
-Assume that $w$ is smooth and
-of compact support, with zero mean. Then all derivatives of $w$ are
-necessarily linearly independent, as any dependence would amount to a
-differential equation with constant coefficients, the solution of
-which would be entire analytic hence not of compact support.
+The first component satisfies
+\begin{eqnarray}
+  \label{eqn:dawe1ds}
+  \frac{\partial \delta p_s}{\partial t} +\kappa\frac{\partial
+  \delta v_s}{\partial z}   &=&  \frac{\delta
+  \kappa(z_s)}{\rho c}H(t)\delta(z-z_s)\nonumber\\
+  \rho \frac{\partial \delta v_s}{\partial t} + \frac{\partial \delta p_s}{\partial
+  z}&=&0\nonumber\\
+  \delta p_s,\delta v_s&=&0, t < 0.
+\end{eqnarray}
+The initial conditions are free to be chosen, and the choice above is convenient.
 
-In the following calculations, the arguments of the coefficients
-$a_0,a_1,b_0,$ and $b_1$ are $\left(t + \frac{z}{c}\right)$ in all
-cases, and are suppressed for the sake of brevity, until the final result.
+Note that \ref{eqn:dawe1ds} is precisely a system of the form
+\ref{eqn:awe1d}, so from the formulae \ref{eqn:sol1d} with $w(t)=\frac{\delta
+  \kappa(z_s)}{\rho c}H(t)$
+\begin{eqnarray}
+\label{eqn:dsol1ds}
+\delta p_s(z,t) &=& \frac{\delta
+  \kappa(z_s)}{2\rho c}H\left(t - \frac{|z-z_s|}{c}\right) \nonumber \\
+\delta v_s(z,t) &=& \pm\frac{\delta \kappa(z_s)}{2\rho^2 c^2}H\left(t - \frac{|z-z_s|}{c}\right)
+\end{eqnarray}
 
-Inserting the ansatz \ref{eqn:ansatz1d} in the second equation in \ref{eqn:dawe1d}, obtain
+Define $P_d(z,t;\zeta), V_d(z,t;\zeta)$ to be the solution of the system
+\begin{eqnarray}
+  \label{eqn:dawe1dd}
+  \frac{\partial P_d}{\partial t} +\kappa\frac{\partial
+  V_d}{\partial z}   &=&   \frac{\delta\kappa(\zeta)}{2\kappa}\delta(z-\zeta)\delta\left(t - \frac{|\zeta-z_s|}{c}\right)\nonumber\\
+  \rho \frac{\partial V_d}{\partial t} + \frac{\partial P_d}{\partial
+  z}&=&0\nonumber\\
+  P_d,V_d&=&0, t < 0.
+\end{eqnarray}
+Then
+\begin{eqnarray}
+  \label{eqn:intrep1}
+  \delta p_d(z,t) & = & \int d\zeta\, P_d(z,t;\zeta)\\
+  \delta v_d(z,t) & = & \int d\zeta\, V_d(z,t;\zeta)
+\end{eqnarray}
+
+The system \ref{eqn:dawe1dd} is precisely of the form
+\ref{eqn:awe1d} with $z_s$ replaced by $\zeta$ and
 \[
-0=\rho \frac{\partial \delta v}{\partial t} + \frac{\partial \delta p}{\partial
-  z} = 
+  w(t)= \frac{\delta\kappa(\zeta)}{2\kappa}\delta\left(t - \frac{|\zeta-z_s|}{c}\right)
 \]
-\[
-\left(\rho b_0 - \frac{a_0}{c}\right) \delta'\left(t -
-  \frac{z}{c}\right)
-\]
-\[
-+\left(\rho b_0' + \frac{a_0'}{c}+\rho b_1 - \frac{a_1}{c} \right)\left(t + \frac{z}{c}\right) \delta\left(t -
-  \frac{z}{c}\right)
-\]
-\[
- + \left(\rho b_1' + \frac{a_1'}{c}\right) \left(t + \frac{z}{c}\right) H\left(t -
-  \frac{z}{c}\right)
-\]
-Since $\delta', \delta$ and $H$ are linearly independent, this equation is
-equivalent to the system
-\begin{eqnarray}
-\label{eqn:asnbs}
-\rho b_0 - \frac{a_0}{c} &=& 0\nonumber\\
-\rho b_0' + \frac{a_0'}{c}+\rho b_1 - \frac{a_1}{c} &=&0 \nonumber\\
-\rho b_1' + \frac{a_1'}{c} &=&0.
-\end{eqnarray}
-The first and third equations imply that 
-\begin{eqnarray}
-\label{eqn:handy}
-\rho c b_0&=&a_0,\nonumber\\
-\rho c b_1&=&-a_1,
-\end{eqnarray}
- so the second equation becomes
-\begin{equation}
-\label{eqn:as}
-a_0' = a_1
-\end{equation}
-Now insert the asatz \ref{eqn:ansatz1d} into the first of the
-dynamical equations \ref{eqn:dawe1d}. Note that \ref{eqn:sol1d}
-implies
-\[
-\delta \kappa \frac{\partial v}{\partial z} = - \frac{\delta
-  \kappa}{2\kappa}\delta\left(t - \frac{z}{c}\right)
-\]
-so that
-\[
-0 = \left(\frac{\partial \delta p}{\partial t} +\kappa\frac{\partial
-  \delta v}{\partial z} + \delta \kappa\frac{\partial
-  v}{\partial z}\right)(z,t)=
-\]
-\[
-\left(a_0 - b_0\frac{\kappa}{c}\right)\delta'\left(t - \frac{z}{c}\right) +
-\]
-\[
-\left(a_0' + b_0'\frac{\kappa}{c} + a_1-b_1\frac{\kappa}{c}\right)\delta\left(t -
-  \frac{z}{c}\right)+
-\]
-\[
-\left(a_1' + b_1'\frac{\kappa}{c}\right) H\left(t - \frac{z}{c}\right)
-\]
-\[
-- \frac{\delta \kappa}{2\kappa}\delta\left(t - \frac{z}{c}\right)
-\]
-The first and last equations from \ref{eqn:handy} imply that the first
-and third terms in the sum above vanish, and that the second and fourth
-amount to
-\[
-=\left(2 (a_0'+a_1) -\frac{\delta \kappa}{2\kappa}\right) \delta\left(t -
-  \frac{z}{c}\right)
-\]
-Taking the identity \ref{eqn:as} into account and re-inserting the
-argument of $a_1$, the upshot is
-\[
-4a_1\left(t + \frac{z}{c}\right) \delta\left(t -
-  \frac{z}{c}\right) = \frac{\delta \kappa(z)}{2\kappa} \delta\left(t
-  -  \frac{z}{c}\right)
-\]
-In view of the delta function factor, $z=ct$ in the arguments of $a_1$
-and $\delta \kappa $, so the above equation is equivalent to
-\begin{equation}
-\label{eqn:asol}
-a_1(t)=\frac{\delta \kappa\left(\frac{ct}{2}\right)}{8\kappa}
-\end{equation}
-It remains to determine the integration constant for $a_0$. In view of
-the condition, mentioned above, that $\delta \kappa$ vanish in $z \le
-0$, it follows from \ref{eqn:dawe1d} that $\delta p(0,0)=0$, so
-$a_0(0)=0$. Consequently the ansatz \ref{eqn:ansatz1d} is validated,
-and 
+Thus from \ref{eqn:sol1d}
+\begin{eqnarray*}
+  P_d(z,t;\zeta) &=& \frac{\delta\kappa(\zeta)}{4\kappa} \delta\left(t - \frac{|z-\zeta|+|\zeta-z_s|}{c}\right) \nonumber \\
+  V_d(z,t;\zeta) &=& \pm\frac{\delta\kappa(\zeta)}{4\kappa \rho c} \delta\left(t - \frac{|z-\zeta|+|\zeta-z_s|}{c}\right) 
+\end{eqnarray*}
+
 \begin{equation}

### 2019-12-16 repro/papers/sserefl/paper.tex
prelim section

@@
 \begin{eqnarray}
+  \label{eqn:defpt}
  f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s) & \Rightarrow & S[\kappa]f = S_p
-  [\kappa]w\\ \nonumber
-  \label{eqn:specsrc}
+  [\kappa]w\\ 
+  \label{eqn:defss}
 f(\bx,t;\bx_s) = u(x,y,t;x_s,y_s)\delta(z-z_s) & \Rightarrow & S[\kappa]f =
@@
 \]
+or
+\[
+  Au(x,t;x_s) = (x-x_s)u(x,t;x_s)
+\]
+for 2D. 
 The problem formulation is
@@
 only one shot, for which I take the location to be
-$\bx_s=(0,0,0)$ and suppress $\bx_s$ from the notation.
+$\bx_s=(0,0,z_s)$.
 
@@
 \begin{eqnarray}
- \tf(\xi,z,t) = w(t)\delta(z) & \Rightarrow & \tS[\kappa]\tf = \tS_p
-  [\kappa]w\\ \nonumber
-  \label{eqn:specsrc}
+  \label{eqn:pwpt}
+ \tf(\xi,z,t) = w(t)\delta(z-z_s) & \Rightarrow & \tS[\kappa]\tf = \tS_p
+  [\kappa]w\\ 
+  \label{eqn:pwss}
 \tf(\xi,z,t) = \tu(\xi,t)\delta(z-z_s) & \Rightarrow & \tS[\kappa]\tf =
@@
 acoustics system detailed in Appendices A and B, yields
-\begin{equation}
-\label{eqn:modhom}
-\tS_s[\kappa_0]\tu(\xi,t) = \frac{1}{2(1-c_0^2\xi^2)}\tu
-\left(\xi,t-z_r\sqrt{\frac{1}{c_0^2}-\xi^2}\right).
-\end{equation}
-That is, at a homogeneous model, the plane wave modeling operator is
-simply scaling by a $\xi$-dependent factor.
-
+\begin{eqnarray}
+  \label{eqn:ptmodhompw}
+  \tS_p[\kappa_0]w(\xi,t) & = &\frac{1}{2(1-c_0^2\xi^2)}w
+                                \left(t-\frac{|z_r-z_s|}{\tc_0(\xi)}\right)\\
+    \label{eqn:ssmodhompw}
+  \tS_s[\kappa_0]\tu(\xi,t) & = &\frac{1}{2(1-c_0^2\xi^2)}\tu 
+                                  \left(\xi,t-\frac{|z_r-z_s|}{\tc_0(\xi)}\right)\\
+    \label{eqn:stmodhompw}
+  \tS[\kappa_0]\tf(\xi,t) & = &  \frac{1}{2}(1-c_0^2\xi^2)^{-1}\int dz \tf\left(\xi,z,t -
+                                \frac{|z_r-z|}{\tc_0(\xi)}\right)                              
+\end{eqnarray}
+Here
+\begin{equation}
+  \label{eqn:vertvel}
+  \tc(\xi,z) = \frac{c(z)}{\sqrt{1-c(z)^2\xi^2}}
+    \end{equation}
+is the vertical plane wave velocity, and we write
+$\tc_0(\xi)=\tc(\xi,z)$ for any $z$, when $c=c_0$.
+
+\subsection{Extended FWI in Plane Wave Domain}
+Of the various extended FWI formulation explained above, three do not
+require the storage of full space-time fields:
+\begin{itemize}
+\item Least-squares FWI;
+\item Reformulated WRI; and
+\item Surface Source Extension.
+\end{itemize}
+The remainder of this paper is devoted to analysis of these three
+approaches.
+
+It is possible to transform the objectives posed in the space-time
+domain problem statements above, but this leads to technical
+complications that do not affect the conclusions. Therefore I pose
+analogous problems directly in the plane wave domain. The $L^2$ norms
+(integral root mean squares) below involve integration over a range of
+plane wave slownesses $\xi$, which for simplicity I assume symmetric,
+that is, $-\xi_{\rm max} \le \xi \le \xi_{\rm max}$, and over
+appropriate ranges of other variables which I do not state explicitly
+here (see the analysis sections later in the paper where all such
+choices are made explicitly). As before, the $L^2$-norm squared symbol
+$\|\cdot\|^2$ means the integral over all variables of the square of
+whatever is inside.
+
+\subsubsection{Least Squares FWI}
+
+\begin{equation}
+  \label{eqn:defolspw}
+  \mbox{Given } \td, \mbox{ find }\kappa \mbox{ and }w \mbox{ to
+    minimize }
+  \| \td -  \tS_p[\kappa]w\delta(\cdot-z_s)\|^2.
+\end{equation}
+
+\subsubsection{Reformulated WRI}
+
+\begin{equation}
+  \label{eqn:defwrireformpw}
+  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }e \mbox{ to
+    minimize }
+  \| \td -  \tS[\kappa]w_*\delta(\cdot-z_s) - \tS[\kappa]\tS[\kappa]^Te\|^2+\lambda^2\|\tS[\kappa]^Te\|^2.
+\end{equation}
+
+The reformulated WRI objective involves 
+the product $\tS[\kappa]\tS[\kappa]^T$ in which $\tS[\kappa]^T$ is the
+formal or $L^2$ adjoint. For the homogeneous case,
+$\kappa=\kappa_0,\rho=\rho_0,c=c_0$,
+\begin{equation}
+  \label{eqn:stmodhompwtrans}
+  (\tS[\kappa]^T\te)(\xi,z,t)=\frac{1}{2}(1-c_0^2\xi^2)^{-1}\te\left(\xi,t +
+    \frac{|z_r-z|}{\tc_0(\xi)}\right).
+\end{equation}
+
+While I have endeavoured to limit mathematical technicalities in this background
+presentation, one such technicality intrudes at this point: the
+output of the transpose operator $\tS[\kappa]^T$ is nonzero and non-decaying arbitrarily
+far in space and in negative time. As will be obvious shortly, this
+fact prevents the composition $\tS[\kappa]\tS[\kappa]^T$ from being
+well-defined, without some further constraint being imposed on the
+domain of $\tS[\kappa]$. The simplest appears to be the assumption
+that all plane wave sources act over a limited depth range, $0 \le z
+\le z_{\rm max}$. Then the $z$ integrals appearing in the expressions
+for $S$ and $\tS$ have limits $[0,z_{\rm max}]$
+
+With this amendment, compute the composition:
+\[
+  \tS[\kappa]\tS[\kappa]^T\te(\xi,t) = \frac{1}{2}(1-c_0^2\xi^2)^{-1}
+\]
+\[
+  \times 
+  \int_0^{z_{\rm max}} dz\, \frac{1}{2}(1-c_0^2\xi^2)^{-1}\te\left(\xi,t' +
+    \frac{|z_r-z|}{\tc_0(\xi)}\right)_{t'=t-\frac{|z_r-z|}{\tc_0(\xi)}}
+\]
+\begin{equation}
+  \label{eqn:stmodhompwnorm}
+ = \frac{z_{\rm max}}{4}(1-c_0^2\xi^2)^{-2}\te(\xi,t)
+\end{equation}
+
+\subsubsection{Surface Source Extension}
+
+The surface source space-time formulation employs an
+annihilator $A$, defined in \ref{eqn:defste}. The corresponding
+operator in the plane wave domain is somewhat easier to compute for the 2D
+case: one obtains 
+\[
+\tA = D_{\xi}I_t,
+\]
+that is, differentiation in $\xi$ and integration in $t$, in either
+order. The 3D expression is somewhat more complicated. Since the
+examples below will be computed in 2D, the above definition suffices.
+
+The surface source extended FWI problem in the plane wave domain is then
+\begin{equation}
+  \label{eqn:defssepw}
+  \mbox{Given } \td, \mbox{ find }\kappa \mbox{ and }\tu \mbox{ to
+    minimize }
+  \| \td -  \tS_s[\kappa]\tu\|^2+\lambda^2\|\tA\tu\|^2.
+\end{equation}
+
+\subsection{Inversion Velocity Analysis: the zero penalty limit}
+Examination of the operators appearing in the extended inversion
+problems considered here (equations \ref{eqn:stmodhompwnorm} for
+reformulated WRI, \ref{eqn:ssmodhompw} for SSE) reveals that they
+are both invertible. This property permits a simplification of the
+penalty formulations \ref{eqn:defwrireformpw} and
+\ref{eqn:defssepw}. This simplification has been used several times in
+prior work on extended FWI
+\cite[]{Symes:EAGE15,Herve2017,HouSymes:Geo18}, and has come to be
+called Inversion Velocity Analysis (IVA) in the exploration seismology
+literature.
+
+The idea is well-known in numerical optimization, and can most easily
+be explained in the context of the SSE objective function, in
+plane-wave domain:
+\begin{equation}
+  \label{eqn:ssepen}
+  \kappa,\tu \mapsto \|\td-\tS_s[\kappa]\tu\|^2 + \lambda^2\|\tA\tu\|^2
+\end{equation}
+
+The transformation to IVA takes place in two steps. The first is
+trivial: simply divide the objective function by $\lambda^2$, to
+obtain
+\begin{equation}
+  \label{eqn:ssepeninv}
+  \kappa,\tu \mapsto \lambda^{-2}\|\td-\tS_s[\kappa]\tu\|^2 + \|\tA\tu\|^2
+\end{equation}
+Only the minimum values change: for any particular $\lambda \in
+\bR^+$, the sets of local and global minimizers are exactly the same.
+
+The second step in moving from this objective to the corresponding IVA
+objective is of value in itself: it consists of eliminating the linear
+variable $\tu$ by minimizing the function defined in \ref{eqn:ssepen}
+over $\tu$. The resulting minimum is a function of $\kappa$: {\em a
+  priori}, a local minimizer is the same as the $\kappa$ component of
+a local minimizer of the original objective. This reduction of the
+objective to a function of some of its variables is known as the {\em
+  Variable Projection Method} (VPM), especially if the eliminated variables
+appear linearly in the original problem, as they do here
+\cite[]{GolubPereyra:03}. For the plane wave surface source objective,
+the result of VPM reduction is
+\begin{equation}
+  \label{eqn:ssevpm}
+  \kappa \mapsto \lambda^{-2}\|\td-\tS_s[\kappa]\tu[\kappa]\|^2 + \|\tA\tu[\kappa]\|^2
+\end{equation}
+in which $\tu[\kappa]$ is the minimizer of the base objective
+\ref{eqn:ssepeninv} over $\tu$, with $\kappa$ held fixed:
+$\tu[\kappa]$ solves the {\em normal equation}
+\begin{equation}
+  \label{eqn:ssenorm}
+  (\lambda^{-2}\tS_s[\kappa]^T\tS_s[\kappa] +
+  \tA^T\tA)\tu[\kappa]=\lambda^{-2}\tS_s[\kappa]^T\td.
+\end{equation}
+
+The major import of VPM is that the gradient
+of the reduced objective, a key element in any calculus-based
+optimization algorithm, is relatively simple to compute: formally, it
+is simply the gradient of the original objective with the
+to-be-eliminated variables held fixed. See \cite{GolubPereyra:03} for
+an explanation, and
+\cite{vanLeeuwenMulder:09,Rickett:SEG12,LiRickettAbubakar:13} for
+applications to geophysical inversion.
+
+The final step is to observe that the VPM objective converges as
+$\lambda \rightarrow 0$. Examination of \ref{eqn:ssepeninv} suggests
+that for this to happen, the data must be fit perfectly, at least in
+the limit. This is likely to happen only if the linear modeling
+operator ($\tS_s[\kappa]$, in this case) is invertible - which it is,
+at least in the case of constant $\kappa=\kappa_0$,
+see equation \ref{eqn:ssmodhompw}.  In fact $\tS[\kappa]$ is a
+continuous function of $\kappa$, as is explained in Appendix \appSingular, so in fact data is fit exactly and by precisely one $\tu$ for
+all $\kappa$ sufficiently close to constant. It is not hard to see that the
+$\lambda = 0$ limit of the VPM objective obtained from 
+\ref{eqn:ssepeninv} is
+\begin{equation}
+  \label{eqn:ssiva}
+  \kappa \mapsto \|A \tS_s[\kappa]^{-1}\td\|^2
+\end{equation}
+That is, the IVA objective is the mean square of the annihilator ouput
+from the necessarily unique extended source field $\tu =
+\tS_s[\kappa]^{-1}\td$ that precisely fits the data.
+
+Appendix \appIVA  presents a proof that equation \ref{eqn:ssiva}
+actually describes the $\lambda \rightarrow 0$ limit. The version of
+surface source extended FWI whose properties are described in the next
+section is the minimization problem for this objective, to which I
+give a name:
+\begin{equation}
+  \label{eqn:defssiva}
+  J_{\rm SSE} = \frac{1}{2}\|A\tu[\kappa]\|^2, \,
+  \tu[\kappa]=\tS_s[\kappa]^{-1}\td
+\end{equation}
+
+The story for reformulated WRI is similar. From equation
+\ref{eqn:defwrireformpw}, the VPM objective after scaling by
+$\lambda^{-2}$ is 
+\begin{equation}
+  \label{eqn:wrivpm}
+  \kappa, \te \mapsto \lambda^{-2}\| \td -  \tS[\kappa]w_*\delta(\cdot-z_s) -
+  \tS[\kappa]\tS[\kappa]^T\te[\kappa]\|^2+\|\tS[\kappa]^T\te[\kappa]\|^2,
+\end{equation}
+in which $\te[\kappa]$ solves the normal equation
+\begin{equation}
+  \label{eqn:wrinorm}
+  (\lambda^{-2}(\tS[\kappa]\tS[\kappa]^T)^2 +
+  \tS[\kappa]\tS[\kappa]^T)\te[\kappa] =
+  \lambda^{-2}\tS[\kappa]\tS[\kappa]^T(\td -
+  \tS[\kappa]w_*\delta(\cdot-z_s) )
+\end{equation}
+If $\kappa$ is close to $\kappa_0$, then $\tS[\kappa]\tS[\kappa]^T$ is
+invertible, as follows from equation \ref{eqn:stmodhompwnorm} and the
+continuity of $\tS[\kappa]$ with respect to $\kappa$, mentioned
+above. Therefore the normal equation is equivalent to
+\begin{equation}
+  \label{eqn:wrinorm1}
+  (\lambda^{-2}\tS[\kappa]\tS[\kappa]^T +
+  I)\te[\kappa] =
+  \lambda^{-2}(\td -
+  \tS[\kappa]w_*\delta(\cdot-z_s) )
+\end{equation}
+The $\lambda \rightarrow 0$ limit of the resulting function of
+$\kappa$ is
+\[
+  J_{\rm WRI}[\kappa] = \frac{1}{2}\|\tS[\kappa]^T((\tS[\kappa]\tS[\kappa]^T)^{-1}(\td -
+  \tS[\kappa]w_*\delta(\cdot-z_s)))\|^2
+\]
+\begin{equation}
+  \label{eqn:wripwiva}
+  = \frac{1}{2}\langle \td -
+  \tS[\kappa]w_*\delta(\cdot-z_s), (\tS[\kappa]\tS[\kappa]^T)^{-1}( \td -
+  \tS[\kappa]w_*\delta(\cdot-z_s)) \rangle
+\end{equation}
+
+Finally, it is entirely possible to apply VPM to the least squares FWI
+objective \ref{eqn:defolspw}, as has been suggested as an algorithmic
+approach to determining source parameters as part of FWI
+\cite{Rickett:SEG12,LiRickettAbubakar:13}. However in this instance it
+is more relevant to pose the known-source version of least squares
+FWI, as most parallel to WRI (in any of its forms): recalling the
+identity \ref{eqn:pwpt}, 
+\[
+  J_{\rm FWI}[\kappa] = \frac{1}{2}\|\td-\tS_p[\kappa]w_*\|^2
+\]
+\begin{equation}
+  \label{eqn:olspwobj}
+  = \frac{1}{2}\langle \td -
+  \tS[\kappa]w_*\delta(\cdot-z_s), \td -
+  \tS[\kappa]w_*\delta(\cdot-z_s)\rangle.
+\end{equation}
+Comparing equations \ref{eqn:wripwiva} and \ref{eqn:olspwobj}, reach
+the surprising conclusion that
+\begin{quote}
+  The $\lambda \rightarrow 0$ limit of the WRI-VPM objective
+  \ref{eqn:wrivpm} is a weighted-norm variant of the FWI objective
+  \ref{eqn:olspwobj}.
+\end{quote}
+This conclusion is surprising because one would normally expect the
+opposite limit, that is $\lambda \rightarrow \infty$, of the WRI-VPM
+objective \ref{eqn:wrivpm} to be equivalent in some sense to the FWI
+objective \ref{eqn:olspwobj}, rather than the $\lambda \rightarrow 0$
+limit. Indeed, the limit $\lambda \rightarrow \infty$ can only make
+sense if $\tS[\kappa]\tS[\kappa]^T\te \rightarrow 0$: since
+$\tS[\kappa]\tS[\kappa]^T$ is invertible for $\kappa$ near $\kappa_0$,
+in that case $\te \rightarrow 0$ necessarily, so the objective
+\ref{eqn:wrivpm} presumably tends to \ref{eqn:olspwobj}. This conclusion
+is in fact correct, as follows from a variant of the argument
+presented in Appendix \appIVA.
+
+The import of the reasoning presented in Appendix \appIVA is that for
+$\lambda$ sufficiently small, the behaviour of the VPM objective in
+each of these cases is close to that of the $\lambda \rightarrow 0$
+limit. As is well-known, and will be recapitulated in the Main Results
+section, the least squares FWI objective does not generate
+constructive model updates. Since all of the VPM objectives explored
+here become close to the least squares FWI objective for large
+$\lambda$, it is the case of small $\lambda$ that holds any possible
+promise of generating substantially more constructive updates, and we
+will see below that indeed in some cases that promise is realized. The
+$\lambda \rightarrow 0$ limit provides a convenient window into
+small $\lambda$ behaviour, so the focus in the remainder of the paper
+will be on the limiting problems defined in \ref{eqn:olspwobj},
+\ref{eqn:wripwiva}, and \ref{eqn:defssiva}.
 \section{Main Results}
@@
 
-Assume that $f$ is square-integrable, $\kappa = \kappa(z)$,
+Assume that $\kappa = \kappa(z)$,
 $\rho=\rho(z)$ (layered medium). Then the modeled shots differ from
@@
  If $\log \kappa$ and $\log \rho$ are globally
-bounded, there is $\xi_{\rm max} > 0$ so that for $\|\xi\|<\xi_{\rm
+bounded, and $f$ is square-integrable, there is $\xi_{\rm max} > 0$ so that for $\|\xi\|<\xi_{\rm
   max}$, the integrals defining slant stack fields
@@
 $\tf(\xi,z,t) = \tw(\xi,t)\delta(z)$. The plane wave domain modeling
-operator samples $\tp$ on $z=z_r=0$:
+operator samples $\tp$ on $z=z_r$:
 \begin{equation}
 \label{eqn:modpw}
-\tS[\kappa]\tw = \tp|_{z=0}
+\tS[\kappa]\tw = \tp|_{z=z_r}
 \end{equation}
@@
 identity, $\tS[\kappa]$ is invertible in sup-norm neighborhood of
-$\kappa$. Introduce the the scaled reduced penalty function objective
-\[
-\kappa \mapsto \min_{\tw}(\frac{1}{\alpha^2}\|\tS[\kappa]\tw-\td\|^2 + \|\tA
-\tw\|^2)
-\]
-\[
-=\frac{1}{\alpha^2}\|\tS[\kappa]\tw_{\alpha}[\kappa]-\td\|^2 + \|\tA
-\tw_{\alpha}[\kappa]\|^2
-\]
-where $\tw_{\alpha}[\kappa]$ solves the normal equation:
-\[
-\tS[\kappa]^T\tS[\kappa]\tw  + \alpha^2\tA^T\tA\tw = \tS[\kappa]^T\td.
-\]
-Since $\tS[\kappa]$ is invertible, $\tw_{\alpha}[\kappa]$ is uniformly
-bounded as $\alpha \rightarrow 0$. Let $\tw_0[\kappa] =
-\tS[\kappa]^{-1}\td$. Then
-\[
-\tS[\kappa]^T\tS[\kappa](\tw_{\alpha}[\kappa]-\tw_0[\kappa])+\alpha^2\tA^T\tA\tw_{\alpha}[\kappa]
-= 0,
-\]
-so $\tw_{\alpha}[\kappa] \rightarrow \tw_0[\kappa]$ as $\alpha
-\rightarrow 0$. Consequently the scaled objective above becomes
-\[
-\frac{1}{\alpha^2}\|\tS[\kappa]\tw_{\alpha}[\kappa]-\td\|^2 + \|\tA 
-\tw_{\alpha}[\kappa]\|^2
-=\frac{1}{\alpha^2}\|\tS[\kappa](\tw_{\alpha}[\kappa]-\tw_0[\kappa])\|^2 + \|\tA 
-\tw_{\alpha}[\kappa]\|^2
-\]
-\[
-=\frac{1}{\alpha^2}\langle (\tw_{\alpha}[\kappa]-\tw_0[\kappa]),
-\tS[\kappa]^T\tS[\kappa](\tw_{\alpha}[\kappa]-\tw_0[\kappa])\rangle+ \|\tA 
-\tw_{\alpha}[\kappa]\|^2
-\]
-\[
-=\frac{1}{\alpha^2}\langle (\tw_{\alpha}[\kappa]-\tw_0[\kappa]),
--\alpha^2\tA^T\tA \tw_{\alpha}[\kappa] \rangle + \|\tA\tw_{\alpha}[\kappa]\|^2
-=\langle (\tw_{\alpha}[\kappa]-\tw_0[\kappa]),
--\tA^T\tA \tw_{\alpha}[\kappa] \rangle + \|\tA\tw_{\alpha}[\kappa]\|^2
-\]
-\begin{equation}
-\label{eqn:j0def}
-\rightarrow \|\tA\tw_0[\kappa]\|^2 \equiv 2 J_0[\kappa,\td].
-\end{equation}
-This $\alpha \rightarrow 0$ limit is the simplified objective used by
-\cite{Symes:EAGE15} and mentioned in the introduction.
-
-%\[
-%DJ[\kappa_0]\delta \kappa = -\langle D(\tS[\kappa_0]\tw)\delta \kappa,
-%\tA^T\tA S[\kappa_0]^{-1}\td \rangle , \,\tw=\tS[\kappa_0]^{-1}\td
-%\]
+$\kappa$.
 Now assume that $\td$ is data computed from a simple reflector as
@@
 \begin{eqnarray}
-\label{eqn:awe1d}
+\label{eqn:awe1dptsrc}
 \frac{\partial p}{\partial t} +\kappa\frac{\partial 
@@
 Since the right hand side is singular, so is the solution, so it must
-be a solution in the weak sense; since it is causal, it is uniquely
-determined by the system above. It follows from the weak solution
+be a solution in the weak sense. It follows from the weak solution
 conditions that the pressure is continuous at $p=0$, whence $v$ must
@@
 
-Now assume that $\kappa, \rho$ are constant. 
+Now assume that $\kappa=\kappa_0 \in \bR^+, \rho=\rho_0 \in \bR^+, c_0
+= \sqrt{\kappa_0/rho_0}$ are constant. 
 In $z \ne z_s$, the right hand side 
@@
 \[
-p(z,t)=a\left(t \mp \frac{z-z_s}{c}\right), \, v(z,t)=\pm b\left(t \mp
-  \frac{z_s}{c}\right)
+p(z,t)=a\left(t \mp \frac{z-z_s}{c_0}\right), \, v(z,t)=\pm b\left(t \mp
+  \frac{z_s}{c_0}\right)
 \]
 From the second dynamical equation (Newton's law) it follows that $b =
-a/(\rho c)$. The singularity on the LHS of the first dynamical
+a/(\rho_0 c_0)$. The singularity on the LHS of the first dynamical
 equation (constitutive law) is
@@
 \kappa [v]_{z=z_s}\delta(z-z_s) =
-2\frac{\kappa}{c}b\delta(z-z_s) = 2a\delta(z-z_s).
+2\frac{\kappa}{c_0}b\delta(z-z_s) = 2a\delta(z-z_s).
 \] 
@@
 \begin{eqnarray}
+\label{eqn:sol1dptsrc}
+p(z,t) &=& \frac{1}{2}w\left(t - \frac{|z-z_s|}{c_0}\right) \nonumber \\
+v(z,t) &=& \mbox{sgn}(z-z_s)\frac{1}{2\rho_0 c_0}w\left(t -\frac{|z-z_s|}{c_0}\right)
+           \nonumber \\
+\end{eqnarray}
+This result (computation of the Green's function for the acoustic
+system) permits an explicit expression for the system with a
+space-time source:
+\begin{eqnarray}
+\label{eqn:awe1d}
+\frac{\partial p}{\partial t} +\kappa_0\frac{\partial 
+  v}{\partial z} &=& f(z,t) \nonumber\\
+\rho_0 \frac{\partial v}{\partial t} + \frac{\partial p}{\partial 
+  z}&=&0\nonumber\\
+ p,v&=&0, t \ll 0. 
+\end{eqnarray}
+Since
+\[
+  f(z,t) = \int dz_s\,f(z_s,t)\delta(z-z_s)
+\]
+obtain
+\begin{eqnarray}
 \label{eqn:sol1d}
-p(z,t) &=& \frac{1}{2}w\left(t \mp \frac{z-z_s}{c}\right) \nonumber \\
-v(z,t) &=& \pm\frac{1}{2\rho c}w\left(t \mp \frac{z-z_s}{c}\right)
+p(z,t) &=& \frac{1}{2}\int dz_s f\left(z_s,t - \frac{|z-z_s|}{c_0}\right) \nonumber \\
+v(z,t) &=& \frac{1}{2\rho_0 c_0} \int dz_s \mbox{sgn} (z-z_s) f\left(z_s,t - \frac{|z-z_s|}{c_0}\right)
            \nonumber \\
-\pm z-z_s &>& 0.
 \end{eqnarray}
@@
 
-\append{Plane Wave Modeling of Reflection Traces}
+\append{Plane Waves}
 Re-write the plane wave system \ref{eqn:awepw} as
@@
 \frac{\partial \tp}{\partial t} + \tk\frac{\partial
-  \tv}{\partial z} &=& (1-c^2\xi^2)^{-1}\tf \nonumber\\
+  \tv}{\partial z} &=& (1-c_0^2\xi^2)^{-1}\tf \nonumber\\
 \rho \frac{\partial \tv}{\partial t} + \frac{\partial \tp}{\partial
@@
 in which $\tk(\xi,z)=(1-c(z)^2\xi^2)^{-1}\kappa(z)$ is a scaled version of
-bulk modulus. Similarly define the vertical plane wave velocity as
+bulk modulus. Define the vertical plane wave velocity as
 \[
@@
 \]
+Comparing the plane wave system \ref{eqn:awepw1} with the 1D system
+\ref{eqn:awe1d}, and using the 1D solution \ref{eqn:sol1d}, obtain
+\begin{eqnarray}
+\label{eqn:solpw}
+\tp(\xi,z,t) &=& \frac{1}{2}(1-c_0^2\xi^2)^{-1}\int dz_s \tf\left(\xi,z_s,t - \frac{|z-z_s|}{\tc_0(\xi)}\right) \nonumber \\
+\tv(\xi,z,t) &=& \frac{1}{2\rho_0 c_0}(1-c_0^2\xi^2)^{-1} \int dz_s \mbox{sgn} (z-z_s)\tf\left(\xi,z_s,t - \frac{|z-z_s|}{\tc_0(\xi)}\right)
+           \nonumber \\
+\end{eqnarray}
+
+Accordingly, the plane-wave forward map is given by
+\[
+  \tS[\kappa_0]\tf(\xi,t) =\tp(\xi,z_r,t)
+\]
+\begin{equation}
+\label{eqn:fwdpw1}
+= \frac{1}{2}(1-c_0^2\xi^2)^{-1}\int dz_s \tf\left(\xi,z_s,t -
+  \frac{|z-z_s|}{\tc_0(\xi)}\right)
+\end{equation}
+
 If $\tf(\xi,z,t)=\tw(\xi,t)\delta(z)$ is a surface
@@
 
+\append{Zero-Weight Limit of Variable Projection}
+As in the Preliminaries section, this discussion focuses on the plane wave
+Surface Source objective, in the scaled variant \ref{eqn:ssepeninv},
+reproduced here for convenience:
+\begin{equation}
+  \label{eqn:ssepeninv1}
+  \kappa,\tu \mapsto \lambda^{-2}\|\td-\tS_s[\kappa]\tu\|^2 + \|\tA\tu\|^2
+\end{equation}
+Variable projection creates a function of $\kappa$ by minimizing this
+function over $\tu$, to obtain
+\[
+\kappa \mapsto \frac{1}{\lambda^2}\|\tS[\kappa]\tu_{\lambda}[\kappa]-\td\|^2 + \|\tA
+\tu_{\lambda}[\kappa]\|^2
+\]
+where $\tu_{\lambda}[\kappa]$ solves the normal equation:
+\[
+\tS[\kappa]^T\tS[\kappa]\tu  + \lambda^2\tA^T\tA\tu = \tS[\kappa]^T\td.
+\]
+Since $\tS[\kappa]$ is invertible, $\tu_{\lambda}[\kappa]$ is
+bounded uniformly in $\lambda$ as $\lambda \rightarrow 0$. Let $\tu_0[\kappa] =
+\tS[\kappa]^{-1}\td$. Then
+\[
+\tS[\kappa]^T\tS[\kappa](\tu_{\lambda}[\kappa]-\tu_0[\kappa])+\lambda^2\tA^T\tA\tu_{\lambda}[\kappa]
+= 0,
+\]
+so $\tu_{\lambda}[\kappa] \rightarrow \tu_0[\kappa]$ as $\lambda
+\rightarrow 0$. Consequently the scaled objective above becomes
+\[
+\frac{1}{\lambda^2}\|\tS[\kappa]\tu_{\lambda}[\kappa]-\td\|^2 + \|\tA 
+\tu_{\lambda}[\kappa]\|^2
+=\frac{1}{\lambda^2}\|\tS[\kappa](\tu_{\lambda}[\kappa]-\tu_0[\kappa])\|^2 + \|\tA 
+\tu_{\lambda}[\kappa]\|^2
+\]
+\[
+=\frac{1}{\lambda^2}\langle (\tu_{\lambda}[\kappa]-\tu_0[\kappa]),
+\tS[\kappa]^T\tS[\kappa](\tu_{\lambda}[\kappa]-\tu_0[\kappa])\rangle+ \|\tA 
+\tu_{\lambda}[\kappa]\|^2
+\]
+\[
+=\frac{1}{\lambda^2}\langle (\tu_{\lambda}[\kappa]-\tu_0[\kappa]),
+-\lambda^2\tA^T\tA \tu_{\lambda}[\kappa] \rangle + \|\tA\tu_{\lambda}[\kappa]\|^2
+=\langle (\tu_{\lambda}[\kappa]-\tu_0[\kappa]),
+-\tA^T\tA \tu_{\lambda}[\kappa] \rangle + \|\tA\tu_{\lambda}[\kappa]\|^2
+\]
+\begin{equation}
+\label{eqn:j0def}
+\rightarrow \|\tA\tu_0[\kappa]\|^2 \equiv 2 J_0[\kappa,\td].
+\end{equation}
+This $\lambda \rightarrow 0$ limit is the simplified objective used by
+\cite{Symes:EAGE15} and mentioned in the Preliminaries section.
+
 \bibliographystyle{seg}

### 2019-12-17 repro/papers/sserefl/paper.tex
more edits

@@
 \[
-  f(z,t) = \int dz_s\,f(z_s,t)\delta(z-z_s)
+  f(z,t) = \int dz_1\,f(z_s,t)\delta(z-z_1)
 \]
@@
 \label{eqn:sol1d}
-p(z,t) &=& \frac{1}{2}\int dz_s f\left(z_s,t - \frac{|z-z_s|}{c_0}\right) \nonumber \\
-v(z,t) &=& \frac{1}{2\rho_0 c_0} \int dz_s \mbox{sgn} (z-z_s) f\left(z_s,t - \frac{|z-z_s|}{c_0}\right)
-           \nonumber \\
+p(z,t) &=& \frac{1}{2}\int dz_1 f\left(z_1,t - \frac{|z-z_1|}{c_0}\right) \\
+v(z,t) &=& \frac{1}{2\rho_0 c_0} \int dz_1 \mbox{sgn} (z-z_1) f\left(z_1,t - \frac{|z-z_1|}{c_0}\right)\\
 \end{eqnarray}
@@
 
-The key is to understand the perturbational impulse response. That is,
-set $w(t)=H(t)$. Then the identity \ref{eqn:sol1d} implies that the
-inhomogeneous term in the first equation of the system \ref{eqn:dawe1d}
-is 
-\begin{equation}
-  \label{eqn:pertrhs}
--\delta \kappa(z) \frac{\partial v}{\partial z}(z,t) = \frac{\delta
-  \kappa(z_s)}{\rho c}H(t)\delta(z-z_s) +
-\frac{\delta\kappa(z)}{2\kappa}\delta\left(t - \frac{|z-z_s|}{c}\right)
-\end{equation}
-Correspondingly, the solution $(\delta p, \delta v)$ is the sum of
-three components:
-\begin{eqnarray}
-  \label{eqn:pertrhsterms}
-  \delta p & = & \delta p_s + \delta p_d \nonumber\\
-  \delta v & = & \delta v_s + \delta v_d 
-\end{eqnarray}
-The first component satisfies
-\begin{eqnarray}
-  \label{eqn:dawe1ds}
-  \frac{\partial \delta p_s}{\partial t} +\kappa\frac{\partial
-  \delta v_s}{\partial z}   &=&  \frac{\delta
-  \kappa(z_s)}{\rho c}H(t)\delta(z-z_s)\nonumber\\
-  \rho \frac{\partial \delta v_s}{\partial t} + \frac{\partial \delta p_s}{\partial
-  z}&=&0\nonumber\\
-  \delta p_s,\delta v_s&=&0, t < 0.
-\end{eqnarray}
-The initial conditions are free to be chosen, and the choice above is convenient.
-
-Note that \ref{eqn:dawe1ds} is precisely a system of the form
-\ref{eqn:awe1d}, so from the formulae \ref{eqn:sol1d} with $w(t)=\frac{\delta
-  \kappa(z_s)}{\rho c}H(t)$
-\begin{eqnarray}
-\label{eqn:dsol1ds}
-\delta p_s(z,t) &=& \frac{\delta
-  \kappa(z_s)}{2\rho c}H\left(t - \frac{|z-z_s|}{c}\right) \nonumber \\
-\delta v_s(z,t) &=& \pm\frac{\delta \kappa(z_s)}{2\rho^2 c^2}H\left(t - \frac{|z-z_s|}{c}\right)
-\end{eqnarray}
-
-Define $P_d(z,t;\zeta), V_d(z,t;\zeta)$ to be the solution of the system
-\begin{eqnarray}
-  \label{eqn:dawe1dd}
-  \frac{\partial P_d}{\partial t} +\kappa\frac{\partial
-  V_d}{\partial z}   &=&   \frac{\delta\kappa(\zeta)}{2\kappa}\delta(z-\zeta)\delta\left(t - \frac{|\zeta-z_s|}{c}\right)\nonumber\\
-  \rho \frac{\partial V_d}{\partial t} + \frac{\partial P_d}{\partial
-  z}&=&0\nonumber\\
-  P_d,V_d&=&0, t < 0.
-\end{eqnarray}
-Then
-\begin{eqnarray}
-  \label{eqn:intrep1}
-  \delta p_d(z,t) & = & \int d\zeta\, P_d(z,t;\zeta)\\
-  \delta v_d(z,t) & = & \int d\zeta\, V_d(z,t;\zeta)
-\end{eqnarray}
-
-The system \ref{eqn:dawe1dd} is precisely of the form
-\ref{eqn:awe1d} with $z_s$ replaced by $\zeta$ and
-\[
-  w(t)= \frac{\delta\kappa(\zeta)}{2\kappa}\delta\left(t - \frac{|\zeta-z_s|}{c}\right)
-\]
-Thus from \ref{eqn:sol1d}
-\begin{eqnarray*}
-  P_d(z,t;\zeta) &=& \frac{\delta\kappa(\zeta)}{4\kappa} \delta\left(t - \frac{|z-\zeta|+|\zeta-z_s|}{c}\right) \nonumber \\
-  V_d(z,t;\zeta) &=& \pm\frac{\delta\kappa(\zeta)}{4\kappa \rho c} \delta\left(t - \frac{|z-\zeta|+|\zeta-z_s|}{c}\right) 
-\end{eqnarray*}
-
-\begin{equation}
-\label{eqn:imp1d}
-D(S[\kappa]H)\delta \kappa (t) = \delta p(0,t)=  \frac{\delta 
-  \kappa\left(\frac{ct}{2}\right)}{8\kappa} 
-\end{equation}
-whence for arbitrary causal wavelet $w$,
-\begin{equation}
-\label{eqn:fwd1d}
-D(S[\kappa]w)\delta \kappa (t) = \int_0^t\,ds\,w'(t-s)\frac{\delta 
-  \kappa\left(\frac{cs}{2}\right)}{8\kappa} 
+From the second equation in the pair \ref{eqn:sol1d}
+\[
+  \frac{\partial v}{\partial z}(z,t) = \frac{1}{\rho_0 c_0} f(z,t) 
+\]
+\begin{equation}
+  \label{eqn:sol1ddvdz}
+ - \frac{1}{2\kappa_0} \int\, dz_2\, \frac{\partial f}{\partial t}\left(z_2,t - \frac{|z-z_2|}{c_0}\right)
+\end{equation}
+Thus using the first equations in \ref{eqn:sol1d} and \ref{eqn:dawe1d}, 
+\[
+  \delta p(z,t) =- \frac{1}{2} \int \, dz_1 \, \delta
+  \kappa(z_1)\frac{\partial v}{\partial z}\left(z_1,t - \frac{|z-z_1|}{c_0}\right)
+\]
+\[
+  =-\frac{1}{2\rho_0c_0} \int \,dz_1\, \delta
+  \kappa(z_1) f \left(z_1,t - \frac{|z-z_1|}{c_0}\right)
+  \]
+\begin{equation}
+  \label{eqn:sol1ddp}
+  + \frac{1}{4\kappa_0} \int\,dz_2 \, \int \,dz_1\,\delta 
+  \kappa(z_1) \frac{\partial f}{\partial t}\left(z_2,t - \frac{|z_2-z_1|+|z_1-z|}{c_0}\right)
+\end{equation}
+
+An important special case is the 1D point source,
+$f(z,t)=w(t)\delta(z-z_s)$: then
+\[
+  \delta p(z,t) = -\frac{\delta \kappa(z_s)}{2\rho_0c_0}w\left(t - \frac{|z-z_s|}{c_0}\right)
+\]
+\begin{equation}
+  \label{eqn:sol1ddppw}
+  +\frac{1}{4\kappa_0} \int \,dz_1\,\delta 
+  \kappa(z_1) \frac{\partial w}{\partial t}\left(t - \frac{|z_s-z_1|+|z_1-z|}{c_0}\right)
 \end{equation}
@@
 \[
-\tc(\xi,z)=\frac{c(z)}{1-c(z)^2\xi^2} =
+\tc(\xi,z)=\frac{c(z)}{\sqrt{1-c(z)^2\xi^2}} =
 \left(\frac{1}{c(z)^2}-\xi^2\right)^{-1/2}.

### 2019-12-20 repro/papers/sserefl/paper.tex
first results

@@
     minimize }
-  \| d -  S[\kappa]f\|^2+\lambda^2\|f-w_*\delta(\cdot-\bx_s)\|^2.
+  \| d -  S[\kappa]f\|^2+\alpha^2\|f-w_*\delta(\cdot-\bx_s)\|^2.
 \end{equation}
@@
     minimize }
-  \| d -  S[\kappa]w_*\delta(\cdot-\bx_s) - S[\kappa]S[\kappa]^Te\|^2+\lambda^2\|S[\kappa]^Te\|^2.
+  \| d -  S[\kappa]w_*\delta(\cdot-\bx_s) - S[\kappa]S[\kappa]^Te\|^2+\alpha^2\|S[\kappa]^Te\|^2.
 \end{equation}
@@
     minimize }
-  \| d -  S[\kappa]f\|^2+\lambda^2\|Af\|^2.
+  \| d -  S[\kappa]f\|^2+\alpha^2\|Af\|^2.
 \end{equation}
@@
   \label{eqn:defstereform}
-  \| d -  S[\kappa]w_*\delta(\cdot-\bx_s) - S[\kappa]S[\kappa]^Te\|^2+\lambda^2\|A(S[\kappa]^Te+n)\|^2.
+  \| d -  S[\kappa]w_*\delta(\cdot-\bx_s) - S[\kappa]S[\kappa]^Te\|^2+\alpha^2\|A(S[\kappa]^Te+n)\|^2.
 \end{equation}
@@
     minimize }
-  \| d -  S_s[\kappa]u\|^2+\lambda^2\|Au\|^2.
+  \| d -  S_s[\kappa]u\|^2+\alpha^2\|Au\|^2.
 \end{equation}
@@
   \label{eqn:ptmodhompw}
-  \tS_p[\kappa_0]w(\xi,t) & = &\frac{1}{2(1-c_0^2\xi^2)}w
+  \tS_p[\kappa_0]w(\xi,t) & = &\frac{1}{2c_0(1-c_0^2\xi^2)^{1/2}}w
                                 \left(t-\frac{|z_r-z_s|}{\tc_0(\xi)}\right)\\
     \label{eqn:ssmodhompw}
-  \tS_s[\kappa_0]\tu(\xi,t) & = &\frac{1}{2(1-c_0^2\xi^2)}\tu 
+  \tS_s[\kappa_0]\tu(\xi,t) & = &\frac{1}{2c_0(1-c_0^2\xi^2)^{1/2}}\tu 
                                   \left(\xi,t-\frac{|z_r-z_s|}{\tc_0(\xi)}\right)\\
     \label{eqn:stmodhompw}
-  \tS[\kappa_0]\tf(\xi,t) & = &  \frac{1}{2}(1-c_0^2\xi^2)^{-1}\int dz \tf\left(\xi,z,t -
+  \tS[\kappa_0]\tf(\xi,t) & = &  \frac{1}{2c_0(1-c_0^2\xi^2)^{1/2}}\int dz \tf\left(\xi,z,t -
                                 \frac{|z_r-z|}{\tc_0(\xi)}\right)                              
@@
     minimize }
-  \| \td -  \tS[\kappa]w_*\delta(\cdot-z_s) - \tS[\kappa]\tS[\kappa]^Te\|^2+\lambda^2\|\tS[\kappa]^Te\|^2.
+  \| \td -  \tS[\kappa]w_*\delta(\cdot-z_s) - \tS[\kappa]\tS[\kappa]^Te\|^2+\alpha^2\|\tS[\kappa]^Te\|^2.
 \end{equation}
@@
     minimize }
-  \| \td -  \tS_s[\kappa]\tu\|^2+\lambda^2\|\tA\tu\|^2.
+  \| \td -  \tS_s[\kappa]\tu\|^2+\alpha^2\|\tA\tu\|^2.
 \end{equation}
@@
   \label{eqn:ssepen}
-  \kappa,\tu \mapsto \|\td-\tS_s[\kappa]\tu\|^2 + \lambda^2\|\tA\tu\|^2
+  \kappa,\tu \mapsto \|\td-\tS_s[\kappa]\tu\|^2 + \alpha^2\|\tA\tu\|^2
 \end{equation}
@@
 The transformation to IVA takes place in two steps. The first is
-trivial: simply divide the objective function by $\lambda^2$, to
+trivial: simply divide the objective function by $\alpha^2$, to
 obtain
@@
   \label{eqn:ssepeninv}
-  \kappa,\tu \mapsto \lambda^{-2}\|\td-\tS_s[\kappa]\tu\|^2 + \|\tA\tu\|^2
-\end{equation}
-Only the minimum values change: for any particular $\lambda \in
+  \kappa,\tu \mapsto \alpha^{-2}\|\td-\tS_s[\kappa]\tu\|^2 + \|\tA\tu\|^2
+\end{equation}
+Only the minimum values change: for any particular $\alpha \in
 \bR^+$, the sets of local and global minimizers are exactly the same.
@@
   \label{eqn:ssevpm}
-  \kappa \mapsto \lambda^{-2}\|\td-\tS_s[\kappa]\tu[\kappa]\|^2 + \|\tA\tu[\kappa]\|^2
+  \kappa \mapsto \alpha^{-2}\|\td-\tS_s[\kappa]\tu[\kappa]\|^2 + \|\tA\tu[\kappa]\|^2
 \end{equation}
@@
   \label{eqn:ssenorm}
-  (\lambda^{-2}\tS_s[\kappa]^T\tS_s[\kappa] +
-  \tA^T\tA)\tu[\kappa]=\lambda^{-2}\tS_s[\kappa]^T\td.
+  (\alpha^{-2}\tS_s[\kappa]^T\tS_s[\kappa] +
+  \tA^T\tA)\tu[\kappa]=\alpha^{-2}\tS_s[\kappa]^T\td.
 \end{equation}
@@
 The final step is to observe that the VPM objective converges as
-$\lambda \rightarrow 0$. Examination of \ref{eqn:ssepeninv} suggests
+$\alpha \rightarrow 0$. Examination of \ref{eqn:ssepeninv} suggests
 that for this to happen, the data must be fit perfectly, at least in
@@
 all $\kappa$ sufficiently close to constant. It is not hard to see that the
-$\lambda = 0$ limit of the VPM objective obtained from 
+$\alpha = 0$ limit of the VPM objective obtained from 
 \ref{eqn:ssepeninv} is
@@
 Appendix \appIVA  presents a proof that equation \ref{eqn:ssiva}
-actually describes the $\lambda \rightarrow 0$ limit. The version of
+actually describes the $\alpha \rightarrow 0$ limit. The version of
 surface source extended FWI whose properties are described in the next
@@
 \ref{eqn:defwrireformpw}, the VPM objective after scaling by
-$\lambda^{-2}$ is 
+$\alpha^{-2}$ is 
 \begin{equation}
   \label{eqn:wrivpm}
-  \kappa, \te \mapsto \lambda^{-2}\| \td -  \tS[\kappa]w_*\delta(\cdot-z_s) -
+  \kappa, \te \mapsto \alpha^{-2}\| \td -  \tS[\kappa]w_*\delta(\cdot-z_s) -
   \tS[\kappa]\tS[\kappa]^T\te[\kappa]\|^2+\|\tS[\kappa]^T\te[\kappa]\|^2,
@@
   \label{eqn:wrinorm}
-  (\lambda^{-2}(\tS[\kappa]\tS[\kappa]^T)^2 +
+  (\alpha^{-2}(\tS[\kappa]\tS[\kappa]^T)^2 +
   \tS[\kappa]\tS[\kappa]^T)\te[\kappa] =
-  \lambda^{-2}\tS[\kappa]\tS[\kappa]^T(\td -
+  \alpha^{-2}\tS[\kappa]\tS[\kappa]^T(\td -
   \tS[\kappa]w_*\delta(\cdot-z_s) )
@@
   \label{eqn:wrinorm1}
-  (\lambda^{-2}\tS[\kappa]\tS[\kappa]^T +
+  (\alpha^{-2}\tS[\kappa]\tS[\kappa]^T +
   I)\te[\kappa] =
-  \lambda^{-2}(\td -
+  \alpha^{-2}(\td -
   \tS[\kappa]w_*\delta(\cdot-z_s) )
 \end{equation}
-The $\lambda \rightarrow 0$ limit of the resulting function of
+The $\alpha \rightarrow 0$ limit of the resulting function of
 $\kappa$ is
@@
 \begin{quote}
-  The $\lambda \rightarrow 0$ limit of the WRI-VPM objective
+  The $\alpha \rightarrow 0$ limit of the WRI-VPM objective
   \ref{eqn:wrivpm} is a weighted-norm variant of the FWI objective
@@
 This conclusion is surprising because one would normally expect the
-opposite limit, that is $\lambda \rightarrow \infty$, of the WRI-VPM
+opposite limit, that is $\alpha \rightarrow \infty$, of the WRI-VPM
 objective \ref{eqn:wrivpm} to be equivalent in some sense to the FWI
-objective \ref{eqn:olspwobj}, rather than the $\lambda \rightarrow 0$
-limit. Indeed, the limit $\lambda \rightarrow \infty$ can only make
+objective \ref{eqn:olspwobj}, rather than the $\alpha \rightarrow 0$
+limit. Indeed, the limit $\alpha \rightarrow \infty$ can only make
 sense if $\tS[\kappa]\tS[\kappa]^T\te \rightarrow 0$: since
@@
 The import of the reasoning presented in Appendix \appIVA is that for
-$\lambda$ sufficiently small, the behaviour of the VPM objective in
-each of these cases is close to that of the $\lambda \rightarrow 0$
+$\alpha$ sufficiently small, the behaviour of the VPM objective in
+each of these cases is close to that of the $\alpha \rightarrow 0$
 limit. As is well-known, and will be recapitulated in the Main Results
@@
 here become close to the least squares FWI objective for large
-$\lambda$, it is the case of small $\lambda$ that holds any possible
+$\alpha$, it is the case of small $\alpha$ that holds any possible
 promise of generating substantially more constructive updates, and we
 will see below that indeed in some cases that promise is realized. The
-$\lambda \rightarrow 0$ limit provides a convenient window into
-small $\lambda$ behaviour, so the focus in the remainder of the paper
+$\alpha \rightarrow 0$ limit provides a convenient window into
+small $\alpha$ behaviour, so the focus in the remainder of the paper
 will be on the limiting problems defined in \ref{eqn:olspwobj},
 \ref{eqn:wripwiva}, and \ref{eqn:defssiva}.
+
 \section{Main Results}
 
-The space dimension is not terribly important in the results to
-follow. Generally I will assume it to take the physical value $d=3$,
-though the numerical examples will use $d=2$ for reasons of
-computational cost. For spatial position, write $\bx = (x,y,z)$ (or $\bx = (x,z)$ for
-2D).
-
-\cite{BlazekStolkSymes:13} give conditions under which the system
-\ref{eqn:awe} has a strong solution, that is, in which all of the
-terms in equations above are interpretable as functions of space-time. Assume that
-
-\indent Assumption A: $\log \kappa, \log \rho$ are bounded and
-measureable. \\
-
-\indent Assumption B: $f$ is square-integrable, has a square-integrable time
-  derivative, and \\has compact support.\\
-
-Then Theorem 1 in \cite{BlazekStolkSymes:13} implies that there exists
-a unique causal (vanishing for $t \ll 0$) strong solution
-$(p,\bv)$ of the system \ref{eqn:awe}, that is, $p$ and $\bv$ are
-$C^1$ functions of time with square-integrable (in space) values,
-$\nabla p$ and $\nabla \cdot \bv$ are $C^0$ functions of time with
-square-integrable values,  and the equations \ref{eqn:awe} are
-satisfied pointwise a. e.
-
-The central object in this study is the relation between the
-coefficients $\kappa$ and $\rho$, on the one hand, and the fields $p$
-and $\bv$ on the other. The influence of the wave velocity $c$ is the
-most important, so from here on $\rho$ will be fixed and constant:
-$\rho(\bx) = \rho_0 \in \bR^+$. Thus view $p$ and $\bv$ as functions of
-$\kappa$ and $f$.
-
-In seismic and ultrasonic experiments, the acoustic
-fields are recorded at or near a hypersurface (the surface of the
-earth, or the skin, or...). Theorem 2 in \cite[]{BlazekStolkSymes:13}
-assures that the trace of $p$ on a cylindrical hypersurface $\Gamma
-\times [0,T]$ is well-defined, as is the trace of the normal component
-of $\bv$. These traces will serve as idealized data in the discussion
-to follow, ignoring issues of sampling and other details of actual
-measurement processes.
-
-This discussion further idealizes acquisition by setting $\Gamma =
-\{\bx: z=z_r\}$. This is not a bad approximation for some field
-settings, and is convenient for the eventual use of the Radon
-transform.
-
-Active (artificial) sources (right-hand sides in the wave equations)
-are frequently localized in space, relative to other active length scales. The
-simplest such localized source is the {\em isotropic radiator},
-\begin{equation}
-  \label{eqn:ipr}
-  f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s).  
-\end{equation} 
-in which $\bx_s$ is the location of the radiator. The radiator is a
-special case of the {\em surface source}, in which the support of the
-right-hand side in \ref{eqn:awe} is confined to a hypersurface. For
-present purposes, it suffices to consider sources localized on the
-hyperplane $z=z_s$:
-\begin{equation}
-  \label{eqn:ss}
-  f(\bx,t;\bx_s) = w(x,y,t)\delta(z-z_s). 
-\end{equation}
-Neither distribution defined by \ref{eqn:ipr}, \ref{eqn:ss} is 
-square-integrable, so the theory in \cite[]{BlazekStolkSymes:13} does
-not apply directly to the system \ref{eqn:awe} with $f$ defined by
-equation \ref{eqn:ipr}. However a simple trick produces a unique
-(necessarily weak) solution in both cases, provided an additional
-constraint is placed on $\kappa$:
-
-\indent Assumption C: There exists $\epsilon >0$ so that $\kappa =
-\kappa_0 \in \bR^+$ in $\{\bx: |z-z_s|<\epsilon\}$.
-
-\noindent That is, $\kappa$ is constant near the plane $z=z_s$. See
-Appendix \appSingular for a proof of this statement.
-
-Define the coefficient-to-data map (``forward map'', ``modeling
-operator'') $S$ by
-\begin{equation}
-\label{eqn:fwd}
-S[\kappa]f = p|_{z=z_r}
-\end{equation}
-The notation reminds the reader that $S$ is linear in $f$ but
-nonlinear in $\kappa$.
-
-Theorem 2 in \cite[]{BlazekStolkSymes:13} assures that $S[\kappa]$ is
-well-defined as a map from right-hand sides $f$ having
-square-integrable time derivatives to square-integrable traces. With
-more regularity in time, the map is also differentiable in its
-dependence on $\kappa$. In Appendix \ref{appSingular}, this result is
-extended to the singular sources \ref{eqn:ipr}, \ref{eqn:ss}.
-
-In fact, stronger results hold under Assumption C. The special cases
-considered below have stronger continuity properties.
-
-After Radon transform, this source takes the form
-$\tw(\xi,t)\delta(z)$, where $\xi$ is the plane wave slowness.
-
-The underlying hypothesis is that the data comes from a point
-source. Extension methods encode such information in an annihilator,
-an operator that vanishes for models that satisfy the hypothesis. For
-SSE-FWI, a natural annilator is a variant on the operator introduced
-for this purpose by \cite{HuangSymes:Geo18a,HuangSymes:Geo18b} for two
-similar source extension methods, namely multiplication by offset,
-also reminiscent of a now-standard annihilator for subsurface offset
-introduced by \cite{stolk2001c}. In the plane wave domain,
-multiplication by offset becomes 
-\[
-\tA = D_{\xi}I_t,
-\]
-that is, differentiation in $\xi$ and integration in $t$, in either
-order.
-
-The forward map $\tS$ for plane-wave transformed inverse problem maps
-extended plane wave source wavelets $\tw(\xi,t)$ to plane wave data
-traces (sampling of the plane-wave field at $z=0$, assuming this to be the
-depth of sources and receivers, as I do here). I will regard this operator as a
-function of the bulk modulus $\kappa(z)$, by assuming the density
-$\rho$ to be fixed and homogeneous.
-
-$\tS[\kappa]$ is 
-actually invertible, at least for $\kappa$ close to homogeneous. This conclusion follows for instance
-from explicit expressions in two special cases: for homogeneous models
-$\kappa=\kappa_0$, 
-\[
-\tS[\kappa_0]\tw(\xi,t) = a_0(\xi)\tw(\xi,t),
-\]
-and for a smooth model $\kappa_1$ with an embedded reflector at $z=z_1$,
-\[
-\tS[\kappa_1](\xi,t) = a_0(\xi)\wl(t) +
-R(\xi)\wl\left(t-2\tau(\xi,z_1)\right) + O(\lambda)
-\]
-Here the source is non-extended: $\tw(\xi,t)=\wl(t)$, and 
+As explained in the introduction, the wave propagation regime implies
+a separation of scales, between the propagation distance ($O(1)$ in
+appropriate units), and the (smaller) wavelength scale. The results of
+this paper are valid for ``small'' wavelengths, that is, as the
+wavelength parameter $\lambda \rightarrow 0$. Wavelength is a data
+property, induced by a corresponding feature of the source. Recall
+the assumption that the data is (well-approximated by) point-source
+data for an appropriate choice of acoustic parameter fields $\kappa$,
+$\rho$. A convenient way of introducing wavelength as a parameter in
+the acoustic dynamic fields is via a {\em family} of source wavelets $\wl$:
 \[
@@
 \]
-is a family of wavelets depending on central wavelength $\lambda$. The
-vertical travel time $\tau(\xi,z_1)$ is given by
-\[
-\tau(\xi,z_1) = \int_0^{z_1}\sqrt{\frac{1}{c^2}-\xi^2},\,\,c=\sqrt{\kappa/rho}.
-\]
-
-Since $\tS$ is invertible, it is possible to use the
-simplified objective presented by
-\cite{Symes:EAGE15}:
-\begin{equation}
-\label{eqn:dsobj}
-J_0[\kappa]=\frac{1}{2}\|\tA \tS[\kappa]^{-1}\td\|^2
-\end{equation}
-A derivation of this limit case from the standard penalty formulation
-of extended FWI 
-appears in the last section. Inversion based on this type of objective
-has been called ``inversion velocity analysis'' in the subsurface
-offset context \cite[]{Herve2017},\cite[]{HouSymes:Geo18}, and is also the
-basis of Adaptive Waveform Inversion
-\cite{Warner:14,Warner:16}.
-
-The main result pertains to the gradient at homogeneous
-$\kappa=\kappa_0$ for data resulting from a single flat reflector
-embedded in a smooth background:
-\[
-\tdl = \tS[\kappa_1]\wl
-\]
-I calculate the gradient of $J_0$ in this configuration.
-The result is given in equation \ref{eqn:jgrad} in the last
-section. Converted to spatial frequency via the Fourier transform, it
-becomes
-\begin{equation}
-\label{eqn:jgradft-intro}
-\widehat{\nabla J[\kappa_0]}(k) = -\frac{c_0}{16}\int\, d\xi\,a_1(\xi) 
-\frac{1}{ik}\hat{u}_{\lambda}(\xi,k)D_{\xi}^2\overline{\hat{u}_{\lambda}(\xi,k)}
-\end{equation}
-where
-\[
-\hat{u}_{\lambda}(\xi,k) = \left(1 +
-  \frac{R(\xi)}{a_0(\xi)}e^{4ik\tau(\xi,z_1)/\tc(\xi)}\right)\hat{\wl}\left(\frac{2k}{\tc(\xi)}\right) 
-\]
-and 
-\[
-\tc(\xi) = \left(\frac{1}{c_0}^2-\xi^2\right)^{-1/2}
-\]
-is the vertical plane wave velocity for homogenous wave velocity
-$c_0 = \sqrt{\kappa_0/rho}$.
-
-As is reasonable, even mandatory, for simulation of seismic data,
+These wavelets are scaled to have constant power (mean-square)
+independent of $\lambda$. As is reasonable, even mandatory, for simulation of seismic data, 
 assume that the ``mother wavelet'' $w_1$ has at least two vanishing 
-moments. Then for any $k_{\rm max} >0$, $\xi_{\rm max} > 0$,
-\[
-\max \{\left|\hat{w}_{\lambda}\left(2k/\tc{\xi}\right)\right|:
-|k|\le k_{\rm max}, |\xi|\le \xi_{\rm max} \} \rightarrow 0\mbox{ as }
-\lambda \rightarrow 0.
-\]
-It follows immediately from equation \ref{eqn:jgradft-intro} that the
-same is true of the gradient:
-\begin{equation}
-\label{eqn:bigdeal}
-\max \{\left|\widehat{\nabla J[\kappa_0]}(k)\right|:
-|k|\le k_{\rm max}, |\xi|\le \xi_{\rm max} \} \rightarrow 0\mbox{ as }
-\lambda \rightarrow 0.
-\end{equation}
-
-Therefore the rate at which traveltime changes for any plane wave
-component in the specified range, as the bulk modulus is changed along
-the direction of the negative gradient, is vanishingly small as
-$\lambda \rightarrow 0$.
-
-Otherwise put, for high-frequency reflection data, this method cannot
-update the velocity away from a homogenous initial guess in a way that
-is kinematically constructive.
-
-It may even be possible to show that the update may be destructive, by
-more careful examination of the gradient approximation. However, who
-cares, really: it is established that SSE-FWI has similar behaviour to
-FWI for at least one simple instance of reflection data.
-
-\section{Plane Wave Modeling}
-
-Assume that $\kappa = \kappa(z)$,
-$\rho=\rho(z)$ (layered medium). Then the modeled shots differ from
-each other by a horizontal translation, so that there is in effect
-only one shot, for which I take the location to be
-$\bx_s=(0,0,z_s)$ and suppress $\bx_s$ from the notation.
-
-To make things even simpler, set $z_s=z_r=0$. Permitting arbitrary
-values just makes the arithmetic more complicated, without affecting
-the essential results.
-
- If $\log \kappa$ and $\log \rho$ are globally
-bounded, and $f$ is square-integrable, there is $\xi_{\rm max} > 0$ so that for $\|\xi\|<\xi_{\rm
-  max}$, the integrals defining slant stack fields
-$\tilde{p},\tilde{\bv}$, namely
-\begin{eqnarray}
-\label{eqn:slant}
-\tp(\xi,z,t)& = &\int dx p(x,z,t+\xi x),\nonumber\\
-\tv(\xi,z,t)&=&\int dx \bv(x,z,t+\xi x),\nonumber\\
-\tf(\xi,z,t)&=&\int dx f(x,z,t+\xi x),
-\end{eqnarray}
-are absolutely convergent, and $\tp, \tv$ satisfy 
-\begin{eqnarray}
-\label{eqn:awepw}
-(1-c^2\xi^2)\frac{\partial \tp}{\partial t} + \kappa\frac{\partial
-  \tv}{\partial z} &=& \tf \nonumber\\
-\rho \frac{\partial \tv}{\partial t} + \frac{\partial \tp}{\partial
-  z}&=&0\nonumber\\
-\tp,\tv&=&0, t \ll 0.
-\end{eqnarray}
-Here $c=\sqrt{\kappa/\rho}$ is the sound velocity. I use $\xi$ instead
-of the more usual $p$ for plane wave slowness since $p$ already has
-been assigned to mean pressure. Note that if
-$f(\bx,t;\bx_s)=w(t)\delta(\bx)$, then
-$\tf(\xi,z,t)=w(t)\delta(z)$. Likewise for a surface extended source,
-$f(\bx,t;\bx_s) = w(x,t)\delta(z)$, $\tf(\xi,z,t) = \tw(\xi,t)\delta(z)$.
-
-From now on assume a surface extended source, so in plane wave domain
-$\tf(\xi,z,t) = \tw(\xi,t)\delta(z)$. The plane wave domain modeling
-operator samples $\tp$ on $z=z_r$:
-\begin{equation}
-\label{eqn:modpw}
-\tS[\kappa]\tw = \tp|_{z=z_r}
-\end{equation}
-
-For homogeneous $\kappa=\kappa_0, \rho=\rho_0, c=c_0$, a computation
-recapitulated in the Appendix C, relying on computations for the 1D
-acoustics system detailed in Appendices A and B, yields
-\begin{equation}
-\label{eqn:modhom}
-\tS[\kappa_0]\tw(\xi,t) = \frac{1}{2(1-c_0^2\xi^2)}\tw(\xi,t).
-\end{equation}
-That is, at a homogeneous model, the plane wave modeling operator is
-simply scaling by a $\xi$-dependent factor.
-
-The derivative of the plane wave modeling operator $\tS$ with respect
-to $\kappa$ is the $z=0$ trace of the pressure perturbation field
-$\delta \tp$, a solution component of the perturbational plane wave
-system
-\begin{eqnarray}
-\label{eqn:dawepw}
-(1-c^2\xi^2)\frac{\partial \delta \tp}{\partial t} -
-  \delta \kappa \frac{\xi^2}{\rho}\frac{\partial \tp}{\partial t}
-  \nonumber\\
-+\kappa\frac{\partial
-  \delta \tv}{\partial z} + \delta \kappa\frac{\partial
-  \tv}{\partial z} &=& 0 \nonumber\\
-\rho \frac{\partial \delta \tv}{\partial t} + \frac{\partial \delta \tp}{\partial
-  z}&=&0\nonumber\\
-\delta \tp,\delta \tv&=&0, t \ll 0.
-\end{eqnarray}
-The derivative of the modeling operator $\tS$ is simply the trace at
-$z=0$ of the plane wave pressure field $\tp$:
-\begin{equation}
-\label{eqn:dmodpw}
-D(\tS[\kappa]\tw)\delta \kappa = \delta \tp|_{z=0}
-\end{equation}
-For homogeneous background model $\kappa=\kappa_0$, the appendices
-explain how to develop an explicit expression for this operator: one
-obtains
-\begin{equation}
-\label{eqn:dmodhom}
-D(\tS[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
-\frac{c_0}{16\kappa_0 (1-c_0^2\xi^2)^{5/2}}\int \,ds\,\tw(\xi,t-s)(\delta \kappa)'  \left(\frac{c_0
-      s}{2\sqrt{1-c_0^2\xi^2}}\right)
-\end{equation}
-To compute the gradient of the SSE objective function, the transpose
-of this operator (as a linear map on $\delta \kappa$) is required:
-that is
+moments, that is, vanishes at zero frequency, and is smooth, that is,
+also vanishes at zero wavelength. Also assume that the source acts
+over a finite time interval.
+
+Correspondingly, the target data is also a $\lambda$-dependent family:
+\begin{equation}
+  \label{eqn:datadef}
+  \tdl = \tS_p[\kappa_*]\wl.
+\end{equation}
+in which $\kappa_*$ is the target bulk modulus, an estimate of which
+is the object of the inversion exercise.
+
+For each of the approaches mentioned in the last section, I will
+analyze two cases, transmission and reflection. In all cases, $\kappa$
+obeys Assumptions A - C, so that the system \ref{eqn:awepw} and its relatives have
+well-defined solutions, thanks to the constructions presented in
+Appendix \appSingular.
+
+\paragraph{Transmission:} $z_s < z_r$, $\kappa_*$ is smooth and $ = \kappa_0$ for
+$z$ near $z_s$, $\ge \kappa_0$ for all $z$, and $>\kappa_0$ for $z$
+near $z_r$, Geometric acoustics implies that the solution is given by
+the progressing wave expansion [REFERENCE] in terms of the vertical
+travel time $\tau(\xi,z_1,z_2)$ and geometric amplitude
+$a(\xi,z_1,z_2)$ between depths $z_1$ and $z_2$. The traveltime is given by
+\[
+\tau(\xi,z_1,z_2) = \int_{z_1}^{z_2}\,dz\, \frac{1}{\tc(\xi,z)}.
+\]
+I do not give an explicit expression for the amplitude, noting only
+that it is smooth in all three variables and positive. The geometric
+acoustics approximation of the data is
+\begin{equation}
+  \label{eqn:geosoltrans}
+  \tdl(\xi,t) = a(\xi,z_s,z_r)\wl(t-\tau(\xi,z_s,z_r)).
+\end{equation}
+The explicit expression \ref{eqn:ptmodhompw} for $\kappa_0$ may be
+viewed as a special case of \ref{eqn:geosoltrans}, and gives reference
+data for that case:
+\begin{equation}
+  \label{eqn:analsoltrans}
+  \td_{0,\lambda}(\xi,t) =  \tS_p[\kappa_0]\wl(\xi,t)  = \frac{1}{2c_0(1-c_0^2\xi^2)^{1/2}}\wl
+                                \left(t-\frac{|z_r-z_s|}{\tc_0(\xi)}\right)
+\end{equation}
+
+\paragraph{Reflection:}
+In the reflection configuration, $z_r=z_s$, $\kappa_*(z)=\kappa_0$ for
+$z<z_s$, and $\kappa_*$ is smooth except for a non-zero jump at $z=z_j
+> z_s$. Geometric acoustics provides an approximate solution
+containing a reflection for $z<z_j$, in particular at $z=z_s=z_r:$
+\[
+\tS[\kappa_*](\xi,t) = a(\xi,z_s,z_s)\wl(t) +
+R(\xi)\wl\left(t-2\tau(\xi,z_s,z_j)\right) + O(\lambda)
+\]
+The reflection coefficient $R(\xi)$ is a function of the values of
+$\kappa_*$ on either side of the jump, and of $\rho_0$. It is
+generically non-zero except possibly for isolated values of $\xi$. Its precise
+value is of no concern.
+
+Having described these two configurations, I will now discuss the
+behaviour of the three approaches to waveform inversion described in
+the previous section, in each configuration. The crux of the matter is
+the bahaviour of the gradient of the objective function in each case,
+computed at the reference model $\kappa_0$,
+as the wavelength parameter $\lambda \rightarrow 0$. Each objective -
+the basic least squares objective in the first case, the IVA
+objectives for the other two - is stable, in fact approaches a non-
+zero limit. If the size of the gradient tends to zero, on the other
+hand, then the gradient contains vanishingly little information about
+constructive updates of $\kappa$ and gradient-based optimization
+algorithms can be expected to perform poorly.
+
+\subsection{Least Squares FWI}
+
+wri
+
+sse
+
+\section{Derivations of Main Results}
+
+\subsection{Least Squares FWI, Transmission Configuration}
+
+The gradient of the objective, given in equation \ref{eqn:olspwobj},
+is
+\[
+  \nabla J_{\rm FWI}[\kappa] = D\tS[\kappa]^T(\tS[\kappa]\tw_*-d)
+\]
+
+For homogeneous background model $\kappa=\kappa_0$, Appendix \appPlane
+develops an explicit expression 
+\[
+D(\tS_p[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
+  -\frac{1}{2\kappa_0c_0(1-c_0^2\xi^2)^{1/2}} \delta 
+  \kappa(z_s) w\left(t - \frac{|z_r-z_s|}{\tc_0(\xi)}\right)
+\]
+\begin{equation}
+\label{eqn:dmodhompwpt}
+    + \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int \,dz\,\delta 
+  \kappa(z) \frac{\partial w}{\partial
+    t}\left(t - \frac{|z_s-z|+|z-z_r|}{\tc_0(\xi)}\right)
+\end{equation}
+The conditions imposed on the transmission configuration imply that
+$\delta \kappa(z) = 0$ for $z$ near $z_s$. This constraint implies that
+\begin{equation}
+  \label{eqn:dmodhompwpt0}
+  D(\tS_p[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
+     \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int \,dz\,\delta 
+  \kappa(z)  \frac{\partial w}{\partial
+    t}\left(t - \frac{|z_s-z|+|z-z_r|}{\tc_0(\xi)}\right).
+\end{equation}
+Thus
 \begin{equation}
@@
 D(\tS[\kappa]\tw)_{\kappa=\kappa_0}^T\td(z) =
-- 
-\frac{1}{8} \int \,d\xi\,\frac{1}{\kappa_0 (1-c_0^2\xi^2)^2}\frac{\partial}{\partial z} \int \,dt\,
-  w\left(\xi,t-\frac{2z\sqrt{1-c_0^2\xi^2}}{c_0}\right)d(t).
+\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\,\int\,dt\,
+\frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)}
+\frac{\partial w}{\partial
+    t}\left(t - \frac{|z_s-z|+|z-z_r|}{\tc_0(\xi)}\right) \td(\xi,t).
 \end{equation}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Simple Reflection Data}
-
-Inclusion of reflections simply means permitting $\kappa$ to be
-non-smooth. The simplest example is the layer-over-half-space. The
-data to be described in this section belongs to a model having the
-same properties at the measurement surface $z=0$ as the homogeneous
-reference model: that is, $\kappa(0)=\kappa_0$. So to introduce a
-reflector with different moveout than that implied by the reference
-model, it is necessary to allow $\kappa$ to be non-constant between
-the surface an the reflector depth, say $z_1 > 0$.
-
-Suppose that $\kappa_1$ is smooth in the intervals $[0,z_1)$ and
-$(z_1,\infty)$, but undergoes a nonzero jump at $z=z_1$, 
-\[
-[\kappa_1]_{z=z_1} \ne 0.
-\]
-The proper notion of solution for wave equations such as equation
-\ref{eqn:awe} or \ref{eqn:awepw} with discontinuous coefficients such
-as $\kappa_1$ is that of {\em weak solution}, see
-\cite{BlazekStolkSymes:13}. For the point source case, $\tw(\xi,t) =
-w(t)$, with smooth $w$, a weak solution can be constructed that is
-smooth in $z \ne z_1$: this construction is the familiar combination
-of reflected and transmitted waves. Since causal weak solutions are
-unique, this is {\em the} solution. The smooth fields on either side
-of the reflector at $z=z_1$ can approximated via geometric optics,
-leading to an asymptotic
-approximation to the solution. To express the asymptotic nature of
-this approximation, consider a family of source wavelets, rather than
-a single wavelet
-\[
-\wl = \frac{1}{\sqrt{\lambda}}w_1\left(\frac{t}{\lambda}\right)
-\]
-where $w_1\in C_0^{\infty}(\bR)$ has zero mean. Then [REFERENCE - WHERE?]
-
-The pressure field at the receiver 
-location $z_r=0$ takes the familiar form  
-\begin{equation}
-\label{eqn:modlohs}
-\tS[\kappa_1]\wl(\xi,t) = \frac{1}{2c_0\sqrt{1-c_0^2\xi^2}}\wl(t) +
-R(\xi)\wl\left(t-2\tau(\xi,z_1)\right) + O(\lambda)
-\end{equation}
-where $\tau$ is the vertical plane wave travel time:
-\begin{equation}
-\label{eqn:pwtt}
-\tau(\xi,z) = \int_0^{z} \sqrt{\frac{1}{c(z)^2}-\xi^2}.
-\end{equation}
-The reflection coefficient $R$ depends on $\xi$ and on
-the values of $\kappa_1$ and $\rho$ near $z=z_1$: the precise form is not
-important for the argument to follow.
-
-The remainder in \ref{eqn:modlohs} is $O(\lambda)$ in the sense of
-$L^2$
 
@@
 be a solution in the weak sense. It follows from the weak solution
-conditions that the pressure is continuous at $p=0$, whence $v$ must
+conditions that the pressure is continuous at $z=z_s$, whence $v$ must
 have a discontinuity.
@@
 Now assume that $\kappa=\kappa_0 \in \bR^+, \rho=\rho_0 \in \bR^+, c_0
-= \sqrt{\kappa_0/rho_0}$ are constant. 
+= \sqrt{\kappa_0/\rho_0}$ are constant. 
 In $z \ne z_s$, the right hand side 
@@
 \[
-p(z,t)=a\left(t \mp \frac{z-z_s}{c_0}\right), \, v(z,t)=\pm b\left(t \mp
-  \frac{z_s}{c_0}\right)
+p(z,t)=a\left(t -\frac{|z-z_s|}{c_0}\right), \, v(z,t)=\mbox{sgn}(z-z_s) b\left(t -
+  \frac{|z-z_s|}{c_0}\right)
 \]
@@
 \[
-\kappa [v]_{z=z_s}\delta(z-z_s) =
-2\frac{\kappa}{c_0}b\delta(z-z_s) = 2a\delta(z-z_s).
+\kappa_0 [v]_{z=z_s}\delta(z-z_s) =
+2\kappa_0 b\delta(z-z_s) = 2c_0 a\delta(z-z_s).
 \] 
 This must in turn equal the RHS of the constitutive law, whence
-$a=w/2$. Thus
+$a=w/(2c_0)$. Thus
 \begin{eqnarray}
 \label{eqn:sol1dptsrc}
-p(z,t) &=& \frac{1}{2}w\left(t - \frac{|z-z_s|}{c_0}\right) \nonumber \\
-v(z,t) &=& \mbox{sgn}(z-z_s)\frac{1}{2\rho_0 c_0}w\left(t -\frac{|z-z_s|}{c_0}\right)
+p(z,t) &=& \frac{1}{2c_0}w\left(t - \frac{|z-z_s|}{c_0}\right) \nonumber \\
+v(z,t) &=& \mbox{sgn}(z-z_s)\frac{1}{2\kappa_0}w\left(t -\frac{|z-z_s|}{c_0}\right)
            \nonumber \\
@@
 \label{eqn:sol1d}
-p(z,t) &=& \frac{1}{2}\int dz_1 f\left(z_1,t - \frac{|z-z_1|}{c_0}\right) \\
-v(z,t) &=& \frac{1}{2\rho_0 c_0} \int dz_1 \mbox{sgn} (z-z_1) f\left(z_1,t - \frac{|z-z_1|}{c_0}\right)\\
+p(z,t) &=& \frac{1}{2c_0}\int dz_1 f\left(z_1,t - \frac{|z-z_1|}{c_0}\right) \\
+v(z,t) &=& \frac{1}{2\kappa_0} \int dz_1 \mbox{sgn} (z-z_1) f\left(z_1,t - \frac{|z-z_1|}{c_0}\right)\\
 \end{eqnarray}
@@
 
-If $\kappa, \rho $ are constant, it is possible to develop explicit
-expressions for the solution of this system. Only $\kappa$ will be
+If $\kappa=\kappa_0, \rho=\rho_0 $ are constant, it is possible to develop explicit
+expressions for the solution of this system. As before, only $\kappa$ will be
 subject to perturbation in any of the problems considered in this
@@
 \[
-  \frac{\partial v}{\partial z}(z,t) = \frac{1}{\rho_0 c_0} f(z,t) 
+  \frac{\partial v}{\partial z}(z,t) = \frac{1}{\kappa_0} f(z,t) 
 \]
@@
   \label{eqn:sol1ddvdz}
- - \frac{1}{2\kappa_0} \int\, dz_2\, \frac{\partial f}{\partial t}\left(z_2,t - \frac{|z-z_2|}{c_0}\right)
+ - \frac{1}{2\kappa_0c_0} \int\, dz_2\, \frac{\partial f}{\partial t}\left(z_2,t - \frac{|z-z_2|}{c_0}\right)
 \end{equation}
@@
 \[
-  \delta p(z,t) =- \frac{1}{2} \int \, dz_1 \, \delta
+  \delta p(z,t) =- \frac{1}{2c_0} \int \, dz_1 \, \delta
   \kappa(z_1)\frac{\partial v}{\partial z}\left(z_1,t - \frac{|z-z_1|}{c_0}\right)
@@
 \[
-  =-\frac{1}{2\rho_0c_0} \int \,dz_1\, \delta
+  =-\frac{1}{2\kappa_0c_0} \int \,dz_1\, \delta
   \kappa(z_1) f \left(z_1,t - \frac{|z-z_1|}{c_0}\right)
@@
   \label{eqn:sol1ddp}
-  + \frac{1}{4\kappa_0} \int\,dz_2 \, \int \,dz_1\,\delta 
+  + \frac{1}{4\kappa_0c_0^2} \int\,dz_2 \, \int \,dz_1\,\delta 
   \kappa(z_1) \frac{\partial f}{\partial t}\left(z_2,t - \frac{|z_2-z_1|+|z_1-z|}{c_0}\right)
@@
 \[
-  \delta p(z,t) = -\frac{\delta \kappa(z_s)}{2\rho_0c_0}w\left(t - \frac{|z-z_s|}{c_0}\right)
+  \delta p(z,t) = -\frac{\delta \kappa(z_s)}{2\kappa_0c_0}w\left(t - \frac{|z-z_s|}{c_0}\right)
 \]
@@
   \label{eqn:sol1ddppw}
-  +\frac{1}{4\kappa_0} \int \,dz_1\,\delta 
+  +\frac{1}{4\kappa_0c_0^2} \int \,dz_1\,\delta 
   \kappa(z_1) \frac{\partial w}{\partial t}\left(t - \frac{|z_s-z_1|+|z_1-z|}{c_0}\right)
@@
 \append{Plane Waves}
+
 Re-write the plane wave system \ref{eqn:awepw} as
@@
 \frac{\partial \tp}{\partial t} + \tk\frac{\partial
-  \tv}{\partial z} &=& (1-c_0^2\xi^2)^{-1}\tf \nonumber\\
+  \tv}{\partial z} &=& (1-c^2\xi^2)^{-1}\tf \nonumber\\
 \rho \frac{\partial \tv}{\partial t} + \frac{\partial \tp}{\partial
@@
 in which $\tk(\xi,z)=(1-c(z)^2\xi^2)^{-1}\kappa(z)$ is a scaled version of
-bulk modulus. Define the vertical plane wave velocity as
-\[
-\tc(\xi,z)=\frac{c(z)}{\sqrt{1-c(z)^2\xi^2}} =
+bulk modulus. The corresponding wave velocity $\tc$ is given by
+\[
+\tc(\xi,z)=\sqrt{\tk/\rho} = \frac{c(z)}{\sqrt{1-c(z)^2\xi^2}} =
 \left(\frac{1}{c(z)^2}-\xi^2\right)^{-1/2}.
 \]
-Comparing the plane wave system \ref{eqn:awepw1} with the 1D system
-\ref{eqn:awe1d}, and using the 1D solution \ref{eqn:sol1d}, obtain
-\begin{eqnarray}
-\label{eqn:solpw}
-\tp(\xi,z,t) &=& \frac{1}{2}(1-c_0^2\xi^2)^{-1}\int dz_s \tf\left(\xi,z_s,t - \frac{|z-z_s|}{\tc_0(\xi)}\right) \nonumber \\
+
+In effect, systen \ref{eqn:awepw1} is a parametrized 1D (space) wave
+system, so the results of the last two appendices apply to the
+homogeneous case, in which $\kappa=\kappa_0$ and $\rho=\rho_0$ are
+independent of $z$. Comparing the plane wave system \ref{eqn:awepw1} with the 1D system
+\ref{eqn:awe1d}, and using the 1D solution for homogeneous fluids
+\ref{eqn:sol1d}, obtain
+\begin{eqnarray*}
+\tp(\xi,z,t) &=& \frac{1}{2\tc_0(\xi)}(1-c_0^2\xi^2)^{-1}\int dz_s
+                 \tf\left(\xi,z_s,t -
+                 \frac{|z-z_s|}{\tc_0(\xi)}\right) \nonumber \\
 \tv(\xi,z,t) &=& \frac{1}{2\rho_0 c_0}(1-c_0^2\xi^2)^{-1} \int dz_s \mbox{sgn} (z-z_s)\tf\left(\xi,z_s,t - \frac{|z-z_s|}{\tc_0(\xi)}\right)
            \nonumber \\
-\end{eqnarray}
+\end{eqnarray*}
+which becomes
+\begin{eqnarray}
+\label{eqn:solpw}
+ \tp(\xi,z,t)   &=&\frac{1}{2c_0(1-c_0^2\xi^2)^{1/2}}\int dz_s 
+                 \tf\left(\xi,z_s,t -
+                 \frac{|z-z_s|}{\tc_0(\xi)}\right) \nonumber \\
+\tv(\xi,z,t) &=& \frac{1}{2\kappa_0} \int dz_s \mbox{sgn} (z-z_s)\tf\left(\xi,z_s,t - \frac{|z-z_s|}{\tc_0(\xi)}\right)
+                 \nonumber \\
+\end{eqnarray}
+after some simplification.
 
@@
 \label{eqn:fwdpw1}
-= \frac{1}{2}(1-c_0^2\xi^2)^{-1}\int dz_s \tf\left(\xi,z_s,t -
-  \frac{|z-z_s|}{\tc_0(\xi)}\right)
-\end{equation}
-
-If $\tf(\xi,z,t)=\tw(\xi,t)\delta(z)$ is a surface
-extended source, plane wave version, this is equivalent to replacing
-$\tw(\xi,t)$ by $(1-c(0)^2\xi^2)^{-1}\tw(\xi,t)$.
-
-Since the system \ref{eqn:awepw1} has the form of the 1D system
-\ref{eqn:awe1d} for each $\xi $, both $\tS[\kappa_0] $ and
-$D\tS[\kappa_0]$ can be computed from the corresponding 1D
-results. From \ref{eqn:sol1d},
-\[
-\tS[\kappa_0]\tw(\xi,t)=\frac{1}{2}(1-c_0^2\xi^2)^{-1}\tw(\xi,t)
-\]
-which is equation \ref{eqn:modhom}.
-
-From \ref{eqn:fwd1d},
-\[
-D(\tS[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa = \int_0^t\,ds\,(1-c_0^2\xi^2)\tw'(\xi,t-s)\frac{\delta 
-  \tk \left(\frac{\tc s}{2}\right)}{8\tk} 
-\]
+= \frac{1}{2c_0(1-c_0^2\xi^2)^{1/2}}\int dz \tf\left(\xi,z,t -
+  \frac{|z_r-z|}{\tc_0(\xi)}\right)
+\end{equation}
+which replicates equation \ref{eqn:stmodhompw}. Equations
+\ref{eqn:ssmodhompw} and \ref{eqn:ptmodhompw} follow by simple specializations.
+
+The linearized forward map $D\tS[\kappa]$ is defined by
+\begin{equation}
+  \label{eqn:dfwdpw}
+  D\tS[\kappa]\delta \kappa(\xi,t) = \delta \tp(\xi,z_r,t)
+\end{equation}
+Here $\tp$ and $\tv$ solve the linearized plane-wave system
+\begin{eqnarray}
+\label{eqn:dawepw1}
+\frac{\partial \delta\tp}{\partial t} + \tk\frac{\partial
+  \delta \tv}{\partial z} +  \delta \tk\frac{\partial
+  \delta \tv}{\partial z}  &=& (1-c^2\xi^2)^{-2}\delta (c^2)\xi^2\tf \nonumber\\
+\rho \frac{\partial \delta \tv}{\partial t} + \frac{\partial \delta \tp}{\partial
+  z}&=&0\nonumber\\
+\delta \tp,\delta \tv&=&0, t \ll 0,
+\end{eqnarray}
+
+Comparing \ref{eqn:dawepw1} with identity
+\ref{eqn:dawe1d}, observe that the solution of \ref{eqn:dawepw1} for is
+the sum of
+\begin{itemize}
+\item the solution of \ref{eqn:dawe1d} with $\tk, \delta \tk$
+  substituted for $\kappa, \delta \kappa$, and the right hand side in
+  the first equation of the system \ref{eqn:awe1d} for
+  the reference fields $\tp, \tv$ replaced by $(1-c^2\xi^2)^{-1}\tf$;
+\item the solution of \ref{eqn:awe1d} with $\tf$ replaced by
+  $(1-c^2\xi^2)^{-2}\delta (c^2)\xi^2\tf$, and $\kappa_0$ replaced by $\tk_0$..
+\end{itemize}
+
+Expressions for linearized forward maps at homogeneous reference
+models, $D\tS[\kappa_0]\delta \kappa$, follow this observation and the
+results of the last two appendices. For the first summand, in
+\ref{eqn:sol1ddp}, substitute $\tk_0$ for $\kappa_0$, $\delta \tk$ for
+$\delta \kappa$, and $(1-c_0^2\xi^2)\tf$ for $\tf$ to obtain
+\[
+-\frac{1}{2\tk_0(\xi)\tc_0(\xi)} \int \,dz_1\, \delta 
+\tk(\xi,z_1)(1-c_0\xi^2)^{-1} \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right)
+\]
+\[
+  + \frac{1}{4\tk_0(\xi)\tc_0(\xi)} \int\,dz_2 \, \int \,dz_1\,\delta 
+  \tk(\xi,z_1) (1-c_0\xi^2)^{-1}\frac{\partial \tf}{\partial t}\left(\xi,z_2,t - \frac{|z_2-z_1|+|z_1-z|}{\tc_0(\xi)}\right) 
+\]
+Since
+\[
+\delta \tk = \delta \left(\frac{\kappa}{1-\kappa \xi^2/\rho}\right) = \delta \kappa (1-c^2\xi^2)^{-2}
+\]
+this amounts to
+\[
+=  -\frac{1}{2\kappa_0c_0 1-c_0\xi^2)^{3/2}} \int \,dz_1\, \delta 
+\kappa(z_1) (1-c_0^2\xi^2)^{-2} \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right) 
+\]
+\begin{equation}
+  \label{eqn:solpwdp1}
+  + \frac{1}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int\,dz_2 \, \int \,dz_1\,\delta 
+  \kappa(z_1) (1-c_0^2\xi^2)^{-2}  \frac{\partial \tf}{\partial t}\left(\xi,z_2,t - \frac{|z_2-z_1|+|z_1-z|}{\tc_0(\xi)}\right) 
+\end{equation}
+
 Note that
 \[
-\delta \tk = \delta \left(\frac{\kappa}{1-\kappa \xi^2/\rho}\right) = 
-\]
-\[
-\delta \kappa (1-c^2\xi^2)^{-2}
-\]
+  (1-c^2\xi^2)^{-2}\delta (c^2)\xi^2= (1-c^2\xi^2)^{-2}\delta \kappa
+  /\rho
+\]
+Therefore, from \ref{eqn:sol1d}, the second summand identified above
+is
+\[
+  =\frac{1}{2\tc(\xi)\rho_0}\int \,dz_1\, \delta 
+\kappa(z_1)(1-c_0^2\xi^2)^{-2} \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right)
+\]
+\begin{equation}
+  \label{eqn:solpwdp2}
+=\frac{c_0^2\xi)^2}{2\kappa_0c_0(1-c_0^2\xi^2)^{3/2}}\int \,dz_1\, \delta 
+\kappa(z_1) \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right)
+\end{equation}
+Adding the quantities \ref{eqn:solpwdp1} and \ref{eqn:solpwdp2},
+obtain
+\[
+  \delta \tp(\xi,z,t) = -\frac{1}{2\kappa_0c_0(1-c_0^2\xi^2)^{1/2}}\int \,dz_1\, \delta 
+  \kappa(z_1) \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right)
+  \]
+\begin{equation}
+  \label{eqn:solpwdp}
+  + \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int\,dz_2 \, \int \,dz_1\,\delta 
+  \kappa(z_1)  \frac{\partial \tf}{\partial
+    t}\left(\xi,z_2,t - \frac{|z_2-z_1|+|z_1-z|}{\tc_0(\xi)}\right)
+  \end{equation}
 Consequently 
@@
   D(\tS[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
-\int_0^t\,ds\,\tw'(\xi,t-s)\frac{\delta \kappa  \left(\frac{\tc 
-      s}{2}\right)}{8\kappa_0 (1-c_0^2\xi^2)^2} 
-\]
-\[
-= \frac{c_0}{2}\int_0^t\,ds\,\tw(\xi,t-s)\frac{(\delta \kappa)'  \left(\frac{\tc 
-      s}{2}\right)}{8\kappa_0 (1-c_0^2\xi^2)^{5/2}} 
-\]
-which is equation \ref{eqn:dmodhom}.
+  -\frac{1}{2\kappa_0c_0(1-c_0^2\xi^2)^{1/2}}\int \,dz_1\, \delta 
+  \kappa(z_1) \tf \left(\xi,z_1,t - \frac{|z_r-z_1|}{\tc_0(\xi)}\right)
+\]
+\begin{equation}
+  \label{eqn:dmodhomst1}
+    + \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int\,dz_2 \, \int \,dz_1\,\delta 
+  \kappa(z_1)  \frac{\partial \tf}{\partial
+    t}\left(\xi,z_2,t - \frac{|z_2-z_1|+|z_1-z_r|}{\tc_0(\xi)}\right)
+\end{equation}
+which is equation \ref{eqn:dmodhompwst}. For a point source
+$\tf(\xi,z,t) = w(t)\delta(z-z_s)$, this reduces to
+\[
+  D(\tS[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
+  -\frac{1}{2\kappa_0c_0(1-c_0^2\xi^2)^{1/2}} \delta 
+  \kappa(z_s) w\left(t - \frac{|z_r-z_s|}{\tc_0(\xi)}\right)
+\]
+\begin{equation}
+  \label{eqn:dmodhompt1}
+    + \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int \,dz\,\delta 
+  \kappa(z) \frac{\partial w}{\partial
+    t}\left(t - \frac{|z_s-z|+|z-z_r|}{\tc_0(\xi)}\right)
+\end{equation}
+which is equation \ref{eqn:dmodhompwpt}.
 
@@
 
+\append{OLD RESULTS}
+
+The underlying hypothesis is that the data comes from a point
+source. Extension methods encode such information in an annihilator,
+an operator that vanishes for models that satisfy the hypothesis. For
+SSE-FWI, a natural annilator is a variant on the operator introduced
+for this purpose by \cite{HuangSymes:Geo18a,HuangSymes:Geo18b} for two
+similar source extension methods, namely multiplication by offset,
+also reminiscent of a now-standard annihilator for subsurface offset
+introduced by \cite{stolk2001c}. In the plane wave domain,
+multiplication by offset becomes 
+\[
+\tA = D_{\xi}I_t,
+\]
+that is, differentiation in $\xi$ and integration in $t$, in either
+order.
+
+The forward map $\tS$ for plane-wave transformed inverse problem maps
+extended plane wave source wavelets $\tw(\xi,t)$ to plane wave data
+traces (sampling of the plane-wave field at $z=0$, assuming this to be the
+depth of sources and receivers, as I do here). I will regard this operator as a
+function of the bulk modulus $\kappa(z)$, by assuming the density
+$\rho$ to be fixed and homogeneous.
+
+$\tS[\kappa]$ is 
+actually invertible, at least for $\kappa$ close to homogeneous. This conclusion follows for instance
+from explicit expressions in two special cases: for homogeneous models
+$\kappa=\kappa_0$, 
+\[
+\tS[\kappa_0]\tw(\xi,t) = a_0(\xi)\tw(\xi,t),
+\]
+and for a smooth model $\kappa_1$ with an embedded reflector at $z=z_1$,
+
+\[
+\wl(t)=\frac{1}{\sqrt{\lambda}}w_1\left(\frac{t}{\lambda}\right)
+\]
+is a family of wavelets depending on central wavelength $\lambda$. The
+vertical travel time $\tau(\xi,z_1)$ is given by
+\[
+\tau(\xi,z_1) = \int_0^{z_1}\sqrt{\frac{1}{c^2}-\xi^2},\,\,c=\sqrt{\kappa/rho}.
+\]
+
+Since $\tS$ is invertible, it is possible to use the
+simplified objective presented by
+\cite{Symes:EAGE15}:
+\begin{equation}
+\label{eqn:dsobj}
+J_0[\kappa]=\frac{1}{2}\|\tA \tS[\kappa]^{-1}\td\|^2
+\end{equation}
+A derivation of this limit case from the standard penalty formulation
+of extended FWI 
+appears in the last section. Inversion based on this type of objective
+has been called ``inversion velocity analysis'' in the subsurface
+offset context \cite[]{Herve2017},\cite[]{HouSymes:Geo18}, and is also the
+basis of Adaptive Waveform Inversion
+\cite{Warner:14,Warner:16}.
+
+The main result pertains to the gradient at homogeneous
+$\kappa=\kappa_0$ for data resulting from a single flat reflector
+embedded in a smooth background:
+\[
+\tdl = \tS[\kappa_1]\wl
+\]
+I calculate the gradient of $J_0$ in this configuration.
+The result is given in equation \ref{eqn:jgrad} in the last
+section. Converted to spatial frequency via the Fourier transform, it
+becomes
+\begin{equation}
+\label{eqn:jgradft-intro}
+\widehat{\nabla J[\kappa_0]}(k) = -\frac{c_0}{16}\int\, d\xi\,a_1(\xi) 
+\frac{1}{ik}\hat{u}_{\lambda}(\xi,k)D_{\xi}^2\overline{\hat{u}_{\lambda}(\xi,k)}
+\end{equation}
+where
+\[
+\hat{u}_{\lambda}(\xi,k) = \left(1 +
+  \frac{R(\xi)}{a_0(\xi)}e^{4ik\tau(\xi,z_1)/\tc(\xi)}\right)\hat{\wl}\left(\frac{2k}{\tc(\xi)}\right) 
+\]
+and 
+\[
+\tc(\xi) = \left(\frac{1}{c_0}^2-\xi^2\right)^{-1/2}
+\]
+is the vertical plane wave velocity for homogenous wave velocity
+$c_0 = \sqrt{\kappa_0/rho}$.
+
+Then for any $k_{\rm max} >0$, $\xi_{\rm max} > 0$,
+\[
+\max \{\left|\hat{w}_{\lambda}\left(2k/\tc{\xi}\right)\right|:
+|k|\le k_{\rm max}, |\xi|\le \xi_{\rm max} \} \rightarrow 0\mbox{ as }
+\lambda \rightarrow 0.
+\]
+It follows immediately from equation \ref{eqn:jgradft-intro} that the
+same is true of the gradient:
+\begin{equation}
+\label{eqn:bigdeal}
+\max \{\left|\widehat{\nabla J[\kappa_0]}(k)\right|:
+|k|\le k_{\rm max}, |\xi|\le \xi_{\rm max} \} \rightarrow 0\mbox{ as }
+\lambda \rightarrow 0.
+\end{equation}
+
+Therefore the rate at which traveltime changes for any plane wave
+component in the specified range, as the bulk modulus is changed along
+the direction of the negative gradient, is vanishingly small as
+$\lambda \rightarrow 0$.
+
+Otherwise put, for high-frequency reflection data, this method cannot
+update the velocity away from a homogenous initial guess in a way that
+is kinematically constructive.
+
+It may even be possible to show that the update may be destructive, by
+more careful examination of the gradient approximation. However, who
+cares, really: it is established that SSE-FWI has similar behaviour to
+FWI for at least one simple instance of reflection data.
 \bibliographystyle{seg}

### 2020-01-23 repro/papers/sserefl/paper.tex
finished WRI trans comp

@@
 domain of $\tS[\kappa]$. The simplest appears to be the assumption
-that all plane wave sources act over a limited depth range, $0 \le z
+that all plane wave sources act over a limited depth range, $z_{\rm min} \le z
 \le z_{\rm max}$. Then the $z$ integrals appearing in the expressions
-for $S$ and $\tS$ have limits $[0,z_{\rm max}]$
+for $S$ and $\tS$ have limits $[z_{\rm min},z_{\rm max}]$
 
@@
 \[
-  \tS[\kappa]\tS[\kappa]^T\te(\xi,t) = \frac{1}{2}(1-c_0^2\xi^2)^{-1}
+  \tS[\kappa_0]\tS[\kappa_0]^T\te(\xi,t) = \frac{1}{2}(1-c_0^2\xi^2)^{-1}
 \]
@@
   \times 
-  \int_0^{z_{\rm max}} dz\, \frac{1}{2}(1-c_0^2\xi^2)^{-1}\te\left(\xi,t' +
+  \int_{z_{\rm min}}^{z_{\rm max}} dz\, \frac{1}{2}(1-c_0^2\xi^2)^{-1}\te\left(\xi,t' +
     \frac{|z_r-z|}{\tc_0(\xi)}\right)_{t'=t-\frac{|z_r-z|}{\tc_0(\xi)}}
@@
   \label{eqn:stmodhompwnorm}
- = \frac{z_{\rm max}}{4}(1-c_0^2\xi^2)^{-2}\te(\xi,t)
+ = \frac{z_{\rm max}-z_{\rm min}}{4}(1-c_0^2\xi^2)^{-2}\te(\xi,t) 
 \end{equation}
@@
 
-\paragraph{Transmission:} $z_s < z_r$, $\kappa_*$ is smooth and $ = \kappa_0$ for
-$z$ near $z_s$, $\ge \kappa_0$ for all $z$, and $>\kappa_0$ for $z$
-near $z_r$, Geometric acoustics implies that the solution is given by
-the progressing wave expansion [REFERENCE] in terms of the vertical
-travel time $\tau(\xi,z_1,z_2)$ and geometric amplitude
-$a(\xi,z_1,z_2)$ between depths $z_1$ and $z_2$. The traveltime is given by
-\[
-\tau(\xi,z_1,z_2) = \int_{z_1}^{z_2}\,dz\, \frac{1}{\tc(\xi,z)}.
+\paragraph{Transmission:} $z_s < z_r$, $\kappa_*$ is smooth and and
+differs from $\kappa_0$ only between $z_s$ and $z_r$: that is, $\kappa = \kappa_0$ for
+$z < z_s+\epsilon$, $z>z_r-\epsilon$, suitable $\epsilon>0$. To ensure
+that the data computed from $\kappa_*$ is distinct from the reference
+data computed from $\kappa_0$, also assume that $\kappa \ge \kappa_0$,
+and that $\kappa > \kappa_0$ over some subinterval of $[z_s,z_r]$ of
+positive length.
+
+Geometric acoustics implies that the solution for $\kappa_*$ is
+asymptotic to the first term of the progressing wave expansion
+[REFERENCE], which depends on the vertical
+travel time $\tau_*(\xi,z_1,z_2)$ and geometric amplitude
+$a_*(\xi,z_1,z_2)$ between depths $z_1$ and $z_2$. The traveltime is given by
+\[
+\tau(\xi,z_1,z_2) = \int_{z_1}^{z_2}\,dz\, \frac{1}{\tc_*(\xi,z)}.
 \]
 I do not give an explicit expression for the amplitude, noting only
-that it is smooth in all three variables and positive. The geometric
+that it is smooth in all three variables and positive.
+
+The geometric
 acoustics approximation of the data is
@@
   \label{eqn:geosoltrans}
-  \tdl(\xi,t) = a(\xi,z_s,z_r)\wl(t-\tau(\xi,z_s,z_r)).
+  \tdl(\xi,t) = a(\xi,z_s,z_r)\wl(t-\tau(\xi,z_s,z_r)) + O(\lambda). 
 \end{equation}
@@
 \end{equation}
+The vertical travel time for $\kappa=\kappa_0$ is
+\[
+  \tau_0(\xi,z_1,z_2) = \frac{z_2-z_1}{\tc_0(\xi)}.
+\]
+A consequence of the assumptions about the relation between $\kappa_*$
+and $\kappa_0$, stated above, is that $\tau_*(\xi,z_s,z_r) <
+\tau_0(\xi,z_s,z_r)$ for $|\xi|\le \xi_{\rm max}$.
 
@@
 is
-\[
-  \nabla J_{\rm FWI}[\kappa] = D\tS[\kappa]^T(\tS[\kappa]\tw_*-d)
-\]
-
+\begin{equation}
+  \label{eqn:fwigrad}
+  \nabla J_{\rm FWI}[\kappa] = D\tS_p[\kappa]^T(\tS_p[\kappa]\wl-\tdl)
+\end{equation}
+The first step in this computation is 
 For homogeneous background model $\kappa=\kappa_0$, Appendix \appPlane
@@
 \[
-D(\tS_p[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
+D(\tS_p[\kappa_0]\wl)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
   -\frac{1}{2\kappa_0c_0(1-c_0^2\xi^2)^{1/2}} \delta 
-  \kappa(z_s) w\left(t - \frac{|z_r-z_s|}{\tc_0(\xi)}\right)
+  \kappa(z_s) \wl\left(t - \frac{|z_r-z_s|}{\tc_0(\xi)}\right)
 \]
@@
 \label{eqn:dmodhompwpt}
-    + \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int \,dz\,\delta 
-  \kappa(z) \frac{\partial w}{\partial
+    + \frac{1}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int \,dz\,\delta 
+  \kappa(z) \frac{\partial \wl}{\partial
     t}\left(t - \frac{|z_s-z|+|z-z_r|}{\tc_0(\xi)}\right)
@@
   \label{eqn:dmodhompwpt0}
-  D(\tS_p[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
-     \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int \,dz\,\delta 
-  \kappa(z)  \frac{\partial w}{\partial
+  D(\tS_p[\kappa]\wl)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
+     \frac{1}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int \,dz\,\delta 
+  \kappa(z)  \frac{\partial \wl}{\partial 
     t}\left(t - \frac{|z_s-z|+|z-z_r|}{\tc_0(\xi)}\right). 
 \end{equation}
+Furthermore, $\delta \kappa = 0$ for $z>z_r$ as well, which means that
+the integral may be restricted to the interval $z_s \le z \le
+z_r$. Thus in the range of integration, $|z_s-z| + |z_r-z|=z_r-z_s$
+whence equation \ref{eqn:dmodhompwpt0} may be rewritten as
+\begin{equation}
+  \label{eqn:dmodhompwpt1}
+  D(\tS_p[\kappa]\wl)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
+     \frac{1}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \frac{\partial \wl}{\partial 
+    t}\left(t - \frac{z_r-z_s}{\tc_0(\xi)}\right) \left(\int_{z_s}^{z_r} \,dz\,\delta 
+  \kappa(z) \right). 
+\end{equation}
+Note that the linearized transmission seismogram is only sensitive to
+the average of $\delta \kappa$ between $z_s$ and $z_r$, which is in
+turn proportional by a $\xi$-dependent factor to the perturbation in
+travel time between source and receiver. Accordingly, the adjoint of
+the linearized forward map outputs a $z$-independent bulk modulus perturbation:
+\begin{equation}
+\label{eqn:dmodhompwptadj}
+D(\tS[\kappa]\wl)_{\kappa=\kappa_0}^T\td(z) =
+\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\,\int\,dt\,
+\frac{1}{4\kappa_0c_0^2(1-c_0^2\xi^2)}
+\frac{\partial w}{\partial
+    t}\left(t - \frac{z_r-z_s}{\tc_0(\xi)}\right) \td(\xi,t).
+\end{equation}
+
+To compute the gradient \ref{eqn:fwigrad} at the homogeneous bulk
+modulus $\kappa_0$, recall that the target data is
+given by \ref{eqn:geosoltrans}, and the trial data for
+$\kappa=\kappa_0$ by \ref{eqn:analsoltrans}. Insert these into the
+right hand side of the identity \ref{eqn:fwigrad}:
+\[
+  \nabla J_{\rm FWI}[\kappa_0](z) =
+  D(\tS[\kappa]\wl)_{\kappa=\kappa_0}^T(a_0\wl(t-\tau_0)
+  -a\wl(t-\tau))(z) + O(\lambda)
+\]
+\[
+=  \int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\,\int\,dt\,
+\frac{1}{4\kappa_0c_0^2(1-c_0^2\xi^2)}
+\frac{\partial \wl}{\partial
+    t}(t - \tau_0(\xi,z_s,z_r))
+  \left(a_0(\xi,z_s,z_r)\wl(t-\tau_0(\xi,z_s,z_r) )\right.
+\]
+\[
+-
+\left.  a(\xi,z_s,z_r)\wl(t-\tau(\xi,z_s,z_r))\right) + O(\lambda)
+\]
+\[
+  = \int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\,
+\frac{1}{4\kappa_0c_0^2(1-c_0^2\xi^2)}\left[a_0(\xi,z_s,z_r) \int \,
+  dt \, \left(\frac{\partial \wl}{\partial t}\wl  \right)
+  (t-\tau_0(\xi,z_s,z_r))
+\right.
+\]
+\[
+  \left.
+- a(\xi,z_s,z_r) \int\,dt\,\frac{\partial \wl}{\partial t} 
+  (t-\tau_0(\xi,z_s,z_r)) \wl(t-\tau(\xi,z_s,z_r)\right] + O(\lambda)
+\]
+The first of the two $t$ integrals vanishes because the integrand is
+an exact derivative (of $\wl^2/2$) . The second vanishes when
+$\lambda$ is small enough that the support of $\wl$ is smaller than
+half of $\tau_0(\xi,z_s,z_r)-\tau(\xi,z_s,z_r)$ for $\xi$ in the
+interval of integration. Therefore we conclude that
+\[
+  \nabla J_{\rm FWI}[\kappa_0] = O(\lambda)
+\]
+as stated in the last section.
+
+\subsection{WRI, Transission Configuration}
+In this configuration, $z_{\rm min}=z_s < z_r=z_{\rm max}$ and
+\begin{equation}
+  \label{eqn:writransupp}
+  \mbox{supp }\tf \subset [-\xi_{\rm max},\xi_{\rm max}]\times
+  [z_s,z_r]\times \bR.
+\end{equation}
+Recall that (equation \ref{eqn:wripwiva1} with $w_*=\wl$)
+\begin{equation}
+  \label{eqn:wripwiva1}
+  J_{\rm WRI}[\kappa]   = \frac{1}{2}\langle \td -
+  \tS_p[\kappa]\wl, (\tS[\kappa]\tS[\kappa]^T)^{-1}( \td -
+  \tS[\kappa]\wl) \rangle  
+\end{equation} 
 Thus
-\begin{equation}
-\label{eqn:dmodhomadj}
-D(\tS[\kappa]\tw)_{\kappa=\kappa_0}^T\td(z) =
-\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\,\int\,dt\,
-\frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)}
-\frac{\partial w}{\partial
-    t}\left(t - \frac{|z_s-z|+|z-z_r|}{\tc_0(\xi)}\right) \td(\xi,t).
+\[
+  D J_{\rm WRI}[\kappa]\delta \kappa|_{\kappa=\kappa_0}   =
+  DJ_1[\kappa]\delta \kappa|_{\kappa=\kappa_0} +
+  DJ_2[\kappa]\delta \kappa|_{\kappa=\kappa_0},
+\]
+in which
+\[
+  J_1[\kappa]   = \frac{1}{2}\langle \td -
+  \tS_p[\kappa]\wl, (\tS[\kappa_0]\tS[\kappa_0]^T)^{-1}( \td -
+  \tS_p[\kappa]\wl) \rangle 
+\]
+and
+\[
+ J_2[\kappa] =  \frac{1}{2}\langle \td -
+  \tS_p[\kappa_0]\wl, (\tS[\kappa]\tS[\kappa]^T)^{-1}( \td -
+  \tS_p[\kappa_0]\wl) \rangle
+\]
+Recall that 
+\[
+  \tS[\kappa_0]\tS[\kappa_0]^T\te(\xi,t) = \frac{z_{\rm max}-z_{\rm min}}{4}(1-c_0^2\xi^2)^{-2}\te(\xi,t).
+\]
+(equation \ref{eqn:stmodhompwnorm}). Thus $J_1$ is a statically
+eqweighted version of $J_{\rm FWI}$. The estimation of $D J_{\rm
+  FWI}[\kappa]\delta \kappa|_{\kappa=\kappa_0}$ follows exactly the
+same pattern as the 
+estimation of $D J_{\rm
+  FWI}[\kappa]\delta \kappa|_{\kappa=\kappa_0}$ in the last section,
+and yields
+\begin{equation}
+  \label{eqn:j1est}
+  \nabla J_1[\kappa_0] = O(\lambda).
+\end{equation}
+
+To estimate $\nabla J_2[\kappa_0]$, abbreviate
+\[
+  \te=\td -
+  \tS_p[\kappa_0]\wl 
+\]
+and note that
+\[
+  DJ_2[\kappa]\delta \kappa|_{\kappa=\kappa_0} = -
+  \frac{1}{2}\langle \te,
+  (\tS[\kappa_0]\tS[\kappa_0]^T)^{-1}\left((D\tS[\kappa]\delta
+  \kappa)_{\kappa=\kappa_0}\tS[\kappa_0]^T\right.
+\]
+\[
+  \left.+\tS[\kappa_0](D\tS[\kappa]\delta
+  \kappa)_{\kappa=\kappa_0}^T\right) (\tS[\kappa_0]\tS[\kappa_0]^T)^{-1} \te\rangle
+\]
+\begin{equation}
+  \label{eqn:dj2}
+ = -\langle \te,(\tS[\kappa_0]\tS[\kappa_0]^T)^{-1}
+  ((D\tS[\kappa]\delta
+  \kappa)_{\kappa=\kappa_0}\tS[\kappa_0]^T)
+  (\tS[\kappa_0]\tS[\kappa_0]^T)^{-1}\te\rangle
+\end{equation}
+Recall that $\mbox{supp }(\tf) \subset \{z_s \le z \le
+z_r\}$, that is, the domain of $\tS$ is actually
+$L^2([-\xi_{\rm max},\xi_{\rm max}] \times [z_s,z_r] \times
+\bR)$. Injection of this domain into $L^2(\bR^3)$ is equivalent to
+placing limits on the various integrals that appear in the definitions
+of $\tS$ and related operators. For the adjoint, the dual projection
+is part of the definition:
+\begin{equation}
+  (\tS[\kappa_0]^T\te)(\xi,z,t)=\frac{1}{2}(1-c_0^2\xi^2)^{-1}\te\left(\xi,t +
+    \frac{|z_r-z|}{\tc_0(\xi)}\right), \,|\xi|\le \xi_{rm max}, z_s
+  \le z \le z_r, \, =0 \mbox{ else}.
+\end{equation}
+(equation \ref{eqn:stmodhompwtrans}). From equation \ref{eqn:dmodhomst1app},
+\[
+  D(\tS[\kappa_0]\tf)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
+  -\frac{1}{2\kappa_0c_0(1-c_0^2\xi^2)^{1/2}}\int \,dz_1\, \delta 
+  \kappa(z_1) \tf \left(\xi,z_1,t - \frac{|z_r-z_1|}{\tc_0(\xi)}\right)
+\]
+\[
+    + \frac{1}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int\,dz_2 \, \int \,dz_1\,\delta 
+  \kappa(z_1)  \frac{\partial \tf}{\partial
+    t}\left(\xi,z_2,t - \frac{|z_2-z_1|+|z_1-z_r|}{\tc_0(\xi)}\right)
+\]
+So
+\[
+ ( D\tS[\kappa]\delta \kappa) \tS^T[\kappa]\te(\xi,t) =
+  -\frac{1}{4\kappa_0c_0}(1-c_0^2\xi^2)^{-3/2} \te(\xi,t)
+  \int_{z_s}^{z_r}\,dz\,\delta \kappa (z)
+\]
+\begin{equation}
+  \label{eqn:dmodhompwst2}
+     + \frac{1}{8\kappa_0c_0^2(1-c_0^2\xi^2)^2} \int_{z_s}^{z_r}\,dz_2 \, \int_{z_s}^{z_r} \,dz_1\,\delta 
+  \kappa(z_1)  \frac{\partial \te}{\partial
+    t}\left(\xi,t - \frac{|z_2-z_1|+|z_1-z_r|-|z_r-z_2|}{\tc_0(\xi)}\right)
+\end{equation}
+The integral in the secont term splits naturally into the subdomains
+\begin{eqnarray*}
+ z_s \le z_2 \le z_1 \le z_r & \Rightarrow &
+                                            |z_2-z_1|+|z_1-z_r|-|z_r-z_2|
+                                            = z_1-z_2 + z_r-z_1 -
+                                           ( z_r-z_2) = 0\\
+z_s \le z_1 \le z_2\le z_r  & \Rightarrow
+                                          &|z_2-z_1|+|z_1-z_r|-|z_r-z_2|
+                                            = z_2-z_1 + z_r-z_1 -
+                                            (z_r-z_2) = 2(z_2-z_1)
+\end{eqnarray*}
+So
+\[
+  \int_{z_s}^{z_r} \,dz_2\, \frac{\partial \te}{\partial 
+    t}\left(\xi,t -
+    \frac{|z_2-z_1|+|z_1-z_r|-|z_r-z_2|}{\tc_0(\xi)}\right) 
+\]
+\[
+  = \int_{z_1}^{z_r} \,dz_2\, \frac{\partial
+    \te}{\partial t}\left(\xi,t-\frac{2(z_2-z_1)}{\tc_0(\xi)}\right)
+  + (z_1-z_s) \frac{\partial
+    \te}{\partial t}(\xi,t)
+\]
+\[
+=-\frac{\tc_0}{2}\int_{z_1}^{z_r} \,dz_2\, \frac{\partial}{\partial z_2}\te\left(\xi,t-\frac{2(z_2-z_1)}{\tc_0(\xi)}\right)
+  + (z_1-z_s) \frac{\partial
+    \te}{\partial t}(\xi,t)
+\]
+\[
+  = -\frac{\tc_0}{2}\left(\te\left(\xi,t-\frac{2(z_r-z_1)}{\tc_0(\xi)}\right)-\te(\xi,t)\right)
+    + (z_1-z_s) \frac{\partial
+    \te}{\partial t}(\xi,t)
+\]
+so 
+\[
+  ( D\tS[\kappa]\delta \kappa) \tS^T[\kappa]\te(\xi,t) =
+  \frac{1}{4\kappa_0c_0}\left(\frac{1}{4}(1-c_0^2\xi^2)^{-5/2}- (1-c_0^2\xi^2)^{-3/2}\right) \te(\xi,t)
+  \int_{z_s}^{z_r}\,dz\,\delta \kappa (z)
+\]
+\begin{equation}
+  \label{eqn:dmodhompwst3}
+     + \frac{1}{8\kappa_0c_0^2(1-c_0^2\xi^2)^2}
+     \int_{z_s}^{z_r}\,dz_1\,\delta\kappa(z_1) \left((z_1-z_s) \frac{\partial
+    \te}{\partial t}(\xi,t) -\frac{\tc_0}{2}\te\left(\xi,t-\frac{2(z_r-z_1)}{\tc_0(\xi)}\right)\right)
+\end{equation}
+Therefore 
+\[
+  DJ_2[\kappa]\delta \kappa|_{\kappa=\kappa_0} = \left(\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi \,W_0(\xi)
+  \int\,dt\, \te(\xi,t)^2\right)\int_{z_s}^{z_r}\,dz\,\delta \kappa
+\]
+\[
+     -\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\,W_1(\xi)\int\,dt\, \int_{z_s}^{z_r}\,dz_1\,\delta\kappa(z_1) \left((z_1-z_s) \te(\xi,t)\frac{\partial 
+         \te}{\partial t}(\xi,t) \right.
+     \]
+\begin{equation}
+  \label{eqn:dj2e1}
+ \left.    -\frac{\tc_0}{2}\te(\xi,t)\te\left(\xi,t-\frac{2(z_r-z_1)}{\tc_0(\xi)}\right)\right)
+\end{equation}
+In the right-hand side of equation \ref{eqn:dj2e1}, $W_0$ and $W_1$
+are functions of $\xi$, parametrized by $\kappa_0$ and $\rho_0$. They
+have no depedence on the target model $\kappa_*,\rho_*$ used to
+generate the target data $d$.
+
+Note that the first summand under the integral sign in the second term
+integrates to zero in $t$, as it is an exact derivative of a smooth
+compactly supported function. Therefore combining equations
+\ref{eqn:j1est} and \ref{eqn:dj2e1} and the definition of the formal
+or $L^2$ gradient, obtain
+\[
+  \nabla J_{\rm WRI}[\kappa_0](z) = \left(\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi \,W_0(\xi)
+    \int\,dt\, \te(\xi,t)^2\right) {\bf 1}_{[z_s,z_r]}(z)
+\]
+\begin{equation}
+  \label{eqn:gradwri1}
+  +\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\,W_2(\xi)\int\,dt\, 
+  \te(\xi,t)\te\left(\xi,t-\frac{2(z_r-z)}{\tc_0(\xi)}\right) + O(\lambda)
+\end{equation}
+in which $W_2 = (\tc_0/2)W_1$.
+
+The inner ($t$) integral in the second term on the RHS of
+\ref{eqn:gradwri1} is the autocorrelation of $\te$ at lag
+$2(z_r-z)/\tc_0(\xi)$. Introducing the Fourier transform $\hat{\te}$
+of $\te$ in $t$,
+\[
+  \nabla J_{\rm WRI}[\kappa_0](z) = \left(\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi \,W_0(\xi)
+    \int\,dt\, \te(\xi,t)^2\right) {\bf 1}_{[z_s,z_r]}(z)
+\]
+\begin{equation}
+  \label{eqn:gradwri2}
+  +\frac{1}{2\pi}\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\,W_2(\xi)\int\,d\omega\, 
+  |\hat{\te}(\xi,\omega)|^2e^{-i\omega\frac{2(z_r-z)}{\tc_0(\xi)}} + O(\lambda)
+\end{equation}
+Recall that
+\[
+  \te=\tdl -
+  \td_{0,\lambda}
+\]
+and
+\begin{equation}
+  \label{eqn:geosoltrans1}
+  \tdl(\xi,t) = a(\xi,z_s,z_r)\wl(t-\tau(\xi,z_s,z_r)) + O(\lambda). 
+\end{equation}
+The explicit expression \ref{eqn:ptmodhompw} for $\kappa_0$ may be 
+viewed as a special case of \ref{eqn:geosoltrans}, and gives reference 
+data for that case: 
+\begin{equation}
+  \label{eqn:analsoltrans1}
+  \td_{0,\lambda}(\xi,t) =  \tS_p[\kappa_0]\wl(\xi,t)  = a_0(\xi,z_s,z_r)\wl 
+  \left(t-\frac{|z_r-z_s|}{\tc_0(\xi)}\right) 
+\end{equation}
+where $a_0$ is given in \ref{eqn:anasoltrans}. So
+\[
+  \te(\xi,t) = \tdl(\xi,t)-\td_{0,\lambda}(\xi,t) =
+  \frac{1}{\sqrt{\lambda}}\te_1\left(\xi,\frac{t}{\lambda}\right) + O(\lambda),
+\]
+in which
+\begin{equation}
+  \label{eqn:e1def}
+  \te_1 = \td_1 -\td_{0,1}.
+\end{equation}
+In terms of the Fourier transform,
+\[
+  \hat{\te}(\xi,\omega) = \sqrt{\lambda}\hat{\te}_1(\xi,\lambda\omega)
+  + O(\lambda)
+\]
+
+Calculating the rate of change of traveltime between $z_s$ and $z_r$:
+\[
+  \delta \int_{z_s}^{z_r}\,\frac{1}{\tc}|_{\kappa=\kappa_0} =
+  -\frac{\rho \tc_0}{\kappa_0^2} \int_{z_s}^{z_r}\,\delta \kappa
+\]
+The rate at $\kappa=\kappa_0$ in the direction $\delta \kappa =
+-\nabla J_{\rm WRI}(\kappa_0)$ is therefore
+\[
+  = \left[(z_r-z_s) \left(\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi \, \frac{\rho \tc_0(\xi)}{\kappa_0^2} W_0(\xi)
+    \int\,dt\, \te(\xi,t)^2\right)\right.
+\]
+\[
+ \left. +\frac{1}{2\pi}\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\,W_2(\xi)\int\,d\omega\, 
+  |\hat{\te}(\xi,\omega)|^2 \int_{z_s}^{z_r}\, dz\,e^{-i\omega\frac{2(z_r-z)}{\tc_0(\xi)}} \right]+ O(\lambda)
+\]
+(all integrals are absolutely convergent hence can be interchanged)
+\[
+  = \left[(z_r-z_s) \left(\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi \, \frac{\rho \tc_0(\xi)}{\kappa_0^2} W_0(\xi) 
+    \int\,dt\, \te(\xi,t)^2\right)\right. 
+\]
+\[
+ \left. +\frac{1}{2\pi}\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\, \frac{\rho \tc_0(\xi)}{\kappa_0^2} W_2(\xi)\int\,d\omega\, 
+  \lambda|\hat{\te}_1(\xi,\lambda\omega)|^2 \int_{z_s}^{z_r}\, dz\,e^{-i\omega\frac{2(z_r-z)}{\tc_0(\xi)}} \right]+ O(\lambda) 
+\]
+\[
+  = \left[(z_r-z_s) \left(\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi \, \frac{\rho \tc_0(\xi)}{\kappa_0^2} W_0(\xi) 
+    \int\,dt\, \te(\xi,t)^2\right)\right. 
+\]
+\[
+ \left. +\frac{1}{2\pi}\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\, \frac{\rho \tc_0(\xi)}{\kappa_0^2} W_2(\xi)\lambda\int\,\frac{d\sigma}{\sigma}\, 
+  |\hat{\te}_1(\xi,\sigma)|^2 \int_{z_s}^{z_r}\, dz\,e^{-i\sigma\frac{2(z_r-z)}{\lambda\tc_0(\xi)}} \right]+ O(\lambda) 
+\]
+\[
+  = \left[(z_r-z_s) \left(\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi \, \frac{\rho \tc_0(\xi)}{\kappa_0^2} W_0(\xi) 
+    \int\,dt\, \te(\xi,t)^2\right)\right. 
+\]
+\[
+ \left. +\frac{1}{2\pi}\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\, \frac{\rho \tc_0(\xi)}{\kappa_0^2} W_2(\xi)\lambda\int\,\frac{d\sigma}{\sigma}\, 
+   |\hat{\te}_1(\xi,\sigma)|^2 \int_{z_s}^{z_r}\,
+   dz\,\frac{\lambda\tc_0(\xi)}{2i\sigma}\frac{\partial}{\partial
+       z}e^{-i\sigma\frac{2(z_r-z)}{\lambda\tc_0(\xi)}} \right]+ O(\lambda) 
+\]
+\[
+  = \left[(z_r-z_s) \left(\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi \, \frac{\rho \tc_0(\xi)}{\kappa_0^2} W_0(\xi) 
+    \int\,dt\, \te(\xi,t)^2\right)\right. 
+\]
+\[
+ \left. +\frac{\lambda^2}{2\pi}\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\, \frac{\rho \tc_0(\xi)^2}{\kappa_0^2} W_2(\xi)\int\,d\sigma\, 
+   |\frac{\hat{\te}_1(\xi,\sigma)}{i\sigma}|^2 \frac{1}{2i}(1-e^{-i\sigma\frac{2(z_r-z_s)}{\lambda\tc_0(\xi)}}) \right]+ O(\lambda) 
+\]
+
+In view of the relation \ref{eqn:e1def}, the assumption that $w_1$ is
+the derivative of a $C_0^{\infty}$ primitive, and the uniform boundedness of
+the other factors in the integral with respect to $\lambda$, it
+follows finally that the rate is
+\begin{equation}
+  \label{eqn:dttwri}
+  =(z_r-z_s) \left(\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi \, \frac{\rho \tc_0(\xi)}{\kappa_0^2} W_0(\xi) 
+    \int\,dt\, \te(\xi,t)^2\right) + O(\lambda)
 \end{equation}
@@
 \begin{eqnarray}
-\label{eqn:sol1d}
-p(z,t) &=& \frac{1}{2c_0}\int dz_1 f\left(z_1,t - \frac{|z-z_1|}{c_0}\right) \\
-v(z,t) &=& \frac{1}{2\kappa_0} \int dz_1 \mbox{sgn} (z-z_1) f\left(z_1,t - \frac{|z-z_1|}{c_0}\right)\\
+\label{eqn:sol1dp}
+p(z,t) &=& \frac{1}{2c_0}\int dz_1 f\left(z_1,t -
+           \frac{|z-z_1|}{c_0}\right) \\
+  \label{eqn:sol1dv}
+v(z,t) &=& \frac{1}{2\kappa_0} \int dz_1 \mbox{sgn} (z-z_1) f\left(z_1,t - \frac{|z-z_1|}{c_0}\right)
 \end{eqnarray}
@@
 
-From the second equation in the pair \ref{eqn:sol1d}
+From \ref{eqn:sol1dv}
 \[
@@
 \end{equation}
-Thus using the first equations in \ref{eqn:sol1d} and \ref{eqn:dawe1d}, 
+Thus using equation \ref{eqn:sol1dp} and \ref{eqn:dawe1d}, 
 \[
@@
   + \frac{1}{4\kappa_0c_0^2} \int\,dz_2 \, \int \,dz_1\,\delta 
-  \kappa(z_1) \frac{\partial f}{\partial t}\left(z_2,t - \frac{|z_2-z_1|+|z_1-z|}{c_0}\right)
-\end{equation}
-
-An important special case is the 1D point source,
+  \kappa(z_2) \frac{\partial f}{\partial t}\left(z_1,t - \frac{|z_1-z_2|+|z_2-z|}{c_0}\right)
+\end{equation}
+
+A convenient check on this calculation is available for constant
+$\delta \kappa = \delta \kappa_0$. Directly perturb the RHS of equation 
+\ref{eqn:sol1dp}, to obtain
+\[
+  \delta p(z,t) = -\frac{\delta c_0}{2 c_0^2} \int
+  \,dz_1\,f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
+  +\frac{\delta c_0}{2c_0^3}\int
+  \,dz_1\,\frac{\partial f}{\partial
+    t}\left(z_1,t-\frac{|z-z_1|}{c_0}\right)|z-z_1|
+\]
+Since
+\[
+  \delta c_0 = \delta \left(\frac{\kappa_0}{\rho_0}\right)^{1/2} =
+  \frac{1}{2} \left(\frac{\kappa_0}{\rho_0}\right)^{-1/2}\frac{\delta
+    \kappa_0}{\rho_0}
+\]
+\[
+  = \frac{1}{2}\frac{\delta \kappa \rho_0^{1/2}}{\kappa_0^{1/2}\rho_0}
+  \]
+\[
+=\frac{1}{2} \frac{\delta
+  \kappa_0}{\kappa_0}\left(\frac{\kappa_0}{\rho_0}\right)^{1/2}
+=\frac{1}{2} \frac{c_0\delta \kappa_0}{\kappa_0}
+\]
+the identity above can be rewritten as
+\[
+  \delta p(z,t) = -\frac{\delta \kappa_0}{4 \kappa_0 c_0} \int
+  \,dz_1\,f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
+  \]
+\begin{equation}
+  \label{eqn:dpraw} 
+  +\frac{\delta \kappa_0}{4\kappa_0c_0^2}\int
+  \,dz_1\,\frac{\partial f}{\partial
+    t}\left(z_1,t-\frac{|z-z_1|}{c_0}\right)|z-z_1|
+\end{equation}
+The critical quantity is the integral in the second term:
+\[
+  \int \,dz_1\,\frac{\partial f}{\partial t}\left(z_1,t-\frac{|z-z_1|}{c_0}\right)|z-z_1|
+\]
+\[
+  = \int_{-\infty}^z \,dz_1\,\frac{\partial f}{\partial
+    t}\left(z_1,t-\frac{z-z_1}{c_0}\right)(z-z_1)
+\]
+\[
+   + \int^{\infty}_z \,dz_1\,\frac{\partial f}{\partial t}\left(z_1,t-\frac{z_1-z}{c_0}\right)(z_1-z)
+ \]
+ \[
+   =\int_{-\infty}^z \,dz_1\,\int_{z_1}^{z} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{z-z_2+z_2-z_1}{c_0}\right)
+ \]
+ \[
+   + \int^{\infty}_z \,dz_1\,\int_{z}^{z_1}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{z_1-z_2 + z_2-z}{c_0}\right)
+ \]
+ Checking signs, one sees that this sum is equal to
+ \[
+   =\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   + \int^{\infty}_z \,dz_1\,\int_{-\infty}^{\infty}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{|z_1-z_2| + |z_2-z|}{c_0}\right)
+ \]
+ \[
+   -\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{z_1} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+\[
+   -\int_{-\infty}^z \,dz_1\,\int_{z}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   - \int^{\infty}_z \,dz_1\,\int_{-\infty}^{z}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{|z_1-z_2| + |z_2-z|}{c_0}\right)
+ \]
+ \[
+   - \int^{\infty}_z \,dz_1\,\int_{z_1}^{\infty}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{|z_1-z_2| + |z_2-z|}{c_0}\right)
+ \]
+\[
+   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   -\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{z_1} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{z-z_2+z_1-z_2}{c_0}\right)
+ \]
+\[
+   -\int_{-\infty}^z \,dz_1\,\int_{z}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{z_2-z+z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   - \int^{\infty}_z \,dz_1\,\int_{-\infty}^{z}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{z_1-z_2 + z-z_2}{c_0}\right)
+ \]
+ \[
+   - \int^{\infty}_z \,dz_1\,\int_{z_1}^{\infty}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{z_2-z_1 + z_2-z}{c_0}\right)
+ \]
+\[
+   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   -\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{z_1} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{z+z_1-2z_2}{c_0}\right)
+ \]
+\[
+   -\int_{-\infty}^z \,dz_1\,\int_{z}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{2z_2-z-z_1}{c_0}\right)
+ \]
+ \[
+   - \int^{\infty}_z \,dz_1\,\int_{-\infty}^{z}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{z_1+ z-2z_2}{c_0}\right)
+ \]
+ \[
+   - \int^{\infty}_z \,dz_1\,\int_{z_1}^{\infty}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{2z_2-z_1-z}{c_0}\right)
+ \]
+\[
+   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   -\frac{c_0}{2}\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{z_1} \,dz_2\, \frac{\partial }{\partial
+     z_2}f\left(z_1,t-\frac{z+z_1-2z_2}{c_0}\right)
+ \]
+\[
+   +\frac{c_0}{2}\int_{-\infty}^z \,dz_1\,\int_{z}^{\infty} \,dz_2\, \frac{\partial}{\partial
+     z_2}f\left(z_1,t-\frac{2z_2-z-z_1}{c_0}\right)
+ \]
+ \[
+   - \frac{c_0}{2}\int^{\infty}_z \,dz_1\,\int_{-\infty}^{z}\,dz_2\,\frac{\partial
+     }{\partial z_2}f\left(z_1,t-\frac{z_1+ z-2z_2}{c_0}\right)
+ \]
+ \[
+   +\frac{c_0}{2} \int^{\infty}_z \,dz_1\,\int_{z_1}^{\infty}\,dz_2\,\frac{\partial
+     }{\partial z_2}f\left(z_1,t-\frac{2z_2-z_1-z}{c_0}\right)
+ \]
+\[
+   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   -\frac{c_0}{2}\int_{-\infty}^z \,dz_1\, f\left(z_1,t-\frac{z-z_1}{c_0}\right)
+   -\frac{c_0}{2}\int_{-\infty}^z \,dz_1\,f\left(z_1,t-\frac{z-z_1}{c_0}\right)
+ \]
+ \[
+   - \frac{c_0}{2}\int^{\infty}_z \,dz_1\,f\left(z_1,t-\frac{z_1-z}{c_0}\right)
+   -\frac{c_0}{2} \int^{\infty}_z \,dz_1\,f\left(z_1,t-\frac{z_1-z}{c_0}\right)
+ \]
+ \[
+   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   -c_0\int_{-\infty}^{\infty} \,dz_1\, f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
+ \]
+ Insert the identity just established in the second term on the RHS of
+ \ref{eqn:dpraw}:
+ \[
+  \delta p(z,t) = -\frac{\delta \kappa_0}{4 \kappa_0 c_0} \int
+  \,dz_1\,f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
+\]
+\[
+  +\frac{\delta \kappa_0}{4\kappa_0c_0^2}\left(\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+      t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right) \right.
+\]
+\[
+  \left. -c_0\int_{-\infty}^{\infty} \,dz_1\,
+    f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)\right)
+\]
+\[
+=-\frac{\delta \kappa_0}{2\kappa_0c_0}\int_{-\infty}^{\infty} \,dz_1\,
+f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
+\]
+\begin{equation}
+  \label{eqn:dpraw}
++  \frac{\delta \kappa_0}{4\kappa_0c_0^2}\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+      t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+  \end{equation}
+which is identical to the expression \ref{eqn:sol1ddp} for the case of constant
+$\delta \kappa = \delta \kappa_0$.
+
+An important special case of the identity \ref{eqn:sol1ddp} is the 1D point source,
 $f(z,t)=w(t)\delta(z-z_s)$: then
@@
 \ref{eqn:awe1d}, and using the 1D solution for homogeneous fluids
-\ref{eqn:sol1d}, obtain
+\ref{eqn:sol1dp},\ref{eqn:sol1dv}, obtain
 \begin{eqnarray*}
@@
 \begin{eqnarray}
-\label{eqn:solpw}
+\label{eqn:solpwp}
  \tp(\xi,z,t)   &=&\frac{1}{2c_0(1-c_0^2\xi^2)^{1/2}}\int dz_s 
                  \tf\left(\xi,z_s,t -
-                 \frac{|z-z_s|}{\tc_0(\xi)}\right) \nonumber \\
+                    \frac{|z-z_s|}{\tc_0(\xi)}\right)  \\
+  \label{eqn:solpwv}
 \tv(\xi,z,t) &=& \frac{1}{2\kappa_0} \int dz_s \mbox{sgn} (z-z_s)\tf\left(\xi,z_s,t - \frac{|z-z_s|}{\tc_0(\xi)}\right)
-                 \nonumber \\
 \end{eqnarray}
@@
 \ref{eqn:sol1ddp}, substitute $\tk_0$ for $\kappa_0$, $\delta \tk$ for
-$\delta \kappa$, and $(1-c_0^2\xi^2)\tf$ for $\tf$ to obtain
+$\delta \kappa$, and $(1-c_0^2\xi^2)^{-1}\tf$ for $\tf$ to obtain
 \[
@@
 \[
-  + \frac{1}{4\tk_0(\xi)\tc_0(\xi)} \int\,dz_2 \, \int \,dz_1\,\delta 
+  + \frac{1}{4\tk_0(\xi)\tc_0(\xi)^2} \int\,dz_2 \, \int \,dz_1\,\delta 
   \tk(\xi,z_1) (1-c_0\xi^2)^{-1}\frac{\partial \tf}{\partial t}\left(\xi,z_2,t - \frac{|z_2-z_1|+|z_1-z|}{\tc_0(\xi)}\right) 
@@
 \[
-\delta \tk = \delta \left(\frac{\kappa}{1-\kappa \xi^2/\rho}\right) = \delta \kappa (1-c^2\xi^2)^{-2}
-\]
+\delta \tk = \delta \left(\frac{\kappa}{1-\kappa \xi^2/\rho}\right) = \delta \kappa (1-c_0^2\xi^2)^{-2}
+\]
+at $\kappa=\kappa_0$,
 this amounts to
 \[
-=  -\frac{1}{2\kappa_0c_0 1-c_0\xi^2)^{3/2}} \int \,dz_1\, \delta 
-\kappa(z_1) (1-c_0^2\xi^2)^{-2} \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right) 
+-\frac{1}{2\kappa_0(1-c_0^2\xi^2)^{-1}c_0(1-c_0^2\xi^2)^{-1/2}} \int \,dz_1\, \delta 
+\kappa(z_1)(1-c_0\xi^2)^{-3} \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right)
+\]
+\[
+  + \frac{1}{4\kappa_0(1-c_0^2\xi^2)^{-1}c_0^2(1-c_0^2\xi^2)^{-1}} \int\,dz_2 \, \int \,dz_1\,\delta 
+  \kappa(z_1) (1-c_0\xi^2)^{-3}\frac{\partial \tf}{\partial t}\left(\xi,z_2,t - \frac{|z_2-z_1|+|z_1-z|}{\tc_0(\xi)}\right) 
+\]
+\[
+=  -\frac{1}{2\kappa_0c_0 (1-c_0\xi^2)^{3/2}} \int \,dz_1\, \delta 
+\kappa(z_1) \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right) 
 \]
@@
   + \frac{1}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int\,dz_2 \, \int \,dz_1\,\delta 
-  \kappa(z_1) (1-c_0^2\xi^2)^{-2}  \frac{\partial \tf}{\partial t}\left(\xi,z_2,t - \frac{|z_2-z_1|+|z_1-z|}{\tc_0(\xi)}\right) 
+  \kappa(z_1) \frac{\partial \tf}{\partial t}\left(\xi,z_2,t - \frac{|z_2-z_1|+|z_1-z|}{\tc_0(\xi)}\right) 
 \end{equation}
@@
 \[
-  (1-c^2\xi^2)^{-2}\delta (c^2)\xi^2= (1-c^2\xi^2)^{-2}\delta \kappa
+  (1-c^2\xi^2)^{-2}\delta (c^2)\xi^2= (1-c^2\xi^2)^{-2}\xi^2\delta \kappa
   /\rho
 \]
-Therefore, from \ref{eqn:sol1d}, the second summand identified above
+Therefore, from \ref{eqn:sol1dp}, the second summand identified above
 is
 \[
-  =\frac{1}{2\tc(\xi)\rho_0}\int \,dz_1\, \delta 
+  =\frac{\xi^2}{2\tc(\xi)\rho_0}\int \,dz_1\, \delta 
 \kappa(z_1)(1-c_0^2\xi^2)^{-2} \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right)
 \]
+\[
+  = \frac{\xi^2}{2c_0(1-c_0^2\xi^2)^{3/2}\rho_0}\int \,dz_1\, \delta 
+\kappa(z_1) \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right)
+\]
+
 \begin{equation}
   \label{eqn:solpwdp2}
-=\frac{c_0^2\xi)^2}{2\kappa_0c_0(1-c_0^2\xi^2)^{3/2}}\int \,dz_1\, \delta 
+=\frac{c_0^2\xi^2}{2\kappa_0c_0(1-c_0^2\xi^2)^{3/2}}\int \,dz_1\, \delta 
 \kappa(z_1) \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right)
@@
   \label{eqn:solpwdp}
-  + \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int\,dz_2 \, \int \,dz_1\,\delta 
+  + \frac{1}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int\,dz_2 \, \int \,dz_1\,\delta 
   \kappa(z_1)  \frac{\partial \tf}{\partial
@@
 \[
-  D(\tS[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
+  D(\tS[\kappa]\tf)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
   -\frac{1}{2\kappa_0c_0(1-c_0^2\xi^2)^{1/2}}\int \,dz_1\, \delta 
@@
 \begin{equation}
-  \label{eqn:dmodhomst1}
-    + \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int\,dz_2 \, \int \,dz_1\,\delta 
+  \label{eqn:dmodhomst1app}
+    + \frac{1}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int\,dz_2 \, \int \,dz_1\,\delta 
   \kappa(z_1)  \frac{\partial \tf}{\partial
@@
 \end{equation}
-which is equation \ref{eqn:dmodhompwst}. For a point source
+which is equation \ref{eqn:dmodhompwst1}. For a point source
 $\tf(\xi,z,t) = w(t)\delta(z-z_s)$, this reduces to
 \[
-  D(\tS[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
+  D(\tS_p[\kappa]w)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
   -\frac{1}{2\kappa_0c_0(1-c_0^2\xi^2)^{1/2}} \delta 
@@
   \label{eqn:dmodhompt1}
-    + \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int \,dz\,\delta 
+    + \frac{1}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int \,dz\,\delta 
   \kappa(z) \frac{\partial w}{\partial

### 2020-02-02 repro/papers/ssevelan/attempt.tex
trans to srcextcv

Another approach: write $w = L[c]^{-1}_-P^T h$. The object is to compute $u=L[c]^{-1}_+w$, that is, the solution of 
\[
(c^{-2}\partial_t^2 - \nabla^2)u(t) = w(t), 0\le t \le T; \,u(0) = \partial_t u(0) = 0
\]
Duhamel's principle would provide a solution in the form of integrals over time of solutions of homogeneous initial value problems:
\[
u(t) = \int_0^t d\tau v(t,\tau), \, (c^{-2}\partial_t^2 - \nabla^2)v(t,\tau) = 0, 0\le t \le T; \,v(t,\tau)_{t=\tau} = 0, \partial_t v(t,\tau)_{t=\tau} = w(\tau)
\]
Note that $w$ solves the homogenous wave equation except at $z=0$. That makes it tempting to use $w$ directly to define $v$. In fact, set 
\[
W(t) = -int_t^T ds w(s),\,v_0(t,\tau) = \frac{1}{2}(W(t) - W(2\tau-t)) 
\]
Then $v_0$ satisfies the Duhamel initial conditions. Note that because $w$ has zero Cauchy data at $t=T$,
\[
\partial_t W(t)= -\int_t^T ds \partial_t w(s),\,\partial_t^2 W(t)=-\int_t^T ds \partial^2_t w(s)
\]
so 
\[
(c^{-2}\partial_t^2 - \nabla^2)v_0(t,\tau) = -\int_t^T(c^{-2}\partial^2_t -\nabla^2)w + \int_{2\tau-t}^T(c^{-2}\partial^2_t -\nabla^2)
\]
For $t>\tau$, $t > 2\tau-t$, so 
\[
= -\int_{2\tau-t}^t(c^{-2}\partial^2_t -\nabla^2)w = -\int_{2\tau-t}^t P^Th
\]
and 
\[ 
v_0(t,\tau) = \int_{2\tau-t}^t w  
\] 
Consequently $v=v_0 + v_1$, in which $v_1$ is defined by 
\[
(c^{-2}\partial_t^2 - \nabla^2)v_1(t,\tau) = \int_{2\tau-t}^t P^Th, \,  v_1(t,\tau)_{t=\tau} = 0, \,  \partial_t v_1(t,\tau)_{t=\tau} = 0
\]

### 2020-02-02 repro/papers/ssevelan/paper.tex
trans to srcextcv

@@
 \end{equation}
-Bulk modulus is depicted in Figure \ref{fig:bml0}.
+Bulk modulus is depicted in Figure XXX \ref{fig:bml0}.
 
-\plot{bml0}{width=\textwidth}{Lens example: target bulk modulus $\kappa$}
+%\plot{bml0}{width=\textwidth}{Lens example: target bulk modulus $\kappa$}
 
@@
 %\includegraphics[height=1.5in]{Fig/pulse00.pdf} %\includegraphics[height=1.5in]{Fig/ptpwindl0.pdf}\\
-\plot{pulse00}{width=0.5\textwidth}{Bandpass filter pulse $f$}
-\plot{ptpwindl0}{width=\textwidth}{Receiver line  $S(\kappa)$  at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km.
-bulk modulus $\kappa$ with low velocity lens as in Figure \ref{fig:bml0}, $\rho$=1 g/cm$^{3}$.
-Coarse grid: $dx=dz=20$ m, $dt=$ 8 ms, (2,4) staggered grid scheme.
-}
+%\plot{pulse00}{width=0.5\textwidth}{Bandpass filter pulse $f$}
+%\plot{ptpwindl0}{width=\textwidth}{Receiver line  $S(\kappa)$  at
+%$z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km. bulk modulus
+%$\kappa$ with low velocity lens as in Figure \ref{fig:bml0}, $\rho$=1 g/cm$^{3}$. Coarse grid: $dx=dz=20$ m, $dt=$ 8 ms, (2,4) staggered grid scheme. }
 % at receiver position $1000 m$ for source at $x=3500$ m, $z=3000$ m}
@@
 %\includegraphics[height=1.2in]{Fig/ptpwindl0.pdf}\includegraphics[height=1.2in]{Fig/cgwhestsourceplh0.pdf}\\
-\plot{cgwhestsourceplh0}{width=\textwidth}{$\of=\oS(\kappa_0)^{\dagger}d$ for $\kappa_0 \equiv 4.0$ GPa, approx. via CG - $r=2000 m$}
+%%\plot{cgwhestsourceplh0}{width=\textwidth}{$\of=\oS(\kappa_0)^{\dagger}d$ for $\kappa_0 \equiv 4.0$ GPa, approx. via CG - $r=2000 m$}
 

### 2020-02-02 repro/papers/ssevelan/talkfigs.tex
trans to srcextcv

\section{Figures for talk}
\plot{pulse00}{width=\textwidth}{Bandpass filter pulse}
\plot{fwitrbulkmodinvlh0}{width=\textwidth}{Inversion for bulk modulus of data displayed in Figure \ref{fig:ptpwindl0}. Initial model: constant bulk modulus = 4.0 GPa. Algorithm: 20 iterations of trust region Krylov-Gauss-Newton algorithm with preconditioning by box smoother of radius 5 points, repeated 10 times.  Each KGN step computed by up to 20 Conjugate Gradient iterations.} 
\plot{fwitrdatapreslh0}{width=\textwidth}{Data residual produced by 20 itersions of KGN FWI for bulk modulus, starting at constant bulk modulus. Residual RMS approximately 26\% of data RMS.}
\plot{cgwhestsourceplh0}{width=\textwidth}{Extended inversion for source on 1500 m $\le x \le$ 5500 m, $z$=3000 m, data from Figure \ref{fig:ptpwindl0}, homogeneous bulk modulus = 4.0 GPA.}
\plot{cgwhresdataplh0}{width=\textwidth}{Residual data from extended source inversion shown in Figure \ref{fig:cgwhestsourceplh0}. RMS is 3\% of data RMS.}
\plot{bmh0s}{width=\textwidth}{Iteration 1}
\plot{cgw1bulkupdlh0}{width=\textwidth}{Iteration 1 SSE bulkmod estimate}
\plot{cgw2bulkupdlh0}{width=\textwidth}{Iteration 2 SSE bulkmod estimate}
\plot{cgw3bulkupdlh0}{width=\textwidth}{Iteration 3 SSE bulkmod estimate}
\plot{cgw4bulkupdlh0}{width=\textwidth}{Iteration 4 SSE bulkmod estimate}
\plot{cgw5bulkupdlh0}{width=\textwidth}{Iteration 5 SSE bulkmod estimate}
\plot{cgw6bulkupdlh0}{width=\textwidth}{Iteration 6 SSE bulkmod estimate}
\plot{cgw7bulkupdlh0}{width=\textwidth}{Iteration 7 SSE bulkmod estimate}
\plot{cgw8bulkupdlh0}{width=\textwidth}{Iteration 8 SSE bulkmod estimate}
\plot{cgw9bulkupdlh0}{width=\textwidth}{Iteration 9 SSE bulkmod estimate}
\plot{cgw10bulkupdlh0}{width=\textwidth}{Iteration 10 SSE bulkmod estimate}
\plot{cgw11bulkupdlh0}{width=\textwidth}{Iteration 11 SSE bulkmod estimate}
\plot{cgw12bulkupdlh0}{width=\textwidth}{Iteration 12 SSE bulkmod estimate}
\plot{cgw13bulkupdlh0}{width=\textwidth}{Iteration 13 SSE bulkmod estimate}
\plot{cgw14bulkupdlh0}{width=\textwidth}{Iteration 14 SSE bulkmod estimate}
\plot{cgw15bulkupdlh0}{width=\textwidth}{Iteration 15 SSE bulkmod estimate}
\plot{cgw22bulkupdlh0}{width=\textwidth}{Iteration 22 SSE bulkmod estimate}
\plot{cgw30bulkupdlh0}{width=\textwidth}{Iteration 30 SSE bulkmod estimate}
\plot{fwitrsse1bulkmodinvlh0}{width=\textwidth}{Inversion for bulk modulus of data displayed in Figure \ref{fig:ptpwindl0}. Initial model: result of 15 steepest descent steps of SSE, Figure \ref{fig:cgw15bulkupdlh0}. Algorithm: 20 iterations of trust region Krylov-Gauss-Newton algorithm with preconditioning by box smoother of radius 5 points, repeated 10 times.  Each KGN step computed by up to 20 Conjugate Gradient iterations.} 
\plot{fwitrsse1datapreslh0}{width=\textwidth}{Data residual produced by 20 itersions of KGN FWI for bulk modulus, starting at output of 15 steepest descent steps of SSE, Figure \ref{fig:cgw15bulkupdlh0}. Residual RMS approximately 1.2\% of data RMS.}
\plot{cgw0estsourceplh0}{width=\textwidth}{Iteration 1 SSE source estimate alpha=4.36 $\times 10^{-6}$}
\plot{cgw1estsourceplh0}{width=\textwidth}{Iteration 2 SSE source estimate alpha=5.93 $\times 10^{-6}$}
\plot{cgw2estsourceplh0}{width=\textwidth}{Iteration 3 SSE source estimate alpha=6.82 $\times 10^{-6}$}
\plot{cgw3estsourceplh0}{width=\textwidth}{Iteration 4 SSE source estimat alpha=7.93 $\times 10^{-6}$e}
\plot{cgw4estsourceplh0}{width=\textwidth}{Iteration 5 SSE source estimate alpha=9.29 $\times 10^{-6}$}
\plot{cgw5estsourceplh0}{width=\textwidth}{Iteration 6 SSE source estimate alpha=1.1 $\times 10^{-5}$}
\plot{cgw6estsourceplh0}{width=\textwidth}{Iteration 7 SSE source estimate alpha=1.32 $\times 10^{-5}$}
\plot{cgw7estsourceplh0}{width=\textwidth}{Iteration 8 SSE source estimate alpha=1.32 $\times 10^{-5}$}
\plot{cgw8estsourceplh0}{width=\textwidth}{Iteration 9 SSE source estimate alpha=1.72 $\times 10^{-5}$}
\plot{cgw9estsourceplh0}{width=\textwidth}{Iteration 10 SSE source estimate alpha=2.2 $\times 10^{-5}$}
\plot{cgw10estsourceplh0}{width=\textwidth}{Iteration 11 SSE source estimate alpha=2.79 $\times 10^{-5}$}
\plot{cgw11estsourceplh0}{width=\textwidth}{Iteration 12 SSE source estimate alpha=2.79 $\times 10^{-5}$}
\plot{cgw12estsourceplh0}{width=\textwidth}{Iteration 13 SSE source estimate alpha=2.79 $\times 10^{-5}$}
\plot{cgw13estsourceplh0}{width=\textwidth}{Iteration 14 SSE source estimate alpha=2.79 $\times 10^{-5}$}
\plot{cgw14estsourceplh0}{width=\textwidth}{Iteration 15 SSE source estimate alpha=2.79 $\times 10^{-5}$}
\plot{cgw21estsourceplh0}{width=\textwidth}{Iteration 22 SSE source estimate alpha=5.73 $\times 10^{-5}$}
\plot{cgw29estsourceplh0}{width=\textwidth}{Iteration 30 SSE source estimate alpha=9.1 $\times 10^{-5}$}
\plot{cgw0resdataplh0}{width=\textwidth}{Iteration 1 SSE data residual}
\plot{cgw1resdataplh0}{width=\textwidth}{Iteration 2 SSE data residual}
\plot{cgw2resdataplh0}{width=\textwidth}{Iteration 3 SSE data residual}
\plot{cgw3resdataplh0}{width=\textwidth}{Iteration 4 SSE data residual}
\plot{cgw4resdataplh0}{width=\textwidth}{Iteration 5 SSE data residual}
\plot{cgw5resdataplh0}{width=\textwidth}{Iteration 6 SSE data residual}
\plot{cgw6resdataplh0}{width=\textwidth}{Iteration 7 SSE data residual}
\plot{cgw7resdataplh0}{width=\textwidth}{Iteration 8 SSE data residual}
\plot{cgw8resdataplh0}{width=\textwidth}{Iteration 9 SSE data residual}
\plot{cgw9resdataplh0}{width=\textwidth}{Iteration 10 SSE data residual}
\plot{cgw10resdataplh0}{width=\textwidth}{Iteration 11 SSE data residual}
\plot{cgw11resdataplh0}{width=\textwidth}{Iteration 12 SSE data residual}
\plot{cgw12resdataplh0}{width=\textwidth}{Iteration 13 SSE data residual}
\plot{cgw13resdataplh0}{width=\textwidth}{Iteration 14 SSE data residual}
\plot{cgw14resdataplh0}{width=\textwidth}{Iteration 15 SSE data residual}

### 2020-02-03 repro/papers/wri1d/paper.tex
first commit wri expl

--- 2019-12-20 repro/papers/sserefl/paper.tex
@@
-\title{Surface Source Extension, Reflection Case}
+\title{Wavefield Reconstruction Inversion: an example}
 \author{William. W. Symes \thanks{The Rice Inversion Project,
@@
 
-\righthead{Reflection SSE}
-
-\maketitle
-\begin{abstract}
-FWI via surface source extension (SSE) applies equally well to reflection or transmission data.
-In the transmission case, it is asymptotically equivalent to a form of travel time tomography, fitting event slowness rather than arrival time.
-Use of this method for reflection data simply involves permitting reflectors in the velocity and other subsurface mechanical parameters.
-It is possible to analyze a simple layered medium special case in the plane wave domain.
-The analysis suggests that in fact reflection SSE cannot constructively update velocity in absence of correctly positioned reflectors, so more closely resembles so-called Reflection FWI than transmission SSE in its velocity updating capabilities.
-This theoretical result is supported by numerical experiments with 2D reflection data.
-\end{abstract}
+\righthead{1D WRI}
 
 \section{Introduction}
-FWI based on source extension has been around since the early 1990's,
-and some versions are provably robust - converge from substantially
-incorrect initial velocity models. However the theoretical robustness
-results all pertain to purely transmission configurations, such as
-crosswell or diving wave geometry. It has been an open question
-whether source extended FWI remains robust if all kinematic
-information in the data is in the form of reflection
-moveout. Some numerical evidence for such robustness has appeared, but
-in each case it is not obvious that post-critical events did not play
-a role in the inversion.
 
-The goal of this paper is to produce a proof that in one precisely
-defined case and in a precisely defined sense, one particular variant
-of source-extended FWI is {\em not} robust for inversion of reflection
-data. This counterexample negates any general claim of robustness for
-such methods of reflection inversion, of course.
+\section{Full Waveform Inversion}
+Full waveform inversion (FWI) is the current nomenclature for data-fitting parameter estimation driven by wavefield modeling.
+The aim of FWI is to fit a vector of data observations $d$ with a predicted data vector of the same form.
+The prediction is accomplished by application of a modeling operator $S$ to a vector of input observations $f$, representing sources of energy that initiate waves in the material model underlying $S$.
+$S$ propagates these waves to measurement locations at which the data samples are extracted from the wave fields, generating the output of $S$.
+The propagation dynamics depend on a vector of material parameters $c$, hence $S=S[c]$ also depends on $c$.
 
-Surface source extension, defined below, is the method investigated
-here. The objective of the reasoning laid out in the following sections is
-to explicitly calculate the gradient of a particular surface source
-extended FWI objective. A context that permits such explicit
-calculation is plane-wave propagation in a layered medium. An explicit
-computation of the gradient at a homogeneous background medium with
-data containing a single reflector shows that the gradient has
-vanishing low-frequency components as the central data frequency goes
-to infinity: in fact, the mean square in any fixed frequency band goes
-to zero. Therefore a step in the negative gradient direction cannot
-affect traveltimes significantly and therefore this technique cannot
-produce a productive velocity update in this case.
+A restatement of the goal: estimation of the materal parameters $c$ (hence the modeling operator $S[c]$) and the  from the data $d$, via the requirement that $S[c]f \approx d$.
+A typical mehod for enforcing this requirement is the minimization of an objective misfit measure (objective, for short), the most common choice being the square norm of a Hilbert space in which the data is presumed to reside:
+\begin{equation}
+  \label{eqn:deffwifull}
+  J_{\rm FWI}[c,f] = \frac{1}{2}\|S[c]f -d\|^2.
+\end{equation}
+This approach was first suggested in the 1980's (\cite{BamChavLai:79,Tara:84a,KolbColLai:86,Crasetal:90},
+and many other papers since then). Usually some form of regularization
+is applied, to compensate for poorly determined aspects of $c$
+and/or $f$, as is explained in Tarantola's influential book
+\cite[]{Tarantola:05}. Also, $f$ may be constrained in one way or another to embody characteristics of field energy sources (an example of this is given below).
 
-It is important to note the contrast with other extension methods. The
-surface source extended FWI objective in the closest analogue for
-transmission, namely the estimation of a homogeneous velocity from one
-or more transmission traces, has been shown to be locally convex
-independent of data frequency content. Also, the subsurface offset
-extended FWI objective for surface reflection data has been shown to
-have similar properties.
+The example of FWI to be explored in this paper is one of the simplest possible, based on the 1D acoustics system connecting excess pressure $p$, particle velocity $v$, sonstitutive law defect (``source'') $f$, density $\rho$, and wave velocity $c$:
+\begin{eqnarray}
+\label{eqn:awe1d}
+\frac{\partial p}{\partial t} + \rho c^2\frac{\partial 
+  v}{\partial z} &=& f \nonumber\\
+\rho \frac{\partial v}{\partial t} + \frac{\partial p}{\partial 
+  z}&=&0\nonumber\\
+ p,v&=&0, t \ll 0. 
+\end{eqnarray}
+The fields $p,v,f$ are functions of spatial position $z \in \bR$ and time $t \in \bR$, whereas $c, \rho$ are functions of $z$ alone, so that the system \ref{eqn:awe1d} is autonomous.
 
-The first of the following sections sets out the framework of this
-study. It does {\em not} include mathematically complete definitions
-or fully qualified statements of results - these may be found in their
-entirety in later sections. Instead, it is intended to introduce the
-concepts that figure in the main results, which are stated succinctly
-in the third section. The  I will review plane-wave modeling, and construct
-an asymptotic approximation to single-reflector plane wave data. In
-the final section I will present the surface source extended FWI
-objective and compute its gradient at a homogeneous background with
-single reflector data. The main result will follow immediately from
-this calculation.
+The system \ref{eqn:awe1d} has classical (smooth) solutions $(p,v)$ when $c, \rho,$ and $f$ are smooth, and $\log c$ and $\log \rho$ are bounded on $\bR$, as is well-established \cite[]{Lax:PDENotes}.
+In this paper, for reasons to be discussed below, $c$ and $\rho$ are constrained to be constant ($z$-independent) in which case solutions may be constructed by elementary methods (Appendix A).
 
-I include several appendices, on 1D wave propagation and plane wave
-reflection modeling, as this material is difficult to find in
-sufficiently precise form.
+Choose $l_c, l_{\rho} >0$ and admit $c, \rho$ as coefficients in \ref{eqn:awe1d} for which $|\log c| \le l_c, |\log \rho| \le l_{\rho}$.
 
-Whenever convenient, I abbreviate ``surface source extended FWI'' as SSE-FWI.
+Limit observations to the time interval $[0,T]$, at the spatial (``receiver'')  location $z_r$. The modeling operator outputs the pressure trace $p(z_r,t)$ over the time interval $[0,T]$:
+\begin{equation}
+  \label{eqn:defmod}
+  S[c]f = p|_{\{z_r\}\times [0,T]}
+\end{equation}
 
-\section{Preliminaries}
+The formulation of the inverse problem via least-squares requires a choice of Hilbert space structure for the domain and range of $S[c]$. A natural choice is
+\begin{equation}
 
-This paper will use linear acoustics in velocity-pressure form as its
-model for wave motion. In this model, the space-time fields $p$ is excess
-pressure, $\bv$ is particle velocity about equilibrium, $f$
-represents an energy source as a defect in the acoustic constitutive
-law. The bulk modulus $\kappa$ and density $\rho$ are positive functions of
-spatial position. The acoustic wave system couples these quantities:
-\begin{eqnarray}
-\label{eqn:awe}
-\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
-                                                                f \nonumber\\
-\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
-p,{\bf v}&=&0, t \ll 0
-\end{eqnarray}
-In terms of the solution of the system \ref{eqn:awe}, the modeling
-operator is defined by
+  S[c]: L^2([z_{\rm min},z_{\rm max}] \times [0,T]) \rightarrow L^2[0,T]
+\end{equation}
+For homogeneous ($z$-independent) $\kappa$, Appendix A provides an explicit expression:
 \begin{equation}
-\label{eqn:fwd}
-S[\kappa]f = p|_{z=z_r}
+  \label{eqn:defmodhom}
+  S[c]f(t) = \frac{1}{2c}\int dz f\left(z,t - \frac{|z_r-z|}{c}\right) 
 \end{equation}
-Two special types of right-hand side are important in the discussion
-to follow: isotropic point sources, of the form
+From this expression it is simple to see that $S[c]$ is a bounded operator with the domain and range described in \ref{eqn:defmoddr}.
+
+With this framework, the data is presumed to lie in the range of $S[c]$: $d \in L^2[0,T]$, and all components of the basic FWI definition \ref{eqn:deffwifull} are defined.
+
+The version of FWI discussed here presumes that the source field corresponding to the data $d$ is supported at a point $z_s \ne z_r$, and is known.
+This field will be denoted depends on a function of time (``wavelet'') $w \in L^2[0,T]$.
+Formally, the resulting acoustic field satisfies the system \ref{eqn:awe1d} with $f=w(t)\delta(z-z_s)$.
+Inserting this expression in the explicit expression \ref{eqn:defmodhom}, obtain
 \begin{equation}
-  \label{eqn:ipr}
-  f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s).  
+  \label{eqn:defmodpt}
+  S[c](w\delta(\cdot-\z_s)) = \frac{1}{2c}w\left(t - \frac{|z_r-z_s|}{c}\right) = S_p[c]w(t)
 \end{equation}
-and surface sources, of the form
-\begin{equation}
-  \label{eqn:ss}
-  f(\bx,t;\bx_s) = u(x,y,t;x_s,y_s)\delta(z-z_s). 
-\end{equation}
-Point sources are a special case of surface sources. Introduce appropriately
-parametrized versions of the modeling operator for these two cases:
-\begin{eqnarray}
-  \label{eqn:defpt}
- f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s) & \Rightarrow & S[\kappa]f = S_p
-  [\kappa]w\\ 
-  \label{eqn:defss}
-f(\bx,t;\bx_s) = u(x,y,t;x_s,y_s)\delta(z-z_s) & \Rightarrow & S[\kappa]f =
-  S_s[\kappa]u
-\end{eqnarray}
 
-The standing assumption in this work is that observed data to be fit
-is to good approximation the result of a point source. That is, there
-are bulk modulus $\kappa_*(\bx)$ and source wavelet or pulse $w_*(t)$
-so that
-\begin{equation}
-  \label{eqn:datadef}
-  d = S_p[\kappa_*]w_*
-\end{equation}
-In equation \ref{eqn:datadef}, $d$ is a function of receiver position
-$\bx_r$, source position $\bx_s$, and time $t$.
+\section{Wavefield Reconstruction Inversion}
+This section will introduce Wavefield Reconstruction Inversion (WRI) and develop some of its formal algebraic properties.
+The notation $S[c]$ will represent a modeling operator based on wave dynamics of some sort, depending on a vector $c$ of material parameters.
+The conclusions developed here in fact apply to WRI in any such setting.
+These conclusions will be applied to 1D acoustics in the following section.
 
-Full waveform inversion (FWI), as construed in this paper, means:
-\begin{equation}
-  \label{eqn:deffwi}
-  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }w \mbox{ so that }d
-  \approx S_p[\kappa]w.
-\end{equation}
-In the following paragraphs, I will discuss four approaches to FWI:
-
-\subsection{``Straight'' FWI}
-
-The standard approach to FWI is to formulate the task \ref{eqn:deffwi}
-as a nonlinear least squares problem:
-\begin{equation}
-  \label{eqn:defols}
-  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }w \mbox{ to
-    minimize }
-  \| d -  S_p[\kappa]w\|^2.
-\end{equation}
-Here the symbol $\|\cdot\|^2$ signifies {\em mean-square}, that is,
-sum the squares of every value of the field between the vertical bars.
-
-This concept was introduced in the 1980's
-(\cite{BamChavLai:79,Tara:84a,KolbColLai:86,Crasetal:90},
-and many other papers since then). Usually some form of regularization
-is applied, to compensate for poorly determined aspects of $\kappa$
-and/or $w$, as is explained in Tarantola's influential book
-\cite[]{Tarantola:05}. The approach was of only academic interest
-until roughly fifteen years ago, when 3D modeling and related computations
-first became feasible at industrial scale, and evidence began to emerge 
-that FWI offered potentially decisive improvements in imaging accuracy
-over other inversion (imaging) methods, especially for construction of
-large-scale earth model features
-\cite[]{Plessix:10,Vigh:10}. In the last ten years, FWI has become
-both a production service offered by the geophysical industry, and a
-leading topic of both industry-oriented and academic research.
-
-As was understood very early in the evolution of FWI, the basic
-least-squares formulation \ref{eqn:defols} suffers from a major
-defect: the parameter spaces tend to be so large that only iterative
-optimization (that is, a variant
-of Newton's method) may feasibly be applied, due to computational cost
-- this remains true today and for the foreseeable future - but these
-method tend not to produce usable results unless the initial iterate
-is in some sense ``good'' (predicts travel times to within a
-half-wavelength) \cite[]{GauTarVir:86,Plessix:10}.``Good'' initial models are not always
-available. Accordingly, a number of alternative approaches have been
-proposed, many of them variations on the least squares theme, amongst
-which are the three approaches described below.
-
-\subsection{Source Extensions}
-
-Careful analysis of the defect mentioned in the last paragraph shows
-that the failure of iterative optimization applied to \ref{eqn:defols}
-to converge (or at least to converge usefully quickly) appears to
-arise from failure to fit the data: some parts of the oscillatory time
-series $d(\cdot,t)$ overlap oscillations in $S_p[\kappa]w$, while
-others do not. It seems reasonable to guess that this ``cycle skipping'' phenomenon 
-could be avoided if the data could be fit throughout the inversion
-process. That requirement in turn implies a modeling operator with
-more degrees of freedom, that is, a bigger domain, than the basic
-modeling operator $S_p$, which should produce the same results as
-$S_p$ when applied to pairs of bulk moduli and isotropic point sources \ref{eqn:ipr}.
-
-There is
-well-established mathematical terminology for this relationship:
-operator A is an {\em extension} of operator B if (i) the domain of $A$
-contains the domain of $B$, and whenever $v$ is in the domain of $B$,
-$Av = Bv$. Two extensions of $S_p$ already appear
-above: the isotropic point radiator \ref{eqn:ipr} is a special case of
-both the surface source \ref{eqn:ss}, which is in turn a special case
-of the generic right-hand side in the system \ref{eqn:awe}. So the
-domain of $S_s$ contains the domain of $S_p$, as does the domain of
-$S$. From the relations \ref{eqn:specsrc}, $S$ and $S_s$ produce the
-same result as $S_p$ when applied to sources in their common domain,
-so these operators are both extensions of $S_p$.
-
-Since these extensions of $S_p$ involve additional degrees of freedom
-for the energy source in the system \ref{eqn:awe}, they have come to
-be called {\em source extensions}. These are not the only source
-extensions - see \cite{HuangNammourSymesDollizal:SEG19} for an
-overview of source extensions and their properties. Extensions may
-also be constructed by adding degrees of freedom to the coefficients
-in the wave system (bulk modulus, in the present case), 
-rather than the right hand side - see \cite{geoprosp:2008} for a
-description of some of these {\em medium extensions}.
-
-The purpose of any extension is to add options for data fitting,
-thereby reducing the chances of cycle-skipping. As will be seen below,
-both extension $S$ and $S_s$ permit arbitrarily accurate fitting of
-essentially any data with essentially any bulk modulus field $\kappa$,
-with proper choice of $f$ or $w$ respectively. However, the standing
-assumption on the data $d$ is that it derives from a point source, per
-\ref{eqn:datadef}. So the added degrees of freedom represented by the
-extensions $S$ and $S_s$ must be suppressed somehow. As explained by
-\cite{geoprosp:2008}, a convenient mechanism for suppression of the
-extension degrees of freedom is a penalty term added to the objective
-of the extended
-version of the least squares problem \ref{eqn:defols}. Choice of penalty
-term, along with choice of extension, characterizes each of the extended
-inversion methods detailed in the next few paragraphs.
-
-\subsubsection{Wavefield Reconstruction Inversion}
-This concept was introduced by \cite{LeeuwenHerrmannWRI:13}, and further
+WRI was introduced by \cite{LeeuwenHerrmannWRI:13}, and further
 developed by \cite{LeeuwenHerrmann:16,WangYingst:SEG16} and other
-authors. It is based on the extension $S$ (equation \ref{eqn:fwd}), with full
+authors. It is based on the presumption that the correct source $q$ is known.  and extension $S$ (equation \ref{eqn:fwd}), with full
 space-time volume source $f$, and presumes that the correct point
@@
 differently, in terms of the second order wave equation for the
-pressure wavefield $p$. From this viewpoint, the 
-second terms becomes a scaled mean square of the residual in the
-wave equation. The formulation presented here is equivalent, and was
-introduced by \cite{WangYingst:SEG16}. \cite{LeeuwenHerrmann:16} used
+pressure wavefield $p$.
+From this viewpoint, $\partial/\partial t(f-w_*\delta(\cdot-z_s))$ is the residual, that is, the image of the 2nd order wave operator on $p$, so the second term penalizes the failure of $p$ to solve the wave equation with source $\partial w_*/\partial t \delta(\cdot-z_s)$.
+The formulation presented here is equivalent, and was
+introduced by \cite{WangYingst:SEG16}.
+
+\cite{LeeuwenHerrmann:16} used
 the variable projection method \cite[]{GolubPereyra:03}, eliminating the source $f$ in the
 inner step and updating the bulk modulus (or an equivalent quantity)
-in the outer step. In any penalty method, control of the penalty
+in the outer step. That is, the problem \ref{eqn:defwri} is equivalent to minimization of
+\begin{equation}
+\label{eqn:defvpm}
+  J[\kappa] =
+  \mbox{min}_f \frac{1}{2}(\| d-S[\kappa]f\|^2+\alpha^2\|f-w_*\delta(\cdot-\bx_s)\|^2)
+\end{equation}
+over $\kappa$. This is the approach that I shall pursue here.
+
+In any penalty method, control of the penalty
 parameter has a large influence on the speed of
@@
 to minimize the influence of the penalty weight choice.
+Alternatively, one can use a version of the discrepancy principle to adjust $\alpha$ dynamically \cite[]{FuSymes2017discrepancy}, as the WRI problem has the necessary features described in that paper.
 
-The Achilles' heel of Wavefield Reconstruction Inversion (WRI), in any of
-the formulations described so far, is the necessity of storing a full
-space time field, namely $f$ (or $p$). Especially for time-domain
-formulation and high frequency waves, this storage requirement can be
-formidable. In fact, ``most'' source fields $f$ are non-radiating,
-that is, $S[\kappa]f=0$. The orthocomplement of this subspace of
+``Most'' source fields $f$ are non-radiating,
+that is, $S[\kappa]f=0$. The orthocomplement of the subspace of
 non-radiating sources is the range of the adjoint operator
-$S[\kappa]^T$. Therefore the sources may be re-parametrized as images of
+$S[\kappa]^T$. This observation suggest that the sources may be re-parametrized as images of
 $S[\kappa]^T$. Write $f = g + w_*\delta(\cdot-\bx_s)$: then the second term in
@@
 is, $n$ is a non-radiating
-source), and note that the decomposition is orthogonal. Then the WRI problem \ref{eqn:defwri} may be reformulated as
-\begin{equation}
-  \label{eqn:defwrireform}
-  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }e \mbox{ to
-    minimize }
-  \| d -  S[\kappa]w_*\delta(\cdot-\bx_s) - S[\kappa]S[\kappa]^Te\|^2+\alpha^2\|S[\kappa]^Te\|^2.
-\end{equation}
-Note that $n$ has disappeared: {\em a priori} it does not contribute to the first
-term, and $n \ne 0$ makes the second term larger. So minimization
-necessarily entails the choice $n=0$.
-
-This reformulation does {\em not} require storage of $d+1$-dimensional
-fields, as does the original WRI formulation \ref{eqn:defwri}. The
-field $e$ has taken the place of the space-time source $f$, and its
-size is precisely the same as that of the data field $d$. Therefore
-the extra memory cost over modeling is the same as another copy of
-data. A close examination reveals that in computing the composition
-$SS^T$ via time-stepping methods, the same simultaneity issue arises
-as in all applications of the adjoint state method. However good
-solutions are known \cite[]{Griewank:book}, and since precisely the
-same issue arises in time-domain FWI, the cost is the same. Therefore,
-it is this reformulation of WRI that will be analyzed below.
-
-\subsubsection{Space-Time Extension}
-\cite{HuangSymes:Geo18a} introduced a variant of WRI, also using the space-time extension, but not
-requiring knowledge of the target point source wavelet $w_*$: in fact,
-an estimate of this wavelet is a by-product of the algorithm. This
-approach relies on an {\em annihilator}, a linear operator whose null
-space is precisely the desired class of source models, in this case
-isotropic point radiators. Such an operator can be used to generate a
-quadratic form whose zero set is also the desired class, hence can be
-used as a penalty in a nonlinear least squares objective, designed to
-drive the source towards to the desired type. See \cite{geoprosp:2008}
-for a discussion of annihilators in extended inversion.
-
-The annihilator proposed by \cite{HuangSymes:Geo18a} is multiplication
-by $|\bx-\bx_s|$, that is,
-\begin{equation}
-  \label{eqn:defann}
-  Af(\bx,t;\bx_s) = |\bx-\bx_s|f(\bx,t;\bx_s).
-\end{equation}
-The space-time extension approach consists of solving the optimization
-problem:
-\begin{equation}
-  \label{eqn:defste}
-  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }f \mbox{ to
-    minimize }
-  \| d -  S[\kappa]f\|^2+\alpha^2\|Af\|^2.
-\end{equation}
-This problem can be reformulated in the same way as the WRI problem. Write
-$f = w\delta(\cdot-\bx_s) + S[\kappa]^Te + n$, where $n$ is
-non-radiating as before: then the problem \ref{eqn:defste} is
-equivalent to
+source), and note that the decomposition is orthogonal. Then
 \[
-   \mbox{Given } d, \mbox{ find }\kappa, w, e \mbox{ and }n \mbox{ to
-     minimize }
+  J[\kappa] =
+  \mbox{min}_{e,n} \frac{1}{2}(\| d-S[\kappa](S[\kappa]^Te+w_*\delta(\cdot-z_s))\|^2+\alpha^2(\|S[\kappa]^Te\|^2+\|n\|^2))
   \]
 \begin{equation}
-  \label{eqn:defstereform}
-  \| d -  S[\kappa]w_*\delta(\cdot-\bx_s) - S[\kappa]S[\kappa]^Te\|^2+\alpha^2\|A(S[\kappa]^Te+n)\|^2.
+  \label{eqn:defvpmred}
+   \mbox{min}_{e} \frac{1}{2}(\|r-S[\kappa]S[\kappa]^Te\|^2+\alpha^2\|S[\kappa]^Te\|^2)
  \end{equation}
-Note that the non-radiating component $n$ must remain in the second
-term of the objective, because there is no reason for $n$ to be
-$A^TA$-orthogonal to the range of $S[\kappa]^T$. Also, the isotropic
-point radiator component $w\delta(\cdot-\bx_s)$ must also remain as
-part of the unknowns - simply because it is unknown! It does not
-appear in the second term because $A$ annihilates it.
+ in which $r[\kappa]=d-S[\kappa]w_*\delta(\cdot-z_s)$ is the data residual.
 
-\subsubsection{Surface Source Extension}
-Beside the reformulation of WRI described above, it is also possible
-to avoid the storage problem of WRI by limiting the support of the
-extended source $f$ to lower dimensional sets. Of several
-possibilities discussed in the literature, I will discuss the 
-  surface source extension approach introduced by
-\cite{HuangNammourSymesDollizal:SEG19}. The extension in question is
-defined in equation \ref{eqn:defss}, and denoted $S_s$. The same
-annihilator \ref{eqn:defann} is used in this approach, this time
-interpreted as acting in the plane $z=z_s$, that is,
+This reformulation has some computational advantages \cite[]{WangYingst:SEG16,Herrmann:SEG19}, but also leads to a useful analytic transformation of the WRI problem.
+The minimizer on the RHS of equation \ref{eqn:defvpmred} is the solution $e=e_{\alpha}[\kappa]$ of the normal equation
 \[
-  Au(x,y,t;x_s,y_s) = \sqrt{(x-x_s)^2+(y-y_s)^2}u(x,y,t;x_s,y_s).
-\]
-or
-\[
-  Au(x,t;x_s) = (x-x_s)u(x,t;x_s)
-\]
-for 2D. 
-The problem formulation is
-\begin{equation}
-  \label{eqn:defsse}
-  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }u \mbox{ to
-    minimize }
-  \| d -  S_s[\kappa]u\|^2+\alpha^2\|Au\|^2.
-\end{equation}
-Evidently the extended source has as many parameters as the data, so
-the memory complexity of this problem is comparable to that of FWI or
-reformulated WRI. The computational complexity compares favorably to
-that of reformulated WRI.
-
-\subsection{Layered Media, Plane Waves}
-As explained in the introduction, this study explores the properties
-of the various FWI variants explained above for the special case of
-layered media: $\kappa = \kappa(z)$,
-$\rho=\rho(z)$. Then the modeled shots differ from
-each other by a horizontal translation, so that there is in effect
-only one shot, for which I take the location to be
-$\bx_s=(0,0,z_s)$.
-
-If $\log \kappa$ and $\log \rho$ are globally bounded and $f$ has
-compact support (and some other properties that will be made precise
-later), there is $\xi_{\rm max} > 0$ so that for
-$\|\xi\|<\xi_{\rm max}$, the integrals defining slant stack fields
-$\tilde{p},\tilde{\bv}$, namely
-\begin{eqnarray}
-\label{eqn:slant}
-\tp(\xi,z,t)& = &\int dx\, dy\, p(x,y,z,t+\xi x),\nonumber\\
-\tv(\xi,z,t)&=&\int dx\, dy\, v_z(x,y,z,t+\xi x),\nonumber\\
-\tf(\xi,z,t)&=&\int dx\,dy\, f(x,y,z,t+\xi x),
-\end{eqnarray}
-are absolutely convergent, and $\tp, \tv$ satisfy the plane wave
-version of the acoustic system \ref{eqn:awe}:
-\begin{eqnarray}
-\label{eqn:awepw}
-(1-c^2\xi^2)\frac{\partial \tp}{\partial t} + \kappa\frac{\partial
-  \tv}{\partial z} &=& \tf \nonumber\\
-\rho \frac{\partial \tv}{\partial t} + \frac{\partial \tp}{\partial
-  z}&=&0\nonumber\\
-\tp,\tv&=&0, t \ll 0.
-\end{eqnarray}
-Here $c=\sqrt{\kappa/\rho}$ is the sound velocity. I use $\xi$ instead
-of the more usual $p$ for plane wave slowness since $p$ already has
-been assigned to mean pressure. Note that for a isotropic point
-source, respectively a surface source,
-\begin{eqnarray}
-  f(\bx,t;\bx_s)=w(t)\delta(\bx) & \Rightarrow &
-                                               \tf(\xi,z,t)=w(t)\delta(z)\\ \nonumber 
-  f(\bx,t;\bx_s) = u(x,y,t)\delta(z) & \Rightarrow &
-                                                     \tf(\xi,z,t) =
-                                                     \tu(\xi,t)\delta(z)\\
-  \nonumber
-  \tu(\xi,t) &=& \int dx\,dy\, u(x,y,t+\xi x)
-\end{eqnarray}
-The plane wave domain modeling
-operator for space-time extended sources samples $\tp$ on $z=z_r$:
-\begin{equation}
-\label{eqn:modpw}
-\tS[\kappa]\tf = \tp|_{z=z_r}
-\end{equation}
-Similarly, the plane wave domain versions of $S_p$ and $S_s$ are
-\begin{eqnarray}
-  \label{eqn:pwpt}
- \tf(\xi,z,t) = w(t)\delta(z-z_s) & \Rightarrow & \tS[\kappa]\tf = \tS_p
-  [\kappa]w\\ 
-  \label{eqn:pwss}
-\tf(\xi,z,t) = \tu(\xi,t)\delta(z-z_s) & \Rightarrow & \tS[\kappa]\tf =
-  \tS_s[\kappa]\tu
-\end{eqnarray}
-
-For homogeneous $\kappa=\kappa_0, \rho=\rho_0, c=c_0$, a computation
-recapitulated in the Appendix C, relying on computations for the 1D
-acoustics system detailed in Appendices A and B, yields
-\begin{eqnarray}
-  \label{eqn:ptmodhompw}
-  \tS_p[\kappa_0]w(\xi,t) & = &\frac{1}{2c_0(1-c_0^2\xi^2)^{1/2}}w
-                                \left(t-\frac{|z_r-z_s|}{\tc_0(\xi)}\right)\\
-    \label{eqn:ssmodhompw}
-  \tS_s[\kappa_0]\tu(\xi,t) & = &\frac{1}{2c_0(1-c_0^2\xi^2)^{1/2}}\tu 
-                                  \left(\xi,t-\frac{|z_r-z_s|}{\tc_0(\xi)}\right)\\
-    \label{eqn:stmodhompw}
-  \tS[\kappa_0]\tf(\xi,t) & = &  \frac{1}{2c_0(1-c_0^2\xi^2)^{1/2}}\int dz \tf\left(\xi,z,t -
-                                \frac{|z_r-z|}{\tc_0(\xi)}\right)                              
-\end{eqnarray}
-Here
-\begin{equation}
-  \label{eqn:vertvel}
-  \tc(\xi,z) = \frac{c(z)}{\sqrt{1-c(z)^2\xi^2}}
-    \end{equation}
-is the vertical plane wave velocity, and we write
-$\tc_0(\xi)=\tc(\xi,z)$ for any $z$, when $c=c_0$.
-
-\subsection{Extended FWI in Plane Wave Domain}
-Of the various extended FWI formulation explained above, three do not
-require the storage of full space-time fields:
-\begin{itemize}
-\item Least-squares FWI;
-\item Reformulated WRI; and
-\item Surface Source Extension.
-\end{itemize}
-The remainder of this paper is devoted to analysis of these three
-approaches.
-
-It is possible to transform the objectives posed in the space-time
-domain problem statements above, but this leads to technical
-complications that do not affect the conclusions. Therefore I pose
-analogous problems directly in the plane wave domain. The $L^2$ norms
-(integral root mean squares) below involve integration over a range of
-plane wave slownesses $\xi$, which for simplicity I assume symmetric,
-that is, $-\xi_{\rm max} \le \xi \le \xi_{\rm max}$, and over
-appropriate ranges of other variables which I do not state explicitly
-here (see the analysis sections later in the paper where all such
-choices are made explicitly). As before, the $L^2$-norm squared symbol
-$\|\cdot\|^2$ means the integral over all variables of the square of
-whatever is inside.
-
-\subsubsection{Least Squares FWI}
-
-\begin{equation}
-  \label{eqn:defolspw}
-  \mbox{Given } \td, \mbox{ find }\kappa \mbox{ and }w \mbox{ to
-    minimize }
-  \| \td -  \tS_p[\kappa]w\delta(\cdot-z_s)\|^2.
-\end{equation}
-
-\subsubsection{Reformulated WRI}
-
-\begin{equation}
-  \label{eqn:defwrireformpw}
-  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }e \mbox{ to
-    minimize }
-  \| \td -  \tS[\kappa]w_*\delta(\cdot-z_s) - \tS[\kappa]\tS[\kappa]^Te\|^2+\alpha^2\|\tS[\kappa]^Te\|^2.
-\end{equation}
-
-The reformulated WRI objective involves 
-the product $\tS[\kappa]\tS[\kappa]^T$ in which $\tS[\kappa]^T$ is the
-formal or $L^2$ adjoint. For the homogeneous case,
-$\kappa=\kappa_0,\rho=\rho_0,c=c_0$,
-\begin{equation}
-  \label{eqn:stmodhompwtrans}
-  (\tS[\kappa]^T\te)(\xi,z,t)=\frac{1}{2}(1-c_0^2\xi^2)^{-1}\te\left(\xi,t +
-    \frac{|z_r-z|}{\tc_0(\xi)}\right).
-\end{equation}
-
-While I have endeavoured to limit mathematical technicalities in this background
-presentation, one such technicality intrudes at this point: the
-output of the transpose operator $\tS[\kappa]^T$ is nonzero and non-decaying arbitrarily
-far in space and in negative time. As will be obvious shortly, this
-fact prevents the composition $\tS[\kappa]\tS[\kappa]^T$ from being
-well-defined, without some further constraint being imposed on the
-domain of $\tS[\kappa]$. The simplest appears to be the assumption
-that all plane wave sources act over a limited depth range, $0 \le z
-\le z_{\rm max}$. Then the $z$ integrals appearing in the expressions
-for $S$ and $\tS$ have limits $[0,z_{\rm max}]$
-
-With this amendment, compute the composition:
-\[
-  \tS[\kappa]\tS[\kappa]^T\te(\xi,t) = \frac{1}{2}(1-c_0^2\xi^2)^{-1}
-\]
-\[
-  \times 
-  \int_0^{z_{\rm max}} dz\, \frac{1}{2}(1-c_0^2\xi^2)^{-1}\te\left(\xi,t' +
-    \frac{|z_r-z|}{\tc_0(\xi)}\right)_{t'=t-\frac{|z_r-z|}{\tc_0(\xi)}}
-\]
-\begin{equation}
-  \label{eqn:stmodhompwnorm}
- = \frac{z_{\rm max}}{4}(1-c_0^2\xi^2)^{-2}\te(\xi,t)
-\end{equation}
-
-\subsubsection{Surface Source Extension}
-
-The surface source space-time formulation employs an
-annihilator $A$, defined in \ref{eqn:defste}. The corresponding
-operator in the plane wave domain is somewhat easier to compute for the 2D
-case: one obtains 
-\[
-\tA = D_{\xi}I_t,
-\]
-that is, differentiation in $\xi$ and integration in $t$, in either
-order. The 3D expression is somewhat more complicated. Since the
-examples below will be computed in 2D, the above definition suffices.
-
-The surface source extended FWI problem in the plane wave domain is then
-\begin{equation}
-  \label{eqn:defssepw}
-  \mbox{Given } \td, \mbox{ find }\kappa \mbox{ and }\tu \mbox{ to
-    minimize }
-  \| \td -  \tS_s[\kappa]\tu\|^2+\alpha^2\|\tA\tu\|^2.
-\end{equation}
-
-\subsection{Inversion Velocity Analysis: the zero penalty limit}
-Examination of the operators appearing in the extended inversion
-problems considered here (equations \ref{eqn:stmodhompwnorm} for
-reformulated WRI, \ref{eqn:ssmodhompw} for SSE) reveals that they
-are both invertible. This property permits a simplification of the
-penalty formulations \ref{eqn:defwrireformpw} and
-\ref{eqn:defssepw}. This simplification has been used several times in
-prior work on extended FWI
-\cite[]{Symes:EAGE15,Herve2017,HouSymes:Geo18}, and has come to be
-called Inversion Velocity Analysis (IVA) in the exploration seismology
-literature.
-
-The idea is well-known in numerical optimization, and can most easily
-be explained in the context of the SSE objective function, in
-plane-wave domain:
-\begin{equation}
-  \label{eqn:ssepen}
-  \kappa,\tu \mapsto \|\td-\tS_s[\kappa]\tu\|^2 + \alpha^2\|\tA\tu\|^2
-\end{equation}
-
-The transformation to IVA takes place in two steps. The first is
-trivial: simply divide the objective function by $\alpha^2$, to
-obtain
-\begin{equation}
-  \label{eqn:ssepeninv}
-  \kappa,\tu \mapsto \alpha^{-2}\|\td-\tS_s[\kappa]\tu\|^2 + \|\tA\tu\|^2
-\end{equation}
-Only the minimum values change: for any particular $\alpha \in
-\bR^+$, the sets of local and global minimizers are exactly the same.
-
-The second step in moving from this objective to the corresponding IVA
-objective is of value in itself: it consists of eliminating the linear
-variable $\tu$ by minimizing the function defined in \ref{eqn:ssepen}
-over $\tu$. The resulting minimum is a function of $\kappa$: {\em a
-  priori}, a local minimizer is the same as the $\kappa$ component of
-a local minimizer of the original objective. This reduction of the
-objective to a function of some of its variables is known as the {\em
-  Variable Projection Method} (VPM), especially if the eliminated variables
-appear linearly in the original problem, as they do here
-\cite[]{GolubPereyra:03}. For the plane wave surface source objective,
-the result of VPM reduction is
-\begin{equation}
-  \label{eqn:ssevpm}
-  \kappa \mapsto \alpha^{-2}\|\td-\tS_s[\kappa]\tu[\kappa]\|^2 + \|\tA\tu[\kappa]\|^2
-\end{equation}
-in which $\tu[\kappa]$ is the minimizer of the base objective
-\ref{eqn:ssepeninv} over $\tu$, with $\kappa$ held fixed:
-$\tu[\kappa]$ solves the {\em normal equation}
-\begin{equation}
-  \label{eqn:ssenorm}
-  (\alpha^{-2}\tS_s[\kappa]^T\tS_s[\kappa] +
-  \tA^T\tA)\tu[\kappa]=\alpha^{-2}\tS_s[\kappa]^T\td.
-\end{equation}
-
-The major import of VPM is that the gradient
-of the reduced objective, a key element in any calculus-based
-optimization algorithm, is relatively simple to compute: formally, it
-is simply the gradient of the original objective with the
-to-be-eliminated variables held fixed. See \cite{GolubPereyra:03} for
-an explanation, and
-\cite{vanLeeuwenMulder:09,Rickett:SEG12,LiRickettAbubakar:13} for
-applications to geophysical inversion.
-
-The final step is to observe that the VPM objective converges as
-$\alpha \rightarrow 0$. Examination of \ref{eqn:ssepeninv} suggests
-that for this to happen, the data must be fit perfectly, at least in
-the limit. This is likely to happen only if the linear modeling
-operator ($\tS_s[\kappa]$, in this case) is invertible - which it is,
-at least in the case of constant $\kappa=\kappa_0$,
-see equation \ref{eqn:ssmodhompw}.  In fact $\tS[\kappa]$ is a
-continuous function of $\kappa$, as is explained in Appendix \appSingular, so in fact data is fit exactly and by precisely one $\tu$ for
-all $\kappa$ sufficiently close to constant. It is not hard to see that the
-$\alpha = 0$ limit of the VPM objective obtained from 
-\ref{eqn:ssepeninv} is
-\begin{equation}
-  \label{eqn:ssiva}
-  \kappa \mapsto \|A \tS_s[\kappa]^{-1}\td\|^2
-\end{equation}
-That is, the IVA objective is the mean square of the annihilator ouput
-from the necessarily unique extended source field $\tu =
-\tS_s[\kappa]^{-1}\td$ that precisely fits the data.
-
-Appendix \appIVA  presents a proof that equation \ref{eqn:ssiva}
-actually describes the $\alpha \rightarrow 0$ limit. The version of
-surface source extended FWI whose properties are described in the next
-section is the minimization problem for this objective, to which I
-give a name:
-\begin{equation}
-  \label{eqn:defssiva}
-  J_{\rm SSE} = \frac{1}{2}\|A\tu[\kappa]\|^2, \,
-  \tu[\kappa]=\tS_s[\kappa]^{-1}\td
-\end{equation}
-
-The story for reformulated WRI is similar. From equation
-\ref{eqn:defwrireformpw}, the VPM objective after scaling by
-$\alpha^{-2}$ is 
-\begin{equation}
-  \label{eqn:wrivpm}
-  \kappa, \te \mapsto \alpha^{-2}\| \td -  \tS[\kappa]w_*\delta(\cdot-z_s) -
-  \tS[\kappa]\tS[\kappa]^T\te[\kappa]\|^2+\|\tS[\kappa]^T\te[\kappa]\|^2,
-\end{equation}
-in which $\te[\kappa]$ solves the normal equation
-\begin{equation}
-  \label{eqn:wrinorm}
-  (\alpha^{-2}(\tS[\kappa]\tS[\kappa]^T)^2 +
-  \tS[\kappa]\tS[\kappa]^T)\te[\kappa] =
-  \alpha^{-2}\tS[\kappa]\tS[\kappa]^T(\td -
-  \tS[\kappa]w_*\delta(\cdot-z_s) )
-\end{equation}
-If $\kappa$ is close to $\kappa_0$, then $\tS[\kappa]\tS[\kappa]^T$ is
-invertible, as follows from equation \ref{eqn:stmodhompwnorm} and the
-continuity of $\tS[\kappa]$ with respect to $\kappa$, mentioned
-above. Therefore the normal equation is equivalent to
-\begin{equation}
-  \label{eqn:wrinorm1}
-  (\alpha^{-2}\tS[\kappa]\tS[\kappa]^T +
-  I)\te[\kappa] =
-  \alpha^{-2}(\td -
-  \tS[\kappa]w_*\delta(\cdot-z_s) )
-\end{equation}
-The $\alpha \rightarrow 0$ limit of the resulting function of
-$\kappa$ is
-\[
-  J_{\rm WRI}[\kappa] = \frac{1}{2}\|\tS[\kappa]^T((\tS[\kappa]\tS[\kappa]^T)^{-1}(\td -
-  \tS[\kappa]w_*\delta(\cdot-z_s)))\|^2
-\]
-\begin{equation}
-  \label{eqn:wripwiva}
-  = \frac{1}{2}\langle \td -
-  \tS[\kappa]w_*\delta(\cdot-z_s), (\tS[\kappa]\tS[\kappa]^T)^{-1}( \td -
-  \tS[\kappa]w_*\delta(\cdot-z_s)) \rangle
-\end{equation}
-
-Finally, it is entirely possible to apply VPM to the least squares FWI
-objective \ref{eqn:defolspw}, as has been suggested as an algorithmic
-approach to determining source parameters as part of FWI
-\cite{Rickett:SEG12,LiRickettAbubakar:13}. However in this instance it
-is more relevant to pose the known-source version of least squares
-FWI, as most parallel to WRI (in any of its forms): recalling the
-identity \ref{eqn:pwpt}, 
-\[
-  J_{\rm FWI}[\kappa] = \frac{1}{2}\|\td-\tS_p[\kappa]w_*\|^2 
-\]
-\begin{equation}
-  \label{eqn:olspwobj}
-  = \frac{1}{2}\langle \td -
-  \tS[\kappa]w_*\delta(\cdot-z_s), \td -
-  \tS[\kappa]w_*\delta(\cdot-z_s)\rangle.
-\end{equation}
-Comparing equations \ref{eqn:wripwiva} and \ref{eqn:olspwobj}, reach
-the surprising conclusion that
-\begin{quote}
-  The $\alpha \rightarrow 0$ limit of the WRI-VPM objective
-  \ref{eqn:wrivpm} is a weighted-norm variant of the FWI objective
-  \ref{eqn:olspwobj}.
-\end{quote}
-This conclusion is surprising because one would normally expect the
-opposite limit, that is $\alpha \rightarrow \infty$, of the WRI-VPM
-objective \ref{eqn:wrivpm} to be equivalent in some sense to the FWI
-objective \ref{eqn:olspwobj}, rather than the $\alpha \rightarrow 0$
-limit. Indeed, the limit $\alpha \rightarrow \infty$ can only make
-sense if $\tS[\kappa]\tS[\kappa]^T\te \rightarrow 0$: since
-$\tS[\kappa]\tS[\kappa]^T$ is invertible for $\kappa$ near $\kappa_0$,
-in that case $\te \rightarrow 0$ necessarily, so the objective
-\ref{eqn:wrivpm} presumably tends to \ref{eqn:olspwobj}. This conclusion
-is in fact correct, as follows from a variant of the argument
-presented in Appendix \appIVA.
-
-The import of the reasoning presented in Appendix \appIVA is that for
-$\alpha$ sufficiently small, the behaviour of the VPM objective in
-each of these cases is close to that of the $\alpha \rightarrow 0$
-limit. As is well-known, and will be recapitulated in the Main Results
-section, the least squares FWI objective does not generate
-constructive model updates. Since all of the VPM objectives explored
-here become close to the least squares FWI objective for large
-$\alpha$, it is the case of small $\alpha$ that holds any possible
-promise of generating substantially more constructive updates, and we
-will see below that indeed in some cases that promise is realized. The
-$\alpha \rightarrow 0$ limit provides a convenient window into
-small $\alpha$ behaviour, so the focus in the remainder of the paper
-will be on the limiting problems defined in \ref{eqn:olspwobj},
-\ref{eqn:wripwiva}, and \ref{eqn:defssiva}.
-
-\section{Main Results}
-
-As explained in the introduction, the wave propagation regime implies
-a separation of scales, between the propagation distance ($O(1)$ in
-appropriate units), and the (smaller) wavelength scale. The results of
-this paper are valid for ``small'' wavelengths, that is, as the
-wavelength parameter $\lambda \rightarrow 0$. Wavelength is a data
-property, induced by a corresponding feature of the source. Recall
-the assumption that the data is (well-approximated by) point-source
-data for an appropriate choice of acoustic parameter fields $\kappa$,
-$\rho$. A convenient way of introducing wavelength as a parameter in
-the acoustic dynamic fields is via a {\em family} of source wavelets $\wl$:
-\[
-\wl(t)=\frac{1}{\sqrt{\lambda}}w_1\left(\frac{t}{\lambda}\right)
-\]
-These wavelets are scaled to have constant power (mean-square)
-independent of $\lambda$. As is reasonable, even mandatory, for simulation of seismic data, 
-assume that the ``mother wavelet'' $w_1$ has at least two vanishing 
-moments, that is, vanishes at zero frequency, and is smooth, that is,
-also vanishes at zero wavelength. Also assume that the source acts
-over a finite time interval.
-
-Correspondingly, the target data is also a $\lambda$-dependent family:
-\begin{equation}
-  \label{eqn:datadef}
-  \tdl = \tS_p[\kappa_*]\wl.
-\end{equation}
-in which $\kappa_*$ is the target bulk modulus, an estimate of which
-is the object of the inversion exercise.
-
-For each of the approaches mentioned in the last section, I will
-analyze two cases, transmission and reflection. In all cases, $\kappa$
-obeys Assumptions A - C, so that the system \ref{eqn:awepw} and its relatives have
-well-defined solutions, thanks to the constructions presented in
-Appendix \appSingular.
-
-\paragraph{Transmission:} $z_s < z_r$, $\kappa_*$ is smooth and $ = \kappa_0$ for
-$z$ near $z_s$, $\ge \kappa_0$ for all $z$, and $>\kappa_0$ for $z$
-near $z_r$, Geometric acoustics implies that the solution is given by
-the progressing wave expansion [REFERENCE] in terms of the vertical
-travel time $\tau(\xi,z_1,z_2)$ and geometric amplitude
-$a(\xi,z_1,z_2)$ between depths $z_1$ and $z_2$. The traveltime is given by
-\[
-\tau(\xi,z_1,z_2) = \int_{z_1}^{z_2}\,dz\, \frac{1}{\tc(\xi,z)}.
-\]
-I do not give an explicit expression for the amplitude, noting only
-that it is smooth in all three variables and positive. The geometric
-acoustics approximation of the data is
-\begin{equation}
-  \label{eqn:geosoltrans}
-  \tdl(\xi,t) = a(\xi,z_s,z_r)\wl(t-\tau(\xi,z_s,z_r)).
-\end{equation}
-The explicit expression \ref{eqn:ptmodhompw} for $\kappa_0$ may be
-viewed as a special case of \ref{eqn:geosoltrans}, and gives reference
-data for that case:
-\begin{equation}
-  \label{eqn:analsoltrans}
-  \td_{0,\lambda}(\xi,t) =  \tS_p[\kappa_0]\wl(\xi,t)  = \frac{1}{2c_0(1-c_0^2\xi^2)^{1/2}}\wl
-                                \left(t-\frac{|z_r-z_s|}{\tc_0(\xi)}\right)
-\end{equation}
-
-\paragraph{Reflection:}
-In the reflection configuration, $z_r=z_s$, $\kappa_*(z)=\kappa_0$ for
-$z<z_s$, and $\kappa_*$ is smooth except for a non-zero jump at $z=z_j
-> z_s$. Geometric acoustics provides an approximate solution
-containing a reflection for $z<z_j$, in particular at $z=z_s=z_r:$
-\[
-\tS[\kappa_*](\xi,t) = a(\xi,z_s,z_s)\wl(t) +
-R(\xi)\wl\left(t-2\tau(\xi,z_s,z_j)\right) + O(\lambda)
-\]
-The reflection coefficient $R(\xi)$ is a function of the values of
-$\kappa_*$ on either side of the jump, and of $\rho_0$. It is
-generically non-zero except possibly for isolated values of $\xi$. Its precise
-value is of no concern.
-
-Having described these two configurations, I will now discuss the
-behaviour of the three approaches to waveform inversion described in
-the previous section, in each configuration. The crux of the matter is
-the bahaviour of the gradient of the objective function in each case,
-computed at the reference model $\kappa_0$,
-as the wavelength parameter $\lambda \rightarrow 0$. Each objective -
-the basic least squares objective in the first case, the IVA
-objectives for the other two - is stable, in fact approaches a non-
-zero limit. If the size of the gradient tends to zero, on the other
-hand, then the gradient contains vanishingly little information about
-constructive updates of $\kappa$ and gradient-based optimization
-algorithms can be expected to perform poorly.
-
-\subsection{Least Squares FWI}
-
-wri
-
-sse
-
-\section{Derivations of Main Results}
-
-\subsection{Least Squares FWI, Transmission Configuration}
-
-The gradient of the objective, given in equation \ref{eqn:olspwobj},
-is
-\[
-  \nabla J_{\rm FWI}[\kappa] = D\tS[\kappa]^T(\tS[\kappa]\tw_*-d)
-\]
-
-For homogeneous background model $\kappa=\kappa_0$, Appendix \appPlane
-develops an explicit expression 
-\[
-D(\tS_p[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
-  -\frac{1}{2\kappa_0c_0(1-c_0^2\xi^2)^{1/2}} \delta 
-  \kappa(z_s) w\left(t - \frac{|z_r-z_s|}{\tc_0(\xi)}\right)
-\]
-\begin{equation}
-\label{eqn:dmodhompwpt}
-    + \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int \,dz\,\delta 
-  \kappa(z) \frac{\partial w}{\partial
-    t}\left(t - \frac{|z_s-z|+|z-z_r|}{\tc_0(\xi)}\right)
-\end{equation}
-The conditions imposed on the transmission configuration imply that
-$\delta \kappa(z) = 0$ for $z$ near $z_s$. This constraint implies that
-\begin{equation}
-  \label{eqn:dmodhompwpt0}
-  D(\tS_p[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
-     \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int \,dz\,\delta 
-  \kappa(z)  \frac{\partial w}{\partial
-    t}\left(t - \frac{|z_s-z|+|z-z_r|}{\tc_0(\xi)}\right).
-\end{equation}
-Thus
-\begin{equation}
-\label{eqn:dmodhomadj}
-D(\tS[\kappa]\tw)_{\kappa=\kappa_0}^T\td(z) =
-\int_{-\xi_{\rm max}}^{\xi_{\rm max}}\,d\xi\,\int\,dt\,
-\frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)}
-\frac{\partial w}{\partial
-    t}\left(t - \frac{|z_s-z|+|z-z_r|}{\tc_0(\xi)}\right) \td(\xi,t).
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{SSE Objective and Gradient: vanishing low frequencies}
-In this section I will formulate the SSE objective, and calculate its
-gradient at homogeneous background for single reflector data as
-described in the last section. I will show that this gradient has
-low-frequency content as the data wavelength $\lambda$, introduced in
-the last section and in the introduction, tends to zero.
-
-FWI via SSE requires specification of an {\em annihilator}, the null
-space of which consists of non-extended models, as explained in the
-introduction.  Construct a suitable plane-wave domain annihilator
-$\tA$ for
-SSE by intertwining the Radon transform $R$ with the space-domain
-annihilator $A = $ multiplication by offset. In order to do this
-properly, it is necessary to introduce a cutoff in offset: not a
-hardship, since infinite offsets are unavailable. Denote by $\chi$ a
-offset mute. Define
-\[
-Rw(\chi,t) = \int \,dx\,\chi(x) w(x,t+\xi x)
-\]
-\[
-Aw(x,t)=x w(x,t)
-\]
-\[
-\tilde{A} Rw (\xi,t) = R Aw (\xi,t)
-\]
-\[
-=\int dx x \chi(x) w(x,t+\xi x) = \int dx \chi(x)\frac{\partial}{\partial \xi}
-I_tw(x,t+\xi x)
-\]
-\[
-= D_{\xi} I_t R  w \equiv \tilde{A} Rw (\xi,t) 
-\]
-
-Since $\tS[\kappa]$ is continuous in $\kappa$
-\cite[]{BlazekStolkSymes:13}, and $\tS[\kappa_0]$ is a multiple of the
-identity, $\tS[\kappa]$ is invertible in sup-norm neighborhood of
-$\kappa$.
-Now assume that $\td$ is data computed from a simple reflector as
-described in the previous section, depending on the wavelength
-parameter $\lambda$:
-\[
-\td_{\lambda}(\xi,t) = a_0(\xi)\wl(t) +
-R(\xi)\wl\left(t-2\tau(\xi,z_1)\right) + O(\lambda)
-\]
-Then
-\begin{equation}
-\label{eqn:wtildedef}
-\tS[\kappa_0]^{-1}\td_{\lambda}(\xi,t) = \wl(t) +
-\frac{R(\xi)}{a_0(\xi)}\wl\left(t-2\tau(\xi,z_1)\right) + O(\lambda)
-\equiv \tw(\xi,t) + O(\lambda)
-\end{equation}
-Write 
-\[
-a_1(\xi)=\frac{1}{\kappa_0 (1-c_0^2\xi^2)^2}, \tc_0(\xi) =
-\frac{c_0}{\sqrt{1-c_0^2\xi^2}}
-\]
-%Then
-%\[
-%DJ[\kappa_0]\delta \kappa =
-%\left\langle \frac{c_0}{16\kappa_0 (1-c_0^2\xi^2)^{5/2}}\int
-%\,ds\,\tw(\xi,t-s)(\delta \kappa)'
-%\left(\frac{c_0s}{2\sqrt{1-c_0^2\xi^2}}\right), D_{\xi}^2I_t^2
-%\tw(\xi,t) \right\rangle
-%\]
-%Per the derivation of simple reflection, presume that (1) $w_1(t) = 0$
-%for $t>1$, and $w_1$ (hence $\wl$) has two vanishing moments, and (2) $\delta \kappa = 0 $ in an interval of positive
-%length about $z=0$, and (3) $\lambda$ is small enough that $\wl(t)=0$
-%if $\delta(\kappa)(t\tc/2) \ne 0$ for all $\xi$ for which $\chi(\xi)
-%\ne 0$ (I.e. the range of the $\xi$ integral).
-
-%Then the convolution of $\delta \kappa(t\tc/2)$ and $\wl$ vanishes for
-%small $t$, hence the product with $\wl(t)$
-
-%and
-
-From the definition \ref{eqn:j0def} of $J_0$ and a simple calculation,
-\[
-\nabla J[\kappa_0,\td] = (D\tS[\kappa_0]\tw)^T(\tS[\kappa_0]^T)^{-1}\tA^T\tA
-\tS[\kappa_0]^{-1}\td,  \,\tw=\tS[\kappa_0]^{-1}\td
-\]
-Since $D_{\xi}$ and $I_t$ are both skew-symmetric and commute, $\tA$
-is symmetric. $\tS$ is $\xi-$ dependent scaling. For simplicity
-abbreviated the relation \ref{eqn:modhom} as
-\[
-\tS[\kappa_0]\tw(\xi,t) = a_0(\xi)\tw(\xi,t)
-\]
-Then
-\[
-\nabla J[\kappa_0](z) =
-- 
-\frac{1}{8} \int \,d\xi\,a_1(\xi)\frac{\partial}{\partial z} \int \,dt\, 
-  \tw\left(\xi,t-\frac{2z}{\tc_0}\right)\left(D_{\xi}^2I_t^2 \tw(\xi,t)\right)
-\]
-
-\begin{equation}
-\label{eqn:jgrad}
-=-\frac{c_0}{16} \int \,d\xi\,a_1(\xi)\frac{\partial}{\partial z} \int \,dz'\, 
-  \tw\left(\xi,\frac{2(z'-z)}{\tc_0}\right)\left(D_{\xi}^2I_z^2
-    \tw\left(\xi,\frac{2z'}{\tc_0}\right)\right)
-\end{equation} 
-Set $u(\xi,z) = \tw(\xi,2z/\tc)$. Denote by $\hat{f}$ the Fourier
-transform of $f$.  
-\begin{equation}
-\label{eqn:jgradft}
-\widehat{\nabla J[\kappa_0]}(k) = -\frac{c_0}{16}\int\, d\xi\,a_1(\xi) 
-\frac{1}{ik}\hat{u}(\xi,k)D_{\xi}^2\overline{\hat{u}(\xi,k)}
-\end{equation}
-From the definition \ref{eqn:wtildedef} of $\tw$,
-\[
-\hat{\tw}(\xi,\omega) = \left(1 + \frac{R(\xi)}{a_0(\xi)}e^{2i\omega\tau(\xi,z_1)}\right)\hat{\wl}(\omega)
+ ( (S[\kappa]S[\kappa]^T)^2 + \alpha^2S[\kappa](S[\kappa]^T)e = S[\kappa]S[\kappa]^Tr[\kappa]
 \]
@@
 \[
-\hat{u}(\xi,k) = \left(1 +
-  \frac{R(\xi)}{a_0(\xi)}e^{4ik\tau(\xi,z_1)/\tc(\xi)}\right)\hat{\wl}\left(\frac{2k}{\tc(\xi)}\right) 
+  S[\kappa]S[\kappa]^Te_{\alpha}[\kappa] = S[\kappa]S[\kappa]^T(S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]
 \]
-Together with the expression \ref{eqn:jgradft} for the Fourier
-transform of the gradient, this observation establishes the assertion
-stated in the introduction.
+Assume that $S[\kappa]S[\kappa]^T$ is invertible - this is in fact for the problem outlined in the introduction, as will be verified shortly. Then
+\begin{equation}
+  \label{eqn:norsol}
+  e_{\alpha}[\kappa]=(S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]
+\end{equation}
+Consequently
+\[
+  J[\kappa] =
+  \frac{1}{2}(\|r[\kappa]-S[\kappa]S[\kappa]^Te_{\alpha}[\kappa]\|^2+\alpha^2\|S[\kappa]^Te_{\alpha}[\kappa]\|^2)
+\]
+\[
+  = \frac{1}{2}(\|r[\kappa]-S[\kappa]S[\kappa]^T (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]\|^2+\langle (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa],
+  S[\kappa]S[\kappa]^T (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]
+  \rangle
+\]
+\[
+  =\frac{1}{2}\left(\|\alpha^2 (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]\|^2
+  +\langle (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa],
+  -\alpha^2 (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]\rangle\right.
+\]
+\[
+  + \left.\langle (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa],r[\kappa]\rangle\right)
+\]
+\[
+  =  \frac{1}{2}\left(
+\langle (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa],r[\kappa]\rangle +
+\alpha^2(1-\alpha^2)\| (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]\|^2\right) 
+\]
+\[
+  =  \frac{1}{2}
+  \langle ((S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}+\alpha^2(1-\alpha^2)(S[\kappa]S[\kappa]^T+
+  \alpha^2I)^{-2})r[\kappa],r[\kappa]\rangle
+\]
+That is,
+\begin{equation}
+  \label{eqn:defwrialt}
+  J[\kappa] = \frac{1}{2}\langle r[\kappa], W[\kappa] r[\kappa]\rangle
+\end{equation}
+with
+\begin{equation}
+  \label{eqn:defwriwt}
+  W[\kappa] = (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}+\alpha^2(1-\alpha^2)(S[\kappa]S[\kappa]^T+
+  \alpha^2I)^{-2}.
+\end{equation}
+This remarkable identity shows that the WRI objective function is a {\em weighted norm of the data residual $r[\kappa]$}.
+Note that this conclusion depends only on the presumption that the operator $S[\kappa]S[\kappa]^T$ is invertible (or, more precisely, injective), a condition that will be verified below for the problem considered in this paper, and which holds in many other settings in which WRI has been advocated.
 
-\append{Singular RIght-hand Sides and Weak Solutions}
-Theorem 1 in \cite[]{BlazekStolkSymes:13} provides a strong solution of the system \ref{eqn:awe}: $p$, respectively $\bv$, is continuous
-in time with values in the Sobolev space $H^1(\bR^d)$, respectively
-$H^1_{\rm div}(\bR^d)$ and
-also continuously differentiable with values in $L^2(\bR^d)$,
-respectively $(L^2(\bR^d))^d$. Therefore all of the terms in
-\ref{eqn:awe} are square-integrable functions.
+\section{WRI for 1D Acoustics}
+The preceding section provides the necessary ingredients for an assessment of the relation between WRI and FWI.
+While the conclusion reached below applies to many wave propagation settings, the 1D acoustic setting is particularly simple and yet illustrates clearly the nature of this relation.
 
-Theorem 4 in \cite[]{BlazekStolkSymes:13} implies that for each $t$, $p(\cdot,t)$ and $\bv(\cdot,t)$
-have compact support, and that the support expands in $t$ at the rate
-bounded by the sup of the velocity $c=\sqrt{\kappa/\rho}$, as one
-would expect. Therefore, for any fixed max time $T$, it is possible to
-treat the problem for $0\le t \le T$ as a Dirichlet problem in a
-domain $\Omega$ of sufficiently large diameter. Without loss of
-generality, assume that $\mbox{supp }f \subset \Omega \times [0,T]$.
-
-The isotropic point radiator \ref{eqn:ipr} is not
-square-integrable, so the theory in \cite[]{BlazekStolkSymes:13} does
-not apply directly to the system
-\begin{eqnarray}
-\label{eqn:aweptsrc}
-\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
-                                                                w(t)\delta(\bx-\bx_s) \nonumber\\
-\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
-p,{\bf v}&=&0, t \ll 0
-\end{eqnarray}
-However there is a simple trick that makes this
-theory applicable, provided one assumes that $\kappa$ is constant
-($=\kappa_0 \in \bR^+$) near
-$\bx=\bx_s$ (Assumption C). With this constraint, the problem
-\ref{eqn:aweptsrc} may be shown to have a unique causal weak solution.,
-
-Note that the existence of weak solutions for problems of this type
-has been 
-well-known for decades, with the additional assumption that the coefficients are
-smooth, see for example \cite{Lax:PDENotes}. Using the progressing
-wave expansion of the fundamental solution for smooth coefficient
-problems (for example, \cite{Friedlander:75}) in place of the analytic
-expression for the constant-coefficient point radiator solution given
-below, it is possible to obtain a similar result under the weaker
-\begin{itemize}
-\item[Assumption C']: $\kappa$ is smooth near $\bx_s$.
-\end{itemize}
-
-Denote by $(p_0,\bv_0)$ the (weak) solution of the
-system \ref{eqn:awe} with $\kappa$ replaced by $\kappa_0$. For $d=3$,
-\[
- p_0(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-\frac{ |\bx-\bx_s|}{c_0}\right)
-\]
-in which $c_0=\sqrt{\kappa_0/\rho_0}$. There is a similar expression
-for $\bv_0$, and for other space dimensions $d$. Because $w$ is smooth, both $p$ and $\bv$ are smooth
-except at $\bx=\bx_s$. Pick $\phi \in C_0^{\infty}(\bR^d)$ so that
-$\phi = 1$ in a neighborhood of $\bx_s$ and $\kappa=\kappa_0$ on
-$\mbox{supp }\phi$. Then
-\begin{eqnarray}
-\label{eqn:awecut}
-\frac{\partial (\phi p_0)}{\partial t} + \kappa \nabla \cdot (\phi {\bf v}_0) &=&
-                           w(t)\delta(\bx-\bx_s) + f_0  \nonumber\\
-\rho \frac{\partial (\phi {\bf v}_0)}{\partial t} + \nabla (\phi p_0) = {\bf g}_0 \nonumber\\
-\end{eqnarray}
-in which $f_0$ and ${\bf g}_0$ are smooth of compact support. Denote by
-$(p_1,\bv_1)$ the strong solution of
-\begin{eqnarray}
-\label{eqn:awesupp}
-\frac{\partial p_1}{\partial t} + \kappa \nabla \cdot {\bf v}_1 &=&
-                                                                -f_0 \nonumber\\
-\rho \frac{\partial {\bf v}_1}{\partial t} + \nabla p_1 = -{\bf g}_0 \nonumber\\
-p_1,{\bf v}_1&=&0, t \ll 0
-\end{eqnarray}
-as provided by Theorem 1 of \cite{BlazekStolkSymes:13}. Then
-\begin{eqnarray*}
-  p & = & \phi p_0 + p_1\\
-  \bv & = & \phi \bv_0 + \bv_1
-\end{eqnarray*}
-defines the unique weak solution of the system \ref{eqn:aweptsrc}.
-
-\cite{SantosaSymes:00} use an analogous construction for the
-second-order wave equation to give $L^2$ bounds on the point radiator
-solution.
-
-As mentioned in the Preliminaries section, a similar argument
-establishes the existence of weak solutions for the surface source problem
-\begin{eqnarray}
-\label{eqn:awess}
-\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
-                                                                w(x,y,t)\delta(z-z_s) \nonumber\\
-\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
-p,{\bf v}&=&0, t \ll 0
-\end{eqnarray}
-under Assumption D, that $\kappa=\kappa_0$ near $z=z_s$. An analytic
-solution for $\kappa \equiv \kappa_0$ follows from that for the point
-radiator problem, by superposition. Then a similar cutoff construction
-and appeal to Theorem 1 of \cite{BlazekStolkSymes:13} proveds a weak
-solution to \ref{eqn:awess}.
-
-This construction also works via recourse to the progressing wave
-expansion for the weaker
-\begin{itemize}
-\item[Assumption D']: $\epsilon > 0$ exists for which $\kappa$ is
-smooth in the collar neigborhood $\{\bx: |z-z_s|<\epsilon\}$.
-\end{itemize}
+The first task is to give an explicit expression for the operator $S[c]S[c]^T$ appearing repeatedly in the expression \ref{eqn:defwriwt}.
+From the definition \ref{eqn:defmodhom}, it follows immediatlely that
+\begin{equation}
+  \label{eqn:stmodhompwtrans}
+S[c]^Te(z,t)=\frac{1}{2c}e\left(t +  \frac{|z_r-z|}{c}\right). 
+\end{equation}
+whence
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
   + \frac{1}{4\kappa_0c_0^2} \int\,dz_2 \, \int \,dz_1\,\delta 
-  \kappa(z_1) \frac{\partial f}{\partial t}\left(z_2,t - \frac{|z_2-z_1|+|z_1-z|}{c_0}\right)
+  \kappa(z_2) \frac{\partial f}{\partial t}\left(z_1,t - \frac{|z_1-z_2|+|z_2-z|}{c_0}\right)
 \end{equation}
 
-An important special case is the 1D point source,
+A convenient check on this calculation is available for constant
+$\delta \kappa = \delta \kappa_0$. Directly perturb the RHS of equation 
+\ref{eqn:sol1dp}, to obtain
+\[
+  \delta p(z,t) = -\frac{\delta c_0}{2 c_0^2} \int
+  \,dz_1\,f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
+  +\frac{\delta c_0}{2c_0^3}\int
+  \,dz_1\,\frac{\partial f}{\partial
+    t}\left(z_1,t-\frac{|z-z_1|}{c_0}\right)|z-z_1|
+\]
+Since
+\[
+  \delta c_0 = \delta \left(\frac{\kappa_0}{\rho_0}\right)^{1/2} =
+  \frac{1}{2} \left(\frac{\kappa_0}{\rho_0}\right)^{-1/2}\frac{\delta
+    \kappa_0}{\rho_0}
+\]
+\[
+  = \frac{1}{2}\frac{\delta \kappa \rho_0^{1/2}}{\kappa_0^{1/2}\rho_0}
+  \]
+\[
+=\frac{1}{2} \frac{\delta
+  \kappa_0}{\kappa_0}\left(\frac{\kappa_0}{\rho_0}\right)^{1/2}
+=\frac{1}{2} \frac{c_0\delta \kappa_0}{\kappa_0}
+\]
+the identity above can be rewritten as
+\[
+  \delta p(z,t) = -\frac{\delta \kappa_0}{4 \kappa_0 c_0} \int
+  \,dz_1\,f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
+  \]
+\begin{equation}
+  \label{eqn:dpraw} 
+  +\frac{\delta \kappa_0}{4\kappa_0c_0^2}\int
+  \,dz_1\,\frac{\partial f}{\partial
+    t}\left(z_1,t-\frac{|z-z_1|}{c_0}\right)|z-z_1|
+\end{equation}
+The critical quantity is the integral in the second term:
+\[
+  \int \,dz_1\,\frac{\partial f}{\partial t}\left(z_1,t-\frac{|z-z_1|}{c_0}\right)|z-z_1|
+\]
+\[
+  = \int_{-\infty}^z \,dz_1\,\frac{\partial f}{\partial
+    t}\left(z_1,t-\frac{z-z_1}{c_0}\right)(z-z_1)
+\]
+\[
+   + \int^{\infty}_z \,dz_1\,\frac{\partial f}{\partial t}\left(z_1,t-\frac{z_1-z}{c_0}\right)(z_1-z)
+ \]
+ \[
+   =\int_{-\infty}^z \,dz_1\,\int_{z_1}^{z} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{z-z_2+z_2-z_1}{c_0}\right)
+ \]
+ \[
+   + \int^{\infty}_z \,dz_1\,\int_{z}^{z_1}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{z_1-z_2 + z_2-z}{c_0}\right)
+ \]
+ Checking signs, one sees that this sum is equal to
+ \[
+   =\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   + \int^{\infty}_z \,dz_1\,\int_{-\infty}^{\infty}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{|z_1-z_2| + |z_2-z|}{c_0}\right)
+ \]
+ \[
+   -\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{z_1} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+\[
+   -\int_{-\infty}^z \,dz_1\,\int_{z}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   - \int^{\infty}_z \,dz_1\,\int_{-\infty}^{z}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{|z_1-z_2| + |z_2-z|}{c_0}\right)
+ \]
+ \[
+   - \int^{\infty}_z \,dz_1\,\int_{z_1}^{\infty}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{|z_1-z_2| + |z_2-z|}{c_0}\right)
+ \]
+\[
+   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   -\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{z_1} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{z-z_2+z_1-z_2}{c_0}\right)
+ \]
+\[
+   -\int_{-\infty}^z \,dz_1\,\int_{z}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{z_2-z+z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   - \int^{\infty}_z \,dz_1\,\int_{-\infty}^{z}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{z_1-z_2 + z-z_2}{c_0}\right)
+ \]
+ \[
+   - \int^{\infty}_z \,dz_1\,\int_{z_1}^{\infty}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{z_2-z_1 + z_2-z}{c_0}\right)
+ \]
+\[
+   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   -\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{z_1} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{z+z_1-2z_2}{c_0}\right)
+ \]
+\[
+   -\int_{-\infty}^z \,dz_1\,\int_{z}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{2z_2-z-z_1}{c_0}\right)
+ \]
+ \[
+   - \int^{\infty}_z \,dz_1\,\int_{-\infty}^{z}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{z_1+ z-2z_2}{c_0}\right)
+ \]
+ \[
+   - \int^{\infty}_z \,dz_1\,\int_{z_1}^{\infty}\,dz_2\,\frac{\partial
+     f}{\partial t}\left(z_1,t-\frac{2z_2-z_1-z}{c_0}\right)
+ \]
+\[
+   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   -\frac{c_0}{2}\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{z_1} \,dz_2\, \frac{\partial }{\partial
+     z_2}f\left(z_1,t-\frac{z+z_1-2z_2}{c_0}\right)
+ \]
+\[
+   +\frac{c_0}{2}\int_{-\infty}^z \,dz_1\,\int_{z}^{\infty} \,dz_2\, \frac{\partial}{\partial
+     z_2}f\left(z_1,t-\frac{2z_2-z-z_1}{c_0}\right)
+ \]
+ \[
+   - \frac{c_0}{2}\int^{\infty}_z \,dz_1\,\int_{-\infty}^{z}\,dz_2\,\frac{\partial
+     }{\partial z_2}f\left(z_1,t-\frac{z_1+ z-2z_2}{c_0}\right)
+ \]
+ \[
+   +\frac{c_0}{2} \int^{\infty}_z \,dz_1\,\int_{z_1}^{\infty}\,dz_2\,\frac{\partial
+     }{\partial z_2}f\left(z_1,t-\frac{2z_2-z_1-z}{c_0}\right)
+ \]
+\[
+   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   -\frac{c_0}{2}\int_{-\infty}^z \,dz_1\, f\left(z_1,t-\frac{z-z_1}{c_0}\right)
+   -\frac{c_0}{2}\int_{-\infty}^z \,dz_1\,f\left(z_1,t-\frac{z-z_1}{c_0}\right)
+ \]
+ \[
+   - \frac{c_0}{2}\int^{\infty}_z \,dz_1\,f\left(z_1,t-\frac{z_1-z}{c_0}\right)
+   -\frac{c_0}{2} \int^{\infty}_z \,dz_1\,f\left(z_1,t-\frac{z_1-z}{c_0}\right)
+ \]
+ \[
+   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+ \]
+ \[
+   -c_0\int_{-\infty}^{\infty} \,dz_1\, f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
+ \]
+ Insert the identity just established in the second term on the RHS of
+ \ref{eqn:dpraw}:
+ \[
+  \delta p(z,t) = -\frac{\delta \kappa_0}{4 \kappa_0 c_0} \int
+  \,dz_1\,f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
+\]
+\[
+  +\frac{\delta \kappa_0}{4\kappa_0c_0^2}\left(\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+      t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right) \right.
+\]
+\[
+  \left. -c_0\int_{-\infty}^{\infty} \,dz_1\,
+    f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)\right)
+\]
+\[
+=-\frac{\delta \kappa_0}{2\kappa_0c_0}\int_{-\infty}^{\infty} \,dz_1\,
+f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
+\]
+\begin{equation}
+  \label{eqn:dpraw}
++  \frac{\delta \kappa_0}{4\kappa_0c_0^2}\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
+      t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
+  \end{equation}
+which is identical to the expression \ref{eqn:sol1ddp} for the case of constant
+$\delta \kappa = \delta \kappa_0$.
+
+An important special case of the identity \ref{eqn:sol1ddp} is the 1D point source,
 $f(z,t)=w(t)\delta(z-z_s)$: then
@@
 
-\append{Plane Waves}
-
-Re-write the plane wave system \ref{eqn:awepw} as
-\begin{eqnarray}
-\label{eqn:awepw1}
-\frac{\partial \tp}{\partial t} + \tk\frac{\partial
-  \tv}{\partial z} &=& (1-c^2\xi^2)^{-1}\tf \nonumber\\
-\rho \frac{\partial \tv}{\partial t} + \frac{\partial \tp}{\partial
-  z}&=&0\nonumber\\
-\tp,\tv&=&0, t \ll 0,
-\end{eqnarray}
-in which $\tk(\xi,z)=(1-c(z)^2\xi^2)^{-1}\kappa(z)$ is a scaled version of
-bulk modulus. The corresponding wave velocity $\tc$ is given by
-\[
-\tc(\xi,z)=\sqrt{\tk/\rho} = \frac{c(z)}{\sqrt{1-c(z)^2\xi^2}} =
-\left(\frac{1}{c(z)^2}-\xi^2\right)^{-1/2}.
-\]
-
-In effect, systen \ref{eqn:awepw1} is a parametrized 1D (space) wave
-system, so the results of the last two appendices apply to the
-homogeneous case, in which $\kappa=\kappa_0$ and $\rho=\rho_0$ are
-independent of $z$. Comparing the plane wave system \ref{eqn:awepw1} with the 1D system
-\ref{eqn:awe1d}, and using the 1D solution for homogeneous fluids
-\ref{eqn:sol1d}, obtain
-\begin{eqnarray*}
-\tp(\xi,z,t) &=& \frac{1}{2\tc_0(\xi)}(1-c_0^2\xi^2)^{-1}\int dz_s
-                 \tf\left(\xi,z_s,t -
-                 \frac{|z-z_s|}{\tc_0(\xi)}\right) \nonumber \\
-\tv(\xi,z,t) &=& \frac{1}{2\rho_0 c_0}(1-c_0^2\xi^2)^{-1} \int dz_s \mbox{sgn} (z-z_s)\tf\left(\xi,z_s,t - \frac{|z-z_s|}{\tc_0(\xi)}\right)
-           \nonumber \\
-\end{eqnarray*}
-which becomes
-\begin{eqnarray}
-\label{eqn:solpw}
- \tp(\xi,z,t)   &=&\frac{1}{2c_0(1-c_0^2\xi^2)^{1/2}}\int dz_s 
-                 \tf\left(\xi,z_s,t -
-                 \frac{|z-z_s|}{\tc_0(\xi)}\right) \nonumber \\
-\tv(\xi,z,t) &=& \frac{1}{2\kappa_0} \int dz_s \mbox{sgn} (z-z_s)\tf\left(\xi,z_s,t - \frac{|z-z_s|}{\tc_0(\xi)}\right)
-                 \nonumber \\
-\end{eqnarray}
-after some simplification.
-
-Accordingly, the plane-wave forward map is given by
-\[
-  \tS[\kappa_0]\tf(\xi,t) =\tp(\xi,z_r,t)
-\]
-\begin{equation}
-\label{eqn:fwdpw1}
-= \frac{1}{2c_0(1-c_0^2\xi^2)^{1/2}}\int dz \tf\left(\xi,z,t -
-  \frac{|z_r-z|}{\tc_0(\xi)}\right)
-\end{equation}
-which replicates equation \ref{eqn:stmodhompw}. Equations
-\ref{eqn:ssmodhompw} and \ref{eqn:ptmodhompw} follow by simple specializations.
-
-The linearized forward map $D\tS[\kappa]$ is defined by
-\begin{equation}
-  \label{eqn:dfwdpw}
-  D\tS[\kappa]\delta \kappa(\xi,t) = \delta \tp(\xi,z_r,t)
-\end{equation}
-Here $\tp$ and $\tv$ solve the linearized plane-wave system
-\begin{eqnarray}
-\label{eqn:dawepw1}
-\frac{\partial \delta\tp}{\partial t} + \tk\frac{\partial
-  \delta \tv}{\partial z} +  \delta \tk\frac{\partial
-  \delta \tv}{\partial z}  &=& (1-c^2\xi^2)^{-2}\delta (c^2)\xi^2\tf \nonumber\\
-\rho \frac{\partial \delta \tv}{\partial t} + \frac{\partial \delta \tp}{\partial
-  z}&=&0\nonumber\\
-\delta \tp,\delta \tv&=&0, t \ll 0,
-\end{eqnarray}
-
-Comparing \ref{eqn:dawepw1} with identity
-\ref{eqn:dawe1d}, observe that the solution of \ref{eqn:dawepw1} for is
-the sum of
-\begin{itemize}
-\item the solution of \ref{eqn:dawe1d} with $\tk, \delta \tk$
-  substituted for $\kappa, \delta \kappa$, and the right hand side in
-  the first equation of the system \ref{eqn:awe1d} for
-  the reference fields $\tp, \tv$ replaced by $(1-c^2\xi^2)^{-1}\tf$;
-\item the solution of \ref{eqn:awe1d} with $\tf$ replaced by
-  $(1-c^2\xi^2)^{-2}\delta (c^2)\xi^2\tf$, and $\kappa_0$ replaced by $\tk_0$..
-\end{itemize}
-
-Expressions for linearized forward maps at homogeneous reference
-models, $D\tS[\kappa_0]\delta \kappa$, follow this observation and the
-results of the last two appendices. For the first summand, in
-\ref{eqn:sol1ddp}, substitute $\tk_0$ for $\kappa_0$, $\delta \tk$ for
-$\delta \kappa$, and $(1-c_0^2\xi^2)\tf$ for $\tf$ to obtain
-\[
--\frac{1}{2\tk_0(\xi)\tc_0(\xi)} \int \,dz_1\, \delta 
-\tk(\xi,z_1)(1-c_0\xi^2)^{-1} \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right)
-\]
-\[
-  + \frac{1}{4\tk_0(\xi)\tc_0(\xi)} \int\,dz_2 \, \int \,dz_1\,\delta 
-  \tk(\xi,z_1) (1-c_0\xi^2)^{-1}\frac{\partial \tf}{\partial t}\left(\xi,z_2,t - \frac{|z_2-z_1|+|z_1-z|}{\tc_0(\xi)}\right) 
-\]
-Since
-\[
-\delta \tk = \delta \left(\frac{\kappa}{1-\kappa \xi^2/\rho}\right) = \delta \kappa (1-c^2\xi^2)^{-2}
-\]
-this amounts to
-\[
-=  -\frac{1}{2\kappa_0c_0 1-c_0\xi^2)^{3/2}} \int \,dz_1\, \delta 
-\kappa(z_1) (1-c_0^2\xi^2)^{-2} \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right) 
-\]
-\begin{equation}
-  \label{eqn:solpwdp1}
-  + \frac{1}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int\,dz_2 \, \int \,dz_1\,\delta 
-  \kappa(z_1) (1-c_0^2\xi^2)^{-2}  \frac{\partial \tf}{\partial t}\left(\xi,z_2,t - \frac{|z_2-z_1|+|z_1-z|}{\tc_0(\xi)}\right) 
-\end{equation}
-
-Note that
-\[
-  (1-c^2\xi^2)^{-2}\delta (c^2)\xi^2= (1-c^2\xi^2)^{-2}\delta \kappa
-  /\rho
-\]
-Therefore, from \ref{eqn:sol1d}, the second summand identified above
-is
-\[
-  =\frac{1}{2\tc(\xi)\rho_0}\int \,dz_1\, \delta 
-\kappa(z_1)(1-c_0^2\xi^2)^{-2} \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right)
-\]
-\begin{equation}
-  \label{eqn:solpwdp2}
-=\frac{c_0^2\xi)^2}{2\kappa_0c_0(1-c_0^2\xi^2)^{3/2}}\int \,dz_1\, \delta 
-\kappa(z_1) \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right)
-\end{equation}
-Adding the quantities \ref{eqn:solpwdp1} and \ref{eqn:solpwdp2},
-obtain
-\[
-  \delta \tp(\xi,z,t) = -\frac{1}{2\kappa_0c_0(1-c_0^2\xi^2)^{1/2}}\int \,dz_1\, \delta 
-  \kappa(z_1) \tf \left(\xi,z_1,t - \frac{|z-z_1|}{\tc_0(\xi)}\right)
-  \]
-\begin{equation}
-  \label{eqn:solpwdp}
-  + \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int\,dz_2 \, \int \,dz_1\,\delta 
-  \kappa(z_1)  \frac{\partial \tf}{\partial
-    t}\left(\xi,z_2,t - \frac{|z_2-z_1|+|z_1-z|}{\tc_0(\xi)}\right)
-  \end{equation}
-Consequently 
-\[
-  D(\tS[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
-  -\frac{1}{2\kappa_0c_0(1-c_0^2\xi^2)^{1/2}}\int \,dz_1\, \delta 
-  \kappa(z_1) \tf \left(\xi,z_1,t - \frac{|z_r-z_1|}{\tc_0(\xi)}\right)
-\]
-\begin{equation}
-  \label{eqn:dmodhomst1}
-    + \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int\,dz_2 \, \int \,dz_1\,\delta 
-  \kappa(z_1)  \frac{\partial \tf}{\partial
-    t}\left(\xi,z_2,t - \frac{|z_2-z_1|+|z_1-z_r|}{\tc_0(\xi)}\right)
-\end{equation}
-which is equation \ref{eqn:dmodhompwst}. For a point source
-$\tf(\xi,z,t) = w(t)\delta(z-z_s)$, this reduces to
-\[
-  D(\tS[\kappa]\tw)_{\kappa=\kappa_0}\delta \kappa(\xi,t) =
-  -\frac{1}{2\kappa_0c_0(1-c_0^2\xi^2)^{1/2}} \delta 
-  \kappa(z_s) w\left(t - \frac{|z_r-z_s|}{\tc_0(\xi)}\right)
-\]
-\begin{equation}
-  \label{eqn:dmodhompt1}
-    + \frac{}{4\kappa_0c_0^2(1-c_0^2\xi^2)} \int \,dz\,\delta 
-  \kappa(z) \frac{\partial w}{\partial
-    t}\left(t - \frac{|z_s-z|+|z-z_r|}{\tc_0(\xi)}\right)
-\end{equation}
-which is equation \ref{eqn:dmodhompwpt}.
-
-\append{Zero-Weight Limit of Variable Projection}
-As in the Preliminaries section, this discussion focuses on the plane wave
-Surface Source objective, in the scaled variant \ref{eqn:ssepeninv},
-reproduced here for convenience:
-\begin{equation}
-  \label{eqn:ssepeninv1}
-  \kappa,\tu \mapsto \lambda^{-2}\|\td-\tS_s[\kappa]\tu\|^2 + \|\tA\tu\|^2
-\end{equation}
-Variable projection creates a function of $\kappa$ by minimizing this
-function over $\tu$, to obtain
-\[
-\kappa \mapsto \frac{1}{\lambda^2}\|\tS[\kappa]\tu_{\lambda}[\kappa]-\td\|^2 + \|\tA
-\tu_{\lambda}[\kappa]\|^2
-\]
-where $\tu_{\lambda}[\kappa]$ solves the normal equation:
-\[
-\tS[\kappa]^T\tS[\kappa]\tu  + \lambda^2\tA^T\tA\tu = \tS[\kappa]^T\td.
-\]
-Since $\tS[\kappa]$ is invertible, $\tu_{\lambda}[\kappa]$ is
-bounded uniformly in $\lambda$ as $\lambda \rightarrow 0$. Let $\tu_0[\kappa] =
-\tS[\kappa]^{-1}\td$. Then
-\[
-\tS[\kappa]^T\tS[\kappa](\tu_{\lambda}[\kappa]-\tu_0[\kappa])+\lambda^2\tA^T\tA\tu_{\lambda}[\kappa]
-= 0,
-\]
-so $\tu_{\lambda}[\kappa] \rightarrow \tu_0[\kappa]$ as $\lambda
-\rightarrow 0$. Consequently the scaled objective above becomes
-\[
-\frac{1}{\lambda^2}\|\tS[\kappa]\tu_{\lambda}[\kappa]-\td\|^2 + \|\tA 
-\tu_{\lambda}[\kappa]\|^2
-=\frac{1}{\lambda^2}\|\tS[\kappa](\tu_{\lambda}[\kappa]-\tu_0[\kappa])\|^2 + \|\tA 
-\tu_{\lambda}[\kappa]\|^2
-\]
-\[
-=\frac{1}{\lambda^2}\langle (\tu_{\lambda}[\kappa]-\tu_0[\kappa]),
-\tS[\kappa]^T\tS[\kappa](\tu_{\lambda}[\kappa]-\tu_0[\kappa])\rangle+ \|\tA 
-\tu_{\lambda}[\kappa]\|^2
-\]
-\[
-=\frac{1}{\lambda^2}\langle (\tu_{\lambda}[\kappa]-\tu_0[\kappa]),
--\lambda^2\tA^T\tA \tu_{\lambda}[\kappa] \rangle + \|\tA\tu_{\lambda}[\kappa]\|^2
-=\langle (\tu_{\lambda}[\kappa]-\tu_0[\kappa]),
--\tA^T\tA \tu_{\lambda}[\kappa] \rangle + \|\tA\tu_{\lambda}[\kappa]\|^2
-\]
-\begin{equation}
-\label{eqn:j0def}
-\rightarrow \|\tA\tu_0[\kappa]\|^2 \equiv 2 J_0[\kappa,\td].
-\end{equation}
-This $\lambda \rightarrow 0$ limit is the simplified objective used by
-\cite{Symes:EAGE15} and mentioned in the Preliminaries section.
-
-\append{OLD RESULTS}
-
-The underlying hypothesis is that the data comes from a point
-source. Extension methods encode such information in an annihilator,
-an operator that vanishes for models that satisfy the hypothesis. For
-SSE-FWI, a natural annilator is a variant on the operator introduced
-for this purpose by \cite{HuangSymes:Geo18a,HuangSymes:Geo18b} for two
-similar source extension methods, namely multiplication by offset,
-also reminiscent of a now-standard annihilator for subsurface offset
-introduced by \cite{stolk2001c}. In the plane wave domain,
-multiplication by offset becomes 
-\[
-\tA = D_{\xi}I_t,
-\]
-that is, differentiation in $\xi$ and integration in $t$, in either
-order.
-
-The forward map $\tS$ for plane-wave transformed inverse problem maps
-extended plane wave source wavelets $\tw(\xi,t)$ to plane wave data
-traces (sampling of the plane-wave field at $z=0$, assuming this to be the
-depth of sources and receivers, as I do here). I will regard this operator as a
-function of the bulk modulus $\kappa(z)$, by assuming the density
-$\rho$ to be fixed and homogeneous.
-
-$\tS[\kappa]$ is 
-actually invertible, at least for $\kappa$ close to homogeneous. This conclusion follows for instance
-from explicit expressions in two special cases: for homogeneous models
-$\kappa=\kappa_0$, 
-\[
-\tS[\kappa_0]\tw(\xi,t) = a_0(\xi)\tw(\xi,t),
-\]
-and for a smooth model $\kappa_1$ with an embedded reflector at $z=z_1$,
-
-\[
-\wl(t)=\frac{1}{\sqrt{\lambda}}w_1\left(\frac{t}{\lambda}\right)
-\]
-is a family of wavelets depending on central wavelength $\lambda$. The
-vertical travel time $\tau(\xi,z_1)$ is given by
-\[
-\tau(\xi,z_1) = \int_0^{z_1}\sqrt{\frac{1}{c^2}-\xi^2},\,\,c=\sqrt{\kappa/rho}.
-\]
-
-Since $\tS$ is invertible, it is possible to use the
-simplified objective presented by
-\cite{Symes:EAGE15}:
-\begin{equation}
-\label{eqn:dsobj}
-J_0[\kappa]=\frac{1}{2}\|\tA \tS[\kappa]^{-1}\td\|^2
-\end{equation}
-A derivation of this limit case from the standard penalty formulation
-of extended FWI 
-appears in the last section. Inversion based on this type of objective
-has been called ``inversion velocity analysis'' in the subsurface
-offset context \cite[]{Herve2017},\cite[]{HouSymes:Geo18}, and is also the
-basis of Adaptive Waveform Inversion
-\cite{Warner:14,Warner:16}.
-
-The main result pertains to the gradient at homogeneous
-$\kappa=\kappa_0$ for data resulting from a single flat reflector
-embedded in a smooth background:
-\[
-\tdl = \tS[\kappa_1]\wl
-\]
-I calculate the gradient of $J_0$ in this configuration.
-The result is given in equation \ref{eqn:jgrad} in the last
-section. Converted to spatial frequency via the Fourier transform, it
-becomes
-\begin{equation}
-\label{eqn:jgradft-intro}
-\widehat{\nabla J[\kappa_0]}(k) = -\frac{c_0}{16}\int\, d\xi\,a_1(\xi) 
-\frac{1}{ik}\hat{u}_{\lambda}(\xi,k)D_{\xi}^2\overline{\hat{u}_{\lambda}(\xi,k)}
-\end{equation}
-where
-\[
-\hat{u}_{\lambda}(\xi,k) = \left(1 +
-  \frac{R(\xi)}{a_0(\xi)}e^{4ik\tau(\xi,z_1)/\tc(\xi)}\right)\hat{\wl}\left(\frac{2k}{\tc(\xi)}\right) 
-\]
-and 
-\[
-\tc(\xi) = \left(\frac{1}{c_0}^2-\xi^2\right)^{-1/2}
-\]
-is the vertical plane wave velocity for homogenous wave velocity
-$c_0 = \sqrt{\kappa_0/rho}$.
-
-Then for any $k_{\rm max} >0$, $\xi_{\rm max} > 0$,
-\[
-\max \{\left|\hat{w}_{\lambda}\left(2k/\tc{\xi}\right)\right|:
-|k|\le k_{\rm max}, |\xi|\le \xi_{\rm max} \} \rightarrow 0\mbox{ as }
-\lambda \rightarrow 0.
-\]
-It follows immediately from equation \ref{eqn:jgradft-intro} that the
-same is true of the gradient:
-\begin{equation}
-\label{eqn:bigdeal}
-\max \{\left|\widehat{\nabla J[\kappa_0]}(k)\right|:
-|k|\le k_{\rm max}, |\xi|\le \xi_{\rm max} \} \rightarrow 0\mbox{ as }
-\lambda \rightarrow 0.
-\end{equation}
-
-Therefore the rate at which traveltime changes for any plane wave
-component in the specified range, as the bulk modulus is changed along
-the direction of the negative gradient, is vanishingly small as
-$\lambda \rightarrow 0$.
-
-Otherwise put, for high-frequency reflection data, this method cannot
-update the velocity away from a homogenous initial guess in a way that
-is kinematically constructive.
-
-It may even be possible to show that the update may be destructive, by
-more careful examination of the gradient approximation. However, who
-cares, really: it is established that SSE-FWI has similar behaviour to
-FWI for at least one simple instance of reflection data.
 \bibliographystyle{seg}

### 2020-02-06 repro/papers/wri1d/paper.tex
first draft

@@
 
+\begin{abstract}
+  Full waveform inversion, or nonlinear least squares data-fitting
+  driven by wave equation simulation, has proven its ability to
+  extract detailed maps of earth structture from near-surface seismic
+  observations. However, the least squares error function at the heart
+  of this method tends to develop a high degree of nonconvexity, so
+  that local optimization methods (the only numerical methods feasible
+  for field-scale problems) tend to fail, unless provided with initial
+  estimates of a quality not always available. A number of alternative
+  optimization principles have been advanced that promise some degree
+  of release from the multimodality of Full Waveform Inversion,
+  amongst them Wavefield Reconstruction Inversion, the focus of this
+  paper. Applied to a simple 1D acoustic transmission problem, both
+  Full Waveform and Wavefield Reconstruction Inversion methods reduce
+  to minimization of explicitly computable functions, in an asymptotic
+  sense. The analysis presented here shows that Wavefield
+  Reconstruction Inversion can be vulnerable to the same
+  ``cycle-skipping'' failure mode as Full Waveform Inversion.
+\end{abstract}
+
 \section{Introduction}
 
-\section{Full Waveform Inversion}
-Full waveform inversion (FWI) is the current nomenclature for data-fitting parameter estimation driven by wavefield modeling.
-The aim of FWI is to fit a vector of data observations $d$ with a predicted data vector of the same form.
-The prediction is accomplished by application of a modeling operator $S$ to a vector of input observations $f$, representing sources of energy that initiate waves in the material model underlying $S$.
-$S$ propagates these waves to measurement locations at which the data samples are extracted from the wave fields, generating the output of $S$.
-The propagation dynamics depend on a vector of material parameters $c$, hence $S=S[c]$ also depends on $c$.
+Full waveform inversion (FWI) is the current nomenclature in the
+seismology literature for data-fitting earth structure estimation
+driven by wavefield modeling. The earth properties to be estimated
+(material densities, stiffnesses, attenuation rates,...) form a vector
+$c$ of spatially fields that appear as coefficients in systems of hyperbolic
+partial differential equations, modeling seismic wave
+propagation. The wavefields used in structure estimation (``imaging'')
+are small motion disturbances of the earth's equilibrium state, so the
+equations of motion are typically 
+linear(ized). Right-hand side vectors $f$ in these systems model energy
+input that initiates waves (earthquakes, man-made sources such as
+explosives or mechanical vibrators). Data vectors $d$ are simulated by
+sampling the solution fields at the locations of measurement devices
+(accelerometers, microphones,...)  over appropriate time
+intervals. The relation between the energy source $f$ and the
+simulated data is linear in $f$, but nonlinear in the coefficient
+vector $c$, so is naturally represented by a family of linear
+operators $S[c]$ parametrized by the coefficient vector $c$.
 
-A restatement of the goal: estimation of the materal parameters $c$ (hence the modeling operator $S[c]$) and the  from the data $d$, via the requirement that $S[c]f \approx d$.
-A typical mehod for enforcing this requirement is the minimization of an objective misfit measure (objective, for short), the most common choice being the square norm of a Hilbert space in which the data is presumed to reside:
+The objective of FWI is to find $c$ and $f$, given $d$, so that
+$S[c]f \approx d$. A typical method for enforcing this requirement is
+the minimization of an objective misfit measure (objective, for
+short), the most common choice being the square norm of a Hilbert
+space in which the data is presumed to reside:
 \begin{equation}
-  \label{eqn:deffwifull}
-  J_{\rm FWI}[c,f] = \frac{1}{2}\|S[c]f -d\|^2.
+  \label{eqn:deffwi}
+  \mbox{Given } d, \mbox{ find }c \mbox{ and }f \mbox{ to
+    minimize }
+  \|d -  S[c]f\|^2
 \end{equation}
-This approach was first suggested in the 1980's (\cite{BamChavLai:79,Tara:84a,KolbColLai:86,Crasetal:90},
-and many other papers since then). Usually some form of regularization
-is applied, to compensate for poorly determined aspects of $c$
-and/or $f$, as is explained in Tarantola's influential book
-\cite[]{Tarantola:05}. Also, $f$ may be constrained in one way or another to embody characteristics of field energy sources (an example of this is given below).
+in which $\|\cdot\|$ is the norm in a suitable Hilbert space. This
+approach was first suggested in the 1980's
+(\cite{BamChavLai:79,Tara:84a,KolbColLai:86,Crasetal:90}, and many
+other papers since then). Usually some form of regularization is
+applied, to compensate for poorly determined aspects of $c$ and/or
+$f$, as is explained in Tarantola's influential book
+\cite[]{Tarantola:05}. Also, $f$ may be constrained in one way or
+another to embody characteristics of field energy sources, or even
+regarded as known (an example of this is given below).
 
+Within a few years of its introduction into quantitative seismology,
+FWI was understood to suffer from a severe limitation. Because of the
+typical dimensions of earth models, and consequent cost of accurate
+computation of $S$, iterative local optimization provides the only
+feasible route to estimation of $c$ via solution of the optimization
+problem \ref{eqn:deffwi}. However the objective function of this
+optimization problem (the mean-square residual appearing in display \ref{eqn:deffwi}) has many local minima in general, most
+having nothing to do with a usable estimate of earth structure. [An
+explicit example of this multi-modal behaviour appears below.]
+Reliable estimation of $c$ via iterative local optimization requires
+that the initial estimate predict the correct arrival time of waves,
+as they appears in the data, within a wavelength at dominant
+frequencies \cite[]{GauTarVir:86,VirieuxOperto:09}.
+
+Despite this severe constraint, FWI has shown enough promise as a tool
+for both industrial and academic seismology that it is now a
+mainstream research topic, and to some extent a commercial
+product. Estimation of sufficiently accurate initial models via
+non-FWI methods is common practice, though what ``sufficiently
+accurate'' means may be difficult to discern
+\cite[]{Plessix:10}. The wavelength criterion mentioned in the last
+paragraph may be made less onerous by collection of relatively
+low-frequency data \cite[]{Wolfspar:16}. Finally, many alternatives to
+straightforward least-squares data fitting have been suggested, some
+of which appear to exhibit less tendency to develop local minima than
+does the problem described in \ref{eqn:deffwi} \cite[]{Symes:09}.
+
+This topic of this paper is one of these alternative approaches,
+Wavefield Reconstruction Inversion (``WRI''). WRI was introduced by \cite{LeeuwenHerrmannWRI:13}, and further
+developed by \cite{LeeuwenHerrmann:16,WangYingst:SEG16} and other
+authors. It is based on the presumption that the correct source (right
+hand side in the equations of motion) $q$ is
+known, and combines a penalty for data misfit with a penalty for
+failing to solve the equations of motion with the correct right hand side:
+\begin{equation}
+  \label{eqn:defwri}
+  \mbox{Given } d \mbox{ and }q, \mbox{ find }c \mbox{ and }f \mbox{ to
+    minimize }
+  \|d -  S[c]f\|^2+\alpha^2\|f-q\|^2.
+\end{equation}
+The penalty parameter $\alpha$ may be chosen small, and examples given
+by \cite{LeeuwenHerrmannWRI:13} and others suggest that the problem
+defined in display \ref{eqn:defwri} is less likely to develop
+uninformative local minima if $\alpha$ is small. This might be so for
+several reasons: one is that when less weight is
+put on making the residual in the wave equation ($f-q$) small, 
+$f$ may be chosen to make the data residual $d-S[c]f$ small
+instead. One aspect of failure to predict the arrival times of waves
+accurately is that small data residual is then difficult to
+achieve. By maintaining fidelity to
+the data, local minimization of the WRI objective might provide a
+spurious-minimum-free path to a satisfactory estimate of $c$.
+
+The main result of this paper is that in one simple case, in which all
+of the necessary computations can be carried out by hand, this hope is
+not realized: minimization of the WRI objective is just as likely to
+be trapped in a spurious local minimizer as is the FWI objective.  The
+context of this conclusion is a simple transmission inverse problem
+for the 1D acoustic wave system, which models pulse transmission along
+a 1D continuum from a source point to a receiver point. The pulses
+used in this thought experiment are short, so the main information
+content of the data is the time of transit from source to
+receiver. The predominant information about the material model, in
+this case reduced to the wave velocity (a scalar function of
+position), is just this transit time, so I constrain both the target
+wave velocity $c_*$ generating the data and the trial wave velocity
+$c$ to be constant, that is, independent of position along the 1D
+continuum. I introduce a family of inverse problems, parametrized by a
+parameter $\lambda$ playing the role of wavelength. For sufficiently small
+$\lambda$, it is possible to show explicitly that the FWI problem
+possesses local minimizers far from the global minimizer at the target
+$c_*$, and that initiating a local iterative optimization, such as
+steepest descent or Newton's method, at a distance from $c_*$ bounded
+below by a multiple of the ``wavelength'' $\lambda$ will result in
+convergence to these spurious local minima. That is, FWI 
+behaves in exactly the manner described in much of the literature on
+this topic. However, an analysis of WRI applied to the same context
+yields the same result: spurious local minima exist for sufficiently
+small $\lambda$, and will be found by local optimization unless the
+starting point is within $O(\lambda)$ of the target. That is, WRI
+behaves in a manner qualitatively indistinguishable from FWI. In
+particular, its ability to allow good fit to data for small $\alpha$
+does not safeguard it from failure to converge globally to a ``good''
+local minimum.
+
+This paper begins with a description of the 1D inverse transmission
+problem and explicit computation of various components of the FWI
+approach, based on explicit solution of the 1D acoustic system
+presented in Appendix A. In the third section I introduce the $\lambda-$dependent
+family of problems, and establish the asymptotic properties of FWI as
+$\lambda \rightarrow 0$. The fourth section develops the algebraic
+structure of WRI, culminating in a remarkable identity revealing WRI
+to be equivalent to minimization of a weighted norm of the data
+residual, with a weight operator depending on the coefficient vector
+$c$. This result is quite general, applying to essentially any
+realization of WRI, and already shows that its behaviour must be
+closely related to that of FWI. In the fifth section I return to 
+1D acoustics problem and the $\lambda$-dependent family of inverse
+problems, compute that weight operator explicitly, and deduce the
+global behaviour of WRI in this instance. The paper ends with a
+discussion of the relation of the analysis presented here to a
+previous analysis of optimization formulations of wave inversion
+problems based on parameter-dependent quadratic forms
+\cite[]{StolkSymes:03}, and a brief discussion of some other
+alternatives to FWI.
+
+\section{FWI for 1D Acoustics}
 The example of FWI to be explored in this paper is one of the simplest possible, based on the 1D acoustics system connecting excess pressure $p$, particle velocity $v$, sonstitutive law defect (``source'') $f$, density $\rho$, and wave velocity $c$:
@@
 
-The system \ref{eqn:awe1d} has classical (smooth) solutions $(p,v)$ when $c, \rho,$ and $f$ are smooth, and $\log c$ and $\log \rho$ are bounded on $\bR$, as is well-established \cite[]{Lax:PDENotes}.
-In this paper, for reasons to be discussed below, $c$ and $\rho$ are constrained to be constant ($z$-independent) in which case solutions may be constructed by elementary methods (Appendix A).
+The system \ref{eqn:awe1d} has classical (smooth) solutions $(p,v)$
+when $c, \rho,$ and $f$ are smooth, and $\log c$ and $\log \rho$ are
+bounded on $\bR$, as is well-established \cite[]{Lax:PDENotes}. In
+this paper, for reasons to be discussed below, $c$ and $\rho$ are
+constrained to be constant ($z$-independent) in which case solutions
+may be constructed by elementary methods (Appendix A).
 
-Choose $l_c, l_{\rho} >0$ and admit $c, \rho$ as coefficients in \ref{eqn:awe1d} for which $|\log c| \le l_c, |\log \rho| \le l_{\rho}$.
+In fact, I shall assume $\rho>0$ to be fixed for the remainder of this
+paper, that is, not updated in the inversion process. The wave
+velocity $c$ will range over an interval: it is the parameter to be inverted.
+To be specific, choose $c_{\rm max} > c_{\rm min} > 0$, and require
+that $c$ satisfy $c_{\rm min} \le c \le c_{\rm max}$.
 
@@
 \begin{equation}
+  \label{eqn:defmoddom}
+  S[c]: L^2([z_{\rm min},z_{\rm max}] \times \bR) \rightarrow L^2[0,T]
+\end{equation}
+As it will turn out, the choice of spatial interval $[z_{\rm min},z_{\rm max}]$ is arbitrary, so long as it has positive length and contains $z_s$ in its interior.
 
-  S[c]: L^2([z_{\rm min},z_{\rm max}] \times [0,T]) \rightarrow L^2[0,T]
-\end{equation}
 For homogeneous ($z$-independent) $\kappa$, Appendix A provides an explicit expression:
@@
   \label{eqn:defmodhom}
-  S[c]f(t) = \frac{1}{2c}\int dz f\left(z,t - \frac{|z_r-z|}{c}\right) 
+  S[c]f(t) = \frac{1}{2c}\int_{z_{\rm min}}^{z_{\rm max}} dz f\left(z,t - \frac{|z_r-z|}{c}\right) 
 \end{equation}
-From this expression it is simple to see that $S[c]$ is a bounded operator with the domain and range described in \ref{eqn:defmoddr}.
+From this expression it is simple to see that $S[c]$ is a bounded
+operator with the domain and range described in
+\ref{eqn:defmoddom}. The role of compact support in $z$ in ensuring
+boundedness is also evident. % It is also evident from that limiting source to have bounded support in $z$ is necessary make $S[c]$ bounded.
 
-With this framework, the data is presumed to lie in the range of $S[c]$: $d \in L^2[0,T]$, and all components of the basic FWI definition \ref{eqn:deffwifull} are defined.
+With this framework, the data is presumed to lie in the range space of $S[c]$: $d \in L^2[0,T]$.
 
 The version of FWI discussed here presumes that the source field corresponding to the data $d$ is supported at a point $z_s \ne z_r$, and is known.
-This field will be denoted depends on a function of time (``wavelet'') $w \in L^2[0,T]$.
+This field depends on a function of time (``wavelet'') $w \in L^2[0,T]$.
 Formally, the resulting acoustic field satisfies the system \ref{eqn:awe1d} with $f=w(t)\delta(z-z_s)$.
@@
   \label{eqn:defmodpt}
-  S[c](w\delta(\cdot-\z_s)) = \frac{1}{2c}w\left(t - \frac{|z_r-z_s|}{c}\right) = S_p[c]w(t)
+  S[c](w\delta(\cdot-z_s)) = \frac{1}{2c}w\left(t - \frac{|z_r-z_s|}{c}\right) = S_p[c]w(t)
+\end{equation}
+Thus for this restricted class of source fields, the FWI objective can be redefined as
+\begin{equation}
+  \label{eqn:deffwipt}
+  J_{\rm FWI}[c,d.w] = \frac{1}{2}\|S_p[c]w -d\|^2.
+\end{equation}
+$S_p[c]$ is a bounded operator with domain $L^2(\bR)$ and range $L^2([0,T])$.
+
+To end this section, it is necessary to address an irritating technical point: the point source defined above is not a member of the domain of $S[c]$, as it was defined in display \ref{eqn:defmoddom}, so the left-hand side of equation \ref{eqn:defmodpt} does not actually make sense.
+The fix for this incompatibility actually elucidates the relation between $S[c]$ and $S_p[c]$.
+Appendix B describes the construction of a family of bounded injective operators  $:L^2(\bR) \rightarrow L^2([z_{\rm min},z_{\rm max}] \times \bR)$ for which
+\begin{equation}
+  \label{eqn:ext}
+  S_p[c] = S[c] \circ E[c]
+\end{equation}
+This relation exhibits $S[c]$ as an {\em extension} of $S_p[c]$ as
+described by \cite{Symes:09}, and identifies WRI as an extended
+inversion method.
+
+The construction described in Appendix B requires that $z_s \in
+(z_{\rm min},z_{\rm max})$, as mentioned above.
+
+\section{Global Asymptotics of FWI}
+The FWI objective is well-known to exhibit non-convexity unless
+data frequency content is limited to a small range near 0 Hz, how
+small being determined by other scales and by the extent to which the
+initial wave velocity differs from the target.
+
+To understand the non-convexity phenomenon and the
+relation of the various scales, it is advantageous to introduce a
+family of source wavelets, depending on a parameter $\lambda$ playing
+the role of wavelength:
+\begin{equation}
+  \label{eqn:wfam}
+  \wl(t) = \frac{1}{\sqrt{\lambda}}w_1\left(\frac{t}{\lambda}\right).
+\end{equation}
+In the definition \ref{eqn:wfam}, the ``mother wavelet'' $w_1\in
+C_0^{\infty}(0,T)$, and the scaling is chosen so fhat
+$\|\wl\|_{L^2(\bR)}$ is independent of $\lambda>0$.
+
+Evidently there is no control of $c$ at all in the data if the time
+interval of the observation, namely $[0,T]$, is so short that no
+signal arrives within it. Accordingly, add to the other assumptions
+made so far
+the requirement that the transit time between source and receiver at
+the slowest permitted velocity is less than $T$:
+\begin{equation}
+  \label{eqn:time}
+  \frac{|z_s-z_r|}{c_{\rm min}} < T
 \end{equation}
 
+Choose a target wave velocity $c_* \in [c_{\rm min},c_{\rm
+  max}]$. Introduce a family of consisten data $\dl$, generated by
+$c_*$ and the wavelet family $\wl$:
+\begin{equation}
+  \label{eqn:dfam}
+  \dl = S_p[c_*]\wl
+\end{equation}
+and a corresponding family of FWI objectives:
+\[
+  J_{\rm FWI}[c,\dl,\wl] = \frac{1}{2}\|\dl-S_p[c]\wl\|^2
+\]
+\begin{equation}
+  \label{eqn:fwiexpl}
+  =
+  \int_0^T\,dt\,\left|\frac{1}{2c_*}\wl\left(t-\frac{z_s-z_r}{c_*}\right)
+    - \frac{1}{2c}\wl\left(t-\frac{z_s-z_r}{c}\right)\right|^2
+\end{equation}
+Note that $\mbox{supp }\wl \subset [0,\lambda T]$, so
+\begin{equation}
+  \label{eqn:wlsupps}
+  \mbox{supp }S_p[c]\wl \subset \left[\frac{z_s-z_r}{c},\lambda
+    T+\frac{z_s-z_r}{c}\right] \cap [0,T]
+\end{equation}
+The transit time condition \ref{eqn:time} implies that there exists
+$\lambda_*>0$ so that for $\lambda < \lambda_*$,
+\[
+  \lambda T+\frac{z_s-z_r}{c} < T
+\]
+for all admissible $c$. That is, 
+for $\lambda <\lambda_*$
+\begin{equation}
+  \label{eqn:constnorm}
+  \|S_p[c]\wl\|^2 = \frac{1}{4c^2}\int_{-\infty}^{\infty}\,dt\,\left|
+    \wl\left(t-\frac{z_s-z_r}{c}\right)\right|^2 =
+  \frac{\|w_1\|^2}{4c^2}.
+\end{equation}
+
+Recall that the object of this study is the global behaviour of
+objective functions for velocity estimation: in this context, that
+means the behaviour for $c$ far from $c_*$.  Pick $\delta c >0$ for
+which $[c_*-\delta c, c_* + \delta c] \subset [c_{\rm min},c_{\rm
+  max}]$, that is, that the set of admissible $c$ further than $\delta
+c$ from $c_*$ is non-void.
+
+Then there is $\lambda_{**}$ depending on $\delta c, T, and
+|z_s-z_r|$ so that if $|c-c_*|>\delta c$ and $\lambda<\lambda_{**}$, then
+\[
+  \left|\frac{|z_s-z_r|}{c}-\frac{|z_s-z_r|}{c_*}\right| > 2\lambda T,
+\]
+that is, the infima of the supports of $S_p[c]\wl, \dl=S_p[c_*]\wl$
+are further apart than the lengths of these supports. In that case,
+necessarily
+\[
+  \mbox{supp }S_p[c]\wl \cap \mbox{supp }S_p[c_*]\wl = \emptyset.
+\]
+Therefore $S_p[c]\wl$ and $S_p[c_*]\wl$ are orthogonal in $L^2[0,T]$,
+and
+\begin{equation}
+  \label{eqn:fwiscrewed}
+  J_{\rm FWI,\lambda}[c] =
+  \frac{1}{2}\left(\frac{1}{4c^2}+\frac{1}{4c_*^2}\right)\|w_1\|^2.
+\end{equation}
+Amongst other consequences, one immediately deduces from the
+expression \ref{eqn:fwiscrewed} the non-convexity result:
+
+\begin{theorem}
+  \label{thm:thm1}
+  There exists $C>0$, $\lambda_0$ so that for $\lambda <
+  \lambda_0$, the minimizer of $J_{\rm FWI, \lambda}[c]$ on the
+  complement of $[c_*-C\lambda, c_*+C\lambda]$ is $c=c_{\rm max}$.
+\end{theorem}
+
+That is, outside of a neighborhood of width proportional to a
+wavelength, minimization of $J_{\rm FWI}$ yields a local minimizer far from
+the target $c_*$ which generates the (noise-free) data.
+
+For this 1D problem, a happy 1D accident occurs: a descent
+minimization starting at $c_0 < c_*$ will at least initially proceed
+in the right direction. With sufficiently small steps, it is possible
+that an interation might land in the (small) domain of attraction
+around $c_*$. However neither this nor various other accidental
+advantages stemming from the very special form of this problem should
+be regarded as of any importance.
+
 \section{Wavefield Reconstruction Inversion}
-This section will introduce Wavefield Reconstruction Inversion (WRI) and develop some of its formal algebraic properties.
-The notation $S[c]$ will represent a modeling operator based on wave dynamics of some sort, depending on a vector $c$ of material parameters.
-The conclusions developed here in fact apply to WRI in any such setting.
-These conclusions will be applied to 1D acoustics in the following section.
+This section will describe Wavefield Reconstruction Inversion (WRI)
+and develop some of its formal algebraic properties. The notation
+$S[c]$ will represent a modeling operator based on wave dynamics of
+some sort, depending on a vector $c$ of material parameters. The
+conclusions developed here in fact apply to WRI in any such
+setting. These conclusions will be applied to 1D acoustics in the
+following section, with $c$ specialized to a scalar (wave velocity).
 
-WRI was introduced by \cite{LeeuwenHerrmannWRI:13}, and further
-developed by \cite{LeeuwenHerrmann:16,WangYingst:SEG16} and other
-authors. It is based on the presumption that the correct source $q$ is known.  and extension $S$ (equation \ref{eqn:fwd}), with full
-space-time volume source $f$, and presumes that the correct point
-source wavelet $w_*$ is known. 
-\begin{equation}
-  \label{eqn:defwri}
-  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }f \mbox{ to
-    minimize }
-  \| d -  S[\kappa]f\|^2+\alpha^2\|f-w_*\delta(\cdot-\bx_s)\|^2.
-\end{equation}
-Evidently the difference $f-w_*\delta(\cdot-\bx_s)$ should be such
-that the scaled sum-of-squares makes sense - this is always the case
-for discretizations, but for the continuum limit entails an actual
-assumption.
+In the application below, the target source $q$ will be a point source
+as in the previous section, however for the development of the basic
+properties of WRI, that is immaterial.
 
-\cite{LeeuwenHerrmannWRI:13} posed this problem slightly
-differently, in terms of the second order wave equation for the
-pressure wavefield $p$.
-From this viewpoint, $\partial/\partial t(f-w_*\delta(\cdot-z_s))$ is the residual, that is, the image of the 2nd order wave operator on $p$, so the second term penalizes the failure of $p$ to solve the wave equation with source $\partial w_*/\partial t \delta(\cdot-z_s)$.
-The formulation presented here is equivalent, and was
+Note that in general $\alpha$ is a dimensional parameter, having the
+same dimensions as $S$.
+
+\cite{LeeuwenHerrmannWRI:13} posed this problem slightly differently:
+instead of the first order acoustics system (equation \ref{eqn:awe1d}
+for the 1D case), they pose the wave dynamics in terms of the second
+order wave equation for the pressure wavefield $p$. From this
+viewpoint, $\partial(f-q) /\partial t$ is the residual, that is, the
+difference between the image $\partial f/\partial t$ of the 2nd order
+wave operator on $p$, and the assumed right-hand side
+$\partial q/\partial t$. So in this form, the second term penalizes
+the failure of $p$ to solve the wave equation with the assumed
+source. The formulation presented here is equivalent, and was
 introduced by \cite{WangYingst:SEG16}.
 
-\cite{LeeuwenHerrmann:16} used
-the variable projection method \cite[]{GolubPereyra:03}, eliminating the source $f$ in the
-inner step and updating the bulk modulus (or an equivalent quantity)
-in the outer step. That is, the problem \ref{eqn:defwri} is equivalent to minimization of
+\cite{LeeuwenHerrmann:16} used the variable projection method
+\cite[]{GolubPereyra:03}, eliminating the source $f$ in the inner step
+and updating the bulk modulus (or an equivalent quantity) in the outer
+step. That is, the problem \ref{eqn:defwri} is equivalent to
+minimization of
 \begin{equation}
 \label{eqn:defvpm}
-  J[\kappa] =
-  \mbox{min}_f \frac{1}{2}(\| d-S[\kappa]f\|^2+\alpha^2\|f-w_*\delta(\cdot-\bx_s)\|^2)
+  J_{\rm WRI}[c,d,q] =
+  \mbox{min}_f \frac{1}{2}(\| d-S[c]f\|^2+\alpha^2\|f-q\|^2)
 \end{equation}
-over $\kappa$. This is the approach that I shall pursue here.
+over $c$. This is the approach that I shall pursue here.
 
@@
 ``Most'' source fields $f$ are non-radiating,
-that is, $S[\kappa]f=0$. The orthocomplement of the subspace of
+that is, $S[c]f=0$. The orthocomplement of the subspace of
 non-radiating sources is the range of the adjoint operator
-$S[\kappa]^T$. This observation suggest that the sources may be re-parametrized as images of
-$S[\kappa]^T$. Write $f = g + w_*\delta(\cdot-\bx_s)$: then the second term in
+$S[c]^T$. This observation suggest that the sources may be re-parametrized as images of
+$S[c]^T$. Write $f = g + q$: then the second term in
 the objective displayed in \ref{eqn:defwri} is simply $\|g\|^2$.
-Decompose $g = S[\kappa]^Te + n$, in
-which $e$ is the same type of object as $d$ and $S[\kappa]n=0$ (that
+Decompose $g = S[c]^Te + n$, in
+which $e$ is the same type of object as $d$ and $S[c]n=0$ (that
 is, $n$ is a non-radiating
@@
 \[
-  J[\kappa] =
-  \mbox{min}_{e,n} \frac{1}{2}(\| d-S[\kappa](S[\kappa]^Te+w_*\delta(\cdot-z_s))\|^2+\alpha^2(\|S[\kappa]^Te\|^2+\|n\|^2))
+  J_{\rm WRI}[c,d,q] =
+  \mbox{min}_{e,n} \frac{1}{2}(\| d-S[c](S[c]^Te+q)\|^2+\alpha^2(\|S[c]^Te\|^2+\|n\|^2))
   \]
@@
   \label{eqn:defvpmred}
-   \mbox{min}_{e} \frac{1}{2}(\|r-S[\kappa]S[\kappa]^Te\|^2+\alpha^2\|S[\kappa]^Te\|^2)
+   \mbox{min}_{e} \frac{1}{2}(\|r-S[c]S[c]^Te\|^2+\alpha^2\|S[c]^Te\|^2)
  \end{equation}
- in which $r[\kappa]=d-S[\kappa]w_*\delta(\cdot-z_s)$ is the data residual.
+ in which $r[c]=d-S[c]q$ is the data residual.
 
 This reformulation has some computational advantages \cite[]{WangYingst:SEG16,Herrmann:SEG19}, but also leads to a useful analytic transformation of the WRI problem.
-The minimizer on the RHS of equation \ref{eqn:defvpmred} is the solution $e=e_{\alpha}[\kappa]$ of the normal equation
+The minimizer on the RHS of equation \ref{eqn:defvpmred} is the solution $e=e_{\alpha}[c]$ of the normal equation
 \[
- ( (S[\kappa]S[\kappa]^T)^2 + \alpha^2S[\kappa](S[\kappa]^T)e = S[\kappa]S[\kappa]^Tr[\kappa]
+ ( (S[c]S[c]^T)^2 + \alpha^2S[c](S[c]^T)e = S[c]S[c]^Tr[c]
 \]
@@
 \[
-  S[\kappa]S[\kappa]^Te_{\alpha}[\kappa] = S[\kappa]S[\kappa]^T(S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]
+  S[c]S[c]^Te_{\alpha}[c] = S[c]S[c]^T(S[c]S[c]^T+\alpha^2I)^{-1}r[c]
 \]
-Assume that $S[\kappa]S[\kappa]^T$ is invertible - this is in fact for the problem outlined in the introduction, as will be verified shortly. Then
+Assume that $S[c]S[c]^T$ is invertible - this is in fact for the problem outlined in the introduction, as will be verified shortly. Then
 \begin{equation}
   \label{eqn:norsol}
-  e_{\alpha}[\kappa]=(S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]
+  e_{\alpha}[c]=(S[c]S[c]^T+\alpha^2I)^{-1}r[c]
 \end{equation}
@@
 \[
-  J[\kappa] =
-  \frac{1}{2}(\|r[\kappa]-S[\kappa]S[\kappa]^Te_{\alpha}[\kappa]\|^2+\alpha^2\|S[\kappa]^Te_{\alpha}[\kappa]\|^2)
+  J_{\rm WRI}[c,d,q] =
+  \frac{1}{2}(\|r[c]-S[c]S[c]^Te_{\alpha}[c]\|^2+\alpha^2\|S[c]^Te_{\alpha}[c]\|^2)
 \]
 \[
-  = \frac{1}{2}(\|r[\kappa]-S[\kappa]S[\kappa]^T (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]\|^2+\langle (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa],
-  S[\kappa]S[\kappa]^T (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]
-  \rangle
+  = \frac{1}{2}\left(\|r[c]-S[c]S[c]^T
+    (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\|^2 \right.
 \]
 \[
-  =\frac{1}{2}\left(\|\alpha^2 (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]\|^2
-  +\langle (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa],
-  -\alpha^2 (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]\rangle\right.
+  \left.+\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],
+  S[c]S[c]^T (S[c]S[c]^T+\alpha^2I)^{-1}r[c]
+  \rangle\right)
 \]
 \[
-  + \left.\langle (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa],r[\kappa]\rangle\right)
+  =\frac{1}{2}\left(\|\alpha^2 (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\|^2\right.
 \]
 \[
-  =  \frac{1}{2}\left(
-\langle (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa],r[\kappa]\rangle +
-\alpha^2(1-\alpha^2)\| (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}r[\kappa]\|^2\right) 
+  \left.
+  +\alpha^2\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],
+  -\alpha^2 (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\rangle\right.
 \]
 \[
-  =  \frac{1}{2}
-  \langle ((S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}+\alpha^2(1-\alpha^2)(S[\kappa]S[\kappa]^T+
-  \alpha^2I)^{-2})r[\kappa],r[\kappa]\rangle
+  + \left.\alpha^2\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],r[c]\rangle\right)
+\]
+\[
+  =  \frac{\alpha^2}{2}
+\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],r[c]\rangle 
 \]
@@
   \label{eqn:defwrialt}
-  J[\kappa] = \frac{1}{2}\langle r[\kappa], W[\kappa] r[\kappa]\rangle
+  J_{\rm WRI}[c,d,q] = \frac{1}{2}\langle r[c], W_{\alpha}[c] r[c]\rangle
 \end{equation}
@@
   \label{eqn:defwriwt}
-  W[\kappa] = (S[\kappa]S[\kappa]^T+\alpha^2I)^{-1}+\alpha^2(1-\alpha^2)(S[\kappa]S[\kappa]^T+
-  \alpha^2I)^{-2}.
+  W_{\alpha}[c] = \frac{\alpha^2}{2}(S[c]S[c]^T+\alpha^2I)^{-1}
 \end{equation}
-This remarkable identity shows that the WRI objective function is a {\em weighted norm of the data residual $r[\kappa]$}.
-Note that this conclusion depends only on the presumption that the operator $S[\kappa]S[\kappa]^T$ is invertible (or, more precisely, injective), a condition that will be verified below for the problem considered in this paper, and which holds in many other settings in which WRI has been advocated.
+This remarkable identity shows that the WRI objective function is a {\em weighted norm of the data residual $r[c]$}.
+Note that this conclusion depends only on the presumption that the operator $S[c]S[c]^T$ is invertible (or, more precisely, injective), a condition that will be verified below for the problem considered in this paper, and which holds in many other settings in which WRI has been advocated.
 
@@
 \begin{equation}
-  \label{eqn:stmodhompwtrans}
-S[c]^Te(z,t)=\frac{1}{2c}e\left(t +  \frac{|z_r-z|}{c}\right). 
+  \label{eqn:modhomtrans}
+S[c]^Te(z,t)=
+\left\{
+    \begin{array}{c}
+      \frac{1}{2c}e\left(t +  \frac{|z_r-z|}{c}\right), \,z_{\rm min} \le z \le z_{\rm max};\\
+      0, else.
+    \end{array}
+  \right.
 \end{equation}
 whence
+\[
+  S[c]S[c]^Te(t) = \frac{z_{\rm max}-z_{\rm min}}{4c^2}e(t),
+\]
+that is,
+\begin{equation}
+  \label{eqn:modhomnormal}
+  S[c]S[c]^T = \frac{z_{\rm max}-z_{\rm min}}{4c^2}I
+\end{equation}
+Thus the weight operator $W[c]$ appearing in \ref{eqn:defwrialt} takes
+the form
+\[
+  W_{\alpha}[c] = u(c) I,
+\]
+\begin{equation}
+  \label{eqn:wriwthom}
+u(c) = \frac{\alpha^2}{2}\left(\frac{z_{\rm max}-z_{\rm min}}{4c^2}+\alpha^2\right)^{-1}.
+\end{equation}
+
+Next, suppose that $q = w\delta(z-z_s)$, that is, the target source is
+a point source, so that $S[c]q = S_p[c]w$ in the notation used in
+the discussion of FWI. Thus \ref{eqn:defwrialt} can be re-written as
+\begin{equation}
+  \label{eqn:wriscrewed}
+  J_{\rm WRI}[c,d,q]=u[c]J_{\rm FWI}[c,d,w]
+\end{equation}
+
+Recall the wavelength-dependent family of problems introduced in the derivation and
+statement of Theorem \ref{thm:thm1}: target wave velocity $c_*$, wavelength parameter
+$\lambda$, parametrized family of wavelets $\wl$ and corresponding
+data $\dl$.
+
+Define
+\begin{equation}
+  \label{eqn:defbeta}
+\beta = \frac{z_{\rm max}-z_{\rm min}}{c_*^2} - 4 \alpha^2
+\end{equation}
+\begin{theorem}
+  \label{thm:thm2}
+  There exist $C>0$, $\lambda_0$ so that for $\lambda <
+  \lambda_0$, the minimizer of $J_{\rm WRI}[c,\dl,\wl \delta(\cdot-z_s)]$ on the
+  complement of $[c_*-C\lambda, c_*+C\lambda]$ is
+  \begin{itemize}
+  \item $c=c_{\rm max}$ if $\beta<0$;
+  \item $c=c_{\rm min}$ if $\beta>0$;
+  \item any $c < c_*-C\lambda$ or $>c_*+C\lambda$ if $\beta=0$.
+  \end{itemize}
+\end{theorem}
+\begin{proof}
+  From \ref{eqn:wriscrewed}, \ref{eqn:wriwthom}, and
+  \ref{eqn:fwiscrewed},
+  \[
+    J_{\rm WRI}[c,\dl,\wl\delta(\cdot-z_s)] =
+    \frac{\alpha^2}{2}\left(\frac{z_{\rm max}-z_{\rm
+          min}}{4c^2}+\alpha^2\right)^{-1}\frac{1}{2}\left(\frac{1}{4c^2}+\frac{1}{4c_*^2}\right)\|w_1\|^2
+  \]
+  \[
+    =
+    \frac{\alpha^2}{4}\frac{1+\frac{c^2}{c_*^2}}{z_{\rm 
+        max}-z_{\rm min}+4c^2\alpha^2}
+  \]
+  This linear fractional function of $c^2$ is increasing, decreasing,
+  or constant if $\beta>0, \beta<0$ or $\beta=0$, respectively.
+\end{proof}
+
+In other words, $J_{\rm WRI}$ has local minima far from the target
+velocity $c_*$, in the same way as does $J_{\rm FWI}$. One of the
+local minima will be the result of a local optimization almost surely,
+unless the initial estimate of $c$ is ``within a wavelength'' of the
+target velocity.
+
+\section{Discussion}
+
+(TBA)
+
+\section{Conclusion}
+
+(TBA)
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\append{1D Born Approximation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\append{Equivalence of Force and Defect}
 
-The perturbational system for \ref{eqn:awe1d} is:
-\begin{eqnarray}
-\label{eqn:dawe1d}
-\frac{\partial \delta p}{\partial t} +\kappa\frac{\partial
-  \delta v}{\partial z} + \delta \kappa\frac{\partial
-  v}{\partial z} &=& 0 \nonumber\\
-\rho \frac{\partial \delta v}{\partial t} + \frac{\partial \delta p}{\partial
-  z}&=&0\nonumber\\
-\delta p,\delta v&=&0, t \ll 0.
-\end{eqnarray}
-
-If $\kappa=\kappa_0, \rho=\rho_0 $ are constant, it is possible to develop explicit
-expressions for the solution of this system. As before, only $\kappa$ will be
-subject to perturbation in any of the problems considered in this
-report, so assume that that $\delta\rho=0$.
-
-From \ref{eqn:sol1dv}
-\[
-  \frac{\partial v}{\partial z}(z,t) = \frac{1}{\kappa_0} f(z,t) 
-\]
-\begin{equation}
-  \label{eqn:sol1ddvdz}
- - \frac{1}{2\kappa_0c_0} \int\, dz_2\, \frac{\partial f}{\partial t}\left(z_2,t - \frac{|z-z_2|}{c_0}\right)
-\end{equation}
-Thus using equation \ref{eqn:sol1dp} and \ref{eqn:dawe1d}, 
-\[
-  \delta p(z,t) =- \frac{1}{2c_0} \int \, dz_1 \, \delta
-  \kappa(z_1)\frac{\partial v}{\partial z}\left(z_1,t - \frac{|z-z_1|}{c_0}\right)
-\]
-\[
-  =-\frac{1}{2\kappa_0c_0} \int \,dz_1\, \delta
-  \kappa(z_1) f \left(z_1,t - \frac{|z-z_1|}{c_0}\right)
-  \]
-\begin{equation}
-  \label{eqn:sol1ddp}
-  + \frac{1}{4\kappa_0c_0^2} \int\,dz_2 \, \int \,dz_1\,\delta 
-  \kappa(z_2) \frac{\partial f}{\partial t}\left(z_1,t - \frac{|z_1-z_2|+|z_2-z|}{c_0}\right)
-\end{equation}
-
-A convenient check on this calculation is available for constant
-$\delta \kappa = \delta \kappa_0$. Directly perturb the RHS of equation 
-\ref{eqn:sol1dp}, to obtain
-\[
-  \delta p(z,t) = -\frac{\delta c_0}{2 c_0^2} \int
-  \,dz_1\,f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
-  +\frac{\delta c_0}{2c_0^3}\int
-  \,dz_1\,\frac{\partial f}{\partial
-    t}\left(z_1,t-\frac{|z-z_1|}{c_0}\right)|z-z_1|
-\]
-Since
-\[
-  \delta c_0 = \delta \left(\frac{\kappa_0}{\rho_0}\right)^{1/2} =
-  \frac{1}{2} \left(\frac{\kappa_0}{\rho_0}\right)^{-1/2}\frac{\delta
-    \kappa_0}{\rho_0}
-\]
-\[
-  = \frac{1}{2}\frac{\delta \kappa \rho_0^{1/2}}{\kappa_0^{1/2}\rho_0}
-  \]
-\[
-=\frac{1}{2} \frac{\delta
-  \kappa_0}{\kappa_0}\left(\frac{\kappa_0}{\rho_0}\right)^{1/2}
-=\frac{1}{2} \frac{c_0\delta \kappa_0}{\kappa_0}
-\]
-the identity above can be rewritten as
-\[
-  \delta p(z,t) = -\frac{\delta \kappa_0}{4 \kappa_0 c_0} \int
-  \,dz_1\,f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
-  \]
-\begin{equation}
-  \label{eqn:dpraw} 
-  +\frac{\delta \kappa_0}{4\kappa_0c_0^2}\int
-  \,dz_1\,\frac{\partial f}{\partial
-    t}\left(z_1,t-\frac{|z-z_1|}{c_0}\right)|z-z_1|
-\end{equation}
-The critical quantity is the integral in the second term:
-\[
-  \int \,dz_1\,\frac{\partial f}{\partial t}\left(z_1,t-\frac{|z-z_1|}{c_0}\right)|z-z_1|
-\]
-\[
-  = \int_{-\infty}^z \,dz_1\,\frac{\partial f}{\partial
-    t}\left(z_1,t-\frac{z-z_1}{c_0}\right)(z-z_1)
-\]
-\[
-   + \int^{\infty}_z \,dz_1\,\frac{\partial f}{\partial t}\left(z_1,t-\frac{z_1-z}{c_0}\right)(z_1-z)
- \]
- \[
-   =\int_{-\infty}^z \,dz_1\,\int_{z_1}^{z} \,dz_2\, \frac{\partial f}{\partial
-     t}\left(z_1,t-\frac{z-z_2+z_2-z_1}{c_0}\right)
- \]
- \[
-   + \int^{\infty}_z \,dz_1\,\int_{z}^{z_1}\,dz_2\,\frac{\partial
-     f}{\partial t}\left(z_1,t-\frac{z_1-z_2 + z_2-z}{c_0}\right)
- \]
- Checking signs, one sees that this sum is equal to
- \[
-   =\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
-     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
- \]
- \[
-   + \int^{\infty}_z \,dz_1\,\int_{-\infty}^{\infty}\,dz_2\,\frac{\partial
-     f}{\partial t}\left(z_1,t-\frac{|z_1-z_2| + |z_2-z|}{c_0}\right)
- \]
- \[
-   -\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{z_1} \,dz_2\, \frac{\partial f}{\partial
-     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
- \]
-\[
-   -\int_{-\infty}^z \,dz_1\,\int_{z}^{\infty} \,dz_2\, \frac{\partial f}{\partial
-     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
- \]
- \[
-   - \int^{\infty}_z \,dz_1\,\int_{-\infty}^{z}\,dz_2\,\frac{\partial
-     f}{\partial t}\left(z_1,t-\frac{|z_1-z_2| + |z_2-z|}{c_0}\right)
- \]
- \[
-   - \int^{\infty}_z \,dz_1\,\int_{z_1}^{\infty}\,dz_2\,\frac{\partial
-     f}{\partial t}\left(z_1,t-\frac{|z_1-z_2| + |z_2-z|}{c_0}\right)
- \]
-\[
-   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
-     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
- \]
- \[
-   -\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{z_1} \,dz_2\, \frac{\partial f}{\partial
-     t}\left(z_1,t-\frac{z-z_2+z_1-z_2}{c_0}\right)
- \]
-\[
-   -\int_{-\infty}^z \,dz_1\,\int_{z}^{\infty} \,dz_2\, \frac{\partial f}{\partial
-     t}\left(z_1,t-\frac{z_2-z+z_2-z_1|}{c_0}\right)
- \]
- \[
-   - \int^{\infty}_z \,dz_1\,\int_{-\infty}^{z}\,dz_2\,\frac{\partial
-     f}{\partial t}\left(z_1,t-\frac{z_1-z_2 + z-z_2}{c_0}\right)
- \]
- \[
-   - \int^{\infty}_z \,dz_1\,\int_{z_1}^{\infty}\,dz_2\,\frac{\partial
-     f}{\partial t}\left(z_1,t-\frac{z_2-z_1 + z_2-z}{c_0}\right)
- \]
-\[
-   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
-     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
- \]
- \[
-   -\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{z_1} \,dz_2\, \frac{\partial f}{\partial
-     t}\left(z_1,t-\frac{z+z_1-2z_2}{c_0}\right)
- \]
-\[
-   -\int_{-\infty}^z \,dz_1\,\int_{z}^{\infty} \,dz_2\, \frac{\partial f}{\partial
-     t}\left(z_1,t-\frac{2z_2-z-z_1}{c_0}\right)
- \]
- \[
-   - \int^{\infty}_z \,dz_1\,\int_{-\infty}^{z}\,dz_2\,\frac{\partial
-     f}{\partial t}\left(z_1,t-\frac{z_1+ z-2z_2}{c_0}\right)
- \]
- \[
-   - \int^{\infty}_z \,dz_1\,\int_{z_1}^{\infty}\,dz_2\,\frac{\partial
-     f}{\partial t}\left(z_1,t-\frac{2z_2-z_1-z}{c_0}\right)
- \]
-\[
-   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
-     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
- \]
- \[
-   -\frac{c_0}{2}\int_{-\infty}^z \,dz_1\,\int_{-\infty}^{z_1} \,dz_2\, \frac{\partial }{\partial
-     z_2}f\left(z_1,t-\frac{z+z_1-2z_2}{c_0}\right)
- \]
-\[
-   +\frac{c_0}{2}\int_{-\infty}^z \,dz_1\,\int_{z}^{\infty} \,dz_2\, \frac{\partial}{\partial
-     z_2}f\left(z_1,t-\frac{2z_2-z-z_1}{c_0}\right)
- \]
- \[
-   - \frac{c_0}{2}\int^{\infty}_z \,dz_1\,\int_{-\infty}^{z}\,dz_2\,\frac{\partial
-     }{\partial z_2}f\left(z_1,t-\frac{z_1+ z-2z_2}{c_0}\right)
- \]
- \[
-   +\frac{c_0}{2} \int^{\infty}_z \,dz_1\,\int_{z_1}^{\infty}\,dz_2\,\frac{\partial
-     }{\partial z_2}f\left(z_1,t-\frac{2z_2-z_1-z}{c_0}\right)
- \]
-\[
-   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
-     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
- \]
- \[
-   -\frac{c_0}{2}\int_{-\infty}^z \,dz_1\, f\left(z_1,t-\frac{z-z_1}{c_0}\right)
-   -\frac{c_0}{2}\int_{-\infty}^z \,dz_1\,f\left(z_1,t-\frac{z-z_1}{c_0}\right)
- \]
- \[
-   - \frac{c_0}{2}\int^{\infty}_z \,dz_1\,f\left(z_1,t-\frac{z_1-z}{c_0}\right)
-   -\frac{c_0}{2} \int^{\infty}_z \,dz_1\,f\left(z_1,t-\frac{z_1-z}{c_0}\right)
- \]
- \[
-   =\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
-     t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
- \]
- \[
-   -c_0\int_{-\infty}^{\infty} \,dz_1\, f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
- \]
- Insert the identity just established in the second term on the RHS of
- \ref{eqn:dpraw}:
- \[
-  \delta p(z,t) = -\frac{\delta \kappa_0}{4 \kappa_0 c_0} \int
-  \,dz_1\,f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
-\]
-\[
-  +\frac{\delta \kappa_0}{4\kappa_0c_0^2}\left(\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
-      t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right) \right.
-\]
-\[
-  \left. -c_0\int_{-\infty}^{\infty} \,dz_1\,
-    f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)\right)
-\]
-\[
-=-\frac{\delta \kappa_0}{2\kappa_0c_0}\int_{-\infty}^{\infty} \,dz_1\,
-f\left(z_1,t-\frac{|z-z_1|}{c_0}\right)
-\]
-\begin{equation}
-  \label{eqn:dpraw}
-+  \frac{\delta \kappa_0}{4\kappa_0c_0^2}\int_{-\infty}^{\infty} \,dz_1\,\int_{-\infty}^{\infty} \,dz_2\, \frac{\partial f}{\partial
-      t}\left(z_1,t-\frac{|z-z_2|+|z_2-z_1|}{c_0}\right)
-  \end{equation}
-which is identical to the expression \ref{eqn:sol1ddp} for the case of constant
-$\delta \kappa = \delta \kappa_0$.
-
-An important special case of the identity \ref{eqn:sol1ddp} is the 1D point source,
-$f(z,t)=w(t)\delta(z-z_s)$: then
-\[
-  \delta p(z,t) = -\frac{\delta \kappa(z_s)}{2\kappa_0c_0}w\left(t - \frac{|z-z_s|}{c_0}\right)
-\]
-\begin{equation}
-  \label{eqn:sol1ddppw}
-  +\frac{1}{4\kappa_0c_0^2} \int \,dz_1\,\delta 
-  \kappa(z_1) \frac{\partial w}{\partial t}\left(t - \frac{|z_s-z_1|+|z_1-z|}{c_0}\right)
-\end{equation}
+(TBA)
 

### 2020-02-14 repro/papers/wri1d/paper.tex
first draft

@@
   driven by wave equation simulation, has proven its ability to
-  extract detailed maps of earth structture from near-surface seismic
+  extract detailed maps of earth structure from near-surface seismic
   observations. However, the least squares error function at the heart
@@
 approach was first suggested in the 1980's
-(\cite{BamChavLai:79,Tara:84a,KolbColLai:86,Crasetal:90}, and many
+(\cite[]{BamChavLai:79,Tara:84a,KolbColLai:86,Crasetal:90}, and many
 other papers since then). Usually some form of regularization is
@@
 \end{equation}
-The penalty parameter $\alpha$ may be chosen small, and examples given
+Examples given
 by \cite{LeeuwenHerrmannWRI:13} and others suggest that the problem
@@
 \end{equation}
+That is, the support of f will be assumed to lie in the strip $[z_{\rm min},z_{\rm max}] \times \bR$.
 As it will turn out, the choice of spatial interval $[z_{\rm min},z_{\rm max}]$ is arbitrary, so long as it has positive length and contains $z_s$ in its interior.
@@
 \ref{eqn:defmoddom}. The role of compact support in $z$ in ensuring
-boundedness is also evident. % It is also evident from that limiting source to have bounded support in $z$ is necessary make $S[c]$ bounded.
+boundedness is also evident.
+%It is also evident from that limiting source to have bounded support in $z$ is necessary make $S[c]$ bounded.
 
-With this framework, the data is presumed to lie in the range space of $S[c]$: $d \in L^2[0,T]$.
+With this framework, the data is presumed to lie in the range space of
+$S[c]$: $d \in L^2[0,T]$. It is simple to verify from the identity
+\ref{eqn:defmodhom} that $S[c]$ is also surjective. That is, any data
+at all can be fit by $S[c]$ with an appropriate choice of input. This
+observation is important in the development of WRI, to be explained
+below.
 
-The version of FWI discussed here presumes that the source field corresponding to the data $d$ is supported at a point $z_s \ne z_r$, and is known.
-This field depends on a function of time (``wavelet'') $w \in L^2[0,T]$.
-Formally, the resulting acoustic field satisfies the system \ref{eqn:awe1d} with $f=w(t)\delta(z-z_s)$.
-Inserting this expression in the explicit expression \ref{eqn:defmodhom}, obtain
+The version of FWI discussed here presumes that the source field
+corresponding to the data $d$ is supported at a point $z_s \ne z_r$,
+and is known. This field depends on a function of time (``wavelet'')
+$w \in L^2(\bR)$. Formally, the resulting acoustic field satisfies the
+system \ref{eqn:awe1d} with $f=w(t)\delta(z-z_s)$. Inserting this
+expression in the explicit expression \ref{eqn:defmodhom}, obtain
 \begin{equation}
@@
 relation of the various scales, it is advantageous to introduce a
-family of source wavelets, depending on a parameter $\lambda$ playing
+family of source wavelets, depending on a parameter $\lambda$, having
+dimensions of time and playing
 the role of wavelength:
@@
 In the definition \ref{eqn:wfam}, the ``mother wavelet'' $w_1\in
-C_0^{\infty}(0,T)$, and the scaling is chosen so fhat
+C_0^{\infty}(0,1)$, and the scaling is chosen so fhat
 $\|\wl\|_{L^2(\bR)}$ is independent of $\lambda>0$.
@@
 \end{equation}
-Note that $\mbox{supp }\wl \subset [0,\lambda T]$, so
+Note that $\mbox{supp }\wl \subset [0,\lambda]$, so
 \begin{equation}
@@
   \mbox{supp }S_p[c]\wl \subset \left[\frac{z_s-z_r}{c},\lambda
-    T+\frac{z_s-z_r}{c}\right] \cap [0,T]
+    +\frac{z_s-z_r}{c}\right] \cap [0,T]
 \end{equation}
 The transit time condition \ref{eqn:time} implies that there exists
-$\lambda_*>0$ so that for $\lambda < \lambda_*$,
+$\lambda_0>0$ so that for $\lambda < \lambda_0$,
 \[
-  \lambda T+\frac{z_s-z_r}{c} < T
+  \lambda +\frac{z_s-z_r}{c} < T
 \]
 for all admissible $c$. That is, 
-for $\lambda <\lambda_*$
+for $\lambda <\lambda_0$, $\mbox{supp }S_p[c]\wl \subset(0,T)\mbox{
+  for }c \in [c_{\rm min},c_{\rm max}]$, and 
 \begin{equation}
@@
 objective functions for velocity estimation: in this context, that
-means the behaviour for $c$ far from $c_*$.  Pick $\delta c >0$ for
-which $[c_*-\delta c, c_* + \delta c] \subset [c_{\rm min},c_{\rm
-  max}]$, that is, that the set of admissible $c$ further than $\delta
-c$ from $c_*$ is non-void.
-
-Then there is $\lambda_{**}$ depending on $\delta c, T, and
-|z_s-z_r|$ so that if $|c-c_*|>\delta c$ and $\lambda<\lambda_{**}$, then
+means the behaviour for $c$ far from $c_*$. Define
+\begin{equation}
+  \label{eqn:defL}
+  L = \frac{2 c_{\rm max}^2}{|z_s-z_r|}.
+\end{equation}
+Then if $|c_*-c| > L\lambda$,
 \[
-  \left|\frac{|z_s-z_r|}{c}-\frac{|z_s-z_r|}{c_*}\right| > 2\lambda T,
+  \left|\frac{|z_s-z_r|}{c}-\frac{|z_s-z_r|}{c_*}\right|
+  =\frac{|c-c_*||z_s-z_r|}{cc_*} 
 \]
-that is, the infima of the supports of $S_p[c]\wl, \dl=S_p[c_*]\wl$
+\[
+\ge \frac{|c-c_*||z_s-z_r|}{c_{\rm max}^2} > L\lambda
+\frac{|z_s-z_r|}{c_{\rm max}^2} = 2\lambda.
+\]
+That is, the infima of the supports of $S_p[c]\wl, \dl=S_p[c_*]\wl$
 are further apart than the lengths of these supports. In that case,
@@
   \label{thm:thm1}
-  There exists $C>0$, $\lambda_0$ so that for $\lambda <
-  \lambda_0$, the minimizer of $J_{\rm FWI, \lambda}[c]$ on the
-  complement of $[c_*-C\lambda, c_*+C\lambda]$ is $c=c_{\rm max}$.
+  For $L>0$ given by equation \ref{eqn:defL} and $\lambda <
+  \lambda_0$, the minimizer of $J_{\rm FWI}[c,\dl,\wl]$ on the
+  complement of $[c_*-L\lambda, c_*+L\lambda]$ is $c=c_{\rm max}$.
 \end{theorem}
@@
 wavelength, minimization of $J_{\rm FWI}$ yields a local minimizer far from
-the target $c_*$ which generates the (noise-free) data.
+the target velocity $c_*$ that generates the (noise-free) data.
 
@@
 minimization of
+\[
+  J_{\rm WRI}[c,d,q] =
+  \mbox{min}_f \frac{1}{2}(\| d-S[c]f\|^2+\alpha^2\|f-q\|^2)
+\]
+over $c$. This is the approach that I shall pursue here.
+
+It will turn out to be convenient to define the residual with the
+target source $q$ as $r[c]=d-S[c]q$, and set $g=f-q$. Then this
+definition can be rewritten as
 \begin{equation}
 \label{eqn:defvpm}
-  J_{\rm WRI}[c,d,q] =
-  \mbox{min}_f \frac{1}{2}(\| d-S[c]f\|^2+\alpha^2\|f-q\|^2)
+  J_{\rm WRI}[c,d,q] =\mbox{min}_g \frac{1}{2}(\| r[c]-S[c]g\|^2+\alpha^2\|g\|^2)
 \end{equation}
-over $c$. This is the approach that I shall pursue here.
 
@@
 
-``Most'' source fields $f$ are non-radiating,
-that is, $S[c]f=0$. The orthocomplement of the subspace of
-non-radiating sources is the range of the adjoint operator
-$S[c]^T$. This observation suggest that the sources may be re-parametrized as images of
-$S[c]^T$. Write $f = g + q$: then the second term in
-the objective displayed in \ref{eqn:defwri} is simply $\|g\|^2$.
-Decompose $g = S[c]^Te + n$, in
-which $e$ is the same type of object as $d$ and $S[c]n=0$ (that
-is, $n$ is a non-radiating
-source), and note that the decomposition is orthogonal. Then
+``Most'' source fields $f$ are non-radiating, that is, $S[c]f=0$, and
+such sources contribute nothing to the data fit term in the definition
+of $J_{\rm WRI}$. If the domain and range of $S$ were finite
+dimensional (which of course they are, after discretization), then the
+Fundamental Theorem of Linear Algebra identifies the null space of
+$S[c]$ (the non-radiating sources) as the orthocomplement of the range
+of the transpose $S[c]^T$ \cite[]{Strang:93}. In the
+infinite-dimensional setting of this paper, the Closed Range Theorem
+\cite[]{Yosida} states that the same is true if the range of $S[c]$ is
+closed. There are various ways to ensure this property, but the
+simplest is relevant here:
+\begin{quote}
+  For all admissible models $c$, $S[c]$ is surjective.
+\end{quote}
+That is, {em any data can be fit exactly using the extended model
+  space (the domain of $S$)}.
+This property is a characteristic of of extended modeling
+methods \cite[]{geoprosp:2008}: the ability to fit any data appears to
+be a essential for such methods to produce objectives
+without spurious local minima. It holds for the problems for which WRI
+has been advocated. For the simple model problem considered
+here, surjectivity follows from the explicit expression for $S[c]f$,
+as was noted in the discussion following equation \ref{eqn:defmodhom}.
+
+Assuming that $S[c]$ is surjective for any admissible $c$, the
+orthocomplement of the subspace of non-radiating sources is the range
+of the adjoint operator $S[c]^T$. In the
+definition \ref{eqn:defvpm}, decompose $g = S[c]^Te + n$, in
+which $e$ is the same type of object as $d$ and $S[c]n=0$ (that is,
+$n$ is a non-radiating source), and note that the decomposition is
+orthogonal. Then
 \[
@@
   \label{eqn:defvpmred}
-   \mbox{min}_{e} \frac{1}{2}(\|r-S[c]S[c]^Te\|^2+\alpha^2\|S[c]^Te\|^2)
+   \mbox{min}_{e} \frac{1}{2}(\|r[c]-S[c]S[c]^Te\|^2+\alpha^2\|S[c]^Te\|^2).
  \end{equation}
- in which $r[c]=d-S[c]q$ is the data residual.
 
@@
 \]
-Assume that $S[c]S[c]^T$ is invertible - this is in fact for the problem outlined in the introduction, as will be verified shortly. Then
+Since the null space of $S[c]$ is orthogonal to the range of $S[c]^T$
+under the surjectivity assumption, $S[c]S[c]^T$ is injective, whence
 \begin{equation}
@@
 This remarkable identity shows that the WRI objective function is a {\em weighted norm of the data residual $r[c]$}.
-Note that this conclusion depends only on the presumption that the operator $S[c]S[c]^T$ is invertible (or, more precisely, injective), a condition that will be verified below for the problem considered in this paper, and which holds in many other settings in which WRI has been advocated.
 
@@
   \label{thm:thm2}
-  There exist $C>0$, $\lambda_0$ so that for $\lambda <
+  For $L$ as defined in \ref{eqn:defL}, and $\lambda <
   \lambda_0$, the minimizer of $J_{\rm WRI}[c,\dl,\wl \delta(\cdot-z_s)]$ on the
-  complement of $[c_*-C\lambda, c_*+C\lambda]$ is
+  complement of $[c_*-L\lambda, c_*+L\lambda]$ is
   \begin{itemize}
@@
   \item $c=c_{\rm min}$ if $\beta>0$;
-  \item any $c < c_*-C\lambda$ or $>c_*+C\lambda$ if $\beta=0$.
+  \item any $c < c_*-L\lambda$ or $>c_*+L\lambda$ if $\beta=0$.
   \end{itemize}
@@
   From \ref{eqn:wriscrewed}, \ref{eqn:wriwthom}, and
-  \ref{eqn:fwiscrewed},
+  \ref{eqn:fwiscrewed}, if $|c-c_*| > L\lambda$,
   \[
@@
 \section{Discussion}
+Theorems \ref{thm:thm1} and \ref{thm:thm2} call out the chief
+conclusions of this work: that at least for the 1D acoustic
+transmission inverse problem, both $J_{\rm FWI}[c,d,w]$ and
+$J_{\rm WRI}[c,d,q]$ exhibit local minima (in $c$) far from the global
+minimum for consistent data, and the domain of attraction of the
+global minimizer can be arbitrarily small. These are striking
+conclusions, but the 1D acoustic transmission problem is very special
+and lacks fidelity to field practice. I shall show how these
+approaches to solving this special problem share properties with a
+much larger family of inversion methods. The theory developed to
+explain these properties suggests methods that may not suffer the
+non-convexity of FWI and WRI, and in fact do not in several cases that
+I will mention.
 
-(TBA)
+First, a consequence of the results proven here: $J_{\rm FWI}[c,d,w]$
+and $J_{\rm WRI}[c,d,q]$ are not smooth as joint functions of model
+($c$) and data ($d$) vectors. If they were, their derivatives would be
+bounded uniformly over bounded sets in $c,d$, but the two main results
+show that this is not the case. As $\lambda \rightarrow 0$, $d=\dl$
+varies within a ball $B \subset L^2([0,T])$ of radius $\|d_1\|$ (since
+the $\|\dl\|$ is independent of $\lambda$), but the value of either
+objective changes from a positive value (bounded away from zero
+independently of $\lambda$) to zero over an interval of $c$ of length
+$O(\lambda)$. Therefore the derivatives with respect to $c$ of both
+objective functions are not bounded over
+$[c_{\rm min},c_{\rm max}] \times B$.
+
+While lack of smoothness is not in itself the most important property
+established in the preceding sections, it is a necessary condition for
+stable and reliable parameter recovery via local
+optimization. Moreover, necessary conditions for smoothness are
+known for a much wider class of quadratic form objectives for inverse
+problems.
+
+These results
+concern optimization problems of the genral form
+\begin{equation}
+  \label{eqn:gen1}
+  \mbox{Given }d\mbox{, find }c\mbox{ to extremize }
+   J[c,d] = \langle G[c]d, A[c] G[c]d \rangle
+\end{equation}
+In this prescription, $d$ is a data vector, as in the examples above,
+$c$ is a vector of material parameters to be estimated, $G[c]$ is a
+$c$-dependent family of operators having the same nature as the
+simulation operators $S[c]$ and $S_p[c]$ figuring in the preceding
+discussion, and $A[c]$ is an operator-valued function of $c$, on the range
+of $G[c]$.
+
+\cite{StolkSymes:03} assume that the
+operator-valiued function $G[c]$ is of a class typical of modeling
+operators for wave equation inverse problems, or their inverses or
+adjoints. The precise characterization of the so-called {\em microlocally elliptic Fourier Integral
+Operators} is quite technical \cite[]{Dui:95}. Roughly speaking, such operators map high-frequency localized wave packets to other
+such packets with well-defined changes of position and direction of
+oscillation. If the dependence of such an operator $G[c]$ on $c$ is of sufficiently full rank, in
+the sense that destination packets can be shifted in any direction by
+changing $c$ appropriately, along with a couple of other technical
+assumptions, one can conclude that $J$ as defined in \ref{eqn:gen1} is
+smooth in $c$ and $d$ jointly if and only if the operator $A[c]$ is a 
+{\em pseudodifferential operator} - again, a class of operators whose
+precise definition is quite technical. However these operators also
+have a rough characterization: they do not change the location of
+oscillatory wave packets
+or alter their direction of oscillation, but only scale such packets
+by smooth functions, to good approximation.
+
+With a bit of fiddling, the FWI problem \ref{eqn:deffwipt} for 1D 
+acoustic transmission inversion can be rewritten in the form \ref{eqn:gen1}. Note
+that $S_p[c]$ is invertible (more specifically, has a right inverse; if [0,T] were
+extended to $(-\infty,\infty)$ it would have a left inverse too). From
+the defintion \ref{eqn:dfam} of the data family $\dl$, one sees that
+$\wl = S_p[c_*]^{-1}\dl$. Therefore
+\[
+  J_{\rm FWI}[c,\dl,\wl] = \frac{1}{2}\|(I-S_p[c]S_p[c_*]^{-1})\dl\|^2
+\]
+\[
+  =\frac{1}{2}(\|\dl\|^2 + \|S_p[c]S_p[c_*]^{-1}\dl\|^2) + \langle \dl,
+  S_p[c]S_p[c_*]^{-1}\dl\rangle
+\]
+For the operator family $S_p[c]$ defined above, it is easy to see that 
+the second term is smooth in $c$, and the first is constant. The third 
+can be rewritten as 
+\[
+  \langle \dl,  S_p[c]S_p[c_*]^{-1}\dl\rangle = \langle S[c]^T\dl, 
+  (S_p[c]^T S_p[c_*])^{-1}S_p[c]^Td\rangle 
+\]
+which has the form \ref{eqn:gen1} with the choices $G[c]=S_p[c]^T$,
+$A[c]=(S_p[c]^T S_p[c_*])^{-1}$. A similar manipulation exhibits
+$J_{\rm WRI}$ as the sum of harmless terms and a quadratic form of the
+form \ref{eqn:gen1}.
+
+Given the rough understanding of the results of \cite{StolkSymes:03}
+sketched above, one would conclude that neither $J_{\rm FWI}$ nor
+$J_{\rm WRI}$ are likely to be smooth jointly in $c$ and $d$. Indeed,
+apart from scale, $S_p[c]$ is composition with a shift (translation)
+by $(z_{\rm max}-z_{\rm min})c^{-1}$, so
+$A[c]=(S_p[c]^T S_p[c_*])^{-1}$ is composition with a shift by
+$(z_{\rm max}-z_{\rm min})(c^{-1}-c_*^{-1})$. Thus application of
+$A[c]$ does not leave the position of a wave packet fixed, unless
+$c=c_*$ - and indeed $A[c]$ is not a pseudodifferential operator
+unless $c=c_*$. On the other hand, $G[c]$ is a shift operator, the
+simplest prototype of an elliptic Fourier Integral Operator. Therefore
+the conclusion, derived directly from Theorem \ref{thm:thm1}, that
+$J_{\rm FWI}$ is not smooth jointly in $c$ and $d$ would also appear
+to follow from the main result of \cite[]{StolkSymes:03}. This
+conclusion can be made precise by proper attention to detail, and the
+same is true of $J_{\rm WRI}$.
+
+In the context of the 1D acoustic transmission problem as formulated
+here, the question immediately arises: do quadratic forms \ref{eqn:gen1}
+exist that are smooth jointly in $c$ and $d$, and whose global
+minimizer is the correct velocity $c=c_*$? An affirmative answer is
+provided for precisely this example problem in
+\cite[]{wwsorcas:19-02}. The operator $G[c]=S_p[c]^T$ is precisely the
+same as appeared in the reformulation of $J_{\rm FWI}$. Ignoring a
+$c-$dependent multiplier,
+\[
+  A[c]u(t) = t^2u(t).
+\]
+Applied to the wavelength-dependent family of data $\dl$ and source
+wavelets $\wl$ used repeatedly throughout this paper, $A[c]$ yields a
+vanishing result as $\lambda \rightarrow 0$ for the correct velocity
+$c=c_*$ and a stably non-zero result otherwise. For these choices, it
+can be established that
+\[
+  \frac{d}{dc}J[c,\dl]
+  \left\{\begin{array}{c}
+           > 0 \mbox{ if }c<c_*+O(\lambda),\\
+           < 0 \mbox{ if }c>c_*+O(\lambda)
+         \end{array}
+       \right.
+     \]
+That is, all local minima of $J[c,\dl]$ lie within $O(\lambda)$ (``a
+wavelength'') of the target velocity $c_*$. Not surprisingly the
+wavelength parameter also regulates the accuracy of the inversion.
+
+The reader is directed to \cite[]{HuangNammourSymesDollizal:SEG19} for
+an extensive discussion of other similar source extension methods for
+various wave inversion problems, and to references cited there for
+references to earlier work on this topic.
 
 \section{Conclusion}
+The tendency of iterative FWI to become trapped in uninformative local
+minima has been much discussed and still drives a substantial
+worldwide research program, almost 35 years after the phenomenon was
+first identified. WRI is amongst the many remedies proposed for this
+pathology, and numerical experiments have appeared to suggest that it
+may succeed. The example investigated in this report is simple enough
+to allow for rigorous mathematical conclusions regarding the behaviour
+of both FWI and WRI. The complete explanation for the behaviour of FWI
+is no surprise. As it turns out, the same conclusion may be reached
+for WRI: in this example at least, iterative minimization of the WRI
+objective it is no more likely to produce a useful estimate of wave
+velocity than is FWI, and for the same reason - indeed, the two are
+very closely linked (equations \ref{eqn:defwrialt},
+\ref{eqn:defwriwt}).
 
-(TBA)
+While these specific conclusions are of course tied to the extremely
+simple homogeneous acoustic 1D transmission inverse problem studied
+here, the relations \ref{eqn:defwrialt}, \ref{eqn:defwriwt} are
+straightforward algebraic properties of WRI and appear to link it
+closely to FWI in any wave propagation setting. As explained in the
+discussion section, even mere smoothness of a quadratic form objective
+function in both the model and data parameters may impose restrictions
+on the operators involved in the construction of the form. These
+restrictions are generally not met by any version of FWI. An
+examination of other versions of WRI from this point of view may prove
+informative.
 
@@
 \label{eqn:awe1dptsrc}
-\frac{\partial p}{\partial t} +\kappa\frac{\partial 
+\frac{\partial p}{\partial t} +\rho c^2\frac{\partial 
   v}{\partial z} &=& w(t)\delta(z-z_s) \nonumber\\
@@
 
-Now assume that $\kappa=\kappa_0 \in \bR^+, \rho=\rho_0 \in \bR^+, c_0
-= \sqrt{\kappa_0/\rho_0}$ are constant. 
 In $z \ne z_s$, the right hand side 
@@
 \[
-p(z,t)=a\left(t -\frac{|z-z_s|}{c_0}\right), \, v(z,t)=\mbox{sgn}(z-z_s) b\left(t -
-  \frac{|z-z_s|}{c_0}\right)
+p(z,t)=a\left(t -\frac{|z-z_s|}{c}\right), \, v(z,t)=\mbox{sgn}(z-z_s) b\left(t -
+  \frac{|z-z_s|}{c}\right)
 \]
 From the second dynamical equation (Newton's law) it follows that $b =
-a/(\rho_0 c_0)$. The singularity on the LHS of the first dynamical
+a/(\rho c)$. The singularity on the LHS of the first dynamical
 equation (constitutive law) is
 \[
-\kappa_0 [v]_{z=z_s}\delta(z-z_s) =
-2\kappa_0 b\delta(z-z_s) = 2c_0 a\delta(z-z_s).
+\rho c^2 [v]_{z=z_s}\delta(z-z_s) =
+2\rho c^2 b\delta(z-z_s) = 2c a\delta(z-z_s).
 \] 
 This must in turn equal the RHS of the constitutive law, whence
-$a=w/(2c_0)$. Thus
+$a=w/(2c)$. Thus
 \begin{eqnarray}
 \label{eqn:sol1dptsrc}
-p(z,t) &=& \frac{1}{2c_0}w\left(t - \frac{|z-z_s|}{c_0}\right) \nonumber \\
-v(z,t) &=& \mbox{sgn}(z-z_s)\frac{1}{2\kappa_0}w\left(t -\frac{|z-z_s|}{c_0}\right)
+p(z,t) &=& \frac{1}{2c}w\left(t - \frac{|z-z_s|}{c}\right) \nonumber \\
+v(z,t) &=& \mbox{sgn}(z-z_s)\frac{1}{2\rho c^2}w\left(t -\frac{|z-z_s|}{c}\right)
            \nonumber \\
@@
 \label{eqn:awe1d}
-\frac{\partial p}{\partial t} +\kappa_0\frac{\partial 
+\frac{\partial p}{\partial t} +\rho c^2\frac{\partial 
   v}{\partial z} &=& f(z,t) \nonumber\\
-\rho_0 \frac{\partial v}{\partial t} + \frac{\partial p}{\partial 
+\rho \frac{\partial v}{\partial t} + \frac{\partial p}{\partial 
   z}&=&0\nonumber\\
@@
 \label{eqn:sol1dp}
-p(z,t) &=& \frac{1}{2c_0}\int dz_1 f\left(z_1,t -
-           \frac{|z-z_1|}{c_0}\right) \\
+p(z,t) &=& \frac{1}{2c}\int dz_1 f\left(z_1,t -
+           \frac{|z-z_1|}{c}\right) \\
   \label{eqn:sol1dv}
-v(z,t) &=& \frac{1}{2\kappa_0} \int dz_1 \mbox{sgn} (z-z_1) f\left(z_1,t - \frac{|z-z_1|}{c_0}\right)
+v(z,t) &=& \frac{1}{2\rho c^2} \int dz_1 \mbox{sgn} (z-z_1) f\left(z_1,t - \frac{|z-z_1|}{c}\right)
 \end{eqnarray}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\append{Equivalence of Force and Defect}
+\append{Equivalence of point and non-point sources}
 
-(TBA)
+As noted in the text, the point source $w(t)\delta(z-z_s)$ is not a
+member of the domain of the simulation operator $S[c]$, as it is not
+square-integrable. The object of this appendix is to construct a square-integral right-hand side in the
+system \ref{eqn:awe1d} for which the pressure field $p$
+is the same as that of the weak solution to the point source problem \ref{eqn:awe1dptsrc}
+constructed in the last section, near the receiver point $z=z_r$, and
+to exhibit this square-integrable replacement for the point source as
+the image of the point source wavelet under a bounded extension map,
+as in equation \ref{eqn:ext}.
+
+One step in this construction involve buidling a constitutive defect
+(pressure) source that is equivalent, in the sense of generating the
+same solution, to a force (velocity) source. This construction is
+presented here in the context of the 1D acoustic problem, but is a
+special case of a much more general construction of considerable
+interest in its own right \cite[]{BurridgeKnopoff:64}.
+
+Let
+$\epsilon$ be any positive number $<|z_r-z_s|$. 
+Denote by $(p,\bv)$ the (weak) solution \ref{eqn:sol1dptsrc} of the point source
+problem constructed in the last section.
+Pick $\phi \in C_0^{\infty}(\bR)$ so that
+$\phi = 1$ if $|z-z_s| \le \epsilon/2$ and $\phi(z)=0$ if $|z-z_s|
+\ge\epsilon$. Set $p_0=p(1-\phi)$, $v_0=v(1-\phi)$. Then
+\begin{eqnarray}
+\label{eqn:awecut}
+\frac{\partial p_0}{\partial t} + \rho c^2 \frac{\partial
+  v_0}{\partial z} &=&
+                            f_0  \nonumber\\
+\rho \frac{\partial v_0}{\partial t} +\frac{\partial p_0}{\partial z} &=& g_0 \nonumber\\
+\end{eqnarray}
+in which
+\begin{eqnarray}
+  \label{eqn:rhscut}
+  f_0(z,t) &=& -\rho c^2 v(z,t) \frac{\partial}{\partial z}(1-\phi(z))\nonumber\\
+  g_0(z,t) &=& -p(z,t) \frac{\partial}{\partial z}(1-\phi(z))
+\end{eqnarray}
+vanish near $z=z_s$. If $w \in L^2(\bR)$ and vanishes for large
+negative $t$ (as it must, for the system \ref{eqn:awe1dptsrc} to be
+compatible), then from expressions \ref{eqn:sol1dptsrc} the
+distributions $p,v$ are locally square-integrable in
+$\{z:|z-z_s| \ge \epsilon/2\} \times \bR$ and vanish for large
+negative $t$, whence the same is true of $f_0,g_0$.
+
+Assume for the moment that $w \in C^{\infty}_0(\bR)$, so that $p,v$
+are smooth away from $z=z_s$ and $p_0, v_0, f_0, g_0$ are smooth. Then
+$p_0$ is also the solution of the second-order initial value problem
+\begin{eqnarray}
+  \label{eqn:awe2ord}
+  \frac{1}{\rho c^2} \frac{\partial^2 p_0}{\partial t^2} -
+  \frac{1}{\rho}\frac{\partial^2 p_0}{\partial z^2} &=& F\nonumber \\
+  p &=& 0, t\ll 0
+\end{eqnarray}
+
+with the right-hand side $F$ given by
+\begin{equation}
+  \label{eqn:rhs2ord}
+  F = \frac{1}{\rho c^2}\frac{\partial f_0}{\partial t} -
+  \frac{1}{\rho}\frac{\partial g_0}{\partial z}
+\end{equation}
+Define $f$ by
+\begin{equation}
+  \label{eqn:defequivrhs}
+  f(z,t) = \rho c^2 \int_{-\infty}^t \,ds\,F(z,t) =  f_0(z,t) - c^2\int_{-\infty}^{t}\,ds\,\frac{\partial
+    g_0}{\partial z}(z,s)
+\end{equation}
+Then setting
+\begin{eqnarray*}
+  p_1 &=& p_0,\\
+  v_1 &=& \frac{1}{\rho}\int_{-\infty}^t\frac{\partial p_0}{\partial z}
+\end{eqnarray*}
+it follows from \ref{eqn:awe2ord} and \ref{eqn:defequivrhs} that
+$p_1,v_1$ solves \ref{eqn:awe1d} with $f$ as given above. Since
+$p_1=p_0$, and $p_0=p$ in a neighborhood of $z=z_r$, it follows that
+\begin{equation}
+  \label{eqn:impext}
+  S[c]f = S_p[w],
+\end{equation}
+that is, that using RHS $f$ in \ref{eqn:awe1d} produces the same
+pressure field near $z=z_r$ as does the point source in
+\ref{eqn:awe1dptsrc}. Also
+\[
+  f(z,t) = -\rho c^2 v(z,t) \frac{\partial}{\partial z}(1-\phi(z)) - c^2
+  \int_{-\infty}^t\,ds\, \frac{\partial}{\partial z}\left(-p(z,s) \frac{\partial}{\partial z}(1-\phi(z))\right)
+\]
+\[
+  = - c^2 \left(\int_{-\infty}^t\,ds\,\left( \rho \frac{\partial v}{\partial
+      t} -\frac{\partial p}{\partial
+      z}\right)(z,s)\frac{\partial}{\partial z}(1-\phi(z))
+  -p(z,s)\frac{\partial^2}{\partial z^2}(1-\phi(z)) \right)
+\]
+\[
+  = -2\rho c^2 v(z,t)\frac{\partial}{\partial z}(1-\phi(z)) +
+  c^2\frac{\partial^2}{\partial z^2}(1-\phi(z))\int_{-\infty}^t
+  \,ds\,p(z,s)
+\]
+using the second equation (momentum balance) in the system
+\ref{eqn:awe1dptsrc}. Use \ref{eqn:sol1dptsrc} to replace $p,v$ by
+explicit expressons in $w$:
+\[
+    = -\mbox{sgn}(z-z_s)w\left(t -\frac{|z-z_s|}{c}\right)\frac{\partial}{\partial z}(1-\phi(z)) +
+  \frac{\partial^2}{\partial z^2}(1-\phi(z))\frac{c}{2}\left(\int_{-\infty}^tw\right)\left(t - \frac{|z-z_s|}{c}\right)
+\]
+\begin{equation}
+  \label{eqn:defextop}
+  =E[c]w(z,t)
+\end{equation}
+whence the image of $w$ under $E[c]$ is square-integrable,
+$E[c]$ extends to a bounded operator $L^2(\bR) \rightarrow
+L^2([z_{\rm min},z_{\rm max}]\times \bR)$, and from equation \ref{eqn:impext}
+\begin{equation}
+  \label{eqn:extapp}
+S[c]\circ E[c] = S_p[c]
+\end{equation}
+as asserted in equation \ref{eqn:ext}.
 

### 2020-02-23 repro/papers/wri1d/paper.tex
near final

@@
 $c$ to be constant, that is, independent of position along the 1D
-continuum. I introduce a family of inverse problems, parametrized by a
+continuum. I introduce a family of inverse problems, depending on a
 parameter $\lambda$ playing the role of wavelength. For sufficiently small
@@
 does not safeguard it from failure to converge globally to a ``good''
-local minimum.
+local minimum. In fact, the $\alpha \rightarrow 0$ limit of the WRI
+objective is well-defined (after scaling by $1/alpha^2$ and also behaves in
+qualitatively the same way as the FWI objective. So the apparent
+ability to maintain better data fit via reduction of $\alpha$ does not
+lead to global behaviour asymptotically more amenable to local optimization.
 
@@
 residual, with a weight operator depending on the coefficient vector
-$c$. This result is quite general, applying to essentially any
+$c$. This identity has also been derived by \cite{Leeuwen2019note},
+using a different argument. The result is quite general, applying to essentially any
 realization of WRI, and already shows that its behaviour must be
@@
 That is, the support of f will be assumed to lie in the strip $[z_{\rm min},z_{\rm max}] \times \bR$.
-As it will turn out, the choice of spatial interval $[z_{\rm min},z_{\rm max}]$ is arbitrary, so long as it has positive length and contains $z_s$ in its interior.
+As it will turn out, the choice of spatial interval $[z_{\rm min},z_{\rm max}]$ is arbitrary, so long as it has positive length.
 
@@
 The version of FWI discussed here presumes that the source field
-corresponding to the data $d$ is supported at a point $z_s \ne z_r$,
-and is known. This field depends on a function of time (``wavelet'')
+corresponding to the data $d$ is supported at a point $z_s \in \bR$,
+and is known. The source position $z_s$ must satisfy $z_{\rm min} \le
+z_s \le z_{\rm max}$ and $z_s \ne z_r$, but is otherwise arbitrary.
+
+This point source field depends on a function of time (``wavelet'')
 $w \in L^2(\bR)$. Formally, the resulting acoustic field satisfies the
@@
 
-To end this section, it is necessary to address an irritating technical point: the point source defined above is not a member of the domain of $S[c]$, as it was defined in display \ref{eqn:defmoddom}, so the left-hand side of equation \ref{eqn:defmodpt} does not actually make sense.
-The fix for this incompatibility actually elucidates the relation between $S[c]$ and $S_p[c]$.
-Appendix B describes the construction of a family of bounded injective operators  $:L^2(\bR) \rightarrow L^2([z_{\rm min},z_{\rm max}] \times \bR)$ for which
+To end this section, it is necessary to address an irritating
+technical point: the point source defined above is not a member of the
+domain of $S[c]$, as it was defined in display \ref{eqn:defmoddom}, so
+the left-hand side of equation \ref{eqn:defmodpt} does not actually
+make sense. The fix for this incompatibility actually elucidates the
+relation between $S[c]$ and $S_p[c]$. Appendix B describes the
+construction of a family of bounded injective operators  $E[c]: L^2(\bR) \rightarrow L^2([z_{\rm min},z_{\rm max}] \times \bR)$ for which
 \begin{equation}
@@
 This relation exhibits $S[c]$ as an {\em extension} of $S_p[c]$ as
-described by \cite{Symes:09}, and identifies WRI as an extended
-inversion method.
+described by \cite{Symes:09}. WRI for 1D acoustic transmission is
+based on $S[c]$, and so is identified as an extended version of FWI.
 
@@
 
-\section{Global Asymptotics of FWI}
+\section{Global Asymptotics of 1D transmission FWI}
 The FWI objective is well-known to exhibit non-convexity unless
@@
 In the definition \ref{eqn:wfam}, the ``mother wavelet'' $w_1\in
-C_0^{\infty}(0,1)$, and the scaling is chosen so fhat
+C_0^{\infty}(0,1)$ has dimensionless argument, and the scaling is chosen so fhat
 $\|\wl\|_{L^2(\bR)}$ is independent of $\lambda>0$.
@@
 Choose a target wave velocity $c_* \in [c_{\rm min},c_{\rm
-  max}]$. Introduce a family of consisten data $\dl$, generated by
+  max}]$. Introduce a family of consistent data $\dl$, generated by
 $c_*$ and the wavelet family $\wl$:
@@
   =
-  \int_0^T\,dt\,\left|\frac{1}{2c_*}\wl\left(t-\frac{z_s-z_r}{c_*}\right)
-    - \frac{1}{2c}\wl\left(t-\frac{z_s-z_r}{c}\right)\right|^2
+  \int_0^T\,dt\,\left|\frac{1}{2c_*}\wl\left(t-\frac{|z_s-z_r|}{c_*}\right)
+    - \frac{1}{2c}\wl\left(t-\frac{|z_s-z_r|}{c}\right)\right|^2
 \end{equation}
@@
   \label{eqn:wlsupps}
-  \mbox{supp }S_p[c]\wl \subset \left[\frac{z_s-z_r}{c},\lambda
-    +\frac{z_s-z_r}{c}\right] \cap [0,T]
+  \mbox{supp }S_p[c]\wl \subset \left[\frac{|z_s-z_r|}{c},\lambda
+    +\frac{|z_s-z_r|}{c}\right] \cap [0,T]
 \end{equation}
@@
 \[
-  \lambda +\frac{z_s-z_r}{c} < T
+  \lambda +\frac{|z_s-z_r|}{c} < T
 \]
@@
   \|S_p[c]\wl\|^2 = \frac{1}{4c^2}\int_{-\infty}^{\infty}\,dt\,\left|
-    \wl\left(t-\frac{z_s-z_r}{c}\right)\right|^2 =
+    \wl\left(t-\frac{|z_s-z_r|}{c}\right)\right|^2 =
   \frac{\|w_1\|^2}{4c^2}.
@@
 \[
-  J_{\rm WRI}[c,d,q] =
+  J_{\rm WRI}^{\alpha}[c,d,q] =
   \mbox{min}_f \frac{1}{2}(\| d-S[c]f\|^2+\alpha^2\|f-q\|^2)
@@
 \label{eqn:defvpm}
-  J_{\rm WRI}[c,d,q] =\mbox{min}_g \frac{1}{2}(\| r[c]-S[c]g\|^2+\alpha^2\|g\|^2)
+  J_{\rm WRI}^{\alpha}[c,d,q] =\mbox{min}_g \frac{1}{2}(\| r[c]-S[c]g\|^2+\alpha^2\|g\|^2)
 \end{equation}
@@
 such sources contribute nothing to the data fit term in the definition
-of $J_{\rm WRI}$. If the domain and range of $S$ were finite
+of $J_{\rm WRI}^{\alpha}$. If the domain and range of $S$ were finite
 dimensional (which of course they are, after discretization), then the
@@
 \[
-  J_{\rm WRI}[c,d,q] =
+  J_{\rm WRI}^{\alpha}[c,d,q] =
   \mbox{min}_{e,n} \frac{1}{2}(\| d-S[c](S[c]^Te+q)\|^2+\alpha^2(\|S[c]^Te\|^2+\|n\|^2))
@@
   \label{eqn:defvpmred}
-   \mbox{min}_{e} \frac{1}{2}(\|r[c]-S[c]S[c]^Te\|^2+\alpha^2\|S[c]^Te\|^2).
+ =  \mbox{min}_{e} \frac{1}{2}(\|r[c]-S[c]S[c]^Te\|^2+\alpha^2\|S[c]^Te\|^2).
  \end{equation}
@@
 \[
-  J_{\rm WRI}[c,d,q] =
+  J_{\rm WRI}^{\alpha}[c,d,q] =
   \frac{1}{2}(\|r[c]-S[c]S[c]^Te_{\alpha}[c]\|^2+\alpha^2\|S[c]^Te_{\alpha}[c]\|^2)
@@
   \label{eqn:defwrialt}
-  J_{\rm WRI}[c,d,q] = \frac{1}{2}\langle r[c], W_{\alpha}[c] r[c]\rangle
+  J_{\rm WRI}^{\alpha}[c,d,q] = \frac{1}{2}\langle r[c], W_{\alpha}[c] r[c]\rangle
 \end{equation}
@@
 \end{equation}
-This remarkable identity shows that the WRI objective function is a {\em weighted norm of the data residual $r[c]$}.
+This remarkable identity shows that the WRI objective function is a
+{\em weighted norm of the data residual $r[c]$}.
 
-\section{WRI for 1D Acoustics}
+\cite{Leeuwen2019note} gives a different derivation of an identity
+equivalent to equations \ref{eqn:defwrialt}, \ref{eqn:defwriwt}.
+
+\section{Global Asymptotics of 1D transmission WRI}
 The preceding section provides the necessary ingredients for an assessment of the relation between WRI and FWI.
@@
   \label{eqn:wriscrewed}
-  J_{\rm WRI}[c,d,q]=u[c]J_{\rm FWI}[c,d,w]
+  J_{\rm WRI}^{\alpha}[c,d,q]=u[c]J_{\rm FWI}[c,d,w]
 \end{equation}
@@
   For $L$ as defined in \ref{eqn:defL}, and $\lambda <
-  \lambda_0$, the minimizer of $J_{\rm WRI}[c,\dl,\wl \delta(\cdot-z_s)]$ on the
+  \lambda_0$, the minimizer of $J_{\rm WRI}^{\alpha}[c,\dl,\wl \delta(\cdot-z_s)]$ on the
   complement of $[c_*-L\lambda, c_*+L\lambda]$ is
@@
   \[
-    J_{\rm WRI}[c,\dl,\wl\delta(\cdot-z_s)] =
+    J_{\rm WRI}^{\alpha}[c,\dl,\wl\delta(\cdot-z_s)] =
     \frac{\alpha^2}{2}\left(\frac{z_{\rm max}-z_{\rm
@@
 
-In other words, $J_{\rm WRI}$ has local minima far from the target
+In other words, $J_{\rm WRI}^{\alpha}$ has local minima far from the target
 velocity $c_*$, in the same way as does $J_{\rm FWI}$. One of the
@@
 
+Note that $L$ is independent of $\alpha$ (definition \ref{eqn:defL},
+and for small enough $\alpha$, $\beta > 0$ (definition
+\ref{eqn:defbeta}. Conclude that the region $|c-c_*| > L\lambda$ is
+independent of $L$, and the minimizer of $J_{\rm
+  WRI}[c,\dl,\wl\delta(\cdot-z_s)]$ is $c=c_{\rm min}$ for small
+enough $\alpha$. Therefore taking $\alpha$ small does not change the
+multimodal nature of $J_{\rm WRI}^{\alpha}$: there remain multiple far-apart
+local minima, no matter how small $\alpha$ may be.
+
 \section{Discussion}
@@
 transmission inverse problem, both $J_{\rm FWI}[c,d,w]$ and
-$J_{\rm WRI}[c,d,q]$ exhibit local minima (in $c$) far from the global
-minimum for consistent data, and the domain of attraction of the
-global minimizer can be arbitrarily small. These are striking
+$J_{\rm WRI}^{\alpha}[c,d,q]$ exhibit local minima (in $c$) far from the global
+minimum for consistent data, the domain of attraction of the
+global minimizer can be arbitrarily small, and these properties
+persist as $\alpha \rightarrow 0$. These are striking
 conclusions, but the 1D acoustic transmission problem is very special
@@
 \[
-  A[c]u(t) = t^2u(t).
+  A[c]u(t) = tu(t).
 \]
@@
 an extensive discussion of other similar source extension methods for
-various wave inversion problems, and to references cited there for
+various wave inversion problems, and for
 references to earlier work on this topic.
@@
 \[
-  f(z,t) = \int dz_1\,f(z_s,t)\delta(z-z_1)
+  f(z,t) = \int dz_1\,f(z_1,t)\delta(z-z_1)
 \]
@@
 
-One step in this construction involve buidling a constitutive defect
-(pressure) source that is equivalent, in the sense of generating the
-same solution, to a force (velocity) source. This construction is
+One step in this construction involves building a constitutive defect
+(pressure) source that is equivalent to a force (velocity) source, in the sense of generating the
+same solution outside of the source support. This construction is
 presented here in the context of the 1D acoustic problem, but is a

### 2020-02-23 repro/papers/wri1d/paper.tex
near final

@@
 
-Note that $L$ is independent of $\alpha$ (definition \ref{eqn:defL},
+Note that $L$ is independent of $\alpha$ (definition \ref{eqn:defL}),
 and for small enough $\alpha$, $\beta > 0$ (definition
-\ref{eqn:defbeta}. Conclude that the region $|c-c_*| > L\lambda$ is
-independent of $L$, and the minimizer of $J_{\rm
-  WRI}[c,\dl,\wl\delta(\cdot-z_s)]$ is $c=c_{\rm min}$ for small
-enough $\alpha$. Therefore taking $\alpha$ small does not change the
-multimodal nature of $J_{\rm WRI}^{\alpha}$: there remain multiple far-apart
-local minima, no matter how small $\alpha$ may be.
+\ref{eqn:defbeta}). Conclude that the region $\{c \in [c_{\rm min},
+c_{\rm max}]: |c-c_*| > L\lambda$ is
+independent of $\alpha$, and the minimizer of
+$J_{\rm WRI}[c,\dl,\wl\delta(\cdot-z_s)]$ in this region (away from
+$c_*$) is $c=c_{\rm min}$ for small enough $\alpha$. Therefore taking
+$\alpha$ small does not change the multimodal nature of
+$J_{\rm WRI}^{\alpha}$: there remain multiple far-apart local minima,
+no matter how small $\alpha$ may be.
 

### 2020-03-04 repro/papers/dso1d/paper.tex
more

@@
-\title{Why it works}
+\title{Full Waveform Inversion by Source Extension: Why it works}
 \author{William. W. Symes \thanks{The Rice Inversion Project,
@@
 \section{Introduction}
-Many of the extension concepts introduced to combat cycle-skipping in FWI sound good, and appear to work at least to some extent as one might hope from their heuristic justifications.
-However very few of these approaches have been assured of success via mathematical argument: in essence, they are mostly justified only ``in the rear-view mirror'', with no assurance that failure is not just around the corner, at the next example.
-On top of that, some of these approaches, for example those based on the computationally attractive Variable Projection Method (``VPM'') of \cite{GolubPereyra:03} have attributes such as gradient formulae cast in such form that the reasons for success are not readily apparent.
+Full Waveform Inversion (FWI) is now well-established as
+a useful tool for probing the earth's subsurface
+\cite[]{VirieuxOperto:09,Fichtner:10}, but its use is still
+impeded by so-called ``cycle-skipping''. Because the computational
+size of field inversion tasks is very large, only iterative local (descent)
+minimization of the data misfit function is computationally
+feasible. However such local descent methods tend to stagnate as
+suboptimal and geologically uninformative earth models, unless initial
+models are already quite close to optimal, in the sense of predicting
+the arrival times of seismic events to within a small multiple of a
+dominant wavelenght \cite[]{GauTarVir:86,Plessix:10}.
 
-This note shows exactly how VPM leads to successful velocity updates for a particular extension of a very simple inverse problem, which asks that a homogeneous velocity model be deduced from one trace at known offset.
-I put this computation forward not because there are not simpler ways of answering the question it poses - there certainly are - but because the formal ingredients of waveform-based velocity estimation in this very simple setting are common to many similar extended inversion algorithms, and because in this case every computation can be done analytically, nearly to completion.
-In particular, it becomes clear why the VPM gradient formula produces a constructive update, despite not incorporating explicitly the extension measure (``annihilator'') central to the approach.
+This paper concerns one of the many ideas that have been advanced to
+overcome cycle-skipping, namely so-called extended inversion
+\cite[]{geoprosp:2008}. ``Extended'' signifies that addional degrees
+of freedom are provided to the modeling process, in the hope of
+opening up more effective routes to geologically informative models
+with acceptable data fit. Since these extended degrees of freedom are
+not part of the basic physics chosen to model the data acquisition
+process, they should be suppressed in the eventual solution. Extended
+inversion methods differ by the choice of additional degrees of
+freedom, and by choice of penalty applied to eliminate them in the final
+result.
 
-\section{Theory}
-Suppose that a single trace is recorded, at distance $r$ from a source, and both source and receiver locations far enough from physical boundaries that the recorded wave has effectively propagated in free space.
+Many of these extension concepts sound
+plausible, and appear to work at least to some extent as one might
+hope from their heuristic justifications. However very few of these
+approaches have been underwritten by mathematical argument: in
+essence, they are mostly justified only ``in the rear-view mirror'',
+with no assurance that failure is not just around the corner, at the
+next example. On top of that, some of these approaches, for example
+those based on the computationally attractive Variable Projection
+Method (``VPM'') of \cite{GolubPereyra:03}, are cast in such form that the reasons for success are
+not readily apparent.
 
-Assume acoustic propagation, constant density, and isotropic point source and receiver.
-With a single trace, at best the average velocity in a region between source and receiver can be estimated, so assume that the velocity is constant.
-Then the modeling operator depends on $v$ and maps the source pulse $f$ to the predicted trace $S[v]f$ by
+This note shows exactly how VPM leads to successful velocity updates
+for a particular extended inversion approach to a very simple inverse
+problem, which asks that a homogeneous velocity field be deduced from
+one trace at known offset. I put forward this inverse problem and
+extension-based solution not because there are not simpler
+ways of answering the question it poses - there certainly are - but
+because the formal ingredients of waveform-based velocity estimation
+in this very simple setting are common to many similar extended
+inversion algorithms, and because in this case every computation can
+be done analytically, nearly to completion. In particular, it becomes
+clear why the VPM gradient formula produces a constructive update,
+with no possibility of stagnation at local minimizers.
+
+The extended inversion approach
+developed here uses a {\em source extension}, in which source
+parameters form the additional degrees of freedom. This type of
+extension presumes that the actual or target source is constrained in
+some way; the extended source is allowed to violate the
+constraint. For the problem considered here, the source model reduces
+to a wavelet, and the target wavelet is assumed to be non-zero only in
+a short time
+interval (an approximate impulse). The extension consists in permitting energy to spread
+in time at intermediate iterations of the inversion. A simple penalty
+for energy spread (second moment of square amplitude) drives the
+extended source towards a focused source satisfying the assumed
+constraint. Not all penalties are created equal: the penalty used here
+has a critical {\em (pseudo-)differential} attribute necessary for avoidance of cycle-skipping.
+
+%despite not incorporating explicitly the extension measure
+%(``annihilator'') central to the approach.
+
+I begin with a quick sketch of constant density acoustics,
+and describe the single-trace transmission inverse problem. To make
+the role of data frequency content clear, I introduce a family of
+noise-free data parametrized by wavelength. For completeness, I show how
+the standard FWI approach to this problem generates multiple local
+minima that will be found by any descent method unless the initial
+estimate predicts travel time from source to receiver with an error
+on the order of a wavelength. The next section describes the source
+extension objective, and the reduced objective
+produced by VPM. As VPM eliminates the extended source, this
+function depends only on the velocity, just as does the FWI objective.  A
+nearly-explicit calculation of the VPM gradient 
+shows that the only stationary points are ``within a wavelength'' of
+the correct velocity, used to build the data: that is, cycle-skipping
+cannot occur. The paper ends with a discussion of the parallels
+between the calculations presented here and the structure of other
+extended inversion methods applicable to field-scale velocity
+estimation, and the critical role that the differential nature of the
+extension penalty plays in the success of this and other extension methods.
+
+\section{Preliminaries}
+Assume small amplitude (linearized) acoustic propagation, constant
+density, and isotropic point source and receiver. Denote by $m(\bx)$
+the slowness (reciprocal velocity) at spatial position $\bx$, $f(t)$
+the time dependence of the point source (``wavelet'') at location
+$\bx=\bx_s$. Then the (excess) pressure field $p(\bx,t)$ obeys a
+scalar wave equation:
+\begin{eqnarray}
+  \label{eqn:awe}
+  \left(m(\bx)^2\frac{\partial^2 p}{\partial t^2} - \nabla^2\right) p(\bx,t) &=&
+                                                                         f(t)\delta(\bx-\bx_s) \nonumber\\
+  p(\bx,t)&=&0, t\ll 0
+\end{eqnarray}
+Suppose that a single trace is recorded, at distance $r>0$ from the
+source position $\bx_s$. The dominant information in a single
+trace is the transient signal time of arrival, constraining only the mean slowness in the
+region
+between source and receiver, so assume that the
+slowness is constant, that is, independent of position $\bx$. The pressure field is simply the  the source
+wavelet $f(t)$ convolved with the
+acoustic Green's function, for which an analytic expression is
+available in the constant $m$ case \cite[]{CourHil:62}:
+\begin{equation}
+  \label{eqn:homsol}
+  p(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}f\left(t-m|\bx-\bx_s|\right).
+\end{equation}
+
+The receiver location $\bx_r$ lies at distance $r$ from the source
+location $\bx_s$, that is, $|\bx_r-\bx_s|=r$. The predicted signal at
+$p(\bx_r,t)$ depends nonlinearly on the slowness $m$ and linearly on the
+source wavelet $f$. Therefore it is naturally represented as the
+action of a $m$-dependent linear operator $S[m]$ on $f$:
 \begin{equation}
 \label{eqn:mod}
-S[v]f (t)= \frac{1}{4\pi r}f\left(t-\frac{r}{v}\right).
+S[m]f (t)= p(\bx_r,t) = \frac{1}{4\pi r}f\left(t-mr\right).
 \end{equation}
-Ignoring amplitude, this map implements a $v$-dependent time shift, which is everyone's favorite example to show why FWI cycle-skips and get stuck.
 
-An obvious extended model for this example is $\{v,f\}$, that is, pairs of velocity and source wavelet, with the extended modeling operator already defined by equation \ref{eqn:mod}.
+Ignoring amplitude, this map implements a $m$-dependent time
+shift. This time shift operator is the basis of many descriptions of
+the cycle-skipping phenomenon (for example, \cite{VirieuxOperto:09},
+Figure 7), so it is unsurprising that an analysis of cycle-skipping
+can be based on the simple modeling operator described above, which
+amounts essentially to a time shift. To make the link with wavelet
+frequency content manifest, I introduce a family $\{\fl\}$ of
+wavelets parametrized by a nondimensional wavelength $\lambda$,
+\begin{equation}
+  \label{eqn:deffl}
+  \fl(t) = \frac{1}{\lambda}f_1\left(\frac{t}{\lambda}\right).
+\end{equation}
+The ``mother wavelet'' $f_1$ vanishes for $t>1$, and has positive
+mean-square. No other constraints need be placed on $f_1$. Note that
+the scaling is such that the mean-square
+\[
+  \|\fl\|^2 = \int\,dt\,|\fl(t)|^2
+\]
+is independent of $\lambda$.
+
+Note that in relations such as ``$t>1$'' a time unit is implicit in
+the right-hand side.
+
+To this family of wavelets and a choice of target slowness $m_*$  corresponds a family of noise-free data
+\begin{equation}
+  \label{eqn:defdata}
+  \dl=S[m_*]\fl.
+\end{equation}
+This family of data in turn defines a family of inverse problems, to
+which I now turn.
+
+\section{Full Waveform Inversion}
+The preceding section provided all of the raw ingredients to define
+full waveform inversion for estimation of $m$ from a single trace.
+It is only $m$ that is to be determined: the $\lambda-$dependent family of wavelets
+$\{\fl\}$ is regarded as known, along with the data family $\{\dl\}$. The aim
+is to chose $m$ to minimize 
+\begin{equation}
+  \label{eqn:FWIfix}
+  J_{\rm fix}[m] = \frac{1}{2}\|S[m]\fl-\dl\|^2.
+\end{equation}
+for all values of $\lambda > 0$.
+
+Written out in detail, this objective function is
+\[
+ J_{\rm fix}[m] =  \frac{1}{32\pi^2
+    r^2}\int\,dt\,\left|\fl\left(t-mr\right)-\fl\left(t-m_*r\right)\right|^2
+\]
+Since $f_1$ vanishes for $|t|>1$, $\fl$ vanishes for $|t|>\lambda$,
+and $S[m]\fl$ vanishes if $|t-mr|>\lambda$. So if $|mr-m_*r|
+= |m-m_*|r > 2\lambda$, then $|t-mr|+|t-m_*r| \ge |mr-m_*r| >
+2\lambda$ so either $|t-mr|>\lambda$ or $|t-m_*r|>\lambda$, that is,
+either $S[m]\fl(t)=0$ or $S[m_*]\fl(t)=0$. Therefore $S[m]\fl$ and
+$S[m_*]\fl$ are orthogonal in the sense of the $L^2$ inner product:
+\begin{equation}
+  \label{eqn:ortho}
+  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \langle S[m]\fl,
+  S[m_*]\fl\rangle = \int\,dt\,S[m]\fl(t)S[m_*]\fl(t) = 0
+\end{equation}
+But $\dl = S[m_*]\fl$, so this is the same as saying that $\dl$ is
+orthogonal to $S[m]\fl$. So conclude that
+\begin{equation}
+  \label{eqn:iso}
+  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, J_{\rm fix}[m]=\frac{1}{16\pi^2
+    r^2}\|f_1\|^2.
+\end{equation}
+using the previously observed independence of $\|\fl\|$ from
+$\lambda$.
+
+That is, for slowness $m$ in error by more than $\lambda/r$ from the
+target slowness $m_*$, the FWI objective \ref{eqn:FWIfix} is perfectly
+flat: all nearby values of $m$ are local minima. Therefore the local
+exploration of the objective
+gives no useful information whatever about constructive search
+directions towards the global minimizer $m=m_*$, for which the
+objective value is of course = 0, if the initial estimate of $m$ in
+error by an amount greater than a fixed multiple of $\lambda$. This is
+precisely the behaviour of FWI noted many times in the literature: the
+model must be known to ``within a wavelength''.
+
+\section{Source Extension Inversion}
+An obvious extended model for this example is $\{m,f\}$, that is, pairs of velocity and source wavelet, with the extended modeling operator already defined by equation \ref{eqn:mod}.
 With an annihilator $A$, to be chosen below, the penalty form of extended inversion is: minimize over $\{v,f\}$
@@
 \label{eqn:obj}
-J[v,f] = \frac{1}{2}(\|S[v]f-d\|^2 + \alpha^2 \|Af\|^2).
+J[v,f] = \frac{1}{2}(\|S[m]f-d\|^2 + \alpha^2 \|Af\|^2).
 \end{equation}
@@
 where $f[v]$ is the minimizer of $J$ over $f$ for given $v$.
-For this example, $J_{\rm red}$ is simple to compute. First observe that apart from amplitude, $S[v]$ is unitary:
+For this example, $J_{\rm red}$ is simple to compute. First observe that apart from amplitude, $S[m]$ is unitary:
 \begin{equation}
 \label{eqn:tran}
-S[v]^T g (t) = \frac{1}{4\pi r}g\left(t+\frac{r}{v}\right)
+S[m]^T g (t) = \frac{1}{4\pi r}g\left(t+\frac{r}{v}\right)
 \end{equation}
@@
 \label{eqn:unit}
-S[v]^TS[v] f (t) = \frac{1}{(4\pi r)^2} f(t).
+S[m]^TS[m] f (t) = \frac{1}{(4\pi r)^2} f(t).
 \end{equation}
@@
 \label{eqn:norm1}
-\left(\frac{1}{(4\pi r)^2} I + \alpha^2 A^TA\right)f = S[v]^Td.
+\left(\frac{1}{(4\pi r)^2} I + \alpha^2 A^TA\right)f = S[m]^Td.
 \end{equation}
@@
 \label{eqn:resid}
-(S[v]f[v]-d)(t) =\frac{1}{4\pi r}[ (1+(4\pi r)^2 \alpha^2 (t-r/v)^2)^{-1} - 1]f_*(t-r/v_*)
+(S[m]f[v]-d)(t) =\frac{1}{4\pi r}[ (1+(4\pi r)^2 \alpha^2 (t-r/v)^2)^{-1} - 1]f_*(t-r/v_*)
 \end{equation}
@@
 \label{eqn:grad}
-\nabla J[v] = (DS[v]f[v])^*(S[v]f[v]-d)
+\nabla J[v] = (DS[m]f[v])^*(S[m]f[v]-d)
 \end{equation}.
-In this formula, $DS[v]$ is the derivative with respect to $v$ - since we can make explicit use of the Green's function, easy to see that
+In this formula, $DS[m]$ is the derivative with respect to $v$ - since we can make explicit use of the Green's function, easy to see that
 \begin{equation}
 \label{eqn:deriv}
-(DS[v]\delta v)f (t) = \frac{\delta v}{4\pi v^2}\frac{df}{dt}(t-r/v) = S[v](Q[v]\delta v)f (t),
+(DS[m]\delta v)f (t) = \frac{\delta v}{4\pi v^2}\frac{df}{dt}(t-r/v) = S[m](Q[v]\delta v)f (t),
 \end{equation}
@@
 That is, $Q[v]\delta v$ is a skew-adjoint operator depending linearly on $\delta v$ - more on this below.
-The adjoint $(DS[v]f)^*$ is the adjoint of the map from model space to data space:
+The adjoint $(DS[m]f)^*$ is the adjoint of the map from model space to data space:
 \[
-(DS[v]f): \delta v \mapsto (DS[v]\delta v)f
+(DS[m]f): \delta v \mapsto (DS[m]\delta v)f
 \]
-and is NOT the same as the adjoint denoted by $S[v]^T$, which is the data-space-to-data-space adjoint (got that?).
+and is NOT the same as the adjoint denoted by $S[m]^T$, which is the data-space-to-data-space adjoint (got that?).
 That is, this adjoint make this relation true, for all $\delta v,f,$ and $d$:
@@
 \label{eqn:vadj}
-\delta v \cdot (DS[v]f)^*d = \int \,dt\,[(DS[v]\delta v)f] (t) d(t) 
+\delta v \cdot (DS[m]f)^*d = \int \,dt\,[(DS[m]\delta v)f] (t) d(t) 
 \end{equation}
@@
 \[
-\delta v \cdot \nabla J[v] =  \int \,dt\,[(DS[v]\delta v)f[v]] (t) (S[v]f[v]-d)(t)
+\delta v \cdot \nabla J[v] =  \int \,dt\,[(DS[m]\delta v)f[v]] (t) (S[m]f[v]-d)(t)
 \]
 \[
-= \int \,dt\, [S[v](Q[v]\delta v)f[v](t)] (S[v]f[v]-d)(t) = \int \,dt\,(Q[v]\delta v)f[v](t) S[v]^T(S[v]f[v]-d)(t)
+= \int \,dt\, [S[m](Q[v]\delta v)f[v](t)] (S[m]f[v]-d)(t) = \int
+\,dt\,(Q[v]\delta v)f[v](t) S[m]^T(S[m]  f[v]-d)(t)
 \]

### 2020-03-28 repro/papers/wri1d/paper.tex
near final - need DSO ref

@@
 local minimum. In fact, the $\alpha \rightarrow 0$ limit of the WRI
-objective is well-defined (after scaling by $1/alpha^2$ and also behaves in
+objective is well-defined (after scaling by $1/\alpha^2$) and also behaves in
 qualitatively the same way as the FWI objective. So the apparent
@@
   \label{eqn:fwiscrewed}
-  J_{\rm FWI,\lambda}[c] =
+   J_{\rm FWI}[c,\dl,\wl]  =
   \frac{1}{2}\left(\frac{1}{4c^2}+\frac{1}{4c_*^2}\right)\|w_1\|^2.
@@
 member of the domain of the simulation operator $S[c]$, as it is not
-square-integrable. The object of this appendix is to construct a square-integral right-hand side in the
+square-integrable. The object of this appendix is to construct a square-integrable right-hand side in the
 system \ref{eqn:awe1d} for which the pressure field $p$

### 2020-03-28 repro/papers/dso1d/paper.tex
near final - need WRI ref

@@
 \begin{abstract}
-An extremely simple example shows how extended modeling plus variable projection plus a suitable annihilator lead to constructive velocity updates.
+  An extremely simple single-trace transmission example shows how an
+  extended source formulation of full waveform inversion can produce
+  an optimization problem without spurious local minima (``cycle
+  skipping''). The data consist of a single trace recorded at a given
+  distance from a point source. The velocity or slowness is presumed
+  homogeneous, and the target source wavelet is presumed
+  quasi-impulsive or focused at zero time lag. The source is extended
+  by permitting energy to spread in time, and the spread is controlled
+  by adding a weighted mean square of the extended source wavelet to
+  the data misfit, to produce the extended inversion objective. The
+  objective function and its gradient can be computed explicitly, and
+  it is easily seen that all local minimizers must be within a
+  wavelength of the correct slowness. The derivation shows several
+  important features of all similar extended source algorithms. For
+  example, nested optimization, with the source estimation in the
+  inner optimization (variable projection method), is essential. The
+  choice of the weight operator, controlling the extended source
+  degrees of freedom, is critical: the choice presented here is a
+  differential operator, and that property is crucial for production
+  of an objective immune from cycle-skipping.
 \end{abstract}
@@
 \section{Introduction}
-Full Waveform Inversion (FWI) is now well-established as
-a useful tool for probing the earth's subsurface
-\cite[]{VirieuxOperto:09,Fichtner:10}, but its use is still
-impeded by so-called ``cycle-skipping''. Because the computational
-size of field inversion tasks is very large, only iterative local (descent)
-minimization of the data misfit function is computationally
-feasible. However such local descent methods tend to stagnate as
-suboptimal and geologically uninformative earth models, unless initial
-models are already quite close to optimal, in the sense of predicting
-the arrival times of seismic events to within a small multiple of a
-dominant wavelenght \cite[]{GauTarVir:86,Plessix:10}.
+Full Waveform Inversion (FWI), or estimation of earth structure by
+model-driven least squares data fitting, is now well-established as a
+useful tool for probing the earth's subsurface
+\cite[]{VirieuxOperto:09,Fichtner:10}. However, so-called ``cycle-skipping'', the tendency of iterative FWI
+algorithms to stagnate at suboptimal and geologically uninformative
+earth models, still impedes its use. Because the computational size of field inversion tasks
+is very large, only iterative local (descent) minimization of the data
+misfit function is computationally feasible. However local
+descent methods avoid suboptimal stagnation only if initial models are
+already quite close to optimal, in the sense of predicting the arrival
+times of seismic events to within a small multiple of a dominant
+wavelength \cite[]{GauTarVir:86,Plessix:10}.
 
-This paper concerns one of the many ideas that have been advanced to
+This paper concerns one of the several ideas that have been advanced to
 overcome cycle-skipping, namely so-called extended inversion
@@
 
-Many of these extension concepts sound
-plausible, and appear to work at least to some extent as one might
-hope from their heuristic justifications. However very few of these
-approaches have been underwritten by mathematical argument: in
-essence, they are mostly justified only ``in the rear-view mirror'',
-with no assurance that failure is not just around the corner, at the
-next example. On top of that, some of these approaches, for example
-those based on the computationally attractive Variable Projection
-Method (``VPM'') of \cite{GolubPereyra:03}, are cast in such form that the reasons for success are
-not readily apparent.
+Many of these extended inversion concepts sound plausible, and appear
+to work at least to some extent as one might hope from their heuristic
+justifications. However very few of these approaches have been
+underwritten by mathematical argument: in essence, they are mostly
+justified only ``in the rear-view mirror'', with no assurance that
+failure is not just around the corner, at the next example. On top of
+that, some of these approaches, for example those based on the
+computationally attractive Variable Projection Method (``VPM'') of
+\cite{GolubPereyra:03}, are cast in such form that the reasons for
+success are not readily apparent.
 
@@
 clear why the VPM gradient formula produces a constructive update,
-with no possibility of stagnation at local minimizers.
+with no possibility of stagnation away from the global minimum.
 
@@
 some way; the extended source is allowed to violate the
-constraint. For the problem considered here, the source model reduces
+constraint. For recent overview of source extension methods, see
+\cite{HuangNammourSymesDollizal:SEG19}. Source
+extension methods have computational complexity approximately the same
+as that of FWI, a signal advantage over the alterantive medium
+extension class described for instance in \cite[]{geoprosp:2008}.
+
+For the problem considered here, the source model amounts 
 to a wavelet, and the target wavelet is assumed to be non-zero only in
 a short time
-interval (an approximate impulse). The extension consists in permitting energy to spread
+interval (an approximate impulse, perhaps as the result of signature deconvolution). The extension consists in permitting energy to spread
 in time at intermediate iterations of the inversion. A simple penalty
 for energy spread (second moment of square amplitude) drives the
-extended source towards a focused source satisfying the assumed
+extended source towards a focused source approximately satisfying the assumed
 constraint. Not all penalties are created equal: the penalty used here
-has a critical {\em (pseudo-)differential} attribute necessary for avoidance of cycle-skipping.
+has the critical {\em (pseudo-)differential} attribute necessary for
+avoidance of cycle-skipping, as will be explained in the Discussion section.
 
@@
 frequency content manifest, I introduce a family $\{\fl\}$ of
-wavelets parametrized by a nondimensional wavelength $\lambda$,
+wavelets indexed by $\lambda$, a parameter having dimensions of time,
 \begin{equation}
@@
 \end{equation}
-The ``mother wavelet'' $f_1$ vanishes for $t>1$, and has positive
-mean-square. No other constraints need be placed on $f_1$. Note that
+The argument $s$ of the ``mother wavelet'' $f_1$ is
+nondimensional. The only constraints placed on $f_1$ are that (i) 
+$f_1(s)=0$ for $|s|\ge 1$, and (ii) $f_1$  has positive
+mean-square, that is, does not vanish identically. Note that
 the scaling is such that the mean-square
@@
 
-Note that in relations such as ``$t>1$'' a time unit is implicit in
-the right-hand side.
+I shall refer to $\lambda$ as ``wavelength'': if $f_1$ has a dominant
+period of oscillation, then so does $\fl$, and it is proportional to
+$\lambda$.
 
@@
   \label{eqn:FWIfix}
-  J_{\rm fix}[m] = \frac{1}{2}\|S[m]\fl-\dl\|^2.
+  J_{\rm FWI}[m] = \frac{1}{2}\|S[m]\fl-\dl\|^2.
 \end{equation}
@@
 \[
- J_{\rm fix}[m] =  \frac{1}{32\pi^2
+ J_{\rm FWI}[m] =  \frac{1}{32\pi^2
     r^2}\int\,dt\,\left|\fl\left(t-mr\right)-\fl\left(t-m_*r\right)\right|^2
@@
   \label{eqn:iso}
-  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, J_{\rm fix}[m]=\frac{1}{16\pi^2
+  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, J_{\rm FWI}[m]=\frac{1}{16\pi^2
     r^2}\|f_1\|^2.
@@
 
-That is, for slowness $m$ in error by more than $\lambda/r$ from the
-target slowness $m_*$, the FWI objective \ref{eqn:FWIfix} is perfectly
+That is, for slowness $m$ in error by more than $2\lambda/r$ from the
+target slowness $m_*$, the FWI objective $J_{\rm FWI}$ is perfectly
 flat: all nearby values of $m$ are local minima. Therefore the local
@@
 
-\section{Source Extension Inversion}
-An obvious extended model for this example is $\{m,f\}$, that is, pairs of velocity and source wavelet, with the extended modeling operator already defined by equation \ref{eqn:mod}.
-With an annihilator $A$, to be chosen below, the penalty form of extended inversion is: minimize over $\{v,f\}$
+\section{Extended Source Inversion}
+As mentioned in the introduction, the modeling operator introduced in
+the last section, $m \mapsto S[m]\fl$, may be extended simply by
+including the source wavelet as one of the model parameters: that is,
+the model vector becomes $(m,f)$, and the modeling operator, $(m,f)
+\mapsto S[m]f$.
+
+The reader will have no trouble seeing that the data misfit using this
+extended modeling operator can always be made to vanish entirely by
+proper choice of wavelet $f$, unless $f$ must satisfy some additional
+constraints. \cite{HuangNammourSymesDollizal:SEG19} describe a
+plethora of possible constraints for this and similar source
+extensions. Many (but not all) take the form of a quadratic penalty,
+that is, the mean square of $Af$, $A$ being a suitable operator,
+commonly dubbed an {\em annihilator}: in many examples the ideal
+output for $A$ applied to a source obeying the target constraints
+is the zero vector \cite[]{geoprosp:2008}. With an annihilator $A$, to
+be chosen below, the penalty form of extended inversion is: minimize
+over $\{m,f\}$
 \begin{equation}
 \label{eqn:obj}
-J[v,f] = \frac{1}{2}(\|S[m]f-d\|^2 + \alpha^2 \|Af\|^2).
+J_{\rm ESI}[m,f] = \frac{1}{2}(\|S[m]f-\dl\|^2 + \alpha^2 \|Af\|^2).
 \end{equation}
-VPM \cite[]{GolubPereyra:03} takes advantage of $J$ being quadratic in $f$ to solve for $f$ given $v$, thus producing a reducted objective of $v$ alone:
+
+The choice of penalty weight $\alpha$ has a profound influence on the
+character of this optimization problem. For the moment, I will mandate
+only that $\alpha > 0$. A detailed discussion of methods for setting
+$\alpha$ is beyond the scope of this brief paper.  In the Discussion
+section, I mention an effective method for choosing $\alpha$ to
+optimize convergence of iterative minimization, that applies (and has
+been applied) to much larger-scale inversion problems with features
+similar to those of this simple example.
+
+While minimization of $J_{\rm ESI}$ might be tackled directly - by
+alternately minimizations between $m$ and $f$, or by computing updates
+for $m$ and $f$ simultaneously - such joint mimization performs
+poorly, as \cite{YinHuang:16} has shown. The reason for this poor
+performance is that $J_{\rm ESI}$ has dramatically different
+sensitivity to $m$ versus $f$, especially for high frequency $f$, as
+the reader will see below. Instead, a nested approach, in which $f$ is
+eliminated in an inner optimization,
+generally gives far better numerical performance.  This Variable
+Projection Method (VPM) \cite[]{GolubPereyra:03} takes advantage of
+$J_{\rm ESI}$ being quadratic in
+$f$ to solve for $f$ given $m$, thus producing a reduced objective of
+$m$ alone:
 \begin{equation}
 \label{eqn:red}
-J_{\rm red}[v] = \min_f J[v,f] = J[v,f[v]],
+J_{\rm VPM}[m] = \min_f J_{\rm ESI}[m,f] = J_{\rm ESI}[m,f[m]],
 \end{equation}
-where $f[v]$ is the minimizer of $J$ over $f$ for given $v$.
-For this example, $J_{\rm red}$ is simple to compute. First observe that apart from amplitude, $S[m]$ is unitary:
+where $f[m]$ is the minimizer of $J_{\rm ESI}$ over $f$ for given $m$.
+For the problem considered here, $J_{\rm VPM}$ is explicitly computable. First observe that apart from amplitude, $S[m]$ is unitary:
 \begin{equation}
 \label{eqn:tran}
-S[m]^T g (t) = \frac{1}{4\pi r}g\left(t+\frac{r}{v}\right)
+S[m]^T g (t) = \frac{1}{4\pi r}g\left(t+mr\right)
 \end{equation}
@@
 \label{eqn:norm1}
-\left(\frac{1}{(4\pi r)^2} I + \alpha^2 A^TA\right)f = S[m]^Td.
+\left(\frac{1}{(4\pi r)^2} I + \alpha^2 A^TA\right)f[m] = S[m]^T\dl.
 \end{equation}
 
-At this point I have to come clean about the actual choice of $A$.
-As usual with source extensions, the proper annihilator depends on modeling assumptions, not fundamental physics.
-Assume that signature decon has been applied and the target source $f_*$ is a zero-phase bandpass filter, reasonably well focused at time $t=0$.
-Since there is no spatial structure to exploit in this hyper-simple problem, we construct $A$ to penalize energy away from $t=0$:
+At this point I have to come clean about the actual choice of
+$A$. Recall that $A$ is called an ``annihilator'', as it vanishes when
+applied to the target source field, at least in an idealized
+limit. Thus the choice of $A$ depends on modeling assumptions, not
+fundamental physics, as is characteristic of extended source
+inversion. As $\lambda \rightarrow 0$, the target source wavelet $\fl$
+focuses at time $t=0$, in the sense that it vanishes for
+$|t| > \lambda$. Therefore I choose $A$ to penalize energy away from
+$t=0$:
 \begin{equation}
@@
 \end{equation}
-Then the normal operator on the LHS of \ref{eqn:norm1} is simply multiplication by a positive function of time, and can be inverted by inspection.
+This
+particular annihilator has been employed in earlier papers on extended
+source inversion
+\cite[]{Plessix:00a,LuoSava:11,Warner:14,HuangSymes:SEG15a,Warner:16,HuangSymes:GEO17}.
+
+With this choice of $A$, the normal operator on the LHS of \ref{eqn:norm1} is simply multiplication by a positive function of time, and can be inverted by inspection.
 Using the identity \ref{eqn:tran} for the adjoint operator, obtain
-\begin{equation}
-\label{eqn:soln} 
-f[v](t) = \left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2\right)^{-1}\frac{1}{4\pi r}d\left(t+\frac{r}{v}\right)  
-\end{equation} 
-Suppose that the data is noise-free: 
-\begin{equation}
-\label{eqn:nonoise}
-d(t) = S[v_*]f_*(t) = \frac{1}{4\pi r}f_*\left(t-\frac{r}{v_*}\right)
-\end{equation}
-Then
+\[
+f[m](t) = \left(\frac{1}{(4\pi r)^2} + \alpha^2
+  t^2\right)^{-1}\frac{1}{4\pi r}\dl\left(t+mr\right)
+\]
 \begin{equation}
 \label{eqn:solnnon}
-f[v](t) = \left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2\right)^{-1}\frac{1}{(4\pi r)^2}f_*\left(t-\frac{r}{v_*}+\frac{r}{v}\right) 
+= \left(1+ (4\pi r)^2\alpha^2 t^2\right)^{-1}\fl\left(t+(m-m_*)r\right) 
 \end{equation}
-so the residual is, after a little algebra,
+thanks to the definition of the data $\dl$ (equation \ref{eqn:defdata}).
+After a little algebra,
 \begin{equation}
 \label{eqn:resid}
-(S[m]f[v]-d)(t) =\frac{1}{4\pi r}[ (1+(4\pi r)^2 \alpha^2 (t-r/v)^2)^{-1} - 1]f_*(t-r/v_*)
+(S[m]f[m]-d)(t) =\frac{1}{4\pi r}[ (1+(4\pi r)^2 \alpha^2 (t-mr)^2)^{-1} - 1]\fl(t-m_*r)
 \end{equation}
-The gradient of $J_{\rm red}$ is given by the famous VPM formula (well-known to those who know it well):
+and
+\begin{equation}
+  \label{eqn:anniexp}
+  Af[m] = t \left(1+ (4\pi r)^2\alpha^2
+    t^2\right)^{-1}\fl\left(t+(m-m_*)r\right).
+\end{equation}
+So
+\[
+  J_{\rm VPM}[m] = \frac{1}{2}\int\,dt\, \left( \frac{1}{(4\pi r)^2}[
+    (1+(4\pi r)^2 \alpha^2 (t+(m_*-m)r)^2)^{-1} - 1]^2 +
+  \right.
+  \]
+  \[
+   \left. \alpha^2(t+(m_*-m)r)^2 \left(1+ (4\pi r)^2\alpha^2
+      (t+(m_*-m)r)^2\right)^{-2}\right)|\fl(t)|^2
+\]
+\begin{equation}
+\label{eqn:vpmexpl}
+=\frac{1}{2(4\pi r)^2}\int\,dt\, [1-(1+(4\pi r)^2 \alpha^2
+(t+(m_*-m)r)^2)^{-1}]|\fl(t)|^2
+\end{equation}
+
+The gradient of $J_{\rm VPM}$ can be extracted by elementary means
+from the identity \ref{eqn:vpmexpl}, but instead we will derive it in
+using some important features that this inverse problem shares with
+other inverse wave problems with more complex physics. To begin with, the gradient of a VPM
+objective of the form \ref{eqn:red} is given by the formula 
 \begin{equation}
 \label{eqn:grad}
-\nabla J[v] = (DS[m]f[v])^*(S[m]f[v]-d)
-\end{equation}.
-In this formula, $DS[m]$ is the derivative with respect to $v$ - since we can make explicit use of the Green's function, easy to see that
+\nabla J_{\rm VPM}[m] = (DS[m]f[m])^*(S[m]f[m]-d).
+\end{equation}
+This easily derived result is in some sense the main content of \cite[]{GolubPereyra:03}.  In this formula, $DS[m]f$ is the derivative
+of the modeling operator $S[m]f$
+with respect to $m$, that is, with $f$ held fixed. The adjoint $(DS[m]f)^*$ is the adjoint of the map from model space to data space:
+\[
+(DS[m]f): \delta m \mapsto (DS[m]\delta m)f
+\]
+and is NOT the same as the adjoint denoted by $S[m]^T$, which is the source-space-to-data-space adjoint.
+That is, this adjoint make this relation true, for all $m, \delta m,f,$ and $d$:
+\begin{equation}
+\label{eqn:vadj}
+\delta m \cdot (DS[m]f)^*d = \int \,dt\,[(DS[m]\delta m)f] (t) d(t) 
+\end{equation}
+On the left side is the dot product in model space - since model space
+is just 1D in this example, that's just the numerical product. On the
+right is the dot product in data space, in the idealized continuum
+limit.
+
+Note that the VPM gradient formula is remarkable in two ways. First,
+it is exactly the same as the FWI gradient, that is, the gradient of
+$J_{\rm FWI}$ if you happen to insert the solution $f[m]$ of the
+normal equation for $f$. Second, it does not mention the annihilator
+$A$ at all: its impact is locked up in $f[m]$.
+
+The key to unlocking the meaning of the VPM gradient formula for this
+and similar problems is a remarkable feature of the derivative
+operator $DS[m]$: from the definition \ref{eqn:mod},
 \begin{equation}
 \label{eqn:deriv}
-(DS[m]\delta v)f (t) = \frac{\delta v}{4\pi v^2}\frac{df}{dt}(t-r/v) = S[m](Q[v]\delta v)f (t),
+(DS[m]\delta m)f (t) = -\frac{\delta m}{4\pi}\frac{df}{dt}(t-mr) = S[m](Q[m]\delta m)f (t),
 \end{equation}
@@
 \label{eqn:defq}
-(Q[v]\delta v)f = \frac{r \delta v}{v^2} \frac{df}{dt}.
+(Q[m]\delta m)f = -r\delta m \frac{df}{dt}.
 \end{equation}
-That is, $Q[v]\delta v$ is a skew-adjoint operator depending linearly on $\delta v$ - more on this below.
-The adjoint $(DS[m]f)^*$ is the adjoint of the map from model space to data space:
+That is, $Q[m]\delta m$ is a skew-adjoint operator depending linearly
+on $\delta m$ - more on this below.
+
+A calculation, detailed in Appendix A, yields an expression for the
+gradient \ref{eqn:grad} in terms of $Q$:
+\begin{equation}
+\label{eqn:gradcomm}
+\delta m \cdot \nabla J_{\rm VPM}[m] = \frac{1}{2}\alpha^2\int \,dt\,f[m](t)[(Q[m]\delta m),A^TA]f[m](t)
+\end{equation}
+Here, the symbol $[L,M]$ denotes the {\em commutator} of the operators
+$L$ and $M$: $[L,M]=LM-ML$.
+
+Note that the annihilator $A$ is explicitly present in
+\ref{eqn:gradcomm}. The structure displayed in this expression is
+common to many other extended inversion methods. The extreme
+simplicity of the factorization \ref{eqn:deriv}, \ref{eqn:defq} and
+the gradient expression \ref{eqn:gradcomm} is
+modified for more complex inversion problems by asymptotically
+negligible corrections \cite[]{Symes:IPTA14,tenKroode:IPTA14,Symes:Madrid}.
+
+%%%%%%%%%%%
+
+Remember that $A^TA$ amounts to multiplying by $t^2$, and $Q$ is the scaled time derivative (equation \ref{eqn:defq}), so
+\begin{equation}
+\label{eqn:comm3}
+[(Q[m]\delta m),A^TA]=-2r\delta mt
+\end{equation}
+Insert this identity into equation \ref{eqn:gradcomm} to obtain
+\begin{equation}
+\label{eqn:comm4}
+\delta m \cdot \nabla J_{\rm VPM}[m] =  -r\delta m \alpha^2\int \,dt\,tf[m]^2(t)
+\end{equation}
+Combine this identity with the formula \ref{eqn:solnnon} for the
+solution of the inner problem, and divide out the common factor
+$\delta m $, to obtain
+\begin{equation}
+\label{eqn:gradfinal}
+\nabla J_{\rm VPM}[m] = -r\alpha^2\int \,dt\,t\left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2\right)^{-1}\frac{1}{(4\pi r)^2}\fl\left(t+(m-m_*)r\right)^2
+\end{equation}
+Recall that $f_1(s)$ vanishes if $|s|\ge 1$, so $\fl(t+(m-m_*)r)$
+vanishes if $|t+(m-m_*)r| > \lambda$. Therefore the integral on the
+RHS of equation \ref{eqn:gradfinal} can be re-written
 \[
-(DS[m]f): \delta v \mapsto (DS[m]\delta v)f
+  = -r\alpha^2\int_{-(m-m_*)r-\lambda}^{-(m-m_*)r+\lambda}
+  \,dt\, \frac{t}{1+(4\pi r)^2\alpha^2 t^2}\fl\left(t+(m-m_*)r\right)^2
 \]
-and is NOT the same as the adjoint denoted by $S[m]^T$, which is the data-space-to-data-space adjoint (got that?).
-That is, this adjoint make this relation true, for all $\delta v,f,$ and $d$:
-\begin{equation}
-\label{eqn:vadj}
-\delta v \cdot (DS[m]f)^*d = \int \,dt\,[(DS[m]\delta v)f] (t) d(t) 
-\end{equation}
-On the left side is the dot product in model space - since model space is just 1D in this example, that's just the numerical product.
-On the right is the dot product in data space, in the idealized continuum limit.
+If $m > m_*+\lambda/r$, then the entire interval of integration is a proper
+subset of the negative half-axis. Consequently the first factor in the
+integrand satisfies
+\[
+ \frac{t}{1+(4\pi r)^2\alpha^2 t^2}
+  \le -(m-m_*)r+\lambda < 0
+\]
+over the interval of integration. Consequently, the integral is
+negative (since $\fl$ has a positive mean square) and so the gradient
+is positive. Similar reasoning applies to the case $m <
+m_*-\lambda/r$.
+
+To summarize,
+\begin{itemize}
+\item if $m > m_*+\lambda/r$, then $\nabla J_{\rm VPM}[m] > 0$, and
+\item if $m < m_*-\lambda/r$, then $\nabla J_{\rm VPM}[m] < 0$.
+\end{itemize}
+That is, {\em $J_{\rm VPM}$ has no local minima
+  further than  $O(\lambda)$ from the global minimum:} the gradient
+has the correct sign and slowness updates computed from it will be
+constructive, unless the slowness estimate is already ``within a
+wavelength'' of being correct.
+
+On the other hand, careful examination of equation \ref{eqn:gradfinal}
+shows that $J_{\rm VPM}$ is not convex: there is an inflection point
+$O(1/\alpha)$ from the global minimizer $m_*$, and in fact as $\alpha
+\rightarrow \infty$ for fixed $\lambda$, the $J_{\rm VPM}$
+approximates $J_{\rm FWI}$.
+
+\section{Discussion}
+There remain several important points to be made.
+
+First, the formal computations centering on the operator $Q$
+(equations \ref{eqn:deriv} through \ref{eqn:gradcomm}) depend only on
+the relation \ref{eqn:deriv} and the skew-symmetry of $Q$, which hold
+with minor modifications for other more complex waveform inversion
+problems. These other more complex problems do not submit to such a
+simple treatment as is shown in the equations following
+\ref{eqn:gradcomm}, but for example it is possible in some cases to
+use a relation analogous to \ref{eqn:deriv} to extract a relation between extended waveform inversion and traveltime tomography, via analysis of the Hessian at a zero-residual global minimizer.
+See for instance \cite{tenKroode:IPTA14,Symes:IPTA14,Symes:Madrid}.
+Up to that point the reasoning is quite general, and central to the understanding gained so far of extended inversion methods.
+
+Another issue mentioned in the last section is that the penalty weight
+$\alpha$ must be chosen somehow, and its choice influences heavily the
+convergence rate of iterative optimization algorithms applied to $J_{
+  \rm VPM}$ (or $J_{\rm ESI}$, though (as also mentioned above) it is a
+poor candidate for local optimization). Two solutions to the weight
+assignment problem has emerged in the last few years, both quite effective. One applies
+directly to VPM problem, via use of the Discrepancy Principle
+\cite[]{Morozov:84}. This algorithm controls the size of the data
+misfit term in $J_{\rm ESI}$ to lie in an interval chosen to contain
+the assumed level of data noise, by adjusting $\alpha$ sporadically as
+the optimization proceeds. While this approach requires an estimate of
+data misfit to be achieved at the optimal model, it is very effective
+in maintaining convergence \cite[]{FuSymes2017discrepancy}. Another
+approach modifies the VPM problem by means of the Augmented Lagrangian
+method \cite[]{NocedalWright}. This reformulation appears to suppress
+much of the sensitivity to $\alpha$ of VPM optimization, and has been
+successfully used in extended inversion \cite[]{Aghamiry:19}.
+
+Perhaps the most important general message implicit in the example
+presented here is that the choice of the annihilator $A$ determines
+whether the extended inversion algorithm achieves global or semiglobal
+convexity, as is accomplished in the present example. The annihilator
+used here has a property whose importance can be guessed by
+examining the gradient formula \ref{eqn:gradcomm}. The operator $Q$ is
+a first order differential operator. The gradient is a quadratic form
+whose Hessian is the communator $[Q,A^TA]$, and whose argument is
+$f[m]$. In order that the VPM objective be continuous for any model
+$m$ and finite energy data $d$, this form should admit any
+finite-energy source as argument: in technical terms, it should be a
+bounded (or continuous) operator on the Hilbert space of finite energy
+traces. If one asks for a bit more, namely that VPM objective function
+have derivatives of arbitrary order, then it is not too hard to see
+that the iterated commutators $[Q,...[Q,A^TA]...]$ must all be bounded
+operators. This is a very strong restriction on $A^TA$: this operator
+must be {\em pseudodifferential}, that is, a combination of
+differential operators and powers of the Laplace operator
+\cite[]{Tay:81}. For the present operator, $A^TA$ is
+multiplication by $t^2$, a very simple pseudodifferential
+operator. For more discussion of this requirement in the context of
+annihilators in extended inverseion, see \cite{geoprosp:2008}, where
+you can also find references to the technical backstory.
+
+This constraint actually rules out some popular approaches to FWI. To
+begin with, basic least-squares FWI as formulated in the third section
+above can be reformulated as a quadratic form whose Hessian turns out
+not to be pseudodifferential. Therefore the FWI objective is not
+smooth in data and model jointly, a fact that is linked to the
+cycle-skipping behaviour demonstrated above. More surprising, perhaps,
+the same turns out to be true for Wavefield Reconstruction Inversion (WRI),
+an extended source inversion algorithm introduced by
+\cite{LeeuwenHerrmannWRI:13}, and further developed by
+\cite{LeeuwenHerrmann:16}, \cite{WangYingst:SEG16}, and
+\cite{Aghamiry:19}, amongst others. This approach turns out to be closely
+linked to basic FWI, and can be formulated as minimization of a similar quadratic form
+which once again has a non-pseudodifferential Hessian. Not
+coincidentally, it also exhibits cycle-skipping behaviour. Just as for
+FWI, in simple cases such as the problem studied in this paper, WRI
+can be shown explicitly to have local minima far from the global
+minimum, and a region of attraction for the global minimum on the
+order of a wavelength in diameter, just as does FWI.
+
+\section{Conclusion}
+Desipite its simplicity, the single-trace transmission inversion
+problem proves typical of many more complex waveform inversion
+problems. The structure of the derivative is similar in many of these
+problems, and for the particularly simple one explained here, can be
+analysed on paper to the point of showing explicitly why a simple
+extended source approach to waveform inversion works - that is,
+generates an objective all of whose local minima are ``within a
+wavelength'' of the global minimizer. Otherwise stated, this
+particular extended source inversion is genuinely immune to
+cycle-skipping. The simple structure of this problem showcases the
+importance of the variable projection reduction (elimination of the
+extended source) and a proper choice of annihilator in the formulation
+of the basic objective. It also makes clear the central role played by
+a factorization of the linearized modeling operator, a feature shared
+with many more complex extended source methods applicable at field scale.
+
+\append{Calculation of the VPM Gradient}
 Using equations \ref{eqn:deriv}, \ref{eqn:defq} and \ref{eqn:vadj},
+and \ref{eqn:grad},
 \[
-\delta v \cdot \nabla J[v] =  \int \,dt\,[(DS[m]\delta v)f[v]] (t) (S[m]f[v]-d)(t)
+\delta m \cdot \nabla J_{\rm VPM}[m] =  \int \,dt\,[(DS[m]\delta m)f[m]] (t) (S[m]f[m]-d)(t)
 \]
 \[
-= \int \,dt\, [S[m](Q[v]\delta v)f[v](t)] (S[m]f[v]-d)(t) = \int
-\,dt\,(Q[v]\delta v)f[v](t) S[m]^T(S[m]  f[v]-d)(t)
+= \int \,dt\, [S[m](Q[m]\delta m)f[m](t)] (S[m]f[m]-d)(t) = \int
+\,dt\,(Q[m]\delta m)f[m](t) S[m]^T(S[m]  f[m]-d)(t)
 \]
@@
 \label{eqn:comm1}
-= -\alpha^2\int \,dt\, (Q[v]\delta v)f[v](t)A^TAf[v](t)
+= -\alpha^2\int \,dt\, (Q[m]\delta m)f[m](t)A^TAf[m](t)
 \end{equation}
-Since $Q[v]\delta v$ is skew-symmetric, shift it onto the other factor in this $L^2$ inner product (why not):
+Since $Q[m]\delta m$ is skew-symmetric, shift it onto the other factor in this $L^2$ inner product (why not):
 \[
-= \alpha^2\int \,dt\, f[v](t)(Q[v]\delta v)A^TAf[v](t) 
+= \alpha^2\int \,dt\, f[m](t)(Q[m]\delta m)A^TAf[m](t) 
 \]
@@
 \label{eqn:comm2}
-= \alpha^2\int \,dt\, (A^TAf[v](t)(Q[v]\delta v)f[v](t) + f[v](t)[(Q[v]\delta v),A^TA]f[v](t))
+= \alpha^2\int \,dt\, (A^TAf[m](t)(Q[m]\delta m)f[m](t) + f[m](t)[(Q[m]\delta m),A^TA]f[m](t))
 \end{equation}
@@
 \[
--\alpha^2\int \,dt\, (Q[v]\delta v)f[v](t)A^TAf[v](t) = \frac{1}{2}\alpha^2\int \,dt\,f[v](t)[(Q[v]\delta v),A^TA]f[v](t)
+-\alpha^2\int \,dt\, (Q[m]\delta m)f[m](t)A^TAf[m](t) = \frac{1}{2}\alpha^2\int \,dt\,f[m](t)[(Q[m]\delta m),A^TA]f[m](t)
 \]
@@
 \begin{equation}
-\label{eqn:gradcomm}
-\delta v \cdot \nabla J[v] = \frac{1}{2}\alpha^2\int \,dt\,f[v](t)[(Q[v]\delta v),A^TA]f[v](t)
+\label{eqn:gradcomm-app}
+\delta m \cdot \nabla J_{\rm VPM}[m] = \frac{1}{2}\alpha^2\int \,dt\,f[m](t)[(Q[m]\delta m),A^TA]f[m](t)
 \end{equation}
-Remember that $A^TA$ amounts to multiplying by $t^2$, and $Q$ is the scaled time derivative (equation \ref{eqn:defq}), so
-\begin{equation}
-\label{eqn:comm3}
-[(Q[v]\delta v),A^TA]=\frac{2r\delta v}{v^2}t
-\end{equation}
-Insert this identity into equation \ref{eqn:gradcomm} to obtain
-\begin{equation}
-\label{eqn:comm4}
-\delta v \cdot \nabla J_{\rm red}[v] =  \delta v\frac{r \alpha^2}{v^2}\int \,dt\,tf[v]^2(t)
-\end{equation}
-Combine this identity with the formula \ref{eqn:solnnon} for the solution of the inner problem, and divide out the common factor $\delta v$, to obtain
-\[
-\nabla J_{\rm red}[v] = \frac{r\alpha^2}{v^2}\int \,dt\,t\left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2\right)^{-1}\frac{1}{(4\pi r)^2}f_*\left(t-\frac{r}{v_*}+\frac{r}{v}\right)^2
-\]
-\begin{equation}
-\label{eqn:gradfinal}
-= \frac{r\alpha^2}{v^2}\int \,dt\,\left(t+\frac{r}{v_*}-\frac{r}{v}\right)\left(1 + (4\pi r)^2\alpha^2 \left(t+\frac{r}{v_*}-\frac{r}{v}\right)^2\right)^{-1}f_*(t)^2
-\end{equation}
-
-Recall that the definition of $A$ (multiplication by $t$) was justified by a modeling assumption, that the actual source is (possibly after designature) highly localized at $t=0$ - ideally a Dirac delta, but because of bandlimitation not quite.
-Still, a high degree of localization, as implicit in this choice of annihilator, means that the integral above is approximately proportional to the integrand at $t=0$, that is,
-\[
-\approx \frac{r\alpha^2}{v^2}\left(\frac{r}{v_*}-\frac{r}{v}\right)\left(1 + (4\pi r)^2\alpha^2 \left(\frac{r}{v_*}-\frac{r}{v}\right)^2\right)^{-1}
-\]
-Note that to the extent that this approximation is accurate,
-\begin{itemize}
-\item if $v > v_*$, then $\nabla J_{\rm red}[v] > 0$, and
-\item if $v < v_*$, then $\nabla J_{\rm red}[v] < 0$, and
-\end{itemize}
-That is, within the domain of validity of this approximation (setting $t=0$ in equation \ref{eqn:gradfinal}), {\em there are no local mins:} the gradient has the correct sign and velocity updates computed from it will be constructive.
-
-On the other hand, careful examination of equation \ref{eqn:gradfinal} (which involves no approximations) shows that if $\alpha$ is too large, then the finite-frequency spread of $f_*$ is guaranteed to result in non-convex behaviour, since the side-lobes will dominate.
-This message is consistent with the import of many other similar computations.
-
-\section{Discussion}
-There remain two important points to be made.
-
-First, the formal computations centering on the operator $Q$ (equations \ref{eqn:deriv} through \ref{eqn:gradcomm}) depend only on the relation \ref{eqn:deriv} and the skew-symmetry of $Q$, which hold with minor modifications for other more complex waveform inversion problems.
-These other more complex problems do not submit to such a simple treatment as is shown in the equations following \ref{eqn:gradcomm}, but for example it is possible in some cases to use the relation \ref{eqn:deriv} to extract a relation between extended waveform inversion and traveltime tomography, via analysis of the Hessian at a zero-residual global minimizer.
-See for instance \cite{tenKroode:IPTA14,Symes:IPTA14,Symes:Madrid}.
-Up to that point the reasoning is quite general, and central to the understanding gained so far of extended inversion methods.
-
-The cited papers are set in the context of subsurface-offset extended Born inversion, and demand a single scattering approximation and high-frequency asymptotics - that is, microlocal anallysis.
-Somewhat surprisingly, it has recently proven possible to establish the existence of a variant of the operator $Q$ for acoustic problems with transmission configuration but without the assumed smoothness of the material parameter fields essential for microlocal analysis.
-That is, at least in some circumstances $DS=SQ$ with approximately skew-symmetric $Q$, even when reflections (even multiple) occur in the material model.
-It remains to be seen whether enough can be established about these more general versions of $Q$ (which are for example in general not pseudodifferential0 to gain any hold over the corresponding inverse problems.
-
-The second point to be made has to do with the necessity of nested optimization, of which VPM is a variant.
-The aim is to find a minimizer of the bivariate objective $J[v,f]$ (and ultimately of data fitting, that is, FWI), but the bivariate problem is computationally intractable.
-The framework developed here can be used to explain why this is so: optimization in $v$ and $f$ simultaneously is very ill-conditioned (in the continuum limit, infinitely so), whereas reduced objetives (of which VPM produces one possibility) are much easier to optimize.
-Yin Huang's thesis \cite[]{YinHuang:16} shows this contrast between global and nested optimization very clearly, using a different extension.
-
-\section{Conclusion}
-Desipite its simplicity, the single-trace inversion transmission problem proves typical of many more complex waveform inversion problems.
-The structure of the derivative is similar in many of these problems, and for the particularly simple one explained here, can be analysed on paper to the point of showing explicitly why this approach to waveform inversino works.
-
+which is identical to equation \ref{eqn:gradcomm}.
 \bibliographystyle{seg}

### 2020-04-18 repro/papers/dso1d/paper.tex
more

@@
   \label{eqn:deffl}
-  \fl(t) = \frac{1}{\lambda}f_1\left(\frac{t}{\lambda}\right).
+  \fl(t) = \frac{1}{\sqrt{\lambda}}f_1\left(\frac{t}{\lambda}\right).
 \end{equation}
@@
 minimum, and a region of attraction for the global minimum on the
-order of a wavelength in diameter, just as does FWI.
+order of a wavelength in diameter, just as does FWI \cite[]{wwsorcas:20-01}.
 

### 2020-05-19 repro/papers/dso1d/junk.tex
stuff

The second version of the FWI objective responds to the observation
that seismic sources are themselves subject to much uncertainty, and
therefore could reasonably be included in the unknowns of the inverse
problem \cite[]{SymMink:97,VirieuxOperto:09}. Thus: chose $m$ {\em
  and} $f$ to minimize
\begin{equation}
  \label{eqn:FWIvar}
  J_{\rm var}[m,f] = \frac{1}{2}\|S[m]f-\dl\|^2.
\end{equation}
It might be thought that the additional degrees of freedom represented
by the unknown source function in \ref{eqn:FWIvar} would ``relax'' the
minimization and open up more paths to the global
solution. Paradoxically, a good way to use these extra degrees of
freedom is to get rid of them: define
\begin{equation}
  \label{eqn:FWIvarvpm}
  J_{\rm var}^{\rm vpm}[m] = \min_f \frac{1}{2}\|S[m]f-\dl\|^2
\end{equation}
Replacement of $J_{\rm var}$ by $J_{\rm var}^{\rm vpm}$ is the
so-called Variable Projection Method \cite[]{GolubPereyra:03}, applied to the problem at
hand. Note that local, respectively global minimizers $m$ of $J_{\rm
  var}^{\rm vpm}$ are necessarily the first components of local,
respectively global, minimizers of $J_{\rm var}$, therefore
minimization of $J_{\rm var}$ is logically equivalent to minimization
of $J_{\rm var}^{\rm vpm}$. A number of authors have specifically
suggested that this approach be applied eliminate ``nuisance''
parameters such as seismic source models that do not represent the
main information content to be extracted from the data
\cite[]{MulderVanLeeuwen:09,Rickett:SEG12,Aravkin:12,LiRickettAbubakar:13}.

For the present problem, it is easy to explicitly compute $ J_{\rm
  var}^{\rm vpm}$. If $f[m]$ denotes the minimizer of $J_{\rm
  var}[m,f]$ over $f$, then
\[
  J_{\rm  var}^{\rm vpm}[m]=J_{\var}[m,f[m]].
\]
Since $J_{\var}[m,f]$ is quadratic in $f$, $f[m]$ is a solution of the
normal equation
\begin{equation}
  \label{eqn:defnormal}
  S[m]^TS[m] f[m] = S[m]^T\dl
\end{equation}
It follows immediately from the definition \ref{eqn:mod} that
\begin{equation}
\label{eqn:tran}
S[m]^T g (t) = \frac{1}{4\pi r}g\left(t+mr\right)
\end{equation}
so
\begin{equation}
\label{eqn:unit}
S[m]^TS[m] f (t) = \frac{1}{(4\pi r)^2} f(t).
\end{equation}
From the definition \ref{eqn:defdata} of the data and the identity \ref{eqn:tran},
\[
  S[m]^T\dl = \frac{1}{(4\pi r)^2}\fl(t-(m_*-m)r)
\]
so from the identity \ref{eqn:unit} and the normal equatioh
\ref{eqn:defnormal} obtain
\[
  f[m] = \fl(t-(m_*-m)r).
\]
Therefore $S[m]f[m] = \fl(t-m_*r)

### 2020-06-15 repro/papers/discrep/gradstruct.tex
getting there

\section{Structure of the Gradient}
As was mentioned in the introduction, proper choice of norms in domain
and range of the example modeling operator $F$ described there renders
it unitary. If $F$ were also differentiable (rather than
``differentiable with loss of a derivative'', as explained above),
$Q[m,\delta m] = F[m]^TDF[m]\delta m$ would factor $DF[m]\delta m =
F[m]Q[m,\delta m]$, and $Q[m,\delta m]$ is skew-symmetric. This
section presents a precise variant of this structure that is shared by
the simple example studied in this paper and by many other inverse
problems in wave propagation.

The substitute for unitarity that can be achieved in for many such
problems is

\noindent {\bf Assumption D:} For each $m \in M$, $F[m]$ is
{\em approximately unitary}: there exist
\begin{eqnarray}
  W_X&\in& C^{\infty}(M,\cap_{s=0}^{\infty}{\cal I}(X^s,X^s)),\\\nonumber
  W_D&\in& C^{\infty}(M,\cap_{s=0}^{\infty}{\cal I}(D^s,D^s))
  \label{eqn:w}
\end{eqnarray}
for all $s \ge 0$, so that
\begin{equation}
\label{eqn:appinvdef}
F^{\dagger} = W_X^{-1}F^TW_D: M \rightarrow {\cal B}(D^s,X^s),
\end{equation}
satisfies
\begin{equation}
  \label{eqn:appinvprop}
  F^{\dagger}F = I +R_X,\,FF^{\dagger}=I+R_D
\end{equation}
with
\begin{eqnarray}
   R_X&\in& C^{\infty}(M,\cap_{s,t=0}^{\infty}{\cal B}(X^s,X^{s+t})),\\\nonumber
  R_D&\in& C^{\infty}(M,\cap_{s,t=0}^{\infty}{\cal B}(D^s,D^{s+t}))
  \label{eqn:r}
\end{eqnarray}

Note that the Assumption D together with condition \ref{eqn:reg1} from
Assumption C imply that $F^{\dagger}$ is also ``differentiable with loss of a derivative'':
\begin{equation}
\label{eqn:appinvderiv}
F^{\dagger} \in C^k(M, {\cal B}(D^s,X^t)) \mbox{ provided that
}s > t+k, t.
\end{equation}
$F^{\dagger}$ is the adjoint of $F$ with respect to the $M$-dependent
norms
\begin{eqnarray}
  \|x\|^2_m &=& \langle x, W_X[m]x\rangle_0,\\ \nonumber
  \|d\|^2_m &=& \langle d, W_D[m]d\rangle_0
  \label{eqn:unit}
\end{eqnarray}

\noindent {\bf Remark:} As mentioned earlier, in concrete instances,
$X^s,D^s$ are$L^2$-based  Sobolev spaces, for
which the index $s$ indicates the number of square-integrable
derivatives. The assertion \ref{eqn:r} thus means that the
``remainder'' operators $R_X,R_D$ are smoothing: outputs are
arbitrarily smoother than inputs. For the very simple example
described in the introductions, in fact $R_X=0=R_D$.

\begin{theorem}
  \label{thm:q} Under Assumptions A-D, 
the derivative $DF$ admits the factorization
\begin{equation}
  \label{eqn:decomp}
  DF[m]\delta m = F[m]Q[m,\delta m] + E[m,\delta m],
\end{equation}
in which $ Q \in \cap_{s \ge 0}C^{\infty}(M \times TM, {\cal
  B}(X^{s+1},X^s))$ and $E \in C^{\infty}(M \times TM, ,\cap_{s,t=0}^{\infty}{\cal B}(X^s,D^{s+t}))$
are linear in their second arguments. Also, $Q$ is {\em essentially skew-adjoint}, in the sense that for any $m \in M,  \delta m \in TM$, and $x_1,x_2 \in X^1$,
\begin{equation}
  \label{eqn:essskew}
  \langle Q[m,\delta m] x_1,x_2 \rangle_{X^0} = -\langle x_1, Q[m,\delta
  m]x_2\rangle_{X^0}
  + \langle R[m,\delta m]x_1,x_2\rangle_{X^0},
\end{equation}
with $R \in \cap_{s,t \ge 0}C^{\infty}(M \times TM, {\cal B}(X^s,X^{s+t}))$,
also linear in its second argument.
\end{theorem}

\begin{proof} Define $Q = F^{\dagger}DF$. Then
  \[
    FQ = (I+R_D)DF = DF + E
  \]
  \[
    D (F^{\dagger}F) = DF^{\dagger}F+F^{\dagger}DF 
  \]
  \[
    = D(W_X^{-1}F^TW_D)F +Q = -W_X^{-1}DW_X W_X^{-1}F^TW_DF
  \]
  \[+
  W_X^{-1}(Q^TF^T-E)W_DF + W_X^{-1}F^TW_D W_D^{-1}DW_DF +Q
\]
\[
=-W_X^{-1}DW_XF^{\dagger}F + Q^{\dagger}F^{\dagger}F - E^{\dagger}F + F^{\dagger}W_D^{-1}DW_DF+Q
\]
\[
  = -W_X^{-1}DW_X(I+R_X) + Q(I+R_X) - E^{\dagger}F +
  F^{\dagger}W_D^{-1}DW_DF+Q
\]
\[
  = Q^{\dagger} + Q -W_X^{-1}DW_X(I+R_X) + QR_X - E^{\dagger}F +
  F^{\dagger}W_D^{-1}DW_DF
  \]
\end{proof}

From equations \ref{eqn:grad1} and \ref{eqn:decomp},
\[
    D\tJa[m;d]\delta m = \langle 
  (DF[m]\delta m)\ax[m,d],F[m]\ax[m;d]-d\rangle_{D^0}
\]
\[
  =\langle F[m]Q[m,\delta m]\ax[m;d],F[m]\ax[m;d]-d\rangle_{D^0}
\]
\[
  = \langle Q[m,\delta m]\ax[m;d],F[m]^T(F[m]\ax[m;d]-d)\rangle_{D^0}
\]
\[
  = -\langle Q[m,\delta m]\ax[m;d],A^TA\ax[m;d]\rangle_{X^0}
\]
\[
  = -\alpha^2\langle A^TA Q[m,\delta m]\ax[m;d],\ax[m;d]\rangle_{X^0}
\]
\[
  =\alpha^2\langle [Q[m,\delta m],A^TA]\ax[m;d],\ax[m;d]\rangle_{X^0} -\alpha^2\langle Q[m,\delta m]A^TA\ax[m;d],\ax[m;d]\rangle_{X^0}
\]
\[
  =\alpha^2\langle [Q[m,\delta m],A^TA]\ax[m;d],\ax[m;d]\rangle_{X^0}-\alpha^2\langle Q[m,\delta m]^T\ax[m;d], A^TA\ax[m;d]\rangle_{X^0}
\]

### 2020-06-15 repro/papers/discrep/paper.tex
getting there

\title{A Discrepancy-Based Algorithm for Extended Inversion}
\author{William. W. Symes \thanks{P. O. Box 43, Orcas WA USA, email {\tt symes@rice.edu}.}}

\lefthead{Symes}

\righthead{Discrepancy}

\maketitle
\begin{abstract}
  Study of an extremely simple single-trace transmission example shows
  how an extended source formulation of full waveform inversion can
  produce an optimization problem without spurious local minima
  (``cycle skipping''), hence efficiently solvable via Newton-like
  local optimization methods. The data consist of a single trace
  recorded at a given distance from a point source. The velocity or
  slowness is presumed homogeneous, and the target source wavelet is
  presumed quasi-impulsive or localized near zero time lag. The source
  is extended by permitting energy to spread in time, and the spread
  is controlled by adding a weighted mean square of the extended
  source wavelet to the data misfit, to produce the extended inversion
  objective. The objective function and its gradient can be computed
  explicitly, and it is easily seen that the error in any local
  minimizer is bounded by a multiple of the permitted energy spread,
  which may be interpreted as a wavelength. The derivation shows several
  important features of all similar extended source algorithms. For
  example, nested optimization, with the source estimation in the
  inner optimization (variable projection method), is essential. The
  choice of the penalty operator, controlling the extended source
  degrees of freedom, is critical: in order to produce an objective
  immune from cycle-skipping, the penalty operator must be
  (pseudo-)differential. The penalty weight is chosen dynamically,
  according to a version of the Discrepancy Principle, in order to
  accelerate convergence to a near-global minimum. Like many other
  extended formulations, the example problem considered here permits
  the discrepancy criterion to be satisfied with zero penalty
  weight. This unusual feature removes a common obstacle to the use of
  the discrepancy principle, namely the determination of a
  satisfactory estimate of the penalty weight: the initial weight may
  be set to zero, and the dynamic algorithm updates it until
  convergence is achieved.

\end{abstract}

\section{Introduction}
Full Waveform Inversion (FWI), or estimation of earth structure by
model-driven least squares data fitting, is now well-established as a
useful tool for probing the earth's subsurface
\cite[]{VirieuxOperto:09,Fichtner:10}. However, so-called ``cycle-skipping'', the tendency of iterative FWI
algorithms to stagnate at suboptimal and geologically uninformative
earth models, still impedes its use. Because the computational size of field inversion tasks
is very large, only iterative local (descent) minimization of the data
misfit function is computationally feasible. However local
descent methods avoid suboptimal stagnation only if initial models are
already quite close to optimal, in the sense of predicting the arrival
times of seismic events to within a small multiple of a dominant
wavelength \cite[]{GauTarVir:86,Plessix:10}.

This paper concerns one of the several ideas that have been advanced to
overcome cycle-skipping, namely so-called extended inversion
\cite[]{geoprosp:2008}. ``Extended'' signifies that addional degrees
of freedom are provided to the modeling process, in the hope of
opening up more effective routes to geologically informative models
with acceptable data fit. Since these extended degrees of freedom are
not part of the basic physics chosen to model the data acquisition
process, they should be suppressed in the eventual solution. Extended
inversion methods differ by the choice of additional degrees of
freedom, and by choice of penalty applied to eliminate them in the final
result.

Many of these extended inversion concepts sound plausible, and appear
to work at least to some extent as one might hope from their heuristic
justifications. However very few of these approaches have been
underwritten by mathematical argument: in essence, they are mostly
justified only ``in the rear-view mirror'', with no assurance that
failure is not just around the corner, at the next example. On top of
that, some of these approaches, for example those based on the
computationally attractive Variable Projection Method (``VPM'') of
\cite{GolubPereyra:03}, are cast in such form that the reasons for
success are not readily apparent.

This note shows exactly how extended inversion leads to successful solution
of a very simple wave propagation inverse
problem, which asks that a homogeneous velocity field be deduced from
one trace at known offset. I put forward this inverse problem and
extension-based solution not because there are not simpler
ways of answering the question it poses - there certainly are - but
because the formal ingredients of waveform-based velocity estimation
in this very simple setting are common to many similar extended
inversion algorithms, and because in this case every computation can
be done analytically, nearly to completion. In particular, it becomes
clear why the VPM gradient formula produces a constructive update,
with no possibility of stagnation away from the global minimum.

The extended inversion approach
developed here uses a {\em source extension}, in which source
parameters form the additional degrees of freedom. This type of
extension presumes that the actual or target source is constrained in
some way; the extended source is allowed to violate the
constraint. For recent overview of source extension methods, see
\cite{HuangNammourSymesDollizal:SEG19}. Source
extension methods have computational complexity approximately the same
as that of FWI, a signal advantage over the alterantive medium
extension class described for instance in \cite[]{geoprosp:2008}.

For the problem considered here, the source model amounts to a
wavelet, and the target wavelet is assumed to be non-zero only in a
short time interval (an approximate impulse, perhaps as the result of
signature deconvolution). The extension consists in permitting energy
to spread in time at intermediate iterations of the inversion. A
penalty for energy spread drives the extended source towards a focused
source approximately satisfying the assumed constraint. Not all
penalties are created equal. Two plausible penalty operators are
studied here, both multiplication operators: one by the characteristic
function of the target support, the other by a smooth function of
time. Only the smooth multiplier avoids cycle-skip. This distinction
is quite general, as will be explained in the Discussion section.

%despite not incorporating explicitly the extension measure
%(``annihilator'') central to the approach.

I begin with a quick sketch of constant density acoustics, and
describe the single-trace transmission inverse problem.. For
completeness, I show how the standard FWI approach to this problem,
with the small support of the source imposed as a hard constraint,
generates multiple local minima that will be found by any descent
method unless the initial slowness estimate predicts travel time from
source to receiver with an error on the order of a wavelength. The
next section describes the extended source objective, and the reduced
objective produced by VPM, in general terms. As VPM eliminates the
extended source, this function depends only on the velocity, just as
does the FWI objective.  Of course, it also depends on the operator
used to define the penalty (``annihilator''). I propose two choices: a
soft version of the support constraint, and an even softer constraint,
namely multiplication by time. A nearly-explicit calculation of the
VPM gradient for these two choices shows that the first leads again to
multiple local minima far from the target solution, whereas the second
does not. In fact, in the second case, the only stationary points are
``within a wavelength'' of the correct velocity, used to build the
data: that is, cycle-skipping cannot occur. For this second, more
successful penalty, I introduce the discrepancy based dynamic
weighting algorithm, and show how it accelerates the convergence
towards a local minimizer, which has been shown to be an acceptable
approximation to a global minimizer. The paper ends with a discussion
of the parallels between the calculations presented here and the
structure of other extended inversion methods applicable to
field-scale velocity estimation, and the critical role that the
differential nature of the extension penalty plays in the success of
this and other extension methods.

I shall use the abbreviations ${\cal B}(X,Y)$ and ${\cal I}(X,Y)$ for
the algebra of bounded linear operators from the Hilbert space $X$ to
the Hilbert space $Y$, and its subalgebra of invertible operators.

%\section{Theory}

%\cite{FuSymes2017discrepancy} introduce a version of the discrepancy
%principle and its application to separable nonlinear least squares
%problems in penalty form \ref{eqn:basic}. The algorithm developed
%there is incomplete, however: it lacks  termination criteria. In this
%section, I review the discrepancy-based algorithm and introduce an
%appropriate stopping rule, thus completing the development of the
%algorithm presented in \cite{FuSymes2017discrepancy}.

\section{An Example: Single Trace Modeling}
Assume small amplitude (linearized) acoustic propagation, constant
density, and isotropic point source and receiver. Denote by $m(\bx)$
the slowness (reciprocal velocity) at spatial position $\bx$, $w(t)$
the time dependence of the point source (``wavelet'') at location
$\bx=\bx_s$. Then the (excess) pressure field $p(\bx,t)$ obeys a
scalar wave equation:
\begin{eqnarray}
  \label{eqn:awe}
  \left(m(\bx)^2\frac{\partial^2 p}{\partial t^2} - \nabla^2\right) p(\bx,t) &=&
                                                                         w(t)\delta(\bx-\bx_s) \nonumber\\
  p(\bx,t)&=&0, t\ll 0
\end{eqnarray}
Suppose that a single trace is recorded, at distance $r>0$ from the
source position $\bx_s$. The dominant information in a single
trace is the transient signal time of arrival, constraining only the mean slowness in the
region
between source and receiver, so assume that the
slowness is constant, that is, independent of position $\bx$. The pressure field is simply the  the source
wavelet $w(t)$ convolved with the
acoustic Green's function, for which an analytic expression is
available in the constant $m$ case \cite[]{CourHil:62}:
\begin{equation}
  \label{eqn:homsol}
  p(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-m|\bx-\bx_s|\right).
\end{equation}

The receiver location $\bx_r$ lies at distance $r$ from the source
location $\bx_s$, that is, $|\bx_r-\bx_s|=r$. The predicted signal at
$p(\bx_r,t)$ depends nonlinearly on the slowness $m$ and linearly on the
source wavelet $w$. Therefore it is naturally represented as the
action of a $m$-dependent linear operator $S[m]$ on $w$:
\begin{equation}
\label{eqn:mod}
F[m]w(t)= p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right).
\end{equation}

Ignoring amplitude, this map implements a $m$-dependent time
shift. This time shift operator is the basis of many descriptions of
the cycle-skipping phenomenon (for example, \cite{VirieuxOperto:09},
Figure 7), so it is unsurprising that an analysis of cycle-skipping
can be based on the simple modeling operator described above, which
amounts essentially to a time shift.

The slowness $m$ must be positive, as follows from basic acoustics,
and in fact reside in a range $M=(m_{\rm min}, m_{\rm max})$ characteristic of the
material model: for crustal rock, a reasonable choice would be
$m_{\rm min}=0.125, m_{\rm max}=0.6$ s/km. The wavelet $w$ is
naturally assumed to be 
square-integrable, corresponding to finite power transfer from the
source process to the propagating wavefield $p$
\cite[]{SantosaSymes:00}. The predicted data $F[m]w$ is then also
square-integrable.

FIeld recording takes place over finite time intervals (and with
finite sample rate, a detail I shall ignore in this paper). Denote by
$[t_{\rm min},t_{\rm max}]$ the recording interval for the
hypothetical single trace data. Then natural choices for domain and
range of $F[m]$ are $W = L^2(\bR)$ and $D=L^2([t_{\rm min},t_{\rm
  max}])$ respectively (that is, interpret the definition \ref{eqn:mod}
as holding for $t \in [t_{\rm min},t_{\rm  max}]$).

\section{The Inverse Problem}

Part of the definition of an inverse problem based on the model
definition \ref{eqn:mod} is that observed data $d \in D$ be fit: that
is, $m, w$ must be found so that $F[m]w \approx d$. However this
formulation does not provide enough information to determine the model
$(m,w) \in M \times W$: indeed, the definition
\ref{eqn:mod} shows that $F[m]$ is surjective for any $m \in M$, so
the data can always be fit precisely by appropriate choice of $w$, and
$m$ is completely
unconstrained.  In order that data fit constrain $m$, it is necessary
to further constrain $w$ {\em a priori} - that is, to add information
beyond the trace data $d$.

One natural constraint is to assume that the support of $w$ is
limited. An assumption of small support may be justified as
follows. In practical seismic data processing, actual sources may act
over considerable time intervals - for example, the most commonly used
marine source (airgun) produces an oscillating pulse that dies away
slowly. It is commonplace to estimate this pulse then deconvolve it
from the data by safeguarded Fourier division or other means. This
so-called signature deconvolution results in modified data
corresponding to a source equal to signature deconvolution of the
pulse estimate itself. The deconvolved source pulse approximates an
impulse ($\delta(t)$): it cannot have point support, due to the finite
frequency nature of seismic data, hence the bandlimited nature of the
pulse estimate, but its support is ultimately much smaller than that
of the original pulse estimate.

Therefore, supplement the observed data $d \in D$ with a presumed 
source support radius $\lambda$. Define 
$\lW=\{w \in W: \mbox{ supp }w \subset [-\lambda,\lambda]\}$, and 
denote by $\lF$ the restriction of $F$, defined in equation 
\ref{eqn:mod}, to $M \times \lW$.

Note that for $w \in \lW$, $m \in M$,
$\mbox{supp }F[m]w \subset [ m_{\rm min}r-\lambda,
m_{\rm max}r+\lambda] \cap
[t_{\rm min},t_{\rm max}]$. Also, $\lF[m]$ is coercive for all $m \in
M$ (so that $w$ is uniquely determined) only if
$[ m_{\rm min}r-\lambda,m_{\rm max}r+\lambda] \subset
[t_{\rm min},t_{\rm max}]$. Define $\lambda_{\rm max}
>0$, so that $M, r,$ and $T$ by
\begin{equation}
  \label{eqn:fullrec}
  [ m_{\rm min}r-\lambda_{\rm max}, m_{\rm max}r+\lambda_{\rm max}]
  \subset [t_{\rm min},t_{\rm max}].
\end{equation}
Then for 
$m \in M$, $0 <\lambda \le \lambda_{\rm max}$, $F[m] \in {\cal B}(\lW,D)$ is coercive.

Note that condition \ref{eqn:fullrec} implies
\begin{eqnarray}
  \lambda_{\rm max} & \le & t_{\rm max}-m_{\rm max}r,\nonumber \\
  t_{\rm min}-m_{\rm min}r & \le & -\lambda_{\rm max} \label{eqn:lminc}
\end{eqnarray}
hence for any $m \in M$,
\begin{equation}
  [-\lambda_{\rm max}, -\lambda_{\rm max}] \subset[ t_{\rm min}-mr ,
  t_{\rm max}-mr].
  \label{eqn:zeroinc}
\end{equation}

In terms of this infrastructure, the inverse problem studied in 
this paper is:

\noindent {\bf Problem Statement:} given data $d \in D$, relative 
error level $\epsilon >0$, and support radius $0 <\lambda \le 
\lambda_{\rm max}$, find 
$(m,w) \in M \times \lW$ for which 
\begin{equation}
  \label{eqn:probstat0}
  \|\lF[m]w-d\| \le \epsilon\|d\|. 
\end{equation}

The support constraint is closely linked to the folk theorem about FWI noted many
times in the literature: convergence of a descent method requires that
the initial slowness must be known to ``within a (fraction of a) wavelength''. The
relation is a consequence of Heisenberg's inequality
(see for example \cite{Folland:07}, p. 255). Assume that $w \in L^2(\bR)$, and that both $tw$ and derivative $dw/dt$ are
square-integrable (an essential additional assumption, as will be
explained below). Then for any $t_0, k_0 \in \bR$,
\[
  \left(\int\,dt\, (t-t_0)^2|w(t)|^2\right)
  \left(\int\,dk\,(k-k_0)^2|\hat{w}(k)|^2\right)
  \ge\frac{\|w\|^4}{16 \pi^2}
\]
in which $\hat{w}$ denotes the Fourier transform of $w$. Rearrange
slightly and choosing $t_0=0, k_0=0$,
\[
  \left(\int\,dt\, t^2\frac{|w(t)|^2}{\|w\|^2}\right)
  \left(\int\,dk\,k^2\frac{|\hat{w}(k)|^2}{\|\hat{w}\|^2}\right)
  \ge\frac{1}{16 \pi^2}
\]
If $w \in \lW$, then the first factor is $\le \lambda^2$, whereas the
second factor is the power spectrum mean square estimate of
frequency squared $k^2$. That is the power spectrum RMS estimate of
frequency is at least $1/(4 \pi \lambda)$. In the sense of the RMS
spectral estimator of frequency, then, $\lambda$ is proportional to an upper
bound on representative wavelength: substantial energy must be present
at frequencies larger than $1/\lambda$.

The best case for data fitting is clearly the one in which the data
can be fit precisely: that is, there exists $(m_*,w_*) \in M\times \lW$ so that
\begin{equation}
  \label{eqn:defdatanonoise}
  d=\lF[m_*]w_*.
\end{equation}
Such data $d$ is {\em noise-free}, in the range of the map $\lF$, and a solution of the Problem Statement
exists with arbitrarily small $\epsilon>0$. I will relax the noise-free assumption
later in the paper: until further notice, assume that $d \in
D$ satisfies the condition \ref{eqn:defdatanonoise}.

\section{Full Waveform Inversion: hard constraint}
A natural formulation of the Problem Statement \ref{eqn:probstat0} is
via least squares: that is, given $d \in D$, $\lambda >0$, find $(m,w)
\in M \times \lW$ to minimize $e: M \times W
 \rightarrow \bR$, defined by
\begin{equation}
  \label{eqn:FWI}
  e[m,w;d] = \frac{1}{2}\|\lF[m]w-d\|^2.
\end{equation}
A pair $(m,w) \in M\times \lW$
satisfies the condition \ref{eqn:probstat0} iff $e[m,w;d] \le \epsilon^2/2$, so a solution exists iff a 
global minimizer of $e$ satisfies this inequality.

In this discussion, I will presume that $d$ is 
noise-free, as defined at the end of the last section (equation
\ref{eqn:defdatanonoise}). Thus {\em a priori} a global minimizer of
$e$ is a solution of the Problem Statement for any
$\epsilon > 0$. The only
question is how to find such a global minimizer. As mentioned in the
introduction, the simple model problem studied here is a proxy for
large-scale data fitting problems for which the only computationally
feastible approach to minimization of analogs of $e$
is some form of descent-based iterative local optimization. Any such
algorithm requires the specification of an initial iterate
$(m_0,w_0)$. Therefore the key question is: what
constraints must be imposed on $(m_0,w_0)$ in order that the resulting
sequence of iterates convertes to a global minimizer?

To answer this question, note that 
\[
 e[m,w;d] =  \frac{1}{32\pi^2
    r^2}\int_0^T\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2
\]
Since $w_*, w$ vanish for $|t|>\lambda$,
$\lF[m_*]w_*(t)$ vanishes if $|t-m_*r|>\lambda$ and $\lF[m]w$ vanishes if $|t-mr|>\lambda$. So if $|mr-m_*r|
= |m-m_*|r > 2\lambda$, then $|t-mr|+|t-m_*r| \ge |mr-m_*r| >
2\lambda$ so either $|t-mr|>\lambda$ or $|t-m_*r|>\lambda$, that is,
either $\lF[m]w(t)=0$ or $\lF[m_*]w_*=0$. Therefore $\lF[m]w$ and
$\lF[m_*]w_*$ are orthogonal in the sense of the $L^2$ inner product
$\langle \cdot,\cdot \rangle_D$ on $D$:
\begin{equation}
  \label{eqn:ortho}
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \langle F[m]w,
  F[m_*]w_*\rangle_D = 0
\end{equation}
But $d = \lF[m_*]w_*$, so this is the same as saying that $d$ is
orthogonal to $F[m]w$. So conclude after a minor manipulation that
\begin{equation}
  \label{eqn:iso}
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, e[m,w;d]=\frac{1}{32\pi^2
    r^2}(\|w\|^2 + \|w_*\|^2).
\end{equation}
That is, for slowness $m$ in error by more than $2\lambda/r$ from the 
target slowness $m_*$, the means square error (FWI objective) $e$ is independent of
$m$, and its minimum over $w$ is attained for $w=0$:
\begin{theorem}
  \label{thm:fwi}
  Suppose that $0 <\lambda\le \lambda_{\rm max}$,  $m_* \in M, w_*
  \in \lW, d=\lF[m_*]w_*$ is noise-free data per definition \ref{eqn:defdatanonoise},
  Under assumption \ref{eqn:fullrec}, for any $m \in M$ with $r|m-m_*|>2\lambda$,
\begin{equation}
  \label{eqn:isovpm}
 \min_w e[m,w;d]=e[m,0,d]=\frac{1}{32\pi^2 r^2}\|w_*\|^2,
\end{equation}
and any such $(m,0)$ is a stationary point of $e$.
\end{theorem}

Therefore local minimizers of $e$ abound, as far as you like from the
global minimizer $(m_*,w_*)$. Local exploration of the FWI objective
$e$ gives no useful information whatever about constructive search
directions, and descent-based optimization tends to fail if the
initial estimate $m_0$ is in error by more than $2\lambda/r$
(``further than a multiple of a wavelength'', per discussion
above). In fact the actual behaviour of FWI itererations is worse
(failure if $m_0$ is in error by ``half a wavelength''), as follows
from a more refined analysis of the ``cycle-skipping'' local behaviour of $e$ near its
global minimizer.

\section{Soft constraints and penalty formulation}

For inverse problems based on wave propagation, enforcing hard
constraints on the solution ($w \in \lW$) leads to the pathology
similar to that described in the last
section for the single-trace transmission problem, and to failure of
local optimization methods to converge to a useful solution. A soft
constraint or penalty approach can avoid this obstacle.

This
discussion presumes that the penalty is quadratic, constrains only
$w\in W$
(as it is a soft version of the condition $w \in \lW$), and is
$m-$independent. Therefore the penalty function takes the form
\begin{equation}
\label{eqn:pen}
p[w] = \frac{1}{2}\|Aw\|_Y^2 
\end{equation}
where $A \in {\cal B}(W,Y)$, $Y$ being another suitable Hilbert
space.

The penalty approach asks that a linear combination of $e,p$,
depending on a relative weight $\alpha$:
\begin{equation}
\Ja[m, w;d] = e[m,w;d] + \alpha^2 p[w] = \frac{1}{2}(\|F[m]w-d\|^2 + \alpha^2 \|Aw\|^2).
\label{eqn:pen}
\end{equation}
be minimized over $m\in M$ and $w\in W$.

The choice of penalty weight $\alpha$ has a profound influence on the
character of this optimization problem. For the moment, I will mandate
only that $\alpha \ge 0$. A complete description of an algorithm using
the penalty function \ref{eqn:pen} to solve the inverse problem
\ref{eqn:probstat0} must include a selection principle for $\alpha$. I
shall describe one such principle below.

While minimization of $\Ja$ might be tackled directly - by
alternately minimizations between $m$ and $w$, or by computing updates
for $m$ and $w$ simultaneously - such joint mimization performs
poorly, as \cite{YinHuang:16} has shown. The reason for this poor
performance is that $\Ja$ has dramatically different
sensitivity to $m$ versus $w$,
%especially for high frequency $f$, as
%the reader will see below.
as will be established below.
Instead, a nested approach, in which $w$ is
eliminated in an inner optimization,
generally gives far better numerical performance.  This {\em Variable
Projection Method} (VPM) \cite[]{GolubPereyra:03} takes advantage of
$\Ja$ being quadratic in
$w$ to eliminate $w$ from the problem statement, as described in the
following theorem.

In stating this result, I will make use of the the Hessian operator of
$\Ja[m,\cdot]$, which is $F[m]^TF[m] + \alpha^2 A^TA$. The
superscript $T$ denotes the adjoint with respect to the inner products
in $W=L^2(\bR)$ and $D=L^2([t_{\rm min},t_{\rm max}])$.

\begin{theorem}
\label{thm:red}
Assume that
\begin{equation}
\label{eqn:norminv}
F[m]^TF[m] +\alpha^2 A^TA \in {\cal I}(W,W).
\end{equation}
Then the unique stationary point $\aw[m;d] \in W$ of
$\Ja[m,\cdot;d]$ is the solution of the {\em normal equation}:
\begin{equation}
  \label{eqn:norm}
  (F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td,
\end{equation}
Define the reduced or variable projection objective by
\begin{equation}
  \label{eqn:redexp}
  \tJa[m;d] =\Ja[m,w_{\alpha}[m;d];d] = \min_w \Ja[m,w;d]
\end{equation}
Then
\begin{equation}
  \label{eqn:objexp}
  \tilde{J}_{\alpha}[m;d] =\langle d, (I-F[m]
  (F[m]^TF[m]+\alpha^2A^TA)^{-1}F[m]^T)d\rangle_D
\end{equation}
\end{theorem}

\begin{proof}
Substitute $w_{\alpha}[m;d]$ for $w$ in the definition \ref{eqn:pen}
of $J_{\alpha}$, using the normal equation \ref{eqn:norm} and 
rearranging, obtain the result.
\end{proof}

\section{All stationary points are (near) global minimizers -
  sometimes}
It is not immediately apparent why use of a soft constraint should
produce an optimization problem more amenable to descent methods than
does the hard-constrained inverse problem (FWI). In this section, I
will show that the penalty formulation can yield a reduced problem
with for which any stationary point is close to a global minimizer,
{\em for some choices of penalty operator, but not for other choices
  that appear equally or even more suitable}.

For the single-trace transmission inverse problem, $\tilde{J}_{\alpha}$ is explicitly
computable. First observe that apart from amplitude, $F[m]$ is
unitary: for $g \in D$,
\begin{equation}
\label{eqn:tran}
F[m]^T g (t) =
\left\{
  \begin{array}{c}
    \frac{1}{4\pi r}g\left(t+mr\right), \, t \in [t_{\rm min}-mr,
    t_{\rm max}-mr],\\
    0, \mbox{ else.}
  \end{array}
\right.
\end{equation}
so
\begin{equation}
  \label{eqn:unit}
  F[m]^TF[m] = \frac{1}{(4\pi r)^2}{\bf 1}_{[t_{\rm min}-mr,  
    t_{\rm max}-mr]}
\end{equation}
in which ${\bf 1}_{S}$ denotes
multiplication by the characteristic function of a measurable 
$S \subset \bR$.

Therefore the normal equation for the minimizer on the RHS of equation \ref{eqn:red} is
\begin{equation}
  \label{eqn:norm1}
  \left(\frac{1}{(4\pi r)^2} {\bf 1}_{[t_{\rm min}-mr,  
      t_{\rm max}-mr]} + \alpha^2 A^TA\right)w= F[m]^Td.
\end{equation}

At this point I have to come clean about the actual choice of
$A$. The role of $A$ is to penalize distance from the constrained
subspace $\lW$, in some sense.

I will introduce two options for $A$. Both are scalar 
multiplication operators defined by a choice of $a \in L^{\infty}(\bR)$:
\begin{equation}
  \label{eqn:annmult}
  A w(t)= a(t)w(t), \, t\in \bR.
\end{equation}

An
operator $A_{\lambda}$ of which $\lW$ is the
null space would be a natural 
choice. Such operators have come to be called ``annihilators'', since
they map all members of the constraint subspace $\lW$ to zero.
Define 
\begin{eqnarray}
  A = E^c_{\lambda}&=&I - E_{\lambda},\mbox{ where } \nonumber \\
  E_{\lambda}w(t) &=&{\bf 1}_{[-\lambda,\lambda]}(t)w(t). 
                      \label{eqn:ann0}
\end{eqnarray}
That is, $E_{\lambda}$ is the orthogonal projector onto $\lW$,
and $E_{\lambda}^c$ is the orthogonal projector onto its
orthocomplement, an operator of the form \ref{eqn:annmult} with $a
= 1 - {\bf 1}_{[-\lambda,\lambda]}$.

A second possible
penalty operator penalizes energy away from
$t=0$: choose $\tau > 0$ and set
\begin{equation}
  \label{eqn:ann}
  a(t) = \min(|t|, \tau).
\end{equation}
The cutoff $\tau$ will be chosen large enough to be effectively inactive:
specifically, hindsight suggests
\begin{equation}
  \label{eqn:taudef}
  \tau = \max \{t_{\rm min}-m_{\rm max}r, t_{\rm max}-m_{\rm min}r\}.
\end{equation}
This
particular annihilator has been employed in earlier papers on extended
source inversion
\cite[]{Plessix:00a,LuoSava:11,Warner:14,HuangSymes:SEG15a,Warner:16,HuangSymes:GEO17}.

With these choices, the normal equation \ref{eqn:norm1} becomes
\begin{equation}
\label{eqn:norm2}
\left(\frac{1}{(4\pi r)^2}  {\bf 1}_{[t_{\rm min}-mr,  
      t_{\rm max}-mr]} + \alpha^2a^2\right)w= F[m]^Td.
\end{equation}

\begin{theorem}
  \label{thm:norminv}
  With the assumptions \ref{eqn:mod}, \ref{eqn:fullrec},
  \ref{eqn:annmult} and either choice of penalty multiplier $a$
  described above,
  \begin{itemize}
  \item[1. ]The normal operator $F[m]^TF[m] + \alpha^2A^TA$ is
    invertible for any $m \in M$, $\alpha > 0$;
  \item[2. ] For any $d \in D$,
    \begin{equation}
      \aw[m;d](t) = \left\{
        \begin{array}{c}
          \left(\frac{1}{(4\pi r)^2} + \alpha^2
          a^2(t)\right)^{-1}\frac{1}{4 \pi r}d(t+mr), t \in [t_{\rm
          min}-mr, t_{\rm max}-mr];\\
          0, \mbox{ else.}
        \end{array}
      \right.
    \end{equation}
  \item[3. ] If $d$ is noise-free, as in equation
    \ref{eqn:datadefnonoise},
    \begin{equation}
      \label{eqn:solnnon}
      \aw[m,d](t)= \left(1+ (4\pi r)^2\alpha^2 a(t)^2\right)^{-1}w_*\left(t+(m-m_*)r\right) 
    \end{equation}
  \end{itemize}
\end{theorem}

Sincee normal operator is invertible, the reduced
objective $\tJa$ is given by the expression \ref{eqn:objexp}.
Employing the expression \ref{eqn:tran} for the action of $F[m]^T$,
the definition \ref{eqn:defdatanonoise} of noise-free data,
and the realization \ref{eqn:norm2} of the normal operator as a
multiplication operator, obtain after some simplification

\begin{equation}
  \label{eqn:residnorm}
  e[m,\aw[m,d];d] 
= 8 \pi^2 r^2 \alpha^4\int\,dt\,a(t-(m-m_*)r)^4(1+(4\pi r)^2 \alpha^2 
    a(t-(m-m^*)r)^2)^{-2}w_*(t)^2.
\end{equation}
\begin{equation}
  \label{eqn:anninorm}
  p[m,\aw[m,d];d] = \frac{1}{2}\int \,dt\,  a(t-(m-m_*)r)^2 \left(1+ (4\pi r)^2\alpha^2
    a(t-(m-m_*)r)^2\right)^{-2}w_*(t)^2.
\end{equation}
so finally
\begin{equation}
\label{eqn:expj}
\tJa[m;d] = =\frac{\alpha^2}{2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
  a(t-(m-m^*)r)^2)^{-1}w_*(t)^2.
\end{equation}

\begin{theorem}
  \label{thm:boxcarbad}
  For $A=E^c_{\lambda}$, that is, $a=1-{\bf 1}_{-\lambda,\lambda]}$ in
  the definition \ref{eqn:expj}. Then if $|m-m_*| >  2\lambda/r$,
  \[
    \tJa[m;d] = \frac{\alpha^2}{2(1+(4 \pi r)^2 \alpha^2)}\|w_*\|^2.
  \]
\end{theorem}

\begin{proof}
  This is clear, as the supports of $w_*$ and $t \rightarrow {\bf
    1}_{[-\lambda,\lambda]}(t-(m-m_*))$ are disjoint under this
  condition.
\end{proof}

\noindent {\bf Remark.}
One might have thought that $A=E^c_{\lambda}$ would be a better choice
of annihilator, as for noise-free
data, the solution set defined by the problem statement
\ref{eqn:probstat0} is the same as the set of global minimizers of
$\tJa$ in this case. However, for this choice of annihilator,
$\tJa$ exhibits the same feature as the mean square error $e$, namely
a continuum of stationary points at any distance from the global
minimizer $m_*$ greater than a multiple of $\lambda$. Therefore the
extended inversion with this choice of annihilator is no more amenable
to local optimization than is FWI.

Surprisingly, the choice $A=S$ yields an optimization
problem has better global convergence properties:

\begin{theorem}
  \label{thm:rampgood}
  For $A=S$, that is, $a(t)=t$ in the definition \ref{eqn:expj}, all
  stationary points of $\tJa$ lie in the interval
  \[
    [m_*-\lambda/r, m_*+\lambda/r].
  \]
\end{theorem}
\begin{proof}
  A brief calculation yields
\begin{equation}
\label{eqn:gradfinal}
\frac{\partial}{\partial m}\tJa[m;d] = -r\alpha^2\int \,dt\,t\left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2\right)^{-2}\frac{1}{(4\pi r)^2}w_*\left(t+(m-m_*)r\right)^2
\end{equation}
Recall that $w_*(t+(m-m_*)r)$
vanishes if $|t+(m-m_*)r| > \lambda$. Therefore the integral on the
RHS of equation \ref{eqn:gradfinal} can be re-written
\[
  = -r\alpha^2\int_{-(m-m_*)r-\lambda}^{-(m-m_*)r+\lambda}
  \,dt\, \frac{t}{(1+(4\pi r)^2\alpha^2 t^2)^2}w_*\left(t+(m-m_*)r\right)^2
\]
If $m > m_*+\lambda/r$, then the entire interval of integration is a proper
subset of the negative half-axis. Consequently the first factor in the
integrand satisfies
\[
 \frac{t}{1+(4\pi r)^2\alpha^2 t^2}
  \le -(m-m_*)r+\lambda < 0
\]
over the interval of integration. The integral is
negative, so the derivative
is positive. Similar reasoning applies to the case $m <
m_*-\lambda/r$. Thus
\begin{eqnarray}
  m > m_*+\lambda/r & \Rightarrow & \frac{\partial \tJa[m;d]}{\partial
    m}> 0; \nonumber\\
  m < m_*-\lambda/r & \Rightarrow & \frac{\partial \tJa[m;d]}{\partial
                                    m}< 0.
\label{eqn:leftright}
\end{eqnarray}
\end{proof}

\noindent {\bf Remark.} That is, {\em with the choice $A=S$, $\tJa$ has no local minima
  further than  $O(\lambda)$ from the global minimum:} the gradient
has the correct sign and slowness updates computed from it will be
constructive, unless the slowness estimate is already ``within a
wavelength'' of being correct.

\section{Extended Inversion}
The preceding theorem established that a proper choice of annihilator leads to a
reduced penalty objective all of whose stationary points are within
$O(\lambda)$ of the target slowness $m_*$, provided that the data are
noise-free. This result leaves open two questions:
\begin{itemize}
\item how does one use this reduced penalty minimization to produce
  a solution of the inverse problem as in problem statement
  \ref{eqn:probstat0}? There is no reason to think that the minmizer
  $\bar{m}$ of $\tJa$ with produce $\aw[\bar{m};d] \in \lW$, as
  required by the problem statement.
\item how does one answer the same question for noisy data?
\end{itemize}

A partial resolution of these issues lies in the relation between the
annihilator $S$ (multiplication by $|t|$) and the family of projectors
$\{E_{\lambda}: 0 \le \lambda \le \lambda_{\rm max}\}$. In fact, this
family of operators is a resolution of the identity for the positive
semidefinite self-adjoint $S$:
\begin{eqnarray}
  \label{eqn:res1}
  S &=& \int_{0}^{\|A\|} \lambda d\lE,\nonumber\\
  S^2=S^TS  &=& \int_{0}^{\|A\|} \lambda^2 d\lE.
\end{eqnarray}
This relation is similar to that explained in the example on p. 313 of \cite{Yosida}.

An immediate consequence  is that under some
circumstance, establishing a bound on $\Ja[m,w;d]$ delivers a solution
of the inverse problem in the sense of the Problem Statement,
above.  This conclusion follows from a simple fact about quadratic
penalty functions posed in terms of the spectral resolution of the
penalty operator:

\begin{theorem}
  \label{thm:minsol}
  Suppose $W$ and $D$ are Hilbert spaces, $S \in {\cal B}(W,W)$, $R
  \in {\cal B}(W,D)$, $d \in D, w \in W$, and $\alpha, \epsilon>0, \gamma >
  1$. Suppose $\lambda_{\rm min} \le \lambda_{\rm max}$ and that
  $\{\lE: \lambda \in [\lambda_{\rm min},\lambda_{\rm max}]\}$ is the
  spectral resolution of $\sqrt{S^TS}$. Then for any $\lambda \in  [\lambda_{\rm min},\lambda_{\rm max}]$,
  \begin{equation}
    \label{eqn:gentol}
    \|R\lE w-d\| \le \epsilon\|d\|
  \end{equation}
  provided that 
  \begin{equation}
    \label{eqn:epslam}
    \epsilon \ge \gamma \frac{(\|Rw-d\|^2 +
      \alpha^2\|Sw\|^2)^{1/2}}{\|d\|}, \,\,\lambda \ge 
    \frac{\|R\|}{\alpha (\gamma-1)}
  \end{equation}
\end{theorem}
\begin{proof}
First note that
\[
  \|Sw\|^2 = \|S(\lE w + (I-\lE)w)\|^2 =
  \langle \lE w + (I-\lE)w, S^TS (\lE w + (I-\lE)w)\rangle
\]
\[
  =  \langle \lE w + (I-\lE)w, (\lE w + (I-\lE) S^TS w)\rangle
\]
\[
  = \langle \lE w,\lE S^TSw \rangle+ \langle(I-\lE)w,(I-\lE) S^TS w\rangle
\]
(since $\lE$ is a projector)
\[
\ge  \langle (I-\lE)w,S^TS (I-\lE)w \rangle
\]
\[
  =\langle w, S^TS (I-\lE)w\rangle
\]
(since $(I-\lE)$ commutes with $S^TS$)
\[
  =\int_{\lambda}^{\lambda_{\rm max}} \mu^2 d\langle w,E_{\mu}w\rangle 
\]
\[
  \ge \lambda_{\rm min}^2 \int_{\lambda}^{\|A\|} d\langle w,E_{\mu}w\rangle
\]
\begin{equation}
  \label{eqn:lb}
  =\lambda^2\langle w, (I-\lE)w\rangle   =\lambda^2\|(I-\lE)w\|^2
\end{equation}

Set
\[
  \delta =  \frac{(\|Rw-d\|^2+\alpha^2\|Sw\|^2)^{1/2}}{\|d\|}.
\]
For any $\lambda >0$,
\[
  \alpha \lambda \|(I-\lE)w\| \le \alpha \|Sw\| \le \delta \|d\|^2
\]
whence
\[
  \|R\lE w-d\| \le \|Rw-d\|+\|R(I-\lE)w\|
\]
\[
  \le \delta\|d\| + \|R\|\frac{\delta}{\alpha \lambda}\|d\|
\]
\[
  \le \left(1 + \frac{\|R\|}{\alpha\lambda}\right)\delta\|d\|
\]
The presumed bounds \ref{eqn:epslam} on $\epsilon$ and $\lambda$ imply that this is
\[
  \le \gamma \delta \|d\| \le \epsilon\|d\|.
\]
\end{proof}

In the context of the single trace transmission problem, the operators

\begin{corollary}
  Suppose that $m \in M, d\in D, \gamma>1, \epsilon>0, \lambda >0, \alpha
  > 0$ satisfy
    \begin{equation}
    \label{eqn:epslam}
    \epsilon \ge \gamma \frac{(2\tJa[m;d])^{1/2}}{\|d\|}, \,\,\lambda \ge 
    \frac{\|F[m]\|}{\alpha (\gamma-1)}
  \end{equation}
  Then $(m,\lE \aw[m;d])$ is a solution of the inverse problem as stated
  in condition \ref{eqn:probstat0}, that is, $\lE\aw[m;d] \in \lW$ and
  \[
    e[m,\aw[m;d];d] \le \frac{\epsilon^2}{2} \|d\|^2
  \]
\end{corollary}

That is, the minimizer of $\Ja$ is a solution of the constrained
problem, as stated in the Problem Statement \ref{eqn:probstat0}, for
$\lambda = O(1/\alpha)$, provided that the target noise level
($\epsilon$) is higher by a fudge factor $\gamma>1$ than the minimum
value of $\Ja$.

It remains to establish how

\section{The Discrepancy Principle}
As described in the introduction, the discrepancy principle (in one of
its guises) involves setting an acceptable range of data misfit
$[e_-,e_+]$, and adjusting the penalty weight $\alpha$ in the reduced
objective definition \ref{eqn:red} so that the data error term at the
inner solution, $e[m,\aw[m;d]]$, lies in this range. Since updating
$m$ changes the inner problem, the appropriate condition for such
separable problems is that $e$ {\em stays} in the range $[e_-,e_+]$ as
$m$ is updated. In this section, we examine the dependence of $e$
on $\alpha$ with a view to understanding how to reset $\alpha$ when
$m$ changes.

Accordingly, regard $m$ as fixed and suppress it from the notation for the remainder of this section, and introduce the abbreviations
\begin{eqnarray}
\label{eqn:erralph}
e(\alpha) &=& e[m,\aw[m;d]]\\
\label{eqn:penalph}
p(\alpha) &=& p[m,\aw[m;d]]
\end{eqnarray}
Since the normal matrix $F^T F + \alpha^2 A^T A$ has been assumed
invertible (condition \ref{eqn:nopco}), the solution of $\aw$ of the
normal equation \ref{eqn:condition} is smooth in $\alpha^2$.
Differentiating equation \ref{eqn:condition} with respect to  $\alpha^2$ leads to the relation
\begin{equation}
(F^T F + \alpha^2 A^T A ) \frac{dw}{d\alpha^2} = -A^T A w
\label{eqn:dnorm}
\end{equation}
whence
\begin{align}
\frac{de}{d\alpha^2} 
&=\left\langle\frac{dw}{d\alpha^2},F^T(Fw-d) \right\rangle \nonumber \\
&=-\alpha^2\left\langle\frac{dw}{d\alpha^2},A^TAw\right\rangle \nonumber \\ 
&=\alpha^2 \langle A^TAw,(F^TF + \alpha^2 A^TA)^{-1}A^TAw\rangle \nonumber \\
&\ge 0
\label{eqn:de}
\end{align}
Note that the inequality in equation \ref{eqn:de} is {\em strict}  if $p > 0$ hence $A^TAw \ne 0$, since the normal operator is assumed to be positive definite.
The derivative of $p(\alpha^2)$ with respect to $\alpha^2$ is
\begin{align}
\frac{dp}{d\alpha^2} &=  -\langle A^T Aw,(F^TF + \alpha^2 A^TA)^{-1}A^TAw \rangle \nonumber \\
&\leq 0
\label{eqn:dp}
\end{align}
similarly a strict inequality if $p > 0$.

Equations \ref{eqn:de} and \ref{eqn:dp} show that increasing $\alpha^2$ implies increasing $e$ while decreasing $p$, and
\begin{align}
&\langle A^TA w,(F^TF + \alpha^2 A^TA)^{-1} A^TA w \rangle \nonumber \\ 
=& \langle (A^TA)^{1/2}w,[(A^TA)^{-1/2}F^TF(A^TA)^{-1/2} + \alpha^2 I]^{-1}(A^TA)^{1/2}w \rangle \nonumber \\ 
\le& \frac{1}{\alpha^2} \langle A^TA w, w\rangle = \frac{2}{\alpha^2}p
\end{align}
In view of equation \ref{eqn:de},
\begin{equation}
\label{eqn:lep}
\frac{de}{d\alpha^2}  \le 2p.
\end{equation}
with this inequality also being strict if $p > 0$.

Suppose the current weight is $\alpha^2_c$ and denote a candidate for an updated weight by $\alpha^2_+$. Then from inequality \ref{eqn:lep},
\begin{equation}
e(\alpha^2_+)-e(\alpha^2_c) \le \int_{\alpha^2_c}^{\alpha^2_+} 2p d\alpha^2
\end{equation}
If $\alpha^2_+ \ge \alpha^2_c$, then in view of inequality \ref{eqn:dp}, the above is
\begin{equation}
\label{eqn:basic}
\le 2 p(\alpha^2_c) (\alpha^2_+-\alpha^2_c)  
\end{equation}

Let us suppose that $e(\alpha^2_c) < e_{+}$. Then setting 
\begin{equation}
\label{eqn:alphasecant}
\alpha^2_+ = \alpha^2_c + \frac{e_{+}-e(\alpha^2_c)}{2p(\alpha^2_c)} 
\end{equation}
implies via inequality \ref{eqn:basic} that 
\begin{equation}
e(\alpha^2_+)-e(\alpha^2_c) \le e_{+}-e(\alpha^2_c)
\end{equation}

Assuming that $p(\alpha^2_+) > 0$, hence $p(\alpha^2)>0$ for $\alpha^2_c \le \alpha^2 \le \alpha^2_+$,  we conclude that if $\alpha^2_+$ is given by the rule \ref{eqn:alphasecant},
then
\begin{equation}
\label{eqn:assert}
e(\alpha^2_c) < e(\alpha^2_+) \le e_{+}
\end{equation}
That is, unless $p(\alpha^2_+) = 0$ (in which case a physical solution of the inverse problem has been reached),  $e(\alpha^2_+)$ is larger than $e(\alpha^2_c)$ but in any case does not exceed $e_+$. The rule \ref{eqn:alphasecant} therefore provides a feasible updated $\alpha^2$ consistent with the upper bound in the discrepancy principle.

\section{stuff}
In many cases, the constraint posed by membership of the second component $w$ in the subspace $\lW$ is an approximation to an idealized limit $\lambda \rightarrow 0$.
It may be that $W_0 = \cap_{\lambda > 0}\lW$ is non-trivial, in which
case $F$ may be viewed as an extension of $F_0 = F|_{ M \times
  W_0}$. In that case, $F_0$ and $F$ realize the idealized extension
structure discussed in many prior works on extended inversion (for
example, \cite[]{geoprosp:2008}): $F_0$ is the physical modeling
operator, and $W_0$ consists of second components of physical or
feasible models. However, in many of these examples, $W_0$ is not
computationally accessible, or - even if it is - gives results that
closely approximate those obtained with $\lW$ for small $\lambda$. For
other examples, for instance that  discussed later in this paper (and
implicitly in \cite[]{Warner:16}, for instance), $W_0=\{0\}$ and the
family $\{\lW: \lambda > 0\}$ actually embodies all of the necessary
physics.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2020-06-15 repro/papers/discrep/vpmloss.tex
getting there

\section{Variable projection method with loss of derivative}

While minimization of $J_{\alpha}$ might be tackled directly - by
alternately minimizations between $m$ and $w$, or by computing updates
for $m$ and $w$ simultaneously - such joint mimization performs
poorly, as \cite{YinHuang:16} has shown. The reason for this poor
performance is that $J_{\alpha}$ has dramatically different
sensitivity to $m$ versus $w$,
as will be explained below.
The separable nature of the least-squares inverse problems defined in \ref{eqn:obj} invites use of the variable projection method, a nested optimization approach. First, in the inner loop, the objective function is optimized over linear parameter $w$ with the nonlinear parameter $m$ fixed. 
The gradient of the objective function $\Ja[m, w]$ with respect to $w$ is
\begin{equation}
\nabla _{w} \Ja[m, w;d] =F[m]^T (F[m] w - d) 
+ \alpha A^T A  w 
\label{eqn:grad1}
\end{equation}
where $T$ denotes transpose.  A stationary point of equation \ref{eqn:obj} satisfies the {\em normal equation}
\begin{equation}
(F[m]^T F[m] + \alpha A^T A ) w = F[m]^T d.
\label{eqn:condition}
\end{equation}
In view of Assumption A above, the Lax-Milgram Theorem (\cite{Yosida}, p. 92)
applies to show that the system \ref{eqn:condition} has a
unique solution. Since $m$, $d$ and $\alpha$ determine the operator on the LHS
of equation \ref{eqn:condition} and its RHS, its solution becomes a
function $\aw[m;d]$ of these quantities. Continuity in $d$ is a
consequence of the Lax-Milgram result.

As observed by \cite{GolubPereyra:73,GolubPereyra:03}, $\Ja$ may be minimized over $m,w$ by minimizing the reduced objective
\begin{equation}
  \label{eqn:red}
  \tJa[m; d] = \Ja[m,\aw[m;d];d]
\end{equation}
over $m$. It is natural, and the goal of this paper, to use local,
Newton-like methods to minimize. However such
methods require at least a gradient, and therefore derivatives of $F$,
and so far I have not specified any regularity properties of $F$.

In fact, if the theory is to encompass inverse problems in wave
propagation in which wave velocities are amongst the model parameters
to be estimated, then $F$ cannot be presumed to be differentiable, or
even locally uniformly continuous, as a ${\cal B}(W,D)$-valued
function on $M$. Instead, it is {\em differentiable with loss of
  regularity}, as I will now explain.

The mapping properties of typical modeling operators $F$ may be
described with the help of a {\em scales of Hilbert spaces}
$\{W^s, D^s\}_{s=0}^{\infty}$, with the following properties:

\noindent {\bf Assumption C:}
\begin{itemize}
\item $\{W^s\}_{s=0}^{\infty}$ and $\{D^s\}_{s=0}^{\infty}$ are {\em decreasing}, that is,
  $W^{s+1} \subset W^s$ for $s \ge 0$, and similarly for $\{D^s\}_{s=0}^{\infty}$;
\item denoting by $\|\cdot\|_s$ the norm in $W^s$, $\|w\|_t \ge
  \|w\|_s$ if $t \ge s$ and $w \in W^t$, with a similar property for $\{D^s\}_{s=0}^{\infty}$;
\item $W^0=W$, $D^0=D$;
\item 
  \begin{equation}
    \label{eqn:reg0}
    F[m] \in {\cal B}(W^s,D^s),\,F[m]^T \in {\cal B}(D^s,W^s) \mbox{
      for any } s\ge 0, m\in M
  \end{equation}
  (here $F[m]^T$ denotes the adjoint of $F[m] \in {\cal B}(W^0,D^0)$,
  that is, $\langle F[m]w,d \rangle_{D^0} = \langle w,F[m]^T
  \rangle_{W^0}$ for $w \in W^0=W, d \in D^0=D$);
\item $A \in {\cal B}(W^s,W^s)$ for any $s \ge 0$;
\item For each $\alpha > 0$, 
  \begin{equation}
    \label{eqn:normsm}
    F^TF+\alpha^2A^TA \in C^{\infty}(M, {\cal I}(W^s,W^s)).
  \end{equation}
  There exists $C_{s,\alpha}>0$ so  that for all
    $m\in M$, 
    \begin{equation}
      \label{eqn:nopco}
      \|(F[m]^TF[m]+\alpha^2A^TA)^{-1}\|_{{\cal B}(W^s,W^s)} \le
      C_{s,\alpha}.
    \end{equation}
\item 
  \begin{eqnarray}
    F &\in& C^k(M, {\cal B}(W^s,D^t)) \mbox{ provided that }s > t+k, t, 
    k \ge 0;\\
    \label{eqn:reg1}
    F^T &\in& C^k(M, {\cal B}(D^s,W^t)) \mbox{ provided that }s > t+k, t, 
    k \ge 0;
    \label{eqn:reg2}
  \end{eqnarray}
\end{itemize}
Note that the inequality of indices in the condition \ref{eqn:reg1} is {\em
  strict}. In particular, $F$ is not assumed to be continuous as a map
from $M$ to ${\cal B}(W^s,D^s)$ for any $s$. This constraint is
necessary, as even the simplest examples show. For instance, the
example considered later in this paper has this property, as I will
show explicitly. On the other hand,
\cite{BlazekStolkSymes:13} show that versions of conditions
\ref{eqn:reg0}, \ref{eqn:reg1} hold in
general for mappings defined by symmetric hyperbolic
integro-differential systems, encompassing most common physical models
of wave propagation.

Note also that, despite the relatively weak differentiability properties of $F$, as
indicated in condition \ref{eqn:reg1}, its normal operator is actually
smooth as a ${\cal I}(W^s,W^s)$-valued function for all $s\ge 0$
(condition \ref{eqn:nopco}). This condition also abstracts properties
common to many (though not all) extended inversion frameworks. It 
certainly holds for the example described in this paper.

The scales appearing in the examples that motivated Assumption C
calibrate regularity: that is, members of $W^t$ are more regular as
functions on space-time than members of $W^s$ if $t>s$. Thus the
condition \ref{eqn:reg1} expresses differentiability (of $F$ as a
function on $M$) with loss of regularity (of the arguments as
functions on space-time).

A first consequence of Assumption C is that the solution $\aw[m;d]$ of
the normal equation \ref{eqn:condition} is a continuously differentiable $W^s$-valued
function on $M \times D^{s+2}$ for any $s \ge 0, \alpha>0$. Indeed, condition
\ref{eqn:nopco} implies that the inverse of the normal operator is a
smooth ${\cal B}(W^s,W^s)$-valued function on $M$, and condition
\ref{eqn:reg2} that the RHS of the normal equation \ref{eqn:condition}
depends differentiably on $(m,d) \in M \times D^t$ so long as $t>s+1$.

From its definition
the definition of $\Ja$ (\ref{eqn:obj}), and the conclusion of the
preceding paragraph, conclude that $\tJa[m;d] \in C^1(M \times D^4)$.
Indeed, $\aw \in C^1(M \times D^4,W^2)$ from the preceding paragraph,
and the data residual $(m,w,d) \mapsto F[m]w-d$ is in $C^1(M \times
W^2 \times D^0)$ according to condition \ref{eqn:reg1}.

The main result of \cite[]{GolubPereyra:73} is a formula for the
gradient of the reduced objective \ref{eqn:red}.  Justification of this
formula in the present context rests on the mapping properties of the
derivative $DF$, which are naturally posed in
terms of the tangent space of the domain of $F$.  Recall that $M$ is
an open subset of a Hilbert space, which has so far gone
un-named. This ambient Hilbert space can be identified with the
tangent space $T_mM$ of $M$ at any point $m \in M$, and these tangent
spaces are all isomophic to the ambient space. Taking advantage of
this fact, I will drop the subscript and refer to $TM (=T_mM$ for any
$m \in M)$ as the tangent space of $M$. Thus assumption \ref{eqn:reg1}
implies that for $m \in M$, $DF[m] \in {\cal B}(TM,{\cal B}(W,D))$.

A particular instance of \ref{eqn:reg1} is the assertion that $F \in
C^1(M,{\cal B}(W^2,D^0))$. Consequently for $m \in M$, $(\delta m,w)
\rightarrow (DF[m]\delta m)w$ is continuous and bilinear $:TM \times
W^2 \rightarrow D^0$. The derivative of $\tJa$ is
\[
    D\tJa[m;d]\delta m = \langle 
  (DF[m]\delta m)\aw[m,d],F[m]\aw[m;d]-d\rangle_{D^0}
\]
\begin{equation}
  \label{eqn:grad1}
 +\langle D\aw[m;d]\delta m,F[m]^T(F[m]\aw[m;d]-d)+\alpha^2A^TA\aw[m;d]\rangle_{W^0}
\end{equation}
for any $\delta m \in TM$, provided that the derivatives appearing on
the RHS make sense, which they do if $d \in D^4$, as noted above. Note
that the second term on the RHS of equation \ref{eqn:grad1} vanishes
thanks to the normal equation \ref{eqn:condition}.

Given $w \in
W^2$, the adjoint of the continuous linear map $T_mM \rightarrow D^0$
given by $\delta m \rightarrow
(DF[m]\delta m)w$ will be denoted $DF[m]^*(w,\cdot)$: that is,
\begin{equation}
  \label{eqn:madj}
  \langle (DF[m]\delta m)w,d\rangle_{D^0} = \langle \delta m,
  DF[m]^*(w,d)\rangle_{TM}.
\end{equation}

With this convention, the result \ref{eqn:grad1} can be re-written
\begin{equation}
  \label{eqn:deriv}
  D\tJa[m;d]\delta m = \langle \delta m, DF[m]^*(\aw[m;d], F[m]\aw[m;d]-d)\rangle_{TM}
\end{equation}
Therefore

\begin{theorem}\label{eqn:vpmgrad}
  Under Assumptions A, B, and C,  $\tJa[\cdot;d] \in C^k(M)$ provided
  that $k \ge 1$ and $d \in D^{3+k}$. Its gradient is given by
 \begin{equation}
   \nabla _{m} \tJa[m; d] = D F[m]^*(\aw[m;d],F[m]\aw[m;d]-d))
   \label{eqn:grad2}
 \end{equation}
 \end{theorem}
The identity \ref{eqn:grad2} has precisely the form of the formula derived
\cite{GolubPereyra:03,GolubPereyra:73} for differentiable separable
least squares problems on $\bR^n$. The existence of higher derivatives
than the first under the circumstances indicated follows from very
similar reasoning to that used to establish the assertion for $k=1$.

### 2020-07-10 repro/papers/discrep/paper.tex
interim

@@
 \title{A Discrepancy-Based Algorithm for Extended Inversion}
-\author{William. W. Symes \thanks{P. O. Box 43, Orcas WA USA, email {\tt symes@rice.edu}.}}
+\author{William W. Symes, Orcas Island, Huiyi Chen* and Susan E. Minkoff, The University of Texas at Dallas}
 
@@
   max}])$ respectively (that is, interpret the definition \ref{eqn:mod}
-as holding for $t \in [t_{\rm min},t_{\rm  max}]$).
+as holding for $t \in [t_{\rm min},t_{\rm  max}]$). So
+\begin{equation}
+  \label{eqn:mapprop}
+  \mbox{for }m \in M, F[m] \in {\cal B}(W,D),\mbox{ and }\|F[m]\| =
+  \frac{1}{4\pi r}.
+\end{equation}
 
@@
 
-The support constraint is closely linked to the folk theorem about FWI noted many
-times in the literature: convergence of a descent method requires that
-the initial slowness must be known to ``within a (fraction of a) wavelength''. The
-relation is a consequence of Heisenberg's inequality
-(see for example \cite{Folland:07}, p. 255). Assume that $w \in L^2(\bR)$, and that both $tw$ and derivative $dw/dt$ are
-square-integrable (an essential additional assumption, as will be
-explained below). Then for any $t_0, k_0 \in \bR$,
-\[
-  \left(\int\,dt\, (t-t_0)^2|w(t)|^2\right)
-  \left(\int\,dk\,(k-k_0)^2|\hat{w}(k)|^2\right)
-  \ge\frac{\|w\|^4}{16 \pi^2}
-\]
-in which $\hat{w}$ denotes the Fourier transform of $w$. Rearrange
-slightly and choosing $t_0=0, k_0=0$,
-\[
-  \left(\int\,dt\, t^2\frac{|w(t)|^2}{\|w\|^2}\right)
-  \left(\int\,dk\,k^2\frac{|\hat{w}(k)|^2}{\|\hat{w}\|^2}\right)
-  \ge\frac{1}{16 \pi^2}
-\]
-If $w \in \lW$, then the first factor is $\le \lambda^2$, whereas the
-second factor is the power spectrum mean square estimate of
-frequency squared $k^2$. That is the power spectrum RMS estimate of
-frequency is at least $1/(4 \pi \lambda)$. In the sense of the RMS
-spectral estimator of frequency, then, $\lambda$ is proportional to an upper
-bound on representative wavelength: substantial energy must be present
-at frequencies larger than $1/\lambda$.
-
-The best case for data fitting is clearly the one in which the data
-can be fit precisely: that is, there exists $(m_*,w_*) \in M\times \lW$ so that
+\noindent{\bf Remark:} The support constraint is closely linked to the
+folk theorem about FWI noted many times in the literature: convergence
+of a descent method requires that the initial slowness must be known
+to ``within a (fraction of a) wavelength''. The relation is a
+consequence of Heisenberg's inequality, and will be reviewed later in
+the paper.
+
+The best case for data fitting
+is clearly the one in which the data can be fit precisely: that is,
+there exists $(m_*,w_*) \in M\times \lW$ so that
 \begin{equation}
@@
 $A$. The role of $A$ is to penalize distance from the constrained
-subspace $\lW$, in some sense.
-
-I will introduce two options for $A$. Both are scalar 
+subspace $\lW$, in some sense. All choices for $A$ considered here are scalar 
 multiplication operators defined by a choice of $a \in L^{\infty}(\bR)$:
@@
 
-An
-operator $A_{\lambda}$ of which $\lW$ is the
-null space would be a natural 
-choice. Such operators have come to be called ``annihilators'', since
-they map all members of the constraint subspace $\lW$ to zero.
-Define 
+With these choices, the normal equation \ref{eqn:norm1} becomes
+\begin{equation}
+\label{eqn:norm2}
+\left(\frac{1}{(4\pi r)^2}  {\bf 1}_{[t_{\rm min}-mr,  
+      t_{\rm max}-mr]} + \alpha^2a^2\right)w= F[m]^Td.
+\end{equation}
+
+\begin{theorem}
+  \label{thm:norminv}
+  Assume the conditions \ref{eqn:mod}, \ref{eqn:fullrec},
+  \ref{eqn:annmult}. Also assume that $\lambda \in (0,\lambda_{\rm
+    max}], \alpha > 0,$ and that $a \in L^{\rm infty}(\bR)$
+  mentioned in condition \ref{eqn:annmult} satisfies
+  \begin{equation}
+    \label{eqn:abnd}
+    |t| > \lambda \Rightarrow a(t) \ge C,
+  \end{equation}
+  in which $C>0$ may depend on $\lambda$.
+  Then
+  \begin{itemize}
+  \item[1. ]the normal operator $F[m]^TF[m] + \alpha^2A^TA$ is
+    invertible for any $m \in M$, $\alpha > 0$;
+  \item[2. ]for any $d \in D$, the normal equation \ref{eqn:norm} has
+    a unique solution $\aw[m;d]\in W$, and 
+    \begin{equation}
+      \label{eqn:normsol}
+      \aw[m;d](t) = \left\{
+        \begin{array}{c}
+          \left(\frac{1}{(4\pi r)^2} + \alpha^2
+          a^2(t)\right)^{-1}\frac{1}{4 \pi r}d(t+mr), t \in [t_{\rm
+          min}-mr, t_{\rm max}-mr];\\
+          0, \mbox{ else;}
+        \end{array}
+      \right.
+    \end{equation}
+  \item[3. ]if in addition $d=F[m_*]w_*, w_* \in \lW$ is noise-free, as in equation
+    \ref{eqn:defdatanonoise},
+    \begin{equation}
+      \label{eqn:solnnongen}
+      \aw[m,d](t)= \left(1+ (4\pi r)^2\alpha^2 a(t)^2\right)^{-1}w_*\left(t+(m-m_*)r\right).
+    \end{equation}
+  \end{itemize}
+\end{theorem}
+
+\begin{proof}
+  \begin{itemize}
+  \item[1. ]Note that thanks to \ref{eqn:zeroinc}, if $|t|\le
+    \lambda \le \lambda_{\rm max}$, then ${\bf 1}_{[t_{\rm min}-mr,  
+      t_{\rm max}-mr]}(t) = 1$, whereas if $|t|>\lambda$,
+    then $a(t) \ge C$, whence
+    \[
+      \frac{1}{(4\pi r)^2}  {\bf 1}_{[t_{\rm min}-mr,  
+        t_{\rm max}-mr]} + \alpha^2a^2  \ge \min\{(4\pi r)^2,
+      \alpha^2\min\{1/(4\pi r)^2,C^2\} \}> 0.
+    \]
+    Therefore the normal operator is invertible under the stated
+    conditions.
+
+  \item[2. ]From the identity \ref{eqn:tran}.
+    \[
+      \mbox{supp }F[m]^Td \subset [t_{\rm min}-mr,t_{\rm max}-mr].
+    \]
+    Define $w_{\rm tmp}$ to be the right-hand side of equation \ref{eqn:normsol}. Then
+    from the previous observation and identity \ref{eqn:tran},
+    \[
+      \mbox{supp }w_{\rm tmp} \subset [t_{\rm min}-mr,t_{\rm max}-mr].
+    \]
+    From the identity \ref{eqn:unit}, for any $w \in W$,
+    \[
+      t \in [t_{\rm min}-mr,t_{\rm max}-mr] \Rightarrow F[m]^TF[m]w(t)
+      = \frac{1}{(4 \pi r)^2}w(t).
+    \]
+    It follows from this and the previous two observations that
+    $w_{\rm tmp}$ solves the normal equation \ref{eqn:norm}, and
+    therefore that $\aw[m;d]=w_{\rm tmp}$.
+
+  \item[3. ]Follows by inserting the definition
+    \ref{eqn:defdatanonoise} of $d$ in \ref{eqn:normsol} and
+    rearranging.
+  \end{itemize}
+\end{proof}
+
+Since the normal operator is invertible, the reduced
+objective $\tJa$ is given by the expression \ref{eqn:objexp}.
+The results of the previous theorem imply those of
+
+\begin{theorem}
+  \label{thm:epjnonoise}
+  Assume the hypotheses of Theorem \ref{thm:norminv}, item 3. Then 
+\begin{equation}
+  \label{eqn:residnorm}
+  e[m,\aw[m,d];d] 
+= 8 \pi^2 r^2 \alpha^4\int\,dt\,a(t-(m-m_*)r)^4(1+(4\pi r)^2 \alpha^2 
+    a(t-(m-m^*)r)^2)^{-2}w_*(t)^2.
+\end{equation}
+\begin{equation}
+  \label{eqn:anninorm}
+  p[m,\aw[m,d];d] = \frac{1}{2}\int \,dt\,  a(t-(m-m_*)r)^2 \left(1+ (4\pi r)^2\alpha^2
+    a(t-(m-m_*)r)^2\right)^{-2}w_*(t)^2.
+\end{equation}
+so finally
+\begin{equation}
+\label{eqn:expjgen}
+\tJa[m;d] = =\frac{\alpha^2}{2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
+  a(t-(m-m^*)r)^2)^{-1}w_*(t)^2.
+\end{equation}
+\end{theorem}
+
+I willl now examine two possible choices of penalty multipler $a$ in
+defintion \ref{eqn:annmult}.
+
+A penalty operator $A$ of which $\lW$ is the null space would be a
+natural choice. Such operators have come to be called
+``annihilators'', since they map all members of the constraint
+subspace $\lW$ to zero.  A simple example is
 \begin{eqnarray}
@@
 orthocomplement, an operator of the form \ref{eqn:annmult} with $a
-= 1 - {\bf 1}_{[-\lambda,\lambda]}$.
-
-A second possible
-penalty operator penalizes energy away from
-$t=0$: choose $\tau > 0$ and set
-\begin{equation}
-  \label{eqn:ann}
-  a(t) = \min(|t|, \tau).
-\end{equation}
-The cutoff $\tau$ will be chosen large enough to be effectively inactive:
-specifically, hindsight suggests
-\begin{equation}
-  \label{eqn:taudef}
-  \tau = \max \{t_{\rm min}-m_{\rm max}r, t_{\rm max}-m_{\rm min}r\}.
-\end{equation}
-This
-particular annihilator has been employed in earlier papers on extended
-source inversion
-\cite[]{Plessix:00a,LuoSava:11,Warner:14,HuangSymes:SEG15a,Warner:16,HuangSymes:GEO17}.
-
-With these choices, the normal equation \ref{eqn:norm1} becomes
-\begin{equation}
-\label{eqn:norm2}
-\left(\frac{1}{(4\pi r)^2}  {\bf 1}_{[t_{\rm min}-mr,  
-      t_{\rm max}-mr]} + \alpha^2a^2\right)w= F[m]^Td.
-\end{equation}
-
-\begin{theorem}
-  \label{thm:norminv}
-  With the assumptions \ref{eqn:mod}, \ref{eqn:fullrec},
-  \ref{eqn:annmult} and either choice of penalty multiplier $a$
-  described above,
-  \begin{itemize}
-  \item[1. ]The normal operator $F[m]^TF[m] + \alpha^2A^TA$ is
-    invertible for any $m \in M$, $\alpha > 0$;
-  \item[2. ] For any $d \in D$,
-    \begin{equation}
-      \aw[m;d](t) = \left\{
-        \begin{array}{c}
-          \left(\frac{1}{(4\pi r)^2} + \alpha^2
-          a^2(t)\right)^{-1}\frac{1}{4 \pi r}d(t+mr), t \in [t_{\rm
-          min}-mr, t_{\rm max}-mr];\\
-          0, \mbox{ else.}
-        \end{array}
-      \right.
-    \end{equation}
-  \item[3. ] If $d$ is noise-free, as in equation
-    \ref{eqn:datadefnonoise},
-    \begin{equation}
-      \label{eqn:solnnon}
-      \aw[m,d](t)= \left(1+ (4\pi r)^2\alpha^2 a(t)^2\right)^{-1}w_*\left(t+(m-m_*)r\right) 
-    \end{equation}
-  \end{itemize}
-\end{theorem}
-
-Sincee normal operator is invertible, the reduced
-objective $\tJa$ is given by the expression \ref{eqn:objexp}.
-Employing the expression \ref{eqn:tran} for the action of $F[m]^T$,
-the definition \ref{eqn:defdatanonoise} of noise-free data,
-and the realization \ref{eqn:norm2} of the normal operator as a
-multiplication operator, obtain after some simplification
-
-\begin{equation}
-  \label{eqn:residnorm}
-  e[m,\aw[m,d];d] 
-= 8 \pi^2 r^2 \alpha^4\int\,dt\,a(t-(m-m_*)r)^4(1+(4\pi r)^2 \alpha^2 
-    a(t-(m-m^*)r)^2)^{-2}w_*(t)^2.
-\end{equation}
-\begin{equation}
-  \label{eqn:anninorm}
-  p[m,\aw[m,d];d] = \frac{1}{2}\int \,dt\,  a(t-(m-m_*)r)^2 \left(1+ (4\pi r)^2\alpha^2
-    a(t-(m-m_*)r)^2\right)^{-2}w_*(t)^2.
-\end{equation}
-so finally
-\begin{equation}
-\label{eqn:expj}
-\tJa[m;d] = =\frac{\alpha^2}{2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
-  a(t-(m-m^*)r)^2)^{-1}w_*(t)^2.
-\end{equation}
+= 1 - {\bf 1}_{[-\lambda,\lambda]}$. Note that this choice of $a$
+satisfies condition \ref{eqn:abnd} with $C \equiv 1$.
 
@@
   For $A=E^c_{\lambda}$, that is, $a=1-{\bf 1}_{-\lambda,\lambda]}$ in
-  the definition \ref{eqn:expj}. Then if $|m-m_*| >  2\lambda/r$,
+  the definition \ref{eqn:expjgen}. Then if $|m-m_*| >  2\lambda/r$,
   \[
@@
 
-Surprisingly, the choice $A=S$ yields an optimization
-problem has better global convergence properties:
+A second possible 
+penalty operator penalizes energy away from 
+$t=0$: choose $\tau > 0$ and set 
+\begin{equation}
+  \label{eqn:ann}
+  a(t) = \min(|t|, \tau). 
+\end{equation}
+The cutoff $\tau$ will be chosen large enough to be effectively inactive: 
+specifically, hindsight suggests 
+\begin{equation}
+  \label{eqn:taudef}
+  \tau = \max\{|t_{\rm min}-m_{\rm min}r|,|t_{\rm min}-m_{\rm max}r|, |t_{\rm max}-m_{\rm min}r|, |t_{\rm max}-m_{\rm max}r|\}. 
+\end{equation}
+This 
+particular annihilator has been employed in earlier papers on extended 
+source inversion 
+\cite[]{Plessix:00a,LuoSava:11,Warner:14,HuangSymes:SEG15a,Warner:16,HuangSymes:GEO17}. It
+satisfies condition \ref{eqn:abnd} with $C=\lambda$.
+
+Surprisingly, the choice \ref{eqn:ann} yields an optimization
+problem with better global convergence properties:
 
@@
   \label{thm:rampgood}
-  For $A=S$, that is, $a(t)=t$ in the definition \ref{eqn:expj}, all
+  For $a(t)=\min\{|t|,\tau\}$ in the definition \ref{eqn:annmult},
+  with $\tau$ given by equation \ref{eqn:taudef}, all
   stationary points of $\tJa$ lie in the interval
@@
 \begin{proof}
+  As observed before, $\mbox{supp }\aw[m;d] \subset [t_{\rm
+    min}-mr,t_{\rm max}-mr]\subset [-\tau,\tau]$, with $\tau$ defined
+  in \ref{eqn:taudef}. Therefore for $t \in  [t_{\rm min}-mr,t_{\rm
+    max}-mr]$, $a(t) = |t|$, and reason analogous to the proof of
+  Theorem \ref{thm:norminv}, item 2, shows that
+  \begin{equation}
+    \label{eqn:solnnon}
+    \aw[m,d](t)= \left(1+ (4\pi r)^2\alpha^2 t^2\right)^{-1}w_*\left(t+(m-m_*)r\right) 
+  \end{equation}
+  Straightforward modification of the proof of Theorem
+  \ref{thm:epjnonoise} gives
+  \begin{equation}
+    \label{eqn:expj}
+    \tJa[m;d] =\frac{\alpha^2}{2}\int\,dt\,(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
+    (t-(m-m^*)r)^2)^{-1}w_*(t)^2.
+  \end{equation}
   A brief calculation yields
@@
     \label{eqn:gradfinal}
-\frac{\partial}{\partial m}\tJa[m;d] = -r\alpha^2\int \,dt\,t\left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2\right)^{-2}\frac{1}{(4\pi r)^2}w_*\left(t+(m-m_*)r\right)^2
+    \frac{\partial}{\partial m}\tJa[m;d] = -r\alpha^2\int \,dt\,\frac{t}{(1+(4\pi r)^2\alpha^2 t^2)^2}w_*(t+(m-m_*)r)^2.
   \end{equation}
@@
   \[
- \frac{t}{1+(4\pi r)^2\alpha^2 t^2}
-  \le -(m-m_*)r+\lambda < 0
+    \frac{t}{(1+(4\pi r)^2\alpha^2 t^2)^2}< 0
   \]
@@
   \begin{eqnarray}
-  m > m_*+\lambda/r & \Rightarrow & \frac{\partial \tJa[m;d]}{\partial
-    m}> 0; \nonumber\\
-  m < m_*-\lambda/r & \Rightarrow & \frac{\partial \tJa[m;d]}{\partial
-                                    m}< 0.
+    m > m_*+\lambda/r & \Rightarrow & \frac{\partial \tJa}{\partial
+                                      m}[m;d]> 0; \nonumber\\
+    m < m_*-\lambda/r & \Rightarrow & \frac{\partial \tJa}{\partial
+                                      m}[m;d]< 0.
                                       \label{eqn:leftright}
@@
 
-\noindent {\bf Remark.} That is, {\em with the choice $A=S$, $\tJa$ has no local minima
+\noindent {\bf Remark.} That is, {\em with the choice of penalty
+  multiplier $a$ given in equation \ref{eqn:ann}, $\tJa$ has no local minima
   further than  $O(\lambda)$ from the global minimum:} the gradient
@@
 
+\noindent {\bf Remark.} The estimate $|m-m_*|r<\lambda$ for stationary
+points $m$ of $\tJa$ is sharp: it is possible to choose $w_* \in \lW$
+so that $\lambda - |m-m_*|r$ is as small as you like. In particular,
+the ``exact'' or ``true'' slowness $m_*$ is not necessarily the only 
+stationary point, or even a stationary point at all.
+
+HERE INSERT RESULT: TJA DIFFBL RIGHTARROW A DIFFBLE - relate in
+Discussion to Stolk 03
+
 \section{Extended Inversion}
-The preceding theorem established that a proper choice of annihilator leads to a
-reduced penalty objective all of whose stationary points are within
-$O(\lambda)$ of the target slowness $m_*$, provided that the data are
-noise-free. This result leaves open two questions:
+The preceding theorem established that a proper choice of annihilator
+leads to a reduced penalty objective all of whose stationary points
+are within $O(\lambda)$ of the target slowness $m_*$, provided that
+the data are noise-free in the sense of equation
+\ref{eqn:defdatanonoise}. This result leaves open two questions:
 \begin{itemize}
@@
   a solution of the inverse problem as in problem statement
-  \ref{eqn:probstat0}? There is no reason to think that the minmizer
-  $\bar{m}$ of $\tJa$ with produce $\aw[\bar{m};d] \in \lW$, as
-  required by the problem statement.
+  \ref{eqn:probstat0}? 
 \item how does one answer the same question for noisy data?
@@
 
-A partial resolution of these issues lies in the relation between the
-annihilator $S$ (multiplication by $|t|$) and the family of projectors
-$\{E_{\lambda}: 0 \le \lambda \le \lambda_{\rm max}\}$. In fact, this
-family of operators is a resolution of the identity for the positive
-semidefinite self-adjoint $S$:
+\subsection{Noise-free Data}
+\begin{theorem}
+  \label{thm:ipnonoisesuf}
+  Suppose that $a$ is given by definition \ref{eqn:ann}, $\alpha$,
+  $\mu \in (0,\lambda_{\rm max}]$,
+  $d$ is given by
+  \ref{eqn:defdatanonoise} with $w_* \in W_{\mu}$, and  $m_{\infty}$ is a stationary
+  point of $\tJa[\cdot;d]$. Then $(m_{\infty},\aw[m_{\infty};d])$ is a
+  solution of the inverse problem \ref{eqn:probstat0} for any $\lambda
+  \ge 2\mu$ and $\epsilon \ge (4\pi r \lambda \alpha)^2$.
+\end{theorem}
+
+\begin{proof}
+From the
+inequalities \ref{eqn:leftright}, $|(m-m_*)r|\le \mu$. From the
+identity \ref{eqn:solnnon}, $\mbox{supp }\aw[m_{\infty};d] \subset
+[(m_{\infty}-m_*)r-\mu,(m_{\infty}-m_*)r+\mu] \subset
+[-2\mu,2\mu]$. Because of the support limitation, $a(t)=|t|$
+in the interval of integration appearing in \ref{eqn:residnorm}, so
+\[
+  e[m_{\infty},\aw[m_{\infty},d];d] 
+= 8 \pi^2 r^2 \alpha^4\int^{\mu}_{-\mu}\,dt\,|t-(m_{\infty}-m_*)r|^4(1+(4\pi r)^2 \alpha^2 
+|t-(m_{\infty}-m^*)r|^2)^{-2}w_*(t)^2.
+\]
+and therefore
+\begin{equation}
+  \label{eqn:estresidnorm}
+e[m_{\infty},\aw[m_{\infty};d];d] \le 8 \pi^2 r^2 (2\mu\alpha)^4 \|w_*\|^2 =
+\frac{1}{2}(8 \pi r \mu \alpha)^4 \|d\|^2
+\end{equation}
+\end{proof}
+
+The inequality \ref{eqn:estresidnorm} can be interpreted as a bound 
+on $\alpha$, given $\epsilon$ and $\lambda$, for a
+stationary point of $\tJa$ to yield a solution of the inverse
+problem: one obtains a solution, provided that $\alpha$ is
+sufficiently small. On the other hand, it is clear that $\alpha$
+cannot be too large if stationary points of $\tJa$ are to yield
+solutions: the integrand in \ref{eqn:residnorm} is increasing in
+$\alpha$ for every $t$ and $m$, and the multiplier
+\[
+t \mapsto (4\pi r \alpha(t-(m_{\infty}-m_*)r))^4(1+(4\pi r)^2 \alpha^2 
+|t-(m_{\infty}-m^*)r|^2)^{-2}
+\]
+tends monotonically to $1$ as $\alpha \rightarrow \infty$, uniformly
+on the complement of any open interval containing
+$t=(m_{\infty}-m_*)r$. Therefore
+\begin{equation}
+  \label{eqn:elimit}
+  \lim_{\alpha \rightarrow \infty}e[m_{\infty},\aw[m_{\infty};d];d] =
+  \frac{1}{2}\frac{1}{(4 \pi r)^2}\|w_*\|^2 = \frac{1}{2}\|d\|^2.
+\end{equation}
+Consequently, there exists $\alpha_{\rm max}(\epsilon,\lambda,d)$ so
+that
+\[
+  e[m_{\infty},\aw[m_{\infty};d];d]  \le \frac{1}{2}\epsilon^2\|d\|^2
+  \Rightarrow \alpha \le \alpha_{\rm max}(\epsilon,\lambda,d).
+\]
+The existence of this limiting penalty weight has been inferred
+indirectly; the next section is devoted to a constructive algorithm
+for its approximation.
+
+[THE NEXT BIT MAY OR MAY NOT BE WORTH KEEPING - IT WAS INTERESTING TO
+WORK OUT]
+
+Without further contraints on data or solution, nothing more can be
+said about bounds on $\alpha$. If the data $d$ (and the target wavelet
+$w_*$) are assumed to have a square integrable derivative, then a
+necessary condition follows from the Heisenberg
+inequality (see for example \cite{Folland:07}, p. 255). To formulate
+this result in its most general form, introduce the Hilbert subspaces
+$V^0 \subset L^2(\bR), V^1 \subset H^1(\bR)$:
 \begin{eqnarray}
-  \label{eqn:res1}
-  S &=& \int_{0}^{\|A\|} \lambda d\lE,\nonumber\\
-  S^2=S^TS  &=& \int_{0}^{\|A\|} \lambda^2 d\lE.
+  V^0 & = & \{ f \in L^2(\bR): Af \in L^2(\bR)\}, \nonumber\\
+  V^1 & = & V^0 \cap H^1(\bR).
+            \label{eqn:vdef}
 \end{eqnarray}
-This relation is similar to that explained in the example on p. 313 of \cite{Yosida}.
-
-An immediate consequence  is that under some
-circumstance, establishing a bound on $\Ja[m,w;d]$ delivers a solution
-of the inverse problem in the sense of the Problem Statement,
-above.  This conclusion follows from a simple fact about quadratic
-penalty functions posed in terms of the spectral resolution of the
-penalty operator:
+$V^0$ is the domain of $A$, and equipped with the graph norm of $A$. A
+natural norm in $V^1$ is
+\[
+  \|f\|^2_{V^1} = \|f\|_{V^0}^2 + \|f\|_{H^1}^2.
+\]
+$V^j$ is the completion of $C_0^{\infty}(\bR)$ in the corresponding
+norm, j=0,1.
+
+\begin{proposition}
+  \label{thm:heis}
+For $w \in V^1$,
+  \begin{equation}
+    \label{eqn:heis}
+    \|Aw\|\|w'\| \ge \frac{1}{2}\|w\|^2
+  \end{equation}
+\end{proposition}
+
+\begin{proof}
+  For $w \in C_0^{\infty}(\bR)$,
+  \[
+    \int w^2 = \left|\int\,dt\, t (w(t)^2)' \right|= \left|2\int\,dt\,tw(t)w'(t)\right| \le
+    2\|Aw\| \|w'\|
+  \]
+  by the Cauchy-Schwarz inequality. Since $C_0^{\infty}(\bR)$ is dense
+  in $V^1$, the conclusion follows by continuity.
+\end{proof}
+
+In the conventional formulation of the Heisenberg inequality, the $L^2$ norm of
+$w'$ is replaced by the its equivalent in terms of the Fourier
+transform $\hat{w}$. Adopting temporarily the use of dummy variables
+in the expression of functions, the identity \ref{eqn:heis} turns into
+the usual form of the Heisenberg inequality: for $w \in V^1$,
+\begin{equation}
+\label{eqn:fheis}
+\|tw(t)\|\|k\hat{w}(k)\| \ge \frac{1}{4\pi}\|w\|^2.
+\end{equation}
+
+Define $\krms[w]$, the root mean square estimator of frequency of $w
+\in V^1$, by
+\begin{equation}
+  \label{eqn:krms}
+  \krms[w]=\frac{1}{2\pi}\frac{\|w'\|}{\|w\|} = \left(\int
+    \,dk\,\frac{|\hat{w}(k)|^2}{\|\hat{w}\|^2} k^2\right)^{1/2}.
+\end{equation}
+Then the inequalities \ref{eqn:heis}, \ref{eqn:fheis} can be rewritten as
+\begin{equation}
+  \label{eqn:frmsheis}
+  \|Aw\| \ge \frac{\|w\|}{4\pi \krms[w]}.
+\end{equation}
+
+For $\lambda >0$, define
+\begin{equation}
+  \label{eqn:wlam1}
+  \lW^1 = \lW \cap H^1(\bR).
+\end{equation}
+Note that $\lW^1 \subset V^1$ is a closed subspace for any
+$\lambda>0$.
+
+\begin{proposition}
+  \label{thm:klam}
+  For $w \in \lW^1$,
+  \[
+    \krms[w] \ge \frac{1}{4 \pi \lambda}
+  \]
+\end{proposition}
+
+\begin{proof}
+  Follows directly from inequality \ref{eqn:frmsheis} and the obvious bound
+  $\|A|_{\lW}\|\le \lambda$.
+\end{proof}
+
+\noindent{\bf Remark:} This result is the link mentioned earlier between the support
+constraint and the well-known frequency-based criteria for success of
+FWI. The result of Theorem \ref{thm:fwi} can be rephrased as showing the
+existence of many stationary points of the mean-square error function
+for which the travel time is in error by more than
+$\lambda_{\krms[w]|}$. In fact, the usual error criterion mentioned in the
+literature is that the initial estimate of travel time must be in error by at most ``half a
+wavelength'' if FWI is to converge reliably. This is correct in some
+circumstances, depending on features of the target wavelet $w_*$ of
+which the arguments in this paper do not take account.
+
+\begin{proposition}
+  \label{thm:heis2}
+  For $\lambda>0, w \in \lW^1$,
+  \begin{equation}
+    \label{eqn:heis2}
+    \|A^2w\| \ge \frac{\|w\|}{24 \lambda (2\pi\krms[w])^3}.
+  \end{equation}
+\end{proposition}
+
+\begin{proof}
+  Since $A$  preserves $\lW^1$, inequality \ref{eqn:heis} implies
+  \begin{equation}
+    \label{eqn:heissq}
+    \|A^2w\| \ge \frac{\|Aw\|^2}{2\|(Aw)'\|}.
+  \end{equation}
+  From the definition of $A$, $(Aw)'=w +Aw'$, so for $w \in \lW^1$,
+  \[
+    \|(Aw)'\| \le \|w\|+ \lambda\|w'\|.
+  \]
+  Applying \ref{eqn:heis} again,
+  \[
+    \|w\|^2 \le 2\|w'\|\|Aw\|\le 2\lambda \|w'\|\|w\|,
+  \]
+  so obtain the 1D Poincar\'{e} inequality: for $w \in \lW^1$,
+  \[
+    \|w\| \le 2\lambda\|w'\|.
+  \]
+  Thus
+  \[
+    \|(Aw)'\| \le 3\lambda\|w'\|.
+  \]
+  Apply this estimate together with the basic Heisenberg estimate
+  \ref{eqn:heis} to the inequality \ref{eqn:heissq} to obtain
+  \[
+    \|A^2w\| \ge \frac{\|w\|^4}{24\lambda \|w'\|^3}
+  \]
+  Rearranging and using the definition \ref{eqn:krms} of $\krms[w]$,
+  arrive at inequality \ref{eqn:heis2}.
+\end{proof}
 
 \begin{theorem}
-  \label{thm:minsol}
-  Suppose $W$ and $D$ are Hilbert spaces, $S \in {\cal B}(W,W)$, $R
-  \in {\cal B}(W,D)$, $d \in D, w \in W$, and $\alpha, \epsilon>0, \gamma >
-  1$. Suppose $\lambda_{\rm min} \le \lambda_{\rm max}$ and that
-  $\{\lE: \lambda \in [\lambda_{\rm min},\lambda_{\rm max}]\}$ is the
-  spectral resolution of $\sqrt{S^TS}$. Then for any $\lambda \in  [\lambda_{\rm min},\lambda_{\rm max}]$,
-  \begin{equation}
-    \label{eqn:gentol}
-    \|R\lE w-d\| \le \epsilon\|d\|
-  \end{equation}
-  provided that 
-  \begin{equation}
-    \label{eqn:epslam}
-    \epsilon \ge \gamma \frac{(\|Rw-d\|^2 +
-      \alpha^2\|Sw\|^2)^{1/2}}{\|d\|}, \,\,\lambda \ge 
-    \frac{\|R\|}{\alpha (\gamma-1)}
+  \label{thm:ipnonoisenec}
+  Suppose that $a$ is given by definition \ref{eqn:ann}, $\alpha$,
+  $\mu \in (0,\lambda_{\rm max}/2]$, $d$ is given by
+  \ref{eqn:defdatanonoise} with $w_* \in W_{\mu}^1$, and that a
+  stationary point $m_{\infty}$ of $\tJa[\cdot;d]$ yields a solution
+  of the inverse problem \ref{eqn:probstat0} for $\lambda \ge 2\mu$,
+  $\epsilon>0$. Then
+\begin{equation}
+  \label{eqn:epsalpha}
+  3 (4\pi \lambda \krms[w_*])^3 \epsilon \ge \frac{(4\pi  r\alpha\lambda)^2}
+  {(1+(4\pi r \alpha \lambda)^2)}
 \end{equation}
 \end{theorem}
+
 \begin{proof}
-First note that
-\[
-  \|Sw\|^2 = \|S(\lE w + (I-\lE)w)\|^2 =
-  \langle \lE w + (I-\lE)w, S^TS (\lE w + (I-\lE)w)\rangle
-\]
-\[
-  =  \langle \lE w + (I-\lE)w, (\lE w + (I-\lE) S^TS w)\rangle
-\]
-\[
-  = \langle \lE w,\lE S^TSw \rangle+ \langle(I-\lE)w,(I-\lE) S^TS w\rangle
-\]
-(since $\lE$ is a projector)
-\[
-\ge  \langle (I-\lE)w,S^TS (I-\lE)w \rangle
-\]
-\[
-  =\langle w, S^TS (I-\lE)w\rangle
-\]
-(since $(I-\lE)$ commutes with $S^TS$)
-\[
-  =\int_{\lambda}^{\lambda_{\rm max}} \mu^2 d\langle w,E_{\mu}w\rangle 
-\]
-\[
-  \ge \lambda_{\rm min}^2 \int_{\lambda}^{\|A\|} d\langle w,E_{\mu}w\rangle
-\]
-\begin{equation}
-  \label{eqn:lb}
-  =\lambda^2\langle w, (I-\lE)w\rangle   =\lambda^2\|(I-\lE)w\|^2
-\end{equation}
-
-Set
-\[
-  \delta =  \frac{(\|Rw-d\|^2+\alpha^2\|Sw\|^2)^{1/2}}{\|d\|}.
-\]
-For any $\lambda >0$,
-\[
-  \alpha \lambda \|(I-\lE)w\| \le \alpha \|Sw\| \le \delta \|d\|^2
-\]
-whence
-\[
-  \|R\lE w-d\| \le \|Rw-d\|+\|R(I-\lE)w\|
-\]
-\[
-  \le \delta\|d\| + \|R\|\frac{\delta}{\alpha \lambda}\|d\|
-\]
-\[
-  \le \left(1 + \frac{\|R\|}{\alpha\lambda}\right)\delta\|d\|
-\]
-The presumed bounds \ref{eqn:epslam} on $\epsilon$ and $\lambda$ imply that this is
-\[
-  \le \gamma \delta \|d\| \le \epsilon\|d\|.
-\]
+  Rearranging the identity \ref{eqn:residnorm}, and observing as in
+  the proof of Theorem \ref{thm:ipnonoisesuf} that the support of the
+  integrand is contained in $[-2\mu, 2\mu] \subset [\lambda,\lambda]$,
+  \[
+  e[m_{\infty},\aw[m_{\infty},d];d] 
+= 8 \pi^2 r^2 \alpha^4\int\,dt\,t^4(1+(4\pi r)^2 \alpha^2 t^2)^{-2}w_*(t+(m_{\infty}-m_*))^2
+\]
+\[
+  \ge \frac{8\pi^2r^2\alpha^4}{(1+(4\pi r)^2 \alpha^2
+    \lambda^2)^{2}}\int\,dt\,t^4w_*(t+(m_{\infty}-m_*))^2
+\]
+\begin{equation}
+  \label{eqn:residnormbis}
+  \ge \frac{8\pi^2r^2\alpha^4}{(1+(4\pi r)^2 \alpha^2
+    \lambda^2)^{2}}\|A^2w_*(\cdot + (m_{\infty}-m_*)r)\|^2.
+ \end{equation}
+Since $\krms[w]$ is invariant under translation of $w \in V^1$,
+Proposition \ref{thm:heis2} implies that this is
+\[
+  \ge \frac{8\pi^2r^2\alpha^4}{(1+(4\pi r)^2 \alpha^2
+    \lambda^2)^{2}}\frac{\|w_*\|^2}{(24)^2 \lambda^2 (2\pi\krms[w_*])^6}
+\]
+\[
+  = \frac{1}{2}\frac{(4\pi r\alpha\lambda)^4}{(1+(4\pi r \alpha
+    \lambda)^2)^{2}}\frac{\|d\|^2}{(24)^2 (2\pi
+    \lambda \krms[w_*])^6}
+\]
+The pair $(m_{\infty}, \aw[m_{\infty},d])$ is presumed to solve the inverse
+problem as stated in \ref{eqn:probstat0}, in particular
+\[
+  \epsilon \ge (2 e[m_{\infty}, \aw[m_{\infty},d],d])^{1/2}/\|d\|
+\]
+the inequality \ref{eqn:epsalpha} follows.
+
 \end{proof}
 
-In the context of the single trace transmission problem, the operators
-
-\begin{corollary}
-  Suppose that $m \in M, d\in D, \gamma>1, \epsilon>0, \lambda >0, \alpha
-  > 0$ satisfy
-    \begin{equation}
-    \label{eqn:epslam}
-    \epsilon \ge \gamma \frac{(2\tJa[m;d])^{1/2}}{\|d\|}, \,\,\lambda \ge 
-    \frac{\|F[m]\|}{\alpha (\gamma-1)}
-  \end{equation}
-  Then $(m,\lE \aw[m;d])$ is a solution of the inverse problem as stated
-  in condition \ref{eqn:probstat0}, that is, $\lE\aw[m;d] \in \lW$ and
-  \[
-    e[m,\aw[m;d];d] \le \frac{\epsilon^2}{2} \|d\|^2
-  \]
-\end{corollary}
-
-That is, the minimizer of $\Ja$ is a solution of the constrained
-problem, as stated in the Problem Statement \ref{eqn:probstat0}, for
-$\lambda = O(1/\alpha)$, provided that the target noise level
-($\epsilon$) is higher by a fudge factor $\gamma>1$ than the minimum
-value of $\Ja$.
-
-It remains to establish how
+Inequality \ref{eqn:epsalpha} couples the dimensionless quantites
+$\epsilon$,  $4\pi r \lambda \alpha$, and $4 \pi \lambda
+krms[w_*]$. Proposition \ref{thm:klam}, implies that the left
+hand side is $\ge 3\epsilon$. Since the right hand side is $\le 1$,
+the inequality implies no limitation on $\alpha$ if the left hand side
+is $\ge 1$. For small $\epsilon$, the the largest permissible $\alpha$
+is $O(\sqrt{\epsilon})$. The permissible range of $\alpha$ increases with
+nondimensionalized RMS frequency $4\pi \lambda \krms[w_*]$. Since
+$\krms$ is invariant under translation and scaling of its argument,
+$\krms[w_*]=\krms[d]$, that is, the nondimensionalized RMS frequency
+is an observable property of the data, in the noise-free case at
+least.
+
+\subsection{The Effect of Noise}
+Suppose that the data trace $d$ takes the form
+\begin{equation}
+  \label{eqn:defdatanoisy}
+  d = F[m_*]w_* + n
+\end{equation}
+with $m_* \in M, w_* \in \lW$ as before, and noise trace $n \in
+D$. Since no support assumptions can be made about $n$, equation
+\ref{eqn:normsol} implies that $\aw[m;d] \notin \lW$ for any value of
+$\alpha$.  Therefore minimization of $\tJa$ cannot by itself yield a
+solution of the inverse problem as defined in the problem statement
+\ref{eqn:probstat0}.
+
+Recall that $\lE$ is the projection of $W$ onto $\lW$
+(i.e. multiplication by ${\bf 1}_{[-\lambda,\lambda]}$.
+
+\begin{theorem}
+  \label{thm:ipnoisesuf}
+  Suppose that $a$ is given by definition \ref{eqn:ann}, $\alpha$,
+  $\mu \in (0,\lambda_{\rm max}]$, $d$ is given by
+  \ref{eqn:defdatanoise} with $w_* \in W_{\mu}$, $n \in D$, and
+  $m_{\infty}$ is a stationary point of $\tJa[\cdot;d]$. Then
+  $(m_{\infty},\lE\aw[m_{\infty};d])$ is a solution of the inverse
+  problem \ref{eqn:probstat0} for any $\lambda \ge 2\mu$ and
+  $\epsilon \ge (4\pi r \lambda \alpha)^2 + C\|n\|/\alpha$.
+\end{theorem}
 

### 2020-08-01 repro/papers/wri1d/rev1/response.tex
rev1 and submission versions

\address{email: {\tt symes@rice.edu}}
\title{Responses to reviews of ``Wavefield Reconstruction Inversion: an example'', submitted to Inverse Problems}
\author{William. W. Symes}

\lefthead{Symes}

\righthead{1D WRI}

\section{Responses to Referee Comments}
I very much appreciate the careful and constructive criticisms provided by the referees, and have attempted to address every point they have made, at the cost of lengthening the paper somewhat.

\section{Referee 1}

\noindent Comment: (1)The theory is credible, however, I also suggest that you’d better supply some numerical examples;

\noindent Response: I have added a short section with a numerical example.

\noindent Comment: (2)In the section of abstract, I suggest that you should focus on our new method;

\noindent Response: Indeed, I would love to do so, but you have the advantage of me: I don't know who you are, and even more don't know what your ``new method'' is. Also note that abstracts should not include references.

\noindent Comment: (3)Equations (1),(4)-(7), (9), (12)-(14), (19), (21), (23)-(26), (28)-(32): lack of punctuation marks;

\noindent Response: thank you - I have attempted to properly punctuate every equation, and appreciate you calling these out.

\noindent Comment: (4)Pg. 2, line 12: “((Bamberger et al.,” Please delete a “(”.

\noindent Response: thank you for pointing this out.

\noindent Comment: (5)DISCUSSION: operator-valiued-->operator-valued;

\noindent Response: Fixed, thanks.

\noindent Comment: (6)In the section of Introduction, please supply the researches in the last three years.

\noindent Response: I have added more references, to make clear tha work on WRI continues right up to the present. This paper is not meant to be a review of WRI - that is for someone more involved in it to write. I just want to make the point that the claims offered for WRI as a cycle-skip remedy are clearly wrong, by showing a counter-example to which these claims should also apply.

\section{Referee 2}

\noindent Comment: on the relation between WRI and MSWI.

\noindent Response: You have made me aware that my first draft was too glib about the relation between these various approaches. I have added a full description of the relation between ``traditional'' WRI as formulated by van Leeuwen and Herrmann, and the variant of MSWI that I call WRI in this paper. In fact they are completely equivalent under conditions that hold for the problem I consider, in that (1) the original formulation of the vL-H GJI 2013 paper and the VPM reduction as in their 2016 IP paper have stationary points and local minima in 1-1 correspondence, and (2) the VPM-WRI of the 2016 IP paper and the VPM reduction of the problem stated in display 2 of my paper also have stationary points and local minima in 1-1 correspondence - it could not be otherwise, as these two reduced objectives are the same function! The equivalence (1) is a consequence of the separable part of the objective (that is, the dependence on the state variable) being strictly convex quadratic. In any case, while this discussion bulks up the paper a bit, I agree that it is really necessary to fully support my conclusions - namely, it would not matter what version of WRI you use, you will wind up with the same behaviour.

I also gave a more detailed (but still rough) account of the theoretical results of van Leeuwen and Herrmann from their 2016 IP paper, and tried to identify clearly the reason advanced in most papers on the topic to think WRI might overcome cycle-skipping - namely bigger search space, that is, merely the fact that it uses extended modeling. Further, I pointed out that the theory developed in my paper applies precisely to the 1D transmission problem, and not to any other WRI implementation (necessarily). However the arguments in favor of WRI also apply, and therefore are shown to be (at least) incomplete.

\noindent Comment: on other approaches to convexifying FWI.

\noindent Response: I have added a few sentences in the introduction, with some recent references, on the use of the Wasserstein metric. Also, I identified WRI as being based on extended modeling, and gave a reference which includes a discussion of MVA and related ideas (and some historical stuff, as extended modeling is really an old story). You are right that these things should be mentioned, and of course more could be said, eg. about tomography, but I hesitate to do muich more than this, as the paper begins already to become a bit top-heavy.

\noindent Detailed Comments:

\noindent Response: many thanks! I believe I have addressed them all.

\noindent [Except for the first: ``...requires that the intial estimate predict...'' is a weird bit of English grammar called a modal verb phrase - ``requires that'' is the modal verb (a compound, unlike ``must'' or ``should'') and ``predict'' is the auxiliary verb, which is always used in its base (infinitive) form. So no ``s''. Another example: ``he must dig a hole'' vs. ``he digs a hole''. This is really obscure, sorry, there is nothing more confusing than English grammar.]

### 2020-08-01 repro/papers/wri1d/rev1/madfigs.tex
rev1 and submission versions

\inputdir{project}

\plot{wavelet}{width=\textwidth}{Ricker Wavelet, zero phase, 20 Hz
  center frequency.}

\plot{data}{width=\textwidth}{Predicted data for 1D transmission
  problem, source wavelet = 20 Hz Ricker wavelet of Figure
  \ref{fig:wavelet}, source-receiver offset = 1 km, velocity = 2.5
  km/s.}

\plot{j0}{width=\textwidth}{Blue curve = FWI objective, red curve =
  VPM-WRI objective, plotted as functions of slowness, $\alpha$= 0.1
  s/$\sqrt{\rm km}$.}

\plot{j3}{width=\textwidth}{Blue curve = FWI objective, red curve =
  VPM-WRI objective, plotted as functions of slowness, $\alpha$= 1.0
  s/$\sqrt{\rm km}$.}

### 2020-08-01 repro/papers/wri1d/rev1/paper.tex
rev1 and submission versions

--- 2020-03-28 repro/papers/wri1d/paper.tex
@@
+\address{email: {\tt symes@rice.edu}}
 \title{Wavefield Reconstruction Inversion: an example}
-\author{William. W. Symes \thanks{The Rice Inversion Project,
-Department of Computational and Applied Mathematics, Rice University,
-Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}
+\author{William. W. Symes\\Rice University}
 
@@
 \begin{abstract}
-  Full waveform inversion, or nonlinear least squares data-fitting
-  driven by wave equation simulation, has proven its ability to
+  Nonlinear least squares data-fitting driven by physical process
+  simulation is a classic and widely successful technique for the
+  solution of inverse problems in science and engineering. Known as
+  ``Full Waveform Inversion'' in application to seismology, it can
   extract detailed maps of earth structure from near-surface seismic
-  observations. However, the least squares error function at the heart
+  observations, but also suffers from a defect not always encountered
+  in other applications: the least squares error function at the heart
   of this method tends to develop a high degree of nonconvexity, so
   that local optimization methods (the only numerical methods feasible
-  for field-scale problems) tend to fail, unless provided with initial
+  for field-scale problems) may fail to produce geophysically useful
+  final estimates of earth structure, unless provided with initial
   estimates of a quality not always available. A number of alternative
@@
   to minimization of explicitly computable functions, in an asymptotic
-  sense. The analysis presented here shows that Wavefield
+  sense. The analysis presented here shows explicitly how multiple
+  local minima arise in Full Waveform Inversion, and that Wavefield
   Reconstruction Inversion can be vulnerable to the same
-  ``cycle-skipping'' failure mode as Full Waveform Inversion.
+  ``cycle-skipping'' failure mode.
 \end{abstract}
@@
 (material densities, stiffnesses, attenuation rates,...) form a vector
-$c$ of spatially fields that appear as coefficients in systems of hyperbolic
+$c$ of spatially varying fields that appear as coefficients in systems of hyperbolic
 partial differential equations, modeling seismic wave
@@
 The objective of FWI is to find $c$ and $f$, given $d$, so that
-$S[c]f \approx d$. A typical method for enforcing this requirement is
+$S[c]f \approx d$. A typical method for achieving this goal is
 the minimization of an objective misfit measure (objective, for
@@
     minimize }
-  \|d -  S[c]f\|^2
-\end{equation}
-in which $\|\cdot\|$ is the norm in a suitable Hilbert space. This
-approach was first suggested in the 1980's
-(\cite[]{BamChavLai:79,Tara:84a,KolbColLai:86,Crasetal:90}, and many
+  \|d -  S[c]f\|^2.
+\end{equation}
+In equation \ref{eqn:deffwi}, $\|\cdot\|$ is the norm in a suitable
+Hilbert space.
+
+This approach was first suggested in the 1980's
+(\cite{BamChavLai:79,Tara:84a,KolbColLai:86,Crasetal:90}, and many
 other papers since then). Usually some form of regularization is
@@
 that the initial estimate predict the correct arrival time of waves,
-as they appears in the data, within a wavelength at dominant
+as they appears in the data, within a wavelength (in fact,
+conventionally a half wavelength) at dominant
 frequencies \cite[]{GauTarVir:86,VirieuxOperto:09}.
@@
 of which appear to exhibit less tendency to develop local minima than
-does the problem described in \ref{eqn:deffwi} \cite[]{Symes:09}.
-
-This topic of this paper is one of these alternative approaches,
+does the problem described in \ref{eqn:deffwi}. For example, other
+metrics than the $L^2$ distance employed in definition
+\ref{eqn:deffwi} may be used to characterize the size
+of the data residual $S[c]f-d$, and some of these result in larger
+regions of convexity about global minima than possessed by the $L^2$ based
+objective \ref{eqn:deffwi} . A particularly promising class of
+alternative distance measures derives from optimal transport theory
+\cite[]{EngquistYang:GEO18,Metivier:GEO18}. Of course, practical
+seismology has estimated earth mechanical parameters, including wave
+velocities, for decades. Some of these practical methods are based on
+model extension, that is, the addition of extra degrees of freedom
+beyond those required by basic physics. More recently, the extension
+concept has been used to reformulate the FWI
+problem. \cite{geoprosp:2008} describes several variants of
+extension-based inversion and its origins in seismic data processing.
+
+This topic of this paper is one of these extension-based inversion approaches,
 Wavefield Reconstruction Inversion (``WRI''). WRI was introduced by \cite{LeeuwenHerrmannWRI:13}, and further
-developed by \cite{LeeuwenHerrmann:16,WangYingst:SEG16} and other
+developed by \cite{LeeuwenHerrmann:16,WangYingst:SEG16,Lietal:18, Leeuwen2019note,Herrmann:SEG19,Aghamiry:19,FangWangHerrmann:20,Aghmiryetal:20,Louboutinetal:20}; and other
 authors. It is based on the presumption that the correct source (right
@@
 \end{equation}
-Examples given
-by \cite{LeeuwenHerrmannWRI:13} and others suggest that the problem
-defined in display \ref{eqn:defwri} is less likely to develop
-uninformative local minima if $\alpha$ is small. This might be so for
-several reasons: one is that when less weight is
-put on making the residual in the wave equation ($f-q$) small, 
-$f$ may be chosen to make the data residual $d-S[c]f$ small
-instead. One aspect of failure to predict the arrival times of waves
-accurately is that small data residual is then difficult to
-achieve. By maintaining fidelity to
-the data, local minimization of the WRI objective might provide a
-spurious-minimum-free path to a satisfactory estimate of $c$.
+The formulation of WRI given by \cite{LeeuwenHerrmannWRI:13}
+appears to differ from the alternative formulation \ref{eqn:defwri},
+introduced by \cite{WangYingst:SEG16} but in fact the two are
+completely equivalent, in the sense of having corresponding local
+minimizers. The original form and its equivalence with the problem
+\ref{eqn:defwri} are explained in the fourth section of this paper. As
+noted by \cite{Leeuwen2019note}, either formulation is available not
+just for seismic inverse problems, but for any inverse problem based
+on a separable (partly linear) modeling operator.
+
+Numerical examples provided in the cited literature on WRI suggest
+that the problem defined in display \ref{eqn:defwri} is less likely to
+develop uninformative local minima (that is, is ``less nonlinear'') than
+is the least squares problem \ref{eqn:deffwi} in application to
+seismic inversion. The most-often cited justification for this point
+of view is that the WRI problem \ref{eqn:defwri} involves a search for an
+optimum over many more degrees of freedom than does the FWI problem
+\ref{eqn:deffwi}, since the trial source $f$ is also to be
+determined. More concretely, \cite{LeeuwenHerrmann:16} show that if
+$(c_*,f_*)$ is a
+stationary point of the objective defined in problem
+\ref{eqn:defwri}, then $c_*$ is within (roughly) the data residual scaled by the
+reciprocal penalty parameter (that is, $O(\alpha^{-1}\|d-S[c_*]f_*\|)$) of a
+global FWI minimizer. [The precise statement
+of this result, which follows from Theorem 4.3 in
+\cite[]{LeeuwenHerrmann:16}, depends on analysis of the FWI problem
+\ref{eqn:deffwi} as an equality-constrained optimization; we refer the
+reader to the cited reference for details.] This is however a {\em
+  local} characterization of WRI stationary points: it shows that
+these are close to solutions of the FWI problem only
+for those that produce a small data residual, and does not address the {\em
+  global} question of whether large-residual stationary points exist,
+as appears to be the case for FWI.
+% , if $\alpha$ is sufficiently small. 
+%This might be so for several reasons: one is that
+%when less weight is put on making the residual in the wave equation
+%($f-q$) small, $f$ may be chosen to make the data residual $d-S[c]f$
+%small instead. One aspect of failure to predict the arrival times of
+%waves accurately is that small data residual is then difficult to
+%achieve. By maintaining fidelity to the data, local minimization of
+%the WRI objective might provide a spurious-minimum-free path to a
+%satisfactory estimate of $c$.
 
 The main result of this paper is that in one simple case, in which all
-of the necessary computations can be carried out by hand, this hope is
-not realized: minimization of the WRI objective is just as likely to
+of the necessary computations can be carried out by hand, the WRI
+objective does indeed possess stationary points with large data residual.
+% this hope is not realized:
+In fact, in this case, minimization of the WRI objective is just as likely to
 be trapped in a spurious local minimizer as is the FWI objective.  The
@@
 ability to maintain better data fit via reduction of $\alpha$ does not
-lead to global behaviour asymptotically more amenable to local optimization.
+lead to global behaviour asymptotically more amenable to local
+optimization.
+
+I emphasize that this conclusion holds {\em a priori} for the 1D
+transmission inverse problem described in the next section, and {\em
+  only} for this problem. The results presented here imply nothing
+about the behaviour of any other instance of WRI. In particular, the
+analysis takes advantage of the diagonal (in fact, scalar) nature of
+the operator $S[c]S[c]^T$ (the central role played by this operator
+will become clear in the fourth section below). For other
+problems, for instance the simple example presented in
+\cite{Leeuwen2019note}, this operator is not diagonal, so the the
+short cuts used in the analysis given here are not available. However,
+it should be understood that the various heuristic justifications
+found in the literature for expecting WRI to avoid cycle-skipping, and
+the theoretical analysis in \cite{LeeuwenHerrmann:16}, all apply to
+the 1D transmission problem as well, and therefore in themselves are
+insufficient to imply avoidance of cycle-skipping.
 
@@
 approach, based on explicit solution of the 1D acoustic system
-presented in Appendix A. In the third section I introduce the $\lambda-$dependent
-family of problems, and establish the asymptotic properties of FWI as
-$\lambda \rightarrow 0$. The fourth section develops the algebraic
-structure of WRI, culminating in a remarkable identity revealing WRI
-to be equivalent to minimization of a weighted norm of the data
-residual, with a weight operator depending on the coefficient vector
-$c$. This identity has also been derived by \cite{Leeuwen2019note},
-using a different argument. The result is quite general, applying to essentially any
-realization of WRI, and already shows that its behaviour must be
-closely related to that of FWI. In the fifth section I return to 
-1D acoustics problem and the $\lambda$-dependent family of inverse
-problems, compute that weight operator explicitly, and deduce the
-global behaviour of WRI in this instance. The paper ends with a
-discussion of the relation of the analysis presented here to a
-previous analysis of optimization formulations of wave inversion
-problems based on parameter-dependent quadratic forms
-\cite[]{StolkSymes:03}, and a brief discussion of some other
-alternatives to FWI.
+presented in Appendix A. In the third section I introduce the
+$\lambda-$dependent family of problems, and establish the asymptotic
+properties of FWI as $\lambda \rightarrow 0$. The fourth section
+develops the algebraic structure of WRI, culminating in a remarkable
+identity revealing WRI to be equivalent to minimization of a weighted
+norm of the data residual, with a weight operator depending on the
+coefficient vector $c$. This identity has also been derived by
+\cite{Leeuwen2019note}, using a different argument. The result is
+quite general, applying to essentially any realization of WRI, and
+already shows that its behaviour must be closely related to that of
+FWI. In the fifth section I return to 1D acoustics problem and the
+$\lambda$-dependent family of inverse problems, compute that weight
+operator explicitly, and deduce the global behaviour of WRI in this
+instance. This behaviour is illustrated by a very simple numerical
+example in the sixth section. The paper ends with a discussion of the
+relation of the analysis presented here to a previous analysis of
+optimization formulations of wave inversion problems based on
+parameter-dependent quadratic forms \cite[]{StolkSymes:03}, and a
+brief discussion of an extension-based inversion approach using a different
+penalty than does WRI, which in application to the 1D transmission
+problem provably avoids cycle-skipping.
 
 \section{FWI for 1D Acoustics}
-The example of FWI to be explored in this paper is one of the simplest possible, based on the 1D acoustics system connecting excess pressure $p$, particle velocity $v$, sonstitutive law defect (``source'') $f$, density $\rho$, and wave velocity $c$:
+The example of FWI to be explored in this paper is one of the simplest possible, based on the 1D acoustics system connecting excess pressure $p$, particle velocity $v$, constitutive law defect (``source'') $f$, density $\rho$, and wave velocity $c$:
 \begin{eqnarray}
@@
   \label{eqn:defmod}
-  S[c]f = p|_{\{z_r\}\times [0,T]}
-\end{equation}
-
-The formulation of the inverse problem via least-squares requires a choice of Hilbert space structure for the domain and range of $S[c]$. A natural choice is
+  S[c]f = p|_{\{z_r\}\times [0,T]}.
+\end{equation}
+This trace is well-defined if $f$ is smooth, since then the solution
+$(p,v)$ is smooth.
+For homogeneous ($z$-independent) $c$, Appendix A provides an explicit expression:
+\begin{equation}
+  \label{eqn:defmodhom}
+  S[c]f(t) = \frac{1}{2c}\int_{z_{\rm min}}^{z_{\rm max}} dz f\left(z,t - \frac{|z_r-z|}{c}\right). 
+\end{equation}
+From this expression it clear that $S[c]$ extends a bounded
+operator
 \begin{equation}
   \label{eqn:defmoddom}
-  S[c]: L^2([z_{\rm min},z_{\rm max}] \times \bR) \rightarrow L^2[0,T]
-\end{equation}
-That is, the support of f will be assumed to lie in the strip $[z_{\rm min},z_{\rm max}] \times \bR$.
-As it will turn out, the choice of spatial interval $[z_{\rm min},z_{\rm max}]$ is arbitrary, so long as it has positive length.
-
-For homogeneous ($z$-independent) $\kappa$, Appendix A provides an explicit expression:
-\begin{equation}
-  \label{eqn:defmodhom}
-  S[c]f(t) = \frac{1}{2c}\int_{z_{\rm min}}^{z_{\rm max}} dz f\left(z,t - \frac{|z_r-z|}{c}\right) 
-\end{equation}
-From this expression it is simple to see that $S[c]$ is a bounded
-operator with the domain and range described in
-\ref{eqn:defmoddom}. The role of compact support in $z$ in ensuring
-boundedness is also evident.
-%It is also evident from that limiting source to have bounded support in $z$ is necessary make $S[c]$ bounded.
-
-With this framework, the data is presumed to lie in the range space of
+  S[c]: L^2([z_{\rm min},z_{\rm max}] \times \bR) \rightarrow L^2[0,T].
+\end{equation}
+That is, the support of f will be assumed to lie in the strip $[z_{\rm
+  min},z_{\rm max}] \times \bR$.  The choice of spatial interval
+$[z_{\rm min},z_{\rm max}]$ is arbitrary, so long as it has positive
+length. The role of compact support in $z$ in ensuring
+boundedness of $S[c]$ is also evident from equation \ref{eqn:defmodhom}.
+
+It should be mentioned that the triviality of the conclusion
+\ref{eqn:defmoddom} is a happy accident of one space dimension: for
+analogous hyperbolic initial-boundary value problems in more than one
+space dimension, simple integral representations such as
+\ref{eqn:defmodhom} are not to be expected, and the regularity of
+traces becomes a much more involved matter - see for example
+\cite{Symes:83}, \cite{Lasi:87}, and \cite{BaoSy:91b}.
+
+The formulation of the inverse problem via least-squares requires a
+choice of Hilbert space structure for the domain and range of $S[c]$,
+given for this problem in display \ref{eqn:defmoddom}.
+The data $d$ is presumed to lie in the range space of
 $S[c]$: $d \in L^2[0,T]$. It is simple to verify from the identity
@@
 $w \in L^2(\bR)$. Formally, the resulting acoustic field satisfies the
-system \ref{eqn:awe1d} with $f=w(t)\delta(z-z_s)$. Inserting this
+system \ref{eqn:awe1d} with $f(z,t)=w(t)\delta(z-z_s)$. Inserting this
 expression in the explicit expression \ref{eqn:defmodhom}, obtain
-\begin{equation}
+\begin{eqnarray}
+  f(z,t) & = & w(t)\delta(z-z_s),\nonumber\\
+  (S[c]f)(t) &=& \frac{1}{2c}w\left(t -
+                              \frac{|z_r-z_s|}{c}\right)\nonumber\\
+  &= &(S_p[c]w)(t).
   \label{eqn:defmodpt}
-  S[c](w\delta(\cdot-z_s)) = \frac{1}{2c}w\left(t - \frac{|z_r-z_s|}{c}\right) = S_p[c]w(t)
-\end{equation}
-Thus for this restricted class of source fields, the FWI objective can be redefined as
+\end{eqnarray}
+$S_p[c]$ is a bounded operator with domain $L^2(\bR)$ and range
+$L^2([0,T])$.
+
+For this restricted class of source fields, the FWI problem
+\ref{eqn:deffwi} can be redefined as minimization of the objective
 \begin{equation}
   \label{eqn:deffwipt}
-  J_{\rm FWI}[c,d.w] = \frac{1}{2}\|S_p[c]w -d\|^2.
-\end{equation}
-$S_p[c]$ is a bounded operator with domain $L^2(\bR)$ and range $L^2([0,T])$.
+  J_{\rm FWI}[c;d,w] = \frac{1}{2}\|S_p[c]w -d\|^2.
+\end{equation}
+over $c$, with $d$ and $w$ treated as parameters.
 
@@
   \label{eqn:ext}
-  S_p[c] = S[c] \circ E[c]
+  S_p[c] = S[c] \circ E[c].
 \end{equation}
@@
 In the definition \ref{eqn:wfam}, the ``mother wavelet'' $w_1\in
-C_0^{\infty}(0,1)$ has dimensionless argument, and the scaling is chosen so fhat
+C_0^{\infty}(0,1)$ has dimensionless argument, and the scaling is chosen so that
 $\|\wl\|_{L^2(\bR)}$ is independent of $\lambda>0$.
@@
    \label{eqn:time}
-  \frac{|z_s-z_r|}{c_{\rm min}} < T
+  \frac{|z_s-z_r|}{c_{\rm min}} < T.
 \end{equation}
@@
 and a corresponding family of FWI objectives:
-\[
-  J_{\rm FWI}[c,\dl,\wl] = \frac{1}{2}\|\dl-S_p[c]\wl\|^2
-\]
-\begin{equation}
+\begin{eqnarray}
+  J_{\rm FWI}[c;\dl,\wl] &=& \frac{1}{2}\|\dl-S_p[c]\wl\|^2\nonumber\\
+  &=& \frac{1}{2} \int_0^T\,dt\,\left|\frac{1}{2c_*}\wl\left(t-\frac{|z_s-z_r|}{c_*}\right)
+    - \frac{1}{2c}\wl\left(t-\frac{|z_s-z_r|}{c}\right)\right|^2.
   \label{eqn:fwiexpl}
-  =
-  \int_0^T\,dt\,\left|\frac{1}{2c_*}\wl\left(t-\frac{|z_s-z_r|}{c_*}\right)
-    - \frac{1}{2c}\wl\left(t-\frac{|z_s-z_r|}{c}\right)\right|^2
-\end{equation}
+\end{eqnarray}
 Note that $\mbox{supp }\wl \subset [0,\lambda]$, so
@@
   \mbox{supp }S_p[c]\wl \subset \left[\frac{|z_s-z_r|}{c},\lambda
-    +\frac{|z_s-z_r|}{c}\right] \cap [0,T]
+    +\frac{|z_s-z_r|}{c}\right] \cap [0,T].
 \end{equation}
@@
 Then if $|c_*-c| > L\lambda$,
-\[
+\begin{eqnarray}
   \left|\frac{|z_s-z_r|}{c}-\frac{|z_s-z_r|}{c_*}\right|
-  =\frac{|c-c_*||z_s-z_r|}{cc_*} 
-\]
-\[
-\ge \frac{|c-c_*||z_s-z_r|}{c_{\rm max}^2} > L\lambda
-\frac{|z_s-z_r|}{c_{\rm max}^2} = 2\lambda.
-\]
-That is, the infima of the supports of $S_p[c]\wl, \dl=S_p[c_*]\wl$
-are further apart than the lengths of these supports. In that case,
+  &=&\frac{|c-c_*||z_s-z_r|}{cc_*} \nonumber\\
+  &\ge& \frac{|c-c_*||z_s-z_r|}{c_{\rm max}^2}\nonumber\\
+  &>& L\lambda\frac{|z_s-z_r|}{c_{\rm max}^2} \nonumber\\
+  &=& 2\lambda.
+      \label{eqn:shift}
+\end{eqnarray}
+That is, equation \ref{eqn:shift} shows that when the condition
+\ref{eqn:defL} is satisfied, the infima of the supports of $S_p[c]\wl, \dl=S_p[c_*]\wl$
+are further apart than the lengths of these supports. Then
 necessarily
-\[
-  \mbox{supp }S_p[c]\wl \cap \mbox{supp }S_p[c_*]\wl = \emptyset.
-\]
-Therefore $S_p[c]\wl$ and $S_p[c_*]\wl$ are orthogonal in $L^2[0,T]$,
+  $\mbox{supp }S_p[c]\wl \cap \mbox{supp }S_p[c_*]\wl = \emptyset$, so
+$S_p[c]\wl$ and $S_p[c_*]\wl$ are orthogonal in $L^2[0,T]$,
 and
@@
   \label{eqn:fwiscrewed}
-   J_{\rm FWI}[c,\dl,\wl]  =
+   J_{\rm FWI}[c;\dl,\wl]  =
   \frac{1}{2}\left(\frac{1}{4c^2}+\frac{1}{4c_*^2}\right)\|w_1\|^2.
@@
   For $L>0$ given by equation \ref{eqn:defL} and $\lambda <
-  \lambda_0$, the minimizer of $J_{\rm FWI}[c,\dl,\wl]$ on the
+  \lambda_0$, the minimizer of $J_{\rm FWI}[c;\dl,\wl]$ on the
   complement of $[c_*-L\lambda, c_*+L\lambda]$ is $c=c_{\rm max}$.
@@
 This section will describe Wavefield Reconstruction Inversion (WRI)
-and develop some of its formal algebraic properties. The notation
-$S[c]$ will represent a modeling operator based on wave dynamics of
-some sort, depending on a vector $c$ of material parameters. The
-conclusions developed here in fact apply to WRI in any such
-setting. These conclusions will be applied to 1D acoustics in the
-following section, with $c$ specialized to a scalar (wave velocity).
-
-In the application below, the target source $q$ will be a point source
-as in the previous section, however for the development of the basic
-properties of WRI, that is immaterial.
-
-Note that in general $\alpha$ is a dimensional parameter, having the
-same dimensions as $S$.
-
-\cite{LeeuwenHerrmannWRI:13} posed this problem slightly differently:
-instead of the first order acoustics system (equation \ref{eqn:awe1d}
-for the 1D case), they pose the wave dynamics in terms of the second
-order wave equation for the pressure wavefield $p$. From this
-viewpoint, $\partial(f-q) /\partial t$ is the residual, that is, the
-difference between the image $\partial f/\partial t$ of the 2nd order
-wave operator on $p$, and the assumed right-hand side
-$\partial q/\partial t$. So in this form, the second term penalizes
-the failure of $p$ to solve the wave equation with the assumed
-source. The formulation presented here is equivalent, and was
-introduced by \cite{WangYingst:SEG16}.
-
-\cite{LeeuwenHerrmann:16} used the variable projection method
-\cite[]{GolubPereyra:03}, eliminating the source $f$ in the inner step
-and updating the bulk modulus (or an equivalent quantity) in the outer
-step. That is, the problem \ref{eqn:defwri} is equivalent to
-minimization of
-\[
-  J_{\rm WRI}^{\alpha}[c,d,q] =
-  \mbox{min}_f \frac{1}{2}(\| d-S[c]f\|^2+\alpha^2\|f-q\|^2)
-\]
-over $c$. This is the approach that I shall pursue here.
-
-It will turn out to be convenient to define the residual with the
-target source $q$ as $r[c]=d-S[c]q$, and set $g=f-q$. Then this
-definition can be rewritten as
+and develop some of its formal algebraic properties. I will follow the
+development of the basic concept and its variable projection
+refinement from the papers of van Leeuwen and Herrmann
+\cite[]{LeeuwenHerrmannWRI:13,LeeuwenHerrmann:16}, then show how
+WRI may be cast as an extended source inversion. In this section,
+consistent with the literature on this topic, I will abjure precise
+definitions of spaces and operators involved in the problem
+formulations. I introduce just enough mathematical structure to
+conclude that these three variants of WRI are {\em strictly
+  equivalent}: all three are optimization formulations, and a local
+minimizer of any one of the corresponds one-to-one with local
+minimizers of the others by well-defined transformations.
+
+These conclusions will be applied to 1D acoustics in the following
+section, with $c$ specialized to a scalar (wave velocity) and all of
+the missing definitions filled in.
+
+In the formulation of \cite{LeeuwenHerrmannWRI:13}, WRI involves a
+dynamical wavefield $u$, a data vector $d$, a data sampling operator
+$P$, a vector of material parameters $c$, a wave operator $L[c]$
+depending on $c$, a penalty weight $\alpha$, and a target source
+$q$. The domains and ranges of $P$ and $L[c]$ are presumed to be
+Hilbert spaces, in which the various vectors sit, and the symbol
+$\|\cdot\|$ will represent the appropriate Hilbert norm in each place
+where it appears.
+
+The WRI objective function is defined as
+\begin{equation}
+  \label{eqn:wriorig}
+  \phi_{\alpha}[c,u;d,q] = \frac{1}{2}(\|Pu-d\|^2 +
+  \alpha^2\|L[c]u-q\|^2).
+\end{equation}
+Note that I have changed the notation of \cite{LeeuwenHerrmannWRI:13}
+slightly to conform more closely to that used in the rest of this
+paper, however the meaning is exactly the same as their equation 5.
+
+Note also that $d$ and $q$ appear as parameters in $\phi_{\alpha}$, which
+is to be optimized over $c,u$. In particular, the target source $q$
+must be known. \cite{FangWangHerrmann:20} describe an algorithm for
+incorporating estimation of $q$ into WRI.
+
+The objective $\phi_{\alpha}[c,u;d,q]$ is quadratic in $u$, so given
+the other three fields, estimation of $u$ results from solution of a
+linear least squares problem. \cite{LeeuwenHerrmannWRI:13} assume that
+$L[c]$ is the Helmholtz operator and $c$ is the square slowness. In
+that case (and some others), $\phi_{\alpha}[c,u;d,q]$ is quadratic in
+$c$, so minimization over $c$ with fixed $u$ is also a linear least
+squares problem (in fact, a very simple one, in the Helmholtz case, as
+it may be solved by pointwise division). Based on these observations,
+\cite{LeeuwenHerrmannWRI:13} minimize $\phi_{\alpha}$ via
+coordinate search, alternating minimization over $u$ with minimization
+over $c$.
+
+\cite{LeeuwenHerrmann:16} used the Variable Projection Method (VPM)
+\cite[]{GolubPereyra:03}, rather than coordinate search, to reduce the
+optimization of $\phi_{\alpha}[c,u;d,q]$ to an optimization in $c$
+alone. Define
+\begin{equation}
+  \label{eqn:ulambda}
+  u_{\alpha}[c;d,q]= \mbox{arg}\min_u \phi_{\alpha}[c,u;d,q].
+\end{equation}
+Then assuming (as do \cite{LeeuwenHerrmann:16}) that $L[c]$ is
+invertible for any admissible $c$, $L[c]^TL[c]$ is symmetric positive
+definite, and therefore $u_{\alpha}$ is uniquely determined as the
+solution of a normal equation (equation after display (20) in
+\cite[]{LeeuwenHerrmann:16})
+\begin{equation}
+  \label{eqn:unorm}
+  u_{\alpha}[c;d,q] = (L[c]^TL[c] + \alpha^{-2} P^TP)^{-1}(L[c]^Tq  +
+  \alpha^{-2}P^Td).
+\end{equation}
+The VPM reduction of the objective function is
+\begin{equation}
+  \label{eqn:uvpm}
+  \phi_{\alpha}^{\rm
+    red}[c;d,q]=\phi_{\alpha}[c,u_{\alpha}[c;d,q];d,q].
+\end{equation}
+Note that $u_{\alpha}$ is defined by solution of an unconstrained
+convex quadratic minimization \ref{eqn:ulambda}. Therefore $c,u$ is a
+local minimizer of $\phi_{\alpha}[\cdot,\cdot;d,q]$ {\em if and only
+  if} $c$ is a local minimizer of $\phi_{\alpha}^{\rm red}[\cdot;d,q]$
+{\em and} $u = u_{\alpha}[c;d,q]$. That is, local minimizers of the
+reduced objective $\phi_{\alpha}^{\rm red}[\cdot;d,q]$ are in
+one-to-one correspondence with local minimizers of the basic WRI
+objective $\phi_{\alpha}[\cdot,\cdot;d,q]$. In that sense, the
+original and reduced WRI problems are completely equivalent.
+
+The assumption that $L[c]$ is invertible is essential to the
+arguments of \cite{LeeuwenHerrmannWRI:13,LeeuwenHerrmann:16}. That is, the
+equation $L[c]u=f$ has a unique solution $u=L[c]^{-1}f$ for any $f$ in
+the range space of $L[c]$ (as in the cited papers, I will leave the
+actual identity of both domain and range of $L[c]$ a bit vague). With
+$u_{\alpha}$ defined as in equation \ref{eqn:unorm}, compute
+\[
+  f_{\alpha}[c;d,q] \equiv L[c]u_{\alpha}[c;d,q]
+\]
+\[
+  = L[c](L[c]^TL[c] + \alpha^{-2} P^TP)^{-1}L[c]^T(L[c]^T)^{-1}(L[c]^Tq  +
+  \alpha^{-2}P^Td)
+\]
+\[
+ = (I + \alpha^{-2}(PL[c]^{-1})^T(PL[c]^{-1}))^{-1}(q +
+  \alpha^{-2}(PL[c]^{-1})^Td).
+\]
+Define $S[c] = PL[c]^{-1}$. Then with a little rearranging,
+\begin{equation}
+  \label{eqn:fnorm}
+   f_{\alpha}[c;d,q] = (S[c]^TS[c] + \alpha^2 I)^{-1}(\alpha^2q + S[c]^Td).
+\end{equation}
+Inspection shows that $f_{\alpha}[c;d,q]$ is the solution of a least
+squares problem:
+\begin{equation}
+  \label{eqn:fprob}
+  f_{\alpha}[c;d,q]=\mbox{arg}\
+ min_f \frac{1}{2}(\|S[c]f-d\|^2 +
+  \alpha^2\|f-q\|^2).
+\end{equation}
+VPM (elimination of $f$) applied to the objective on the right hand
+side of equation \ref{eqn:fprob} yields the reduced objective
+\[
+  J_{\rm WRI}^{\alpha}[c;d,q] =
+  \min_f \frac{1}{2}(\| d-S[c]f\|^2+\alpha^2\|f-q\|^2)
+\]
+\begin{equation}
+  \label{eqn:fvpm}
+  =\frac{1}{2}(\|d-S[c]f_{\alpha}[c;d,q]\|^2 +
+  \alpha^2\|f_{\alpha}[c;d,q]-q\|^2).
+\end{equation}
+
+The main point to be made here is that the objectives
+$\phi_{\alpha}^{\rm red}$ and $J_{\rm WRI}^{\alpha}$ are {\em the
+  same}:
+\[
+  \phi_{\alpha}^{\rm red}[c;d,q] =
+  \frac{1}{2}(\|L[c]u_{\alpha}[c;d,q]-q\|^2
+  +\alpha^2\|d-Pu_{\alpha}[c;d,q]\|^2)
+\]
+\[
+  =\frac{1}{2}(\|f_{\alpha}[c;d,q]-q\|^2+\alpha^2\|d-PL[c]^{-1}f_{\alpha}[c;d,q]\|^2)
+\]
+\begin{equation}
+  \label{eqn:fu}
+  =J^{\alpha}_{\rm WRI}[c;d,q].
+\end{equation}
+Since these are the same functions, they have the same local
+minimizers. As noted earlier, $c$ is a local minimizer of
+$\phi_{\alpha}^{\rm red}[c;d,q]$ if and only if
+$(c,u_{\alpha}[c;d,q])$ is a local minimizer of
+$\phi_{\alpha}[c,u;d,q]$. Since $c \mapsto u_{\alpha}[c;d,q]$ is
+injective, conclude that the sets of local minimizers of the three
+optimization problems defined so far in this section are in one-to-one correspondence.
+
+Define the residual with the
+target source $q$ as $r[c]=d-S[c]q$, and set $g=f-q$. Then the
+definition \ref{eqn:fvpm} can be rewritten as
 \begin{equation}
 \label{eqn:defvpm}
-  J_{\rm WRI}^{\alpha}[c,d,q] =\mbox{min}_g \frac{1}{2}(\| r[c]-S[c]g\|^2+\alpha^2\|g\|^2)
-\end{equation}
-
-In any penalty method, control of the penalty
-parameter has a large influence on the speed of
-convergence. \cite{Aghamiry:19} use an augmented Lagrangian algorithm
-to minimize the influence of the penalty weight choice.
-Alternatively, one can use a version of the discrepancy principle to adjust $\alpha$ dynamically \cite[]{FuSymes2017discrepancy}, as the WRI problem has the necessary features described in that paper.
+  J_{\rm WRI}^{\alpha}[c;d,q] =\min_g \frac{1}{2}(\| r[c]-S[c]g\|^2+\alpha^2\|g\|^2).
+\end{equation}
+This formulation was introduced by \cite{WangYingst:SEG16}.
 
@@
 \end{quote}
-That is, {em any data can be fit exactly using the extended model
+That is, {\em any data can be fit exactly using the extended model
   space (the domain of $S$)}.
-This property is a characteristic of of extended modeling
+This property is a characteristic of extended modeling
 methods \cite[]{geoprosp:2008}: the ability to fit any data appears to
-be a essential for such methods to produce objectives
+be essential for such methods to produce objectives
 without spurious local minima. It holds for the problems for which WRI
@@
 \[
-  J_{\rm WRI}^{\alpha}[c,d,q] =
-  \mbox{min}_{e,n} \frac{1}{2}(\| d-S[c](S[c]^Te+q)\|^2+\alpha^2(\|S[c]^Te\|^2+\|n\|^2))
+  J_{\rm WRI}^{\alpha}[c;d,q] =
+  \min_{e,n} \frac{1}{2}(\| d-S[c](S[c]^Te+q)\|^2+\alpha^2(\|S[c]^Te\|^2+\|n\|^2))
   \]
@@
   \label{eqn:defvpmred}
- =  \mbox{min}_{e} \frac{1}{2}(\|r[c]-S[c]S[c]^Te\|^2+\alpha^2\|S[c]^Te\|^2).
+ =  \min_{e} \frac{1}{2}(\|r[c]-S[c]S[c]^Te\|^2+\alpha^2\|S[c]^Te\|^2).
  \end{equation}
@@
 \[
- ( (S[c]S[c]^T)^2 + \alpha^2S[c](S[c]^T)e = S[c]S[c]^Tr[c]
+ ( (S[c]S[c]^T)^2 + \alpha^2S[c](S[c]^T)e = S[c]S[c]^Tr[c],
 \]
@@
 \[
-  S[c]S[c]^Te_{\alpha}[c] = S[c]S[c]^T(S[c]S[c]^T+\alpha^2I)^{-1}r[c]
+  S[c]S[c]^Te_{\alpha}[c] = S[c]S[c]^T(S[c]S[c]^T+\alpha^2I)^{-1}r[c].
 \]
@@
   \label{eqn:norsol}
-  e_{\alpha}[c]=(S[c]S[c]^T+\alpha^2I)^{-1}r[c]
+  e_{\alpha}[c]=(S[c]S[c]^T+\alpha^2I)^{-1}r[c].
 \end{equation}
 Consequently
-\[
-  J_{\rm WRI}^{\alpha}[c,d,q] =
-  \frac{1}{2}(\|r[c]-S[c]S[c]^Te_{\alpha}[c]\|^2+\alpha^2\|S[c]^Te_{\alpha}[c]\|^2)
-\]
-\[
-  = \frac{1}{2}\left(\|r[c]-S[c]S[c]^T
-    (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\|^2 \right.
-\]
-\[
+\begin{eqnarray}
+  J_{\rm WRI}^{\alpha}[c;d,q] &=&
+  \frac{1}{2}(\|r[c]-S[c]S[c]^Te_{\alpha}[c]\|^2+\alpha^2\|S[c]^Te_{\alpha}[c]\|^2)\nonumber\\
+  &=& \frac{1}{2}\left(\|r[c]-S[c]S[c]^T
+    (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\|^2\right. \nonumber\\
+  & &
   \left.+\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],
   S[c]S[c]^T (S[c]S[c]^T+\alpha^2I)^{-1}r[c]
-  \rangle\right)
-\]
-\[
-  =\frac{1}{2}\left(\|\alpha^2 (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\|^2\right.
-\]
-\[
-  \left.
-  +\alpha^2\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],
-  -\alpha^2 (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\rangle\right.
-\]
-\[
-  + \left.\alpha^2\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],r[c]\rangle\right)
-\]
-\[
-  =  \frac{\alpha^2}{2}
-\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],r[c]\rangle 
-\]
-That is,
+  \rangle\right)  \nonumber\\
+  &=&\frac{1}{2}\left(\|\alpha^2 (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\|^2\right. \nonumber\\
+  & &
+  + \left. \alpha^2 \langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],
+  - \alpha^2 (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\rangle\right. \nonumber\\
+  & &  + \left. \alpha^2\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],r[c]\rangle\right) \nonumber\\
+  &=&  \frac{\alpha^2}{2}\langle
+      (S[c]S[c]^T+\alpha^2I)^{-1}r[c],r[c]\rangle.
+   \label{eqn:longone}
+\end{eqnarray}
+Rearranging the RHS of equation \ref{eqn:longone}, obtain
 \begin{equation}
   \label{eqn:defwrialt}
-  J_{\rm WRI}^{\alpha}[c,d,q] = \frac{1}{2}\langle r[c], W_{\alpha}[c] r[c]\rangle
+  J_{\rm WRI}^{\alpha}[c;d,q] = \frac{1}{2}\langle r[c], W_{\alpha}[c] r[c]\rangle,
 \end{equation}
@@
   \label{eqn:defwriwt}
-  W_{\alpha}[c] = \frac{\alpha^2}{2}(S[c]S[c]^T+\alpha^2I)^{-1}
+  W_{\alpha}[c] = \frac{\alpha^2}{2}(S[c]S[c]^T+\alpha^2I)^{-1}.
 \end{equation}
@@
 equivalent to equations \ref{eqn:defwrialt}, \ref{eqn:defwriwt}.
+
+Note that $\alpha^2$ must have the same units as $S[c]S[c]^T$ (or
+$S[c]^TS[c]$), hence {\em $W_{\alpha}[c]$ is dimensionless}. Of course
+this also follows from the definition of
+$ J_{\rm WRI}^{\alpha}$, which must have the same units as $J_{\rm FWI}$.
+
+To end this discussion, it must stipulated that in any penalty method,
+control of the penalty parameter has a large influence on the speed of
+convergence. \cite{Aghamiry:19} use an augmented Lagrangian algorithm
+to minimize the influence of the penalty weight choice. Alternatively,
+one can use a version of the Discrepancy Principle to adjust $\alpha$
+dynamically \cite[]{FuSymes2017discrepancy}, as the WRI problem has
+the necessary features described in that paper. Since the properties
+of WRI to be established in the next section are actually independent
+of $\alpha>0$, I leave the matter of $\alpha$ control at that.
 
@@
       \frac{1}{2c}e\left(t +  \frac{|z_r-z|}{c}\right), \,z_{\rm min} \le z \le z_{\rm max};\\
-      0, else.
+      0, \mbox{ else.}
     \end{array}
@@
   \label{eqn:modhomnormal}
-  S[c]S[c]^T = \frac{z_{\rm max}-z_{\rm min}}{4c^2}I
+  S[c]S[c]^T = \frac{z_{\rm max}-z_{\rm min}}{4c^2}I.
 \end{equation}
@@
 \[
-  W_{\alpha}[c] = u(c) I,
+  W_{\alpha}[c] = u_{\alpha}[c] I,
 \]
@@
   \label{eqn:wriwthom}
-u(c) = \frac{\alpha^2}{2}\left(\frac{z_{\rm max}-z_{\rm min}}{4c^2}+\alpha^2\right)^{-1}.
+u_{\alpha}[c] = \frac{\alpha^2}{2}\left(\frac{z_{\rm max}-z_{\rm min}}{4c^2}+\alpha^2\right)^{-1}.
 \end{equation}
@@
   \label{eqn:wriscrewed}
-  J_{\rm WRI}^{\alpha}[c,d,q]=u[c]J_{\rm FWI}[c,d,w]
+  J_{\rm WRI}^{\alpha}[c;d,q]=u_{\alpha}[c]J_{\rm FWI}[c;d,w].
 \end{equation}
@@
    \label{eqn:defbeta}
-\beta = \frac{z_{\rm max}-z_{\rm min}}{c_*^2} - 4 \alpha^2
+\beta = \frac{z_{\rm max}-z_{\rm min}}{c_*^2} - 4 \alpha^2.
 \end{equation}
@@
   For $L$ as defined in \ref{eqn:defL}, and $\lambda <
-  \lambda_0$, the minimizer of $J_{\rm WRI}^{\alpha}[c,\dl,\wl \delta(\cdot-z_s)]$ on the
+  \lambda_0$, the minimizer of $J_{\rm WRI}^{\alpha}[c;\dl,\wl \delta(\cdot-z_s)]$ on the
   complement of $[c_*-L\lambda, c_*+L\lambda]$ is
@@
   \ref{eqn:fwiscrewed}, if $|c-c_*| > L\lambda$,
-  \[
-    J_{\rm WRI}^{\alpha}[c,\dl,\wl\delta(\cdot-z_s)] =
+\begin{eqnarray}
+    J_{\rm WRI}^{\alpha}[c;\dl,\wl\delta(\cdot-z_s)] &=&
     \frac{\alpha^2}{2}\left(\frac{z_{\rm max}-z_{\rm
-          min}}{4c^2}+\alpha^2\right)^{-1}\frac{1}{2}\left(\frac{1}{4c^2}+\frac{1}{4c_*^2}\right)\|w_1\|^2
-  \]
-  \[
-    =
-    \frac{\alpha^2}{4}\frac{1+\frac{c^2}{c_*^2}}{z_{\rm 
-        max}-z_{\rm min}+4c^2\alpha^2}
-  \]
-  This linear fractional function of $c^2$ is increasing, decreasing,
+          min}}{4c^2}+\alpha^2\right)^{-1}\frac{1}{2}\left(\frac{1}{4c^2}+\frac{1}{4c_*^2}\right)\|w_1\|^2\nonumber\\
+   & =&    \frac{\alpha^2}{4}\frac{1+\frac{c^2}{c_*^2}}{z_{\rm 
+        max}-z_{\rm min}+4c^2\alpha^2}\|w_1\|^2.
+  \label{eqn:yetanutta}
+\end{eqnarray}
+  The linear fractional function of $c^2$ on the RHS of equation
+  \ref{eqn:yetanutta} is increasing, decreasing,
   or constant if $\beta>0, \beta<0$ or $\beta=0$, respectively.
@@
 independent of $\alpha$, and the minimizer of
-$J_{\rm WRI}[c,\dl,\wl\delta(\cdot-z_s)]$ in this region (away from
+$J_{\rm WRI}[c;\dl,\wl\delta(\cdot-z_s)]$ in this region (away from
 $c_*$) is $c=c_{\rm min}$ for small enough $\alpha$. Therefore taking
@@
 
+\section{Numerical Illustration}
+The Ricker wavelet (normalized second derivative of a Gaussian) is a
+commonly used synthetic seismic signal, as it closely resembles short
+segments of field data traces. Figure \ref{fig:wavelet} displays a Ricker
+wavelet of 20 Hz peak frequency and zero phase
+(symmetric about $t=0$). The Ricker wavelet
+of 1 Hz peak frequency decreases by almost three orders of magnitude at $t=\pm
+1$, relative to its peak at $t=0$, so could reasonably be truncated to
+$[-1,1]$ with a bit of smoothing, hence could be taken as the ``mother
+wavelet'' $w_1$ of the discussion in section 2. Then the signal
+displayed in Figure \ref{fig:wavelet} is $w_{\lambda}$ for $\lambda=0.05$.
+
+Figure \ref{fig:data} shows the simulated data $d=S[c_*]w_{\lambda}$
+for $c_*$ = 2.5 km/s and source-receiver offset $|z_r-z_s|=1$ km. The
+time shift is calculated via piecewise linear interpolation on a grid
+with time interval $\Delta t = 0.001$ s.
+
+Figures \ref{fig:j0} and \ref{fig:j3} show FWI and VPM-reduced WRI
+objectives ($J_{\rm FWI}$ and $J_{\rm WRI}^{\alpha}$ respectively) as functions of slowness
+(reciprocal velocity). The FWI (mean square data residual) values are
+computed directly from equation \ref{eqn:deffwi}, using the
+trapezoidal rule to approximate the integral. The VPM-reduced WRI
+values are calculated on the basis of the weighted-residual formulas
+\ref{eqn:defwrialt} and \ref{eqn:defwriwt}, using the result
+\ref{eqn:wriwthom} which holds for the 1D transmission problem (that
+is, the relation \ref{eqn:wriscrewed}).
+
+The Figures \ref{fig:j0} and \ref{fig:j3} differ in choice of penalty
+weight $\alpha$ for the WRI objective. For the 1D transmission
+problem, the unit of $\alpha$ is (time)/$\sqrt{\rm (length)}$, so
+s/$\sqrt{\rm km}$ in the units used here. For Figure \ref{fig:j0},
+$\alpha = 0.1$ s/$\sqrt{\rm km}$, whereas for Figure \ref{fig:j3},
+$\alpha = 1.0$ s/$\sqrt{\rm km}$, For $J_{\rm FWI}$, $c=c_{\rm min}$
+is a local minimizer, consistent with Theorem \ref{thm:thm1}, and corresponding to the largest admissible slowness.
+For the parameters
+of Figure \ref{fig:j0}, the quantity $\beta$ defined in equation
+\ref{eqn:defbeta} is positive, so $c=c_{\rm min}$ is a local minimizer
+of $J_{\rm WRI}^{\alpha}$, corresponding to the largest admissible slowness, per Theorem \ref{thm:thm2}. For
+the parameters pertaining to Figure \ref{fig:j3}, $\beta$ is
+negative, so $c=c_{\rm max}$ is a local minimizer of $J_{\rm
+  WRI}^{\alpha}$, corresponding to the smallest admissible slowness.
+
+In all cases, the global minimizer occurs at the target velocity
+$c_*$=2.5 km/s (slowness = 0.4 s/km). It is ``defended'' by close-in
+local minimizers roughly half a wavelength on ether side. Our analysis
+would be just as valid for non-oscillatory wavelets, so cannot predict
+this smaller-scale structure.
+
+\input{madfigs.tex}
+%\input{subfigs.tex}
+
 \section{Discussion}
@@
 conclusions of this work: that at least for the 1D acoustic
-transmission inverse problem, both $J_{\rm FWI}[c,d,w]$ and
-$J_{\rm WRI}^{\alpha}[c,d,q]$ exhibit local minima (in $c$) far from the global
+transmission inverse problem, both $J_{\rm FWI}[c;d,w]$ and
+$J_{\rm WRI}^{\alpha}[c;d,q]$ exhibit local minima (in $c$) far from the global
 minimum for consistent data, the domain of attraction of the
@@
 
-First, a consequence of the results proven here: $J_{\rm FWI}[c,d,w]$
-and $J_{\rm WRI}[c,d,q]$ are not smooth as joint functions of model
+First, a consequence of the results proven here: $J_{\rm FWI}[c;d,w]$
+and $J_{\rm WRI}^{\alpha}[c,;d,q]$ are not smooth as joint functions of model
 ($c$) and data ($d$) vectors. If they were, their derivatives would be
@@
 These results
-concern optimization problems of the genral form
+concern optimization problems of the general form
 \begin{equation}
@@
   \mbox{Given }d\mbox{, find }c\mbox{ to extremize }
-   J[c,d] = \langle G[c]d, A[c] G[c]d \rangle
+   J[c,d] = \langle G[c]d, A[c] G[c]d \rangle.
 \end{equation}
@@
 $c$ is a vector of material parameters to be estimated, $G[c]$ is a
-$c$-dependent family of operators having the same nature as the
-simulation operators $S[c]$ and $S_p[c]$ figuring in the preceding
-discussion, and $A[c]$ is an operator-valued function of $c$, on the range
-of $G[c]$.
+$c$-dependent family of operators, whose common range is a Hilbert
+space with inner product $\langle \cdot, \cdot \rangle$.
+$A[c]$ is an operator-valued function of $c$, with domain and the
+range equal to the range of $G[c]$.
 
 \cite{StolkSymes:03} assume that the
-operator-valiued function $G[c]$ is of a class typical of modeling
+operator-valued function $G[c]$ is of a class typical of modeling
 operators for wave equation inverse problems, or their inverses or
-adjoints. The precise characterization of the so-called {\em microlocally elliptic Fourier Integral
+adjoints. The precise characterization of these so-called {\em microlocally elliptic Fourier Integral
 Operators} is quite technical \cite[]{Dui:95}. Roughly speaking, such operators map high-frequency localized wave packets to other
 such packets with well-defined changes of position and direction of
-oscillation. If the dependence of such an operator $G[c]$ on $c$ is of sufficiently full rank, in
+oscillation. The simulation operators $S[c]$ and $S_p[c]$ figuring in the preceding
+discussion are particularly simple examples of this type.
+
+If the dependence of such an operator $G[c]$ on $c$ is of sufficiently full rank, in
 the sense that destination packets can be shifted in any direction by
@@
 {\em pseudodifferential operator} - again, a class of operators whose
-precise definition is quite technical. However these operators also
+precise definition is quite technical \cite[]{Dui:95,Tay:81}. However these operators also
 have a rough characterization: they do not change the location of
@@
 extended to $(-\infty,\infty)$ it would have a left inverse too). From
-the defintion \ref{eqn:dfam} of the data family $\dl$, one sees that
+the definition \ref{eqn:dfam} of the data family $\dl$, one sees that
 $\wl = S_p[c_*]^{-1}\dl$. Therefore
-\[
-  J_{\rm FWI}[c,\dl,\wl] = \frac{1}{2}\|(I-S_p[c]S_p[c_*]^{-1})\dl\|^2
-\]
-\[
-  =\frac{1}{2}(\|\dl\|^2 + \|S_p[c]S_p[c_*]^{-1}\dl\|^2) + \langle \dl,
-  S_p[c]S_p[c_*]^{-1}\dl\rangle
-\]
+\begin{eqnarray}
+  \label{eqn:anutta}
+  J_{\rm FWI}[c;\dl,\wl] &=&
+                             \frac{1}{2}\|(I-S_p[c]S_p[c_*]^{-1})\dl\|^2 \nonumber\\
+ & =&\frac{1}{2}(\|\dl\|^2 + \|S_p[c]S_p[c_*]^{-1}\dl\|^2) + \langle \dl,
+  S_p[c]S_p[c_*]^{-1}\dl\rangle.
+\end{eqnarray}
 For the operator family $S_p[c]$ defined above, it is easy to see that
-the second term is smooth in $c$, and the first is constant. The third 
-can be rewritten as 
-\[
+the second term in the right hand side of equation \ref{eqn:anutta} is
+smooth in $c$, and the first is constant. The third can be rewritten
+as
+\begin{equation}
+  \label{eqn:evenmore}
   \langle \dl,  S_p[c]S_p[c_*]^{-1}\dl\rangle = \langle S[c]^T\dl, 
-  (S_p[c]^T S_p[c_*])^{-1}S_p[c]^Td\rangle 
-\]
-which has the form \ref{eqn:gen1} with the choices $G[c]=S_p[c]^T$,
+  (S_p[c]^T S_p[c_*])^{-1}S_p[c]^Td\rangle.
+\end{equation}
+The RHS of equation \ref{eqn:evenmore} has the form \ref{eqn:gen1} with the choices $G[c]=S_p[c]^T$,
 $A[c]=(S_p[c]^T S_p[c_*])^{-1}$. A similar manipulation exhibits
@@
            < 0 \mbox{ if }c>c_*+O(\lambda)
-         \end{array}
+         \end{array}.
        \right.
@@
 straightforward algebraic properties of WRI and appear to link it
-closely to FWI in any wave propagation setting. As explained in the
+closely to FWI in any wave propagation setting. Since the commonly
+cited arguments for reduced nonlinearity of WRI over FWI also apply in
+the context of the 1D transmission problem, these arguments cannot be
+regarded as having any actual force. As explained in the
 discussion section, even mere smoothness of a quadratic form objective
@@
 
+\section{Acknowledgements}
+The author thanks Felix Herrmann, Tristan van Leeuwen, David Yingst,
+and Chao Wang for informative conversations about Wavefield
+Reconstruction Inversion. Comments from Joakim Blanch and John
+Washbourne were very helpful in improving the manuscript, as were
+excellent suggestions made by the two anonymous reviewers. The sponsors of The Rice
+Inversion Project provided long-term support for the author's
+research, leading to the work reported here.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 p(z,t)=a\left(t -\frac{|z-z_s|}{c}\right), \, v(z,t)=\mbox{sgn}(z-z_s) b\left(t -
-  \frac{|z-z_s|}{c}\right)
+  \frac{|z-z_s|}{c}\right).
 \]
@@
 \label{eqn:sol1dptsrc}
-p(z,t) &=& \frac{1}{2c}w\left(t - \frac{|z-z_s|}{c}\right) \nonumber \\
-v(z,t) &=& \mbox{sgn}(z-z_s)\frac{1}{2\rho c^2}w\left(t -\frac{|z-z_s|}{c}\right)
+p(z,t) &=& \frac{1}{2c}w\left(t - \frac{|z-z_s|}{c}\right), \nonumber \\
+v(z,t) &=& \mbox{sgn}(z-z_s)\frac{1}{2\rho c^2}w\left(t -\frac{|z-z_s|}{c}\right).
            \nonumber \\
@@
 \begin{eqnarray}
-\label{eqn:awe1d}
+\label{eqn:awe1dapp}
 \frac{\partial p}{\partial t} +\rho c^2\frac{\partial 
@@
 \[
-  f(z,t) = \int dz_1\,f(z_1,t)\delta(z-z_1)
+  f(z,t) = \int dz_1\,f(z_1,t)\delta(z-z_1),
 \]
@@
 p(z,t) &=& \frac{1}{2c}\int dz_1 f\left(z_1,t -
-           \frac{|z-z_1|}{c}\right) \\
+           \frac{|z-z_1|}{c}\right), \\
   \label{eqn:sol1dv}
-v(z,t) &=& \frac{1}{2\rho c^2} \int dz_1 \mbox{sgn} (z-z_1) f\left(z_1,t - \frac{|z-z_1|}{c}\right)
+v(z,t) &=& \frac{1}{2\rho c^2} \int dz_1 \mbox{sgn} (z-z_1) f\left(z_1,t - \frac{|z-z_1|}{c}\right).
 \end{eqnarray}
@@
   v_0}{\partial z} &=&
-                            f_0  \nonumber\\
-\rho \frac{\partial v_0}{\partial t} +\frac{\partial p_0}{\partial z} &=& g_0 \nonumber\\
+                            f_0,  \nonumber\\
+\rho \frac{\partial v_0}{\partial t} +\frac{\partial p_0}{\partial z} &=& g_0, \nonumber\\
 \end{eqnarray}
@@
   \label{eqn:rhscut}
-  f_0(z,t) &=& -\rho c^2 v(z,t) \frac{\partial}{\partial z}(1-\phi(z))\nonumber\\
+  f_0(z,t) &=& -\rho c^2 v(z,t) \frac{\partial}{\partial z}(1-\phi(z)),\nonumber\\
   g_0(z,t) &=& -p(z,t) \frac{\partial}{\partial z}(1-\phi(z))
@@
   \frac{1}{\rho c^2} \frac{\partial^2 p_0}{\partial t^2} -
-  \frac{1}{\rho}\frac{\partial^2 p_0}{\partial z^2} &=& F\nonumber \\
-  p &=& 0, t\ll 0
+  \frac{1}{\rho}\frac{\partial^2 p_0}{\partial z^2} &=& F,\nonumber \\
+  p &=& 0, t\ll 0,
 \end{eqnarray}
@@
   F = \frac{1}{\rho c^2}\frac{\partial f_0}{\partial t} -
-  \frac{1}{\rho}\frac{\partial g_0}{\partial z}
+  \frac{1}{\rho}\frac{\partial g_0}{\partial z}.
 \end{equation}
@@
   f(z,t) = \rho c^2 \int_{-\infty}^t \,ds\,F(z,t) =  f_0(z,t) - c^2\int_{-\infty}^{t}\,ds\,\frac{\partial
-    g_0}{\partial z}(z,s)
+    g_0}{\partial z}(z,s).
 \end{equation}
@@
   p_1 &=& p_0,\\
-  v_1 &=& \frac{1}{\rho}\int_{-\infty}^t\frac{\partial p_0}{\partial z}
+  v_1 &=& \frac{1}{\rho}\int_{-\infty}^t\frac{\partial p_0}{\partial z},
 \end{eqnarray*}
@@
 \end{equation}
-that is, that using RHS $f$ in \ref{eqn:awe1d} produces the same
+that is, that using RHS $f$ in \ref{eqn:awe1dapp} produces the same
 pressure field near $z=z_r$ as does the point source in
@@
   c^2\frac{\partial^2}{\partial z^2}(1-\phi(z))\int_{-\infty}^t
-  \,ds\,p(z,s)
+  \,ds\,p(z,s),
 \]
@@
   \label{eqn:defextop}
-  =E[c]w(z,t)
+  =E[c]w(z,t),
 \end{equation}
@@
   \label{eqn:extapp}
-S[c]\circ E[c] = S_p[c]
+S[c]\circ E[c] = S_p[c],
 \end{equation}
@@
 \bibliographystyle{seg}
-\bibliography{../../bib/masterref}
+\bibliography{../../../bib/masterref}

### 2020-08-01 repro/papers/wri1d/rev1/submit/subfigs.tex
rev1 and submission versions

\begin{figure}[ht]   
\begin{center}
\includegraphics[width=\textwidth]{wavelet.pdf}
\caption{Ricker Wavelet, zero phase, 20 Hz center frequency}
\label{fig:wavelet}
\end{center}
\end{figure}

\begin{figure}[ht]   
\begin{center}
\includegraphics[width=\textwidth]{data.pdf}
\caption{Predicted data for 1D transmission 
  problem, source wavelet = 20 Hz Ricker wavelet of Figure 
  \ref{fig:wavelet}, source-receiver offset = 1 km, velocity = 2.5 
  km/s.}
\label{fig:data}
\end{center}
\end{figure}

\begin{figure}[ht]   
\begin{center}
\includegraphics[width=\textwidth]{j0.pdf}
\caption{Blue curve = FWI objective, red curve =
  VPM-WRI objective, plotted as functions of slowness, $\alpha$= 0.1
  s/$\sqrt{\rm km}$.}
\label{fig:j0}
\end{center}
\end{figure}

\begin{figure}[ht]   
\begin{center}
\includegraphics[width=\textwidth]{j3.pdf}
\caption{Blue curve = FWI objective, red curve =
  VPM-WRI objective, plotted as functions of slowness, $\alpha$= 1.0
  s/$\sqrt{\rm km}$.}
\label{fig:j3}
\end{center}
\end{figure}

### 2020-08-03 repro/papers/discrep/paper.tex
interim

@@
 
+Note also that $F[m]$ is surjective for every $m \in M$.
+
+In computational practice, $W$ will have to be replaced by a
+finite-dimensional subspace of $L^2(\bR)$. Many such choices will
+implicitly limit the support of $w \in W$ to a bounded interval, say
+$[T_{\rm min},T_{\rm max}]$. To maintain the surjective property,
+these bounds should be chosen so that $[t_{\rm min}, t_{\rm max}]
+\subset [T_{\rm min}+mr,T_{\rm max}+mr]$ for all $m \in M$, that is,
+\begin{eqnarray}
+  \label{eqn:dombounds}
+  t_{\rm min} &\ge & T_{\rm min}+m_{\rm max}r,\nonumber\\
+  t_{\rm max} &\le & T_{\rm max}+m_{\rm min}r.
+\end{eqnarray}
+We will ignore these computational necessities in this work,
+maintaining the defintion $W=L^2(\bR)$.
+
 \section{The Inverse Problem}
@@
 Then for 
-$m \in M$, $0 <\lambda \le \lambda_{\rm max}$, $F[m] \in {\cal B}(\lW,D)$ is coercive.
+$m \in M$, $0 <\lambda \le \lambda_{\rm max}$, $\lF[m] \in {\cal B}(\lW,D)$ is coercive.
 
@@
 \begin{theorem}
+  \label{thm:epjgen}
+  Assume the hypotheses of Theorem \ref{thm:norminv}. Then
+  \begin{equation}
+  \label{eqn:residnormgen}
+  e[m,\aw[m,d];d] = \frac{1}{2}\int \,dt\,(4\pi r \alpha a(t-mr))^4(1 +
+  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)^2
+\end{equation}
+\begin{equation}
+  \label{eqn:anninormgen}
+  p[m,\aw[m,d];d] = \frac{1}{2}\int \,dt\,(4\pi r a(t-mr))^2(1 +
+  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)
+\end{equation}
+
+\begin{equation}
+  \label{eqn:expjgen}
+\tJa[m;d] = \frac{1}{2}\int\,dt\,(4\pi r \alpha a(t-mr))^2(1+(4\pi r \alpha 
+a(t-mr))^2)^{-1}d(t)^2. 
+\end{equation}
+Finally, if $a \in W^{1,1}(\bR)$ (distributions with integrable 
+derivatives), then $\tJa[\cdot;d]$ is differentiable, and 
+\begin{equation}
+  \label{eqn:dexpjgen}
+  \frac{d}{dm}\tJa[m;d] = -(4 \pi r \alpha)^2 \int \,dt \, 
+  \left(a\frac{da}{dt}\right)(t-mr)(1+(4\pi r \alpha 
+  a(t-mr))^2)^{-2}d(t)^2. 
+\end{equation}
+\end{theorem}
+
+\begin{corollary}
   \label{thm:epjnonoise}
@@
 \end{equation}
-so finally
-\begin{equation}
-\label{eqn:expjgen}
-\tJa[m;d] = =\frac{\alpha^2}{2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
-  a(t-(m-m^*)r)^2)^{-1}w_*(t)^2.
-\end{equation}
-\end{theorem}
+so 
+\begin{equation}
+\label{eqn:expjnonoise}
+\tJa[m;d] = \frac{\alpha^2}{2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
+  a(t-(m-m_*)r)^2)^{-1}w_*(t)^2. 
+\end{equation}
+Finally, if $a \in W^{1,1}(\bR)$, then $\tJa[\cdot;d]$ is differentiable, and 
+\begin{equation}
+  \label{eqn:dexpjnonoise}
+  \frac{d}{dm}\tJa[m;d] = -r \alpha^2 \int \,dt \, 
+  \left(a\frac{da}{dt}\right)(t-(m-m_*)r)(1+(4\pi r)^2 \alpha^2 
+  a(t-(m-m_*)r)^2)^{-2}w_*(t)^2. 
+\end{equation}
+\end{corollary}
 
 I willl now examine two possible choices of penalty multipler $a$ in
-defintion \ref{eqn:annmult}.
+definition \ref{eqn:annmult}.
 
@@
 \cite[]{Plessix:00a,LuoSava:11,Warner:14,HuangSymes:SEG15a,Warner:16,HuangSymes:GEO17}. It
-satisfies condition \ref{eqn:abnd} with $C=\lambda$.
+satisfies condition \ref{eqn:abnd} with $C=\lambda$, and is a member
+of $W^{1,1}(\bR)$, so that the resulting objective $\tJa$ is
+differentiable, with derivative given by equation \ref{eqn:dexpgen},
+or \ref{eqn:dexpjnonoise} in the noise-free case.
 
@@
     min}-mr,t_{\rm max}-mr]\subset [-\tau,\tau]$, with $\tau$ defined
-  in \ref{eqn:taudef}. Therefore for $t \in  [t_{\rm min}-mr,t_{\rm
-    max}-mr]$, $a(t) = |t|$, and reason analogous to the proof of
-  Theorem \ref{thm:norminv}, item 2, shows that
-  \begin{equation}
-    \label{eqn:solnnon}
-    \aw[m,d](t)= \left(1+ (4\pi r)^2\alpha^2 t^2\right)^{-1}w_*\left(t+(m-m_*)r\right) 
-  \end{equation}
-  Straightforward modification of the proof of Theorem
-  \ref{thm:epjnonoise} gives
-  \begin{equation}
-    \label{eqn:expj}
-    \tJa[m;d] =\frac{\alpha^2}{2}\int\,dt\,(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
-    (t-(m-m^*)r)^2)^{-1}w_*(t)^2.
-  \end{equation}
-  A brief calculation yields
+  in \ref{eqn:taudef}. Therefore, $a(t) = |t|$ $a a'(t) = t$ in the
+  support of the integrand on the RHS of equation
+  \ref{eqn:depjnonoise}, which therefore 
+  becomes (after change of integration variable)
   \begin{equation}
     \label{eqn:gradfinal}
-    \frac{\partial}{\partial m}\tJa[m;d] = -r\alpha^2\int \,dt\,\frac{t}{(1+(4\pi r)^2\alpha^2 t^2)^2}w_*(t+(m-m_*)r)^2.
+    \frac{d}{dm}\tJa[m;d] = -r \alpha^2 \int \,dt \, 
+  t(1+(4\pi r)^2 \alpha^2 
+  t^2)^{-2}w_*(t+(m-m_*))^2.
   \end{equation}
@@
   \]
-  If $m > m_*+\lambda/r$, then the entire interval of integration is a proper
+  Suppose that $\mu < \lambda$ and $w_* \in W_{\mu}$. Then for
+  $t+(m-m_*)r \in \mbox{supp }w_*$, $t\le\mu-(m-m_*)r$.
+  If $m > m_*+\lambda/r$, then $t >\mu-\lambda<0$, so the support of
+  the $[-\mu-(m-m_*)r, \mu-(m-m_*)r]$ is a proper
   subset of the negative half-axis. Consequently the first factor in the
@@
   negative, so the derivative
-  is positive. Similar reasoning applies to the case $m <
+  is positive. In fact, the quantity displayed above is
+  \[
+    \le 
+  \]
+  Similar reasoning applies to the case $m <
   m_*-\lambda/r$. Thus
@@
   \label{eqn:defdatanoisy}
-  d = F[m_*]w_* + n
+  d = F[m_*]w_* + n = d_*+n,
 \end{equation}
@@
 \ref{eqn:probstat0}.
+
+Recall that $F[m]$ is surjective for every $m \in M$. Define for each
+$m \in M$
+\begin{equation}
+  \label{eqn:defnoisewavelet}
+  w_{n}[m] = \mbox{arg}\min_w\{\|w\|: F[m]w=n\}.
+\end{equation}
+That is,
+\[
+  w_n[m](t) =
+  \left\{
+    \begin{array}{c}
+      4\pi r n(t+mr),\,t \in [t_{\rm min}-mr,t_{\rm max}-mr];\\
+      0, \mbox{ else.}
+    \end{array}
+  \right.
+\]
+Set
+\begin{equation}
+  \label{eqn:defwflat}
+  w_{\flat} = w_* + w_n[m_*];
+\end{equation}
+then
+\begin{equation}
+  \label{eqn:datacomplete}
+  d = F[m_*]w_{\flat}.
+\end{equation}
+That is, we can treat the data as if it were noise-free, for the price
+of altering the wavelet. The altered wavelet is of course not a member
+of $\lW$.
+
+From the proof of Theorem \ref{thm:norminv}, for the choice
+$a(t)=\min(|t|,\tau)$ with $\tau$ defined in \ref{eqn:taudef},
+\begin{equation}
+  \label{eqn:solnoise}
+  \aw[m,d](t)= \left(1+ (4\pi r)^2\alpha^2 t^2\right)^{-1}w_{\flat}\left(t+(m-m_*)r\right).
+\end{equation}
+and from Theorem \ref{thm:epjnonoise}
+\begin{equation}
+\label{eqn:expjgen}
+\tJa[m;d] = =\frac{\alpha^2}{2}\int\,dt\,(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
+  (t-(m-m^*)r)^2)^{-1}w_{\flat}(t)^2.
+\end{equation}
+whence via the same short computation as before
+\[
+  \frac{\partial}{\partial m}\tJa[m;d] = -r\alpha^2\int \,dt\,\frac{t-(m-m_*)r}{(1+(4\pi r)^2\alpha^2 (t-(m-m_*)r)^2)^2}w_{\flat}(t)^2. 
+\]
+\[
+  = -r\alpha^2 \int \, dt \, (t-(m-m_*)r)\aw[m;d](t)^2 
+\]  
+\[
+  = -r\alpha^2\left\{ \int_{-\lambda}^{\lambda} \, dt \, 
+    (t-(m-m_*)r)(w_*(t)^2 +2w_*(t)w_n[m_*](t))\right\}
+\]
+\begin{equation}
+  \label{eqn:gradnoise}
+-r\alpha^2\int_{-\infty}^{\infty} \, dt \, a(t-(m-m_*)r) w_n[m_*](t)^2.
+\end{equation}
+The last line recalls the definition of $a$ for this example
+(\ref{eqn:deftau}).
+
+We will follow the pattern laid down in the proof of Theorem
+\ref{thm:rampgood}. However to obtain an estimate depending on some
+measure of noise, it is necessary to explicitly quantify the bounds
+within which stationary points of $\tJa$ are located. Pick $\dlam >0$,
+and suppose that $|m-m_*|r > \lambda + \dlam$. Then for $t \in
+[-\lambda,\lambda]$,
+\begin{eqnarray*}
+  (m-m_*)r >   \lambda +\dlam & \Rightarrow & t-(m-m_*)r < -\dl, \\
+  (m-m_*)r < -\lambda +\dlam & \Rightarrow & t-(m-m_*)r >   \dl.
+\end{eqnarray*}
+In the first case,
 

### 2020-08-19 repro/papers/discrep/paper.tex
interim

--- 2020-07-10 repro/papers/discrep/paper.tex
@@
-\title{A Discrepancy-Based Algorithm for Extended Inversion}
+\title{Solution of an Acoustic Transmission Inverse Problem by Extended Inversion}
 \author{William W. Symes, Orcas Island, Huiyi Chen* and Susan E. Minkoff, The University of Texas at Dallas}
@@
   the discrepancy principle, namely the determination of a
-  satisfactory estimate of the penalty weight: the initial weight may
-  be set to zero, and the dynamic algorithm updates it until
+  satisfactory initial estimate of the penalty weight: the initial weight may
+  be set to zero, and the dynamic algorithm updates the weight until
   convergence is achieved.
@@
 \section{Introduction}
-Full Waveform Inversion (FWI), or estimation of earth structure by
-model-driven least squares data fitting, is now well-established as a
-useful tool for probing the earth's subsurface
-\cite[]{VirieuxOperto:09,Fichtner:10}. However, so-called ``cycle-skipping'', the tendency of iterative FWI
-algorithms to stagnate at suboptimal and geologically uninformative
-earth models, still impedes its use. Because the computational size of field inversion tasks
-is very large, only iterative local (descent) minimization of the data
-misfit function is computationally feasible. However local
-descent methods avoid suboptimal stagnation only if initial models are
-already quite close to optimal, in the sense of predicting the arrival
-times of seismic events to within a small multiple of a dominant
-wavelength \cite[]{GauTarVir:86,Plessix:10}.
-
-This paper concerns one of the several ideas that have been advanced to
-overcome cycle-skipping, namely so-called extended inversion
-\cite[]{geoprosp:2008}. ``Extended'' signifies that addional degrees
-of freedom are provided to the modeling process, in the hope of
-opening up more effective routes to geologically informative models
-with acceptable data fit. Since these extended degrees of freedom are
-not part of the basic physics chosen to model the data acquisition
-process, they should be suppressed in the eventual solution. Extended
-inversion methods differ by the choice of additional degrees of
-freedom, and by choice of penalty applied to eliminate them in the final
-result.
-
-Many of these extended inversion concepts sound plausible, and appear
-to work at least to some extent as one might hope from their heuristic
-justifications. However very few of these approaches have been
-underwritten by mathematical argument: in essence, they are mostly
-justified only ``in the rear-view mirror'', with no assurance that
-failure is not just around the corner, at the next example. On top of
-that, some of these approaches, for example those based on the
-computationally attractive Variable Projection Method (``VPM'') of
-\cite{GolubPereyra:03}, are cast in such form that the reasons for
-success are not readily apparent.
-
-This note shows exactly how extended inversion leads to successful solution
-of a very simple wave propagation inverse
-problem, which asks that a homogeneous velocity field be deduced from
-one trace at known offset. I put forward this inverse problem and
-extension-based solution not because there are not simpler
-ways of answering the question it poses - there certainly are - but
-because the formal ingredients of waveform-based velocity estimation
-in this very simple setting are common to many similar extended
-inversion algorithms, and because in this case every computation can
-be done analytically, nearly to completion. In particular, it becomes
-clear why the VPM gradient formula produces a constructive update,
-with no possibility of stagnation away from the global minimum.
-
-The extended inversion approach
-developed here uses a {\em source extension}, in which source
-parameters form the additional degrees of freedom. This type of
-extension presumes that the actual or target source is constrained in
-some way; the extended source is allowed to violate the
-constraint. For recent overview of source extension methods, see
-\cite{HuangNammourSymesDollizal:SEG19}. Source
-extension methods have computational complexity approximately the same
-as that of FWI, a signal advantage over the alterantive medium
-extension class described for instance in \cite[]{geoprosp:2008}.
-
-For the problem considered here, the source model amounts to a
-wavelet, and the target wavelet is assumed to be non-zero only in a
-short time interval (an approximate impulse, perhaps as the result of
-signature deconvolution). The extension consists in permitting energy
-to spread in time at intermediate iterations of the inversion. A
-penalty for energy spread drives the extended source towards a focused
-source approximately satisfying the assumed constraint. Not all
-penalties are created equal. Two plausible penalty operators are
-studied here, both multiplication operators: one by the characteristic
-function of the target support, the other by a smooth function of
-time. Only the smooth multiplier avoids cycle-skip. This distinction
-is quite general, as will be explained in the Discussion section.
-
-%despite not incorporating explicitly the extension measure
-%(``annihilator'') central to the approach.
-
-I begin with a quick sketch of constant density acoustics, and
-describe the single-trace transmission inverse problem.. For
-completeness, I show how the standard FWI approach to this problem,
-with the small support of the source imposed as a hard constraint,
-generates multiple local minima that will be found by any descent
-method unless the initial slowness estimate predicts travel time from
-source to receiver with an error on the order of a wavelength. The
-next section describes the extended source objective, and the reduced
-objective produced by VPM, in general terms. As VPM eliminates the
-extended source, this function depends only on the velocity, just as
-does the FWI objective.  Of course, it also depends on the operator
-used to define the penalty (``annihilator''). I propose two choices: a
-soft version of the support constraint, and an even softer constraint,
-namely multiplication by time. A nearly-explicit calculation of the
-VPM gradient for these two choices shows that the first leads again to
-multiple local minima far from the target solution, whereas the second
-does not. In fact, in the second case, the only stationary points are
-``within a wavelength'' of the correct velocity, used to build the
-data: that is, cycle-skipping cannot occur. For this second, more
-successful penalty, I introduce the discrepancy based dynamic
-weighting algorithm, and show how it accelerates the convergence
-towards a local minimizer, which has been shown to be an acceptable
-approximation to a global minimizer. The paper ends with a discussion
-of the parallels between the calculations presented here and the
-structure of other extended inversion methods applicable to
-field-scale velocity estimation, and the critical role that the
-differential nature of the extension penalty plays in the success of
-this and other extension methods.
-
-I shall use the abbreviations ${\cal B}(X,Y)$ and ${\cal I}(X,Y)$ for
+Model-based parameter estimation by least squares data fitting is a
+widely used and successful technique for data analysis in science and
+engineering \cite[]{Bertero:98,Vogel:02}, particularly in the
+geosciences \cite[]{Parker:94,Tarantola:05}. In its application to
+seismology, least-squares data fitting has come to be known as Full
+Waveform Inversion (FWI), and is now well-established as a useful tool
+for probing the earth's subsurface
+\cite[]{VirieuxOperto:09,Fichtner:10,Schuster:17}. However, in its
+application to seismology, FWI encounters a serious
+practical obstacle, so-called ``cycle-skipping''. This term signifies
+the tendency of the mean-square error function to exhibit many
+stationary points, and local minima far from the global minimizer (we
+show an explicit example of this phenomenon later in this paper). Because of the
+typical dimensions of field FWI problems, only local gradient-based
+minimization algorithms are computationally practical, and these find
+local minimizers, so may stagnate at suboptimal and geologically
+uninformative earth models. Local descent methods avoid suboptimal
+stagnation only if initial models are already quite close to optimal,
+in the sense of predicting the arrival times of seismic events to
+within a small multiple of the dominant data wavelength
+\cite[]{GauTarVir:86,VirieuxOperto:09,Plessix:10}.
+
+This paper examines one of a number of proposed alternatives to least squares
+FWI, based on an extension of the model-to-data mapping.  We show
+exactly how an example of this {\em extended inversion} approach leads
+to successful solution of a very simple acoustic transmission inverse
+problem. Despite its simplicity, this problem exhibits several
+important characteristics of all inverse problems in wave propagation
+in which wave velocities are amongst the unknowns. It has the
+advantage of permitting explicit calculations that reveal clearly the
+failure mechanism for FWI, that is, straightforward nonlinear least
+squares, and also the reasons for success of an extended inversion
+approach.
+
+In the remainder of this introduction, we will briefly describe the conceptual
+framework, the inverse problem, and sketch our main results. We leave out
+some technical details that are supplied in the next section,
+containing precise statements of the results.
+
+The idealized physical setting is the propagation of wave through a
+homogeneous acoustic fluid occuyping $\bR^3$ \cite[]{Frie:58}. The
+fluid's salient characteristic for the present discussion is the
+slowness, or reciprocal wave velocity, $m > 0$. A source of acoustic
+energy localized at a point in $\bR^3$ radiates uniformly in all
+directions with time-dependent intensity $w(t)$. Both the acoustic
+pressure field and the intensity (``wavelet'') $w$ vanish at large
+negative time. A receiver records the pressure values at a distance
+$r>0$ from the source position, between the times $t_{\rm min}$ and
+$t_{\rm max}$. As explained in the next section, the
+recorded presssure is predicted to be
+\begin{equation}
+\label{eqn:mod}
+\frac{1}{4\pi r}w\left(t-mr\right)  = F[m]w(t)
+\end{equation}
+Ignoring the amplitude factor $1/(4\pi r)$, this map implements a $m$-dependent time
+shift. This time shift operator is the basis of many descriptions of
+the cycle-skipping phenomenon (for example, \cite{VirieuxOperto:09},
+Figure 7), so it is unsurprising that an analysis of cycle-skipping
+can be based on the simple modeling operator described above.
+
+Evidently if the wavelet $w$ is square-integrable, so is the predicted
+recorded pressure - its $L^2$ norm is the cumulative power trasferred
+from the fluid to the receiver \cite[]{SantosaSymes:00}. Thus this relation defines a map
+$F: \bR^+ \times L^2(\bR) \rightarrow
+L^2([t_{\rm min},t_{\rm max}])$, linear in its second argument.
+
+A naive version of the inverse problem investigated in this paper
+would be: given recorded data
+$d \in L^2([t_{\rm min},t_{\rm max}])$, find $m \in \bR$ and $w \in L^2(\bR)$ so that
+$F[m]w \approx d$. Clearly this problem statement is not interesting as it
+stands, since it is always possible to solve it with no error and nonuniquely,
+and it does not constrain the slowness $m$ at all. To be useful, the statement must
+be augmented with a constraint on the model $(m,w)$. We choose 
+$\lambda > 0$ and search for $w$ with $\mbox{supp }w \subset
+[-\lambda,\lambda]$. The quantity $\lambda$ is to be considered
+``small'': as we shall show, it is related to a mean wavelength of
+$w$.
+
+Abusing notation slightly, we identify $L^2([-\lambda,\lambda])$ with
+a subspace of $L^2(\bR)$ via extension by $0$. Denote by $\lF$
+restriction of $F$ to $\bR^+ \times L^2([-\lambda,\lambda])$ defined by equation
+\ref{eqn:mod}. Then the FWI problem with support constraint,
+described in the last paragraph, can be crudely phrased: given $d \in
+L^2([t_{\rm min},t_{\rm max}])$, find $(m,w)$ in the domain of $\lF$ so that $\lF[m,w]
+\approx d$.
+
+Despite its simplicity, this problem exhibits the
+cycle-skipping pathology characteristic of field-scale FWI. Specifically, we show that the reduced mean square
+error function, the objective function of FWI for this problem, 
+\begin{equation}
+  \label{eqn:redms}
+  (m,w) \mapsto \lerr[m,w,d]=\frac{1}{2}\|\lF[m]w-d\|^2,
+\end{equation}
+has local minimizers far from any global minimizer, even for noise-free
+data (Theorem \ref{thm:fwi}).
+
+This fact implies that local optimization must fail to solve a precise
+version of the FWI problem: given $\epsilon, \lambda > 0$, $(m,w)$ is
+a solution if it lies in the domain of $\lF$ and
+$\lerr[m,w,d] \le 1/2 \epsilon^2\|d\|^2$ (Problem Statement
+\ref{eqn:probstat0} below).  In fact, we show
+that``large residual'' local minima of $\lerr[\cdot,\cdot,d]$ exist,
+at which its value is much greater than its global minimum
+value. Therefore if $\epsilon$ is near the global minimum value, then
+local minimization of $\lerr[\cdot,\cdot,d]$ can fail to find a
+solution of the inverse problem. This phenomenon is what is meant by
+``cycle-skipping''.
+
+Extended inversion involves an extension of the model-to-data map, in
+this case $\lF$.  The extension used here consists in dropping the
+support constraint on $w$, so the extended map is simply $F$ as
+defined above ($\lF$ is a restriction of $F$, so $F$ is an extension
+of $\lF$). To compensate for the resulting indeterminacy in the
+estimation of $m$, we add a penalty term to the objective: for $m \in
+\bR^+,w\in L^2(\bR), d \in L^2([t_{\rm min},t_{\rm max}])$, define
+\begin{eqnarray}
+  \label{eqn:pen}
+  e[m,w;d] & = & \frac{1}{2}\|F[m]w-d\|^2,\\
+  g[w] & = & \frac{1}{2}\|A\|^2, \\
+  \Ja[m, w;d] &=& e[m,w;d] + \alpha^2 g[w] \nonumber\\
+  &=& \frac{1}{2}(\|F[m]w-d\|^2 + \alpha^2 \|Aw\|^2). 
+\label{eqn:pen}
+\end{eqnarray}
+The only obvious requirement on the penalty operator $A$ is that it be
+nonzero for large $|t|$, so that minimizers of $\Ja$ should be
+concentrated near $t=0$. Thus the penalty is a soft stand-in for the
+support constraint. We
+show that a good choice for $A$ is given by $(Aw)(t)=tw(t)$, and the
+remainder of the discussion in this section makes this assumption. The
+penalty weight $\alpha$ must be chosen somehow; we review a simple
+algorithm for controlling $\alpha$ later in the paper.
+
+Assume that $\mu>0$ and $d = F[m_*]w_* + n$, with $\mbox{supp }w_*
+\subset [-\mu,\mu]$, and denote by $\eta = \|n\|/\|F[m_*w_*]$
+the noise-to-signal ratio. We show that any local minimizer $(m,w)$
+of $\Ja$ satisfies 
+\begin{itemize}
+\item If $\eta \le 0.6$,
+  \begin{equation}
+    \label{eqn:mbnd}
+    |m-m_*| \le (1+f(\eta))\frac{\lambda}{r},
+  \end{equation}
+  where $f(\eta)=2\eta+O(\eta^2)$ (Theorem \ref{thm:mnoiseres});
+\item 
+  $(m,{\bf 1}_{[-\lambda,\lambda]}w)$ is a solution of the FWI problem
+  as stated above, that is,
+  \begin{equation}
+    \label{eqn:success}
+    \lerr[m, {\bf 1}_{[-\lambda,\lambda]}w,d] \le \frac{1}{2}\epsilon^2,
+  \end{equation}
+  so long as
+  \begin{eqnarray}
+    \lambda & \ge & (2+f(\eta))\mu,\nonumber\\
+    \epsilon &\ge & (4\pi r \alpha \lambda)^2 +\eta.
+    \label{eqn:solcond}
+  \end{eqnarray}
+  (Theorem \ref{thm:ipnoisesuf})
+\end{itemize}.
+That is, the minimization of the extended objective $\Ja$ provides a
+solution of the FWI problem, with constraints on the achievable
+performance and bounds on the errors. All of these
+bounds are shown to be sharp, and the constraints cannot be
+significantly weakened. For example, the support radius $\lambda$ in
+the projection onto $L^2([-\lambda,\lambda])$ must in general be at least twice the
+support radius of the ``noise free'' wavelet $w_*$, plus an additional
+allowance for noise.
+
+The remainder of the paper begins with a complete description of the acoustic transmission problem and precise
+statements of our main results, filling in the
+few details not discussed in the preceding sketch. Then we provide an
+discussion of the potential theoretical and practical import of these
+results, including relations with prior work, some numerical
+illustrations, and directions for further research. Following this
+discussion section, we give complete proofs of all of our results.
+An appendix describes the computational algorithm used in producing
+the numerical examples, including a method for controlling the penalty
+weight $\alpha$ based on the Discrepancy Principle
+\cite[]{EnglHankeNeubauer,FuSymes2017discrepancy}. A second appendix
+explains the relation between the support constraint and well-known
+criteria for success of FWI based on data frequency content. The basis
+for this relation is the Heisenberg inequality \cite[]{Folland:07}.
+
+\section{An Acoustic Transmission Inverse Problem}
+In this section, we describe solution of the radiation
+problem for acoustic wave
+propagation in a homogeneous fluid, and various operators based on
+this solution. Using these operators, we formulate the single-trace
+acoustic transmission inverse problem. Finally, we state the main
+results of our paper.
+
+We shall use the abbreviations ${\cal B}(X,Y)$ and ${\cal I}(X,Y)$ for 
 the algebra of bounded linear operators from the Hilbert space $X$ to 
@@
 
-%\section{Theory}
-
-%\cite{FuSymes2017discrepancy} introduce a version of the discrepancy
-%principle and its application to separable nonlinear least squares
-%problems in penalty form \ref{eqn:basic}. The algorithm developed
-%there is incomplete, however: it lacks  termination criteria. In this
-%section, I review the discrepancy-based algorithm and introduce an
-%appropriate stopping rule, thus completing the development of the
-%algorithm presented in \cite{FuSymes2017discrepancy}.
-
-\section{An Example: Single Trace Modeling}
+\subsection{Acoustic transmission}
 Assume small amplitude (linearized) acoustic propagation, constant
@@
 source wavelet $w$. Therefore it is naturally represented as the
-action of a $m$-dependent linear operator $S[m]$ on $w$:
-\begin{equation}
-\label{eqn:mod}
-F[m]w(t)= p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right).
-\end{equation}
-
-Ignoring amplitude, this map implements a $m$-dependent time
-shift. This time shift operator is the basis of many descriptions of
-the cycle-skipping phenomenon (for example, \cite{VirieuxOperto:09},
-Figure 7), so it is unsurprising that an analysis of cycle-skipping
-can be based on the simple modeling operator described above, which
-amounts essentially to a time shift.
+action of a $m$-dependent linear operator $F[m]$ on $w$, as described
+in \ref{eqn:mod}.
+%\begin{equation}
+%\label{eqn:mod}
+%F[m]w(t)= p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right).
+%\end{equation}
 
@@
 material model: for crustal rock, a reasonable choice would be
-$m_{\rm min}=0.125, m_{\rm max}=0.6$ s/km. The wavelet $w$ is
-naturally assumed to be 
-square-integrable, corresponding to finite power transfer from the
-source process to the propagating wavefield $p$
-\cite[]{SantosaSymes:00}. The predicted data $F[m]w$ is then also
-square-integrable.
-
-FIeld recording takes place over finite time intervals (and with
-finite sample rate, a detail I shall ignore in this paper). Denote by
+$m_{\rm min}=0.125, m_{\rm max}=0.6$ s/km.
+
+Field recording takes place over finite time intervals (and with
+finite sample rate, a detail we shall ignore in this paper). Denote by
 $[t_{\rm min},t_{\rm max}]$ the recording interval for the
@@
 \end{equation}
-
-\section{The Inverse Problem}
+Note also that $F[m]$ is surjective for every $m \in M$.
+
+In computational practice, $W$ will have to be replaced by a
+finite-dimensional subspace of $L^2(\bR)$. Many such choices will
+implicitly limit the support of $w \in W$ to a bounded interval, say
+$[T_{\rm min},T_{\rm max}]$. To maintain the surjective property,
+these bounds should be chosen so that $[t_{\rm min}, t_{\rm max}]
+\subset [T_{\rm min}+mr,T_{\rm max}+mr]$ for all $m \in M$, that is,
+\begin{eqnarray}
+  \label{eqn:dombounds}
+  t_{\rm min} &\ge & T_{\rm min}+m_{\rm max}r,\nonumber\\
+  t_{\rm max} &\le & T_{\rm max}+m_{\rm min}r.
+\end{eqnarray}
+We will ignore these computational necessities in this work,
+maintaining the definition $W=L^2(\bR)$.
+
+\subsection{An inverse problem}
 
@@
 [t_{\rm min},t_{\rm max}]$. Also, $\lF[m]$ is coercive for all $m \in
-M$ (so that $w$ is uniquely determined) only if
+M$ only if
 $[ m_{\rm min}r-\lambda,m_{\rm max}r+\lambda] \subset
-[t_{\rm min},t_{\rm max}]$. Define $\lambda_{\rm max}
->0$, so that $M, r,$ and $T$ by
+[t_{\rm min},t_{\rm max}]$.
+
+We will assume that $\lambda_{\rm max}
+>0$ is chosen so that $M, r,\lambda_{\rm max}$ and $[t_{\rm min},t_{\rm max}]$ satisfy
 \begin{equation}
@@
 \noindent {\bf Problem Statement:} given data $d \in D$, relative 
-error level $\epsilon >0$, and support radius $0 <\lambda \le 
-\lambda_{\rm max}$, find 
+error level $\epsilon \in [0,1)$, and support radius $\lambda \in (0, 
+\lambda_{\rm max}]$, find 
 $(m,w) \in M \times \lW$ for which 
@@
 \end{equation}
+
+\noindent{\bf Remark:} The constraint $\epsilon < 1$ imposed on the
+target noise level is eliminates the obvious choice $(m,0)$, which
+satisfies the data misfit constraint for any $m \in M$.
 
@@
 to ``within a (fraction of a) wavelength''. The relation is a
-consequence of Heisenberg's inequality, and will be reviewed later in
-the paper.
+consequence of Heisenberg's inequality, and will be reviewed in
+Appendix B.
 
@@
 \end{equation}
-Such data $d$ is {\em noise-free}, in the range of the map $\lF$, and a solution of the Problem Statement
-exists with arbitrarily small $\epsilon>0$. I will relax the noise-free assumption
-later in the paper: until further notice, assume that $d \in
-D$ satisfies the condition \ref{eqn:defdatanonoise}.
+Such data $d$ is {\em noise-free}, in the range of the map $\lF$. For
+such data a solution of the Problem Statement
+exists with arbitrarily small $\epsilon>0$.
+
+\subsection{Soft constraints and penalty formulation}
+
+As mentioned in the introduction, a straightforward nonlinear least
+squares formulation of the problem stated in \ref{eqn:probstat0}
+exhibits local minimizers with large fit error, and these will be
+found by any local optimization algorithm unless it is initialized at
+a quite accurate initial model. The first result stated in the next
+section establishes this assertion for the single trace transmission
+problem, however it appears to be valid for field scale problems, for
+essentially the same reasons, as widely reported in the literature on
+seismic inversion.
+
+One (of many) alternative optimization formulation of this and similar
+problems involves introducing an extension of the model-to-data
+operator, and compensating for the physically spurious degrees of
+freedom so introduced by adding a penalty term controlling them to the
+objective.  The application of this idea presented here extends $\lF$
+by dropping the support constraint on its second argument, that is,
+using the expression \ref{eqn:mod} to define a map $F: M \times W
+\rightarrow D$. The objective begins with an extended version of the
+mean-square error,
+\begin{equation}
+  \label{eqn:mserr}
+  e[m,w;d]=\frac{1}{2}\|F[m]w-d\|^2.
+\end{equation}
+We will choose a penalty to control the extra degrees of freedom, that
+is, $w$ outside of $[-\lambda,\lambda]$. This  penalty is quadratic, constrains only $w\in W$ (as it is a soft
+version of the condition $w \in \lW$), and is
+$m-$independent. Therefore the penalty function takes the form
+\begin{equation}
+\label{eqn:pen}
+p[w] = \frac{1}{2}\|Aw\|_Y^2 
+\end{equation}
+where $A \in {\cal B}(W,Y)$, $Y$ being another suitable Hilbert
+space.
+
+The penalty approach asks that a linear combination of $e,p$,
+depending on a relative weight $\alpha$:
+\begin{equation}
+\Ja[m, w;d] = e[m,w;d] + \alpha^2 p[w] = \frac{1}{2}(\|F[m]w-d\|^2 + \alpha^2 \|Aw\|^2)
+\label{eqn:pen}
+\end{equation}
+be minimized over $(m,w)\in M \times W$.
+
+The choice of penalty weight $\alpha$ has a profound influence on the
+character of this optimization problem. For the moment, we will mandate
+only that $\alpha \ge 0$. A complete description of an algorithm using
+the penalty function \ref{eqn:pen} to solve the inverse problem
+\ref{eqn:probstat0} must include a selection principle for $\alpha$. We
+describe one such principle in Appendix A.
+
+The main theoretical device used in the proofs of our main results on
+extended inversion is the reduction of $\Ja$ to a function $\tJa$ of
+$m$ alone, by minimization over $w$. This reduction is the foundation
+of the {\em Variable Projection Method} (VPM)
+\cite[]{GolubPereyra:03} for solution of partly linear (``separable'')
+least squares problems. We use the VPM in our numerical illustrations
+of the theory, as described in Discussion section and Appendix A.
+
+\begin{theorem}
+\label{thm:red}
+Assume that the Hessian operator of
+$\Ja[m,\cdot;d]$ is invertible:
+\begin{equation}
+\label{eqn:norminv}
+F[m]^TF[m] +\alpha^2 A^TA \in {\cal I}(W,W).
+\end{equation}
+(The superscript $T$ denotes the adjoint with respect to the inner products 
+in $W=L^2(\bR)$ and $D=L^2([t_{\rm min},t_{\rm max}])$.)
+Then the unique stationary point $\aw[m;d] \in W$ of
+$\Ja[m,\cdot;d]$ is the solution of the {\em normal equation}:
+\begin{equation}
+  \label{eqn:norm}
+  (F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td,
+\end{equation}
+Define the reduced or variable projection objective by
+\begin{equation}
+  \label{eqn:redexp}
+  \tJa[m;d] =\Ja[m,w_{\alpha}[m;d];d] = \min_w \Ja[m,w;d]
+\end{equation}
+Then $(m,w)$ is a local minimizer of $\Ja[\cdot,\cdot;d]$ if and only
+if $m$ is a local minimizer of $\tJa[\cdot;d]$ and $w=\aw[m;d]$.
+\end{theorem}
+\noindent{\bf Remark:} If $\Ja[\cdot,\cdot;d]$ and $\tJa[\cdot;d]$ are differentiable, then
+  ``local minimizer'' in the conclusion of the preceding theorem can
+  be replaced by ``stationary point''. However, for the problem
+  addressed in this paper, $\Ja[\cdot,\cdot;d]$ {\em is not}
+  differentiable without added smoothness constraints on $w$, whereas
+  $\tJa[\cdot;d]$ {\em is} differentiable for proper choice of penalty
+  operator $A$, as will be shown
+  below. This phenomenon is replicated for a wide variety of extended
+  approaches for inverse problems in wave propagation. We will have
+  more to say about it in the Discussion section.
+
+\subsection{Main Results}
+The first main result establishes the existence of large (100 \%)
+residual local minimizers for the basic FWI objective $\lerr$, even
+for noise-free data.
+\begin{theorem}
+  \label{thm:fwi}
+  Suppose that $0 <\lambda\le \lambda_{\rm max}$,  $m_* \in M, w_*
+  \in \lW, d=\lF[m_*]w_*$ is noise-free data per definition \ref{eqn:defdatanonoise},
+  Under assumption \ref{eqn:fullrec}, for any $m \in M$ with $|m-m_*|r>2\lambda$,
+\begin{equation}
+  \label{eqn:isovpm}
+ \min_w \lerr[m,w;d]=\lerr[m,0,d]=\frac{1}{2}\|d\|^2,
+\end{equation}
+and any such $(m,0)$ is a local minimizer of $e$.
+\end{theorem}
+
+Turning to extended inversion, we examine two choices of penalty
+operators. For each choice, we ask first whether local minimizers 
+of the resulting VPM
+objective $\tJa[\cdot;d]$ occur far from a slowness $m_*$. Both choices for $A$ considered here are scalar 
+multiplication operators defined by a choice of positive multiplier $a \in L^{\infty}(\bR)$:
+\begin{equation}
+  \label{eqn:annmult}
+  A w(t)= a(t)w(t), \, t\in \bR: \,a(t) > 0 \mbox{ for }t \ne 0.
+\end{equation}
+
+A penalty operator $A$ of which $\lW$ is the null space would be a
+natural choice. Such operators have come to be called
+``annihilators'', since they map all members of the constraint
+subspace $\lW$ to zero.  A simple example is
+\begin{eqnarray}
+  A = E^c_{\lambda}&=&I - E_{\lambda},\mbox{ where } \nonumber \\
+  E_{\lambda}w(t) &=&{\bf 1}_{[-\lambda,\lambda]}(t)w(t). 
+                      \label{eqn:ann0}
+\end{eqnarray}
+That is, $E_{\lambda}$ is the orthogonal projector onto $\lW$,
+and $E_{\lambda}^c$ is the orthogonal projector onto its
+orthocomplement, an operator of the form \ref{eqn:annmult} with $a
+= 1 - {\bf 1}_{[-\lambda,\lambda]}$. %Note that this choice of $a$
+%satisfies condition \ref{eqn:abnd} with $C \equiv 1$.
+
+\begin{theorem}
+  \label{thm:boxcarbad}
+  Suppose that
+  \begin{itemize}
+  \item[1. ] $\lambda \in (0,\lambda_{\rm max}]$;
+  \item[2. ] $m_*\in M, w_*\in \lW$, and $d=F[m_*]w_*$ (noise-free
+    data);
+  \item[3. ] $A=E^c_{\lambda}$, that is, $a=1-{\bf 1}_{-\lambda,\lambda]}$ in
+    the definition \ref{eqn:annmult}.
+  \end{itemize}
+  Then if $|m-m_*| >  2\lambda/r$,
+  \[
+    \tJa[m;d] = \frac{\alpha^2}{2(1+(4 \pi r)^2 \alpha^2)}\|w_*\|^2.
+  \]
+\end{theorem}
+
+\noindent {\bf Remark.}
+One might have thought that $A=E^c_{\lambda}$ would be a better choice
+of annihilator, as for noise-free
+data, the solution set defined by the problem statement
+\ref{eqn:probstat0} is the same as the set of global minimizers of
+$\tJa$ in this case. However, for this choice of annihilator,
+$\tJa$ exhibits the same feature as the mean square error $e$, namely
+a continuum of stationary points at any distance from the global
+minimizer $m_*$ greater than a multiple of $\lambda$. Therefore the
+extended inversion with this choice of annihilator is no more amenable
+to local optimization than is FWI.
+
+A second possible 
+penalty operator penalizes energy away from 
+$t=0$: choose $\tau > 0$ and set 
+\begin{equation}
+  \label{eqn:ann}
+  a(t) = \min(|t|, \tau). 
+\end{equation}
+The cutoff $\tau$ will be chosen large enough to be effectively inactive: 
+specifically, hindsight suggests 
+\begin{equation}
+  \label{eqn:taudef}
+  \tau = \max\{|t_{\rm min}-m_{\rm min}r|,|t_{\rm min}-m_{\rm max}r|, |t_{\rm max}-m_{\rm min}r|, |t_{\rm max}-m_{\rm max}r|\}. 
+\end{equation}
+This 
+particular annihilator has been employed in earlier papers on extended 
+source inversion 
+\cite[]{Plessix:00a,LuoSava:11,Warner:14,HuangSymes:SEG15a,Warner:16,HuangSymes:GEO17}.
+
+\begin{theorem}
+  \label{thm:mnoiseres}
+  Asumme that
+  \begin{itemize}
+  \item[1. ] $\alpha, \mu> 0$,
+  \item[2. ] $m_* \in M, w_* \in W_{\mu}$,
+  \item[3. ] $d_* = F[m_*]w_*$,
+  \item[4. ] $n \in D$ and $d = d_* + n$.
+  \end{itemize}
+  Set $\eta = \|n\|/\|d_*\|$, and assume further that
+  \begin{itemize}
+  \item[5. ]
+    \begin{equation}
+      \label{eqn:mnoisecond}
+      \eta < \frac{\sqrt{5}-1}{2}, and
+    \end{equation}
+  \item[6. ] $m$ is a stationary point of $\tJa[\cdot;d]$.
+  \end{itemize}
+  Then
+  \begin{equation}
+    \label{eqn:mnoisesuff}
+    |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}.
+  \end{equation}
+\end{theorem}
+
+\noindent {\bf Remark.} That is, {\em with the choice of penalty
+  multiplier $a$ given in equation \ref{eqn:ann}, and support radius
+  $\mu$ of the ``noise-free'' wavelet $w_*$, $\tJa$ has no local
+  minima further than $(1+O(\eta))\mu/r)$ from the slowness used to
+  generate the data}.
+
+\noindent {\bf Remark.} The estimate $|m-m_*|r<\mu(1+O(\eta))$ for stationary
+points $m$ of $\tJa$ is sharp: it is possible to choose $w_* \in W_{\mu}$
+so that $\mu - |m-m_*|r$ is as small as you like. In particular,
+the ``exact'' or ``true'' slowness $m_*$ is not necessarily the only 
+stationary point, or even a stationary point at all, and in particular
+is not (necessarily) a global minimizer of $\tJa$.
+
+\noindent {\bf Remark.} No similar bound could hold for much larger
+noise levels than specified in condition \ref{eqn:mnoisecond}, the
+right-hand side of which is a bit larger than 0.6. For example, if the noise is
+the predicted data for the same wavelet $w_*$ with a substantially different
+slowness $m_{\flat}$, that is, $n=F[m_{\flat}]w_*$, then by symmetry
+the average slowness $0.5(m_*+m_{\flat})$ is a stationary point of
+$\tJa[\cdot,d_*+n]$. The difference with $m_*$ is not constrained at
+all by the support of $w_*$. So for this example with 100\% noise, no
+bound of the type given by conclusion 2 could possibly hold. We will
+illustrate this phenomenon numerically later in the paper.
+
+\noindent{\bf Remark.} Note that $\alpha$ plays no role in the
+conclusions of this theorem. It is only required that $\alpha >0$.
+
+Unless the data is noise-free, there is no reason to suppose that the
+estimated wavelet $\aw[m;d]$ (Theorem \ref{thm:red}) will lie in
+$\lW$, unless the support of the noise $n$ is not restricted. In order
+to construct a solution of the inverse problem \ref{eqn:probstat0}, we
+project $\aw[m;d]$ onto $\lW$. For sufficiently large $\lambda,
+\epsilon$, the result is a solution of the inverse problem:
+
+\begin{theorem}
+  \label{thm:ipnoisesuf}
+  Assume the hypotheses of Theorem \ref{thm:mnoiseres}, and that
+  inequality \ref{eqn:mnoisecond} holds, and $\mu \in
+  (0,\lambda_{\rm max}]$. Then the pair
+  \[
+    (m,{\bf 1}_{[-\lambda,\lambda]}\aw[m,d])
+  \]
+  solves the inverse problem as stated in \ref{eqn:probstat0} if
+  \begin{equation}
+    \label{eqn:ipnoiselam}
+    \left( 2+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\mu \le \lambda
+    \le \lambda_{\rm max}, 
+  \end{equation}
+  and
+  \begin{equation}
+    \label{eqn:ipnoiseeps}
+    \epsilon \ge (4 \pi r \alpha\lambda)^2(1+\eta^2)^{1/2} 
+  \end{equation}    
+\end{theorem}
+
+\noindent {\bf Remark:} Note that the sufficient condition \ref{eqn:ipnoiselam} for
+$\lambda$ is independent of $\alpha$. It follows that for any choice
+of $\lambda$ consistent with this bound, $(m,{\bf
+  1}_{[-\lambda,\lambda]}\aw[m,d])$ is a solution of the inverse
+problem for any $\epsilon > 0$ provided that $\alpha$ is chosen
+sufficiently small ($O(\sqrt{epsilon})$).
+
+\section{Discussion}
+
+Use of model extension to reveal the influence of earth mechanical
+parameters on data is an old idea, and underlies several classes of
+seismic data processing methods in widespread use for decades
+\cite[]{geoprosp:2008}. Use of model extension to create optimization
+principles for inversion is a more recent development
+\cite[]{SymesCar:91,Plessix:00a,Symes:09,LuoSava:11,BiondiAlmomin:SEG12,LeeuwenHerrmannWRI:13,LiuSymesLi:14,Warner:14,ChaurisGP:14,Warner:16,LeeuwenHerrmann:16,Herve2017,HouSymes:Geo18}. Extended
+inversion methods differ by the choice of additional degrees of
+freedom, and by choice of penalty applied to eliminate them in the
+final result. The approach studied in this paper belongs to the {\em
+  source extension} variety: the extra parameters are provided by
+permitting the energy source component of the experimental model to
+have more, or less constrained, parameters than the experimental
+design suggests. \cite{HuangNammourSymesDollizal:SEG19} give an
+overview and taxonomy of source extension methods, with extensive
+references.
+
+Extended inversion is not the only alternative to straightforward
+least-squares data fitting that may overcome cycle-skipping. For
+example, evidence has recently emerged that the Wasserstein metric
+from arising in the theory of optimal transport may provide a measure
+of error between model-predicted and observed seismic data less
+oscillatory than the mean-square
+\cite[]{EngquistYang:GEO18,Metivier:GEO18}.
+
+Many inversion concepts sound plausible, and appear
+to work at least to some extent as one might hope from their heuristic
+justifications. However very few of these approaches have been
+underwritten by mathematical argument: in essence, they are mostly
+justified only by numerical example, ``in the rear-view mirror'', with no assurance that
+failure is not just around the corner, at the next example. On top of
+that, some of these approaches, for example those based on the
+computationally attractive Variable Projection Method (``VPM'') of
+\cite{GolubPereyra:03}, are cast in such form that the reasons for
+success are not readily apparent.
+
+In many cases, the constraint posed by membership of the second component $w$ in the subspace $\lW$ is an approximation to an idealized limit $\lambda \rightarrow 0$.
+It may be that $W_0 = \cap_{\lambda > 0}\lW$ is non-trivial, in which
+case $F$ may be viewed as an extension of $F_0 = F|_{ M \times
+  W_0}$. In that case, $F_0$ and $F$ realize the idealized extension
+structure discussed in many prior works on extended inversion (for
+example, \cite[]{geoprosp:2008}): $F_0$ is the physical modeling
+operator, and $W_0$ consists of second components of physical or
+feasible models. However, in many of these examples, $W_0$ is not
+computationally accessible, or - even if it is - gives results that
+closely approximate those obtained with $\lW$ for small $\lambda$. For
+other examples, for instance that  discussed later in this paper (and
+implicitly in \cite[]{Warner:16}, for instance), $W_0=\{0\}$ and the
+family $\{\lW: \lambda > 0\}$ actually embodies all of the necessary
+physics.
+
+While minimization of $\Ja$ might be tackled directly - by
+alternating minimizations over $m$ and $w$, or by computing updates
+for $m$ and $w$ simultaneously - such joint minimization performs
+poorly, as \cite{YinHuang:16} has shown. The reason for this poor
+performance is that $\Ja$ has dramatically different
+sensitivity to $m$ versus $w$, as observed in the cited reference and elsewhere.
+%especially for high frequency $f$, as
+%the reader will see below.
+%as will be established below.
+Instead, a nested approach, in which $w$ is
+eliminated in an inner optimization,
+generally gives far better numerical performance.  This takes advantage of
+$\Ja$ being quadratic in
+$w$ to eliminate $w$ from the problem statement, as described in the
+following theorem.
+%\section{Theory}
+
+%\cite{FuSymes2017discrepancy} introduce a version of the discrepancy
+%principle and its application to separable nonlinear least squares
+%problems in penalty form \ref{eqn:basic}. The algorithm developed
+%there is incomplete, however: it lacks  termination criteria. In this
+%section, I review the discrepancy-based algorithm and introduce an
+%appropriate stopping rule, thus completing the development of the
+%algorithm presented in \cite{FuSymes2017discrepancy}.
 
@@
 A natural formulation of the Problem Statement \ref{eqn:probstat0} is
-via least squares: that is, given $d \in D$, $\lambda >0$, find $(m,w)
-\in M \times \lW$ to minimize $e: M \times W
- \rightarrow \bR$, defined by
-\begin{equation}
-  \label{eqn:FWI}
-  e[m,w;d] = \frac{1}{2}\|\lF[m]w-d\|^2.
-\end{equation}
-A pair $(m,w) \in M\times \lW$
-satisfies the condition \ref{eqn:probstat0} iff $e[m,w;d] \le \epsilon^2/2$, so a solution exists iff a 
-global minimizer of $e$ satisfies this inequality.
+via least squares: that is, given $d \in D$, $\lambda >0$, find
+$(m,w) \in M \times \lW$ to minimize
+$\lerr: M \times W \rightarrow \bR$, defined in display
+\ref{eqn:redms}.  A pair $(m,w) \in M\times \lW$ satisfies the
+condition \ref{eqn:probstat0} iff $\lerr[m,w;d] \le \epsilon^2/2$, so a
+solution exists iff a global minimizer of $\lerr$ satisfies this
+inequality.
 
@@
 \[
- e[m,w;d] =  \frac{1}{32\pi^2
+ \lerr[m,w;d] =  \frac{1}{32\pi^2
     r^2}\int_0^T\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2
@@
 
-\section{Soft constraints and penalty formulation}
-
-For inverse problems based on wave propagation, enforcing hard
-constraints on the solution ($w \in \lW$) leads to the pathology
-similar to that described in the last
-section for the single-trace transmission problem, and to failure of
-local optimization methods to converge to a useful solution. A soft
-constraint or penalty approach can avoid this obstacle.
-
-This
-discussion presumes that the penalty is quadratic, constrains only
-$w\in W$
-(as it is a soft version of the condition $w \in \lW$), and is
-$m-$independent. Therefore the penalty function takes the form
-\begin{equation}
-\label{eqn:pen}
-p[w] = \frac{1}{2}\|Aw\|_Y^2 
-\end{equation}
-where $A \in {\cal B}(W,Y)$, $Y$ being another suitable Hilbert
-space.
-
-The penalty approach asks that a linear combination of $e,p$,
-depending on a relative weight $\alpha$:
-\begin{equation}
-\Ja[m, w;d] = e[m,w;d] + \alpha^2 p[w] = \frac{1}{2}(\|F[m]w-d\|^2 + \alpha^2 \|Aw\|^2).
-\label{eqn:pen}
-\end{equation}
-be minimized over $m\in M$ and $w\in W$.
-
-The choice of penalty weight $\alpha$ has a profound influence on the
-character of this optimization problem. For the moment, I will mandate
-only that $\alpha \ge 0$. A complete description of an algorithm using
-the penalty function \ref{eqn:pen} to solve the inverse problem
-\ref{eqn:probstat0} must include a selection principle for $\alpha$. I
-shall describe one such principle below.
-
-While minimization of $\Ja$ might be tackled directly - by
-alternately minimizations between $m$ and $w$, or by computing updates
-for $m$ and $w$ simultaneously - such joint mimization performs
-poorly, as \cite{YinHuang:16} has shown. The reason for this poor
-performance is that $\Ja$ has dramatically different
-sensitivity to $m$ versus $w$,
-%especially for high frequency $f$, as
-%the reader will see below.
-as will be established below.
-Instead, a nested approach, in which $w$ is
-eliminated in an inner optimization,
-generally gives far better numerical performance.  This {\em Variable
-Projection Method} (VPM) \cite[]{GolubPereyra:03} takes advantage of
-$\Ja$ being quadratic in
-$w$ to eliminate $w$ from the problem statement, as described in the
-following theorem.
-
-In stating this result, I will make use of the the Hessian operator of
-$\Ja[m,\cdot]$, which is $F[m]^TF[m] + \alpha^2 A^TA$. The
-superscript $T$ denotes the adjoint with respect to the inner products
-in $W=L^2(\bR)$ and $D=L^2([t_{\rm min},t_{\rm max}])$.
-
-\begin{theorem}
-\label{thm:red}
-Assume that
-\begin{equation}
-\label{eqn:norminv}
-F[m]^TF[m] +\alpha^2 A^TA \in {\cal I}(W,W).
-\end{equation}
-Then the unique stationary point $\aw[m;d] \in W$ of
-$\Ja[m,\cdot;d]$ is the solution of the {\em normal equation}:
-\begin{equation}
-  \label{eqn:norm}
-  (F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td,
-\end{equation}
-Define the reduced or variable projection objective by
-\begin{equation}
-  \label{eqn:redexp}
-  \tJa[m;d] =\Ja[m,w_{\alpha}[m;d];d] = \min_w \Ja[m,w;d]
-\end{equation}
-Then
-\begin{equation}
-  \label{eqn:objexp}
-  \tilde{J}_{\alpha}[m;d] =\langle d, (I-F[m]
-  (F[m]^TF[m]+\alpha^2A^TA)^{-1}F[m]^T)d\rangle_D
-\end{equation}
-\end{theorem}
-
-\begin{proof}
-Substitute $w_{\alpha}[m;d]$ for $w$ in the definition \ref{eqn:pen}
-of $J_{\alpha}$, using the normal equation \ref{eqn:norm} and 
-rearranging, obtain the result.
-\end{proof}
-
 \section{All stationary points are (near) global minimizers -
@@
 {\em for some choices of penalty operator, but not for other choices
-  that appear equally or even more suitable}.
+  that might appear equally or even more suitable}.
 
@@
 
-Therefore the normal equation for the minimizer on the RHS of equation \ref{eqn:red} is
+Therefore the normal equation for the minimizer on the RHS of equation \ref{eqn:redexp} is
 \begin{equation}
@@
 
-At this point I have to come clean about the actual choice of
-$A$. The role of $A$ is to penalize distance from the constrained
-subspace $\lW$, in some sense. All choices for $A$ considered here are scalar 
-multiplication operators defined by a choice of $a \in L^{\infty}(\bR)$:
-\begin{equation}
-  \label{eqn:annmult}
-  A w(t)= a(t)w(t), \, t\in \bR.
-\end{equation}
+The role of $A$ is to penalize distance from the constrained
+subspace $\lW$, in some sense.
 
@@
   \ref{eqn:annmult}. Also assume that $\lambda \in (0,\lambda_{\rm
-    max}], \alpha > 0,$ and that $a \in L^{\rm infty}(\bR)$
+    max}], \alpha > 0,$ and that $a \in L^{\infty}(\bR)$
   mentioned in condition \ref{eqn:annmult} satisfies
@@
     \begin{equation}
-      \label{eqn:solnnongen}
+      \label{eqn:solnonoise}
       \aw[m,d](t)= \left(1+ (4\pi r)^2\alpha^2 a(t)^2\right)^{-1}w_*\left(t+(m-m_*)r\right).
@@
 \begin{theorem}
+  \label{thm:epjgen}
+  Assume the hypotheses of Theorem \ref{thm:norminv}. Then
+  \begin{equation}
+  \label{eqn:residnormgen}
+  e[m,\aw[m,d];d] = \frac{1}{2}\int \,dt\,(4\pi r \alpha a(t-mr))^4(1 +
+  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)^2
+\end{equation}
+\begin{equation}
+  \label{eqn:anninormgen}
+  p[m,\aw[m,d];d] = \frac{1}{2}\int \,dt\,(4\pi r a(t-mr))^2(1 +
+  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)
+\end{equation}
+
+\begin{equation}
+  \label{eqn:expjgen}
+\tJa[m;d] = \frac{1}{2}\int\,dt\,(4\pi r \alpha a(t-mr))^2(1+(4\pi r \alpha 
+a(t-mr))^2)^{-1}d(t)^2. 
+\end{equation}
+Finally, if $a \in W^{1,1}_{\rm loc}(\bR)$ (distributions with integrable 
+derivatives), then $\tJa[\cdot;d]$ is differentiable, and 
+\begin{equation}
+  \label{eqn:dexpjgen}
+  \frac{d}{dm}\tJa[m;d] = -(4 \pi r \alpha)^2 \int \,dt \, 
+  \left(a\frac{da}{dt}\right)(t-mr)(1+(4\pi r \alpha 
+  a(t-mr))^2)^{-2}d(t)^2. 
+\end{equation}
+\end{theorem}
+
+\begin{proof}
+  From equation \ref{eqn:normsol},
+  \[
+    F[m]\aw[m;d](t) = 
+    \frac{1}{4 \pi r}\left(\frac{1}{(4\pi r)^2} + \alpha^2
+      a^2(t-mr)\right)^{-1}\frac{1}{4 \pi r}d(t),
+  \]
+  so
+  \[
+    (F[m]\aw[m;d]-d)(t) = (1 + (4 \pi r\alpha
+    a(t-mr))^2)^{-1}-1)d(t)
+  \]
+  \[
+    =(4 \pi r\alpha a(t-mr))^2(1 + (4 \pi r\alpha
+    a(t-mr))^2)^{-1}d(t).
+  \]
+  Half the integral of the square of this data residual is
+  $e[m,\aw[m;d],d]$, which proves identity \ref{eqn:residnormgen}.
+
+  To compute $p[m,\aw[m;d],d]$, note that
+  \[
+    A\aw[m;d](t)=a(t) \left(\frac{1}{(4\pi r)^2} + \alpha^2
+      a^2(t)\right)^{-1}\frac{1}{4 \pi r}d(t+mr)
+  \]
+  \[
+    = 4\pi r a(t) (1 + (4\pi r \alpha a(t))^2)^{-1}d(t+mr)
+  \]
+  for $ t \in [t_{\rm min}-mr, t_{\rm max}-mr]$, so squaring,
+  integrating, and changing integration variables $t \mapsto t-mr$
+  gives the result \ref{eqn:anninormgen}
+
+  That the VPM objective $\tJa$ is given by \ref{eqn:expjgen} follows from equations \ref{eqn:pen},
+  \ref{eqn:redexp}, \ref{eqn:residnormgen}, and
+  \ref{eqn:anninormgen}. Differentiation under the integral sign
+  yields the expression \ref{eqn:dexpjgen} for its derivative with
+  respect to $m$, for smooth $a$, and then by a limiting argument for
+  $a \in W^{1,1}_{\rm loc}(\bR)$.
+\end{proof}
+
+\begin{corollary}
   \label{thm:epjnonoise}
@@
 \end{equation}
-so finally
-\begin{equation}
-\label{eqn:expjgen}
-\tJa[m;d] = =\frac{\alpha^2}{2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
-  a(t-(m-m^*)r)^2)^{-1}w_*(t)^2.
-\end{equation}
-\end{theorem}
-
-I willl now examine two possible choices of penalty multipler $a$ in
-defintion \ref{eqn:annmult}.
-
-A penalty operator $A$ of which $\lW$ is the null space would be a
-natural choice. Such operators have come to be called
-``annihilators'', since they map all members of the constraint
-subspace $\lW$ to zero.  A simple example is
-\begin{eqnarray}
-  A = E^c_{\lambda}&=&I - E_{\lambda},\mbox{ where } \nonumber \\
-  E_{\lambda}w(t) &=&{\bf 1}_{[-\lambda,\lambda]}(t)w(t). 
-                      \label{eqn:ann0}
-\end{eqnarray}
-That is, $E_{\lambda}$ is the orthogonal projector onto $\lW$,
-and $E_{\lambda}^c$ is the orthogonal projector onto its
-orthocomplement, an operator of the form \ref{eqn:annmult} with $a
-= 1 - {\bf 1}_{[-\lambda,\lambda]}$. Note that this choice of $a$
-satisfies condition \ref{eqn:abnd} with $C \equiv 1$.
-
-\begin{theorem}
-  \label{thm:boxcarbad}
-  For $A=E^c_{\lambda}$, that is, $a=1-{\bf 1}_{-\lambda,\lambda]}$ in
-  the definition \ref{eqn:expjgen}. Then if $|m-m_*| >  2\lambda/r$,
-  \[
-    \tJa[m;d] = \frac{\alpha^2}{2(1+(4 \pi r)^2 \alpha^2)}\|w_*\|^2.
-  \]
-\end{theorem}
-
-\begin{proof}
-  This is clear, as the supports of $w_*$ and $t \rightarrow {\bf
-    1}_{[-\lambda,\lambda]}(t-(m-m_*))$ are disjoint under this
-  condition.
+so 
+\begin{equation}
+\label{eqn:expjnonoise}
+\tJa[m;d] = \frac{\alpha^2}{2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
+  a(t-(m-m_*)r)^2)^{-1}w_*(t)^2. 
+\end{equation}
+Finally, if $a \in W^{1,1}_{\rm loc}(\bR)$, then $\tJa[\cdot;d]$ is differentiable, and 
+\begin{equation}
+  \label{eqn:dexpjnonoise}
+  \frac{d}{dm}\tJa[m;d] = -r \alpha^2 \int \,dt \, 
+  \left(a\frac{da}{dt}\right)(t-(m-m_*)r)(1+(4\pi r)^2 \alpha^2 
+  a(t-(m-m_*)r)^2)^{-2}w_*(t)^2. 
+\end{equation}
+\end{corollary}
+
+\begin{proof} of Theorem \ref{thm:boxcarbad}
+
+ This is clear from the definition $a = 1-{\bf 1}_{-\lambda,\lambda]}$ and equation \ref{eqn:expjnonoise}, as the supports of $w_*$ and ${\bf
+    1}_{[-\lambda,\lambda]}(t-(m-m_*))$ are disjoint for the range of
+  $m$ identified in the theorem.
 \end{proof}
-
-\noindent {\bf Remark.}
-One might have thought that $A=E^c_{\lambda}$ would be a better choice
-of annihilator, as for noise-free
-data, the solution set defined by the problem statement
-\ref{eqn:probstat0} is the same as the set of global minimizers of
-$\tJa$ in this case. However, for this choice of annihilator,
-$\tJa$ exhibits the same feature as the mean square error $e$, namely
-a continuum of stationary points at any distance from the global
-minimizer $m_*$ greater than a multiple of $\lambda$. Therefore the
-extended inversion with this choice of annihilator is no more amenable
-to local optimization than is FWI.
-
-A second possible 
-penalty operator penalizes energy away from 
-$t=0$: choose $\tau > 0$ and set 
-\begin{equation}
-  \label{eqn:ann}
-  a(t) = \min(|t|, \tau). 
-\end{equation}
-The cutoff $\tau$ will be chosen large enough to be effectively inactive: 
-specifically, hindsight suggests 
-\begin{equation}
-  \label{eqn:taudef}
-  \tau = \max\{|t_{\rm min}-m_{\rm min}r|,|t_{\rm min}-m_{\rm max}r|, |t_{\rm max}-m_{\rm min}r|, |t_{\rm max}-m_{\rm max}r|\}. 
-\end{equation}
-This 
-particular annihilator has been employed in earlier papers on extended 
-source inversion 
-\cite[]{Plessix:00a,LuoSava:11,Warner:14,HuangSymes:SEG15a,Warner:16,HuangSymes:GEO17}. It
-satisfies condition \ref{eqn:abnd} with $C=\lambda$.
-
-Surprisingly, the choice \ref{eqn:ann} yields an optimization
-problem with better global convergence properties:
 
@@
   \label{thm:rampgood}
-  For $a(t)=\min\{|t|,\tau\}$ in the definition \ref{eqn:annmult},
-  with $\tau$ given by equation \ref{eqn:taudef}, all
-  stationary points of $\tJa$ lie in the interval
-  \[
-    [m_*-\lambda/r, m_*+\lambda/r].
-  \]
+  Suppose that
+  \begin{enumerate}
+  \item $m_* \in M$;
+  \item $0 < \mu \le \lambda$, and $w_* \in W_{\mu}$;
+  \item $d_* = F[m_*]w_*$;
+  \item $a(t)=\min\{|t|,\tau\}$ in the definition \ref{eqn:annmult},
+    with $\tau$ given by equation \ref{eqn:taudef}; and
+  \item $\alpha > 0$.
+  \end{enumerate}
+  Then for any $m \in M$, 
+  \begin{equation}
+    | (m - m_*)r| > \lambda  \Rightarrow  \left|\frac{d}{dm}\tJa[m;d_*]\right| > \alpha^2 
+    \frac{\lambda-\mu}{(1+(4\pi r)^2\alpha^2 
+      (\lambda+\mu)^2)^{2}} \|w_*\|^2 
+    \label{eqn:gradbndnonoise}
+  \end{equation}
 \end{theorem}
 \begin{proof}
-  As observed before, $\mbox{supp }\aw[m;d] \subset [t_{\rm
+  As observed before, $\mbox{supp }\aw[m;d_*] \subset [t_{\rm
     min}-mr,t_{\rm max}-mr]\subset [-\tau,\tau]$, with $\tau$ defined
-  in \ref{eqn:taudef}. Therefore for $t \in  [t_{\rm min}-mr,t_{\rm
-    max}-mr]$, $a(t) = |t|$, and reason analogous to the proof of
-  Theorem \ref{thm:norminv}, item 2, shows that
-  \begin{equation}
-    \label{eqn:solnnon}
-    \aw[m,d](t)= \left(1+ (4\pi r)^2\alpha^2 t^2\right)^{-1}w_*\left(t+(m-m_*)r\right) 
-  \end{equation}
-  Straightforward modification of the proof of Theorem
-  \ref{thm:epjnonoise} gives
-  \begin{equation}
-    \label{eqn:expj}
-    \tJa[m;d] =\frac{\alpha^2}{2}\int\,dt\,(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
-    (t-(m-m^*)r)^2)^{-1}w_*(t)^2.
-  \end{equation}
-  A brief calculation yields
+  in \ref{eqn:taudef}. Therefore, $a(t) = |t|$ $a a'(t) = t$ in the
+  support of the integrand on the RHS of equation
+  \ref{eqn:dexpjnonoise}, which therefore 
+  becomes (after change of integration variable)
   \begin{equation}
     \label{eqn:gradfinal}
-    \frac{\partial}{\partial m}\tJa[m;d] = -r\alpha^2\int \,dt\,\frac{t}{(1+(4\pi r)^2\alpha^2 t^2)^2}w_*(t+(m-m_*)r)^2.
+    \frac{d}{dm}\tJa[m;d_*] = -r \alpha^2 \int \,dt \, 
+  t(1+(4\pi r)^2 \alpha^2 
+  t^2)^{-2}w_*(t+(m-m_*))^2.
   \end{equation}
@@
     = -r\alpha^2\int_{-(m-m_*)r-\lambda}^{-(m-m_*)r+\lambda}
-    \,dt\, \frac{t}{(1+(4\pi r)^2\alpha^2 t^2)^2}w_*\left(t+(m-m_*)r\right)^2
-  \]
-  If $m > m_*+\lambda/r$, then the entire interval of integration is a proper
-  subset of the negative half-axis. Consequently the first factor in the
-  integrand satisfies
-  \[
-    \frac{t}{(1+(4\pi r)^2\alpha^2 t^2)^2}< 0
-  \]
-  over the interval of integration. The integral is
-  negative, so the derivative
-  is positive. Similar reasoning applies to the case $m <
-  m_*-\lambda/r$. Thus
-  \begin{eqnarray}
-    m > m_*+\lambda/r & \Rightarrow & \frac{\partial \tJa}{\partial
-                                      m}[m;d]> 0; \nonumber\\
-    m < m_*-\lambda/r & \Rightarrow & \frac{\partial \tJa}{\partial
-                                      m}[m;d]< 0.
-                                      \label{eqn:leftright}
-  \end{eqnarray}
+    \,dt\, t(1+(4\pi r)^2\alpha^2 t^2)^{-2}w_*\left(t+(m-m_*)r\right)^2
+  \]
+  Suppose that $\mu \le \lambda$ and $w_* \in W_{\mu}$. 
+  If $m > m_*+\lambda/r$, then $t+(m-m_*)r \in \mbox{supp }w_*$
+  implies $-\mu - \lambda < t < \mu-\lambda<0$, so 
+  \[
+    t(1+(4\pi r)^2\alpha^2 t^2)^{-2} < (\mu-\lambda)(1+(4\pi r)^2\alpha^2 (\mu+\lambda)^2)^{-2}<0
+  \]
+  in the support of the integrand in equation
+  \ref{eqn:gradfinal}. Arguing similarly for $m<m_*-\lambda/r$, obtain
+  a similar inequality, implying the conclusion \ref{eqn:gradbndnonoise}.
 \end{proof}
 
-\noindent {\bf Remark.} That is, {\em with the choice of penalty
-  multiplier $a$ given in equation \ref{eqn:ann}, $\tJa$ has no local minima
-  further than  $O(\lambda)$ from the global minimum:} the gradient
-has the correct sign and slowness updates computed from it will be
-constructive, unless the slowness estimate is already ``within a
-wavelength'' of being correct.
-
-\noindent {\bf Remark.} The estimate $|m-m_*|r<\lambda$ for stationary
-points $m$ of $\tJa$ is sharp: it is possible to choose $w_* \in \lW$
-so that $\lambda - |m-m_*|r$ is as small as you like. In particular,
-the ``exact'' or ``true'' slowness $m_*$ is not necessarily the only 
-stationary point, or even a stationary point at all.
+\begin{corollary}
+  \label{thm:rampreallygood}
+  Suppose that
+  \begin{enumerate}
+  \item $m_* \in M$;
+  \item $0 <  \lambda$, and $w_* \in \lW$;
+  \item $d_* = F[m_*]w_*$;
+  \item $a(t)=\min\{|t|,\tau\}$ in the definition \ref{eqn:annmult},
+    with $\tau$ given by equation \ref{eqn:taudef}; 
+  \item $\alpha > 0$; and
+  \item$m \in M$ is a stationary point of $\tJa[\cdot;d_*]$.
+  \end{enumerate}
+  Then $|m-m_*| < \lambda /r$.
+\end{corollary}
+
+\begin{proof} Follows directly from Theorem \ref{thm:rampgood} by
+  taking $\mu=\lambda$.
+\end{proof}
 
@@
 \begin{proof}
-From the
-inequalities \ref{eqn:leftright}, $|(m-m_*)r|\le \mu$. From the
-identity \ref{eqn:solnnon}, $\mbox{supp }\aw[m_{\infty};d] \subset
+  From the assumption $w_* \in W_{\mu}$ and Corollary
+  \ref{thm:rampreallygood}, $|(m_{\infty}-m_*)r|\le \mu$. From the
+  identity \ref{eqn:solnonoise},
+  $\mbox{supp }\aw[m_{\infty};d] \subset
   [(m_{\infty}-m_*)r-\mu,(m_{\infty}-m_*)r+\mu] \subset
-[-2\mu,2\mu]$. Because of the support limitation, $a(t)=|t|$
-in the interval of integration appearing in \ref{eqn:residnorm}, so
+  [-2\mu,2\mu]$. Because of the support limitation, $a(t)=|t|$ in the
+  interval of integration appearing in \ref{eqn:residnorm}, so
 \[
@@
 
-[THE NEXT BIT MAY OR MAY NOT BE WORTH KEEPING - IT WAS INTERESTING TO
-WORK OUT]
-
-Without further contraints on data or solution, nothing more can be
-said about bounds on $\alpha$. If the data $d$ (and the target wavelet
-$w_*$) are assumed to have a square integrable derivative, then a
-necessary condition follows from the Heisenberg
-inequality (see for example \cite{Folland:07}, p. 255). To formulate
-this result in its most general form, introduce the Hilbert subspaces
-$V^0 \subset L^2(\bR), V^1 \subset H^1(\bR)$:
-\begin{eqnarray}
-  V^0 & = & \{ f \in L^2(\bR): Af \in L^2(\bR)\}, \nonumber\\
-  V^1 & = & V^0 \cap H^1(\bR).
-            \label{eqn:vdef}
-\end{eqnarray}
-$V^0$ is the domain of $A$, and equipped with the graph norm of $A$. A
-natural norm in $V^1$ is
-\[
-  \|f\|^2_{V^1} = \|f\|_{V^0}^2 + \|f\|_{H^1}^2.
-\]
-$V^j$ is the completion of $C_0^{\infty}(\bR)$ in the corresponding
-norm, j=0,1.
-
-\begin{proposition}
-  \label{thm:heis}
-For $w \in V^1$,
-  \begin{equation}
-    \label{eqn:heis}
-    \|Aw\|\|w'\| \ge \frac{1}{2}\|w\|^2
-  \end{equation}
-\end{proposition}
-
-\begin{proof}
-  For $w \in C_0^{\infty}(\bR)$,
-  \[
-    \int w^2 = \left|\int\,dt\, t (w(t)^2)' \right|= \left|2\int\,dt\,tw(t)w'(t)\right| \le
-    2\|Aw\| \|w'\|
-  \]
-  by the Cauchy-Schwarz inequality. Since $C_0^{\infty}(\bR)$ is dense
-  in $V^1$, the conclusion follows by continuity.
-\end{proof}
-
-In the conventional formulation of the Heisenberg inequality, the $L^2$ norm of
-$w'$ is replaced by the its equivalent in terms of the Fourier
-transform $\hat{w}$. Adopting temporarily the use of dummy variables
-in the expression of functions, the identity \ref{eqn:heis} turns into
-the usual form of the Heisenberg inequality: for $w \in V^1$,
-\begin{equation}
-\label{eqn:fheis}
-\|tw(t)\|\|k\hat{w}(k)\| \ge \frac{1}{4\pi}\|w\|^2.
-\end{equation}
-
-Define $\krms[w]$, the root mean square estimator of frequency of $w
-\in V^1$, by
-\begin{equation}
-  \label{eqn:krms}
-  \krms[w]=\frac{1}{2\pi}\frac{\|w'\|}{\|w\|} = \left(\int
-    \,dk\,\frac{|\hat{w}(k)|^2}{\|\hat{w}\|^2} k^2\right)^{1/2}.
-\end{equation}
-Then the inequalities \ref{eqn:heis}, \ref{eqn:fheis} can be rewritten as
-\begin{equation}
-  \label{eqn:frmsheis}
-  \|Aw\| \ge \frac{\|w\|}{4\pi \krms[w]}.
-\end{equation}
-
-For $\lambda >0$, define
-\begin{equation}
-  \label{eqn:wlam1}
-  \lW^1 = \lW \cap H^1(\bR).
-\end{equation}
-Note that $\lW^1 \subset V^1$ is a closed subspace for any
-$\lambda>0$.
-
-\begin{proposition}
-  \label{thm:klam}
-  For $w \in \lW^1$,
-  \[
-    \krms[w] \ge \frac{1}{4 \pi \lambda}
-  \]
-\end{proposition}
-
-\begin{proof}
-  Follows directly from inequality \ref{eqn:frmsheis} and the obvious bound
-  $\|A|_{\lW}\|\le \lambda$.
-\end{proof}
-
-\noindent{\bf Remark:} This result is the link mentioned earlier between the support
-constraint and the well-known frequency-based criteria for success of
-FWI. The result of Theorem \ref{thm:fwi} can be rephrased as showing the
-existence of many stationary points of the mean-square error function
-for which the travel time is in error by more than
-$\lambda_{\krms[w]|}$. In fact, the usual error criterion mentioned in the
-literature is that the initial estimate of travel time must be in error by at most ``half a
-wavelength'' if FWI is to converge reliably. This is correct in some
-circumstances, depending on features of the target wavelet $w_*$ of
-which the arguments in this paper do not take account.
-
-\begin{proposition}
-  \label{thm:heis2}
-  For $\lambda>0, w \in \lW^1$,
-  \begin{equation}
-    \label{eqn:heis2}
-    \|A^2w\| \ge \frac{\|w\|}{24 \lambda (2\pi\krms[w])^3}.
-  \end{equation}
-\end{proposition}
-
-\begin{proof}
-  Since $A$  preserves $\lW^1$, inequality \ref{eqn:heis} implies
-  \begin{equation}
-    \label{eqn:heissq}
-    \|A^2w\| \ge \frac{\|Aw\|^2}{2\|(Aw)'\|}.
-  \end{equation}
-  From the definition of $A$, $(Aw)'=w +Aw'$, so for $w \in \lW^1$,
-  \[
-    \|(Aw)'\| \le \|w\|+ \lambda\|w'\|.
-  \]
-  Applying \ref{eqn:heis} again,
-  \[
-    \|w\|^2 \le 2\|w'\|\|Aw\|\le 2\lambda \|w'\|\|w\|,
-  \]
-  so obtain the 1D Poincar\'{e} inequality: for $w \in \lW^1$,
-  \[
-    \|w\| \le 2\lambda\|w'\|.
-  \]
-  Thus
-  \[
-    \|(Aw)'\| \le 3\lambda\|w'\|.
-  \]
-  Apply this estimate together with the basic Heisenberg estimate
-  \ref{eqn:heis} to the inequality \ref{eqn:heissq} to obtain
-  \[
-    \|A^2w\| \ge \frac{\|w\|^4}{24\lambda \|w'\|^3}
-  \]
-  Rearranging and using the definition \ref{eqn:krms} of $\krms[w]$,
-  arrive at inequality \ref{eqn:heis2}.
-\end{proof}
-
-\begin{theorem}
-  \label{thm:ipnonoisenec}
-  Suppose that $a$ is given by definition \ref{eqn:ann}, $\alpha$,
-  $\mu \in (0,\lambda_{\rm max}/2]$, $d$ is given by
-  \ref{eqn:defdatanonoise} with $w_* \in W_{\mu}^1$, and that a
-  stationary point $m_{\infty}$ of $\tJa[\cdot;d]$ yields a solution
-  of the inverse problem \ref{eqn:probstat0} for $\lambda \ge 2\mu$,
-  $\epsilon>0$. Then
-\begin{equation}
-  \label{eqn:epsalpha}
-  3 (4\pi \lambda \krms[w_*])^3 \epsilon \ge \frac{(4\pi  r\alpha\lambda)^2}
-  {(1+(4\pi r \alpha \lambda)^2)}
-\end{equation}
-\end{theorem}
-
-\begin{proof}
-  Rearranging the identity \ref{eqn:residnorm}, and observing as in
-  the proof of Theorem \ref{thm:ipnonoisesuf} that the support of the
-  integrand is contained in $[-2\mu, 2\mu] \subset [\lambda,\lambda]$,
-  \[
-  e[m_{\infty},\aw[m_{\infty},d];d] 
-= 8 \pi^2 r^2 \alpha^4\int\,dt\,t^4(1+(4\pi r)^2 \alpha^2 t^2)^{-2}w_*(t+(m_{\infty}-m_*))^2
-\]
-\[
-  \ge \frac{8\pi^2r^2\alpha^4}{(1+(4\pi r)^2 \alpha^2
-    \lambda^2)^{2}}\int\,dt\,t^4w_*(t+(m_{\infty}-m_*))^2
-\]
-\begin{equation}
-  \label{eqn:residnormbis}
-  \ge \frac{8\pi^2r^2\alpha^4}{(1+(4\pi r)^2 \alpha^2
-    \lambda^2)^{2}}\|A^2w_*(\cdot + (m_{\infty}-m_*)r)\|^2.
- \end{equation}
-Since $\krms[w]$ is invariant under translation of $w \in V^1$,
-Proposition \ref{thm:heis2} implies that this is
-\[
-  \ge \frac{8\pi^2r^2\alpha^4}{(1+(4\pi r)^2 \alpha^2
-    \lambda^2)^{2}}\frac{\|w_*\|^2}{(24)^2 \lambda^2 (2\pi\krms[w_*])^6}
-\]
-\[
-  = \frac{1}{2}\frac{(4\pi r\alpha\lambda)^4}{(1+(4\pi r \alpha
-    \lambda)^2)^{2}}\frac{\|d\|^2}{(24)^2 (2\pi
-    \lambda \krms[w_*])^6}
-\]
-The pair $(m_{\infty}, \aw[m_{\infty},d])$ is presumed to solve the inverse
-problem as stated in \ref{eqn:probstat0}, in particular
-\[
-  \epsilon \ge (2 e[m_{\infty}, \aw[m_{\infty},d],d])^{1/2}/\|d\|
-\]
-the inequality \ref{eqn:epsalpha} follows.
-
-\end{proof}
-
-Inequality \ref{eqn:epsalpha} couples the dimensionless quantites
-$\epsilon$,  $4\pi r \lambda \alpha$, and $4 \pi \lambda
-krms[w_*]$. Proposition \ref{thm:klam}, implies that the left
-hand side is $\ge 3\epsilon$. Since the right hand side is $\le 1$,
-the inequality implies no limitation on $\alpha$ if the left hand side
-is $\ge 1$. For small $\epsilon$, the the largest permissible $\alpha$
-is $O(\sqrt{\epsilon})$. The permissible range of $\alpha$ increases with
-nondimensionalized RMS frequency $4\pi \lambda \krms[w_*]$. Since
-$\krms$ is invariant under translation and scaling of its argument,
-$\krms[w_*]=\krms[d]$, that is, the nondimensionalized RMS frequency
-is an observable property of the data, in the noise-free case at
-least.
-
 \subsection{The Effect of Noise}
@@
   \label{eqn:defdatanoisy}
-  d = F[m_*]w_* + n
-\end{equation}
-with $m_* \in M, w_* \in \lW$ as before, and noise trace $n \in
+  d = F[m_*]w_* + n = d_*+n,
+\end{equation}
+with $m_* \in M, w_* \in W_{\mu}$, $0<\mu<\lambda$, and noise trace $n \in
 D$. Since no support assumptions can be made about $n$, equation
-\ref{eqn:normsol} implies that $\aw[m;d] \notin \lW$ for any value of
-$\alpha$.  Therefore minimization of $\tJa$ cannot by itself yield a
+\ref{eqn:normsol} implies that $\aw[m;d] \notin \lW$ for any values of
+$\alpha$ and $\lambda$.  Therefore minimization of $\tJa$ cannot by itself yield a
 solution of the inverse problem as defined in the problem statement
-\ref{eqn:probstat0}.
-
-Recall that $\lE$ is the projection of $W$ onto $\lW$
-(i.e. multiplication by ${\bf 1}_{[-\lambda,\lambda]}$.
+\ref{eqn:probstat0}. In this section, we explain how a solution may
+nontheless be constructed from a stationary point of $\tJa$.
+
+First we examine the effect of additive noise on the estimation of the
+slowness $m$. In expressing the result, we use the dimensionless
+relative data error
+\begin{equation}
+  \label{eqn:defeta}
+  \eta = \frac{\|n\|}{\|d_*\|}. 
+\end{equation}
 
 \begin{theorem}
-  \label{thm:ipnoisesuf}
-  Suppose that $a$ is given by definition \ref{eqn:ann}, $\alpha$,
-  $\mu \in (0,\lambda_{\rm max}]$, $d$ is given by
-  \ref{eqn:defdatanoise} with $w_* \in W_{\mu}$, $n \in D$, and
-  $m_{\infty}$ is a stationary point of $\tJa[\cdot;d]$. Then
-  $(m_{\infty},\lE\aw[m_{\infty};d])$ is a solution of the inverse
-  problem \ref{eqn:probstat0} for any $\lambda \ge 2\mu$ and
-  $\epsilon \ge (4\pi r \lambda \alpha)^2 + C\|n\|/\alpha$.
+  \label{thm:mnoise}
+  Assume the hypotheses of Theorem \ref{thm:rampgood}, and that $d$ is
+  given by definition \ref{eqn:defdatanoisy}. Suppose that $m \in M$
+  is a stationary point of $\tJa[\cdot;d]$, and that
+  \begin{equation}
+    \label{eqn:mnoisebnd}
+    \eta(1+\eta) \le \frac{16}{3\sqrt{3}}\frac{4\pi r \alpha
+      (\lambda-\mu)}{(1+(4\pi r\alpha(\lambda+\mu))^2)^2}
+  \end{equation}
+  Then $|m-m_*| \le \lambda /r$.
 \end{theorem}
 
-\section{The Discrepancy Principle}
+\begin{proof}
+  From equation \ref{eqn:dexpjgen}, $d \tJa / dm$ is the
+value of a quadratic form in $d$ with (indefinite) symmetric operator
+$B = $ multiplication by
+\[
+  b(t;m,\alpha)  = -\frac{(4 \pi r \alpha)^2 (t-mr)}{(1+(4\pi r \alpha (t-mr))^2)^{2}}
+\]
+Therefore
+\begin{equation}
+  \label{eqn:gradlip}
+  \left|\frac{d}{dm}\tJa[m;d]-\frac{d}{dm}\tJa[m;d_*]\right| =
+  |\langle (d+d_*),B(d-d_*)\rangle| \le \max_{t \in
+    \bR}|b(t;m,\alpha)|\eta(1+\eta)\|d_*\|^2
+\end{equation}
+A straightforward calculation shows that
+\[
+  \max_{t \in \bR} b(t;m,\alpha) = \frac{3\sqrt{3}}{16} 4\pi r\alpha.
+\]
+For a stationary point $m$ of
+$\tJa[\cdot;d]$, the inequality \ref{eqn:gradlip} implies
+\[
+  \left|\frac{d}{dm}\tJa[m;d_*]\right| \le \frac{3\sqrt{3}}{16} 4\pi
+  r\alpha \eta(1+\eta)\|d_*\|^2
+\]
+On the other hand, the conclusion \ref{eqn:gradbndnonoise} of Theorem
+\ref{thm:rampgood} implies that if also
+\[
+  \frac{3\sqrt{3}}{16} 4\pi r\alpha \eta(1+\eta)\|d_*\|^2 \le (4 \pi r
+  \alpha)^2 \frac{\lambda-\mu}{(1+(4\pi r)^2\alpha^2
+    (\lambda+\mu)^2)^{2}} \|d_*\|^2
+\]
+then $|m-m_*|\le \lambda/r$. Rearranging, obtain the conclusion.
+\end{proof}
+
+\begin{corollary}
+  \label{thm:mnoisecor}
+  Asumme the hypotheses of Theorem \ref{thm:mnoise}, in particular
+  that $m$ is a stationary point of $\tJa[\cdot;d]$, $d=d_*+n$. Then
+  \begin{itemize}
+    \item[1. ] the bound \ref{eqn:mnoisebnd} holds only if 
+      noise level $\eta=\|n\|/\|d_*\|$ satisfies
+      \begin{equation}
+        \label{eqn:mnoisecond}
+        \eta < \frac{1+\sqrt{5}}{2} < 1;
+      \end{equation}
+    \item[2. ] if condition \ref{eqn:mnoisecond} holds, then
+%      $|m-m_*|<\lambda/r$ provided that
+%      \begin{equation}
+%        \label{eqn:mnoisesuff}
+%        \frac{\lambda}{\mu} \ge 1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}. 
+%      \end{equation}
+      \begin{equation}
+        \label{eqn:mnoisesuff}
+        |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}. 
+      \end{equation}
+    \end{itemize}
+\end{corollary}
+
+\begin{proof}
+  Write $\lambda = (1+\delta)\mu$, and $x=4 \pi r \alpha \mu$. Then
+  the right-hand side of equation \ref{eqn:mnoisebnd} may be written as
+  \begin{equation}
+    \label{eqn:mnoisebndrev}
+    \frac{16}{3\sqrt{3}}\frac{4\pi r \alpha
+      (\lambda-\mu)}{(1+(4\pi r\alpha(\lambda+\mu))^2)^2} = D
+    \frac{x}{(1+C^2 x^2)^2},
+  \end{equation}
+  where
+  \[
+    D=\frac{16}{3\sqrt{3}}\delta,\,C=2+\delta.
+  \]
+  The positive stationary point of the quantity on the right-hand side
+  of \ref{eqn:mnoisebndrev} is a maximum, and occurs at
+  $x=1/(\sqrt{3}C)$, that is
+  \[
+    4 \pi r \alpha \mu = \frac{1}{\sqrt{3}(2+\delta)}.
+  \]
+  Thus
+  \[
+    1+C^2x^2 = \frac{4}{3}
+  \]
+  hence the maximum value is
+  \[
+    \frac{D3\sqrt{3}}{16C} = \frac{\delta}{2+\delta}.
+  \]
+  This maximum value must be larger than the left hand side of inequality
+  \ref{eqn:mnoisebnd} in order that there be any solutions at all, but
+  is less than $1$. This observation establishes conclusion 1 of the
+  theorem. Also, existence of $\lambda$ satisfying \ref{eqn:mnoisebnd}
+  implies that
+  \[
+    \eta(1+\eta) \le \frac{\delta}{2+\delta}.
+  \]
+  Solving this inequality for $\delta$ and unwinding the definitions
+  leads to conclusion 2.
+\end{proof}
+
+\noindent {\bf Remark:} We emphasize that Theorem \ref{thm:mnoise} and
+Corollary \ref{thm:mnoisecor} state {\em sufficient} conditions for a bound on
+the slowness error $|m-m_*|$ in terms of the relative data noise level $\eta$,
+giving an additional ``fudge factor'' beyond the support size $\mu$
+of the noise-free wavelet $w_*$ for an interval within which the slowness error is
+guaranteed to lie.
+
+Conclusion 1 in Corollary \ref{thm:mnoisecor} constrains the range of
+noise level to which these results apply to a bit more than 60\%. That
+is, the bound given by conclusion 2 is useful only for small noise. In
+the limit as $\eta \rightarrow 0$, conclusion 2 becomes
+$\lambda/\mu \gtrsim 1 + 2\eta$, that is, the ``fudge factor'' beyond
+the noise-free bound is approximatly twice the noise level.
+
+On the other hand, stronger bounds than given by Corollary
+\ref{thm:mnoisecor} are possible, given additional constraints on the
+noise $n$. A natural example is uniformly distributed random noise,
+filtered to have the same spectrum as the source. The expression
+\ref{eqn:dexpjgen} implies that the interaction of noise $n$ and
+signal $d_*$ in the derivative of $\tJa$ is local, so that the
+coefficient of $\eta$ on the left-hand side of inequality
+\ref{eqn:mnoisebnd} is effectively much less that 1, resulting in a
+larger range of allowable $\eta$. While we will not formulate such a
+result, one of the numerical examples below suggests its feasibility.
+
+    % Recall that $\lE$ is the projection of $W$ onto $\lW$
+%(i.e. multiplication by ${\bf 1}_{[-\lambda,\lambda]}$.
+
+\begin{proof} of Theorem \ref{thm:ipnoisesuf}:
+From Theorem \ref{thm:norminv}, 
+\[
+  \aw[m;d](t) = (1+ (4 \pi r \alpha t)^2)^{-1}(w_*(t+(m-m_*)) + 4\pi r 
+  n(t+mr)) 
+\]
+\[
+  = \aw[m;d_*] + (1+ (4 \pi r \alpha t)^2)^{-1}4\pi r n(t+mr)
+\]
+
+From Corollary \ref{thm:mnoisecor},
+\[
+  |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r} 
+\]
+\[
+  =\left(2+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}
+  -\frac{\mu}{r} \le \left(\frac{\lambda}{\mu}-1\right)\frac{\mu}{r}
+\]
+That is,
+\[
+  |m-m_*|r \le \lambda-\mu.
+\]
+From Theorem \ref{thm:norminv},
+\[
+  \mbox{supp }\aw[m;d_*] \subset [-\mu-(m-m_*)r, \mu-(m-m_*)r]
+  \subset [-\lambda,\lambda]
+\]
+so
+\[
+  E_{\lambda}\aw[m;d](t) = \aw[m;d_*](t) + E_{\lambda}(1+ (4 \pi r \alpha
+  t)^2)^{-1}4\pi r n(t+mr)
+\]
+From the definition of $F[m]$, for any $t_1<t_2$, $w \in W$,  
+\[
+  F[m]{\bf 1}_{[t_1.t_2]}w = {\bf 1}_{[t_1+mr,t_2+mr]}F[m]w  
+\]
+Thus the data residual after projection is
+\[
+  F[m]E_{\lambda}\aw[m,d](t) -d(t) = F[m]\aw[m,d_*](t) -d_*(t)  
+\]
+\[
+  + {\bf 1}_{[-\lambda+mr,\lambda+mr]}(4 \pi r \alpha (t-mr))^2 (1+ (4 \pi
+  r \alpha (t-mr))^2)^{-1} n(t)
+\]
+\[
+  -(1- {\bf 1}_{[-\lambda+mr,\lambda+mr]})n(t)
+\]
+From \ref{eqn:residnorm} and the bound on $m-m_*$,
+\[
+  \| F[m]E_{\lambda}\aw[m,d_*] -d*\|^2 = (4 \pi r \alpha)^4
+  \int_{-\lambda+(m-m_*)r}^{\lambda+(m-m_*)r}\,dt\, (t-(m-m_*)r)^4
+\]
+\[
+  \times (1+(4\pi r \alpha)^2 
+  (t-(m-m^*)r)^2)^{-2}d_*(t)^2
+\]
+\[
+  \le (4 \pi r \alpha \lambda)^4\|d_*\|^2
+\]
+Similarly, the norm squared of the sum of the last two terms is 
+\[
+  \le (4 \pi r \alpha \lambda)^2 \|{\bf 1}_{[-\lambda+mr,\lambda+mr]}
+  n\|^2 + \|(1 - {\bf 1}_{[-\lambda+mr,\lambda+mr]}n\|^2
+\]
+Without additional hypotheses to outlaw the accumulation of $n$ near
+$t=mr$, all that can be said is that this is
+\[
+  \le \max \{(4 \pi r \alpha \lambda)^2, 1\} \|n\|^2
+\]
+Putting this all together,
+\[
+  \|F[m]E_{\lambda}\aw[m,d]-d\| \le 4 \pi r \alpha
+  \lambda(\max\{1,(4\pi r \alpha \lambda)^{-2}\}\eta^2 + (4
+  \pi r \alpha \lambda)^2)^{1/2}\|d_*\|
+\]
+as asserted.
+\end{proof}
+
+\bibliographystyle{seg}
+\bibliography{../../bib/masterref}
+
+\append{Numerics and penalty selection via the Discrepancy Principle}
 As described in the introduction, the discrepancy principle (in one of
@@
 objective definition \ref{eqn:red} so that the data error term at the
-inner solution, $e[m,\aw[m;d]]$, lies in this range. Since updating
+inner solution, $e[m,\aw[m;d],d]$, lies in this range. Since updating
 $m$ changes the inner problem, the appropriate condition for such
@@
 
-Accordingly, regard $m$ as fixed and suppress it from the notation for the remainder of this section, and introduce the abbreviations
+Accordingly, regard $m$ as fixed and suppress it from the notation for the remainder of ;dthis section, and introduce the abbreviations
 \begin{eqnarray}
-\label{eqn:erralph}
-e(\alpha) &=& e[m,\aw[m;d]]\\
+\label{eqn:erralph}a
+e(\alpha) &=& e[m,\aw[m;d],d]\\
 \label{eqn:penalph}
-p(\alpha) &=& p[m,\aw[m;d]]
+p(\alpha) &=& p[m,\aw[m;d],d]
 \end{eqnarray}
@@
 
-\section{stuff}
-In many cases, the constraint posed by membership of the second component $w$ in the subspace $\lW$ is an approximation to an idealized limit $\lambda \rightarrow 0$.
-It may be that $W_0 = \cap_{\lambda > 0}\lW$ is non-trivial, in which
-case $F$ may be viewed as an extension of $F_0 = F|_{ M \times
-  W_0}$. In that case, $F_0$ and $F$ realize the idealized extension
-structure discussed in many prior works on extended inversion (for
-example, \cite[]{geoprosp:2008}): $F_0$ is the physical modeling
-operator, and $W_0$ consists of second components of physical or
-feasible models. However, in many of these examples, $W_0$ is not
-computationally accessible, or - even if it is - gives results that
-closely approximate those obtained with $\lW$ for small $\lambda$. For
-other examples, for instance that  discussed later in this paper (and
-implicitly in \cite[]{Warner:16}, for instance), $W_0=\{0\}$ and the
-family $\{\lW: \lambda > 0\}$ actually embodies all of the necessary
-physics.
-
-\bibliographystyle{seg}
-\bibliography{../../bib/masterref}
+\append{Heisenberg, support, and wavelength}
+
+[THE NEXT BIT MAY OR MAY NOT BE WORTH KEEPING - IT WAS INTERESTING TO
+WORK OUT]
+
+Without further contraints on data or solution, nothing more can be
+said about bounds on $\alpha$. If the data $d$ (and the target wavelet
+$w_*$) are assumed to have a square integrable derivative, then a
+necessary condition follows from the Heisenberg
+inequality (see for example \cite{Folland:07}, p. 255). To formulate
+this result in its most general form, introduce the Hilbert subspaces
+$V^0 \subset L^2(\bR), V^1 \subset H^1(\bR)$:
+\begin{eqnarray}
+  V^0 & = & \{ f \in L^2(\bR): Af \in L^2(\bR)\}, \nonumber\\
+  V^1 & = & V^0 \cap H^1(\bR).
+            \label{eqn:vdef}
+\end{eqnarray}
+$V^0$ is the domain of $A$, and equipped with the graph norm of $A$. A
+natural norm in $V^1$ is
+\[
+  \|f\|^2_{V^1} = \|f\|_{V^0}^2 + \|f\|_{H^1}^2.
+\]
+$V^j$ is the completion of $C_0^{\infty}(\bR)$ in the corresponding
+norm, j=0,1.
+
+\begin{proposition}
+  \label{thm:heis}
+For $w \in V^1$,
+  \begin{equation}
+    \label{eqn:heis}
+    \|Aw\|\|w'\| \ge \frac{1}{2}\|w\|^2
+  \end{equation}
+\end{proposition}
+
+\begin{proof}
+  For $w \in C_0^{\infty}(\bR)$,
+  \[
+    \int w^2 = \left|\int\,dt\, t (w(t)^2)' \right|= \left|2\int\,dt\,tw(t)w'(t)\right| \le
+    2\|Aw\| \|w'\|
+  \]
+  by the Cauchy-Schwarz inequality. Since $C_0^{\infty}(\bR)$ is dense
+  in $V^1$, the conclusion follows by continuity.
+\end{proof}
+
+In the conventional formulation of the Heisenberg inequality, the $L^2$ norm of
+$w'$ is replaced by the its equivalent in terms of the Fourier
+transform $\hat{w}$. Adopting temporarily the use of dummy variables
+in the expression of functions, the identity \ref{eqn:heis} turns into
+the usual form of the Heisenberg inequality: for $w \in V^1$,
+\begin{equation}
+\label{eqn:fheis}
+\|tw(t)\|\|k\hat{w}(k)\| \ge \frac{1}{4\pi}\|w\|^2.
+\end{equation}
+
+Define $\krms[w]$, the root mean square estimator of frequency of $w
+\in V^1$, by
+\begin{equation}
+  \label{eqn:krms}
+  \krms[w]=\frac{1}{2\pi}\frac{\|w'\|}{\|w\|} = \left(\int
+    \,dk\,\frac{|\hat{w}(k)|^2}{\|\hat{w}\|^2} k^2\right)^{1/2}.
+\end{equation}
+Then the inequalities \ref{eqn:heis}, \ref{eqn:fheis} can be rewritten as
+\begin{equation}
+  \label{eqn:frmsheis}
+  \|Aw\| \ge \frac{\|w\|}{4\pi \krms[w]}.
+\end{equation}
+
+For $\lambda >0$, define
+\begin{equation}
+  \label{eqn:wlam1}
+  \lW^1 = \lW \cap H^1(\bR).
+\end{equation}
+Note that $\lW^1 \subset V^1$ is a closed subspace for any
+$\lambda>0$.
+
+\begin{proposition}
+  \label{thm:klam}
+  For $w \in \lW^1$,
+  \[
+    \krms[w] \ge \frac{1}{4 \pi \lambda}
+  \]
+\end{proposition}
+
+\begin{proof}
+  Follows directly from inequality \ref{eqn:frmsheis} and the obvious bound
+  $\|A|_{\lW}\|\le \lambda$.
+\end{proof}
+
+\noindent{\bf Remark:} This result is the link mentioned earlier between the support
+constraint and the well-known frequency-based criteria for success of
+FWI. The result of Theorem \ref{thm:fwi} can be rephrased as showing the
+existence of many stationary points of the mean-square error function
+for which the travel time is in error by more than
+$\lambda \ge 1/{4 \pi \krms[w]|}$. In fact, the usual error criterion mentioned in the
+literature is that the initial estimate of travel time must be in error by at most ``half a
+wavelength'' if FWI is to converge reliably. This is correct in some
+circumstances, depending on features of the target wavelet $w_*$ of
+which the arguments in this paper do not take account.
+
+\begin{proposition}
+  \label{thm:heis2}
+  For $\lambda>0, w \in \lW^1$,
+  \begin{equation}
+    \label{eqn:heis2}
+    \|A^2w\| \ge \frac{\|w\|}{24 \lambda (2\pi\krms[w])^3}.
+  \end{equation}
+\end{proposition}
+
+\begin{proof}
+  Since $A$  preserves $\lW^1$, inequality \ref{eqn:heis} implies
+  \begin{equation}
+    \label{eqn:heissq}
+    \|A^2w\| \ge \frac{\|Aw\|^2}{2\|(Aw)'\|}.
+  \end{equation}
+  From the definition of $A$, $(Aw)'=w +Aw'$, so for $w \in \lW^1$,
+  \[
+    \|(Aw)'\| \le \|w\|+ \lambda\|w'\|.
+  \]
+  Applying \ref{eqn:heis} again,
+  \[
+    \|w\|^2 \le 2\|w'\|\|Aw\|\le 2\lambda \|w'\|\|w\|,
+  \]
+  so obtain the 1D Poincar\'{e} inequality: for $w \in \lW^1$,
+  \[
+    \|w\| \le 2\lambda\|w'\|.
+  \]
+  Thus
+  \[
+    \|(Aw)'\| \le 3\lambda\|w'\|.
+  \]
+  Apply this estimate together with the basic Heisenberg estimate
+  \ref{eqn:heis} to the inequality \ref{eqn:heissq} to obtain
+  \[
+    \|A^2w\| \ge \frac{\|w\|^4}{24\lambda \|w'\|^3}
+  \]
+  Rearranging and using the definition \ref{eqn:krms} of $\krms[w]$,
+  arrive at inequality \ref{eqn:heis2}.
+\end{proof}
+
+\begin{theorem}
+  \label{thm:ipnonoisenec}
+  Suppose that $a$ is given by definition \ref{eqn:ann}, $\alpha$,
+  $\mu \in (0,\lambda_{\rm max}/2]$, $d$ is given by
+  \ref{eqn:defdatanonoise} with $w_* \in W_{\mu}^1$, and that a
+  stationary point $m_{\infty}$ of $\tJa[\cdot;d]$ yields a solution
+  of the inverse problem \ref{eqn:probstat0} for $\lambda \ge 2\mu$,
+  $\epsilon>0$. Then
+\begin{equation}
+  \label{eqn:epsalpha}
+  3 (4\pi \lambda \krms[w_*])^3 \epsilon \ge \frac{(4\pi  r\alpha\lambda)^2}
+  {(1+(4\pi r \alpha \lambda)^2)}
+\end{equation}
+\end{theorem}
+
+\begin{proof}
+  Rearranging the identity \ref{eqn:residnorm}, and observing as in
+  the proof of Theorem \ref{thm:ipnonoisesuf} that the support of the
+  integrand is contained in $[-2\mu, 2\mu] \subset [\lambda,\lambda]$,
+  \[
+  e[m_{\infty},\aw[m_{\infty},d];d] 
+= 8 \pi^2 r^2 \alpha^4\int\,dt\,t^4(1+(4\pi r)^2 \alpha^2 t^2)^{-2}w_*(t+(m_{\infty}-m_*))^2
+\]
+\[
+  \ge \frac{8\pi^2r^2\alpha^4}{(1+(4\pi r)^2 \alpha^2
+    \lambda^2)^{2}}\int\,dt\,t^4w_*(t+(m_{\infty}-m_*))^2
+\]
+\begin{equation}
+  \label{eqn:residnormbis}
+  \ge \frac{8\pi^2r^2\alpha^4}{(1+(4\pi r)^2 \alpha^2
+    \lambda^2)^{2}}\|A^2w_*(\cdot + (m_{\infty}-m_*)r)\|^2.
+ \end{equation}
+Since $\krms[w]$ is invariant under translation of $w \in V^1$,
+Proposition \ref{thm:heis2} implies that this is
+\[
+  \ge \frac{8\pi^2r^2\alpha^4}{(1+(4\pi r)^2 \alpha^2
+    \lambda^2)^{2}}\frac{\|w_*\|^2}{(24)^2 \lambda^2 (2\pi\krms[w_*])^6}
+\]
+\[
+  = \frac{1}{2}\frac{(4\pi r\alpha\lambda)^4}{(1+(4\pi r \alpha
+    \lambda)^2)^{2}}\frac{\|d\|^2}{(24)^2 (2\pi
+    \lambda \krms[w_*])^6}
+\]
+The pair $(m_{\infty}, \aw[m_{\infty},d])$ is presumed to solve the inverse
+problem as stated in \ref{eqn:probstat0}, in particular
+\[
+  \epsilon \ge (2 e[m_{\infty}, \aw[m_{\infty},d],d])^{1/2}/\|d\|
+\]
+the inequality \ref{eqn:epsalpha} follows.
+
+\end{proof}
+
+Inequality \ref{eqn:epsalpha} couples the dimensionless quantites
+$\epsilon$,  $4\pi r \lambda \alpha$, and $4 \pi \lambda
+\krms[w_*]$. Proposition \ref{thm:klam}, implies that the left
+hand side is $\ge 3\epsilon$. Since the right hand side is $\le 1$,
+the inequality implies no limitation on $\alpha$ if the left hand side
+is $\ge 1$. For small $\epsilon$, the the largest permissible $\alpha$
+is $O(\sqrt{\epsilon})$. The permissible range of $\alpha$ increases with
+nondimensionalized RMS frequency $4\pi \lambda \krms[w_*]$. Since
+$\krms$ is invariant under translation and scaling of its argument,
+$\krms[w_*]=\krms[d]$, that is, the nondimensionalized RMS frequency
+is an observable property of the data, in the noise-free case at
+least.

### 2020-09-27 repro/papers/discrep/alg.tex
added alg, figs, makefigs, modified fig captions

\title{Discrepancy Algorithm for Single-Trace Transmission Inverse Problem}
\author{William W. Symes}

\lefthead{Symes}

\righthead{Algorithm}

\maketitle
\begin{abstract}
  Concise statement of the discrepancy algorithm for solving the single-trace transmission inverse problem.
\end{abstract}

\section{Definitions}
``The paper'' throughout refers to ``Solution of an Acoustic Transmission Inverse Problem by Extended Inversion'', current draft.

Note that I have changed the notation for the penalty term to $g$, rather than $p$ as has been the case before, in order to avoid conflict with the conventional notation for the acoustic pressure field.

The mean square error $e$, penalty term $g$, and extended objective $\Ja $ are defined in equations 4, 5, and 6. These quantities are normalized via division by the norm-squared of the data. In recent drafts I also introduced a notation for the relative $L^2$ error = noise-to-signal ratio:
\begin{equation}
  \label{eqn:rmsgen}
  r[m,w;d] = \|F[m]w-d\|/\|d\| = \sqrt{2 e[m,w;d]}
\end{equation}
The algorithm presented here assumes that the multiplier $a$ appearing in equation 39 (and in the definition of the annihilator $A$) is given by $a(t)=|t|$. The paper explains how this choice can be regarded as defining a bounded $A$, due to the bounded range of time allowed in the data.

The minimizer of $\Ja $ over the wavelet $w$ is given by solution of the normal equation. Under the conditions stated in Proposition 2, this minimizer is given in equation 39 of the paper:
\begin{equation}
  \label{eqn:normsol}
  \aw[m;d](t) = \left\{
    \begin{array}{c}
      \left(\frac{1}{(4\pi r)^2} + \alpha^2
      t^2\right)^{-1}\frac{1}{4 \pi r}d(t+mr), t \in [t_{\rm
      min}-mr, t_{\rm max}-mr];\\
      0, \mbox{ else;}
    \end{array}
  \right.
\end{equation}
The reduced quantities for the Variable Projection Method are obtained by substituting $\aw[m;d]$ for $w$ in equations 4, 5, and 6 to obtain equations 41, 42, and 43, then using $a(t)=|t|$:
\begin{equation}
  \label{eqn:residnormgen}
  e[m,\aw[m,d];d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r \alpha (t-mr))^4(1 +
  (4\pi r \alpha (t-mr))^2)^{-2}d(t)^2
\end{equation}
\begin{equation}
  \label{eqn:anninormgen}
  g[m,\aw[m,d];d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r (t-mr))^2(1 +
  (4\pi r \alpha (t-mr))^2)^{-2}d(t)
\end{equation}

\begin{equation}
  \label{eqn:expjgen}
\tJa[m;d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}}\,dt\,(4\pi r \alpha (t-mr))^2(1+(4\pi r \alpha 
(t-mr))^2)^{-1}d(t)^2. 
\end{equation}
The derivative with respect to $m$ of $\tJa $ is
\begin{equation}
  \label{eqn:dexpjgen}
  \frac{d}{dm}\tJa[m;d] = -\frac{(4 \pi r \alpha)^2}{\|d\|^2} \int_{t_{\rm min}}^{t_{\rm max}} \,dt \,  (t-mr)(1+(4\pi r \alpha (t-mr))^2)^{-2}d(t)^2. 
\end{equation}

The data $d$ is sampled on a uniform grid of cell size $dt$ (in matlab notation, $t_{\rm min}:dt:t_{\rm max}$). The integrals in the preceding equations, and the integral in the defintion of $\|d\|^2$, should be approximated by a quadrature rule. The trapezoidal rule should be accurate enough, provided that $d$ is bandlimited. In our numerical experiments, both noise-free data and noise are obtained by convolving numerical delta functions (``spikes'') with sampled Ricker wavelets. If the center frequency of the Ricker wavelet is $f$, then the usual rule of thumb is that Fourier components above frequencies of $2.5f$ are negligible. Sampling at 10 gridpoints per wavelength is a roughly adequate choice for the trapezoidal rule.

For example, if the center frequency is 20 Hz, Fourier components at frequencies above 50 Hz are negligible. 50 Hz corresponds to a wavelength of 0.02 s, so a suitable sample interval for our calculations is $dt = 0.002$ s.

\section{Algorithm}
The ideal goal of the discrepancy algorithm is:

\begin{quote}
  Given data $d$, a target noise-to-signal ratio $r_{\rm tgt}$ (or equivalently a target signal-to-noise ratio of $1/r_{\rm tgt}$), and a fudge factor $C<1$, find a stationary point $m_{\infty}$ of the VPM objective $\tJa[\cdot;d]$ so that
\begin{equation}
  \label{eqn:discrep}
  r_-=Cr_{\rm tgt} \le r[m,\aw[m;d];d] \le r_+= C^{-1}r_{\rm tgt}, 
\end{equation}
or equivalently
\begin{equation}
  \label{eqn:discrepsq}
  e_-=C^2e_{\rm tgt} \le e[m,\aw[m;d];d] \le e_+=C^{-2}e_{\rm tgt}.
\end{equation}
\end{quote}
Note: in the Fu paper, $e_{\pm}$ were denoted $X_{\pm}$, and $e_{\rm tgt}$ was denoted $X$.

This statement is ideal because an actual stationary point cannot be found exactly. Instead, accept an approximate stationary point $m_{\rm approx}$ defined by a choice of tolerance $\delta$ for the norm of the gradient:
\begin{equation}
  \label{eqn:approx}
  \left|\frac{d \tJa}{dm}[m_{\rm approx};d]\right| \le \delta.
\end{equation}

Schematic statement of the discrepancy algorithm:
\begin{algorithm}[H]
\caption{Discrepancy algorithm for updating $m, \alpha$}
\begin{algorithmic}[1]
  \State print parameters: $e_{\rm tgt},e_-,e_+,\delta$, max number of iterations allowed for outer loop, $\alpha$ update loop, $m$ update loop.
  \State print target slowness $m_*$, plot target wavelet $w_*$, noise-free data $d_*=F[m_*]w_*$, noise $n$, data $d=d_*+n$
  \State Choose $m\in M$, $\alpha=0$;
  \Repeat
  \Repeat
  \State print iteration number, $m$, $\alpha$, $r$, $e$, $g$, $\tJa$ before update
  \State update $\alpha$ for fixed $m$
  \State print $m$, $\alpha$, $r$, $e$, $g$, $\tJa$ after update  
  \Until inequalities \ref{eqn:discrepsq} are satisfied
  \Repeat
  \State print iteration number, $m$, $\alpha$, $r$, $e$, $g$, $\tJa$ before update
  \State update $m$ for fixed $\alpha$
  \State print $m$, $\alpha$, $r$, $e$, $g$, $\tJa$ after update  
  \Until{inequalities \ref{eqn:discrepsq} are not satisfied, or $\|\nabla \tJa[m;d]\|$
    sufficiently small}
  \Until{inequalities \ref{eqn:discrepsq} are satisfied}
  \State print final values of $m$, $\alpha$, $r$, $e$, $g$, $\tJa$
  \State plot estimated wavelet $\aw[m;d]$, estimated data $F[m]\aw[m;d]$, and residual $F[m]\aw[m;d]-d$
\end{algorithmic}
\end{algorithm}

Note: Each loop (outer, $m$, $\alpha$ needs to be supplied with a limit on the number of iterations. If a loop comes to the limit without satisfying the convergence criterion for that loop (``until''), then the algorithm should exit with an appropriate failure message.

Note: plots of data ($d_*$, $n$, $d$, $F[m]\aw[m;d]$, $F[m]\aw[m;d]-d$) should have same vertical scale. Same for plots of wavelets ($w_*$, $\aw[m;d]$).

\section{Update $\alpha $}
The Fu paper (repeated in Appendix A in the current draft) sketches a proof that if $e_{\rm curr} = e[m,w_{\alpha_{\rm curr}}[m;d];d]< e_+$, then  $\alpha_{\rm next}$ given by  
\begin{equation}
\label{eqn:alphasecant}
\alpha^2_{\rm next} = \alpha^2_{\rm curr} + \frac{e_{+}-e_{\rm curr}}{2p[m,w_{\alpha_{\rm curr}}[m;d];d]} 
\end{equation}
satisfies
\[
  e_{\rm curr}  < e_{\rm next} =  e[m,w_{\alpha_{\rm next}}[m;d];d]  < e_+.
\]
That is, the update rule \ref{eqn:alphasecant} produces an increasing sequence of values of $e$ in the interval $[0,e_+)$.

We are currently lacking a proof that in fact the sequence of $e$ values so generated converges to $e_+$ for fixed $m$, but numerical evidence is very strong that this is so. Therefore we expect the stopping criterion in line 9 of the algorithm to be satisfied after some number of iterations, so this is a suitable update rule for $\alpha$.

This iteration does appear to be fairly slow, compared to some alternatives. However it is reliable. Since the current computations are very inexpensive, use it.

\section{Update $m$}

Any local optimization algorithm may be used, provided that is has the property called ``global convergence'' in the literature of optimization. This property does {\em not} mean ``converges to the global optimum''. In fact it means ``converges globally (starting from any initial guess) to a local minimum''.

An example of such an algorithm is steepest descent with backtracking line search. The line search should include a ``sufficient decrease'' test. Huiyi's recent algorithm is a good choice. Also, good implementations of Brent's method for one-variable minimization are also globally convergent.

The intent of the discrepancy algorithm design is that eventually the gradient decrease criterion will be satisfied, that is, that the iteration will converge before the bounds \ref{eqn:discrepsq} are violated. That is why we use membership in an interval containing $e_{\rm tgt}$, rather than equality with $e_{\rm tgt}$, as the termination criterion. When the $m$ update converges (small gradient) while the bounds \ref{eqn:discrepsq} are satisfied, the outer iteration (over both $m$ and $\alpha$) will terminate, with a satisfactory solution (line 15).

Again, there is convincing numerical evidence that this will generally happen, but a proof is currently lacking.

### 2020-09-27 repro/papers/discrep/paper.tex
interim

@@
 
-This paper examines one of a number of proposed alternatives to least squares
-FWI, based on an extension of the model-to-data mapping.  We show
-exactly how an example of this {\em extended inversion} approach leads
-to successful solution of a very simple acoustic transmission inverse
-problem. Despite its simplicity, this problem exhibits several
-important characteristics of all inverse problems in wave propagation
-in which wave velocities are amongst the unknowns. It has the
-advantage of permitting explicit calculations that reveal clearly the
-failure mechanism for FWI, that is, straightforward nonlinear least
-squares, and also the reasons for success of an extended inversion
-approach.
+This paper examines one of a number of proposed alternatives to least
+squares FWI, based on an extension of the model-to-data mapping.  We
+show exactly how an example of this {\em extended inversion} approach
+leads to solution of a very simple acoustic transmission
+inverse problem {\em via} local gradient-based optimization. Despite
+its simplicity, this problem exhibits several important
+characteristics of all inverse problems in wave propagation in which
+wave velocities are amongst the unknowns. It has the advantage of
+permitting explicit calculations that reveal clearly the failure
+mechanism for FWI, that is, straightforward nonlinear least squares,
+and also the reasons for success of an extended inversion approach.
 
 In the remainder of this introduction, we will briefly describe the conceptual
-framework, the inverse problem, and sketch our main results. We leave out
-some technical details that are supplied in the next section,
-containing precise statements of the results.
-
-The idealized physical setting is the propagation of wave through a
-homogeneous acoustic fluid occuyping $\bR^3$ \cite[]{Frie:58}. The
-fluid's salient characteristic for the present discussion is the
+framework of this inverse problem, and sketch our main results. We leave out
+some technical details that are supplied in the next section, where we give
+precise statements of our results. In particular, theorems referenced
+in the rest of this section are to be found, stated in full, in the
+next section.
+
+The physical setting for our discussion is the propagation of waves through a
+homogeneous acoustic fluid occupying $\bR^3$ \cite[]{Frie:58}. The
+fluid's salient characteristic for present purposes is the
 slowness, or reciprocal wave velocity, $m > 0$. A source of acoustic
@@
 Ignoring the amplitude factor $1/(4\pi r)$, this map implements a $m$-dependent time
-shift. This time shift operator is the basis of many descriptions of
+shift. This time shift operator is the basis of many illustrations of
 the cycle-skipping phenomenon (for example, \cite{VirieuxOperto:09},
@@
 
-Evidently if the wavelet $w$ is square-integrable, so is the predicted
-recorded pressure - its $L^2$ norm is the cumulative power trasferred
+If the wavelet $w$ is square-integrable, so is the predicted
+recorded pressure $F[m]w$ - its $L^2$ norm is the cumulative power trasferred
 from the fluid to the receiver \cite[]{SantosaSymes:00}. Thus this relation defines a map
@@
 would be: given recorded data
-$d \in L^2([t_{\rm min},t_{\rm max}])$, find $m \in \bR$ and $w \in L^2(\bR)$ so that
+$d \in L^2([t_{\rm min},t_{\rm max}])$, find $m \in \bR^+$ and $w \in L^2(\bR)$ so that
 $F[m]w \approx d$. Clearly this problem statement is not interesting as it
-stands, since it is always possible to solve it with no error and nonuniquely,
-and it does not constrain the slowness $m$ at all. To be useful, the statement must
-be augmented with a constraint on the model $(m,w)$. We choose 
-$\lambda > 0$ and search for $w$ with $\mbox{supp }w \subset
-[-\lambda,\lambda]$. The quantity $\lambda$ is to be considered
-``small'': as we shall show, it is related to a mean wavelength of
-$w$.
+stands, since $F[m]$ is surjective for every $m \in \bR^+$: fitting
+the data does not constrain the slowness $m$ at all.
+
+To be useful, the problem statement must be augmented with a
+constraint on the model $(m,w)$.  One natural constraint is to assume
+that the support of $w$ is limited: choose $\lambda > 0$ and search
+for $w$ with $\mbox{supp }w \subset [-\lambda,\lambda]$. The quantity
+$\lambda$ is to be considered ``small''. Under some circumstances, it
+is related to a mean wavelength of $w$ (Appendix B).
+
+An assumption of small support may be justified as
+follows. In practical seismic data processing, actual sources may act
+over considerable time intervals. An example from exploration
+seismology is the airgun, the most commonly used
+marine source: an airgun (or array of airguns, the usual apparatus)
+produces an oscillating pressure pulse that dies away
+slowly. It is commonplace to estimate this pulse then deconvolve it
+from the data by safeguarded Fourier division or other means. This
+so-called signature deconvolution results in modified data
+corresponding to a source equal to signature deconvolution of the
+pulse estimate itself. The deconvolved source pulse approximates an
+impulse ($\delta(t)$): it cannot have point support, due to the finite
+frequency nature of seismic data, hence the bandlimited nature of the
+pulse estimate, but its support is ultimately much smaller than that
+of the original pulse estimate.
 
@@
   \label{eqn:redms}
-  (m,w) \mapsto \lerr[m,w,d]=\frac{1}{2}\|\lF[m]w-d\|^2,
+  \lerr[m,w;d]=\frac{1}{2}\|\lF[m]w-d\|^2/\|d\|^2,
 \end{equation}
@@
 a solution if it lies in the domain of $\lF$ and
-$\lerr[m,w,d] \le 1/2 \epsilon^2\|d\|^2$ (Problem Statement
+$\|F[m]w-d\| \le \epsilon\|d\|$ (Problem Statement
 \ref{eqn:probstat0} below).  In fact, we show
@@
 at which its value is much greater than its global minimum
-value. Therefore if $\epsilon$ is near the global minimum value, then
-local minimization of $\lerr[\cdot,\cdot,d]$ can fail to find a
+value. Therefore if $1/2\epsilon^2$ is near the global minimum value
+of $\lerr[\cdot,cdot;d]$, then
+local minimization of $\lerr[\cdot,\cdot;d]$ can fail to find a
 solution of the inverse problem. This phenomenon is what is meant by
@@
 \begin{eqnarray}
-  \label{eqn:pen}
-  e[m,w;d] & = & \frac{1}{2}\|F[m]w-d\|^2,\\
-  g[w] & = & \frac{1}{2}\|A\|^2, \\
+  r[m,w;d] & = & \|F[m]w-d\|/\|d\|, \label{eqn:rmserr}\\
+  e[m,w;d] & = & r[m,w;d]^2, \label{eqn:mserr}\\
+  g[w] & = & \frac{1}{2}\|Aw\|^2/\|d\|^2, \label{eqn:pen} \\
   \Ja[m, w;d] &=& e[m,w;d] + \alpha^2 g[w] \nonumber\\
-  &=& \frac{1}{2}(\|F[m]w-d\|^2 + \alpha^2 \|Aw\|^2). 
-\label{eqn:pen}
+  &=& \frac{1}{2}(\|F[m]w-d\|^2 + \alpha^2 \|Aw\|^2)/\|d\|^2.
+\label{eqn:penerr}
 \end{eqnarray}
-The only obvious requirement on the penalty operator $A$ is that it be
-nonzero for large $|t|$, so that minimizers of $\Ja$ should be
+
+The normalizations by the data norm $\|d\|$ in the definitions
+\ref{eqn:rmserr}, \ref{eqn:mserr}, \ref{eqn:pen}, and \ref{eqn:penerr} make the values
+of $r$, $e$, $\alpha^2 g$, and $\Ja$ nondimensional.
+
+The only obvious requirement on the penalty operator $A$ is that it
+penalize 
+nonzero values of $w$ for large $|t|$, so that minimizers $(m,w)$ of
+$\Ja$ should have $w$
 concentrated near $t=0$. Thus the penalty is a soft stand-in for the
 support constraint. We
-show that a good choice for $A$ is given by $(Aw)(t)=tw(t)$, and the
+show that a good choice for $A$ is given (essentially) by
+$(Aw)(t)=tw(t)$ (Proposition \ref{thm:rampgood}), and the
 remainder of the discussion in this section makes this assumption. The
@@
 Assume that $\mu>0$ and $d = F[m_*]w_* + n$, with $\mbox{supp }w_*
-\subset [-\mu,\mu]$, and denote by $\eta = \|n\|/\|F[m_*w_*]$
+\subset [-\mu,\mu]$, and denote by $\eta = \|n\|/\|F[m_*,w_*]\|$
 the noise-to-signal ratio. We show that any local minimizer $(m,w)$
@@
     \label{eqn:mbnd}
-    |m-m_*| \le (1+f(\eta))\frac{\lambda}{r},
+    |m-m_*| \le (1+f(\eta))\frac{\mu}{r},
   \end{equation}
@@
     \label{eqn:success}
-    \lerr[m, {\bf 1}_{[-\lambda,\lambda]}w,d] \le \frac{1}{2}\epsilon^2,
+    \lerr[m, {\bf 1}_{[-\lambda,\lambda]}w;d] \le \frac{1}{2}\epsilon^2,
   \end{equation}
@@
 bounds are shown to be sharp, and the constraints cannot be
-significantly weakened. For example, the support radius $\lambda$ in
+significantly weakened. For example, in conclusion \ref{eqn:solcond}, the support radius $\lambda$ in
 the projection onto $L^2([-\lambda,\lambda])$ must in general be at least twice the
@@
 statements of our main results, filling in the
-few details not discussed in the preceding sketch. Then we provide an
+few details not discussed in the preceding sketch. Then we provide a
 discussion of the potential theoretical and practical import of these
-results, including relations with prior work, some numerical
+results, including relation with prior work, some numerical
 illustrations, and directions for further research. Following this
@@
 the algebra of bounded linear operators from the Hilbert space $X$ to 
-the Hilbert space $Y$, and its subalgebra of invertible operators.
+the Hilbert space $Y$, and its subalgebra of invertible
+operators. All of the operators appearing in this discussion are
+members of ${\cal B}(X,Y)$ for some choice of Hilbert spaces $X$ and
+$Y$, and superscript $T$ denotes the transpose or adjoint of a such an
+operator in the sense specified by its domain and range Hilbert structure.
 
@@
 the slowness (reciprocal velocity) at spatial position $\bx$, $w(t)$
-the time dependence of the point source (``wavelet'') at location
+the time dependence of the point source intensity (``wavelet'' or
+``source wavelet'') at location
 $\bx=\bx_s$. Then the (excess) pressure field $p(\bx,t)$ obeys a
-scalar wave equation:
+scalar wave equation \cite[]{Frie:58}:
 \begin{eqnarray}
@@
 \end{eqnarray}
-Suppose that a single trace is recorded, at distance $r>0$ from the
-source position $\bx_s$. The dominant information in a single
+Suppose that the pressure at a receiver position $\bx=\bx_r$ is recorded, at distance $r>0$ from the
+source position $\bx_s$. We shall refer to the resulting function of
+time as a {\em trace}, from common seismic usage. The dominant information in a single
 trace is the transient signal time of arrival, constraining only the mean slowness in the
@@
 
-The receiver location $\bx_r$ lies at distance $r$ from the source
-location $\bx_s$, that is, $|\bx_r-\bx_s|=r$. The predicted signal at
+Define $r=|\bx_r-\bx_s| $ to be the source-receiver distance
+(``offset''). The predicted signal at
 $p(\bx_r,t)$ depends nonlinearly on the slowness $m$ and linearly on the
@@
 action of a $m$-dependent linear operator $F[m]$ on $w$, as described
-in \ref{eqn:mod}.
-%\begin{equation}
-%\label{eqn:mod}
-%F[m]w(t)= p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right).
-%\end{equation}
+in \ref{eqn:mod}, which we repeat here for convenience:
+\begin{equation}
+\label{eqn:mod1}
+F[m]w(t)= p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right).
+\end{equation}
 
 The slowness $m$ must be positive, as follows from basic acoustics,
-and in fact reside in a range $M=(m_{\rm min}, m_{\rm max})$ characteristic of the
+and in fact reside in a range characteristic of the
 material model: for crustal rock, a reasonable choice would be
@@
 $[t_{\rm min},t_{\rm max}]$ the recording interval for the
-hypothetical single trace data. Then natural choices for domain and
-range of $F[m]$ are $W = L^2(\bR)$ and $D=L^2([t_{\rm min},t_{\rm
-  max}])$ respectively (that is, interpret the definition \ref{eqn:mod}
-as holding for $t \in [t_{\rm min},t_{\rm  max}]$). So
+hypothetical single trace data.
+
+Natural choices for domain and
+range of $F$ are
+\begin{itemize}
+\item $M=(m_{\rm min}, m_{\rm max}),\,0 < m_{\rm min} \le m_{\rm
+    max}$;
+\item $W = L^2(\bR)$;
+\item $D=L^2([t_{\rm min},t_{\rm max}]),\, t_{\rm min}<t_{\rm max}$;
+\item $F: M \times W \rightarrow D$ as specified in \ref{eqn:mod}.
+\end{itemize}
+It is immediately evident from these choices and from the definition
+\ref{eqn:mod} that
 \begin{equation}
@@
 
-In computational practice, $W$ will have to be replaced by a
+\noindent{\bf Remark:} In computational practice, $W$ will have to be replaced by a
 finite-dimensional subspace of $L^2(\bR)$. Many such choices will
@@
 \subsection{An inverse problem}
-
-Part of the definition of an inverse problem based on the model
-definition \ref{eqn:mod} is that observed data $d \in D$ be fit: that
-is, $m, w$ must be found so that $F[m]w \approx d$. However this
-formulation does not provide enough information to determine the model
-$(m,w) \in M \times W$: indeed, the definition
-\ref{eqn:mod} shows that $F[m]$ is surjective for any $m \in M$, so
-the data can always be fit precisely by appropriate choice of $w$, and
-$m$ is completely
-unconstrained.  In order that data fit constrain $m$, it is necessary
-to further constrain $w$ {\em a priori} - that is, to add information
-beyond the trace data $d$.
-
-One natural constraint is to assume that the support of $w$ is
-limited. An assumption of small support may be justified as
-follows. In practical seismic data processing, actual sources may act
-over considerable time intervals - for example, the most commonly used
-marine source (airgun) produces an oscillating pulse that dies away
-slowly. It is commonplace to estimate this pulse then deconvolve it
-from the data by safeguarded Fourier division or other means. This
-so-called signature deconvolution results in modified data
-corresponding to a source equal to signature deconvolution of the
-pulse estimate itself. The deconvolved source pulse approximates an
-impulse ($\delta(t)$): it cannot have point support, due to the finite
-frequency nature of seismic data, hence the bandlimited nature of the
-pulse estimate, but its support is ultimately much smaller than that
-of the original pulse estimate.
-
-Therefore, supplement the observed data $d \in D$ with a presumed 
-source support radius $\lambda$. Define 
-$\lW=\{w \in W: \mbox{ supp }w \subset [-\lambda,\lambda]\}$, and 
-denote by $\lF$ the restriction of $F$, defined in equation 
-\ref{eqn:mod}, to $M \times \lW$.
-
-Note that for $w \in \lW$, $m \in M$,
-$\mbox{supp }F[m]w \subset [ m_{\rm min}r-\lambda,
-m_{\rm max}r+\lambda] \cap
-[t_{\rm min},t_{\rm max}]$. Also, $\lF[m]$ is coercive for all $m \in
-M$ only if
-$[ m_{\rm min}r-\lambda,m_{\rm max}r+\lambda] \subset
-[t_{\rm min},t_{\rm max}]$.
-
-We will assume that $\lambda_{\rm max}
->0$ is chosen so that $M, r,\lambda_{\rm max}$ and $[t_{\rm min},t_{\rm max}]$ satisfy
-\begin{equation}
-  \label{eqn:fullrec}
-  [ m_{\rm min}r-\lambda_{\rm max}, m_{\rm max}r+\lambda_{\rm max}]
-  \subset [t_{\rm min},t_{\rm max}].
-\end{equation}
-Then for 
-$m \in M$, $0 <\lambda \le \lambda_{\rm max}$, $\lF[m] \in {\cal B}(\lW,D)$ is coercive.
-
-Note that condition \ref{eqn:fullrec} implies
-\begin{eqnarray}
-  \lambda_{\rm max} & \le & t_{\rm max}-m_{\rm max}r,\nonumber \\
-  t_{\rm min}-m_{\rm min}r & \le & -\lambda_{\rm max} \label{eqn:lminc}
-\end{eqnarray}
-hence for any $m \in M$,
-\begin{equation}
-  [-\lambda_{\rm max}, -\lambda_{\rm max}] \subset[ t_{\rm min}-mr ,
-  t_{\rm max}-mr].
-  \label{eqn:zeroinc}
-\end{equation}
+As mentioned earlier, $F[m]$ is surjective for every
+$m \in M$. Since all possible data lie in the range of $F[m]$ for any
+$m \in M$, some restriction of the domain
+of $F$ is necessary in order that fitting the data constrain
+$m$. A more precise version of the constraint proposed in the
+introduction involves a selection of a maxium support radius
+$\lambda_{\rm max} >0$. Then define for $\lambda \in (0,\lambda_{\rm max}]$:
+\begin{itemize}
+\item $\lW = \{w \in W:
+  \mbox{supp }w \subset [-\lambda,\lambda]\}$;
+\item $\lF = F|_{M \times \lW}$.
+\end{itemize}
 
@@
 
-\noindent {\bf Problem Statement:} given data $d \in D$, relative 
-error level $\epsilon \in [0,1)$, and support radius $\lambda \in (0, 
-\lambda_{\rm max}]$, find 
-$(m,w) \in M \times \lW$ for which 
-\begin{equation}
-  \label{eqn:probstat0}
-  \|\lF[m]w-d\| \le \epsilon\|d\|. 
-\end{equation}
+\begin{quote}
+\noindent {\bf Problem Statement:}
+  given data $d \in D$, relative error level $\epsilon \in
+  [0,1)$, and support radius $\lambda \in (0, \lambda_{\rm
+    max}]$, find $(m,w) \in M \times \lW$ for which 
+\begin{equation}
+  \label{eqn:probstat0}  \|\lF[m]w-d\| \le \epsilon\|d\|,
+\end{equation}
+or equivalently
+\begin{equation}
+  \label{eqn:probstat1}
+  \lerr[m,w;d] \le \frac{1}{2}\epsilon^2,
+\end{equation}
+with $\lerr$ defined in equation \ref{eqn:redms}.
+\end{quote}
 
@@
 Such data $d$ is {\em noise-free}, in the range of the map $\lF$. For
-such data a solution of the Problem Statement
+such data a solution of the Problem Statement \ref{eqn:probstat0}
 exists with arbitrarily small $\epsilon>0$.
 
-\subsection{Soft constraints and penalty formulation}
-
-As mentioned in the introduction, a straightforward nonlinear least
-squares formulation of the problem stated in \ref{eqn:probstat0}
-exhibits local minimizers with large fit error, and these will be
-found by any local optimization algorithm unless it is initialized at
-a quite accurate initial model. The first result stated in the next
-section establishes this assertion for the single trace transmission
-problem, however it appears to be valid for field scale problems, for
-essentially the same reasons, as widely reported in the literature on
-seismic inversion.
-
-One (of many) alternative optimization formulation of this and similar
-problems involves introducing an extension of the model-to-data
-operator, and compensating for the physically spurious degrees of
-freedom so introduced by adding a penalty term controlling them to the
-objective.  The application of this idea presented here extends $\lF$
-by dropping the support constraint on its second argument, that is,
-using the expression \ref{eqn:mod} to define a map $F: M \times W
-\rightarrow D$. The objective begins with an extended version of the
-mean-square error,
-\begin{equation}
-  \label{eqn:mserr}
-  e[m,w;d]=\frac{1}{2}\|F[m]w-d\|^2.
-\end{equation}
-We will choose a penalty to control the extra degrees of freedom, that
-is, $w$ outside of $[-\lambda,\lambda]$. This  penalty is quadratic, constrains only $w\in W$ (as it is a soft
-version of the condition $w \in \lW$), and is
-$m-$independent. Therefore the penalty function takes the form
-\begin{equation}
-\label{eqn:pen}
-p[w] = \frac{1}{2}\|Aw\|_Y^2 
-\end{equation}
-where $A \in {\cal B}(W,Y)$, $Y$ being another suitable Hilbert
-space.
-
-The penalty approach asks that a linear combination of $e,p$,
-depending on a relative weight $\alpha$:
-\begin{equation}
-\Ja[m, w;d] = e[m,w;d] + \alpha^2 p[w] = \frac{1}{2}(\|F[m]w-d\|^2 + \alpha^2 \|Aw\|^2)
-\label{eqn:pen}
-\end{equation}
-be minimized over $(m,w)\in M \times W$.
-
-The choice of penalty weight $\alpha$ has a profound influence on the
-character of this optimization problem. For the moment, we will mandate
-only that $\alpha \ge 0$. A complete description of an algorithm using
-the penalty function \ref{eqn:pen} to solve the inverse problem
-\ref{eqn:probstat0} must include a selection principle for $\alpha$. We
-describe one such principle in Appendix A.
-
-The main theoretical device used in the proofs of our main results on
-extended inversion is the reduction of $\Ja$ to a function $\tJa$ of
-$m$ alone, by minimization over $w$. This reduction is the foundation
-of the {\em Variable Projection Method} (VPM)
-\cite[]{GolubPereyra:03} for solution of partly linear (``separable'')
-least squares problems. We use the VPM in our numerical illustrations
-of the theory, as described in Discussion section and Appendix A.
-
-\begin{theorem}
-\label{thm:red}
-Assume that the Hessian operator of
-$\Ja[m,\cdot;d]$ is invertible:
-\begin{equation}
-\label{eqn:norminv}
-F[m]^TF[m] +\alpha^2 A^TA \in {\cal I}(W,W).
-\end{equation}
-(The superscript $T$ denotes the adjoint with respect to the inner products 
-in $W=L^2(\bR)$ and $D=L^2([t_{\rm min},t_{\rm max}])$.)
-Then the unique stationary point $\aw[m;d] \in W$ of
-$\Ja[m,\cdot;d]$ is the solution of the {\em normal equation}:
-\begin{equation}
-  \label{eqn:norm}
-  (F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td,
-\end{equation}
-Define the reduced or variable projection objective by
-\begin{equation}
-  \label{eqn:redexp}
-  \tJa[m;d] =\Ja[m,w_{\alpha}[m;d];d] = \min_w \Ja[m,w;d]
-\end{equation}
-Then $(m,w)$ is a local minimizer of $\Ja[\cdot,\cdot;d]$ if and only
-if $m$ is a local minimizer of $\tJa[\cdot;d]$ and $w=\aw[m;d]$.
-\end{theorem}
-\noindent{\bf Remark:} If $\Ja[\cdot,\cdot;d]$ and $\tJa[\cdot;d]$ are differentiable, then
-  ``local minimizer'' in the conclusion of the preceding theorem can
-  be replaced by ``stationary point''. However, for the problem
-  addressed in this paper, $\Ja[\cdot,\cdot;d]$ {\em is not}
-  differentiable without added smoothness constraints on $w$, whereas
-  $\tJa[\cdot;d]$ {\em is} differentiable for proper choice of penalty
-  operator $A$, as will be shown
-  below. This phenomenon is replicated for a wide variety of extended
-  approaches for inverse problems in wave propagation. We will have
-  more to say about it in the Discussion section.
-
 \subsection{Main Results}
+While $F[m]$ is surjective for every $m \in M$, as noted above, it is
+very far from injective. On the other hand, under a constraint that
+will be assumed throughout, $\lF[m]$ is injective (in fact, $4 \pi r
+\lF[m]$ is an isometry):
+\begin{proposition}
+  \label{thm:fullrec}
+  Suppose that 
+  \begin{equation}
+    \label{eqn:fullrec}
+    [ m_{\rm min}r-\lambda_{\rm max}, m_{\rm max}r+\lambda_{\rm max}]
+    \subset [t_{\rm min},t_{\rm max}].
+  \end{equation}
+  Then $\lF[m]\in {\cal B}(\lW,D)$ is coercive for every $m \in M, \lambda \in
+  (0,\lambda_{\rm max}]$.
+\end{proposition}
+
 The first main result establishes the existence of large (100 \%)
@@
   \label{eqn:isovpm}
- \min_w \lerr[m,w;d]=\lerr[m,0,d]=\frac{1}{2}\|d\|^2,
-\end{equation}
-and any such $(m,0)$ is a local minimizer of $e$.
+ \min_w \lerr[m,w;d]=\lerr[m,0;d] = \frac{1}{2},
+\end{equation}
+and any such $(m,0)$ is a local minimizer of $\lerr$ with relative RMS
+error = 1.0.
 \end{theorem}
 
-Turning to extended inversion, we examine two choices of penalty
-operators. For each choice, we ask first whether local minimizers 
-of the resulting VPM
-objective $\tJa[\cdot;d]$ occur far from a slowness $m_*$. Both choices for $A$ considered here are scalar 
-multiplication operators defined by a choice of positive multiplier $a \in L^{\infty}(\bR)$:
+The main theoretical device used in the proofs of our main results on 
+extended inversion is the reduction of the penalty objective $\Ja$
+(equation \ref{eqn:penerr}) to a function $\tJa$ of 
+$m$ alone, by minimization over $w$: 
+\begin{equation}
+  \label{eqn:redexp}
+  \tJa[m;d] = \inf_w \Ja[m,w;d]
+\end{equation}
+This reduction is the foundation 
+of the {\em Variable Projection Method} (VPM) 
+\cite[]{GolubPereyra:03} for solution of partly linear (``separable'') 
+least squares problems.
+
+A minimizer $w$ on the right-hand side of definition
+\ref{eqn:redexp} must solve the {\em normal equation}
+\begin{equation}
+  \label{eqn:norm}
+  (F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td, 
+\end{equation}
+
+The choices for the penalty operator $A$ considered here are scalar 
+multiplication operators on $W$ defined by a choice of multiplier $a \in L^{\infty}(\bR)$:
 \begin{equation}
   \label{eqn:annmult}
-  A w(t)= a(t)w(t), \, t\in \bR: \,a(t) > 0 \mbox{ for }t \ne 0.
-\end{equation}
+  A w(t)= a(t)w(t), \, t\in \bR.
+\end{equation}
+
+\begin{theorem}
+  \label{thm:norminv}
+  Assume the condition \ref{eqn:fullrec}, $C>0$, and suppose that $A$ is
+  given by equation \ref{eqn:annmult} for $a \in L^{\infty}(\bR)$
+  satisfying
+  \begin{equation}
+    \label{eqn:abnd} 
+    a \ge 0; \, a(t) \ge C\mbox{ for }|t| \ge \lambda_{\rm max}.
+  \end{equation}
+  Then \begin{itemize}
+  \item[1. ]the reduced objective $\tJa$ is given by
+    \begin{equation}
+      \label{eqn:redexp1}
+      \tJa[m;d] = \Ja[m,\aw[m;d];d],
+    \end{equation}
+    in which $\aw[m;d] \in W$ is the unique solution of the normal
+    equation \ref{eqn:norm}.
+  \item[2. ]The following are equivalent:
+    \begin{itemize}
+    \item[i. ]$(m,w) \in M \times W$ is a local minimizer of
+      $\Ja[\cdot,\cdot;d]$, and
+    \item[ii. ]$m$ is a local minimizer of $\tJa[\cdot;d]$ and
+      $w=\aw[m;d]$.
+    \end{itemize}
+  \end{itemize}
+\end{theorem}
+
+\noindent{\bf Remark:} If $\Ja[\cdot,\cdot;d]$ and $\tJa[\cdot;d]$
+were differentiable, then ``local minimizer'' in the conclusion of the
+preceding theorem could be replaced by ``stationary point''. However,
+for the problem addressed in this paper, $\Ja[\cdot,\cdot;d]$ {\em is
+  not} differentiable without added smoothness constraints on $w$,
+whereas $\tJa[\cdot;d]$ {\em is} differentiable for proper choice of
+penalty operator $A$:
+
+\begin{theorem}
+  \label{thm:diffobj}
+  Suppose that in addition to the hypotheses of Theorem
+  \ref{thm:norminv}, $a \in W^{1,\infty}_{\rm loc}(\bR)$, then $\tJa[\cdot;d]
+  \in C^1(M)$.
+\end{theorem}
+
+This phenomenon - nondifferentiable objective, differentiable reduced
+objective - is replicated for a wide variety of extended
+approaches for inverse problems in wave propagation. We will have
+more to say about it in the Discussion section.
+
+We examine two choices for $A$. For each choice, we ask first whether local minimizers 
+of the resulting VPM objective $\tJa[\cdot;d]$ occur far from a slowness $m_*$.
 
@@
   \[
-    \tJa[m;d] = \frac{\alpha^2}{2(1+(4 \pi r)^2 \alpha^2)}\|w_*\|^2.
+    \tJa[m;d] = \frac{(4\pi r \alpha)^2}{2(1+(4 \pi r \alpha)^2)}.
   \]
@@
 \ref{eqn:probstat0} is the same as the set of global minimizers of
-$\tJa$ in this case. However, for this choice of annihilator,
-$\tJa$ exhibits the same feature as the mean square error $e$, namely
-a continuum of stationary points at any distance from the global
+$\Ja$ in this case. However, for this choice of annihilator,
+$t\Ja$ exhibits the same feature as the mean square error $e$, namely
+a continuum of local minimizers at any distance from the global
 minimizer $m_*$ greater than a multiple of $\lambda$. Therefore the
@@
 \end{equation}
-The cutoff $\tau$ will be chosen large enough to be effectively inactive: 
+Note that $a \ge 0$ and $a \in L^{\infty}(\bR) \cap W^{1,1}_{\rm
+  loc}(\bR)$. The cutoff $\tau$ will be chosen large enough to be effectively inactive: 
 specifically, hindsight suggests 
@@
   multiplier $a$ given in equation \ref{eqn:ann}, and support radius
-  $\mu$ of the ``noise-free'' wavelet $w_*$, $\tJa$ has no local
-  minima further than $(1+O(\eta))\mu/r)$ from the slowness used to
+  $\mu$ of the ``noise-free'' wavelet $w_*$, $\Ja$ has no local
+  minima with slownesses further than $(1+O(\eta))\mu/r)$ from the slowness used to
   generate the data}.
 
-\noindent {\bf Remark.} The estimate $|m-m_*|r<\mu(1+O(\eta))$ for stationary
-points $m$ of $\tJa$ is sharp: it is possible to choose $w_* \in W_{\mu}$
+\noindent {\bf Remark.} The estimate $|m-m_*|r<\mu(1+O(\eta))$ for
+local minima of $\Ja$ is sharp: it is possible to choose $w_* \in W_{\mu}$
 so that $\mu - |m-m_*|r$ is as small as you like. In particular,
 the ``exact'' or ``true'' slowness $m_*$ is not necessarily the only
-stationary point, or even a stationary point at all, and in particular
-is not (necessarily) a global minimizer of $\tJa$.
+slowness component of a local minimizer, or even the slowness
+component of any local minimizer, and in particular
+is not (necessarily) the slowness component of a global minimizer of $\Ja$.
 
@@
 the predicted data for the same wavelet $w_*$ with a substantially different
-slowness $m_{\flat}$, that is, $n=F[m_{\flat}]w_*$, then by symmetry
-the average slowness $0.5(m_*+m_{\flat})$ is a stationary point of
-$\tJa[\cdot,d_*+n]$. The difference with $m_*$ is not constrained at
-all by the support of $w_*$. So for this example with 100\% noise, no
+slowness $m_{\flat}$, that is, $n=F[m_{\flat}]w_*$, then a simple
+symmetry argument shows that if there is a local minimizer of $\Ja[\cdot,\cdot;d_*+n]$. with
+slowness near $m_*$,
+there must also be a minimizer with slowness near $m_{\flat}$.
+so that the difference with $m_*$ is not constrained at
+all by the assumed support radius of $w_*$. So for this example with 100\% noise, no
 bound of the type given by conclusion 2 could possibly hold. We will
@@
 Unless the data is noise-free, there is no reason to suppose that the
-estimated wavelet $\aw[m;d]$ (Theorem \ref{thm:red}) will lie in
+estimated wavelet $\aw[m;d]$ (Theorem \ref{thm:norminv}) will lie in
 $\lW$, unless the support of the noise $n$ is not restricted. In order
@@
     \label{eqn:ipnoiseeps}
-    \epsilon \ge (4 \pi r \alpha\lambda)^2(1+\eta^2)^{1/2} 
+    \epsilon \ge (4 \pi r \alpha\lambda)^2+\eta. 
   \end{equation}    
@@
 problem for any $\epsilon > 0$ provided that $\alpha$ is chosen
-sufficiently small ($O(\sqrt{epsilon})$).
+sufficiently small ($O(\sqrt{\epsilon})$).
 
@@
 
-\section{Full Waveform Inversion: hard constraint}
-A natural formulation of the Problem Statement \ref{eqn:probstat0} is
-via least squares: that is, given $d \in D$, $\lambda >0$, find
-$(m,w) \in M \times \lW$ to minimize
-$\lerr: M \times W \rightarrow \bR$, defined in display
-\ref{eqn:redms}.  A pair $(m,w) \in M\times \lW$ satisfies the
-condition \ref{eqn:probstat0} iff $\lerr[m,w;d] \le \epsilon^2/2$, so a
-solution exists iff a global minimizer of $\lerr$ satisfies this
-inequality.
-
-In this discussion, I will presume that $d$ is 
-noise-free, as defined at the end of the last section (equation
-\ref{eqn:defdatanonoise}). Thus {\em a priori} a global minimizer of
-$e$ is a solution of the Problem Statement for any
-$\epsilon > 0$. The only
-question is how to find such a global minimizer. As mentioned in the
-introduction, the simple model problem studied here is a proxy for
-large-scale data fitting problems for which the only computationally
-feastible approach to minimization of analogs of $e$
-is some form of descent-based iterative local optimization. Any such
-algorithm requires the specification of an initial iterate
-$(m_0,w_0)$. Therefore the key question is: what
-constraints must be imposed on $(m_0,w_0)$ in order that the resulting
-sequence of iterates convertes to a global minimizer?
-
-To answer this question, note that 
+The inverse problem stated at the beginning of this paper concerns
+three types of
+objects: a set of vectors of measured or observed parameters (``the
+data''); a set of vectors of non-observed parameters (``the model''); and a
+modeling or prediction operator, which accepts model vectors as
+input and produces predicted data vectors as output.
+
+The ``inverse problem'' is really a family of questions, parametrized
+by a scalar tolerance. The question implicitly posed in the statement
+\ref{eqn:probstat0} is: given a tolerance and an observed data vector,
+is there a model vector for which the predicted data vector is within
+the tolerance of the observed data vector, in the sense of a distance
+function on the collection of all possible data vectors.
+
+This deterministic definition of inverse problems has been presented by
+many authors, for instance \cite{ddjackson:72,Parker:94}. Most variants 
+presume that the data and model vectors form Hilbert spaces, or
+possibly subsets of Hilbert spaces, and that the distance between
+predicted and observed data vectors is the norm of the difference. Our
+approach fits squarely in this tradition.
+
+In any concrete instance, a ``yes'' answer can only be justified by
+exhibition of a model vector satisfying the fit criterion, hence the
+form of the statement \ref{eqn:probstat0}. The extension approach
+discussed in this paper relies on use of an extended modeling or
+prediction operator, satisfying two criteria:
+\begin{itemize}
+\item the extended operator is an extension of the modeling
+  operator, that is, its domain is a superset of the model vector
+  space and, when fed a model vector, it replicates the action of
+  the modeling operator;
+\item the extended operator is {\em surjective}, that is, all data
+  vectors can be fit without error.
+\end{itemize}
+
+\section{Derivations}
+\noindent{\bf Remark:} A useful consequence of the condition 
+\ref{eqn:fullrec}: for every $m \in M$, 
+\begin{equation}
+  [-\lambda_{\rm max}, -\lambda_{\rm max}] \subset[ t_{\rm min}-mr , 
+  t_{\rm max}-mr]. 
+  \label{eqn:zeroinc}
+\end{equation}
+
+\subsection{Proof of Theorem \ref{thm:fwi}}
+%A natural formulation of the Problem Statement \ref{eqn:probstat0} is
+%via least squares: that is, given $d \in D$, $\lambda >0$, find
+%$(m,w) \in M \times \lW$ to minimize
+%$\lerr: M \times W \rightarrow \bR$, defined in display
+%\ref{eqn:redms}.  A pair $(m,w) \in M\times \lW$ satisfies the
+%condition \ref{eqn:probstat0} iff $\lerr[m,w;d] \le \epsilon^2/2$, so a
+%solution exists iff a global minimizer of $\lerr$ satisfies this
+%inequality.
+
+%In this discussion, I will presume that $d$ is 
+%noise-free, as defined at the end of the last section (equation
+%\ref{eqn:defdatanonoise}). Thus {\em a priori} a global minimizer of
+%$e$ is a solution of the Problem Statement for any
+%$\epsilon > 0$. The only
+%question is how to find such a global minimizer. As mentioned in the
+%introduction, the simple model problem studied here is a proxy for
+%large-scale data fitting problems for which the only computationally
+%feastible approach to minimization of analogs of $e$
+%is some form of descent-based iterative local optimization. Any such
+%algorithm requires the specification of an initial iterate
+%$(m_0,w_0)$. Therefore the key question is: what
+%constraints must be imposed on $(m_0,w_0)$ in order that the resulting
+%sequence of iterates convertes to a global minimizer?
+
+\begin{proof} of Theorem \ref{thm:fwi}: From the definition \ref{eqn:mod},
 \[
@@
 target slowness $m_*$, the means square error (FWI objective) $e$ is independent of
-$m$, and its minimum over $w$ is attained for $w=0$:
-\begin{theorem}
-  \label{thm:fwi}
-  Suppose that $0 <\lambda\le \lambda_{\rm max}$,  $m_* \in M, w_*
-  \in \lW, d=\lF[m_*]w_*$ is noise-free data per definition \ref{eqn:defdatanonoise},
-  Under assumption \ref{eqn:fullrec}, for any $m \in M$ with $r|m-m_*|>2\lambda$,
-\begin{equation}
-  \label{eqn:isovpm}
- \min_w e[m,w;d]=e[m,0,d]=\frac{1}{32\pi^2 r^2}\|w_*\|^2,
-\end{equation}
-and any such $(m,0)$ is a stationary point of $e$.
-\end{theorem}
-
-Therefore local minimizers of $e$ abound, as far as you like from the
+$m$, and its minimum over $w$ is attained for $w=0$
+\end{proof}
+%\begin{theorem}
+%  \label{thm:fwi}
+%  Suppose that $0 <\lambda\le \lambda_{\rm max}$,  $m_* \in M, w_*
+%  \in \lW, d=\lF[m_*]w_*$ is noise-free data per definition \ref{eqn:defdatanonoise},
+%  Under assumption \ref{eqn:fullrec}, for any $m \in M$ with $r|m-m_*|>2\lambda$,
+%\begin{equation}
+%  \label{eqn:isovpm}
+% \min_w e[m,w;d]=e[m,0,d]=\frac{1}{32\pi^2 r^2}\|w_*\|^2,
+%\end{equation}
+%and any such $(m,0)$ is a stationary point of $e$.
+%\end{theorem}
+
+Therefore local minimizers of $\lerr$ abound, as far as you like from the
 global minimizer $(m_*,w_*)$. Local exploration of the FWI objective
@@
 (failure if $m_0$ is in error by ``half a wavelength''), as follows
-from a more refined analysis of the ``cycle-skipping'' local behaviour of $e$ near its
+from a more refined analysis of the ``cycle-skipping'' local behaviour of $\lerr$ near its
 global minimizer.
 
-\section{All stationary points are (near) global minimizers -
-  sometimes}
-It is not immediately apparent why use of a soft constraint should
-produce an optimization problem more amenable to descent methods than
-does the hard-constrained inverse problem (FWI). In this section, I
-will show that the penalty formulation can yield a reduced problem
-with for which any stationary point is close to a global minimizer,
-{\em for some choices of penalty operator, but not for other choices
-  that might appear equally or even more suitable}.
-
+%\begin{theorem}
+%\label{thm:red}
+%$\Ja[m,\cdot;d]$ is invertible:
+%\begin{equation}
+%\label{eqn:norminv}
+%F[m]^TF[m] +\alpha^2 A^TA \in {\cal I}(W,W).
+%\end{equation}
+%(The superscript $T$ denotes the adjoint with respect to the inner products 
+%in $W=L^2(\bR)$ and $D=L^2([t_{\rm min},t_{\rm max}])$.)
+%Then the unique stationary point $\aw[m;d] \in W$ of
+%$\Ja[m,\cdot;d]$ is a global minimizer, and is the solution of the {\em normal equation}:
+%\begin{equation}
+ % \label{eqn:norm}
+  %(F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td, 
+%\end{equation}
+%The reduced or variable projection objective $\tJa$ defined in
+%equation \ref{eqn:redexp} is also given by
+%\begin{equation}
+ % \label{eqn:redexp1}
+ % \tJa[m;d] =\Ja[m,w_{\alpha}[m;d];d] = \min_w \Ja[m,w;d]
+%\end{equation}
+%Then $(m,w)$ is a local minimizer of $\Ja[\cdot,\cdot;d]$ if and only
+%if $m$ is a local minimizer of $\tJa[\cdot;d]$ and $w=\aw[m;d]$.
+%\end{theorem}
+
+%\noindent{\bf Remark:} The choice
+%of penalty weight $\alpha$ has a significant influence on the
+%performance of numerical optimization algorithms in minimizing
+%$\Ja$ and/or $\tJa$, but relatively little influence on the error estimates to
+%follow. For the discussion of these estimates, we assume only that
+%$\alpha>0$. Appendix A contains a description of an algorithm for
+%constructive determination of $\alpha$.
+
+\subsection{Proof of Theorem \ref{thm:norminv}}
 For the single-trace transmission inverse problem, $\tilde{J}_{\alpha}$ is explicitly
@@
 
-\begin{theorem}
-  \label{thm:norminv}
+\begin{proposition}
+  \label{thm:norminvexp}
   Assume the conditions \ref{eqn:mod}, \ref{eqn:fullrec},
   \ref{eqn:annmult}. Also assume that $\lambda \in (0,\lambda_{\rm
-    max}], \alpha > 0,$ and that $a \in L^{\infty}(\bR)$
-  mentioned in condition \ref{eqn:annmult} satisfies
-  \begin{equation}
-    \label{eqn:abnd}
-    |t| > \lambda \Rightarrow a(t) \ge C,
-  \end{equation}
-  in which $C>0$ may depend on $\lambda$.
+    max}], \alpha > 0,$ and that  $C>0$ exists so that $a \in L^{\infty}(\bR)$
+  mentioned in condition \ref{eqn:annmult} satisfies condition
+  \ref{eqn:abnd}.
+%  \begin{equation}
+%   \label{eqn:abnd}
+%    |t| > \lambda \Rightarrow a(t) \ge C,
+%  \end{equation}
+%  in which $C>0$ may depend on $\lambda$.
   Then
@@
     invertible for any $m \in M$, $\alpha > 0$;
-  \item[2. ]for any $d \in D$, the normal equation \ref{eqn:norm} has
-    a unique solution $\aw[m;d]\in W$, and 
+  \item[2. ]the solution $\aw[m;d]\in W$ of the normal equation
+    \ref{eqn:norm} is given by
     \begin{equation}
@@
   \end{itemize}
-\end{theorem}
+\end{proposition}
 
@@
 
-Since the normal operator is invertible, the reduced
-objective $\tJa$ is given by the expression \ref{eqn:objexp}.
-The results of the previous theorem imply those of
-
-\begin{theorem}
+\begin{proof} of Theorem \ref{thm:norminv}: Since the normal operator
+  is invertible (item 1 of Proposition \ref{thm:norminvexp}), the
+  conclusions of this theorem follow from the Lax-Milgram Theorem
+  (\cite{Yosida}, III.7).
+\end{proof}
+
+\subsection{Proof of Theorem \ref{thm:diffobj}}
+
+\begin{proposition}
   \label{thm:epjgen}
-  Assume the hypotheses of Theorem \ref{thm:norminv}. Then
+  Assume the hypotheses of Proposition \ref{thm:norminvexp}. Then
   \begin{equation}
   \label{eqn:residnormgen}
-  e[m,\aw[m,d];d] = \frac{1}{2}\int \,dt\,(4\pi r \alpha a(t-mr))^4(1 +
+  e[m,\aw[m,d];d] = \frac{1}{2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r \alpha a(t-mr))^4(1 +
   (4\pi r \alpha a(t-mr))^2)^{-2}d(t)^2
@@
   \label{eqn:anninormgen}
-  p[m,\aw[m,d];d] = \frac{1}{2}\int \,dt\,(4\pi r a(t-mr))^2(1 +
+  p[m,\aw[m,d];d] = \frac{1}{2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r a(t-mr))^2(1 +
   (4\pi r \alpha a(t-mr))^2)^{-2}d(t)
@@
   \label{eqn:expjgen}
-\tJa[m;d] = \frac{1}{2}\int\,dt\,(4\pi r \alpha a(t-mr))^2(1+(4\pi r \alpha 
+\tJa[m;d] = \frac{1}{2}\int_{t_{\rm min}}^{t_{\rm max}}\,dt\,(4\pi r \alpha a(t-mr))^2(1+(4\pi r \alpha 
 a(t-mr))^2)^{-1}d(t)^2. 
 \end{equation}
-Finally, if $a \in W^{1,1}_{\rm loc}(\bR)$ (distributions with integrable 
-derivatives), then $\tJa[\cdot;d]$ is differentiable, and 
-\begin{equation}
-  \label{eqn:dexpjgen}
-  \frac{d}{dm}\tJa[m;d] = -(4 \pi r \alpha)^2 \int \,dt \, 
-  \left(a\frac{da}{dt}\right)(t-mr)(1+(4\pi r \alpha 
-  a(t-mr))^2)^{-2}d(t)^2. 
-\end{equation}
-\end{theorem}
+\end{proposition}
 
@@
   \[
-    =(4 \pi r\alpha a(t-mr))^2(1 + (4 \pi r\alpha
+    = -(4 \pi r\alpha a(t-mr))^2(1 + (4 \pi r\alpha
     a(t-mr))^2)^{-1}d(t).
@@
   \ref{eqn:redexp}, \ref{eqn:residnormgen}, and
-  \ref{eqn:anninormgen}. Differentiation under the integral sign
-  yields the expression \ref{eqn:dexpjgen} for its derivative with
-  respect to $m$, for smooth $a$, and then by a limiting argument for
-  $a \in W^{1,1}_{\rm loc}(\bR)$.
+  \ref{eqn:anninormgen}.
 \end{proof}
 
+\begin{proof} of Theorem \ref{thm:diffobj}:
+Suppose first that $a \in C^1(\bR)$. Differentiation under the integral sign  
+  yields the expression for its derivative:
+\begin{equation}
+  \label{eqn:dexpjgen}
+  \frac{d}{dm}\tJa[m;d] = -(4 \pi r \alpha)^2 \int_{t_{\rm min}}^{t_{\rm max}} \,dt \, 
+  \left(a\frac{da}{dt}\right)(t-mr)(1+(4\pi r \alpha 
+  a(t-mr))^2)^{-2}d(t)^2. 
+\end{equation}
+For $a \in W^{1,\infty}_{\rm loc}(\bR)$ a limiting argument shows that the
+same expression gives the derivative of $\tJa$.
+\end{proof}
+
+\subsection{Proof of Theorem \ref{thm:boxcarbad}}
 \begin{corollary}
   \label{thm:epjnonoise}
-  Assume the hypotheses of Theorem \ref{thm:norminv}, item 3. Then 
+  Assume the hypotheses of Proposition \ref{thm:norminvexp}, item 3. Then 
 \begin{equation}
@@
 
-\begin{theorem}
+\subsection{Proof of Theorem \ref{thm:mnoiseres}}
+\begin{proposition}
   \label{thm:rampgood}
@@
   \end{equation}
-\end{theorem}
+\end{proposition}
 \begin{proof}
@@
 
-HERE INSERT RESULT: TJA DIFFBL RIGHTARROW A DIFFBLE - relate in
-Discussion to Stolk 03
-
-\section{Extended Inversion}
 The preceding theorem established that a proper choice of annihilator
@@
 
-\subsection{Noise-free Data}
-\begin{theorem}
+The next result answers the first question, in the case of noise-free data:
+\begin{proposition}
   \label{thm:ipnonoisesuf}
@@
   \ge 2\mu$ and $\epsilon \ge (4\pi r \lambda \alpha)^2$.
-\end{theorem}
+\end{proposition}
 
@@
 
-\subsection{The Effect of Noise}
-Suppose that the data trace $d$ takes the form
+We turn now to the second issue identified above, the effect of
+noise. Suppose that the data trace $d$ takes the form
 \begin{equation}
@@
 
-\begin{theorem}
+\begin{proposition}
   \label{thm:mnoise}
-  Assume the hypotheses of Theorem \ref{thm:rampgood}, and that $d$ is
+  Assume the hypotheses of Proposition \ref{thm:rampgood}, and that $d$ is
   given by definition \ref{eqn:defdatanoisy}. Suppose that $m \in M$
@@
   \end{equation}
-  Then $|m-m_*| \le \lambda /r$.
-\end{theorem}
+  Then
+  \begin{equation}
+    \label{eqn:mnoisebndfin}
+    |m-m_*| \le \frac{\lambda}{r}
+  \end{equation}.
+\end{proposition}
 
@@
 \]
-On the other hand, the conclusion \ref{eqn:gradbndnonoise} of Theorem
+On the other hand, the conclusion \ref{eqn:gradbndnonoise} of Proposition
 \ref{thm:rampgood} implies that if also
@@
   that $m$ is a stationary point of $\tJa[\cdot;d]$, $d=d_*+n$. Then
-  \begin{itemize}
-    \item[1. ] the bound \ref{eqn:mnoisebnd} holds only if 
-      noise level $\eta=\|n\|/\|d_*\|$ satisfies
-      \begin{equation}
-        \label{eqn:mnoisecond}
-        \eta < \frac{1+\sqrt{5}}{2} < 1;
-      \end{equation}
-    \item[2. ] if condition \ref{eqn:mnoisecond} holds, then
+  \begin{equation}
+    \label{eqn:mnoisecor}
+    |m-m_*| \le \frac{\mu}{r} + \frac{\eta}{\alpha} \left(\frac{3\sqrt{3}(1+\eta)}{64\pi r^2}(1+(8\pi r \alpha
+      \lambda_{\rm max})^2)^2\right)
+  \end{equation}
+\end{corollary}
+
+\begin{proof} Assume that $\lambda$ is chosen to obtain equality in
+  the condition \ref{eqn:mnoisebnd}, substitute the bound $2
+  \lambda_{\rm max}$ for $\lambda + \mu$ in the denominator, solve for
+  $\lambda$ and substitute in inequality\ref{eqn:mnoisebndfin}.
+\end{proof}
+
+\noindent {\bf Remark:} This is the only error bound in our treatment
+of this problem in which the size of $\alpha$ plays a role. It
+suggests that the estimate of $m$ becomes more accurate in the
+presence of noise for larger $\alpha$. This result is intuitively
+appealing, and is supported by numerical evidence, as shown below.
+
+%  \begin{itemize}
+%    \item[1. ] the bound \ref{eqn:mnoisebnd} holds only if 
+%      noise level $\eta=\|n\|/\|d_*\|$ satisfies
+%      \begin{equation}
+%        \label{eqn:mnoisecond}
+%        \eta < \frac{1+\sqrt{5}}{2} < 1;
+%      \end{equation}
+%    \item[2. ] if condition \ref{eqn:mnoisecond} holds, then
 %      $|m-m_*|<\lambda/r$ provided that
@@
 %      \end{equation}
-      \begin{equation}
-        \label{eqn:mnoisesuff}
-        |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}. 
-      \end{equation}
-    \end{itemize}
-\end{corollary}
-
-\begin{proof}
+%      \begin{equation}
+%        \label{eqn:mnoisesuff}
+%        |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}. 
+%      \end{equation}
+%    \end{itemize}
+%\end{corollary}
+
+\begin{proof} of Theorem \ref{thm:mnoiseres}:
   Write $\lambda = (1+\delta)\mu$, and $x=4 \pi r \alpha \mu$. Then
@@
   This maximum value must be larger than the left hand side of inequality
-  \ref{eqn:mnoisebnd} in order that there be any solutions at all, but
-  is less than $1$. This observation establishes conclusion 1 of the
-  theorem. Also, existence of $\lambda$ satisfying \ref{eqn:mnoisebnd}
-  implies that
-  \[
-    \eta(1+\eta) \le \frac{\delta}{2+\delta}.
-  \]
-  Solving this inequality for $\delta$ and unwinding the definitions
-  leads to conclusion 2.
+  \ref{eqn:mnoisebnd}, that is,
+  \[
+    \eta(1+\eta) \le \frac{\delta}{2+\delta},
+  \]
+  in order that there be any solutions at all, but the right hand side
+  is less than $1$. This observation establishes the necessity of
+  hypothesis \ref{eqn:mnoisecond} of the
+  theorem. Solving this above inequality for $\delta$ and unwinding
+  the definitions, one finds that the right-hand side of inequality
+  \ref{eqn:mnoisebnd} is bounded by \ref{eqn:mnoisesuff}, so appeal to
+  Proposition \ref{thm:mnoise} finishes the proof.
 \end{proof}
 
-\noindent {\bf Remark:} We emphasize that Theorem \ref{thm:mnoise} and
-Corollary \ref{thm:mnoisecor} state {\em sufficient} conditions for a bound on
+\noindent {\bf Remark:} We emphasize that Theorem \ref{thm:mnoiseres} states {\em sufficient} conditions for a bound on
 the slowness error $|m-m_*|$ in terms of the relative data noise level $\eta$,
@@
 
-Conclusion 1 in Corollary \ref{thm:mnoisecor} constrains the range of
+Conclusion 1 in Theorem \ref{thm:mnoiseres} constrains the range of
 noise level to which these results apply to a bit more than 60\%. That
@@
 
-On the other hand, stronger bounds than given by Corollary
-\ref{thm:mnoisecor} are possible, given additional constraints on the
+On the other hand, stronger bounds than given by Theorem
+\ref{thm:mnoiseres} are possible, given additional constraints on the
 noise $n$. A natural example is uniformly distributed random noise,
@@
 
+\subsection{Proof of Theorem \ref{thm:ipnoisesuf}}
 \begin{proof} of Theorem \ref{thm:ipnoisesuf}:
@@
 
-From Corollary \ref{thm:mnoisecor},
+From Theorem \ref{thm:mnoiseres},
 \[
@@
 \]
+which from assumption \ref{eqn:ipnoiselam} is
 \[
@@
 \[
-  \| F[m]E_{\lambda}\aw[m,d_*] -d*\|^2 = (4 \pi r \alpha)^4
+  \| F[m]E_{\lambda}\aw[m,d_*] -d_*\|^2 = (4 \pi r \alpha)^4
   \int_{-\lambda+(m-m_*)r}^{\lambda+(m-m_*)r}\,dt\, (t-(m-m_*)r)^4
@@
 \[
-  \|F[m]E_{\lambda}\aw[m,d]-d\| \le 4 \pi r \alpha
-  \lambda(\max\{1,(4\pi r \alpha \lambda)^{-2}\}\eta^2 + (4
-  \pi r \alpha \lambda)^2)^{1/2}\|d_*\|
-\]
-as asserted.
+  \|F[m]E_{\lambda}\aw[m,d]-d\| \le (4 \pi r \alpha
+  \lambda)^2\|d_*\| + \max \{4\pi r \alpha \lambda, 1\}\|n\|
+\]
+\[
+  \le ((4 \pi r \alpha \lambda)^2 +\max \{4\pi r \alpha \lambda, 1\}
+  \eta)\|d_*\|.
+\]
+If the right-hand side is to be less than $\|d_*\|$ as required by the
+definition \ref{eqn:probstat0} of the inverse problem, then
+necessarily $4\pi r\alpha \lambda < 1$, so the right hand side in the
+preceding inequality is bounded by the right hand side of assumption
+\ref{eqn:ipnoiseeps} of the theorem. Therefore this assumption implies
+that the relative residual is $\le \epsilon$.
 \end{proof}
@@
 \append{Numerics and penalty selection via the Discrepancy Principle}
-As described in the introduction, the discrepancy principle (in one of
-its guises) involves setting an acceptable range of data misfit
-$[e_-,e_+]$, and adjusting the penalty weight $\alpha$ in the reduced
-objective definition \ref{eqn:red} so that the data error term at the
-inner solution, $e[m,\aw[m;d],d]$, lies in this range. Since updating
-$m$ changes the inner problem, the appropriate condition for such
-separable problems is that $e$ {\em stays} in the range $[e_-,e_+]$ as
-$m$ is updated. In this section, we examine the dependence of $e$
-on $\alpha$ with a view to understanding how to reset $\alpha$ when
-$m$ changes.
+
+Modified Problem Statement:
+\begin{quote}
+  \label{eqn:probmod}
+  given $d \in D$ and $0 < e_- < e_+$,  find  $m \in M, \alpha \in \bR^+$ so that
+  \begin{itemize}
+  \item[(i) ]$m$ is a stationary point of $\tJa[\cdot;d]$, and
+  \item[(ii) ]$e[m,\aw[m;d];d] \in [e_-,e_+]$.
+  \end{itemize}
+\end{quote}
+
+\noindent {\bf Remark:} Given a solution $m_{\infty}, \alpha_{\infty}$, the pair
+$(m_{\infty},w_{\alpha_{\infty}}[m_{\infty};d])$ is a local minimizer
+of $J_{\alpha_{\infty}}[\cdot,\cdot;d]$ (Theorem \ref{thm:norminv}).
+%.
+%so is also a minimizer of $g[w]$ over the set
+%\[
+%  \{(m,w): e[m,w;d] \ge
+%  e[m_{\infty},w_{\alpha_{\infty}}[m_{\infty};d]]\}
+%\]
+%or equivalently a maximizer of $e[m,w,d]$ over the set
+%\[
+%  \{(m,w): m \in M, g[w] \le g[w_{\alpha_{\infty}}[m_{\infty};d]]\}
+%\]
+%according to standard Lagrange multiplier theory. That is,
+
+We describe an alternating, or coordinate search, algorithm for
+solution of problem \ref{eqn:probmod}, combining a local optimization
+algorithm for updating $m$, and a second algorithm for updating
+$\alpha$. A first version of this algorithm appeared in \cite{FuSymes2017discrepancy}.
+
+Given an
+objective function $\Phi: M \rightarrow \bR$, a local optimization algorithm generates 
+generates a map $G[\Phi,...]: M \rightarrow M$, mapping a current estimate
+of $m$ to an updated estimate. The update rule may depend only on the
+current estimate, as is the case for steepest descent, the
+Gauss-Newton method, or Newton's method, or may also depend on
+information generated during earlier updates, as for secant-type
+methods such as BFGS. The notation is intended to allow for this
+latter possibility. Under ``standard conditions'' on $\Phi$ and the initial estimate for
+$m$, and equipped with so-called globalization safeguards to ensure
+satisfaction of sufficient decrease conditions
+(see \cite{NocedalWright}), the iteration produces a sequence in $M$
+converging to a local minimizer of $\Phi$. An example of such a local
+optimization algorithm is steepest descent with backtracking line
+search.
+
+A normal stopping criterion for such an algorithm would be a tolerance
+test for the norm of $\nabla \Phi$ (and a limit on iteration count, of
+course). In our application to $\Ja = e + \alpha^2 g$, we add another
+stopping criterion:
+\begin{quote}
+  stop if $e \notin [e_-,e_+]$.
+\end{quote}
+Since $e$ is a summand in $\tJa$, $e$ will typically decrease along a
+minimizing sequence, so the expected condition invoking this stopping
+rule is $e<e_-$. We denote a update rule with this enhancement,
+applied to $\tJa$, by $G[\alpha,e_-,e_+,...]$, still allowing the
+possibility of information in addition to the current iterate.
+
+We will also require a rule for updating $\alpha$, defining a map $H: M \times E
+\rightarrow \bR^+$, in which $E=\{(e_-,e_+) \in \bR^2:
+e_-<e_+\}$. For any $m \in M$ the rule is required to produce an $\alpha$ for which
+the error bounds are satisfied: If $(m, e_-, e_+) \in M \times E$ and $\alpha = H(m, e_-,
+e_+)$, then $e[m,\aw[m,d];d] \in [e_-,e_+]$. We will describe two
+such rules below.
+
+In outline, the algorithm is as follows:
+\begin{algorithm}[H]
+\caption{Scheme for updating $m, \alpha$}
+\begin{algorithmic}[1]
+  \State Choose $m\in M$
+  \Repeat
+  \State $\alpha \gets H[m,e_-,e_+]$
+  \Repeat
+  \State $m \gets G[\alpha,e_-,e_+,...](m)$
+  \Until{$e[m,\aw[m,d];d] \notin [e_-,e_+]$ or $\|\nabla \tJa[m;d]\|$
+    sufficiently small, or...}
+  \Until{$e[m,\aw[m,d];d] \in [e_-,e_+]$} 
+\end{algorithmic}
+\end{algorithm}
+Note that after step 3, the error bounds are satisfied, that is,
+$e[m,\aw[m;d];d] \in [e_-,e_+]$, but after step 4, that is likely not
+to be the case: the alteration of $m$ is likely to reduce $e$, as it is a
+summand in the definition of $\tJa$. If the $e$ is reduced below
+$e_-$, or if an approximate local minimizer is detected, then the condition in step 6 is satisfied, so control returns
+to step 7. If it is satisfied, the algorithm terminates, elsewhich
+control loops back to step 2, $\alpha$ is updated, and the inner $m$
+update loop is entered again. Termination requires that the bounds on
+$e$ are satisfied (step 7), hence that the $m$ update loop terminates
+by finding a local min which satisfies these bounds.
 

### 2020-10-05 repro/papers/sseprecond/plots.tex
interim

@@
+\title{Plots for ``Efficient Computation of Extended Sources''}
+\author{William. W. Symes \thanks{Orcas Island, WA 98280, email {\tt symes@rice.edu}.}}
+
+\lefthead{Symes}
+
+\righthead{Approximate Source Inversion}
+
+\maketitle
+\begin{abstract}
+Examples illustrating preconditioner developed in the text.
+\end{abstract}
+
+\inputdir{project}
+
 These examples are based on two different material models.
@@
 
-\plot{ptpwindh0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km.
+\plot{ptpwindhh0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km.
 Homogeneous material model, $\kappa=4$ GPa, $\rho$=1 g/cm$^{3}$.
@@
 }
-\plot{ptvzwindhh0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindh0}.}
+\plot{ptvzwindhh0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindhh0}.}
 \plot{srcvzghh0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindhh0} by time reversal (approximate inverse) algorithm explained in text.}
 \plot{reptpwindhh0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzghh0}.} 
-\plot{trcftrhh0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindh0}, \ref{fig:reptpwindhh0}.}.
+\plot{trcftrhh0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindhh0}, \ref{fig:reptpwindhh0}.}.
 
 \plot{bml0}{width=15cm}{Inhomogenous bulk modulus field, with low velocity lens.}
-\plot{ptpwindl0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km.
+\plot{ptpwindll0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km.
 bulk modulus with low velocity lens as in Figure \ref{fig:bml0}, $\rho$=1 g/cm$^{3}$.
@@
 \plot{reptpwindll0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzgll0}, both approximate inversion and re-simulation in bulk modulus of Figure \ref{fig:bml0} (``correct velocity'').}
-\plot{trcftrll0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindl0}, \ref{fig:reptpwindll0}.}
+\plot{trcftrll0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindll0}, \ref{fig:reptpwindll0}.}

### 2020-11-01 repro/papers/discrep/paper.tex
getting close

@@
 the tendency of the mean-square error function to exhibit many
-stationary points, and local minima far from the global minimizer (we
+stationary points far from the global minimizer (we
 show an explicit example of this phenomenon later in this paper). Because of the
 typical dimensions of field FWI problems, only local gradient-based
-minimization algorithms are computationally practical, and these find
-local minimizers, so may stagnate at suboptimal and geologically
+minimization algorithms are computationally practical. These find
+stationary points (or local minima), so may stagnate at suboptimal and geologically
 uninformative earth models. Local descent methods avoid suboptimal
@@
 This paper examines one of a number of proposed alternatives to least
-squares FWI, based on an extension of the model-to-data mapping.  We
-show exactly how an example of this {\em extended inversion} approach
-leads to solution of a very simple acoustic transmission
-inverse problem {\em via} local gradient-based optimization. Despite
-its simplicity, this problem exhibits several important
-characteristics of all inverse problems in wave propagation in which
-wave velocities are amongst the unknowns. It has the advantage of
-permitting explicit calculations that reveal clearly the failure
-mechanism for FWI, that is, straightforward nonlinear least squares,
-and also the reasons for success of an extended inversion approach.
-
-In the remainder of this introduction, we will briefly describe the conceptual
-framework of this inverse problem, and sketch our main results. We leave out
-some technical details that are supplied in the next section, where we give
-precise statements of our results. In particular, theorems referenced
-in the rest of this section are to be found, stated in full, in the
-next section.
-
-The physical setting for our discussion is the propagation of waves through a
+squares FWI, based on an extension of the model-to-data mapping
+(``extended inversion''), in the context of 
+a very simple acoustic transmission inverse problem: the sound
+velocity in a fluid is to be determined from a single recording of a
+pressure wave propagated over a known distance, originating at a known
+location around a known time. We show precisely
+how descent-based FWI fails to solve this problem absent a very good
+initial model estimate, and precisely why extended inversion succeeds, using the same type
+of optimization, and from essentially arbitrary initial model
+estimate. We give rigorous estimates for the accuracy of the sound
+velocity determination in terms of the time extent of the energy
+excitation and the signal-to-noise ratio of the data. We also illustrate
+these mathematical conclusions with numerical
+examples.
+
+The single-trace transmission problem studied in this paper is a
+drastically simplified cartoon of actual seismic prospecting. It seems
+worth careful study nonetheless for two reasons: it exhibits the
+phenomenon of cycle-skipping, and its simplicity allows the properties
+of both FWI and extended inversion approaches to be established
+rigorously. The underlying mathematical structures are common to many
+inverse problems in wave propagation in which wave velocities are
+amongst the unknowns. For example, we offer an ad hoc calculation of
+the gradient of the extended inversion objective, taking advantage of
+the simplicity of the problem context. However we also show, in an
+appendix, how that calculation hides an abstract structure shared by
+many other problems.% - see for example
+%\cite{Symes:IPTA14,tenKroode:IPTA14,Symes:EAGE15,HuangSymes:SEG15a}.
+
+In the next (``Overview'') section, we describe the conceptual
+framework of this inverse problem, review the related literature, and
+state the essence of our main results, leaving out a few technical
+details that are supplied later. Numerical illustrations of these
+results occupy the following (``Examples'') section. The fourth
+(``Findings'') section of the paper contains precise statements of the
+mathematical results and supporting discussion, followed by
+derivations. In particular, theorems referenced in the Overview
+section are to be found, stated in full with their proofs, in the this
+final part of the paper. Three appendices describe the algorithm used to
+control the penalty weight in our formulation of extended
+inversion, further refinement of error estimates based ond the
+Heisenberg inequality, and the abstract structure of objective derivatives
+inherent in extended inversion based on wave propagation.
+
+\section{Overview}
+
+The physical setting for our discussion is the propagation of
+small-amplitude pressure waves through a
 homogeneous acoustic fluid occupying $\bR^3$ \cite[]{Frie:58}. The
@@
 slowness, or reciprocal wave velocity, $m > 0$. A source of acoustic
-energy localized at a point in $\bR^3$ radiates uniformly in all
-directions with time-dependent intensity $w(t)$. Both the acoustic
-pressure field and the intensity (``wavelet'') $w$ vanish at large
-negative time. A receiver records the pressure values at a distance
-$r>0$ from the source position, between the times $t_{\rm min}$ and
-$t_{\rm max}$. As explained in the next section, the
-recorded presssure is predicted to be
+energy localized at a (``source'') point $\bx_s \in \bR^3$ radiates uniformly in all
+directions with time-dependent intensity $w(t)$, generating a pressure
+field $p(\bx,t)$. Both the 
+pressure field $p$ and the intensity (``wavelet'') $w$ vanish at large
+negative time. A receiver records the pressure values at a point
+$\bx_r \in \bR^3$, a distance
+$r>0$ from the source position $\bx_s$, between the times $t_{\rm min}$ and
+$t_{\rm max}$. According to linear acoustics, the pressure field
+solves the the wave equation \cite[]{Frie:58}:
+\begin{eqnarray}
+  \label{eqn:awe}
+  \left(m^2\frac{\partial^2 p}{\partial t^2} - \nabla^2\right) p(\bx,t) &=&
+                                                                         w(t)\delta(\bx-\bx_s) \nonumber\\
+  p(\bx,t)&=&0, t\ll 0.
+\end{eqnarray}
+As the system \ref{eqn:awe} is linear and autonomous, the solution is
+the convolution of the source
+wavelet $w$ with the Green's function or fundamental solution. For
+constant $m$, the result is the well-known expression
+(\cite{CourHil:62}, Chapter VI, section 12, equation 47):
+\begin{equation}
+  \label{eqn:homsol}
+  p(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-m|\bx-\bx_s|\right),
+\end{equation}
+so the predicted pressure at the receiver point $\bx_r$ is
 \begin{equation}
 \label{eqn:mod}
-\frac{1}{4\pi r}w\left(t-mr\right)  = F[m]w(t)
-\end{equation}
-Ignoring the amplitude factor $1/(4\pi r)$, this map implements a $m$-dependent time
-shift. This time shift operator is the basis of many illustrations of
+p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right)  = F[m]w(t)
+\end{equation}
+The mapping $F[m]$ so defined is (ignoring the amplitude factor
+$1/(4\pi r)$), an $m$-dependent time shift. This time shift operator is the basis of many illustrations of
 the cycle-skipping phenomenon (for example, \cite{VirieuxOperto:09},
@@
 produces an oscillating pressure pulse that dies away
-slowly. It is commonplace to estimate this pulse then deconvolve it
+slowly. It is commonplace to estimate this pulse, usually by recording
+the emitted pressure field at a position where it should nominally be isolated from other
+signal, then deconvolve it
 from the data by safeguarded Fourier division or other means. This
@@
 impulse ($\delta(t)$): it cannot have point support, due to the finite
-frequency nature of seismic data, hence the bandlimited nature of the
-pulse estimate, but its support is ultimately much smaller than that
-of the original pulse estimate.
+frequency nature of seismic data and the Heisenberg principle (see
+Appendix B), but its amplitudes are typically small outside of a small
+much smaller time interval than is the case for the original pulse
+estimate. Thus after a small modification, the deconvolved source
+pulse has small support.
+
+For a contemporary description of view on signature deconvolution as
+practiced in the seismic industry, see \cite{SheriffGeldart:1995},
+section 9.5 and \cite{Yil:01}, Chapter 2.
 
@@
 
-Despite its simplicity, this problem exhibits the
-cycle-skipping pathology characteristic of field-scale FWI. Specifically, we show that the reduced mean square
-error function, the objective function of FWI for this problem, 
+Despite its simplicity, this problem exhibits the cycle-skipping
+pathology characteristic of field-scale FWI. Specifically, we show
+that the reduced mean square error function, the objective function of
+FWI for this problem,
 \begin{equation}
@@
 This fact implies that local optimization must fail to solve a precise
-version of the FWI problem: given $\epsilon, \lambda > 0$, $(m,w)$ is
-a solution if it lies in the domain of $\lF$ and
-$\|F[m]w-d\| \le \epsilon\|d\|$ (Problem Statement
-\ref{eqn:probstat0} below).  In fact, we show
+version of the FWI problem:
+\begin{quote} given $\epsilon, \lambda > 0$, fine $(m,w)$ in the
+domain of $\lF$ so that
+\begin{equation}
+  \label{eqn:probstat}
+  \|\lF[m]w-d\| \le \epsilon\|d\|.
+\end{equation}
+\end{quote}
+In fact, we show
 that``large residual'' local minima of $\lerr[\cdot,\cdot,d]$ exist,
@@
 defined above ($\lF$ is a restriction of $F$, so $F$ is an extension
-of $\lF$). To compensate for the resulting indeterminacy in the
-estimation of $m$, we add a penalty term to the objective: for $m \in
+of $\lF$).
+
+To compensate for the resulting indeterminacy in the
+estimation of $m$, we add a penalty term to the objective, defined by
+a penalty operator $A$ with domain and range $L^2(\bR)$: for $m \in
 \bR^+,w\in L^2(\bR), d \in L^2([t_{\rm min},t_{\rm max}])$, define
@@
   r[m,w;d] & = & \|F[m]w-d\|/\|d\|, \label{eqn:rmserr}\\
-  e[m,w;d] & = & r[m,w;d]^2, \label{eqn:mserr}\\
+  e[m,w;d] & = & \frac{1}{2}r[m,w;d]^2, \label{eqn:mserr}\\
   g[w] & = & \frac{1}{2}\|Aw\|^2/\|d\|^2, \label{eqn:pen} \\
@@
 \ref{eqn:rmserr}, \ref{eqn:mserr}, \ref{eqn:pen}, and \ref{eqn:penerr} make the values
-of $r$, $e$, $\alpha^2 g$, and $\Ja$ nondimensional.
-
-The only obvious requirement on the penalty operator $A$ is that it
+of $r$, $e$, $\alpha^2 g$, and $\Ja$ nondimensional, and simplifies
+both the statements of theoretical results and the interpretation of
+numerical examples.
+
+Apart from dimensional correctness, the only obvious requirement on the penalty operator $A$ is that it
 penalize 
@@
 Assume that $\mu>0$ and $d = F[m_*]w_* + n$, with $\mbox{supp }w_*
-\subset [-\mu,\mu]$, and denote by $\eta = \|n\|/\|F[m_*,w_*]\|$
+\subset [-\mu,\mu]$, and denote by $\eta = \|n\|/\|F[m_*]w_*\|$
 the noise-to-signal ratio. We show that any local minimizer $(m,w)$
@@
 
-The remainder of the paper begins with a complete description of the acoustic transmission problem and precise
-statements of our main results, filling in the
-few details not discussed in the preceding sketch. Then we provide a
-discussion of the potential theoretical and practical import of these
-results, including relation with prior work, some numerical
-illustrations, and directions for further research. Following this
-discussion section, we give complete proofs of all of our results.
-An appendix describes the computational algorithm used in producing
-the numerical examples, including a method for controlling the penalty
-weight $\alpha$ based on the Discrepancy Principle
-\cite[]{EnglHankeNeubauer,FuSymes2017discrepancy}. A second appendix
-explains the relation between the support constraint and well-known
-criteria for success of FWI based on data frequency content. The basis
-for this relation is the Heisenberg inequality \cite[]{Folland:07}.
-
-\section{An Acoustic Transmission Inverse Problem}
-In this section, we describe solution of the radiation
-problem for acoustic wave
-propagation in a homogeneous fluid, and various operators based on
-this solution. Using these operators, we formulate the single-trace
-acoustic transmission inverse problem. Finally, we state the main
-results of our paper.
-
-We shall use the abbreviations ${\cal B}(X,Y)$ and ${\cal I}(X,Y)$ for 
-the algebra of bounded linear operators from the Hilbert space $X$ to 
-the Hilbert space $Y$, and its subalgebra of invertible
-operators. All of the operators appearing in this discussion are
-members of ${\cal B}(X,Y)$ for some choice of Hilbert spaces $X$ and
-$Y$, and superscript $T$ denotes the transpose or adjoint of a such an
-operator in the sense specified by its domain and range Hilbert structure.
-
-\subsection{Acoustic transmission}
-Assume small amplitude (linearized) acoustic propagation, constant
-density, and isotropic point source and receiver. Denote by $m(\bx)$
-the slowness (reciprocal velocity) at spatial position $\bx$, $w(t)$
-the time dependence of the point source intensity (``wavelet'' or
-``source wavelet'') at location
-$\bx=\bx_s$. Then the (excess) pressure field $p(\bx,t)$ obeys a
-scalar wave equation \cite[]{Frie:58}:
-\begin{eqnarray}
-  \label{eqn:awe}
-  \left(m(\bx)^2\frac{\partial^2 p}{\partial t^2} - \nabla^2\right) p(\bx,t) &=&
-                                                                         w(t)\delta(\bx-\bx_s) \nonumber\\
-  p(\bx,t)&=&0, t\ll 0
-\end{eqnarray}
-Suppose that the pressure at a receiver position $\bx=\bx_r$ is recorded, at distance $r>0$ from the
-source position $\bx_s$. We shall refer to the resulting function of
-time as a {\em trace}, from common seismic usage. The dominant information in a single
-trace is the transient signal time of arrival, constraining only the mean slowness in the
-region
-between source and receiver, so assume that the
-slowness is constant, that is, independent of position $\bx$. The pressure field is simply the  the source
-wavelet $w(t)$ convolved with the
-acoustic Green's function, for which an analytic expression is
-available in the constant $m$ case \cite[]{CourHil:62}:
-\begin{equation}
-  \label{eqn:homsol}
-  p(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-m|\bx-\bx_s|\right).
-\end{equation}
-
-Define $r=|\bx_r-\bx_s| $ to be the source-receiver distance
-(``offset''). The predicted signal at
-$p(\bx_r,t)$ depends nonlinearly on the slowness $m$ and linearly on the
-source wavelet $w$. Therefore it is naturally represented as the
-action of a $m$-dependent linear operator $F[m]$ on $w$, as described
-in \ref{eqn:mod}, which we repeat here for convenience:
-\begin{equation}
-\label{eqn:mod1}
-F[m]w(t)= p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right).
-\end{equation}
-
-The slowness $m$ must be positive, as follows from basic acoustics,
-and in fact reside in a range characteristic of the
-material model: for crustal rock, a reasonable choice would be
-$m_{\rm min}=0.125, m_{\rm max}=0.6$ s/km.
-
-Field recording takes place over finite time intervals (and with
-finite sample rate, a detail we shall ignore in this paper). Denote by
-$[t_{\rm min},t_{\rm max}]$ the recording interval for the
-hypothetical single trace data.
-
-Natural choices for domain and
-range of $F$ are
-\begin{itemize}
-\item $M=(m_{\rm min}, m_{\rm max}),\,0 < m_{\rm min} \le m_{\rm
-    max}$;
-\item $W = L^2(\bR)$;
-\item $D=L^2([t_{\rm min},t_{\rm max}]),\, t_{\rm min}<t_{\rm max}$;
-\item $F: M \times W \rightarrow D$ as specified in \ref{eqn:mod}.
-\end{itemize}
-It is immediately evident from these choices and from the definition
-\ref{eqn:mod} that
-\begin{equation}
-  \label{eqn:mapprop}
-  \mbox{for }m \in M, F[m] \in {\cal B}(W,D),\mbox{ and }\|F[m]\| =
-  \frac{1}{4\pi r}.
-\end{equation}
-Note also that $F[m]$ is surjective for every $m \in M$.
-
-\noindent{\bf Remark:} In computational practice, $W$ will have to be replaced by a
-finite-dimensional subspace of $L^2(\bR)$. Many such choices will
-implicitly limit the support of $w \in W$ to a bounded interval, say
-$[T_{\rm min},T_{\rm max}]$. To maintain the surjective property,
-these bounds should be chosen so that $[t_{\rm min}, t_{\rm max}]
-\subset [T_{\rm min}+mr,T_{\rm max}+mr]$ for all $m \in M$, that is,
-\begin{eqnarray}
-  \label{eqn:dombounds}
-  t_{\rm min} &\ge & T_{\rm min}+m_{\rm max}r,\nonumber\\
-  t_{\rm max} &\le & T_{\rm max}+m_{\rm min}r.
-\end{eqnarray}
-We will ignore these computational necessities in this work,
-maintaining the definition $W=L^2(\bR)$.
-
-\subsection{An inverse problem}
-As mentioned earlier, $F[m]$ is surjective for every
-$m \in M$. Since all possible data lie in the range of $F[m]$ for any
-$m \in M$, some restriction of the domain
-of $F$ is necessary in order that fitting the data constrain
-$m$. A more precise version of the constraint proposed in the
-introduction involves a selection of a maxium support radius
-$\lambda_{\rm max} >0$. Then define for $\lambda \in (0,\lambda_{\rm max}]$:
-\begin{itemize}
-\item $\lW = \{w \in W:
-  \mbox{supp }w \subset [-\lambda,\lambda]\}$;
-\item $\lF = F|_{M \times \lW}$.
-\end{itemize}
-
-In terms of this infrastructure, the inverse problem studied in 
-this paper is:
-
-\begin{quote}
-\noindent {\bf Problem Statement:}
-  given data $d \in D$, relative error level $\epsilon \in
-  [0,1)$, and support radius $\lambda \in (0, \lambda_{\rm
-    max}]$, find $(m,w) \in M \times \lW$ for which 
-\begin{equation}
-  \label{eqn:probstat0}  \|\lF[m]w-d\| \le \epsilon\|d\|,
-\end{equation}
-or equivalently
-\begin{equation}
-  \label{eqn:probstat1}
-  \lerr[m,w;d] \le \frac{1}{2}\epsilon^2,
-\end{equation}
-with $\lerr$ defined in equation \ref{eqn:redms}.
-\end{quote}
-
-\noindent{\bf Remark:} The constraint $\epsilon < 1$ imposed on the
-target noise level is eliminates the obvious choice $(m,0)$, which
-satisfies the data misfit constraint for any $m \in M$.
-
-\noindent{\bf Remark:} The support constraint is closely linked to the
-folk theorem about FWI noted many times in the literature: convergence
-of a descent method requires that the initial slowness must be known
-to ``within a (fraction of a) wavelength''. The relation is a
-consequence of Heisenberg's inequality, and will be reviewed in
-Appendix B.
-
-The best case for data fitting
-is clearly the one in which the data can be fit precisely: that is,
-there exists $(m_*,w_*) \in M\times \lW$ so that
-\begin{equation}
-  \label{eqn:defdatanonoise}
-  d=\lF[m_*]w_*.
-\end{equation}
-Such data $d$ is {\em noise-free}, in the range of the map $\lF$. For
-such data a solution of the Problem Statement \ref{eqn:probstat0}
-exists with arbitrarily small $\epsilon>0$.
-
-\subsection{Main Results}
-While $F[m]$ is surjective for every $m \in M$, as noted above, it is
-very far from injective. On the other hand, under a constraint that
-will be assumed throughout, $\lF[m]$ is injective (in fact, $4 \pi r
-\lF[m]$ is an isometry):
-\begin{proposition}
-  \label{thm:fullrec}
-  Suppose that 
-  \begin{equation}
-    \label{eqn:fullrec}
-    [ m_{\rm min}r-\lambda_{\rm max}, m_{\rm max}r+\lambda_{\rm max}]
-    \subset [t_{\rm min},t_{\rm max}].
-  \end{equation}
-  Then $\lF[m]\in {\cal B}(\lW,D)$ is coercive for every $m \in M, \lambda \in
-  (0,\lambda_{\rm max}]$.
-\end{proposition}
-
-The first main result establishes the existence of large (100 \%)
-residual local minimizers for the basic FWI objective $\lerr$, even
-for noise-free data.
-\begin{theorem}
-  \label{thm:fwi}
-  Suppose that $0 <\lambda\le \lambda_{\rm max}$,  $m_* \in M, w_*
-  \in \lW, d=\lF[m_*]w_*$ is noise-free data per definition \ref{eqn:defdatanonoise},
-  Under assumption \ref{eqn:fullrec}, for any $m \in M$ with $|m-m_*|r>2\lambda$,
-\begin{equation}
-  \label{eqn:isovpm}
- \min_w \lerr[m,w;d]=\lerr[m,0;d] = \frac{1}{2},
-\end{equation}
-and any such $(m,0)$ is a local minimizer of $\lerr$ with relative RMS
-error = 1.0.
-\end{theorem}
-
-The main theoretical device used in the proofs of our main results on 
-extended inversion is the reduction of the penalty objective $\Ja$
-(equation \ref{eqn:penerr}) to a function $\tJa$ of 
-$m$ alone, by minimization over $w$: 
-\begin{equation}
-  \label{eqn:redexp}
-  \tJa[m;d] = \inf_w \Ja[m,w;d]
-\end{equation}
-This reduction is the foundation 
-of the {\em Variable Projection Method} (VPM) 
-\cite[]{GolubPereyra:03} for solution of partly linear (``separable'') 
-least squares problems.
-
-A minimizer $w$ on the right-hand side of definition
-\ref{eqn:redexp} must solve the {\em normal equation}
-\begin{equation}
-  \label{eqn:norm}
-  (F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td, 
-\end{equation}
-
-The choices for the penalty operator $A$ considered here are scalar 
-multiplication operators on $W$ defined by a choice of multiplier $a \in L^{\infty}(\bR)$:
-\begin{equation}
-  \label{eqn:annmult}
-  A w(t)= a(t)w(t), \, t\in \bR.
-\end{equation}
-
-\begin{theorem}
-  \label{thm:norminv}
-  Assume the condition \ref{eqn:fullrec}, $C>0$, and suppose that $A$ is
-  given by equation \ref{eqn:annmult} for $a \in L^{\infty}(\bR)$
-  satisfying
-  \begin{equation}
-    \label{eqn:abnd} 
-    a \ge 0; \, a(t) \ge C\mbox{ for }|t| \ge \lambda_{\rm max}.
-  \end{equation}
-  Then \begin{itemize}
-  \item[1. ]the reduced objective $\tJa$ is given by
-    \begin{equation}
-      \label{eqn:redexp1}
-      \tJa[m;d] = \Ja[m,\aw[m;d];d],
-    \end{equation}
-    in which $\aw[m;d] \in W$ is the unique solution of the normal
-    equation \ref{eqn:norm}.
-  \item[2. ]The following are equivalent:
-    \begin{itemize}
-    \item[i. ]$(m,w) \in M \times W$ is a local minimizer of
-      $\Ja[\cdot,\cdot;d]$, and
-    \item[ii. ]$m$ is a local minimizer of $\tJa[\cdot;d]$ and
-      $w=\aw[m;d]$.
-    \end{itemize}
-  \end{itemize}
-\end{theorem}
-
-\noindent{\bf Remark:} If $\Ja[\cdot,\cdot;d]$ and $\tJa[\cdot;d]$
-were differentiable, then ``local minimizer'' in the conclusion of the
-preceding theorem could be replaced by ``stationary point''. However,
-for the problem addressed in this paper, $\Ja[\cdot,\cdot;d]$ {\em is
-  not} differentiable without added smoothness constraints on $w$,
-whereas $\tJa[\cdot;d]$ {\em is} differentiable for proper choice of
-penalty operator $A$:
-
-\begin{theorem}
-  \label{thm:diffobj}
-  Suppose that in addition to the hypotheses of Theorem
-  \ref{thm:norminv}, $a \in W^{1,\infty}_{\rm loc}(\bR)$, then $\tJa[\cdot;d]
-  \in C^1(M)$.
-\end{theorem}
-
-This phenomenon - nondifferentiable objective, differentiable reduced
-objective - is replicated for a wide variety of extended
-approaches for inverse problems in wave propagation. We will have
-more to say about it in the Discussion section.
-
-We examine two choices for $A$. For each choice, we ask first whether local minimizers 
-of the resulting VPM objective $\tJa[\cdot;d]$ occur far from a slowness $m_*$.
-
-A penalty operator $A$ of which $\lW$ is the null space would be a
-natural choice. Such operators have come to be called
-``annihilators'', since they map all members of the constraint
-subspace $\lW$ to zero.  A simple example is
-\begin{eqnarray}
-  A = E^c_{\lambda}&=&I - E_{\lambda},\mbox{ where } \nonumber \\
-  E_{\lambda}w(t) &=&{\bf 1}_{[-\lambda,\lambda]}(t)w(t). 
-                      \label{eqn:ann0}
-\end{eqnarray}
-That is, $E_{\lambda}$ is the orthogonal projector onto $\lW$,
-and $E_{\lambda}^c$ is the orthogonal projector onto its
-orthocomplement, an operator of the form \ref{eqn:annmult} with $a
-= 1 - {\bf 1}_{[-\lambda,\lambda]}$. %Note that this choice of $a$
-%satisfies condition \ref{eqn:abnd} with $C \equiv 1$.
-
-\begin{theorem}
-  \label{thm:boxcarbad}
-  Suppose that
-  \begin{itemize}
-  \item[1. ] $\lambda \in (0,\lambda_{\rm max}]$;
-  \item[2. ] $m_*\in M, w_*\in \lW$, and $d=F[m_*]w_*$ (noise-free
-    data);
-  \item[3. ] $A=E^c_{\lambda}$, that is, $a=1-{\bf 1}_{-\lambda,\lambda]}$ in
-    the definition \ref{eqn:annmult}.
-  \end{itemize}
-  Then if $|m-m_*| >  2\lambda/r$,
-  \[
-    \tJa[m;d] = \frac{(4\pi r \alpha)^2}{2(1+(4 \pi r \alpha)^2)}.
-  \]
-\end{theorem}
-
-\noindent {\bf Remark.}
-One might have thought that $A=E^c_{\lambda}$ would be a better choice
-of annihilator, as for noise-free
-data, the solution set defined by the problem statement
-\ref{eqn:probstat0} is the same as the set of global minimizers of
-$\Ja$ in this case. However, for this choice of annihilator,
-$t\Ja$ exhibits the same feature as the mean square error $e$, namely
-a continuum of local minimizers at any distance from the global
-minimizer $m_*$ greater than a multiple of $\lambda$. Therefore the
-extended inversion with this choice of annihilator is no more amenable
-to local optimization than is FWI.
-
-A second possible 
-penalty operator penalizes energy away from 
-$t=0$: choose $\tau > 0$ and set 
-\begin{equation}
-  \label{eqn:ann}
-  a(t) = \min(|t|, \tau). 
-\end{equation}
-Note that $a \ge 0$ and $a \in L^{\infty}(\bR) \cap W^{1,1}_{\rm
-  loc}(\bR)$. The cutoff $\tau$ will be chosen large enough to be effectively inactive: 
-specifically, hindsight suggests 
-\begin{equation}
-  \label{eqn:taudef}
-  \tau = \max\{|t_{\rm min}-m_{\rm min}r|,|t_{\rm min}-m_{\rm max}r|, |t_{\rm max}-m_{\rm min}r|, |t_{\rm max}-m_{\rm max}r|\}. 
-\end{equation}
-This 
-particular annihilator has been employed in earlier papers on extended 
-source inversion 
-\cite[]{Plessix:00a,LuoSava:11,Warner:14,HuangSymes:SEG15a,Warner:16,HuangSymes:GEO17}.
-
-\begin{theorem}
-  \label{thm:mnoiseres}
-  Asumme that
-  \begin{itemize}
-  \item[1. ] $\alpha, \mu> 0$,
-  \item[2. ] $m_* \in M, w_* \in W_{\mu}$,
-  \item[3. ] $d_* = F[m_*]w_*$,
-  \item[4. ] $n \in D$ and $d = d_* + n$.
-  \end{itemize}
-  Set $\eta = \|n\|/\|d_*\|$, and assume further that
-  \begin{itemize}
-  \item[5. ]
-    \begin{equation}
-      \label{eqn:mnoisecond}
-      \eta < \frac{\sqrt{5}-1}{2}, and
-    \end{equation}
-  \item[6. ] $m$ is a stationary point of $\tJa[\cdot;d]$.
-  \end{itemize}
-  Then
-  \begin{equation}
-    \label{eqn:mnoisesuff}
-    |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}.
-  \end{equation}
-\end{theorem}
-
-\noindent {\bf Remark.} That is, {\em with the choice of penalty
-  multiplier $a$ given in equation \ref{eqn:ann}, and support radius
-  $\mu$ of the ``noise-free'' wavelet $w_*$, $\Ja$ has no local
-  minima with slownesses further than $(1+O(\eta))\mu/r)$ from the slowness used to
-  generate the data}.
-
-\noindent {\bf Remark.} The estimate $|m-m_*|r<\mu(1+O(\eta))$ for
-local minima of $\Ja$ is sharp: it is possible to choose $w_* \in W_{\mu}$
-so that $\mu - |m-m_*|r$ is as small as you like. In particular,
-the ``exact'' or ``true'' slowness $m_*$ is not necessarily the only
-slowness component of a local minimizer, or even the slowness
-component of any local minimizer, and in particular
-is not (necessarily) the slowness component of a global minimizer of $\Ja$.
-
-\noindent {\bf Remark.} No similar bound could hold for much larger
-noise levels than specified in condition \ref{eqn:mnoisecond}, the
-right-hand side of which is a bit larger than 0.6. For example, if the noise is
-the predicted data for the same wavelet $w_*$ with a substantially different
-slowness $m_{\flat}$, that is, $n=F[m_{\flat}]w_*$, then a simple
-symmetry argument shows that if there is a local minimizer of $\Ja[\cdot,\cdot;d_*+n]$. with
-slowness near $m_*$,
-there must also be a minimizer with slowness near $m_{\flat}$.
-so that the difference with $m_*$ is not constrained at
-all by the assumed support radius of $w_*$. So for this example with 100\% noise, no
-bound of the type given by conclusion 2 could possibly hold. We will
-illustrate this phenomenon numerically later in the paper.
-
-\noindent{\bf Remark.} Note that $\alpha$ plays no role in the
-conclusions of this theorem. It is only required that $\alpha >0$.
-
-Unless the data is noise-free, there is no reason to suppose that the
-estimated wavelet $\aw[m;d]$ (Theorem \ref{thm:norminv}) will lie in
-$\lW$, unless the support of the noise $n$ is not restricted. In order
-to construct a solution of the inverse problem \ref{eqn:probstat0}, we
-project $\aw[m;d]$ onto $\lW$. For sufficiently large $\lambda,
-\epsilon$, the result is a solution of the inverse problem:
-
-\begin{theorem}
-  \label{thm:ipnoisesuf}
-  Assume the hypotheses of Theorem \ref{thm:mnoiseres}, and that
-  inequality \ref{eqn:mnoisecond} holds, and $\mu \in
-  (0,\lambda_{\rm max}]$. Then the pair
-  \[
-    (m,{\bf 1}_{[-\lambda,\lambda]}\aw[m,d])
-  \]
-  solves the inverse problem as stated in \ref{eqn:probstat0} if
-  \begin{equation}
-    \label{eqn:ipnoiselam}
-    \left( 2+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\mu \le \lambda
-    \le \lambda_{\rm max}, 
-  \end{equation}
-  and
-  \begin{equation}
-    \label{eqn:ipnoiseeps}
-    \epsilon \ge (4 \pi r \alpha\lambda)^2+\eta. 
-  \end{equation}    
-\end{theorem}
-
-\noindent {\bf Remark:} Note that the sufficient condition \ref{eqn:ipnoiselam} for
-$\lambda$ is independent of $\alpha$. It follows that for any choice
-of $\lambda$ consistent with this bound, $(m,{\bf
-  1}_{[-\lambda,\lambda]}\aw[m,d])$ is a solution of the inverse
-problem for any $\epsilon > 0$ provided that $\alpha$ is chosen
-sufficiently small ($O(\sqrt{\epsilon})$).
-
-\section{Discussion}
-
 Use of model extension to reveal the influence of earth mechanical
@@
 
-Many inversion concepts sound plausible, and appear
-to work at least to some extent as one might hope from their heuristic
-justifications. However very few of these approaches have been
-underwritten by mathematical argument: in essence, they are mostly
-justified only by numerical example, ``in the rear-view mirror'', with no assurance that
-failure is not just around the corner, at the next example. On top of
-that, some of these approaches, for example those based on the
-computationally attractive Variable Projection Method (``VPM'') of
-\cite{GolubPereyra:03}, are cast in such form that the reasons for
-success are not readily apparent.
-
-In many cases, the constraint posed by membership of the second component $w$ in the subspace $\lW$ is an approximation to an idealized limit $\lambda \rightarrow 0$.
-It may be that $W_0 = \cap_{\lambda > 0}\lW$ is non-trivial, in which
-case $F$ may be viewed as an extension of $F_0 = F|_{ M \times
-  W_0}$. In that case, $F_0$ and $F$ realize the idealized extension
-structure discussed in many prior works on extended inversion (for
-example, \cite[]{geoprosp:2008}): $F_0$ is the physical modeling
-operator, and $W_0$ consists of second components of physical or
-feasible models. However, in many of these examples, $W_0$ is not
-computationally accessible, or - even if it is - gives results that
-closely approximate those obtained with $\lW$ for small $\lambda$. For
-other examples, for instance that  discussed later in this paper (and
-implicitly in \cite[]{Warner:16}, for instance), $W_0=\{0\}$ and the
-family $\{\lW: \lambda > 0\}$ actually embodies all of the necessary
-physics.
+%In many cases, the constraint posed by membership of the second component $w$ in the subspace $\lW$ is an approximation to an idealized limit $\lambda \rightarrow 0$.
+%It may be that $W_0 = \cap_{\lambda > 0}\lW$ is non-trivial, in which
+%case $F$ may be viewed as an extension of $F_0 = F|_{ M \times
+%  W_0}$. In that case, $F_0$ and $F$ realize the idealized extension
+%structure discussed in many prior works on extended inversion (for
+%example, \cite[]{geoprosp:2008}): $F_0$ is the physical modeling
+%operator, and $W_0$ consists of second components of physical or
+%feasible models. However, in many of these examples, $W_0$ is not
+%computationally accessible, or - even if it is - gives results that
+%closely approximate those obtained with $\lW$ for small $\lambda$. For
+%other examples, for instance that  discussed later in this paper (and
+%implicitly in \cite[]{Warner:16}, for instance), $W_0=\{0\}$ and the
+%family $\{\lW: \lambda > 0\}$ actually embodies all of the necessary
+%physics.
+
+\section{Examples}
+This section presents numerical examples of the estimates of $m$ and
+$w$ obtained by minimization of $\Ja$. We have selected these examples
+to illustrate the conclusions described in the Overview section.
 
@@
 sensitivity to $m$ versus $w$, as observed in the cited reference and elsewhere.
-%especially for high frequency $f$, as
-%the reader will see below.
-%as will be established below.
 Instead, a nested approach, in which $w$ is
 eliminated in an inner optimization,
-generally gives far better numerical performance.  This takes advantage of
-$\Ja$ being quadratic in
-$w$ to eliminate $w$ from the problem statement, as described in the
-following theorem.
+generally gives far better numerical performance. This approach was
+formalized by \cite{GolubPereyra:73,GolubPereyra:03} as the {\em
+  Variable Projection Method} (``VPM''). Besides being the main
+computational device used in our examples, it also plays a central
+role in the theory developed in the next section.
+
+The minimization of $\Ja$ over $w$ yields a function $\tJa$ of $m$
+alone:
+\begin{equation}
+  \label{eqn:redexp0}
+  \tJa[m;d] = \inf_w \Ja[m,w;d]
+\end{equation}
+Under appropriate conditions, met in our examples (Theorem
+\ref{thm:norminv}, $\tJa$ has an alternate expression
+\begin{equation}
+  \label{eqn:redexp10}
+  \tJa[m;d] = \Ja[m,\aw[m;d];d],
+\end{equation}
+in which $\aw[m;d] \in W$ is the solution of the normal
+equation
+\begin{equation}
+  \label{eqn:norm0}
+  (F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td.
+\end{equation}
+
+[ examples and discussion - Huiyi]
+
+[ may need to repeat some of the explicit formulas for $\aw, e, g,
+\tJa$ from the next section - refer to theorems]
+
 %\section{Theory}
@@
 
-The inverse problem stated at the beginning of this paper concerns
-three types of
-objects: a set of vectors of measured or observed parameters (``the
-data''); a set of vectors of non-observed parameters (``the model''); and a
-modeling or prediction operator, which accepts model vectors as
-input and produces predicted data vectors as output.
-
-The ``inverse problem'' is really a family of questions, parametrized
-by a scalar tolerance. The question implicitly posed in the statement
-\ref{eqn:probstat0} is: given a tolerance and an observed data vector,
-is there a model vector for which the predicted data vector is within
-the tolerance of the observed data vector, in the sense of a distance
-function on the collection of all possible data vectors.
-
-This deterministic definition of inverse problems has been presented by
-many authors, for instance \cite{ddjackson:72,Parker:94}. Most variants 
-presume that the data and model vectors form Hilbert spaces, or
-possibly subsets of Hilbert spaces, and that the distance between
-predicted and observed data vectors is the norm of the difference. Our
-approach fits squarely in this tradition.
-
-In any concrete instance, a ``yes'' answer can only be justified by
-exhibition of a model vector satisfying the fit criterion, hence the
-form of the statement \ref{eqn:probstat0}. The extension approach
-discussed in this paper relies on use of an extended modeling or
-prediction operator, satisfying two criteria:
+\section{Findings}
+This section describes the mathematical infrastructure necessary to
+make precise the conclusions described in the Overview section, then
+gives complete statements and proofs, along with those of
+necessary auxiliary results.
+
+We shall use the abbreviations ${\cal B}(X,Y)$ and ${\cal I}(X,Y)$ for 
+the algebra of bounded linear operators from the Hilbert space $X$ to 
+the Hilbert space $Y$, and its subalgebra of invertible
+operators. All of the operators appearing in this discussion are
+members of ${\cal B}(X,Y)$ for some choice of Hilbert spaces $X$ and
+$Y$, and superscript $T$ denotes the transpose or adjoint of a such an
+operator in the sense specified by its domain and range Hilbert structure.
+
+\subsection{An Acoustic transmission Inverse Problem}
+The data prediction or modeling operator for the single-trace acoustic
+transmission problem has been defined above, in equation
+\ref{eqn:mod}, which we restate here for convenience:
+\begin{equation}
+\label{eqn:mod1}
+F[m]w(t)  = p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right) 
+\end{equation}
+The slowness $m$ must be positive, as follows from basic acoustics,
+and in fact reside in a range characteristic of the
+material model: for crustal rock, a reasonable choice would be
+$m_{\rm min}=0.125, m_{\rm max}=0.6$ s/km.
+
+Natural choices for domain and
+range of $F$ are thus
 \begin{itemize}
-\item the extended operator is an extension of the modeling
-  operator, that is, its domain is a superset of the model vector
-  space and, when fed a model vector, it replicates the action of
-  the modeling operator;
-\item the extended operator is {\em surjective}, that is, all data
-  vectors can be fit without error.
+\item $M=(m_{\rm min}, m_{\rm max}),\,0 < m_{\rm min} \le m_{\rm
+    max}$;
+\item $W = L^2(\bR)$;
+\item $D=L^2([t_{\rm min},t_{\rm max}]),\, t_{\rm min}<t_{\rm max}$;
+\item $F: M \times W \rightarrow D$ as specified in \ref{eqn:mod}.
 \end{itemize}
-
-\section{Derivations}
+It is immediately evident from these choices and from the definition
+\ref{eqn:mod} that
+\begin{equation}
+  \label{eqn:mapprop}
+  \mbox{for }m \in M, F[m] \in {\cal B}(W,D),\mbox{ and }\|F[m]\| =
+  \frac{1}{4\pi r}.
+\end{equation}
+Note also that $F[m]$ is surjective for every $m \in M$.
+
+\noindent{\bf Remark:} In computational practice, $W$ will have to be replaced by a
+finite-dimensional subspace of $L^2(\bR)$. Many such choices will
+implicitly limit the support of $w \in W$ to a bounded interval, say
+$[T_{\rm min},T_{\rm max}]$. To maintain the surjective property,
+these bounds should be chosen so that $[t_{\rm min}, t_{\rm max}]
+\subset [T_{\rm min}+mr,T_{\rm max}+mr]$ for all $m \in M$, that is,
+\begin{eqnarray}
+  \label{eqn:dombounds}
+  t_{\rm min} &\ge & T_{\rm min}+m_{\rm max}r,\nonumber\\
+  t_{\rm max} &\le & T_{\rm max}+m_{\rm min}r.
+\end{eqnarray}
+We will ignore these computational necessities in this work,
+maintaining the definition $W=L^2(\bR)$.
+
+As mentioned earlier, $F[m]$ is surjective for every
+$m \in M$. Since all possible data lie in the range of $F[m]$ for any
+$m \in M$, some restriction of the domain
+of $F$ is necessary in order that fitting the data constrain
+$m$. The constraint proposed in the
+Overview section involves a selection of a maxium support radius
+$\lambda_{\rm max} >0$. Then define for $\lambda \in (0,\lambda_{\rm max}]$:
+\begin{itemize}
+\item $\lW = \{w \in W:
+  \mbox{supp }w \subset [-\lambda,\lambda]\}$;
+\item $\lF = F|_{M \times \lW}$.
+\end{itemize}
+
+In terms of this infrastructure, the inverse problem studied in 
+this paper may be stated as a slightly refined version of the problem statement
+\ref{eqn:probstat}:
+
+\begin{quote}
+\noindent {\bf Problem Statement:}
+  given data $d \in D$, relative error level $\epsilon \in
+  [0,1)$, and support radius $\lambda \in (0, \lambda_{\rm
+    max}]$, find $(m,w) \in M \times \lW$ for which 
+\begin{equation}
+  \label{eqn:probstat0}  \|\lF[m]w-d\| \le \epsilon\|d\|,
+\end{equation}
+or equivalently
+\begin{equation}
+  \label{eqn:probstat1}
+  \lerr[m,w;d] \le \frac{1}{2}\epsilon^2,
+\end{equation}
+with $\lerr$ defined in equation \ref{eqn:redms}.
+\end{quote}
+
+\noindent{\bf Remark:} The constraint $\epsilon < 1$ imposed on the
+target noise level is eliminates the obvious choice $(m,0)$, which
+satisfies the data misfit constraint for any $m \in M$.
+
+\noindent{\bf Remark:} The support constraint is closely linked to the
+folk theorem about FWI noted many times in the literature: convergence
+of a descent method requires that the initial slowness must be known
+to ``within a (fraction of a) wavelength''. The relation is a
+consequence of Heisenberg's inequality, and will be reviewed in
+Appendix B.
+
+The best case for data fitting
+is clearly the one in which the data can be fit precisely: that is,
+there exists $(m_*,w_*) \in M\times \lW$ so that
+\begin{equation}
+  \label{eqn:defdatanonoise}
+  d=\lF[m_*]w_*.
+\end{equation}
+Such data $d$ is {\em noise-free}, in the range of the map $\lF$. For
+such data a solution of the Problem Statement \ref{eqn:probstat0}
+exists with arbitrarily small $\epsilon>0$.
+
+\subsection{Results}
+While $F[m]$ is surjective for every $m \in M$, as noted above, it is
+very far from injective. On the other hand, under a constraint that
+will be assumed throughout, $\lF[m]$ is injective (in fact, $4 \pi r
+\lF[m]$ is an isometry):
+\begin{proposition}
+  \label{thm:fullrec}
+  Suppose that 
+  \begin{equation}
+    \label{eqn:fullrec}
+    [ m_{\rm min}r-\lambda_{\rm max}, m_{\rm max}r+\lambda_{\rm max}]
+    \subset [t_{\rm min},t_{\rm max}].
+  \end{equation}
+  Then $\lF[m]\in {\cal B}(\lW,D)$ is coercive for every $m \in M, \lambda \in
+  (0,\lambda_{\rm max}]$.
+\end{proposition}
+
+The first main result establishes the existence of large (100 \%)
+residual local minimizers for the basic FWI objective $\lerr$, even
+for noise-free data.
+\begin{theorem}
+  \label{thm:fwi}
+  Suppose that $0 <\lambda\le \lambda_{\rm max}$,  $m_* \in M, w_*
+  \in \lW, d=\lF[m_*]w_*$ is noise-free data per definition \ref{eqn:defdatanonoise},
+  Under assumption \ref{eqn:fullrec}, for any $m \in M$ with $|m-m_*|r>2\lambda$,
+\begin{equation}
+  \label{eqn:isovpm}
+ \min_w \lerr[m,w;d]=\lerr[m,0;d] = \frac{1}{2},
+\end{equation}
+and any such $(m,0)$ is a local minimizer of $\lerr$ with relative RMS
+error = 1.0.
+\end{theorem}
+
+The main theoretical device used in the proofs of our main results on 
+extended inversion is Variable Projection reduction of the penalty objective $\Ja$
+(equation \ref{eqn:penerr}) to a function $\tJa$ of 
+$m$ alone, by minimization over $w$, as described in the Examples
+section. We repeat the definition here for the reader's convenience:
+\begin{equation}
+  \label{eqn:redexp}
+  \tJa[m;d] = \inf_w \Ja[m,w;d]
+\end{equation}
+
+A minimizer $w$ on the right-hand side of definition
+\ref{eqn:redexp} must solve the {\em normal equation}
+\begin{equation}
+  \label{eqn:norm}
+  (F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td, 
+\end{equation}
+
+The choices for the penalty operator $A$ considered here are scalar 
+multiplication operators on $W$ defined by a choice of multiplier $a \in L^{\infty}(\bR)$:
+\begin{equation}
+  \label{eqn:annmult}
+  A w(t)= a(t)w(t), \, t\in \bR.
+\end{equation}
+
+\begin{theorem}
+  \label{thm:norminv}
+  Assume the condition \ref{eqn:fullrec}, $C>0$, and suppose that $A$ is
+  given by equation \ref{eqn:annmult} for $a \in L^{\infty}(\bR)$
+  satisfying
+  \begin{equation}
+    \label{eqn:abnd} 
+    a \ge 0; \, a(t) \ge C\mbox{ for }|t| \ge \lambda_{\rm max}.
+  \end{equation}
+  Then \begin{itemize}
+  \item[1. ]the reduced objective $\tJa$ is given by
+    \begin{equation}
+      \label{eqn:redexp1}
+      \tJa[m;d] = \Ja[m,\aw[m;d];d],
+    \end{equation}
+    in which $\aw[m;d] \in W$ is the unique solution of the normal
+    equation \ref{eqn:norm}.
+  \item[2. ]The following are equivalent:
+    \begin{itemize}
+    \item[i. ]$(m,w) \in M \times W$ is a local minimizer of
+      $\Ja[\cdot,\cdot;d]$, and
+    \item[ii. ]$m$ is a local minimizer of $\tJa[\cdot;d]$ and
+      $w=\aw[m;d]$.
+    \end{itemize}
+  \end{itemize}
+\end{theorem}
+
+\noindent{\bf Remark:} If $\Ja[\cdot,\cdot;d]$ and $\tJa[\cdot;d]$
+were differentiable, then ``local minimizer'' in the conclusion of the
+preceding theorem could be replaced by ``stationary point''. However,
+for the problem addressed in this paper, $\Ja[\cdot,\cdot;d]$ {\em is
+  not} differentiable without added smoothness constraints on $w$,
+whereas $\tJa[\cdot;d]$ {\em is} differentiable for proper choice of
+penalty operator $A$:
+
+\begin{theorem}
+  \label{thm:diffobj}
+  Suppose that in addition to the hypotheses of Theorem
+  \ref{thm:norminv}, $a \in W^{1,\infty}_{\rm loc}(\bR)$, then $\tJa[\cdot;d]
+  \in C^1(M)$.
+\end{theorem}
+
+This phenomenon - nondifferentiable objective, differentiable reduced
+objective - is replicated for a wide variety of extended
+approaches for inverse problems in wave propagation. See for example \cite{Symes:EAGE15}.
+
+We examine two choices for $A$. For each choice, we ask first whether local minimizers 
+of the resulting VPM objective $\tJa[\cdot;d]$ occur far from a slowness $m_*$.
+
+A penalty operator $A$ of which $\lW$ is the null space would be a
+natural choice. Such operators have come to be called
+``annihilators'', since they map all members of the constraint
+subspace $\lW$ to zero.  A simple example is
+\begin{eqnarray}
+  A = E^c_{\lambda}&=&I - E_{\lambda},\mbox{ where } \nonumber \\
+  E_{\lambda}w(t) &=&{\bf 1}_{[-\lambda,\lambda]}(t)w(t). 
+                      \label{eqn:ann0}
+\end{eqnarray}
+That is, $E_{\lambda}$ is the orthogonal projector onto $\lW$,
+and $E_{\lambda}^c$ is the orthogonal projector onto its
+orthocomplement, an operator of the form \ref{eqn:annmult} with $a
+= 1 - {\bf 1}_{[-\lambda,\lambda]}$. %Note that this choice of $a$
+%satisfies condition \ref{eqn:abnd} with $C \equiv 1$.
+
+\begin{theorem}
+  \label{thm:boxcarbad}
+  Suppose that
+  \begin{itemize}
+  \item[1. ] $\lambda \in (0,\lambda_{\rm max}]$;
+  \item[2. ] $m_*\in M, w_*\in \lW$, and $d=F[m_*]w_*$ (noise-free
+    data);
+  \item[3. ] $A=E^c_{\lambda}$, that is, $a=1-{\bf 1}_{-\lambda,\lambda]}$ in
+    the definition \ref{eqn:annmult}.
+  \end{itemize}
+  Then if $|m-m_*| >  2\lambda/r$,
+  \[
+    \tJa[m;d] = \frac{(4\pi r \alpha)^2}{2(1+(4 \pi r \alpha)^2)}.
+  \]
+\end{theorem}
+
+\noindent {\bf Remark.}
+One might have thought that $A=E^c_{\lambda}$ would be a better choice
+of annihilator, as for noise-free
+data, the solution set defined by the problem statement
+\ref{eqn:probstat0} is the same as the set of global minimizers of
+$\Ja$ in this case. However, for this choice of annihilator,
+$t\Ja$ exhibits the same feature as the mean square error $e$, namely
+a continuum of local minimizers at any distance from the global
+minimizer $m_*$ greater than a multiple of $\lambda$. Therefore the
+extended inversion with this choice of annihilator is no more amenable
+to local optimization than is FWI.
+
+A second possible 
+penalty operator penalizes energy away from 
+$t=0$: choose $\tau > 0$ and set 
+\begin{equation}
+  \label{eqn:ann}
+  a(t) = \min(|t|, \tau). 
+\end{equation}
+Note that $a \ge 0$ and $a \in L^{\infty}(\bR) \cap W^{1,1}_{\rm
+  loc}(\bR)$. The cutoff $\tau$ will be chosen large enough to be effectively inactive: 
+specifically, hindsight suggests 
+\begin{equation}
+  \label{eqn:taudef}
+  \tau = \max\{|t_{\rm min}-m_{\rm min}r|,|t_{\rm min}-m_{\rm max}r|, |t_{\rm max}-m_{\rm min}r|, |t_{\rm max}-m_{\rm max}r|\}. 
+\end{equation}
+This 
+particular annihilator has been employed in earlier papers on extended 
+source inversion 
+\cite[]{Plessix:00a,LuoSava:11,Warner:14,HuangSymes:SEG15a,Warner:16,HuangSymes:GEO17}.
+
+\begin{theorem}
+  \label{thm:mnoiseres}
+  Asumme that
+  \begin{itemize}
+  \item[1. ] $\alpha, \mu> 0$,
+  \item[2. ] $m_* \in M, w_* \in W_{\mu}$,
+  \item[3. ] $d_* = F[m_*]w_*$,
+  \item[4. ] $n \in D$ and $d = d_* + n$.
+  \end{itemize}
+  Set $\eta = \|n\|/\|d_*\|$, and assume further that
+  \begin{itemize}
+  \item[5. ]
+    \begin{equation}
+      \label{eqn:mnoisecond}
+      \eta < \frac{\sqrt{5}-1}{2}, and
+    \end{equation}
+  \item[6. ] $m$ is a stationary point of $\tJa[\cdot;d]$.
+  \end{itemize}
+  Then
+  \begin{equation}
+    \label{eqn:mnoisesuff}
+    |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}.
+  \end{equation}
+\end{theorem}
+
+\noindent {\bf Remark.} That is, {\em with the choice of penalty
+  multiplier $a$ given in equation \ref{eqn:ann}, and support radius
+  $\mu$ of the ``noise-free'' wavelet $w_*$, $\Ja$ has no local
+  minima with slownesses further than $(1+O(\eta))\mu/r)$ from the slowness used to
+  generate the data}.
+
+\noindent {\bf Remark.} The estimate $|m-m_*|r<\mu(1+O(\eta))$ for
+local minima of $\Ja$ is sharp: it is possible to choose $w_* \in W_{\mu}$
+so that $\mu - |m-m_*|r$ is as small as you like. In particular,
+the ``exact'' or ``true'' slowness $m_*$ is not necessarily the only
+slowness component of a local minimizer, or even the slowness
+component of any local minimizer, and in particular
+is not (necessarily) the slowness component of a global minimizer of $\Ja$.
+
+\noindent {\bf Remark.} No similar bound could hold for much larger
+noise levels than specified in condition \ref{eqn:mnoisecond}, the
+right-hand side of which is a bit larger than 0.6. For example, if the noise is
+the predicted data for the same wavelet $w_*$ with a substantially different
+slowness $m_{\flat}$, that is, $n=F[m_{\flat}]w_*$, then a simple
+symmetry argument shows that if there is a local minimizer of $\Ja[\cdot,\cdot;d_*+n]$. with
+slowness near $m_*$,
+there must also be a minimizer with slowness near $m_{\flat}$.
+so that the difference with $m_*$ is not constrained at
+all by the assumed support radius of $w_*$. So for this example with 100\% noise, no
+bound of the type given by conclusion 2 could possibly hold. We will
+illustrate this phenomenon numerically later in the paper.
+
+\noindent{\bf Remark.} Note that $\alpha$ plays no role in the
+conclusions of this theorem. It is only required that $\alpha >0$.
+
+Unless the data is noise-free, there is no reason to suppose that the
+estimated wavelet $\aw[m;d]$ (Theorem \ref{thm:norminv}) will lie in
+$\lW$, unless the support of the noise $n$ is not restricted. In order
+to construct a solution of the inverse problem \ref{eqn:probstat0}, we
+project $\aw[m;d]$ onto $\lW$. For sufficiently large $\lambda,
+\epsilon$, the result is a solution of the inverse problem:
+
+\begin{theorem}
+  \label{thm:ipnoisesuf}
+  Assume the hypotheses of Theorem \ref{thm:mnoiseres}, and that
+  inequality \ref{eqn:mnoisecond} holds, and $\mu \in
+  (0,\lambda_{\rm max}]$. Then the pair
+  \[
+    (m,{\bf 1}_{[-\lambda,\lambda]}\aw[m,d])
+  \]
+  solves the inverse problem as stated in \ref{eqn:probstat0} if
+  \begin{equation}
+    \label{eqn:ipnoiselam}
+    \left( 2+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\mu \le \lambda
+    \le \lambda_{\rm max}, 
+  \end{equation}
+  and
+  \begin{equation}
+    \label{eqn:ipnoiseeps}
+    \epsilon \ge (4 \pi r \alpha\lambda)^2+\eta. 
+  \end{equation}    
+\end{theorem}
+
+\noindent {\bf Remark:} Note that the sufficient condition \ref{eqn:ipnoiselam} for
+$\lambda$ is independent of $\alpha$. It follows that for any choice
+of $\lambda$ consistent with this bound, $(m,{\bf
+  1}_{[-\lambda,\lambda]}\aw[m,d])$ is a solution of the inverse
+problem for any $\epsilon > 0$ provided that $\alpha$ is chosen
+sufficiently small ($O(\sqrt{\epsilon})$).
+
+\subsection{Proof of Theorem \ref{thm:fwi}}
 \noindent{\bf Remark:} A useful consequence of the condition 
@@
 
-\subsection{Proof of Theorem \ref{thm:fwi}}
 %A natural formulation of the Problem Statement \ref{eqn:probstat0} is
@@
  \lerr[m,w;d] =  \frac{1}{32\pi^2
-    r^2}\int_0^T\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2
+    r^2\|d\|^2}\int_0^T\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2
 \]
@@
 orthogonal to $F[m]w$. So conclude after a minor manipulation that
+\[
+  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \lerr[m,w;d]=\frac{1}{32\pi^2 
+    r^2\|d\|^2}(\|w\|^2 + \|w_*\|^2).
+\]
 \begin{equation}
   \label{eqn:iso}
-  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, e[m,w;d]=\frac{1}{32\pi^2
-    r^2}(\|w\|^2 + \|w_*\|^2).
+  = \frac{1}{2}\left(\frac{\|w\|^2}{\|w_*\|^2} + 1 \right)
 \end{equation}
 That is, for slowness $m$ in error by more than $2\lambda/r$ from the 
-target slowness $m_*$, the means square error (FWI objective) $e$ is independent of
+target slowness $m_*$, the means square error (FWI objective) $\lerr$ is independent of
 $m$, and its minimum over $w$ is attained for $w=0$
@@
 
-The role of $A$ is to penalize distance from the constrained
-subspace $\lW$, in some sense.
-
 With these choices, the normal equation \ref{eqn:norm1} becomes
@@
 
-\begin{proof}
+\begin{proof} of Theorem \ref{thm:norminv}:
   \begin{itemize}
@@
 
-\begin{proof} of Theorem \ref{thm:norminv}: Since the normal operator
-  is invertible (item 1 of Proposition \ref{thm:norminvexp}), the
-  conclusions of this theorem follow from the Lax-Milgram Theorem
-  (\cite{Yosida}, III.7).
-\end{proof}
+%\begin{proof} of Theorem \ref{thm:norminv}: Since the normal operator
+%  is invertible (item 1 of Proposition \ref{thm:norminvexp}), the
+%  conclusions of this theorem follow from the Lax-Milgram Theorem
+%  (\cite{Yosida}, III.7).
+%\end{proof}
 
@@
   \label{eqn:residnormgen}
-  e[m,\aw[m,d];d] = \frac{1}{2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r \alpha a(t-mr))^4(1 +
+  e[m,\aw[m,d];d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r \alpha a(t-mr))^4(1 +
   (4\pi r \alpha a(t-mr))^2)^{-2}d(t)^2
@@
   \label{eqn:anninormgen}
-  p[m,\aw[m,d];d] = \frac{1}{2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r a(t-mr))^2(1 +
+  p[m,\aw[m,d];d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r a(t-mr))^2(1 +
   (4\pi r \alpha a(t-mr))^2)^{-2}d(t)
@@
   \label{eqn:expjgen}
-\tJa[m;d] = \frac{1}{2}\int_{t_{\rm min}}^{t_{\rm max}}\,dt\,(4\pi r \alpha a(t-mr))^2(1+(4\pi r \alpha 
+\tJa[m;d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}}\,dt\,(4\pi r \alpha a(t-mr))^2(1+(4\pi r \alpha 
 a(t-mr))^2)^{-1}d(t)^2. 
@@
   \label{eqn:dexpjgen}
-  \frac{d}{dm}\tJa[m;d] = -(4 \pi r \alpha)^2 \int_{t_{\rm min}}^{t_{\rm max}} \,dt \, 
+  \frac{d}{dm}\tJa[m;d] = -\frac{(4 \pi r \alpha)^2}{\|d\|^2} \int_{t_{\rm min}}^{t_{\rm max}} \,dt \, 
   \left(a\frac{da}{dt}\right)(t-mr)(1+(4\pi r \alpha 
@@
   \label{thm:epjnonoise}
-  Assume the hypotheses of Proposition \ref{thm:norminvexp}, item 3. Then 
+  Assume the hypotheses of Proposition \ref{thm:norminvexp}, item
+  3. Then noting that $\|d\| = \|w_*\|/(4 \pi r)$
 \begin{equation}
@@
   e[m,\aw[m,d];d] 
-= 8 \pi^2 r^2 \alpha^4\int\,dt\,a(t-(m-m_*)r)^4(1+(4\pi r)^2 \alpha^2 
+= \frac{\alpha^4}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^4(1+(4\pi r)^2 \alpha^2 
     a(t-(m-m^*)r)^2)^{-2}w_*(t)^2.
@@
   \label{eqn:anninorm}
-  p[m,\aw[m,d];d] = \frac{1}{2}\int \,dt\,  a(t-(m-m_*)r)^2 \left(1+ (4\pi r)^2\alpha^2
+  p[m,\aw[m,d];d] = \frac{(4\pi r)^2}{2\|w_*\|^2}\int \,dt\,  a(t-(m-m_*)r)^2 \left(1+ (4\pi r)^2\alpha^2
     a(t-(m-m_*)r)^2\right)^{-2}w_*(t)^2.
@@
 \label{eqn:expjnonoise}
-\tJa[m;d] = \frac{\alpha^2}{2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
+\tJa[m;d] = \frac{(4\pi r \alpha)^2}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
   a(t-(m-m_*)r)^2)^{-1}w_*(t)^2. 
@@
   \label{eqn:dexpjnonoise}
-  \frac{d}{dm}\tJa[m;d] = -r \alpha^2 \int \,dt \, 
+  \frac{d}{dm}\tJa[m;d] = -\frac{r (4\pi r \alpha)^2}{\|w_*\|^2} \int \,dt \, 
   \left(a\frac{da}{dt}\right)(t-(m-m_*)r)(1+(4\pi r)^2 \alpha^2 
@@
     min}-mr,t_{\rm max}-mr]\subset [-\tau,\tau]$, with $\tau$ defined
-  in \ref{eqn:taudef}. Therefore, $a(t) = |t|$ $a a'(t) = t$ in the
+  in \ref{eqn:taudef}. Therefore, $a(t) = |t|$, $a a'(t) = t$ in the
   support of the integrand on the RHS of equation
@@
   becomes (after change of integration variable)
+  %%%%%%%%%%%%%%%%%%%%%%%%%
   \begin{equation}
@@
 
-\noindent {\bf Remark:} This is the only error bound in our treatment
-of this problem in which the size of $\alpha$ plays a role. It
-suggests that the estimate of $m$ becomes more accurate in the
-presence of noise for larger $\alpha$. This result is intuitively
+\noindent {\bf Remark:} This is the only error bound on $m$ in which
+the size of $\alpha$ plays a significant role. It
+suggests that error in the estimate of $m$ due to data noise is a
+decreasing function of $\alpha$, at least for small $\alpha$. This result is intuitively
 appealing, and is supported by numerical evidence, as shown below.
 
+%%%%%%%%%%%
+\begin{proposition}
+  \label{thm:wnoiseres}
+  Asumme that
+  \begin{itemize}
+  \item[1. ] $\alpha, \mu> 0$,
+  \item[2. ] $m_* \in M, w_* \in W_{\mu}$,
+  \item[3. ] $d_* = F[m_*]w_*$,
+  \item[4. ] $n \in D$ and $d = d_* + n$.
+  \end{itemize}
+  Set $\eta = \|n\|/\|d_*\|$, and assume further that
+  \begin{itemize}
+  \item[5. ]
+    \begin{equation}
+      \label{eqn:mnoisecond}
+      \eta < \frac{\sqrt{5}-1}{2}, and
+    \end{equation}
+  \item[6. ] $m$ is a stationary point of $\tJa[\cdot;d]$.
+  \end{itemize}
+  Then
+  \begin{equation}
+    \label{eqn:mnoisesuff}
+    |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}.
+  \end{equation}
+\end{proposition}
+
+\begin{proof}
+  From equation \ref{eqn:normsol},
+  \[
+    \aw[m;d](t) = 
+    \left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2)\right)^{-1}\frac{1}{4 \pi r}d(t+mr)
+  \]
+  \[
+    =4\pi r(1 + (4\pi r \alpha t)^2)^{-1} (\frac{1}{4\pi r}
+    w_*(t+(m-m_*)r) + n(t+mr) 
+  \]
+  so
+  \begin{equation}
+    \label{eqn:wres}
+    \aw[m;d](t) - w_*(t) = (4\pi r \alpha t)^2 (1+ (4\pi r \alpha t)^2)^{-1}w_*(t+(m-m_*)r)
+    +w_*(t+(m-m_*)r)-w_*(t) + 4 \pi r (1 + (4\pi r \alpha t)^2)^{-1}n(t+mr)
+  \end{equation}
+\end{proof}
 %  \begin{itemize}
@@
 \append{Numerics and penalty selection via the Discrepancy Principle}
+This appendix describes a method for controlling the penalty
+weight $\alpha$ based on the Discrepancy Principle
+\cite[]{EnglHankeNeubauer,FuSymes2017discrepancy}.
 
@@
 \append{Heisenberg, support, and wavelength}
-
-[THE NEXT BIT MAY OR MAY NOT BE WORTH KEEPING - IT WAS INTERESTING TO
-WORK OUT]
-
 Without further contraints on data or solution, nothing more can be

### 2020-11-23 repro/papers/sseprecond/paper.tex
more

@@
 
-The class of non-point sources investigated here are those confined a hypersurface assumed to contain the locations of physical (point) sources, but not a priori required to have point support.
+The class of non-point sources investigated here are those confined a hypersurface assumed to contain the locations of physical (point) sources, but not {\em a priori} required to have point support.
 These {\em surface extended sources} are able to yield a relatively small residual in the source subproblem \ref{eqn:esis} for more or less arbitrary coefficient array $c$.
@@
 
-The next section gives a precise description of an acoustic version of problem \ref{eqn:esis}, and points out two interesting special cases, the {\em crosswell} and {\em diving wave} configurations.
+The next section gives a precise description of an acoustic version of problem \ref{eqn:esis} for an idealized {\em crosswell} configuration, in which source located on a surface $z=z_s$ generate waves that propagate to to a receiver surface $z=z_r > z_s$.
+In principle, {\em diving waves} can be described with similar mathematics, but with some additional complications, so I do not treat the diving wave case explicitly in this paper.
 The assumptions detailed here imply that the data $d$ of the inverse problems \ref{eqn:esi}, \ref{eqn:esis} have the physical character of {\em transmitted waves}.
@@
 
-%The penalty operator $A$ is chosen to penalize non-physical aspects of the trial source $g$: physical sources are supposed to constitute the null space of $A$.
+\section{Operators and Adjoints}
 
-%In some cases, the resulting  relaxed data fit problem has been observed to be more amenable to solution by local optimization.
-The main goal of this paper is explain why this should be so,  in a particular instance of source-extended FWI.
-I will show that for a specific choice of {\em annihilator} $A$, the optimization problem \ref{eqn:esi} possesses a {\em qualitatively} larger domain of convexity for consistent data ($PL[c]^{-1}g = d, Ag=0$) than does the FWI problem \ref{eqn:ols}, while possessing the same solution.
-
-%For sake of argument (and because almost all work so far has assumed it), adopt the isotropic point source as the model: for the data gather at source positon $\bx_s$, the source takes the form
-%\begin{equation}
-%\label{eqn:ptsrc}
-%f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s). 
-%\end{equation}
-%in which the {\em wavelet} $w(t)$ has whatever structure is necessary for $f$ to lie in the range of the wave operator $L[c]$.
-FWI can also accommodate more complex source models reflecting the nontrivial radiation patterns of actual sources.
-%The data projection operator $P$ involves the restriction of one or more components of the field $u$ to a hypersurface $X_r$.
-The simplest interesting choice of $X_r$ is a hyperplane, say $z=z_r$ (with the convention that $\bx=(x,y,z)$).
-To emulate the limited aperture of any experimental data system, include multiplication by a smooth cut-off or taper function $\phi$.
-Component selection is a special case of multiplication by a matrix $P_c \in \bR^{n_d \times n_s}$, in which $n_d$ is the number of data components and $n_s$ is the dimension of the state vector $u$.
-%\begin{equation}
-%\label{eqn:data}
-%Pu(x,y,t) = \phi(x,y,z_r)P_cu(x,y,z_r,t)
-%\end{equation}
-%in which $P_c$ is a
-
-%Source extension replaces $f(\bx,t;\bx_s)$ with an artificial source $g(\bx,t; \bx_s)$ with more degrees of freedom than merely a single common source wavelet $w(t)$ as in definition \ref{eqn:ptsrc}.
-Assuming that \ref{eqn:ptsrc} is actually correct physics, these extra degrees of freedom have to be suppressed in the optimal solution, via an {\em annihilator} $A$, an operator acting on extended sources $g$, whose null space is precisely the physical sources (of the form \ref{eqn:ptsrc}).
-Extended source inversion replaces the FWI objective with a penalty function
-
-%The way in which degrees of freedom are introduced, and the choice of annihilator, distinguish the various source extension methods, of which there are many - see \cite{HuangNammourSymesDollizal:SEG19}, who introduce the {\em surface source} extension.
-It is this surface source extension that is the topic of this paper.
-It is similar  to the volume extension \cite[]{HuangSymes:Geo18b}, but the source is spread over the source space-time surface, rather than over space at $t=0$.
-%Similar to the space-time and volume extensions \cite[]{HuangSymes:Geo18a,HuangSymes:Geo18b}, the annihilator is given by the {\em localization penalty}:
-%\begin{equation}
-%\label{eqn:msste}
-%A[g](\bx,t;\bx_s) = |\bx-\bx_s|g(\bx,t;\bx_s). 
-%\end{equation} 
-%Therefore an estimate of the the source wavelet is a by-product of the inversion, and does not need to be specified {\em a priori}.
-Like the volume extension (and unlike, for example, Wavefield Reconstruction Inversion \cite[]{LeeuwenHerrmannWRI:13,LeeuwenHerrmann:16}), the additional data volume required is on the same order of size as the data.
-
-%The volume and surface extensions have a well-defined relation to tomographic principles (slope or traveltime, depending on type of annihilator).
-This relation holds regardless of ray geometry (with or without triplication) in transmission configuration.
-
-%Again, nothing is known theoretically about the behaviour of any of these methods for reflection data, and numerical evidence is ambiguous.
-
-\section{Extended Source Modeling}
-For acoustic wave physics, the coefficient vector $c$ is the pair $c=(\kappa,\beta)$ of bulk modulus $\kappa$ and buoyancy (reciprocal density) $\beta$, and the state vector $u=(p,\bv)$ consists of pressure $p$ (a scalar space-time field) and particle velocity $\bv$ (a vector space-time field).
+For acoustic wave physics, the coefficient vector $c$ is the pair $c=(\kappa,\beta)$ of bulk modulus $\kappa$ and density $\rho$, and the state vector $u=(p,\bv)^T$ consists of pressure $p$ (a scalar space-time field) and particle velocity $\bv$ (a vector space-time field).
 The wave operator $L[c]$ is given by the linear acoustics system:
 \begin{equation}
-\label{eqn:awe}
+\label{eqn:aweop}
 L[c]u = 
@@
 \begin{array}{c}
-\partial_t p  + \kappa \nabla \cdot \bv \\
-\partial_t \bv + \beta \nabla p
+\frac{1}{\kappa}\frac{\partial p}{\partial t}  + \nabla \cdot \bv, \\
+\rho\frac{\partial \bv}{\partial t} + \nabla p.
+\end{array}
+\right) 
+\end{equation}
+That is,
+\begin{equation}
+  \label{eqn:awemat}
+  L[c] = \left(
+    \begin{array}{cc}
+      \frac{1}{\kappa}\frac{\partial}{\partial t} & \nabla \cdot \\
+      \nabla & \rho \frac{\partial}{\partial t}
     \end{array}
@@
 
-The space dimension will be denoted by $d$. Most of what follows is valid for any $d >0$. The choice $d=2$ is computationally convenient, if non-physical.
+The space dimension is $d$. Most of what follows is valid for any $d >0$. Examples later in this paper will use $d=2$ for computationally convenience.
 
@@
 
-Real-world sampling is discrete and direction-sensitive.
-This paper ignores such issues and presumes that the sampling operator $P$ simply restricts its argument to a cylindrical space-time hypersurface $\Gamma_r \times \bR$.
-Here $\Gamma_r \subset \bR^d$ is an open subset of a smooth codimension 1 oriented hypersurface.
-To distinguish this output or receiver sampling operator from another similar operator, add the subscript $r$ to the notation, that is, write $P_r$ rather than $P$.
-For technical reasons (and to emulate actual acquisition practice) include multiplication by a cutoff or mute $\phi_r \in C_0^{\infty}(\bR^{d+1})$, and extract only samples of the pressure field:
+The idealized ``crosswell'' configuration involves sources confined to $z=z_s$.
+I will mostly represent sources as constitutive law defects, that is, right-hand sides in the system $L[c]u=f$ of the form $f(\bx,t) = (h_s(x,y,t)\delta(z-z_s), {\bf 0})^T$.
+With the choice $L[c]$ given in \ref{eqn:aweop}, the causal wave system $L[c]u^+=f$ takes the form
+\begin{eqnarray}
+\label{eqn:awep}
+\frac{1}{\kappa}\frac{\partial p^{+}}{\partial t} & = & - \nabla \cdot \bv^{+} +
+h_s \delta(z-z_s), \nonumber \\
+\rho\frac{\partial \bv^{+}}{\partial t} & = & - \nabla p^{+},\nonumber \\
+p^{+} & =& 0 \mbox{ for } t \ll 0,\nonumber\\ 
+\bv^{+} & = & 0 \mbox{ for } t \ll 0.
+\end{eqnarray}
+
+Similarly, the anti-causal wave system $L[c]u^-=f$ takes the form
+\begin{eqnarray}
+ \label{eqn:awem}
+\frac{1}{\kappa}\frac{\partial p^{-}}{\partial t} & = & - \nabla \cdot \bv^{-} +
+h_s \delta(z-z_s), \nonumber \\
+\rho \frac{\partial \bv^{-}}{\partial t} & = & - \nabla p^{-},\nonumber \\
+p^{-} & =& 0 \mbox{ for } t \gg 0,\nonumber\\ 
+\bv^{-} & = & 0 \mbox{ for } t \gg 0.
+\end{eqnarray}
+Sampling the causal pressure field $p^{+}$ on the surface $z=z_r$ defines the causal forward modeling operator $S^{+}_{z_s,z_r}[c]$:
 \begin{equation}
-\label{eqn:recp}
-(P_ru)(\bx,t) = \phi_r(\bx,t)p(\bx,t),\,\bx \in \Gamma_r, t \in \bR. 
+  \label{eqn:fwd}
+  S^{+}_{z_s,z_r}[c]h_s = p^+|_{z=z_r}
 \end{equation}
-A similar sampling operator for the velocity field will also be required below. Denote by ${\bf n}_r$ the outward unit normal field on $\Gamma_r$, and define
-\begin{equation}
-\label{eqn:recv}
-(V_ru)(\bx,t) = \phi_r(\bx,t){\bf n}_r(\bx)\cdot \bv(\bx,t),\,\bx \in \Gamma_r, t \in \bR.
-\end{equation}
+The anti-causal forward modeling operator $S^-_{z_s,z_r}$ has a similar definition.
+The subscript signifies that sources are located on $z=z_s$, the receivers on $z=z_r$.
+It's necessary to include this information in the notation, as versions of $S$ with sources and receivers in several locations will be needed in the discussion below.
 
-Physical sources are modeled as an isotropic constitutive law defects of point support.
-As mentioned earlier, it is simple to accommodate more complex source models.
-Assume that all source lie in an open subset $\Gamma_s \subset \bR^d$ of a smooth orientable codimension 1 hypersurface in $\bR^d$, possibly different from the hypersurface containing $\Gamma_r$.
-Denote by $P_s, V_s$ sampling operators on $\Gamma_s \times \bR$ defined precisely analogously to $P_r$ and $V_r$ (equations \ref{eqn:recp}, \ref{eqn:recv}):
-\begin{equation}
-\label{eqn:recp}
-(P_su)(\bx,t) = \phi_s(\bx,t)p(\bx,t),\,\bx \in \Gamma_s, t \in \bR. 
-\end{equation}
-and similarly for $V_s$.
-Denote by $\delta_{\bx_s}$ the delta distribution on $\Gamma_s$ at $\bx_s \in \Gamma_s$, and by $w \in C_0^{\infty}(\bR)$ a choice of source wavelet.
-The corresponding physical source is then the right-hand side in the acoustic system
-\begin{equation}
-\label{eqn:awe-src}
-L[c]u = P_s^T(w\delta_{\bx_s}) 
-\end{equation}
-With dummy space and time variables, the right hand side above is exactly the form $w(t)e_1\delta(\bx-\bx_s)$ given in the point source definition \ref{eqn:ptsrc}, where , ${\bf e}_1$ the standard basis vector $e_1=(1,0,...,0) \in \bR^{d+1}$.
-
-As explained in the introduction, unless the data $d$ and the coefficient array $c$ are compatible, the source subproblem \ref{eqn:esis} has no low-residual solution of the form $f = P_s^T(w\delta_{\bx_s}) $.
-A larger class of source models is required to fit data.
-A precise description of the class described in the introduction is simply the range of $P_s^T$: that is, for $h \in C^{\infty}(\Gamma_s \times \bR)$, define
-\begin{equation}
-S[c]h = P_rL[c]^{-1}P_s^T h.
-\end{equation}
-$S[c]$ is the surface source extension modeling operator. For this class of sources, the source subproblem becomes 
-\begin{equation}
-\label{eqn:esis-sse}
-\mbox{given } c, \mbox{ choose } h  \mbox{ to minimize } \|S[c]h -d \|^2 + \alpha \|Ah\|^2 
-\end{equation}
-To complete the definition of this problem, a choice of {\em annihilator} $A$ is required. A simple choice is
-\begin{equation}
-\label{eqn:msste}
-Ah(\bx,t) = |\bx-\bx_s|h(\bx,t). 
-\end{equation}
-
-Formally, the solution of the source subproblem is also the solution of the normal equation
-\begin{equation}
-\label{eqn:normal}
-(S[c]^TS[c] + \alpha^2 A^TA) h = S[c]^T d 
-\end{equation}
-This relation holds, and a solution $h$ is guaranteed to exist, if the $\alpha=0$ normal operator $S[c]^TS[c]$ is well-defined and continuous in $L^2(\bar{\Gamma}_s \times \bR)$, and there exists $C>0$ so that
-\begin{equation}
-\label{eqn:coercive}
-\|S[c]h\|^2 \le C|h|^2,\,\,\|S[c]h\|^2 + \alpha^2 \|Ah\|^2 \ge C^{-1}\|h\|^2
-\end{equation}
-for all $h \in L^2(\bar{\Gamma}_s)$. These estimates hold under various conditions. One particular setting in which the least squares problem \ref{eqn:esis} is well-posed is given by
-
-\noindent {\bf Assumption A:} There exists $\theta_*>0$ so that for every ray connecting a point in $\mbox{supp }\phi_s$ with a point in $\mbox{supp }\phi_r$, the angles $\theta_s$ and $\theta_r$ subtended by the ray and the tangent planes of $\Gamma_s$ and $\Gamma_r$ respectively are both greater than $\theta_*$ in absolute value.
-
-The first appendix provides a proof of the following statement:
-
-\begin{theorem} 
-\label{thm:one}
-Suppose that $\log \kappa, \log \beta \in C^{\infty}(\bR^d) \cap L^{\infty}(\bR^d)$, and that Assumption A holds.
-Then the estimates \ref{eqn:coercive} hold for $h \in C^{\infty}_0(\Gamma_s \times \bR)$, hence $S[c]$ extends to $L^2(\bar{\Gamma}_s \times \bR)$ with the same estimates.
-\end{theorem}
-
-The assumptions made in the hypotheses of \ref{thm:one} imply that the inverse wave problems under consideration in this paper are of the transmission type, that is, do not involve reflected waves.
-The reflected wave concept is well-defined only in terms of high-frequency asymptotics: because the coefficients are assumed smooth, there are no material singularities to generate reflections.
-Two configurations of sources and receivers that naturally fit this framework, at least in an ideal sense, are
-\begin{itemize}
-\item the {\em crosswell} configuration, in which $\Gamma_s$ and $\Gamma_r$ are parallel hyperplanes.
-In reality, wells are approximately 1D structures embedded in 3D ambient media, so this configuration does not really model them well.
-However it has become commonplace to use the 2D version of this configuration as a model, assuming that the waves propagate in-plane between the wells  \cite[]{BregChapBai:86,Harr:93,Zhouetal:93,Pratt:99b,Plessix:00}.
-\item the {\em diving wave} configuration, $\Gamma_s$ and $\Gamma_r$ are both subsets of a horizontal plane at the source, respectively receiver, depth (assumed well-defined).
-In marine towed streamer and ocean bottom seismometer surveys, a set of refracted waves arrives earlier than the direct wave in the water column, at sufficient distance from the source.
-The mutes $\phi_s, \phi_r$ isolate these refacted (``diving wave'') arrivals.
-This data is the focus of the ``classic'' period of FWI \cite[]{SirgPratt02,BrendersPratt:06a,BrendersPratt:06b,VirieuxOperto:09,Vigh:10,Vigh13}.
-\end{itemize}
-In both cases, it is possible (though not necessary) that Assumption A is satisfied.
-Note that in the diving wave configuration, the mutes $\phi_S,\phi_r$ are responsible for eliminating direct wave energy, hence enforcing the acute incidence of the waves connecting the admitted source and receiver points.
-The numerical examples presented below will fall into the crosswell category.
-
-Assumption A is not by any means the most general hypothesis under which the source subproblem has a well-behaved solution.
-However weaker hypotheses typically involve multidimensional filters applied to data (and extended sources) to eliminate waves propagating tangent to the source and receiver surfaces, so require more involved implementations.
-
-%Denote by $J_{\alpha}[g,c;d]$ the objective function defined in \ref{eqn:esi}:
-%\begin{equation}
-%\label{eqn:jdef}
-%J_{\alpha}[g,c;d] = \frac{1}{2}( \|PL[c]^{-1}g -d \|^2 + \alpha \|A[g]\|^2) 
-%\end{equation}
-%To compute the objective 
-%\begin{itemize}
-%\item solve the acoustic wave system \ref{eqn:awe} with $g$ as in \ref{eqn:mssur}
-%\item sample at receiver locations and compute mean-square difference with pressure data traces, take mean square
-%\item add scaled mean square of annihilator applied to the source
-%\end{itemize}
-
-%The variable projection method \cite[]{GolubPereyra:03} applied to the objective \ref{eqn:esi} approaches its minimization via a nested algorithm: first minimize \ref{eqn:esi} over the source $g$ for given coefficients $c$, thus making $g = g[c]$ a function of $c$ (inner minimization); then substitute $g[c]$ in $J$ and minimize the resulting function of $c$ (outer minimization).
-The optimal $g[c]$ solves the first order necessary condition for minimization of $J$ over $g$: $\nabla_g J_{\alpha}[g,c;d] = 0$.
-For linear annihilators such as that defined by \ref{eqn:mssur}, $J$ is quadratic in $g$, so the first order necessary condition is also sufficient, and is equivalent to the linear {\em normal equation}:
-
-%where I have written $S[c]=PL[c]^{-1}$ for the operator that maps the (extended) source to the data traces. Substitution of $g[c]$ in $J$ results in the {\em reduced objective}
-%\begin{equation}
-%\label{eqn:jred}
-%\tilde{J}_{\alpha}[c;d] = J_{\alpha}[g[c],c;d].
-%\end{equation}
-%The gradient would appear to involve the derivative of the inner solution $g[c]$ with respect to $c$, computation of which implies solution of additional linear systems.
-However by fortunate accident, the chain rule and the normal equation \ref{eqn:normal}
-%conspire to eliminate all terms involving derivatives of $g[c]$. Consequently the direction derivative of $\tilde{J}_{\alpha}$ in the direction $\delta c$ is
-%\begin{equation}
-%\label{eqn:tildejderiv}
-%D_c\tilde{J}_{\alpha}[c;d]\delta c = \langle D(S[c]g)\delta c,S[c]g-d \rangle|_{g=g[c]}
-%\end{equation}
-%whence
-%\begin{equation}
-%\label{eqn:l2grad}
-%\nabla \tilde{J}_{\alpha}[c;d] = D_c(S[c]g)^T|_{g=g[c]}(S[c]g[c]-d)
-%\end{equation}
-%Let us field-strip the expression on the right.
-The rightmost factor is the data residual, which is a by-product of many methods for the solution of the normal equation \ref{eqn:normal}.
-$D_cS$ is the linearization or Born approximation of $S$, or perturbation with respect to $c$, with $g$ fixed:
-%\[
-%S[c+\delta c]g \approx S[c]g + D_c(S[c]g)\delta c
-%\]
-%That is, $D_c(S[c]g)$ is the object commonly called the Born modeling operator, with source wavefield $g$.
-Its adjoint or transpose is $D_c(S[c]g)^T$ - as is well known, this is the RTM operator (or rather a version of it), computed via the adjoint state method in the time domain.
-Note that the Born and RTM operators need to match, i.e.
-both be derived from the same underlying full waveform modeling operator (in this case acoustodynamics, but the same would apply for any other wave theory).
-
-%The upshot is that the computation of the gradient goes like this:
-%\begin{itemize}
-%\item solve the normal equation \ref{eqn:normal} for $g[c]$
-%\item compute (or retain) the data residual $S[c]g[c]-d$
-%\item reverse-time migrate the data residual, using $g[c]$ to generate the source wavefield in the adjoint state method - the output is the gradient of $\tilde{J}_{\alpha}$.
-%\end{itemize}
-
-%To make this algorithm practical and efficient, effective preconditioning needs to be applied to both inner and outer problems.
-
-\subsection{Time Reversal}
-%In contrast to the outer problem, a preconditioner for the inner problem is available that works regardless of medium smoothness, so for both transmission and reflection (the latter with some caveats).
-%First the transmission case - a sort of rotated crosswell configuration. Idealize the receiver set as a continuum plane $z=z_r$, with $z_r > z_s$ - not essential but simpler.
-
-%NB: the definition of the forward map (and the data) should include a taper $\phi(\bx_r,\bx_s)$ masking out the shot gather $d_r(\cdot,\cdot;\bx_s)$ for each source position $\bx_s$.
-However I will not explicitly mention this operation.
-
-Change notation slightly, and write $S[c]h_s$ rather than $S[c]g$ with $g = h_s \delta(z-z_s)$, as $z_s$ will be presumed ot be the same for all shots.
-The pressure and velocity fields occuring in the computation of $d_r=S[c]h_s$ form the solution of
-\begin{eqnarray}
-\label{eqn:awedata}
-\frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv +
-h \delta(z-z_s), \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
-p & =& 0 \mbox{ for } t<0\nonumber\\ 
-\bv & = & 0 \mbox{ for } t<0 \nonumber\\
- p|_{z=z_r}& = & d_r
-\end{eqnarray}
-Suppose that the recording time interval $[0,T]$ is long enough that $ d_r \approx 0$ for $t > T$.
-Note that $p,\bv$ are {\em incoming} (downgoing) in $z>z_s$.
-Therefore in $z<z_r$ {\em the same fields} approximately solve
-\begin{eqnarray}
-\label{eqn:awetr}
-\frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
-p & =& 0 \mbox{ for } t>T\nonumber\\ 
-\bv & = & 0 \mbox{ for } t>T \nonumber\\
-p|_{z=z_r} &=&  d_r
-\end{eqnarray} 
-The last equation defines $p$ on the entire plane $z=z_r$ for $t<T$, and the previous two define final conditions at $t=T$.
-So the {\em time-reversed} system \ref{eqn:awetr} has unique solutions in $z<z_r, t<T$ and $z>z_r,t<T$.
-The solution in $z<z_r$ is incoming (downgoing) at $z=z_s$, and is {\em the same vector of fields} (approximately) as the solution of \ref{eqn:awedata} in the region $z_s<z<z_r$ between the source and receiver surfaces.
-
-Define $d_s =p|_{z=z_s}$ for the solution of \ref{eqn:awedata}. Then $p,\bv$ is also the solution of
-\begin{eqnarray}
-\label{eqn:awedir}
-\frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
-p &=&0, t< 0,\nonumber\\
-\bv &=&0,t<0,\nonumber\\
-p|_{z=z_s }& = &d_s
-\end{eqnarray}
-in both $z>z_s$ and $z<z_s$, since the pressure field for the solution of \ref{eqn:awedata} is continuous across $z=z_s$.
-Since the same fields solve the two different systems, there must be a relation between the data.
-In fact, the normal component of velocity $v_z$ is discontinuous, and (recalling the $g(\bx,t;\bx_s) = h_s(x,y)\delta(z-z_s)$).
-A calculation using integration by parts shows that $h_s$ is precisely the jump in $v_z$ across $z=z_s$:
-\begin{equation}
-\label{eqn:vjump}
-h_s = [v_z]|_{z=z_s}.
-\end{equation}
-A further simplification occurs if $\kappa, \rho$ are constant near $z=z_s$, then by symmetry
-\begin{equation}
-\label{eqn:vsimp}
-[v_z]|_{z=z_s}=2 \lim_{z \rightarrow z_s^+} v_z
-\end{equation}
-
-With these preparations, define 
-\begin{equation}
-\label{eqn:psinv}
-S[c]^{\dagger}d_r= [v_z]|_{z=z_s}.
-\end{equation}
-where the jump in $v_z$ at $z=z_s$ is computed by
-\begin{itemize}
-\item solving \ref{eqn:awetr} backwards in time,
-\item reading off $d_s=p$ at $z=z_s$,
-\item solving \ref{eqn:awedir} and computing $[v_z]|_{z=z_s}$, or if appropriate simply using \ref{eqn:vsimp}.
-\end{itemize}
- The relations developed in the last few paragraphs may be summarized as follows:
-\begin{equation}
-\label{eqn:appinv}
-d_r = S[c]h_s \,\,\Leftrightarrow \,\, h_s \approx S[c]^{\dagger}d_r
-\end{equation}
-that is, $S[c]^{\dagger}$ is an approximate inverse of $S[c]$.
-
-As I will show in the example section, the approximation may be accurate enough without any further improvement.
-However, a closer examination shows that in fact it is possible to express the relation between $S[c]$ and $S[c]^{\dagger}$ in terms of a change of norm, that is, in terms of a preconditioner, and this relation can be used to considerably accelerate an iterative solution of the normal equation \ref{eqn:normal}.
-
-To see this, first compute the adjoint $S[c]^T$, by a variant of the adjoint state method, in this case a by-product of the conservation of energy. Suppose that 
-$\tilde{p},\tilde{\bv}$ solve
-\begin{eqnarray}
-\label{eqn:aweadj}
-\frac{\partial \tilde{p}}{\partial t} & = & \kappa (-\nabla \cdot \tilde{\bv} + 
- d_r \delta(z-z_r))\nonumber \\
-\frac{\partial \tilde{\bv}}{\partial t} & = & - \beta \nabla \tilde{p},\nonumber \\
-\tilde{p} & =& 0 \mbox{ for } t>T\nonumber\\ 
-\tilde{\bv} & = & 0 \mbox{ for } t>T 
-\end{eqnarray} 
+The adjoint of $S^+_{z_s,z_r}[c]$ can be computed
+by a variant of the adjoint state method, in this case a by-product of the conservation of energy.
+Suppose that $p^-,\bv^-$ solve \ref{eqn:awem} with $h_s\delta(z-z_s)$ replaced by $ h_r \delta(z-z_r)$.
 Then
@@
 0 = 
-\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  
-\frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=T}
+\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
+\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
 -
-\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=0}
+\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
 \]
@@
 = 
-\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)
+\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
 \]
@@
 = 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \tilde {p} + \kappa p \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} \right.
+\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
 \]
@@
 +
-\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \tilde{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \tilde{\bv}}{\partial t} \right)
+\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
 \]
@@
 = 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
- \frac{h_s}{\kappa} \delta(z-z_s)\right) \tilde{p} + p \left(- \nabla \cdot \tilde{\bv} + 
- d_r \delta(z-z_r)\right) \right.
+\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
+ h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
+ h_r \delta(z-z_r)\right) \right.
 \]
@@
 +
-\left.  (- \nabla p) \cdot \tilde{\bv} + \bv \cdot (-\nabla \tilde{p}) \right)
+\left.  (- \nabla p^+) \cdot \bv^- + \bv^+ \cdot (-\nabla p^-) \right)
 \]
@@
 = 
-\int_{0}^{T}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
- \frac{h_s}{\kappa} \delta(z-z_s)\right) \tilde{p} + p \left(- \nabla \cdot \tilde{\bv} + 
- d_r \delta(z-z_r)\right) \right.
+\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
+ h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
+ h_r \delta(z-z_r)\right) \right.
 \]
@@
 +
-\left.  p (\nabla \cdot \tilde{\bv}) + (\nabla \cdot \bv) \tilde{p} \right)
+\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- \right)
 \]
@@
 \[
-0 = \int\,dt\,dx\,dy\, h_s \left(\frac{\tilde{p}}{\kappa}\right)_{z=z_s} + (p  d_r)_{z=z_r}
+0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, h_sp^-|_{z=z_s} +  h_rp^+|_{z=z_r} = \langle h_s, p^-|_{z=z_s}\rangle + \langle h_r, S[c]^+_{z_s,z_r}[c]h_s \rangle
 \]
@@
 \label{eqn:sadj}
- S[c]^T d_r = -\left(\frac{\tilde{p}}{\kappa}\right)|_{z=z_s}
+ S^+_{z_s,z_r}[c]^T h_r = -p^-|_{z=z_s}
 \end{equation}
-Both $S[c]^T$ and $S[c]^{\dagger}$ are defined by time-reversed solutions of the same acoustic system, except with different data.
-System \ref{eqn:aweadj} places $d_r$ as a pressure source on $z=z_r$, with an additional factor of $\kappa$, and its trace on $z=z_s$ must be scaled by $1/\kappa$.
-System \ref{eqn:awetr} employs $d_r$ as a Dirichlet boundary condition, and the trace of its pressure component must be employed as a Dirichlet boundary condition for the forward-in-time problem \ref{eqn:awedir}, after which the jump in the normal velocity is extracted.
+
+The systems \ref{eqn:awep} and \ref{eqn:awem} differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
+$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
+\begin{equation}
+  \label{eqn:trdef}
+  {\cal R} \left(
+    \begin{array}{c}
+      p\\
+      \bv
+    \end{array}
+  \right) =
+  \left(
+    \begin{array}{c}
+      Rp\\
+      -R\bv
+    \end{array}
+  \right)
+\end{equation}
+Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awep} with $h_s\delta(z-z_s)$ replaced by $-Rh_r\delta(z-z_r)$. That is,
+\begin{equation}
+  \label{eqn:trsadj}
+  Rp^-|_{z=z_s} = -S^+_{z_r,z_s}[c]Rh_r
+\end{equation}
+Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that
+\begin{equation} 
+  \label{eqn:trtr}
+  S^+_{z_s,z_r}[c]^T = RS^+_{z_r,z_s}[c]R.
+\end{equation}
+
+\section{Surface source-to-pressure operator}
+
+A particular case of the relation \ref{eqn:trtr} is important in its own right, when $z_r=z_s$. Define surface source-to-pressure operator $\Lambda_{z_s}$ by
+\begin{equation}
+  \label{eqn:ntoddef}
+  \Lambda_{z_s} = S^+_{z_s,z_s}.
+\end{equation}
+Then \ref{eqn:trtr} reads
+\begin{equation}
+  \label{eqn:ntodtr}
+  \Lambda_{z_s}^T = R\Lambda_{z_s}R.
+\end{equation}
+This operator is closely related to the operator mapping $\lim_{z\rightarrow z_s^+} \bv_z^+$ to $\lim_{z\rightarrow z_s^+}  p^+$, often called the {\em Neumann-to-Dirichlet} operator.
+The connection is explained below.
+
+The quadratic form defined by $\Lambda_{z_s}$ has fundamental physical significance.
+Suppose that the surface source $h_s$ vanishes outside of a bounded time interval $[t_{\rm min},t_{\rm max}]$.
+Then the causal acoustic field $(p^+,\bv^+)$, solution of the system \ref{eqn:awep}, vanishes for $t<t_{\rm min}$.
+Define the total acoustic energy $E(t)$ at time $t$ by
+\begin{equation}
+  \label{eqn:defae}
+  E(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^+)^2}{\kappa} + \rho |\bv^+|^2\right)
+\end{equation}
+Then since $E(t_{\rm min})=0$,
+\[
+  E(t_{\rm max}) = \int_{t_{\rm min}}^{t_{\rm max}} \frac{dE}{dt}
+\]
+\[
+  = \int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\, \left(p^+ \frac{1}{\kappa}\frac{\partial p^+}{\partial t} + \bv^+ \cdot \rho \frac{\partial \bv^+}{\partial t}\right)
+\]
+\[
+  =\int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\,\left(p^+(-\nabla \cdot \bv^{+} + h_s\delta(z-z_s)) + \bv^+ \cdot \rho \frac{\partial \bv^+}{\partial t}\right)
+\]
+\[
+  \int_{t_{\rm min}}^{t_{\rm max}}\int\,d\bx\, \left(\left(\nabla p^+\rho \frac{\partial \bv^+}{\partial t}\right) + p^+h_s\delta(z-z_s)\right)
+\]
+\[
+  =  \int_{t_{\rm min}}^{t_{\rm max}}\int\,dxdy h_s p^+|_{z=z_s} = \langle h_s, \Lambda_{z_s} h_s \rangle.
+\]
+Since the source $h_s\delta(z-z_s)$ no longer acts on the fluid for $t >t_{\rm max}$, $E_{\rm tot}= E(t_{\rm max})$ is the total energy transferred from the source to the fluid, and
+\begin{equation}
+  \label{eqn:surfenergy}
+  E_{\rm tot} = \langle h_s, \Lambda_{z_s} h_s \rangle.
+\end{equation}
+We can conclude from identity \ref{eqn:surfenergy} that the symmetric part of $\Lambda_{z_s}$ is positive semidefinite:
+\begin{equation}
+  \label{eqn:lamsd}
+  \frac{1}{2}(\Lambda_{z_s}^T + \Lambda_{z_s}) \ge 0.
+\end{equation}
+
+Observe that the causal acoustic system \ref{eqn:awep} implies directly that
+\begin{equation}
+  \label{eqn:sourcevjump}
+  h_s= [v_z]|_{z_s}.
+\end{equation}
+
+\section{Progressing Waves and Symbols}
+
+The symbol $\sigma(L[c])$ of a matrix operator $L[c]$ is a complex scalar-valued matrix of the same size. In terms of its product with an arbitrary vector $\bu \in \bR^{d+1}$, it is
+\[
+  \sigma(L[c]) \bu = e^{-i(\omega t + \bk \cdot \bx)} (L[c]e^{i(\omega t + \bk \cdot \bx)}\bu).
+\]
+For $L[c]$ given by the definition \ref{eqn:awemat}, obtain
+\begin{equation}
+  \sigma(L[c])(t,\bx,\omega,\bk) = i\left(
+    \begin{array}{cc}
+      \frac{\omega}{\kappa(\bx)} & \bk^T \\
+      \bk & \rho(\bx) \omega)
+    \end{array}
+  \right)
+\end{equation}
+
+The characteristic equation of $L[c]$ is $\det \sigma(L[c]) = 0$. The significance of the characteristic equation is its role in constraining the phase $\psi(\bx)$ in the progressing wave ansatz,
+\begin{equation}
+  \label{eqn:go}
+  \bu(\bx,t) = e^{i\omega (t-\psi(\bx))} \ba (\bx).
+\end{equation}
+Applying $L[c]$ to this ansatz, obtain
+\[
+  L[c]\bu(\bx,t)=i \omega
+  \left(
+    \begin{array}{c}
+      \frac{1}{\kappa}a_p -  \nabla \psi \cdot \ba_v \\
+      -a_p \nabla \psi + \rho \ba_v
+    \end{array}
+  \right) + ...
+\]
+\[
+  = i \omega \sigma(L[c])(t,\bx,1,- \nabla \psi(\bx))ba + ...
+\]
+where ``$...$'' denotes terms of lower order in $\omega$. Thus $\bu$ as defined in equation \ref{eqn:go} is a high-frequency asymptotic solution, $L[c]\bu \approx 0$, if
+\begin{itemize}
+\item
+  \begin{equation}
+    \label{eqn:eik}
+    \det(\sigma(L[c])(t,\bx,1,-\nabla \psi(\bx))) = 0,
+  \end{equation}
+  (eikonal equation), and
+\item
+  $\ba$ is a null vector of $\sigma(L[c])(t,\bx,1,-\nabla \psi)$.
+\end{itemize}
+Explicitly, the eikonal equation reads
+\[
+  0 = \rho^2\left(\frac{\rho}{\kappa}-|\nabla \psi|^2\right),
+\]
+equivalent to the familar form
+\begin{equation}
+  \label{eqn:usualeik}
+  c|\nabla \psi| = 1,
+\end{equation}
+in which $c = \rho/\kappa$ is the wave velocity. Solution via the method of characteristics gives
+\[
+  \psi(\bx(t)) = \psi(\bx(0)) + t
+\]
+for a ray $t \mapsto \bx(t)$ satisfying
+\[
+  \frac{d\bx}{dt}(t) = -c(\bx)^2\nabla \psi(\bx(t))
+\]
+Finally, $\ba$ is a null vector of $\sigma(L[c])(\cdot,\cdot,1,-\nabla \psi)$ iff
+\begin{equation}
+  \label{eqn:godton}
+  \ba_v = \frac{a_p}{\rho}\nabla \psi
+\end{equation}
+The scalar $a_p$ evolves along the ray according to the transport equation, which I will not derive here; suffice it to say that $a_p(t) = a_p(0) \gamma(t)$, where $\gamma > 0$ is smooth and $\gamma(0)=1$.
+
+Suppose that $\bx_0 = (x,y,z_s)^T$ and $(\xi_x,\xi_y) \in \bR^2$ satisfies $c(\bx_0)|(\xi_x,\xi_y)|<1$.
+Then there is a neighborhood $\Omega$ of $\bx_0$ and $\xi_z \in C^{\infty}(\Omega), \xi_z>0$, so that $\bxi^{\pm}(\bx)=(\xi_x,\xi_y,\pm \xi_z(\bx))$ satisfy $c(\bx)|\bxi^{\pm}(\bx)|=1$ for $\bx \in \Omega$.
+
+A standard argument shows that there exists $\epsilon > 0$ for which a solution $\psi^{\pm}$ of the eikonal equation exists in $\Omega \cap \{\bx: |z-z_s|<\epsilon\}$ with
+\begin{equation}
+  \label{eqn:eikic}
+  \psi^{\pm}(x,y,z_s,t) = t -x \xi_x - y \xi_y, \, \frac{\partial \psi}{\partial z}(x,y,z_s,t) = \pm \xi_z(x,y,z_s).
+\end{equation}
+Let $\chi \in C_0^{\infty}, \chi(\bx_0)=1$, and set $p_0^{\pm}(x,y,t) = \chi(x,y,z_s)e^{i\omega(t-x\xi_x - y\xi_y)}$. Define $p_{\rm go}^{\pm} \in C^{\infty}(\Omega \cap \{\bx: |z-z_s|<\epsilon\})$ by
+\[
+  p_{\rm go}^{\pm}(\bx,t) = a^{\pm}_p(\bx) e^{i\omega (t-\psi^{\pm}(\bx))}.
+\]
+where $a^{\pm}_p$ solves the transport equation with $a^{\pm}_p|_{z=z_s} = \chi$. Set
+\[
+  \bv^{\pm}_{\rm go}(\bx,t) = \frac{a^{\pm}_p(\bx)}{\rho(\bx)}\nabla \psi^{\pm}(\bx) e^{i\omega (t-\psi^{\pm}(\bx))}
+\]
+Then $\bu^{\pm}_{\rm go} = (p^{\pm}_{\rm go}, \bv^{\pm}_{\rm go})$ is an asymptotic solution of $L[c]\bu = 0$ with $p^{\pm}_{\rm go}|_{z=z_s} = \chi e^{i\omega(t-x\xi_x - y\xi_y)}$.
+
+From equations \ref{eqn:godton} and \ref{eqn:eikic},
+\[
+  v_{z,{\rm go}}^{\pm}(x,y,z_s,t) = \pm \frac{\chi(x,y,t)}{\rho(x,y,z_s)}\xi_z  e^{i\omega(t-x\xi_x - y\xi_y)}
+\]
+\begin{equation}
+  \label{eqn:asymdton}
+= \pm \frac{\chi(x,y,t)}{\rho(x,y,z_s)c(x,y,z_s)}\sqrt{1-c(x,y,z_s)^2(\xi_x^2+\xi_y^2)}
+\end{equation}
+
+Integration over frequency produces a progressing wave traveling (at least near $z=z_s$) downwards, that is, increasing $z$ with increasing $t$, for the choice of sight $\pm = +$, and upwards, that is, decreasing $z$ with increasing $t$, for the other sign.
+Therefore a causal progressing wave with $p$ given on $z=z_s$ is obtained by synthesizing distorted plane waves as just described with $\pm = +$ in $z>z_s$, $\pm=-$ in $z<z_s$.
+From equation \ref{eqn:asymdton}, the normal components of $\bv$ have limits equal in magnitude but of opposite sign as $z\rightarrow 0^{\pm}$, up to a lower frequency error.
+That is, the jump in $\bv$ satisfies
+\begin{equation}
+  \label{eqn:asymjump}
+  [v_z]\approx \lim_{z\rightarrow 0^+} v_{z}
+\end{equation}
+where $\approx$ denotes ``up to a low frequency error'', from here on.
+
+This construction also implies an approximation of the source-to-pressure operator as a pseudodifferential operator.
+Choose a partition of unity in $z=z_s$, for which the support of every member is a domain $\Omega$ for which the construction above can be carried out: this implies that there be a uniform $\delta > 0$ so that
+\begin{equation}
+  \label{eqn:nongraze}
+  1>\delta + c(x,y,z_s)\sqrt{\xi_x^2+\xi_y^2}
+\end{equation}
+for all $(x,y,t,1,\xi_x,\xi_y) \in WF(p_0)$. Denote by $\Gamma$ the subset of $(x,y,t,k_x,k_y,\omega) \in T^*(\bR^3)$ for which $x,y,\xi_{xy}=k_{xy}/\omega$ satisfy the inequality \ref{eqn:nongraze}.
+
+Then near $z=z_s$, $p$ is approximately a sum of geometric optics solutions of the type displayed above, integrated over $\omega,k_x,k_y$, with $k_{x,y} = \omega \xi_{x,y}$.  Thus
+\[
+  p(x,y,z_s,t) \approx \frac{1}{2}\int_{\Gamma(x,y)} \,d\omega\,dk_x\,dk_y \hat{[v_z]}\rho(x,y,z_s)c(x,y,z_s)
+\]
+\[
+   \times \left(1-c^2(x,y,z_s)\left(\left(\frac{k_x}{\omega}\right)^2 + \left(\frac{k_y}{\omega}\right) ^2\right)\right)^{-1/2}e^{\i(\omega t + k_x x+ k_y y)}
+\]
+Since $p$ is the pressure component of a solution of \ref{eqn:awep}, equation \ref{eqn:sourcevjump} implies that
+\begin{equation}
+  \label{eqn:psidodton}
+  \Lambda_s h_s(x,y,t) \approx \frac{1}{2}\int_{\Gamma(x,y)} \,d\omega\,dk_x\,dk_y
+  \lambda_s(x,y,t,k_x,k_y,\omega) \hat{h_s}(k_x,k_y,\omega) e^{\i(\omega t + k_x x+ k_y y)}
+\end{equation}
+with
+\begin{equation}
+  \label{eqn:symboldton}
+  \lambda_s(x,y,t,k_x,k_y,\omega) =\frac{1}{2}\rho(x,y,z_s)c(x,y,z_s)\left(1-c^2(x,y,z_s)\left(\left(\frac{k_x}{\omega}\right)^2 + \left(\frac{k_y}{\omega}\right) ^2\right)\right)^{-1/2}
+\end{equation}
+
+Restricted to $\Gamma$, $\lambda \in S^0_{0,1}(\bR^2)$ is real-valued, hence defines an essentially symmetric operator. That is,
+\begin{equation}
+  \label{eqn:approxsymm}
+  \Lambda_s^T \approx \Lambda_s.
+\end{equation}
+
+\section{Parametrix via Time-reversal}
+
+Both $S[c]^T$ and $S[c]^{\dagger}$ are defined by time-reversed solutions of the same acoustic system, except with different data:
+\begin{itemize}
+\item system \ref{eqn:aweadj} inputs $d_r$ as a pressure source on $z=z_r$, with an additional factor of $\kappa$, and outputs the trace of the reverse-time propagated field on $z=z_s$, scaled by $1/\kappa$, to obtaint $S[c]^Td_r$;
+\item system \ref{eqn:awetr} inputs $d_r$ as a Dirichlet boundary condition, and the trace of its pressure component must be employed as a Dirichlet boundary condition for the forward-in-time problem \ref{eqn:awedir}, after which the jump in the normal velocity is extracted.
+\end{itemize}
 
@@
 \end{quote}
-Call the operator that produces the jump in normal velocity from Dirichlet pressure data $\Lambda[c]_s$ for the source surface $z=z_s$, $\Lambda[c]_r$ for the receiver surface $z=z_r$ (in the math literature, this is the ``Dirichlet-to-Neumann operator'', more or less).
-The discussion so far can be summarized as follows:
+Call the operator that produces the jump in normal velocity from Dirichlet pressure data $\Lambda[c]_s$ for the source surface $z=z_s$ and forward in time, $\Lambda[c]_r$ for the receiver surface $z=z_r$ backwards in time (in the math literature, these are ``Dirichlet-to-Neumann operators'', more or less).
+
+\cite{HouSymes:EAGE16} demonstrated a very similar preconditioner for Least Squares Migration, also for its subsurface offset extension \cite[]{HouSymes:16}, motivated by \cite{tenKroode:12}.
+These constructions all involve the Dirichlet-to-Neumann operator.
+This concept also turns up in hidden form in the work of Yu Zhang and collaborators on true amplitude migration \cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,Zhang YuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.
+
+\section{Efficient Preconditioned CG}
+
+The discussion so far can be summarized as follows: an efficient pseudoinverse for $S[c]$ is the weighted adjoint operator
 \begin{equation}
 \label{eqn:wadj}
-S[c]^{\dagger} = W[c]_m^{-1}S[c]^TW[c]_d 
+S[c]^{\dagger} = W[c]_m^{-1}S[c]^TW[c]_d,
 \end{equation}
@@
 \label{eqn:weights}
-W[c]_m = \kappa^{-1},\,\, W[c]_d = \kappa^{-1}\Lambda[c]_r
+W[c]_m = \kappa^{-1}\Lambda[c]^{-1}_s,\,\, W[c]_d = \kappa^{-1}\Lambda[c]_r
 \end{equation}
-in which ``$\kappa$'' means the operator ``multiply by $\kappa$''. This means that $S[c]^{\dagger}$ {\em is} the adjoint of $S[c]$ with respect to weighted norms 
+Here ``$\kappa$'' means the operator ``multiply by $\kappa$''.
+
+The statement that $S[c]^{\dagger}$ is the weighted adjoint of $S[c]$ means that $S[c]^{\dagger}$ {\em is} the adjoint of $S[c]$ with respect to weighted norms 
 \begin{itemize}
@@
 \end{itemize}
-Since $S[c]^{\dagger}S[c] \approx I$ as shown above, $S[c]$ is approximately unitary in the Hilbert spaces of source and data traces with norms defined by the weighting operators $W_m$ and $W_d$ respectively.
-Therefore a Krylov space method employing these norms will converge rapidly.
+Since $S[c]^{\dagger}S[c] \approx I$ in subspaces of the domain space, as shown above, $S[c]$ is approximately unitary in (subspaces of) the Hilbert spaces of source and data traces with norms defined by the weighting operators $W_m$ and $W_d$ respectively.
+Therefore a Krylov space method employing these norms will converge rapidly, at least for the well-determined components of the solution.
 
-\cite{HouSymes:EAGE16} demonstrated a very similar preconditioner for Least Squares Migration, also for its subsurface offset extension \cite[]{HouSymes:16}, motivated by \cite{tenKroode:12}.
-These constructions all involve the Dirichlet-to-Neumann operator.
-This concept also turns up in hidden form in the work of Yu Zhang and collaborators on true amplitude migration \cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,ZhangYuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.
+The most convenient arrangement the CG algorithm taking advantage of the structure \ref{eqn:wadj} is the {\em Preconditioned CG}.
+Since the coefficient array $c$ will remain unchanged during the iteration, write $S$ instead of $S[c]$, etc.
+
+Allowing that the fit error will be measured by the data space norm, the least squares problem to be solved is not just $Sh \approx d$, but a regularized version:
+\begin{equation}
+  \label{eqn:einv}
+  \mbox{minimize}_h \|Sh-d\|^2_d + \alpha^2 \|Ah\|^2_m
+\end{equation}
+
+[The modified data space norm $\|d\|_d^2 = \langle d, W_d d\rangle$ has physical meaning: for acoustics, it is proportional to the power transmitted through the surface $z=z_r$.]
+
+The minimizer of the objective defined in equation \ref{eqn:einv} solves the normal equation
+\begin{equation}
+  \label{eqn:norm}
+  (S^{\dagger}S + \alpha^2 A^{\dagger}A)h = S^{\dagger}d 
+\end{equation}
+
+where the weighted adjoint $S^{\dagger}$ has already been defined in equation \ref{eqn:wadj}, and $A^{\dagger}$ is the adjoint of $A$ in the weighted model space norm defined by $W_m$, namely
+\begin{equation}
+  \label{eqn:aadj}
+  A^{\dagger} = W_m^{-1}A W_m.
+\end{equation}
+
+Introducing these definitions in the normal equation \ref{eqn:norm},
+\begin{equation}
+  \label{eqn:norm1}
+  W_m^{-1}(S^TW_dS + \alpha^2 A^TW_mA)h = W_m^{-1}S^TW_md 
+\end{equation}
+Since $W_m$ is self-adjoint and positive definite, the common factor on both sides of \ref{eqn:norm1} can be re-written as
+
+write $S^*, A^*$ for the adjoints with the original (Euclidean) inner product in the domains but the weighted inner product in data space:
+\begin{equation}
+  \label{eqn:normpart}
+  Nh = (S^*S + \alpha^2 A^*A)h = S^*d 
+\end{equation}
+using the ``partly weighted'' adjoints
+\begin{eqnarray}
+  \label{eqn:sadj}
+  S^* &=& S^T W_d,\\
+  A^* &=& A^T W_m.
+\end{eqnarray}
+The Preconditioned Conjugate Gradient (``PCG'') algorithm for solution of equation \ref{eqn:normpart} with preconditioner $W_m$ is usually written as Algorithm 1.
+
+\begin{algorithm}[H]
+\caption{Preconditioned Conjugate Gradient Algorithm, Standard Version}
+\begin{algorithmic}[1]
+\State Choose $h_0=0$ 
+  \State $r_0 \gets S^*d$
+  \State $p_0 \gets W_m^{-1}r_0$
+  \State $g_0 \gets p_0$
+  \State $q_0 \gets Np_0$
+  \State $k \gets 0$
+  \Repeat
+  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle p_k,q_k\rangle}$
+  \State $h_{k+1} \gets h_k + \alpha_k p_k$
+  \State $r_{k+1} \gets r_k - \alpha_kq_k$
+  \State $g_{k+1} \gets W_m^{-1}r_{k+1}$
+  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
+  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
+  \State $q_{k+1} \gets Np_{k+1}$
+  \State $k \gets k+1$
+  \Until{Error is sufficiently small, or max iteration count exceeded} 
+\end{algorithmic}
+\end{algorithm}
+
+The operator whose eigenvalue spectrum determines the speed of convergence here is the preconditioned normal operator
+\begin{equation}
+  \label{eqn:pno}
+  W_m^{-1}N = S^{\dagger}S +\alpha^2 A^{\dagger}A
+\end{equation}
+This isn't just an approximate identity on a subspace, as $S^{\dagger}S$ is.
+The addition of the penalty term affects the spectrum, pushing parts of it away from $1$.
+I have not yet analyzed this effect, however for the ``in-aperture'' part of the model space,
+\[
+  W_m^{-1}N \approx I +\alpha^2 W_m^{-1}A^TW_mA
+\]
+For small $\alpha$, this operator is a perturbation of the spectrum of the identity, so convergence should be rapid.
+
+In the preceding section, I showed that effective choices for the weight operators are $W_m = \Lambda_s^{-1}$ and $W_d = \Lambda_r$, which are resepctively the ``Neumann-to-Dirichlet'' operator mapping pressure to normal particle velocity on the source surface $\{z=z_s\}$, and the ``Dirichlet-to-Neumann'' operator mapping pressure to normal particle velocity on the receiver surface.
+Application of these operators are expensive and/or awkward, so should be avoided if at all possible.
+
+A very important point: the weighted ``partial'' adjoint $S^*=S^TW_d$ appears {\em only} in combination with $S$ in the algorithm listed above, {\em except} for the very first line.
+That is, one can calculate $W_dS$ followed by $S^T$, rather than $S$ followed by $S^TW_d$, provided that you can take care of the first line.
+$W_dS$ For any variant of elasticity, $W_d$ relates two combinations of dynamical field components - for acoustics, it returns the normal velocity for a given pressure field on the receiver surface.
+In the solution of the dynamical equations (for instance \ref{eqn:awe}), all components are calculated.
+Therefore obtaining the output of $W_d$ amounts to extracting the appropriate field combination.
+For acoustics, simply record the normal velocity rather than the pressure: then you have computed $W_dS$.
+Therefore no additional expensive computations are required.
+It is only the very first instance of $S^*$, on the first line, where an actual computation of $W_d$ is required, unless cheap workarounds like that explained by \cite{HouSymes:15} are available.
 
@@
 
-\append{Proof of Theorem 1}
-Following the notation of \cite{BaoSy:91b}, denote by $\gamma_s$ ($\gamma_r$)the closed conic subset of $T^*(\Gamma_s\times \bR)|_{\mbox{supp } \phi_s}$ ($T^*(\Gamma_r\times \bR)|_{\mbox{supp } \phi_r}$) consisting of tangential components of null bicharacteristics passing over points of $\mbox{supp }\phi_s$ and $\mbox{supp }\phi_r$.
-That is, the ray with initial condition $(\bx,t,\xi,\omega) \in T^*(\Gamma_s\times \bR)|_{\mbox{supp } \phi_s}$ passes over $\mbox{supp }\phi_r$ if and only if $(\bx,t,\xi,\omega) \in \gamma_s$.
-Since each such ray is paired with the ray with the same spatial projection traversed backwards, it is also the case that $(\bx,t,\xi,\omega) \in T^*(\Gamma_r\times \bR)|_{\mbox{supp } \phi_r}$ passes over $\mbox{supp }\phi_s$ if and only if $(\bx,t,\xi,\omega) \in \gamma_r$.
-
-Define $s=\sqrt{\kappa/\rho}$ to be the wave slowness (reciprocal velocity), and
-\[
-c(\bx,t,\xi,\omega) = s(\bx)^2\omega^2-|\xi|^2
-\]
-to be the characteristic polynomial of the acoustic wave system, and $C=c^{-1}(0)$ the characteristic variety.
-The angle condition in Assumption A implies the existence of a conic neighborhood $\tilde{\gamma}_s$ of $\gamma_s$, so that for  $(\bx,t,\xi,\omega) \in \tilde{\gamma}_s$, the polynomial $\tau \mapsto c(\bx,t,\xi + \tau{\bf n}_s(\bx),\omega)$ has two distinct roots, $\pm \tau_s$.
-Let $\gamma_{1,s}$ be a conic neighborhood of the complement $\tilde{\gamma}_s^C$, chosen so that $\gamma_{1,s} \cap \gamma_s = \emptyset$.
-Let $\gamma_s' = \Pi_s^*(\gamma_{1,s})$, where $\Pi_s^*$ is the pullback map: $T^*(\Gamma_s\times \bR) \rightarrow T^*(\bR^{d+1})$.
-Similarly, construct $\gamma_r'$ and $\gamma_{1,r}$.
-
-The null bicharacteristic through $(\bx,\xi,\omega,\xi) \in \gamma_r' \cap C$ does not pass over $\mbox{supp } \phi_s$ by construction, hence the RHS of the system $L[c]u=P_r^Th$ is microlocally of class $H^1$ (in fact, smooth) near this bicharacteristic,  thanks to the microlocal propagation of singularities theorem \cite[]{BealsReed:82}: that is, $p \in H^1_{ml}(\gamma_r')$.
-
-Need to know $p \in L^2$.
-Restart: consider $F_{r,s}$ mapping Cauchy data at $t=T$ by reverse time to $P_{r,s} p$, $\mbox{supp }\phi_s \cup \mbox{supp }\phi_r \subset \{t<T\}$.
-Only relevant rays are continuations of those in $\gamma_r$.
-Represent $S$ as composition: $S=F_rF_s^T$.
-$F_s$ $L^2$-bounded $\Rightarrow$ for $u, v \in C_0^{\infty}$,
-\[
-|\langle F_s u, v \rangle| = \le C\|u\|\|v\|=|\langle u, F_s^T v\rangle|
-\]
-Indep arg establishes that $F_s^Tv$
-
-%for which $(\bx,t,\xi\pm \tau_s{\bf n}_s(\bx),\omega)$ is the initial condition on $\Gamma_s \times \bR$ of a ray passing over $\mbox{supp }\phi_r$.
-Choose a conic neighborhood $\tilde{\gamma}_s$ of $\gamma_s$
-
-Denote by
-
-%,  $T^*(\Gamma_r\times \bR)|_{\mbox{supp \phi_r}}$ for which Note that the angle condition mentioned in Theorem \ref{thm:one} implies that if $(\bx,t,\xi,\omega) \in \gamma' $ and $\bx \in \Gamma_s$ or $\Gamma_r$, then $\omega$ is a simple root of are distinct, and even a lower bound on the difference proportional to $\epsilon$.
-It also implies that $\gamma'$ is disjoint from a neighborhood of the normal bundles of $\Gamma_s \times \bR$, $\Gamma_r \times \bR$.
-
-\append{Transpose of Full Acoustic Modeling Operator}
-
-In this section I compute the adjoint operator form modeling with both pressure and velocity source inputs and pressure and velocity trace outputs. The forward modeling operator is defined by solving
-\begin{eqnarray}
-\label{eqn:fawedata}
-\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
-h_s \delta(z-z_s)), \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta ( \nabla p + l_s {\bf e}_z\delta(z-z_s)),\nonumber \\
-p & =& 0 \mbox{ for } t<0\nonumber\\ 
-\bv & = & 0 \mbox{ for } t<0 \nonumber\\
-\end{eqnarray}
-The new elements here are the point load time function $l_s(\bx,t;\bx_s)$, and the vertical unit vector ${\bf e}_z=(0,1)^T$ for 2D, $=(0,0,1)^T$ for 3D.
-The forward map ${\bf F}$ is defined as implicit in the discussion of the Intrinsically Preconditioned CG algorithm: let ${\bf h}_s=(h_s,l_s)^T$, then
-\begin{equation}
-\label{eqn:ffwd}
-{\bf F}{\bf h}_s =\left\{ \left[
-\begin{array}{c}
-p(\bx,t;\bx_s)\\
-v_z(\bx,t;\bx_s)
-\end{array}
-\right]_{z=z_r}\right\}
-\end{equation}
-
-Let $\tilde{\bf h}_r=(\tilde{h}_r,\tilde{l}_r)^T$ be a function on $z=z_r$, 
-and 
-$\tilde{p},\tilde{\bv}$ solve the backwards-in-time boundary value problem:
-\begin{eqnarray}
-\label{eqn:faweadj}
-\frac{\partial \tilde{p}}{\partial t} & = & \kappa (-\nabla \cdot \tilde{\bv} + 
- \tilde{h}_r \delta(z-z_r))\nonumber \\
-\frac{\partial \tilde{\bv}}{\partial t} & = & \beta (\nabla \tilde{p} + \tilde{l}_r{\bf e}_z\delta(z-z_r),\nonumber \\
-\tilde{p} & =& 0 \mbox{ for } t>T\nonumber\\ 
-\tilde{\bv} & = & 0 \mbox{ for } t>T 
-\end{eqnarray} 
-Then
-\[
-0 = 
-\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  
-\frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=T}
--
-\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)|_{t=0}
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \tilde {p} +  p \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} \right.
-\]
-\[
-+
-\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \tilde{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \tilde{\bv}}{\partial t} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
- h_s \delta(z-z_s)\right) \tilde{p} + p \left(- \nabla \cdot \tilde{\bv} + 
- \tilde{h}_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  (- \nabla p + l_s{\bf e}_z\delta(z-z_s)) \cdot \tilde{\bv} + \bv \cdot (-\nabla \tilde{p} + \tilde{l}_r{\bf e}_z\delta(z-z_r)) \right)
-\]
-\[
-= 
-\int_{0}^{T}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv \tilde{p} + 
- h_s\tilde{p}|_{z=z_s} - p\nabla \cdot \tilde{\bv} + 
- \tilde{h}_r p|_{z=z_r}\right) \right.
-\]
-\[
-+
-\left.  p (\nabla \cdot \tilde{\bv}) + l_s \tilde{v}_z|_{z=z_r}+ (\nabla \cdot \bv) \tilde{p}  + l_r  v_z|_{z=z_r}\right)
-\]
-after integration by parts in the last two terms. Most of what is left cancels, leaving 
-\[
-0 = \int\,dt\,dx\,dy\, (h_s \tilde{p})_{z=z_s} + (p  \tilde{h}_r)_{z=z_r} + (l_s\tilde{v}_z)_{z=z_s} + (v_z \tilde{l}_r)_{z=z_r},
-\]
-whence
-\begin{equation}
-\label{eqn:fadj}
- {\bf F}^T {\bf h}_r = -\left[
-\begin{array}{c}
-\tilde{p}(\bx,t;\bx_s)\\
-\tilde{\bf v}(\bx,t;\bx_s)
-\end{array}
-\right]_{z=z_s}
-\end{equation}
-
-\append{Transpose of Cauchy Data to Trace operator}
-
-For $u_0=(p_0,\bv_0) \in (C_0^{\infty}(\bR^d))^{d+1}$, define ${\bf F}u_0 = (P_r p, P_r {\bf n}_r\cdot \bv)$ where $u=(p,\bv)$ solves
-\begin{eqnarray}
-\label{eqn:fawedata}
-\frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv, \nonumber \\
-\frac{\partial \bv}{\partial t} & = & - \beta \nabla p ,\nonumber \\
-p(\cdot,T) & =& p_0 \\ 
-\bv(\cdot,T) & = & \bv_0\\
-\end{eqnarray}
-(Note: assumed that $\mbox{supp }\phi_r \subset \bR^d \times (0,T)$.)
-
-Let $\tilde{\bf h}_r=(\tilde{p}_r,\tilde{v_{nr}})^T \in (C_0^{\infty}(\Gamma_r \times (0,T)))^2$ 
-and 
-$\tilde{u}=(\tilde{p},\tilde{\bv})$ solve 
-\begin{eqnarray}
-\label{eqn:faweadj}
-\frac{\partial \tilde{p}}{\partial t} & = & \kappa (-\nabla \cdot \tilde{\bv} + 
- \tilde{p}_r \phi_r \delta_{\Gamma_r \times \bR})\nonumber \\
-\frac{\partial \tilde{\bv}}{\partial t} & = & \beta (-\nabla \tilde{p} + \tilde{v}_{nr}{\bf n}_r \phi_r \delta_{\Gamma_r \times \bR},)\nonumber \\
-\tilde{p} & =& 0 \mbox{ for } t<0\nonumber\\ 
-\tilde{\bv} & = & 0 \mbox{ for } t<0 
-\end{eqnarray} 
-Then
-\[
-0 = 
-\left(\int\, dx\,dy\,dz\, \frac{p_0 \tilde{p}}{\kappa} +  
-\frac{\bv_0 \cdot \tilde{\bv}}{\beta} \right)|_{t=T}
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \tilde{p}}{\kappa} +  \frac{\bv \cdot \tilde{\bv}}{\beta} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \tilde {p} +  p \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} \right.
-\]
-\[
-+
-\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \tilde{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \tilde{\bv}}{\partial t} \right)
-\]
-\[
-= 
-\int_{0}^{T} \,dt\, \int\, dx\,dy\,dz\, \left((- \nabla \cdot \bv) \tilde{p} + p (- \nabla \cdot \tilde{\bv})\right.
-\]
-\[
-\left.- \nabla p  \cdot \tilde{\bv} + \bv \cdot( -\nabla \tilde{p})\right)
-\]
-\[
-+ \int_0^T \,dt \, \int_{\Gamma_r}\, dS_{\Gamma_r}( p \tilde{p}_r + \bv \cdot {\bf n}_r \tilde{v}_{nr})
-\]
-The first term vanishes after integration by parts. Accordingly,
-\begin{equation}
-\label{eqn:fadj}
- {\bf F}^T {\bf h}_r= -\tilde{u}(\cdot,T).
-\end{equation}
-
-\append{$L^2$ estimates}
-This appendix lays out the proof of
-
-\begin{theorem} 
-\label{thm:one}
-Suppose that $\log \kappa, \log \beta \in C^{\infty}(\bR^d) \cap L^{\infty}(\bR^d)$, and that Assumption A holds.
-Then the estimates \ref{eqn:coercive} hold for $h \in C^{\infty}_0(\Gamma_s \times \bR)$, hence $S[c]$ extends to $L^2(\bar{\Gamma}_s \times \bR)$ with the same estimates.
-\end{theorem}
-
-For convenience, I restate
-
-\noindent {\bf Assumption A:} There exists $\theta_*>0$ so that for every ray connecting a point in $\mbox{supp }\phi_s$ with a point in $\mbox{supp }\phi_r$, the angles $\theta_s$ and $\theta_r$ subtended by the ray and the tangent planes of $\Gamma_s$ and $\Gamma_r$ respectively are both greater than $\theta_*$ in absolute value.
-
-Recall that $\Gamma_r \subset \bR^d$ is a smooth embedded surface containing the receiver locations. Denote by 
-${\cal I}_r$ the injection of $\Gamma_r \times \bR$ into $\bR^{d+1}$. Define the angle $\theta(\xi)$ made by a covector $\xi \in T^*(\bR^d)|_{\Gamma_r}$ with $\Gamma_r$ by 
-\begin{equation}
-\label{eqn:angledef}
-\cos \theta(\xi) = \frac{\|  (D{\cal I}_r)^T\xi\|}{\|\xi\|}.
-\end{equation}
-in which $\| \cdot \|$ denotes the induced Euclidean norm.
-
-For a covector $(\xi,\omega) \in  T^*(\bR^{d+1})$, define the angle with $\Gamma_r \times \bR$ to be the angle made by $\xi$ with $\Gamma_r$.
-
-Choose $\theta^*>0$, and denote by $\lambda_r$ the open conic subset of $T^*(\bR^{d+1}) |_{\Gamma_r \times \bR}$ of covectors making angle $> \theta*$ with $\Gamma_r \times \bR$.
-The projection $\gamma_r = (D{\cal I}_r)^T\lambda_r$ consists of tangential components of these covectors; it is easy to see that $\gamma_r$ is also open.
-
-Let $P$ denote the symbol of the acoustic system
-\begin{equation}
-P(\bx,\omega,\bk) = 
-i\left(
-\begin{array}{cc}
-\frac{\omega}{\kappa(\bx)} & \bk^T\\
-\bk &  \rho(\bx)\omega I_{3\times3}
-\end{array}
-\right)
-\end{equation}
-For a normal covector $\nu \in {\cal N}(\Gamma_r \times \bR)  = ((D{\cal I}_r)^T)^{-1}(0)$ and $\bk \in \gamma_r$, the polynomial
-\[
-\omega \mapsto \det P(\bx,\omega,\bk)
-\]
-has two distinct nonzero roots $\pm \omega(\bx,\bk)$.
-
-Let $\beta_r$ be an open conic neighborhood of $\gamma_r^C$, and define $\alpha_r = ((D{\cal I}_r)^T)^{-1}\beta_r \subset T^*(\bR^{d+1}) |_{\Gamma_r \times \bR}$.
-
-Let $H \in OPS^0(\bR^{d+1})$ be elliptic with symbol asymptotic to 1 near the characteristic variety 
-\begin{equation}
-\label{eqn:chardef}
-C = \{(\bx,t,\bk,\omega): \det P(\bx,\bk,\omega)=0\}
-\end{equation}. That is,
- $ES(H)$ is a small nbhd of characteristic variety. Note that
-\[ ES(H) \cap {\cal N}(\Gamma_r \times \bR) = \emptyset\]
-From the second of these facts and the definition of $\beta_r$, the Theorem
-in \cite{BaoSy:91b} implies that for $\Phi \in C_0^{\infty}(\bR^{d+1})$,
-\begin{equation}
-\label{eqn:baosymes}
-Pu=0 \mbox{ and } u \in L^2(\bR^{d+1}) \cap H^1_{ml}(ES(H) \cap \beta_r) \,\Rightarrow\, {\cal I_r}^*(\Phi u) \in L^2(\Gamma_r \times \bR)
-\end{equation}
-
-Unpack this: $u \in H^1_{ml}(ES(H) \cap \beta_r),\, s \ge 0$ means that there exists $A \in OPS^1(\bR^{d+1})$, elliptic in $ES(H) \cap \beta_r$, so that $Au \in L^2_{\rm loc}(\bR^{d+1})$.
-Choose $\Phi \in C_0^{\infty}(\bR^{d+1})$, that is, $\Phi Au \in L^2_{\rm loc}(\bR^{d+1})$, that is, $u \in {\rm Dom}(\Phi A)$.
-Denote by $\| \cdot \|_{\Phi A}$ the graph norm of $\Phi A$, and by $H^s_{\Phi A}$ the completion of $C_0^{\infty}(\bR^{d+1})$ with respect to $\| \cdot \|_{\Phi A}$, with its natural injection into $L^2(\bR^{d+1})$.
-Then
-\[
-u \in H^s_{ml}(ES(H) \cap \beta_r) \,\Leftrightarrow \, u \in H^s_{\Phi A}
-\]
-for any $\Phi$, suitable choice of $A$. Further, the proof of the Theorem in \cite{BaoSy:91b} actually implies that for $\Phi$, $A$ as above with $s=1$,
-\[
-\|{\cal I}_r^*(\Phi u)\| \le C \|u\|_{\Phi A},
-\]
-That is, the localized restriction map is continuous on $H^1_{\Phi A}$.
-
- $\lambda_T \subset T^*(\bR^{d+1})_{t=T}$ = intersection with flow-out of $\lambda_r$, ${\cal I}_t =$ injection $\bR^d \rightarrow \bR^d \times \{t\} \subset \bR^{d+1}$, $\gamma_T = (D{\cal I})^T\lambda_T$.
-$\gamma_T$ = projection of flow-out to $\{t=T\}$ of lift of $\gamma_r$.
-
-$R_T \in OPS^1(\bR^d)$ elliptic of order 1 in nbhd of $\bar{\gamma}_T^c$ so $ES(R_T)^c \subset \gamma_T$. $R \in C^{\infty}(\bR, OPS^1(\bR^d))$ flow-out of $R_T$. Write (lifted essential support) 
-\[
-LES(R) = \cup_{t \in \bR} ((D({\cal I}_t)^T)^{-1}ES(R_t). 
-\]
-\begin{lemma}
-\label{thm:lem1} $(D{\cal I}_r)^T (LES(R) \cap T^*(\bR^{d+1})_{\Gamma_r \times \bR})$ contains a neighborhood of $\bar{\gamma}_r^c$.
-\end{lemma}
-\begin{proof}
-$(\bx,t,\xi,\omega) \in T^*(\bR^{d+1})_{\Gamma_r \times \bR}, \notin LES(R)$ means that $(D({\cal I}_r)^T(\bx,t,\xi,\omega) = (\bx,\xi) \notin ES(R_t)$.
-Let ${\bf H}$ denote the flow.
-Then ${\bf H}_{T-t}(\bx,\xi) \notin ES(R_T)$ so ${\bf H}_{T-t}(\bx,\xi) = (\bx_T,\xi_T) \in \gamma_T$ so there are $\omega_T, \omega_r$ for which $(\bx_T,T,\xi_T,\omega_T) \in \lambda_T \cap \mbox{ Char }$ is on ray through $(\bx,t,\xi,\omega_r) \in \lambda_r$, whence $(D{\cal I}_r)^T (\bx,t,\xi,\omega_r) \in \gamma_r$.
-\end{proof}
-
-Then Prop.
-2 in BaoSymes implies that for $ \Phi \in C_0^{\infty}(\bR^{d+1})$, $R\Phi H \in OPS^1(\bR^{d+1})$ and $ \lambda_r^C \subset ES(R\Phi H) \subset LES(R) \cap ES(H)$.
-Standard energy estimates imply if $R_Tu_T \in L^2(\bR^d)$ then $Ru \in L^2(\bR^{d+1})$.
-Also $(I-H)\Psi u \in H^m_{\rm loc}(\bR^{d+1})$ for any $m>0$ whence $R\Phi(I-H)\Psi u \in H^m_{\rm loc}(\bR^{d+1})$ for any $m \ge 0$ assume $\Psi \Phi = \Phi $.
-So $R\Phi Hu \in L^2(\bR^{d+1})$.
-
 \bibliographystyle{seg}

### 2020-11-23 repro/papers/sseprecond/plots.tex
interim

@@
 \inputdir{project}
+
+\section{Approximate Inversion}
 
@@
 \plot{trcftrll0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindll0}, \ref{fig:reptpwindll0}.}
+
+\section{Dirichlet-to-Neumann map}
+
+If the  (causal) pressure field $p$ is specified and continuous across $z=z_s$, then the acoustic fields are uniquely determined in $z<z_s$ and $z>z_s$.
+In particular, limits of  the normal ($z$) component of velocity  as $z\rightarrow z_s^{\pm}$ are determined, and the jump discontinuity in $z$ is equal to the surface pressure source generating the fields.
+The limits from the two sides are equal except for sign, therefore the surface source is equal to twice the limiting boundary value of the normal velocity.
+
+Similarly, specification of the (continuous) normal velocity uniquely determines the acoustic fields on either side of the surface $z=z_s$.
+The limits of the pressure as $z \rightarrow z_s^{\pm}$ have opposite signs and the jump discontinuity is equal to the surface normal velocity source generating the fields.
+
+The last observation leads to a prescription for computing the map from pressure trace on $z=z_s$ to normal velocity on $z=z_s$: form the surface source equal to twice the pressure trace (to account for the change of sign across the surface); solve the acoustic problem with this source; read off the normal component of velocity on $z=z_s$.
+
+Note that IWAVE treats every right-hand side point as defining a point source, that is, an amplitude divided by $\Delta x \Delta z$.
+Since the surface source is a surface $\delta$, that is (numerical approximation) an amplitude divided by $\Delta z$, it is necessary to precondition the source by multiplying the amplitude by $\Delta x$.
+Since the discontinuity is equal to twice the source amplitude, another factor of 2 is required.
+Finally, IWAVE does not include the bulk modulus and bulk modulus coefficients in the source injection, so these must also be multiplied in.
+
+The upshot is the cycle represented in the figures \ref{fig:srcvzhh0},  \ref{fig:resrcvzhh0}, \ref{fig:srcvzlh0},  \ref{fig:resrcvzlh0}.
+Figure \ref{fig:srcvzhh0} is the normal component of velocity at $z_s=3$ km of backpropagation, from the field with pressure trace displayed in Figure \ref{fig:ptpwindh0} at $z_r=1$ km, propagation and backpropagation in homogeneous material model with $\kappa = $ 4 Gpa, $\rho = $ 1 g/m$^{-6}$.The pressure trace at $z=z_s$ was also recorded, and used to recompute the normal velocity trace \ref{fig:resrcvzhh0}.
+The difference is displayed as figure \ref{fig:dresrcvzhh0}.
+
+Similarly, Figure \ref{fig:srcvzlh0} is the normal component of velocity at $z_s=3$ km of backpropagation, from the field with pressure trace displayed in Figure \ref{ptpwindl0} at $z_r=1$ km, propagation in the model with the model with bulk modulus displayed in Figure \ref{fig:bml0} and density $\rho = $ 1 g/m$^{-6}$, and backpropagation in the homogeneous model with bulk modulus = 4 GPa.
+The pressure trace at $z=z_s$ was also recorded, and used to recompute the normal velocity trace \ref{fig:resrcvzlh0}.
+The difference is displayed as figure \ref{fig:dresrcvzlh0}.
+
+%%%%%%%%%%%%%%%%%%%
+
+\plot{srcptphh0}{width=15cm}{Pressure traces corresponding to the pressure traces of Figure \ref{fig:ptpwindhh0}, backpropagated in homogeneous medium with $\kappa = $ 4 GPa, density = 1 g/cm$^3$.}
+
+\plot{srcvzhh0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindhh0}, backpropagated in homogeneous medium with $\kappa = $ 4 GPa, density = 1 g/cm$^3$.}
+
+\plot{resrcvzhh0}{width=15cm}{Output of synthetic DtoN map: $v_z$ traces for velocity source scaled from display \ref{fig:srcptphh0}.}
+
+\plot{dresrcvzhh0}{width=15cm}{Difference of figures \ref{fig:srcvzhh0}, \ref{fig:resrcvzhh0}.}
+
+\plot{srcptplh0}{width=15cm}{Pressure traces corresponding to the pressure traces of Figure \ref{fig:ptpwindll0}, backpropagated in homogeneous medium with $\kappa = $ 4 GPa, density = 1 g/cm$^3$.}
+
+\plot{srcvzlh0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindll0}, backpropagated in homogeneous medium with $\kappa = $ 4 GPa, density = 1 g/cm$^3$.}
+
+\plot{resrcvzlh0}{width=15cm}{Output of synthetic DtoN map: $v_z$ traces for source scaled from display \ref{fig:srcptplh0}.}
+
+\plot{dresrcvzlh0}{width=15cm}{Difference of figures \ref{fig:srcvzlh0}, \ref{fig:resrcvzlh0}.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\plot{srcvzhh0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindhh0}, backpropagated in homogeneous medium with $\kappa = $ 4 GPa, density = 1 g/cm$^3$.}
+
+\plot{resrcptphh0}{width=15cm}{Output of synthetic  NtoD map: pressure traces for source scaled from display \ref{fig:srcvzhh0}.}
+
+\plot{dresrcptphh0}{width=15cm}{Difference of figures \ref{fig:srcptphh0}, \ref{fig:resrcptphh0}.}
+
+\plot{srcvzlh0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindll0}, backpropagated in homogeneous medium with $\kappa = $ 4 GPa, density = 1 g/cm$^3$.}
+
+\plot{resrcptplh0}{width=15cm}{Output of synthetic NtoD map: pressure traces for source scaled from display \ref{fig:srcvzlh0}.}
+
+\plot{dresrcptplh0}{width=15cm}{Difference of figures \ref{fig:srcptplh0}, \ref{fig:resrcptplh0}.}

### 2020-11-23 repro/papers/discrep/figs.tex
interim

\title{Discrepancy Algorithm Examples}
\author{William W. Symes}

\lefthead{Symes}

\righthead{Examples}

\maketitle
\begin{abstract}
Numerical examples of discrepancy inversion.
\end{abstract}

\inputdir{project}
\newpage
\section{0. nominal N/S=0.2, target N/S=0.4, radius=0.1} 
\plot{datanp0}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =0.2}

\plot{resnp0}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 3.243320, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.400000, achieved noise-to-signal ratio =0.407452, estimated slowness=0.396000.}

\plot{resimp0}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 3.243320, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.400000, achieved noise-to-signal ratio =0.407452, estimated slowness=0.396000.}

\plot{wavp0}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =3.243320, assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presimp0}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp0}, penalty weight $\alpha$ =3.243320, actual noise-to-signal ratio 0.200000, target noise-to-signal ratio 0.400000, achieved noise-to-signal ratio = , assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presnp0}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp0}, data from Figure \ref{fig:datanp0}, actual noise-to-signal ratio = 0.200000, target noise-to-signal ratio = 0.400000, achieved noise-to-signal ratio = 0.289792, wavelet from Figure \ref{fig:wavp0}, penalty weight $\alpha$ =3.243320, assumed support radius =0.100000, estimated slowness =0.396000.}
\newpage
\section{1. nominal N/S=0.2, target N/S=0.2, radius=0.1} 
\plot{datanp1}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =0.2}

\plot{resnp1}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 1.000780, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.200000, achieved noise-to-signal ratio =0.206226, estimated slowness=0.396000.}

\plot{resimp1}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 1.000780, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.200000, achieved noise-to-signal ratio =0.206226, estimated slowness=0.396000.}

\plot{wavp1}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =1.000780, assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presimp1}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp1}, penalty weight $\alpha$ =1.000780, actual noise-to-signal ratio 0.200000, target noise-to-signal ratio 0.200000, achieved noise-to-signal ratio = , assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presnp1}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp1}, data from Figure \ref{fig:datanp1}, actual noise-to-signal ratio = 0.200000, target noise-to-signal ratio = 0.200000, achieved noise-to-signal ratio = 0.169006, wavelet from Figure \ref{fig:wavp1}, penalty weight $\alpha$ =1.000780, assumed support radius =0.100000, estimated slowness =0.396000.}
\newpage
\section{2. nominal N/S=0.2, target N/S=0.1, radius=0.1} 
\plot{datanp2}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =0.2}

\plot{resnp2}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 0.329086, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.100000, achieved noise-to-signal ratio =0.103459, estimated slowness=0.396000.}

\plot{resimp2}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 0.329086, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.100000, achieved noise-to-signal ratio =0.103459, estimated slowness=0.396000.}

\plot{wavp2}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =0.329086, assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presimp2}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp2}, penalty weight $\alpha$ =0.329086, actual noise-to-signal ratio 0.200000, target noise-to-signal ratio 0.100000, achieved noise-to-signal ratio = , assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presnp2}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp2}, data from Figure \ref{fig:datanp2}, actual noise-to-signal ratio = 0.200000, target noise-to-signal ratio = 0.100000, achieved noise-to-signal ratio = 0.159951, wavelet from Figure \ref{fig:wavp2}, penalty weight $\alpha$ =0.329086, assumed support radius =0.100000, estimated slowness =0.396000.}
\newpage
\section{3. nominal N/S=0.2, target N/S=0.05, radius=0.1} 
\plot{datanp3}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =0.2}

\plot{resnp3}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 0.176535, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.050000, achieved noise-to-signal ratio =0.050243, estimated slowness=0.396000.}

\plot{resimp3}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 0.176535, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.050000, achieved noise-to-signal ratio =0.050243, estimated slowness=0.396000.}

\plot{wavp3}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =0.176535, assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presimp3}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp3}, penalty weight $\alpha$ =0.176535, actual noise-to-signal ratio 0.200000, target noise-to-signal ratio 0.050000, achieved noise-to-signal ratio = , assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presnp3}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp3}, data from Figure \ref{fig:datanp3}, actual noise-to-signal ratio = 0.200000, target noise-to-signal ratio = 0.050000, achieved noise-to-signal ratio = 0.159736, wavelet from Figure \ref{fig:wavp3}, penalty weight $\alpha$ =0.176535, assumed support radius =0.100000, estimated slowness =0.396000.}
\newpage
\section{4. nominal N/S=0.2, target N/S=0.4, radius=0.05} 
\plot{datanp4}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =0.2}

\plot{resnp4}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 3.243320, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.400000, achieved noise-to-signal ratio =0.407452, estimated slowness=0.396000.}

\plot{resimp4}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 3.243320, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.400000, achieved noise-to-signal ratio =0.407452, estimated slowness=0.396000.}

\plot{wavp4}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =3.243320, assumed support radius =0.050000, estimated slowness =0.396000.}

\plot{presimp4}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp4}, penalty weight $\alpha$ =3.243320, actual noise-to-signal ratio 0.200000, target noise-to-signal ratio 0.400000, achieved noise-to-signal ratio = , assumed support radius =0.050000, estimated slowness =0.396000.}

\plot{presnp4}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp4}, data from Figure \ref{fig:datanp4}, actual noise-to-signal ratio = 0.200000, target noise-to-signal ratio = 0.400000, achieved noise-to-signal ratio = 0.291543, wavelet from Figure \ref{fig:wavp4}, penalty weight $\alpha$ =3.243320, assumed support radius =0.050000, estimated slowness =0.396000.}
\newpage
\section{5. nominal N/S=0.2, target N/S=0.2, radius=0.05} 
\plot{datanp5}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =0.2}

\plot{resnp5}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 1.000780, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.200000, achieved noise-to-signal ratio =0.206226, estimated slowness=0.396000.}

\plot{resimp5}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 1.000780, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.200000, achieved noise-to-signal ratio =0.206226, estimated slowness=0.396000.}

\plot{wavp5}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =1.000780, assumed support radius =0.050000, estimated slowness =0.396000.}

\plot{presimp5}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp5}, penalty weight $\alpha$ =1.000780, actual noise-to-signal ratio 0.200000, target noise-to-signal ratio 0.200000, achieved noise-to-signal ratio = , assumed support radius =0.050000, estimated slowness =0.396000.}

\plot{presnp5}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp5}, data from Figure \ref{fig:datanp5}, actual noise-to-signal ratio = 0.200000, target noise-to-signal ratio = 0.200000, achieved noise-to-signal ratio = 0.181122, wavelet from Figure \ref{fig:wavp5}, penalty weight $\alpha$ =1.000780, assumed support radius =0.050000, estimated slowness =0.396000.}
\newpage
\section{6. nominal N/S=0.2, target N/S=0.1, radius=0.05} 
\plot{datanp6}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =0.2}

\plot{resnp6}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 0.329086, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.100000, achieved noise-to-signal ratio =0.103459, estimated slowness=0.396000.}

\plot{resimp6}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 0.329086, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.100000, achieved noise-to-signal ratio =0.103459, estimated slowness=0.396000.}

\plot{wavp6}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =0.329086, assumed support radius =0.050000, estimated slowness =0.396000.}

\plot{presimp6}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp6}, penalty weight $\alpha$ =0.329086, actual noise-to-signal ratio 0.200000, target noise-to-signal ratio 0.100000, achieved noise-to-signal ratio = , assumed support radius =0.050000, estimated slowness =0.396000.}

\plot{presnp6}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp6}, data from Figure \ref{fig:datanp6}, actual noise-to-signal ratio = 0.200000, target noise-to-signal ratio = 0.100000, achieved noise-to-signal ratio = 0.176239, wavelet from Figure \ref{fig:wavp6}, penalty weight $\alpha$ =0.329086, assumed support radius =0.050000, estimated slowness =0.396000.}
\newpage
\section{7. nominal N/S=0.2, target N/S=0.05, radius=0.05} 
\plot{datanp7}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =0.2}

\plot{resnp7}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 0.176535, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.050000, achieved noise-to-signal ratio =0.050243, estimated slowness=0.396000.}

\plot{resimp7}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 0.176535, data noise-to-signal ratio $\eta$ =0.200000, target noise-to-signal ratio =0.050000, achieved noise-to-signal ratio =0.050243, estimated slowness=0.396000.}

\plot{wavp7}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =0.176535, assumed support radius =0.050000, estimated slowness =0.396000.}

\plot{presimp7}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp7}, penalty weight $\alpha$ =0.176535, actual noise-to-signal ratio 0.200000, target noise-to-signal ratio 0.050000, achieved noise-to-signal ratio = , assumed support radius =0.050000, estimated slowness =0.396000.}

\plot{presnp7}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp7}, data from Figure \ref{fig:datanp7}, actual noise-to-signal ratio = 0.200000, target noise-to-signal ratio = 0.050000, achieved noise-to-signal ratio = 0.176169, wavelet from Figure \ref{fig:wavp7}, penalty weight $\alpha$ =0.176535, assumed support radius =0.050000, estimated slowness =0.396000.}
\newpage
\section{8. nominal N/S=1.0, target N/S=1.0, radius=0.1} 
\plot{datanp8}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =1.0}

\plot{resnp8}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 5.505000, data noise-to-signal ratio $\eta$ =1.000000, target noise-to-signal ratio =1.000000, achieved noise-to-signal ratio =0.973389, estimated slowness=0.396000.}

\plot{resimp8}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 5.505000, data noise-to-signal ratio $\eta$ =1.000000, target noise-to-signal ratio =1.000000, achieved noise-to-signal ratio =0.973389, estimated slowness=0.396000.}

\plot{wavp8}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =5.505000, assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presimp8}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp8}, penalty weight $\alpha$ =5.505000, actual noise-to-signal ratio 1.000000, target noise-to-signal ratio 1.000000, achieved noise-to-signal ratio = , assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presnp8}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp8}, data from Figure \ref{fig:datanp8}, actual noise-to-signal ratio = 1.000000, target noise-to-signal ratio = 1.000000, achieved noise-to-signal ratio = 0.691375, wavelet from Figure \ref{fig:wavp8}, penalty weight $\alpha$ =5.505000, assumed support radius =0.100000, estimated slowness =0.396000.}
\newpage
\section{9. nominal N/S=1.0, target N/S=0.5, radius=0.1} 
\plot{datanp9}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =1.0}

\plot{resnp9}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 0.546527, data noise-to-signal ratio $\eta$ =1.000000, target noise-to-signal ratio =0.500000, achieved noise-to-signal ratio =0.523341, estimated slowness=0.388000.}

\plot{resimp9}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 0.546527, data noise-to-signal ratio $\eta$ =1.000000, target noise-to-signal ratio =0.500000, achieved noise-to-signal ratio =0.523341, estimated slowness=0.388000.}

\plot{wavp9}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =0.546527, assumed support radius =0.100000, estimated slowness =0.388000.}

\plot{presimp9}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp9}, penalty weight $\alpha$ =0.546527, actual noise-to-signal ratio 1.000000, target noise-to-signal ratio 0.500000, achieved noise-to-signal ratio = , assumed support radius =0.100000, estimated slowness =0.388000.}

\plot{presnp9}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp9}, data from Figure \ref{fig:datanp9}, actual noise-to-signal ratio = 1.000000, target noise-to-signal ratio = 0.500000, achieved noise-to-signal ratio = 0.559738, wavelet from Figure \ref{fig:wavp9}, penalty weight $\alpha$ =0.546527, assumed support radius =0.100000, estimated slowness =0.388000.}
\newpage
\section{10. nominal N/S=1.0, target N/S=0.1, radius=0.1} 
\plot{datanp10}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =1.0}

\plot{resnp10}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 0.133101, data noise-to-signal ratio $\eta$ =1.000000, target noise-to-signal ratio =0.100000, achieved noise-to-signal ratio =0.102812, estimated slowness=0.364000.}

\plot{resimp10}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 0.133101, data noise-to-signal ratio $\eta$ =1.000000, target noise-to-signal ratio =0.100000, achieved noise-to-signal ratio =0.102812, estimated slowness=0.364000.}

\plot{wavp10}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =0.133101, assumed support radius =0.100000, estimated slowness =0.364000.}

\plot{presimp10}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp10}, penalty weight $\alpha$ =0.133101, actual noise-to-signal ratio 1.000000, target noise-to-signal ratio 0.100000, achieved noise-to-signal ratio = , assumed support radius =0.100000, estimated slowness =0.364000.}

\plot{presnp10}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp10}, data from Figure \ref{fig:datanp10}, actual noise-to-signal ratio = 1.000000, target noise-to-signal ratio = 0.100000, achieved noise-to-signal ratio = 0.529947, wavelet from Figure \ref{fig:wavp10}, penalty weight $\alpha$ =0.133101, assumed support radius =0.100000, estimated slowness =0.364000.}
\newpage
\section{11. nominal N/S=0.0, target N/S=0.1, radius=0.1} 
\plot{datanp11}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =0.0}

\plot{resnp11}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 1.389380, data noise-to-signal ratio $\eta$ =0.000000, target noise-to-signal ratio =0.100000, achieved noise-to-signal ratio =0.101633, estimated slowness=0.396000.}

\plot{resimp11}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 1.389380, data noise-to-signal ratio $\eta$ =0.000000, target noise-to-signal ratio =0.100000, achieved noise-to-signal ratio =0.101633, estimated slowness=0.396000.}

\plot{wavp11}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =1.389380, assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presimp11}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp11}, penalty weight $\alpha$ =1.389380, actual noise-to-signal ratio 0.000000, target noise-to-signal ratio 0.100000, achieved noise-to-signal ratio = , assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presnp11}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp11}, data from Figure \ref{fig:datanp11}, actual noise-to-signal ratio = 0.000000, target noise-to-signal ratio = 0.100000, achieved noise-to-signal ratio = 0.071865, wavelet from Figure \ref{fig:wavp11}, penalty weight $\alpha$ =1.389380, assumed support radius =0.100000, estimated slowness =0.396000.}
\newpage
\section{12. nominal N/S=0.0, target N/S=0.01, radius=0.1} 
\plot{datanp12}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =0.0}

\plot{resnp12}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 0.410373, data noise-to-signal ratio $\eta$ =0.000000, target noise-to-signal ratio =0.010000, achieved noise-to-signal ratio =0.010434, estimated slowness=0.400000.}

\plot{resimp12}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 0.410373, data noise-to-signal ratio $\eta$ =0.000000, target noise-to-signal ratio =0.010000, achieved noise-to-signal ratio =0.010434, estimated slowness=0.400000.}

\plot{wavp12}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =0.410373, assumed support radius =0.100000, estimated slowness =0.400000.}

\plot{presimp12}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp12}, penalty weight $\alpha$ =0.410373, actual noise-to-signal ratio 0.000000, target noise-to-signal ratio 0.010000, achieved noise-to-signal ratio = , assumed support radius =0.100000, estimated slowness =0.400000.}

\plot{presnp12}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp12}, data from Figure \ref{fig:datanp12}, actual noise-to-signal ratio = 0.000000, target noise-to-signal ratio = 0.010000, achieved noise-to-signal ratio = 0.007378, wavelet from Figure \ref{fig:wavp12}, penalty weight $\alpha$ =0.410373, assumed support radius =0.100000, estimated slowness =0.400000.}
\newpage
\section{13. nominal N/S=0.0, target N/S=0.001, radius=0.1} 
\plot{datanp13}{width=\textwidth}{Data plus noise, noise-to-signal ratio $\eta$ =0.0}

\plot{resnp13}{width=\textwidth}{Residual at VPM inversion, penalty weight $\alpha$ = 0.126322, data noise-to-signal ratio $\eta$ =0.000000, target noise-to-signal ratio =0.001000, achieved noise-to-signal ratio =0.001005, estimated slowness=0.396000.}

\plot{resimp13}{width=\textwidth}{Resimulation at VPM inversion, penalty weight $\alpha$ = 0.126322, data noise-to-signal ratio $\eta$ =0.000000, target noise-to-signal ratio =0.001000, achieved noise-to-signal ratio =0.001005, estimated slowness=0.396000.}

\plot{wavp13}{width=\textwidth}{Estimated wavelet at VPM inversion (blue) vs. noise-free wavelet (red), penalty weight $\alpha$ =0.126322, assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presimp13}{width=\textwidth}{Resimulation at projected VPM inversion (blue), noise-free data trace (red). Wavelet from Figure \ref{fig:wavp13}, penalty weight $\alpha$ =0.126322, actual noise-to-signal ratio 0.000000, target noise-to-signal ratio 0.001000, achieved noise-to-signal ratio = , assumed support radius =0.100000, estimated slowness =0.396000.}

\plot{presnp13}{width=\textwidth}{Residual at projected VPM inversion, resimulation from Figure \ref{fig:presimp13}, data from Figure \ref{fig:datanp13}, actual noise-to-signal ratio = 0.000000, target noise-to-signal ratio = 0.001000, achieved noise-to-signal ratio = 0.000710, wavelet from Figure \ref{fig:wavp13}, penalty weight $\alpha$ =0.126322, assumed support radius =0.100000, estimated slowness =0.396000.}

### 2020-11-23 repro/papers/discrep/hc.tex
interim

\title{Example Plots}
\author{William. W. Symes \email{\tt symes@rice.edu}}

\lefthead{WWS}

\righthead{HC Examples}

\maketitle
\begin{abstract}
Examples from HC, 2020.10.08.
\end{abstract}

\section{Introduction}
Data process:

For example {\tt n, n=1, 3, 4} and {\tt type} = {\tt data}, {\tt residual}:
\begin{itemize}
\item read HC file {\tt file.txt} = ex{\tt n\_type\_}(appropriate figure number){\tt.txt}, use {\tt grep -v} to remove comment ilne
\item use SU utility {\tt a2b} to convert ascii file {\tt file.txt} to binary file {\tt file.bin}
\item use SU utility {\tt suaddhead} to add headers {\tt ns} and {\tt dt} to produce SEGY file {\tt file.su}
\item use Madagascar utility {\tt sfsuread} to convert SEGY file {\tt file.su} to RSF file {\tt file.rsf} for plotting
\item use IWAVE utility {\tt SEGYNorm} to output L2 norm of the data in {\tt file.su} to terminal.
\end{itemize}

For example  {\tt n, n=1, 3, 4}, use Madagascar utility {\tt sfcat} to concatenate the data and residual {\tt .rsf} files, then use {\tt sfgraph} to plot two curves, with data in blue and residual in red.

{\tt SEGYNorm} computes the scaled $l^2$ norm of a vector $x$ (interpreted as the vector of samples of a function of a real variable on a regular grid) with length $ns$ and sample interval $dt$:
\[
  \|x\| = \left(dt \sum_{i=0}^{ns-1} |x[i]|^2\right)^{1/2}
\]
Note that if the extreme samples are $=0$, then this expession is equivalent to the trapezoidal rule.

There are (at least) two obvious ways of defining SNR. By ``SNR'' in the captions below, I mean the norm of the data divided by the norm of the residual. The meaning in the current draft paper might be called ``ideal SNR'', meaning the norm of the {\em noise-free} data (the data of Example 1 in all cases), divided by the difference of the predicted data minus the noise-free data (rather than the input data, which the algorithm is trying to fit). There shouldn't be much difference between these two, though they are certainly not the same. I have only given the straightforward SNR since in any non-synthetic situation you would not know the ``noise-free'' data.

\section{Examples}

\inputdir{hc}

\plot{ex1}{width=\textwidth}{Example 1 (combines Figures 3 and 5 in HC notes of 2020.10.05): data = blue curve, residual = red curve. Data norm = 1.3765, residual norm = 0.724783, achieved signal-noise ratio = 1.899.}

\plot{ex3}{width=\textwidth}{Example 3 (combines Figures 18 and 21 in HC notes of 2020.10.05): data = blue curve, residual = red curve. Data norm = 2.01366, residual norm = 0.595673, achieved signal-noise ratio = 3.38.}

\plot{ex4}{width=\textwidth}{Example 4 (combines Figures 27 and 30 in HC notes of 2020.10.05): data = blue curve, residual = red curve. Data norm = 1.60526, residual norm = 0.685794, achieved signal-noise ratio = 2.341.}

### 2020-11-23 repro/papers/discrep/paper.tex
interim

@@
 (``Findings'') section of the paper contains precise statements of the
-mathematical results and supporting discussion, followed by
-derivations. In particular, theorems referenced in the Overview
-section are to be found, stated in full with their proofs, in the this
-final part of the paper. Three appendices describe the algorithm used to
-control the penalty weight in our formulation of extended
-inversion, further refinement of error estimates based ond the
-Heisenberg inequality, and the abstract structure of objective derivatives
-inherent in extended inversion based on wave propagation.
+mathematical results, derivations, and supporting discussion. Three
+appendices describe the algorithm used to control the penalty weight
+in our formulation of extended inversion, further refinement of error
+estimates based ond the Heisenberg inequality, and the abstract
+structure of extended inversion implicit in the derivations presented
+in the ``Findings'' section.
 
@@
 
-\subsection{Results}
+\subsection{Full Waveform Inversion}
 While $F[m]$ is surjective for every $m \in M$, as noted above, it is
@@
 \end{proposition}
+
+\noindent{\bf Remark:} A useful consequence of the condition 
+\ref{eqn:fullrec}: for every $m \in M$, 
+\begin{equation}
+  [-\lambda_{\rm max}, -\lambda_{\rm max}] \subset[ t_{\rm min}-mr , 
+  t_{\rm max}-mr]. 
+  \label{eqn:zeroinc}
+\end{equation}
 
@@
 
+\begin{proof} From the definition \ref{eqn:mod},
+\[
+ \lerr[m,w;d] =  \frac{1}{32\pi^2
+    r^2\|d\|^2}\int_0^T\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2
+\]
+Since $w_*, w$ vanish for $|t|>\lambda$,
+$\lF[m_*]w_*(t)$ vanishes if $|t-m_*r|>\lambda$ and $\lF[m]w$ vanishes if $|t-mr|>\lambda$. So if $|mr-m_*r|
+= |m-m_*|r > 2\lambda$, then $|t-mr|+|t-m_*r| \ge |mr-m_*r| >
+2\lambda$ so either $|t-mr|>\lambda$ or $|t-m_*r|>\lambda$, that is,
+either $\lF[m]w(t)=0$ or $\lF[m_*]w_*=0$. Therefore $\lF[m]w$ and
+$\lF[m_*]w_*$ are orthogonal in the sense of the $L^2$ inner product
+$\langle \cdot,\cdot \rangle_D$ on $D$:
+\begin{equation}
+  \label{eqn:ortho}
+  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \langle F[m]w,
+  F[m_*]w_*\rangle_D = 0
+\end{equation}
+But $d = \lF[m_*]w_*$, so this is the same as saying that $d$ is
+orthogonal to $F[m]w$. So conclude after a minor manipulation that
+\[
+  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \lerr[m,w;d]=\frac{1}{32\pi^2 
+    r^2\|d\|^2}(\|w\|^2 + \|w_*\|^2).
+\]
+\begin{equation}
+  \label{eqn:iso}
+  = \frac{1}{2}\left(\frac{\|w\|^2}{\|w_*\|^2} + 1 \right)
+\end{equation}
+That is, for slowness $m$ in error by more than $2\lambda/r$ from the 
+target slowness $m_*$, the means square error (FWI objective) $\lerr$ is independent of
+$m$, and its minimum over $w$ is attained for $w=0$
+\end{proof}
+
+Therefore local minimizers of $\lerr$ abound, as far as you like from the
+global minimizer $(m_*,w_*)$. Local exploration of the FWI objective
+$e$ gives no useful information whatever about constructive search
+directions, and descent-based optimization tends to fail if the
+initial estimate $m_0$ is in error by more than $2\lambda/r$
+(``further than a multiple of a wavelength'', per discussion
+in the second Appendix). In fact the actual behaviour of FWI itererations is worse
+(failure if $m_0$ is in error by ``half a wavelength''), as follows
+from a more refined analysis of the ``cycle-skipping'' local behaviour of $\lerr$ near its
+global minimizer.
+
+\subsection{Variable Projection}
 The main theoretical device used in the proofs of our main results on 
@@
 \end{equation}
+With $A$ of this form, $\tilde{J}_{\alpha}$ is explicitly
+computable. First observe that apart from amplitude, $F[m]$ is
+unitary: for $g \in D$,
+\begin{equation}
+\label{eqn:tran}
+F[m]^T g (t) =
+\left\{
+  \begin{array}{c}
+    \frac{1}{4\pi r}g\left(t+mr\right), \, t \in [t_{\rm min}-mr,
+    t_{\rm max}-mr],\\
+    0, \mbox{ else.}
+  \end{array}
+\right.
+\end{equation}
+so
+\begin{equation}
+  \label{eqn:unit}
+  F[m]^TF[m] = \frac{1}{(4\pi r)^2}{\bf 1}_{[t_{\rm min}-mr,  
+    t_{\rm max}-mr]}
+\end{equation}
+in which ${\bf 1}_{S}$ denotes
+multiplication by the characteristic function of a measurable 
+$S \subset \bR$.
+
+Therefore the normal equation for the minimizer on the RHS of equation \ref{eqn:redexp} is
+\begin{equation}
+  \label{eqn:norm1}
+  \left(\frac{1}{(4\pi r)^2} {\bf 1}_{[t_{\rm min}-mr,  
+      t_{\rm max}-mr]} + \alpha^2 A^TA\right)w= F[m]^Td.
+\end{equation}
+
+With these choices, the normal equation \ref{eqn:norm1} becomes
+\begin{equation}
+\label{eqn:norm2}
+\left(\frac{1}{(4\pi r)^2}  {\bf 1}_{[t_{\rm min}-mr,  
+      t_{\rm max}-mr]} + \alpha^2a^2\right)w= F[m]^Td.
+\end{equation}
+
+\begin{proposition}
+  \label{thm:norminvexp}
+  Assume the conditions \ref{eqn:mod}, \ref{eqn:fullrec},
+  \ref{eqn:annmult}. Also assume that $\lambda \in (0,\lambda_{\rm
+    max}], \alpha > 0,$ and that  $C>0$ exists so that $a \in L^{\infty}(\bR)$
+  mentioned in condition \ref{eqn:annmult} satisfies condition
+  \ref{eqn:abnd}.
+%  \begin{equation}
+%   \label{eqn:abnd}
+%    |t| > \lambda \Rightarrow a(t) \ge C,
+%  \end{equation}
+%  in which $C>0$ may depend on $\lambda$.
+  Then
+  \begin{itemize}
+  \item[1. ]the normal operator $F[m]^TF[m] + \alpha^2A^TA$ is
+    invertible for any $m \in M$, $\alpha > 0$;
+  \item[2. ]the solution $\aw[m;d]\in W$ of the normal equation
+    \ref{eqn:norm} is given by
+    \begin{equation}
+      \label{eqn:normsol}
+      \aw[m;d](t) = \left\{
+        \begin{array}{c}
+          \left(\frac{1}{(4\pi r)^2} + \alpha^2
+          a^2(t)\right)^{-1}\frac{1}{4 \pi r}d(t+mr), t \in [t_{\rm
+          min}-mr, t_{\rm max}-mr];\\
+          0, \mbox{ else;}
+        \end{array}
+      \right.
+    \end{equation}
+  \item[3. ]if in addition $d=F[m_*]w_*, w_* \in \lW$ is noise-free, as in equation
+    \ref{eqn:defdatanonoise},
+    \begin{equation}
+      \label{eqn:solnonoise}
+      \aw[m,d](t)= \left(1+ (4\pi r)^2\alpha^2 a(t)^2\right)^{-1}w_*\left(t+(m-m_*)r\right).
+    \end{equation}
+  \end{itemize}
+\end{proposition}
+
+\begin{proof} 
+  \begin{itemize}
+  \item[1. ]Note that thanks to \ref{eqn:zeroinc}, if $|t|\le
+    \lambda \le \lambda_{\rm max}$, then ${\bf 1}_{[t_{\rm min}-mr,  
+      t_{\rm max}-mr]}(t) = 1$, whereas if $|t|>\lambda$,
+    then $a(t) \ge C$, whence
+    \[
+      \frac{1}{(4\pi r)^2}  {\bf 1}_{[t_{\rm min}-mr,  
+        t_{\rm max}-mr]} + \alpha^2a^2  \ge \min\{(4\pi r)^2,
+      \alpha^2\min\{1/(4\pi r)^2,C^2\} \}> 0.
+    \]
+    Therefore the normal operator is invertible under the stated
+    conditions.
+
+  \item[2. ]From the identity \ref{eqn:tran}.
+    \[
+      \mbox{supp }F[m]^Td \subset [t_{\rm min}-mr,t_{\rm max}-mr].
+    \]
+    Define $w_{\rm tmp}$ to be the right-hand side of equation \ref{eqn:normsol}. Then
+    from the previous observation and identity \ref{eqn:tran},
+    \[
+      \mbox{supp }w_{\rm tmp} \subset [t_{\rm min}-mr,t_{\rm max}-mr].
+    \]
+    From the identity \ref{eqn:unit}, for any $w \in W$,
+    \[
+      t \in [t_{\rm min}-mr,t_{\rm max}-mr] \Rightarrow F[m]^TF[m]w(t)
+      = \frac{1}{(4 \pi r)^2}w(t).
+    \]
+    It follows from this and the previous two observations that
+    $w_{\rm tmp}$ solves the normal equation \ref{eqn:norm}, and
+    therefore that $\aw[m;d]=w_{\rm tmp}$.
+
+  \item[3. ]Follows by inserting the definition
+    \ref{eqn:defdatanonoise} of $d$ in \ref{eqn:normsol} and
+    rearranging.
+  \end{itemize}
+\end{proof}
 
@@
 
+\begin{proof} These conclusions follow immediately from Proposition
+  \ref{thm:norminvexp}.
+\end{proof}
+
 \noindent{\bf Remark:} If $\Ja[\cdot,\cdot;d]$ and $\tJa[\cdot;d]$
@@
 whereas $\tJa[\cdot;d]$ {\em is} differentiable for proper choice of
-penalty operator $A$:
+penalty operator $A$.
+
+This conclusion follows from properties of the modeling operator $F$
+shared with many other inverse problems in wave propagation, as
+explained in the third appendix. Here, we derive it from explicit
+expressions for $\tJa$ and its components.
+
+\begin{proposition}
+  \label{thm:epjgen}
+  Assume the hypotheses of Proposition \ref{thm:norminvexp}. Then
+  \begin{equation}
+  \label{eqn:residnormgen}
+  e[m,\aw[m,d];d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r \alpha a(t-mr))^4(1 +
+  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)^2
+\end{equation}
+\begin{equation}
+  \label{eqn:anninormgen}
+  p[m,\aw[m,d];d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r a(t-mr))^2(1 +
+  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)
+\end{equation}
+
+\begin{equation}
+  \label{eqn:expjgen}
+\tJa[m;d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}}\,dt\,(4\pi r \alpha a(t-mr))^2(1+(4\pi r \alpha 
+a(t-mr))^2)^{-1}d(t)^2. 
+\end{equation}
+\end{proposition}
+
+\begin{proof}
+  From equation \ref{eqn:normsol},
+  \[
+    F[m]\aw[m;d](t) = 
+    \frac{1}{4 \pi r}\left(\frac{1}{(4\pi r)^2} + \alpha^2
+      a^2(t-mr)\right)^{-1}\frac{1}{4 \pi r}d(t),
+  \]
+  so
+  \[
+    (F[m]\aw[m;d]-d)(t) = (1 + (4 \pi r\alpha
+    a(t-mr))^2)^{-1}-1)d(t)
+  \]
+  \[
+    = -(4 \pi r\alpha a(t-mr))^2(1 + (4 \pi r\alpha
+    a(t-mr))^2)^{-1}d(t).
+  \]
+  Half the integral of the square of this data residual is
+  $e[m,\aw[m;d],d]$, which proves identity \ref{eqn:residnormgen}.
+
+  To compute $p[m,\aw[m;d],d]$, note that
+  \[
+    A\aw[m;d](t)=a(t) \left(\frac{1}{(4\pi r)^2} + \alpha^2
+      a^2(t)\right)^{-1}\frac{1}{4 \pi r}d(t+mr)
+  \]
+  \[
+    = 4\pi r a(t) (1 + (4\pi r \alpha a(t))^2)^{-1}d(t+mr)
+  \]
+  for $ t \in [t_{\rm min}-mr, t_{\rm max}-mr]$, so squaring,
+  integrating, and changing integration variables $t \mapsto t-mr$
+  gives the result \ref{eqn:anninormgen}
+
+  That the VPM objective $\tJa$ is given by \ref{eqn:expjgen} follows from equations \ref{eqn:pen},
+  \ref{eqn:redexp}, \ref{eqn:residnormgen}, and
+  \ref{eqn:anninormgen}.
+\end{proof}
 
@@
 
-This phenomenon - nondifferentiable objective, differentiable reduced
-objective - is replicated for a wide variety of extended
-approaches for inverse problems in wave propagation. See for example \cite{Symes:EAGE15}.
+\begin{proof}
+Suppose first that $a \in C^1(\bR)$. Differentiation under the integral sign  
+  yields the expression for its derivative:
+\begin{equation}
+  \label{eqn:dexpjgen}
+  \frac{d}{dm}\tJa[m;d] = -\frac{(4 \pi r \alpha)^2}{\|d\|^2} \int_{t_{\rm min}}^{t_{\rm max}} \,dt \, 
+  \left(a\frac{da}{dt}\right)(t-mr)(1+(4\pi r \alpha 
+  a(t-mr))^2)^{-2}d(t)^2. 
+\end{equation}
+For $a \in W^{1,\infty}_{\rm loc}(\bR)$ a limiting argument shows that the
+same expression gives the derivative of $\tJa$.
+\end{proof}
+
+It will be useful to record expressions for the various componenets of
+$\tJa$ when the data is noise-free, that is, the context of
+Proposition \ref{thm:norminvexp}, item 3.
+
+\begin{corollary}
+  \label{thm:epjnonoise}
+  Assume the hypotheses of Proposition \ref{thm:norminvexp}, item
+  3. Then noting that $\|d\| = \|w_*\|/(4 \pi r)$
+\begin{equation}
+  \label{eqn:residnorm}
+  e[m,\aw[m,d];d] 
+= \frac{\alpha^4}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^4(1+(4\pi r)^2 \alpha^2 
+    a(t-(m-m^*)r)^2)^{-2}w_*(t)^2.
+\end{equation}
+\begin{equation}
+  \label{eqn:anninorm}
+  p[m,\aw[m,d];d] = \frac{(4\pi r)^2}{2\|w_*\|^2}\int \,dt\,  a(t-(m-m_*)r)^2 \left(1+ (4\pi r)^2\alpha^2
+    a(t-(m-m_*)r)^2\right)^{-2}w_*(t)^2.
+\end{equation}
+so 
+\begin{equation}
+\label{eqn:expjnonoise}
+\tJa[m;d] = \frac{(4\pi r \alpha)^2}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
+  a(t-(m-m_*)r)^2)^{-1}w_*(t)^2. 
+\end{equation}
+Finally, if $a \in W^{1,1}_{\rm loc}(\bR)$, then $\tJa[\cdot;d]$ is differentiable, and 
+\begin{equation}
+  \label{eqn:dexpjnonoise}
+  \frac{d}{dm}\tJa[m;d] = -\frac{r (4\pi r \alpha)^2}{\|w_*\|^2} \int \,dt \, 
+  \left(a\frac{da}{dt}\right)(t-(m-m_*)r)(1+(4\pi r)^2 \alpha^2 
+  a(t-(m-m_*)r)^2)^{-2}w_*(t)^2. 
+\end{equation}
+\end{corollary}
+
+\subsection{Choice of Penalty Operator}
 
@@
 \subsection{Proof of Theorem \ref{thm:fwi}}
-\noindent{\bf Remark:} A useful consequence of the condition 
-\ref{eqn:fullrec}: for every $m \in M$, 
-\begin{equation}
-  [-\lambda_{\rm max}, -\lambda_{\rm max}] \subset[ t_{\rm min}-mr , 
-  t_{\rm max}-mr]. 
-  \label{eqn:zeroinc}
-\end{equation}
 
@@
 
-\begin{proof} of Theorem \ref{thm:fwi}: From the definition \ref{eqn:mod},
-\[
- \lerr[m,w;d] =  \frac{1}{32\pi^2
-    r^2\|d\|^2}\int_0^T\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2
-\]
-Since $w_*, w$ vanish for $|t|>\lambda$,
-$\lF[m_*]w_*(t)$ vanishes if $|t-m_*r|>\lambda$ and $\lF[m]w$ vanishes if $|t-mr|>\lambda$. So if $|mr-m_*r|
-= |m-m_*|r > 2\lambda$, then $|t-mr|+|t-m_*r| \ge |mr-m_*r| >
-2\lambda$ so either $|t-mr|>\lambda$ or $|t-m_*r|>\lambda$, that is,
-either $\lF[m]w(t)=0$ or $\lF[m_*]w_*=0$. Therefore $\lF[m]w$ and
-$\lF[m_*]w_*$ are orthogonal in the sense of the $L^2$ inner product
-$\langle \cdot,\cdot \rangle_D$ on $D$:
-\begin{equation}
-  \label{eqn:ortho}
-  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \langle F[m]w,
-  F[m_*]w_*\rangle_D = 0
-\end{equation}
-But $d = \lF[m_*]w_*$, so this is the same as saying that $d$ is
-orthogonal to $F[m]w$. So conclude after a minor manipulation that
-\[
-  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \lerr[m,w;d]=\frac{1}{32\pi^2 
-    r^2\|d\|^2}(\|w\|^2 + \|w_*\|^2).
-\]
-\begin{equation}
-  \label{eqn:iso}
-  = \frac{1}{2}\left(\frac{\|w\|^2}{\|w_*\|^2} + 1 \right)
-\end{equation}
-That is, for slowness $m$ in error by more than $2\lambda/r$ from the 
-target slowness $m_*$, the means square error (FWI objective) $\lerr$ is independent of
-$m$, and its minimum over $w$ is attained for $w=0$
-\end{proof}
 %\begin{theorem}
@@
 %\end{theorem}
-
-Therefore local minimizers of $\lerr$ abound, as far as you like from the
-global minimizer $(m_*,w_*)$. Local exploration of the FWI objective
-$e$ gives no useful information whatever about constructive search
-directions, and descent-based optimization tends to fail if the
-initial estimate $m_0$ is in error by more than $2\lambda/r$
-(``further than a multiple of a wavelength'', per discussion
-above). In fact the actual behaviour of FWI itererations is worse
-(failure if $m_0$ is in error by ``half a wavelength''), as follows
-from a more refined analysis of the ``cycle-skipping'' local behaviour of $\lerr$ near its
-global minimizer.
 
@@
 \subsection{Proof of Theorem \ref{thm:norminv}}
-For the single-trace transmission inverse problem, $\tilde{J}_{\alpha}$ is explicitly
-computable. First observe that apart from amplitude, $F[m]$ is
-unitary: for $g \in D$,
-\begin{equation}
-\label{eqn:tran}
-F[m]^T g (t) =
-\left\{
-  \begin{array}{c}
-    \frac{1}{4\pi r}g\left(t+mr\right), \, t \in [t_{\rm min}-mr,
-    t_{\rm max}-mr],\\
-    0, \mbox{ else.}
-  \end{array}
-\right.
-\end{equation}
-so
-\begin{equation}
-  \label{eqn:unit}
-  F[m]^TF[m] = \frac{1}{(4\pi r)^2}{\bf 1}_{[t_{\rm min}-mr,  
-    t_{\rm max}-mr]}
-\end{equation}
-in which ${\bf 1}_{S}$ denotes
-multiplication by the characteristic function of a measurable 
-$S \subset \bR$.
-
-Therefore the normal equation for the minimizer on the RHS of equation \ref{eqn:redexp} is
-\begin{equation}
-  \label{eqn:norm1}
-  \left(\frac{1}{(4\pi r)^2} {\bf 1}_{[t_{\rm min}-mr,  
-      t_{\rm max}-mr]} + \alpha^2 A^TA\right)w= F[m]^Td.
-\end{equation}
-
-With these choices, the normal equation \ref{eqn:norm1} becomes
-\begin{equation}
-\label{eqn:norm2}
-\left(\frac{1}{(4\pi r)^2}  {\bf 1}_{[t_{\rm min}-mr,  
-      t_{\rm max}-mr]} + \alpha^2a^2\right)w= F[m]^Td.
-\end{equation}
-
-\begin{proposition}
-  \label{thm:norminvexp}
-  Assume the conditions \ref{eqn:mod}, \ref{eqn:fullrec},
-  \ref{eqn:annmult}. Also assume that $\lambda \in (0,\lambda_{\rm
-    max}], \alpha > 0,$ and that  $C>0$ exists so that $a \in L^{\infty}(\bR)$
-  mentioned in condition \ref{eqn:annmult} satisfies condition
-  \ref{eqn:abnd}.
-%  \begin{equation}
-%   \label{eqn:abnd}
-%    |t| > \lambda \Rightarrow a(t) \ge C,
-%  \end{equation}
-%  in which $C>0$ may depend on $\lambda$.
-  Then
-  \begin{itemize}
-  \item[1. ]the normal operator $F[m]^TF[m] + \alpha^2A^TA$ is
-    invertible for any $m \in M$, $\alpha > 0$;
-  \item[2. ]the solution $\aw[m;d]\in W$ of the normal equation
-    \ref{eqn:norm} is given by
-    \begin{equation}
-      \label{eqn:normsol}
-      \aw[m;d](t) = \left\{
-        \begin{array}{c}
-          \left(\frac{1}{(4\pi r)^2} + \alpha^2
-          a^2(t)\right)^{-1}\frac{1}{4 \pi r}d(t+mr), t \in [t_{\rm
-          min}-mr, t_{\rm max}-mr];\\
-          0, \mbox{ else;}
-        \end{array}
-      \right.
-    \end{equation}
-  \item[3. ]if in addition $d=F[m_*]w_*, w_* \in \lW$ is noise-free, as in equation
-    \ref{eqn:defdatanonoise},
-    \begin{equation}
-      \label{eqn:solnonoise}
-      \aw[m,d](t)= \left(1+ (4\pi r)^2\alpha^2 a(t)^2\right)^{-1}w_*\left(t+(m-m_*)r\right).
-    \end{equation}
-  \end{itemize}
-\end{proposition}
-
-\begin{proof} of Theorem \ref{thm:norminv}:
-  \begin{itemize}
-  \item[1. ]Note that thanks to \ref{eqn:zeroinc}, if $|t|\le
-    \lambda \le \lambda_{\rm max}$, then ${\bf 1}_{[t_{\rm min}-mr,  
-      t_{\rm max}-mr]}(t) = 1$, whereas if $|t|>\lambda$,
-    then $a(t) \ge C$, whence
-    \[
-      \frac{1}{(4\pi r)^2}  {\bf 1}_{[t_{\rm min}-mr,  
-        t_{\rm max}-mr]} + \alpha^2a^2  \ge \min\{(4\pi r)^2,
-      \alpha^2\min\{1/(4\pi r)^2,C^2\} \}> 0.
-    \]
-    Therefore the normal operator is invertible under the stated
-    conditions.
-
-  \item[2. ]From the identity \ref{eqn:tran}.
-    \[
-      \mbox{supp }F[m]^Td \subset [t_{\rm min}-mr,t_{\rm max}-mr].
-    \]
-    Define $w_{\rm tmp}$ to be the right-hand side of equation \ref{eqn:normsol}. Then
-    from the previous observation and identity \ref{eqn:tran},
-    \[
-      \mbox{supp }w_{\rm tmp} \subset [t_{\rm min}-mr,t_{\rm max}-mr].
-    \]
-    From the identity \ref{eqn:unit}, for any $w \in W$,
-    \[
-      t \in [t_{\rm min}-mr,t_{\rm max}-mr] \Rightarrow F[m]^TF[m]w(t)
-      = \frac{1}{(4 \pi r)^2}w(t).
-    \]
-    It follows from this and the previous two observations that
-    $w_{\rm tmp}$ solves the normal equation \ref{eqn:norm}, and
-    therefore that $\aw[m;d]=w_{\rm tmp}$.
-
-  \item[3. ]Follows by inserting the definition
-    \ref{eqn:defdatanonoise} of $d$ in \ref{eqn:normsol} and
-    rearranging.
-  \end{itemize}
-\end{proof}
 
@@
 
-\begin{proposition}
-  \label{thm:epjgen}
-  Assume the hypotheses of Proposition \ref{thm:norminvexp}. Then
-  \begin{equation}
-  \label{eqn:residnormgen}
-  e[m,\aw[m,d];d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r \alpha a(t-mr))^4(1 +
-  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)^2
-\end{equation}
-\begin{equation}
-  \label{eqn:anninormgen}
-  p[m,\aw[m,d];d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r a(t-mr))^2(1 +
-  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)
-\end{equation}
-
-\begin{equation}
-  \label{eqn:expjgen}
-\tJa[m;d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}}\,dt\,(4\pi r \alpha a(t-mr))^2(1+(4\pi r \alpha 
-a(t-mr))^2)^{-1}d(t)^2. 
-\end{equation}
-\end{proposition}
-
-\begin{proof}
-  From equation \ref{eqn:normsol},
-  \[
-    F[m]\aw[m;d](t) = 
-    \frac{1}{4 \pi r}\left(\frac{1}{(4\pi r)^2} + \alpha^2
-      a^2(t-mr)\right)^{-1}\frac{1}{4 \pi r}d(t),
-  \]
-  so
-  \[
-    (F[m]\aw[m;d]-d)(t) = (1 + (4 \pi r\alpha
-    a(t-mr))^2)^{-1}-1)d(t)
-  \]
-  \[
-    = -(4 \pi r\alpha a(t-mr))^2(1 + (4 \pi r\alpha
-    a(t-mr))^2)^{-1}d(t).
-  \]
-  Half the integral of the square of this data residual is
-  $e[m,\aw[m;d],d]$, which proves identity \ref{eqn:residnormgen}.
-
-  To compute $p[m,\aw[m;d],d]$, note that
-  \[
-    A\aw[m;d](t)=a(t) \left(\frac{1}{(4\pi r)^2} + \alpha^2
-      a^2(t)\right)^{-1}\frac{1}{4 \pi r}d(t+mr)
-  \]
-  \[
-    = 4\pi r a(t) (1 + (4\pi r \alpha a(t))^2)^{-1}d(t+mr)
-  \]
-  for $ t \in [t_{\rm min}-mr, t_{\rm max}-mr]$, so squaring,
-  integrating, and changing integration variables $t \mapsto t-mr$
-  gives the result \ref{eqn:anninormgen}
-
-  That the VPM objective $\tJa$ is given by \ref{eqn:expjgen} follows from equations \ref{eqn:pen},
-  \ref{eqn:redexp}, \ref{eqn:residnormgen}, and
-  \ref{eqn:anninormgen}.
-\end{proof}
-
-\begin{proof} of Theorem \ref{thm:diffobj}:
-Suppose first that $a \in C^1(\bR)$. Differentiation under the integral sign  
-  yields the expression for its derivative:
-\begin{equation}
-  \label{eqn:dexpjgen}
-  \frac{d}{dm}\tJa[m;d] = -\frac{(4 \pi r \alpha)^2}{\|d\|^2} \int_{t_{\rm min}}^{t_{\rm max}} \,dt \, 
-  \left(a\frac{da}{dt}\right)(t-mr)(1+(4\pi r \alpha 
-  a(t-mr))^2)^{-2}d(t)^2. 
-\end{equation}
-For $a \in W^{1,\infty}_{\rm loc}(\bR)$ a limiting argument shows that the
-same expression gives the derivative of $\tJa$.
-\end{proof}
-
 \subsection{Proof of Theorem \ref{thm:boxcarbad}}
-\begin{corollary}
-  \label{thm:epjnonoise}
-  Assume the hypotheses of Proposition \ref{thm:norminvexp}, item
-  3. Then noting that $\|d\| = \|w_*\|/(4 \pi r)$
-\begin{equation}
-  \label{eqn:residnorm}
-  e[m,\aw[m,d];d] 
-= \frac{\alpha^4}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^4(1+(4\pi r)^2 \alpha^2 
-    a(t-(m-m^*)r)^2)^{-2}w_*(t)^2.
-\end{equation}
-\begin{equation}
-  \label{eqn:anninorm}
-  p[m,\aw[m,d];d] = \frac{(4\pi r)^2}{2\|w_*\|^2}\int \,dt\,  a(t-(m-m_*)r)^2 \left(1+ (4\pi r)^2\alpha^2
-    a(t-(m-m_*)r)^2\right)^{-2}w_*(t)^2.
-\end{equation}
-so 
-\begin{equation}
-\label{eqn:expjnonoise}
-\tJa[m;d] = \frac{(4\pi r \alpha)^2}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
-  a(t-(m-m_*)r)^2)^{-1}w_*(t)^2. 
-\end{equation}
-Finally, if $a \in W^{1,1}_{\rm loc}(\bR)$, then $\tJa[\cdot;d]$ is differentiable, and 
-\begin{equation}
-  \label{eqn:dexpjnonoise}
-  \frac{d}{dm}\tJa[m;d] = -\frac{r (4\pi r \alpha)^2}{\|w_*\|^2} \int \,dt \, 
-  \left(a\frac{da}{dt}\right)(t-(m-m_*)r)(1+(4\pi r)^2 \alpha^2 
-  a(t-(m-m_*)r)^2)^{-2}w_*(t)^2. 
-\end{equation}
-\end{corollary}
 

### 2020-12-07 repro/papers/skip/abstract.tex
initial

Speaker: William W. Symes

Title: Extended Inversion: when it works, and why

Abstract: Model extension is the basis for several classic seismic velocity analysis techniques. In the last couple of decades, various inversion methods based on model extension have been suggested as approaches to eliminating or reducing the tendency of Full Waveform Inversion to cycle-skip. This talk will illustrate the extended inversion concept by applying it to perhaps the simplest waveform inversion problem imaginable: to estimate the sound velocity in a homogeneous fluid from recording of a transient pulse propagated from a single source to a single receiver at known offset.  While cartoonishly simple, this single trace transmission problem exhibits key difficulties of industrial scale inversion - in particular, it provides an often-used illustration of cycle-skipping. I will present complete analyses of Full Waveform Inversion, Wavefield Reconstruction Inversion, and another extended inversion method similar to Adaptive Waveform Inversion, applied to single trace transmission. The mathematics is straightforward enough that the analysis can be performed ``on paper'': it is possible to determine precisely why each method overcomes cycle-skipping - or does not. In fact, only one of these approaches ``works'', that is, does not cycle-skip. To conclude, I'll present a few preliminary results from a generalization of the ``good'' method to a 2D crosswell velocity estimation problem, and suggest that this type of extended inversion may eventually be feasible on an industrial scale, with cost comparable to FWI's.

Acknowledgements: I thank many students, colleagues and collaborators who have helped me learn whatever it is that I know about this subject. The work reported in this talk is in part the result of a collaboration with Prof. Susan E. Minkoff and graduate assistant Huiyi Chen at University of Texas - Dallas.

### 2020-12-07 repro/papers/skip/slides.tex
initial

\documentclass[xcolor=dvipsnames,12pt,aspectratio=169]{beamer}
\usecolortheme[named=violet]{structure}

%\usetheme{umbc2} 
\usepackage{pgf}
%\usepackage[pdftex]{graphicx}
\usepackage{color}
\usepackage{amssymb,amsmath}
\usepackage[english]{babel}
\mode<presentation>

 \usepackage{multimedia}
\usefonttheme{structurebold}

%\useinnertheme{umbcboxes}

%\setbeamercolor{umbcboxes}{bg=violet!15,fg=black}

\setbeamercovered{transparent}
\setbeamertemplate{navigation symbols}{}
\input{./math-macros}
\setlength{\parskip}{2ex}   % place a blank line between paragraphs
\def\Logo{\@ifnextchar(\beamer@Logo\logo}
\def\beamer@Logo(#1,#2){\logo}

\AtBeginSection[]{
{\setbeamertemplate{logo}{}
 \begin{frame}<beamer>
    \frametitle{Agenda}
    \tableofcontents[currentsection]
  \end{frame}
}
}
% Images
% Macros
\newcommand{\peq}{\,+\hspace{-0.15cm}=}
\newcommand{\bm}{\bar{m}}
\newcommand{\dm}{\delta m}
\newcommand{\dom}{\delta \bar{m}}
\newcommand{\doma}{\delta \bar{m}_{\alpha}}
\newcommand{\om}{\bar{m}}
\newcommand{\bH}{{\bf H}}
\newcommand{\bR}{{\bf R}}
\newcommand{\bcF}{{\bar{\cal F}}}
\newcommand{\cF}{{\cal F}}
\newcommand{\oF}{\bar{F}}
\newcommand{\odF}{\bar{F}^{\dagger}}
\newcommand{\oS}{\bar{S}}
\newcommand{\oM}{\bar{M}}
\newcommand{\Ja}{J_{\alpha}}
\newcommand{\Na}{N_{\alpha}}
\newcommand{\dd}{\delta d}
\newcommand{\ddl}{\delta d_{\alpha}}

\title[]{Extended Inversion: when it works, and why}
\author[]{William W. Symes}
\institute[]{Rice University}
\date{December 2020}

\begin{document}
\frame{
\titlepage
}

\begin{frame}\frametitle{How to Cycle-Skip, in one easy lesson}
\only<1>{
Single trace transmission in a fluid:

Point source at $\bx=\bx_s$ radiates with transient intensity $w(t)$ (``wavelet'')

Resulting pressure wave propagates to point receiver at $\bx=\bx_r$, offset $|\bx_r-\bx_s|=h$

Pressure trace predicted by linear acoustics:
\[
p(t) = \frac{1}{4\pi h}w(t-mh) = (F[m]w)(t)
\]
where $m =$ pressure wave slowness ($=1/v_p$), $F$ = modeling operator - linear in w, nonlinear in m
}
\only<2>{
Inverse problem: given pressure trace $d(t)$, wavelet $w(t)$, and offset $h$, find slowness $m$ so that
\[
F[m] w  \approx d
\]
Standard approach: turn it into an optimization - minimize over $m$ the {\em mean square error}
\[
\frac{1}{2}\int\,dt\, (F[m]w(t) -d(t))^2 = \frac{1}{2}\|F[m]w-d\|^2 = J_{\rm FWI}[m;w,d]
\]

= world's simplest FWI task
}
\only<3>{
Asymptotic transient-ness:

Mother wavelet $w_1$: $w_1(t)=0, |t|>1$ and $\int w_1^2 = 1$

Scaled wavelets $w_{\lambda} = \frac{1}{\sqrt{\lambda}}w_1(t/\lambda)$: $w_{\lambda}(t)=0, |t|>\lambda$ and $\int w_{\lambda}^2 = 1$

NB: if center frequency of $w_1 = \omega_1$, then center frequency of $w_{\lambda} = \omega_1/\lambda$

Theorem Uh-Oh: If $d = F[m_*]w_{\lambda}$ (noise-free transient data) and $|m-m_*| > 2\lambda/r$, then
\[
J_{\rm FWI}[m;w_{\lambda},d] = \|d\|^2, \,\, \nabla_m J_{\rm FWI}[m;w_{\lambda},d] = 0
\]
}
\only<4>{
$J_{\rm FWI}[m;w_{\lambda},d]$, $d=F[m_*]w_{\lambda}$, $w_{\lambda}$ = Ricker wavelet center frequency = $1/\lambda$, $m_*=0.4$ s/km, $h=1 km$
\begin{center}
\includegraphics[height=4cm]{Fig/FWI.png}
\end{center}
Red: $\lambda=0.05$ (20 Hz), Blue: $\lambda = 0.25$ (40 Hz) [H. Chen et al. SEG 20]
}
\only<5>{
Postmortem:

\begin{itemize}
\item for most $m$, predicted data $F[m]w$ is not close to target data $d$
\item for most $m$, gradient is not a useful search direction
\item shape of objective function depends strongly on data frequency content - large residual, useless gradient more common for high frequencies (small $\lambda$)
\item flip side: low frequency (big $\lambda$) $\Rightarrow$ useful updates for larger set of initial models
\end{itemize}
}
\end{frame}

\begin{frame}\frametitle{Wavefield Reconstruction Inversion}
\only<1>{
Extended inversion: {\em enlarge the model space} = add (nonphysical) degrees of freedom, converge to point in original model space

WRI - van Leeuwen \& Herrmann GJI 13, IP 16, many other papers - most studied extended inversion approach

Idea: view wave equation as {\em weak constraint} $\sim$ {\em add source parameters}

Rationale: more likely to be able to 
\begin{itemize}
\item find low residual (extended) models (``hug your data''),
\item explore useful update directions
\end{itemize}
}
\only<2>{
Cast: 
\begin{itemize}
\item model parameters $m$ (eg. $m$=slowness)
\item Wave operator $L[m]$ (eg. constant density acoustic op
\[
\left. L[m]=m^2\frac{\partial^2}{\partial t^2} - \nabla^2\right)
\]
\item Known (!) source field $q(\bx,t)$ (eg. point source $w(t)\delta(\bx-\bx_s)$
\item dynamic fields $u(\bx,t)$ solve $L[m]u=q$ (eg. $u=$ pressure
\item Sampling operator $P$ extracts data trace(s) (pressure or ...) from $u$
\item Modeling operator $F[m]=PL[m]^{-1}$
\end{itemize}
}
\only<3>{
FWI: given $d$, $q$, 
\[
\min_m \|Pu-d\|^2 \mbox{ subj to } L[m]u=q
\]
\[
\Leftrightarrow \min_m \|F[m]q-d\|^2
\]

WRI: given $d$, $q$,
\[
\min_{m,u} \|Pu-d\|^2 + \alpha^2 \|L[m]u-q\|^2
\]

(NB: $\alpha$ small $\Rightarrow$ emphasis on 1st term $\Rightarrow$ ``hug your data'')
}
\only<4>{
equivalent (eg. Yingst \& Wang SEG 16):

given $d$, $q$, set $g=L[m]u-q$,

then $Pu-d = PL[m]^{-1}(g+q)-d = F[m]g + e[m]$

$e[m]=F[m]q-d$ = usual data residual

WRI: given $q,d$, 
\[
\min_{m,g} \|F[m]g+e[m]|^2 + \alpha^2 \|g|^2
\]
}
\only<5>{
Simplification 1:

Variable Projection (Golub \& Pereyra 73, 03): eliminate $g$ by solving quadratic minimization
\[
J_{\rm WRI}[m;q,d] = \min_{g} \|F[m]g+e[m]\|^2 + \alpha^2 \|g\|^2
\]
minimizer = $g[m;q,d]$, then
\[
=\|F[m]g[m;q,d]+e[m]|^2 + \alpha^2 \|g[m;q,d]\|^2
\]
$m$ minimizes $J_{\rm WRI}$ $\Leftrightarrow$ $(m,g[m;q,d])$ solves WRI problem 
}
\only<6>{
Simplification 2:

Non-radiating sources = $\{g: F[m]g=0\}$ = null space of $F[m]$

Rank-Nullity Theorem: $g = n + F[m]^Ts$, $s \in $ range of $F$ = data, $n$ = non-radiating, {\em and}
$n$ and $F[m]^Ts$ are orthogonal
\[
\|F[m]g+e[m]\|^2 + \alpha^2 \|g\|^2 = \|F[m](F[m]^Ts + n)+e[m]\|^2 + \alpha^2\|F[m]^Ts + n\|^2
\]
\[
= \|F[m]F[m]^Ts+e[m]\|^2 + \alpha^2\|F[m]^Ts\|^2 + \alpha^2\|n\|^2
\]
so
\[
J_{\rm WRI}[m;q,d] = \min_s \|F[m]F[m]^Ts-e[m]\|^2 + \alpha^2\|F[m]^Ts\|^2 
\]
}
\only<7>{
Simplification 3:

Minimization over $s$ $\Leftrightarrow$ solution of normal equation
\[
((F[m]F[m]^T)^2 + \alpha^2 F[m]F[m]^T)s = F[m]F[m]^Te[m]
\]
Substitute solution into def of $J_{\rm WRI}$, do a page of algebra (IP, v. 36 no. 10 (2020)), then...
\[
J_{\rm WRI} = \alpha^2 e[m]^T(F[m]F[m]^T + \alpha^2 I)^{-1} e[m]
\]
compare:
\[
J_{\rm FWI} = e[m]^T e[m]
\]
{\color{blue} $J_{\rm WRI}$ is weighted variant of $J_{\rm FWI}$, weight operator = $\alpha^2(F[m]F[m]^T + \alpha^2 I)^{-1} $}
}
\only<8>{
Apply to single-trace transmission:
\[
F[m]g(t) = \int_{r \le R}\,dx\,\frac{g(t-mr)}{4 \pi r}, \, F[m]^Ts(\bx,t) = 
\left\{\begin{array}{c}
\frac{s(t+mr)}{4 \pi r}, r\le R,\\
0, else
\end{array}
\right.
\]
($r = |\bx-\bx_r|$ and $g=0$ if $r > R$), so 
\[
F[m]F[m]^T s(t) = \frac{R}{4\pi} s(t),\,\,\alpha^2(F[m]F[m]^2 + \alpha^2 I)^{-1} =\frac{4\pi\alpha^2}{R + 4\pi\alpha^2}I
\]
}
\only<9>{
Theorem Uh-Oh \# 2:
\[
J_{\rm WRI}[m;w(t)\delta(\bx-\bx_s),d] = \frac{4\pi\alpha^2}{R + 4\pi\alpha^2} J_{\rm FWI}[m;w,d]
\]

Postmortem:
\begin{itemize}
\item WRI just as skippy as FWI!!!
\item enlarging the search space not enough
\item ``hugging your data'' also not enough: WRI cycle-skips for small $\alpha$
\item conclusion not limited to single trace transmission: similar for other transmission IPs, eg. diving wave inversion
\end{itemize}
}
\end{frame}

\begin{frame}\frametitle{Matched Source Extension}
\only<1>{
Back to FWI setting:
\[
F[m]w(t) = \frac{1}{4 \pi h}w(t-mh)
\]
A different extended inversion: add wavelet to unknowns

However, $F[m]$ is invertible for every $m$, so need constraint on $w$

Idea: after signature decon, wavelet should be {\em compact} $\sim$ nonzero only near $t=0$ - inspires {\em Matched Source Objective}
\[
\min_{m,w} \|F[m]w-d\|^2 + \|Aw\|^2, \,\,Aw(t)=tw(t)
\]
(S. 94, Plessix et al. 99, similar to Adaptive Waveform Inversion (AWI) Warner \& Guatsch 14)
}
\end{frame}
\end{document}

### 2020-12-07 repro/papers/discrep/paper.tex
current

@@
   local optimization methods. The data consist of a single trace
-  recorded at a given distance from a point source. The velocity or
-  slowness is presumed homogeneous, and the target source wavelet is
-  presumed quasi-impulsive or localized near zero time lag. The source
-  is extended by permitting energy to spread in time, and the spread
-  is controlled by adding a weighted mean square of the extended
-  source wavelet to the data misfit, to produce the extended inversion
-  objective. The objective function and its gradient can be computed
-  explicitly, and it is easily seen that the error in any local
-  minimizer is bounded by a multiple of the permitted energy spread,
-  which may be interpreted as a wavelength. The derivation shows several
-  important features of all similar extended source algorithms. For
-  example, nested optimization, with the source estimation in the
-  inner optimization (variable projection method), is essential. The
-  choice of the penalty operator, controlling the extended source
-  degrees of freedom, is critical: in order to produce an objective
-  immune from cycle-skipping, the penalty operator must be
-  (pseudo-)differential. The penalty weight is chosen dynamically,
-  according to a version of the Discrepancy Principle, in order to
-  accelerate convergence to a near-global minimum. Like many other
-  extended formulations, the example problem considered here permits
-  the discrepancy criterion to be satisfied with zero penalty
-  weight. This unusual feature removes a common obstacle to the use of
-  the discrepancy principle, namely the determination of a
-  satisfactory initial estimate of the penalty weight: the initial weight may
-  be set to zero, and the dynamic algorithm updates the weight until
-  convergence is achieved.
+  extracted from a causal pressure field, propagating in a fluid
+  according to linear acoustics and recorded at a given distance from
+  a transient point energy source. The velocity or slowness of the
+  fluid is presumed homogeneous, and the target source intensity
+  (``wavelet'') is presumed quasi-impulsive, with zero energy for time
+  lags greater than a specified maximum lag (specified support). The
+  data are assumed noisy, with known RMS signal-to-noise ratio. The
+  inverse problem is: from the recorded trace, recover both the sound
+  velocity and source wavelet with specified support, so that the data
+  is fit with prescribed RMS relative error. After showing explicitly that
+  the obvious least squares objective has multiple large residual
+  minimizers, we introduce an extended approach. The source is extended by
+  permitting energy to spread in time. The spread is controlled by
+  adding a weighted mean square of the extended source wavelet to the
+  data misfit, to produce the extended inversion objective. We choose
+  a weight operator to penalize source energy away from zero lag via
+  multiplication by the time lag. The resulting objective function and
+  its gradient can be computed explicitly. We show that the error in 
+  slowness component of any local minimizer is bounded by multiple of
+  the maximum lag and the assumed noise level. In particular, there
+  are no spurious local minima. We also show that the projection of
+  the extended wavelet onto the interval of support yields a solution
+  of the inverse problem as posed above, with appropriate choice of
+  maxiumum lag and RMS relative error. In several appendices, we
+  describe a constructive algorithm for controlling the penalty
+  weight, the relation between maxiumum lag and RMS central frequency
+  via the Heisenberg principle, and some structural details which the
+  very simple inverse problem discussed here shares with more
+  prototypical problems in industrial and academic seismology.
+
+  %which may be interpreted as a wavelength.
+  %The derivation shows
+  %several important features of all similar extended source
+  %algorithms. For example, nested optimization, with the source
+  %estimation in the inner optimization (variable projection method),
+  %is essential. The choice of the penalty operator, controlling the
+  %extended source degrees of freedom, is critical: in order to produce
+  %an objective immune from cycle-skipping, the penalty operator must
+  %be (pseudo-)differential. The penalty weight is chosen dynamically,
+  %according to a version of the Discrepancy Principle, in order to
+  %accelerate convergence to a near-global minimum. Like many other
+  %extended formulations, the example problem considered here permits
+  %the discrepancy criterion to be satisfied with zero penalty
+ % weight. This unusual feature removes a common obstacle to the use of
+  %the discrepancy principle, namely the determination of a
+  %satisfactory initial estimate of the penalty weight: the initial
+ % weight may be set to zero, and the dynamic algorithm updates the
+ % weight until convergence is achieved.
 
@@
 
+\begin{proof} of Theorem \ref{thm:boxcarbad}
+
+ This is clear from the definition $a = 1-{\bf 1}_{-\lambda,\lambda]}$ and equation \ref{eqn:expjnonoise}, as the supports of $w_*$ and ${\bf
+    1}_{[-\lambda,\lambda]}(t-(m-m_*))$ are disjoint for the range of
+  $m$ identified in the theorem.
+\end{proof}
+
 \noindent {\bf Remark.}
@@
 
-\begin{theorem}
-  \label{thm:mnoiseres}
-  Asumme that
-  \begin{itemize}
-  \item[1. ] $\alpha, \mu> 0$,
-  \item[2. ] $m_* \in M, w_* \in W_{\mu}$,
-  \item[3. ] $d_* = F[m_*]w_*$,
-  \item[4. ] $n \in D$ and $d = d_* + n$.
-  \end{itemize}
-  Set $\eta = \|n\|/\|d_*\|$, and assume further that
-  \begin{itemize}
-  \item[5. ]
-    \begin{equation}
-      \label{eqn:mnoisecond}
-      \eta < \frac{\sqrt{5}-1}{2}, and
-    \end{equation}
-  \item[6. ] $m$ is a stationary point of $\tJa[\cdot;d]$.
-  \end{itemize}
-  Then
-  \begin{equation}
-    \label{eqn:mnoisesuff}
-    |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}.
-  \end{equation}
-\end{theorem}
-
-\noindent {\bf Remark.} That is, {\em with the choice of penalty
-  multiplier $a$ given in equation \ref{eqn:ann}, and support radius
-  $\mu$ of the ``noise-free'' wavelet $w_*$, $\Ja$ has no local
-  minima with slownesses further than $(1+O(\eta))\mu/r)$ from the slowness used to
-  generate the data}.
-
-\noindent {\bf Remark.} The estimate $|m-m_*|r<\mu(1+O(\eta))$ for
-local minima of $\Ja$ is sharp: it is possible to choose $w_* \in W_{\mu}$
-so that $\mu - |m-m_*|r$ is as small as you like. In particular,
-the ``exact'' or ``true'' slowness $m_*$ is not necessarily the only
-slowness component of a local minimizer, or even the slowness
-component of any local minimizer, and in particular
-is not (necessarily) the slowness component of a global minimizer of $\Ja$.
-
-\noindent {\bf Remark.} No similar bound could hold for much larger
-noise levels than specified in condition \ref{eqn:mnoisecond}, the
-right-hand side of which is a bit larger than 0.6. For example, if the noise is
-the predicted data for the same wavelet $w_*$ with a substantially different
-slowness $m_{\flat}$, that is, $n=F[m_{\flat}]w_*$, then a simple
-symmetry argument shows that if there is a local minimizer of $\Ja[\cdot,\cdot;d_*+n]$. with
-slowness near $m_*$,
-there must also be a minimizer with slowness near $m_{\flat}$.
-so that the difference with $m_*$ is not constrained at
-all by the assumed support radius of $w_*$. So for this example with 100\% noise, no
-bound of the type given by conclusion 2 could possibly hold. We will
-illustrate this phenomenon numerically later in the paper.
-
-\noindent{\bf Remark.} Note that $\alpha$ plays no role in the
-conclusions of this theorem. It is only required that $\alpha >0$.
-
-Unless the data is noise-free, there is no reason to suppose that the
-estimated wavelet $\aw[m;d]$ (Theorem \ref{thm:norminv}) will lie in
-$\lW$, unless the support of the noise $n$ is not restricted. In order
-to construct a solution of the inverse problem \ref{eqn:probstat0}, we
-project $\aw[m;d]$ onto $\lW$. For sufficiently large $\lambda,
-\epsilon$, the result is a solution of the inverse problem:
-
-\begin{theorem}
-  \label{thm:ipnoisesuf}
-  Assume the hypotheses of Theorem \ref{thm:mnoiseres}, and that
-  inequality \ref{eqn:mnoisecond} holds, and $\mu \in
-  (0,\lambda_{\rm max}]$. Then the pair
-  \[
-    (m,{\bf 1}_{[-\lambda,\lambda]}\aw[m,d])
-  \]
-  solves the inverse problem as stated in \ref{eqn:probstat0} if
-  \begin{equation}
-    \label{eqn:ipnoiselam}
-    \left( 2+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\mu \le \lambda
-    \le \lambda_{\rm max}, 
-  \end{equation}
-  and
-  \begin{equation}
-    \label{eqn:ipnoiseeps}
-    \epsilon \ge (4 \pi r \alpha\lambda)^2+\eta. 
-  \end{equation}    
-\end{theorem}
-
-\noindent {\bf Remark:} Note that the sufficient condition \ref{eqn:ipnoiselam} for
-$\lambda$ is independent of $\alpha$. It follows that for any choice
-of $\lambda$ consistent with this bound, $(m,{\bf
-  1}_{[-\lambda,\lambda]}\aw[m,d])$ is a solution of the inverse
-problem for any $\epsilon > 0$ provided that $\alpha$ is chosen
-sufficiently small ($O(\sqrt{\epsilon})$).
-
-\subsection{Proof of Theorem \ref{thm:fwi}}
-
-%A natural formulation of the Problem Statement \ref{eqn:probstat0} is
-%via least squares: that is, given $d \in D$, $\lambda >0$, find
-%$(m,w) \in M \times \lW$ to minimize
-%$\lerr: M \times W \rightarrow \bR$, defined in display
-%\ref{eqn:redms}.  A pair $(m,w) \in M\times \lW$ satisfies the
-%condition \ref{eqn:probstat0} iff $\lerr[m,w;d] \le \epsilon^2/2$, so a
-%solution exists iff a global minimizer of $\lerr$ satisfies this
-%inequality.
-
-%In this discussion, I will presume that $d$ is 
-%noise-free, as defined at the end of the last section (equation
-%\ref{eqn:defdatanonoise}). Thus {\em a priori} a global minimizer of
-%$e$ is a solution of the Problem Statement for any
-%$\epsilon > 0$. The only
-%question is how to find such a global minimizer. As mentioned in the
-%introduction, the simple model problem studied here is a proxy for
-%large-scale data fitting problems for which the only computationally
-%feastible approach to minimization of analogs of $e$
-%is some form of descent-based iterative local optimization. Any such
-%algorithm requires the specification of an initial iterate
-%$(m_0,w_0)$. Therefore the key question is: what
-%constraints must be imposed on $(m_0,w_0)$ in order that the resulting
-%sequence of iterates convertes to a global minimizer?
-
-%\begin{theorem}
-%  \label{thm:fwi}
-%  Suppose that $0 <\lambda\le \lambda_{\rm max}$,  $m_* \in M, w_*
-%  \in \lW, d=\lF[m_*]w_*$ is noise-free data per definition \ref{eqn:defdatanonoise},
-%  Under assumption \ref{eqn:fullrec}, for any $m \in M$ with $r|m-m_*|>2\lambda$,
-%\begin{equation}
-%  \label{eqn:isovpm}
-% \min_w e[m,w;d]=e[m,0,d]=\frac{1}{32\pi^2 r^2}\|w_*\|^2,
-%\end{equation}
-%and any such $(m,0)$ is a stationary point of $e$.
-%\end{theorem}
-
-%\begin{theorem}
-%\label{thm:red}
-%$\Ja[m,\cdot;d]$ is invertible:
-%\begin{equation}
-%\label{eqn:norminv}
-%F[m]^TF[m] +\alpha^2 A^TA \in {\cal I}(W,W).
-%\end{equation}
-%(The superscript $T$ denotes the adjoint with respect to the inner products 
-%in $W=L^2(\bR)$ and $D=L^2([t_{\rm min},t_{\rm max}])$.)
-%Then the unique stationary point $\aw[m;d] \in W$ of
-%$\Ja[m,\cdot;d]$ is a global minimizer, and is the solution of the {\em normal equation}:
-%\begin{equation}
- % \label{eqn:norm}
-  %(F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td, 
-%\end{equation}
-%The reduced or variable projection objective $\tJa$ defined in
-%equation \ref{eqn:redexp} is also given by
-%\begin{equation}
- % \label{eqn:redexp1}
- % \tJa[m;d] =\Ja[m,w_{\alpha}[m;d];d] = \min_w \Ja[m,w;d]
-%\end{equation}
-%Then $(m,w)$ is a local minimizer of $\Ja[\cdot,\cdot;d]$ if and only
-%if $m$ is a local minimizer of $\tJa[\cdot;d]$ and $w=\aw[m;d]$.
-%\end{theorem}
-
-%\noindent{\bf Remark:} The choice
-%of penalty weight $\alpha$ has a significant influence on the
-%performance of numerical optimization algorithms in minimizing
-%$\Ja$ and/or $\tJa$, but relatively little influence on the error estimates to
-%follow. For the discussion of these estimates, we assume only that
-%$\alpha>0$. Appendix A contains a description of an algorithm for
-%constructive determination of $\alpha$.
-
-\subsection{Proof of Theorem \ref{thm:norminv}}
-
-%\begin{proof} of Theorem \ref{thm:norminv}: Since the normal operator
-%  is invertible (item 1 of Proposition \ref{thm:norminvexp}), the
-%  conclusions of this theorem follow from the Lax-Milgram Theorem
-%  (\cite{Yosida}, III.7).
-%\end{proof}
-
-\subsection{Proof of Theorem \ref{thm:diffobj}}
-
-\subsection{Proof of Theorem \ref{thm:boxcarbad}}
-
-\begin{proof} of Theorem \ref{thm:boxcarbad}
-
- This is clear from the definition $a = 1-{\bf 1}_{-\lambda,\lambda]}$ and equation \ref{eqn:expjnonoise}, as the supports of $w_*$ and ${\bf
-    1}_{[-\lambda,\lambda]}(t-(m-m_*))$ are disjoint for the range of
-  $m$ identified in the theorem.
-\end{proof}
-
-\subsection{Proof of Theorem \ref{thm:mnoiseres}}
 \begin{proposition}
@@
 
+\begin{theorem}
+  \label{thm:mnoiseres}
+  Asumme that
+  \begin{itemize}
+  \item[1. ] $\alpha, \mu> 0$,
+  \item[2. ] $m_* \in M, w_* \in W_{\mu}$,
+  \item[3. ] $d_* = F[m_*]w_*$,
+  \item[4. ] $n \in D$ and $d = d_* + n$.
+  \end{itemize}
+  Set $\eta = \|n\|/\|d_*\|$, and assume further that
+  \begin{itemize}
+  \item[5. ]
+    \begin{equation}
+      \label{eqn:mnoisecond}
+      \eta < \frac{\sqrt{5}-1}{2}, and
+    \end{equation}
+  \item[6. ] $m$ is a stationary point of $\tJa[\cdot;d]$.
+  \end{itemize}
+  Then
+  \begin{equation}
+    \label{eqn:mnoisesuff}
+    |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}.
+  \end{equation}
+\end{theorem}
+
 \begin{proof} of Theorem \ref{thm:mnoiseres}:
@@
 \end{proof}
+
+\noindent {\bf Remark.} That is, {\em with the choice of penalty
+  multiplier $a$ given in equation \ref{eqn:ann}, and support radius
+  $\mu$ of the ``noise-free'' wavelet $w_*$, $\Ja$ has no local
+  minima with slownesses further than $(1+O(\eta))\mu/r)$ from the slowness used to
+  generate the data}.
+
+\noindent {\bf Remark.} The estimate $|m-m_*|r<\mu(1+O(\eta))$ for
+local minima of $\Ja$ is sharp: it is possible to choose $w_* \in W_{\mu}$
+so that $\mu - |m-m_*|r$ is as small as you like. In particular,
+the ``exact'' or ``true'' slowness $m_*$ is not necessarily the only
+slowness component of a local minimizer, or even the slowness
+component of any local minimizer, and in particular
+is not (necessarily) the slowness component of a global minimizer of $\Ja$.
+
+\noindent {\bf Remark.} No similar bound could hold for much larger
+noise levels than specified in condition \ref{eqn:mnoisecond}, the
+right-hand side of which is a bit larger than 0.6. For example, if the noise is
+the predicted data for the same wavelet $w_*$ with a substantially different
+slowness $m_{\flat}$, that is, $n=F[m_{\flat}]w_*$, then a simple
+symmetry argument shows that if there is a local minimizer of $\Ja[\cdot,\cdot;d_*+n]$. with
+slowness near $m_*$,
+there must also be a minimizer with slowness near $m_{\flat}$.
+so that the difference with $m_*$ is not constrained at
+all by the assumed support radius of $w_*$. So for this example with 100\% noise, no
+bound of the type given by conclusion 2 could possibly hold. We will
+illustrate this phenomenon numerically later in the paper.
+
+\noindent{\bf Remark.} Note that $\alpha$ plays no role in the
+conclusions of this theorem. It is only required that $\alpha >0$.
 
@@
 
-    % Recall that $\lE$ is the projection of $W$ onto $\lW$
-%(i.e. multiplication by ${\bf 1}_{[-\lambda,\lambda]}$.
-
-\subsection{Proof of Theorem \ref{thm:ipnoisesuf}}
+Unless the data is noise-free, there is no reason to suppose that the
+estimated wavelet $\aw[m;d]$ (Theorem \ref{thm:norminv}) will lie in
+$\lW$, unless the support of the noise $n$ is not restricted. In order
+to construct a solution of the inverse problem \ref{eqn:probstat0}, we
+project $\aw[m;d]$ onto $\lW$. For sufficiently large $\lambda,
+\epsilon$, the result is a solution of the inverse problem:
+
+\begin{theorem}
+  \label{thm:ipnoisesuf}
+  Assume the hypotheses of Theorem \ref{thm:mnoiseres}, and that
+  inequality \ref{eqn:mnoisecond} holds, and $\mu \in
+  (0,\lambda_{\rm max}]$. Then the pair
+  \[
+    (m,{\bf 1}_{[-\lambda,\lambda]}\aw[m,d])
+  \]
+  solves the inverse problem as stated in \ref{eqn:probstat0} if
+  \begin{equation}
+    \label{eqn:ipnoiselam}
+    \left( 2+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\mu \le \lambda
+    \le \lambda_{\rm max}, 
+  \end{equation}
+  and
+  \begin{equation}
+    \label{eqn:ipnoiseeps}
+    \epsilon \ge (4 \pi r \alpha\lambda)^2+\eta. 
+  \end{equation}    
+\end{theorem}
+
 \begin{proof} of Theorem \ref{thm:ipnoisesuf}:
@@
 \end{proof}
+
+\noindent {\bf Remark:} Note that the sufficient condition \ref{eqn:ipnoiselam} for
+$\lambda$ is independent of $\alpha$. It follows that for any choice
+of $\lambda$ consistent with this bound, $(m,{\bf
+  1}_{[-\lambda,\lambda]}\aw[m,d])$ is a solution of the inverse
+problem for any $\epsilon > 0$ provided that $\alpha$ is chosen
+sufficiently small ($O(\sqrt{\epsilon})$).
 

### 2020-12-08 repro/papers/skip/slides.tex
current
near done

@@
 
-Theorem Uh-Oh: If $d = F[m_*]w_{\lambda}$ (noise-free transient data) and $|m-m_*| > 2\lambda/r$, then
+Theorem Uh-Oh: If $d = F[m_*]w_{\lambda}$ (noise-free transient data) and $|m-m_*| > 2\lambda/h$, then
 \[
@@
 \only<4>{
-$J_{\rm FWI}[m;w_{\lambda},d]$, $d=F[m_*]w_{\lambda}$, $w_{\lambda}$ = Ricker wavelet center frequency = $1/\lambda$, $m_*=0.4$ s/km, $h=1 km$
+$J_{\rm FWI}$: $d=F[m_*]w_{\lambda}$, $m_*=0.4$ s/km, $w_{\lambda}$ = Ricker ctr freq = $1/\lambda$ Hz, $h=1$ km
+\vspace{-0.5cm}
 \begin{center}
-\includegraphics[height=4cm]{Fig/FWI.png}
+\includegraphics[height=5cm]{Fig/FWI.png}
 \end{center}
-Red: $\lambda=0.05$ (20 Hz), Blue: $\lambda = 0.25$ (40 Hz) [H. Chen et al. SEG 20]
+\vspace{-0.5cm}
+Red: $\lambda=0.05$ s (20 Hz), Blue: $\lambda = 0.025$ s (40 Hz) [H. Chen et al. SEG 20]
 }
@@
 \item model parameters $m$ (eg. $m$=slowness)
-\item Wave operator $L[m]$ (eg. constant density acoustic op
-\[
-\left. L[m]=m^2\frac{\partial^2}{\partial t^2} - \nabla^2\right)
-\]
+\item Wave operator $L[m]$ (eg. constant density acoustic op $L[m]=m^2\partial_t^2 - \nabla^2$)
 \item Known (!) source field $q(\bx,t)$ (eg. point source $w(t)\delta(\bx-\bx_s)$
-\item dynamic fields $u(\bx,t)$ solve $L[m]u=q$ (eg. $u=$ pressure
+\item dynamic fields $u(\bx,t)$ solve $L[m]u=q$ (eg. $u=$ pressure)
 \item Sampling operator $P$ extracts data trace(s) (pressure or ...) from $u$
@@
 \[
-\min_{m,g} \|F[m]g+e[m]|^2 + \alpha^2 \|g|^2
+\min_{m,g} \|F[m]g+e[m]\|^2 + \alpha^2 \|g\|^2
 \]
@@
 \[
-=\|F[m]g[m;q,d]+e[m]|^2 + \alpha^2 \|g[m;q,d]\|^2
+=\|F[m]g[m;q,d]+e[m]\|^2 + \alpha^2 \|g[m;q,d]\|^2
 \]
@@
 \]
+NB: $g$ is space-time field, $s$ is data - vast memory reduction, practical 3D time domain (Yingst \& Wang SEG 16, Rizzuti et al. SEG 19)
 }
@@
 
-\begin{frame}\frametitle{Matched Source Extension}
+\begin{frame}\frametitle{Matched Source Inversion}
 \only<1>{
@@
 \[
-\min_{m,w} \|F[m]w-d\|^2 + \|Aw\|^2, \,\,Aw(t)=tw(t)
+\min_{m,w} \|F[m]w-d\|^2 + \alpha^2\|Aw\|^2, \,\,Aw(t)=tw(t)
 \]
@@
 }
+\only<2>{
+Variable Projection: 
+\[
+J_{\rm MSI}[m;d] = \min_w \|F[m]w-d\|^2 +\alpha^2 \|Aw\|^2
+\]
+Minimizer $w[m;d]$ solves normal equation
+\[
+(F[m]^TF[m] + \alpha^2A^TA) w = F[m]^T d
+\]
+Explicit calculation:
+\[
+F[m]^TF[m] + \alpha^2A^TA = \left(\frac{1}{(4\pi h)^2} + \alpha^2 t^2\right)I, \,\,F[m]^Td(t) = \frac{1}{4\pi h}d(t+mh)
+\]
+}
+\only<3>{
+Plug it in, get explicit formulas for everything - in particular, if data is noise-free, $d=F[m_*]w_{\lambda}$, then
+\[
+\nabla_m J_{\rm MSI}[m;d] = -h \alpha^2 \int \,dt\, \frac{t-(m-m_*)h}{(1+(4\pi h \alpha (t-(m-m_*)h))^2)^2}w_{\lambda}(t)^2
+\]
+
+Theorem Ah-Ha! If $d=F[m_*]w_{\lambda}$,
+\[
+\nabla_m J_{\rm MSI}[m;d] \left\{\begin{array}{c}
+< 0 \mbox{ if }m < m_*-\lambda/h\\
+> 0 \mbox{ if }m > m_*+\lambda/h
+\end{array}
+\right.
+\]
+$\Rightarrow$ if $m$ is stationary point of $J_{\rm MSI}[\cdot;d]$, then $|m-m_*| \le \lambda$.
+}
+\only<4>{
+$J_{\rm MSI}$: $d=F[m_*]w_{\lambda}$, $m_*=0.4$ s/km, $w_{\lambda}$ = Ricker ctr freq = $1/\lambda$ Hz, $h=1$ km, $\alpha=1 s^{-1}km^{-1}$
+\vspace{-0.5cm}
+\begin{center}
+\includegraphics[height=5cm]{Fig/VPM.png}
+\end{center}
+\vspace{-0.5cm}
+Red: $\lambda=0.05$ s (20 Hz), Blue: $\lambda = 0.025$ s (40 Hz) [H. Chen et al. SEG 20]
+}
+\only<5>{
+$J_{\rm MSI} vs. J_{\rm FWI}$: $d=F[m_*]w_{\lambda}$, $m_*=0.4$ s/km, $w_{\lambda}$ = Ricker ctr freq = 40 Hz ($\lambda = 0.025$ s), $h=1$ km, $\alpha=1 s^{-1}km^{-1}$
+\vspace{-0.5cm}
+\begin{center}
+\includegraphics[height=5cm]{Fig/FWI-VPM-40.png}
+\end{center}
+\vspace{-0.5cm}
+Red: MSI, Blue: FWI [H. Chen et al. SEG 20]
+}
+\only<6>{
+$J_{\rm MSI} vs. J_{\rm FWI}$: $d=F[m_*]w_{\lambda}+n$, $n$ = 100\% RMS filtered stationary random noise, otherwise same
+\vspace{-0.5cm}
+\begin{center}
+\includegraphics[height=5cm]{Fig/FWI-VPM-noisy.pdf}
+\end{center}
+\vspace{-0.5cm}
+Red: MSI, Blue: FWI [H. Chen]
+}
 \end{frame}
+
 \end{document}

### 2020-12-09 repro/papers/skip/slides.tex
final

@@
 
-\begin{frame}\frametitle{How to Cycle-Skip, in one easy lesson}
+\begin{frame}\frametitle{Overview}
+Apply sophisticated methods to a simple inversion problem
+
+Why:
+\begin{itemize}
+\item clear understanding of what works and what doesn't, and why
+\item all details ``on paper''
+\item conclusions not justified only ``in rear view mirror'' by a few numerical examples
+\item road map for application to industrially relevant technology
+\end{itemize}
+
+{\em ``You can do this at home!''}
+
+Thanks to Prof. Sue Minkoff and GRA Huiyi Chen, UT Dallas
+\end{frame}
+
+\begin{frame}\frametitle{World's simplest waveform inversion problem}
 \only<1>{
@@
 \]
-where $m =$ pressure wave slowness ($=1/v_p$), $F$ = modeling operator - linear in w, nonlinear in m
+where $m =$ pressure wave slowness ($=1/v_p$), $F$ = modeling operator - linear in w, nonlinear in $m$
 }
@@
 \]
-Standard approach: turn it into an optimization - minimize over $m$ the {\em mean square error}
+Can solve by inspection! Use a ruler to measure time shift between $w(t)$ and $w(t-mh)$
+
+BUT let's use the standard approach instead: turn it into an optimization - minimize over $m$ the {\em mean square error}
 \[
@@
 \]
-
 = world's simplest FWI task
 }
-\only<3>{
+\end{frame}
+
+\begin{frame}\frametitle{How to cycle-skip, in one easy lesson}
+\only<1>{
 Asymptotic transient-ness:
@@
 NB: if center frequency of $w_1 = \omega_1$, then center frequency of $w_{\lambda} = \omega_1/\lambda$
+}
+\only<2>{
 
@@
 \]
+Proof.
+\begin{center}
+\vspace{-0.5cm}
+\includegraphics[height=3cm]{Fig/datas.pdf}\\
+Data: same wavelet, $m$ = 0.25 s/km (red) and $m_*$ = 0.4 s/km (blue)
+\end{center}
+\vspace{-0.5cm}
+$F[m]w_{\lambda}$ orthogonal to $F[m_*]w_{\lambda}$ if $|m-m_*| > 2\lambda/h$
+\hfill Q. E. D.
 }
-\only<4>{
+\only<3>{
 $J_{\rm FWI}$: $d=F[m_*]w_{\lambda}$, $m_*=0.4$ s/km, $w_{\lambda}$ = Ricker ctr freq = $1/\lambda$ Hz, $h=1$ km
@@
 }
-\only<5>{
+\only<4>{
 Postmortem:
@@
 
-WRI - van Leeuwen \& Herrmann GJI 13, IP 16, many other papers - most studied extended inversion approach
-
-Idea: view wave equation as {\em weak constraint} $\sim$ {\em add source parameters}
-
 Rationale: more likely to be able to 
@@
 \end{itemize}
+
+WRI - van Leeuwen \& Herrmann GJI 13, IP 16, many other papers - most studied extended inversion approach
+
+Idea: view wave equation as {\em weak constraint} $\sim$ {\em add source parameters}
+
 }
@@
 \item Wave operator $L[m]$ (eg. constant density acoustic op $L[m]=m^2\partial_t^2 - \nabla^2$)
-\item Known (!) source field $q(\bx,t)$ (eg. point source $w(t)\delta(\bx-\bx_s)$
+\item Known (!) source field $q(\bx,t)$ (eg. point source $w(t)\delta(\bx-\bx_s)$)
 \item dynamic fields $u(\bx,t)$ solve $L[m]u=q$ (eg. $u=$ pressure)
@@
 
-given $d$, $q$, set $g=L[m]u-q$,
+given $d$, $q$, set $g=L[m]u-q$ (``residual source''),
 
@@
 \]
-minimizer = $g[m;q,d]$, then
-\[
-=\|F[m]g[m;q,d]+e[m]\|^2 + \alpha^2 \|g[m;q,d]\|^2
-\]
-$m$ minimizes $J_{\rm WRI}$ $\Leftrightarrow$ $(m,g[m;q,d])$ solves WRI problem 
+Since $\min_{m,g}(...) = \min_m(\min_g (...))$, WRI problem $\equiv$ minimize $J_{\rm WRI}$
 }
@@
 \]
-Substitute solution into def of $J_{\rm WRI}$, do a page of algebra (IP, v. 36 no. 10 (2020)), then...
+Substitute solution into def of $J_{\rm WRI}$, do a page of algebra (S., IP 20), then...
 \[
@@
 \item ``hugging your data'' also not enough: WRI cycle-skips for small $\alpha$
-\item conclusion not limited to single trace transmission: similar for other transmission IPs, eg. diving wave inversion
+\item conclusion not limited to single trace transmission: similar for other transmission IPs, eg. diving wave inversion (Fang \& Demanet, SEG 20)
 \end{itemize}
@@
 }
+\only<7>{
+Verdict: MSI
+\begin{itemize}
+\item appears to avoid cycle-skipping
+\item ``hugs data'', expands search space - but effectively
+\item Lots of multi-D relatives show good cycle-skip busting examples:  matched source extension (Huang \& S.
+16), volume and space-time source extensions (Huang et al GEO 18), surface source extension (Huang et al.
+SEG 19), receiver shift extension (M\'{e}tivier SEG 20),  AWI (Warner \& Guatsch GEO 16 - field scale 3D, commercial)
+\end{itemize}
+}
+\end{frame}
+
+\begin{frame}\frametitle{So why?}
+\only<1>{
+Common penalty form for extended source inversion methods:
+\[
+\min_{m,w} \|F[m]w-d\|^2 + \alpha^2 \|Aw\|^2
+\]
+\begin{itemize}
+\item $m = $ vector of parameter {\em fields} incl. velocities
+\item $F[m] = $ modeling operator
+\item $d=$ multichannel data 
+\item $w$ =  extended source, $A = $ annihilator: physical source satisfy $Aw \approx 0$
+\end{itemize}
+Example: (AWI/MSI) point radiator  $w(t)\delta(\bx-\bx_s)$: physical source = impulsive, focused at $t=0$; extended source: spread in $t$; $Aw(t)=tw(t)$.
+}
+\only<2>{
+Necessary condition: reduced objective (VPM) must be regular (differentiable) in $m, d$ {\em jointly}
+
+For penalty functions as above: $A$ {\em must} be (pseudo)differential (Stolk \& S 03)
+
+Re-write of FWI, WRI reduced objectives: analogous op {\em not} pseudodifferential
+
+\begin{itemize}
+\item FWI, WRI {\em fail}, MSI {\em passes} (S. IP 20)
+\end{itemize}
+
+}
+\only<3>{
+Sufficient condition: gradient related to traveltime tomography gradient
+\begin{itemize}
+\item Structure of wave-based modeling operators:
+\[
+D_m (F[m]w)\delta m = {\color{blue}Q[m,\delta m]}F[m]w
+\]
+$Q$ = order 1 approx. skew-symm. (pseudo)differential op
+\end{itemize}
+Example: (AWI/MSI) $F[m]w(\bx_r,t) \approx a(\bx_s,\bx_r)w(t-\tau[m](\bx_s,\bx_r))$, so
+\[
+D_m (F[m]w)\delta m \approx {\color{blue}-D\tau(\bx_s,\bx_r)\delta m\frac{\partial}{\partial t}}F[m]w
+\] 
+(chain rule!)
+}
+\only<4>{
+(...bunch of algebra...) $\Rightarrow$ 
+\[
+\nabla J_{\rm MSI} \propto \nabla_m \|\tau[m]-\tau_{\rm data}\|^2
+\]
+so MSI $\approx$ travel time inversion
+
+And that's why.
+
+(Chen et al. SEG 20, Huang \& S. SEG 16) 
+}
+\end{frame}
+
+\begin{frame}\frametitle{Thanks to...}
+\begin{itemize}
+\item Prof. Susan. E. Minkoff, GRA Huiyi Chen, UT Dallas
+\item Guanghui Huang (PGS), Rami Nammour, Mohamed Dolliazal, Paul Williamson (Total)
+\item many other students, postdocs, and colleagues
+\item former sponsors of The Rice Inversion Project (1992-2019)
+\item you!
+\end{itemize}
+\end{frame}
+
+\begin{frame}\frametitle{Key References}
+G. Huang et al., ``Waveform Inversion by Source Extension'': SEG 2019
+
+H. Chen et al., ``Full waveform inversion by source extension: why it works", SEG 2020
+
+W. Symes, ``Wavefield reconstruction inversion: an example'', {\em Inverse Problems} v. 36(10), 2020
 \end{frame}

### 2020-12-15 repro/papers/discrep/discrep/paper.tex
tar dir prep for SEM to use overleaf
overleaf 20.12.14

--- 2020-12-07 repro/papers/discrep/paper.tex
@@
 \title{Solution of an Acoustic Transmission Inverse Problem by Extended Inversion}
-\author{William W. Symes, Orcas Island, Huiyi Chen* and Susan E. Minkoff, The University of Texas at Dallas}
+\author{William W. Symes\footnotemark[1], Huiyi Chen\footnotemark[2],
+  and Susan E. Minkoff\footnotemark[2]}
+\address{\footnotemark[1]PO Box 43, Orcas, WA 98280 USA\\
+  \footnotemark[2]Department of Mathematical Sciences, FO 35, 
+  University of Texas at Dallas, Richardson, TX 75080 USA}
 
@@
 widely used and successful technique for data analysis in science and
-engineering \cite[]{Bertero:98,Vogel:02}, particularly in the
-geosciences \cite[]{Parker:94,Tarantola:05}. In its application to
+engineering \cite{Bertero:98,Vogel:02}, particularly in the
+geosciences \cite{Parker:94,Tarantola:05}. In its application to
 seismology, least-squares data fitting has come to be known as Full
@@
   \label{eqn:anninorm}
-  p[m,\aw[m,d];d] = \frac{(4\pi r)^2}{2\|w_*\|^2}\int \,dt\,  a(t-(m-m_*)r)^2 \left(1+ (4\pi r)^2\alpha^2
-    a(t-(m-m_*)r)^2\right)^{-2}w_*(t)^2.
+  p[m,\aw[m,d];d] = \frac{(4\pi r)^2}{2\|w_*\|^2}\int \,dt\,  
+  \frac{a(t-(m-m_*)r)^2}{(1+ (4\pi r)^2\alpha^2
+    a(t-(m-m_*)r)^2)^{2}}w_*(t)^2.
 \end{equation}
@@
   \frac{d}{dm}\tJa[m;d] = -\frac{r (4\pi r \alpha)^2}{\|w_*\|^2} \int \,dt \, 
-  \left(a\frac{da}{dt}\right)(t-(m-m_*)r)(1+(4\pi r)^2 \alpha^2 
-  a(t-(m-m_*)r)^2)^{-2}w_*(t)^2. 
+  \frac{\left(a\frac{da}{dt}\right)(t-(m-m_*)r)}{(1+(4\pi r)^2 \alpha^2 
+  a(t-(m-m_*)r)^2)^{2}}w_*(t)^2. 
 \end{equation}
@@
   e[m_{\infty},\aw[m_{\infty},d];d] 
-= 8 \pi^2 r^2 \alpha^4\int^{\mu}_{-\mu}\,dt\,|t-(m_{\infty}-m_*)r|^4(1+(4\pi r)^2 \alpha^2 
-|t-(m_{\infty}-m^*)r|^2)^{-2}w_*(t)^2.
+= 8 \pi^2 r^2 \alpha^4\int^{\mu}_{-\mu}\,dt\,\frac{|t-(m_{\infty}-m_*)r|^4}{(1+(4\pi r)^2 \alpha^2 
+|t-(m_{\infty}-m^*)r|^2)^{2}}w_*(t)^2.
 \]
@@
 appealing, and is supported by numerical evidence, as shown below.
-
-%%%%%%%%%%%
-\begin{proposition}
-  \label{thm:wnoiseres}
-  Asumme that
-  \begin{itemize}
-  \item[1. ] $\alpha, \mu> 0$,
-  \item[2. ] $m_* \in M, w_* \in W_{\mu}$,
-  \item[3. ] $d_* = F[m_*]w_*$,
-  \item[4. ] $n \in D$ and $d = d_* + n$.
-  \end{itemize}
-  Set $\eta = \|n\|/\|d_*\|$, and assume further that
-  \begin{itemize}
-  \item[5. ]
-    \begin{equation}
-      \label{eqn:mnoisecond}
-      \eta < \frac{\sqrt{5}-1}{2}, and
-    \end{equation}
-  \item[6. ] $m$ is a stationary point of $\tJa[\cdot;d]$.
-  \end{itemize}
-  Then
-  \begin{equation}
-    \label{eqn:mnoisesuff}
-    |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}.
-  \end{equation}
-\end{proposition}
-
-\begin{proof}
-  From equation \ref{eqn:normsol},
-  \[
-    \aw[m;d](t) = 
-    \left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2)\right)^{-1}\frac{1}{4 \pi r}d(t+mr)
-  \]
-  \[
-    =4\pi r(1 + (4\pi r \alpha t)^2)^{-1} (\frac{1}{4\pi r}
-    w_*(t+(m-m_*)r) + n(t+mr) 
-  \]
-  so
-  \begin{equation}
-    \label{eqn:wres}
-    \aw[m;d](t) - w_*(t) = (4\pi r \alpha t)^2 (1+ (4\pi r \alpha t)^2)^{-1}w_*(t+(m-m_*)r)
-    +w_*(t+(m-m_*)r)-w_*(t) + 4 \pi r (1 + (4\pi r \alpha t)^2)^{-1}n(t+mr)
-  \end{equation}
-\end{proof}
-%  \begin{itemize}
-%    \item[1. ] the bound \ref{eqn:mnoisebnd} holds only if 
-%      noise level $\eta=\|n\|/\|d_*\|$ satisfies
-%      \begin{equation}
-%        \label{eqn:mnoisecond}
-%        \eta < \frac{1+\sqrt{5}}{2} < 1;
-%      \end{equation}
-%    \item[2. ] if condition \ref{eqn:mnoisecond} holds, then
-%      $|m-m_*|<\lambda/r$ provided that
-%      \begin{equation}
-%        \label{eqn:mnoisesuff}
-%        \frac{\lambda}{\mu} \ge 1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}. 
-%      \end{equation}
-%      \begin{equation}
-%        \label{eqn:mnoisesuff}
-%        |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}. 
-%      \end{equation}
-%    \end{itemize}
-%\end{corollary}
 
@@
   \end{itemize}
-  Set $\eta = \|n\|/\|d_*\|$, and assume further that
-  \begin{itemize}
-  \item[5. ]
-    \begin{equation}
+  Set $\eta = \|n\|/\|d_*\|$. Assume that $\eta$ satisfies inequality \begin{equation}
   \label{eqn:mnoisecond}    
-      \eta < \frac{\sqrt{5}-1}{2}, and
-    \end{equation}
-  \item[6. ] $m$ is a stationary point of $\tJa[\cdot;d]$.
-  \end{itemize}
+  \eta < \frac{\sqrt{5}-1}{2},
+  \end{equation}
+  and that $m$ is a stationary point of $\tJa[\cdot;d]$.
   Then
@@
 \bibliographystyle{seg}
-\bibliography{../../bib/masterref}
-
-\append{Numerics and penalty selection via the Discrepancy Principle}
+\bibliography{masterref}
+
+\append{Penalty weight selection via the Discrepancy Principle}
 This appendix describes a method for controlling the penalty
-weight $\alpha$ based on the Discrepancy Principle
+weight $\alpha$ based on a version the Discrepancy Principle
 \cite[]{EnglHankeNeubauer,FuSymes2017discrepancy}.
@@
 of $J_{\alpha_{\infty}}[\cdot,\cdot;d]$ (Theorem \ref{thm:norminv}).
-%.
-%so is also a minimizer of $g[w]$ over the set
-%\[
-%  \{(m,w): e[m,w;d] \ge
-%  e[m_{\infty},w_{\alpha_{\infty}}[m_{\infty};d]]\}
-%\]
-%or equivalently a maximizer of $e[m,w,d]$ over the set
-%\[
-%  \{(m,w): m \in M, g[w] \le g[w_{\alpha_{\infty}}[m_{\infty};d]]\}
-%\]
-%according to standard Lagrange multiplier theory. That is,
 
@@
 Given an
-objective function $\Phi: M \rightarrow \bR$, a local optimization algorithm generates 
+objective function $\Phi: M \rightarrow \bR$, a local optimization algorithm 
 generates a map $G[\Phi,...]: M \rightarrow M$, mapping a current estimate
@@
 $e_-$, or if an approximate local minimizer is detected, then the condition in step 6 is satisfied, so control returns
-to step 7. If it is satisfied, the algorithm terminates, elsewhich
+to step 7. If it is satisfied, the algorithm terminates, else
 control loops back to step 2, $\alpha$ is updated, and the inner $m$
@@
 
-Accordingly, regard $m$ as fixed and suppress it from the notation for the remainder of ;dthis section, and introduce the abbreviations
+Accordingly, regard $m$ as fixed and suppress it from the notation for the remainder of this section, and introduce the abbreviations
 \begin{eqnarray}
@@
 \end{eqnarray}
-Since the normal matrix $F^T F + \alpha^2 A^T A$ has been assumed
-invertible (condition \ref{eqn:nopco}), the solution of $\aw$ of the
-normal equation \ref{eqn:condition} is smooth in $\alpha^2$.
-Differentiating equation \ref{eqn:condition} with respect to  $\alpha^2$ leads to the relation
+Assume that the normal operator $F^T F + \alpha^2 A^T A$ is boundedly invertible, as for example in Theorem \ref{thm:norminv}. Then the solution of $\aw$ of the
+normal equation \ref{eqn:norm}
+is smooth in $\alpha^2$.
+Differentiating equation the normal equation  
+with respect to  $\alpha^2$ leads to the relation
 \begin{equation}

### 2021-01-03 repro/papers/discrep/discrep/paper.tex
app a finished

@@
 
-\begin{proof} Follows directly from Theorem \ref{thm:rampgood} by
+\begin{proof} Follows directly from Proposition \ref{thm:rampgood} by
   taking $\mu=\lambda$.
@@
 The existence of this limiting penalty weight has been inferred
-indirectly; the next section is devoted to a constructive algorithm
+indirectly; Appendix A describes a constructive algorithm
 for its approximation.
@@
 \bibliographystyle{seg}
-\bibliography{masterref}
+\bibliography{masterref,local}
 
 \append{Penalty weight selection via the Discrepancy Principle}
-This appendix describes a method for controlling the penalty
+This appendix describes an algorithm for controlling the penalty
 weight $\alpha$ based on a version the Discrepancy Principle
-\cite[]{EnglHankeNeubauer,FuSymes2017discrepancy}.
-
-Modified Problem Statement:
+\cite[]{EnglHankeNeubauer,Hanke:17,FuSymes2017discrepancy}. 
+The algorithm solves the problem
 \begin{quote}
@@
   \item[(i) ]$m$ is a stationary point of $\tJa[\cdot;d]$, and
-  \item[(ii) ]$e[m,\aw[m;d];d] \in [e_-,e_+]$.
+  \item[(ii) ]$e[m,\aw[m;d];d] \in (e_-,e_+)$.
   \end{itemize}
 \end{quote} 
-
-\noindent {\bf Remark:} Given a solution $m_{\infty}, \alpha_{\infty}$, the pair
-$(m_{\infty},w_{\alpha_{\infty}}[m_{\infty};d])$ is a local minimizer
-of $J_{\alpha_{\infty}}[\cdot,\cdot;d]$ (Theorem \ref{thm:norminv}).
-
 We describe an alternating, or coordinate search, algorithm for
-solution of problem \ref{eqn:probmod}, combining a local optimization
+solution of this problem, combining a local optimization
 algorithm for updating $m$, and a second algorithm for updating
-$\alpha$. A first version of this algorithm appeared in \cite{FuSymes2017discrepancy}.
+$\alpha$.
+A first version of this algorithm appeared in \cite{FuSymes2017discrepancy}.Note that the mean square error $e$ lies in an open interval $(e_-,e_+)$ at a solution of this problem.
+Use of an interval, rather than a single target error level, accomplishes two objectives:
+\begin{itemize}
+    \item it is consistent with the general lack of precise knowledge of data error in most applications;
+    \item it permits a local optimization algorithm to make several small updates of $m$ before an update of $\alpha$ is required, as is required for good performance of algorithms such as BFGS.
+\end{itemize}
 
@@
 satisfaction of sufficient decrease conditions
-(see \cite{NocedalWright}), the iteration produces a sequence in $M$
-converging to a local minimizer of $\Phi$. An example of such a local
-optimization algorithm is steepest descent with backtracking line
-search.
+(see \cite{NocedalWright}), iteration of $G$ produces a sequence in $M$
+converging to a local minimizer of $\Phi$.
 
@@
 We will also require a rule for updating $\alpha$, defining a map $H: M \times E
-\rightarrow \bR^+$, in which $E=\{(e_-,e_+) \in \bR^2:
+\rightarrow \bR^+$, in which $E=\{(e_-,e_+) \in \bR^2: 0 < 
 e_-<e_+\}$. For any $m \in M$ the rule is required to produce an $\alpha$ for which
@@
 
-Accordingly, regard $m$ as fixed and suppress it from the notation for the remainder of this section, and introduce the abbreviations
+The penalty parameter update strategies are based on the following fact about linear combinations of quadratic forms, similar to well-known results from the theory of Tihonov regularization \cite[]{Hanke:17}:
+
+\begin{theorem}
+  \label{thm:tich}
+  Suppose that $W, D$ are Hilbert spaces, $F \in \cal{B}(W,D)$, $A \in \cal{B}(W,W)$, $F^TF + \alpha^2 A^TA > 0$ for any $\alpha \ge 0$ (in particular, $F^TF>0$), and $A$ is injective.
+For $d \in D$, define
+  $w: D \times \bR^+ \rightarrow W$,
+  $e: D \times \bR^+ \rightarrow \bR^+$, and $g: D \times \bR^+ \rightarrow \bR^+$ by
   \begin{eqnarray}
-\label{eqn:erralph}a
-e(\alpha) &=& e[m,\aw[m;d],d]\\
-\label{eqn:penalph}
-p(\alpha) &=& p[m,\aw[m;d],d]
+    w[d,\alpha] &=& (F^TF + \alpha^2 A^TA)^{-1}F^Td \nonumber \\
+    e[d,\alpha] &=& \frac{1}{2\|d\|^2}\|Fw[d,\alpha]-d\|^2 \nonumber \\
+    g[d,\alpha] &=& \frac{1}{2\|d\|^2}\|Aw[d,\alpha]\|^2
+    \label{eqn:tichdefs}
   \end{eqnarray}
-Assume that the normal operator $F^T F + \alpha^2 A^T A$ is boundedly invertible, as for example in Theorem \ref{thm:norminv}. Then the solution of $\aw$ of the
-normal equation \ref{eqn:norm}
-is smooth in $\alpha^2$.
-Differentiating equation the normal equation  
-with respect to  $\alpha^2$ leads to the relation
+  Then for any $d\in D$, $w[d,\cdot]$, $e[d,\cdot]$, and $g[d,\cdot]$ are of class $ C^{\infty}(\bR^+)$, and 
+  \begin{itemize}
+      \item[0. ] EITHER $d$ is perpindicular to the range of $F$, $e \equiv \frac{1}{2}, g \equiv 0$ for all $\alpha \ge 0$, 
+      \item[1. ] OR $e$ is positive and strictly increasing, $g$ is positive and strictly decreasing for  $\alpha\ge 0$, and 
+      \begin{equation}
+        \label{eqn:lep}
+        \frac{de}{d\alpha^2}  \le 2g.
+      \end{equation}
+  \end{itemize}
+\end{theorem}
+
+\noindent {\bf Remark:} The conditions on $F$ and $A$ in the statement of this theorem might seem unduly restrictive.
+Both hold for the cartoon problem studied in the body of this paper.
+In fact, in most cases studied in the literature, the extended modeling operator represented here by $F$ is not coercive, and must be modified by regularization to meet the conditions of the theorem.
+Instances of the penalty operator $A$ studied in the literature are in fact usually injective, though this fact is commonly overlooked.
+See \cite{Symes:09} for some discussion.
+
+\begin{proof}
+Since the normal operator $F^T F + \alpha^2 A^T A$ is boundedly invertible and smooth in $\alpha \ge 0$, $w$ is smooth in $\alpha$, as are $e$ and $g$. 
+Differentiate the definition of $w$ 
+with respect to  $\alpha^2$ to obtain
 \begin{equation}
@@
 \end{align}
-Note that the inequality in equation \ref{eqn:de} is {\em strict}  if $p > 0$ hence $A^TAw \ne 0$, since the normal operator is assumed to be positive definite.
-The derivative of $p(\alpha^2)$ with respect to $\alpha^2$ is
+Note that the inequality in equation \ref{eqn:de} is {\em strict}  if $g[d,\alpha] > 0$ hence $A^TAw[d,\alpha] \ne 0$, since the normal operator is assumed to be positive definite.
+Also,
 \begin{align}
-\frac{dp}{d\alpha^2} &=  -\langle A^T Aw,(F^TF + \alpha^2 A^TA)^{-1}A^TAw \rangle \nonumber \\
+\frac{dg}{d\alpha^2} &=  -\langle A^T Aw,(F^TF + \alpha^2 A^TA)^{-1}A^TAw \rangle \nonumber \\
 &\leq 0
@@
 \end{align}
-similarly a strict inequality if $p > 0$.
-
-Equations \ref{eqn:de} and \ref{eqn:dp} show that increasing $\alpha^2$ implies increasing $e$ while decreasing $p$, and
+similarly a strict inequality if $g \ne 0$.
+
+It follows that either $g[d,\alpha]>0$ for all $\alpha\ge 0$, or there exists $\alpha_0\ge 0$ for which $g[d,\alpha]>0$ for $0
+\le \alpha<\alpha_0$ and $g[d,\alpha]=0$ for $\alpha \ge \alpha_0$, hence $Aw[d,\alpha]=0$ for $\alpha \ge \alpha_0$.
+Since $A$ is assumed injective, $w[d,\alpha]=0$ for $\alpha > \alpha_0$.
+From the definition of $w$, it follows that $F^Td=0$, that is, $d$ is orthogonal to the range of $F$, and in fact $\alpha_0=0$, $g \equiv 0$ and $e \equiv \frac{1}{2}$ for all $\alpha\ge 0$.
+
+In the first case, that is, $F^Td \ne 0$, equations \ref{eqn:de} and \ref{eqn:dp} show that increasing $\alpha^2$ implies increasing $e$ while decreasing $g$, and
 \begin{align}
 &\langle A^TA w,(F^TF + \alpha^2 A^TA)^{-1} A^TA w \rangle \nonumber \\ 
-=& \langle (A^TA)^{1/2}w,[(A^TA)^{-1/2}F^TF(A^TA)^{-1/2} + \alpha^2 I]^{-1}(A^TA)^{1/2}w \rangle \nonumber \\ 
-\le& \frac{1}{\alpha^2} \langle A^TA w, w\rangle = \frac{2}{\alpha^2}p
+=& \lim_{\epsilon \rightarrow 0}\langle (A^TA+\epsilon^2I)^{1/2}w,[(A^TA+\epsilon^2I)^{-1/2}F^TF(A^TA+\epsilon^2I)^{-1/2} + \alpha^2 I]^{-1}(A^TA+\epsilon^2I)^{1/2}w \rangle \nonumber \\ 
+\le& \lim_{\epsilon \rightarrow 0}\frac{1}{\alpha^2} \langle (A^TA+\epsilon^2I) w, w\rangle = \frac{2}{\alpha^2}g.
 \end{align}
-In view of equation \ref{eqn:de},
-\begin{equation}
-\label{eqn:lep}
-\frac{de}{d\alpha^2}  \le 2p.
-\end{equation}
-with this inequality also being strict if $p > 0$.
-
-Suppose the current weight is $\alpha^2_c$ and denote a candidate for an updated weight by $\alpha^2_+$. Then from inequality \ref{eqn:lep},
-\begin{equation}
-e(\alpha^2_+)-e(\alpha^2_c) \le \int_{\alpha^2_c}^{\alpha^2_+} 2p d\alpha^2
-\end{equation}
-If $\alpha^2_+ \ge \alpha^2_c$, then in view of inequality \ref{eqn:dp}, the above is
+which establishes inequality \ref{eqn:lep}.
+\end{proof}
+
+\begin{proposition}
+  \label{thm:alphainf}
+  Under the hypotheses of Theorem \ref{thm:tich}, 
+  \begin{eqnarray}
+  \label{eqn:walphainf}
+  \lim_{\alpha \rightarrow \infty} w(d,\alpha) & = & 0,\\
+  \label{eqn:ealphainf} 
+  \lim_{\alpha \rightarrow \infty} e(d,\alpha) & = & 1/2.
+  \end{eqnarray}
+\end{proposition}
+\begin{proof}
+  Since $F^TF$ is positive-definite, it has a positive-definite square root, and the definition \ref{eqn:tichdefs} of $w(d,\alpha)$ is equivalent to
+  \begin{equation}\label{eqn:ftfw}
+  (F^TF)^{1/2}w(d,\alpha) = (I+\alpha^2 (F^TF)^{-1/2}A^TA(F^TF)^{-1/2})^{-1}(F^TF)^{-1/2}F^Td.
+  \end{equation}
+  Denote by $E:\bR \rightarrow {\cal B}(W,W)$ the resolution of the identity for the self-adjoint bounded operator $(F^TF)^{-1/2}A^TA(F^TF)^{-1/2}$.
+According to the theorem on spectral representation of functions of self-adjoint operators (\cite{Yosida}, section XI.5, Theorem 1),
+  \begin{equation}\label{eqn:specw}
+  \|(F^TF)^{1/2}w(d,\alpha)\|^2 = \int_0^{\|F^TF\|}\frac{1}{1+\alpha^2 \lambda}d \|E(\lambda)(F^TF)^{-1/2}F^Td\|^2.
+  \end{equation}
+  Since $A^TA$ is assumed injective, so is $F^TF)^{-1/2}A^TA(F^TF)^{-1/2}$, whence $0$ is either a member of the resolvent set of the latter operator, or of its continuous spectrum.
+In either case, $\{0\}$ is a set of measure zero with respect to $\|E(\lambda)(F^TF)^{-1/2}F^Td\|^2$, so the integrand $(1+\alpha^2 \lambda)$ converges to zero $\|E(\lambda)(F^TF)^{-1/2}F^Td\|^2$-almost everywhere as $\alpha \rightarrow \infty$, and is bounded above by $1$ on the spectrum.
+The Dominated Convergence Theorem implies that $\|(F^TF)^{1/2}w(d,\alpha)\|^2 \rightarrow 0$ as $\alpha \rightarrow 0$, which establishes the first conclusion.
+The second follows from the definition of $e$ and the continuity of the operators involved in it.
+\end{proof}
+
+\begin{proposition}
+  \label{thm:psidef}
+  In addition to the hypotheses of Theorem \ref{thm:tich}, assume that $F^Td \ne 0$, and that $e[d,0]<e_+ \le 1/2$.
+Then there exists a unique $\alpha_+ > 0$ satisfying $e_+ = e[d,\alpha_+]$, and $e[d,\alpha]<e_+$ for $\alpha \in [0,\alpha_+)$.
+Define $\Psi: [0,\alpha_+] \rightarrow \bR$ by
+ \begin{equation}
+\label{eqn:alphasecant}
+\Psi(\alpha)= \left(\alpha^2 + \frac{e_{+}-e[d,\alpha]}{2g[d,\alpha]} \right)^{1/2}
+\end{equation} 
+Then $\Psi([0,\alpha_+]) \subset (0,\alpha_+]$ and $\alpha<\Psi(\alpha)<\alpha_+$ for $\alpha \in [0,\alpha_+)$. 
+\end{proposition}
+
+\noindent {\bf Remark:} Note that $\Psi(0)$ is well-defined and $>0$.
+
+\begin{proof} Existence and uniqueness of $\alpha_+$ follows from Theorem \ref{thm:tich}, Proposition \ref{thm:alphainf}, and the assumption that $e[d,0]<e_+\le 1/2$.
+Since $e$ is strictly increasing under the assumption that $d$ is not perpindicular to the range of $F$, $e[d,\alpha]<e_+$ for $\alpha \in [0,\alpha_+)$.
+
+Suppose that $0 \le \alpha \le \psi \le \alpha_+$. From inequality \ref{eqn:lep}, 
+\[
+e[d,\psi]-e[d,\alpha] \le \int^{\psi^2}_{\alpha^2} d\tau^2 g[d,\tau] 
+\]
 \begin{equation}
 \label{eqn:basic}
-\le 2 p(\alpha^2_c) (\alpha^2_+-\alpha^2_c)  
-\end{equation}
-
-Let us suppose that $e(\alpha^2_c) < e_{+}$. Then setting 
-\begin{equation}
-\label{eqn:alphasecant}
-\alpha^2_+ = \alpha^2_c + \frac{e_{+}-e(\alpha^2_c)}{2p(\alpha^2_c)} 
-\end{equation}
-implies via inequality \ref{eqn:basic} that 
-\begin{equation}
-e(\alpha^2_+)-e(\alpha^2_c) \le e_{+}-e(\alpha^2_c)
-\end{equation}
-
-Assuming that $p(\alpha^2_+) > 0$, hence $p(\alpha^2)>0$ for $\alpha^2_c \le \alpha^2 \le \alpha^2_+$,  we conclude that if $\alpha^2_+$ is given by the rule \ref{eqn:alphasecant},
-then
+< 2 g[d,\alpha] (\psi^2-\alpha^2)  
+\end{equation}
+since $g$ is positive and strictly decreasing.
+From the definition \ref{eqn:alphasecant}, $\Psi(\alpha) \ge \alpha$ for $\alpha \in [0,\alpha_+]$, and $\Psi(\alpha)>\alpha$ if $\alpha \in [0,\alpha_+)$.
+Setting $\psi=\Psi(\alpha)$,
+inequality \ref{eqn:basic} and definition \ref{eqn:alphasecant} imply
+\[
+e[d,\Psi(\alpha)]-e[d,\alpha] < e_{+}-e[d,\alpha]
+\]
+so
 \begin{equation}
 \label{eqn:assert}
-e(\alpha^2_c) < e(\alpha^2_+) \le e_{+}
-\end{equation}
-That is, unless $p(\alpha^2_+) = 0$ (in which case a physical solution of the inverse problem has been reached),  $e(\alpha^2_+)$ is larger than $e(\alpha^2_c)$ but in any case does not exceed $e_+$.
-The rule \ref{eqn:alphasecant} therefore provides a feasible updated $\alpha^2$ consistent with the upper bound in the discrepancy principle.
-
+\alpha \in [0,\alpha_+) \Rightarrow e[d,\alpha] < e[d,\Psi(\alpha)] < e_{+}.
+\end{equation}
+The last inequality ans the strict increase of $e$ imply that $\Psi(\alpha)<e_+$ if $\alpha \in [0,\alpha_+)$.
+\end{proof}
+
+These results suggest an algorithm to determine $\alpha$: select an initial $\alpha_0 \ge 0$ for which $e[d,\alpha_0]<e_-$, then set $\alpha_n=\Psi(\alpha_{n-1}), n=1,2,...$, until $e[d,\alpha_n] > e_-$.
+Since $e_-<e_+$, the preceding result implies that $e[d,\alpha_n]<e_+$ also.
+
+That is, iteration of $\Psi$ produces an increasing sequence in $[0,\alpha_+]$.
+Two further ingredients are required to view it as an algorithm for production of an $\alpha$ satisfying the discrepancy principle:  a method for selection of an initial $\alpha_0$, and assurance that the error $e$ will eventually exceed the lower bound $e_-$.
+
+\begin{theorem}
+  \label{thm:discrepalg} In addition to the hypotheses of Proposition\ref{thm:psidef}, assume that $e[d,0] <e_-<e_+\le 1/2$. Define the sequence $\{\alpha_n: n \in {\bf Z}_+\}$ by
+  \begin{itemize}
+      \item $\alpha_0=0$, and
+      \item $\alpha_{n+1}=\Psi(\alpha_n), n \in {\bf Z}_+$.
+  \end{itemize}
+  Then there exists $N=N(d,e_-,e_+) \in {\bf Z}_+$ for which 
+  \begin{itemize}
+      \item $e(d,\alpha_n) \le e_-$ for $0 \le n <N$, and
+      \item $e(d,\alpha_N) \in (e_-,e_+)$
+  \end{itemize}
+\end{theorem}
+
+\begin{proof}
+  By hypothesis, the first conclusion is satisfied for $N=0$.
+From the definition \ref{eqn:alphasecant} of $\Psi$ and the conclusions of Proposition \ref{thm:psidef}, $\{\alpha_n\}$ is strictly increasing and $\alpha_n \le \alpha_+$ for all $n \ge 0$, so has a limit point $\alpha_{\infty} \in (0,\alpha_+]$.
+If $\alpha_{\infty}<\alpha_+$, $\epsilon = \Psi(\alpha_{\infty})-\alpha_{\infty} >0$.
+Since $\alpha \mapsto \Psi(\alpha)-\alpha$ is continuous, there is $\delta>0$ so that if $|\alpha-\alpha_{\infty}|<\delta$, then $\Psi(\alpha)-\alpha >\epsilon/2$.
+Since $\alpha_{\infty}=\lim_{n\rightarrow \infty}\alpha_n$, there is $m \in {\bf Z}_+$ for which $\alpha_m \in (\alpha_{\infty}-\min(\epsilon/2,\delta),\alpha_{\infty})$, so that $\alpha_{m+1} = \Psi(\alpha_m) >\alpha_m + \epsilon/2>\alpha_{\infty}$, contradicting the definition of $\alpha_{\infty}$ as the limit of the increasing sequence $\{\alpha_n\}$.
+Therefore, we conclude that $\alpha_{\infty}=\alpha_+$.
+
+  Since $\alpha_n <\alpha_+$ for all $n$, according to Proposition \ref{thm:psidef}, there must exist a least $n = N(d,e_-,e_+)$ for which $\alpha_n \in (e_-,e_+)$.
+\end{proof}
+
+\noindent {\bf Remark:} The condition $e(d,0)\le e_-$ means that $\sqrt{2e_-}$ is an upper bound for the distance from $d$ to the range of $F$.
+Recall that $e_-$ represents an underestimate for data noise level.
+That is, it is assumed that the data is within data noise level of the range of $F$.
+Such an assumption is unlikely to be tenable for the physical (non-extended) modeling operator, unless the nonlinear model parameters ($m$ in this discussion) are chosen near-optimally.
+The ability to fit data, even when the parameters represented by $m$ are greatly in error, is an essential characteristic of successful extended inversion strategies.
+
+\noindent {\bf Remark:} The update rule \ref{eqn:alphasecant} is convergent to a satisfactory value, as just shown, but in practice is rather slow.
+Accelerated updates using variants of the secant rule are possible, but an analysis of those possibilities is beyond the scope of this paper.
+The numerical examples shown above use the rule \ref{eqn:alphasecant}.
+
+A satisfactory penalty parameter selection rule can now be defined: in the notation introduced at the beginning of this appendix, define 
+\begin{equation}
+    H[m,e_-,e_+] = \alpha_{N(d,e_-,e_+)}.
+\end{equation}
+Choosing $F=F[m]$ in Theorem \ref{thm:discrepalg}, conclude that $H$, so defined, has the required properties.
+
+This is as far as we can go without more concrete assumptions on $F$ and $A$.
+For a few problems, analysis of the gradient of the reduced objective $\tJa$ shows that its  domain of convexity is large in some sense for a range of $\alpha$, then the discrepancy algorithm sketched here can efficiently drive $\alpha$ to near its largest feasible value, thus giving the best resolution near the global minimizer.
+That program has been carried out for the simple cartoon problem of this paper, and partly for some other more protypical inverse wave problem, as mentioned in the introduction.
 \append{Heisenberg, support, and wavelength}

### 2021-03-28 repro/papers/sseprecond/paper.tex
snapshot

@@
 \begin{enumerate}
-\item a linear wave operator $L[c]$, depending on an array of space-dependent coefficients $c$ and acting on wavefields $u$ vanishing in negative time:
+\item a linear wave operator $L[\bf{c}]$, depending on an array of space-dependent coefficients $\bf{c}$ and acting on wavefields $\bu$ vanishing in negative time:
 \begin{equation}
 \label{eqn:init}
-u \equiv 0, t \ll 0; 
+\bu \equiv 0, t \ll 0; 
 \end{equation}
 \item a trace sampling operator $P$ acting on wavefields and producing data traces;
-\item and a source function (of space and time) $f$ representing energy input to the system. 
+\item and a source function (of space and time) $\bff$ representing energy input to the system. 
 \end{enumerate}
-The basic FWI problem is: given data $d$, find $c$ so that 
+The basic FWI problem is: given data $d$, find $\bf{c}$ so that 
 \begin{equation}
 \label{eqn:fwi}
-Pu \approx d \mbox{ and } L[c]u = f.
+P\bu \approx d \mbox{ and } L[\bf{c}]\bu = \bff.
 \end{equation}
 
-The energy source $f$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time.
+The energy source $\bff$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time.
 In fact, additional source degrees of freedom, beyond those needed to describe physically realized sources, may be useful in rendering the FWI problem \ref{eqn:fwi} more amenable to numerical solution, via so-called extended modeling (see \cite{geoprosp:2008}, \cite{LeeuwenHerrmannWRI:13}, \cite{HuangNammourSymesDollizal:SEG19}, and many references cited there).
-Therefore it is natural to view $f$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
+Therefore it is natural to view $\bff$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
 %A typical nonlinear least squares formulation is:
@@
 \label{eqn:esi}
-\mbox{choose } c, f \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2 + \alpha^2 \|Af\|^2 
+\mbox{choose } \bf{c}, \bff \mbox{ to minimize } \|PL[\bf{c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
 \end{equation}
@@
 \label{eqn:esis}
-\mbox{given } c, \mbox{ choose } f \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2 + \alpha \|Af\|^2 
+\mbox{given } \bf{c}, \mbox{ choose } \bff \mbox{ to minimize } \|PL[\bf{c}]^{-1}\bff -d \|^2 + \alpha \|A\bff\|^2 
 \end{equation}
@@
 The source subproblem is a necessary step in solving the overall FWI problem \ref{eqn:fwi}.
-In particular, the nonlinear least squares realization \ref{eqn:esi} of source-extended FWI is naturally approached via the {\em variable projection method} \cite[]{GolubPereyra:73,GolubPereyra:03,vanLeeuwenMulder:09,Rickett:SEG12}, in which the minimum over $f$ in \ref{eqn:esis} is treated as a function of $c$, which is in turn minimized.
-Variable projection thus treats the source subproblem as an inner problem, which must be solved for an estimate of $f$ at every step of an iterative method for the outer problem of minimization over $c$.
+In particular, the nonlinear least squares realization \ref{eqn:esi} of source-extended FWI is naturally approached via the {\em variable projection method} \cite[]{GolubPereyra:73,GolubPereyra:03,vanLeeuwenMulder:09,Rickett:SEG12}, in which the minimum over $\bff$ in \ref{eqn:esis} is treated as a function of $\bf{c}$, which is in turn minimized.
+Variable projection thus treats the source subproblem as an inner problem, which must be solved for an estimate of $\bff$ at every step of an iterative method for the outer problem of minimization over $\bf{c}$.
 Efficiency in solving the source subproblem \ref{eqn:esis} is critical to efficient implementation of variable projection for solving the regularized FWI problem \ref{eqn:esi}.
 
-This paper restricts the choice of wave physics embodied in $L[c]$ to linear acoustics.
-The principal characteristics assumed of sources are localization and isotropy.
+This paper restricts the choice of wave physics embodied in
+$L[\bf{c}]$ to linear acoustics.  Thus $\bf{c}=(\kappa,\rho)$ is the
+pair (bulk modulus, density) of positive functions of spatial
+position.
+The principal characteristics assumed of sources $\bff=$ are localization and isotropy.
 That is, physical sources will be assumed to be isotropic point radiators, described by the functional form
@@
 
-The following section describes the construction of the approximate inverse and explains the conditions under which accuracy should be expected.
+The following section describes the construction of the approximate
+inverse and explains the conditions under which accuracy should be expected.
 I observe that the approximate inverse is approximately the adjoint of the modeling operator $PL[c]^{-1}$ with respect to weighted norms in its domain and range (see \cite{HouSymes:15} for a similar observation in a different context).
@@
 
-\section{Operators and Adjoints}
+\section{Surface Sources and Solutions}
 
-For acoustic wave physics, the coefficient vector $c$ is the pair $c=(\kappa,\beta)$ of bulk modulus $\kappa$ and density $\rho$, and the state vector $u=(p,\bv)^T$ consists of pressure $p$ (a scalar space-time field) and particle velocity $\bv$ (a vector space-time field).
-The wave operator $L[c]$ is given by the linear acoustics system:
+For acoustic wave physics, the coefficient vector $c$ is the pair $\bf{c}=(\kappa,\rho)$ of bulk modulus $\kappa$ and density $\rho$, and the state vector $\bu=(p,\bv)^T$ consists of pressure $p$ (a scalar space-time field) and particle velocity $\bv$ (a vector space-time field).
+The wave operator $L[\bf{c}]$ is given by the linear acoustics system:
 \begin{equation}
@@
 
-The space dimension is $d$. Most of what follows is valid for any $d >0$. Examples later in this paper will use $d=2$ for computationally convenience.
+Most of what follows is valid for any space dimension $n >0$. Examples
+later in this paper will use $n=2$ for computationally convenience. In
+the mathematical development that follows, I will set $n=3$.
 
-As mentioned earlier, in this paper the coefficients $c$ is regarded as defined throughout space $\bR^d$, the state vector $u$ throughout space-time $\bR^{d+1}$.
+As mentioned earlier, in this paper the coefficients $\bf{c}$ is regarded as defined throughout space $\bR^d$, the state vector $\bu$ throughout space-time $\bR^{d+1}$.
 
-The idealized ``crosswell'' configuration involves sources confined to $z=z_s$.
-I will mostly represent sources as constitutive law defects, that is, right-hand sides in the system $L[c]u=f$ of the form $f(\bx,t) = (h_s(x,y,t)\delta(z-z_s), {\bf 0})^T$.
-With the choice $L[c]$ given in \ref{eqn:aweop}, the causal wave system $L[c]u^+=f$ takes the form
+The idealized ``crosswell'' configuration involves sources confined to
+$z=z_s$. Such sources are combinations of constitutive law defects and
+loads normal to the surface, that is, right-hand sides in the system
+$L[c]\bf{u}=\bf{f}$ of the form $\bf{f}(\bx,t) =
+(h_s(x,y,t)\delta(z-z_s), f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar
+defect $h_s$ and normal force $f_s$ and $\bf{e}_z=(0,0,1)^T$. With the choice $L[\bf{c}]$ given in \ref{eqn:aweop}, the causal wave system $L[\bf{c}]\bu^+=\bff$ takes the form
 \begin{eqnarray}
@@
 h_s \delta(z-z_s), \nonumber \\
-\rho\frac{\partial \bv^{+}}{\partial t} & = & - \nabla p^{+},\nonumber \\
+\rho\frac{\partial \bv^{+}}{\partial t} & = & - \nabla p^{+} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
 p^{+} & =& 0 \mbox{ for } t \ll 0,\nonumber\\ 
@@
 h_s \delta(z-z_s), \nonumber \\
-\rho \frac{\partial \bv^{-}}{\partial t} & = & - \nabla p^{-},\nonumber \\
+\rho \frac{\partial \bv^{-}}{\partial t} & = & - \nabla p^{-} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
 p^{-} & =& 0 \mbox{ for } t \gg 0,\nonumber\\ 
@@
 \end{eqnarray}
-Sampling the causal pressure field $p^{+}$ on the surface $z=z_r$ defines the causal forward modeling operator $S^{+}_{z_s,z_r}[c]$:
-\begin{equation}
+
+\section{Operators and Adjoints}
+All of the operators discussed in this section depend on the
+coefficient array $\bf{c}$, which I will suppress from the notation.
+Denote by $P_s,P_r$ the trace (sampling) operators on $z=z_s$, $z=z_r$
+respectively. The causal and anticausal modeling operators
+${\cal S}^{+}_{z_s,z_r}$ and ${\cal S}^-_{z_s,z_r}$ are defined in
+terms of the solutions $(p^{\pm},\bv^{\pm})$ of the systems
+\ref{eqn:awep} and \ref{eqn:awem} by
+\begin{eqnarray}
+  {\cal S}^{+}_{z_s,z_r}(h_s,f_s) & = & (P_rp^+,P_r v_z^+)|_{z=z_r},
+                                        \nonumber\\
+  {\cal S}^{-}_{z_s,z_r}(h_s,f_s) & = & (P_rp^-,P_r v_z^-)|_{z=z_r}.
   \label{eqn:fwd}
-  S^{+}_{z_s,z_r}[c]h_s = p^+|_{z=z_r}
-\end{equation}
-The anti-causal forward modeling operator $S^-_{z_s,z_r}$ has a similar definition.
-The subscript signifies that sources are located on $z=z_s$, the receivers on $z=z_r$.
-It's necessary to include this information in the notation, as versions of $S$ with sources and receivers in several locations will be needed in the discussion below.
+\end{eqnarray}
+The subscript signifies that sources are located on $z=z_s$, the
+receivers on $z=z_r$. It is necessary to include this information in
+the notation, as versions of ${\cal S}^{\pm}$ with sources and receivers in
+several locations will be needed in the discussion below.
 
-The adjoint of $S^+_{z_s,z_r}[c]$ can be computed
-by a variant of the adjoint state method, in this case a by-product of the conservation of energy.
-Suppose that $p^-,\bv^-$ solve \ref{eqn:awem} with $h_s\delta(z-z_s)$ replaced by $ h_r \delta(z-z_r)$.
-Then
+The adjoint of ${\cal S}^+_{z_s,z_r}$ can be computed by a variant of
+the adjoint state method, in this case a by-product of the
+conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awem}
+with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
+$ (h_r,f_r\bf{e}_z) \delta(z-z_r)$. Then
 \[
@@
 +
-\left.  (- \nabla p^+) \cdot \bv^- + \bv^+ \cdot (-\nabla p^-) \right)
+\left.  (- \nabla p^++f_s\bf{e}_z) \cdot \bv^- + \bv^+ \cdot (-\nabla
+  p^- + f_r \bf{e_z}) \right)
 \]
@@
 +
-\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- \right)
+\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- 
+  +f_s \delta(z-z_s) v_z^- + v_z^+f_r \delta(z-z_r) \right)
 \]
@@
 \[
-0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, h_sp^-|_{z=z_s} +  h_rp^+|_{z=z_r} = \langle h_s, p^-|_{z=z_s}\rangle + \langle h_r, S[c]^+_{z_s,z_r}[c]h_s \rangle
+0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
+( h_rP_rp^++f_rP_rv_z^+) = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
 \]
@@
 \label{eqn:sadj}
- S^+_{z_s,z_r}[c]^T h_r = -p^-|_{z=z_s}
+ ({\cal S}^+_{z_s,z_r})^T = -{\cal S}^{-}_{z_r,z_s}.
 \end{equation}
@@
 \end{equation}
-Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awep} with $h_s\delta(z-z_s)$ replaced by $-Rh_r\delta(z-z_r)$. That is,
+Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awep} with
+$(h_s\delta(z-z_s),f_s\bf{e_z}\delta(z-z_s))$ replaced by
+$(-Rh_r\delta(z-z_r), Rf_r\bf{e_z}\delta(z-z_r))$. That is,
 \begin{equation}
   \label{eqn:trsadj}
-  Rp^-|_{z=z_s} = -S^+_{z_r,z_s}[c]Rh_r
+  {\cal R}{\cal S}^- = -{\cal S}^+_{z_r,z_s}{\cal R}
 \end{equation}
@@
   \label{eqn:trtr}
-  S^+_{z_s,z_r}[c]^T = RS^+_{z_r,z_s}[c]R.
+ ({\cal S}^+_{z_s,z_r})^T = {\cal R}{\cal S}_{z_r,z_s}^+{\cal R}=
+ -{\cal S}^-_{z_r,z_s}.
 \end{equation}
@@
   \label{eqn:ntoddef}
-  \Lambda_{z_s} = S^+_{z_s,z_s}.
+  \Lambda^{\pm}_{z_s} = S^{\pm}_{z_s,z_s}.
 \end{equation}
@@
   \label{eqn:ntodtr}
-  \Lambda_{z_s}^T = R\Lambda_{z_s}R.
+  (\Lambda^+_{z_s})^T = R\Lambda_{z_s}^+R = -\Lambda^-_{z_s}
 \end{equation}
-This operator is closely related to the operator mapping $\lim_{z\rightarrow z_s^+} \bv_z^+$ to $\lim_{z\rightarrow z_s^+}  p^+$, often called the {\em Neumann-to-Dirichlet} operator.
+This operator is closely related to the operator mapping $\lim_{z\rightarrow z_s^{\pm}} \bv_z^+$ to $\lim_{z\rightarrow z_s^{\pm}}  p^+$, often called the {\em Neumann-to-Dirichlet} operator.
 The connection is explained below.
 
-The quadratic form defined by $\Lambda_{z_s}$ has fundamental physical significance.
-Suppose that the surface source $h_s$ vanishes outside of a bounded time interval $[t_{\rm min},t_{\rm max}]$.
-Then the causal acoustic field $(p^+,\bv^+)$, solution of the system \ref{eqn:awep}, vanishes for $t<t_{\rm min}$.
-Define the total acoustic energy $E(t)$ at time $t$ by
+The quadratic form defined by $\Lambda^{\pm}_{z_s}$ has fundamental physical
+significance. Define the total acoustic energy $E^{\pm}(t)$ of the field $(p^{\pm},\bv^{\pm})$, at time $t$ by
 \begin{equation}
   \label{eqn:defae}
-  E(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^+)^2}{\kappa} + \rho |\bv^+|^2\right)
+  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)
 \end{equation}
-Then since $E(t_{\rm min})=0$,
+Then for $t_{\min}<t_{\max}$,
 \[
-  E(t_{\rm max}) = \int_{t_{\rm min}}^{t_{\rm max}} \frac{dE}{dt}
+  E^{\pm}(t_{\rm max})-E^{\pm}(t_{\rm min}) = \int_{t_{\rm min}}^{t_{\rm max}} \frac{dE^{\pm}}{dt}
 \]
 \[
-  = \int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\, \left(p^+ \frac{1}{\kappa}\frac{\partial p^+}{\partial t} + \bv^+ \cdot \rho \frac{\partial \bv^+}{\partial t}\right)
+  = \int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\, \left(p^{\pm} \frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} + \bv^{\pm} \cdot \rho \frac{\partial \bv^{\pm}}{\partial t}\right)
 \]
 \[
-  =\int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\,\left(p^+(-\nabla \cdot \bv^{+} + h_s\delta(z-z_s)) + \bv^+ \cdot \rho \frac{\partial \bv^+}{\partial t}\right)
+  =\int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\,\left(p^{\pm}(-\nabla \cdot \bv^{+} + h_s\delta(z-z_s)) + \bv^{\pm} \cdot \rho \frac{\partial \bv^{\pm}}{\partial t}\right)
 \]
 \[
-  \int_{t_{\rm min}}^{t_{\rm max}}\int\,d\bx\, \left(\left(\nabla p^+\rho \frac{\partial \bv^+}{\partial t}\right) + p^+h_s\delta(z-z_s)\right)
+=  \int_{t_{\rm min}}^{t_{\rm max}}\int\,d\bx\, \left(\left(\nabla p^{\pm}+\rho \frac{\partial \bv^{\pm}}{\partial t}\right) + p^{\pm}h_s\delta(z-z_s)\right)
+\]
+\begin{equation}
+  \label{eqn:eident}
+  =  \int_{t_{\rm min}}^{t_{\rm max}}\int\,dxdy h_s p^{\pm}|_{z=z_s} = \int_{t_{\min}}^{t_{\max}}\langle h_s(t), (\Lambda^{\pm}_{z_s} h_s)(t) \rangle_{L^2(\bR^2)}.
+\end{equation}
+That is, $\langle h_s(t), (\Lambda^{\pm}_{z_s} h_s)(t)
+\rangle_{L^2(\bR^2)}$ is the rate of energy transfer from source to
+the fluid at time $t$.
+
+Assume that $h_s$ has compact support in time. Then $E^{\pm}
+\rightarrow 0$ as $t \rightarrow \mp \infty$, and is eventually
+constant as $t \rightarrow \pm \infty$. Then the energy identity
+\ref{eqn:eident} implies that
+\begin{equation}
+  \label{eqn:etot}
+  \pm \lim_{t \rightarrow \pm \infty} E^{\pm}(t) = \int \,dt\, \langle
+  h_s(t), (\Lambda^{\pm}_{z_s} h_s)(t)\rangle_{L^2(\bR^2)} = \langle
+  h_s,\Lambda^{\pm}_{z_s} h_s\rangle
+\end{equation}
+By virtue of identity \ref{eqn:ntodtr},
+\[
+  \lim_{t \rightarrow -\infty}E^{-}(t) =-\langle
+  h_s,\Lambda^{-}_{z_s} h_s\rangle
 \]
 \[
-  =  \int_{t_{\rm min}}^{t_{\rm max}}\int\,dxdy h_s p^+|_{z=z_s} = \langle h_s, \Lambda_{z_s} h_s \rangle.
+  =  \langle  h_s,(\Lambda^{^+})^T_{z_s} h_s\rangle
 \]
-Since the source $h_s\delta(z-z_s)$ no longer acts on the fluid for $t >t_{\rm max}$, $E_{\rm tot}= E(t_{\rm max})$ is the total energy transferred from the source to the fluid, and
+\[
+  =  \langle  h_s,\Lambda^{^+}_{z_s} h_s\rangle
+\]
+so
 \begin{equation}
   \label{eqn:surfenergy}
-  E_{\rm tot} = \langle h_s, \Lambda_{z_s} h_s \rangle.
+  = \lim_{t \rightarrow \infty}E^{+}(t)
 \end{equation}
-We can conclude from identity \ref{eqn:surfenergy} that the symmetric part of $\Lambda_{z_s}$ is positive semidefinite:
+That is, the energy total energy transferred from the source to the
+causal acoustic field as it evolves forward in time is the same as the
+total energy transferred to the anti-causal acoustic field as it
+evolves backwards in time.
+
+Also onclude from identity \ref{eqn:etot} that the symmetric part of $\Lambda^+_{z_s}$ is positive semidefinite:
 \begin{equation}
   \label{eqn:lamsd}
-  \frac{1}{2}(\Lambda_{z_s}^T + \Lambda_{z_s}) \ge 0.
+  \frac{1}{2}((\Lambda_{z_s}^+)^T + \Lambda^+_{z_s}) \ge 0.
 \end{equation}
 
-Observe that the causal acoustic system \ref{eqn:awep} implies directly that
+%Observe that the causal acoustic system \ref{eqn:awep} implies directly that
+%\begin{equation}
+%  \label{eqn:sourcevjump}
+%  h_s= [v_z]|_{z=z_s}.
+%\end{equation}
+
+\section{Parametrix via Time Reversal}
+
+Define $H$ to be the subspace $\{u \in (L^2(\bR^3))^4: \mbox{supp
+}u\subset \{z<z_s\}\}$, by analogy with \cite{Lax:PDENotes},
+p. 122. Note: must add microlocal $H^1$ constraint per
+\cite[]{BaoSy:91b} so that trace operator is well-defined - postpone
+this.
+
+Given $u=(\bar{p}_0,\bar{\bv}_0) \in H$, extend in $t$ by solving the acoustic system
+\begin{eqnarray}
+\label{eqn:awe0}
+  \frac{1}{\kappa}\frac{\partial p_0}{\partial t} & = & - \nabla \cdot \bv_0, \nonumber \\
+  \rho\frac{\partial \bv_0}{\partial t} & = & - \nabla p_0,\nonumber \\
+  p_0 & =& \bar{p}_0, t=0,\nonumber\\ 
+  \bv_0 & = & \bar{\bv}_0, t=0,\nonumber\\
+\end{eqnarray}
+Note that $(p_0,\bv_0)=0$ near $t=0$ in the region $z>z_s$.
+
+Denote by $P_s$, respectively $P_r$, the trace operator on $z=z_s$,
+respectively $z=z_r$. Given $(p_0,\bv_0)$ as above, define
+$(p_1,\bv_1)$ as the solution in $z \ne z_s$ of
+\begin{eqnarray}
+\label{eqn:awe1}
+  \frac{1}{\kappa}\frac{\partial p_1}{\partial t} & = & - \nabla \cdot \bv_1, \nonumber \\
+  \rho\frac{\partial \bv_1}{\partial t} & = & - \nabla p_1,\nonumber \\
+  p_1 & =& 0,  \mbox{ for } 0,\nonumber\\ 
+  \bv_1 & = & 0 \mbox{ for } 0,\nonumber\\
+  P_sp_1 = P_sp_0. 
+\end{eqnarray}
+Extended to form a distribution on all of $\bR^4$, $(p_1,\bv_1)$ solve the
+inhomogeneous system
+\begin{eqnarray}
+\label{eqn:awe1}
+  \frac{1}{\kappa}\frac{\partial p_1}{\partial t} & = &
+                                                        ((\Lambda^+_{z_s})^{-1}P_sp_0)\delta(z-z_s) - \nabla \cdot \bv_1, \nonumber \\
+  \rho\frac{\partial \bv_1}{\partial t} & = & - \nabla p_1,\nonumber \\
+  p_1 & =& 0,  \mbox{ for } t=0,\nonumber\\ 
+  \bv_1 & = & 0 \mbox{ for } t=0.\nonumber\\
+\end{eqnarray}
+where
 \begin{equation}
-  \label{eqn:sourcevjump}
-  h_s= [v_z]|_{z_s}.
+  \label{eqn:defhs}
+  (\Lambda^+_{z_s})^{-1}P_sp_0 = P_s[(\bv_1)_z].
 \end{equation}
+
+Since $p_0$ also solves the acoustic system \ref{eqn:awe1} in $z>z_s$
+and has the same data at $t=0, z>z_s$ and trace on $z=z_s$, standard
+uniqueness theorem shows that $(p_1,\bv_1) = (p_0, \bv_0)$ in $z>z_s$.
+
+Introduce the operator $T^+_{z_s,z_r}$ mapping pressure on $z=z_s$ to
+pressure on $z=z_r$: that is, $T^{+}_{z_s,z_r}P_sp_0 = P_rp_0$. From
+the previous observation, $T^{+}_{z_s,z_r}P_sp_0 = P_rp_1$, that is,
+you can compute $T^{+}$ by solving the system defining $p_1$, with
+data $P_sp_0$. From the definition of $S^{+}_{z_s,z_r}$, the definition
+\ref{eqn:defhs}, and the boundary condition for $p_1$,
+\begin{equation}
+  \label{eqn:relnST}
+  T^+_{z_s,z_r} = S^+_{z_s,z_r} (\Lambda^+_{z_s})^{-1}.
+\end{equation}
+
+Let $\epsilon>0$ satisfy
+\begin{equation}
+  \label{eqn:defeps}
+  E[{\bf 1}_{\bR^2 \times (-\infty,z_r]}(p_0,\bv_0)](T) \le \epsilon E[(p_0,\bv_0)]
+\end{equation}
+Omit the time on the RHS because the energy of the full field is
+time-invariant. Inequality \ref{eqn:defeps} states that at time
+$t=T$, the energy in
+the region $z<z_r$ is at most $\epsilon$ times the energy in all of
+space. For small $\epsilon$, in other words, most of the
+energy has left $z<z_r$, at least temporarily, at $t=T$. This
+assertion is similar to the outgoing field condition in Lax-Phillips
+scattering theory, but is time-local and approximate.
+
+Define $p_2$ to be the solution in $z \le z_r$ of
+\begin{eqnarray}
+\label{eqn:awe2}
+  \frac{1}{\kappa}\frac{\partial p_2}{\partial t} & = & - \nabla \cdot \bv_2, \nonumber \\
+  \rho\frac{\partial \bv_2}{\partial t} & = & - \nabla p_2,\nonumber \\
+  p_2 & =& 0,  \mbox{ for } t=T,\nonumber\\ 
+  \bv_2 & = & 0 \mbox{ for } t=T,\nonumber\\
+  P_rp_2 = P_rp_0. 
+\end{eqnarray}
+Then $(\delta p_2,\delta \bv_2)=(p_0-p_2,\bv_0-\bv_2)$ solves
+\begin{eqnarray}
+\label{eqn:awe2}
+  \frac{1}{\kappa}\frac{\partial \delta p_2}{\partial t} & = & -
+                                                               \nabla
+                                                               \cdot
+                                                               \delta \bv_2, \nonumber \\
+  \rho\frac{\partial \delta \bv_2}{\partial t} & = & - \nabla \delta p_2,\nonumber \\
+  \delta p_2 & =& p_0,  \mbox{ for } t=T,\nonumber\\ 
+  \delta \bv_2 & = & \bv_0 \mbox{ for } t=T,\nonumber\\
+  P_r\delta p_2 = 0. 
+\end{eqnarray}
+Using the standard energy estimate for homogenous Dirichlet
+problem,
+\begin{equation}
+  \label{eqn:errtr}
+  E[{\bf 1}_{\bR^2 \times (-\infty,z_r]}(\delta p_2,\delta \bv_2)](t) \le
+  \epsilon E[(p_0,\bv_0)], \, 0 \le t \le T.
+\end{equation}
+
+Define $T^{-}P_rp_0 = P_sp_2$. Then
+\[
+  T_{z_r,z_s}^{-}P_rp_0-P_sp_0 = P_s\delta p_2.
+\]
+For the time being, write $O(\epsilon)$ for errors in the trace at
+$z_s$ (output of $P_s$) for solutions of the forward Dirichlet problem
+for initial data satisfying the estimate \ref{eqn:errtr}. Of course
+this does not imply any $L^2$ bound on the trace - for that,
+additional constraints on the initial data are required, to be
+discussed below. With this convention,
+\[
+  T_{z_r,z_s}^{-}P_rp_0 =
+  S^{-}_{z_r,z_s}(\Lambda^{-}_{z_r})^{-1}P_rp_0
+\]
+\[
+  =  S^{-}_{z_r,z_s}(\Lambda^{-}_{z_r})^{-1}T_{z_s,z_r}^+P_sp_0
+\]
+\[
+=
+S^{-}_{z_r,z_s}(\Lambda^{-}_{z_r})^{-1}S_{z_s,z_r}^+(\Lambda^{+}_{z_s})^{-1}P_sp_0
+\]
+\[
+  = P_s p_0 + O(\epsilon)
+\]
+Setting $h_s= +(\Lambda^{+}_{z_s})^{-1}P_sp_0$ and applying
+$+(\Lambda^{+}_{z_s})^{-1}$ to both sides, obtain
+\[
+  h_s =
+  (\Lambda^{+}_{z_s})^{-1}S^{-}_{z_r,z_s}(\Lambda^{-}_{z_r})^{-1}S_{z_s,z_r}^+h_s
+  + O(\epsilon)
+\]
+\[
+  =
+ -(\Lambda^{+}_{z_s})^{-1}(S^{+}_{z_r,z_s})^T(\Lambda^{-}_{z_r})^{-1}S_{z_s,z_r}^+h_s
+  +O(\epsilon)
+\]
+\begin{equation}
+  \label{eqn:approxunit}
+ =(\Lambda^{+}_{z_s})^{-1}(S^{+}_{z_r,z_s})^T((\Lambda^{+}_{z_r})^{-1})^TS_{z_s,z_r}^+h_s
+  +O(\epsilon)
+\end{equation}
+The last two steps are justified by appeal to identities
+\ref{eqn:trtr} and \ref{eqn:ntodtr}.
+
+Define
+\begin{equation}
+  \label{eqn:appinv}
+  (S^+_{z_s,z_r})^{\dagger} =
+  (\Lambda^{+}_{z_s})^{-1}(S^{+}_{z_r,z_s})^T((\Lambda^{+}_{z_r})^{-1})^T.
+\end{equation}
+Then equation \ref{eqn:approxunit} asserts that
+$(S^+_{z_s,z_r})^{\dagger}$ is an approximate inverse to
+$S^+_{z_s,z_r}$, in the sense that the error is the trace of a
+ssolution with small Cauchy condition at $t=0$. Later will specify
+conditions under which the output error is small in $L^2$.
 
@@
   \label{eqn:asymjump}
-  [v_z]\approx \lim_{z\rightarrow 0^+} v_{z}
+  [v_z]\approx 2\lim_{z\rightarrow 0^+} v_{z}
 \end{equation}
@@
   \label{eqn:psidodton}
-  \Lambda_s h_s(x,y,t) \approx \frac{1}{2}\int_{\Gamma(x,y)} \,d\omega\,dk_x\,dk_y
+  \Lambda^+_{z_s} h_s(x,y,t) \approx \frac{1}{2}\int_{\Gamma(x,y)} \,d\omega\,dk_x\,dk_y
   \lambda_s(x,y,t,k_x,k_y,\omega) \hat{h_s}(k_x,k_y,\omega) e^{\i(\omega t + k_x x+ k_y y)}
@@
   \label{eqn:symboldton}
-  \lambda_s(x,y,t,k_x,k_y,\omega) =\frac{1}{2}\rho(x,y,z_s)c(x,y,z_s)\left(1-c^2(x,y,z_s)\left(\left(\frac{k_x}{\omega}\right)^2 + \left(\frac{k_y}{\omega}\right) ^2\right)\right)^{-1/2}
+  \lambda^+_{z_s}(x,y,t,k_x,k_y,\omega) =\frac{1}{2}\rho(x,y,z_s)c(x,y,z_s)\left(1-c^2(x,y,z_s)\left(\left(\frac{k_x}{\omega}\right)^2 + \left(\frac{k_y}{\omega}\right) ^2\right)\right)^{-1/2}
 \end{equation}
 
-Restricted to $\Gamma$, $\lambda \in S^0_{0,1}(\bR^2)$ is real-valued, hence defines an essentially symmetric operator. That is,
+Restricted to $\Gamma$, $\lambda^+_{z_s} \in S^0_{0,1}(\bR^2)$ is real-valued, hence defines an essentially symmetric operator. That is,
 \begin{equation}
   \label{eqn:approxsymm}
-  \Lambda_s^T \approx \Lambda_s.
+  (\Lambda^+_{z_s})^T \approx \Lambda^{+}_{z_s}.
 \end{equation}
 
-\section{Parametrix via Time-reversal}
+Note that the defintion \ref{eqn:appinv} of $ (S^+_{z_s,z_r})^{\dagger} $ is that of
+the transpose of $S^+_{z_s,z_r}$ in the sense of weighted norms, as
+follows. Define trace spaces $W=L^2(\bR^2 \times [0,T])$, with inner product
+\[
+  \langle h_1,h_2\rangle_s = \langle
+  h_1,\frac{1}{2}(\Lambda^+_{z_s}+(\Lambda^+_{z_s})^T)h_2\rangle
+\]
+and $D=L^2(\bR^2 \times [0,T])$ with inner product
+\[
+  \langle d_1,d_2\rangle_r = \langle
+  d_1,\left(\frac{1}{2}(\Lambda^+_{z_s}+(\Lambda^+_{z_s})^T)\right)^{-1}d_2\rangle
+\]
+(These definitions must be supplemented with constraints on
+out-of-aperture smoothness). Then the adjoint of $S^+_{z_s,z_r}: W
+\rightarrow D$ is
+\[
+ \left(\frac{1}{2}
+   (\Lambda^+_{z_s}+(\Lambda^+_{z_s})^T)\right)^{-1}(S^+_{z_s,z_r})^T
+ \left(\frac{1}{2}(\Lambda^+_{z_r}+(\Lambda^+_{z_r})^T)\right)^{-1}
+\]
+ By virtue of the approximation \ref{eqn:approxsymm}, this is
+\[
+   \approx
+   (\Lambda^{+}_{z_s})^{-1}(S^{+}_{z_r,z_s})^T((\Lambda^{+}_{z_r})^{-1})^T
+   = (S^+_{z_s,z_r})^{\dagger}.
+\]
+which from the last section is approximately inverse to
+$S^+_{z_s,z_r}$. That is, with the domain and range described above,
+$S^+_{z_s,z_r}$ is approximately unitary.
 
-Both $S[c]^T$ and $S[c]^{\dagger}$ are defined by time-reversed solutions of the same acoustic system, except with different data:
-\begin{itemize}
-\item system \ref{eqn:aweadj} inputs $d_r$ as a pressure source on $z=z_r$, with an additional factor of $\kappa$, and outputs the trace of the reverse-time propagated field on $z=z_s$, scaled by $1/\kappa$, to obtaint $S[c]^Td_r$;
-\item system \ref{eqn:awetr} inputs $d_r$ as a Dirichlet boundary condition, and the trace of its pressure component must be employed as a Dirichlet boundary condition for the forward-in-time problem \ref{eqn:awedir}, after which the jump in the normal velocity is extracted.
-\end{itemize}
-
-The relation between a Dirichlet boundary condition and a source has already been explained, in the context of justifying the approximation property of $S[c]^{\dagger}$:
-\begin{quote}
-The surface source producing the same field as a Dirichlet boundary condition (on either side of the boundary surface) is the jump in the normal velocity component of the Dirichlet field. 
-\end{quote}
-Call the operator that produces the jump in normal velocity from Dirichlet pressure data $\Lambda[c]_s$ for the source surface $z=z_s$ and forward in time, $\Lambda[c]_r$ for the receiver surface $z=z_r$ backwards in time (in the math literature, these are ``Dirichlet-to-Neumann operators'', more or less).
+\section{Efficient Preconditioned CG}
 
@@
 This concept also turns up in hidden form in the work of Yu Zhang and collaborators on true amplitude migration \cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,Zhang YuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.
-
-\section{Efficient Preconditioned CG}
 

### 2021-05-04 repro/papers/sseprecond/paper.tex
interim

@@
 
-The energy source $\bff$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time.
+%The energy source $\bff$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time.
 In fact, additional source degrees of freedom, beyond those needed to describe physically realized sources, may be useful in rendering the FWI problem \ref{eqn:fwi} more amenable to numerical solution, via so-called extended modeling (see \cite{geoprosp:2008}, \cite{LeeuwenHerrmannWRI:13}, \cite{HuangNammourSymesDollizal:SEG19}, and many references cited there).
 Therefore it is natural to view $\bff$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
-%A typical nonlinear least squares formulation is:
-%\begin{equation}
-%\label{eqn:ols}
-%\mbox{choose } c \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2.
-%\end{equation}
-
-%As is well-known, local optimization methods are the only feasible approach given the dimensions of a typical instance of \ref{eqn:fwi}, and those have a tendency to stall due to ``cycle-skipping''.
-Source extension is one approach to avoiding this problem.
-It consists in imposing the wave equation as a soft as opposed to hard constraint, by allowing the source to have more degrees of freedom than is permitted by a faithful  model model of the seismic experiment, and constraining these additional degrees of freedom by means of an additive penalty modifying the probem \ref{eqn:ols}:
+A simple nonlinear least squares formulation is:
+\begin{equation}
+\label{eqn:ols}
+\mbox{choose } \bf{c} \mbox{ to minimize } \|PL[\bf{c}]^{-1}\bff -d \|^2.
+\end{equation}
+Practical optimization formulations typically augment the objective in
+\ref{eqn:ols} by additive penalties or other constraints.
+
+As is well-known, local optimization methods are the only feasible
+approach given the dimensions of a typical instance of \ref{eqn:fwi},
+and those have a tendency to stall due to ``cycle-skipping''. Source
+extension is one approach to avoiding this problem. It consists in
+imposing the wave equation as a soft as opposed to hard constraint, by
+allowing the source field $\bff$ to have more degrees of freedom than
+is permitted by a faithful model model of the seismic experiment, and
+constraining these additional degrees of freedom by means of an
+additive penalty modifying the probem \ref{eqn:ols}:
 \begin{equation}
@@
 \end{equation}
-The operator $A$ penalizes deviation from known (or assumed) characteristics of the source function - its null space consists of feasible source models.
+The operator $A$ penalizes deviation from known (or assumed)
+characteristics of the source function - its null space consists of
+feasible (or ``physical'') source models.
+
+\cite{HuangNammourSymesDollizal:SEG19} present an overview of the
+literature on source extension methods, describing a variety of
+methods to add degrees of freedom to physical source model. The present paper
+concerns {\em surface source extension}: physical sources are
+presumed to be concentrated at points $\bx_s$ in space, whereas their extended
+counterparts are permitted to spread energy over surfaces containing
+the physical source locations. A simple choice for the penalty
+operator $A$ is then multiplication by the distance $|\bx-\bx_s|$ to the physical
+source location:
+\begin{equation}
+  \label{eqn:penop}
+  (A\bff)(\bx,t) = |\bx-\bx_s|\bff(\bx,t)
+\end{equation}
+I shall use this choice of penalty operator whenever a specific choice
+is necessary in the development of the theory below.
 
@@
 \end{equation}
-
-The penalty operator $A$ is assumed linear in this work, so the source subproblem is a linear least squares problem.
-Under some additional assumptions to be described below, I shall show how to construct an accurate approximate solution operator for problem \ref{eqn:esis}.
-This approximate solution operator may be used to accelerate Krylov space methods for the solution of the source subproblem \ref{eqn:esis}.
-Numerical examples suggest the effectiveness of this acceleration.
-
-The source subproblem is a necessary step in solving the overall FWI problem \ref{eqn:fwi}.
-In particular, the nonlinear least squares realization \ref{eqn:esi} of source-extended FWI is naturally approached via the {\em variable projection method} \cite[]{GolubPereyra:73,GolubPereyra:03,vanLeeuwenMulder:09,Rickett:SEG12}, in which the minimum over $\bff$ in \ref{eqn:esis} is treated as a function of $\bf{c}$, which is in turn minimized.
-Variable projection thus treats the source subproblem as an inner problem, which must be solved for an estimate of $\bff$ at every step of an iterative method for the outer problem of minimization over $\bf{c}$.
-Efficiency in solving the source subproblem \ref{eqn:esis} is critical to efficient implementation of variable projection for solving the regularized FWI problem \ref{eqn:esi}.
-
-This paper restricts the choice of wave physics embodied in
-$L[\bf{c}]$ to linear acoustics.  Thus $\bf{c}=(\kappa,\rho)$ is the
-pair (bulk modulus, density) of positive functions of spatial
-position.
-The principal characteristics assumed of sources $\bff=$ are localization and isotropy.
-That is, physical sources will be assumed to be isotropic point radiators, described by the functional form
-\begin{equation}
-\label{eqn:ptsrc}
-f_{\rm pt}(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s). 
-\end{equation}
-Source locations $\bx_s$ are assumed known (as is natural for active source methods), and the source {\em wavelet} $w(t)$ is to be found as part of the solution of \ref{eqn:esi}.
-A convenient choice of penalty operator $A$ is then multiplication by the {\em offset} $|\bx-\bx_s|$: its null space consists precisely of distributions of the form \ref{eqn:ptsrc}.
-
-More complex radiation characteristics may be accommodated with similar but more complex choice for the operator $A$.
-Since the core analysis is similar, this paper deals only with the simplest case \ref{eqn:ptsrc}.
-
-The wavefield $u$ in this work is presumed to exist in all of Euclicean space-time: in other words, boundary effects have been removed from $u$, by surface-related multiple elimination or other techniques.
-Predicted data traces are simply the samples of one or more components of $u$ at receiver points $\bx_r$.
-
-%The domain of wave propagation is denoted $\Omega$, and two subsets of its boundary are identified: $\Gamma_s$ contains the support (non-zeroes) of the (extended) sources and in particular the locations $\bx_s$ of physical sources.
-Data trace locations $\bx_r$ lie in another subset $\Gamma_r$.
-The data projection operator $P$ samples the acoustic field $u$ on $\Gamma_r$, and includes a finite aperture mute.
-
-Achieving a small residual in the source subproblem \ref{eqn:esis} for arbitrary $c$ is a critical part of the convergence mechanism for the extended source approach to FWI \cite[]{HuangNammourSymesDollizal:SEG19}.
-However, for arbitrary data $d$, coefficient array $c$, and source location $\bx_s$, there does not exist an isotropic point source $f_{\rm pt}$ of the form \ref{eqn:ptsrc} for which $PL[c]^{-1}f_{\rm pt} \approx d$.
-This is so even if there does exist a coefficient array $c^*$ and point source $f_{\rm pt}^*$ located at $\bx_s$ with wavelet $w^*(t)$ for which $d = PL[c^*]^{-1}f_{\rm pt}^*$, if $c$ differs substantially from $c^*$.
-It is this fact that underlies the effectiveness of the extended source approach to FWI:  driving the residual in the problem \ref{eqn:esi} towards zero necessarily requires modifying $c$ to resemble $c^*$, assuming the latter exists, as well as pushing the optimal $f$ towards the null space of $A$.
-Conversely, the source subproblem \ref{eqn:esis} does not have a small residual solution of point source form \ref{eqn:ptsrc}, in general, for a given choice of $c$, even if that were the case for a different choice ($c^*$).
-Therefore a more general class of source models than that specified by the condition \ref{eqn:ptsrc} must be admitted to the feasible set for the source subproblem \ref{eqn:esis} if it is to serve as the inner problem in a variable projection formulation of extended source FWI.
-
-The class of non-point sources investigated here are those confined a hypersurface assumed to contain the locations of physical (point) sources, but not {\em a priori} required to have point support.
-These {\em surface extended sources} are able to yield a relatively small residual in the source subproblem \ref{eqn:esis} for more or less arbitrary coefficient array $c$.
-This observation is a by-product of the approximate solution for \ref{eqn:esis} constructed below.
-The construction of the approximate solution is closely related to the {\em time reversal} method of photoacoustic tomography (see \cite{StefanovUhlmannIP:09} and references cited there).
-
-The next section gives a precise description of an acoustic version of problem \ref{eqn:esis} for an idealized {\em crosswell} configuration, in which source located on a surface $z=z_s$ generate waves that propagate to to a receiver surface $z=z_r > z_s$.
-In principle, {\em diving waves} can be described with similar mathematics, but with some additional complications, so I do not treat the diving wave case explicitly in this paper.
-The assumptions detailed here imply that the data $d$ of the inverse problems \ref{eqn:esi}, \ref{eqn:esis} have the physical character of {\em transmitted waves}.
-There are no mathematical results at present concerning reflected wave inverse problems treated by source extension methods, though there are tantalizing numerical clues \cite[]{LeeuwenHerrmannWRI:13,Warner:14,Warner:16,LeeuwenHerrmann:16,HuangSymes:Geo17,HuangSymes:Geo18a,HuangSymes:Geo18b}.
-
-The following section describes the construction of the approximate
-inverse and explains the conditions under which accuracy should be expected.
-I observe that the approximate inverse is approximately the adjoint of the modeling operator $PL[c]^{-1}$ with respect to weighted norms in its domain and range (see \cite{HouSymes:15} for a similar observation in a different context).
-This fact immediately suggests use of the approximate inverse as a preconditioner to accelerate convergence of Krylov space methods such as conjugate gradient iteration applied to the least squares problem \ref{eqn:esis}.
-However, its straightforward application in the preconditioned conjugate gradient (PCG) method \cite[]{Golub:2012} involves explicitly a version the so called Dirichlet-to-Neumann (D2N) operator, which is computationally awkward.
-I show how to reorganize the PCG iteration so that only solutions of $L[c]u=f$ are required.
-The penultimate section presents some 2D examples in the crosswell configuration, illustrating the accuracy of the approximate inversion and the accelerated convergence of PCG.
-The examples are chosen to emphasize that surface extended sources can be constructed to fit more or less arbitrary data, unlike point sources - as mentioned before, this approximate invertibility property of the modeling operator is of critical importance in the application to extended FWI.
-The paper ends with a discussion of several important matters not addressed here, and a restatement of the conclusions.
-
-The intent of the main body of this paper is to present the formal structure of the extended source subproblem and the way in which this structure leads to accelerated numerical solution.
-This formal structure is however supported by a rigorous foundation.
-In order to avoid disrupting the formal account, this foundation material is relegated to an appendix.
-
-\section{Surface Sources and Solutions}
-
-For acoustic wave physics, the coefficient vector $c$ is the pair $\bf{c}=(\kappa,\rho)$ of bulk modulus $\kappa$ and density $\rho$, and the state vector $\bu=(p,\bv)^T$ consists of pressure $p$ (a scalar space-time field) and particle velocity $\bv$ (a vector space-time field).
-The wave operator $L[\bf{c}]$ is given by the linear acoustics system:
+Solution of this subproblem is an essential component of {\em variable
+  projection} algorithms for solution of the nonlinear inverse problem
+\ref{eqn:esi}. Variable projection is not merely a convenient choice
+of algorithm for this purpose: it is in some sense essential, see for
+example \cite{Symes:SEG20}. It replaces the nonlinear
+least squares problem \ref{eqn:esi} with a {\em reduced} problem, to
+be solved iteratively. Each iteration involves solution of the
+subproblem \ref{eqn:esis}. Therefore efficient solution of the
+subproblem is essential to efficient solution of the nonlinear problem
+via variable projection.
+
+The penalty operator $A$ defined in \ref{eqn:penop} is linear, so the source
+subproblem is a linear least squares problem. Under some additional
+assumptions to be described below, I shall show how to construct an
+accurate approximate solution operator for problem
+\ref{eqn:esis}. This approximate solution operator may be used to
+accelerate Krylov space methods for the solution of the source
+subproblem \ref{eqn:esis}. Numerical examples suggest the
+effectiveness of this acceleration.
+
+\section{Operators and Adjoints}
+
+For acoustic wave physics, the coefficient vector is
+$\bf{c}=(\kappa,\rho)$, with components bulk modulus $\kappa$ and
+density $\rho$, and the state vector $\bu=(p,\bv)^T$ consists of
+pressure $p$ (a scalar space-time field) and particle velocity $\bv$
+(a vector space-time field). The wave operator $L[\bf{c}]$ is:
 \begin{equation}
 \label{eqn:aweop}
-L[c]u = 
+L[\bf{c}]\bf{u} = 
 \left(
@@
   \label{eqn:awemat}
-  L[c] = \left(
+  L[\bf{c}] = \left(
     \begin{array}{cc}
@@
 
-Most of what follows is valid for any space dimension $n >0$. Examples
-later in this paper will use $n=2$ for computationally convenience. In
-the mathematical development that follows, I will set $n=3$.
-
-As mentioned earlier, in this paper the coefficients $\bf{c}$ is regarded as defined throughout space $\bR^d$, the state vector $\bu$ throughout space-time $\bR^{d+1}$.
-
-The idealized ``crosswell'' configuration involves sources confined to
-$z=z_s$. Such sources are combinations of constitutive law defects and
+Most of what follows is valid for any space dimension $n >0$. The
+coefficient vector $\bf{c}=(\kappa,\rho)$ is defined throughout space
+$\bR^n$, the state vector $\bu$ throughout space-time
+$\bR^{n+1}$. Whenever convenient for mathematical manipulations,
+$n=3$: for instance, I will write $\bx=(x,y,z)^T$ for the spatial
+coordinate vector, and refer to the third (vertical) coordinate of
+particle velocity as $v_z$. Examples
+later in this paper will use $n=2$ for computational convenience.
+
+Since all of the operators in the discussion that follows depend on
+$\bf{c}$, I will suppress it from the notation, for example, $L=L[\bf{c}]$.
+
+The extended ``crosswell'' configuration involves sources confined to
+a spatial hypersurface. For convenience, choose this to be the hyperplane
+$\{(x,y,z,t): z=z_s\}$. Such sources are combinations of constitutive law defects and
 loads normal to the surface, that is, right-hand sides in the system
-$L[c]\bf{u}=\bf{f}$ of the form $\bf{f}(\bx,t) =
+$L\bu=\bff$ of the form $\bff(\bx,t) =
 (h_s(x,y,t)\delta(z-z_s), f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar
-defect $h_s$ and normal force $f_s$ and $\bf{e}_z=(0,0,1)^T$. With the choice $L[\bf{c}]$ given in \ref{eqn:aweop}, the causal wave system $L[\bf{c}]\bu^+=\bff$ takes the form
-\begin{eqnarray}
-\label{eqn:awep}
-\frac{1}{\kappa}\frac{\partial p^{+}}{\partial t} & = & - \nabla \cdot \bv^{+} +
+defect $h_s$ and normal force $f_s$  ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in \ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$ takes the form
+\begin{eqnarray}
+\label{eqn:awepm}
+\frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} & = & - \nabla \cdot \bv^{\pm} +
 h_s \delta(z-z_s), \nonumber \\
-\rho\frac{\partial \bv^{+}}{\partial t} & = & - \nabla p^{+} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
-p^{+} & =& 0 \mbox{ for } t \ll 0,\nonumber\\ 
-\bv^{+} & = & 0 \mbox{ for } t \ll 0.
+\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
+p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
+\bv^{\pm} & = & 0 \mbox{ for } \pm t \ll 0.
 \end{eqnarray}
 
-Similarly, the anti-causal wave system $L[c]u^-=f$ takes the form
-\begin{eqnarray}
- \label{eqn:awem}
-\frac{1}{\kappa}\frac{\partial p^{-}}{\partial t} & = & - \nabla \cdot \bv^{-} +
-h_s \delta(z-z_s), \nonumber \\
-\rho \frac{\partial \bv^{-}}{\partial t} & = & - \nabla p^{-} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
-p^{-} & =& 0 \mbox{ for } t \gg 0,\nonumber\\ 
-\bv^{-} & = & 0 \mbox{ for } t \gg 0.
-\end{eqnarray}
-
-\section{Operators and Adjoints}
-All of the operators discussed in this section depend on the
-coefficient array $\bf{c}$, which I will suppress from the notation.
-Denote by $P_s,P_r$ the trace (sampling) operators on $z=z_s$, $z=z_r$
-respectively. The causal and anticausal modeling operators
-${\cal S}^{+}_{z_s,z_r}$ and ${\cal S}^-_{z_s,z_r}$ are defined in
+\noindent {\bf Remark:} In system \ref{eqn:awepm} and many similar
+systems to follow, I will use the shorthand
+\[
+  p^+ = 0 \mbox{ for } t \ll 0
+\]
+to mean
+\[
+  \mbox{For some } T \in \bR, p^+(\cdot,t) = 0 \mbox{ for all } t \ge
+  T.
+\]
+Extended forward modeling consists in solving \ref{eqn:awepm} and
+sampling the solution components on another spatial hyperplane $\{(x,y,z,t): z=z_r\}$. Denote by $P_s,P_r$ the trace (sampling) operators on $z=z_s$, $z=z_r$
+respectively. The causal/anti-causal modeling operators
+${\cal S}^{\pm}_{z_s,z_r}$ are defined in
 terms of the solutions $(p^{\pm},\bv^{\pm})$ of the systems
-\ref{eqn:awep} and \ref{eqn:awem} by
-\begin{eqnarray}
-  {\cal S}^{+}_{z_s,z_r}(h_s,f_s) & = & (P_rp^+,P_r v_z^+)|_{z=z_r},
-                                        \nonumber\\
-  {\cal S}^{-}_{z_s,z_r}(h_s,f_s) & = & (P_rp^-,P_r v_z^-)|_{z=z_r}.
+\ref{eqn:awepm} by
+\begin{equation}
+  {\cal S}^{\pm}_{z_s,z_r}(h_s,f_s)  = (P_rp^{\pm},P_r v_z^{\pm}),
   \label{eqn:fwd}
-\end{eqnarray}
+\end{equation}
 The subscript signifies that sources are located on $z=z_s$, the
@@
 
-The adjoint of ${\cal S}^+_{z_s,z_r}$ can be computed by a variant of
-the adjoint state method, in this case a by-product of the
-conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awem}
-with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
-$ (h_r,f_r\bf{e}_z) \delta(z-z_r)$. Then
-\[
-0 = 
-\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
-\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
--
-\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
-\]
-\[
-+
-\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
- h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
- h_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  (- \nabla p^++f_s\bf{e}_z) \cdot \bv^- + \bv^+ \cdot (-\nabla
-  p^- + f_r \bf{e_z}) \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
- h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
- h_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- 
-  +f_s \delta(z-z_s) v_z^- + v_z^+f_r \delta(z-z_r) \right)
-\]
-after integration by parts in the last two terms. Most of what is left cancels, leaving 
-\[
-0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
-( h_rP_rp^++f_rP_rv_z^+) = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
-\]
-whence
-\begin{equation}
-\label{eqn:sadj}
- ({\cal S}^+_{z_s,z_r})^T = -{\cal S}^{-}_{z_r,z_s}.
-\end{equation}
-
-The systems \ref{eqn:awep} and \ref{eqn:awem} differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
-$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
+It follows from the adjoint state method (see Derivations section
+below for details) that
+\begin{equation}
+  \label{eqn:sadj1}
+  ({\cal S}^{\pm}_{z_s, z_r})^T = -{\cal S}^{\mp}_{z_r,z_s}
+\end{equation}
+
+Define $R$ to be the {\em time-reversal operator} on functions of
+space-time, $Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em
+  acoustic field time-reversal operator}
 \begin{equation}
@@
 \end{equation}
+Then 
+\begin{equation}
+  \label{eqn:trsadj}
+  {\cal R}{\cal S}^{\mp} = -{\cal S}^{\pm}_{z_r,z_s}{\cal R}
+\end{equation}
+Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that
+
+\begin{equation} 
+  \label{eqn:trtr}
+ ({\cal S}^{\pm}_{z_s,z_r})^T = {\cal R}{\cal S}_{z_r,z_s}^{\pm}{\cal R}=
+ -{\cal S}^{\mp}_{z_r,z_s}.
+\end{equation}
+
+Define $\Pi_i, i=0,1$ to be the projection on the first,
+respectively second, component of a vector in $\bR^2$. Define the
+forward modeling operator from pressure source to pressure trace by
+\begin{equation}
+  \label{eqn:sdef}
+  S^{\pm}_{z_s,z_r} = \Pi_0 {\cal S}^{\pm}_{z_s,z_r} \Pi_0^T 
+\end{equation}
+and the forward modeling operator from velocity source (normal force)
+to velocity trace by 
+\begin{equation}
+  \label{eqn:vdef}
+  V^{\pm}_{z_s,z_r} = \Pi_1 {\cal S}^{\pm}_{z_s,z_r} \Pi_1^T 
+\end{equation}
+The relation \ref{eqn:trtr} implies that
+\begin{eqnarray}
+  (S^{\pm}_{z_s,z_r})^T &=& -S^{\mp}_{z_r,z_s} \nonumber\\
+                        &=& R S^{\pm}_{z_r,z_s}R, \nonumber\\
+    (V^{\pm}_{z_s,z_r})^T &=& -V^{\mp}_{z_r,z_s} \nonumber\\
+                        &=& R V^{\pm}_{z_r,z_s}R.
+                            \label{eqn:trtrcomp}
+\end{eqnarray}
+
+\section{Fields, Traces and Sources}
+In this section, $(p^{\pm}_0,\bv^{\pm}_0)$ denotes an acoustic field,
+continuous in the region $z_s \le z $, obeying the homogeneous wave
+equation in the interior $z_s < z $:
+\begin{eqnarray}
+\label{eqn:awe1}
+  \frac{1}{\kappa}\frac{\partial p^{\pm}_0}{\partial t} & = & - \nabla \cdot \bv^{\pm}_0, \nonumber \\
+  \rho\frac{\partial \bv^{\pm}_0}{\partial t} & = & - \nabla
+                                                    p^{\pm}_0, \nonumber \\
+  p^{\pm}_0 & =& 0,  \mbox{ for } t \ll 0, \nonumber\\ 
+  \bv^{\pm}_0 & = & 0 \mbox{ for } t \ll 0.
+\end{eqnarray}
+I will show how to view $(p^{\pm}_0,\bv^{\pm}_0)$ as the $z>z_s$ part
+of a solution of the system \ref{eqn:awepm} in two different ways,
+corresponding to two different extensions to the whole space. An
+important relation between the components of the operator
+${\cal S}^{\pm}_{z_s,z_r}$ follows (equation \ref{eqn:lamnull} below).
+
+Given a function $\phi$ on $z=z_s$, define $(p^{\pm}_1,\bv^{\pm}_1)$
+as the solution in the two half-spaces $z <z_s, z>z_s$ of
+\begin{eqnarray}
+\label{eqn:awe1}
+  \frac{1}{\kappa}\frac{\partial p_1}{\partial t} & = & - \nabla \cdot \bv_1, \nonumber \\
+  \rho\frac{\partial \bv_1}{\partial t} & = & - \nabla p_1,\nonumber \\
+  p^{\pm}_1 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
+  \bv^{\pm}_1 & = & 0 \mbox{ for } \pm t \ll 0,\nonumber\\
+  \lim_{z \rightarrow z_s} P_sp^{\pm}_1& =& \phi.
+\end{eqnarray}
+By construction, $p^{\pm}_1$ is continuous across $z=z_s$. The
+discontinuity $[v^{\pm}_{1,z}]|_{z=z_s}$ is a linear function of
+$\phi$. Define the {\em pressure-to-source map} $\Lambda^{\pm}_{z_s}$ by
+\begin{equation}
+  \label{eqn:deflam}
+  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]|_{z=z_s}
+\end{equation}
+Then as a distribution on all of $\bR^4$, $(p^{\pm}_1,\bv^{\pm}_1)$ solve the
+inhomogeneous system
+\begin{eqnarray}
+\label{eqn:sawe1}
+  \frac{1}{\kappa}\frac{\partial p^{\pm}_1}{\partial t} & = &
+                                                        (\Lambda^{\pm}_{z_s}\phi)\delta(z-z_s) - \nabla \cdot \bv^{\pm}_1, \nonumber \\
+  \rho\frac{\partial \bv^{\pm}_1}{\partial t} & = & - \nabla p^{\pm}_1,\nonumber \\
+  p^{\pm}_1 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
+  \bv^{\pm}_1 & = & 0 \mbox{ for } \pm t \ll 0.
+\end{eqnarray}
+
+This is precisely the causal option of system \ref{eqn:awepm} for
+$f_s=0$ and $h_s=\Lambda^{\pm}_{z_s}\phi$. From \ref{eqn:awe1} and the definition
+\ref{eqn:fwd},
+\begin{equation}
+  \label{eqn:lamfwd}
+  \phi = S^{\pm}_{z_s,z_s}\Lambda^{\pm}_{z_s}\phi,
+\end{equation}
+that is,
+\begin{equation}
+  \label{eqn:lamfwd1}
+  S^{\pm}_{z_s,z_s} = (\Lambda^{\pm}_{z_s})^{-1}.
+\end{equation}
+
+Similarly, suppose $\psi$ is a function on $z=z_s$, and that
+$(p^{\pm}_2,\bv^{\pm}_2)$ solves
+\begin{eqnarray}
+\label{eqn:awe2}
+  \frac{1}{\kappa}\frac{\partial p^{\pm}_2}{\partial t} & = & - \nabla \cdot \bv^{\pm}_2, \nonumber \\
+  \rho\frac{\partial \bv^{\pm}_2}{\partial t} & = & - \nabla p^{\pm}_2,\nonumber \\
+  p^{\pm}_2 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
+  \bv^{\pm}_2 & = & 0 \mbox{ for } \pm t \ll 0,\nonumber\\
+  \lim_{z \rightarrow z_s} P_sv^{\pm}_{2,z} &=& \psi.
+\end{eqnarray}
+in the two half-spaces $z<z_s, z > z_s$. In this case, $v^{\pm}_{2,z}$
+is continuous at $z=z_s$ by construction. Viewed as a
+distribution on all of $\bR^4$, $(p^{\pm}_2,\bv^{\pm}_2)$ solve the
+system \ref{eqn:awepm} with $h_s=0$, and the discontinuity in pressure
+across $z=z_s$ as the inhomogeneous term in Newton's law:
+\begin{eqnarray}
+\label{eqn:sawe2}
+  \frac{1}{\kappa}\frac{\partial p^{\pm}_2}{\partial t} & = &
+                                                        - \nabla \cdot \bv^{\pm}_2, \nonumber \\
+  \rho\frac{\partial \bv^{\pm}_2}{\partial t} & = & 
+                                                    [p^{\pm}_2]|_{z=z_s}{\bf e_z}\delta(z-z_s) - \nabla p^{\pm}_2,\nonumber \\
+  p^{\pm}_2 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
+  \bv^{\pm}_2 & = & 0 \mbox{ for } \pm t \ll 0.
+\end{eqnarray}
+This relation defines a map from the velocity boundary data $\psi$ to
+the source in the velocity equation, having units of force/volume,
+similar to $\Lambda^{\pm}_{z_s}$. As it's not needed in the sequel, I
+won't name this map.
+
+The quadratic form defined by $\Lambda^{\pm}_{z_s}$ has fundamental
+physical significance. Define the total acoustic energy $E^{\pm}(t)$ of the
+field $(p^{\pm},\bv^{\pm})$, at time $t$ by
+\begin{equation}
+  \label{eqn:defae0}
+  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)
+\end{equation}
+Then
+\begin{equation}
+  \label{eqn:elim}
+  \pm \lim_{\pm t \rightarrow \infty} E^{\pm}(t) =  \langle P_sp^{\pm},
+  (\Lambda^{\pm}_{z_s} P_sp^{\pm}) \rangle_{L^2(z=z_s)}.
+\end{equation}
+That is, the value of the quadratic form defined by
+$\Lambda^{\pm}_{z_s}$, evaluated at the pressure trace on $z=z_s$,
+gives the total energy transferred from the source to the
+acoustic field over time. Since $E$ is itself a positive definite
+quadratic form in the acoustic field, it follows that $\pm
+\Lambda^{\pm}_{z_s}$ is positive semi-definite.
+
+A closer relation between traces and sources follows from the
+geometric optics analysis of the half-space solution, presented
+below. I will show that, for propagating wave data,
+\begin{equation}
+  \label{eqn:odd1}
+  \lim_{z\rightarrow z_s^+} v_{1,z} \approx - \lim_{z\rightarrow z_s^-}
+  v_{1,z}
+\end{equation}
+Similarly,
+\begin{equation}
+  \label{eqn:odd2}
+  \lim_{z\rightarrow z_s^+} p_{2} \approx - \lim_{z\rightarrow z_s^-}
+  p_{2}
+\end{equation}
+for the velocity field solving \ref{eqn:awe1}, or equivalently
+\ref{eqn:awe1}. Here ``$\approx$'' means in the sense of high
+frequency asymptotics, that is, that the difference between the two
+sides is relatively smooth, hence small if the data is highly
+oscillatory. Similarly, for the pressure field solving \ref{eqn:awe2}
+or \ref{eqn:sawe2}, the jump at $z=z_s$ is twice the limiting value
+(trace) from above, up to a smooth error. Therefore
+\begin{eqnarray}
+  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]|_{z=z_s} &\approx& 2
+                              \lim_{z\rightarrow z_s}v^{\pm}_{1,z},\\  \label{eqn:tracejump1}
+  [p^{\pm}_2]|_{z=z_s} &\approx& 2 \lim_{z\rightarrow z_s^+}p^{\pm}_{2},\\  \label{eqn:tracejump2}
+\end{eqnarray}
+
+These apparently harmless relations have a profound consequence,
+because of a final step: identify the boundary data $\phi$ and $\psi$,
+so far arbitrary, as the traces of $p^{\pm}_0$ and $v^{\pm}_{0,z}$ on $z=z_s$,
+that is, set
+\begin{eqnarray}
+  \phi & = & P_s p^{\pm}_0, \nonumber \\
+  \psi & = & P_s v^{\pm}_{0,z}.
+             \label{eqn:bv0}
+\end{eqnarray}
+Since the traces on $z=z_s$ of the pressure fields $p^{\pm}_0$ and $p^{\pm}_1$ are
+the same, and both are pressure components of solutions of the same homogeneous
+acoustic system in $z>z_s$, the standard energy identity implies that
+$(p^{\pm}_0,\bv^{\pm}_0) = (p^{\pm}_1,\bv^{\pm}_1)$ in
+$z > z_s$. Similarly, $P_sv^{\pm}_{0,z}$ and $P_sv^{\pm}_{2,z}$ are
+the same, whence $(p^{\pm}_0,\bv^{\pm}_0) = (p^{\pm}_2,\bv^{\pm}_2)$ in
+$z>z_s$.
+
+Therefore the assumption \ref{eqn:bv0} implies that
+\begin{equation}
+  \label{eqn:jump0}
+  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]_{z=z_s} \approx 2P_s v^{\pm}_{0,z}  = 2\psi,
+\end{equation}
+and
+\begin{equation}
+  \label{eqn:jump1}
+[p_2^{\pm}]_{z=z_s} \approx  2P_s p^{\pm}_{0}= 2 \phi.           
+\end{equation}
+In the notation of the last section, set $2 \phi = 2P_sp^{\pm}_0= f_s,
+2\psi = 2P_s v^{\pm}_{0,z}= h_s$. Then relation \ref{eqn:jump0}
+implies that
+\begin{equation}
+  \label{eqn:lamfh}
+  \Lambda^{\pm}_{z_s}f_s \approx h_s.
+\end{equation}
+
+From the definitions in the last section and the system \ref{eqn:sawe1},
+\begin{equation}
+  \label{eqn:pident0}
+P_r p^{\pm}_0 = \Pi_0{\cal S}^{\pm}(h_s,0)^T = \Pi_0{\cal
+  S}^{\pm}\Pi_0^Th_s,
+\end{equation}
+and
+\begin{equation}
+  \label{eqn:pident1}
+P_r p^{\pm}_0 \approx \Pi_0{\cal S}^{\pm}_{z_s,z_r} (0,f_s)^T = \Pi_0{\cal
+  S}^{\pm}_{z_s,z_r} \Pi_1^Tf_s,
+\end{equation}
+Similarly, from the relations above and \ref{eqn:sawe2},
+\begin{equation}
+  \label{eqn:vident0}
+  P_r  v^{\pm}_{0,z}  \approx \Pi_1{\cal S}^{\pm}_{z_s,z_r}(h_s,0)^T  =
+  \Pi_1{\cal S}^{\pm}_{z_s,z_r} \Pi_0^Th_s 
+\end{equation}
+and
+\begin{equation}
+  \label{eqn:vident1}
+  P_r  v^{\pm}_{0,z}  = \Pi_1{\cal S}^{\pm}_{z_s,z_r}(0,f_s)^T  =
+  \Pi_1{\cal S}^{\pm}_{z_s,z_r} \Pi_1^Tf_s 
+\end{equation}
+Together with \ref{eqn:lamfh}, these equations imply that
+\begin{equation}
+  \label{eqn:lamnull}
+{\cal  S}_{z_s,z_r}^{\pm}P_0^T\Lambda^{\pm}_{z_s} \approx {\cal S}^{\pm}_{z_s,z_r}P_1^T.
+\end{equation}
+Consequently, ${\cal S}_{z_s,z_r}$ has an ``approximate null space'',
+in the sense that its composition with the block operator
+$(\Lambda^{\pm}_{z_s},-I)^T$ is asymptotically negligible.
+
+\section{Parametrix via Time Reversal}
+
+Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
+efficient iterative solution of the linear subproblem
+\ref{eqn:esis}. Direct computation, for instance by solving
+\ref{eqn:sawe2} and reading off $P_sv^{\pm}_{2,z}$, turns out to be
+numerically ill-behaved. An accurate computation is available via the
+computation of a {\em parametrix} (high frequency approximate inverse)
+for $S^{\pm}_{z_s,z_r}$. This parametrix also justifies the use of the
+symmetric part of $\Lambda^{\pm}_{z_s}$ as a preconditioner. In this
+section I will explain the parametrix construction, and also show how
+to extract the symmetric part of $\Lambda^{\pm}_{z_s}$.
+
+The key ingredient in this {\em time reversal} construction is a
+{\em local energy decay} assumption. Define the local energy in the
+acoustic field $(p,\bv)$ at
+time $t$ in a region $\Omega \subset \bR^3$ by imiting the integration
+in the definition of total energy \ref{eqn:defae0} to $\Omega$:
+\begin{equation}
+  \label{eqn:defaeloc}
+  E_{\Omega}[(p,\bv)](t) = \frac{1}{2} \int_{\Omega} \,d\bx \, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx) |\bv(\bx,t)|^2\right)
+\end{equation}
+
+We assume that $(p^+,\bv^+)$ is {\em downgoing} in the half-space
+$\Omega = \{(x,y,z): z \le z_r\}$: that is, there exists a solution of
+the acoustic system $(\tilde{p}^+,\tilde{\bv}^+)$ in space-time $\bR^4$
+for which
+\begin{itemize}
+\item[1. ] $p^+ \approx \tilde{p}^+, \bv^+ \approx \tilde{\bv}^+$ in
+  the slab $\Gamma = \{(x,y,z): z_s \le z \le z_r\}$, and
+\item[2. ] The energy in $(\tilde{p}^+,\tilde{\bv}^+)$ decays in
+  forward time, in the half-space $\Omega$:  $E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](t) \rightarrow 0$
+  as $t \rightarrow \infty$.
+\end{itemize}
+
+Conditions for local energy decay have been widely studied in various
+settings \cite[]{MorawetzPhillipsLax:05,Hristova:09}. For the half-space
+geometries considered here, local energy does not decay in general,
+because of the existence of ``trapped'' rays traveling parallel to the
+boundary. Instead, high frequency data associated with rays crossing
+the boundary $\{z=z_r\}$ transversally, and never returning to the slab, have
+the local energy decay property.
+
+The downgoing constraint has mostly been studied in relation to
+seismology, mostly as a ray-theoretic concept: downgoing fields are
+concentrated in phase space along rays for which depth $z$ is
+increasing with $t$. The version above is stated without reference to
+ray theory, but the setting of this paper allows its use and in a
+later section I will explicitly construct the downgoing approximation
+under appropriate ray-theoretic conditions. The numerical examples will
+illustrate the necessary assumptions. When ray-theory is appropriate,
+the downgoing constraint can be {\em ex post facto} enforced via dip
+filtering. For example, the application of the ideas developed here to
+diving wave inversion will require such a filtering step.
+
+An immediate consequence of the energy decay assumption is that in the
+slab $\Gamma$, the field $(p^+,\bv^+)$ approximates the solution of an
+anti-causal evolution equation. For any choice of $T$, define
+$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
+$\Omega$ of
+\begin{eqnarray}
+\label{eqn:revawe}
+  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
+  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
+  \tilde{p}^- & =& 0,  \mbox{ for } t=T,\nonumber\\ 
+  \tilde{\bv}^- & = & 0 \mbox{ for } t=T,\nonumber\\
+  P_r\tilde{p}^- &=& P_rp^+ \mbox { for } t \le T . 
+\end{eqnarray}
+As will be shown later, under the downgoing assumtion, $P_r p^+
+\approx P_r \tilde{p}^+$.
+Therefore $(\delta p,\delta \bv)=(\tilde{p}^+-\tilde{p}^-,\tilde{\bv}^+-\tilde{\bv}^-)$
+differs negligibly from the solution of
+\begin{eqnarray}
+\label{eqn:deltaawe}
+  \frac{1}{\kappa}\frac{\partial \delta p}{\partial t} & = & -
+                                                               \nabla
+                                                               \cdot
+                                                               \delta \bv, \nonumber \\
+  \rho\frac{\partial \delta \bv}{\partial t} & = & - \nabla \delta p,\nonumber \\
+  \delta p & =& \tilde{p}^+,  \mbox{ for } t=T,\nonumber\\ 
+  \delta \bv & = & \tilde{\bv}^+ \mbox{ for } t=T,\nonumber\\
+  P_r\delta p & = & 0,
+\end{eqnarray}
+in $\Omega \times \bR$. Since
+$E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](T) \rightarrow 0$ as $T
+\rightarrow \infty$, the usual energy estimate implies that the energy $\delta
+E$ in this field,
+\[
+  \delta E_{\Omega}(t) = \frac{1}{2} \int_{\Omega} \,d\bx \,
+ \left(\frac{(\delta p)^2}{\kappa} + \rho |\delta \bv|^2\right)
+\]
+satisfies $\delta E_{\Omega}(t) \rightarrow 0$ as $T \rightarrow
+\infty$. This decay does not generally imply that $P_s\delta p, P_s \delta \bv)
+\rightarrow 0$ (for instance) as $T \rightarrow \infty$, since the trace operator
+$P_s$ is not continuous, but it does so imply for purely downgoing
+fields, as will be demonstrated later, so
+\begin{equation}
+  \label{eqn:approxinv0}
+  \lim_{T \rightarrow \infty} P_s\tilde{v}_z^- \approx \lim_{z \rightarrow
+    0^+}\tilde{v}_z^+ \approx \lim_{z \rightarrow 0^+}v^+_z
+\end{equation}
+
+Recall (equations \ref{eqn:deflam}, \ref{eqn:jump0}) that
+\[
+  \Lambda^+_{z_s} P_s p^+ = 2 \lim_{z \rightarrow z_s^+}v^+_z 
+\]
+and that with $h_s= \Lambda^+_{z_s}P_sp^+$,
+\[
+  S^+_{z_s,z_r}h_s = P_r p^+
+\]
+Together with the last two equations, the relation
+\ref{eqn:approxinv0} implies that
+\begin{equation}
+  \label{eqn:approxinv1}
+  S^+_{z_s,z_r}(2 P_s \tilde{v}_z^-) = P_rp^+.
+\end{equation}
+As shown in the last section, the solution of the system
+\ref{eqn:revawe} differs negligibly from the anti-causal solution $(p^-_2,\bv^-_2)$
+of the system \ref{eqn:sawe2} with (i) $z_s$ replaced by $z_r$, and
+(ii) $\psi = [p^-_2]|_{z=z_r} = -2 lim_{z \rightarrow z_r^-} p^-_2
+\approx -2 P_rp^+$. Therefore (definitions \ref{eqn:sdef}, \ref{eqn:vdef})
+\begin{equation}
+  \label{eqn:approxinv2}
+  P_s \tilde{v}_z^- \approx V^-_{z_r,z_s}(-2 P_rp^+).
+\end{equation}
+Taken together, relations \ref{eqn:approxinv1} and
+\ref{eqn:approxinv2} yield the first equation in the first main result of this section:
+\begin{eqnarray}
+  \label{eqn:approxinv}
+  V^-_{z_r,z_s} S^+_{z_s,z_r} &=& \frac{1}{4}I, \nonumber\\
+  S^-_{z_r,z_s} V^+_{z_s,z_r} &=& \frac{1}{4}I.
+\end{eqnarray}.
+The second equation follows by an exactly analogous argument.
+
+\section{Derivations}
+\subsection{Extension to whole space}
+If $\phi$ is smooth and vanishes for large negative $t$, then
+$(p_1,\bv_1)$ can be constructed by pasting together the smooth
+solutions of the inhomogeneous Dirichlet problems in the half-spaces
+$\pm z > z_s$. Construct these by extending $\phi$ smoothly to
+$\pm z>z_s$, and solve systems similar to \ref{eqn:awe1} for the
+difference with $(p_1,\bv_1)$ by quoting the main results from
+\cite{BlazekStolkSymes:13}. Since the right hand side in this
+difference system is smooth, and the system is causal, the solution is
+smooth and vanishes on $z=z_s$. Adding back the extension of $\phi$
+produces smooth solutions in $\pm z > z_s$; define $(p_1,\bv)_1$ to be
+these smooth functions in $z \ne z_s$. Since the boundary values for
+pressure are the same, $p_1$ extends continuously to $z=z_s$. 
+The pressure boundary value $\phi$ uniquely determines the solutions
+in the two half-spaces, by standard energy estimates,
+
+\subsection{Adjoint Computation}
+The adjoint of ${\cal S}^+_{z_s,z_r}$ can be computed by a variant of
+the adjoint state method, in this case a by-product of the
+conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awem}
+with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
+$ (h_r,f_r\bf{e}_z) \delta(z-z_r)$. Then
+\[
+0 = 
+\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
+\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
+-
+\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
+\]
+\[
+= 
+\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
+\]
+\[
+= 
+\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
+\]
+\[
++
+\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
+\]
+\[
+= 
+\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
+ h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
+ h_r \delta(z-z_r)\right) \right.
+\]
+\[
++
+\left.  (- \nabla p^++f_s\bf{e}_z) \cdot \bv^- + \bv^+ \cdot (-\nabla
+  p^- + f_r \bf{e_z}) \right)
+\]
+\[
+= 
+\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
+ h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
+ h_r \delta(z-z_r)\right) \right.
+\]
+\[
++
+\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- 
+  +f_s \delta(z-z_s) v_z^- + v_z^+f_r \delta(z-z_r) \right)
+\]
+after integration by parts in the last two terms. Most of what is left cancels, leaving 
+\[
+0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
+( h_rP_rp^++f_rP_rv_z^+) = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
+\]
+whence
+\begin{equation}
+\label{eqn:sadj}
+ ({\cal S}^+_{z_s,z_r})^T = -{\cal S}^{-}_{z_r,z_s}.
+\end{equation}
+
+The systems \ref{eqn:awep} and \ref{eqn:awem} differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
+$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
+\begin{equation}
+  \label{eqn:trdef}
+  {\cal R} \left(
+    \begin{array}{c}
+      p\\
+      \bv
+    \end{array}
+  \right) =
+  \left(
+    \begin{array}{c}
+      Rp\\
+      -R\bv
+    \end{array}
+  \right)
+\end{equation}
 Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awep} with
@@
 Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that
-\begin{equation} 
-  \label{eqn:trtr}
- ({\cal S}^+_{z_s,z_r})^T = {\cal R}{\cal S}_{z_r,z_s}^+{\cal R}=
- -{\cal S}^-_{z_r,z_s}.
-\end{equation}
 
@@
 
-\section{Parametrix via Time Reversal}
-
-Define $H$ to be the subspace $\{u \in (L^2(\bR^3))^4: \mbox{supp
-}u\subset \{z<z_s\}\}$, by analogy with \cite{Lax:PDENotes},
-p. 122. Note: must add microlocal $H^1$ constraint per
-\cite[]{BaoSy:91b} so that trace operator is well-defined - postpone
-this.
-
-Given $u=(\bar{p}_0,\bar{\bv}_0) \in H$, extend in $t$ by solving the acoustic system
+%######################################
+\section{Propagating Waves}
+Define $H$ to be the subspace $\{\bu \in (L^2(\bR^3))^4: \bu=0,z\ge
+z_s\}$, by analogy with \cite{Lax:PDENotes}, p. 122.
+%Note: must add microlocal $H^1$ constraint per
+%\cite[]{BaoSy:91b} so that trace operator is well-defined - postpone
+%this.
+
+Given $u=(\bar{p}_0,\bar{\bv}_0) \in H$, extend in $t$ by solving the
+acoustic initial value problem
 \begin{eqnarray}
@@
 \end{eqnarray}
-Note that $(p_0,\bv_0)=0$ near $t=0$ in the region $z>z_s$.
-
-Denote by $P_s$, respectively $P_r$, the trace operator on $z=z_s$,
-respectively $z=z_r$. Given $(p_0,\bv_0)$ as above, define
-$(p_1,\bv_1)$ as the solution in $z \ne z_s$ of
-\begin{eqnarray}
-\label{eqn:awe1}
-  \frac{1}{\kappa}\frac{\partial p_1}{\partial t} & = & - \nabla \cdot \bv_1, \nonumber \\
-  \rho\frac{\partial \bv_1}{\partial t} & = & - \nabla p_1,\nonumber \\
-  p_1 & =& 0,  \mbox{ for } 0,\nonumber\\ 
-  \bv_1 & = & 0 \mbox{ for } 0,\nonumber\\
-  P_sp_1 = P_sp_0. 
-\end{eqnarray}
-Extended to form a distribution on all of $\bR^4$, $(p_1,\bv_1)$ solve the
-inhomogeneous system
-\begin{eqnarray}
-\label{eqn:awe1}
-  \frac{1}{\kappa}\frac{\partial p_1}{\partial t} & = &
-                                                        ((\Lambda^+_{z_s})^{-1}P_sp_0)\delta(z-z_s) - \nabla \cdot \bv_1, \nonumber \\
-  \rho\frac{\partial \bv_1}{\partial t} & = & - \nabla p_1,\nonumber \\
-  p_1 & =& 0,  \mbox{ for } t=0,\nonumber\\ 
-  \bv_1 & = & 0 \mbox{ for } t=0.\nonumber\\
-\end{eqnarray}
-where
-\begin{equation}
-  \label{eqn:defhs}
-  (\Lambda^+_{z_s})^{-1}P_sp_0 = P_s[(\bv_1)_z].
-\end{equation}
+Standard results on hyperbolic systems assure existence of a unique
+square-integrable (weak) solution, which is smooth if $(p_0,\bv_0)$ is
+smooth at $t=0$ \cite[]{CourHil:62,LaxPDENotes}.
+
+Note that $(p_0,\bv_0)=0$ near $t=0$ in the half-space $z>z_s$.
+
+For any suitable space-time function $u$, define
+\begin{equation}
+  \label{eqn:postrace}
+  P^+_{z_s}u(x,y,t) =
+  \left\{
+    \begin{array}{c}
+      u(x,y,z_s,t), t>0,\\
+      0, t\le 0.
+    \end{array}
+  \right\}
+\end{equation}
+(Here, ``suitable'' means: the trace on $z=z_s$ makes sense. More on
+this later.)
 
@@
 uniqueness theorem shows that $(p_1,\bv_1) = (p_0, \bv_0)$ in $z>z_s$.
-
-Introduce the operator $T^+_{z_s,z_r}$ mapping pressure on $z=z_s$ to
-pressure on $z=z_r$: that is, $T^{+}_{z_s,z_r}P_sp_0 = P_rp_0$. From
-the previous observation, $T^{+}_{z_s,z_r}P_sp_0 = P_rp_1$, that is,
-you can compute $T^{+}$ by solving the system defining $p_1$, with
-data $P_sp_0$. From the definition of $S^{+}_{z_s,z_r}$, the definition
-\ref{eqn:defhs}, and the boundary condition for $p_1$,
-\begin{equation}
-  \label{eqn:relnST}
-  T^+_{z_s,z_r} = S^+_{z_s,z_r} (\Lambda^+_{z_s})^{-1}.
-\end{equation}
-
-Let $\epsilon>0$ satisfy
-\begin{equation}
-  \label{eqn:defeps}
-  E[{\bf 1}_{\bR^2 \times (-\infty,z_r]}(p_0,\bv_0)](T) \le \epsilon E[(p_0,\bv_0)]
-\end{equation}
-Omit the time on the RHS because the energy of the full field is
-time-invariant. Inequality \ref{eqn:defeps} states that at time
-$t=T$, the energy in
-the region $z<z_r$ is at most $\epsilon$ times the energy in all of
-space. For small $\epsilon$, in other words, most of the
-energy has left $z<z_r$, at least temporarily, at $t=T$. This
-assertion is similar to the outgoing field condition in Lax-Phillips
-scattering theory, but is time-local and approximate.
-
-Define $p_2$ to be the solution in $z \le z_r$ of
-\begin{eqnarray}
-\label{eqn:awe2}
-  \frac{1}{\kappa}\frac{\partial p_2}{\partial t} & = & - \nabla \cdot \bv_2, \nonumber \\
-  \rho\frac{\partial \bv_2}{\partial t} & = & - \nabla p_2,\nonumber \\
-  p_2 & =& 0,  \mbox{ for } t=T,\nonumber\\ 
-  \bv_2 & = & 0 \mbox{ for } t=T,\nonumber\\
-  P_rp_2 = P_rp_0. 
-\end{eqnarray}
-Then $(\delta p_2,\delta \bv_2)=(p_0-p_2,\bv_0-\bv_2)$ solves
-\begin{eqnarray}
-\label{eqn:awe2}
-  \frac{1}{\kappa}\frac{\partial \delta p_2}{\partial t} & = & -
-                                                               \nabla
-                                                               \cdot
-                                                               \delta \bv_2, \nonumber \\
-  \rho\frac{\partial \delta \bv_2}{\partial t} & = & - \nabla \delta p_2,\nonumber \\
-  \delta p_2 & =& p_0,  \mbox{ for } t=T,\nonumber\\ 
-  \delta \bv_2 & = & \bv_0 \mbox{ for } t=T,\nonumber\\
-  P_r\delta p_2 = 0. 
-\end{eqnarray}
-Using the standard energy estimate for homogenous Dirichlet
-problem,
-\begin{equation}
-  \label{eqn:errtr}
-  E[{\bf 1}_{\bR^2 \times (-\infty,z_r]}(\delta p_2,\delta \bv_2)](t) \le
-  \epsilon E[(p_0,\bv_0)], \, 0 \le t \le T.
-\end{equation}
-
-Define $T^{-}P_rp_0 = P_sp_2$. Then
-\[
-  T_{z_r,z_s}^{-}P_rp_0-P_sp_0 = P_s\delta p_2.
-\]
-For the time being, write $O(\epsilon)$ for errors in the trace at
-$z_s$ (output of $P_s$) for solutions of the forward Dirichlet problem
-for initial data satisfying the estimate \ref{eqn:errtr}. Of course
-this does not imply any $L^2$ bound on the trace - for that,
-additional constraints on the initial data are required, to be
-discussed below. With this convention,
-\[
-  T_{z_r,z_s}^{-}P_rp_0 =
-  S^{-}_{z_r,z_s}(\Lambda^{-}_{z_r})^{-1}P_rp_0
-\]
-\[
-  =  S^{-}_{z_r,z_s}(\Lambda^{-}_{z_r})^{-1}T_{z_s,z_r}^+P_sp_0
-\]
-\[
-=
-S^{-}_{z_r,z_s}(\Lambda^{-}_{z_r})^{-1}S_{z_s,z_r}^+(\Lambda^{+}_{z_s})^{-1}P_sp_0
-\]
-\[
-  = P_s p_0 + O(\epsilon)
-\]
-Setting $h_s= +(\Lambda^{+}_{z_s})^{-1}P_sp_0$ and applying
-$+(\Lambda^{+}_{z_s})^{-1}$ to both sides, obtain
-\[
-  h_s =
-  (\Lambda^{+}_{z_s})^{-1}S^{-}_{z_r,z_s}(\Lambda^{-}_{z_r})^{-1}S_{z_s,z_r}^+h_s
-  + O(\epsilon)
-\]
-\[
-  =
- -(\Lambda^{+}_{z_s})^{-1}(S^{+}_{z_r,z_s})^T(\Lambda^{-}_{z_r})^{-1}S_{z_s,z_r}^+h_s
-  +O(\epsilon)
-\]
-\begin{equation}
-  \label{eqn:approxunit}
- =(\Lambda^{+}_{z_s})^{-1}(S^{+}_{z_r,z_s})^T((\Lambda^{+}_{z_r})^{-1})^TS_{z_s,z_r}^+h_s
-  +O(\epsilon)
-\end{equation}
-The last two steps are justified by appeal to identities
-\ref{eqn:trtr} and \ref{eqn:ntodtr}.
-
-Define
-\begin{equation}
-  \label{eqn:appinv}
-  (S^+_{z_s,z_r})^{\dagger} =
-  (\Lambda^{+}_{z_s})^{-1}(S^{+}_{z_r,z_s})^T((\Lambda^{+}_{z_r})^{-1})^T.
-\end{equation}
-Then equation \ref{eqn:approxunit} asserts that
-$(S^+_{z_s,z_r})^{\dagger}$ is an approximate inverse to
-$S^+_{z_s,z_r}$, in the sense that the error is the trace of a
-ssolution with small Cauchy condition at $t=0$. Later will specify
-conditions under which the output error is small in $L^2$.
 
@@
 \end{equation}
-This is the approximate inverse computed in the examples.
-Note that only the values of $\kappa$ near the source datum $z=z_s$ play a role in the relation \ref{eqn:sreln} or in the definition \ref{eqn:asginv} of the IWAVE ASG approximate inverse.
+This is the approximate inverse computed in the examples. Note that
+only the values of $\kappa$ near the source datum $z=z_s$ play a role
+in the relation \ref{eqn:sreln} or in the definition \ref{eqn:asginv}
+of the IWAVE ASG approximate inverse.
+
+\section{Discussion}
+
+The source subproblem is a necessary step in solving the overall FWI problem \ref{eqn:fwi}.
+In particular, the nonlinear least squares realization \ref{eqn:esi} of source-extended FWI is naturally approached via the {\em variable projection method} \cite[]{GolubPereyra:73,GolubPereyra:03,vanLeeuwenMulder:09,Rickett:SEG12}, in which the minimum over $\bff$ in \ref{eqn:esis} is treated as a function of $\bf{c}$, which is in turn minimized.
+Variable projection thus treats the source subproblem as an inner problem, which must be solved for an estimate of $\bff$ at every step of an iterative method for the outer problem of minimization over $\bf{c}$.
+Efficiency in solving the source subproblem \ref{eqn:esis} is critical to efficient implementation of variable projection for solving the regularized FWI problem \ref{eqn:esi}.
+
+This paper restricts the choice of wave physics embodied in
+$L[\bf{c}]$ to linear acoustics.  Thus $\bf{c}=(\kappa,\rho)$ is the
+pair (bulk modulus, density) of positive functions of spatial
+position.
+The principal characteristics assumed of sources $\bff=$ are localization and isotropy.
+That is, physical sources will be assumed to be isotropic point radiators, described by the functional form
+\begin{equation}
+\label{eqn:ptsrc}
+f_{\rm pt}(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s). 
+\end{equation}
+Source locations $\bx_s$ are assumed known (as is natural for active source methods), and the source {\em wavelet} $w(t)$ is to be found as part of the solution of \ref{eqn:esi}.
+A convenient choice of penalty operator $A$ is then multiplication by the {\em offset} $|\bx-\bx_s|$: its null space consists precisely of distributions of the form \ref{eqn:ptsrc}.
+
+More complex radiation characteristics may be accommodated with similar but more complex choice for the operator $A$.
+Since the core analysis is similar, this paper deals only with the simplest case \ref{eqn:ptsrc}.
+
+The wavefield $u$ in this work is presumed to exist in all of Euclicean space-time: in other words, boundary effects have been removed from $u$, by surface-related multiple elimination or other techniques.
+Predicted data traces are simply the samples of one or more components of $u$ at receiver points $\bx_r$.
+
+%The domain of wave propagation is denoted $\Omega$, and two subsets of its boundary are identified: $\Gamma_s$ contains the support (non-zeroes) of the (extended) sources and in particular the locations $\bx_s$ of physical sources.
+Data trace locations $\bx_r$ lie in another subset $\Gamma_r$.
+The data projection operator $P$ samples the acoustic field $u$ on $\Gamma_r$, and includes a finite aperture mute.
+
+Achieving a small residual in the source subproblem \ref{eqn:esis} for arbitrary $c$ is a critical part of the convergence mechanism for the extended source approach to FWI \cite[]{HuangNammourSymesDollizal:SEG19}.
+However, for arbitrary data $d$, coefficient array $c$, and source location $\bx_s$, there does not exist an isotropic point source $f_{\rm pt}$ of the form \ref{eqn:ptsrc} for which $PL[c]^{-1}f_{\rm pt} \approx d$.
+This is so even if there does exist a coefficient array $c^*$ and point source $f_{\rm pt}^*$ located at $\bx_s$ with wavelet $w^*(t)$ for which $d = PL[c^*]^{-1}f_{\rm pt}^*$, if $c$ differs substantially from $c^*$.
+It is this fact that underlies the effectiveness of the extended source approach to FWI:  driving the residual in the problem \ref{eqn:esi} towards zero necessarily requires modifying $c$ to resemble $c^*$, assuming the latter exists, as well as pushing the optimal $f$ towards the null space of $A$.
+Conversely, the source subproblem \ref{eqn:esis} does not have a small residual solution of point source form \ref{eqn:ptsrc}, in general, for a given choice of $c$, even if that were the case for a different choice ($c^*$).
+Therefore a more general class of source models than that specified by the condition \ref{eqn:ptsrc} must be admitted to the feasible set for the source subproblem \ref{eqn:esis} if it is to serve as the inner problem in a variable projection formulation of extended source FWI.
+
+The class of non-point sources investigated here are those confined a hypersurface assumed to contain the locations of physical (point) sources, but not {\em a priori} required to have point support.
+These {\em surface extended sources} are able to yield a relatively small residual in the source subproblem \ref{eqn:esis} for more or less arbitrary coefficient array $c$.
+This observation is a by-product of the approximate solution for \ref{eqn:esis} constructed below.
+The construction of the approximate solution is closely related to the {\em time reversal} method of photoacoustic tomography (see \cite{StefanovUhlmannIP:09} and references cited there).
+
+The next section gives a precise description of an acoustic version of problem \ref{eqn:esis} for an idealized {\em crosswell} configuration, in which source located on a surface $z=z_s$ generate waves that propagate to to a receiver surface $z=z_r > z_s$.
+In principle, {\em diving waves} can be described with similar mathematics, but with some additional complications, so I do not treat the diving wave case explicitly in this paper.
+The assumptions detailed here imply that the data $d$ of the inverse problems \ref{eqn:esi}, \ref{eqn:esis} have the physical character of {\em transmitted waves}.
+There are no mathematical results at present concerning reflected wave inverse problems treated by source extension methods, though there are tantalizing numerical clues \cite[]{LeeuwenHerrmannWRI:13,Warner:14,Warner:16,LeeuwenHerrmann:16,HuangSymes:Geo17,HuangSymes:Geo18a,HuangSymes:Geo18b}.
+
+The following section describes the construction of the approximate
+inverse and explains the conditions under which accuracy should be expected.
+I observe that the approximate inverse is approximately the adjoint of the modeling operator $PL[c]^{-1}$ with respect to weighted norms in its domain and range (see \cite{HouSymes:15} for a similar observation in a different context).
+This fact immediately suggests use of the approximate inverse as a preconditioner to accelerate convergence of Krylov space methods such as conjugate gradient iteration applied to the least squares problem \ref{eqn:esis}.
+However, its straightforward application in the preconditioned conjugate gradient (PCG) method \cite[]{Golub:2012} involves explicitly a version the so called Dirichlet-to-Neumann (D2N) operator, which is computationally awkward.
+I show how to reorganize the PCG iteration so that only solutions of $L[c]u=f$ are required.
+The penultimate section presents some 2D examples in the crosswell configuration, illustrating the accuracy of the approximate inversion and the accelerated convergence of PCG.
+The examples are chosen to emphasize that surface extended sources can be constructed to fit more or less arbitrary data, unlike point sources - as mentioned before, this approximate invertibility property of the modeling operator is of critical importance in the application to extended FWI.
+The paper ends with a discussion of several important matters not addressed here, and a restatement of the conclusions.
+
+The intent of the main body of this paper is to present the formal structure of the extended source subproblem and the way in which this structure leads to accelerated numerical solution.
+This formal structure is however supported by a rigorous foundation.
+In order to avoid disrupting the formal account, this foundation material is relegated to an appendix.
+
 \bibliographystyle{seg}
 \bibliography{../../bib/masterref}
-
-\bibliographystyle{seg}
-\bibliography{../../bib/masterref}

### 2021-05-05 repro/papers/sseprecond/paper.tex
interim

@@
 \end{equation}
-Then as a distribution on all of $\bR^4$, $(p^{\pm}_1,\bv^{\pm}_1)$ solve the
+Then as a distribution on all of $\bR^4$, $(p^{\pm}_1,\bv^{\pm}_1)$ solves the
 inhomogeneous system
@@
   \label{eqn:lamnull}
-{\cal  S}_{z_s,z_r}^{\pm}P_0^T\Lambda^{\pm}_{z_s} \approx {\cal S}^{\pm}_{z_s,z_r}P_1^T.
+{\cal  S}_{z_s,z_r}^{\pm}\Pi_0^T\Lambda^{\pm}_{z_s} \approx {\cal S}^{\pm}_{z_s,z_r}\Pi_1^T.
 \end{equation}
@@
   \label{eqn:approxinv}
-  V^-_{z_r,z_s} S^+_{z_s,z_r} &=& \frac{1}{4}I, \nonumber\\
-  S^-_{z_r,z_s} V^+_{z_s,z_r} &=& \frac{1}{4}I.
+  V^-_{z_r,z_s} S^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
+  S^-_{z_r,z_s} V^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
+  V^+_{z_s,z_r} S^-_{z_r,z_s} & \approx & \frac{1}{4}I, \nonumber\\
+  S^+_{z_s,z_r} V^-_{z_r,z_s} & \approx & \frac{1}{4}I.\\
 \end{eqnarray}.
-The second equation follows by an exactly analogous argument.
+The second equation follows by an exactly analogous argument, and the
+last two by time reversal and interchange of the roles of $z_s$ and
+$z_r$.
+
+Having identified approximate two-sided inverses for $S^{\pm},
+V^{\pm}$, the approximate null space identity \ref{eqn:lamnull} leads
+to a relation between $\Lambda^{\pm}$ and ${\cal S}^{\pm}$ that
+provides the foundation for practical computations. Applying $\Pi_0$
+to both sides of \ref{eqn:lamnull} gives
+\[
+ \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_0^T
+ \Lambda^+_{z_s} = S^+{z_s,z_r} \Lambda^+_{z_s}.
+\]
+From the first approximate identity in \ref{eqn:approxinv},
+\begin{equation}
+  \label{eqn:lamident}
+  V^{-}_{z_r,z_s}\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \frac{1}{4}\Lambda^+_{z_s}.
+\end{equation}
+This identity is the second major result of this section: it shows how
+to compute that action of $\Lambda^+_{z_s}$ by propagating the input
+pressure trace, identified as a source for the velocity evolution,
+forward in time from $z_s$ to $z_r$
+reading off the pressure trace on $z=z_r$, identifying it once more as
+a source for velocity, propagating it backwards in time from $z_r$ to
+$z_s$, and finally reading off the velocity trace, interpreted as a
+pressure evolution source on $z_s$. I will give a detailed description
+of this computation in the section on numerical results.
+
+The importance of this result lies in the failure of the obvious
+method for computing the action of $\Lambda^{\pm}_{z_s}$, namely to
+employ the pressure trace as a source in the velocity equation ($f_s$,
+in the notation used above) at $z=z_s$, and read off the velocity
+field also at $z=z_s$. This difficulty is related to the existence of
+tangentially propagating waves and the lack of continuity of the trace
+operator. The method implicit in equation \ref{eqn:lamident} avoids
+this difficulty by propagating the fields a positive distance in $z$:
+assuming as always that the causal fields are downgoing, this step
+eliminates any tangentially propagating fields from consideration.
+
+Computation of the transpose of $\Lambda^+$ (exact, not approximate in
+the high frequency sense) will be important in the
+construction of a preconditioner. The relation \ref{eqn:lamident} also
+provides an computation for this operator. The idea is to treat the
+left-hand side of equation \ref{eqn:lamident} as an approximation to
+$\Lambda$, and compute its exact transposes via the adjoint state
+method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
+\begin{equation}
+  \label{eqn:lamtransp}
+   \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r} \approx \frac{1}{4}(\Lambda^+_{z_s})^T.
+\end{equation} 
+The left-hand sides of \ref{eqn:lamident} and \ref{eqn:lamtransp} will
+simply be substituted for $\Lambda$ and its transpose in the
+preconditioner construction: preconditioners must be (precisely)
+self-adjoint, but need only be approximated to perform well.
 

### 2021-05-14 repro/papers/sseprecond/paper.tex
interim

@@
 For acoustic wave physics, the coefficient vector is
-$\bf{c}=(\kappa,\rho)$, with components bulk modulus $\kappa$ and
+$\bf{c}=(\kappa,\rho)^T$, with components bulk modulus $\kappa$ and
 density $\rho$, and the state vector $\bu=(p,\bv)^T$ consists of
@@
 Since all of the operators in the discussion that follows depend on
+the coefficient vector 
 $\bf{c}$, I will suppress it from the notation, for example, $L=L[\bf{c}]$.
 
-The extended ``crosswell'' configuration involves sources confined to
-a spatial hypersurface. For convenience, choose this to be the hyperplane
-$\{(x,y,z,t): z=z_s\}$. Such sources are combinations of constitutive law defects and
-loads normal to the surface, that is, right-hand sides in the system
-$L\bu=\bff$ of the form $\bff(\bx,t) =
-(h_s(x,y,t)\delta(z-z_s), f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar
-defect $h_s$ and normal force $f_s$  ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in \ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$ takes the form
+The surface source extension replaces point sources on or near a
+surface in $\bR^3$ with source functions confined to the same
+surface. The simplest example of this extended geometry specifies a
+plane $\{(x,y,z,t): z=z_s\}$ at source depth $z_s$ as the surface. For
+acoustic modeling, surface sources are combinations of constitutive law
+defects and loads normal to the surface, localized on $z=z_s$. That
+is, right-hand sides in the system $L\bu=\bff$ take the form
+$\bff(\bx,t) = (h_s(x,y,t)\delta(z-z_s),
+f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar defect $h_s$ and normal
+force $f_s$ ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in
+\ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$
+takes the form
 \begin{eqnarray}
@@
 \]
-to mean
-\[
-  \mbox{For some } T \in \bR, p^+(\cdot,t) = 0 \mbox{ for all } t \ge
+to mean that $p^+$ is {\em causal}, that is,
+\[
+  \mbox{For some } T \in \bR, p^+(\cdot,t) = 0 \mbox{ for all } t <
   T.
 \]
+Similarly,
+\[
+  p^- = 0 \mbox{ for } t \gg 0 
+\]
+signifies that $p^-$ is anti-causal.
+
 Extended forward modeling consists in solving \ref{eqn:awepm} and
-sampling the solution components on another spatial hyperplane $\{(x,y,z,t): z=z_r\}$. Denote by $P_s,P_r$ the trace (sampling) operators on $z=z_s$, $z=z_r$
-respectively. The causal/anti-causal modeling operators
-${\cal S}^{\pm}_{z_s,z_r}$ are defined in
-terms of the solutions $(p^{\pm},\bv^{\pm})$ of the systems
-\ref{eqn:awepm} by
-\begin{equation}
-  {\cal S}^{\pm}_{z_s,z_r}(h_s,f_s)  = (P_rp^{\pm},P_r v_z^{\pm}),
+sampling the solution components at receiver locations. For
+simplicity, throughout this paper I will assume that the receivers are
+located on another spatial hyperplane $\{(x,y,z,t): z=z_r\}$ at
+receiver depth $z_r>z_s$.
+
+Denote by $P_s,P_r$ the trace (sampling)
+operators on $z=z_s$, $z=z_r$ respectively.
+Under the standing assumptions to be made on
+the solutions of system \ref{eqn:awepm}, the key components of the
+solution ($p^{\pm}$ and $v^{\pm}_z$) are continuous functions of $z$
+in the open slab $z_s<z<z_r$, but may be discontinuous at the source
+plane $z=z_s$. Similarly, the roles of $z_s$ and $z_r$ will be
+interchanged in some of the constructions to come, and the
+corresponding solutions may be discontinuous at $z=z_r$. Accordingly,
+interpret $P_s$, $P_r$ as the limit from right and left respectively:
+for $u=p^{\pm}$ or $v^{\pm}_z$,
+\begin{eqnarray}
+  \label{eqn:defsamp}
+  P_su(x,y,t) &=& \lim_{z \rightarrow z_s^+} u(x,y,z,t),\nonumber \\
+  P_ru(x,y,t) &=& \lim_{z \rightarrow z_r^-} u(x,y,z,t).                  
+\end{eqnarray}
+
+The causal/anti-causal vector
+modeling operators ${\cal S}^{\pm}_{z_s,z_r}$ are defined in terms of
+the solutions $(p^{\pm},\bv^{\pm})$ of the systems \ref{eqn:awepm} by
+\begin{equation}
+  {\cal S}^{\pm}_{z_s,z_r}(h_s,f_s)^T  = (P_rp^{\pm},P_r v_z^{\pm})^T,
   \label{eqn:fwd}
@@
 several locations will be needed in the discussion below.
+
+As will be shown later, ${\cal S}^{\pm}$ is not stably invertible, in fact
+having rank 1 (in an approximate sense). The diagonal components of
+${\cal S}^{\pm}$ thus carry essentially all of its information.
+Denote by $\Pi_i, i=0,1$ the projection on the first,
+respectively second, component of a vector in $\bR^2$. The 
+forward modeling operator from pressure source to pressure trace is
+\begin{equation}
+  \label{eqn:sdef}
+  S^{\pm}_{z_s,z_r} = \Pi_0 {\cal S}^{\pm}_{z_s,z_r} \Pi_0^T 
+\end{equation}
+and the forward modeling operator from velocity source (normal force)
+to velocity trace is
+\begin{equation}
+  \label{eqn:vdef}
+  V^{\pm}_{z_s,z_r} = \Pi_1 {\cal S}^{\pm}_{z_s,z_r} \Pi_1^T 
+\end{equation}
 
@@
 
-Define $\Pi_i, i=0,1$ to be the projection on the first,
-respectively second, component of a vector in $\bR^2$. Define the
-forward modeling operator from pressure source to pressure trace by
-\begin{equation}
-  \label{eqn:sdef}
-  S^{\pm}_{z_s,z_r} = \Pi_0 {\cal S}^{\pm}_{z_s,z_r} \Pi_0^T 
-\end{equation}
-and the forward modeling operator from velocity source (normal force)
-to velocity trace by 
-\begin{equation}
-  \label{eqn:vdef}
-  V^{\pm}_{z_s,z_r} = \Pi_1 {\cal S}^{\pm}_{z_s,z_r} \Pi_1^T 
-\end{equation}
 The relation \ref{eqn:trtr} implies that
@@
 \[
-  \Lambda^+_{z_s} P_s p^+ = 2 \lim_{z \rightarrow z_s^+}v^+_z 
+  \Lambda^+_{z_s} P_s p^+ \approx 2 \lim_{z \rightarrow z_s^+}v^+_z 
 \]
@@
   \label{eqn:approxinv1}
-  S^+_{z_s,z_r}(2 P_s \tilde{v}_z^-) = P_rp^+.
+  S^+_{z_s,z_r}(2 P_s \tilde{v}_z^-) \approx P_rp^+.
 \end{equation}
@@
 of the system \ref{eqn:sawe2} with (i) $z_s$ replaced by $z_r$, and
-(ii) $\psi = [p^-_2]|_{z=z_r} = -2 lim_{z \rightarrow z_r^-} p^-_2
+(ii) $\psi = [p^-_2]|_{z=z_r} = -2 \lim_{z \rightarrow z_r^-} p^-_2
 \approx -2 P_rp^+$. Therefore (definitions \ref{eqn:sdef}, \ref{eqn:vdef})
@@
   \label{eqn:approxinv}
-  V^-_{z_r,z_s} S^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
-  S^-_{z_r,z_s} V^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
-  V^+_{z_s,z_r} S^-_{z_r,z_s} & \approx & \frac{1}{4}I, \nonumber\\
-  S^+_{z_s,z_r} V^-_{z_r,z_s} & \approx & \frac{1}{4}I.\\
+  V^-_{z_r,z_s} S^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
+  S^-_{z_r,z_s} V^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
+  V^+_{z_s,z_r} S^-_{z_r,z_s} & \approx & -\frac{1}{4}I, \nonumber\\
+  S^+_{z_s,z_r} V^-_{z_r,z_s} & \approx & -\frac{1}{4}I.\\
 \end{eqnarray}.
@@
  \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_0^T
- \Lambda^+_{z_s} = S^+{z_s,z_r} \Lambda^+_{z_s}.
+ \Lambda^+_{z_s} = S^+_{z_s,z_r} \Lambda^+_{z_s}.
 \]
@@
   \label{eqn:lamident}
-  V^{-}_{z_r,z_s}\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \frac{1}{4}\Lambda^+_{z_s}.
+  V^{-}_{z_r,z_s}\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx -\frac{1}{4}\Lambda^+_{z_s}.
 \end{equation}
@@
   \label{eqn:lamtransp}
-   \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r} \approx \frac{1}{4}(\Lambda^+_{z_s})^T.
+   \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r} \approx -\frac{1}{4}(\Lambda^+_{z_s})^T.
 \end{equation} 
@@
 self-adjoint, but need only be approximated to perform well.
+
+\section{Approximate Unitarity}
+Recall that the fields $(p^{\pm},\bv^{\pm})$ in $z>z_s$ can be viewed
+as either the solution of the system \ref{eqn:awepm} with pressure source $h_s=0$ and
+a suitable choice of velocity source $f_s$, or of the same system with
+suitable choice of $h_s$ and $f_s=0$, and that these choices are
+related by equation \ref{eqn:lamfh}: $\Lambda^{\pm}f_s = h_s$. The
+pressure traces at $z=z_r>z_s$ are the same: from the definitions
+\ref{eqn:sdef}, \ref{eqn:vdef},
+\begin{eqnarray}
+  \label{eqn:sandv}
+  S^{\pm}_{z_s,z_r}h_s& =& P_rp^{\pm},\nonumber \\
+  V^{\pm}_{z_s,z_r}f_s &=& P_rv_z^{\pm}.
+\end{eqnarray}
+Applying the same reasoning to traces on $z=z_r$: create pressure and
+velocity sources $h_r$ and $f_r$ at $z=z_r$ by odd extension of the
+velocity, respectively pressure, fields, and using the jumps as source
+coefficients. Then as before
+$\Lambda^{\pm}_{z_r}f_r=h_r=-P_rv^{\pm}_z$ (the sign change is due to
+$Lambda$ producing the jump, whereas $P_rv^{\pm}_z$ is the limit from
+the left of the field $v^{\pm}_{1,z}$ of system \ref{eqn:awe1}), hence
+$\Lambda^{\pm}_{z_r}P_rp^{\pm} = P_rv_z^{\pm}$. Combining this
+observation with equations
+\ref{eqn:lamfh}, \ref{eqn:sandv}, obtain (with roles of $z_s,z_r$ interchanged)
+\[
+V^{\pm}_{z_r,z_w}(-2P_rp^{\pm}) =
+  \Lambda^{\pm}_{z_s}S^{\pm}_{z_r,z_s}(-2P_rv_z^{\pm})  =
+  \Lambda^{\pm}_{z_s}S^{\pm}_{z_r,z_s}(-2\Lambda^{\pm}_{z_r}P_rp^{\pm})
+\]
+whence
+\begin{equation}
+  \label{eqn:preunit}
+ V^{\pm}_{z_r,z_w}
+ =\Lambda^{\pm}_{z_s}S^{\pm}_{z_r,z_s}\Lambda^{\pm}_{z_r}.
+\end{equation}
+
+In view of the relations \ref{eqn:trtrcomp}, equation
+\ref{eqn:preunit} implies that
+\[
+  V^-_{z_r,z_s}=-(V^+_{z_s,z_r})^T = -\Lambda^{-}_{z_s}(S^+_{z_s,z_r})^T\Lambda^{-}_{z_r}.
+\]
+\[
+  =-(\Lambda^{+}_{z_s})^T(S^+_{z_s,z_r})^T(\Lambda^{+}_{z_r})^T.
+\]
+Therefore the first equation in display \ref{eqn:approxinv} can be re-written
+\[
+  -4V^-_{z_r,z_s}S^+_{z_s,z_r} =
+\]
+\begin{equation}
+  \label{eqn:baseunit}
+ 4 (\Lambda^{+}_{z_s})^T(S^+_{z_s,z_r})^T(\Lambda^{+}_{z_r})^T
+  S^+_{z_s,z_r} \approx I
+\end{equation}
 

### 2021-05-18 repro/papers/sseprecond/paper.tex
interim

@@
 \begin{enumerate}
-\item a linear wave operator $L[\bf{c}]$, depending on an array of space-dependent coefficients $\bf{c}$ and acting on wavefields $\bu$ vanishing in negative time:
+\item a linear wave operator $L[{\bf c}]$, depending on a vector of
+  space-dependent coefficients ${\bf c}$ and acting on causal vector wavefields $\bu$ vanishing in negative time:
 \begin{equation}
@@
 \item a trace sampling operator $P$ acting on wavefields and producing data traces;
-\item and a source function (of space and time) $\bff$ representing energy input to the system. 
+\item and a (vector) source function (of space and time) $\bff$ representing energy input to the system. 
 \end{enumerate}
-The basic FWI problem is: given data $d$, find $\bf{c}$ so that 
+The basic FWI problem is: given data $d$, find ${\bf c}$ so that 
 \begin{equation}
@@
 \end{equation}
+In this formulation, the source function $\bff$ may be given, or
+to be determined subject to some constraints.
 
@@
 \label{eqn:ols}
-\mbox{choose } \bf{c} \mbox{ to minimize } \|PL[\bf{c}]^{-1}\bff -d \|^2.
+\mbox{choose } {\bf c} \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2.
 \end{equation}
@@
 approach given the dimensions of a typical instance of \ref{eqn:fwi},
-and those have a tendency to stall due to ``cycle-skipping''. Source
+and those have a tendency to stall due to ``cycle-skipping''. See for
+exampe \cite{VirieuxOperto:09} and many references cited there. Source
 extension is one approach to avoiding this problem. It consists in
@@
 \label{eqn:esi}
-\mbox{choose } \bf{c}, \bff \mbox{ to minimize } \|PL[\bf{c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
+\mbox{choose } {\bf c}, \bff \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
 \end{equation}
@@
 
-This paper presents a numerically efficient approach to solving the {\em source subproblem}
+This paper presents a numerically efficient approach to solving the
+{\em source subproblem} of problem \ref{eqn:esi}:
 \begin{equation}
 \label{eqn:esis}
-\mbox{given } \bf{c}, \mbox{ choose } \bff \mbox{ to minimize } \|PL[\bf{c}]^{-1}\bff -d \|^2 + \alpha \|A\bff\|^2 
+\mbox{given } {\bf c}, \mbox{ choose } \bff \mbox{ to minimize }
+\|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha \|A\bff\|^2 
 \end{equation}
@@
 
-\section{Operators and Adjoints}
+I will fully describe the preconditioner construction for a special
+case of the source subproblem \ref{eqn:esis}, in which ${\bf u}$ is an
+acoustic field, $L[{\bf c}]$ is the wave operator of linear
+acoustodynamics, the spatial positions of traces extracted by $P$ lie
+on a plane, and the positions at which the extended
+source $\bff$ is nonzero lie on another, parallel, plane. The data and
+sources are further limited to pressure traces and constitutive law
+defects (``pressure sources''). While this
+``crosswell'' configuration simplifies the analysis underlying the
+construction of approximate solutions for the source subproblem
+\ref{eqn:esis}, it is only one of many transmission configurations for
+which similar developments are possible. Perhaps the most important
+alternative example is the diving wave configuration, which plays a
+central role in contemporary FWI. I will discuss the generalization of
+surface source extention to diving wave inversion at the end of the
+paper.
+
+The preconditioner construction is very similar to the time-reversal
+method in photoacoustic tomography
+\cite[]{StefanovUhlmannIP:09}. Preconditioning amounts to a change of
+norm in the domain and range spaces of the modeling operator. In this
+case, the modfied norms are weighted $L^2$, and the weight operators
+map pressure to corresponding surface source on the source and
+receiver planes. These are essentially the same as the ``hyperbolic
+Dirichlet-to-Neumann'' map that plays a prominent role in
+thermoacoustic tomography and other wave inverse problems
+\cite[]{Rachele:00,StefUhl:05}. The obvious computation of these maps
+- prescribe the pressure, solve the wave equation with this boundary
+condition, read off the equivalent source - suffers from intrinsic
+numerical inaccuracy. I suggest an alternative computationally
+feasible approach.
+
+\section{Operators}
 
@@
   \label{eqn:awemat}
-  L[\bf{c}] = \left(
+  L[{\bf c}] = \left(
     \begin{array}{cc}
@@
 \end{equation}
+$L[{\bf c}]$ has a well-defined inverse if it is restricted to either
+causal or anti-causal vector wavefields.
 
@@
 located on another spatial hyperplane $\{(x,y,z,t): z=z_r\}$ at
-receiver depth $z_r>z_s$.
-
-Denote by $P_s,P_r$ the trace (sampling)
-operators on $z=z_s$, $z=z_r$ respectively.
-Under the standing assumptions to be made on
-the solutions of system \ref{eqn:awepm}, the key components of the
-solution ($p^{\pm}$ and $v^{\pm}_z$) are continuous functions of $z$
-in the open slab $z_s<z<z_r$, but may be discontinuous at the source
-plane $z=z_s$. Similarly, the roles of $z_s$ and $z_r$ will be
-interchanged in some of the constructions to come, and the
-corresponding solutions may be discontinuous at $z=z_r$. Accordingly,
-interpret $P_s$, $P_r$ as the limit from right and left respectively:
-for $u=p^{\pm}$ or $v^{\pm}_z$,
+receiver depth $z_r>z_s$. The constructions to follow involve interchange
+of the roles of $z_s$ and $z_r$ (that is, locating sources on $z=z_r$
+and receivers at $z=z_s$), so rather than the sampling operator $P$ of
+the introduction, I will denote by $P_s,P_r$ the sampling 
+operators on $z=z_s$, $z=z_r$ respectively. In practice, sampling
+occurs at a discrete array of points (trace locations) on these
+surfaces, and over a zone of finite extent. In this theoretical
+discussion, I will neglect both finite sample rate and extent, and
+regard the data, for example $P_rp^+$, as continuously sampled and
+extending over the entire plane $z=z_r$.
+
+A technical difficulty with spatial sampling must be
+addressed. Acoustic field energy is defined in terms of the mean
+squares of the pressure and particle velocity fields. However fields
+with finite mean square do not in general have well-defined
+restrictions to lower-dimensional sets: for example, the pressure
+field of a finite-energy acoustic vector field may not have a
+well-defined restriction to the space-time plane $z=z_r$. This
+phenomenon is related to the ill-posedness of wave equations as
+evolution equations in spatial variables, an observation attributed to
+Hadamard (see \cite{CourHil:62}, Chapter 6, section 17, also
+\cite{Payn:75,Symes:83}). Some constraint on the acoustic field,
+beyond finite energy, is mandatory in formulating the inverse problems
+\ref{eqn:esi} and \ref{eqn:esis}. In fact, the natural constraint in
+the ``crosswell'' geometry of this paper is that high-frequency energy
+{\em not} travel along rays parallel or nearly parallel to the
+surfaces $z=z_s, z=z_r$. I will call fields with this property {\em
+  downgoing} (even though the concept also encompasses {\em upcoming}
+fields). \cite{BaoSy:91b} give a mathematically complete discussion of
+downgoing wave field properties.
+
+For downgoing solutions of system \ref{eqn:awepm}, the key
+components ($p^{\pm}$ and $v^{\pm}_z$) are continuous functions of $z$
+in the open slab $z_s<z<z_r$ with well-defined limits at the boundary
+planes, but may be discontinuous at the source plane
+$z=z_s$. Similarly, the roles of $z_s$ and $z_r$ will be interchanged
+in some of the constructions to come, and the corresponding solutions
+may be discontinuous at $z=z_r$. Accordingly, interpret $P_s$, $P_r$
+as the limit from right and left respectively: for $u=p^{\pm}$ or
+$v^{\pm}_z$,
 \begin{eqnarray}
@@
 
-As will be shown later, ${\cal S}^{\pm}$ is not stably invertible, in fact
-having rank 1 (in an approximate sense). The diagonal components of
-${\cal S}^{\pm}$ thus carry essentially all of its information.
+\noindent {\bf Remark:} To connect with the formulation presented in
+the introduction, note that for continuous $u$,
+$P_su(x,y,t)=u(x,y,z_s,t)$, and therefore the adjoint of $P_s$ (in the
+sense of distributions) is $P_s^Th(x,y,z,t) =
+h(x,y,t)\delta(z-z_s)$. Write ${\cal P}_s = \mbox{diag }(P_s,P_s)$ and
+similarly for ${\cal P}_r$. Then
+\[
+  {\cal S}^{+}_{z_s,z_r} = {\cal P}_r L^{-1}({\cal P}_s)^T,
+\]
+in which $L^{-1}$ is interpreted in the causal sense, and similarly
+for ${\cal S}^{-}$. Sources confined to $z=z_s$ are precisely those
+functions (distributions, really) output by ${\cal P}_s^T$, so the
+problem statements \ref{eqn:esi} and \ref{eqn:esis} can be rewritten
+in terms of ${\cal S}^+_{z_s,z_r}$, with $P$ identified with ${\cal P}_r$.
+
+${\cal S}^{\pm}$ is not stably invertible, in fact having rank 1 (in
+an approximate sense - see the identity \ref{eqn:lamnull}). The
+diagonal components of ${\cal S}^{\pm}$ thus carry essentially all of
+its information, and it is in terms of these that a sensible inverse problem
+is defined.
+
 Denote by $\Pi_i, i=0,1$ the projection on the first,
@@
 \end{equation}
+
+With these conventions, we can write the version of the source
+subproblem studied in this paper as
+\begin{equation}
+  \label{eqn:esisp}
+  \mbox{find }h_s\mbox{ to minimize }\|S^{+}_{z_s,z_r}h_s- d\|^2 +
+  \alpha^2\|Ah_s\|^2.
+\end{equation}
+
+\section{2D Examples}
+
+\inputdir{project}
+
+To illustrate the structure described in the preceding section, I
+introduce two 2D acoustic models, one spatially homogeneous, the other
+with highly refractive. The first, homogenous model has $\kappa = 4$
+GPa and $\rho = 1$ g/cm$^3$ throughout a rectangular domain of size 8 km ($x$) $\times$ 4 km
+($z$). The second, refractive, model is a perturbation of the first by
+a low-velocity acoustic lens positioned in the center of the
+rectangle. To produce this structure, the density is chosen
+homogeneous as in the first model, while the bulk modulus decreases to
+from 4 GPa outside the lens to 1.6 GPA in its center, as shown in Figure \ref{fig:bml0}.
+
+Discretization is conventional, with a rectangular grid and staggered
+finite difference scheme \cite[]{Virieux:84} of order 2 in time and
+2$k$ in space; for most of the experiments reported below, $k=4$.
+Sampling operators such as $P_r$ are implemented via linear
+interpolation, and source insertion via adjoint linear interpolation
+(as noted above, in the continuum limit, sources are represented via
+adjoint sampling). Steps in $x$ and $z$ are the same. Most results use
+$\Delta x = 20$ m, and limit the temporal frequency of the computed
+traces to 12.5 Hz. We also explore the dependence of a few results on frequency, using
+$\Delta x = 10$ m and $5$ m, to accomodate 25 and 50 Hz respectively,
+maintaining 8 samples per wavelength.
+
+The horizontal line of receivers sits at depth $z_r = $ 1000 m, as
+shown in Figure \ref{fig:bml0}. Receiver $x$ ranges from $2000$ to
+$6000$ m. A single point (physical) source appears in these
+experiments, located at $x_s=3500$ m, $z_s=3000$ m, as also indicated
+in the figure. The source time function is a bandpass filter with
+corner frequencies $1, 2.5, 7.5, 12.5$ Hz for the lowest frequency
+source, and scaled as appropriate for examples with higher frequency
+and finer sampling.  Extended sources are confined to the horizontal
+line through the physical source position, that is $z_s = 3000$ m,
+over a 4 km interval starting at $x_r=$ 2000 m. Note that we have
+reversed the order relation between $z_s$ and $z_r$ described in the
+text ($z_s<z_r$). This difference is immaterial.
+
+\plot{bml0}{width=\textwidth}{Bulk modulus, lens model. Color scale is
+in GPa. Positions of point source and receiver line indicated.}
+
+The point source pressure data generated by this configuration for the
+homogeneous model is displayed in Figure \ref{fig:recphh0}, for the
+lens model in \ref{fig:recplh0}. Triplication of arrivals is evident in the latter.
+
+\plot{recphh0}{width=\textwidth}{Output pressure gather for point source in 
+  homogeneous medium, configuration as described in Figure \ref{fig:bml0} and in
+  the text.}
+
+\plot{recplh0}{width=\textwidth}{Output pressure gather for point source in 
+  refractive (lens) medium, configuration as described in Figure \ref{fig:bml0} and in
+  the text.}
+
+While inversion of pressure data is the main object of this exercise,
+normal velocity ($v_z$) data will play an important role, so I display
+the corresponding gathers in Figures \ref{fig:recvzhh0} and \ref{fig:recvzlh0}.
+
+\plot{recvzhh0}{width=\textwidth}{Output vertical velocity gather for point source in 
+  homogeneous medium, configuration as described in Figure \ref{fig:bml0} and in
+  the text.}
+
+\plot{recvzlh0}{width=\textwidth}{Output vertical velocity gather for point source in 
+  refractive (lens) medium, configuration as described in Figure \ref{fig:bml0} and in
+  the text.}
+
+\section{Adjoints}
 

### 2021-05-18 repro/papers/sseprecond/paper.tex
interim

@@
 $\Delta x = 10$ m and $5$ m, to accomodate 25 and 50 Hz respectively,
-maintaining 8 samples per wavelength.
+maintaining 8 samples per wavelength. All computations are carried out
+in single precision.
+
+\cite{GeoPros:11} gives a description of the code
+implementation, out-of-date in a few respects but overall accurate.
 
@@
 \end{eqnarray}
+
+The implemented code uses the adjoint state method on the discrete
+level, to assure that the computed adjoint operators are adjoint at
+the level of machine precision to the computed operators. The issue of
+reverse-time storage is resolved through the optimal checkpointing
+technique \cite[]{Griewank:book,Symes:06a-pub}, again without loss of
+precision.
+
+Typical results with pseudorandom input traces $d_r, h_z, w_r, f_s$ are:
+\begin{itemize}
+\item computed $\langle d_r, S^+_{z_s,z_r}h_s\rangle = -2.41069174,
+  \langle (S^+_{z_s,z_r})^Td_r, h_s \rangle = -2.41069388.$
+\item  computed $\langle w_r, V^+_{z_s,z_r}f_s\rangle = -2.73362470,
+  \langle (V^+_{z_s,z_r})^Tw_r, f_s \rangle = -2.73362136.$
+\end{itemize}
+Ideally, the differences of these inner products should be at most a
+relatively small multiple of machine precision, {\em relative} to the
+products $\|d_r\|\| S^+_{z_s,z_r}\|\|h_s\|$ and
+$\|w_r\|\|V^+_{z_s,z_r}\|\|f_s\|$ (thus making the result
+dimensionless). The operator norm $\| S^+_{z_s,z_r}\|$ is
+computationally inaccessible, so instead I used the smaller quantities
+$\|d_r\|\| S^+_{z_s,z_r}h_s\|$ etc. as stand-ins - thereby
+overestimating the relative error between the inner products. In all
+cases, over a very large number of random inputs, the largest observed
+relative error estimate was $O(10^{-9}$, well under the appropriate limit for
+single precision.
+
+\section{Approximate Inversion}
+
+\section{Iterative Inversion}
 

### 2021-05-21 repro/papers/sseprecond/paper.tex
interim

--- 2021-05-18 repro/papers/sseprecond/paper.tex
@@
 h_s \delta(z-z_s), \nonumber \\
-\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
+\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} +
+                                                f_s{\bf e} \delta(z-z_s),\nonumber \\
 p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
@@
 
-${\cal S}^{\pm}$ is not stably invertible, in fact having rank 1 (in
-an approximate sense - see the identity \ref{eqn:lamnull}). The
+${\cal S}^{\pm}$ is not stably invertible (equation \ref{eqn:lamnull}). The
 diagonal components of ${\cal S}^{\pm}$ thus carry essentially all of
@@
 Discretization is conventional, with a rectangular grid and staggered
-finite difference scheme \cite[]{Virieux:84} of order 2 in time and
+finite difference scheme \cite[]{Vir:84} of order 2 in time and
 2$k$ in space; for most of the experiments reported below, $k=4$.
@@
 \end{eqnarray}
+
+The implemented code uses the discrete adjoint state method and auto-generated code \cite[]{TapenadeRef13}, to
+assure that the computed adjoint operators are adjoint at the level of
+machine precision to the computed operators. The reverse-time storage
+issue is resolved through the optimal checkpointing technique
+\cite[]{Griewank:book,Symes:06a-pub}, again without loss of precision.
+
+Typical results with pseudorandom input traces $d_r, h_z, w_r, f_s$ are:
+\begin{itemize}
+\item computed $\langle d_r, S^+_{z_s,z_r}h_s\rangle = -2.41069174,
+  \langle (S^+_{z_s,z_r})^Td_r, h_s \rangle = -2.41069388.$
+\item  computed $\langle w_r, V^+_{z_s,z_r}f_s\rangle = -2.73362470,
+  \langle (V^+_{z_s,z_r})^Tw_r, f_s \rangle = -2.73362136.$
+\end{itemize}
+Ideally, the differences of these inner products should be at most a
+relatively small multiple of machine precision, {\em relative} to the
+products $\|d_r\|\| S^+_{z_s,z_r}\|\|h_s\|$ and
+$\|w_r\|\|V^+_{z_s,z_r}\|\|f_s\|$ (division by these quantities makes the result
+dimensionless and scale-independent). The operator norm $\| S^+_{z_s,z_r}\|$ is
+computationally inaccessible, so instead I used the smaller quantities
+$\|d_r\|\| S^+_{z_s,z_r}h_s\|$ etc. as stand-ins - thereby
+overestimating the relative error between the inner products. In all
+cases, over a very large number of random inputs, the largest observed
+relative error estimate was $O(10^{-9})$, well under the appropriate limit for
+single precision.
+
+\section{Pressure-to-Source}
+
+Since the system \ref{eqn:awepm} has a unique solution by standard
+theory \cite[]{Lax:PDENotes}, the source vector field $(h_s,f_s)$
+determines the acoustic field $(p^{\pm},\bv^{\pm})$ in space time, and
+in particular the limits from the right at $z=z_s$, $P_sp^{\pm}$ and
+$P_sv_z^{\pm}$. This relation is not invertible: it is not possible to
+prescribe both pressure and normal velocity on a surface such as
+$z=z_s$. So the columns of the matrix operator
+${\cal S}^{\pm}_{z_s,z_r}$ must satisfy a linear relation. In this
+section I will explain this relation; it involves the {\em
+  pressure-to-source} map. This operator also turns out to be the
+principle component of a preconditioning strategy for iterative
+solution of the optimization problem \ref{eqn:esis}, so I will devote
+some effort to its proper definition. It is closely related to the
+Dirichlet-to-Neumann operator mentioned in the introduction.
+
+While it is not possible to prescribe both pressure and velocity on
+$z=z_s$ in solutions of \ref{eqn:awepm}, it is possible to
+prescribe pressure only, for instance: if the function $\phi$ on
+the surface $z=z_s$ satisfies suitable conditions, for
+example the downgoing constraint mentioned earlier, a unique solution
+exists for the acoustic system in both half-spaces $\pm z > z_s$:
+\begin{eqnarray}
+\label{eqn:awe0}
+  \frac{1}{\kappa}\frac{\partial p_{\pm}}{\partial t} & = & - \nabla \cdot \bv_{\pm}, \nonumber \\
+  \rho\frac{\partial \bv_{\pm}}{\partial t} & = & - \nabla
+                                                    p_{\pm}, \nonumber \\
+  p_{\pm} & =& 0,  \mbox{ for } t \ll 0, \nonumber\\ 
+  \bv_{\pm} & = & 0 \mbox{ for } t \ll 0, \nonumber\\
+  \lim_{z \rightarrow z_s^{\pm}}p_{\pm}(x,y,t,z)& =& \phi(x,y,t).
+\end{eqnarray}
+Note that the subscript $\pm$ here refers to the sign of $z-z_s$, as opposed
+to the superscript ${\pm}$, which refers to the sign of $t$ throughout
+this paper.
+
+From the boundary condition (last equation in \ref{eqn:awe0}), one
+sees that the pressures $p_{\pm}$ in the two half-spaces have the same
+limit at the boundary $z=z_s$. Stick the two half-space
+solutions together to form an acoustic field $(p^+,\bv^+)$ in all of
+space-time, that is,
+\begin{equation}
+  \label{eqn:awealt}
+  p^+(x,y,z,t) =
+  \left\{
+    \begin{array}{c}
+      p_+(x,y,z,t) \mbox{ if } z>0,\\
+      p_-(x,y,z,t) \mbox{ if } z<0,
+    \end{array}
+  \right.
+\end{equation}
+and a similar definition for $\bv^+$. Then $p^+$ is continuous across
+$z=z_s$, and the boundary condition in system \ref{eqn:awe0} may be
+written as $P_sp^+=\phi$.
+
+The same construction can be carried out in the anti-causal sense,
+with anti-causal half-space solutions glued together to form a
+full-space distribution solution $(p^-,\bv^-)$, with the property that
+$p^-$ is continuous across $z=z_s$ and $P_sp^-=\phi$.
+
+The reader may object that the notation $(p^\pm,\bv^\pm)$ is already in
+use, for the solution of \ref{eqn:awepm}. This objection is
+valid. However, {\em in the sense
+  of distributions}, $(p^{\pm},\bv^{\pm})$ as defined in display
+\ref{eqn:awealt}, is {\em exactly} the causal solution of \ref{eqn:awepm}
+for the choice $h_s = -[v^{\pm}_{z}]|_{z=z_s}, f_s=0$, as follows from a
+simple integration-by-parts calculation. So the notation is consistent!
+
+The negative jump $-[v^{\pm}_{z}]|_{z=z_s}$ is thus a function of $\phi$. Define
+the {\em pressure-to-source} operator $\Lambda^{\pm}_{z_s}$ by
+\begin{equation}
+  \label{eqn:deflam}
+  \Lambda^{\pm}_{z_s}\phi = -[v^{\pm}_{1,z}]|_{z=z_s}
+\end{equation}
+The conclusion: if $h_s = \Lambda^{\pm}_{z_s}\phi$ and $f_s=0$ in the
+system \ref{eqn:awepm}, then $\phi=P_sp^{\pm}$.
+
+Otherwise put, $S^{\pm}_{z_s,z_s}\Lambda^{\pm}_{z_s} \phi = \phi$, so
+$\Lambda^{\pm}_{z_s}$ is inverse to $S^{\pm}_{z_s,z_s}$. The relation
+\ref{eqn:trtrcomp} implies in turn that
+\begin{equation}
+  \label{eqn:lamadj}
+  (\Lambda^{\pm}_{z_s})^T = - \Lambda^{\mp}_{z_s}
+\end{equation}
+
+There is also a {\em velocity-to-source} operator. For the solution
+$(p^{\pm},\bv^{\pm})$ of system \ref{eqn:awepm} with $h_s=0$, the
+normal component of velocity, $v^{\pm}_z$, is continuous across
+$z=z_s$, and the velocity source (vertical load)
+$f_s=-[p^{\pm}]_{z=z_s}$. I will not name the velocity-to-source
+operator, as it does not appear explicitly in the developments to
+follow. As will be seen, it is essentially the inverse of the
+pressure-to-source operator.
+
+The quadratic form defined by $\Lambda^{\pm}_{z_s}$ has fundamental
+physical significance. Define the total acoustic energy $E^{\pm}(t)$ of the
+field $(p^{\pm},\bv^{\pm})$, at time $t$ by
+\begin{equation}
+  \label{eqn:defae0}
+  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)
+\end{equation}
+Then
+\begin{equation}
+  \label{eqn:elim}
+  \pm \lim_{\pm t \rightarrow \infty} E^{\pm}(t) =  \langle P_sp^{\pm},
+  (\Lambda^{\pm}_{z_s} P_sp^{\pm}) \rangle_{L^2(z=z_s)}.
+\end{equation}
+That is, the value of the quadratic form defined by
+$\Lambda^{\pm}_{z_s}$, evaluated at the pressure trace on $z=z_s$,
+gives the total energy transferred from the source to the
+acoustic field over time. Since $E$ is itself a positive definite
+quadratic form in the acoustic field, it follows that $\pm
+\Lambda^{\pm}_{z_s}$ is positive semi-definite.
+
+While $\Lambda^{\pm}_{z_s}$ is positive semi-definite, it is not
+symmetric. However, it is {\em approximately symmetric} in the
+high-frequency sense. This fact follows from a
+geometric optics analysis of the half-space solution,
+assuming (as always) downgoing data:
+\begin{equation}
+  \label{eqn:lamappsim}
+  (\Lambda^{\pm}_{z_s})^T \approx \Lambda^{\pm}_{z_s}.
+\end{equation}
+
+The analysis also reveals that the solution components not continuous
+at $z=z_s$ are odd there:
+\begin{equation}
+  \label{eqn:odd1}
+  \lim_{z\rightarrow z_s^+} v^{\pm}_{z} \approx - \lim_{z\rightarrow z_s^-}
+  v^{\pm}_{z}
+\end{equation}
+for the solution of \ref{eqn:awepm} with $f_s=0$.
+Similarly, 
+\begin{equation}
+  \label{eqn:odd2}
+  \lim_{z\rightarrow z_s^+} p^{\pm}\approx - \lim_{z\rightarrow z_s^-}
+  p^{\pm}
+\end{equation}
+for the solution of \ref{eqn:awepm} with $h_s=0$. Here ``$\approx$''
+means in the sense of high frequency asymptotics, that is, that the
+difference between the two sides is relatively smooth, hence small if
+the data is highly oscillatory. Therefore if $f_s=0$ in system \ref{eqn:awepm},
+\begin{equation}
+  h_s = \Lambda^{\pm}_{z_s}P_sp^{\pm} = -[v^{\pm}_{z}]|_{z=z_s} \approx -2
+  P_sv^{\pm}_{z}
+  \label{eqn:tracejump10}
+\end{equation}
+Similarly, if $h_s=0$ in system \ref{eqn:awepm}, then
+\begin{equation}
+  \label{eqn:tracejump20}
+  f_s = -[p^{\pm}]|_{z=z_s} \approx -2 P_s p^{\pm}.
+\end{equation}
+Thus $f_s$ determines approximately the boundary value of $p^{\pm}$,
+as a solution of the acoustic wave system in the half-space
+$z>z_s$. However, as repeated in equation \ref{eqn:tracejump10}, a
+solution with this boundary value is also the restriction to $z>z_s$
+of a solution to \ref{eqn:awepm} with $f_s=0$ and $h_s=
+\Lambda^{\pm}_{z_s}P_sp^{\pm}$. Therefore if
+\begin{equation}
+  \label{eqn:hfcondn}
+  h_s =\Lambda^{\pm}_{z_s}f_s,
+\end{equation}
+then the pressure boundary value $P_sp^{\pm}$ is the
+same for the solutions of \ref{eqn:awepm} for source vectors $(h_s,0)$
+and $(0,f_s)$. Since the pressure boundary values are the same, the solutions
+in $z>z_s$ are the same. In particular, since $z_r>z_s$ and ${\cal
+  S}^{\pm}_{z_s,z_r}(h_s,f_s)^T = (P_rp^{\pm},P_rv^{\pm}_z)^T$, it follows
+that
+\begin{equation}
+  \label{eqn:snull}
+  {\cal S}^{\pm}_{z_s,z_r}(-\Lambda^{\pm}_{z_s}f_s,f_s)^T \approx 0.
+\end{equation}
+
+Equation \ref{eqn:snull} states the relation between the columns of $
+{\cal S}^{\pm}_{z_s,z_r}$ mentioned in the introduction to this
+section.
+
+To give an example of \ref{eqn:snull} in action, it is necessary to
+create downgoing data and the action of the operator $\Lambda^{\pm}$
+on a downdoing field data $f_s$ on $z=z_s$. Note that a point source
+on $z=z_s$ creates high frequency energy traveling on rays parallel
+and nearly parallel to $z=z_s$, so that won't do. One simple approach
+is to create pressure and normal velocity data on $z=z_s$ by placing a
+point source at a depth $z_d<z_s$. Since the examples used here are
+homogeneous in $z<z_s$, the traces extracted from the resulting
+causal pressure and velocity data on $z=z_s$ are {\em a priori}
+downgoing, hence related by the operator $\Lambda^+_{z_s}$.
+
+For a point source at $z_d=3500$ m, $x_d=3500$ m, same bandpass filter
+wavelet as used for prior examples, causal data on $z=z_s=3000$ m are
+shown in Figures \ref{fig:dsrcphh0} and \ref{fig:dsrcvzhh0}. Since the
+mechanical parameters in the homogeneous and lens models are the same
+for $z<z_s$, and no rays return to this zone in either model, these
+data are asymptotically the same for both models, and I show only the
+homogenous medium results. Regard these gathers as the trace
+$(P_sp^+,P_sv^+_z)$ on $z=z_s$ of a downgoing acoustic field in $z>z_s$.
+Equations \ref{eqn:tracejump10} and
+\ref{eqn:tracejump20} show that these differ by a factor of -2 from
+source functions $f_s$ and $h_s$ as in the system \ref{eqn:awepm},
+with $h_s=0$ and $f_s=0$ respectively.
+
+These source functions satisfy
+the relation \ref{eqn:hfcondn}, therefore source vectors $(h_s,0)$ and
+$(0, f_s)$ generate {\em the same acoustic field} $(p^+,\bv^+)$ in $z>z_s$. Figures
+\ref{fig:drecphh0}, \ref{fig:dfwdphh0}, \ref{fig:daltphh0} (homogeneous model) and
+\ref{fig:drecplh0}, \ref{fig:dfwdplh0}, \ref{fig:daltplh0} (lens model) show the pressure
+gathers extracted at $z_r=1000$ m for the point source at $z=z_d$ and
+for the two choices of extended source at $z=z_s$, on the same color
+scale. The obvious similarity between the fields generated by the two
+extended sources,
+predicted by equation \ref{eqn:snull}, is confirmed by trace
+comparisons in figures \ref{fig:drecphh0tr21}-\ref{fig:daltplh0tr81}.
+
+\plot{dsrcphh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
+  pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
+  bandpass filter source.}
+
+\plot{dsrcvzhh0}{width=\textwidth}{Trace $P_sv_z^+$ on $z=z_s=3000$ m of
+  vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
+  bandpass filter source.}
+
+\plot{drecphh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the homogeneous model described in the 
+  text, with point pressure source (constitutive defect) at $z_d=3500$
+  m, $x_d=3500$ m.}
+
+\plot{dfwdphh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the homogeneous model described in the 
+  text, with extended pressure source (constitutive defect) on $z=z_s=3000$ m given by $-2$
+  times the field depicted in Figure \ref{fig:dsrcvzhh0}
+  ($h_s=-2P_sv_z^+$) and zero velocity source (vertical load) 
+  ($f_s=0$).}
+
+\plot{daltphh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the homogeneous model described in the 
+  text, with extended velocity source (vertical load) on $z=z_s=3000$ m given by $-2$
+  times the field depicted in Figure \ref{fig:dsrcphh0}
+  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
+  ($h_s=0$).}
+
+\plot{drecplh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the lens model described in the 
+  text, with point pressure source (constitutive defect) at $z_d=3500$
+  m, $x_d=3500$ m.}
+
+\plot{dfwdplh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the lens model described in the 
+  text, with extended pressure source (constitutive defect) on $z=z_s=3000$ m given by $-2$
+  times the field depicted in Figure \ref{fig:dsrcvzhh0}
+  ($h_s=-2P_sv_z^+$) and zero velocity source (vertical load) 
+  ($f_s=0$).}
+
+\plot{daltplh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the lens model described in the 
+  text, with extended velocity source (vertical load) on $z=z_s=3000$ m given by $-2$
+  times the field depicted in Figure \ref{fig:dsrcphh0}
+  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
+  ($h_s=0$).}
+
+\plot{drecphh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$)
+  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
+  (red).}
+
+\plot{daltplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$)
+  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
+  (red).}
+\section{Approximate Inversion}
+
+\section{Accelerated Iterative Inversion}
 
@@
 
-The quadratic form defined by $\Lambda^{\pm}_{z_s}$ has fundamental
-physical significance. Define the total acoustic energy $E^{\pm}(t)$ of the
-field $(p^{\pm},\bv^{\pm})$, at time $t$ by
-\begin{equation}
-  \label{eqn:defae0}
-  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)
-\end{equation}
-Then
-\begin{equation}
-  \label{eqn:elim}
-  \pm \lim_{\pm t \rightarrow \infty} E^{\pm}(t) =  \langle P_sp^{\pm},
-  (\Lambda^{\pm}_{z_s} P_sp^{\pm}) \rangle_{L^2(z=z_s)}.
-\end{equation}
-That is, the value of the quadratic form defined by
-$\Lambda^{\pm}_{z_s}$, evaluated at the pressure trace on $z=z_s$,
-gives the total energy transferred from the source to the
-acoustic field over time. Since $E$ is itself a positive definite
-quadratic form in the acoustic field, it follows that $\pm
-\Lambda^{\pm}_{z_s}$ is positive semi-definite.
-
 A closer relation between traces and sources follows from the
@@
 
+%%%%%%%%%%%%%%%%%%%
 These apparently harmless relations have a profound consequence,

### 2021-05-22 repro/papers/sseprecond/paper.tex
interim

@@
   (red).}
+
 \section{Approximate Inversion}
+
+Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
+efficient iterative solution of the linear subproblem
+\ref{eqn:esis}. Direct computation, for instance by solving
+\ref{eqn:sawe2} and reading off $P_sv^{\pm}_{2,z}$, turns out to be
+numerically ill-behaved. The relation \ref{eqn:snull} provides and
+alternative approach, {\em provided} that an accurate approximate
+inverse to $S^+_{z_s,z_r}$ is available. For instance, the first row
+of the vector equation \ref{eqn:snull} reads (slightly rearranged):
+\[
+\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1 f_s \approx \Pi_0{\cal
+  S}^+_{z_s,z_r}Pi_0\Lambda^+_{z_s}f_s
+\]
+\begin{equation}
+  \label{eqn:lamidea}
+  = S^++_{z_s,z_r}\Lambda^+_{z_s}f_s.
+\end{equation}
+An approximate inverse for $S^++_{z_s,z_r}$ would permit solution of
+this equation for $\Lambda^+_{z_s}f_s$.
+
+In this section I will explain how to build the approximate inverse.
+The key ingredient in this {\em time reversal} construction is a
+{\em local energy decay} assumption. Define the local energy in the
+acoustic field $(p,\bv)$ at
+time $t$ in a region $\Omega \subset \bR^3$ by limiting the integration
+in the definition of total energy \ref{eqn:defae0} to $\Omega$:
+\begin{equation}
+  \label{eqn:defaeloc}
+  E_{\Omega}[(p,\bv)](t) = \frac{1}{2} \int_{\Omega} \,d\bx \, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx) |\bv(\bx,t)|^2\right)
+\end{equation}
+The necessary assumption is: local energy decay in the half-space
+$\Omega = \{(x,y,z): z \le z_r\}$. The local energy in $\Omega$ of the
+causal solution $(p^+,\bv^+)$ of system \ref{eqn:awepm} cannot decay,
+since the source on $z=z_s<z_r$ radiates in both positive and negative
+$z$ in forward time. That obstacle is resolved by a redefinition of
+the ``downgoing'' concept. I will call  downgoing if there is a
+source-free acoustic field $(\tilde{p}^+,\tilde{\bv}^+)$ in all of space-time for which
+\begin{itemize}
+\item[1. ] $p^+ \approx \tilde{p}^+, \bv^+ \approx \tilde{\bv}^+$ in
+  the slab $\{(x,y,z,t): z_s \le z \le z_r\}$, and
+\item[2. ] The energy in $(\tilde{p}^+,\tilde{\bv}^+)$ decays in
+  forward time, in the half-space $\Omega$:  $E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](t) \rightarrow 0$
+  as $t \rightarrow \infty$.
+\end{itemize}
+Conditions for local energy decay have been widely studied in various
+settings, for instance obstacle scattering
+\cite[]{MorawetzPhillipsLax:05}. \cite{Pauen2000NonTr-6079 gives a
+comprehensive overview of this subject. For smooth (non-scattering)
+acoustic parameters, as is assumed in this paper, the crux of the
+matter is that rays carrying high-frequence energy exit the half-space
+$\Omega$ in finite time. For the half-space
+geometries considered here, local energy does not decay in general,
+because of the existence of ``trapped'' rays traveling parallel to the
+boundary and remaining interior to $\Omega$ for all time. If an
+acoustic field is downgoing in the usual sense, that is, its
+high-frequency energy is concentrated on rays that exit $\Omega$
+within some common maxiumum time, then the results cited above show
+that local energy in $\Omega$ decays (exponentially for $n=3$).
+
+Starting from this observation, it is possible to show that if the
+source vector $(h_s,f_s)^T$ has the property termed ``downgoing'' in
+preceding sections, that is, emanates high-frequency energy only along
+rays that leave $\Omega$ within a common maximum time, then the
+solution $(p^+,\bv^+)$ of \ref{eqn:awepm} is downgoing in the sense
+defined here. All of our numerical examples behave as one would expect
+for downgoing fields.
+
+An immediate consequence of the downgoing assumption stated above is that in the
+slab $z_s<z<z_r$, the field $(p^+,\bv^+)$ approximates the solution of an
+anti-causal evolution equation. For any choice of $T$, define
+$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
+$\Omega$ of
+\begin{eqnarray}
+\label{eqn:revawe}
+  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
+  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
+  \tilde{p}^- & =& 0,  \mbox{ for } t=T,\nonumber\\ 
+  \tilde{\bv}^- & = & 0 \mbox{ for } t=T,\nonumber\\
+  P_r\tilde{p}^- &=& P_rp^+ \mbox { for } t \le T . 
+\end{eqnarray}
+As will be shown later, under the downgoing assumption, $P_r p^+
+\approx P_r \tilde{p}^+$.
+Therefore $(\delta p,\delta \bv)=(\tilde{p}^+-\tilde{p}^-,\tilde{\bv}^+-\tilde{\bv}^-)$
+differs negligibly from the solution of
+\begin{eqnarray}
+\label{eqn:deltaawe}
+  \frac{1}{\kappa}\frac{\partial \delta p}{\partial t} & = & -
+                                                               \nabla
+                                                               \cdot
+                                                               \delta \bv, \nonumber \\
+  \rho\frac{\partial \delta \bv}{\partial t} & = & - \nabla \delta p,\nonumber \\
+  \delta p & =& \tilde{p}^+,  \mbox{ for } t=T,\nonumber\\ 
+  \delta \bv & = & \tilde{\bv}^+ \mbox{ for } t=T,\nonumber\\
+  P_r\delta p & = & 0,
+\end{eqnarray}
+in $\Omega \times \bR$. Since
+$E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](T) \rightarrow 0$ as $T
+\rightarrow \infty$, the usual energy estimate implies that the energy $\delta
+E$ in this field,
+\[
+  \delta E_{\Omega}(t) = \frac{1}{2} \int_{\Omega} \,d\bx \,
+ \left(\frac{(\delta p)^2}{\kappa} + \rho |\delta \bv|^2\right)
+\]
+satisfies $\delta E_{\Omega}(t) \rightarrow 0$ as $T \rightarrow
+\infty$. This decay does not generally imply that $P_s\delta p, P_s \delta \bv)
+\rightarrow 0$ (for instance) as $T \rightarrow \infty$, since the trace operator
+$P_s$ is not continuous, but it does so imply for purely downgoing
+fields, as will be demonstrated later, so
+\begin{equation}
+  \label{eqn:approxinv0}
+  \lim_{T \rightarrow \infty} P_s\tilde{v}_z^- \approx \lim_{z \rightarrow
+    0^+}\tilde{v}_z^+ \approx \lim_{z \rightarrow 0^+}v^+_z
+\end{equation}
+
+Recall (equations \ref{eqn:deflam}, \ref{eqn:jump0}) that
+\[
+  \Lambda^+_{z_s} P_s p^+ \approx 2 \lim_{z \rightarrow z_s^+}v^+_z 
+\]
+and that with $h_s= \Lambda^+_{z_s}P_sp^+$,
+\[
+  S^+_{z_s,z_r}h_s = P_r p^+
+\]
+Together with the last two equations, the relation
+\ref{eqn:approxinv0} implies that
+\begin{equation}
+  \label{eqn:approxinv1}
+  S^+_{z_s,z_r}(2 P_s \tilde{v}_z^-) \approx P_rp^+.
+\end{equation}
+As shown in the last section, the solution of the system
+\ref{eqn:revawe} differs negligibly from the anti-causal solution $(p^-_2,\bv^-_2)$
+of the system \ref{eqn:sawe2} with (i) $z_s$ replaced by $z_r$, and
+(ii) $\psi = [p^-_2]|_{z=z_r} = -2 \lim_{z \rightarrow z_r^-} p^-_2
+\approx -2 P_rp^+$. Therefore (definitions \ref{eqn:sdef}, \ref{eqn:vdef})
+\begin{equation}
+  \label{eqn:approxinv2}
+  P_s \tilde{v}_z^- \approx V^-_{z_r,z_s}(-2 P_rp^+).
+\end{equation}
+Taken together, relations \ref{eqn:approxinv1} and
+\ref{eqn:approxinv2} yield the first equation in the first main result of this section:
+\begin{eqnarray}
+  \label{eqn:approxinv}
+  V^-_{z_r,z_s} S^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
+  S^-_{z_r,z_s} V^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
+  V^+_{z_s,z_r} S^-_{z_r,z_s} & \approx & -\frac{1}{4}I, \nonumber\\
+  S^+_{z_s,z_r} V^-_{z_r,z_s} & \approx & -\frac{1}{4}I.\\
+\end{eqnarray}.
+The second equation follows by an exactly analogous argument, and the
+last two by time reversal and interchange of the roles of $z_s$ and
+$z_r$.
+
+Having identified approximate two-sided inverses for $S^{\pm},
+V^{\pm}$, the approximate null space identity \ref{eqn:lamnull} leads
+to a relation between $\Lambda^{\pm}$ and ${\cal S}^{\pm}$ that
+provides the foundation for practical computations. Applying $\Pi_0$
+to both sides of \ref{eqn:lamnull} gives
+\[
+ \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_0^T
+ \Lambda^+_{z_s} = S^+_{z_s,z_r} \Lambda^+_{z_s}.
+\]
+From the first approximate identity in \ref{eqn:approxinv},
+\begin{equation}
+  \label{eqn:lamident}
+  V^{-}_{z_r,z_s}\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx -\frac{1}{4}\Lambda^+_{z_s}.
+\end{equation}
+This identity is the second major result of this section: it shows how
+to compute that action of $\Lambda^+_{z_s}$ by propagating the input
+pressure trace, identified as a source for the velocity evolution,
+forward in time from $z_s$ to $z_r$
+reading off the pressure trace on $z=z_r$, identifying it once more as
+a source for velocity, propagating it backwards in time from $z_r$ to
+$z_s$, and finally reading off the velocity trace, interpreted as a
+pressure evolution source on $z_s$. I will give a detailed description
+of this computation in the section on numerical results.
+
+The importance of this result lies in the failure of the obvious
+method for computing the action of $\Lambda^{\pm}_{z_s}$, namely to
+employ the pressure trace as a source in the velocity equation ($f_s$,
+in the notation used above) at $z=z_s$, and read off the velocity
+field also at $z=z_s$. This difficulty is related to the existence of
+tangentially propagating waves and the lack of continuity of the trace
+operator. The method implicit in equation \ref{eqn:lamident} avoids
+this difficulty by propagating the fields a positive distance in $z$:
+assuming as always that the causal fields are downgoing, this step
+eliminates any tangentially propagating fields from consideration.
+
+Computation of the transpose of $\Lambda^+$ (exact, not approximate in
+the high frequency sense) will be important in the
+construction of a preconditioner. The relation \ref{eqn:lamident} also
+provides an computation for this operator. The idea is to treat the
+left-hand side of equation \ref{eqn:lamident} as an approximation to
+$\Lambda$, and compute its exact transposes via the adjoint state
+method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
+\begin{equation}
+  \label{eqn:lamtransp}
+   \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r} \approx -\frac{1}{4}(\Lambda^+_{z_s})^T.
+\end{equation} 
+The left-hand sides of \ref{eqn:lamident} and \ref{eqn:lamtransp} will
+simply be substituted for $\Lambda$ and its transpose in the
+preconditioner construction: preconditioners must be (precisely)
+self-adjoint, but need only be approximated to perform well.
 
@@
 
-Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
-efficient iterative solution of the linear subproblem
-\ref{eqn:esis}. Direct computation, for instance by solving
-\ref{eqn:sawe2} and reading off $P_sv^{\pm}_{2,z}$, turns out to be
-numerically ill-behaved. An accurate computation is available via the
-computation of a {\em parametrix} (high frequency approximate inverse)
-for $S^{\pm}_{z_s,z_r}$. This parametrix also justifies the use of the
-symmetric part of $\Lambda^{\pm}_{z_s}$ as a preconditioner. In this
-section I will explain the parametrix construction, and also show how
-to extract the symmetric part of $\Lambda^{\pm}_{z_s}$.
-
-The key ingredient in this {\em time reversal} construction is a
-{\em local energy decay} assumption. Define the local energy in the
-acoustic field $(p,\bv)$ at
-time $t$ in a region $\Omega \subset \bR^3$ by imiting the integration
-in the definition of total energy \ref{eqn:defae0} to $\Omega$:
-\begin{equation}
-  \label{eqn:defaeloc}
-  E_{\Omega}[(p,\bv)](t) = \frac{1}{2} \int_{\Omega} \,d\bx \, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx) |\bv(\bx,t)|^2\right)
-\end{equation}
-
-We assume that $(p^+,\bv^+)$ is {\em downgoing} in the half-space
-$\Omega = \{(x,y,z): z \le z_r\}$: that is, there exists a solution of
-the acoustic system $(\tilde{p}^+,\tilde{\bv}^+)$ in space-time $\bR^4$
-for which
-\begin{itemize}
-\item[1. ] $p^+ \approx \tilde{p}^+, \bv^+ \approx \tilde{\bv}^+$ in
-  the slab $\Gamma = \{(x,y,z): z_s \le z \le z_r\}$, and
-\item[2. ] The energy in $(\tilde{p}^+,\tilde{\bv}^+)$ decays in
-  forward time, in the half-space $\Omega$:  $E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](t) \rightarrow 0$
-  as $t \rightarrow \infty$.
-\end{itemize}
-
-Conditions for local energy decay have been widely studied in various
-settings \cite[]{MorawetzPhillipsLax:05,Hristova:09}. For the half-space
-geometries considered here, local energy does not decay in general,
-because of the existence of ``trapped'' rays traveling parallel to the
-boundary. Instead, high frequency data associated with rays crossing
-the boundary $\{z=z_r\}$ transversally, and never returning to the slab, have
-the local energy decay property.
-
-The downgoing constraint has mostly been studied in relation to
-seismology, mostly as a ray-theoretic concept: downgoing fields are
-concentrated in phase space along rays for which depth $z$ is
-increasing with $t$. The version above is stated without reference to
-ray theory, but the setting of this paper allows its use and in a
-later section I will explicitly construct the downgoing approximation
-under appropriate ray-theoretic conditions. The numerical examples will
-illustrate the necessary assumptions. When ray-theory is appropriate,
-the downgoing constraint can be {\em ex post facto} enforced via dip
-filtering. For example, the application of the ideas developed here to
-diving wave inversion will require such a filtering step.
-
-An immediate consequence of the energy decay assumption is that in the
-slab $\Gamma$, the field $(p^+,\bv^+)$ approximates the solution of an
-anti-causal evolution equation. For any choice of $T$, define
-$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
-$\Omega$ of
-\begin{eqnarray}
-\label{eqn:revawe}
-  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
-  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
-  \tilde{p}^- & =& 0,  \mbox{ for } t=T,\nonumber\\ 
-  \tilde{\bv}^- & = & 0 \mbox{ for } t=T,\nonumber\\
-  P_r\tilde{p}^- &=& P_rp^+ \mbox { for } t \le T . 
-\end{eqnarray}
-As will be shown later, under the downgoing assumtion, $P_r p^+
-\approx P_r \tilde{p}^+$.
-Therefore $(\delta p,\delta \bv)=(\tilde{p}^+-\tilde{p}^-,\tilde{\bv}^+-\tilde{\bv}^-)$
-differs negligibly from the solution of
-\begin{eqnarray}
-\label{eqn:deltaawe}
-  \frac{1}{\kappa}\frac{\partial \delta p}{\partial t} & = & -
-                                                               \nabla
-                                                               \cdot
-                                                               \delta \bv, \nonumber \\
-  \rho\frac{\partial \delta \bv}{\partial t} & = & - \nabla \delta p,\nonumber \\
-  \delta p & =& \tilde{p}^+,  \mbox{ for } t=T,\nonumber\\ 
-  \delta \bv & = & \tilde{\bv}^+ \mbox{ for } t=T,\nonumber\\
-  P_r\delta p & = & 0,
-\end{eqnarray}
-in $\Omega \times \bR$. Since
-$E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](T) \rightarrow 0$ as $T
-\rightarrow \infty$, the usual energy estimate implies that the energy $\delta
-E$ in this field,
-\[
-  \delta E_{\Omega}(t) = \frac{1}{2} \int_{\Omega} \,d\bx \,
- \left(\frac{(\delta p)^2}{\kappa} + \rho |\delta \bv|^2\right)
-\]
-satisfies $\delta E_{\Omega}(t) \rightarrow 0$ as $T \rightarrow
-\infty$. This decay does not generally imply that $P_s\delta p, P_s \delta \bv)
-\rightarrow 0$ (for instance) as $T \rightarrow \infty$, since the trace operator
-$P_s$ is not continuous, but it does so imply for purely downgoing
-fields, as will be demonstrated later, so
-\begin{equation}
-  \label{eqn:approxinv0}
-  \lim_{T \rightarrow \infty} P_s\tilde{v}_z^- \approx \lim_{z \rightarrow
-    0^+}\tilde{v}_z^+ \approx \lim_{z \rightarrow 0^+}v^+_z
-\end{equation}
-
-Recall (equations \ref{eqn:deflam}, \ref{eqn:jump0}) that
-\[
-  \Lambda^+_{z_s} P_s p^+ \approx 2 \lim_{z \rightarrow z_s^+}v^+_z 
-\]
-and that with $h_s= \Lambda^+_{z_s}P_sp^+$,
-\[
-  S^+_{z_s,z_r}h_s = P_r p^+
-\]
-Together with the last two equations, the relation
-\ref{eqn:approxinv0} implies that
-\begin{equation}
-  \label{eqn:approxinv1}
-  S^+_{z_s,z_r}(2 P_s \tilde{v}_z^-) \approx P_rp^+.
-\end{equation}
-As shown in the last section, the solution of the system
-\ref{eqn:revawe} differs negligibly from the anti-causal solution $(p^-_2,\bv^-_2)$
-of the system \ref{eqn:sawe2} with (i) $z_s$ replaced by $z_r$, and
-(ii) $\psi = [p^-_2]|_{z=z_r} = -2 \lim_{z \rightarrow z_r^-} p^-_2
-\approx -2 P_rp^+$. Therefore (definitions \ref{eqn:sdef}, \ref{eqn:vdef})
-\begin{equation}
-  \label{eqn:approxinv2}
-  P_s \tilde{v}_z^- \approx V^-_{z_r,z_s}(-2 P_rp^+).
-\end{equation}
-Taken together, relations \ref{eqn:approxinv1} and
-\ref{eqn:approxinv2} yield the first equation in the first main result of this section:
-\begin{eqnarray}
-  \label{eqn:approxinv}
-  V^-_{z_r,z_s} S^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
-  S^-_{z_r,z_s} V^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
-  V^+_{z_s,z_r} S^-_{z_r,z_s} & \approx & -\frac{1}{4}I, \nonumber\\
-  S^+_{z_s,z_r} V^-_{z_r,z_s} & \approx & -\frac{1}{4}I.\\
-\end{eqnarray}.
-The second equation follows by an exactly analogous argument, and the
-last two by time reversal and interchange of the roles of $z_s$ and
-$z_r$.
-
-Having identified approximate two-sided inverses for $S^{\pm},
-V^{\pm}$, the approximate null space identity \ref{eqn:lamnull} leads
-to a relation between $\Lambda^{\pm}$ and ${\cal S}^{\pm}$ that
-provides the foundation for practical computations. Applying $\Pi_0$
-to both sides of \ref{eqn:lamnull} gives
-\[
- \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_0^T
- \Lambda^+_{z_s} = S^+_{z_s,z_r} \Lambda^+_{z_s}.
-\]
-From the first approximate identity in \ref{eqn:approxinv},
-\begin{equation}
-  \label{eqn:lamident}
-  V^{-}_{z_r,z_s}\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx -\frac{1}{4}\Lambda^+_{z_s}.
-\end{equation}
-This identity is the second major result of this section: it shows how
-to compute that action of $\Lambda^+_{z_s}$ by propagating the input
-pressure trace, identified as a source for the velocity evolution,
-forward in time from $z_s$ to $z_r$
-reading off the pressure trace on $z=z_r$, identifying it once more as
-a source for velocity, propagating it backwards in time from $z_r$ to
-$z_s$, and finally reading off the velocity trace, interpreted as a
-pressure evolution source on $z_s$. I will give a detailed description
-of this computation in the section on numerical results.
-
-The importance of this result lies in the failure of the obvious
-method for computing the action of $\Lambda^{\pm}_{z_s}$, namely to
-employ the pressure trace as a source in the velocity equation ($f_s$,
-in the notation used above) at $z=z_s$, and read off the velocity
-field also at $z=z_s$. This difficulty is related to the existence of
-tangentially propagating waves and the lack of continuity of the trace
-operator. The method implicit in equation \ref{eqn:lamident} avoids
-this difficulty by propagating the fields a positive distance in $z$:
-assuming as always that the causal fields are downgoing, this step
-eliminates any tangentially propagating fields from consideration.
-
-Computation of the transpose of $\Lambda^+$ (exact, not approximate in
-the high frequency sense) will be important in the
-construction of a preconditioner. The relation \ref{eqn:lamident} also
-provides an computation for this operator. The idea is to treat the
-left-hand side of equation \ref{eqn:lamident} as an approximation to
-$\Lambda$, and compute its exact transposes via the adjoint state
-method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
-\begin{equation}
-  \label{eqn:lamtransp}
-   \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r} \approx -\frac{1}{4}(\Lambda^+_{z_s})^T.
-\end{equation} 
-The left-hand sides of \ref{eqn:lamident} and \ref{eqn:lamtransp} will
-simply be substituted for $\Lambda$ and its transpose in the
-preconditioner construction: preconditioners must be (precisely)
-self-adjoint, but need only be approximated to perform well.
-
 \section{Approximate Unitarity}

### 2021-05-24 repro/papers/sseprecond/oldstuff.tex
interim

--- 2021-05-18 repro/papers/sseprecond/paper.tex
@@
-\title{Efficient Computation of Extended Sources}
-\author{William. W. Symes \thanks{The Rice Inversion Project,
-Department of Computational and Applied Mathematics, Rice University,
-Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}
-
-\lefthead{Symes}
-
-\righthead{Approximate Source Inversion}
-
-\maketitle
-\begin{abstract}
-Source extension is a reformulation of inverse problems in wave propagation, that at least in some cases leads to computationally tractable iterative solution methods.
-The core subproblem in all source extension methods is the solution of a linear inverse problem for a source (right hand side in a system of wave equations) through minimization of data error in the least squares sense with soft imposition of physical constraints on the source via an additive quadratic penalty.
-A variant of the time reversal method from photoacoustic tomography provides an approximate solution that can be used to precondition Krylov space iteration for rapid convergence to the solution of this subproblem.
-An acoustic 2D example for sources supported on a surface, with a soft contraint enforcing point support, illustrates the effectiveness of this preconditioner.
-\end{abstract}
-
-\section{Introduction}
-Full Waveform Inversion (FWI) can be described in terms of 
-\begin{enumerate}
-\item a linear wave operator $L[{\bf c}]$, depending on a vector of
-  space-dependent coefficients ${\bf c}$ and acting on causal vector wavefields $\bu$ vanishing in negative time:
-\begin{equation}
-\label{eqn:init}
-\bu \equiv 0, t \ll 0; 
-\end{equation}
-\item a trace sampling operator $P$ acting on wavefields and producing data traces;
-\item and a (vector) source function (of space and time) $\bff$ representing energy input to the system. 
-\end{enumerate}
-The basic FWI problem is: given data $d$, find ${\bf c}$ so that 
-\begin{equation}
-\label{eqn:fwi}
-P\bu \approx d \mbox{ and } L[\bf{c}]\bu = \bff.
-\end{equation}
-In this formulation, the source function $\bff$ may be given, or
-to be determined subject to some constraints.
-
-%The energy source $\bff$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time.
-In fact, additional source degrees of freedom, beyond those needed to describe physically realized sources, may be useful in rendering the FWI problem \ref{eqn:fwi} more amenable to numerical solution, via so-called extended modeling (see \cite{geoprosp:2008}, \cite{LeeuwenHerrmannWRI:13}, \cite{HuangNammourSymesDollizal:SEG19}, and many references cited there).
-Therefore it is natural to view $\bff$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
-A simple nonlinear least squares formulation is:
-\begin{equation}
-\label{eqn:ols}
-\mbox{choose } {\bf c} \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2.
-\end{equation}
-Practical optimization formulations typically augment the objective in
-\ref{eqn:ols} by additive penalties or other constraints.
-
-As is well-known, local optimization methods are the only feasible
-approach given the dimensions of a typical instance of \ref{eqn:fwi},
-and those have a tendency to stall due to ``cycle-skipping''. See for
-exampe \cite{VirieuxOperto:09} and many references cited there. Source
-extension is one approach to avoiding this problem. It consists in
-imposing the wave equation as a soft as opposed to hard constraint, by
-allowing the source field $\bff$ to have more degrees of freedom than
-is permitted by a faithful model model of the seismic experiment, and
-constraining these additional degrees of freedom by means of an
-additive penalty modifying the probem \ref{eqn:ols}:
-\begin{equation}
-\label{eqn:esi}
-\mbox{choose } {\bf c}, \bff \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
-\end{equation}
-The operator $A$ penalizes deviation from known (or assumed)
-characteristics of the source function - its null space consists of
-feasible (or ``physical'') source models.
-
-\cite{HuangNammourSymesDollizal:SEG19} present an overview of the
-literature on source extension methods, describing a variety of
-methods to add degrees of freedom to physical source model. The present paper
-concerns {\em surface source extension}: physical sources are
-presumed to be concentrated at points $\bx_s$ in space, whereas their extended
-counterparts are permitted to spread energy over surfaces containing
-the physical source locations. A simple choice for the penalty
-operator $A$ is then multiplication by the distance $|\bx-\bx_s|$ to the physical
-source location:
-\begin{equation}
-  \label{eqn:penop}
-  (A\bff)(\bx,t) = |\bx-\bx_s|\bff(\bx,t)
-\end{equation}
-I shall use this choice of penalty operator whenever a specific choice
-is necessary in the development of the theory below.
-
-This paper presents a numerically efficient approach to solving the
-{\em source subproblem} of problem \ref{eqn:esi}:
-\begin{equation}
-\label{eqn:esis}
-\mbox{given } {\bf c}, \mbox{ choose } \bff \mbox{ to minimize }
-\|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha \|A\bff\|^2 
-\end{equation}
-Solution of this subproblem is an essential component of {\em variable
-  projection} algorithms for solution of the nonlinear inverse problem
-\ref{eqn:esi}. Variable projection is not merely a convenient choice
-of algorithm for this purpose: it is in some sense essential, see for
-example \cite{Symes:SEG20}. It replaces the nonlinear
-least squares problem \ref{eqn:esi} with a {\em reduced} problem, to
-be solved iteratively. Each iteration involves solution of the
-subproblem \ref{eqn:esis}. Therefore efficient solution of the
-subproblem is essential to efficient solution of the nonlinear problem
-via variable projection.
-
-The penalty operator $A$ defined in \ref{eqn:penop} is linear, so the source
-subproblem is a linear least squares problem. Under some additional
-assumptions to be described below, I shall show how to construct an
-accurate approximate solution operator for problem
-\ref{eqn:esis}. This approximate solution operator may be used to
-accelerate Krylov space methods for the solution of the source
-subproblem \ref{eqn:esis}. Numerical examples suggest the
-effectiveness of this acceleration.
-
-I will fully describe the preconditioner construction for a special
-case of the source subproblem \ref{eqn:esis}, in which ${\bf u}$ is an
-acoustic field, $L[{\bf c}]$ is the wave operator of linear
-acoustodynamics, the spatial positions of traces extracted by $P$ lie
-on a plane, and the positions at which the extended
-source $\bff$ is nonzero lie on another, parallel, plane. The data and
-sources are further limited to pressure traces and constitutive law
-defects (``pressure sources''). While this
-``crosswell'' configuration simplifies the analysis underlying the
-construction of approximate solutions for the source subproblem
-\ref{eqn:esis}, it is only one of many transmission configurations for
-which similar developments are possible. Perhaps the most important
-alternative example is the diving wave configuration, which plays a
-central role in contemporary FWI. I will discuss the generalization of
-surface source extention to diving wave inversion at the end of the
-paper.
-
-The preconditioner construction is very similar to the time-reversal
-method in photoacoustic tomography
-\cite[]{StefanovUhlmannIP:09}. Preconditioning amounts to a change of
-norm in the domain and range spaces of the modeling operator. In this
-case, the modfied norms are weighted $L^2$, and the weight operators
-map pressure to corresponding surface source on the source and
-receiver planes. These are essentially the same as the ``hyperbolic
-Dirichlet-to-Neumann'' map that plays a prominent role in
-thermoacoustic tomography and other wave inverse problems
-\cite[]{Rachele:00,StefUhl:05}. The obvious computation of these maps
-- prescribe the pressure, solve the wave equation with this boundary
-condition, read off the equivalent source - suffers from intrinsic
-numerical inaccuracy. I suggest an alternative computationally
-feasible approach.
-
-\section{Operators}
-
-For acoustic wave physics, the coefficient vector is
-$\bf{c}=(\kappa,\rho)^T$, with components bulk modulus $\kappa$ and
-density $\rho$, and the state vector $\bu=(p,\bv)^T$ consists of
-pressure $p$ (a scalar space-time field) and particle velocity $\bv$
-(a vector space-time field). The wave operator $L[\bf{c}]$ is:
-\begin{equation}
-\label{eqn:aweop}
-L[\bf{c}]\bf{u} = 
-\left(
-\begin{array}{c}
-\frac{1}{\kappa}\frac{\partial p}{\partial t}  + \nabla \cdot \bv, \\
-\rho\frac{\partial \bv}{\partial t} + \nabla p.
-\end{array}
-\right) 
-\end{equation}
-That is,
-\begin{equation}
-  \label{eqn:awemat}
-  L[{\bf c}] = \left(
-    \begin{array}{cc}
-      \frac{1}{\kappa}\frac{\partial}{\partial t} & \nabla \cdot \\
-      \nabla & \rho \frac{\partial}{\partial t}
-    \end{array}
-  \right)
-\end{equation}
-$L[{\bf c}]$ has a well-defined inverse if it is restricted to either
-causal or anti-causal vector wavefields.
-
-Most of what follows is valid for any space dimension $n >0$. The
-coefficient vector $\bf{c}=(\kappa,\rho)$ is defined throughout space
-$\bR^n$, the state vector $\bu$ throughout space-time
-$\bR^{n+1}$. Whenever convenient for mathematical manipulations,
-$n=3$: for instance, I will write $\bx=(x,y,z)^T$ for the spatial
-coordinate vector, and refer to the third (vertical) coordinate of
-particle velocity as $v_z$. Examples
-later in this paper will use $n=2$ for computational convenience.
-
-Since all of the operators in the discussion that follows depend on
-the coefficient vector 
-$\bf{c}$, I will suppress it from the notation, for example, $L=L[\bf{c}]$.
-
-The surface source extension replaces point sources on or near a
-surface in $\bR^3$ with source functions confined to the same
-surface. The simplest example of this extended geometry specifies a
-plane $\{(x,y,z,t): z=z_s\}$ at source depth $z_s$ as the surface. For
-acoustic modeling, surface sources are combinations of constitutive law
-defects and loads normal to the surface, localized on $z=z_s$. That
-is, right-hand sides in the system $L\bu=\bff$ take the form
-$\bff(\bx,t) = (h_s(x,y,t)\delta(z-z_s),
-f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar defect $h_s$ and normal
-force $f_s$ ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in
-\ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$
-takes the form
-\begin{eqnarray}
-\label{eqn:awepm}
-\frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} & = & - \nabla \cdot \bv^{\pm} +
-h_s \delta(z-z_s), \nonumber \\
-\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
-p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
-\bv^{\pm} & = & 0 \mbox{ for } \pm t \ll 0.
-\end{eqnarray}
-
-\noindent {\bf Remark:} In system \ref{eqn:awepm} and many similar
-systems to follow, I will use the shorthand
-\[
-  p^+ = 0 \mbox{ for } t \ll 0 
-\]
-to mean that $p^+$ is {\em causal}, that is,
-\[
-  \mbox{For some } T \in \bR, p^+(\cdot,t) = 0 \mbox{ for all } t <
-  T.
-\]
-Similarly,
-\[
-  p^- = 0 \mbox{ for } t \gg 0 
-\]
-signifies that $p^-$ is anti-causal.
-
-Extended forward modeling consists in solving \ref{eqn:awepm} and
-sampling the solution components at receiver locations. For
-simplicity, throughout this paper I will assume that the receivers are
-located on another spatial hyperplane $\{(x,y,z,t): z=z_r\}$ at
-receiver depth $z_r>z_s$. The constructions to follow involve interchange
-of the roles of $z_s$ and $z_r$ (that is, locating sources on $z=z_r$
-and receivers at $z=z_s$), so rather than the sampling operator $P$ of
-the introduction, I will denote by $P_s,P_r$ the sampling 
-operators on $z=z_s$, $z=z_r$ respectively. In practice, sampling
-occurs at a discrete array of points (trace locations) on these
-surfaces, and over a zone of finite extent. In this theoretical
-discussion, I will neglect both finite sample rate and extent, and
-regard the data, for example $P_rp^+$, as continuously sampled and
-extending over the entire plane $z=z_r$.
-
-A technical difficulty with spatial sampling must be
-addressed. Acoustic field energy is defined in terms of the mean
-squares of the pressure and particle velocity fields. However fields
-with finite mean square do not in general have well-defined
-restrictions to lower-dimensional sets: for example, the pressure
-field of a finite-energy acoustic vector field may not have a
-well-defined restriction to the space-time plane $z=z_r$. This
-phenomenon is related to the ill-posedness of wave equations as
-evolution equations in spatial variables, an observation attributed to
-Hadamard (see \cite{CourHil:62}, Chapter 6, section 17, also
-\cite{Payn:75,Symes:83}). Some constraint on the acoustic field,
-beyond finite energy, is mandatory in formulating the inverse problems
-\ref{eqn:esi} and \ref{eqn:esis}. In fact, the natural constraint in
-the ``crosswell'' geometry of this paper is that high-frequency energy
-{\em not} travel along rays parallel or nearly parallel to the
-surfaces $z=z_s, z=z_r$. I will call fields with this property {\em
-  downgoing} (even though the concept also encompasses {\em upcoming}
-fields). \cite{BaoSy:91b} give a mathematically complete discussion of
-downgoing wave field properties.
-
-For downgoing solutions of system \ref{eqn:awepm}, the key
-components ($p^{\pm}$ and $v^{\pm}_z$) are continuous functions of $z$
-in the open slab $z_s<z<z_r$ with well-defined limits at the boundary
-planes, but may be discontinuous at the source plane
-$z=z_s$. Similarly, the roles of $z_s$ and $z_r$ will be interchanged
-in some of the constructions to come, and the corresponding solutions
-may be discontinuous at $z=z_r$. Accordingly, interpret $P_s$, $P_r$
-as the limit from right and left respectively: for $u=p^{\pm}$ or
-$v^{\pm}_z$,
-\begin{eqnarray}
-  \label{eqn:defsamp}
-  P_su(x,y,t) &=& \lim_{z \rightarrow z_s^+} u(x,y,z,t),\nonumber \\
-  P_ru(x,y,t) &=& \lim_{z \rightarrow z_r^-} u(x,y,z,t).                  
-\end{eqnarray}
-
-The causal/anti-causal vector
-modeling operators ${\cal S}^{\pm}_{z_s,z_r}$ are defined in terms of
-the solutions $(p^{\pm},\bv^{\pm})$ of the systems \ref{eqn:awepm} by
-\begin{equation}
-  {\cal S}^{\pm}_{z_s,z_r}(h_s,f_s)^T  = (P_rp^{\pm},P_r v_z^{\pm})^T,
-  \label{eqn:fwd}
-\end{equation}
-The subscript signifies that sources are located on $z=z_s$, the
-receivers on $z=z_r$. It is necessary to include this information in
-the notation, as versions of ${\cal S}^{\pm}$ with sources and receivers in
-several locations will be needed in the discussion below.
-
-\noindent {\bf Remark:} To connect with the formulation presented in
-the introduction, note that for continuous $u$,
-$P_su(x,y,t)=u(x,y,z_s,t)$, and therefore the adjoint of $P_s$ (in the
-sense of distributions) is $P_s^Th(x,y,z,t) =
-h(x,y,t)\delta(z-z_s)$. Write ${\cal P}_s = \mbox{diag }(P_s,P_s)$ and
-similarly for ${\cal P}_r$. Then
-\[
-  {\cal S}^{+}_{z_s,z_r} = {\cal P}_r L^{-1}({\cal P}_s)^T,
-\]
-in which $L^{-1}$ is interpreted in the causal sense, and similarly
-for ${\cal S}^{-}$. Sources confined to $z=z_s$ are precisely those
-functions (distributions, really) output by ${\cal P}_s^T$, so the
-problem statements \ref{eqn:esi} and \ref{eqn:esis} can be rewritten
-in terms of ${\cal S}^+_{z_s,z_r}$, with $P$ identified with ${\cal P}_r$.
-
-${\cal S}^{\pm}$ is not stably invertible, in fact having rank 1 (in
-an approximate sense - see the identity \ref{eqn:lamnull}). The
-diagonal components of ${\cal S}^{\pm}$ thus carry essentially all of
-its information, and it is in terms of these that a sensible inverse problem
-is defined.
-
-Denote by $\Pi_i, i=0,1$ the projection on the first,
-respectively second, component of a vector in $\bR^2$. The 
-forward modeling operator from pressure source to pressure trace is
-\begin{equation}
-  \label{eqn:sdef}
-  S^{\pm}_{z_s,z_r} = \Pi_0 {\cal S}^{\pm}_{z_s,z_r} \Pi_0^T 
-\end{equation}
-and the forward modeling operator from velocity source (normal force)
-to velocity trace is
-\begin{equation}
-  \label{eqn:vdef}
-  V^{\pm}_{z_s,z_r} = \Pi_1 {\cal S}^{\pm}_{z_s,z_r} \Pi_1^T 
-\end{equation}
-
-With these conventions, we can write the version of the source
-subproblem studied in this paper as
-\begin{equation}
-  \label{eqn:esisp}
-  \mbox{find }h_s\mbox{ to minimize }\|S^{+}_{z_s,z_r}h_s- d\|^2 +
-  \alpha^2\|Ah_s\|^2.
-\end{equation}
-
-\section{2D Examples}
-
-\inputdir{project}
-
-To illustrate the structure described in the preceding section, I
-introduce two 2D acoustic models, one spatially homogeneous, the other
-with highly refractive. The first, homogenous model has $\kappa = 4$
-GPa and $\rho = 1$ g/cm$^3$ throughout a rectangular domain of size 8 km ($x$) $\times$ 4 km
-($z$). The second, refractive, model is a perturbation of the first by
-a low-velocity acoustic lens positioned in the center of the
-rectangle. To produce this structure, the density is chosen
-homogeneous as in the first model, while the bulk modulus decreases to
-from 4 GPa outside the lens to 1.6 GPA in its center, as shown in Figure \ref{fig:bml0}.
-
-Discretization is conventional, with a rectangular grid and staggered
-finite difference scheme \cite[]{Virieux:84} of order 2 in time and
-2$k$ in space; for most of the experiments reported below, $k=4$.
-Sampling operators such as $P_r$ are implemented via linear
-interpolation, and source insertion via adjoint linear interpolation
-(as noted above, in the continuum limit, sources are represented via
-adjoint sampling). Steps in $x$ and $z$ are the same. Most results use
-$\Delta x = 20$ m, and limit the temporal frequency of the computed
-traces to 12.5 Hz. We also explore the dependence of a few results on frequency, using
-$\Delta x = 10$ m and $5$ m, to accomodate 25 and 50 Hz respectively,
-maintaining 8 samples per wavelength. All computations are carried out
-in single precision.
-
-\cite{GeoPros:11} gives a description of the code
-implementation, out-of-date in a few respects but overall accurate.
-
-The horizontal line of receivers sits at depth $z_r = $ 1000 m, as
-shown in Figure \ref{fig:bml0}. Receiver $x$ ranges from $2000$ to
-$6000$ m. A single point (physical) source appears in these
-experiments, located at $x_s=3500$ m, $z_s=3000$ m, as also indicated
-in the figure. The source time function is a bandpass filter with
-corner frequencies $1, 2.5, 7.5, 12.5$ Hz for the lowest frequency
-source, and scaled as appropriate for examples with higher frequency
-and finer sampling.  Extended sources are confined to the horizontal
-line through the physical source position, that is $z_s = 3000$ m,
-over a 4 km interval starting at $x_r=$ 2000 m. Note that we have
-reversed the order relation between $z_s$ and $z_r$ described in the
-text ($z_s<z_r$). This difference is immaterial.
-
-\plot{bml0}{width=\textwidth}{Bulk modulus, lens model. Color scale is
-in GPa. Positions of point source and receiver line indicated.}
-
-The point source pressure data generated by this configuration for the
-homogeneous model is displayed in Figure \ref{fig:recphh0}, for the
-lens model in \ref{fig:recplh0}. Triplication of arrivals is evident in the latter.
-
-\plot{recphh0}{width=\textwidth}{Output pressure gather for point source in 
-  homogeneous medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-\plot{recplh0}{width=\textwidth}{Output pressure gather for point source in 
-  refractive (lens) medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-While inversion of pressure data is the main object of this exercise,
-normal velocity ($v_z$) data will play an important role, so I display
-the corresponding gathers in Figures \ref{fig:recvzhh0} and \ref{fig:recvzlh0}.
-
-\plot{recvzhh0}{width=\textwidth}{Output vertical velocity gather for point source in 
-  homogeneous medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-\plot{recvzlh0}{width=\textwidth}{Output vertical velocity gather for point source in 
-  refractive (lens) medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-\section{Adjoints}
-
-It follows from the adjoint state method (see Derivations section
-below for details) that
-\begin{equation}
-  \label{eqn:sadj1}
-  ({\cal S}^{\pm}_{z_s, z_r})^T = -{\cal S}^{\mp}_{z_r,z_s}
-\end{equation}
-
-Define $R$ to be the {\em time-reversal operator} on functions of
-space-time, $Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em
-  acoustic field time-reversal operator}
-\begin{equation}
-  \label{eqn:trdef}
-  {\cal R} \left(
-    \begin{array}{c}
-      p\\
-      \bv
-    \end{array}
-  \right) =
-  \left(
-    \begin{array}{c}
-      Rp\\
-      -R\bv
-    \end{array}
-  \right)
-\end{equation}
-Then 
-\begin{equation}
-  \label{eqn:trsadj}
-  {\cal R}{\cal S}^{\mp} = -{\cal S}^{\pm}_{z_r,z_s}{\cal R}
-\end{equation}
-Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that
-
-\begin{equation} 
-  \label{eqn:trtr}
- ({\cal S}^{\pm}_{z_s,z_r})^T = {\cal R}{\cal S}_{z_r,z_s}^{\pm}{\cal R}=
- -{\cal S}^{\mp}_{z_r,z_s}.
-\end{equation}
-
-The relation \ref{eqn:trtr} implies that
-\begin{eqnarray}
-  (S^{\pm}_{z_s,z_r})^T &=& -S^{\mp}_{z_r,z_s} \nonumber\\
-                        &=& R S^{\pm}_{z_r,z_s}R, \nonumber\\
-    (V^{\pm}_{z_s,z_r})^T &=& -V^{\mp}_{z_r,z_s} \nonumber\\
-                        &=& R V^{\pm}_{z_r,z_s}R.
-                            \label{eqn:trtrcomp}
-\end{eqnarray}
-
-The implemented code uses the adjoint state method on the discrete
-level, to assure that the computed adjoint operators are adjoint at
-the level of machine precision to the computed operators. The issue of
-reverse-time storage is resolved through the optimal checkpointing
-technique \cite[]{Griewank:book,Symes:06a-pub}, again without loss of
-precision.
-
-Typical results with pseudorandom input traces $d_r, h_z, w_r, f_s$ are:
-\begin{itemize}
-\item computed $\langle d_r, S^+_{z_s,z_r}h_s\rangle = -2.41069174,
-  \langle (S^+_{z_s,z_r})^Td_r, h_s \rangle = -2.41069388.$
-\item  computed $\langle w_r, V^+_{z_s,z_r}f_s\rangle = -2.73362470,
-  \langle (V^+_{z_s,z_r})^Tw_r, f_s \rangle = -2.73362136.$
-\end{itemize}
-Ideally, the differences of these inner products should be at most a
-relatively small multiple of machine precision, {\em relative} to the
-products $\|d_r\|\| S^+_{z_s,z_r}\|\|h_s\|$ and
-$\|w_r\|\|V^+_{z_s,z_r}\|\|f_s\|$ (thus making the result
-dimensionless). The operator norm $\| S^+_{z_s,z_r}\|$ is
-computationally inaccessible, so instead I used the smaller quantities
-$\|d_r\|\| S^+_{z_s,z_r}h_s\|$ etc. as stand-ins - thereby
-overestimating the relative error between the inner products. In all
-cases, over a very large number of random inputs, the largest observed
-relative error estimate was $O(10^{-9}$, well under the appropriate limit for
-single precision.
-
-\section{Approximate Inversion}
-
-\section{Iterative Inversion}
-
 \section{Fields, Traces and Sources}
@@
 \section{Parametrix via Time Reversal}
-
-Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
-efficient iterative solution of the linear subproblem
-\ref{eqn:esis}. Direct computation, for instance by solving
-\ref{eqn:sawe2} and reading off $P_sv^{\pm}_{2,z}$, turns out to be
-numerically ill-behaved. An accurate computation is available via the
-computation of a {\em parametrix} (high frequency approximate inverse)
-for $S^{\pm}_{z_s,z_r}$. This parametrix also justifies the use of the
-symmetric part of $\Lambda^{\pm}_{z_s}$ as a preconditioner. In this
-section I will explain the parametrix construction, and also show how
-to extract the symmetric part of $\Lambda^{\pm}_{z_s}$.
-
-The key ingredient in this {\em time reversal} construction is a
-{\em local energy decay} assumption. Define the local energy in the
-acoustic field $(p,\bv)$ at
-time $t$ in a region $\Omega \subset \bR^3$ by imiting the integration
-in the definition of total energy \ref{eqn:defae0} to $\Omega$:
-\begin{equation}
-  \label{eqn:defaeloc}
-  E_{\Omega}[(p,\bv)](t) = \frac{1}{2} \int_{\Omega} \,d\bx \, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx) |\bv(\bx,t)|^2\right)
-\end{equation}
-
-We assume that $(p^+,\bv^+)$ is {\em downgoing} in the half-space
-$\Omega = \{(x,y,z): z \le z_r\}$: that is, there exists a solution of
-the acoustic system $(\tilde{p}^+,\tilde{\bv}^+)$ in space-time $\bR^4$
-for which
-\begin{itemize}
-\item[1. ] $p^+ \approx \tilde{p}^+, \bv^+ \approx \tilde{\bv}^+$ in
-  the slab $\Gamma = \{(x,y,z): z_s \le z \le z_r\}$, and
-\item[2. ] The energy in $(\tilde{p}^+,\tilde{\bv}^+)$ decays in
-  forward time, in the half-space $\Omega$:  $E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](t) \rightarrow 0$
-  as $t \rightarrow \infty$.
-\end{itemize}
-
-Conditions for local energy decay have been widely studied in various
-settings \cite[]{MorawetzPhillipsLax:05,Hristova:09}. For the half-space
-geometries considered here, local energy does not decay in general,
-because of the existence of ``trapped'' rays traveling parallel to the
-boundary. Instead, high frequency data associated with rays crossing
-the boundary $\{z=z_r\}$ transversally, and never returning to the slab, have
-the local energy decay property.
-
-The downgoing constraint has mostly been studied in relation to
-seismology, mostly as a ray-theoretic concept: downgoing fields are
-concentrated in phase space along rays for which depth $z$ is
-increasing with $t$. The version above is stated without reference to
-ray theory, but the setting of this paper allows its use and in a
-later section I will explicitly construct the downgoing approximation
-under appropriate ray-theoretic conditions. The numerical examples will
-illustrate the necessary assumptions. When ray-theory is appropriate,
-the downgoing constraint can be {\em ex post facto} enforced via dip
-filtering. For example, the application of the ideas developed here to
-diving wave inversion will require such a filtering step.
-
-An immediate consequence of the energy decay assumption is that in the
-slab $\Gamma$, the field $(p^+,\bv^+)$ approximates the solution of an
-anti-causal evolution equation. For any choice of $T$, define
-$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
-$\Omega$ of
-\begin{eqnarray}
-\label{eqn:revawe}
-  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
-  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
-  \tilde{p}^- & =& 0,  \mbox{ for } t=T,\nonumber\\ 
-  \tilde{\bv}^- & = & 0 \mbox{ for } t=T,\nonumber\\
-  P_r\tilde{p}^- &=& P_rp^+ \mbox { for } t \le T . 
-\end{eqnarray}
-As will be shown later, under the downgoing assumtion, $P_r p^+
-\approx P_r \tilde{p}^+$.
-Therefore $(\delta p,\delta \bv)=(\tilde{p}^+-\tilde{p}^-,\tilde{\bv}^+-\tilde{\bv}^-)$
-differs negligibly from the solution of
-\begin{eqnarray}
-\label{eqn:deltaawe}
-  \frac{1}{\kappa}\frac{\partial \delta p}{\partial t} & = & -
-                                                               \nabla
-                                                               \cdot
-                                                               \delta \bv, \nonumber \\
-  \rho\frac{\partial \delta \bv}{\partial t} & = & - \nabla \delta p,\nonumber \\
-  \delta p & =& \tilde{p}^+,  \mbox{ for } t=T,\nonumber\\ 
-  \delta \bv & = & \tilde{\bv}^+ \mbox{ for } t=T,\nonumber\\
-  P_r\delta p & = & 0,
-\end{eqnarray}
-in $\Omega \times \bR$. Since
-$E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](T) \rightarrow 0$ as $T
-\rightarrow \infty$, the usual energy estimate implies that the energy $\delta
-E$ in this field,
-\[
-  \delta E_{\Omega}(t) = \frac{1}{2} \int_{\Omega} \,d\bx \,
- \left(\frac{(\delta p)^2}{\kappa} + \rho |\delta \bv|^2\right)
-\]
-satisfies $\delta E_{\Omega}(t) \rightarrow 0$ as $T \rightarrow
-\infty$. This decay does not generally imply that $P_s\delta p, P_s \delta \bv)
-\rightarrow 0$ (for instance) as $T \rightarrow \infty$, since the trace operator
-$P_s$ is not continuous, but it does so imply for purely downgoing
-fields, as will be demonstrated later, so
-\begin{equation}
-  \label{eqn:approxinv0}
-  \lim_{T \rightarrow \infty} P_s\tilde{v}_z^- \approx \lim_{z \rightarrow
-    0^+}\tilde{v}_z^+ \approx \lim_{z \rightarrow 0^+}v^+_z
-\end{equation}
-
-Recall (equations \ref{eqn:deflam}, \ref{eqn:jump0}) that
-\[
-  \Lambda^+_{z_s} P_s p^+ \approx 2 \lim_{z \rightarrow z_s^+}v^+_z 
-\]
-and that with $h_s= \Lambda^+_{z_s}P_sp^+$,
-\[
-  S^+_{z_s,z_r}h_s = P_r p^+
-\]
-Together with the last two equations, the relation
-\ref{eqn:approxinv0} implies that
-\begin{equation}
-  \label{eqn:approxinv1}
-  S^+_{z_s,z_r}(2 P_s \tilde{v}_z^-) \approx P_rp^+.
-\end{equation}
-As shown in the last section, the solution of the system
-\ref{eqn:revawe} differs negligibly from the anti-causal solution $(p^-_2,\bv^-_2)$
-of the system \ref{eqn:sawe2} with (i) $z_s$ replaced by $z_r$, and
-(ii) $\psi = [p^-_2]|_{z=z_r} = -2 \lim_{z \rightarrow z_r^-} p^-_2
-\approx -2 P_rp^+$. Therefore (definitions \ref{eqn:sdef}, \ref{eqn:vdef})
-\begin{equation}
-  \label{eqn:approxinv2}
-  P_s \tilde{v}_z^- \approx V^-_{z_r,z_s}(-2 P_rp^+).
-\end{equation}
-Taken together, relations \ref{eqn:approxinv1} and
-\ref{eqn:approxinv2} yield the first equation in the first main result of this section:
-\begin{eqnarray}
-  \label{eqn:approxinv}
-  V^-_{z_r,z_s} S^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
-  S^-_{z_r,z_s} V^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
-  V^+_{z_s,z_r} S^-_{z_r,z_s} & \approx & -\frac{1}{4}I, \nonumber\\
-  S^+_{z_s,z_r} V^-_{z_r,z_s} & \approx & -\frac{1}{4}I.\\
-\end{eqnarray}.
-The second equation follows by an exactly analogous argument, and the
-last two by time reversal and interchange of the roles of $z_s$ and
-$z_r$.
-
-Having identified approximate two-sided inverses for $S^{\pm},
-V^{\pm}$, the approximate null space identity \ref{eqn:lamnull} leads
-to a relation between $\Lambda^{\pm}$ and ${\cal S}^{\pm}$ that
-provides the foundation for practical computations. Applying $\Pi_0$
-to both sides of \ref{eqn:lamnull} gives
-\[
- \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_0^T
- \Lambda^+_{z_s} = S^+_{z_s,z_r} \Lambda^+_{z_s}.
-\]
-From the first approximate identity in \ref{eqn:approxinv},
-\begin{equation}
-  \label{eqn:lamident}
-  V^{-}_{z_r,z_s}\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx -\frac{1}{4}\Lambda^+_{z_s}.
-\end{equation}
-This identity is the second major result of this section: it shows how
-to compute that action of $\Lambda^+_{z_s}$ by propagating the input
-pressure trace, identified as a source for the velocity evolution,
-forward in time from $z_s$ to $z_r$
-reading off the pressure trace on $z=z_r$, identifying it once more as
-a source for velocity, propagating it backwards in time from $z_r$ to
-$z_s$, and finally reading off the velocity trace, interpreted as a
-pressure evolution source on $z_s$. I will give a detailed description
-of this computation in the section on numerical results.
-
-The importance of this result lies in the failure of the obvious
-method for computing the action of $\Lambda^{\pm}_{z_s}$, namely to
-employ the pressure trace as a source in the velocity equation ($f_s$,
-in the notation used above) at $z=z_s$, and read off the velocity
-field also at $z=z_s$. This difficulty is related to the existence of
-tangentially propagating waves and the lack of continuity of the trace
-operator. The method implicit in equation \ref{eqn:lamident} avoids
-this difficulty by propagating the fields a positive distance in $z$:
-assuming as always that the causal fields are downgoing, this step
-eliminates any tangentially propagating fields from consideration.
-
-Computation of the transpose of $\Lambda^+$ (exact, not approximate in
-the high frequency sense) will be important in the
-construction of a preconditioner. The relation \ref{eqn:lamident} also
-provides an computation for this operator. The idea is to treat the
-left-hand side of equation \ref{eqn:lamident} as an approximation to
-$\Lambda$, and compute its exact transposes via the adjoint state
-method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
-\begin{equation}
-  \label{eqn:lamtransp}
-   \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r} \approx -\frac{1}{4}(\Lambda^+_{z_s})^T.
-\end{equation} 
-The left-hand sides of \ref{eqn:lamident} and \ref{eqn:lamtransp} will
-simply be substituted for $\Lambda$ and its transpose in the
-preconditioner construction: preconditioners must be (precisely)
-self-adjoint, but need only be approximated to perform well.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+In this section I will explain how to build an approximate inverse.
+The key ingredient in this {\em time reversal} construction is a
+{\em local energy decay} assumption. Define the local energy in the
+acoustic field $(p,\bv)$ at
+time $t$ in a region $\Omega \subset \bR^3$ by limiting the integration
+in the definition of total energy \ref{eqn:defae0} to $\Omega$:
+\begin{equation}
+  \label{eqn:defaeloc}
+  E_{\Omega}[(p,\bv)](t) = \frac{1}{2} \int_{\Omega} \,d\bx \, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx) |\bv(\bx,t)|^2\right)
+\end{equation}
+The necessary assumption is: local energy decay in the half-space
+$\Omega = \{(x,y,z): z \le z_r\}$. The local energy in $\Omega$ of the
+causal solution $(p^+,\bv^+)$ of system \ref{eqn:awepm} cannot decay,
+since the source on $z=z_s<z_r$ radiates in both positive and negative
+$z$ in forward time. That obstacle is resolved by a redefinition of
+the ``downgoing'' concept. I will call  downgoing if there is a
+source-free acoustic field $(\tilde{p}^+,\tilde{\bv}^+)$ in all of space-time for which
+\begin{itemize}
+\item[1. ] $p^+ \approx \tilde{p}^+, \bv^+ \approx \tilde{\bv}^+$ in
+  the slab $\{(x,y,z,t): z_s \le z \le z_r\}$, and
+\item[2. ] The energy in $(\tilde{p}^+,\tilde{\bv}^+)$ decays in
+  forward time, in the half-space $\Omega$:  $E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](t) \rightarrow 0$
+  as $t \rightarrow \infty$.
+\end{itemize}
+Conditions for local energy decay have been widely studied in various
+settings, for instance obstacle scattering
+\cite[]{MorawetzPhillipsLax:05}. \cite{Pauen2000NonTr-6079 gives a
+comprehensive overview of this subject. For smooth (non-scattering)
+acoustic parameters, as is assumed in this paper, the crux of the
+matter is that rays carrying high-frequence energy exit the half-space
+$\Omega$ in finite time. For the half-space
+geometries considered here, local energy does not decay in general,
+because of the existence of ``trapped'' rays traveling parallel to the
+boundary and remaining interior to $\Omega$ for all time. If an
+acoustic field is downgoing in the usual sense, that is, its
+high-frequency energy is concentrated on rays that exit $\Omega$
+within some common maxiumum time, then the results cited above show
+that local energy in $\Omega$ decays (exponentially for $n=3$).
+
+Starting from this observation, it is possible to show that if the
+source vector $(h_s,f_s)^T$ has the property termed ``downgoing'' in
+preceding sections,
 \section{Efficient Preconditioned CG}
-
-\cite{HouSymes:EAGE16} demonstrated a very similar preconditioner for Least Squares Migration, also for its subsurface offset extension \cite[]{HouSymes:16}, motivated by \cite{tenKroode:12}.
-These constructions all involve the Dirichlet-to-Neumann operator.
-This concept also turns up in hidden form in the work of Yu Zhang and collaborators on true amplitude migration \cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,Zhang YuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.
-
-The discussion so far can be summarized as follows: an efficient pseudoinverse for $S[c]$ is the weighted adjoint operator
-\begin{equation}
-\label{eqn:wadj}
-S[c]^{\dagger} = W[c]_m^{-1}S[c]^TW[c]_d,
-\end{equation}
-in which 
-\begin{equation}
-\label{eqn:weights}
-W[c]_m = \kappa^{-1}\Lambda[c]^{-1}_s,\,\, W[c]_d = \kappa^{-1}\Lambda[c]_r
-\end{equation}
-Here ``$\kappa$'' means the operator ``multiply by $\kappa$''.
-
-The statement that $S[c]^{\dagger}$ is the weighted adjoint of $S[c]$ means that $S[c]^{\dagger}$ {\em is} the adjoint of $S[c]$ with respect to weighted norms 
-\begin{itemize}
-\item on the source space ($h_s$): weight $W[c]_m$
-\item on the data space ($d_r$): weight $W[c]_d$
-\end{itemize}
-Since $S[c]^{\dagger}S[c] \approx I$ in subspaces of the domain space, as shown above, $S[c]$ is approximately unitary in (subspaces of) the Hilbert spaces of source and data traces with norms defined by the weighting operators $W_m$ and $W_d$ respectively.
-Therefore a Krylov space method employing these norms will converge rapidly, at least for the well-determined components of the solution.
-
-The most convenient arrangement the CG algorithm taking advantage of the structure \ref{eqn:wadj} is the {\em Preconditioned CG}.
-Since the coefficient array $c$ will remain unchanged during the iteration, write $S$ instead of $S[c]$, etc.
-
-Allowing that the fit error will be measured by the data space norm, the least squares problem to be solved is not just $Sh \approx d$, but a regularized version:
-\begin{equation}
-  \label{eqn:einv}
-  \mbox{minimize}_h \|Sh-d\|^2_d + \alpha^2 \|Ah\|^2_m
-\end{equation}
-
-[The modified data space norm $\|d\|_d^2 = \langle d, W_d d\rangle$ has physical meaning: for acoustics, it is proportional to the power transmitted through the surface $z=z_r$.]
-
-The minimizer of the objective defined in equation \ref{eqn:einv} solves the normal equation
-\begin{equation}
-  \label{eqn:norm}
-  (S^{\dagger}S + \alpha^2 A^{\dagger}A)h = S^{\dagger}d 
-\end{equation}
-
-where the weighted adjoint $S^{\dagger}$ has already been defined in equation \ref{eqn:wadj}, and $A^{\dagger}$ is the adjoint of $A$ in the weighted model space norm defined by $W_m$, namely
-\begin{equation}
-  \label{eqn:aadj}
-  A^{\dagger} = W_m^{-1}A W_m.
-\end{equation}
-
-Introducing these definitions in the normal equation \ref{eqn:norm},
-\begin{equation}
-  \label{eqn:norm1}
-  W_m^{-1}(S^TW_dS + \alpha^2 A^TW_mA)h = W_m^{-1}S^TW_md 
-\end{equation}
-Since $W_m$ is self-adjoint and positive definite, the common factor on both sides of \ref{eqn:norm1} can be re-written as
-
-write $S^*, A^*$ for the adjoints with the original (Euclidean) inner product in the domains but the weighted inner product in data space:
-\begin{equation}
-  \label{eqn:normpart}
-  Nh = (S^*S + \alpha^2 A^*A)h = S^*d 
-\end{equation}
-using the ``partly weighted'' adjoints
-\begin{eqnarray}
-  \label{eqn:sadj}
-  S^* &=& S^T W_d,\\
-  A^* &=& A^T W_m.
-\end{eqnarray}
-The Preconditioned Conjugate Gradient (``PCG'') algorithm for solution of equation \ref{eqn:normpart} with preconditioner $W_m$ is usually written as Algorithm 1.
-
-\begin{algorithm}[H]
-\caption{Preconditioned Conjugate Gradient Algorithm, Standard Version}
-\begin{algorithmic}[1]
-\State Choose $h_0=0$ 
-  \State $r_0 \gets S^*d$
-  \State $p_0 \gets W_m^{-1}r_0$
-  \State $g_0 \gets p_0$
-  \State $q_0 \gets Np_0$
-  \State $k \gets 0$
-  \Repeat
-  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle p_k,q_k\rangle}$
-  \State $h_{k+1} \gets h_k + \alpha_k p_k$
-  \State $r_{k+1} \gets r_k - \alpha_kq_k$
-  \State $g_{k+1} \gets W_m^{-1}r_{k+1}$
-  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
-  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
-  \State $q_{k+1} \gets Np_{k+1}$
-  \State $k \gets k+1$
-  \Until{Error is sufficiently small, or max iteration count exceeded} 
-\end{algorithmic}
-\end{algorithm}
-
-The operator whose eigenvalue spectrum determines the speed of convergence here is the preconditioned normal operator
-\begin{equation}
-  \label{eqn:pno}
-  W_m^{-1}N = S^{\dagger}S +\alpha^2 A^{\dagger}A
-\end{equation}
-This isn't just an approximate identity on a subspace, as $S^{\dagger}S$ is.
-The addition of the penalty term affects the spectrum, pushing parts of it away from $1$.
-I have not yet analyzed this effect, however for the ``in-aperture'' part of the model space,
-\[
-  W_m^{-1}N \approx I +\alpha^2 W_m^{-1}A^TW_mA
-\]
-For small $\alpha$, this operator is a perturbation of the spectrum of the identity, so convergence should be rapid.
-
-In the preceding section, I showed that effective choices for the weight operators are $W_m = \Lambda_s^{-1}$ and $W_d = \Lambda_r$, which are resepctively the ``Neumann-to-Dirichlet'' operator mapping pressure to normal particle velocity on the source surface $\{z=z_s\}$, and the ``Dirichlet-to-Neumann'' operator mapping pressure to normal particle velocity on the receiver surface.
-Application of these operators are expensive and/or awkward, so should be avoided if at all possible.
-
-A very important point: the weighted ``partial'' adjoint $S^*=S^TW_d$ appears {\em only} in combination with $S$ in the algorithm listed above, {\em except} for the very first line.
-That is, one can calculate $W_dS$ followed by $S^T$, rather than $S$ followed by $S^TW_d$, provided that you can take care of the first line.
-$W_dS$ For any variant of elasticity, $W_d$ relates two combinations of dynamical field components - for acoustics, it returns the normal velocity for a given pressure field on the receiver surface.
-In the solution of the dynamical equations (for instance \ref{eqn:awe}), all components are calculated.
-Therefore obtaining the output of $W_d$ amounts to extracting the appropriate field combination.
-For acoustics, simply record the normal velocity rather than the pressure: then you have computed $W_dS$.
-Therefore no additional expensive computations are required.
-It is only the very first instance of $S^*$, on the first line, where an actual computation of $W_d$ is required, unless cheap workarounds like that explained by \cite{HouSymes:15} are available.
-
-\section{Prototype Numerical Examples}
-I present a collection of simple examples that illustrate the features of the surface source extension claimed in preceding sections.
-
-I used the IWAVE acoustic staggered grid package to carry out these calculations.
-This package implements (2,2k) schemes for k=1,2,..., and outputs traces (of either velocity or pressure at any point in space via multilinear interpolation.
-The discretized modeling operator is thus of second order accuracy, though as usual I have used higher order in space to reduce grid dispersion.
-
-Source injection is implemented as the adjoint of trace sampling, resulting in another second-order error [REFERENCES].
-
-The data is a single shot gather, with a source at coordinates $x_s=z_s=3000$ m (units of length are meters in all cases).
-The receiver line occupies $1500 \le x_r \le 5500$ m, with receiver depth $z_r=1000$ m.
-Extended sources occupy $1500 \le x_s \le 5500$ m, with the same depth $z_s=3000$ m as the ``physical'' source used to generate the data.
-This region turns out to be adequate to represent the extended sources that approximately invert the data, for the cases examined below.
-An algorithm to automatically identify an appropriate region can be based on the ideas developed by \cite{Fu:Geo17}.
-
-I have used absorbing boundary conditions (split-field PML) on all four sides of the 4000 m (vertical) $\times$ 8000 m (horizontal) simulation domain.
-Evidently inclusion of a free surface is important to the application of the ideas explained here to diving wave marine data, and I have not addressed the necessary modifications here.
-
-The Dirichlet-to-Neumann operator $\Lambda$ is an essential part of the inner problem preconditioner just presented. Implementation can be accomplished in several ways:
-\begin{itemize}
-\item \cite{tenKroode:12} suggests using a one-way operator;
-\item if both pressure and normal partical velocity are measured (or simulated), then the two are related by $\Lambda$ and the velocity component can simply be used as the output;
-\item presence of a free surface implies all of the usual problems, such as the need for removal of receiver-side ghosts.
-On the other hand, if the free surface is within a quarter-wavelength throughout the useful bandwidth of the data, then the ghosted data differs from $\Lambda p$ by a time integration and a scale factor, a fact used to good effect by \cite{HouSymes:15}.
-\end{itemize}
-
-In the examples, I have used the second observation.
-With a finite difference implementation of the pressure-velocity system \ref{eqn:awe}, velocity components are available ``for free'', short-circuiting explicit computation of $\Lambda$.
-
-The  {\tt project/SConstruct} script is set up to carry out the necessary computations on grids with spacings $\Delta x = \Delta z = $ 20, 10, and 5 m, with a jump of roughly 8 in computation time resulting from each refinement.
-For present purposes, the coarsest (20 m) grid seems to be sufficient, and that is the grid used in the examples presented below.
-The source pulses are chosen so that the computation is reasonably accurate.
-For the 20 m grid case, I use a zero-phase trapezoidal bandpass filter source with corner frequencies of 1.0, 2.0, 7.5, and 12.5 Hz.
-
-The IWAVE asg driver has been set up to recognize the case {\tt deriv=0} as defining the map from source (right-hand side in the pressure equation) to data (pressure) traces.
-The adjoint to this map, as explained above, is reverse-time propagation of the data traces as pressure sources, followed by scaling (formula \ref{eqn:sadj}).
-The approximate inverse is computed by application of the Dirichlet-to-Neumann operator to the pressure traces to produce corresponding velocity traces, followed by injection as pressure sources and reverse time propagation, followed by another application of the Dirichlet-to-Neumann map and scaling (formulas \ref{eqn:appinv}, \ref{eqn:adj}).
-
-The script implements these operatations step-by-step via calls to IWAVE, Madagascar, and SU commands.
-A peculiarity of the {\tt asg.x} driver needs to be mentioned: it is based on an un-scaled version of the constitutive law defect source representation, that is, {\tt asg.x} approximately computes the solution of the system \ref{eqn:awedata} with the first equation replaced by
-\begin{equation}
-\label{eqn:asgdata}
-\frac{\partial p}{\partial t}  =  - \kappa \nabla \cdot \bv +
-h \delta(z-z_s).
-\end{equation}
-Denote by $S_{\rm asg}[c]$ the forward map produced by {\tt asg.x}. Then comparison of \ref{eqn:awedata} and \ref{eqn:asgdata} reveals that
-\begin{equation}
-\label{eqn:sreln}
-S[c]=S_{\rm asg}[c]\kappa
-\end{equation}
-where $\kappa$ is shorthand for the operator of multiplication by $\kappa$. Accordingly, and approximate inverse for $S_{\rm asg}[c]$ is
-\[
-I \approx S_{\rm asg}[c]^{\dagger}S_{\rm asg}[c] = S_{\rm asg}[c]^{\dagger}S[c]\kappa^{-1}
-\]
-Since $S[c]S[c]^{\dagger} \approx I \approx S[c]^{\dagger}S[c]$, it follows that
-\[
-S_{\rm asg}[c]^{\dagger}=\kappa S[c]^{\dagger} = \kappa\Lambda[c]_s S[c]^T \Lambda[c]_r 
-\]
-\begin{equation}
-\label{eqn:asginv}
-= \kappa \Lambda[c]_s \kappa S_{\rm asg}[c]^T\Lambda[c]_r
-\end{equation}
-This is the approximate inverse computed in the examples. Note that
-only the values of $\kappa$ near the source datum $z=z_s$ play a role
-in the relation \ref{eqn:sreln} or in the definition \ref{eqn:asginv}
-of the IWAVE ASG approximate inverse.
-
-\section{Discussion}
-
-The source subproblem is a necessary step in solving the overall FWI problem \ref{eqn:fwi}.
-In particular, the nonlinear least squares realization \ref{eqn:esi} of source-extended FWI is naturally approached via the {\em variable projection method} \cite[]{GolubPereyra:73,GolubPereyra:03,vanLeeuwenMulder:09,Rickett:SEG12}, in which the minimum over $\bff$ in \ref{eqn:esis} is treated as a function of $\bf{c}$, which is in turn minimized.
-Variable projection thus treats the source subproblem as an inner problem, which must be solved for an estimate of $\bff$ at every step of an iterative method for the outer problem of minimization over $\bf{c}$.
-Efficiency in solving the source subproblem \ref{eqn:esis} is critical to efficient implementation of variable projection for solving the regularized FWI problem \ref{eqn:esi}.
-
-This paper restricts the choice of wave physics embodied in
-$L[\bf{c}]$ to linear acoustics.  Thus $\bf{c}=(\kappa,\rho)$ is the
-pair (bulk modulus, density) of positive functions of spatial
-position.
-The principal characteristics assumed of sources $\bff=$ are localization and isotropy.
-That is, physical sources will be assumed to be isotropic point radiators, described by the functional form
-\begin{equation}
-\label{eqn:ptsrc}
-f_{\rm pt}(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s). 
-\end{equation}
-Source locations $\bx_s$ are assumed known (as is natural for active source methods), and the source {\em wavelet} $w(t)$ is to be found as part of the solution of \ref{eqn:esi}.
-A convenient choice of penalty operator $A$ is then multiplication by the {\em offset} $|\bx-\bx_s|$: its null space consists precisely of distributions of the form \ref{eqn:ptsrc}.
-
-More complex radiation characteristics may be accommodated with similar but more complex choice for the operator $A$.
-Since the core analysis is similar, this paper deals only with the simplest case \ref{eqn:ptsrc}.
-
-The wavefield $u$ in this work is presumed to exist in all of Euclicean space-time: in other words, boundary effects have been removed from $u$, by surface-related multiple elimination or other techniques.
-Predicted data traces are simply the samples of one or more components of $u$ at receiver points $\bx_r$.
-
-%The domain of wave propagation is denoted $\Omega$, and two subsets of its boundary are identified: $\Gamma_s$ contains the support (non-zeroes) of the (extended) sources and in particular the locations $\bx_s$ of physical sources.
-Data trace locations $\bx_r$ lie in another subset $\Gamma_r$.
-The data projection operator $P$ samples the acoustic field $u$ on $\Gamma_r$, and includes a finite aperture mute.
-
-Achieving a small residual in the source subproblem \ref{eqn:esis} for arbitrary $c$ is a critical part of the convergence mechanism for the extended source approach to FWI \cite[]{HuangNammourSymesDollizal:SEG19}.
-However, for arbitrary data $d$, coefficient array $c$, and source location $\bx_s$, there does not exist an isotropic point source $f_{\rm pt}$ of the form \ref{eqn:ptsrc} for which $PL[c]^{-1}f_{\rm pt} \approx d$.
-This is so even if there does exist a coefficient array $c^*$ and point source $f_{\rm pt}^*$ located at $\bx_s$ with wavelet $w^*(t)$ for which $d = PL[c^*]^{-1}f_{\rm pt}^*$, if $c$ differs substantially from $c^*$.
-It is this fact that underlies the effectiveness of the extended source approach to FWI:  driving the residual in the problem \ref{eqn:esi} towards zero necessarily requires modifying $c$ to resemble $c^*$, assuming the latter exists, as well as pushing the optimal $f$ towards the null space of $A$.
-Conversely, the source subproblem \ref{eqn:esis} does not have a small residual solution of point source form \ref{eqn:ptsrc}, in general, for a given choice of $c$, even if that were the case for a different choice ($c^*$).
-Therefore a more general class of source models than that specified by the condition \ref{eqn:ptsrc} must be admitted to the feasible set for the source subproblem \ref{eqn:esis} if it is to serve as the inner problem in a variable projection formulation of extended source FWI.
-
-The class of non-point sources investigated here are those confined a hypersurface assumed to contain the locations of physical (point) sources, but not {\em a priori} required to have point support.
-These {\em surface extended sources} are able to yield a relatively small residual in the source subproblem \ref{eqn:esis} for more or less arbitrary coefficient array $c$.
-This observation is a by-product of the approximate solution for \ref{eqn:esis} constructed below.
-The construction of the approximate solution is closely related to the {\em time reversal} method of photoacoustic tomography (see \cite{StefanovUhlmannIP:09} and references cited there).
-
-The next section gives a precise description of an acoustic version of problem \ref{eqn:esis} for an idealized {\em crosswell} configuration, in which source located on a surface $z=z_s$ generate waves that propagate to to a receiver surface $z=z_r > z_s$.
-In principle, {\em diving waves} can be described with similar mathematics, but with some additional complications, so I do not treat the diving wave case explicitly in this paper.
-The assumptions detailed here imply that the data $d$ of the inverse problems \ref{eqn:esi}, \ref{eqn:esis} have the physical character of {\em transmitted waves}.
-There are no mathematical results at present concerning reflected wave inverse problems treated by source extension methods, though there are tantalizing numerical clues \cite[]{LeeuwenHerrmannWRI:13,Warner:14,Warner:16,LeeuwenHerrmann:16,HuangSymes:Geo17,HuangSymes:Geo18a,HuangSymes:Geo18b}.
-
-The following section describes the construction of the approximate
-inverse and explains the conditions under which accuracy should be expected.
-I observe that the approximate inverse is approximately the adjoint of the modeling operator $PL[c]^{-1}$ with respect to weighted norms in its domain and range (see \cite{HouSymes:15} for a similar observation in a different context).
-This fact immediately suggests use of the approximate inverse as a preconditioner to accelerate convergence of Krylov space methods such as conjugate gradient iteration applied to the least squares problem \ref{eqn:esis}.
-However, its straightforward application in the preconditioned conjugate gradient (PCG) method \cite[]{Golub:2012} involves explicitly a version the so called Dirichlet-to-Neumann (D2N) operator, which is computationally awkward.
-I show how to reorganize the PCG iteration so that only solutions of $L[c]u=f$ are required.
-The penultimate section presents some 2D examples in the crosswell configuration, illustrating the accuracy of the approximate inversion and the accelerated convergence of PCG.
-The examples are chosen to emphasize that surface extended sources can be constructed to fit more or less arbitrary data, unlike point sources - as mentioned before, this approximate invertibility property of the modeling operator is of critical importance in the application to extended FWI.
-The paper ends with a discussion of several important matters not addressed here, and a restatement of the conclusions.
-
-The intent of the main body of this paper is to present the formal structure of the extended source subproblem and the way in which this structure leads to accelerated numerical solution.
-This formal structure is however supported by a rigorous foundation.
-In order to avoid disrupting the formal account, this foundation material is relegated to an appendix.
-
-\bibliographystyle{seg}
-\bibliography{../../bib/masterref}

### 2021-05-24 repro/papers/sseprecond/paper.tex
interim

@@
 
-I will fully describe the preconditioner construction for a special
+I will fully describe a preconditioner for a special
 case of the source subproblem \ref{eqn:esis}, in which ${\bf u}$ is an
@@
 thermoacoustic tomography and other wave inverse problems
-\cite[]{Rachele:00,StefUhl:05}. The obvious computation of these maps
+\cite[]{Rachele:00,StefUhl:05}. \cite{HouSymes:EAGE16} demonstrated a
+very similar preconditioner for Least Squares Migration, also for its
+subsurface offset extension \cite[]{HouSymes:16}, motivated by
+\cite{tenKroode:12}. These constructions also involve the
+Dirichlet-to-Neumann operator. This concept also turns up in hidden
+form in the work of Yu Zhang and collaborators on true amplitude
+migration
+\cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,Zhang
+  YuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.
+
+The obvious computation of the pressure-to-source map
 - prescribe the pressure, solve the wave equation with this boundary
@@
 
-\plot{dsrcphh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
+\plot{dsrcpghh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
   pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
@@
 
-\plot{dsrcvzhh0}{width=\textwidth}{Trace $P_sv_z^+$ on $z=z_s=3000$ m of
+\plot{dsrcvzghh0}{width=\textwidth}{Trace $P_sv_z^+$ on $z=z_s=3000$ m of
   vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
@@
 \end{equation}
-An approximate inverse for $S^++_{z_s,z_r}$ would permit solution of
-this equation for $\Lambda^+_{z_s}f_s$.
-
-In this section I will explain how to build the approximate inverse.
-The key ingredient in this {\em time reversal} construction is a
-{\em local energy decay} assumption. Define the local energy in the
-acoustic field $(p,\bv)$ at
-time $t$ in a region $\Omega \subset \bR^3$ by limiting the integration
-in the definition of total energy \ref{eqn:defae0} to $\Omega$:
-\begin{equation}
-  \label{eqn:defaeloc}
-  E_{\Omega}[(p,\bv)](t) = \frac{1}{2} \int_{\Omega} \,d\bx \, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx) |\bv(\bx,t)|^2\right)
-\end{equation}
-The necessary assumption is: local energy decay in the half-space
-$\Omega = \{(x,y,z): z \le z_r\}$. The local energy in $\Omega$ of the
-causal solution $(p^+,\bv^+)$ of system \ref{eqn:awepm} cannot decay,
-since the source on $z=z_s<z_r$ radiates in both positive and negative
-$z$ in forward time. That obstacle is resolved by a redefinition of
-the ``downgoing'' concept. I will call  downgoing if there is a
-source-free acoustic field $(\tilde{p}^+,\tilde{\bv}^+)$ in all of space-time for which
-\begin{itemize}
-\item[1. ] $p^+ \approx \tilde{p}^+, \bv^+ \approx \tilde{\bv}^+$ in
-  the slab $\{(x,y,z,t): z_s \le z \le z_r\}$, and
-\item[2. ] The energy in $(\tilde{p}^+,\tilde{\bv}^+)$ decays in
-  forward time, in the half-space $\Omega$:  $E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](t) \rightarrow 0$
-  as $t \rightarrow \infty$.
-\end{itemize}
-Conditions for local energy decay have been widely studied in various
-settings, for instance obstacle scattering
-\cite[]{MorawetzPhillipsLax:05}. \cite{Pauen2000NonTr-6079 gives a
-comprehensive overview of this subject. For smooth (non-scattering)
-acoustic parameters, as is assumed in this paper, the crux of the
-matter is that rays carrying high-frequence energy exit the half-space
-$\Omega$ in finite time. For the half-space
-geometries considered here, local energy does not decay in general,
-because of the existence of ``trapped'' rays traveling parallel to the
-boundary and remaining interior to $\Omega$ for all time. If an
-acoustic field is downgoing in the usual sense, that is, its
-high-frequency energy is concentrated on rays that exit $\Omega$
-within some common maxiumum time, then the results cited above show
-that local energy in $\Omega$ decays (exponentially for $n=3$).
-
-Starting from this observation, it is possible to show that if the
-source vector $(h_s,f_s)^T$ has the property termed ``downgoing'' in
-preceding sections, that is, emanates high-frequency energy only along
-rays that leave $\Omega$ within a common maximum time, then the
-solution $(p^+,\bv^+)$ of \ref{eqn:awepm} is downgoing in the sense
-defined here. All of our numerical examples behave as one would expect
-for downgoing fields.
+An approximate inverse for $S^++_{z_s,z_r}$ would permit (approximate)
+solution of this equation for $\Lambda^+_{z_s}f_s$.
+
+Recall that the source vector $(h_s,f_s)$ is assumed to produce a
+downgoing field $(p^+,\bv^+)$, that is, emanates high-frequency energy only along
+rays that make an angle with the vertical bounded below by a common
+minimum angle. Such rays leave $\Omega$ within a common maximum time.
 
@@
 slab $z_s<z<z_r$, the field $(p^+,\bv^+)$ approximates the solution of an
-anti-causal evolution equation. For any choice of $T$, define
+anti-causal evolution equation. Indeed, for any choice of $T$, define
 $(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
-$\Omega$ of
+$\Omega \times (-\infty,T]$ of
 \begin{eqnarray}
@@
 \end{eqnarray}
+That is, $\tilde{p}^-$ has the same boundary value as
+$p^+$ for $t<T$, but $(\tilde{p}^-,\tilde{\bv}^-)$ vanishes at
+$t=T$. Suppose that $T$ is chosen so that the rays carrying
+significant energy in $(p^+,\bv^+)$ have left the half-space $\Omega$
+at time $T$. Then
+
 As will be shown later, under the downgoing assumption, $P_r p^+
@@
 
-\section{Fields, Traces and Sources}
-In this section, $(p^{\pm}_0,\bv^{\pm}_0)$ denotes an acoustic field,
-continuous in the region $z_s \le z $, obeying the homogeneous wave
-equation in the interior $z_s < z $:
-\begin{eqnarray}
-\label{eqn:awe1}
-  \frac{1}{\kappa}\frac{\partial p^{\pm}_0}{\partial t} & = & - \nabla \cdot \bv^{\pm}_0, \nonumber \\
-  \rho\frac{\partial \bv^{\pm}_0}{\partial t} & = & - \nabla
-                                                    p^{\pm}_0, \nonumber \\
-  p^{\pm}_0 & =& 0,  \mbox{ for } t \ll 0, \nonumber\\ 
-  \bv^{\pm}_0 & = & 0 \mbox{ for } t \ll 0.
-\end{eqnarray}
-I will show how to view $(p^{\pm}_0,\bv^{\pm}_0)$ as the $z>z_s$ part
-of a solution of the system \ref{eqn:awepm} in two different ways,
-corresponding to two different extensions to the whole space. An
-important relation between the components of the operator
-${\cal S}^{\pm}_{z_s,z_r}$ follows (equation \ref{eqn:lamnull} below).
-
-Given a function $\phi$ on $z=z_s$, define $(p^{\pm}_1,\bv^{\pm}_1)$
-as the solution in the two half-spaces $z <z_s, z>z_s$ of
-\begin{eqnarray}
-\label{eqn:awe1}
-  \frac{1}{\kappa}\frac{\partial p_1}{\partial t} & = & - \nabla \cdot \bv_1, \nonumber \\
-  \rho\frac{\partial \bv_1}{\partial t} & = & - \nabla p_1,\nonumber \\
-  p^{\pm}_1 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
-  \bv^{\pm}_1 & = & 0 \mbox{ for } \pm t \ll 0,\nonumber\\
-  \lim_{z \rightarrow z_s} P_sp^{\pm}_1& =& \phi.
-\end{eqnarray}
-By construction, $p^{\pm}_1$ is continuous across $z=z_s$. The
-discontinuity $[v^{\pm}_{1,z}]|_{z=z_s}$ is a linear function of
-$\phi$. Define the {\em pressure-to-source map} $\Lambda^{\pm}_{z_s}$ by
-\begin{equation}
-  \label{eqn:deflam}
-  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]|_{z=z_s}
-\end{equation}
-Then as a distribution on all of $\bR^4$, $(p^{\pm}_1,\bv^{\pm}_1)$ solves the
-inhomogeneous system
-\begin{eqnarray}
-\label{eqn:sawe1}
-  \frac{1}{\kappa}\frac{\partial p^{\pm}_1}{\partial t} & = &
-                                                        (\Lambda^{\pm}_{z_s}\phi)\delta(z-z_s) - \nabla \cdot \bv^{\pm}_1, \nonumber \\
-  \rho\frac{\partial \bv^{\pm}_1}{\partial t} & = & - \nabla p^{\pm}_1,\nonumber \\
-  p^{\pm}_1 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
-  \bv^{\pm}_1 & = & 0 \mbox{ for } \pm t \ll 0.
-\end{eqnarray}
-
-This is precisely the causal option of system \ref{eqn:awepm} for
-$f_s=0$ and $h_s=\Lambda^{\pm}_{z_s}\phi$. From \ref{eqn:awe1} and the definition
-\ref{eqn:fwd},
-\begin{equation}
-  \label{eqn:lamfwd}
-  \phi = S^{\pm}_{z_s,z_s}\Lambda^{\pm}_{z_s}\phi,
-\end{equation}
-that is,
-\begin{equation}
-  \label{eqn:lamfwd1}
-  S^{\pm}_{z_s,z_s} = (\Lambda^{\pm}_{z_s})^{-1}.
-\end{equation}
-
-Similarly, suppose $\psi$ is a function on $z=z_s$, and that
-$(p^{\pm}_2,\bv^{\pm}_2)$ solves
-\begin{eqnarray}
-\label{eqn:awe2}
-  \frac{1}{\kappa}\frac{\partial p^{\pm}_2}{\partial t} & = & - \nabla \cdot \bv^{\pm}_2, \nonumber \\
-  \rho\frac{\partial \bv^{\pm}_2}{\partial t} & = & - \nabla p^{\pm}_2,\nonumber \\
-  p^{\pm}_2 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
-  \bv^{\pm}_2 & = & 0 \mbox{ for } \pm t \ll 0,\nonumber\\
-  \lim_{z \rightarrow z_s} P_sv^{\pm}_{2,z} &=& \psi.
-\end{eqnarray}
-in the two half-spaces $z<z_s, z > z_s$. In this case, $v^{\pm}_{2,z}$
-is continuous at $z=z_s$ by construction. Viewed as a
-distribution on all of $\bR^4$, $(p^{\pm}_2,\bv^{\pm}_2)$ solve the
-system \ref{eqn:awepm} with $h_s=0$, and the discontinuity in pressure
-across $z=z_s$ as the inhomogeneous term in Newton's law:
-\begin{eqnarray}
-\label{eqn:sawe2}
-  \frac{1}{\kappa}\frac{\partial p^{\pm}_2}{\partial t} & = &
-                                                        - \nabla \cdot \bv^{\pm}_2, \nonumber \\
-  \rho\frac{\partial \bv^{\pm}_2}{\partial t} & = & 
-                                                    [p^{\pm}_2]|_{z=z_s}{\bf e_z}\delta(z-z_s) - \nabla p^{\pm}_2,\nonumber \\
-  p^{\pm}_2 & =& 0,  \mbox{ for } \pm t \ll 0,\nonumber\\ 
-  \bv^{\pm}_2 & = & 0 \mbox{ for } \pm t \ll 0.
-\end{eqnarray}
-This relation defines a map from the velocity boundary data $\psi$ to
-the source in the velocity equation, having units of force/volume,
-similar to $\Lambda^{\pm}_{z_s}$. As it's not needed in the sequel, I
-won't name this map.
-
-A closer relation between traces and sources follows from the
-geometric optics analysis of the half-space solution, presented
-below. I will show that, for propagating wave data,
-\begin{equation}
-  \label{eqn:odd1}
-  \lim_{z\rightarrow z_s^+} v_{1,z} \approx - \lim_{z\rightarrow z_s^-}
-  v_{1,z}
-\end{equation}
-Similarly,
-\begin{equation}
-  \label{eqn:odd2}
-  \lim_{z\rightarrow z_s^+} p_{2} \approx - \lim_{z\rightarrow z_s^-}
-  p_{2}
-\end{equation}
-for the velocity field solving \ref{eqn:awe1}, or equivalently
-\ref{eqn:awe1}. Here ``$\approx$'' means in the sense of high
-frequency asymptotics, that is, that the difference between the two
-sides is relatively smooth, hence small if the data is highly
-oscillatory. Similarly, for the pressure field solving \ref{eqn:awe2}
-or \ref{eqn:sawe2}, the jump at $z=z_s$ is twice the limiting value
-(trace) from above, up to a smooth error. Therefore
-\begin{eqnarray}
-  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]|_{z=z_s} &\approx& 2
-                              \lim_{z\rightarrow z_s}v^{\pm}_{1,z},\\  \label{eqn:tracejump1}
-  [p^{\pm}_2]|_{z=z_s} &\approx& 2 \lim_{z\rightarrow z_s^+}p^{\pm}_{2},\\  \label{eqn:tracejump2}
-\end{eqnarray}
-
-%%%%%%%%%%%%%%%%%%%
-These apparently harmless relations have a profound consequence,
-because of a final step: identify the boundary data $\phi$ and $\psi$,
-so far arbitrary, as the traces of $p^{\pm}_0$ and $v^{\pm}_{0,z}$ on $z=z_s$,
-that is, set
-\begin{eqnarray}
-  \phi & = & P_s p^{\pm}_0, \nonumber \\
-  \psi & = & P_s v^{\pm}_{0,z}.
-             \label{eqn:bv0}
-\end{eqnarray}
-Since the traces on $z=z_s$ of the pressure fields $p^{\pm}_0$ and $p^{\pm}_1$ are
-the same, and both are pressure components of solutions of the same homogeneous
-acoustic system in $z>z_s$, the standard energy identity implies that
-$(p^{\pm}_0,\bv^{\pm}_0) = (p^{\pm}_1,\bv^{\pm}_1)$ in
-$z > z_s$. Similarly, $P_sv^{\pm}_{0,z}$ and $P_sv^{\pm}_{2,z}$ are
-the same, whence $(p^{\pm}_0,\bv^{\pm}_0) = (p^{\pm}_2,\bv^{\pm}_2)$ in
-$z>z_s$.
-
-Therefore the assumption \ref{eqn:bv0} implies that
-\begin{equation}
-  \label{eqn:jump0}
-  \Lambda^{\pm}_{z_s}\phi = [v^{\pm}_{1,z}]_{z=z_s} \approx 2P_s v^{\pm}_{0,z}  = 2\psi,
-\end{equation}
-and
-\begin{equation}
-  \label{eqn:jump1}
-[p_2^{\pm}]_{z=z_s} \approx  2P_s p^{\pm}_{0}= 2 \phi.           
-\end{equation}
-In the notation of the last section, set $2 \phi = 2P_sp^{\pm}_0= f_s,
-2\psi = 2P_s v^{\pm}_{0,z}= h_s$. Then relation \ref{eqn:jump0}
-implies that
-\begin{equation}
-  \label{eqn:lamfh}
-  \Lambda^{\pm}_{z_s}f_s \approx h_s.
-\end{equation}
-
-From the definitions in the last section and the system \ref{eqn:sawe1},
-\begin{equation}
-  \label{eqn:pident0}
-P_r p^{\pm}_0 = \Pi_0{\cal S}^{\pm}(h_s,0)^T = \Pi_0{\cal
-  S}^{\pm}\Pi_0^Th_s,
-\end{equation}
-and
-\begin{equation}
-  \label{eqn:pident1}
-P_r p^{\pm}_0 \approx \Pi_0{\cal S}^{\pm}_{z_s,z_r} (0,f_s)^T = \Pi_0{\cal
-  S}^{\pm}_{z_s,z_r} \Pi_1^Tf_s,
-\end{equation}
-Similarly, from the relations above and \ref{eqn:sawe2},
-\begin{equation}
-  \label{eqn:vident0}
-  P_r  v^{\pm}_{0,z}  \approx \Pi_1{\cal S}^{\pm}_{z_s,z_r}(h_s,0)^T  =
-  \Pi_1{\cal S}^{\pm}_{z_s,z_r} \Pi_0^Th_s 
-\end{equation}
-and
-\begin{equation}
-  \label{eqn:vident1}
-  P_r  v^{\pm}_{0,z}  = \Pi_1{\cal S}^{\pm}_{z_s,z_r}(0,f_s)^T  =
-  \Pi_1{\cal S}^{\pm}_{z_s,z_r} \Pi_1^Tf_s 
-\end{equation}
-Together with \ref{eqn:lamfh}, these equations imply that
-\begin{equation}
-  \label{eqn:lamnull}
-{\cal  S}_{z_s,z_r}^{\pm}\Pi_0^T\Lambda^{\pm}_{z_s} \approx {\cal S}^{\pm}_{z_s,z_r}\Pi_1^T.
-\end{equation}
-Consequently, ${\cal S}_{z_s,z_r}$ has an ``approximate null space'',
-in the sense that its composition with the block operator
-$(\Lambda^{\pm}_{z_s},-I)^T$ is asymptotically negligible.
-
-\section{Parametrix via Time Reversal}
-
-\section{Approximate Unitarity}
-Recall that the fields $(p^{\pm},\bv^{\pm})$ in $z>z_s$ can be viewed
-as either the solution of the system \ref{eqn:awepm} with pressure source $h_s=0$ and
-a suitable choice of velocity source $f_s$, or of the same system with
-suitable choice of $h_s$ and $f_s=0$, and that these choices are
-related by equation \ref{eqn:lamfh}: $\Lambda^{\pm}f_s = h_s$. The
-pressure traces at $z=z_r>z_s$ are the same: from the definitions
-\ref{eqn:sdef}, \ref{eqn:vdef},
-\begin{eqnarray}
-  \label{eqn:sandv}
-  S^{\pm}_{z_s,z_r}h_s& =& P_rp^{\pm},\nonumber \\
-  V^{\pm}_{z_s,z_r}f_s &=& P_rv_z^{\pm}.
-\end{eqnarray}
-Applying the same reasoning to traces on $z=z_r$: create pressure and
-velocity sources $h_r$ and $f_r$ at $z=z_r$ by odd extension of the
-velocity, respectively pressure, fields, and using the jumps as source
-coefficients. Then as before
-$\Lambda^{\pm}_{z_r}f_r=h_r=-P_rv^{\pm}_z$ (the sign change is due to
-$Lambda$ producing the jump, whereas $P_rv^{\pm}_z$ is the limit from
-the left of the field $v^{\pm}_{1,z}$ of system \ref{eqn:awe1}), hence
-$\Lambda^{\pm}_{z_r}P_rp^{\pm} = P_rv_z^{\pm}$. Combining this
-observation with equations
-\ref{eqn:lamfh}, \ref{eqn:sandv}, obtain (with roles of $z_s,z_r$ interchanged)
-\[
-V^{\pm}_{z_r,z_w}(-2P_rp^{\pm}) =
-  \Lambda^{\pm}_{z_s}S^{\pm}_{z_r,z_s}(-2P_rv_z^{\pm})  =
-  \Lambda^{\pm}_{z_s}S^{\pm}_{z_r,z_s}(-2\Lambda^{\pm}_{z_r}P_rp^{\pm})
-\]
-whence
-\begin{equation}
-  \label{eqn:preunit}
- V^{\pm}_{z_r,z_w}
- =\Lambda^{\pm}_{z_s}S^{\pm}_{z_r,z_s}\Lambda^{\pm}_{z_r}.
-\end{equation}
-
-In view of the relations \ref{eqn:trtrcomp}, equation
-\ref{eqn:preunit} implies that
-\[
-  V^-_{z_r,z_s}=-(V^+_{z_s,z_r})^T = -\Lambda^{-}_{z_s}(S^+_{z_s,z_r})^T\Lambda^{-}_{z_r}.
-\]
-\[
-  =-(\Lambda^{+}_{z_s})^T(S^+_{z_s,z_r})^T(\Lambda^{+}_{z_r})^T.
-\]
-Therefore the first equation in display \ref{eqn:approxinv} can be re-written
-\[
-  -4V^-_{z_r,z_s}S^+_{z_s,z_r} =
-\]
-\begin{equation}
-  \label{eqn:baseunit}
- 4 (\Lambda^{+}_{z_s})^T(S^+_{z_s,z_r})^T(\Lambda^{+}_{z_r})^T
-  S^+_{z_s,z_r} \approx I
-\end{equation}
-
-\section{Derivations}
-\subsection{Extension to whole space}
-If $\phi$ is smooth and vanishes for large negative $t$, then
-$(p_1,\bv_1)$ can be constructed by pasting together the smooth
-solutions of the inhomogeneous Dirichlet problems in the half-spaces
-$\pm z > z_s$. Construct these by extending $\phi$ smoothly to
-$\pm z>z_s$, and solve systems similar to \ref{eqn:awe1} for the
-difference with $(p_1,\bv_1)$ by quoting the main results from
-\cite{BlazekStolkSymes:13}. Since the right hand side in this
-difference system is smooth, and the system is causal, the solution is
-smooth and vanishes on $z=z_s$. Adding back the extension of $\phi$
-produces smooth solutions in $\pm z > z_s$; define $(p_1,\bv)_1$ to be
-these smooth functions in $z \ne z_s$. Since the boundary values for
-pressure are the same, $p_1$ extends continuously to $z=z_s$. 
-The pressure boundary value $\phi$ uniquely determines the solutions
-in the two half-spaces, by standard energy estimates,
-
-\subsection{Adjoint Computation}
-The adjoint of ${\cal S}^+_{z_s,z_r}$ can be computed by a variant of
-the adjoint state method, in this case a by-product of the
-conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awem}
-with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
-$ (h_r,f_r\bf{e}_z) \delta(z-z_r)$. Then
-\[
-0 = 
-\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
-\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
--
-\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
-\]
-\[
-+
-\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
- h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
- h_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  (- \nabla p^++f_s\bf{e}_z) \cdot \bv^- + \bv^+ \cdot (-\nabla
-  p^- + f_r \bf{e_z}) \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
- h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
- h_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- 
-  +f_s \delta(z-z_s) v_z^- + v_z^+f_r \delta(z-z_r) \right)
-\]
-after integration by parts in the last two terms. Most of what is left cancels, leaving 
-\[
-0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
-( h_rP_rp^++f_rP_rv_z^+) = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
-\]
-whence
-\begin{equation}
-\label{eqn:sadj}
- ({\cal S}^+_{z_s,z_r})^T = -{\cal S}^{-}_{z_r,z_s}.
-\end{equation}
-
-The systems \ref{eqn:awep} and \ref{eqn:awem} differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
-$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
-\begin{equation}
-  \label{eqn:trdef}
-  {\cal R} \left(
-    \begin{array}{c}
-      p\\
-      \bv
-    \end{array}
-  \right) =
-  \left(
-    \begin{array}{c}
-      Rp\\
-      -R\bv
-    \end{array}
-  \right)
-\end{equation}
-Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awep} with
-$(h_s\delta(z-z_s),f_s\bf{e_z}\delta(z-z_s))$ replaced by
-$(-Rh_r\delta(z-z_r), Rf_r\bf{e_z}\delta(z-z_r))$. That is,
-\begin{equation}
-  \label{eqn:trsadj}
-  {\cal R}{\cal S}^- = -{\cal S}^+_{z_r,z_s}{\cal R}
-\end{equation}
-Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that
-
-\section{Surface source-to-pressure operator}
-
-A particular case of the relation \ref{eqn:trtr} is important in its own right, when $z_r=z_s$. Define surface source-to-pressure operator $\Lambda_{z_s}$ by
-\begin{equation}
-  \label{eqn:ntoddef}
-  \Lambda^{\pm}_{z_s} = S^{\pm}_{z_s,z_s}.
-\end{equation}
-Then \ref{eqn:trtr} reads
-\begin{equation}
-  \label{eqn:ntodtr}
-  (\Lambda^+_{z_s})^T = R\Lambda_{z_s}^+R = -\Lambda^-_{z_s}
-\end{equation}
-This operator is closely related to the operator mapping $\lim_{z\rightarrow z_s^{\pm}} \bv_z^+$ to $\lim_{z\rightarrow z_s^{\pm}}  p^+$, often called the {\em Neumann-to-Dirichlet} operator.
-The connection is explained below.
-
-The quadratic form defined by $\Lambda^{\pm}_{z_s}$ has fundamental physical
-significance. Define the total acoustic energy $E^{\pm}(t)$ of the field $(p^{\pm},\bv^{\pm})$, at time $t$ by
-\begin{equation}
-  \label{eqn:defae}
-  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)
-\end{equation}
-Then for $t_{\min}<t_{\max}$,
-\[
-  E^{\pm}(t_{\rm max})-E^{\pm}(t_{\rm min}) = \int_{t_{\rm min}}^{t_{\rm max}} \frac{dE^{\pm}}{dt}
-\]
-\[
-  = \int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\, \left(p^{\pm} \frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} + \bv^{\pm} \cdot \rho \frac{\partial \bv^{\pm}}{\partial t}\right)
-\]
-\[
-  =\int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\,\left(p^{\pm}(-\nabla \cdot \bv^{+} + h_s\delta(z-z_s)) + \bv^{\pm} \cdot \rho \frac{\partial \bv^{\pm}}{\partial t}\right)
-\]
-\[
-=  \int_{t_{\rm min}}^{t_{\rm max}}\int\,d\bx\, \left(\left(\nabla p^{\pm}+\rho \frac{\partial \bv^{\pm}}{\partial t}\right) + p^{\pm}h_s\delta(z-z_s)\right)
-\]
-\begin{equation}
-  \label{eqn:eident}
-  =  \int_{t_{\rm min}}^{t_{\rm max}}\int\,dxdy h_s p^{\pm}|_{z=z_s} = \int_{t_{\min}}^{t_{\max}}\langle h_s(t), (\Lambda^{\pm}_{z_s} h_s)(t) \rangle_{L^2(\bR^2)}.
-\end{equation}
-That is, $\langle h_s(t), (\Lambda^{\pm}_{z_s} h_s)(t)
-\rangle_{L^2(\bR^2)}$ is the rate of energy transfer from source to
-the fluid at time $t$.
-
-Assume that $h_s$ has compact support in time. Then $E^{\pm}
-\rightarrow 0$ as $t \rightarrow \mp \infty$, and is eventually
-constant as $t \rightarrow \pm \infty$. Then the energy identity
-\ref{eqn:eident} implies that
-\begin{equation}
-  \label{eqn:etot}
-  \pm \lim_{t \rightarrow \pm \infty} E^{\pm}(t) = \int \,dt\, \langle
-  h_s(t), (\Lambda^{\pm}_{z_s} h_s)(t)\rangle_{L^2(\bR^2)} = \langle
-  h_s,\Lambda^{\pm}_{z_s} h_s\rangle
-\end{equation}
-By virtue of identity \ref{eqn:ntodtr},
-\[
-  \lim_{t \rightarrow -\infty}E^{-}(t) =-\langle
-  h_s,\Lambda^{-}_{z_s} h_s\rangle
-\]
-\[
-  =  \langle  h_s,(\Lambda^{^+})^T_{z_s} h_s\rangle
-\]
-\[
-  =  \langle  h_s,\Lambda^{^+}_{z_s} h_s\rangle
-\]
-so
-\begin{equation}
-  \label{eqn:surfenergy}
-  = \lim_{t \rightarrow \infty}E^{+}(t)
-\end{equation}
-That is, the energy total energy transferred from the source to the
-causal acoustic field as it evolves forward in time is the same as the
-total energy transferred to the anti-causal acoustic field as it
-evolves backwards in time.
-
-Also onclude from identity \ref{eqn:etot} that the symmetric part of $\Lambda^+_{z_s}$ is positive semidefinite:
-\begin{equation}
-  \label{eqn:lamsd}
-  \frac{1}{2}((\Lambda_{z_s}^+)^T + \Lambda^+_{z_s}) \ge 0.
-\end{equation}
-
-%Observe that the causal acoustic system \ref{eqn:awep} implies directly that
-%\begin{equation}
-%  \label{eqn:sourcevjump}
-%  h_s= [v_z]|_{z=z_s}.
-%\end{equation}
-
-%######################################
-\section{Propagating Waves}
-Define $H$ to be the subspace $\{\bu \in (L^2(\bR^3))^4: \bu=0,z\ge
-z_s\}$, by analogy with \cite{Lax:PDENotes}, p. 122.
-%Note: must add microlocal $H^1$ constraint per
-%\cite[]{BaoSy:91b} so that trace operator is well-defined - postpone
-%this.
-
-Given $u=(\bar{p}_0,\bar{\bv}_0) \in H$, extend in $t$ by solving the
-acoustic initial value problem
-\begin{eqnarray}
-\label{eqn:awe0}
-  \frac{1}{\kappa}\frac{\partial p_0}{\partial t} & = & - \nabla \cdot \bv_0, \nonumber \\
-  \rho\frac{\partial \bv_0}{\partial t} & = & - \nabla p_0,\nonumber \\
-  p_0 & =& \bar{p}_0, t=0,\nonumber\\ 
-  \bv_0 & = & \bar{\bv}_0, t=0,\nonumber\\
-\end{eqnarray}
-Standard results on hyperbolic systems assure existence of a unique
-square-integrable (weak) solution, which is smooth if $(p_0,\bv_0)$ is
-smooth at $t=0$ \cite[]{CourHil:62,LaxPDENotes}.
-
-Note that $(p_0,\bv_0)=0$ near $t=0$ in the half-space $z>z_s$.
-
-For any suitable space-time function $u$, define
-\begin{equation}
-  \label{eqn:postrace}
-  P^+_{z_s}u(x,y,t) =
-  \left\{
-    \begin{array}{c}
-      u(x,y,z_s,t), t>0,\\
-      0, t\le 0.
-    \end{array}
-  \right\}
-\end{equation}
-(Here, ``suitable'' means: the trace on $z=z_s$ makes sense. More on
-this later.)
-
-Since $p_0$ also solves the acoustic system \ref{eqn:awe1} in $z>z_s$
-and has the same data at $t=0, z>z_s$ and trace on $z=z_s$, standard
-uniqueness theorem shows that $(p_1,\bv_1) = (p_0, \bv_0)$ in $z>z_s$.
-
-\section{Progressing Waves and Symbols}
-
-The symbol $\sigma(L[c])$ of a matrix operator $L[c]$ is a complex scalar-valued matrix of the same size. In terms of its product with an arbitrary vector $\bu \in \bR^{d+1}$, it is
-\[
-  \sigma(L[c]) \bu = e^{-i(\omega t + \bk \cdot \bx)} (L[c]e^{i(\omega t + \bk \cdot \bx)}\bu).
-\]
-For $L[c]$ given by the definition \ref{eqn:awemat}, obtain
-\begin{equation}
-  \sigma(L[c])(t,\bx,\omega,\bk) = i\left(
-    \begin{array}{cc}
-      \frac{\omega}{\kappa(\bx)} & \bk^T \\
-      \bk & \rho(\bx) \omega)
-    \end{array}
-  \right)
-\end{equation}
-
-The characteristic equation of $L[c]$ is $\det \sigma(L[c]) = 0$. The significance of the characteristic equation is its role in constraining the phase $\psi(\bx)$ in the progressing wave ansatz,
-\begin{equation}
-  \label{eqn:go}
-  \bu(\bx,t) = e^{i\omega (t-\psi(\bx))} \ba (\bx).
-\end{equation}
-Applying $L[c]$ to this ansatz, obtain
-\[
-  L[c]\bu(\bx,t)=i \omega
-  \left(
-    \begin{array}{c}
-      \frac{1}{\kappa}a_p -  \nabla \psi \cdot \ba_v \\
-      -a_p \nabla \psi + \rho \ba_v
-    \end{array}
-  \right) + ...
-\]
-\[
-  = i \omega \sigma(L[c])(t,\bx,1,- \nabla \psi(\bx))ba + ...
-\]
-where ``$...$'' denotes terms of lower order in $\omega$. Thus $\bu$ as defined in equation \ref{eqn:go} is a high-frequency asymptotic solution, $L[c]\bu \approx 0$, if
-\begin{itemize}
-\item
-  \begin{equation}
-    \label{eqn:eik}
-    \det(\sigma(L[c])(t,\bx,1,-\nabla \psi(\bx))) = 0,
-  \end{equation}
-  (eikonal equation), and
-\item
-  $\ba$ is a null vector of $\sigma(L[c])(t,\bx,1,-\nabla \psi)$.
-\end{itemize}
-Explicitly, the eikonal equation reads
-\[
-  0 = \rho^2\left(\frac{\rho}{\kappa}-|\nabla \psi|^2\right),
-\]
-equivalent to the familar form
-\begin{equation}
-  \label{eqn:usualeik}
-  c|\nabla \psi| = 1,
-\end{equation}
-in which $c = \rho/\kappa$ is the wave velocity. Solution via the method of characteristics gives
-\[
-  \psi(\bx(t)) = \psi(\bx(0)) + t
-\]
-for a ray $t \mapsto \bx(t)$ satisfying
-\[
-  \frac{d\bx}{dt}(t) = -c(\bx)^2\nabla \psi(\bx(t))
-\]
-Finally, $\ba$ is a null vector of $\sigma(L[c])(\cdot,\cdot,1,-\nabla \psi)$ iff
-\begin{equation}
-  \label{eqn:godton}
-  \ba_v = \frac{a_p}{\rho}\nabla \psi
-\end{equation}
-The scalar $a_p$ evolves along the ray according to the transport equation, which I will not derive here; suffice it to say that $a_p(t) = a_p(0) \gamma(t)$, where $\gamma > 0$ is smooth and $\gamma(0)=1$.
-
-Suppose that $\bx_0 = (x,y,z_s)^T$ and $(\xi_x,\xi_y) \in \bR^2$ satisfies $c(\bx_0)|(\xi_x,\xi_y)|<1$.
-Then there is a neighborhood $\Omega$ of $\bx_0$ and $\xi_z \in C^{\infty}(\Omega), \xi_z>0$, so that $\bxi^{\pm}(\bx)=(\xi_x,\xi_y,\pm \xi_z(\bx))$ satisfy $c(\bx)|\bxi^{\pm}(\bx)|=1$ for $\bx \in \Omega$.
-
-A standard argument shows that there exists $\epsilon > 0$ for which a solution $\psi^{\pm}$ of the eikonal equation exists in $\Omega \cap \{\bx: |z-z_s|<\epsilon\}$ with
-\begin{equation}
-  \label{eqn:eikic}
-  \psi^{\pm}(x,y,z_s,t) = t -x \xi_x - y \xi_y, \, \frac{\partial \psi}{\partial z}(x,y,z_s,t) = \pm \xi_z(x,y,z_s).
-\end{equation}
-Let $\chi \in C_0^{\infty}, \chi(\bx_0)=1$, and set $p_0^{\pm}(x,y,t) = \chi(x,y,z_s)e^{i\omega(t-x\xi_x - y\xi_y)}$. Define $p_{\rm go}^{\pm} \in C^{\infty}(\Omega \cap \{\bx: |z-z_s|<\epsilon\})$ by
-\[
-  p_{\rm go}^{\pm}(\bx,t) = a^{\pm}_p(\bx) e^{i\omega (t-\psi^{\pm}(\bx))}.
-\]
-where $a^{\pm}_p$ solves the transport equation with $a^{\pm}_p|_{z=z_s} = \chi$. Set
-\[
-  \bv^{\pm}_{\rm go}(\bx,t) = \frac{a^{\pm}_p(\bx)}{\rho(\bx)}\nabla \psi^{\pm}(\bx) e^{i\omega (t-\psi^{\pm}(\bx))}
-\]
-Then $\bu^{\pm}_{\rm go} = (p^{\pm}_{\rm go}, \bv^{\pm}_{\rm go})$ is an asymptotic solution of $L[c]\bu = 0$ with $p^{\pm}_{\rm go}|_{z=z_s} = \chi e^{i\omega(t-x\xi_x - y\xi_y)}$.
-
-From equations \ref{eqn:godton} and \ref{eqn:eikic},
-\[
-  v_{z,{\rm go}}^{\pm}(x,y,z_s,t) = \pm \frac{\chi(x,y,t)}{\rho(x,y,z_s)}\xi_z  e^{i\omega(t-x\xi_x - y\xi_y)}
-\]
-\begin{equation}
-  \label{eqn:asymdton}
-= \pm \frac{\chi(x,y,t)}{\rho(x,y,z_s)c(x,y,z_s)}\sqrt{1-c(x,y,z_s)^2(\xi_x^2+\xi_y^2)}
-\end{equation}
-
-Integration over frequency produces a progressing wave traveling (at least near $z=z_s$) downwards, that is, increasing $z$ with increasing $t$, for the choice of sight $\pm = +$, and upwards, that is, decreasing $z$ with increasing $t$, for the other sign.
-Therefore a causal progressing wave with $p$ given on $z=z_s$ is obtained by synthesizing distorted plane waves as just described with $\pm = +$ in $z>z_s$, $\pm=-$ in $z<z_s$.
-From equation \ref{eqn:asymdton}, the normal components of $\bv$ have limits equal in magnitude but of opposite sign as $z\rightarrow 0^{\pm}$, up to a lower frequency error.
-That is, the jump in $\bv$ satisfies
-\begin{equation}
-  \label{eqn:asymjump}
-  [v_z]\approx 2\lim_{z\rightarrow 0^+} v_{z}
-\end{equation}
-where $\approx$ denotes ``up to a low frequency error'', from here on.
-
-This construction also implies an approximation of the source-to-pressure operator as a pseudodifferential operator.
-Choose a partition of unity in $z=z_s$, for which the support of every member is a domain $\Omega$ for which the construction above can be carried out: this implies that there be a uniform $\delta > 0$ so that
-\begin{equation}
-  \label{eqn:nongraze}
-  1>\delta + c(x,y,z_s)\sqrt{\xi_x^2+\xi_y^2}
-\end{equation}
-for all $(x,y,t,1,\xi_x,\xi_y) \in WF(p_0)$. Denote by $\Gamma$ the subset of $(x,y,t,k_x,k_y,\omega) \in T^*(\bR^3)$ for which $x,y,\xi_{xy}=k_{xy}/\omega$ satisfy the inequality \ref{eqn:nongraze}.
-
-Then near $z=z_s$, $p$ is approximately a sum of geometric optics solutions of the type displayed above, integrated over $\omega,k_x,k_y$, with $k_{x,y} = \omega \xi_{x,y}$.  Thus
-\[
-  p(x,y,z_s,t) \approx \frac{1}{2}\int_{\Gamma(x,y)} \,d\omega\,dk_x\,dk_y \hat{[v_z]}\rho(x,y,z_s)c(x,y,z_s)
-\]
-\[
-   \times \left(1-c^2(x,y,z_s)\left(\left(\frac{k_x}{\omega}\right)^2 + \left(\frac{k_y}{\omega}\right) ^2\right)\right)^{-1/2}e^{\i(\omega t + k_x x+ k_y y)}
-\]
-Since $p$ is the pressure component of a solution of \ref{eqn:awep}, equation \ref{eqn:sourcevjump} implies that
-\begin{equation}
-  \label{eqn:psidodton}
-  \Lambda^+_{z_s} h_s(x,y,t) \approx \frac{1}{2}\int_{\Gamma(x,y)} \,d\omega\,dk_x\,dk_y
-  \lambda_s(x,y,t,k_x,k_y,\omega) \hat{h_s}(k_x,k_y,\omega) e^{\i(\omega t + k_x x+ k_y y)}
-\end{equation}
-with
-\begin{equation}
-  \label{eqn:symboldton}
-  \lambda^+_{z_s}(x,y,t,k_x,k_y,\omega) =\frac{1}{2}\rho(x,y,z_s)c(x,y,z_s)\left(1-c^2(x,y,z_s)\left(\left(\frac{k_x}{\omega}\right)^2 + \left(\frac{k_y}{\omega}\right) ^2\right)\right)^{-1/2}
-\end{equation}
-
-Restricted to $\Gamma$, $\lambda^+_{z_s} \in S^0_{0,1}(\bR^2)$ is real-valued, hence defines an essentially symmetric operator. That is,
-\begin{equation}
-  \label{eqn:approxsymm}
-  (\Lambda^+_{z_s})^T \approx \Lambda^{+}_{z_s}.
-\end{equation}
-
-Note that the defintion \ref{eqn:appinv} of $ (S^+_{z_s,z_r})^{\dagger} $ is that of
-the transpose of $S^+_{z_s,z_r}$ in the sense of weighted norms, as
-follows. Define trace spaces $W=L^2(\bR^2 \times [0,T])$, with inner product
-\[
-  \langle h_1,h_2\rangle_s = \langle
-  h_1,\frac{1}{2}(\Lambda^+_{z_s}+(\Lambda^+_{z_s})^T)h_2\rangle
-\]
-and $D=L^2(\bR^2 \times [0,T])$ with inner product
-\[
-  \langle d_1,d_2\rangle_r = \langle
-  d_1,\left(\frac{1}{2}(\Lambda^+_{z_s}+(\Lambda^+_{z_s})^T)\right)^{-1}d_2\rangle
-\]
-(These definitions must be supplemented with constraints on
-out-of-aperture smoothness). Then the adjoint of $S^+_{z_s,z_r}: W
-\rightarrow D$ is
-\[
- \left(\frac{1}{2}
-   (\Lambda^+_{z_s}+(\Lambda^+_{z_s})^T)\right)^{-1}(S^+_{z_s,z_r})^T
- \left(\frac{1}{2}(\Lambda^+_{z_r}+(\Lambda^+_{z_r})^T)\right)^{-1}
-\]
- By virtue of the approximation \ref{eqn:approxsymm}, this is
-\[
-   \approx
-   (\Lambda^{+}_{z_s})^{-1}(S^{+}_{z_r,z_s})^T((\Lambda^{+}_{z_r})^{-1})^T
-   = (S^+_{z_s,z_r})^{\dagger}.
-\]
-which from the last section is approximately inverse to
-$S^+_{z_s,z_r}$. That is, with the domain and range described above,
-$S^+_{z_s,z_r}$ is approximately unitary.
-
-\section{Efficient Preconditioned CG}
-
-\cite{HouSymes:EAGE16} demonstrated a very similar preconditioner for Least Squares Migration, also for its subsurface offset extension \cite[]{HouSymes:16}, motivated by \cite{tenKroode:12}.
-These constructions all involve the Dirichlet-to-Neumann operator.
-This concept also turns up in hidden form in the work of Yu Zhang and collaborators on true amplitude migration \cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,Zhang YuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.
-
 The discussion so far can be summarized as follows: an efficient pseudoinverse for $S[c]$ is the weighted adjoint operator

### 2021-05-29 repro/papers/sseprecond/paper.tex
interim

@@
 migration
-\cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,Zhang
-  YuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.
+\cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09}.
 
@@
 
-\plot{bml0}{width=\textwidth}{Bulk modulus, lens model. Color scale is
-in GPa. Positions of point source and receiver line indicated.}
-
 The point source pressure data generated by this configuration for the
@@
 
-\plot{recphh0}{width=\textwidth}{Output pressure gather for point source in 
-  homogeneous medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-\plot{recplh0}{width=\textwidth}{Output pressure gather for point source in 
-  refractive (lens) medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
 While inversion of pressure data is the main object of this exercise,
@@
 the corresponding gathers in Figures \ref{fig:recvzhh0} and \ref{fig:recvzlh0}.
-
-\plot{recvzhh0}{width=\textwidth}{Output vertical velocity gather for point source in 
-  homogeneous medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-\plot{recvzlh0}{width=\textwidth}{Output vertical velocity gather for point source in 
-  refractive (lens) medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
 
@@
   \label{eqn:deflam}
-  \Lambda^{\pm}_{z_s}\phi = -[v^{\pm}_{1,z}]|_{z=z_s}
+  \Lambda^{\pm}_{z_s}\phi = -[v^{\pm}_{z}]|_{z=z_s}
 \end{equation}
@@
   \label{eqn:hfcondn}
-  h_s =\Lambda^{\pm}_{z_s}f_s,
+  h_s =-\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,
 \end{equation}
@@
   \label{eqn:snull}
-  {\cal S}^{\pm}_{z_s,z_r}(-\Lambda^{\pm}_{z_s}f_s,f_s)^T \approx 0.
+  {\cal S}^{\pm}_{z_s,z_r}\left(-\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,f_s\right)^T \approx 0.
 \end{equation}
@@
 
-For a point source at $z_d=3500$ m, $x_d=3500$ m, same bandpass filter
-wavelet as used for prior examples, causal data on $z=z_s=3000$ m are
-shown in Figures \ref{fig:dsrcphh0} and \ref{fig:dsrcvzhh0}. Since the
+Since the
 mechanical parameters in the homogeneous and lens models are the same
@@
 source functions $f_s$ and $h_s$ as in the system \ref{eqn:awepm},
-with $h_s=0$ and $f_s=0$ respectively.
+with $h_s=0$ and $f_s=0$ respectively. For a point source at $z_d=3500$ m, $x_d=3500$ m, same bandpass filter
+wavelet as used for prior examples, causal data on $z=z_s=3000$ m are
+shown in Figures \ref{fig:dsrcpghh0} and \ref{fig:dsrcvzghh0}.
 
@@
 
-\plot{dsrcpghh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
-  pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
-  bandpass filter source.}
-
-\plot{dsrcvzghh0}{width=\textwidth}{Trace $P_sv_z^+$ on $z=z_s=3000$ m of
-  vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
-  bandpass filter source.}
-
-\plot{drecphh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the homogeneous model described in the 
-  text, with point pressure source (constitutive defect) at $z_d=3500$
-  m, $x_d=3500$ m.}
-
-\plot{dfwdphh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the homogeneous model described in the 
-  text, with extended pressure source (constitutive defect) on $z=z_s=3000$ m given by $-2$
-  times the field depicted in Figure \ref{fig:dsrcvzhh0}
-  ($h_s=-2P_sv_z^+$) and zero velocity source (vertical load) 
-  ($f_s=0$).}
-
-\plot{daltphh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the homogeneous model described in the 
-  text, with extended velocity source (vertical load) on $z=z_s=3000$ m given by $-2$
-  times the field depicted in Figure \ref{fig:dsrcphh0}
-  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
-  ($h_s=0$).}
-
-\plot{drecplh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the lens model described in the 
-  text, with point pressure source (constitutive defect) at $z_d=3500$
-  m, $x_d=3500$ m.}
-
-\plot{dfwdplh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the lens model described in the 
-  text, with extended pressure source (constitutive defect) on $z=z_s=3000$ m given by $-2$
-  times the field depicted in Figure \ref{fig:dsrcvzhh0}
-  ($h_s=-2P_sv_z^+$) and zero velocity source (vertical load) 
-  ($f_s=0$).}
-
-\plot{daltplh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the lens model described in the 
-  text, with extended velocity source (vertical load) on $z=z_s=3000$ m given by $-2$
-  times the field depicted in Figure \ref{fig:dsrcphh0}
-  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
-  ($h_s=0$).}
-
-\plot{drecphh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$)
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
-  (red).}
-
-\plot{daltplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$)
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
-  (red).}
-
 \section{Approximate Inversion}
@@
 \[
-\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1 f_s \approx \Pi_0{\cal
-  S}^+_{z_s,z_r}Pi_0\Lambda^+_{z_s}f_s
+\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx \frac{1}{2}\Pi_0{\cal
+  S}^+_{z_s,z_r}\Pi_0\Lambda^+_{z_s}f_s
 \]
@@
   \label{eqn:lamidea}
-  = S^++_{z_s,z_r}\Lambda^+_{z_s}f_s.
+  = \frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
 \end{equation}
@@
 rays that make an angle with the vertical bounded below by a common
-minimum angle. Such rays leave $\Omega$ within a common maximum time.
-
-An immediate consequence of the downgoing assumption stated above is that in the
+minimum angle. Such rays leave $\Omega$ within a common maximum
+time. Consequently (Appendix B), in the
 slab $z_s<z<z_r$, the field $(p^+,\bv^+)$ approximates the solution of an
-anti-causal evolution equation. Indeed, for any choice of $T$, define
+anti-causal evolution equation. Choose $\chi(t)$ to be a smooth function
+that is $= 0$ for $t \gg 0$ and $=1$ at times when near rays carrying
+high-frequency energy in $(p^+,\bv^+)$ cross $z=z_r$. Define 
 $(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
-$\Omega \times (-\infty,T]$ of
+$\Omega \times \bR$ of
 \begin{eqnarray}
@@
   \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
-  \tilde{p}^- & =& 0,  \mbox{ for } t=T,\nonumber\\ 
-  \tilde{\bv}^- & = & 0 \mbox{ for } t=T,\nonumber\\
-  P_r\tilde{p}^- &=& P_rp^+ \mbox { for } t \le T . 
+  \tilde{p}^- & =& 0,  \mbox{ for } t \gg 0\\ 
+  \tilde{\bv}^- & = & 0 \mbox{ for } t \gg 0\\
+  P_r\tilde{p}^- &=& \chi P_rp^+ . 
 \end{eqnarray}
-That is, $\tilde{p}^-$ has the same boundary value as
-$p^+$ for $t<T$, but $(\tilde{p}^-,\tilde{\bv}^-)$ vanishes at
-$t=T$. Suppose that $T$ is chosen so that the rays carrying
-significant energy in $(p^+,\bv^+)$ have left the half-space $\Omega$
-at time $T$. Then
-
-As will be shown later, under the downgoing assumption, $P_r p^+
-\approx P_r \tilde{p}^+$.
-Therefore $(\delta p,\delta \bv)=(\tilde{p}^+-\tilde{p}^-,\tilde{\bv}^+-\tilde{\bv}^-)$
-differs negligibly from the solution of
-\begin{eqnarray}
-\label{eqn:deltaawe}
-  \frac{1}{\kappa}\frac{\partial \delta p}{\partial t} & = & -
-                                                               \nabla
-                                                               \cdot
-                                                               \delta \bv, \nonumber \\
-  \rho\frac{\partial \delta \bv}{\partial t} & = & - \nabla \delta p,\nonumber \\
-  \delta p & =& \tilde{p}^+,  \mbox{ for } t=T,\nonumber\\ 
-  \delta \bv & = & \tilde{\bv}^+ \mbox{ for } t=T,\nonumber\\
-  P_r\delta p & = & 0,
-\end{eqnarray}
-in $\Omega \times \bR$. Since
-$E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](T) \rightarrow 0$ as $T
-\rightarrow \infty$, the usual energy estimate implies that the energy $\delta
-E$ in this field,
-\[
-  \delta E_{\Omega}(t) = \frac{1}{2} \int_{\Omega} \,d\bx \,
- \left(\frac{(\delta p)^2}{\kappa} + \rho |\delta \bv|^2\right)
-\]
-satisfies $\delta E_{\Omega}(t) \rightarrow 0$ as $T \rightarrow
-\infty$. This decay does not generally imply that $P_s\delta p, P_s \delta \bv)
-\rightarrow 0$ (for instance) as $T \rightarrow \infty$, since the trace operator
-$P_s$ is not continuous, but it does so imply for purely downgoing
-fields, as will be demonstrated later, so
-\begin{equation}
-  \label{eqn:approxinv0}
-  \lim_{T \rightarrow \infty} P_s\tilde{v}_z^- \approx \lim_{z \rightarrow
-    0^+}\tilde{v}_z^+ \approx \lim_{z \rightarrow 0^+}v^+_z
-\end{equation}
-
-Recall (equations \ref{eqn:deflam}, \ref{eqn:jump0}) that
-\[
-  \Lambda^+_{z_s} P_s p^+ \approx 2 \lim_{z \rightarrow z_s^+}v^+_z 
-\]
-and that with $h_s= \Lambda^+_{z_s}P_sp^+$,
-\[
-  S^+_{z_s,z_r}h_s = P_r p^+
-\]
-Together with the last two equations, the relation
-\ref{eqn:approxinv0} implies that
-\begin{equation}
-  \label{eqn:approxinv1}
-  S^+_{z_s,z_r}(2 P_s \tilde{v}_z^-) \approx P_rp^+.
-\end{equation}
-As shown in the last section, the solution of the system
-\ref{eqn:revawe} differs negligibly from the anti-causal solution $(p^-_2,\bv^-_2)$
-of the system \ref{eqn:sawe2} with (i) $z_s$ replaced by $z_r$, and
-(ii) $\psi = [p^-_2]|_{z=z_r} = -2 \lim_{z \rightarrow z_r^-} p^-_2
-\approx -2 P_rp^+$. Therefore (definitions \ref{eqn:sdef}, \ref{eqn:vdef})
-\begin{equation}
-  \label{eqn:approxinv2}
-  P_s \tilde{v}_z^- \approx V^-_{z_r,z_s}(-2 P_rp^+).
-\end{equation}
-Taken together, relations \ref{eqn:approxinv1} and
-\ref{eqn:approxinv2} yield the first equation in the first main result of this section:
+That is, $\tilde{p}^-$ has the same boundary value on $z=z_r$ as
+$p^+$, except for low-frequency residue that is muted by
+$\chi$. Therefore
+$p^+ \approx \tilde{p}^-, \bv^+ \approx \tilde{\bv}^-$ near
+$z=z_r$. Since the right-hand sides in the system \ref{eqn:awepm} are
+singular only on $z=z_s$, and the high-frequency components of
+$(p^+,\bv^+)$ are carried by downgoing rays, these differ negligibly
+from the the high-frequency components of
+$(\tilde{p}^-,\tilde{\bv}^-)$ in the space-time slab $z_s<z<z_r$, and
+the approximation holds throughout this region. In particular
+$P_sv^+_z \approx P_s \tilde{v}^-_z$. In view of the relation
+\ref{eqn:tracejump10},
+\begin{equation}
+  \label{eqn:tildevtohsubs}
+  -2P_s\tilde{v}^-_z \approx h_s,
+\end{equation}
+so solution
+of the system \ref{eqn:revawe} followed by restriction to $z=z_s$ and
+multiplication by $-2$ 
+approximately inverts the map $S^+{z_s,z_r}: h_s \mapsto P_rp^+$.
+
+Next observe that in view of the relation \ref{eqn:tracejump20}, and
+the downgoing nature of the ray system carrying the high frequency
+energy in $(p^+,\bv^+)$, the field $(\tilde{p}^-,\tilde{\bv}^-)$ is
+actually the restriction to $z<z_r$ of the anti-causal solution of \ref{eqn:awepm}
+with $z_s$ replaced by $z_r$, zero constitutive defect, and vertical
+load given by the jump in pressure at $z=z_r$ - for this field, use
+the same notation. Continuity of vertical
+velocity $\tilde{v}^-_z$ at $z=z_r$ implies that the vertical load is
+\[
+  f_r = -[\tilde{p}^-]|_{z=z_r} = 2 P_r \tilde{p}^- = 2 P_r p^+
+\]
+(from the definition \ref{eqn:defsamp}, $P_r$ is the limit from the
+left). Thus
+\[
+  P_s \tilde{v^-_z} \approx V^-_{z_r,z_s}(2 P_rp^+) \approx
+  2V^-_{z_r,z_s}S^+_{z_r,z_s}h_s.
+\]
+so
+\[
+  h_s \approx -2 P_s v^+_z \approx -2 P_s \tilde{v}^-_z \approx
+  -4V^-_{z_r,z_s}S^+_{z_r,z_s}h_s
+\]
+Combine this observation with \ref{eqn:tildevtohsubs} to obtain
+\[
+ -4  V^-_{z_r,z_s} S^+_{z_s,z_r}  \approx  I,
+\]
+This relation combines with the identity \ref{eqn:trtrcomp} to
+yield the first main result of this section:
 \begin{eqnarray}
   \label{eqn:approxinv}
-  V^-_{z_r,z_s} S^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
-  S^-_{z_r,z_s} V^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
-  V^+_{z_s,z_r} S^-_{z_r,z_s} & \approx & -\frac{1}{4}I, \nonumber\\
-  S^+_{z_s,z_r} V^-_{z_r,z_s} & \approx & -\frac{1}{4}I.\\
+  (V^+_{z_s,z_r})^T S^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
+  (S^+_{z_s,z_r})^T V^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
+  V^+_{z_s,z_r} (S^+_{z_s,z_r})^T & \approx & \frac{1}{4}I, \nonumber\\
+  S^+_{z_s,z_r} (V^+_{z_s,z_r})^T & \approx & \frac{1}{4}I.\\
 \end{eqnarray}.
-The second equation follows by an exactly analogous argument, and the
-last two by time reversal and interchange of the roles of $z_s$ and
-$z_r$.
-
-Having identified approximate two-sided inverses for $S^{\pm},
-V^{\pm}$, the approximate null space identity \ref{eqn:lamnull} leads
-to a relation between $\Lambda^{\pm}$ and ${\cal S}^{\pm}$ that
-provides the foundation for practical computations. Applying $\Pi_0$
-to both sides of \ref{eqn:lamnull} gives
-\[
- \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_0^T
- \Lambda^+_{z_s} = S^+_{z_s,z_r} \Lambda^+_{z_s}.
-\]
-From the first approximate identity in \ref{eqn:approxinv},
+The second equation is simply the transpose of the first, and the
+last two follow by by an exactly analogous argument using time
+reversal and interchange of the roles of $z_s$ and$z_r$.
+
+The conclusion is significant enough to merit restating in English:
+provided that high-frequency energy in the various fields is carried
+along downgoing ray fields, the transpose of $V^+$ is an approximate
+inverse to $S^+$, modulo a factor of 4. To recover the pressure source
+$h_s$ generating a pressure gather $P_rp$ at $z=z_r$, multiply the
+latter by -2, then apply the transpose of $V^+_{z_s,z_r}$ to this
+gather, reading out a vertical velocity field at $z=z_s$. Multiply
+again by -2 and you have a high-frequency approximation to $h_s$.
+
+I have applied this procedure to the pressure gathers shown in Figures
+\ref{fig:recphh0} and \ref{fig:recplh0}, generated by a point source
+at $z_s=3000$ m, $x_s=3500$ m, propagating in the homogeneous and lens
+models respectively. The source time dependence is again a
+[1, 2.5, 10, 12.5] Hz trapezoidal bandpass filter. As noted earlier,
+a point source at depth $z=z_s$ generates a non-downgoing acoustic
+field in the slab $z_s < z < z_r$. However the pressure gather at
+$z=z_r$ contains only downgoing energy after a suitable mute: this is
+evident in the figures. Therefore one would expect inversion of this
+data to produce an aperture-limited (in effect, dip-filtered) version
+of the point source.
+
+Figure \ref{fig:srcvzghh0} shows the approximate inversion suggested
+by the first equation in \ref{eqn:approxinv}. The result indeed looks
+like a dip-filtered point source. To see that this is indeed an
+inversion, I apply $S^{+}_{z_s,z_r}$ to the source shown in Figure
+\ref{fig:srcvzghh0}, to produce the re-simulation shown in Figure
+\ref{fig:fwdphh0}. Comparison with the gather in Figure
+\ref{fig:recphh0} via a difference plot \ref{fig:diffrecphh0} verifies
+that a very close match has been obtained, so the source in Figure
+\ref{fig:srcvzghh0} is indeed a satisfactory inversion, even though
+the original point source has not been recovered: the difference is in
+the null space of the modeling operator, arising from the limited
+aperture of the acquisition geometry.
+
+As mentioned at the beginning of this section, the availability of approximate two-sided inverses for $S^{+},
+V^{+}$ leads to a practical computation for $\Lambda^{+}$: apply
+$4(V^{+}_{z_s,z_r})^T$ to both sides of equation \ref{eqn:lamidea} and
+use the first equation in the list \ref{eqn:approxinv} to get
 \begin{equation}
   \label{eqn:lamident}
-  V^{-}_{z_r,z_s}\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx -\frac{1}{4}\Lambda^+_{z_s}.
+  \Lambda^+_{z_s} \approx 8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
 \end{equation}
@@
 reading off the pressure trace on $z=z_r$, identifying it once more as
-a source for velocity, propagating it backwards in time from $z_r$ to
+a point load (source for velocity), propagating it backwards in time from $z_r$ to
 $z_s$, and finally reading off the velocity trace, interpreted as a
-pressure evolution source on $z_s$. I will give a detailed description
-of this computation in the section on numerical results.
+pressure evolution source on $z_s$.
 
@@
 Computation of the transpose of $\Lambda^+$ (exact, not approximate in
-the high frequency sense) will be important in the
-construction of a preconditioner. The relation \ref{eqn:lamident} also
-provides an computation for this operator. The idea is to treat the
-left-hand side of equation \ref{eqn:lamident} as an approximation to
-$\Lambda$, and compute its exact transposes via the adjoint state
+the high frequency sense) is critical to the successful
+construction of the preconditioner. The relation \ref{eqn:lamident}
+does not provide a computation for this operator. However set
+\begin{equation}
+  \label{eqn:lamtilde}
+  \tilde{\Lambda}^+_{z_s} = 8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
+\end{equation}
+Then \ref{eqn:lamident} can be rewritten
+\[
+  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
+\]
+Since successful preconditioning requires only approximate inversion,
+use of $\tilde{\Lambda}^+_{z_s} $ in place of $\Lambda^+_{z_s}$ will
+still yield a working preconditioner, and the former can be transposed
+to machine precision via the definition \ref{eqn:lamtilde} and the adjoint state
 method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
@@
   \label{eqn:lamtransp}
-   \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r} \approx -\frac{1}{4}(\Lambda^+_{z_s})^T.
-\end{equation} 
-The left-hand sides of \ref{eqn:lamident} and \ref{eqn:lamtransp} will
-simply be substituted for $\Lambda$ and its transpose in the
-preconditioner construction: preconditioners must be (precisely)
-self-adjoint, but need only be approximated to perform well.
+   (\tilde{\Lambda}^+_{z_s})^T = 8 \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r}
+\end{equation}
 
@@
 
+\plot{bml0}{width=\textwidth}{Bulk modulus, lens model. Color scale is 
+in GPa. Positions of point source and receiver line indicated.}
+
+\plot{recphh0}{width=\textwidth}{Output pressure gather for point source in 
+  homogeneous medium, configuration as described in Figure \ref{fig:bml0} and in
+  the text.}
+
+\plot{recplh0}{width=\textwidth}{Output pressure gather for point source in 
+  refractive (lens) medium, configuration as described in Figure \ref{fig:bml0} and in
+  the text.}
+
+\plot{recvzhh0}{width=\textwidth}{Output vertical velocity gather for point source in 
+  homogeneous medium, configuration as described in Figure \ref{fig:bml0} and in
+  the text.}
+
+\plot{recvzlh0}{width=\textwidth}{Output vertical velocity gather for point source in 
+  refractive (lens) medium, configuration as described in Figure \ref{fig:bml0} and in
+  the text.}
+
+\plot{dsrcpghh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
+  pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
+  bandpass filter source. Scaled to represent point load (velocity
+  source). }
+
+\plot{dsrcvzghh0}{width=\textwidth}{$\Lambda^+_{z_s}P_sp^+$ = scaled
+  version of trace $P_sv_z^+$ on $z=z_s=3000$ m of
+  vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
+  bandpass filter source.}
+
+\plot{drecphh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the homogeneous model described in the 
+  text, with point pressure source (constitutive defect) at $z_d=3500$
+  m, $x_d=3500$ m.}
+
+\plot{dfwdphh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the homogeneous model described in the 
+  text, with extended pressure source (constitutive defect) $h_s$ on $z=z_s=3000$ m given by  field depicted in Figure \ref{fig:dsrcvzghh0}
+  ($h_s=-2P_sv_z^+=\Lambda^+_{z_s}P_sp^+$) and zero velocity source (vertical load) 
+  ($f_s=0$).}
+
+\plot{daltphh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the homogeneous model described in the 
+  text, with extended velocity source (vertical load) $f_s$ on $z=z_s=3000$
+  m given by the field depicted in Figure \ref{fig:dsrcpghh0}
+  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
+  ($h_s=0$).}
+
+\plot{drecplh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the lens model described in the 
+  text, with point pressure source (constitutive defect) at $z_d=3500$
+  m, $x_d=3500$ m.}
+
+\plot{dfwdplh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the lens model described in the 
+  text, with extended pressure source (constitutive defect) on $z=z_s=3000$ m given bythe field depicted in Figure \ref{fig:dsrcvzghh0}
+   ($h_s=-2P_sv_z^+=\Lambda^+_{z_s}P_sp^+$) and zero velocity source (vertical load) 
+  ($f_s=0$).}
+
+\plot{daltplh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the lens model described in the 
+  text, with extended velocity source (vertical load) on $z=z_s=3000$
+  m given by the field depicted in Figure \ref{fig:dsrcpghh0}
+  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
+  ($h_s=0$).}
+
+\plot{drecphh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$)
+  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
+  (red).}
+
+\plot{daltplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$)
+  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
+  (red).}
+
+\plot{recphh0}{width=\textwidth}{Pressure gather from point source at
+  $z_s=3000$ m, $x_s=3500$ m, homogeneous model. Mute applied near cable ends.}
+
+\plot{srcvzghh0}{width=\textwidth}{Approximate inversion of the gather
+  in Figure \ref{fig:recphh0} suggested by the first equation in
+  display \ref{eqn:approxinv}. Scaled version of output $v_z$ field
+  obtained by applying transpose of $V^+_{z_s,z_r}$.}
+
+\plot{fwdphh0}{width=\textwidth}{Pressure gather produced from source
+  shown in Figure \ref{fig:srcvzghh0}.}
+
+\plot{diffrecphh0}{width=\textwidth}{Difference between gathers
+  displayed in Figures \ref{fig:recphh0} and \ref{fig:fwdphh0},
+  plotted on same color scale.}
+
+\plot{invvzghh0}{width=\textwidth}{Inverted pressure source (= scaled
+  $v_z$) gather obtained by another application of $(V^+_{z_s,z_r})^T$
+  to data shown in Figure \ref{fig:fwdphh0}. Compare Figure
+  \ref{fig:srcvzghh0}.}
+
 \bibliographystyle{seg}

### 2021-05-31 repro/papers/sseprecond/paper.tex
interim

@@
   \label{eqn:snull}
-  {\cal S}^{\pm}_{z_s,z_r}\left(-\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,f_s\right)^T \approx 0.
+  {\cal S}^{\pm}_{z_s,z_r}\left(\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,f_s\right)^T \approx 0.
 \end{equation}
@@
 predicted by equation \ref{eqn:snull}, is confirmed by trace
-comparisons in figures \ref{fig:drecphh0tr21}-\ref{fig:daltplh0tr81}.
-
-\section{Approximate Inversion}
-
-Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
-efficient iterative solution of the linear subproblem
-\ref{eqn:esis}. Direct computation, for instance by solving
-\ref{eqn:sawe2} and reading off $P_sv^{\pm}_{2,z}$, turns out to be
-numerically ill-behaved. The relation \ref{eqn:snull} provides and
-alternative approach, {\em provided} that an accurate approximate
-inverse to $S^+_{z_s,z_r}$ is available. For instance, the first row
-of the vector equation \ref{eqn:snull} reads (slightly rearranged):
-\[
-\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx \frac{1}{2}\Pi_0{\cal
-  S}^+_{z_s,z_r}\Pi_0\Lambda^+_{z_s}f_s
-\]
-\begin{equation}
-  \label{eqn:lamidea}
-  = \frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
-\end{equation}
-An approximate inverse for $S^++_{z_s,z_r}$ would permit (approximate)
-solution of this equation for $\Lambda^+_{z_s}f_s$.
+comparisons in figures \ref{fig:drecphh0tr21}-\ref{fig:daltphh0tr81}
+and
+\ref{fig:drecplh0tr21}-\ref{fig:daltplh0tr81}.
+
+\section{Time Reversal}
 
@@
 I have applied this procedure to the pressure gathers shown in Figures
-\ref{fig:recphh0} and \ref{fig:recplh0}, generated by a point source
-at $z_s=3000$ m, $x_s=3500$ m, propagating in the homogeneous and lens
-models respectively. The source time dependence is again a
-[1, 2.5, 10, 12.5] Hz trapezoidal bandpass filter. As noted earlier,
-a point source at depth $z=z_s$ generates a non-downgoing acoustic
-field in the slab $z_s < z < z_r$. However the pressure gather at
-$z=z_r$ contains only downgoing energy after a suitable mute: this is
-evident in the figures. Therefore one would expect inversion of this
-data to produce an aperture-limited (in effect, dip-filtered) version
-of the point source.
-
-Figure \ref{fig:srcvzghh0} shows the approximate inversion suggested
-by the first equation in \ref{eqn:approxinv}. The result indeed looks
-like a dip-filtered point source. To see that this is indeed an
-inversion, I apply $S^{+}_{z_s,z_r}$ to the source shown in Figure
-\ref{fig:srcvzghh0}, to produce the re-simulation shown in Figure
-\ref{fig:fwdphh0}. Comparison with the gather in Figure
-\ref{fig:recphh0} via a difference plot \ref{fig:diffrecphh0} verifies
-that a very close match has been obtained, so the source in Figure
-\ref{fig:srcvzghh0} is indeed a satisfactory inversion, even though
-the original point source has not been recovered: the difference is in
-the null space of the modeling operator, arising from the limited
-aperture of the acquisition geometry.
-
-As mentioned at the beginning of this section, the availability of approximate two-sided inverses for $S^{+},
-V^{+}$ leads to a practical computation for $\Lambda^{+}$: apply
+\ref{fig:drecphh0} and \ref{fig:drecplh0}, generated by a point source
+at $z_d=3500$ m, $x_s=3500$ m, propagating in the homogeneous and lens
+models respectively. The source time dependence is again a [1, 2.5,
+10, 12.5] Hz trapezoidal bandpass filter. As noted earlier, in the
+region $z < z_s=3000 m$, these fields are downgoing and can be
+regarded as the result of either pressure or velocity source at $z=z_s$: the
+pressure source $h_s$ (Figure \ref{fig:dhshh0}) is -2 times the
+vertical velocity gather depicted in Figure \ref{fig:dsrcvzhh0}, the
+velocity source $f_s$ (Figure \ref{fig:dfshh0}) is -2 times the
+pressure gather depicted in Figure \ref{fig:dsrcphh0}.
+
+Figure \ref{fig:dinvhshh0} shows the approximate inversion (via the
+first equation in display \ref{eqn:approxinv}) of the
+pressure gather shown in Figure \ref{fig:drecphh0}, inverted in the homogeneous model
+used to generate the data. The result is a scaled, dip-filtered
+version of the pressure source shown in Figure \ref{fig:dhshh0}.
+The dip filter effect is unavoidable and is caused by the aperture
+limitation of the acquisition geometry: the steeper dips in the source
+gather \ref{fig:dhshh0} do not contribute to the data, and are not
+recovered in the inversion.
+
+To see that this is indeed an inversion, I apply $S^{+}_{z_s,z_r}$ to
+the source shown in Figure \ref{fig:dinvhshh0}, to produce the
+re-simulation shown in Figure \ref{fig:drerecphh0}. Simulation is carried
+out in the homogeneous model used in the inversion. Comparison with the
+gather in Figure \ref{fig:drecphh0} via a difference plot
+\ref{fig:ddiffrecphh0} verifies that a very close match has been
+obtained, so the source in Figure \ref{fig:dinvhshh0} is indeed a
+satisfactory inversion, even it differs considerably from the original
+point source: the difference lies in an approximate null space of the modeling
+operator, arising from the limited aperture of the acquisition
+geometry.
+
+The remainder of this section is devoted to illustrating two important
+features of the approximate inversion developed above. The
+first is the validity of the inversion approach given in display
+\ref{eqn:approxinv} so long as the fields generated by the
+model used in simulation and inversion are downgoing. The models
+may be quite far from homogeneous, and even generate triplications,
+whereas the inversions remain accurate modulo the dip filtering effect
+of aperture. Figure \ref{fig:dinvhsll0} shows the pressure source at $z=z_s$
+inverted from the data Figure
+\ref{fig:drecpll0}, in which a triplication is obvious. Recall that
+this data is generated by propagating a point source at $z_d=3500$ m
+in the lens model. The inversion uses the same model. Figure
+\ref{fig:dinvhsll0} is an aperture-limited version of the pressure
+source at $z=z_s=3000$ m
+(Figure \ref{fig:dhsll0}) that generates this data by a dip filter, in
+the same way that the sources in Figures \ref{fig:dhsll0} and
+\ref{fig:dinvhsll0} are related. The inverted source is an accurate
+inversion: it generates the data depicted in Figure
+\ref{fig:drerecpll0}, which is very close to the input data (Figure
+\ref{fig:drecpll0}). The difference is displayed on the same color
+scale in Figure \ref{ddiffrecpll0}.
+
+The second feature is a decoupling of the models used (implicitly or
+otherwise) in simulation and those used in inversion: it is possible
+to {\em accurately invert pressure data for pressure source using the
+  wrong model}, in the sense that the inverted source will
+generate an accurate recovery of the input data re-simulated in the
+same (wrong) model. For example, inversion of the data of the previous
+example (Figure \ref{fig:drecpll0}) {\em in the homogeneous model}
+results in an inverted source gather at $z=z_s$ (Figure \ref{fig:dinvhslh0}
+that is very far from a dip-filtered version of the source used
+generate the data \ref{eqn:dhshh0}. However, this result is an
+accurate inversion: re-simulation {\em using the same (homogeneous)
+  model as used in the inversion} results in accurate recovery of the
+input pressure gather (Figure \ref{fig:drecpll0}). The difference is
+shown on the same color scale in Figure \ref{fig:ddiffrecplh0}.
+
+\section{Computing $\Lambda$}
+
+Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
+efficient iterative solution of the linear subproblem
+\ref{eqn:esis}. Direct computation of the pressure-to-source operator $\Lambda^{\pm}_{z_s}$, for instance by solving
+\ref{eqn:awepm} and reading off $P_sv^{\pm}_{z}$, turns out to be
+numerically ill-behaved. The relation \ref{eqn:snull} provides and
+alternative approach, taking advantage of the accurate approximate
+inverse to $S^+_{z_s,z_r}$ is available. For instance, the first row
+of the vector equation \ref{eqn:snull} reads (slightly rearranged):
+\[
+\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx -\frac{1}{2}\Pi_0{\cal
+  S}^+_{z_s,z_r}\Pi_0\Lambda^+_{z_s}f_s
+\]
+\begin{equation}
+  \label{eqn:lamidea}
+  = -\frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
+\end{equation}
+
+The approximate inverse construction for $S^++_{z_s,z_r}$ would permits (approximate)
+solution of this equation for $\Lambda^+_{z_s}f_s$: apply
 $4(V^{+}_{z_s,z_r})^T$ to both sides of equation \ref{eqn:lamidea} and
@@
   \label{eqn:lamident}
-  \Lambda^+_{z_s} \approx 8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
+  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
 \end{equation}
@@
 
+Figure \ref{fig:preddinvhshh0} shows the pressure source gather produced
+from the pressure data gather in Figure \ref{drecphh0} by application
+of the operator on the right-hand side of
+formula \ref{eqn:lamident}. The scale is the same as for Figures
+\ref{fig:dinvhshh0} and \ref{fig:dhshh0}. In fact the former
+appears nearly identical, and both appear to be dip-filtered versions
+of the latter. The difference plot (Figure \ref{fig:ddiffinvhshh0},
+using the same color scale, shows that the similarity is quantitative.
+
+Figures \ref{fig:preddinvhsll0} and \ref{fig:ddiffinvhsll0} accomplish
+the same comparison for the $\Lambda^+_{z_s}$ approximation in
+equation \ref{eqn:lamident}. A deeper study of the pressure-to-source
+operator shows that it is approximately dependent only on the model
+coefficients near the source surface ($z=z_s$ in this case). Since the
+homogeneous and lens models are identical near this surface, it is
+unsurprising that these figures are very close to the previous two.
+
 Computation of the transpose of $\Lambda^+$ (exact, not approximate in
@@
   \label{eqn:lamtilde}
-  \tilde{\Lambda}^+_{z_s} = 8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
+  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
 \end{equation}
@@
   \label{eqn:lamtransp}
-   (\tilde{\Lambda}^+_{z_s})^T = 8 \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r}
-\end{equation}
+   (\tilde{\Lambda}^+_{z_s})^T = -8 \Pi_1 ({\cal S}^+_{z_r,z_s})^T\Pi_0^T V^{+}_{z_s,z_r}
+ \end{equation}
+
+\section{Unitarity}
 
@@
 
-\plot{dsrcpghh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
+\plot{dsrcphh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
   pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
-  bandpass filter source. Scaled to represent point load (velocity
-  source). }
-
-\plot{dsrcvzghh0}{width=\textwidth}{$\Lambda^+_{z_s}P_sp^+$ = scaled
-  version of trace $P_sv_z^+$ on $z=z_s=3000$ m of
+  bandpass filter source. }
+
+\plot{dsrcvzhh0}{width=\textwidth}{Trace $P_sv_z^+$ on $z=z_s=3000$ m of
   vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
@@
   system \ref{eqn:awepm} in the homogeneous model described in the
-  text, with extended pressure source (constitutive defect) $h_s$ on $z=z_s=3000$ m given by  field depicted in Figure \ref{fig:dsrcvzghh0}
-  ($h_s=-2P_sv_z^+=\Lambda^+_{z_s}P_sp^+$) and zero velocity source (vertical load) 
-  ($f_s=0$).}
+  text, with extended pressure source (constitutive defect)
+  on $z=z_s=3000$ m given by $h_s=-2P_sv_z^+=\Lambda^+_{z_s}P_sp^+$,
+  where $P_sp^+$ is shown in Figure \ref{fig:dsrcphh0} and $P_sv^+_z$
+  in Figure \ref{fig:dsrcvzhh0}. In this simulation, the velocity
+  velocity source (vertical load) vanishes: $f_s=0$.}
 
@@
   system \ref{eqn:awepm} in the homogeneous model described in the
-  text, with extended velocity source (vertical load) $f_s$ on $z=z_s=3000$
-  m given by the field depicted in Figure \ref{fig:dsrcpghh0}
-  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
-  ($h_s=0$).}
+  text, with extended velocity source (vertical load) $f_s=-2P_sp^+$
+  on $z=z_s=3000$ m, where $P_sp^+$ is shown in Figure
+  \ref{fig:dsrcphh0}.  In this simulation, the zero pressure source
+  (constitutive defect vanishes: $h_s=0$.}
 
@@
 
+\plot{drecphh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$) 
+  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
+  (red).}
+
+\plot{daltphh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$) 
+  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:daltphh0}
+  (red).}
+
+\plot{drecphh0tr51}{width=\textwidth}{Overplot of traces 51 ($x=3000$) 
+  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
+  (red).}
+
+\plot{daltphh0tr51}{width=\textwidth}{Overplot of traces 51 ($x=3000$) 
+  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:daltphh0}
+  (red).}
+
+\plot{drecphh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$) 
+  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
+  (red).}
+
+\plot{daltphh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$) 
+  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:daltphh0}
+  (red).}
+
 \plot{dfwdplh0}{width=\textwidth}{Pressure gather at receiver depth 
@@
   system \ref{eqn:awepm} in the lens model described in the 
-  text, with extended pressure source (constitutive defect) on $z=z_s=3000$ m given bythe field depicted in Figure \ref{fig:dsrcvzghh0}
+  text, with extended pressure source (constitutive defect) on
+  $z=z_s=3000$ m given bythe field depicted in Figure
+  \ref{fig:dsrcvzhh0} scaled by -2 
    ($h_s=-2P_sv_z^+=\Lambda^+_{z_s}P_sp^+$) and zero velocity source (vertical load) 
@@
   text, with extended velocity source (vertical load) on $z=z_s=3000$
-  m given by the field depicted in Figure \ref{fig:dsrcpghh0}
+  m given by the field depicted in Figure \ref{fig:dsrcphh0} scaled by -2
   ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
@@
 
-\plot{drecphh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$)
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
+\plot{drecplh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$) 
+  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:dfwdplh0}
+  (red).}
+
+\plot{daltplh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$) 
+  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
+  (red).}
+
+\plot{drecplh0tr51}{width=\textwidth}{Overplot of traces 51 ($x=3000$) 
+  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:dfwdplh0}
+  (red).}
+
+\plot{daltplh0tr51}{width=\textwidth}{Overplot of traces 51 ($x=3000$) 
+  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
+  (red).}
+
+\plot{drecplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$) 
+  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:dfwdplh0}
   (red).}
@@
 
-\plot{recphh0}{width=\textwidth}{Pressure gather from point source at
-  $z_s=3000$ m, $x_s=3500$ m, homogeneous model. Mute applied near cable ends.}
-
-\plot{srcvzghh0}{width=\textwidth}{Approximate inversion of the gather
-  in Figure \ref{fig:recphh0} suggested by the first equation in
-  display \ref{eqn:approxinv}. Scaled version of output $v_z$ field
-  obtained by applying transpose of $V^+_{z_s,z_r}$.}
-
-\plot{fwdphh0}{width=\textwidth}{Pressure gather produced from source
-  shown in Figure \ref{fig:srcvzghh0}.}
-
-\plot{diffrecphh0}{width=\textwidth}{Difference between gathers
-  displayed in Figures \ref{fig:recphh0} and \ref{fig:fwdphh0},
+\plot{dinvhshh0}{width=\textwidth}{Approximate inversion via first 
+  equation in display \ref{eqn:approxinv}. Inversion in homogenous model of 
+  the pressure gather in Figure \ref{fig:drecphh0}, simulated with 
+  homogeneous model. Scaled version of output $v_z$ field obtained by 
+  applying transpose of $4V^+_{z_s,z_r}$.}
+
+\plot{drerecphh0}{width=\textwidth}{Re-simulated pressure gather
+  produced from inverted source
+  shown in Figure \ref{fig:dinvhshh0}. Simulation in homogeneous model.}
+
+\plot{ddiffrecphh0}{width=\textwidth}{Difference between gathers
+  displayed in Figures \ref{fig:drecphh0} and \ref{fig:drerecphh0},
   plotted on same color scale.}
 
-\plot{invvzghh0}{width=\textwidth}{Inverted pressure source (= scaled
-  $v_z$) gather obtained by another application of $(V^+_{z_s,z_r})^T$
-  to data shown in Figure \ref{fig:fwdphh0}. Compare Figure
-  \ref{fig:srcvzghh0}.}
+\plot{dinvhsll0}{width=\textwidth}{Approximate inversion via first 
+  equation in display \ref{eqn:approxinv}. Inversion in lens model of 
+  the pressure gather in Figure \ref{fig:drecpll0}, simulated with 
+  lens model. Scaled version of output $v_z$ field obtained by 
+  applying transpose of $4V^+_{z_s,z_r}$.}
+
+\plot{drerecphh0}{width=\textwidth}{Re-simulated pressure gather
+  produced from inverted source
+  shown in Figure \ref{fig:dinvhsll0}. Simulation in lens model.}
+
+\plot{ddiffrecphh0}{width=\textwidth}{Difference between gathers
+  displayed in Figures \ref{fig:drecpll0} and \ref{fig:drerecpll0},
+  plotted on same color scale.}
+
+\plot{dinvhslh0}{width=\textwidth}{Approximate inversion via first 
+  equation in display \ref{eqn:approxinv}. Inversion in homogeneous model of 
+  the pressure gather in Figure \ref{fig:drecplh0}, simulated with 
+  lens model. Scaled version of output $v_z$ field obtained by 
+  applying transpose of $4V^+_{z_s,z_r}$.}
+
+\plot{drerecplh0}{width=\textwidth}{Re-simulated pressure gather produced from
+  inverted source
+  shown in Figure \ref{fig:dinvhslh0}. Simulation in homogeneous model.}
+
+\plot{ddiffrecplh0}{width=\textwidth}{Difference between gathers
+  displayed in Figures \ref{fig:drecplh0} and \ref{fig:drerecplh0},
+  plotted on same color scale.}
+
+\plot{preddinvhshh0}{width=\textwidth}{Pressure source gather = image
+  of pressure-to-source operator $\Lambda^+_{z_s}$ applied to pressure gather
+  shown in Figure \ref{fig:dsrcphh0}, homogeneous model. Compare Figure
+  \ref{fig:dinvhshh0}.}
+
+\plot{ddiffinvhshh0}{width=\textwidth}{Difference between image
+  (Figure \ref{fig:predinvhshh0}) of $\Lambda^+_{z_s}$ applied to
+  pressure gather (Figure \ref{fig:dsrcphh0}) and source gather
+  (Figure \ref{fig:dinvhshh0}) inverted from receiver pressure gather
+  (Figure \ref{fig:drecphh0}), homogeneous model used in all
+  propagations. Same color scale as in Figure \ref{fig:predinvhshh0}. }
+
+\plot{preddinvhsll0}{width=\textwidth}{Pressure source gather = image
+  of pressure-to-source operator $\Lambda^+_{z_s}$ applied to pressure gather
+  shown in Figure \ref{fig:dsrcphh0}, lens model. Compare Figure
+  \ref{fig:dinvhsll0}.}
+
+\plot{ddiffinvhsll0}{width=\textwidth}{Difference between image
+  (Figure \ref{fig:predinvhslh0}) of $\Lambda^+_{z_s}$ applied to
+  pressure gather (Figure \ref{fig:dsrcpph0}) and source gather
+  (Figure \ref{fig:dinvhsll0}) inverted from receiver pressure gather
+  (Figure \ref{fig:drecplh0}), lens model used in all propagations.
+ Same color scale as in Figure \ref{fig:predinvhsll0}.}
 

### 2021-06-07 repro/papers/sseprecond/paper.tex
getting there

@@
 
-${\cal S}^{\pm}$ is not stably invertible (equation \ref{eqn:lamnull}). The
+${\cal S}^{\pm}$ is not stably invertible: its columns are
+approximately linearly dependent, as will be verified below. The
 diagonal components of ${\cal S}^{\pm}$ thus carry essentially all of
@@
 With these conventions, we can write the version of the source
-subproblem studied in this paper as
+subproblem \ref{eqn:esis} studied in this paper as
 \begin{equation}
@@
 
-It follows from the adjoint state method (see Derivations section
-below for details) that
+It follows from the adjoint state method (see Appendix A for details) that
 \begin{equation}
@@
 wavelet as used for prior examples, causal data on $z=z_s=3000$ m are
-shown in Figures \ref{fig:dsrcpghh0} and \ref{fig:dsrcvzghh0}.
+shown in Figures \ref{fig:dsrcphh0} and \ref{fig:dsrcvzhh0}.
 
@@
 multiplication by $-2$ 
-approximately inverts the map $S^+{z_s,z_r}: h_s \mapsto P_rp^+$.
+approximately inverts the map $S^+_{z_s,z_r}: h_s \mapsto P_rp^+$.
 
@@
 \[
-  f_r = -[\tilde{p}^-]|_{z=z_r} = 2 P_r \tilde{p}^- = 2 P_r p^+
+  f_r = -[\tilde{p}^-]|_{z=z_r} =-(\lim_{z\rightarrow
+    z_r^+}\tilde{p}^- - \lim_{z\rightarrow
+    z_r^-}\tilde{p}^-)
+\]
+\[
+  \approx 2 P_r \tilde{p}^- = 2 P_r p^+
 \]
@@
 regarded as the result of either pressure or velocity source at $z=z_s$: the
-pressure source $h_s$ (Figure \ref{fig:dhshh0}) is -2 times the
+pressure source gather $h_s$ (Figure \ref{fig:dhshh0})  is -2 times the
 vertical velocity gather depicted in Figure \ref{fig:dsrcvzhh0}, the
-velocity source $f_s$ (Figure \ref{fig:dfshh0}) is -2 times the
+velocity source gather $f_s$ (Figure \ref{fig:dfshh0}) is -2 times the
 pressure gather depicted in Figure \ref{fig:dsrcphh0}.
@@
 inverted from the data Figure
-\ref{fig:drecpll0}, in which a triplication is obvious. Recall that
+\ref{fig:drecplh0}, in which a triplication is obvious. Recall that
 this data is generated by propagating a point source at $z_d=3500$ m
@@
 source at $z=z_s=3000$ m
-(Figure \ref{fig:dhsll0}) that generates this data by a dip filter, in
-the same way that the sources in Figures \ref{fig:dhsll0} and
-\ref{fig:dinvhsll0} are related. The inverted source is an accurate
+(Figure \ref{fig:dhshh0}) that generates this data by a dip filter, in
+the same way that the sources in Figures \ref{fig:dhshh0} and
+\ref{fig:dinvhshh0} are related. The inverted source is an accurate
 inversion: it generates the data depicted in Figure
 \ref{fig:drerecpll0}, which is very close to the input data (Figure
-\ref{fig:drecpll0}). The difference is displayed on the same color
-scale in Figure \ref{ddiffrecpll0}.
+\ref{fig:drecplh0}). The difference is displayed on the same color
+scale in Figure \ref{fig:ddiffrecpll0}.
 
@@
 same (wrong) model. For example, inversion of the data of the previous
-example (Figure \ref{fig:drecpll0}) {\em in the homogeneous model}
+example (Figure \ref{fig:drecplh0}) {\em in the homogeneous model}
 results in an inverted source gather at $z=z_s$ (Figure \ref{fig:dinvhslh0}
 that is very far from a dip-filtered version of the source used
-generate the data \ref{eqn:dhshh0}. However, this result is an
+generate the data \ref{fig:dhshh0}. However, this result is an
 accurate inversion: re-simulation {\em using the same (homogeneous)
   model as used in the inversion} results in accurate recovery of the
-input pressure gather (Figure \ref{fig:drecpll0}). The difference is
+input pressure gather (Figure \ref{fig:drecplh0}). The difference is
 shown on the same color scale in Figure \ref{fig:ddiffrecplh0}.
 
-\section{Computing $\Lambda$}
+\section{Unitarity}
+
+The next chapter in this story recognizes the relations in display
+\ref{eqn:approxinv} as asserting the approximate unitarity of
+$S^+_{z_s,z_r}$.
+
+The matrix identity \ref{eqn:snull} implies a relation between $S, V,$
+and $\Lambda$ of some interest in itself. After minor re-arrangement, the second row of reads
+\begin{equation}
+  \label{eqn:snull2}
+-\frac{1}{2}\Pi_1{\cal S}^{\pm}_{z_s,z_r}\Pi_0^T\Lambda^{\pm}_{z_s}  \approx
+V^{\pm}_{z_s,z_r}.
+\end{equation}
+In these relations, the projection on the left picks out the vertical velocity component
+of a downgoing wavefield at $z=z_r$: that is,
+\[
+-\frac{1}{2}\Pi_1{\cal S}^{\pm}_{z_s,z_r}\Pi_0^T\Lambda^{\pm}_{z_s}P_sp^+
+=-\frac{1}{2}P_r v_z^+,
+\]
+where $(p^+, \bv^+)$ solve the system \ref{eqn:awepm} with $f_s=0$ and
+$h_s = \Lambda^{\pm}_{z_s}P_sp^+$. On the other hand, from relation
+\ref{eqn:tracejump10},
+\[
+  P_r v_z^+ = -\frac{1}{2}\Lambda^+_{z_r}P_r p^+
+\]
+where
+\[
+  P_r p^+ = \Pi_0{\cal S}^{+}_{z_s,z_r}\Pi_0^T\Lambda^{+}_{z_s}P_s
+  p^+
+\]
+\[
+  = S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_sp^+
+\]
+Therefore combining the last two equations with \ref{eqn:snull2},
+obtain
+\begin{equation}
+  \label{eqn:sv}
+  \frac{1}{4}\Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s} = V^+_{z_s,z_r}.
+\end{equation}
+This is the promised relation.
+
+A minor extension of an earlier example, based on the gathers
+displayed in Figures \ref{fig:dsrcphh0} ($P_sp^+$) and
+\ref{fig:dsrcvzhh0} ($P_sv_z^+$) illustrates the identity
+\ref{eqn:sv}.
+The pressure gathers at $z=z_r$ produced by application of
+$S^+_{z_r,z_s}$ to $h_s=-2P_s v^+_z$, and $\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T$ to $f_s
+=-2P_sp^+$, have already been displayed (Figures \ref{fig:dfwdphh0}
+and \ref{fig:daltphh0}), and are identical to several digits. The same 
+is true of the vertical velocity gathers, shown in \ref{fig:dfwdvzhh0}
+and \ref{fig:daltvzhh0}; their difference, plotted on the same scale,
+appears in Figure \ref{fig:dsvcomphh0}. Once again, this relation
+expresses the interchangeability of pressure and velocity sources,
+related by $\Lambda$.
+
+Relation \ref{eqn:tracejump10} characterizes
+$\Lambda^+_{z_s}$ as connecting the pressure and velocity components
+of downgoing fields restricted to $z=z_s$. That is, the pressure
+source gather
+$h_s=\Lambda^{+}_{z_s}P_s p^+ = -2
+P_{z_s}v_z$, displayed in Figure \ref{fig:dhshh0}, is the image of the
+pressure gather in Figure \ref{fig:dsrcphh0} under
+$\Lambda^{+}_{z_s}$.
+The pressure gather
+\ref{fig:dfwdphh0} is the image of this pressure source gather under
+$S^+_{z_s,z_r}$. The corresponding vertical velocity gather
+(Figure \ref{fig:dfwdvzhh0}) is $-1/2$ times $\Lambda^+_{z_r}P_r
+p$. Therefore scaling the data in Figure \ref{fig:dfwdvzhh0} by
+$-2$ produces
+$Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_s p^+$. On the other
+hand, figure \ref{fig:daltvzhh0} shows the result of applying
+$V^+_{z_s,z_r}$ to $f_s=-2P_s p^+$. Therefore scaling the gather in
+Figure \ref{fig:daltvzhh0} by $-\frac{1}{2}$ produces $V^+_{z_s,z_r}P_zp^+$.
+Since the data in Figures \ref{fig:dfwdvzhh0} and \ref{fig:daltvzhh0}
+are essentially identical, the relation \ref{eqn:sv} holds for this
+example.
+
+As shown in the last section, $4(V_{z_s,z_r}^+)^T$ is approximately
+inverse to $S^{+}_{z_s,z_r}$. Therefore, transposing both sides of
+equation \ref{eqn:sv} and using \ref{eqn:approxinv}, obtain
+\begin{equation}
+  \label{eqn:almostunitary}
+  4(V_{z_s,z_r}^+)^TS^+_{z_s,z_r} = [ (\Lambda^+_{z_r})^T
+  (S^{+}_{z_s,z_r})^T(\Lambda^+_{z_s})^T]S^{+}_{z_s,z_r} \approx I.
+\end{equation}
+
+The remarkable feature of the identity \ref{eqn:almostunitary} is that
+it exhibits an approximate right inverse of $S^+$ as an adjoint with
+respect to a weighted inner product - or it would, if the operators
+$(\Lambda^+)$ were symmetric positive definite. As noted earlier,
+these operators are only approximately symmetric, though they are
+positive semi-definite. That is not a great obstacle, however:
+symmetrizing them in the obvious way commits a negligible error, of
+the sort that this paper already neglects wholesale. That is,
+\begin{equation}
+  \label{eqn:unitary}
+  [ \frac{1}{2}((\Lambda^+_{z_r})^T+ \Lambda^+_{z_r})
+  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_s})^T+
+  \Lambda^+_{z_s})]S^{+}_{z_s,z_r} \approx I.
+\end{equation}
+The symmetrized $\Lambda$ operators are at least positive
+semi-definite, hence define (at least) semi-norms.
+
+Similar relations have been derived for other scattering operators,
+and have been used to accelerate iterative solutions of inverse
+scatering problems: \cite{DafniSymes:SEG18b} review some of this
+literature.
+
+\section{Accelerated Iterative Inversion}
+
+For convenience, in this section write $S$ in place of
+$S^+_{z_s,z_r}$. Also abbreviate the symmetrized $\Lambda$ operators
+using notation suggesting weight
+operators in model and data spaces:
+\begin{eqnarray}
+  W_m^{-1}&=& \frac{1}{2}((\Lambda^+_{z_s})^T+
+              \Lambda^+_{z_s}),\nonumber \\
+  W_d &=& \frac{1}{2}((\Lambda^+_{z_r})^T+ \Lambda^+_{z_r}).
+          \label{eqn:wdef}
+\end{eqnarray}
+The identification of the symmetrized $\Lambda^+_{z_s}$ as the inverse
+of another operator $W_m$ is formal, since the former operator is
+likely to have null (or nearly-null) vectors due to aperture-related
+amplitude loss. I will proceed as if this were sensible, since the
+mathematical issues can be defused and the only the application of
+$W_m^{-1}$ is required in the algorithms described here.
+
+Adopting Hilbert norms defined by the operators $W_m$ and $W_d$ in its
+domain and range respectively, the adjoint of $S$ is given by
+\begin{equation}
+\label{eqn:wadj}
+S^{\dagger} = W_m^{-1}S^TW_d,
+\end{equation}
+
+In this notation, the relation \ref{eqn:unitary} takes the form
+\begin{equation}
+  \label{eqn:wunitary}
+  S^{\dagger}S \approx I.
+\end{equation}
+That is to say, $S$ is approximately unitary with respect to the
+weighted norms defined by $W_m$ and $W_d$. Therefore a Krylov space
+method employing these norms will converge rapidly, at least for the
+well-determined components of the solution.
+
+Similar relations have been derived for other scattering operators,  
+and have been used to accelerate iterative solutions of inverse  
+scattering problems: \cite{DafniSymes:SEG18b} review some of this  
+literature.
+
+The most convenient arrangement the Conjugate Gradient (CG) algorithm
+taking advantage of the structure \ref{eqn:wadj} is the {\em
+  Preconditioned CG}.
+
+Allowing that the fit error will be measured by the data space norm,
+the least squares problem to be solved is not just $Sh \approx d$, but
+a regularized version:
+\begin{equation}
+  \label{eqn:einv}
+  \mbox{minimize}_h \|Sh-d\|^2_d + \alpha^2 \|Ah\|^2_m
+\end{equation}
+Remark: recall that the modified data space norm $\|d\|_d^2 = \langle
+d, W_d d\rangle$ has physical meaning: for acoustics, it is
+proportional to the power transmitted to the fluid by the source.
+
+The minimizer of the objective defined in equation \ref{eqn:einv}
+solves the normal equation
+\begin{equation}
+  \label{eqn:norm0}
+  (S^{\dagger}S + \alpha^2 A^{\dagger}A)h = S^{\dagger}d 
+\end{equation}
+where the weighted adjoint $S^{\dagger}$ has already been defined in equation \ref{eqn:wadj}, and $A^{\dagger}$ is the adjoint of $A$ in the weighted model space norm defined by $W_m$, namely
+\begin{equation}
+  \label{eqn:aadj}
+  A^{\dagger} = W_m^{-1}A^TW_m.
+\end{equation}
+
+Note that the normal operator appearing on the left-hand side of
+\ref{eqn:norm0} is not an approximate identity, due to the presence of
+the regularization term: the spectrum increases in spread with
+increasing $\alpha$, leading to slower convergence. Fortunately for
+the present setting, the operators $W_m^{-1}$, $A$, and $W_m$
+approximately commute (they are scalar {\em pseudodifferential}, once
+the difficulties with the definition of $W_m$, mentioned above, are
+taken care of), so that $A^{\dagger} \approx A^T$. Therefore
+\begin{equation}
+  \label{eqn:normapprox}
+  S^{\dagger}S + \alpha^2 A^{\dagger}A \approx I + \alpha^2A^TA
+\end{equation}
+Recall that $A$ is simply multiplication by the Euclidean distance to
+the physical source point $\bx_s$: $A u (\bx) = |\bx-\bx_s|u(\bx),
+A^TAu(\bx) = |\bx-\bx_s|^2u(\bx)$. So the equation $(I+\alpha^2
+A^TA)u=b$ is trivial to solve, and this is a key characteristic of a
+good preconditioner. However this observation must be combined with
+the wegithed norm structure.
+
+Rewrite the normal equation \ref{eqn:norm} as
+\begin{equation}
+  \label{eqn:norm1}
+  W_m^{-1}(S^TW_dS + \alpha^2 A^TW_mA)h = W_m^{-1}S^TW_md 
+\end{equation}
+Since $W_m$ is self-adjoint and positive definite, the common factor on both sides of \ref{eqn:norm1} can be re-written as
+\begin{equation}
+  \label{eqn:normpart}
+  Nh = (S^*S + \alpha^2 A^*A)h = S^*d 
+\end{equation}
+in which $S^*, A^*$ are the adjoints with the original (Euclidean)
+inner product in the domains but the weighted inner product in data
+space:
+\begin{eqnarray}
+  \label{eqn:sadj}
+  S^* &=& S^T W_d,\\
+  A^* &=& A^T W_m.
+\end{eqnarray}
+Note the $S^*S$ and $A^*A$ are symmetric in the Euclidean sense, so
+equation \ref{eqn:normpart} is a symmetric positive (semi-)definite
+linear system, just the sort of thing for which the 
+The Preconditioned Conjugate Gradient (``PCG'') algorithm was
+designed. PCG for solution
+of equation \ref{eqn:normpart} with preconditioner $M$is usually
+written as Algorithm 1 (see for example \cite{Golub:2012}):
+
+\begin{algorithm}[H]
+\caption{Preconditioned Conjugate Gradient Algorithm, Standard Version}
+\begin{algorithmic}[1]
+\State Choose $h_0=0$ 
+  \State $r_0 \gets S^*d$
+  \State $p_0 \gets M^{-1} r_0$
+  \State $g_0 \gets p_0$
+  \State $q_0 \gets Np_0$
+  \State $k \gets 0$
+  \Repeat
+  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle p_k,q_k\rangle}$
+  \State $h_{k+1} \gets h_k + \alpha_k p_k$
+  \State $r_{k+1} \gets r_k - \alpha_kq_k$
+  \State $g_{k+1} \gets M^{-1} r_{k+1}$
+  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
+  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
+  \State $q_{k+1} \gets Np_{k+1}$
+  \State $k \gets k+1$
+  \Until{Error is sufficiently small, or max iteration count exceeded} 
+\end{algorithmic}
+\end{algorithm}
+The iteration converges rapidly if $M^{-1}N \approx I$. This is true
+if and only if the symmetrized operator $M^{-1/2}NM^{-1/2} \approx I$,
+which is in turn true if the eigenvalues of $M^{-1/2}NM^{-1/2}$ are
+close to 1 (actually works well is most of these eigenvalues are close
+to 1, and the rest are small - which is the case for the current
+problem)..  Further, PCG is computationally effective is M is easy to
+invert.
+
+From \ref{eqn:normapprox} and \ref{eqn:norm1}, it follows that
+\[
+  W_m^{-1}(S^TW_dS + \alpha^2 A^TW_mA) \approx I + \alpha^2 A^TA.
+\]
+This observation suggests using $M=W_m(I+\alpha^2A^TA)$. This choice
+is not symmetric, but since the operators on the right-hand side are
+scalar pseudodifferential hence commute, it is equivalent to use of
+\begin{eqnarray}
+  M         &=&(I+\alpha^2A^TA)^{1/2}W_m(I+\alpha^2A^TA)^{1/2},\nonumber \\
+  M^{-1}
+            &=&(I+\alpha^2A^TA)^{-1/2}W_m^{-1}(I+\alpha^2A^TA)^{-1/2}.
+                \label{eqn:defprecond}
+\end{eqnarray}
+ With this choice, \ref{eqn:normapprox} implies that
+$M^{-1}N \approx I$, also $M$ is symmetric. As already mentioned,
+powers of $I + \alpha^2A^TA$ are trivial to compute, given the choice
+of $A$ made here. We will examine fast algorithms for computing
+$W_m^{-1}$ = the symmetrized pressure-to-source operator in the next
+section. Note that only $M^{-1}$, hence only $W_m^{-1}$, appears in
+Algorithm 1.
+
+[Example]
+
+\section{Computing and Symmetrizing $\Lambda$}
 
@@
 alternative approach, taking advantage of the accurate approximate
-inverse to $S^+_{z_s,z_r}$ is available. For instance, the first row
-of the vector equation \ref{eqn:snull} reads (slightly rearranged):
-\[
-\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx -\frac{1}{2}\Pi_0{\cal
-  S}^+_{z_s,z_r}\Pi_0\Lambda^+_{z_s}f_s
-\]
+inverse to $S^+_{z_s,z_r}$ is available. The first row of
+\ref{eqn:snull} is given explicitly in \ref{eqn:snull1}, which I
+repeat for the reader's convenience:
 \begin{equation}
   \label{eqn:lamidea}
-  = -\frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
-\end{equation}
-
-The approximate inverse construction for $S^++_{z_s,z_r}$ would permits (approximate)
+  \Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx  -\frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
+\end{equation}
+
+The approximate inverse construction for $S^++_{z_s,z_r}$ permits (approximate)
 solution of this equation for $\Lambda^+_{z_s}f_s$: apply
@@
 
-\section{Unitarity}
-
-\section{Accelerated Iterative Inversion}
-
-The discussion so far can be summarized as follows: an efficient pseudoinverse for $S[c]$ is the weighted adjoint operator
-\begin{equation}
-\label{eqn:wadj}
-S[c]^{\dagger} = W[c]_m^{-1}S[c]^TW[c]_d,
-\end{equation}
-in which 
-\begin{equation}
-\label{eqn:weights}
-W[c]_m = \kappa^{-1}\Lambda[c]^{-1}_s,\,\, W[c]_d = \kappa^{-1}\Lambda[c]_r
-\end{equation}
-Here ``$\kappa$'' means the operator ``multiply by $\kappa$''.
-
-The statement that $S[c]^{\dagger}$ is the weighted adjoint of $S[c]$ means that $S[c]^{\dagger}$ {\em is} the adjoint of $S[c]$ with respect to weighted norms 
-\begin{itemize}
-\item on the source space ($h_s$): weight $W[c]_m$
-\item on the data space ($d_r$): weight $W[c]_d$
-\end{itemize}
-Since $S[c]^{\dagger}S[c] \approx I$ in subspaces of the domain space, as shown above, $S[c]$ is approximately unitary in (subspaces of) the Hilbert spaces of source and data traces with norms defined by the weighting operators $W_m$ and $W_d$ respectively.
-Therefore a Krylov space method employing these norms will converge rapidly, at least for the well-determined components of the solution.
-
-The most convenient arrangement the CG algorithm taking advantage of the structure \ref{eqn:wadj} is the {\em Preconditioned CG}.
-Since the coefficient array $c$ will remain unchanged during the iteration, write $S$ instead of $S[c]$, etc.
-
-Allowing that the fit error will be measured by the data space norm, the least squares problem to be solved is not just $Sh \approx d$, but a regularized version:
-\begin{equation}
-  \label{eqn:einv}
-  \mbox{minimize}_h \|Sh-d\|^2_d + \alpha^2 \|Ah\|^2_m
-\end{equation}
-
-[The modified data space norm $\|d\|_d^2 = \langle d, W_d d\rangle$ has physical meaning: for acoustics, it is proportional to the power transmitted through the surface $z=z_r$.]
-
-The minimizer of the objective defined in equation \ref{eqn:einv} solves the normal equation
-\begin{equation}
-  \label{eqn:norm}
-  (S^{\dagger}S + \alpha^2 A^{\dagger}A)h = S^{\dagger}d 
-\end{equation}
-
-where the weighted adjoint $S^{\dagger}$ has already been defined in equation \ref{eqn:wadj}, and $A^{\dagger}$ is the adjoint of $A$ in the weighted model space norm defined by $W_m$, namely
-\begin{equation}
-  \label{eqn:aadj}
-  A^{\dagger} = W_m^{-1}A W_m.
-\end{equation}
-
-Introducing these definitions in the normal equation \ref{eqn:norm},
-\begin{equation}
-  \label{eqn:norm1}
-  W_m^{-1}(S^TW_dS + \alpha^2 A^TW_mA)h = W_m^{-1}S^TW_md 
-\end{equation}
-Since $W_m$ is self-adjoint and positive definite, the common factor on both sides of \ref{eqn:norm1} can be re-written as
-
-write $S^*, A^*$ for the adjoints with the original (Euclidean) inner product in the domains but the weighted inner product in data space:
-\begin{equation}
-  \label{eqn:normpart}
-  Nh = (S^*S + \alpha^2 A^*A)h = S^*d 
-\end{equation}
-using the ``partly weighted'' adjoints
-\begin{eqnarray}
-  \label{eqn:sadj}
-  S^* &=& S^T W_d,\\
-  A^* &=& A^T W_m.
-\end{eqnarray}
-The Preconditioned Conjugate Gradient (``PCG'') algorithm for solution of equation \ref{eqn:normpart} with preconditioner $W_m$ is usually written as Algorithm 1.
-
-\begin{algorithm}[H]
-\caption{Preconditioned Conjugate Gradient Algorithm, Standard Version}
-\begin{algorithmic}[1]
-\State Choose $h_0=0$ 
-  \State $r_0 \gets S^*d$
-  \State $p_0 \gets W_m^{-1}r_0$
-  \State $g_0 \gets p_0$
-  \State $q_0 \gets Np_0$
-  \State $k \gets 0$
-  \Repeat
-  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle p_k,q_k\rangle}$
-  \State $h_{k+1} \gets h_k + \alpha_k p_k$
-  \State $r_{k+1} \gets r_k - \alpha_kq_k$
-  \State $g_{k+1} \gets W_m^{-1}r_{k+1}$
-  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
-  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
-  \State $q_{k+1} \gets Np_{k+1}$
-  \State $k \gets k+1$
-  \Until{Error is sufficiently small, or max iteration count exceeded} 
-\end{algorithmic}
-\end{algorithm}
-
-The operator whose eigenvalue spectrum determines the speed of convergence here is the preconditioned normal operator
-\begin{equation}
-  \label{eqn:pno}
-  W_m^{-1}N = S^{\dagger}S +\alpha^2 A^{\dagger}A
-\end{equation}
-This isn't just an approximate identity on a subspace, as $S^{\dagger}S$ is.
-The addition of the penalty term affects the spectrum, pushing parts of it away from $1$.
-I have not yet analyzed this effect, however for the ``in-aperture'' part of the model space,
-\[
-  W_m^{-1}N \approx I +\alpha^2 W_m^{-1}A^TW_mA
-\]
-For small $\alpha$, this operator is a perturbation of the spectrum of the identity, so convergence should be rapid.
-
-In the preceding section, I showed that effective choices for the weight operators are $W_m = \Lambda_s^{-1}$ and $W_d = \Lambda_r$, which are resepctively the ``Neumann-to-Dirichlet'' operator mapping pressure to normal particle velocity on the source surface $\{z=z_s\}$, and the ``Dirichlet-to-Neumann'' operator mapping pressure to normal particle velocity on the receiver surface.
-Application of these operators are expensive and/or awkward, so should be avoided if at all possible.
-
-A very important point: the weighted ``partial'' adjoint $S^*=S^TW_d$ appears {\em only} in combination with $S$ in the algorithm listed above, {\em except} for the very first line.
-That is, one can calculate $W_dS$ followed by $S^T$, rather than $S$ followed by $S^TW_d$, provided that you can take care of the first line.
-$W_dS$ For any variant of elasticity, $W_d$ relates two combinations of dynamical field components - for acoustics, it returns the normal velocity for a given pressure field on the receiver surface.
-In the solution of the dynamical equations (for instance \ref{eqn:awe}), all components are calculated.
-Therefore obtaining the output of $W_d$ amounts to extracting the appropriate field combination.
-For acoustics, simply record the normal velocity rather than the pressure: then you have computed $W_dS$.
-Therefore no additional expensive computations are required.
-It is only the very first instance of $S^*$, on the first line, where an actual computation of $W_d$ is required, unless cheap workarounds like that explained by \cite{HouSymes:15} are available.
-
-\section{Prototype Numerical Examples}
-I present a collection of simple examples that illustrate the features of the surface source extension claimed in preceding sections.
-
-I used the IWAVE acoustic staggered grid package to carry out these calculations.
-This package implements (2,2k) schemes for k=1,2,..., and outputs traces (of either velocity or pressure at any point in space via multilinear interpolation.
-The discretized modeling operator is thus of second order accuracy, though as usual I have used higher order in space to reduce grid dispersion.
-
-Source injection is implemented as the adjoint of trace sampling, resulting in another second-order error [REFERENCES].
-
-The data is a single shot gather, with a source at coordinates $x_s=z_s=3000$ m (units of length are meters in all cases).
-The receiver line occupies $1500 \le x_r \le 5500$ m, with receiver depth $z_r=1000$ m.
-Extended sources occupy $1500 \le x_s \le 5500$ m, with the same depth $z_s=3000$ m as the ``physical'' source used to generate the data.
-This region turns out to be adequate to represent the extended sources that approximately invert the data, for the cases examined below.
-An algorithm to automatically identify an appropriate region can be based on the ideas developed by \cite{Fu:Geo17}.
-
-I have used absorbing boundary conditions (split-field PML) on all four sides of the 4000 m (vertical) $\times$ 8000 m (horizontal) simulation domain.
-Evidently inclusion of a free surface is important to the application of the ideas explained here to diving wave marine data, and I have not addressed the necessary modifications here.
-
-The Dirichlet-to-Neumann operator $\Lambda$ is an essential part of the inner problem preconditioner just presented. Implementation can be accomplished in several ways:
-\begin{itemize}
-\item \cite{tenKroode:12} suggests using a one-way operator;
-\item if both pressure and normal partical velocity are measured (or simulated), then the two are related by $\Lambda$ and the velocity component can simply be used as the output;
-\item presence of a free surface implies all of the usual problems, such as the need for removal of receiver-side ghosts.
-On the other hand, if the free surface is within a quarter-wavelength throughout the useful bandwidth of the data, then the ghosted data differs from $\Lambda p$ by a time integration and a scale factor, a fact used to good effect by \cite{HouSymes:15}.
-\end{itemize}
-
-In the examples, I have used the second observation.
-With a finite difference implementation of the pressure-velocity system \ref{eqn:awe}, velocity components are available ``for free'', short-circuiting explicit computation of $\Lambda$.
-
-The  {\tt project/SConstruct} script is set up to carry out the necessary computations on grids with spacings $\Delta x = \Delta z = $ 20, 10, and 5 m, with a jump of roughly 8 in computation time resulting from each refinement.
-For present purposes, the coarsest (20 m) grid seems to be sufficient, and that is the grid used in the examples presented below.
-The source pulses are chosen so that the computation is reasonably accurate.
-For the 20 m grid case, I use a zero-phase trapezoidal bandpass filter source with corner frequencies of 1.0, 2.0, 7.5, and 12.5 Hz.
-
-The IWAVE asg driver has been set up to recognize the case {\tt deriv=0} as defining the map from source (right-hand side in the pressure equation) to data (pressure) traces.
-The adjoint to this map, as explained above, is reverse-time propagation of the data traces as pressure sources, followed by scaling (formula \ref{eqn:sadj}).
-The approximate inverse is computed by application of the Dirichlet-to-Neumann operator to the pressure traces to produce corresponding velocity traces, followed by injection as pressure sources and reverse time propagation, followed by another application of the Dirichlet-to-Neumann map and scaling (formulas \ref{eqn:appinv}, \ref{eqn:adj}).
-
-The script implements these operatations step-by-step via calls to IWAVE, Madagascar, and SU commands.
-A peculiarity of the {\tt asg.x} driver needs to be mentioned: it is based on an un-scaled version of the constitutive law defect source representation, that is, {\tt asg.x} approximately computes the solution of the system \ref{eqn:awedata} with the first equation replaced by
-\begin{equation}
-\label{eqn:asgdata}
-\frac{\partial p}{\partial t}  =  - \kappa \nabla \cdot \bv +
-h \delta(z-z_s).
-\end{equation}
-Denote by $S_{\rm asg}[c]$ the forward map produced by {\tt asg.x}. Then comparison of \ref{eqn:awedata} and \ref{eqn:asgdata} reveals that
-\begin{equation}
-\label{eqn:sreln}
-S[c]=S_{\rm asg}[c]\kappa
-\end{equation}
-where $\kappa$ is shorthand for the operator of multiplication by $\kappa$. Accordingly, and approximate inverse for $S_{\rm asg}[c]$ is
-\[
-I \approx S_{\rm asg}[c]^{\dagger}S_{\rm asg}[c] = S_{\rm asg}[c]^{\dagger}S[c]\kappa^{-1}
-\]
-Since $S[c]S[c]^{\dagger} \approx I \approx S[c]^{\dagger}S[c]$, it follows that
-\[
-S_{\rm asg}[c]^{\dagger}=\kappa S[c]^{\dagger} = \kappa\Lambda[c]_s S[c]^T \Lambda[c]_r 
-\]
-\begin{equation}
-\label{eqn:asginv}
-= \kappa \Lambda[c]_s \kappa S_{\rm asg}[c]^T\Lambda[c]_r
-\end{equation}
-This is the approximate inverse computed in the examples. Note that
-only the values of $\kappa$ near the source datum $z=z_s$ play a role
-in the relation \ref{eqn:sreln} or in the definition \ref{eqn:asginv}
-of the IWAVE ASG approximate inverse.
-
-\section{Discussion}
-
-The source subproblem is a necessary step in solving the overall FWI problem \ref{eqn:fwi}.
-In particular, the nonlinear least squares realization \ref{eqn:esi} of source-extended FWI is naturally approached via the {\em variable projection method} \cite[]{GolubPereyra:73,GolubPereyra:03,vanLeeuwenMulder:09,Rickett:SEG12}, in which the minimum over $\bff$ in \ref{eqn:esis} is treated as a function of $\bf{c}$, which is in turn minimized.
-Variable projection thus treats the source subproblem as an inner problem, which must be solved for an estimate of $\bff$ at every step of an iterative method for the outer problem of minimization over $\bf{c}$.
-Efficiency in solving the source subproblem \ref{eqn:esis} is critical to efficient implementation of variable projection for solving the regularized FWI problem \ref{eqn:esi}.
-
-This paper restricts the choice of wave physics embodied in
-$L[\bf{c}]$ to linear acoustics.  Thus $\bf{c}=(\kappa,\rho)$ is the
-pair (bulk modulus, density) of positive functions of spatial
-position.
-The principal characteristics assumed of sources $\bff=$ are localization and isotropy.
-That is, physical sources will be assumed to be isotropic point radiators, described by the functional form
-\begin{equation}
-\label{eqn:ptsrc}
-f_{\rm pt}(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s). 
-\end{equation}
-Source locations $\bx_s$ are assumed known (as is natural for active source methods), and the source {\em wavelet} $w(t)$ is to be found as part of the solution of \ref{eqn:esi}.
-A convenient choice of penalty operator $A$ is then multiplication by the {\em offset} $|\bx-\bx_s|$: its null space consists precisely of distributions of the form \ref{eqn:ptsrc}.
-
-More complex radiation characteristics may be accommodated with similar but more complex choice for the operator $A$.
-Since the core analysis is similar, this paper deals only with the simplest case \ref{eqn:ptsrc}.
-
-The wavefield $u$ in this work is presumed to exist in all of Euclicean space-time: in other words, boundary effects have been removed from $u$, by surface-related multiple elimination or other techniques.
-Predicted data traces are simply the samples of one or more components of $u$ at receiver points $\bx_r$.
-
-%The domain of wave propagation is denoted $\Omega$, and two subsets of its boundary are identified: $\Gamma_s$ contains the support (non-zeroes) of the (extended) sources and in particular the locations $\bx_s$ of physical sources.
-Data trace locations $\bx_r$ lie in another subset $\Gamma_r$.
-The data projection operator $P$ samples the acoustic field $u$ on $\Gamma_r$, and includes a finite aperture mute.
-
-Achieving a small residual in the source subproblem \ref{eqn:esis} for arbitrary $c$ is a critical part of the convergence mechanism for the extended source approach to FWI \cite[]{HuangNammourSymesDollizal:SEG19}.
-However, for arbitrary data $d$, coefficient array $c$, and source location $\bx_s$, there does not exist an isotropic point source $f_{\rm pt}$ of the form \ref{eqn:ptsrc} for which $PL[c]^{-1}f_{\rm pt} \approx d$.
-This is so even if there does exist a coefficient array $c^*$ and point source $f_{\rm pt}^*$ located at $\bx_s$ with wavelet $w^*(t)$ for which $d = PL[c^*]^{-1}f_{\rm pt}^*$, if $c$ differs substantially from $c^*$.
-It is this fact that underlies the effectiveness of the extended source approach to FWI:  driving the residual in the problem \ref{eqn:esi} towards zero necessarily requires modifying $c$ to resemble $c^*$, assuming the latter exists, as well as pushing the optimal $f$ towards the null space of $A$.
-Conversely, the source subproblem \ref{eqn:esis} does not have a small residual solution of point source form \ref{eqn:ptsrc}, in general, for a given choice of $c$, even if that were the case for a different choice ($c^*$).
-Therefore a more general class of source models than that specified by the condition \ref{eqn:ptsrc} must be admitted to the feasible set for the source subproblem \ref{eqn:esis} if it is to serve as the inner problem in a variable projection formulation of extended source FWI.
-
-The class of non-point sources investigated here are those confined a hypersurface assumed to contain the locations of physical (point) sources, but not {\em a priori} required to have point support.
-These {\em surface extended sources} are able to yield a relatively small residual in the source subproblem \ref{eqn:esis} for more or less arbitrary coefficient array $c$.
-This observation is a by-product of the approximate solution for \ref{eqn:esis} constructed below.
-The construction of the approximate solution is closely related to the {\em time reversal} method of photoacoustic tomography (see \cite{StefanovUhlmannIP:09} and references cited there).
-
-The next section gives a precise description of an acoustic version of problem \ref{eqn:esis} for an idealized {\em crosswell} configuration, in which source located on a surface $z=z_s$ generate waves that propagate to to a receiver surface $z=z_r > z_s$.
-In principle, {\em diving waves} can be described with similar mathematics, but with some additional complications, so I do not treat the diving wave case explicitly in this paper.
-The assumptions detailed here imply that the data $d$ of the inverse problems \ref{eqn:esi}, \ref{eqn:esis} have the physical character of {\em transmitted waves}.
-There are no mathematical results at present concerning reflected wave inverse problems treated by source extension methods, though there are tantalizing numerical clues \cite[]{LeeuwenHerrmannWRI:13,Warner:14,Warner:16,LeeuwenHerrmann:16,HuangSymes:Geo17,HuangSymes:Geo18a,HuangSymes:Geo18b}.
-
-The following section describes the construction of the approximate
-inverse and explains the conditions under which accuracy should be expected.
-I observe that the approximate inverse is approximately the adjoint of the modeling operator $PL[c]^{-1}$ with respect to weighted norms in its domain and range (see \cite{HouSymes:15} for a similar observation in a different context).
-This fact immediately suggests use of the approximate inverse as a preconditioner to accelerate convergence of Krylov space methods such as conjugate gradient iteration applied to the least squares problem \ref{eqn:esis}.
-However, its straightforward application in the preconditioned conjugate gradient (PCG) method \cite[]{Golub:2012} involves explicitly a version the so called Dirichlet-to-Neumann (D2N) operator, which is computationally awkward.
-I show how to reorganize the PCG iteration so that only solutions of $L[c]u=f$ are required.
-The penultimate section presents some 2D examples in the crosswell configuration, illustrating the accuracy of the approximate inversion and the accelerated convergence of PCG.
-The examples are chosen to emphasize that surface extended sources can be constructed to fit more or less arbitrary data, unlike point sources - as mentioned before, this approximate invertibility property of the modeling operator is of critical importance in the application to extended FWI.
-The paper ends with a discussion of several important matters not addressed here, and a restatement of the conclusions.
-
-The intent of the main body of this paper is to present the formal structure of the extended source subproblem and the way in which this structure leads to accelerated numerical solution.
-This formal structure is however supported by a rigorous foundation.
-In order to avoid disrupting the formal account, this foundation material is relegated to an appendix.
+ \newpage
+
+\section{Figures}
 
@@
 
+\plot{dhshh0}{width=\textwidth}{Pressure source gather = -2 $\times$
+  vertical velocity gather (Figure \ref{fig:dsrcvzhh0}).}
+
+\plot{dfshh0}{width=\textwidth}{Velocity source gather = -2 $\times$
+  pressure gather (Figure \ref{fig:dsrcphh0}).}
+
 \plot{dinvhshh0}{width=\textwidth}{Approximate inversion via first 
@@
   homogeneous model. Scaled version of output $v_z$ field obtained by 
-  applying transpose of $4V^+_{z_s,z_r}$.}
+  applying transpose of $4V^+_{z_s,z_r}$. Approximates dip-filtered
+  version of pressure source gather (Figure \ref{fig:dhshh0}).}
 
@@
   equation in display \ref{eqn:approxinv}. Inversion in lens model of 
-  the pressure gather in Figure \ref{fig:drecpll0}, simulated with 
+  the pressure gather in Figure \ref{fig:drecplh0}, simulated with 
   lens model. Scaled version of output $v_z$ field obtained by 
-  applying transpose of $4V^+_{z_s,z_r}$.}
-
-\plot{drerecphh0}{width=\textwidth}{Re-simulated pressure gather
+  applying transpose of $4V^+_{z_s,z_r}$. Approximates dip-filtered
+  version of pressure source gather (Figure \ref{fig:dhshh0}).}
+
+\plot{drerecpll0}{width=\textwidth}{Re-simulated pressure gather
   produced from inverted source
@@
 
-\plot{ddiffrecphh0}{width=\textwidth}{Difference between gathers
+\plot{ddiffrecpll0}{width=\textwidth}{Difference between gathers
   displayed in Figures \ref{fig:drecpll0} and \ref{fig:drerecpll0},
@@
 \plot{dinvhslh0}{width=\textwidth}{Approximate inversion via first 
-  equation in display \ref{eqn:approxinv}. Inversion in homogeneous model of 
+  equation in display \ref{eqn:approxinv}. Inversion in homogenous model of 
   the pressure gather in Figure \ref{fig:drecplh0}, simulated with 
   lens model. Scaled version of output $v_z$ field obtained by 
-  applying transpose of $4V^+_{z_s,z_r}$.}
+  applying transpose of $4V^+_{z_s,z_r}$. Approximates dip-filtered
+  version of pressure source gather (Figure \ref{fig:dhshh0})}
 
@@
   inverted source
-  shown in Figure \ref{fig:dinvhslh0}. Simulation in homogeneous model.}
+  shown in Figure \ref{fig:dinvhslh0}. Simulation in lens model.}
 
@@
   plotted on same color scale.}
+
+\plot{dfwdvzhh0}{width=\textwidth}{Vertical velocity gather, generated
+  with a pressure source in the homogeneous model, 
+  corresponding to pressure gather \ref{fig:dfwdphh0}.}
+
+\plot{daltvzhh0}{width=\textwidth}{Vertical velocity gather, generated
+  with a velocity source in the homogenous model, corresponding to
+  pressure gather \ref{fig:daltphh0}.}
+
+\plot{dsvcomphh0}{width=\textwidth}{Difference between velocity gathers
+  shown in Figures \ref{fig:dfwdvzhh0} and \ref{fig:daltvzhh0},
+  plotted on the same color scale as these figures.}
 
@@
 
-\plot{ddiffinvhshh0}{width=\textwidth}{Difference between image
-  (Figure \ref{fig:predinvhshh0}) of $\Lambda^+_{z_s}$ applied to
-  pressure gather (Figure \ref{fig:dsrcphh0}) and source gather
+\plot{ddiffinvhshh0}{width=\textwidth}{Difference between (a) image
+  (Figure \ref{fig:preddinvhshh0}) of $\Lambda^+_{z_s}$ applied to
+  pressure gather (Figure \ref{fig:dsrcphh0}), and (b) source gather
   (Figure \ref{fig:dinvhshh0}) inverted from receiver pressure gather
-  (Figure \ref{fig:drecphh0}), homogeneous model used in all
-  propagations. Same color scale as in Figure \ref{fig:predinvhshh0}. }
+  (Figure \ref{fig:drecphh0}). Homogeneous model used in all
+  propagations. Same color scale as in Figure \ref{fig:preddinvhshh0}. }
 
@@
 
-\plot{ddiffinvhsll0}{width=\textwidth}{Difference between image
-  (Figure \ref{fig:predinvhslh0}) of $\Lambda^+_{z_s}$ applied to
-  pressure gather (Figure \ref{fig:dsrcpph0}) and source gather
+\plot{ddiffinvhsll0}{width=\textwidth}{Difference between (a) image
+  (Figure \ref{fig:preddinvhsll0}) of $\Lambda^+_{z_s}$ applied to
+  pressure gather (Figure \ref{fig:dsrcphh0}), and (b) source gather
   (Figure \ref{fig:dinvhsll0}) inverted from receiver pressure gather
-  (Figure \ref{fig:drecplh0}), lens model used in all propagations.
- Same color scale as in Figure \ref{fig:predinvhsll0}.}
+  (Figure \ref{fig:drecplh0}). Lens model used in all propagations.
+  Same color scale as in Figure \ref{fig:preddinvhsll0}.}
+
+\newpage
+
+\append{Adjoint Computation}
+
+The adjoint of ${\cal S}^+_{z_s,z_r}$ can be computed by a variant of
+the adjoint state method, in this case a by-product of the
+conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awepm}
+with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
+$ (h_r,f_r\bf{e}_z) \delta(z-z_r)$. Then
+\[
+0 = 
+\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
+\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
+-
+\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
+\]
+\[
+= 
+\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
+\]
+\[
+= 
+\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
+\]
+\[
++
+\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
+\]
+\[
+= 
+\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
+ h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
+ h_r \delta(z-z_r)\right) \right.
+\]
+\[
++
+\left.  (- \nabla p^++f_s\bf{e}_z) \cdot \bv^- + \bv^+ \cdot (-\nabla
+  p^- + f_r \bf{e_z}) \right)
+\]
+\[
+= 
+\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
+ h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
+ h_r \delta(z-z_r)\right) \right.
+\]
+\[
++
+\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- 
+  +f_s \delta(z-z_s) v_z^- + v_z^+f_r \delta(z-z_r) \right)
+\]
+after integration by parts in the last two terms. Most of what is left cancels, leaving 
+\[
+0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
+( h_rP_rp^++f_rP_rv_z^+) = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
+\]
+whence
+\begin{equation}
+\label{eqn:sadj}
+ ({\cal S}^+_{z_s,z_r})^T = -{\cal S}^{-}_{z_r,z_s}.
+\end{equation}
+
+The systems \ref{eqn:awepm} for $(p^{\pm},\bv^{\pm})$ with the two
+signs of superscript differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
+$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
+\begin{equation}
+  \label{eqn:trdef}
+  {\cal R} \left(
+    \begin{array}{c}
+      p\\
+      \bv
+    \end{array}
+  \right) =
+  \left(
+    \begin{array}{c}
+      Rp\\
+      -R\bv
+    \end{array}
+  \right)
+\end{equation}
+Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awepm} with
+$(h_s\delta(z-z_s),f_s\bf{e_z}\delta(z-z_s))$ replaced by
+$(-Rh_r\delta(z-z_r), Rf_r\bf{e_z}\delta(z-z_r))$. That is,
+\begin{equation}
+  \label{eqn:trsadj}
+  {\cal R}{\cal S}^- = -{\cal S}^+_{z_r,z_s}{\cal R}
+\end{equation}
+Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj}
+and \ref{eqn:trsadj} imply equation \ref{eqn:trtr}.
 

### 2021-06-07 repro/papers/sseprecond/oldstuff.tex
initial commit - pruned stuff

@@
+\section{Discussion}
+
+The source subproblem is a necessary step in solving the overall FWI problem \ref{eqn:fwi}.
+In particular, the nonlinear least squares realization \ref{eqn:esi} of source-extended FWI is naturally approached via the {\em variable projection method} \cite[]{GolubPereyra:73,GolubPereyra:03,vanLeeuwenMulder:09,Rickett:SEG12}, in which the minimum over $\bff$ in \ref{eqn:esis} is treated as a function of $\bf{c}$, which is in turn minimized.
+Variable projection thus treats the source subproblem as an inner problem, which must be solved for an estimate of $\bff$ at every step of an iterative method for the outer problem of minimization over $\bf{c}$.
+Efficiency in solving the source subproblem \ref{eqn:esis} is critical to efficient implementation of variable projection for solving the regularized FWI problem \ref{eqn:esi}.
+
+This paper restricts the choice of wave physics embodied in
+$L[\bf{c}]$ to linear acoustics.  Thus $\bf{c}=(\kappa,\rho)$ is the
+pair (bulk modulus, density) of positive functions of spatial
+position.
+The principal characteristics assumed of sources $\bff=$ are localization and isotropy.
+That is, physical sources will be assumed to be isotropic point radiators, described by the functional form
+\begin{equation}
+\label{eqn:ptsrc}
+f_{\rm pt}(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s). 
+\end{equation}
+Source locations $\bx_s$ are assumed known (as is natural for active source methods), and the source {\em wavelet} $w(t)$ is to be found as part of the solution of \ref{eqn:esi}.
+A convenient choice of penalty operator $A$ is then multiplication by the {\em offset} $|\bx-\bx_s|$: its null space consists precisely of distributions of the form \ref{eqn:ptsrc}.
+
+More complex radiation characteristics may be accommodated with similar but more complex choice for the operator $A$.
+Since the core analysis is similar, this paper deals only with the simplest case \ref{eqn:ptsrc}.
+
+The wavefield $u$ in this work is presumed to exist in all of Euclicean space-time: in other words, boundary effects have been removed from $u$, by surface-related multiple elimination or other techniques.
+Predicted data traces are simply the samples of one or more components of $u$ at receiver points $\bx_r$.
+
+%The domain of wave propagation is denoted $\Omega$, and two subsets of its boundary are identified: $\Gamma_s$ contains the support (non-zeroes) of the (extended) sources and in particular the locations $\bx_s$ of physical sources.
+Data trace locations $\bx_r$ lie in another subset $\Gamma_r$.
+The data projection operator $P$ samples the acoustic field $u$ on $\Gamma_r$, and includes a finite aperture mute.
+
+Achieving a small residual in the source subproblem \ref{eqn:esis} for arbitrary $c$ is a critical part of the convergence mechanism for the extended source approach to FWI \cite[]{HuangNammourSymesDollizal:SEG19}.
+However, for arbitrary data $d$, coefficient array $c$, and source location $\bx_s$, there does not exist an isotropic point source $f_{\rm pt}$ of the form \ref{eqn:ptsrc} for which $PL[c]^{-1}f_{\rm pt} \approx d$.
+This is so even if there does exist a coefficient array $c^*$ and point source $f_{\rm pt}^*$ located at $\bx_s$ with wavelet $w^*(t)$ for which $d = PL[c^*]^{-1}f_{\rm pt}^*$, if $c$ differs substantially from $c^*$.
+It is this fact that underlies the effectiveness of the extended source approach to FWI:  driving the residual in the problem \ref{eqn:esi} towards zero necessarily requires modifying $c$ to resemble $c^*$, assuming the latter exists, as well as pushing the optimal $f$ towards the null space of $A$.
+Conversely, the source subproblem \ref{eqn:esis} does not have a small residual solution of point source form \ref{eqn:ptsrc}, in general, for a given choice of $c$, even if that were the case for a different choice ($c^*$).
+Therefore a more general class of source models than that specified by the condition \ref{eqn:ptsrc} must be admitted to the feasible set for the source subproblem \ref{eqn:esis} if it is to serve as the inner problem in a variable projection formulation of extended source FWI.
+
+The class of non-point sources investigated here are those confined a hypersurface assumed to contain the locations of physical (point) sources, but not {\em a priori} required to have point support.
+These {\em surface extended sources} are able to yield a relatively small residual in the source subproblem \ref{eqn:esis} for more or less arbitrary coefficient array $c$.
+This observation is a by-product of the approximate solution for \ref{eqn:esis} constructed below.
+The construction of the approximate solution is closely related to the {\em time reversal} method of photoacoustic tomography (see \cite{StefanovUhlmannIP:09} and references cited there).
+
+The next section gives a precise description of an acoustic version of problem \ref{eqn:esis} for an idealized {\em crosswell} configuration, in which source located on a surface $z=z_s$ generate waves that propagate to to a receiver surface $z=z_r > z_s$.
+In principle, {\em diving waves} can be described with similar mathematics, but with some additional complications, so I do not treat the diving wave case explicitly in this paper.
+The assumptions detailed here imply that the data $d$ of the inverse problems \ref{eqn:esi}, \ref{eqn:esis} have the physical character of {\em transmitted waves}.
+There are no mathematical results at present concerning reflected wave inverse problems treated by source extension methods, though there are tantalizing numerical clues \cite[]{LeeuwenHerrmannWRI:13,Warner:14,Warner:16,LeeuwenHerrmann:16,HuangSymes:Geo17,HuangSymes:Geo18a,HuangSymes:Geo18b}.
+
+The following section describes the construction of the approximate
+inverse and explains the conditions under which accuracy should be expected.
+I observe that the approximate inverse is approximately the adjoint of the modeling operator $PL[c]^{-1}$ with respect to weighted norms in its domain and range (see \cite{HouSymes:15} for a similar observation in a different context).
+This fact immediately suggests use of the approximate inverse as a preconditioner to accelerate convergence of Krylov space methods such as conjugate gradient iteration applied to the least squares problem \ref{eqn:esis}.
+However, its straightforward application in the preconditioned conjugate gradient (PCG) method \cite[]{Golub:2012} involves explicitly a version the so called Dirichlet-to-Neumann (D2N) operator, which is computationally awkward.
+I show how to reorganize the PCG iteration so that only solutions of $L[c]u=f$ are required.
+The penultimate section presents some 2D examples in the crosswell configuration, illustrating the accuracy of the approximate inversion and the accelerated convergence of PCG.
+The examples are chosen to emphasize that surface extended sources can be constructed to fit more or less arbitrary data, unlike point sources - as mentioned before, this approximate invertibility property of the modeling operator is of critical importance in the application to extended FWI.
+The paper ends with a discussion of several important matters not addressed here, and a restatement of the conclusions.
+
+The intent of the main body of this paper is to present the formal structure of the extended source subproblem and the way in which this structure leads to accelerated numerical solution.
+This formal structure is however supported by a rigorous foundation.
+In order to avoid disrupting the formal account, this foundation material is relegated to an appendix.
+
+\section{Prototype Numerical Examples}
+I present a collection of simple examples that illustrate the features of the surface source extension claimed in preceding sections.
+
+I used the IWAVE acoustic staggered grid package to carry out these calculations.
+This package implements (2,2k) schemes for k=1,2,..., and outputs traces (of either velocity or pressure at any point in space via multilinear interpolation.
+The discretized modeling operator is thus of second order accuracy, though as usual I have used higher order in space to reduce grid dispersion.
+
+Source injection is implemented as the adjoint of trace sampling, resulting in another second-order error [REFERENCES].
+
+The data is a single shot gather, with a source at coordinates $x_s=z_s=3000$ m (units of length are meters in all cases).
+The receiver line occupies $1500 \le x_r \le 5500$ m, with receiver depth $z_r=1000$ m.
+Extended sources occupy $1500 \le x_s \le 5500$ m, with the same depth $z_s=3000$ m as the ``physical'' source used to generate the data.
+This region turns out to be adequate to represent the extended sources that approximately invert the data, for the cases examined below.
+An algorithm to automatically identify an appropriate region can be based on the ideas developed by \cite{Fu:Geo17}.
+
+I have used absorbing boundary conditions (split-field PML) on all four sides of the 4000 m (vertical) $\times$ 8000 m (horizontal) simulation domain.
+Evidently inclusion of a free surface is important to the application of the ideas explained here to diving wave marine data, and I have not addressed the necessary modifications here.
+
+The Dirichlet-to-Neumann operator $\Lambda$ is an essential part of the inner problem preconditioner just presented. Implementation can be accomplished in several ways:
+\begin{itemize}
+\item \cite{tenKroode:12} suggests using a one-way operator;
+\item if both pressure and normal partical velocity are measured (or simulated), then the two are related by $\Lambda$ and the velocity component can simply be used as the output;
+\item presence of a free surface implies all of the usual problems, such as the need for removal of receiver-side ghosts.
+On the other hand, if the free surface is within a quarter-wavelength throughout the useful bandwidth of the data, then the ghosted data differs from $\Lambda p$ by a time integration and a scale factor, a fact used to good effect by \cite{HouSymes:15}.
+\end{itemize}
+
+In the examples, I have used the second observation.
+With a finite difference implementation of the pressure-velocity system \ref{eqn:awe}, velocity components are available ``for free'', short-circuiting explicit computation of $\Lambda$.
+
+The  {\tt project/SConstruct} script is set up to carry out the necessary computations on grids with spacings $\Delta x = \Delta z = $ 20, 10, and 5 m, with a jump of roughly 8 in computation time resulting from each refinement.
+For present purposes, the coarsest (20 m) grid seems to be sufficient, and that is the grid used in the examples presented below.
+The source pulses are chosen so that the computation is reasonably accurate.
+For the 20 m grid case, I use a zero-phase trapezoidal bandpass filter source with corner frequencies of 1.0, 2.0, 7.5, and 12.5 Hz.
+
+The IWAVE asg driver has been set up to recognize the case {\tt deriv=0} as defining the map from source (right-hand side in the pressure equation) to data (pressure) traces.
+The adjoint to this map, as explained above, is reverse-time propagation of the data traces as pressure sources, followed by scaling (formula \ref{eqn:sadj}).
+The approximate inverse is computed by application of the Dirichlet-to-Neumann operator to the pressure traces to produce corresponding velocity traces, followed by injection as pressure sources and reverse time propagation, followed by another application of the Dirichlet-to-Neumann map and scaling (formulas \ref{eqn:appinv}, \ref{eqn:adj}).
+
+The script implements these operatations step-by-step via calls to IWAVE, Madagascar, and SU commands.
+A peculiarity of the {\tt asg.x} driver needs to be mentioned: it is based on an un-scaled version of the constitutive law defect source representation, that is, {\tt asg.x} approximately computes the solution of the system \ref{eqn:awedata} with the first equation replaced by
+\begin{equation}
+\label{eqn:asgdata}
+\frac{\partial p}{\partial t}  =  - \kappa \nabla \cdot \bv +
+h \delta(z-z_s).
+\end{equation}
+Denote by $S_{\rm asg}[c]$ the forward map produced by {\tt asg.x}. Then comparison of \ref{eqn:awedata} and \ref{eqn:asgdata} reveals that
+\begin{equation}
+\label{eqn:sreln}
+S[c]=S_{\rm asg}[c]\kappa
+\end{equation}
+where $\kappa$ is shorthand for the operator of multiplication by $\kappa$. Accordingly, and approximate inverse for $S_{\rm asg}[c]$ is
+\[
+I \approx S_{\rm asg}[c]^{\dagger}S_{\rm asg}[c] = S_{\rm asg}[c]^{\dagger}S[c]\kappa^{-1}
+\]
+Since $S[c]S[c]^{\dagger} \approx I \approx S[c]^{\dagger}S[c]$, it follows that
+\[
+S_{\rm asg}[c]^{\dagger}=\kappa S[c]^{\dagger} = \kappa\Lambda[c]_s S[c]^T \Lambda[c]_r 
+\]
+\begin{equation}
+\label{eqn:asginv}
+= \kappa \Lambda[c]_s \kappa S_{\rm asg}[c]^T\Lambda[c]_r
+\end{equation}
+This is the approximate inverse computed in the examples. Note that
+only the values of $\kappa$ near the source datum $z=z_s$ play a role
+in the relation \ref{eqn:sreln} or in the definition \ref{eqn:asginv}
+of the IWAVE ASG approximate inverse.
+
 \section{Fields, Traces and Sources}

### 2021-06-08 repro/papers/sseprecond/paper.tex
interim

@@
 
-Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
-efficient iterative solution of the linear subproblem
-\ref{eqn:esis}. Direct computation of the pressure-to-source operator $\Lambda^{\pm}_{z_s}$, for instance by solving
+Computations of $\Lambda^{\pm}_{z_s}$ and its transpose are clearly critical steps in an
+implementation of the PCG algorithm outlined in the preceding section.
+Direct computation of the pressure-to-source operator $\Lambda^{\pm}_{z_s}$, for instance by solving
 \ref{eqn:awepm} and reading off $P_sv^{\pm}_{z}$, turns out to be
@@
 alternative approach, taking advantage of the accurate approximate
-inverse to $S^+_{z_s,z_r}$ is available. The first row of
+inverse to $S^+_{z_s,z_r}$ constructed above. The first row of
 \ref{eqn:snull} is given explicitly in \ref{eqn:snull1}, which I
@@
 
-Figure \ref{fig:preddinvhshh0} shows the pressure source gather produced
-from the pressure data gather in Figure \ref{drecphh0} by application
-of the operator on the right-hand side of
-formula \ref{eqn:lamident}. The scale is the same as for Figures
-\ref{fig:dinvhshh0} and \ref{fig:dhshh0}. In fact the former
-appears nearly identical, and both appear to be dip-filtered versions
-of the latter. The difference plot (Figure \ref{fig:ddiffinvhshh0},
-using the same color scale, shows that the similarity is quantitative.
-
-Figures \ref{fig:preddinvhsll0} and \ref{fig:ddiffinvhsll0} accomplish
+Figure \ref{fig:preddinvhshh0} shows the pressure source gather
+produced from the pressure data gather in Figure \ref{dsrcphh0} by
+application of the operator on the right-hand side of formula
+\ref{eqn:lamident} to the source surface pressure gather depicted in
+\ref{fig:dsrcphh0}. The homogeneous model is used in all propagations
+implicit in the prescription \ref{eqn:lamident}. The scale is the same
+as for Figures \ref{fig:dinvhshh0} and \ref{fig:dhshh0}. In fact
+Figure \ref{fig:preddinvhshh0} appears to be nearly identical to
+Figure \ref{fig:dinvhshh0}, and both appear to be dip-filtered
+versions of Figure \ref{fig:dhshh0. The difference plot (Figure
+  \ref{fig:ddiffinvhshh0}, using the same color scale, shows that the
+  similarity is quantitative. The recovered source gather (Figure
+  \ref{fig:preddinvhshh0}) also generates the same acoustic fields as
+  the point source at $z=z_d$: the pressure gather at $z=z_r=1000$ m
+  depth is shown in Figure \ref{fig:dpredhsrecphh0}, and its
+  difference with the point source gather (Figure \ref{fig:drecphh0})
+  plotted on the same color scale in Figure
+  \ref{fig:ddiffpredhsrecphh0}.
+
+Figures \ref{fig:preddinvhsll0}, \ref{fig:ddiffinvhsll0},
+\ref{fig:dpredhsrecpll0}, and \ref{fig:ddiffpredhsrecpll0} accomplish
 the same comparison for the $\Lambda^+_{z_s}$ approximation in
-equation \ref{eqn:lamident}. A deeper study of the pressure-to-source
-operator shows that it is approximately dependent only on the model
+equation \ref{eqn:lamident}, with all propagation taking place in the
+lens model.
+
+A deeper study of the pressure-to-source
+operator (Appendix B) shows that it is approximately dependent only on the model
 coefficients near the source surface ($z=z_s$ in this case). Since the
@@
 unsurprising that these figures are very close to the previous two.
+However an even more useful observation is that the calculations in
+the approximation \ref{eqn:lamident} could just as well be carried out
+in a much smaller region around the source surface, and produce a
+result that is functionally identical in that it will serve as a
+source for the same acoustic fields globally, with small error. In
+effect, equation \ref{eqn:lamident} involving propagation from source
+($z=z_s$) to receiver ($z=z_r$) surfaces is altered by replacing $z_r$
+with a receiver datum $z_s+\Delta z$ considerably closers to $z_s$:
+\begin{equation}
+  \label{eqn:lamnear}
+  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
+    S}^+_{z_s,z_z_s+\Delta z}\Pi_1^T.
+\end{equation}
+Using a receiver datum closer to the source surface has two favorable consequences:
+\begin{itemize}
+\item The computational domain can be smaller than is necessary to
+  simulate the target data, as it need only contain the source
+  surface and the receiver datum implicit in
+  equation \ref{eqn:lamident}. This shrinkage of the computational
+  domain can lead to substantial improvements in computational
+  efficiency.
+\item Since the receiver data may be chosen much closer to the
+  source surface that is the case for the target data, the effective
+  aperture active in the relation \ref{eqn:lamident} can be much
+  larger, producing an estimated source gather much less affected by
+  aperture limitation.
+\end{itemize}
+
+To illustrate these points, I have re-computed the source gather
+estimates shown in Figures \ref{fig:preddinvhshh0} and
+\ref{fig:preddinvhsll0}. In place of the receiver geometry in the
+target data (Figures \ref{fig:drecphh0} and \ref{fig:drecpll0}, I
+place the receiver array at $z_s+\Delta z=2900$ m depth, just 100 m above the
+source surface at $z_s=3000$ m. For the discretization used to create
+the examples shown so far, that is just a 5 gridpoint difference in
+depth, as opposed to 100 gridpoints between the source and receiver
+depths for examples such as shown in Figure \ref{fig:recphh0} or
+\ref{fig:drecphh0}. The resulting approximation to $\Lambda_{z_s}^+$
+via equation \ref{eqn:lamident} for this configuration is evaluated in
+the same way as before in Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
+\ref{fig:dprednshsrecphh0}, and \ref{fig:ddiffprednshsrecphh0}
+(homogeneous model) and Figures \ref{fig:preddnshsll0}, \ref{fig:ddiffnshsll0},
+\ref{fig:dprednshsrecpll0}, and \ref{fig:ddiffprednshsrecpll0} (lens
+model). The effect of aperture limitation is clearly diminished: the
+second figure in each series compares the full-aperture pressure source
+gathers (Figures \ref{fig:dhshh0.su} and \ref{fig:dhsll0.su} with the
+approximated image of the corresponding pressure gather (Figure
+\ref{fig:dsrcphh0}) under the approximation to $\Lambda_{z_s}^+$, and
+the last figures show that the approximated source gathers accurately
+predict the point-source pressure gather at the receiver datum
+$z_r=1000$ m.
 
@@
   \label{eqn:lamtilde}
-  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
-\end{equation}
-Then \ref{eqn:lamident} can be rewritten
+  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
+    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
+\end{equation}
+Then \ref{eqn:lamnear} can be rewritten
 \[
@@
   \label{eqn:lamtransp}
-   (\tilde{\Lambda}^+_{z_s})^T = -8 \Pi_1 ({\cal S}^+_{z_r,z_s})^T\Pi_0^T V^{+}_{z_s,z_r}
+   (\tilde{\Lambda}^+_{z_s})^T = -8 \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
+     z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}
+ \end{equation}
+Thus the model space weight operator introduced in the last section is
+replaced by
+\[
+  \tilde{W}_m = \frac{1}{2}(\tilde{\Lambda}^+_{z_s} +
+  (\tilde{\Lambda}^+_{z_s})^T)
+\]
+\[
+  = -8 \left((V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
+    S}^+_{z_s,z_s+\Delta z}\Pi_1^T \right.+
+\]
+\[
+  \left. \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
+      z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}\right)
+\]
+\begin{equation}
+  \label{eqn:wcomp}
+  = -4(\Pi_1 ({\cal S}^+_{z_s,z_s+\Delta z})^T(\Pi_0^T\Pi_1 + \Pi_1^T\Pi_0) {\cal S}^+_{z_s,z_s+\Delta z}\Pi_1^T)
 \end{equation}

### 2021-06-10 repro/papers/sseprecond/paper.tex
interim

@@
 Figure \ref{fig:dinvhshh0}, and both appear to be dip-filtered
-versions of Figure \ref{fig:dhshh0. The difference plot (Figure
+versions of Figure \ref{fig:dhshh0}. The difference plot (Figure
   \ref{fig:ddiffinvhshh0}, using the same color scale, shows that the
@@
   \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_z_s+\Delta z}\Pi_1^T.
+    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
 \end{equation}
@@
 \end{equation}
+with a similar definition for the replacement $\tilde{W}_d$ of $W_d$.
+
+This identity shows that only one forward and one adjoint simulation
+are necessary to compute the action of $\tilde{W}_{m,d}$. The operator
+in the center of the expression on the right-hand side, $\Pi_0^T\Pi_1
++ \Pi_1^T\Pi_0$, simply exchanges the components of the acoustic
+fields, passing the velocity field as a pressure source and the
+pressure field as a velocity source.
 

### 2021-06-14 repro/papers/sseprecond/paper.tex
discuss lam transpose

@@
 $z_r=1000$ m.
-
-Computation of the transpose of $\Lambda^+$ (exact, not approximate in
+As mentioned in the last section, computation of the transpose of $\Lambda^+$ (exact, not approximate in
 the high frequency sense) is critical to the successful
@@
 \]
+Of course, all of the examples so far show images of
+$\tilde{\Lambda}^+_{z_s}$.
+
 Since successful preconditioning requires only approximate inversion,
@@
 \end{equation}
-Thus the model space weight operator introduced in the last section is
-replaced by
+
+Figure \ref{fig:preddnshstrhh0} shows the image of the pressure gather
+in Figure \ref{fig:dsrcphh0} under $(\tilde{\Lambda}^+_{z_s})^T$,
+using the ``near'' traces at $z=2900$, that is, $\Delta z = 100$ m in
+expression \ref{eqn:lamtransp}, and propagation in the
+homogeneous model. Note
+the close resemblance to the image of the same pressure gather under
+$\tilde{\Lambda}^+_{z_s}$ displayed in Figure
+\ref{fig:preddnshshh0}. The difference of these two images is
+displayed in \ref{fig:ddiffnslamtrhh0}, on the same color scale as the
+images themselves. Since the propagation takes place entirely in a
+region where all of the mechanical parameters are homogenous, I do not
+offer a similar comparison for the lens model.
+
+The model space weight operator $W_m$ introduced in the last section is
+replaced by its asymptotic approximation
 \[
@@
 pressure field as a velocity source.
+
+Figure \ref{fig:symmdnshshh0} shows the output of the symmetrized
+approximate source-to-pressure operator per equation \ref{eqn:wcomp},
+applied once again to the pressure data in Figure
+\ref{fig:dsrcp}. Note the resemblance to Figures
+\ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}. These are all
+asymptotic approximations of each other.
 
@@
   Same color scale as in Figure \ref{fig:preddinvhsll0}.}
+
+\plot{preddnshshh0}{width=\textwidth}{Pressure source gather = image
+  under pressure-to-source operator $\Lambda^+_{z_s}$ of pressure gather
+  shown in Figure \ref{fig:dsrcphh0}, homogeneous model, using
+  ``near'' receiver traces at $z=2900$ m. Compare Figure
+  \ref{fig:dhshh0} and \ref{fig:dinvhshh0}: because the sources and
+  receivers are close, little aperture is lost in this case.}
+
+\plot{ddiffnshshh0}{width=\textwidth}{Difference between (a) image
+  (Figure \ref{fig:preddnshshh0}) of $\Lambda^+_{z_s}$ applied to
+  pressure gather (Figure \ref{fig:dsrcphh0}) using a near receiver
+  array to implement formulat \ref{eqn:lamident}, and (b) source gather
+  (Figure \ref{fig:dhshh0}) inferred from vertical velocity.
+  Homogeneous model used in all
+  propagations. Same color scale as in Figure \ref{fig:preddnshshh0}. }
+
+\plot{preddnshstrhh0}{width=\textwidth}{Pressure source gather = image
+  under {\em transpose} of pressure-to-source operator
+  $\Lambda^+_{z_s}$ of pressure gather shown in Figure
+  \ref{fig:dsrcphh0}, homogeneous model, using ``near'' receiver
+  traces at $z=2900$ m. Compare Figure \ref{fig:dhshh0} and
+  \ref{fig:preddnshshh0}: as noted in the text, $\Lambda^+_{z_s}$ is
+  asymptotically symmetric, so the resemblance is not a surprise.}
+
+\plot{ddiffnslamtrhh0}{width=\textwidth}{Difference between data in
+  Figures \ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}, plotted
+  on the same scale as these figures, showing that the asymptotic
+  symmetry of $\Lambda^+_{z_s}$ is actually quantitative for the
+  length, time and frequency scales of these examples.}
+
+\plot{symmdnshshh0}{width=\textwidth}{Pressure source gather = image
+    under {\em symmetrized} pressure-to-source operator
+    $\frac{1}{2}\left(\Lambda^+_{z_s}+(\Lambda^+_{z_s})^T\right)$ of
+    pressure gather shown in Figure \ref{fig:dsrcphh0}, homogeneous
+    model, using ``near'' receiver traces at $z=2900$ m. Compare
+    Figure \ref{fig:preddnshshh0}.}
 

### 2021-06-17 repro/papers/sseprecond/paper.tex
includes cg vs pcg plots, but no text

@@
   V^+_{z_s,z_r} (S^+_{z_s,z_r})^T & \approx & \frac{1}{4}I, \nonumber\\
-  S^+_{z_s,z_r} (V^+_{z_s,z_r})^T & \approx & \frac{1}{4}I.\\
+  S^+_{z_s,z_r} (V^+_{z_s,z_r})^T & \approx & \frac{1}{4}I.
 \end{eqnarray}.
@@
 limitation of the acquisition geometry: the steeper dips in the source
-gather \ref{fig:dhshh0} do not contribute to the data, and are not
+gather (Figure \ref{fig:dhshh0}) do not contribute to the data, and are not
 recovered in the inversion.
@@
 that is very far from a dip-filtered version of the source used
-generate the data \ref{fig:dhshh0}. However, this result is an
+generate the data (Figure \ref{fig:dhshh0}). However, this result is an
 accurate inversion: re-simulation {\em using the same (homogeneous)
@@
 $-2$ produces
-$Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_s p^+$. On the other
+$\Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_s p^+$. On the other
 hand, figure \ref{fig:daltvzhh0} shows the result of applying
@@
   \label{eqn:almostunitary}
-  4(V_{z_s,z_r}^+)^TS^+_{z_s,z_r} = [ (\Lambda^+_{z_r})^T
-  (S^{+}_{z_s,z_r})^T(\Lambda^+_{z_s})^T]S^{+}_{z_s,z_r} \approx I.
+  4(V_{z_s,z_r}^+)^TS^+_{z_s,z_r} = [ (\Lambda^+_{z_s})^T
+  (S^{+}_{z_s,z_r})^T(\Lambda^+_{z_r})^T]S^{+}_{z_s,z_r} \approx I.
 \end{equation}
@@
   \label{eqn:unitary}
-  [ \frac{1}{2}((\Lambda^+_{z_r})^T+ \Lambda^+_{z_r})
-  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_s})^T+
-  \Lambda^+_{z_s})]S^{+}_{z_s,z_r} \approx I.
-\end{equation}
+  [ \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
+  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
+  \Lambda^+_{z_r})]S^{+}_{z_s,z_r} \approx I.
+\end{equation}
+
+To illustrate this unitary property of $S^+_{z_s,z_r}$, I apply the
+operator
+\[
+  \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
+  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
+  \Lambda^+_{z_r})
+\]
+to the data $S^+_{z_s,z_r}h_s$ (Figure \ref{fig:drerecphh0}), in which
+$h_s$ is the dip-filtered version (Figure \ref{fig:dinvhshh0}) of the
+downgoing source created earlier (Figure \ref{fig:dhshh0}), produced
+by inversion of the data in Figure \ref{fig:drecphh0} via the first
+equation in display \ref{eqn:approxinv}, propagation in the homogenous
+model.  The operator above is computed via the technique explained in
+the section ``Computing and Symmetrizing $\Lambda$'', below, using
+auxiliary receiver arrays 100 m above the data source and receiver arrays.
+
+The output is shown in Figure
+\ref{fig:lamsstlamrdrecphh0}. The difference with the actual source is
+shown in Figure \ref{fig:difflamsstlamrdrecphh0}. A similar exercise
+using the lens data \ref{fig:drecplh0} but inversion in the homogenous
+model produces the extended source depicted in
+\ref{fig:lamsstlamrdrecplh0}, which differs from the inversion result
+via equation \ref{eqn:approxinv} by the gather displayed in Figure
+\ref{fig:difflamsstlamrdrecplh0}.
+
 The symmetrized $\Lambda$ operators are at least positive
 semi-definite, hence define (at least) semi-norms.
-
 Similar relations have been derived for other scattering operators,
@@
 likely to have null (or nearly-null) vectors due to aperture-related
-amplitude loss. I will proceed as if this were sensible, since the
-mathematical issues can be defused and the only the application of
-$W_m^{-1}$ is required in the algorithms described here.
+amplitude loss. Since some version of $W_m$ is essential in the
+formulation for effective preconditioning, I will derive a usable
+candidate to stand in for it below.
 
@@
 good preconditioner. However this observation must be combined with
-the wegithed norm structure.
-
-Rewrite the normal equation \ref{eqn:norm} as
+the weighted norm structure.
+
+Rewrite the normal equation \ref{eqn:norm0} as
 \begin{equation}
@@
 inverse to $S^+_{z_s,z_r}$ constructed above. The first row of
-\ref{eqn:snull} is given explicitly in \ref{eqn:snull1}, which I
-repeat for the reader's convenience:
+\ref{eqn:snull}, slightly rearranged, is
 \begin{equation}
@@
 \end{equation}
-
 The approximate inverse construction for $S^++_{z_s,z_r}$ permits (approximate)
@@
 \end{equation}
-This identity is the second major result of this section: it shows how
+This identity is the major result of this section: it shows how
 to compute that action of $\Lambda^+_{z_s}$ by propagating the input
@@
 Figure \ref{fig:preddinvhshh0} shows the pressure source gather
-produced from the pressure data gather in Figure \ref{dsrcphh0} by
+produced from the pressure data gather in Figure \ref{fig:dsrcphh0} by
 application of the operator on the right-hand side of formula
@@
 versions of Figure \ref{fig:dhshh0}. The difference plot (Figure
-  \ref{fig:ddiffinvhshh0}, using the same color scale, shows that the
+  \ref{fig:ddiffinvhshh0}), using the same color scale, shows that the
   similarity is quantitative. The recovered source gather (Figure
@@
 To illustrate these points, I have re-computed the source gather
-estimates shown in Figures \ref{fig:preddinvhshh0} and
-\ref{fig:preddinvhsll0}. In place of the receiver geometry in the
-target data (Figures \ref{fig:drecphh0} and \ref{fig:drecpll0}, I
-place the receiver array at $z_s+\Delta z=2900$ m depth, just 100 m above the
+estimates shown in Figure \ref{fig:preddinvhshh0}. In place of the receiver geometry in the
+target data (Figure \ref{fig:drecphh0}), I
+place a receiver array at $z_s+\Delta z=2900$ m depth, just 100 m above the
 source surface at $z_s=3000$ m. For the discretization used to create
@@
 depths for examples such as shown in Figure \ref{fig:recphh0} or
-\ref{fig:drecphh0}. The resulting approximation to $\Lambda_{z_s}^+$
-via equation \ref{eqn:lamident} for this configuration is evaluated in
-the same way as before in Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
-\ref{fig:dprednshsrecphh0}, and \ref{fig:ddiffprednshsrecphh0}
-(homogeneous model) and Figures \ref{fig:preddnshsll0}, \ref{fig:ddiffnshsll0},
-\ref{fig:dprednshsrecpll0}, and \ref{fig:ddiffprednshsrecpll0} (lens
-model). The effect of aperture limitation is clearly diminished: the
-second figure in each series compares the full-aperture pressure source
-gathers (Figures \ref{fig:dhshh0.su} and \ref{fig:dhsll0.su} with the
-approximated image of the corresponding pressure gather (Figure
-\ref{fig:dsrcphh0}) under the approximation to $\Lambda_{z_s}^+$, and
-the last figures show that the approximated source gathers accurately
-predict the point-source pressure gather at the receiver datum
-$z_r=1000$ m.
-As mentioned in the last section, computation of the transpose of $\Lambda^+$ (exact, not approximate in
-the high frequency sense) is critical to the successful
-construction of the preconditioner. The relation \ref{eqn:lamident}
-does not provide a computation for this operator. However set
+\ref{fig:drecphh0}.
+
+Note that the homogeneous and lens models are identical in the depth
+range $2900 < z < 3000$ m, so the computed $\Lambda^+_{z_s}$ operators
+will be the same for both models. Hence I show only results for the
+homogenous model.
+
+The approximation to $\Lambda_{z_s}^+$ via equation \ref{eqn:lamident}
+for this configuration is evaluated in the same way as before in
+Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
+\ref{fig:dprednshsrecphh0}, and \ref{fig:ddiffprednshsrecphh0}. The effect of aperture
+limitation is clearly diminished: the second figure in this series
+compares the full-aperture pressure source gather (Figure
+\ref{fig:dhshh0}) with the 
+image of the corresponding pressure gather (Figure \ref{fig:dsrcphh0})
+under the approximation to $\Lambda_{z_s}^+$, and the last figures
+show that the approximated source gathers accurately predict the
+point-source pressure gather at the receiver datum $z_r=1000$ m.
+
+As mentioned in the last section, computation of the transpose of
+$\Lambda^+$ (exact, not approximate in the high frequency sense) is
+critical to the successful construction of the preconditioner. The
+relation \ref{eqn:lamident} does not provide a computation for this
+operator. However set
 \begin{equation}
@@
 applied once again to the pressure data in Figure
-\ref{fig:dsrcp}. Note the resemblance to Figures
+\ref{fig:dsrcphh0}. Note the resemblance to Figures
 \ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}. These are all
@@
 
+\plot{lamsstlamrdrecphh0}{width=\textwidth}{Inversion of data shown in
+  Figure \ref{fig:drecphh0}, simulated in homogeneous model, using the
+  approximate unitarity relation \ref{eqn:unitary} and propagation in
+  homogenous model.}
+
+\plot{difflamsstlamrdrecphh0}{width=\textwidth}{Difference between
+  data displayed in Figures \ref{fig:dinvhshh0} and
+  \ref{fig:lamsstlamrdrecphh0}, plotted on the same color scale.}
+
+\plot{lamsstlamrdrecplh0}{width=\textwidth}{Inversion of data shown in
+  Figure \ref{fig:drecplh0}, simulated in lens model, using the
+  approximate unitarity relation \ref{eqn:unitary} and propagation in
+  homogenous model.}
+
+\plot{difflamsstlamrdrecplh0}{width=\textwidth}{Difference between
+  data displayed in Figures \ref{fig:dinvhslh0} and
+  \ref{fig:lamsstlamrdrecplh0}, plotted on the same color scale.}
+
 \plot{preddinvhshh0}{width=\textwidth}{Pressure source gather = image
@@
   (Figure \ref{fig:drecphh0}). Homogeneous model used in all
-  propagations. Same color scale as in Figure \ref{fig:preddinvhshh0}. }
+  propagations. Same color scale as in Figure
+  \ref{fig:preddinvhshh0}. }
+
+\plot{dpredhsrecphh0}{width=\textwidth}{Pressure gather at receiver 
+  datum $z=z_r=1000$ m simulated in homogeneous model from source 
+  gather shown in Figure \ref{fig:preddinvhshh0}, also based on the 
+  homogeneous model. Compare with point source pressure gather (Figure 
+  \ref{fig:drecphh0}).}
+
+\clearpage
+
+\plot{ddiffpredhsrecphh0}{width=\textwidth}{Plot of difference 
+  between data shown in Figures \ref{fig:drecphh0} and 
+  \ref{fig:dpredhsrecphh0}, plotted on same color scale as the latter 
+  two figures.}
 
@@
 
+\plot{dpredhsrecpll0}{width=\textwidth}{Pressure gather at receiver 
+  datum $z=z_r=1000$ m simulated in lens model from source 
+  gather shown in Figure \ref{fig:preddinvhsll0}, also based on the 
+  homogeneous model. Compare with point source pressure gather (Figure 
+  \ref{fig:drecplh0}).}
+
+\plot{ddiffpredhsrecpll0}{width=\textwidth}{Plot of difference 
+  between data shown in Figures \ref{fig:drecplh0} and 
+  \ref{fig:dpredhsrecpll0}, plotted on same color scale as the latter 
+  two figures.}
+
+\clearpage
+
 \plot{preddnshshh0}{width=\textwidth}{Pressure source gather = image
@@
   pressure gather (Figure \ref{fig:dsrcphh0}) using a near receiver
-  array to implement formulat \ref{eqn:lamident}, and (b) source gather
-  (Figure \ref{fig:dhshh0}) inferred from vertical velocity.
+  array to implement formula \ref{eqn:lamident}, and (b) source gather
+1650  (Figure \ref{fig:dhshh0}) inferred from vertical velocity.
   Homogeneous model used in all
-  propagations. Same color scale as in Figure \ref{fig:preddnshshh0}. }
+  propagations. Same color scale as in Figure
+  \ref{fig:preddnshshh0}. }
+
+\plot{dprednshsrecphh0}{width=\textwidth}{Pressure gather at receiver
+  datum $z=z_r=1000$ m simulated from source gather shown in Figure
+  \ref{fig:preddnshshh0}. Compare with point source pressure gather
+  (Figure \ref{fig:drecplh0}).}
+
+\clearpage
+
+\plot{ddiffprednshsrecphh0}{width=\textwidth}{Plot of difference 
+  between data shown in Figures \ref{fig:drecphh0} and 
+  \ref{fig:dpredhsrecphh0}, plotted on same color scale as the latter 
+  two figures.}
 
@@
     Figure \ref{fig:preddnshshh0}.}
+
+\plot{compreshh0}{width=\textwidth}{Comparison of residual Euclidean 
+  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
+  homogeneous model, point source (Figure \ref{fig:recphh0}), inversion 
+  in homogeneous model.}
+
+\plot{compreslh0}{width=\textwidth}{Comparison of residual Euclidean 
+  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
+  lens model, point source (Figure \ref{fig:recplh0}), inversion 
+  in homogenous model.}
+
+\plot{compresll0}{width=\textwidth}{Comparison of residual Euclidean 
+  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
+  lens model, point source (Figure \ref{fig:recplh0}), inversion 
+  in lens model.}
+
+\plot{compnreshh0}{width=\textwidth}{Comparison of normal residual
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted
+  vs. iteration. Data = homogeneous model, point source (Figure
+  \ref{fig:recphh0}), inversion in homogenous model.}
+
+\plot{compnreslh0}{width=\textwidth}{Comparison of normal residual 
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+  vs. iteration. Data = lens model, point source (Figure 
+  \ref{fig:recplh0}), inversion in homogenous model.}
+
+\plot{compnresll0}{width=\textwidth}{Comparison of normal residual 
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+  vs. iteration. Data = lens model, point source (Figure 
+  \ref{fig:recplh0}), inversion in lens model.}
 

### 2021-06-22 repro/papers/sseprecond/paper.tex
cg/pcg results

@@
 
-The model space weight operator $W_m$ introduced in the last section is
+The model space weight operator $W_m^{-1}$ introduced in the last section is
 replaced by its asymptotic approximation
 \[
-  \tilde{W}_m = \frac{1}{2}(\tilde{\Lambda}^+_{z_s} +
+\frac{1}{2}(\tilde{\Lambda}^+_{z_s} +
   (\tilde{\Lambda}^+_{z_s})^T)
@@
 \[
-  = -8 \left((V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
+  \approx -8 \left((V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
     S}^+_{z_s,z_s+\Delta z}\Pi_1^T \right.+
@@
 \begin{equation}
-  \label{eqn:wcomp}
-  = -4(\Pi_1 ({\cal S}^+_{z_s,z_s+\Delta z})^T(\Pi_0^T\Pi_1 + \Pi_1^T\Pi_0) {\cal S}^+_{z_s,z_s+\Delta z}\Pi_1^T)
+  \label{eqn:winvcomp}
+  = -4(\Pi_1 ({\cal S}^+_{z_s,z_s+\Delta z})^T(\Pi_0^T\Pi_1 +
+  \Pi_1^T\Pi_0) {\cal S}^+_{z_s,z_s+\Delta z}\Pi_1^T) = \tilde{W}_m^{-1} .
 \end{equation}
@@
 Figure \ref{fig:symmdnshshh0} shows the output of the symmetrized
-approximate source-to-pressure operator per equation \ref{eqn:wcomp},
+approximate source-to-pressure operator per equation \ref{eqn:winvcomp},
 applied once again to the pressure data in Figure
@@
 \ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}. These are all
-asymptotic approximations of each other.
-
- \newpage
+asymptotic approximations of each other. Figure
+\ref{fig:ddiffsymmdxnshsll0} shows the
+ the difference between the pressure gather at $z=z_r$ produced from
+ the pressure source output by the symmetrized $\Lambda$, and the point source
+simulation (Figure \ref{fig:dfwdplh0}), plotted on the same scale as
+the latter, in both cases with all propagations in the lens model.
+
+One more computation is required for the full implementation of the
+preconditioning strategy explained in the last section: $W_m$ is
+required, not just $W_m^{-1}$. Note that $W_m$ plays two roles in the
+second term in equation \ref{eqn:norm1}: it is the weight matrix for
+both the domain and range norms for $A$. It is perfectly OK for one of
+these to be replaced by an asymptotic approximation, so long as it is
+symmetric and computable (and at least semi-definite). The second row
+in equation \ref{eqn:snull} appears as \ref{eqn:snull2} above:
+introducing (formally) the inverse of $\Lambda^+$,
+\begin{equation}
+  \label{eqn:snull2mod}
+-\frac{1}{2}\Pi_1{\cal S}^{+}_{z_s,z_r}\Pi_0^T  \approx
+V^{+}_{z_s,z_r}(\Lambda^{+}_{z_s})^{-1}
+\end{equation}
+whence from the second line in display \ref{eqn:approxinv}
+\begin{equation}
+  \label{eqn:laminv}
+-\frac{1}{8}(S^+_{z_s,z_r})^T\Pi_1{\cal S}^{+}_{z_s,z_r}\Pi_0^T  
+\approx (\Lambda^{+}_{z_s})^{-1}
+\end{equation} 
+and
+\begin{equation}
+  \label{eqn:laminvtr}
+-\frac{1}{8}\Pi_0({\cal S}^{+}_{z_s,z_r})^T\Pi_1^T S^+_{z_s,z_r}
+\approx ((\Lambda^{+}_{z_s})^{-1})^T.
+\end{equation}
+Using the definition \ref{eqn:sdef} of $S^+_{z_s,z_r}$, the
+symmetrized $\Lambda^{-1}$ is
+\begin{equation}
+  \label{wcomp}
+ \tilde{W}_m = -\frac{1}{16}\left(\Pi_0 ({\cal S}^{+}_{z_s,z_r})^T
+   (\Pi_1^T \Pi_0 + \Pi_0^T \Pi_1) {\cal S}^{+}_{z_s,z_r} \Pi_0^T\right)
+   \approx \frac{1}{2}((\Lambda^{+}_{z_s})^{-1}
+   +((\Lambda^{+}_{z_s})^{-1})^T) .
+\end{equation}
+Comparison with the definition \ref{eqn:winvcomp} shows that
+$\tilde{W}_m$ and $\tilde{W}_m^{-1}$ differ only in the initial and
+final projection factors (and overall scale), and in particular either
+can be computed for the cost of a forward/adjoint operator pair. Note
+that $\tilde{W}_m^{-1}$ is inverse to $\tilde{W}_m$ only in an
+approximate (asymptotic, aperture-limited) sense.
 
@@
 
+\plot{ddiffsymmdnshsll0}{width=\textwidth}{Difference between point
+  source simulation (Figure \ref{fig:dfwdplh0}) and pressure gather at
+  $z=z_r=1000$ m produced by simulation with the source shown in
+  Figure \ref{fig:symmdnshshh0},
+  propagation in the lens model.}
+
 \plot{compreshh0}{width=\textwidth}{Comparison of residual Euclidean 
@@
   homogeneous model, point source (Figure \ref{fig:recphh0}), inversion 
-  in homogeneous model.}
-
-\plot{compreslh0}{width=\textwidth}{Comparison of residual Euclidean 
+  in homogeneous model. Penalty weight
+  $\alpha=0$.}
+
+\plot{compres0lh0}{width=\textwidth}{Comparison of residual Euclidean 
   norms: CG (blue), PCG (red), plotted vs. iteration. Data =
   lens model, point source (Figure \ref{fig:recplh0}), inversion 
-  in homogenous model.}
+  in homogenous model. Penalty weight 
+  $\alpha=0$.}
+
+\plot{compres1lh0}{width=\textwidth}{Comparison of residual Euclidean 
+  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
+  lens model, point source (Figure \ref{fig:recplh0}), inversion 
+  in homogenous model. Penalty weight 
+  $\alpha=10^{-3}$.}
 
@@
   lens model, point source (Figure \ref{fig:recplh0}), inversion 
-  in lens model.}
+  in lens model. Penalty weight
+  $\alpha=0$.}
 
@@
   vs. iteration. Data = homogeneous model, point source (Figure
-  \ref{fig:recphh0}), inversion in homogenous model.}
-
-\plot{compnreslh0}{width=\textwidth}{Comparison of normal residual 
+  \ref{fig:recphh0}), inversion in homogenous model. Penalty weight $\alpha=0$.}
+
+\plot{compnres0lh0}{width=\textwidth}{Comparison of normal residual 
   (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
   vs. iteration. Data = lens model, point source (Figure 
-  \ref{fig:recplh0}), inversion in homogenous model.}
+  \ref{fig:recplh0}), inversion in homogenous model. Penalty weight
+  $\alpha=0$.}
+
+\plot{compnres1lh0}{width=\textwidth}{Comparison of normal residual 
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+  vs. iteration. Data = lens model, point source (Figure 
+  \ref{fig:recplh0}), inversion in homogenous model. Penalty weight
+  $\alpha=10^{-3}$.}
 
@@
   vs. iteration. Data = lens model, point source (Figure 
-  \ref{fig:recplh0}), inversion in lens model.}
+  \ref{fig:recplh0}), inversion in lens model. Penalty weight
+  $\alpha=0$.}
 

### 2021-07-13 repro/papers/sseprecond/paper.tex
updated paper.tex and project/SConstruct as transmitted to Mohamed

@@
 
-\plot{compreshh0}{width=\textwidth}{Comparison of residual Euclidean 
+\plot{compres0hh0}{width=\textwidth}{Comparison of residual Euclidean 
   norms: CG (blue), PCG (red), plotted vs. iteration. Data =
@@
 
+\plot{compres0ll0}{width=\textwidth}{Comparison of residual Euclidean 
+  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
+  lens model, point source (Figure \ref{fig:recplh0}), inversion 
+  in lens model. Penalty weight 
+  $\alpha=0$.}
+
+\plot{compres1hh0}{width=\textwidth}{Comparison of residual Euclidean 
+  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
+  homogeneous model, point source (Figure \ref{fig:recphh0}), inversion 
+  in homogenous model. Penalty weight 
+  $\alpha=10^{-3}$.}
+
 \plot{compres1lh0}{width=\textwidth}{Comparison of residual Euclidean 
@@
 
-\plot{compresll0}{width=\textwidth}{Comparison of residual Euclidean 
+\plot{compres1ll0}{width=\textwidth}{Comparison of residual Euclidean 
   norms: CG (blue), PCG (red), plotted vs. iteration. Data =
@@
 
-\plot{compnreshh0}{width=\textwidth}{Comparison of normal residual
+\plot{compnres0hh0}{width=\textwidth}{Comparison of normal residual
   (gradient) Euclidean norms: CG (blue), PCG (red), plotted
@@
 
+\plot{compnres0ll0}{width=\textwidth}{Comparison of normal residual 
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+  vs. iteration. Data = lens model, point source (Figure 
+  \ref{fig:recplh0}), inversion in lens model. Penalty weight 
+  $\alpha=0$.}
+
+\plot{compnres1hh0}{width=\textwidth}{Comparison of normal residual 
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+  vs. iteration. Data = homogeneous model, point source (Figure 
+  \ref{fig:recphh0}), inversion in homogenous model. Penalty weight 
+  $\alpha=10^{-3}$.}
+
 \plot{compnres1lh0}{width=\textwidth}{Comparison of normal residual 
@@
 
-\plot{compnresll0}{width=\textwidth}{Comparison of normal residual 
+\plot{compnres1ll0}{width=\textwidth}{Comparison of normal residual 
   (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
@@
   \ref{fig:recplh0}), inversion in lens model. Penalty weight
-  $\alpha=0$.}
+  $\alpha=10^{-3}$.}
 

### 2021-08-28 repro/papers/EAGE21/cartoon.cc
slides done

@@ -1,0 +1,84 @@
+#include <sys/types.h>
+#include <sys/uio.h>
+#include <string.h>
+#include <unistd.h>
+#include <fcntl.h>
+#include <sys/stat.h>
+#include <stdio.h>
+#include <float.h>
+#include <math.h>
+#include <stdlib.h>
+#include <assert.h>
+#include <ctype.h>
+#include <time.h>
+
+#include <iostream>
+#include <fstream>
+#include <sstream>
+#include <iomanip>
+#include <limits>
+#include <exception>
+#include <stdexcept>
+#include <typeinfo>
+#include <string>
+#include <map>
+#include <list>
+#include <vector>
+#include <valarray>
+#include <complex>
+#include <memory>
+
+using namespace std;
+
+#define NV 101
+#define VE 2.0
+#define V0 0.5
+#define DV 0.01
+#define C0 1.0
+#define C1 0.1
+#define C2 4.0
+#define C3 0.05
+#define C4 0.2
+#define C5 7.0
+#define C6 50.0
+#define C7 500.0
+
+int main(int argc, char ** argv) {
+
+  float * v   = new float[NV];
+  float * j   = new float[NV];
+  float * h   = new float[NV];
+  float * t   = new float[NV];
+  float * f   = new float[NV];
+
+  for (int i=0; i< NV; i++) {
+    v[i] = i*DV - V0;
+    h[i] = C0*v[i]*v[i];
+    j[i] = h[i] - C3*sin(C2*v[i])*sin(C2*v[i])*sin(C2*v[i])*sin(C2*v[i])
+      + C1*sin(C5*v[i])*sin(C5*v[i])*sin(C5*v[i])*C3*sin(C5*v[i])*sin(C5*v[i])*sin(C5*v[i]);
+    t[i]= h[i] + C4*sin(C2*v[i])*sin(C2*v[i])*sin(C2*v[i])*sin(C2*v[i])
+      - C1*sin(C2*v[i])*sin(C2*v[i])*sin(C2*v[i])*C1*sin(C2*v[i])*sin(C2*v[i])*sin(C2*v[i]);
+    f[i]= 0.5*(C7*h[i]+C0*sin(C6*v[i])*sin(C6*v[i]))/(1.0+C7*h[i]);
+  }
+
+  FILE * dat = fopen("cartoon.rsf@","w");
[…]

### 2021-10-26 repro/papers/EFWIreview/InitialThoughts.tex
initial

\documentclass[10pt]{article}
\usepackage{array}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage[all]{xy}
\usepackage{hyperref}
\hypersetup{ colorlinks = true, linkcolor = blue, citecolor = blue}

\parindent=0cm
\parskip=.25cm
\def\accskip{\hskip.25em\relax}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{result}{Result}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{assumption}{Assumption}
\newcommand{\qed}{\hfill \ensuremath{\Box}}

\begin{document}
Hi Bill,

I have started thinking about the possible content of a joined review paper on extension concepts for velocity inversion. Here are some initial thoughts for discussion. Feel free to disagree of course -- everything below is very rough and you may have completely different ideas about content of the paper. Shall we set up a zoom meeting again to discuss?

One of the first questions I have is whether you are thinking about a pure review, or would consider including (partially) new material. For example, you mentioned you have a proof for the asymptotic vanishing of the gradient with respect to velocity of the SSE-functional on reflection data. Have you published that proof already? I will also mention some rough ideas below that we may want to work out as part of this paper.

Assuming that Frontiers of Earth Science will come up with financial support for the cost related to open access publication, we will need to prepare an abstract, so we better agree on the central theme of the paper. The tentative title is telling a lot -- the infamous cycle skipping problem in standard FWI may be overcome by non-physical extensions of the inverse problem. The true physical model is enforced in a soft manner through a penalization term, annihilating non-physical behaviour. Another important common denominator in extended inverse problems is the requirement that first order perturbations of extended forward modeling can be written as the product of that operator and a 1st order skew symmetric pseudo-differential operator, something we have both established for the subsurface offset extended Born operator in 2014.

As far as I can see, there are two classes of extended inversion schemes, one for reflection data (data misfit+DSO term), the other for transmission data (data misfit+SSE term). This dichotomy is at the forefront of attention again with the advent of long offset nodal surveys like Amendment. Such surveys are obviously motivated by the desire to measure transmission data, but vendors typically try to make their FWI schemes work for both reflections and transmissions in these data. Since everybody starts at the lowest possible frequency (typically 1.6-1.7 Hz), their is not much of a discussion initially, as the earth is not very reflective at such low frequencies. The debate starts when vendors push up the maximum frequency in the inversion, sometimes up to 12 Hz, where one would certainly have a mix between the two data types. Sofar, the difference between  FWI on diving/refracted waves only and on the full data has been small by the way.

I would be in favour of connecting the mathematical insights to the new data becoming available in the industry, maybe already in the abstract. In the paper itself we could review the two classes mentioned above and possibly speculate on schemes working on both reflection and transmission data simultaneously. I very much realize this may require a next review paper in 10 years time;).

Below are some more concrete thoughts around the two classes with lots of questions.

You have formulated extended inversion schemes for reflection data of the form
\begin{align}\label{ExtendedReflectionInversion}
  J[c,i] & = \frac{1}{2}\left\Vert Fi-d \right\Vert^2 + \frac{1}{2}\alpha^2\left\Vert Ai\right\Vert^2,
\end{align}
with $F=F[c]$ a subsurface offset extended Born operator, $i=i(x,h)$ a subsurface offset extended image and $A$ an annihilator. I first saw this extension in your work from many years ago and recall you making comments on the number of iterations required by early pactitioners. Where did this topic go since then? I presume this will feature in our review?

I fiddled a bit with this inverse problem by using the variable projection method described in your WRI and SSE papers and the relation
\begin{align}\label{PerturbationExtendedBornOperator}
  \delta F[c] & = F[c]P[c,\delta c],
\end{align}
which we both discovered and found these to be the essential ingredients for asymptotic local convexity of $J_{VPM}[c]$. Up to second order in $\delta c=c-c_0$ one has the asymptotic result
\begin{align}\label{ExtendedReflectionInversion_AsymptoticConvexity}
  J_{VPM}[c] & = \frac{1}{2}\Vert F[c_0]B[c_0,\delta c]i_0\Vert^2+
  \frac{1}{2}\alpha^2\Vert[A,P[c_0,\delta c]]i_0\Vert^2,
\end{align}
 where $i_0(x,h)=i[c_0]=r(x)\delta(h)$ is a focused image gather and $B[c_0,\delta c]$ is a symmetric order $0$ pseudo-differential operator, linear in $\delta c$. I presume you have done similar/better calculations?

For transmission data one should use the surface source extended inverse problem, which takes the form
\begin{align}\label{ExtendedTransmissionInversion}
  J[c,f] & = \frac{1}{2}\left\Vert Sf-d \right\Vert^2 + \frac{1}{2}\alpha^2\left\Vert Af\right\Vert^2,
\end{align}
with $S=S[c]$ the solution operator for the full wave equation, $f=f(x_s,y_s,x,y,t)\break\times\delta(z-z_s)$ a surface extended source, $A$ an annihilator trying to focus the source in $x=x_s, y=y_s)$. Your analysis of this problem for the 1D constant velocity case is very instructive. Could we consider the more general situation of diving waves in an arbitrary smooth velocity model? Clearly, we would need a relation of the form (\ref{PerturbationExtendedBornOperator}) again, which would look like
\begin{align}\label{PerturbationExtendedTransmissionOperator}
  \delta S[c] & = S[c]Q[c,\delta c].
\end{align}
I have not done the precise math, but it seems to me that such a relation ought to hold. $S[c]$ is an FIO mapping source functions to data and equation (\ref{PerturbationExtendedTransmissionOperator}) should be true for
\begin{align}\label{symbolQ}
{\rm sym}\, Q=i\omega(x,k)\delta T(x,k)
\end{align} 
with $\delta T(x,k)$ the traveltime perturbation due to a velocity perturbation $\delta c(y)$ along a diving ray starting in $x$ with take-off direction determined by the horizontal slowness $k$. We should have access to $k$ in the case of dense source (or receiver) sampling. If this can be demonstrated properly, would the same reasoning as above lead to asymptotic local convexity of the functional (\ref{ExtendedTransmissionInversion})?

In practice, data will of course be a mix of reflections and diving waves in a ratio depending on frequency and on acquisition type. For very low frequencies, diving/refracted waves start to dominate. The new long offset nodal data are obviously also much richer in diving/refracted wave content. Is there anything that can be said about this situation, if only in a forward looking manner? A pedestrian approach may be to try to separate reflections and transmissions in processing. This may suffer for large angle reflections, which are hard to distinguish from diving waves. Maybe your analysis of the vanishing of the gradient of the SSE functional with respect to velocity when applied to reflection data would show that there is no need to separate reflection and transmission data, i.e.\, that the SSE functional (\ref{PerturbationExtendedTransmissionOperator}) is essentially only sensitive to the diving waves in the data? If so, we could apply the surface source extended FWI first and follow up by differential semblance velocity analysis on the reflections. This would be a better version of current practice, which consists of finding the large scale deep velocity updates by standard FWI on transmission data and following up by reflection tomography (either standard ray based TTI, or wave equation based RMO updating as in Shell's Wave Path Tomography).

\end{document}

### 2021-10-29 repro/papers/discreptheory/paper.tex
arXiv

--- 2021-01-03 repro/papers/discrep/discrep/paper.tex
@@
-\title{Solution of an Acoustic Transmission Inverse Problem by Extended Inversion}
-\author{William W. Symes\footnotemark[1], Huiyi Chen\footnotemark[2],
-  and Susan E. Minkoff\footnotemark[2]}
-\address{\footnotemark[1]PO Box 43, Orcas, WA 98280 USA\\
-  \footnotemark[2]Department of Mathematical Sciences, FO 35, 
-  University of Texas at Dallas, Richardson, TX 75080 USA}
+\title{Solution of an Acoustic Transmission Inverse Problem by
+  Extended Inversion: Theory}
+\author{William W. Symes\footnotemark[1]}
+\address{\footnotemark[1]PO Box 43, Orcas, WA 98280 USA, {\tt symes@rice.edu}}
 
@@
 
-\righthead{Discrepancy}
+\righthead{1D Acoustic Inversion: Theory}
 
@@
 \begin{abstract}
-  Study of an extremely simple single-trace transmission example shows
-  how an extended source formulation of full waveform inversion can
-  produce an optimization problem without spurious local minima
-  (``cycle skipping''), hence efficiently solvable via Newton-like
-  local optimization methods. The data consist of a single trace
-  extracted from a causal pressure field, propagating in a fluid
-  according to linear acoustics and recorded at a given distance from
-  a transient point energy source. The velocity or slowness of the
-  fluid is presumed homogeneous, and the target source intensity
-  (``wavelet'') is presumed quasi-impulsive, with zero energy for time
-  lags greater than a specified maximum lag (specified support). The
-  data are assumed noisy, with known RMS signal-to-noise ratio. The
-  inverse problem is: from the recorded trace, recover both the sound
-  velocity and source wavelet with specified support, so that the data
-  is fit with prescribed RMS relative error. After showing explicitly that
-  the obvious least squares objective has multiple large residual
-  minimizers, we introduce an extended approach. The source is extended by
-  permitting energy to spread in time. The spread is controlled by
-  adding a weighted mean square of the extended source wavelet to the
-  data misfit, to produce the extended inversion objective. We choose
-  a weight operator to penalize source energy away from zero lag via
-  multiplication by the time lag. The resulting objective function and
-  its gradient can be computed explicitly. We show that the error in 
-  slowness component of any local minimizer is bounded by multiple of
-  the maximum lag and the assumed noise level. In particular, there
-  are no spurious local minima. We also show that the projection of
-  the extended wavelet onto the interval of support yields a solution
-  of the inverse problem as posed above, with appropriate choice of
-  maxiumum lag and RMS relative error. In several appendices, we
-  describe a constructive algorithm for controlling the penalty
-  weight, the relation between maxiumum lag and RMS central frequency
-  via the Heisenberg principle, and some structural details which the
-  very simple inverse problem discussed here shares with more
-  prototypical problems in industrial and academic seismology.
-
-  %which may be interpreted as a wavelength.
-  %The derivation shows
-  %several important features of all similar extended source
-  %algorithms. For example, nested optimization, with the source
-  %estimation in the inner optimization (variable projection method),
-  %is essential. The choice of the penalty operator, controlling the
-  %extended source degrees of freedom, is critical: in order to produce
-  %an objective immune from cycle-skipping, the penalty operator must
-  %be (pseudo-)differential. The penalty weight is chosen dynamically,
-  %according to a version of the Discrepancy Principle, in order to
-  %accelerate convergence to a near-global minimum. Like many other
-  %extended formulations, the example problem considered here permits
-  %the discrepancy criterion to be satisfied with zero penalty
- % weight. This unusual feature removes a common obstacle to the use of
-  %the discrepancy principle, namely the determination of a
-  %satisfactory initial estimate of the penalty weight: the initial
- % weight may be set to zero, and the dynamic algorithm updates the
- % weight until convergence is achieved.
+  A single-trace transmission inverse problem for the wave equation
+  seeks to determine both the wave velocity in a homogenous acoustic
+  medium and the transient waveform of an isotropic point source. The
+  duration (support) of the source waveform and the source-to-receiver
+  distance are assumed known. A least squares formulation of this
+  problem exhibits the ``cycle-skipping'' behaviour observed in field
+  scale problems of this type, with many local minima differing
+  greatly from the global minimizer. This behaviour is eliminated by
+  dropping the hard support constraint on the source waveform,
+  replacing it by a soft penalty implemented as a weighted mean-square
+  of the source waveform. For properly chosen weight function,
+  penalizing nonzero values away from $t=0$, the velocity component of
+  any stationary point differs from the global minimizer of the
+  constrained least-squares formulation by a linear combination of the
+  source waveform support radius and data noise-to-signal ratio. Given
+  an estimate of data noise, the penalty weight can be dynamically
+  adjusteded during iterative optimization to maximize predicted data
+  accuracy and closely approximate a support-constrained source waveform.
 
@@
 \section{Introduction}
-Model-based parameter estimation by least squares data fitting is a
-widely used and successful technique for data analysis in science and
-engineering \cite{Bertero:98,Vogel:02}, particularly in the
-geosciences \cite{Parker:94,Tarantola:05}. In its application to
-seismology, least-squares data fitting has come to be known as Full
-Waveform Inversion (FWI), and is now well-established as a useful tool
-for probing the earth's subsurface
-\cite[]{VirieuxOperto:09,Fichtner:10,Schuster:17}. However, in its
-application to seismology, FWI encounters a serious
-practical obstacle, so-called ``cycle-skipping''. This term signifies
-the tendency of the mean-square error function to exhibit many
-stationary points far from the global minimizer (we
-show an explicit example of this phenomenon later in this paper). Because of the
-typical dimensions of field FWI problems, only local gradient-based
-minimization algorithms are computationally practical. These find
-stationary points (or local minima), so may stagnate at suboptimal and geologically
-uninformative earth models. Local descent methods avoid suboptimal
-stagnation only if initial models are already quite close to optimal,
-in the sense of predicting the arrival times of seismic events to
-within a small multiple of the dominant data wavelength
-\cite[]{GauTarVir:86,VirieuxOperto:09,Plessix:10}.
-
-This paper examines one of a number of proposed alternatives to least
-squares FWI, based on an extension of the model-to-data mapping
-(``extended inversion''), in the context of 
-a very simple acoustic transmission inverse problem: the sound
-velocity in a fluid is to be determined from a single recording of a
-pressure wave propagated over a known distance, originating at a known
-location around a known time. We show precisely
-how descent-based FWI fails to solve this problem absent a very good
-initial model estimate, and precisely why extended inversion succeeds, using the same type
-of optimization, and from essentially arbitrary initial model
-estimate. We give rigorous estimates for the accuracy of the sound
-velocity determination in terms of the time extent of the energy
-excitation and the signal-to-noise ratio of the data. We also illustrate
-these mathematical conclusions with numerical
-examples.
-
-The single-trace transmission problem studied in this paper is a
-drastically simplified cartoon of actual seismic prospecting. It seems
-worth careful study nonetheless for two reasons: it exhibits the
-phenomenon of cycle-skipping, and its simplicity allows the properties
-of both FWI and extended inversion approaches to be established
-rigorously. The underlying mathematical structures are common to many
-inverse problems in wave propagation in which wave velocities are
-amongst the unknowns. For example, we offer an ad hoc calculation of
-the gradient of the extended inversion objective, taking advantage of
-the simplicity of the problem context. However we also show, in an
-appendix, how that calculation hides an abstract structure shared by
-many other problems.% - see for example
-%\cite{Symes:IPTA14,tenKroode:IPTA14,Symes:EAGE15,HuangSymes:SEG15a}.
-
-In the next (``Overview'') section, we describe the conceptual
-framework of this inverse problem, review the related literature, and
-state the essence of our main results, leaving out a few technical
-details that are supplied later. Numerical illustrations of these
-results occupy the following (``Examples'') section. The fourth
-(``Findings'') section of the paper contains precise statements of the
-mathematical results, derivations, and supporting discussion. Three
-appendices describe the algorithm used to control the penalty weight
-in our formulation of extended inversion, further refinement of error
-estimates based ond the Heisenberg inequality, and the abstract
-structure of extended inversion implicit in the derivations presented
-in the ``Findings'' section.
-
-\section{Overview}
-
-The physical setting for our discussion is the propagation of
-small-amplitude pressure waves through a
-homogeneous acoustic fluid occupying $\bR^3$ \cite[]{Frie:58}. The
-fluid's salient characteristic for present purposes is the
-slowness, or reciprocal wave velocity, $m > 0$. A source of acoustic
-energy localized at a (``source'') point $\bx_s \in \bR^3$ radiates uniformly in all
-directions with time-dependent intensity $w(t)$, generating a pressure
-field $p(\bx,t)$. Both the 
-pressure field $p$ and the intensity (``wavelet'') $w$ vanish at large
-negative time. A receiver records the pressure values at a point
-$\bx_r \in \bR^3$, a distance
-$r>0$ from the source position $\bx_s$, between the times $t_{\rm min}$ and
-$t_{\rm max}$. According to linear acoustics, the pressure field
+Inverse problems based on wave equations (acoustic, elastic,
+Maxwell's...) are commonplace in geophysics, nondestructive materials
+testing, medical imaging, and other areas of science and engineering
+in which wave motion plays an important role. Often these problems are
+formulated as optimization of mean square (or other scalar measure)
+functions of misfit between observed and predicted data. In many
+contexts, such as exploration seismology, the computational size of
+these problems is very large, and even the prediction (or modeling) of
+the data requires use of high-performance computing
+environments. Accordingly, rapidly convergent optimization algorithms
+based on Newton's method and its relatives to update model parameters
+are the numerical methods of choice. These however are local, in the
+sense that they converge to stationary points of the objective
+functions, usually near the initial model parameters. For
+high-frequency (hence high resolution) data regimes, stagnation of
+local optimization methods at physically irrelevant model vectors is
+frequently observed. This ``cycle-skipping'' is one of the main
+obstacles to widespread successful deployment of optimization-based
+methods for inverse problems in wave propagation
+\cite[]{GauTarVir:86,VirieuxOperto:09,Fichtner:10,Plessix:10,Schuster:17}.
+
+This paper shows how a modification of least-squares data fitting via
+{\em modeling operator extension} overcomes cycle-skipping in a simple
+example. The problem studied here may be the simplest inverse wave
+problem that exhibits cycle-skipping - in fact, essentially this
+example is routinely used to illustrate cycle-skipping (see for
+example \cite{VirieuxOperto:09}, Figure 7). Its data is a single trace
+at a point in $\bR^3$ (``receiver location'') of the causal linear
+acoustic pressure field, generated by an isotropic radiator acting at
+another point (``source location'') at a known distance from the
+receiver location. The acoustic material is supposed spatially
+homogeneous, hence characterized by wave slowness (reciprocal
+velocity) and density. The object of the inverse problem is to recover
+the slowness and the time history of the isotropic radiator
+(``wavelet'') from the recorded trace. In order that the data
+constrain the slowness, the wavelet is presumed to have known compact
+support. Thus the (``physical'') modeling operator has as its domain
+the Cartesian product of an interval of positive slownesses, and the
+subspace of $L^2(\bR)$ consisting of wavelets with the prescribed
+support. Even for noise-free data (in the range of the modeling
+operator), the obvious least squares problem formulated in terms of
+this modeling operator exhibits many local minimizers having nothing
+to do with the (``target'') slowness and wavelet used to create the
+data - that is, cycle-skipping occurs (Theorem \ref{thm:fwi}).
+
+Dropping the support constraint on the wavelet creates an extension of
+the modeling map for which any data may be fit exactly, for any choice
+of slowness. The link between data and slowness is restored by adding
+a quadratic penalty on non-zero values of the wavelet outside of the
+physical support interval to the extended mean-square data misfit. The
+first main result of this paper is that for noise-free data, the
+slowness component of {\em any} stationary point of this extended
+penalty function differs from the target slowness by a multiple of the
+target wavelet radius (Theorem \ref{thm:rampreallygood}). That is, cycle-skipping does not occur, and the
+wavelet support radius determines the resolution of slowness. For
+noisy data (that is, sum of a noise-free signal trace in the range of
+the physical modeling operator and a square-integrable noise trace),
+any stationary point differs from the target slowness by a combination
+of the wavelet support radius and the noise-to-signal ratio (Theorem \ref{thm:mnoiseres}). Since the
+wavelet component of the extended domain does not constrain support,
+stationary points of the extended penalty function do not provide
+solutions of the original, physical inverse problem, which includes
+the support constraint. Such solutions can be obtained, with estimates
+on support radius and data error, simply by truncating the wavelet
+components of stationary points (Theorem \ref{thm:ipnoisesuf}).
+
+The inverse problem studied here is far too simple to have any direct
+use in applications. However it seems worthy of careful study for two
+reasons. First, its simplicity allows an unusually complete account of
+its properties. The present paper details its mathematical behaviour,
+including explicit and sharp error estimates rarely available for nonlinear
+inverse problems. A companion paper \cite[]{SymesChenMinkoff:21}
+describes numerical examples that illustrate the mathematical
+conclusions.
+
+Second, the analysis reveals several important features shared with
+similar approaches to other wave inverse problems, and so provides an
+at least partial pattern for reaching similar conclusions with
+immediate practical implications. For example, nested optimization
+(variable projection method, \cite{GolubPereyra:03}) based on a model
+decomposition into inner and outer variables seems to be very
+important: in some cases, such as the simple problem presented here,
+the decomposition is obvious (inner variable = wavelet, outer variable
+= slowness), in others less so
+\cite[]{geoprosp:2008,Terentyev:thesis}. The extended modeling
+operator must be surjective, or at least have a dense range, {\em for
+  each value of the outer variable}, so that data may be fit well even
+for a poor initial guess of the outer variable. for the extended
+inversion approach to be successful. The derivative of the extended
+modeling operator with respect to the outer variable is
+well-approximated by the composition of the extended modeling operator
+itself and a pseudodifferential operator of order 1
+\cite[]{Symes:IPTA14,tenKroode:IPTA14}. The choice of the penalty
+operator, controlling the extended degrees of freedom, is critical: in
+order to produce an objective immune from cycle-skipping, this penalty
+operator must be (pseudo-)differential
+\cite[]{StolkSymes:03}. Generally, the extended modeling operator is
+not differentiable in the outer variable. However these last two facts
+together imply the differentiability of the reduced objective function
+from variable projection (Theorem \ref{thm:diffobj}, Appendix C). Because of the data-fitting
+assumption, a straightforward algorithm for scaling the quadratic
+penalty, based on the Discrepancy Principle, is available - see
+\cite{FuSymes2017discrepancy,SymesChenMinkoff:21}, also Appendix
+A. This scaling varies dynamically during iterative optimization, in
+effect changing the objective function sporadically as the iteration
+converges.
+
+The approach explained here is an example of {\em extended source}
+inversion, see \cite{HuangNammourSymesDollizal:SEG19} for a review,
+and \cite{MetivierBrossier:SEG20} for a recent variation. Adding
+degrees of freedom to the coefficients in the wave equation in various
+ways produces so-called {\em extended model} inversion approaches,
+which actually have a longer history in seismology - see
+\cite{geoprosp:2008} for details. The need to overcome cycle-skipping
+has spawned a great number of concepts, of which extension-based
+inversion is only one. Least-squares inversion succeeds with
+sufficiently good initial estimates of wave velocity, and the dominant
+techniques in industrial and academic seismology exploit data other
+than direct measurements of waves, for example measurement of wave
+time-of-flight, to supply such initial estimates
+\cite[]{VirieuxOperto:09,Fichtner:10,Schuster:17}. Use of a data
+misfit measure other than least squares, such as versions of the
+Wasserstein metric from optimal transport theory, also shows promise
+\cite[]{Metivier:GEO18,EngquistYang:GEO18}.
+
+The next section gives precise statements of the inverse problem
+studied here, and its various components. The following sections
+contain statements and proofs of the main results: the existence of
+muliple stationary points of the least squares problem, properties of
+the extended modeling operator and the reduced objective, choice of
+the penalty operator, the main results mentioned above about
+stationary points of the reduced objective, and the discrepancy-based
+algorithm for adjusting the penalty scale. Three appendices justify
+the discrepancy-based scale adjustment, review the link between
+constrained support and spectrum provided by the Heisenberg inequality,
+and expose the abstract structure underlying the differentiability of
+the reduced objective.
+
+I use the abbreviations ${\cal B}(X,Y)$ and ${\cal I}(X,Y)$ for 
+the algebra of bounded linear operators from the Hilbert space $X$ to 
+the Hilbert space $Y$, and its subalgebra of invertible
+operators. All of the operators appearing in this discussion are
+members of ${\cal B}(X,Y)$ for some choice of Hilbert spaces $X$ and
+$Y$, and superscript $T$ denotes the transpose or adjoint of a such an
+operator in the sense specified by its domain and range Hilbert structure.
+
+\section{An Acoustic transmission Inverse Problem}
+According to linear acoustics, the causal pressure field due to an
+isotropic point radiator at $\bx_s \in \bR^3$ 
 solves the the wave equation \cite[]{Frie:58}:
@@
 \end{eqnarray}
-As the system \ref{eqn:awe} is linear and autonomous, the solution is
-the convolution of the source
-wavelet $w$ with the Green's function or fundamental solution. For
-constant $m$, the result is the well-known expression
-(\cite{CourHil:62}, Chapter VI, section 12, equation 47):
+The solution is well-known, see for instance
+\cite{CourHil:62}, Chapter VI, section 12, equation 47:
 \begin{equation}
@@
 \end{equation}
-so the predicted pressure at the receiver point $\bx_r$ is
+This trace of $p$ at $\bx_r \in \bR^3$ can be viewed as the result of
+applying an m-dependent linear operator $F[m]$ to the wavelet $w$:
 \begin{equation}
 \label{eqn:mod}
-p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right)  = F[m]w(t)
-\end{equation}
-The mapping $F[m]$ so defined is (ignoring the amplitude factor
-$1/(4\pi r)$), an $m$-dependent time shift. This time shift operator is the basis of many illustrations of
-the cycle-skipping phenomenon (for example, \cite{VirieuxOperto:09},
-Figure 7), so it is unsurprising that an analysis of cycle-skipping
-can be based on the simple modeling operator described above.
-
-If the wavelet $w$ is square-integrable, so is the predicted
-recorded pressure $F[m]w$ - its $L^2$ norm is the cumulative power trasferred
-from the fluid to the receiver \cite[]{SantosaSymes:00}. Thus this relation defines a map
-$F: \bR^+ \times L^2(\bR) \rightarrow
-L^2([t_{\rm min},t_{\rm max}])$, linear in its second argument.
-
-A naive version of the inverse problem investigated in this paper
-would be: given recorded data
-$d \in L^2([t_{\rm min},t_{\rm max}])$, find $m \in \bR^+$ and $w \in L^2(\bR)$ so that
-$F[m]w \approx d$. Clearly this problem statement is not interesting as it
-stands, since $F[m]$ is surjective for every $m \in \bR^+$: fitting
-the data does not constrain the slowness $m$ at all.
-
-To be useful, the problem statement must be augmented with a
-constraint on the model $(m,w)$.  One natural constraint is to assume
-that the support of $w$ is limited: choose $\lambda > 0$ and search
-for $w$ with $\mbox{supp }w \subset [-\lambda,\lambda]$. The quantity
-$\lambda$ is to be considered ``small''. Under some circumstances, it
-is related to a mean wavelength of $w$ (Appendix B).
-
-An assumption of small support may be justified as
-follows. In practical seismic data processing, actual sources may act
-over considerable time intervals. An example from exploration
-seismology is the airgun, the most commonly used
-marine source: an airgun (or array of airguns, the usual apparatus)
-produces an oscillating pressure pulse that dies away
-slowly. It is commonplace to estimate this pulse, usually by recording
-the emitted pressure field at a position where it should nominally be isolated from other
-signal, then deconvolve it
-from the data by safeguarded Fourier division or other means. This
-so-called signature deconvolution results in modified data
-corresponding to a source equal to signature deconvolution of the
-pulse estimate itself. The deconvolved source pulse approximates an
-impulse ($\delta(t)$): it cannot have point support, due to the finite
-frequency nature of seismic data and the Heisenberg principle (see
-Appendix B), but its amplitudes are typically small outside of a small
-much smaller time interval than is the case for the original pulse
-estimate. Thus after a small modification, the deconvolved source
-pulse has small support.
-
-For a contemporary description of view on signature deconvolution as
-practiced in the seismic industry, see \cite{SheriffGeldart:1995},
-section 9.5 and \cite{Yil:01}, Chapter 2.
-
-Abusing notation slightly, we identify $L^2([-\lambda,\lambda])$ with
-a subspace of $L^2(\bR)$ via extension by $0$. Denote by $\lF$
-restriction of $F$ to $\bR^+ \times L^2([-\lambda,\lambda])$ defined by equation
-\ref{eqn:mod}. Then the FWI problem with support constraint,
-described in the last paragraph, can be crudely phrased: given $d \in
-L^2([t_{\rm min},t_{\rm max}])$, find $(m,w)$ in the domain of $\lF$ so that $\lF[m,w]
-\approx d$.
-
-Despite its simplicity, this problem exhibits the cycle-skipping
-pathology characteristic of field-scale FWI. Specifically, we show
-that the reduced mean square error function, the objective function of
-FWI for this problem,
-\begin{equation}
-  \label{eqn:redms}
-  \lerr[m,w;d]=\frac{1}{2}\|\lF[m]w-d\|^2/\|d\|^2,
-\end{equation}
-has local minimizers far from any global minimizer, even for noise-free
-data (Theorem \ref{thm:fwi}).
-
-This fact implies that local optimization must fail to solve a precise
-version of the FWI problem:
-\begin{quote} given $\epsilon, \lambda > 0$, fine $(m,w)$ in the
-domain of $\lF$ so that
-\begin{equation}
-  \label{eqn:probstat}
-  \|\lF[m]w-d\| \le \epsilon\|d\|.
-\end{equation}
-\end{quote}
-In fact, we show
-that``large residual'' local minima of $\lerr[\cdot,\cdot,d]$ exist,
-at which its value is much greater than its global minimum
-value. Therefore if $1/2\epsilon^2$ is near the global minimum value
-of $\lerr[\cdot,cdot;d]$, then
-local minimization of $\lerr[\cdot,\cdot;d]$ can fail to find a
-solution of the inverse problem. This phenomenon is what is meant by
-``cycle-skipping''.
-
-Extended inversion involves an extension of the model-to-data map, in
-this case $\lF$.  The extension used here consists in dropping the
-support constraint on $w$, so the extended map is simply $F$ as
-defined above ($\lF$ is a restriction of $F$, so $F$ is an extension
-of $\lF$).
-
-To compensate for the resulting indeterminacy in the
-estimation of $m$, we add a penalty term to the objective, defined by
-a penalty operator $A$ with domain and range $L^2(\bR)$: for $m \in
-\bR^+,w\in L^2(\bR), d \in L^2([t_{\rm min},t_{\rm max}])$, define
-\begin{eqnarray}
-  r[m,w;d] & = & \|F[m]w-d\|/\|d\|, \label{eqn:rmserr}\\
-  e[m,w;d] & = & \frac{1}{2}r[m,w;d]^2, \label{eqn:mserr}\\
-  g[w] & = & \frac{1}{2}\|Aw\|^2/\|d\|^2, \label{eqn:pen} \\
-  \Ja[m, w;d] &=& e[m,w;d] + \alpha^2 g[w] \nonumber\\
-  &=& \frac{1}{2}(\|F[m]w-d\|^2 + \alpha^2 \|Aw\|^2)/\|d\|^2.
-\label{eqn:penerr}
-\end{eqnarray}
-
-The normalizations by the data norm $\|d\|$ in the definitions
-\ref{eqn:rmserr}, \ref{eqn:mserr}, \ref{eqn:pen}, and \ref{eqn:penerr} make the values
-of $r$, $e$, $\alpha^2 g$, and $\Ja$ nondimensional, and simplifies
-both the statements of theoretical results and the interpretation of
-numerical examples.
-
-Apart from dimensional correctness, the only obvious requirement on the penalty operator $A$ is that it
-penalize 
-nonzero values of $w$ for large $|t|$, so that minimizers $(m,w)$ of
-$\Ja$ should have $w$
-concentrated near $t=0$. Thus the penalty is a soft stand-in for the
-support constraint. We
-show that a good choice for $A$ is given (essentially) by
-$(Aw)(t)=tw(t)$ (Proposition \ref{thm:rampgood}), and the
-remainder of the discussion in this section makes this assumption. The
-penalty weight $\alpha$ must be chosen somehow; we review a simple
-algorithm for controlling $\alpha$ later in the paper.
-
-Assume that $\mu>0$ and $d = F[m_*]w_* + n$, with $\mbox{supp }w_*
-\subset [-\mu,\mu]$, and denote by $\eta = \|n\|/\|F[m_*]w_*\|$
-the noise-to-signal ratio. We show that any local minimizer $(m,w)$
-of $\Ja$ satisfies 
-\begin{itemize}
-\item If $\eta \le 0.6$,
-  \begin{equation}
-    \label{eqn:mbnd}
-    |m-m_*| \le (1+f(\eta))\frac{\mu}{r},
-  \end{equation}
-  where $f(\eta)=2\eta+O(\eta^2)$ (Theorem \ref{thm:mnoiseres});
-\item 
-  $(m,{\bf 1}_{[-\lambda,\lambda]}w)$ is a solution of the FWI problem
-  as stated above, that is,
-  \begin{equation}
-    \label{eqn:success}
-    \lerr[m, {\bf 1}_{[-\lambda,\lambda]}w;d] \le \frac{1}{2}\epsilon^2,
-  \end{equation}
-  so long as
-  \begin{eqnarray}
-    \lambda & \ge & (2+f(\eta))\mu,\nonumber\\
-    \epsilon &\ge & (4\pi r \alpha \lambda)^2 +\eta.
-    \label{eqn:solcond}
-  \end{eqnarray}
-  (Theorem \ref{thm:ipnoisesuf})
-\end{itemize}.
-That is, the minimization of the extended objective $\Ja$ provides a
-solution of the FWI problem, with constraints on the achievable
-performance and bounds on the errors. All of these
-bounds are shown to be sharp, and the constraints cannot be
-significantly weakened. For example, in conclusion \ref{eqn:solcond}, the support radius $\lambda$ in
-the projection onto $L^2([-\lambda,\lambda])$ must in general be at least twice the
-support radius of the ``noise free'' wavelet $w_*$, plus an additional
-allowance for noise.
-
-Use of model extension to reveal the influence of earth mechanical
-parameters on data is an old idea, and underlies several classes of
-seismic data processing methods in widespread use for decades
-\cite[]{geoprosp:2008}. Use of model extension to create optimization
-principles for inversion is a more recent development
-\cite[]{SymesCar:91,Plessix:00a,Symes:09,LuoSava:11,BiondiAlmomin:SEG12,LeeuwenHerrmannWRI:13,LiuSymesLi:14,Warner:14,ChaurisGP:14,Warner:16,LeeuwenHerrmann:16,Herve2017,HouSymes:Geo18}. Extended
-inversion methods differ by the choice of additional degrees of
-freedom, and by choice of penalty applied to eliminate them in the
-final result. The approach studied in this paper belongs to the {\em
-  source extension} variety: the extra parameters are provided by
-permitting the energy source component of the experimental model to
-have more, or less constrained, parameters than the experimental
-design suggests. \cite{HuangNammourSymesDollizal:SEG19} give an
-overview and taxonomy of source extension methods, with extensive
-references.
-
-Extended inversion is not the only alternative to straightforward
-least-squares data fitting that may overcome cycle-skipping. For
-example, evidence has recently emerged that the Wasserstein metric
-from arising in the theory of optimal transport may provide a measure
-of error between model-predicted and observed seismic data less
-oscillatory than the mean-square
-\cite[]{EngquistYang:GEO18,Metivier:GEO18}.
-
-%In many cases, the constraint posed by membership of the second component $w$ in the subspace $\lW$ is an approximation to an idealized limit $\lambda \rightarrow 0$.
-%It may be that $W_0 = \cap_{\lambda > 0}\lW$ is non-trivial, in which
-%case $F$ may be viewed as an extension of $F_0 = F|_{ M \times
-%  W_0}$. In that case, $F_0$ and $F$ realize the idealized extension
-%structure discussed in many prior works on extended inversion (for
-%example, \cite[]{geoprosp:2008}): $F_0$ is the physical modeling
-%operator, and $W_0$ consists of second components of physical or
-%feasible models. However, in many of these examples, $W_0$ is not
-%computationally accessible, or - even if it is - gives results that
-%closely approximate those obtained with $\lW$ for small $\lambda$. For
-%other examples, for instance that  discussed later in this paper (and
-%implicitly in \cite[]{Warner:16}, for instance), $W_0=\{0\}$ and the
-%family $\{\lW: \lambda > 0\}$ actually embodies all of the necessary
-%physics.
-
-\section{Examples}
-This section presents numerical examples of the estimates of $m$ and
-$w$ obtained by minimization of $\Ja$. We have selected these examples
-to illustrate the conclusions described in the Overview section.
-
-While minimization of $\Ja$ might be tackled directly - by
-alternating minimizations over $m$ and $w$, or by computing updates
-for $m$ and $w$ simultaneously - such joint minimization performs
-poorly, as \cite{YinHuang:16} has shown. The reason for this poor
-performance is that $\Ja$ has dramatically different
-sensitivity to $m$ versus $w$, as observed in the cited reference and elsewhere.
-Instead, a nested approach, in which $w$ is
-eliminated in an inner optimization,
-generally gives far better numerical performance. This approach was
-formalized by \cite{GolubPereyra:73,GolubPereyra:03} as the {\em
-  Variable Projection Method} (``VPM''). Besides being the main
-computational device used in our examples, it also plays a central
-role in the theory developed in the next section.
-
-The minimization of $\Ja$ over $w$ yields a function $\tJa$ of $m$
-alone:
-\begin{equation}
-  \label{eqn:redexp0}
-  \tJa[m;d] = \inf_w \Ja[m,w;d]
-\end{equation}
-Under appropriate conditions, met in our examples (Theorem
-\ref{thm:norminv}, $\tJa$ has an alternate expression
-\begin{equation}
-  \label{eqn:redexp10}
-  \tJa[m;d] = \Ja[m,\aw[m;d];d],
-\end{equation}
-in which $\aw[m;d] \in W$ is the solution of the normal
-equation
-\begin{equation}
-  \label{eqn:norm0}
-  (F[m]^TF[m]+\alpha^2A^TA)w= F[m]^Td.
-\end{equation}
-
-[ examples and discussion - Huiyi]
-
-[ may need to repeat some of the explicit formulas for $\aw, e, g,
-\tJa$ from the next section - refer to theorems]
-
-%\section{Theory}
-
-%\cite{FuSymes2017discrepancy} introduce a version of the discrepancy
-%principle and its application to separable nonlinear least squares
-%problems in penalty form \ref{eqn:basic}. The algorithm developed
-%there is incomplete, however: it lacks  termination criteria. In this
-%section, I review the discrepancy-based algorithm and introduce an
-%appropriate stopping rule, thus completing the development of the
-%algorithm presented in \cite{FuSymes2017discrepancy}.
-
-\section{Findings}
-This section describes the mathematical infrastructure necessary to
-make precise the conclusions described in the Overview section, then
-gives complete statements and proofs, along with those of
-necessary auxiliary results.
-
-We shall use the abbreviations ${\cal B}(X,Y)$ and ${\cal I}(X,Y)$ for 
-the algebra of bounded linear operators from the Hilbert space $X$ to 
-the Hilbert space $Y$, and its subalgebra of invertible
-operators. All of the operators appearing in this discussion are
-members of ${\cal B}(X,Y)$ for some choice of Hilbert spaces $X$ and
-$Y$, and superscript $T$ denotes the transpose or adjoint of a such an
-operator in the sense specified by its domain and range Hilbert structure.
-
-\subsection{An Acoustic transmission Inverse Problem}
-The data prediction or modeling operator for the single-trace acoustic
-transmission problem has been defined above, in equation
-\ref{eqn:mod}, which we restate here for convenience:
-\begin{equation}
-\label{eqn:mod1}
 F[m]w(t)  = p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right) 
@@
 \end{eqnarray}
-We will ignore these computational necessities in this work,
+I will ignore these computational necessities in this work,
 maintaining the definition $W=L^2(\bR)$.
 
-As mentioned earlier, $F[m]$ is surjective for every
-$m \in M$. Since all possible data lie in the range of $F[m]$ for any
-$m \in M$, some restriction of the domain
-of $F$ is necessary in order that fitting the data constrain
-$m$. The constraint proposed in the
-Overview section involves a selection of a maxium support radius
-$\lambda_{\rm max} >0$. Then define for $\lambda \in (0,\lambda_{\rm max}]$:
+As mentioned earlier, $F[m]$ is surjective for every $m \in M$. Since
+all possible data lie in the range of $F[m]$ for any $m \in M$, some
+restriction of the domain of $F$ is necessary in order that fitting
+the data constrain $m$. The constraint employed is the specification
+of a maxium support radius $\lambda_{\rm max} >0$. Then define for
+$\lambda \in (0,\lambda_{\rm max}]$:
 \begin{itemize}
@@
 In terms of this infrastructure, the inverse problem studied in 
-this paper may be stated as a slightly refined version of the problem statement
-\ref{eqn:probstat}:
+this paper may be stated as
 
 \begin{quote}
-\noindent {\bf Problem Statement:}
+\noindent {\bf Inverse Problem:}
   given data $d \in D$, relative error level $\epsilon \in
@@
 \end{equation}
-or equivalently
+\end{quote}
+
+Define the relative mean-square error $\lerr: M \times \lW \times D
+\rightarrow \bR^+$ by
+\begin{equation}
+  \label{eqn:redms}
+  \lerr[m,w;d]=\frac{1}{2}\|\lF[m]w-d\|^2/\|d\|^2,
+\end{equation}
+so that inequality \ref{eqn:probstat0} is equivalent to
 \begin{equation}
@@
 \end{equation}
-with $\lerr$ defined in equation \ref{eqn:redms}.
-\end{quote}
+Minimization of $\lerr$ is the standard least-squares formulation of
+the Inverse Problem defined above: if the global minimum value of
+$\lerr$ is less than $\frac{1}{2}\epsilon^2$, then the global
+minimizer is a solution of the Inverse Problem.
 
@@
 satisfies the data misfit constraint for any $m \in M$.
+
+\noindent{\bf Remark:} I shall refer to the minimization of $\lerr$ as
+``Full Waveform Inversion'' or ``FWI'', as this is the
+terminology used in the seismology literature to identify this and
+similar optimization problems.
 
@@
 
-\subsection{Full Waveform Inversion}
-While $F[m]$ is surjective for every $m \in M$, as noted above, it is
+\section{Full Waveform Inversion}
+While $F$ is surjective, as noted above, it is
 very far from injective. On the other hand, under a constraint that
-will be assumed throughout, $\lF[m]$ is injective (in fact, $4 \pi r
+will be assumed throughout, $\lF[m]$ is injective for each $m \in M$ (in fact, $4 \pi r
 \lF[m]$ is an isometry):
@@
 (failure if $m_0$ is in error by ``half a wavelength''), as follows
-from a more refined analysis of the ``cycle-skipping'' local behaviour of $\lerr$ near its
+from a more refined analysis of the cycle-skipping local behaviour of $\lerr$ near its
 global minimizer.
 
-\subsection{Variable Projection}
+\section{Extended Source Inversion}
+The phenomenon explained in the last section can be avoided by
+reformulating the inverse problem via an extended modeling operator
+and a soft (penalty) constraint to replace the support
+requirement. The extension simply amounts to dropping the support
+constraint, and replacing $\lW$ and $\lF$ by $W$ and $F$.
+
+The hard
+support constraint implicit in the choice of $\lW$ as domain for the
+modeling operator is replaced by a soft constraint in the form of a
+quadratic penalty, with weight operator $A:W \rightarrow W$.
+The choices for the penalty operator $A$ considered here are scalar 
+multiplication operators on $W$ defined by a choice of multiplier $a \in L^{\infty}(\bR)$:
+\begin{equation}
+  \label{eqn:annmult}
+  A w(t)= a(t)w(t), \, t\in \bR.
+\end{equation}
+Explicit choices
+for $a$ are discuss below.
+
+With these choices, define
+\begin{eqnarray}
+  \label{eqn:edef}
+  e[m,w;d] & = & \frac{1}{2}\|F[m]w-d\|^2/\|d\|^2;\\
+  \label{eqn:gdef}
+  g[w;d] & = & \frac{1}{2}\|Aw\|^2/\|d\|^2;\\
+  \label{eqn:jdef}
+  \Ja[m,w;d] & = & e[m,w;d] + \alpha^2g[w;d].
+\end{eqnarray}
+
+\section{Variable Projection}
 The main theoretical device used in the proofs of our main results on 
@@
 (equation \ref{eqn:penerr}) to a function $\tJa$ of 
-$m$ alone, by minimization over $w$, as described in the Examples
-section. We repeat the definition here for the reader's convenience:
+$m$ alone, by minimization over $w$:
 \begin{equation}
@@
 
-The choices for the penalty operator $A$ considered here are scalar 
-multiplication operators on $W$ defined by a choice of multiplier $a \in L^{\infty}(\bR)$:
-\begin{equation}
-  \label{eqn:annmult}
-  A w(t)= a(t)w(t), \, t\in \bR.
-\end{equation}
-With $A$ of this form, $\tilde{J}_{\alpha}$ is explicitly
+With $A$ of the form \ref{eqn:annmult}, $\tilde{J}_{\alpha}$ is explicitly
 computable. First observe that apart from amplitude, $F[m]$ is
@@
 shared with many other inverse problems in wave propagation, as
-explained in the third appendix. Here, we derive it from explicit
+explained in the third appendix. Here, I derive it from explicit
 expressions for $\tJa$ and its components.
@@
   p[m,\aw[m,d];d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r a(t-mr))^2(1 +
-  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)
+  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)^2
 \end{equation}
@@
 
-\subsection{Choice of Penalty Operator}
-
-We examine two choices for $A$. For each choice, we ask first whether local minimizers 
+\section{Choice of Penalty Operator}
+
+I examine two choices for $A$. For each choice, I ask first whether local minimizers 
 of the resulting VPM objective $\tJa[\cdot;d]$ occur far from a slowness $m_*$.
@@
 source inversion 
-\cite[]{Plessix:00a,LuoSava:11,Warner:14,HuangSymes:SEG15a,Warner:16,HuangSymes:GEO17}.
+\cite[]{Plessix:00a,LuoSava:11,Warner:14,HuangSymes:SEG15a,Warner:16,HuangSymes:Geo17}.
+
+\section{Stationary Points}
 
@@
   \begin{equation}
-    | (m - m_*)r| > \lambda  \Rightarrow  \left|\frac{d}{dm}\tJa[m;d_*]\right| > \alpha^2 
-    \frac{\lambda-\mu}{(1+(4\pi r)^2\alpha^2 
-      (\lambda+\mu)^2)^{2}} \|w_*\|^2 
+    | (m - m_*)r| > \lambda  \Rightarrow  \left|\frac{d}{dm}\tJa[m;d_*]\right| >  
+    \frac{r(4 \pi r \alpha)^2(\lambda-\mu)}{(1+(4\pi r\alpha)^2 
+      (\lambda+\mu)^2)^{2}}  
     \label{eqn:gradbndnonoise}
@@
     \label{eqn:gradfinal}
-    \frac{d}{dm}\tJa[m;d_*] = -r \alpha^2 \int \,dt \, 
+    \frac{d}{dm}\tJa[m;d_*] = -\frac{r (4\pi r\alpha)^2}{\|w_*\|^2} \int \,dt \, 
   t(1+(4\pi r)^2 \alpha^2 
@@
   \[
-    = -r\alpha^2\int_{-(m-m_*)r-\lambda}^{-(m-m_*)r+\lambda}
+    = -\frac{r(4 \pi r \alpha)^2}{\|w_*\|^2}\int_{-(m-m_*)r-\lambda}^{-(m-m_*)r+\lambda}
     \,dt\, t(1+(4\pi r)^2\alpha^2 t^2)^{-2}w_*\left(t+(m-m_*)r\right)^2
@@
 
-\begin{corollary}
+\begin{theorem}
   \label{thm:rampreallygood}
@@
   Then $|m-m_*| < \lambda /r$.
-\end{corollary}
+\end{theorem}
 
@@
   solution of the inverse problem \ref{eqn:probstat0} for any $\lambda
-  \ge 2\mu$ and $\epsilon \ge (4\pi r \lambda \alpha)^2$.
+  \ge 2\mu$ and
+  \begin{equation}
+    \label{eqn:estresidnorm}
+    \epsilon \ge \frac{(8\pi r \mu \alpha)^2}{1 + (8\pi r \mu\alpha)^2}.
+  \end{equation}
 \end{proposition}
@@
 \begin{proof}
-  From the assumption $w_* \in W_{\mu}$ and Corollary
+  From the assumption $w_* \in W_{\mu}$ and Theorem
   \ref{thm:rampreallygood}, $|(m_{\infty}-m_*)r|\le \mu$. From the
@@
 = 8 \pi^2 r^2 \alpha^4\int^{\mu}_{-\mu}\,dt\,\frac{|t-(m_{\infty}-m_*)r|^4}{(1+(4\pi r)^2 \alpha^2 
-|t-(m_{\infty}-m^*)r|^2)^{2}}w_*(t)^2.
-\]
-and therefore
-\begin{equation}
-  \label{eqn:estresidnorm}
-e[m_{\infty},\aw[m_{\infty};d];d] \le 8 \pi^2 r^2 (2\mu\alpha)^4 \|w_*\|^2 =
-\frac{1}{2}(8 \pi r \mu \alpha)^4 \|d\|^2
-\end{equation}
+|t-(m_{\infty}-m^*)r|^2)^{2}}w_*(t)^2
+\]
+\[
+  \frac{1}{2} (4\pi r \alpha)^4\int^{\mu}_{-\mu}\,dt\,\frac{|t-(m_{\infty}-m_*)r|^4}{(1+(4\pi r)^2 \alpha^2 
+|t-(m_{\infty}-m^*)r|^2)^{2}}d(t+m_*r)^2
+\]
+\[
+  \le \frac{1}{2} \|d\|^2  \left(\frac{(8\pi r \alpha \mu)^4}{(1+(8\pi
+      r \alpha \mu)^2)^2}\right).
+  \]
 \end{proof}
@@
 
-We turn now to the second issue identified above, the effect of
+I turn now to the second issue identified above, the effect of
 noise. Suppose that the data trace $d$ takes the form
@@
 solution of the inverse problem as defined in the problem statement
-\ref{eqn:probstat0}. In this section, we explain how a solution may
+\ref{eqn:probstat0}. In this section, I explain how a solution may
 nontheless be constructed from a stationary point of $\tJa$.
 
-First we examine the effect of additive noise on the estimation of the
-slowness $m$. In expressing the result, we use the dimensionless
+First, examine the effect of additive noise on the estimation of the
+slowness $m$. In expressing the result, use the dimensionless
 relative data error
@@
   \lambda_{\rm max}$ for $\lambda + \mu$ in the denominator, solve for
-  $\lambda$ and substitute in inequality\ref{eqn:mnoisebndfin}.
-\end{proof}
-
-\noindent {\bf Remark:} This is the only error bound on $m$ in which
-the size of $\alpha$ plays a significant role. It
+  $\lambda$ and substitute in inequality \ref{eqn:mnoisebndfin}.
+\end{proof}
+
+\noindent {\bf Remark:} This bound 
 suggests that error in the estimate of $m$ due to data noise is a
 decreasing function of $\alpha$, at least for small $\alpha$. This result is intuitively
-appealing, and is supported by numerical evidence, as shown below.
+appealing, and is supported by numerical evidence. A similar but stronger bound will be demonstrated in the next section.
 
@@
 all by the assumed support radius of $w_*$. So for this example with 100\% noise, no
-bound of the type given by conclusion 2 could possibly hold. We will
-illustrate this phenomenon numerically later in the paper.
+bound of the type given by conclusion 2 could possibly hold. The
+companion paper \cite[]{SymesChenMinkoff:21}
+illustrates this phenomenon numerically.
 
@@
 
-\noindent {\bf Remark:} We emphasize that Theorem \ref{thm:mnoiseres} states {\em sufficient} conditions for a bound on
+\noindent {\bf Remark:} I emphasize that Theorem \ref{thm:mnoiseres} states {\em sufficient} conditions for a bound on
 the slowness error $|m-m_*|$ in terms of the relative data noise level $\eta$,
@@
 \ref{eqn:mnoisebnd} is effectively much less that 1, resulting in a
-larger range of allowable $\eta$. While we will not formulate such a
-result, one of the numerical examples below suggests its feasibility.
+larger range of allowable $\eta$. While I will not formulate such a
+result, one of the numerical examples in the companion paper
+\cite[]{SymesChenMinkoff:21} suggests its feasibility.
 
@@
 $\lW$, unless the support of the noise $n$ is not restricted. In order
-to construct a solution of the inverse problem \ref{eqn:probstat0}, we
+to construct a solution of the inverse problem \ref{eqn:probstat0}, 
 project $\aw[m;d]$ onto $\lW$. For sufficiently large $\lambda,
@@
     \label{eqn:ipnoiseeps}
-    \epsilon \ge (4 \pi r \alpha\lambda)^2+\eta. 
+    \epsilon \ge \frac{(8 \pi r \alpha\lambda)^2}{1 + (8 \pi r \alpha\lambda)^2}+\eta. 
   \end{equation}    
@@
 
+\section{Selection of $\alpha$}
+
+So far in this story, the selection of the penalty weight $\alpha$ has played a relatively minor role.
+In practical calculation, on the other hand, selection of $\alpha$ strongly influences the convergence of iterative methods and the quality of the results.
+
+Concentration of the wavelet near $t=0$ clearly favors larger $\alpha$: as will be established in Appendix A, the reduced penalty term $g[m,\aw[m;d];d]$ is a decreasing function of $\alpha$ for any $m$.
+As $g$ measures the dispersion of the wavelet away from $t=0$, larger $\alpha$ is to be preferred, all else being equal.
+
+However all else is not equal: the choice of $\alpha$ affects to VPM reduced objective $\tJa[m;d]$ and therefore the estimation of the nonlinear variable $m$ as well.
+To express this connection, observe that for any $d \in D$, $\lambda \in (0,\lambda_{\rm max}]$ and $m \in M$,
+the orthogonal projection of $d$ onto the range $F[m]W_{\lambda}$ is
+$({\bf 1}_{[mr-\lambda,mr+\lambda]}d)(t+mr)$, whence the distance from $d$ to the range of $F$ restricted to $W_{\lambda}$ is
+\[
+\mbox{inf}_{w \in W_{\lambda}} \|F[m]w-d\| = \|(1-{\bf 1}_{[mr-\lambda,mr+\lambda]})d\|.
+\]
+Denote by $r_{\lambda}[m;d]$ this distance normalized by $\|d\|$:
+\[
+r_{\lambda}[m;d] = \frac{ \|(1-{\bf 1}_{[mr-\lambda,mr+\lambda]})d\|}{\|d\|}.
+\]
+Denote by $v_{\lambda}[m;d]$ the pre-image under
+$F$ of the projection of $d$ onto $F[m]W_{\lambda}$:
+\[
+v_{\lambda}[m;d](t) = 4\pi r {\bf 1}_{[-\lambda,\lambda]}(t)d(t+mr).
+\]
+Then using notation introduced earlier in this section for the residual norm,
+\[
+r_{\lambda}[m;d] = r[m,v_{\lambda}[m;d];d]
+\]
+
+\begin{theorem}
+  \label{thm:malpha}
+  Suppose $d \in D$, $\alpha>0$, $0<\lambda \le \lambda_{\rm max}$, $m,m_* \in M$, and condition \ref{eqn:fullrec} holds. Also assume that
+  \begin{equation}
+      \label{eqn:rcomp}
+  \gamma = \frac{r[m,\aw[m;d];d]}{\sqrt{1-r_{\lambda}[m_*;d]^2}} < 1.
+  \end{equation}
+  Then
+  \begin{equation}
+      \label{eqn:mbnde}
+      |m-m_*| \le \lambda/r + \frac{1}{4\pi r^2 \alpha}\left(\frac{\gamma}{1-\gamma}\right)^{1/2}
+   \end{equation}
+\end{theorem}
+
+\begin{proof}
+From the definition, $0 \le r_{\lambda}[m;d] < 1 $ for all $m \in M$, so $\gamma$ is well-defined.
+
+Since condition \ref{eqn:fullrec} is assumed to hold, the conclusions of Proposition \ref{thm:epjgen} are available, in particular, equation \ref{eqn:residnormgen}:
+  \[
+  e[m,\aw[m;d];d] = \frac{1}{2}r[m,\aw[m;d];d]^2 = \frac{1}{2\|d\|^2} \int\, dt\, H(4\pi r\alpha|t-mr|)d(t)^2
+  \]
+  in which 
+  \[
+  H(x) = \left( \frac{x^2}{1+x^2}\right)^2.
+  \]
+  Note that $0 \le H \le 1$ and $H$ is strictly increasing on $\bR^+$.
+
+  Either $|m-m_*|r > \lambda$, or not. In the latter case, the conclusion \ref{eqn:mbnde} follows trivially. In the former case,
+  \[
+  r[m,\aw[m;d];d]^2 = \frac{1}{\|d\|^2} \int\, dt\, H(4\pi r\alpha|t-mr|)d(t)^2
+  \]
+  \[
+  \ge \frac{1}{\|d\|^2} \int\, dt\, H(4\pi r\alpha|t-mr|)({\bf 1}_{[m_*r-\lambda,m_*r+\lambda]}d(t))^2
+  \]
+  \[
+  = \frac{1}{\|d\|^2} \int\, dt\, H(4\pi r\alpha|t-(m-m_*)r|)({\bf 1}_{[-\lambda,\lambda]}d(t+m_*r))^2
+  \]
+  \[
+  \ge \frac{H(4\pi r\alpha(|m-m_*|r-\lambda))}{\|d\|^2} \int\, dt\, ({\bf 1}_{[-\lambda,\lambda]}d(t+m_*r))^2
+  \]
+  \[
+   = \frac{H(4\pi r\alpha(|m-m_*|r-\lambda))}{\|d\|^2} \int\, dt\, ({\bf 1}_{[m_*r-\lambda,m_*r+\lambda]}d(t))^2
+  \]
+  \[
+  = H(4\pi r\alpha(|m-m_*|r-\lambda))(1 - r_{\lambda}[m_*;d]^2)
+  \]
+  whence from the definition \ref{eqn:rcomp} of $\gamma$
+  \begin{equation}
+      \label{eqn:hineq}
+    H(4\pi r\alpha(|m-m_*|r-\lambda)) \le \gamma.
+  \end{equation}
+$H$ is bijective: $\bR^+ \rightarrow [0,1)$. Solving inequality \ref{eqn:hineq} for $4 \pi r \alpha(|m-m_*|r-\lambda)$ and rearranging, obtain the conclusion \ref{eqn:mbnde}. 
+\end{proof}
+
+\noindent {\bf Remark:} The parameters $\alpha$ and $\lambda$ play completely independent roles in this result.
+Given $d \in D$, $m_* \in M$, and $\lambda \in (,\lambda_{\rm max}]$, $r_{\lambda}[m_*;d]$ is the lower bound on relative RMS data error with slowness $m_*$, attained at $w = v_{\lambda}[m;d]$.
+Select $r_*>0$ so that $r_*^2 < 1 - r_{\lambda}[m_*;d]^2$, and suppose the for each $\alpha>0$, it is possible to find $m_{\alpha} \in M$ so that $r[m_{\alpha},w_{\alpha}[m_{\alpha};d];d] = r_*$.
+Then from \ref{eqn:mbnde},
+\[
+|m_{\alpha}-m_*| \le \lambda/r + O(1/\alpha).
+\]
+That is, this bound suggests that amongst $\alpha>0$ that permit attainment (via proper choice of m) of a prescribed value for the RMS data error at the minimizer of $J_{\alpha}[m,\cdot;d]$, then the bound \ref{eqn:mbnde} on the error between this $m$ and the ``target'' slowness $m_*$ is smaller if $\alpha$ is larger.
+This observation motivates the Discrepancy Algorithm described in Appendix A.
+
 \bibliographystyle{seg}
-\bibliography{masterref,local}
+\bibliography{../../bib/masterref}
 
@@
 \end{quote} 
-We describe an alternating, or coordinate search, algorithm for
+I describe an alternating, or coordinate search, algorithm for
 solution of this problem, combining a local optimization
@@
 test for the norm of $\nabla \Phi$ (and a limit on iteration count, of
-course). In our application to $\Ja = e + \alpha^2 g$, we add another
+course). In application to $\Ja = e + \alpha^2 g$, add another
 stopping criterion:
@@
 minimizing sequence, so the expected condition invoking this stopping
-rule is $e<e_-$. We denote a update rule with this enhancement,
+rule is $e<e_-$. Denote a update rule with this enhancement,
 applied to $\tJa$, by $G[\alpha,e_-,e_+,...]$, still allowing the
@@
 
-We will also require a rule for updating $\alpha$, defining a map $H: M \times E
+The algorithm also requires a rule for updating $\alpha$, defining a map $H: M \times E
 \rightarrow \bR^+$, in which $E=\{(e_-,e_+) \in \bR^2: 0 < 
@@
 the error bounds are satisfied: If $(m, e_-, e_+) \in M \times E$ and $\alpha = H(m, e_-,
-e_+)$, then $e[m,\aw[m,d];d] \in [e_-,e_+]$. We will describe two
-such rules below.
+e_+)$, then $e[m,\aw[m,d];d] \in [e_-,e_+]$. A usable rule is
+described below (equation \ref{eqn:alphasecant}).
 
@@
 \end{equation}
-The last inequality ans the strict increase of $e$ imply that $\Psi(\alpha)<e_+$ if $\alpha \in [0,\alpha_+)$.
+The last inequality ans the strict increase of $e$ imply that $\Psi(\alpha)<\alpha_+$ if $\alpha \in [0,\alpha_+)$.
 \end{proof}
@@
       \item $e(d,\alpha_n) \le e_-$ for $0 \le n <N$, and
-      \item $e(d,\alpha_N) \in (e_-,e_+)$
+      \item $e[d,\alpha_N] \in (e_-,e_+)$
   \end{itemize}
@@
 Since $\alpha_{\infty}=\lim_{n\rightarrow \infty}\alpha_n$, there is $m \in {\bf Z}_+$ for which $\alpha_m \in (\alpha_{\infty}-\min(\epsilon/2,\delta),\alpha_{\infty})$, so that $\alpha_{m+1} = \Psi(\alpha_m) >\alpha_m + \epsilon/2>\alpha_{\infty}$, contradicting the definition of $\alpha_{\infty}$ as the limit of the increasing sequence $\{\alpha_n\}$.
-Therefore, we conclude that $\alpha_{\infty}=\alpha_+$.
-
-  Since $\alpha_n <\alpha_+$ for all $n$, according to Proposition \ref{thm:psidef}, there must exist a least $n = N(d,e_-,e_+)$ for which $\alpha_n \in (e_-,e_+)$.
+Therefore, conclude that $\alpha_{\infty}=\alpha_+$.
+
+  According to Proposition \ref{thm:psidef}, $\alpha_n \rightarrow \alpha_+$ and $\alpha_n$ increases with $n$,  whence $e[d,\alpha_n] \rightarrow e_+$ and $e[d,\alpha_n]$ increases with $n$.
+Therefore there must exist a least $n = N(d,e_-,e_+)$ for which $e[d,\alpha_n] \in (e_-,e_+)$.
 \end{proof}
@@
 Accelerated updates using variants of the secant rule are possible, but an analysis of those possibilities is beyond the scope of this paper.
-The numerical examples shown above use the rule \ref{eqn:alphasecant}.
 
@@
 
-This is as far as we can go without more concrete assumptions on $F$ and $A$.
-For a few problems, analysis of the gradient of the reduced objective $\tJa$ shows that its  domain of convexity is large in some sense for a range of $\alpha$, then the discrepancy algorithm sketched here can efficiently drive $\alpha$ to near its largest feasible value, thus giving the best resolution near the global minimizer.
-That program has been carried out for the simple cartoon problem of this paper, and partly for some other more protypical inverse wave problem, as mentioned in the introduction.
+This is as far as one can go without more concrete assumptions on $F$
+and $A$. If the reduced
+objective $\tJa$ possesses a large (in some sense) domain of convexity for a range of $\alpha$, then the discrepancy algorithm sketched
+here can efficiently drive $\alpha$ to near its largest feasible
+value, thus giving the best resolution near the global minimizer. That
+program has been carried out for the simple cartoon problem of this
+paper, and partly for some other more protypical inverse wave problems \cite[]{FuSymes2017discrep}.
+
 \append{Heisenberg, support, and wavelength}
@@
 least.
+
+\append{Abstract Structure of the Gradient}
+The expression \ref{eqn:dexpjgen} for the derivative of the reduced
+objective $\tJa$ is the result of elementary manipulations, based on the
+explicit expression \ref{eqn:mod} for the modeling operator $F$. In
+this appendix I give an alternative derivation that generalizes to
+extended inversion formulations for much less constrained physics. I
+will point out the additional reasoning necessary to reach similar
+conclusions in these more complex instances of extended inversion, as
+presented for example in \cite{StolkSymes:03,StolkDeHoopSymes:09,Symes:IPTA14,tenKroode:IPTA14,HuangSymes:Geo17,HuangSymes:Geo18a,HuangSymes:Geo18b,HuangNammourSymesDollizal:SEG19}.
+
+Recall that $\aw[m;d]$ is the solution of the normal equation
+\ref{eqn:norm}. The reduced objective $\tJa$ is given by
+\[
+  \tJa[m;d] = \Ja[m,\aw[m;d];d]
+\]
+\[
+  = \frac{1}{2}(\|F[m]\aw[m;d]-d\|^2 + \alpha^2\|Aw[m;d]\|^2)
+\]
+(equation \ref{eqn:redexp1})
+\begin{equation}
+  \label{eqn:redsimple}
+  =\frac{1}{2}(\|d\|^2 - \langle d,F[m]\aw[m;d]\rangle),
+\end{equation}
+after a little algebra.
+
+As mentioned above, $F:M \times W \rightarrow D$ is {\em not}
+differentiable. Neither is $\aw: M \times D \rightarrow W$, as follows
+immediately from the identity \ref{eqn:normsol}. However $F\aw: M
+\times D \rightarrow D$ is differentiable, hence so is $\tJa: M \times
+D \rightarrow \bR$ thanks
+to the identity \ref{eqn:redsimple},
+under the conditions on the multiplier $a$ identified in Theorem
+\ref{thm:diffobj}.
+
+For the remainder of this appendix, assume that
+$a \in C^{\infty}(\bR)\cap L^{\infty}(\bR)$ and $a(t)=t$ for $|t|<\tau$ and $|a(t)| \ge
+\tau$ for $|t| \ge \tau$, with $\tau$
+defined in equation \ref{eqn:taudef}.
+
+Proposition \ref{thm:norminvexp}, item [1], implies that
+\begin{equation}
+  \label{eqn:faw}
+  F[m]\aw[m;d] = F[m](F[m]^TF[m] + \alpha^2 A^TA)^{-1}F[m]^T d.
+\end{equation}
+The normal operator and its inverse are multiplication
+operators, whereas $F$ and $F^T$ are scaled shift operators, inverse
+to each other except for scale. From this it is easy to see that the
+operator on the RHS of equation \ref{eqn:faw} is multiplication by a
+smooth function, with its arguments shifted by $mr$. Such an operator
+is smooth in $m$, hence so is $F\aw$.
+
+In the other extended inversion settings mentioned at the beginning of
+this appendix, $F$ is a microlocally elliptic Fourier Integral
+Operator (FIO) \cite[]{Dui:95}. The canonical relation takes on the role
+of the shift operator $t \rightarrow t-mr$, and has the properties
+necessary to conclude that the composition of $F$ and $F^T$ in both
+ordersis pseudodifferential, at least in an open conic subset of the
+cotangent bundle. If $A^TA$ is pseudodifferential, then so is the
+normal operator. The inverse in the RHS of equation \ref{eqn:faw} must
+be replaced by a microlocal parametrix, and a smooth error added to
+the RHS. Thanks to Egorov's Theorem \cite[]{Tay:81}, a special case of
+the rules for composing FIOs, the operator on the RHS of equation
+\ref{eqn:faw} is a pseudodifferential operator whose symbol is
+algebraic in geometric optics quantities, hence depends smoothly on
+the coefficients in the wave equation, as does $F\aw$.
+
+Since $\tJa$ and its gradient depend smoothly on their arguments, it
+is possible to derive an alternate expression assuming that $d \in
+H^1(\bR)$, then extend it by continuity to $d \in L^2(\bR)$. Note that
+if $w \in H^1(\bR)$, then
+  differentiable, and 
+\begin{equation}
+\label{eqn:deriv}
+(D(F[m]w)\delta m)(t) = F[m](Q[m]\delta m)w (t), 
+\end{equation}
+where 
+\begin{equation}
+\label{eqn:defq}
+(Q[m]\delta m)w = -r\delta m \frac{dw}{dt}. 
+\end{equation}
+
+That is, $DF[m]\delta m$ factors into $F[m]$ following $Q[m]\delta m$,
+where the latter a skew-adjoint differential operator of order 1,
+depending linearly on $\delta m$.
+
+Assume that $d \in H^1(\bR)$. From equation \ref{eqn:normsol}, it
+follows that $w[m;d] \in H^1(\bR)$ and moreover that $m \mapsto
+\aw[m;d]$ is differentiable as a map from $\bR^+$ to
+$L^2(\bR)$. From equation \ref{eqn:redexp1},
+\[
+  \tJa[m;d] = \frac{1}{2}(\|F[m]w[m;d]-d\|^2 + \alpha^2 \|Aw[m;d]\|.
+\]
+whence $m \mapsto \tJa[m;d]$ is differentiable. A standard calculation
+invoking the normal equation \ref{eqn:norm} shows that
+\[
+  d\tJa[m;d]\delta m = \langle D(F[m]\aw[m;d])\delta m, F[m]\aw[m;d]-d\rangle
+\]
+\[
+  = \langle F[m](Q[m]\delta m)\aw[m;d], F[m]\aw[m;d]-d\rangle
+\]
+\[
+  = \langle (Q[m]\delta m)\aw[m;d], F[m]^T(F[m]\aw[m;d]-d)\rangle
+\]
+\[
+  = -\alpha^2\langle (Q[m]\delta m)\aw[m;d],A^TA\aw[m;d] \rangle
+\]
+\[
+  = \alpha^2 \langle \aw[m;d], (Q[m]\delta m)A^TA \aw[m;d]\rangle
+\]
+(using antisymmetry of $Q$)
+\[
+  = \alpha^2 \langle \aw[m;d],[ (Q[m]\delta m), A^TA]\aw[m;d] \rangle
+  + \alpha^2 \langle (Q[m]\delta m)\aw[m;d],A^TA\aw[m;d] \rangle
+\]
+Rearranging,
+\begin{equation}
+  \label{eqn:djq}
+  d\tJa[m;d]\delta m = \frac{1}{2}\alpha^2 \langle \aw[m;d],[
+  (Q[m]\delta m), A^TA]\aw[m;d] \rangle.
+\end{equation}
+Since $Q$ is a differential operator of order 1, and $A^TA$ an
+operator of order 0, the commutator has order 0. That is, the RHS of
+equation \ref{eqn:djq} defines a continuous quadratic form in $d$ with
+respect to the $L^2$ norm. As shown above, the same is true of the
+left hand side. Therefore their extensions by continuity to $d \in
+L^2(\bR)$ are the same, and the identity \ref{eqn:djq} holds for $d
+\in L^2(\bR)$.
+
+Making the identity \ref{eqn:djq} explicit by means of equation
+\ref{eqn:defq} and $A^TA = t^2$, substituting the expression
+\ref{eqn:normsol} for $\aw[m;d]$, and rearranging: one obtains
+precisely the expression \ref{eqn:dexpjgen}.
+
+For the more general contexts mentioned earlier, a similar derivation
+is possible. The analog of $Q$ is a pseudodifferential operator of
+order 1, which is essentially (to leading order) skew-symmetric, and
+the more general version of the identity \ref{eqn:djq} is approximate,
+with lower-order (smoother) error terms.
+
+These ideas were used in \cite{tenKroode:IPTA14,Symes:IPTA14} to show
+that the analog of $\tJa$ is tangent to second order to a convex
+quadratic form, related to the Hessian of an objective formulated in
+terms of traveltime. This relation explains the ability of the variant
+of extended inversion studied in the cited references to recover
+kinematically accurate velocity fields. Of course, that is what is
+shown in detail in this paper for the very simple model problem
+studied here.

### 2022-01-11 repro/papers/discreptheorypub/paper.tex
more

--- 2021-10-29 repro/papers/discreptheory/paper.tex
@@
-\title{Solution of an Acoustic Transmission Inverse Problem by
-  Extended Inversion: Theory}
-\author{William W. Symes\footnotemark[1]}
-\address{\footnotemark[1]PO Box 43, Orcas, WA 98280 USA, {\tt symes@rice.edu}}
-
-\lefthead{Symes}
-
-\righthead{1D Acoustic Inversion: Theory}
+\title{Error Bounds for Extended Source Inversion applied to an Acoustic Transmission Inverse Problem}
+\author[Symes]{William W. Symes$^1$\\
+$^1$ PO Box 43, Orcas, WA 98280 USA, {\tt symes@rice.edu}}
 
@@
 \begin{abstract}
-  A single-trace transmission inverse problem for the wave equation
-  seeks to determine both the wave velocity in a homogenous acoustic
-  medium and the transient waveform of an isotropic point source. The
-  duration (support) of the source waveform and the source-to-receiver
-  distance are assumed known. A least squares formulation of this
-  problem exhibits the ``cycle-skipping'' behaviour observed in field
-  scale problems of this type, with many local minima differing
-  greatly from the global minimizer. This behaviour is eliminated by
-  dropping the hard support constraint on the source waveform,
-  replacing it by a soft penalty implemented as a weighted mean-square
-  of the source waveform. For properly chosen weight function,
-  penalizing nonzero values away from $t=0$, the velocity component of
-  any stationary point differs from the global minimizer of the
-  constrained least-squares formulation by a linear combination of the
-  source waveform support radius and data noise-to-signal ratio. Given
-  an estimate of data noise, the penalty weight can be dynamically
-  adjusteded during iterative optimization to maximize predicted data
-  accuracy and closely approximate a support-constrained source waveform.
+  A simple inverse problem for the wave equation requires
+  determination of both the wave velocity in a homogenous acoustic
+  material and the transient waveform of an isotropic point radiator,
+  given the time history of the wavefield at a remote point in
+  space. The duration (support) of the source waveform and the
+  source-to-receiver distance are assumed known. A least squares
+  formulation of this problem exhibits the ``cycle-skipping''
+  behaviour observed in field scale problems of this type, with many
+  local minima differing greatly from the global minimizer. An
+  extended formulation, dropping the support constraint on the source
+  waveform in favor of a weighted quadratic penalty, eliminates this
+  misbehaviour. With proper choice of the weight operator, the
+  velocity component at {\em any} local minimizer of this extended
+  objective function differs from the global
+  minimizer of the least-squares formulation by less than a linear
+  combination of the source waveform support radius and data
+  noise-to-signal ratio.
 
@@
 testing, medical imaging, and other areas of science and engineering
-in which wave motion plays an important role. Often these problems are
-formulated as optimization of mean square (or other scalar measure)
-functions of misfit between observed and predicted data. In many
-contexts, such as exploration seismology, the computational size of
-these problems is very large, and even the prediction (or modeling) of
-the data requires use of high-performance computing
-environments. Accordingly, rapidly convergent optimization algorithms
-based on Newton's method and its relatives to update model parameters
-are the numerical methods of choice. These however are local, in the
-sense that they converge to stationary points of the objective
-functions, usually near the initial model parameters. For
-high-frequency (hence high resolution) data regimes, stagnation of
-local optimization methods at physically irrelevant model vectors is
-frequently observed. This ``cycle-skipping'' is one of the main
-obstacles to widespread successful deployment of optimization-based
-methods for inverse problems in wave propagation
+in which wave motion plays an important role. These problems may be
+formulated as nonlinear least squares problems, requiring optimization
+of mean square misfit between observed and predicted data. Iterative
+local numerical optimization methods (relatives of Newton's method)
+are feasible solution approaches for the computationally large
+variants of these nonlinear least squares problems that occur in
+seismology and medical imaging. However, local optimization produces
+only approximate stationary points, and these exhibit a tendency to
+stagnate at physically irrelevant solutions, under typical conditions
+of data acquisition. This ``cycle-skipping''
+pathology can sometimes be overcome through use of information other than
+measured wave data, however it remains a serious obstacle to effective
+solution of field-scale inverse problems in wave propagation
 \cite[]{GauTarVir:86,VirieuxOperto:09,Fichtner:10,Plessix:10,Schuster:17}.
 
-This paper shows how a modification of least-squares data fitting via
-{\em modeling operator extension} overcomes cycle-skipping in a simple
-example. The problem studied here may be the simplest inverse wave
-problem that exhibits cycle-skipping - in fact, essentially this
-example is routinely used to illustrate cycle-skipping (see for
-example \cite{VirieuxOperto:09}, Figure 7). Its data is a single trace
-at a point in $\bR^3$ (``receiver location'') of the causal linear
-acoustic pressure field, generated by an isotropic radiator acting at
-another point (``source location'') at a known distance from the
-receiver location. The acoustic material is supposed spatially
-homogeneous, hence characterized by wave slowness (reciprocal
-velocity) and density. The object of the inverse problem is to recover
-the slowness and the time history of the isotropic radiator
-(``wavelet'') from the recorded trace. In order that the data
-constrain the slowness, the wavelet is presumed to have known compact
-support. Thus the (``physical'') modeling operator has as its domain
-the Cartesian product of an interval of positive slownesses, and the
-subspace of $L^2(\bR)$ consisting of wavelets with the prescribed
-support. Even for noise-free data (in the range of the modeling
-operator), the obvious least squares problem formulated in terms of
-this modeling operator exhibits many local minimizers having nothing
-to do with the (``target'') slowness and wavelet used to create the
-data - that is, cycle-skipping occurs (Theorem \ref{thm:fwi}).
-
-Dropping the support constraint on the wavelet creates an extension of
-the modeling map for which any data may be fit exactly, for any choice
-of slowness. The link between data and slowness is restored by adding
-a quadratic penalty on non-zero values of the wavelet outside of the
-physical support interval to the extended mean-square data misfit. The
-first main result of this paper is that for noise-free data, the
-slowness component of {\em any} stationary point of this extended
-penalty function differs from the target slowness by a multiple of the
-target wavelet radius (Theorem \ref{thm:rampreallygood}). That is, cycle-skipping does not occur, and the
-wavelet support radius determines the resolution of slowness. For
-noisy data (that is, sum of a noise-free signal trace in the range of
-the physical modeling operator and a square-integrable noise trace),
-any stationary point differs from the target slowness by a combination
-of the wavelet support radius and the noise-to-signal ratio (Theorem \ref{thm:mnoiseres}). Since the
-wavelet component of the extended domain does not constrain support,
-stationary points of the extended penalty function do not provide
-solutions of the original, physical inverse problem, which includes
-the support constraint. Such solutions can be obtained, with estimates
-on support radius and data error, simply by truncating the wavelet
-components of stationary points (Theorem \ref{thm:ipnoisesuf}).
-
-The inverse problem studied here is far too simple to have any direct
-use in applications. However it seems worthy of careful study for two
-reasons. First, its simplicity allows an unusually complete account of
-its properties. The present paper details its mathematical behaviour,
-including explicit and sharp error estimates rarely available for nonlinear
-inverse problems. A companion paper \cite[]{SymesChenMinkoff:21}
-describes numerical examples that illustrate the mathematical
-conclusions.
-
-Second, the analysis reveals several important features shared with
-similar approaches to other wave inverse problems, and so provides an
-at least partial pattern for reaching similar conclusions with
-immediate practical implications. For example, nested optimization
-(variable projection method, \cite{GolubPereyra:03}) based on a model
-decomposition into inner and outer variables seems to be very
-important: in some cases, such as the simple problem presented here,
-the decomposition is obvious (inner variable = wavelet, outer variable
-= slowness), in others less so
-\cite[]{geoprosp:2008,Terentyev:thesis}. The extended modeling
-operator must be surjective, or at least have a dense range, {\em for
-  each value of the outer variable}, so that data may be fit well even
-for a poor initial guess of the outer variable. for the extended
-inversion approach to be successful. The derivative of the extended
-modeling operator with respect to the outer variable is
-well-approximated by the composition of the extended modeling operator
-itself and a pseudodifferential operator of order 1
-\cite[]{Symes:IPTA14,tenKroode:IPTA14}. The choice of the penalty
-operator, controlling the extended degrees of freedom, is critical: in
-order to produce an objective immune from cycle-skipping, this penalty
-operator must be (pseudo-)differential
-\cite[]{StolkSymes:03}. Generally, the extended modeling operator is
-not differentiable in the outer variable. However these last two facts
-together imply the differentiability of the reduced objective function
-from variable projection (Theorem \ref{thm:diffobj}, Appendix C). Because of the data-fitting
-assumption, a straightforward algorithm for scaling the quadratic
-penalty, based on the Discrepancy Principle, is available - see
-\cite{FuSymes2017discrepancy,SymesChenMinkoff:21}, also Appendix
-A. This scaling varies dynamically during iterative optimization, in
-effect changing the objective function sporadically as the iteration
-converges.
-
-The approach explained here is an example of {\em extended source}
-inversion, see \cite{HuangNammourSymesDollizal:SEG19} for a review,
-and \cite{MetivierBrossier:SEG20} for a recent variation. Adding
-degrees of freedom to the coefficients in the wave equation in various
-ways produces so-called {\em extended model} inversion approaches,
-which actually have a longer history in seismology - see
-\cite{geoprosp:2008} for details. The need to overcome cycle-skipping
-has spawned a great number of concepts, of which extension-based
-inversion is only one. Least-squares inversion succeeds with
-sufficiently good initial estimates of wave velocity, and the dominant
-techniques in industrial and academic seismology exploit data other
-than direct measurements of waves, for example measurement of wave
-time-of-flight, to supply such initial estimates
-\cite[]{VirieuxOperto:09,Fichtner:10,Schuster:17}. Use of a data
-misfit measure other than least squares, such as versions of the
-Wasserstein metric from optimal transport theory, also shows promise
-\cite[]{Metivier:GEO18,EngquistYang:GEO18}.
-
-The next section gives precise statements of the inverse problem
-studied here, and its various components. The following sections
-contain statements and proofs of the main results: the existence of
-muliple stationary points of the least squares problem, properties of
-the extended modeling operator and the reduced objective, choice of
-the penalty operator, the main results mentioned above about
-stationary points of the reduced objective, and the discrepancy-based
-algorithm for adjusting the penalty scale. Three appendices justify
-the discrepancy-based scale adjustment, review the link between
-constrained support and spectrum provided by the Heisenberg inequality,
-and expose the abstract structure underlying the differentiability of
-the reduced objective.
-
-I use the abbreviations ${\cal B}(X,Y)$ and ${\cal I}(X,Y)$ for 
-the algebra of bounded linear operators from the Hilbert space $X$ to 
-the Hilbert space $Y$, and its subalgebra of invertible
-operators. All of the operators appearing in this discussion are
-members of ${\cal B}(X,Y)$ for some choice of Hilbert spaces $X$ and
-$Y$, and superscript $T$ denotes the transpose or adjoint of a such an
-operator in the sense specified by its domain and range Hilbert structure.
+Since numerical feasibility limits practical optimization approaches
+to search for stationary points, it is natural to explore alternatives
+to nonlinear least squares formulations for which approximate
+stationary points might perforce be acceptable solutions. Many such
+alternatives have been suggested and applied in both synthetic and
+field data tests, in some cases with excellent results (see
+\cite{HuangNammourSymesDollizal:SEG19},
+\cite{PladysBrossierLiMetivier:GEO21} for recent reviews). However
+numerical tests are inevitably ``a look in the rear view mirror'',
+with no guarantee that the next example will not behave quite
+differently. See \cite{Symes:2020} for an example of such an unpleasant
+surprise. Theoretical guarantees that stationary points are
+satisfactory solutions of specific inverse problems, in some specific
+sense and under specified conditions, are almost entirely absent from
+the literature.
+
+This paper provides such a theoretical guarantee for a modification of
+least-squares data fitting via {\em modeling operator extension},
+applied to a very simple example: recovery of the (spatially
+homogeneous) wave speed of an acoustic material model, together with
+the waveform of an isotropic point radiator (``wavelet'') with
+specified support, from the time history (``trace'') of the resulting
+pressure field at a single remote location. This is perhaps the
+simplest inverse wave propagation problem that exhibits
+cycle-skipping. While far too simple to have immediate
+practical application, it is a special case, or subproblem, of many
+field-scale inverse problems in industrial and academic seismology. It
+is also routinely used to illustrate cycle-skipping (see for example
+\cite{VirieuxOperto:09}, Figure 7). Modeling operator extension ideas underlie
+widely used techniques in seismic data processing, and have been
+applied to inversion for several decades
+\cite[]{geoprosp:2008}. The particular approach used in this work
+is an example of {\em source extension}.
+\cite{HuangNammourSymesDollizal:SEG19} review recent source extension
+contributions, some of which are closely related to the approach to
+single-trace transmission inversion studied here.
+
+Model extension is only one of several ideas proposed to remedy
+cycle-skipping. For instance, measuring data misfit with versions of the Wasserstein
+metric from optimal transport theory, rather than the $L^2$ norm, also
+shows promise in mitigating cycle-skipping
+\cite[]{Metivier:GEO18,EngquistYang:GEO18,Ramos-Martinez:SEG18,Wang:SEG19,EngquistYang:CPAM21}. \cite{MahankaliYang:21}
+have established conditions under which the Wasserstein misfit is
+convex in a few parameters of simple models. Aside from the present
+paper, their work is one of the very few to give a complete mathematical
+justification (albeit also in very
+restricted circumstances) for application of local optimization to a nonlinear
+least squares alternative.
+
+The next section precisely defines the single-trace transmission
+inverse problem, and the following section confirms the
+cycle-skipping behaviour of the obvious nonlinear least squares
+formulation, that is, that stationary points exist at arbitrarily
+large distance from any global minimizer (Theorem \ref{thm:fwi}). The
+modeling operator extension used here simply consists in dropping the
+support constraint, thus allowing fit to any data with any wavespeed,
+by suitable choice of wavelet. The link between wave speed and data is
+re-established by supplementing the mean-square data misfit function
+with a suitable quadratic penalty on nonzero values of the wavelet far
+from the specified support. I show that any stationary point of the
+resulting penalty function deviates from the global solution of the
+original, constrained nonlinear least squares problem by an amount
+proportional to a combination of the support diameter and the relative
+data error, and this bound is sharp
+(Theorems \ref{thm:rampreallygood} and \ref{thm:mnoiseres}). A
+solution of the original inverse problem (with wavelet support
+constraint) is recovered by truncating the wavelet obtained in the
+course of minimizing the penalty function (Theorem
+\ref{thm:ipnoisesuf}).
+
+The elementary
+analysis developed here leans on the special features of the problem
+under study to achieve sharper results than will likely be possible in
+more general contexts. The Appendix recasts some of these
+constructions in a form that applies, at least formally, to more
+prototypical, larger-scale problems
+\cite[]{Symes:IPTA14,tenKroode:IPTA14}, and suggests a route to
+achieving a similar, mathematically complete understanding of
+extension-based approaches to applied wave propagation inverse
+problems.
 
 \section{An Acoustic transmission Inverse Problem}
-According to linear acoustics, the causal pressure field due to an
-isotropic point radiator at $\bx_s \in \bR^3$ 
-solves the the wave equation \cite[]{Frie:58}:
+According to linear acoustics, the causal pressure field $p(\bx,t)$ due to an
+isotropic point radiator at $\bx_s \in \bR^3$ with time-varying
+intensity (``wavelet'') $w(t)$, propagating in an elastic fluid with
+slowness (reciprocal velocity) $m$ and density $1$ (in appropriate
+units), is the solution of the the initial value problem for the wave equation \cite[]{Frie:58}:
 \begin{eqnarray}
@@
 material model: for crustal rock, a reasonable choice would be
-$m_{\rm min}=0.125, m_{\rm max}=0.6$ s/km.
+$m_{\rm min}=0.125, m_{\rm max}=0.6$ s/km. The pressure trace
+$p(\bx_r,\cdot)$ is square-integrable if and only if the same is true
+of the wavelet $w$. Since the square-integral of the pressure trace is
+proportional to the accumulated energy transferred from the fluid to
+the sensor \cite[]{SantosaSymes:00}, assume that $w \in L^2(\bR)$.
 
@@
   \label{eqn:mapprop}
-  \mbox{for }m \in M, F[m] \in {\cal B}(W,D),\mbox{ and }\|F[m]\| =
+  \mbox{for }m \in M, F[m] \mbox{ is bounded, and }\|F[m]\| =
   \frac{1}{4\pi r}.
 \end{equation}
-Note also that $F[m]$ is surjective for every $m \in M$.
-
-\noindent{\bf Remark:} In computational practice, $W$ will have to be replaced by a
-finite-dimensional subspace of $L^2(\bR)$. Many such choices will
-implicitly limit the support of $w \in W$ to a bounded interval, say
-$[T_{\rm min},T_{\rm max}]$. To maintain the surjective property,
-these bounds should be chosen so that $[t_{\rm min}, t_{\rm max}]
-\subset [T_{\rm min}+mr,T_{\rm max}+mr]$ for all $m \in M$, that is,
-\begin{eqnarray}
-  \label{eqn:dombounds}
-  t_{\rm min} &\ge & T_{\rm min}+m_{\rm max}r,\nonumber\\
-  t_{\rm max} &\le & T_{\rm max}+m_{\rm min}r.
-\end{eqnarray}
-I will ignore these computational necessities in this work,
-maintaining the definition $W=L^2(\bR)$.
-
-As mentioned earlier, $F[m]$ is surjective for every $m \in M$. Since
+
+Since
 all possible data lie in the range of $F[m]$ for any $m \in M$, some
 restriction of the domain of $F$ is necessary in order that fitting
-the data constrain $m$. The constraint employed is the specification
-of a maxium support radius $\lambda_{\rm max} >0$. Then define for
-$\lambda \in (0,\lambda_{\rm max}]$:
+the data constrain $m$. The constraint employed in this work is the specification
+of a maxium support radius $\lambda_{\rm max} >0$ (see the companion
+paper \cite{SymesChenMinkoff:21} for a justification of this choice).
+
+For
+$\lambda \in (0,\lambda_{\rm max}]$, define
 \begin{itemize}
@@
 \noindent{\bf Remark:} The constraint $\epsilon < 1$ imposed on the
-target noise level is eliminates the obvious choice $(m,0)$, which
-satisfies the data misfit constraint for any $m \in M$.
+target noise level eliminates the obvious choice $(m,0)$, which
+satisfies the data misfit constraint for any $m \in M$ if $\epsilon
+\ge 1$.
 
@@
 similar optimization problems.
-
-\noindent{\bf Remark:} The support constraint is closely linked to the
-folk theorem about FWI noted many times in the literature: convergence
-of a descent method requires that the initial slowness must be known
-to ``within a (fraction of a) wavelength''. The relation is a
-consequence of Heisenberg's inequality, and will be reviewed in
-Appendix B.
 
@@
   \end{equation}
-  Then $\lF[m]\in {\cal B}(\lW,D)$ is coercive for every $m \in M, \lambda \in
+  Then $\lF[m]$ is coercive for every $m \in M, \lambda \in
   (0,\lambda_{\rm max}]$.
@@
  \lerr[m,w;d] =  \frac{1}{32\pi^2
-    r^2\|d\|^2}\int_0^T\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2
+    r^2\|d\|^2}\int\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2
 \]
@@
 directions, and descent-based optimization tends to fail if the
-initial estimate $m_0$ is in error by more than $2\lambda/r$
-(``further than a multiple of a wavelength'', per discussion
-in the second Appendix). In fact the actual behaviour of FWI itererations is worse
+initial estimate $m_0$ is in error by more than $2\lambda/r$. In fact the actual behaviour of FWI itererations is worse
 (failure if $m_0$ is in error by ``half a wavelength''), as follows
@@
 
-\section{Variable Projection}
 The main theoretical device used in the proofs of our main results on 
 extended inversion is Variable Projection reduction of the penalty objective $\Ja$
-(equation \ref{eqn:penerr}) to a function $\tJa$ of 
+(equation \ref{eqn:jdef}) to a function $\tJa$ of 
 $m$ alone, by minimization over $w$:
@@
 \end{equation}
-
 A minimizer $w$ on the right-hand side of definition
@@
 \end{equation}
+(here and in the following, the superscript $T$ signifies adjoint in
+the sense of domain and range Hilbert structures).
 
@@
 \end{equation}
+Evidently, $F^TF$ (the first summand in equation \ref{eqn:norm2}) is
+not coercive. If $A$ is to yield small output for input localized near
+$t=0$, as it must to play its desired role in replacing the hard
+support constraint, then $A^TA$ must not be coercive either, that is,
+its spectrum must have zero as its infimum. Therefore an additional
+condition on $A$ is required to enable application of the Lax-Milgram
+constrruction to the normal
+equation \ref{eqn:norm}. This condition is an hypothesis of
 
@@
     max}], \alpha > 0,$ and that  $C>0$ exists so that $a \in L^{\infty}(\bR)$
-  mentioned in condition \ref{eqn:annmult} satisfies condition
-  \ref{eqn:abnd}.
+  mentioned in equation \ref{eqn:annmult} satisfies the condition
+\begin{equation}
+    \label{eqn:abnd} 
+    a \ge 0; \, a(t) \ge C\mbox{ for }|t| \ge \lambda_{\rm max}.
+  \end{equation}
 %  \begin{equation}
@@
   given by equation \ref{eqn:annmult} for $a \in L^{\infty}(\bR)$
-  satisfying
-  \begin{equation}
-    \label{eqn:abnd} 
-    a \ge 0; \, a(t) \ge C\mbox{ for }|t| \ge \lambda_{\rm max}.
-  \end{equation}
-  Then \begin{itemize}
+  satisfying condition \ref{eqn:abnd}.  Then
+  \begin{itemize}
   \item[1. ]the reduced objective $\tJa$ is given by
@@
 
-\noindent{\bf Remark:} If $\Ja[\cdot,\cdot;d]$ and $\tJa[\cdot;d]$
+If $\Ja[\cdot,\cdot;d]$ and $\tJa[\cdot;d]$
 were differentiable, then ``local minimizer'' in the conclusion of the
@@
 penalty operator $A$.
-
 This conclusion follows from properties of the modeling operator $F$
 shared with many other inverse problems in wave propagation, as
-explained in the third appendix. Here, I derive it from explicit
+explained in the Appendix. Here, I derive differentiability from explicit
 expressions for $\tJa$ and its components.
@@
 
-  That the VPM objective $\tJa$ is given by \ref{eqn:expjgen} follows from equations \ref{eqn:pen},
-  \ref{eqn:redexp}, \ref{eqn:residnormgen}, and
+  That the VPM objective $\tJa$ is given by \ref{eqn:expjgen} follows from equations \ref{eqn:redexp1}, \ref{eqn:residnormgen}, and
   \ref{eqn:anninormgen}.
@@
 $\tJa$ when the data is noise-free, that is, the context of
-Proposition \ref{thm:norminvexp}, item 3.
-
+Proposition \ref{thm:norminvexp}, item 3:
 \begin{corollary}
@@
 = \frac{\alpha^4}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^4(1+(4\pi r)^2 \alpha^2 
-    a(t-(m-m^*)r)^2)^{-2}w_*(t)^2.
+    a(t-(m-m_*)r)^2)^{-2}w_*(t)^2.
 \end{equation}
@@
 \end{equation}
-Finally, if $a \in W^{1,1}_{\rm loc}(\bR)$, then $\tJa[\cdot;d]$ is differentiable, and 
+Finally, if $a \in W^{1,\infty}(\bR)$, then $\tJa[\cdot;d]$ is differentiable, and 
 \begin{equation}
@@
 
-\section{Choice of Penalty Operator}
-
-I examine two choices for $A$. For each choice, I ask first whether local minimizers 
-of the resulting VPM objective $\tJa[\cdot;d]$ occur far from a slowness $m_*$.
-
-A penalty operator $A$ of which $\lW$ is the null space would be a
-natural choice. Such operators have come to be called
-``annihilators'', since they map all members of the constraint
-subspace $\lW$ to zero.  A simple example is
-\begin{eqnarray}
-  A = E^c_{\lambda}&=&I - E_{\lambda},\mbox{ where } \nonumber \\
-  E_{\lambda}w(t) &=&{\bf 1}_{[-\lambda,\lambda]}(t)w(t). 
-                      \label{eqn:ann0}
-\end{eqnarray}
-That is, $E_{\lambda}$ is the orthogonal projector onto $\lW$,
-and $E_{\lambda}^c$ is the orthogonal projector onto its
-orthocomplement, an operator of the form \ref{eqn:annmult} with $a
-= 1 - {\bf 1}_{[-\lambda,\lambda]}$. %Note that this choice of $a$
-%satisfies condition \ref{eqn:abnd} with $C \equiv 1$.
-
-\begin{theorem}
-  \label{thm:boxcarbad}
-  Suppose that
-  \begin{itemize}
-  \item[1. ] $\lambda \in (0,\lambda_{\rm max}]$;
-  \item[2. ] $m_*\in M, w_*\in \lW$, and $d=F[m_*]w_*$ (noise-free
-    data);
-  \item[3. ] $A=E^c_{\lambda}$, that is, $a=1-{\bf 1}_{-\lambda,\lambda]}$ in
-    the definition \ref{eqn:annmult}.
-  \end{itemize}
-  Then if $|m-m_*| >  2\lambda/r$,
-  \[
-    \tJa[m;d] = \frac{(4\pi r \alpha)^2}{2(1+(4 \pi r \alpha)^2)}.
-  \]
-\end{theorem}
-
-\begin{proof} of Theorem \ref{thm:boxcarbad}
-
- This is clear from the definition $a = 1-{\bf 1}_{-\lambda,\lambda]}$ and equation \ref{eqn:expjnonoise}, as the supports of $w_*$ and ${\bf
-    1}_{[-\lambda,\lambda]}(t-(m-m_*))$ are disjoint for the range of
-  $m$ identified in the theorem.
-\end{proof}
-
-\noindent {\bf Remark.}
-One might have thought that $A=E^c_{\lambda}$ would be a better choice
-of annihilator, as for noise-free
-data, the solution set defined by the problem statement
-\ref{eqn:probstat0} is the same as the set of global minimizers of
-$\Ja$ in this case. However, for this choice of annihilator,
-$t\Ja$ exhibits the same feature as the mean square error $e$, namely
-a continuum of local minimizers at any distance from the global
-minimizer $m_*$ greater than a multiple of $\lambda$. Therefore the
-extended inversion with this choice of annihilator is no more amenable
-to local optimization than is FWI.
-
-A second possible 
-penalty operator penalizes energy away from 
-$t=0$: choose $\tau > 0$ and set 
+Recall that the purpose of the penalty operator is to penalize energy away from 
+$t=0$. A simple multiplier of class $W^{1,\infty}$ that accomplishes
+this goal (and produces a differentiable reduced objective $\tJa$,
+thanks to Theorem \ref{thm:diffobj}) is
 \begin{equation}
@@
 \end{equation}
-Note that $a \ge 0$ and $a \in L^{\infty}(\bR) \cap W^{1,1}_{\rm
-  loc}(\bR)$. The cutoff $\tau$ will be chosen large enough to be effectively inactive: 
-specifically, hindsight suggests 
+for suitable $\tau>0$. In fact, 
+the cutoff $\tau$ will be chosen large enough to be effectively inactive: 
+hindsight suggests 
 \begin{equation}
@@
 \end{equation}
-This 
-particular annihilator has been employed in earlier papers on extended 
+Essentially this 
+particular annihilator has been employed in earlier work on extended 
 source inversion 
@@
   $d$ is given by
-  \ref{eqn:defdatanonoise} with $w_* \in W_{\mu}$, and  $m_{\infty}$ is a stationary
-  point of $\tJa[\cdot;d]$. Then $(m_{\infty},\aw[m_{\infty};d])$ is a
+  \ref{eqn:defdatanonoise} with $w_* \in W_{\mu}$, and  $m$ is a stationary
+  point of $\tJa[\cdot;d]$. Then $(m,\aw[m;d])$ is a
   solution of the inverse problem \ref{eqn:probstat0} for any $\lambda
@@
   From the assumption $w_* \in W_{\mu}$ and Theorem
-  \ref{thm:rampreallygood}, $|(m_{\infty}-m_*)r|\le \mu$. From the
+  \ref{thm:rampreallygood}, $|(m-m_*)r|\le \mu$. From the
   identity \ref{eqn:solnonoise},
-  $\mbox{supp }\aw[m_{\infty};d] \subset
-  [(m_{\infty}-m_*)r-\mu,(m_{\infty}-m_*)r+\mu] \subset
+  $\mbox{supp }\aw[m;d] \subset
+  [(m-m_*)r-\mu,(m-m_*)r+\mu] \subset
   [-2\mu,2\mu]$. Because of the support limitation, $a(t)=|t|$ in the
@@
 \[
-  e[m_{\infty},\aw[m_{\infty},d];d] 
-= 8 \pi^2 r^2 \alpha^4\int^{\mu}_{-\mu}\,dt\,\frac{|t-(m_{\infty}-m_*)r|^4}{(1+(4\pi r)^2 \alpha^2 
-|t-(m_{\infty}-m^*)r|^2)^{2}}w_*(t)^2
-\]
-\[
-  \frac{1}{2} (4\pi r \alpha)^4\int^{\mu}_{-\mu}\,dt\,\frac{|t-(m_{\infty}-m_*)r|^4}{(1+(4\pi r)^2 \alpha^2 
-|t-(m_{\infty}-m^*)r|^2)^{2}}d(t+m_*r)^2
+  e[m,\aw[m,d];d] 
+= 8 \pi^2 r^2 \alpha^4\int^{\mu}_{-\mu}\,dt\,\frac{|t-(m-m_*)r|^4}{(1+(4\pi r)^2 \alpha^2 
+|t-(m-m_*)r|^2)^{2}}w_*(t)^2
+\]
+\[
+  \frac{1}{2} (4\pi r \alpha)^4\int^{\mu}_{-\mu}\,dt\,\frac{|t-(m-m_*)r|^4}{(1+(4\pi r)^2 \alpha^2 
+|t-(m-m_*)r|^2)^{2}}d(t+m_*r)^2
 \]
@@
 \[
-t \mapsto (4\pi r \alpha(t-(m_{\infty}-m_*)r))^4(1+(4\pi r)^2 \alpha^2 
-|t-(m_{\infty}-m^*)r|^2)^{-2}
+t \mapsto (4\pi r \alpha(t-(m-m_*)r))^4(1+(4\pi r)^2 \alpha^2 
+|t-(m-m_*)r|^2)^{-2}
 \]
@@
 on the complement of any open interval containing
-$t=(m_{\infty}-m_*)r$. Therefore
+$t=(m-m_*)r$. Therefore
 \begin{equation}
   \label{eqn:elimit}
-  \lim_{\alpha \rightarrow \infty}e[m_{\infty},\aw[m_{\infty};d];d] =
+  \lim_{\alpha \rightarrow \infty}e[m,\aw[m;d];d] =
   \frac{1}{2}\frac{1}{(4 \pi r)^2}\|w_*\|^2 = \frac{1}{2}\|d\|^2.
@@
 \[
-  e[m_{\infty},\aw[m_{\infty};d];d]  \le \frac{1}{2}\epsilon^2\|d\|^2
+  e[m,\aw[m;d];d]  \le \frac{1}{2}\epsilon^2\|d\|^2
   \Rightarrow \alpha \le \alpha_{\rm max}(\epsilon,\lambda,d).
@@
 The existence of this limiting penalty weight has been inferred
-indirectly; Appendix A describes a constructive algorithm
-for its approximation.
+indirectly. \cite{FuSymes2017discrepancy} describe a constructive
+algorithm for its approximation, which is used in the companion paper
+\cite{SymesChenMinkoff:21} to dynamically adjust $\alpha$ in the
+course of optimization of $\tJa$.
 
@@
 \ref{eqn:probstat0}. In this section, I explain how a solution may
-nontheless be constructed from a stationary point of $\tJa$.
+nonetheless be constructed from a stationary point of $\tJa$.
 
@@
   \label{thm:mnoisecor}
-  Asumme the hypotheses of Theorem \ref{thm:mnoise}, in particular
+  Asumme the hypotheses of Proposition \ref{thm:mnoise}, in particular
   that $m$ is a stationary point of $\tJa[\cdot;d]$, $d=d_*+n$. Then
@@
 
-\noindent {\bf Remark:} This bound 
-suggests that error in the estimate of $m$ due to data noise is a
-decreasing function of $\alpha$, at least for small $\alpha$. This result is intuitively
-appealing, and is supported by numerical evidence. A similar but stronger bound will be demonstrated in the next section.
+\noindent {\bf Remark:} This bound suggests that error in the estimate
+of $m$ due to data noise is a decreasing function of $\alpha$, at
+least for small $\alpha$. This result is intuitively appealing, and is
+supported by numerical evidence. It may be taken as a justification
+for the discrepancy-based algorithm for adjusting $\alpha$ during
+inversion \cite[]{FuSymes2017discrepancy}, used in the companion paper
+\cite[]{SymesChenMinkoff:21}.
 
@@
 estimated wavelet $\aw[m;d]$ (Theorem \ref{thm:norminv}) will lie in
-$\lW$, unless the support of the noise $n$ is not restricted. In order
+$\lW$, unless the support of the noise $n$ is restricted. In order
 to construct a solution of the inverse problem \ref{eqn:probstat0}, 
@@
   \times (1+(4\pi r \alpha)^2 
-  (t-(m-m^*)r)^2)^{-2}d_*(t)^2
+  (t-(m-m_*)r)^2)^{-2}d_*(t)^2
 \]
@@
 
-\section{Selection of $\alpha$}
-
-So far in this story, the selection of the penalty weight $\alpha$ has played a relatively minor role.
-In practical calculation, on the other hand, selection of $\alpha$ strongly influences the convergence of iterative methods and the quality of the results.
-
-Concentration of the wavelet near $t=0$ clearly favors larger $\alpha$: as will be established in Appendix A, the reduced penalty term $g[m,\aw[m;d];d]$ is a decreasing function of $\alpha$ for any $m$.
-As $g$ measures the dispersion of the wavelet away from $t=0$, larger $\alpha$ is to be preferred, all else being equal.
-
-However all else is not equal: the choice of $\alpha$ affects to VPM reduced objective $\tJa[m;d]$ and therefore the estimation of the nonlinear variable $m$ as well.
-To express this connection, observe that for any $d \in D$, $\lambda \in (0,\lambda_{\rm max}]$ and $m \in M$,
-the orthogonal projection of $d$ onto the range $F[m]W_{\lambda}$ is
-$({\bf 1}_{[mr-\lambda,mr+\lambda]}d)(t+mr)$, whence the distance from $d$ to the range of $F$ restricted to $W_{\lambda}$ is
-\[
-\mbox{inf}_{w \in W_{\lambda}} \|F[m]w-d\| = \|(1-{\bf 1}_{[mr-\lambda,mr+\lambda]})d\|.
-\]
-Denote by $r_{\lambda}[m;d]$ this distance normalized by $\|d\|$:
-\[
-r_{\lambda}[m;d] = \frac{ \|(1-{\bf 1}_{[mr-\lambda,mr+\lambda]})d\|}{\|d\|}.
-\]
-Denote by $v_{\lambda}[m;d]$ the pre-image under
-$F$ of the projection of $d$ onto $F[m]W_{\lambda}$:
-\[
-v_{\lambda}[m;d](t) = 4\pi r {\bf 1}_{[-\lambda,\lambda]}(t)d(t+mr).
-\]
-Then using notation introduced earlier in this section for the residual norm,
-\[
-r_{\lambda}[m;d] = r[m,v_{\lambda}[m;d];d]
-\]
-
-\begin{theorem}
-  \label{thm:malpha}
-  Suppose $d \in D$, $\alpha>0$, $0<\lambda \le \lambda_{\rm max}$, $m,m_* \in M$, and condition \ref{eqn:fullrec} holds. Also assume that
-  \begin{equation}
-      \label{eqn:rcomp}
-  \gamma = \frac{r[m,\aw[m;d];d]}{\sqrt{1-r_{\lambda}[m_*;d]^2}} < 1.
-  \end{equation}
-  Then
-  \begin{equation}
-      \label{eqn:mbnde}
-      |m-m_*| \le \lambda/r + \frac{1}{4\pi r^2 \alpha}\left(\frac{\gamma}{1-\gamma}\right)^{1/2}
-   \end{equation}
-\end{theorem}
-
-\begin{proof}
-From the definition, $0 \le r_{\lambda}[m;d] < 1 $ for all $m \in M$, so $\gamma$ is well-defined.
-
-Since condition \ref{eqn:fullrec} is assumed to hold, the conclusions of Proposition \ref{thm:epjgen} are available, in particular, equation \ref{eqn:residnormgen}:
-  \[
-  e[m,\aw[m;d];d] = \frac{1}{2}r[m,\aw[m;d];d]^2 = \frac{1}{2\|d\|^2} \int\, dt\, H(4\pi r\alpha|t-mr|)d(t)^2
-  \]
-  in which 
-  \[
-  H(x) = \left( \frac{x^2}{1+x^2}\right)^2.
-  \]
-  Note that $0 \le H \le 1$ and $H$ is strictly increasing on $\bR^+$.
-
-  Either $|m-m_*|r > \lambda$, or not. In the latter case, the conclusion \ref{eqn:mbnde} follows trivially. In the former case,
-  \[
-  r[m,\aw[m;d];d]^2 = \frac{1}{\|d\|^2} \int\, dt\, H(4\pi r\alpha|t-mr|)d(t)^2
-  \]
-  \[
-  \ge \frac{1}{\|d\|^2} \int\, dt\, H(4\pi r\alpha|t-mr|)({\bf 1}_{[m_*r-\lambda,m_*r+\lambda]}d(t))^2
-  \]
-  \[
-  = \frac{1}{\|d\|^2} \int\, dt\, H(4\pi r\alpha|t-(m-m_*)r|)({\bf 1}_{[-\lambda,\lambda]}d(t+m_*r))^2
-  \]
-  \[
-  \ge \frac{H(4\pi r\alpha(|m-m_*|r-\lambda))}{\|d\|^2} \int\, dt\, ({\bf 1}_{[-\lambda,\lambda]}d(t+m_*r))^2
-  \]
-  \[
-   = \frac{H(4\pi r\alpha(|m-m_*|r-\lambda))}{\|d\|^2} \int\, dt\, ({\bf 1}_{[m_*r-\lambda,m_*r+\lambda]}d(t))^2
-  \]
-  \[
-  = H(4\pi r\alpha(|m-m_*|r-\lambda))(1 - r_{\lambda}[m_*;d]^2)
-  \]
-  whence from the definition \ref{eqn:rcomp} of $\gamma$
-  \begin{equation}
-      \label{eqn:hineq}
-    H(4\pi r\alpha(|m-m_*|r-\lambda)) \le \gamma.
-  \end{equation}
-$H$ is bijective: $\bR^+ \rightarrow [0,1)$. Solving inequality \ref{eqn:hineq} for $4 \pi r \alpha(|m-m_*|r-\lambda)$ and rearranging, obtain the conclusion \ref{eqn:mbnde}. 
-\end{proof}
-
-\noindent {\bf Remark:} The parameters $\alpha$ and $\lambda$ play completely independent roles in this result.
-Given $d \in D$, $m_* \in M$, and $\lambda \in (,\lambda_{\rm max}]$, $r_{\lambda}[m_*;d]$ is the lower bound on relative RMS data error with slowness $m_*$, attained at $w = v_{\lambda}[m;d]$.
-Select $r_*>0$ so that $r_*^2 < 1 - r_{\lambda}[m_*;d]^2$, and suppose the for each $\alpha>0$, it is possible to find $m_{\alpha} \in M$ so that $r[m_{\alpha},w_{\alpha}[m_{\alpha};d];d] = r_*$.
-Then from \ref{eqn:mbnde},
-\[
-|m_{\alpha}-m_*| \le \lambda/r + O(1/\alpha).
-\]
-That is, this bound suggests that amongst $\alpha>0$ that permit attainment (via proper choice of m) of a prescribed value for the RMS data error at the minimizer of $J_{\alpha}[m,\cdot;d]$, then the bound \ref{eqn:mbnde} on the error between this $m$ and the ``target'' slowness $m_*$ is smaller if $\alpha$ is larger.
-This observation motivates the Discrepancy Algorithm described in Appendix A.
-
 \bibliographystyle{seg}
@@
 
-\append{Penalty weight selection via the Discrepancy Principle}
-This appendix describes an algorithm for controlling the penalty
-weight $\alpha$ based on a version the Discrepancy Principle
-\cite[]{EnglHankeNeubauer,Hanke:17,FuSymes2017discrepancy}. 
-The algorithm solves the problem
-\begin{quote}
-  \label{eqn:probmod}
-  given $d \in D$ and $0 < e_- < e_+$,  find  $m \in M, \alpha \in \bR^+$ so that
+% \appendix{Abstract Structure of the Gradient}
+\section*{APPENDIX: ABSTRACT STRUCTURE OF THE GRADIENT}
+
+The inverse problem studied here is far too simple to have any direct
+use in applications. Its simplicity allows a rather complete account of
+its properties, as the reader has seen in the preceding pages. However
+this discussion has touched on features found in more complex and
+prototypical problems:
 \begin{itemize}
-  \item[(i) ]$m$ is a stationary point of $\tJa[\cdot;d]$, and
-  \item[(ii) ]$e[m,\aw[m;d];d] \in (e_-,e_+)$.
+\item Nested optimization (variable projection method,
+  \cite{GolubPereyra:03}) based on a model decomposition into inner
+  and outer variables seems to be very important: in some cases, such
+  as the simple problem presented here, the decomposition is obvious
+  (inner variable = wavelet, outer variable = slowness), in others
+  less so \cite[]{geoprosp:2008,Terentyev:thesis}.
+\item The extended modeling operator must be surjective, or at least
+  have a dense range, {\em for each value of the outer variable}, so
+  that data may be fit well even for a poor initial guess of the outer
+  variable. for the extended inversion approach to be successful.
+\item Because of this data-fitting assumption, a straightforward
+  algorithm for scaling the quadratic penalty, based on the
+  Discrepancy Principle, is available - see
+  \cite{FuSymes2017discrepancy,Symes:21a,SymesChenMinkoff:21}. This
+  scaling varies dynamically during iterative optimization, in effect
+  changing the objective function sporadically as the iteration
+  converges.
+\item The derivative of the extended modeling operator with respect to
+  the outer variable is well-approximated by the composition of the
+  extended modeling operator itself and a pseudodifferential operator
+  of order 1 \cite[]{Symes:IPTA14,tenKroode:IPTA14}.
+\item The choice of the penalty operator, controlling the extended
+  degrees of freedom, is critical: in order to produce an objective
+  immune from cycle-skipping, this penalty operator must be
+  (pseudo-) differential of order zero \cite[]{StolkSymes:03}. A smooth
+  multiplier $a$, as in equation \ref{eqn:annmult}, is a special case.
 \end{itemize}
-\end{quote} 
-I describe an alternating, or coordinate search, algorithm for
-solution of this problem, combining a local optimization
-algorithm for updating $m$, and a second algorithm for updating
-$\alpha$.
-A first version of this algorithm appeared in \cite{FuSymes2017discrepancy}.Note that the mean square error $e$ lies in an open interval $(e_-,e_+)$ at a solution of this problem.
-Use of an interval, rather than a single target error level, accomplishes two objectives:
-\begin{itemize}
-    \item it is consistent with the general lack of precise knowledge of data error in most applications;
-    \item it permits a local optimization algorithm to make several small updates of $m$ before an update of $\alpha$ is required, as is required for good performance of algorithms such as BFGS.
-\end{itemize}
-
-Given an
-objective function $\Phi: M \rightarrow \bR$, a local optimization algorithm 
-generates a map $G[\Phi,...]: M \rightarrow M$, mapping a current estimate
-of $m$ to an updated estimate. The update rule may depend only on the
-current estimate, as is the case for steepest descent, the
-Gauss-Newton method, or Newton's method, or may also depend on
-information generated during earlier updates, as for secant-type
-methods such as BFGS. The notation is intended to allow for this
-latter possibility. Under ``standard conditions'' on $\Phi$ and the initial estimate for
-$m$, and equipped with so-called globalization safeguards to ensure
-satisfaction of sufficient decrease conditions
-(see \cite{NocedalWright}), iteration of $G$ produces a sequence in $M$
-converging to a local minimizer of $\Phi$.
-
-A normal stopping criterion for such an algorithm would be a tolerance
-test for the norm of $\nabla \Phi$ (and a limit on iteration count, of
-course). In application to $\Ja = e + \alpha^2 g$, add another
-stopping criterion:
-\begin{quote}
-  stop if $e \notin [e_-,e_+]$.
-\end{quote}
-Since $e$ is a summand in $\tJa$, $e$ will typically decrease along a
-minimizing sequence, so the expected condition invoking this stopping
-rule is $e<e_-$. Denote a update rule with this enhancement,
-applied to $\tJa$, by $G[\alpha,e_-,e_+,...]$, still allowing the
-possibility of information in addition to the current iterate.
-
-The algorithm also requires a rule for updating $\alpha$, defining a map $H: M \times E
-\rightarrow \bR^+$, in which $E=\{(e_-,e_+) \in \bR^2: 0 < 
-e_-<e_+\}$. For any $m \in M$ the rule is required to produce an $\alpha$ for which
-the error bounds are satisfied: If $(m, e_-, e_+) \in M \times E$ and $\alpha = H(m, e_-,
-e_+)$, then $e[m,\aw[m,d];d] \in [e_-,e_+]$. A usable rule is
-described below (equation \ref{eqn:alphasecant}).
-
-In outline, the algorithm is as follows:
-\begin{algorithm}[H]
-\caption{Scheme for updating $m, \alpha$}
-\begin{algorithmic}[1]
-  \State Choose $m\in M$
-  \Repeat
-  \State $\alpha \gets H[m,e_-,e_+]$
-  \Repeat
-  \State $m \gets G[\alpha,e_-,e_+,...](m)$
-  \Until{$e[m,\aw[m,d];d] \notin [e_-,e_+]$ or $\|\nabla \tJa[m;d]\|$
-    sufficiently small, or...}
-  \Until{$e[m,\aw[m,d];d] \in [e_-,e_+]$} 
-\end{algorithmic}
-\end{algorithm}
-Note that after step 3, the error bounds are satisfied, that is,
-$e[m,\aw[m;d];d] \in [e_-,e_+]$, but after step 4, that is likely not
-to be the case: the alteration of $m$ is likely to reduce $e$, as it is a
-summand in the definition of $\tJa$. If the $e$ is reduced below
-$e_-$, or if an approximate local minimizer is detected, then the condition in step 6 is satisfied, so control returns
-to step 7. If it is satisfied, the algorithm terminates, else
-control loops back to step 2, $\alpha$ is updated, and the inner $m$
-update loop is entered again. Termination requires that the bounds on
-$e$ are satisfied (step 7), hence that the $m$ update loop terminates
-by finding a local min which satisfies these bounds.
-
-The penalty parameter update strategies are based on the following fact about linear combinations of quadratic forms, similar to well-known results from the theory of Tihonov regularization \cite[]{Hanke:17}:
-
-\begin{theorem}
-  \label{thm:tich}
-  Suppose that $W, D$ are Hilbert spaces, $F \in \cal{B}(W,D)$, $A \in \cal{B}(W,W)$, $F^TF + \alpha^2 A^TA > 0$ for any $\alpha \ge 0$ (in particular, $F^TF>0$), and $A$ is injective.
-For $d \in D$, define
-  $w: D \times \bR^+ \rightarrow W$,
-  $e: D \times \bR^+ \rightarrow \bR^+$, and $g: D \times \bR^+ \rightarrow \bR^+$ by
-  \begin{eqnarray}
-    w[d,\alpha] &=& (F^TF + \alpha^2 A^TA)^{-1}F^Td \nonumber \\
-    e[d,\alpha] &=& \frac{1}{2\|d\|^2}\|Fw[d,\alpha]-d\|^2 \nonumber \\
-    g[d,\alpha] &=& \frac{1}{2\|d\|^2}\|Aw[d,\alpha]\|^2
-    \label{eqn:tichdefs}
-  \end{eqnarray}
-  Then for any $d\in D$, $w[d,\cdot]$, $e[d,\cdot]$, and $g[d,\cdot]$ are of class $ C^{\infty}(\bR^+)$, and 
-  \begin{itemize}
-      \item[0. ] EITHER $d$ is perpindicular to the range of $F$, $e \equiv \frac{1}{2}, g \equiv 0$ for all $\alpha \ge 0$, 
-      \item[1. ] OR $e$ is positive and strictly increasing, $g$ is positive and strictly decreasing for  $\alpha\ge 0$, and 
-      \begin{equation}
-        \label{eqn:lep}
-        \frac{de}{d\alpha^2}  \le 2g.
-      \end{equation}
-  \end{itemize}
-\end{theorem}
-
-\noindent {\bf Remark:} The conditions on $F$ and $A$ in the statement of this theorem might seem unduly restrictive.
-Both hold for the cartoon problem studied in the body of this paper.
-In fact, in most cases studied in the literature, the extended modeling operator represented here by $F$ is not coercive, and must be modified by regularization to meet the conditions of the theorem.
-Instances of the penalty operator $A$ studied in the literature are in fact usually injective, though this fact is commonly overlooked.
-See \cite{Symes:09} for some discussion.
-
-\begin{proof}
-Since the normal operator $F^T F + \alpha^2 A^T A$ is boundedly invertible and smooth in $\alpha \ge 0$, $w$ is smooth in $\alpha$, as are $e$ and $g$. 
-Differentiate the definition of $w$ 
-with respect to  $\alpha^2$ to obtain
-\begin{equation}
-(F^T F + \alpha^2 A^T A ) \frac{dw}{d\alpha^2} = -A^T A w
-\label{eqn:dnorm}
-\end{equation}
-whence
-\begin{align}
-\frac{de}{d\alpha^2} 
-&=\left\langle\frac{dw}{d\alpha^2},F^T(Fw-d) \right\rangle \nonumber \\
-&=-\alpha^2\left\langle\frac{dw}{d\alpha^2},A^TAw\right\rangle \nonumber \\ 
-&=\alpha^2 \langle A^TAw,(F^TF + \alpha^2 A^TA)^{-1}A^TAw\rangle \nonumber \\
-&\ge 0
-\label{eqn:de}
-\end{align}
-Note that the inequality in equation \ref{eqn:de} is {\em strict}  if $g[d,\alpha] > 0$ hence $A^TAw[d,\alpha] \ne 0$, since the normal operator is assumed to be positive definite.
-Also,
-\begin{align}
-\frac{dg}{d\alpha^2} &=  -\langle A^T Aw,(F^TF + \alpha^2 A^TA)^{-1}A^TAw \rangle \nonumber \\
-&\leq 0
-\label{eqn:dp}
-\end{align}
-similarly a strict inequality if $g \ne 0$.
-
-It follows that either $g[d,\alpha]>0$ for all $\alpha\ge 0$, or there exists $\alpha_0\ge 0$ for which $g[d,\alpha]>0$ for $0
-\le \alpha<\alpha_0$ and $g[d,\alpha]=0$ for $\alpha \ge \alpha_0$, hence $Aw[d,\alpha]=0$ for $\alpha \ge \alpha_0$.
-Since $A$ is assumed injective, $w[d,\alpha]=0$ for $\alpha > \alpha_0$.
-From the definition of $w$, it follows that $F^Td=0$, that is, $d$ is orthogonal to the range of $F$, and in fact $\alpha_0=0$, $g \equiv 0$ and $e \equiv \frac{1}{2}$ for all $\alpha\ge 0$.
-
-In the first case, that is, $F^Td \ne 0$, equations \ref{eqn:de} and \ref{eqn:dp} show that increasing $\alpha^2$ implies increasing $e$ while decreasing $g$, and
-\begin{align}
-&\langle A^TA w,(F^TF + \alpha^2 A^TA)^{-1} A^TA w \rangle \nonumber \\ 
-=& \lim_{\epsilon \rightarrow 0}\langle (A^TA+\epsilon^2I)^{1/2}w,[(A^TA+\epsilon^2I)^{-1/2}F^TF(A^TA+\epsilon^2I)^{-1/2} + \alpha^2 I]^{-1}(A^TA+\epsilon^2I)^{1/2}w \rangle \nonumber \\ 
-\le& \lim_{\epsilon \rightarrow 0}\frac{1}{\alpha^2} \langle (A^TA+\epsilon^2I) w, w\rangle = \frac{2}{\alpha^2}g.
-\end{align}
-which establishes inequality \ref{eqn:lep}.
-\end{proof}
-
-\begin{proposition}
-  \label{thm:alphainf}
-  Under the hypotheses of Theorem \ref{thm:tich}, 
-  \begin{eqnarray}
-  \label{eqn:walphainf}
-  \lim_{\alpha \rightarrow \infty} w(d,\alpha) & = & 0,\\
-  \label{eqn:ealphainf} 
-  \lim_{\alpha \rightarrow \infty} e(d,\alpha) & = & 1/2.
-  \end{eqnarray}
-\end{proposition}
-\begin{proof}
-  Since $F^TF$ is positive-definite, it has a positive-definite square root, and the definition \ref{eqn:tichdefs} of $w(d,\alpha)$ is equivalent to
-  \begin{equation}\label{eqn:ftfw}
-  (F^TF)^{1/2}w(d,\alpha) = (I+\alpha^2 (F^TF)^{-1/2}A^TA(F^TF)^{-1/2})^{-1}(F^TF)^{-1/2}F^Td.
-  \end{equation}
-  Denote by $E:\bR \rightarrow {\cal B}(W,W)$ the resolution of the identity for the self-adjoint bounded operator $(F^TF)^{-1/2}A^TA(F^TF)^{-1/2}$.
-According to the theorem on spectral representation of functions of self-adjoint operators (\cite{Yosida}, section XI.5, Theorem 1),
-  \begin{equation}\label{eqn:specw}
-  \|(F^TF)^{1/2}w(d,\alpha)\|^2 = \int_0^{\|F^TF\|}\frac{1}{1+\alpha^2 \lambda}d \|E(\lambda)(F^TF)^{-1/2}F^Td\|^2.
-  \end{equation}
-  Since $A^TA$ is assumed injective, so is $F^TF)^{-1/2}A^TA(F^TF)^{-1/2}$, whence $0$ is either a member of the resolvent set of the latter operator, or of its continuous spectrum.
-In either case, $\{0\}$ is a set of measure zero with respect to $\|E(\lambda)(F^TF)^{-1/2}F^Td\|^2$, so the integrand $(1+\alpha^2 \lambda)$ converges to zero $\|E(\lambda)(F^TF)^{-1/2}F^Td\|^2$-almost everywhere as $\alpha \rightarrow \infty$, and is bounded above by $1$ on the spectrum.
-The Dominated Convergence Theorem implies that $\|(F^TF)^{1/2}w(d,\alpha)\|^2 \rightarrow 0$ as $\alpha \rightarrow 0$, which establishes the first conclusion.
-The second follows from the definition of $e$ and the continuity of the operators involved in it.
-\end{proof}
-
-\begin{proposition}
-  \label{thm:psidef}
-  In addition to the hypotheses of Theorem \ref{thm:tich}, assume that $F^Td \ne 0$, and that $e[d,0]<e_+ \le 1/2$.
-Then there exists a unique $\alpha_+ > 0$ satisfying $e_+ = e[d,\alpha_+]$, and $e[d,\alpha]<e_+$ for $\alpha \in [0,\alpha_+)$.
-Define $\Psi: [0,\alpha_+] \rightarrow \bR$ by
- \begin{equation}
-\label{eqn:alphasecant}
-\Psi(\alpha)= \left(\alpha^2 + \frac{e_{+}-e[d,\alpha]}{2g[d,\alpha]} \right)^{1/2}
-\end{equation} 
-Then $\Psi([0,\alpha_+]) \subset (0,\alpha_+]$ and $\alpha<\Psi(\alpha)<\alpha_+$ for $\alpha \in [0,\alpha_+)$. 
-\end{proposition}
-
-\noindent {\bf Remark:} Note that $\Psi(0)$ is well-defined and $>0$.
-
-\begin{proof} Existence and uniqueness of $\alpha_+$ follows from Theorem \ref{thm:tich}, Proposition \ref{thm:alphainf}, and the assumption that $e[d,0]<e_+\le 1/2$.
-Since $e$ is strictly increasing under the assumption that $d$ is not perpindicular to the range of $F$, $e[d,\alpha]<e_+$ for $\alpha \in [0,\alpha_+)$.
-
-Suppose that $0 \le \alpha \le \psi \le \alpha_+$. From inequality \ref{eqn:lep}, 
-\[
-e[d,\psi]-e[d,\alpha] \le \int^{\psi^2}_{\alpha^2} d\tau^2 g[d,\tau] 
-\]
-\begin{equation}
-\label{eqn:basic}
-< 2 g[d,\alpha] (\psi^2-\alpha^2)  
-\end{equation}
-since $g$ is positive and strictly decreasing.
-From the definition \ref{eqn:alphasecant}, $\Psi(\alpha) \ge \alpha$ for $\alpha \in [0,\alpha_+]$, and $\Psi(\alpha)>\alpha$ if $\alpha \in [0,\alpha_+)$.
-Setting $\psi=\Psi(\alpha)$,
-inequality \ref{eqn:basic} and definition \ref{eqn:alphasecant} imply
-\[
-e[d,\Psi(\alpha)]-e[d,\alpha] < e_{+}-e[d,\alpha]
-\]
-so
-\begin{equation}
-\label{eqn:assert}
-\alpha \in [0,\alpha_+) \Rightarrow e[d,\alpha] < e[d,\Psi(\alpha)] < e_{+}.
-\end{equation}
-The last inequality ans the strict increase of $e$ imply that $\Psi(\alpha)<\alpha_+$ if $\alpha \in [0,\alpha_+)$.
-\end{proof}
-
-These results suggest an algorithm to determine $\alpha$: select an initial $\alpha_0 \ge 0$ for which $e[d,\alpha_0]<e_-$, then set $\alpha_n=\Psi(\alpha_{n-1}), n=1,2,...$, until $e[d,\alpha_n] > e_-$.
-Since $e_-<e_+$, the preceding result implies that $e[d,\alpha_n]<e_+$ also.
-
-That is, iteration of $\Psi$ produces an increasing sequence in $[0,\alpha_+]$.
-Two further ingredients are required to view it as an algorithm for production of an $\alpha$ satisfying the discrepancy principle:  a method for selection of an initial $\alpha_0$, and assurance that the error $e$ will eventually exceed the lower bound $e_-$.
-
-\begin{theorem}
-  \label{thm:discrepalg} In addition to the hypotheses of Proposition\ref{thm:psidef}, assume that $e[d,0] <e_-<e_+\le 1/2$. Define the sequence $\{\alpha_n: n \in {\bf Z}_+\}$ by
-  \begin{itemize}
-      \item $\alpha_0=0$, and
-      \item $\alpha_{n+1}=\Psi(\alpha_n), n \in {\bf Z}_+$.
-  \end{itemize}
-  Then there exists $N=N(d,e_-,e_+) \in {\bf Z}_+$ for which 
-  \begin{itemize}
-      \item $e(d,\alpha_n) \le e_-$ for $0 \le n <N$, and
-      \item $e[d,\alpha_N] \in (e_-,e_+)$
-  \end{itemize}
-\end{theorem}
-
-\begin{proof}
-  By hypothesis, the first conclusion is satisfied for $N=0$.
-From the definition \ref{eqn:alphasecant} of $\Psi$ and the conclusions of Proposition \ref{thm:psidef}, $\{\alpha_n\}$ is strictly increasing and $\alpha_n \le \alpha_+$ for all $n \ge 0$, so has a limit point $\alpha_{\infty} \in (0,\alpha_+]$.
-If $\alpha_{\infty}<\alpha_+$, $\epsilon = \Psi(\alpha_{\infty})-\alpha_{\infty} >0$.
-Since $\alpha \mapsto \Psi(\alpha)-\alpha$ is continuous, there is $\delta>0$ so that if $|\alpha-\alpha_{\infty}|<\delta$, then $\Psi(\alpha)-\alpha >\epsilon/2$.
-Since $\alpha_{\infty}=\lim_{n\rightarrow \infty}\alpha_n$, there is $m \in {\bf Z}_+$ for which $\alpha_m \in (\alpha_{\infty}-\min(\epsilon/2,\delta),\alpha_{\infty})$, so that $\alpha_{m+1} = \Psi(\alpha_m) >\alpha_m + \epsilon/2>\alpha_{\infty}$, contradicting the definition of $\alpha_{\infty}$ as the limit of the increasing sequence $\{\alpha_n\}$.
-Therefore, conclude that $\alpha_{\infty}=\alpha_+$.
-
-  According to Proposition \ref{thm:psidef}, $\alpha_n \rightarrow \alpha_+$ and $\alpha_n$ increases with $n$,  whence $e[d,\alpha_n] \rightarrow e_+$ and $e[d,\alpha_n]$ increases with $n$.
-Therefore there must exist a least $n = N(d,e_-,e_+)$ for which $e[d,\alpha_n] \in (e_-,e_+)$.
-\end{proof}
-
-\noindent {\bf Remark:} The condition $e(d,0)\le e_-$ means that $\sqrt{2e_-}$ is an upper bound for the distance from $d$ to the range of $F$.
-Recall that $e_-$ represents an underestimate for data noise level.
-That is, it is assumed that the data is within data noise level of the range of $F$.
-Such an assumption is unlikely to be tenable for the physical (non-extended) modeling operator, unless the nonlinear model parameters ($m$ in this discussion) are chosen near-optimally.
-The ability to fit data, even when the parameters represented by $m$ are greatly in error, is an essential characteristic of successful extended inversion strategies.
-
-\noindent {\bf Remark:} The update rule \ref{eqn:alphasecant} is convergent to a satisfactory value, as just shown, but in practice is rather slow.
-Accelerated updates using variants of the secant rule are possible, but an analysis of those possibilities is beyond the scope of this paper.
-
-A satisfactory penalty parameter selection rule can now be defined: in the notation introduced at the beginning of this appendix, define 
-\begin{equation}
-    H[m,e_-,e_+] = \alpha_{N(d,e_-,e_+)}.
-\end{equation}
-Choosing $F=F[m]$ in Theorem \ref{thm:discrepalg}, conclude that $H$, so defined, has the required properties.
-
-This is as far as one can go without more concrete assumptions on $F$
-and $A$. If the reduced
-objective $\tJa$ possesses a large (in some sense) domain of convexity for a range of $\alpha$, then the discrepancy algorithm sketched
-here can efficiently drive $\alpha$ to near its largest feasible
-value, thus giving the best resolution near the global minimizer. That
-program has been carried out for the simple cartoon problem of this
-paper, and partly for some other more protypical inverse wave problems \cite[]{FuSymes2017discrep}.
-
-\append{Heisenberg, support, and wavelength}
-Without further contraints on data or solution, nothing more can be
-said about bounds on $\alpha$. If the data $d$ (and the target wavelet
-$w_*$) are assumed to have a square integrable derivative, then a
-necessary condition follows from the Heisenberg
-inequality (see for example \cite{Folland:07}, p. 255). To formulate
-this result in its most general form, introduce the Hilbert subspaces
-$V^0 \subset L^2(\bR), V^1 \subset H^1(\bR)$:
-\begin{eqnarray}
-  V^0 & = & \{ f \in L^2(\bR): Af \in L^2(\bR)\}, \nonumber\\
-  V^1 & = & V^0 \cap H^1(\bR).
-            \label{eqn:vdef}
-\end{eqnarray}
-$V^0$ is the domain of $A$, and equipped with the graph norm of $A$. A
-natural norm in $V^1$ is
-\[
-  \|f\|^2_{V^1} = \|f\|_{V^0}^2 + \|f\|_{H^1}^2.
-\]
-$V^j$ is the completion of $C_0^{\infty}(\bR)$ in the corresponding
-norm, j=0,1.
-
-\begin{proposition}
-  \label{thm:heis}
-For $w \in V^1$,
-  \begin{equation}
-    \label{eqn:heis}
-    \|Aw\|\|w'\| \ge \frac{1}{2}\|w\|^2
-  \end{equation}
-\end{proposition}
-
-\begin{proof}
-  For $w \in C_0^{\infty}(\bR)$,
-  \[
-    \int w^2 = \left|\int\,dt\, t (w(t)^2)' \right|= \left|2\int\,dt\,tw(t)w'(t)\right| \le
-    2\|Aw\| \|w'\|
-  \]
-  by the Cauchy-Schwarz inequality. Since $C_0^{\infty}(\bR)$ is dense
-  in $V^1$, the conclusion follows by continuity.
-\end{proof}
-
-In the conventional formulation of the Heisenberg inequality, the $L^2$ norm of
-$w'$ is replaced by the its equivalent in terms of the Fourier
-transform $\hat{w}$. Adopting temporarily the use of dummy variables
-in the expression of functions, the identity \ref{eqn:heis} turns into
-the usual form of the Heisenberg inequality: for $w \in V^1$,
-\begin{equation}
-\label{eqn:fheis}
-\|tw(t)\|\|k\hat{w}(k)\| \ge \frac{1}{4\pi}\|w\|^2.
-\end{equation}
-
-Define $\krms[w]$, the root mean square estimator of frequency of $w
-\in V^1$, by
-\begin{equation}
-  \label{eqn:krms}
-  \krms[w]=\frac{1}{2\pi}\frac{\|w'\|}{\|w\|} = \left(\int
-    \,dk\,\frac{|\hat{w}(k)|^2}{\|\hat{w}\|^2} k^2\right)^{1/2}.
-\end{equation}
-Then the inequalities \ref{eqn:heis}, \ref{eqn:fheis} can be rewritten as
-\begin{equation}
-  \label{eqn:frmsheis}
-  \|Aw\| \ge \frac{\|w\|}{4\pi \krms[w]}.
-\end{equation}
-
-For $\lambda >0$, define
-\begin{equation}
-  \label{eqn:wlam1}
-  \lW^1 = \lW \cap H^1(\bR).
-\end{equation}
-Note that $\lW^1 \subset V^1$ is a closed subspace for any
-$\lambda>0$.
-
-\begin{proposition}
-  \label{thm:klam}
-  For $w \in \lW^1$,
-  \[
-    \krms[w] \ge \frac{1}{4 \pi \lambda}
-  \]
-\end{proposition}
-
-\begin{proof}
-  Follows directly from inequality \ref{eqn:frmsheis} and the obvious bound
-  $\|A|_{\lW}\|\le \lambda$.
-\end{proof}
-
-\noindent{\bf Remark:} This result is the link mentioned earlier between the support
-constraint and the well-known frequency-based criteria for success of
-FWI. The result of Theorem \ref{thm:fwi} can be rephrased as showing the
-existence of many stationary points of the mean-square error function
-for which the travel time is in error by more than
-$\lambda \ge 1/{4 \pi \krms[w]|}$. In fact, the usual error criterion mentioned in the
-literature is that the initial estimate of travel time must be in error by at most ``half a
-wavelength'' if FWI is to converge reliably. This is correct in some
-circumstances, depending on features of the target wavelet $w_*$ of
-which the arguments in this paper do not take account.
-
-\begin{proposition}
-  \label{thm:heis2}
-  For $\lambda>0, w \in \lW^1$,
-  \begin{equation}
-    \label{eqn:heis2}
-    \|A^2w\| \ge \frac{\|w\|}{24 \lambda (2\pi\krms[w])^3}.
-  \end{equation}
-\end{proposition}
-
-\begin{proof}
-  Since $A$  preserves $\lW^1$, inequality \ref{eqn:heis} implies
-  \begin{equation}
-    \label{eqn:heissq}
-    \|A^2w\| \ge \frac{\|Aw\|^2}{2\|(Aw)'\|}.
-  \end{equation}
-  From the definition of $A$, $(Aw)'=w +Aw'$, so for $w \in \lW^1$,
-  \[
-    \|(Aw)'\| \le \|w\|+ \lambda\|w'\|.
-  \]
-  Applying \ref{eqn:heis} again,
-  \[
-    \|w\|^2 \le 2\|w'\|\|Aw\|\le 2\lambda \|w'\|\|w\|,
-  \]
-  so obtain the 1D Poincar\'{e} inequality: for $w \in \lW^1$,
-  \[
-    \|w\| \le 2\lambda\|w'\|.
-  \]
-  Thus
-  \[
-    \|(Aw)'\| \le 3\lambda\|w'\|.
-  \]
-  Apply this estimate together with the basic Heisenberg estimate
-  \ref{eqn:heis} to the inequality \ref{eqn:heissq} to obtain
-  \[
-    \|A^2w\| \ge \frac{\|w\|^4}{24\lambda \|w'\|^3}
-  \]
-  Rearranging and using the definition \ref{eqn:krms} of $\krms[w]$,
-  arrive at inequality \ref{eqn:heis2}.
-\end{proof}
-
-\begin{theorem}
-  \label{thm:ipnonoisenec}
-  Suppose that $a$ is given by definition \ref{eqn:ann}, $\alpha$,
-  $\mu \in (0,\lambda_{\rm max}/2]$, $d$ is given by
-  \ref{eqn:defdatanonoise} with $w_* \in W_{\mu}^1$, and that a
-  stationary point $m_{\infty}$ of $\tJa[\cdot;d]$ yields a solution
-  of the inverse problem \ref{eqn:probstat0} for $\lambda \ge 2\mu$,
-  $\epsilon>0$. Then
-\begin{equation}
-  \label{eqn:epsalpha}
-  3 (4\pi \lambda \krms[w_*])^3 \epsilon \ge \frac{(4\pi  r\alpha\lambda)^2}
-  {(1+(4\pi r \alpha \lambda)^2)}
-\end{equation}
-\end{theorem}
-
-\begin{proof}
-  Rearranging the identity \ref{eqn:residnorm}, and observing as in
-  the proof of Theorem \ref{thm:ipnonoisesuf} that the support of the
-  integrand is contained in $[-2\mu, 2\mu] \subset [\lambda,\lambda]$,
-  \[
-  e[m_{\infty},\aw[m_{\infty},d];d] 
-= 8 \pi^2 r^2 \alpha^4\int\,dt\,t^4(1+(4\pi r)^2 \alpha^2 t^2)^{-2}w_*(t+(m_{\infty}-m_*))^2
-\]
-\[
-  \ge \frac{8\pi^2r^2\alpha^4}{(1+(4\pi r)^2 \alpha^2
-    \lambda^2)^{2}}\int\,dt\,t^4w_*(t+(m_{\infty}-m_*))^2
-\]
-\begin{equation}
-  \label{eqn:residnormbis}
-  \ge \frac{8\pi^2r^2\alpha^4}{(1+(4\pi r)^2 \alpha^2
-    \lambda^2)^{2}}\|A^2w_*(\cdot + (m_{\infty}-m_*)r)\|^2.
- \end{equation}
-Since $\krms[w]$ is invariant under translation of $w \in V^1$,
-Proposition \ref{thm:heis2} implies that this is
-\[
-  \ge \frac{8\pi^2r^2\alpha^4}{(1+(4\pi r)^2 \alpha^2
-    \lambda^2)^{2}}\frac{\|w_*\|^2}{(24)^2 \lambda^2 (2\pi\krms[w_*])^6}
-\]
-\[
-  = \frac{1}{2}\frac{(4\pi r\alpha\lambda)^4}{(1+(4\pi r \alpha
-    \lambda)^2)^{2}}\frac{\|d\|^2}{(24)^2 (2\pi
-    \lambda \krms[w_*])^6}
-\]
-The pair $(m_{\infty}, \aw[m_{\infty},d])$ is presumed to solve the inverse
-problem as stated in \ref{eqn:probstat0}, in particular
-\[
-  \epsilon \ge (2 e[m_{\infty}, \aw[m_{\infty},d],d])^{1/2}/\|d\|
-\]
-the inequality \ref{eqn:epsalpha} follows.
-
-\end{proof}
-
-Inequality \ref{eqn:epsalpha} couples the dimensionless quantites
-$\epsilon$,  $4\pi r \lambda \alpha$, and $4 \pi \lambda
-\krms[w_*]$. Proposition \ref{thm:klam}, implies that the left
-hand side is $\ge 3\epsilon$. Since the right hand side is $\le 1$,
-the inequality implies no limitation on $\alpha$ if the left hand side
-is $\ge 1$. For small $\epsilon$, the the largest permissible $\alpha$
-is $O(\sqrt{\epsilon})$. The permissible range of $\alpha$ increases with
-nondimensionalized RMS frequency $4\pi \lambda \krms[w_*]$. Since
-$\krms$ is invariant under translation and scaling of its argument,
-$\krms[w_*]=\krms[d]$, that is, the nondimensionalized RMS frequency
-is an observable property of the data, in the noise-free case at
-least.
-
-\append{Abstract Structure of the Gradient}
+
 The expression \ref{eqn:dexpjgen} for the derivative of the reduced
@@
 
-For the remainder of this appendix, assume that
+For the model problem studied in the body of this paper, a simple
+argument justifies this conclusion.  To sideline some technical
+details, assume that
 $a \in C^{\infty}(\bR)\cap L^{\infty}(\bR)$ and $a(t)=t$ for $|t|<\tau$ and $|a(t)| \ge
 \tau$ for $|t| \ge \tau$, with $\tau$
-defined in equation \ref{eqn:taudef}.
-
+defined in equation \ref{eqn:taudef}. Then
 Proposition \ref{thm:norminvexp}, item [1], implies that
@@
 
-In the other extended inversion settings mentioned at the beginning of
-this appendix, $F$ is a microlocally elliptic Fourier Integral
-Operator (FIO) \cite[]{Dui:95}. The canonical relation takes on the role
-of the shift operator $t \rightarrow t-mr$, and has the properties
-necessary to conclude that the composition of $F$ and $F^T$ in both
-ordersis pseudodifferential, at least in an open conic subset of the
-cotangent bundle. If $A^TA$ is pseudodifferential, then so is the
-normal operator. The inverse in the RHS of equation \ref{eqn:faw} must
-be replaced by a microlocal parametrix, and a smooth error added to
-the RHS. Thanks to Egorov's Theorem \cite[]{Tay:81}, a special case of
-the rules for composing FIOs, the operator on the RHS of equation
-\ref{eqn:faw} is a pseudodifferential operator whose symbol is
-algebraic in geometric optics quantities, hence depends smoothly on
-the coefficients in the wave equation, as does $F\aw$.
-
 Since $\tJa$ and its gradient depend smoothly on their arguments, it
@@
 \[
-  \tJa[m;d] = \frac{1}{2}(\|F[m]w[m;d]-d\|^2 + \alpha^2 \|Aw[m;d]\|.
+  \tJa[m;d] = \frac{1}{2}(\|F[m]w[m;d]-d\|^2 + \alpha^2 \|Aw[m;d]\|),
 \]
@@
 \[
-  d\tJa[m;d]\delta m = \langle D(F[m]\aw[m;d])\delta m, F[m]\aw[m;d]-d\rangle
+  D\tJa[m;d]\delta m = \langle D(F[m]\aw[m;d])\delta m, F[m]\aw[m;d]-d\rangle
 \]
@@
 
-For the more general contexts mentioned earlier, a similar derivation
-is possible. The analog of $Q$ is a pseudodifferential operator of
-order 1, which is essentially (to leading order) skew-symmetric, and
-the more general version of the identity \ref{eqn:djq} is approximate,
-with lower-order (smoother) error terms.
-
-These ideas were used in \cite{tenKroode:IPTA14,Symes:IPTA14} to show
-that the analog of $\tJa$ is tangent to second order to a convex
-quadratic form, related to the Hessian of an objective formulated in
-terms of traveltime. This relation explains the ability of the variant
-of extended inversion studied in the cited references to recover
+Similar reasoning can be applied to some of the other extended
+inversion settings mentioned at the beginning of this
+appendix. Provided that the outer parameter $m$ consists of (or
+parametrizes) smooth coefficients in the principal part of the
+governing system of wave equations, $F$ is a microlocally elliptic
+Fourier Integral Operator (FIO) \cite[]{Dui:95}. The canonical
+relation of $F$ takes on the role of the shift operator
+$t \rightarrow t-mr$, and has the properties necessary to conclude
+that the composition of $F$ and $F^T$ in both orders is
+pseudodifferential, at least in an open conic subset of the cotangent
+bundle. If $A^TA$ is pseudodifferential, then so is the normal
+operator, and it is microlocally elliptic, so positive-definite when
+restricted to a suitable subspace of the inner parameter $w$, which
+appears in the hyperbolic system as components of a right-hand side or
+boundary condition. The inverse in the RHS of equation \ref{eqn:faw}
+must be replaced by a microlocal parametrix. Thanks to Egorov's
+Theorem \cite[]{Tay:81}, a special case of the rules for composing
+FIOs, the operator on the RHS of equation \ref{eqn:faw} is a
+pseudodifferential operator whose symbol is algebraic in geometric
+optics quantities, hence depends smoothly on the coefficients in the
+system of wave equations, and therefore on $m$, as does $F\aw$.
+
+Microlocal ellipticity of $F$ implies an analogue of relation
+\ref{eqn:deriv}, in which $Q$ is an essentially skew-symmetric
+pseudodifferential operator of order 1. Since the derivation 
+the gradient identity \ref{eqn:djq} is algebraic, it holds
+approximately in these more general cases, with lower-order (smoother)
+error terms. See 
+\cite{Symes:Madrid,tenKroode:IPTA14,Symes:IPTA14} for details in
+several specific cases, and for a similar calculation of the Hessian.
+For these examples, the principal symbol of $Q$ can be computed
+explicitly, as a function of the geodesic distance (traveltime)
+function. Consequently, the analog of $\tJa$ is tangent to second
+order to the Hessian of an objective formulated in terms of
+traveltime. This relation explains the ability of the variant of
+extended inversion studied in the cited references to recover
 kinematically accurate velocity fields. Of course, that is what is

### 2022-03-28 repro/papers/GEM22/newstuff.tex
first commit

--- 2021-06-17 repro/papers/sseprecond/paper.tex
@@
 
-%The energy source $\bff$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time.
-In fact, additional source degrees of freedom, beyond those needed to describe physically realized sources, may be useful in rendering the FWI problem \ref{eqn:fwi} more amenable to numerical solution, via so-called extended modeling (see \cite{geoprosp:2008}, \cite{LeeuwenHerrmannWRI:13}, \cite{HuangNammourSymesDollizal:SEG19}, and many references cited there).
-Therefore it is natural to view $\bff$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
-A simple nonlinear least squares formulation is:
+A simple nonlinear least squares formulation, including estimation of the source, is:
 \begin{equation}
 \label{eqn:ols}
-\mbox{choose } {\bf c} \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2.
+\mbox{choose } {\bf c}, \bff \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2.
 \end{equation}
@@
 and those have a tendency to stall due to ``cycle-skipping''. See for
-exampe \cite{VirieuxOperto:09} and many references cited there. Source
+example \cite{VirieuxOperto:09} and many references cited there. Source
 extension is one approach to avoiding this problem. It consists in
@@
 \cite{HuangNammourSymesDollizal:SEG19} present an overview of the
-literature on source extension methods, describing a variety of
+literature on source extension, describing a variety of
 methods to add degrees of freedom to physical source model. The present paper
@@
 example \cite{Symes:SEG20}. It replaces the nonlinear
-least squares problem \ref{eqn:esi} with a {\em reduced} problem, to
-be solved iteratively. Each iteration involves solution of the
-subproblem \ref{eqn:esis}. Therefore efficient solution of the
+least squares problem \ref{eqn:esi} with a {\em reduced} problem, namely the function of ${\bf c}$ defined by the minimum of the subproblem objective \ref{eqn:esis}.
+Evaluation of this reduced objective requires solution of the
+subproblem \ref{eqn:esis}. Therefore efficient solution of this
 subproblem is essential to efficient solution of the nonlinear problem
@@
 \ref{eqn:esis}. This approximate solution operator may be used to
-accelerate Krylov space methods for the solution of the source
+accelerate (``precondition'') Krylov space methods for the solution of the source
 subproblem \ref{eqn:esis}. Numerical examples suggest the
 effectiveness of this acceleration.
+
+Beyond accelerated Krylov iteration for solution of problem \ref{eqn:esis}, the structure of the preconditioner is key to understanding the relation of the extended problem \ref{eqn:esi} to traveltime tomography, which in turn explains its ability to avoid cycle-skipping \cite[]{Symes:22b}.
+The same structure enables efficient and accurate evaluation of the reduced objective gradient, another essential task in the computation solution of the extended inverse problem \cite[]{Symes:22c}.
 
@@
 
-The obvious computation of the pressure-to-source map
-- prescribe the pressure, solve the wave equation with this boundary
-condition, read off the equivalent source - suffers from intrinsic
-numerical inaccuracy. I suggest an alternative computationally
-feasible approach.
+%The obvious computation of the pressure-to-source map
+%- prescribe the pressure, solve the wave equation with this boundary
+%condition, read off the equivalent source - suffers from intrinsic
+%numerical inaccuracy. I suggest an alternative computationally
+%feasible approach.
 
@@
 coefficient vector $\bf{c}=(\kappa,\rho)$ is defined throughout space
-$\bR^n$, the state vector $\bu$ throughout space-time
-$\bR^{n+1}$. Whenever convenient for mathematical manipulations,
+$\bR^n$, and are assumed to be smooth, in fact $\log \kappa, \log \rho \in C^{\infty}(\bR^d)$ and uniformly bounded.
+The same is true of the wave velocity $c = \sqrt{\kappa/\rho}$.
+The state vector $\bu$ is a vector-valued distribution on space-time
+$\bR^{n+1}$: $\bu = (p,\bv)$, in which the pressure field $p$ is scalar and the (particle) velocity field $\bv$ is an $n$-vector field. Whenever convenient for mathematical manipulations,
 $n=3$: for instance, I will write $\bx=(x,y,z)^T$ for the spatial
-coordinate vector, and refer to the third (vertical) coordinate of
+coordinate vector, or $(\bx,t) = (x,y,z,t)$ for the space-time coordinate vector, and refer to the third (vertical) coordinate of
 particle velocity as $v_z$. Examples
@@
 
+\noindent {\bf Remark:} I will use the shorthand
+\[
+  \bu  = 0 \mbox{ for } t \ll 0 
+\]
+to mean that $\bu$ is {\em causal}, that is,
+\[
+  \mbox{For some } T \in \bR, \bu(\cdot,t) = 0 \mbox{ for all } t <
+  T.
+\]
+Similarly,
+\[
+  \bu = 0 \mbox{ for } t \gg 0 
+\]
+signifies that $\bu$ is anti-causal.
+
+It has been understood for decades that initial/boundary value problems for the system $L\bu=\bff$ have well-behaved solutions.
+In particular, \cite{BlazekStolkSymes:13} state results which imply the following
+
+\begin{theorem}
+  \label{thm:basicp}
+  Denote by $\Omega \subset \bR^n$ a bounded domain with smoothly embedded oriented boundary $\partial \Omega$, and $V=H^1_0(\Omega) \times H^1_{\rm div}(\Omega)$. 
+  \begin{itemize}
+  \item[i.
+] Suppose that $\bff = (f,{\bf g}) \in (L^2_{\rm comp}(\Omega \times {\bf R}))^{n+1}$.
+Then the system $L\bu=\bff$ has a unique causal weak solution $\bu = (p,\bv) \in (L_{\rm loc}^2(\Omega \times \bR))^{n+1}$: for any ${\bf \phi} = (\phi, {\bf \psi}) \in (C_0^{\infty}(V))^{n+1}$,
+    \begin{equation}
+      \label{eqn:weak}
+      \int \left(\frac{1}{\kappa} p \frac{\partial \phi}{\partial t} + \rho \bv \cdot \frac{\partial{\bf \psi}}{\partial t} + \bv \cdot \nabla \phi + p \nabla \cdot {\bf \psi} \right) = \int (f\phi + {\bf g} \cdot {\bf \psi}).
+    \end{equation}
+  \item[ii. ] If in addition $1 \le k \in {\bf N}$ and
+    $\partial^k_t\bff \in (L^2(\Omega \times \bR))^{n+1}$, then $\bu
+    \in C^k(\bR, (L^2(\Omega))^{n+1}) \cap C^{k-1}(\bR,V)$: that is,
+    $\bu$ is a strong solution, satisfying $L\bu=\bff$ almost everywhere..
+  \item[iii. ] 
+    Suppose that $R>0$ and $\mbox{supp }\bff \subset [T_0, T_1] \times B_R(\bx_0)$.
+Then $\bu(\cdot,t)=0$ for $t<T_0$, and for each $t \in [T_0,\infty)$, $\mbox{supp }\bu \subset B_R(\bx_0) + B_{c_{\rm max}(t-T_0)}({\bf 0})$.
+  \end{itemize}
+\end{theorem}
+
+\begin{cor}
+  \label{thm:weakcor}
+  Let $\alpha_{\epsilon}$ be a Dirac family in $C_0^{\infty}(\bR)$,
+  and $\bu$ be a causal weak solution of $L\bu=\bff \in (L^2_{\rm
+    comp}(\Omega \times \bR))^{n+1}$. Then for all $\epsilon > 0$,
+  $\alpha_{\epsilon} *_t \bu$ is a strong solution with well-defined and vanishing restriction
+  to $\partial \Omega \times \bR$.
+\end{cor}
+
+Denote by ${\bf \nu}(\bx)$ the outward unit normal (co)vector at $ \bx
+\in \partial \Omega$.
+
+\begin{theorem}
+  \label{thm:basicv}
+  Let  $H^1_{{\rm div},0}(\Omega)$ denote the subspace of $H^1_{\rm
+    div}(\Omega)$ consisting of vector fields $\bv$ satisfying $\bv
+  \cdot {\bf \nu} = 0$ on $\partial \Omega$. Define
+  $W = H^1(\Omega) \times H^1_{{\rm div},0}(\Omega)$.
+Then the same conclusions as listed in Theorem \ref{thm:basicp} hold,
+with $V$ replaced by $W$.
+\end{theorem}
+\begin{proof}
+  This result follows directly from Theorem 1 in
+  \cite[]{BlazekStolkSymes:13}, once it is shown that the spatial part
+  of $L$ is skew-adjoint on $W$. The analogous statement for
+  $V$ is proven in \cite[]{BlazekStolkSymes:13}, Appendix A, and a
+  similar proof establishes the same property of $W$.
+\end{proof}
+
+\begin{remark} The cited reference \cite[]{BlazekStolkSymes:13} requires only that the coefficients ${\bf c}$ be bounded and measureable.
+The arguments developed here, however, depend on approximation of solutions by means of geometric optics, hence the additional requirement of coefficient smoothness.
+\end{remark}
+
 The surface source extension replaces point sources on or near a
-surface in $\bR^3$ with source functions confined to the same
+surface in $\bR^3$ with distributions supported on the same
 surface. The simplest example of this extended geometry specifies a
@@
 \rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} +
-                                                f_s{\bf e} \delta(z-z_s),\nonumber \\
+                                                f_s{\bf e}_z \delta(z-z_s),\nonumber \\
 p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
@@
 
-\noindent {\bf Remark:} In system \ref{eqn:awepm} and many similar
-systems to follow, I will use the shorthand
+The singular right-hand sides appearing in the systems \ref{eqn:awepm} do not conform to the conditions of the basic well-posedness statement \ref{thm:basicp}, so the first task is to formulate conditions under which well-behaved solutions exist.
+This is most conveniently done by first developing conditions under which an inhomogeneous boundary value problem has well-behaved solutions.
+The pressure field on the surface $\partial \Omega \times \bR$ uniquely determines the finite energy ($L^2$) causal acoustic field in the interior, as follows from simple energy estimates, but cannot be specified arbitrarily.
+A finite energy solution exists if the pressure field on the boundary is {\em incoming} (corresponding to {\em downgoing} in seismic parlance), in the sense that its Fourier components decrease rapidly with increasing frequency except at initial conditions for rays entering $\Omega$ at a positive angle, with distance to the boundary increasing with increasing time.
+A precise statement depends on a parameter $\delta >0$, essentially a bound on the direction cosine relative to the tangent plane of the boundary.
+
+\begin{definition}
+  \label{thm:definc}
+  Given $\delta >0$, $d \in L^2_{\rm comp}(\partial \Omega \times \bR)$ is $\delta-${\em incoming} if there exists a conic subset $\Gamma \subset T^*(\partial \Omega \times \bR)$ and $H \in OPS^0(\partial \Omega \times \bR)$ satsifying the conditions:
+  \begin{itemize}
+  \item[1. ] $H$ is elliptic in $\Gamma$;
+  \item[2. ] for every $(\bx,t,\xi,\omega) \in \Gamma$, there is a
+    unique $z \in \bR$ and $\psi(\bx,{\bf \xi},\omega) = \frac{{\bf \xi}}{\omega} + z {\bf \nu}(\bx)  \in
+    \bR^n$ so that $(\bx,t,\psi(\bx,{\bf \xi},\omega),\omega)$ is
+    characteristic for $L$, $|z| > \delta$, and $z\omega < 0$; and
+  \item[3. ] $(I-H)d \in H^2(\partial \Omega \times \bR)$.
+  \end{itemize}
+\end{definition}
+
+For $\epsilon > 0$, denote by $\Omega_{\epsilon}$ the collar neigborhood of $\partial \Omega$ defined by
+\begin{equation}
+  \label{eqn:collar}
+  \Omega_{\epsilon} = \{\bx + z {\bf \nu}(\bx): 0 \ge z \ge  -\epsilon\}.
+\end{equation}
+For $\epsilon$ sufficiently small, $\Omega_{\epsilon}$ is a smoothly bounded submanifold of $\Omega$.
+
+\begin{theorem}
+  \label{thm:dirprob}
+  Suppose that $\delta>0$ and $d \in L^2_{\rm comp}(\partial \Omega
+  \times \bR)$ is $\delta-${\em incoming}. Then there exists a unique
+  causal weak solution $\bu = (p,\bv) \in (L^2_{\rm loc}(\Omega \times
+  \bR)^{n+1}$ of $L\bu=0$ in $\Omega$, for which $p$ assumes the boundary value $d$, in the sense that for every $\phi \in C_0^{\infty}(\bR^{n+1})$,
+  \begin{equation}
+    \label{eqn:weakbdry}
+    \lim_{\epsilon \rightarrow 0}(\alpha_{\epsilon} *_t p)|_{\partial
+      \Omega \times \bR}  = d.
+  \end{equation}
+\end{theorem}
+
+\begin{proof}
+  Choosing $\epsilon$ sufficiently small, construct an extension $d_1\in H^2(\Omega_{\epsilon}\times \bR)$ of $(I-H)d$, and cut it off smoothly to produce an extension, also denoted $d_1$, to all of $\Omega \times \bR$.
+Define $(p_1,\bv_1)$ to be the weak solution of
+\begin{eqnarray}
+\label{eqn:awepm1}
+\frac{1}{\kappa}\frac{\partial p_1}{\partial t} & = & - \nabla \cdot \bv_1 -
+\frac{1}{\kappa}\frac{\partial d_1}{\partial t}, \nonumber \\
+\rho\frac{\partial \bv_1}{\partial t} & = & - \nabla p_1 \\
+p_1& =& 0 \mbox{ for } t \ll 0,\nonumber\\ 
+\bv_1 & = & 0 \mbox{ for } t \ll 0.
+\end{eqnarray}
+From Theorem \ref{thm:basicp} with $k=1$, conclude that $(p_1,\bv_1)
+\in H^1_0(\Omega \times \bR) \times H^1_{\rm div}(\Omega \times \bR)$
+is a strong solution and in particular that the trace of
+$p_1|_{\partial \Omega \times \bR} \in L^2_{\rm loc}(\partial \Omega \times \bR)$ is
+well-defined, and in fact
+\begin{equation}
+  \label{eqn:smparttrace}
+  p_1|_{\partial \Omega \times \bR} = (I-H)d.
+\end{equation}
+
+Let $\{({\cal U}_i, \chi_i): i=1,...,N\}$ be a coordinate atlas for
+the oriented submanifold $\partial \Omega \subset \bR^n$. That is,
+$\{{\cal U}_i: i=1,...N\}$ is a locally finite open cover of $\partial \Omega$, 
+$\chi_i:{\cal U}_i \rightarrow \bR^{n-1}$ is a diffeomorphism for each $i$, and
+$\partial \Omega \cap {\cal U}_i = \chi_i^{-1}\{\bx \in \bR^n:
+x_n=0\}$. Define $\tilde{\chi}_i: {\cal U}_i \times \bR \rightarrow
+\bR^{n+1}$ by $\tilde{\chi}_i(\bx,t) = (\chi_i(\bx),t)$. Then $\{({\cal U}_i \times \bR, \tilde{\chi_i}):
+i=1,...,N\}$ is a coordinate atlas for $\partial \Omega \times \bR \subset \bR^n$.
+% From \cite{Tay:81}, Ch. II, Theorem 5.1, for $u \in
+%C_0^{\infty}({\cal U}_i)$,
+%\begin{equation}
+ % \label{eqn:pullback}
+ % (H(u \circ \chi_i))\circ \chi_i^{-1} = \tilde{H}_i u,
+%\end{equation}
+%with $\tilde{H}_i \in OPS^0({\cal U}_i)$ with principal symbol
+%\begin{equation}
+ % \sigma(\tilde{H}_i)_0(\bx,{\bf \xi}) =
+  %\sigma(H)_0(\chi_i^{-1}(\bx),D\chi_i(\chi_i^{-1}(\bx))^T{\bf \xi})
+%\end{equation}
+%Also $\tilde{H}_i$ is essentially supported in $\chi_i^*\Gamma_1$
+
+Every point $\bx \in \overline{\cal U}_i \cap \partial \Omega$ has a
+neighborhoods $\bx \in {\cal U}\subset \overline{\cal U} \subset {\cal
+  V}$ in which the
+restrictions of $(\tilde{\chi}_i^{-1})^*\Gamma$ to $\chi_i({\cal V}) \times \bR$ are submanifolds of product conic
+submanifolds $(\chi_i({\cal V}) \times \bR) \times \gamma_i
+\subset T^*(\bR^{n+1})$, so that the pull-backs
+$\tilde{\chi}_i^*((\chi_i({\cal V}) \times \bR) \times \gamma_i)
+\supset \Gamma|_{{\cal V} \times \bR}$ 
+satisfy condition 2 in Definition \ref{thm:definc} with $\delta$
+replaced by $\delta/2$. 
+Taking a finite subcover and intersections, identify the base open sets ${\cal U}$ with
+the coordinate neigborhoods ${\cal U}_i$, with corresponding ${\cal
+  V}_i \supset \overline{\cal U}_i$.
+Choose a partition of unity $\{\phi_i: i=1,...,N\}$ subordinate to
+$\{{\cal U}_i:i=1,...,N\}$, and for each $i=1,...N$, a cutoff function
+$\psi_i\in C_0^{\infty}({\cal V}_i)$ that is $\equiv 1$ on ${\cal U}_i$.
+
+The local product structure means that for each $i=1,...,N$, $({\bf \xi,}\omega) \in \gamma_{i}$,  condition 2 in
+Definition \ref{thm:definc} is satisfied at $\tilde{\chi}_i^*(\bx,t,{\bf \xi},\omega)$ for all $(\bx,t) \in
+(\chi_i({\cal V}_i) \times \bR)$  (with $\delta$ replaced by
+$\delta/2$).
+A standard local geometric asysmptotics construction implies the existence of a 
+neighborhood of the closure of $(\partial \Omega \cap {\cal
+  V}_i)\times \bR$,  in which for each $({\bf \xi,}\omega) \in
+\gamma_{1,i}$, there exists a unique smooth solution of the eikonal
+equation $\tau(\bx,t,{\bf \xi}/\omega)$ with $\tau_i(\bx,t {\bf \xi}/\omega) = t+\chi_i(\bx) \cdot {\bf
+  \xi}/\omega$, for $\bx \in \partial \Omega \cap {\cal  V}_i$, $t \in
+\bR$.
+
+Choose $b \in C_0^{\infty}(\bR)$ so that $b(t)d(\bx,t)=d(\bx,t)$ for
+$\bx \in \partial \Omega, t \in \bR$. Denote by $a_i$ the solution to order 2 of
+the transport system  corresponding to $\tau_i$, with $a_i=b\psi_i$ on $(\partial \Omega \cap
+{\cal  V}_i) \times \bR$. Because of
+the smooth cut-off of $a_i$ at the lateral boundary of a ray tube, and
+because of the uniformity of condition 2 in Definition
+\ref{thm:definc} mentioned above,  the pair
+$(\tilde{p}_i,\tilde{\bv}_i)$ given by
+\begin{eqnarray}
+  \label{eqn:pgo}
+  \tilde{p}(\bx,t,{\bf \xi},\omega) &=& a_i(\bx,t,{\bf
+                                        \xi}/\omega)e^{i\omega\tau(\bx,t,{\bf
+                                        \xi}/\omega)},\\
+  \label{eqn:vgo}
+  \tilde{\bv}(\bx,t,{\bf \xi},\omega) & = & -\frac{1}{\rho(\bx)}\int_{-infty}^t dt \nabla \tilde{p}(\bx,t,{\bf \xi},\omega)
+\end{eqnarray}
+are well-defined and smooth in a neighborhood of $\partial \Omega
+\times \bR$, and satisfy
+\begin{eqnarray}
+\label{eqn:aweloc}
+\frac{1}{\kappa}\frac{\partial \tilde{p}_i}{\partial t} & = & - \nabla \cdot \tilde{\bv}_i +
+\frac{1}{\omega^2}\tilde{r}_i, \nonumber \\
+\rho\frac{\partial \tilde{\bv}_i}{\partial t} & = & - \nabla \tilde{p}_i
+                                               \nonumber \\
+\tilde{p}_i & =& 0 \mbox{ for }  t \ll 0,\nonumber\\ 
+\tilde{bv}_i & = & 0 \mbox{ for } t \ll 0.
+\end{eqnarray}
+The remainder $\tilde{r}_i(\cdot,\cdot,{\bf \xi},\omega)$ is locally
+$L^2$-bounded in a neighborhood of $\partial \Omega \times \bR$,
+uniformly over $({\bf \xi},\omega) \in \gamma_{i}$.
+
+Denote by $\tilde{\gamma}_i$ the complement of the n-dimensional unit
+ball in $\gamma_i$. Let
 \[
-  p^+ = 0 \mbox{ for } t \ll 0 
+  \tilde{d}_i = (\phi_iHd) \circ \chi_i^{-1}
 \]
-to mean that $p^+$ is {\em causal}, that is,
+The essential support of $H$ is strictly contained in
+$\Gamma$, and $(\chi_i^{-1})^*\Gamma|_{{\cal U}_i} \subset
+\chi_i({\cal U}_i) \times \gamma_i$, so the Fourier transform of
+$\tilde{d}_i$ decreases rapidly outside of $\gamma_i$: that is,
+\begin{equation}
+  \label{eqn:smpart}
+  \tilde{e}_i(\by,t) = \int_{\tilde{\gamma}_i^c} d\xi d\omega
+  \hat{\tilde{d}_i}(\xi,\omega)e^{i(\by \cdot {\bf \xi} + \omega t)}
+\end{equation}
+is smooth. Recall that $a_i=b\psi_i=1$ on $\mbox{supp
+}\phi_i Hd = \mbox{supp }\tilde{d}_i\circ \tilde{\chi}_i$, so for $\bx \in {\cal U}_i$
 \[
-  \mbox{For some } T \in \bR, p^+(\cdot,t) = 0 \mbox{ for all } t <
-  T.
+(\tilde{d_i}- \tilde{e}_i) (\chi_i(\bx),t)  = \int_{\tilde{\gamma}_i} d\xi d\omega
+\hat{\tilde{d}_i}(\xi,\omega)e^{i(\chi_i(\bx) \cdot {\bf \xi} + \omega  t)}
 \]
-Similarly,
 \[
-  p^- = 0 \mbox{ for } t \gg 0 
+  = \int_{\tilde{\gamma}_i} d\xi d\omega
+\hat{\tilde{d}_i}(\xi,\omega)(a_i(\bx,t,{\bf
+  \xi},\omega)e^{i\tau_i(\bx,t,{\bf \xi}/\omega)} - \tilde{e}_i(\chi_i(\bx),t)
 \]
-signifies that $p^-$ is anti-causal.
+\begin{equation}
+  \label{eqn:prepsynth}
+  = \int_{\tilde{\gamma}_i} d\xi d\omega \hat{\tilde{d}_i}({\bf
+    \xi},\omega)\tilde{p}_i(\bx,t,{\bf \xi}, \omega) - \tilde{e}_i(\chi_i(\bx),t).
+\end{equation}
+Define 
+\begin{equation}
+  \label{eqn:psynth}
+  \bar{p}_i(\bx,t) = \int_{\tilde{\gamma}_i} d\xi d\omega \hat{\tilde{d}_i}({\bf
+    \xi},\omega)\tilde{p}_i(\bx,t,{\bf \xi}, \omega)
+\end{equation}
+for $\bx \in \partial \Omega \cap {\cal U}_i$. Extend the definition
+of $\bar{p}_i$ to a neighborhood of $\partial \Omega$ in $\Omega$ by
+imposing the condition \ref{eqn:psynth}, and
+define $\bar{\bf{v}}_i$ in the same neighborhood by
+\begin{equation}
+  label{eqn:vsynth}
+  \bar{\bf{v}}_i(\bx,t) = -\frac{1}{\rho(\bx)}\int_{-infty}^t dt \nabla \bar{p}_i(\bx,t)
+\end{equation}
+Then $(\bar{p}_i,{\bar{\bf v}}_i)$ solves the acoustic system with the right hand side
+\begin{equation}
+\label{eqn:rhs}
+\bar{r}_i(\bx,t) = \int_{\tilde{\gamma}_i} d\xi d\omega
+\frac{1}{\omega}^2\tilde{r}_i(\bx,t,{\bf \xi},\omega) -
+\kappa(\bx)\frac{\partial}{\partial t}\tilde{e}_i(\chi_i(\bx),t) .
+\end{equation}
+On the boundary, for $\bx \in \partial \Omega$, the definition
+\ref{eqn:psynth} implies that
+\begin{equation}
+  \label{eqn:locbc}
+  \bar{p}_i(\bx,t) = \tilde{d}_i(\chi(\bx),t)
+  =(\phi_iHd)(\bx,t)
+\end{equation}
+The sums
+\begin{equation}
+  \label{eqn:psynth2}
+  p_2 = \sum_{i=1}^N\bar{p}_i, \bv_2 = \sum_{i=1}^N\bar{\bf v}_i
+\end{equation}
+solve the acoustic system with right-hand side
+\begin{equation}
+  \label{eqn:rsynth}
+  r_2 = \sum_{i=1}^N\bar{r}_i
+\end{equation}
+in a neighborhood of the boundary $\partial \Omega \times \bR$, which
+may be taken to be a collar neighborhood $\Omega_{\epsilon}$ for
+sufficiently small $\epsilon$. Since $\{\phi_i: i=1,...,N\}$ is a
+partition of unity, equation \ref{eqn:locbc} implies that
+\begin{equation}
+  \label{eqn:bc2}
+  p_2|_{\partial \Omega \times \bR} = Hd.
+\end{equation}
+
+Since $\tilde{r}_i$ is uniformly $L^2$-bounded over $\gamma_i$ and
+$\tilde{e}$ is smooth, $\partial_t^2\bar{r}_i$ is
+locally square-integrable in $\Omega_{\epsilon}
+\times \bR$. Denote by $\bu_3=(p_3,\bv_3) \in C^1(\bR,(L^2(\Omega))^{n+1})
+\cap C^0(\bR,V)$ the strong causal solution of the system
+$L\bu=(r_2,0)$. Since the restriction of
+$p_3$ to $\partial \Omega \times \bR$ is well-defined and vanishes,
+$\bu_4 = (p_4,\bv_4) = (p_2-p_3,\bv_2-\bv_3)$ is a causal weak
+solution of $L\bu=0$ in $\Omega \times \bR$ with $p_4|_{\partial \Omega \times \bR}=Hd$.
+
+ Let $\mu \in C_0^{\infty}(\bR^n)$ so that
+$\mu=0$ in $\Omega_{\epsilon/2}$, $\mu=1$ in $\Omega - \Omega_{\epsilon}$.
+
+\end{proof}
 
@@
 
-\section{2D Examples}
-
-\inputdir{project}
-
-To illustrate the structure described in the preceding section, I
-introduce two 2D acoustic models, one spatially homogeneous, the other
-with highly refractive. The first, homogenous model has $\kappa = 4$
-GPa and $\rho = 1$ g/cm$^3$ throughout a rectangular domain of size 8 km ($x$) $\times$ 4 km
-($z$). The second, refractive, model is a perturbation of the first by
-a low-velocity acoustic lens positioned in the center of the
-rectangle. To produce this structure, the density is chosen
-homogeneous as in the first model, while the bulk modulus decreases to
-from 4 GPa outside the lens to 1.6 GPA in its center, as shown in Figure \ref{fig:bml0}.
-
-Discretization is conventional, with a rectangular grid and staggered
-finite difference scheme \cite[]{Vir:84} of order 2 in time and
-2$k$ in space; for most of the experiments reported below, $k=4$.
-Sampling operators such as $P_r$ are implemented via linear
-interpolation, and source insertion via adjoint linear interpolation
-(as noted above, in the continuum limit, sources are represented via
-adjoint sampling). Steps in $x$ and $z$ are the same. Most results use
-$\Delta x = 20$ m, and limit the temporal frequency of the computed
-traces to 12.5 Hz. We also explore the dependence of a few results on frequency, using
-$\Delta x = 10$ m and $5$ m, to accomodate 25 and 50 Hz respectively,
-maintaining 8 samples per wavelength. All computations are carried out
-in single precision.
-
-\cite{GeoPros:11} gives a description of the code
-implementation, out-of-date in a few respects but overall accurate.
-
-The horizontal line of receivers sits at depth $z_r = $ 1000 m, as
-shown in Figure \ref{fig:bml0}. Receiver $x$ ranges from $2000$ to
-$6000$ m. A single point (physical) source appears in these
-experiments, located at $x_s=3500$ m, $z_s=3000$ m, as also indicated
-in the figure. The source time function is a bandpass filter with
-corner frequencies $1, 2.5, 7.5, 12.5$ Hz for the lowest frequency
-source, and scaled as appropriate for examples with higher frequency
-and finer sampling.  Extended sources are confined to the horizontal
-line through the physical source position, that is $z_s = 3000$ m,
-over a 4 km interval starting at $x_r=$ 2000 m. Note that we have
-reversed the order relation between $z_s$ and $z_r$ described in the
-text ($z_s<z_r$). This difference is immaterial.
-
-The point source pressure data generated by this configuration for the
-homogeneous model is displayed in Figure \ref{fig:recphh0}, for the
-lens model in \ref{fig:recplh0}. Triplication of arrivals is evident in the latter.
-
-While inversion of pressure data is the main object of this exercise,
-normal velocity ($v_z$) data will play an important role, so I display
-the corresponding gathers in Figures \ref{fig:recvzhh0} and \ref{fig:recvzlh0}.
-
-\section{Adjoints}
-
-It follows from the adjoint state method (see Appendix A for details) that
-\begin{equation}
-  \label{eqn:sadj1}
-  ({\cal S}^{\pm}_{z_s, z_r})^T = -{\cal S}^{\mp}_{z_r,z_s}
-\end{equation}
-
-Define $R$ to be the {\em time-reversal operator} on functions of
-space-time, $Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em
-  acoustic field time-reversal operator}
-\begin{equation}
-  \label{eqn:trdef}
-  {\cal R} \left(
-    \begin{array}{c}
-      p\\
-      \bv
-    \end{array}
-  \right) =
-  \left(
-    \begin{array}{c}
-      Rp\\
-      -R\bv
-    \end{array}
-  \right)
-\end{equation}
-Then 
-\begin{equation}
-  \label{eqn:trsadj}
-  {\cal R}{\cal S}^{\mp} = -{\cal S}^{\pm}_{z_r,z_s}{\cal R}
-\end{equation}
-Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that
-
-\begin{equation} 
-  \label{eqn:trtr}
- ({\cal S}^{\pm}_{z_s,z_r})^T = {\cal R}{\cal S}_{z_r,z_s}^{\pm}{\cal R}=
- -{\cal S}^{\mp}_{z_r,z_s}.
-\end{equation}
-
-The relation \ref{eqn:trtr} implies that
-\begin{eqnarray}
-  (S^{\pm}_{z_s,z_r})^T &=& -S^{\mp}_{z_r,z_s} \nonumber\\
-                        &=& R S^{\pm}_{z_r,z_s}R, \nonumber\\
-    (V^{\pm}_{z_s,z_r})^T &=& -V^{\mp}_{z_r,z_s} \nonumber\\
-                        &=& R V^{\pm}_{z_r,z_s}R.
-                            \label{eqn:trtrcomp}
-\end{eqnarray}
-
-The implemented code uses the discrete adjoint state method and auto-generated code \cite[]{TapenadeRef13}, to
-assure that the computed adjoint operators are adjoint at the level of
-machine precision to the computed operators. The reverse-time storage
-issue is resolved through the optimal checkpointing technique
-\cite[]{Griewank:book,Symes:06a-pub}, again without loss of precision.
-
-Typical results with pseudorandom input traces $d_r, h_z, w_r, f_s$ are:
-\begin{itemize}
-\item computed $\langle d_r, S^+_{z_s,z_r}h_s\rangle = -2.41069174,
-  \langle (S^+_{z_s,z_r})^Td_r, h_s \rangle = -2.41069388.$
-\item  computed $\langle w_r, V^+_{z_s,z_r}f_s\rangle = -2.73362470,
-  \langle (V^+_{z_s,z_r})^Tw_r, f_s \rangle = -2.73362136.$
-\end{itemize}
-Ideally, the differences of these inner products should be at most a
-relatively small multiple of machine precision, {\em relative} to the
-products $\|d_r\|\| S^+_{z_s,z_r}\|\|h_s\|$ and
-$\|w_r\|\|V^+_{z_s,z_r}\|\|f_s\|$ (division by these quantities makes the result
-dimensionless and scale-independent). The operator norm $\| S^+_{z_s,z_r}\|$ is
-computationally inaccessible, so instead I used the smaller quantities
-$\|d_r\|\| S^+_{z_s,z_r}h_s\|$ etc. as stand-ins - thereby
-overestimating the relative error between the inner products. In all
-cases, over a very large number of random inputs, the largest observed
-relative error estimate was $O(10^{-9})$, well under the appropriate limit for
-single precision.
-
-\section{Pressure-to-Source}
-
-Since the system \ref{eqn:awepm} has a unique solution by standard
-theory \cite[]{Lax:PDENotes}, the source vector field $(h_s,f_s)$
-determines the acoustic field $(p^{\pm},\bv^{\pm})$ in space time, and
-in particular the limits from the right at $z=z_s$, $P_sp^{\pm}$ and
-$P_sv_z^{\pm}$. This relation is not invertible: it is not possible to
-prescribe both pressure and normal velocity on a surface such as
-$z=z_s$. So the columns of the matrix operator
-${\cal S}^{\pm}_{z_s,z_r}$ must satisfy a linear relation. In this
-section I will explain this relation; it involves the {\em
-  pressure-to-source} map. This operator also turns out to be the
-principle component of a preconditioning strategy for iterative
-solution of the optimization problem \ref{eqn:esis}, so I will devote
-some effort to its proper definition. It is closely related to the
-Dirichlet-to-Neumann operator mentioned in the introduction.
-
-While it is not possible to prescribe both pressure and velocity on
-$z=z_s$ in solutions of \ref{eqn:awepm}, it is possible to
-prescribe pressure only, for instance: if the function $\phi$ on
-the surface $z=z_s$ satisfies suitable conditions, for
-example the downgoing constraint mentioned earlier, a unique solution
-exists for the acoustic system in both half-spaces $\pm z > z_s$:
-\begin{eqnarray}
-\label{eqn:awe0}
-  \frac{1}{\kappa}\frac{\partial p_{\pm}}{\partial t} & = & - \nabla \cdot \bv_{\pm}, \nonumber \\
-  \rho\frac{\partial \bv_{\pm}}{\partial t} & = & - \nabla
-                                                    p_{\pm}, \nonumber \\
-  p_{\pm} & =& 0,  \mbox{ for } t \ll 0, \nonumber\\ 
-  \bv_{\pm} & = & 0 \mbox{ for } t \ll 0, \nonumber\\
-  \lim_{z \rightarrow z_s^{\pm}}p_{\pm}(x,y,t,z)& =& \phi(x,y,t).
-\end{eqnarray}
-Note that the subscript $\pm$ here refers to the sign of $z-z_s$, as opposed
-to the superscript ${\pm}$, which refers to the sign of $t$ throughout
-this paper.
-
-From the boundary condition (last equation in \ref{eqn:awe0}), one
-sees that the pressures $p_{\pm}$ in the two half-spaces have the same
-limit at the boundary $z=z_s$. Stick the two half-space
-solutions together to form an acoustic field $(p^+,\bv^+)$ in all of
-space-time, that is,
-\begin{equation}
-  \label{eqn:awealt}
-  p^+(x,y,z,t) =
-  \left\{
-    \begin{array}{c}
-      p_+(x,y,z,t) \mbox{ if } z>0,\\
-      p_-(x,y,z,t) \mbox{ if } z<0,
-    \end{array}
-  \right.
-\end{equation}
-and a similar definition for $\bv^+$. Then $p^+$ is continuous across
-$z=z_s$, and the boundary condition in system \ref{eqn:awe0} may be
-written as $P_sp^+=\phi$.
-
-The same construction can be carried out in the anti-causal sense,
-with anti-causal half-space solutions glued together to form a
-full-space distribution solution $(p^-,\bv^-)$, with the property that
-$p^-$ is continuous across $z=z_s$ and $P_sp^-=\phi$.
-
-The reader may object that the notation $(p^\pm,\bv^\pm)$ is already in
-use, for the solution of \ref{eqn:awepm}. This objection is
-valid. However, {\em in the sense
-  of distributions}, $(p^{\pm},\bv^{\pm})$ as defined in display
-\ref{eqn:awealt}, is {\em exactly} the causal solution of \ref{eqn:awepm}
-for the choice $h_s = -[v^{\pm}_{z}]|_{z=z_s}, f_s=0$, as follows from a
-simple integration-by-parts calculation. So the notation is consistent!
-
-The negative jump $-[v^{\pm}_{z}]|_{z=z_s}$ is thus a function of $\phi$. Define
-the {\em pressure-to-source} operator $\Lambda^{\pm}_{z_s}$ by
-\begin{equation}
-  \label{eqn:deflam}
-  \Lambda^{\pm}_{z_s}\phi = -[v^{\pm}_{z}]|_{z=z_s}
-\end{equation}
-The conclusion: if $h_s = \Lambda^{\pm}_{z_s}\phi$ and $f_s=0$ in the
-system \ref{eqn:awepm}, then $\phi=P_sp^{\pm}$.
-
-Otherwise put, $S^{\pm}_{z_s,z_s}\Lambda^{\pm}_{z_s} \phi = \phi$, so
-$\Lambda^{\pm}_{z_s}$ is inverse to $S^{\pm}_{z_s,z_s}$. The relation
-\ref{eqn:trtrcomp} implies in turn that
-\begin{equation}
-  \label{eqn:lamadj}
-  (\Lambda^{\pm}_{z_s})^T = - \Lambda^{\mp}_{z_s}
-\end{equation}
-
-There is also a {\em velocity-to-source} operator. For the solution
-$(p^{\pm},\bv^{\pm})$ of system \ref{eqn:awepm} with $h_s=0$, the
-normal component of velocity, $v^{\pm}_z$, is continuous across
-$z=z_s$, and the velocity source (vertical load)
-$f_s=-[p^{\pm}]_{z=z_s}$. I will not name the velocity-to-source
-operator, as it does not appear explicitly in the developments to
-follow. As will be seen, it is essentially the inverse of the
-pressure-to-source operator.
-
-The quadratic form defined by $\Lambda^{\pm}_{z_s}$ has fundamental
-physical significance. Define the total acoustic energy $E^{\pm}(t)$ of the
-field $(p^{\pm},\bv^{\pm})$, at time $t$ by
-\begin{equation}
-  \label{eqn:defae0}
-  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)
-\end{equation}
-Then
-\begin{equation}
-  \label{eqn:elim}
-  \pm \lim_{\pm t \rightarrow \infty} E^{\pm}(t) =  \langle P_sp^{\pm},
-  (\Lambda^{\pm}_{z_s} P_sp^{\pm}) \rangle_{L^2(z=z_s)}.
-\end{equation}
-That is, the value of the quadratic form defined by
-$\Lambda^{\pm}_{z_s}$, evaluated at the pressure trace on $z=z_s$,
-gives the total energy transferred from the source to the
-acoustic field over time. Since $E$ is itself a positive definite
-quadratic form in the acoustic field, it follows that $\pm
-\Lambda^{\pm}_{z_s}$ is positive semi-definite.
-
-While $\Lambda^{\pm}_{z_s}$ is positive semi-definite, it is not
-symmetric. However, it is {\em approximately symmetric} in the
-high-frequency sense. This fact follows from a
-geometric optics analysis of the half-space solution,
-assuming (as always) downgoing data:
-\begin{equation}
-  \label{eqn:lamappsim}
-  (\Lambda^{\pm}_{z_s})^T \approx \Lambda^{\pm}_{z_s}.
-\end{equation}
-
-The analysis also reveals that the solution components not continuous
-at $z=z_s$ are odd there:
-\begin{equation}
-  \label{eqn:odd1}
-  \lim_{z\rightarrow z_s^+} v^{\pm}_{z} \approx - \lim_{z\rightarrow z_s^-}
-  v^{\pm}_{z}
-\end{equation}
-for the solution of \ref{eqn:awepm} with $f_s=0$.
-Similarly, 
-\begin{equation}
-  \label{eqn:odd2}
-  \lim_{z\rightarrow z_s^+} p^{\pm}\approx - \lim_{z\rightarrow z_s^-}
-  p^{\pm}
-\end{equation}
-for the solution of \ref{eqn:awepm} with $h_s=0$. Here ``$\approx$''
-means in the sense of high frequency asymptotics, that is, that the
-difference between the two sides is relatively smooth, hence small if
-the data is highly oscillatory. Therefore if $f_s=0$ in system \ref{eqn:awepm},
-\begin{equation}
-  h_s = \Lambda^{\pm}_{z_s}P_sp^{\pm} = -[v^{\pm}_{z}]|_{z=z_s} \approx -2
-  P_sv^{\pm}_{z}
-  \label{eqn:tracejump10}
-\end{equation}
-Similarly, if $h_s=0$ in system \ref{eqn:awepm}, then
-\begin{equation}
-  \label{eqn:tracejump20}
-  f_s = -[p^{\pm}]|_{z=z_s} \approx -2 P_s p^{\pm}.
-\end{equation}
-Thus $f_s$ determines approximately the boundary value of $p^{\pm}$,
-as a solution of the acoustic wave system in the half-space
-$z>z_s$. However, as repeated in equation \ref{eqn:tracejump10}, a
-solution with this boundary value is also the restriction to $z>z_s$
-of a solution to \ref{eqn:awepm} with $f_s=0$ and $h_s=
-\Lambda^{\pm}_{z_s}P_sp^{\pm}$. Therefore if
-\begin{equation}
-  \label{eqn:hfcondn}
-  h_s =-\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,
-\end{equation}
-then the pressure boundary value $P_sp^{\pm}$ is the
-same for the solutions of \ref{eqn:awepm} for source vectors $(h_s,0)$
-and $(0,f_s)$. Since the pressure boundary values are the same, the solutions
-in $z>z_s$ are the same. In particular, since $z_r>z_s$ and ${\cal
-  S}^{\pm}_{z_s,z_r}(h_s,f_s)^T = (P_rp^{\pm},P_rv^{\pm}_z)^T$, it follows
-that
-\begin{equation}
-  \label{eqn:snull}
-  {\cal S}^{\pm}_{z_s,z_r}\left(\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,f_s\right)^T \approx 0.
-\end{equation}
-
-Equation \ref{eqn:snull} states the relation between the columns of $
-{\cal S}^{\pm}_{z_s,z_r}$ mentioned in the introduction to this
-section.
-
-To give an example of \ref{eqn:snull} in action, it is necessary to
-create downgoing data and the action of the operator $\Lambda^{\pm}$
-on a downdoing field data $f_s$ on $z=z_s$. Note that a point source
-on $z=z_s$ creates high frequency energy traveling on rays parallel
-and nearly parallel to $z=z_s$, so that won't do. One simple approach
-is to create pressure and normal velocity data on $z=z_s$ by placing a
-point source at a depth $z_d<z_s$. Since the examples used here are
-homogeneous in $z<z_s$, the traces extracted from the resulting
-causal pressure and velocity data on $z=z_s$ are {\em a priori}
-downgoing, hence related by the operator $\Lambda^+_{z_s}$.
-
-Since the
-mechanical parameters in the homogeneous and lens models are the same
-for $z<z_s$, and no rays return to this zone in either model, these
-data are asymptotically the same for both models, and I show only the
-homogenous medium results. Regard these gathers as the trace
-$(P_sp^+,P_sv^+_z)$ on $z=z_s$ of a downgoing acoustic field in $z>z_s$.
-Equations \ref{eqn:tracejump10} and
-\ref{eqn:tracejump20} show that these differ by a factor of -2 from
-source functions $f_s$ and $h_s$ as in the system \ref{eqn:awepm},
-with $h_s=0$ and $f_s=0$ respectively. For a point source at $z_d=3500$ m, $x_d=3500$ m, same bandpass filter
-wavelet as used for prior examples, causal data on $z=z_s=3000$ m are
-shown in Figures \ref{fig:dsrcphh0} and \ref{fig:dsrcvzhh0}.
-
-These source functions satisfy
-the relation \ref{eqn:hfcondn}, therefore source vectors $(h_s,0)$ and
-$(0, f_s)$ generate {\em the same acoustic field} $(p^+,\bv^+)$ in $z>z_s$. Figures
-\ref{fig:drecphh0}, \ref{fig:dfwdphh0}, \ref{fig:daltphh0} (homogeneous model) and
-\ref{fig:drecplh0}, \ref{fig:dfwdplh0}, \ref{fig:daltplh0} (lens model) show the pressure
-gathers extracted at $z_r=1000$ m for the point source at $z=z_d$ and
-for the two choices of extended source at $z=z_s$, on the same color
-scale. The obvious similarity between the fields generated by the two
-extended sources,
-predicted by equation \ref{eqn:snull}, is confirmed by trace
-comparisons in figures \ref{fig:drecphh0tr21}-\ref{fig:daltphh0tr81}
-and
-\ref{fig:drecplh0tr21}-\ref{fig:daltplh0tr81}.
-
-\section{Time Reversal}
-
-Recall that the source vector $(h_s,f_s)$ is assumed to produce a
-downgoing field $(p^+,\bv^+)$, that is, emanates high-frequency energy only along
-rays that make an angle with the vertical bounded below by a common
-minimum angle. Such rays leave $\Omega$ within a common maximum
-time. Consequently (Appendix B), in the
-slab $z_s<z<z_r$, the field $(p^+,\bv^+)$ approximates the solution of an
-anti-causal evolution equation. Choose $\chi(t)$ to be a smooth function
-that is $= 0$ for $t \gg 0$ and $=1$ at times when near rays carrying
-high-frequency energy in $(p^+,\bv^+)$ cross $z=z_r$. Define 
-$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
-$\Omega \times \bR$ of
-\begin{eqnarray}
-\label{eqn:revawe}
-  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
-  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
-  \tilde{p}^- & =& 0,  \mbox{ for } t \gg 0\\ 
-  \tilde{\bv}^- & = & 0 \mbox{ for } t \gg 0\\
-  P_r\tilde{p}^- &=& \chi P_rp^+ . 
-\end{eqnarray}
-That is, $\tilde{p}^-$ has the same boundary value on $z=z_r$ as
-$p^+$, except for low-frequency residue that is muted by
-$\chi$. Therefore
-$p^+ \approx \tilde{p}^-, \bv^+ \approx \tilde{\bv}^-$ near
-$z=z_r$. Since the right-hand sides in the system \ref{eqn:awepm} are
-singular only on $z=z_s$, and the high-frequency components of
-$(p^+,\bv^+)$ are carried by downgoing rays, these differ negligibly
-from the the high-frequency components of
-$(\tilde{p}^-,\tilde{\bv}^-)$ in the space-time slab $z_s<z<z_r$, and
-the approximation holds throughout this region. In particular
-$P_sv^+_z \approx P_s \tilde{v}^-_z$. In view of the relation
-\ref{eqn:tracejump10},
-\begin{equation}
-  \label{eqn:tildevtohsubs}
-  -2P_s\tilde{v}^-_z \approx h_s,
-\end{equation}
-so solution
-of the system \ref{eqn:revawe} followed by restriction to $z=z_s$ and
-multiplication by $-2$ 
-approximately inverts the map $S^+_{z_s,z_r}: h_s \mapsto P_rp^+$.
-
-Next observe that in view of the relation \ref{eqn:tracejump20}, and
-the downgoing nature of the ray system carrying the high frequency
-energy in $(p^+,\bv^+)$, the field $(\tilde{p}^-,\tilde{\bv}^-)$ is
-actually the restriction to $z<z_r$ of the anti-causal solution of \ref{eqn:awepm}
-with $z_s$ replaced by $z_r$, zero constitutive defect, and vertical
-load given by the jump in pressure at $z=z_r$ - for this field, use
-the same notation. Continuity of vertical
-velocity $\tilde{v}^-_z$ at $z=z_r$ implies that the vertical load is
-\[
-  f_r = -[\tilde{p}^-]|_{z=z_r} =-(\lim_{z\rightarrow
-    z_r^+}\tilde{p}^- - \lim_{z\rightarrow
-    z_r^-}\tilde{p}^-)
-\]
-\[
-  \approx 2 P_r \tilde{p}^- = 2 P_r p^+
-\]
-(from the definition \ref{eqn:defsamp}, $P_r$ is the limit from the
-left). Thus
-\[
-  P_s \tilde{v^-_z} \approx V^-_{z_r,z_s}(2 P_rp^+) \approx
-  2V^-_{z_r,z_s}S^+_{z_r,z_s}h_s.
-\]
-so
-\[
-  h_s \approx -2 P_s v^+_z \approx -2 P_s \tilde{v}^-_z \approx
-  -4V^-_{z_r,z_s}S^+_{z_r,z_s}h_s
-\]
-Combine this observation with \ref{eqn:tildevtohsubs} to obtain
-\[
- -4  V^-_{z_r,z_s} S^+_{z_s,z_r}  \approx  I,
-\]
-This relation combines with the identity \ref{eqn:trtrcomp} to
-yield the first main result of this section:
-\begin{eqnarray}
-  \label{eqn:approxinv}
-  (V^+_{z_s,z_r})^T S^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
-  (S^+_{z_s,z_r})^T V^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
-  V^+_{z_s,z_r} (S^+_{z_s,z_r})^T & \approx & \frac{1}{4}I, \nonumber\\
-  S^+_{z_s,z_r} (V^+_{z_s,z_r})^T & \approx & \frac{1}{4}I.
-\end{eqnarray}.
-The second equation is simply the transpose of the first, and the
-last two follow by by an exactly analogous argument using time
-reversal and interchange of the roles of $z_s$ and$z_r$.
-
-The conclusion is significant enough to merit restating in English:
-provided that high-frequency energy in the various fields is carried
-along downgoing ray fields, the transpose of $V^+$ is an approximate
-inverse to $S^+$, modulo a factor of 4. To recover the pressure source
-$h_s$ generating a pressure gather $P_rp$ at $z=z_r$, multiply the
-latter by -2, then apply the transpose of $V^+_{z_s,z_r}$ to this
-gather, reading out a vertical velocity field at $z=z_s$. Multiply
-again by -2 and you have a high-frequency approximation to $h_s$.
-
-I have applied this procedure to the pressure gathers shown in Figures
-\ref{fig:drecphh0} and \ref{fig:drecplh0}, generated by a point source
-at $z_d=3500$ m, $x_s=3500$ m, propagating in the homogeneous and lens
-models respectively. The source time dependence is again a [1, 2.5,
-10, 12.5] Hz trapezoidal bandpass filter. As noted earlier, in the
-region $z < z_s=3000 m$, these fields are downgoing and can be
-regarded as the result of either pressure or velocity source at $z=z_s$: the
-pressure source gather $h_s$ (Figure \ref{fig:dhshh0})  is -2 times the
-vertical velocity gather depicted in Figure \ref{fig:dsrcvzhh0}, the
-velocity source gather $f_s$ (Figure \ref{fig:dfshh0}) is -2 times the
-pressure gather depicted in Figure \ref{fig:dsrcphh0}.
-
-Figure \ref{fig:dinvhshh0} shows the approximate inversion (via the
-first equation in display \ref{eqn:approxinv}) of the
-pressure gather shown in Figure \ref{fig:drecphh0}, inverted in the homogeneous model
-used to generate the data. The result is a scaled, dip-filtered
-version of the pressure source shown in Figure \ref{fig:dhshh0}.
-The dip filter effect is unavoidable and is caused by the aperture
-limitation of the acquisition geometry: the steeper dips in the source
-gather (Figure \ref{fig:dhshh0}) do not contribute to the data, and are not
-recovered in the inversion.
-
-To see that this is indeed an inversion, I apply $S^{+}_{z_s,z_r}$ to
-the source shown in Figure \ref{fig:dinvhshh0}, to produce the
-re-simulation shown in Figure \ref{fig:drerecphh0}. Simulation is carried
-out in the homogeneous model used in the inversion. Comparison with the
-gather in Figure \ref{fig:drecphh0} via a difference plot
-\ref{fig:ddiffrecphh0} verifies that a very close match has been
-obtained, so the source in Figure \ref{fig:dinvhshh0} is indeed a
-satisfactory inversion, even it differs considerably from the original
-point source: the difference lies in an approximate null space of the modeling
-operator, arising from the limited aperture of the acquisition
-geometry.
-
-The remainder of this section is devoted to illustrating two important
-features of the approximate inversion developed above. The
-first is the validity of the inversion approach given in display
-\ref{eqn:approxinv} so long as the fields generated by the
-model used in simulation and inversion are downgoing. The models
-may be quite far from homogeneous, and even generate triplications,
-whereas the inversions remain accurate modulo the dip filtering effect
-of aperture. Figure \ref{fig:dinvhsll0} shows the pressure source at $z=z_s$
-inverted from the data Figure
-\ref{fig:drecplh0}, in which a triplication is obvious. Recall that
-this data is generated by propagating a point source at $z_d=3500$ m
-in the lens model. The inversion uses the same model. Figure
-\ref{fig:dinvhsll0} is an aperture-limited version of the pressure
-source at $z=z_s=3000$ m
-(Figure \ref{fig:dhshh0}) that generates this data by a dip filter, in
-the same way that the sources in Figures \ref{fig:dhshh0} and
-\ref{fig:dinvhshh0} are related. The inverted source is an accurate
-inversion: it generates the data depicted in Figure
-\ref{fig:drerecpll0}, which is very close to the input data (Figure
-\ref{fig:drecplh0}). The difference is displayed on the same color
-scale in Figure \ref{fig:ddiffrecpll0}.
-
-The second feature is a decoupling of the models used (implicitly or
-otherwise) in simulation and those used in inversion: it is possible
-to {\em accurately invert pressure data for pressure source using the
-  wrong model}, in the sense that the inverted source will
-generate an accurate recovery of the input data re-simulated in the
-same (wrong) model. For example, inversion of the data of the previous
-example (Figure \ref{fig:drecplh0}) {\em in the homogeneous model}
-results in an inverted source gather at $z=z_s$ (Figure \ref{fig:dinvhslh0}
-that is very far from a dip-filtered version of the source used
-generate the data (Figure \ref{fig:dhshh0}). However, this result is an
-accurate inversion: re-simulation {\em using the same (homogeneous)
-  model as used in the inversion} results in accurate recovery of the
-input pressure gather (Figure \ref{fig:drecplh0}). The difference is
-shown on the same color scale in Figure \ref{fig:ddiffrecplh0}.
-
-\section{Unitarity}
-
-The next chapter in this story recognizes the relations in display
-\ref{eqn:approxinv} as asserting the approximate unitarity of
-$S^+_{z_s,z_r}$.
-
-The matrix identity \ref{eqn:snull} implies a relation between $S, V,$
-and $\Lambda$ of some interest in itself. After minor re-arrangement, the second row of reads
-\begin{equation}
-  \label{eqn:snull2}
--\frac{1}{2}\Pi_1{\cal S}^{\pm}_{z_s,z_r}\Pi_0^T\Lambda^{\pm}_{z_s}  \approx
-V^{\pm}_{z_s,z_r}.
-\end{equation}
-In these relations, the projection on the left picks out the vertical velocity component
-of a downgoing wavefield at $z=z_r$: that is,
-\[
--\frac{1}{2}\Pi_1{\cal S}^{\pm}_{z_s,z_r}\Pi_0^T\Lambda^{\pm}_{z_s}P_sp^+
-=-\frac{1}{2}P_r v_z^+,
-\]
-where $(p^+, \bv^+)$ solve the system \ref{eqn:awepm} with $f_s=0$ and
-$h_s = \Lambda^{\pm}_{z_s}P_sp^+$. On the other hand, from relation
-\ref{eqn:tracejump10},
-\[
-  P_r v_z^+ = -\frac{1}{2}\Lambda^+_{z_r}P_r p^+
-\]
-where
-\[
-  P_r p^+ = \Pi_0{\cal S}^{+}_{z_s,z_r}\Pi_0^T\Lambda^{+}_{z_s}P_s
-  p^+
-\]
-\[
-  = S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_sp^+
-\]
-Therefore combining the last two equations with \ref{eqn:snull2},
-obtain
-\begin{equation}
-  \label{eqn:sv}
-  \frac{1}{4}\Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s} = V^+_{z_s,z_r}.
-\end{equation}
-This is the promised relation.
-
-A minor extension of an earlier example, based on the gathers
-displayed in Figures \ref{fig:dsrcphh0} ($P_sp^+$) and
-\ref{fig:dsrcvzhh0} ($P_sv_z^+$) illustrates the identity
-\ref{eqn:sv}.
-The pressure gathers at $z=z_r$ produced by application of
-$S^+_{z_r,z_s}$ to $h_s=-2P_s v^+_z$, and $\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T$ to $f_s
-=-2P_sp^+$, have already been displayed (Figures \ref{fig:dfwdphh0}
-and \ref{fig:daltphh0}), and are identical to several digits. The same 
-is true of the vertical velocity gathers, shown in \ref{fig:dfwdvzhh0}
-and \ref{fig:daltvzhh0}; their difference, plotted on the same scale,
-appears in Figure \ref{fig:dsvcomphh0}. Once again, this relation
-expresses the interchangeability of pressure and velocity sources,
-related by $\Lambda$.
-
-Relation \ref{eqn:tracejump10} characterizes
-$\Lambda^+_{z_s}$ as connecting the pressure and velocity components
-of downgoing fields restricted to $z=z_s$. That is, the pressure
-source gather
-$h_s=\Lambda^{+}_{z_s}P_s p^+ = -2
-P_{z_s}v_z$, displayed in Figure \ref{fig:dhshh0}, is the image of the
-pressure gather in Figure \ref{fig:dsrcphh0} under
-$\Lambda^{+}_{z_s}$.
-The pressure gather
-\ref{fig:dfwdphh0} is the image of this pressure source gather under
-$S^+_{z_s,z_r}$. The corresponding vertical velocity gather
-(Figure \ref{fig:dfwdvzhh0}) is $-1/2$ times $\Lambda^+_{z_r}P_r
-p$. Therefore scaling the data in Figure \ref{fig:dfwdvzhh0} by
-$-2$ produces
-$\Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_s p^+$. On the other
-hand, figure \ref{fig:daltvzhh0} shows the result of applying
-$V^+_{z_s,z_r}$ to $f_s=-2P_s p^+$. Therefore scaling the gather in
-Figure \ref{fig:daltvzhh0} by $-\frac{1}{2}$ produces $V^+_{z_s,z_r}P_zp^+$.
-Since the data in Figures \ref{fig:dfwdvzhh0} and \ref{fig:daltvzhh0}
-are essentially identical, the relation \ref{eqn:sv} holds for this
-example.
-
-As shown in the last section, $4(V_{z_s,z_r}^+)^T$ is approximately
-inverse to $S^{+}_{z_s,z_r}$. Therefore, transposing both sides of
-equation \ref{eqn:sv} and using \ref{eqn:approxinv}, obtain
-\begin{equation}
-  \label{eqn:almostunitary}
-  4(V_{z_s,z_r}^+)^TS^+_{z_s,z_r} = [ (\Lambda^+_{z_s})^T
-  (S^{+}_{z_s,z_r})^T(\Lambda^+_{z_r})^T]S^{+}_{z_s,z_r} \approx I.
-\end{equation}
-
-The remarkable feature of the identity \ref{eqn:almostunitary} is that
-it exhibits an approximate right inverse of $S^+$ as an adjoint with
-respect to a weighted inner product - or it would, if the operators
-$(\Lambda^+)$ were symmetric positive definite. As noted earlier,
-these operators are only approximately symmetric, though they are
-positive semi-definite. That is not a great obstacle, however:
-symmetrizing them in the obvious way commits a negligible error, of
-the sort that this paper already neglects wholesale. That is,
-\begin{equation}
-  \label{eqn:unitary}
-  [ \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
-  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
-  \Lambda^+_{z_r})]S^{+}_{z_s,z_r} \approx I.
-\end{equation}
-
-To illustrate this unitary property of $S^+_{z_s,z_r}$, I apply the
-operator
-\[
-  \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
-  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
-  \Lambda^+_{z_r})
-\]
-to the data $S^+_{z_s,z_r}h_s$ (Figure \ref{fig:drerecphh0}), in which
-$h_s$ is the dip-filtered version (Figure \ref{fig:dinvhshh0}) of the
-downgoing source created earlier (Figure \ref{fig:dhshh0}), produced
-by inversion of the data in Figure \ref{fig:drecphh0} via the first
-equation in display \ref{eqn:approxinv}, propagation in the homogenous
-model.  The operator above is computed via the technique explained in
-the section ``Computing and Symmetrizing $\Lambda$'', below, using
-auxiliary receiver arrays 100 m above the data source and receiver arrays.
-
-The output is shown in Figure
-\ref{fig:lamsstlamrdrecphh0}. The difference with the actual source is
-shown in Figure \ref{fig:difflamsstlamrdrecphh0}. A similar exercise
-using the lens data \ref{fig:drecplh0} but inversion in the homogenous
-model produces the extended source depicted in
-\ref{fig:lamsstlamrdrecplh0}, which differs from the inversion result
-via equation \ref{eqn:approxinv} by the gather displayed in Figure
-\ref{fig:difflamsstlamrdrecplh0}.
-
-The symmetrized $\Lambda$ operators are at least positive
-semi-definite, hence define (at least) semi-norms.
-Similar relations have been derived for other scattering operators,
-and have been used to accelerate iterative solutions of inverse
-scatering problems: \cite{DafniSymes:SEG18b} review some of this
-literature.
-
-\section{Accelerated Iterative Inversion}
-
-For convenience, in this section write $S$ in place of
-$S^+_{z_s,z_r}$. Also abbreviate the symmetrized $\Lambda$ operators
-using notation suggesting weight
-operators in model and data spaces:
-\begin{eqnarray}
-  W_m^{-1}&=& \frac{1}{2}((\Lambda^+_{z_s})^T+
-              \Lambda^+_{z_s}),\nonumber \\
-  W_d &=& \frac{1}{2}((\Lambda^+_{z_r})^T+ \Lambda^+_{z_r}).
-          \label{eqn:wdef}
-\end{eqnarray}
-The identification of the symmetrized $\Lambda^+_{z_s}$ as the inverse
-of another operator $W_m$ is formal, since the former operator is
-likely to have null (or nearly-null) vectors due to aperture-related
-amplitude loss. Since some version of $W_m$ is essential in the
-formulation for effective preconditioning, I will derive a usable
-candidate to stand in for it below.
-
-Adopting Hilbert norms defined by the operators $W_m$ and $W_d$ in its
-domain and range respectively, the adjoint of $S$ is given by
-\begin{equation}
-\label{eqn:wadj}
-S^{\dagger} = W_m^{-1}S^TW_d,
-\end{equation}
-
-In this notation, the relation \ref{eqn:unitary} takes the form
-\begin{equation}
-  \label{eqn:wunitary}
-  S^{\dagger}S \approx I.
-\end{equation}
-That is to say, $S$ is approximately unitary with respect to the
-weighted norms defined by $W_m$ and $W_d$. Therefore a Krylov space
-method employing these norms will converge rapidly, at least for the
-well-determined components of the solution.
-
-Similar relations have been derived for other scattering operators,  
-and have been used to accelerate iterative solutions of inverse  
-scattering problems: \cite{DafniSymes:SEG18b} review some of this  
-literature.
-
-The most convenient arrangement the Conjugate Gradient (CG) algorithm
-taking advantage of the structure \ref{eqn:wadj} is the {\em
-  Preconditioned CG}.
-
-Allowing that the fit error will be measured by the data space norm,
-the least squares problem to be solved is not just $Sh \approx d$, but
-a regularized version:
-\begin{equation}
-  \label{eqn:einv}
-  \mbox{minimize}_h \|Sh-d\|^2_d + \alpha^2 \|Ah\|^2_m
-\end{equation}
-Remark: recall that the modified data space norm $\|d\|_d^2 = \langle
-d, W_d d\rangle$ has physical meaning: for acoustics, it is
-proportional to the power transmitted to the fluid by the source.
-
-The minimizer of the objective defined in equation \ref{eqn:einv}
-solves the normal equation
-\begin{equation}
-  \label{eqn:norm0}
-  (S^{\dagger}S + \alpha^2 A^{\dagger}A)h = S^{\dagger}d 
-\end{equation}
-where the weighted adjoint $S^{\dagger}$ has already been defined in equation \ref{eqn:wadj}, and $A^{\dagger}$ is the adjoint of $A$ in the weighted model space norm defined by $W_m$, namely
-\begin{equation}
-  \label{eqn:aadj}
-  A^{\dagger} = W_m^{-1}A^TW_m.
-\end{equation}
-
-Note that the normal operator appearing on the left-hand side of
-\ref{eqn:norm0} is not an approximate identity, due to the presence of
-the regularization term: the spectrum increases in spread with
-increasing $\alpha$, leading to slower convergence. Fortunately for
-the present setting, the operators $W_m^{-1}$, $A$, and $W_m$
-approximately commute (they are scalar {\em pseudodifferential}, once
-the difficulties with the definition of $W_m$, mentioned above, are
-taken care of), so that $A^{\dagger} \approx A^T$. Therefore
-\begin{equation}
-  \label{eqn:normapprox}
-  S^{\dagger}S + \alpha^2 A^{\dagger}A \approx I + \alpha^2A^TA
-\end{equation}
-Recall that $A$ is simply multiplication by the Euclidean distance to
-the physical source point $\bx_s$: $A u (\bx) = |\bx-\bx_s|u(\bx),
-A^TAu(\bx) = |\bx-\bx_s|^2u(\bx)$. So the equation $(I+\alpha^2
-A^TA)u=b$ is trivial to solve, and this is a key characteristic of a
-good preconditioner. However this observation must be combined with
-the weighted norm structure.
-
-Rewrite the normal equation \ref{eqn:norm0} as
-\begin{equation}
-  \label{eqn:norm1}
-  W_m^{-1}(S^TW_dS + \alpha^2 A^TW_mA)h = W_m^{-1}S^TW_md 
-\end{equation}
-Since $W_m$ is self-adjoint and positive definite, the common factor on both sides of \ref{eqn:norm1} can be re-written as
-\begin{equation}
-  \label{eqn:normpart}
-  Nh = (S^*S + \alpha^2 A^*A)h = S^*d 
-\end{equation}
-in which $S^*, A^*$ are the adjoints with the original (Euclidean)
-inner product in the domains but the weighted inner product in data
-space:
-\begin{eqnarray}
-  \label{eqn:sadj}
-  S^* &=& S^T W_d,\\
-  A^* &=& A^T W_m.
-\end{eqnarray}
-Note the $S^*S$ and $A^*A$ are symmetric in the Euclidean sense, so
-equation \ref{eqn:normpart} is a symmetric positive (semi-)definite
-linear system, just the sort of thing for which the 
-The Preconditioned Conjugate Gradient (``PCG'') algorithm was
-designed. PCG for solution
-of equation \ref{eqn:normpart} with preconditioner $M$is usually
-written as Algorithm 1 (see for example \cite{Golub:2012}):
-
-\begin{algorithm}[H]
-\caption{Preconditioned Conjugate Gradient Algorithm, Standard Version}
-\begin{algorithmic}[1]
-\State Choose $h_0=0$ 
-  \State $r_0 \gets S^*d$
-  \State $p_0 \gets M^{-1} r_0$
-  \State $g_0 \gets p_0$
-  \State $q_0 \gets Np_0$
-  \State $k \gets 0$
-  \Repeat
-  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle p_k,q_k\rangle}$
-  \State $h_{k+1} \gets h_k + \alpha_k p_k$
-  \State $r_{k+1} \gets r_k - \alpha_kq_k$
-  \State $g_{k+1} \gets M^{-1} r_{k+1}$
-  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
-  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
-  \State $q_{k+1} \gets Np_{k+1}$
-  \State $k \gets k+1$
-  \Until{Error is sufficiently small, or max iteration count exceeded} 
-\end{algorithmic}
-\end{algorithm}
-The iteration converges rapidly if $M^{-1}N \approx I$. This is true
-if and only if the symmetrized operator $M^{-1/2}NM^{-1/2} \approx I$,
-which is in turn true if the eigenvalues of $M^{-1/2}NM^{-1/2}$ are
-close to 1 (actually works well is most of these eigenvalues are close
-to 1, and the rest are small - which is the case for the current
-problem)..  Further, PCG is computationally effective is M is easy to
-invert.
-
-From \ref{eqn:normapprox} and \ref{eqn:norm1}, it follows that
-\[
-  W_m^{-1}(S^TW_dS + \alpha^2 A^TW_mA) \approx I + \alpha^2 A^TA.
-\]
-This observation suggests using $M=W_m(I+\alpha^2A^TA)$. This choice
-is not symmetric, but since the operators on the right-hand side are
-scalar pseudodifferential hence commute, it is equivalent to use of
-\begin{eqnarray}
-  M         &=&(I+\alpha^2A^TA)^{1/2}W_m(I+\alpha^2A^TA)^{1/2},\nonumber \\
-  M^{-1}
-            &=&(I+\alpha^2A^TA)^{-1/2}W_m^{-1}(I+\alpha^2A^TA)^{-1/2}.
-                \label{eqn:defprecond}
-\end{eqnarray}
-With this choice, \ref{eqn:normapprox} implies that
-$M^{-1}N \approx I$, also $M$ is symmetric. As already mentioned,
-powers of $I + \alpha^2A^TA$ are trivial to compute, given the choice
-of $A$ made here. We will examine fast algorithms for computing
-$W_m^{-1}$ = the symmetrized pressure-to-source operator in the next
-section. Note that only $M^{-1}$, hence only $W_m^{-1}$, appears in
-Algorithm 1.
-
-[Example]
-
-\section{Computing and Symmetrizing $\Lambda$}
-
-Computations of $\Lambda^{\pm}_{z_s}$ and its transpose are clearly critical steps in an
-implementation of the PCG algorithm outlined in the preceding section.
-Direct computation of the pressure-to-source operator $\Lambda^{\pm}_{z_s}$, for instance by solving
-\ref{eqn:awepm} and reading off $P_sv^{\pm}_{z}$, turns out to be
-numerically ill-behaved. The relation \ref{eqn:snull} provides and
-alternative approach, taking advantage of the accurate approximate
-inverse to $S^+_{z_s,z_r}$ constructed above. The first row of
-\ref{eqn:snull}, slightly rearranged, is
-\begin{equation}
-  \label{eqn:lamidea}
-  \Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx  -\frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
-\end{equation}
-The approximate inverse construction for $S^++_{z_s,z_r}$ permits (approximate)
-solution of this equation for $\Lambda^+_{z_s}f_s$: apply
-$4(V^{+}_{z_s,z_r})^T$ to both sides of equation \ref{eqn:lamidea} and
-use the first equation in the list \ref{eqn:approxinv} to get
-\begin{equation}
-  \label{eqn:lamident}
-  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
-\end{equation}
-This identity is the major result of this section: it shows how
-to compute that action of $\Lambda^+_{z_s}$ by propagating the input
-pressure trace, identified as a source for the velocity evolution,
-forward in time from $z_s$ to $z_r$
-reading off the pressure trace on $z=z_r$, identifying it once more as
-a point load (source for velocity), propagating it backwards in time from $z_r$ to
-$z_s$, and finally reading off the velocity trace, interpreted as a
-pressure evolution source on $z_s$.
-
-The importance of this result lies in the failure of the obvious
-method for computing the action of $\Lambda^{\pm}_{z_s}$, namely to
-employ the pressure trace as a source in the velocity equation ($f_s$,
-in the notation used above) at $z=z_s$, and read off the velocity
-field also at $z=z_s$. This difficulty is related to the existence of
-tangentially propagating waves and the lack of continuity of the trace
-operator. The method implicit in equation \ref{eqn:lamident} avoids
-this difficulty by propagating the fields a positive distance in $z$:
-assuming as always that the causal fields are downgoing, this step
-eliminates any tangentially propagating fields from consideration.
-
-Figure \ref{fig:preddinvhshh0} shows the pressure source gather
-produced from the pressure data gather in Figure \ref{fig:dsrcphh0} by
-application of the operator on the right-hand side of formula
-\ref{eqn:lamident} to the source surface pressure gather depicted in
-\ref{fig:dsrcphh0}. The homogeneous model is used in all propagations
-implicit in the prescription \ref{eqn:lamident}. The scale is the same
-as for Figures \ref{fig:dinvhshh0} and \ref{fig:dhshh0}. In fact
-Figure \ref{fig:preddinvhshh0} appears to be nearly identical to
-Figure \ref{fig:dinvhshh0}, and both appear to be dip-filtered
-versions of Figure \ref{fig:dhshh0}. The difference plot (Figure
-  \ref{fig:ddiffinvhshh0}), using the same color scale, shows that the
-  similarity is quantitative. The recovered source gather (Figure
-  \ref{fig:preddinvhshh0}) also generates the same acoustic fields as
-  the point source at $z=z_d$: the pressure gather at $z=z_r=1000$ m
-  depth is shown in Figure \ref{fig:dpredhsrecphh0}, and its
-  difference with the point source gather (Figure \ref{fig:drecphh0})
-  plotted on the same color scale in Figure
-  \ref{fig:ddiffpredhsrecphh0}.
-
-Figures \ref{fig:preddinvhsll0}, \ref{fig:ddiffinvhsll0},
-\ref{fig:dpredhsrecpll0}, and \ref{fig:ddiffpredhsrecpll0} accomplish
-the same comparison for the $\Lambda^+_{z_s}$ approximation in
-equation \ref{eqn:lamident}, with all propagation taking place in the
-lens model.
-
-A deeper study of the pressure-to-source
-operator (Appendix B) shows that it is approximately dependent only on the model
-coefficients near the source surface ($z=z_s$ in this case). Since the
-homogeneous and lens models are identical near this surface, it is
-unsurprising that these figures are very close to the previous two.
-However an even more useful observation is that the calculations in
-the approximation \ref{eqn:lamident} could just as well be carried out
-in a much smaller region around the source surface, and produce a
-result that is functionally identical in that it will serve as a
-source for the same acoustic fields globally, with small error. In
-effect, equation \ref{eqn:lamident} involving propagation from source
-($z=z_s$) to receiver ($z=z_r$) surfaces is altered by replacing $z_r$
-with a receiver datum $z_s+\Delta z$ considerably closers to $z_s$:
-\begin{equation}
-  \label{eqn:lamnear}
-  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
-\end{equation}
-Using a receiver datum closer to the source surface has two favorable consequences:
-\begin{itemize}
-\item The computational domain can be smaller than is necessary to
-  simulate the target data, as it need only contain the source
-  surface and the receiver datum implicit in
-  equation \ref{eqn:lamident}. This shrinkage of the computational
-  domain can lead to substantial improvements in computational
-  efficiency.
-\item Since the receiver data may be chosen much closer to the
-  source surface that is the case for the target data, the effective
-  aperture active in the relation \ref{eqn:lamident} can be much
-  larger, producing an estimated source gather much less affected by
-  aperture limitation.
-\end{itemize}
-
-To illustrate these points, I have re-computed the source gather
-estimates shown in Figure \ref{fig:preddinvhshh0}. In place of the receiver geometry in the
-target data (Figure \ref{fig:drecphh0}), I
-place a receiver array at $z_s+\Delta z=2900$ m depth, just 100 m above the
-source surface at $z_s=3000$ m. For the discretization used to create
-the examples shown so far, that is just a 5 gridpoint difference in
-depth, as opposed to 100 gridpoints between the source and receiver
-depths for examples such as shown in Figure \ref{fig:recphh0} or
-\ref{fig:drecphh0}.
-
-Note that the homogeneous and lens models are identical in the depth
-range $2900 < z < 3000$ m, so the computed $\Lambda^+_{z_s}$ operators
-will be the same for both models. Hence I show only results for the
-homogenous model.
-
-The approximation to $\Lambda_{z_s}^+$ via equation \ref{eqn:lamident}
-for this configuration is evaluated in the same way as before in
-Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
-\ref{fig:dprednshsrecphh0}, and \ref{fig:ddiffprednshsrecphh0}. The effect of aperture
-limitation is clearly diminished: the second figure in this series
-compares the full-aperture pressure source gather (Figure
-\ref{fig:dhshh0}) with the 
-image of the corresponding pressure gather (Figure \ref{fig:dsrcphh0})
-under the approximation to $\Lambda_{z_s}^+$, and the last figures
-show that the approximated source gathers accurately predict the
-point-source pressure gather at the receiver datum $z_r=1000$ m.
-
-As mentioned in the last section, computation of the transpose of
-$\Lambda^+$ (exact, not approximate in the high frequency sense) is
-critical to the successful construction of the preconditioner. The
-relation \ref{eqn:lamident} does not provide a computation for this
-operator. However set
-\begin{equation}
-  \label{eqn:lamtilde}
-  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
-\end{equation}
-Then \ref{eqn:lamnear} can be rewritten
-\[
-  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
-\]
-Of course, all of the examples so far show images of
-$\tilde{\Lambda}^+_{z_s}$.
-
-Since successful preconditioning requires only approximate inversion,
-use of $\tilde{\Lambda}^+_{z_s} $ in place of $\Lambda^+_{z_s}$ will
-still yield a working preconditioner, and the former can be transposed
-to machine precision via the definition \ref{eqn:lamtilde} and the adjoint state
-method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
-\begin{equation}
-  \label{eqn:lamtransp}
-  (\tilde{\Lambda}^+_{z_s})^T = -8 \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
-    z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}
-\end{equation}
-
-Figure \ref{fig:preddnshstrhh0} shows the image of the pressure gather
-in Figure \ref{fig:dsrcphh0} under $(\tilde{\Lambda}^+_{z_s})^T$,
-using the ``near'' traces at $z=2900$, that is, $\Delta z = 100$ m in
-expression \ref{eqn:lamtransp}, and propagation in the
-homogeneous model. Note
-the close resemblance to the image of the same pressure gather under
-$\tilde{\Lambda}^+_{z_s}$ displayed in Figure
-\ref{fig:preddnshshh0}. The difference of these two images is
-displayed in \ref{fig:ddiffnslamtrhh0}, on the same color scale as the
-images themselves. Since the propagation takes place entirely in a
-region where all of the mechanical parameters are homogenous, I do not
-offer a similar comparison for the lens model.
-
-The model space weight operator $W_m$ introduced in the last section is
-replaced by its asymptotic approximation
-\[
-  \tilde{W}_m = \frac{1}{2}(\tilde{\Lambda}^+_{z_s} +
-  (\tilde{\Lambda}^+_{z_s})^T)
-\]
-\[
-  = -8 \left((V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T \right.+
-\]
-\[
-  \left. \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
-      z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}\right)
-\]
-\begin{equation}
-  \label{eqn:wcomp}
-  = -4(\Pi_1 ({\cal S}^+_{z_s,z_s+\Delta z})^T(\Pi_0^T\Pi_1 + \Pi_1^T\Pi_0) {\cal S}^+_{z_s,z_s+\Delta z}\Pi_1^T)
-\end{equation}
-with a similar definition for the replacement $\tilde{W}_d$ of $W_d$.
-
-This identity shows that only one forward and one adjoint simulation
-are necessary to compute the action of $\tilde{W}_{m,d}$. The operator
-in the center of the expression on the right-hand side, $\Pi_0^T\Pi_1
-+ \Pi_1^T\Pi_0$, simply exchanges the components of the acoustic
-fields, passing the velocity field as a pressure source and the
-pressure field as a velocity source.
-
-Figure \ref{fig:symmdnshshh0} shows the output of the symmetrized
-approximate source-to-pressure operator per equation \ref{eqn:wcomp},
-applied once again to the pressure data in Figure
-\ref{fig:dsrcphh0}. Note the resemblance to Figures
-\ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}. These are all
-asymptotic approximations of each other.
-
- \newpage
-
-\section{Figures}
-
-\plot{bml0}{width=\textwidth}{Bulk modulus, lens model. Color scale is 
-in GPa. Positions of point source and receiver line indicated.}
-
-\plot{recphh0}{width=\textwidth}{Output pressure gather for point source in 
-  homogeneous medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-\plot{recplh0}{width=\textwidth}{Output pressure gather for point source in 
-  refractive (lens) medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-\plot{recvzhh0}{width=\textwidth}{Output vertical velocity gather for point source in 
-  homogeneous medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-\plot{recvzlh0}{width=\textwidth}{Output vertical velocity gather for point source in 
-  refractive (lens) medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-\plot{dsrcphh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
-  pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
-  bandpass filter source. }
-
-\plot{dsrcvzhh0}{width=\textwidth}{Trace $P_sv_z^+$ on $z=z_s=3000$ m of
-  vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
-  bandpass filter source.}
-
-\plot{drecphh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the homogeneous model described in the 
-  text, with point pressure source (constitutive defect) at $z_d=3500$
-  m, $x_d=3500$ m.}
-
-\plot{dfwdphh0}{width=\textwidth}{Pressure gather at receiver depth
-  $z_r=1000$ m from field generated by causal solution of acoustic
-  system \ref{eqn:awepm} in the homogeneous model described in the
-  text, with extended pressure source (constitutive defect)
-  on $z=z_s=3000$ m given by $h_s=-2P_sv_z^+=\Lambda^+_{z_s}P_sp^+$,
-  where $P_sp^+$ is shown in Figure \ref{fig:dsrcphh0} and $P_sv^+_z$
-  in Figure \ref{fig:dsrcvzhh0}. In this simulation, the velocity
-  velocity source (vertical load) vanishes: $f_s=0$.}
-
-\plot{daltphh0}{width=\textwidth}{Pressure gather at receiver depth
-  $z_r=1000$ m from field generated by causal solution of acoustic
-  system \ref{eqn:awepm} in the homogeneous model described in the
-  text, with extended velocity source (vertical load) $f_s=-2P_sp^+$
-  on $z=z_s=3000$ m, where $P_sp^+$ is shown in Figure
-  \ref{fig:dsrcphh0}.  In this simulation, the zero pressure source
-  (constitutive defect vanishes: $h_s=0$.}
-
-\plot{drecplh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the lens model described in the 
-  text, with point pressure source (constitutive defect) at $z_d=3500$
-  m, $x_d=3500$ m.}
-
-\plot{drecphh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$) 
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
-  (red).}
-
-\plot{daltphh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$) 
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:daltphh0}
-  (red).}
-
-\plot{drecphh0tr51}{width=\textwidth}{Overplot of traces 51 ($x=3000$) 
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
-  (red).}
-
-\plot{daltphh0tr51}{width=\textwidth}{Overplot of traces 51 ($x=3000$) 
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:daltphh0}
-  (red).}
-
-\plot{drecphh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$) 
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
-  (red).}
-
-\plot{daltphh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$) 
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:daltphh0}
-  (red).}
-
-\plot{dfwdplh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the lens model described in the 
-  text, with extended pressure source (constitutive defect) on
-  $z=z_s=3000$ m given bythe field depicted in Figure
-  \ref{fig:dsrcvzhh0} scaled by -2 
-   ($h_s=-2P_sv_z^+=\Lambda^+_{z_s}P_sp^+$) and zero velocity source (vertical load) 
-  ($f_s=0$).}
-
-\plot{daltplh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the lens model described in the 
-  text, with extended velocity source (vertical load) on $z=z_s=3000$
-  m given by the field depicted in Figure \ref{fig:dsrcphh0} scaled by -2
-  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
-  ($h_s=0$).}
-
-\plot{drecplh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$) 
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:dfwdplh0}
-  (red).}
-
-\plot{daltplh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$) 
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
-  (red).}
-
-\plot{drecplh0tr51}{width=\textwidth}{Overplot of traces 51 ($x=3000$) 
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:dfwdplh0}
-  (red).}
-
-\plot{daltplh0tr51}{width=\textwidth}{Overplot of traces 51 ($x=3000$) 
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
-  (red).}
-
-\plot{drecplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$) 
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:dfwdplh0}
-  (red).}
-
-\plot{daltplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$)
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
-  (red).}
-
-\plot{dhshh0}{width=\textwidth}{Pressure source gather = -2 $\times$
-  vertical velocity gather (Figure \ref{fig:dsrcvzhh0}).}
-
-\plot{dfshh0}{width=\textwidth}{Velocity source gather = -2 $\times$
-  pressure gather (Figure \ref{fig:dsrcphh0}).}
-
-\plot{dinvhshh0}{width=\textwidth}{Approximate inversion via first 
-  equation in display \ref{eqn:approxinv}. Inversion in homogenous model of 
-  the pressure gather in Figure \ref{fig:drecphh0}, simulated with 
-  homogeneous model. Scaled version of output $v_z$ field obtained by 
-  applying transpose of $4V^+_{z_s,z_r}$. Approximates dip-filtered
-  version of pressure source gather (Figure \ref{fig:dhshh0}).}
-
-\plot{drerecphh0}{width=\textwidth}{Re-simulated pressure gather
-  produced from inverted source
-  shown in Figure \ref{fig:dinvhshh0}. Simulation in homogeneous model.}
-
-\plot{ddiffrecphh0}{width=\textwidth}{Difference between gathers
-  displayed in Figures \ref{fig:drecphh0} and \ref{fig:drerecphh0},
-  plotted on same color scale.}
-
-\plot{dinvhsll0}{width=\textwidth}{Approximate inversion via first 
-  equation in display \ref{eqn:approxinv}. Inversion in lens model of 
-  the pressure gather in Figure \ref{fig:drecplh0}, simulated with 
-  lens model. Scaled version of output $v_z$ field obtained by 
-  applying transpose of $4V^+_{z_s,z_r}$. Approximates dip-filtered
-  version of pressure source gather (Figure \ref{fig:dhshh0}).}
-
-\plot{drerecpll0}{width=\textwidth}{Re-simulated pressure gather
-  produced from inverted source
-  shown in Figure \ref{fig:dinvhsll0}. Simulation in lens model.}
-
-\plot{ddiffrecpll0}{width=\textwidth}{Difference between gathers
-  displayed in Figures \ref{fig:drecpll0} and \ref{fig:drerecpll0},
-  plotted on same color scale.}
-
-\plot{dinvhslh0}{width=\textwidth}{Approximate inversion via first 
-  equation in display \ref{eqn:approxinv}. Inversion in homogenous model of 
-  the pressure gather in Figure \ref{fig:drecplh0}, simulated with 
-  lens model. Scaled version of output $v_z$ field obtained by 
-  applying transpose of $4V^+_{z_s,z_r}$. Approximates dip-filtered
-  version of pressure source gather (Figure \ref{fig:dhshh0})}
-
-\plot{drerecplh0}{width=\textwidth}{Re-simulated pressure gather produced from
-  inverted source
-  shown in Figure \ref{fig:dinvhslh0}. Simulation in lens model.}
-
-\plot{ddiffrecplh0}{width=\textwidth}{Difference between gathers
-  displayed in Figures \ref{fig:drecplh0} and \ref{fig:drerecplh0},
-  plotted on same color scale.}
-
-\plot{dfwdvzhh0}{width=\textwidth}{Vertical velocity gather, generated
-  with a pressure source in the homogeneous model, 
-  corresponding to pressure gather \ref{fig:dfwdphh0}.}
-
-\plot{daltvzhh0}{width=\textwidth}{Vertical velocity gather, generated
-  with a velocity source in the homogenous model, corresponding to
-  pressure gather \ref{fig:daltphh0}.}
-
-\plot{dsvcomphh0}{width=\textwidth}{Difference between velocity gathers
-  shown in Figures \ref{fig:dfwdvzhh0} and \ref{fig:daltvzhh0},
-  plotted on the same color scale as these figures.}
-
-\plot{lamsstlamrdrecphh0}{width=\textwidth}{Inversion of data shown in
-  Figure \ref{fig:drecphh0}, simulated in homogeneous model, using the
-  approximate unitarity relation \ref{eqn:unitary} and propagation in
-  homogenous model.}
-
-\plot{difflamsstlamrdrecphh0}{width=\textwidth}{Difference between
-  data displayed in Figures \ref{fig:dinvhshh0} and
-  \ref{fig:lamsstlamrdrecphh0}, plotted on the same color scale.}
-
-\plot{lamsstlamrdrecplh0}{width=\textwidth}{Inversion of data shown in
-  Figure \ref{fig:drecplh0}, simulated in lens model, using the
-  approximate unitarity relation \ref{eqn:unitary} and propagation in
-  homogenous model.}
-
-\plot{difflamsstlamrdrecplh0}{width=\textwidth}{Difference between
-  data displayed in Figures \ref{fig:dinvhslh0} and
-  \ref{fig:lamsstlamrdrecplh0}, plotted on the same color scale.}
-
-\plot{preddinvhshh0}{width=\textwidth}{Pressure source gather = image
-  of pressure-to-source operator $\Lambda^+_{z_s}$ applied to pressure gather
-  shown in Figure \ref{fig:dsrcphh0}, homogeneous model. Compare Figure
-  \ref{fig:dinvhshh0}.}
-
-\plot{ddiffinvhshh0}{width=\textwidth}{Difference between (a) image
-  (Figure \ref{fig:preddinvhshh0}) of $\Lambda^+_{z_s}$ applied to
-  pressure gather (Figure \ref{fig:dsrcphh0}), and (b) source gather
-  (Figure \ref{fig:dinvhshh0}) inverted from receiver pressure gather
-  (Figure \ref{fig:drecphh0}). Homogeneous model used in all
-  propagations. Same color scale as in Figure
-  \ref{fig:preddinvhshh0}. }
-
-\plot{dpredhsrecphh0}{width=\textwidth}{Pressure gather at receiver 
-  datum $z=z_r=1000$ m simulated in homogeneous model from source 
-  gather shown in Figure \ref{fig:preddinvhshh0}, also based on the 
-  homogeneous model. Compare with point source pressure gather (Figure 
-  \ref{fig:drecphh0}).}
-
-\clearpage
-
-\plot{ddiffpredhsrecphh0}{width=\textwidth}{Plot of difference 
-  between data shown in Figures \ref{fig:drecphh0} and 
-  \ref{fig:dpredhsrecphh0}, plotted on same color scale as the latter 
-  two figures.}
-
-\plot{preddinvhsll0}{width=\textwidth}{Pressure source gather = image
-  of pressure-to-source operator $\Lambda^+_{z_s}$ applied to pressure gather
-  shown in Figure \ref{fig:dsrcphh0}, lens model. Compare Figure
-  \ref{fig:dinvhsll0}.}
-
-\plot{ddiffinvhsll0}{width=\textwidth}{Difference between (a) image
-  (Figure \ref{fig:preddinvhsll0}) of $\Lambda^+_{z_s}$ applied to
-  pressure gather (Figure \ref{fig:dsrcphh0}), and (b) source gather
-  (Figure \ref{fig:dinvhsll0}) inverted from receiver pressure gather
-  (Figure \ref{fig:drecplh0}). Lens model used in all propagations.
-  Same color scale as in Figure \ref{fig:preddinvhsll0}.}
-
-\plot{dpredhsrecpll0}{width=\textwidth}{Pressure gather at receiver 
-  datum $z=z_r=1000$ m simulated in lens model from source 
-  gather shown in Figure \ref{fig:preddinvhsll0}, also based on the 
-  homogeneous model. Compare with point source pressure gather (Figure 
-  \ref{fig:drecplh0}).}
-
-\plot{ddiffpredhsrecpll0}{width=\textwidth}{Plot of difference 
-  between data shown in Figures \ref{fig:drecplh0} and 
-  \ref{fig:dpredhsrecpll0}, plotted on same color scale as the latter 
-  two figures.}
-
-\clearpage
-
-\plot{preddnshshh0}{width=\textwidth}{Pressure source gather = image
-  under pressure-to-source operator $\Lambda^+_{z_s}$ of pressure gather
-  shown in Figure \ref{fig:dsrcphh0}, homogeneous model, using
-  ``near'' receiver traces at $z=2900$ m. Compare Figure
-  \ref{fig:dhshh0} and \ref{fig:dinvhshh0}: because the sources and
-  receivers are close, little aperture is lost in this case.}
-
-\plot{ddiffnshshh0}{width=\textwidth}{Difference between (a) image
-  (Figure \ref{fig:preddnshshh0}) of $\Lambda^+_{z_s}$ applied to
-  pressure gather (Figure \ref{fig:dsrcphh0}) using a near receiver
-  array to implement formula \ref{eqn:lamident}, and (b) source gather
-1650  (Figure \ref{fig:dhshh0}) inferred from vertical velocity.
-  Homogeneous model used in all
-  propagations. Same color scale as in Figure
-  \ref{fig:preddnshshh0}. }
-
-\plot{dprednshsrecphh0}{width=\textwidth}{Pressure gather at receiver
-  datum $z=z_r=1000$ m simulated from source gather shown in Figure
-  \ref{fig:preddnshshh0}. Compare with point source pressure gather
-  (Figure \ref{fig:drecplh0}).}
-
-\clearpage
-
-\plot{ddiffprednshsrecphh0}{width=\textwidth}{Plot of difference 
-  between data shown in Figures \ref{fig:drecphh0} and 
-  \ref{fig:dpredhsrecphh0}, plotted on same color scale as the latter 
-  two figures.}
-
-\plot{preddnshstrhh0}{width=\textwidth}{Pressure source gather = image
-  under {\em transpose} of pressure-to-source operator
-  $\Lambda^+_{z_s}$ of pressure gather shown in Figure
-  \ref{fig:dsrcphh0}, homogeneous model, using ``near'' receiver
-  traces at $z=2900$ m. Compare Figure \ref{fig:dhshh0} and
-  \ref{fig:preddnshshh0}: as noted in the text, $\Lambda^+_{z_s}$ is
-  asymptotically symmetric, so the resemblance is not a surprise.}
-
-\plot{ddiffnslamtrhh0}{width=\textwidth}{Difference between data in
-  Figures \ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}, plotted
-  on the same scale as these figures, showing that the asymptotic
-  symmetry of $\Lambda^+_{z_s}$ is actually quantitative for the
-  length, time and frequency scales of these examples.}
-
-\plot{symmdnshshh0}{width=\textwidth}{Pressure source gather = image
-    under {\em symmetrized} pressure-to-source operator
-    $\frac{1}{2}\left(\Lambda^+_{z_s}+(\Lambda^+_{z_s})^T\right)$ of
-    pressure gather shown in Figure \ref{fig:dsrcphh0}, homogeneous
-    model, using ``near'' receiver traces at $z=2900$ m. Compare
-    Figure \ref{fig:preddnshshh0}.}
-
-\plot{compreshh0}{width=\textwidth}{Comparison of residual Euclidean 
-  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
-  homogeneous model, point source (Figure \ref{fig:recphh0}), inversion 
-  in homogeneous model.}
-
-\plot{compreslh0}{width=\textwidth}{Comparison of residual Euclidean 
-  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
-  lens model, point source (Figure \ref{fig:recplh0}), inversion 
-  in homogenous model.}
-
-\plot{compresll0}{width=\textwidth}{Comparison of residual Euclidean 
-  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
-  lens model, point source (Figure \ref{fig:recplh0}), inversion 
-  in lens model.}
-
-\plot{compnreshh0}{width=\textwidth}{Comparison of normal residual
-  (gradient) Euclidean norms: CG (blue), PCG (red), plotted
-  vs. iteration. Data = homogeneous model, point source (Figure
-  \ref{fig:recphh0}), inversion in homogenous model.}
-
-\plot{compnreslh0}{width=\textwidth}{Comparison of normal residual 
-  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
-  vs. iteration. Data = lens model, point source (Figure 
-  \ref{fig:recplh0}), inversion in homogenous model.}
-
-\plot{compnresll0}{width=\textwidth}{Comparison of normal residual 
-  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
-  vs. iteration. Data = lens model, point source (Figure 
-  \ref{fig:recplh0}), inversion in lens model.}
-
-\newpage
-
-\append{Adjoint Computation}
-
-The adjoint of ${\cal S}^+_{z_s,z_r}$ can be computed by a variant of
-the adjoint state method, in this case a by-product of the
-conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awepm}
-with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
-$ (h_r,f_r\bf{e}_z) \delta(z-z_r)$. Then
-\[
-0 = 
-\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
-\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
--
-\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
-\]
-\[
-+
-\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
- h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
- h_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  (- \nabla p^++f_s\bf{e}_z) \cdot \bv^- + \bv^+ \cdot (-\nabla
-  p^- + f_r \bf{e_z}) \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
- h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
- h_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- 
-  +f_s \delta(z-z_s) v_z^- + v_z^+f_r \delta(z-z_r) \right)
-\]
-after integration by parts in the last two terms. Most of what is left cancels, leaving 
-\[
-0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
-( h_rP_rp^++f_rP_rv_z^+) = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
-\]
-whence
-\begin{equation}
-\label{eqn:sadj}
- ({\cal S}^+_{z_s,z_r})^T = -{\cal S}^{-}_{z_r,z_s}.
-\end{equation}
-
-The systems \ref{eqn:awepm} for $(p^{\pm},\bv^{\pm})$ with the two
-signs of superscript differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
-$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
-\begin{equation}
-  \label{eqn:trdef}
-  {\cal R} \left(
-    \begin{array}{c}
-      p\\
-      \bv
-    \end{array}
-  \right) =
-  \left(
-    \begin{array}{c}
-      Rp\\
-      -R\bv
-    \end{array}
-  \right)
-\end{equation}
-Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awepm} with
-$(h_s\delta(z-z_s),f_s\bf{e_z}\delta(z-z_s))$ replaced by
-$(-Rh_r\delta(z-z_r), Rf_r\bf{e_z}\delta(z-z_r))$. That is,
-\begin{equation}
-  \label{eqn:trsadj}
-  {\cal R}{\cal S}^- = -{\cal S}^+_{z_r,z_s}{\cal R}
-\end{equation}
-Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj}
-and \ref{eqn:trsadj} imply equation \ref{eqn:trtr}.
-
 \bibliographystyle{seg}

### 2022-03-28 repro/papers/GEM22/paper.tex
first commit

--- 2021-07-13 repro/papers/sseprecond/paper.tex
@@
 
+\inputdir{../sseprecond/project}
+
 \section{Introduction}
@@
 constraining these additional degrees of freedom by means of an
-additive penalty modifying the probem \ref{eqn:ols}:
+additive quadratic penalty modifying the problem \ref{eqn:ols}:
 \begin{equation}
@@
 \ref{eqn:esis}. This approximate solution operator may be used to
-accelerate Krylov space methods for the solution of the source
+accelerate Krylov space methods for the solution of the surface source
 subproblem \ref{eqn:esis}. Numerical examples suggest the
@@
 acoustodynamics, the spatial positions of traces extracted by $P$ lie
-on a plane, and the positions at which the extended
-source $\bff$ is nonzero lie on another, parallel, plane. The data and
-sources are further limited to pressure traces and constitutive law
-defects (``pressure sources''). While this
+on a depth plane $z=z_r$, and the positions at which the extended
+source $\bff$ is nonzero lie on another, parallel, depth plane $z=z_s$. 
+%That is, sources take the form of a combination of constitutive law
+%defect and normal force,
+%\begin{equation}
+%  \label{eqn:surfsrc}
+%  \bff(x,y,z,t) = h_s(x,y,t)\delta(z-z_s)(1,0,0,0)^T + f_s(x,y,z,t)(0,0,0,1)^T
+%\end{equation}
+%and the sampling operator $P$ is defined by
+%\begin{equation}
+%  \label{eqn:surfsam}
+%  P\bu(x,y,t) =
+%\end{equation}
+%supported on $z=z_s$, and similar
+%surface vertical loads.
+This
 ``crosswell'' configuration simplifies the analysis underlying the
 construction of approximate solutions for the source subproblem
-\ref{eqn:esis}, it is only one of many transmission configurations for
+\ref{eqn:esis}. It is only one of many transmission configurations for
 which similar developments are possible. Perhaps the most important
 alternative example is the diving wave configuration, which plays a
-central role in contemporary FWI. I will discuss the generalization of
-surface source extention to diving wave inversion at the end of the
-paper.
+central role in contemporary FWI.
 
@@
 map pressure to corresponding surface source on the source and
-receiver planes. These are essentially the same as the ``hyperbolic
-Dirichlet-to-Neumann'' map that plays a prominent role in
-thermoacoustic tomography and other wave inverse problems
+receiver planes. This pressure-to-source map is closely related to the ``hyperbolic
+Dirichlet-to-Neumann'' operator that plays a prominent role in
+photoacoustic tomography and other wave inverse problems
 \cite[]{Rachele:00,StefUhl:05}. \cite{HouSymes:EAGE16} demonstrated a
@@
 
-The obvious computation of the pressure-to-source map
-- prescribe the pressure, solve the wave equation with this boundary
-condition, read off the equivalent source - suffers from intrinsic
-numerical inaccuracy. I suggest an alternative computationally
-feasible approach.
+The obvious computation of the pressure-to-source map - prescribe the
+pressure, solve the wave equation with this boundary condition, read
+off the equivalent source - suffers from intrinsic numerical
+inaccuracy. I suggest an alternative computationally feasible
+approach, via economical short-distance wave propagation. Since the
+map is symmetric only in an approximate, asymptotic sense, it must be
+symmetrized for use as a Krylov preconditioner. I describe a
+symmetrization procedure that requires no further wave computations
+beyond those necessary to compute the action of the operator itself.
+
+The discussion in this paper is formal and incomplete, in the sense
+that some important mathematical underpinnings are only
+sketched. I will treat the modeling operator $PL[{\bf c}]^{-1}$ as if
+it mapped square integrable surface sources to square integrable sampled
+data. This is not true in full generality: while the surface source
+problem has distribution solutions, they are not generally square
+integrable (finite acoustic field energy). Even if the solutions have
+finite energy, they do not in general have well-defined
+restrictions to lower-dimensional sets. In other words, the action of
+the sampling
+operator $P$ on the space-time plane $z=z_r$ is not well-defined for arbitrary
+finite-energy acoustic fields. Thus the modeling operator envisioned above
+may not be well-defined.
+
+This
+phenomenon is related to the ill-posedness of wave equations as
+evolution equations in spatial variables, an observation attributed to
+Hadamard (see \cite{CourHil:62}, Chapter 6, section 17). Some constraint on the acoustic field,
+beyond finite energy, is mandatory in any precise mathematical formulation the inverse problems
+\ref{eqn:esi} and \ref{eqn:esis}. In fact, the natural constraint in
+the ``crosswell'' geometry of this paper is that high-frequency energy
+travel {\em only} along rays crossing the
+surfaces $z=z_s, z=z_r$ transversally. That is, source functions on
+$z=z_s$ generate waves with energy traveling along rays leaving the surface at a non-zero angle, and
+energy arrives at the recording surface $z=z_r$ along rays making
+non-zero angle with it. I will call sources, sampled data, and
+acoustic fields with this property {\em
+  downgoing} (even though the concept also encompasses {\em upcoming}
+propagation). Note that the downgoing property restricts the behaviour of
+acoustic fields near the source and receiver surfaces - what the
+fields do elsewhere is their own business.
+
+Several works have explored the mathematics of the
+downgoing condition and its consequences in the context of the scalar second order wave equation, see for instance
+\cite{Payn:75,Symes:83,Lasi:86,LasLionsTrig:86,Lasi:87, BaoSy:91b}. 
+Since the purpose of this paper is to explore the algorithmic
+consequences of the mathematical structure implied by the downgoing
+condition, I leave the elaboration of the mathematical details to
+another place.
+
+The next section defines the modeling operator $PL[{\bf c}]$, its
+adjoint, and important specializations (pressure vs. normal velocity
+sources and data). The sections to follow define the
+source-to-pressure operator, construct an approximate inverse of the
+modeling operator by time reversal (as suggested by work in
+photoacoustic tomography), use the source-to-pressure operator to
+express the approximate inverse as the modeling operator adjoint in
+weighted norms (thus establishing that the modeling operator is {\em
+  unitary} in the sense of these norms), explain how to use this
+construction to precondition Conjugate Gradient iteration, and
+organize the preconditioning computation so as to involve only one
+wave propagation calculation. The penultimate section displays simple
+2D numerical examples of all of the key steps, culminating in a
+comparison of straight vs. preconditioned CG iteration. The paper ends
+with a brief discussion-and-conclusion section, reviewing what has
+been accomplished and listing a few of the many questions left open.
 
@@
 
-A technical difficulty with spatial sampling must be
-addressed. Acoustic field energy is defined in terms of the mean
-squares of the pressure and particle velocity fields. However fields
-with finite mean square do not in general have well-defined
-restrictions to lower-dimensional sets: for example, the pressure
-field of a finite-energy acoustic vector field may not have a
-well-defined restriction to the space-time plane $z=z_r$. This
-phenomenon is related to the ill-posedness of wave equations as
-evolution equations in spatial variables, an observation attributed to
-Hadamard (see \cite{CourHil:62}, Chapter 6, section 17, also
-\cite{Payn:75,Symes:83}). Some constraint on the acoustic field,
-beyond finite energy, is mandatory in formulating the inverse problems
-\ref{eqn:esi} and \ref{eqn:esis}. In fact, the natural constraint in
-the ``crosswell'' geometry of this paper is that high-frequency energy
-{\em not} travel along rays parallel or nearly parallel to the
-surfaces $z=z_s, z=z_r$. I will call fields with this property {\em
-  downgoing} (even though the concept also encompasses {\em upcoming}
-fields). \cite{BaoSy:91b} give a mathematically complete discussion of
-downgoing wave field properties.
-
 For downgoing solutions of system \ref{eqn:awepm}, the key
@@
 
-\section{2D Examples}
-
-\inputdir{project}
-
-To illustrate the structure described in the preceding section, I
-introduce two 2D acoustic models, one spatially homogeneous, the other
-with highly refractive. The first, homogenous model has $\kappa = 4$
-GPa and $\rho = 1$ g/cm$^3$ throughout a rectangular domain of size 8 km ($x$) $\times$ 4 km
-($z$). The second, refractive, model is a perturbation of the first by
-a low-velocity acoustic lens positioned in the center of the
-rectangle. To produce this structure, the density is chosen
-homogeneous as in the first model, while the bulk modulus decreases to
-from 4 GPa outside the lens to 1.6 GPA in its center, as shown in Figure \ref{fig:bml0}.
-
-Discretization is conventional, with a rectangular grid and staggered
-finite difference scheme \cite[]{Vir:84} of order 2 in time and
-2$k$ in space; for most of the experiments reported below, $k=4$.
-Sampling operators such as $P_r$ are implemented via linear
-interpolation, and source insertion via adjoint linear interpolation
-(as noted above, in the continuum limit, sources are represented via
-adjoint sampling). Steps in $x$ and $z$ are the same. Most results use
-$\Delta x = 20$ m, and limit the temporal frequency of the computed
-traces to 12.5 Hz. We also explore the dependence of a few results on frequency, using
-$\Delta x = 10$ m and $5$ m, to accomodate 25 and 50 Hz respectively,
-maintaining 8 samples per wavelength. All computations are carried out
-in single precision.
-
-\cite{GeoPros:11} gives a description of the code
-implementation, out-of-date in a few respects but overall accurate.
-
-The horizontal line of receivers sits at depth $z_r = $ 1000 m, as
-shown in Figure \ref{fig:bml0}. Receiver $x$ ranges from $2000$ to
-$6000$ m. A single point (physical) source appears in these
-experiments, located at $x_s=3500$ m, $z_s=3000$ m, as also indicated
-in the figure. The source time function is a bandpass filter with
-corner frequencies $1, 2.5, 7.5, 12.5$ Hz for the lowest frequency
-source, and scaled as appropriate for examples with higher frequency
-and finer sampling.  Extended sources are confined to the horizontal
-line through the physical source position, that is $z_s = 3000$ m,
-over a 4 km interval starting at $x_r=$ 2000 m. Note that we have
-reversed the order relation between $z_s$ and $z_r$ described in the
-text ($z_s<z_r$). This difference is immaterial.
-
-The point source pressure data generated by this configuration for the
-homogeneous model is displayed in Figure \ref{fig:recphh0}, for the
-lens model in \ref{fig:recplh0}. Triplication of arrivals is evident in the latter.
-
-While inversion of pressure data is the main object of this exercise,
-normal velocity ($v_z$) data will play an important role, so I display
-the corresponding gathers in Figures \ref{fig:recvzhh0} and \ref{fig:recvzlh0}.
-
-\section{Adjoints}
-
 It follows from the adjoint state method (see Appendix A for details) that
@@
 
-The implemented code uses the discrete adjoint state method and auto-generated code \cite[]{TapenadeRef13}, to
-assure that the computed adjoint operators are adjoint at the level of
-machine precision to the computed operators. The reverse-time storage
-issue is resolved through the optimal checkpointing technique
-\cite[]{Griewank:book,Symes:06a-pub}, again without loss of precision.
-
-Typical results with pseudorandom input traces $d_r, h_z, w_r, f_s$ are:
-\begin{itemize}
-\item computed $\langle d_r, S^+_{z_s,z_r}h_s\rangle = -2.41069174,
-  \langle (S^+_{z_s,z_r})^Td_r, h_s \rangle = -2.41069388.$
-\item  computed $\langle w_r, V^+_{z_s,z_r}f_s\rangle = -2.73362470,
-  \langle (V^+_{z_s,z_r})^Tw_r, f_s \rangle = -2.73362136.$
-\end{itemize}
-Ideally, the differences of these inner products should be at most a
-relatively small multiple of machine precision, {\em relative} to the
-products $\|d_r\|\| S^+_{z_s,z_r}\|\|h_s\|$ and
-$\|w_r\|\|V^+_{z_s,z_r}\|\|f_s\|$ (division by these quantities makes the result
-dimensionless and scale-independent). The operator norm $\| S^+_{z_s,z_r}\|$ is
-computationally inaccessible, so instead I used the smaller quantities
-$\|d_r\|\| S^+_{z_s,z_r}h_s\|$ etc. as stand-ins - thereby
-overestimating the relative error between the inner products. In all
-cases, over a very large number of random inputs, the largest observed
-relative error estimate was $O(10^{-9})$, well under the appropriate limit for
-single precision.
-
 \section{Pressure-to-Source}
@@
   pressure-to-source} map. This operator also turns out to be the
-principle component of a preconditioning strategy for iterative
+principal component of a preconditioning strategy for iterative
 solution of the optimization problem \ref{eqn:esis}, so I will devote
@@
 section.
-
-To give an example of \ref{eqn:snull} in action, it is necessary to
-create downgoing data and the action of the operator $\Lambda^{\pm}$
-on a downdoing field data $f_s$ on $z=z_s$. Note that a point source
-on $z=z_s$ creates high frequency energy traveling on rays parallel
-and nearly parallel to $z=z_s$, so that won't do. One simple approach
-is to create pressure and normal velocity data on $z=z_s$ by placing a
-point source at a depth $z_d<z_s$. Since the examples used here are
-homogeneous in $z<z_s$, the traces extracted from the resulting
-causal pressure and velocity data on $z=z_s$ are {\em a priori}
-downgoing, hence related by the operator $\Lambda^+_{z_s}$.
-
-Since the
-mechanical parameters in the homogeneous and lens models are the same
-for $z<z_s$, and no rays return to this zone in either model, these
-data are asymptotically the same for both models, and I show only the
-homogenous medium results. Regard these gathers as the trace
-$(P_sp^+,P_sv^+_z)$ on $z=z_s$ of a downgoing acoustic field in $z>z_s$.
-Equations \ref{eqn:tracejump10} and
-\ref{eqn:tracejump20} show that these differ by a factor of -2 from
-source functions $f_s$ and $h_s$ as in the system \ref{eqn:awepm},
-with $h_s=0$ and $f_s=0$ respectively. For a point source at $z_d=3500$ m, $x_d=3500$ m, same bandpass filter
-wavelet as used for prior examples, causal data on $z=z_s=3000$ m are
-shown in Figures \ref{fig:dsrcphh0} and \ref{fig:dsrcvzhh0}.
-
-These source functions satisfy
-the relation \ref{eqn:hfcondn}, therefore source vectors $(h_s,0)$ and
-$(0, f_s)$ generate {\em the same acoustic field} $(p^+,\bv^+)$ in $z>z_s$. Figures
-\ref{fig:drecphh0}, \ref{fig:dfwdphh0}, \ref{fig:daltphh0} (homogeneous model) and
-\ref{fig:drecplh0}, \ref{fig:dfwdplh0}, \ref{fig:daltplh0} (lens model) show the pressure
-gathers extracted at $z_r=1000$ m for the point source at $z=z_d$ and
-for the two choices of extended source at $z=z_s$, on the same color
-scale. The obvious similarity between the fields generated by the two
-extended sources,
-predicted by equation \ref{eqn:snull}, is confirmed by trace
-comparisons in figures \ref{fig:drecphh0tr21}-\ref{fig:daltphh0tr81}
-and
-\ref{fig:drecplh0tr21}-\ref{fig:daltplh0tr81}.
 
@@
 
-I have applied this procedure to the pressure gathers shown in Figures
-\ref{fig:drecphh0} and \ref{fig:drecplh0}, generated by a point source
-at $z_d=3500$ m, $x_s=3500$ m, propagating in the homogeneous and lens
-models respectively. The source time dependence is again a [1, 2.5,
-10, 12.5] Hz trapezoidal bandpass filter. As noted earlier, in the
-region $z < z_s=3000 m$, these fields are downgoing and can be
-regarded as the result of either pressure or velocity source at $z=z_s$: the
-pressure source gather $h_s$ (Figure \ref{fig:dhshh0})  is -2 times the
-vertical velocity gather depicted in Figure \ref{fig:dsrcvzhh0}, the
-velocity source gather $f_s$ (Figure \ref{fig:dfshh0}) is -2 times the
-pressure gather depicted in Figure \ref{fig:dsrcphh0}.
-
-Figure \ref{fig:dinvhshh0} shows the approximate inversion (via the
-first equation in display \ref{eqn:approxinv}) of the
-pressure gather shown in Figure \ref{fig:drecphh0}, inverted in the homogeneous model
-used to generate the data. The result is a scaled, dip-filtered
-version of the pressure source shown in Figure \ref{fig:dhshh0}.
-The dip filter effect is unavoidable and is caused by the aperture
-limitation of the acquisition geometry: the steeper dips in the source
-gather (Figure \ref{fig:dhshh0}) do not contribute to the data, and are not
-recovered in the inversion.
-
-To see that this is indeed an inversion, I apply $S^{+}_{z_s,z_r}$ to
-the source shown in Figure \ref{fig:dinvhshh0}, to produce the
-re-simulation shown in Figure \ref{fig:drerecphh0}. Simulation is carried
-out in the homogeneous model used in the inversion. Comparison with the
-gather in Figure \ref{fig:drecphh0} via a difference plot
-\ref{fig:ddiffrecphh0} verifies that a very close match has been
-obtained, so the source in Figure \ref{fig:dinvhshh0} is indeed a
-satisfactory inversion, even it differs considerably from the original
-point source: the difference lies in an approximate null space of the modeling
-operator, arising from the limited aperture of the acquisition
-geometry.
-
-The remainder of this section is devoted to illustrating two important
-features of the approximate inversion developed above. The
-first is the validity of the inversion approach given in display
-\ref{eqn:approxinv} so long as the fields generated by the
-model used in simulation and inversion are downgoing. The models
-may be quite far from homogeneous, and even generate triplications,
-whereas the inversions remain accurate modulo the dip filtering effect
-of aperture. Figure \ref{fig:dinvhsll0} shows the pressure source at $z=z_s$
-inverted from the data Figure
-\ref{fig:drecplh0}, in which a triplication is obvious. Recall that
-this data is generated by propagating a point source at $z_d=3500$ m
-in the lens model. The inversion uses the same model. Figure
-\ref{fig:dinvhsll0} is an aperture-limited version of the pressure
-source at $z=z_s=3000$ m
-(Figure \ref{fig:dhshh0}) that generates this data by a dip filter, in
-the same way that the sources in Figures \ref{fig:dhshh0} and
-\ref{fig:dinvhshh0} are related. The inverted source is an accurate
-inversion: it generates the data depicted in Figure
-\ref{fig:drerecpll0}, which is very close to the input data (Figure
-\ref{fig:drecplh0}). The difference is displayed on the same color
-scale in Figure \ref{fig:ddiffrecpll0}.
-
-The second feature is a decoupling of the models used (implicitly or
-otherwise) in simulation and those used in inversion: it is possible
-to {\em accurately invert pressure data for pressure source using the
-  wrong model}, in the sense that the inverted source will
-generate an accurate recovery of the input data re-simulated in the
-same (wrong) model. For example, inversion of the data of the previous
-example (Figure \ref{fig:drecplh0}) {\em in the homogeneous model}
-results in an inverted source gather at $z=z_s$ (Figure \ref{fig:dinvhslh0}
-that is very far from a dip-filtered version of the source used
-generate the data (Figure \ref{fig:dhshh0}). However, this result is an
-accurate inversion: re-simulation {\em using the same (homogeneous)
-  model as used in the inversion} results in accurate recovery of the
-input pressure gather (Figure \ref{fig:drecplh0}). The difference is
-shown on the same color scale in Figure \ref{fig:ddiffrecplh0}.
-
 \section{Unitarity}
@@
 This is the promised relation.
-
-A minor extension of an earlier example, based on the gathers
-displayed in Figures \ref{fig:dsrcphh0} ($P_sp^+$) and
-\ref{fig:dsrcvzhh0} ($P_sv_z^+$) illustrates the identity
-\ref{eqn:sv}.
-The pressure gathers at $z=z_r$ produced by application of
-$S^+_{z_r,z_s}$ to $h_s=-2P_s v^+_z$, and $\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T$ to $f_s
-=-2P_sp^+$, have already been displayed (Figures \ref{fig:dfwdphh0}
-and \ref{fig:daltphh0}), and are identical to several digits. The same 
-is true of the vertical velocity gathers, shown in \ref{fig:dfwdvzhh0}
-and \ref{fig:daltvzhh0}; their difference, plotted on the same scale,
-appears in Figure \ref{fig:dsvcomphh0}. Once again, this relation
-expresses the interchangeability of pressure and velocity sources,
-related by $\Lambda$.
-
-Relation \ref{eqn:tracejump10} characterizes
-$\Lambda^+_{z_s}$ as connecting the pressure and velocity components
-of downgoing fields restricted to $z=z_s$. That is, the pressure
-source gather
-$h_s=\Lambda^{+}_{z_s}P_s p^+ = -2
-P_{z_s}v_z$, displayed in Figure \ref{fig:dhshh0}, is the image of the
-pressure gather in Figure \ref{fig:dsrcphh0} under
-$\Lambda^{+}_{z_s}$.
-The pressure gather
-\ref{fig:dfwdphh0} is the image of this pressure source gather under
-$S^+_{z_s,z_r}$. The corresponding vertical velocity gather
-(Figure \ref{fig:dfwdvzhh0}) is $-1/2$ times $\Lambda^+_{z_r}P_r
-p$. Therefore scaling the data in Figure \ref{fig:dfwdvzhh0} by
-$-2$ produces
-$\Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_s p^+$. On the other
-hand, figure \ref{fig:daltvzhh0} shows the result of applying
-$V^+_{z_s,z_r}$ to $f_s=-2P_s p^+$. Therefore scaling the gather in
-Figure \ref{fig:daltvzhh0} by $-\frac{1}{2}$ produces $V^+_{z_s,z_r}P_zp^+$.
-Since the data in Figures \ref{fig:dfwdvzhh0} and \ref{fig:daltvzhh0}
-are essentially identical, the relation \ref{eqn:sv} holds for this
-example.
 
@@
 \end{equation}
-
-To illustrate this unitary property of $S^+_{z_s,z_r}$, I apply the
-operator
-\[
-  \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
-  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
-  \Lambda^+_{z_r})
-\]
-to the data $S^+_{z_s,z_r}h_s$ (Figure \ref{fig:drerecphh0}), in which
-$h_s$ is the dip-filtered version (Figure \ref{fig:dinvhshh0}) of the
-downgoing source created earlier (Figure \ref{fig:dhshh0}), produced
-by inversion of the data in Figure \ref{fig:drecphh0} via the first
-equation in display \ref{eqn:approxinv}, propagation in the homogenous
-model.  The operator above is computed via the technique explained in
-the section ``Computing and Symmetrizing $\Lambda$'', below, using
-auxiliary receiver arrays 100 m above the data source and receiver arrays.
-
-The output is shown in Figure
-\ref{fig:lamsstlamrdrecphh0}. The difference with the actual source is
-shown in Figure \ref{fig:difflamsstlamrdrecphh0}. A similar exercise
-using the lens data \ref{fig:drecplh0} but inversion in the homogenous
-model produces the extended source depicted in
-\ref{fig:lamsstlamrdrecplh0}, which differs from the inversion result
-via equation \ref{eqn:approxinv} by the gather displayed in Figure
-\ref{fig:difflamsstlamrdrecplh0}.
 
@@
 
-[Example]
-
 \section{Computing and Symmetrizing $\Lambda$}
@@
 eliminates any tangentially propagating fields from consideration.
-
-Figure \ref{fig:preddinvhshh0} shows the pressure source gather
-produced from the pressure data gather in Figure \ref{fig:dsrcphh0} by
-application of the operator on the right-hand side of formula
-\ref{eqn:lamident} to the source surface pressure gather depicted in
-\ref{fig:dsrcphh0}. The homogeneous model is used in all propagations
-implicit in the prescription \ref{eqn:lamident}. The scale is the same
-as for Figures \ref{fig:dinvhshh0} and \ref{fig:dhshh0}. In fact
-Figure \ref{fig:preddinvhshh0} appears to be nearly identical to
-Figure \ref{fig:dinvhshh0}, and both appear to be dip-filtered
-versions of Figure \ref{fig:dhshh0}. The difference plot (Figure
-  \ref{fig:ddiffinvhshh0}), using the same color scale, shows that the
-  similarity is quantitative. The recovered source gather (Figure
-  \ref{fig:preddinvhshh0}) also generates the same acoustic fields as
-  the point source at $z=z_d$: the pressure gather at $z=z_r=1000$ m
-  depth is shown in Figure \ref{fig:dpredhsrecphh0}, and its
-  difference with the point source gather (Figure \ref{fig:drecphh0})
-  plotted on the same color scale in Figure
-  \ref{fig:ddiffpredhsrecphh0}.
-
-Figures \ref{fig:preddinvhsll0}, \ref{fig:ddiffinvhsll0},
-\ref{fig:dpredhsrecpll0}, and \ref{fig:ddiffpredhsrecpll0} accomplish
-the same comparison for the $\Lambda^+_{z_s}$ approximation in
-equation \ref{eqn:lamident}, with all propagation taking place in the
-lens model.
 
@@
 
-To illustrate these points, I have re-computed the source gather
-estimates shown in Figure \ref{fig:preddinvhshh0}. In place of the receiver geometry in the
-target data (Figure \ref{fig:drecphh0}), I
+As mentioned in the last section, computation of the transpose of
+$\Lambda^+$ (exact, not approximate in the high frequency sense) is
+critical to the successful construction of the preconditioner. The
+relation \ref{eqn:lamident} does not provide a computation for this
+operator. However set
+\begin{equation}
+  \label{eqn:lamtilde}
+  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
+    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
+\end{equation}
+Then \ref{eqn:lamnear} can be rewritten
+\[
+  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
+\]
+Of course, all of the examples so far show images of
+$\tilde{\Lambda}^+_{z_s}$.
+
+Since successful preconditioning requires only approximate inversion,
+use of $\tilde{\Lambda}^+_{z_s} $ in place of $\Lambda^+_{z_s}$ will
+still yield a working preconditioner, and the former can be transposed
+to machine precision via the definition \ref{eqn:lamtilde} and the adjoint state
+method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
+\begin{equation}
+  \label{eqn:lamtransp}
+  (\tilde{\Lambda}^+_{z_s})^T = -8 \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
+    z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}
+\end{equation}
+
+The model space weight operator $W_m^{-1}$ introduced in the last section is
+replaced by its asymptotic approximation
+\[
+\frac{1}{2}(\tilde{\Lambda}^+_{z_s} +
+  (\tilde{\Lambda}^+_{z_s})^T)
+\]
+\[
+  \approx -8 \left((V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
+    S}^+_{z_s,z_s+\Delta z}\Pi_1^T \right.+
+\]
+\[
+  \left. \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
+      z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}\right)
+\]
+\begin{equation}
+  \label{eqn:winvcomp}
+  = -4(\Pi_1 ({\cal S}^+_{z_s,z_s+\Delta z})^T(\Pi_0^T\Pi_1 +
+  \Pi_1^T\Pi_0) {\cal S}^+_{z_s,z_s+\Delta z}\Pi_1^T) = \tilde{W}_m^{-1} .
+\end{equation}
+with a similar definition for the replacement $\tilde{W}_d$ of $W_d$.
+
+This identity shows that only one forward and one adjoint simulation
+are necessary to compute the action of $\tilde{W}_{m,d}$. The operator
+in the center of the expression on the right-hand side, $\Pi_0^T\Pi_1
++ \Pi_1^T\Pi_0$, simply exchanges the components of the acoustic
+fields, passing the velocity field as a pressure source and the
+pressure field as a velocity source.
+
+One more computation is required for the full implementation of the
+preconditioning strategy explained in the last section: $W_m$ is
+required, not just $W_m^{-1}$. Note that $W_m$ plays two roles in the
+second term in equation \ref{eqn:norm1}: it is the weight matrix for
+both the domain and range norms for $A$. It is perfectly OK for one of
+these to be replaced by an asymptotic approximation, so long as it is
+symmetric and computable (and at least semi-definite). The second row
+in equation \ref{eqn:snull} appears as \ref{eqn:snull2} above:
+introducing (formally) the inverse of $\Lambda^+$,
+\begin{equation}
+  \label{eqn:snull2mod}
+-\frac{1}{2}\Pi_1{\cal S}^{+}_{z_s,z_r}\Pi_0^T  \approx
+V^{+}_{z_s,z_r}(\Lambda^{+}_{z_s})^{-1}
+\end{equation}
+whence from the second line in display \ref{eqn:approxinv}
+\begin{equation}
+  \label{eqn:laminv}
+-\frac{1}{8}(S^+_{z_s,z_r})^T\Pi_1{\cal S}^{+}_{z_s,z_r}\Pi_0^T  
+\approx (\Lambda^{+}_{z_s})^{-1}
+\end{equation} 
+and
+\begin{equation}
+  \label{eqn:laminvtr}
+-\frac{1}{8}\Pi_0({\cal S}^{+}_{z_s,z_r})^T\Pi_1^T S^+_{z_s,z_r}
+\approx ((\Lambda^{+}_{z_s})^{-1})^T.
+\end{equation}
+Using the definition \ref{eqn:sdef} of $S^+_{z_s,z_r}$, the
+symmetrized $\Lambda^{-1}$ is
+\begin{equation}
+  \label{wcomp}
+ \tilde{W}_m = -\frac{1}{16}\left(\Pi_0 ({\cal S}^{+}_{z_s,z_r})^T
+   (\Pi_1^T \Pi_0 + \Pi_0^T \Pi_1) {\cal S}^{+}_{z_s,z_r} \Pi_0^T\right)
+   \approx \frac{1}{2}((\Lambda^{+}_{z_s})^{-1}
+   +((\Lambda^{+}_{z_s})^{-1})^T) .
+\end{equation}
+Comparison with the definition \ref{eqn:winvcomp} shows that
+$\tilde{W}_m$ and $\tilde{W}_m^{-1}$ differ only in the initial and
+final projection factors (and overall scale), and in particular either
+can be computed for the cost of a forward/adjoint operator pair. Note
+that $\tilde{W}_m^{-1}$ is inverse to $\tilde{W}_m$ only in an
+approximate (asymptotic, aperture-limited) sense.
+
+\section{Numerical Examples}
+This section illustrates the most important conclusions developed in
+the preceding sections by finite difference wavefield simulation.
+
+\subsection{Synthetic models and simulation}
+To illustrate the structure described in the preceding section, I
+introduce two 2D acoustic models, one spatially homogeneous, the other
+highly refractive. The first, homogenous model has $\kappa = 4$
+GPa and $\rho = 1$ g/cm$^3$ throughout a rectangular domain of size 8 km ($x$) $\times$ 4 km
+($z$). The second, refractive, model is a perturbation of the first by
+a low-velocity acoustic lens positioned in the center of the
+rectangle. To produce this structure, the density is chosen
+homogeneous as in the first model, while the bulk modulus decreases to
+from 4 GPa outside the lens to 1.6 GPA in its center, as shown in Figure \ref{fig:bml0}.
+
+Discretization is conventional, with a rectangular grid and staggered
+finite difference scheme \cite[]{Vir:84} of order 2 in time and
+2$k$ in space; for most of the experiments reported below,
+$k=4$. Absorbing boudary conditions of perfectly matched layer type
+are applied at all boundaries of the simulation rectangle \cite[]{Habashy:07}.
+Sampling operators such as $P_r$ are implemented via linear
+interpolation, and source insertion via adjoint linear interpolation
+(as noted above, in the continuum limit, sources are represented via
+adjoint sampling). Steps in $x$ and $z$ are the same. Most results use
+$\Delta x = 20$ m, and limit the temporal frequency of the computed
+traces to 12.5 Hz.
+
+\cite{GeoPros:11} gives a description of the code
+implementation, out-of-date in a few respects but overall accurate.
+The implementation uses the discrete adjoint state method and
+auto-generated code \cite[]{TapenadeRef13}, to assure that the
+computed adjoint operators are adjoint at the level of machine
+precision to the computed operators. The reverse-time storage issue is
+resolved through the optimal checkpointing technique
+\cite[]{Griewank:book,Symes:06a-pub}, again without loss of
+precision. This procedure results in computed adjoints for
+$S^+_{z_s,z_r}$ and other operators that pass usual test for adjoint
+accuracy, comparing inner products with pseudorandom input vectors,
+with errors well under machine precision.
+
+%Typical results with pseudorandom input traces $d_r, h_z, w_r, f_s$ are:
+%\begin{itemize}
+%\item computed $\langle d_r, S^+_{z_s,z_r}h_s\rangle = -2.41069174,
+%  \langle (S^+_{z_s,z_r})^Td_r, h_s \rangle = -2.41069388.$
+%\item  computed $\langle w_r, V^+_{z_s,z_r}f_s\rangle = -2.73362470,
+%  \langle (V^+_{z_s,z_r})^Tw_r, f_s \rangle = -2.73362136.$
+%\end{itemize}
+%Ideally, the differences of these inner products should be at most a
+%relatively small multiple of machine precision, {\em relative} to the
+%products $\|d_r\|\| S^+_{z_s,z_r}\|\|h_s\|$ and
+%$\|w_r\|\|V^+_{z_s,z_r}\|\|f_s\|$ (division by these quantities makes the result
+%dimensionless and scale-independent). The operator norm $\| S^+_{z_s,z_r}\|$ is
+%computationally inaccessible, so instead I used the smaller quantities
+%$\|d_r\|\| S^+_{z_s,z_r}h_s\|$ etc. as stand-ins - thereby
+%overestimating the relative error between the inner products. In all
+%cases, over a very large number of random inputs, the largest observed
+%relative error estimate was $O(10^{-9})$, well under the appropriate limit for
+%single precision.
+
+%We also explore the dependence of a few results on frequency, using
+%$\Delta x = 10$ m and $5$ m, to accomodate 25 and 50 Hz respectively,
+%maintaining 8 samples per wavelength. All computations are carried out
+%in single precision.
+
+The horizontal line of receivers sits at depth $z_r = $ 1000 m, the
+(extended) sources at $z_s=3000$ m.  %as
+% shown in Figure \ref{fig:bml0}.
+Source and receiver $x$ ranges from $2000$ to
+$6000$ m. Note that we have
+reversed the order relation between $z_s$ and $z_r$ described in the
+text ($z_s<z_r$). This difference is immaterial for the purpose of
+illustrating the mathematical structures developed in the preceding paragraphs.
+%A single point (physical) source appears in these
+%experiments, located at $x_s=3500$ m, $z_s=3000$ m, as also indicated
+%in the figure.
+
+%Extended sources are confined to the horizontal
+%line through the physical source position, that is $z_s = 3000$ m,
+%over a 4 km interval starting at $x_r=$ 2000 m..
+
+%The point source pressure data generated by this configuration for the
+%homogeneous model is displayed in Figure \ref{fig:recphh0}, for the
+%lens model in \ref{fig:recplh0}. Triplication of arrivals is evident in the latter.
+
+%While inversion of pressure data is the main object of this exercise,
+%normal velocity ($v_z$) data will play an important role, so I display
+%the corresponding gathers in Figures \ref{fig:recvzhh0} and \ref{fig:recvzlh0}.
+
+\subsection{Creating downgoing fields}
+The downgoing condition constrains high-frequency energy of localized
+plane wave components, hence could be enforced by dip
+filtering. However, a simpler approach is to construct fields that
+must be entirely downgoing at the source and receiver surfaces by
+virtue of ray geometry.
+
+Note that a point source
+on $z=z_s$ creates high frequency energy traveling on rays parallel
+and nearly parallel to $z=z_s$, so that won't do. However, placing a
+point source at a depth $z_d<z_s$ will work. Since the examples used here are
+homogeneous in $z<z_s$, and the sampling region for extended sources
+is a finite interval, all rays carrying high frequency energy cross
+the source surface $z=z_s$ at a postive angle, and the field and its
+traces are {\em a priori} downgoing. The same is obviously true at the
+receiver surface for the homogeneous model, but is also true for the
+lens model, as no rays are refracted horizontally {\em at the receiver surface}.
+
+The choice of a point source at
+$z_d=3500$ m, $x_d=3500$ m,  bandpass filter wavelet with
+corner frequencies $1, 2.5, 7.5, 12.5$ Hz, gives the causal pressure and
+velocity gathers at $z=z_s=3000$ m
+shown in Figures \ref{fig:dsrcphh0} and \ref{fig:dsrcvzhh0}.  Since the
+mechanical parameters in the homogeneous and lens models are the same
+for $z<z_s$, and no rays return to this zone in either model, these
+data are asymptotically the same for both models, and I show only the
+homogenous medium results.
+%for the lowest frequency
+%source, and scaled as appropriate for examples with higher frequency
+%and finer sampling.
+%wavelet as used for prior examples,
+
+\subsection{Equivalence of pressure, velocity sources}
+These gathers are the pressure and
+velocity traces
+$(P_sp^+,P_sv^+_z)$ on $z=z_s$ of a downgoing acoustic field in
+$z<z_s$, hence related by the operator $\Lambda^+_{z_s}$.
+Equations \ref{eqn:snull}, \ref{eqn:tracejump10} and
+\ref{eqn:tracejump20} show that these differ by a factor of -2 from
+source functions $f_s$ and $h_s$ in the system \ref{eqn:awepm},
+with $h_s=0$ and $f_s=0$ respectively, that generate the same acoustic
+field in $z<z_s$, and in particular the same receiver traces on
+$z=z_r$, at least asymptotically.
+
+%To give an example of \ref{eqn:snull} in action, it is necessary to
+%create downgoing data and the action of the operator $\Lambda^{\pm}$
+%on a downdoing field data $f_s$ on $z=z_s$.
+
+%These source functions satisfy
+%the relation \ref{eqn:hfcondn}, therefore source vectors $(h_s,0)$ and
+%$(0, f_s)$ generate {\em the same acoustic field} $(p^+,\bv^+)$ in
+%$z>z_s$.
+Figures
+%\ref{fig:drecphh0}, \ref{fig:dfwdphh0}, \ref{fig:daltphh0} (homogeneous model) and
+\ref{fig:drecplh0}, \ref{fig:dfwdplh0}, \ref{fig:daltplh0} show the pressure
+gathers extracted at $z_r=1000$ m for the point source at $z=z_d$ and
+for the two choices of extended source at $z=z_s$, on the same color
+scale. The obvious similarity between the fields generated by the two
+extended sources,
+predicted by equation \ref{eqn:snull}, is confirmed by trace
+comparisons in figures %\ref{fig:drecphh0tr21}-\ref{fig:daltphh0tr81}
+%and
+\ref{fig:drecplh0tr81},\ref{fig:daltplh0tr81}. Other traces are
+equally similar.
+
+% time rev
+
+\subsection{Inversion by time reversal}
+I have applied the approximate inversion procedure suggested in
+equation \ref{eqn:approxinv} to the pressure gather shown in Figure
+% \ref{fig:drecphh0} and
+\ref{fig:drecplh0}, generated by a point source
+at $z_d=3500$ m, $x_s=3500$ m, propagating in the %homogeneous and
+lens
+model (Figure \ref{fig:bml0}). I choose this example for two
+reasons. First, the success of the inversion demostrates the
+insensitivity of the time reversal method to ray multipathing
+(triplication), evident in the data (Figure \ref{fig:drecplh0}).
+%s respectively
+Second, I will invert this data in the homogeneous model, that is,
+construct sources that (approximately) reproduce the data using a
+different material model than the one in which it was produced. This
+capability is critically important in the application of the
+approximate inversion in nonlinear inversion, where the early iterations
+involve solution of the source estimation problem \ref{eqn:esis} at
+(possibly very) wrong material models ${\bf c}$. Successful extension
+methods maintain data fit throughout the course of the inversion.
+%The source time dependence is again a [1, 2.5,
+%10, 12.5] Hz trapezoidal bandpass filter.
+
+As noted earlier, the acoustic field in this example is downgoing
+throughout the simulation range. It can be
+regarded as the result of either pressure or velocity source at $z=z_s$: the
+pressure source gather $h_s$ (Figure \ref{fig:dhshh0})  is -2 times the
+vertical velocity gather depicted in Figure \ref{fig:dsrcvzhh0}, the
+velocity source gather $f_s$ (Figure \ref{fig:dfshh0}) is -2 times the
+pressure gather depicted in Figure \ref{fig:dsrcphh0}.
+
+Figure \ref{fig:dinvhslh0} shows the approximate inversion (via the
+first equation in display \ref{eqn:approxinv}) of the
+pressure gather shown in Figure \ref{fig:drecplh0}, inverted in the homogeneous model
+(rather than in the lens model used to generate the data). The result
+differs greatly from the pressure source shown in Figure
+\ref{fig:dhshh0}, as it must since it results from inversion in the
+wrong material model.
+Some dip filter effect is unavoidable and is caused by the aperture
+limitation of the acquisition geometry: the steeper dips in the source
+gather (Figure \ref{fig:dhshh0}) do not contribute to the data, nor to
+the inversion. Also, the limited receiver aperture causes truncation
+artifacts in the inversion. However, this result is an
+accurate inversion: re-simulation (application of $S^{+}_{z_s,z_r}$) {\em using the same (homogeneous)
+  model as used in the inversion} results in accurate recovery (Figure \ref{fig:drerecplh0}) of the
+input pressure gather (Figure \ref{fig:drecplh0}). The difference is
+shown on the same color scale in Figure \ref{fig:ddiffrecplh0}.
+
+%To see that this is indeed an inversion, I apply $S^{+}_{z_s,z_r}$ to
+%the source shown in Figure \ref{fig:dinvhshh0}, to produce the
+%re-simulation shown in Figure \ref{fig:drerecphh0}. Simulation is carried
+%out in the homogeneous model used in the inversion. Comparison with the
+%gather in Figure \ref{fig:drecphh0} via a difference plot
+%\ref{fig:ddiffrecphh0} verifies that a very close match has been
+%obtained, so the source in Figure \ref{fig:dinvhshh0} is indeed a
+%satisfactory inversion, even it differs considerably from the original
+%point source: the difference lies in an approximate null space of the modeling
+%operator, arising from the limited aperture of the acquisition
+%geometry.
+
+%The remainder of this section is devoted to illustrating two important
+%features of the approximate inversion developed above. The
+%first is the validity of the inversion approach given in display
+%\ref{eqn:approxinv} so long as the fields generated by the
+%model used in simulation and inversion are downgoing. The models
+%may be quite far from homogeneous, and even generate triplications,
+%whereas the inversions remain accurate modulo the dip filtering effect
+%of aperture. Figure \ref{fig:dinvhsll0} shows the pressure source at $z=z_s$
+%inverted from the data Figure
+%\ref{fig:drecplh0}, in which a triplication is obvious. Recall that
+%this data is generated by propagating a point source at $z_d=3500$ m
+%in the lens model. The inversion uses the same model. Figure
+%\ref{fig:dinvhsll0} is an aperture-limited version of the pressure
+%source at $z=z_s=3000$ m
+%(Figure \ref{fig:dhshh0}) that generates this data by a dip filter, in
+%the same way that the sources in Figures \ref{fig:dhshh0} and
+%\ref{fig:dinvhshh0} are related. The inverted source is an accurate
+%inversion: it generates the data depicted in Figure
+%\ref{fig:drerecpll0}, which is very close to the input data (Figure
+%\ref{fig:drecplh0}). The difference is displayed on the same color
+%scale in Figure \ref{fig:ddiffrecpll0}.
+
+%The second feature is a decoupling of the models used (implicitly or
+%otherwise) in simulation and those used in inversion: it is possible
+%to {\em accurately invert pressure data for pressure source using the
+%  wrong model}, in the sense that the inverted source will
+%generate an accurate recovery of the input data re-simulated in the
+%same (wrong) model. For example, inversion of the data of the previous
+%example (Figure \ref{fig:drecplh0}) {\em in the homogeneous model}
+%results in an inverted source gather at $z=z_s$ (Figure \ref{fig:dinvhslh0}
+%that is very far from a dip-filtered version of the source used
+%generate the data (Figure \ref{fig:dhshh0}).
+
+% unitarity
+
+\subsection{Quasi-unitary property of the modeling operator}
+The identities \ref{eqn:approxinv} and \ref{eqn:sv} would together
+establish the approximately unitary property of $S^+_{z_s,z_r}$, if
+$\Lambda$ were symmetric. Identity \ref{eqn:approxinv} was illustrated
+in the last subsection. Setting the symmetric issue aside for the
+moment, an illustration of the relation \ref{eqn:sv} proceeds as follows.
+%A minor extension of an earlier example, based on the gathers
+%displayed in Figures \ref{fig:dsrcphh0} ($P_sp^+$) and
+%\ref{fig:dsrcvzhh0} ($P_sv_z^+$) illustrates the identity
+%\ref{eqn:sv}.
+%The pressure gathers at $z=z_r$ produced by application of
+%$S^+_{z_r,z_s}$ to $h_s=-2P_s v^+_z$, and $\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T$ to $f_s
+%=-2P_sp^+$, have already been displayed (Figures \ref{fig:dfwdphh0}
+%and \ref{fig:daltphh0}), and are identical to several digits. The same 
+%is true of the vertical velocity gathers, shown in \ref{fig:dfwdvzhh0}
+%and \ref{fig:daltvzhh0}; their difference, plotted on the same scale,
+%appears in Figure \ref{fig:dsvcomphh0}. Once again, this relation
+%expresses the interchangeability of pressure and velocity sources,
+%related by $\Lambda$.
+
+Relation \ref{eqn:tracejump10} characterizes
+$\Lambda^+_{z_s}$ as connecting the pressure and velocity components
+of downgoing fields restricted to $z=z_s$. That is, the pressure
+source gather
+$h_s=\Lambda^{+}_{z_s}P_s p^+ = -2
+P_{z_s}v_z$, displayed in Figure \ref{fig:dhshh0}, is the image of the
+pressure gather in Figure \ref{fig:dsrcphh0} under
+$\Lambda^{+}_{z_s}$.
+The pressure gather
+\ref{fig:dfwdplh0} is the image of this pressure source gather under
+$S^+_{z_s,z_r}$ (using the lens model). The corresponding vertical velocity gather
+(Figure \ref{fig:dfwdvzlh0}) is $-1/2$ times $\Lambda^+_{z_r}P_r
+p$. Therefore scaling the data in Figure \ref{fig:dfwdvzlh0} by
+$-2$ produces
+$\Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_s p^+$. On the other
+hand, figure \ref{fig:daltvzlh0} shows the result of applying
+$V^+_{z_s,z_r}$ to $f_s=-2P_s p^+$. Therefore scaling the gather in
+Figure \ref{fig:daltvzlh0} by $-\frac{1}{2}$ produces $V^+_{z_s,z_r}P_zp^+$.
+Since the data in Figures \ref{fig:dfwdvzlh0} and \ref{fig:daltvzlh0}
+are essentially identical, the relation \ref{eqn:sv} holds for this
+example.
+
+\subsection{Accurate computation of $\Lambda$}
+%Figure \ref{fig:preddinvhshh0} shows the pressure source gather
+%produced from the pressure data gather in Figure \ref{fig:dsrcphh0} by
+%application of the operator on the right-hand side of formula
+%\ref{eqn:lamident} to the source surface pressure gather depicted in
+%\ref{fig:dsrcphh0}. The homogeneous model is used in all propagations
+%implicit in the prescription \ref{eqn:lamident}. The scale is the same
+%as for Figures \ref{fig:dinvhshh0} and \ref{fig:dhshh0}. In fact
+%Figure \ref{fig:preddinvhshh0} appears to be nearly identical to
+%Figure \ref{fig:dinvhshh0}, and both appear to be dip-filtered
+%versions of Figure \ref{fig:dhshh0}. The difference plot (Figure
+ % \ref{fig:ddiffinvhshh0}), using the same color scale, shows that the
+ % similarity is quantitative. The recovered source gather (Figure
+ % \ref{fig:preddinvhshh0}) also generates the same acoustic fields as
+  %the point source at $z=z_d$: the pressure gather at $z=z_r=1000$ m
+%  depth (propagated in the lens model) is shown in Figure \ref{fig:dpredhsrecplh0}, and its
+%  difference with the point source gather (Figure \ref{fig:drecplh0})
+%  plotted on the same color scale in Figure
+%  \ref{fig:ddiffpredhsrecplh0}.
+
+%Figures \ref{fig:preddinvhsll0}, \ref{fig:ddiffinvhsll0},
+%\ref{fig:dpredhsrecpll0}, and \ref{fig:ddiffpredhsrecpll0} accomplish
+%the same comparison for the $\Lambda^+_{z_s}$ approximation in
+%equation \ref{eqn:lamident}, with all propagation taking place in the
+%lens model.
+
+\subsection{Economical computation of $\Lambda$ in ``thin'' subdomain}
+Equation \ref{eqn:lamnear} suggests a thin-slab computation of the
+$\Lambda$ action, which is both accurate and economical. 
+estimates shown in Figure \ref{fig:preddinvhshh0}. This calculations
 place a receiver array at $z_s+\Delta z=2900$ m depth, just 100 m above the
@@
 depth, as opposed to 100 gridpoints between the source and receiver
-depths for examples such as shown in Figure \ref{fig:recphh0} or
-\ref{fig:drecphh0}.
-
-Note that the homogeneous and lens models are identical in the depth
+depths for examples such as shown in Figure \ref{fig:drecphh0}.
+
+Asymptotically, $\Lambda^{\pm}_{z_s}$ depends only on the medium
+coefficients ${\bf c}$ in an arbirarily small region containing the
+source surface $z=z_s$. In this example, the homogeneous and lens models are identical in the depth
 range $2900 < z < 3000$ m, so the computed $\Lambda^+_{z_s}$ operators
-will be the same for both models. Hence I show only results for the
+will be precisely the same for both models. Hence I show only results for the
 homogenous model.
 
-The approximation to $\Lambda_{z_s}^+$ via equation \ref{eqn:lamident}
+The approximation to $\Lambda_{z_s}^+$ via equation \ref{eqn:lamnear}
 for this configuration is evaluated in the same way as before in
-Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
-\ref{fig:dprednshsrecphh0}, and \ref{fig:ddiffprednshsrecphh0}. The effect of aperture
+Figures \ref{fig:preddnshslh0}, \ref{fig:ddiffnshslh0},
+\ref{fig:dprednshsrecplh0}, and \ref{fig:ddiffprednshsrecplh0}. The effect of aperture
 limitation is clearly diminished: the second figure in this series
@@
 
-As mentioned in the last section, computation of the transpose of
-$\Lambda^+$ (exact, not approximate in the high frequency sense) is
-critical to the successful construction of the preconditioner. The
-relation \ref{eqn:lamident} does not provide a computation for this
-operator. However set
-\begin{equation}
-  \label{eqn:lamtilde}
-  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
-\end{equation}
-Then \ref{eqn:lamnear} can be rewritten
-\[
-  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
-\]
-Of course, all of the examples so far show images of
-$\tilde{\Lambda}^+_{z_s}$.
-
-Since successful preconditioning requires only approximate inversion,
-use of $\tilde{\Lambda}^+_{z_s} $ in place of $\Lambda^+_{z_s}$ will
-still yield a working preconditioner, and the former can be transposed
-to machine precision via the definition \ref{eqn:lamtilde} and the adjoint state
-method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
-\begin{equation}
-  \label{eqn:lamtransp}
-  (\tilde{\Lambda}^+_{z_s})^T = -8 \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
-    z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}
-\end{equation}
+\subsection{Symmetrizing $\Lambda$}
 
@@
 
-The model space weight operator $W_m^{-1}$ introduced in the last section is
-replaced by its asymptotic approximation
-\[
-\frac{1}{2}(\tilde{\Lambda}^+_{z_s} +
-  (\tilde{\Lambda}^+_{z_s})^T)
-\]
-\[
-  \approx -8 \left((V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T \right.+
-\]
-\[
-  \left. \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
-      z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}\right)
-\]
-\begin{equation}
-  \label{eqn:winvcomp}
-  = -4(\Pi_1 ({\cal S}^+_{z_s,z_s+\Delta z})^T(\Pi_0^T\Pi_1 +
-  \Pi_1^T\Pi_0) {\cal S}^+_{z_s,z_s+\Delta z}\Pi_1^T) = \tilde{W}_m^{-1} .
-\end{equation}
-with a similar definition for the replacement $\tilde{W}_d$ of $W_d$.
-
-This identity shows that only one forward and one adjoint simulation
-are necessary to compute the action of $\tilde{W}_{m,d}$. The operator
-in the center of the expression on the right-hand side, $\Pi_0^T\Pi_1
-+ \Pi_1^T\Pi_0$, simply exchanges the components of the acoustic
-fields, passing the velocity field as a pressure source and the
-pressure field as a velocity source.
+\subsection{Asymptotic symmetry of $\Lambda$}
 
@@
 
-One more computation is required for the full implementation of the
-preconditioning strategy explained in the last section: $W_m$ is
-required, not just $W_m^{-1}$. Note that $W_m$ plays two roles in the
-second term in equation \ref{eqn:norm1}: it is the weight matrix for
-both the domain and range norms for $A$. It is perfectly OK for one of
-these to be replaced by an asymptotic approximation, so long as it is
-symmetric and computable (and at least semi-definite). The second row
-in equation \ref{eqn:snull} appears as \ref{eqn:snull2} above:
-introducing (formally) the inverse of $\Lambda^+$,
-\begin{equation}
-  \label{eqn:snull2mod}
--\frac{1}{2}\Pi_1{\cal S}^{+}_{z_s,z_r}\Pi_0^T  \approx
-V^{+}_{z_s,z_r}(\Lambda^{+}_{z_s})^{-1}
-\end{equation}
-whence from the second line in display \ref{eqn:approxinv}
-\begin{equation}
-  \label{eqn:laminv}
--\frac{1}{8}(S^+_{z_s,z_r})^T\Pi_1{\cal S}^{+}_{z_s,z_r}\Pi_0^T  
-\approx (\Lambda^{+}_{z_s})^{-1}
-\end{equation} 
-and
-\begin{equation}
-  \label{eqn:laminvtr}
--\frac{1}{8}\Pi_0({\cal S}^{+}_{z_s,z_r})^T\Pi_1^T S^+_{z_s,z_r}
-\approx ((\Lambda^{+}_{z_s})^{-1})^T.
-\end{equation}
-Using the definition \ref{eqn:sdef} of $S^+_{z_s,z_r}$, the
-symmetrized $\Lambda^{-1}$ is
-\begin{equation}
-  \label{wcomp}
- \tilde{W}_m = -\frac{1}{16}\left(\Pi_0 ({\cal S}^{+}_{z_s,z_r})^T
-   (\Pi_1^T \Pi_0 + \Pi_0^T \Pi_1) {\cal S}^{+}_{z_s,z_r} \Pi_0^T\right)
-   \approx \frac{1}{2}((\Lambda^{+}_{z_s})^{-1}
-   +((\Lambda^{+}_{z_s})^{-1})^T) .
-\end{equation}
-Comparison with the definition \ref{eqn:winvcomp} shows that
-$\tilde{W}_m$ and $\tilde{W}_m^{-1}$ differ only in the initial and
-final projection factors (and overall scale), and in particular either
-can be computed for the cost of a forward/adjoint operator pair. Note
-that $\tilde{W}_m^{-1}$ is inverse to $\tilde{W}_m$ only in an
-approximate (asymptotic, aperture-limited) sense.
+\subsection{Unitary property of modeling operator}
+To illustrate this unitary property of $S^+_{z_s,z_r}$, I apply the
+operator
+\[
+  \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
+  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
+  \Lambda^+_{z_r})
+\]
+to the data $S^+_{z_s,z_r}h_s$ (Figure \ref{fig:drerecplh0}), in which
+$h_s$ is the version (Figure \ref{fig:dinvhslh0}) of the
+downgoing source created earlier (Figure \ref{fig:dhslh0}), produced
+by inversion of the data in Figure \ref{fig:drecplh0} via the first
+equation in display \ref{eqn:approxinv}, propagation in the homogenous
+model.  The operator above is computed via the technique explained in
+the preceding subsection, below, using
+auxiliary receiver arrays 100 m above the data source and receiver arrays.
+
+The output is shown in Figure
+\ref{fig:lamsstlamrdrecplh0}. The difference with the actual source is
+shown in Figure \ref{fig:difflamsstlamrdrecplh0}.
+%A similar exercise
+%using the lens data \ref{fig:drecplh0} but inversion in the homogenous
+%model produces the extended source depicted in
+%\ref{fig:lamsstlamrdrecplh0}, which differs from the inversion result
+%via equation \ref{eqn:approxinv} by the gather displayed in Figure
+%\ref{fig:difflamsstlamrdrecplh0}.
+
+\subsection{Preconditioned CG iteration}
+This final subsection shows that result of Conjugate Gradient
+iteration, with and without preconditioning, applied to the source
+estimation problem \ref{eqn:esis}, with zero and non-zero penalty
+weight $\alpha$. The data $d$ is the gather shown in
+\ref{fig:drecplh0}, simulated using the lens model with source shown
+in Figure \ref{fig:dhshh0}, or, alternatively, a point source with
+bandpass filter wavelet located at $x_d=3500$ m, $z_d=3500$ m. In the inversion,
+the material model is taken to be homogeneous, as has been the case in
+all of the previous examples.
+
+Figure \ref{fig:compnreslh0} shows the progress of the normal residual
+(Euclidean norm of the difference of the two sides of equation \ref{eqn:norm1}),
+for Conjugate Gradient and Preconditioned Conjugate Gradient
+(Algorithm 1) iterations, applied to solution of the optimization
+problem \ref{eqn:einv} with $\alpha=0$. For CG, the norms are both the ordinary
+Euclidean norm, $W_m=W_d=I$. For PCG, $W_m$ and $W_d$ are given in
+display \ref{eqn:wdef}, with the symmetrized $\Lambda$s computed as
+indicated in the preceding subsections. Convergence for the
+preconditioned algorithm is roughly 4 times as fast.
+
+Figure \ref{fig:compnreslh0} shows the same comparison with non-zero
+penalty weight, $\alpha=10^{-3}$. The PCG normal residual curve is
+almost identical with that in the $\alpha=0$ case, wheras the CG
+convergence has slowed down noticeably, being about five times as slow
+as the preconditioned algorithm.
+
+\section{Conclusion}
 

### 2022-03-29 repro/papers/GEM22/paper.tex
near final

@@
 a low-velocity acoustic lens positioned in the center of the
-rectangle. To produce this structure, the density is chosen
+rectangle (Figure \ref{fig:bml0}. To produce this structure, the density is chosen
 homogeneous as in the first model, while the bulk modulus decreases to
 from 4 GPa outside the lens to 1.6 GPA in its center, as shown in Figure \ref{fig:bml0}.
+
+\plot{bml0}{width=\textwidth}{Bulk modulus, lens model. Color scale is 
+in GPa. Positions of point source and receiver line indicated.}
 
@@
 The downgoing condition constrains high-frequency energy of localized
-plane wave components, hence could be enforced by dip
+plane wave components, hence could be enforced by dip>
 filtering. However, a simpler approach is to construct fields that
@@
 %wavelet as used for prior examples,
+
+\plot{dsrcphh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
+  pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
+  bandpass filter source. }
+
+\plot{dsrcvzhh0}{width=\textwidth}{Trace $P_sv_z^+$ on $z=z_s=3000$ m of
+  vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
+  bandpass filter source.}
 
@@
 
+\plot{drecplh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the lens model described in the 
+  text, with point pressure source (constitutive defect) at $z_d=3500$
+  m, $x_d=3500$ m.}
+
+\plot{dfwdplh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the lens model described in the 
+  text, with extended pressure source (constitutive defect) on
+  $z=z_s=3000$ m given bythe field depicted in Figure
+  \ref{fig:dsrcvzhh0} scaled by -2 
+   ($h_s=-2P_sv_z^+=\Lambda^+_{z_s}P_sp^+$) and zero velocity source (vertical load) 
+  ($f_s=0$).}
+
+\plot{daltplh0}{width=\textwidth}{Pressure gather at receiver depth 
+  $z_r=1000$ m from field generated by causal solution of acoustic 
+  system \ref{eqn:awepm} in the lens model described in the 
+  text, with extended velocity source (vertical load) on $z=z_s=3000$
+  m given by the field depicted in Figure \ref{fig:dsrcphh0} scaled by -2
+  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
+  ($h_s=0$).}
+
+\plot{drecplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$) 
+  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:dfwdplh0}
+  (red).}
+
+\plot{daltplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$)
+  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
+  (red).}
 % time rev
@@
 
+\plot{dhshh0}{width=\textwidth}{Pressure source gather = -2 $\times$
+  vertical velocity gather (Figure \ref{fig:dsrcvzhh0}).}
+
+\plot{dfshh0}{width=\textwidth}{Velocity source gather = -2 $\times$
+  pressure gather (Figure \ref{fig:dsrcphh0}).}
+
 Figure \ref{fig:dinvhslh0} shows the approximate inversion (via the
@@
 shown on the same color scale in Figure \ref{fig:ddiffrecplh0}.
+
+\plot{dinvhslh0}{width=\textwidth}{Approximate inversion via first 
+  equation in display \ref{eqn:approxinv}. Inversion in homogenous model of 
+  the pressure gather in Figure \ref{fig:drecplh0}, simulated with 
+  lens model. Scaled version of output $v_z$ field obtained by 
+  applying transpose of $4V^+_{z_s,z_r}$. Approximates dip-filtered
+  version of pressure source gather (Figure \ref{fig:dhshh0})}
+
+\plot{drerecplh0}{width=\textwidth}{Re-simulated pressure gather produced from
+  inverted source
+  shown in Figure \ref{fig:dinvhslh0}. Simulation in homogenous model
+  used for inversion.}
+
+\plot{ddiffrecplh0}{width=\textwidth}{Difference between gathers
+  displayed in Figures \ref{fig:drecplh0} and \ref{fig:drerecplh0},
+  plotted on same color scale.}
 
@@
 
-\subsection{Accurate computation of $\Lambda$}
+\plot{dfwdvzlh0}{width=\textwidth}{Vertical velocity gather, generated
+  with a pressure source in the lens model, 
+  corresponding to pressure gather \ref{fig:dfwdplh0}.}
+
+\plot{daltvzlh0}{width=\textwidth}{Vertical velocity gather, generated
+  with a velocity source in the lens model, corresponding to
+  pressure gather \ref{fig:daltphh0}.}
+
+\plot{dsvcomplh0}{width=\textwidth}{Difference between velocity gathers
+  shown in Figures \ref{fig:dfwdvzhh0} and \ref{fig:daltvzhh0},
+  plotted on the same color scale as these figures.}
+
 %Figure \ref{fig:preddinvhshh0} shows the pressure source gather
@@
 $\Lambda$ action, which is both accurate and economical. 
-estimates shown in Figure \ref{fig:preddinvhshh0}. This calculations
+This calculations
 place a receiver array at $z_s+\Delta z=2900$ m depth, just 100 m above the
@@
 depth, as opposed to 100 gridpoints between the source and receiver
-depths for examples such as shown in Figure \ref{fig:drecphh0}.
+depths for examples such as shown in Figure \ref{fig:drecplh0}.
 
@@
 The approximation to $\Lambda_{z_s}^+$ via equation \ref{eqn:lamnear}
-for this configuration is evaluated in the same way as before in
-Figures \ref{fig:preddnshslh0}, \ref{fig:ddiffnshslh0},
+for this configuration is evaluated in
+Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
 \ref{fig:dprednshsrecplh0}, and \ref{fig:ddiffprednshsrecplh0}. The effect of aperture
@@
 
+\plot{preddnshshh0}{width=\textwidth}{Pressure source gather = image
+  under pressure-to-source operator $\Lambda^+_{z_s}$ of pressure gather
+  shown in Figure \ref{fig:dsrcphh0}, homogeneous model, using
+  ``near'' receiver traces at $z=2900$ m. Compare Figure
+  \ref{fig:dhshh0} and \ref{fig:dinvhshh0}: because the sources and
+  receivers are close, little aperture is lost in this case.}
+
+\plot{ddiffnshshh0}{width=\textwidth}{Difference between (a) image
+  (Figure \ref{fig:preddnshshh0}) of $\Lambda^+_{z_s}$ applied to
+  pressure gather (Figure \ref{fig:dsrcphh0}) using a near receiver
+  array to implement formula \ref{eqn:lamident}, and (b) source gather
+  (Figure \ref{fig:dhshh0}) inferred from vertical velocity.
+  Homogeneous model used in all
+  propagations. Same color scale as in Figure
+  \ref{fig:preddnshshh0}. }
+
+%\plot{dprednshsrecplh0}{width=\textwidth}{Pressure gather at receiver
+%  datum $z=z_r=1000$ m simulated in lens model from source gather shown in Figure
+%  \ref{fig:preddnshshh0}. Compare with point source pressure gather
+%  (Figure \ref{fig:drecplh0}).}
+
+%\plot{ddiffprednshsrecplh0}{width=\textwidth}{Plot of difference 
+%  between data shown in Figures \ref{fig:drecplh0} and 
+%  \ref{fig:dprednshsrecplh0}, plotted on same color scale as the latter 
+% two figures.}
+
+\plot{dprednshsrecpll0}{width=\textwidth}{Pressure gather at receiver
+  datum $z=z_r=1000$ m simulated in lens model from source gather shown in Figure
+  \ref{fig:preddnshshh0}. Compare with point source pressure gather
+  (Figure \ref{fig:drecplh0}).}
+
+\plot{ddiffprednshsrecpll0}{width=\textwidth}{Plot of difference 
+  between data shown in Figures \ref{fig:drecplh0} and 
+  \ref{fig:dprednshsrecplh0}, plotted on same color scale as the latter 
+  two figures.}
 \subsection{Symmetrizing $\Lambda$}
@@
 
+\plot{preddnshstrhh0}{width=\textwidth}{Pressure source gather = image
+  under {\em transpose} of pressure-to-source operator
+  $\Lambda^+_{z_s}$ of pressure gather shown in Figure
+  \ref{fig:dsrcphh0}, homogeneous model, using ``near'' receiver
+  traces at $z=2900$ m. Compare Figure \ref{fig:dhshh0} and
+  \ref{fig:preddnshshh0}: as noted in the text, $\Lambda^+_{z_s}$ is
+  asymptotically symmetric, so the resemblance is not a surprise.}
+
+\plot{ddiffnslamtrhh0}{width=\textwidth}{Difference between data in
+  Figures \ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}, plotted
+  on the same scale as these figures, showing that the asymptotic
+  symmetry of $\Lambda^+_{z_s}$ is actually quantitative for the
+  length, time and frequency scales of these examples.}
+
 \subsection{Asymptotic symmetry of $\Lambda$}
@@
 
+\plot{symmdnshshh0}{width=\textwidth}{Pressure source gather = image
+    under {\em symmetrized} pressure-to-source operator
+    $\frac{1}{2}\left(\Lambda^+_{z_s}+(\Lambda^+_{z_s})^T\right)$ of
+    pressure gather shown in Figure \ref{fig:dsrcphh0}, homogeneous
+    model, using ``near'' receiver traces at $z=2900$ m. Compare
+    Figure \ref{fig:preddnshshh0}.}
+
+\plot{ddiffsymmdnshsll0}{width=\textwidth}{Difference between point
+  source simulation (Figure \ref{fig:dfwdplh0}) and pressure gather at
+  $z=z_r=1000$ m produced by simulation with the source shown in
+  Figure \ref{fig:symmdnshshh0},
+  propagation in the lens model.}
+
 \subsection{Unitary property of modeling operator}
@@
 %\ref{fig:difflamsstlamrdrecplh0}.
+
+\plot{lamsstlamrdrecplh0}{width=\textwidth}{Inversion of data shown in
+  Figure \ref{fig:drecplh0}, simulated in lens model, using the
+  approximate unitarity relation \ref{eqn:unitary} and propagation in
+  homogenous model.}
+
+\plot{difflamsstlamrdrecplh0}{width=\textwidth}{Difference between
+  data displayed in Figures \ref{fig:dinvhslh0} and
+  \ref{fig:lamsstlamrdrecplh0}, plotted on the same color scale.}
 
@@
 
-Figure \ref{fig:compnreslh0} shows the same comparison with non-zero
-penalty weight, $\alpha=10^{-3}$. The PCG normal residual curve is
-almost identical with that in the $\alpha=0$ case, wheras the CG
-convergence has slowed down noticeably, being about five times as slow
-as the preconditioned algorithm.
-
-\section{Conclusion}
-
-\section{Figures}
-
-\plot{bml0}{width=\textwidth}{Bulk modulus, lens model. Color scale is 
-in GPa. Positions of point source and receiver line indicated.}
-
-\plot{recphh0}{width=\textwidth}{Output pressure gather for point source in 
-  homogeneous medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-\plot{recplh0}{width=\textwidth}{Output pressure gather for point source in 
-  refractive (lens) medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-\plot{recvzhh0}{width=\textwidth}{Output vertical velocity gather for point source in 
-  homogeneous medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-\plot{recvzlh0}{width=\textwidth}{Output vertical velocity gather for point source in 
-  refractive (lens) medium, configuration as described in Figure \ref{fig:bml0} and in
-  the text.}
-
-\plot{dsrcphh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
-  pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
-  bandpass filter source. }
-
-\plot{dsrcvzhh0}{width=\textwidth}{Trace $P_sv_z^+$ on $z=z_s=3000$ m of
-  vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
-  bandpass filter source.}
-
-\plot{drecphh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the homogeneous model described in the 
-  text, with point pressure source (constitutive defect) at $z_d=3500$
-  m, $x_d=3500$ m.}
-
-\plot{dfwdphh0}{width=\textwidth}{Pressure gather at receiver depth
-  $z_r=1000$ m from field generated by causal solution of acoustic
-  system \ref{eqn:awepm} in the homogeneous model described in the
-  text, with extended pressure source (constitutive defect)
-  on $z=z_s=3000$ m given by $h_s=-2P_sv_z^+=\Lambda^+_{z_s}P_sp^+$,
-  where $P_sp^+$ is shown in Figure \ref{fig:dsrcphh0} and $P_sv^+_z$
-  in Figure \ref{fig:dsrcvzhh0}. In this simulation, the velocity
-  velocity source (vertical load) vanishes: $f_s=0$.}
-
-\plot{daltphh0}{width=\textwidth}{Pressure gather at receiver depth
-  $z_r=1000$ m from field generated by causal solution of acoustic
-  system \ref{eqn:awepm} in the homogeneous model described in the
-  text, with extended velocity source (vertical load) $f_s=-2P_sp^+$
-  on $z=z_s=3000$ m, where $P_sp^+$ is shown in Figure
-  \ref{fig:dsrcphh0}.  In this simulation, the zero pressure source
-  (constitutive defect vanishes: $h_s=0$.}
-
-\plot{drecplh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the lens model described in the 
-  text, with point pressure source (constitutive defect) at $z_d=3500$
-  m, $x_d=3500$ m.}
-
-\plot{drecphh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$) 
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
-  (red).}
-
-\plot{daltphh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$) 
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:daltphh0}
-  (red).}
-
-\plot{drecphh0tr51}{width=\textwidth}{Overplot of traces 51 ($x=3000$) 
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
-  (red).}
-
-\plot{daltphh0tr51}{width=\textwidth}{Overplot of traces 51 ($x=3000$) 
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:daltphh0}
-  (red).}
-
-\plot{drecphh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$) 
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:dfwdphh0}
-  (red).}
-
-\plot{daltphh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$) 
-  from gathers shown in \ref{fig:drecphh0} (blue), \ref{fig:daltphh0}
-  (red).}
-
-\plot{dfwdplh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the lens model described in the 
-  text, with extended pressure source (constitutive defect) on
-  $z=z_s=3000$ m given bythe field depicted in Figure
-  \ref{fig:dsrcvzhh0} scaled by -2 
-   ($h_s=-2P_sv_z^+=\Lambda^+_{z_s}P_sp^+$) and zero velocity source (vertical load) 
-  ($f_s=0$).}
-
-\plot{daltplh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the lens model described in the 
-  text, with extended velocity source (vertical load) on $z=z_s=3000$
-  m given by the field depicted in Figure \ref{fig:dsrcphh0} scaled by -2
-  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
-  ($h_s=0$).}
-
-\plot{drecplh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$) 
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:dfwdplh0}
-  (red).}
-
-\plot{daltplh0tr21}{width=\textwidth}{Overplot of traces 21 ($x=2400$) 
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
-  (red).}
-
-\plot{drecplh0tr51}{width=\textwidth}{Overplot of traces 51 ($x=3000$) 
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:dfwdplh0}
-  (red).}
-
-\plot{daltplh0tr51}{width=\textwidth}{Overplot of traces 51 ($x=3000$) 
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
-  (red).}
-
-\plot{drecplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$) 
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:dfwdplh0}
-  (red).}
-
-\plot{daltplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$)
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
-  (red).}
-
-\plot{dhshh0}{width=\textwidth}{Pressure source gather = -2 $\times$
-  vertical velocity gather (Figure \ref{fig:dsrcvzhh0}).}
-
-\plot{dfshh0}{width=\textwidth}{Velocity source gather = -2 $\times$
-  pressure gather (Figure \ref{fig:dsrcphh0}).}
-
-\plot{dinvhshh0}{width=\textwidth}{Approximate inversion via first 
-  equation in display \ref{eqn:approxinv}. Inversion in homogenous model of 
-  the pressure gather in Figure \ref{fig:drecphh0}, simulated with 
-  homogeneous model. Scaled version of output $v_z$ field obtained by 
-  applying transpose of $4V^+_{z_s,z_r}$. Approximates dip-filtered
-  version of pressure source gather (Figure \ref{fig:dhshh0}).}
-
-\plot{drerecphh0}{width=\textwidth}{Re-simulated pressure gather
-  produced from inverted source
-  shown in Figure \ref{fig:dinvhshh0}. Simulation in homogeneous model.}
-
-\plot{ddiffrecphh0}{width=\textwidth}{Difference between gathers
-  displayed in Figures \ref{fig:drecphh0} and \ref{fig:drerecphh0},
-  plotted on same color scale.}
-
-\plot{dinvhsll0}{width=\textwidth}{Approximate inversion via first 
-  equation in display \ref{eqn:approxinv}. Inversion in lens model of 
-  the pressure gather in Figure \ref{fig:drecplh0}, simulated with 
-  lens model. Scaled version of output $v_z$ field obtained by 
-  applying transpose of $4V^+_{z_s,z_r}$. Approximates dip-filtered
-  version of pressure source gather (Figure \ref{fig:dhshh0}).}
-
-\plot{drerecpll0}{width=\textwidth}{Re-simulated pressure gather
-  produced from inverted source
-  shown in Figure \ref{fig:dinvhsll0}. Simulation in lens model.}
-
-\plot{ddiffrecpll0}{width=\textwidth}{Difference between gathers
-  displayed in Figures \ref{fig:drecpll0} and \ref{fig:drerecpll0},
-  plotted on same color scale.}
-
-\plot{dinvhslh0}{width=\textwidth}{Approximate inversion via first 
-  equation in display \ref{eqn:approxinv}. Inversion in homogenous model of 
-  the pressure gather in Figure \ref{fig:drecplh0}, simulated with 
-  lens model. Scaled version of output $v_z$ field obtained by 
-  applying transpose of $4V^+_{z_s,z_r}$. Approximates dip-filtered
-  version of pressure source gather (Figure \ref{fig:dhshh0})}
-
-\plot{drerecplh0}{width=\textwidth}{Re-simulated pressure gather produced from
-  inverted source
-  shown in Figure \ref{fig:dinvhslh0}. Simulation in lens model.}
-
-\plot{ddiffrecplh0}{width=\textwidth}{Difference between gathers
-  displayed in Figures \ref{fig:drecplh0} and \ref{fig:drerecplh0},
-  plotted on same color scale.}
-
-\plot{dfwdvzhh0}{width=\textwidth}{Vertical velocity gather, generated
-  with a pressure source in the homogeneous model, 
-  corresponding to pressure gather \ref{fig:dfwdphh0}.}
-
-\plot{daltvzhh0}{width=\textwidth}{Vertical velocity gather, generated
-  with a velocity source in the homogenous model, corresponding to
-  pressure gather \ref{fig:daltphh0}.}
-
-\plot{dsvcomphh0}{width=\textwidth}{Difference between velocity gathers
-  shown in Figures \ref{fig:dfwdvzhh0} and \ref{fig:daltvzhh0},
-  plotted on the same color scale as these figures.}
-
-\plot{lamsstlamrdrecphh0}{width=\textwidth}{Inversion of data shown in
-  Figure \ref{fig:drecphh0}, simulated in homogeneous model, using the
-  approximate unitarity relation \ref{eqn:unitary} and propagation in
-  homogenous model.}
-
-\plot{difflamsstlamrdrecphh0}{width=\textwidth}{Difference between
-  data displayed in Figures \ref{fig:dinvhshh0} and
-  \ref{fig:lamsstlamrdrecphh0}, plotted on the same color scale.}
-
-\plot{lamsstlamrdrecplh0}{width=\textwidth}{Inversion of data shown in
-  Figure \ref{fig:drecplh0}, simulated in lens model, using the
-  approximate unitarity relation \ref{eqn:unitary} and propagation in
-  homogenous model.}
-
-\plot{difflamsstlamrdrecplh0}{width=\textwidth}{Difference between
-  data displayed in Figures \ref{fig:dinvhslh0} and
-  \ref{fig:lamsstlamrdrecplh0}, plotted on the same color scale.}
-
-\plot{preddinvhshh0}{width=\textwidth}{Pressure source gather = image
-  of pressure-to-source operator $\Lambda^+_{z_s}$ applied to pressure gather
-  shown in Figure \ref{fig:dsrcphh0}, homogeneous model. Compare Figure
-  \ref{fig:dinvhshh0}.}
-
-\plot{ddiffinvhshh0}{width=\textwidth}{Difference between (a) image
-  (Figure \ref{fig:preddinvhshh0}) of $\Lambda^+_{z_s}$ applied to
-  pressure gather (Figure \ref{fig:dsrcphh0}), and (b) source gather
-  (Figure \ref{fig:dinvhshh0}) inverted from receiver pressure gather
-  (Figure \ref{fig:drecphh0}). Homogeneous model used in all
-  propagations. Same color scale as in Figure
-  \ref{fig:preddinvhshh0}. }
-
-\plot{dpredhsrecphh0}{width=\textwidth}{Pressure gather at receiver 
-  datum $z=z_r=1000$ m simulated in homogeneous model from source 
-  gather shown in Figure \ref{fig:preddinvhshh0}, also based on the 
-  homogeneous model. Compare with point source pressure gather (Figure 
-  \ref{fig:drecphh0}).}
-
-\clearpage
-
-\plot{ddiffpredhsrecphh0}{width=\textwidth}{Plot of difference 
-  between data shown in Figures \ref{fig:drecphh0} and 
-  \ref{fig:dpredhsrecphh0}, plotted on same color scale as the latter 
-  two figures.}
-
-\plot{preddinvhsll0}{width=\textwidth}{Pressure source gather = image
-  of pressure-to-source operator $\Lambda^+_{z_s}$ applied to pressure gather
-  shown in Figure \ref{fig:dsrcphh0}, lens model. Compare Figure
-  \ref{fig:dinvhsll0}.}
-
-\plot{ddiffinvhsll0}{width=\textwidth}{Difference between (a) image
-  (Figure \ref{fig:preddinvhsll0}) of $\Lambda^+_{z_s}$ applied to
-  pressure gather (Figure \ref{fig:dsrcphh0}), and (b) source gather
-  (Figure \ref{fig:dinvhsll0}) inverted from receiver pressure gather
-  (Figure \ref{fig:drecplh0}). Lens model used in all propagations.
-  Same color scale as in Figure \ref{fig:preddinvhsll0}.}
-
-\plot{dpredhsrecpll0}{width=\textwidth}{Pressure gather at receiver 
-  datum $z=z_r=1000$ m simulated in lens model from source 
-  gather shown in Figure \ref{fig:preddinvhsll0}, also based on the 
-  homogeneous model. Compare with point source pressure gather (Figure 
-  \ref{fig:drecplh0}).}
-
-\plot{ddiffpredhsrecpll0}{width=\textwidth}{Plot of difference 
-  between data shown in Figures \ref{fig:drecplh0} and 
-  \ref{fig:dpredhsrecpll0}, plotted on same color scale as the latter 
-  two figures.}
-
-\clearpage
-
-\plot{preddnshshh0}{width=\textwidth}{Pressure source gather = image
-  under pressure-to-source operator $\Lambda^+_{z_s}$ of pressure gather
-  shown in Figure \ref{fig:dsrcphh0}, homogeneous model, using
-  ``near'' receiver traces at $z=2900$ m. Compare Figure
-  \ref{fig:dhshh0} and \ref{fig:dinvhshh0}: because the sources and
-  receivers are close, little aperture is lost in this case.}
-
-\plot{ddiffnshshh0}{width=\textwidth}{Difference between (a) image
-  (Figure \ref{fig:preddnshshh0}) of $\Lambda^+_{z_s}$ applied to
-  pressure gather (Figure \ref{fig:dsrcphh0}) using a near receiver
-  array to implement formula \ref{eqn:lamident}, and (b) source gather
-1650  (Figure \ref{fig:dhshh0}) inferred from vertical velocity.
-  Homogeneous model used in all
-  propagations. Same color scale as in Figure
-  \ref{fig:preddnshshh0}. }
-
-\plot{dprednshsrecphh0}{width=\textwidth}{Pressure gather at receiver
-  datum $z=z_r=1000$ m simulated from source gather shown in Figure
-  \ref{fig:preddnshshh0}. Compare with point source pressure gather
-  (Figure \ref{fig:drecplh0}).}
-
-\clearpage
-
-\plot{ddiffprednshsrecphh0}{width=\textwidth}{Plot of difference 
-  between data shown in Figures \ref{fig:drecphh0} and 
-  \ref{fig:dpredhsrecphh0}, plotted on same color scale as the latter 
-  two figures.}
-
-\plot{preddnshstrhh0}{width=\textwidth}{Pressure source gather = image
-  under {\em transpose} of pressure-to-source operator
-  $\Lambda^+_{z_s}$ of pressure gather shown in Figure
-  \ref{fig:dsrcphh0}, homogeneous model, using ``near'' receiver
-  traces at $z=2900$ m. Compare Figure \ref{fig:dhshh0} and
-  \ref{fig:preddnshshh0}: as noted in the text, $\Lambda^+_{z_s}$ is
-  asymptotically symmetric, so the resemblance is not a surprise.}
-
-\plot{ddiffnslamtrhh0}{width=\textwidth}{Difference between data in
-  Figures \ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}, plotted
-  on the same scale as these figures, showing that the asymptotic
-  symmetry of $\Lambda^+_{z_s}$ is actually quantitative for the
-  length, time and frequency scales of these examples.}
-
-\plot{symmdnshshh0}{width=\textwidth}{Pressure source gather = image
-    under {\em symmetrized} pressure-to-source operator
-    $\frac{1}{2}\left(\Lambda^+_{z_s}+(\Lambda^+_{z_s})^T\right)$ of
-    pressure gather shown in Figure \ref{fig:dsrcphh0}, homogeneous
-    model, using ``near'' receiver traces at $z=2900$ m. Compare
-    Figure \ref{fig:preddnshshh0}.}
-
-\plot{ddiffsymmdnshsll0}{width=\textwidth}{Difference between point
-  source simulation (Figure \ref{fig:dfwdplh0}) and pressure gather at
-  $z=z_r=1000$ m produced by simulation with the source shown in
-  Figure \ref{fig:symmdnshshh0},
-  propagation in the lens model.}
-
-\plot{compres0hh0}{width=\textwidth}{Comparison of residual Euclidean 
-  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
-  homogeneous model, point source (Figure \ref{fig:recphh0}), inversion 
-  in homogeneous model. Penalty weight
-  $\alpha=0$.}
-
-\plot{compres0lh0}{width=\textwidth}{Comparison of residual Euclidean 
-  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
-  lens model, point source (Figure \ref{fig:recplh0}), inversion 
-  in homogenous model. Penalty weight 
-  $\alpha=0$.}
-
-\plot{compres0ll0}{width=\textwidth}{Comparison of residual Euclidean 
-  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
-  lens model, point source (Figure \ref{fig:recplh0}), inversion 
-  in lens model. Penalty weight 
-  $\alpha=0$.}
-
-\plot{compres1hh0}{width=\textwidth}{Comparison of residual Euclidean 
-  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
-  homogeneous model, point source (Figure \ref{fig:recphh0}), inversion 
-  in homogenous model. Penalty weight 
-  $\alpha=10^{-3}$.}
-
-\plot{compres1lh0}{width=\textwidth}{Comparison of residual Euclidean 
-  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
-  lens model, point source (Figure \ref{fig:recplh0}), inversion 
-  in homogenous model. Penalty weight 
-  $\alpha=10^{-3}$.}
-
-\plot{compres1ll0}{width=\textwidth}{Comparison of residual Euclidean 
-  norms: CG (blue), PCG (red), plotted vs. iteration. Data =
-  lens model, point source (Figure \ref{fig:recplh0}), inversion 
-  in lens model. Penalty weight
-  $\alpha=0$.}
-
-\plot{compnres0hh0}{width=\textwidth}{Comparison of normal residual
-  (gradient) Euclidean norms: CG (blue), PCG (red), plotted
-  vs. iteration. Data = homogeneous model, point source (Figure
-  \ref{fig:recphh0}), inversion in homogenous model. Penalty weight $\alpha=0$.}
-
 \plot{compnres0lh0}{width=\textwidth}{Comparison of normal residual 
@@
 
-\plot{compnres0ll0}{width=\textwidth}{Comparison of normal residual 
-  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
-  vs. iteration. Data = lens model, point source (Figure 
-  \ref{fig:recplh0}), inversion in lens model. Penalty weight 
-  $\alpha=0$.}
-
-\plot{compnres1hh0}{width=\textwidth}{Comparison of normal residual 
-  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
-  vs. iteration. Data = homogeneous model, point source (Figure 
-  \ref{fig:recphh0}), inversion in homogenous model. Penalty weight 
-  $\alpha=10^{-3}$.}
+Figure \ref{fig:compnres1lh0} shows the same comparison with non-zero
+penalty weight, $\alpha=10^{-3}$. The PCG normal residual curve is
+almost identical with that in the $\alpha=0$ case, wheras the CG
+convergence has slowed down noticeably, being about five times as slow
+as the preconditioned algorithm.
 
@@
 
-\plot{compnres1ll0}{width=\textwidth}{Comparison of normal residual 
-  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
-  vs. iteration. Data = lens model, point source (Figure 
-  \ref{fig:recplh0}), inversion in lens model. Penalty weight
-  $\alpha=10^{-3}$.}
-
-\newpage
+\section{Conclusion}
 

### 2022-03-29 repro/papers/sseprecond/paper.tex
test commit
added some figs

@@
 
+\plot{dfwdvzlh0}{width=\textwidth}{Vertical velocity gather, generated 
+  with a pressure source in the lens model, 
+  corresponding to pressure gather \ref{fig:dfwdphh0}.}
+
 \plot{daltvzhh0}{width=\textwidth}{Vertical velocity gather, generated 
@@
 
+\plot{daltvzlh0}{width=\textwidth}{Vertical velocity gather, generated 
+  with a velocity source in the lens model, corresponding to 
+  pressure gather \ref{fig:daltphh0}.}
+
 \plot{dsvcomphh0}{width=\textwidth}{Difference between velocity gathers
   shown in Figures \ref{fig:dfwdvzhh0} and \ref{fig:daltvzhh0},
+  plotted on the same color scale as these figures.}
+
+\plot{dsvcomplh0}{width=\textwidth}{Difference between velocity gathers
+  shown in Figures \ref{fig:dfwdvzlh0} and \ref{fig:daltvzlh0},
   plotted on the same color scale as these figures.}
@@
 
+\plot{dprednshsrecpll0}{width=\textwidth}{Pressure gather at receiver
+  datum $z=z_r=1000$ m simulated from source gather shown in Figure
+  \ref{fig:preddnshshh0}. Compare with point source pressure gather
+  (Figure \ref{fig:drecplh0}).}
+
 \plot{ddiffprednshsrecphh0}{width=\textwidth}{Plot of difference 
   between data shown in Figures \ref{fig:drecphh0} and 
-  \ref{fig:dpredhsrecphh0}, plotted on same color scale as the latter 
+  \ref{fig:dprednshsrecphh0}, plotted on same color scale as the latter 
+  two figures.}
+
+\plot{ddiffprednshsrecpll0}{width=\textwidth}{Plot of difference 
+  between data shown in Figures \ref{fig:drecplh0} and 
+  \ref{fig:dprednshsrecplh0}, plotted on same color scale as the latter 
   two figures.}

### 2022-04-03 repro/papers/RevGEO22-0070/review.tex
final version

%% This file is automatically generated. Do not edit!
% \documentclass[georeport,12pt]{geophysics}
\documentclass[12pt]{article}

\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amscd}
\usepackage{setspace}
\usepackage{wasysym}
\usepackage{algorithm}
\usepackage{algpseudocode}

\newcommand{\mb}{\mathbf}
\newcommand{\ba}{\mathbf{a}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\bxi}{\mathbf{\xi}}
\newcommand{\bk}{\mathbf{k}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bff}{\mathbf{f}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bz}{\mathbf{z}}
\newcommand{\bh}{\mathbf{h}}
\newcommand{\bR}{\mathbf{R}}

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{alg}{Algorithm}
\newtheorem{remark}{Remark}
\newtheorem{cor}{Corollary}
\newtheorem{example}{Example}
\newtheorem{definition}{Defenition}

\begin{document}
\title{Review of ``Waveform inversion via reduced order modeling'', GEO22-0070}
\maketitle
%\author{anonymous}

%\begin{abstract}
%  blah blah blah (edit out of .ltx)
%\end{abstract}

Reduced order modeling (ROM) refers to a collection of techniques for control or parameter identification of dynamical systems that approximate the dynamics by projection of the system state onto a collection of time samples, or snapshots. In effect, the collection of snapshots taken from the time history itself is used as an approximation to the full time history, and a reduced dynamics deduced that operates on this collection. The dimension (order) is smaller, perhaps much smaller, and control of the dynamics and or parameter estimation is computationally simpler that is the case with the original model. Observation of the snapshots is ordinarily a requirement.

Application of ROM to waveform inversion would seem to face an immediate obstacle: the system state is the seismic wavefield, unobservable except (perhaps) in a very few places in the earth's interior, and on or near the surface. The approach to inversion described in this work relies on two clever observations about the structure of wavefields that permit the construction of a Galerkin projection of wave dynamics onto the span of the snapshots, directly from surface data - {\em without access to the snapshots}. The idea has been described in a number of publications, and applied in various ways. The contribution of this paper is to show how the ROM construction can lead to velocity estimation, perhaps avoiding the notorious cycle-skipping obstacle of FWI. The paper describes in detail a velocity inversion algorihm based on ROM, and illustrates it with several examples, based on the well-known Tarantola Camembert and Marmousi 2D models.

The ROM construction requires a good deal of set-up, converting a familiar type of acoustic model for seismic data into a rather unfamiliar one, and relying (apparently) on a rather strict set of assumptions. I will first give an overview of the inversion approach, concentrating on the core constructions that make it quite different from other approaches to inversion. In effect, the result is a remarkable representation of the data in the form of a reduced-order model of the acoustic wave operator. The rest of the algorithm is involved but more conventional, estimating the velocity field by regularized least-squares fit to this data-derived reduced-order operator representation. After this overview of the methodology, I make some observations and pose some questions for the authors to consider.

\section{Method}

The acoustic pressure field is assumed to be a solution of the constant-density wave equation. Sources (right-hand sides) take the form form $\theta(\bx-\bx_s)\partial_tf(t)$, in which $f$ has a non-negative, even Fourier transform, that is, is an autocorrelation (always achievable, assuming the the physical source is known). The spatial factor $\theta(\bx)$ is zero outside of a ``small'' ball centered at $\bx=0$, so that the sensors and sources have localized support near their nominal locations. The pressure field is scaled by $1/c(\bx)$, which has the effect of replacing the spatial operator $c(x)^2\nabla^2$ in the wave equation by the symmetric operator ${\cal A}=c(\bx)\nabla^2c(\bx)$. The causal pressure field $p^{(s)}$ for the source at $\bx_s$ is replaced by the symmetric-in-time field $w^{(s)}(\bx,t)=(p^{(s)}(\bx,t) + p^{(s)}(\bx,-t))/c(\bx)$.

Data traces are defined by averaging the time-symmetrized pressure field with the same kernel $\theta$ used to define the source:
\[
  D^{(r,s)}(t) = \int dx \frac{\theta(\bx-\bx_r)}{c(\bx_r)}w^{(s)}(\bx,t)
\]
It is assumed tht the sets $\{\bx_s\}$ and $\{\bx_r\}$ of source and receiver positions are the same, so that $D$ is a square $m \times m$ matrix of traces. Note that these are not the traces of a causal pressure field resulting from localized sources, but can be computed from such ``normal'' traces, given the validity of the assumptions outlined above.

Finally, $w$ is filtered by a local operator defined as a function of ${\cal A}$ (equation 11). This transformation (based on Duhamel's principle) converts $w$, which satisfies a wave equation with homogeneous initial values and a energy input through a right-hand side, into $u$, which satisfies a wave equation with zero right hand side and energy input through an initial condition:
\begin{eqnarray}
  \partial_t^2 u^{(s)} + {\cal A}u^{(s)} & = & 0, \nonumber \\
  u^{(s)}(\bx,0) &=& u^{(s)}_0(\bx), \nonumber \\
  \partial_t u^{(s)}(\bx,0) &=& 0.
  \label{eqn:ivp}
\end{eqnarray}
A consequence of this construction is a key relation between the data and the transformed field $u$ (equation 16, more or less):
\[
  D^{(r,s)}(t) = \int dx u^{(r)}_0(\bx) u^{(s)}(\bx,t) = \langle u^{(r)}_0, u^{(s)}(t) \rangle
\]
where the angle brackets on the right signify the $L^2$ (Euclidean) inner product.
That is, averaging $w^{(s)}$ against $\theta(\bx-\bx_r)/c(\bx_r)$ is equivalent to averaging $u^{(s)}$ against the $t=0$ field $u_0^{(r)}$.

One of the fundamental ideas of this work is that you can represent the Green's function of problem \ref{eqn:ivp} via the so-called functional calculus, as an operator cosine:
\begin{equation}
  \label{eqn:cost}
  u^{(s)}(t,\bx) = \cos ( t\sqrt{{\cal A}})u_0^{(s)}(\bx)
\end{equation}
This would be obvious if ${\cal A}$ were a number. It is also true for symmetric operators like ${\cal A}$ (officialliy, once they are made into self-adjoint operators by proper choice of domain), as a consequence of the spectral theory of such operators, as is explained in the manuscript.

After these preliminaries, the ROM is based on orthogonal projection of the vector of fields $u^{(s)}(\bx,t), s=0,...m-1$ onto the subspace spanned by the first $n$ snapshots at time interval $\tau$, i.e. $u^{(s)}(\bx,j\tau), s=0,...m-1,j=0,...n-1$. Organize the fields into a block vector $u(\bx,t) = (u^{(0)}(\bx,t),..., u^{(m-1)}(\bx,t))$, and the snapshots into the $nm$ block (row) vector $U(\bx)$, $U_{sm+j}(\bx)= u^{(s)}(\bx,j\tau)$. The block matrix projection (Galerkin approximation) may be written
\[
  u^{\rm gal}(\bx,t) = U(\bx)g(t)
\]
in which $g(t)=(g_0(t),...,g_{n-1}(t))$ is the $nm\times m$ matrix of Galerkin coefficients, chosen so that the residual in the wave operator applied to $u^{\rm gal}$ is orthogonal to the span of the elements of $U$. That is,
\[
 \langle U^T,(\partial_t^2 Ug + {\cal A}Ug)\rangle = \partial_t^2 \langle U^T,U \rangle \partial_t^2 g + \langle U^T, {\cal A}U\rangle g
\]
\begin{equation}
  \label{eqn:gal}
  = {\bf M}\partial_t^2 g + \tilde{\bf S}g = 0
\end{equation}
where the $nm \times nm$ Galerkin mass and stiffness matrices are
\[
  {\bf M} = \langle U^T,U \rangle = \int dx U^T(\bx)U(\bx), 
\]
\[
\tilde{\bf S} = \langle U^T,{\cal A}U\rangle
\]

This is all very well, and the stiffness matrix $\tilde{\bf S}$ represents ${\cal A}$ and contains information about the velocity field $c$, though in a somewhat non-transparent way, since the snapshot vector $U$ also depends on $c$. However the snapshots $U$ are precisely the missing ingredient - they represent the acoustic field in the interior of the earth. They could be calculated if one knew $c$, by solving the problem \ref{eqn:ivp} numerically, however of course $c$ is precisely the unknown being sought.

This difficulty is overcome through two very clever observations. The first rests on the high-school identity
\[
  \cos a \cos b = \frac{1}{2}(\cos(a+b) + \cos(a-b))
\]
From the Green's function representation \ref{eqn:cost}, suppressing $\bx$) and grouping the fields for all sources together into the $m$-vector $u$,
\begin{equation}
  \label{eqn:cosj}
  u_j=u(j\tau) = \cos(j\tau\sqrt{{\cal A}})u_0
\end{equation}
So the $(i,j)$ $m \times m$ block of the $nm \times nm$ mass matrix ${\bf M}$ can be re-written (equation 31)
\[
  {\bf M}_{i,j} = \langle u_i, u_j \rangle = \langle \cos(i\tau\sqrt{{\cal A}})u_0, \cos(j\tau\sqrt{{\cal A}})u_0\rangle
\]
\[
  =\langle u_0,\cos(i\tau\sqrt{{\cal A}})\cos(j\tau\sqrt{{\cal A}})u_0\rangle
\]
(since real functions of the symmetric operator ${\cal A}$ are symmetric and commute)
\[
  =\frac{1}{2}\langle u_0,[\cos((i+j)\tau\sqrt{{\cal A}})+\cos(|i-j|\tau\sqrt{{\cal A}})u_0\rangle
\]
\[
  =\frac{1}{2}(D_{i+j} + D_{|i-j|}),
\]
where we have written $D_j=D(j\tau)$.

In other words, {\em the mass matrix, of inner products of acoustic fields in the interior of the earth, can be calculated directly from the data}.

The same is true of the stiffness matrix $\tilde{\bf S}$, by a similar calculation (equation
49):
\[
  \tilde{\bf S}_{i,j}=-\frac{1}{2}((\partial_t^2 D)_{i+j} + (\partial_t^2 D)_{|i-j|}),
\]
where the second time derivative traces $\partial_t^2 D$ are computed by any convenient numerical approximation.

The stiffness matrix contains information about $c$, but that is still hidden behind the interior fields $U$, in view of the defintiion. To tease it out, introduce block  Gram-Schmidt orthogonalization of the vector $U = VR$. Here $V$ has orthogonal $m$-dimensional time blocks, and $R$ is an $nm \times nm$ dimensional block upper triangular block matrix. Then you can re-write the left-hand side of equation \ref{eqn:gal} as
\[
  U^T(U\partial^2_t g + {\cal A} Ug) = R^TR \partial^2_t g + R^TV^T{\cal A}VR g = 0
\]
Note in particular,
\begin{equation}
  \label{eqn:chol}
  R^TR = {\bf M}.
\end{equation}
Define the ROM fields $\tilde{u}^{\rm ROM} = Rg$. Then
\[
  \partial^2 \tilde{u}^{\rm ROM} + {\cal A}^{\rm ROM} \tilde{u}^{\rm ROM} = 0
\]
where ${\cal A}^{\rm ROM} = V^T{\cal A}V$ is the ``ROM of the wave operator'', out of which will be extracted an estimate of $c$.

This still appears to be unsatisfactory, as the definition of ${\cal A}^{\rm ROM}$ involves the orthogonalized interior fields. Now the second very clever observation intervenes: $R$ is block upper triangular, so equation \ref{eqn:chol} expresses the block Cholesky decomposition of ${\bf M}$. Thus $R$ can be computed from ${\bf M}$ by a block version of Gram-Schmidt orthogonalization, for instance. But ${\bf M}$ is computable directly from the data, therefore the same is true of $R$. $\tilde{\bf S}$ is also computable directly from the data, and
\[
  {\cal A}^{\rm ROM} = V^T{\cal A}V = (R^T)^{-1}\tilde{\bf S}R^{-1},
\]
whence ${\cal A}^{\rm ROM}$ {\em is computable directly from the data, entirely without knowledge of the interior fields}.

It now remains to extract the velocity, or an approximation, from ${\cal A}^{\rm ROM}$. The method used is basically straightforward nonlinear least squares with Tikhonov regularization. The version of ${\cal A}^{\rm ROM}$ computed from the data as just described is treated as the data in this least squares problem. The prediction operator maps a trial velocity $v$ to synthetic data, whence to a synthetic ROM operator ${\cal A}^{\rm ROM}(v)$. The Frobenius norm of the difference of these two operators is then minimized by updating $v$ using a combination of layer-stripping and Gauss-Newton algorithms. Of course there are number of twists: for instance, the decay of the matrix entries of ${\cal A}^{\rm ROM}$ away from the diagonal (shown in Appendix B) is used to reduce the computational complexity of the algorithm.

\section{Observations}
I offer a few questions and comments.

\subsection{Cycle-skipping mitigation - why?}
The examples, based on Tarantola's Camembert and the Marmousi model, are impressive and certainly suggest that avoidance of cycle-skipping is somehow built into this approach to velocity inversion. However the mechanics of the ROM construction are sufficiently unfamiliar to this reviewer that he has no intuitive understanding of why this should be so. I believe the same will be true for most, or even all, readers of {\em Geophysics} who tackle this work. What is the mechanism by which cycle-skipping is avoided? I am not asking for proof - there are very few informative theoretical results about convexity or monomodality in the literature on wave inverse problems, and the results that do exist are mostly not so useful. I merely suggest that a plausibility argument in favor of cycle-skipping mitigation would make this paper considerably more valuable.

\subsection{Data restrictions}
From the outset, this paper presumes that sources and receivers occupy the same positions and are parametrized in the same way. Such assumptions are natural and more or less correct in some laboratory ultrasonics settings, in which the sources and receivers are the same transducers, and perhaps in some other settings, for example in ocean acoustics. However the same cannot be said of seismic data acquisition, where source and receiver positions often do not coincide and physical characteristics of sources and receviers (frequency response, radiation pattern,...) usually differ, often considerably. That is, Green's functions are reciprocal, but data often are not. Can this approach be modified to deal directly with the non-reciprocal data usually encountered in all forms of seismology? The algebraic rigidity of the ROM construction suggests that the answer is ``no''. If that is the case, then a form of data completion and direction-dependent signature modification would be required as preprocess. As a simple initial exercise without the complication of radiation pattern modification, the authors could consider what modifications would be necessary to use data in the original Marmousi streamer geometry. This is not a minor point - it bears on whether this approach can ever be practical for seismic inversion.

Note that, for all of its defects, FWI naturally accommodates more or less arbitrary data geometry and source/receiver characteristics. Of course, the latter may not be known with the necessary precision, but in fact these are nowadays often incorporated into the parameters to be determined by FWI. To become useful, the ROM construction will have to acquire some of this flexibility.

\subsection{Linear independence of snapshots}
A major assumption, that the snapshots are linearly independent, appears as a footnote on p. 10: ``The space S is nm-dimensional because the snapshots $u_j(\bx)$ for $j=0,..,n$ are assumed linearly independent.'' Linear independence is an open property, and seems reasonable to assume, but an assumption on which the entire construction rests does not belong in a footnote. Also, one must ask why this assumption is reasonable. A hint appears on p. 31, in a footnote to the statement ``Because in the ROM construction we need the wavefront to progress downward at all times...''; the footnote is ``If this does not hold, the snapshots that define the approximation space S may be linearly dependent.'' That is extremely interesting. First, what wavefronts? Does the remark refer to the wavefronts of geometric asymptotics, i.e. level sets of geodesic distance, emitted by the sources? In that case, it sounds like the so-called ``DSR'' assumption underlying one-way prestack migration. The Marmousi model produces many overturned wavefronts, yet the ROM-based inversion succeeds. Is it only some part of the geometric wavefronts that must ``progress downwards''? Or does this statement not pertain to geometric wavefronts at all? This is a fundamental point, and the authors really need to clear it up.

\subsection{Computational complexity}
The authors do not directly address this topic at all. The algorithm involves a number of disparate subsidiary calculations - time domain finite difference simulation, QR factorization of large matrices, Gauss-Newton iteration, etc. The reader cannot be expeccted to estimate the computational cost of this algorithm or the factors on which it depends. The authors should add a serious attempt to assess the cost - maybe per Gauss-Newton itertion in each layer computation with reasonable off-diagoal truncation, and some rough estimate of overall cost and its critical parameters.

The straw-man algorithm, FWI, is now a commercial product, applied many times to 3D field data. What is a rough estimate of cost for a 3D ROM inversion?

\subsection{Audience}
As currently written, comprehending this paper requires some mathematical background, at the level of basic graduate courses in functional and numerical analysis. The vast majority of {\em Geophysics} readers have not acquired this background. Some attempt to compensate for this disparity would likely enlarge the pool of interested readers. For example, a graphical representation of block Cholesky decomposition would give the reader with minimal background in numerical linear algebra a fighting chance.

\subsection{Miscellaneous}
The construction of the ROM propagator is really irrelevant, and could usefully be removed.

Can this construction be applied to transmission data? For instance, what about one of the other Camembert experiments from the Gauthier et al. paper, in which the receivers are on the opposite side (in addition, to keep the coincident source-receiver hypothesis valid)?

\section{Conclusion}
This paper presents truly interesting and original work, using concepts from numerical linear algebra and functional analysis to recast the velocity inversion problem in a form that appears to avoid cycle-skipping, at least in some cases. In its current form, it has two major drawbacks as a contribution to {\em Geophysics}: it presumes background that most readers do not have, and makes assumptions about data that are simply not true of almost all data collected in the field. If the authors address the questions and comments I have listed here, they will go a long ways towards making a revised version suitable for publication in {\em Geophysics}.

\end{document}

### 2022-04-03 repro/papers/GEM22/submit/paper.tex
original submission

--- 2022-03-29 repro/papers/GEM22/paper.tex
@@
-\title{Efficient Computation of Extended Sources}
-\author{William. W. Symes \thanks{The Rice Inversion Project,
-Department of Computational and Applied Mathematics, Rice University,
-Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}
+\documentclass[georeport,12pt]{geophysics}
+
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{amsthm}
+\usepackage{amscd}
+\usepackage{setspace}
+\usepackage{wasysym}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+
+\newcommand{\mb}{\mathbf}
+\newcommand{\ba}{\mathbf{a}}
+\newcommand{\bx}{\mathbf{x}}
+\newcommand{\bxi}{\mathbf{\xi}}
+\newcommand{\bk}{\mathbf{k}}
+\newcommand{\bv}{\mathbf{v}}
+\newcommand{\bff}{\mathbf{f}}
+\newcommand{\bu}{\mathbf{u}}
+\newcommand{\by}{\mathbf{y}}
+\newcommand{\bz}{\mathbf{z}}
+\newcommand{\bh}{\mathbf{h}}
+\newcommand{\bR}{\mathbf{R}}
+
+\newtheorem{lemma}{Lemma}
+\newtheorem{theorem}{Theorem}
+\newtheorem{alg}{Algorithm}
+\newtheorem{remark}{Remark}
+\newtheorem{cor}{Corollary}
+\newtheorem{example}{Example}
+\newtheorem{definition}{Defenition}
+
+\setfigdir{.}
+
+\begin{document}
+\title{Efficient Computation of Extended Surface Sources}
+\author{William W. Symes\\
+ %\thanks{The Rice Inversion Project,
+  Department of Computational and Applied Mathematics,\\
+  Rice University,Houston TX 77251-1892 USA,\\
+  email {\tt symes@rice.edu},\\
+ORCID 0000-0001-6213-4272}
 
@@
 
-\inputdir{../sseprecond/project}
+\noindent {\bf Keywords:} inverse problems, wave propagation, time
+reversal, Krylov subspace methods, preconditioning
+
+% \inputdir{../sseprecond/project}
+\inputdir{.}
 
@@
 \cite{Payn:75,Symes:83,Lasi:86,LasLionsTrig:86,Lasi:87, BaoSy:91b}. 
-Since the purpose of this paper is to explore the algorithmic
+Elaboration of these mathematical details is beyond the scope of this
+paper, which aims instead to explore the algorithmic
 consequences of the mathematical structure implied by the downgoing
-condition, I leave the elaboration of the mathematical details to
-another place.
+condition.
 
@@
 express the approximate inverse as the modeling operator adjoint in
-weighted norms (thus establishing that the modeling operator is {\em
+weighted norms (thus establishing that the modeling operator is {\em approximately
   unitary} in the sense of these norms), explain how to use this
@@
 organize the preconditioning computation so as to involve only one
+extra and relatively inexpensive
 wave propagation calculation. The penultimate section displays simple
@@
 \end{equation}
-$L[{\bf c}]$ has a well-defined inverse if it is restricted to either
-causal or anti-causal vector wavefields.
+$L[{\bf c}]$ has a well-defined inverse in the sense of distributions
+if it is restricted to either causal or anti-causal vector wavefields.
 
@@
 
+As explained in the Introduction, the downgoing constraint on the square-integrable
+source functions $h_s, f_s$ is
+essential both for finite energy solutions of the system
+\ref{eqn:awepm} to exist, and for these solutions to have well-defined
+traces on the receiver surface $z=z_r$. This constraint will be
+assumed throughout, often tacitly.
 For downgoing solutions of system \ref{eqn:awepm}, the key
@@
 \end{equation}
-Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that
-
+Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj1} and \ref{eqn:trsadj} imply that
 \begin{equation} 
@@
 \end{equation}
-
 The relation \ref{eqn:trtr} implies that
@@
   \label{eqn:defae0}
-  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)
+  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)(\bx,t).
 \end{equation}
@@
 high-frequency sense. This fact follows from a
-geometric optics analysis of the half-space solution,
-assuming (as always) downgoing data:
+geometric optics analysis of the half-space solution. This leads to
+the identification of $\Lambda^{\pm}_{z_s}$ as a {\em
+  pseudodifferential operator} of order zero on $z=z_s$, with principal symbol
+\begin{equation}
+  \label{eqn:lamsym}
+  \sigma_0(\Lambda^{\pm}_{z_s}) = \pm 2(\kappa(\bx)
+  \rho(\bx))^{1/2}\left(1-\frac{\kappa(\bx)(\xi^2+\eta^2)}{\rho(\bx)\omega^2}\right)^{-1/2}.
+\end{equation}
+Here $\xi$, $\eta$, and $\omega$ are the dual Fourier variables to
+$x$, $y$, and $t$ respectively. The downgoing assumptions means that
+for local planewave components of $P_sp$, the quantity inside the
+square root is positive. Thus $\Lambda^{\pm}_{z_s}$ has real principal
+symbol (in fact, the entire symbol is real) hence defines an
+asymptotically symmetric operator:
 \begin{equation}
@@
 \end{equation}
-
+(For more on this, see \cite{StefUhl:05}.)
 The analysis also reveals that the solution components not continuous
@@
 
-Similar relations have been derived for other scattering operators,  
-and have been used to accelerate iterative solutions of inverse  
-scattering problems: \cite{DafniSymes:SEG18b} review some of this  
-literature.
-
 The most convenient arrangement the Conjugate Gradient (CG) algorithm
 taking advantage of the structure \ref{eqn:wadj} is the {\em
-  Preconditioned CG}.
-
-Allowing that the fit error will be measured by the data space norm,
-the least squares problem to be solved is not just $Sh \approx d$, but
-a regularized version:
+  Preconditioned CG}. Allowing that the fit error will be measured by
+the data space norm, the least squares problem to be solved is not
+just $Sh \approx d$, but a regularized version:
 \begin{equation}
@@
 \end{equation}
-Remark: recall that the modified data space norm $\|d\|_d^2 = \langle
+
+\noindent {\bf Remark:} recall that the modified data space norm $\|d\|_d^2 = \langle
 d, W_d d\rangle$ has physical meaning: for acoustics, it is
@@
 the difficulties with the definition of $W_m$, mentioned above, are
-taken care of), so that $A^{\dagger} \approx A^T$. Therefore
+taken care of). Scalar pseudodifferential operators approximately
+commute, so $A^{\dagger} \approx A^T$. Therefore
 \begin{equation}
@@
 \end{equation}
-Since $W_m$ is self-adjoint and positive definite, the common factor on both sides of \ref{eqn:norm1} can be re-written as
+Since $W_m$ is self-adjoint and positive semidefinite, the common factor on both sides of \ref{eqn:norm1} can be re-written as
 \begin{equation}
@@
 \begin{eqnarray}
-  \label{eqn:sadj}
+  \label{eqn:sadjwt}
   S^* &=& S^T W_d,\\
@@
 designed. PCG for solution
-of equation \ref{eqn:normpart} with preconditioner $M$is usually
+of equation \ref{eqn:normpart} with preconditioner $M$ is usually
 written as Algorithm 1 (see for example \cite{Golub:2012}):
@@
 
-A deeper study of the pressure-to-source
-operator (Appendix B) shows that it is approximately dependent only on the model
-coefficients near the source surface ($z=z_s$ in this case). Since the
-homogeneous and lens models are identical near this surface, it is
-unsurprising that these figures are very close to the previous two.
-However an even more useful observation is that the calculations in
-the approximation \ref{eqn:lamident} could just as well be carried out
-in a much smaller region around the source surface, and produce a
-result that is functionally identical in that it will serve as a
-source for the same acoustic fields globally, with small error. In
-effect, equation \ref{eqn:lamident} involving propagation from source
-($z=z_s$) to receiver ($z=z_r$) surfaces is altered by replacing $z_r$
-with a receiver datum $z_s+\Delta z$ considerably closers to $z_s$:
+A deeper study of the pressure-to-source operator (or of the closely
+related Dirichlet-to-Neumann operator for the second order wave
+equation, see \cite{StefUhl:05}) shows that it is approximately
+dependent only on the model coefficients near the source surface
+($z=z_s$ in this case). Since the homogeneous and lens models are
+identical near this surface, it is unsurprising that these figures are
+very close to the previous two.  However an even more useful
+observation is that the calculations in the approximation
+\ref{eqn:lamident} could just as well be carried out in a much smaller
+region around the source surface, and produce a result that is
+functionally identical in that it will serve as a source for the same
+acoustic fields globally, with small error. In effect, equation
+\ref{eqn:lamident} involving propagation from source ($z=z_s$) to
+receiver ($z=z_r$) surfaces is altered by replacing $z_r$ with a
+receiver datum $z_s+\Delta z$ considerably closers to $z_s$:
 \begin{equation}
@@
 (as noted above, in the continuum limit, sources are represented via
-adjoint sampling). Steps in $x$ and $z$ are the same. Most results use
-$\Delta x = 20$ m, and limit the temporal frequency of the computed
-traces to 12.5 Hz.
+adjoint sampling). Steps in $x$ and $z$ are the same. In the following
+examples, $\Delta x = 20$ m. This choice limits the temporal frequency
+of accurately computed fields to rougly 12 Hz.
 
@@
   lens model. Scaled version of output $v_z$ field obtained by 
-  applying transpose of $4V^+_{z_s,z_r}$. Approximates dip-filtered
-  version of pressure source gather (Figure \ref{fig:dhshh0})}
+  applying transpose of $4V^+_{z_s,z_r}$. Quite different from 
+  pressure source gather (Figure \ref{fig:dhshh0}) used to generate
+  data - since inversion takes place in a different material model!}
 
@@
   with a velocity source in the lens model, corresponding to
-  pressure gather \ref{fig:daltphh0}.}
+  pressure gather \ref{fig:daltplh0}.}
 
 \plot{dsvcomplh0}{width=\textwidth}{Difference between velocity gathers
-  shown in Figures \ref{fig:dfwdvzhh0} and \ref{fig:daltvzhh0},
+  shown in Figures \ref{fig:dfwdvzlh0} and \ref{fig:daltvzlh0},
   plotted on the same color scale as these figures.}
@@
 Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
-\ref{fig:dprednshsrecplh0}, and \ref{fig:ddiffprednshsrecplh0}. The effect of aperture
+\ref{fig:dprednshsrecpll0}, and \ref{fig:ddiffprednshsrecpll0}. The effect of aperture
 limitation is clearly diminished: the second figure in this series
@@
   ``near'' receiver traces at $z=2900$ m. Compare Figure
-  \ref{fig:dhshh0} and \ref{fig:dinvhshh0}: because the sources and
+  \ref{fig:dhshh0}: because the sources and
   receivers are close, little aperture is lost in this case.}
@@
   between data shown in Figures \ref{fig:drecplh0} and 
-  \ref{fig:dprednshsrecplh0}, plotted on same color scale as the latter 
+  \ref{fig:dprednshsrecpll0}, plotted on same color scale as the latter 
   two figures.}
@@
 asymptotic approximations of each other. Figure
-\ref{fig:ddiffsymmdxnshsll0} shows the
+\ref{fig:ddiffsymmdnshsll0} shows the
  the difference between the pressure gather at $z=z_r$ produced from
@@
 to the data $S^+_{z_s,z_r}h_s$ (Figure \ref{fig:drerecplh0}), in which
-$h_s$ is the version (Figure \ref{fig:dinvhslh0}) of the
-downgoing source created earlier (Figure \ref{fig:dhslh0}), produced
-by inversion of the data in Figure \ref{fig:drecplh0} via the first
-equation in display \ref{eqn:approxinv}, propagation in the homogenous
-model.  The operator above is computed via the technique explained in
+$h_s$ is the %version (Figure \ref{fig:dinvhslh0}) of the
+downgoing source created earlier (Figure \ref{fig:dhshh0})%, produced
+%by inversion of the data in Figure \ref{fig:drecplh0} via the first
+%equation in display \ref{eqn:approxinv}, propagation in the homogenous
+%model.
+The operator above is computed via the technique explained in
 the preceding subsection, below, using
@@
 
-Figure \ref{fig:compnreslh0} shows the progress of the normal residual
+Figure \ref{fig:compnres0lh0} shows the progress of the normal residual
 (Euclidean norm of the difference of the two sides of equation \ref{eqn:norm1}),
@@
   vs. iteration. Data = lens model, point source (Figure 
-  \ref{fig:recplh0}), inversion in homogenous model. Penalty weight 
+  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
   $\alpha=0$.}
@@
   vs. iteration. Data = lens model, point source (Figure 
-  \ref{fig:recplh0}), inversion in homogenous model. Penalty weight 
+  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
   $\alpha=10^{-3}$.}
@@
 \section{Conclusion}
+
+The linear modeling operator of surface source extended acoustic
+waveform inversion is approximately invertible, and this paper has
+shown how to approximately invert it. The construction is based on
+reverse time propagation of data, as inspired by the literature on
+photoacoustic tomography. However, since the input energy comes from a
+surface source, rather than a pressure boundary value, the
+pressure-to-source operator intervenes. It provides not just an
+approximate inverse, but a definition of weighted norms in domain and
+range spaces of the modeling operator, in terms of which that operator
+is approximately unitary. Accordingly, Krylov space iteration defined
+in terms of these weighted norms, or equivalently preconditioned
+Conjugate Gradient iteration, gives a rapidly convergent solution
+method for the linear subproblem.
+
+The existence of an approximate unitary representation of the modeling
+operator is not merely a computational convenience, however. It
+reveals fundamental aspects of the operator's structure that enable
+an explanation for the mitigation of cycle-skipping, a feature of the
+{\em nonlinear} extended inverse problem. This fact echoes earlier observations
+concerned a reflected wave inverse problem, involving a modeling
+operator with a similar approximate inverse
+\cite[]{tenKroode:IPTA14,Symes:IPTA14}. Also, the approximate inverse
+leads to a stable computation of the gradient of the nonlinear
+objective function \ref{eqn:esi}, resolving a difficulty first noted
+also for reflected wave inversion \cite[]{KerSy:94}.
+
+The transmission inverse problem figuring most prominently in
+contemporary applied seismology is surely the Full Waveform Inversion
+(FWI) of diving wave data. This is essentially the same problem as the
+one discussed in this paper, and can be formulated and treated the
+same way, at least for acoustic wave physics. All of the topics
+treated here are open for elastic wave physics - the analogue of the
+pressure-to-source map would is the map from surface velocity field to
+corresponding constitutive defect, analogous to the elastic
+Dirichlet-to-Neumann map investigated by \cite{Rachele:00}.
+
+The underlying tool in the ideas developed here is geometric optics
+(or ray theory), without which the very concept of downgoing waves
+would be meaningless. The physics of actual earth materials includes
+material heterogeneity on all scales, which appears to leave little
+room for the assumption of scale separation underlying geometric
+optics. Moreover, earth materials are anelastic, with elastic wave
+energy being converted to and from thermal excitation, pore fluid
+motion, and so on. A truly satisfactory understanding of inverse wave
+problems will eventually need to accommodate heterogenity and
+anelasticity beyond the current capabilities of the ray-based theory.
 
@@
 the adjoint state method, in this case a by-product of the
-conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awepm}
+conservation of energy. This calculation leads to equation
+\ref{eqn:sadj1}, from which the other statements about adjoints made in the second
+section of the paper follow.
+
+Suppose that $p^-,\bv^-$ solve \ref{eqn:awepm}
 with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
@@
 0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
-( h_rP_rp^++f_rP_rv_z^+) = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
-\]
-whence
-\begin{equation}
-\label{eqn:sadj}
- ({\cal S}^+_{z_s,z_r})^T = -{\cal S}^{-}_{z_r,z_s}.
-\end{equation}
-
-The systems \ref{eqn:awepm} for $(p^{\pm},\bv^{\pm})$ with the two
-signs of superscript differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
-$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
-\begin{equation}
-  \label{eqn:trdef}
-  {\cal R} \left(
-    \begin{array}{c}
-      p\\
-      \bv
-    \end{array}
-  \right) =
-  \left(
-    \begin{array}{c}
-      Rp\\
-      -R\bv
-    \end{array}
-  \right)
-\end{equation}
-Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awepm} with
-$(h_s\delta(z-z_s),f_s\bf{e_z}\delta(z-z_s))$ replaced by
-$(-Rh_r\delta(z-z_r), Rf_r\bf{e_z}\delta(z-z_r))$. That is,
-\begin{equation}
-  \label{eqn:trsadj}
-  {\cal R}{\cal S}^- = -{\cal S}^+_{z_r,z_s}{\cal R}
-\end{equation}
-Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj}
-and \ref{eqn:trsadj} imply equation \ref{eqn:trtr}.
+( h_rP_rp^++f_rP_rv_z^+)
+\]
+\[
+  = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
+\]
+whence \ref{eqn:sadj1} follows immediately.
+
+\section{Declarations}
+\subsection{Funding}
+The author did not receive support from any organization for the
+submitted work.
+\subsection{Competing Interests}
+The author certifies that he has no affiliations with or involvement
+in any organization or entity with any financial interest or
+non-financial interest in the subject matter or materials discussed in
+this manuscript.
+\subsection{Data, Material, and Code Availability}
+The computational examples reported in this work were written in the
+Madacascar reproducible research framework ({\tt
+  http:www.reproducibility.org}). Code and data source is available
+from the author on request.
 
@@
 \bibliography{../../bib/masterref}
+
+\end{document}

### 2022-09-08 repro/papers/RevGEO22-0350/paper.tex
initial

%% This file is automatically generated. Do not edit!
% \documentclass[georeport,12pt]{geophysics}
\documentclass[12pt]{article}

\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amscd}
\usepackage{setspace}
\usepackage{wasysym}
\usepackage{algorithm}
\usepackage{algpseudocode}

\newcommand{\mb}{\mathbf}
\newcommand{\bB}{\mathbf{B}}
\newcommand{\tB}{\tilde{\mathbf{B}}}
\newcommand{\bm}{\mathbf{m}}
\newcommand{\tm}{\tilde{\mathbf{m}}}
\newcommand{\bp}{\mathbf{p}}
\newcommand{\tp}{\tilde{\mathbf{p}}}
\newcommand{\bd}{\mathbf{d}}
\newcommand{\bD}{\mathbf{D}}
\newcommand{\ba}{\mathbf{a}}
\newcommand{\bA}{\mathbf{A}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\bxi}{\mathbf{\xi}}
\newcommand{\bk}{\mathbf{k}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bff}{\mathbf{f}}
\newcommand{\tff}{\tilde{\mathbf{f}}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bz}{\mathbf{z}}
\newcommand{\bh}{\mathbf{h}}
\newcommand{\bR}{\mathbf{R}}

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{alg}{Algorithm}
\newtheorem{remark}{Remark}
\newtheorem{cor}{Corollary}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}

\begin{document}
\title{Review of ``Full waveform inversion by model extension: theory, design and optimization'', GEO22-0350}
\maketitle

\section{Summary}
This paper and its companion (22-0382) make a major contribution to the literature on full waveform inversion (FWI) and avoidance of cycle-skipping. The exposition has a number of easily-corrected deficiencies, and raises several questions implicitly which should be explicitly recognised.

\section{Overview}
FWI is often (though not always) formulated as a least squares problem, the minimization of
\begin{equation}
  \label{eqn:fwi}
  {\bf \Phi}^{\rm FWI}(\bm) = \|\bff(\bm)-\bd\|^2.
\end{equation}
In the definition \ref{eqn:fwi}, $\bff$ is a modeling operator, computed by solving an initial-boundary value problem for a system of wave equations, $\bm$ is a model vector consisting of (functions of) the coefficients of these equations, and $\bd$ is a field of data traces. The definition is often augmented by constraints and regularization terms designed to enforce physically sensible qualities of optimal models.
Starting with the pioneering contributions of Tarantola (1984), Lailly (1983) [these and other citations appear in the manuscript bibliography, unless otherwise noted], and others in the 1980s, minimization of ${\bf \Phi}^{\rm FWI}$ and relatives has been based continuous optimization, that is, on Newton's method and its relatives, which are in turn based on on linearization of the modeling operator: in the notation of this paper,
\begin{equation}
  \label{eqn:born}
  \bff(\bm + \bp) \approx \bff(\bm) + \bB(\bm)\bp
\end{equation}
in which $\bB $ is the linearization or derivative of the forward modeling operator $\bff $ (known in this business as the Born modeling operator, hence the use of ``B''), and $\bp$ is a model perturbation. As nicely illustrated in Figure 4 and the discussion on pp. 15-16, the least-squares solution arising from this approximation is unable to fit data with any choice of model perturbation $\bp$ if the slowness $\bm $ is substantiallly incorrect in its prediction of data kinematics. Therefore any attempt to update $\bm $ through reducing the linearized (Born) fit error may fail through inability to access kinematic information in the data, unless the initial estimate of slowness is already near-correct in the kinematic sense. This failure mode has come to be (somewhat loosely) termed ``cycle-skipping'', because of the typical appearance of predicted waveforms (in $\bff(\bm)$) shifted by more than a half-cycle from observed waveforms.

The work discussed in this paper is in the {\em extended modeling} genre. That is, it provides additional parameters to the model and modeling operator so that local improvement in data fit becomes possible. Using this manuscript's convention for extended quantities, denote the extended model parameter field by $\tm $, the extended modeling operator by $\tff $, and its derivative by $\tB $. The extensions used in the present work amounts to permitting some of the components of $\tm $, which are coefficients in the wave equation, to become {\em operators}, rather than merely multiplication by functions. This extension is physically equivalent to relaxing the no-action-at-a-distance axiom of continuum mechanics.  [For example, the square slowness is proportional to the compliance (reciprocal bulk modulus) for constant density acoustics, the physical setting of this paper. The bulk modulus is the ratio of stress and strain (pressure and divergence of displacement) in acoustics, so permitting it to become a (non-local) operator represents action-at-a-distance.]

The first order approximation to the extended modeling operator is
\begin{equation}
  \label{eqn:extborn}
  \tff(\tm + \tp) \approx \tff(\tm) + \tB(\tm)\tp.
\end{equation}
As has been verified in some of the cited references, and illustrated in Figure 4, for suitable choices of model extension, any reference extended model $\tm $ and data field $\bd$, an extended model perturbation $\tp$ can be found so that the right hand side of equation \ref{eqn:extborn} closely approximates $\bd$. Thus various approximation to Newton's method can at least achieve local improvement in data fit.

Since the (action-at-a-distance) extension violates basic continuum mechanics, it must be suppressed at the solution somehow. One approach to ensuring that the ultimate solution is physical (that is, not extended) uses an operator (``annihilator'') $\bA$ that yields a vanishing result when applied to a physical model field $\bm$: $\bA\bm = 0$, so that any action at non-zero distance is penalized. A simple way to employ such a penalty is in the form of a ``penalty function'':
\begin{equation}
  \label{eqn:penfun}
  {\bf \Phi}^{\rm NL}_{\epsilon}(\tm) = \|\tff(\tm)-\bd\|^2 + \epsilon^2\|\bA \tm\|^2.
\end{equation}
in which the vertical bars denote the root mean square, or $L^2$ norm  (Symes, 2008). If $\bA \tm$ is small at a minimizer of ${\bf \Phi}^{\rm NL}$, then $\tm$ is close to physical, and approximates a (non-extended) minimizer of ${\bf \Phi}^{\rm FWI}$.

A major disadvantage to this approach is computational: the extended model field $\tm$ acts as an operator on the wavefield and its space and time derivatives, so is represented after discretization by a full matrix, which must be applied at every time step in a typical time domain wave solver. A physical model field $\bm$, in contrast, is represented by a diagonal matrix, much cheaper to apply.

This computational disadvantage can be ameliorated to some extent by {\em replacing} the extended modeling operator $\tff$ with its extended Born approximation \ref{eqn:extborn} at a physical model $\tm=\bm$, with extended perturbation $\tp$. Since physical models are (special cases of) extended models, the first order perturbation approximation \ref{eqn:extborn} with $\tm=\bm$ still approximates essentially arbitrary data with suitable choice of extended perturbation $\tp$. At the solution, the model estimate $\bm + \tp$ should be physical. Since the annihilator is linear (in most version of this approach), this condition is equivalent to $\bA \tp = 0$.  A penalty function capturing this idea reads
\begin{equation}
  \label{eqn:wemva}
  {\bf \Phi}^{\rm MVA}_{\epsilon}(\bm,\tp) = \|\bff(\bm) + \tB(\bm)\tp-\bd\|^2 + \epsilon^2\|\bA \tp\|^2.
\end{equation}

Minimization of ${\bf \Phi}^{\rm MVA}_{\epsilon}$ is a version of {\em Wave Equation Migration Velocity Analysis}, or WEMVA. Because the extended modeling operator $\tff$ has been replaced by its linearization about physical models, even a noise-free solution at which the objective value = 0 is not an FWI solution, but rather a Born-FWI solution, that is, $\bd \approx \bff(\bm) + \bB(\bm)\bp$. This approximation is satisfactory exactly when the Born approximation is satisfactory, meaning that the data consists essentially of primary reflections away from critical angle. This status can sometimes be achieved by preprocessing, but field data often contain significant amounts of refracted, diving wave, and non-primary reflection energy, all carrying information about the subsurface but treated as noise by WEMVA. This is one possible reason that WEMVA has not migrated into common industrial practice. Another is that the computation of the extended Born operator $\tB$ involves full matrix multiplies in each time step of a simulation loop, just as does the extended modeling operator $\tff$. This intrinsic expense dominates the cost of WEMVA.

The manuscript under review explains a clever modification of WEMVA that frees it from the Born limitation, and delivers an approximate FWI solution instead. This modification has two components. The first component is the observation that if $\tp \approx 0$ at the solution, then $\bm$  is a minimizer of ${\bf \Phi}^{\rm FWI}$. To force this happy circumstance to occur, the annihilator $\bA$ is augmented by a multiple of the identity operator:
\begin{equation}
  \label{eqn:defd}
  \bD =
  \left[
    \begin{array}{c}
      \alpha {\bf I} \\
      \bA
    \end{array}
  \right]
\end{equation}
with $\alpha > 0$ (see Figure 11). Thus $\bD \tp = 0$ implies not just that $\tp$ is physical, but that $\tp = 0$. The modified penalty term in the FWIME objective is then $\epsilon^2\|\bD \tp\|^2$, rather than $\epsilon^2\|\bA \tp\|^2$:
\begin{equation}
  \label{eqn:fwime}
  {\bf \Phi}^{\rm IME}_{\epsilon}(\bm,\tp) = \|\bff(\bm) + \tB(\bm)\tp-\bd\|^2 + \epsilon^2\|\bD \tp\|^2.
\end{equation}

Figures 14 - 18 show that this innovation may not be quite enough, in that the search direction (gradient) generated by the modified penalty function may have a tendency to overemphasize local oscillatory perturbations, similar to the the gradient of  ${\bf \Phi}^{\rm FWI}$. It's not obvious that iteration would lead to stagnation, but the search direction displayed in Figure 18 (c) is certainly not efficient.

The second innovation is designed to produce a more efficient search direction, by application of a smoothing operator whose effective width decreases as the FWIME iteration proceeds, eventually permitting features of all scales to migrate into $\bm$. Presumably the justification for model smoothing is the familiar observation that the longest spatial frequencies in the model field have the strongest influence (for the most part) on travel times. Therefore to match kinematics, the smoothest parts of the model should be favored, at least initially. The authors accomplish this goal by representing $\bm$ on a (possibly spatially non-uniform) spline space. The effect is to smooth the gradient of the FWIME objective function by the operator ${\bf S}{\bf S}^T$, ${\bf S}$ being the spline representation operator (spline coefficients to fine grid values). Figures 21 (a)-(c) show the effect of spline smoothing on the gradient components displayed in Figures 18 (a)-(c), which is indeed constructive.

To achieve eventual equivalence to FWI, fine scale features must migrate into the model estimate. This migration is achieved by repeated refinement of the spline grid. Refinements occur when the model updates for a given spline grid stagnate. The current model estimate is then projected onto a finer spline grid and the iteration restarted. In the final block of iterations, the spline grid is identical to the fine simulation grid. The three major examples presented at the end of the paper illustrate the effectiveness of this strategy, the eventual vanishing of the extended perturbation $\tp$, and the recovery of an effective FWI solution.

In this overview I have neglected several concepts emphasized in the exposition, such as Born and tomographic gradients. Also it should be mentioned that the variable projection method plays a critical role in the formulation and performance of the algorithm. However I believe that in some sense these are side issues, and the essential ideas are those I have described.

\section{Comments}

The approach described in this paper represents a significant advance in the literature on anti-cycle-skip methods. The paper itself however suffers from two defects: (1) not enough information, and (2) too much non-information. The purpose of a scientific paper is to present new results, and to describe the method used to produce these results in enough detail that a knowledgeable reader could (at least in principle) reproduce the results, {\em and} integrate the methodology into their own work. Note that availability of a computer program or programs with input data does not necessarily accomplish the second goal: if the reader can {\em only} reproduce the results of the paper, then the reader has learned nothing of any use. Too much information is left out of the current manuscript to allow even a knowledgeable reader to transfer the methodology to his/her own projects. On the other hand, repeated and repetitive claims for the virtues of the approach are distracting, and in some cases misleading. In the following couple of pages, I will try to give a representative list of missing information and unnecessary misinformation, hopefully in enough detail that the authors can easily fix both types of defect.

\subsection{Model Specification}

The physics used throughout is introduced on p. 7, in English: $\bff$ is the ``discretized acoustic isotropic constant-density forward modeling operator''.  Next up (skipping the WEMVA subsection on p. 8, about which more later) is $\tB$, the ``extended Born modeling operator''. Then $\bD$, ``an invertible modified form of the DSO operator''.

It's likely true that most interested readers will understand what is meant by ``discretized acoustic isotropic constant-density forward modeling operator''. However, one cannot tell from the manuscript what discretization is used here, nor what source representation (isotropic point source, probably, but with or without correction for off-grid placement? And with what wavelets, in the various examples?).

Fewer will be familiar with the ``extended Born modeling operator''. For one thing, what extension? And for that matter, what is an extension? Only those marinated in the literature on anti-cycle-skip are likely to know.

Finally, only specialists are likely to recognize ``an invertible modified form of the DSO operator'', especially since the references cited on p 10 discuss extensions and DSO operators quite different from those used here. Furthermore the invertible modification is not described at all - what invertible modification?

The paper really needs to be modified to include explicitly all of this missing information. But this is easy: an appendix is a natural solution. Start with the acoustic constant density wave equation, introduce sampling with isotropic point sources and receivers. Take the slowness derivative to obtain the Born operator. Define the model extensions you actually use (subsurface horizontal offset and time shift). Describe briefly the finite difference approach you use to turn these into discrete problems, for both extended and non-extended contexts. Then define the operator $\bD$ (I offer display \ref{eqn:defd} at no extra charge).

A reader who already knows all of this stuff, or doesn't care, can ignore the appendix, but one who really wants to be sure of following your reasoning will read it.

\subsection{Hyperparameters, hyperparameters everywhere...}

The authors assert that their method is ``simple to use'' (p. 1), its features are mathematically consistent ``thereby reducing the number of optimization hyper-parameters to two'' (p. 5), and result in a method that ``can invert any type of seismic data using the same framework and without the need for intensive hyper-parameter tuning'' (p. 6). Also, ``...FWIME is formulated in a compact and mathematically consistent manner that only requires a simple tuning of one hyperparameter at the initial step'' (p. 21). Many other statements throughout the paper echo this notion that the method is largely automated so that non-expert users can easily adopt it.

While this goal (automation via good default or computed choices of essential parameters) is very laudable, it's not yet achieved in this work. There are far more than one (or two) ``hyper-parameters'' that must be chosen as input to Algorithm 2. I address a number of them in this subsection, beginning with the two that are actually called out.

\subsubsection{Penalty Parameter $\epsilon$}
This ``trade-off'' parameter is one of the two called out explicitly in the manuscript. It is to be selected by ``a trial and error approach based on examining a subset of the CIGs extracted from the optimal extended perturbation $\tp_{\epsilon}^{\rm opt}$ computed at the initial step''. Very well, but how? What aspect of the CIGs should be used to set this parameter? How many CIGs need to be examined? Note that any correction requires re-computation of  
$\tp_{\epsilon}^{\rm opt}$, which (as later discussed) is the principal computational bottleneck of the method. So this step is manual {\em and} computationally expensive, but the authors do not describe how to carry it out!

I suggest including an example that involves an initial incorrect (too large or too small) choice of $\epsilon$, and its correction based on whatever the authors actually do by trial and error. The examples included in the current manuscript do not show this process - and it is absolutely critical! Too small, and not enough resolution of slowness; too large, and the method behaves like FWI. How do you tell where you are in this spectrum by looking at CIGs? The behaviour of the iteration may be relatively insensitive to $\epsilon$, as noted on p. 21, however apparently it cannot be off by much more than an order of magnitude. Note that $\epsilon$ is a dimensional parameter, which makes its tuning even more difficult.

\subsubsection{Extent of Extension}
p. 29 bottom: ``Figure 17
illustrates how $\tp^{\rm opt}$ contains crucial kinematic information and it shows the importance of using an extended perturbation with a large-enough extension (otherwise the information would be lost).''

The obvious question is, how does the user determine what is ``large enough''?  It appears that the answer is ``trial and error'', by spot-checking gathers. However even more important: what quality should the extended gather possess? In principle, this question has an easy answer: the extension axis (axes) should be large enough that the data is well-fit at the beginning. However, that is not the whole story. Should the test be that the data is well-fit with $\epsilon=0$? Since $\epsilon > 0$ is in tension with data fit, perhaps that's the right answer, but note that this once again involves computing $\tp^{\rm opt}$, which is the single most expensive part of the algorithm.

Again, I think the simplest way to address this deficiency is to include a non-trivial example, in which an inadequate extension axis is adjusted. Since looking at individual data gathers is not going to reduce the expense, presumably the relative residual of the entire data set
\[
  \|\bff(\bm) + \tB(\bm)\tp-\bd\|/\|\bd\|
\]
should be below some suitable tolerance. What tolerance? Or perhaps the authors use some other criterion. Please explain!

\subsubsection{Type of Extension}
The examples use the time-shift extension as in Biondi and Almomin (2014), which is fine for 2D examples. This choice gets little discussion. On p. 26, we find  ``...for 3D field applications, the time-lag extension requires a single additional axis (compared to two axes for space lags), thereby reducing the computational cost and memory footprint of the method.''. However, the purpose of the extension is to permit the extended model to fit the data, thus transporting kinematic information from data space to model space. In 3D (full azimuth), the number of data axes is 5, whereas with the time-shift extension the number of extended model axes is 4. This parameter count suggests that at least in some cases a time-shift extended model will not be adequate for 3D application.

\subsubsection{Penalty parameter $\alpha$}
The operator $\bD$ is implicitly defined in Figure 11, and depends on a parameter $\alpha$. I have taken the liberty of creating an actual definition \ref{eqn:defd}, which may describe the operator $\bD$ used in this work. It depends on a parameter $\alpha$. If $\alpha=0$, then $\bD$ is effectively a WEMVA penalty operator. As stated on p. 20, ``...since the goal is not to obtain a well-cofused image (but to make $\tp_{\epsilon}^{\rm opt}$ vanish), we modify the DSO operator by also penalizing energy located on the physical plane of $\tp_{\epsilon}^{\rm opt}$.''

So: how do we choose $\alpha$? So far as I can tell, the choice of $\alpha$ used in the examples is nowhere given, nor is any principle suggested to make this choice. The authors need to include information about this. It's fine if they just say $\alpha=1$ or some such arbitrary choice, even better if they give a reason for a specific choice, but they need to tell us what they chose, and why.

This choice raises another question: if the goal is to suppress all the energy in the extended part of the model $\tp$, why not just dispense with the WEMVA part of $\bD$ altogether, and choose $\bD={\bf I}$? The authors might try to answer this question as part of describing the effect of various choices of $\alpha$.

\subsubsection{Number of CG iterations, accuracy of $\tp_{\epsilon}^{\rm opt}$}
The number of CG iterations used to compute  $\tp_{\epsilon}^{\rm opt}$ is mentioned in each example, but not the reason for the choice (which varies from example to example). In fact, this is another number that the user much choose. How did the authors make this choice, and what advice do they have for a potential user?

This question is more subtle than it might at first appear.

A natural criterion for truncation of the CG loop is that the gradient of $\Phi_{\epsilon,\bm}(\tp)$ (equation 6, p. 10) decrease in length below a prescribed maximum (absolute tolerance) or below a prescribed factor of its length at iteration 0 (relative tolerance). The first iteration of CG is a steepest descent step along the gradient, which could also be viewed as an extended RTM image. Would that search direction be sufficient? If so, a great deal of the expense in carrying out FWIME could be eliminated. Actually, using RTM rather than a more accurate minimization of $\Phi_{\epsilon,\bm}(\tp)$ almost surely won't deliver an accurate gradient of $\Phi_{\epsilon}(\bm)$, which is the main goal of computing $\tp_{\epsilon}^{\rm opt}$.

This topic is ripe for further research, but there have been a couple of studies. Symes and Kern (1994) pointed out that RTM is insufficient to yield a useful gradient, but so is CG with any of the common truncation criteria. The reason is that the expression 16 (p. 22) of $\nabla \Phi_{\epsilon}(\bm)$ is incorrect. It is missing the term
\[
  (D \tp_{\epsilon}^{\rm opt}(\bm))^* [\tB(m)^*(\bff(\bm)+\tB(\bm) \tp_{\epsilon}^{\rm opt}(\bm)-d) + \epsilon^2\bD^*\bD \tp_{\epsilon}^{\rm opt}(\bm)],
\]
which comes from applying the chain rule to the right-hand side of equation 6. Of course, the quantity in square brackets is precisely the gradient of $\Phi_{\epsilon,\bm}(\tp)$ (with respect to $\tp$. An iterative method such as CG eventually makes this gradient small (in the $L^2$ sense) but does not make it vanish, in any finite number of iterations. That would be fine, {\em except} that the operator $(D \tp_{\epsilon}^{\rm opt}(\bm))^*$ involves the derivative of $\tp_{\epsilon}^{\rm opt}(\bm)$ with respect to $\bm$. This $\bm$ derivative has a similar effect to spatial derivatives, that is, enhances high frequency content without limitation, and can therefore give a large result even when applied to an $L^2$-small $\nabla \Phi_{\epsilon,\bm}(\tp)$. Thus the missing term in equation 16 can be quite large.

Symes and Kern (1994) give examples of this phenomenon showing that it can cause the computed gradient via the VPM formulae (of which equation 16 is an example), with the equivalent of $\tp_{\epsilon}^{\rm opt}(\bm)$ computed via CG, to be uselessly inaccurate. They also describe a correction that compensates for the missing term using Chebyshev (rather than CG) iteration, and show examples of accurate gradient calculations using this alternative technique. Twenty years later, the group of Herv\'{e} Chauris revisited this topic, proposing several remedies. The {\em function}  $\Phi_{\epsilon}$ with CG-derived $\tp_{\epsilon}^{\rm opt}(\bm)$ is differentiable, and the errors in its values are controlled by the $L^2$ error achieved by CG. Chuaris et al. applied automatic differentiation to calculate the adjoint derivative $(D \tp_{\epsilon}^{\rm opt}(\bm))^*$, and thus computed the gradient of $\Phi_{\epsilon}$ with CG-derived $\tp_{\epsilon}^{\rm opt}(\bm)$ to machine precision \cite[]{CocherChaurisPlessix:COMG17}. Alternatively, they showed that a stable gradient could be obtained by application of a suitable filter to the naive VPM gradient, without however giving a convergence proof \cite[]{CocherChaurisPlessix:GEO18}.

Apart from these references, the issue is ignored altogether in the literature on WEMVA and related topics. I speculate that the neglect of gradient instability may be in some part responsible for the failure of extended model techniques to have a serious impact on geophysical practice.

\subsubsection{Spline grids, refinement schedule}
The authors are good about telling us what spline grids are used in the examples (though for the last example we learn only the first and the last (the finite difference grid) and that there is another grid in between). As for how a user might construct the spline grid schedule - how many grids, what degree of refinement from one to the next - I cannot find any advice. It would be appreciated. Too few grid levels could presumably result in failure to converge to an optimal model. Too many is just expensive. It may well be that the authors don't have a comprehensive theory for managing the grid schedule, but they clearly have some idea about how to do it. There are least two parameters here that need choosing - the coarsest grid level, and the refinement factor. That's for the simplest case, of similar grids.

The introduction of local refinements would introduce an entirely new set of choices. None of the FWIME examples use non-uniform grids, nor discuss how to include non-uniform refinement in FWIME. I think that even the minor amount of discussion of non-uniform grids, centering around Figures 19 and 20, is out of place here and should be reserved for another paper, where nonuniform grids are actually used.

A side-issue: the dismissal of a spatial filter as in Biondi and Almomin (2014) on p. 30 is misleading. Algorithm 2 precisely applies a spatial filter. The use of a spline grid to construct it is a convenience, not an essential feature. Some mechanism to control the degree of smoothing, or width of the averaging kernel, is essential in this application, however there are many ways (other than the spline construction) to do that. The essential point is the coupling of the degree of smoothing to the optimization, reducing the spatial scale of $\bm$ towards zero as the iteration proceeds.

\subsubsection{Optimization Parameters}
Finally, every optimization algorithm depends on parameter choices. For CG, it suffices to specify the number of iterations, as already discussed, though that may not be optimal, as also discussed. For LBFGS, the rank of the approximate inverse Hessian and several line search parameters need specification. This is true for any application of these methods, and is not specific to FWIME.

\subsubsection{Summary}
I count at least 10 important user-selected parameters. Maybe 5 (mostly the optimization parameters, and maybe $\alpha$) could be given usable defaults, leaving another 5 (at least, more in 3D) as user choices. It's not just one, or even two.

\subsection{Forms of WEMVA}
The objective $\Phi_{\epsilon}^{\rm MVA}$ defined in equation \ref{eqn:wemva}, which I claimed to be a form of WEMVA, differs from the function $\Phi_{\rm WEMVA}$ discussed on p. 8. However, the latter is a limiting case of the former: just as the FWIME function tends to the FWI objective as $\epsilon \rightarrow \infty$, so does $\epsilon^{-2}\Phi_{\epsilon}^{\rm MVA}$ tend to $\Phi_{\rm WEMVA}$ as $\epsilon \rightarrow 0$, provided that we make two identifications: (1) ${\bf E} = \bA$, and (2) we solve for $\tp$ using one CG iteration, so that $\tp$ is the RTM image ${\bf I}$ as defined in equation (3). The earliest accounts of VPM in extended inversion, such as Symes and Kern (1994), used the penalty function form given in equation \ref{eqn:wemva} of this review, along with much more accurate solution of the inner problem.

In any case, the relation between FWIME and WEMVA in the form of definition \ref{eqn:wemva} in this review seems more evident than in the form of equation 2 in the manuscript, and the latter is actually a limiting special case of the former. The authors might consider revising this discussion accordinly.

\subsection{Transmitted and refracted wave inversion}
The first and third examples are excellent, and clearly differentiate FWIME from the standard viewpoint of MVA as a process applied to primary reflections. However the observation that extended inversion of the type considered here can accomodated transmitted and diving waves is not new. \cite{Shen2013} and \cite{Lameloise2016} independently observed that subsurface offset extension provides a framework for  model-domain analysis of velocity error from transmitted wavefields, including diving waves.

This earlier work did not produce algorithms that accommodate all wave types within one workflow. FWIME, through its two innovations (the disappearing extended perturbation, and integrated management of spatial scale) does exactly that.

\subsection{FWIME vs. RFWI}
How does FWIME actually compare to RFWI, in ability to converge to a good nonlinear least squares solution? RFWI is similar to WEMVA in that it splits the model into kinematic and dynamic components, though the decomposition changes during the iteration. It is reduced to a problem over the kinematic component by VPM, just as some WEMVAs and FWIME do, and ultimately returns an approximate FWI solution, as does FWIME. The remark at the end of p. 25 suggests that FWIME is ``more robust''. Can you cite a specific comparison? If not it would be better to remove this assertion.

\subsection{Nonlinear inversion}
Example 2 illustrates FWIME inversion of reflection data. Similar examples appear in accounts of various versions of WEMVA. However even construed as an inversion method, as I did earlier, WEMVA produces a solution of the Born inversion problem, with separate velocity and reflectivity models, usually differing in scale spectrum. Thus WEMVA does not actually solve the FWI problem, whereas FWIME does.

The most obvious difference between acoustic data and its Born approximation is the (possible) presence in the former of multiple reflections. These are commonplace in field data, and are routinely suppressed in standard data processing. Such suppression of multiples is often imperfect, and renders the data unfaithful to the acoustic model. Since FWIME ultimately solves the FWI problem, fitting data via the acoustic model, it is irresistable to ask: does FWIME succeed in the presence of multiple reflections?

The data of example 2 may contain multiple reflections, though of small amplitude, since the reflection coefficients are all relatively small, and absorbing boundaries appear to have been used througout. A free surface would have produced considerably more multiple energy. Would FWIME have succeeded then?

While this question - can FWIME deal directly with multiples? - is fascinating, it opens up several aspects of inversion that this paper has not considered. In particular, source calibration is essential in interpreting multiply reflected energy. Since most sources in field use are anisotropic, this amounts to more than estimating a wavelet: radiation pattern must also be pinned down.

Still, even a ``cheating'' synthetic example, with a known isotropic point source and a nice free-surface multiple series, would represent a decisive advance beyond WEMVA. Can FWIME do it?

\subsection{A final thought}
The comments I have offered so far mostly concern clear expositional shortcomings of the current draft, and I would expect to see them addressed in a revision. My final comment is more stylistic in nature, and I offer it more as a statement of personal preference than as a requirement.

I find the material in the part of the paper between pages 12 and 30 to be redundant and mostly motivational. I suspect I am typical of the potential readers of this paper in that I don't need quite so much motivation, or ``a high-level description of the main intuition'' (p. 12): the ideas and eamples in this paper are sufficient motivation already.

This is a long and complicated piece of writing, so I will just pick out two examples. The section ``FWIME inversion workflow'' presents an fully described algorithm (Algorithm 1), then proceeds to show that it doesn't always work by a special example introduced only for this purpose. This takes five pages. The point of this discussion is the necessity of a spatial multi-scale approach, at least in some cases (but not example 1!). This point is made in the section ``A model-space multi-scale approach for FWI'', and the illustration could be folded in with example 2, which also does not work without multiple scales (or so the discussion suggests). My suggestion would be: get rid of Algorithm 1, which is not the algorithm ultimately used and is anyway a special case of Algorithm 2 with one spatial scale, and fold the illustration in with the example section.

Another example is equation 21 on p. 23 and surrounding discussion, which echoes the earlier subsection title ``FWIME $\approx$ FWI + WEMVA''.  The text following the equation is an interesting description of (what the authors claim is) typical behaviour of the algorithm, in terms of the two components of the gradient. Part of this behaviour is illustrated in example 2, which shows the components of the initial FWIME search direction. Clearly the rest of the behaviour could be illustrated in by this example as well, instead of introducing a different but very similar example (Figure 12 etc.) to make the illustration. You could even get rabbit ears from Mora!

There are many other ways in which I feel that this mid-section of the paper is too repetitive and motivational, in a way that might be appropriate in a PhD thesis or a textbook, but which just gets in the way of the main message in a research paper. As stated earlier, I don't ask for specific revisions to address this issue, and am not even sure that every reader would react the same way. I do wish the authors to think about what really belongs in this exposition, and realize that unnecessary length will reduce the impact of their contribution.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

\end{document}

### 2022-10-22 repro/papers/GEM22/submitR1/oldexpls.tex
initial revision commit

--- 2022-03-29 repro/papers/GEM22/paper.tex
@@
-\title{Efficient Computation of Extended Sources}
-\author{William. W. Symes \thanks{The Rice Inversion Project,
-Department of Computational and Applied Mathematics, Rice University,
-Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}
-
-\lefthead{Symes}
-
-\righthead{Approximate Source Inversion}
-
-\maketitle
-\begin{abstract}
-Source extension is a reformulation of inverse problems in wave propagation, that at least in some cases leads to computationally tractable iterative solution methods.
-The core subproblem in all source extension methods is the solution of a linear inverse problem for a source (right hand side in a system of wave equations) through minimization of data error in the least squares sense with soft imposition of physical constraints on the source via an additive quadratic penalty.
-A variant of the time reversal method from photoacoustic tomography provides an approximate solution that can be used to precondition Krylov space iteration for rapid convergence to the solution of this subproblem.
-An acoustic 2D example for sources supported on a surface, with a soft contraint enforcing point support, illustrates the effectiveness of this preconditioner.
-\end{abstract}
-
-\inputdir{../sseprecond/project}
-
-\section{Introduction}
-Full Waveform Inversion (FWI) can be described in terms of 
-\begin{enumerate}
-\item a linear wave operator $L[{\bf c}]$, depending on a vector of
-  space-dependent coefficients ${\bf c}$ and acting on causal vector wavefields $\bu$ vanishing in negative time:
-\begin{equation}
-\label{eqn:init}
-\bu \equiv 0, t \ll 0; 
-\end{equation}
-\item a trace sampling operator $P$ acting on wavefields and producing data traces;
-\item and a (vector) source function (of space and time) $\bff$ representing energy input to the system. 
-\end{enumerate}
-The basic FWI problem is: given data $d$, find ${\bf c}$ so that 
-\begin{equation}
-\label{eqn:fwi}
-P\bu \approx d \mbox{ and } L[\bf{c}]\bu = \bff.
-\end{equation}
-In this formulation, the source function $\bff$ may be given, or
-to be determined subject to some constraints.
-
-%The energy source $\bff$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time.
-In fact, additional source degrees of freedom, beyond those needed to describe physically realized sources, may be useful in rendering the FWI problem \ref{eqn:fwi} more amenable to numerical solution, via so-called extended modeling (see \cite{geoprosp:2008}, \cite{LeeuwenHerrmannWRI:13}, \cite{HuangNammourSymesDollizal:SEG19}, and many references cited there).
-Therefore it is natural to view $\bff$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
-A simple nonlinear least squares formulation is:
-\begin{equation}
-\label{eqn:ols}
-\mbox{choose } {\bf c} \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2.
-\end{equation}
-Practical optimization formulations typically augment the objective in
-\ref{eqn:ols} by additive penalties or other constraints.
-
-As is well-known, local optimization methods are the only feasible
-approach given the dimensions of a typical instance of \ref{eqn:fwi},
-and those have a tendency to stall due to ``cycle-skipping''. See for
-example \cite{VirieuxOperto:09} and many references cited there. Source
-extension is one approach to avoiding this problem. It consists in
-imposing the wave equation as a soft as opposed to hard constraint, by
-allowing the source field $\bff$ to have more degrees of freedom than
-is permitted by a faithful model model of the seismic experiment, and
-constraining these additional degrees of freedom by means of an
-additive quadratic penalty modifying the problem \ref{eqn:ols}:
-\begin{equation}
-\label{eqn:esi}
-\mbox{choose } {\bf c}, \bff \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
-\end{equation}
-The operator $A$ penalizes deviation from known (or assumed)
-characteristics of the source function - its null space consists of
-feasible (or ``physical'') source models.
-
-\cite{HuangNammourSymesDollizal:SEG19} present an overview of the
-literature on source extension methods, describing a variety of
-methods to add degrees of freedom to physical source model. The present paper
-concerns {\em surface source extension}: physical sources are
-presumed to be concentrated at points $\bx_s$ in space, whereas their extended
-counterparts are permitted to spread energy over surfaces containing
-the physical source locations. A simple choice for the penalty
-operator $A$ is then multiplication by the distance $|\bx-\bx_s|$ to the physical
-source location:
-\begin{equation}
-  \label{eqn:penop}
-  (A\bff)(\bx,t) = |\bx-\bx_s|\bff(\bx,t)
-\end{equation}
-I shall use this choice of penalty operator whenever a specific choice
-is necessary in the development of the theory below.
-
-This paper presents a numerically efficient approach to solving the
-{\em source subproblem} of problem \ref{eqn:esi}:
-\begin{equation}
-\label{eqn:esis}
-\mbox{given } {\bf c}, \mbox{ choose } \bff \mbox{ to minimize }
-\|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha \|A\bff\|^2 
-\end{equation}
-Solution of this subproblem is an essential component of {\em variable
-  projection} algorithms for solution of the nonlinear inverse problem
-\ref{eqn:esi}. Variable projection is not merely a convenient choice
-of algorithm for this purpose: it is in some sense essential, see for
-example \cite{Symes:SEG20}. It replaces the nonlinear
-least squares problem \ref{eqn:esi} with a {\em reduced} problem, to
-be solved iteratively. Each iteration involves solution of the
-subproblem \ref{eqn:esis}. Therefore efficient solution of the
-subproblem is essential to efficient solution of the nonlinear problem
-via variable projection.
-
-The penalty operator $A$ defined in \ref{eqn:penop} is linear, so the source
-subproblem is a linear least squares problem. Under some additional
-assumptions to be described below, I shall show how to construct an
-accurate approximate solution operator for problem
-\ref{eqn:esis}. This approximate solution operator may be used to
-accelerate Krylov space methods for the solution of the surface source
-subproblem \ref{eqn:esis}. Numerical examples suggest the
-effectiveness of this acceleration.
-
-I will fully describe a preconditioner for a special
-case of the source subproblem \ref{eqn:esis}, in which ${\bf u}$ is an
-acoustic field, $L[{\bf c}]$ is the wave operator of linear
-acoustodynamics, the spatial positions of traces extracted by $P$ lie
-on a depth plane $z=z_r$, and the positions at which the extended
-source $\bff$ is nonzero lie on another, parallel, depth plane $z=z_s$. 
-%That is, sources take the form of a combination of constitutive law
-%defect and normal force,
-%\begin{equation}
-%  \label{eqn:surfsrc}
-%  \bff(x,y,z,t) = h_s(x,y,t)\delta(z-z_s)(1,0,0,0)^T + f_s(x,y,z,t)(0,0,0,1)^T
-%\end{equation}
-%and the sampling operator $P$ is defined by
-%\begin{equation}
-%  \label{eqn:surfsam}
-%  P\bu(x,y,t) =
-%\end{equation}
-%supported on $z=z_s$, and similar
-%surface vertical loads.
-This
-``crosswell'' configuration simplifies the analysis underlying the
-construction of approximate solutions for the source subproblem
-\ref{eqn:esis}. It is only one of many transmission configurations for
-which similar developments are possible. Perhaps the most important
-alternative example is the diving wave configuration, which plays a
-central role in contemporary FWI.
-
-The preconditioner construction is very similar to the time-reversal
-method in photoacoustic tomography
-\cite[]{StefanovUhlmannIP:09}. Preconditioning amounts to a change of
-norm in the domain and range spaces of the modeling operator. In this
-case, the modfied norms are weighted $L^2$, and the weight operators
-map pressure to corresponding surface source on the source and
-receiver planes. This pressure-to-source map is closely related to the ``hyperbolic
-Dirichlet-to-Neumann'' operator that plays a prominent role in
-photoacoustic tomography and other wave inverse problems
-\cite[]{Rachele:00,StefUhl:05}. \cite{HouSymes:EAGE16} demonstrated a
-very similar preconditioner for Least Squares Migration, also for its
-subsurface offset extension \cite[]{HouSymes:16}, motivated by
-\cite{tenKroode:12}. These constructions also involve the
-Dirichlet-to-Neumann operator. This concept also turns up in hidden
-form in the work of Yu Zhang and collaborators on true amplitude
-migration
-\cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09}.
-
-The obvious computation of the pressure-to-source map - prescribe the
-pressure, solve the wave equation with this boundary condition, read
-off the equivalent source - suffers from intrinsic numerical
-inaccuracy. I suggest an alternative computationally feasible
-approach, via economical short-distance wave propagation. Since the
-map is symmetric only in an approximate, asymptotic sense, it must be
-symmetrized for use as a Krylov preconditioner. I describe a
-symmetrization procedure that requires no further wave computations
-beyond those necessary to compute the action of the operator itself.
-
-The discussion in this paper is formal and incomplete, in the sense
-that some important mathematical underpinnings are only
-sketched. I will treat the modeling operator $PL[{\bf c}]^{-1}$ as if
-it mapped square integrable surface sources to square integrable sampled
-data. This is not true in full generality: while the surface source
-problem has distribution solutions, they are not generally square
-integrable (finite acoustic field energy). Even if the solutions have
-finite energy, they do not in general have well-defined
-restrictions to lower-dimensional sets. In other words, the action of
-the sampling
-operator $P$ on the space-time plane $z=z_r$ is not well-defined for arbitrary
-finite-energy acoustic fields. Thus the modeling operator envisioned above
-may not be well-defined.
-
-This
-phenomenon is related to the ill-posedness of wave equations as
-evolution equations in spatial variables, an observation attributed to
-Hadamard (see \cite{CourHil:62}, Chapter 6, section 17). Some constraint on the acoustic field,
-beyond finite energy, is mandatory in any precise mathematical formulation the inverse problems
-\ref{eqn:esi} and \ref{eqn:esis}. In fact, the natural constraint in
-the ``crosswell'' geometry of this paper is that high-frequency energy
-travel {\em only} along rays crossing the
-surfaces $z=z_s, z=z_r$ transversally. That is, source functions on
-$z=z_s$ generate waves with energy traveling along rays leaving the surface at a non-zero angle, and
-energy arrives at the recording surface $z=z_r$ along rays making
-non-zero angle with it. I will call sources, sampled data, and
-acoustic fields with this property {\em
-  downgoing} (even though the concept also encompasses {\em upcoming}
-propagation). Note that the downgoing property restricts the behaviour of
-acoustic fields near the source and receiver surfaces - what the
-fields do elsewhere is their own business.
-
-Several works have explored the mathematics of the
-downgoing condition and its consequences in the context of the scalar second order wave equation, see for instance
-\cite{Payn:75,Symes:83,Lasi:86,LasLionsTrig:86,Lasi:87, BaoSy:91b}. 
-Since the purpose of this paper is to explore the algorithmic
-consequences of the mathematical structure implied by the downgoing
-condition, I leave the elaboration of the mathematical details to
-another place.
-
-The next section defines the modeling operator $PL[{\bf c}]$, its
-adjoint, and important specializations (pressure vs. normal velocity
-sources and data). The sections to follow define the
-source-to-pressure operator, construct an approximate inverse of the
-modeling operator by time reversal (as suggested by work in
-photoacoustic tomography), use the source-to-pressure operator to
-express the approximate inverse as the modeling operator adjoint in
-weighted norms (thus establishing that the modeling operator is {\em
-  unitary} in the sense of these norms), explain how to use this
-construction to precondition Conjugate Gradient iteration, and
-organize the preconditioning computation so as to involve only one
-wave propagation calculation. The penultimate section displays simple
-2D numerical examples of all of the key steps, culminating in a
-comparison of straight vs. preconditioned CG iteration. The paper ends
-with a brief discussion-and-conclusion section, reviewing what has
-been accomplished and listing a few of the many questions left open.
-
-\section{Operators}
-
-For acoustic wave physics, the coefficient vector is
-$\bf{c}=(\kappa,\rho)^T$, with components bulk modulus $\kappa$ and
-density $\rho$, and the state vector $\bu=(p,\bv)^T$ consists of
-pressure $p$ (a scalar space-time field) and particle velocity $\bv$
-(a vector space-time field). The wave operator $L[\bf{c}]$ is:
-\begin{equation}
-\label{eqn:aweop}
-L[\bf{c}]\bf{u} = 
-\left(
-\begin{array}{c}
-\frac{1}{\kappa}\frac{\partial p}{\partial t}  + \nabla \cdot \bv, \\
-\rho\frac{\partial \bv}{\partial t} + \nabla p.
-\end{array}
-\right) 
-\end{equation}
-That is,
-\begin{equation}
-  \label{eqn:awemat}
-  L[{\bf c}] = \left(
-    \begin{array}{cc}
-      \frac{1}{\kappa}\frac{\partial}{\partial t} & \nabla \cdot \\
-      \nabla & \rho \frac{\partial}{\partial t}
-    \end{array}
-  \right)
-\end{equation}
-$L[{\bf c}]$ has a well-defined inverse if it is restricted to either
-causal or anti-causal vector wavefields.
-
-Most of what follows is valid for any space dimension $n >0$. The
-coefficient vector $\bf{c}=(\kappa,\rho)$ is defined throughout space
-$\bR^n$, the state vector $\bu$ throughout space-time
-$\bR^{n+1}$. Whenever convenient for mathematical manipulations,
-$n=3$: for instance, I will write $\bx=(x,y,z)^T$ for the spatial
-coordinate vector, and refer to the third (vertical) coordinate of
-particle velocity as $v_z$. Examples
-later in this paper will use $n=2$ for computational convenience.
-
-Since all of the operators in the discussion that follows depend on
-the coefficient vector 
-$\bf{c}$, I will suppress it from the notation, for example, $L=L[\bf{c}]$.
-
-The surface source extension replaces point sources on or near a
-surface in $\bR^3$ with source functions confined to the same
-surface. The simplest example of this extended geometry specifies a
-plane $\{(x,y,z,t): z=z_s\}$ at source depth $z_s$ as the surface. For
-acoustic modeling, surface sources are combinations of constitutive law
-defects and loads normal to the surface, localized on $z=z_s$. That
-is, right-hand sides in the system $L\bu=\bff$ take the form
-$\bff(\bx,t) = (h_s(x,y,t)\delta(z-z_s),
-f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar defect $h_s$ and normal
-force $f_s$ ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in
-\ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$
-takes the form
-\begin{eqnarray}
-\label{eqn:awepm}
-\frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} & = & - \nabla \cdot \bv^{\pm} +
-h_s \delta(z-z_s), \nonumber \\
-\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} +
-                                                f_s{\bf e} \delta(z-z_s),\nonumber \\
-p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
-\bv^{\pm} & = & 0 \mbox{ for } \pm t \ll 0.
-\end{eqnarray}
-
-\noindent {\bf Remark:} In system \ref{eqn:awepm} and many similar
-systems to follow, I will use the shorthand
-\[
-  p^+ = 0 \mbox{ for } t \ll 0 
-\]
-to mean that $p^+$ is {\em causal}, that is,
-\[
-  \mbox{For some } T \in \bR, p^+(\cdot,t) = 0 \mbox{ for all } t <
-  T.
-\]
-Similarly,
-\[
-  p^- = 0 \mbox{ for } t \gg 0 
-\]
-signifies that $p^-$ is anti-causal.
-
-Extended forward modeling consists in solving \ref{eqn:awepm} and
-sampling the solution components at receiver locations. For
-simplicity, throughout this paper I will assume that the receivers are
-located on another spatial hyperplane $\{(x,y,z,t): z=z_r\}$ at
-receiver depth $z_r>z_s$. The constructions to follow involve interchange
-of the roles of $z_s$ and $z_r$ (that is, locating sources on $z=z_r$
-and receivers at $z=z_s$), so rather than the sampling operator $P$ of
-the introduction, I will denote by $P_s,P_r$ the sampling 
-operators on $z=z_s$, $z=z_r$ respectively. In practice, sampling
-occurs at a discrete array of points (trace locations) on these
-surfaces, and over a zone of finite extent. In this theoretical
-discussion, I will neglect both finite sample rate and extent, and
-regard the data, for example $P_rp^+$, as continuously sampled and
-extending over the entire plane $z=z_r$.
-
-For downgoing solutions of system \ref{eqn:awepm}, the key
-components ($p^{\pm}$ and $v^{\pm}_z$) are continuous functions of $z$
-in the open slab $z_s<z<z_r$ with well-defined limits at the boundary
-planes, but may be discontinuous at the source plane
-$z=z_s$. Similarly, the roles of $z_s$ and $z_r$ will be interchanged
-in some of the constructions to come, and the corresponding solutions
-may be discontinuous at $z=z_r$. Accordingly, interpret $P_s$, $P_r$
-as the limit from right and left respectively: for $u=p^{\pm}$ or
-$v^{\pm}_z$,
-\begin{eqnarray}
-  \label{eqn:defsamp}
-  P_su(x,y,t) &=& \lim_{z \rightarrow z_s^+} u(x,y,z,t),\nonumber \\
-  P_ru(x,y,t) &=& \lim_{z \rightarrow z_r^-} u(x,y,z,t).                  
-\end{eqnarray}
-
-The causal/anti-causal vector
-modeling operators ${\cal S}^{\pm}_{z_s,z_r}$ are defined in terms of
-the solutions $(p^{\pm},\bv^{\pm})$ of the systems \ref{eqn:awepm} by
-\begin{equation}
-  {\cal S}^{\pm}_{z_s,z_r}(h_s,f_s)^T  = (P_rp^{\pm},P_r v_z^{\pm})^T,
-  \label{eqn:fwd}
-\end{equation}
-The subscript signifies that sources are located on $z=z_s$, the
-receivers on $z=z_r$. It is necessary to include this information in
-the notation, as versions of ${\cal S}^{\pm}$ with sources and receivers in
-several locations will be needed in the discussion below.
-
-\noindent {\bf Remark:} To connect with the formulation presented in
-the introduction, note that for continuous $u$,
-$P_su(x,y,t)=u(x,y,z_s,t)$, and therefore the adjoint of $P_s$ (in the
-sense of distributions) is $P_s^Th(x,y,z,t) =
-h(x,y,t)\delta(z-z_s)$. Write ${\cal P}_s = \mbox{diag }(P_s,P_s)$ and
-similarly for ${\cal P}_r$. Then
-\[
-  {\cal S}^{+}_{z_s,z_r} = {\cal P}_r L^{-1}({\cal P}_s)^T,
-\]
-in which $L^{-1}$ is interpreted in the causal sense, and similarly
-for ${\cal S}^{-}$. Sources confined to $z=z_s$ are precisely those
-functions (distributions, really) output by ${\cal P}_s^T$, so the
-problem statements \ref{eqn:esi} and \ref{eqn:esis} can be rewritten
-in terms of ${\cal S}^+_{z_s,z_r}$, with $P$ identified with ${\cal P}_r$.
-
-${\cal S}^{\pm}$ is not stably invertible: its columns are
-approximately linearly dependent, as will be verified below. The
-diagonal components of ${\cal S}^{\pm}$ thus carry essentially all of
-its information, and it is in terms of these that a sensible inverse problem
-is defined.
-
-Denote by $\Pi_i, i=0,1$ the projection on the first,
-respectively second, component of a vector in $\bR^2$. The 
-forward modeling operator from pressure source to pressure trace is
-\begin{equation}
-  \label{eqn:sdef}
-  S^{\pm}_{z_s,z_r} = \Pi_0 {\cal S}^{\pm}_{z_s,z_r} \Pi_0^T 
-\end{equation}
-and the forward modeling operator from velocity source (normal force)
-to velocity trace is
-\begin{equation}
-  \label{eqn:vdef}
-  V^{\pm}_{z_s,z_r} = \Pi_1 {\cal S}^{\pm}_{z_s,z_r} \Pi_1^T 
-\end{equation}
-
-With these conventions, we can write the version of the source
-subproblem \ref{eqn:esis} studied in this paper as
-\begin{equation}
-  \label{eqn:esisp}
-  \mbox{find }h_s\mbox{ to minimize }\|S^{+}_{z_s,z_r}h_s- d\|^2 +
-  \alpha^2\|Ah_s\|^2.
-\end{equation}
-
-It follows from the adjoint state method (see Appendix A for details) that
-\begin{equation}
-  \label{eqn:sadj1}
-  ({\cal S}^{\pm}_{z_s, z_r})^T = -{\cal S}^{\mp}_{z_r,z_s}
-\end{equation}
-
-Define $R$ to be the {\em time-reversal operator} on functions of
-space-time, $Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em
-  acoustic field time-reversal operator}
-\begin{equation}
-  \label{eqn:trdef}
-  {\cal R} \left(
-    \begin{array}{c}
-      p\\
-      \bv
-    \end{array}
-  \right) =
-  \left(
-    \begin{array}{c}
-      Rp\\
-      -R\bv
-    \end{array}
-  \right)
-\end{equation}
-Then 
-\begin{equation}
-  \label{eqn:trsadj}
-  {\cal R}{\cal S}^{\mp} = -{\cal S}^{\pm}_{z_r,z_s}{\cal R}
-\end{equation}
-Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that
-
-\begin{equation} 
-  \label{eqn:trtr}
- ({\cal S}^{\pm}_{z_s,z_r})^T = {\cal R}{\cal S}_{z_r,z_s}^{\pm}{\cal R}=
- -{\cal S}^{\mp}_{z_r,z_s}.
-\end{equation}
-
-The relation \ref{eqn:trtr} implies that
-\begin{eqnarray}
-  (S^{\pm}_{z_s,z_r})^T &=& -S^{\mp}_{z_r,z_s} \nonumber\\
-                        &=& R S^{\pm}_{z_r,z_s}R, \nonumber\\
-    (V^{\pm}_{z_s,z_r})^T &=& -V^{\mp}_{z_r,z_s} \nonumber\\
-                        &=& R V^{\pm}_{z_r,z_s}R.
-                            \label{eqn:trtrcomp}
-\end{eqnarray}
-
-\section{Pressure-to-Source}
-
-Since the system \ref{eqn:awepm} has a unique solution by standard
-theory \cite[]{Lax:PDENotes}, the source vector field $(h_s,f_s)$
-determines the acoustic field $(p^{\pm},\bv^{\pm})$ in space time, and
-in particular the limits from the right at $z=z_s$, $P_sp^{\pm}$ and
-$P_sv_z^{\pm}$. This relation is not invertible: it is not possible to
-prescribe both pressure and normal velocity on a surface such as
-$z=z_s$. So the columns of the matrix operator
-${\cal S}^{\pm}_{z_s,z_r}$ must satisfy a linear relation. In this
-section I will explain this relation; it involves the {\em
-  pressure-to-source} map. This operator also turns out to be the
-principal component of a preconditioning strategy for iterative
-solution of the optimization problem \ref{eqn:esis}, so I will devote
-some effort to its proper definition. It is closely related to the
-Dirichlet-to-Neumann operator mentioned in the introduction.
-
-While it is not possible to prescribe both pressure and velocity on
-$z=z_s$ in solutions of \ref{eqn:awepm}, it is possible to
-prescribe pressure only, for instance: if the function $\phi$ on
-the surface $z=z_s$ satisfies suitable conditions, for
-example the downgoing constraint mentioned earlier, a unique solution
-exists for the acoustic system in both half-spaces $\pm z > z_s$:
-\begin{eqnarray}
-\label{eqn:awe0}
-  \frac{1}{\kappa}\frac{\partial p_{\pm}}{\partial t} & = & - \nabla \cdot \bv_{\pm}, \nonumber \\
-  \rho\frac{\partial \bv_{\pm}}{\partial t} & = & - \nabla
-                                                    p_{\pm}, \nonumber \\
-  p_{\pm} & =& 0,  \mbox{ for } t \ll 0, \nonumber\\ 
-  \bv_{\pm} & = & 0 \mbox{ for } t \ll 0, \nonumber\\
-  \lim_{z \rightarrow z_s^{\pm}}p_{\pm}(x,y,t,z)& =& \phi(x,y,t).
-\end{eqnarray}
-Note that the subscript $\pm$ here refers to the sign of $z-z_s$, as opposed
-to the superscript ${\pm}$, which refers to the sign of $t$ throughout
-this paper.
-
-From the boundary condition (last equation in \ref{eqn:awe0}), one
-sees that the pressures $p_{\pm}$ in the two half-spaces have the same
-limit at the boundary $z=z_s$. Stick the two half-space
-solutions together to form an acoustic field $(p^+,\bv^+)$ in all of
-space-time, that is,
-\begin{equation}
-  \label{eqn:awealt}
-  p^+(x,y,z,t) =
-  \left\{
-    \begin{array}{c}
-      p_+(x,y,z,t) \mbox{ if } z>0,\\
-      p_-(x,y,z,t) \mbox{ if } z<0,
-    \end{array}
-  \right.
-\end{equation}
-and a similar definition for $\bv^+$. Then $p^+$ is continuous across
-$z=z_s$, and the boundary condition in system \ref{eqn:awe0} may be
-written as $P_sp^+=\phi$.
-
-The same construction can be carried out in the anti-causal sense,
-with anti-causal half-space solutions glued together to form a
-full-space distribution solution $(p^-,\bv^-)$, with the property that
-$p^-$ is continuous across $z=z_s$ and $P_sp^-=\phi$.
-
-The reader may object that the notation $(p^\pm,\bv^\pm)$ is already in
-use, for the solution of \ref{eqn:awepm}. This objection is
-valid. However, {\em in the sense
-  of distributions}, $(p^{\pm},\bv^{\pm})$ as defined in display
-\ref{eqn:awealt}, is {\em exactly} the causal solution of \ref{eqn:awepm}
-for the choice $h_s = -[v^{\pm}_{z}]|_{z=z_s}, f_s=0$, as follows from a
-simple integration-by-parts calculation. So the notation is consistent!
-
-The negative jump $-[v^{\pm}_{z}]|_{z=z_s}$ is thus a function of $\phi$. Define
-the {\em pressure-to-source} operator $\Lambda^{\pm}_{z_s}$ by
-\begin{equation}
-  \label{eqn:deflam}
-  \Lambda^{\pm}_{z_s}\phi = -[v^{\pm}_{z}]|_{z=z_s}
-\end{equation}
-The conclusion: if $h_s = \Lambda^{\pm}_{z_s}\phi$ and $f_s=0$ in the
-system \ref{eqn:awepm}, then $\phi=P_sp^{\pm}$.
-
-Otherwise put, $S^{\pm}_{z_s,z_s}\Lambda^{\pm}_{z_s} \phi = \phi$, so
-$\Lambda^{\pm}_{z_s}$ is inverse to $S^{\pm}_{z_s,z_s}$. The relation
-\ref{eqn:trtrcomp} implies in turn that
-\begin{equation}
-  \label{eqn:lamadj}
-  (\Lambda^{\pm}_{z_s})^T = - \Lambda^{\mp}_{z_s}
-\end{equation}
-
-There is also a {\em velocity-to-source} operator. For the solution
-$(p^{\pm},\bv^{\pm})$ of system \ref{eqn:awepm} with $h_s=0$, the
-normal component of velocity, $v^{\pm}_z$, is continuous across
-$z=z_s$, and the velocity source (vertical load)
-$f_s=-[p^{\pm}]_{z=z_s}$. I will not name the velocity-to-source
-operator, as it does not appear explicitly in the developments to
-follow. As will be seen, it is essentially the inverse of the
-pressure-to-source operator.
-
-The quadratic form defined by $\Lambda^{\pm}_{z_s}$ has fundamental
-physical significance. Define the total acoustic energy $E^{\pm}(t)$ of the
-field $(p^{\pm},\bv^{\pm})$, at time $t$ by
-\begin{equation}
-  \label{eqn:defae0}
-  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)
-\end{equation}
-Then
-\begin{equation}
-  \label{eqn:elim}
-  \pm \lim_{\pm t \rightarrow \infty} E^{\pm}(t) =  \langle P_sp^{\pm},
-  (\Lambda^{\pm}_{z_s} P_sp^{\pm}) \rangle_{L^2(z=z_s)}.
-\end{equation}
-That is, the value of the quadratic form defined by
-$\Lambda^{\pm}_{z_s}$, evaluated at the pressure trace on $z=z_s$,
-gives the total energy transferred from the source to the
-acoustic field over time. Since $E$ is itself a positive definite
-quadratic form in the acoustic field, it follows that $\pm
-\Lambda^{\pm}_{z_s}$ is positive semi-definite.
-
-While $\Lambda^{\pm}_{z_s}$ is positive semi-definite, it is not
-symmetric. However, it is {\em approximately symmetric} in the
-high-frequency sense. This fact follows from a
-geometric optics analysis of the half-space solution,
-assuming (as always) downgoing data:
-\begin{equation}
-  \label{eqn:lamappsim}
-  (\Lambda^{\pm}_{z_s})^T \approx \Lambda^{\pm}_{z_s}.
-\end{equation}
-
-The analysis also reveals that the solution components not continuous
-at $z=z_s$ are odd there:
-\begin{equation}
-  \label{eqn:odd1}
-  \lim_{z\rightarrow z_s^+} v^{\pm}_{z} \approx - \lim_{z\rightarrow z_s^-}
-  v^{\pm}_{z}
-\end{equation}
-for the solution of \ref{eqn:awepm} with $f_s=0$.
-Similarly, 
-\begin{equation}
-  \label{eqn:odd2}
-  \lim_{z\rightarrow z_s^+} p^{\pm}\approx - \lim_{z\rightarrow z_s^-}
-  p^{\pm}
-\end{equation}
-for the solution of \ref{eqn:awepm} with $h_s=0$. Here ``$\approx$''
-means in the sense of high frequency asymptotics, that is, that the
-difference between the two sides is relatively smooth, hence small if
-the data is highly oscillatory. Therefore if $f_s=0$ in system \ref{eqn:awepm},
-\begin{equation}
-  h_s = \Lambda^{\pm}_{z_s}P_sp^{\pm} = -[v^{\pm}_{z}]|_{z=z_s} \approx -2
-  P_sv^{\pm}_{z}
-  \label{eqn:tracejump10}
-\end{equation}
-Similarly, if $h_s=0$ in system \ref{eqn:awepm}, then
-\begin{equation}
-  \label{eqn:tracejump20}
-  f_s = -[p^{\pm}]|_{z=z_s} \approx -2 P_s p^{\pm}.
-\end{equation}
-Thus $f_s$ determines approximately the boundary value of $p^{\pm}$,
-as a solution of the acoustic wave system in the half-space
-$z>z_s$. However, as repeated in equation \ref{eqn:tracejump10}, a
-solution with this boundary value is also the restriction to $z>z_s$
-of a solution to \ref{eqn:awepm} with $f_s=0$ and $h_s=
-\Lambda^{\pm}_{z_s}P_sp^{\pm}$. Therefore if
-\begin{equation}
-  \label{eqn:hfcondn}
-  h_s =-\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,
-\end{equation}
-then the pressure boundary value $P_sp^{\pm}$ is the
-same for the solutions of \ref{eqn:awepm} for source vectors $(h_s,0)$
-and $(0,f_s)$. Since the pressure boundary values are the same, the solutions
-in $z>z_s$ are the same. In particular, since $z_r>z_s$ and ${\cal
-  S}^{\pm}_{z_s,z_r}(h_s,f_s)^T = (P_rp^{\pm},P_rv^{\pm}_z)^T$, it follows
-that
-\begin{equation}
-  \label{eqn:snull}
-  {\cal S}^{\pm}_{z_s,z_r}\left(\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,f_s\right)^T \approx 0.
-\end{equation}
-
-Equation \ref{eqn:snull} states the relation between the columns of $
-{\cal S}^{\pm}_{z_s,z_r}$ mentioned in the introduction to this
-section.
-
-\section{Time Reversal}
-
-Recall that the source vector $(h_s,f_s)$ is assumed to produce a
-downgoing field $(p^+,\bv^+)$, that is, emanates high-frequency energy only along
-rays that make an angle with the vertical bounded below by a common
-minimum angle. Such rays leave $\Omega$ within a common maximum
-time. Consequently (Appendix B), in the
-slab $z_s<z<z_r$, the field $(p^+,\bv^+)$ approximates the solution of an
-anti-causal evolution equation. Choose $\chi(t)$ to be a smooth function
-that is $= 0$ for $t \gg 0$ and $=1$ at times when near rays carrying
-high-frequency energy in $(p^+,\bv^+)$ cross $z=z_r$. Define 
-$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
-$\Omega \times \bR$ of
-\begin{eqnarray}
-\label{eqn:revawe}
-  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
-  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
-  \tilde{p}^- & =& 0,  \mbox{ for } t \gg 0\\ 
-  \tilde{\bv}^- & = & 0 \mbox{ for } t \gg 0\\
-  P_r\tilde{p}^- &=& \chi P_rp^+ . 
-\end{eqnarray}
-That is, $\tilde{p}^-$ has the same boundary value on $z=z_r$ as
-$p^+$, except for low-frequency residue that is muted by
-$\chi$. Therefore
-$p^+ \approx \tilde{p}^-, \bv^+ \approx \tilde{\bv}^-$ near
-$z=z_r$. Since the right-hand sides in the system \ref{eqn:awepm} are
-singular only on $z=z_s$, and the high-frequency components of
-$(p^+,\bv^+)$ are carried by downgoing rays, these differ negligibly
-from the the high-frequency components of
-$(\tilde{p}^-,\tilde{\bv}^-)$ in the space-time slab $z_s<z<z_r$, and
-the approximation holds throughout this region. In particular
-$P_sv^+_z \approx P_s \tilde{v}^-_z$. In view of the relation
-\ref{eqn:tracejump10},
-\begin{equation}
-  \label{eqn:tildevtohsubs}
-  -2P_s\tilde{v}^-_z \approx h_s,
-\end{equation}
-so solution
-of the system \ref{eqn:revawe} followed by restriction to $z=z_s$ and
-multiplication by $-2$ 
-approximately inverts the map $S^+_{z_s,z_r}: h_s \mapsto P_rp^+$.
-
-Next observe that in view of the relation \ref{eqn:tracejump20}, and
-the downgoing nature of the ray system carrying the high frequency
-energy in $(p^+,\bv^+)$, the field $(\tilde{p}^-,\tilde{\bv}^-)$ is
-actually the restriction to $z<z_r$ of the anti-causal solution of \ref{eqn:awepm}
-with $z_s$ replaced by $z_r$, zero constitutive defect, and vertical
-load given by the jump in pressure at $z=z_r$ - for this field, use
-the same notation. Continuity of vertical
-velocity $\tilde{v}^-_z$ at $z=z_r$ implies that the vertical load is
-\[
-  f_r = -[\tilde{p}^-]|_{z=z_r} =-(\lim_{z\rightarrow
-    z_r^+}\tilde{p}^- - \lim_{z\rightarrow
-    z_r^-}\tilde{p}^-)
-\]
-\[
-  \approx 2 P_r \tilde{p}^- = 2 P_r p^+
-\]
-(from the definition \ref{eqn:defsamp}, $P_r$ is the limit from the
-left). Thus
-\[
-  P_s \tilde{v^-_z} \approx V^-_{z_r,z_s}(2 P_rp^+) \approx
-  2V^-_{z_r,z_s}S^+_{z_r,z_s}h_s.
-\]
-so
-\[
-  h_s \approx -2 P_s v^+_z \approx -2 P_s \tilde{v}^-_z \approx
-  -4V^-_{z_r,z_s}S^+_{z_r,z_s}h_s
-\]
-Combine this observation with \ref{eqn:tildevtohsubs} to obtain
-\[
- -4  V^-_{z_r,z_s} S^+_{z_s,z_r}  \approx  I,
-\]
-This relation combines with the identity \ref{eqn:trtrcomp} to
-yield the first main result of this section:
-\begin{eqnarray}
-  \label{eqn:approxinv}
-  (V^+_{z_s,z_r})^T S^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
-  (S^+_{z_s,z_r})^T V^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
-  V^+_{z_s,z_r} (S^+_{z_s,z_r})^T & \approx & \frac{1}{4}I, \nonumber\\
-  S^+_{z_s,z_r} (V^+_{z_s,z_r})^T & \approx & \frac{1}{4}I.
-\end{eqnarray}.
-The second equation is simply the transpose of the first, and the
-last two follow by by an exactly analogous argument using time
-reversal and interchange of the roles of $z_s$ and$z_r$.
-
-The conclusion is significant enough to merit restating in English:
-provided that high-frequency energy in the various fields is carried
-along downgoing ray fields, the transpose of $V^+$ is an approximate
-inverse to $S^+$, modulo a factor of 4. To recover the pressure source
-$h_s$ generating a pressure gather $P_rp$ at $z=z_r$, multiply the
-latter by -2, then apply the transpose of $V^+_{z_s,z_r}$ to this
-gather, reading out a vertical velocity field at $z=z_s$. Multiply
-again by -2 and you have a high-frequency approximation to $h_s$.
-
-\section{Unitarity}
-
-The next chapter in this story recognizes the relations in display
-\ref{eqn:approxinv} as asserting the approximate unitarity of
-$S^+_{z_s,z_r}$.
-
-The matrix identity \ref{eqn:snull} implies a relation between $S, V,$
-and $\Lambda$ of some interest in itself. After minor re-arrangement, the second row of reads
-\begin{equation}
-  \label{eqn:snull2}
--\frac{1}{2}\Pi_1{\cal S}^{\pm}_{z_s,z_r}\Pi_0^T\Lambda^{\pm}_{z_s}  \approx
-V^{\pm}_{z_s,z_r}.
-\end{equation}
-In these relations, the projection on the left picks out the vertical velocity component
-of a downgoing wavefield at $z=z_r$: that is,
-\[
--\frac{1}{2}\Pi_1{\cal S}^{\pm}_{z_s,z_r}\Pi_0^T\Lambda^{\pm}_{z_s}P_sp^+
-=-\frac{1}{2}P_r v_z^+,
-\]
-where $(p^+, \bv^+)$ solve the system \ref{eqn:awepm} with $f_s=0$ and
-$h_s = \Lambda^{\pm}_{z_s}P_sp^+$. On the other hand, from relation
-\ref{eqn:tracejump10},
-\[
-  P_r v_z^+ = -\frac{1}{2}\Lambda^+_{z_r}P_r p^+
-\]
-where
-\[
-  P_r p^+ = \Pi_0{\cal S}^{+}_{z_s,z_r}\Pi_0^T\Lambda^{+}_{z_s}P_s
-  p^+
-\]
-\[
-  = S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_sp^+
-\]
-Therefore combining the last two equations with \ref{eqn:snull2},
-obtain
-\begin{equation}
-  \label{eqn:sv}
-  \frac{1}{4}\Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s} = V^+_{z_s,z_r}.
-\end{equation}
-This is the promised relation.
-
-As shown in the last section, $4(V_{z_s,z_r}^+)^T$ is approximately
-inverse to $S^{+}_{z_s,z_r}$. Therefore, transposing both sides of
-equation \ref{eqn:sv} and using \ref{eqn:approxinv}, obtain
-\begin{equation}
-  \label{eqn:almostunitary}
-  4(V_{z_s,z_r}^+)^TS^+_{z_s,z_r} = [ (\Lambda^+_{z_s})^T
-  (S^{+}_{z_s,z_r})^T(\Lambda^+_{z_r})^T]S^{+}_{z_s,z_r} \approx I.
-\end{equation}
-
-The remarkable feature of the identity \ref{eqn:almostunitary} is that
-it exhibits an approximate right inverse of $S^+$ as an adjoint with
-respect to a weighted inner product - or it would, if the operators
-$(\Lambda^+)$ were symmetric positive definite. As noted earlier,
-these operators are only approximately symmetric, though they are
-positive semi-definite. That is not a great obstacle, however:
-symmetrizing them in the obvious way commits a negligible error, of
-the sort that this paper already neglects wholesale. That is,
-\begin{equation}
-  \label{eqn:unitary}
-  [ \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
-  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
-  \Lambda^+_{z_r})]S^{+}_{z_s,z_r} \approx I.
-\end{equation}
-
-The symmetrized $\Lambda$ operators are at least positive
-semi-definite, hence define (at least) semi-norms.
-Similar relations have been derived for other scattering operators,
-and have been used to accelerate iterative solutions of inverse
-scatering problems: \cite{DafniSymes:SEG18b} review some of this
-literature.
-
-\section{Accelerated Iterative Inversion}
-
-For convenience, in this section write $S$ in place of
-$S^+_{z_s,z_r}$. Also abbreviate the symmetrized $\Lambda$ operators
-using notation suggesting weight
-operators in model and data spaces:
-\begin{eqnarray}
-  W_m^{-1}&=& \frac{1}{2}((\Lambda^+_{z_s})^T+
-              \Lambda^+_{z_s}),\nonumber \\
-  W_d &=& \frac{1}{2}((\Lambda^+_{z_r})^T+ \Lambda^+_{z_r}).
-          \label{eqn:wdef}
-\end{eqnarray}
-The identification of the symmetrized $\Lambda^+_{z_s}$ as the inverse
-of another operator $W_m$ is formal, since the former operator is
-likely to have null (or nearly-null) vectors due to aperture-related
-amplitude loss. Since some version of $W_m$ is essential in the
-formulation for effective preconditioning, I will derive a usable
-candidate to stand in for it below.
-
-Adopting Hilbert norms defined by the operators $W_m$ and $W_d$ in its
-domain and range respectively, the adjoint of $S$ is given by
-\begin{equation}
-\label{eqn:wadj}
-S^{\dagger} = W_m^{-1}S^TW_d,
-\end{equation}
-
-In this notation, the relation \ref{eqn:unitary} takes the form
-\begin{equation}
-  \label{eqn:wunitary}
-  S^{\dagger}S \approx I.
-\end{equation}
-That is to say, $S$ is approximately unitary with respect to the
-weighted norms defined by $W_m$ and $W_d$. Therefore a Krylov space
-method employing these norms will converge rapidly, at least for the
-well-determined components of the solution.
-
-Similar relations have been derived for other scattering operators,  
-and have been used to accelerate iterative solutions of inverse  
-scattering problems: \cite{DafniSymes:SEG18b} review some of this  
-literature.
-
-The most convenient arrangement the Conjugate Gradient (CG) algorithm
-taking advantage of the structure \ref{eqn:wadj} is the {\em
-  Preconditioned CG}.
-
-Allowing that the fit error will be measured by the data space norm,
-the least squares problem to be solved is not just $Sh \approx d$, but
-a regularized version:
-\begin{equation}
-  \label{eqn:einv}
-  \mbox{minimize}_h \|Sh-d\|^2_d + \alpha^2 \|Ah\|^2_m
-\end{equation}
-Remark: recall that the modified data space norm $\|d\|_d^2 = \langle
-d, W_d d\rangle$ has physical meaning: for acoustics, it is
-proportional to the power transmitted to the fluid by the source.
-
-The minimizer of the objective defined in equation \ref{eqn:einv}
-solves the normal equation
-\begin{equation}
-  \label{eqn:norm0}
-  (S^{\dagger}S + \alpha^2 A^{\dagger}A)h = S^{\dagger}d 
-\end{equation}
-where the weighted adjoint $S^{\dagger}$ has already been defined in equation \ref{eqn:wadj}, and $A^{\dagger}$ is the adjoint of $A$ in the weighted model space norm defined by $W_m$, namely
-\begin{equation}
-  \label{eqn:aadj}
-  A^{\dagger} = W_m^{-1}A^TW_m.
-\end{equation}
-
-Note that the normal operator appearing on the left-hand side of
-\ref{eqn:norm0} is not an approximate identity, due to the presence of
-the regularization term: the spectrum increases in spread with
-increasing $\alpha$, leading to slower convergence. Fortunately for
-the present setting, the operators $W_m^{-1}$, $A$, and $W_m$
-approximately commute (they are scalar {\em pseudodifferential}, once
-the difficulties with the definition of $W_m$, mentioned above, are
-taken care of), so that $A^{\dagger} \approx A^T$. Therefore
-\begin{equation}
-  \label{eqn:normapprox}
-  S^{\dagger}S + \alpha^2 A^{\dagger}A \approx I + \alpha^2A^TA
-\end{equation}
-Recall that $A$ is simply multiplication by the Euclidean distance to
-the physical source point $\bx_s$: $A u (\bx) = |\bx-\bx_s|u(\bx),
-A^TAu(\bx) = |\bx-\bx_s|^2u(\bx)$. So the equation $(I+\alpha^2
-A^TA)u=b$ is trivial to solve, and this is a key characteristic of a
-good preconditioner. However this observation must be combined with
-the weighted norm structure.
-
-Rewrite the normal equation \ref{eqn:norm0} as
-\begin{equation}
-  \label{eqn:norm1}
-  W_m^{-1}(S^TW_dS + \alpha^2 A^TW_mA)h = W_m^{-1}S^TW_md 
-\end{equation}
-Since $W_m$ is self-adjoint and positive definite, the common factor on both sides of \ref{eqn:norm1} can be re-written as
-\begin{equation}
-  \label{eqn:normpart}
-  Nh = (S^*S + \alpha^2 A^*A)h = S^*d 
-\end{equation}
-in which $S^*, A^*$ are the adjoints with the original (Euclidean)
-inner product in the domains but the weighted inner product in data
-space:
-\begin{eqnarray}
-  \label{eqn:sadj}
-  S^* &=& S^T W_d,\\
-  A^* &=& A^T W_m.
-\end{eqnarray}
-Note the $S^*S$ and $A^*A$ are symmetric in the Euclidean sense, so
-equation \ref{eqn:normpart} is a symmetric positive (semi-)definite
-linear system, just the sort of thing for which the 
-The Preconditioned Conjugate Gradient (``PCG'') algorithm was
-designed. PCG for solution
-of equation \ref{eqn:normpart} with preconditioner $M$is usually
-written as Algorithm 1 (see for example \cite{Golub:2012}):
-
-\begin{algorithm}[H]
-\caption{Preconditioned Conjugate Gradient Algorithm, Standard Version}
-\begin{algorithmic}[1]
-\State Choose $h_0=0$ 
-  \State $r_0 \gets S^*d$
-  \State $p_0 \gets M^{-1} r_0$
-  \State $g_0 \gets p_0$
-  \State $q_0 \gets Np_0$
-  \State $k \gets 0$
-  \Repeat
-  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle p_k,q_k\rangle}$
-  \State $h_{k+1} \gets h_k + \alpha_k p_k$
-  \State $r_{k+1} \gets r_k - \alpha_kq_k$
-  \State $g_{k+1} \gets M^{-1} r_{k+1}$
-  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
-  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
-  \State $q_{k+1} \gets Np_{k+1}$
-  \State $k \gets k+1$
-  \Until{Error is sufficiently small, or max iteration count exceeded} 
-\end{algorithmic}
-\end{algorithm}
-The iteration converges rapidly if $M^{-1}N \approx I$. This is true
-if and only if the symmetrized operator $M^{-1/2}NM^{-1/2} \approx I$,
-which is in turn true if the eigenvalues of $M^{-1/2}NM^{-1/2}$ are
-close to 1 (actually works well is most of these eigenvalues are close
-to 1, and the rest are small - which is the case for the current
-problem)..  Further, PCG is computationally effective is M is easy to
-invert.
-
-From \ref{eqn:normapprox} and \ref{eqn:norm1}, it follows that
-\[
-  W_m^{-1}(S^TW_dS + \alpha^2 A^TW_mA) \approx I + \alpha^2 A^TA.
-\]
-This observation suggests using $M=W_m(I+\alpha^2A^TA)$. This choice
-is not symmetric, but since the operators on the right-hand side are
-scalar pseudodifferential hence commute, it is equivalent to use of
-\begin{eqnarray}
-  M         &=&(I+\alpha^2A^TA)^{1/2}W_m(I+\alpha^2A^TA)^{1/2},\nonumber \\
-  M^{-1}
-            &=&(I+\alpha^2A^TA)^{-1/2}W_m^{-1}(I+\alpha^2A^TA)^{-1/2}.
-                \label{eqn:defprecond}
-\end{eqnarray}
-With this choice, \ref{eqn:normapprox} implies that
-$M^{-1}N \approx I$, also $M$ is symmetric. As already mentioned,
-powers of $I + \alpha^2A^TA$ are trivial to compute, given the choice
-of $A$ made here. We will examine fast algorithms for computing
-$W_m^{-1}$ = the symmetrized pressure-to-source operator in the next
-section. Note that only $M^{-1}$, hence only $W_m^{-1}$, appears in
-Algorithm 1.
-
-\section{Computing and Symmetrizing $\Lambda$}
-
-Computations of $\Lambda^{\pm}_{z_s}$ and its transpose are clearly critical steps in an
-implementation of the PCG algorithm outlined in the preceding section.
-Direct computation of the pressure-to-source operator $\Lambda^{\pm}_{z_s}$, for instance by solving
-\ref{eqn:awepm} and reading off $P_sv^{\pm}_{z}$, turns out to be
-numerically ill-behaved. The relation \ref{eqn:snull} provides and
-alternative approach, taking advantage of the accurate approximate
-inverse to $S^+_{z_s,z_r}$ constructed above. The first row of
-\ref{eqn:snull}, slightly rearranged, is
-\begin{equation}
-  \label{eqn:lamidea}
-  \Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx  -\frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
-\end{equation}
-The approximate inverse construction for $S^++_{z_s,z_r}$ permits (approximate)
-solution of this equation for $\Lambda^+_{z_s}f_s$: apply
-$4(V^{+}_{z_s,z_r})^T$ to both sides of equation \ref{eqn:lamidea} and
-use the first equation in the list \ref{eqn:approxinv} to get
-\begin{equation}
-  \label{eqn:lamident}
-  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
-\end{equation}
-This identity is the major result of this section: it shows how
-to compute that action of $\Lambda^+_{z_s}$ by propagating the input
-pressure trace, identified as a source for the velocity evolution,
-forward in time from $z_s$ to $z_r$
-reading off the pressure trace on $z=z_r$, identifying it once more as
-a point load (source for velocity), propagating it backwards in time from $z_r$ to
-$z_s$, and finally reading off the velocity trace, interpreted as a
-pressure evolution source on $z_s$.
-
-The importance of this result lies in the failure of the obvious
-method for computing the action of $\Lambda^{\pm}_{z_s}$, namely to
-employ the pressure trace as a source in the velocity equation ($f_s$,
-in the notation used above) at $z=z_s$, and read off the velocity
-field also at $z=z_s$. This difficulty is related to the existence of
-tangentially propagating waves and the lack of continuity of the trace
-operator. The method implicit in equation \ref{eqn:lamident} avoids
-this difficulty by propagating the fields a positive distance in $z$:
-assuming as always that the causal fields are downgoing, this step
-eliminates any tangentially propagating fields from consideration.
-
-A deeper study of the pressure-to-source
-operator (Appendix B) shows that it is approximately dependent only on the model
-coefficients near the source surface ($z=z_s$ in this case). Since the
-homogeneous and lens models are identical near this surface, it is
-unsurprising that these figures are very close to the previous two.
-However an even more useful observation is that the calculations in
-the approximation \ref{eqn:lamident} could just as well be carried out
-in a much smaller region around the source surface, and produce a
-result that is functionally identical in that it will serve as a
-source for the same acoustic fields globally, with small error. In
-effect, equation \ref{eqn:lamident} involving propagation from source
-($z=z_s$) to receiver ($z=z_r$) surfaces is altered by replacing $z_r$
-with a receiver datum $z_s+\Delta z$ considerably closers to $z_s$:
-\begin{equation}
-  \label{eqn:lamnear}
-  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
-\end{equation}
-Using a receiver datum closer to the source surface has two favorable consequences:
-\begin{itemize}
-\item The computational domain can be smaller than is necessary to
-  simulate the target data, as it need only contain the source
-  surface and the receiver datum implicit in
-  equation \ref{eqn:lamident}. This shrinkage of the computational
-  domain can lead to substantial improvements in computational
-  efficiency.
-\item Since the receiver data may be chosen much closer to the
-  source surface that is the case for the target data, the effective
-  aperture active in the relation \ref{eqn:lamident} can be much
-  larger, producing an estimated source gather much less affected by
-  aperture limitation.
-\end{itemize}
-
-As mentioned in the last section, computation of the transpose of
-$\Lambda^+$ (exact, not approximate in the high frequency sense) is
-critical to the successful construction of the preconditioner. The
-relation \ref{eqn:lamident} does not provide a computation for this
-operator. However set
-\begin{equation}
-  \label{eqn:lamtilde}
-  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
-\end{equation}
-Then \ref{eqn:lamnear} can be rewritten
-\[
-  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
-\]
-Of course, all of the examples so far show images of
-$\tilde{\Lambda}^+_{z_s}$.
-
-Since successful preconditioning requires only approximate inversion,
-use of $\tilde{\Lambda}^+_{z_s} $ in place of $\Lambda^+_{z_s}$ will
-still yield a working preconditioner, and the former can be transposed
-to machine precision via the definition \ref{eqn:lamtilde} and the adjoint state
-method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
-\begin{equation}
-  \label{eqn:lamtransp}
-  (\tilde{\Lambda}^+_{z_s})^T = -8 \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
-    z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}
-\end{equation}
-
-The model space weight operator $W_m^{-1}$ introduced in the last section is
-replaced by its asymptotic approximation
-\[
-\frac{1}{2}(\tilde{\Lambda}^+_{z_s} +
-  (\tilde{\Lambda}^+_{z_s})^T)
-\]
-\[
-  \approx -8 \left((V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T \right.+
-\]
-\[
-  \left. \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
-      z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}\right)
-\]
-\begin{equation}
-  \label{eqn:winvcomp}
-  = -4(\Pi_1 ({\cal S}^+_{z_s,z_s+\Delta z})^T(\Pi_0^T\Pi_1 +
-  \Pi_1^T\Pi_0) {\cal S}^+_{z_s,z_s+\Delta z}\Pi_1^T) = \tilde{W}_m^{-1} .
-\end{equation}
-with a similar definition for the replacement $\tilde{W}_d$ of $W_d$.
-
-This identity shows that only one forward and one adjoint simulation
-are necessary to compute the action of $\tilde{W}_{m,d}$. The operator
-in the center of the expression on the right-hand side, $\Pi_0^T\Pi_1
-+ \Pi_1^T\Pi_0$, simply exchanges the components of the acoustic
-fields, passing the velocity field as a pressure source and the
-pressure field as a velocity source.
-
-One more computation is required for the full implementation of the
-preconditioning strategy explained in the last section: $W_m$ is
-required, not just $W_m^{-1}$. Note that $W_m$ plays two roles in the
-second term in equation \ref{eqn:norm1}: it is the weight matrix for
-both the domain and range norms for $A$. It is perfectly OK for one of
-these to be replaced by an asymptotic approximation, so long as it is
-symmetric and computable (and at least semi-definite). The second row
-in equation \ref{eqn:snull} appears as \ref{eqn:snull2} above:
-introducing (formally) the inverse of $\Lambda^+$,
-\begin{equation}
-  \label{eqn:snull2mod}
--\frac{1}{2}\Pi_1{\cal S}^{+}_{z_s,z_r}\Pi_0^T  \approx
-V^{+}_{z_s,z_r}(\Lambda^{+}_{z_s})^{-1}
-\end{equation}
-whence from the second line in display \ref{eqn:approxinv}
-\begin{equation}
-  \label{eqn:laminv}
--\frac{1}{8}(S^+_{z_s,z_r})^T\Pi_1{\cal S}^{+}_{z_s,z_r}\Pi_0^T  
-\approx (\Lambda^{+}_{z_s})^{-1}
-\end{equation} 
-and
-\begin{equation}
-  \label{eqn:laminvtr}
--\frac{1}{8}\Pi_0({\cal S}^{+}_{z_s,z_r})^T\Pi_1^T S^+_{z_s,z_r}
-\approx ((\Lambda^{+}_{z_s})^{-1})^T.
-\end{equation}
-Using the definition \ref{eqn:sdef} of $S^+_{z_s,z_r}$, the
-symmetrized $\Lambda^{-1}$ is
-\begin{equation}
-  \label{wcomp}
- \tilde{W}_m = -\frac{1}{16}\left(\Pi_0 ({\cal S}^{+}_{z_s,z_r})^T
-   (\Pi_1^T \Pi_0 + \Pi_0^T \Pi_1) {\cal S}^{+}_{z_s,z_r} \Pi_0^T\right)
-   \approx \frac{1}{2}((\Lambda^{+}_{z_s})^{-1}
-   +((\Lambda^{+}_{z_s})^{-1})^T) .
-\end{equation}
-Comparison with the definition \ref{eqn:winvcomp} shows that
-$\tilde{W}_m$ and $\tilde{W}_m^{-1}$ differ only in the initial and
-final projection factors (and overall scale), and in particular either
-can be computed for the cost of a forward/adjoint operator pair. Note
-that $\tilde{W}_m^{-1}$ is inverse to $\tilde{W}_m$ only in an
-approximate (asymptotic, aperture-limited) sense.
-
 \section{Numerical Examples}
@@
   vs. iteration. Data = lens model, point source (Figure 
-  \ref{fig:recplh0}), inversion in homogenous model. Penalty weight 
+  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
   $\alpha=10^{-3}$.}
-
-\section{Conclusion}
-
-\append{Adjoint Computation}
-
-The adjoint of ${\cal S}^+_{z_s,z_r}$ can be computed by a variant of
-the adjoint state method, in this case a by-product of the
-conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awepm}
-with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
-$ (h_r,f_r\bf{e}_z) \delta(z-z_r)$. Then
-\[
-0 = 
-\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
-\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
--
-\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
-\]
-\[
-+
-\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
- h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
- h_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  (- \nabla p^++f_s\bf{e}_z) \cdot \bv^- + \bv^+ \cdot (-\nabla
-  p^- + f_r \bf{e_z}) \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
- h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
- h_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- 
-  +f_s \delta(z-z_s) v_z^- + v_z^+f_r \delta(z-z_r) \right)
-\]
-after integration by parts in the last two terms. Most of what is left cancels, leaving 
-\[
-0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
-( h_rP_rp^++f_rP_rv_z^+) = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
-\]
-whence
-\begin{equation}
-\label{eqn:sadj}
- ({\cal S}^+_{z_s,z_r})^T = -{\cal S}^{-}_{z_r,z_s}.
-\end{equation}
-
-The systems \ref{eqn:awepm} for $(p^{\pm},\bv^{\pm})$ with the two
-signs of superscript differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
-$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
-\begin{equation}
-  \label{eqn:trdef}
-  {\cal R} \left(
-    \begin{array}{c}
-      p\\
-      \bv
-    \end{array}
-  \right) =
-  \left(
-    \begin{array}{c}
-      Rp\\
-      -R\bv
-    \end{array}
-  \right)
-\end{equation}
-Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awepm} with
-$(h_s\delta(z-z_s),f_s\bf{e_z}\delta(z-z_s))$ replaced by
-$(-Rh_r\delta(z-z_r), Rf_r\bf{e_z}\delta(z-z_r))$. That is,
-\begin{equation}
-  \label{eqn:trsadj}
-  {\cal R}{\cal S}^- = -{\cal S}^+_{z_r,z_s}{\cal R}
-\end{equation}
-Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj}
-and \ref{eqn:trsadj} imply equation \ref{eqn:trtr}.
-
-\bibliographystyle{seg}
-\bibliography{../../bib/masterref}

### 2022-10-22 repro/papers/GEM22/submitR1/paper.tex
initial revision commit

--- 2022-04-03 repro/papers/GEM22/submit/paper.tex
@@
 \end{enumerate}
-The basic FWI problem is: given data $d$, find ${\bf c}$ so that 
+The basic FWI problem is: given data $d$, find ${\bf c}$ so that data
+is fit and wave physics are honored, that is, 
 \begin{equation}
@@
 \end{equation}
-In this formulation, the source function $\bff$ may be given, or
-to be determined subject to some constraints.
-
+The source function $\bff$ may be given, or to be determined along
+with $\bf{c}$.
 %The energy source $\bff$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time.
@@
 Therefore it is natural to view $\bff$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
-A simple nonlinear least squares formulation is:
+The task \ref{eqn:fwi} may be cast as a nonlinear least squares problem: 
 \begin{equation}
@@
 \end{equation}
-Practical optimization formulations typically augment the objective in
-\ref{eqn:ols} by additive penalties or other constraints.
+Practical variants of the least squares problem \ref{eqn:ols}
+typically augment the objective with additive penalties or other
+constraints \cite[]{VirieuxOperto:09,Fichtner:10,Schuster:17}.
 
@@
 approach given the dimensions of a typical instance of \ref{eqn:fwi},
-and those have a tendency to stall due to ``cycle-skipping''. See for
-example \cite{VirieuxOperto:09} and many references cited there. Source
-extension is one approach to avoiding this problem. It consists in
-imposing the wave equation as a soft as opposed to hard constraint, by
-allowing the source field $\bff$ to have more degrees of freedom than
-is permitted by a faithful model model of the seismic experiment, and
-constraining these additional degrees of freedom by means of an
-additive quadratic penalty modifying the problem \ref{eqn:ols}:
+and those have a tendency to stall at uninformative model estimates
+due to ``cycle-skipping''. This phenomenon is often interpreted as the
+capture of optimizing sequences at local minima of the nonlinear objective
+function in \ref{eqn:ols} far from the global minimum, or at least not
+possessing the characteristics of a useful solution. See for
+example \cite{GauTarVir:86,VirieuxOperto:09,PladysBrossierLiMetivier:GEO21}.
+
+Source extension is one approach to avoiding this``cycle-skipping''
+obstacle. It consists in imposing the wave equation as a soft
+constraint, allowing the source field $\bff$ to have more degrees of
+freedom than is permitted by a faithful model model of the seismic
+experiment, and constraining these additional degrees of freedom by
+means of an additive quadratic penalty modifying the problem
+\ref{eqn:ols}:
 \begin{equation}
@@
 \end{equation}
-The operator $A$ penalizes deviation from known (or assumed)
+The linear operator $A$ penalizes deviation from known (or assumed)
 characteristics of the source function - its null space consists of
 feasible (or ``physical'') source models.
-
 \cite{HuangNammourSymesDollizal:SEG19} present an overview of the
 literature on source extension methods, describing a variety of
-methods to add degrees of freedom to physical source model. The present paper
-concerns {\em surface source extension}: physical sources are
-presumed to be concentrated at points $\bx_s$ in space, whereas their extended
-counterparts are permitted to spread energy over surfaces containing
-the physical source locations. A simple choice for the penalty
-operator $A$ is then multiplication by the distance $|\bx-\bx_s|$ to the physical
-source location:
+methods to add degrees of freedom to physical source model.
+
+The present paper concerns {\em surface source extension}: physical
+sources are presumed to be concentrated spatially at points $\bx_s$
+in space, whereas their extended counterparts are permitted to spread
+energy over a {\em source surface} containing the physical source spatial
+locations. Similarly, the receiver spatial locations are confined to a
+{\em receiver surface}. A simple choice for the penalty operator $A$ is then
+multiplication by the distance $|\bx-\bx_s|$ to the physical source
+location:
 \begin{equation}
@@
 
-The penalty operator $A$ defined in \ref{eqn:penop} is linear, so the source
+The modeling operator $PL[{\bf c}]^{-1}$ and the penalty operator $A$ defined in \ref{eqn:penop} are linear, so the source
 subproblem is a linear least squares problem. Under some additional
@@
 \ref{eqn:esis}. This approximate solution operator may be used to
-accelerate Krylov space methods for the solution of the surface source
-subproblem \ref{eqn:esis}. Numerical examples suggest the
-effectiveness of this acceleration.
-
-I will fully describe a preconditioner for a special
+accelerate (``precondition'') Krylov space methods for the solution of the surface source
+subproblem \ref{eqn:esis}. I will fully describe a preconditioner for a special
 case of the source subproblem \ref{eqn:esis}, in which ${\bf u}$ is an
 acoustic field, $L[{\bf c}]$ is the wave operator of linear
-acoustodynamics, the spatial positions of traces extracted by $P$ lie
-on a depth plane $z=z_r$, and the positions at which the extended
-source $\bff$ is nonzero lie on another, parallel, depth plane $z=z_s$. 
+acoustodynamics, and the source and receiver surfaces are horizontal
+planes. Numerical examples in this setting suggest the
+effectiveness of this acceleration.
+
+I will use two 2D numerical models 
+throughout to illustrate the theory. In both, horizontal lines serve as
+source and receiver surfaces. The first is a ``crosswell'' or slab
+configuration with an acoustic lens positioned between a deeps source
+and a shallower line of 
+receivers, resulting in markedly triplicated arrivals. The goal in
+this first example is to construct a surface source that explains the data in a
+homogeneous medium (that is, inversion in a wrong velocity, emulating
+the early iterations of FWI based on extended modeling). The second is a layered
+model in which velocity increases with depth, resulting the formation
+of diving waves. This configuration simulates an
+ocean-bottom node and a line of near-surface sources: the roles of
+source and receiver are switched for computational convenience. In
+this second example, the diving wave arrivals are isolated (by an
+appropriate mute operator) and a
+source constructed that explains them alone.
+
+The next section defines the modeling operator $PL[{\bf c}]$, its
+adjoint, and important specializations (pressure vs. normal velocity
+sources and data). It also introduces the two 2D examples mentioned
+above.  The following section constructs an approximate inverse of the
+modeling operator by time reversal (as suggested by work in
+photoacoustic tomography), and illustrates its efficacy in the two
+examples. This construction requires extraction of
+velocity data, or equivalently a surface source, from pressure data.
+The subsequent sections describe this pressure-to-source operator,
+express the approximate inverse as the modeling operator adjoint in
+weighted norms (thus establishing that the modeling operator is {\em
+  approximately unitary} in the sense of these norms), explain how to
+use this construction to precondition Conjugate Gradient iteration,
+and organize the preconditioning computation so as to involve only one
+extra and relatively inexpensive wave propagation calculation. I use
+the 2D examples to illustrate each of
+these developments. The paper ends with a brief
+discussion-and-conclusion section, reviewing what has been
+accomplished and listing a few of the many questions left open.
+
+%I will fully describe a preconditioner for a special
+%case of the source subproblem \ref{eqn:esis}, in which ${\bf u}$ is an
+%acoustic field, $L[{\bf c}]$ is the wave operator of linear
+%acoustodynamics, and the source and receiver surfaces are horizontal
+%planes.
+%spatial positions of traces extracted by $P$ lie
+%on a depth plane $z=z_r$, and the positions at which the extended
+%source $\bff$ is nonzero lie on another, parallel, depth plane $z=z_s$. 
 %That is, sources take the form of a combination of constitutive law
@@
 %surface vertical loads.
-This
-``crosswell'' configuration simplifies the analysis underlying the
-construction of approximate solutions for the source subproblem
-\ref{eqn:esis}. It is only one of many transmission configurations for
-which similar developments are possible. Perhaps the most important
-alternative example is the diving wave configuration, which plays a
-central role in contemporary FWI.
-
-The preconditioner construction is very similar to the time-reversal
-method in photoacoustic tomography
-\cite[]{StefanovUhlmannIP:09}. Preconditioning amounts to a change of
-norm in the domain and range spaces of the modeling operator. In this
-case, the modfied norms are weighted $L^2$, and the weight operators
-map pressure to corresponding surface source on the source and
-receiver planes. This pressure-to-source map is closely related to the ``hyperbolic
+%This configuration simplifies the presentation of the
+%approximate solution construction.
+
+\section{Overview}
+
+The preconditioner construction is closely related to the time
+reversal method in photoacoustic tomography
+\cite[]{StefanovUhlmannIP:09,Hristova:09}. A simplified mathematical
+translation of this medical imaging task is to infer the initial
+excess pressure distribution over a fluid-containing region at time
+$t=0$ from measurements of the pressure on a surface enclosing the
+fluid over a time interval $0 \le t \le t_{\rm max}$. This problem
+clearly has strong similarities to, but also differences from, the
+problem studied here. The time reversal method presumes that the
+pressure field has returned to equilibrium (zero excess pressure), or close to it, at the
+final time $t=t_{\rm max}$ Then the field can be (at least
+approximately) viewed as the backwards-in-time initial boundary value
+problem with zero final conditions at $t=t_{\rm max}$, and boundary
+values given by the measurements. Evolving the field backwards in time
+backwards to $t=0$ thus solves the problem. Except in special
+circumstances, the pressure field never actually vanishes at finite
+time, so the solution is approximate.
+
+The seismic surface source extension proplem \ref{eqn:esis} differs in
+several obvious ways from the photoacoustic setting. The measurement
+or receiver surface does not surround the region of wave
+propagation. It is not the initial pressure time-slice at $t=0$ that
+is to be determined, but a time-extended source $\bff$ confined to a
+surface. The penalty operator has no analogue in the basic
+photoacoustic problem description. Finally, once these obstacles are
+overcome, using the approximate solution operator to accelerate Krylov
+iteration for solution of the optimization problem \ref{eqn:esis}
+requires that the operator be identified as the adjoint of the
+modeling operator with respect to suitable inner products in its domain
+and range. In the pages to follow, I will address each of these issues.
+
+First, reverse-time propagation can be localized via ray theory,
+within high-frequency asymptotic approximation. This step requires
+some assumptions about the ray fields: all rays responsible for significant
+energy in the receiver data must arrive from the source surface, and must cross the source and receiver
+surfaces transversally, that is, making non-zero angles. Also, the
+rays must approach source and receiver surfaces from one side or the
+other (``the inside''), so that locally the surfaces can be treated as boundaries of
+propagation domains. With these assumptions, reverse-time propagation
+of the receiver data
+closely approximates the acoustic fields near the source surface. Next
+observe that for a causal field with a surface pressure source (with a
+continuous pressure field across the surface), the source is
+proportional to the jump in the (particle) velocity field. Moreover,
+to leading order in frequency, the velocity field switches sign at the
+surface, so the jump is just twice the sampled value on the
+surface. Thus reverse-time propagation and reading off the velocity
+field on the source surface yields
+a source that reproduces the pressure data, within an asymptotically small error.
+%Preconditioning amounts to a change of
+%norm in the domain and range spaces of the modeling operator. In this
+%case, the modfied norms are weighted $L^2$, and the weight operators
+%map pressure to corresponding surface source on the source and
+%receiver planes.
+
+This is the essence of the approximate solution of the problem
+\ref{eqn:esis} for penalty weight $\alpha=0$. To see how Krylov
+iteration might be accelerated, note that time-reverse propagation of
+the receiver pressure values may be represented by a time-reverse
+source propagation, with the source constructed from the velocity
+field at the receiver surface. Moreover, the reverse-time
+source-to-pressure propagation is the transpose (adjoint) of the
+forward-time source-to-pressure propagation - this is a very simple
+version of the adjoint state construction \cite[]{Plessix:06}. So the
+approximate inverse is: conversion of pressure to source at the
+receiver surface, followed by the adjoint of the modeling operator,
+followed by the conversion of pressure to source at the source
+surface. This sequence precisely describes the adjoint of the modeling
+operator with respect to weighted norms in the spaces of source and
+receiver data, with the weight operators being the
+pressure-to-source operators. That this construction results in an
+approximate inverse means that the modeling operator is approximately
+unitary with respect to these weighted norms, which in turn implies
+that the preconditioned conjugate gradient algorithm should converge rapidly.
+
+This pressure-to-source map is closely related to the ``hyperbolic
 Dirichlet-to-Neumann'' operator that plays a prominent role in
@@
 
-The obvious computation of the pressure-to-source map - prescribe the
-pressure, solve the wave equation with this boundary condition, read
-off the equivalent source - suffers from intrinsic numerical
-inaccuracy. I suggest an alternative computationally feasible
-approach, via economical short-distance wave propagation. Since the
-map is symmetric only in an approximate, asymptotic sense, it must be
-symmetrized for use as a Krylov preconditioner. I describe a
-symmetrization procedure that requires no further wave computations
-beyond those necessary to compute the action of the operator itself.
+FInally, the $\alpha=0$ case is not sufficient: the penalty operator
+$A$ is an essential component of the nonlinear inverse propblem
+\ref{eqn:esi}. As it turns out, $A$ commutes with the other operators
+involved - approximately, but that is enough. Therefore the effect of
+$A$ can be compensated with an easily-computed factor, whence the
+preconditioned acceleration extended to the case $\alpha > 0$.
+
+%The obvious computation of the pressure-to-source map - prescribe the
+%pressure, solve the wave equation with this boundary condition, read
+%off the equivalent source - suffers from intrinsic numerical
+%inaccuracy. I suggest an alternative computationally feasible
+%approach, via economical short-distance wave propagation. Since the
+%map is symmetric only in an approximate, asymptotic sense, it must be
+%symmetrized for use as a Krylov preconditioner. I describe a
+%symmetrization procedure that requires no further wave computations
+%beyond those necessary to compute the action of the operator itself.
 
@@
 the sampling
-operator $P$ on the space-time plane $z=z_r$ is not well-defined for arbitrary
+operator $P$ on the receiver surface is not well-defined for arbitrary
 finite-energy acoustic fields. Thus the modeling operator envisioned above
-may not be well-defined.
-
-This
-phenomenon is related to the ill-posedness of wave equations as
+is not be well-defined, strictly speaking.
+
+This phenomenon is related to the ill-posedness of wave equations as
 evolution equations in spatial variables, an observation attributed to
-Hadamard (see \cite{CourHil:62}, Chapter 6, section 17). Some constraint on the acoustic field,
-beyond finite energy, is mandatory in any precise mathematical formulation the inverse problems
-\ref{eqn:esi} and \ref{eqn:esis}. In fact, the natural constraint in
-the ``crosswell'' geometry of this paper is that high-frequency energy
-travel {\em only} along rays crossing the
-surfaces $z=z_s, z=z_r$ transversally. That is, source functions on
-$z=z_s$ generate waves with energy traveling along rays leaving the surface at a non-zero angle, and
-energy arrives at the recording surface $z=z_r$ along rays making
-non-zero angle with it. I will call sources, sampled data, and
-acoustic fields with this property {\em
-  downgoing} (even though the concept also encompasses {\em upcoming}
-propagation). Note that the downgoing property restricts the behaviour of
-acoustic fields near the source and receiver surfaces - what the
-fields do elsewhere is their own business.
-
-Several works have explored the mathematics of the
-downgoing condition and its consequences in the context of the scalar second order wave equation, see for instance
+Hadamard (see \cite{CourHil:62}, Chapter 6, section 17). A number of
+authors have described precise forms of the ray conditions described
+above, and shown how these conditions lead to the desired behaviour of
+the modeling operator, that is, mapping of finite energy sources to
+finite energy data, or equivalent properties
 \cite{Payn:75,Symes:83,Lasi:86,LasLionsTrig:86,Lasi:87, BaoSy:91b}.
 Elaboration of these mathematical details is beyond the scope of this
-paper, which aims instead to explore the algorithmic
-consequences of the mathematical structure implied by the downgoing
-condition.
-
-The next section defines the modeling operator $PL[{\bf c}]$, its
-adjoint, and important specializations (pressure vs. normal velocity
-sources and data). The sections to follow define the
-source-to-pressure operator, construct an approximate inverse of the
-modeling operator by time reversal (as suggested by work in
-photoacoustic tomography), use the source-to-pressure operator to
-express the approximate inverse as the modeling operator adjoint in
-weighted norms (thus establishing that the modeling operator is {\em approximately
-  unitary} in the sense of these norms), explain how to use this
-construction to precondition Conjugate Gradient iteration, and
-organize the preconditioning computation so as to involve only one
-extra and relatively inexpensive
-wave propagation calculation. The penultimate section displays simple
-2D numerical examples of all of the key steps, culminating in a
-comparison of straight vs. preconditioned CG iteration. The paper ends
-with a brief discussion-and-conclusion section, reviewing what has
-been accomplished and listing a few of the many questions left open.
+paper, which aims instead to explore the algorithmic consequences of
+the mathematical structure implied by the nongrazing hypothesis.
+
+%Some
+%constraint on the acoustic field, beyond finite energy, is mandatory
+%in any precise mathematical formulation the inverse problems
+%\ref{eqn:esi} and \ref{eqn:esis}. I assume throughout that that
+%high-frequency energy contributing to the measured data
+%(output of the modeling operator) travels {\em only} along rays crossing the source
+%and receiver surfaces transversally, that is, making non-zero angles
+%with the tangent planes. I will call sources, sampled data, and
+%acoustic fields with this property {\em non-grazing}. Note that the
+%non-grazing property restricts the behaviour of acoustic fields near the
+%source and receiver surfaces - what the fields do elsewhere is their
+%own business.
+
+%A further restriction turns out to be necessary: {\em all} of the rays
+%associated with significan energy in the data must originate on
+%the source surface and have the {\em incoming} property (essentially
+%what is known in seismology as {\em downgoing}). This assumption uses
+%an orientation of the source surface to distinguish inward and outward
+%directions, and mandates that the rays in question all cross the
+%source surface and propagate inwards in forward time. By symmetry,
+%rays arriving at the receiver surface should propagate outward, with
+%proper choice of orientation. These assumptions are also local at the
+%source and receiver surfaces, and are often a natural consequence of
+%the acquisition and ray geometry, as is the case in the two examples
+%mentioned above.
 
@@
 
-Most of what follows is valid for any space dimension $n >0$. The
-coefficient vector $\bf{c}=(\kappa,\rho)$ is defined throughout space
-$\bR^n$, the state vector $\bu$ throughout space-time
-$\bR^{n+1}$. Whenever convenient for mathematical manipulations,
-$n=3$: for instance, I will write $\bx=(x,y,z)^T$ for the spatial
+Since all of the operators in the discussion that follows depend on
+the coefficient vector $\bf{c}$, I will suppress it from the notation,
+for example, $L=L[\bf{c}]$.
+
+Most of what follows is valid for any space dimension $n >0$. I will
+describe the theory for $n=3$, write $\bx=(x,y,z)^T$ for the spatial
 coordinate vector, and refer to the third (vertical) coordinate of
-particle velocity as $v_z$. Examples
-later in this paper will use $n=2$ for computational convenience.
-
-Since all of the operators in the discussion that follows depend on
-the coefficient vector 
-$\bf{c}$, I will suppress it from the notation, for example, $L=L[\bf{c}]$.
+particle velocity $\bv$ as $v_z$. For computational convenience, the
+examples are two-dimensional.
 
 The surface source extension replaces point sources on or near a
-surface in $\bR^3$ with source functions confined to the same
+surface in $\bR^3$ with source functions confined to the 
 surface. The simplest example of this extended geometry specifies a
@@
 simplicity, throughout this paper I will assume that the receivers are
-located on another spatial hyperplane $\{(x,y,z,t): z=z_r\}$ at
-receiver depth $z_r>z_s$. The constructions to follow involve interchange
-of the roles of $z_s$ and $z_r$ (that is, locating sources on $z=z_r$
-and receivers at $z=z_s$), so rather than the sampling operator $P$ of
-the introduction, I will denote by $P_s,P_r$ the sampling 
-operators on $z=z_s$, $z=z_r$ respectively. In practice, sampling
-occurs at a discrete array of points (trace locations) on these
-surfaces, and over a zone of finite extent. In this theoretical
-discussion, I will neglect both finite sample rate and extent, and
-regard the data, for example $P_rp^+$, as continuously sampled and
-extending over the entire plane $z=z_r$.
-
-As explained in the Introduction, the downgoing constraint on the square-integrable
-source functions $h_s, f_s$ is
-essential both for finite energy solutions of the system
-\ref{eqn:awepm} to exist, and for these solutions to have well-defined
-traces on the receiver surface $z=z_r$. This constraint will be
-assumed throughout, often tacitly.
-For downgoing solutions of system \ref{eqn:awepm}, the key
-components ($p^{\pm}$ and $v^{\pm}_z$) are continuous functions of $z$
-in the open slab $z_s<z<z_r$ with well-defined limits at the boundary
-planes, but may be discontinuous at the source plane
-$z=z_s$. Similarly, the roles of $z_s$ and $z_r$ will be interchanged
-in some of the constructions to come, and the corresponding solutions
-may be discontinuous at $z=z_r$. Accordingly, interpret $P_s$, $P_r$
-as the limit from right and left respectively: for $u=p^{\pm}$ or
-$v^{\pm}_z$,
-\begin{eqnarray}
-  \label{eqn:defsamp}
-  P_su(x,y,t) &=& \lim_{z \rightarrow z_s^+} u(x,y,z,t),\nonumber \\
-  P_ru(x,y,t) &=& \lim_{z \rightarrow z_r^-} u(x,y,z,t).                  
-\end{eqnarray}
+located on another spatial plane $\{(x,y,z,t): z=z_r\}$ at receiver
+depth $z_r$. The constructions to follow involve interchange of the
+roles of $z_s$ and $z_r$ (that is, locating sources on $z=z_r$ and
+receivers at $z=z_s$), so both require sampling operators: $P_s,P_r$
+are the sampling operators on $z=z_s$, $z=z_r$ respectively. In
+practice, sampling necessarily occurs at a discrete array of points
+(trace locations), and over a zone of finite extent. In this
+theoretical discussion, I will neglect the finite sample rate, and
+regard the data, for example $P_rp^+$, as continuously sampled. The
+output samples are necessarily muted, that is, non-zero only over a space-time domain
+of finite extent. This mute, and any tapering applied to the data
+traces, are regarded as part of the sampling operators $P_s,P_r$.
 
@@
 \end{equation}
-The subscript signifies that sources are located on $z=z_s$, the
+The subscript signifies that sources are located on $z=z_s$,
 receivers on $z=z_r$. It is necessary to include this information in
 the notation, as versions of ${\cal S}^{\pm}$ with sources and receivers in
-several locations will be needed in the discussion below.
-
-\noindent {\bf Remark:} To connect with the formulation presented in
-the introduction, note that for continuous $u$,
-$P_su(x,y,t)=u(x,y,z_s,t)$, and therefore the adjoint of $P_s$ (in the
-sense of distributions) is $P_s^Th(x,y,z,t) =
-h(x,y,t)\delta(z-z_s)$. Write ${\cal P}_s = \mbox{diag }(P_s,P_s)$ and
-similarly for ${\cal P}_r$. Then
-\[
-  {\cal S}^{+}_{z_s,z_r} = {\cal P}_r L^{-1}({\cal P}_s)^T,
-\]
-in which $L^{-1}$ is interpreted in the causal sense, and similarly
-for ${\cal S}^{-}$. Sources confined to $z=z_s$ are precisely those
-functions (distributions, really) output by ${\cal P}_s^T$, so the
-problem statements \ref{eqn:esi} and \ref{eqn:esis} can be rewritten
-in terms of ${\cal S}^+_{z_s,z_r}$, with $P$ identified with ${\cal P}_r$.
-
-${\cal S}^{\pm}$ is not stably invertible: its columns are
-approximately linearly dependent, as will be verified below. The
-diagonal components of ${\cal S}^{\pm}$ thus carry essentially all of
-its information, and it is in terms of these that a sensible inverse problem
-is defined.
+several locations will be needed later in the discussion.
 
@@
 
-It follows from the adjoint state method (see Appendix A for details) that
-\begin{equation}
-  \label{eqn:sadj1}
-  ({\cal S}^{\pm}_{z_s, z_r})^T = -{\cal S}^{\mp}_{z_r,z_s}
-\end{equation}
-
-Define $R$ to be the {\em time-reversal operator} on functions of
-space-time, $Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em
-  acoustic field time-reversal operator}
-\begin{equation}
-  \label{eqn:trdef}
-  {\cal R} \left(
-    \begin{array}{c}
-      p\\
-      \bv
-    \end{array}
-  \right) =
-  \left(
-    \begin{array}{c}
-      Rp\\
-      -R\bv
-    \end{array}
-  \right)
-\end{equation}
-Then 
-\begin{equation}
-  \label{eqn:trsadj}
-  {\cal R}{\cal S}^{\mp} = -{\cal S}^{\pm}_{z_r,z_s}{\cal R}
-\end{equation}
-Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj1} and \ref{eqn:trsadj} imply that
-\begin{equation} 
-  \label{eqn:trtr}
- ({\cal S}^{\pm}_{z_s,z_r})^T = {\cal R}{\cal S}_{z_r,z_s}^{\pm}{\cal R}=
- -{\cal S}^{\mp}_{z_r,z_s}.
-\end{equation}
-The relation \ref{eqn:trtr} implies that
-\begin{eqnarray}
-  (S^{\pm}_{z_s,z_r})^T &=& -S^{\mp}_{z_r,z_s} \nonumber\\
-                        &=& R S^{\pm}_{z_r,z_s}R, \nonumber\\
-    (V^{\pm}_{z_s,z_r})^T &=& -V^{\mp}_{z_r,z_s} \nonumber\\
-                        &=& R V^{\pm}_{z_r,z_s}R.
-                            \label{eqn:trtrcomp}
-\end{eqnarray}
-
-\section{Pressure-to-Source}
-
-Since the system \ref{eqn:awepm} has a unique solution by standard
-theory \cite[]{Lax:PDENotes}, the source vector field $(h_s,f_s)$
-determines the acoustic field $(p^{\pm},\bv^{\pm})$ in space time, and
-in particular the limits from the right at $z=z_s$, $P_sp^{\pm}$ and
-$P_sv_z^{\pm}$. This relation is not invertible: it is not possible to
-prescribe both pressure and normal velocity on a surface such as
-$z=z_s$. So the columns of the matrix operator
-${\cal S}^{\pm}_{z_s,z_r}$ must satisfy a linear relation. In this
-section I will explain this relation; it involves the {\em
-  pressure-to-source} map. This operator also turns out to be the
-principal component of a preconditioning strategy for iterative
-solution of the optimization problem \ref{eqn:esis}, so I will devote
-some effort to its proper definition. It is closely related to the
-Dirichlet-to-Neumann operator mentioned in the introduction.
-
-While it is not possible to prescribe both pressure and velocity on
-$z=z_s$ in solutions of \ref{eqn:awepm}, it is possible to
-prescribe pressure only, for instance: if the function $\phi$ on
-the surface $z=z_s$ satisfies suitable conditions, for
-example the downgoing constraint mentioned earlier, a unique solution
-exists for the acoustic system in both half-spaces $\pm z > z_s$:
-\begin{eqnarray}
-\label{eqn:awe0}
-  \frac{1}{\kappa}\frac{\partial p_{\pm}}{\partial t} & = & - \nabla \cdot \bv_{\pm}, \nonumber \\
-  \rho\frac{\partial \bv_{\pm}}{\partial t} & = & - \nabla
-                                                    p_{\pm}, \nonumber \\
-  p_{\pm} & =& 0,  \mbox{ for } t \ll 0, \nonumber\\ 
-  \bv_{\pm} & = & 0 \mbox{ for } t \ll 0, \nonumber\\
-  \lim_{z \rightarrow z_s^{\pm}}p_{\pm}(x,y,t,z)& =& \phi(x,y,t).
-\end{eqnarray}
-Note that the subscript $\pm$ here refers to the sign of $z-z_s$, as opposed
-to the superscript ${\pm}$, which refers to the sign of $t$ throughout
-this paper.
-
-From the boundary condition (last equation in \ref{eqn:awe0}), one
-sees that the pressures $p_{\pm}$ in the two half-spaces have the same
-limit at the boundary $z=z_s$. Stick the two half-space
-solutions together to form an acoustic field $(p^+,\bv^+)$ in all of
-space-time, that is,
-\begin{equation}
-  \label{eqn:awealt}
-  p^+(x,y,z,t) =
-  \left\{
-    \begin{array}{c}
-      p_+(x,y,z,t) \mbox{ if } z>0,\\
-      p_-(x,y,z,t) \mbox{ if } z<0,
-    \end{array}
-  \right.
-\end{equation}
-and a similar definition for $\bv^+$. Then $p^+$ is continuous across
-$z=z_s$, and the boundary condition in system \ref{eqn:awe0} may be
-written as $P_sp^+=\phi$.
-
-The same construction can be carried out in the anti-causal sense,
-with anti-causal half-space solutions glued together to form a
-full-space distribution solution $(p^-,\bv^-)$, with the property that
-$p^-$ is continuous across $z=z_s$ and $P_sp^-=\phi$.
-
-The reader may object that the notation $(p^\pm,\bv^\pm)$ is already in
-use, for the solution of \ref{eqn:awepm}. This objection is
-valid. However, {\em in the sense
-  of distributions}, $(p^{\pm},\bv^{\pm})$ as defined in display
-\ref{eqn:awealt}, is {\em exactly} the causal solution of \ref{eqn:awepm}
-for the choice $h_s = -[v^{\pm}_{z}]|_{z=z_s}, f_s=0$, as follows from a
-simple integration-by-parts calculation. So the notation is consistent!
-
-The negative jump $-[v^{\pm}_{z}]|_{z=z_s}$ is thus a function of $\phi$. Define
-the {\em pressure-to-source} operator $\Lambda^{\pm}_{z_s}$ by
-\begin{equation}
-  \label{eqn:deflam}
-  \Lambda^{\pm}_{z_s}\phi = -[v^{\pm}_{z}]|_{z=z_s}
-\end{equation}
-The conclusion: if $h_s = \Lambda^{\pm}_{z_s}\phi$ and $f_s=0$ in the
-system \ref{eqn:awepm}, then $\phi=P_sp^{\pm}$.
-
-Otherwise put, $S^{\pm}_{z_s,z_s}\Lambda^{\pm}_{z_s} \phi = \phi$, so
-$\Lambda^{\pm}_{z_s}$ is inverse to $S^{\pm}_{z_s,z_s}$. The relation
-\ref{eqn:trtrcomp} implies in turn that
-\begin{equation}
-  \label{eqn:lamadj}
-  (\Lambda^{\pm}_{z_s})^T = - \Lambda^{\mp}_{z_s}
-\end{equation}
-
-There is also a {\em velocity-to-source} operator. For the solution
-$(p^{\pm},\bv^{\pm})$ of system \ref{eqn:awepm} with $h_s=0$, the
-normal component of velocity, $v^{\pm}_z$, is continuous across
-$z=z_s$, and the velocity source (vertical load)
-$f_s=-[p^{\pm}]_{z=z_s}$. I will not name the velocity-to-source
-operator, as it does not appear explicitly in the developments to
-follow. As will be seen, it is essentially the inverse of the
-pressure-to-source operator.
-
-The quadratic form defined by $\Lambda^{\pm}_{z_s}$ has fundamental
-physical significance. Define the total acoustic energy $E^{\pm}(t)$ of the
-field $(p^{\pm},\bv^{\pm})$, at time $t$ by
-\begin{equation}
-  \label{eqn:defae0}
-  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)(\bx,t).
-\end{equation}
-Then
-\begin{equation}
-  \label{eqn:elim}
-  \pm \lim_{\pm t \rightarrow \infty} E^{\pm}(t) =  \langle P_sp^{\pm},
-  (\Lambda^{\pm}_{z_s} P_sp^{\pm}) \rangle_{L^2(z=z_s)}.
-\end{equation}
-That is, the value of the quadratic form defined by
-$\Lambda^{\pm}_{z_s}$, evaluated at the pressure trace on $z=z_s$,
-gives the total energy transferred from the source to the
-acoustic field over time. Since $E$ is itself a positive definite
-quadratic form in the acoustic field, it follows that $\pm
-\Lambda^{\pm}_{z_s}$ is positive semi-definite.
-
-While $\Lambda^{\pm}_{z_s}$ is positive semi-definite, it is not
-symmetric. However, it is {\em approximately symmetric} in the
-high-frequency sense. This fact follows from a
-geometric optics analysis of the half-space solution. This leads to
-the identification of $\Lambda^{\pm}_{z_s}$ as a {\em
-  pseudodifferential operator} of order zero on $z=z_s$, with principal symbol
-\begin{equation}
-  \label{eqn:lamsym}
-  \sigma_0(\Lambda^{\pm}_{z_s}) = \pm 2(\kappa(\bx)
-  \rho(\bx))^{1/2}\left(1-\frac{\kappa(\bx)(\xi^2+\eta^2)}{\rho(\bx)\omega^2}\right)^{-1/2}.
-\end{equation}
-Here $\xi$, $\eta$, and $\omega$ are the dual Fourier variables to
-$x$, $y$, and $t$ respectively. The downgoing assumptions means that
-for local planewave components of $P_sp$, the quantity inside the
-square root is positive. Thus $\Lambda^{\pm}_{z_s}$ has real principal
-symbol (in fact, the entire symbol is real) hence defines an
-asymptotically symmetric operator:
-\begin{equation}
-  \label{eqn:lamappsim}
-  (\Lambda^{\pm}_{z_s})^T \approx \Lambda^{\pm}_{z_s}.
-\end{equation}
-(For more on this, see \cite{StefUhl:05}.)
-The analysis also reveals that the solution components not continuous
-at $z=z_s$ are odd there:
-\begin{equation}
-  \label{eqn:odd1}
-  \lim_{z\rightarrow z_s^+} v^{\pm}_{z} \approx - \lim_{z\rightarrow z_s^-}
-  v^{\pm}_{z}
-\end{equation}
-for the solution of \ref{eqn:awepm} with $f_s=0$.
-Similarly, 
-\begin{equation}
-  \label{eqn:odd2}
-  \lim_{z\rightarrow z_s^+} p^{\pm}\approx - \lim_{z\rightarrow z_s^-}
-  p^{\pm}
-\end{equation}
-for the solution of \ref{eqn:awepm} with $h_s=0$. Here ``$\approx$''
-means in the sense of high frequency asymptotics, that is, that the
-difference between the two sides is relatively smooth, hence small if
-the data is highly oscillatory. Therefore if $f_s=0$ in system \ref{eqn:awepm},
-\begin{equation}
-  h_s = \Lambda^{\pm}_{z_s}P_sp^{\pm} = -[v^{\pm}_{z}]|_{z=z_s} \approx -2
-  P_sv^{\pm}_{z}
-  \label{eqn:tracejump10}
-\end{equation}
-Similarly, if $h_s=0$ in system \ref{eqn:awepm}, then
-\begin{equation}
-  \label{eqn:tracejump20}
-  f_s = -[p^{\pm}]|_{z=z_s} \approx -2 P_s p^{\pm}.
-\end{equation}
-Thus $f_s$ determines approximately the boundary value of $p^{\pm}$,
-as a solution of the acoustic wave system in the half-space
-$z>z_s$. However, as repeated in equation \ref{eqn:tracejump10}, a
-solution with this boundary value is also the restriction to $z>z_s$
-of a solution to \ref{eqn:awepm} with $f_s=0$ and $h_s=
-\Lambda^{\pm}_{z_s}P_sp^{\pm}$. Therefore if
-\begin{equation}
-  \label{eqn:hfcondn}
-  h_s =-\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,
-\end{equation}
-then the pressure boundary value $P_sp^{\pm}$ is the
-same for the solutions of \ref{eqn:awepm} for source vectors $(h_s,0)$
-and $(0,f_s)$. Since the pressure boundary values are the same, the solutions
-in $z>z_s$ are the same. In particular, since $z_r>z_s$ and ${\cal
-  S}^{\pm}_{z_s,z_r}(h_s,f_s)^T = (P_rp^{\pm},P_rv^{\pm}_z)^T$, it follows
-that
-\begin{equation}
-  \label{eqn:snull}
-  {\cal S}^{\pm}_{z_s,z_r}\left(\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,f_s\right)^T \approx 0.
-\end{equation}
-
-Equation \ref{eqn:snull} states the relation between the columns of $
-{\cal S}^{\pm}_{z_s,z_r}$ mentioned in the introduction to this
-section.
+The two examples mentioned in the introduction illustrate the setting
+just described. The first example (Figure \ref{fig:olensrays0}) embeds
+an acoustic (low-velocity) lens between a source surface at $z_s=3000$
+(orange horizontal line) 
+m and a receiver surface at $z_r=1000$ m (yellow horizontal line).  The data to be used in this
+example results from a point source at $z_s=3500, x_s=3500$ m (orange
+star). I have overlain the rays connecting this source point with the
+data portion of the receiver surface: evidently all of the rays
+involved cross the source and receiver surfaces transversally. Also,
+regarding ``inwards'' as being ``upwards'' at the source surface,
+``downwards'' at the receiver surface, it is clear that the data is
+incoming at the source surface, outgoing at the receiver
+surface.
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{olensrays0.pdf}
+  \caption{Bulk modulus with acoustic lens. Color scale unit is
+    GPa. Orange horizontal line is source surface at depth $z_s= 3000$
+    m, yellow horizontal
+    line is receiver surface at depth $z_r=1000$ m. Point source location for data
+    generation at $(3500, 3500)$ m indicated with star. Overlain with rays from point
+    source to receiver surface.}
+  \label{fig:olensrays0}
+\end{figure}
+
+%\ref{fig:drecplh0}, \ref{fig:dfwdplh0}
+Figure \ref{fig:drecplh0} shows the gather $P_rp_{\rm pt}^{+}$, where
+$p_{\rm pt}^{+}$ is causal field generated by the point source at
+$(3500,3500)$ m with a trapezoidal bandpass filter wavelet having
+significant energy between 1 and 12.5 Hz. The mute embedded in $P_r$
+limits trace positions to 2000 m $ \le x_r \le $ 6000 m, and times to
+1.2 s $ \le t \le $ 3 s.
+
+Figure \ref{fig:dsrcphh0} displays an extended (pressure) source $h_s$, in the form of
+traces in the spatial range 2000 m $ \le x_r \le $ 6000 m and time
+range 0 s $\le t \le $ 2 s. The modeling operator output $S_{z_s,z_r}^+h_s$,
+obtained from the causal solution $(p^+, \bv^+)$ of \ref{eqn:awepm}
+with this choice of $h_s$ and $f_s=0,$ by application of the sampling
+operator $P_r$ to $p^+$, are shown in Figure \ref{fig:dfwdplh0}.
+
+\noindent {\bf Remark:} The close resemblance between Figures
+\ref{fig:drecplh0} and \ref{fig:dfwdplh0} is not accidental  - the
+extended source $h_ s$ shown in Figure \ref{fig:dsrcphh0} is constructed
+so that $S^++_{z_s,z_r}h_s$ (Figure \ref{fig:dfwdplh0}) closely
+approximates the point source gather $P_r p^+_{\rm pt}$ (Figure
+\ref{fig:drecplh0}). This construction will be explained in the next section.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:drecplh0}]{\includegraphics[width=0.32\textwidth]{drecplh0.pdf}}
+  \subfloat[\label{fig:dsrcphh0}]{\includegraphics[width=0.32\textwidth]{dsrcphh0.pdf}}
+  \subfloat[\label{fig:dfwdplh0}]{\includegraphics[width=0.32\textwidth]{dfwdplh0.pdf}}
+  \caption{Data generated using configuration in Figure
+    \ref{fig:olensrays0}. (a) traces from point source at $(3500,3500)$
+    with $[1, 2, 7.5, 12.5]$ Hz zero phase trapezoidal bandpass
+    filter wavelet, delayed by $0.5$ s. (b) Equivalent extended source on
+    the source surface at depth $z_s=3000$ m. (c) Gather generated by
+    equivalent source shown in (b). Color scale is same for (a) and (c).}
+\end{figure}
+
+The second example is intended as a cartoon of long-offset node
+acquisition. It features a depth-dependent increasting
+velocity. Figure \ref{fig:ooplrays0} shows bulk modulus field (once again, the
+density is constant and $= 1$ g/cm$^3$). The source and receiver
+surfaces are horizontal lines, as in the first example, at depths
+$z_s=500$ m and $z_r= 100$ m respectively. A point source used to
+generate a data traces is positioned at $z_s=500, x_s=10000$ m. The
+source wavelet is the same bandpass filter as in
+the previous example.
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{ooplrays0.pdf}
+  \caption{Bulk modulus generating diving waves. Color scale unit is
+    GPa. Orange horizontal line is source surface at depth $z_s= 500$
+    m, yellow horizontal
+    line is receiver surface at depth $z_r=100$ m. Point source location for data
+    generation at $(500, 10000)$ m indicated with star. Overlain with rays from point
+    source to receiver surface.}
+  \label{fig:ooplrays0}
+\end{figure}
+
+A diving wave arrival is clearly visible in Figure
+\ref{fig:ndwraw0}, which displays data traces over the position range $14000 \le
+x_r \le 20000$ m. The ray field connecting the point source to the
+receiver surface is plotted with the diving rays in blue, the rest in
+red. Note that at the source, the diving rays are incoming if the
+orientation is chosen with inwards = down, in contrast to the previous
+example, while they are outgoing at
+the receiver surface if it is oriented so that inwards = down also.
+
+This geometry permits inversion of the diving wave data alone, which
+is isolated via a mute, depicted in (Figure \ref{fig:ndwmute0}. The
+mute includes time truncation at 6 s, and offsets between -10000 and
+10000 m. The isolated
+diving wave (window between offsets 4000 and 10000) is shown in
+Figure \ref{fig:ndwdata0}.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:ndwraw0}]{\includegraphics[width=0.32\textwidth]{ndwraw0.pdf}}
+  \subfloat[\label{fig:ndwmute0}]{\includegraphics[width=0.32\textwidth]{ndwmute0.pdf}}
+  \subfloat[\label{fig:ndwdata0}]{\includegraphics[width=0.32\textwidth]{ndwdata0.pdf}}
+  \caption{Data generated using configuration in Figure
+    \ref{fig:ooplrays0}. (a) traces from point source at $(500,10000)$ m
+    with $[1, 2, 7.5, 12.5]$ Hz zero phase trapezoidal bandpass
+    filter wavelet, delayed by $0.4$ s. (b) Mute function,
+    incorporated in the sampling operator $P_r$. (c) Diving wave
+    gather. Color scale is same for (a) and (c).}
+\end{figure}
+%As explained in the Introduction, the downgoing constraint on the square-integrable
+%source functions $h_s, f_s$ is
+%essential both for finite energy solutions of the system
+%\ref{eqn:awepm} to exist, and for these solutions to have well-defined
+%traces on the receiver surface $z=z_r$. This constraint will be
+%assumed throughout, often tacitly.
+%For downgoing solutions of system \ref{eqn:awepm}, the key
+%components ($p^{\pm}$ and $v^{\pm}_z$) are continuous functions of $z$
+%in the open slab $z_s<z<z_r$ with well-defined limits at the boundary
+%planes, but may be discontinuous at the source plane
+%$z=z_s$. Similarly, the roles of $z_s$ and $z_r$ will be interchanged
+%in some of the constructions to come, and the corresponding solutions
+%may be discontinuous at $z=z_r$. Accordingly, interpret $P_s$, $P_r$
+%as the limit from right and left respectively: for $u=p^{\pm}$ or
+%$v^{\pm}_z$,
+%\begin{eqnarray}
+ % \label{eqn:defsamp}
+ % P_su(x,y,t) &=& \lim_{z \rightarrow z_s^+} u(x,y,z,t),\nonumber \\
+ % P_ru(x,y,t) &=& \lim_{z \rightarrow z_r^-} u(x,y,z,t).                  
+%\end{eqnarray}
+
+%\noindent {\bf Remark:} To connect with the formulation presented in
+%the introduction, note that for continuous $u$,
+%$P_su(x,y,t)=u(x,y,z_s,t)$, and therefore the adjoint of $P_s$ (in the
+%sense of distributions) is $P_s^Th(x,y,z,t) =
+%h(x,y,t)\delta(z-z_s)$. Write ${\cal P}_s = \mbox{diag }(P_s,P_s)$ and
+%similarly for ${\cal P}_r$. Then
+%\[
+ % {\cal S}^{+}_{z_s,z_r} = {\cal P}_r L^{-1}({\cal P}_s)^T,
+%\]
+%in which $L^{-1}$ is interpreted in the causal sense, and similarly
+%for ${\cal S}^{-}$. Sources confined to $z=z_s$ are precisely those
+%functions (distributions, really) output by ${\cal P}_s^T$, so the
+%problem statements \ref{eqn:esi} and \ref{eqn:esis} can be rewritten
+%in terms of ${\cal S}^+_{z_s,z_r}$, with $P$ identified with ${\cal P}_r$.
+
+%${\cal S}^{\pm}$ is not stably invertible: its columns are
+%approximately linearly dependent, as will be verified below. The
+%diagonal components of ${\cal S}^{\pm}$ thus carry essentially all of
+%its information, and it is in terms of these that a sensible inverse problem
+%is defined.
 
@@
 
+It follows from the adjoint state method (see Appendix A for details) that
+\begin{equation}
+  \label{eqn:sadj1}
+  ({\cal S}^{\pm}_{z_s, z_r})^T = -{\cal S}^{\mp}_{z_r,z_s}
+\end{equation}
+
+Define $R$ to be the {\em time-reversal operator} on functions of
+space-time, $Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em
+  acoustic field time-reversal operator}
+\begin{equation}
+  \label{eqn:trdef}
+  {\cal R} \left(
+    \begin{array}{c}
+      p\\
+      \bv
+    \end{array}
+  \right) =
+  \left(
+    \begin{array}{c}
+      Rp\\
+      -R\bv
+    \end{array}
+  \right)
+\end{equation}
+Then 
+\begin{equation}
+  \label{eqn:trsadj}
+  {\cal R}{\cal S}^{\mp} = -{\cal S}^{\pm}_{z_r,z_s}{\cal R}
+\end{equation}
+Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj1} and \ref{eqn:trsadj} imply that
+\begin{equation} 
+  \label{eqn:trtr}
+ ({\cal S}^{\pm}_{z_s,z_r})^T = {\cal R}{\cal S}_{z_r,z_s}^{\pm}{\cal R}=
+ -{\cal S}^{\mp}_{z_r,z_s}.
+\end{equation}
+The relation \ref{eqn:trtr} implies that
+\begin{eqnarray}
+  (S^{\pm}_{z_s,z_r})^T &=& -S^{\mp}_{z_r,z_s} \nonumber\\
+                        &=& R S^{\pm}_{z_r,z_s}R, \nonumber\\
+    (V^{\pm}_{z_s,z_r})^T &=& -V^{\mp}_{z_r,z_s} \nonumber\\
+                        &=& R V^{\pm}_{z_r,z_s}R.
+                            \label{eqn:trtrcomp}
+\end{eqnarray}
+
+\section{Pressure-to-Source}
+
+Since the system \ref{eqn:awepm} has a unique solution by standard
+theory \cite[]{Lax:PDENotes}, the source vector field $(h_s,f_s)$
+determines the acoustic field $(p^{\pm},\bv^{\pm})$ in space time, and
+in particular the limits from the right at $z=z_s$, $P_sp^{\pm}$ and
+$P_sv_z^{\pm}$. This relation is not invertible: it is not possible to
+prescribe both pressure and normal velocity on a surface such as
+$z=z_s$. So the columns of the matrix operator
+${\cal S}^{\pm}_{z_s,z_r}$ must satisfy a linear relation. In this
+section I will explain this relation; it involves the {\em
+  pressure-to-source} map. This operator also turns out to be the
+principal component of a preconditioning strategy for iterative
+solution of the optimization problem \ref{eqn:esis}, so I will devote
+some effort to its proper definition. It is closely related to the
+Dirichlet-to-Neumann operator mentioned in the introduction.
+
+While it is not possible to prescribe both pressure and velocity on
+$z=z_s$ in solutions of \ref{eqn:awepm}, it is possible to
+prescribe pressure only, for instance: if the function $\phi$ on
+the surface $z=z_s$ satisfies suitable conditions, for
+example the downgoing constraint mentioned earlier, a unique solution
+exists for the acoustic system in both half-spaces $\pm z > z_s$:
+\begin{eqnarray}
+\label{eqn:awe0}
+  \frac{1}{\kappa}\frac{\partial p_{\pm}}{\partial t} & = & - \nabla \cdot \bv_{\pm}, \nonumber \\
+  \rho\frac{\partial \bv_{\pm}}{\partial t} & = & - \nabla
+                                                    p_{\pm}, \nonumber \\
+  p_{\pm} & =& 0,  \mbox{ for } t \ll 0, \nonumber\\ 
+  \bv_{\pm} & = & 0 \mbox{ for } t \ll 0, \nonumber\\
+  \lim_{z \rightarrow z_s^{\pm}}p_{\pm}(x,y,t,z)& =& \phi(x,y,t).
+\end{eqnarray}
+Note that the subscript $\pm$ here refers to the sign of $z-z_s$, as opposed
+to the superscript ${\pm}$, which refers to the sign of $t$ throughout
+this paper.
+
+From the boundary condition (last equation in \ref{eqn:awe0}), one
+sees that the pressures $p_{\pm}$ in the two half-spaces have the same
+limit at the boundary $z=z_s$. Stick the two half-space
+solutions together to form an acoustic field $(p^+,\bv^+)$ in all of
+space-time, that is,
+\begin{equation}
+  \label{eqn:awealt}
+  p^+(x,y,z,t) =
+  \left\{
+    \begin{array}{c}
+      p_+(x,y,z,t) \mbox{ if } z>0,\\
+      p_-(x,y,z,t) \mbox{ if } z<0,
+    \end{array}
+  \right.
+\end{equation}
+and a similar definition for $\bv^+$. Then $p^+$ is continuous across
+$z=z_s$, and the boundary condition in system \ref{eqn:awe0} may be
+written as $P_sp^+=\phi$.
+
+The same construction can be carried out in the anti-causal sense,
+with anti-causal half-space solutions glued together to form a
+full-space distribution solution $(p^-,\bv^-)$, with the property that
+$p^-$ is continuous across $z=z_s$ and $P_sp^-=\phi$.
+
+The reader may object that the notation $(p^\pm,\bv^\pm)$ is already in
+use, for the solution of \ref{eqn:awepm}. This objection is
+valid. However, {\em in the sense
+  of distributions}, $(p^{\pm},\bv^{\pm})$ as defined in display
+\ref{eqn:awealt}, is {\em exactly} the causal solution of \ref{eqn:awepm}
+for the choice $h_s = -[v^{\pm}_{z}]|_{z=z_s}, f_s=0$, as follows from a
+simple integration-by-parts calculation. So the notation is consistent!
+
+The negative jump $-[v^{\pm}_{z}]|_{z=z_s}$ is thus a function of $\phi$. Define
+the {\em pressure-to-source} operator $\Lambda^{\pm}_{z_s}$ by
+\begin{equation}
+  \label{eqn:deflam}
+  \Lambda^{\pm}_{z_s}\phi = -[v^{\pm}_{z}]|_{z=z_s}
+\end{equation}
+The conclusion: if $h_s = \Lambda^{\pm}_{z_s}\phi$ and $f_s=0$ in the
+system \ref{eqn:awepm}, then $\phi=P_sp^{\pm}$.
+
+Otherwise put, $S^{\pm}_{z_s,z_s}\Lambda^{\pm}_{z_s} \phi = \phi$, so
+$\Lambda^{\pm}_{z_s}$ is inverse to $S^{\pm}_{z_s,z_s}$. The relation
+\ref{eqn:trtrcomp} implies in turn that
+\begin{equation}
+  \label{eqn:lamadj}
+  (\Lambda^{\pm}_{z_s})^T = - \Lambda^{\mp}_{z_s}
+\end{equation}
+
+There is also a {\em velocity-to-source} operator. For the solution
+$(p^{\pm},\bv^{\pm})$ of system \ref{eqn:awepm} with $h_s=0$, the
+normal component of velocity, $v^{\pm}_z$, is continuous across
+$z=z_s$, and the velocity source (vertical load)
+$f_s=-[p^{\pm}]_{z=z_s}$. I will not name the velocity-to-source
+operator, as it does not appear explicitly in the developments to
+follow. As will be seen, it is essentially the inverse of the
+pressure-to-source operator.
+
+The quadratic form defined by $\Lambda^{\pm}_{z_s}$ has fundamental
+physical significance. Define the total acoustic energy $E^{\pm}(t)$ of the
+field $(p^{\pm},\bv^{\pm})$, at time $t$ by
+\begin{equation}
+  \label{eqn:defae0}
+  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)(\bx,t).
+\end{equation}
+Then
+\begin{equation}
+  \label{eqn:elim}
+  \pm \lim_{\pm t \rightarrow \infty} E^{\pm}(t) =  \langle P_sp^{\pm},
+  (\Lambda^{\pm}_{z_s} P_sp^{\pm}) \rangle_{L^2(z=z_s)}.
+\end{equation}
+That is, the value of the quadratic form defined by
+$\Lambda^{\pm}_{z_s}$, evaluated at the pressure trace on $z=z_s$,
+gives the total energy transferred from the source to the
+acoustic field over time. Since $E$ is itself a positive definite
+quadratic form in the acoustic field, it follows that $\pm
+\Lambda^{\pm}_{z_s}$ is positive semi-definite.
+
+While $\Lambda^{\pm}_{z_s}$ is positive semi-definite, it is not
+symmetric. However, it is {\em approximately symmetric} in the
+high-frequency sense. This fact follows from a
+geometric optics analysis of the half-space solution. This leads to
+the identification of $\Lambda^{\pm}_{z_s}$ as a {\em
+  pseudodifferential operator} of order zero on $z=z_s$, with principal symbol
+\begin{equation}
+  \label{eqn:lamsym}
+  \sigma_0(\Lambda^{\pm}_{z_s}) = \pm 2(\kappa(\bx)
+  \rho(\bx))^{1/2}\left(1-\frac{\kappa(\bx)(\xi^2+\eta^2)}{\rho(\bx)\omega^2}\right)^{-1/2}.
+\end{equation}
+Here $\xi$, $\eta$, and $\omega$ are the dual Fourier variables to
+$x$, $y$, and $t$ respectively. The downgoing assumptions means that
+for local planewave components of $P_sp$, the quantity inside the
+square root is positive. Thus $\Lambda^{\pm}_{z_s}$ has real principal
+symbol (in fact, the entire symbol is real) hence defines an
+asymptotically symmetric operator:
+\begin{equation}
+  \label{eqn:lamappsim}
+  (\Lambda^{\pm}_{z_s})^T \approx \Lambda^{\pm}_{z_s}.
+\end{equation}
+(For more on this, see \cite{StefUhl:05}.)
+The analysis also reveals that the solution components not continuous
+at $z=z_s$ are odd there:
+\begin{equation}
+  \label{eqn:odd1}
+  \lim_{z\rightarrow z_s^+} v^{\pm}_{z} \approx - \lim_{z\rightarrow z_s^-}
+  v^{\pm}_{z}
+\end{equation}
+for the solution of \ref{eqn:awepm} with $f_s=0$.
+Similarly, 
+\begin{equation}
+  \label{eqn:odd2}
+  \lim_{z\rightarrow z_s^+} p^{\pm}\approx - \lim_{z\rightarrow z_s^-}
+  p^{\pm}
+\end{equation}
+for the solution of \ref{eqn:awepm} with $h_s=0$. Here ``$\approx$''
+means in the sense of high frequency asymptotics, that is, that the
+difference between the two sides is relatively smooth, hence small if
+the data is highly oscillatory. Therefore if $f_s=0$ in system \ref{eqn:awepm},
+\begin{equation}
+  h_s = \Lambda^{\pm}_{z_s}P_sp^{\pm} = -[v^{\pm}_{z}]|_{z=z_s} \approx -2
+  P_sv^{\pm}_{z}
+  \label{eqn:tracejump10}
+\end{equation}
+Similarly, if $h_s=0$ in system \ref{eqn:awepm}, then
+\begin{equation}
+  \label{eqn:tracejump20}
+  f_s = -[p^{\pm}]|_{z=z_s} \approx -2 P_s p^{\pm}.
+\end{equation}
+Thus $f_s$ determines approximately the boundary value of $p^{\pm}$,
+as a solution of the acoustic wave system in the half-space
+$z>z_s$. However, as repeated in equation \ref{eqn:tracejump10}, a
+solution with this boundary value is also the restriction to $z>z_s$
+of a solution to \ref{eqn:awepm} with $f_s=0$ and $h_s=
+\Lambda^{\pm}_{z_s}P_sp^{\pm}$. Therefore if
+\begin{equation}
+  \label{eqn:hfcondn}
+  h_s =-\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,
+\end{equation}
+then the pressure boundary value $P_sp^{\pm}$ is the
+same for the solutions of \ref{eqn:awepm} for source vectors $(h_s,0)$
+and $(0,f_s)$. Since the pressure boundary values are the same, the solutions
+in $z>z_s$ are the same. In particular, since $z_r>z_s$ and ${\cal
+  S}^{\pm}_{z_s,z_r}(h_s,f_s)^T = (P_rp^{\pm},P_rv^{\pm}_z)^T$, it follows
+that
+\begin{equation}
+  \label{eqn:snull}
+  {\cal S}^{\pm}_{z_s,z_r}\left(\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,f_s\right)^T \approx 0.
+\end{equation}
+
+Equation \ref{eqn:snull} states the relation between the columns of $
+{\cal S}^{\pm}_{z_s,z_r}$ mentioned in the introduction to this
+section.
+
 \section{Unitarity}
@@
 
-\section{Numerical Examples}
-This section illustrates the most important conclusions developed in
-the preceding sections by finite difference wavefield simulation.
-
-\subsection{Synthetic models and simulation}
-To illustrate the structure described in the preceding section, I
-introduce two 2D acoustic models, one spatially homogeneous, the other
-highly refractive. The first, homogenous model has $\kappa = 4$
-GPa and $\rho = 1$ g/cm$^3$ throughout a rectangular domain of size 8 km ($x$) $\times$ 4 km
-($z$). The second, refractive, model is a perturbation of the first by
-a low-velocity acoustic lens positioned in the center of the
-rectangle (Figure \ref{fig:bml0}. To produce this structure, the density is chosen
-homogeneous as in the first model, while the bulk modulus decreases to
-from 4 GPa outside the lens to 1.6 GPA in its center, as shown in Figure \ref{fig:bml0}.
-
-\plot{bml0}{width=\textwidth}{Bulk modulus, lens model. Color scale is 
-in GPa. Positions of point source and receiver line indicated.}
-
-Discretization is conventional, with a rectangular grid and staggered
-finite difference scheme \cite[]{Vir:84} of order 2 in time and
-2$k$ in space; for most of the experiments reported below,
-$k=4$. Absorbing boudary conditions of perfectly matched layer type
-are applied at all boundaries of the simulation rectangle \cite[]{Habashy:07}.
-Sampling operators such as $P_r$ are implemented via linear
-interpolation, and source insertion via adjoint linear interpolation
-(as noted above, in the continuum limit, sources are represented via
-adjoint sampling). Steps in $x$ and $z$ are the same. In the following
-examples, $\Delta x = 20$ m. This choice limits the temporal frequency
-of accurately computed fields to rougly 12 Hz.
-
-\cite{GeoPros:11} gives a description of the code
-implementation, out-of-date in a few respects but overall accurate.
-The implementation uses the discrete adjoint state method and
-auto-generated code \cite[]{TapenadeRef13}, to assure that the
-computed adjoint operators are adjoint at the level of machine
-precision to the computed operators. The reverse-time storage issue is
-resolved through the optimal checkpointing technique
-\cite[]{Griewank:book,Symes:06a-pub}, again without loss of
-precision. This procedure results in computed adjoints for
-$S^+_{z_s,z_r}$ and other operators that pass usual test for adjoint
-accuracy, comparing inner products with pseudorandom input vectors,
-with errors well under machine precision.
-
-%Typical results with pseudorandom input traces $d_r, h_z, w_r, f_s$ are:
-%\begin{itemize}
-%\item computed $\langle d_r, S^+_{z_s,z_r}h_s\rangle = -2.41069174,
-%  \langle (S^+_{z_s,z_r})^Td_r, h_s \rangle = -2.41069388.$
-%\item  computed $\langle w_r, V^+_{z_s,z_r}f_s\rangle = -2.73362470,
-%  \langle (V^+_{z_s,z_r})^Tw_r, f_s \rangle = -2.73362136.$
-%\end{itemize}
-%Ideally, the differences of these inner products should be at most a
-%relatively small multiple of machine precision, {\em relative} to the
-%products $\|d_r\|\| S^+_{z_s,z_r}\|\|h_s\|$ and
-%$\|w_r\|\|V^+_{z_s,z_r}\|\|f_s\|$ (division by these quantities makes the result
-%dimensionless and scale-independent). The operator norm $\| S^+_{z_s,z_r}\|$ is
-%computationally inaccessible, so instead I used the smaller quantities
-%$\|d_r\|\| S^+_{z_s,z_r}h_s\|$ etc. as stand-ins - thereby
-%overestimating the relative error between the inner products. In all
-%cases, over a very large number of random inputs, the largest observed
-%relative error estimate was $O(10^{-9})$, well under the appropriate limit for
-%single precision.
-
-%We also explore the dependence of a few results on frequency, using
-%$\Delta x = 10$ m and $5$ m, to accomodate 25 and 50 Hz respectively,
-%maintaining 8 samples per wavelength. All computations are carried out
-%in single precision.
-
-The horizontal line of receivers sits at depth $z_r = $ 1000 m, the
-(extended) sources at $z_s=3000$ m.  %as
-% shown in Figure \ref{fig:bml0}.
-Source and receiver $x$ ranges from $2000$ to
-$6000$ m. Note that we have
-reversed the order relation between $z_s$ and $z_r$ described in the
-text ($z_s<z_r$). This difference is immaterial for the purpose of
-illustrating the mathematical structures developed in the preceding paragraphs.
-%A single point (physical) source appears in these
-%experiments, located at $x_s=3500$ m, $z_s=3000$ m, as also indicated
-%in the figure.
-
-%Extended sources are confined to the horizontal
-%line through the physical source position, that is $z_s = 3000$ m,
-%over a 4 km interval starting at $x_r=$ 2000 m..
-
-%The point source pressure data generated by this configuration for the
-%homogeneous model is displayed in Figure \ref{fig:recphh0}, for the
-%lens model in \ref{fig:recplh0}. Triplication of arrivals is evident in the latter.
-
-%While inversion of pressure data is the main object of this exercise,
-%normal velocity ($v_z$) data will play an important role, so I display
-%the corresponding gathers in Figures \ref{fig:recvzhh0} and \ref{fig:recvzlh0}.
-
-\subsection{Creating downgoing fields}
-The downgoing condition constrains high-frequency energy of localized
-plane wave components, hence could be enforced by dip>
-filtering. However, a simpler approach is to construct fields that
-must be entirely downgoing at the source and receiver surfaces by
-virtue of ray geometry.
-
-Note that a point source
-on $z=z_s$ creates high frequency energy traveling on rays parallel
-and nearly parallel to $z=z_s$, so that won't do. However, placing a
-point source at a depth $z_d<z_s$ will work. Since the examples used here are
-homogeneous in $z<z_s$, and the sampling region for extended sources
-is a finite interval, all rays carrying high frequency energy cross
-the source surface $z=z_s$ at a postive angle, and the field and its
-traces are {\em a priori} downgoing. The same is obviously true at the
-receiver surface for the homogeneous model, but is also true for the
-lens model, as no rays are refracted horizontally {\em at the receiver surface}.
-
-The choice of a point source at
-$z_d=3500$ m, $x_d=3500$ m,  bandpass filter wavelet with
-corner frequencies $1, 2.5, 7.5, 12.5$ Hz, gives the causal pressure and
-velocity gathers at $z=z_s=3000$ m
-shown in Figures \ref{fig:dsrcphh0} and \ref{fig:dsrcvzhh0}.  Since the
-mechanical parameters in the homogeneous and lens models are the same
-for $z<z_s$, and no rays return to this zone in either model, these
-data are asymptotically the same for both models, and I show only the
-homogenous medium results.
-%for the lowest frequency
-%source, and scaled as appropriate for examples with higher frequency
-%and finer sampling.
-%wavelet as used for prior examples,
-
-\plot{dsrcphh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
-  pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
-  bandpass filter source. }
-
-\plot{dsrcvzhh0}{width=\textwidth}{Trace $P_sv_z^+$ on $z=z_s=3000$ m of
-  vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
-  bandpass filter source.}
-
-\subsection{Equivalence of pressure, velocity sources}
-These gathers are the pressure and
-velocity traces
-$(P_sp^+,P_sv^+_z)$ on $z=z_s$ of a downgoing acoustic field in
-$z<z_s$, hence related by the operator $\Lambda^+_{z_s}$.
-Equations \ref{eqn:snull}, \ref{eqn:tracejump10} and
-\ref{eqn:tracejump20} show that these differ by a factor of -2 from
-source functions $f_s$ and $h_s$ in the system \ref{eqn:awepm},
-with $h_s=0$ and $f_s=0$ respectively, that generate the same acoustic
-field in $z<z_s$, and in particular the same receiver traces on
-$z=z_r$, at least asymptotically.
-
-%To give an example of \ref{eqn:snull} in action, it is necessary to
-%create downgoing data and the action of the operator $\Lambda^{\pm}$
-%on a downdoing field data $f_s$ on $z=z_s$.
-
-%These source functions satisfy
-%the relation \ref{eqn:hfcondn}, therefore source vectors $(h_s,0)$ and
-%$(0, f_s)$ generate {\em the same acoustic field} $(p^+,\bv^+)$ in
-%$z>z_s$.
-Figures
-%\ref{fig:drecphh0}, \ref{fig:dfwdphh0}, \ref{fig:daltphh0} (homogeneous model) and
-\ref{fig:drecplh0}, \ref{fig:dfwdplh0}, \ref{fig:daltplh0} show the pressure
-gathers extracted at $z_r=1000$ m for the point source at $z=z_d$ and
-for the two choices of extended source at $z=z_s$, on the same color
-scale. The obvious similarity between the fields generated by the two
-extended sources,
-predicted by equation \ref{eqn:snull}, is confirmed by trace
-comparisons in figures %\ref{fig:drecphh0tr21}-\ref{fig:daltphh0tr81}
-%and
-\ref{fig:drecplh0tr81},\ref{fig:daltplh0tr81}. Other traces are
-equally similar.
-
-\plot{drecplh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the lens model described in the 
-  text, with point pressure source (constitutive defect) at $z_d=3500$
-  m, $x_d=3500$ m.}
-
-\plot{dfwdplh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the lens model described in the 
-  text, with extended pressure source (constitutive defect) on
-  $z=z_s=3000$ m given bythe field depicted in Figure
-  \ref{fig:dsrcvzhh0} scaled by -2 
-   ($h_s=-2P_sv_z^+=\Lambda^+_{z_s}P_sp^+$) and zero velocity source (vertical load) 
-  ($f_s=0$).}
-
-\plot{daltplh0}{width=\textwidth}{Pressure gather at receiver depth 
-  $z_r=1000$ m from field generated by causal solution of acoustic 
-  system \ref{eqn:awepm} in the lens model described in the 
-  text, with extended velocity source (vertical load) on $z=z_s=3000$
-  m given by the field depicted in Figure \ref{fig:dsrcphh0} scaled by -2
-  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
-  ($h_s=0$).}
-
-\plot{drecplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$) 
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:dfwdplh0}
-  (red).}
-
-\plot{daltplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$)
-  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
-  (red).}
-% time rev
-
-\subsection{Inversion by time reversal}
-I have applied the approximate inversion procedure suggested in
-equation \ref{eqn:approxinv} to the pressure gather shown in Figure
-% \ref{fig:drecphh0} and
-\ref{fig:drecplh0}, generated by a point source
-at $z_d=3500$ m, $x_s=3500$ m, propagating in the %homogeneous and
-lens
-model (Figure \ref{fig:bml0}). I choose this example for two
-reasons. First, the success of the inversion demostrates the
-insensitivity of the time reversal method to ray multipathing
-(triplication), evident in the data (Figure \ref{fig:drecplh0}).
-%s respectively
-Second, I will invert this data in the homogeneous model, that is,
-construct sources that (approximately) reproduce the data using a
-different material model than the one in which it was produced. This
-capability is critically important in the application of the
-approximate inversion in nonlinear inversion, where the early iterations
-involve solution of the source estimation problem \ref{eqn:esis} at
-(possibly very) wrong material models ${\bf c}$. Successful extension
-methods maintain data fit throughout the course of the inversion.
-%The source time dependence is again a [1, 2.5,
-%10, 12.5] Hz trapezoidal bandpass filter.
-
-As noted earlier, the acoustic field in this example is downgoing
-throughout the simulation range. It can be
-regarded as the result of either pressure or velocity source at $z=z_s$: the
-pressure source gather $h_s$ (Figure \ref{fig:dhshh0})  is -2 times the
-vertical velocity gather depicted in Figure \ref{fig:dsrcvzhh0}, the
-velocity source gather $f_s$ (Figure \ref{fig:dfshh0}) is -2 times the
-pressure gather depicted in Figure \ref{fig:dsrcphh0}.
-
-\plot{dhshh0}{width=\textwidth}{Pressure source gather = -2 $\times$
-  vertical velocity gather (Figure \ref{fig:dsrcvzhh0}).}
-
-\plot{dfshh0}{width=\textwidth}{Velocity source gather = -2 $\times$
-  pressure gather (Figure \ref{fig:dsrcphh0}).}
-
-Figure \ref{fig:dinvhslh0} shows the approximate inversion (via the
-first equation in display \ref{eqn:approxinv}) of the
-pressure gather shown in Figure \ref{fig:drecplh0}, inverted in the homogeneous model
-(rather than in the lens model used to generate the data). The result
-differs greatly from the pressure source shown in Figure
-\ref{fig:dhshh0}, as it must since it results from inversion in the
-wrong material model.
-Some dip filter effect is unavoidable and is caused by the aperture
-limitation of the acquisition geometry: the steeper dips in the source
-gather (Figure \ref{fig:dhshh0}) do not contribute to the data, nor to
-the inversion. Also, the limited receiver aperture causes truncation
-artifacts in the inversion. However, this result is an
-accurate inversion: re-simulation (application of $S^{+}_{z_s,z_r}$) {\em using the same (homogeneous)
-  model as used in the inversion} results in accurate recovery (Figure \ref{fig:drerecplh0}) of the
-input pressure gather (Figure \ref{fig:drecplh0}). The difference is
-shown on the same color scale in Figure \ref{fig:ddiffrecplh0}.
-
-\plot{dinvhslh0}{width=\textwidth}{Approximate inversion via first 
-  equation in display \ref{eqn:approxinv}. Inversion in homogenous model of 
-  the pressure gather in Figure \ref{fig:drecplh0}, simulated with 
-  lens model. Scaled version of output $v_z$ field obtained by 
-  applying transpose of $4V^+_{z_s,z_r}$. Quite different from 
-  pressure source gather (Figure \ref{fig:dhshh0}) used to generate
-  data - since inversion takes place in a different material model!}
-
-\plot{drerecplh0}{width=\textwidth}{Re-simulated pressure gather produced from
-  inverted source
-  shown in Figure \ref{fig:dinvhslh0}. Simulation in homogenous model
-  used for inversion.}
-
-\plot{ddiffrecplh0}{width=\textwidth}{Difference between gathers
-  displayed in Figures \ref{fig:drecplh0} and \ref{fig:drerecplh0},
-  plotted on same color scale.}
-
-%To see that this is indeed an inversion, I apply $S^{+}_{z_s,z_r}$ to
-%the source shown in Figure \ref{fig:dinvhshh0}, to produce the
-%re-simulation shown in Figure \ref{fig:drerecphh0}. Simulation is carried
-%out in the homogeneous model used in the inversion. Comparison with the
-%gather in Figure \ref{fig:drecphh0} via a difference plot
-%\ref{fig:ddiffrecphh0} verifies that a very close match has been
-%obtained, so the source in Figure \ref{fig:dinvhshh0} is indeed a
-%satisfactory inversion, even it differs considerably from the original
-%point source: the difference lies in an approximate null space of the modeling
-%operator, arising from the limited aperture of the acquisition
-%geometry.
-
-%The remainder of this section is devoted to illustrating two important
-%features of the approximate inversion developed above. The
-%first is the validity of the inversion approach given in display
-%\ref{eqn:approxinv} so long as the fields generated by the
-%model used in simulation and inversion are downgoing. The models
-%may be quite far from homogeneous, and even generate triplications,
-%whereas the inversions remain accurate modulo the dip filtering effect
-%of aperture. Figure \ref{fig:dinvhsll0} shows the pressure source at $z=z_s$
-%inverted from the data Figure
-%\ref{fig:drecplh0}, in which a triplication is obvious. Recall that
-%this data is generated by propagating a point source at $z_d=3500$ m
-%in the lens model. The inversion uses the same model. Figure
-%\ref{fig:dinvhsll0} is an aperture-limited version of the pressure
-%source at $z=z_s=3000$ m
-%(Figure \ref{fig:dhshh0}) that generates this data by a dip filter, in
-%the same way that the sources in Figures \ref{fig:dhshh0} and
-%\ref{fig:dinvhshh0} are related. The inverted source is an accurate
-%inversion: it generates the data depicted in Figure
-%\ref{fig:drerecpll0}, which is very close to the input data (Figure
-%\ref{fig:drecplh0}). The difference is displayed on the same color
-%scale in Figure \ref{fig:ddiffrecpll0}.
-
-%The second feature is a decoupling of the models used (implicitly or
-%otherwise) in simulation and those used in inversion: it is possible
-%to {\em accurately invert pressure data for pressure source using the
-%  wrong model}, in the sense that the inverted source will
-%generate an accurate recovery of the input data re-simulated in the
-%same (wrong) model. For example, inversion of the data of the previous
-%example (Figure \ref{fig:drecplh0}) {\em in the homogeneous model}
-%results in an inverted source gather at $z=z_s$ (Figure \ref{fig:dinvhslh0}
-%that is very far from a dip-filtered version of the source used
-%generate the data (Figure \ref{fig:dhshh0}).
-
-% unitarity
-
-\subsection{Quasi-unitary property of the modeling operator}
-The identities \ref{eqn:approxinv} and \ref{eqn:sv} would together
-establish the approximately unitary property of $S^+_{z_s,z_r}$, if
-$\Lambda$ were symmetric. Identity \ref{eqn:approxinv} was illustrated
-in the last subsection. Setting the symmetric issue aside for the
-moment, an illustration of the relation \ref{eqn:sv} proceeds as follows.
-%A minor extension of an earlier example, based on the gathers
-%displayed in Figures \ref{fig:dsrcphh0} ($P_sp^+$) and
-%\ref{fig:dsrcvzhh0} ($P_sv_z^+$) illustrates the identity
-%\ref{eqn:sv}.
-%The pressure gathers at $z=z_r$ produced by application of
-%$S^+_{z_r,z_s}$ to $h_s=-2P_s v^+_z$, and $\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T$ to $f_s
-%=-2P_sp^+$, have already been displayed (Figures \ref{fig:dfwdphh0}
-%and \ref{fig:daltphh0}), and are identical to several digits. The same 
-%is true of the vertical velocity gathers, shown in \ref{fig:dfwdvzhh0}
-%and \ref{fig:daltvzhh0}; their difference, plotted on the same scale,
-%appears in Figure \ref{fig:dsvcomphh0}. Once again, this relation
-%expresses the interchangeability of pressure and velocity sources,
-%related by $\Lambda$.
-
-Relation \ref{eqn:tracejump10} characterizes
-$\Lambda^+_{z_s}$ as connecting the pressure and velocity components
-of downgoing fields restricted to $z=z_s$. That is, the pressure
-source gather
-$h_s=\Lambda^{+}_{z_s}P_s p^+ = -2
-P_{z_s}v_z$, displayed in Figure \ref{fig:dhshh0}, is the image of the
-pressure gather in Figure \ref{fig:dsrcphh0} under
-$\Lambda^{+}_{z_s}$.
-The pressure gather
-\ref{fig:dfwdplh0} is the image of this pressure source gather under
-$S^+_{z_s,z_r}$ (using the lens model). The corresponding vertical velocity gather
-(Figure \ref{fig:dfwdvzlh0}) is $-1/2$ times $\Lambda^+_{z_r}P_r
-p$. Therefore scaling the data in Figure \ref{fig:dfwdvzlh0} by
-$-2$ produces
-$\Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_s p^+$. On the other
-hand, figure \ref{fig:daltvzlh0} shows the result of applying
-$V^+_{z_s,z_r}$ to $f_s=-2P_s p^+$. Therefore scaling the gather in
-Figure \ref{fig:daltvzlh0} by $-\frac{1}{2}$ produces $V^+_{z_s,z_r}P_zp^+$.
-Since the data in Figures \ref{fig:dfwdvzlh0} and \ref{fig:daltvzlh0}
-are essentially identical, the relation \ref{eqn:sv} holds for this
-example.
-
-\plot{dfwdvzlh0}{width=\textwidth}{Vertical velocity gather, generated
-  with a pressure source in the lens model, 
-  corresponding to pressure gather \ref{fig:dfwdplh0}.}
-
-\plot{daltvzlh0}{width=\textwidth}{Vertical velocity gather, generated
-  with a velocity source in the lens model, corresponding to
-  pressure gather \ref{fig:daltplh0}.}
-
-\plot{dsvcomplh0}{width=\textwidth}{Difference between velocity gathers
-  shown in Figures \ref{fig:dfwdvzlh0} and \ref{fig:daltvzlh0},
-  plotted on the same color scale as these figures.}
-
-%Figure \ref{fig:preddinvhshh0} shows the pressure source gather
-%produced from the pressure data gather in Figure \ref{fig:dsrcphh0} by
-%application of the operator on the right-hand side of formula
-%\ref{eqn:lamident} to the source surface pressure gather depicted in
-%\ref{fig:dsrcphh0}. The homogeneous model is used in all propagations
-%implicit in the prescription \ref{eqn:lamident}. The scale is the same
-%as for Figures \ref{fig:dinvhshh0} and \ref{fig:dhshh0}. In fact
-%Figure \ref{fig:preddinvhshh0} appears to be nearly identical to
-%Figure \ref{fig:dinvhshh0}, and both appear to be dip-filtered
-%versions of Figure \ref{fig:dhshh0}. The difference plot (Figure
- % \ref{fig:ddiffinvhshh0}), using the same color scale, shows that the
- % similarity is quantitative. The recovered source gather (Figure
- % \ref{fig:preddinvhshh0}) also generates the same acoustic fields as
-  %the point source at $z=z_d$: the pressure gather at $z=z_r=1000$ m
-%  depth (propagated in the lens model) is shown in Figure \ref{fig:dpredhsrecplh0}, and its
-%  difference with the point source gather (Figure \ref{fig:drecplh0})
-%  plotted on the same color scale in Figure
-%  \ref{fig:ddiffpredhsrecplh0}.
-
-%Figures \ref{fig:preddinvhsll0}, \ref{fig:ddiffinvhsll0},
-%\ref{fig:dpredhsrecpll0}, and \ref{fig:ddiffpredhsrecpll0} accomplish
-%the same comparison for the $\Lambda^+_{z_s}$ approximation in
-%equation \ref{eqn:lamident}, with all propagation taking place in the
-%lens model.
-
-\subsection{Economical computation of $\Lambda$ in ``thin'' subdomain}
-Equation \ref{eqn:lamnear} suggests a thin-slab computation of the
-$\Lambda$ action, which is both accurate and economical. 
-This calculations
-place a receiver array at $z_s+\Delta z=2900$ m depth, just 100 m above the
-source surface at $z_s=3000$ m. For the discretization used to create
-the examples shown so far, that is just a 5 gridpoint difference in
-depth, as opposed to 100 gridpoints between the source and receiver
-depths for examples such as shown in Figure \ref{fig:drecplh0}.
-
-Asymptotically, $\Lambda^{\pm}_{z_s}$ depends only on the medium
-coefficients ${\bf c}$ in an arbirarily small region containing the
-source surface $z=z_s$. In this example, the homogeneous and lens models are identical in the depth
-range $2900 < z < 3000$ m, so the computed $\Lambda^+_{z_s}$ operators
-will be precisely the same for both models. Hence I show only results for the
-homogenous model.
-
-The approximation to $\Lambda_{z_s}^+$ via equation \ref{eqn:lamnear}
-for this configuration is evaluated in
-Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
-\ref{fig:dprednshsrecpll0}, and \ref{fig:ddiffprednshsrecpll0}. The effect of aperture
-limitation is clearly diminished: the second figure in this series
-compares the full-aperture pressure source gather (Figure
-\ref{fig:dhshh0}) with the 
-image of the corresponding pressure gather (Figure \ref{fig:dsrcphh0})
-under the approximation to $\Lambda_{z_s}^+$, and the last figures
-show that the approximated source gathers accurately predict the
-point-source pressure gather at the receiver datum $z_r=1000$ m.
-
-\plot{preddnshshh0}{width=\textwidth}{Pressure source gather = image
-  under pressure-to-source operator $\Lambda^+_{z_s}$ of pressure gather
-  shown in Figure \ref{fig:dsrcphh0}, homogeneous model, using
-  ``near'' receiver traces at $z=2900$ m. Compare Figure
-  \ref{fig:dhshh0}: because the sources and
-  receivers are close, little aperture is lost in this case.}
-
-\plot{ddiffnshshh0}{width=\textwidth}{Difference between (a) image
-  (Figure \ref{fig:preddnshshh0}) of $\Lambda^+_{z_s}$ applied to
-  pressure gather (Figure \ref{fig:dsrcphh0}) using a near receiver
-  array to implement formula \ref{eqn:lamident}, and (b) source gather
-  (Figure \ref{fig:dhshh0}) inferred from vertical velocity.
-  Homogeneous model used in all
-  propagations. Same color scale as in Figure
-  \ref{fig:preddnshshh0}. }
-
-%\plot{dprednshsrecplh0}{width=\textwidth}{Pressure gather at receiver
-%  datum $z=z_r=1000$ m simulated in lens model from source gather shown in Figure
-%  \ref{fig:preddnshshh0}. Compare with point source pressure gather
-%  (Figure \ref{fig:drecplh0}).}
-
-%\plot{ddiffprednshsrecplh0}{width=\textwidth}{Plot of difference 
-%  between data shown in Figures \ref{fig:drecplh0} and 
-%  \ref{fig:dprednshsrecplh0}, plotted on same color scale as the latter 
-% two figures.}
-
-\plot{dprednshsrecpll0}{width=\textwidth}{Pressure gather at receiver
-  datum $z=z_r=1000$ m simulated in lens model from source gather shown in Figure
-  \ref{fig:preddnshshh0}. Compare with point source pressure gather
-  (Figure \ref{fig:drecplh0}).}
-
-\plot{ddiffprednshsrecpll0}{width=\textwidth}{Plot of difference 
-  between data shown in Figures \ref{fig:drecplh0} and 
-  \ref{fig:dprednshsrecpll0}, plotted on same color scale as the latter 
-  two figures.}
-\subsection{Symmetrizing $\Lambda$}
-
-Figure \ref{fig:preddnshstrhh0} shows the image of the pressure gather
-in Figure \ref{fig:dsrcphh0} under $(\tilde{\Lambda}^+_{z_s})^T$,
-using the ``near'' traces at $z=2900$, that is, $\Delta z = 100$ m in
-expression \ref{eqn:lamtransp}, and propagation in the
-homogeneous model. Note
-the close resemblance to the image of the same pressure gather under
-$\tilde{\Lambda}^+_{z_s}$ displayed in Figure
-\ref{fig:preddnshshh0}. The difference of these two images is
-displayed in \ref{fig:ddiffnslamtrhh0}, on the same color scale as the
-images themselves. Since the propagation takes place entirely in a
-region where all of the mechanical parameters are homogenous, I do not
-offer a similar comparison for the lens model.
-
-\plot{preddnshstrhh0}{width=\textwidth}{Pressure source gather = image
-  under {\em transpose} of pressure-to-source operator
-  $\Lambda^+_{z_s}$ of pressure gather shown in Figure
-  \ref{fig:dsrcphh0}, homogeneous model, using ``near'' receiver
-  traces at $z=2900$ m. Compare Figure \ref{fig:dhshh0} and
-  \ref{fig:preddnshshh0}: as noted in the text, $\Lambda^+_{z_s}$ is
-  asymptotically symmetric, so the resemblance is not a surprise.}
-
-\plot{ddiffnslamtrhh0}{width=\textwidth}{Difference between data in
-  Figures \ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}, plotted
-  on the same scale as these figures, showing that the asymptotic
-  symmetry of $\Lambda^+_{z_s}$ is actually quantitative for the
-  length, time and frequency scales of these examples.}
-
-\subsection{Asymptotic symmetry of $\Lambda$}
-
-Figure \ref{fig:symmdnshshh0} shows the output of the symmetrized
-approximate source-to-pressure operator per equation \ref{eqn:winvcomp},
-applied once again to the pressure data in Figure
-\ref{fig:dsrcphh0}. Note the resemblance to Figures
-\ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}. These are all
-asymptotic approximations of each other. Figure
-\ref{fig:ddiffsymmdnshsll0} shows the
- the difference between the pressure gather at $z=z_r$ produced from
- the pressure source output by the symmetrized $\Lambda$, and the point source
-simulation (Figure \ref{fig:dfwdplh0}), plotted on the same scale as
-the latter, in both cases with all propagations in the lens model.
-
-\plot{symmdnshshh0}{width=\textwidth}{Pressure source gather = image
-    under {\em symmetrized} pressure-to-source operator
-    $\frac{1}{2}\left(\Lambda^+_{z_s}+(\Lambda^+_{z_s})^T\right)$ of
-    pressure gather shown in Figure \ref{fig:dsrcphh0}, homogeneous
-    model, using ``near'' receiver traces at $z=2900$ m. Compare
-    Figure \ref{fig:preddnshshh0}.}
-
-\plot{ddiffsymmdnshsll0}{width=\textwidth}{Difference between point
-  source simulation (Figure \ref{fig:dfwdplh0}) and pressure gather at
-  $z=z_r=1000$ m produced by simulation with the source shown in
-  Figure \ref{fig:symmdnshshh0},
-  propagation in the lens model.}
-
-\subsection{Unitary property of modeling operator}
-To illustrate this unitary property of $S^+_{z_s,z_r}$, I apply the
-operator
-\[
-  \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
-  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
-  \Lambda^+_{z_r})
-\]
-to the data $S^+_{z_s,z_r}h_s$ (Figure \ref{fig:drerecplh0}), in which
-$h_s$ is the %version (Figure \ref{fig:dinvhslh0}) of the
-downgoing source created earlier (Figure \ref{fig:dhshh0})%, produced
-%by inversion of the data in Figure \ref{fig:drecplh0} via the first
-%equation in display \ref{eqn:approxinv}, propagation in the homogenous
-%model.
-The operator above is computed via the technique explained in
-the preceding subsection, below, using
-auxiliary receiver arrays 100 m above the data source and receiver arrays.
-
-The output is shown in Figure
-\ref{fig:lamsstlamrdrecplh0}. The difference with the actual source is
-shown in Figure \ref{fig:difflamsstlamrdrecplh0}.
-%A similar exercise
-%using the lens data \ref{fig:drecplh0} but inversion in the homogenous
-%model produces the extended source depicted in
-%\ref{fig:lamsstlamrdrecplh0}, which differs from the inversion result
-%via equation \ref{eqn:approxinv} by the gather displayed in Figure
-%\ref{fig:difflamsstlamrdrecplh0}.
-
-\plot{lamsstlamrdrecplh0}{width=\textwidth}{Inversion of data shown in
-  Figure \ref{fig:drecplh0}, simulated in lens model, using the
-  approximate unitarity relation \ref{eqn:unitary} and propagation in
-  homogenous model.}
-
-\plot{difflamsstlamrdrecplh0}{width=\textwidth}{Difference between
-  data displayed in Figures \ref{fig:dinvhslh0} and
-  \ref{fig:lamsstlamrdrecplh0}, plotted on the same color scale.}
-
-\subsection{Preconditioned CG iteration}
-This final subsection shows that result of Conjugate Gradient
-iteration, with and without preconditioning, applied to the source
-estimation problem \ref{eqn:esis}, with zero and non-zero penalty
-weight $\alpha$. The data $d$ is the gather shown in
-\ref{fig:drecplh0}, simulated using the lens model with source shown
-in Figure \ref{fig:dhshh0}, or, alternatively, a point source with
-bandpass filter wavelet located at $x_d=3500$ m, $z_d=3500$ m. In the inversion,
-the material model is taken to be homogeneous, as has been the case in
-all of the previous examples.
-
-Figure \ref{fig:compnres0lh0} shows the progress of the normal residual
-(Euclidean norm of the difference of the two sides of equation \ref{eqn:norm1}),
-for Conjugate Gradient and Preconditioned Conjugate Gradient
-(Algorithm 1) iterations, applied to solution of the optimization
-problem \ref{eqn:einv} with $\alpha=0$. For CG, the norms are both the ordinary
-Euclidean norm, $W_m=W_d=I$. For PCG, $W_m$ and $W_d$ are given in
-display \ref{eqn:wdef}, with the symmetrized $\Lambda$s computed as
-indicated in the preceding subsections. Convergence for the
-preconditioned algorithm is roughly 4 times as fast.
-
-\plot{compnres0lh0}{width=\textwidth}{Comparison of normal residual 
-  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
-  vs. iteration. Data = lens model, point source (Figure 
-  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
-  $\alpha=0$.}
-
-Figure \ref{fig:compnres1lh0} shows the same comparison with non-zero
-penalty weight, $\alpha=10^{-3}$. The PCG normal residual curve is
-almost identical with that in the $\alpha=0$ case, wheras the CG
-convergence has slowed down noticeably, being about five times as slow
-as the preconditioned algorithm.
-
-\plot{compnres1lh0}{width=\textwidth}{Comparison of normal residual 
-  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
-  vs. iteration. Data = lens model, point source (Figure 
-  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
-  $\alpha=10^{-3}$.}
-
 \section{Conclusion}
@@
 
-The transmission inverse problem figuring most prominently in
-contemporary applied seismology is surely the Full Waveform Inversion
-(FWI) of diving wave data. This is essentially the same problem as the
-one discussed in this paper, and can be formulated and treated the
-same way, at least for acoustic wave physics. All of the topics
+All of the topics
 treated here are open for elastic wave physics - the analogue of the
@@
 \bibliographystyle{seg}
-\bibliography{../../bib/masterref}
+\bibliography{../../../bib/masterref}
 

### 2022-10-25 repro/papers/GEM22/submitR1/paper.tex
mods

@@
 feasible (or ``physical'') source models.
-\cite{HuangNammourSymesDollizal:SEG19} present an overview of the
-literature on source extension methods, describing a variety of
-methods to add degrees of freedom to physical source model.
+
+Well-studied examples of extended source approaches to FWI include
+Wavefield Reconstruction Inversion (WRI)  \cite[]{LeeuwenHerrmannWRI:13,LeeuwenHerrmann:16,Lietal:18,Aghmiryetal:20,Louboutinetal:20} and Adaptive Waveform
+Inversion (AWI)
+\cite[]{Warner:16,GuachWarnerRavaut:GEO19,Guaschetal:NPJDM20,Yongetal:EAGE21,Warneretal:SEG21}. \cite{HuangNammourSymesDollizal:SEG19}
+present an overview of the recent
+literature on source extension methods.
 
 The present paper concerns {\em surface source extension}: physical
-sources are presumed to be concentrated spatially at points $\bx_s$
-in space, whereas their extended counterparts are permitted to spread
-energy over a {\em source surface} containing the physical source spatial
-locations. Similarly, the receiver spatial locations are confined to a
-{\em receiver surface}. A simple choice for the penalty operator $A$ is then
-multiplication by the distance $|\bx-\bx_s|$ to the physical source
-location:
+sources are presumed to be concentrated at points in space, whereas
+their extended counterparts are permitted to spread energy over a {\em
+  source surface}. Similarly, receivers are located on a {\em receiver
+  surface}. Assuming that the presumed point source is at location
+$\bx_s$, a simple (though certainluy not the only) choice for the
+penalty operator $A$ is then multiplication by the distance
+$|\bx-\bx_s|$ to the physical source location:
 \begin{equation}
@@
 \label{eqn:esis}
-\mbox{given } {\bf c}, \mbox{ choose } \bff \mbox{ to minimize }
+\mbox{given } {\bf c}, \mbox{ choose surface source } \bff \mbox{ to minimize }
 \|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha \|A\bff\|^2 
@@
 throughout to illustrate the theory. In both, horizontal lines serve as
-source and receiver surfaces. The first is a ``crosswell'' or slab
-configuration with an acoustic lens positioned between a deeps source
+source and receiver surfaces. The first is a ``horizontal crosswell'' or slab
+configuration with an acoustic lens positioned between a deep source
 and a shallower line of 
@@
 
-The next section defines the modeling operator $PL[{\bf c}]$, its
-adjoint, and important specializations (pressure vs. normal velocity
+Since the constructions described here involve a significant number of
+components and ideas, I begin the paper with an overview, mapping out
+the key steps and their relation to other work.
+The next section defines the modeling operator $PL[{\bf c}]^{-1}$ and important specializations (pressure vs. normal velocity
 sources and data). It also introduces the two 2D examples mentioned
@@
 modeling operator by time reversal (as suggested by work in
-photoacoustic tomography), and illustrates its efficacy in the two
+photoacoustic tomography), and illustrates its efficacy via the two
 examples. This construction requires extraction of
@@
 
-The preconditioner construction is closely related to the time
+The preconditioner construction is partly inspired by the time
 reversal method in photoacoustic tomography
-\cite[]{StefanovUhlmannIP:09,Hristova:09}. A simplified mathematical
+\cite[]{StefanovUhlmannIP:09,Hristova:09}. This problem
+clearly has strong similarities to, but also differences from, the
+problem studied here.
+
+A simplified mathematical
 translation of this medical imaging task is to infer the initial
@@
 $t=0$ from measurements of the pressure on a surface enclosing the
-fluid over a time interval $0 \le t \le t_{\rm max}$. This problem
-clearly has strong similarities to, but also differences from, the
-problem studied here. The time reversal method presumes that the
-pressure field has returned to equilibrium (zero excess pressure), or close to it, at the
+fluid over a time interval $0 \le t \le t_{\rm max}$. The time reversal method presumes that the
+pressure field has returned to equilibrium (zero excess pressure), or
+close to it, in the fluid-containing region at the
 final time $t=t_{\rm max}$ Then the field can be (at least
@@
 is to be determined, but a time-extended source $\bff$ confined to a
-surface. The penalty operator has no analogue in the basic
+surface. The penalty operator $A$ has no analogue in the basic
 photoacoustic problem description. Finally, once these obstacles are
@@
 requires that the operator be identified as the adjoint of the
-modeling operator with respect to suitable inner products in its domain
-and range. In the pages to follow, I will address each of these issues.
+modeling operator with respect to suitable inner products in its
+domain and range. The next few paragraphs sketch resolutions of each
+of these issues.
 
@@
 within high-frequency asymptotic approximation. This step requires
-some assumptions about the ray fields: all rays responsible for significant
-energy in the receiver data must arrive from the source surface, and must cross the source and receiver
-surfaces transversally, that is, making non-zero angles. Also, the
-rays must approach source and receiver surfaces from one side or the
-other (``the inside''), so that locally the surfaces can be treated as boundaries of
-propagation domains. With these assumptions, reverse-time propagation
-of the receiver data
+some assumptions about the ray fields: all rays responsible for
+significant energy in the receiver data must arrive from the source
+surface, and must cross the source and receiver surfaces
+transversally, that is, making non-zero angles. Also, the rays must
+approach source and receiver surfaces from one side or the other
+(``the inside''), so that locally the surfaces can be treated as
+boundaries of propagation domains. Specifically, rays must emanate
+from the source surface towards the inside (``incoming'' rays), and
+arrive at the receiver surface from the inside (``outgoing'' rays).
+With these assumptions, reverse-time propagation of the receiver data
 closely approximates the acoustic fields near the source surface. Next
@@
 surface. Thus reverse-time propagation and reading off the velocity
-field on the source surface yields
-a source that reproduces the pressure data, within an asymptotically small error.
+field on the source surface yields a source that reproduces the
+pressure data, within an asymptotically small error.
 %Preconditioning amounts to a change of
@@
 version of the adjoint state construction \cite[]{Plessix:06}. So the
-approximate inverse is: conversion of pressure to source at the
+approximate inverse may be described as: conversion of pressure to source at the
 receiver surface, followed by the adjoint of the modeling operator,
@@
 unitary with respect to these weighted norms, which in turn implies
-that the preconditioned conjugate gradient algorithm should converge rapidly.
+that a properly constructed preconditioned conjugate gradient algorithm should converge
+rapidly.
+
+FInally, the $\alpha=0$ case is not sufficient: the penalty operator
+$A$ is an essential component of the nonlinear inverse propblem
+\ref{eqn:esi}. As it turns out, $A$ commutes with the other operators
+involved - approximately, but that is enough. Therefore the effect of
+$A$ can be compensated with an easily-computed factor, whence the
+preconditioned acceleration extended to the case $\alpha > 0$.
 
@@
 \cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09}.
-
-FInally, the $\alpha=0$ case is not sufficient: the penalty operator
-$A$ is an essential component of the nonlinear inverse propblem
-\ref{eqn:esi}. As it turns out, $A$ commutes with the other operators
-involved - approximately, but that is enough. Therefore the effect of
-$A$ can be compensated with an easily-computed factor, whence the
-preconditioned acceleration extended to the case $\alpha > 0$.
 
@@
 \end{equation}
-$L[{\bf c}]$ has a well-defined inverse in the sense of distributions
+$L[{\bf c}]$ has a well-defined inverse %in the sense of distributions
 if it is restricted to either causal or anti-causal vector wavefields.
@@
 describe the theory for $n=3$, write $\bx=(x,y,z)^T$ for the spatial
-coordinate vector, and refer to the third (vertical) coordinate of
-particle velocity $\bv$ as $v_z$. For computational convenience, the
+coordinate vector, and similarly for the 
+particle velocity $\bv = (v_x,v_y,v_z)$. For computational convenience, the
 examples are two-dimensional.
@@
 The surface source extension replaces point sources on or near a
-surface in $\bR^3$ with source functions confined to the 
-surface. The simplest example of this extended geometry specifies a
-plane $\{(x,y,z,t): z=z_s\}$ at source depth $z_s$ as the surface. For
+surface $\Sigma_s$ in $\bR^3$ with source functions confined to the
+surface . The surface should be supplied with a choice of unit normal
+vector field ${\bf n}_s$. Since the surface separates space into two
+non-overlapping parts, at least locally, designate the part into which
+the normal vector points as the ``outside'', so that ${\bf n}_s$ is
+the outward normal.
+
+For
 acoustic modeling, surface sources are combinations of constitutive law
-defects and loads normal to the surface, localized on $z=z_s$. That
+defects and loads normal to the surface, localized on $\Sigma_s$. That
 is, right-hand sides in the system $L\bu=\bff$ take the form
-$\bff(\bx,t) = (h_s(x,y,t)\delta(z-z_s),
-f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar defect $h_s$ and normal
-force $f_s$ ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in
-\ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$
+$\bff(\bx,t) = (h_s\delta_{\Sigma_s},
+f_s{\bf n}_s\delta_{\Sigma_s})^T$ for scalar defect $h_s$ and normal
+load $f_s$. With the choice $L$ given in
+\ref{eqn:awemat}, the causal system $L\bu=\bff$
 takes the form
@@
 \label{eqn:awepm}
-\frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} & = & - \nabla \cdot \bv^{\pm} +
-h_s \delta(z-z_s), \nonumber \\
-\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} +
-                                                f_s{\bf e} \delta(z-z_s),\nonumber \\
-p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
-\bv^{\pm} & = & 0 \mbox{ for } \pm t \ll 0.
+\frac{1}{\kappa}\frac{\partial p}{\partial t} & = & - \nabla \cdot \bv +
+h_s \delta_{\Sigma_s}, \nonumber \\
+\rho\frac{\partial \bv}{\partial t} & = & - \nabla p +
+                                                f_s{\bf n}_s \delta_{\Sigma_s},\nonumber \\
+p & =& 0 \mbox{ for }  t \ll 0,\nonumber\\ 
+\bv & = & 0 \mbox{ for } t \ll 0.
 \end{eqnarray}
@@
 \noindent {\bf Remark:} In system \ref{eqn:awepm} and many similar
-systems to follow, I will use the shorthand
-\[
-  p^+ = 0 \mbox{ for } t \ll 0 
-\]
-to mean that $p^+$ is {\em causal}, that is,
-\[
-  \mbox{For some } T \in \bR, p^+(\cdot,t) = 0 \mbox{ for all } t <
+systems to follow, I will use the shorthand such as
+\[
+  p = 0 \mbox{ for } t \ll 0 
+\]
+to mean that $p$ is {\em causal}, that is,
+\[
+  \mbox{For some } T \in \bR, p(\cdot,t) = 0 \mbox{ for all } t <
   T.
@@
 \[
-  p^- = 0 \mbox{ for } t \gg 0 
-\]
-signifies that $p^-$ is anti-causal.
+  p = 0 \mbox{ for } t \gg 0 
+\]
+signifies that $p$ is anti-causal.
 
 Extended forward modeling consists in solving \ref{eqn:awepm} and
-sampling the solution components at receiver locations. For
-simplicity, throughout this paper I will assume that the receivers are
-located on another spatial plane $\{(x,y,z,t): z=z_r\}$ at receiver
-depth $z_r$. The constructions to follow involve interchange of the
-roles of $z_s$ and $z_r$ (that is, locating sources on $z=z_r$ and
-receivers at $z=z_s$), so both require sampling operators: $P_s,P_r$
-are the sampling operators on $z=z_s$, $z=z_r$ respectively. In
+sampling the solution components at a receiver surface $\Sigma_r$ with
+(``outward'') unit normal field ${\bf n}_r$.
+$P_s$ and $P_r$ are the sampling operators on $\Sigma_s$ and
+$\Sigma_r$ respectively. In
 practice, sampling necessarily occurs at a discrete array of points
@@
 theoretical discussion, I will neglect the finite sample rate, and
-regard the data, for example $P_rp^+$, as continuously sampled. The
+regard the data, for example $P_rp$, as continuously sampled. The
 output samples are necessarily muted, that is, non-zero only over a space-time domain
@@
 
-The causal/anti-causal vector
-modeling operators ${\cal S}^{\pm}_{z_s,z_r}$ are defined in terms of
-the solutions $(p^{\pm},\bv^{\pm})$ of the systems \ref{eqn:awepm} by
-\begin{equation}
-  {\cal S}^{\pm}_{z_s,z_r}(h_s,f_s)^T  = (P_rp^{\pm},P_r v_z^{\pm})^T,
+The vector modeling operator ${\cal S}$ relates the source amplitudes
+$h_s,f_s$ on $\Sigma_s$ to the sampled pressure and normal velocity
+$p$, $\bv \cdot {\bf n}_r$ for 
+the solution $(p,\bv)$ of the systems \ref{eqn:awepm} by
+\begin{equation}
+  {\cal S}(h_s,f_s)^T  = (P_rp,P_r(\bv \cdot {\bf n}_r) )^T,
   \label{eqn:fwd}
 \end{equation}
-The subscript signifies that sources are located on $z=z_s$,
-receivers on $z=z_r$. It is necessary to include this information in
-the notation, as versions of ${\cal S}^{\pm}$ with sources and receivers in
-several locations will be needed later in the discussion.
-
-Denote by $\Pi_i, i=0,1$ the projection on the first,
-respectively second, component of a vector in $\bR^2$. The 
-forward modeling operator from pressure source to pressure trace is
+
+Denote by $\Pi_i, i=0,1$ the projection on the first,respectively
+second, component of a vector in $\bR^2$. The forward modeling
+operator from pressure source to pressure trace is
 \begin{equation}
   \label{eqn:sdef}
-  S^{\pm}_{z_s,z_r} = \Pi_0 {\cal S}^{\pm}_{z_s,z_r} \Pi_0^T 
-\end{equation}
-and the forward modeling operator from velocity source (normal force)
-to velocity trace is
-\begin{equation}
-  \label{eqn:vdef}
-  V^{\pm}_{z_s,z_r} = \Pi_1 {\cal S}^{\pm}_{z_s,z_r} \Pi_1^T 
-\end{equation}
+  S = \Pi_0 {\cal S}\Pi_0^T 
+\end{equation}
+%and the forward modeling operator from velocity source (normal force)
+%to velocity trace is
+%\begin{equation}
+%  \label{eqn:vdef}
+%  V_{z_s,z_r} = \Pi_1 {\cal S}_{z_s,z_r} \Pi_1^T 
+%\end{equation}
 
 With these conventions, we can write the version of the source
-subproblem \ref{eqn:esis} studied in this paper as
+subproblem \ref{eqn:esis} studied in this paper as: given a pressure
+gather $d$ on $\Sigma_r$,
 \begin{equation}
   \label{eqn:esisp}
-  \mbox{find }h_s\mbox{ to minimize }\|S^{+}_{z_s,z_r}h_s- d\|^2 +
+  \mbox{find }h_s\mbox{ to minimize }\|Sh_s- d\|^2 +
   \alpha^2\|Ah_s\|^2.
@@
 just described. The first example (Figure \ref{fig:olensrays0}) embeds
-an acoustic (low-velocity) lens between a source surface at $z_s=3000$
-(orange horizontal line) 
-m and a receiver surface at $z_r=1000$ m (yellow horizontal line).  The data to be used in this
-example results from a point source at $z_s=3500, x_s=3500$ m (orange
-star). I have overlain the rays connecting this source point with the
-data portion of the receiver surface: evidently all of the rays
-involved cross the source and receiver surfaces transversally. Also,
-regarding ``inwards'' as being ``upwards'' at the source surface,
-``downwards'' at the receiver surface, it is clear that the data is
-incoming at the source surface, outgoing at the receiver
-surface.
+an acoustic (low-velocity) lens between a source surface
+$\Sigma_s= \{z = z_s=3000 \mbox{ m}\}$ (orange horizontal line) m and
+a receiver surface $\Sigma_r= \{z=z_r=1000 \mbox{ m}\}$ (yellow
+horizontal line).  The data to be used in this example results from a
+point source at $(3500, 3500)$ m (orange star). I have overlain
+the rays connecting this source point with the data portion of the
+receiver surface: evidently all of the rays involved cross the source
+and receiver surfaces transversally. Also, regarding ``inwards'' as
+being ``upwards'' at the source surface, ``downwards'' at the receiver
+surface, it is clear that the data is incoming at the source surface,
+outgoing at the receiver surface.
 
@@
 %\ref{fig:drecplh0}, \ref{fig:dfwdplh0}
-Figure \ref{fig:drecplh0} shows the gather $P_rp_{\rm pt}^{+}$, where
-$p_{\rm pt}^{+}$ is causal field generated by the point source at
+Figure \ref{fig:drecplh0} shows the gather $P_rp_{\rm pt}$, where
+$p_{\rm pt}$ is causal field generated by the point source at
 $(3500,3500)$ m with a trapezoidal bandpass filter wavelet having
@@
 traces in the spatial range 2000 m $ \le x_r \le $ 6000 m and time
-range 0 s $\le t \le $ 2 s. The modeling operator output $S_{z_s,z_r}^+h_s$,
-obtained from the causal solution $(p^+, \bv^+)$ of \ref{eqn:awepm}
+range 0 s $\le t \le $ 2 s. The modeling operator output $Sh_s$,
+obtained from the causal solution $(p, \bv)$ of \ref{eqn:awepm}
 with this choice of $h_s$ and $f_s=0,$ by application of the sampling
-operator $P_r$ to $p^+$, are shown in Figure \ref{fig:dfwdplh0}.
+operator $P_r$ to $p$, is shown in Figure \ref{fig:dfwdplh0}.
 
@@
 extended source $h_ s$ shown in Figure \ref{fig:dsrcphh0} is constructed
-so that $S^++_{z_s,z_r}h_s$ (Figure \ref{fig:dfwdplh0}) closely
-approximates the point source gather $P_r p^+_{\rm pt}$ (Figure
+so that $Sh_s$ (Figure \ref{fig:dfwdplh0}) closely
+approximates the point source gather $P_r p_{\rm pt}$ (Figure
 \ref{fig:drecplh0}). This construction will be explained in the next section.
@@
 This geometry permits inversion of the diving wave data alone, which
-is isolated via a mute, depicted in (Figure \ref{fig:ndwmute0}. The
+is isolated via a mute, depicted in Figure \ref{fig:ndwmute0}. The
 mute includes time truncation at 6 s, and offsets between -10000 and
@@
 \end{figure}
-%As explained in the Introduction, the downgoing constraint on the square-integrable
-%source functions $h_s, f_s$ is
-%essential both for finite energy solutions of the system
-%\ref{eqn:awepm} to exist, and for these solutions to have well-defined
-%traces on the receiver surface $z=z_r$. This constraint will be
-%assumed throughout, often tacitly.
-%For downgoing solutions of system \ref{eqn:awepm}, the key
-%components ($p^{\pm}$ and $v^{\pm}_z$) are continuous functions of $z$
-%in the open slab $z_s<z<z_r$ with well-defined limits at the boundary
-%planes, but may be discontinuous at the source plane
-%$z=z_s$. Similarly, the roles of $z_s$ and $z_r$ will be interchanged
-%in some of the constructions to come, and the corresponding solutions
-%may be discontinuous at $z=z_r$. Accordingly, interpret $P_s$, $P_r$
-%as the limit from right and left respectively: for $u=p^{\pm}$ or
-%$v^{\pm}_z$,
-%\begin{eqnarray}
- % \label{eqn:defsamp}
- % P_su(x,y,t) &=& \lim_{z \rightarrow z_s^+} u(x,y,z,t),\nonumber \\
- % P_ru(x,y,t) &=& \lim_{z \rightarrow z_r^-} u(x,y,z,t).                  
-%\end{eqnarray}
-
-%\noindent {\bf Remark:} To connect with the formulation presented in
-%the introduction, note that for continuous $u$,
-%$P_su(x,y,t)=u(x,y,z_s,t)$, and therefore the adjoint of $P_s$ (in the
-%sense of distributions) is $P_s^Th(x,y,z,t) =
-%h(x,y,t)\delta(z-z_s)$. Write ${\cal P}_s = \mbox{diag }(P_s,P_s)$ and
-%similarly for ${\cal P}_r$. Then
-%\[
- % {\cal S}^{+}_{z_s,z_r} = {\cal P}_r L^{-1}({\cal P}_s)^T,
-%\]
-%in which $L^{-1}$ is interpreted in the causal sense, and similarly
-%for ${\cal S}^{-}$. Sources confined to $z=z_s$ are precisely those
-%functions (distributions, really) output by ${\cal P}_s^T$, so the
-%problem statements \ref{eqn:esi} and \ref{eqn:esis} can be rewritten
-%in terms of ${\cal S}^+_{z_s,z_r}$, with $P$ identified with ${\cal P}_r$.
-
-%${\cal S}^{\pm}$ is not stably invertible: its columns are
-%approximately linearly dependent, as will be verified below. The
-%diagonal components of ${\cal S}^{\pm}$ thus carry essentially all of
-%its information, and it is in terms of these that a sensible inverse problem
-%is defined.
 
@@
 
-Recall that the source vector $(h_s,f_s)$ is assumed to produce a
-downgoing field $(p^+,\bv^+)$, that is, emanates high-frequency energy only along
-rays that make an angle with the vertical bounded below by a common
-minimum angle. Such rays leave $\Omega$ within a common maximum
-time. Consequently (Appendix B), in the
-slab $z_s<z<z_r$, the field $(p^+,\bv^+)$ approximates the solution of an
-anti-causal evolution equation. Choose $\chi(t)$ to be a smooth function
-that is $= 0$ for $t \gg 0$ and $=1$ at times when near rays carrying
-high-frequency energy in $(p^+,\bv^+)$ cross $z=z_r$. Define 
-$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
-$\Omega \times \bR$ of
+Time reversal as found in the literature on photoacoustic tomography
+must be localized to apply to the surface source inversion problem,
+since there is no measurement surface surrounding the region of
+propagation. It must also be modified to produce a source, rather than
+a restriction of the pressure field. The assumption of incoming rays at the source surface and
+outgoing rays at the receiver surface makes both modifications  
+possible.
+
+The construction takes place in three stages: first, an
+acoustic field with high-frequency energy propagating along an
+incoming ray family is identified as the causal solution of an inhomogeneous boundary value problem {\em locally and on
+  the inside}. The boundary value is simply the restriction of the
+acoustic field to the surface. Second, the same causal solution is
+identified as the causal solution of an acoustic problem with source,
+singular on the surface in the fashion of equation
+\ref{eqn:awepm}. The source amplitude is proportional to 
+the velocity field sampled on the surface. Third, this identification is used twice, first to
+propagate the receiver data backwards in time as the solution of an
+anti-causal problem,  the velocity field of which is sampled on the
+source surface; then the causal problem is solved with the
+source built from the velocity samples. This solution must be
+identical to the field generating the data, along the ray fields
+carrying all of the high-frequency energy. Therefore the reconstructed
+source is an approximate solution of the problem \ref{eqn:esis} for
+$\alpha=0$.
+
+%For the bulk of this section, the source and receiver surfaces are
+%arbitrary smooth surfaces, as there is no conceptual penalty in
+%dropping the assumption that they are coordinate surfaces.
+
+\subsection{Field localization}
+The first task is to view an acoustic field locally as a solution of a
+boundary value problem with prescribed pressure on the boundary of a
+domain, as is the case in the photoacoustic tomography configuration.
+
+In this subsection $(p,\bv)$ is a solution of the homogeneous acoustic system
 \begin{eqnarray}
-\label{eqn:revawe}
-  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
-  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
-  \tilde{p}^- & =& 0,  \mbox{ for } t \gg 0\\ 
-  \tilde{\bv}^- & = & 0 \mbox{ for } t \gg 0\\
-  P_r\tilde{p}^- &=& \chi P_rp^+ . 
+\label{eqn:awebase}
+  \frac{1}{\kappa}\frac{\partial p}{\partial t} & = & - \nabla \cdot \bv, \nonumber \\
+  \rho\frac{\partial \bv}{\partial t} & = & - \nabla p,
 \end{eqnarray}
-That is, $\tilde{p}^-$ has the same boundary value on $z=z_r$ as
-$p^+$, except for low-frequency residue that is muted by
-$\chi$. Therefore
-$p^+ \approx \tilde{p}^-, \bv^+ \approx \tilde{\bv}^-$ near
-$z=z_r$. Since the right-hand sides in the system \ref{eqn:awepm} are
-singular only on $z=z_s$, and the high-frequency components of
-$(p^+,\bv^+)$ are carried by downgoing rays, these differ negligibly
-from the the high-frequency components of
-$(\tilde{p}^-,\tilde{\bv}^-)$ in the space-time slab $z_s<z<z_r$, and
-the approximation holds throughout this region. In particular
-$P_sv^+_z \approx P_s \tilde{v}^-_z$. In view of the relation
-\ref{eqn:tracejump10},
-\begin{equation}
-  \label{eqn:tildevtohsubs}
-  -2P_s\tilde{v}^-_z \approx h_s,
-\end{equation}
-so solution
-of the system \ref{eqn:revawe} followed by restriction to $z=z_s$ and
-multiplication by $-2$ 
-approximately inverts the map $S^+_{z_s,z_r}: h_s \mapsto P_rp^+$.
-
-Next observe that in view of the relation \ref{eqn:tracejump20}, and
-the downgoing nature of the ray system carrying the high frequency
-energy in $(p^+,\bv^+)$, the field $(\tilde{p}^-,\tilde{\bv}^-)$ is
-actually the restriction to $z<z_r$ of the anti-causal solution of \ref{eqn:awepm}
-with $z_s$ replaced by $z_r$, zero constitutive defect, and vertical
-load given by the jump in pressure at $z=z_r$ - for this field, use
-the same notation. Continuity of vertical
-velocity $\tilde{v}^-_z$ at $z=z_r$ implies that the vertical load is
-\[
-  f_r = -[\tilde{p}^-]|_{z=z_r} =-(\lim_{z\rightarrow
-    z_r^+}\tilde{p}^- - \lim_{z\rightarrow
-    z_r^-}\tilde{p}^-)
-\]
-\[
-  \approx 2 P_r \tilde{p}^- = 2 P_r p^+
-\]
-(from the definition \ref{eqn:defsamp}, $P_r$ is the limit from the
-left). Thus
-\[
-  P_s \tilde{v^-_z} \approx V^-_{z_r,z_s}(2 P_rp^+) \approx
-  2V^-_{z_r,z_s}S^+_{z_r,z_s}h_s.
-\]
-so
-\[
-  h_s \approx -2 P_s v^+_z \approx -2 P_s \tilde{v}^-_z \approx
-  -4V^-_{z_r,z_s}S^+_{z_r,z_s}h_s
-\]
-Combine this observation with \ref{eqn:tildevtohsubs} to obtain
-\[
- -4  V^-_{z_r,z_s} S^+_{z_s,z_r}  \approx  I,
-\]
-This relation combines with the identity \ref{eqn:trtrcomp} to
-yield the first main result of this section:
+The source surface $\Sigma$ is given a choice of unit normal field ${\bf
+  n}$. Rays carrying
+high-frequency energy in $(p,\bv)$ pass through $\Sigma$, and
+intersect $\Sigma$ only at times
+$>0$. Also, the velocity
+vectors of all such rays have a negative ${\bf n}$ component where
+they cross $\Sigma$. Construct a time interval $[0,\tau]$ and
+a region in space $\Omega^+$ containing $\Sigma$ so that $\Sigma$ forms
+part of the boundary $\partial \Omega^+$ of $\Omega^+$, ${\bf n}$ is the
+outward unit normal to $\Omega^+$ on $\Sigma$, and the energetic
+rays cross $\partial \Omega^+$ {\em only} at
+points of $\Sigma$ in the time interval $[0,\tau]$ - that is, in time
+less than $\tau$ the rays do not reach other parts of $\partial
+\Omega^+$. Note that the rays are {\em incoming} to $\Omega^+$ in the terminology
+introduced earlier. See Figure \ref{fig:omegarays}.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.8\textwidth]{omegarays.pdf}
+  \caption{Localization near crossing points of rays. $\Sigma$ is
+    source surface.  Dashed line is part of a ray carrying high
+    frequency energy. Ray velocity vector has negative component in
+    direction of unit normal ${\bf n}$ (outward, in chosen
+    orientation). $\Omega^+$ contains all ray segments for times
+    $0 \le t \le \tau$ (green arrow). Rays intersect boundary
+    $\partial \Omega^+$ only in $\Sigma$ for this time
+    range. $\Omega^-$ lies on outward side of $\Sigma$, which is part
+    of its boundary. The domain $\Omega$ is the union of $\Omega^+$
+    and $\Omega^-$.}
+  \label{fig:omegarays}
+\end{figure}
+
+Consequent to this ray geometry, conclude that the field $(p,\bv)$  is asymptotically negligible both
+throughout $\Omega^+$ at $t=0$ and at all points of 
+$\partial \Omega^+$ except for $\Sigma$, for $0\le t \le \tau$. So $(p,\bv)$ differs
+negligibly from the solution of the boundary value problem 
 \begin{eqnarray}
-  \label{eqn:approxinv}
-  (V^+_{z_s,z_r})^T S^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
-  (S^+_{z_s,z_r})^T V^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
-  V^+_{z_s,z_r} (S^+_{z_s,z_r})^T & \approx & \frac{1}{4}I, \nonumber\\
-  S^+_{z_s,z_r} (V^+_{z_s,z_r})^T & \approx & \frac{1}{4}I.
-\end{eqnarray}.
-The second equation is simply the transpose of the first, and the
-last two follow by by an exactly analogous argument using time
-reversal and interchange of the roles of $z_s$ and$z_r$.
-
-The conclusion is significant enough to merit restating in English:
-provided that high-frequency energy in the various fields is carried
-along downgoing ray fields, the transpose of $V^+$ is an approximate
-inverse to $S^+$, modulo a factor of 4. To recover the pressure source
-$h_s$ generating a pressure gather $P_rp$ at $z=z_r$, multiply the
-latter by -2, then apply the transpose of $V^+_{z_s,z_r}$ to this
-gather, reading out a vertical velocity field at $z=z_s$. Multiply
-again by -2 and you have a high-frequency approximation to $h_s$.
-
-It follows from the adjoint state method (see Appendix A for details) that
-\begin{equation}
-  \label{eqn:sadj1}
-  ({\cal S}^{\pm}_{z_s, z_r})^T = -{\cal S}^{\mp}_{z_r,z_s}
-\end{equation}
-
-Define $R$ to be the {\em time-reversal operator} on functions of
-space-time, $Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em
-  acoustic field time-reversal operator}
-\begin{equation}
-  \label{eqn:trdef}
-  {\cal R} \left(
-    \begin{array}{c}
-      p\\
-      \bv
-    \end{array}
-  \right) =
-  \left(
-    \begin{array}{c}
-      Rp\\
-      -R\bv
-    \end{array}
-  \right)
-\end{equation}
-Then 
-\begin{equation}
-  \label{eqn:trsadj}
-  {\cal R}{\cal S}^{\mp} = -{\cal S}^{\pm}_{z_r,z_s}{\cal R}
-\end{equation}
-Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj1} and \ref{eqn:trsadj} imply that
-\begin{equation} 
-  \label{eqn:trtr}
- ({\cal S}^{\pm}_{z_s,z_r})^T = {\cal R}{\cal S}_{z_r,z_s}^{\pm}{\cal R}=
- -{\cal S}^{\mp}_{z_r,z_s}.
-\end{equation}
-The relation \ref{eqn:trtr} implies that
+\label{eqn:aweloc}
+  \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} & = & - \nabla \cdot \tilde{\bv}, \nonumber \\
+  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla
+                                                    \tilde{p}, \\
+  \tilde{p}(\bx,t) & = & p(\bx,t) \mbox{ for } \bx \in \Sigma, 0 \le t \le 
+  \tau,\\
+  \tilde{p}(\bx,t) & = & 0 \mbox{ for } \bx \in \partial \Omega^+ \setminus \Sigma, 0 \le t \le
+ \tau.
+\end{eqnarray}
+
+\subsection{Source representation}
+
+The fields involved in the surface source extension are causal and
+source-generated, rather than solutions of boundary value problems
+such as the system \ref{eqn:aweloc}. To transition to a source
+representation, begin by extending the domain $\Omega^+$, lying on the
+incoming side of the surface $\Sigma$, to a full neighborhood $\Omega$
+of sigma
+by adding a domain $\Omega^-$ on the outside (see Figure
+\ref{fig:omegarays}). Once again, construct $\Omega^-$ so that none of
+the rays carrying high-frequency energy cross the boundary of
+$\Omega^-$ except at $\Sigma$. Extend the field $(\tilde{p},\tilde{\bv})$ to the
+full neigborhood $\Omega$ by solving the system \ref{eqn:aweloc} with
+$\Omega^+$ replaced by $\Omega^-$. Thus the extended
+field is pieced together out of causal fields solving the boundary value
+problem on either side of the boudary, with {\em the same} boundary
+condition, that is, the trace of $p$ on $\Sigma$. Thus the extended
+field is continuous in $p$. it is not however continuous in
+$\bv$, as will be seen shortly. More precisely, the normal component
+$\bv\cdot {\bf n}$
+has a discontinuity, while the tangential components are continuous.
+The jump in
+$\bv\cdot {\bf n}$ contributes a delta function to the gradient: as a
+field in the whole neigborhood $\Omega$, $(\tilde{p},\tilde{\bv})$
+solves
 \begin{eqnarray}
-  (S^{\pm}_{z_s,z_r})^T &=& -S^{\mp}_{z_r,z_s} \nonumber\\
-                        &=& R S^{\pm}_{z_r,z_s}R, \nonumber\\
-    (V^{\pm}_{z_s,z_r})^T &=& -V^{\mp}_{z_r,z_s} \nonumber\\
-                        &=& R V^{\pm}_{z_r,z_s}R.
-                            \label{eqn:trtrcomp}
+\label{eqn:awebase}
+  \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} & = & [\bv \cdot {\bf n}]_{\Sigma}\delta_{\Sigma}
+                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
+  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p},\\
+  \tilde{p}(\bx,t) & = & 0 \mbox{ for } \bx \in \partial \Omega, 0 \le t \le 
+  \tau,\\
+  \tilde{p} & = & 0 \mbox{ for } \bx \in \Omega^+, t , \\
+  \tilde{\bv} & = & {\bf 0} \mbox{ for } \bx \in \Omega^+, t = 0.       
 \end{eqnarray}
+in which $\delta_{\Sigma}$ is the delta function on $\Sigma$. The
+square bracket denotes the jump in the quantity enclosed across
+$\Sigma$, from the outside to the inside. Since ${\bf n}$ points from
+the inside to the outside, this convention contributes another minus
+sign.
+
+This representation is even better than it looks: since the ray
+families in $\Omega^{\pm}$ cross the boundary only at $\Sigma$ within
+the prescribed time interval, the
+solution changes only by an asymptotically negligible amount in
+$\Omega$ for $0 \le t \le \tau$ if the boundary condition is removed
+and the system is solved for all of space-time.
+
+There is a downside, namely the apparent necessity of constructing the
+solution in $\Omega^-$ so that the jump of $\bv \cdot {\bf n}$ may be
+computed. That is actually not necessary, however.
+This is best seen by analyzing a generic geometric optics
+component associated to the phase function $\psi^+$ in $\Omega^+$ and its family of
+rays. The acoustic field is assembled by a sum (in the continuum, an
+integral) over such solutions. A particular summand looks like
+\begin{equation}
+  \label{eqn:go}
+   a(\bx)e^{i\omega (t - \psi(\bx))}.
+\end{equation}
+for $\bx \in \Omega^+$.
+The phase $\psi$ solves the eikonal equation, the amplitude $a$ the
+related transport equation. The restriction of $p$
+to $\Sigma$ determines $\psi$ (and $a$) restricted to $\Sigma$, essentially by
+Fourier synthesis. Since the eikonal
+equation is even in $\nabla \psi$,  and $\psi$ restricted to $\Sigma$
+determines the tangential derivatives of $\psi$, the normal component
+of $\nabla \psi$ is determined up to sign by the boundary values of
+$p$. The spatial component of a ray $X(t)$ associated to $\psi$ solves
+the ordinary differential equation
+\begin{equation}
+  \label{eqn:ray}
+  \frac{dX}{dt} = \nabla \psi(X).
+\end{equation}
+The normal component of this ray must be negative (in
+$\Omega^+$) - this is the incoming condition, and guarantees that the
+ray does not intersect $t=0$ in $\Omega^+$ near $\Sigma$, justifying the initial
+condition in \ref{eqn:aweloc}. Similarly, the normal component must be
+positive in $\Omega^-$ near $\Sigma$. But both must have the same magnitude, since
+$\psi$ solves the eikonal equation - so they are negatives of each
+other. For causal solutions such as $(\tilde{p},\tilde{\bv})$, the
+Newton's law equation in \ref{eqn:aweloc} is equivalent to
+\begin{equation}\label{eqn:newt}
+  \tilde{\bv}(\bx,t) = -\frac{1}{\rho}\int_{-\infty}^t ds \nabla \tilde{p}(\bx,s),
+\end{equation}
+except at $\Sigma$.
+it follows that the normal components of $\tilde{\bv}$ have limits at
+point of $\Sigma$ 
+that are opposite in sign: for $\bx \in \Sigma$,
+\[
+  \lim_{\zeta \rightarrow 0} (\tilde{\bv}(\bx,t) \cdot {\bf
+    n}(\bx+\zeta {\bf n})) = 
+-\lim_{\zeta \rightarrow 0} (\tilde{\bv}(\bx,t) \cdot {\bf
+  n}(\bx -\zeta {\bf n}))
+\]
+whence
+\begin{equation}
+[\tilde{\bv} \cdot {\bf n}] = 2  \label{eqn:vjump} \lim_{\zeta
+  \rightarrow 0} (\tilde{\bv}(\bx,t) \cdot {\bf n}(\bx -\zeta {\bf
+  n}))
+\approx 2 \bv \cdot {\bf n}
+\end{equation}
+That is, at least over the short time interval $0 \le t \le \tau$,
+$(p,\bv)$ is asymptotically approximated in $\Omega^+$ by
+$(\tilde{p},\tilde{\bv})$, which solves
+\begin{eqnarray}
+\label{eqn:awesrc}
+  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & f \delta_{\Sigma}
+                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
+  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p},\\
+  \tilde{p} & = & 0 \mbox{ for } \bx \in \Omega, t=0, \\
+  \tilde{\bv} & = & {\bf 0} \mbox{ for } \bx \in \Omega, t = 0.       
+\end{eqnarray}
+with $f = 2 \bv \cdot {\bf n}$ on $\Sigma$.
+
+This result appears limited by the short time assumption (on $\tau$)
+and the corresponding spatial localization to $\Omega^+$. However
+neither of these apparent limitations is real. The short time
+assumption may be relaxed by windowing the fields to short time
+windows then adding up the results. The approximating fields
+$(\tilde{p}, \tilde{\bv})$ are themselves approximated by ray
+theory solutions. The assumptions on rays crossing source and receiver
+surfaces are global, and extend the approximation \ref{eqn:awesrc}
+globally. Thus the last two conditions may be replaced by the basic
+causality condition
+\begin{eqnarray}
+  \label{eqn:awesrcmod}
+  ... & = & ...\\
+  \tilde{p} & = & 0 \mbox{ for }  t \ll 0, \\
+  \tilde{\bv} & = & {\bf 0} \mbox{ for } t \ll 0.       
+\end{eqnarray}
+The effect of this change on the receiver surface traces of $\tilde{p}$ and
+$\tilde{\bv}$ is negligible.
+
+An example has already appeared in Figure \ref{fig:dsrcphh0}. In this
+example, the field $(p, \bv)$ is the point source solution with the
+point source depicted in Figure \ref{fig:olensrays0}, located at
+$(3500, 3500)$ m. The role of $\Sigma$ is played by the source surface
+$z=z_s=3000$ m, and near this surface the field solves the source-free
+system \ref{eqn:awebase}. The ``inside'' in this example is the region
+$z > z_s$, and the outward unit normal is therefore ${\bf n}=(0,0,-1)$.
+Thus $(p,\bv)$ should be asymptotically close to the solution
+$(\tilde{p},\tilde{\bv})$ of the
+system \ref{eqn:awesrc} in the region $z< z_s$, at least at locations
+connected by the ray field shown in Figure \ref{fig:olenrays0} to the
+source surface, provided that the source amplitude $f$ is chosen $=
+-2v_z$ on $\Sigma=\{z=z_s\}$. Figure \ref{fig:dsrcphh0} displays exactly this $f$, and
+\ref{fig:dfwdplh0} the traces of $\tilde{p}$ on the
+  receiver surface $z=z_r=1000$ m, which satisfies the ray condition. The extremely
+close match between Figures \ref{fig:drecplh0} (traces of $p$) and
+\ref{fig:dfwdplh0} is consistent with the conclusion reached here.
+
+\subsection{Approximate Inversion}
+
+The time-reversal construction can now be translated into the context
+of seismic surface source inversion. Assume that $(p,\bv)$ is a
+solution of the free-space acoustic system \ref{eqn:awebase}, with
+high-frequency energy carried by a ray family that satisfies the
+criteria described earlier: incoming at the source surface $\Sigma_s$
+with normal field ${\bf n}_s$, outging at the receiver surface
+$\Sigma_r$ with normal field ${\bf n}_r$, and every ray crossing
+$\Sigma_r$ meets $\Sigma_s$. The goal of this subsection is to
+describe a procedure that constructs a source field $h_s$ on
+$\Sigma_s$ so that $Sh_s \approx P_rp$, thus approximately inverting
+the modeling operator $S$.
+
+The sampled fields $(P_rp,P_r\bv)$ on $\Sigma_r$
+is negligible for large $t$. Due the outgoing assumption,
+$(p,\bv)$ are negligible for large $t$ near $\Sigma_r$. The
+integrated Newton's law \ref{eqn:newt} acquires a minus sign if the
+integration from $-\infty$ to $t$ is replaced by integration from $t$
+to $\infty$. Apply the construction of the last subsection in reverse
+time, to conclude
+that $(p,\bv)$ is approximated by $(\tilde{p},\tilde{\bv})$, where the
+latter solves the time-reversed problem
+\begin{eqnarray}
+\label{eqn:awerec}
+  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & h_r \delta_{\Sigma_r}
+                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
+  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p},\\
+  \tilde{p} & = & 0 \mbox{ for } t\gg 0, \\
+  \tilde{\bv} & = & {\bf 0} \mbox{ for } t \gg 0.       
+\end{eqnarray}
+with $h_r = -2P_r(\bv \cdot {\bf n}_r)$.
+
+By construction, high-frequency energy in $(\tilde{p},\tilde{\bv})$ is
+carried by the same ray family as for $(p,\bv)$. Every ray passes over
+$\Sigma_s$ and is incoming there. Since $(p,\bv)$ vanishes for $t \ll
+0$ near $\Sigma_s$, $(\tilde{p},\tilde{\bv})$ is negligible there,  So
+the construction detailed in the last two sections is applicable
+again, and $(\tilde{p},\tilde{\bv})$ is approximated by the solution
+of 
+\begin{eqnarray}
+\label{eqn:awesrc}
+  \frac{1}{\kappa}\frac{\partial\hat{p}}{\partial t} & = & h_s \delta_{\Sigma_s}
+                                                      - \nabla \cdot \hat{\bv}, \nonumber \\
+  \rho\frac{\partial \hat{\bv}}{\partial t} & = & - \nabla \hat{p},\\
+  \hat{p} & = & 0 \mbox{ for } t\ll 0, \\
+  \hat{\bv} & = & {\bf 0} \mbox{ for } t \ll 0.       
+\end{eqnarray}
+with $h_s = 2P_s(\tilde{\bv} \cdot {\bf n}_r) \approx 2P_s(\bv \cdot
+{\bf n}_r)$.
+
+\subsection{Summary}
+The surface source version of the time reversal construction involves three steps: given
+pressure data $P_rp$ on the receiver surface $\Sigma_r$ for a solution
+$(p,\bv)$ of the system \ref{eqn:awebase} satisfying the ray conditions,
+\begin{itemize}
+\item[1. ] create a source $h_r$ on $\Sigma_r$
+  by multiplying the corresponding normal velocity trace by $-2$: $h_r = -2P_r(\bv
+  \cdot {\bf n}_r)$;
+\item[2. ] propagate $h_r$ backwards in time by solving the system
+  \ref{eqn:awerec} for the acoustic field $(\tilde{p},\tilde{\bv})$;
+\item[3. ] create a source $h_s$ on $\Sigma_s$ by  multiplying tne
+  normal velocity trace of $(\tilde{p},\tilde{\bv})$ by 2: $h_s = 2P_s(\tilde{\bv}
+  \cdot {\bf n}_s)$.
+\end{itemize}
+Then $P_r p \approx Sh_s$.
+
+The surface source representaion part of the construction shows that
+under the assumed ray conditions, a solution of \ref{eqn:awebase} is
+asymptotically the same near the receive surface $\Sigma_r$ as the
+solution of the system \ref{eqn:awepm} with a suitable source
+$\tilde{f}_s$ (and zero velocity source, $h_s=0$. Therefore this
+construction also inverts the modeling operator $S$ in the sense that
+$Sh_s\approx S\tilde{h_s}$ if $P_rp \approx S\tilde{h_s}$. That is,
+this construction produces an asymptotic rightt inverse of $S$: it
+does not necessily (approximately) recover the source that generates
+the data, but recovers some source that generates the data.
+
+\subsection{Example: Lens}
+The first example uses the acoustic lens model (Figure
+\ref{fig:olensrays0}, with $\Sigma_s$ = horizontal line at depth 
+3000 m, and $\Sigma_r$ = horizontal line at depth 1000 m. The pressure
+data to be inverted is the sampling of the pressure field on
+$\Sigma_r$, generated by the point source shown in the figure, at
+$(3500, 3500)$ m (Figure \ref{fig:drecplh0}. However, I will invert
+this data in a homogeneous medium with $\kappa \equiv 4$ GPa. These
+choice illustrate two facts about surface source inversion and the
+approximate inversion based on time reversal. First, the success of the inversion demostrates the
+insensitivity of the time reversal method to ray multipathing
+(triplication), evident in the data (Figure \ref{fig:drecplh0}).
+%s respectively
+Second, it is entirely possible to invert data in a material model
+other than the one in which it was produced (in the case of synthetic
+data, of course). This
+capability is critically important in the application of the
+approximate inversion in nonlinear inversion, where the early iterations
+involve solution of the source estimation problem \ref{eqn:esis} at
+(possibly very) wrong material models ${\bf c}$. Successful extension
+methods maintain data fit throughout the course of the inversion.
+%The source time dependence is again a [1, 2.5,
+%10, 12.5] Hz trapezoidal bandpass filter.
+
+As noted earlier (and visible in Figure \ref{fig:olensrays0}), the acoustic field in this example satisfies the ray
+conditions on which the time reversal construction is based: The rays
+are incoming at $\Sigma_s$, outgoing at $\Sigma_r$ and all rays
+carrying significant energy at points of $\Sigma_r$ pass over
+$\Sigma_s$. The outward unit normal at $\Sigma_s$ is ${\bf n}_s ={\bf
+  e}_z=(0,1)$; at $\Sigma_r$, it's ${\bf n}_r =-{\bf
+  e}_z=(0,-1)$. So the source on the receiver surface $\Sigma_r$, input to reverse time propagation
+(system \ref{eqn:awerec}, generating the fields $(\tilde{p},\tilde{\bv})$) is
+$h_r=2P_rv_z$ (the field $P_rv_z$ is depicted in Figure \ref{fig:dfwdvzlh0}). After reverse time propagation , the approximate
+inversion $\tilde{h}_s=2P_s\tilde{v}_z$ is shown in Figure
+\ref{dinvhslh0}. It differs considerably from the source field used to
+generate the data (Figure \ref{fig:dsrcphh0}, however when used as the
+source in forward modeling (system \ref{eqn:awesrc}) {\em with the
+  same material parameters used in the inversion}, it generates a
+pressure gather on $\Sigma_r$ (Figure \ref{fig:drerecplh0}) very
+closely 
+approximating the input gather (Figure \ref{fig:dfwdplh0}). The
+difference is plotted on the same color scale in Figure
+\ref{fig:ddiffrecplh0}.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:dfwdvzlh0}]{\includegraphics[width=0.45\textwidth]{dfwdvzlh0.pdf}}
+  \subfloat[\label{fig:dinvhslh0}]{\includegraphics[width=0.45\textwidth]{dinvhslh0.pdf}}
+  \caption{(a) Normal velocity data corresponding to pressure data in
+    Figure \ref{fig:dfwdplh0}. (b) Source function resulting from time-reversal inversion
+    of pressure data in Figure \ref{fig:dfwdplh0}. Propagation in homogeneous
+    bulk modulus model (different from the model used to generate the
+    data!). Plotted with same color scale as Figure \ref{fig:dsrcphh0}.}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:drerecplh0}]{\includegraphics[width=0.45\textwidth]{drerecplh0.pdf}}
+  \subfloat[\label{fig:ddiffrecplh0}]{\includegraphics[width=0.45\textwidth]{ddiffrecplh0.pdf}}
+  \caption{(a) Resimulated data: pressure gather generated from
+    inverted source (Figure \ref{fig:dinvhslh0}, using same
+    homogeneous bulk modulus as used in inversion. (b) Difference
+    between data depicted in Figure \ref{fig:drerecplh0} and in Figure \ref{fig:dfwdplh0}.
+    Both (a) and (b) plotted with same color scale as Figure \ref{fig:dfwdplh0}.}
+\end{figure}
+
+\subsection{Example: Diving Wave}
+
+The second example is based on the diving wave model (Figure
+\ref{fig:ooplrays0}). In this example, unlike the previous one, the source
+and receiver surfaces do not bound the propagating region, even in the
+vertical direction. Thus this exercise differs even more than the
+first from time reversal in photoacoustic tomography. Inspection of
+the blue part of the ray field in Figure \ref{fig:ooplrays0} reveals that
+the rays are incoming at the source surface, and outgoing at the receiver surface, if the
+outward normal is chosen as $-{\bf e}_z$. The mute shown in Figure
+\ref{fig:ndwmute0} eliminates the rest of the ray field. What remains
+satisfies the condition that every ray passing the receiver surface
+originates in the source surface.
+
+In this example, the inversion is performed with the same model as
+used to generate the data. Note that sufficiently small changes in the model would
+result in similar ray configurations, so time reversal inversion could
+proceed as is shown here. However sufficiently large changes in the
+bulk modulus model could result in fundamental changes in the ray
+field: for instance, change to a homogenous model would completely
+eliminate the diving rays and the corresponding data component, thus
+making an inversion of the type presented here impossible.
+
+The inversion proceeds otherwise as before. The source $h_r=-2 P_rv_z$
+on $\Sigma_r$ is built from the normal component of velocity, shown in
+Figure \ref{fig:ndwdatavz0}. Reverse time propagation to create the
+fields $(\tilde{p},\tilde{\bv})$, extraction of
+the normal component of velocity (note the sign difference with the
+previous example!) and scaling by the factor 2 produces the inverted
+source shown in Figure \ref{fig:ndwdataresrcplbulksm0}. Forward
+propagation yields the pressure gather shown in Figure
+\ref{fig:ndwdataresimplbulksm0}, which closely matches the original
+diving wave pressure data (Figure \ref{fig:ndwdata0}). The difference
+is displayed on the same color scale in Figure
+\ref{fig:ndwdatarediffplbulksm0}.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:ndwdatavz0}]{\includegraphics[width=0.45\textwidth]{ndwdatavz0.pdf}}
+  \subfloat[\label{fig:ndwdataresrcplbulksm0}]{\includegraphics[width=0.45\textwidth]{ndwdataresrcplbulksm0.pdf}}
+  \caption{(a) Normal velocity data corresponding to pressure data in
+    Figure \ref{fig:ndwdata0}. (b) Source function resulting from time-reversal inversion
+    of pressure data in Figure \ref{fig:ndwdata0}. Propagation in
+    same model (Figure \ref{fig:ooplrays0}) as used to generate the data.}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:ndwdataresimplbulksm0}]{\includegraphics[width=0.45\textwidth]{ndwdataresimplbulksm0.pdf}}
+  \subfloat[\label{fig:ndwdatarediffplbulksm0}]{\includegraphics[width=0.45\textwidth]{ndwdatarediffplbulksm0.pdf}}
+  \caption{(a) Resimulated data: pressure gather generated from
+    inverted source (Figure \ref{fig: ndwdataresrcplbulksm0}, using same
+    bulk modulus (Figure \ref{fig:ooplrays0}) as used in data
+    generation and inversion. (b) Difference
+    between data depicted in Figure \ref{fig:ndwdataresimplbulksm0} and in Figure \ref{fig:ndwdata0}.
+    Both (a) and (b) plotted with same color scale as Figure \ref{fig:ndwdata0}.}
+\end{figure}
 

### 2023-01-25 repro/papers/GEM22/submitR2/response.tex
pre-sub

COMMENTS FOR THE AUTHOR:

Dear Author, 
Thanks for this revised version of your  study. Both reviewers acknowledge the significant improvement of this new version. They still point some typos, and comments/questions I think would be beneficial to consider in a further revised version of your study before we accept it for publication. See these comments appended below. 
Thanks a lot for your work and this interesting study, 
Best regards,
Ludovic Métivier.

=============================================================================================================================================================================================================================

Reviewer #1: The manuscript has been considerably revised. The revised version is much better. The early introduction of the examples and the overview of the time reversal method significantly improve the readability. I would like to recommend the manuscript for publication. 
I have, however, two questions and I have detected typos (I am afraid that there are many more).

Comment 1. what are the blue and red rays in Figure 3 ?
the blue rays start from the source, the red rays start from a point located just below the source. As a result, some red and blue rays are very close to each other.

Response: I have created a clearer ray diagram overlain on the bulk modulus color scale plot (Figure 3) and added another figure and some discussion to explain the ray geometry. The red and blue rays start at the same source point and are simply different parts of the same point source ray field. The diving wave example is similar to that depicted in Figure 4.20a in Sheriff and Geldart, {\em Exploration Seismology}. It features a homogeneous upper layer, a middle layer in which the velocity increases rapidly, and a homogeneous lower layer, with smooth transitions between the layers. As Sheriff and Geldart say, ``A velocity gradient in a layer that is substantially higher than the velocity gradient above the bed, as in fig. 4.20a, causes a very sharp increase in ray path curvature and a folding back of the time-distance curve. Such a triplication of branches of the time-distance curve (fig. 4.20c) is usually difficult to see because the later branches become lost in the later cycles of the earlier arrivals.'' To show clearly how this triplication happens, I've included a plot of arrival times in my introduction of this example (new Figure 4b). The part of the arrival time curve in red corresponds to the those red rays in the ray plot (Figure 3) that refract back to the receiver surface, and similarly for the blue rays. Note that the red branch is close enough to the direct arrival time that its chief effect is an apparent wavelet change in the direct arrival. Both red and blue branches are really diving wave arrival times, but only the waveform arrival corresponding to the blue branch can be isolated by a simple mute.

Comment: on page 23, I do not understand the following sentence:
The function a_0 is a frequency-independent envelope, equal to 1 where h is non-zero.

Response: This section skipped over a number of steps, which I have included in the revision. I believe that I have explained this idea.

typos:

page 2
certainluy
->
certainly

page 5
veloity
->
velocity

propblem
->
problem

tahnks 
->
thanks to

is proportion to 
->
is proportional to

page 6

slowiy
->
slowly

page 8
increasting
->
increasing

page 10

to generate a data traces

page 13

of sigma 
-> 
of $\Sigma$

page 16
representaion 
->
representation

page 21
too many s in Psshat{p}

page 22
Speclializing
->
Specializing

last equation: confusion between s and t
partial v partial t
the use of s is very confusing (it is the symbol of surface !)

page 23
first equation: confusion between s and t

SInce 
->
Since

check the sentence:
is asymptotic the pressure field of to an incoming solution

last equation: confusion between s and t

page 24
parenthesis in lambda(...

page 26
parenthesis in (tilde{p},tilde{v}

negligbly -> negligibly

defintiion -> definition

page 29
as it need  
->
as it needs

to ne discussed
->
to be discussed

page 31
parenthesis in (equations 28,29

page 33
typo in: and equation 48, .

page 34
computationa
->
computational

page 36
ismeasured 
->
is measured

page 37
parenthesis in (Figure 15b

page 39
Madacascar
->
Madagascar

=============================================================================================================================================================================================================================

Reviewer #2: Overall the paper is much improved and I now found it relatively easy to follow and understand.  I particularly liked the 'steps' introduced as a way to summarize what you've done so far and to give the reader further help in actually implementing these ideas.  (I also liked the roadmap section at the beginning.)  I also found the new figures much easier to compare and understand.

There are a few areas where I think things could still be somewhat clearer, which are:

Comment 1. In the introduction p 3 line 44: Perhaps say in a few words why you isolate the diving rays.  Also later when you show your ray picture (Figure 3) with the diving waves, first I think I see red rays beside the blue ones, and second just out of curiosity, what is your definition of diving wave, because I see things that turn back to the surface that you do not deem diving rays.

Response: You are of course correct that the ``things that turn back'' are another diving wave branch. I've considerably revised the figure, its caption, and the discussion in the text, intending to answer all of these concerns.

Comment 2. You relate what you are doing to photo-acoustic tomography, which is neat and certainly was the motivation.  As I read the paper though, I can't help thinking that what you are doing is a lot like running RTM back to the source.  I expect that other readers may also wonder what is really fundamentally different between RTM and this approach, so I think that adding something to make this comparison explicit would be helpful.

Response: this is a very good question. I have added a paragraph describing the role of the D-to-N map in ``true amplitude RTM'', aka approximate Born inversion. In fact it's quite similar to the role described in this paper for the inverse surface source problem. In both cases the core idea is to backpropagate the actual data field. In the present case, that is a transmitted field; in true-amplitude migration, it's the scatterered field. Reconstructing these from the data is the key to inversion in both cases, and the D-to-N map is the key to reconstruction.

Comment 3. In figure 2 and the lens example in general, why is the source you used not on the source surface?  Perhaps I have misunderstood the meaning of the source surface, and that could be clarified.  I guess that the source should be below that surface so the rays come into it from below and then leave it 'from above' and enter the receiver surface from below?   I am sure that one could phrase that much better than I have here, but as I was reading it was odd to see the source below what's called the source surface.

Response: I could have placed the source on the source surface. I created a parallel sequence of experiments for the lens example using that configuration. However a point source cannot be reconstructed, at least not with the simple transmission geometry used in this paper, since a lot of energy propagates away ``to China'' and is not available in the inversion. You can reconstruct a source that generates the data, but it isn't a point source - if the velocity is right, it is more or less a dip-filtered version of the point source. Instead, I chose to somewhat artificially create a source that can actually be reconstructed fairly accurately in this example: the rays carrying significant energy actually make it to the receiver surface within the time limit.

The diving wave example uses a point source on the source surface. You will notice that the reconstructed source (Figure 9b) appears to be a dip-filtered point source, regarded as a function on the source surface.

Comment 4, In Figure 3, there are some rays that look weird to me, at around (8000,1000) and (12000,1000) This is just a gradient so I am expecting more-or-less circular wavefronts.  I don't know that this really requires a comment or anything, but maybe a mention as to what is going on there.

Response: Figure 3 and surrounding discussion needed some tuning up. Note that the gradient-to-constant transition is smoothed, and the gradient is large enough to generate a triplication. Having homogeneous layers top and bottom really messes things up. Sheriff and Geldart, {\em Exploration Geophysics}, discussion of fig. 4.20a, have a nice discussion of this behaviour. There are segments of circles in there, but overall these rays are not close to circles. I think the description now in the text, and the inclusion of the arrival time plot, will make things clear enough.

Comment 5. p 11 line 32, what do you mean by 'must be localized' in which variables?

Response: The PAT construction does not apply to the surface source inversion problem, since there is no bounding surface on all of which the pressure field is measured/known. I get around this by localizing in space-time. For short-time propagation, the signal does not propagate far from the source surface, so an artificial bounding surface can be added on which the field is automatically zero. So the field can be viewed as the solution of a local pressure boundary value problem, for a little while. Then it can be converted to a local solution of a problem with surface source, as described. Since the field vanishes near the artificial boundary, you can simply get rid of it, and if the pressure field on the boundary is only non-zero for a short time, that's it - just let it propagate in space time. Finally, any pressure distribution on the source surface can be broken up in to a sum of terms, each only non-zero for a short interval. Apply the short-time-window construction to each summand, and add them up.

Comment 6. Figure 6b has quite a different source than that shown in Figure 2b.  I think that when you talk about Figure 6 it would be good to re-iterate why you are getting a different source here (I think velocity vs pressure, but remembering that many people only look at the pictures this will be confusing).

Response: I think this is actually explained in the caption of Figure 7 (formerly Figure 6), also in the text discussion. This is the result of inversion {\em in the wrong velocity}. The data (Figure 2a or 2c) is inverted here in a homogeneous model - so the source does not look at all like that (2b) used to generate the data in the lens model.

Typos: I am {\em very} grateful to both referees for taking the trouble to list many of the embarassingly extensive typos. I believe that I have fixed all of these, and some others which we won't mention. I only hope I have not created yet more...

A very-incomplete typo list (I do suggest getting out your grammatical fine-toothed comb and checking for more of these):

-Spelling of constraint in the abstract
-p 6 line 23 "is not be well-defined"
-p 13 line 24 there is a sigma without a \
-p 16 line 59 spelling of representation
-P 21 line 40 The sentence beginning "Differentiate" is missing some important part-of-speech for which I should have a name but don't, perhaps a subject?
-p26 line 28 missing )
-p26 line 48 sentence fragment beginning "To illustrate"
-p26 line 62 missing space in is measured

### 2023-01-25 repro/papers/GEM22/submitR2/paper.tex
pre-sub

--- 2022-10-25 repro/papers/GEM22/submitR1/paper.tex
@@
-\documentclass[georeport,12pt]{geophysics}
+\documentclass[12pt]{geophysics}
 
@@
 \newtheorem{example}{Example}
-\newtheorem{definition}{Defenition}
+\newtheorem{definition}{Definition}
 
@@
 \begin{abstract}
-Source extension is a reformulation of inverse problems in wave propagation, that at least in some cases leads to computationally tractable iterative solution methods.
-The core subproblem in all source extension methods is the solution of a linear inverse problem for a source (right hand side in a system of wave equations) through minimization of data error in the least squares sense with soft imposition of physical constraints on the source via an additive quadratic penalty.
-A variant of the time reversal method from photoacoustic tomography provides an approximate solution that can be used to precondition Krylov space iteration for rapid convergence to the solution of this subproblem.
-An acoustic 2D example for sources supported on a surface, with a soft contraint enforcing point support, illustrates the effectiveness of this preconditioner.
+Source extension is a reformulation of inverse problems in wave
+propagation, that at least in some cases leads to computationally
+tractable iterative solution methods. The core subproblem in all
+source extension methods is the solution of a linear inverse problem
+for a source (right hand side in a system of wave equations) through
+minimization of data error in the least squares sense with soft
+imposition of physical constraints on the source via an additive
+quadratic penalty. For an acoustic formulation for sources supported
+on a surface, with a soft contraint enforcing concentration at a
+point, a variant of the time reversal method from photoacoustic
+tomography provides an approximate solution. This approximate inverse
+can be used to precondition Krylov space iteration for rapid
+convergence to the solution of the core subproblem in this
+setting. Numerical examples illustrate the effectiveness of this preconditioner.
 \end{abstract}
@@
 acoustic field, $L[{\bf c}]$ is the wave operator of linear
-acoustodynamics, and the source and receiver surfaces are horizontal
-planes. Numerical examples in this setting suggest the
+acoustodynamics. Numerical examples in this setting suggest the
 effectiveness of this acceleration.
@@
 final time $t=t_{\rm max}$ Then the field can be (at least
-approximately) viewed as the backwards-in-time initial boundary value
+approximately) viewed as the solution of the backwards-in-time initial boundary value
 problem with zero final conditions at $t=t_{\rm max}$, and boundary
 values given by the measurements. Evolving the field backwards in time
-backwards to $t=0$ thus solves the problem. Except in special
+to $t=0$ thus solves the problem. Except in special
 circumstances, the pressure field never actually vanishes at finite
@@
 surface, and must cross the source and receiver surfaces
-transversally, that is, making non-zero angles. Also, the rays must
-approach source and receiver surfaces from one side or the other
-(``the inside''), so that locally the surfaces can be treated as
-boundaries of propagation domains. Specifically, rays must emanate
-from the source surface towards the inside (``incoming'' rays), and
-arrive at the receiver surface from the inside (``outgoing'' rays).
+transversally, that is, making non-zero angles. The surfaces must be
+oriented, with a choice of (``outward'') unit normal vector. The
+energetic rays must emanate
+from the source surface inwards (``incoming'' rays), and
+arrive at the receiver surface outwards (``outgoing'' rays). Then both surfaces
+can be treated locally as
+boundaries of propagation domains. 
 With these assumptions, reverse-time propagation of the receiver data
-closely approximates the acoustic fields near the source surface. Next
-observe that for a causal field with a surface pressure source (with a
-continuous pressure field across the surface), the source is
+closely approximates the acoustic fields near the source
+surface. Next, I show that for a causal field with a surface pressure source (with a
+continuous pressure field across the surface), the source amplitude is
 proportional to the jump in the (particle) velocity field. Moreover,
@@
 surface, so the jump is just twice the sampled value on the
-surface. Thus reverse-time propagation and reading off the velocity
+surface. Thus constructing a source from the velocity trace on the
+receiver surface, reverse-time propagation of this source, and reading off the velocity
 field on the source surface yields a source that reproduces the
 pressure data, within an asymptotically small error.
-%Preconditioning amounts to a change of
-%norm in the domain and range spaces of the modeling operator. In this
-%case, the modfied norms are weighted $L^2$, and the weight operators
-%map pressure to corresponding surface source on the source and
-%receiver planes.
 
@@
 \ref{eqn:esis} for penalty weight $\alpha=0$. To see how Krylov
-iteration might be accelerated, note that time-reverse propagation of
-the receiver pressure values may be represented by a time-reverse
-source propagation, with the source constructed from the velocity
-field at the receiver surface. Moreover, the reverse-time
-source-to-pressure propagation is the transpose (adjoint) of the
+iteration might be accelerated, I note that reverse-time propagation of
+the receiver surface source to pressure on the source surface is the transpose (adjoint) of the
 forward-time source-to-pressure propagation - this is a very simple
@@
 approximate inverse may be described as: conversion of pressure to source at the
-receiver surface, followed by the adjoint of the modeling operator,
+receiver surface, followed by application of the modeling operator adjoint,
 followed by the conversion of pressure to source at the source
@@
 receiver data, with the weight operators being the
-pressure-to-source operators. That this construction results in an
+pressure-to-source operators on the two surfaces. That this construction results in an
 approximate inverse means that the modeling operator is approximately
@@
 
-FInally, the $\alpha=0$ case is not sufficient: the penalty operator
+Finally, the $\alpha=0$ case is not sufficient: the penalty operator
 $A$ is an essential component of the nonlinear inverse propblem
@@
 photoacoustic tomography and other wave inverse problems
-\cite[]{Rachele:00,StefUhl:05}. \cite{HouSymes:EAGE16} demonstrated a
-very similar preconditioner for Least Squares Migration, also for its
-subsurface offset extension \cite[]{HouSymes:16}, motivated by
-\cite{tenKroode:12}. These constructions also involve the
-Dirichlet-to-Neumann operator. This concept also turns up in hidden
-form in the work of Yu Zhang and collaborators on true amplitude
+\cite[]{Rachele:00,StefUhl:05}. In this terminology, ``Dirichlet''
+refers to pressure restricted to a surface, and ``Neumann'' to the
+normal derivative of pressure on that surface. Since the latter is
+proportional to normal particle acceleration thanks the equations of motion,
+and normal particle velocity is proportion to source, the relation to the
+pressure-to-source map as described in this paper is clear. A number
+of authors have described preconditioners for Least Squares Migration
+with subsurface offset extension in which the Dirichlet-to-Neumann or
+pressure-to-source map plays a central role
+\cite[]{tenKroode:12,HouSymes:15,Herve2017}. It also turns up in hidden
+form in the work of Yu Zhang and collaborators on ``true amplitude''
 migration
@@
 
-%The obvious computation of the pressure-to-source map - prescribe the
-%pressure, solve the wave equation with this boundary condition, read
-%off the equivalent source - suffers from intrinsic numerical
-%inaccuracy. I suggest an alternative computationally feasible
-%approach, via economical short-distance wave propagation. Since the
-%map is symmetric only in an approximate, asymptotic sense, it must be
-%symmetrized for use as a Krylov preconditioner. I describe a
-%symmetrization procedure that requires no further wave computations
-%beyond those necessary to compute the action of the operator itself.
+The role of the pressure-to-source map in these constructions is quite
+similar to the role described here. Reverse Time Migration, in its
+basic form, broadcasts pressure traces into the propagation region,
+backwards in time. Where this backward propagated energy intersects
+the forward propagated Green's function, an imaging condition outputs
+an image amplitude map, which is not however an approximate inversion,
+as is widely understood, except in a crude kinematic sense. The
+various ``true amplitude'' or approximate least squares migrations,
+apply the pressure-to-source operator (either explicitly, or
+implicitly via switching boundary conditions) {\em before}
+reverse-time broadcasting. Since the data are presumed to be traces of
+the forward scattered (Born) pressure field, the backward propagated
+wavefield must approximate the forward scattered field,
+time-reversed. This reconstructed scattered field combines at
+scattering points with the Green's function to yield an approximate
+inversion for the materal parameter perturbations. Moreover, this
+approximate Born inversion is also adjoint to the Born scattering operator in weighted norms, one
+of the weight operators being the pressure-to-source map, hence an
+effective preconditioner for Krylov space methods for linear least
+squares inversion, that is, Least Squares Migration
+\cite[]{HouSymes:16}. Thus the present construction bears a
+considerable resemblance to these ``true-amplitude'' reverse-time
+migration methods.
 
@@
 that some important mathematical underpinnings are only
-sketched. I will treat the modeling operator $PL[{\bf c}]^{-1}$ as if
+sketched. The transport of high-frequency energy along rays is a
+crucial ingredient in the constructions explained here. Of course,
+such transport only occurs in material parameter models that vary
+slowiy in space on the wavelength scale, and even then only
+approximately. Rather than provide explicit
+smoothness constraints and asymptotic error estimates, I will use
+these concepts in an intuitive way, using ``approximate'' and
+``asymptotic'' and the symbol ``$\approx$'' to suggest asymptotically
+negligible errors not explicitly assessed, and refer to such errors as
+``negligible''.
+
+I will treat the modeling operator $PL[{\bf c}]^{-1}$ as if
 it mapped square integrable surface sources to square integrable sampled
@@
 finite-energy acoustic fields. Thus the modeling operator envisioned above
-is not be well-defined, strictly speaking.
-
+is not well-defined, strictly speaking.
 This phenomenon is related to the ill-posedness of wave equations as
@@
 finite energy data, or equivalent properties
-\cite{Payn:75,Symes:83,Lasi:86,LasLionsTrig:86,Lasi:87, BaoSy:91b}.
+\cite[]{Payn:75,Symes:83,Lasi:86,LasLionsTrig:86,Lasi:87, BaoSy:91b}.
 Elaboration of these mathematical details is beyond the scope of this
 paper, which aims instead to explore the algorithmic consequences of
-the mathematical structure implied by the nongrazing hypothesis.
+the mathematical structure implied by the incoming/outgoing ray constrants..
 
@@
 coordinate vector, and similarly for the 
-particle velocity $\bv = (v_x,v_y,v_z)$. For computational convenience, the
+particle velocity $\bv = (v_x,v_y,v_z)^T$. For computational convenience, the
 examples are two-dimensional.
@@
 \end{eqnarray}
+in which $\delta_{\Sigma_s}$ is the delta function on $\Sigma_s$, and
+similarly for $\Sigma_r$.
 
@@
 sampling the solution components at a receiver surface $\Sigma_r$ with
-(``outward'') unit normal field ${\bf n}_r$.
+(outward) unit normal field ${\bf n}_r$.
 $P_s$ and $P_r$ are the sampling operators on $\Sigma_s$ and
@@
 practice, sampling necessarily occurs at a discrete array of points
-(trace locations), and over a zone of finite extent. In this
+(trace locations). In this
 theoretical discussion, I will neglect the finite sample rate, and
@@
 Denote by $\Pi_i, i=0,1$ the projection on the first,respectively
-second, component of a vector in $\bR^2$. The forward modeling
+second, component of a vector in $\bR^2$:
+\begin{equation}
+  \label{eqn:pidef}
+  \Pi_0 (a,b)^T = a, \, \Pi_1(a,b)^T= b.
+\end{equation}
+The forward modeling
 operator from pressure source to pressure trace is
@@
 $\Sigma_s= \{z = z_s=3000 \mbox{ m}\}$ (orange horizontal line) m and
-a receiver surface $\Sigma_r= \{z=z_r=1000 \mbox{ m}\}$ (yellow
+receiver surface $\Sigma_r= \{z=z_r=1000 \mbox{ m}\}$ (yellow
 horizontal line).  The data to be used in this example results from a
@@
 receiver surface: evidently all of the rays involved cross the source
-and receiver surfaces transversally. Also, regarding ``inwards'' as
-being ``upwards'' at the source surface, ``downwards'' at the receiver
-surface, it is clear that the data is incoming at the source surface,
+and receiver surfaces transversally. The outward normal vectors are chosen as
+${\bf n}_s={\bf e}_z=(0,1)^T$ on the source surface and ${\bf
+  n}_r=-{\bf e}_z$ on the receiver surface.
+One sees from the Figure that the ray field carrying high frequency
+energy is incoming at the source surface,
 outgoing at the receiver surface.
@@
     generation at $(3500, 3500)$ m indicated with star. Overlain with rays from point
-    source to receiver surface.}
+    source to receiver surface. Outward normal points down at source
+    surface, up at receiver surface, so rays are incoming at source
+    surface, outgoing at receiver surface.}
   \label{fig:olensrays0}
@@
 
-Figure \ref{fig:dsrcphh0} displays an extended (pressure) source $h_s$, in the form of
+Figure \ref{fig:dhshh0} displays an extended (pressure) source $h_s$, in the form of
 traces in the spatial range 2000 m $ \le x_r \le $ 6000 m and time
@@
   \subfloat[\label{fig:drecplh0}]{\includegraphics[width=0.32\textwidth]{drecplh0.pdf}}
-  \subfloat[\label{fig:dsrcphh0}]{\includegraphics[width=0.32\textwidth]{dsrcphh0.pdf}}
+  \subfloat[\label{fig:dhshh0}]{\includegraphics[width=0.32\textwidth]{dhshh0.pdf}}
   \subfloat[\label{fig:dfwdplh0}]{\includegraphics[width=0.32\textwidth]{dfwdplh0.pdf}}
@@
 The second example is intended as a cartoon of long-offset node
-acquisition. It features a depth-dependent increasting
+acquisition. It features a depth-dependent increasing
 velocity. Figure \ref{fig:ooplrays0} shows bulk modulus field (once again, the
@@
 \begin{figure}
-  \includegraphics[width=\textwidth]{ooplrays0.pdf}
+  \includegraphics[width=\textwidth]{ooplrays0.png}
   \caption{Bulk modulus generating diving waves. Color scale unit is
@@
     generation at $(500, 10000)$ m indicated with star. Overlain with rays from point
-    source to receiver surface.}
+    source to receiver surface. Outward normal points up at both source
+    and receiver surfaces, so depicted rays are incoming at source
+    surface, outgoing at receiver surface. Significance of red
+    vs. blue coloring explained in text and caption of Figure \ref{fig:ooxtraysend}.}
   \label{fig:ooplrays0}
@@
 A diving wave arrival is clearly visible in Figure
-\ref{fig:ndwraw0}, which displays data traces over the position range $14000 \le
-x_r \le 20000$ m. The ray field connecting the point source to the
-receiver surface is plotted with the diving rays in blue, the rest in
-red. Note that at the source, the diving rays are incoming if the
+\ref{fig:ndwraw0}, which displays data traces over the positive offset range $10000 \le
+x_r \le 20000$ m. This part of the wavefield is connected to some
+refracted rays from the source, those shown in blue in Figure
+\ref{fig:ooplrays0}.  Other refracted rays, shown in red in that
+Figure, either do not arrive at the receiver surface within the time
+range of the data, or are associated with another branch of refracted
+arrivals. These two collections of refracted rays result in two
+arrival time branches, shown in blue and red in Figure
+\ref{fig:ooxtraysend}. The refracted arrival times shown in red,
+corresponding to some of the rays shown in red in Figure
+\ref{fig:ooplrays0}, corresponds to a refracted arrival that almost
+overlaps the direct wave, as is clearly visible in Figure
+\ref{fig:ndwraw0}. The arrival combines with the direct wave to result
+in an apparent waveform change in the latter. The refracted arrival
+times shown in blue in Figure \ref{fig:ooxtraysend} correspond to the
+blue rays in Figure \ref{fig:ooplrays0}, and to the faster event
+extending to the right of the direct wave.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:ndwraw0}]{\includegraphics[width=0.4\textwidth]{ndwraw0.pdf}}
+  \subfloat[\label{fig:ooxtraysend}]{\includegraphics[width=0.4\textwidth]{ooxtraysend.pdf}}
+  \caption{Data generated using configuration in Figure
+    \ref{fig:ooplrays0}. (a) Traces from point source at $(500,10000)$ m
+    with $[1, 2, 7.5, 12.5]$ Hz zero phase trapezoidal bandpass
+    filter wavelet, delayed by $0.4$ s. Note that direct arrival
+    waveform appears to change where it intersects the diving wave
+    arrival visible on the right. (b) Arrival times at surface $z=0$
+    of rays leaving source point ($(500,10000)$ m) at $t=0$. Blue
+    and red branches corresponds to blue and red rays in Figure
+    \ref{fig:ooplrays0}, respectively. The diving wave carried by
+    the red rays arrives at the red time branch, which is nearly
+    coincident with the direct arrival branch (the black line). This
+    part of the refracted wave field combines with
+    the direct wave to produce the apparent wavelet change shown in 
+    subfigure (a). The blue arrival time branch corresponds to the diving
+    wave energy carried by the blue rays in Figure
+    \ref{fig:ooplrays0}, displayed in Figure \ref{fig:ndwdata0}.}
+\end{figure}
+
+Note that at the source, all of the rays shown in Figure
+\ref{fig:ooplrays0} are incoming if the source surface
 orientation is chosen with inwards = down, in contrast to the previous
-example, while they are outgoing at
-the receiver surface if it is oriented so that inwards = down also.
-
-This geometry permits inversion of the diving wave data alone, which
-is isolated via a mute, depicted in Figure \ref{fig:ndwmute0}. The
+example. The ``blue'' rays, as well as those of the ``red'' rays that
+are refracted back towards $z=0$, are outgoing at
+the receiver surface if it is oriented so that inwards = down
+also.
+
+The ``blue'' arrival can be  isolated via a mute, depicted in Figure
+\ref{fig:ndwmute0}.  The
 mute includes time truncation at 6 s, and offsets between -10000 and
 10000 m. The isolated
-diving wave (window between offsets 4000 and 10000) is shown in
-Figure \ref{fig:ndwdata0}.
+diving wave (for positive offsets) is shown in
+Figure \ref{fig:ndwdata0}. This muted field can be inverted via the
+techniques exploained below, that is, a source field on $z=500$ m
+constructed that regenerates approximately the data shown in
+\ref{fig:ndwdata0}. In principle the ``red'' arrival, or the ``red''
+and ``blue'' arrivals together, could also be inverted by the same
+technique, however the ``red'' arrival can only be isolated by
+subtraction of the direct wave, which is not itself incoming (or
+outgoing). In this paper, the simpler option is pursued, of inverting
+the muted diving wave arrival in Figure \ref{fig:ndwdata0}, in order
+to illustrated the localization mechanism at the heart of the inversion.
 
@@
   \centering
-  \subfloat[\label{fig:ndwraw0}]{\includegraphics[width=0.32\textwidth]{ndwraw0.pdf}}
-  \subfloat[\label{fig:ndwmute0}]{\includegraphics[width=0.32\textwidth]{ndwmute0.pdf}}
-  \subfloat[\label{fig:ndwdata0}]{\includegraphics[width=0.32\textwidth]{ndwdata0.pdf}}
-  \caption{Data generated using configuration in Figure
-    \ref{fig:ooplrays0}. (a) traces from point source at $(500,10000)$ m
-    with $[1, 2, 7.5, 12.5]$ Hz zero phase trapezoidal bandpass
-    filter wavelet, delayed by $0.4$ s. (b) Mute function,
-    incorporated in the sampling operator $P_r$. (c) Diving wave
-    gather. Color scale is same for (a) and (c).}
+  \subfloat[\label{fig:ndwmute0}]{\includegraphics[width=0.4\textwidth]{ndwmute0.pdf}}
+  \subfloat[\label{fig:ndwdata0}]{\includegraphics[width=0.4\textwidth]{ndwdata0.pdf}}
+  \caption{(a) Mute function,
+    incorporated in the sampling operator $P_r$, designed to isolate
+    the diving wave branch visible in Figure \ref{fig:ndwraw0}. (b) Diving wave
+    gather, mute in (a) applied to data in Figure
+    \ref{fig:ndwraw0}. Color scale is 5 tmies that of Figure \ref{fig:ndwraw0}.}
 \end{figure}
@@
 
-Time reversal as found in the literature on photoacoustic tomography
-must be localized to apply to the surface source inversion problem,
-since there is no measurement surface surrounding the region of
-propagation. It must also be modified to produce a source, rather than
-a restriction of the pressure field. The assumption of incoming rays at the source surface and
-outgoing rays at the receiver surface makes both modifications  
-possible.
-
-The construction takes place in three stages: first, an
-acoustic field with high-frequency energy propagating along an
-incoming ray family is identified as the causal solution of an inhomogeneous boundary value problem {\em locally and on
-  the inside}. The boundary value is simply the restriction of the
-acoustic field to the surface. Second, the same causal solution is
-identified as the causal solution of an acoustic problem with source,
-singular on the surface in the fashion of equation
-\ref{eqn:awepm}. The source amplitude is proportional to 
-the velocity field sampled on the surface. Third, this identification is used twice, first to
-propagate the receiver data backwards in time as the solution of an
-anti-causal problem,  the velocity field of which is sampled on the
-source surface; then the causal problem is solved with the
-source built from the velocity samples. This solution must be
-identical to the field generating the data, along the ray fields
-carrying all of the high-frequency energy. Therefore the reconstructed
-source is an approximate solution of the problem \ref{eqn:esis} for
-$\alpha=0$.
+There is no measurement surface surrounding the region of propagation
+in the surface source inversion problem. Therefore the time reversal
+construction, as found in the literature on photoacoustic tomography,
+must be localized in space-time, so that an artificial surrounding
+region may be constructed. It must also be modified to produce a
+source, rather than a slice of the pressure field. The assumption of
+incoming rays at the source surface and outgoing rays at the receiver
+surface makes both modifications possible.
+
+The construction takes place in three stages: first, an acoustic field
+with high-frequency energy propagating along an incoming ray family is
+identified as the causal solution of an inhomogeneous boundary value
+problem {\em locally and on the inside}. The boundary value is simply
+the restriction of the pressure field to the surface. Second, the same
+causal solution is identified as the causal solution of an acoustic
+problem with source, singular on the surface in the fashion of
+equation \ref{eqn:awepm}. The source amplitude is proportional to the
+velocity field sampled on the surface. Third, this identification is
+used twice, first to propagate the receiver data backwards in time as
+the solution of an anti-causal problem with source on the receiver
+surface; the velocity field of which is sampled on the source surface;
+then the causal problem is solved with the source built from the
+velocity samples. This solution must asymptotically approximate the
+field generating the data, along the ray fields carrying all of the
+high-frequency energy. Therefore the reconstructed source is an
+approximate solution of the problem \ref{eqn:esis} for $\alpha=0$.
 
@@
 \begin{eqnarray}
-\label{eqn:awebase}
+\label{eqn:awefree}
   \frac{1}{\kappa}\frac{\partial p}{\partial t} & = & - \nabla \cdot \bv, \nonumber \\
@@
   \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla
-                                                    \tilde{p}, \\
+                                                    \tilde{p}, \nonumber \\
   \tilde{p}(\bx,t) & = & p(\bx,t) \mbox{ for } \bx \in \Sigma, 0 \le t \le 
-  \tau,\\
+  \tau, \nonumber \\
   \tilde{p}(\bx,t) & = & 0 \mbox{ for } \bx \in \partial \Omega^+ \setminus \Sigma, 0 \le t \le
@@
 incoming side of the surface $\Sigma$, to a full neighborhood $\Omega$
-of sigma
+of $\Sigma$
 by adding a domain $\Omega^-$ on the outside (see Figure
@@
                                                       - \nabla \cdot \tilde{\bv}, \nonumber \\
-  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p},\\
+  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p}, \nonumber \\
   \tilde{p}(\bx,t) & = & 0 \mbox{ for } \bx \in \partial \Omega, 0 \le t \le 
-  \tau,\\
-  \tilde{p} & = & 0 \mbox{ for } \bx \in \Omega^+, t , \\
+  \tau, \nonumber \\
+  \tilde{p} & = & 0 \mbox{ for } \bx \in \Omega^+, t , \nonumber \\
   \tilde{\bv} & = & {\bf 0} \mbox{ for } \bx \in \Omega^+, t = 0.       
 \end{eqnarray}
-in which $\delta_{\Sigma}$ is the delta function on $\Sigma$. The
-square bracket denotes the jump in the quantity enclosed across
+The square bracket denotes the jump in the quantity enclosed across
 $\Sigma$, from the outside to the inside. Since ${\bf n}$ points from
@@
 This is best seen by analyzing a generic geometric optics
-component associated to the phase function $\psi^+$ in $\Omega^+$ and its family of
+component associated to the phase function $\psi$ in $\Omega^+$ and its family of
 rays. The acoustic field is assembled by a sum (in the continuum, an
@@
   \label{eqn:go}
-   a(\bx)e^{i\omega (t - \psi(\bx))}.
+   a(\bx,t)e^{i\omega (t - \psi(\bx))}.
 \end{equation}
@@
 \[
-  \lim_{\zeta \rightarrow 0} (\tilde{\bv}(\bx,t) \cdot {\bf
-    n}(\bx+\zeta {\bf n})) = 
--\lim_{\zeta \rightarrow 0} (\tilde{\bv}(\bx,t) \cdot {\bf
-  n}(\bx -\zeta {\bf n}))
+  \lim_{\zeta \rightarrow 0} (\tilde{\bv}(\bx+\zeta {\bf n}(\bx),t) \cdot {\bf
+    n}(\bx)) = 
+-\lim_{\zeta \rightarrow 0} (\tilde{\bv}(\bx-\zeta {\bf n}(\bx),t) \cdot {\bf
+  n}(\bx))
 \]
@@
 \begin{equation}
-[\tilde{\bv} \cdot {\bf n}] = 2  \label{eqn:vjump} \lim_{\zeta
-  \rightarrow 0} (\tilde{\bv}(\bx,t) \cdot {\bf n}(\bx -\zeta {\bf
-  n}))
-\approx 2 \bv \cdot {\bf n}
+   \label{eqn:vjump}
+   [\tilde{\bv} \cdot {\bf n}] = 2 \lim_{\zeta
+  \rightarrow 0} (\tilde{\bv}(\bx -\zeta {\bf
+  n}(\bx),t) \cdot {\bf n}(\bx))
+\approx 2 \bv(\bx) \cdot {\bf n}(\bx)
 \end{equation}
@@
 \label{eqn:awesrc}
-  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & f \delta_{\Sigma}
+  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & h \delta_{\Sigma}
                                                       - \nabla \cdot \tilde{\bv}, \nonumber \\
-  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p},\\
-  \tilde{p} & = & 0 \mbox{ for } \bx \in \Omega, t=0, \\
+  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p}, \nonumber \\
+  \tilde{p} & = & 0 \mbox{ for } \bx \in \Omega, t=0, \nonumber \\
   \tilde{\bv} & = & {\bf 0} \mbox{ for } \bx \in \Omega, t = 0.       
 \end{eqnarray}
-with $f = 2 \bv \cdot {\bf n}$ on $\Sigma$.
+with $h = 2 \bv \cdot {\bf n}$ on $\Sigma$.
 
@@
   \label{eqn:awesrcmod}
-  ... & = & ...\\
-  \tilde{p} & = & 0 \mbox{ for }  t \ll 0, \\
+  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & h \delta_{\Sigma}
+                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
+  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p}, \nonumber \\
+  \tilde{p} & = & 0 \mbox{ for }  t \ll 0, \nonumber \\
   \tilde{\bv} & = & {\bf 0} \mbox{ for } t \ll 0.       
@@
 
-An example has already appeared in Figure \ref{fig:dsrcphh0}. In this
+An example has already appeared in Figure \ref{fig:dhshh0}. In this
 example, the field $(p, \bv)$ is the point source solution with the
 point source depicted in Figure \ref{fig:olensrays0}, located at
-$(3500, 3500)$ m. The role of $\Sigma$ is played by the source surface
-$z=z_s=3000$ m, and near this surface the field solves the source-free
-system \ref{eqn:awebase}. The ``inside'' in this example is the region
-$z > z_s$, and the outward unit normal is therefore ${\bf n}=(0,0,-1)$.
+$(3500, 3500)$ m. The role of $\Sigma$ is played by the source surface $\Sigma_s=
+\{z=z_s=3000 \mbox{ m}\}$, and near this surface the field solves the source-free
+system \ref{eqn:awefree}. The ``inside'' in this example is the region
+$z > z_s$, and the outward unit normal is therefore ${\bf n}=(0,0,-1)^T$.
 Thus $(p,\bv)$ should be asymptotically close to the solution
@@
 system \ref{eqn:awesrc} in the region $z< z_s$, at least at locations
-connected by the ray field shown in Figure \ref{fig:olenrays0} to the
-source surface, provided that the source amplitude $f$ is chosen $=
--2v_z$ on $\Sigma=\{z=z_s\}$. Figure \ref{fig:dsrcphh0} displays exactly this $f$, and
+connected by the ray field shown in Figure \ref{fig:olensrays0} to the
+source surface, provided that the source amplitude $h$ is chosen $=
+-2v_z$ on $\Sigma_s=\{z=z_s\}$. Figure \ref{fig:dhshh0} displays exactly this $h$, and
 \ref{fig:dfwdplh0} the traces of $\tilde{p}$ on the
   receiver surface $z=z_r=1000$ m, which satisfies the ray condition. The extremely
-close match between Figures \ref{fig:drecplh0} (traces of $p$) and
+close match between Figures \ref{fig:drecplh0} (traces of $p$ on $\Sigma_r$) and
 \ref{fig:dfwdplh0} is consistent with the conclusion reached here.
@@
 of seismic surface source inversion. Assume that $(p,\bv)$ is a
-solution of the free-space acoustic system \ref{eqn:awebase}, with
+solution of the free-space acoustic system \ref{eqn:awefree}, with
 high-frequency energy carried by a ray family that satisfies the
@@
 time, to conclude
-that $(p,\bv)$ is approximated by $(\tilde{p},\tilde{\bv})$, where the
+that $(p,\bv)$ is approximated by $(\hat{p},\hat{\bv})$, where the
 latter solves the time-reversed problem
 \begin{eqnarray}
-\label{eqn:awerec}
-  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & h_r \delta_{\Sigma_r}
-                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
-  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p},\\
-  \tilde{p} & = & 0 \mbox{ for } t\gg 0, \\
-  \tilde{\bv} & = & {\bf 0} \mbox{ for } t \gg 0.       
+\label{eqn:awerecglob}
+  \frac{1}{\kappa}\frac{\partial\hat{p}}{\partial t} & = & h_r \delta_{\Sigma_r}
+                                                      - \nabla \cdot \hat{\bv}, \nonumber \\
+  \rho\frac{\partial \hat{\bv}}{\partial t} & = & - \nabla \hat{p}, \nonumber \\
+  \hat{p} & = & 0 \mbox{ for } t\gg 0, \nonumber \\
+  \hat{\bv} & = & {\bf 0} \mbox{ for } t \gg 0.       
 \end{eqnarray}
@@
 
-By construction, high-frequency energy in $(\tilde{p},\tilde{\bv})$ is
+By construction, high-frequency energy in $(\hat{p},\hat{\bv})$ is
 carried by the same ray family as for $(p,\bv)$. Every ray passes over
 $\Sigma_s$ and is incoming there. Since $(p,\bv)$ vanishes for $t \ll
-0$ near $\Sigma_s$, $(\tilde{p},\tilde{\bv})$ is negligible there,  So
+0$ near $\Sigma_s$, $(\hat{p},\hat{\bv})$ is negligible there,  So
 the construction detailed in the last two sections is applicable
-again, and $(\tilde{p},\tilde{\bv})$ is approximated by the solution
-of 
+again, and the reverse-time field $(\hat{p},\hat{\bv})$ is
+approximated by the forward-time field $(\tilde{p},\tilde{\bv})$ that solves
 \begin{eqnarray}
-\label{eqn:awesrc}
-  \frac{1}{\kappa}\frac{\partial\hat{p}}{\partial t} & = & h_s \delta_{\Sigma_s}
-                                                      - \nabla \cdot \hat{\bv}, \nonumber \\
-  \rho\frac{\partial \hat{\bv}}{\partial t} & = & - \nabla \hat{p},\\
-  \hat{p} & = & 0 \mbox{ for } t\ll 0, \\
-  \hat{\bv} & = & {\bf 0} \mbox{ for } t \ll 0.       
+\label{eqn:awesrcglob}
+  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & h_s \delta_{\Sigma_s}
+                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
+  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p}, \nonumber \\
+  \tilde{p} & = & 0 \mbox{ for } t\ll 0, \nonumber \\
+  \tilde{\bv} & = & {\bf 0} \mbox{ for } t \ll 0.       
 \end{eqnarray}
-with $h_s = 2P_s(\tilde{\bv} \cdot {\bf n}_r) \approx 2P_s(\bv \cdot
+with $h_s = 2P_s(\hat{\bv} \cdot {\bf n}_r) \approx 2P_s(\bv \cdot
 {\bf n}_r)$.
@@
 pressure data $P_rp$ on the receiver surface $\Sigma_r$ for a solution
-$(p,\bv)$ of the system \ref{eqn:awebase} satisfying the ray conditions,
+$(p,\bv)$ of the system \ref{eqn:awefree} satisfying the ray conditions,
 \begin{itemize}
@@
 \item[2. ] propagate $h_r$ backwards in time by solving the system
-  \ref{eqn:awerec} for the acoustic field $(\tilde{p},\tilde{\bv})$;
+  \ref{eqn:awerecglob} for the acoustic field $(\hat{p},\hat{\bv})$;
 \item[3. ] create a source $h_s$ on $\Sigma_s$ by  multiplying tne
-  normal velocity trace of $(\tilde{p},\tilde{\bv})$ by 2: $h_s = 2P_s(\tilde{\bv}
+  normal velocity trace of $\hat{\bv}\cdot {\bf n}_s$ by 2: $h_s = 2P_s(\hat{\bv}
   \cdot {\bf n}_s)$.
@@
 
-The surface source representaion part of the construction shows that
-under the assumed ray conditions, a solution of \ref{eqn:awebase} is
-asymptotically the same near the receive surface $\Sigma_r$ as the
+The surface source representation part of the construction shows that
+under the assumed ray conditions, a solution of \ref{eqn:awefree} is
+asymptotically the same near the receiver surface $\Sigma_r$ as the
 solution of the system \ref{eqn:awepm} with a suitable source
-$\tilde{f}_s$ (and zero velocity source, $h_s=0$. Therefore this
+$h_s$ (and zero velocity source, $f_s=0$). Therefore this
 construction also inverts the modeling operator $S$ in the sense that
 $Sh_s\approx S\tilde{h_s}$ if $P_rp \approx S\tilde{h_s}$. That is,
-this construction produces an asymptotic rightt inverse of $S$: it
+this construction produces an asymptotic right inverse of $S$: it
 does not necessily (approximately) recover the source that generates
@@
 The first example uses the acoustic lens model (Figure
-\ref{fig:olensrays0}, with $\Sigma_s$ = horizontal line at depth 
+\ref{fig:olensrays0}), with $\Sigma_s$ = horizontal line at depth 
 3000 m, and $\Sigma_r$ = horizontal line at depth 1000 m. The pressure
@@
 $\Sigma_r$, generated by the point source shown in the figure, at
-$(3500, 3500)$ m (Figure \ref{fig:drecplh0}. However, I will invert
-this data in a homogeneous medium with $\kappa \equiv 4$ GPa. These
-choice illustrate two facts about surface source inversion and the
-approximate inversion based on time reversal. First, the success of the inversion demostrates the
+$(3500, 3500)$ m (Figure \ref{fig:drecplh0}). However, I will invert
+this data in a homogeneous medium with $\kappa \equiv 4$ GPa. This
+choice illustrates two facts about surface source inversion and
+approximate inversion based on time reversal. First, the success of the inversion demonstrates the
 insensitivity of the time reversal method to ray multipathing
@@
 data, of course). This
-capability is critically important in the application of the
-approximate inversion in nonlinear inversion, where the early iterations
+capability is critically important in the application of 
+approximate source inversion in nonlinear extended inversion, where the early iterations
 involve solution of the source estimation problem \ref{eqn:esis} at
@@
   e}_z=(0,-1)$. So the source on the receiver surface $\Sigma_r$, input to reverse time propagation
-(system \ref{eqn:awerec}, generating the fields $(\tilde{p},\tilde{\bv})$) is
+(system \ref{eqn:awerecglob}, generating the fields $(\hat{p},\hat{\bv})$) is
 $h_r=2P_rv_z$ (the field $P_rv_z$ is depicted in Figure \ref{fig:dfwdvzlh0}). After reverse time propagation , the approximate
-inversion $\tilde{h}_s=2P_s\tilde{v}_z$ is shown in Figure
-\ref{dinvhslh0}. It differs considerably from the source field used to
-generate the data (Figure \ref{fig:dsrcphh0}, however when used as the
-source in forward modeling (system \ref{eqn:awesrc}) {\em with the
+inversion $\hat{h}_s=2P_s\hat{v}_z$ is shown in Figure
+\ref{fig:dinvhslh0}. It differs considerably from the source field used to
+generate the data (Figure \ref{fig:dsrcphh0}), however when used as the
+source in forward modeling (system \ref{eqn:awesrcglob}) {\em with the
   same material parameters used in the inversion}, it generates a
@@
   \caption{(a) Resimulated data: pressure gather generated from
-    inverted source (Figure \ref{fig:dinvhslh0}, using same
+    inverted source (Figure \ref{fig:dinvhslh0}), using same
     homogeneous bulk modulus as used in inversion. (b) Difference
@@
   \caption{(a) Resimulated data: pressure gather generated from
-    inverted source (Figure \ref{fig: ndwdataresrcplbulksm0}, using same
+    inverted source (Figure \ref{fig:ndwdataresrcplbulksm0}), using same
     bulk modulus (Figure \ref{fig:ooplrays0}) as used in data
@@
     between data depicted in Figure \ref{fig:ndwdataresimplbulksm0} and in Figure \ref{fig:ndwdata0}.
-    Both (a) and (b) plotted with same color scale as Figure \ref{fig:ndwdata0}.}
+    Both (a) and (b) plotted with same color scale as Figure
+    \ref{fig:ndwdata0}. Note that low-amplitude ``ghosts'' of other
+    arrival branches appear in the difference plot.}
 \end{figure}
@@
 
-Since the system \ref{eqn:awepm} has a unique solution by standard
-theory \cite[]{Lax:PDENotes}, the source vector field $(h_s,f_s)$
-determines the acoustic field $(p^{\pm},\bv^{\pm})$ in space time, and
-in particular the limits from the right at $z=z_s$, $P_sp^{\pm}$ and
-$P_sv_z^{\pm}$. This relation is not invertible: it is not possible to
-prescribe both pressure and normal velocity on a surface such as
-$z=z_s$. So the columns of the matrix operator
-${\cal S}^{\pm}_{z_s,z_r}$ must satisfy a linear relation. In this
-section I will explain this relation; it involves the {\em
-  pressure-to-source} map. This operator also turns out to be the
-principal component of a preconditioning strategy for iterative
-solution of the optimization problem \ref{eqn:esis}, so I will devote
-some effort to its proper definition. It is closely related to the
-Dirichlet-to-Neumann operator mentioned in the introduction.
-
-While it is not possible to prescribe both pressure and velocity on
-$z=z_s$ in solutions of \ref{eqn:awepm}, it is possible to
-prescribe pressure only, for instance: if the function $\phi$ on
-the surface $z=z_s$ satisfies suitable conditions, for
-example the downgoing constraint mentioned earlier, a unique solution
-exists for the acoustic system in both half-spaces $\pm z > z_s$:
-\begin{eqnarray}
-\label{eqn:awe0}
-  \frac{1}{\kappa}\frac{\partial p_{\pm}}{\partial t} & = & - \nabla \cdot \bv_{\pm}, \nonumber \\
-  \rho\frac{\partial \bv_{\pm}}{\partial t} & = & - \nabla
-                                                    p_{\pm}, \nonumber \\
-  p_{\pm} & =& 0,  \mbox{ for } t \ll 0, \nonumber\\ 
-  \bv_{\pm} & = & 0 \mbox{ for } t \ll 0, \nonumber\\
-  \lim_{z \rightarrow z_s^{\pm}}p_{\pm}(x,y,t,z)& =& \phi(x,y,t).
-\end{eqnarray}
-Note that the subscript $\pm$ here refers to the sign of $z-z_s$, as opposed
-to the superscript ${\pm}$, which refers to the sign of $t$ throughout
-this paper.
-
-From the boundary condition (last equation in \ref{eqn:awe0}), one
-sees that the pressures $p_{\pm}$ in the two half-spaces have the same
-limit at the boundary $z=z_s$. Stick the two half-space
-solutions together to form an acoustic field $(p^+,\bv^+)$ in all of
-space-time, that is,
-\begin{equation}
-  \label{eqn:awealt}
-  p^+(x,y,z,t) =
-  \left\{
-    \begin{array}{c}
-      p_+(x,y,z,t) \mbox{ if } z>0,\\
-      p_-(x,y,z,t) \mbox{ if } z<0,
-    \end{array}
-  \right.
-\end{equation}
-and a similar definition for $\bv^+$. Then $p^+$ is continuous across
-$z=z_s$, and the boundary condition in system \ref{eqn:awe0} may be
-written as $P_sp^+=\phi$.
-
-The same construction can be carried out in the anti-causal sense,
-with anti-causal half-space solutions glued together to form a
-full-space distribution solution $(p^-,\bv^-)$, with the property that
-$p^-$ is continuous across $z=z_s$ and $P_sp^-=\phi$.
-
-The reader may object that the notation $(p^\pm,\bv^\pm)$ is already in
-use, for the solution of \ref{eqn:awepm}. This objection is
-valid. However, {\em in the sense
-  of distributions}, $(p^{\pm},\bv^{\pm})$ as defined in display
-\ref{eqn:awealt}, is {\em exactly} the causal solution of \ref{eqn:awepm}
-for the choice $h_s = -[v^{\pm}_{z}]|_{z=z_s}, f_s=0$, as follows from a
-simple integration-by-parts calculation. So the notation is consistent!
-
-The negative jump $-[v^{\pm}_{z}]|_{z=z_s}$ is thus a function of $\phi$. Define
-the {\em pressure-to-source} operator $\Lambda^{\pm}_{z_s}$ by
-\begin{equation}
-  \label{eqn:deflam}
-  \Lambda^{\pm}_{z_s}\phi = -[v^{\pm}_{z}]|_{z=z_s}
-\end{equation}
-The conclusion: if $h_s = \Lambda^{\pm}_{z_s}\phi$ and $f_s=0$ in the
-system \ref{eqn:awepm}, then $\phi=P_sp^{\pm}$.
-
-Otherwise put, $S^{\pm}_{z_s,z_s}\Lambda^{\pm}_{z_s} \phi = \phi$, so
-$\Lambda^{\pm}_{z_s}$ is inverse to $S^{\pm}_{z_s,z_s}$. The relation
-\ref{eqn:trtrcomp} implies in turn that
-\begin{equation}
-  \label{eqn:lamadj}
-  (\Lambda^{\pm}_{z_s})^T = - \Lambda^{\mp}_{z_s}
-\end{equation}
-
-There is also a {\em velocity-to-source} operator. For the solution
-$(p^{\pm},\bv^{\pm})$ of system \ref{eqn:awepm} with $h_s=0$, the
-normal component of velocity, $v^{\pm}_z$, is continuous across
-$z=z_s$, and the velocity source (vertical load)
-$f_s=-[p^{\pm}]_{z=z_s}$. I will not name the velocity-to-source
-operator, as it does not appear explicitly in the developments to
-follow. As will be seen, it is essentially the inverse of the
-pressure-to-source operator.
-
-The quadratic form defined by $\Lambda^{\pm}_{z_s}$ has fundamental
-physical significance. Define the total acoustic energy $E^{\pm}(t)$ of the
-field $(p^{\pm},\bv^{\pm})$, at time $t$ by
-\begin{equation}
-  \label{eqn:defae0}
-  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)(\bx,t).
-\end{equation}
-Then
-\begin{equation}
-  \label{eqn:elim}
-  \pm \lim_{\pm t \rightarrow \infty} E^{\pm}(t) =  \langle P_sp^{\pm},
-  (\Lambda^{\pm}_{z_s} P_sp^{\pm}) \rangle_{L^2(z=z_s)}.
-\end{equation}
+The result of the last section has a couple of apparent
+defects. First, the approximate inversion of pressure data requires
+knowledge of velocity data, which is not prescribed in the problem
+statement \ref{eqn:esisp}. Second, the significance of the
+reverse-time field $(\hat{p},\hat{\bv})$ is obscure. In this section
+these two issues are addressed.
+
+\subsection{Pressure-to-source operator}
+The first step is to formalize the relation between pressure and
+source (or equally well, pressure and normal velocity) on as surface
+$\Sigma$ as the application of an operator.
+
+Observe that in principle it's not necessary to
+supply the normal velocity at the source (or receiver) surface, since
+it's determined locally by the pressure on the surface, or causal (or
+anti-causal) fields. This follows from the observation that the system
+\ref{eqn:aweloc} has a unique solution, a by-product of the
+conservation of energy (see for example \cite{Lax:PDENotes}, section
+4.3). That is, in the language of that discussion, knowledge of $p$ on
+$\Sigma$ for $0\le t \le \tau$ determines $(\tilde{p},\tilde{\bv})$ in
+the entire domain $\Omega^+$, $0\le t \le \tau$. Therefore knowledge
+of $p$ on $\Sigma$ for $0\le t \le \tau$ determines
+$\bv \cdot {\bf n}$, up to a negligible error. As in the last section,
+it follows that knowledge of $p$ on $\Sigma$, any $t$, determines a
+source $h$ on $\Sigma$ for which the pressure field $\tilde{p}$
+(solution of the system \ref{eqn:awesrcmod}) differs negligibly from
+$p$.
+
+This relation defines a causal {\em pressure-to-source operator}
+$\Lambda: p \mapsto h$, yielding the source on a surface $\Sigma$ that
+reproduces the (causal) pressure field from its trace on $\Sigma$, at
+least locally on the inward side as defined by the selection of normal
+field on $\Sigma$, under the ray conditions defined earlier. Since
+$h \approx 2 \bv \cdot {\bf n}$ (limiting value of $\bv$ from the
+inward direction), $\Lambda$ is effectively the
+pressure-to-normal-velocity operator.
+
+With this convention, two of the three steps in
+the approximate inversion procedure are operator applications. The
+operators involved are the causal pressure-to-source operator for the
+source surface $\Sigma_s$, denoted $\Lambda_s$, and an analogous
+anti-causal operator for the receiver surface $\Sigma_r$. 
+The characterization of the the source field $h_r$ in equation
+\ref{eqn:awerecglob} shows that this anti-causal pressure-to-source
+operator differs from its causal analogue only by sign. Denote the
+causal operator by $\Lambda_r$.
+
+\subsection{Time Reversal as Adjoint State}
+The middle step is solution of the reverse-time system
+\ref{eqn:awerecglob} for the field $(\hat{p},\hat{\bv})$, with input
+pressure source
+$h_r$ on $\Sigma_r$ and output $P_s\hat{p}$ on $\Sigma_s$. To
+understand the significance of this mapping, it is just as easy, and
+will be useful later, to include a velocity source $f_r$, that is,
+modify the second equation in \ref{eqn:awerecglob} by adding $f_r {\bf
+  n}_r \delta_{\Sigma_r}$, so that it becomes a reverse-time version
+of the basic system \ref{eqn:awepm}. Recall the definition of energy
+in the acoustic field $(p,\bv)$:
+\begin{equation}
+  \label{eqn:energydef}
+   E[p,\bv](t) = \frac{1}{2} \int \,dx\,dy\,dz\, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx)\bv(\bx,t)
+   \cdot \bv(\bx,t)\right)
+\end{equation}
+This is a quadratic form. Polarization makes a bilinear form, into
+which one can plug a pair of fields. Use the solutions $(p,\bv)$ of
+the forward-time system \ref{eqn:awepm} for one of the pair, and
+the solution $(\hat{p},\hat{\bv})$ of system \ref{eqn:awerecglob}
+(augmented with the velocity surface source) for the other. Since $(p,\bv)$ vanishes
+for large negative time, $(\hat{p},\hat{\bv})$ for large positive
+time, 
+\[
+0 = 
+\left(\int\, dx\,dy\,dz\, \frac{p \hat{p}}{\kappa} +  
+\rho \bv \cdot \hat{\bv} \right)|_{t \rightarrow \infty}
+-
+\left(\int\, dx\,dy\,dz\, \frac{p \hat{p}}{\kappa} +  \rho \bv \cdot \hat{\bv} \right)|_{t \rightarrow -\infty}
+\]
+\[
+= 
+\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \hat{p}}{\kappa} +  \rho \bv \cdot \hat{\bv} \right)
+\]
+Carry out the differentiation under the integral sign, replace time derivatives by space derivatives and
+source terms using the equations of motion in the systems
+\ref{eqn:awepm} and \ref{eqn:awerecglob}, and integrate by parts to
+move all of the space derivatives onto $p$ and $\bv$.  Most of the
+remaining expressions cancel, leaving
+\[
+  = \int_{-\infty}^{\infty}\,dt \, \left[
+    \int\,d\sigma_s(\bx)\,(h_sP_s\hat{p}+f_sP_s(\hat{\bv}\cdot{\bf
+      n}_s)) +
+    \int\,d\sigma_r(\bx)\,(h_rP_rp+f_rP_r(\bv\cdot{\bf n}_r)) \right]
+\]
+Here $\sigma_s$ and $\sigma_r$ are the area elements on $\Sigma_s$ and
+$\Sigma_r$. These integrals are inner products of time traces on $\Sigma_s$ and $\Sigma_r$. Denote these inner products by $\langle \cdot,\cdot
+\rangle_s$ and $\langle \cdot,\cdot
+\rangle_r$ respectively. Then the above is written in terms of the
+vector modeling operator ${\cal S}$ (definition \ref{eqn:fwd}) as 
+\[
+  = \langle (h_s,f_s)^T, (P_s\hat{p},P_s(\hat{\bv}\cdot{\bf
+    n}_s))^T \rangle_s+ \langle (h_r,f_r)^T, {\cal S}(h_s,f_s)^T \rangle_r,
+\]
+whence identify the adjoint of ${\cal S}$ as
+\begin{equation}
+  \label{eqn:vecadj}
+  {\cal S}^T(h_r,f_r) = -(P_s\hat{p},P_s(\hat{\bv}\cdot{\bf
+    n}_s))^T.
+\end{equation}
+Specializing to the pressure modeling operator $S$ (equation
+\ref{eqn:sdef}),
+\begin{equation}
+  \label{eqn:padj}
+  S^T = (\Pi_0 {\cal S} \Pi_0^T)^T = \Pi_0 {\cal S}^T \Pi_0^T
+\end{equation}
+so to calculate the adjoint of $S$ applied to $h_r$, solve system
+\ref{eqn:awerecglob} (with $f_r=0$) and extract the pressure at $\Sigma_s$:
+\[
+  S^T(h_r) = -P_s \hat{p}.
+\]
+
+Combined with the definitions of the pressure-to-source operators
+$\Lambda_s$ and $\Lambda_r$, these results enable a translation of the
+approximate inverse construction in the last section into a chain of
+operator applications: to invert a pressure gather $d$ on $\Sigma_r$
+for a source $h_s$ on $\Sigma_s$ that approximately reproduces it ($d
+\approx Sh_s$),
+\begin{itemize}
+\item[1. ] create a source $h_r = -2\Lambda_r d$ on $\Sigma_r$;
+\item[2. ] apply the adjoint $S^T$ (solve the reverse time system
+  \ref{eqn:awerecglob}) to $h_r$;
+\item[3. ] apply $\Lambda_s$ to $-S^Td$.
+\end{itemize}
+Putting this all together, an approximate right inverse operator
+$S^{\dagger}$ to $S$ may be expressed as
+\begin{equation}
+  \label{eqn:baseinv}
+  S^{\dagger} = \Lambda_s S^T \Lambda_r,
+\end{equation}
+that is,
+\begin{equation}
+  \label{eqn:rightinv}
+  SS^{\dagger} \approx I.
+\end{equation}
+
+\subsection{Positivity of $\Lambda$}
+
+Equation \ref{eqn:baseinv} is remarkable, in that the form of
+$S^{\dagger}$ is that of an adjoint with respect to weighted norms,
+defined by the operators $\Lambda_s^{-1}$ in the domain of $S$ and
+$\Lambda_r$ in the range. This only makes sense if the
+pressure-to-source operators are positive definite and symmetric, the
+necessary conditions to define weighted norms.
+
+In fact the pressure-to-source operator is positive semi-definite, and that
+fact has fundamental physical significance. Suppose that $(p,\bv)$
+solves the basic system \ref{eqn:awepm} with $f_s=0$. From the
+definition of energy \ref{eqn:energydef} and causality of the fields,
+\[
+ E[p,\bv](t_1) = \frac{1}{2} \int_{-\infty}^{t_1} \,dt \int \,dx\,dy\,dz\,
+ \frac{d}{dt}\left(\frac{p(\bx,t)^2}{\kappa(\bx)} +
+   \rho(\bx)\bv(\bx,t) \cdot \bv(\bx,t)\right)
+\]
+\[
+  = \int_{-\infty}^{t_1} \,dt \int \,dx\,dy\,dz\,
+ \left(\frac{p (\bx,t)\frac{\partial p}{\partial t}(\bx,t)^2}{\kappa(\bx)} +
+   \rho(\bx)\bv(\bx,t) \cdot \partial{\bv}{\partial t}(\bx,t)\right)
+\]
+Using \ref{eqn:awepm}, replace the time derivatives with spatial derivatives and the delta
+function on ths source surface, integrate by parts to shift the space
+derivatives to $p$. Everything cancels except for
+\[
+  =  \int_{-\infty}^{t_1} \,dt \int \,dx\,dy\,dz\, p(\bx,t) h_s(\bx,t)
+  \delta_{\Sigma_s}(\bx) = \int_{-\infty}^{t_1} \,dt \int_{\Sigma_s} \,d\sigma_s \, p h_s.
+\]
+Take the limit as $t_1 \rightarrow \infty$ to get
+\[
+\lim_{t_1 \rightarrow \infty} E[p,\bv](t_1) = \langle h_s, P_s p \rangle_s = \langle P_s p, \Lambda_s P_s p \rangle_s. 
+\]
 That is, the value of the quadratic form defined by
-$\Lambda^{\pm}_{z_s}$, evaluated at the pressure trace on $z=z_s$,
+$\Lambda_s$, evaluated at the pressure trace on $\Sigma_s$,
 gives the total energy transferred from the source to the
 acoustic field over time. Since $E$ is itself a positive definite
-quadratic form in the acoustic field, it follows that $\pm
-\Lambda^{\pm}_{z_s}$ is positive semi-definite.
-
-While $\Lambda^{\pm}_{z_s}$ is positive semi-definite, it is not
+quadratic form in the acoustic field, it follows that $
+\Lambda_s$ is positive semi-definite.
+
+\subsection{Approximate symmetry of $\Lambda$}
+Except in special cases, $\Lambda_{s,r}$ is not
 symmetric. However, it is {\em approximately symmetric} in the
-high-frequency sense. This fact follows from a
-geometric optics analysis of the half-space solution. This leads to
-the identification of $\Lambda^{\pm}_{z_s}$ as a {\em
-  pseudodifferential operator} of order zero on $z=z_s$, with principal symbol
-\begin{equation}
-  \label{eqn:lamsym}
-  \sigma_0(\Lambda^{\pm}_{z_s}) = \pm 2(\kappa(\bx)
-  \rho(\bx))^{1/2}\left(1-\frac{\kappa(\bx)(\xi^2+\eta^2)}{\rho(\bx)\omega^2}\right)^{-1/2}.
-\end{equation}
-Here $\xi$, $\eta$, and $\omega$ are the dual Fourier variables to
-$x$, $y$, and $t$ respectively. The downgoing assumptions means that
-for local planewave components of $P_sp$, the quantity inside the
-square root is positive. Thus $\Lambda^{\pm}_{z_s}$ has real principal
-symbol (in fact, the entire symbol is real) hence defines an
-asymptotically symmetric operator:
+high-frequency sense. This fact follows from a high-frequency
+asymptotic analysis, a sketch of which follows for the special case of
+a planar source surface
+$\Lambda_s=\{z=z_s\}$, as in the two examples. Since the same reasoning
+applies to $\Lambda_r$, I will omit the subscript for the remainder of
+this subsection.
+
+The pressure trace
+$p(x,y,t)$ can be expressed by Fourier synthesis as
+\begin{equation}
+  \label{eqn:four}
+p(x,y,t) = \frac{1}{(2\pi)^3}\int \,d \omega \,d\xi \, d\eta \,\omega^2
+\,d\eta\,\hat{p}(\omega,\xi,\eta) e^{i\omega(t-x\xi - y\eta)}
+\end{equation}
+in which the spatial frequencies have been expresed as slownesses
+$\xi,\eta$ scaled by temporal frequency $\omega$.
+Choose $a_0(x,y,t)$ so that $a_0(x,y,t) = 1$ if $p(x,y,t) \ne 0$. Then
+$p(x,y,t) = a_0(x,y,t)p(x,y.t)$
+\begin{equation}
+  \label{eqn:fourloc}
+=\frac{1}{(2\pi)^3}\int \,d\omega \, d\xi \, d\eta \, \omega^2
+\,d\eta\,\hat{p}(\omega,\xi,\eta) [a_0(x,y,t) e^{i\omega(t-x\xi -
+  y\eta)}]
+\end{equation}
+This identity expresses $p(x,y,t)$ as a sythesis of
+oscillatory wave packets of the form
+\begin{equation}
+  \label{eqn:pwp}
+ a_0(x,y,t) e^{i\omega(t-x\xi-y\eta)}
+\end{equation}
+in which the spatial frequencies have been expressed as slownesses
+$\xi,\eta$ scaled by temporal frequency $\omega$. The function $a_0$
+is a frequency-independent envelope, equal to 1 where $h$ is
+non-zero.
+
+The wave packet $a_0(x,y,t) e^{i\omega(t-x\xi-y\eta)}$ can be
+viewed as the boundary value on $\Sigma$ of a
+high-frequency asymptotic solution of the form \ref{eqn:go}, provided that the
+slowness vector $(\xi, \eta)$ satisfies the propagating condition
+$c^2(\bx)(\xi^2+\eta^2) < 1$ at all $\bx$ for
+which $a_0(\bx,t)$ is non-zero for any $t$. [In general, this is a
+strong condition, and might not even have any solutions. However it
+always has solutions if $a_0(x,y,t)$ is nonzero only over a sufficiently
+small region, which is the case is $p(x,y,t)$ is non-zero over a
+sufficiently small region. An arbitrary $p(x,y,t)$ can always be
+broken up into a sum, the summands of which are non-zero only over
+small regions, so in fact the requirement mentioned above is not a
+significant constraint: just carry out the following construction for
+summands, and add up the results.] In that case, two
+solutions $\psi$ of the eikonal equation $|\nabla \psi|=c$ exist, at least locally near
+$\Sigma$, for which $\psi(x,y,0)=x\xi+y\eta$. Choose the solution
+that increases in the direction of the outward unit normal ${\bf
+  n}$, and solve the corresponding transport equation for an
+amplitude $a$ satisfying $a(x,y,0,t) = a_0(x,y,t)$. Then the geometric optics solution $a e^{i\omega(t-\psi)}$ is
+asymptotic to the pressure field of an incoming solution. From the
+evolution equation for $\bv$, integrated in time as in equation
+\ref{eqn:newt},  an expression follows for the corresponding normal
+velocity:
+\[
+  -{\bf n} \cdot \frac{1}{\rho}\int_{-\infty}^t ds \nabla (a
+  e^{i\omega(s-\psi)})=\frac{1}{\rho}\int_{-\infty}^t ds {\bf n}
+  \cdot(i\omega a\nabla \psi -\nabla a) e^{i\omega(s-\psi)})
+\]
+Now use the identity
+\[
+  e^{i\omega(s-\psi)} = \frac{1}{i\omega}\frac{d}{ds}e^{i\omega(s-\psi)}
+\]
+and integrate by parts in $s$ to obtain the expression for
+the normal velocity field 
+\[
+  = \frac{a}{\rho} ({\bf n} \cdot \nabla \psi)  e^{i\omega(t-\psi)} + O(1/\omega).
+\]
+The terms not written explicitly begin with one of order -1 in
+$\omega$. The remainder appears to be also of order -1, but
+application of the integration-by-parts trick described above yields
+another term of order -2 plus another remainder of apparent order
+-2. Continuing this process develops the $O(1/\omega)$ remainder in an
+asymptotic series of homogeneous terms in $1/\omega$ of arbitrary
+length.
+
+An expression for the normal velocity on
+$\Sigma$ follows by restriction. Note that ${\bf n} \cdot \nabla \psi =( c^{-2}(x,y,z)-
+(\xi^2+\eta^2))^{1/2}$ because of the choice of sign and the eikonal
+equation, and the root is real because of the propagating assumption.
+
+The upshot of this calculation is that to leading order in frequency,
+the normal velocity corresponding to a pressure wave packet differs by a position- and slowness-dependent
+factor. Accounting for the factor of 2 difference between the source
+amplitude and velocity trace, this calculation may be summarized:
+\begin{equation}
+  \label{eqn:lamwp}
+  \Lambda (a_0e^{i\omega(t-x\xi-y\eta)}) = \lambda
+  (a_0e^{i\omega(t-x\xi-y\eta)}),
+\end{equation}
+where $\lambda(x,y,t,\xi,\eta,\omega)$ is
+defined by the asymptotic series in $\omega$ developed above. Its dominant term,
+$O(1)$ (order 0) in $\omega$, is
+\[
+  \lambda^0(x,y,t,\xi,\eta,\omega) = 2\frac{( c^{-2}(x,y,z)-
+    \xi^2+\eta^2)^{1/2}}{\rho} + O(1/\omega)
+\]
+\begin{equation}
+  \label{eqn:lamsymb}
+= 2(\kappa
+\rho)^{-1/2}\left(1-\frac{\kappa(k_x^2+k_y^2)}{\rho \omega^2}\right)^{1/2} + O(1/\omega).
+\end{equation}
+In the final expression, the slownesses $\xi,\eta$ have been replaced
+by the Fourier variables $k_x=x\xi, k_y=y\eta$. Note that as a
+function of $k_x,k_y,\omega$, $\lambda^0$ is homogeneous of order
+0. It is easy to see that the $N$th term of the asymptotic series
+defining $\lambda$ is of homogeneous order $-N$.
+
+%%%%%%%%%%%
+% discussion of psidos
+%
+Operators that affect oscillatory wave packets by multiplication with
+functions having the properties of $\lambda$ have come to
+be called ``pseudodifferential'' \cite[]{Nir:72,Tay:81,SaiR:91}. The multiplier ($\lambda$ in this
+case) is called the {\em symbol} of the operator, and the dominant
+term in an asymptotic development of the multiplier
+(for instance $\lambda^0$) is called the {\em principal symbol}.
+Amongst other special properties of these operators, their adjoints
+are operators of the same type. In particular, the principal symbol of the adjoint is
+the Hermitian (matrix) adjoint of the principal symbol. For scalar
+operators like $\Lambda$ (mapping scalar-valued functions to
+scalar-valued functions), the symbols are scalar ($1 \times 1$), and
+the Hermitian adjoint is simply the complex conjugate (\cite{Tay:81},
+section II.4). Therefore scalar operators with real principal symbol,
+like $\Lambda$, have the same principal symbols as their adjoints. So
+$\Lambda$ differs from $\Lambda^T$ by a pseudodifferential operator
+whose symbol is an asymptotic sum of terms of negative order in
+frequency. Therefore the difference is negligible in the limit of high
+frequencies. This observation establishes the conclusion of this subsection: 
 \begin{equation}
   \label{eqn:lamappsim}
-  (\Lambda^{\pm}_{z_s})^T \approx \Lambda^{\pm}_{z_s}.
-\end{equation}
-(For more on this, see \cite{StefUhl:05}.)
-The analysis also reveals that the solution components not continuous
-at $z=z_s$ are odd there:
-\begin{equation}
-  \label{eqn:odd1}
-  \lim_{z\rightarrow z_s^+} v^{\pm}_{z} \approx - \lim_{z\rightarrow z_s^-}
-  v^{\pm}_{z}
-\end{equation}
-for the solution of \ref{eqn:awepm} with $f_s=0$.
-Similarly, 
-\begin{equation}
-  \label{eqn:odd2}
-  \lim_{z\rightarrow z_s^+} p^{\pm}\approx - \lim_{z\rightarrow z_s^-}
-  p^{\pm}
-\end{equation}
-for the solution of \ref{eqn:awepm} with $h_s=0$. Here ``$\approx$''
-means in the sense of high frequency asymptotics, that is, that the
-difference between the two sides is relatively smooth, hence small if
-the data is highly oscillatory. Therefore if $f_s=0$ in system \ref{eqn:awepm},
-\begin{equation}
-  h_s = \Lambda^{\pm}_{z_s}P_sp^{\pm} = -[v^{\pm}_{z}]|_{z=z_s} \approx -2
-  P_sv^{\pm}_{z}
-  \label{eqn:tracejump10}
-\end{equation}
-Similarly, if $h_s=0$ in system \ref{eqn:awepm}, then
-\begin{equation}
-  \label{eqn:tracejump20}
-  f_s = -[p^{\pm}]|_{z=z_s} \approx -2 P_s p^{\pm}.
-\end{equation}
-Thus $f_s$ determines approximately the boundary value of $p^{\pm}$,
-as a solution of the acoustic wave system in the half-space
-$z>z_s$. However, as repeated in equation \ref{eqn:tracejump10}, a
-solution with this boundary value is also the restriction to $z>z_s$
-of a solution to \ref{eqn:awepm} with $f_s=0$ and $h_s=
-\Lambda^{\pm}_{z_s}P_sp^{\pm}$. Therefore if
-\begin{equation}
-  \label{eqn:hfcondn}
-  h_s =-\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,
-\end{equation}
-then the pressure boundary value $P_sp^{\pm}$ is the
-same for the solutions of \ref{eqn:awepm} for source vectors $(h_s,0)$
-and $(0,f_s)$. Since the pressure boundary values are the same, the solutions
-in $z>z_s$ are the same. In particular, since $z_r>z_s$ and ${\cal
-  S}^{\pm}_{z_s,z_r}(h_s,f_s)^T = (P_rp^{\pm},P_rv^{\pm}_z)^T$, it follows
-that
+  \Lambda^T \approx \Lambda,
+\end{equation}
+
+Before leaving this topic, recall that the computation outlined above
+was predicated on the propagating assumption, namely that the slowness
+parameters $\xi,\eta$ of the wave packet defined in \ref{eqn:pwp}
+satisfy $c(x,y,z_s)^2(\xi^2 + \eta^2) < 1$ whereever the amplitude
+$a_0(x,y,t)$ is non-zero. Therefore the conclusion \ref{eqn:lamappsim}
+holds only for application to such wave packets. However, the restriction of
+propagating waves (such as those that are output by the modeling
+operator $S$ or its adjoint $S^T$) are well-approximated by sums over
+propagating wave packets: this is part of the assumption that the rays
+carrying high-frequency energy cross the source and receiver surfaces
+at non-zero angles. To make the definition of $\Lambda$ complete, it
+must be extended to non-propagating wave packets, for instance by
+making its effect negligible via a dip filter. I will ignore this refinement, since the
+application to non-propagating components appears to play little role
+in approximate inversion and its applications as developed here.
+
+\section{Computing $\Lambda$}
+
+The missing ingredient in the preceding discussion is any indication
+of how the action of the pressure-to-source operators $\Lambda$ is to be
+computed. As will be seen in the next section, an (approximate)
+inverse of this operator is also required in the formulation of
+preconditioned Krylov iteration.
+
+Direct computation of the pressure-to-source operator $\Lambda_s$, for instance, by solving
+\ref{eqn:awepm} and reading off $P_sv_{z}$, turns out to be
+numerically ill-behaved. An alternate approach begins with the
+observation that a pressure gather on $\Sigma_r$ produced by a
+pressure source $h_s$ on $\Sigma_s$ can also be well-approximated by
+the field from a velocity source. To see this, recall the first major
+step in the time reversal construction, that a solution of the wave
+equation near $\Sigma_s$ can be represented as the solution of a
+boundary value problem with $p$ given on $\Sigma_s$ (system
+\ref{eqn:aweloc}). The derivation combined the unique determination of
+the field by its pressure boundary values on an enclosing domain with
+the consequences of the standing assumptions about rays carrying high
+frequency energy. Precisely the same logic could have been applied to
+the normal velocity boundary values, as these also uniquely determine
+the field in an enclosing domain. Proceeding further, a source
+representation based on continuous extension of normal velocity, with
+a jump in the pressure follows by logic precisely analogous to that
+leading to the system \ref{eqn:awesrc}: instead, obtain a field
+$(\tilde{p},\tilde{\bv})$ differing negligibly from the field of the
+same name in the earlier discussion, solving 
+\begin{eqnarray}
+\label{eqn:awesrcvzmod}
+  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & 
+                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
+  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & f_s {\bf n}\delta_{\Sigma}- \nabla \tilde{p}, \nonumber \\
+  \tilde{p} & = & 0 \mbox{ for }  t \ll 0, \nonumber \\
+  \tilde{\bv} & = & {\bf 0} \mbox{ for }  t \ll  0.       
+\end{eqnarray}
+(that is, system \ref{eqn:awepm} with $h=0$) with $f_s = 2P_sp$.
+
+This fact can usefully be rewritten using the vector modeling
+operator ${\cal S}$ (definition \ref{eqn:fwd}), and the projection
+operators $\Pi_i$ (definition \ref{eqn:pidef}), which insert scalar
+sources into the pairs that occur in the acoustic system
+\ref{eqn:awepm}: explicitly, $\Pi_0(h,f)^T = h, \Pi_1(h,f)^T = f,
+\Pi_0^Th = (h,0), \Pi_1^Tf = (0,f)$. Then
+\begin{eqnarray}
+  P_r\tilde{p} &=& \Pi_0 {\cal S} \Pi_1^Tf_s, \nonumber \\
+  P_r({\bf n}_r \cdot \tilde{\bv}) &=& \Pi_1 {\cal S} \Pi_1^Tf_s.
+                                       \label{eqn:ptildef}
+\end{eqnarray}
+Since the field $(\tilde{p},\tilde{\bv})$ constructed here differs negligibly from that
+constructed earlier with the same name, which in turn differs negligibly from $p$, it
+follows that $P_r\tilde{p} \approx P_rp = Sh_s$. By definition of $S$
+(\ref{eqn:sdef}), this is the same as $=\Pi_0{\cal
+  S}\Pi_0^Th_s$. Since the velocity fields also differ negligibly,
+obtain
+\begin{eqnarray}
+  P_r\tilde{p} & \approx & \Pi_0 {\cal S} \Pi_0^Th_s, \nonumber \\
+  P_r({\bf n}_r \cdot \tilde{\bv}) & \approx & \Pi_1 {\cal S} \Pi_0^Th_s.
+                                       \label{eqn:ptildeh}
+\end{eqnarray}
+
+From the definition of $\Lambda_s$, $h_s=\Lambda_sP_sp =
+\frac{1}{2}\Lambda_sf_s$. Combining \ref{eqn:ptildef},
+\ref{eqn:ptildeh}, and this last identity, obtain
 \begin{equation}
   \label{eqn:snull}
-  {\cal S}^{\pm}_{z_s,z_r}\left(\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,f_s\right)^T \approx 0.
-\end{equation}
-
-Equation \ref{eqn:snull} states the relation between the columns of $
-{\cal S}^{\pm}_{z_s,z_r}$ mentioned in the introduction to this
-section.
-
-\section{Unitarity}
-
-The next chapter in this story recognizes the relations in display
-\ref{eqn:approxinv} as asserting the approximate unitarity of
-$S^+_{z_s,z_r}$.
-
-The matrix identity \ref{eqn:snull} implies a relation between $S, V,$
-and $\Lambda$ of some interest in itself. After minor re-arrangement, the second row of reads
-\begin{equation}
-  \label{eqn:snull2}
--\frac{1}{2}\Pi_1{\cal S}^{\pm}_{z_s,z_r}\Pi_0^T\Lambda^{\pm}_{z_s}  \approx
-V^{\pm}_{z_s,z_r}.
-\end{equation}
-In these relations, the projection on the left picks out the vertical velocity component
-of a downgoing wavefield at $z=z_r$: that is,
+  {\cal S}\left(-\frac{1}{2}\Lambda_{z_s}f_s,f_s\right)^T \approx 0.
+\end{equation}
+
+That is, the same data gather on $\Sigma_r$ can be obtained either
+with a pressure source $h_s$ or with a velocity source $f_s$, and
+there is a definite relation betwen the two. To illustrate this
+relation, recall that for the first of the two examples introduced
+above, the velocity source $f_s$ is twice the pressure trace shown in
+Figure \ref{fig:dsrcphh0}, and the pressure source (Figure \ref{fig:dhshh0}) is twice the normal
+velocity trace shown in Figure \ref{fig:dsrcvzhh0}.
+Figure \ref{fig:daltplh0} shows $\Pi_0{\cal S}\Pi_1^Tf_s$, Figure
+\ref{fig:dfwdplh0bis} shows $\Pi_0{\cal S}\Pi_0^Th_s$, and Figure
+\ref{fig:ddifffwdaltplh0} shows the difference on the same color scale.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:dsrcphh0}]{\includegraphics[width=0.45\textwidth]{dsrcphh0.pdf}}
+  \subfloat[\label{fig:dsrcvzhh0}]{\includegraphics[width=0.45\textwidth]{dsrcvzhh0.pdf}}
+  \caption{(a) Pressure gather for Lens example on source surface
+    $\Sigma_s=\{z=1000\mbox{ m}\}$, generated by point source at
+    $(3500,3500)$ m. (b) Normal velocity gather for Lens example on source surface
+    $\Sigma_s=\{z=1000\mbox{ m}\}$, generated by point source at
+    $(3500,3500)$ m.}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:dfwdplh0bis}]{\includegraphics[width=0.32\textwidth]{dfwdplh0.pdf}}
+  \subfloat[\label{fig:daltplh0}]{\includegraphics[width=0.32\textwidth]{daltplh0.pdf}} 
+  \subfloat[\label{fig:ddifffwdaltplh0}]{\includegraphics[width=0.32\textwidth]{ddifffwdaltplh0.pdf}}
+  \caption{(a) Pressure gather on receiver surface
+    $\Sigma_r=\{z=1000\mbox{ m}\}$ due to extended pressure source = 2
+    $\times$ normal velocity gather shown in Figure
+    \ref{fig:dsrcvzhh0}. This is the same gather as shown in Figure
+    \ref{fig:dfwdplh0}, and the extended pressure source has been
+    shown in Figure \ref{fig:dhshh0}.
+    (b) Pressure gather on receiver surface
+    $\Sigma_r=\{z=1000\mbox{ m}\}$ due to extended velocity source = 2
+    $\times$ pressure gather shown in Figure
+    \ref{fig:dsrcvzhh0}. (c) Difference (b)-(a), plotted on same color
+    scale.}
+\end{figure}
+
+One other important consequence of this duality is an alternate view
+of the time reversal construction of the approximate inverse. The
+entire construction can be carried out with traces of the normal
+velocity in place of trace of the pressure. The steps are precisely
+analogous, and justified in exactly the same way. Therefore I will
+simply summarize the result:
+to invert a pressure gather $d$ on $\Sigma_r$
+for a source $f_s$ on $\Sigma_s$ that approximately reproduces it ($d
+\approx \Pi_0{\cal S}\Pi_1^Tf_s$):
+\begin{itemize}
+\item[1. ] create a source $f_r = - 2d$ on $\Sigma_r$;
+\item[2. ] apply the adjoint $(\Pi_1{\cal S}\Pi_1^T)^T=\Pi_1{\cal
+    S}^T\Pi_1^T$ (solve the reverse time system
+  \begin{eqnarray}
+  \label{eqn:awerecglobvz}
+  \frac{1}{\kappa}\frac{\partial\hat{p}}{\partial t} & = & 
+                                                      - \nabla \cdot \hat{\bv}, \nonumber \\
+  \rho\frac{\partial \hat{\bv}}{\partial t} & = & f_r {\bf n}r
+  \delta_{\Sigma_r} - \nabla \hat{p}, \nonumber \\
+  \hat{p} & = & 0 \mbox{ for } t\gg 0, \nonumber \\
+    \hat{\bv} & = & {\bf 0} \mbox{ for } t \gg 0.
+  \end{eqnarray}
+\item[3. ]
+  then extract $h_s=2{\bf n_s}\cdot P_s\hat{\bv}. = 2\Pi_1{\cal
+    S}^T\Pi_1^Tf_r=-4\Pi_1{\cal S}^T\Pi_1^Td$.
+\end{itemize}
+The upshot of this construction is the alternate computation for the
+approximate inverse:
+\begin{equation}
+  \label{eqn:altinv}
+  S^{\dagger}  \approx -4\Pi_1{\cal S}^T\Pi_1^T.
+\end{equation}
+The first row of
+\ref{eqn:snull}, slightly rearranged, reads
+\begin{equation}
+  \label{eqn:lamidea}
+  \Pi_0{\cal S}\Pi_1^T f_s \approx  \frac{1}{2} S\Lambda_sf_s.
+\end{equation}
+Apply the approximate inverse construction in the form
+\ref{eqn:altinv} to solve \ref{eqn:lamidea} for $\Lambda_s$:
+\begin{equation}
+  \label{eqn:lamident}
+  \Lambda_s \approx -8\Pi_1{\cal S}^T\Pi_1^T\Pi_0 {\cal S}\Pi_1^T.
+\end{equation}
+This identity is the major result of this section: it shows how
+to compute that action of $\Lambda_s$ by propagating the input
+pressure trace, identified as a source for the velocity evolution,
+forward in time 
+reading off the pressure trace on $\Sigma_r$, identifying it once more as
+a point load (source for velocity), propagating it backwards in time, and finally reading off the velocity trace, scaled to be a
+pressure evolution source on $\Sigma_s$.
+
+The importance of this result lies in the failure of the obvious
+method for computing the action of $\Lambda_s$, namely to
+employ the pressure trace as a source in the velocity equation ($f_s$,
+in the notation used above) at $\Sigma_s$, and read off the velocity
+field also at $\Sigma_s$. This difficulty is related to the existence of
+tangentially propagating waves and the lack of continuity of the trace
+operator. The method implicit in equation \ref{eqn:lamident} avoids
+this difficulty by propagating the fields a positive distance in $z$:
+assuming as always that the causal fields are downgoing, this step
+eliminates any tangentially propagating fields from consideration.
+
+It is implicit in the discussion above of the pressure-to-source operator 
+depends only on the model coefficients near the source surface
+$\Sigma_s$. However an even more useful
+observation is that the calculations in the approximation
+\ref{eqn:lamident} could just as well be carried out in a much smaller
+region around the source surface, and produce a result that is
+functionally identical in that it will serve as a source for the same
+acoustic fields globally, with small error. Take advantage of this
+observation by replacing - for the purpose of evaluating the
+expression \ref{eqn:lamident} only -  the receiver surface by an
+auxiliary receiver surface $\Sigma_a$ near $\Sigma_s$ but a small positive
+distance from it. Define the auxiliary operator ${\cal S}_a$ by
+sampling solution of the system \ref{eqn:awepm} on $\Sigma_a$ rather
+than $\Sigma_r$. Thus obtain an approximation for $\Lambda_s$
+\begin{equation}
+  \label{eqn:lamnear}
+ \Lambda_s \approx -8\Pi_1{\cal S}_a^T\Pi_1^T\Pi_0 {\cal S}_a\Pi_1^T. 
+\end{equation}
+
+Using an auxiliary receiver datum closer to the source surface to
+compute that pressure-to-source maps (only!) has two favorable consequences:
+\begin{itemize}
+\item The computational domain can be smaller than is necessary to
+  simulate the target data, as it need only contain the source
+  surface and the receiver datum implicit in
+  equation \ref{eqn:lamident}. If the domain is small enough, the ray
+  conditions imply that no ray re-enters the region between the source
+  suface and the auxiliary surface, therefore an absorbing boundary
+  condition can be applied outside of this region, considerably
+  shrinking the computational domain and substantial improving computational
+  efficiency. The results shown below used such a truncation strategy.
+\item Since the receiver data may be chosen much closer to the
+  source surface that is the case for the target data, the effective
+  aperture active in the relation \ref{eqn:lamident} can be much
+  larger, producing an estimated source gather much less affected by
+  aperture limitation.
+\end{itemize}
+
+Figures
+\ref{fig:preddnshshh0} and \ref{fig:ddiffnshshh0} show an example of
+this construction.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:preddnshshh0}]{\includegraphics[width=0.45\textwidth]{preddnshshh0.pdf}}
+  \subfloat[\label{fig:ddiffnshshh0}]{\includegraphics[width=0.45\textwidth]{ddiffnshshh0.pdf}}
+  \caption{(a) Pressure source gather = image
+  under pressure-to-source operator $\Lambda_s$ of pressure gather
+  shown in Figure \ref{fig:dsrcphh0}, computed using
+  auxiliary receiver surface``near''  at $z=2900$ m and equation \ref{eqn:lamnear}. Compare Figure
+  \ref{fig:dhshh0}: because the sources and
+  receivers are close, little aperture is lost in this case. (b)
+  Difference between data plotted in (a)
+  and source gather
+  inferred from vertical velocity at source
+  surface ($z=3000$ m) (Figure \ref{fig:dhshh0}). Same color scale as in Figure
+  (a).}
+\end{figure}
+
+In the iterative methods to be discussed in the next section, it will
+be important to be able to compute the transpose of the
+pressure-to-source operator - not approximately, but to machine
+precision. The prescription \ref{eqn:lamnear} only
+defines an approximation. However simply substituting the right-hand
+side of equation \ref{eqn:lamnear} - in effect, substituting the
+approximation for the operator that it approximates - gives an
+operator variant of $\Lambda_s$ with an accessible
+machine-precision adjoint:
+\begin{equation}
+  \label{eqn:lamneartr}
+ \Lambda_s^T \approx -8\Pi_1{\cal S}_a^T\Pi_0^T\Pi_1 {\cal S}_a\Pi_1^T. 
+\end{equation}
+From now on, when $\Lambda_s$ or $\Lambda_s^T$ appear, the symbols are
+understood to mean the approximations \ref{eqn:lamnear} and
+\ref{eqn:lamneartr}.
+
+Figure \ref{fig:preddnshstrhh0} shows the image of the pressure gather
+in Figure \ref{fig:dsrcphh0} under $\Lambda_s^T$ approximated as in
+equation \ref{eqn:lamneartr},
+using the auxiliary receiver surface at $z=2900$. Note
+the close resemblance to the image of the same pressure gather under
+$\Lambda_s$ displayed in Figure
+\ref{fig:preddnshshh0}. The difference of these two images is
+displayed in \ref{fig:ddiffnslamtrhh0}, on the same color scale as the
+images themselves. Note that propagation takes place entirely in a
+region where the mechanical parameters are homogenous.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:preddnshstrhh0}]{\includegraphics[width=0.45\textwidth]{preddnshstrhh0.pdf}}
+  \subfloat[\label{fig:ddiffnslamtrhh0}]{\includegraphics[width=0.45\textwidth]{ddiffnslamtrhh0.pdf}}
+  \caption{(a) Pressure source gather = image
+  under transposed pressure-to-source operator $\Lambda_s^T$ of pressure gather
+  shown in Figure \ref{fig:dsrcphh0}, computed using
+  auxiliary receiver surface``near''  at $z=2900$ m and equation \ref{eqn:lamneartr}. Compare Figure
+  \ref{fig:dhshh0}: because the sources and
+  receivers are close, little aperture is lost in this case. (b) Difference between (a) 
+  and source gather inferred from vertical velocity at source
+  surface ($z=3000$ m) (Figure \ref{fig:dhshh0}). Same color scale as (a).}
+\end{figure}
+
+\section{Accelerated Iterative Inversion}
+
+The preceding sections provide an approximate inverse for the modeling
+operator $S$ and a means to compute it effectively. These constitute
+the basic ingredients of an effective preconditioning strategy. The
+remaining tasks are: the construction of weighted norms making $S$
+approximately unitary (insofar as possible), definition of a formal
+preconditioned conjugate gradient algorithm; and modification of the
+symmetrized approximate inverse to account for the presence of the
+penalty term. Examples illustrate the effectiveness of the resulting
+preconditioning strategy.
+
+\subsection{Symmetrization and Weighted Norms}
+Since $\Lambda_s$ approximates its transpose (equation \ref{eqn:lamappsim}), it also approximates its
+ symmetrization:
+\begin{equation}
+  \label{eqn:lamsymm}
+  \frac{1}{2}(\Lambda_s^T+\Lambda_s) \approx \Lambda_s.
+\end{equation}
+and similarly for $\Lambda_r$.
+
+Define
+\begin{eqnarray}
+  W_m^{-1}&=& \frac{1}{2}(\Lambda_s^T+
+              \Lambda_s),\nonumber \\
+  W_d &=& \frac{1}{2}(\Lambda_r^T+ \Lambda_r)
+          \label{eqn:wdef}
+\end{eqnarray}
+(the subscripts signify ``model space'' and ``data space'' of $S$).
+These operators define (semi)-norms in the spaces of finite energy
+pressure fields on $\Sigma_s$ and $\Sigma_r$:
 \[
--\frac{1}{2}\Pi_1{\cal S}^{\pm}_{z_s,z_r}\Pi_0^T\Lambda^{\pm}_{z_s}P_sp^+
-=-\frac{1}{2}P_r v_z^+,
+  \|p_s\|^2 = \langle p_s, W_m p_s\rangle,  \|p_r\|^2 = \langle p_r,
+  W_d p_r\rangle,
 \] 
-where $(p^+, \bv^+)$ solve the system \ref{eqn:awepm} with $f_s=0$ and
-$h_s = \Lambda^{\pm}_{z_s}P_sp^+$. On the other hand, from relation
-\ref{eqn:tracejump10},
+The identification of the symmetrized $\Lambda_s$ as the inverse
+of another operator $W_m$ is formal at this point, since the former operator is
+likely to have null (or nearly-null) vectors due to aperture-related
+amplitude loss, and certainly due to asymptotic vanishing at
+non-propagating waves. Since some version of $W_m$ is essential in the
+formulation for effective preconditioning, I will derive a usable
+candidate to stand in for it below.
+
+The adjoint of $S$ with respect to the weighted norms is given
+approximately by
+\begin{equation}
+\label{eqn:wadj}
+S^{\dagger} \approx  W_m^{-1}S^TW_d,
+\end{equation}
+As we have shown (equations \ref{eqn:baseinv}, \ref{eqn:rightinv}),
+$S^{\dagger}$ is an approximate right inverse to $S$: repeating
+equation \ref{eqn:rightinv} for convenience,
+\begin{equation}
+  \label{eqn:wunitary}
+  SS^{\dagger} \approx I.
+\end{equation}
+This is not quite enough to establish that
+$S$ is approximately unitary with respect to the
+weighted norms defined by $W_m$ and $W_d$, since that would require
+that $S^{\dagger}$ also be an approximate left inverse. The ray
+conditions assumed throughout include the presumption that all rays
+carrying high frequency energy arriving at the receiver surface
+(within the support of the sampling operator) pass
+over the source surface. Consequently a source supported on the source
+surface exists that produces this data. However it was not assumed
+that all rays passing over the source surface arrive at the receiver
+surface. Therefore high frequency energy can be emitted from the
+source surface without arriving at the receiver surface. We have seen
+examples of this phenomenon already. The extended source shown in Figure
+\ref{fig:ndwdataresrcplbulksm0} generates the diving wave data shown
+in Figure \ref{fig:ndwdata0} to close approximation, even though the
+latter was generated by a point source. The difference between the
+extended source in Figure \ref{fig:ndwdataresrcplbulksm0} and the
+point source depicted in Figure \ref{fig:ooplrays0} is an approximate
+null vector for the modeling operator $S$ as defined for this
+example.
+
+In fact, that is true for the difference between $S^{\dagger}Sh$ and
+$h$ for any extended source $h$:
 \[
-  P_r v_z^+ = -\frac{1}{2}\Lambda^+_{z_r}P_r p^+
+  S(S^{\dagger}Sh - h) = (SS^{\dagger}-I)Sh \approx 0.
 \]
-where
+Suppose that $n$ is an approximate null vector for $S$, and $h$ is any
+other extended source. Then
 \[
-  P_r p^+ = \Pi_0{\cal S}^{+}_{z_s,z_r}\Pi_0^T\Lambda^{+}_{z_s}P_s
-  p^+
+  \langle S^{\dagger}S h, n \rangle_m = \langle Sh, Sn \rangle_d
+  \approx 0.
 \]
-\[
-  = S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_sp^+
-\]
-Therefore combining the last two equations with \ref{eqn:snull2},
-obtain
-\begin{equation}
-  \label{eqn:sv}
-  \frac{1}{4}\Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s} = V^+_{z_s,z_r}.
-\end{equation}
-This is the promised relation.
-
-As shown in the last section, $4(V_{z_s,z_r}^+)^T$ is approximately
-inverse to $S^{+}_{z_s,z_r}$. Therefore, transposing both sides of
-equation \ref{eqn:sv} and using \ref{eqn:approxinv}, obtain
-\begin{equation}
-  \label{eqn:almostunitary}
-  4(V_{z_s,z_r}^+)^TS^+_{z_s,z_r} = [ (\Lambda^+_{z_s})^T
-  (S^{+}_{z_s,z_r})^T(\Lambda^+_{z_r})^T]S^{+}_{z_s,z_r} \approx I.
-\end{equation}
-
-The remarkable feature of the identity \ref{eqn:almostunitary} is that
-it exhibits an approximate right inverse of $S^+$ as an adjoint with
-respect to a weighted inner product - or it would, if the operators
-$(\Lambda^+)$ were symmetric positive definite. As noted earlier,
-these operators are only approximately symmetric, though they are
-positive semi-definite. That is not a great obstacle, however:
-symmetrizing them in the obvious way commits a negligible error, of
-the sort that this paper already neglects wholesale. That is,
-\begin{equation}
-  \label{eqn:unitary}
-  [ \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
-  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
-  \Lambda^+_{z_r})]S^{+}_{z_s,z_r} \approx I.
-\end{equation}
-
-The symmetrized $\Lambda$ operators are at least positive
-semi-definite, hence define (at least) semi-norms.
-Similar relations have been derived for other scattering operators,
+That is, $S^{\dagger}S$ is the (approximate) orthogonal projector onto
+the orthocomplement of the null space of $S$. So iterations such as CG
+based on $S^{\dagger}S$ construct approximation sequences lying close
+to this orthocomplement, as if the null space did not exist. This
+observation allows us to treat $S$ as if it were unitary in discussing
+the behaviour of Krylov space algorithms based on the weighted spaces,
+even though it actually is not.
+
+Similar relations have been derived for other wave-theoretic modeling operators,
 and have been used to accelerate iterative solutions of inverse
@@
 
-\section{Accelerated Iterative Inversion}
-
-For convenience, in this section write $S$ in place of
-$S^+_{z_s,z_r}$. Also abbreviate the symmetrized $\Lambda$ operators
-using notation suggesting weight
-operators in model and data spaces:
-\begin{eqnarray}
-  W_m^{-1}&=& \frac{1}{2}((\Lambda^+_{z_s})^T+
-              \Lambda^+_{z_s}),\nonumber \\
-  W_d &=& \frac{1}{2}((\Lambda^+_{z_r})^T+ \Lambda^+_{z_r}).
-          \label{eqn:wdef}
-\end{eqnarray}
-The identification of the symmetrized $\Lambda^+_{z_s}$ as the inverse
-of another operator $W_m$ is formal, since the former operator is
-likely to have null (or nearly-null) vectors due to aperture-related
-amplitude loss. Since some version of $W_m$ is essential in the
-formulation for effective preconditioning, I will derive a usable
-candidate to stand in for it below.
-
-Adopting Hilbert norms defined by the operators $W_m$ and $W_d$ in its
-domain and range respectively, the adjoint of $S$ is given by
-\begin{equation}
-\label{eqn:wadj}
-S^{\dagger} = W_m^{-1}S^TW_d,
-\end{equation}
-
-In this notation, the relation \ref{eqn:unitary} takes the form
-\begin{equation}
-  \label{eqn:wunitary}
-  S^{\dagger}S \approx I.
-\end{equation}
-That is to say, $S$ is approximately unitary with respect to the
-weighted norms defined by $W_m$ and $W_d$. Therefore a Krylov space
-method employing these norms will converge rapidly, at least for the
-well-determined components of the solution.
+Concerning computation, just as for the pressure-to-source operators,
+the weight operators are to be understood as replaced by their
+approximations via equations \ref{eqn:lamnear} and \ref{eqn:lamneartr}:
+\[
+W_m^{-1} = \frac{1}{2}(\Lambda_s +
+\Lambda_s^T)
+\]
+\begin{equation}
+  \label{eqn:winvcomp}
+  -4\Pi_1 {\cal S}_a^T(\Pi_0^T\Pi_1+ \Pi_1^T\Pi_0){\cal S}_a\Pi_0.
+\end{equation}
+with a similar approximate realization of $W_d$.
+
+Figure \ref{fig:symmdnshshh0} shows the output of the symmetrized
+approximate pressure-to-source operator per equation \ref{eqn:winvcomp},
+applied once again to the pressure data in Figure
+\ref{fig:dsrcphh0}. Note the resemblance to Figures
+\ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}. These are all
+asymptotic approximations of each other. Figure
+\ref{fig:ddiffsymmdnshsll0} shows the
+ the difference between the pressure gather at $z=z_r$ produced from
+ the pressure source output by the symmetrized $\Lambda$, and the point source
+simulation (Figure \ref{fig:dfwdplh0}), plotted on the same scale as
+the latter, in both cases with all propagations in the lens model.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:symmdnshshh0}]{\includegraphics[width=0.45\textwidth]{symmdnshshh0.pdf}}
+  \subfloat[\label{fig:ddiffsymmdnshsll0}]{\includegraphics[width=0.45\textwidth]{ddiffsymmdnshsll0.pdf}}
+  \caption{(a) Pressure source gather = image
+  under symmetrized pressure-to-source operator $W_m^{-1}$ applied to pressure gather
+  shown in Figure \ref{fig:dsrcphh0}, computed using
+  auxiliary receiver surface``near''  at $z=2900$ m and 
+  equation \ref{eqn:winvcomp}. Compare Figures
+  \ref{fig:dhshh0}, \ref{fig:preddnshshh0} and
+  \ref{fig:preddnshstrhh0}: these are all asymptotic approximations of
+  each other. (b) Difference between pressure gather at the receiver
+  surface $\Sigma_r$  ($z=1000$ m) modeled using symmetrized
+  pressure-to-source gather in (a), and pressure gather shown in Figure \ref{fig:dfwdplh0},
+  propagated in lens model from 
+  source gather shown in Figure \ref{fig:dhshh0} (inferred from vertical velocity at source
+  surface ($z=3000$ m)). Same color scale as
+  Figures \ref{fig:drecplh0} and \ref{fig:dfwdplh0}.}
+\end{figure}
+
+The identity \ref{eqn:winvcomp} shows that only one forward and one adjoint simulation
+are necessary to compute the action of $W_m^{-1}$ or $W_d$, and only
+over the truncated computational domain between the source or receiver
+surface and its auxiliary, nearby surface. The operator
+in the center of the expression on the right-hand side, $\Pi_0^T\Pi_1
++ \Pi_1^T\Pi_0$, simply exchanges the components of the acoustic
+fields, passing the velocity field as a pressure source and the
+pressure field as a velocity source.
+
+One more computation is required for the full implementation of the
+preconditioning strategy explained in the last section: $W_m$ is
+required, not just $W_m^{-1}$. Note that $W_m$ plays two roles in the
+second term in equation \ref{eqn:norm1}, below: it is the weight matrix for
+both the domain and range norms for $A$. It is perfectly OK for one of
+these to be replaced by an asymptotic approximation, so long as it is
+symmetric and computable (and at least semi-definite). Manipulations
+very similar to those used to obtain equation \ref{eqn:winvcomp} lead
+to the approximation
+\begin{equation}
+  \label{wcomp}
+ W_m \approx -\frac{1}{16}\left(\Pi_0 {\cal S}_a^T
+   (\Pi_1^T \Pi_0 + \Pi_0^T \Pi_1) {\cal S}_a \Pi_0^T\right)
+\end{equation}
+Comparison with the definition \ref{eqn:winvcomp} shows that
+$W_m$ and $W_m^{-1}$ differ only in the initial and
+final projection factors (and overall scale), and in particular either
+can be computed for the cost of a forward/adjoint operator pair. Note
+that $W_m^{-1}$ is inverse to $W_m$ only in an
+approximate (asymptotic, aperture-limited) sense.
+
+\subsection{Preconditioned Conjugate Gradient Iteration}
 
@@
 d, W_d d\rangle$ has physical meaning: for acoustics, it is
-proportional to the power transmitted to the fluid by the source.
+proportional to the power transmitted to the fluid by the source
+corresponding to $d$.
 
@@
 \end{equation}
-
-Note that the normal operator appearing on the left-hand side of
-\ref{eqn:norm0} is not an approximate identity, due to the presence of
-the regularization term: the spectrum increases in spread with
-increasing $\alpha$, leading to slower convergence. Fortunately for
-the present setting, the operators $W_m^{-1}$, $A$, and $W_m$
-approximately commute (they are scalar {\em pseudodifferential}, once
-the difficulties with the definition of $W_m$, mentioned above, are
-taken care of). Scalar pseudodifferential operators approximately
-commute, so $A^{\dagger} \approx A^T$. Therefore
-\begin{equation}
-  \label{eqn:normapprox}
-  S^{\dagger}S + \alpha^2 A^{\dagger}A \approx I + \alpha^2A^TA
-\end{equation}
-Recall that $A$ is simply multiplication by the Euclidean distance to
-the physical source point $\bx_s$: $A u (\bx) = |\bx-\bx_s|u(\bx),
-A^TAu(\bx) = |\bx-\bx_s|^2u(\bx)$. So the equation $(I+\alpha^2
-A^TA)u=b$ is trivial to solve, and this is a key characteristic of a
-good preconditioner. However this observation must be combined with
-the weighted norm structure.
 
@@
 to 1, and the rest are small - which is the case for the current
-problem)..  Further, PCG is computationally effective is M is easy to
+problem)..  Further, PCG is computationally effective if M is easy to
 invert.
+
+\subsection{Preconditioning with Penalty}
+
+Note that the normal operator appearing on the left-hand side of
+\ref{eqn:norm0} is not an approximate identity, due to the presence of
+the regularization term: the spectrum increases in spread with
+increasing $\alpha$, leading to slower convergence. We have already
+observed that the pressure-to-source operators $\Lambda_s$ and
+$\Lambda_r$ are pseudodifferential, and $A$ is pseudodifferential as
+well. The rules for calculation of
+compositions and inverses for such operators show that the operators
+$W_m^{-1}$ (properly interpreted) and $W_m$ are also pseudodifferential.
+Scalar pseudodifferential operators approximately
+commute, so $A^{\dagger} \approx A^T$. Therefore
+\begin{equation}
+  \label{eqn:normapprox}
+  S^{\dagger}S + \alpha^2 A^{\dagger}A \approx I + \alpha^2A^TA
+\end{equation}
+(restricted to the orthocomplement of the null space of $S$).
+Recall that $A$ is simply multiplication by the Euclidean distance to
+the physical source point $\bx_s$: $A u (\bx) = |\bx-\bx_s|u(\bx),
+A^TAu(\bx) = |\bx-\bx_s|^2u(\bx)$. So the equation $(I+\alpha^2
+A^TA)u=b$ is trivial to solve, and this is a key characteristic of a
+good preconditioner.
 
@@
 powers of $I + \alpha^2A^TA$ are trivial to compute, given the choice
-of $A$ made here. We will examine fast algorithms for computing
-$W_m^{-1}$ = the symmetrized pressure-to-source operator in the next
-section. Note that only $M^{-1}$, hence only $W_m^{-1}$, appears in
-Algorithm 1.
-
-\section{Computing and Symmetrizing $\Lambda$}
-
-Computations of $\Lambda^{\pm}_{z_s}$ and its transpose are clearly critical steps in an
-implementation of the PCG algorithm outlined in the preceding section.
-Direct computation of the pressure-to-source operator $\Lambda^{\pm}_{z_s}$, for instance by solving
-\ref{eqn:awepm} and reading off $P_sv^{\pm}_{z}$, turns out to be
-numerically ill-behaved. The relation \ref{eqn:snull} provides and
-alternative approach, taking advantage of the accurate approximate
-inverse to $S^+_{z_s,z_r}$ constructed above. The first row of
-\ref{eqn:snull}, slightly rearranged, is
-\begin{equation}
-  \label{eqn:lamidea}
-  \Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx  -\frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
-\end{equation}
-The approximate inverse construction for $S^++_{z_s,z_r}$ permits (approximate)
-solution of this equation for $\Lambda^+_{z_s}f_s$: apply
-$4(V^{+}_{z_s,z_r})^T$ to both sides of equation \ref{eqn:lamidea} and
-use the first equation in the list \ref{eqn:approxinv} to get
-\begin{equation}
-  \label{eqn:lamident}
-  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
-\end{equation}
-This identity is the major result of this section: it shows how
-to compute that action of $\Lambda^+_{z_s}$ by propagating the input
-pressure trace, identified as a source for the velocity evolution,
-forward in time from $z_s$ to $z_r$
-reading off the pressure trace on $z=z_r$, identifying it once more as
-a point load (source for velocity), propagating it backwards in time from $z_r$ to
-$z_s$, and finally reading off the velocity trace, interpreted as a
-pressure evolution source on $z_s$.
-
-The importance of this result lies in the failure of the obvious
-method for computing the action of $\Lambda^{\pm}_{z_s}$, namely to
-employ the pressure trace as a source in the velocity equation ($f_s$,
-in the notation used above) at $z=z_s$, and read off the velocity
-field also at $z=z_s$. This difficulty is related to the existence of
-tangentially propagating waves and the lack of continuity of the trace
-operator. The method implicit in equation \ref{eqn:lamident} avoids
-this difficulty by propagating the fields a positive distance in $z$:
-assuming as always that the causal fields are downgoing, this step
-eliminates any tangentially propagating fields from consideration.
-
-A deeper study of the pressure-to-source operator (or of the closely
-related Dirichlet-to-Neumann operator for the second order wave
-equation, see \cite{StefUhl:05}) shows that it is approximately
-dependent only on the model coefficients near the source surface
-($z=z_s$ in this case). Since the homogeneous and lens models are
-identical near this surface, it is unsurprising that these figures are
-very close to the previous two.  However an even more useful
-observation is that the calculations in the approximation
-\ref{eqn:lamident} could just as well be carried out in a much smaller
-region around the source surface, and produce a result that is
-functionally identical in that it will serve as a source for the same
-acoustic fields globally, with small error. In effect, equation
-\ref{eqn:lamident} involving propagation from source ($z=z_s$) to
-receiver ($z=z_r$) surfaces is altered by replacing $z_r$ with a
-receiver datum $z_s+\Delta z$ considerably closers to $z_s$:
-\begin{equation}
-  \label{eqn:lamnear}
-  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
-\end{equation}
-Using a receiver datum closer to the source surface has two favorable consequences:
-\begin{itemize}
-\item The computational domain can be smaller than is necessary to
-  simulate the target data, as it need only contain the source
-  surface and the receiver datum implicit in
-  equation \ref{eqn:lamident}. This shrinkage of the computational
-  domain can lead to substantial improvements in computational
-  efficiency.
-\item Since the receiver data may be chosen much closer to the
-  source surface that is the case for the target data, the effective
-  aperture active in the relation \ref{eqn:lamident} can be much
-  larger, producing an estimated source gather much less affected by
-  aperture limitation.
-\end{itemize}
-
-As mentioned in the last section, computation of the transpose of
-$\Lambda^+$ (exact, not approximate in the high frequency sense) is
-critical to the successful construction of the preconditioner. The
-relation \ref{eqn:lamident} does not provide a computation for this
-operator. However set
-\begin{equation}
-  \label{eqn:lamtilde}
-  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
-\end{equation}
-Then \ref{eqn:lamnear} can be rewritten
-\[
-  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
-\]
-Of course, all of the examples so far show images of
-$\tilde{\Lambda}^+_{z_s}$.
-
-Since successful preconditioning requires only approximate inversion,
-use of $\tilde{\Lambda}^+_{z_s} $ in place of $\Lambda^+_{z_s}$ will
-still yield a working preconditioner, and the former can be transposed
-to machine precision via the definition \ref{eqn:lamtilde} and the adjoint state
-method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
-\begin{equation}
-  \label{eqn:lamtransp}
-  (\tilde{\Lambda}^+_{z_s})^T = -8 \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
-    z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}
-\end{equation}
-
-The model space weight operator $W_m^{-1}$ introduced in the last section is
-replaced by its asymptotic approximation
-\[
-\frac{1}{2}(\tilde{\Lambda}^+_{z_s} +
-  (\tilde{\Lambda}^+_{z_s})^T)
-\]
-\[
-  \approx -8 \left((V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T \right.+
-\]
-\[
-  \left. \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
-      z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}\right)
-\]
-\begin{equation}
-  \label{eqn:winvcomp}
-  = -4(\Pi_1 ({\cal S}^+_{z_s,z_s+\Delta z})^T(\Pi_0^T\Pi_1 +
-  \Pi_1^T\Pi_0) {\cal S}^+_{z_s,z_s+\Delta z}\Pi_1^T) = \tilde{W}_m^{-1} .
-\end{equation}
-with a similar definition for the replacement $\tilde{W}_d$ of $W_d$.
-
-This identity shows that only one forward and one adjoint simulation
-are necessary to compute the action of $\tilde{W}_{m,d}$. The operator
-in the center of the expression on the right-hand side, $\Pi_0^T\Pi_1
-+ \Pi_1^T\Pi_0$, simply exchanges the components of the acoustic
-fields, passing the velocity field as a pressure source and the
-pressure field as a velocity source.
-
-One more computation is required for the full implementation of the
-preconditioning strategy explained in the last section: $W_m$ is
-required, not just $W_m^{-1}$. Note that $W_m$ plays two roles in the
-second term in equation \ref{eqn:norm1}: it is the weight matrix for
-both the domain and range norms for $A$. It is perfectly OK for one of
-these to be replaced by an asymptotic approximation, so long as it is
-symmetric and computable (and at least semi-definite). The second row
-in equation \ref{eqn:snull} appears as \ref{eqn:snull2} above:
-introducing (formally) the inverse of $\Lambda^+$,
-\begin{equation}
-  \label{eqn:snull2mod}
--\frac{1}{2}\Pi_1{\cal S}^{+}_{z_s,z_r}\Pi_0^T  \approx
-V^{+}_{z_s,z_r}(\Lambda^{+}_{z_s})^{-1}
-\end{equation}
-whence from the second line in display \ref{eqn:approxinv}
-\begin{equation}
-  \label{eqn:laminv}
--\frac{1}{8}(S^+_{z_s,z_r})^T\Pi_1{\cal S}^{+}_{z_s,z_r}\Pi_0^T  
-\approx (\Lambda^{+}_{z_s})^{-1}
-\end{equation} 
-and
-\begin{equation}
-  \label{eqn:laminvtr}
--\frac{1}{8}\Pi_0({\cal S}^{+}_{z_s,z_r})^T\Pi_1^T S^+_{z_s,z_r}
-\approx ((\Lambda^{+}_{z_s})^{-1})^T.
-\end{equation}
-Using the definition \ref{eqn:sdef} of $S^+_{z_s,z_r}$, the
-symmetrized $\Lambda^{-1}$ is
-\begin{equation}
-  \label{wcomp}
- \tilde{W}_m = -\frac{1}{16}\left(\Pi_0 ({\cal S}^{+}_{z_s,z_r})^T
-   (\Pi_1^T \Pi_0 + \Pi_0^T \Pi_1) {\cal S}^{+}_{z_s,z_r} \Pi_0^T\right)
-   \approx \frac{1}{2}((\Lambda^{+}_{z_s})^{-1}
-   +((\Lambda^{+}_{z_s})^{-1})^T) .
-\end{equation}
-Comparison with the definition \ref{eqn:winvcomp} shows that
-$\tilde{W}_m$ and $\tilde{W}_m^{-1}$ differ only in the initial and
-final projection factors (and overall scale), and in particular either
-can be computed for the cost of a forward/adjoint operator pair. Note
-that $\tilde{W}_m^{-1}$ is inverse to $\tilde{W}_m$ only in an
-approximate (asymptotic, aperture-limited) sense.
+of $A$ made here.
+
+\subsection{Example}
+This final subsection shows that result of Conjugate Gradient
+iteration, with and without preconditioning, applied to the source
+estimation problem \ref{eqn:esis}, with zero and non-zero penalty
+weight $\alpha$. The data $d$ is the gather shown in
+\ref{fig:drecplh0}, simulated using the lens model (Figure \ref{fig:ooplrays0}) with source shown
+in Figure \ref{fig:dsrcphh0}, or, alternatively, a point source with
+bandpass filter wavelet located at $x_d=3500$ m, $z_d=3500$ m. In the inversion,
+the material model is taken to be homogeneous, as has been the case in
+all of the previous examples.
+
+Figure \ref{fig:compnres0lh0} shows the progress of the normal residual
+(Euclidean norm of the difference of the two sides of equation \ref{eqn:norm1}),
+for Conjugate Gradient and Preconditioned Conjugate Gradient
+(Algorithm 1) iterations, applied to solution of the optimization
+problem \ref{eqn:einv} with $\alpha=0$. Because the normal residual is 
+measured in the Euclidean norm for both CG and PCG, the results are
+comparable. For PCG, the preconditioner $M$ is given by equation
+\ref{eqn:defprecond}, with the symmetrized $\Lambda$s computed as
+indicated in the preceding section with auxiliary source and
+receiver surfaces shifted by 100 m. Convergence for the
+preconditioned algorithm is roughly 4 times as fast: this speed
+comparison is much more obvious in a vertically stretched version of
+the figure (Figure \ref{fig:compnrestrunc0lh0}).
+
+%\plot{compnres0lh0}{width=\textwidth}{Comparison of normal residual 
+%  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+%  vs. iteration. Data = lens model, point source (Figure 
+%  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
+%  $\alpha=0$.}
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:compnres0lh0}]{\includegraphics[width=0.45\textwidth]{compnres0lh0.pdf}}
+  \subfloat[\label{fig:compnrestrunc0lh0}]{\includegraphics[width=0.45\textwidth]{compnrestrunc0lh0.pdf}}
+  \caption{(a) Comparison of normal residual 
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+  vs. iteration. Data = lens model, point source (Figure 
+  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
+  $\alpha=0$. (b) Same comparison, with expanded vertical scale, to
+  show more clearly that PCG ir roughly 4-5 times as fast as (in
+  achieving a given normal residual) as CG.}
+\end{figure}
+
+Figure \ref{fig:compnres1lh0} shows the same comparison with non-zero
+penalty weight, $\alpha=10^{-3}$. Convergence for both CG and PCG is
+somewhat slower than in the $\alpha=0$ case. However the relation is
+about the same, still 4-5 times as many iterations to achieve a given
+normal residual level with CG as with PCG.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:compnres1lh0}]{\includegraphics[width=0.45\textwidth]{compnres1lh0.pdf}}
+  \subfloat[\label{fig:compnrestrunc1lh0}]{\includegraphics[width=0.45\textwidth]{compnrestrunc1lh0.pdf}}
+  \caption{(a) Comparison of normal residual 
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+  vs. iteration. Data = lens model, point source (Figure 
+  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
+  $\alpha=10^{-3}$. (b) Same comparison, with expanded vertical scale, to
+  show more clearly that PCG ir roughly 4-5 times as fast as (in
+  achieving a given normal residual) as CG.}
+\end{figure}
+
+%\plot{compnres1lh0}{width=\textwidth}{Comparison of normal residual 
+%  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+%  vs. iteration. Data = lens model, point source (Figure 
+%  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
+%  $\alpha=10^{-3}$.}
 
@@
 The underlying tool in the ideas developed here is geometric optics
-(or ray theory), without which the very concept of downgoing waves
-would be meaningless. The physics of actual earth materials includes
+(or ray theory), without which the very concept of incoming/outgoing waves
+(on which these results rest) would be meaningless. The physics of actual earth materials includes
 material heterogeneity on all scales, which appears to leave little
@@
 
-\append{Adjoint Computation}
-
-The adjoint of ${\cal S}^+_{z_s,z_r}$ can be computed by a variant of
-the adjoint state method, in this case a by-product of the
-conservation of energy. This calculation leads to equation
-\ref{eqn:sadj1}, from which the other statements about adjoints made in the second
-section of the paper follow.
-
-Suppose that $p^-,\bv^-$ solve \ref{eqn:awepm}
-with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
-$ (h_r,f_r\bf{e}_z) \delta(z-z_r)$. Then
-\[
-0 = 
-\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
-\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
--
-\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
-\]
-\[
-+
-\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
- h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
- h_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  (- \nabla p^++f_s\bf{e}_z) \cdot \bv^- + \bv^+ \cdot (-\nabla
-  p^- + f_r \bf{e_z}) \right)
-\]
-\[
-= 
-\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
- h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
- h_r \delta(z-z_r)\right) \right.
-\]
-\[
-+
-\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- 
-  +f_s \delta(z-z_s) v_z^- + v_z^+f_r \delta(z-z_r) \right)
-\]
-after integration by parts in the last two terms. Most of what is left cancels, leaving 
-\[
-0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
-( h_rP_rp^++f_rP_rv_z^+)
-\]
-\[
-  = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
-\]
-whence \ref{eqn:sadj1} follows immediately.
+\section{Acknowledgements}
+The author thanks the anonymous reviewers for their very helpful
+comments and suggestions.
 
@@
 The computational examples reported in this work were written in the
-Madacascar reproducible research framework ({\tt
+Madagascar reproducible research framework ({\tt
   http:www.reproducibility.org}). Code and data source is available

### 2023-01-26 pyrvl/linalg.py
added clip to simplot

@@ -278,9 +278,9 @@ def hdrcomp(vec1, vec2):
         cmd = os.path.join(TRIP,'iwave/trace/main/SEGYCmpHdrs.x')
         ret = os.system(cmd + ' in1=' + vec1 + ' in2=' + vec2)
         if ret != 0:
-#            print('command:')
-#            print(cmd + ' in1=' + vec1 + ' in2=' + vec2)
-#            print('failed with return value ' + str(ret))
+            #print('command:')
+            #print(cmd + ' in1=' + vec1 + ' in2=' + vec2)
+            #print('failed with return value ' + str(ret))
             return False
     else:
         return False
@@ -367,7 +367,7 @@ def rsfcomp(vec1, vec2):
 
         return False
     
-def simplot(f, addcb=False, width=7, asprat=-1):
+def simplot(f, addcb=False, clip=None, width=7, asprat=-1):
 
     RSFROOT = os.getenv('RSFROOT')
     if not os.path.exists(RSFROOT):
@@ -411,6 +411,10 @@ def simplot(f, addcb=False, width=7, asprat=-1):
     datafile=str(inp.get('in'))
     # print('datafile=' + datafile[2:len(datafile)-3])
     data=inp.read()
+
+    # start with the clip
+    if clip is not None:
+        np.clip(a=data, a_min=-clip, a_max=clip, out=data)
 
     # line plot
     if n2==1:

### 2023-01-26 repro/papers/GEM22/submitR2/response.tex
final mods maybe

@@
-COMMENTS FOR THE AUTHOR:
+\documentclass[12pt]{article}
+
+\begin{document}
+\title{Response to Comments on ``Efficient Computation of Extended Surface Sources'', Revision 1, submitted to IJGE}
+\maketitle
+
+Once again, I thank the reviewers for a very helpful comments and suggestions, and for amazingly diligent typo detection. Comments and respponses follow.
+
+\noindent {\bf Editor:}
 
@@
 
-=============================================================================================================================================================================================================================
+Response: Once again the reviews have been quite helpful. I have tried to address each point raised by the reviewers.
 
-Reviewer #1: The manuscript has been considerably revised.
+\noindent {\bf Reviewer 1:}
+
+Comment: The manuscript has been considerably revised.
 The revised version is much better.
@@
 I have, however, two questions and I have detected typos (I am afraid that there are many more).
+
+Response: I have tried to answer the questions, and correct the typos without however introducing as many new ones!
 
@@
 Comment: on page 23, I do not understand the following sentence:
-The function a_0 is a frequency-independent envelope, equal to 1 where h is non-zero.
+The function $a_0$ is a frequency-independent envelope, equal to 1 where h is non-zero.
 
-Response: This section skipped over a number of steps, which I have included in the revision. I believe that I have explained this idea.
+Response: This section skipped over a number of steps, which I have included in the revision. I hope that this more detailed account explains this idea.
 
-typos:
+\noindent {\bf Reviewer 2:}
 
-page 2
-certainluy
-->
-certainly
-
-page 5
-veloity
-->
-velocity
-
-propblem
-->
-problem
-
-tahnks 
-->
-thanks to
-
-is proportion to 
-->
-is proportional to
-
-page 6
-
-slowiy
-->
-slowly
-
-page 8
-increasting
-->
-increasing
-
-page 10
-
-to generate a data traces
-
-page 13
-
-of sigma 
--> 
-of $\Sigma$
-
-page 16
-representaion 
-->
-representation
-
-page 21
-too many s in Psshat{p}
-
-page 22
-Speclializing
-->
-Specializing
-
-last equation: confusion between s and t
-partial v partial t
-the use of s is very confusing (it is the symbol of surface !)
-
-page 23
-first equation: confusion between s and t
-
-SInce 
-->
-Since
-
-check the sentence:
-is asymptotic the pressure field of to an incoming solution
-
-last equation: confusion between s and t
-
-page 24
-parenthesis in lambda(...
-
-page 26
-parenthesis in (tilde{p},tilde{v}
-
-negligbly -> negligibly
-
-defintiion -> definition
-
-page 29
-as it need  
-->
-as it needs
-
-to ne discussed
-->
-to be discussed
-
-page 31
-parenthesis in (equations 28,29
-
-page 33
-typo in: and equation 48, .
-
-page 34
-computationa
-->
-computational
-
-page 36
-ismeasured 
-->
-is measured
-
-page 37
-parenthesis in (Figure 15b
-
-page 39
-Madacascar
-->
-Madagascar
-
-=============================================================================================================================================================================================================================
-
-Reviewer #2: Overall the paper is much improved and I now found it relatively easy to follow and understand.
+Comment: Overall the paper is much improved and I now found it relatively easy to follow and understand.
 I particularly liked the 'steps' introduced as a way to summarize what you've done so far and to give the reader further help in actually implementing these ideas.
 (I also liked the roadmap section at the beginning.)  I also found the new figures much easier to compare and understand.
+There are a few areas where I think things could still be somewhat clearer, which are:
 
-There are a few areas where I think things could still be somewhat clearer, which are:
+Response: Thank you for your comments, which inspired these changes in R1.
 
@@
 Response: You are of course correct that the ``things that turn back'' are another diving wave branch.
-I've considerably revised the figure, its caption, and the discussion in the text, intending to answer all of these concerns.
+I've considerably revised the figure, its caption, and the discussion in the text, intending to answer all of these questions.
 
@@
 I created a parallel sequence of experiments for the lens example using that configuration.
-However a point source cannot be reconstructed, at least not with the simple transmission geometry used in this paper, since a lot of energy propagates away ``to China'' and is not available in the inversion.
+However a point source cannot be reconstructed, at least not with the simple transmission geometry used in this paper, since a lot of its energy propagates out of aperture and is not available in the inversion.
 You can reconstruct a source that generates the data, but it isn't a point source - if the velocity is right, it is more or less a dip-filtered version of the point source.
@@
 
-The diving wave example uses a point source on the source surface.
-You will notice that the reconstructed source (Figure 9b) appears to be a dip-filtered point source, regarded as a function on the source surface.
+The diving wave example uses a point source on the source surface. You will notice that the reconstructed source (Figure 9b) appears to be a dip-filtered point source, as expected.
 
@@
 Note that the gradient-to-constant transition is smoothed, and the gradient is large enough to generate a triplication.
-Having homogeneous layers top and bottom really messes things up.
+Having homogeneous layers top and bottom really messes things up in an interesting way.
 Sheriff and Geldart, {\em Exploration Geophysics}, discussion of fig.
-4.20a, have a nice discussion of this behaviour.
+4.20a, mention this behaviour.
 There are segments of circles in there, but overall these rays are not close to circles.
-I think the description now in the text, and the inclusion of the arrival time plot, will make things clear enough.
+I think the description now in the text, and the inclusion of the arrival time plot, will make things clear.
 
@@
 
-Response: The PAT construction does not apply to the surface source inversion problem, since there is no bounding surface on all of which the pressure field is measured/known.
+Response: I have added a bit of discussion that I hope will justify and explain this assertion.
+The PAT construction does not apply to the surface source inversion problem, since there is no bounding surface on all of which the pressure field is measured/known.
 I get around this by localizing in space-time.
@@
 Then it can be converted to a local solution of a problem with surface source, as described.
-Since the field vanishes near the artificial boundary, you can simply get rid of it, and if the pressure field on the boundary is only non-zero for a short time, that's it - just let it propagate in space time.
+Having taken that step, you can simply get rid of the artificial boundary, since the field vanishes near it.
+If the pressure field on the source (or receiver) surface is only non-zero for a short time, that's it - just let it propagate in forward time.
 Finally, any pressure distribution on the source surface can be broken up in to a sum of terms, each only non-zero for a short interval.
-Apply the short-time-window construction to each summand, and add them up.
+Apply the short-time-window construction to each summand, and add up the resulting fields.
 
@@
 
-Typos: I am {\em very} grateful to both referees for taking the trouble to list many of the embarassingly extensive typos.
+\noindent {\bf Typos and other corrections:}
+
+I am {\em very} grateful to both referees for taking the trouble to list many of the embarassingly extensive typos.
 I believe that I have fixed all of these, and some others which we won't mention.
-I only hope I have not created yet more...
+I only hope I have not created many new typos...
 
-A very-incomplete typo list (I do suggest getting out your grammatical fine-toothed comb and checking for more of these):
+I also corrected a number if errors in figure and equation references, some of which may once have been right but no longer were.
 
--Spelling of constraint in the abstract
--p 6 line 23 "is not be well-defined"
--p 13 line 24 there is a sigma without a \
--p 16 line 59 spelling of representation
--P 21 line 40 The sentence beginning "Differentiate" is missing some important part-of-speech for which I should have a name but don't, perhaps a subject?
--p26 line 28 missing )
--p26 line 48 sentence fragment beginning "To illustrate"
--p26 line 62 missing space in is measured
+The section ``Computing $\Lambda$'' needed revision, as I had glossed over the fact that the approximate inverse is only a {\em right} approximate inverse - necessarily, since the modeling operator $S$, as I have defined it, can have small singular values corresponding to out-of-aperture transmission, or possibly even an outright null space.
+A consequence is that the approximation of $\Lambda$ computed in this section is accurate up to an error that is neglible after application of $S$, and that needed to be stated clearly.
+
+\end{document}

### 2023-01-26 repro/papers/GEM22/submitR2/paper.tex
maybe final

@@
 constraint, allowing the source field $\bff$ to have more degrees of
-freedom than is permitted by a faithful model model of the seismic
+freedom than is permitted by a faithful model of the seismic
 experiment, and constraining these additional degrees of freedom by
@@
 \ref{fig:dinvhslh0}. It differs considerably from the source field used to
-generate the data (Figure \ref{fig:dsrcphh0}), however when used as the
+generate the data (Figure \ref{fig:dhshh0}), however when used as the
 source in forward modeling (system \ref{eqn:awesrcglob}) {\em with the
@@
     bulk modulus model (different from the model used to generate the
-    data!). Plotted with same color scale as Figure \ref{fig:dsrcphh0}.}
+    data!). Plotted with same color scale as Figure \ref{fig:dhshh0}.}
 \end{figure}
@@
 supply the normal velocity at the source (or receiver) surface, since
-it's determined locally by the pressure on the surface, or causal (or
+it's determined locally by the pressure on the surface, for causal (or
 anti-causal) fields. This follows from the observation that the system
@@
 \[
-  S^T(h_r) = -P_s \hat{p}.
+  S^Th_r = -P_s \hat{p}.
 \]
@@
 a planar source surface
-$\Lambda_s=\{z=z_s\}$, as in the two examples. Since the same reasoning
+$\Sigma_s=\{z=z_s\}$, as in the two examples. Since the same reasoning
 applies to $\Lambda_r$, I will omit the subscript for the remainder of
@@
   \lambda^0(x,y,t,\xi,\eta,\omega) = 2\frac{( c^{-2}(x,y,z)-
-    \xi^2+\eta^2)^{1/2}}{\rho} + O(1/\omega)
+    (\xi^2+\eta^2))^{1/2}}{\rho} + O(1/\omega)
 \]
@@
 In the final expression, the slownesses $\xi,\eta$ have been replaced
-by the Fourier variables $k_x=x\xi, k_y=y\eta$. Note that as a
+by the Fourier variables $k_x=\omega\xi, k_y=\omega\eta$. Note that as a
 function of $k_x,k_y,\omega$, $\lambda^0$ is homogeneous of order
@@
                                                       - \nabla \cdot \tilde{\bv}, \nonumber \\
-  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & f_s {\bf n}\delta_{\Sigma}- \nabla \tilde{p}, \nonumber \\
+  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & f_s {\bf n}\delta_{\Sigma_s}- \nabla \tilde{p}, \nonumber \\
   \tilde{p} & = & 0 \mbox{ for }  t \ll 0, \nonumber \\
@@
   \caption{(a) Pressure gather for Lens example on source surface
-    $\Sigma_s=\{z=1000\mbox{ m}\}$, generated by point source at
+    $\Sigma_s=\{z=3000\mbox{ m}\}$, generated by point source at
     $(3500,3500)$ m. (b) Normal velocity gather for Lens example on source surface
-    $\Sigma_s=\{z=1000\mbox{ m}\}$, generated by point source at
+    $\Sigma_s=\{z=3000\mbox{ m}\}$, generated by point source at
     $(3500,3500)$ m.}
@@
     $\times$ pressure gather shown in Figure
-    \ref{fig:dsrcvzhh0}. (c) Difference (b)-(a), plotted on same color
+    \ref{fig:dsrcphh0}. (c) Difference (b)-(a), plotted on same color
     scale.}
@@
 \end{equation}
-Apply the approximate inverse construction in the form
-\ref{eqn:altinv} to solve \ref{eqn:lamidea} for $\Lambda_s$:
+Apply $S^{\dagger}$ in the form \ref{eqn:altinv} to both sides to obtain
 \begin{equation}
   \label{eqn:lamident}
-  \Lambda_s \approx -8\Pi_1{\cal S}^T\Pi_1^T\Pi_0 {\cal S}\Pi_1^T.
-\end{equation}
+ S^{\dagger} S \Lambda_s \approx -8\Pi_1{\cal S}^T\Pi_1^T\Pi_0 {\cal S}\Pi_1^T.
+\end{equation}
+As earlier noted, $S^{\dagger}$ is an approximate right inverse to
+$S$, not an approximate left inverse, so the right-hand side of
+\ref{eqn:lamident} is not (necessarily) an approximation to
+$\Lambda_s$. However, apply $S$ to both sides and use the relation
+\ref{eqn:rightinv}:
+\begin{equation}
+  \label{eqn:goodenuf}
+ S \Lambda_s \approx SS^{\dagger} S \Lambda_s \approx S(-8\Pi_1{\cal S}^T\Pi_1^T\Pi_0 {\cal S}\Pi_1^T).
+\end{equation}
+That is, the right-hand side of equation \ref{eqn:lamident} differs
+from $\Lambda_s$ by an operator approximately annihilated by $S$.
+
 This identity is the major result of this section: it shows how
-to compute that action of $\Lambda_s$ by propagating the input
+to compute that action of $\Lambda_s$, up to an error with negligible
+effect on the subsequent
+action of $S$, by propagating the input
 pressure trace, identified as a source for the velocity evolution,
@@
 \end{equation}
+In view of the result \ref{eqn:goodenuf}, the meaning of this
+approximation is that the difference of the outputs of $S$, applied to
+the two sides, is negligible. Since $\Lambda_s$ is used only in
+combination with $S$, it is tempting to take the right-hand side of
+equation \ref{eqn:lamnear} as the {\em definition} of $\Lambda_s$. This
+apparently dramatic step will be fully justified in the next section.
 
@@
 pressure-to-source operator - not approximately, but to machine
-precision. The prescription \ref{eqn:lamnear} only
-defines an approximation. However simply substituting the right-hand
-side of equation \ref{eqn:lamnear} - in effect, substituting the
-approximation for the operator that it approximates - gives an
-operator variant of $\Lambda_s$ with an accessible
-machine-precision adjoint:
+precision. The prescription \ref{eqn:lamnear} only defines an
+approximation. However simply substituting the right-hand side of
+equation \ref{eqn:lamnear} for $\Lambda_s$, as suggested above - in
+effect, substituting the approximation for the operator that it
+approximates - gives an operator variant of $\Lambda_s$ with an
+accessible machine-precision adjoint:
 \begin{equation}

### 2023-07-27 repro/papers/awi/paper.tex
complete rewrite from scratch

\title{Notes on AWI}
\author{William W. Symes}

\begin{abstract}
goes here.
\end{abstract}

\section{Introduction}

\section{AWI and why it works}

The version of AWI introduced by Michael Warner and Lluis Guasch \cite[]{Warner:16,GuaschWarnerRavaut:GEO19} rests on the assumption that seismic waves are governed by linear acoustics, and that each shot is associated to an isotropic point source with known location $\bx_s$ and wavelet $w(t;\bx_s)$. That is, the pressure and velocity fields $p({\bf x},t;{\bf x}_s)$, ${\bf v}({\bf x},t;{\bf x}_s)$ for the shot location ${\bf x}_s$ depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf x})$ (reciprocal of the density $\rho({\bf x})$), and wavelet $w(t;{\bf x}_s)$ through the acoustic system
 $$
 \frac{\partial p}{\partial t} = - \kappa \nabla \cdot {\bf v} +
w(t;{\bf x}_s) \delta({\bf x}-{\bf x}_s);
\frac{\partial {\bf v}}{\partial t} = - \beta \nabla p; 
p, {\bf v} = 0 \mbox{ for }  t \ll 0.
$$
For sake of brevity, define the model vector $m=(\kappa,\rho)$. The forward map or {\em modeling operator} is $F[m]w = \{p({\bf x}_r,t;{\bf x}_s)\}$, in which shot and receiver positions ${\bf x}_s, {\bf x}_r$ define the acquisition geometry.

Acoustic inversion, for present purposes, means: given time interval $[0,T]$, source and receiver locations $(\bx_s,\bx_r)$, source wavelet $w_* \in C_0^{\infty}(\bR)$ and data traces $d(\bx_r,\cdot;\bx_s) \in L^2([0,T])$, find a model $m$ so that $F[m]w_* \approx d$. The simplest version of FWI concretizes this task by asking for a model $m$ minimizing the mean square error
\begin{equation}
  \label{eqn:fwi}
  J_{\rm FWI}[m;d] = \frac{1}{2}\|F[m]w_*-d\|^2
\end{equation}
Practical versions of FWI include various types of regularization and homotopy, see the sources cited in the Introduction. As mentioned there, local iterative optimization methods tend to stagnate at highly sub-optimal (``cycle-skipped'') estimates of $m$ unless the initial estimate is of quite high quality, often available only at considerable cost in effort and data collection.
%For now, assume that the data $d({\bf x}_r,t;{\bf x}_s)$ is the output of the modeling operator for "true" model $m_*$, that is, the "true" bulk modulus, buoyancy, and wavelet $\kappa_*, \beta_*, w_*$: that is, $d = F[m_*]w_*$.

AWI is based on extending the modeling operator $F$, that is, enlarging its domain.
The extended modeling operator ${\bar F}$ maps extended sources $\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure field. That is, the extended source depends on the receiver location as well as the source location - so there is one acoustic system for each source *and* receiver position - a lot of wave equations! If all of the wavelets for each source are the same, that is, ${\bar w}({\bf x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver position, then ${\bar F}[m]\bar{w} = F[m]w$. That is, $F$ is a special case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of $F$. This is the *source-recever* extension, in the terminology of Huang et al. 2015.

AWI assumes that the extended sources are time convolutions of the (known) exact source with an {\em adaptive kernel} $u({\bf x}_r,t;{\bf x}_s)$: that is, $\bar{w} = u * w_*$ - the asterisk denotes convolution in time. Since linear acoustics is time-translation invariant, its solution commutes with time convolution, that is,
$$
\bar{F}[m]\bar{w} = \bar{F}[m](u* w_*) = (F[m]w_*)*u
$$
For notational convenience, introduce the notation $S[m,w_*]$ for the convolution operator with kernel $F[m]w_*$:
\begin{equation}
  \label{eqn:defs}
  S[m,w_*]u = (F[m]w_*)*u
\end{equation}
Warner and Guasch assume (implicitly) that there is always a (near-)zero-residual solution of the extended inversion problem. That is, they assume that for any $m$, there is an extended source $\bar{w}$, or equivalently an adaptive kernel $u_0$ with $\bar{w} = w_**u_0$ for which $\bar{F}[m]\bar{w} = S[m,w_*]u_0\approx d$. If $\kappa \approx \kappa_*, \beta \approx \beta_*$, then $\bar{w}$ should be $\approx w_*$, so the adaptive kernel $u_0$ should be approximately $\delta(t)$ and independent of ${\bf x}_s,{\bf x}_r$. Such a kernel $u_0$ is (approximately) in the null space of multiplication by $t$.

This observation leads to the definition of the Warner-Guasch AWI objective function $\tJz$:

\begin{itemize}
\item[1. ]Given $m$, solve the deconvolution problem $(F[m]w_*)*u_0 = d$ for $u_0$, if necessary in a least squares sense and with Tihonov regularization;. 
\item[3. ]Compute the objective 
  \begin{equation}
    \label{eqn:tJz}
    \tJz[m] = \sum_{x_s,x_r} \left(\frac{\int dt |tu_0|^2}{\int dt |u_0|^2}\right)
  \end{equation}
and its gradient, then update $\kappa,\beta$ using a gradient descent method to minimize $\tJz$.
\end{itemize}

The normalization of the integrand per source and receiver by the $L^2$ norm of $u_0$ is quite important, as will be explained shortly.

It may not be obvious at first glance why this approach to estimating $m$ should be any more effective than FWI. To see why AWI works, in at least some settings, we assume that the spatial dimension is three, and make use of the geometric optics approximation to the solution of the wave equation, in a form due to Hadamard. This approximation is {\em local}: it applies when sources and receivers are sufficiently close. We will discuss its global failure and the consequences for AWI in a later section.

Transient acoustic sources have no DC (zero frequency) component and are relatively small at low frequencies. One was to enforce these characteristics is to assume that the wavelet $w_*$ is the time derivative of a $v_* \in C_0^{\infty}(\bR)$. Set $\lambda(w_*) = \|v_*\|/\|w_*\|$ ($L^2$ norms). $\lambda$ is a proxy for wavelength: it is the reciprocal of the RMS frequency.

For a particular source-receiver pair ${\bf x}_s,{\bf x}_r$ with $\bx_r$ "not too far" from $\bx_s$,, suppressed from the notation,
\begin{equation}
  \label{eqn:hadamard}
  F[m]w_*(t) = a[m]w_*(t-\tau[m]) + Rw_*(t)
\end{equation}
where $R(\bx_r,\bx_s)$ is a bounded operator on $L^2((-\infty,T])$ bounded by a multiple of $\lambda(w_*)$. This ``single arrival'' approximation  follows from the expansion of the fundamental solution of the wave equation due to Hadamard, see for instance \cite{Friedlander:75}. This construction is sketched in Appendix B. Note that in general the approximation \ref{eqn:hadamard} is only local: for smooth $m$, it holds for $\bx_r$ sufficiently close to $\bx_s$, but fails at larger distances.

For approximations with $O(\lambda(w_*)$ remainders, write
\begin{equation}
  \label{eqn:pwe}
  F[m]w_*(t) \approx a[m]w_*(t-\tau[m])
\end{equation}
Thanks to the approximation \ref{eqn:pwe},
\begin{equation}
  \label{eqn:spwe}
S[m,w_*]u(t) \approx a[m](w_**u)(t-\tau[m])
\end{equation}
Assume that the data is noise-free:
$$
d(t)=F[m_*]w_*(t) \approx a[m_*]w_*(t-\tau[m_*])
$$
The (approximate) solution of $S[m,w_*]u_0=d$ is obvious by inspection: $u_0(t) = a[m_*]/a[m]\delta(t-\tau[m_*]+\tau[m])$. There is immediately a big problem: this $u_0$ is not square-integrable, and the AWI objective $\tJz$ is undefined. To fix this, it's necessary to treat $u_0$ as the solution of a regularized least squares problem: define
$$
u_0^{\epsilon} = \mbox{argmin}_u (\|S[m,w_*]u-d\|^2 + \epsilon^2\|u\|^2)
$$
$$
\approx \mbox{argmin}_u (\|a[m](w_**u)(\cdot-\tau[m])-a[m_*]w_*(\cdot-\tau[m_*])\|^2 + \epsilon^2\|u\|^2)
$$
Use the temporary abbreviations $a_*=a[m_*], a=a[m], \tau_*=\tau[m_*], \tau=\tau[m]$, and write in terms of the Fourier transforms $\hat{u}_{\epsilon},\hat{w_*}$ of $u_0^{\epsilon}$ and $w_*$:
$$
\hat{u}_{\epsilon} \approx \mbox{argmin}_{\hat{u}}\left(\int d\omega |a\hat{w_*}\hat{u}e^{i\omega \tau}-a_*\hat{w_*}e^{i\omega\tau_*}|^2 + \epsilon^2 |\hat{u}|^2\right)
$$
The normal equation is
$$
(a^2 |\hat{w}_*|^2 +\epsilon^2)\hat{u} = a a_*|\hat{w}_*|^2e^{i\omega(\tau_*-\tau)}
$$
the solution of which is
$$
\hat{u}_{\epsilon} \approx \frac{a^*}{a}\hat{g}_{\frac{a^*}{a}\epsilon} e^{i\omega(\tau_*-\tau)}
$$
where
$$ 
\hat{g}_{\epsilon} = \frac{a_*^2|\hat{w}_*|^2}{a_*^2|\hat{w}_*|^2 + \epsilon^2}.
$$
Since $w_* \in C_0^{\infty}$, $\hat{w_*}(\omega) \rightarrow 0$ faster than any negative power of $\omega$ as $|\omega| \rightarrow \infty$ and is analytic, whence $\hat{g}_{\epsilon}: \epsilon > 0$ tends to $1$ almost everywhere as $\epsilon \rightarrow 0$. Hence the inverse Fourier transform $g_{\epsilon}$ tends to $\delta$ in the sense of distributions, and 
$$
u_0^{\epsilon}(t) \approx \frac{a^*}{a}g_{\frac{a^*}{a}\epsilon}(t-(\tau_*-\tau))
$$
to a multiple of a shifted $\delta$, in fact exactly the distribution solution of $\bar{S}[m,w_*]u=d$. However for $\epsilon>0$, $u_0^{\epsilon}$ is square integrable.

\cite{HuangSymes:SEG15a} presented an analysis of the unweighted AWI objective (numerator of the right hand side in the definition \ref{eqn:tJz}), based in turn on ideas presented in Song's thesis \cite[]{Song:94c}. With $u$ replaced by $u_0^{\epsilon}$,  this unweighted objective is
$$
\int dt |tu_0^{\epsilon}|^2 \approx \frac{a_*^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\epsilon}(t-(\tau_*-\tau))|^2
$$
$$
=\frac{a_*^2}{a^2} \int dt\, (t+(\tau_*-\tau))^2|g_{\frac{a^*}{a}\epsilon}(t)|^2
$$
The linear (in t) term vanishes, since $g$ is necessarily even in time (since it Fourier transform is real). So this is
$$
=\frac{a_*^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\epsilon}(t)|^2
+(\tau_*-\tau)^2\frac{a_*^2}{a^2}\|g_{\frac{a^*}{a}\epsilon}\|^2
$$
The first term tends to zero as $\epsilon \rightarrow 0$, since $g_{\epsilon}$ is a delta-family, so you can make it negligible with proper choice of $\epsilon$. The second term suggests that this function is related to the travel-time error $\tau_*-\tau$.

The relation is obscured by the presence of of the amplitude ratio $a_*/a$ both as a multiplier and in the scaling of $\epsilon$. Since the amplitudes depend on $m$, their presence may influence the location (or even the presence) of  stationary points of $\tJz$ for models differing in their travel time predictions. So stationary points of the unweighted AWI objective may not yield slowness models that match travel-times with the data, even in the noise-free case (even though the examples reported in Huang and Symes SEG 2015, and other works cited there, suggest that sometimes this is the case).

The scaling by trace norms of $u_0^{\epsilon}$ in the definition \ref{eqn:tJz} eliminates the possibility of non-travel-time stationary points. Since the objective function is built up trace-by-trace, use the same approximation for the norm of $u_0^{\epsilon}$: 
$$
\|u_0^{\epsilon}\| \approx \frac{a^*}{a}\|g_{\frac{a^*}{a}\epsilon}\|
$$
So
$$
J_0[m,w_*] \approx \sum_{{\bf x}_s,{\bf x}_r} (\tau[m_*]({\bf x}_s,{\bf x}_r)-\tau[m]({\bf x}_s,{\bf x}_r))^2
$$
ignoring the $\|tg_{...}\|^2$ term as being negligible for small enough $\epsilon$.

In other words, the AWI objective is *exactly* an approximation to the travel-time tomography objective. And that's why it works - when geometric optics in the form given here is an accurate approximation to the Green's function.

It remains to see what geometric optics has to say about the penalty objective $\tilde{J}_{\alpha}[m,w_*]$, how geometric optics might fail, and what happens to AWI when it does.

\section{A Penalty Version of AWI}
The adaptive kernel $u_0^{\epsilon}$ appearing in the definition \ref{eqn:tJz} is the solution of a regularized least squares problem. This observation suggests augmenting the regularized least squares objective by a penalty proportional to the right-hand side of \ref{eqn:tJz}:
\begin{equation}
  \label{eqn:Ja}
  \Ja[m,u;d] = \frac{1}{2}\left(\|S[m,w_*]u-d\|_d^2 + \epsilon^2\|u\| + \alpha^2\|tu\|_m\right).
\end{equation}
Here $\|u\|^2 = \langle u, u \rangle$ is the (unweighted) $L^2$ norm, and $\|\cdot\|_m$ is an $(m,d)$-dependent (weighted) norm: with
\[
  u_0^{\epsilon}[m,d] = \mbox{arg min}_u \Jz[m,u;d],
\]
define
\[
  Wu (\bx_r,t;\bx_s) = \sum_{\bx_r,\bx_s}\frac{1}
  {\int\,ds\,|u_0^{\epsilon}[m,d](\bx_r,s;\bx_s)|^2}u(\bx_r,t;\bx_s)
\]
and
\begin{equation}
  \label{eqn:defmnorm}
  \|u\|^2_m = \langle u, W u \rangle.
\end{equation}
With this notation,
\[
  \tJz[m;d] = \frac{1}{2}\|tu^{\epsilon}_0[m;d]\|^2_m
\]
In the definition of $\tJz$, a quadratic is evaluated at the minimizer of $\Jz[m,u;d]$. This suggests the possibility of evaluating the quadratic objective $\Ja[m,u;d]$ at its own minimizer:
\begin{equation}
  \label{eqn:tJa}
  \tJa[m;d] = \min_u \Ja[m,u,d] = \Ja[m,\ua[m;d];d]
\end{equation}
This is the so-called variable projection reduction of the quadratic $\Ja$.
Note that the right-hand side of the definition of $\tJz$ is the last term on the right-hand side of $\Ja$, scaled by $\alpha^{-2}$ and evaluated at the minimizer of the sum of the first two terms, that is, $\Ja$ for $\alpha=0$.

\begin{theorem}
  \label{thm:basic}
Suppose that $S \in {\cal L}(U,D)$, the is a bounded linear maps from (Hilbert space) domain $U$ (``model space'') with inner product $\langle \cdot, \cdot \rangle_m$ to range $D$ with inner product $\langle \cdot, \cdot \rangle_d$. Assume that $S$ is coercive, that is, there exists $\epsilon >0$ so that $\|Su\|_d \ge \epsilon\|u\|_m$ for $u \in U$.
Suppose also that $A \in {\cal L}(U,U)$, and that $d \in D$. Define the function $J_{\alpha}$ on $U$ by 
\begin{equation}
  \label{eqn:eq1}
  J_{\alpha}(u) = \frac{1}{2}(\|Su-d\|^2_d + \alpha^2\|Au\|_m^2).
\end{equation}
Then for $\alpha \ge 0$, $J_{\alpha}$ possesses a unique minimizer $u_{\alpha} \in U$, for which
\begin{equation}
  \label{eqn:eq0}
  \tilde{J}_{\alpha} = \min_{u \in U} J_{\alpha}(u) = J_{\alpha}(u_{\alpha}),
\end{equation}
and
\begin{equation}
  \label{eqn:eq4}
  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tilde{J}_{\alpha}-\tilde{J}_0).= \frac{1}{2}\|Au_0\|^2
\end{equation}
This limit is uniform over bounded sets of $(S, A) \in {\cal L}(U,D) \times {\cal L}(U, U)$.
\end{theorem}

For completeness, the proof is given in an appendix.

This relation will be applied in the context of a model space $U$ possessing a reference Hilbert norm $\|\cdot\|$ and a family of equivalent norms $\|\cdot\|_m$ depending on a parameter $m \in M$, in which $M$ is an open subset of a Banach space $X$. The dependence is through a an operator-valued function $W$ on $M$, taking values in cone ${\cal B}(U)$ of bounded self-adjoint positive-definite operators on $U$. On the other hand, we assume that the norm on $D$ is fixed, independent of $M$, and treat it as a reference norm.

\begin{theorem}
  \label{thm:uniform}
Assume that $W: M \rightarrow {\cal B}(U)$ is of class $C^1$ and uniformly positive definite, and that the norm $\|\cdot\|_m$ is given by
\begin{equation}
  \label{eqn:normdef}
  \|u\|_m^2 = \langle u, W[m] u \rangle.
\end{equation}
Suppose also that $S: M \rightarrow {\cal L}(U,D)$ is of class $C^1$ and uniformly coercive, and that $A \in {\cal L}(U,U)$ as before. For $\alpha \ge 0$, define $\tJa: M \rightarrow \bR$ by
\[
  \tilde{J}_{\alpha}[m] = \min_{u \in U}\frac{1}{2}(\|S[m]u-d\|^2_d + \alpha^2\|Au\|_m^2)
\]
\begin{equation}
  \label{eqn:tjadef}
  = \frac{1}{2}(\|S[m]u_{\alpha}[m]-d\|^2_d + \alpha^2\|Au_{\alpha}[m]\|_m^2)
\end{equation}
where
\begin{equation}
  \label{eqn:ua}
  u_{\alpha}[m] = \mbox{arg min}_{u \in U}\frac{1}{2}(\|S[m]u-d\|^2_d + \alpha^2\|Au\|_m^2)
\end{equation}
for $m \in M$. Then $\tilde{J}_{\alpha}$ is of class $C^1(M)$, and
\begin{equation}
  \label{eqn:lim}
  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tJa[m]-\tJz[m]).= \frac{1}{2}\|A\uz[m]\|_m^2.
\end{equation}
The limit is in the sense of $C^1(M)$, that is, both the left-hand side and its $m$ derivative converge to the right-hand side and its $m$ derivative, respectively. Moreover $u_{\alpha}$ is of class $C^1(M,U)$.
\end{theorem}

\begin{proof} All conclusions but the last follow immediately from Theorem \ref{thm:basic} and the uniformity assumptions on $S$ and $W$.

Since the norm in $D$ is fixed for the discussion, drop the subscript and treat it as a reference norm. Denote the adjoint with respect to the norms $\|\cdot\|_m$ on $U$, $\|\cdot\|_d = \|\cdot\|$ on $D$ by a superscript $*$, and with respect to $\|\cdot\|$ on $U$ by a superscript $T$. The normal equation defining $\ua$ is
  \[
    (S[m]^*S[m] + \alpha^2 A[m]^*A) \ua[m] = S[m]^*d
  \]
  \begin{equation}
    \label{eqn:normal}
= (W[m]^{-1}S[m]^TS[m] + \alpha^2 W[m]^{-1}A^TW[m]A)\ua[m]= W[m]^{-1}S[m]^Td
\end{equation}
or equivalently
\begin{equation}
  \label{eqn:normal1}
  (S[m]^TS[m] + \alpha^2 A^TW[m]A)\ua[m] = S[m]^Td
\end{equation}
Since $S$ is uniformly coercive, and both $S$ and $W$ are of class $C^1$, the normal operator on the left hand side is boundedly invertible and of class $C^1$. Therefore the solution $\ua$ is of class $C^1$.

[need to add proof of $C^1$ convergence.]
\end{proof}

Apply this relation in the context of the AWI penalty problem defined earlier. $M$ is an open set in $C^k(\Omega, \bR^2)$. $U$ is $\oplus_{\bx_r,\bx_s}L^2([-T_u,T_u])$, $D$ is the data space $\oplus_{\bx_r,\bx_s}L^2([0,T])$,. $S$ is the regularized modeling operator $(S[m,w_*],\epsilon I)^T$, $d$ is replaced by the $(d,0) \in D \times U$,  $W$ is the weight operator on $U$ defined in \ref{eqn:defmnorm}.

As noted before [but it wasn't - fix!], stationary points of $\tJz$ are $\lambda(w_*)$-close to stationary points of the travel time mean-square error, in the single-arrival case. The result of this section implies that stationary points of $\tJa$ are close to those of $\tJz$. Therefore, at least for small $\alpha$, minimization of $\tJa$ produces acoustic models that closely match the traveltimes predicted in single-arrival data.

\append{$\alpha \rightarrow 0$ scaled limit of penalty function}
Define
\begin{equation}
  \label{eqn:eq0}
  \tJa = \min_u J_{\alpha}(u) = J_{\alpha}(\ua)
\end{equation}
Since $S$ is positive definite, the minimum is well-defined for any $\alpha \ge 0$, and the minimizer $\ua \in U$ satisfies the normal equation.

The claim to be established is that
\begin{equation}
  \label{eqn:eq4}
  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tJa-\tJz).= \frac{1}{2}\|A\uz\|_m^2
\end{equation}

To see this, use the normal equation to write
\[
  \ua = (S^TS + \alpha^2 A^TA)^{-1}S^Td = (S^TS + \alpha^2 A^TA)^{-1}S^TS \uz
\]
\begin{equation}
  \label{eqn:eq2}
  = \uz - \alpha^2 (S^TS + \alpha^2 A^TA)^{-1}A^TA\uz = \uz-\alpha^2 \va
\end{equation}
Note that $\va = (S^TS + \alpha^2 A^TA)^{-1}A^TA\uz$ is uniformly bounded in $\alpha \ge 0$.

Stuff the RHS of equation \ref{eqn:eq2} into the definition \ref{eqn:eq1} to obtain
\begin{equation}
  \label{eqn:eq3}
  \tJa = \frac{1}{2}(\|S\uz-d\|_d^2 - 2 \alpha^2\langle S\uz-d, S\va\rangle_d + \alpha^2\|A\uz\|_m^2 + O(\alpha^4))
\end{equation}
The second term on the RHS of equation \ref{eqn:eq3} vanishes thanks to the normal equation, and the first term is precisely $\tJz$. The conclusion \ref{eqn:eq4} follows.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2023-10-21 repro/papers/awi/paper.tex
update

@@
 AWI is based on extending the modeling operator $F$, that is, enlarging its domain.
-The extended modeling operator ${\bar F}$ maps extended sources $\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure field.
-That is, the extended source depends on the receiver location as well as the source location - so there is one acoustic system for each source *and* receiver position - a lot of wave equations!
-If all of the wavelets for each source are the same, that is, ${\bar w}({\bf x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver position, then ${\bar F}[m]\bar{w} = F[m]w$.
-That is, $F$ is a special case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of $F$.
-This is the *source-recever* extension, in the terminology of Huang et al.
-2015.
+The extended modeling operator ${\bar F}$ maps extended sources
+$\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
+field. That is, the extended source depends on the receiver location
+as well as the source location - so there is one acoustic system for
+each source *and* receiver position - a lot of wave equations! If all
+of the wavelets for each source are the same, that is, ${\bar w}({\bf
+  x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver
+position, then ${\bar F}[m]\bar{w} = F[m]w$. That is, $F$ is a special
+case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
+$F$. This is the {\em source-receiver} extension, in the terminology of \cite{HuangSymes2015SEG}.
 
-AWI assumes that the extended sources are time convolutions of the (known) exact source with an {\em adaptive kernel} $u({\bf x}_r,t;{\bf x}_s)$: that is, $\bar{w} = u * w_*$ - the asterisk denotes convolution in time.
+AWI assumes that the extended sources are time convolutions of the
+(known) exact source $w_*$ with an {\em adaptive filter} $u({\bf x}_r,t;{\bf x}_s)$: that is, $\bar{w} = u * w_*$ - the asterisk denotes convolution in time.
 Since linear acoustics is time-translation invariant, its solution commutes with time convolution, that is,
-$$
+\[
  \bar{F}[m]\bar{w} = \bar{F}[m](u* w_*) = (F[m]w_*)*u
-$$
-For notational convenience, introduce the notation $S[m,w_*]$ for the convolution operator with kernel $F[m]w_*$:
+\]
+Since the known source $w_*$ is not adjusted during the inversion, 
+introduce the notation $S[m]$ for the convolution operator with kernel $F[m]w_*$:
 \begin{equation}
-  \label{eqn:defs}
-  S[m,w_*]u = (F[m]w_*)*u
+  \label{eqn:sdef}
+  \equiv S[m]u
 \end{equation}
-Warner and Guasch assume (implicitly) that there is always a (near-)zero-residual solution of the extended inversion problem.
-That is, they assume that for any $m$, there is an extended source $\bar{w}$, or equivalently an adaptive kernel $u_0$ with $\bar{w} = w_**u_0$ for which $\bar{F}[m]\bar{w} = S[m,w_*]u_0\approx d$.
+Warner and Guasch assume (implicitly) that there is always a
+(near-)zero-residual solution of the extended inversion problem. That
+is, they assume that for any $m$, there is an extended source
+$\bar{w}$, or equivalently an adaptive filter $u_0$ with $\bar{w}=w_**u_0$, for which $S[m]u_0\approx d$.
 If $\kappa \approx \kappa_*, \beta \approx \beta_*$, then $\bar{w}$ should be $\approx w_*$, so the adaptive kernel $u_0$ should be approximately $\delta(t)$ and independent of ${\bf x}_s,{\bf x}_r$.
@@
 \begin{itemize}
-\item[1. ]Given $m$, solve the deconvolution problem $(F[m]w_*)*u_0 = d$ for $u_0$, if necessary in a least squares sense and with Tihonov regularization;. 
+\item[1. ]Given $m$, solve the deconvolution problem $S[m]u_0 = d$ for
+  $u_0$, if necessary in a least squares sense and with Tihonov
+  regularization (``prewhitening'');
 \item[3. ]Compute the objective 
@@
     \label{eqn:tJz}
-    \tJz[m] = \sum_{x_s,x_r} \left(\frac{\int dt |tu_0|^2}{\int dt |u_0|^2}\right)
+    \tJz[m] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu_0(\bx_r,t;\bx_s)|^2}{\int dt |u_0(\bx_r,t;\bx_s)|^2}\right)
   \end{equation}
@@
 
-The normalization of the integrand per source and receiver by the $L^2$ norm of $u_0$ is quite important, as will be explained shortly.
+The trace-by-trace normalization of the integrand by the
+$L^2$ norm of $u_0$ is quite important, as will be explained shortly.
 
@@
 
-For a particular source-receiver pair ${\bf x}_s,{\bf x}_r$ with $\bx_r$ "not too far" from $\bx_s$,, suppressed from the notation,
+For a particular source-receiver pair ${\bf x}_s,{\bf x}_r$ with $\bx_r$ "not too far" from $\bx_s$, suppressed from the notation,
 \begin{equation}
@@
   \label{eqn:spwe}
-S[m,w_*]u(t) \approx a[m](w_**u)(t-\tau[m])
+S[m]u(t) \approx a[m](w_**u)(t-\tau[m])
 \end{equation}
@@
 $$
-The (approximate) solution of $S[m,w_*]u_0=d$ is obvious by inspection: $u_0(t) = a[m_*]/a[m]\delta(t-\tau[m_*]+\tau[m])$.
+The (approximate) solution of $S[m]u_0=d$ is obvious by inspection: $u_0(t) = a[m_*]/a[m]\delta(t-\tau[m_*]+\tau[m])$.
 There is immediately a big problem: this $u_0$ is not square-integrable, and the AWI objective $\tJz$ is undefined.
@@
 $$
-u_0^{\epsilon} = \mbox{argmin}_u (\|S[m,w_*]u-d\|^2 + \epsilon^2\|u\|^2)
+u_0^{\epsilon} = \mbox{argmin}_u (\|S[m]u-d\|^2 + \epsilon^2\|u\|^2)
 $$
@@
   \label{eqn:Ja}
-  \Ja[m,u;d] = \frac{1}{2}\left(\|S[m,w_*]u-d\|_d^2 + \epsilon^2\|u\| + \alpha^2\|tu\|_m\right).
+  \Ja[m,u;d] = \frac{1}{2}\left(\|S[m]u-d\|_d^2 + \epsilon^2\|u\| + \alpha^2\|tu\|_m\right).
 \end{equation}
@@
 $U$ is $\oplus_{\bx_r,\bx_s}L^2([-T_u,T_u])$, $D$ is the data space $\oplus_{\bx_r,\bx_s}L^2([0,T])$,.
-$S$ is the regularized modeling operator $(S[m,w_*],\epsilon I)^T$, $d$ is replaced by the $(d,0) \in D \times U$,  $W$ is the weight operator on $U$ defined in \ref{eqn:defmnorm}.
+$S$ is the regularized modeling operator $(S[m],\epsilon I)^T$, $d$ is replaced by the $(d,0) \in D \times U$,  $W$ is the weight operator on $U$ defined in \ref{eqn:defmnorm}.
 
-As noted before [but it wasn't - fix!], stationary points of $\tJz$ are $\lambda(w_*)$-close to stationary points of the travel time mean-square error, in the single-arrival case.
-The result of this section implies that stationary points of $\tJa$ are close to those of $\tJz$.
-Therefore, at least for small $\alpha$, minimization of $\tJa$ produces acoustic models that closely match the traveltimes predicted in single-arrival data.
+As noted before [but it wasn't - fix!], stationary points of $\tJz$
+are $\lambda(w_*)$-close to stationary points of the travel time
+mean-square error, in the single-arrival case. The result of this
+section implies that stationary points of $\tJa$ are close to those of
+$\tJz$. Therefore, at least for small $\alpha$, minimization of $\tJa$
+produces acoustic models that closely match the traveltimes predicted
+in single-arrival data.
+
+\section{Computation of the Reduced Penalty Function and its Gradient}
+Like all penalty functions, this one is best viewed as a least-squares objective of a map with values in a product space: define
+$$
+A_{\alpha,\epsilon}[m] =  
+\left[
+\begin{array}{c}
+S[m]\\
+\epsilon I\\
+\alpha T_{\epsilon}[m]
+\end{array}
+\right]
+, \, b = \left[
+\begin{array}{c}
+d\\
+0\\
+0
+\end{array}
+\right]
+$$
+Then
+$$
+J_{\alpha,\epsilon}[m,w_*,u] = 0.5\|A_{\alpha,\epsilon}[m]u-b\|^2
+$$
+
+Both Newton's method with the Kaufman Hessian approximation (described in the VCL notebook) and LBFGS or similar first-order methods apply to minimization of $\tilde{J}_{\alpha,\epsilon}[m]$.
+As shown in the VCL notebook, application of these algorithms to a VPM reduction like $\tilde{J}_{\alpha,\epsilon}$ requires access to two linear operator valued functions of $m$, namely $A_{\alpha,\epsilon}[m]$ and the (partial) derivative $D_m(A_{\alpha,\epsilon}[m]u)$.
+The former is the main actor in computation of the objective via solution of the normal equations.
+The latter appears in the VPM gradient formula:
+$$
+\nabla \tilde{J}_{\alpha,\epsilon}[m] = (D_m A_{\alpha,\epsilon}[m]u)|_{u=\tilde{u}_{\alpha,\epsilon}[m]}^T(A_{\alpha,\epsilon}[m]\tilde{u}_{\alpha,\epsilon}[m] -b)
+$$
+Note that $D_m A_{\alpha,\epsilon}[m]u$ is a map from the model parameter space to the product range space of $A$, and the transpose is a map in the other direction.
+
+\subsection{Computing $D_mA_{\alpha,\epsilon}[m]^T$}
+The main goal of this and the next two subsections is to express the
+operator in the subsection title in terms of
+\begin{itemize}
+  \item trace-by-trace convolution and cross-correlation operators,
+  \item the derivative $D_mF[m]w_*$ and its adjoint, and
+  \item trace-by-trace scaling.
+\end{itemize}
+These the first two bullets describe common components of FWI.The
+third is necessary to express the AWI penalty operator.
+
+Note that the second-mentioned operator describes the derivative of
+the acoustic forward modeling operator for a single point source per
+receiver gather, not the extended modeling operator $\bar{F}$. The
+latter is not a modular component of FWI, and a self-contained
+implementation is not likely to be available.
+
+Write $\nabla \tilde{J}_{\alpha,\epsilon}$ in terms of its components:
+\begin{equation}
+= D_mS[m]^T e_0 + \alpha D_mT_{\epsilon}[m]^Te_2
+\label{eqn:tildejgrad}
+\end{equation}
+where $e=(e_0,e_1,e_2)^T$ is the residual vector:
+$$
+e = \left[
+\begin{array}{c}
+S[m]u-d\\
+\epsilon u\\
+\alpha T[m]u
+\end{array}
+\right],\, u=\tilde{u}_{\alpha,\epsilon}[m].
+$$
+Note that the regularization term does not appear in the VPM gradient expression, since it does not depend explicitly on $m$.
+
+Evidently the key calculations are $D_mS^T$ and $D_m T^T$.
+
+\subsection{Convolution Revisited}
+
+To keep track of the data flow in subsequent calculations, it is convenient to introduce a notation for the operator of convolution by $u$.
+This notation must include the time ranges for input and output traces, as these define the index limits in the sums that implement the convolution operator.
+Looked at another way, these ranges define the domain and range of the operator, and that is how the definition in *segyvc.ConvolutionOperator* is arranged: the arguments to the class constructor are the domain and range (as *segyvc.Spaces*, type-checked) and the filename for the convolution kernel.
+Domain, range, and kernel must share spatial geometry (source and receiver positions, in particular number of traces, and time step; these conditions are checked.
+In this section, the convolution operator with compatible trace space domain $X$ and range $Y$, and kernel $z \in Z$ is denoted $C[X,Y,Z,z]$.
+The kernel space $Z$ is also compatible with $X$ and $Y$.
+
+The domain, range, and kernel spaces considered here are all coordinate spaces, defined by finite time index ranges (or, in the continuum case, bounded time ranges).
+For such a coordinate space $X$, denote by $\Pi_X$ the orthogonal projection of the ambient space $l^2$ (or $L^2({\bf R})$ onto $X$.
+In the discrete case, this projection $\Pi_X$ simply extracts the values inside the characterizing index range $I_X$:
+$$
+(\Pi_X f)_i = f_i, \, i \in I_X,
+$$
+and similarly for the continuum case.
+The transpose projection $\Pi_X^T$ injects the array of values within the range into $l^2$, with zeros outside.
+So $\Pi_X^T\Pi_X$ is the projection onto the subspace of $l^2$ (or $L^2({\bf R})$) corresponding to $X$, i.e.
+the range of $\Pi_X^T$, and $\Pi_X \Pi_X^T = I_X$.
+
+Define convolution $f*g$ by the usual sum or integral, for arguments $f$ and $g$ of bounded support. With these conventions,
+$$
+C[X,Y,Z,z]x = \Pi_Y(\Pi_Z^T z * \Pi_X^T x).
+$$
+The class *segyvc.ConvolutionOperator* implements precisely the discrete version of this definition.
+
+Both discrete and continuous convolution are commutative.
+The same is true of the discrete, finite version of convolution implemented in *segyvc.ConvolutionOperator*, and can be expressed as the follow obvious consequence of the definition: if $X, Y, Z$ are compatible trace spaces, and $z \in Z, x \in X$, then
+\begin{equation}
+C[X,Y,Z,z]x = C[Z,Y,X,x]z
+\label{eqn:convcomm}
+\end{equation}
+The transpose of convolution is expressed via the time reversal operator $f \mapsto \check{f}\, \check{f}_i = f_{-i}$:  
+$$
+\langle f*g,h \rangle = \langle g, \check{f}*h \rangle.
+$$ 
+The right-hand side defines the cross-correlation of $f$ and $h$. From the definition of $C$,
+$$
+\langle C[X,Y,Z,z]x, y\rangle_Y = \langle \Pi_Y(\Pi_Z^T z * \Pi_X^T x), y \rangle_Y
+$$
+$$
+= \langle \Pi_Z^T z * \Pi_X^T x, \Pi_y^T y \rangle = \langle \Pi_X^T x, \check{(\Pi_Z^T z)}* \Pi_Y^T y \rangle
+$$
+Denote by $\check{Z}$ the coordinate space produced by time reversal applied to $Z$, that is, $I_{\check{Z}}$ consists of the negatives of members of $I_Z$.
+Then $\check{(\Pi_Z^T z)}=\Pi_{\check{Z}}\check{z}$, so the above is
+$$
+=\langle \Pi_X^T x, \Pi_{\check{Z}}^T \check{z} * \Pi_Y^T y \rangle =\langle x, \Pi_X(\Pi_{\check{Z}}^T \check{z} * \Pi_Y^T y) \rangle_X
+$$
+$$
+= \langle x, C[Y,X,\check{Z},\check{z}]y \rangle_X.
+$$
+That is, $C[X,Y,Z,z]^T = C[Y,X,\check{Z},\check{z}]$.
+
+Combining the commutativity relation $\ref{eqn:convcomm}$ with this identity yields the useful identity:
+\begin{equation}
+C[X,Y,Z,z]^Ty = C[Y,X,\check{Z},\check{z}]y =
+C[\check{Z},X,Y,y]\check{z} =
+C[X,\check{Z},\check{Y},\check{y}]^T\check{z}.
+\label{eqn:convcommtransp}
+\end{equation}
+
+\subsection{Back to AWI penalty gradient}
+
+The SEGY trace spaces involved in the AWI calculation are the data space $D$ and the adaptive filter space $U$.
+Minimization of $J_{\alpha,\epsilon}$ amounts to solution of a least squares problem with operator $C[U,D,D,F[m]w]$, that is, convolution with the predicted data:
+$$
+S[m]u = C[U,D,D,F[m]w_*]u = C[D,D,U,u]F[m]w_*
+$$
+The second expression, equivalent by the commutativity relation \ref{eqn:convcomm} to the first, is more convenient for computing the derivative with respect to $m$ of $S[m]u$:
+$$
+D_m(S[m]u) \delta m = C[D,D,U,u](D_mF[m]\delta m)
+$$
+whence ($e_0 = S[m]u_{\alpha,\epsilon}[m]-d \in D$) the first summand in the gradient of $\tilde{J}_{\alpha,\epsilon}$ is
+$$
+D_m(S[m]u)^T e_0 = D_mF[m]^TC[D,D,U,u]^Te_0.
+$$
+%The derivative $D_mF[m]$ and its adjoint are defined in the *vcl.Function* subclass *asg.fsbop*, and the convolution operator $C$ and its transpose in *segyvc.ConvolutionOperator*.
+
+Introduce the set $SR$ of source-receiver pairs active in the data
+$D$, and scale factor space $P = \{f:SR \rightarrow \bR\}$. Write
+\[
+  T[m]u = B_2 \circ B_1 \circ B_0[m],
+\]
+where 
+\[
+B_2: P \times U \rightarrow U, \, B_2[f,u](bx_r,t;\bx_s) = f(\bx_r,\bx_s) tu(\bx_r,t;\bx_s)
+\]
+\[
+B_1: U \rightarrow P, \,B_1[u](\bx_r,t,\bx_s) = \|u(\bx_r,\cdot; \bx_s)\|^{-1}
+\]
+\[
+B_0: M \rightarrow U, \, B_0[m] = u_{0,\epsilon}[m]
+\]
+Then
+\[
+ \frac{\alpha^2}{2} D_m(\|T[m]u\|_U^2)\delta m = \alpha^2\langle D_m(
+ T[m]u)\delta m, T[m]u \rangle_U
+\]
+\[
+  =\alpha^2 \langle DB_2(B_1\circ B_0[m],u)DB_1(B_0[m]) DB_0[m]\delta m,
+  T[m]u\rangle_U
+\]
+\begin{equation}
+  \label{eqn:compgrad}
+  =\alpha^2 \langle DB_0[m]\delta m, DB_1(B_0[m])^TDB_2(B_1\circ
+  B_0[m],u)^TT[m]u\rangle_U
+\end{equation}
+Next display the derivative and adjoint derivative of $B_1$ and $B_2$:
+$$
+(D_fB_2[f,u]\delta f)(\bx_r,t;\bx_s) = tu(\bx_r,t;\bx_s)\delta f(\bx_r,\bx_s)
+$$
+$$
+D_f(B_2[f,u])^Te_2(\bx_r,\bx_s) = \int dt\,te_2(\bx_r,t;\bx_s)u(\bx_r,t;\bx_S);
+$$
+$$
+D_u B_1[u]\delta u(\bx_r,\bx_s) = -\|u(\bx_r,\cdot;\bx_s)\|^{-3}\langle u(bx_r,\cdot;\bx_s), \delta u(\bx_r,\cdot;\bx_s) \rangle
+$$
+$$
+(D_u B_1[u])^Tf(\bx_r,t;\bx_s) =
+-f(\bx_r,\bx_s)\|u(\bx_r,\cdot;\bx_s\|^{-3}u(\bx_r,t;\bx_s)
+$$
+With these expressions, compose various components appearing on the
+right hand side of equation \ref{eqn:compgrad}:
+\[
+  B_1\circ B_0[m] (\bx_r,\bx_s) =  \|\tilde{u}_{0,\epsilon}(\bx_r,\cdot; \bx_s)\|^{-1},
+\]
+\[
+ DB_2(B_1\circ B_0[m],u)^Te_2 = \int dt\,te_2(\bx_r,t;\bx_s)u(\bx_r,t;\bx_s),
+\]
+(Note that the result is independent of the first argument.)
+\[
+DB_1(B_0[m])^Tf(\bx_r,t:\bx_s)=
+-f(\bx_r,\bx_s)\|u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\|^{-3}u_{0,\epsilon}(\bx_r,\cdot;\bx_s),
+\]
+and
+\[
+  DB_1(B_0[m])^TDB_2(B_1\circ  B_0[m],u)^TT[m]u =
+\]
+\begin{equation}
+  \label{eqn:compgrad1}
+  -\int dt\,t(T[m]u)(\bx_r,t;\bx_s)u(\bx_r,t;\bx_s)
+  \|u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\|^{-3}u_{0,\epsilon}(\bx_r,\cdot;\bx_s)
+\end{equation}
+$$
+u_{0,\epsilon}[m] = (S[m]^TS[m] + \epsilon^2 I)^{-1}S[m]^Td
+$$
+$$
+= (C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*] + \epsilon^2 I)^{-1}C[U,D,D,F[m]w_*]^Td.
+$$
+So
+$$
+D_mB_0[m]\delta m = (C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*] + \epsilon^2 I)^{-1}
+$$
+$$
+\times ((D_mC[U,D,D,F[m]w_*]\delta m)^Td -(D_m(C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*])\delta m)u_{0,\epsilon}[m])
+$$
+$$
+=(C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*] + \epsilon^2 I)^{-1}
+$$
+$$
+\times
+(D_m(C[U,D,D,F[m]w_*]^Tr_0)\delta m  - C[U,D,D,F[m]w_*]^TD_m(C[U,D,D,F[m]w_*]u_0)\delta m)
+$$ 
+where the derivatives are to be evaluated before the insertion of the definitions 
+$$
+r_0 = d -C[U,D,D,F[m]w_*]u_{0,\epsilon}[m], \, u_0 = u_{0,\epsilon}[m]
+$$
+From the commutativity relation,
+$$
+D_m(C[U,D,D,F[m]w_*]u_0)\delta m = D_m(C[D,D,U,u_0]F[m]w_*)\delta m
+=C[D,D,U,u_0]D_mF[m]\delta m.
+$$
+Similarly, use the identity \ref{eqn:convcommtransp} to write
+$$
+C[U,D,D,F[m]w_*]^Tr_0 = C[U,\check{D},\check{D},\check{r_0}]^T({F[m]w_*})\check, 
+$$
+so
+$$
+D_m(C[U,D,D,F[m]w_*]^Tr_0)\delta m=
+C[U,\check{D},\check{D},\check{r_0}]^T(D_mF[m]w_* \delta m)\check.
+$$
+
+Putting these calculations together,
 

### 2023-10-21 repro/papers/awigrad/paper.tex
initial

--- 2023-10-21 repro/papers/awi/paper.tex
@@
-\title{Notes on AWI}
+\title{Computation of the Reduced AWI Penalty Function and its Gradient}
 \author{William W. Symes}
@@
 
-\section{Introduction}
+\section{Definitions}
+Hilbert space $X$ has inner product $\langle \cdot,\cdot \rangle_X$ and norm $\|\cdot\|_X$.
 
-\section{AWI and why it works}
+$M$ = model set, an open set in a Banach space also pre-Hilbert structure
 
-The version of AWI introduced by Michael Warner and Lluis Guasch \cite[]{Warner:16,GuaschWarnerRavaut:GEO19} rests on the assumption that seismic waves are governed by linear acoustics, and that each shot is associated to an isotropic point source with known location $\bx_s$ and wavelet $w(t;\bx_s)$.
-That is, the pressure and velocity fields $p({\bf x},t;{\bf x}_s)$, ${\bf v}({\bf x},t;{\bf x}_s)$ for the shot location ${\bf x}_s$ depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf x})$ (reciprocal of the density $\rho({\bf x})$), and wavelet $w(t;{\bf x}_s)$ through the acoustic system
- $$
- \frac{\partial p}{\partial t} = - \kappa \nabla \cdot {\bf v} +
-w(t;{\bf x}_s) \delta({\bf x}-{\bf x}_s);
-\frac{\partial {\bf v}}{\partial t} = - \beta \nabla p; 
-p, {\bf v} = 0 \mbox{ for }  t \ll 0.
-$$
-For sake of brevity, define the model vector $m=(\kappa,\rho)$.
-The forward map or {\em modeling operator} is $F[m]w = \{p({\bf x}_r,t;{\bf x}_s)\}$, in which shot and receiver positions ${\bf x}_s, {\bf x}_r$ define the acquisition geometry.
+$Q$ = finite ``acquisition'' set of source-receiver location pairs
 
-Acoustic inversion, for present purposes, means: given time interval $[0,T]$, source and receiver locations $(\bx_s,\bx_r)$, source wavelet $w_* \in C_0^{\infty}(\bR)$ and data traces $d(\bx_r,\cdot;\bx_s) \in L^2([0,T])$, find a model $m$ so that $F[m]w_* \approx d$.
-The simplest version of FWI concretizes this task by asking for a model $m$ minimizing the mean square error
+$P = \bR^{|Q|}$ = real functions on $Q$, with Euclidean norm (no attempt to scale for geometry of acquisition).
+
+$[0,t_d]$ = acquisition time interval
+
+$[-t_u,t_u]$ = adaptive filter time interval
+
+$D$ = data Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([0,t_d])$ (again, no attempt to scale for geometry of acquisition).
+
+$U$ = adaptive filter Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([-t_u,t_u])$
+
+$w_* \in C_0^{\infty}(\bR)$ = known point source wavelet
+
+$F: M \rightarrow D$ = forward modeling operator, use of $w_*$ implicit. Since $w_*\in C_0^{\infty}(\bR)$  and distance from source points to receiver points is bounded below, $F \in C^1(M,D)$.
+
+Modeling Bound Assumption: $\|F[m](\bx_r,\cdot;\bx_s)\|_{L^2([0,t_d])}$ is uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
+
+$S: M \rightarrow {\cal B}(U,D)$ = trace-by-trace convolution with predicted data: $S[m]u = F[m] * u$.
+Due to smoothness of $F$, $S \in C^1(M,{\cal B}(U,D))$.
+Operation is trace-by-trace, that is block-diagonal: $S[m] = \bigoplus_{(\bx_r,\bx_s) \in Q}S[m]_{\bx_r,\bx_s}$.
+The Modeling Bound Assumption implies that $\|S[m](\bx_r,\cdot;\bx_s)\|_{L^2([0,t_d])}$ is also uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
+
+No-zero-traces Assumption: there exists $C>0$ so that for all $(\bx_r,\bx_s) \in Q$,
+$\|d(\bx_r,\cdot;\bx_s)\|_{L^2([0,t_d])} \ge C.$
+
+Observable Data Assumption: there exists $C>0$ so that for all $m \in M, (\bx_r,\bx_s) \in Q$,
 \begin{equation}
-  \label{eqn:fwi}
-  J_{\rm FWI}[m;d] = \frac{1}{2}\|F[m]w_*-d\|^2
+  \label{eqn:obsdata}
+  \|S[m]_{\bx_r,\bx_s}^Td(\bx_r,\cdot;\bx_s)\|_{L^2[-t_u,t_u]} \ge C \|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_d]}
+\end{equation}.
+That is, the projection of each trace onto the range of convolution of the corresponding predicted data trace is coercive.
+
+$\sigma > 0$ = Tihonov regularization weight
+
+$\alpha \ge 0$ = penalty weight
+
+Definition of AWI penalty objective divides into two parts:
+
+Part 1: Unpenalized adaptive filter $u_{0,\sigma}[m,d]$ solving regularized least squares data fitting problem:  given $m \in M$, minimizes
+\[
+ J_{0,\sigma}[u,m,d] = \frac{1}{2}(\|S[m]u - d\|_D^2 + \sigma^2 \|u\|^2_U).
+\]
+Since $\sigma > 0$, normal operator is uniformly bounded below and of class $C^1$ in $M$ and $D$. So unique solution exists, and $u_{0,\sigma} \in C^1(M \times D, U)$.
+
+Also, the normal operator is block-diagonal, so each trace solves a least-squares problem, and
+\[
+  (S[m]_{\bx_r,\bx_s}^TS[m]_{\bx_r,\bx_s} + \sigma^2 I)u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)= S[m]_{\bx_r,\bx_s}^Td(\bx_r,\cdot;\bx_s)
+\]
+According to the Modeling Bound Assumption, observable data assumption, the left-hand side is bounded by a multiple of $\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_u,t_u]}$, uniformly in $m \in M, (\bx_r,\bx_s) \in Q$.
+According to the Observable Data Assumption, the right-hand side is bounded below by a multiple of $\|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_d]}$, uniform in the same sense.
+Conclude that there exists $C>0$ so that
+\begin{equation}
+  \label{eqn:u0lower}
+  \|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_u,t_u]} \ge C \|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_d]}
 \end{equation}
-Practical versions of FWI include various types of regularization and homotopy, see the sources cited in the Introduction.
-As mentioned there, local iterative optimization methods tend to stagnate at highly sub-optimal (``cycle-skipped'') estimates of $m$ unless the initial estimate is of quite high quality, often available only at considerable cost in effort and data collection.
-%For now, assume that the data $d({\bf x}_r,t;{\bf x}_s)$ is the output of the modeling operator for "true" model $m_*$, that is, the "true" bulk modulus, buoyancy, and wavelet $\kappa_*, \beta_*, w_*$: that is, $d = F[m_*]w_*$.
+uniformly in $m \in M, (\bx_r,\bx_s) \in Q$.
 
-AWI is based on extending the modeling operator $F$, that is, enlarging its domain.
-The extended modeling operator ${\bar F}$ maps extended sources
-$\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
-field. That is, the extended source depends on the receiver location
-as well as the source location - so there is one acoustic system for
-each source *and* receiver position - a lot of wave equations! If all
-of the wavelets for each source are the same, that is, ${\bar w}({\bf
-  x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver
-position, then ${\bar F}[m]\bar{w} = F[m]w$. That is, $F$ is a special
-case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
-$F$. This is the {\em source-receiver} extension, in the terminology of \cite{HuangSymes2015SEG}.
+Part 2:
+Preconditioned penalty operator: block diagonal, 
+\begin{equation}
+  \label{eqn:ppf}
+  T_{\sigma}[m]u (\bx_r,t;\bx_s) = \frac{u(\bx_r,t;\bx_s)}{\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_u,t_u]} }
+\end{equation}
+Well-defined and $C^1$ as function of $m$ due to the bound \ref{eqn:u0lower} and the the No-zero-trace assumption.
 
-AWI assumes that the extended sources are time convolutions of the
-(known) exact source $w_*$ with an {\em adaptive filter} $u({\bf x}_r,t;{\bf x}_s)$: that is, $\bar{w} = u * w_*$ - the asterisk denotes convolution in time.
-Since linear acoustics is time-translation invariant, its solution commutes with time convolution, that is,
+Define for $\alpha \ge 0$
 \[
- \bar{F}[m]\bar{w} = \bar{F}[m](u* w_*) = (F[m]w_*)*u
+ J_{\alpha,\sigma}[u,m,d] = \frac{1}{2}(\|S[m]u - d\|_D^2 + \alpha^2\|T_\sigma[m]u\|_U^2 + \sigma^2 \|u\|^2_U).
 \]
-Since the known source $w_*$ is not adjusted during the inversion, 
-introduce the notation $S[m]$ for the convolution operator with kernel $F[m]w_*$:
-\begin{equation}
-  \label{eqn:sdef}
-  \equiv S[m]u
-\end{equation}
-Warner and Guasch assume (implicitly) that there is always a
-(near-)zero-residual solution of the extended inversion problem. That
-is, they assume that for any $m$, there is an extended source
-$\bar{w}$, or equivalently an adaptive filter $u_0$ with $\bar{w}=w_**u_0$, for which $S[m]u_0\approx d$.
-If $\kappa \approx \kappa_*, \beta \approx \beta_*$, then $\bar{w}$ should be $\approx w_*$, so the adaptive kernel $u_0$ should be approximately $\delta(t)$ and independent of ${\bf x}_s,{\bf x}_r$.
-Such a kernel $u_0$ is (approximately) in the null space of multiplication by $t$.
-
-This observation leads to the definition of the Warner-Guasch AWI objective function $\tJz$:
-
-\begin{itemize}
-\item[1. ]Given $m$, solve the deconvolution problem $S[m]u_0 = d$ for
-  $u_0$, if necessary in a least squares sense and with Tihonov
-  regularization (``prewhitening'');
-\item[3. ]Compute the objective 
-  \begin{equation}
-    \label{eqn:tJz}
-    \tJz[m] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu_0(\bx_r,t;\bx_s)|^2}{\int dt |u_0(\bx_r,t;\bx_s)|^2}\right)
-  \end{equation}
-and its gradient, then update $\kappa,\beta$ using a gradient descent method to minimize $\tJz$.
-\end{itemize}
-
-The trace-by-trace normalization of the integrand by the
-$L^2$ norm of $u_0$ is quite important, as will be explained shortly.
-
-It may not be obvious at first glance why this approach to estimating $m$ should be any more effective than FWI.
-To see why AWI works, in at least some settings, we assume that the spatial dimension is three, and make use of the geometric optics approximation to the solution of the wave equation, in a form due to Hadamard.
-This approximation is {\em local}: it applies when sources and receivers are sufficiently close.
-We will discuss its global failure and the consequences for AWI in a later section.
-
-Transient acoustic sources have no DC (zero frequency) component and are relatively small at low frequencies.
-One was to enforce these characteristics is to assume that the wavelet $w_*$ is the time derivative of a $v_* \in C_0^{\infty}(\bR)$.
-Set $\lambda(w_*) = \|v_*\|/\|w_*\|$ ($L^2$ norms).
-$\lambda$ is a proxy for wavelength: it is the reciprocal of the RMS frequency.
-
-For a particular source-receiver pair ${\bf x}_s,{\bf x}_r$ with $\bx_r$ "not too far" from $\bx_s$, suppressed from the notation,
-\begin{equation}
-  \label{eqn:hadamard}
-  F[m]w_*(t) = a[m]w_*(t-\tau[m]) + Rw_*(t)
-\end{equation}
-where $R(\bx_r,\bx_s)$ is a bounded operator on $L^2((-\infty,T])$ bounded by a multiple of $\lambda(w_*)$.
-This ``single arrival'' approximation  follows from the expansion of the fundamental solution of the wave equation due to Hadamard, see for instance \cite{Friedlander:75}.
-This construction is sketched in Appendix B.
-Note that in general the approximation \ref{eqn:hadamard} is only local: for smooth $m$, it holds for $\bx_r$ sufficiently close to $\bx_s$, but fails at larger distances.
-
-For approximations with $O(\lambda(w_*)$ remainders, write
-\begin{equation}
-  \label{eqn:pwe}
-  F[m]w_*(t) \approx a[m]w_*(t-\tau[m])
-\end{equation}
-Thanks to the approximation \ref{eqn:pwe},
-\begin{equation}
-  \label{eqn:spwe}
-S[m]u(t) \approx a[m](w_**u)(t-\tau[m])
-\end{equation}
-Assume that the data is noise-free:
-$$
-d(t)=F[m_*]w_*(t) \approx a[m_*]w_*(t-\tau[m_*])
-$$
-The (approximate) solution of $S[m]u_0=d$ is obvious by inspection: $u_0(t) = a[m_*]/a[m]\delta(t-\tau[m_*]+\tau[m])$.
-There is immediately a big problem: this $u_0$ is not square-integrable, and the AWI objective $\tJz$ is undefined.
-To fix this, it's necessary to treat $u_0$ as the solution of a regularized least squares problem: define
-$$
-u_0^{\epsilon} = \mbox{argmin}_u (\|S[m]u-d\|^2 + \epsilon^2\|u\|^2)
-$$
-$$
-\approx \mbox{argmin}_u (\|a[m](w_**u)(\cdot-\tau[m])-a[m_*]w_*(\cdot-\tau[m_*])\|^2 + \epsilon^2\|u\|^2)
-$$
-Use the temporary abbreviations $a_*=a[m_*], a=a[m], \tau_*=\tau[m_*], \tau=\tau[m]$, and write in terms of the Fourier transforms $\hat{u}_{\epsilon},\hat{w_*}$ of $u_0^{\epsilon}$ and $w_*$:
-$$
-\hat{u}_{\epsilon} \approx \mbox{argmin}_{\hat{u}}\left(\int d\omega |a\hat{w_*}\hat{u}e^{i\omega \tau}-a_*\hat{w_*}e^{i\omega\tau_*}|^2 + \epsilon^2 |\hat{u}|^2\right)
-$$
-The normal equation is
-$$
-(a^2 |\hat{w}_*|^2 +\epsilon^2)\hat{u} = a a_*|\hat{w}_*|^2e^{i\omega(\tau_*-\tau)}
-$$
-the solution of which is
-$$
-\hat{u}_{\epsilon} \approx \frac{a^*}{a}\hat{g}_{\frac{a^*}{a}\epsilon} e^{i\omega(\tau_*-\tau)}
-$$
-where
-$$ 
-\hat{g}_{\epsilon} = \frac{a_*^2|\hat{w}_*|^2}{a_*^2|\hat{w}_*|^2 + \epsilon^2}.
-$$
-Since $w_* \in C_0^{\infty}$, $\hat{w_*}(\omega) \rightarrow 0$ faster than any negative power of $\omega$ as $|\omega| \rightarrow \infty$ and is analytic, whence $\hat{g}_{\epsilon}: \epsilon > 0$ tends to $1$ almost everywhere as $\epsilon \rightarrow 0$.
-Hence the inverse Fourier transform $g_{\epsilon}$ tends to $\delta$ in the sense of distributions, and
-$$
-u_0^{\epsilon}(t) \approx \frac{a^*}{a}g_{\frac{a^*}{a}\epsilon}(t-(\tau_*-\tau))
-$$
-to a multiple of a shifted $\delta$, in fact exactly the distribution solution of $\bar{S}[m,w_*]u=d$. However for $\epsilon>0$, $u_0^{\epsilon}$ is square integrable.
-
-\cite{HuangSymes:SEG15a} presented an analysis of the unweighted AWI objective (numerator of the right hand side in the definition \ref{eqn:tJz}), based in turn on ideas presented in Song's thesis \cite[]{Song:94c}.
-With $u$ replaced by $u_0^{\epsilon}$,  this unweighted objective is
-$$
-\int dt |tu_0^{\epsilon}|^2 \approx \frac{a_*^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\epsilon}(t-(\tau_*-\tau))|^2
-$$
-$$
-=\frac{a_*^2}{a^2} \int dt\, (t+(\tau_*-\tau))^2|g_{\frac{a^*}{a}\epsilon}(t)|^2
-$$
-The linear (in t) term vanishes, since $g$ is necessarily even in time (since it Fourier transform is real). So this is
-$$
-=\frac{a_*^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\epsilon}(t)|^2
-+(\tau_*-\tau)^2\frac{a_*^2}{a^2}\|g_{\frac{a^*}{a}\epsilon}\|^2
-$$
-The first term tends to zero as $\epsilon \rightarrow 0$, since $g_{\epsilon}$ is a delta-family, so you can make it negligible with proper choice of $\epsilon$.
-The second term suggests that this function is related to the travel-time error $\tau_*-\tau$.
-
-The relation is obscured by the presence of of the amplitude ratio $a_*/a$ both as a multiplier and in the scaling of $\epsilon$.
-Since the amplitudes depend on $m$, their presence may influence the location (or even the presence) of  stationary points of $\tJz$ for models differing in their travel time predictions.
-So stationary points of the unweighted AWI objective may not yield slowness models that match travel-times with the data, even in the noise-free case (even though the examples reported in Huang and Symes SEG 2015, and other works cited there, suggest that sometimes this is the case).
-
-The scaling by trace norms of $u_0^{\epsilon}$ in the definition \ref{eqn:tJz} eliminates the possibility of non-travel-time stationary points.
-Since the objective function is built up trace-by-trace, use the same approximation for the norm of $u_0^{\epsilon}$:
-$$
-\|u_0^{\epsilon}\| \approx \frac{a^*}{a}\|g_{\frac{a^*}{a}\epsilon}\|
-$$
-So
-$$
-J_0[m,w_*] \approx \sum_{{\bf x}_s,{\bf x}_r} (\tau[m_*]({\bf x}_s,{\bf x}_r)-\tau[m]({\bf x}_s,{\bf x}_r))^2
-$$
-ignoring the $\|tg_{...}\|^2$ term as being negligible for small enough $\epsilon$.
-
-In other words, the AWI objective is *exactly* an approximation to the travel-time tomography objective.
-And that's why it works - when geometric optics in the form given here is an accurate approximation to the Green's function.
-
-It remains to see what geometric optics has to say about the penalty objective $\tilde{J}_{\alpha}[m,w_*]$, how geometric optics might fail, and what happens to AWI when it does.
-
-\section{A Penalty Version of AWI}
-The adaptive kernel $u_0^{\epsilon}$ appearing in the definition \ref{eqn:tJz} is the solution of a regularized least squares problem.
-This observation suggests augmenting the regularized least squares objective by a penalty proportional to the right-hand side of \ref{eqn:tJz}:
-\begin{equation}
-  \label{eqn:Ja}
-  \Ja[m,u;d] = \frac{1}{2}\left(\|S[m]u-d\|_d^2 + \epsilon^2\|u\| + \alpha^2\|tu\|_m\right).
-\end{equation}
-Here $\|u\|^2 = \langle u, u \rangle$ is the (unweighted) $L^2$ norm, and $\|\cdot\|_m$ is an $(m,d)$-dependent (weighted) norm: with
-\[
-  u_0^{\epsilon}[m,d] = \mbox{arg min}_u \Jz[m,u;d],
-\]
-define
-\[
-  Wu (\bx_r,t;\bx_s) = \sum_{\bx_r,\bx_s}\frac{1}
-  {\int\,ds\,|u_0^{\epsilon}[m,d](\bx_r,s;\bx_s)|^2}u(\bx_r,t;\bx_s)
-\]
-and
-\begin{equation}
-  \label{eqn:defmnorm}
-  \|u\|^2_m = \langle u, W u \rangle.
-\end{equation}
-With this notation,
-\[
-  \tJz[m;d] = \frac{1}{2}\|tu^{\epsilon}_0[m;d]\|^2_m
-\]
-In the definition of $\tJz$, a quadratic is evaluated at the minimizer of $\Jz[m,u;d]$. This suggests the possibility of evaluating the quadratic objective $\Ja[m,u;d]$ at its own minimizer:
-\begin{equation}
-  \label{eqn:tJa}
-  \tJa[m;d] = \min_u \Ja[m,u,d] = \Ja[m,\ua[m;d];d]
-\end{equation}
-This is the so-called variable projection reduction of the quadratic $\Ja$.
-Note that the right-hand side of the definition of $\tJz$ is the last term on the right-hand side of $\Ja$, scaled by $\alpha^{-2}$ and evaluated at the minimizer of the sum of the first two terms, that is, $\Ja$ for $\alpha=0$.
-
-\begin{theorem}
-  \label{thm:basic}
-Suppose that $S \in {\cal L}(U,D)$, the is a bounded linear maps from (Hilbert space) domain $U$ (``model space'') with inner product $\langle \cdot, \cdot \rangle_m$ to range $D$ with inner product $\langle \cdot, \cdot \rangle_d$.
-Assume that $S$ is coercive, that is, there exists $\epsilon >0$ so that $\|Su\|_d \ge \epsilon\|u\|_m$ for $u \in U$.
-Suppose also that $A \in {\cal L}(U,U)$, and that $d \in D$. Define the function $J_{\alpha}$ on $U$ by 
-\begin{equation}
-  \label{eqn:eq1}
-  J_{\alpha}(u) = \frac{1}{2}(\|Su-d\|^2_d + \alpha^2\|Au\|_m^2).
-\end{equation}
-Then for $\alpha \ge 0$, $J_{\alpha}$ possesses a unique minimizer $u_{\alpha} \in U$, for which
-\begin{equation}
-  \label{eqn:eq0}
-  \tilde{J}_{\alpha} = \min_{u \in U} J_{\alpha}(u) = J_{\alpha}(u_{\alpha}),
-\end{equation}
-and
-\begin{equation}
-  \label{eqn:eq4}
-  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tilde{J}_{\alpha}-\tilde{J}_0).= \frac{1}{2}\|Au_0\|^2
-\end{equation}
-This limit is uniform over bounded sets of $(S, A) \in {\cal L}(U,D) \times {\cal L}(U, U)$.
-\end{theorem}
-
-For completeness, the proof is given in an appendix.
-
-This relation will be applied in the context of a model space $U$ possessing a reference Hilbert norm $\|\cdot\|$ and a family of equivalent norms $\|\cdot\|_m$ depending on a parameter $m \in M$, in which $M$ is an open subset of a Banach space $X$.
-The dependence is through a an operator-valued function $W$ on $M$, taking values in cone ${\cal B}(U)$ of bounded self-adjoint positive-definite operators on $U$.
-On the other hand, we assume that the norm on $D$ is fixed, independent of $M$, and treat it as a reference norm.
-
-\begin{theorem}
-  \label{thm:uniform}
-Assume that $W: M \rightarrow {\cal B}(U)$ is of class $C^1$ and uniformly positive definite, and that the norm $\|\cdot\|_m$ is given by
-\begin{equation}
-  \label{eqn:normdef}
-  \|u\|_m^2 = \langle u, W[m] u \rangle.
-\end{equation}
-Suppose also that $S: M \rightarrow {\cal L}(U,D)$ is of class $C^1$ and uniformly coercive, and that $A \in {\cal L}(U,U)$ as before. For $\alpha \ge 0$, define $\tJa: M \rightarrow \bR$ by
-\[
-  \tilde{J}_{\alpha}[m] = \min_{u \in U}\frac{1}{2}(\|S[m]u-d\|^2_d + \alpha^2\|Au\|_m^2)
-\]
-\begin{equation}
-  \label{eqn:tjadef}
-  = \frac{1}{2}(\|S[m]u_{\alpha}[m]-d\|^2_d + \alpha^2\|Au_{\alpha}[m]\|_m^2)
-\end{equation}
-where
-\begin{equation}
-  \label{eqn:ua}
-  u_{\alpha}[m] = \mbox{arg min}_{u \in U}\frac{1}{2}(\|S[m]u-d\|^2_d + \alpha^2\|Au\|_m^2)
-\end{equation}
-for $m \in M$. Then $\tilde{J}_{\alpha}$ is of class $C^1(M)$, and
-\begin{equation}
-  \label{eqn:lim}
-  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tJa[m]-\tJz[m]).= \frac{1}{2}\|A\uz[m]\|_m^2.
-\end{equation}
-The limit is in the sense of $C^1(M)$, that is, both the left-hand side and its $m$ derivative converge to the right-hand side and its $m$ derivative, respectively.
-Moreover $u_{\alpha}$ is of class $C^1(M,U)$.
-\end{theorem}
-
-\begin{proof} All conclusions but the last follow immediately from Theorem \ref{thm:basic} and the uniformity assumptions on $S$ and $W$.
-
-Since the norm in $D$ is fixed for the discussion, drop the subscript and treat it as a reference norm.
-Denote the adjoint with respect to the norms $\|\cdot\|_m$ on $U$, $\|\cdot\|_d = \|\cdot\|$ on $D$ by a superscript $*$, and with respect to $\|\cdot\|$ on $U$ by a superscript $T$.
-The normal equation defining $\ua$ is
-  \[
-    (S[m]^*S[m] + \alpha^2 A[m]^*A) \ua[m] = S[m]^*d
-  \]
-  \begin{equation}
-    \label{eqn:normal}
-= (W[m]^{-1}S[m]^TS[m] + \alpha^2 W[m]^{-1}A^TW[m]A)\ua[m]= W[m]^{-1}S[m]^Td
-\end{equation}
-or equivalently
-\begin{equation}
-  \label{eqn:normal1}
-  (S[m]^TS[m] + \alpha^2 A^TW[m]A)\ua[m] = S[m]^Td
-\end{equation}
-Since $S$ is uniformly coercive, and both $S$ and $W$ are of class $C^1$, the normal operator on the left hand side is boundedly invertible and of class $C^1$.
-Therefore the solution $\ua$ is of class $C^1$.
-
-[need to add proof of $C^1$ convergence.]
-\end{proof}
-
-Apply this relation in the context of the AWI penalty problem defined earlier.
-$M$ is an open set in $C^k(\Omega, \bR^2)$.
-$U$ is $\oplus_{\bx_r,\bx_s}L^2([-T_u,T_u])$, $D$ is the data space $\oplus_{\bx_r,\bx_s}L^2([0,T])$,.
-$S$ is the regularized modeling operator $(S[m],\epsilon I)^T$, $d$ is replaced by the $(d,0) \in D \times U$,  $W$ is the weight operator on $U$ defined in \ref{eqn:defmnorm}.
-
-As noted before [but it wasn't - fix!], stationary points of $\tJz$
-are $\lambda(w_*)$-close to stationary points of the travel time
-mean-square error, in the single-arrival case. The result of this
-section implies that stationary points of $\tJa$ are close to those of
-$\tJz$. Therefore, at least for small $\alpha$, minimization of $\tJa$
-produces acoustic models that closely match the traveltimes predicted
-in single-arrival data.
-
-\section{Computation of the Reduced Penalty Function and its Gradient}
-Like all penalty functions, this one is best viewed as a least-squares objective of a map with values in a product space: define
-$$
-A_{\alpha,\epsilon}[m] =  
-\left[
-\begin{array}{c}
-S[m]\\
-\epsilon I\\
-\alpha T_{\epsilon}[m]
-\end{array}
-\right]
-, \, b = \left[
-\begin{array}{c}
-d\\
-0\\
-0
-\end{array}
-\right]
-$$
-Then
-$$
-J_{\alpha,\epsilon}[m,w_*,u] = 0.5\|A_{\alpha,\epsilon}[m]u-b\|^2
-$$
-
-Both Newton's method with the Kaufman Hessian approximation (described in the VCL notebook) and LBFGS or similar first-order methods apply to minimization of $\tilde{J}_{\alpha,\epsilon}[m]$.
-As shown in the VCL notebook, application of these algorithms to a VPM reduction like $\tilde{J}_{\alpha,\epsilon}$ requires access to two linear operator valued functions of $m$, namely $A_{\alpha,\epsilon}[m]$ and the (partial) derivative $D_m(A_{\alpha,\epsilon}[m]u)$.
-The former is the main actor in computation of the objective via solution of the normal equations.
-The latter appears in the VPM gradient formula:
-$$
-\nabla \tilde{J}_{\alpha,\epsilon}[m] = (D_m A_{\alpha,\epsilon}[m]u)|_{u=\tilde{u}_{\alpha,\epsilon}[m]}^T(A_{\alpha,\epsilon}[m]\tilde{u}_{\alpha,\epsilon}[m] -b)
-$$
-Note that $D_m A_{\alpha,\epsilon}[m]u$ is a map from the model parameter space to the product range space of $A$, and the transpose is a map in the other direction.
+Note that the notation is consistent, that is, reduces to $J_{0,\sigma}$ as defined above for $\alpha=0$.
 
@@
 \[
-  DB_1(B_0[m])^TDB_2(B_1\circ  B_0[m],u)^TT[m]u =
+  (DB_1(B_0[m])^TDB_2(B_1\circ  B_0[m],u)^TT[m]u )(\bx_r,\cdot,\bx_s)
 \]
+\[
+  = -\int dt\,t(T[m]u)(\bx_r,t;\bx_s)u(\bx_r,t;\bx_s)
+  \|u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\|^{-3}u_{0,\epsilon}(\bx_r,\cdot;\bx_s)
+\]
+In view of the definition of $T[m]$, this is
 \begin{equation}
   \label{eqn:compgrad1}
-  -\int dt\,t(T[m]u)(\bx_r,t;\bx_s)u(\bx_r,t;\bx_s)
-  \|u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\|^{-3}u_{0,\epsilon}(\bx_r,\cdot;\bx_s)
+  = -\||T[m]u(\bx_r,\cdot;\bx_s)\|^2
+  \|u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\|^{-2}u_{0,\epsilon}(\bx_r,\cdot;\bx_s)
 \end{equation}
+It remains to compute $ DB_0[m]\delta m,$:
 $$
-u_{0,\epsilon}[m] = (S[m]^TS[m] + \epsilon^2 I)^{-1}S[m]^Td
+B_0[m] = u_{0,\epsilon}[m] = (S[m]^TS[m] + \epsilon^2 I)^{-1}S[m]^Td
 $$
@@
 $$
-r_0 = d -C[U,D,D,F[m]w_*]u_{0,\epsilon}[m], \, u_0 = u_{0,\epsilon}[m]
+r_0 = d -C[U,D,D,F[m]w_*]u_{0,\epsilon}[m], \, u_0 = u_{0,\epsilon}[m].
 $$
-From the commutativity relation,
-$$
+From the commutativity relation \ref{eqn:convcomm},
+\begin{equation}
+  \label{eqn:compgrad2}
 D_m(C[U,D,D,F[m]w_*]u_0)\delta m = D_m(C[D,D,U,u_0]F[m]w_*)\delta m
 =C[D,D,U,u_0]D_mF[m]\delta m.
-$$
+\end{equation}
 Similarly, use the identity \ref{eqn:convcommtransp} to write
@@
 so
+\begin{equation} \label{eqn:compgrad3}
+  D_m(C[U,D,D,F[m]w_*]^Tr_0)\delta m
+  =C[U,\check{D},\check{D},\check{r_0}]^T(D_mF[m]w_* \delta m)\check= C[U,D,D,F[m]w_*\delta m]^Tr_0
+\end{equation}
+
+Putting \ref{eqn:compgrad}, \ref{eqn:compgrad1}, \ref{eqn:compgrad2},
+and \ref{eqn:compgrad3} together,
+\[
+  \frac{\alpha^2}{2} D_m(\|T[m]u\|_U^2)\delta m =
+=-\alpha^2 \langle C[U,D,D,F[m]w_*]^TC[D,D,U,u_0]D_mF[m]\delta m +C[U,D,D, D_mF[m]w_*
+\delta m]^Tr_0, 
+\]
+\[
+ \||T[m]u(\bx_r,\cdot;\bx_s)\|^2
+  \|u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\|^{-2}u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\rangle_U
+  DB_1(B_0[m])^TDB_2(B_1\circ
+  B_0[m],u)^TT[m]u\rangle_U
+\]
+
+Like all penalty functions, this one is best viewed as a least-squares objective of a map with values in a product space: define
 $$
-D_m(C[U,D,D,F[m]w_*]^Tr_0)\delta m=
-C[U,\check{D},\check{D},\check{r_0}]^T(D_mF[m]w_* \delta m)\check.
+A_{\alpha,\epsilon}[m] =  
+\left[
+\begin{array}{c}
+S[m]\\
+\epsilon I\\
+\alpha T_{\epsilon}[m]
+\end{array}
+\right]
+, \, b = \left[
+\begin{array}{c}
+d\\
+0\\
+0
+\end{array}
+\right]
+$$
+Then
+$$
+J_{\alpha,\epsilon}[m,w_*,u] = 0.5\|A_{\alpha,\epsilon}[m]u-b\|^2
 $$
 
-Putting these calculations together,
-
-\append{$\alpha \rightarrow 0$ scaled limit of penalty function}
-Define
-\begin{equation}
-  \label{eqn:eq0}
-  \tJa = \min_u J_{\alpha}(u) = J_{\alpha}(\ua)
-\end{equation}
-Since $S$ is positive definite, the minimum is well-defined for any $\alpha \ge 0$, and the minimizer $\ua \in U$ satisfies the normal equation.
-
-The claim to be established is that
-\begin{equation}
-  \label{eqn:eq4}
-  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tJa-\tJz).= \frac{1}{2}\|A\uz\|_m^2
-\end{equation}
-
-To see this, use the normal equation to write
-\[
-  \ua = (S^TS + \alpha^2 A^TA)^{-1}S^Td = (S^TS + \alpha^2 A^TA)^{-1}S^TS \uz
-\]
-\begin{equation}
-  \label{eqn:eq2}
-  = \uz - \alpha^2 (S^TS + \alpha^2 A^TA)^{-1}A^TA\uz = \uz-\alpha^2 \va
-\end{equation}
-Note that $\va = (S^TS + \alpha^2 A^TA)^{-1}A^TA\uz$ is uniformly bounded in $\alpha \ge 0$.
-
-Stuff the RHS of equation \ref{eqn:eq2} into the definition \ref{eqn:eq1} to obtain
-\begin{equation}
-  \label{eqn:eq3}
-  \tJa = \frac{1}{2}(\|S\uz-d\|_d^2 - 2 \alpha^2\langle S\uz-d, S\va\rangle_d + \alpha^2\|A\uz\|_m^2 + O(\alpha^4))
-\end{equation}
-The second term on the RHS of equation \ref{eqn:eq3} vanishes thanks to the normal equation, and the first term is precisely $\tJz$. The conclusion \ref{eqn:eq4} follows.
-
-\bibliographystyle{seg}
-\bibliography{../../bib/masterref}
+Both Newton's method with the Kaufman Hessian approximation (described in the VCL notebook) and LBFGS or similar first-order methods apply to minimization of $\tilde{J}_{\alpha,\epsilon}[m]$.
+As shown in the VCL notebook, application of these algorithms to a VPM reduction like $\tilde{J}_{\alpha,\epsilon}$ requires access to two linear operator valued functions of $m$, namely $A_{\alpha,\epsilon}[m]$ and the (partial) derivative $D_m(A_{\alpha,\epsilon}[m]u)$.
+The former is the main actor in computation of the objective via solution of the normal equations.
+The latter appears in the VPM gradient formula:
+$$
+\nabla \tilde{J}_{\alpha,\epsilon}[m] = (D_m A_{\alpha,\epsilon}[m]u)|_{u=\tilde{u}_{\alpha,\epsilon}[m]}^T(A_{\alpha,\epsilon}[m]\tilde{u}_{\alpha,\epsilon}[m] -b)
+$$
+Note that $D_m A_{\alpha,\epsilon}[m]u$ is a map from the model parameter space to the product range space of $A$, and the transpose is a map in the other direction.

### 2023-10-25 repro/papers/awi/old.tex
switch to different approach

\title{Notes on AWI}
\author{William W. Symes}

\begin{abstract}
Adaptive Waveform Inversion is the technique closest to our approach in the single-trace transmission project that applies to ``real'' data and has a literature. In fact, it inspired parts of our project. Here are some ideas on how to use AWI as the next step, and what we might accomplish. 
\end{abstract}

\section{Background}
Like our project, AWI treats each data trace (i.e. choice of source and receiver) independently. It also uses the same annihilator,  that is, the goal is to explain the data with a source wavelet as compact as possible and centered at zero, and uses multiplication by time as the penalty operator. It differs in assuming multidimensional data (many traces) and multidimensional models (slowness is a 2D or 3D field). In the form presented by Mike Warner and his group, it assumes a known (``true'') source wavelet. Also, their objective function is just the norm-squared of the annihilator output, normalized by the input wavelet norm-squared, as explained in the AWI notebook. In fact, Warner's objective function is the $\alpha \rightarrow 0$ limit of a scaled penalty function, see below for explanation.

I've sketched a proof in the AWI notebook that AWI ``works'', that is, has a ``good'' slowness model as its only stationary point (or points - uniqueness wasn't part of the story), provided that the data samples transmitted waves through a smooth material model without caustics.. Specifically, any stationary point approximates a stationary point of the travel time tomography objective. Warner has been quiet about the limitations of this result. I am pretty sure that AWI is essentially equivalent to FWI for reflection data, no special good properties in that case.

The successes I know about for AWI are consistent with this picture. In particular, Warner has used AWI to invert diving wave data. Diving waves are transmitted waves refracted back to the surface, and usually appear at longer offsets because the p-wave velocity in the earth increases with depth. Inversion of diving waves is also the``conventional'' use of FWI, so AWI has some real utility.

Diving waves have limited penetration depth - the larger the offset, then greater the depth. This fact has driven the development of super long cables, node surveys, and the like. It also explains the way in which Warner \& Co. use AWI. Starting with the Chevron blind test data in 2014, through a number of published 3D field data examples, Warner typically runs AWI first, pinning down the depth range probed by diving waves. Then he uses FWI (or, more recently RFWI, a variant) to further refine the model at greater depths. So far as I know he has never discussed the reasons for this work flow, just presented examples.

The overall picture is that AWI is a useful and usable velocity estimation technique that applies to ``real'' data, is as close as possible to our prior work, and presents the quickest avenue I can see to produce something both ``practical'' and somewhat familiar. Your sponsors should be quite interested, and there should be several publishable papers in it, eventually.

\section{Goals}
\begin{itemize}
\item State and implement penalty form of AWI;
\item careful write-up of the analysis in the AWI notebook, also for the penalty form;
\item illustrate the properties of AWI: improvement over FWI for single-arrival transmission data including diving waves, failure to improve over FWI for strongly triplicated transmission data and reflection data; This requires
  \begin{itemize}
  \item implement FWI
  \item create several good examples, of both FWI success and failure
  \item implement AWI, and apply to various examples to illustrate the features listed above.
  \end{itemize}
\item for ``good'' case, discrepancy-based control of penalty parameter
\item explore small-$\alpha$ vs. discrepancy-controlled $\alpha$: resolution of slowness, sensitivity to noise;
\item also for ``good'' case, implementation and examples of noise estimation algorithm
\end{itemize}
I propose that we do all of this using my acoustic 2D code. This means that the noise estimation algorithm won't make sense for field data, since the model will be totally wrong hence the noise level too large. But we will be able to produce much more interesting synthetic examples.

\append{$\alpha \rightarrow 0$ scaled limit of penalty function}
This appendix sketches a justification for the relation between the penalty function and original formulations of AWI asserted in the AWI notebook. I worked this out for the review of a paper about another extension inversion method, and can't imagine that it's not somewhere, in some form - it's too simple not to be - but I don't know where (else), so here it is.

Start with the penalty function
\begin{equation}
  \label{eqn:eq1}
  J_{\alpha}(u) = \frac{1}{2}(\|Su-d\|_d + \alpha^2\|Au\|_m^2).
\end{equation}
Here $S$ is a bounded linear map with (Hilbert space) domain $U$ (``model space'') with inner product $\langle \cdot, \cdot \rangle_m$ and range $D$ with inner product $\langle \cdot, \cdot \rangle_d$. Assume that $S$ is positive definite, that is, there exists $\epsilon >0$ so that $\|Su\|_d \ge \epsilon\|u\|_m$. The ``annihilator'' $A$ is a bounded linear map on $U$.

In the notation of the AWI notebook, $S$ is the regularized modeling operator $(\tilde{S}(m,w_*),\epsilon I)^T$, $u$ is the adaptive kernel $\tilde{u}$, $d$ is the augmented data $(d,0)^T$,  and $A$ is the multiply-by-$t$ operator $T$. The limit relation in question does not involve the arguments $m, w_*$ of $\tilde{S}$, so I've suppressed them. The relation in question pertains only to the linear least-squares function $J_{\alpha}$ defined in equation \ref{eqn:eq1}, not to the nonlinear dependence of $\tilde{S}$ on $(m,w_*)$.

Define
\begin{equation}
  \label{eqn:eq0}
  \tJa = \min_u J_{\alpha}(u) = J_{\alpha}(\ua)
\end{equation}
Since $S$ is positive definite, the minimum is well-defined for any $\alpha \ge 0$, and the minimizer $\ua \in U$ satisfies the normal equation.

The claim to be established is that
\begin{equation}
  \label{eqn:eq4}
  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tJa-\tJz).= \frac{1}{2}\|A\uz\|^2
\end{equation}

To see this, use the normal equation to write
\[
  \ua = (S^TS + \alpha^2 A^TA)^{-1}S^Td = (S^TS + \alpha^2 A^TA)^{-1}S^TS \uz
\]
\begin{equation}
  \label{eqn:eq2}
  = \uz - \alpha^2 (S^TS + \alpha^2 A^TA)^{-1}A^TA\uz = \uz-\alpha^2 \va
\end{equation}
Note that $\va = (S^TS + \alpha^2 A^TA)^{-1}A^TA\uz$ is uniformly bounded in $\alpha \ge 0$.

Stuff the RHS of equation \ref{eqn:eq2} into the definition \ref{eqn:eq1} to obtain
\begin{equation}
  \label{eqn:eq3}
  \tJa = \frac{1}{2}(\|S\uz-d\|_d^2 - 2 \alpha^2\langle S\uz-d, S\va\rangle_d + \alpha^2\|A\uz\|_m^2 + O(\alpha^4))
\end{equation}
The second term on the RHS of equation \ref{eqn:eq3} vanishes thanks to the normal equation, and the first term is precisely $\tJz$. The conclusion \ref{eqn:eq4} follows.

Note that this conclusion is not quite what is stated in the notebook (which I will modify shortly!), since the $\alpha=0$ limit of the VPM objective $\tJa$ is non-zero and has to be subtracted. That's not a big deal: in the notation of the notebook,
\[
  \tJz = \frac{1}{2}(\|\tilde{S}[m,w_*]u_{\epsilon} -d\|^2 + \epsilon^2 \|u_{\epsilon}\|^2),
\]
that is, the minimum value of the regularized least squares function.

I included the model space norm $\|\cdot\|_m$ in this derivation because it is not just the $L^2$ norm - it's how the scaling by the norm of $u_{\epsilon}$ naturally enters.
%\bibliographystyle{seg}
%\bibliography{../../bib/masterref}

### 2023-10-25 repro/papers/awigrad/paper_old.tex
switch to different approach

\title{Computation of the Reduced AWI Penalty Function and its Gradient}
\author{William W. Symes}

\begin{abstract}
goes here.
\end{abstract}

\section{Definitions}
Hilbert space $X$ has inner product $\langle \cdot,\cdot \rangle_X$ and norm $\|\cdot\|_X$.

$Q \subset \bR^3 \times \bR^3$ = finite ``acquisition'' set of source-receiver location pairs $(\bx_r,\bx_s)$, assumed disjoint from the diagonal (sources and receivers non-coincident). $Q_s$ = projection onto second factor (source locations occurring in $Q$).

$P = \bR^{|Q|}$ = real functions on $Q$, with Euclidean norm (no attempt to scale for geometry of acquisition).

$[0,t_d]$ = acquisition time interval

$[-t_U,t_U]$ = adaptive filter time interval

$D$ = data Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([0,t_D])$ (again, no attempt to scale for geometry of acquisition).

$U$ = adaptive filter Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([-t_U,t_U])$

$w_* \in C_0^{\infty}(\bR)$ = known point source wavelet

$M = $ model set, bounded open subset of $ C^{\infty}(\bR^3) \times C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3) \times L^{\infty}(\bR^3)$.

\begin{theorem}
  \label{thm:eu}
For $(\log \kappa, \log \beta) \in M$, shot location ${\bf x}_s\in Q_s$, there exist distributions (pressure and velocity fields) $p(\cdot,\cdot,t;\bx_s)$, ${\bf v}(\cdot,\cdot;\bx_s)$ satisfying
\begin{itemize}
\item
  \begin{eqnarray}
    \label{eqn:awe}
    \frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv +
                                        w(t;\bx_s) \delta(\bx-\bx_s); \\
    \frac{\partial \bv}{\partial t} & = & - \beta \nabla p; \\
    p, \bv & = & 0 \mbox{ for }  t \ll 0.
  \end{eqnarray}
\item
  $p, \bv$ are smooth in the punctured space-time $\bR^4 \setminus \{(\bx_s,t): t \in \bR\}$.
\end{itemize}
\end{theorem}
\begin{proof}
  See Appendix A.
\end{proof}

Define $F: M \rightarrow D$ by
\[
  F[m](\bx_r,t;\bx_s) = p(\bx_r,t;\bx_s), \, (\bx_r,\bx_s) \in Q, \, t \in [0,t_D],
\]
for the pressure field $p$ solving \ref{eqn:awe}.

\begin{theorem}
  \label{thm:mbd}
  $\|F[m](\bx_r,\cdot;\bx_s)\|_{L^2([0,t_D])}$ is uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
\end{theorem}
\begin{proof}
  See Appendix A.
\end{proof}

$S: M \rightarrow {\cal B}(U,D)$ = trace-by-trace convolution with predicted data: $S[m]u = F[m] * u$. Due to smoothness of $F$, $S \in C^1(M,{\cal B}(U,D))$. Operation is trace-by-trace, that is block-diagonal: $S[m] = \bigoplus_{(\bx_r,\bx_s) \in Q}S[m]_{\bx_r,\bx_s}$. The Modeling Bound Assumption implies that $\|S[m](\bx_r,\cdot;\bx_s)\|_{{\cal B}(L^2([-t_U,t_U]),L^2([0,t_D]))}$ is also uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.

No-zero-traces Assumption: there exists $C>0$ so that for all $(\bx_r,\bx_s) \in Q$,
$\|d(\bx_r,\cdot;\bx_s)\|_{L^2([0,t_D])} \ge C.$

Observable Data Assumption: there exists $C>0$ so that for all $m \in M, (\bx_r,\bx_s) \in Q$,
\begin{equation}
  \label{eqn:obsdata}
  \|S[m]_{\bx_r,\bx_s}^Td(\bx_r,\cdot;\bx_s)\|_{L^2[-t_U,t_U]} \ge C \|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_D]}
\end{equation}.
That is, the projection of each trace onto the range of convolution of the corresponding predicted data trace is coercive.

$\sigma > 0$ = Tihonov regularization weight

$\alpha \ge 0$ = penalty weight

Definition of AWI penalty objective divides into two parts:

Part 1: Unpenalized adaptive filter $u_{0,\sigma}[m,d]$ solving regularized least squares data fitting problem:  given $m \in M$, minimizes
\[
 J_{0,\sigma}[u,m,d] = \frac{1}{2}(\|S[m]u - d\|_D^2 + \sigma^2 \|u\|^2_U).
\]
Since $\sigma > 0$, normal operator is uniformly bounded below and of class $C^1$ in $M$ and $D$. So unique solution exists, and $u_{0,\sigma} \in C^1(M \times D, U)$.

Also, the normal operator is block-diagonal, so each trace solves a least-squares problem, and
\[
  (S[m]_{\bx_r,\bx_s}^TS[m]_{\bx_r,\bx_s} + \sigma^2 I)u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)= S[m]_{\bx_r,\bx_s}^Td(\bx_r,\cdot;\bx_s)
\]
According to the Modeling Bound Assumption, observable data assumption, the left-hand side is bounded by a multiple of $\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]}$, uniformly in $m \in M, (\bx_r,\bx_s) \in Q$. According to the Observable Data Assumption, the right-hand side is bounded below by a multiple of $\|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_D]}$, uniform in the same sense. Conclude that there exists $C>0$ so that
\begin{equation}
  \label{eqn:u0lower}
  \|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]} \ge C \|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_D]}
\end{equation}
uniformly in $m \in M, (\bx_r,\bx_s) \in Q$.

Part 2:
Preconditioned penalty operator: block diagonal, 
\begin{equation}
  \label{eqn:ppf}
  T_{\sigma}[m]u (\bx_r,t;\bx_s) = \frac{t u(\bx_r,t;\bx_s)}{\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]} }
\end{equation}
Well-defined and $C^1$ as function of $m$ due to the bound \ref{eqn:u0lower} and the the No-zero-trace assumption.

\section{Objectives}
Define for $\alpha \ge 0$
\[
 J_{\alpha,\sigma}[u,m,d] = \frac{1}{2}(\|S[m]u - d\|_D^2 + \alpha^2\|T_\sigma[m]u\|_U^2 + \sigma^2 \|u\|^2_U).
\]
Note that the notation is consistent, that is, reduces to $J_{0,\sigma}$ as defined above for $\alpha=0$.

The variable projection reduction of $J_{\alpha,\sigma}$ is
\begin{equation}
  \label{eqn:jtilde}
 \tilde{J}_{\alpha,\sigma}[m,d] = \frac{1}{2}(\|S[m]u_{\alpha,\sigma}[m,d] - d\|_D^2 + \alpha^2\|T_\sigma[m]u_{\alpha,\sigma}[m,d]\|_U^2 + \sigma^2 \|u_{\alpha,\sigma}\|^2_U).
\end{equation}
in which $u_{\alpha,\sigma}[m,d]$ is the minimizer of $J_{\alpha\sigma}[u,m,d]$ over $u \in U$. The fundamental result of variable projection theory \cite[]{GolubPeyreyra:73} applies under the conditions presented here: the stationary points of $J_{\alpha,\sigma}$ and $\tilde{J}_{\alpha,\sigma}$ are in bijective correspondence.

Computing $\tilde{J}_{\alpha,\sigma}$ takes place in two stages, corresponding to the two-stage construction of $J_{\alpha,\sigma}$:
\begin{itemize}
\item[1. ] minimize $J_{0,\sigma}$ to compute $u_{0,\sigma}$;
\item[2. ] use $u_{0,\sigma}$ to compute $u \mapsto T_{\sigma}[m]u$, and compute $u_{\alpha,\sigma}$ by minimizing $J_{\alpha,\sigma}$.
\end{itemize}
Therefore evaluating $\tilde{J}_{\alpha,\sigma}$ involves solution of two least-squares problems, in sequence.

\section{Derivatives}
This section presents the derivative of $\tilde{J}_{\alpha,\sigma}$, expressed in terms of
\begin{itemize}
  \item trace-by-trace convolution and cross-correlation operators,
  \item the derivative $D_mF[m]$ and its adjoint, and
  \item trace-by-trace scaling
\end{itemize}
These the first two items are common components of typical FWI implementations. The
third is necessary to express the AWI penalty operator.

As pointed out by \cite{GolubPeyreyra:73}, the directional derivative $D_m\tilde{J}_{\alpha,\sigma}[m,d]\delta m$ is the same as the partial directional derivative $D_mJ[u,m,d]\delta m$ with $u = u_{\alpha,\sigma}[m,d]$. Since 
\begin{equation}
D_mJ[u,m,d]\delta m = D_mS[m]^T (S[m]u-d) + \alpha^2 D_m(T_{\sigma}[m]u)^TT_{\sigma}[m]u,
\label{eqn:basederiv}
\end{equation}
the key calculations are $D_mS^T$ and $D_m T^T$.

Note that the regularization term does not appear in the VPM gradient expression, since it does not depend explicitly on $m$.

\subsection{Convolution Revisited}

To keep track of the data flow in subsequent calculations, it is convenient to introduce a notation for the operator of convolution by $u$. This notation must include the time ranges for input and output traces, as these define the index limits in the sums that implement the convolution operator. Looked at another way, these ranges define the domain and range of the operator, and that is how the definition in {\em segyvc.ConvolutionOperator} is arranged: the arguments to the class constructor are the domain and range (as {\em segyvc.Space}s, type-checked) and the filename for the convolution kernel. Domain, range, and kernel must share spatial geometry (source and receiver positions, in particular number of traces, and time step; these conditions are checked). In this section, the trace-by-trace convolution operator with compatible trace space domain $X$ and range $Y$, and kernel $z \in Z$ is denoted $C[X,Y,Z,z]$. 
The kernel space $Z$ is also assumed compatible with $X$ and $Y$.

The domain, range, and kernel spaces considered here are all coordinate spaces, defined by finite time index ranges (or, in the continuum case, bounded time ranges). For such a coordinate space $X$, denote by $I_X$ the index range (for the discrete case) associated with $X$, and $\Pi_X$ the orthogonal projection of the ambient space $l^2$ (or $L^2({\bf R})$ onto $X$. That is, in the discrete case, for $i \in {\bf Z}, (\bx_r,\bx_s) \in Q$,
\[
  \Pi_Xu(\bx_r,\bx_s)_i  =
  \left\{
    \begin{array}{c}
      u(\bx_r,\bx_s)_i, i \in I_X, \\
      0, \mbox{ else.}
    \end{array}
  \right.
\]  
and similarly for the continuum case. The transpose projection $\Pi_X^T$ injects the array of values within the range into $l^2$, with zeros outside. So $\Pi_X^T\Pi_X$ is the projection onto the subspace of $l^2$ (or $L^2({\bf R})$) corresponding to $X$, i.e. the range of $\Pi_X^T$, and $\Pi_X \Pi_X^T = I_X$.

Define convolution $f*g$ by the usual sum or integral, interpreted as block-diagonal (trace-by-trace) for  $f$ and $g$ of bounded support. With these conventions,
\begin{equation}
  \label{eqn:convdef}
C[X,Y,Z,z]x = \Pi_Y(\Pi_Z^T z * \Pi_X^T x).
\end{equation}
The class {\em segyvc.ConvolutionOperator} implements precisely the discrete version of this definition.

Both discrete and continuous convolution are commutative. The same is true of the discrete, finite version of convolution implemented in {\em segyvc.ConvolutionOperator}, and can be expressed as the follow obvious consequence of the definition: if $X, Y, Z$ are compatible trace spaces, and $z \in Z, x \in X$, then
\begin{equation}
C[X,Y,Z,z]x = C[Z,Y,X,x]z
\label{eqn:convcomm}
\end{equation}
The transpose of convolution is expressed via the time reversal operator $f \mapsto \check{f}\, \check{f}_i = f_{-i}$:  
$$
\langle f*g,h \rangle = \langle g, \check{f}*h \rangle.
$$ 
The right-hand side defines the cross-correlation of $f$ and $h$. From the definition of $C$,
$$
\langle C[X,Y,Z,z]x, y\rangle_Y = \langle \Pi_Y(\Pi_Z^T z * \Pi_X^T x), y \rangle_Y
$$
$$
= \langle \Pi_Z^T z * \Pi_X^T x, \Pi_y^T y \rangle = \langle \Pi_X^T x, \check{(\Pi_Z^T z)}* \Pi_Y^T y \rangle
$$
Denote by $\check{Z}$ the coordinate space produced by time reversal applied to $Z$, that is, $I_{\check{Z}}$ consists of the negatives of members of $I_Z$. Then $\check{(\Pi_Z^T z)}=\Pi_{\check{Z}}\check{z}$, so the above is
$$
=\langle \Pi_X^T x, \Pi_{\check{Z}}^T \check{z} * \Pi_Y^T y \rangle =\langle x, \Pi_X(\Pi_{\check{Z}}^T \check{z} * \Pi_Y^T y) \rangle_X
$$
$$
= \langle x, C[Y,X,\check{Z},\check{z}]y \rangle_X.
$$
That is, $C[X,Y,Z,z]^T = C[Y,X,\check{Z},\check{z}]$.

Combining the commutativity relation $\ref{eqn:convcomm}$ with this identity yields the useful identity:
\begin{equation}
C[X,Y,Z,z]^Ty = C[Y,X,\check{Z},\check{z}]y =
C[\check{Z},X,Y,y]\check{z} =
C[X,\check{Z},\check{Y},\check{y}]^T\check{z}.
\label{eqn:convcommtransp}
\end{equation}

\subsection{Back to AWI penalty gradient}

The SEGY trace spaces involved in the AWI calculation are the data space $D$ and the adaptive filter space $U$. Minimization of $J_{\alpha,\sigma}$ amounts to solution of a least squares problem with operator $C[U,D,D,F[m]w]$, that is, convolution with the predicted data:
$$
S[m]u = C[U,D,D,F[m]w_*]u = C[D,D,U,u]F[m]w_*
$$
The second expression, equivalent by the commutativity relation \ref{eqn:convcomm} to the first, is more convenient for computing the derivative with respect to $m$ of $S[m]u$:
$$
D_m(S[m]u) \delta m = C[D,D,U,u](D_mF[m]\delta m)
$$
whence ($e_0 = S[m]u_{\alpha,\sigma}[m]-d \in D$) the first summand in the gradient of $\tilde{J}_{\alpha,\sigma}$ is
$$
D_m(S[m]u)^T e_0 = D_mF[m]^TC[D,D,U,u]^Te_0.
$$
%The derivative $D_mF[m]$ and its adjoint are defined in the *vcl.Function* subclass *asg.fsbop*, and the convolution operator $C$ and its transpose in *segyvc.ConvolutionOperator*.

Recall the set $Q$ of source-receiver pairs active in the data
$D$, and scale factor space $P = \{f:Q \rightarrow \bR\}$. Write
\[
  T_{\sigma}[m]u = B_2 \circ B_1 \circ B_0[m],
\]
where 
\[
B_2: P \times U \rightarrow U, \, B_2[f,u](bx_r,t;\bx_s) = f(\bx_r,\bx_s) tu(\bx_r,t;\bx_s)
\]
\[
B_1: U \rightarrow P, \,B_1[u](\bx_r,t,\bx_s) = \|u(\bx_r,\cdot; \bx_s)\|^{-1}
\]
\[
B_0: M \rightarrow U, \, B_0[m] = u_{0,\sigma}[m]
\]
Then
\[
 \frac{\alpha^2}{2} D_m(\|T_{\sigma}[m]u\|_U^2)\delta m = \alpha^2\langle D_m(
 T_{\sigma}[m]u)\delta m, T_{\sigma}[m]u \rangle_U
\]
\[
  =\alpha^2 \langle D_fB_2(B_1\circ B_0[m],u)D_uB_1(B_0[m]) D_mB_0[m]\delta m,
  T_{\sigma}[m]u\rangle_U
\]
\begin{equation}
  \label{eqn:compgrad}
  =\alpha^2 \langle D_mB_0[m]\delta m, D_uB_1(B_0[m])^TD_fB_2(B_1\circ
  B_0[m],u)^TT_{\sigma}[m]u\rangle_U
\end{equation}
Next display the derivative and adjoint derivative of $B_1$ and $B_2$:
$$
(D_fB_2[f,u]\delta f)(\bx_r,t;\bx_s) = tu(\bx_r,t;\bx_s)\delta f(\bx_r,\bx_s)
$$
$$
D_f(B_2[f,u])^Te_2(\bx_r,\bx_s) = \int dt\,te_2(\bx_r,t;\bx_s)u(\bx_r,t;\bx_S);
$$
$$
D_u B_1[u]\delta u(\bx_r,\bx_s) = -\|u(\bx_r,\cdot;\bx_s)\|^{-3}\langle u(bx_r,\cdot;\bx_s), \delta u(\bx_r,\cdot;\bx_s) \rangle
$$
$$
(D_u B_1[u])^Tf(\bx_r,t;\bx_s) =
-f(\bx_r,\bx_s)\|u(\bx_r,\cdot;\bx_s\|^{-3}u(\bx_r,t;\bx_s)
$$
With these expressions, compose various components appearing on the
right hand side of equation \ref{eqn:compgrad}:
\[
  B_1\circ B_0[m] (\bx_r,\bx_s) =  \|\tilde{u}_{0,\sigma}(\bx_r,\cdot; \bx_s)\|^{-1},
\]
\[
 D_fB_2(B_1\circ B_0[m],u)^Te_2 = \int dt\,te_2(\bx_r,t;\bx_s)u(\bx_r,t;\bx_s),
\]
(Note that the result is independent of the first argument.)
\[
D_uB_1(B_0[m])^Tf(\bx_r,t:\bx_s)=
-f(\bx_r,\bx_s)\|u_{0,\sigma}(\bx_r,\cdot;\bx_s)\|^{-3}u_{0,\sigma}(\bx_r,\cdot;\bx_s),
\]
and
\[
  (DB_1(B_0[m])^TDB_2(B_1\circ  B_0[m],u)^TT_{\sigma}[m]u )(\bx_r,\cdot,\bx_s)
\]
\[
  = -\int dt\,t(T_{\sigma}[m]u)(\bx_r,t;\bx_s)u(\bx_r,t;\bx_s)
  \|u_{0,\sigma}(\bx_r,\cdot;\bx_s)\|^{-3}u_{0,\sigma}(\bx_r,\cdot;\bx_s)
\]
In view of the definition of $T_{\sigma}[m]$, this is
\begin{equation}
  \label{eqn:compgrad1}
  = -\||T_{\sigma}[m]u(\bx_r,\cdot;\bx_s)\|^2
  \|u_{0,\sigma}(\bx_r,\cdot;\bx_s)\|^{-2}u_{0,\sigma}(\bx_r,\cdot;\bx_s)
\end{equation}
It remains to compute $ DB_0[m]\delta m,$:
$$
B_0[m] = u_{0,\sigma}[m] = (S[m]^TS[m] + \sigma^2 I)^{-1}S[m]^Td
$$
\begin{equation}
  \label{eqn:compgrad4}
  DB_0[m]\delta m = (S[m]^TS[m] + \sigma^2 I)^{-1}(D_m(S[m]^T r_0)\delta m - S[m]^TD_m(S[m]u_0)\delta m)
\end{equation}
%$$
%= (C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*] + \sigma^2 I)^{-1}C[U,D,D,F[m]w_*]^Td.
%$$
%So
%$$
%D_mB_0[m]\delta m = (C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*] + \sigma^2 I)^{-1}
%$$
%$$
%\times ((D_mC[U,D,D,F[m]w_*]\delta m)^Td -(D_m(C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*])\delta m)u_{0,\sigma}[m])
%$$
%$$
%=(C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*] + \sigma^2 I)^{-1}
%$$
%$$
%\times
%(D_m(C[U,D,D,F[m]w_*]^Tr_0)\delta m  - C[U,D,D,F[m]w_*]^TD_m(C[U,D,D,F[m]w_*]u_0)\delta m)
%$$ 
where the derivatives are to be evaluated before the insertion of the definitions 
$$
r_0 = d -S[m]u_{0,\sigma}[m], \, u_0 = u_{0,\sigma}[m].
$$
From the commutativity relation \ref{eqn:convcomm},
\begin{equation}
  \label{eqn:compgrad2}
D_m(S[m]u_0)\delta m = D_m(C[U,D,D,F[m]w_*]u_0)\delta m = D_m(C[D,D,U,u_0]F[m]w_*)\delta m
=C[D,D,U,u_0]D_mF[m]\delta m.
\end{equation}
Similarly, use the identity \ref{eqn:convcommtransp} to write
$$
S[m]^Tr_0 = C[U,D,D,F[m]w_*]^Tr_0 = C[U,\check{D},\check{D},\check{r_0}]^T({F[m]w_*})\check, 
$$
so
 \[
  D_m(S[m]^Tr_0)\delta m =  D_m(C[U,D,D,F[m]w_*]^Tr_0)\delta m
\]
\begin{equation} \label{eqn:compgrad3}
  =C[U,\check{D},\check{D},\check{r_0}]^T(D_mF[m]w_* \delta m)\check= C[U,D,D,F[m]w_*\delta m]^Tr_0
\end{equation}
From \ref{eqn:compgrad4}, \ref{eqn:compgrad2}, and \ref{eqn:compgrad3},
\[
  DB_0[m]\delta m = (S[m]^TS[m] + \sigma^2 I)^{-1}
\]  
\begin{equation}
  \label{eqn:compgrad5}
\times (C[U,D,D,DF[m]\delta m]^T(S[m]u_{0,\sigma}[m]-d) - C[D,D,U,u_{0,\sigma}[m]]D_mF[m]\delta m).
\end{equation}
Putting \ref{eqn:compgrad}, \ref{eqn:compgrad1}, \ref{eqn:compgrad2},
and \ref{eqn:compgrad3} together,
\[
  \frac{\alpha^2}{2} D_m(\|T_{\sigma}[m]u\|_U^2)\delta m 
  =\alpha^2 \sum_{(\bx_r,\bx_s) \in Q} \langle  (S[m]_{\bx_r,\bx_s}^TS[m]_{\bx_r,\bx_s} + \sigma^2 I)^{-1}
\]
\[
  \times (C[U,D,D,DF[m)\delta m]_{\bx_r,\bx_s}^T(S[m _{\bx_r,\bx_s}]u_{0,\sigma}[m](\bx_r,\cdot;\bx_S)-d(\bx_r,\cdot;\bx_s))
\]
\[
  - S[m]_{\bx_r,\bx_s} C[U,D,D, D_mF[m]\delta m]_{\bx_r,\bx_s} u_{0,\sigma} (\bx_r,\cdot;\bx_S),
\]
\[
  \||T_{\sigma}[m]u(\bx_r,\cdot;\bx_s)\|^2
  \|u_{0,\sigma}(\bx_r,\cdot;\bx_s)\|^{-2}u_{0,\sigma}(\bx_r,\cdot;\bx_s)\rangle
\]

\[
  = \alpha^2 \sum_{(\bx_r,\bx_s) \in Q} \||T_{\sigma}[m]u(\bx_r,\cdot;\bx_s)\|^2\|u_{0,\sigma}(\bx_r,\cdot;\bx_s)\|^{-2}
\]
\[
  \times \left( \langle S[m _{\bx_r,\bx_s}]u_{0,\sigma}[m](\bx_r,\cdot;\bx_S)-d(\bx_r,\cdot;\bx_s),\right.
\]
\[
  \left. C[U,D,D,DF[m)\delta m]_{\bx_r,\bx_s} (S[m]_{\bx_r,\bx_s}^TS[m]_{\bx_r,\bx_s} + \sigma^2 I)^{-1}u_{0,\sigma}[m](\bx_r,\cdot;\bx_S) \right. \rangle_D
\]
\[
  - \left. u_{0,\sigma}[m](\bx_r,\cdot;\bx_S),\right.
\]
\[
\left. (S[m]_{\bx_r,\bx_s}^TS[m]_{\bx_r,\bx_s} + \sigma^2 I)^{-1}u_{0,\sigma}[m](\bx_r,\cdot;\bx_S) \rangle \right)
\]

### 2023-10-25 pyrvl/awiop.py

@@ -1,73 +1,0 @@
-import vcl
-import segyvc
-import awi
-
-class awiop(vcl.LinearOperator):
-    '''
-    Penalty-AWI operator based on FD solution of wave equation and
-    scale-by-t penalty with optional preconditioning by zero-penalty
-    solution norms.
-
-    Parameters:
-    F (vcl.Function): simulation operator
-    d (vcl.Vector): trace data
-    awisp (vcl.Space): AWI adaptive kernel space
-    awisol (vcl.LSSolver): computes adaptive kernel
-    data (vcl.Vector): optional data trace, flags preconditioning
-
-    if data is provided, then it is used to compute the preconditioned
-    verstion of the AWI penalty operator. Note that in the application,
-    this vector should be the same as the data vector of the inverse
-    problem.
-    '''
-
-    def __init__(self, p, awisp, alpha, awisol=None, data=None):
-        # p is predicted data
-        self.p = p
-        # range is product (range of F x awisp)
-        self.rng = vcl.ProductSpace([p.space, awisp])
-        self.sp = awisp
-        self.alpha = alpha
-        self.sol = awisol
-        self.d = data
-        self.Sm0 = segyvc.ConvolutionOperator(self.sp, self.p.space, self.p.data)
-        self.alphaT = awi.awipensol(self.sp, self.alpha, solver=self.sol, d=data, p=self.p)
-
-    def getDomain(self):
-        return self.sp
-
-    def getRange(self):
-        return self.rng
-
-    def applyFwd(self, dx, dy):
-        try:
-            self.Sm0.applyFwd(dx,dy[0])
-            self.alphaT.applyFwd(dx,dy[1])
-        except Exception as ex:
-            print(ex)
-            raise Exception('called from awiop.applyFwd')
-
-    def applyAdj(self, dx, dy):
-        try:
-            cy = vcl.Vector(self.getDomain())
-            self.Sm0.applyAdj(dx[0],cy)
-            self.alphaT.applyAdj(dx[1],dy)
-            dy.linComb(1.0,cy)
-        except Exception as ex:
-            print(ex)
-            raise Exception('called from awiop.applyAdj')
-
-    def myNameIs(self):
-        print('AWI operator')
-        print('domain = ')
-        self.dom.myNameIs()
-        print('range = data space OPLUS domain')
-        if self.d is not None:
-            print('observed data:')
-            self.d.myNameIs()
-        print('predicted data:')
-        self.p.myNameIs()
[…]

### 2023-10-26 repro/papers/awigrad/paper.tex
revision

@@
 
-$M$ = model set, an open set in a Banach space also pre-Hilbert structure
-
-$Q$ = finite ``acquisition'' set of source-receiver location pairs
+$Q \subset \bR^3 \times \bR^3$ = finite ``acquisition'' set of source-receiver location pairs $(\bx_r,\bx_s)$, assumed disjoint from the diagonal (sources and receivers non-coincident).
+$Q_s$ = projection onto second factor (source locations occurring in $Q$).
 
@@
 
-$[-t_u,t_u]$ = adaptive filter time interval
+$[-t_U,t_U]$ = adaptive filter time interval
 
-$D$ = data Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([0,t_d])$ (again, no attempt to scale for geometry of acquisition).
+$D$ = data Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([0,t_D])$ (again, no attempt to scale for geometry of acquisition).
 
-$U$ = adaptive filter Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([-t_u,t_u])$
+$U$ = adaptive filter Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([-t_U,t_U])$
 
@@
 
-$F: M \rightarrow D$ = forward modeling operator, use of $w_*$ implicit. Since $w_*\in C_0^{\infty}(\bR)$  and distance from source points to receiver points is bounded below, $F \in C^1(M,D)$.
+$M = $ model set, bounded open subset of $ C^{\infty}(\bR^3) \times C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3) \times L^{\infty}(\bR^3)$.
 
-Modeling Bound Assumption: $\|F[m](\bx_r,\cdot;\bx_s)\|_{L^2([0,t_d])}$ is uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
+\begin{theorem}
+  \label{thm:eu}
+For $(\log \kappa, \log \beta) \in M$, shot location ${\bf x}_s\in Q_s$, there exist distributions (pressure and velocity fields) $p(\cdot,\cdot,t;\bx_s)$, ${\bf v}(\cdot,\cdot;\bx_s)$ satisfying
+\begin{itemize}
+\item
+  \begin{eqnarray}
+    \label{eqn:awe}
+    \frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv +
+                                        w(t;\bx_s) \delta(\bx-\bx_s); \\
+    \frac{\partial \bv}{\partial t} & = & - \beta \nabla p; \\
+    p, \bv & = & 0 \mbox{ for }  t \ll 0.
+  \end{eqnarray}
+\item
+  $p, \bv$ are smooth in the punctured space-time $\bR^4 \setminus \{(\bx_s,t): t \in \bR\}$.
+\end{itemize}
+\end{theorem}
+\begin{proof}
+  See Appendix A.
+\end{proof}
+
+Define $F: M \rightarrow D$ by
+\[
+  F[m](\bx_r,t;\bx_s) = p(\bx_r,t;\bx_s), \, (\bx_r,\bx_s) \in Q, \, t \in [0,t_D],
+\]
+for the pressure field $p$ solving \ref{eqn:awe}.
+
+\begin{theorem}
+  \label{thm:mbd}
+  $\|F[m](\bx_r,\cdot;\bx_s)\|_{L^2([0,t_D])}$ is uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
+\end{theorem}
+\begin{proof}
+  See Appendix A.
+\end{proof}
 
 $S: M \rightarrow {\cal B}(U,D)$ = trace-by-trace convolution with predicted data: $S[m]u = F[m] * u$.
+This notation for convolution in time hides a number of important details: none of the arguments are defined on the entire real line, so extension and truncation are involved in the definition.
+I make these details explicit later in the paper.
+
 Due to smoothness of $F$, $S \in C^1(M,{\cal B}(U,D))$.
 Operation is trace-by-trace, that is block-diagonal: $S[m] = \bigoplus_{(\bx_r,\bx_s) \in Q}S[m]_{\bx_r,\bx_s}$.
-The Modeling Bound Assumption implies that $\|S[m](\bx_r,\cdot;\bx_s)\|_{L^2([0,t_d])}$ is also uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
+Theorem \ref{thm:mbd} implies that $\|S[m](\bx_r,\cdot;\bx_s)\|_{{\cal B}(L^2([-t_U,t_U]),L^2([0,t_D]))}$ is also uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
 
 No-zero-traces Assumption: there exists $C>0$ so that for all $(\bx_r,\bx_s) \in Q$,
-$\|d(\bx_r,\cdot;\bx_s)\|_{L^2([0,t_d])} \ge C.$
+$\|d(\bx_r,\cdot;\bx_s)\|_{L^2([0,t_D])} \ge C.$
 
@@
   \label{eqn:obsdata}
-  \|S[m]_{\bx_r,\bx_s}^Td(\bx_r,\cdot;\bx_s)\|_{L^2[-t_u,t_u]} \ge C \|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_d]}
+  \|S[m]_{\bx_r,\bx_s}^Td(\bx_r,\cdot;\bx_s)\|_{L^2[-t_U,t_U]} \ge C \|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_D]}
 \end{equation}.
@@
 
-Also, the normal operator is block-diagonal, so each trace solves a least-squares problem, and
-\[
+Also, the normal operator is block-diagonal, so each trace solves a least-squares problem:
+\begin{equation}
+  \label{eqn:blocknormal}
   (S[m]_{\bx_r,\bx_s}^TS[m]_{\bx_r,\bx_s} + \sigma^2 I)u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)= S[m]_{\bx_r,\bx_s}^Td(\bx_r,\cdot;\bx_s)
-\]
-According to the Modeling Bound Assumption, observable data assumption, the left-hand side is bounded by a multiple of $\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_u,t_u]}$, uniformly in $m \in M, (\bx_r,\bx_s) \in Q$.
-According to the Observable Data Assumption, the right-hand side is bounded below by a multiple of $\|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_d]}$, uniform in the same sense.
+\end{equation}
+According to Theorem \ref{thm:mbd} and the Observable Data Assumption, the left-hand side is bounded by a multiple of $\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]}$, uniformly in $m \in M, (\bx_r,\bx_s) \in Q$.
+According to the Observable Data Assumption, the right-hand side is bounded below by a multiple of $\|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_D]}$, uniform in the same sense.
 Conclude that there exists $C>0$ so that
@@
   \label{eqn:u0lower}
-  \|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_u,t_u]} \ge C \|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_d]}
+  \|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]} \ge C \|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_D]}
 \end{equation}
@@
   \label{eqn:ppf}
-  T_{\sigma}[m]u (\bx_r,t;\bx_s) = \frac{u(\bx_r,t;\bx_s)}{\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_u,t_u]} }
+  T_{\sigma}[m]u (\bx_r,t;\bx_s) = \frac{t u(\bx_r,t;\bx_s)}{\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]} }
 \end{equation}
@@
 
+\section{Objectives}
 Define for $\alpha \ge 0$
@@
 
-\subsection{Computing $D_mA_{\alpha,\epsilon}[m]^T$}
-The main goal of this and the next two subsections is to express the
-operator in the subsection title in terms of
+The variable projection reduction of $J_{\alpha,\sigma}$ is
+\begin{equation}
+  \label{eqn:jtilde}
+ \tilde{J}_{\alpha,\sigma}[m,d] = \frac{1}{2}(\|S[m]u_{\alpha,\sigma}[m,d] - d\|_D^2 + \alpha^2\|T_\sigma[m]u_{\alpha,\sigma}[m,d]\|_U^2 + \sigma^2 \|u_{\alpha,\sigma}\|^2_U).
+\end{equation}
+in which $u_{\alpha,\sigma}[m,d]$ is the minimizer of $J_{\alpha\sigma}[u,m,d]$ over $u \in U$.
+The fundamental result of variable projection theory \cite[]{GolubPeyreyra:73} applies under the conditions presented here: the stationary points of $J_{\alpha,\sigma}$ and $\tilde{J}_{\alpha,\sigma}$ are in bijective correspondence.
+
+Computing $\tilde{J}_{\alpha,\sigma}$ takes place in two stages, corresponding to the two-stage construction of $J_{\alpha,\sigma}$:
+\begin{itemize}
+\item[1. ] minimize $J_{0,\sigma}$ to compute $u_{0,\sigma}$;
+\item[2. ] use $u_{0,\sigma}$ to compute $u \mapsto T_{\sigma}[m]u$, and compute $u_{\alpha,\sigma}$ by minimizing $J_{\alpha,\sigma}$.
+\end{itemize}
+Therefore evaluating $\tilde{J}_{\alpha,\sigma}$ involves solution of two least-squares problems, in sequence.
+
+\section{Derivatives}
+This section presents the derivative of $\tilde{J}_{\alpha,\sigma}$, expressed in terms of
 \begin{itemize}
   \item trace-by-trace convolution and cross-correlation operators,
-  \item the derivative $D_mF[m]w_*$ and its adjoint, and
-  \item trace-by-trace scaling.
+  \item the derivative $D_mF[m]$ and its adjoint, and
+  \item trace-by-trace scaling
 \end{itemize}
-These the first two bullets describe common components of FWI.The
+These the first two items are common components of typical FWI implementations. The
 third is necessary to express the AWI penalty operator.
 
-Note that the second-mentioned operator describes the derivative of
-the acoustic forward modeling operator for a single point source per
-receiver gather, not the extended modeling operator $\bar{F}$. The
-latter is not a modular component of FWI, and a self-contained
-implementation is not likely to be available.
+As pointed out by \cite{GolubPeyreyra:73}, the directional derivative $D_m\tilde{J}_{\alpha,\sigma}[m,d]\delta m$ is the same as the partial directional derivative $D_mJ[u,m,d]\delta m$ with $u = u_{\alpha,\sigma}[m,d]$.
+Since
+\begin{equation}
+D_mJ[u,m,d]\delta m = \langle D_m(S[m]u)\delta m, S[m]u-d \rangle + \alpha^2 \langle D_m(T_{\sigma}[m]u)\delta m,T_{\sigma}[m]u\rangle_U,
+\label{eqn:basederiv}
+\end{equation}
+the key calculations are $D_mS^T$ and $D_m T^T$.
 
-Write $\nabla \tilde{J}_{\alpha,\epsilon}$ in terms of its components:
-\begin{equation}
-= D_mS[m]^T e_0 + \alpha D_mT_{\epsilon}[m]^Te_2
-\label{eqn:tildejgrad}
-\end{equation}
-where $e=(e_0,e_1,e_2)^T$ is the residual vector:
-$$
-e = \left[
-\begin{array}{c}
-S[m]u-d\\
-\epsilon u\\
-\alpha T[m]u
-\end{array}
-\right],\, u=\tilde{u}_{\alpha,\epsilon}[m].
-$$
 Note that the regularization term does not appear in the VPM gradient expression, since it does not depend explicitly on $m$.
 
-Evidently the key calculations are $D_mS^T$ and $D_m T^T$.
+\subsection{Single Trace}
 
-\subsection{Convolution Revisited}
+Since both $S$ and $T$ are block-diagonal, acting trace-by-trace, I will first compute the derivatives in the single-trace case.
+For the remainder of this section, $S$, $T$, and $u_0$ denote the restriction of the corresponding operators to a trace subspace, that is, for some $(\bx_r,\bx_s) \in Q$,
+\begin{eqnarray*}
+  S[m] = S[m]_{\bx_r,\bx_s}&: & L^2[-t_U,t_U] \rightarrow L^2[0,t_D] \\
+  T[m] = T[m]_{\bx_r,\bx_s}&: & L^2[-t_U,t_U] \rightarrow L^2[-t_U,t_U] \\
+  u_{0,\sigma} &=& u_{0,\sigma}(\bx_r,\cdot;\bx_s).
+\end{eqnarray*}
+Inner products and transposes are implicit in these definitions.
 
-To keep track of the data flow in subsequent calculations, it is convenient to introduce a notation for the operator of convolution by $u$.
-This notation must include the time ranges for input and output traces, as these define the index limits in the sums that implement the convolution operator.
-Looked at another way, these ranges define the domain and range of the operator, and that is how the definition in *segyvc.ConvolutionOperator* is arranged: the arguments to the class constructor are the domain and range (as *segyvc.Spaces*, type-checked) and the filename for the convolution kernel.
-Domain, range, and kernel must share spatial geometry (source and receiver positions, in particular number of traces, and time step; these conditions are checked.
-In this section, the convolution operator with compatible trace space domain $X$ and range $Y$, and kernel $z \in Z$ is denoted $C[X,Y,Z,z]$.
-The kernel space $Z$ is also compatible with $X$ and $Y$.
-
-The domain, range, and kernel spaces considered here are all coordinate spaces, defined by finite time index ranges (or, in the continuum case, bounded time ranges).
-For such a coordinate space $X$, denote by $\Pi_X$ the orthogonal projection of the ambient space $l^2$ (or $L^2({\bf R})$ onto $X$.
-In the discrete case, this projection $\Pi_X$ simply extracts the values inside the characterizing index range $I_X$:
-$$
-(\Pi_X f)_i = f_i, \, i \in I_X,
-$$
-and similarly for the continuum case.
-The transpose projection $\Pi_X^T$ injects the array of values within the range into $l^2$, with zeros outside.
-So $\Pi_X^T\Pi_X$ is the projection onto the subspace of $l^2$ (or $L^2({\bf R})$) corresponding to $X$, i.e.
-the range of $\Pi_X^T$, and $\Pi_X \Pi_X^T = I_X$.
-
-Define convolution $f*g$ by the usual sum or integral, for arguments $f$ and $g$ of bounded support. With these conventions,
-$$
-C[X,Y,Z,z]x = \Pi_Y(\Pi_Z^T z * \Pi_X^T x).
-$$
-The class *segyvc.ConvolutionOperator* implements precisely the discrete version of this definition.
-
-Both discrete and continuous convolution are commutative.
-The same is true of the discrete, finite version of convolution implemented in *segyvc.ConvolutionOperator*, and can be expressed as the follow obvious consequence of the definition: if $X, Y, Z$ are compatible trace spaces, and $z \in Z, x \in X$, then
+With these notational conventions, the normal equation \ref{eqn:blocknormal} for $u_{0,\sigma}$ implies
 \begin{equation}
-C[X,Y,Z,z]x = C[Z,Y,X,x]z
-\label{eqn:convcomm}
+  \label{eqn:blocksingle}
+  u_{0,\sigma}[m] = (S[m]^TS[m] + \sigma^2I)^{-1}S[m]^Td.
 \end{equation}
-The transpose of convolution is expressed via the time reversal operator $f \mapsto \check{f}\, \check{f}_i = f_{-i}$:  
-$$
-\langle f*g,h \rangle = \langle g, \check{f}*h \rangle.
-$$ 
-The right-hand side defines the cross-correlation of $f$ and $h$. From the definition of $C$,
-$$
-\langle C[X,Y,Z,z]x, y\rangle_Y = \langle \Pi_Y(\Pi_Z^T z * \Pi_X^T x), y \rangle_Y
-$$
-$$
-= \langle \Pi_Z^T z * \Pi_X^T x, \Pi_y^T y \rangle = \langle \Pi_X^T x, \check{(\Pi_Z^T z)}* \Pi_Y^T y \rangle
-$$
-Denote by $\check{Z}$ the coordinate space produced by time reversal applied to $Z$, that is, $I_{\check{Z}}$ consists of the negatives of members of $I_Z$.
-Then $\check{(\Pi_Z^T z)}=\Pi_{\check{Z}}\check{z}$, so the above is
-$$
-=\langle \Pi_X^T x, \Pi_{\check{Z}}^T \check{z} * \Pi_Y^T y \rangle =\langle x, \Pi_X(\Pi_{\check{Z}}^T \check{z} * \Pi_Y^T y) \rangle_X
-$$
-$$
-= \langle x, C[Y,X,\check{Z},\check{z}]y \rangle_X.
-$$
-That is, $C[X,Y,Z,z]^T = C[Y,X,\check{Z},\check{z}]$.
-
-Combining the commutativity relation $\ref{eqn:convcomm}$ with this identity yields the useful identity:
+So
 \begin{equation}
-C[X,Y,Z,z]^Ty = C[Y,X,\check{Z},\check{z}]y =
-C[\check{Z},X,Y,y]\check{z} =
-C[X,\check{Z},\check{Y},\check{y}]^T\check{z}.
-\label{eqn:convcommtransp}
+  \label{eqn:blockderiv}
+  D_mu_{0,\sigma}[m] \delta m = (S[m]^TS[m] + \sigma^2I)^{-1}((DS[m]^{T}\delta m) (d - S[m]u_{0,\sigma}[m]) - S[m]^T(DS[m]\delta) m u_{0,\sigma}[m])
 \end{equation}
 
-\subsection{Back to AWI penalty gradient}
-
-The SEGY trace spaces involved in the AWI calculation are the data space $D$ and the adaptive filter space $U$.
-Minimization of $J_{\alpha,\epsilon}$ amounts to solution of a least squares problem with operator $C[U,D,D,F[m]w]$, that is, convolution with the predicted data:
-$$
-S[m]u = C[U,D,D,F[m]w_*]u = C[D,D,U,u]F[m]w_*
-$$
-The second expression, equivalent by the commutativity relation \ref{eqn:convcomm} to the first, is more convenient for computing the derivative with respect to $m$ of $S[m]u$:
-$$
-D_m(S[m]u) \delta m = C[D,D,U,u](D_mF[m]\delta m)
-$$
-whence ($e_0 = S[m]u_{\alpha,\epsilon}[m]-d \in D$) the first summand in the gradient of $\tilde{J}_{\alpha,\epsilon}$ is
-$$
-D_m(S[m]u)^T e_0 = D_mF[m]^TC[D,D,U,u]^Te_0.
-$$
-%The derivative $D_mF[m]$ and its adjoint are defined in the *vcl.Function* subclass *asg.fsbop*, and the convolution operator $C$ and its transpose in *segyvc.ConvolutionOperator*.
-
-Introduce the set $SR$ of source-receiver pairs active in the data
-$D$, and scale factor space $P = \{f:SR \rightarrow \bR\}$. Write
+The derivative of $S$ is related to the derivative of $F$ by convolution. The relation with $T$ is more complicated.
+The single-trace version of the definition \ref{eqn:ppf} is
 \[
-  T[m]u = B_2 \circ B_1 \circ B_0[m],
+  (T_{\sigma}[m]u)(t) = \frac{t u(t)}{\|u_{0,\sigma}[m]\|},\,u \in L^2[-t_U,t_U],
 \]
-where 
+so
 \[
-B_2: P \times U \rightarrow U, \, B_2[f,u](bx_r,t;\bx_s) = f(\bx_r,\bx_s) tu(\bx_r,t;\bx_s)
+D(T_{\sigma}[m] u)\delta m = -\frac{t u}{\|u_{0,\sigma}[m]\|^3}\langle D_mu_{0,\sigma}[m]\delta u, u_{0,\sigma}[m]\rangle.
 \]
 \[
-B_1: U \rightarrow P, \,B_1[u](\bx_r,t,\bx_s) = \|u(\bx_r,\cdot; \bx_s)\|^{-1}
+  =-\langle  (S[m]^TS[m] + \sigma^2I)^{-1}((DS[m]^{T}\delta m) (d - S[m]u_{0,\sigma}[m]) - S[m]^T(DS[m]\delta m) u_{0,\sigma}[m]),
+\]
+\[ u_{0,\sigma}[m] \rangle tu \|u_{0,\sigma}[m] \|^{-3}.
+\]
+Since $\|tu\|^2\| u_{0,\sigma}[m]\|^{-4} = \|T_{\sigma}[m]u\|^2\| u_{0,\sigma}[m]\|^{-2}$,
+\[
+  D_m \frac{\alpha^2}{2}\|T_{\sigma}[m]u\|\delta m =-\alpha^2\left( \langle d-S[m]u_{0,\sigma}[m], DS[m]\delta m (S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]\rangle \right.
 \]
 \[
-B_0: M \rightarrow U, \, B_0[m] = u_{0,\epsilon}[m]
-\]
-Then
-\[
- \frac{\alpha^2}{2} D_m(\|T[m]u\|_U^2)\delta m = \alpha^2\langle D_m(
- T[m]u)\delta m, T[m]u \rangle_U
+  \left. - \langle (DS[m]\delta m)u_{0,\sigma}[m],S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]\rangle \right)
 \]
 \[
-  =\alpha^2 \langle DB_2(B_1\circ B_0[m],u)DB_1(B_0[m]) DB_0[m]\delta m,
-  T[m]u\rangle_U
+  \times  \|T_{\sigma}[m]u\|^2\| u_{0,\sigma}[m]\|^{-2}.
+\]
+
+Recall that $S[m]u$ is the convolution of the pertinant trace of $u \in U$ with the corresponding trace of $F[m]$.
+These are functions defined on bounded time intervals: in particular, $F:M \rightarrow L^2([0,t_D])$.
+So restriction and extension must also be involved.
+Define $\Pi_D: L^2(\bR) \rightarrow L^2([0,t_D])$ to be the restriction operator.
+Then its adjoint $\Pi_D^T$ is the zero extension operator: that is, $\Pi_D^Tf(t) = f(t)$ if $t \in [0,t_D]$, $= 0$ else.
+Define $\Pi_U:  L^2(\bR) \rightarrow L^2([-t_U,t_U])$ similarly.
+Then
+\begin{equation}
+  \label{eqn:sdefsingle}
+  S[m]u = \Pi_D(\Pi_D^TF[m] * \Pi_U^Tu)
+\end{equation}
+
+The first term in the expression \ref{eqn:basederiv} for the derivative of $\tilde{J}_{\alpha,\sigma}$ can then be re-written
+\[
+  \langle DS[m]u, S[m]u-d\rangle = \langle \Pi_D(\Pi_D^TDF[m]\delta m * \Pi_U^Tu), S[m]u-d\rangle
+\]
+\[
+  = \langle \Pi_D^TDF[m]\delta m * \Pi_U^Tu, \Pi_D^T (S[m]u-d) \rangle
+\]
+\[
+  = \langle \Pi_D^TDF[m]\delta m,  R(\Pi_U^Tu)*\Pi_D^T (S[m]u-d) \rangle
 \]
 \begin{equation}
-  \label{eqn:compgrad}
-  =\alpha^2 \langle DB_0[m]\delta m, DB_1(B_0[m])^TDB_2(B_1\circ
-  B_0[m],u)^TT[m]u\rangle_U
+  \label{eqn:firstsingle}
+  = \langle DF[m]\delta m, \Pi_D(R(\Pi_U^Tu)*\Pi_D^T (S[m]u-d))\rangle
 \end{equation}
-Next display the derivative and adjoint derivative of $B_1$ and $B_2$:
-$$
-(D_fB_2[f,u]\delta f)(\bx_r,t;\bx_s) = tu(\bx_r,t;\bx_s)\delta f(\bx_r,\bx_s)
-$$
-$$
-D_f(B_2[f,u])^Te_2(\bx_r,\bx_s) = \int dt\,te_2(\bx_r,t;\bx_s)u(\bx_r,t;\bx_S);
-$$
-$$
-D_u B_1[u]\delta u(\bx_r,\bx_s) = -\|u(\bx_r,\cdot;\bx_s)\|^{-3}\langle u(bx_r,\cdot;\bx_s), \delta u(\bx_r,\cdot;\bx_s) \rangle
-$$
-$$
-(D_u B_1[u])^Tf(\bx_r,t;\bx_s) =
--f(\bx_r,\bx_s)\|u(\bx_r,\cdot;\bx_s\|^{-3}u(\bx_r,t;\bx_s)
-$$
-With these expressions, compose various components appearing on the
-right hand side of equation \ref{eqn:compgrad}:
-\[
-  B_1\circ B_0[m] (\bx_r,\bx_s) =  \|\tilde{u}_{0,\epsilon}(\bx_r,\cdot; \bx_s)\|^{-1},
-\]
-\[
- DB_2(B_1\circ B_0[m],u)^Te_2 = \int dt\,te_2(\bx_r,t;\bx_s)u(\bx_r,t;\bx_s),
-\]
-(Note that the result is independent of the first argument.)
-\[
-DB_1(B_0[m])^Tf(\bx_r,t:\bx_s)=
--f(\bx_r,\bx_s)\|u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\|^{-3}u_{0,\epsilon}(\bx_r,\cdot;\bx_s),
-\]
-and
-\[
-  (DB_1(B_0[m])^TDB_2(B_1\circ  B_0[m],u)^TT[m]u )(\bx_r,\cdot,\bx_s)
-\]
-\[
-  = -\int dt\,t(T[m]u)(\bx_r,t;\bx_s)u(\bx_r,t;\bx_s)
-  \|u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\|^{-3}u_{0,\epsilon}(\bx_r,\cdot;\bx_s)
-\]
-In view of the definition of $T[m]$, this is
-\begin{equation}
-  \label{eqn:compgrad1}
-  = -\||T[m]u(\bx_r,\cdot;\bx_s)\|^2
-  \|u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\|^{-2}u_{0,\epsilon}(\bx_r,\cdot;\bx_s)
-\end{equation}
-It remains to compute $ DB_0[m]\delta m,$:
-$$
-B_0[m] = u_{0,\epsilon}[m] = (S[m]^TS[m] + \epsilon^2 I)^{-1}S[m]^Td
-$$
-$$
-= (C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*] + \epsilon^2 I)^{-1}C[U,D,D,F[m]w_*]^Td.
-$$
-So
-$$
-D_mB_0[m]\delta m = (C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*] + \epsilon^2 I)^{-1}
-$$
-$$
-\times ((D_mC[U,D,D,F[m]w_*]\delta m)^Td -(D_m(C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*])\delta m)u_{0,\epsilon}[m])
-$$
-$$
-=(C[U,D,D,F[m]w_*]^TC[U,D,D,F[m]w_*] + \epsilon^2 I)^{-1}
-$$
-$$
-\times
-(D_m(C[U,D,D,F[m]w_*]^Tr_0)\delta m  - C[U,D,D,F[m]w_*]^TD_m(C[U,D,D,F[m]w_*]u_0)\delta m)
-$$ 
-where the derivatives are to be evaluated before the insertion of the definitions 
-$$
-r_0 = d -C[U,D,D,F[m]w_*]u_{0,\epsilon}[m], \, u_0 = u_{0,\epsilon}[m].
-$$
-From the commutativity relation \ref{eqn:convcomm},
-\begin{equation}
-  \label{eqn:compgrad2}
-D_m(C[U,D,D,F[m]w_*]u_0)\delta m = D_m(C[D,D,U,u_0]F[m]w_*)\delta m
-=C[D,D,U,u_0]D_mF[m]\delta m.
-\end{equation}
-Similarly, use the identity \ref{eqn:convcommtransp} to write
-$$
-C[U,D,D,F[m]w_*]^Tr_0 = C[U,\check{D},\check{D},\check{r_0}]^T({F[m]w_*})\check, 
-$$
-so
-\begin{equation} \label{eqn:compgrad3}
-  D_m(C[U,D,D,F[m]w_*]^Tr_0)\delta m
-  =C[U,\check{D},\check{D},\check{r_0}]^T(D_mF[m]w_* \delta m)\check= C[U,D,D,F[m]w_*\delta m]^Tr_0
-\end{equation}
-
-Putting \ref{eqn:compgrad}, \ref{eqn:compgrad1}, \ref{eqn:compgrad2},
-and \ref{eqn:compgrad3} together,
-\[
-  \frac{\alpha^2}{2} D_m(\|T[m]u\|_U^2)\delta m =
-=-\alpha^2 \langle C[U,D,D,F[m]w_*]^TC[D,D,U,u_0]D_mF[m]\delta m +C[U,D,D, D_mF[m]w_*
-\delta m]^Tr_0, 
-\]
-\[
- \||T[m]u(\bx_r,\cdot;\bx_s)\|^2
-  \|u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\|^{-2}u_{0,\epsilon}(\bx_r,\cdot;\bx_s)\rangle_U
-  DB_1(B_0[m])^TDB_2(B_1\circ
-  B_0[m],u)^TT[m]u\rangle_U
-\]
-
-Like all penalty functions, this one is best viewed as a least-squares objective of a map with values in a product space: define
-$$
-A_{\alpha,\epsilon}[m] =  
-\left[
-\begin{array}{c}
-S[m]\\
-\epsilon I\\
-\alpha T_{\epsilon}[m]
-\end{array}
-\right]
-, \, b = \left[
-\begin{array}{c}
-d\\
-0\\
-0
-\end{array}
-\right]
-$$
-Then
-$$
-J_{\alpha,\epsilon}[m,w_*,u] = 0.5\|A_{\alpha,\epsilon}[m]u-b\|^2
-$$
-
-Both Newton's method with the Kaufman Hessian approximation (described in the VCL notebook) and LBFGS or similar first-order methods apply to minimization of $\tilde{J}_{\alpha,\epsilon}[m]$.
-As shown in the VCL notebook, application of these algorithms to a VPM reduction like $\tilde{J}_{\alpha,\epsilon}$ requires access to two linear operator valued functions of $m$, namely $A_{\alpha,\epsilon}[m]$ and the (partial) derivative $D_m(A_{\alpha,\epsilon}[m]u)$.
-The former is the main actor in computation of the objective via solution of the normal equations.
-The latter appears in the VPM gradient formula:
-$$
-\nabla \tilde{J}_{\alpha,\epsilon}[m] = (D_m A_{\alpha,\epsilon}[m]u)|_{u=\tilde{u}_{\alpha,\epsilon}[m]}^T(A_{\alpha,\epsilon}[m]\tilde{u}_{\alpha,\epsilon}[m] -b)
-$$
-Note that $D_m A_{\alpha,\epsilon}[m]u$ is a map from the model parameter space to the product range space of $A$, and the transpose is a map in the other direction.
+In the previous two expressions, $R: L^2(\bR) \rightarrow L^2(\bR)$ is the time-reversal operator: $Rf(t) = f(-t)$.

### 2023-11-10 repro/papers/awigrad/paper.tex
more

@@
-\title{Computation of the Reduced AWI Penalty Function and its Gradient}
-\author{William W. Symes}
+\title{Implementation of Adaptive Waveform Inversion via a Penalty Function}
+\author{Huiyi Chen, Susan E. Minkoff, and William W. Symes}
 
@@
 
-\section{Definitions}
-Hilbert space $X$ has inner product $\langle \cdot,\cdot \rangle_X$ and norm $\|\cdot\|_X$.
+\section{Introduction}
 
-$Q \subset \bR^3 \times \bR^3$ = finite ``acquisition'' set of source-receiver location pairs $(\bx_r,\bx_s)$, assumed disjoint from the diagonal (sources and receivers non-coincident).
-$Q_s$ = projection onto second factor (source locations occurring in $Q$).
+\section{Preliminaries}
+
+$Q \subset \bR^3 \times \bR^3$ is the finite ``acquisition'' set of source-receiver location pairs $(\bx_r,\bx_s)$, assumed disjoint from the diagonal (sources and receivers non-coincident).
+$Q_s$ denotes projection onto second factor (source locations occurring in $Q$).
 
@@
 
-$M = $ model set, bounded open subset of $ C^{\infty}(\bR^3) \times C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3) \times L^{\infty}(\bR^3)$.
+$M = $ bounded open subset of ${\cal M} = C^{\infty}(\bR^3) \times C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3) \times L^{\infty}(\bR^3) \cap L^2(\bR^3) \times L^2(\bR^3)$.
 
@@
     \frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv +
-                                        w(t;\bx_s) \delta(\bx-\bx_s); \\
+                                        w_*(t) \delta(\bx-\bx_s); \\
     \frac{\partial \bv}{\partial t} & = & - \beta \nabla p; \\
@@
   \label{thm:mbd}
-  $\|F[m](\bx_r,\cdot;\bx_s)\|_{L^2([0,t_D])}$ is uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
+  Under the hypotheses of Theorem \label{thm:eu},
+  \begin{itemize}
+  \item [1, ] $\|F[m](\bx_r,\cdot;\bx_s)\|_{L^2([0,t_D])}$ is uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
+  \item [2. ] $F$ has a directional derivative $DF[m]: {\cal M} \rightarrow D$ for every $m \in M$.
+    \item [3. ] for every $m \in M$, $DF[m]$ extends to a bounded map on ${\cal H} = L^2(\bR^3) \times L^2(\bR^3)$
+  \end{itemize}
 \end{theorem}
@@
 
-$S: M \rightarrow {\cal B}(U,D)$ = trace-by-trace convolution with predicted data: $S[m]u = F[m] * u$.
-This notation for convolution in time hides a number of important details: none of the arguments are defined on the entire real line, so extension and truncation are involved in the definition.
-I make these details explicit later in the paper.
+$F$ is the mapping implemented (in finite difference approximation) in {\tt asg.fsbop}, with the additional constraint that the buoyancy field is fixed so that the computational version of ${\cal M}$ consists of (gridded) bulk modulus fields.
 
-Due to smoothness of $F$, $S \in C^1(M,{\cal B}(U,D))$.
-Operation is trace-by-trace, that is block-diagonal: $S[m] = \bigoplus_{(\bx_r,\bx_s) \in Q}S[m]_{\bx_r,\bx_s}$.
-Theorem \ref{thm:mbd} implies that $\|S[m](\bx_r,\cdot;\bx_s)\|_{{\cal B}(L^2([-t_U,t_U]),L^2([0,t_D]))}$ is also uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
+The central object in AWI is the adaptive-filter-to-data operator $S: M \rightarrow {\cal B}(U,D)$, computed by trace-by-trace convolution with predicted data: $S[m]u = F[m] * u$.
+This notation for convolution in time hides a number of important details: none of the arguments are defined on the entire real line, so extension and restriction must be involved in the definition.
+Define $\Pi_D: \bigoplus_{(\bx_r,\bx_s) \in Q} L^2(\bR) \rightarrow D$ to be the restriction operator: that is, for $t \in [0,t_D]$, $\Pi_Df(\bx_r,t;\bx_s) = f(\bx_r,t;\bx_s)$.
+Then its adjoint $\Pi_D^T$ is the zero extension operator: that is, $\Pi_D^Tf(\bx_r,t;\bx_s) = f(\bx_r,t;\bx_s)$ if $t \in [0,t_D]$, $= 0$ else.
+Define $\Pi_U:  \bigoplus_{(\bx_r,\bx_s) \in Q} L^2(\bR) \rightarrow U$ similarly.
+
+The meaning of convolution in the definition of $S$ is: extension followed by (trace-by-trace) convolution followed by restriction. That is, for $u \in U, f \in D$,
+\begin{equation}
+  \label{eqn:convdef}
+  K[u]f(\bx_r,t;\bx_s) = \Pi_D\left(\int\,ds\,(\Pi_U^Tu)(\bx_rt-s;\bx_s) (\Pi_D^Tf)(\bx_r,s;\bx_s)\right)
+\end{equation}
+So defined, $K:U \rightarrow {\cal B}(D,D)$ is continuous and bilinear thanks to Young's inequality.
+
+Convolution is commutative, so the right-hand side of the definition \ref{eqn:convdef} could also be viewed as defining $L: D \rightarrow {\cal B}(U,D)$, for $u \in U, f \in D$,
+\begin{equation}
+  \label{eqn:altconvdef}
+  L[f]u = K[u]f
+\end{equation}
+
+The convolution operator implemented in the module {\tt segyvc} actually implements a time-discrete version of $K$ (or $L$).
+Note that the transpose $K[u]^T]$ is a restricted version of cross-correlation, implemented as the transpose of {\tt segyvc.ConvolutionOperator}.
+
+A proper definition of $S[m]$ is thus
+\begin{equation}
+  \label{eqn:sdef}
+  S[m]u = K[u]F[m] = L[F[m]]u.
+\end{equation}
+$S$ has a directional partial derivative with respect to its first argument, given by
+\begin{equation}
+  \label{eqn:dsdef}
+  D_m(S[m]u) \delta m = K[u]DF[m] \delta m,\,m\in M, u \in U, \delta m \in {\cal M}.
+\end{equation}
+Since $DF[m]$ extends to a continuous linear map: ${\cal H} \rightarrow D$, the same is true of $D_mS[m]u$, $m \in M, u \in U$..
+
+Since operation is trace-by-trace, that is block-diagonal, $S[m] = \mbox{diag}_{(\bx_r,\bx_s) \in Q}S[m]_{\bx_r,\bx_s}$.
+Theorem \ref{thm:mbd} implies that $\|S[m]_{\bx_r,\bx_s}\|_{{\cal B}(L^2([-t_U,t_U]),L^2([0,t_D]))}$ is also uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
+A similar statement applies to the derivative.
 
@@
 
-Part 1: Unpenalized adaptive filter $u_{0,\sigma}[m,d]$ solving regularized least squares data fitting problem:  given $m \in M$, minimizes
+\subsection{ Unpenalized adaptive filter}
+The unpenalized filter $u_{0,\sigma}[m]$ solves the regularized least squares data fitting problem:  given $m \in M$, minimizes
 \[
- J_{0,\sigma}[u,m,d] = \frac{1}{2}(\|S[m]u - d\|_D^2 + \sigma^2 \|u\|^2_U).
+ J_{0,\sigma}[u,m] = \frac{1}{2}(\|S[m]u - d\|_D^2 + \sigma^2 \|u\|^2_U).
 \]
-Since $\sigma > 0$, normal operator is uniformly bounded below and of class $C^1$ in $M$ and $D$. So unique solution exists, and $u_{0,\sigma} \in C^1(M \times D, U)$.
+Since $\sigma > 0$, the normal operator is uniformly bounded below and of class $C^1$ in $M$ and $D$. So a unique solution $u_{0,\sigma} \in C^1(M, U)$ exists.
 
@@
   \label{eqn:blocknormal}
-  (S[m]_{\bx_r,\bx_s}^TS[m]_{\bx_r,\bx_s} + \sigma^2 I)u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)= S[m]_{\bx_r,\bx_s}^Td(\bx_r,\cdot;\bx_s)
+  (S[m]_{\bx_r,\bx_s}^TS[m]_{\bx_r,\bx_s} + \sigma^2 I)u_{0,\sigma}[m](\bx_r,\cdot; \bx_s)= S[m]_{\bx_r,\bx_s}^Td(\bx_r,\cdot;\bx_s)
 \end{equation}
-According to Theorem \ref{thm:mbd} and the Observable Data Assumption, the left-hand side is bounded by a multiple of $\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]}$, uniformly in $m \in M, (\bx_r,\bx_s) \in Q$.
+
+According to Theorem \ref{thm:mbd} and the Observable Data Assumption, the left-hand side of equation \ref{eqn:blocknormal} is bounded by a multiple of $\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]}$, uniformly in $m \in M, (\bx_r,\bx_s) \in Q$.
 According to the Observable Data Assumption, the right-hand side is bounded below by a multiple of $\|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_D]}$, uniform in the same sense.
@@
 
-Part 2:
-Preconditioned penalty operator: block diagonal, 
+\subsection{AWI Penalty Operator}
+The penalty operator is also block diagonal, that is, operates independently on each trace. For the trace at $(\bx_r,\bx_s) \in Q$, 
 \begin{equation}
   \label{eqn:ppf}
-  T_{\sigma}[m]u (\bx_r,t;\bx_s) = \frac{t u(\bx_r,t;\bx_s)}{\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]} }
+  T_{\sigma}[m]u (\bx_r,t;\bx_s) = \frac{t u(\bx_r,t;\bx_s)}{\|u_{0,\sigma}[m](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]} }
 \end{equation}
-Well-defined and $C^1$ as function of $m$ due to the bound \ref{eqn:u0lower} and the the No-zero-trace assumption.
+That is, this penalty operator scales the trace by time, so that the output is small if the nonzero samples are concentrated near $t=0$.
+The second scaling, by the reciprocal of norm of the trace $u_{0,\sigma}(\bx_r,\cdot;\bx_s)$, plays an essential role in linking the AWI objective to the travel-time residual, as is explained in \ref{eqn:HCSMWS:23a}
+This operator is well-defined and $C^1$ as function of $m$ due to the bound \ref{eqn:u0lower} and the the No-zero-trace assumption.
 
@@
   \label{eqn:jtilde}
- \tilde{J}_{\alpha,\sigma}[m,d] = \frac{1}{2}(\|S[m]u_{\alpha,\sigma}[m,d] - d\|_D^2 + \alpha^2\|T_\sigma[m]u_{\alpha,\sigma}[m,d]\|_U^2 + \sigma^2 \|u_{\alpha,\sigma}\|^2_U).
+ \tilde{J}_{\alpha,\sigma}[m,d] = \frac{1}{2}(\|S[m]u_{\alpha,\sigma}[m,d] - d\|_D^2 + \alpha^2\|T_{\sigma}[m]u_{\alpha,\sigma}[m,d]\|_U^2 + \sigma^2 \|u_{\alpha,\sigma}\|^2_U).
 \end{equation}
-in which $u_{\alpha,\sigma}[m,d]$ is the minimizer of $J_{\alpha\sigma}[u,m,d]$ over $u \in U$.
-The fundamental result of variable projection theory \cite[]{GolubPeyreyra:73} applies under the conditions presented here: the stationary points of $J_{\alpha,\sigma}$ and $\tilde{J}_{\alpha,\sigma}$ are in bijective correspondence.
+in which $u_{\alpha,\sigma}[m,d]$ is the minimizer of $J_{\alpha\sigma}[u,m,d]$ over $u \in U$, that is, the solution of the normal equation
+\begin{equation}
+  \label{eqn:normal}
+  (S[m]^TS[m] + \alpha^2T_{\sigma}[m]^TT_{\sigma}[m] + \sigma^2I)u_{\alpha,\sigma}[m] = S[m]^Td
+\end{equation}
+Note that $u_{0,\sigma}$ solves the above system for $\alpha=0$, and that the left-hand side is block-diagonal, similar to the left-hand side of equation \ref{eqn:blocknormal}.
 
-Computing $\tilde{J}_{\alpha,\sigma}$ takes place in two stages, corresponding to the two-stage construction of $J_{\alpha,\sigma}$:
-\begin{itemize}
-\item[1. ] minimize $J_{0,\sigma}$ to compute $u_{0,\sigma}$;
-\item[2. ] use $u_{0,\sigma}$ to compute $u \mapsto T_{\sigma}[m]u$, and compute $u_{\alpha,\sigma}$ by minimizing $J_{\alpha,\sigma}$.
-\end{itemize}
-Therefore evaluating $\tilde{J}_{\alpha,\sigma}$ involves solution of two least-squares problems, in sequence.
+The fundamental result of variable projection theory \cite[]{GolubPereyra:73} applies under the conditions presented here: the stationary points of $J_{\alpha,\sigma}$ and $\tilde{J}_{\alpha,\sigma}$ are in bijective correspondence.
+
+Computing $\tilde{J}_{\alpha,\sigma}$ requires the following steps:
+\begin{enumerate}
+\item given $m$, $d$, $\alpha$, and $\sigma$, compute $u_{0,\sigma}[m]$ by solving the normal equations \ref{eqn:normal} for $\alpha=0$;
+\item compute $v_0 \in U$ by solving $(S[m]^TS[m] + \sigma^2 I)v_0 = u_{0,\sigma}[m]$
+\item use $u_{0,\sigma}[m]$ to construct $T_{\sigma}[m]$, a linear operator on $U$  that scales each input trace at $(\bx_r,\bx_s) \in Q$  by $t\|u_{0,\sigma}(\bx_r,\cdot;\bx_s)\|^{-1}$;
+\item use $S[m]$, $T_{\sigma}[m]$ to compute $u_{\alpha,\sigma}[m]$ by solving the normal equations \ref{eqn:normal} with the given value of $\alpha$;
+\item assemble $\tilde{J}_{\alpha,\sigma}[m]$ according to equation \ref{eqn:jtilde}.
+\end{enumerate}
+
+Therefore evaluating $\tilde{J}_{\alpha,\sigma}$ involves solution of two least-squares problems, in sequence:
+
+{\bf Note:} The implementation {\tt vcalg.conjgrad} of the Conjugate Gradient algorithm for the normal equations optionally returns the residual vector (parameter {\tt e}).
+The obvious way to use this is to create a 3-component product space for the output of the block-column operator $(S[m],\alpha T[m], \sigma I)^T$, and set the right-hand side vector (parameter {\tt b})
+to $(d,0,0)^T$. Then on return one-half the norm-squared of {\tt e} is precisely the (approximate) value of $\tilde{J}_{\alpha,\sigma}[m]$.
 
@@
   \item trace-by-trace convolution and cross-correlation operators,
-  \item the derivative $D_mF[m]$ and its adjoint, and
+  \item the derivative $DF[m]$ and its adjoint, and
   \item trace-by-trace scaling
@@
 
-As pointed out by \cite{GolubPeyreyra:73}, the directional derivative $D_m\tilde{J}_{\alpha,\sigma}[m,d]\delta m$ is the same as the partial directional derivative $D_mJ[u,m,d]\delta m$ with $u = u_{\alpha,\sigma}[m,d]$.
+As pointed out by \cite{GolubPereyra:73}, the directional derivative $D_m\tilde{J}_{\alpha,\sigma}[m,d]\delta m$ is the same as the partial directional derivative $D_mJ[u,m,d]\delta m$ with $u = u_{\alpha,\sigma}[m,d]$.
 Since
@@
 With these notational conventions, the normal equation \ref{eqn:blocknormal} for $u_{0,\sigma}$ implies
-\begin{equation}
-  \label{eqn:blocksingle}
-  u_{0,\sigma}[m] = (S[m]^TS[m] + \sigma^2I)^{-1}S[m]^Td.
-\end{equation}
-So
+\[
+  D_mu_{0,\sigma}[m] \delta m = (S[m]^TS[m] + \sigma^2I)^{-1}
+\]
 \begin{equation}
   \label{eqn:blockderiv}
-  D_mu_{0,\sigma}[m] \delta m = (S[m]^TS[m] + \sigma^2I)^{-1}((DS[m]^{T}\delta m) (d - S[m]u_{0,\sigma}[m]) - S[m]^T(DS[m]\delta) m u_{0,\sigma}[m])
+  \times ((DS[m]^{T}\delta m) (d - S[m]u_{0,\sigma}[m]) - S[m]^T(DS[m]\delta) m u_{0,\sigma}[m])
 \end{equation}
 
-The derivative of $S$ is related to the derivative of $F$ by convolution. The relation with $T$ is more complicated.
+The derivative of $S$ is related to the derivative of $F$ by convolution (definition \ref{eqn:sdef}). The relation with $T$ is more complicated.
 The single-trace version of the definition \ref{eqn:ppf} is
@@
 \[
-  D_m \frac{\alpha^2}{2}\|T_{\sigma}[m]u\|\delta m =-\alpha^2\left( \langle d-S[m]u_{0,\sigma}[m], DS[m]\delta m (S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]\rangle \right.
+  D_m \frac{\alpha^2}{2}\|T_{\sigma}[m]u\|^2\delta m =-\alpha^2\left( \langle d-S[m]u_{0,\sigma}[m], DS[m]\delta m (S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]\rangle \right.
 \]
@@
 \]
+\begin{equation}
+  \label{eqn:penderiv}
+  \times  \|T_{\sigma}[m]u\|^2\| u_{0,\sigma}[m]\|^{-2}.
+\end{equation}
+
+The first term on the right-hand side of the derivative \ref{eqn:basederiv} is
+\begin{equation}
+  \label{eqn:firstsinglebis}
+  \langle DS[m]u, S[m]u-d\rangle = \langle DF[m]\delta m, K[u]^T(S[m]u-d) \rangle
+\end{equation}
+The right-hand side of \ref{eqn:firstsinglebis} can be re-writtenn as
+\begin{equation}
+  \label{eqn:resderiv1}
+  = \langle \delta m, DF[m]^TK[u]^T(S[m]-d)\rangle_M
+\end{equation}
+which identifies the part of the gradient coming from the first term in expression \ref{eqn:basederiv}, in the case of a single trace.
+
+\noindent{\bf Remark:} ``Gradient'' should really be in quotes, as $F$ is not actually differentiable, or even well-defined, in any open subset of ${\cal H}$.
+It is however well-defined in the intersection of $M$ with any ${\cal H}$-closed subspace of ${\cal M}$.
+The latter is necessarily finite-dimensional, but all feasible calculations are carried out in finite-dimensional settings, so that's OK.
+
+Applying the same manipulations to the penalty derivative expression \ref{eqn:penderiv}:
+\[
+  D_m \frac{\alpha^2}{2}\|T_{\sigma}[m] u\|^2\delta m 
+\]
+\[
+  =-\alpha^2\left( \langle d-S[m]u_{0,\sigma}[m], K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]] DF[m]\delta m \rangle \right.
+\]
+\[
+  \left. - \langle K[u_{0,\sigma}[m] ]DF[m]\delta m, S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]\rangle \right)
+\]
 \[
@@
 \]
-
-Recall that $S[m]u$ is the convolution of the pertinant trace of $u \in U$ with the corresponding trace of $F[m]$.
-These are functions defined on bounded time intervals: in particular, $F:M \rightarrow L^2([0,t_D])$.
-So restriction and extension must also be involved.
-Define $\Pi_D: L^2(\bR) \rightarrow L^2([0,t_D])$ to be the restriction operator.
-Then its adjoint $\Pi_D^T$ is the zero extension operator: that is, $\Pi_D^Tf(t) = f(t)$ if $t \in [0,t_D]$, $= 0$ else.
-Define $\Pi_U:  L^2(\bR) \rightarrow L^2([-t_U,t_U])$ similarly.
-Then
+\[
+  = -\alpha^2 \langle \delta m, \|T_{\sigma}[m]u\|^2\| u_{0,\sigma}[m]\|^{-2}
+\]
+\[
+  \times DF[m]^T(K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]]^T (d-S[m]u_{0,\sigma}[m]) 
+\]
 \begin{equation}
-  \label{eqn:sdefsingle}
-  S[m]u = \Pi_D(\Pi_D^TF[m] * \Pi_U^Tu)
+  \label{eqn:penderiv1}
+  -K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]) \rangle
+\end{equation}
+Combining equations \ref{eqn:resderiv1} and \ref{eqn:penderiv1} and substituting $u = u_{\alpha,\sigma}$, obtain for the gradient of $\tilde{J}_{\alpha,\sigma}$ in the case of a single trace:
+\[
+  \nabla \tilde{J}_{\alpha,\sigma}[m] = DF[m]^T\left( K[u_{\alpha,\sigma}[m]]^T(S[m]u_{\alpha,\sigma}[m]-d) - \alpha^2 \|T_{\sigma}[m]u_{\alpha,\sigma}[m]\|^2\| u_{0,\sigma}[m]\|^{-2}\right.
+\]
+\[
+  \left. \times( K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]]^T (d-S[m]u_{0,\sigma}[m]) \right.
+\]
+\begin{equation}
+  \label{eqn:singlegrad}
+\left.-K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]) \right).
 \end{equation}
 
-The first term in the expression \ref{eqn:basederiv} for the derivative of $\tilde{J}_{\alpha,\sigma}$ can then be re-written
+\subsection{Multiple Traces}
+The general case follows immediately from the single trace case.
+It is just necessary to insert restriction to various (receiver, source) pairs in the relation \ref{eqn:singlegrad}, and sum.
+Note that the sum is implicit in the definition of $DF[m]^T$, and of $K^T$ and $S$, but must be carried out explicitly for the part of this expression that amounts to trace-by-trace scaling:
 \[
-  \langle DS[m]u, S[m]u-d\rangle = \langle \Pi_D(\Pi_D^TDF[m]\delta m * \Pi_U^Tu), S[m]u-d\rangle
+  \nabla \tilde{J}_{\alpha,\sigma}[m] = DF[m]^T\left( K[u_{\alpha,\sigma}[m]]^T(S[m]u_{\alpha,\sigma}[m]-d)\right.
 \]
 \[
-  = \langle \Pi_D^TDF[m]\delta m * \Pi_U^Tu, \Pi_D^T (S[m]u-d) \rangle
+  \left.  +\alpha^2 \sum_{(\bx_r,\bx_s)\in Q}\|T_{\sigma}[m]u_{\alpha,\sigma}[m](\bx_r,\cdot;\bx_s)\|^2\| u_{0,\sigma}[m](\bx_r,\cdot;\bx_s\|^{-2}\right.
 \]
 \[
-  = \langle \Pi_D^TDF[m]\delta m,  R(\Pi_U^Tu)*\Pi_D^T (S[m]u-d) \rangle
+  \left. \times( K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]]^T (S[m]u_{0,\sigma}[m]-d)_{(\bx_r,\bx_s)} \right.
 \]
 \begin{equation}
-  \label{eqn:firstsingle}
-  = \langle DF[m]\delta m, \Pi_D(R(\Pi_U^Tu)*\Pi_D^T (S[m]u-d))\rangle
+  \label{eqn:singlegrad}
+\left.+K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]) _{(\bx_r,\bx_s)} \right).
 \end{equation}
-In the previous two expressions, $R: L^2(\bR) \rightarrow L^2(\bR)$ is the time-reversal operator: $Rf(t) = f(-t)$.
+This expression suggestions the following algorithm outline:
+\begin{enumerate}
+\item given $m$, $d$, $\alpha$, and $\sigma$, compute $u_{0,\sigma}[m]$ by solving the normal equations \ref{eqn:normal} for $\alpha=0$;
+\item compute $v_0 \in U$ by solving $(S[m]^TS[m] + \sigma^2 I)v_0 = u_{0,\sigma}[m]$
+\item use $u_{0,\sigma}[m]$ to construct $T_{\sigma}[m]$, a linear operator on $U$  that scales each input trace at $(\bx_r,\bx_s) \in Q$  by $t\|u_{0,\sigma}(\bx_r,\cdot;\bx_s)\|^{-1}$;
+\item use $S[m]$, $T_{\sigma}[m]$ to compute $u_{\alpha,\sigma}[m]$ by solving the normal equations \ref{eqn:normal};
+\item compute $r_0 = S[m]u_{0,\sigma}[m]-d,\,r_{\alpha} = S[m]u_{\alpha,\sigma}[m]-d$ (these may be by-products of steps 1 and 4;
+\item construct linear operator $W_{\alpha,\sigma}[m]$ on $D$ that scales each trace at $(\bx_r,\bx_s) \in Q$  by $ \|tu_{\alpha,\sigma}[m](\bx_r,\cdot;\bx_s)\|^2\|u_{0,\sigma}[m](\bx_r,\cdot;\bx_s)\|^{-4}$
+\item compute four vectors in $D$:
+  \begin{enumerate}
+  \item $G_1 = K[u_{\alpha,\sigma}[m]]^Tr_{\alpha}$
+  \item $G_2 = K[v_0]^Tr_0$
+  \item $G_3 = K[u_{0,\sigma}[m]]^TS[m]v_0$
+  \item $G_0 = G_1 + \alpha^2W_{\alpha,\sigma}[m](G_2 + G_3)$
+  \end{enumerate}
+\item assemble
+  $\nabla \tilde{J}_{\alpha,\sigma}[m] = DF[m]^TG_0$
+\end{enumerate}
+
+\bibliographystyle{seg}
+\bibliography{../../bib/masterref}

### 2023-11-14 iwave/trace/main/txgain.cc
power of t and file-driven x scaling

@@ -1,0 +1,205 @@
+#include <su.h>
+#include <segy.h>
+#include <header.h>
+#include <cubic.h>
+#include <parser.h>
+#include <utility.hh>
+
+const char * sdoc[] = { 
+  "Usage: gain.x in= scale= out= spower= tpower=",
+  "",
+  "Purpose: apply scaling of input traces by trace-dep factor read from ",
+  "segy file with trace length 1, raised to specified power, and also by a",
+  "specified power of t",
+  "",
+  "NOTE: if it exists on call output file is overwritten, else it is created.",
+  "",
+  "Required parameters:",
+  "  in [string]  = filename for input",
+  "  sfile [string] = filename for rms values (data[0] in each trace)",
+  "  out [string] = filename for output",
+  "  spower [int] = 0 power of scale factor",
+  "  tpower [int] = 1 power of t (must be non-negative)",
+  "",
+  "output samples = input samples scaled by the tpower-th power of t, and",
+  "optionally by a power of a factor read from sfile.",
+  "",
+  "If spower=0, then sfile is an inactive argument (and can be left out), ",
+  "and further scaling is not applied. If spower != 0, then the file ",
+  "named by sfile must exist and have at least the number of traces in the ",
+  "input file. The first data sample in each trace should be positive, and ",
+  "output samples in corresponding trace are also scaled by the",
+  "spower-th power of that sfile sample. So",
+  "",
+  "output(t) = input(t) * (t**tpower) * (sfile.data[0]**spower)",
+  "",
+  "NOTE: this function is required for an implementation of AWI.",
+  NULL};
+
+
+int main(int argc, char ** argv) {
+  
+  /* ARGUMENTS */
+  char * in;       /* input file name */
+  char * sfile;    /* scale data file name */
+  char * out;      /* output file name */
+  int spower;      /* power of scale factor */
+  int tpower;      /* power of t */
+
+  /* INTERNAL VARIABLES */
+  PARARRAY * par;  /* param array */
+  FILE * fpin;     /* input file pointer */
+  FILE * fpsc;     /* scale file */
[…]

### 2023-11-14 pyrvl/txgaintest.py
compare old and new calcs

@@ -1,0 +1,91 @@
+import awi
+import vcl
+import vcalg
+import segyvc
+import rsfvc
+import data
+import asg
+import os
+
+TRIP = os.getenv('TRIP')
+
+try:
+
+    # first part: build adaptive filter for data mis-matched to velo
+    
+    # bulk modulus with lens
+    data.model(bulkfile='m.rsf', bulk=4.0, nx=401, nz=201, dx=20, dz=20, lensfac=0.7)
+    # bulk modulus without lens
+    data.model(bulkfile='m0.rsf', bulk=4.0, nx=401, nz=201, dx=20, dz=20, lensfac=1.0)
+
+    # bandpass filter source at sx=4200, sz=3000 (single trace)
+    data.bpfilt(file='wstar.su',nt=251,dt=8.0,s=1.0,f1=1.0,f2=2.5,f3=7.5,f4=12,sx=4200,sz=3000)
+    
+    # create zero data file with same source position, rz=500, rx=[2000,6000]
+    data.rechdr(file='g.su',nt=626,dt=8.0,ntr=201,rx=2000,rz=1000,sx=4200,sz=3000,drx=20)
+    
+    bulksp = rsfvc.Space('m.rsf')
+    datasp = segyvc.Space('g.su')
+    
+    m=vcl.Vector(bulksp,'m.rsf')
+    m0=vcl.Vector(bulksp,'m0.rsf')
+    
+    ######################$ operator ############################
+    F = asg.fsbop(dom=bulksp, rng=datasp, \
+                buoyancy='bym.rsf', source_p='wstar.su', \
+                order=2, sampord=1, nsnaps=20,\
+                cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
+                nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0)
+    ########################## modeling ############################
+    
+    # evaluate F[m]
+    Fm = F(m)
+    Fm0 = F(m0)
+
+    data.rechdr(file='baru.su',nt=251,dt=8.0,\
+            ntr=201,rx=2000,rz=1000,sx=4200,sz=3000,drx=20,delrt=-1000)
+    usp=segyvc.Space('baru.su')
+    baru=vcl.Vector(usp,'baru.su')
+    data.rechdr(file='caps1.su',nt=251,dt=8.0,\
+            ntr=201,rx=2000,rz=1000,sx=4200,sz=3000,drx=20,delrt=-1000)
[…]

### 2023-11-16 repro/papers/awigrad/paper.tex
interim

@@
 
-$P = \bR^{|Q|}$ = real functions on $Q$, with Euclidean norm (no attempt to scale for geometry of acquisition).
+%$P = \bR^{|Q|}$ = real functions on $Q$, with Euclidean norm (no attempt to scale for geometry of acquisition).
 
-$[0,t_d]$ = acquisition time interval
+The data to be inverted consists of $|Q|$ functions of time (``traces'').
+The time interval of recording is assumed to be the same for all traces, namely $[0,t_D]$.
+Traces are assumed square-integrable, members of $ L^2([0,t_D])$ - indeed the square of the trace $L^2$ norm is essentially the energy transferred from the surrounding acoustic fluid to the recording instrument \cite[]{SantosaSymes:00}.
+The set of traces forms the Hilbert space $D$ = data Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([0,t_D])$.
+Note that no attempt is made to weight the traces for the spatial distribution of the acquisition coordinates.
 
-$[-t_U,t_U]$ = adaptive filter time interval
-
-$D$ = data Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([0,t_D])$ (again, no attempt to scale for geometry of acquisition).
-
-$U$ = adaptive filter Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([-t_U,t_U])$
-
-$w_* \in C_0^{\infty}(\bR)$ = known point source wavelet
-
-$M = $ bounded open subset of ${\cal M} = C^{\infty}(\bR^3) \times C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3) \times L^{\infty}(\bR^3) \cap L^2(\bR^3) \times L^2(\bR^3)$.
-
-\begin{theorem}
-  \label{thm:eu}
-For $(\log \kappa, \log \beta) \in M$, shot location ${\bf x}_s\in Q_s$, there exist distributions (pressure and velocity fields) $p(\cdot,\cdot,t;\bx_s)$, ${\bf v}(\cdot,\cdot;\bx_s)$ satisfying
-\begin{itemize}
-\item
+The pressure field $p$, sampled at the receiver locations, is a component of the solution ${p,\bv}$ of the linear acoustic wave equation
 \begin{eqnarray}
@@
 \end{eqnarray}
-\item
-  $p, \bv$ are smooth in the punctured space-time $\bR^4 \setminus \{(\bx_s,t): t \in \bR\}$.
-\end{itemize}
-\end{theorem}
-\begin{proof}
-  See Appendix A.
-\end{proof}
+The right-hand side is localized in space at a source position $\bx_s \in Q_s$, with time dependence given by a {source wavelet} $w_* \in C_0^{\infty}(\bR)$.
+Details about the source mechanism are in fact difficult to measure directly and could justifiably be included in the parameters to be estimated in the solution of the inverse problem, but for the purposes of this discussion the wavelet $w_*$ is assumed known.
 
-Define $F: M \rightarrow D$ by
+The acoustic parameter fields, bulk modulus ($\kappa$) and buoyancy ($\beta$) appearing in the system \ref{eqn:awe} are assumed smooth and uniformly bounded.
+Since the speed of propagation is bounded as is the set of source locations $Q_s$ and the interval of propagation $[0,t_D]$, the values of $\kappa(\bx)$ and $\beta(\bx)$ for sufficiently lartge $|\bx|$ have no effect on the solution of the system \ref{eqn:awe}, so $\log \kappa$ and $\log \beta$ can be assumed square-integrable without loss of generality.
+The set of feasible models $M$ is a  bounded open subset of ${\cal M} = C^{\infty}(\bR^3) \times C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3) \times L^{\infty}(\bR^3) \cap L^2(\bR^3) \times L^2(\bR^3)$.
+
+Under these conditions, the following conclusions may be drawn about the system \ref{eqn:awe}:
+\begin{itemize}
+  \item[A1.
+] For $m = (\log \kappa, \log \beta) \in M$, shot location ${\bf x}_s\in Q_s$, there exist distributions (pressure and velocity fields) $p(\cdot,\cdot;\bx_s)$, ${\bf v}(\cdot,\cdot;\bx_s)$ satisfying the system \ref{eqn:awe};
+  \item[A2. ] $p(\cdot,\cdot,t;\bx_s)$, ${\bf v}(\cdot,\cdot;\bx_s)$ are smooth in the punctured space-time $\bR^4 \setminus \{(\bx_s,t): t \in \bR\}$;
+    \item[A3. ] For $(\log \kappa, \log \beta) \in M$, define 
 \[
@@
 \]
-for the pressure field $p$ solving \ref{eqn:awe}.
+for the pressure field $p$ solving \ref{eqn:awe}. The mapping ({\em forward map}, {\em modeling operator},...) $F: M \rightarrow D$ so defined satisfies
+\begin{itemize}
+  \item[A3.1 ] $\|F[m](\bx_r,\cdot;\bx_s)\|_{L^2([0,t_D])}$ is uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
+  \item[A3.2 ] $F$ has a directional derivative $DF[m]: {\cal M} \rightarrow D$ for every $m \in M$;
+  \item[A3.3 ] for every $m \in M$, $DF[m]$ extends to a bounded map on ${\cal H} = L^2(\bR^3) \times L^2(\bR^3)$.
+  \end{itemize}
+\end{itemize}
+See \cite{Symes:23a} for detailed statements and justifications of these assertions.
 
-\begin{theorem}
-  \label{thm:mbd}
-  Under the hypotheses of Theorem \label{thm:eu},
-  \begin{itemize}
-  \item [1, ] $\|F[m](\bx_r,\cdot;\bx_s)\|_{L^2([0,t_D])}$ is uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
-  \item [2. ] $F$ has a directional derivative $DF[m]: {\cal M} \rightarrow D$ for every $m \in M$.
-    \item [3. ] for every $m \in M$, $DF[m]$ extends to a bounded map on ${\cal H} = L^2(\bR^3) \times L^2(\bR^3)$
-  \end{itemize}
-\end{theorem}
-\begin{proof}
-  See Appendix A.
-\end{proof}
-
+{\bf Note:}
 $F$ is the mapping implemented (in finite difference approximation) in {\tt asg.fsbop}, with the additional constraint that the buoyancy field is fixed so that the computational version of ${\cal M}$ consists of (gridded) bulk modulus fields.
 
-The central object in AWI is the adaptive-filter-to-data operator $S: M \rightarrow {\cal B}(U,D)$, computed by trace-by-trace convolution with predicted data: $S[m]u = F[m] * u$.
+Adaptive filters are collections of traces defined on a (possibly)
+different time interval, usually conveniently chosen to be
+time-symmetric, that is, $[-t_U,t_U]$. The trace geometry of adaptive
+filters is otherwise the same as that of the data traces. $U$ will denote the  adaptive
+filter Hilbert space: $U= \bigoplus_{(\bx_r,\bx_s) \in Q} L^2([-t_U,t_U])$.
+
+The central object in AWI is the adaptive-filter-to-data operator $S:
+M \rightarrow {\cal B}(U,D)$, computed by trace-by-trace convolution
+with predicted data: $S[m]u = F[m] * u$.
 This notation for convolution in time hides a number of important details: none of the arguments are defined on the entire real line, so extension and restriction must be involved in the definition.
@@
 The convolution operator implemented in the module {\tt segyvc} actually implements a time-discrete version of $K$ (or $L$).
-Note that the transpose $K[u]^T]$ is a restricted version of cross-correlation, implemented as the transpose of {\tt segyvc.ConvolutionOperator}.
+Note that the transpose $K[u]^T$ is a restricted version of cross-correlation, implemented as the transpose of {\tt segyvc.ConvolutionOperator}.
 
@@
 Since operation is trace-by-trace, that is block-diagonal, $S[m] = \mbox{diag}_{(\bx_r,\bx_s) \in Q}S[m]_{\bx_r,\bx_s}$.
-Theorem \ref{thm:mbd} implies that $\|S[m]_{\bx_r,\bx_s}\|_{{\cal B}(L^2([-t_U,t_U]),L^2([0,t_D]))}$ is also uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
+Fact A3.1 implies that $\|S[m]_{\bx_r,\bx_s}\|_{{\cal B}(L^2([-t_U,t_U]),L^2([0,t_D]))}$ is also uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
 A similar statement applies to the derivative.
@@
 
-According to Theorem \ref{thm:mbd} and the Observable Data Assumption, the left-hand side of equation \ref{eqn:blocknormal} is bounded by a multiple of $\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]}$, uniformly in $m \in M, (\bx_r,\bx_s) \in Q$.
+According to Fact A3.1 and the Observable Data Assumption, the left-hand side of equation \ref{eqn:blocknormal} is bounded by a multiple of $\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]}$, uniformly in $m \in M, (\bx_r,\bx_s) \in Q$.
 According to the Observable Data Assumption, the right-hand side is bounded below by a multiple of $\|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_D]}$, uniform in the same sense.
@@
 That is, this penalty operator scales the trace by time, so that the output is small if the nonzero samples are concentrated near $t=0$.
-The second scaling, by the reciprocal of norm of the trace $u_{0,\sigma}(\bx_r,\cdot;\bx_s)$, plays an essential role in linking the AWI objective to the travel-time residual, as is explained in \ref{eqn:HCSMWS:23a}
+The second scaling, by the reciprocal of norm of the trace $u_{0,\sigma}(\bx_r,\cdot;\bx_s)$, plays an essential role in linking the AWI objective to the travel-time residual, as is explained by \cite{HCSMWS:23a}
 This operator is well-defined and $C^1$ as function of $m$ due to the bound \ref{eqn:u0lower} and the the No-zero-trace assumption.
@@
 
-Therefore evaluating $\tilde{J}_{\alpha,\sigma}$ involves solution of two least-squares problems, in sequence:
+Therefore evaluating $\tilde{J}_{\alpha,\sigma}$ involves solution of
+two least-squares problems, in sequence.
 
@@
 to $(d,0,0)^T$. Then on return one-half the norm-squared of {\tt e} is precisely the (approximate) value of $\tilde{J}_{\alpha,\sigma}[m]$.
+
+To express the penalty operator $T_{\sigma}[m]$ conveniently, along
+with similar operators appearing in the expression for the gradient
+developed below, introduce functions
+\begin{enumerate}
+\item $R: U \rightarrow ({\bf R}^{+})^{|Q|}$,
+  \[
+    Ru((\bx_r,\bx_s)) = \|u(\bx_r, \cdot;\bx_s)\|
+  \]
+\item $W: {\bf R}^{|Q|} \times {\bf R} \times {\bf R}^{+}
+  \rightarrow {\cal B}(D,D)$,
+  \[
+    W(q,p_s, p_t)f (\bx_r,t;\bx_s) = t^{p_t}
+    q(\bx_r,\bx_s)^{p_s}f(\bx_r,t;\bx_s)
+  \]
+\end{enumerate}
+Use the same notation for the similar function with $D$ replaced by
+$U$. Note that $W$ is only well-defined if either $p_s \ge 0$ or $q>0$.
+
+With these notations, $T_{\sigma}[m]$ may be expressed as
+\begin{equation}
+  \label{eqn:talt}
+  T_{\sigma}[m] = W(Ru_{0,\sigma}[m],-1, 1)
+\end{equation}
+
+{\bf Note:} IWAVE implements the mappings $R$ and $W$ in the form of
+the commands {\tt iwave/trace/main/rms.x} and {\tt
+  iwave/trace/main/txgain.x}. The natural python implementation of $R$
+would output the vector of trace norms as a python array (or a NumPy
+ndarray). However that would require extracting the number of traces
+from the input, and checking that the inputs are compatible. This is
+already implicitly done in the IWAVE manipulations of SU trace
+data. So {\tt iwave/trace/main/rms.x} stores the vector of trace norms
+as the 0th data sample in a set of traces, one for each trace of the
+input data: all header words are the same except for {\tt ns}, which
+is set to 1.
+
+The python functions {\tt awi.setrms} and {\tt awi.applytxgain} call
+the two IWAVE commands, manipulating the data as SU data files, rather
+than as VCL vectors. That is, these are simply functions, not {\tt
+  vcl.Function} objects. Since VCL vectors defined on a {\tt segyvc.Space}
+have access to their data files (keyword {\tt data}), these is
+convenient to use in the definitions of AWI penalty operator,
+gradient, and so on.
 
@@
 \subsection{Multiple Traces}
-The general case follows immediately from the single trace case.
-It is just necessary to insert restriction to various (receiver, source) pairs in the relation \ref{eqn:singlegrad}, and sum.
-Note that the sum is implicit in the definition of $DF[m]^T$, and of $K^T$ and $S$, but must be carried out explicitly for the part of this expression that amounts to trace-by-trace scaling:
+The general case follows immediately from the single trace case. It is
+just necessary to observe that the scalare multiplication in the
+expression \ref{eqn:singlegrad} is a trace-by-trace scaling that can
+be computed by a suitable combination of the mappings $R$ and $W$,
+defined above. The gradient is unlikely to be needed without the value
+of $\tilde{J}_{\alpha,\sigma}$, so assume that
+$T_{\sigma}[m]u_{\alpha,\sigma}$ is available, as well as the vector
+of trace norms of $Ru_{0,\sigma}[m]$. Then multiplication of a trace
+at $(\bx_r,\bx_s)$ by
+$\|T_{\sigma}[m]u_{\alpha,\sigma}[m](\bx_r,\cdot;\bx_s)\|^2\|
+u_{0,\sigma}[m](\bx_r,\cdot;\bx_s)\|^{-2}$ is precisely application of
+\[
+  W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m] u_{\alpha,\sigma}[m],2,0)
+\]
+So
 \[
@@
 \[
-  \left.  +\alpha^2 \sum_{(\bx_r,\bx_s)\in Q}\|T_{\sigma}[m]u_{\alpha,\sigma}[m](\bx_r,\cdot;\bx_s)\|^2\| u_{0,\sigma}[m](\bx_r,\cdot;\bx_s\|^{-2}\right.
-\]
-\[
-  \left. \times( K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]]^T (S[m]u_{0,\sigma}[m]-d)_{(\bx_r,\bx_s)} \right.
+  \left.  +\alpha^2 W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{\alpha,\sigma},2,0)
+( K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]]^T (S[m]u_{0,\sigma}[m]-d) \right.
 \]
@@
   \label{eqn:singlegrad}
-\left.+K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]) _{(\bx_r,\bx_s)} \right).
+\left.+K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m])  \right).
 \end{equation}
@@
 \begin{enumerate}
-\item given $m$, $d$, $\alpha$, and $\sigma$, compute $u_{0,\sigma}[m]$ by solving the normal equations \ref{eqn:normal} for $\alpha=0$;
-\item compute $v_0 \in U$ by solving $(S[m]^TS[m] + \sigma^2 I)v_0 = u_{0,\sigma}[m]$
-\item use $u_{0,\sigma}[m]$ to construct $T_{\sigma}[m]$, a linear operator on $U$  that scales each input trace at $(\bx_r,\bx_s) \in Q$  by $t\|u_{0,\sigma}(\bx_r,\cdot;\bx_s)\|^{-1}$;
-\item use $S[m]$, $T_{\sigma}[m]$ to compute $u_{\alpha,\sigma}[m]$ by solving the normal equations \ref{eqn:normal};
-\item compute $r_0 = S[m]u_{0,\sigma}[m]-d,\,r_{\alpha} = S[m]u_{\alpha,\sigma}[m]-d$ (these may be by-products of steps 1 and 4;
-\item construct linear operator $W_{\alpha,\sigma}[m]$ on $D$ that scales each trace at $(\bx_r,\bx_s) \in Q$  by $ \|tu_{\alpha,\sigma}[m](\bx_r,\cdot;\bx_s)\|^2\|u_{0,\sigma}[m](\bx_r,\cdot;\bx_s)\|^{-4}$
+\item given $m$, $d$, $\alpha$, and $\sigma$, compute
+  $u_{0,\sigma}[m]$ by solving the normal equations \ref{eqn:normal}
+  for $\alpha=0$, and $T_{\sigma}[m]=W(Ru_{0,\sigma},-1,1)$ (this has
+  likely already been done to compute the value of $\tilde{J}_{\alpha,\sigma}$;
+\item use $S[m]$, $T_{\sigma}[m]$ to compute $u_{\alpha,\sigma}[m]$ by
+  solving the normal equations \ref{eqn:normal}, and record
+  $T_{\sigma}[m]u_{\alpha,\sigma}[m]$ as a by-product;
+\item compute $v_{0\sigma} \in U$ by solving $(S[m]^TS[m] + \sigma^2 I)v_{0,\sigma} = u_{0,\sigma}[m]$
+\item compute $r_0 = S[m]u_{0,\sigma}[m]-d,\,r_{\alpha} = S[m]u_{\alpha,\sigma}[m]-d$ (these may be by-products of steps 1 and 3;
 \item compute four vectors in $D$:
@@
   \item $G_3 = K[u_{0,\sigma}[m]]^TS[m]v_0$
-  \item $G_0 = G_1 + \alpha^2W_{\alpha,\sigma}[m](G_2 + G_3)$
+  \item $G_0 = G_1 + \alpha^2  W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{\alpha,\sigma}[m],2,0) (G_2 + G_3)$
   \end{enumerate}
@@
 
+\subsection{Warner-Guasch AWI}
+The (``forward'') AWI objective defined in \cite{Warner:16} is
+the limit
+\[
+  \tilde{J}_{\sigma}[m] = \lim_{\alpha \rightarrow 0}
+  \frac{1}{\alpha^2}(\tilde{J}_{\alpha,\sigma}[m]-\tilde{J}_{0,\sigma})
+\]
+\begin{equation}
+  \label{eqn:wg}
+  = \frac{1}{2} \|T_{\sigma}[m]u_{0,\sigma}\|^2
+\end{equation}
+\cite{Warner:16} presume that $S[m]u_{0,\sigma} = d$, which is
+manifestly not the case for $\sigma > 0$, however for small $\sigma$
+the error is presumably negligible. Neglecting it, the gradient becomes
+\[
+  \nabla \tilde{J}_{\sigma}[m] = DF[m]^T
+    W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{0,\sigma}[m],2,0)
+\]
+\begin{equation}
+  \label{eqn:wggrad}
+\times K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]).
+\end{equation}
+
 \bibliographystyle{seg}

### 2023-11-16 pyrvl/awi.py
generic power scaling

@@ -34,7 +34,8 @@ def setrms(u, urms):
         raise Exception('called from setrms')
 
 
-def compawipensol(u, p, alpha=1.0, u0rms=None):
+#def compawipensol(u, p, alpha=1.0, u0rms=None):
+def applytxgain(u, p, u0rms=None, spower=0, tpower=1):
     '''
     Compute action of AWI penalty operator on a set of traces. Multiplies 
     trace samples by time and penalty weight, optionally scales by 
@@ -49,42 +50,55 @@ def compawipensol(u, p, alpha=1.0, u0rms=None):
     (u) and output (p)
     precond (int): preconditioning flag. 0 = no scaling, else scaling
     applied
-    alpha (float): penalty weight
     '''
     try:
 #        print('compawipen')
-#        print('compawipen: u=' + u + ' p=' + p + ' u0rms=' + u0rms + ' alpha=' + str(alpha))
 #        print('next')
         if not isinstance(TRIP,str):
             raise Exception('string TRIP not defined')
         if not os.path.exists(TRIP):
             raise Exception('TRIP package = ' + TRIP + ' not found')
+# 23.11.15 old version        
 #        print('sanity')
-        if linalg.sanity(u,'su') and linalg.sanity(p,'su'):
-            cmd = os.path.join(TRIP,'iwave/trace/main/awipen.x') \
-                + ' in=' + u + ' rms=' + u0rms \
-                + ' out=' + p + ' precond=1' \
-                + ' alpha=' + str(alpha)
-            if (u0rms is not None):
-                if linalg.sanity(u0rms,'su'):
+#        if linalg.sanity(u,'su') and linalg.sanity(p,'su'):
+#            cmd = os.path.join(TRIP,'iwave/trace/main/awipen.x') \
+#                + ' in=' + u + ' rms=' + u0rms \
+#                + ' out=' + p + ' precond=1' \
+#                + ' alpha=' + str(alpha)
+#            if (u0rms is not None):
+#                if linalg.sanity(u0rms,'su'):
 #                    print('alpha=' + str(alpha))
-                    ret = os.system(cmd)
-                else:
-                    raise Exception('u0rms does not have suffix .su')
+#                    ret = os.system(cmd)
+#                else:
+#                    raise Exception('u0rms does not have suffix .su')
+#            else:
+#                #print(cmd + ' in=' + u  
+#                #            + ' out=' + p + ' precond=0 alpha='
+#                #            + str(alpha))
[…]

### 2023-12-14 repro/papers/awigrad/paper.tex
update

@@
 \]
-Since $\sigma > 0$, the normal operator is uniformly bounded below and of class $C^1$ in $M$ and $D$. So a unique solution $u_{0,\sigma} \in C^1(M, U)$ exists.
+Since $\sigma > 0$, the normal operator is uniformly bounded below and of class $C^1$ in $M$ and $D$. So a unique solution $u_{0,\sigma}[m] \in C^1(M, U)$ exists.
 
@@
 That is, this penalty operator scales the trace by time, so that the output is small if the nonzero samples are concentrated near $t=0$.
-The second scaling, by the reciprocal of norm of the trace $u_{0,\sigma}(\bx_r,\cdot;\bx_s)$, plays an essential role in linking the AWI objective to the travel-time residual, as is explained by \cite{HCSMWS:23a}
-This operator is well-defined and $C^1$ as function of $m$ due to the bound \ref{eqn:u0lower} and the the No-zero-trace assumption.
-
-\section{Objectives}
-Define for $\alpha \ge 0$
-\[
- J_{\alpha,\sigma}[u,m,d] = \frac{1}{2}(\|S[m]u - d\|_D^2 + \alpha^2\|T_\sigma[m]u\|_U^2 + \sigma^2 \|u\|^2_U).
-\]
-Note that the notation is consistent, that is, reduces to $J_{0,\sigma}$ as defined above for $\alpha=0$.
-
-The variable projection reduction of $J_{\alpha,\sigma}$ is
-\begin{equation}
-  \label{eqn:jtilde}
- \tilde{J}_{\alpha,\sigma}[m,d] = \frac{1}{2}(\|S[m]u_{\alpha,\sigma}[m,d] - d\|_D^2 + \alpha^2\|T_{\sigma}[m]u_{\alpha,\sigma}[m,d]\|_U^2 + \sigma^2 \|u_{\alpha,\sigma}\|^2_U).
-\end{equation}
-in which $u_{\alpha,\sigma}[m,d]$ is the minimizer of $J_{\alpha\sigma}[u,m,d]$ over $u \in U$, that is, the solution of the normal equation
-\begin{equation}
-  \label{eqn:normal}
-  (S[m]^TS[m] + \alpha^2T_{\sigma}[m]^TT_{\sigma}[m] + \sigma^2I)u_{\alpha,\sigma}[m] = S[m]^Td
-\end{equation}
-Note that $u_{0,\sigma}$ solves the above system for $\alpha=0$, and that the left-hand side is block-diagonal, similar to the left-hand side of equation \ref{eqn:blocknormal}.
-
-The fundamental result of variable projection theory \cite[]{GolubPereyra:73} applies under the conditions presented here: the stationary points of $J_{\alpha,\sigma}$ and $\tilde{J}_{\alpha,\sigma}$ are in bijective correspondence.
-
-Computing $\tilde{J}_{\alpha,\sigma}$ requires the following steps:
-\begin{enumerate}
-\item given $m$, $d$, $\alpha$, and $\sigma$, compute $u_{0,\sigma}[m]$ by solving the normal equations \ref{eqn:normal} for $\alpha=0$;
-\item compute $v_0 \in U$ by solving $(S[m]^TS[m] + \sigma^2 I)v_0 = u_{0,\sigma}[m]$
-\item use $u_{0,\sigma}[m]$ to construct $T_{\sigma}[m]$, a linear operator on $U$  that scales each input trace at $(\bx_r,\bx_s) \in Q$  by $t\|u_{0,\sigma}(\bx_r,\cdot;\bx_s)\|^{-1}$;
-\item use $S[m]$, $T_{\sigma}[m]$ to compute $u_{\alpha,\sigma}[m]$ by solving the normal equations \ref{eqn:normal} with the given value of $\alpha$;
-\item assemble $\tilde{J}_{\alpha,\sigma}[m]$ according to equation \ref{eqn:jtilde}.
-\end{enumerate}
-
-Therefore evaluating $\tilde{J}_{\alpha,\sigma}$ involves solution of
-two least-squares problems, in sequence.
-
-{\bf Note:} The implementation {\tt vcalg.conjgrad} of the Conjugate Gradient algorithm for the normal equations optionally returns the residual vector (parameter {\tt e}).
-The obvious way to use this is to create a 3-component product space for the output of the block-column operator $(S[m],\alpha T[m], \sigma I)^T$, and set the right-hand side vector (parameter {\tt b})
-to $(d,0,0)^T$. Then on return one-half the norm-squared of {\tt e} is precisely the (approximate) value of $\tilde{J}_{\alpha,\sigma}[m]$.
+The second scaling, by the reciprocal of norm of the trace $u_{0,\sigma}[m](\bx_r,\cdot;\bx_s)$, plays an essential role in linking the AWI objective to the travel-time residual, as is explained by \cite{HCSMWS:23a}
+This operator is well-defined and $C^1$ as function of $m$ due to the
+bound \ref{eqn:u0lower} and the the No-zero-trace assumption.
 
@@
   vcl.Function} objects. Since VCL vectors defined on a {\tt segyvc.Space}
-have access to their data files (keyword {\tt data}), these is
-convenient to use in the definitions of AWI penalty operator,
+have access to their data files (keyword {\tt data}), these are
+convenient to use to express the action of $R$ and $W$ in AWI penalty operator,
 gradient, and so on.
 
+\section{Objectives}
+Define for $\alpha \ge 0$
+\[
+ J_{\alpha,\sigma}[u,m,d] = \frac{1}{2}(\|S[m]u - d\|_D^2 + \alpha^2\|T_\sigma[m]u\|_U^2 + \sigma^2 \|u\|^2_U).
+\]
+Note that the notation is consistent, that is, reduces to $J_{0,\sigma}$ as defined above for $\alpha=0$.
+
+The variable projection reduction of $J_{\alpha,\sigma}$ is
+\begin{equation}
+  \label{eqn:jtilde}
+ \tilde{J}_{\alpha,\sigma}[m,d] = \frac{1}{2}(\|S[m]u_{\alpha,\sigma}[m,d] - d\|_D^2 + \alpha^2\|T_{\sigma}[m]u_{\alpha,\sigma}[m,d]\|_U^2 + \sigma^2 \|u_{\alpha,\sigma}\|^2_U).
+\end{equation}
+in which $u_{\alpha,\sigma}[m,d]$ is the minimizer of $J_{\alpha\sigma}[u,m,d]$ over $u \in U$, that is, the solution of the normal equation
+\begin{equation}
+  \label{eqn:normal}
+  (S[m]^TS[m] + \alpha^2T_{\sigma}[m]^TT_{\sigma}[m] + \sigma^2I)u_{\alpha,\sigma}[m] = S[m]^Td
+\end{equation}
+Note that $u_{0,\sigma}[m]$ solves the above system for $\alpha=0$, and that the left-hand side is block-diagonal, similar to the left-hand side of equation \ref{eqn:blocknormal}.
+
+Computing $\tilde{J}_{\alpha,\sigma}$ requires the following steps:
+\begin{enumerate}
+\item given $m$, $d$, $\alpha$, and $\sigma$, compute $u_{0,\sigma}[m]$ by solving the normal equations \ref{eqn:normal} for $\alpha=0$;
+\item use $u_{0,\sigma}[m]$ to construct $T_{\sigma}[m]$ by means of
+  equation \ref{eqn:talt};
+\item use $S[m]$, $T_{\sigma}[m]$ to compute $u_{\alpha,\sigma}[m]$ by solving the normal equations \ref{eqn:normal} with the given value of $\alpha$;
+\item assemble $\tilde{J}_{\alpha,\sigma}[m]$ according to equation \ref{eqn:jtilde}.
+\end{enumerate}
+
+Therefore evaluating $\tilde{J}_{\alpha,\sigma}$ involves solution of
+two least-squares problems, in sequence.
+
+{\bf Note:} The implementation {\tt vcalg.conjgrad} of the Conjugate Gradient algorithm for the normal equations optionally returns the residual vector (parameter {\tt e}).
+The obvious way to use this is to create a 3-component product space for the output of the block-column operator $(S[m],\alpha T[m], \sigma I)^T$, and set the right-hand side vector (parameter {\tt b})
+to $(d,0,0)^T$. Then on return one-half the norm-squared of {\tt e} is precisely the (approximate) value of $\tilde{J}_{\alpha,\sigma}[m]$.
+
 \section{Derivatives}
+The fundamental result of variable projection theory
+\cite[]{GolubPereyra:73} applies under the conditions presented here:
+the stationary points of $J_{\alpha,\sigma}$ and
+$\tilde{J}_{\alpha,\sigma}$ are in bijective correspondence. Moreover,
+\[
+D_m\tilde{J}_{\alpha,\sigma}[m]\delta m D_mJ[u,m]\delta m|_{u =
+  u_{\alpha,\sigma}[m,d]}
+\]
+\begin{equation}
+= \langle (D_mS[m]u\delta m)_{u =
+  u_{\alpha,\sigma}[m]} (S[m]u_{\alpha,\sigma}[m]-d \rangle + \alpha^2 \langle D_m(T_{\sigma}[m]u)\delta m)_{u=u_{\alpha,\sigma}[m]},T_{\sigma}[m]u_{\alpha,\sigma}[m]\rangle_U.
+\label{eqn:basederiv}
+\end{equation}
+Note that the regularization term does not appear in the VPM gradient expression, since it does not depend explicitly on $m$.
+
 This section presents the derivative of $\tilde{J}_{\alpha,\sigma}$, expressed in terms of
@@
   \item the derivative $DF[m]$ and its adjoint, and
-  \item trace-by-trace scaling
+  \item trace-by-trace scaling via the operators $R$ and $W$.
 \end{itemize}
@@
 
-As pointed out by \cite{GolubPereyra:73}, the directional derivative $D_m\tilde{J}_{\alpha,\sigma}[m,d]\delta m$ is the same as the partial directional derivative $D_mJ[u,m,d]\delta m$ with $u = u_{\alpha,\sigma}[m,d]$.
-Since
-\begin{equation}
-D_mJ[u,m,d]\delta m = \langle D_m(S[m]u)\delta m, S[m]u-d \rangle + \alpha^2 \langle D_m(T_{\sigma}[m]u)\delta m,T_{\sigma}[m]u\rangle_U,
-\label{eqn:basederiv}
-\end{equation}
-the key calculations are $D_mS^T$ and $D_m T^T$.
-
-Note that the regularization term does not appear in the VPM gradient expression, since it does not depend explicitly on $m$.
-
 \subsection{Single Trace}
 
-Since both $S$ and $T$ are block-diagonal, acting trace-by-trace, I will first compute the derivatives in the single-trace case.
-For the remainder of this section, $S$, $T$, and $u_0$ denote the restriction of the corresponding operators to a trace subspace, that is, for some $(\bx_r,\bx_s) \in Q$,
+Since both $S$ and $T$ are block-diagonal, acting trace-by-trace, I
+will first compute the derivatives in the single-trace case. For the
+remainder of this section, $F$, $S$, $T$, $d$, $u$, and $u_0$ denote
+the restriction of the corresponding operators and vectors to a trace subspace, that is, for some $(\bx_r,\bx_s) \in Q$,
 \begin{eqnarray*}
+  F[m] = F[m]_{\bx_r,\bx_s}&: & M \rightarrow D_{\bx_r,\bx_s} =
+                 L^2[0,t_D] \\
   S[m] = S[m]_{\bx_r,\bx_s}&: & L^2[-t_U,t_U] \rightarrow L^2[0,t_D] \\
   T[m] = T[m]_{\bx_r,\bx_s}&: & L^2[-t_U,t_U] \rightarrow L^2[-t_U,t_U] \\
-  u_{0,\sigma} &=& u_{0,\sigma}(\bx_r,\cdot;\bx_s).
+  u_{0,\sigma}[m] &=& u_{0,\sigma}[m](\bx_r,\cdot;\bx_s).
 \end{eqnarray*}
@@
 
-With these notational conventions, the normal equation \ref{eqn:blocknormal} for $u_{0,\sigma}$ implies
+The derivative of $S$ is related to the derivative of $F$ by
+convolution (definition \ref{eqn:sdef}), so the first term in the
+right-hand side of equation \ref{eqn:basederiv} is
+\[
+  \langle D_m(S[m]u)\delta m, S[m]u-d \rangle = \langle
+  K[u]DF[m]\delta m, S[m]u-d \rangle
+\]
+\begin{equation}
+  \label{eqn:resderiv}
+  = \langle DF[m]\delta m, K[u]^T(S[m]u-d) \rangle
+\end{equation}
+
+With the notational conventions listed above, the normal equation \ref{eqn:blocknormal} for $u_{0,\sigma}[m]$ implies
 \[
@@
 \end{equation}
-
-The derivative of $S$ is related to the derivative of $F$ by convolution (definition \ref{eqn:sdef}). The relation with $T$ is more complicated.
 The single-trace version of the definition \ref{eqn:ppf} is
 \[
-  (T_{\sigma}[m]u)(t) = \frac{t u(t)}{\|u_{0,\sigma}[m]\|},\,u \in L^2[-t_U,t_U],
+  T_{\sigma}[m]u)(t) = \frac{t u(t)}{\|u_{0,\sigma}[m]\|},\,u \in L^2[-t_U,t_U],
 \]
@@
 \[
-  =-\langle  (S[m]^TS[m] + \sigma^2I)^{-1}((DS[m]^{T}\delta m) (d - S[m]u_{0,\sigma}[m]) - S[m]^T(DS[m]\delta m) u_{0,\sigma}[m]),
-\]
-\[ u_{0,\sigma}[m] \rangle tu \|u_{0,\sigma}[m] \|^{-3}.
-\]
-Since $\|tu\|^2\| u_{0,\sigma}[m]\|^{-4} = \|T_{\sigma}[m]u\|^2\| u_{0,\sigma}[m]\|^{-2}$,
-\[
-  D_m \frac{\alpha^2}{2}\|T_{\sigma}[m]u\|^2\delta m =-\alpha^2\left( \langle d-S[m]u_{0,\sigma}[m], DS[m]\delta m (S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]\rangle \right.
+  =-\frac{t u}{\|u_{0,\sigma}[m]\|^3}\langle  (S[m]^TS[m] +
+  \sigma^2I)^{-1}((DS[m]^{T}\delta m) (d - S[m]u_{0,\sigma}[m])
 \]
 \[
-  \left. - \langle (DS[m]\delta m)u_{0,\sigma}[m],S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]\rangle \right)
+   - S[m]^T(DS[m]\delta m) u_{0,\sigma}[m]), u_{0,\sigma}[m] \rangle
+\]
+Define
+\begin{equation}
+  \label{eqn:defvsigma}
+  v_{\sigma} = (S[m]^TS[m]+ \sigma^2I)^{-1} u_{0,\sigma}[m] \in
+  L^2[-t_U,t_U].
+\end{equation}
+Then
+\[
+  D(T_{\sigma}[m] u)\delta m =
+  \frac{t u}{\|u_{0,\sigma}[m]\|^3}[\langle (DS[m]\delta
+  m)v_{\sigma}[m], S[m]u_{0,\sigma}[m]-d\rangle
+\]
+\[+
+  \langle S[m]v_{\sigma}[m], (DS[m]\delta m) u_{0,\sigma}[m] \rangle ]
+\]
+and
+\[
+  \langle D(T_{\sigma}[m]u)\delta m, T_{\sigma}[m]u\rangle =
+\frac{\langle t u, tu \rangle}{\|u_{0,\sigma}[m]\|^4}
+\]
+\[
+\times [\langle (DS[m]\delta
+  m)v_{\sigma}[m], S[m]u_{0,\sigma}[m]-d\rangle
++
+  \langle S[m]v_{\sigma}[m], (DS[m]\delta m) u_{0,\sigma}[m] \rangle ]
+\]
+\[
+  = \frac {\|t u\|^2}{\|u_{0,\sigma}[m]\|^4}
+\]
+\[
+  \times [\langle DF[m]\delta m, K[v_{\sigma}[m]]^T(S[m]u_{0,\sigma}[m]-d)
++
+K[u_{0,\sigma}[m]]^T S[m]v_{\sigma}[m]\rangle ]
+\]
+
+\[
+  = \langle DF[m]\delta m, W(RT_{\sigma}u,2,0)W(Ru_{0,\sigma}[m],-2,0)
 \]
@@
   \label{eqn:penderiv}
-  \times  \|T_{\sigma}[m]u\|^2\| u_{0,\sigma}[m]\|^{-2}.
+  \times ( K[v_{\sigma}[m]]^T(S[m]u_{0,\sigma}[m]-d)
++
+K[u_{0,\sigma}[m]]^T S[m]v_{\sigma}[m])\rangle 
 \end{equation}
 
-The first term on the right-hand side of the derivative \ref{eqn:basederiv} is
+Combine equations \ref{eqn:resderiv} and \ref{eqn:penderiv} with
+equation \ref{eqn:basederiv} to obtain
+\[
+  \nabla \tilde{J}_{\alpha,\sigma}[m] =
+  DF[m]^T\large[ K[u_{\alpha,\sigma}[m]]^T (S[m]u_{\alpha,\sigma}[m]-d)
+\]
+\[
+  + \alpha^2 W(RT_{\sigma}[m]u_{\alpha,\sigma}[m],2,0)W(Ru_{0,\sigma}[m],-2,0) (
+  K[v_{\sigma}[m]]^T(S[m]u_{0,\sigma}[m]-d)
+\]
 \begin{equation}
-  \label{eqn:firstsinglebis}
-  \langle DS[m]u, S[m]u-d\rangle = \langle DF[m]\delta m, K[u]^T(S[m]u-d) \rangle
+  \label{eqn:deriv}
++ K[u_{0,\sigma}[m]]^T S[m]v_{\sigma}[m]) \large]
 \end{equation}
-The right-hand side of \ref{eqn:firstsinglebis} can be re-writtenn as
-\begin{equation}
-  \label{eqn:resderiv1}
-  = \langle \delta m, DF[m]^TK[u]^T(S[m]-d)\rangle_M
-\end{equation}
-which identifies the part of the gradient coming from the first term in expression \ref{eqn:basederiv}, in the case of a single trace.
+with $v_{\sigma}$ defined in equation \ref{eqn:defvsigma}.
 
@@
 
-Applying the same manipulations to the penalty derivative expression \ref{eqn:penderiv}:
-\[
-  D_m \frac{\alpha^2}{2}\|T_{\sigma}[m] u\|^2\delta m 
-\]
-\[
-  =-\alpha^2\left( \langle d-S[m]u_{0,\sigma}[m], K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]] DF[m]\delta m \rangle \right.
-\]
-\[
-  \left. - \langle K[u_{0,\sigma}[m] ]DF[m]\delta m, S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]\rangle \right)
-\]
-\[
-  \times  \|T_{\sigma}[m]u\|^2\| u_{0,\sigma}[m]\|^{-2}.
-\]
-\[
-  = -\alpha^2 \langle \delta m, \|T_{\sigma}[m]u\|^2\| u_{0,\sigma}[m]\|^{-2}
-\]
-\[
-  \times DF[m]^T(K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]]^T (d-S[m]u_{0,\sigma}[m]) 
-\]
-\begin{equation}
-  \label{eqn:penderiv1}
-  -K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]) \rangle
-\end{equation}
-Combining equations \ref{eqn:resderiv1} and \ref{eqn:penderiv1} and substituting $u = u_{\alpha,\sigma}$, obtain for the gradient of $\tilde{J}_{\alpha,\sigma}$ in the case of a single trace:
-\[
-  \nabla \tilde{J}_{\alpha,\sigma}[m] = DF[m]^T\left( K[u_{\alpha,\sigma}[m]]^T(S[m]u_{\alpha,\sigma}[m]-d) - \alpha^2 \|T_{\sigma}[m]u_{\alpha,\sigma}[m]\|^2\| u_{0,\sigma}[m]\|^{-2}\right.
-\]
-\[
-  \left. \times( K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]]^T (d-S[m]u_{0,\sigma}[m]) \right.
-\]
-\begin{equation}
-  \label{eqn:singlegrad}
-\left.-K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]) \right).
-\end{equation}
+\subsection{Multiple Traces}
+The general case follows immediately from the single trace case, since
+every operator ($S[m]$, $K[...]$, $W$, $R$) ,appearing in the right-hand side of expression
+\ref{eqn:deriv} is block-diagonal (trace-by-trace), with the exception
+of $DF[m]^T$. This latter is a reduction operation, i.e. a block-row
+operation, so its expression is also the same. That is, equation
+\ref{eqn:deriv} also correctly expresses the gradient in the
+multi-trace case.
 
-\subsection{Multiple Traces}
-The general case follows immediately from the single trace case. It is
-just necessary to observe that the scalare multiplication in the
-expression \ref{eqn:singlegrad} is a trace-by-trace scaling that can
-be computed by a suitable combination of the mappings $R$ and $W$,
-defined above. The gradient is unlikely to be needed without the value
-of $\tilde{J}_{\alpha,\sigma}$, so assume that
-$T_{\sigma}[m]u_{\alpha,\sigma}$ is available, as well as the vector
-of trace norms of $Ru_{0,\sigma}[m]$. Then multiplication of a trace
-at $(\bx_r,\bx_s)$ by
-$\|T_{\sigma}[m]u_{\alpha,\sigma}[m](\bx_r,\cdot;\bx_s)\|^2\|
-u_{0,\sigma}[m](\bx_r,\cdot;\bx_s)\|^{-2}$ is precisely application of
-\[
-  W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m] u_{\alpha,\sigma}[m],2,0)
-\]
-So
-\[
-  \nabla \tilde{J}_{\alpha,\sigma}[m] = DF[m]^T\left( K[u_{\alpha,\sigma}[m]]^T(S[m]u_{\alpha,\sigma}[m]-d)\right.
-\]
-\[
-  \left.  +\alpha^2 W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{\alpha,\sigma},2,0)
-( K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]]^T (S[m]u_{0,\sigma}[m]-d) \right.
-\]
-\begin{equation}
-  \label{eqn:singlegrad}
-\left.+K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m])  \right).
-\end{equation}
-This expression suggestions the following algorithm outline:
+The gradient is unlikely to be needed without the value
+of $\tilde{J}_{\alpha,\sigma}$, so 
+$T_{\sigma}[m]u_{\alpha,\sigma}$ will have been computed and can be
+cached. The same is true of the vector of trace norms
+$Ru_{0,\sigma}[m]$. So all ingredient for application of
+$W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m] u_{\alpha,\sigma}[m],2,0)$
+
 \begin{enumerate}
@@
   \begin{enumerate}
-  \item $G_1 = K[u_{\alpha,\sigma}[m]]^Tr_{\alpha}$
-  \item $G_2 = K[v_0]^Tr_0$
-  \item $G_3 = K[u_{0,\sigma}[m]]^TS[m]v_0$
+  \item $G_1 = K[u_{\alpha,\sigma}[m]]^T r_{\alpha}$
+  \item $G_2 = K[v_{\sigma}[m]]^Tr_0$
+  \item $G_3 = K[u_{0,\sigma}[m]]^T S[m]v_{\sigma}[m]$
   \item $G_0 = G_1 + \alpha^2  W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{\alpha,\sigma}[m],2,0) (G_2 + G_3)$
@@
   \tilde{J}_{\sigma}[m] = \lim_{\alpha \rightarrow 0}
-  \frac{1}{\alpha^2}(\tilde{J}_{\alpha,\sigma}[m]-\tilde{J}_{0,\sigma})
+  \frac{1}{\alpha^2}(\tilde{J}_{\alpha,\sigma}[m]-\tilde{J}_{0,\sigma}[m])
 \]
@@
 \end{equation}
+
 \cite{Warner:16} presume that $S[m]u_{0,\sigma} = d$, which is

### 2023-12-30 repro/papers/awigrad/paper.tex
near done

@@
 The acoustic parameter fields, bulk modulus ($\kappa$) and buoyancy ($\beta$) appearing in the system \ref{eqn:awe} are assumed smooth and uniformly bounded.
-Since the speed of propagation is bounded as is the set of source locations $Q_s$ and the interval of propagation $[0,t_D]$, the values of $\kappa(\bx)$ and $\beta(\bx)$ for sufficiently lartge $|\bx|$ have no effect on the solution of the system \ref{eqn:awe}, so $\log \kappa$ and $\log \beta$ can be assumed square-integrable without loss of generality.
+Since the speed of propagation is bounded as is the set of source locations $Q_s$ and the interval of propagation $[0,t_D]$, the values of $\kappa(\bx)$ and $\beta(\bx)$ for sufficiently large $|\bx|$ have no effect on the solution of the system \ref{eqn:awe}, so $\log \kappa$ and $\log \beta$ can be assumed square-integrable without loss of generality.
 The set of feasible models $M$ is a  bounded open subset of ${\cal M} = C^{\infty}(\bR^3) \times C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3) \times L^{\infty}(\bR^3) \cap L^2(\bR^3) \times L^2(\bR^3)$.
@@
 Use the same notation for the similar function with $D$ replaced by
-$U$. Note that $W$ is only well-defined if either $p_s \ge 0$ or $q>0$.
+$U$. Note that $W$ is only well-defined if either $p_s \ge 0$ or
+$q>0$.
+
+For convenience, note these obvious properties of $W$:
+\begin{itemize}
+\item[W1. ] $W(\cdot,\cdot,0)$ is block-scalar, hence commutes with any
+  block-diagonal operator;
+\item[W2. ] $W(q,p_{s,1},p_{t,1})W(q,p_{s,2},p_{t,2}) = W(q,p_{s,1}+p_{s,2},p_{t,1}+p_{t,2})$;
+\item[W3. ] $W(q_1,p_s,p_{t,1})W(q_2,p_s,p_{t,2}) = W(q_1,q_2,
+  p_s,p_{t,1}+p_{t,2})$
+\end{itemize}
 
@@
 \[
-D_m\tilde{J}_{\alpha,\sigma}[m]\delta m D_mJ[u,m]\delta m|_{u =
-  u_{\alpha,\sigma}[m,d]}
+D_m\tilde{J}_{\alpha,\sigma}[m]\delta m = D_mJ[u,m]\delta m|_{u =u_{\alpha,\sigma}[m,d]}
+\]
+\[
+  = \langle D_m(S[m]u)\delta m_{u =  u_{\alpha,\sigma}[m]} (S[m]u_{\alpha,\sigma}[m]-d) \rangle
 \]
 \begin{equation}
-= \langle (D_mS[m]u\delta m)_{u =
-  u_{\alpha,\sigma}[m]} (S[m]u_{\alpha,\sigma}[m]-d \rangle + \alpha^2 \langle D_m(T_{\sigma}[m]u)\delta m)_{u=u_{\alpha,\sigma}[m]},T_{\sigma}[m]u_{\alpha,\sigma}[m]\rangle_U.
++ \alpha^2 \langle D_m(T_{\sigma}[m]u)\delta m|_{u=u_{\alpha,\sigma}[m]},T_{\sigma}[m]u_{\alpha,\sigma}[m]\rangle_U.
 \label{eqn:basederiv}
@@
 \[
-  D_mu_{0,\sigma}[m] \delta m = (S[m]^TS[m] + \sigma^2I)^{-1}
+  D_mu_{0,\sigma}[m] \delta m = -(S[m]^TS[m] + \sigma^2I)^{-1}
 \]
@@
   \label{eqn:blockderiv}
-  \times ((DS[m]^{T}\delta m) (d - S[m]u_{0,\sigma}[m]) - S[m]^T(DS[m]\delta) m u_{0,\sigma}[m])
+  \times ((DS[m]^{T}\delta m) (S[m]u_{0,\sigma}[m]-d) + S[m]^T(DS[m]\delta m) u_{0,\sigma}[m])
 \end{equation}
@@
 \begin{equation}
-  \label{eqn:defvsigma}
-  v_{\sigma} = (S[m]^TS[m]+ \sigma^2I)^{-1} u_{0,\sigma}[m] \in
+  \label{eqn:defv0tilde}
+  \tilde{v}_{0,\sigma}[m] = (S[m]^TS[m]+ \sigma^2I)^{-1} u_{0,\sigma}[m] \in
   L^2[-t_U,t_U].
@@
   \frac{t u}{\|u_{0,\sigma}[m]\|^3}[\langle (DS[m]\delta
-  m)v_{\sigma}[m], S[m]u_{0,\sigma}[m]-d\rangle
+  m)\tilde{v}_{0,\sigma}[m], S[m]u_{0,\sigma}[m]-d\rangle
 \]
 \[+
-  \langle S[m]v_{\sigma}[m], (DS[m]\delta m) u_{0,\sigma}[m] \rangle ]
+  \langle S[m]\tilde{v}_{0,\sigma}[m], (DS[m]\delta m) u_{0,\sigma}[m] \rangle ]
 \]
@@
 \times [\langle (DS[m]\delta
-  m)v_{\sigma}[m], S[m]u_{0,\sigma}[m]-d\rangle
+  m)\tilde{v}_{0,\sigma}[m], S[m]u_{0,\sigma}[m]-d\rangle
 +
-  \langle S[m]v_{\sigma}[m], (DS[m]\delta m) u_{0,\sigma}[m] \rangle ]
+  \langle S[m]\tilde{v}_{0,\sigma}[m], (DS[m]\delta m) u_{0,\sigma}[m] \rangle ]
 \]
@@
 \[
-  \times [\langle DF[m]\delta m, K[v_{\sigma}[m]]^T(S[m]u_{0,\sigma}[m]-d)
+  \times [\langle DF[m]\delta m, K[\tilde{v}_{0,\sigma}[m]]^T(S[m]u_{0,\sigma}[m]-d)
 +
-K[u_{0,\sigma}[m]]^T S[m]v_{\sigma}[m]\rangle ]
+K[u_{0,\sigma}[m]]^T S[m]\tilde{v}_{0,\sigma}[m]\rangle ]
 \]
@@
 \[
-  = \langle DF[m]\delta m, W(RT_{\sigma}u,2,0)W(Ru_{0,\sigma}[m],-2,0)
+  = \langle DF[m]\delta m, W(R(W(Ru_{0,\sigma}[m],-2,1)u),2,0)
 \]
@@
   \label{eqn:penderiv}
-  \times ( K[v_{\sigma}[m]]^T(S[m]u_{0,\sigma}[m]-d)
+  \times ( K[\tilde{v}_{0,\sigma}[m]]^T(S[m]u_{0,\sigma}[m]-d)
 +
-K[u_{0,\sigma}[m]]^T S[m]v_{\sigma}[m])\rangle 
+K[u_{0,\sigma}[m]]^T S[m]\tilde{v}_{0,\sigma}[m])\rangle
 \end{equation}
 
+It will turn out to be convenient to take advantage of the {\em
+  block-scalar} property of the operator
+$W(R(W(Ru_{0,\sigma}[m],-2,1)u),2,0)$. The other operators on the RHS
+of equation \ref{eqn:penderiv} are all block-diagonal, and the
+right-hand side is linear in $\tilde{v}_{0,\sigma}[m]$. Thus the
+$W...$ operator can be combined with the definition of $\tilde{v}...$:
+define
+\[
+  v_{0,\sigma}[m,u]=W(R(W(Ru_{0,\sigma}[m],-2,1)u),2,0) \tilde{v}_{0,\sigma}[m]
+\]
+to be the solution of the system
+\begin{equation}
+  \label{eqn:defv0}
+  (S[m]^TS[m]+ \sigma^2I) v_{0,\sigma}[m,u] = W(R(W(Ru_{0,\sigma}[m],-2,1)u),2,0) u_{0,\sigma}[m].
+\end{equation}
+
+Then equation \ref{eqn:penderiv} can be re-written as
+\[
+  \langle D(T_{\sigma}[m]u)\delta m, T_{\sigma}[m]u\rangle = \langle DF[m]\delta m, 
+\]
+\begin{equation}
+  \label{eqn:penderiv}
+   K[v_{0,\sigma}[m,u]]^T(S[m]u_{0,\sigma}[m]-d)
++
+K[u_{0,\sigma}[m]]^T S[m]v_{0,\sigma}[m,u]\rangle
+\end{equation}
 Combine equations \ref{eqn:resderiv} and \ref{eqn:penderiv} with
@@
   \nabla \tilde{J}_{\alpha,\sigma}[m] =
-  DF[m]^T\large[ K[u_{\alpha,\sigma}[m]]^T (S[m]u_{\alpha,\sigma}[m]-d)
-\]
-\[
-  + \alpha^2 W(RT_{\sigma}[m]u_{\alpha,\sigma}[m],2,0)W(Ru_{0,\sigma}[m],-2,0) (
-  K[v_{\sigma}[m]]^T(S[m]u_{0,\sigma}[m]-d)
+  DF[m]^T\large\{ K[u_{\alpha,\sigma}[m]]^T (S[m]u_{\alpha,\sigma}[m]-d)
 \]
@@
   \label{eqn:deriv}
-+ K[u_{0,\sigma}[m]]^T S[m]v_{\sigma}[m]) \large]
+  + \alpha^2 (
+  K[v_{0,\sigma}[m, u_{\alpha,\sigma}[m]]]^T(S[m]u_{0,\sigma}[m]-d)
++ K[u_{0,\sigma}[m]]^T S[m]v_{0,\sigma}[m,u_{\alpha,\sigma}[m]]) \large\}.
 \end{equation}
-with $v_{\sigma}$ defined in equation \ref{eqn:defvsigma}.
 
@@
 The gradient is unlikely to be needed without the value
-of $\tilde{J}_{\alpha,\sigma}$, so 
-$T_{\sigma}[m]u_{\alpha,\sigma}$ will have been computed and can be
-cached. The same is true of the vector of trace norms
-$Ru_{0,\sigma}[m]$. So all ingredient for application of
-$W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m] u_{\alpha,\sigma}[m],2,0)$
+of $\tilde{J}_{\alpha,\sigma}$, so the vector of trace norms
+$Ru_{0,\sigma}[m]$ will have been computed, as it is a key ingredient
+in the definition of the penalty
+$T_{\sigma}[m]u_{\alpha,\sigma}$, and can be
+cached.
 
@@
   $u_{0,\sigma}[m]$ by solving the normal equations \ref{eqn:normal}
-  for $\alpha=0$, and $T_{\sigma}[m]=W(Ru_{0,\sigma},-1,1)$ (this has
-  likely already been done to compute the value of $\tilde{J}_{\alpha,\sigma}$;
-\item use $S[m]$, $T_{\sigma}[m]$ to compute $u_{\alpha,\sigma}[m]$ by
+  for $\alpha=0$, also $Ru_{0,\sigma}$ (this has
+  likely already been done to compute the value of $\tilde{J}_{\alpha,\sigma}$);
+\item compute $u_{\alpha,\sigma}[m]$ by
   solving the normal equations \ref{eqn:normal}, and record
-  $T_{\sigma}[m]u_{\alpha,\sigma}[m]$ as a by-product;
-\item compute $v_{0\sigma} \in U$ by solving $(S[m]^TS[m] + \sigma^2 I)v_{0,\sigma} = u_{0,\sigma}[m]$
+  $R(W(Ru_{0,\sigma}[m],-2,1)u_{\alpha,\sigma}[m])$ as a by-product;
+\item compute $v_{0\sigma}[m, u_{\alpha,\sigma}[m]] \in U$ by solving the linear system
+  \ref{eqn:defv0};
 \item compute $r_0 = S[m]u_{0,\sigma}[m]-d,\,r_{\alpha} = S[m]u_{\alpha,\sigma}[m]-d$ (these may be by-products of steps 1 and 3;
@@
   \item $G_1 = K[u_{\alpha,\sigma}[m]]^T r_{\alpha}$
-  \item $G_2 = K[v_{\sigma}[m]]^Tr_0$
-  \item $G_3 = K[u_{0,\sigma}[m]]^T S[m]v_{\sigma}[m]$
-  \item $G_0 = G_1 + \alpha^2  W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{\alpha,\sigma}[m],2,0) (G_2 + G_3)$
+  \item $G_2 = K[v_{0,\sigma}[m, u_{\alpha,\sigma}[m]]]^Tr_0$
+  \item $G_3 = K[u_{0,\sigma}[m]]^T S[m]v_{0,\sigma}[m, u_{\alpha,\sigma}[m]]$
+  \item $G_0 = G_1 + \alpha^2 (G_2 + G_3)$
   \end{enumerate}
@@
 the limit
-\[
+\begin{equation}
+  \label{eqn:wglim}
   \tilde{J}_{\sigma}[m] = \lim_{\alpha \rightarrow 0}
   \frac{1}{\alpha^2}(\tilde{J}_{\alpha,\sigma}[m]-\tilde{J}_{0,\sigma}[m])
-\]
+\end{equation}
 \begin{equation}
@@
 
-\cite{Warner:16} presume that $S[m]u_{0,\sigma} = d$, which is
-manifestly not the case for $\sigma > 0$, however for small $\sigma$
-the error is presumably negligible. Neglecting it, the gradient becomes
+In the following paragraphs, I will compute the $m$ derivative of
+$\tilde{J}_{\sigma}$ directly from the right-hand side of the
+definition \ref{eqn:wg}. Then I will compute it as the limit of
+derivatives as suggested in the identity \ref{eqn:wglim}, and show
+that these are the same.  A by-product of this comparison is the
+observation that the limit \ref{eqn:wglim} is in the sense of
+$C^1(L^{\infty}(\bR^3) \times L^{\infty}(\bR^3))$.
+
+The derivative of the expression on the right-hand side of
+\ref{eqn:wg} is the sum of two terms, the first of which is given by
+\ref{eqn:penderiv} with $u = u_{0,\sigma}[m]$. The second is
 \[
-  \nabla \tilde{J}_{\sigma}[m] = DF[m]^T
-    W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{0,\sigma}[m],2,0)
+  \langle T_{\sigma}[m]u_{0,\sigma}[m], T_{\sigma}[m]D_m
+  u_{0,\sigma}[m]\delta m \rangle
+\]
+which from identity \ref{eqn:blockderiv} is
+\[
+  = -\langle  T_{\sigma}[m]^TT_{\sigma}[m]u_{0,\sigma}[m], (S[m]^TS[m] + \sigma^2I)^{-1}
+\]
+\[
+  \times ((DS[m]^{T}\delta m) (S[m]u_{0,\sigma}[m]-d) +
+  S[m]^T(DS[m]\delta) m u_{0,\sigma}[m])
+\]
+Since
+$T_{\sigma}[m]^TT_{\sigma}[m] = W(Ru_{0,\sigma}[m],-2,2)$, this is
+\[
+  = -\langle (DS[m]\delta m) (S[m]^TS[m] +
+  \sigma^2I)^{-1}W(Ru_{0,\sigma}[m],-2,2)u_{0,\sigma}[m],
+  S[m]u_{0,\sigma}[m] - d \rangle
 \]
 \begin{equation}
-  \label{eqn:wggrad}
-\times K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]).
+  \label{eqn:term2}
+  - \langle S[m] (S[m]^TS[m] +
+  \sigma^2I)^{-1}W(Ru_{0,\sigma}[m],-2,2)u_{0,\sigma}[m], (DS[m]\delta
+  m)u_{0,\sigma}[m] \rangle
 \end{equation}
+Introduce the abbreviation
+\begin{equation}
+  \label{eqn:defv1}
+  v_{1,\sigma}[m] = (S[m]^TS[m] + \sigma^2I)^{-1}W(Ru_{0,\sigma}[m],-2,2)u_{0,\sigma}[m].
+\end{equation}
+Then the RHS of equation \ref{eqn:term2bis} may be re-written as
+\begin{equation}
+  \label{eqn:term2bis}
+  = -\langle DF[m]\delta m, K[v_{1,\sigma}[m]]^T (S[m]u_{0,\sigma}[m]-d) + 
+  K[u_{0,\sigma}[m]]^T S[m] v_{1,\sigma}[m] \rangle.
+\end{equation}
+Recall that we have just computed the second of two terms, the sum of
+which is the derivative of $\tilde{J}_{\sigma}[m]$. The first is
+\ref{eqn:penderiv} with $u = u_{0,\sigma}[m]$. Combine the two
+terms to obtain
+\[
+  D\tilde{J}_{\sigma}[m] \delta m = \langle DF[m]\delta m,
+\]
+\[
+  K[v_{0,\sigma}[m,u_{0,\sigma}[m]]]^T(S[m]u_{0,\sigma}[m]-d)
++
+K[u_{0,\sigma}[m]]^T S[m]v_{0,\sigma}[m,u_{0,\sigma}[m]]\rangle
+\]
+\[
+  -\langle DF[m]\delta m, K[v_{1,\sigma}[m]]^T (S[m]u_{0,\sigma}[m]-d) + 
+  K[u_{0,\sigma}[m]]^T S[m] v_{1,\sigma}[m] \rangle
+\]
+\[
+  = \langle DF[m]\delta
+  m,(K[v_{0,\sigma}[m,u_{0,\sigma}[m]]]-K[v_{1,\sigma}[m]])^T(S[m]u_{0,\sigma}[m]-d)
+\]
+\[
+  + K[u_{0,\sigma}[m]]^T S[m] (v_{0,\sigma}[m,u_{0,\sigma}[m]] -
+  v_{1,\sigma}[m]) \rangle.
+\]
+To simplify this expression further, and reduce the computational
+complexity of the derivative expression, note that the
+operator-valued function $K$ is linear in its argument, according to
+its definition \ref{eqn:convdef}. So the above is
+\begin{equation}
+  \label{eqn:wgderiv}
+  = \langle DF[m]\delta m, (K[v_{\sigma}[m]]^T(S[m]u_{0,\sigma}[m]-d) 
+  + K[u_{0,\sigma}[m]]^T S[m] v_{\sigma}[m] )\rangle.
+\end{equation}
+in which
+\[
+  v_{\sigma}[m] = v_{0,\sigma}[m,u_{0,\sigma}[m]] -  v_{1,\sigma}[m].
+\]
+The definitions \ref{eqn:defv0} and \ref{eqn:defv1} imply that
+$v_{\sigma}[m]$ is the solution of the linear system
+\begin{equation}
+  \label{eqn:defvsigma}
+  (S[m]^TS[m]+ \sigma^2I) v_{\sigma}[m] =
+  (W(R(W(Ru_{0,\sigma}[m],-2,1)u_{0,\sigma}),2,0) - W(Ru_{0,\sigma}[m],-2,2))
+  u_{0,\sigma}[m].
+\end{equation}
+That is, computation of $v_{\sigma}[m]$, hence of $D
+\tilde{J}_{\sigma}[m]\delta m$, requires the solution of only one additional
+linear system, beyond that necessary to compute the value of $\tilde{J}_{\sigma}[m]$.
+
+\noindent {\bf Remark:} The expression \ref{eqn:wgderiv} immediately
+yields a formal prescription for the gradient of
+$\tilde{J}_{\sigma}[m]$ as the output of the adjoint modeling operator
+$DF[m]^T$ applied to an ``adjoint source'' (the right-hand side of the
+inner product \ref{eqn:wgderiv}. \cite{Warner:16} point out this
+similarity with FWI, in which the adjoint source (in the simplest
+version) is the residual $F[m]-d$. Consequently, FWI implementations
+already contain a great deal of the code necessary to implement
+AWI. \cite{Warner:16} also presume that the regularization weight
+$\sigma$ is sufficiently small that the adaptive filter $u_{0,\sigma}$
+effectively fits the data. Thus $S[m]u_{0,\sigma}-d$ is negligible,
+regardless of $m$, and the first term on the right-hand side of the
+inner product \ref{eqn:wgderiv} drops out, further simplifying the AWI
+adjoint source.
+
+It remains to establish that the same result (expression
+\ref{eqn:wgderiv}) for the derivative can be derived from the
+limit \ref{eqn:wglim}. From from equation \ref{eqn:basederiv},
+\[
+  \frac{1}{\alpha^2}(D \tilde{J}_{\alpha,\sigma}[m] \delta m - D
+  \tilde{J}_{0,\sigma}[m]\delta m) =
+\]
+\[
+  = \frac{1}{\alpha^2} \left(\langle D_m(S[m]u)\delta m_{u =  u_{\alpha,\sigma}[m]}, (S[m]u_{\alpha,\sigma}[m]-d) \rangle\right.
+\]
+\[
+  - \left.\langle D_m(S[m]u)\delta m_{u =  u_{0,\sigma}[m]}, (S[m]u_{0,\sigma}[m]-d) \rangle\right)
+\]
+\[
+  +  \langle D_m(T_{\sigma}[m]u)\delta
+  m|_{u=u_{\alpha,\sigma}[m]},T_{\sigma}[m]u_{\alpha,\sigma}[m]\rangle_U
+\]
+Since the normal operator \ref{eqn:normal} is uniformly positive
+definite hence boundedly invertible uniformly in $\alpha \ge 0$,
+$u_{\alpha,\sigma}[m] \rightarrow u_{0,\sigma}[m]$ in $U$. All of the
+operators appearing in the third term are $U-$continuous, so  the
+third term tends to
+\[
+  \langle D_m(T_{\sigma}[m]u)\delta
+  m|_{u=u_{0,\sigma}[m]},T_{\sigma}[m]u_{0,\sigma}[m]\rangle_U
+\]
+which is identixal to the left-hand side of equation
+\ref{eqn:penderiv} and is the first term in the direct computation of
+the $\tilde{J}_{\sigma}$ derivative, analyzed above.
 

### 2023-12-30 repro/papers/awigrad/paper.tex
near final

@@
 
-{\bf Note:}
-$F$ is the mapping implemented (in finite difference approximation) in {\tt asg.fsbop}, with the additional constraint that the buoyancy field is fixed so that the computational version of ${\cal M}$ consists of (gridded) bulk modulus fields.
-
 Adaptive filters are collections of traces defined on a (possibly)
@@
 \end{equation}
-
-The convolution operator implemented in the module {\tt segyvc} actually implements a time-discrete version of $K$ (or $L$).
-Note that the transpose $K[u]^T$ is a restricted version of cross-correlation, implemented as the transpose of {\tt segyvc.ConvolutionOperator}.
 
@@
 
-{\bf Note:} IWAVE implements the mappings $R$ and $W$ in the form of
-the commands {\tt iwave/trace/main/rms.x} and {\tt
-  iwave/trace/main/txgain.x}. The natural python implementation of $R$
-would output the vector of trace norms as a python array (or a NumPy
-ndarray). However that would require extracting the number of traces
-from the input, and checking that the inputs are compatible. This is
-already implicitly done in the IWAVE manipulations of SU trace
-data. So {\tt iwave/trace/main/rms.x} stores the vector of trace norms
-as the 0th data sample in a set of traces, one for each trace of the
-input data: all header words are the same except for {\tt ns}, which
-is set to 1.
-
-The python functions {\tt awi.setrms} and {\tt awi.applytxgain} call
-the two IWAVE commands, manipulating the data as SU data files, rather
-than as VCL vectors. That is, these are simply functions, not {\tt
-  vcl.Function} objects. Since VCL vectors defined on a {\tt segyvc.Space}
-have access to their data files (keyword {\tt data}), these are
-convenient to use to express the action of $R$ and $W$ in AWI penalty operator,
-gradient, and so on.
-
 \section{Objectives}
 Define for $\alpha \ge 0$
-\[
+\begin{equation}
+  \label{eqn:j}
  J_{\alpha,\sigma}[u,m,d] = \frac{1}{2}(\|S[m]u - d\|_D^2 + \alpha^2\|T_\sigma[m]u\|_U^2 + \sigma^2 \|u\|^2_U).
-\]
+\end{equation}
 Note that the notation is consistent, that is, reduces to $J_{0,\sigma}$ as defined above for $\alpha=0$.
@@
 two least-squares problems, in sequence.
-
-{\bf Note:} The implementation {\tt vcalg.conjgrad} of the Conjugate Gradient algorithm for the normal equations optionally returns the residual vector (parameter {\tt e}).
-The obvious way to use this is to create a 3-component product space for the output of the block-column operator $(S[m],\alpha T[m], \sigma I)^T$, and set the right-hand side vector (parameter {\tt b})
-to $(d,0,0)^T$. Then on return one-half the norm-squared of {\tt e} is precisely the (approximate) value of $\tilde{J}_{\alpha,\sigma}[m]$.
 
@@
 \begin{equation}
-  \label{eqn:penderiv}
+  \label{eqn:penderiv1}
    K[v_{0,\sigma}[m,u]]^T(S[m]u_{0,\sigma}[m]-d)
@@
 \end{equation}
-Combine equations \ref{eqn:resderiv} and \ref{eqn:penderiv} with
+Combine equations \ref{eqn:resderiv} and \ref{eqn:penderiv1} with
 equation \ref{eqn:basederiv} to obtain
@@
   \nabla \tilde{J}_{\alpha,\sigma}[m] =
-  DF[m]^T\large\{ K[u_{\alpha,\sigma}[m]]^T (S[m]u_{\alpha,\sigma}[m]-d)
+  DF[m]^T\{ K[u_{\alpha,\sigma}[m]]^T (S[m]u_{\alpha,\sigma}[m]-d)
 \]
@@
   K[v_{0,\sigma}[m, u_{\alpha,\sigma}[m]]]^T(S[m]u_{0,\sigma}[m]-d)
-+ K[u_{0,\sigma}[m]]^T S[m]v_{0,\sigma}[m,u_{\alpha,\sigma}[m]]) \large\}.
++ K[u_{0,\sigma}[m]]^T S[m]v_{0,\sigma}[m,u_{\alpha,\sigma}[m]]) \}.
 \end{equation}
@@
 \ref{eqn:wg} is the sum of two terms, the first of which is given by
-\ref{eqn:penderiv} with $u = u_{0,\sigma}[m]$. The second is
+\ref{eqn:penderiv1} with $u = u_{0,\sigma}[m]$. The second is
 \[
@@
 which is the derivative of $\tilde{J}_{\sigma}[m]$. The first is
-\ref{eqn:penderiv} with $u = u_{0,\sigma}[m]$. Combine the two
+\ref{eqn:penderiv1} with $u = u_{0,\sigma}[m]$. Combine the two
 terms to obtain
@@
 \]
-\[
+\begin{equation}
+  \label{eqn:limdiff}
   +  \langle D_m(T_{\sigma}[m]u)\delta
   m|_{u=u_{\alpha,\sigma}[m]},T_{\sigma}[m]u_{\alpha,\sigma}[m]\rangle_U
-\]
+\end{equation}
 Since the normal operator \ref{eqn:normal} is uniformly positive
@@
 which is identixal to the left-hand side of equation
-\ref{eqn:penderiv} and is the first term in the direct computation of
+\ref{eqn:penderiv1} with $u=u_{0,\sigma}[m]$ and is the first term in the direct computation of
 the $\tilde{J}_{\sigma}$ derivative, analyzed above.
+
+It remains to understand
+\[
+\frac{1}{\alpha^2} \left(\langle D_m(S[m]u)\delta m_{u =  u_{\alpha,\sigma}[m]}, (S[m]u_{\alpha,\sigma}[m]-d) \rangle\right.
+\]
+\[
+  - \left.\langle D_m(S[m]u)\delta m_{u =  u_{0,\sigma}[m]}, (S[m]u_{0,\sigma}[m]-d) \rangle\right)
+\]
+\[
+  =\frac{1}{\alpha^2} \left(\langle D_m(S[m]u)\delta m_{u =  u_{\alpha,\sigma}[m]-u_{0,\sigma}[m]}, (S[m]u_{\alpha,\sigma}[m]-d) \rangle\right.
+\]
+\begin{equation}
+  \label{eqn:resterms}
+  + \left.\langle D_m(S[m]u)\delta m_{u =  u_{0,\sigma}[m]},
+    (S[m](u_{\alpha,\sigma}[m]-u_{0,\sigma}[m]) \rangle\right)
+\end{equation}
+Observe that the normal equation \ref{eqn:normal} implies that
+\[
+  (S[m]^TS[m] + \sigma^2 I)(u_{\alpha,\sigma}[m]-u_{0,\sigma}[m]) =
+  -\alpha^2 T_{\sigma}[m]^TT_{\sigma}[m]u_{\alpha,\sigma}[m]
+\]
+Since $u_{\alpha,\sigma}[m] \rightarrow u_{0,\sigma}$ in the sense of
+$U$ as $\alpha \rightarrow 0$, it follows from definition \ref{eqn:defv1} that
+\[
+  \frac{1}{\alpha^2 }u_{\alpha,\sigma}[m]-u_{0,\sigma}[m] \rightarrow
+  - v_{1,\sigma}[m], \, \alpha \rightarrow 0
+\]
+Writing as before $D_m(S[m]u)\delta m_{u =  ...} =
+(DS[m]\delta m)...$,  the quantity on the right-hand side of equation
+\ref{eqn:resterms}, obtain
+\[
+  \lim_{\alpha \rightarrow 0} 
+\frac{1}{\alpha^2} \left(\langle D_m(S[m]u)\delta m_{u =  u_{\alpha,\sigma}[m]}, (S[m]u_{\alpha,\sigma}[m]-d) \rangle\right.
+\]
+\[
+  - \left.\langle D_m(S[m]u)\delta m_{u =  u_{0,\sigma}[m]}, (S[m]u_{0,\sigma}[m]-d) \rangle\right)
+\]
+\[
+  = - \langle (DS[m]\delta m)v_{1,\sigma}[m], S[m]u_{0,\sigma}[m]-d
+  \rangle
+\]
+\[
+  - \langle (DS[m]\delta m) u_{0,\sigma}[m],S[m]v_{1,\sigma}[m] \rangle
+\]
+\[
+  =-\langle DF[m]\delta m, K[v_{1,\sigma}[m]]^T (S[m]u_{0,\sigma}[m]-d) + 
+  K[u_{0,\sigma}[m]]^T S[m] v_{1,\sigma}[m] \rangle.
+\]
+which is precisely the right-hand side of equation
+\ref{eqn:term2bis}. That is, the limit of the first two terms on the
+right-hand side of equation \ref{eqn:limdiff} as $\alpha \rightarrow
+0$ is exactly the second term of the direct derivative
+computation. Since we have already shown that the third term in
+\ref{eqn:limdiff} limits to
+the first term in the direct computation, a second derivation of the
+derivative $D\tilde{J}_{\sigma}[m]$ has been accomplished.
+
+We have also shown that the limit of the scaled difference of the
+derivatives is the derivative of the limit of the scaled difference,
+pointwise in $m \in M$. In fact the limit is locally uniform in a
+suitable sense, but that is a topic for another discussion.
+
+\append{Unpreconditioned AWI: Matched Source Waveform Inversion}
+Matched Source Waveform Inversion (MSWI) is a close relative to
+AWI, differing in the use of a simpler penalty operator $T$, defined
+by
+\begin{equation}
+  \label{eqn:upf}
+  Tu (\bx_r,t;\bx_s) = t u(\bx_r,t;\bx_s).
+\end{equation}
+Similarly to AWI, $m$ is estimated by minimizing
+\begin{equation}
+  \label{eqn:junp}
+ J^{\rm unp}_{\alpha,\sigma}[u,m,d] = \frac{1}{2}(\|S[m]u - d\|_D^2 + \alpha^2\|Tu\|_U^2 + \sigma^2 \|u\|^2_U).
+\end{equation}
+or equivalently
+\begin{equation}
+  \label{eqn:jtildeunp}
+ \tilde{J}^{\rm unp}_{\alpha,\sigma}[m,d] = \frac{1}{2}(\|S[m]u_{\alpha,\sigma}[m,d] - d\|_D^2 + \alpha^2\|Tu_{\alpha,\sigma}[m,d]\|_U^2 + \sigma^2 \|u_{\alpha,\sigma}\|^2_U).
+\end{equation}
+in which $u_{\alpha,\sigma}$ is the solution of the normal equation
+\begin{equation}
+  \label{eqn:normalunp}
+  (S[m]^TS[m] + \alpha^2T^TT + \sigma^2I)u_{\alpha,\sigma}[m] = S[m]^Td
+\end{equation}
+The only difference between AWI and MSWI lies in the defintiion of the
+penalty operator.
+
+MSWI was introduced in Hua Song's thesis
+\cite[]{Song:94c} using a different definition of $T$ (but also
+$m$-independent). The penalty operator defined in \ref{eqn:upf}, and
+the associated inversion algorithm, was explored by Huang and Symes
+\cite[]{HuangSymes2015SEG,HuangSymes:Geo17}, who show that at least to
+some extent it avoids the cycle-skipping pathology of FWI. Note that
+the division by $Ru_{0,\sigma}[m]$ in the definition of
+$T_{\sigma}[m]$ is responsible for the global relation between AWI and
+travel time tomography, within a limited class of problems, and this
+relation cannot be asserted for MSWI.
+
+Since the MSWI penalty operator $T$ is $m$-independent, it does not
+appear explicilly in the the gradient
+of the reduced MSWI objective $\tilde{J}^{\rm unp}_{\alpha,\sigma}$, as
+given by the VPM gradient formula:
+\[
+  \nabla \tilde{J}^{\rm unp}_{\alpha,\sigma}[m] =
+  D(S[m]u)^T_{u=u_{\alpha,\sigma}[m]}(S[m]u_{\alpha,\sigma}[m]-d)
+\]
+\begin{equation}
+  \label{eqn:tildejunpgrad}
+  = DF[m]^T(K[u_{\alpha,\sigma}[m]]^T(S[m]u_{\alpha,\sigma}[m]-d).
+\end{equation}
+
+\append{Remarks on VCL Implementation}
+\noindent {\bf 1.} $F$ is the mapping implemented (in finite difference approximation) in {\tt asg.fsbop}, with the additional constraint that the buoyancy field is fixed so that the computational version of ${\cal M}$ consists of (gridded) bulk modulus fields.
+
+\noindent {\bf 2.} The convolution operator implemented in the module {\tt segyvc}
+actually implements a time-discrete version of $K$ (or $L$). Note that
+the transpose $K[u]^T$ is a restricted version of cross-correlation,
+implemented as the transpose of {\tt segyvc.ConvolutionOperator}.
+
+\noindent {\bf 3.} IWAVE implements the mappings $R$ and $W$ in the form of
+the commands \\{\tt iwave/trace/main/rms.x} and {\tt
+  iwave/trace/main/txgain.x}. The natural python implementation of $R$
+would output the vector of trace norms as a python array (or a NumPy
+ndarray). However that would require extracting the number of traces
+from the input, and checking that the inputs are compatible. This is
+already implicitly done in the IWAVE manipulations of SU trace
+data. So {\tt iwave/trace/main/rms.x} stores the vector of trace norms
+as the 0th data sample in a set of traces, one for each trace of the
+input data: all header words are the same except for {\tt ns}, set to 1.
+
+\noindent {\bf 4.} The python functions {\tt awi.setrms} and {\tt awi.applytxgain} call
+the two IWAVE commands, manipulating the data as SU data files, rather
+than as VCL vectors. That is, these are simply functions, not {\tt
+  vcl.Function} objects. Since VCL vectors defined on a {\tt segyvc.Space}
+have access to their data files (keyword {\tt data}), these are
+convenient to use to express the action of $R$ and $W$ in AWI penalty operator,
+gradient, and so on.
+
+\noindent {\bf 5.} The implementation {\tt vcalg.conjgrad} of the Conjugate Gradient algorithm for the normal equations optionally returns the residual vector (parameter {\tt e}).
+The obvious way to use this is to create a 3-component product space for the output of the block-column operator $(S[m],\alpha T[m], \sigma I)^T$, and set the right-hand side vector (parameter {\tt b})
+to $(d,0,0)^T$. Then on return one-half the norm-squared of {\tt e} is precisely the (approximate) value of $\tilde{J}_{\alpha,\sigma}[m]$.
 

### 2024-01-15 repro/papers/vpm/paper.tex
initial

\title{Notes on the Variable Projection Method}
\author{William W. Symes}

\begin{abstract}
The variable projection reduction of a separable least squares objective function has a gradient of simple form. Stationary points of the reduced function and the original, un-reduced function are in 1-1 correspondence. An approximate Hessian analogous to the Gauss-Newton Hessian of nonlinear least squares guarantees a non-increase direction and involves less computation than the full Hessian.
\end{abstract}

\section{Introduction}
The Variable Projection Method (VPM) is a useful algorithm for solving separable least squares problems, that is, those whose objectives are linear in some of their unknowns. It uses an inner solution or reduction of the linear unknowns to eliminate them, leaving a function of the nonlinear unknowns to be minimied. The utility of VPM rests largely on the simplicitly of the derivative, or gradient, of the reduced function: it is the same as the gradient of the original, unreduced function, evaluated with the linear variables fixed at their inner solution value. Furthermore, stationary points of the reduced function are in 1-1 correspondence (under certain widely-met conditions) with stationary points of the unreduced function. These observation due to Golub and Pereyra underlie the widespread popularity of VPM  \cite[]{GolubPereyra:73,GolubPereyra:03}.

VPM has been widely used in conjunction with generalized secant (quasi-Newton) methods such as Limited Memory Broyden-Fletcher-Goldfarb-Shanno \cite[]{NocedalWright} in application to scientific and engineering separable least-squares problems, often without explicit acknowledgement that VPM is involved or with drastic approximation to the reduction \cite[]{Ghattas:IP25,LeeuwenHerrmann:16,vanLeeuwenMulder:09,Warner:16}. Other approximations to Newton's method in the VPM context are less often found, since the Hessian computation is relatively involved compared to the gradient. However \cite{Kaufman:75} developed a simplified Hessian, with roughly the same relation to the full Hessian as the Gauss-Newton Hessian to the full Hessian for ordinary nonlinear least squares,

In this note, I review the construction of the reduced gradient, with somewhat more care than is common in the literature. Then I present a complete derivation of the Kaufman Hessian. I end with a comment on the use of this structure in a Krylov-Newton context, in which all of the inner linear systems are solved by iterative methods such as conjugate gradients.

\section{The Variable Projection Objective}
The variable projection method is a minimization algorithm for a scalar function on a product space $f:X \oplus W \rightarrow Y$ of class $C^2$. I will assume that $f$ is defined on the whole product space; the refinements necessary to accommodate constraints on $x$ are similar to those needed for solution of other nonlinear optimization problems. While not strictly necessary, I will also assume that $f$ is quadratic in the second variable. Put another way, $f(x,w) = 0.5*\|A(x)w-b\|^2$, where the values of $A$ are linear operators $: W \rightarrow Y$. Further, it's usually assumed that $A(x)$ is of full column rank (or coercive, in the infinite dimensional case), so that that for each $x$, there is a unique minimiser $\tilde{w}(x)$ of $w \mapsto f(x,w)$, the solution of the normal equation: $A(x)^T(A(x)\tilde{w}(x) - b)=0$. Define the ``variable projection (VP) reduction'' $\tilde{f}$ by 
$$
\tilde{f}(x) = 0.5*\|A(x)\tilde{w}(x)-b\|^2= \min_w f(x,w)
$$
Since $A(x)$ is assumed coercive for every $x \in X$, $A(x)^TA(x)$ is invertible, and $\tilde{w}(x) = (A(x)^TA(x))^{-1}A(x)^Tb$. So
$$ 
\tilde{f}(x) = 0.5*\|(A(x)(A(x)^TA(x))^{-1}A(x)^T - I)b\|^2
$$
The operator in parenthesis projects $Y$ onto the orthocomplement of the range of $A(x)$: call it $P(x)$. That is,
$$
P(x) = I-A(x)(A(x)^TA(x))^{-1}A(x)^T
$$
and
$$
\tilde{f}(x) = 0.5*\|P(x)b\|^2
$$
So the reduced objective is half the length squared of the projection of the data vector $b$ onto the orthocomplement of the range of $A(x)$, which of course depends on $x$. That fact accounts for the name "variable projection".

\section{Gradient: the Observation of Golub and Peyreyra}

One of the main results of the Golub and Pereyra 1973 paper is that $x$ is a stationary point of $\tilde{f}$ if and only if $(x,\tilde{w}(x))$ is a stationary point of $f$. It's worth spelling out the argument because it highlights several important points about the VP reduction.

Note that $f$ is differentiable, if $(x,w)\mapsto A(x)w$ is differentiable, which I will assume. Suppose $s \in X$. Then the directional derivative of $\tilde{w}(x)=(A(x)^TA(x))^{-1}A(x)^Tb$ at $x$ in direction $s$ is 
$$
\frac{d}{dt}((A(x+ts)^TA(x+ts))^{-1}A(x+ts)^Tb)|_{t=0}
$$
$$
=-(A(x)^TA(x))^{-1}\frac{d}{dt}(A(x+ts)^TA(x+ts)(A(x)^TA(x))^{-1}A(x)^Tb)|_{t=0}
$$
$$
+ (A(x)^TA(x))^{-1}\frac{d}{dt}(A(x+ts)^Tb)|_{t=0}
$$
and in particular $\tilde{w}$ is differentiable. Therefore
$$
\frac{d}{dt}\tilde{f}(x+ts)|_{t=0} = \left\langle\frac{d}{dt}A(x+ts)\tilde{w}(x), A(x)\tilde{w}(x)-b\right\rangle + 
$$
$$
0.5\left\langle A(x)\frac{d}{dt}\tilde{w}(x+ts),A(x)\tilde{w}-b\right\rangle
$$
Since the normal equation is equivalent to the assertion that the residual $A(x)\tilde{w}(x)-b$ is orthogonal to the range of $A(x)$, the last term vanishes.

The first term can be re-written as the directional derivative of $f(x,w)$ for fixed $w=\tilde{w}(x)$, that is,
$$
\frac{d}{dt}\tilde{f}(x+ts)|_{t=0} = \frac{d}{dt}f(x+ts,w)|_{t=0,w=\tilde{w}(x)}.
$$
So $x$ is a stationary point of $\tilde{f}$ if and only if the directional derivative of $f$ at $(x,\tilde{w}(x))$ in all directions $(s,0)$ is zero. But the directional derivative of $f$ at $(x,\tilde{w}(x))$ in all directions $(0,\delta w)$ is also zero - that is the definition of $\tilde{w}(x)$. Since the directional derivative is linear in the direction, the directional derivative of $f$ at $(x,\tilde{w}(x))$ is zero in all directions, that is, $(x,\tilde{w}(x))$ is a stationary point of $f$, if and only if $x$ is a stationary point of $\tilde{f}$.

The derivative of the linear-operator-value function $A$ is naturally a bilinear-operator-valued function, since it's linear in the argument $w$ and in the direction $s$ separately. Call it $DA$:
$$
\frac{d}{dt}A(x+ts)w|_{t=0} = DA(x)(w,s)
$$ 
In terms of $DA$, the directional derivative of $\tilde{f}$ is
$$
\frac{d}{dt}\tilde{f}(x+ts)|_{t=0} = \left\langle DA(x)(\tilde{w}(x),s), A(x)\tilde{w}(x)-b\right\rangle 
$$
The expression on the right is the same as the derivative of the function
$$
x \mapsto f(x,w) = 0.5*\|A(x)w-b\|^2,
$$ 
evaluated at $w=\tilde{w}(x)$, that is, $f(x,w)$ {\em for fixed w'}. 
The gradient of $\tilde{f}$ is the Riesz representer of the directional derivative:
$$
\langle s, \mbox{grad} \tilde{f}(x)\rangle = \frac{d}{dt}\tilde{f}(x+ts)|_{t=0}
$$
$$
= \left\langle s, DA(x)^*(\tilde{w}(x),A(x)\tilde{w}(x)-b)\right\rangle 
$$
in which $DA(x)^*$ denotes the {\em partial adjoint} in the second (nonlinear) argument, defined by
$$
\langle s, DA(x)^*(w,y) \rangle = \langle DA(x)(w,s),y\rangle
$$
That is, $y \mapsto DA(x)^*(w,y)$ is the adjoint of the map $s \mapsto DA(x)(w,s)$, the latter being the derivative of $x \mapsto A(x)w$.

Thus
$$
\mbox{grad} \tilde{f}(x) = DA(x)^*(\tilde{w}(x),A(x)\tilde{w}(x)-b).
$$

In fact, this is also the gradient of a least-squares objective.
For a fixed choice $w \in W$, define $F_w(x) = A(x)w$. Then $DA(x)^*(w,y) = DF_w(x)^Ty$. Moreover, if $f_w(x) = 0.5*\|F_w(x)-b\|^2$, then 
$$
\mbox{grad} f_w(x)|_{w=\tilde{w}(x)} =  \mbox{grad}\tilde{f}(x).
$$
From the preceding section,
$$
\mbox{grad} f_w(x) = DF_w(x)^T(F_w(x)-b).
$$ 
Therefore computing the gradient of the VP reduction can be accomplished by combining a computation of the gradient of a nonlinear least-squares objective with a solution of the normal equation.

What's more, $x$ is a stationary point of the VPM objective $\tilde{f}$ if and only if $(x,\tilde{w})$ is a stationary point of the original objective $f$.

\subsection{Computing the gradient}
To compute the gradient of $\mbox{grad} \tilde{f}(x)$,

1. Calculate the minimizer 
\begin{equation}
  \label{eqn:g1}
\tilde{w}(x) = \mbox{argmin}_w 0.5*|A(x)w-b|^2
\end{equation}
and the optimal residual
\begin{equation}
  \label{eqn:g2}
  \tilde{r}(x) = A(x)\tilde{w}(x) - b.
\end{equation}

2.  Then
$$
\mbox{grad} \tilde{f}(x) = D_x(F(x)w)^T_{w=\tilde{w}(x)} \tilde{r}(x).
$$
\begin{equation}
  \label{eqn:g3}
= DA(x)^*(\tilde{w}(x),\tilde{r}(x))
\end{equation}

\noindent {\bf Notes:}
1. Note again that the gradient is identical to the gradient of the fixed-$w$ nonlinear least squares objective $\mbox{grad}_{x} f(x,w)$ with $w$ set equal to $\tilde{w}(x)$.

2. For large-scale problems, iterative solutions of the minimization \ref{eqn:g1} is likely imperative, hence $\tilde{w}(x)$ will be an approximation.

3. Software for least-squares problems often returns the residual as in equation \ref{eqn:g2} as a by-product of the minimization \ref{eqn:g1}.

\section{The Kaufman Hessian Approximation}

Of course $\tilde{f}$ is itself a nonlinear least squares objective: if you define $F(x)=A(x)\tilde{w}(x)$, then 
$$
\tilde{f}(x) = 0.5*\|F(x)-b\|^2.
$$
so it is natural to use the Gauss-Newton algorithm to minimize $\tilde{f}$. The Gauss-Newton step $s$ solves $DF(x)^T(DF(x)s-(F(x)-b))=DF(x)^TDF(x)s+\mbox{grad}\tilde{f}(x)=0$. This is a simplification over the Newton step, but for the special case of the VP reduction can be simplified still further.
$$
DF(x)s = D(A(x)\tilde{w}(x))s = DA(x)(\tilde{w}(x),s) + A(x)D\tilde{w}(x)s
$$
From the differentiability analysis of $\tilde{w}$,
$$
D\tilde{w}(x)s = 
$$
$$
=-(A(x)^TA(x))^{-1}(DA(x)^T(A(x)((A(x)^TA(x))^{-1}A(x)^Tb),s)
$$
$$
+ A(x)^T DA(x)((A(x)^TA(x)^{-1}A(x)^Tb,s) + (A(x)^TA(x)^{-1}DA(x)^T(b,s)|
$$
$$
= -(A(x)^TA(x))^{-1}[DA(x)^T(A(x)\tilde{w}(x),s) + A(x)^TDA(x)(\tilde{w}(x),s)]
$$
$$
+ (A(x)^TA(x))^{-1}DA(x)^T(b,s)
$$
So 
$$
DF(x)s = DA(x)(\tilde{w},s) - A(x) (A(x)^TA(x))^{-1}
$$
$$
\times [DA(x)^T(A(x)\tilde{w}(x),s) + A(x)^TDA(x)(\tilde{w}(x),s)] + A(x)(A(x)^TA(x))^{-1}DA(x)^T(b,s)
$$
$$
= (I-A(x)(A(x)^TA(x))^{-1}A(x)^T)DA(\tilde{w}(x),s)
$$
$$
+ A(x)(A(x)^TA(x))^{-1}DA(x)^T(b-A(x)\tilde{w}(x),s)
$$
The second term has the residual $b-A(x)\tilde{w}(x)=b-F(x)$ as the first argument of the bilinear operator $DA(x)^T$. Kaufman first pointed this out in 1974, and proposed that this term be dropped with the same justification as underlies the transition from Newton to Gauss-Newton: that is, if the residual is small (nearly noise-free data and close to the solution), this term should be negligible. Accepting this proposal, obtain 
$$
DF(x)s \approx (I-A(x)(A(x)^TA(x))^{-1}A(x)^T)DA(x)(\tilde{w}(x),s)
$$
$$
= P(x)DA(x)(\tilde{w}(x),s)
$$
where $P(x)=I-A(x)(A(x)^TA(x))^{-1}A(x)^T$ is the projection of $Y$ onto the orthocomplement of the range of $A(x)$, introduced earlier.
Since $P(x)$ is a projection, it is symmetric, positive semi-definite, and idempotent, that is $P(x)^TP(x)=P(x)^2=P(x)$. Thus the Gauss-Newton operator is approximately
$$
DF(x)^TDF(x)s \approx \mbox{Hess}_{VP}\tilde{f}(x)s = DA(x)^*(\tilde{w}(x),P(x)DA(x)(\tilde{w}(x),s)).
$$
The right-hand side of this equation is the Kaufman Hessian approximation.

The solution $s$ of the modified Newton step equation $\mbox{Hess}_{VP}\tilde {f}(x)s=-\mbox{grad}\tilde{f}(x)$ is a descent (or at least non-ascent) direction for $\tilde{f}$:
$$
\langle \mbox{grad}\tilde{f}(x), s \rangle 
= -\langle \mbox{Hess}_{VP}\tilde{f}(x)s, s\rangle 
$$
$$
=-\langle DA(x)^*(\tilde{w}(x),P(x)DA(x)(\tilde{w}(x),s)),s\rangle
$$
$$
= - \langle DA(x)(\tilde{w}(x),s),P(x)DA(x)(\tilde{w}(x),s)\rangle
$$
$$
= -\|P(x)DA(x)(\tilde{w}(x),s)\|^2 \le 0
$$
since $P(x)$ is a projector.

\subsection{Computing the Kaufman Hessian}

To calculate the Kaufman Hessian approximation $\mbox{Hess}_{VP}\tilde{f}(x)s$ for a direction vector $s$ at position $x$,

1. Calculate the minimizer $\tilde{w}(x)$ and residual $\tilde{r}(x)$ per equations \ref{eqn:g1} and \ref{eqn:g2};

2. Compute the auxiliary minimizer
\begin{equation}
  \label{eqn:h1}
\tilde{v}(x,s) = \mbox{argmin}_v 0.5*|A(x)v-DA(x)(\tilde{w}(x),s)|^2
\end{equation}
and auxiliary residual
\begin{equation}
  \label{eqn:h2}
  \tilde{q}(x,s) = A(x)\tilde{v}(x,s) - DA(x)(\tilde{w}(x),s).
\end{equation}

3. Then
\begin{equation}
  \label{eqn:h3}
  \mbox{Hess}_{VP}\tilde{f}(x)s = DA(x)^*(\tilde{w}(x),A(x)\tilde{q}(x,s)).
\end{equation}

\section{Krylov-Newton optimization of VP reduced objectives}

The major costs of (approximate) Newton optimization are
\begin{enumerate}
\item $A(x)$ - cost of construction ($C_A$)
\item $w \mapsto A(x)w$, $r \mapsto A(x)^T r$ - cost of application ($C_w$)
\item $DA(x,w)s$ - cost of application ($C_{fwd}$)
\item $DA^*(x,w)q$ - cost of application ($C_{adj}$)
\item number of Newton iterations ($N_N$)
\item for direct inner minimization (equations \ref{eqn:g1}, \ref{eqn:h1}), cost of solve ($C_S$)
\item for iterative inner minimization (equations \ref{eqn:g1}, \ref{eqn:h1}), average number of iterations required ($N_I$).
\end{enumerate}
The cost of gradient (and objective value) compuation is
\[
  C_g = C_A + 2 C_w +C_S + C_{adj}
\]
for direct inner solves, or
\[
   C_g = C_A + 2 (1 + N_I)C_w + C_{adj}
 \]
for iterative inner solves. Similarly, the cost of Hessian computation (assuming that the first step in that computation has been taken in computing the gradient) is
\[
  C_h =  2 C_w +C_S + C_{fwd} + C_{adj}
\]
for direct inner solves, or
\[
   C_h = 2 (1 + N_I)C_w) + C_{fwd} + C_{adj}
 \]
for iterative inner solves. In either case, the cost of Newton optimization is roughly
\[
  N_N (C_g + C_h)
\]
The literature has mostly discussed the cost based on direct solution of the Newton equation. For iterative (Newton-Krylov, for instance) solution, the cost is dominated by the inner iterations for each outer iteration, proportional to $N_NN_I$. This is however roughly similar to Newton-Krylov for nonlinear least squares, which has been used successfully for solution of simulation-driven optimization problems for a long time \cite[]{Ghattas:IP25}.
\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2024-01-17 repro/papers/vpm/paper.tex
first draft final

@@
 \section{Introduction}
-The Variable Projection Method (VPM) is a useful algorithm for solving separable least squares problems, that is, those whose objectives are linear in some of their unknowns.
-It uses an inner solution or reduction of the linear unknowns to eliminate them, leaving a function of the nonlinear unknowns to be minimied.
-The utility of VPM rests largely on the simplicitly of the derivative, or gradient, of the reduced function: it is the same as the gradient of the original, unreduced function, evaluated with the linear variables fixed at their inner solution value.
+The Variable Projection Method (VPM) is a useful algorithm for solving separable least squares problems, that is, those whose objectives $f(x,w)$ are quadratic in some of their unknowns, say $w$, and (possibly) non-quadratic in others, say $x$.
+A common approach to minimizing such objectives is {\em alternating direction minimiazation}.
+This iterative method starts with an initial guess $(x_0,w_0)$ at the solution, and constructs a sequence $\{(x_k,w_k)\}$, guaranteed to converge to a stationary point if it converges (to anything).
+The sequence is constructed as follows: given a current guess $(x_k,w_k)$ at the solution, find $w_{k+1}$ by minimizing $f(x_k,w)$ over $w$, then find $x_{k+1}$ by minimizing $f(x,w_{k+1})$ over $x$.
+This approach is conceptually simple and makes use of whatever might be known about the two separate minimization problems (over $w$, respectively $x$), but tends to be quite inefficient.
+That is, it can require many more function and gradient evaluations to achieve a good approximation of a stationary point that would Newton's method (say) applied directly to $f(x,w)$.
+
+Variable projection is an alternative approach roughly as rapidly convergent as Newton's method but uses solutions of the (computationally smaller) $w$ and $x$ subproblems.
+It takes advantage of special algorithms for quadratic minimization to eliminate the quadratic unknowns ($w$), leaving a function of the nonlinear unknowns ($x$) to be minimied.
+The utility of VPM rests largely on the simplicitly of the derivative, or gradient, of the reduced function: it is the same as the gradient of the original, unreduced function, evaluated with the quadratic variables fixed at their inner solution value.
 Furthermore, stationary points of the reduced function are in 1-1 correspondence (under certain widely-met conditions) with stationary points of the unreduced function.
@@
 \section{The Variable Projection Objective}
-The variable projection method is a minimization algorithm for a scalar function on a product space $f:X \oplus W \rightarrow Y$ of class $C^2$.
+The variable projection method is a minimization algorithm for a scalar function on a product space $f:X \oplus W \rightarrow \bR$ of class $C^2$, $X$ and $W$ being Hilbert spaces.
 I will assume that $f$ is defined on the whole product space; the refinements necessary to accommodate constraints on $x$ are similar to those needed for solution of other nonlinear optimization problems.
 While not strictly necessary, I will also assume that $f$ is quadratic in the second variable.
-Put another way, $f(x,w) = 0.5*\|A(x)w-b\|^2$, where the values of $A$ are linear operators $: W \rightarrow Y$.
+Put another way, $f(x,w) = 0.5*\|A(x)w-b\|^2$, where $b \in Y$ (another Hilbert space with norm $\|\cdot\|$) and the values of $A$ are linear operators $: W \rightarrow Y$.
 Further, it's usually assumed that $A(x)$ is of full column rank (or coercive, in the infinite dimensional case), so that that for each $x$, there is a unique minimiser $\tilde{w}(x)$ of $w \mapsto f(x,w)$, the solution of the normal equation: $A(x)^T(A(x)\tilde{w}(x) - b)=0$.
@@
 The operator in parenthesis projects $Y$ onto the orthocomplement of the range of $A(x)$: call it $P(x)$. That is,
-$$
+\begin{equation}
+  \label{eqn:proj}
 P(x) = I-A(x)(A(x)^TA(x))^{-1}A(x)^T
-$$
+\end{equation}
 and
@@
 Since $P(x)$ is a projection, it is symmetric, positive semi-definite, and idempotent, that is $P(x)^TP(x)=P(x)^2=P(x)$. Thus the Gauss-Newton operator is approximately
-$$
+\begin{equation}
+  \label{eqn:kauf}
 DF(x)^TDF(x)s \approx \mbox{Hess}_{VP}\tilde{f}(x)s = DA(x)^*(\tilde{w}(x),P(x)DA(x)(\tilde{w}(x),s)).
-$$
+\end{equation}
 The right-hand side of this equation is the Kaufman Hessian approximation.
 
-The solution $s$ of the modified Newton step equation $\mbox{Hess}_{VP}\tilde {f}(x)s=-\mbox{grad}\tilde{f}(x)$ is a descent (or at least non-ascent) direction for $\tilde{f}$:
+The solution $s$ of the modified Newton step equation
+\begin{equation}
+  \label{eqn:NewtVP}
+  \mbox{Hess}_{VP}\tilde {f}(x)s=-\mbox{grad}\tilde{f}(x)
+\end{equation}
+is a descent (or at least non-ascent) direction for $\tilde{f}$:
 $$
@@
   \label{eqn:h1}
-\tilde{v}(x,s) = \mbox{argmin}_v 0.5*|A(x)v-DA(x)(\tilde{w}(x),s)|^2
+\tilde{v}(x,s) = \mbox{argmin}_v 0.5*\|A(x)v-DA(x)(\tilde{w}(x),s)\|^2
 \end{equation}
@@
 \section{Krylov-Newton optimization of VP reduced objectives}
+The numerical analysis literature on VPM has mostly discussed the cost based on direct solution of the Newton equation, with all of the linear operators appearing above represented as matrices and the linear systems solved by Gaussian elimination \cite[]{Kaufman:75,GolubPereyra:03}.
+In this case, each Newton iteration has approximately the same cost, so that $N_N$ Newton iterations have cost proportional to $N_N$.
+This direct approach is (by definition!) not practical for large-scale problems.
 
-The major costs of (approximate) Newton optimization are
-\begin{enumerate}
-\item $A(x)$ - cost of construction ($C_A$)
-\item $w \mapsto A(x)w$, $r \mapsto A(x)^T r$ - cost of application ($C_w$)
-\item $DA(x,w)s$ - cost of application ($C_{fwd}$)
-\item $DA^*(x,w)q$ - cost of application ($C_{adj}$)
-\item number of Newton iterations ($N_N$)
-\item for direct inner minimization (equations \ref{eqn:g1}, \ref{eqn:h1}), cost of solve ($C_S$)
-\item for iterative inner minimization (equations \ref{eqn:g1}, \ref{eqn:h1}), average number of iterations required ($N_I$).
-\end{enumerate}
-The cost of gradient (and objective value) compuation is
-\[
-  C_g = C_A + 2 C_w +C_S + C_{adj}
-\]
-for direct inner solves, or
-\[
-   C_g = C_A + 2 (1 + N_I)C_w + C_{adj}
- \]
-for iterative inner solves. Similarly, the cost of Hessian computation (assuming that the first step in that computation has been taken in computing the gradient) is
-\[
-  C_h =  2 C_w +C_S + C_{fwd} + C_{adj}
-\]
-for direct inner solves, or
-\[
-   C_h = 2 (1 + N_I)C_w) + C_{fwd} + C_{adj}
- \]
-for iterative inner solves. In either case, the cost of Newton optimization is roughly
-\[
-  N_N (C_g + C_h)
-\]
-The literature has mostly discussed the cost based on direct solution of the Newton equation.
-For iterative (Newton-Krylov, for instance) solution, the cost is dominated by the inner iterations for each outer iteration, proportional to $N_NN_I$.
-This is however roughly similar to Newton-Krylov for nonlinear least squares, which has been used successfully for solution of simulation-driven optimization problems for a long time \cite[]{Ghattas:IP25}.
+Alternatives to direct Newton solves are generalized secant or quasi-Newton iteration, replacing the Hessian by a low-rank approximation that can be explicitly inverted (\cite[]{NocedalWright}), and Newton-Krylov iteration in which the Newton equation \ref{eqn:NewtVP} is solved by an iterative (Krylov) method.
+The quasi-Newton approach ia most popular for VPM-reduced large scale problems, since it requires neither explicit formation, application, nor inversion of the Hessian (for instance \cite{Warner:16}.
+The value and gradient are required at each quasi-Newton step, and these can be computed by completion of a single inner iteration, as noted above.
+If this inner iteration requires $N_I$ applications of the operator $A(x)$, then the total cost is proportional to $N_NN_I$.
+
+Newton-Krylov iteration for nonlinear least squares problems uses iterative solution of the Newton equation \ref{eqn:NewtVP} via a Krylov subspace algorithm, for example conjugate gradient iteration \cite[]{Ghattas:IP25}.
+If $N_M$ iterations are required for the Newton step, then the total cost is proportional to $N_NN_M$.
+For some least-squares problems based on solution of partial differential equations (simulation-driven optimization), the Newton-Krylov approach can be superior in both quality of output and cost to the quasi-Newton approach \cite[]{Metivier:2014}.
+$N_M$ and $N_I$ are similar for many problems, but $N_N$ can be substantially smaller for Newton-Krylov than for a quasi-Newton algorithm.
+
+To employ the Newton-Krylov approach in the context of VPM, each step in the Krylov iteration requires solution of the Newton equation \ref{eqn:NewtVP}.
+Accomplishing this task by a Krylov iteration as well, as would presumably be necessary for large scale problems, involvs a number $N_H$ of applications of $A(x)$ and its transpose for each step.
+Thus the total cost is roughly proportional to $N_NN_MN_H$.
+If $N_M$ and $N_H$ are both roughly the same as the number of Krylov iterations $N_I$ required for solution of the inner least squares problem \ref{eqn:g1}, then the cost is proportional to $N_NN_I^2$, which appears to compare unfavorably to the cost of the quasi-Newton approach.
+This apparent cost penalty makes Newton-Krylov with the Kaufman Hessian unattracive, or even infeasible.
+
+For some simulation-driven optimization problems, a way around this difficulty might be found in specific choices of norm in the quadratic model space ($W$) and possibly in the data Hilbert space $Y$, making $A(x)$ approximately orthogonal: $A(x)^TA(x) \simeq I$ (the change in norms changes the meaning of the transpose or adjoint, of course).
+Computable norms with this property have been constructed for a number of simulation driven optimization problems based on wave propagation, and used to accelerate convergence of Newton-llike methods \cite[]{HouSymes:16,Herve2017,DafniSymes:SEG18b}.
+For application to the VPM-reduced problem, the projector $P(x)$ (definition \ref{eqn:proj}) in the Kaufman Hessian formula \ref{eqn:kauf} is $\approx I-A(x)A(x)^T$.
+Evaluation of this approximation requires no linear system soluition, and the cost estimate for VPM-Newton-Krylov collapses to $N_nN_M$, that is, the same as Newton-Krylov applied to a similar nonlinear least squares problem.
+
+In practice, some modifications are necessary, for example to ensure that the approximate Hessian is positive definite.
+Also, for any approaches based on iterative solution of the inner problem \ref{eqn:g1}, some criterion must be supplied to stop the iteration.
+Further, the accuracy with which $\tilde{w}(x)$ is computed affects the accuracy of the computed gradient $\mbox{grad} \tilde{f}(x)$ and any version of the Hessian, hence the rate (or even existence) of convergence of any Newton-like method.
+All of these issues require further attention.
 \bibliographystyle{seg}

### 2024-01-19 pyrvl/mswitest.py
all

@@ -119,7 +119,8 @@ try:
             vals.append(j.value())
             errs.append(j.dataerr()/dnorm)
         print('\nvals, rel errors at mtest = (1-t)*m0 + t*m')
-        print('alpha = ' + str(myalpha) + ' sigma = ' + str(mysigma) + ' rho = ' + str(myrho))
+        print('alpha = %10.4e sigma = %10.4e rho = %10.4e'
+                      % (myalpha, mysigma, myrho))
         print('  t       val       relerr')
         for i in range(13):
             print('%2.1f  %10.4e  %10.4e' % (0.1*i, vals[i], errs[i]))
@@ -128,7 +129,7 @@ try:
 
     else:
 
-        if choice == 'grad':
+        if choice == 'grad0':
         # homogeneous bulk modulus 
             j0 = awi.mswi(dom=usp, sim=F, mod=m0, data=d, alpha=myalpha, sigma=mysigma, kmax=mykmax, rho=myrho, verbose=2)
             data.model(bulkfile='grad0.rsf', bulk=4.0, nx=NX[idx], nz=NZ[idx],
@@ -152,10 +153,7 @@ try:
             dm.copy(m)
             dm.linComb(-1.0,m0)
             j0 = awi.mswi(dom=usp, sim=F, mod=m0, data=d, alpha=myalpha, sigma=mysigma, kmax=mykmax, rho=myrho, verbose=2)
-#            data.model(bulkfile='grad0.rsf', bulk=4.0, nx=NX[idx], nz=NZ[idx],
-#                           dx=DX[idx], dz=DX[idx], lensfac=1.0)
-#            data.model(bulkfile='mtest.rsf', bulk=4.0, nx=NX[idx], nz=NZ[idx],
-#                           dx=DX[idx], dz=DX[idx], lensfac=1.0)
+
             mtest = vcl.Vector(bulksp)
             g = vcl.Vector(bulksp)
             g.copy(j0.gradient())

### 2024-01-19 pyrvl/awiwgtest.py
all

@@ -111,7 +111,7 @@ try:
                                             
 # construct jet at m0
 
-    if choice == 'u0' or choice == 'grad0' or choice == 'grad':
+    else:
 
         if choice == 'grad0':
         # homogeneous bulk modulus 
@@ -130,7 +130,34 @@ try:
 
             g = vcl.Vector(bulksp,'grad.rsf')
             g.copy(wg.gradient())
-            
+
+        elif choice == 'gradtest':
+            print('awiwg gradient test')
+            h = 0.1
+            data.model(bulkfile='dm.rsf', bulk=4.0, nx=NX[idx], nz=NZ[idx],
+                    dx=DX[idx], dz=DX[idx], lensfac=1.0)
+            dm = vcl.Vector(bulksp,'dm.rsf')
+            dm.copy(m)
+            dm.linComb(-1.0,m0)
+            wg = awi.awiwg(dom=usp, sim=F, mod=m0, data=d, sigma=mysigma, kmax=mykmax, rho=myrho, precond=1, verbose=2)
+            data.model(bulkfile='grad0.rsf', bulk=4.0, nx=NX[idx], nz=NZ[idx],
+                    dx=DX[idx], dz=DX[idx], lensfac=1.0)
+            g = vcl.Vector(bulksp,'grad0.rsf')
+            g.copy(wg.gradient())
+            v0 = wg.value()
+            print('  v at m0 = ' + str(v0))
+            for ih in range(3):
+                data.model(bulkfile='mtest.rsf', bulk=4.0,
+                               nx=NX[idx], nz=NZ[idx],
+                               dx=DX[idx], dz=DX[idx], lensfac=1.0)
+                mtest = vcl.Vector(bulksp,'mtest.rsf')
+                mtest.copy(m0)
+                mtest.linComb(h,dm)
+                wgtest = awi.awiwg(dom=usp, sim=F, mod=mtest, data=d, sigma=mysigma, kmax=mykmax, rho=myrho, precond=1, verbose=0)
+                print('  v at m0 + ' + str(h) + '*dm = ' + str(wgtest.value()))
+                print('  predicted dv = ' + str(h*g.dot(dm)))
+                print('  actual dv = ' + str(wgtest.value()-v0))
+                h *= 0.5
         else:
             wg = awi.awiwg(dom=usp, sim=F, mod=m0, data=d, sigma=mysigma, kmax=mykmax, rho=myrho, precond=1, verbose=2)
             u0.copy(wg.innersol())

### 2024-01-19 pyrvl/npvc.py

@@ -106,6 +106,11 @@ class MatrixLSSolver(vcl.LSSolver):
             print(ex)
             raise Exception('called from npvc.MatrixLSSolver.solve')
 
+    def myNameIs(self):
+        print('MatrixLSSolver: NumPy-based LS solver')
+        print('  uses numpy.linalg.lstsq')
+        
+
 ###### EXAMPLES
 
 class OpExpl1(vcl.Function):

### 2024-01-19 pyrvl/vpm.py

@@ -71,7 +71,7 @@ class vpmjet(vcl.ScalarJet):
             if not isinstance(S, vcl.LSSolver):
                 raise Exception('fourth arg not vcl.LSSolver')
             if x0.space != F.getDomain()[0]:
-                raise Exception('first arg not in domain of second arg')
+                raise Exception('first arg not in domain of second arg[0]')
             if b.space != F.getRange():
                 raise Exception('third arg not vector in range of second arg')
         except Exception as ex:
@@ -103,6 +103,16 @@ class vpmjet(vcl.ScalarJet):
         else:
             return self.v
 
+    def innersol(self):
+        try:
+            if self.e is None:
+                [self.w, self.e] = self.S.solve(self.F.opfcn(self.x),self.b)
+        except Exception as ex:
+            print(ex)
+            raise Exception('called from vpmjet.innersol')
+        else:
+            return self.w
+        
     def gradient(self):
         try:
             if self.e is None:

### 2024-01-21 repro/papers/awi/paper.tex
final maybe

--- 2023-07-27 repro/papers/awi/paper.tex
@@
-\title{Notes on AWI}
+\title{Adaptive Waveform Inversion for Transmitted Wavefields}
 \author{William W. Symes}
@@
 \begin{abstract}
-goes here.
+Under some restricted circumstances, Adaptive Waveform Inversion
+applied to transmitted wave data yields estimates of index of
+refraction (or wave velocity) similar to those obtained by travel time inversion.
 \end{abstract}
@@
 \section{Introduction}
+Full waveform inversion (FWI) refers to the inference of mechanical
+parameter distribution within a material object  (the earth, a human
+body, a manufactured artifact,...) from the measurement of wave motion
+at its surface. Over the last several decades, this task has come to
+be carried out by minimizing the difference between computationa
+prediction of wave measurement versus actual observed
+measurements. The mechanical parameters to be inferred appear as
+coefficients in systems of partial differential equaitons governing
+the wave motion, and discrete representations of these parameters form
+the unknown vectors in the optimization problems thus generated.
 
-\section{AWI and why it works}
+While widely used in recent years, FWI exhibits a pathology when wave
+velocities or their parametric equivalents are amongst the parameters
+to be estimated: the excessive sensitivity of predicted waveforms to
+changes in velocity leads to oscillations in the difference measures,
+especially for the most commonly used mean-square measure. Local
+optimization algorithms (relatives of Newton;'s method) are the only practical
+methods to generate sequences of approximate optimal
+parameters. However the oscillations in the difference measures lead
+these sequences to become entrapped at near parameter distributions
+far from the optimum, and thus do not extract the maximum information
+from the data \cite[]{VirieuxOperto:09}.
 
-The version of AWI introduced by Michael Warner and Lluis Guasch \cite[]{Warner:16,GuaschWarnerRavaut:GEO19} rests on the assumption that seismic waves are governed by linear acoustics, and that each shot is associated to an isotropic point source with known location $\bx_s$ and wavelet $w(t;\bx_s)$.
-That is, the pressure and velocity fields $p({\bf x},t;{\bf x}_s)$, ${\bf v}({\bf x},t;{\bf x}_s)$ for the shot location ${\bf x}_s$ depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf x})$ (reciprocal of the density $\rho({\bf x})$), and wavelet $w(t;{\bf x}_s)$ through the acoustic system
- $$
- \frac{\partial p}{\partial t} = - \kappa \nabla \cdot {\bf v} +
-w(t;{\bf x}_s) \delta({\bf x}-{\bf x}_s);
-\frac{\partial {\bf v}}{\partial t} = - \beta \nabla p; 
-p, {\bf v} = 0 \mbox{ for }  t \ll 0.
-$$
+Various modifications of straighforward best-fit FWI have been
+suggested to overcome this ``cycle-skipping'' pathology. One of these
+is Adaptive Waveform Inversion \cite[]{Warner:16} (AWI). This
+technique has been successfully applied to industry-scale seismic data
+\cite[]{GuaschWarnerRavaut:GEO19,Warneretal:SEG21} as well as to
+ultrasonic imaging of the the human brain
+\cite[]{Guaschetal:NPJDM20}.
+
+While heuristic justifications for AWI (and
+many details of implementation) are discussed in these and other
+works, a clear explanation for its effectiveness is 
+so far missing from the literature. In this short note, I show that
+for active sensing of acoustic materials with unique arrival times of
+transmitted waves
+(that is unique rays of geometric asymptotics between sources and
+receivers), the AWI objective function is an approximation to the
+mean-square time-of-travel misfit. Thus AWI and travel time fitting
+lead to similar material parameter estimates. Since the latter is
+widely believed to be a near-optimal method for wave velocity
+estimation, this demonstration explains roughly ``why it works''.
+
+The reader will
+note that a normalization component of the AWI objective function
+definition is very
+important in reaching this conclusion. The connection between
+objective normalization and the close relation to travel time
+inversion is completely independent from 
+the reasons for normalizing the objective advanced in the
+literature \cite[]{Warner:16,LiAlkhalifah:21}.
+
+After establishing
+the class of problems to be studied and notations for their
+description in the following section, I present the asymptotic
+analysis leading to the main conclusion in the next. The close relation
+between AWI objective and travel time misfit no longer mecessarily
+holds if multiple rays connect source and receiver. I discuss these
+and other aspects of AWI and related approaches in a brief final section.
+
+\section{Adaptive Waveform Inversion}
+
+The version of AWI introduced by \cite{Warner:16} is based on an
+active source acoustic model of data generation. Each acoustic source
+is modeled an isotropic point radiator with known location $\bx_s$ and
+wave form (``wavelet'') $w(t;\bx_s)$. Similarly, receivers are modeled
+as isotropic point sensors, recording pressure at points ${\bf x}_r$.
+The pressure and velocity fields $p({\bf x},t;{\bf x}_s)$, ${\bf v}({\bf x},t;{\bf x}_s)$ for the source location ${\bf x}_s$ depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf x})$ (reciprocal of the density $\rho({\bf x})$), and wavelet $w(t;{\bf x}_s)$ through the acoustic system
+\begin{eqnarray}
+  \label{eqn:awe}
+ \frac{\partial p}{\partial t} & = &- \kappa \nabla \cdot {\bf v} +
+w(t;{\bf x}_s) \delta({\bf x}-{\bf x}_s); \nonumber \\
+\frac{\partial {\bf v}}{\partial t} & = & - \beta \nabla p; \\ 
+p, {\bf v} & = & 0 \mbox{ for }  t \ll 0.
+\end{eqnarray}
 For sake of brevity, define the model vector $m=(\kappa,\rho)$.
-The forward map or {\em modeling operator} is $F[m]w = \{p({\bf x}_r,t;{\bf x}_s)\}$, in which shot and receiver positions ${\bf x}_s, {\bf x}_r$ define the acquisition geometry.
+The forward map or {\em modeling operator} is $F[m]w = \{p({\bf x}_r,t;{\bf x}_s)\}$, in which source and receiver positions ${\bf x}_s, {\bf x}_r$ define the acquisition geometry.
 
-Acoustic inversion, for present purposes, means: given time interval $[0,T]$, source and receiver locations $(\bx_s,\bx_r)$, source wavelet $w_* \in C_0^{\infty}(\bR)$ and data traces $d(\bx_r,\cdot;\bx_s) \in L^2([0,T])$, find a model $m$ so that $F[m]w_* \approx d$.
-The simplest version of FWI concretizes this task by asking for a model $m$ minimizing the mean square error
+Acoustic inversion, for present purposes, means: given recording time interval
+$[0,T]$, source and receiver locations $(\bx_s,\bx_r)$, source wavelet
+$w(t;\bx_s)$ and data traces $d(\bx_r,\cdot;\bx_s)$, find a model $m$
+so that $F[m]w \approx d$. The
+simplest version of FWI concretizes this task by asking for a model
+$m$ minimizing the mean square error
 \begin{equation}
   \label{eqn:fwi}
-  J_{\rm FWI}[m;d] = \frac{1}{2}\|F[m]w_*-d\|^2
+  J_{\rm FWI}[m;d] =\frac{1}{2} \sum_{{\bf x}_r,{\bf x}_s} \int\,dt\, |(F[m]w)({\bf x}_r,t;{\bf x}_s) - d({\bf x}_r,t;{\bf x}_s)|^2
 \end{equation}
-Practical versions of FWI include various types of regularization and homotopy, see the sources cited in the Introduction.
-As mentioned there, local iterative optimization methods tend to stagnate at highly sub-optimal (``cycle-skipped'') estimates of $m$ unless the initial estimate is of quite high quality, often available only at considerable cost in effort and data collection.
-%For now, assume that the data $d({\bf x}_r,t;{\bf x}_s)$ is the output of the modeling operator for "true" model $m_*$, that is, the "true" bulk modulus, buoyancy, and wavelet $\kappa_*, \beta_*, w_*$: that is, $d = F[m_*]w_*$.
+It is convenient to introduce the $L^2$ norm $\| \cdot \|$ as
+shorthand for the right-hand side of the definition \ref{eqn:fwi}:
+\[
+  J_{\rm FWI}[m;d]= \frac{1}{2}\|F[m]w-d\|^2.
+\]
+In the sequel I will use the $L^2$ norm notation to stand for
+mean-square over whatever variables make sense.
+
+Note that the wavelet $w$ is assumed known in this formulation of the
+FWI problem. In fact, even if actual acoustic sources are
+well-approximated by isotropic point radiators, their wave forms are
+usually poorly determined and may reasonably be
+included as  parameters to be estimated by inversion of the
+data. However for the present discussion we will regard $w$ as known.
+
+Practical versions of FWI include various types of regularization and
+imposition of constraints on $m$, see the sources cited in the
+Introduction. As mentioned there, local iterative optimization methods
+are the only practical approach to minimizing $J_{\rm FWI}$ due to the
+computational cost of approximating the forward map $F$ and related
+operators, but these tend to stagnate at highly sub-optimal
+(``cycle-skipped'') estimates of $m$ unless the initial estimate is of
+quite high quality, often available only at considerable cost in data
+collection and processing.
+%For now, assume that the data $d({\bf x}_r,t;{\bf x}_s)$ is the output of the modeling operator for "true" model $m^*$, that is, the "true" bulk modulus, buoyancy, and wavelet $\kappa^*, \beta^*, w^*$: that is, $d = F[m^*]w^*$.
 
 AWI is based on extending the modeling operator $F$, that is, enlarging its domain.
-The extended modeling operator ${\bar F}$ maps extended sources $\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure field.
-That is, the extended source depends on the receiver location as well as the source location - so there is one acoustic system for each source *and* receiver position - a lot of wave equations!
-If all of the wavelets for each source are the same, that is, ${\bar w}({\bf x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver position, then ${\bar F}[m]\bar{w} = F[m]w$.
-That is, $F$ is a special case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of $F$.
-This is the *source-recever* extension, in the terminology of Huang et al.
-2015.
+The extended modeling operator ${\bar F}$ maps extended sources
+$\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
+field. That is, the extended source depends on the receiver location
+as well as the source location - so there is one acoustic system for
+each source {\em and} receiver position - a lot of wave equations! If all
+of the wavelets for each source are the same, that is, ${\bar w}({\bf
+  x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver
+position, then ${\bar F}[m]\bar{w} = F[m]w$. That is, $F$ is a special
+case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
+$F$. This is the {\em source-receiver} extension, in the terminology of \cite{HuangSymes2015SEG}.
 
-AWI assumes that the extended sources are time convolutions of the (known) exact source with an {\em adaptive kernel} $u({\bf x}_r,t;{\bf x}_s)$: that is, $\bar{w} = u * w_*$ - the asterisk denotes convolution in time.
+AWI assumes that the extended sources are time convolutions of the
+(known) source $w$ with an {\em adaptive filter} $u({\bf
+  x}_r,t;{\bf x}_s)$: that is,
+\[
+  \bar{w}({\bf x}_r,t;{\bf x}_s) = (u * w)({\bf  x}_r,t;{\bf x}_s)
+  = (u({\bf  x}_r,\cdot;{\bf x}_s)*w)(t).
+\]
+The asterisk denotes convolution in time. That is,
+$u*w$ denotes the result of convolving each trace in $u$ by $w$. I
+will also use the asterisk to denote the tracae-by-trace convolution
+of two data sets with the same set of source and receiver coordinates,
+that is, the data set consisting of time convolutions of traces with the same source
+and receiver coordinates. The meaning will be clear from context.
+
 Since linear acoustics is time-translation invariant, its solution commutes with time convolution, that is,
-$$
-\bar{F}[m]\bar{w} = \bar{F}[m](u* w_*) = (F[m]w_*)*u
-$$
-For notational convenience, introduce the notation $S[m,w_*]$ for the convolution operator with kernel $F[m]w_*$:
+\[
+ \bar{F}[m]\bar{w} = \bar{F}[m](u* w) = (F[m]w)*u
+\]
+Since the known source $w$ is not adjusted during the inversion, 
+introduce the notation $S[m]$ for the convolution operator with kernel $F[m]w$:
 \begin{equation}
-  \label{eqn:defs}
-  S[m,w_*]u = (F[m]w_*)*u
+  \label{eqn:sdef}
+  (F[m]w)*u \equiv S[m]u
 \end{equation}
-Warner and Guasch assume (implicitly) that there is always a (near-)zero-residual solution of the extended inversion problem.
-That is, they assume that for any $m$, there is an extended source $\bar{w}$, or equivalently an adaptive kernel $u_0$ with $\bar{w} = w_**u_0$ for which $\bar{F}[m]\bar{w} = S[m,w_*]u_0\approx d$.
-If $\kappa \approx \kappa_*, \beta \approx \beta_*$, then $\bar{w}$ should be $\approx w_*$, so the adaptive kernel $u_0$ should be approximately $\delta(t)$ and independent of ${\bf x}_s,{\bf x}_r$.
-Such a kernel $u_0$ is (approximately) in the null space of multiplication by $t$.
+\cite{Warner:16} assume (implicitly) that there is always an exact
+solution of the extended inversion problem. That is, they assume that
+for any $m$, there is an extended source $\bar{w}$, or equivalently an
+adaptive filter $u[m;d]$ with $\bar{w}=w*u[m;d]$, for which
+$S[m]u[m;d]= d$. [This proposition is dubious, but is easy to work
+around, as will be seen below. Also, \cite{Warner:16} also formulate
+AWI in terms of the inverse filter, that is, $u$ for which
+$u*d = F[m]w$, and in fact recommend that definition. The present
+discussion uses only the forward filter.] If
+$\kappa = \kappa^*, \beta = \beta^*$, then $\bar{w}$ should be $= w$,
+so the adaptive kernel $u[m;d]$ should be $=\delta(t)$ for all
+${\bf x}_s,{\bf x}_r$. Such a kernel $u[m;d]$ is in the null space of
+multiplication by $t$.
 
-This observation leads to the definition of the Warner-Guasch AWI objective function $\tJz$:
-
-\begin{itemize}
-\item[1. ]Given $m$, solve the deconvolution problem $(F[m]w_*)*u_0 = d$ for $u_0$, if necessary in a least squares sense and with Tihonov regularization;. 
-\item[3. ]Compute the objective 
+This observation leads to the definition of the AWI objective function $J_{\rm AWI}$:
 \begin{equation}
   \label{eqn:tJz}
-    \tJz[m] = \sum_{x_s,x_r} \left(\frac{\int dt |tu_0|^2}{\int dt |u_0|^2}\right)
+  J_{\rm AWI}[m;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu[m;d](\bx_r,t;\bx_s)|^2}{\int dt |u[m;d](\bx_r,t;\bx_s)|^2}\right)
 \end{equation}
-and its gradient, then update $\kappa,\beta$ using a gradient descent method to minimize $\tJz$.
-\end{itemize}
+in which $u[m;d]$ is as defined above, that is, the solution of $S[m]u[m;d]
+= d$.
 
-The normalization of the integrand per source and receiver by the $L^2$ norm of $u_0$ is quite important, as will be explained shortly.
+Adaptive waveform inversion consists in iterative minimization of
+$J_{\rm AWI}$: choose an initial estimate $m_0=(\kappa_0,\beta_0)$ of the
+model, then update $m=(\kappa,\beta)$ repeatedly using a gradient
+descent method. \cite{Warner:16} describes the computation of $J_{\rm AWI}$
+and its gradient in detail, and their use in a quasi-Newton
+optimization algorithm.
 
-It may not be obvious at first glance why this approach to estimating $m$ should be any more effective than FWI.
-To see why AWI works, in at least some settings, we assume that the spatial dimension is three, and make use of the geometric optics approximation to the solution of the wave equation, in a form due to Hadamard.
-This approximation is {\em local}: it applies when sources and receivers are sufficiently close.
-We will discuss its global failure and the consequences for AWI in a later section.
+The trace-by-trace normalization in definition \ref{eqn:tJz} of each
+trace in $tu[m;d]$ by the
+$L^2$ norm of the corresponding trace in $u[m;d]$ is quite important, as will be explained shortly.
 
-Transient acoustic sources have no DC (zero frequency) component and are relatively small at low frequencies.
-One was to enforce these characteristics is to assume that the wavelet $w_*$ is the time derivative of a $v_* \in C_0^{\infty}(\bR)$.
-Set $\lambda(w_*) = \|v_*\|/\|w_*\|$ ($L^2$ norms).
-$\lambda$ is a proxy for wavelength: it is the reciprocal of the RMS frequency.
+\section{Why it works}
 
-For a particular source-receiver pair ${\bf x}_s,{\bf x}_r$ with $\bx_r$ "not too far" from $\bx_s$,, suppressed from the notation,
+It may not be obvious at first glance why this approach to estimating
+$m$ should be any more effective than FWI. To see why AWI works, in at
+least some settings, we assume that the spatial dimension is three,
+and make use of the geometric asymptotics (ray theory) approximation
+to the solution of the wave equation. This approximation is {\em
+  local}: it applies when sources and receivers are sufficiently
+close. We will discuss its global failure and the consequences for AWI
+in the final section.
+
+Transient acoustic sources have no DC (zero frequency) component and
+are relatively small at low frequencies. One way to enforce these
+characteristics is to assume that the wavelet $w$ is the time
+derivative of another wavelet $v$, also transient ($v(t;{\bf x}_s)=0$
+for $|t| \gg 0$). Set $\lambda(w) = \|v\|/\|w\|$. $\lambda$ is a proxy for wavelength: it is the reciprocal of RMS frequency.
+
+The geometric asymptotics approximation for solution of the point
+radiator problem \ref{eqn:awe} expresses the pressure field in terms
+of $m$-dependent geometric amplitude $a[m]({\bf x}_s,{\bf x}_r)$ and
+travel time $\tau[m]({\bf x}_s,{\bf x}_r)$, and implies that
 \begin{equation}
   \label{eqn:hadamard}
-  F[m]w_*(t) = a[m]w_*(t-\tau[m]) + Rw_*(t)
+  (F[m]w)({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) w(t-\tau[m]({\bf x}_s,{\bf x}_r)) 
 \end{equation}
-where $R(\bx_r,\bx_s)$ is a bounded operator on $L^2((-\infty,T])$ bounded by a multiple of $\lambda(w_*)$.
-This ``single arrival'' approximation  follows from the expansion of the fundamental solution of the wave equation due to Hadamard, see for instance \cite{Friedlander:75}.
-This construction is sketched in Appendix B.
-Note that in general the approximation \ref{eqn:hadamard} is only local: for smooth $m$, it holds for $\bx_r$ sufficiently close to $\bx_s$, but fails at larger distances.
+In this approximation, and in other uses of the symbol $\approx$, the
+error is $O(\lambda(w))$.
 
-For approximations with $O(\lambda(w_*)$ remainders, write
-\begin{equation}
-  \label{eqn:pwe}
-  F[m]w_*(t) \approx a[m]w_*(t-\tau[m])
-\end{equation}
-Thanks to the approximation \ref{eqn:pwe},
+Approximation \ref{eqn:hadamard} follows from a
+``progressing wave''  expansion of the fundamental
+solution (Green's function) of the wave equation due to Hadamard
+\cite[]{Friedlander:75,Qian:JCP24}. For smooth (slowly varying)
+$m$, it holds for $\bx_r$ sufficiently close to $\bx_s$ that a unique ray of
+geometric acoustics connect $\bx_s$ to $\bx_r$. This ``single
+arrival'' condition fails (in
+general) at larger distance. The precise region of validity depends on
+$m$. For the remainder of this section, its validity is assumed.
+
+Applying the approximation \ref{eqn:hadamard}, obtain
 \begin{equation}
   \label{eqn:spwe}
-S[m,w_*]u(t) \approx a[m](w_**u)(t-\tau[m])
+S[m]u({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) (w*u)(t-\tau[m]({\bf x}_s,{\bf x}_r))
 \end{equation}
@@
 $$
-d(t)=F[m_*]w_*(t) \approx a[m_*]w_*(t-\tau[m_*])
+d({\bf x}_r,t;{\bf x}_s)=(F[m^*]w) ({\bf x}_r,t;{\bf x}_s) \approx a[m^*]({\bf x}_s,{\bf x}_r) w(t-\tau[m^*]({\bf x}_s,{\bf x}_r))
 $$
-The (approximate) solution of $S[m,w_*]u_0=d$ is obvious by inspection: $u_0(t) = a[m_*]/a[m]\delta(t-\tau[m_*]+\tau[m])$.
-There is immediately a big problem: this $u_0$ is not square-integrable, and the AWI objective $\tJz$ is undefined.
-To fix this, it's necessary to treat $u_0$ as the solution of a regularized least squares problem: define
+The (approximate) solution of $S[m]u[m;d]=d$ is obvious by inspection:
+\begin{equation}
+  \label{eqn:notl2}
+  u[m;d]({\bf x}_r,t;{\bf x}_s) = \frac{a[m^*]({\bf x}_s,{\bf x}_r)}{a[m]({\bf x}_s,{\bf x}_r)}\delta(t-\tau[m^*]({\bf x}_s,{\bf x}_r)+\tau[m]({\bf x}_s,{\bf x}_r)).
+\end{equation}
+There is immediately a big problem: this $u[m;d]$ is not square-integrable, and the AWI objective $J_{\rm AWI}$ is undefined.
+To fix this, it's necessary to re-define $u[m;d]$ as the solution of a regularized least squares problem: define
+\begin{equation}
+  \label{eqn:reg}
+  u[m;d] = \mbox{argmin}_u (\|S[m]u-d\|^2 + \sigma^2\|u\|^2)
+\end{equation}
 $$
-u_0^{\epsilon} = \mbox{argmin}_u (\|S[m,w_*]u-d\|^2 + \epsilon^2\|u\|^2)
+\approx \mbox{argmin}_u (\|a[m](w*u)(\cdot-\tau[m])-a[m^*]w(\cdot-\tau[m^*])\|^2 + \sigma^2\|u\|^2)
 $$
+
+Note that in these expressions, the source and receiver
+coordinates ${\bf x}_s$ and ${\bf x}_r$ act as either passive parameters
+or summation indices. Therefore, to simplify the presentation, drop them from the notation
+temporarily. Also,  use the temporary abbreviations $a^*=a[m^*], a=a[m], \tau^*=\tau[m^*],
+\tau=\tau[m]$.
+
+With these conventions, re-write equation \ref{eqn:reg} in terms of the Fourier transforms $\hat{u}[m;d],\hat{w}$ of $u[m;d]$ and $w$:
+\begin{equation}
+  \label{eqn:freqreg}
+\hat{u}[m;d] \approx \mbox{argmin}_{\hat{u}}\left(\int d\omega |a\hat{w}\hat{u}e^{i\omega \tau}-a^*\hat{w}e^{i\omega\tau^*}|^2 + \sigma^2 |\hat{u}|^2\right)
+\end{equation}
+The normal equation for the least squares problem \ref{eqn:freqreg} is
 $$
-\approx \mbox{argmin}_u (\|a[m](w_**u)(\cdot-\tau[m])-a[m_*]w_*(\cdot-\tau[m_*])\|^2 + \epsilon^2\|u\|^2)
-$$
-Use the temporary abbreviations $a_*=a[m_*], a=a[m], \tau_*=\tau[m_*], \tau=\tau[m]$, and write in terms of the Fourier transforms $\hat{u}_{\epsilon},\hat{w_*}$ of $u_0^{\epsilon}$ and $w_*$:
-$$
-\hat{u}_{\epsilon} \approx \mbox{argmin}_{\hat{u}}\left(\int d\omega |a\hat{w_*}\hat{u}e^{i\omega \tau}-a_*\hat{w_*}e^{i\omega\tau_*}|^2 + \epsilon^2 |\hat{u}|^2\right)
-$$
-The normal equation is
-$$
-(a^2 |\hat{w}_*|^2 +\epsilon^2)\hat{u} = a a_*|\hat{w}_*|^2e^{i\omega(\tau_*-\tau)}
+(a^2 |\hat{w}|^2 +\sigma^2)\hat{u} = a a^*|\hat{w}|^2e^{i\omega(\tau^*-\tau)}
 $$
@@
 $$
-\hat{u}_{\epsilon} \approx \frac{a^*}{a}\hat{g}_{\frac{a^*}{a}\epsilon} e^{i\omega(\tau_*-\tau)}
+\hat{u}[m;d] \approx \frac{a^*}{a}\hat{g}_{\frac{a^*}{a}\sigma} e^{i\omega(\tau^*-\tau)}
 $$
@@
 $$ 
-\hat{g}_{\epsilon} = \frac{a_*^2|\hat{w}_*|^2}{a_*^2|\hat{w}_*|^2 + \epsilon^2}.
+\hat{g}_{\sigma} = \frac{(a^*)^2|\hat{w}|^2}{(a^*)^2|\hat{w}|^2 + \sigma^2}.
 $$
-Since $w_* \in C_0^{\infty}$, $\hat{w_*}(\omega) \rightarrow 0$ faster than any negative power of $\omega$ as $|\omega| \rightarrow \infty$ and is analytic, whence $\hat{g}_{\epsilon}: \epsilon > 0$ tends to $1$ almost everywhere as $\epsilon \rightarrow 0$.
-Hence the inverse Fourier transform $g_{\epsilon}$ tends to $\delta$ in the sense of distributions, and
+depends only the ``true'' model $m^*$.
+Transient acoustic sources are effectively bandlimited: $\hat{w}(\omega) \rightarrow 0$ faster than any negative power of $\omega$ as $|\omega| \rightarrow \infty$ and is analytic, whence $\hat{g}_{\sigma}:$ tends to $1$ almost everywhere as $\sigma \rightarrow 0$.
+Hence the inverse Fourier transform $g_{\sigma}$ tends to $\delta$ in the sense of distributions, and
 $$
-u_0^{\epsilon}(t) \approx \frac{a^*}{a}g_{\frac{a^*}{a}\epsilon}(t-(\tau_*-\tau))
+u[m;d](t) \approx \frac{a^*}{a}g_{\frac{a^*}{a}\sigma}(t-(\tau^*-\tau))
 $$
-to a multiple of a shifted $\delta$, in fact exactly the distribution solution of $\bar{S}[m,w_*]u=d$. However for $\epsilon>0$, $u_0^{\epsilon}$ is square integrable.
+to a multiple of a shifted $\delta$, in fact exactly the distribution
+solution of $\bar{S}[m,w]u=d$ given in equation \ref{eqn:notl2}. However for $\sigma>0$, $u[m;d]$ is square integrable.
 
-\cite{HuangSymes:SEG15a} presented an analysis of the unweighted AWI objective (numerator of the right hand side in the definition \ref{eqn:tJz}), based in turn on ideas presented in Song's thesis \cite[]{Song:94c}.
-With $u$ replaced by $u_0^{\epsilon}$,  this unweighted objective is
+\cite{HuangSymes2015SEG} presented an analysis of the Matched Source
+Waveform Inversion (MSWI) objective function, which is essentially the
+numerator of the right hand side in the definition
+\ref{eqn:tJz} of the AWI objective. This analysis was based in turn on ideas presented in Song's thesis
+\cite[]{Song:94c}. MSWI deals with the extended source $\bar{w}$, but
+the same construction applies to the adptive filter $u$. With $u$ replaced by $u[m;d]$,  this MSWI objective is
+\[
+J_{\rm MSWI}[m;d] = \int dt |tu[m;d]|^2 \approx \frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\sigma}(t-(\tau^*-\tau))|^2
+\]
+\[
+=\frac{(a^*)^2}{a^2} \int dt\, (t+(\tau^*-\tau))^2|g_{\frac{a^*}{a}\sigma}(t)|^2
+\]
+\begin{equation}
+  \label{eqn:MSWI}
+  =\frac{(a^*)^2}{a^2} \int dt\, (t^2 + 2t
+  (\tau^*-\tau)+(\tau^*-\tau)^2)|g_{\frac{a^*}{a}\sigma}(t)|^2
+\end{equation}
+As noted by \cite{HuangSymes2015SEG}, the linear term (in t) on the
+right-hand side of \ref{eqn:MSWI} vanishes,
+since $g$ is necessarily even in time (since its Fourier transform is
+real). So
+\begin{equation}
+  \label{eqn:MSWIasym}
+  J_{\rm MSWI}[m;d] \approx
+\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\sigma}(t)|^2
++(\tau^*-\tau)^2\frac{(a^*)^2}{a^2}\|g_{\frac{a^*}{a}\sigma}\|^2
+\end{equation}
+The first term tends to zero as $\sigma \rightarrow 0$, since
+$g_{\sigma}$ is a delta-family, so you can make it negligible with
+proper choice of $\sigma$. The second term suggests that this function
+is related to the travel-time error $\tau^*-\tau$.
+
+The relation is obscured by the presence of of the amplitude ratio
+$a^*/a$ both as a multiplier and in the scaling of $\sigma$. Since the
+amplitudes depend on $m$, their presence may influence the location
+(or even the presence) of stationary points of $J_{\rm MSWI}$ for
+models differing in their travel time predictions. So stationary
+points of the MSWI objective may not yield slowness models that match
+travel-times with the data, even in the noise-free case (though
+the examples reported by \cite{HuangSymes2015SEG, HuangSymes:Geo17}, and other works
+cited there, suggest that sometimes this is the case).
+
+Comparing the definitions \ref{eqn:tJz} of $J_{\rm AWI}$ and
+\ref{eqn:MSWI} of $J_{\rm MSWI}$, ones sees
+that the two functions differ only by the factor $\|u[m;d]\|^{-2}$ in
+the former. Using the geometric asymptotics approximation, 
 $$
-\int dt |tu_0^{\epsilon}|^2 \approx \frac{a_*^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\epsilon}(t-(\tau_*-\tau))|^2
+\|u[m;d]\| \approx \frac{a^*}{a}\|g_{\frac{a^*}{a}\sigma}\| + O(\sigma^2)
 $$
+Therefore, multiplication of the MSWI objective by $\|u[m;d]\|^{-2}$
+precisely removes the amplitude factor in the right-hand side of the
+relation \ref{eqn:MSWIasym}. Restoring the source and receiver
+coordinates to the notation,
 $$
-=\frac{a_*^2}{a^2} \int dt\, (t+(\tau_*-\tau))^2|g_{\frac{a^*}{a}\epsilon}(t)|^2
+J_{\rm AWI}[m;d] \approx \sum_{{\bf x}_s,{\bf x}_r} (\tau[m^*]({\bf
+  x}_s,{\bf x}_r)-\tau[m]({\bf x}_s,{\bf x}_r))^2 + O(\sigma^2)
 $$
-The linear (in t) term vanishes, since $g$ is necessarily even in time (since it Fourier transform is real). So this is
-$$
-=\frac{a_*^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\epsilon}(t)|^2
-+(\tau_*-\tau)^2\frac{a_*^2}{a^2}\|g_{\frac{a^*}{a}\epsilon}\|^2
-$$
-The first term tends to zero as $\epsilon \rightarrow 0$, since $g_{\epsilon}$ is a delta-family, so you can make it negligible with proper choice of $\epsilon$.
-The second term suggests that this function is related to the travel-time error $\tau_*-\tau$.
+In other words, the AWI objective is {\em exactly} an approximation to
+the travel-time tomography objective, with errors proportional to the
+``wavelength'' $\lambda(w)$ and the regularization parameter
+$\sigma^2$.
 
-The relation is obscured by the presence of of the amplitude ratio $a_*/a$ both as a multiplier and in the scaling of $\epsilon$.
-Since the amplitudes depend on $m$, their presence may influence the location (or even the presence) of  stationary points of $\tJz$ for models differing in their travel time predictions.
-So stationary points of the unweighted AWI objective may not yield slowness models that match travel-times with the data, even in the noise-free case (even though the examples reported in Huang and Symes SEG 2015, and other works cited there, suggest that sometimes this is the case).
+And that's ``why it works'' - when geometric asymptotics in the form
+given by equation \ref{eqn:hadamard} is an accurate approximation to the Green's function.
 
-The scaling by trace norms of $u_0^{\epsilon}$ in the definition \ref{eqn:tJz} eliminates the possibility of non-travel-time stationary points.
-Since the objective function is built up trace-by-trace, use the same approximation for the norm of $u_0^{\epsilon}$:
-$$
-\|u_0^{\epsilon}\| \approx \frac{a^*}{a}\|g_{\frac{a^*}{a}\epsilon}\|
-$$
+\section{Multiple Arrivals and Cycle-Skipping}
+The geometric asymptotics approximation \ref{eqn:hadamard} holds only
+in the ``single arrival'' case, that is, when sources and receivers
+are close enough together relative to the length scale of significant
+variation in the wave velocity $((\kappa\beta)^{1/2})$ that a unique ray of
+geometric acoustics connects each source-receiver pair. At larger
+distances, multiple raypaths typically exist, corresponding to
+multiple idenfiable wavefront arrivals. Accordingly, in such settings
+the analysis presented in the last section no longer applies, and the
+relation established there between AWI and travel-time inversion is
+case into doubt.
+
+In fact, the behaviour inherited from travel-time inversion in the
+single arrival case, namely the absence of spurious stationary points
+(``cycle-skipping''), generally fails when multiple arrivals with
+significant energy are present in the data. \cite{Symes:94c}
+illustrated this phenomenon for a version of MSWI. Crosswell
+source-receiver geometry often generates multiple arrivals,
+particularly near low-velocity sedimentary layers that are natural
+targets of seismic exploration and act as wave
+guides. \cite{Plessix:00} successfully applied a version of MSWI to
+crosswell waveform inversion by removing traces with close source and
+receiver depths, thus leaving only data that tended to be dominated by
+single arrivals. \cite{HuangSymes:Geo17} illustrate the failure of the
+version of MSWI reported here to avoid cycle-skipping in inversion of
+strongly refracting models, and ameliorate the this failure to some
+extent by Tihonov regularization. They observe that the mapping from
+adaptive filter or extended source to data (represented by the opertor
+$S[m]$ in the discussion above) may not have a well-behaved inverse in
+the presence of multiple wavefront arrivals. \cite{Yongetal:GJI23}
+have shown that precisely the same issue afflicts AWI, and have
+proposed a remedy by localizing the definition of the objective to
+short time intervals, within which multiple wavefront arrrivals are
+unlikely. \cite{Symes:23} constructs an extension based on surface
+sources, as opposed to the source-receiver extension which is the
+basis of MSWI and AWI. This surface source extension generates an
+invertible map from extended source to data.
+
+A simple calculation reveals the phenomenon discussed in all of these
+works. The analogue of the geometric asymptotics for a source-receiver
+pair with connected by multiple rays is (once again suppressing the
+source and receiver coordinates from the notation):
+\begin{equation}
+  \label{eqn:multi}
+  F[m]w(t) \approx \sum_{i \ge 0} a_i H^{p_i}w(t-\tau_i).
+\end{equation}
+The amplitudes $\{a_i\}$ and arrival times $\{\tau_i\}$ depend on $m$, as
+before, through ray-tracing. $H$ is the Hilbert transform, and $p_i$
+counts the number of times the ray for arrival time $\tau_i$ has
+touched a caustic (counted with multiplicity, in an appropriate
+sense). The earliest arrival has not touched a caustic. If the
+indexing is organized so that $\tau_0 < \tau_i$ for $i>0$, then
+$p_0=0$. See \cite{Friedlander:75} for details.
+
+The right-hand side of equation \ref{eqn:multi} can be expressed as
+\[
+  = (\delta + T) * a_0w(t-\tau_0),
+\]
+in which 
+\[
+  T = \sum_{i \ge 1}
+  \frac{a_i}{a_0}H^{p_i}\delta(t-(\tau_i-\tau_0))
+\]
+Thus
+\begin{equation}
+  \label{eqn:unwrap}
+  (\delta +T_1) * F[m]w \approx  a_0w(t-\tau_0)
+\end{equation}
+where $\delta + T_1$ is the convolution inverse of $\delta + T$,
+defined by 
+\[
+  T_1 = \sum_{n=1}^{\infty} (-1)^n T^n.
+\]
+$T^n$ should be understood as the $n$th convolution power in this
+expression, that is, $T*T*...*T$, $n$ times. $T_1$ is well-defined as a distribution.
+
+Once again, it is possible to write the adaptive filter u for which $S[m]u = (F[m]w)*u = d =
+F[m^*]w$ by inspection. Use the notations $a^*_i, t^*_i, p^*_i$ for
+the analogous quantities computed with the ``true'' model $m^*$, and
+$T^*, T_1^*$ defined as above but with $a^*_i$ in place of $a_i$ and so
+on. From the observation \ref{eqn:notl2},
+\[
+  d = F[m_*]w = S[m]u
+\]
+provided that we choose
+\begin{equation}
+  \label{eqn:notl2again}
+u(t) \approx (\delta + T^*)*(\delta +
+T_1)*\left(\frac{a^*_0}{a_0}\delta(t - (\tau^*_0-\tau_0))\right)
+\end{equation}
+Recalling the ordering of arrival times, 
+\[
+  T^* = \frac{a^*_1}{a^*_0}H^{p_1^*} \delta(t-(\tau^*_1-\tau^*_0)) + ...
+\]
+\[
+  T_1 = -\frac{a_1}{a_0}H^{p_1} \delta(t-(\tau_1-\tau_0)) + ...
+\]
+where the elided terms all involve $\delta$s located at later times.
 So
-$$
-J_0[m,w_*] \approx \sum_{{\bf x}_s,{\bf x}_r} (\tau[m_*]({\bf x}_s,{\bf x}_r)-\tau[m]({\bf x}_s,{\bf x}_r))^2
-$$
-ignoring the $\|tg_{...}\|^2$ term as being negligible for small enough $\epsilon$.
-
-In other words, the AWI objective is *exactly* an approximation to the travel-time tomography objective.
-And that's why it works - when geometric optics in the form given here is an accurate approximation to the Green's function.
-
-It remains to see what geometric optics has to say about the penalty objective $\tilde{J}_{\alpha}[m,w_*]$, how geometric optics might fail, and what happens to AWI when it does.
-
-\section{A Penalty Version of AWI}
-The adaptive kernel $u_0^{\epsilon}$ appearing in the definition \ref{eqn:tJz} is the solution of a regularized least squares problem.
-This observation suggests augmenting the regularized least squares objective by a penalty proportional to the right-hand side of \ref{eqn:tJz}:
-\begin{equation}
-  \label{eqn:Ja}
-  \Ja[m,u;d] = \frac{1}{2}\left(\|S[m,w_*]u-d\|_d^2 + \epsilon^2\|u\| + \alpha^2\|tu\|_m\right).
-\end{equation}
-Here $\|u\|^2 = \langle u, u \rangle$ is the (unweighted) $L^2$ norm, and $\|\cdot\|_m$ is an $(m,d)$-dependent (weighted) norm: with
 \[
-  u_0^{\epsilon}[m,d] = \mbox{arg min}_u \Jz[m,u;d],
-\]
-define
-\[
-  Wu (\bx_r,t;\bx_s) = \sum_{\bx_r,\bx_s}\frac{1}
-  {\int\,ds\,|u_0^{\epsilon}[m,d](\bx_r,s;\bx_s)|^2}u(\bx_r,t;\bx_s)
+  (\delta + T^*)*(\delta + T_1)(t) = \delta(t)  +  \frac{a^*_1}{a^*_0}
+  H^{p_1^*}\delta(t-(\tau^*_1-\tau^*_0)) -\frac{a_1}{a_0}H^p_1
+  \delta(t-(\tau_1-\tau_0)) + ...
 \]
@@
 \begin{equation}
-  \label{eqn:defmnorm}
-  \|u\|^2_m = \langle u, W u \rangle.
+  \label{eqn:notl2explicit}
+  u(t) = \frac{a^*_0}{a_0}\left(\delta(t - (\tau^*_0-\tau_0)) +
+    \frac{a^*_1}{a^*_0}H^{p_1^*}\delta(t-(\tau^*_1-\tau_0)) -
+    \frac{a_1}{a_0}H^{p_1}\delta(t-(\tau_1-\tau_0^*)) + ...\right)
 \end{equation}
-With this notation,
-\[
-  \tJz[m;d] = \frac{1}{2}\|tu^{\epsilon}_0[m;d]\|^2_m
-\]
-In the definition of $\tJz$, a quadratic is evaluated at the minimizer of $\Jz[m,u;d]$. This suggests the possibility of evaluating the quadratic objective $\Ja[m,u;d]$ at its own minimizer:
-\begin{equation}
-  \label{eqn:tJa}
-  \tJa[m;d] = \min_u \Ja[m,u,d] = \Ja[m,\ua[m;d];d]
-\end{equation}
-This is the so-called variable projection reduction of the quadratic $\Ja$.
-Note that the right-hand side of the definition of $\tJz$ is the last term on the right-hand side of $\Ja$, scaled by $\alpha^{-2}$ and evaluated at the minimizer of the sum of the first two terms, that is, $\Ja$ for $\alpha=0$.
+The elided terms involve only travel time differences other than the
+ones appearing here, so nothing cancels.
 
-\begin{theorem}
-  \label{thm:basic}
-Suppose that $S \in {\cal L}(U,D)$, the is a bounded linear maps from (Hilbert space) domain $U$ (``model space'') with inner product $\langle \cdot, \cdot \rangle_m$ to range $D$ with inner product $\langle \cdot, \cdot \rangle_d$.
-Assume that $S$ is coercive, that is, there exists $\epsilon >0$ so that $\|Su\|_d \ge \epsilon\|u\|_m$ for $u \in U$.
-Suppose also that $A \in {\cal L}(U,U)$, and that $d \in D$. Define the function $J_{\alpha}$ on $U$ by 
-\begin{equation}
-  \label{eqn:eq1}
-  J_{\alpha}(u) = \frac{1}{2}(\|Su-d\|^2_d + \alpha^2\|Au\|_m^2).
-\end{equation}
-Then for $\alpha \ge 0$, $J_{\alpha}$ possesses a unique minimizer $u_{\alpha} \in U$, for which
-\begin{equation}
-  \label{eqn:eq0}
-  \tilde{J}_{\alpha} = \min_{u \in U} J_{\alpha}(u) = J_{\alpha}(u_{\alpha}),
-\end{equation}
-and
-\begin{equation}
-  \label{eqn:eq4}
-  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tilde{J}_{\alpha}-\tilde{J}_0).= \frac{1}{2}\|Au_0\|^2
-\end{equation}
-This limit is uniform over bounded sets of $(S, A) \in {\cal L}(U,D) \times {\cal L}(U, U)$.
-\end{theorem}
+Observe that the travel-time differences $\tau^*_1-\tau_0$,
+$\tau_1-\tau_0^*$ do not tend to zero, though they do tend to the same
+limit as $m \rightarrow m^*$ (as do the amplitude quotients), and the
+Hilbert transform powers $p_1, p_1^*$ are the same as soon as $m$ is
+sufficiently close to $m^*$.
 
-For completeness, the proof is given in an appendix.
+Of courese, the adaptive filter just computed is not square
+integrable, and must be regularized for use in a mean-square
+construction like that of AWI or MSWI. That amounts essentially to
+replacing the $\delta$s in the above formulae by a square-integrable
+approximate Dirac function, of roughly a wavelength in width. In the
+preceding section, this approximate Dirac was $g$, so use the same
+notation here. The regularized version of $u(t)$ in
+\ref{eqn:notl2explicit} contains summands with multiples of $g(t-(\tau^*_1-\tau_0))$
+and $g(t-(\tau_1-\tau^*_0))$. These are approximately scaled by
+$\tau^*_1-\tau_0$ and $\tau_1-\tau^*_0$ respectively in the
+definitions \ref{eqn:MSWI} and \ref{eqn:tJz} of the MSWI and AWI
+objectives. For other choices of objective, they contribute similarly
+to the total. They will eventually begin to cancel as $m$ approaches
+$m^*$, however this cancellation happens only when the times are
+within the width of the approximate 
+Dirac waveform $g$ - that is, within a wavelength.
 
-This relation will be applied in the context of a model space $U$ possessing a reference Hilbert norm $\|\cdot\|$ and a family of equivalent norms $\|\cdot\|_m$ depending on a parameter $m \in M$, in which $M$ is an open subset of a Banach space $X$.
-The dependence is through a an operator-valued function $W$ on $M$, taking values in cone ${\cal B}(U)$ of bounded self-adjoint positive-definite operators on $U$.
-On the other hand, we assume that the norm on $D$ is fixed, independent of $M$, and treat it as a reference norm.
+This argument shows that when multiple arrivals carry significant energy, any
+inversion approach based on the source-receiver extension, similar to
+AWI or MSWI, will cycle-skip, that is, change rapidly when the
+estimated model approaches its optimal value.
 
-\begin{theorem}
-  \label{thm:uniform}
-Assume that $W: M \rightarrow {\cal B}(U)$ is of class $C^1$ and uniformly positive definite, and that the norm $\|\cdot\|_m$ is given by
-\begin{equation}
-  \label{eqn:normdef}
-  \|u\|_m^2 = \langle u, W[m] u \rangle.
-\end{equation}
-Suppose also that $S: M \rightarrow {\cal L}(U,D)$ is of class $C^1$ and uniformly coercive, and that $A \in {\cal L}(U,U)$ as before. For $\alpha \ge 0$, define $\tJa: M \rightarrow \bR$ by
-\[
-  \tilde{J}_{\alpha}[m] = \min_{u \in U}\frac{1}{2}(\|S[m]u-d\|^2_d + \alpha^2\|Au\|_m^2)
-\]
-\begin{equation}
-  \label{eqn:tjadef}
-  = \frac{1}{2}(\|S[m]u_{\alpha}[m]-d\|^2_d + \alpha^2\|Au_{\alpha}[m]\|_m^2)
-\end{equation}
-where
-\begin{equation}
-  \label{eqn:ua}
-  u_{\alpha}[m] = \mbox{arg min}_{u \in U}\frac{1}{2}(\|S[m]u-d\|^2_d + \alpha^2\|Au\|_m^2)
-\end{equation}
-for $m \in M$. Then $\tilde{J}_{\alpha}$ is of class $C^1(M)$, and
-\begin{equation}
-  \label{eqn:lim}
-  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tJa[m]-\tJz[m]).= \frac{1}{2}\|A\uz[m]\|_m^2.
-\end{equation}
-The limit is in the sense of $C^1(M)$, that is, both the left-hand side and its $m$ derivative converge to the right-hand side and its $m$ derivative, respectively.
-Moreover $u_{\alpha}$ is of class $C^1(M,U)$.
-\end{theorem}
+It should be emphasized that this phenomenon is primarily a feature of the
+source-receiver extension, rather than of the specific objective
+function defined in terms of it. In that respect, the source-receiver
+extension is similar to the common surface parameter extensions that
+have been studied for inversion of reflection data. Inversion methods
+based on these extensions also fail to avoid cycle-skipping in the presence of
+geometrical ray complexity \cite[]{geoprosp:2008}.
 
-\begin{proof} All conclusions but the last follow immediately from Theorem \ref{thm:basic} and the uniformity assumptions on $S$ and $W$.
-
-Since the norm in $D$ is fixed for the discussion, drop the subscript and treat it as a reference norm.
-Denote the adjoint with respect to the norms $\|\cdot\|_m$ on $U$, $\|\cdot\|_d = \|\cdot\|$ on $D$ by a superscript $*$, and with respect to $\|\cdot\|$ on $U$ by a superscript $T$.
-The normal equation defining $\ua$ is
-  \[
-    (S[m]^*S[m] + \alpha^2 A[m]^*A) \ua[m] = S[m]^*d
-  \]
-  \begin{equation}
-    \label{eqn:normal}
-= (W[m]^{-1}S[m]^TS[m] + \alpha^2 W[m]^{-1}A^TW[m]A)\ua[m]= W[m]^{-1}S[m]^Td
-\end{equation}
-or equivalently
-\begin{equation}
-  \label{eqn:normal1}
-  (S[m]^TS[m] + \alpha^2 A^TW[m]A)\ua[m] = S[m]^Td
-\end{equation}
-Since $S$ is uniformly coercive, and both $S$ and $W$ are of class $C^1$, the normal operator on the left hand side is boundedly invertible and of class $C^1$.
-Therefore the solution $\ua$ is of class $C^1$.
-
-[need to add proof of $C^1$ convergence.]
-\end{proof}
-
-Apply this relation in the context of the AWI penalty problem defined earlier.
-$M$ is an open set in $C^k(\Omega, \bR^2)$.
-$U$ is $\oplus_{\bx_r,\bx_s}L^2([-T_u,T_u])$, $D$ is the data space $\oplus_{\bx_r,\bx_s}L^2([0,T])$,.
-$S$ is the regularized modeling operator $(S[m,w_*],\epsilon I)^T$, $d$ is replaced by the $(d,0) \in D \times U$,  $W$ is the weight operator on $U$ defined in \ref{eqn:defmnorm}.
-
-As noted before [but it wasn't - fix!], stationary points of $\tJz$ are $\lambda(w_*)$-close to stationary points of the travel time mean-square error, in the single-arrival case.
-The result of this section implies that stationary points of $\tJa$ are close to those of $\tJz$.
-Therefore, at least for small $\alpha$, minimization of $\tJa$ produces acoustic models that closely match the traveltimes predicted in single-arrival data.
-
-\append{$\alpha \rightarrow 0$ scaled limit of penalty function}
-Define
-\begin{equation}
-  \label{eqn:eq0}
-  \tJa = \min_u J_{\alpha}(u) = J_{\alpha}(\ua)
-\end{equation}
-Since $S$ is positive definite, the minimum is well-defined for any $\alpha \ge 0$, and the minimizer $\ua \in U$ satisfies the normal equation.
-
-The claim to be established is that
-\begin{equation}
-  \label{eqn:eq4}
-  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tJa-\tJz).= \frac{1}{2}\|A\uz\|_m^2
-\end{equation}
-
-To see this, use the normal equation to write
-\[
-  \ua = (S^TS + \alpha^2 A^TA)^{-1}S^Td = (S^TS + \alpha^2 A^TA)^{-1}S^TS \uz
-\]
-\begin{equation}
-  \label{eqn:eq2}
-  = \uz - \alpha^2 (S^TS + \alpha^2 A^TA)^{-1}A^TA\uz = \uz-\alpha^2 \va
-\end{equation}
-Note that $\va = (S^TS + \alpha^2 A^TA)^{-1}A^TA\uz$ is uniformly bounded in $\alpha \ge 0$.
-
-Stuff the RHS of equation \ref{eqn:eq2} into the definition \ref{eqn:eq1} to obtain
-\begin{equation}
-  \label{eqn:eq3}
-  \tJa = \frac{1}{2}(\|S\uz-d\|_d^2 - 2 \alpha^2\langle S\uz-d, S\va\rangle_d + \alpha^2\|A\uz\|_m^2 + O(\alpha^4))
-\end{equation}
-The second term on the RHS of equation \ref{eqn:eq3} vanishes thanks to the normal equation, and the first term is precisely $\tJz$. The conclusion \ref{eqn:eq4} follows.
+\section{Conclusion}
+In this paper, I have shown that AWI is a good approximation to least
+squares travel time inversion of transmitted wave data, up to errors proportional to a typical
+data wavelength and to a regularization (``pre-whitening'')
+weight. However, the argument leading to this conclusion works only if
+the material model is assumed slowly varying so that geometric
+acoustics accurately describes the solution of the wave equation, and if
+sources and receivers are sufficiently close to one another to be
+connected by unique rays. The first condition may be ignored to some
+extend - the conclusion of this paper likely holds if the data is
+dominated by ballistic transmitted waves. However the second condition
+is essential: in the presence of multiple arrivals, the connection
+with travel time inversion breaks down, and AWI and similar algorithms
+become essentially as prone to cycle-skipping as is FWI.
 

### 2024-01-24 repro/papers/awinum/project/barf.py
more

@@ -1,0 +1,21 @@
+import os
+HOME = os.getenv('HOME')
+print(HOME)
+APPS = os.path.join(HOME,'Library')
+print(APPS)
+CWPROOT = os.path.join(APPS,'CWP/44r16')
+print(CWPROOT)
+RSFROOT = os.path.join(APPS,'madagascar/rsfroot')
+print(RSFROOT)
+TRIP = os.path.join(HOME,'work/trip')
+print(TRIP)
+PYRVL = os.path.join(TRIP,'pyrvl')
+print(PYRVL)
+PYTHONPATH = os.getenv('PYTHONPATH')
+print(PYTHONPATH)
+
+import data
+data.hello()
+
+
+

### 2024-01-24 repro/papers/awinum/project/datah.py

@@ -1,0 +1,28 @@
+import scenv
+import os
+import vcl
+import segyvc
+import rsfvc
+import asg
+
+bulksph = rsfvc.Space('m0h.rsf')
+datasph = segyvc.Space('hh.su')
+
+m0h = vcl.Vector(bulksp,'mhl.rsf')
+os.system('/bin/cp hh.su d0h.su')
+d0h = vcl.Vector(datasp, 'd0h.su')
+
+# homog model
+mhhom = vcl.Vector(bulksp,'mhhom.rsf')
+os.system('/bin/cp hh.su dhhom.su')
+dhhom  = vcl.Vector(datasp, 'dhhom.su')
+
+F = asg.fsbop(dom=bulksph, rng=datasph, \
+        buoyancy='bym0h.rsf', source_p='wh.su', \
+        order=2, sampord=1, nsnaps=20,\
+        cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
+        nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0)
+
+d0h.copy(F(m0h))
+dhhom.copy(F(mhhom))
+

### 2024-01-24 repro/papers/awinum/project/datal.py

@@ -1,0 +1,36 @@
+import scenv
+import os
+import vcl
+import segyvc
+import rsfvc
+import asg
+
+bulkspl = rsfvc.Space('m0l.rsf')
+dataspl = segyvc.Space('hl.su')
+
+m0l = vcl.Vector(bulkspl,'m0l.rsf')
+m1l = vcl.Vector(bulkspl,'m1l.rsf')
+os.system('/bin/cp hl.su d0l.su')
+os.system('/bin/cp hl.su d1l.su')
+d0l = vcl.Vector(dataspl, 'd0l.su')
+d1l = vcl.Vector(dataspl, 'd1l.su')
+
+# homog model
+mlhom = vcl.Vector(bulkspl,'mlhom.rsf')
+os.system('/bin/cp hl.su dlhom.su')
+dlhom  = vcl.Vector(dataspl, 'dlhom.su')
+
+F = asg.fsbop(dom=bulkspl, rng=dataspl, \
+        buoyancy='bym0l.rsf', source_p='wl.su', \
+        order=2, sampord=1, nsnaps=20,\
+        cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
+        nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0)
+
+d0l.copy(F(m0l))
+d1l.copy(F(m1l))
+dlhom.copy(F(mlhom))
+
+
+
+
+

### 2024-01-24 repro/papers/awinum/project/datal2s.py

@@ -1,0 +1,36 @@
+import scenv
+import os
+import vcl
+import segyvc
+import rsfvc
+import asg
+
+bulkspl = rsfvc.Space('m0l.rsf')
+dataspl2s = segyvc.Space('hl2s.su')
+
+m0l = vcl.Vector(bulkspl,'m0l.rsf')
+m1l = vcl.Vector(bulkspl,'m1l.rsf')
+os.system('/bin/cp hl2s.su d0l2s.su')
+os.system('/bin/cp hl2s.su d1l2s.su')
+d0l2s = vcl.Vector(dataspl2s, 'd0l2s.su')
+d1l2s = vcl.Vector(dataspl2s, 'd1l2s.su')
+
+# homog model
+mlhom = vcl.Vector(bulkspl,'mlhom.rsf')
+os.system('/bin/cp hl2s.su dlhom2s.su')
+dlhom2s  = vcl.Vector(dataspl2s, 'dlhom2s.su')
+
+F = asg.fsbop(dom=bulkspl, rng=dataspl2s, \
+        buoyancy='bym0l.rsf', source_p='wl2s.su', \
+        order=2, sampord=1, nsnaps=20,\
+        cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
+        nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0)
+
+d0l2s.copy(F(m0l))
+d1l2s.copy(F(m1l))
+dlhom2s.copy(F(mlhom))
+
+
+
+
+

### 2024-01-24 repro/papers/awinum/project/datasim.py

@@ -1,0 +1,45 @@
+import scenv
+import os
+import vcl
+import segyvc
+import rsfvc
+import asg
+
+import sys
+ORDER   = int(sys.argv[1])
+SAMPORD = int(sys.argv[2])
+NSNAPS  = int(sys.argv[3])
+CFL     = float(sys.argv[4])
+CMIN    = float(sys.argv[5])
+CMAX    = float(sys.argv[6])
+DMIN    = float(sys.argv[7])
+DMAX    = float(sys.argv[8])
+NL1     = int(sys.argv[9])
+NL2     = int(sys.argv[10])
+NR1     = int(sys.argv[11])
+NR2     = int(sys.argv[12])
+PMLAMPL = float(sys.argv[13])
+
+dfile = sys.argv[14]
+wfile = sys.argv[15]
+mfile = sys.argv[16]
+bfile = sys.argv[17]
+
+bulksp = rsfvc.Space(mfile)
+datasp = segyvc.Space(dfile)
+
+m = vcl.Vector(bulksp, mfile)
+d = vcl.Vector(datasp, dfile)
+
+F = asg.fsbop(dom=bulksp, rng=datasp, \
+        buoyancy=bfile, source_p=wfile, \
+        order=ORDER, sampord=SAMPORD, nsnaps=NSNAPS,\
+        cfl=CFL, cmin=CMIN, cmax=CMAX,dmin=DMIN, dmax=DMAX,\
+        nl1=NL1, nr1=NR1, nl2=NL2, nr2=NR2, pmlampl=PMLAMPL)
+
+d.copy(F(m))
+
+
+
+
+

### 2024-01-24 repro/papers/awinum/project/files.py

@@ -1,0 +1,59 @@
+import scenv
+import data
+import linalg
+
+NTU = [251,501]
+NTD = [626,1251]
+NX  = [401, 801]
+NZ  = [201, 401]
+DELRTW = [-1000, -500]
+DT  = [8, 4]
+DX  = [20, 10]
+F1  = [1.0, 2.0]
+F2  = [2.5, 5.0]
+F3  = [7.5, 15.0]
+F4  = [12.5, 25.0]
+
+# adaptive filter prototype
+data.rechdr(file='ul.su',nt=NTU[0],dt=DT[0],
+            ntr=201,rx=2000,rz=1000,sx=4200,sz=3000,drx=20,delrt=DELRTW[0])
+data.rechdr(file='uh.su',nt=NTU[1],dt=DT[1],
+            ntr=201,rx=2000,rz=1000,sx=4200,sz=3000,drx=20,delrt=DELRTW[1])
+
+wsc=1.0e+3
+data.bpfilt(file='wl.su',nt=NTU[0],dt=DT[0],s=1.0,
+            f1=F1[0],f2=F2[0],f3=F3[0],f4=F4[0],
+            sx=4200,sz=3000)
+data.bpfilt(file='wh.su',nt=NTU[1],dt=DT[1],s=1.0,
+            f1=F1[1],f2=F2[1],f3=F3[1],f4=F4[1],
+            sx=4200,sz=3000)
+linalg.scale('wl.su',wsc)
+linalg.scale('wh.su',wsc)
+
+# create zero data files with same source position, rz=500, rx=[2000,6000]
+data.rechdr(file='hl.su',nt=NTD[0],dt=DT[0],ntr=201,
+            rx=2000.0,rz=1000.0,sx=4200,sz=3000,drx=20.0,delrt=0)
+data.rechdr(file='hh.su',nt=NTD[1],dt=DT[1],ntr=201,
+            rx=2000.0,rz=1000.0,sx=4200,sz=3000,drx=20.0,delrt=0)
+
+# bulk modulus with smaller focussing lens
+data.model(bulkfile='m0l.rsf', bulk=4.0, nx=NX[0], nz=NZ[0],
+   dx=DX[0], dz=DX[0], lensfac=0.7)
+data.model(bulkfile='m0h.rsf', bulk=4.0, nx=NX[1], nz=NZ[1],
+   dx=DX[1], dz=DX[1], lensfac=0.7)
+
[…]

### 2024-01-24 repro/papers/awinum/project/headers.py

@@ -1,0 +1,40 @@
+import scenv
+import data
+import linalg
+
+import sys
+
+NTU = int(sys.argv[1])
+NTD = int(sys.argv[2])
+DELRTW = float(sys.argv[3])
+DT  = float(sys.argv[4])
+F1  = float(sys.argv[5])
+F2  = float(sys.argv[6])
+F3  = float(sys.argv[7])
+F4  = float(sys.argv[8])
+NS  = int(sys.argv[9])
+DSX = float(sys.argv[10])
+
+u = sys.argv[11]
+w = sys.argv[12]
+h = sys.argv[13]
+
+### 2 shot headers
+# adaptive filter prototype
+data.rechdr(file=u,nt=NTU,dt=DT,
+            ntr=201,rx=2000,rz=1000,sx=4200,sz=3000,drx=20,delrt=DELRTW,
+            nshot=NS, dsx=DSX)
+
+wsc=1.0e+3
+data.bpfiltgather(file=w,nt=NTU,dt=DT,s=1.0,
+            f1=F1,f2=F2,f3=F3,f4=F4,
+            ntr=NS,sxstart=4200,szstart=3000,dsx=DSX,dsz=0)
+linalg.scale(w,wsc)
+
+# create zero data files with same source position, rz=500, rx=[2000,6000]
+data.rechdr(file=h,nt=NTD,dt=DT, ntr=201,
+            rx=2000.0,rz=1000.0,sx=4200,sz=3000,drx=20.0,delrt=0,
+            nshot=NS, dsx=DSX)
+
+
+

### 2024-01-24 repro/papers/awinum/project/models.py

@@ -1,0 +1,17 @@
+import scenv
+import data
+import linalg
+
+import sys
+
+NX  = int(sys.argv[1])
+NZ  = int(sys.argv[2])
+DX  = float(sys.argv[3])
+FAC = float(sys.argv[4])
+RAD = float(sys.argv[5])
+
+m = sys.argv[6]
+
+data.model(bulkfile=m, bulk=4.0, nx=NX, nz=NZ,
+   dx=DX, dz=DX, lensfac=FAC, lensradd=RAD)
+

### 2024-01-24 repro/papers/awinum/project/mswi.py

@@ -1,0 +1,78 @@
+import os
+import sys
+
+import linalg
+import data
+import vcl
+import vcalg
+import vpm
+import segyvc
+import rsfvc
+import asg
+import awi
+
+ORDER=2
+SAMPORD=1
+NSNAPS=20
+CFL=0.5
+CMIN=1.0
+CMAX=3.0
+DMIN=0.8
+DMAX=3.0
+NL1=250
+NL2=250
+NR1=250
+NR2=250
+PMLAMPL=1.0
+
+KMAX=1000
+VERBOSE=2
+
+def mswigrad(bulk, buoy, wavelet, data, filt, grad. alpha, sigma, rho):
+    try: 
+
+        bulksp = rsfvc.Space(bulk)
+        datasp = segyvc.Space(data)
+        filtsp = segyvc.Space(filt)
+
+        m = vcl.Vector(bulksp, bulk)
+        d = vcl.Vector(datasp, data)
+        g = vcl.Vector(bulksp, grad)
+        u = vcl.Vector(filtsp, filt)
+        
+        F = asg.fsbop(dom=bulksp, rng=datasp, \
+                    buoyancy=buoy, source_p=wavelet, \
+                    order=ORDER, sampord=SAMPORD, nsnaps=NSNAPS,\
+                    cfl=CFL, cmin=CMIN, cmax=CMAX, dmin=DMIN, dmax=DMAX,\
+                    nl1=NL1, nr1=NR1, nl2=NL2, nr2=NR2, pmlampl=PMLAMPL)
+
+        j = awi.mswi(dom=filtsp, sim=F, mod=m, data=d, alpha=alpha,
+                         sigma=sigma, kmax=KMAX, rho=rho, verbose=VERBOSE)
+
+        u.copy(j.filter())
+        g.copy(j.gradient())
+
+    except Exception as ex:
+        print(ex)
+        raise Exception('called from mswigrad')
+
+try:
+        
+    # argv1 = rsf file defining model space and bulk mod
+    # argv2 = rsf file defining buoyancy
+    # argv3 = su file defining wavelet
+    # argv4 = su file defining data space and data
+    # argv5 = su file defining adaptive filter space
+    # argv6 = adaptive filter (out)
+    # argv7 = rsf file defining gradient (out)
+    # argv8 = alpha
+    # argv9 = sigma
+    # argv10 = rho
+
[…]

### 2024-02-14 repro/papers/awinum/data/datasim.py
initial

@@ -1,0 +1,44 @@
+import scenv
+import os
+import vcl
+import segyvc
+import rsfvc
+import asg
+
+import sys
+
+# order=2, sampord=1, nsnaps=20,\
+#                cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
+#                nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0,
+#                boundstest=True, noisy=False
+
+def asgsim(bulkmod, data_p, buoyancy, source_p, order=2, sampord=1, nsnaps=20,\
+               cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
+               nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0,
+               boundstest=True, noisy=False):
+    
+    try:
+    
+        bulksp = rsfvc.Space(bulkmod)
+        datasp = segyvc.Space(data_p)
+
+        m = vcl.Vector(bulksp, bulkmod)
+        d = vcl.Vector(datasp, data_p)
+
+        F = asg.fsbop(bulksp, datasp, buoyancy, source_p, \
+                          order, sampord, nsnaps, cfl, cmin, cmax, dmin, dmax,\
+                          nl1, nr1, nl2, nr2, pmlampl, boundstest, noisy \
+                          )
+
+        d.copy(F(m))
+        
+    except Exception as ex:
+        print(ex)
+        raise Exception('called from sim')
+
+if __name__ == '__main__':
+    args = ", ".join(sys.argv[1:])
+    cmd = 'asgsim(' + args + ')'
+    exec(cmd)
+
+

### 2024-02-15 repro/papers/awinum/project/barf.py
removed adBuffer, adStack

@@ -1,21 +1,0 @@
-import os
-HOME = os.getenv('HOME')
-print(HOME)
-APPS = os.path.join(HOME,'Library')
-print(APPS)
-CWPROOT = os.path.join(APPS,'CWP/44r16')
-print(CWPROOT)
-RSFROOT = os.path.join(APPS,'madagascar/rsfroot')
-print(RSFROOT)
-TRIP = os.path.join(HOME,'work/trip')
-print(TRIP)
-PYRVL = os.path.join(TRIP,'pyrvl')
-print(PYRVL)
-PYTHONPATH = os.getenv('PYTHONPATH')
-print(PYTHONPATH)
-
-import data
-data.hello()
-
-
-

### 2024-02-15 repro/papers/awinum/project/datah.py

@@ -1,28 +1,0 @@
-import scenv
-import os
-import vcl
-import segyvc
-import rsfvc
-import asg
-
-bulksph = rsfvc.Space('m0h.rsf')
-datasph = segyvc.Space('hh.su')
-
-m0h = vcl.Vector(bulksp,'mhl.rsf')
-os.system('/bin/cp hh.su d0h.su')
-d0h = vcl.Vector(datasp, 'd0h.su')
-
-# homog model
-mhhom = vcl.Vector(bulksp,'mhhom.rsf')
-os.system('/bin/cp hh.su dhhom.su')
-dhhom  = vcl.Vector(datasp, 'dhhom.su')
-
-F = asg.fsbop(dom=bulksph, rng=datasph, \
-        buoyancy='bym0h.rsf', source_p='wh.su', \
-        order=2, sampord=1, nsnaps=20,\
-        cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
-        nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0)
-
-d0h.copy(F(m0h))
-dhhom.copy(F(mhhom))
-

### 2024-02-15 repro/papers/awinum/project/datal.py

@@ -1,36 +1,0 @@
-import scenv
-import os
-import vcl
-import segyvc
-import rsfvc
-import asg
-
-bulkspl = rsfvc.Space('m0l.rsf')
-dataspl = segyvc.Space('hl.su')
-
-m0l = vcl.Vector(bulkspl,'m0l.rsf')
-m1l = vcl.Vector(bulkspl,'m1l.rsf')
-os.system('/bin/cp hl.su d0l.su')
-os.system('/bin/cp hl.su d1l.su')
-d0l = vcl.Vector(dataspl, 'd0l.su')
-d1l = vcl.Vector(dataspl, 'd1l.su')
-
-# homog model
-mlhom = vcl.Vector(bulkspl,'mlhom.rsf')
-os.system('/bin/cp hl.su dlhom.su')
-dlhom  = vcl.Vector(dataspl, 'dlhom.su')
-
-F = asg.fsbop(dom=bulkspl, rng=dataspl, \
-        buoyancy='bym0l.rsf', source_p='wl.su', \
-        order=2, sampord=1, nsnaps=20,\
-        cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
-        nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0)
-
-d0l.copy(F(m0l))
-d1l.copy(F(m1l))
-dlhom.copy(F(mlhom))
-
-
-
-
-

### 2024-02-15 repro/papers/awinum/project/datal2s.py

@@ -1,36 +1,0 @@
-import scenv
-import os
-import vcl
-import segyvc
-import rsfvc
-import asg
-
-bulkspl = rsfvc.Space('m0l.rsf')
-dataspl2s = segyvc.Space('hl2s.su')
-
-m0l = vcl.Vector(bulkspl,'m0l.rsf')
-m1l = vcl.Vector(bulkspl,'m1l.rsf')
-os.system('/bin/cp hl2s.su d0l2s.su')
-os.system('/bin/cp hl2s.su d1l2s.su')
-d0l2s = vcl.Vector(dataspl2s, 'd0l2s.su')
-d1l2s = vcl.Vector(dataspl2s, 'd1l2s.su')
-
-# homog model
-mlhom = vcl.Vector(bulkspl,'mlhom.rsf')
-os.system('/bin/cp hl2s.su dlhom2s.su')
-dlhom2s  = vcl.Vector(dataspl2s, 'dlhom2s.su')
-
-F = asg.fsbop(dom=bulkspl, rng=dataspl2s, \
-        buoyancy='bym0l.rsf', source_p='wl2s.su', \
-        order=2, sampord=1, nsnaps=20,\
-        cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
-        nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0)
-
-d0l2s.copy(F(m0l))
-d1l2s.copy(F(m1l))
-dlhom2s.copy(F(mlhom))
-
-
-
-
-

### 2024-02-15 repro/caam641/intro.tex
cleaned up

@@
 \section{Goals}
-This book will explore some of the core concepts and methods of industrial seismic imaging and inversion from a mathematical point of view.
-The content arose from my efforts (and those of many others) to explain {\em precisely} how many clever and effective processing algorithms originate in the physics of waves.
-Evidently, either such explanation is possible, or our understanding of seismic waves is lacking in some important respect.
+This book will explore some of the core concepts and methods of active-source seismic imaging and inversion from a mathematical point of view.
+The content arose from my efforts (and those of many others) to explain {\em precisely} the origin in the physics of waves of many clever and effective processing algorithms, widely used in industrial and scientific practice.
+Evidently, either such explanation is possible, or our understanding of seismic waves is lacking in important respects.
 To some extent this project has been successful: wave theory - linear elastodynamics -  actually does explain {\em precisely} why some important seismic imaging and inversion methods work.
-Seismic processing is a huge field, and the account given here is not comprehensive by any means.
+Even better, the theoretical concepts required to understand modern imaging methods are themselves (relatively) modern - the necessary mathematics has developed in parallel with (and largely independent of) engineering/scientific methodology.
+Rich and beautiful connections have been drawn in both directions: imaging algorithms based on data-driven insight have inspired new mathematics, which has returned the favor by suggesting improved (or, in some cases, brand-new) imaging algorithms.
+
+Active-source seismic surveys use various human-made and -controlled devices, or {\em sources}, to stimulate stress/displacement (sound, or more properly seismic) waves propagating through the earth, and records the resulting displacement or stress fields at positions on or near the earth's surface.
+The recordings encode the acoustic structure of the earth through which the waves has passed on their journey from sources to recording devices, or {\em receivers}.
+Seismic imaging aims to decode the earth's structure, or at least those features of it to which sound waves are sensitive, from the recorded data.
+
+This branch of earth science originated roughly one hundred years ago, in prospecting for ore bodies before the first World War, and in artillery ranging methods developed during it.
+Since the early successes in locating petroleum reservoirs along the Gulf of Mexico coast in the 1920's, active-source seismology has become the primary exploration tool of the oil industry, and has also supplied useful methodology to civil and environmental engineering.
+The topics that form the focus of this book, while at least implicit in the early development of the subject, flowered after the introduction of digital computing in the exploration industry, in the 50's and early 60's.
+Each subsequent twenty-year period saw the development of a distinct collection of topics:
+\begin{itemize}
+\item 1960's through 1970's: trace-by-trace processing: deconvolution, normal moveout, stack, post-stack migration (``classic imaging'');
+\item 1980's through 1990's: pre-stack and reverse-time migration (``linearized inversion'');
+\item last 20 years: full waveform inversion (``nonlinear inversion'').
+\end{itemize}
+This book consists of three parts, each devoted to one of the topics listed above.
+The list above is full of jargon; one of the main goals of this book is to explain each term above (and many others used in this subject) in terms of the mathematics of waves.
+The story told here is not comprehensive by any means.
 However, in the pages to follow you will find a few complete accounts, starting in basic linear elastodynamics and arriving at some practical and important imaging and inversion algorithms.
@@
 In particular, it is about both theory and practice, with practice represented by computational examples illustrating the theory, using open source software packages and public domain data, both synthetic and field.
+
 These examples take the form of {\em reproducible computational research}, another pioneering concept of Claerbout's.
-The reader can rebuild the results from raw data and code by executing a few simple commands.
+The reader can rebuild the results from raw data and code by executing a few simple commands, included as auxiliary material with the book.
 Most importantly, the reader can also modify the examples to explore beyond the content of the text.
+This computational infrastructure is described in detail in Appendix A, and rests on three open source software packages (Seismic Unix, Madagascar, and TRIP), which in turn depend on Python, C/C++, and Unix.
+Madagascar provides the computational ``glue'' tying together tex, data, and code.
+The chapters form a directory tree of Madagascar papers, each one with examples and resulting figures produced from Python scripts in a project subdirectory.
+The participants can rebuild every one of these papers from scratch, and the scripts can (and should!) serve as departure points for additional projects.
 
-\section{Syllabus}
-This book's current incarnation is a set of notes for a course, CAAM 641, Spring 2017 edition.
-The structure of this course it a bit unusual, and requires some explanation.
-It will divide into three segments, each consisting of several in-person lectures and work sessions, followed by an internet-based project period.
-Each segment consists of two parts, one oriented towards data processing, the other concerned with mathematical foundations.
-The two parts actually support each other but develop to some extent independently, with the final story emerging over the course.
-These notes mostly support the processing part.
-The mathematical part will read through some recent (and not-so-recent) papers, or notes in those cases where satisfactory references are not available.
-
-THe three segments are:
+\section{Overview}
 
 \begin{itemize}
-\item {\bf Basic Imaging} (Chapters 2 - 6) - starting with an imaging exercise, using 2D marine survey data: RMS velocity estimation, NMO correction of CMP gathers, stack, and post-stack time and depth migration.
+\item {\bf Part 1: Basic Imaging.} Starts with an imaging exercise, using 2D marine survey data: RMS velocity estimation, NMO correction of CMP gathers, stack, and post-stack time and depth migration.
 The object of this part of the course is to identify the origin in wave theory of each process used in this exercise.
@@
 
-The mathematical focus of this segment is the basic theory of the wave equation, following the lead of J.-L.
-Lions.
-We will work through existence of finite energy solutions, regularity as a function of the right hand side and of the coefficients, and the examples of acoustics and elasticity.
+\item {\bf Part 2: Linear Inversion.} Prestack depth imaging via diffraction sums (Kirchhoff migration), Generalized Radon Transform inversion for acoustics and elasticity, Reverse Time Migration, Least Squares Migration (iterative and true amplitude migration).
+This time we start with the theory, as the computational examples will use Reverse Time and True Amplitude migration, but the justification for these is essentially the theory of the Generalized Radon Transform.
+We end by applying these techniques to the 2D marine survey of Part 1.
 
-At the end we will have an understanding at various levels of the Born approximation (when it is justified!), and of its collaboration with high frequency asymptotics to produce (some) post-stack imaging algorithms.
-
-\item {\bf Linear Inversion:} prestack depth imaging via diffraction sums (Kirchhoff migration), Generalized Radon Transform inversion for acoustics and elasticity, Reverse Time Migration, Least Squares Migration (iterative and true amplitude migration).
-This time we start with the theory, as the computational examples will use Reverse Time and True Amplitude migration, but the justification for these is essentially the theory of the Generalized Radon Transform.
-We end by applying these techniques to the 2D marine survey of segment 1.
-
-The mathematical part of this segment has two goals: first, a rapid survey of the convergence theory of finite difference methods, used throughout the course (and seismology) to generate examples and process data; second, filling in some of the mathematical holes in high-frequency asymptotics as developed in the other part of the course, in particular extension of the asymptotic description beyond caustics.
-
-This section will complete the analysis of asymptotics and its role in post-stack imaging, and extend that understanding to pre-stacck imaging.
-
-\item {\bf Nonlinear Inversion:} Theory of reflection and transmission tomography, relation to migration velocity analysis in extended time and image domains.
+\item {\bf Part 3: Nonlinear Inversion.} Theory of reflection and transmission tomography, relation to migration velocity analysis in extended time and image domains.
 Inversion beyond single scattering - full waveform inversion.
@@
 
-Mathematical topics: extended modeling and inversion, source and medium extensions, analysis of convergence for inversion via iterative solution of inner problems. 
 \end{itemize}
 
-Examples presented in course materials use Seismic Unix, Madagascar, and TRIP software packages to realize the concepts introduced in class.
-The course notes take the form of a directory tree of Madagascar papers, each one with any examples and resulting figures produced from Python scripts in a project subdirectory.
-The participants can rebuild every one of these papers from scratch, and the scripts can (and should!) serve as departure points for additional projects.
-
-I suggest questions at the end of each chapter that can serve as nuclei of projects, and welcome suggestions for others.
-
-\section{Schedule}
-The class schedule divides into three periods, each with two weeks of class meetings and three weeks of internet-based project work.
-Classes will meet on Tuesdays and Thursdays from 1600 to 1830 in Duncan Hall room 3076 on the Rice campus.
-A tentative schedule for class meetings is: January 10, 12, 17, and 19; February 14, 16, 21, and 23; and March 21, 23, 28, and 30.
-I will communicate with participants during the intervening project periods by email and, when appropriate, by skype.
-
-\section{Expectations}
-In view of the advanced graduate nature of this course, no general performance metrics are appropriate: each participant will determine their level of involvement and expectations for the course.
+\bibliographystyle{seg}
+\bibliography{../bib/masterref}

### 2024-02-15 repro/caam641/intro/paper.tex
cleaned up

@@
-\title{0.0: Introduction}
-\author{William W. Symes}
-\address{The Rice Inversion Project,
-Rice University,
-Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}
-
-\maketitle
-\parskip 12pt
-
-\lefthead{Symes}
-\righthead{CAAM 641 Section 0.0}
-
-\begin{abstract}
-For Spring semester 2017, CAAM 641 will be devoted to the foundations and algorithms of seismic imaging and inversion.
-\end{abstract}
+\title{Introduction}
 
 \section{Goals}
-This course will explore some of the core concepts and methods of seismic imaging and inversion from a mathematical point of view. The course will
+This book will explore some of the core concepts and methods of active-source seismic imaging and inversion from a mathematical point of view.
+The content arose from my efforts (and those of many others) to explain {\em precisely} the origin in the physics of waves of many clever and effective processing algorithms, widely used in industrial and scientific practice.
+Evidently, either such explanation is possible, or our understanding of seismic waves is lacking in important respects.
+To some extent this project has been successful: wave theory - linear elastodynamics -  actually does explain {\em precisely} why some important seismic imaging and inversion methods work.
+Even better, the theoretical concepts required to understand modern imaging methods are themselves (relatively) modern - the necessary mathematics has developed in parallel with (and largely independent of) engineering/scientific methodology.
+Rich and beautiful connections have been drawn in both directions: imaging algorithms based on data-driven insight have inspired new mathematics, which has returned the favor by suggesting improved (or, in some cases, brand-new) imaging algorithms.
+
+Active-source seismic surveys use various human-made and -controlled devices, or {\em sources}, to stimulate stress/displacement (sound, or more properly seismic) waves propagating through the earth, and records the resulting displacement or stress fields at positions on or near the earth's surface.
+The recordings encode the acoustic structure of the earth through which the waves has passed on their journey from sources to recording devices, or {\em receivers}.
+Seismic imaging aims to decode the earth's structure, or at least those features of it to which sound waves are sensitive, from the recorded data.
+
+This branch of earth science originated roughly one hundred years ago, in prospecting for ore bodies before the first World War, and in artillery ranging methods developed during it.
+Since the early successes in locating petroleum reservoirs along the Gulf of Mexico coast in the 1920's, active-source seismology has become the primary exploration tool of the oil industry, and has also supplied useful methodology to civil and environmental engineering.
+The topics that form the focus of this book, while at least implicit in the early development of the subject, flowered after the introduction of digital computing in the exploration industry, in the 50's and early 60's.
+Each subsequent twenty-year period saw the development of a distinct collection of topics:
+\begin{itemize}
+\item 1960's through 1970's: trace-by-trace processing: deconvolution, normal moveout, stack, post-stack migration (``classic imaging'');
+\item 1980's through 1990's: pre-stack and reverse-time migration (``linearized inversion'');
+\item last 20 years: full waveform inversion (``nonlinear inversion'').
+\end{itemize}
+This book consists of three parts, each devoted to one of the topics listed above.
+The list above is full of jargon; one of the main goals of this book is to explain each term above (and many others used in this subject) in terms of the mathematics of waves.
+The story told here is not comprehensive by any means.
+However, in the pages to follow you will find a few complete accounts, starting in basic linear elastodynamics and arriving at some practical and important imaging and inversion algorithms.
+
+These notes owe a great deal, in structure, content, and spirit, to Jon Claerbout.
+In preparing the text, I re-read the introduction to Jon's book {\em Image Estimation by Example}.
+In Jon's terms, you might say that this course is about ``the theory of the similarity between theory and practice'' (an emphasis that would clearly not appeal to Jon!).
+In particular, it is about both theory and practice, with practice represented by computational examples illustrating the theory, using open source software packages and public domain data, both synthetic and field.
+
+These examples take the form of {\em reproducible computational research}, another pioneering concept of Claerbout's.
+The reader can rebuild the results from raw data and code by executing a few simple commands, included as auxiliary material with the book.
+Most importantly, the reader can also modify the examples to explore beyond the content of the text.
+This computational infrastructure is described in detail in Appendix A, and rests on three open source software packages (Seismic Unix, Madagascar, and TRIP), which in turn depend on Python, C/C++, and Unix.
+Madagascar provides the computational ``glue'' tying together tex, data, and code.
+The chapters form a directory tree of Madagascar papers, each one with examples and resulting figures produced from Python scripts in a project subdirectory.
+The participants can rebuild every one of these papers from scratch, and the scripts can (and should!) serve as departure points for additional projects.
+
+\section{Overview}
 
 \begin{itemize}
-\item emphasize the close correspondence between mathematical foundations and practical algorithms, while allowing the participants to concentrate their attention on one or the other (or both) of these aspects, and
-\item offer structured opportunities to exercise and explore these concepts computationally, using open source software packages and public domain data, both synthetic and field.
-\end{itemize}
-
-\section{Syllabus}
-This course will divide into three segments, each consisting of several in-person lectures and work sessions, followed by internet-based project periods.
-Each segment consists of two parts, one oriented towards data processing, the other concerned with mathematical foundations.
-The two parts have points of contact but are not necessarily parallel.
-I will provide notes as needed, largely for the processing part.
-The mathematical part will read through some recent (and not-so-recent) papers, or notes in those cases where satisfactory references are not available.
-
-\begin{itemize}
-\item {\bf Basic Imaging:} We will begin with basic processing exercise, using 2D marine survey data: NMO correction of CMP gathers, stack, and post-stack time and depth migration.
+\item {\bf Part 1: Basic Imaging.} Starts with an imaging exercise, using 2D marine survey data: RMS velocity estimation, NMO correction of CMP gathers, stack, and post-stack time and depth migration.
 The object of this part of the course is to identify the origin in wave theory of each process used in this exercise.
-Two approximations to solution of the wave equation are central to this project: linearization, aka single scattering, aka the Born approximation, and geometric optics, aka high frequency asymptotics, aka ray theory.
+Two approximations to solution of the wave equation are central to this story: linearization, aka single scattering, aka the Born approximation, and geometric optics, aka high frequency asymptotics, aka ray theory.
 Applied to ``locally layered'' modeling, these lead to the convolutional model, basic time processing of seismic reflection surveys (deconvolution, statics, NMO, stack), and NMO based velocity analysis.
@@
 
-The mathematical focus of this segment is the basic theory of the wave equation, following the lead of J.-L.
-Lions.
-The main reference is \cite{BlazekStolkSymes:13}, which is made available on the web site.
-We will work through existence of finite energy solutions, regularity as a function of the right hand side and of the coefficients, and the examples of acoustics and elasticity.
+\item {\bf Part 2: Linear Inversion.} Prestack depth imaging via diffraction sums (Kirchhoff migration), Generalized Radon Transform inversion for acoustics and elasticity, Reverse Time Migration, Least Squares Migration (iterative and true amplitude migration).
+This time we start with the theory, as the computational examples will use Reverse Time and True Amplitude migration, but the justification for these is essentially the theory of the Generalized Radon Transform.
+We end by applying these techniques to the 2D marine survey of Part 1.
 
-\item {\bf Linear Inversion:} prestack depth imaging via diffraction sums (Kirchhoff migration), Generalized Radon Transform inversion for acoustics and elasticity, Reverse Time Migration, Least Squares Migration (iterative and true amplitude migration).
-This time we start with the theory, as the main examples will use Reverse Time and True Amplitude migration, but the justification for these is essentially the theory of the Generalized Radon Transform.
-We end by applying these techniques to the 2D marine survey of segment 1.
-
-The mathematical part of this segment has two goals: first, a rapid survey of the convergence theory of finite difference methods, used throughout the course (and seismology) to generate examples and process data; second, to extend high-frequency asymptotics beyond caustics.
-
-\item {\bf Nonlinear Inversion:} Theory of reflection and transmission tomography, relation to migration velocity analysis in extended time and image domains.
+\item {\bf Part 3: Nonlinear Inversion.} Theory of reflection and transmission tomography, relation to migration velocity analysis in extended time and image domains.
 Inversion beyond single scattering - full waveform inversion.
@@
 
-Mathematical topics: extended modeling and inversion, source and medium extensions, analysis of convergence for inversion via iterative solution of inner problems. 
 \end{itemize}
 
-Examples presented in course materials will use Seismic Unix, Madagascar, and TRIP software packages to realize the concepts introduced in class.
-In fact, the course notes are organized in the form of directory tree of Madagascar papers, each one with any examples and resulting figures produced from Python scripts in a project subdirectory.
-The participants can rebuild every one of these papers from scratch, and the scripts should serve as departure points for additional projects.
+\section{Reading}
+Lots of good background reading exists for the topics of this course, some with links to codes and computational exercises, and I will refer participants to them whenever appropriate.
+Probably the closest in ``Courant and Hilbert'' spirit is Laurent Demanet's 18.325 class notes \cite[]{Demanet:325notes}.
+These were written for a graduate course in mathematics at MIT, and it shows: topics such as NMO do not appear, but the foundations of imaging theory and the adjoint state method, amongst other things, receive an elegant and complete treatment.
+The now-classic book by Norm Bleistein, Jack Cohen, and John Stockwell \cite[]{BleisteinCohenStockwell:01} gives a very thorough treatment of high-frequency asymptotics and ``Kirchhoff'' imaging.
+Jerry Schuster's book on seismic imaging basics \cite[]{Schuster:10} is an excellent overview of fundamental geophysical and data processing concepts, with Matlab exercises.
+Gary Margrave's book \cite[]{Margrave:book} gives a comprehensive account of seismic data processing, with links to the extensive CREWES Matlab codes.
 
-I suggest questions in each paper that can serve as nuclei of projects, and welcome suggestions for others.
+Full waveform inversion (FWI) is a new enough (as a topic of widespread interest) that comprehensive references to complement the classic 1987 Tarantola book (republished by SIAM as \cite[]{Tarantola:05}) are only recently beginning to appear.
+See for example \cite{Fichtner:10}.
+FWI is well-known to be only locallly convergent (a phenomenon that this course will illustrate, and to some extent explain).
+In some sense you have to ``know the answer before you ask the question''.
+Amongst many approaches that have been suggested for globalizing FWI convergence, I will dwell on extended inversion, an idea borrowed from velocity analysis.
+The literature on extended inversion has become fairly large.
+The reveiw paper \cite{geoprosp:2008} includes an extensive bibliography up to about 2007.
+Course reading for this segment will include this review paper and several papers appearing in the last 12 months.
 
-\section{Schedule}
-The class schedule divides into three periods, each with two weeks of class meetings and three weeks of internet-based project work.
-Classes will meet on Tuesdays and Thursdays from 1600 to 1830 in Duncan Hall room 3076 on the Rice campus.
-A tentative schedule for class meetings is: January 10, 12, 17, and 19; February 14, 16, 21, and 23; and March 21, 23, 28, and 30.
-I will communicate with participants during the intervening project periods by email and, when appropriate, by skype.
+Excellent general texts on geophysical prospecting, for example those by \cite{Yil:01} and \cite{DobSav:88}, set the topics discussed here in a wider context.
 
-\section{Expectations}
-In view of the advanced graduate nature of this course, no general performance metrics are appropriate: each participant will determine their level of involvement and expectations for the course.
+For the fundamental properties of wave equations and their solutions, the main reference is \cite{BlazekStolkSymes:13}, an open access paper available through the journal's web site, {\tt http://iopscience.iop.org/journal/0266-5611}.
+For finite difference methods, the classic is \cite{RichtMort}, which (so far as I know) remains the only mathematically complete account of convergence in book form, half a century after its publication.
+I will provides notes on a special case using energy estimates, that suffices for for the FD methods used in this course (and most of those used in computational seismology).
+Excellent books on other aspects of finite difference methods include \cite{Cohen:01} and \cite{Leveque:07}.
+For the section on asymptotic inversion, I will refer to my old notes \cite[]{SymesNotes}, and to the papers \cite[]{HouSymes:15}, \cite[]{HouSymes:17}, along with many papers cited there.
+The mathematics of extended inversion exists only in fragmentary form.
+I will present an account in the final part of these notes.
 
-\section{Resources}
-A number of good sources exist for the topics of this course, some with links to codes and computational exercises, and I will refer participants to them whenever appropriate.
-Probably the closest in ``Courant and Hilbert'' spirit is Laurent Demanet's 18.325 class notes \cite[]{Demanet:325notes}.
-These were written for a graduate course in mathematics at MIT, and it shows: topics such as NMO do not show up, but the foundations of imaging theory and the adjoint state method, amongst other things, receive an elegant and complete treatment.
-The now-classic book by Norm Bleistein, Jack Cohen, and John Stockwell \cite[]{BleisteinCohenStockwell:01} gives a very thorough treatment of high-frequency asymptotics and ``Kirchhoff'' imaging.
-Gerry Schuster's book on seismic imaging basics \cite[]{Schuster:10} is an excellent overview of the fundamental geophysical and data processing concepts, with Matlab exercises.
-Gary Margrave's book \cite[]{Margrave:book} gives a comprehensive account of seismic data processing, with links to the extensive CREWES Matlab codes.
-Finally, several conventional texts, for example those by \cite{Yil:01} and \cite{DobSav:88}, are excellent resources.
-
-\bibliographystyle{seg}  % style file is seg.bst 
-\bibliography{../../bib/masterref}
+\bibliographystyle{seg}
+\bibliography{../bib/masterref}

### 2024-02-15 repro/caam641/madfigs/paper.tex
cleaned up

@@
 This book rests on a foundation of science and software provided by many talented researchers.
-The following paragraphs list some important background reading, and discusses the book's software environment and its use.
+This chapter describes some of the key references, most of which appear at one place or another in the preceding text, and discusses the book's software environment and its use.
 
@@
 In some sense you have to ``know the answer before you ask the question''.
-Amongst many approaches that have been suggested for globalizing FWI convergence, I will dwell on extended inversion, an idea borrowed from velocity analysis.
+Amongst many approaches that have been suggested for globalizing FWI convergence, I have dwelt on extended inversion, an idea borrowed from velocity analysis.
 The literature on extended inversion has become fairly large.
-The reveiw paper \cite{geoprosp:2008} includes an extensive bibliography up to about 2007.
-Course reading for this segment will include this review paper and several papers appearing in the last 12 months.
+The review paper \cite{geoprosp:2008} includes an extensive bibliography up to about 2007.
+The subject is overdue for another review at this writing (2020).
 
@@
 For finite difference methods, the classic is \cite{RichtMort}, which (so far as I know) remains the only mathematically complete account of convergence in book form, half a century after its publication.
-I will provides notes on a special case using energy estimates, that suffices for for the FD methods used in this course (and most of those used in computational seismology).
+I provide notes on a special case using energy estimates, that suffices for for the FD methods used in this book (and most of those used in computational seismology).
 Excellent books on other aspects of finite difference methods include \cite{Cohen:01} and \cite{Leveque:07}.
-For the section on asymptotic inversion, I will refer to my old notes \cite[]{SymesNotes}, and to the papers \cite[]{HouSymes:15}, \cite[]{HouSymes:17}, along with many papers cited there.
-The mathematics of extended inversion exists only in fragmentary form.
-I will present an account in the final part of these notes.
+x
 

### 2024-02-15 repro/papers/GEM22/paper.tex
cleaned up

@@
-\title{Efficient Computation of Extended Sources}
-\author{William. W. Symes \thanks{The Rice Inversion Project,
-Department of Computational and Applied Mathematics, Rice University,
-Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}
+\title{Efficient Computation of Extended Surface Sources}
+\author{William W. Symes\\
+ %\thanks{The Rice Inversion Project,
+  Department of Computational and Applied Mathematics,\\
+  Rice University,Houston TX 77251-1892 USA,\\
+  email {\tt symes@rice.edu},\\
+ORCID 0000-0001-6213-4272}
 
@@
 \section{Conclusion}
+
+The linear modeling operator of surface source extended acoustic
+waveform inversion is approximately invertible, and this paper has
+shown how to approximately invert it. The construction is based on
+reverse time propagation of data, as inspired by the literature on
+photoacoustic tomography. However, since the input energy comes from a
+surface source, rather than a pressure boundary value, the
+pressure-to-source operator intervenes. It provides not just an
+approximate inverse, but a definition of weighted norms in domain and
+range spaces of the modeling operator, in terms of which that operator
+is approximately unitary. Accordingly, Krylov space iteration defined
+in terms of these weighted norms, or equivalently preconditioned
+Conjugate Gradient iteration, gives a rapidly convergent solution
+method for the linear subproblem.
+
+The existence of an approximate unitary representation of the modeling
+operator is not merely a computational convenience, however. It
+reveals fundamental aspects of the operator's structure that enable
+an explanation for the mitigation of cycle-skipping, a feature of the
+{\em nonlinear} extended inverse problem. This fact echoes earlier observations
+concerned a reflected wave inverse problem, involving a modeling
+operator with a similar approximate inverse
+\cite[]{tenKroode:IPTA14,Symes:IPTA14}. Also, the approximate inverse
+leads to a stable computation of the gradient of the nonlinear
+objective function \ref{eqn:esi}, resolving a difficulty first noted
+also for reflected wave inversion \cite[]{KerSy:94}.
+
+The transmission inverse problem figuring most prominently in
+contemporary applied seismology is surely the Full Waveform Inversion
+(FWI) of diving wave data. This is essentially the same problem as the
+one discussed in this paper, and can be formulated and treated the
+same way, at least for acoustic wave physics. All of the topics
+treated here are open for elastic wave physics - the analogue of the
+pressure-to-source map would is the map from surface velocity field to
+corresponding constitutive defect, analogous to the elastic
+Dirichlet-to-Neumann map investigated by \cite{Rachele:00}.
+
+The underlying tool in the ideas developed here is geometric optics
+(or ray theory), without which the very concept of downgoing waves
+would be meaningless. The physics of actual earth materials includes
+material heterogeneity on all scales, which appears to leave little
+room for the assumption of scale separation underlying geometric
+optics. Moreover, earth materials are anelastic, with elastic wave
+energy being converted to and from thermal excitation, pore fluid
+motion, and so on. A truly satisfactory understanding of inverse wave
+problems will eventually need to accommodate heterogenity and
+anelasticity beyond the current capabilities of the ray-based theory.
+
+\section{Declarations}
+The author received no funding from any source in the performance of
+the work reported here, nor does he have any financial or
+non-financial interests relevant to the content of this article. The
+computations described were developed in the Madagascar reproducible
+research framework (www.reproducibility.org). The software source is
+available on request from the author.
 
@@
 \]
-whence
-\begin{equation}
-\label{eqn:sadj}
- ({\cal S}^+_{z_s,z_r})^T = -{\cal S}^{-}_{z_r,z_s}.
-\end{equation}
-
-The systems \ref{eqn:awepm} for $(p^{\pm},\bv^{\pm})$ with the two
-signs of superscript differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
-$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
-\begin{equation}
-  \label{eqn:trdef}
-  {\cal R} \left(
-    \begin{array}{c}
-      p\\
-      \bv
-    \end{array}
-  \right) =
-  \left(
-    \begin{array}{c}
-      Rp\\
-      -R\bv
-    \end{array}
-  \right)
-\end{equation}
-Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awepm} with
-$(h_s\delta(z-z_s),f_s\bf{e_z}\delta(z-z_s))$ replaced by
-$(-Rh_r\delta(z-z_r), Rf_r\bf{e_z}\delta(z-z_r))$. That is,
-\begin{equation}
-  \label{eqn:trsadj}
-  {\cal R}{\cal S}^- = -{\cal S}^+_{z_r,z_s}{\cal R}
-\end{equation}
-Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj}
-and \ref{eqn:trsadj} imply equation \ref{eqn:trtr}.
+whence \ref{eqn:sadj1} follows immediately.
 

### 2024-02-15 repro/papers/GEM22/submitR1/paper.tex
cleaned up

@@
 surface, so the jump is just twice the sampled value on the
-surface. Thus reverse-time propagation and reading off the velocity
+surface. Thus constructing a source from the veloity trace on the
+receiver surface, reverse-time propagation of this source, and reading off the velocity
 field on the source surface yields a source that reproduces the
 pressure data, within an asymptotically small error.
-%Preconditioning amounts to a change of
-%norm in the domain and range spaces of the modeling operator. In this
-%case, the modfied norms are weighted $L^2$, and the weight operators
-%map pressure to corresponding surface source on the source and
-%receiver planes.
 
@@
 photoacoustic tomography and other wave inverse problems
-\cite[]{Rachele:00,StefUhl:05}. \cite{HouSymes:EAGE16} demonstrated a
-very similar preconditioner for Least Squares Migration, also for its
-subsurface offset extension \cite[]{HouSymes:16}, motivated by
-\cite{tenKroode:12}. These constructions also involve the
-Dirichlet-to-Neumann operator. This concept also turns up in hidden
+\cite[]{Rachele:00,StefUhl:05}. In this terminology, ``Dirichlet''
+refers to pressure restricted to a surface, and ``Neumann'' to the
+normal derivative of pressure on that surface. Since the latter is
+proportional to normal particle acceleration thanks the equations of motion,
+and normal particle velocity is proportion to source, the relation to the
+pressure-to-source map as described in this paper is clear. A number
+of authors have described preconditioners for Least Squares Migration
+with subsurface offset extension in which the Dirichlet-to-Neumann or
+pressure-to-source map plays a central role
+\cite[]{tenKroode:12,HouSymes:15,Herve2017}. It also turns up in hidden
 form in the work of Yu Zhang and collaborators on true amplitude
@@
 
-%The obvious computation of the pressure-to-source map - prescribe the
-%pressure, solve the wave equation with this boundary condition, read
-%off the equivalent source - suffers from intrinsic numerical
-%inaccuracy. I suggest an alternative computationally feasible
-%approach, via economical short-distance wave propagation. Since the
-%map is symmetric only in an approximate, asymptotic sense, it must be
-%symmetrized for use as a Krylov preconditioner. I describe a
-%symmetrization procedure that requires no further wave computations
-%beyond those necessary to compute the action of the operator itself.
-
 The discussion in this paper is formal and incomplete, in the sense
 that some important mathematical underpinnings are only
-sketched. I will treat the modeling operator $PL[{\bf c}]^{-1}$ as if
+sketched. The transport of high-frequency energy along rays is a
+crucial ingredient in the constructions explained here. Of course,
+such transport only occurs in material parameter models that vary
+slowiy in space on the wavelength scale, and even then only
+approximately. Rather than provide explicit
+smoothness constraints and asymptotic error estimates, I will use
+these concepts in an intuitive way, using ``approximate'' and
+``asymptotic'' and the symbol ``$\approx$'' to suggest asymptotically
+negligible errors not explicitly assessed, and refer to such errors as
+``negligible''.
+
+I will treat the modeling operator $PL[{\bf c}]^{-1}$ as if
 it mapped square integrable surface sources to square integrable sampled
@@
 is not be well-defined, strictly speaking.
-
 This phenomenon is related to the ill-posedness of wave equations as
@@
     generation at $(500, 10000)$ m indicated with star. Overlain with rays from point
-    source to receiver surface.}
+    source to receiver surface. Outward normal points up at both source
+    and receiver surfaces, so rays are incoming at source
+    surface, outgoing at receiver surface.}
   \label{fig:ooplrays0}
@@
 
-Time reversal as found in the literature on photoacoustic tomography
+The time reversal construction, as found in the literature on photoacoustic tomography,
 must be localized to apply to the surface source inversion problem,
@@
 propagation. It must also be modified to produce a source, rather than
-a restriction of the pressure field. The assumption of incoming rays at the source surface and
+a slice of the pressure field. The assumption of incoming rays at the source surface and
 outgoing rays at the receiver surface makes both modifications  
@@
 
-The construction takes place in three stages: first, an
-acoustic field with high-frequency energy propagating along an
-incoming ray family is identified as the causal solution of an inhomogeneous boundary value problem {\em locally and on
-  the inside}. The boundary value is simply the restriction of the
-acoustic field to the surface. Second, the same causal solution is
-identified as the causal solution of an acoustic problem with source,
-singular on the surface in the fashion of equation
-\ref{eqn:awepm}. The source amplitude is proportional to 
-the velocity field sampled on the surface. Third, this identification is used twice, first to
-propagate the receiver data backwards in time as the solution of an
-anti-causal problem,  the velocity field of which is sampled on the
-source surface; then the causal problem is solved with the
-source built from the velocity samples. This solution must be
-identical to the field generating the data, along the ray fields
-carrying all of the high-frequency energy. Therefore the reconstructed
-source is an approximate solution of the problem \ref{eqn:esis} for
-$\alpha=0$.
+The construction takes place in three stages: first, an acoustic field
+with high-frequency energy propagating along an incoming ray family is
+identified as the causal solution of an inhomogeneous boundary value
+problem {\em locally and on the inside}. The boundary value is simply
+the restriction of the pressure field to the surface. Second, the same
+causal solution is identified as the causal solution of an acoustic
+problem with source, singular on the surface in the fashion of
+equation \ref{eqn:awepm}. The source amplitude is proportional to the
+velocity field sampled on the surface. Third, this identification is
+used twice, first to propagate the receiver data backwards in time as
+the solution of an anti-causal problem with source on the receiver
+surface; the velocity field of which is sampled on the source surface;
+then the causal problem is solved with the source built from the
+velocity samples. This solution must asymptotically approximate the
+field generating the data, along the ray fields carrying all of the
+high-frequency energy. Therefore the reconstructed source is an
+approximate solution of the problem \ref{eqn:esis} for $\alpha=0$.
 
@@
 The surface source representaion part of the construction shows that
-under the assumed ray conditions, a solution of \ref{eqn:awebase} is
-asymptotically the same near the receive surface $\Sigma_r$ as the
+under the assumed ray conditions, a solution of \ref{eqn:awefree} is
+asymptotically the same near the receiver surface $\Sigma_r$ as the
 solution of the system \ref{eqn:awepm} with a suitable source
-$\tilde{f}_s$ (and zero velocity source, $h_s=0$. Therefore this
+$h_s$ (and zero velocity source, $f_s=0$). Therefore this
 construction also inverts the modeling operator $S$ in the sense that
 $Sh_s\approx S\tilde{h_s}$ if $P_rp \approx S\tilde{h_s}$. That is,
-this construction produces an asymptotic rightt inverse of $S$: it
+this construction produces an asymptotic right inverse of $S$: it
 does not necessily (approximately) recover the source that generates
@@
 $\Sigma_r$, generated by the point source shown in the figure, at
-$(3500, 3500)$ m (Figure \ref{fig:drecplh0}. However, I will invert
+$(3500, 3500)$ m (Figure \ref{fig:drecplh0}). However, I will invert
 this data in a homogeneous medium with $\kappa \equiv 4$ GPa. These
 choice illustrate two facts about surface source inversion and the
-approximate inversion based on time reversal. First, the success of the inversion demostrates the
+approximate inversion based on time reversal. First, the success of the inversion demonstrates the
 insensitivity of the time reversal method to ray multipathing
@@
 
-Since the system \ref{eqn:awepm} has a unique solution by standard
-theory \cite[]{Lax:PDENotes}, the source vector field $(h_s,f_s)$
-determines the acoustic field $(p^{\pm},\bv^{\pm})$ in space time, and
-in particular the limits from the right at $z=z_s$, $P_sp^{\pm}$ and
-$P_sv_z^{\pm}$. This relation is not invertible: it is not possible to
-prescribe both pressure and normal velocity on a surface such as
-$z=z_s$. So the columns of the matrix operator
-${\cal S}^{\pm}_{z_s,z_r}$ must satisfy a linear relation. In this
-section I will explain this relation; it involves the {\em
-  pressure-to-source} map. This operator also turns out to be the
-principal component of a preconditioning strategy for iterative
-solution of the optimization problem \ref{eqn:esis}, so I will devote
-some effort to its proper definition. It is closely related to the
-Dirichlet-to-Neumann operator mentioned in the introduction.
-
-While it is not possible to prescribe both pressure and velocity on
-$z=z_s$ in solutions of \ref{eqn:awepm}, it is possible to
-prescribe pressure only, for instance: if the function $\phi$ on
-the surface $z=z_s$ satisfies suitable conditions, for
-example the downgoing constraint mentioned earlier, a unique solution
-exists for the acoustic system in both half-spaces $\pm z > z_s$:
-\begin{eqnarray}
-\label{eqn:awe0}
-  \frac{1}{\kappa}\frac{\partial p_{\pm}}{\partial t} & = & - \nabla \cdot \bv_{\pm}, \nonumber \\
-  \rho\frac{\partial \bv_{\pm}}{\partial t} & = & - \nabla
-                                                    p_{\pm}, \nonumber \\
-  p_{\pm} & =& 0,  \mbox{ for } t \ll 0, \nonumber\\ 
-  \bv_{\pm} & = & 0 \mbox{ for } t \ll 0, \nonumber\\
-  \lim_{z \rightarrow z_s^{\pm}}p_{\pm}(x,y,t,z)& =& \phi(x,y,t).
-\end{eqnarray}
-Note that the subscript $\pm$ here refers to the sign of $z-z_s$, as opposed
-to the superscript ${\pm}$, which refers to the sign of $t$ throughout
-this paper.
-
-From the boundary condition (last equation in \ref{eqn:awe0}), one
-sees that the pressures $p_{\pm}$ in the two half-spaces have the same
-limit at the boundary $z=z_s$. Stick the two half-space
-solutions together to form an acoustic field $(p^+,\bv^+)$ in all of
-space-time, that is,
-\begin{equation}
-  \label{eqn:awealt}
-  p^+(x,y,z,t) =
-  \left\{
-    \begin{array}{c}
-      p_+(x,y,z,t) \mbox{ if } z>0,\\
-      p_-(x,y,z,t) \mbox{ if } z<0,
-    \end{array}
-  \right.
-\end{equation}
-and a similar definition for $\bv^+$. Then $p^+$ is continuous across
-$z=z_s$, and the boundary condition in system \ref{eqn:awe0} may be
-written as $P_sp^+=\phi$.
-
-The same construction can be carried out in the anti-causal sense,
-with anti-causal half-space solutions glued together to form a
-full-space distribution solution $(p^-,\bv^-)$, with the property that
-$p^-$ is continuous across $z=z_s$ and $P_sp^-=\phi$.
-
-The reader may object that the notation $(p^\pm,\bv^\pm)$ is already in
-use, for the solution of \ref{eqn:awepm}. This objection is
-valid. However, {\em in the sense
-  of distributions}, $(p^{\pm},\bv^{\pm})$ as defined in display
-\ref{eqn:awealt}, is {\em exactly} the causal solution of \ref{eqn:awepm}
-for the choice $h_s = -[v^{\pm}_{z}]|_{z=z_s}, f_s=0$, as follows from a
-simple integration-by-parts calculation. So the notation is consistent!
-
-The negative jump $-[v^{\pm}_{z}]|_{z=z_s}$ is thus a function of $\phi$. Define
-the {\em pressure-to-source} operator $\Lambda^{\pm}_{z_s}$ by
-\begin{equation}
-  \label{eqn:deflam}
-  \Lambda^{\pm}_{z_s}\phi = -[v^{\pm}_{z}]|_{z=z_s}
-\end{equation}
-The conclusion: if $h_s = \Lambda^{\pm}_{z_s}\phi$ and $f_s=0$ in the
-system \ref{eqn:awepm}, then $\phi=P_sp^{\pm}$.
-
-Otherwise put, $S^{\pm}_{z_s,z_s}\Lambda^{\pm}_{z_s} \phi = \phi$, so
-$\Lambda^{\pm}_{z_s}$ is inverse to $S^{\pm}_{z_s,z_s}$. The relation
-\ref{eqn:trtrcomp} implies in turn that
-\begin{equation}
-  \label{eqn:lamadj}
-  (\Lambda^{\pm}_{z_s})^T = - \Lambda^{\mp}_{z_s}
-\end{equation}
-
-There is also a {\em velocity-to-source} operator. For the solution
-$(p^{\pm},\bv^{\pm})$ of system \ref{eqn:awepm} with $h_s=0$, the
-normal component of velocity, $v^{\pm}_z$, is continuous across
-$z=z_s$, and the velocity source (vertical load)
-$f_s=-[p^{\pm}]_{z=z_s}$. I will not name the velocity-to-source
-operator, as it does not appear explicitly in the developments to
-follow. As will be seen, it is essentially the inverse of the
-pressure-to-source operator.
-
-The quadratic form defined by $\Lambda^{\pm}_{z_s}$ has fundamental
-physical significance. Define the total acoustic energy $E^{\pm}(t)$ of the
-field $(p^{\pm},\bv^{\pm})$, at time $t$ by
-\begin{equation}
-  \label{eqn:defae0}
-  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)(\bx,t).
-\end{equation}
-Then
-\begin{equation}
-  \label{eqn:elim}
-  \pm \lim_{\pm t \rightarrow \infty} E^{\pm}(t) =  \langle P_sp^{\pm},
-  (\Lambda^{\pm}_{z_s} P_sp^{\pm}) \rangle_{L^2(z=z_s)}.
-\end{equation}
+The result of the last section has a couple of apparent
+defects. First, the approximate inversion of pressure data requires
+knowledge of velocity data, which is not prescribed in the problem
+statement \ref{eqn:esisp}. Second, the significance of the
+reverse-time field $(\hat{p},\hat{\bv})$ is obscure. In this section
+these two issues are addressed.
+
+\subsection{Pressure-to-source operator}
+The first step is to formalize the relation between pressure and
+source (or equally well, pressure and normal velocity) on as surface
+$\Sigma$ as the application of an operator.
+
+Observe that in principle it's not necessary to
+supply the normal velocity at the source (or receiver) surface, since
+it's determined locally by the pressure on the surface, or causal (or
+anti-causal) fields. This follows from the observation that the system
+\ref{eqn:aweloc} has a unique solution, a by-product of the
+conservation of energy (see for example \cite{Lax:PDENotes}, section
+4.3). That is, in the language of that discussion, knowledge of $p$ on
+$\Sigma$ for $0\le t \le \tau$ determines $(\tilde{p},\tilde{\bv})$ in
+the entire domain $\Omega^+$, $0\le t \le \tau$. Therefore knowledge
+of $p$ on $\Sigma$ for $0\le t \le \tau$ determines
+$\bv \cdot {\bf n}$, up to a negligible error. As in the last section,
+it follows that knowledge of $p$ on $\Sigma$, any $t$, determines a
+source $h$ on $\Sigma$ for which the pressure field $\tilde{p}$
+(solution of the system \ref{eqn:awesrcmod}) differs negligibly from
+$p$.
+
+This relation defines a causal {\em pressure-to-source operator}
+$\Lambda: p \mapsto h$, yielding the source on a surface $\Sigma$ that
+reproduces the (causal) pressure field from its trace on $\Sigma$, at
+least locally on the inward side as defined by the selection of normal
+field on $\Sigma$, under the ray conditions defined earlier. Since
+$h \approx 2 \bv \cdot {\bf n}$ (limiting value of $\bv$ from the
+inward direction), $\Lambda$ is effectively the
+pressure-to-normal-velocity operator.
+
+With this convention, two of the three steps in
+the approximate inversion procedure are operator applications. The
+operators involved are the causal pressure-to-source operator for the
+source surface $\Sigma_s$, denoted $\Lambda_s$, and an analogous
+anti-causal operator for the receiver surface $\Sigma_r$. 
+The characterization of the the source field $h_r$ in equation
+\ref{eqn:awerecglob} shows that this anti-causal pressure-to-source
+operator differs from its causal analogue only by sign. Denote the
+causal operator by $\Lambda_r$.
+
+\subsection{Time Reversal as Adjoint State}
+The middle step is solution of the reverse-time system
+\ref{eqn:awerecglob} for the field $(\hat{p},\hat{\bv})$, with input
+pressure source
+$h_r$ on $\Sigma_r$ and output $P_ss\hat{p}$ on $\Sigma_s$. To
+understand the significance of this mapping, it is just as easy, and
+will be useful later, to include a velocity source $f_r$, that is,
+modify the second equation in \ref{eqn:awerecglob} by adding $f_r {\bf
+  n}_r \delta_{\Sigma_r}$, so that it becomes a reverse-time version
+of the basic system \ref{eqn:awepm}. Recall the definition of energy
+in the acoustic field $(p,\bv)$:
+\begin{equation}
+  \label{eqn:energydef}
+   E[p,\bv](t) = \frac{1}{2} \int \,dx\,dy\,dz\, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx)\bv(\bx,t)
+   \cdot \bv(\bx,t)\right)
+\end{equation}
+This is a quadratic form. Polarization makes a bilinear form, into
+which one can plug a pair of fields. Use the solutions $(p,\bv)$ of
+the forward-time system \ref{eqn:awepm} for one of the pair, and
+the solution $(\hat{p},\hat{\bv})$ of system \ref{eqn:awerecglob}
+(augmented with the velocity surface source) for the other. Since $(p,\bv)$ vanishes
+for large negative time, $(\hat{p},\hat{\bv})$ for large positive
+time, 
+\[
+0 = 
+\left(\int\, dx\,dy\,dz\, \frac{p \hat{p}}{\kappa} +  
+\rho \bv \cdot \hat{\bv} \right)|_{t \rightarrow \infty}
+-
+\left(\int\, dx\,dy\,dz\, \frac{p \hat{p}}{\kappa} +  \rho \bv \cdot \hat{\bv} \right)|_{t \rightarrow -\infty}
+\]
+\[
+= 
+\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \hat{p}}{\kappa} +  \rho \bv \cdot \hat{\bv} \right)
+\]
+Differentiate, replace time derivatives by space derivatives and
+source terms using the equations of motion in the systems
+\ref{eqn:awepm} and \ref{eqn:awerecglob}, and integrate by parts to
+move all of the space derivatives onto $p$ and $\bv$.  Most of the
+remaining expressions cancel, leaving
+\[
+  = \int_{-\infty}^{\infty}\,dt \, \left[
+    \int\,d\sigma_s(\bx)\,(h_sP_s\hat{p}+f_sP_s(\hat{\bv}\cdot{\bf
+      n}_s)) +
+    \int\,d\sigma_r(\bx)\,(h_rP_rp+f_rP_r(\bv\cdot{\bf n}_r)) \right]
+\]
+Here $\sigma_s$ and $\sigma_r$ are the area elements on $\Sigma_s$ and
+$\Sigma_r$. These integrals are inner products of time traces on $\Sigma_s$ and $\Sigma_r$. Denote these inner products by $\langle \cdot,\cdot
+\rangle_s$ and $\langle \cdot,\cdot
+\rangle_r$ respectively. Then the above is written in terms of the
+vector modeling operator ${\cal S}$ (definition \ref{eqn:fwd}) as 
+\[
+  = \langle (h_s,f_s)^T, (P_s\hat{p},P_s(\hat{\bv}\cdot{\bf
+    n}_s))^T \rangle_s+ \langle (h_r,f_r)^T, {\cal S}(h_s,f_s)^T \rangle_r,
+\]
+whence identify the adjoint of ${\cal S}$ as
+\begin{equation}
+  \label{eqn:vecadj}
+  {\cal S}^T(h_r,f_r) = -(P_s\hat{p},P_s(\hat{\bv}\cdot{\bf
+    n}_s))^T.
+\end{equation}
+Speclializing to the pressure modeling operator $S$ (equation
+\ref{eqn:sdef}),
+\begin{equation}
+  \label{eqn:padj}
+  S^T = (\Pi_0 {\cal S} \Pi_0^T)^T = \Pi_0 {\cal S}^T \Pi_0^T
+\end{equation}
+so to calculate the adjoint of $S$ applied to $h_r$, solve system
+\ref{eqn:awerecglob} (with $f_r=0$) and extract the pressure at $\Sigma_s$:
+\[
+  S^T(h_r) = -P_s \hat{p}.
+\]
+
+Combined with the definitions of the pressure-to-source operators
+$\Lambda_s$ and $\Lambda_r$, these results enable a translation of the
+approximate inverse construction in the last section into a chain of
+operator applications: to invert a pressure gather $d$ on $\Sigma_r$
+for a source $h_s$ on $\Sigma_s$ that approximately reproduces it ($d
+\approx Sh_s$),
+\begin{itemize}
+\item[1. ] create a source $h_r = -2\Lambda_r d$ on $\Sigma_r$;
+\item[2. ] apply the adjoint $S^T$ (solve the reverse time system
+  \ref{eqn:awerecglob}) to $h_r$;
+\item[3. ] apply $\Lambda_s$ to $-S^Td$.
+\end{itemize}
+Putting this all together, an approximate right inverse operator
+$S^{\dagger}$ to $S$ may be expressed as
+\begin{equation}
+  \label{eqn:baseinv}
+  S^{\dagger} = \Lambda_s S^T \Lambda_r,
+\end{equation}
+that is,
+\begin{equation}
+  \label{eqn:rightinv}
+  SS^{\dagger} \approx I.
+\end{equation}
+
+\subsection{Positivity of $\Lambda$}
+
+Equation \ref{eqn:baseinv} is remarkable, in that the form of
+$S^{\dagger}$ is that of an adjoint with respect to weighted norms,
+defined by the operators $\Lambda_s^{-1}$ in the domain of $S$ and
+$\Lambda_r$ in the range. This only makes sense if the
+pressure-to-source operators are positive definite and symmetric, the
+necessary conditions to define weighted norms.
+
+In fact the pressure-to-source operator is positive semi-definite, and that
+fact has fundamental physical significance. Suppose that $(p,\bv)$
+solves the basic system \ref{eqn:awepm} with $f_s=0$. From the
+definition of energy \ref{eqn:energydef} and causality of the fields,
+\[
+ E[p,\bv](t) = \frac{1}{2} \int_{-\infty}^{t} \,dt \int \,dx\,dy\,dz\,
+ \frac{d}{dt}\left(\frac{p(\bx,t)^2}{\kappa(\bx)} +
+   \rho(\bx)\bv(\bx,t) \cdot \bv(\bx,t)\right)
+\]
+\[
+  = \int_{-\infty}^{t} \,ds \int \,dx\,dy\,dz\,
+ \left(\frac{p (\bx,s)\frac{\partial p}{\partial t}(\bx,s)^2}{\kappa(\bx)} +
+   \rho(\bx)\bv(\bx,s) \cdot \partial{\bv}{\partial t}(\bx,s)\right)
+\]
+Using \ref{eqn:awepm}, replace the time derivatives with spatial derivatives and the delta
+function on ths source surface, integrate by parts to shift the space
+derivatives to $p$. Everything cancels except for
+\[
+  =  \int_{-\infty}^{t} \,ds \int \,dx\,dy\,dz\, p(\bx,s) h_s(\bx,s)
+  \delta_{\Sigma_s}(\bx) = \int_{-\infty}^{t} \,dt \int_{\Sigma_s} \,d\sigma_s \, p h_s.
+\]
+Take the limit as $t \rightarrow \infty$ to get
+\[
+\lim_{t \rightarrow \infty} E[p,\bv](t) = \langle h_s, P_s p \rangle_s = \langle P_s p, \Lambda_s P_s p \rangle_s. 
+\]
 That is, the value of the quadratic form defined by
-$\Lambda^{\pm}_{z_s}$, evaluated at the pressure trace on $z=z_s$,
+$\Lambda_s$, evaluated at the pressure trace on $\Sigma_s$,
 gives the total energy transferred from the source to the
 acoustic field over time. Since $E$ is itself a positive definite
-quadratic form in the acoustic field, it follows that $\pm
-\Lambda^{\pm}_{z_s}$ is positive semi-definite.
-
-While $\Lambda^{\pm}_{z_s}$ is positive semi-definite, it is not
+quadratic form in the acoustic field, it follows that $
+\Lambda_s$ is positive semi-definite.
+
+\subsection{Approximate symmetry of $\Lambda$}
+Except in special cases, $\Lambda_{s,r}$ is not
 symmetric. However, it is {\em approximately symmetric} in the
-high-frequency sense. This fact follows from a
-geometric optics analysis of the half-space solution. This leads to
-the identification of $\Lambda^{\pm}_{z_s}$ as a {\em
-  pseudodifferential operator} of order zero on $z=z_s$, with principal symbol
-\begin{equation}
-  \label{eqn:lamsym}
-  \sigma_0(\Lambda^{\pm}_{z_s}) = \pm 2(\kappa(\bx)
-  \rho(\bx))^{1/2}\left(1-\frac{\kappa(\bx)(\xi^2+\eta^2)}{\rho(\bx)\omega^2}\right)^{-1/2}.
-\end{equation}
-Here $\xi$, $\eta$, and $\omega$ are the dual Fourier variables to
-$x$, $y$, and $t$ respectively. The downgoing assumptions means that
-for local planewave components of $P_sp$, the quantity inside the
-square root is positive. Thus $\Lambda^{\pm}_{z_s}$ has real principal
-symbol (in fact, the entire symbol is real) hence defines an
-asymptotically symmetric operator:
+high-frequency sense. This fact follows from a high-frequency
+asymptotic analysis, a sketch of which follows for the special case of
+a planar source surface
+$\Lambda_s=\{z=z_s\}$, as in the two examples. SInce the same reasoning
+applies to $\Lambda_r$, I will omit the subscript for the remainder of
+this subsection.
+
+The source amplitude
+$h=h(x,y,t)$ can be expressed as a Fourier synthesis over
+oscillatory wave packets of the form
+\begin{equation}
+  \label{eqn:pwp}
+ a_0(x,y,t) e^{i\omega(t-x\xi-y\eta)}
+\end{equation}
+in which the spatial frequencies have been expressed as slownesses
+$\xi,\eta$ scaled by temporal frequency $\omega$. The function $a_0$
+is a frequency-independent envelope, equal to 1 where $h$ is
+non-zero.
+
+The wave packet $a_0(x,y,t) e^{i\omega(t-x\xi-y\eta)}$ can be
+viewed as the boundary value on $\Sigma$ of a
+high-frequency asymptotic solution of the form \ref{eqn:go}, provided that the
+slowness vector $(\xi, \eta)$ satisfies the propagating condition
+$c^2(\bx)(\xi^2+\eta^2) < 1$ at all $\bx$ for
+which $a_0(\bx,t)$ is non-zero for any $t$. In that case, two
+solutions $\psi$ of the eikonal equation $|\nabla \psi|=c$ exist, at least locally near
+$\Sigma$, for which $\psi(x,y,0)=x\xi+y\eta$. Choose the solution
+that increases in the direction of the outward unit normal ${\bf
+  n}$, and solve the corresponding transport equation for an
+amplitude $a$ satisfying $a(x,y,0,t) = a_0(x,y,t)$. Then the geometric optics solution $a e^{i\omega(t-\psi)}$ is
+asymptotic the pressure field of to an incoming solution. From the
+evolution equation for $\bv$, integrated in time as in equation
+\ref{eqn:newt},  an expression follows for the corresponding normal
+velocity:
+\[
+  -{\bf n} \cdot \frac{1}{\rho}\int_{-\infty}^t ds \nabla (a
+  e^{i\omega(t-\psi)})=\frac{1}{\rho}\int_{-\infty}^t ds {\bf n}
+  \cdot(i\omega a\nabla \psi -\nabla a) e^{i\omega(t-\psi)})
+\]
+Now use the identity
+\[
+  e^{i\omega(t-\psi)} = \frac{1}{i\omega}\frac{d}{dt}e^{i\omega(t-\psi)}
+\]
+and integrate by parts in $t$ to obtain the expression for
+the normal velocity field 
+\[
+  = \frac{a}{\rho} ({\bf n} \cdot \nabla \psi)  e^{i\omega(t-\psi)} + O(1/\omega).
+\]
+The terms not written explicitly begin with one of order -1 in
+$\omega$. The remainder appears to be also of order -1, but
+application of the integration-by-parts trick described above yields
+another term of order -2 plus another remainder of apparent order
+-2. Continuing this process develops the $O(1/\omega)$ remainder in an
+asymptotic series of homogeneous terms in $1/\omega$ of arbitrary
+length.
+
+An expression for the normal velocity 
+$\Sigma$ follows by restriction. Note that ${\bf n} \cdot \nabla \psi =( c^{-2}(x,y,z)-
+(\xi^2+\eta^2))^{1/2}$ because of the choice of sign and the eikonal
+equation, and the root is real because of the propagating assumption.
+
+The upshot of this calculation is that to leading order in frequency,
+the normal velocity corresponding to a pressure wave packet differs by a position- and slowness-dependent
+factor. Accounting for the factor of 2 difference between the source
+amplitude and velocity trace, this calculation may be summarized:
+\begin{equation}
+  \label{eqn:lamwp}
+  \Lambda (a_0e^{i\omega(t-x\xi-y\eta)}) = \lambda
+  (a_0e^{i\omega(t-x\xi-y\eta)}),
+\end{equation}
+where $\lambda(x,y,t,\xi,\eta,\omega$ is
+defined by the asymptotic series in $\omega$ developed above. Its dominant term,
+$O(1)$ (order 0) in $\omega$, is
+\[
+  \lambda^0(x,y,t,\xi,\eta,\omega) = 2\frac{( c^{-2}(x,y,z)-
+    \xi^2+\eta^2)^{1/2}}{\rho} + O(1/\omega)
+\]
+\begin{equation}
+  \label{eqn:lamsymb}
+= 2(\kappa
+\rho)^{-1/2}\left(1-\frac{\kappa(k_x^2+k_y^2)}{\rho \omega^2}\right)^{1/2} + O(1/\omega).
+\end{equation}
+In the final expression, the slownesses $\xi,\eta$ have been replaced
+by the Fourier variables $k_x=x\xi, k_y=y\eta$. Note that as a
+function of $k_x,k_y,\omega$, $\lambda^0$ is homogeneous of order
+0. It is easy to see that the $N$th term of the asymptotic series
+defining $\lambda$ is of homogeneous order $-N$.
+
+%%%%%%%%%%%
+% discussion of psidos
+%
+Operators that affect oscillatory wave packets by multiplication with
+functions having the properties of $\lambda$ have come to
+be called ``pseudodifferential'' \cite[]{Nir:72,Tay:81,SaiR:91}. The multiplier ($\lambda$ in this
+case) is called the {\em symbol} of the operator, and the dominant
+term in an asymptotic development of the multiplier
+(for instance $\lambda^0$) is called the {\em principal symbol}.
+Amongst other special properties of these operators, their adjoints
+are operators of the same type. In particular, the principal symbol of the adjoint is
+the Hermitian (matrix) adjoint of the principal symbol. For scalar
+operators like $\Lambda$ (mapping scalar-valued functions to
+scalar-valued functions), the symbols are scalar ($1 \times 1$), and
+the Hermitian adjoint is simply the complex conjugate (\cite{Tay:81},
+section II.4). Therefore scalar operators with real principal symbol,
+like $\Lambda$, have the same principal symbols as their adjoints. So
+$\Lambda$ differs from $\Lambda^T$ by a pseudodifferential operator
+whose symbol is an asymptotic sum of terms of negative order in
+frequency. Therefore the difference is negligible in the limit of high
+frequencies. This observation establishes the conclusion of this subsection: 
 \begin{equation}
   \label{eqn:lamappsim}
-  (\Lambda^{\pm}_{z_s})^T \approx \Lambda^{\pm}_{z_s}.
-\end{equation}
-(For more on this, see \cite{StefUhl:05}.)
-The analysis also reveals that the solution components not continuous
-at $z=z_s$ are odd there:
-\begin{equation}
-  \label{eqn:odd1}
-  \lim_{z\rightarrow z_s^+} v^{\pm}_{z} \approx - \lim_{z\rightarrow z_s^-}
-  v^{\pm}_{z}
-\end{equation}
-for the solution of \ref{eqn:awepm} with $f_s=0$.
-Similarly, 
-\begin{equation}
-  \label{eqn:odd2}
-  \lim_{z\rightarrow z_s^+} p^{\pm}\approx - \lim_{z\rightarrow z_s^-}
-  p^{\pm}
-\end{equation}
-for the solution of \ref{eqn:awepm} with $h_s=0$. Here ``$\approx$''
-means in the sense of high frequency asymptotics, that is, that the
-difference between the two sides is relatively smooth, hence small if
-the data is highly oscillatory. Therefore if $f_s=0$ in system \ref{eqn:awepm},
-\begin{equation}
-  h_s = \Lambda^{\pm}_{z_s}P_sp^{\pm} = -[v^{\pm}_{z}]|_{z=z_s} \approx -2
-  P_sv^{\pm}_{z}
-  \label{eqn:tracejump10}
-\end{equation}
-Similarly, if $h_s=0$ in system \ref{eqn:awepm}, then
-\begin{equation}
-  \label{eqn:tracejump20}
-  f_s = -[p^{\pm}]|_{z=z_s} \approx -2 P_s p^{\pm}.
-\end{equation}
-Thus $f_s$ determines approximately the boundary value of $p^{\pm}$,
-as a solution of the acoustic wave system in the half-space
-$z>z_s$. However, as repeated in equation \ref{eqn:tracejump10}, a
-solution with this boundary value is also the restriction to $z>z_s$
-of a solution to \ref{eqn:awepm} with $f_s=0$ and $h_s=
-\Lambda^{\pm}_{z_s}P_sp^{\pm}$. Therefore if
-\begin{equation}
-  \label{eqn:hfcondn}
-  h_s =-\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,
-\end{equation}
-then the pressure boundary value $P_sp^{\pm}$ is the
-same for the solutions of \ref{eqn:awepm} for source vectors $(h_s,0)$
-and $(0,f_s)$. Since the pressure boundary values are the same, the solutions
-in $z>z_s$ are the same. In particular, since $z_r>z_s$ and ${\cal
-  S}^{\pm}_{z_s,z_r}(h_s,f_s)^T = (P_rp^{\pm},P_rv^{\pm}_z)^T$, it follows
-that
+  \Lambda^T \approx \Lambda,
+\end{equation}
+
+Before leaving this topic, recall that the computation outlined above
+was predicated on the propagating assumption, namely that the slowness
+parameters $\xi,\eta$ of the wave packet defined in \ref{eqn:pwp}
+satisfy $c(x,y,z_s)^2(\xi^2 + \eta^2) < 1$ whereever the amplitude
+$a_0(x,y,t)$ is non-zero. Therefore the conclusion \ref{eqn:lamappsim}
+holds only for application to such wave packets. However, the restriction of
+propagating waves (such as those that are output by the modeling
+operator $S$ or its adjoint $S^T$) are well-approximated by sums over
+propagating wave packets: this is part of the assumption that the rays
+carrying high-frequency energy cross the source and receiver surfaces
+at non-zero angles. To make the definition of $\Lambda$ complete, it
+must be extended to non-propagating wave packets, for instance by
+making its effect negligible via a dip filter. I will ignore this refinement, since the
+application to non-propagating components appears to play little role
+in approximate inversion and its applications as developed here.
+
+\section{Computing $\Lambda$}
+
+The missing ingredient in the preceding discussion is any indication
+of how the action of the pressure-to-source operators $\Lambda$ is to be
+computed. As will be seen in the next section, an (approximate)
+inverse of this operator is also required in the formulation of
+preconditioned Krylov iteration.
+
+Direct computation of the pressure-to-source operator $\Lambda_s$, for instance, by solving
+\ref{eqn:awepm} and reading off $P_sv_{z}$, turns out to be
+numerically ill-behaved. An alternate approach begins with the
+observation that a pressure gather on $\Sigma_r$ produced by a
+pressure source $h_s$ on $\Sigma_s$ can also be well-approximated by
+the field from a velocity source. To see this, recall the first major
+step in the time reversal construction, that a solution of the wave
+equation near $\Sigma_s$ can be represented as the solution of a
+boundary value problem with $p$ given on $\Sigma_s$ (system
+\ref{eqn:aweloc}). The derivation combined the unique determination of
+the field by its pressure boundary values on an enclosing domain with
+the consequences of the standing assumptions about rays carrying high
+frequency energy. Precisely the same logic could have been applied to
+the normal velocity boundary values, as these also uniquely determine
+the field in an enclosing domain. Proceeding further, a source
+representation based on continuous extension of normal velocity, with
+a jump in the pressure follows by logic precisely analogous to that
+leading to the system \ref{eqn:awesrc}: instead, obtain a field
+$(\tilde{p},\tilde{\bv})$ differing negligibly from the field of the
+same name in the earlier discussion, solving 
+\begin{eqnarray}
+\label{eqn:awesrcvzmod}
+  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & 
+                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
+  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & f_s {\bf n}\delta_{\Sigma}- \nabla \tilde{p}, \nonumber \\
+  \tilde{p} & = & 0 \mbox{ for }  t \ll 0, \nonumber \\
+  \tilde{\bv} & = & {\bf 0} \mbox{ for }  t \ll  0.       
+\end{eqnarray}
+(that is, system \ref{eqn:awepm} with $h=0$) with $f_s = 2P_sp$.
+
+This fact can usefully be rewritten using the vector modeling
+operator ${\cal S}$ (definition \ref{eqn:fwd}), and the projection
+operators $\Pi_i$ (definition \ref{eqn:pidef}), which insert scalar
+sources into the pairs that occur in the acoustic system
+\ref{eqn:awepm}: explicitly, $\Pi_0(h,f)^T = h, \Pi_1(h,f)^T = f,
+\Pi_0^Th = (h,0), \Pi_1^Tf = (0,f)$. Then
+\begin{eqnarray}
+  P_r\tilde{p} &=& \Pi_0 {\cal S} \Pi_1^Tf_s, \nonumber \\
+  P_r({\bf n}_r \cdot \tilde{\bv}) &=& \Pi_1 {\cal S} \Pi_1^Tf_s.
+                                       \label{eqn:ptildef}
+\end{eqnarray}
+Since the field $(\tilde{p},\tilde{\bv}$ constructed here differs negligbly from that
+constructed earlier with the same name, which in turn differs negligibly from $p$, it
+follows that $P_r\tilde{p} \approx P_rp = Sh_s$. By definition of $S$
+(\ref{eqn:sdef}), this is the same as $=\Pi_0{\cal
+  S}\Pi_0^Th_s$. Since the velocity fields also differ negligibly,
+obtain
+\begin{eqnarray}
+  P_r\tilde{p} & \approx & \Pi_0 {\cal S} \Pi_0^Th_s, \nonumber \\
+  P_r({\bf n}_r \cdot \tilde{\bv}) & \approx & \Pi_1 {\cal S} \Pi_0^Th_s.
+                                       \label{eqn:ptildeh}
+\end{eqnarray}
+
+From the defintiion of $\Lambda_s$, $h_s=\Lambda_sP_sp =
+\frac{1}{2}\Lambda_sf_s$. Combining \ref{eqn:ptildef},
+\ref{eqn:ptildeh}, and this last identity, obtain
 \begin{equation}
   \label{eqn:snull}
-  {\cal S}^{\pm}_{z_s,z_r}\left(\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,f_s\right)^T \approx 0.
-\end{equation}
-
-Equation \ref{eqn:snull} states the relation between the columns of $
-{\cal S}^{\pm}_{z_s,z_r}$ mentioned in the introduction to this
-section.
-
-\section{Unitarity}
-
-The next chapter in this story recognizes the relations in display
-\ref{eqn:approxinv} as asserting the approximate unitarity of
-$S^+_{z_s,z_r}$.
-
-The matrix identity \ref{eqn:snull} implies a relation between $S, V,$
-and $\Lambda$ of some interest in itself. After minor re-arrangement, the second row of reads
-\begin{equation}
-  \label{eqn:snull2}
--\frac{1}{2}\Pi_1{\cal S}^{\pm}_{z_s,z_r}\Pi_0^T\Lambda^{\pm}_{z_s}  \approx
-V^{\pm}_{z_s,z_r}.
-\end{equation}
-In these relations, the projection on the left picks out the vertical velocity component
-of a downgoing wavefield at $z=z_r$: that is,
+  {\cal S}\left(-\frac{1}{2}\Lambda_{z_s}f_s,f_s\right)^T \approx 0.
+\end{equation}
+
+That is, the same data gather on $\Sigma_r$ can be obtained either
+with a pressure source $h_s$ or with a velocity source $f_s$, and
+there is a definite relation betwen the two. To illustrate this
+relation, recall that for the first of the two examples introduced
+above. The velocity source $f_s$ is twice the pressure trace shown in
+Figure \ref{fig:dsrcphh0}, and the pressure source (Figure \ref{fig:dhshh0}) is twice the normal
+velocity trace shown in Figure \ref{fig:dsrcvzhh0}.
+Figure \ref{fig:daltplh0} shows $\Pi_0{\cal S}\Pi_1^Tf_s$, Figure
+\ref{fig:dfwdplh0bis} shows $\Pi_0{\cal S}\Pi_0^Th_s$, and Figure
+\ref{fig:ddifffwdaltplh0} shows the difference on the same color scale.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:dsrcphh0}]{\includegraphics[width=0.45\textwidth]{dsrcphh0.pdf}}
+  \subfloat[\label{fig:dsrcvzhh0}]{\includegraphics[width=0.45\textwidth]{dsrcvzhh0.pdf}}
+  \caption{(a) Pressure gather for Lens example on source surface
+    $\Sigma_s=\{z=1000\mbox{ m}\}$, generated by point source at
+    $(3500,3500)$ m. (b) Normal velocity gather for Lens example on source surface
+    $\Sigma_s=\{z=1000\mbox{ m}\}$, generated by point source at
+    $(3500,3500)$ m.}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:dfwdplh0bis}]{\includegraphics[width=0.32\textwidth]{dfwdplh0.pdf}}
+  \subfloat[\label{fig:daltplh0}]{\includegraphics[width=0.32\textwidth]{daltplh0.pdf}} 
+  \subfloat[\label{fig:ddifffwdaltplh0}]{\includegraphics[width=0.32\textwidth]{ddifffwdaltplh0.pdf}}
+  \caption{(a) Pressure gather on receiver surface
+    $\Sigma_r=\{z=1000\mbox{ m}\}$ due to extended pressure source = 2
+    $\times$ normal velocity gather shown in Figure
+    \ref{fig:dsrcvzhh0}. This is the same gather as shown in Figure
+    \ref{fig:dfwdplh0}, and the extended pressure source has been
+    shown in Figure \ref{fig:dhshh0}.
+    (b) Pressure gather on receiver surface
+    $\Sigma_r=\{z=1000\mbox{ m}\}$ due to extended velocity source = 2
+    $\times$ pressure gather shown in Figure
+    \ref{fig:dsrcvzhh0}. (c) Difference (b)-(a), plotted on same color
+    scale.}
+\end{figure}
+
+One other important consequence of this duality is an alternate view
+of the time reversal construction of the approximate inverse. The
+entire construction can be carried out with traces of the normal
+velocity in place of trace of the pressure. The steps are precisely
+analogous, and justified in exactly the same way. Therefore I will
+simply summarize the result:
+to invert a pressure gather $d$ on $\Sigma_r$
+for a source $f_s$ on $\Sigma_s$ that approximately reproduces it ($d
+\approx \Pi_0{\cal S}\Pi_1^Tf_s$):
+\begin{itemize}
+\item[1. ] create a source $f_r = - 2d$ on $\Sigma_r$;
+\item[2. ] apply the adjoint $(\Pi_1{\cal S}\Pi_1^T)^T=\Pi_1{\cal
+    S}^T\Pi_1^T$ (solve the reverse time system
+  \begin{eqnarray}
+  \label{eqn:awerecglobvz}
+  \frac{1}{\kappa}\frac{\partial\hat{p}}{\partial t} & = & 
+                                                      - \nabla \cdot \hat{\bv}, \nonumber \\
+  \rho\frac{\partial \hat{\bv}}{\partial t} & = & f_r {\bf n}r
+  \delta_{\Sigma_r} - \nabla \hat{p}, \nonumber \\
+  \hat{p} & = & 0 \mbox{ for } t\gg 0, \nonumber \\
+    \hat{\bv} & = & {\bf 0} \mbox{ for } t \gg 0.
+  \end{eqnarray}
+\item[3. ]
+  then extract $h_s=2{\bf n_s}\cdot P_s\hat{\bv}. = 2\Pi_1{\cal
+    S}^T\Pi_1^Tf_r=-4\Pi_1{\cal S}^T\Pi_1^Td$.
+\end{itemize}
+The upshot of this construction is the alternate computation for the
+approximate inverse:
+\begin{equation}
+  \label{eqn:altinv}
+  S^{\dagger}  \approx -4\Pi_1{\cal S}^T\Pi_1^T.
+\end{equation}
+The first row of
+\ref{eqn:snull}, slightly rearranged, reads
+\begin{equation}
+  \label{eqn:lamidea}
+  \Pi_0{\cal S}\Pi_1^T f_s \approx  \frac{1}{2} S\Lambda_sf_s.
+\end{equation}
+Apply the approximate inverse construction in the form
+\ref{eqn:altinv} to solve \ref{eqn:lamidea} for $\Lambda_s$:
+\begin{equation}
+  \label{eqn:lamident}
+  \Lambda_s \approx -8\Pi_1{\cal S}^T\Pi_1^T\Pi_0 {\cal S}\Pi_1^T.
+\end{equation}
+This identity is the major result of this section: it shows how
+to compute that action of $\Lambda_s$ by propagating the input
+pressure trace, identified as a source for the velocity evolution,
+forward in time 
+reading off the pressure trace on $\Sigma_r$, identifying it once more as
+a point load (source for velocity), propagating it backwards in time, and finally reading off the velocity trace, scaled to be a
+pressure evolution source on $\Sigma_s$.
+
+The importance of this result lies in the failure of the obvious
+method for computing the action of $\Lambda_s$, namely to
+employ the pressure trace as a source in the velocity equation ($f_s$,
+in the notation used above) at $\Sigma_s$, and read off the velocity
+field also at $\Sigma_s$. This difficulty is related to the existence of
+tangentially propagating waves and the lack of continuity of the trace
+operator. The method implicit in equation \ref{eqn:lamident} avoids
+this difficulty by propagating the fields a positive distance in $z$:
+assuming as always that the causal fields are downgoing, this step
+eliminates any tangentially propagating fields from consideration.
+
+It is implicit in the discussion above of the pressure-to-source operator 
+depends only on the model coefficients near the source surface
+$\Sigma_s$. However an even more useful
+observation is that the calculations in the approximation
+\ref{eqn:lamident} could just as well be carried out in a much smaller
+region around the source surface, and produce a result that is
+functionally identical in that it will serve as a source for the same
+acoustic fields globally, with small error. Take advantage of this
+observation by replacing - for the purpose of evaluating the
+expression \ref{eqn:lamident} only -  the receiver surface by an
+auxiliary receiver surface $\Sigma_a$ near $\Sigma_s$ but a small positive
+distance from it. Define the auxiliary operator ${\cal S}_a$ by
+sampling solution of the system \ref{eqn:awepm} on $\Sigma_a$ rather
+than $\Sigma_r$. Thus obtain an approximation for $\Lambda_s$
+\begin{equation}
+  \label{eqn:lamnear}
+ \Lambda_s \approx -8\Pi_1{\cal S}_a^T\Pi_1^T\Pi_0 {\cal S}_a\Pi_1^T. 
+\end{equation}
+
+Using an auxiliary receiver datum closer to the source surface to
+compute that pressure-to-source maps (only!) has two favorable consequences:
+\begin{itemize}
+\item The computational domain can be smaller than is necessary to
+  simulate the target data, as it need only contain the source
+  surface and the receiver datum implicit in
+  equation \ref{eqn:lamident}. If the domain is small enough, the ray
+  conditions imply that no ray re-enters the region between the source
+  suface and the auxiliary surface, therefore an absorbing boundary
+  condition can be applied outside of this region, considerably
+  shrinking the computational domain and substantial improving computational
+  efficiency. The results shown below used such a truncation strategy.
+\item Since the receiver data may be chosen much closer to the
+  source surface that is the case for the target data, the effective
+  aperture active in the relation \ref{eqn:lamident} can be much
+  larger, producing an estimated source gather much less affected by
+  aperture limitation.
+\end{itemize}
+
+Figures
+\ref{fig:preddnshshh0} and \ref{fig:ddiffnshshh0} show an example of
+this construction.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:preddnshshh0}]{\includegraphics[width=0.45\textwidth]{preddnshshh0.pdf}}
+  \subfloat[\label{fig:ddiffnshshh0}]{\includegraphics[width=0.45\textwidth]{ddiffnshshh0.pdf}}
+  \caption{(a) Pressure source gather = image
+  under pressure-to-source operator $\Lambda_s$ of pressure gather
+  shown in Figure \ref{fig:dsrcphh0}, computed using
+  auxiliary receiver surface``near''  at $z=2900$ m and equation \ref{eqn:lamnear}. Compare Figure
+  \ref{fig:dhshh0}: because the sources and
+  receivers are close, little aperture is lost in this case. (b)
+  Difference between data plotted in (a)
+  and source gather
+  inferred from vertical velocity at source
+  surface ($z=3000$ m) (Figure \ref{fig:dhshh0}). Same color scale as in Figure
+  (a).}
+\end{figure}
+
+In the iterative methods to ne discussed in the next section, it will
+be important to be able to compute the transpose of the
+pressure-to-source operator - not approximately, but to machine
+precision. The prescription \ref{eqn:lamnear} only
+defines an approximation. However simply substituting the right-hand
+side of equation \ref{eqn:lamnear} - in effect, substituting the
+approximation for the operator that it approximates - gives an
+operator variant of $\Lambda_s$ with an accessible
+machine-precision adjoint:
+\begin{equation}
+  \label{eqn:lamneartr}
+ \Lambda_s^T \approx -8\Pi_1{\cal S}_a^T\Pi_0^T\Pi_1 {\cal S}_a\Pi_1^T. 
+\end{equation}
+From now on, when $\Lambda_s$ or $\Lambda_s^T$ appear, the symbols are
+understood to mean the approximations \ref{eqn:lamnear} and
+\ref{eqn:lamneartr}.
+
+Figure \ref{fig:preddnshstrhh0} shows the image of the pressure gather
+in Figure \ref{fig:dsrcphh0} under $\Lambda_s^T$ approximated as in
+equation \ref{eqn:lamneartr},
+using the auxiliary receiver surface at $z=2900$. Note
+the close resemblance to the image of the same pressure gather under
+$\Lambda_s$ displayed in Figure
+\ref{fig:preddnshshh0}. The difference of these two images is
+displayed in \ref{fig:ddiffnslamtrhh0}, on the same color scale as the
+images themselves. Note that propagation takes place entirely in a
+region where the mechanical parameters are homogenous.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:preddnshstrhh0}]{\includegraphics[width=0.45\textwidth]{preddnshstrhh0.pdf}}
+  \subfloat[\label{fig:ddiffnslamtrhh0}]{\includegraphics[width=0.45\textwidth]{ddiffnslamtrhh0.pdf}}
+  \caption{(a) Pressure source gather = image
+  under transposed pressure-to-source operator $\Lambda_s^T$ of pressure gather
+  shown in Figure \ref{fig:dsrcphh0}, computed using
+  auxiliary receiver surface``near''  at $z=2900$ m and equation \ref{eqn:lamneartr}. Compare Figure
+  \ref{fig:dhshh0}: because the sources and
+  receivers are close, little aperture is lost in this case. (b) Difference between (a) 
+  and source gather inferred from vertical velocity at source
+  surface ($z=3000$ m) (Figure \ref{fig:dhshh0}). Same color scale as (a).}
+\end{figure}
+
+\section{Accelerated Iterative Inversion}
+
+The preceding sections provide an approximate inverse for the modeling
+operator $S$ and a means to compute it effectively. These constitute
+the basic ingredients of an effective preconditioning strategy. The
+remaining tasks are: the construction of weighted norms making $S$
+approximately unitary (insofar as possible), definition of a formal
+preconditioned conjugate gradient algorithm; and modification of the
+symmetrized approximate inverse to account for the presence of the
+penalty term. Examples illustrate the effectiveness of the resulting
+preconditioning strategy.
+
+\subsection{Symmetrization and Weighted Norms}
+Since $\Lambda_s$ approximates its transpose (equation \ref{eqn:lamappsim}), it also approximates its
+ symmetrization:
+\begin{equation}
+  \label{eqn:lamsymm}
+  \frac{1}{2}(\Lambda_s^T+\Lambda_s) \approx \Lambda_s.
+\end{equation}
+and similarly for $\Lambda_r$.
+
+Define
+\begin{eqnarray}
+  W_m^{-1}&=& \frac{1}{2}(\Lambda_s^T+
+              \Lambda_s),\nonumber \\
+  W_d &=& \frac{1}{2}(\Lambda_r^T+ \Lambda_r)
+          \label{eqn:wdef}
+\end{eqnarray}
+(the subscripts signify ``model space'' and ``data space'' of $S$).
+These operators define (semi)-norms in the spaces of finite energy
+pressure fields on $\Sigma_s$ and $\Sigma_r$:
 \[
--\frac{1}{2}\Pi_1{\cal S}^{\pm}_{z_s,z_r}\Pi_0^T\Lambda^{\pm}_{z_s}P_sp^+
-=-\frac{1}{2}P_r v_z^+,
+  \|p_s\|^2 = \langle p_s, W_m p_s\rangle,  \|p_r\|^2 = \langle p_r,
+  W_d p_r\rangle,
 \] 
-where $(p^+, \bv^+)$ solve the system \ref{eqn:awepm} with $f_s=0$ and
-$h_s = \Lambda^{\pm}_{z_s}P_sp^+$. On the other hand, from relation
-\ref{eqn:tracejump10},
+The identification of the symmetrized $\Lambda_s$ as the inverse
+of another operator $W_m$ is formal at this point, since the former operator is
+likely to have null (or nearly-null) vectors due to aperture-related
+amplitude loss, and certainly due to asymptotic vanishing at
+non-propagating waves. Since some version of $W_m$ is essential in the
+formulation for effective preconditioning, I will derive a usable
+candidate to stand in for it below.
+
+The adjoint of $S$ with respect to the weighted norms is given
+approximately by
+\begin{equation}
+\label{eqn:wadj}
+S^{\dagger} \approx  W_m^{-1}S^TW_d,
+\end{equation}
+As we have shown (equations \ref{eqn:baseinv}, \ref{eqn:rightinv},
+$S^{\dagger}$ is an approximate right inverse to $S$: repeating
+equation \ref{eqn:rightinv} for convenience,
+\begin{equation}
+  \label{eqn:wunitary}
+  SS^{\dagger} \approx I.
+\end{equation}
+This is not quite enough to establish that
+$S$ is approximately unitary with respect to the
+weighted norms defined by $W_m$ and $W_d$, since that would require
+that $S^{\dagger}$ also be an approximate left inverse. The ray
+conditions assumed throughout include the presumption that all rays
+carrying high frequency energy arriving at the receiver surface
+(within the support of the sampling operator) pass
+over the source surface. Consequently a source supported on the source
+surface exists that produces this data. However it was not assumed
+that all rays passing over the source surface arrive at the receiver
+surface. Therefore high frequency energy can be emitted from the
+source surface without arriving at the receiver surface. We have seen
+examples of this phenomenon already. The extended source shown in Figure
+\ref{fig:ndwdataresrcplbulksm0} generates the diving wave data shown
+in Figure \ref{fig:ndwdata0} to close approximation, even though the
+latter was generated by a point source. The difference between the
+extended source in Figure \ref{fig:ndwdataresrcplbulksm0} and the
+point source depicted in Figure \ref{fig:ooplrays0} is an approximate
+null vector for the modeling operator $S$ as defined for this
+example.
+
+In fact, that is true for the difference between $S^{\dagger}Sh$ and
+$h$ for any extended source $h$:
 \[
-  P_r v_z^+ = -\frac{1}{2}\Lambda^+_{z_r}P_r p^+
+  S(S^{\dagger}Sh - h) = (SS^{\dagger}-I)Sh \approx 0.
 \]
-where
+Suppose that $n$ is an approximate null vector for $S$, and $h$ is any
+other extended source. Then
 \[
-  P_r p^+ = \Pi_0{\cal S}^{+}_{z_s,z_r}\Pi_0^T\Lambda^{+}_{z_s}P_s
-  p^+
+  \langle S^{\dagger}S h, n \rangle_m = \langle Sh, Sn \rangle_d
+  \approx 0.
 \]
-\[
-  = S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_sp^+
-\]
-Therefore combining the last two equations with \ref{eqn:snull2},
-obtain
-\begin{equation}
-  \label{eqn:sv}
-  \frac{1}{4}\Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s} = V^+_{z_s,z_r}.
-\end{equation}
-This is the promised relation.
-
-As shown in the last section, $4(V_{z_s,z_r}^+)^T$ is approximately
-inverse to $S^{+}_{z_s,z_r}$. Therefore, transposing both sides of
-equation \ref{eqn:sv} and using \ref{eqn:approxinv}, obtain
-\begin{equation}
-  \label{eqn:almostunitary}
-  4(V_{z_s,z_r}^+)^TS^+_{z_s,z_r} = [ (\Lambda^+_{z_s})^T
-  (S^{+}_{z_s,z_r})^T(\Lambda^+_{z_r})^T]S^{+}_{z_s,z_r} \approx I.
-\end{equation}
-
-The remarkable feature of the identity \ref{eqn:almostunitary} is that
-it exhibits an approximate right inverse of $S^+$ as an adjoint with
-respect to a weighted inner product - or it would, if the operators
-$(\Lambda^+)$ were symmetric positive definite. As noted earlier,
-these operators are only approximately symmetric, though they are
-positive semi-definite. That is not a great obstacle, however:
-symmetrizing them in the obvious way commits a negligible error, of
-the sort that this paper already neglects wholesale. That is,
-\begin{equation}
-  \label{eqn:unitary}
-  [ \frac{1}{2}((\Lambda^+_{z_s})^T+ \Lambda^+_{z_s})
-  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_r})^T+
-  \Lambda^+_{z_r})]S^{+}_{z_s,z_r} \approx I.
-\end{equation}
-
-The symmetrized $\Lambda$ operators are at least positive
-semi-definite, hence define (at least) semi-norms.
-Similar relations have been derived for other scattering operators,
+That is, $S^{\dagger}S$ is the (approximate) orthogonal projector onto
+the orthocomplement of the null space of $S$. So iterations such as CG
+based on $S^{\dagger}S$ construct approximation sequences lying close
+to this orthocomplement, as if the null space did not exist. This
+observation allows us to treat $S$ as if it were unitary in discussing
+the behaviour of Krylov space algorithms based on the weighted spaces,
+even though it actually is not.
+
+Similar relations have been derived for other wave-theoretic modeling operators,
 and have been used to accelerate iterative solutions of inverse
@@
 
-\section{Accelerated Iterative Inversion}
-
-For convenience, in this section write $S$ in place of
-$S^+_{z_s,z_r}$. Also abbreviate the symmetrized $\Lambda$ operators
-using notation suggesting weight
-operators in model and data spaces:
-\begin{eqnarray}
-  W_m^{-1}&=& \frac{1}{2}((\Lambda^+_{z_s})^T+
-              \Lambda^+_{z_s}),\nonumber \\
-  W_d &=& \frac{1}{2}((\Lambda^+_{z_r})^T+ \Lambda^+_{z_r}).
-          \label{eqn:wdef}
-\end{eqnarray}
-The identification of the symmetrized $\Lambda^+_{z_s}$ as the inverse
-of another operator $W_m$ is formal, since the former operator is
-likely to have null (or nearly-null) vectors due to aperture-related
-amplitude loss. Since some version of $W_m$ is essential in the
-formulation for effective preconditioning, I will derive a usable
-candidate to stand in for it below.
-
-Adopting Hilbert norms defined by the operators $W_m$ and $W_d$ in its
-domain and range respectively, the adjoint of $S$ is given by
-\begin{equation}
-\label{eqn:wadj}
-S^{\dagger} = W_m^{-1}S^TW_d,
-\end{equation}
-
-In this notation, the relation \ref{eqn:unitary} takes the form
-\begin{equation}
-  \label{eqn:wunitary}
-  S^{\dagger}S \approx I.
-\end{equation}
-That is to say, $S$ is approximately unitary with respect to the
-weighted norms defined by $W_m$ and $W_d$. Therefore a Krylov space
-method employing these norms will converge rapidly, at least for the
-well-determined components of the solution.
+Concerning computation, just as for the pressure-to-source operators,
+the weight operators are to be understood as replaced by their
+approximations via equations \ref{eqn:lamnear} and \ref{eqn:lamneartr}:
+\[
+W_m^{-1} = \frac{1}{2}(\Lambda_s +
+\Lambda_s^T)
+\]
+\begin{equation}
+  \label{eqn:winvcomp}
+  -4\Pi_1 {\cal S}_a^T(\Pi_0^T\Pi_1+ \Pi_1^T\Pi_0){\cal S}_a\Pi_0.
+\end{equation}
+with a similar approximate realization of $W_d$.
+
+Figure \ref{fig:symmdnshshh0} shows the output of the symmetrized
+approximate pressure-to-source operator per equation \ref{eqn:winvcomp},
+applied once again to the pressure data in Figure
+\ref{fig:dsrcphh0}. Note the resemblance to Figures
+\ref{fig:preddnshshh0} and \ref{fig:preddnshstrhh0}. These are all
+asymptotic approximations of each other. Figure
+\ref{fig:ddiffsymmdnshsll0} shows the
+ the difference between the pressure gather at $z=z_r$ produced from
+ the pressure source output by the symmetrized $\Lambda$, and the point source
+simulation (Figure \ref{fig:dfwdplh0}), plotted on the same scale as
+the latter, in both cases with all propagations in the lens model.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:symmdnshshh0}]{\includegraphics[width=0.45\textwidth]{symmdnshshh0.pdf}}
+  \subfloat[\label{fig:ddiffsymmdnshsll0}]{\includegraphics[width=0.45\textwidth]{ddiffsymmdnshsll0.pdf}}
+  \caption{(a) Pressure source gather = image
+  under symmetrized pressure-to-source operator $W_m^{-1}$ applied to pressure gather
+  shown in Figure \ref{fig:dsrcphh0}, computed using
+  auxiliary receiver surface``near''  at $z=2900$ m and 
+  equation \ref{eqn:winvcomp}, . Compare Figures
+  \ref{fig:dhshh0}, \ref{fig:preddnshshh0} and
+  \ref{fig:preddnshstrhh0}: these are all asymptotic approximations of
+  each other. (b) Difference between pressure gather at the receiver
+  surface $\Sigma_r$  ($z=1000$ m) modeled using symmetrized
+  pressure-to-source gather in (a), and pressure gather shown in Figure \ref{fig:dfwdplh0},
+  propagated in lens model from 
+  source gather shown in Figure \ref{fig:dhshh0} (inferred from vertical velocity at source
+  surface ($z=3000$ m)). Same color scale as
+  Figures \ref{fig:drecplh0} and \ref{fig:dfwdplh0}.}
+\end{figure}
+
+The identity \ref{eqn:winvcomp} shows that only one forward and one adjoint simulation
+are necessary to compute the action of $W_m^{-1}$ or $W_d$, and only
+over the truncated computationa domain between the source or receiver
+surface and its auxiliary, nearby surface. The operator
+in the center of the expression on the right-hand side, $\Pi_0^T\Pi_1
++ \Pi_1^T\Pi_0$, simply exchanges the components of the acoustic
+fields, passing the velocity field as a pressure source and the
+pressure field as a velocity source.
+
+One more computation is required for the full implementation of the
+preconditioning strategy explained in the last section: $W_m$ is
+required, not just $W_m^{-1}$. Note that $W_m$ plays two roles in the
+second term in equation \ref{eqn:norm1}, below: it is the weight matrix for
+both the domain and range norms for $A$. It is perfectly OK for one of
+these to be replaced by an asymptotic approximation, so long as it is
+symmetric and computable (and at least semi-definite). Manipulations
+very similar to those used to obtain equation \ref{eqn:winvcomp} lead
+to the approximation
+\begin{equation}
+  \label{wcomp}
+ W_m \approx -\frac{1}{16}\left(\Pi_0 {\cal S}_a^T
+   (\Pi_1^T \Pi_0 + \Pi_0^T \Pi_1) {\cal S}_a \Pi_0^T\right)
+\end{equation}
+Comparison with the definition \ref{eqn:winvcomp} shows that
+$W_m$ and $W_m^{-1}$ differ only in the initial and
+final projection factors (and overall scale), and in particular either
+can be computed for the cost of a forward/adjoint operator pair. Note
+that $W_m^{-1}$ is inverse to $W_m$ only in an
+approximate (asymptotic, aperture-limited) sense.
+
+\subsection{Preconditioned Conjugate Gradient Iteration}
 
@@
 powers of $I + \alpha^2A^TA$ are trivial to compute, given the choice
-of $A$ made here. We will examine fast algorithms for computing
-$W_m^{-1}$ = the symmetrized pressure-to-source operator in the next
-section. Note that only $M^{-1}$, hence only $W_m^{-1}$, appears in
-Algorithm 1.
-
-\section{Computing and Symmetrizing $\Lambda$}
-
-Computations of $\Lambda^{\pm}_{z_s}$ and its transpose are clearly critical steps in an
-implementation of the PCG algorithm outlined in the preceding section.
-Direct computation of the pressure-to-source operator $\Lambda^{\pm}_{z_s}$, for instance by solving
-\ref{eqn:awepm} and reading off $P_sv^{\pm}_{z}$, turns out to be
-numerically ill-behaved. The relation \ref{eqn:snull} provides and
-alternative approach, taking advantage of the accurate approximate
-inverse to $S^+_{z_s,z_r}$ constructed above. The first row of
-\ref{eqn:snull}, slightly rearranged, is
-\begin{equation}
-  \label{eqn:lamidea}
-  \Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx  -\frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
-\end{equation}
-The approximate inverse construction for $S^++_{z_s,z_r}$ permits (approximate)
-solution of this equation for $\Lambda^+_{z_s}f_s$: apply
-$4(V^{+}_{z_s,z_r})^T$ to both sides of equation \ref{eqn:lamidea} and
-use the first equation in the list \ref{eqn:approxinv} to get
-\begin{equation}
-  \label{eqn:lamident}
-  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
-\end{equation}
-This identity is the major result of this section: it shows how
-to compute that action of $\Lambda^+_{z_s}$ by propagating the input
-pressure trace, identified as a source for the velocity evolution,
-forward in time from $z_s$ to $z_r$
-reading off the pressure trace on $z=z_r$, identifying it once more as
-a point load (source for velocity), propagating it backwards in time from $z_r$ to
-$z_s$, and finally reading off the velocity trace, interpreted as a
-pressure evolution source on $z_s$.
-
-The importance of this result lies in the failure of the obvious
-method for computing the action of $\Lambda^{\pm}_{z_s}$, namely to
-employ the pressure trace as a source in the velocity equation ($f_s$,
-in the notation used above) at $z=z_s$, and read off the velocity
-field also at $z=z_s$. This difficulty is related to the existence of
-tangentially propagating waves and the lack of continuity of the trace
-operator. The method implicit in equation \ref{eqn:lamident} avoids
-this difficulty by propagating the fields a positive distance in $z$:
-assuming as always that the causal fields are downgoing, this step
-eliminates any tangentially propagating fields from consideration.
-
-A deeper study of the pressure-to-source operator (or of the closely
-related Dirichlet-to-Neumann operator for the second order wave
-equation, see \cite{StefUhl:05}) shows that it is approximately
-dependent only on the model coefficients near the source surface
-($z=z_s$ in this case). Since the homogeneous and lens models are
-identical near this surface, it is unsurprising that these figures are
-very close to the previous two.  However an even more useful
-observation is that the calculations in the approximation
-\ref{eqn:lamident} could just as well be carried out in a much smaller
-region around the source surface, and produce a result that is
-functionally identical in that it will serve as a source for the same
-acoustic fields globally, with small error. In effect, equation
-\ref{eqn:lamident} involving propagation from source ($z=z_s$) to
-receiver ($z=z_r$) surfaces is altered by replacing $z_r$ with a
-receiver datum $z_s+\Delta z$ considerably closers to $z_s$:
-\begin{equation}
-  \label{eqn:lamnear}
-  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
-\end{equation}
-Using a receiver datum closer to the source surface has two favorable consequences:
-\begin{itemize}
-\item The computational domain can be smaller than is necessary to
-  simulate the target data, as it need only contain the source
-  surface and the receiver datum implicit in
-  equation \ref{eqn:lamident}. This shrinkage of the computational
-  domain can lead to substantial improvements in computational
-  efficiency.
-\item Since the receiver data may be chosen much closer to the
-  source surface that is the case for the target data, the effective
-  aperture active in the relation \ref{eqn:lamident} can be much
-  larger, producing an estimated source gather much less affected by
-  aperture limitation.
-\end{itemize}
-
-As mentioned in the last section, computation of the transpose of
-$\Lambda^+$ (exact, not approximate in the high frequency sense) is
-critical to the successful construction of the preconditioner. The
-relation \ref{eqn:lamident} does not provide a computation for this
-operator. However set
-\begin{equation}
-  \label{eqn:lamtilde}
-  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
-\end{equation}
-Then \ref{eqn:lamnear} can be rewritten
-\[
-  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
-\]
-Of course, all of the examples so far show images of
-$\tilde{\Lambda}^+_{z_s}$.
-
-Since successful preconditioning requires only approximate inversion,
-use of $\tilde{\Lambda}^+_{z_s} $ in place of $\Lambda^+_{z_s}$ will
-still yield a working preconditioner, and the former can be transposed
-to machine precision via the definition \ref{eqn:lamtilde} and the adjoint state
-method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
-\begin{equation}
-  \label{eqn:lamtransp}
-  (\tilde{\Lambda}^+_{z_s})^T = -8 \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
-    z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}
-\end{equation}
-
-The model space weight operator $W_m^{-1}$ introduced in the last section is
-replaced by its asymptotic approximation
-\[
-\frac{1}{2}(\tilde{\Lambda}^+_{z_s} +
-  (\tilde{\Lambda}^+_{z_s})^T)
-\]
-\[
-  \approx -8 \left((V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
-    S}^+_{z_s,z_s+\Delta z}\Pi_1^T \right.+
-\]
-\[
-  \left. \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
-      z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}\right)
-\]
-\begin{equation}
-  \label{eqn:winvcomp}
-  = -4(\Pi_1 ({\cal S}^+_{z_s,z_s+\Delta z})^T(\Pi_0^T\Pi_1 +
-  \Pi_1^T\Pi_0) {\cal S}^+_{z_s,z_s+\Delta z}\Pi_1^T) = \tilde{W}_m^{-1} .
-\end{equation}
-with a similar definition for the replacement $\tilde{W}_d$ of $W_d$.
-
-This identity shows that only one forward and one adjoint simulation
-are necessary to compute the action of $\tilde{W}_{m,d}$. The operator
-in the center of the expression on the right-hand side, $\Pi_0^T\Pi_1
-+ \Pi_1^T\Pi_0$, simply exchanges the components of the acoustic
-fields, passing the velocity field as a pressure source and the
-pressure field as a velocity source.
-
-One more computation is required for the full implementation of the
-preconditioning strategy explained in the last section: $W_m$ is
-required, not just $W_m^{-1}$. Note that $W_m$ plays two roles in the
-second term in equation \ref{eqn:norm1}: it is the weight matrix for
-both the domain and range norms for $A$. It is perfectly OK for one of
-these to be replaced by an asymptotic approximation, so long as it is
-symmetric and computable (and at least semi-definite). The second row
-in equation \ref{eqn:snull} appears as \ref{eqn:snull2} above:
-introducing (formally) the inverse of $\Lambda^+$,
-\begin{equation}
-  \label{eqn:snull2mod}
--\frac{1}{2}\Pi_1{\cal S}^{+}_{z_s,z_r}\Pi_0^T  \approx
-V^{+}_{z_s,z_r}(\Lambda^{+}_{z_s})^{-1}
-\end{equation}
-whence from the second line in display \ref{eqn:approxinv}
-\begin{equation}
-  \label{eqn:laminv}
--\frac{1}{8}(S^+_{z_s,z_r})^T\Pi_1{\cal S}^{+}_{z_s,z_r}\Pi_0^T  
-\approx (\Lambda^{+}_{z_s})^{-1}
-\end{equation} 
-and
-\begin{equation}
-  \label{eqn:laminvtr}
--\frac{1}{8}\Pi_0({\cal S}^{+}_{z_s,z_r})^T\Pi_1^T S^+_{z_s,z_r}
-\approx ((\Lambda^{+}_{z_s})^{-1})^T.
-\end{equation}
-Using the definition \ref{eqn:sdef} of $S^+_{z_s,z_r}$, the
-symmetrized $\Lambda^{-1}$ is
-\begin{equation}
-  \label{wcomp}
- \tilde{W}_m = -\frac{1}{16}\left(\Pi_0 ({\cal S}^{+}_{z_s,z_r})^T
-   (\Pi_1^T \Pi_0 + \Pi_0^T \Pi_1) {\cal S}^{+}_{z_s,z_r} \Pi_0^T\right)
-   \approx \frac{1}{2}((\Lambda^{+}_{z_s})^{-1}
-   +((\Lambda^{+}_{z_s})^{-1})^T) .
-\end{equation}
-Comparison with the definition \ref{eqn:winvcomp} shows that
-$\tilde{W}_m$ and $\tilde{W}_m^{-1}$ differ only in the initial and
-final projection factors (and overall scale), and in particular either
-can be computed for the cost of a forward/adjoint operator pair. Note
-that $\tilde{W}_m^{-1}$ is inverse to $\tilde{W}_m$ only in an
-approximate (asymptotic, aperture-limited) sense.
+of $A$ made here.
+
+\subsection{Example}
+This final subsection shows that result of Conjugate Gradient
+iteration, with and without preconditioning, applied to the source
+estimation problem \ref{eqn:esis}, with zero and non-zero penalty
+weight $\alpha$. The data $d$ is the gather shown in
+\ref{fig:drecplh0}, simulated using the lens model (Figure \ref{fig:ooplrays0}) with source shown
+in Figure \ref{fig:dsrcphh0}, or, alternatively, a point source with
+bandpass filter wavelet located at $x_d=3500$ m, $z_d=3500$ m. In the inversion,
+the material model is taken to be homogeneous, as has been the case in
+all of the previous examples.
+
+Figure \ref{fig:compnres0lh0} shows the progress of the normal residual
+(Euclidean norm of the difference of the two sides of equation \ref{eqn:norm1}),
+for Conjugate Gradient and Preconditioned Conjugate Gradient
+(Algorithm 1) iterations, applied to solution of the optimization
+problem \ref{eqn:einv} with $\alpha=0$. Because the normal residual is%
+measured in the Euclidean norm for both CG and PCG, the results are
+comparable. For PCG, the preconditioner $M$ is given by equation
+\ref{eqn:defprecond}, with the symmetrized $\Lambda$s computed as
+indicated in the preceding section with auxiliary source and
+receiver surfaces shifted by 100 m. Convergence for the
+preconditioned algorithm is roughly 4 times as fast: this speed
+comparison is much more obvious in a vertically stretched version of
+the figure (Figure \ref{fig:compnrestrunc0lh0}.
+
+%\plot{compnres0lh0}{width=\textwidth}{Comparison of normal residual 
+%  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+%  vs. iteration. Data = lens model, point source (Figure 
+%  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
+%  $\alpha=0$.}
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:compnres0lh0}]{\includegraphics[width=0.45\textwidth]{compnres0lh0.pdf}}
+  \subfloat[\label{fig:compnrestrunc0lh0}]{\includegraphics[width=0.45\textwidth]{compnrestrunc0lh0.pdf}}
+  \caption{(a) Comparison of normal residual 
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+  vs. iteration. Data = lens model, point source (Figure 
+  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
+  $\alpha=0$. (b) Same comparison, with expanded vertical scale, to
+  show more clearly that PCG ir roughly 4-5 times as fast as (in
+  achieving a given normal residual) as CG.}
+\end{figure}
+
+Figure \ref{fig:compnres1lh0} shows the same comparison with non-zero
+penalty weight, $\alpha=10^{-3}$. Convergence for both CG and PCG is
+somewhat slower than in the $\alpha=0$ case. However the relation is
+about the same, still 4-5 times as many iterations to achieve a given
+normal residual level with CG as with PCG.
+
+\begin{figure}
+  \centering
+  \subfloat[\label{fig:compnres1lh0}]{\includegraphics[width=0.45\textwidth]{compnres1lh0.pdf}}
+  \subfloat[\label{fig:compnrestrunc1lh0}]{\includegraphics[width=0.45\textwidth]{compnrestrunc1lh0.pdf}}
+  \caption{(a) Comparison of normal residual 
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+  vs. iteration. Data = lens model, point source (Figure 
+  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
+  $\alpha=10^{-3}$. (b) Same comparison, with expanded vertical scale, to
+  show more clearly that PCG ir roughly 4-5 times as fast as (in
+  achieving a given normal residual) as CG.}
+\end{figure}
+
+%\plot{compnres1lh0}{width=\textwidth}{Comparison of normal residual 
+%  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+%  vs. iteration. Data = lens model, point source (Figure 
+%  \ref{fig:drecplh0}), inversion in homogenous model. Penalty weight 
+%  $\alpha=10^{-3}$.}
 

### 2024-02-15 repro/papers/GEM22/submitR2/paper.tex
cleaned up

@@
   surface}. Assuming that the presumed point source is at location
-$\bx_s$, a simple (though certainluy not the only) choice for the
+$\bx_s$, a simple (though certainly not the only) choice for the
 penalty operator $A$ is then multiplication by the distance

### 2024-02-15 repro/papers/awi/paper.tex
cleaned up

@@
 \begin{abstract}
-Under some restricted circumstances, Adaptive Waveform Inversion
-applied to transmitted wave data yields estimates of index of
-refraction (or wave velocity) similar to those obtained by travel time inversion.
+Adaptive Waveform Inversion
+applied to transient transmitted wave data yields estimates of index of
+refraction (or wave velocity) similar to those obtained by travel time
+inversion, provided that the data contain a single smooth wavefront.
 \end{abstract}
@@
 measurements. The mechanical parameters to be inferred appear as
-coefficients in systems of partial differential equaitons governing
+coefficients in systems of partial differential equations governing
 the wave motion, and discrete representations of these parameters form
@@
 technique has been successfully applied to industry-scale seismic data
-\cite[]{GuaschWarnerRavaut:GEO19,Warneretal:SEG21} as well as to
+\cite[]{GuaschWarnerRavaut:GEO19,Warneretal:SEG21} and shows promise for
 ultrasonic imaging of the the human brain
@@
 transmitted waves
-(that is unique rays of geometric asymptotics between sources and
+(that is, unique rays of geometric asymptotics between sources and
 receivers), the AWI objective function is an approximation to the
@@
 widely believed to be a near-optimal method for wave velocity
-estimation, this demonstration explains roughly ``why it works''.
+estimation, this demonstration explains roughly ``why AWI works'', at
+least within this restricted scope.
 
@@
 objective normalization and the close relation to travel time
-inversion is completely independent from 
+inversion has nothing to do with
 the reasons for normalizing the objective advanced in the
@@
 
-Acoustic inversion, for present purposes, means: given recording time interval
-$[0,T]$, source and receiver locations $(\bx_s,\bx_r)$, source wavelet
+Acoustic inversion, for present purposes, means: given source and receiver locations $(\bx_s,\bx_r)$, source wavelet
 $w(t;\bx_s)$ and data traces $d(\bx_r,\cdot;\bx_s)$, find a model $m$
@@
 
-Note that the wavelet $w$ is assumed known in this formulation of the
+Note that in actual data acquisition, recording occurs over finite
+time intervals, so the integrals appearing in the definition
+\ref{eqn:fwi} have finite upper and lower limits, as opposed to the
+$\pm \infty$ limits implicit in the notation. It will be assumed that
+wavefield energy is causal (zero before an onset time) and decays
+rapidly enough as $t \rightarrow \infty$ that the difference between
+finite and infinite range of integration can be ignored throughout
+this discussion.
+
+The wavelet $w$ is assumed known in this formulation of the
 FWI problem. In fact, even if actual acoustic sources are

### 2024-02-15 repro/papers/awigrad/paper.tex
cleaned up

@@
 \section{Introduction}
+\section{A Penalty Version of AWI}
+The adaptive kernel $u_0^{\epsilon}$ appearing in the definition \ref{eqn:tJz} is the solution of a regularized least squares problem.
+This observation suggests augmenting the regularized least squares objective by a penalty proportional to the right-hand side of \ref{eqn:tJz}:
+\begin{equation}
+  \label{eqn:Ja}
+  \Ja[m,u;d] = \frac{1}{2}\left(\|S[m]u-d\|_d^2 + \epsilon^2\|u\| + \alpha^2\|tu\|_m\right).
+\end{equation}
+Here $\|u\|^2 = \langle u, u \rangle$ is the (unweighted) $L^2$ norm, and $\|\cdot\|_m$ is an $(m,d)$-dependent (weighted) norm: with
+\[
+  u_0^{\epsilon}[m,d] = \mbox{arg min}_u \Jz[m,u;d],
+\]
+define
+\[
+  Wu (\bx_r,t;\bx_s) = \sum_{\bx_r,\bx_s}\frac{1}
+  {\int\,ds\,|u_0^{\epsilon}[m,d](\bx_r,s;\bx_s)|^2}u(\bx_r,t;\bx_s)
+\]
+and
+\begin{equation}
+  \label{eqn:defmnorm}
+  \|u\|^2_m = \langle u, W u \rangle.
+\end{equation}
+With this notation,
+\[
+  \tJz[m;d] = \frac{1}{2}\|tu^{\epsilon}_0[m;d]\|^2_m
+\]
+In the definition of $\tJz$, a quadratic is evaluated at the minimizer of $\Jz[m,u;d]$. This suggests the possibility of evaluating the quadratic objective $\Ja[m,u;d]$ at its own minimizer:
+\begin{equation}
+  \label{eqn:tJa}
+  \tJa[m;d] = \min_u \Ja[m,u,d] = \Ja[m,\ua[m;d];d]
+\end{equation}
+This is the so-called variable projection reduction of the quadratic $\Ja$.
+Note that the right-hand side of the definition of $\tJz$ is the last term on the right-hand side of $\Ja$, scaled by $\alpha^{-2}$ and evaluated at the minimizer of the sum of the first two terms, that is, $\Ja$ for $\alpha=0$.
+
+\begin{theorem}
+  \label{thm:basic}
+Suppose that $S \in {\cal L}(U,D)$, the is a bounded linear maps from (Hilbert space) domain $U$ (``model space'') with inner product $\langle \cdot, \cdot \rangle_m$ to range $D$ with inner product $\langle \cdot, \cdot \rangle_d$.
+Assume that $S$ is coercive, that is, there exists $\epsilon >0$ so that $\|Su\|_d \ge \epsilon\|u\|_m$ for $u \in U$.
+Suppose also that $A \in {\cal L}(U,U)$, and that $d \in D$. Define the function $J_{\alpha}$ on $U$ by 
+\begin{equation}
+  \label{eqn:eq1}
+  J_{\alpha}(u) = \frac{1}{2}(\|Su-d\|^2_d + \alpha^2\|Au\|_m^2).
+\end{equation}
+Then for $\alpha \ge 0$, $J_{\alpha}$ possesses a unique minimizer $u_{\alpha} \in U$, for which
+\begin{equation}
+  \label{eqn:eq0}
+  \tilde{J}_{\alpha} = \min_{u \in U} J_{\alpha}(u) = J_{\alpha}(u_{\alpha}),
+\end{equation}
+and
+\begin{equation}
+  \label{eqn:eq4}
+  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tilde{J}_{\alpha}-\tilde{J}_0).= \frac{1}{2}\|Au_0\|^2
+\end{equation}
+This limit is uniform over bounded sets of $(S, A) \in {\cal L}(U,D) \times {\cal L}(U, U)$.
+\end{theorem}
+
+For completeness, the proof is given in an appendix.
+
+This relation will be applied in the context of a model space $U$ possessing a reference Hilbert norm $\|\cdot\|$ and a family of equivalent norms $\|\cdot\|_m$ depending on a parameter $m \in M$, in which $M$ is an open subset of a Banach space $X$.
+The dependence is through a an operator-valued function $W$ on $M$, taking values in cone ${\cal B}(U)$ of bounded self-adjoint positive-definite operators on $U$.
+On the other hand, we assume that the norm on $D$ is fixed, independent of $M$, and treat it as a reference norm.
+
+\begin{theorem}
+  \label{thm:uniform}
+Assume that $W: M \rightarrow {\cal B}(U)$ is of class $C^1$ and uniformly positive definite, and that the norm $\|\cdot\|_m$ is given by
+\begin{equation}
+  \label{eqn:normdef}
+  \|u\|_m^2 = \langle u, W[m] u \rangle.
+\end{equation}
+Suppose also that $S: M \rightarrow {\cal L}(U,D)$ is of class $C^1$ and uniformly coercive, and that $A \in {\cal L}(U,U)$ as before. For $\alpha \ge 0$, define $\tJa: M \rightarrow \bR$ by
+\[
+  \tilde{J}_{\alpha}[m] = \min_{u \in U}\frac{1}{2}(\|S[m]u-d\|^2_d + \alpha^2\|Au\|_m^2)
+\]
+\begin{equation}
+  \label{eqn:tjadef}
+  = \frac{1}{2}(\|S[m]u_{\alpha}[m]-d\|^2_d + \alpha^2\|Au_{\alpha}[m]\|_m^2)
+\end{equation}
+where
+\begin{equation}
+  \label{eqn:ua}
+  u_{\alpha}[m] = \mbox{arg min}_{u \in U}\frac{1}{2}(\|S[m]u-d\|^2_d + \alpha^2\|Au\|_m^2)
+\end{equation}
+for $m \in M$. Then $\tilde{J}_{\alpha}$ is of class $C^1(M)$, and
+\begin{equation}
+  \label{eqn:lim}
+  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tJa[m]-\tJz[m]).= \frac{1}{2}\|A\uz[m]\|_m^2.
+\end{equation}
+The limit is in the sense of $C^1(M)$, that is, both the left-hand side and its $m$ derivative converge to the right-hand side and its $m$ derivative, respectively.
+Moreover $u_{\alpha}$ is of class $C^1(M,U)$.
+\end{theorem}
+
+\begin{proof} All conclusions but the last follow immediately from Theorem \ref{thm:basic} and the uniformity assumptions on $S$ and $W$.
+
+Since the norm in $D$ is fixed for the discussion, drop the subscript and treat it as a reference norm.
+Denote the adjoint with respect to the norms $\|\cdot\|_m$ on $U$, $\|\cdot\|_d = \|\cdot\|$ on $D$ by a superscript $*$, and with respect to $\|\cdot\|$ on $U$ by a superscript $T$.
+The normal equation defining $\ua$ is
+  \[
+    (S[m]^*S[m] + \alpha^2 A[m]^*A) \ua[m] = S[m]^*d
+  \]
+  \begin{equation}
+    \label{eqn:normal}
+= (W[m]^{-1}S[m]^TS[m] + \alpha^2 W[m]^{-1}A^TW[m]A)\ua[m]= W[m]^{-1}S[m]^Td
+\end{equation}
+or equivalently
+\begin{equation}
+  \label{eqn:normal1}
+  (S[m]^TS[m] + \alpha^2 A^TW[m]A)\ua[m] = S[m]^Td
+\end{equation}
+Since $S$ is uniformly coercive, and both $S$ and $W$ are of class $C^1$, the normal operator on the left hand side is boundedly invertible and of class $C^1$.
+Therefore the solution $\ua$ is of class $C^1$.
+
+[need to add proof of $C^1$ convergence.]
+\end{proof}
+
+Apply this relation in the context of the AWI penalty problem defined earlier.
+$M$ is an open set in $C^k(\Omega, \bR^2)$.
+$U$ is $\oplus_{\bx_r,\bx_s}L^2([-T_u,T_u])$, $D$ is the data space $\oplus_{\bx_r,\bx_s}L^2([0,T])$,.
+$S$ is the regularized modeling operator $(S[m],\epsilon I)^T$, $d$ is replaced by the $(d,0) \in D \times U$,  $W$ is the weight operator on $U$ defined in \ref{eqn:defmnorm}.
+
+As noted before [but it wasn't - fix!], stationary points of $\tJz$
+are $\lambda(w_*)$-close to stationary points of the travel time
+mean-square error, in the single-arrival case. The result of this
+section implies that stationary points of $\tJa$ are close to those of
+$\tJz$. Therefore, at least for small $\alpha$, minimization of $\tJa$
+produces acoustic models that closely match the traveltimes predicted
+in single-arrival data.
 
@@
 \begin{equation}
-  \label{eqn:deriv}
+  \label{eqn:tildejgrad}
   + \alpha^2 (
@@
 every operator ($S[m]$, $K[...]$, $W$, $R$) ,appearing in the right-hand side of expression
-\ref{eqn:deriv} is block-diagonal (trace-by-trace), with the exception
+\ref{eqn:tildejgrad} is block-diagonal (trace-by-trace), with the exception
 of $DF[m]^T$. This latter is a reduction operation, i.e. a block-row
 operation, so its expression is also the same. That is, equation
-\ref{eqn:deriv} also correctly expresses the gradient in the
+\ref{eqn:tildejgrad} also correctly expresses the gradient in the
 multi-trace case.
@@
 $v_{\sigma}[m]$ is the solution of the linear system
+\[
+    (S[m]^TS[m]+ \sigma^2I) v_{\sigma}[m] =
+\]
 \begin{equation}
   \label{eqn:defvsigma}
-  (S[m]^TS[m]+ \sigma^2I) v_{\sigma}[m] =
   (W(R(W(Ru_{0,\sigma}[m],-2,1)u_{0,\sigma}),2,0) - W(Ru_{0,\sigma}[m],-2,2))
@@
 \end{equation}
-The only difference between AWI and MSWI lies in the defintiion of the
+The only difference between AWI and MSWI lies in the absence of
+scaling by the reciprocal of $Ru_{0,\sigma}[m]$ in the definition of the MSWI
 penalty operator.
@@
   \nabla \tilde{J}^{\rm unp}_{\alpha,\sigma}[m] =
-  D(S[m]u)^T_{u=u_{\alpha,\sigma}[m]}(S[m]u_{\alpha,\sigma}[m]-d)
+  D_m(S[m]u)^T_{u=u_{\alpha,\sigma}[m]}(S[m]u_{\alpha,\sigma}[m]-d)
 \]
@@
   \label{eqn:tildejunpgrad}
-  = DF[m]^T(K[u_{\alpha,\sigma}[m]]^T(S[m]u_{\alpha,\sigma}[m]-d).
+  = DF[m]^TK[u_{\alpha,\sigma}[m]]^T(S[m]u_{\alpha,\sigma}[m]-d).
 \end{equation}
+
+Note that the right-hand side of equation \ref{eqn:tildejunpgrad} is
+exactly the first term in the right-hand side of equation
+\ref{eqn:tildejgrad}. That is, the gradient of the AWI objective
+function is the gradient of the MSWI objective function, modified by
+terms that account for the $m$-dependence of the AWI penalty operator $T_{\sigma}[m]$.
 
@@
 The obvious way to use this is to create a 3-component product space for the output of the block-column operator $(S[m],\alpha T[m], \sigma I)^T$, and set the right-hand side vector (parameter {\tt b})
-to $(d,0,0)^T$. Then on return one-half the norm-squared of {\tt e} is precisely the (approximate) value of $\tilde{J}_{\alpha,\sigma}[m]$.
+to $(d,0,0)^T$. Then on return one-half the norm-squared of {\tt e} is
+precisely the (approximate) value of $\tilde{J}_{\alpha,\sigma}[m]$.
+
+\append{$\alpha \rightarrow 0$ scaled limit of penalty function}
+Define
+\begin{equation}
+  \label{eqn:eq0}
+  \tJa = \min_u J_{\alpha}(u) = J_{\alpha}(\ua)
+\end{equation}
+Since $S$ is positive definite, the minimum is well-defined for any $\alpha \ge 0$, and the minimizer $\ua \in U$ satisfies the normal equation.
+
+The claim to be established is that
+\begin{equation}
+  \label{eqn:eq4}
+  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tJa-\tJz).= \frac{1}{2}\|A\uz\|_m^2
+\end{equation}
+
+To see this, use the normal equation to write
+\[
+  \ua = (S^TS + \alpha^2 A^TA)^{-1}S^Td = (S^TS + \alpha^2 A^TA)^{-1}S^TS \uz
+\]
+\begin{equation}
+  \label{eqn:eq2}
+  = \uz - \alpha^2 (S^TS + \alpha^2 A^TA)^{-1}A^TA\uz = \uz-\alpha^2 \va
+\end{equation}
+Note that $\va = (S^TS + \alpha^2 A^TA)^{-1}A^TA\uz$ is uniformly bounded in $\alpha \ge 0$.
+
+Stuff the RHS of equation \ref{eqn:eq2} into the definition \ref{eqn:eq1} to obtain
+\begin{equation}
+  \label{eqn:eq3}
+  \tJa = \frac{1}{2}(\|S\uz-d\|_d^2 - 2 \alpha^2\langle S\uz-d, S\va\rangle_d + \alpha^2\|A\uz\|_m^2 + O(\alpha^4))
+\end{equation}
+The second term on the RHS of equation \ref{eqn:eq3} vanishes thanks to the normal equation, and the first term is precisely $\tJz$. The conclusion \ref{eqn:eq4} follows.
 

### 2024-02-15 repro/papers/discreptheory/paper.tex
cleaned up

@@
 extended inversion is Variable Projection reduction of the penalty objective $\Ja$
-(equation \ref{eqn:penerr}) to a function $\tJa$ of 
+(equation \ref{eqn:jdef}) to a function $\tJa$ of 
 $m$ alone, by minimization over $w$:
@@
 program has been carried out for the simple cartoon problem of this
-paper, and partly for some other more protypical inverse wave problems \cite[]{FuSymes2017discrep}.
+paper, and partly for some other more protypical inverse wave problems \cite[]{FuSymes2017discrepancy}.
 

### 2024-02-15 repro/papers/discreptheorypub/paper.tex
more
cleaned up

@@
 \title{Error Bounds for Extended Source Inversion applied to an Acoustic Transmission Inverse Problem}
-\author[Symes]{William W. Symes$^1$\\
-$^1$ PO Box 43, Orcas, WA 98280 USA, {\tt symes@rice.edu}}
-
-\maketitle
+\author{William W. Symes\\
+  Department of Computational and Applied Mathematics, \\
+  Rice University, Houston, Texas 77251, USA; {\tt symes@rice.edu}}
+
+\lefthead{Error Bounds for Extended Source Inversion}
+\righthead{Symes}
+
 \begin{abstract}
@@
   \left(m^2\frac{\partial^2 p}{\partial t^2} - \nabla^2\right) p(\bx,t) &=&
-                                                                         w(t)\delta(\bx-\bx_s) \nonumber\\
+                                                                         w(t)\delta(\bx-\bx_s), \nonumber\\
   p(\bx,t)&=&0, t\ll 0.
@@
   \label{eqn:homsol}
-  p(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-m|\bx-\bx_s|\right),
+  p(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-m|\bx-\bx_s|\right).
 \end{equation}
@@
 \label{eqn:mod}
-F[m]w(t)  = p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right) 
+F[m]w(t)  = p(\bx_r,t) = \frac{1}{4\pi r}w\left(t-mr\right).
 \end{equation}
@@
   \label{eqn:probstat1}
-  \lerr[m,w;d] \le \frac{1}{2}\epsilon^2,
+  \lerr[m,w;d] \le \frac{1}{2}\epsilon^2.
 \end{equation}
@@
 \begin{equation}
-  [-\lambda_{\rm max}, -\lambda_{\rm max}] \subset[ t_{\rm min}-mr , 
-  t_{\rm max}-mr]. 
+  [-\lambda_{\rm max}, \lambda_{\rm max}] \subset[ t_{\rm min}-mr , 
+  t_{\rm max}+mr]. 
   \label{eqn:zeroinc}
@@
  \lerr[m,w;d] =  \frac{1}{32\pi^2
-    r^2\|d\|^2}\int\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2
+    r^2\|d\|^2}\int\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2.
 \]
@@
   |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \langle F[m]w,
-  F[m_*]w_*\rangle_D = 0
+  F[m_*]w_*\rangle_D = 0.
 \end{equation}
@@
   |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \lerr[m,w;d]=\frac{1}{32\pi^2 
-    r^2\|d\|^2}(\|w\|^2 + \|w_*\|^2).
+    r^2\|d\|^2}(\|w\|^2 + \|w_*\|^2)
 \]
@@
   \label{eqn:iso}
-  = \frac{1}{2}\left(\frac{\|w\|^2}{\|w_*\|^2} + 1 \right)
+  = \frac{1}{2}\left(\frac{\|w\|^2}{\|w_*\|^2} + 1 \right).
 \end{equation}
@@
 target slowness $m_*$, the means square error (FWI objective) $\lerr$ is independent of
-$m$, and its minimum over $w$ is attained for $w=0$
+$m$, and its minimum over $w$ is attained for $w=0$.
 \end{proof}
@@
 The main theoretical device used in the proofs of our main results on 
-extended inversion is Variable Projection reduction of the penalty objective $\Ja$
+extended inversion is reduction of the penalty objective $\Ja$
 (equation \ref{eqn:jdef}) to a function $\tJa$ of 
@@
   \label{eqn:redexp}
-  \tJa[m;d] = \inf_w \Ja[m,w;d]
-\end{equation}
+  \tJa[m;d] = \inf_w \Ja[m,w;d].
+\end{equation}
+Reduction of this type is often termed ``Variable Projection'', after \cite{GolubPereyra:73,GolubPereyra:03}.
+
 A minimizer $w$ on the right-hand side of definition
@@
           min}-mr, t_{\rm max}-mr];\\
-          0, \mbox{ else;}
+          0, \mbox{ else.}
         \end{array}
@@
 \begin{proof}
+  The idea here is that $F$ fails to be coercive just where $A$ is
+  coercive, and vis-versa.
   \begin{itemize}
@@
   e[m,\aw[m,d];d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r \alpha a(t-mr))^4(1 +
-  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)^2
+  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)^2,
 \end{equation}
@@
   p[m,\aw[m,d];d] = \frac{1}{2\|d\|^2}\int_{t_{\rm min}}^{t_{\rm max}} \,dt\,(4\pi r a(t-mr))^2(1 +
-  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)^2
-\end{equation}
-
+  (4\pi r \alpha a(t-mr))^2)^{-2}d(t)^2,
+\end{equation}
+and
 \begin{equation}
@@
 
-  That the VPM objective $\tJa$ is given by \ref{eqn:expjgen} follows from equations \ref{eqn:redexp1}, \ref{eqn:residnormgen}, and
+  That the reduced objective $\tJa$ is given by \ref{eqn:expjgen} follows from equations \ref{eqn:redexp1}, \ref{eqn:residnormgen}, and
   \ref{eqn:anninormgen}.
@@
 
-It will be useful to record expressions for the various componenets of
+It will be useful to record expressions for the various components of
 $\tJa$ when the data is noise-free, that is, the context of
@@
 
+\section{Stationary Points}
+
 Recall that the purpose of the penalty operator is to penalize energy away from 
@@
 \cite[]{Plessix:00a,LuoSava:11,Warner:14,HuangSymes:SEG15a,Warner:16,HuangSymes:Geo17}.
-
-\section{Stationary Points}
 
@@
     \frac{r(4 \pi r \alpha)^2(\lambda-\mu)}{(1+(4\pi r\alpha)^2 
-      (\lambda+\mu)^2)^{2}}  
+      (\lambda+\mu)^2)^{2}}.
     \label{eqn:gradbndnonoise}
@@
     = -\frac{r(4 \pi r \alpha)^2}{\|w_*\|^2}\int_{-(m-m_*)r-\lambda}^{-(m-m_*)r+\lambda}
-    \,dt\, t(1+(4\pi r)^2\alpha^2 t^2)^{-2}w_*\left(t+(m-m_*)r\right)^2
+    \,dt\, t(1+(4\pi r)^2\alpha^2 t^2)^{-2}w_*\left(t+(m-m_*)r\right)^2.
   \]
@@
   \label{eqn:elimit}
-  \lim_{\alpha \rightarrow \infty}e[m,\aw[m;d];d] =
-  \frac{1}{2}\frac{1}{(4 \pi r)^2}\|w_*\|^2 = \frac{1}{2}\|d\|^2.
+  \lim_{\alpha \rightarrow \infty}e[m,\aw[m;d];d] = \frac{1}{2}.
 \end{equation}
@@
 \[
-  e[m,\aw[m;d];d]  \le \frac{1}{2}\epsilon^2\|d\|^2
+  e[m,\aw[m;d];d]  \le \frac{1}{2}\epsilon^2
   \Rightarrow \alpha \le \alpha_{\rm max}(\epsilon,\lambda,d).
@@
 D$. Since no support assumptions can be made about $n$, equation
-\ref{eqn:normsol} implies that $\aw[m;d] \notin \lW$ for any values of
-$\alpha$ and $\lambda$.  Therefore minimization of $\tJa$ cannot by itself yield a
+\ref{eqn:normsol} does not imply that $\aw[m;d] \in \lW$ for any values of
+$\alpha$ or $\lambda$.  Therefore minimization of $\tJa$ cannot by itself yield a
 solution of the inverse problem as defined in the problem statement
-\ref{eqn:probstat0}. In this section, I explain how a solution may
-nonetheless be constructed from a stationary point of $\tJa$.
-
-First, examine the effect of additive noise on the estimation of the
-slowness $m$. In expressing the result, use the dimensionless
-relative data error
+\ref{eqn:probstat0}. %\textcolor{red}{In this section, I explain how a solution may
+%nonetheless be constructed from a stationary point of $\tJa$.
+%First, examine the effect of additive noise on the estimation of the
+%slowness $m$. In expressing the result, use the dimensionless
+%relative data error}
+
+%\textcolor{blue}
+{To see how a solution may
+  nonetheless be constructed from a stationary point of $\tJa$,
+  examine the effect of additive noise on the estimation of the
+slowness $m$, in terms of the dimensionless
+relative data error}
 \begin{equation}
@@
     \eta(1+\eta) \le \frac{16}{3\sqrt{3}}\frac{4\pi r \alpha
-      (\lambda-\mu)}{(1+(4\pi r\alpha(\lambda+\mu))^2)^2}
+      (\lambda-\mu)}{(1+(4\pi r\alpha(\lambda+\mu))^2)^2}.
   \end{equation}
@@
     \label{eqn:mnoisebndfin}
-    |m-m_*| \le \frac{\lambda}{r}
+    |m-m_*| \le \frac{\lambda}{r}.
   \end{equation}.
@@
   From equation \ref{eqn:dexpjgen}, $d \tJa / dm$ is the
-value of a quadratic form in $d$ with (indefinite) symmetric operator
+  value of a quadratic form in $d$
+%  \textcolor{red}{with}
+%  \textcolor{blue}
+  {defined by the}
+  (indefinite) symmetric operator
 $B = $ multiplication by
 \[
-  b(t;m,\alpha)  = -\frac{(4 \pi r \alpha)^2 (t-mr)}{(1+(4\pi r \alpha (t-mr))^2)^{2}}
+  b(t;m,\alpha)  = -\frac{(4 \pi r \alpha)^2 (t-mr)}{(1+(4\pi r \alpha (t-mr))^2)^{2}}.
 \]
@@
   \left|\frac{d}{dm}\tJa[m;d_*]\right| \le \frac{3\sqrt{3}}{16} 4\pi
-  r\alpha \eta(1+\eta)\|d_*\|^2
+  r\alpha \eta(1+\eta)\|d_*\|^2.
 \]
@@
   \alpha)^2 \frac{\lambda-\mu}{(1+(4\pi r)^2\alpha^2
-    (\lambda+\mu)^2)^{2}} \|d_*\|^2
+    (\lambda+\mu)^2)^{2}} \|d_*\|^2,
 \]
@@
     |m-m_*| \le \frac{\mu}{r} + \frac{\eta}{\alpha} \left(\frac{3\sqrt{3}(1+\eta)}{64\pi r^2}(1+(8\pi r \alpha
-      \lambda_{\rm max})^2)^2\right)
+      \lambda_{\rm max})^2)^2\right).
   \end{equation}
@@
 
-\begin{proof} of Theorem \ref{thm:mnoiseres}:
+\begin{proof}.
   Write $\lambda = (1+\delta)\mu$, and $x=4 \pi r \alpha \mu$. Then
@@
   \[
-    1+C^2x^2 = \frac{4}{3}
+    1+C^2x^2 = \frac{4}{3},
   \]
@@
 the limit as $\eta \rightarrow 0$, conclusion 2 becomes
-$\lambda/\mu \gtrsim 1 + 2\eta$, that is, the ``fudge factor'' beyond
+$\lambda/\mu \ge 1 + 2\eta$, that is, the ``fudge factor'' beyond
 the noise-free bound is approximatly twice the noise level.
@@
 
-\begin{proof} of Theorem \ref{thm:ipnoisesuf}:
+\begin{proof}:
 From Theorem \ref{thm:norminv}, 
@@
 \[
-  = \aw[m;d_*] + (1+ (4 \pi r \alpha t)^2)^{-1}4\pi r n(t+mr)
-\]
-
+  = \aw[m;d_*] + (1+ (4 \pi r \alpha t)^2)^{-1}4\pi r n(t+mr).
+\]
 From Theorem \ref{thm:mnoiseres},
 \[
-  |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r} 
+  |m-m_*| \le \left(1+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r},
 \]
@@
   =\left(2+\frac{2\eta(1+\eta)}{1-\eta(1+\eta)}\right)\frac{\mu}{r}
-  -\frac{\mu}{r} \le \left(\frac{\lambda}{\mu}-1\right)\frac{\mu}{r}
+  -\frac{\mu}{r} \le \left(\frac{\lambda}{\mu}-1\right)\frac{\mu}{r}.
 \]
@@
   \mbox{supp }\aw[m;d_*] \subset [-\mu-(m-m_*)r, \mu-(m-m_*)r]
-  \subset [-\lambda,\lambda]
+  \subset [-\lambda,\lambda],
 \]
@@
 \[
-  E_{\lambda}\aw[m;d](t) = \aw[m;d_*](t) + E_{\lambda}(1+ (4 \pi r \alpha
-  t)^2)^{-1}4\pi r n(t+mr)
+  {\bf 1}_{[-\lambda,\lambda]}\aw[m;d](t) = \aw[m;d_*](t) + {\bf 1}_{[-\lambda,\lambda]}(1+ (4 \pi r \alpha
+  t)^2)^{-1}4\pi r n(t+mr).
 \]
@@
 \[
-  F[m]{\bf 1}_{[t_1.t_2]}w = {\bf 1}_{[t_1+mr,t_2+mr]}F[m]w  
+  F[m]{\bf 1}_{[t_1.t_2]}w = {\bf 1}_{[t_1+mr,t_2+mr]}F[m]w.
 \]
@@
 \[
-  F[m]E_{\lambda}\aw[m,d](t) -d(t) = F[m]\aw[m,d_*](t) -d_*(t)  
+  F[m]{\bf 1}_{[-\lambda,\lambda]}\aw[m,d](t) -d(t) = F[m]\aw[m,d_*](t) -d_*(t)  
 \]
@@
 \[
-  -(1- {\bf 1}_{[-\lambda+mr,\lambda+mr]})n(t)
+  -(1- {\bf 1}_{[-\lambda+mr,\lambda+mr]})n(t).
 \]
@@
 \[
-  \| F[m]E_{\lambda}\aw[m,d_*] -d_*\|^2 = (4 \pi r \alpha)^4
+  \| F[m]{\bf 1}_{[-\lambda,\lambda]}\aw[m,d_*] -d_*\|^2 = (4 \pi r \alpha)^4
   \int_{-\lambda+(m-m_*)r}^{\lambda+(m-m_*)r}\,dt\, (t-(m-m_*)r)^4
@@
 \[
-  \le (4 \pi r \alpha \lambda)^4\|d_*\|^2
+  \le (4 \pi r \alpha \lambda)^4\|d_*\|^2.
 \]
@@
   \le (4 \pi r \alpha \lambda)^2 \|{\bf 1}_{[-\lambda+mr,\lambda+mr]}
-  n\|^2 + \|(1 - {\bf 1}_{[-\lambda+mr,\lambda+mr]}n\|^2
+  n\|^2 + \|(1 - {\bf 1}_{[-\lambda+mr,\lambda+mr]}n\|^2.
 \]
@@
 \[
-  \le \max \{(4 \pi r \alpha \lambda)^2, 1\} \|n\|^2
+  \le \max \{(4 \pi r \alpha \lambda)^2, 1\} \|n\|^2.
 \]
@@
 \[
-  \|F[m]E_{\lambda}\aw[m,d]-d\| \le (4 \pi r \alpha
+  \|F[m]{\bf 1}_{[-\lambda,\lambda]}\aw[m,d]-d\| \le (4 \pi r \alpha
   \lambda)^2\|d_*\| + \max \{4\pi r \alpha \lambda, 1\}\|n\|
@@
 \bibliographystyle{seg}
-\bibliography{../../bib/masterref}
+\bibliography{../../bib/masterref}<
+
+%\input{paper.bbl}
 
@@
   Discrepancy Principle, is available - see
-  \cite{FuSymes2017discrepancy,Symes:21a,SymesChenMinkoff:21}. This
+  \cite{FuSymes2017discrepancy,SymesChenMinkoff:21}. This
   scaling varies dynamically during iterative optimization, in effect
@@
 \[
-  = \frac{1}{2}(\|F[m]\aw[m;d]-d\|^2 + \alpha^2\|Aw[m;d]\|^2)
+  = \frac{1}{2}(\|F[m]\aw[m;d]-d\|^2 + \alpha^2\|A\aw[m;d]\|^2)
 \]
@@
 
-Since $\tJa$ and its gradient depend smoothly on their arguments, it
-is possible to derive an alternate expression assuming that $d \in
-H^1(\bR)$, then extend it by continuity to $d \in L^2(\bR)$. Note that
-if $w \in H^1(\bR)$, then
-  differentiable, and 
+Theorem \ref{thm:diffobj} also gives an explicit expression for the
+derivative of $\tJa$. An alternative expression
+does not depend on the structure of the single-trace modeling
+operator, hence applies in a much more general setting. Note that if
+$w \in H^1(\bR)$, then $m \mapsto F[m]w$ is differentiable, and
 \begin{equation}
@@
 \end{equation}
-
 That is, $DF[m]\delta m$ factors into $F[m]$ following $Q[m]\delta m$,
@@
 Assume that $d \in H^1(\bR)$. From equation \ref{eqn:normsol}, it
-follows that $w[m;d] \in H^1(\bR)$ and moreover that $m \mapsto
+follows that $\aw[m;d] \in H^1(\bR)$ and moreover that $m \mapsto
 \aw[m;d]$ is differentiable as a map from $\bR^+$ to
@@
 \[
-  \tJa[m;d] = \frac{1}{2}(\|F[m]w[m;d]-d\|^2 + \alpha^2 \|Aw[m;d]\|),
+  \tJa[m;d] = \frac{1}{2}(\|F[m]\aw[m;d]-d\|^2 + \alpha^2 \|A\aw[m;d]\|),
 \]
@@
   = \alpha^2 \langle \aw[m;d],[ (Q[m]\delta m), A^TA]\aw[m;d] \rangle
-  + \alpha^2 \langle (Q[m]\delta m)\aw[m;d],A^TA\aw[m;d] \rangle
+  + \alpha^2 \langle (Q[m]\delta m)\aw[m;d],A^TA\aw[m;d] \rangle.
 \]
@@
   \label{eqn:djq}
-  d\tJa[m;d]\delta m = \frac{1}{2}\alpha^2 \langle \aw[m;d],[
+  D\tJa[m;d]\delta m = \frac{1}{2}\alpha^2 \langle \aw[m;d],[
   (Q[m]\delta m), A^TA]\aw[m;d] \rangle.

### 2024-02-15 repro/papers/iwavecal/paper.tex
cleaned up

@@
 \date{}
-\author{William W. Symes, The Rice Inversion Project}
+\author{William W. Symes}
+
+\inputdir{project}
 
@@
 
+\section{Introduction}
+This note concerns verification of the IWAVE family of finite
+difference wave solvers \cite[]{GeoPros:11}. In the numerics QC vernacular, verification
+means ``show that the equations are solved correctly'', as opposed to
+validation, ``show that the correct equations are solved''. In the
+context of discrete methods for partial differential equations,
+``correctly'' means: with an adequate rate of convergence as grids or
+meshes are refined.
+
+Several types of verification tests exist, none of
+them entirely satisfactory (see for example \cite{FehlerKeliher:11},
+Chapter 3). The approach used here is comparison with
+analytic (or nearly analytic) closed-form solutions. Such solutions
+exist only in special cases, for example, wave equations with constant
+coefficients, the case considererd here. While verification in such
+special cases does not necessarily predict the errors encountered
+under fewer constraints (for example, heterogenous media), it does
+indicate whether overall scaling of the numerical solution is correct.
+
+The example reported below, and similar examples not explicitly shown,
+verify the correct solution of homogeneous medium acoustics problems
+with isotropic point sources using IWAVE finite difference
+simulators. This task requires first that the
+analytic solution for a homogeneous medium be derived. While such
+solutions are found in many publications, most instances take
+advantage of restrictions on the problem parameters to simplify the
+results. Since the point of this exercise is to verify correct
+dependence on parameters, I begin this note with a complete derivation
+of the homogeneous medium point source solution (the ``radiation
+problem'' of \cite{CourHil:62}, Chapter 6), with dependence on
+parameters explicitly described. The solution for the 3D version of
+this problem is actually analytic. Since 2D finite difference
+simulation is much less expensive than 3D, that case is also
+important, but the result involves quadrature, so is not really
+analytic. I pose a 2D homogeneous medium point source problem, and
+solve it using IWAVE's staggered grid finite difference simulator, and
+a trapezoidal rule realization of the 2D quasi-analytic
+solution. Comparison shows convincingly that the finite difference
+simulator is functioning correctly for this class of problems.
+
 \section{Theory}
@@
 \end{eqnarray}
-(\cite[]{gur81}, section 19, or \cite[]{Frie:58}, Chapter 1).
+(\cite{gur81}, section 19, or \cite{Frie:58}, Chapter 1).
 The linear acoustic constitutive law makes relative change of volume and pressure proportional by the the bulk modulus $\kappa$, which therefore has units of pressure = force/area.
@@
 
-In this section we will develop analytic expressions for the pressure field $p$ in \ref{eqn:aws} or \ref{eqn:awe} for 2D and 3D, assuming that $\kappa$ and $\beta$ are constant and $f$ has the form of an {\em isotropic point radiator} located at the origin:
+In this section I will develop analytic expressions for the pressure field $p$ in \ref{eqn:aws} or \ref{eqn:awe} for 2D and 3D, assuming that $\kappa$ and $\beta$ are constant and $f$ has the form of an {\em isotropic point radiator} located at the origin:
 \begin{equation}
@@
 \subsection{3D}
-The solution of the point radiator problem is well-known, see for example \cite{CouHil:62}, Chapter 6 or \cite{GuiSte:79}, Chapter 1.
+The solution of the point radiator problem is well-known, see for example \cite{CourHil:62}, Chapter 6 or \cite{GuiSte:79}, Chapter 1.
 I repeat it here for completeness.
@@
 \subsection{2D}
-I will regard 2D solutions of equations \ref{eqn:aws} and \ref{eqn:awe} as line source solutions of the 3D equations.
-In this subsection, I use the notation $\bx=(x,z)$ for the 2D position vector in the ``inline'' plane perpindicular to the $y$ axis, and $\bx_{3D} = (x,y,z)$.
-The isotropic line radiator appropriate for the ASG version of the acoustic system \ref{eqn:asg} is
+I will regard 2D solutions of equations \ref{eqn:aws} and
+\ref{eqn:awe} as line source solutions of the 3D equations. In this
+subsection, I use the notation $\bx=(x,z)$ for the 2D position vector
+in the ``inline'' plane perpindicular to the $y$ axis, and $\bx_{3D} =
+(x,y,z)$. The isotropic line radiator appropriate for the ASG version
+of the acoustic system
+\ref{eqn:asg} is 
 \begin{equation}
@@
 
-To find an analytic form for the pressure field of an isotropic line radiator, use D'Alembert's method of descent (\cite[]{CouHil:62}, Ch. 6). This approach begins with the observation that
+To find an analytic form for the pressure field of an isotropic line radiator, use D'Alembert's method of descent (\cite{CourHil:62}, Ch. 6). This approach begins with the observation that
 \[
@@
 \]
-so taking the limit $\epsilon \rightarrow 0$ and introducing the Heavyside function to express the integral as a convolution, obtain
+so taking the limit $\epsilon \rightarrow 0$ and introducing the
+Heavyside function to express the integral as a convolution, obtain
+\[
+p_{ASG2D}(\bx,t) = \frac{1}{2\pi c}\int d\tau
+\frac{dw_{ASG2D}}{dt}(t-\tau)\frac{H(c\tau-|\bx|)}{\sqrt{c^2 \tau^2
+    -|\bx|^2}}
+\]
 \begin{equation}
 \label{eqn:asgsol2d}
-p_{ASG2D}(\bx,t) = \frac{1}{2\pi c}\int d\tau \frac{dw_{ASG2D}}{dt}(t-\tau)\frac{H(c\tau-|\bx|)}{\sqrt{c^2 \tau^2 -|\bx|^2}}
+= \frac{d}{dt}(w * g(\cdot,r))
+\end{equation}
+where
+\begin{equation}
+  \label{eqn:green2dsing}
+  g(t,r) = \frac{1}{2\pi c^2}\frac{H(t-r/c)}{\sqrt{t^2 -r^2/c^2}}
 \end{equation}
 
-The integral \ref{eqn:asgsol2d} is singular, and inconvenient for computation.
-The singular factor in the denominator is $\sigma = \sqrt{\tau -|\bx|/c}$, which if used as an alternate integration variable desingularizes the integral.
-Note that
-$\tau = \sigma^2 + |\bx|/c$, $d\tau = 2 \sigma d\sigma$, that the range of $\sigma$ is $[0,\sqrt{t-|\bx|/c}]$, and that
+The convolution kernel \ref{eqn:green2dsing} is singular, so simple
+regular grid 
+discretizations
+of convolution (for instance trapezoidal rule) have sub-optimal accuracy. 
+To produce an equivalent expression involving less singular
+integrands, note that
 \[
-\sqrt{c^2\tau^2-|\bx|^2} = c\sigma \sqrt{\sigma^2 + 2|\bx|/c}.
+  g(t,r) = \frac{1}{2\pi c^2}
+  \left(\frac{H(t-r/c)}{\sqrt{t+r/c}}2\frac{d}{dt}\sqrt(t-r/c)\right)
 \]
-So
-\begin{equation}\label{eqn:asgsol2dalt}
-p_{ASG2D}(\bx,t)=\frac{1}{\pi c^2} \int_0^{\sqrt{t-|\bx|/c}}d\sigma \frac{\frac{dw_{ASG2D}}{dt}(t-\sigma^2-|\bx|/c)}{\sqrt{\sigma^2 + 2|\bx|/c}}. 
+\[
+  =\frac{1}{\pi c^2}\left(\frac{d}{dt} H(t-r/c)
+    \sqrt{\frac{t-r/c}{t+r/c}}
+ - H(t-r/c)\sqrt{t-r/c}
+    \frac{d}{dt}\frac{H(t-r/c)}{\sqrt{t+r/c}}\right)
+\]
+Since $H(t-r/c)\sqrt{t-r/c}$ is continuous and vanishes at $t=r/c$, 
+\[
+  = \frac{d}{dt} \left(\frac{1}{\pi c^2}H(t-r/c)
+    \frac{(t-r/c)^{1/2}}{(t+r/c)^{1/2}} \right)+
+  \left(\frac{1}{2\pi c^2} H(t-r/c)
+    \frac{(t-r/c)^{1/2}}{(t+r/c)^{3/2}} \right)
+\]
+Therefore, the relation \ref{eqn:asgsol2d} is equivalent to
+\begin{equation}
+  \label{eqn:asg2dsolbis}
+  p_{ASG2D}(\bx,t) = \frac{d^2}{dt^2}(g_1(\cdot, r) *w)(t) + \frac{d}{dt}(g_2(\cdot,r)*w)(t),
 \end{equation}
-Discrete evaluation for this expression requires interpolation of $f$, but the integral is proper so ordinary numerical quadrature works well.
+in which
+\begin{eqnarray}
+  \label{eqn:green2dt1}
+  g_1(t,r) &=& \frac{1}{\pi c^2}H(t-r/c)
+               \frac{(t-r/c)^{1/2}}{(t+r/c)^{1/2}},\\
+  \label{eqn:green2dt2}
+  g_2(t,r) &=& \frac{1}{2\pi c^2} H(t-r/c)
+               \frac{(t-r/c)^{1/2}}{(t+r/c)^{3/2}}.
+\end{eqnarray}
 
-From the relation \ref{eqn:asgacd}, 
-\begin{equation}\label{eqn:acdsol2dalt}
-p_{ACD2D}(\bx,t)=\frac{1}{\pi c^2} \int_0^{\sqrt{t-|\bx|/c}}d\sigma \frac{w_{ACD2D}(t-\sigma^2-|\bx|/c)}{\sqrt{\sigma^2 + 2|\bx|/c}}. 
-\end{equation}
-Here $p_{ACD2D}$ solves the 2D version of the system \ref{eqn:acd} with line source pulse $w_{ACD2D}$.
+\section{Example}
+In this section, I present a 2D example showing that pressure traces
+computed via convolution with Green;'s functions, as developed in the
+last section, closely approximate
+those computed via staggered
+grid finite difference approximation of the system \ref{eqn:asg}. I
+present only the 2D case, to avoid computational expense. This example
+verifies that calibration of the finite difference implementation.
 
-It is easily checked that the right-hand sides of equations \ref{eqn:asgsol2d}, \ref{eqn:asgsol2dalt}, and \ref{eqn:acdsol2dalt} have units of pressure.
+Numerical evaluation of the right-hand side in equation
+\ref{eqn:asg2dsolbis} requires a choice of quadrature rule. The
+Green's function components $g_1$ and $g_2$ are continuous but have a
+$t^{1/2}$ singularity at the wavefront $t=r/c$. Sampling on a uniform
+grid effectively zeros $g_1$ and $g_2$ in the sample interval next
+to the wavefront, resulting in a time shift of a fraction of the time
+step. The trapezoidal rule
+applied to such integrands is therefore only first-order accurate. A
+specialized quadrature rule adapted to this singularity would yield
+more accurate results. Alternatively, further
+integration-by-parts manipulations like those that led to the
+expression \ref{eqn:asg2dsolbis} would shift more derivatives onto the
+wavelet, thus restoring second-order accuracy of the trapezoidal rule
+approximate convolution. I will not explore these options, since the
+main aim (calibration of the finite difference code) is achieved
+without them.
+
+The finite difference scheme used to approximate the solution of the
+system \ref{eqn:asg}
+is the staggered grid scheme
+introduced by \cite{GauTarVir:86}, based on leapfrog centered
+difference formulae. Any even formal order of accuracy in space and
+time can be achieved by higher-order leapfrog formulae in space and
+Lax-Wendroff refinement in time \cite[]{moczoetal:06}. The scheme used
+here is 2nd order accurate in time, 4th order in space. I will show
+computational results from three levels of grid refinement, by a
+factor of 2 in both space and time, so the formal order of accuracy is
+2. The use of a higher order of accuracy in the space derivatives
+leads to less grid dispersion, that is, a more accurate result at each
+level of grid refinement.
+
+IWAVE source input assumes that right-hand sides in the wave equation
+are represented by time traces, and that each time trace is the
+coefficient of a spatial delta function. Spatial delta functions are
+discretized by adjoint interpolation, mapping amplitudes at
+(arbitrary) sample points
+to nearby grid points, and scaling by the inverse of the spatial grid
+cell volume. This adjoint sampling and scaling is built into
+IWAVE. Only the time trace - in other words, the wavelet in the
+context of the present problem - need be passed to the simulator, in
+the form of a SEGY trace file.
+
+In this example, $\kappa = $ 4 GPa and $\rho = 1$ g/cm$^3$, so that $c = 2$
+m/ms. A point source is placed at at $x_s=$ 3500 m, $z_s=$ 3000 m with
+respect to an arbitrarly chosen origin, and a receiver at $x_r=$ 3500
+m, $z_r=$ 1000 m, so the source-receiver distance is $r=$
+2061.552... Note that the the wavefront position will not lie on any
+time grid point, which are integral multiples of a
+millisecond, for the sampling used below.
+The source wavelet is a trapezoidal bandpass filter
+with corner frequencies at 1.0, 2.5, 7.5 and 12 Hz, depicted in Figure
+\ref{fig:ptsrc2}.
+
+\plot{ptsrc2}{width=0.8\textwidth}{Zero-phase bandpass filter  with
+  corner frequencies 1, 2.5, 7.5, 12 Hz, delayed by 1 s.}
+
+The computations use three time sample rates for source wavelet and
+data traces, with $\Delta t = 0.008, 0.004, $ and
+$0.002$ s. The finite difference computations use corresponding square spatial grids
+with $\Delta x = \Delta z = $ 20, 10, and 5 m. The finite difference 
+time step is computed internally, and is smaller than the time sample
+rates. Cubic spline interpolation and adjoint interpolation
+transfers data between internal and sample time grids.
+
+Figures \ref{fig:plot0}, \ref{fig:plot1}, and \ref{fig:plot2} show
+the finite difference and convolutional traces computed with these choices for time steps $\Delta t = $
+0.008, 0.004, and 0.002 s respectively, along with the difference
+traces. The first order error is visually obvious, but so is
+convergence. The $L^2$ errors (relative to the finite difference
+traces) are 29\%, 14.8\%, and 7.6\% respectively. Almost all of this
+error is the effective time shift error in the trapezoidal rule evaluation of
+the integrations in equation \ref{eqn:asg2dsolbis}: the relative $L^2$
+error between the finite difference traces for coarsest and finest
+grids is 2.7\%.
+
+\plot{plot0}{width=0.8\textwidth}{Finite difference trace (blue)
+  vs. convolution with Green's function (equation
+  \ref{eqn:asg2dsolbis}) (red) and difference trace (black), coarsest grid
+  ($\Delta t$ = 0.008 s).}
+
+\plot{plot1}{width=0.8\textwidth}{Finite difference trace (blue)
+  vs. convolution with Green's function (equation
+  \ref{eqn:asg2dsolbis}) (red) and difference trace (black), medium grid
+  ($\Delta t$ = 0.004 s).}
+
+\plot{plot2}{width=0.8\textwidth}{Finite difference trace (blue)
+  vs. convolution with Green's function (equation
+  \ref{eqn:asg2dsolbis}) (red) and difference trace (black), finest grid
+  ($\Delta t$ = 0.002 s).}
+
+\section{Conclusion}
+The calculations presented here accomplished two objectives. First,
+the Green's functions constructed here are correctly scaled for linear
+acoustics in homogeneous fluids with arbitrary bulk modulus and
+density. While many references give expressions for these Green's
+functions, most choose a particular reference scale (frequently
+$\kappa = \rho = 1$) hence do not explicitly give results needed for
+comparison to numerical solutions. Second, the problem is formulated
+consistently with the IWAVE finite difference codes for both the
+second-order wave equation and the first-order pressure-velocity
+system. With these tools, verification of correct calibration for the
+finite difference codes is straightforward. The example presented in
+the last section verifies correct calibration of the staggered grid
+scheme for the pressure-velocity system in 2D. Similar verifications
+are possible for the other cases covered by the theory presented above.
+
+\bibliographystyle{seg}
+\bibliography{../../bib/masterref}

### 2024-02-15 repro/papers/ms/paper.tex
cleaned up

@@
 \label{eqn:depths} 
-Z_s & = & Z_s\} \nonumber \\
-Z_r & = & Z_r\} 
+Z_s & = & [z_s^{\rm min}, z_s^{\rm max}] \nonumber \\
+Z_r & = & [z_r^{\rm min},z_r^{\rm max}] 
 \end{eqnarray}
@@
 For physical reasons, $m$ should be positive and bounded uniformly above and 
-below, say by $0 < m_{\rm min} < m_{\rm max}$. We assume that the
+below, say by $0 < m_{\rm min} < m_{\rm max}$.
+
+We assume that the
 material is {\em acoustically transparent}, that is, generates no
@@
 optics approximation, to be summarized below, shows that a convenient
-mathematical proxy for transparency is the assumption that $m$is
+mathematical proxy for transparency is the assumption that $m$ is
 smooth, with derivatives up to
@@
 
-\section{A Synthetic Example: FWI}
-A simple 2D synthetic example illustrates the impact of Theorem
-\ref{thm:FWIconv}. We will also use this example throughout the paper
-to illustrate alternative solution methods for the crosswell inverse
-problem.
-
-The target slowness $m_*$ is the reciprocal of a piecewise
-linear velocity (Figure
-\ref{fig:velbig}). This corresponding velocity has a constant $2$ km/s layer near $x=0$ and
-increases linearly for $x\ge 0.5$ km. Boundary conditions are Dirichlet
-for pressure on all parts of the boundary, although the wavefields
-occuring in the experiments described here interact only with the $x=0$
-surface so that the effective geometry is that of a half-space. Data
-for a single source at depth $x_s=$ 10 m, $z_s=$ 5100 m (coordinates as
-depicted in Figure \ref{fig:velbig}. A line of receivers is positioned
-at depth $x_r=$ 3000 m, with 4000 m $\le z_r \le$ 7000 m, spaced
-$\Delta z_r$=30 m apart.
-
-The source pulse ($f$ in equation (\ref{eqn:pressure})) is a [2.5, 5, 25, 30]
-Hz trapezoidal bandpass filter produces the synthetic data
-depicted in Figure \ref{fig:shot6kmbig}. Note that the finite
-difference scheme used to produce this data approximates 2D wave
-propagation rather than the 3D propagation mentioned in the discussion
-so far, so that the data waveform is asymmetric, unlike the
-source. Also the presence of the pressure-free surface near the source
-contributes effectively another derivative. Nonetheless the data
-resemble the prediction of the constant velocity calculation (equation
-(\ref{eqn:cv})).
-
-For a very smooth $m$ such as that depicted in Figure
-\ref{fig:velbig}, the curve $z_r \mapsto \tau[m](z_s,z_r)$ resembles a hyperbola, as one sees from Figure
-\ref{fig:shot6kmbig}. The onset time of the signal in this figure
-depends on the slowness used in the simulation. For
-$m_0 \equiv 0.5$ s/km, the predicted data $S[m_0]f$ are depicted in
-Figure \ref{fig:shot6kmhom}. The general appearance is similar to   Figure
-\ref{fig:shot6kmbig}, but the onset time of the wave is later, due to
-the slower wave velocity.
-
-Now consider the inverse problem (\ref{eqn:ls}) with $d=S[m_*]f$. Since
-$m_*=m_0$ near the source point, it is not unreasonable to take $m_0$
-as an initial guess for the wave velocity. Conventional FWI uses
-iterative optimization algorithm to minimize $J_{\rm LS}$. Descent
-algorithms related to Newton's method have been favored for FWI, due
-to the computational cost of function evaluation
-\cite[]{VirieuxOperto:09}. Of course, any such algorithm finds a
-stationary point of the objective, assuming convergence, and not
-necessarily a global minimizer.
-
-A plot (Figure \ref{fig:fwicurve}) 
-of $J_{\rm LS}$ along the line
-segment from $m_0$ to $m_*$ immediately shows the difficulty faced by
-a local optimization approach: as one
-would predict from Theorem \ref{thm:FWIconv} and its proof, the predicted data change
-little in size along this segment and have support nearly disjoint
-from that of the target data $d$, hence are orthogonal to it, until the trial slowness becomes
-quite close to the target $m_*$, so that the local behaviour of $J_{\rm LS}$ in a small
-neighborhood of $m_0$ gives little or
-no hint about a constructive update direction. One would not expect a
-local optimization algorithm to be effective under such circumstances.
-
-Most descent methods require computed gradient, as input to the
-computation of a descent step. Assuming that $f \in H^1_0(\bR)$, the gradient of $J_{\rm LS}$ is 
-\be\label{eqn:gradls}
-\nabla J_{\rm LS}[m] = D(S[m]f)^T(w^2(S[m]f-d)).
-\ee
-In the expression \ref{eqn:gradls}, note that $m \rightarrow S[m]f$ is
-a map from $M$ to $L^2(\Omega)$, the derivative of which is given in
-equation  \ref{eqn:preq}. The symbol $D(S[m]f)^T$ denotes the
-transpose of this derivative with respect to a Hilbert space norm
-chosen for $M$, about which we have said nothing up to this point. We
-will mention some sensible possibilities in the Discussion section;
-for now, we simply identify $M$ with $L^2(\bR^2)$ for the purpose of
-computing the transpose, hence the gradient of $J_{\rm LS}$. Thus the
-gradient as computed here is the image of the formal or $L^2$ adjoint
-of $D(S[m]f)$, applied
-to the time-cutoff residual $w^2(S[m]f-d)$. 
-%There is no reason to suppose that this
-%image would be smooth, given the weak constraints on $f$ and $d$, so
-%in general it will not define an acceptable update direction for
-%velocity. One way out of this predicament is to extend the domain of
-%$S$ to a subset of a larger Hilbert space containing
-%$C^{\infty}(\bR^2)$, defined by control of finitely many derivatives,
-%such as a Sobolev space, and define the adjoint with respect to this
-%inner product structure, obtained by concatenating a smoothing
-%operator with the formal adjoint. The reader can consult for instance
-%\cite{KerSy:94} for more on this idea. For the purposes of this
-%discussion we will display the formal gradient (\ref{eqn:gradls})..
-
-Figure \ref{fig:fwigrad6kmbighom} shows the result of the computation
-(\ref{eqn:gradls}) for the example given previously, $m=m_0$, $d=S[m_*]f$. The adjoint $D(S[m]f)^T$ is
-computed by the adjoint state method
-\cite[]{Chavent:74,GauTarVir:86,Plessix:06}. Geometric optics explains
-the apparent interrogation of only the triangle spanning the source
-and the receiver segment - perturbations of velocity outside of it do
-not affect the geometric optics travel times (rays are straight lines,
-in constant velocity) connecting source and receivers. However there
-is very little indication of a productive update: the edge effects one
-sees in this figure (``rabbit ears'') average to very nearly zero when
-weighted by the difference between $m_0$ and $m_*$ - that is, the
-gradient is essentially orthogonal to the best search direction.
-
-If we use the gradient as a search direction, as one would for the
-steepest descent algorithm, for instance, then one obtains the values
-of $J_{\rm LS}$ displayed in Figure \ref{fig:fwigrad} - clearly not a
-productive search direction.
-
 \section{Source-Receiver Extension}
@@
 %objective). Convex combinations of homogeneous (2 km/s) and lens models.}
-\plot{velbig_dec_exp}{width=\textwidth}{Example: $v$ = 2 km/s for $z \le
-  0.5 km$, $=2.0 + (z-0.5)*0.5$ km/s for $z>0.5$ km. Source location
-  $x=5100$, $z=10$ m and receiver locations $x=3000 - 6000$ m,
-$z=3000$ m indicated.}
-\plot{shot6kmhom}{width=\textwidth}{Shot gather for geometry indicated
-  in Figure \ref{fig:velbig_dec_exp}, generated with homogeneous velocity:
-  $v_0=2$ km/s}
-\plot{shot6kmbig}{width=\textwidth}{Shot gather for geometry and
-  $v$ as in Figure \ref{fig:velbig_dec_exp}.}
-\plot{shot6kmbig_decon}{width=\textwidth}{Inverse image of data in 
-  Figure (\ref{shot6kmbig}) under extended modeling operator at $v_0$
-  = 2 km/s}
-\plot{shot6kmbig_bigdecon}{width=\textwidth}{Inverse image of data in 
-  Figure \ref{shot6kmbig} under extended modeling operator at target
-  $v$ (Figure \ref{fig:velbig_dec_exp}).}
-\plot{fwigrad6kmbighom}{width=\textwidth}{FWI gradient for shot gather
-defined in Figure \ref{fig:velbig_dec_exp}, at $v=v_0$ = 2 km/s. Scaled by
-10$^6$ for plotting. Note very
-small values, and sign is positive, indicating that FWI objective {\em
-increases} in the direction of the target velocity.}
-\plot{fwiscan}{width=\textwidth}{Plot of values of the FWI objective
-  along the line segment between $v_0$ and $v$, for the shot gather
-  described in Figure \ref{fig:velbig_dec_exp}. For abscissa $h$, value
-  plotted is FWI objective at $h v_0 + (1-h) v$. Note that objective
-  does in fact initially increase from its value at $v_0$, consistent
-  with the positive average value of the gradient as shown in Figure \ref{fig:fwigrad6kmbighom}. }
-\plot{ivagrad6kmbighom}{width=\textwidth}{Source-receiver Extended FWI
-  gradient, same geometry, data, and velocity model as in Figure
-  \ref{fig:fwigrad6kmbighom}. Note that overall sign is negative, indicating
-  correct conclusion that $v_0$ is slow.}
-\plot{fwistack}{width=\textwidth}{FWI gradient at $v=v_0$ for line of shots 
-  extending from 2100 m to 8000 m. Scaled by
-10$^6$ for plotting. Receiver locations are windowed so 
-  that the range of offsets is the same for all shots. In effect, the 
-  gradient is a stack of translates of the gradient in Figure
-  \ref{fig:fwigrad6kmbighom}.}
-\plot{ivastack}{width=\textwidth}{Source-receiver Extended FWI gradient at $v=v_0$ for line of shots 
-  extending from 2100 m to 8000 m. Receiver locations are windowed so 
-  that the range of offsets is the same for all shots. In effect, the 
-  gradient is a stack of translates of the gradient in Figure
-  \ref{fig:ivagrad6kmbighom}.}
+%\plot{./velbig_dec}{width=\textwidth}{Example: $v$ = 2 km/s for $z \le
+%  0.5 km$, $=2.0 + (z-0.5)*0.5$ km/s for $z>0.5$ km. Source location
+%  $x=5100$, $z=10$ m and receiver locations $x=3000 - 6000$ m,
+%$z=3000$ m indicated.}
+%\plot{shot6kmhom}{width=\textwidth}{Shot gather for geometry indicated
+%  in Figure \ref{fig:velbig_dec_exp}, generated with homogeneous velocity:
+%  $v_0=2$ km/s}
+%\plot{shot6kmbig}{width=\textwidth}{Shot gather for geometry and
+%  $v$ as in Figure \ref{fig:velbig_dec_exp}.}
+%\plot{shot6kmbig_decon}{width=\textwidth}{Inverse image of data in 
+%  Figure (\ref{shot6kmbig}) under extended modeling operator at $v_0$
+%  = 2 km/s}
+%\plot{shot6kmbig_bigdecon}{width=\textwidth}{Inverse image of data in 
+%  Figure \ref{shot6kmbig} under extended modeling operator at target
+%  $v$ (Figure \ref{fig:velbig_dec_exp}).}
+%\plot{fwigrad6kmbighom}{width=\textwidth}{FWI gradient for shot gather
+%defined in Figure \ref{fig:velbig_dec_exp}, at $v=v_0$ = 2 km/s. Scaled by
+%10$^6$ for plotting. Note very
+%small values, and sign is positive, indicating that FWI objective {\em
+%increases} in the direction of the target velocity.}
+%\plot{fwiscan}{width=\textwidth}{Plot of values of the FWI objective
+%  along the line segment between $v_0$ and $v$, for the shot gather
+%  described in Figure \ref{fig:velbig_dec_exp}. For abscissa $h$, value
+%  plotted is FWI objective at $h v_0 + (1-h) v$. Note that objective
+%  does in fact initially increase from its value at $v_0$, consistent
+%  with the positive average value of the gradient as shown in Figure \ref{fig:fwigrad6km%bighom}. }
+%\plot{ivagrad6kmbighom}{width=\textwidth}{Source-receiver Extended FWI
+%  gradient, same geometry, data, and velocity model as in Figure
+%  \ref{fig:fwigrad6kmbighom}. Note that overall sign is negative, indicating
+% correct conclusion that $v_0$ is slow.}
+%\plot{fwistack}{width=\textwidth}{FWI gradient at $v=v_0$ for line of shots 
+%  extending from 2100 m to 8000 m. Scaled by
+%10$^6$ for plotting. Receiver locations are windowed so 
+%  that the range of offsets is the same for all shots. In effect, the 
+%  gradient is a stack of translates of the gradient in Figure
+%  \ref{fig:fwigrad6kmbighom}.}
+%\plot{ivastack}{width=\textwidth}{Source-receiver Extended FWI gradient at $v=v_0$ for line of shots 
+%  extending from 2100 m to 8000 m. Receiver locations are windowed so 
+%  that the range of offsets is the same for all shots. In effect, the 
+%  gradient is a stack of translates of the gradient in Figure
+%  \ref{fig:ivagrad6kmbighom}.}
 

### 2024-02-15 repro/papers/srcextov/trans/prop.tex
cleaned up

--- 2019-10-23 repro/papers/trans/prop.tex
@@
 \end{algorithm}
-
-\section{Prototype Numerical Examples}
-\inputdir{project}
-
-I present a collection of simple examples that illustrate the features of the surface source extension claimed in preceding sections.
-
-I used the IWAVE acoustic staggered grid package to carry out these calculations.
-This package implements (2,2k) schemes for k=1,2,..., and outputs traces (of either velocity or pressure at any point in space via multilinear interpolation.
-The discretized modeling operator is thus of second order accuracy, though as usual I have used higher order in space to reduce grid dispersion.
-
-Source injection is implemented as the adjoint of trace sampling, resulting in another second-order error [REFERENCES].
-
-The data is a single shot gather, with a source at coordinates $x_s=z_s=3000$ m (units of length are meters in all cases).
-The receiver line occupies $1500 \le x_r \le 5500$ m, with receiver depth $z_r=1000$ m.
-Extended sources occupy $1500 \le x_s \le 5500$ m, with the same depth $z_s=3000$ m as the ``physical'' source used to generate the data.
-This region turns out to be adequate to represent the extended sources that approximately invert the data, for the cases examined below.
-An algorithm to automatically identify an appropriate region can be based on the ideas developed by \cite{Fu:Geo17}.
-
-I have used absorbing boundary conditions (split-field PML) on all four sides of the 4000 m (vertical) $\times$ 8000 m (horizontal) simulation domain.
-Evidently includion of a free surface is important to the application of the ideas explained here to diving wave marine data, and I have not addressed the necessary modifications here.
-
-The script is set up to carry out the necessary computations on grids with spacings $\Delta x = \Delta z = $ 20, 10, and 5 m, with a jump of roughly 8 in computation time resulting from each refinement.
-For present purposes, the coarsest (20 m) grid seems to be sufficient, and that is the grid used in the examples presented below.
-The source pulses are chosen so that the computation is reasonably accurate.
-For the 20 m grid case, I use a zero-phase trapezoidal bandpass filter source with corner frequencies of 1.0, 2.0, 7.5, and 12.5 Hz.
-
-The IWAVE asg driver has been set up to recognize the case {\tt deriv=0} as defining the map from source (right-hand side in the pressure equation) to data (pressure) traces.
-The adjoint to this map, as explained above, is reverse-time propagation of the data traces as pressure sources, followed by scaling (formula \ref{eqn:sadj}).
-The approximate inverse is computed by application of the Dirichlet-to-Neumann operator to the pressure traces to produce corresponding velocity traces, followed by injection as pressure sources and reverse time propagation, followed by another application of the Dirichlet-to-Neumann map and scaling (formulas \ref{eqn:appinv}, \ref{eqn:adj}).
-
-The script for the first set of examples implements these operatations step-by-step via calls to IWAVE, Madagascar, and SU commands.
-A peculiarity of the {\tt asg.x} driver needs to be mentioned: it is based on an un-scaled version of the constitutive law defect source representation, that is, {\tt asg.x} approximately computes the solution of the system \ref{eqn:awedata} with the first equation replaced by
-\begin{equation}
-\label{eqn:asgdata}
-\frac{\partial p}{\partial t}  =  - \kappa \nabla \cdot \bv +
-h \delta(z-z_s).
-\end{equation}
-Denote by $S_{\rm asg}[c]$ the forward map produced by {\tt asg.x}. Then comparison of \ref{eqn:awedata} and \ref{eqn:asgdata} reveals that
-\begin{equation}
-\label{eqn:sreln}
-S[c]=S_{\rm asg}[c]\kappa
-\end{equation}
-where $\kappa$ is shorthand for the operator of multiplication by $\kappa$. Accordingly, and approximate inverse for $S_{\rm asg}[c]$ is
-\[
-I \approx S_{\rm asg}[c]^{\dagger}S_{\rm asg}[c] = S_{\rm asg}[c]^{\dagger}S[c]\kappa^{-1}
-\]
-Since $S[c]S[c]^{\dagger} \approx I \approx S[c]^{\dagger}S[c]$, it follows that
-\[
-S_{\rm asg}[c]^{\dagger}=\kappa S[c]^{\dagger} = \kappa\Lambda[c]_s S[c]^T \Lambda[c]_r 
-\]
-\begin{equation}
-\label{eqn:asginv}
-= \kappa \Lambda[c]_s \kappa S_{\rm asg}[c]^T\Lambda[c]_r
-\end{equation}
-This is the approximate inverse computed in the examples.
-Note that only the values of $\kappa$ near the source datum $z=z_s$ play a role in the relation \ref{eqn:sreln} or in the definition \ref{eqn:asginv} of the IWAVE ASG approximate inverse.
-
-These examples are based on two different material models.
-The first is a homogeneous model, with $\kappa=4.0$ GPa and $\rho=1.0$ g/cm$^{3}$.
-Figure \ref{fig:ptpwindh0} shows the shot gather computed in the configuration descritbed above for this choice of propagation medium.
-As mentioned in the preceding section, the velocity traces can be recorded in the same IWAVE workflow: these are displayed in
-\ref{fig:ptvzwindh0}.
-
-The first group of plots show pressure traces for source position $z_s=3$ km, $x_s=3$ km, 201 receivers at $z_r=1$ km, $2 \le x_r \le 6$ km.
-
-\plot{ptpwindh0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km.
-Homogeneous material model, $\kappa=4$ GPa, $\rho$=1 g/cm$^{3}$.
-Coarse grid: $dx=dz=20$ m, $dt=$ 8 ms, (2,4) staggered grid scheme.
-}
-\plot{ptvzwindhh0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindh0}.}
-\plot{srcvzghh0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindhh0} by time reversal (approximate inverse) algorithm explained in text.}
-\plot{reptpwindhh0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzghh0}.} 
-\plot{trcftrhh0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindh0}, \ref{fig:reptpwindhh0}.}.
-
-\plot{bml0}{width=15cm}{Inhomogenous bulk modulus field, with low velocity lens.}
-\plot{ptpwindl0}{width=15cm}{Receiver line at $z_r=1$ km for source at $z_s=3$ km, $x_s=3$ km.
-bulk modulus with low velocity lens as in Figure \ref{fig:bml0}, $\rho$=1 g/cm$^{3}$.
-Coarse grid: $dx=dz=20$ m, $dt=$ 8 ms, (2,4) staggered grid scheme.
-}
-\plot{ptvzwindll0}{width=15cm}{$v_z$ traces corresponding to the pressure traces of Figure \ref{fig:ptpwindl0}.}
-
-\plot{srcvzglh0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindll0} by time reversal (approximate inverse) algorithm explained in text, propagation in homogeneous material model (that is, ``wrong velocity'').}
-\plot{reptpwindlh0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzglh0}, both approximate inversion and re-simulation in  homogeneous material model (``wrong velocity'').}
-\plot{trcftrlh0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindl0}, \ref{fig:reptpwindlh0}.}
-
-\plot{srcvzgll0}{width=15cm}{Source reconstructed from $v_z$ traces in Figure \ref{fig:ptvzwindll0} by time reversal (approximate inverse) algorithm explained in text, propagation in bulk modulus of Figure \ref{fig:bml0} (that is, ``correct velocity'').}
-\plot{reptpwindll0}{width=15cm}{Resimulated pressure data from reconstructed source if Figure \ref{fig:srcvzgll0}, both approximate inversion and re-simulation in bulk modulus of Figure \ref{fig:bml0} (``correct velocity'').}
-\plot{trcftrll0}{width=15cm}{Middle trace from pressure, resimulated pressure traces in Figures \ref{fig:ptpwindl0}, \ref{fig:reptpwindll0}.}

### 2024-02-15 repro/papers/sseprecond/paper.tex
cleaned up

@@
 
+%#############
+
+Another figure \ref{fig:ddifffwdaltplh0}
+More figures: \ref{fig:compnrestrunc0lh0}, \ref{fig:compnrestrunc1lh0}.
+
 \section{Figures}
@@
 
+\plot{compnrestrunc0lh0}{width=\textwidth}{Comparison of normal residual 
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+  vs. iteration. Data = lens model, point source (Figure 
+  \ref{fig:recplh0}), inversion in homogenous model. Penalty weight 
+  $\alpha=0$.}
+
+\plot{compnrestrunc0ll0}{width=\textwidth}{Comparison of normal residual 
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+  vs. iteration. Data = lens model, point source (Figure 
+  \ref{fig:recplh0}), inversion in lens model. Penalty weight 
+  $\alpha=0$.}
+
 \plot{compnres1hh0}{width=\textwidth}{Comparison of normal residual 
@@
 
+\plot{compnrestrunc1lh0}{width=\textwidth}{Comparison of normal residual 
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+  vs. iteration. Data = lens model, point source (Figure 
+  \ref{fig:recplh0}), inversion in homogenous model. Penalty weight 
+  $\alpha=10^{-3}$.}
+
+\plot{compnrestrunc1ll0}{width=\textwidth}{Comparison of normal residual 
+  (gradient) Euclidean norms: CG (blue), PCG (red), plotted 
+  vs. iteration. Data = lens model, point source (Figure 
+  \ref{fig:recplh0}), inversion in lens model. Penalty weight
+  $\alpha=10^{-3}$.}
 \newpage
+
+\plot{ddifffwdaltplh0}{width=\textwidth}{Diff}
 
@@
 
+\newpage
+
+\section{More plots}
+
+A plot of the model with rays overlain, \ref{fig:olensrays0}.
+
+\plot{olensrays0}{width=\textwidth}{Lens model with rays.}
+
 \bibliographystyle{seg}

### 2024-02-15 repro/papers/sseprecond/plots.tex
cleaned up

@@
 The upshot is the cycle represented in the figures \ref{fig:srcvzhh0},  \ref{fig:resrcvzhh0}, \ref{fig:srcvzlh0},  \ref{fig:resrcvzlh0}.
-Figure \ref{fig:srcvzhh0} is the normal component of velocity at $z_s=3$ km of backpropagation, from the field with pressure trace displayed in Figure \ref{fig:ptpwindh0} at $z_r=1$ km, propagation and backpropagation in homogeneous material model with $\kappa = $ 4 Gpa, $\rho = $ 1 g/m$^{-6}$.The pressure trace at $z=z_s$ was also recorded, and used to recompute the normal velocity trace \ref{fig:resrcvzhh0}.
+Figure \ref{fig:srcvzhh0} is the normal component of velocity at $z_s=3$ km of backpropagation, from the field with pressure trace displayed in Figure \ref{fig:ptpwindhh0} at $z_r=1$ km, propagation and backpropagation in homogeneous material model with $\kappa = $ 4 Gpa, $\rho = $ 1 g/m$^{-6}$.The pressure trace at $z=z_s$ was also recorded, and used to recompute the normal velocity trace \ref{fig:resrcvzhh0}.
 The difference is displayed as figure \ref{fig:dresrcvzhh0}.
 
-Similarly, Figure \ref{fig:srcvzlh0} is the normal component of velocity at $z_s=3$ km of backpropagation, from the field with pressure trace displayed in Figure \ref{ptpwindl0} at $z_r=1$ km, propagation in the model with the model with bulk modulus displayed in Figure \ref{fig:bml0} and density $\rho = $ 1 g/m$^{-6}$, and backpropagation in the homogeneous model with bulk modulus = 4 GPa.
+Similarly, Figure \ref{fig:srcvzlh0} is the normal component of velocity at $z_s=3$ km of backpropagation, from the field with pressure trace displayed in Figure \ref{ptpwindll0} at $z_r=1$ km, propagation in the model with the model with bulk modulus displayed in Figure \ref{fig:bml0} and density $\rho = $ 1 g/m$^{-6}$, and backpropagation in the homogeneous model with bulk modulus = 4 GPa.
 The pressure trace at $z=z_s$ was also recorded, and used to recompute the normal velocity trace \ref{fig:resrcvzlh0}.
@@
 \plot{dresrcptplh0}{width=15cm}{Difference of figures \ref{fig:srcptplh0}, \ref{fig:resrcptplh0}.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\plot{recphh0}{width=15cm}{Pressure data, homogeneous velocity model}
+
+\plot{rerecphh0}{width=15cm}{Resimulated pressure traces, Input data simulated in homogeneous velocity model (Figure \ref{fig:recphh0}).
+Resimulation source computed from p traces at receivers only, backpropagated in homogeneous velocity model.}
+
+\plot{reptr21hh0}{width=15cm}{Resimulated pressure trace 21, from data simulated in homogeneous velocity model, backpropagated in homogeneous velocity model.
+Red: target data.
+Blue: resimulated data, with resimulationsource computed from p traces at receivers only.}
+
+\plot{reptr51hh0}{width=15cm}{Resimulated pressure trace 51, from data simulated in homogeneous velocity model, backpropagated in homogeneous velocity model.
+Red: target data.
+Blue: resimulated data, with resimulationsource computed from p traces at receivers only.}
+
+\plot{reptr81hh0}{width=15cm}{Resimulated pressure trace 81, from data simulated in homogeneous velocity model, backpropagated in homogeneous velocity model.
+Red: target data.
+Blue: resimulated data, with resimulationsource computed from p traces at receivers only.}
+
+\plot{recplh0}{width=15cm}{Pressure data, lens velocity model}
+
+\plot{rerecplh0}{width=15cm}{Resimulated pressure traces, Input data simulated in lens velocity model (Figure \ref{fig:recplh0}).
+Resimulation source computed from p traces at receivers only, backpropagated in homogeneous velocity model.}
+
+\plot{reptr21lh0}{width=15cm}{Resimulated pressure trace 21, from data simulated in lens velocity model, backpropagated in homogeneous velocity model.
+Red: target data.
+Blue: resimulated data, with resimulationsource computed from p traces at receivers only.}
+
+\plot{reptr51lh0}{width=15cm}{Resimulated pressure trace 51, from data simulated in lens velocity model, backpropagated in homogeneous velocity model.
+Red: target data.
+Blue: resimulated data, with resimulationsource computed from p traces at receivers only.}
+
+\plot{reptr81lh0}{width=15cm}{Resimulated pressure trace 81, from data simulated in lens velocity model, backpropagated in homogeneous velocity model.
+Red: target data.
+Blue: resimulated data, with resimulationsource computed from p traces at receivers only.}
+
+\plot{rerecpll0}{width=15cm}{Resimulated pressure traces, Input data simulated in lens velocity model (Figure \ref{fig:recplh0}).
+Resimulation source computed from p traces at receivers only, backpropagated in lens velocity model.}
+
+\plot{reptr21ll0}{width=15cm}{Resimulated pressure trace 21, from data simulated in lens velocity model, backpropagated in lens velocity model.
+Red: target data.
+Blue: resimulated data, with resimulationsource computed from p traces at receivers only.}
+
+\plot{reptr51ll0}{width=15cm}{Resimulated pressure trace 51, from data simulated in lens velocity model, backpropagated in lens velocity model.
+Red: target data.
+Blue: resimulated data, with resimulationsource computed from p traces at receivers only.}
+
+\plot{reptr81ll0}{width=15cm}{Resimulated pressure trace 81, from data simulated in lens velocity model, backpropagated in lens velocity model.
+Red: target data.
+Blue: resimulated data, with resimulationsource computed from p traces at receivers only.}

### 2024-02-15 repro/papers/ssevelan/paper.tex
cleaned up

@@
 
-\inputdir{../trans/project}
+\inputdir{project}
 
@@
 \label{eqn:proj}
-\oS_1^{\dagger}\oS &=& \Pi^1_s\\
-\oS \oS_1^{\dagger} &=& \Pi^1_r
+\oS_1^{\dagger}\oS_1 &=& \Pi^1_s\\
+\oS_1 \oS_1^{\dagger} &=& \Pi^1_r
 \end{eqnarray}
@@
 \[
-D^2\Jared = \|[Q,A_r]\ofa\|^2 + O(\lambda^2)
+D^2\Jared = \alpha^2\|[Q,A_r]\ofa\|^2 + O(\lambda^2)
 \]
@@
 
-{\color{blue} To 2nd order at consistent (model, data), SSE inversion $\approx$ traveltime tomography - $J_{\alpha}$ convex in ball of radius $O(1)$ in ``tomographically detectable'' subspace - {\em much} larger region of convexity than FWI}
+{\color{blue} To 2nd order at consistent (model, data), SSE inversion
+  $\approx$ traveltime tomography - $J_{\alpha}$ convex in ball of
+  radius $O(1)$ in ``tomographically detectable'' subspace - {\em
+    much} larger region of convexity than FWI}
+
+% \plot{ptsrc00}{width=\textwidth}{Bandpass filter pulse}
+
+\plot{bml0}{width=\textwidth}{Lens model}
+\plot{plbulksm0}{width=\textwidth}{Layered model}
+\plot{oplrays0}{width=\textwidth}{Diving ray model}
+\plot{ooplrays0}{width=\textwidth}{Diving ray model}
+\plot{ndwdata0}{width=\textwidth}{Muted point source data}
+\plot{ndwdatavz0}{width=\textwidth}{Muted point source data}
+\plot{ndwdataresrcplbulksm0}{width=\textwidth}{Muted point source data}
+\plot{ndwdataresimplbulksm0}{width=\textwidth}{Muted point source
+  data}
+\plot{ndwdatarediffplbulksm0}{width=\textwidth}{Muted point source data}
+\plot{ndwraw0}{width=\textwidth}{Point source data}
+\plot{ndwmute0}{width=\textwidth}{Mute function}
+
+\plot{ooxtraysend}{width=\textwidth}{Ray crossing times $z=0$}

### 2024-02-15 repro/papers/wri1d/paper.tex
cleaned up

@@
+\address{email: {\tt symes@rice.edu}}
 \title{Wavefield Reconstruction Inversion: an example}
-\author{William. W. Symes \thanks{The Rice Inversion Project,
-Department of Computational and Applied Mathematics, Rice University,
-Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}
+\author{William. W. Symes}
 
@@
 \end{equation}
-Examples given
-by \cite{LeeuwenHerrmannWRI:13} and others suggest that the problem
-defined in display \ref{eqn:defwri} is less likely to develop
-uninformative local minima if $\alpha$ is small. This might be so for
-several reasons: one is that when less weight is
-put on making the residual in the wave equation ($f-q$) small, 
-$f$ may be chosen to make the data residual $d-S[c]f$ small
-instead. One aspect of failure to predict the arrival times of waves
-accurately is that small data residual is then difficult to
-achieve. By maintaining fidelity to
-the data, local minimization of the WRI objective might provide a
-spurious-minimum-free path to a satisfactory estimate of $c$.
+\cite{Leeuwen2019note} points out that the formulation
+\ref{eqn:defwri} is available not just for seismic problems, but for
+any inverse problem based on a separable (partly linear) modeling
+operator.
+
+Examples given by \cite{LeeuwenHerrmannWRI:13} and others suggest that
+the problem defined in display \ref{eqn:defwri} is less likely to
+develop uninformative local minima than is the least squares problem
+\ref{eqn:deffwi} in application to seismic inversion, if $\alpha$ is
+sufficiently small. This might be so for several reasons: one is that
+when less weight is put on making the residual in the wave equation
+($f-q$) small, $f$ may be chosen to make the data residual $d-S[c]f$
+small instead. One aspect of failure to predict the arrival times of
+waves accurately is that small data residual is then difficult to
+achieve. By maintaining fidelity to the data, local minimization of
+the WRI objective might provide a spurious-minimum-free path to a
+satisfactory estimate of $c$.
 
@@
 approach, based on explicit solution of the 1D acoustic system
-presented in Appendix A. In the third section I introduce the $\lambda-$dependent
-family of problems, and establish the asymptotic properties of FWI as
-$\lambda \rightarrow 0$. The fourth section develops the algebraic
-structure of WRI, culminating in a remarkable identity revealing WRI
-to be equivalent to minimization of a weighted norm of the data
-residual, with a weight operator depending on the coefficient vector
-$c$. This identity has also been derived by \cite{Leeuwen2019note},
-using a different argument. The result is quite general, applying to essentially any
-realization of WRI, and already shows that its behaviour must be
-closely related to that of FWI. In the fifth section I return to 
-1D acoustics problem and the $\lambda$-dependent family of inverse
-problems, compute that weight operator explicitly, and deduce the
-global behaviour of WRI in this instance. The paper ends with a
-discussion of the relation of the analysis presented here to a
-previous analysis of optimization formulations of wave inversion
-problems based on parameter-dependent quadratic forms
-\cite[]{StolkSymes:03}, and a brief discussion of some other
-alternatives to FWI.
+presented in Appendix A. In the third section I introduce the
+$\lambda-$dependent family of problems, and establish the asymptotic
+properties of FWI as $\lambda \rightarrow 0$. The fourth section
+develops the algebraic structure of WRI, culminating in a remarkable
+identity revealing WRI to be equivalent to minimization of a weighted
+norm of the data residual, with a weight operator depending on the
+coefficient vector $c$. This identity has also been derived by
+\cite{Leeuwen2019note}, using a different argument. The result is
+quite general, applying to essentially any realization of WRI, and
+already shows that its behaviour must be closely related to that of
+FWI. In the fifth section I return to 1D acoustics problem and the
+$\lambda$-dependent family of inverse problems, compute that weight
+operator explicitly, and deduce the global behaviour of WRI in this
+instance. The paper ends with a discussion of the
+relation of the analysis presented here to a previous analysis of
+optimization formulations of wave inversion problems based on
+parameter-dependent quadratic forms \cite[]{StolkSymes:03}, and a
+brief discussion of some other alternatives to FWI.
 
 \section{FWI for 1D Acoustics}
-The example of FWI to be explored in this paper is one of the simplest possible, based on the 1D acoustics system connecting excess pressure $p$, particle velocity $v$, sonstitutive law defect (``source'') $f$, density $\rho$, and wave velocity $c$:
+The example of FWI to be explored in this paper is one of the simplest possible, based on the 1D acoustics system connecting excess pressure $p$, particle velocity $v$, constitutive law defect (``source'') $f$, density $\rho$, and wave velocity $c$:
 \begin{eqnarray}
@@
 
-For homogeneous ($z$-independent) $\kappa$, Appendix A provides an explicit expression:
+For homogeneous ($z$-independent) $c$, Appendix A provides an explicit expression:
 \begin{equation}
@@
 $w \in L^2(\bR)$. Formally, the resulting acoustic field satisfies the
-system \ref{eqn:awe1d} with $f=w(t)\delta(z-z_s)$. Inserting this
+system \ref{eqn:awe1d} with $f(z,t)=w(t)\delta(z-z_s)$. Inserting this
 expression in the explicit expression \ref{eqn:defmodhom}, obtain
-\begin{equation}
+\begin{eqnarray}
+  f(z,t) & = & w(t)\delta(z-z_s)\nonumber\\
+  (S[c]f)(t) &=& \frac{1}{2c}w\left(t -
+                              \frac{|z_r-z_s|}{c}\right)\nonumber\\
+  &= &(S_p[c]w)(t)
   \label{eqn:defmodpt}
-  S[c](w\delta(\cdot-z_s)) = \frac{1}{2c}w\left(t - \frac{|z_r-z_s|}{c}\right) = S_p[c]w(t)
-\end{equation}
+\end{eqnarray}
 Thus for this restricted class of source fields, the FWI objective can be redefined as
@@
 and a corresponding family of FWI objectives:
-\[
-  J_{\rm FWI}[c,\dl,\wl] = \frac{1}{2}\|\dl-S_p[c]\wl\|^2
-\]
-\begin{equation}
+\begin{eqnarray}
+  J_{\rm FWI}[c,\dl,\wl] &=& \frac{1}{2}\|\dl-S_p[c]\wl\|^2\nonumber\\
+  &=&  \int_0^T\,dt\,\left|\frac{1}{2c_*}\wl\left(t-\frac{|z_s-z_r|}{c_*}\right)
+    - \frac{1}{2c}\wl\left(t-\frac{|z_s-z_r|}{c}\right)\right|^2
   \label{eqn:fwiexpl}
-  =
-  \int_0^T\,dt\,\left|\frac{1}{2c_*}\wl\left(t-\frac{|z_s-z_r|}{c_*}\right)
-    - \frac{1}{2c}\wl\left(t-\frac{|z_s-z_r|}{c}\right)\right|^2
-\end{equation}
+\end{eqnarray}
 Note that $\mbox{supp }\wl \subset [0,\lambda]$, so
@@
 \end{quote}
-That is, {em any data can be fit exactly using the extended model
+That is, {\em any data can be fit exactly using the extended model
   space (the domain of $S$)}.
@@
 Consequently
-\[
-  J_{\rm WRI}^{\alpha}[c,d,q] =
-  \frac{1}{2}(\|r[c]-S[c]S[c]^Te_{\alpha}[c]\|^2+\alpha^2\|S[c]^Te_{\alpha}[c]\|^2)
-\]
-\[
-  = \frac{1}{2}\left(\|r[c]-S[c]S[c]^T
-    (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\|^2 \right.
-\]
-\[
+\begin{eqnarray}
+  J_{\rm WRI}^{\alpha}[c,d,q] &=&
+  \frac{1}{2}(\|r[c]-S[c]S[c]^Te_{\alpha}[c]\|^2+\alpha^2\|S[c]^Te_{\alpha}[c]\|^2)\nonumber\\
+  &=& \frac{1}{2}\left(\|r[c]-S[c]S[c]^T
+    (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\|^2\right. \nonumber\\
+  & &
   \left.+\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],
   S[c]S[c]^T (S[c]S[c]^T+\alpha^2I)^{-1}r[c]
-  \rangle\right)
-\]
-\[
-  =\frac{1}{2}\left(\|\alpha^2 (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\|^2\right.
-\]
-\[
-  \left.
-  +\alpha^2\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],
-  -\alpha^2 (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\rangle\right.
-\]
-\[
-  + \left.\alpha^2\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],r[c]\rangle\right)
-\]
-\[
-  =  \frac{\alpha^2}{2}
-\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],r[c]\rangle 
-\]
-That is,
+  \rangle\right)  \nonumber\\
+  &=&\frac{1}{2}\left(\|\alpha^2 (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\|^2\right. \nonumber\\
+  & &
+  + \left. \alpha^2 \langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],
+  - \alpha^2 (S[c]S[c]^T+\alpha^2I)^{-1}r[c]\rangle\right. \nonumber\\
+  & &  + \left. \alpha^2\langle (S[c]S[c]^T+\alpha^2I)^{-1}r[c],r[c]\rangle\right) \nonumber\\
+  &=&  \frac{\alpha^2}{2}\langle
+      (S[c]S[c]^T+\alpha^2I)^{-1}r[c],r[c]\rangle.
+   \label{eqn:longone}
+\end{eqnarray}
+Rearranging the RHS of equation \ref{eqn:longone}, obtain
 \begin{equation}
@@
   \ref{eqn:fwiscrewed}, if $|c-c_*| > L\lambda$,
-  \[
-    J_{\rm WRI}^{\alpha}[c,\dl,\wl\delta(\cdot-z_s)] =
+\begin{eqnarray}
+    J_{\rm WRI}^{\alpha}[c,\dl,\wl\delta(\cdot-z_s)] &=&
     \frac{\alpha^2}{2}\left(\frac{z_{\rm max}-z_{\rm
-          min}}{4c^2}+\alpha^2\right)^{-1}\frac{1}{2}\left(\frac{1}{4c^2}+\frac{1}{4c_*^2}\right)\|w_1\|^2
-  \]
-  \[
-    =
-    \frac{\alpha^2}{4}\frac{1+\frac{c^2}{c_*^2}}{z_{\rm 
+          min}}{4c^2}+\alpha^2\right)^{-1}\frac{1}{2}\left(\frac{1}{4c^2}+\frac{1}{4c_*^2}\right)\|w_1\|^2\nonumber\\
+   & =&    \frac{\alpha^2}{4}\frac{1+\frac{c^2}{c_*^2}}{z_{\rm 
         max}-z_{\rm min}+4c^2\alpha^2}
-  \]
-  This linear fractional function of $c^2$ is increasing, decreasing,
+  \label{eqn:yetanutta}
+\end{eqnarray}
+  The linear fractional function of $c^2$ on the RHS of equation
+  \ref{eqn:yetanutta} is increasing, decreasing,
   or constant if $\beta>0, \beta<0$ or $\beta=0$, respectively.
@@
 First, a consequence of the results proven here: $J_{\rm FWI}[c,d,w]$
-and $J_{\rm WRI}[c,d,q]$ are not smooth as joint functions of model
+and $J_{\rm WRI}^{\alpha}[c,d,q]$ are not smooth as joint functions of model
 ($c$) and data ($d$) vectors. If they were, their derivatives would be
@@
 $c$ is a vector of material parameters to be estimated, $G[c]$ is a
-$c$-dependent family of operators having the same nature as the
-simulation operators $S[c]$ and $S_p[c]$ figuring in the preceding
-discussion, and $A[c]$ is an operator-valued function of $c$, on the range
-of $G[c]$.
+$c$-dependent family of operators, whose common range is a Hilbert
+space with inner product $\langle \cdot, \cdot \rangle$.
+$A[c]$ is an operator-valued function of $c$, with domain and the
+range equal to the range of $G[c]$.
 
@@
 operators for wave equation inverse problems, or their inverses or
-adjoints. The precise characterization of the so-called {\em microlocally elliptic Fourier Integral
+adjoints. The precise characterization of these so-called {\em microlocally elliptic Fourier Integral
 Operators} is quite technical \cite[]{Dui:95}. Roughly speaking, such operators map high-frequency localized wave packets to other
 such packets with well-defined changes of position and direction of
-oscillation. If the dependence of such an operator $G[c]$ on $c$ is of sufficiently full rank, in
+oscillation. The simulation operators $S[c]$ and $S_p[c]$ figuring in the preceding
+discussion are particularly simple examples of this type.
+
+If the dependence of such an operator $G[c]$ on $c$ is of sufficiently full rank, in
 the sense that destination packets can be shifted in any direction by
@@
 extended to $(-\infty,\infty)$ it would have a left inverse too). From
-the defintion \ref{eqn:dfam} of the data family $\dl$, one sees that
+the definition \ref{eqn:dfam} of the data family $\dl$, one sees that
 $\wl = S_p[c_*]^{-1}\dl$. Therefore
-\[
-  J_{\rm FWI}[c,\dl,\wl] = \frac{1}{2}\|(I-S_p[c]S_p[c_*]^{-1})\dl\|^2
-\]
-\[
-  =\frac{1}{2}(\|\dl\|^2 + \|S_p[c]S_p[c_*]^{-1}\dl\|^2) + \langle \dl,
+\begin{eqnarray}
+  \label{eqn:anutta}
+  J_{\rm FWI}[c,\dl,\wl] &=&
+                             \frac{1}{2}\|(I-S_p[c]S_p[c_*]^{-1})\dl\|^2 \nonumber\\
+ & =&\frac{1}{2}(\|\dl\|^2 + \|S_p[c]S_p[c_*]^{-1}\dl\|^2) + \langle \dl,
   S_p[c]S_p[c_*]^{-1}\dl\rangle
-\]
+\end{eqnarray}
 For the operator family $S_p[c]$ defined above, it is easy to see that
-the second term is smooth in $c$, and the first is constant. The third 
-can be rewritten as 
-\[
+the second term in the right hand side of equation \ref{eqn:anutta} is
+smooth in $c$, and the first is constant. The third can be rewritten
+as
+\begin{equation}
+  \label{eqn:evenmore}
   \langle \dl,  S_p[c]S_p[c_*]^{-1}\dl\rangle = \langle S[c]^T\dl, 
-  (S_p[c]^T S_p[c_*])^{-1}S_p[c]^Td\rangle 
-\]
-which has the form \ref{eqn:gen1} with the choices $G[c]=S_p[c]^T$,
+  (S_p[c]^T S_p[c_*])^{-1}S_p[c]^Td\rangle.
+\end{equation}
+The RHS of equation \ref{eqn:evenmore} has the form \ref{eqn:gen1} with the choices $G[c]=S_p[c]^T$,
 $A[c]=(S_p[c]^T S_p[c_*])^{-1}$. A similar manipulation exhibits
@@
 
+\section{Acknowledgements}
+The author thanks Felix Herrmann, Tristan van Leeuwen, David Yingst,
+and Chao Wang for informative conversations about Wavefield
+Reconstruction Inversion. Comments from Joakim Blanch and John
+Washbourne were very helpful in improving the manuscript. The sponsors
+of The Rice Inversion Project provided long-term support for the
+author's research, leading to the work reported here.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2024-02-15 iwave/acd/lib/acd_gfdm.cc

@@ -76,19 +76,19 @@ int acd_tsfm(RDOM * p, RDOM * r, int ia, void * fdpars) {
   IPNT i;
 
   /* pointers for 2D case */
-  register ireal ** restrict uc2;
-  register ireal ** restrict up2;
-  register ireal ** restrict csq2;
-  register ireal ** restrict uc2d;
-  register ireal ** restrict up2d;
-  register ireal ** restrict csq2d;
+  ireal ** restrict uc2;
+  ireal ** restrict up2;
+  ireal ** restrict csq2;
+  ireal ** restrict uc2d;
+  ireal ** restrict up2d;
+  ireal ** restrict csq2d;
   /* pointers for 3D case */
-  //  register ireal *** restrict uc3;
-  //  register ireal *** restrict up3;
-  register ireal *** restrict csq3;
-  //  register ireal *** restrict uc3d;
-  //  register ireal *** restrict up3d;
-  register ireal *** restrict csq3d;
+  //  ireal *** restrict uc3;
+  //  ireal *** restrict up3;
+  ireal *** restrict csq3;
+  //  ireal *** restrict uc3d;
+  //  ireal *** restrict up3d;
+  ireal *** restrict csq3d;
   
   int ndim;                       /* problem dmn */
   IPNT s, s0, cs0;                /* loop starts */ 
@@ -307,19 +307,19 @@ int acd_tsam(RDOM * p, RDOM * r, int ia, void * fdpars) {
   IPNT n;
 
   /* pointers for 2D case */
-  register ireal ** restrict uc2;
-  register ireal ** restrict up2;
-  register ireal ** restrict csq2;
-  register ireal ** restrict uc2d;
-  register ireal ** restrict up2d;
-  register ireal ** restrict csq2d;
+  ireal ** restrict uc2;
+  ireal ** restrict up2;
+  ireal ** restrict csq2;
+  ireal ** restrict uc2d;
+  ireal ** restrict up2d;
+  ireal ** restrict csq2d;
   /* pointers for 3D case */
-  //  register ireal *** restrict uc3;
-  //  register ireal *** restrict up3;
-  register ireal *** restrict csq3;
-  //  register ireal *** restrict uc3d;
-  //  register ireal *** restrict up3d;
-  register ireal *** restrict csq3d;
+  //  ireal *** restrict uc3;
+  //  ireal *** restrict up3;
+  ireal *** restrict csq3;
+  //  ireal *** restrict uc3d;
[…]

### 2024-02-15 iwave/acd/lib/acd_gfdm2.cc

@@ -72,26 +72,26 @@ int acd_tsfm2(RDOM * dd, RDOM *d0, RDOM * d, RDOM * r, int ia, void * fdpars) {
   IPNT i;
 
   /* pointers for 2D case */
-  register ireal ** restrict uc2;
-  register ireal ** restrict up2;
-  register ireal ** restrict csq2;
-  register ireal ** restrict uc2d;
-  register ireal ** restrict up2d;
-  register ireal ** restrict csq2d;
+  ireal ** restrict uc2;
+  ireal ** restrict up2;
+  ireal ** restrict csq2;
+  ireal ** restrict uc2d;
+  ireal ** restrict up2d;
+  ireal ** restrict csq2d;
     
-  register ireal ** restrict uc2d0;
-  register ireal ** restrict up2d0;
-  register ireal ** restrict csq2d0;
-  register ireal ** restrict uc2dd;
-  register ireal ** restrict up2dd;
+  ireal ** restrict uc2d0;
+  ireal ** restrict up2d0;
+  ireal ** restrict csq2d0;
+  ireal ** restrict uc2dd;
+  ireal ** restrict up2dd;
 
   /* pointers for 3D case */
-  //  register ireal *** restrict uc3;
-  //  register ireal *** restrict up3;
-  //  register ireal *** restrict csq3;
-  //  register ireal *** restrict uc3d;
-  //  register ireal *** restrict up3d;
-  register ireal *** restrict csq3d;
+  //  ireal *** restrict uc3;
+  //  ireal *** restrict up3;
+  //  ireal *** restrict csq3;
+  //  ireal *** restrict uc3d;
+  //  ireal *** restrict up3d;
+  ireal *** restrict csq3d;
  
   int ndim;                       /* problem dmn */
   IPNT s, s0, cs0;                /* loop starts */
@@ -298,26 +298,26 @@ int acd_tsam2(RDOM * db, RDOM *b, RDOM * d, RDOM * r, int ia, void * fdpars) {
   IPNT n;
 
   /* pointers for 2D case */
-  register ireal ** restrict uc2;
-  register ireal ** restrict up2;
-  register ireal ** restrict csq2;
-  register ireal ** restrict uc2d;
-  register ireal ** restrict up2d;
-  register ireal ** restrict csq2d;
+  ireal ** restrict uc2;
+  ireal ** restrict up2;
+  ireal ** restrict csq2;
+  ireal ** restrict uc2d;
+  ireal ** restrict up2d;
[…]

### 2024-02-15 iwave/acd/lib/fd_acd.cc

@@ -594,13 +594,13 @@ int acd_timegrid(PARARRAY *pars,
 int acd_step(RDOM* dom, int iv, void * tspars) {
 
   /* pointers for 2D case */
-  register float ** restrict uc2;
-  register float ** restrict up2;
-  register float ** restrict csq2;
+  float ** restrict uc2;
+  float ** restrict up2;
+  float ** restrict csq2;
   /* pointers for 3D case */
-  register float *** restrict uc3;
-  register float *** restrict up3;
-  register float *** restrict csq3;
+  float *** restrict uc3;
+  float *** restrict up3;
+  float *** restrict csq3;
     
   int ndim;                       /* problem dmn */
   IPNT s, s0, cs0;                /* loop starts  */

### 2024-02-15 iwave/ang/include/ADCIG.h

@@ -109,7 +109,7 @@ void HO2ADscat (stADCIG *pADCIG, int nthread, int taperFact) {
       }
     }
     
-    int count = 0;
+    //    int count = 0;
 #ifdef _OPENMP
 #pragma omp parallel default(shared) private(thread_id, n, k, m, j)
 #endif
@@ -131,7 +131,7 @@ void HO2ADscat (stADCIG *pADCIG, int nthread, int taperFact) {
 	  for(m = 0; m < pADCIG->np; m++)
 	    M[k][m][n] = Maux[thread_id][m];
 	}
-	count++;
+	//	count++;
 	//		printf ("thread %d did step %d\n", thread_id, count);
       }
     } // End of parallel region

### 2024-02-15 iwave/ang/include/matrix.h

@@ -610,7 +610,8 @@ int CG(float* x, float* r, float* v, float* Av, float* d, float** A, int N, int 
 {
 
     int i, j, itr;
-    double rs, rr0, vAv, rr, rAv, alpha, beta;
+    //double rs, rr0, vAv, rr, rAv, alpha, beta;
+    double rs, rr0, vAv, rr, alpha, beta;
 
 	//Initiation: r = d-Ax (x=0) ; rr = r*r ; v = r
     rr0 = 0.;
@@ -662,17 +663,17 @@ int CG(float* x, float* r, float* v, float* Av, float* d, float** A, int N, int 
 		//Save rr of cuurent iteration
 		rs  = rr;
 		rr  = 0.;
-		rAv = 0.;
+		//rAv = 0.;
 		//Update: x = x+alpha*v ; r = r-alpha*Av
 		for(i = 0 ; i < N ; ++i){
 		    x[i] += alpha * v[i];
 		    r[i] -= alpha * Av[i];
 		    rr  += (double)(r[i] * r[i]);
-		    rAv += (double)(r[i] * Av[i]);
+		    //rAv += (double)(r[i] * Av[i]);
 		}
 		//Reset the residual every ten iterations
 		if(itr/10*10 == itr){
-		    rAv = 0.;
+		  //rAv = 0.;
 		    rr  = 0.;
 #ifdef _OPENMP
 #pragma omp parallel default(shared) private(i, j) num_threads(8)
@@ -693,7 +694,7 @@ int CG(float* x, float* r, float* v, float* Av, float* d, float** A, int N, int 
 
 		    for(i = 0; i < N; ++i){
 				rr  += (double)(r[i] * r[i]);
-				rAv += (double)(r[i] * Av[i]);
+				//rAv += (double)(r[i] * Av[i]);
 	    	}
 		}
 		printf("CGitr = %d , residue = %lf\n" , itr , rr/rr0);
@@ -723,7 +724,8 @@ int CG_AVA(float* x, float* r, float* v, float* Av, float* d, float** A, int N, 
 
     int i, j, itr;
     int i0, j0, ires, jres;
-    double rs, rr0, vAv, rr, rAv, alpha, beta;
+    double rs, rr0, vAv, rr, alpha, beta;
+    //    double rs, rr0, vAv, rr, rAv, alpha, beta;
 
 	//Initiation: r = d-Ax (x=0) ; rr = r*r ; v = r
[…]

### 2024-02-15 iwave/asg/lib/fd_asg.cc

@@ -467,10 +467,10 @@ void asg_timestep(std::vector<RDOM *> dom,
   i_v[2] = D_V2;
     
   /* fill in pml arrays; these are no-ops after the first call */
-  register ireal * restrict ep[RARR_MAX_NDIM];
-  register ireal * restrict epp[RARR_MAX_NDIM];
-  register ireal * restrict ev[RARR_MAX_NDIM];
-  register ireal * restrict evp[RARR_MAX_NDIM];
+  ireal * restrict ep[RARR_MAX_NDIM];
+  ireal * restrict epp[RARR_MAX_NDIM];
+  ireal * restrict ev[RARR_MAX_NDIM];
+  ireal * restrict evp[RARR_MAX_NDIM];
 
   // local pml widths for this domain
   IPNT nrsloc;
@@ -533,7 +533,7 @@ void asg_timestep(std::vector<RDOM *> dom,
       float *** f3= (dom[0]->_s)[icoeff]._s3;
       //      if (ndim == 2) {
       if ((f2) && (!f3)) {
-	register ireal ** restrict f = (dom[0]->_s)[icoeff]._s2;
+	ireal ** restrict f = (dom[0]->_s)[icoeff]._s2;
 	// extend in dim 1
 	for (int i0=gs[0]+nlsloc[0]; i0<=ge[0]-nrsloc[0]; i0++) {
 	  for (int i1=gs[1]; i1<=gs[1]+nlsloc[1]; i1++)
@@ -551,7 +551,7 @@ void asg_timestep(std::vector<RDOM *> dom,
       }
       //      else if (ndim == 3) {
       else if (f3) {
-	register ireal *** restrict f = (dom[0]->_s)[icoeff]._s3;
+	ireal *** restrict f = (dom[0]->_s)[icoeff]._s3;
 	// extend in dim 2
 	for (int i0=gs[0]+nlsloc[0]; i0<=ge[0]-nrsloc[0]; i0++) {
 	  for (int i1=gs[1]+nlsloc[1]; i1<=ge[1]-nrsloc[1]; i1++) {
@@ -619,17 +619,22 @@ void asg_timestep(std::vector<RDOM *> dom,
   RASN(dha,asgpars->dxs);
   IPNT ihmax;
   IASN(ihmax,asgpars->ihmax);
-  
-  if (ndim==2)
-      
-  // deriv = 0
[…]

### 2024-02-15 iwave/base/lib/parser.cc

@@ -923,7 +923,7 @@ int ps_set(PARARRAY * par, int last, const char * type, const char * key, const 
   else {
     if (!err && !strcmp(type,"char")) {
       memset(l,'\0',MAX_STR_LEN);
-      err=sprintf(l,"%c",*((char *)val));
+      err=snprintf(l,MAX_STR_LEN, "%c",*((char *)val));
       if (0==err || MAX_STR_LEN==err) {
 	err=E_PARSECONVERT;
       }
@@ -931,7 +931,7 @@ int ps_set(PARARRAY * par, int last, const char * type, const char * key, const 
     }
     else if (!err && !strcmp(type,"long")) {
       memset(l,'\0',MAX_STR_LEN);
-      err=sprintf(l,"%ld",*((long *)val));
+      err=snprintf(l,MAX_STR_LEN,"%ld",*((long *)val));
       if (0==err || MAX_STR_LEN==err) {
 	err=E_PARSECONVERT;
       }
@@ -939,7 +939,7 @@ int ps_set(PARARRAY * par, int last, const char * type, const char * key, const 
     }
     else if (!err && !strcmp(type,"ulong")) {
       memset(l,'\0',MAX_STR_LEN);
-      err=sprintf(l,"%lu",*((unsigned long *)val));
+      err=snprintf(l,MAX_STR_LEN, "%lu",*((unsigned long *)val));
       if (0==err || MAX_STR_LEN==err) {
 	err=E_PARSECONVERT;
       }
@@ -948,7 +948,7 @@ int ps_set(PARARRAY * par, int last, const char * type, const char * key, const 
     else if (!err && !strcmp(type,"int")) {
       /*      fprintf(stderr,"in ps_set - int\n"); */
       memset(l,'\0',MAX_STR_LEN);
-      err=sprintf(l,"%d",*((int *)val));
+      err=snprintf(l,MAX_STR_LEN,"%d",*((int *)val));
       /*      fprintf(stderr,"output = %s err=%d\n",l,err); */
       if (0==err || MAX_STR_LEN==err) {
 	err=E_PARSECONVERT;
@@ -957,7 +957,7 @@ int ps_set(PARARRAY * par, int last, const char * type, const char * key, const 
     }
     else if (!err && !strcmp(type,"uint")) {
       memset(l,'\0',MAX_STR_LEN);
-      err=sprintf(l,"%u",*((unsigned int *)val));
[…]

### 2024-02-15 iwave/base/lib/utils.cc

@@ -249,8 +249,8 @@ int getMachineEndianness()
 
 void swapBytes(unsigned char *arr, int arrsize, int atomsize)
 {
-  register int i, j, k, atomsize1;
-  register unsigned char tmp;
+  int i, j, k, atomsize1;
+  unsigned char tmp;
 	
   atomsize1 = atomsize - 1;
 	

### 2024-02-15 iwave/core/lib/fd.cc

@@ -416,7 +416,7 @@ int fd_modelcrea(IPNT cdims, IPNT crank, PARARRAY * par, FILE * stream, IMODEL *
     IPNT gs, ge;
     IPNT dgsr[IWAVE_NNEI], dger[IWAVE_NNEI];                        /*< receive domains */
     IPNT dgsrs[IWAVE_NNEI][RDOM_MAX_NARR], dgers[IWAVE_NNEI][RDOM_MAX_NARR];   /*< all receive domains */
-    int frcvempty[IWAVE_NNEI], rcvne;                       /*< empty receive flag */
+    int frcvempty[IWAVE_NNEI];                       /*< empty receive flag */
   
     /*< grid type in each dimension: primal grid (=0) or dual grid
       (=1) */
@@ -543,9 +543,6 @@ int fd_modelcrea(IPNT cdims, IPNT crank, PARARRAY * par, FILE * stream, IMODEL *
 	    return E_INTERNAL;
 	  }
 
-	  rcvne = 0;
-	  for ( i = 0; i < nnei; ++i ) if ( !(frcvempty[i]) ) ++rcvne;
-	  
 	  for ( i = 0; i < nnei; ++i ) {
 	    //	    fprintf(stream,"iv=%d i=%d dgsr[i][0]=%d dger[i][0]=%d dgsr[i][1]=%d dger[i][1]=%d\n", iv, i, dgsr[i][0],dger[i][0],dgsr[i][1],dger[i][1]); 
 	    IASN(dgsrs[i][iv], dgsr[i]);

### 2024-02-15 iwave/core/lib/iwavefun.cc

@@ -65,8 +65,9 @@ int initoutstream(FILE ** stream, int rk, int sz) {
 
     char *filename = NULL; /* workspace for output filename */
     char *tmp = NULL;   /* ditto */
-    tmp = (char *) usermalloc_(20*sizeof(char));
-    memset(tmp,0,20*sizeof(char));
+    int tmplen = 20;
+    tmp = (char *) usermalloc_(tmplen*sizeof(char));
+    memset(tmp,0,tmplen*sizeof(char));
     filename = (char *)usermalloc_(MAXPATHLEN * sizeof(char));    
     if ( filename == NULL ) {
 	if ( rk == 0 )
@@ -82,19 +83,19 @@ int initoutstream(FILE ** stream, int rk, int sz) {
     size_t aaa = strlen(filename);
     char bbb = filename[aaa-1];
     if ((aaa > 1) && ( bbb != '/')) 
-	strcat(filename,"/");
+      strcat(filename,"/");
     if      ( sz < 10 )
-	sprintf(tmp,"cout%01d.txt",rk);
+      snprintf(tmp,tmplen,"cout%01d.txt",rk);
     else if ( sz < 100 )
-	sprintf(tmp,"cout%02d.txt",rk);
+      snprintf(tmp,tmplen,"cout%02d.txt",rk);
     else if ( sz < 1000 )
-	sprintf(tmp,"cout%03d.txt",rk);
+      snprintf(tmp,tmplen,"cout%03d.txt",rk);
     else if ( sz < 10000 )
-	sprintf(tmp,"cout%04d.txt",rk);
+      snprintf(tmp,tmplen,"cout%04d.txt",rk);
     else if ( sz < 100000 )
-	sprintf(tmp,"cout%05d.txt",rk);
+      snprintf(tmp,tmplen,"cout%05d.txt",rk);
     else
-	sprintf(tmp,"cout%d.txt",rk);
+      snprintf(tmp,tmplen,"cout%d.txt",rk);
 
     strcat(filename,tmp);
 

### 2024-02-15 iwave/core/lib/iwsim.cc

@@ -786,12 +786,14 @@ namespace TSOpt {
 	  drystr<<"    panel="<<panelindex<<" in range=["<<first<<", "<<last<<"], rank="<<retrieveGlobalRank()<<"\n";
 	}
 	else {
+	  // WWS 2024.02.08 changed -1 to 0
 	  if (printact > -1) {
 	    if (panelindex==first && retrieveGlobalRank()==0) {
 	      //	      fprintf(stream,"\n*** IWaveSim: model=%s  fwd=%d deriv=%d\n",ic.iwave_model.c_str(), fwd, order);
 	      fprintf(stderr,"\n*** IWaveSim: model=%s  fwd=%d deriv=%d\n",ic.iwave_model.c_str(), fwd, order);
+	      //	    fprintf(stream,"    panel=%d in range=[%d, %d], rank=%d\n", panelindex,first,last,retrieveGlobalRank());
+	      
 	    }
-	    //	    fprintf(stream,"    panel=%d in range=[%d, %d], rank=%d\n", panelindex,first,last,retrieveGlobalRank());
 	    fprintf(stderr,"    panel=%d in range=[%d, %d], rank=%d\n", panelindex,first,last,retrieveGlobalRank());
 	  }
 

### 2024-02-15 iwave/core/lib/rdomain.cc

@@ -388,14 +388,15 @@ int rd_fwriteslice(RDOM *dom, int iarr, const char *path, int idim, int li)
 
 int rd_a_fsprint(RDOM *dom, const char *path)
 {
-    int a, err;
-    char path2[70];
+  int a, err;
+  const int len=70;
+  char path2[len];
     
     if ( strlen(path) > 60L ) return E_BADINPUT;
         
     for ( a = 0; a < dom->narr; ++a )
     {
-        if ( sprintf(path2, "%s%d", path, a) < 0 ) return E_OTHER;
+      if ( snprintf(path2, len, "%s%d", path, a) < 0 ) return E_OTHER;
             
         err = ra_fprint(dom->_s + a, path2);
         if ( err ) return err;

### 2024-02-15 iwave/explref/lib/fd_asg.cc

@@ -396,10 +396,10 @@ void asg_timestep(std::vector<RDOM *> dom,
   i_v[2] = D_V2;
     
   /* fill in pml arrays; these are no-ops after the first call */
-  register ireal * restrict ep[RARR_MAX_NDIM];
-  register ireal * restrict epp[RARR_MAX_NDIM];
-  register ireal * restrict ev[RARR_MAX_NDIM];
-  register ireal * restrict evp[RARR_MAX_NDIM];
+  ireal * restrict ep[RARR_MAX_NDIM];
+  ireal * restrict epp[RARR_MAX_NDIM];
+  ireal * restrict ev[RARR_MAX_NDIM];
+  ireal * restrict evp[RARR_MAX_NDIM];
 
   // local pml widths for this domain
   IPNT nrsloc;
@@ -459,7 +459,7 @@ void asg_timestep(std::vector<RDOM *> dom,
     for (int icoeff=0;icoeff<ncoeff;icoeff++) {
       rd_a_gse(dom[0], icoeff, gs, ge);
       if (ndim == 2) {
-	register ireal ** restrict f = (dom[0]->_s)[icoeff]._s2;
+	ireal ** restrict f = (dom[0]->_s)[icoeff]._s2;
 	// extend in dim 1
 	for (int i0=gs[0]+nlsloc[0]; i0<=ge[0]-nrsloc[0]; i0++) {
 	  for (int i1=gs[1]; i1<=gs[1]+nlsloc[1]; i1++)
@@ -476,7 +476,7 @@ void asg_timestep(std::vector<RDOM *> dom,
 	}
       }
       else if (ndim == 3) {
-	register ireal *** restrict f = (dom[0]->_s)[icoeff]._s3;
+	ireal *** restrict f = (dom[0]->_s)[icoeff]._s3;
 	// extend in dim 2
 	for (int i0=gs[0]+nlsloc[0]; i0<=ge[0]-nrsloc[0]; i0++) {
 	  for (int i1=gs[1]+nlsloc[1]; i1<=ge[1]-nrsloc[1]; i1++) {
@@ -522,13 +522,13 @@ void asg_timestep(std::vector<RDOM *> dom,
   if (n==1) {
     //    if (fwd == true) {
     if (ndim == 2) {
-      register ireal ** restrict bulk2 = (dom[0]->_s)[D_BULK]._s2;
-      register ireal ** restrict buoy2 = (dom[0]->_s)[D_BUOY]._s2;
[…]

### 2024-02-15 iwave/grid/lib/grid.cc

@@ -311,14 +311,14 @@ int get_panelnum_grid(grid g) {
   _IPNT _id;
   int i;
   int sz=1;
-  int id_check=0;
+  //  int id_check=0;
   get_n(_n,g);
   get_id(_id,g);
   for (i=0;i<g.gdim;i++) {
     // id=dim - time axis id<EXTINT external extended axis
     if ((_id[i] > g.dim-1) && (_id[i] < EXTINT)) {
       sz*=_n[i];
-      id_check++;
+      //  id_check++;
     }
   }
   return sz;

### 2024-02-15 iwave/grid/main/standardmodel.cc

@@ -606,7 +606,7 @@ int main(int argc, char **argv) {
   /* DS */
   if (model == 10){
     for(i = 0; i< 4 ; i++){
-      sprintf(numlname,"m%d_numl",i+1);
+      snprintf(numlname, 10, "m%d_numl",i+1);
       /* set up layer numbers for velocity, density, bulkmod and buoyancy */
       if (ps_ffint(*par,numlname,numl+i)) numl[i] = 1;
        
@@ -619,7 +619,7 @@ int main(int argc, char **argv) {
        
       for (j =0; j< numl[i]; j++){
 	if (j < numl[i]-1){
-	  sprintf(x1rsname,"m%d_rf%d",i+1,j+1);
+	  snprintf(x1rsname, 10, "m%d_rf%d",i+1,j+1);
 	  if(ps_fffloat(*par,x1rsname,x1rs[i]+j)) 
 	    { x1rs[i][j] = e1; 
 	      fprintf(stdout, "WARNING: x1rs[%d][%d] is set to the default value %f!\n",i,j,e1);
@@ -628,7 +628,7 @@ int main(int argc, char **argv) {
 	    fprintf(stdout,"%s=%e\n",x1rsname,*(x1rs[i]+j));
 
 	}
-	sprintf(valname,"m%d_val%d",i+1,j+1);
+	snprintf(valname, 10, "m%d_val%d",i+1,j+1);
 	if(ps_fffloat(*par,valname,values[i]+j)) 
 	  { values[i][j] = Vdefault[i]; 
 	    fprintf(stdout, "WARNING: values[%d][%d] is set to the default value %f!\n",i,j,Vdefault[i]);

### 2024-02-15 iwave/grid/main/standardmodel_elastic.cc

@@ -629,7 +629,7 @@ int main(int argc, char **argv) {
   /* DS */
   if (model == 10){
     for(i = 0; i< 4 ; i++){
-      sprintf(numlname,"m%d_numl",i+1);
+      snprintf(numlname,10,"m%d_numl",i+1);
       /* set up layer numbers for velocity, density, bulkmod and buoyancy */
       if (ps_ffint(*par,numlname,numl+i)) numl[i] = 1;
        
@@ -642,7 +642,7 @@ int main(int argc, char **argv) {
        
       for (j =0; j< numl[i]; j++){
 	if (j < numl[i]-1){
-	  sprintf(x1rsname,"m%d_rf%d",i+1,j+1);
+	  snprintf(x1rsname,10,"m%d_rf%d",i+1,j+1);
 	  if(ps_fffloat(*par,x1rsname,x1rs[i]+j)) 
 	    { x1rs[i][j] = e1; 
 	      fprintf(stderr, "WARNING: x1rs[%d][%d] is set to the default value %f!\n",i,j,e1);
@@ -651,7 +651,7 @@ int main(int argc, char **argv) {
 	    fprintf(stderr,"%s=%e\n",x1rsname,*(x1rs[i]+j));
 
 	}
-	sprintf(valname,"m%d_val%d",i+1,j+1);
+	snprintf(valname, 10, "m%d_val%d",i+1,j+1);
 	if(ps_fffloat(*par,valname,values[i]+j)) 
 	  { values[i][j] = Vdefault[i]; 
 	    fprintf(stderr, "WARNING: values[%d][%d] is set to the default value %f!\n",i,j,Vdefault[i]);

### 2024-02-15 iwave/sucore/include/par.h

@@ -253,9 +253,9 @@ float golden_bracket(float ax, float bx, float cx,
 float brent_bracket(float ax, float bx, float cx,
                 float (*f)(float), float tol, float *xmin);
 
-void linmin(float p[],float xi[],int n,float *fret, float (*func)());
+void linmin(float p[],float xi[],int n,float *fret, float (*func)(void));
 void powell_minimization(float p[], float **xi, int n,
-                float ftol,int *iter,float *fret,float (*func)());
+			 float ftol,int *iter,float *fret,float (*func)(void));
 
 /***** lincoeff -- linearized reflection coefficients */
 /* type definitions */

### 2024-02-15 iwave/trace/main/SEGYMConv.cc

@@ -238,11 +238,28 @@ int main(int argc, char ** argv) {
     while (fgettr(fpker,&trker) && (nker < mxx)) {
       nker++;
 	
-      if (0==fgettr(fpinp,&trinp)) {
+      if (0 == fgettr(fpinp,&trinp)) {
 	RVL::RVLException e;
 	e<<"ERROR: SEGYConv\n";
 	e<<"  fwd mode\n";
 	e<<"  failed to read output trace "<<nker<<"\n";
+	throw e;
+      }
+
+      // read output trace, then back up, to get header right
+      off_t tmpoutoff = ftello(fpout);
+      if (0 == fgettr(fpout,&trout)) {
+	RVL::RVLException e;
+	e<<"ERROR: SEGYConv\n";
+	e<<"  fwd mode\n";
+	e<<"  failed to read output trace "<<nker<<"\n";
+	throw e;
+      }
+      if (fseeko(fpout,tmpoutoff,SEEK_SET)) {
+	RVL::RVLException e;
+	e<<"Error: SEGYConv\n";
+	e<<"  adj mode\n";	  
+	e<<"reset failed on file "<<out<<"\n";
 	throw e;
       }
 	

### 2024-02-15 pyrvl/asg.py

@@ -293,7 +293,8 @@ class fsbop(vcl.Function):
     def __init__(self, dom, rng, \
                      buoyancy, source_p, order, sampord, nsnaps,\
                      cfl, cmin, cmax, dmin, dmax, \
-                     nl1, nr1, nl2, nr2, pmlampl, boundstest=True, noisy=False):
+                     nl1, nr1, nl2, nr2, pmlampl, \
+                     boundstest=True, noisy=False, partask=0):
                      
         try:
             if not isinstance(dom, rsfvc.Space):
@@ -341,6 +342,7 @@ class fsbop(vcl.Function):
                 self.printact = 0
             else:
                 self.printact = -1
+            self.partask = partask
             self.therest = ' order=' + str(self.order) + \
               ' sampord=' + str(self.sampord) + \
               ' nsnaps=' + str(self.nsnaps) + \
@@ -354,7 +356,8 @@ class fsbop(vcl.Function):
               ' nl2=' + str(self.nl2) + \
               ' nr2=' + str(self.nr2) + \
               ' pmlampl=' + str(self.pmlampl) + \
-              ' printact=' + str(self.printact)
+              ' printact=' + str(self.printact) + \
+              ' partask=' + str(self.partask)
 
     def getDomain(self):
         return self.dom

### 2024-02-15 pyrvl/awi.py

@@ -673,11 +673,19 @@ class mswi(vcl.ScalarJet):
         kmax (int): CG iteration limit
         rho (float): CG residual tolerance
         verbose (int): if not = 0, then print current action messages
+        archivedir (string): path to archive directory
+        pointname (string): filename root for model
+        gradname (string): filename root for gradient
+        filtername (string): filename root for filter
+        dataerrname (string): filename root for data error
+        
 =    '''
 
     def __init__(self, mod=None,
                      dom=None, sim=None, data=None,
-                     alpha=0.0, sigma=0.0, kmax=20, rho=0.01, verbose=0):
+                     alpha=0.0, sigma=0.0, kmax=20, rho=0.01, verbose=0,
+                     archivepath=None, pointname=None, gradname=None,
+                     filtername=None, dataerrname=None):
         try:
             if verbose != 0:
                 print('mswi constructor')
@@ -702,6 +710,10 @@ class mswi(vcl.ScalarJet):
                 raise Exception('adaptive filter space not provided')
             if not isinstance(dom,segyvc.Space):
                 raise Exception('adapt filter space not segyvc.Space instance')
+            # check archive path
+            if archivepath is not None:
+                if not os.path.exists(archivepath):
+                    raise Exception('given archive path = ' + archivepath + ' not valid')
             
             # store instance data
             self.dom = dom
@@ -713,6 +725,11 @@ class mswi(vcl.ScalarJet):
             self.kmax = kmax
             self.rho = rho
             self.verbose = verbose
+            self.archivepath = archivepath
+            self.pointname = pointname
+            self.gradname = gradname
+            self.filtername = filtername
+            self.dataerrname = dataerrname
             
             if self.verbose != 0:
                 print('mswi constructor: compute predicted data')
@@ -738,69 +755,86 @@ class mswi(vcl.ScalarJet):
 
     # auxiliary method - solves inner least squares problem 
     def resid(self):
-        if self.ep is None or self.u is None:
-            # convolution by predicted data
-            S = segyvc.ConvolutionOperator(dom=self.dom,
-                                                rng=self.p.space,
-                                                green=self.p.data)
-
-            # penalty op
-            A = awipensol(self.dom)
-
-            # mswi op = awi op witn no precond
-            L = awiop(S, A, self.alpha, self.sigma)
-
-            # penalized regularized adaptive filter
-            self.u = vcl.Vector(L.getDomain())
-
-            # reg rhs
[…]

### 2024-02-15 pyrvl/mswitest.py

@@ -117,7 +117,7 @@ try:
             mm.linComb(0.1*i,m)
             j = awi.mswi(dom=usp, sim=F, mod=mm, data=d, alpha=myalpha, sigma=mysigma, kmax=mykmax, rho=myrho, verbose=2)
             vals.append(j.value())
-            errs.append(j.dataerr()/dnorm)
+            errs.append(j.dataerr().norm()/dnorm)
         print('\nvals, rel errors at mtest = (1-t)*m0 + t*m')
         print('alpha = %10.4e sigma = %10.4e rho = %10.4e'
                       % (myalpha, mysigma, myrho))

### 2024-02-15 pyrvl/nptorsftest.py

@@ -4,16 +4,16 @@ import numpy as np
 try:
     # 1D basic case
     x = np.zeros([21,31])
-#    linalg.ndarraytorsf(x,'nprsf1.rsf')
+    linalg.ndarraytorsf(x,'nprsf1.rsf')
 
     # add unit
-#    linalg.ndarraytorsf(x,'nprsf2.rsf', unit='GPa')
+    linalg.ndarraytorsf(x,'nprsf2.rsf', unit='GPa')
 
     # add axis unit
-#    linalg.ndarraytorsf(x,'nprsf3.rsf', unit='GPa', units=['m'])
+    linalg.ndarraytorsf(x,'nprsf3.rsf', unit='GPa', units=['m'])
 
     # 2D basic case
-#    linalg.ndarraytorsf(x,'nprsf4.rsf', n=[21,31])
+    linalg.ndarraytorsf(x,'nprsf4.rsf', n=[21,31])
 
     # 2D decorated
     linalg.ndarraytorsf(x,'nprsf5.rsf', n=[21,31], d=[10, 10], o=[0.0, 1500], unit='GPa', units=['m', 'm'])

### 2024-02-15 pyrvl/op.py

@@ -68,7 +68,7 @@ def convop(g,w,d,adj=0):
         ' ker=' + g + \
         ' adj=1 >& crud'        
     # execute
-    #    print(cmd)
+    # print(cmd)
     ret = os.system(cmd)
     if ret != 0:
         print('command failed with return value ' + str(ret))

### 2024-02-15 pyrvl/rsfvc.py

@@ -34,7 +34,17 @@ class Space(vcl.Space):
 
     def isData(self,x):
         return ((self.filename == x) or linalg.rsfcomp(self.filename,x))
-    
+
+    def saveto(self,x,y):
+        try:
+            if not self.isData(x):
+                raise Exception('input data ' + str(x) + ' not rsf')
+            if not linalg.copy(x,y):
+                raise Exception('output data ' + str(y) + ' not legal path')
+        except Exception as ex:
+            print(ex)
+            raise Exception('called from rsfvc.Space.saveto')
+        
     # operates on data = filenames
     def raw_linComb(self,a,x,y,b=1.0):
         linalg.lincomb(a,x,y,b)

### 2024-02-15 pyrvl/segyvc.py

@@ -30,6 +30,16 @@ class Space(vcl.Space):
 
     def isData(self,x):
         return linalg.hdrcomp(self.filename,x)
+
+    def saveto(self,x,y):
+        try:
+            if not self.isData(x):
+                raise Exception('input data ' + str(x) + ' not rsf')
+            if not linalg.copy(x,y):
+                raise Exception('output data ' + str(y) + ' not legal path')
+        except Exception as ex:
+            print(ex)
+            raise Exception('called from rsfvc.Space.saveto')
     
     # operates on data = filenames
     def raw_linComb(self,a,x,y,b=1.0):

### 2024-02-15 pyrvl/vcalg.py

@@ -999,7 +999,8 @@ class btls(LineSearch):
             raise Exception('called from vcalg.btls')
 
 # lsargs - see above
-def lsopt(x, J, SD=None, LS=None, descmax=0, desceps=0.01, descverbose=0, lsargs=None, jetargs=None, ddargs=None ):
+def lsopt(x, J, SD=None, LS=None, descmax=0, desceps=0.01, descverbose=0,
+              lsargs=None, jetargs=None, ddargs=None, archargs=None ):
     '''
     Line seach optimization algorithm. Two principal components:
     - SD: SearchDir class: Update methodcomputes search direction from 
@@ -1014,9 +1015,12 @@ def lsopt(x, J, SD=None, LS=None, descmax=0, desceps=0.01, descverbose=0, lsargs
     descmax (int):      max number of descent steps
     desceps (float):    terminates if gradient falls below this proportion of initial
     descverbose (int):  verbosity flag
+
+    keyword dictionaries:
     lsargs:             line search params
     jetargs:            jet params
     ddargs:             search direction params
+    archargs:           archive params
 
     Returns:
     Jx (vcl.ScalarJet): final instance of jet class - final state of search
@@ -1061,8 +1065,10 @@ def lsopt(x, J, SD=None, LS=None, descmax=0, desceps=0.01, descverbose=0, lsargs
         
         # descent iteration
         i = 0
-        gtest = Jx.gradient().norm()
-
+        # archive
+        for k in archargs.keys():
+            Jx.archive(k,str(i),archargs[k])
+        
         if descverbose !=0:
             print('initial value = %10.4e'  % (Jx.value()))
             print('initial step  = %10.4e' % (step))
@@ -1089,6 +1095,9 @@ def lsopt(x, J, SD=None, LS=None, descmax=0, desceps=0.01, descverbose=0, lsargs
                 dr = Jx.gradient().dot(ddir)
                 # update counter
                 i += 1
+                # archive
+                for k in archargs.keys():
+                    Jx.archive(k,str(i),archargs[k])
                 
             else:
                 # bail out

### 2024-02-15 pyrvl/vcl.py

@@ -4,7 +4,12 @@ import os
 from abc import ABC, abstractmethod
 
 class Space(ABC):
-
+    '''
+    Abstract vector space class - core of VCL. Guiding principle: provides 
+    access to data objects of vectors and tools to manipulate them, especially 
+    linear combination and inner product.
+    '''
+    
     # returns a data container handle ("data") - pointer, file name, whatever -
     # consistent with the data structure of vectors in this space
     @abstractmethod
@@ -14,6 +19,11 @@ class Space(ABC):
     # returns True if data object is compatible, else False
     @abstractmethod
     def isData(self,x):
+        pass
+
+    # saves data to appropriate container at specified path.
+    # implementation is optional, so not defined as abstract.
+    def saveto(self, x, path):
         pass
     
     # a, b are scalars, x, y are data objects
@@ -112,6 +122,9 @@ class Space(ABC):
         pass
 
 class ProductSpace(Space):
+    '''
+    Decorator class: turns array of Spaces into a Space.
+    '''
 
     def __init__(self,SpaceList):
         if not isinstance(SpaceList,list):
@@ -195,6 +208,12 @@ class ProductSpace(Space):
 
 class Vector:
 
+    '''
+    Combines data object with reference to a space. The latter
+    specifies how data object is to behave to realize behaviour
+    of vector in space.
+    '''
+
     def __init__(self, sp, data=None):
         try:
             self.space = sp
@@ -227,7 +246,9 @@ class Vector:
         except Exception as ex:
             print(str(ex))
             raise Exception('called from vcl.Vector destructor')
-        
+
+    # replaces self.data with argument. Not particularly useful, should
+    # probably be deprecated.
     def link(self,x):
         try:
             if not self.space.isData(x):
@@ -236,11 +257,26 @@ class Vector:
             print(ex)
             raise Exception("called from vcl.Vector.link")
         else:
-            print('in link: x = ' + x)
+       #     print('in link: x = ' + x)
             if self.own:
                 self.space.cleanup(self.data)            
             self.data = x
             self.own = False
+
+    # copies data to appropriate container second argument serves as
+    # container handle: memory location for in-core data (encoded somehoe),
+    # path to disk location for out-of-core data, cloud location,....
+    # Container should admit treatment as data item in constructor,
+    # so that a vector can be constructed on it.
+    # Differs from link in that handle pointdd to by second arg need not
[…]

### 2024-02-15 repro/papers/awinum/data/datasim.py

@@ -15,7 +15,7 @@ import sys
 def asgsim(bulkmod, data_p, buoyancy, source_p, order=2, sampord=1, nsnaps=20,\
                cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
                nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0,
-               boundstest=True, noisy=False):
+               boundstest=True, noisy=False, partask=0):
     
     try:
     
@@ -27,7 +27,7 @@ def asgsim(bulkmod, data_p, buoyancy, source_p, order=2, sampord=1, nsnaps=20,\
 
         F = asg.fsbop(bulksp, datasp, buoyancy, source_p, \
                           order, sampord, nsnaps, cfl, cmin, cmax, dmin, dmax,\
-                          nl1, nr1, nl2, nr2, pmlampl, boundstest, noisy \
+                          nl1, nr1, nl2, nr2, pmlampl, boundstest, noisy, partask \
                           )
 
         d.copy(F(m))
@@ -38,6 +38,7 @@ def asgsim(bulkmod, data_p, buoyancy, source_p, order=2, sampord=1, nsnaps=20,\
 
 if __name__ == '__main__':
     args = ", ".join(sys.argv[1:])
+    print(args)
     cmd = 'asgsim(' + args + ')'
     exec(cmd)
 

### 2024-02-15 repro/papers/awinum/project/datasim.py

@@ -1,45 +1,0 @@
-import scenv
-import os
-import vcl
-import segyvc
-import rsfvc
-import asg
-
-import sys
-ORDER   = int(sys.argv[1])
-SAMPORD = int(sys.argv[2])
-NSNAPS  = int(sys.argv[3])
-CFL     = float(sys.argv[4])
-CMIN    = float(sys.argv[5])
-CMAX    = float(sys.argv[6])
-DMIN    = float(sys.argv[7])
-DMAX    = float(sys.argv[8])
-NL1     = int(sys.argv[9])
-NL2     = int(sys.argv[10])
-NR1     = int(sys.argv[11])
-NR2     = int(sys.argv[12])
-PMLAMPL = float(sys.argv[13])
-
-dfile = sys.argv[14]
-wfile = sys.argv[15]
-mfile = sys.argv[16]
-bfile = sys.argv[17]
-
-bulksp = rsfvc.Space(mfile)
-datasp = segyvc.Space(dfile)
-
-m = vcl.Vector(bulksp, mfile)
-d = vcl.Vector(datasp, dfile)
-
-F = asg.fsbop(dom=bulksp, rng=datasp, \
-        buoyancy=bfile, source_p=wfile, \
-        order=ORDER, sampord=SAMPORD, nsnaps=NSNAPS,\
-        cfl=CFL, cmin=CMIN, cmax=CMAX,dmin=DMIN, dmax=DMAX,\
-        nl1=NL1, nr1=NR1, nl2=NL2, nr2=NR2, pmlampl=PMLAMPL)
-
-d.copy(F(m))
-
-
-
-
-

### 2024-02-15 repro/papers/awinum/project/msmswi.py

@@ -1,158 +1,0 @@
-import os
-import sys
-
-import linalg
-import data
-import vcl
-import vcalg
-import vpm
-import segyvc
-import rsfvc
-import asg
-import awi
-import tmp
-
-class smoother(vcl.LinearOperator):
-
-    def __init__(self, dom):
-        RSFROOT = os.getenv('RSFROOT')
-        smoother = os.path.join(RSFROOT,'bin/sfsmooth')
-        self.cmd = smoother + ' rect1=10 rect2=10 repeat=2 '
-        self.dom = dom
-        if not isinstance(self.dom,rsfvc.Space):
-            raise Exception('grid smoother: provided domain not rsfvc.Space')
-
-    def getDomain(self):
-        return self.dom
-
-    def getRange(self):
-        return self.dom
-
-    def applyFwd(self,x,y):
-        try:
-            ret = os.system(self.cmd + ' < ' + x.data + ' > ' + y.data)
-            if ret != 0:
-                raise Exception('command failed: ' + self.cmd)
-        except Exception as ex:
-            print(ex)
-            raise Exception('called from smoother.applyFwd')
-
-    def applyAdj(self,x,y):
-        try:
-            ret = os.system(self.cmd + ' < ' + x.data + ' > ' + y.data)
-            if ret != 0:
-                raise Exception('command failed: ' + self.cmd)
-        except Exception as ex:
-            print(ex)
-            raise Exception('called from smoother.applyAdj')
-
-    def myNameIs():
-        print('2D Grid Smoother using sfsmooth')
-        print('rect1=10 rect2=10 repeat=2')
-                  
-try:
-
-    # arg1 = job choice: 'vals', 'grad0', 'grad', or gradtest (default)
-    # arg2 = frequency index
-    # arg3 = lens index
-    # arg4 = alpha
-    # arg5 = sigma
-    # arg6 = rho
-    
-    myalpha = 0.0001
-    mysigma = 0.00001
-    myrho = 0.0025
-
-    wsc=1.0e+3
-    mykmax=1000
-
-    NTU = 251
-    NTD = 626
-    NX  = 401
-    NZ  = 201
[…]

### 2024-02-15 repro/papers/awinum/project/mswi.py

@@ -1,78 +1,0 @@
-import os
-import sys
-
-import linalg
-import data
-import vcl
-import vcalg
-import vpm
-import segyvc
-import rsfvc
-import asg
-import awi
-
-ORDER=2
-SAMPORD=1
-NSNAPS=20
-CFL=0.5
-CMIN=1.0
-CMAX=3.0
-DMIN=0.8
-DMAX=3.0
-NL1=250
-NL2=250
-NR1=250
-NR2=250
-PMLAMPL=1.0
-
-KMAX=1000
-VERBOSE=2
-
-def mswigrad(bulk, buoy, wavelet, data, filt, grad. alpha, sigma, rho):
-    try: 
-
-        bulksp = rsfvc.Space(bulk)
-        datasp = segyvc.Space(data)
-        filtsp = segyvc.Space(filt)
-
-        m = vcl.Vector(bulksp, bulk)
-        d = vcl.Vector(datasp, data)
-        g = vcl.Vector(bulksp, grad)
-        u = vcl.Vector(filtsp, filt)
-        
-        F = asg.fsbop(dom=bulksp, rng=datasp, \
-                    buoyancy=buoy, source_p=wavelet, \
-                    order=ORDER, sampord=SAMPORD, nsnaps=NSNAPS,\
-                    cfl=CFL, cmin=CMIN, cmax=CMAX, dmin=DMIN, dmax=DMAX,\
-                    nl1=NL1, nr1=NR1, nl2=NL2, nr2=NR2, pmlampl=PMLAMPL)
-
-        j = awi.mswi(dom=filtsp, sim=F, mod=m, data=d, alpha=alpha,
-                         sigma=sigma, kmax=KMAX, rho=rho, verbose=VERBOSE)
-
-        u.copy(j.filter())
-        g.copy(j.gradient())
-
-    except Exception as ex:
-        print(ex)
-        raise Exception('called from mswigrad')
-
-try:
-        
-    # argv1 = rsf file defining model space and bulk mod
-    # argv2 = rsf file defining buoyancy
-    # argv3 = su file defining wavelet
-    # argv4 = su file defining data space and data
-    # argv5 = su file defining adaptive filter space
-    # argv6 = adaptive filter (out)
-    # argv7 = rsf file defining gradient (out)
-    # argv8 = alpha
-    # argv9 = sigma
-    # argv10 = rho
-
[…]

### 2024-02-15 rvl/seq/lib/seqfos.cc

@@ -84,7 +84,7 @@ void PolyMult::operator()(SeqDC & y, SeqDC const & x) {
   //  cerr<<"multiply by factor\n";
   //  fac.write(cerr);
 
-  int deg = 0;
+  // int deg = 0;
 //  std::list<double>::iterator iy = y.get().begin();  
 //  std::list<double>::const_iterator ix = x.get().begin();
   std::list<double>::const_iterator ifac = fac.get().begin();
@@ -96,7 +96,7 @@ void PolyMult::operator()(SeqDC & y, SeqDC const & x) {
 
     //    cerr<<"shift workspace to represent multiplication by x^"<<deg<<"\n";
     work.get().push_front(0.0);
-    deg++;
+    //deg++;
     ifac++;
   }
 }
@@ -120,7 +120,7 @@ void PolyMultAdj::operator()(SeqDC & y, SeqDC const & x) {
   //  cerr<<"multiply by factor\n";
   //  fac.write(cerr);
 
-  int deg = 0;
+  //int deg = 0;
 //  std::list<double>::iterator iy = y.get().begin();  
 //  std::list<double>::const_iterator ix = x.get().begin();
   std::list<double>::const_iterator ifac = fac.get().begin();
@@ -134,7 +134,7 @@ void PolyMultAdj::operator()(SeqDC & y, SeqDC const & x) {
     //    cerr<<"shift workspace to represent division by x^"<<deg<<"\n";
     work.get().pop_front();
     //    cerr<<"loop end\n";
-    deg++;
+    //deg++;
     ifac++;
   }
   //  cerr<<"exit PolyMultAdj\n";

### 2024-02-15 pyrvl/mfexec.py
1,2/24 amendments

@@ -1,0 +1,41 @@
+#!/usr/bin/env python3
+
+import sys
+
+if __name__ == '__main__':   
+    try:
+
+
+        #        print('check enough args')
+        if len(sys.argv) < 2:
+            raise Exception('must be called with at least one argument ' + \
+                                '= name of function')
+
+        arga = sys.argv[1].split('.')
+#        print(arga)
+        if len(arga) < 2:
+            raise Exception('first arg must have form mod.fcn')
+
+#        print('arga[0] = ' + arga[0])
+
+        exec('import ' + arga[0])
+
+#        argl = ', '.join(sys.argv[2:])
+
+#        cmd = sys.argv[1] + '(' + argl + ')'
+            
+#        print('function call:')
+#        print(cmd)
+
+        print('sys.argv[1]')
+        print(sys.argv[1])
+
+        exec(str(sys.argv[1]))
+        
+    except Exception as ex:
+        print(ex)
+        sys.exit(1)
+        
+        
+
+        

### 2024-02-15 pyrvl/pyrvl.py

@@ -1,0 +1,52 @@
+#!/usr/bin/env python3
+
+import data
+import sys
+import os
+import data
+
+fkeys = {    'data.bpfilt': ['file','nt','dt','s','f1','f2','f3','f4','sx','sz'],
+             'data.bpfiltgather': ['file','nt','dt','s','f1','f2','f3','f4','ntr','sxstart','szstart','dsx','dsz'],
+             'data.delta': ['file','nt','dt','sx','sz'],
+             'data.deltagather': ['file','nt','dt','ntr','sxstart','szstart','dsx','dsz'],
+             'data.rechdr': ['file','nt','dt','ntr','rx','rz','sx','sz','drx','delrt','nshot','dsx','fixed'],
+             'data.rsffile': ['file', 'datatype', 'unit', 'n1', 'n2', 'd1', 'd2', 'val'],
+             'data.model': ['bulkfile', 'bulk', 'nx', 'nz', 'dx', 'dz', 'lensfac', 'buoy', 'lensradd', 'lensradt'],
+        }
+
+if __name__ == '__main__':
+    try:
+
+        if len(sys.argv) < 2:
+            raise Exception('must be called with at least one argument ' + \
+                                '= name of function')
+        fcn = sys.argv[1]
+        if not fcn in fkeys:
+            raise Exception('function ' + fcn + ' not accessible')
+        cmd = fcn + '('
+
+        arglist = fkeys[fcn]
+
+        args = dict(arg.split('=') for arg in sys.argv[2:])
+        for k in range(len(fkeys[fcn])):
+            if not fkeys[fcn][k] in args:
+                raise Exception('argument ' + fkeys[fcn][k] + ' for function ' + fcn + ' not provided')
+
+            if len(fkeys[fcn][k]) == 1:
+                cmd + fkeys[fcn][k][0]
+        if fcn=='data.bpfilt':
+            data.bpfilt(args['file'],
+                       int(args['nt']),
+                       float(args['dt']),
+                       float(args['s']),
+                       float(args['f1']),
+                       float(args['f2']),
+                       float(args['f3']),
+                       float(args['f4']),
+                       float(args['sx']),
+                       float(args['sz']))
+        if fcm=='data.bpfiltgather':
+            
+        
+    except Exception as ex:
+        print(ex)

### 2024-02-16 pyrvl/asg.py
enabled mpi

@@ -356,8 +356,7 @@ class fsbop(vcl.Function):
               ' nl2=' + str(self.nl2) + \
               ' nr2=' + str(self.nr2) + \
               ' pmlampl=' + str(self.pmlampl) + \
-              ' printact=' + str(self.printact) + \
-              ' partask=' + str(self.partask)
+              ' printact=' + str(self.printact)
 
     def getDomain(self):
         return self.dom
@@ -381,13 +380,33 @@ class fsbop(vcl.Function):
             
             #os.system('ls /var/tmp')
             TRIP = os.getenv('TRIP')
-            cmd = os.path.join(TRIP,'iwave/asg/main/sim.x')
+            if not isinstance(TRIP, str):
+                raise Exception('TRIP not defined in environment')
+            if not os.path.exists(TRIP):
+                raise Exception('TRIP not defined as legal path')
+            SIM = os.path.join(TRIP,'iwave/asg/main/sim.x')
+            if not os.path.exists(SIM):
+                raise Exception(SIM + ' not found')
+            if self.partask > 1:
+                MPIROOT = os.getenv('MPIROOT')
+                if not isinstance(MPIROOT, str):
+                    raise Exception('MPIROOT not defined in environment')
+                if not os.path.exists(MPIROOT):
+                    raise Exception('MPIROOT not defined as legal path')
+                MPIRUN = os.path.join(MPIROOT,'bin/mpirun')
+                if not os.path.exists(MPIRUN):
+                    raise Exception(MPIRUN + ' not found')
+                cmd = MPIRUN + ' -np=' + str(self.partask) + ' ' + SIM
+            else:
+                cmd = SIM
             args = ' bulkmod=' + x.data + \
                    ' buoyancy=' + self.buoyancy + \
                    ' source_p=' + self.source_p + \
                    ' data_p=' + y.data + \
-                   ' deriv=0 adjoint=0' + self.therest
-                   
+                   ' deriv=0 adjoint=0' + self.therest + \
+                   ' partask=' + str(self.partask)
+
+            print(cmd + args)
             ret = os.system(cmd + args)
             if ret != 0:
                 raise Exception('Error: return ' + str(ret) + ' from sim.x')
@@ -400,7 +419,7 @@ class fsbop(vcl.Function):
         try:
             return fsbderiv(self.dom,self.rng,x,self.buoyancy,self.source_p,
                                 self.cmin, self.cmax, self.dmin, self.dmax,
[…]

### 2024-02-17 pyrvl/asg.py
fixed mpirun call

@@ -396,7 +396,7 @@ class fsbop(vcl.Function):
                 MPIRUN = os.path.join(MPIROOT,'bin/mpirun')
                 if not os.path.exists(MPIRUN):
                     raise Exception(MPIRUN + ' not found')
-                cmd = MPIRUN + ' -np=' + str(self.partask) + ' ' + SIM
+                cmd = MPIRUN + ' -np ' + str(self.partask) + ' ' + SIM
             else:
                 cmd = SIM
             args = ' bulkmod=' + x.data + \
@@ -488,7 +488,7 @@ class fsbderiv(vcl.LinearOperator):
                 MPIRUN = os.path.join(MPIROOT,'bin/mpirun')
                 if not os.path.exists(MPIRUN):
                     raise Exception(MPIRUN + ' not found')
-                cmd = MPIRUN + ' -np=' + str(self.partask) + ' ' + SIM
+                cmd = MPIRUN + ' -np ' + str(self.partask) + ' ' + SIM
             else:
                 cmd = SIM
                 
@@ -545,7 +545,7 @@ class fsbderiv(vcl.LinearOperator):
                 MPIRUN = os.path.join(MPIROOT,'bin/mpirun')
                 if not os.path.exists(MPIRUN):
                     raise Exception(MPIRUN + ' not found')
-                cmd = MPIRUN + ' -np=' + str(self.partask) + ' ' + SIM
+                cmd = MPIRUN + ' -np ' + str(self.partask) + ' ' + SIM
             else:
                 cmd = SIM
                 

### 2024-02-27 repro/papers/awinum/paper.tex
much

\title{Matched Source Waveform Inversion for Transmitted Waves}
\author{Huiyi Chen, Susan E. Minkoff, and William W. Symes}

\lefthead{Symes}

\righthead{MSWI}

\maketitle
\parskip 12pt

\begin{abstract}
Matched Source Waveform Inversion applied to transmission data
produces an estimate of refractive index similar to the result of
travel time inversion, but without explicit identification of travel
times. This paper reviews the theoretical justification of this result
and its limitations, and exhibits a numerical illustration. 
\end{abstract}
\setlength{\parindent}{0cm}

\section{Introduction}
Matched Source Waveform Inversion (MSWI) is a variant of Full Waveform
Inversion (FWI) \cite[]{VirieuxOperto:09}, that sometimes overcomes
one of FWI's impediments, namely its tendency to stagnate at
suboptimal model estimates.

(...)
,
demonstrably far from global optima in controlled settings and often
uninformative of material structure in field use. It is possible that
such local descent algorithms are trapped in regions around local,
\section{Summary}
The version of MSWI discussed here uses acoustic wave propagation with
isotropic point sources and receivers.
The pressure and velocity fields $p({\bf x},t;{\bf x}_s)$, ${\bf v}({\bf x},t;{\bf x}_s)$ for the source location ${\bf x}_s$ depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf x})$ (reciprocal of the density $\rho({\bf x})$), and wavelet $w(t;{\bf x}_s)$ through the acoustic system
\begin{eqnarray}
  \label{eqn:awe}
 \frac{\partial p}{\partial t} & = &- \kappa \nabla \cdot {\bf v} +
w(t;{\bf x}_s) \delta({\bf x}-{\bf x}_s); \nonumber \\
\frac{\partial {\bf v}}{\partial t} & = & - \beta \nabla p; \\ 
p, {\bf v} & = & 0 \mbox{ for }  t \ll 0.
\end{eqnarray}
The model vectors $m=(\kappa,\rho)$ make up the domain of the forward
map or {\em modeling operator} is $F[m]w = \{p({\bf x}_r,t;{\bf
  x}_s)\}$, for specified source and receiver positions ${\bf x}_s, {\bf x}_r$ and
recording interval $[0,t_d]$.

In this context, FWI means: given source wavelet
$w(t;\bx_s)$ and data traces $d(\bx_r,\cdot;\bx_s)$, find a model $m$
so that $F[m]w \approx d$. The
simplest version of FWI concretizes this task by asking for a model
$m$ minimizing the mean square error
\begin{equation}
  \label{eqn:fwi}
  J_{\rm FWI}[m;d]= \frac{1}{2}\|F[m]w-d\|^2.
\end{equation}
Note that $w$ is assumed known and treated as a parameter in this
statement of the FWI task. In practice, it is not known (nor are the
sources and receviers necessarily isotropic), and should be estimated
along with the model $m$.

The approach to local optimization taken here (and in most work on FWI
and related topics) is based on the gradient of the objective defined
in equiation \ref{eqn:fwi}:
\begin{equation}
  \label{eqn:fwigrad}
  g = \nabla  J_{\rm FWI}[m;d] = DF[m]^T(F[m]w-d).
\end{equation}
In this formula, $DF[m]$ is the derivative of $F$ at $m$. 
This is the Euclidean (or $L^2$) gradient, that is, the vector $g$ for
which the Euclidean inner product
\begin{equation}
  \label{eqn:eucip}
  \langle g, \delta m\rangle = g^T\delta m
\end{equation}
with any other vector $\delta m$ gives the
rate of change in the direction of that vector of $J_{\rm FWI}$ at m.

For optimization within a set of slowly varying models (on the
wavelength scale),  it is appropriate to penalize oscillation of the
search vector. A convenient way to accompish this goal is the use of a
{\em weighted inner product} to define the gradient, rather than the
Euclidean inner product. A weight operator $W$ should be symmetric and
positive definite: then
\begin{equation}
  \label{eqn:wip}
  \langle g, \delta m\rangle_W = g^TW\delta m
\end{equation}
defines an alternative inner product. Comparing the definitions
\ref{eqn:fwigrad}, \ref{eqn:eucip}, and \ref{eqn:wip}, clearly the
vector $g_W$ for which $\langle g_W, \delta m \rangle_W$ gives the
rate of change of $J_{\rm FWI}$ at $m$ in the direction $\delta m$ is
\begin{equation}
  \label{eqn:fwiwgrad}
  g_W = W^{-1}\nabla  J_{\rm FWI}[m;d] =W^{-1} DF[m]^T(F[m]w-d).
\end{equation}

If $W$ is chosen to greatly
amplify oscillatory components of the vector to which it is applied,
then those components of $g_W$ must be suppressed relative to the
corresponding components of $g$, hence $g_W$ represents a
non-oscillatory search direction. Note that only the inverse operator
$W^{-1}$ appears in the formula \ref{eqn:fwiwgrad}.

As mentioned earlier, application of local optimization methods
directly to $J_{\rm FWI}$ tends to produce unsatisfactory model
estimates. MSWI modifies the measure of distance between predicted and
observed data by inserting an adaptive filter field $u$, consisting of
one filter per trace. Since only finite time intervals of $u$ and $d$ are available in
practice, introduce the {\em truncated filter operator} $K[u]$. This
operator acts by extending its filter $u$ (given on a symmetric
interval $[-t_u,t_u]$ for each source-receiver pair) and the function to which
it is applied (given on the data interval $[0,t_d]$ for each
source-receiver pair) to be zero outside their domains of definition, convolving the
resulting functions on $\bR$, and finally restricting or truncating
the result to the time interval of the input function, that is,
$[0,t_d]$. This operator is applied to the predicted data $F[m]w$ to
produce the filtered predicted data $K[u]F[m]w$.

It is possible to make the error between filtered predicted data and
observed data as small as one likes by choosing an appropriate filter
field $u$, so this error by itself is useless for estimating the
model. If $u(\bx_s,\bx_r,t)=\delta(t)$, on the other hand the filtered
predicted data is identical to the predicted data.
Therefore, some penalty for divergence of the filter $u$ from
$\delta(t)$ needs to supplement the filtered prediction error.. The works referenced in the Introduction mostly use the
mean-square of the filter scaled by $t$, and add it to the mean-square
of the filtered prediction error. This sum is the MSWI objective function:
\begin{equation}
  \label{eqn:filtpen}
  J_{\alpha,\sigma}[m,u;d]=\frac{1}{2}(\|K[u]F[m]w-d\|^2
  +\alpha^2\|tu\|^2 + \sigma^2\|u\|^2).
\end{equation}

\section{A numerical illustration}

\inputdir{project}

This section presents the inversion of acoustic data with dimensions
typical of crustal seismic exploration. The target bulk modulus field ($\kappa$)
is depicted in Figure \ref{fig:m}, and contains an acoustic ``lens''
positioned in the center between 1000 m and 3000 m depth
(``$z$''). The background level outside the lens is 4 GPa; at the
center it is 2 GPa. The
buoyancy ($\beta$)is spatially homogeneous at 1 cm$^3/g$, so the background
wave speed is 2000 m/s.

\plot{m}{width=\textwidth}{Lens model.}

This field is sampled on a 20 $\times$ 20 m grid. Acoustic wave
propagation is simulated via a staggered grid finite difference method
\cite[]{vir86,lev88,Cohen:01}. The time
step is chosen to be safely stable, given the parameter fields,
difference formulae, and
spatial sampling.

Receivers are spaced on the line at depth 1000 m, from 2000 m to 6000
m horizontal coordinate (``$x$''), spaced 20 m apart. Eleven source
positions lie on the line $z=3000$ m, spaced 200 m apart, from
$x=3000$ m to $x = 5000$ m. The source wavelet $w$ (the same for every
source ) is a $[1.0, 2.5, 7.5, 12.5]$ Hz trapezoidal bandpass filter,
with a median frequency of 5.875 Hz corresponding to a median wavelength of
$\approx$ 340 m, and shortest wavelength of 160 m. It is centered at
$t=1$ s.

The discrete pressure field is
sampled at the externally specified time grid and spatial locations
via piecewise linear interpolation in space and cubic spline
interpolation in time. The isotropic point source is added
into the acoustic fields at each time step via the adjoints of these
interpolation operations.

Simulated data from this configuration appears as Figure \ref{fig:d11}.

The initial model $m_0$ for inversion is homogeneous, with $\kappa = $ 4 GPa,
$\beta = $ 1 cm$^3$/g. The corresponding data is depicted in Figure \ref{fig:sim0}.
These plots show the time shift between the homogeneous model data and
the target data, for the central source positions.

\multiplot{2}{d11,sim0}{width=0.45\textwidth}{a: Data for target model $m$
  (Figure \ref{fig:m}). b: Data for homogeneous model $m_0$.}

We apply a version of weighted steepest descent optimization to the
minimization of $J_{\rm FWI}$. The inverse weight operator ($W^{-1}$
in formula \ref{eqn:fwiwgrad}) is a
10-point moving average in both spatial directions, repeated once. As
noted above, the weight operator itself is not required. The weighted
gradient is computed via formula \ref{eqn:fwigrad}. The adjoint
derivative $DF[m]^T$ is computed via the {\em adjoint state method}
\cite[]{Chavent:74,GauTarVir:86}, with time reversal of the acoustic fields
implemented via optimal checkpointing
\cite[]{Griewank:92,Griewank:book,Symes:06a-pub}. The negative of the weighted
gradient ($g_W$, formula \ref{eqn:fwiwgrad}) is the search direction,
and the optimal step in this direction is approximated by a simple
backtracking line search algorithm \cite[]{NocedalWright}.

The value of $J_{\rm FWI}[m_0,d] \approx 3.04$, and the rate of
increase in the weighted gradient direction at $m_0$ is 1.74 $\times
10^{-6}$. Note that the rate of increase is the same as the weighted
gradient norm. After 10 steepest descent steps, the objective value has
decreased by about 30\%, to 1.9, and the weighted gradient norm by
more than an order of magnitude, to $\approx 9.6 \times 10^{-8}$. The
final model appears in Figure \ref{fig:mestfwi0}, and the
corresponding data in Figure \ref{fig:resimfwi0}.

%\multiplot{2}{mestfwi0,resimfwi0}{width=0.45\textwidth}{a: Bulk
%  modulus produced by 10 steepest descent steps to minimize the FWI
%  objective $J_{\rm FWI}[\cdot;d]$ with data $d$ depicted in Figure
%  \ref{fig:sim11}, starting with homogeneous model. b: Data simulated
%  from FWI inversion result.}

\plot{mestfwi0}{width=\textwidth}{a: Bulk modulus produced by 10
  steepest descent steps to minimize the FWI objective \ref{eqn:fwi},
  using the data shown in \ref{fig:d11}.}

\multiplot{2}{resimfwi0,residfwi0}{width=0.45\textwidth}{a: Data
  corresponding to FWI inversion result shown in Figure
  \ref{fig:mestfwi0}. b: Residual or data error - difference between
  data shown in Figure \ref{fig:resimfwi0} and target data
  \ref{fig:d11}. Note the failure to match the later signal in the
  central part of the display.}

While the reduction in the (weighted) gradient norm indicates progress
towards a stationary point, it is not possible to claim that the final
estimate is in the vicinity of a local minimizer. The fit to data
obtained is not at all satisfactory. More discussion of this result
will be found later in the paper.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2024-02-27 repro/papers/awinum/paper_wg.tex
much

\title{AWI-WG Examples}
\author{Huiyi Chen, Susan E. Minkoff, and William W. Symes}

\lefthead{Symes}

\righthead{AWI-WG Examples}

\maketitle
\parskip 12pt

\begin{abstract}
abstract goes here
\end{abstract}
\setlength{\parindent}{0cm}

\section{Summary}
Scans of the WG objective function over the line segment between a homogeneous slowness model $m_0$ and the target model $m$  (the model used to generate the data) show smooth decrease towards a minimum. However the minimizer is often not located at the point on the line segment corresponding to the target model. The line segment is
$$ t \mapsto (1-t)m_0 + t m $$
so $t=1$ is the target.  Depending on exactly which choices of regularization parameter $\sigma$ and normal residual tolerance $\rho$ are used in the computation of the objective function, the minimizer can be close to $t=1$ or in error by more than 10\%, that is, $t > 1.1$.

This error represents a limitation of the resolution obtainable from the AWI-WG objective as a means of estimating slowness. For examples in which the starting model is farther from the target model, the error appears to be proportionally smaller. Since the argument connecting AWI-WG to travel time tomography depends on geometric asymptotic theory, which is approximate, this result is not surprising. It also suggests that the penalty version of AWI, which incorporates a controlled component of the FWI objective, may offer better resolution than the AWI-WG.

\section{Results}

\inputdir{project}

The first example uses for $m$ the ``lens'' model that we have used in many other tests, generated by
\begin{verbatim}
data.model(bulkfile='m0l.rsf', bulk=4.0, nx=401, nz=201,
   dx=20, dz=20, lensfac=0.7)
\end{verbatim}
and plotted in Figure \ref{fig:m0l}:
\plot{m0l}{width=\textwidth}{Lens model, lensfac = 0.7, high aspect ratio'}

The data for these two models  ($m$ and $m_0$) differs mostly by a half-wavelength phase shift: the waveform apparent in both gathers is similar to the first derivative of a gaussian, whereas the waveform of the difference is closer to a second derivative.

\multiplot{2}{d0l,dlhom}{width=0.45\textwidth}{Left: data for target model $m$; Right: data for homogeneous model $m_0$.}
\plot{d0ldiff}{width=0.45\textwidth}{DIfference between gathers plotted in Figures \ref{fig:d0l} and \ref{fig:dlhom}}

The following two tables show the values of the AWI-WG objective function sampled at $\Delta t = 0.1$, for $\sigma=0.001$ and three different (decreasing) choices of $\rho$, resulting in increasingly accurate solution of the normal equations and decreasing error in the data fit by the adaptive filter.

\begin{table}
  \begin{center}
  \begin{tabular}{c|c|c}
    \hline
    t & v & e \\
0.0 &  1.1593e+06 & 2.9150e-02 \\ 
0.1 & 1.0909e+06 & 2.9959e-02 \\
0.2 & 1.0149e+06 & 3.1390e-02 \\
0.3 & 9.7392e+05 & 2.9837e-02 \\
0.4 & 9.2755e+05 & 2.9033e-02 \\
0.5 & 8.5273e+05 & 3.1508e-02 \\
0.6 & 8.2288e+05 & 2.8947e-02 \\
0.7 & 7.8012e+05 & 2.9248e-02 \\
0.8 & 7.5644e+05 & 2.7974e-02 \\
0.9 & 7.2089e+05 & 2.9301e-02 \\
1.0 & 7.0420e+05 & 2.9180e-02 \\
1.1 & 6.8927e+05 & 2.9798e-02 \\
    1.2 & 7.0310e+05 & 2.9222e-02 \\
    \hline 
  \end{tabular}

  \caption{Values (v) of AWI-WG objective and adaptive filter error (e)  based on data shown in Figure \ref{fig:d0l} at $m_{\rm test} = (1-t)m_0 + t m$, computed with $\sigma=0.001, \rho = 0.01$.}
    \label{table:rho01}
  \end{center}
  \end{table}

  \begin{table}
  \begin{center}
  \begin{tabular}{c|c|c}
    \hline
    t & v & e \\
0.0 & 1.2419e+06 & 6.7093e-03 \\
0.1 & 1.1755e+06 & 6.4195e-03 \\
0.2 & 1.1334e+06 & 6.7615e-03 \\
0.3 & 1.1132e+06 & 7.3999e-03 \\
0.4 & 1.0664e+06 & 7.4529e-03 \\
0.5 & 1.0245e+06 & 7.3856e-03 \\
0.6 & 9.9281e+05 & 7.4972e-03 \\
0.7 & 9.4065e+05 & 7.5612e-03 \\
0.8 & 7.6815e+05 & 5.5490e-03 \\
0.9 & 7.5830e+05 & 5.5697e-03 \\
1.0 & 7.5452e+05 & 6.0200e-03 \\
1.1 & 7.6785e+05 & 6.4036e-03 \\
1.2 & 8.1194e+05 & 6.8795e-03 \\
    \hline 
  \end{tabular}
   \caption{Values (v) of AWI-WG objective and adaptive filter error (e) based on data shown in Figure \ref{fig:d0l} at $m_{\rm test} = (1-t)m_0 + t m$, computed with $\sigma=0.001, \rho = 0.001$.}
  \label{table:rho001}    
  \end{center}
\end{table}

\begin{table}
  \begin{center}
  \begin{tabular}{c|c|c}
    \hline
    t & v & e \\
0.0 & 1.4052e+06 & 1.4151e-03 \\
0.1 & 1.3288e+06 & 1.3979e-03 \\
0.2 & 1.2528e+06 & 1.4231e-03 \\
0.3 & 1.1850e+06 & 1.4625e-03 \\
0.4 & 1.1325e+06 & 1.4531e-03 \\
0.5 & 1.0860e+06 & 1.4492e-03 \\
0.6 & 1.0351e+06 & 1.4641e-03 \\
0.7 & 1.0033e+06 & 1.3976e-03 \\
0.8 & 9.9407e+05 & 1.3310e-03 \\
0.9 & 9.8442e+05 & 1.3020e-03 \\
1.0 & 9.6528e+05 & 1.3018e-03 \\
1.1 & 9.4995e+05 & 1.3066e-03 \\
    1.2 & 9.3124e+05 & 1.3467e-03 \\
        \hline 
  \end{tabular}
   \caption{Values (v) of AWI-WG objective and adaptive filter error (e) based on data shown in Figure \ref{fig:d0l} at $m_{\rm test} = (1-t)m_0 + t m$, computed with $\sigma=0.001, \rho = 0.0001$.}
    \label{table:rho0001}
  \end{center}
\end{table}

As $\rho$ decreases by two orders of magnitude, the fit error decreases from about 3 \% ($\rho=0.01$), to 0.7\% ($\rho=0.001$), to 0.15\% $\rho=0.0001$. The last result is probably near-optimal: it requires several hundred CG iterations values of $t$ near 1.0. the positive regularization weight bounds the lowest error achievable away from zero.

Note that the first scan (Table \ref{table:rho01} with $\rho=0.01$) has its minimum sample at $t=1.1$. Assuming that the objective is monotone increasing (decreasing) for $t > 1.2$ (respectively $t < 1.0$), the most one can conclude is that the minimizer on this line segment lies in the open interval $(1.0, 1.2)$. Quadratic fit suggests that it may be close to $t=1.1$. However the precise location of the line minimizer is not the point: that is, rather, that the minimum in any set containing this line segment is not at $t=1.0$, that is not at $m_{\rm test} = m$, the target model.

One might think that the normal equation has not been solved precisely enough to approximate the actual value of the AWI-WG objective. The scan with $\rho=0.001$ (Table \ref{table:rho001}) supports this supposition: the data is fit more accurately by the constructed adaptive filters, and the minimum sample occurs at $t=1.0$ (which of course means only that the line minimum is likely between $0.9$ and $1.1$). However this conclusion would be incorrect: Table \ref{table:rho0001} with an even more accurate solution of the normal equation ($\rho=0.0001$) now shows a minimum at the maximum $t = 1.2$, from which one can only conclude that the line minimizer is $> 1.1$.

These examples lead to the conclusion that the resolution of the AWI-WG slowness estimate is strictly limited: in this example, the error relative to the overall slowness error at the initial estimate ($m_0$) is at least 10\%.

Looked at another way, the difference between $m$ and $m_0$ (with the other acquisition choices and modeling choices made in creating the data and computing the objective) are to small for AWI-WG to detect accurately. To test this hypothesis, I introduce another lens model $m_1$, differing from the previous model in having a fatter and deeper lens:

\plot{m1l}{width=\textwidth}{Lens model $m_1$, lensfac = 0.5, low aspect ratio'}

The command to generate this model is
\begin{verbatim}
data.model(bulkfile='m1.rsf', bulk=4.0, nx=401, nz=201,
   dx=20, dz=20, lensfac=0.5, lensradd=0.4)
\end{verbatim}

(Note the new argument {\tt lensradd} - using this command requires the latest version of the {\tt data} module.)

The time shift between data built with the homogeneous model $m_0$ and the corresponding data built with $m_1$ is considerably larger:

\multiplot{2}{d1l,d1ldiff}{width=0.45\textwidth}{Left: data for target model $m_1$ (Figure \ref{fig:m1l}); Right: difference with data for homogeneous model $m_0$.}

The next three tables show the same experiments as the first three, but the samples are taken from the line segment $t \mapsto (1-t)m_0 + t m_1, t \in [0.0,1.2]$. The choices of $\sigma$ and $\rho$ are the same.

\begin{table}
  \begin{center}
  \begin{tabular}{c|c|c}
    \hline
    t & v & e \\
0.0 & 3.3591e+06 & 2.8394e-02 \\
0.1 & 2.9599e+06 & 2.8493e-02 \\
0.2 & 2.5653e+06 & 2.9214e-02 \\
0.3 & 2.1924e+06 & 3.0009e-02 \\
0.4 & 1.8761e+06 & 2.7319e-02 \\
0.5 & 1.5692e+06 & 2.8048e-02 \\
0.6 & 1.2820e+06 & 2.9795e-02 \\
0.7 & 1.0505e+06 & 2.9463e-02 \\
0.8 & 8.5406e+05 & 2.9257e-02 \\
0.9 & 7.1680e+05 & 3.0119e-02 \\
1.0 & 6.6482e+05 & 2.7938e-02 \\
1.1 & 6.8210e+05 & 2.8238e-02 \\
1.2 & 7.8490e+05 & 2.8498e-02 \\
        \hline 
  \end{tabular}
   \caption{Values (v) of AWI-WG objective and adaptive filter error (e) based on data shown in Figure \ref{fig:d0l} at $m_{\rm test} = (1-t)m_0 + t m_1$, computed with $\sigma=0.001, \rho = 0.01$.}
    \label{table:m1rho01}
  \end{center}
\end{table}
\begin{table}
  \begin{center}
  \begin{tabular}{c|c|c}
    \hline
    t & v & e \\
    0.0 & 3.5153e+06 & 5.9678e-03 \\
0.1 & 3.1136e+06 & 6.1703e-03 \\
0.2 & 2.7146e+06 & 5.8284e-03 \\
0.3 & 2.3423e+06 & 5.9218e-03 \\
0.4 & 2.0156e+06 & 6.2422e-03 \\
0.5 & 1.7010e+06 & 6.1459e-03 \\
0.6 & 1.4255e+06 & 6.3031e-03 \\
0.7 & 1.2071e+06 & 6.4979e-03 \\
0.8 & 1.0679e+06 & 6.8024e-03 \\
0.9 & 9.3836e+05 & 7.2389e-03 \\
1.0 & 7.8422e+05 & 6.3987e-03 \\
1.1 & 8.8863e+05 & 7.0049e-03 \\
1.2 & 1.0187e+06 & 6.3996e-03 \\
        \hline 
  \end{tabular}
   \caption{Values (v) of AWI-WG objective and adaptive filter error (e) based on data shown in Figure \ref{fig:d0l} at $m_{\rm test} = (1-t)m_0 + t m_1$, computed with $\sigma=0.001, \rho = 0.001$.}
    \label{table:m1rho001}
  \end{center}
\end{table}

\begin{table}
  \begin{center}
  \begin{tabular}{c|c|c}
    \hline
    t & v & e \\
    0.0 & 3.5287e+06 & 1.6961e-03 \\
0.1 & 3.0967e+06 & 1.3654e-03 \\
0.2 & 2.7877e+06 & 1.6737e-03 \\
0.3 & 2.3827e+06 & 1.4092e-03 \\
0.4 & 2.0719e+06 & 1.5689e-03 \\
0.5 & 1.7617e+06 & 1.3423e-03 \\
0.6 & 1.4830e+06 & 1.3776e-03 \\
0.7 & 1.2629e+06 & 1.3850e-03 \\
0.8 & 1.0936e+06 & 1.3775e-03 \\
0.9 & 9.4892e+05 & 1.4197e-03 \\
1.0 & 9.0077e+05 & 1.4474e-03 \\
1.1 & 8.9294e+05 & 1.3807e-03 \\
1.2 & 1.0196e+06 & 1.3239e-03 \\
        \hline 
  \end{tabular}
   \caption{Values (v) of AWI-WG objective and adaptive filter error (e) based on data shown in Figure \ref{fig:d0l} at $m_{\rm test} = (1-t)m_0 + t m_1$, computed with $\sigma=0.001, \rho = 0.0001$.}
    \label{table:m1rho0001}
  \end{center}
\end{table}

Qualitatively, these results are similar to those for the closer-to-homogeneous target model $m$. However, this time the scan for gradient tolerance $\rho=0.001$ still has the minimum sample at $t=1.0$, and the scan for $\rho=0.0001$ has its minimum sample at $t=1.1$ rather than $t=1.2$. That is, the error in the location of the minimum sample - or, the ``solution'' of the inverse problem via AWI-WG, if the search is limited to this line segment - is closer, as a proportion of the difference with the starting model, than was the case in the previous example.

These examples and other similar examples suggest that the observed error is a resolution error stemming from the difference between AWI-WG and travel time tomography: the relation is based on geometric asymptotics, which provides approximate solutions of the wave equation.

### 2024-02-27 repro/papers/awinum/mswi/datasim.py

@@ -1,0 +1,44 @@
+import scenv
+import os
+import vcl
+import segyvc
+import rsfvc
+import asg
+
+import sys
+
+# order=2, sampord=1, nsnaps=20,\
+#                cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
+#                nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0,
+#                boundstest=True, noisy=False
+
+def asgsim(bulkmod, data_p, buoyancy, source_p, order=2, sampord=1, nsnaps=20,\
+               cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
+               nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0,
+               boundstest=True, noisy=False, partask=0):
+    
+    try:
+    
+        bulksp = rsfvc.Space(bulkmod)
+        datasp = segyvc.Space(data_p)
+
+        m = vcl.Vector(bulksp, bulkmod)
+        d = vcl.Vector(datasp, data_p)
+
+        F = asg.fsbop(bulksp, datasp, buoyancy, source_p, \
+                          order, sampord, nsnaps, cfl, cmin, cmax, dmin, dmax,\
+                          nl1, nr1, nl2, nr2, pmlampl, boundstest, noisy, partask \
+                          )
+
+        d.copy(F(m))
+        
+    except Exception as ex:
+        print(ex)
+        raise Exception('called from sim')
+
+if __name__ == '__main__':
+    args = ", ".join(sys.argv[1:])
+    cmd = 'asgsim(' + args + ')'
+    exec(cmd)
+
+

### 2024-02-27 repro/papers/awinum/mswi/fwi.py

@@ -1,0 +1,100 @@
+import os
+import sys
+import scenv
+
+import linalg
+import data
+import vcl
+import vcalg
+import vpm
+import segyvc
+import rsfvc
+import asg
+import awi
+import tmp
+
+class smoother(vcl.LinearOperator):
+
+    def __init__(self, dom, rect1=10, rect2=10, repeat=2):
+        RSFROOT = os.getenv('RSFROOT')
+        smoother = os.path.join(RSFROOT,'bin/sfsmooth')
+        self.dom = dom
+        if not isinstance(self.dom,rsfvc.Space):
+            raise Exception('grid smoother: provided domain not rsfvc.Space')
+        self.cmd = smoother + ' rect1=' + str(rect1) + ' rect2=' + str(rect2) + ' repeat=' +str(repeat)
+        
+    def getDomain(self):
+        return self.dom
+
+    def getRange(self):
+        return self.dom
+
+    def applyFwd(self,x,y):
+        try:
+            ret = os.system(self.cmd + ' < ' + x.data + ' > ' + y.data)
+            if ret != 0:
+                raise Exception('command failed: ' + self.cmd)
+        except Exception as ex:
+            print(ex)
+            raise Exception('called from smoother.applyFwd')
+
+    def applyAdj(self,x,y):
+        try:
+            ret = os.system(self.cmd + ' < ' + x.data + ' > ' + y.data)
+            if ret != 0:
+                raise Exception('command failed: ' + self.cmd)
+        except Exception as ex:
+            print(ex)
+            raise Exception('called from smoother.applyAdj')
+
+    def myNameIs():
+        print('2D Grid Smoother using sfsmooth')
+        print('rect1=10 rect2=10 repeat=2')
+
+def fwiopt(bulkmodout, bulkmodin, datain, buoyancy, source, 
+                order=2, sampord=1, nsnaps=20,\
+                cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
+                nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0,
+                boundstest=True, noisy=False,
+                rect1=10, rect2=10, repeat=2,
+                descmax=12, desceps=0.01, descverbose=1, descout=None,
[…]

### 2024-02-27 repro/papers/awinum/mswi/msmswi.py

@@ -1,0 +1,115 @@
+import os
+import sys
+import scenv
+
+import linalg
+import data
+import vcl
+import vcalg
+import vpm
+import segyvc
+import rsfvc
+import asg
+import awi
+import tmp
+
+class smoother(vcl.LinearOperator):
+
+    def __init__(self, dom, rect1=10, rect2=10, repeat=2):
+        RSFROOT = os.getenv('RSFROOT')
+        smoother = os.path.join(RSFROOT,'bin/sfsmooth')
+        self.dom = dom
+        if not isinstance(self.dom,rsfvc.Space):
+            raise Exception('grid smoother: provided domain not rsfvc.Space')
+        self.cmd = smoother + ' rect1=' + str(rect1) + ' rect2=' + str(rect2) + ' repeat=' +str(repeat)
+        
+    def getDomain(self):
+        return self.dom
+
+    def getRange(self):
+        return self.dom
+
+    def applyFwd(self,x,y):
+        try:
+            ret = os.system(self.cmd + ' < ' + x.data + ' > ' + y.data)
+            if ret != 0:
+                raise Exception('command failed: ' + self.cmd)
+        except Exception as ex:
+            print(ex)
+            raise Exception('called from smoother.applyFwd')
+
+    def applyAdj(self,x,y):
+        try:
+            ret = os.system(self.cmd + ' < ' + x.data + ' > ' + y.data)
+            if ret != 0:
+                raise Exception('command failed: ' + self.cmd)
+        except Exception as ex:
+            print(ex)
+            raise Exception('called from smoother.applyAdj')
+
+    def myNameIs():
+        print('2D Grid Smoother using sfsmooth')
+        print('rect1=10 rect2=10 repeat=2')
+
+# arguments:
+# output bulk mod, output filter,
+# input bulk mod, input data, buoyancy, source wavelet
+# fdpars
+# smpars
+# mswipars
+# descpars
+# lspars
+
+def mswiopt(bulkmodout, filterout,
+                bulkmodin, datain, filterin, buoyancy, source, 
+                order=2, sampord=1, nsnaps=20,\
+                cfl=0.5, cmin=1.0, cmax=3.0,dmin=0.8, dmax=3.0,\
+                nl1=250, nr1=250, nl2=250, nr2=250, pmlampl=1.0,
+                boundstest=True, noisy=False,
[…]

### 2024-02-27 repro/papers/awinum/project/files.py

@@ -1,59 +1,0 @@
-import scenv
-import data
-import linalg
-
-NTU = [251,501]
-NTD = [626,1251]
-NX  = [401, 801]
-NZ  = [201, 401]
-DELRTW = [-1000, -500]
-DT  = [8, 4]
-DX  = [20, 10]
-F1  = [1.0, 2.0]
-F2  = [2.5, 5.0]
-F3  = [7.5, 15.0]
-F4  = [12.5, 25.0]
-
-# adaptive filter prototype
-data.rechdr(file='ul.su',nt=NTU[0],dt=DT[0],
-            ntr=201,rx=2000,rz=1000,sx=4200,sz=3000,drx=20,delrt=DELRTW[0])
-data.rechdr(file='uh.su',nt=NTU[1],dt=DT[1],
-            ntr=201,rx=2000,rz=1000,sx=4200,sz=3000,drx=20,delrt=DELRTW[1])
-
-wsc=1.0e+3
-data.bpfilt(file='wl.su',nt=NTU[0],dt=DT[0],s=1.0,
-            f1=F1[0],f2=F2[0],f3=F3[0],f4=F4[0],
-            sx=4200,sz=3000)
-data.bpfilt(file='wh.su',nt=NTU[1],dt=DT[1],s=1.0,
-            f1=F1[1],f2=F2[1],f3=F3[1],f4=F4[1],
-            sx=4200,sz=3000)
-linalg.scale('wl.su',wsc)
-linalg.scale('wh.su',wsc)
-
-# create zero data files with same source position, rz=500, rx=[2000,6000]
-data.rechdr(file='hl.su',nt=NTD[0],dt=DT[0],ntr=201,
-            rx=2000.0,rz=1000.0,sx=4200,sz=3000,drx=20.0,delrt=0)
-data.rechdr(file='hh.su',nt=NTD[1],dt=DT[1],ntr=201,
-            rx=2000.0,rz=1000.0,sx=4200,sz=3000,drx=20.0,delrt=0)
-
-# bulk modulus with smaller focussing lens
-data.model(bulkfile='m0l.rsf', bulk=4.0, nx=NX[0], nz=NZ[0],
-   dx=DX[0], dz=DX[0], lensfac=0.7)
-data.model(bulkfile='m0h.rsf', bulk=4.0, nx=NX[1], nz=NZ[1],
-   dx=DX[1], dz=DX[1], lensfac=0.7)
-
[…]

### 2024-02-27 repro/papers/awinum/project/headers.py

@@ -1,40 +1,0 @@
-import scenv
-import data
-import linalg
-
-import sys
-
-NTU = int(sys.argv[1])
-NTD = int(sys.argv[2])
-DELRTW = float(sys.argv[3])
-DT  = float(sys.argv[4])
-F1  = float(sys.argv[5])
-F2  = float(sys.argv[6])
-F3  = float(sys.argv[7])
-F4  = float(sys.argv[8])
-NS  = int(sys.argv[9])
-DSX = float(sys.argv[10])
-
-u = sys.argv[11]
-w = sys.argv[12]
-h = sys.argv[13]
-
-### 2 shot headers
-# adaptive filter prototype
-data.rechdr(file=u,nt=NTU,dt=DT,
-            ntr=201,rx=2000,rz=1000,sx=4200,sz=3000,drx=20,delrt=DELRTW,
-            nshot=NS, dsx=DSX)
-
-wsc=1.0e+3
-data.bpfiltgather(file=w,nt=NTU,dt=DT,s=1.0,
-            f1=F1,f2=F2,f3=F3,f4=F4,
-            ntr=NS,sxstart=4200,szstart=3000,dsx=DSX,dsz=0)
-linalg.scale(w,wsc)
-
-# create zero data files with same source position, rz=500, rx=[2000,6000]
-data.rechdr(file=h,nt=NTD,dt=DT, ntr=201,
-            rx=2000.0,rz=1000.0,sx=4200,sz=3000,drx=20.0,delrt=0,
-            nshot=NS, dsx=DSX)
-
-
-

### 2024-02-27 repro/papers/awinum/project/models.py

@@ -1,17 +1,0 @@
-import scenv
-import data
-import linalg
-
-import sys
-
-NX  = int(sys.argv[1])
-NZ  = int(sys.argv[2])
-DX  = float(sys.argv[3])
-FAC = float(sys.argv[4])
-RAD = float(sys.argv[5])
-
-m = sys.argv[6]
-
-data.model(bulkfile=m, bulk=4.0, nx=NX, nz=NZ,
-   dx=DX, dz=DX, lensfac=FAC, lensradd=RAD)
-

### 2024-02-27 repro/papers/awinum/data/datasim.py
more

@@ -38,7 +38,6 @@ def asgsim(bulkmod, data_p, buoyancy, source_p, order=2, sampord=1, nsnaps=20,\
 
 if __name__ == '__main__':
     args = ", ".join(sys.argv[1:])
-    print(args)
     cmd = 'asgsim(' + args + ')'
     exec(cmd)
 

### 2024-02-28 repro/papers/awinum/paper.tex
more

@@
 one of FWI's impediments, namely its tendency to stagnate at
-suboptimal model estimates.
+suboptimal model estimates (``cycle-skip''), demonstrably far from global optima in
+controlled settings and often uninformative of material structure in
+field use. It is possible that such local descent algorithms are
+trapped in regions around local minima, though such trapping, or even
+the existance of local, non-global minima, are seldom established with
+any rigor. MSWI
+loosens the bond between predicted and observed data by interposing a
+filter, adapted to map one to the other trace-by-trace, and penalizes
+deviation of the filter from the Dirac delta. This soft penalty
+effectively relaxes the FWI data-fitting problem and permits local
+optimization methods to approximate data kinematics.
 
-(...)
-,
-demonstrably far from global optima in controlled settings and often
-uninformative of material structure in field use. It is possible that
-such local descent algorithms are trapped in regions around local,
-\section{Summary}
+The aim of this paper is to review the theoretical basis of MSWI,
+describe a practical computational framework for MSWI, and use it to
+solve a synthetic acoustic inversion problem for which FWI fails. In
+contrast, MSWI starting at the same initial estimate delivers a
+improved velocity from which FWI succeeds in approaching the global
+minimizer. The example is representative of a case (single arrival or
+simple wavefront transmitted wave data) in which the MSWI objective
+function is demonstrably close to mean square travel time error. Since
+travel time inversion is generally not subject to cycle-skipping,
+successful inversion is expected, and that is what we observe.
+
+MSWI in the form described here was introduced by
+\cite{HuangSymes2015SEG,HuangSymes:Geo17}. It is mathematically
+equivalent to an {\em extended} formulation of the inverse problem, in
+which acoustic point sources are allowed to depend on the receiving
+sensors - a non-physical expansion of the simulation domain. A number
+of other variants of FWI have been based on essentially the same
+extension of acoustic modeling
+\cite[]{Song:94c,Symes:94c,Plessix:00,LuoSava:11,LiAlkhalifah:21}. In
+particular it is the basis of Adaptive Waveform Inversion (AWI) \cite[]{Warner:16,
+  GuaschWarnerRavaut:GEO19,Warneretal:SEG21, Guaschetal:NPJDM20}. MSWI
+is closely related to AWI, but not identical: AWI includes a
+normalization of the adaptive filter, which makes the AWI objective
+function an even closer approximation to travel time mean square error
+than is MSWI, at least for transmitted wave data with a single
+arriving wavefront \cite[]{Symes:24a}.
+
+Transmission data with a single arrival is not just a case in which
+MSWI and AWI are known to closely approximate travel time inversion:
+it is the only case in which this link occurs. In particular, the link
+is broken for transmission data exhibiting multiple arrivals: in the
+presence of complex wavefronts, methods based on the source-receiver
+extension are no less likely than FWI to cycle-skip
+\cite[]{Symes:94c,HuangSymes:Geo17,Symes:24a}. However, the
+source-receiver extension is not the only possible route to FWI
+modification by artificially expanding the definition of energy
+source. \cite{HuangNammourSymesDollizal:SEG19} overview modifications
+of FWI based on various source extensions; some more recent advances
+are described by
+\cite{MetivierBrossier:SEG20,PladysBrossierLiMetivier:GEO21,LiAlkhalifah:21,Yongetal:GJI23,Opertoetal:GEO23}.
+Numerical examples suggest that some of these extensions may avoid
+cycle-skipping for transmitted wave data with complex wavefronts.
+
+The next section describes versions of FWI and MSWI based on
+acoustic wave propagation. The theoretical connection between MSWI and
+travel time inversion is reviewed. The computations required to apply local
+optimization to the objective functions of these inversion methods are
+detailed, as is the variable projection reduction of MSWI and the use
+of weighted norms in the domain of the simulation operator (wave
+velocity or bulk modulus fields, in the setting developed here). The
+third section presents an example, along with a detailed description
+of the numerical methods for simulation and optimization methods used
+in treating this example. The final section discusses some of the many
+unresolved questions around MSWI and similar approaches to inverse
+problems in wave propagation.
+
+\section{Background}
 The version of MSWI discussed here uses acoustic wave propagation with
@@
 recording interval $[0,t_d]$.
+
+In the discussion that follows, the buoyancy $\beta$ will be regarded
+as a fixed parameter. Thus $m$ is effectively the bulk modulus field $\kappa$.
 
@@
   \label{eqn:fwigrad}
-  g = \nabla  J_{\rm FWI}[m;d] = DF[m]^T(F[m]w-d).
+  g = \nabla  J_{\rm FWI}[m;d] = D_m(F[m]w)^T(F[m]w-d).
 \end{equation}
-In this formula, $DF[m]$ is the derivative of $F$ at $m$. 
+In this formula, $D_m(F[m]w)$ is the derivative of $F[m]w$ with
+respect to $m$. 
 This is the Euclidean (or $L^2$) gradient, that is, the vector $g$ for
@@
   \label{eqn:fwiwgrad}
-  g_W = W^{-1}\nabla  J_{\rm FWI}[m;d] =W^{-1} DF[m]^T(F[m]w-d).
+  g_W = W^{-1}\nabla  J_{\rm FWI}[m;d] =W^{-1} D_m(F[m]w)^T(F[m]w-d).
 \end{equation}
@@
 
+Note that the filtered predicted data may be viewed as the predicted
+data for an {\em extended source}, with the source wavelet at location
+$\bx_s$ replaced by $u(\bx_s,\bx_r,\cdot) * w(\bx_s,\cdot)$. That is,
+the adaptive filter construction is equivalent to allowing the source
+to depend on receiver position as well as source position - one source
+wavelet for each source-receiver pair. This is an extension of
+standard modeling, in that the domain (bulk modulus buoyancy, wavelet)
+is larger than in the conventional formulation, and coincides with it
+under the special condition that the source wavelet is independent of
+the receiver location. This  {\em
+  source-receiver extension}\cite[]{HuangSymes2015SEG} is a key ingredient in a number of
+other papers on modifications of FWI, as mentioned in the Introduction.
+
 It is possible to make the error between filtered predicted data and
@@
 \end{equation}
+As with all penalty methods, this definition involves a choice of
+weight ($\alpha$). A second weight ($\sigma$) scales the norm-squared
+of the filter, that is, a Tihonov regularization term, added for
+technical reasons explained by \cite{Warner:16,Symes:24a}. Choice of
+parameters is a critical step in the use of any penalty method. We
+offer some remarks about this choice in the Discussion section.
+
+The domain space (pairs of bulk modulus fields and adaptive filters)
+is very high-dimensional, compared with the domain space for FWI (bulk
+modulus fields alone, in the present context). It is possible to
+optimize $J_{\alpha,\sigma}[m,u;d]$ by alternating updates of $m$ and
+$u$  \cite[]{LiAlkhalifah:21}. However the {\em variable projection} \cite[]{GolubPereyra:73,GolubPereyra:03}
+reduction is generally much more efficient than the altrnating, or
+coordinate search, approach. In this instance, variable projection
+consists in minimizing $J_{\alpha,\sigma}[m,u;d]$ over $u$ (a
+quadratic optimization) to produce an optimal choice
+$u_{\alpha,\sigma}[m:d]$. The reduced objective is
+\begin{equation}
+  \label{eqn:redfiltpen}
+  \tilde{J}_{\alpha,\sigma}[m;d]=\frac{1}{2}(\|K[u_{\alpha,\sigma}[m;d]]F[m]w-d\|^2
+  +\alpha^2\|tu_{\alpha,\sigma}[m;d]\|^2 + \sigma^2\|u_{\alpha,\sigma}[m;d]\|^2).
+\end{equation}
+Note that like the FWI objective, $\tilde{J}_{\alpha,\sigma}[m;d]$
+depends only on $m$, with $d$ as a parameter.
+
+\cite{Symes:24a} explains the relation between the reduced MSWI
+objective and travel time inversion. This relation follows from the
+geometric asymptotics approximation to solutions of the point radiator
+problem \ref{eqn:awe} \cite[]{Friedlander:75}. Presuming that the
+coefficients $\kappa, \beta$ are smooth, there are smooth functions
+$\tau[m](\bx_s,\bx)$ (travel time) and $a[m](\bx_s,\bx)$ (geometric
+amplitude) depending on the model $m$, so that
+\begin{equation}
+  \label{eqn:pwa}
+  p(\bx,t;\bx_s) = a[m](\bx,\bx_s)w(t-\tau[m](\bx,\bx_s),\bx_s) + ...
+\end{equation}
+The elided terms are smoother than $w$, and become small with the
+dominant wavelength in $w$ (\cite{Symes:94a} explains the precise
+meaning of this condition). Thus we can write
+\[
+  F[m]w(\bx_r,t;\bx_s) \approx
+  a[m](\bx_r,\bx_s)w(t-\tau[m](\bx_r,\bx_s))
+\]
+Supposing that the data $d$ is noise-free,
+\[
+  d(\bx_r,t;\bx_s) = F[m^*]w(\bx_r,t;\bx_s) \approx
+  a[m^*](\bx_r,\bx_s)w(t-\tau[m^*](\bx_r,\bx_s))
+\]
+it follows that
+\begin{equation}
+  \label{eqn:tomo}
+  \lim_{\alpha \rightarrow 0}\frac{1}{\alpha^2} (\tilde{J}_{\alpha,\sigma}[m;d]
+  -\tilde{J}_{0,\sigma}) \approx \sum_{\bx_s,\bx_r} \frac{a[m^*]^2}{a[m]^2} (\tau[m^*]-\tau[m])^2\|g_{\frac{a[m^*]}{a[m]}\sigma}\|^2
+\end{equation}
+Here $g_{\sigma}$ is an approximate delta. Its Fourier transform of
+is
+\begin{equation}
+  \label{eqn:gsig}
+  \hat{g}_{\sigma} = \frac{a[m^*]^2 |\hat{w}|^2}{a[m^*]^2 |\hat{w}|^2
+    + \sigma^2}
+\end{equation}
+whence $g_{\sigma}(\bx_s,\bx_r,t) \rightarrow \delta(t)$ in
+the sense of distributions as $\sigma \rightarrow 0$.
+
+If $\alpha$ and $\sigma$ are sufficiently small, then
+$\tilde{J}_{\alpha,\sigma}$ is close to the right-hand side, which is bounded above and below by multiples of the the mean square travel time error between $m$ and
+$m^*$.  This is the relation mentioned in the introduction. 
+It does not
+show that the only stationary points are those of the mean square
+travel time error, even approximately. However if the amplitudes are
+relatively insensive to changes in model, as is true if the
+source-receiver distance is well away from developing multiple
+arrivals, then the difference of stationary points gets small as $m$
+approaches a stationary point of $\tilde{J}_{\alpha,\sigma}$. So we
+would expect minimization $\tilde{J}_{\alpha,\sigma}$ to produce a
+model that closely matches the travel times inherent in the data
+($\tau[m^*]$ in the notation used here).
+
+AWI adds one more feature, namely scaling by the trace norm of the
+filter. This normalization can be interpreted as a choice of weighted
+norm on the space of adaptive filters. It leads to a very similar
+relation to \ref{eqn:tomo} for the AWI penalty function, but without
+the amplitude factors: that is, the right-hand side of \ref{eqn:tomo}
+is {\em just} the mean-square traveltime error. See \cite{Symes:24a}
+for details.
+
+The reduced adaptive filter $u_{\alpha,\sigma}[m;d]$ is the solution
+of the {\em normal equation}
+\begin{equation}
+  \label{eqn:normal}
+  (S[m]^TS[m] + \alpha^2 t^2 + \sigma^2 I)u = S[m]^Td,
+\end{equation}
+in which $S[m]u = K[u]F[m]w$. This positive definite symmetric linear
+system may be solved by various efficient numerical methods. Having
+computed $u_{\alpha,\sigma}[m;d]$ hence the value of 
+$\tilde{J}_{\alpha,\sigma}[m;d]$, its (Euclidean) gradient is given by
+\begin{equation}
+  \label{eqn:gradredfiltpen}
+  \nabla \tilde{J}_{\alpha,\sigma}[m;d] =
+  D_m(F[m]w)^TK[u_{\alpha,\sigma}[m;d]]^T(K[u_{\alpha,\sigma}[m;d]]F[m]w-d)
+\end{equation}
+Apart from the appearance of the truncated filter operator
+$K[u_{\alpha,\sigma}[m;d]]$, this is almost identical to the FWI
+gradient \ref{eqn:fwigrad}. In particular, the last step in the
+computation on the right hand side, the application of the adjoint
+$m$-derivative of F, is exactly the same.
+
+Use of a weighted norm in the model space goes exactly as before: with
+weight operator $W$, the weighted gradient is
+\begin{equation}
+  \label{eqn:wgradredfiltpen}
+  \nabla_W \tilde{J}_{\alpha,\sigma}[m;d] =
+  W^{-1}D_m(F[m]w)^TK[u_{\alpha,\sigma}[m;d]]^T(K[u_{\alpha,\sigma}[m;d]]F[m]w-d)
+\end{equation}
+
+The gradient (or weighted gradient), together with the value, are the
+inputs to first-order methods such as steepest descent and
+Limited-Memory Broyden-Fletcher-Goldfarb-Shanno
+iterations. Methods more closely related to Newton iteration require
+more involved computations (see for instance \cite{Kaufman:75}.
 
@@
 obtained is not at all satisfactory. More discussion of this result
-will be found later in the paper.
+will be found below.
+
+Now apply the same algorithm to the MSWI objective
+$J_{\alpha,\sigma}$. Of course, to do so requires a choice of the
+parameters $\alpha$ and $\sigma$. The purpose of the regularization
+parameter $\sigma > 0$ is
+simply to assure that the system \ref{eqn:normal} is positive
+definite, so it can be chosen quite small - whatever that means. The
+penalty parameter $\alpha > 0$ is the link between the filtered
+residual and the spread in time of the filter: the performance of the
+algorithm depends primarily on its choice. The next section will
+discuss rational approaches to this choice. For now, we use the values
+$\alpha=0.0001, \sigma=0.00001$ based on numerical exploration of the
+objective.
+
+The system \ref{eqn:normal} is far larger than is convenient to solve
+by any variant of Gaussian Elimination, so iterative methods are 
+necessary. We choose the Conjugate Gradient (CG.) method
+\cite[]{Dan:67,Steihaug:83,NocedalWright}, and stop the
+iteration by monitoring the reduction in length of the normal residual
+(the gradient of $J_{\alpha,\sigma}[m,u;d]$ with respect to $u$). The
+reduction threshhold $\rho$ is thus another necessary input parameter:
+we choose $\rho = 0.001$ (note that unlike $\alpha$ and $\sigma$,
+$\rho$ is dimensionless). We use the same inverse weight operator as
+in the FWI example, namely a 10 $\times$ 10 moving average, repeated
+once.
+
+The initial bulk modulus field is once again homogeneous at 4 GPa. The
+initial value of the reduced MSWI objective is $\approx 1.32 \times 10^{-2}$,
+and the length of the (weighted) gradient is $\approx 6.37 \times
+10^{-10}$. After 7 iterations, the normal residual reduction criterion
+is satisfied. The value has decreased to $\approx 9.2 \times 10^{-4}$,
+or a bit more than an order of magnitude reduction. The gradient
+length is $\approx 4.33 \times 10^{-12}$, a reduction of more than two
+orders of magnitude. The resulting bulk modulus is shown in Figure \ref{fig:mest11}.
+
+\plot{mest11}{width=\textwidth}{a: Bulk modulus produced by 7
+  steepest descent steps to minimize the reduced MSWI objective \ref{eqn:redfiltpen},
+  using the data shown in \ref{fig:d11}.}
+
+\multiplot{2}{sim11,resid11}{width=0.45\textwidth}{a: Data
+  corresponding to MSWI result shown in Figure
+  \ref{fig:mest11}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:sim11} and target data
+  \ref{fig:d11}. Note that the later signal in the central portion is
+  still shifted, but by much less than in Figure \ref{fig:residfwi0}.}
+
+Inspection of the residual plot \ref{fig:resid11} suggests that the
+data predicted from the MSWI result is in many places within a
+half-wavelength travel time shift of the data in Figure \ref{fig:d11}.
+A half-wavelength is the conventional threshhold for FWI initial fit
+error: below that it works, above that it doesn't. Thus it is tempting
+to attempt FWI from using the MSWI result as initial estimate.
+
+We applied 11 iterations of steepest descent to the FWI objective,
+with initial estimate of $m$ equal to the final estimate from the MSWI
+inversion just described (Figure \ref{fig:mest11}). The initial FWI
+objective value was $\approx 4.3$, and the weighted gradient norm was
+$\approx 2.15 \times 10^{-5}$.  After 11 steps of steepest descent,
+the prescribed reduction ($10^{-2}$) of the weighted gradient norm was achieved,
+and the objective value had decreased to $\approx 4.41 \times
+10^{-2}$. The resulting bulk modulus field is depicted in Figure
+\ref{fig:mestfwi}. The predicted data at this model, and the residual
+data, are shown in Figures \ref{fig:resimfwi} and \ref{fig:residfwi}
+respectively.
+
+The FWI iteration starting at the MSWI result has effectively achieved
+the global minimum of the FWI objective function: the data match is
+very close. This type of result could perhaps have been achieved by
+systematically increasing $\alpha$, as will be discussed in the next
+section. However, the current result has a remarkable corollary,
+provided that we are willing to assume that the intermediate values of
+the FWI objective between steepest descent iterates are monotonically
+decreasing. This condition could in principle be checked
+computationally. However, for the moment let us accept that it is
+so. Then we can conclude that the homogeneous model at which the first
+FWI inversion was started must lie in a basin of attraction of a
+non-global minimizer, since the value of the objective function at the
+homogeneous model is actually {\em smaller} than that at the result of
+MSWI inversion. However the descent sequence starting at that initial
+estimate arrives at essentially zero residual. No descent path from
+the homogeneous model can
 

### 2024-03-02 pyrvl/vcl.py
least squares jet class

@@ -883,11 +883,109 @@ class StandardJet(ScalarJet):
     def Hessian(self):
         return self.f.Hessian(self.x)
 
+    # null implementation - temporary
+    def archive(self, name, tag, suf):
+        pass
+
     def myNameIs(self):
         print('StandardJet: jet based on ScalarFunction')
         print('Function:')
         self.f.myNameIs()
         print('Evaluation point:')
+        self.x.myNameIs()
+
+class LeastSquaresJet(ScalarJet):
+
+    '''
+    Jet for Least Squares function 0.5*|F(x)-b|^2. Eliminates some intermediate computations
+    necessary in the general case by saving the residuaal vector. Default implementation of 
+    Hessian returns G-N approximation - can be overridden in subclasses
+
+    Constructor parameters:
+    F (Function): map
+    x (Vector): at this point in its domain
+    b (Vector): data vector
+    '''
+
+    def __init__(self, x, F=None, b=None):
+        try:
+            if F is None:
+                raise Exception('second arg is None - cannot have a jet without a function!')
+            if not isinstance(F, Function):
+                raise Exception('second arg not ScalarFunction')
+            if not isinstance(x, Vector):
+                raise Exception('first arg not Vector')
+            if x.space != F.getDomain():
+                raise Exception('first arg not in domain of second arg')
+            if b is None:
+                raise Exception('third arg is None - data vector of least squares function')
+            if not isinstance(b, Vector):
+                raise Exception('third arg not Vector')
+            if b.space != F.getRange():
+                raise Exception('third arg not in range of second arg')
+        except Exception as ex:
+            print(ex)
+            raise Exception('called from StandardJet constructor')
+        else:
+            self.F = F
+            self.x = x
+            self.b = b
+            self.res = None
+            self.resn = None
+            self.grad = None
+            self.DF = None
+            self.hess = None
+
+    def point(self):
+        return self.x
+
+    def value(self):
+        try:
+            if self.res is None:
+                self.res = self.F(self.x)
+                self.res.linComb(-1,self.b)
+                self.resn = self.res.norm()
+            return 0.5*self.resn*self.resn
+        except Exception as ex:
+            print(ex)
+            raise Exception('called from LeastSquaresJet.value')
+
+    def gradient(self):
+        try:
+            if self.grad is None:
+                if self.res is None:
+                    self.res = self.F(self.x)
+                    self.res.linComb(-1,self.b)
[…]

### 2024-03-02 pyrvl/awi.py
more

@@ -687,8 +687,9 @@ class mswi(vcl.ScalarJet):
                      archivepath=None, pointname=None, gradname=None,
                      filtername=None, dataerrname=None):
         try:
-            if verbose != 0:
-                print('mswi constructor')
+
+#            if verbose != 0:
+            print('mswi constructor: verbose = ' + str(verbose))
             # sanity checks
             if data is None:
                 raise Exception('data vector not provided')

### 2024-03-02 pyrvl/vcalg.py
more

@@ -540,7 +540,28 @@ def trgn(x, b, F, imax, eps, kmax, rho, Delta, mured=0.5, muinc=1.8, \
 #    2 - (or more) print it, res at every it
 
 def bcgstep(x, A, k, p, r, gamma, Delta=None, verbose=0):
-#    os.system('echo top; ls /var/tmp')
+    '''
+    Iteration step, basic CG algorithm for SPD systemts.
+
+    Parameters:
+    x (vcl.Vector):           solution estimate to be updated
+    A (vcl.LinearOperator):   SPD operator
+    k (int):                  iteration number
+    p (vcl.Vector):           update direction
+    r (vcl.Vector):           residual on return (optional)
+    gamma (float):            residual norm squared
+    Delta (float):            optional trust radius constraint
+    verbose (int):            verbosity level
+                              0 - no print output
+                              1 - print number of its performed, residual norm
+                              2 - (or more) print it number, res at every it
+
+    Return: [x, k, p, r, gamma, cgend]
+    all updated parameters except
+    cgend (Boolean):          True if trust region transgressed, else False
+
+    '''
+
     try:
             
         # print('#2. $s = A^Tq$')
@@ -880,10 +901,13 @@ class LineSearch(ABC):
     dir (vcl.Vector):     search direction
     step (float):         initial step
     J (vcl.ScalarJet):    objective jet (class)
+    jetargs (dict):       kwargs for jet
+    verbose (int)         verbosity flag (silent = 0)
+    fout (iofile):        output unit, or none for stdout
 #    lsargs:               additional arguments (keyword)
     '''
 
-    def __init__(self, x, val, grad, dir, step, J, jetargs):
+    def __init__(self, x, val, grad, dir, step, J, jetargs, verbose, fout):
         self.x = x
         self.val = val
         self.grad = grad
@@ -891,8 +915,9 @@ class LineSearch(ABC):
         self.step = step
         self.J = J
         self.jetargs = jetargs
-#        self.lsargs = lsargs
-
+        self.verbose = verbose
+        self.fout = fout
+            
     # returns updated jet instance and step
 #    @abstractmethod
     def search(self):
@@ -903,7 +928,7 @@ class btls(LineSearch):
     '''
     Simple acktracking line search. 
 
-    Generic Parameters:
+    Base class parameters:
     x (vcl.Vector):       current solution estimate
     val (float):          current objective value
     grad (vcl.Vector):    current gradient
@@ -911,23 +936,23 @@ class btls(LineSearch):
     step (float):         initial step
[…]

### 2024-03-02 pyrvl/asg.py
more

@@ -406,7 +406,7 @@ class fsbop(vcl.Function):
                    ' deriv=0 adjoint=0' + self.therest + \
                    ' partask=' + str(self.partask)
 
-            print(cmd + args)
+            # print(cmd + args)
             ret = os.system(cmd + args)
             if ret != 0:
                 raise Exception('Error: return ' + str(ret) + ' from sim.x')

### 2024-05-26 repro/papers/awinum/paper.tex
interim

@@
 \begin{abstract}
-Matched Source Waveform Inversion applied to transmission data
+Matched Source Waveform Inversion applied to acoustic transmission data
 produces an estimate of refractive index similar to the result of
@@
 times. This paper reviews the theoretical justification of this result
-and its limitations, and exhibits a numerical illustration. 
+and its limitations, and with 2D numerical illustrations. 
 \end{abstract}
@@
 trapped in regions around local minima, though such trapping, or even
-the existance of local, non-global minima, are seldom established with
+the existence of local, non-global minima, are seldom established with
 any rigor. MSWI
@@
 solve a synthetic acoustic inversion problem for which FWI fails. In
-contrast, MSWI starting at the same initial estimate delivers a
+contrast, MSWI starting at the same initial estimate delivers an
 improved velocity from which FWI succeeds in approaching the global
@@
 
-\section{Background}
+\section{Theory}
 The version of MSWI discussed here uses acoustic wave propagation with
@@
 with any other vector $\delta m$ gives the
-rate of change in the direction of that vector of $J_{\rm FWI}$ at m.
+rate of change in the direction of that vector of $J_{\rm FWI}$ at $m$.
 
@@
 the receiver location. This  {\em
-  source-receiver extension}\cite[]{HuangSymes2015SEG} is a key ingredient in a number of
+  source-receiver extension} \cite[]{HuangSymes2015SEG} is a key ingredient in a number of
 other papers on modifications of FWI, as mentioned in the Introduction.
@@
 The elided terms are smoother than $w$, and become small with the
-dominant wavelength in $w$ (\cite{Symes:94a} explains the precise
+dominant wavelength in $w$ (\cite{Symes:24a} explains the precise
 meaning of this condition). Thus we can write
@@
 source-receiver distance is well away from developing multiple
-arrivals, then the difference of stationary points gets small as $m$
-approaches a stationary point of $\tilde{J}_{\alpha,\sigma}$. So we
+arrivals, then any stationary point of $\tilde{J}_{\alpha,\sigma}$ is
+either close to a tomographic stationary point, or far from $m^*$. So we
 would expect minimization $\tilde{J}_{\alpha,\sigma}$ to produce a
@@
 relation to \ref{eqn:tomo} for the AWI penalty function, but without
-the amplitude factors: that is, the right-hand side of \ref{eqn:tomo}
-is {\em just} the mean-square traveltime error. See \cite{Symes:24a}
+the amplitude factors: that is, the right-hand side of the AWI
+analogue of \ref{eqn:tomo}
+is {\em just} the mean-square traveltime error, up to controllable
+error, for single-arrival transmission data. See \cite{Symes:24a}
 for details.
+
+\cite{Symes:24a} also shows that if energetic multiple arrivals are
+present in the data, for any reason, then cross-talk between travel
+time branches destroys the relation between MSWI (or AWI) and any
+version of the mean-square travel time error. The next sections of
+this paper will illustrate the
+success of MSWI for single-arrival data, and its failure for data with
+multiple arrivals.
+
+We end this section by overviewing the computations required for
+minimization of $\tilde{J}_{\alpha,\sigma}$.
 
@@
 
-\section{A numerical illustration}
+\section{Numerical Illustration}
 
@@
 
-This section presents the inversion of acoustic data with dimensions
-typical of crustal seismic exploration. The target bulk modulus field ($\kappa$)
-is depicted in Figure \ref{fig:m}, and contains an acoustic ``lens''
-positioned in the center between 1000 m and 3000 m depth
-(``$z$''). The background level outside the lens is 4 GPa; at the
-center it is 2 GPa. The
-buoyancy ($\beta$)is spatially homogeneous at 1 cm$^3/g$, so the background
-wave speed is 2000 m/s.
+This section presents application of MSWI to simulated acoustic data with dimensions
+typical of crustal seismic exploration. We use several configurations
+to illustrate the capabilities and limitations of this approach. In
+all cases, we use MSWI to generate an initial bulk modulus estimate
+input to FWI.
 
-\plot{m}{width=\textwidth}{Lens model.}
+\subsection{Experimental setup}
 
-This field is sampled on a 20 $\times$ 20 m grid. Acoustic wave
+The several experiments to be described below share a number of
+features, set out in this section.
+
+Acoustic wave
 propagation is simulated via a staggered grid finite difference method
-\cite[]{vir86,lev88,Cohen:01}. The time
-step is chosen to be safely stable, given the parameter fields,
-difference formulae, and
-spatial sampling.
+\cite[]{vir86,lev88,Cohen:01} of order 2 in time and 8 in space. The time
+step is an internal detail of the simulation, chosen to be safely
+stable, given the parameter fields, difference formulae, and 
+spatial sampling. The discrete pressure field is
+output over the time range $0 < t < 5$ s with sample interval $0.008$
+s at externally specified receiver locations
+via piecewise linear interpolation in space and cubic spline
+interpolation in time. Isotropic point sources are added
+into the acoustic fields at each time step via the adjoints of these
+interpolation operations. These and other details of the simulator are
+discussed in \cite{GeoPros:11}.
 
-Receivers are spaced on the line at depth 1000 m, from 2000 m to 6000
-m horizontal coordinate (``$x$''), spaced 20 m apart. Eleven source
-positions lie on the line $z=3000$ m, spaced 200 m apart, from
-$x=3000$ m to $x = 5000$ m. The source wavelet $w$ (the same for every
-source ) is a $[1.0, 2.5, 7.5, 12.5]$ Hz trapezoidal bandpass filter,
-with a median frequency of 5.875 Hz corresponding to a median wavelength of
+Spatial fields (bulk modulus, buoyancy, gradients) occupy a
+rectangular region of 8 km horizontal (``x'') extent and 4 km vertical
+(``z'') extent, sampled on a 20 $\times$ 20 m grid. For all except the
+third example, receivers are located on the vertical line at
+horizontal coordinate $x$= 5000 m, spaced 20 m apart, starting at
+depth $z$ = 200 m. Twenty isotropic point sources occupy positions lie
+on the line $x=3000$ m, spaced 150 m apart, starting at depth = 500
+m. For the third example, source and receiver lines are moved 2000 m
+further apart, to $x=2000$ m and $x=6000$ m respectively, other
+parameters remaining the same. [Note that this geometry is a 2D
+cartoon of that used in cross-well seismic data acquisition.] The
+source wavelet $w$ (the same for every source ) is a
+$[1.0, 2.5, 7.5, 12.5]$ Hz trapezoidal bandpass filter, with a median
+frequency of 5.875 Hz corresponding to a median wavelength of
 $\approx$ 340 m, and shortest wavelength of 160 m. It is centered at
@@
 
-The discrete pressure field is
-sampled at the externally specified time grid and spatial locations
-via piecewise linear interpolation in space and cubic spline
-interpolation in time. The isotropic point source is added
-into the acoustic fields at each time step via the adjoints of these
-interpolation operations.
+The initial model $m_0$ for MSWI is homogeneous, with $\kappa = $ 4
+GPa, $\beta = $ 1 cm$^3$/g. As noted earlier, we regard $\beta$ as
+fixed in this series of experiments, so mention of it will be
+suppressed. The corresponding data for isotropic point sources at $x=3000$ m,
+point receivers at $x=5000$ m, and the bandpass filter time dependence
+described earlier, is depicted in
+Figure \ref{fig:chwd20}. We use both $m_0$ and the output of MSWI as
+initial models for FWI..
 
-Simulated data from this configuration appears as Figure \ref{fig:d11}.
-
-The initial model $m_0$ for inversion is homogeneous, with $\kappa = $ 4 GPa,
-$\beta = $ 1 cm$^3$/g. The corresponding data is depicted in Figure \ref{fig:sim0}.
-These plots show the time shift between the homogeneous model data and
-the target data, for the central source positions.
-
-\multiplot{2}{d11,sim0}{width=0.45\textwidth}{a: Data for target model $m$
-  (Figure \ref{fig:m}). b: Data for homogeneous model $m_0$.}
+\plot{chwd20}{width=\textwidth}{Data for homogeneous model $m_0$: 20
+  sources at $x=3000$ m, 201 receivers at $x=5000$ m.}
 
 We apply a version of weighted steepest descent optimization to the
-minimization of $J_{\rm FWI}$. The inverse weight operator ($W^{-1}$
+minimization of both $J_{\rm FWI}$ and $\tilde{J}_{\alpha,\sigma}$.
+The inverse weight operator ($W^{-1}$
 in formula \ref{eqn:fwiwgrad}) is a
-10-point moving average in both spatial directions, repeated once. As
-noted above, the weight operator itself is not required. The weighted
+10-point moving average in both spatial directions, repeated once,
+except in the fourth example as discussed below. Recall that the weight operator itself is not required. The weighted
 gradient is computed via formula \ref{eqn:fwigrad}. The adjoint
@@
 and the optimal step in this direction is approximated by a simple
-backtracking line search algorithm \cite[]{NocedalWright}.
+backtracking line search algorithm \cite[]{NocedalWright}. The
+iteration is terminate either when the gradient norm has fallen below
+1\% of its initial value, or when a maximum number of iterations
+(usually 12) is reached.
 
-The value of $J_{\rm FWI}[m_0,d] \approx 3.04$, and the rate of
-increase in the weighted gradient direction at $m_0$ is 1.74 $\times
-10^{-6}$. Note that the rate of increase is the same as the weighted
-gradient norm. After 10 steepest descent steps, the objective value has
-decreased by about 30\%, to 1.9, and the weighted gradient norm by
-more than an order of magnitude, to $\approx 9.6 \times 10^{-8}$. The
-final model appears in Figure \ref{fig:mestfwi0}, and the
-corresponding data in Figure \ref{fig:resimfwi0}.
+The MSWI objective $\tilde{J}_{\alpha,\sigma}$ requires choices of the
+parameters $\alpha$ and $\sigma$, and solution of the normal equation
+\ref{eqn:normal}. This system is far larger than is convenient to solve
+by any variant of Gaussian Elimination, even for our small 2D
+examples, so iterative methods are 
+necessary. We choose the Conjugate Gradient (CG.) method
+\cite[]{Dan:67,Steihaug:83,NocedalWright}, and stop the
+iteration by monitoring the reduction in length of the normal residual
+(the gradient of $J_{\alpha,\sigma}[m,u;d]$ with respect to $u$). The
+reduction threshhold $\rho$ is thus another necessary input
+parameter. For all calculations in this paper, we use
+we choose $\rho = 0.01$ (note that unlike $\alpha$ and $\sigma$,
+$\rho$ is dimensionless).
+
+Choice of regularization weights like $\alpha$ and $\sigma$ is a
+widely studied topic; we mention some methods for this task in the
+discussion section. For the set of examples presented here, we take a
+simpler approach. Since the purpose of $\sigma>0$ is to avoid singularity
+in the system \ref{eqn:normal} even for $\alpha=0$, we choose it a bit
+bigger than floating point tolerance for single precision:
+$\sigma = 10^{-5}$. We choose $\alpha$ by computing the solution
+$u_{\alpha,\sigma}[m_0;d]$ for the data $d$ used in the first example
+below, and several powers of 10 for $\alpha$. The largest such
+$\alpha$ that yields an RMS predicted data error
+($\|S[m_0]u_{\alpha,\sigma}[m_0;d] - d\|$) less than $0.05 \|d\|$ is
+$\alpha = 10^{-4}$. We take this relation as representative, and use
+this value for all MSWI computations.
+
+\subsection{Recovering a circular lens}
+
+The target bulk modulus field ($\kappa$) for the first example 
+is depicted in Figure \ref{fig:m}. It contains an acoustic ``lens''
+positioned in the center between 1000 m and 3000 m depth
+(``$z$''). The background level outside the lens is 4 GPa; at the
+center it is 2.4 GPa. The
+buoyancy ($\beta$)is spatially homogeneous at 1 cm$^3/g$, so the background
+wave speed is 2000 m/s.
+
+\plot{m}{width=\textwidth}{Lens model.}
+
+Simulated data from this configuration appears as Figure
+\ref{fig:cwd20}.
+
+\plot{cwd20}{width=\textwidth}{Data for circular lens model (\ref{fig:m}): 20
+  sources at $x=3000$ m, 201 receivers at $x=5000$ m.}
+
+The value of $J_{\rm FWI}[m_0,d] \approx 4.6$, and the rate of
+increase in the weighted gradient direction at $m_0$ is 4.2 $\times
+10^{-6 }$. Note that the rate of increase is the same as the weighted
+gradient norm. After 12 steepest descent steps, the objective value has
+decreased to 2.6, and the
+weighted gradient norm by more than an order of magnitude, to $\approx 2.5 \times 10^{-7}$. The
+final model appears in Figure \ref{fig:cwlens20mestfwi0}, and the
+corresponding data in Figure \ref{fig:cwlens20resimmestfwi0}.
 
@@
 
-\plot{mestfwi0}{width=\textwidth}{a: Bulk modulus produced by 10
+\plot{cwlens20mestfwi0}{width=\textwidth}{a: Bulk modulus produced by 10
   steepest descent steps to minimize the FWI objective \ref{eqn:fwi},
-  using the data shown in \ref{fig:d11}.}
+  using the data shown in \ref{fig:cwd20}.}
 
-\multiplot{2}{resimfwi0,residfwi0}{width=0.45\textwidth}{a: Data
+\multiplot{2}{cwlens20resimmestfwi0,cwlens20residmestfwi0}{width=0.45\textwidth}{a: Data
   corresponding to FWI inversion result shown in Figure
-  \ref{fig:mestfwi0}. b: Residual or data error - difference between
-  data shown in Figure \ref{fig:resimfwi0} and target data
-  \ref{fig:d11}. Note the failure to match the later signal in the
+  \ref{fig:cwlens20mestfwi0}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:cwlens20resimmestfwi0} and target data
+  \ref{fig:cwd20}. Note the failure to match arrival times of the later signal in the
   central part of the display.}
@@
 
-Now apply the same algorithm to the MSWI objective
-$J_{\alpha,\sigma}$. Of course, to do so requires a choice of the
-parameters $\alpha$ and $\sigma$. The purpose of the regularization
-parameter $\sigma > 0$ is
-simply to assure that the system \ref{eqn:normal} is positive
-definite, so it can be chosen quite small - whatever that means. The
-penalty parameter $\alpha > 0$ is the link between the filtered
-residual and the spread in time of the filter: the performance of the
-algorithm depends primarily on its choice. The next section will
-discuss rational approaches to this choice. For now, we use the values
-$\alpha=0.0001, \sigma=0.00001$ based on numerical exploration of the
-objective.
+This example is the one on which we based the choice $\alpha =
+10^{-4}$. We did this by minimizing $J_{\alpha,\sigma}[m_0, u; d]$
+over $u$ to identify a filter that outputs a good fit to the data
+\ref{fig:cwd20} from the input \ref{fig:chwd20} (simulated from
+$m_0$), and choosing $\alpha$ just small enough to obtain an RMS error
+under 5\%.  The resulting adaptive filter $u_{\alpha,\sigma}[m_0;d]$
+is shown in Figure \ref{fig:cwlens20uest0}. This filter exhibits a lot
+of energy at non-zero times, as is necessary to move the events in
+\ref{fig:chwd20} near those in \ref{fig:cwd20}. Thus the penalty term
+is rather large. We will use the apparent dispersion of energy away
+from $t=0$ in these adaptive filters to judge the quality of
+inversions.
 
-The system \ref{eqn:normal} is far larger than is convenient to solve
-by any variant of Gaussian Elimination, so iterative methods are 
-necessary. We choose the Conjugate Gradient (CG.) method
-\cite[]{Dan:67,Steihaug:83,NocedalWright}, and stop the
-iteration by monitoring the reduction in length of the normal residual
-(the gradient of $J_{\alpha,\sigma}[m,u;d]$ with respect to $u$). The
-reduction threshhold $\rho$ is thus another necessary input parameter:
-we choose $\rho = 0.001$ (note that unlike $\alpha$ and $\sigma$,
-$\rho$ is dimensionless). We use the same inverse weight operator as
-in the FWI example, namely a 10 $\times$ 10 moving average, repeated
-once.
+\plot{cwlens20uest0}{width=\textwidth}{Adaptive filter to match
+  initial model data (Figure \ref{fig:chwd20}) to circular lens
+  data (Figure \ref{fig:cwd20}). Note considerable energy dispersion
+  away from $t=0$.}
 
-The initial bulk modulus field is once again homogeneous at 4 GPa. The
-initial value of the reduced MSWI objective is $\approx 1.32 \times 10^{-2}$,
-and the length of the (weighted) gradient is $\approx 6.37 \times
-10^{-10}$. After 7 iterations, the normal residual reduction criterion
-is satisfied. The value has decreased to $\approx 9.2 \times 10^{-4}$,
-or a bit more than an order of magnitude reduction. The gradient
-length is $\approx 4.33 \times 10^{-12}$, a reduction of more than two
-orders of magnitude. The resulting bulk modulus is shown in Figure \ref{fig:mest11}.
+We applied the same steepest descent algorithm to the MSWI objective
+$J_{\alpha,\sigma}$. The initial bulk modulus field is once again
+homogeneous at 4 GPa. The initial value of the reduced MSWI objective
+is $\approx 1.49 \times 10^{-2}$, and the length of the (weighted)
+gradient is $\approx 4.7 \times 10^{-10}$. After 12 iterations, the value has
+decreased to $\approx 3.25 \times 10^{-3}$, or roughly a factor of 4.5.
+reduction. The gradient length is
+$\approx 5.6 \times 10^{-12}$, a reduction of almost two orders of
+magnitude. The resulting bulk modulus is shown in Figure
+\ref{fig:cwlens20mestmswi}.
 
-\plot{mest11}{width=\textwidth}{a: Bulk modulus produced by 7
+\plot{cwlens20mestmswi}{width=\textwidth}{a: Bulk modulus produced by 12
   steepest descent steps to minimize the reduced MSWI objective \ref{eqn:redfiltpen},
-  using the data shown in \ref{fig:d11}.}
+  using the data shown in \ref{fig:cwd20}.}
 
-\multiplot{2}{sim11,resid11}{width=0.45\textwidth}{a: Data
-  corresponding to MSWI result shown in Figure
-  \ref{fig:mest11}. b: Residual or data error - difference between
-  data shown in Figure \ref{fig:sim11} and target data
-  \ref{fig:d11}. Note that the later signal in the central portion is
-  still shifted, but by much less than in Figure \ref{fig:residfwi0}.}
+The final adaptive filter is shown in Figure
+\ref{fig:cwlens20uestmswi}. Note that energy in the filter has
+migrated towards $t=0$, compared to the adaptive filter produced from
+the homogenous bulk modulus \ref{fig:cwlens20uest0}, and the bulk of
+it is within an apparent half-wavelength of $t=0$. This suggests that
+the MSWI result may be an adequate initial estimate for a successful
+FWI.
 
-Inspection of the residual plot \ref{fig:resid11} suggests that the
+\plot{cwlens20uestmswi}{width=\textwidth}{Adaptive filter to match
+  data for MSWI model (Figure \ref{fig:cwlens20resimmestmswi}) to circular lens
+  data (Figure \ref{fig:cwd20}). Note considerable less energy dispersion
+  away from $t=0$ than is evident in Figure \ref{fig:cwlens20uest0}.}
+
+Another way to look at the improved data kinematics of the MSWI result
+is via the simulate data from the model shown in
+\ref{fig:cwlens20mestmswi}. Figures \ref{fig:cwlens20resimmestmswi},
+\ref{fig:cwlens20residmestmswi} suggest that the
 data predicted from the MSWI result is in many places within a
-half-wavelength travel time shift of the data in Figure \ref{fig:d11}.
+half-wavelength travel time shift of the data in Figure \ref{fig:cwd20}.
 A half-wavelength is the conventional threshhold for FWI initial fit
-error: below that it works, above that it doesn't. Thus it is tempting
-to attempt FWI from using the MSWI result as initial estimate.
+error: below that it works, above that it doesn't.
+
+\multiplot{2}{cwlens20resimmestmswi,cwlens20residmestmswi}{width=0.45\textwidth}{a:
+  Data corresponding to MSWI result shown in Figure
+  \ref{fig:cwlens20mestmswi}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:cwlens20resimmestmswi} and target data
+  \ref{fig:cwd20}. Note that the later signal in the central portion is
+  still shifted, but by much less than in Figure \ref{fig:cwlens20residmestfwi0}.}
 
@@
 with initial estimate of $m$ equal to the final estimate from the MSWI
-inversion just described (Figure \ref{fig:mest11}). The initial FWI
-objective value was $\approx 4.3$, and the weighted gradient norm was
-$\approx 2.15 \times 10^{-5}$.  After 11 steps of steepest descent,
+inversion just described (Figure \ref{fig:cwlens20mestmswi}). The initial FWI
+objective value was $\approx 2.9$, and the weighted gradient norm was
+$\approx 2.0 \times 10^{-4}$.  After 11 steps of steepest descent,
 the prescribed reduction ($10^{-2}$) of the weighted gradient norm was achieved,
-and the objective value had decreased to $\approx 4.41 \times
+and the objective value had decreased to $\approx 9.1 \times
 10^{-2}$. The resulting bulk modulus field is depicted in Figure
-\ref{fig:mestfwi}. The predicted data at this model, and the residual
-data, are shown in Figures \ref{fig:resimfwi} and \ref{fig:residfwi}
+\ref{fig:cwlens20mestmswifwi}. The predicted data at this model, and the residual
+data, are shown in Figures \ref{fig:cwlens20resimmestmswifwi} and \ref{fig:cwlens20residmestmswifwi}
 respectively.
+
+\plot{cwlens20mestmswifwi}{width=\textwidth}{a: Bulk modulus produced by 11
+  steepest descent steps to minimize the FWI objective \ref{eqn:fwi},
+  using the data shown in \ref{fig:cwd20}, and starting at the MSWI
+  result shown in Figure \ref{fig:cwlens20mestmswi}. Reduction in
+  gradient norm of 2 orders of magnitude.}
+
+\multiplot{2}{cwlens20resimmestmswifwi,cwlens20residmestmswifwi}{width=0.45\textwidth}{a: Data
+  corresponding to FWI result shown in Figure
+  \ref{fig:cwlens20mestmswifwi}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:cwlens20resimmestmswifwi} and target data
+  \ref{fig:cwd20}.}
+
+Yet another way to examine the kinematic fit between the target data
+and the resimulated data (Figure \ref{fig:cwlens20resimmestmswifwi})
+is via the adaptive filter computed with the FWI + MSWI model (Figure
+\ref{fig:cwlens20mestmswifwi}).  Figure \ref{fig:cwlens20uest1}
+shows how precisely the kinematics of the resimulation match those of
+the target data.
+
+\plot{cwlens20uest1}{width=\textwidth}{Adaptive filter to match
+  data for FWI + MSWI model (Figure \ref{fig:cwlens20resimmestmswi}) to circular lens
+  data (Figure \ref{fig:cwd20}). Dispersion away from 
+  away from $t=0$ has almost disappeared, showing that the resimulated
+  data is nearly a perfect kinematic match with the target data.}
 
@@
 the global minimum of the FWI objective function: the data match is
-very close. This type of result could perhaps have been achieved by
-systematically increasing $\alpha$, as will be discussed in the next
-section. However, the current result has a remarkable corollary,
-provided that we are willing to assume that the intermediate values of
-the FWI objective between steepest descent iterates are monotonically
-decreasing. This condition could in principle be checked
-computationally. However, for the moment let us accept that it is
-so. Then we can conclude that the homogeneous model at which the first
-FWI inversion was started must lie in a basin of attraction of a
-non-global minimizer, since the value of the objective function at the
-homogeneous model is actually {\em smaller} than that at the result of
-MSWI inversion. However the descent sequence starting at that initial
-estimate arrives at essentially zero residual. No descent path from
-the homogeneous model can
+very close. The combination of MSWI followed by FWI has fit the data
+starting from the homogeneous initial model, a result that cannot be
+reached by FWI alone.
+
+\subsection{Oblate lens: single vs. multiple arrivals}
+
+Our next two examples use another lens model, somewhat deeper (with a
+minimum bulk modulus of 2 GPa in the center), therefore more
+refractive, and extended in the
+horizontal direction. See Figure \ref{fig:cwm}. As is the case with
+the circular lens, this oblate lens is smooth to adequate numerical
+approximation.
+
+\plot{cwm}{width=\textwidth}{Oblate lens model.}
+
+The first example using this model has exactly the same acquisition
+geometry as was used in the circular lens example: 20 sources at x =
+3000 m, 201 receivers at x = 5000 m, spaced as described in the
+previous section. The data is shown in Figure \ref{fig:cowd20}.
+
+\plot{cowd20}{width=\textwidth}{Data using model shown in Figure
+  \ref{fig:cwm}, same source-receiver geometry as for circular lens
+  data (Figure \ref{fig:cwd20}).}
+
+The stronger refraction produced by this model produces a near-focus in
+the center of the source-receiver line. In fact, the ray field is
+triplicated, but the smearing effect of finite bandwidth obscures this
+detail, and the data can be (just!) regarded as exhibiting a single
+arrival on an identifiable wavefront.
+
+In this and subsequent examples, FWI initiated from the homogeneous
+model fails in the same way as shown in the  previous section. So
+there is no point in showing the results, and we don't.
+
+Application of steepest descent to the MSWI objective (with the same
+parameters as in the previoius example) results in the model depicted
+in Figure \ref{fig:nqcaustic20mestmswi}, and reduces the objective by
+almost a factor of 7 (from 2.76 $\times 10^{-2}$ to 4.1 $\times
+10^{-3}$ and the gradient norm by two full orders of magnitude. Data
+simulated with this model is shown in Figure
+\ref{fig:nqcaustic20resimmestmswi}, and the residual in Figure
+\ref{fig:nqcaustic20residmestmswi}.
+
+\plot{nqcaustic20mestmswi}{width=\textwidth}{Result of 8 MSWI
+  iterations applied to the data in Figure \ref{fig:cowd20}, beginning
+  at the homogeneous model.}
+
+\multiplot{2}{nqcaustic20resimmestmswi,nqcaustic20residmestmswi}{width=0.45\textwidth}{a: Data
+  corresponding to MSWI result shown in Figure
+  \ref{fig:nqcaustic20mestmswi}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:nqcaustic20resimmestmswi} and target data
+  \ref{fig:cowd20}.}
+
+The change in adaptive filters between the homogenous initial model
+and the MSWI inversion is shown in Figures
+\ref{fig:nqcaustic20uest0} and \ref{fig:nqcaustic20uestmswi}. The
+reduction in energy dispersion is evident.
+
+\multiplot{2}{nqcaustic20uest0,nqcaustic20uestmswi}{width=0.45\textwidth}{a:
+  adaptive filter estimate for homogeneous initial model, data of
+  Figure \ref{fig:cowd20}; b: adaptive filter estimate for MSWI
+  inversion shown in Figure \ref{fig:nqcaustic20mestmswi}.}
+
+These results render the MSWI inversion result a plausible initial
+estimate for FWI. Twelve FWI iterations applied to the data in Figure \ref{fig:cowd20}, starting at the model of
+Figure \ref{fig:nqcaustic20mestmswi}, result in the
+model shown in Figure \ref{fig:nqcaustic20mestmswifwi}, at which the
+FWI objective has decreased from $\approx 5.5$ to $\approx 0.13$, that
+is, by more than 97\%. The gradient decreases by approximately the
+same factor. This is evident in the plots of simulated data and
+residual (Figures \ref{fig:nqcaustic20resimmestmswifwi} and
+\ref{fig:nqcaustic20residmestmswifwi}). Finally, energy dispresion
+from $t=0$ has nearly disappeared from the adaptive filter computed
+from this FWI +MSWI result (Figure \ref{fig:nqcaustic20uest1}).
+
+\plot{nqcaustic20mestmswifwi}{width=\textwidth}{a: Bulk modulus produced by 12
+  steepest descent steps to minimize the FWI objective \ref{eqn:fwi},
+  using the data shown in \ref{fig:cowd20}, and starting at the MSWI
+  result shown in Figure \ref{fig:nqcaustic20mestmswi}.}
+
+\multiplot{2}{nqcaustic20resimmestmswifwi,nqcaustic20residmestmswifwi}{width=0.45\textwidth}{a: Data
+  corresponding to FWI + MSWI result shown in Figure
+  \ref{fig:nqcaustic20mestmswifwi}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:nqcaustic20resimmestmswifwi} and target data
+  \ref{fig:cowd20}.}
+
+\plot{nqcaustic20uest1}{width=\textwidth}{Adaptive filter to match
+  data for FWI + MSWI model (Figure \ref{fig:nqcaustic20resimmestmswi}) to oblate lens
+  data (Figure \ref{fig:cowd20}). Dispersion away from 
+  away from $t=0$ has almost disappeared, showing that the resimulated
+  data is nearly a perfect kinematic match with the target data.}
+
+The second example based on the target model in Figure \ref{fig:cwm}
+differs from the first {\em only} in that the sources and receivers
+have been moved further away from each other: the sources lie on the
+line $x=2000$, the receivers on $x=6000$. That is, the source-receiver
+offset has been doubled. Otherwise, all aspects of data generation are
+the same, and produce the data shown in Figure \ref{fig:fcwd20}.
+
+\plot{fcwd20}{width=\textwidth}{Data generated using the same inputs
+  as in Figure \ref{fig:cowd20}, but with sources moved to $x=2000$
+  and receivers to $x=6000$.}
+
+Notice that the additional offset has allowed distinct energetic later
+arrivals to appear in the data.
+
+As before, we compute the adaptive filter required to produce the data
+of Figure \ref{fig:fcwd20} from the corresponding homogeneous medium data (Figure
+\ref{fig:chwd20}). This filter, displayed in \ref{fig:caustic20uest0},
+shows that for most traces, two or more energy peaks appear in the
+adaptive filter, forced by the need to match multiple energy peaks in
+the target data with the single energy peak in the homogeneous medium
+data.
+
+Using the same parameters (tolerance of 0.01 for the normal residual
+in the CG iteration, 12 steps of steepest descent) produces the model
+estimate depicted in Figure \ref{fig:caustic20mestmswi}. The overall
+shape suggests decrease in bulk modulus in more or less the right
+place, but the minimum bulk modulus attained (3.15 GPa) is considerably larger
+than the target's, suggesting that the amount of refraction is
+considerably less. That guess is verified by resimulating the data
+(Figure \ref{fig:caustic20resimmestmswi}): the predicted data does not
+exhibit triplication. The computed adaptive filter (Figure 
+\ref{fig:caustic20uestmswi}) appears to attempt a compromise between
+earlier and later arrivals. Most importantly, the final gradient norm
+is approximately 5\% of the initial, suggesting that the minimizer has
+not been approximated as well as was the case in the previous
+examples. FWI initiated at the MSWI estimate gives an evidently
+cycle-skipped model estimate (Figure \ref{fig:caustic20mestmswifwi}),
+with unsatisfactory resimulation and residual (Figures
+\ref{fig:caustic20resimmestmswifwi},
+\ref{fig:caustic20residmestmswifwi}), and considerable energy spread
+ion the estimated adaptive filter (Figure \ref{fig:caustic20uest1}.
+
+\plot{caustic20uest0}{width=\textwidth}{Adaptive filter required to produce the data
+of Figure \ref{fig:fcwd20} from corresponding homogeneous medium data (Figure
+\ref{fig:chwd20}).}
+
+\plot{caustic20mestmswi}{width=\textwidth}{Result of 12 MSWI
+  iterations applied to the data in Figure \ref{fig:fcwd20}, beginning
+  at the homogeneous model.}
+
+\multiplot{2}{caustic20resimmestmswi,caustic20residmestmswi}{width=0.45\textwidth}{a: Data
+  corresponding to MSWI result shown in Figure
+  \ref{fig:caustic20mestmswi}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:caustic20resimmestmswi} and target data
+  \ref{fig:fcwd20}.}
+
+\plot{caustic20uestmswi}{width=\textwidth}{Adaptive filter estimate for MSWI
+  inversion shown in Figure \ref{fig:caustic20mestmswi}.}
+
+\plot{caustic20mestmswifwi}{width=\textwidth}{Bulk modulus produced by 12
+  steepest descent steps to minimize the FWI objective \ref{eqn:fwi},
+  using the data shown in \ref{fig:fcwd20}, and starting at the MSWI
+  result shown in Figure \ref{fig:caustic20mestmswi}.}
+
+\multiplot{2}{caustic20resimmestmswifwi,caustic20residmestmswifwi}{width=0.45\textwidth}{a: Data
+  corresponding to FWI + MSWI result shown in Figure
+  \ref{fig:caustic20mestmswifwi}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:caustic20resimmestmswifwi} and target data
+  \ref{fig:fcwd20}.}
+
+\plot{caustic20uest1}{width=\textwidth}{Adaptive filter to match
+  data for FWI + MSWI model (Figure \ref{fig:caustic20resimmestmswi}) to oblate lens
+  data (Figure \ref{fig:fcwd20}). Considerable energy dispersion away
+  from $t=0$ remains, showing that the inverted model does not match
+  data kinematics well.}
+
+As noted above, the reduction in gradient norm achieved in 12 steepest
+descent iterations for this example is not as great as was the case in
+the two previous exmaples: this appears to be a more difficult
+optimization. Accordingly, we perform 12 additional iterations of
+steepest descent. The final gradient norm is $\approx 1.1 \times
+10^{-11}$, an order of magnitude less than the final norm achieved in
+the first 12 iterations and well over two orders of magnitude less
+than the initial gradient norm of $\approx 5.24 \times 10^{-9}$. The
+depression in the center of the estimated bulk modulus has deepened
+(Figure \ref{fig:caustic20mestmswicont}) (2.53 GPa, vs 3.15 GPa at the
+end of the first 12 iterations). While it is stiil insufficiently
+refractive to produce triplication, the resulting resimulation has
+increased in magnitude for the central tracess, indicating approach to a focus, and delayed by
+roughly a half-wavelength compared to the corresponding result at the
+end of the first MSWI iteration. As the remaining figures show, FWI
+starting from this result is much more satisfactory. The latest
+arrival for the central traces still does not have quite correct
+amplitude, but all data phases are correct.
+
+\plot{caustic20mestmswicont}{width=\textwidth}{Result of 24 MSWI
+  iterations applied to the data in Figure \ref{fig:fcwd20}, beginning
+  at the homogeneous model.}
+
+\multiplot{2}{caustic20resimmestmswicont,caustic20residmestmswicont}{width=0.45\textwidth}{a: Data
+  corresponding to MSWI result shown in Figure
+  \ref{fig:caustic20mestmswicont}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:caustic20resimmestmswicont} and target data
+  \ref{fig:fcwd20}.}
+
+\plot{caustic20uestmswicont}{width=\textwidth}{Adaptive filter estimate for MSWI
+  inversion shown in Figure \ref{fig:caustic20mestmswicont}.}
+
+\plot{caustic20mestmswifwicont}{width=\textwidth}{Bulk modulus produced by 25
+  steepest descent steps to minimize the FWI objective \ref{eqn:fwi},
+  using the data shown in \ref{fig:fcwd20}, and starting at the MSWI
+  result shown in Figure \ref{fig:caustic20mestmswicont}.}
+
+\multiplot{2}{caustic20resimmestmswifwicont,caustic20residmestmswifwicont}{width=0.45\textwidth}{a: Data
+  corresponding to FWI + MSWI result shown in Figure
+  \ref{fig:caustic20mestmswifwicont}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:caustic20resimmestmswifwicont} and target data
+  \ref{fig:fcwd20}.}
+
+\plot{caustic20uest2}{width=\textwidth}{Adaptive filter to match
+  data for FWI + MSWI model (Figure \ref{fig:caustic20resimmestmswicont}) to oblate lens
+  data (Figure \ref{fig:fcwd20}). Dispersion away from 
+  away from $t=0$ has almost disappeared, showing that the resimulated
+  data is nearly a perfect kinematic match with the target data.}
+
+\subsection{Non-smooth models: the Camembert}
+
+\subsection{Noise}
+
+\section{Discussion}
+Several matters deserve attention: character of the FWI result; choice of parameters in the
+definition of the MSWI objective; and application to inverse problems
+defined by other types of wave physics.
+
+\subsection{FWI failure - local mins?}
+The result shown in Figure \ref{fig:cwlens20mestfwi0} is certainly
+unsatisfactory, as is evident from the residual plot (Figure
+\ref{fig:cwlens20residmestfwi0}). However one cannot conclude {\em
+  from this result alone} that further
+iterations of steepest descent, or of a more sophisticated local
+optimization algorithm, starting at the same homogeneous initial
+estimate, would not eventually meander to a global
+minimizer, fitting the data - most importantly, the arrival times -
+with the same sort of accuracy that FWI achieves when started at the
+MSWI result (Figure \ref{fig:cwlens20mestmswi}). Actually, further iterations don't help
+in this instance - at least, not O(10) or O(100) additional
+iterations (we tried!). On the other hand, if O($10^6$)
+additional iterations {\em would} yield a good invertion, it doesn't matter: such vast amount of computational
+work is simply not practical. Nor is it sensible, since alternatives
+exist (FWI + MSWI) that accomplish the inversion goal with far less
+work.
+
+So: we don't know whether Figure \ref{fig:cwlens20mestfwi0} represents an
+approximation to a non-global local minimizer, or not - but it doesn't matter.
+
+\subsection{Parameters}
+The roles of parameters $\sigma$ and $\alpha$ are quite
+different. Regularization, that is, avoidance of singularity in the
+normal equation \ref{eqn:normal}, is the purpose of setting $\sigma >
+0$. As formulated here, $\sigma$ is a dimensional parameter. It is
+simple to make it dimensionless, and that should be done, so that a
+``small'' choice is meaningful.
+
+The penalty parameter $\alpha$ however plays a very different role. In
+principle, $\alpha \rightarrow \infty$ should force the adaptive
+filter to become ``physical'', which in this case means the identity
+operator, i.e. with kernel $\delta(t)$ regardless of source and
+receiver, so that the penalty problem becomes identical to FWI.
+
+There are a variety of rules for setting penalty parameters like
+$\alpha$, of varying degrees of rigor. The discrepancy concept is one
+of these: it is simple to implement, auto-starting for problems like
+MSWI, and steers the penalty problem towards its limit. No magic
+however: this approach requires the assertion of data error. $\alpha$
+is adjusted so that the data error term in \ref{eqn:redfiltpen} or
+similar function is close to the posited size. Doing so is quite
+simple - see \cite{Fu:Geo17b,SymesMinkoffChen:IP22}. It might be objected that
+then one opaque parameter ($\alpha$) has been exchanged for another
+unknown one (data error), however it is even possible to turn this
+construction around to estimate data error
+\cite[]{ChenSymesMinkoff:IMAGE22}. In any case, data error has an
+obvious meaning, whether it is a known quantity or not, whereas
+$\alpha$ is merely a Lagrange multiplier.
+
+\subsection{Other Physics, Other Data}
+Acoustics is seldom exactly the right choice of physics for mechanical
+vibration modeling; in many settings, some variant of elastodynamics,
+perhaps coupled with acoustics in some regions, is the right
+choice. Elastic waves travel at several speeds, intrinsically
+providing multiple wavefronts. Extension of MSWI and similar
+approahces to elastodynamic inverse problems, either by data mode
+separation or some more intrinsic construction, would appear to be a
+very important next step, as would application to electromagnetic imaging.
+
+Sensors (and energy sources) are seldom punctual and/or
+isotropic. Actual source encountered in lab and field can exhibit
+non-trivial radiation patterns or even physical extent non-negligible
+on the wavelength scale. Inclusion of more realistic source modeling
+also seems a very important next step, even for those problems
+admitting a successful approach via source-receiver extension.
+
+As has been emphasized several times, wavefront complexity can prevent
+successful application of inversion methods based on source-receiver
+extension. A variety of other source extension approaches have been
+explored, some with positive numerical results
+\cite[]{HuangNammourSymesDollizal:SEG19}. For many of these
+approaches, the missing piece is a theoretical justification like that
+summarized in equation \ref{eqn:tomo} for MSWI.
+
+Finally, the assumption of smooth material parameter variation is
+unlikely to be justified in actual physical media. Sedimentary rocks,
+human tissue, manufactured objects all exhibit juxtaposition of
+structure at all scales, from long scales modeled by smooth variation,
+to sub-wavelength oscillations. Most likely, settings in which
+ballistic transmitted waves dominate data energetics will permit more
+or less unmodified application of techniques like MSWI, at least to
+gain a low-resolution image of wave velocity, bulk modulus,.... However
+theoretical results in this direction are absent, as are results on
+the related question of applicability of algorithms like MSWI to
+predominantly reflected wave data.
 

### 2024-05-29 repro/papers/awinum/paper.tex
examples with more extreme oblate lens

@@
 very close. The combination of MSWI followed by FWI has fit the data
-starting from the homogeneous initial model, a result that cannot be
-reached by FWI alone.
+starting from the homogeneous initial model, a result that was not obtained
+by FWI alone with the algorithmic choices made here.
 
@@
 refractive, and extended in the
-horizontal direction. See Figure \ref{fig:cwm}. As is the case with
+horizontal direction. See Figure \ref{fig:dcwm}. As is the case with
 the circular lens, this oblate lens is smooth to adequate numerical
@@
 
-\plot{cwm}{width=\textwidth}{Oblate lens model.}
+\plot{dcwm}{width=\textwidth}{Oblate lens model.}
 
@@
 3000 m, 201 receivers at x = 5000 m, spaced as described in the
-previous section. The data is shown in Figure \ref{fig:cowd20}.
+previous section. The data is shown in Figure \ref{fig:dcowd20}.
 
-\plot{cowd20}{width=\textwidth}{Data using model shown in Figure
-  \ref{fig:cwm}, same source-receiver geometry as for circular lens
+\plot{dcowd20}{width=\textwidth}{Data using model shown in Figure
+  \ref{fig:dcwm}, same source-receiver geometry as for circular lens
   data (Figure \ref{fig:cwd20}).}
 
-The stronger refraction produced by this model produces a near-focus in
-the center of the source-receiver line. In fact, the ray field is
-triplicated, but the smearing effect of finite bandwidth obscures this
-detail, and the data can be (just!) regarded as exhibiting a single
+The stronger refraction produced by this model produces a visible triplication
+the center of the source-receiver line, and in fact in all parts of
+the data, although the smearing effect of finite bandwidth obscures this
+detail in many placed. So most of the data can be (just!) regarded as exhibiting a single
 arrival on an identifiable wavefront.
@@
 parameters as in the previoius example) results in the model depicted
-in Figure \ref{fig:nqcaustic20mestmswi}, and reduces the objective by
-almost a factor of 7 (from 2.76 $\times 10^{-2}$ to 4.1 $\times
-10^{-3}$ and the gradient norm by two full orders of magnitude. Data
+in Figure \ref{fig:mnqcaustic20mestmswi}, and reduces the objective to
+roughly 20\% of its initial value, and the gradient norm to 7.5\%. Data
 simulated with this model is shown in Figure
-\ref{fig:nqcaustic20resimmestmswi}, and the residual in Figure
-\ref{fig:nqcaustic20residmestmswi}.
+\ref{fig:mnqcaustic20resimmestmswi}, and the residual in Figure
+\ref{fig:mnqcaustic20residmestmswi}.
 
-\plot{nqcaustic20mestmswi}{width=\textwidth}{Result of 8 MSWI
-  iterations applied to the data in Figure \ref{fig:cowd20}, beginning
+\plot{mnqcaustic20mestmswi}{width=\textwidth}{Result of 8 MSWI
+  iterations applied to the data in Figure \ref{fig:dcowd20}, beginning
   at the homogeneous model.}
 
-\multiplot{2}{nqcaustic20resimmestmswi,nqcaustic20residmestmswi}{width=0.45\textwidth}{a: Data
+\multiplot{2}{mnqcaustic20resimmestmswi,mnqcaustic20residmestmswi}{width=0.45\textwidth}{a: Data
   corresponding to MSWI result shown in Figure
-  \ref{fig:nqcaustic20mestmswi}. b: Residual or data error - difference between
-  data shown in Figure \ref{fig:nqcaustic20resimmestmswi} and target data
-  \ref{fig:cowd20}.}
+  \ref{fig:mnqcaustic20mestmswi}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:mnqcaustic20resimmestmswi} and target data
+  \ref{fig:dcowd20}.}
 
@@
 and the MSWI inversion is shown in Figures
-\ref{fig:nqcaustic20uest0} and \ref{fig:nqcaustic20uestmswi}. The
+\ref{fig:mnqcaustic20uest0} and \ref{fig:mnqcaustic20uestmswi}. The
 reduction in energy dispersion is evident.
 
-\multiplot{2}{nqcaustic20uest0,nqcaustic20uestmswi}{width=0.45\textwidth}{a:
+\multiplot{2}{mnqcaustic20uest0,mnqcaustic20uestmswi}{width=0.45\textwidth}{a:
   adaptive filter estimate for homogeneous initial model, data of
-  Figure \ref{fig:cowd20}; b: adaptive filter estimate for MSWI
-  inversion shown in Figure \ref{fig:nqcaustic20mestmswi}.}
+  Figure \ref{fig:dcowd20}; b: adaptive filter estimate for MSWI
+  inversion shown in Figure \ref{fig:mnqcaustic20mestmswi}.}
 
 These results render the MSWI inversion result a plausible initial
-estimate for FWI. Twelve FWI iterations applied to the data in Figure \ref{fig:cowd20}, starting at the model of
-Figure \ref{fig:nqcaustic20mestmswi}, result in the
-model shown in Figure \ref{fig:nqcaustic20mestmswifwi}, at which the
-FWI objective has decreased from $\approx 5.5$ to $\approx 0.13$, that
-is, by more than 97\%. The gradient decreases by approximately the
-same factor. This is evident in the plots of simulated data and
-residual (Figures \ref{fig:nqcaustic20resimmestmswifwi} and
-\ref{fig:nqcaustic20residmestmswifwi}). Finally, energy dispresion
+estimate for FWI. Twelve FWI iterations applied to the data in Figure \ref{fig:dcowd20}, starting at the model of
+Figure \ref{fig:mnqcaustic20mestmswi}, result in the
+model shown in Figure \ref{fig:mnqcaustic20mestmswifwi}, at which the
+FWI objective has decreased to 4\% of its initial value. That is, the
+RMS error between data and predicted data has been reduced to
+approximately 20\% of its initial value; the gradient
+decreases by a slightly larger factor. The reduction obtained is evident in the plots of simulated data and
+residual (Figures \ref{fig:mnqcaustic20resimmestmswifwi} and
+\ref{fig:mnqcaustic20residmestmswifwi}). Residual energy is
+concentrated in later arrivals in the
+central zone, suggesting that the model shown in Figure
+\ref{fig:mnqcaustic20mestmswifwi} may lie in the basin of global minimization. Finally, energy dispresion
 from $t=0$ has nearly disappeared from the adaptive filter computed
-from this FWI +MSWI result (Figure \ref{fig:nqcaustic20uest1}).
+from this FWI +MSWI result, again except for some traces in the
+central zone (Figure \ref{fig:mnqcaustic20uest1}).
 
-\plot{nqcaustic20mestmswifwi}{width=\textwidth}{a: Bulk modulus produced by 12
+\plot{mnqcaustic20mestmswifwi}{width=\textwidth}{a: Bulk modulus produced by 12
   steepest descent steps to minimize the FWI objective \ref{eqn:fwi},
-  using the data shown in \ref{fig:cowd20}, and starting at the MSWI
-  result shown in Figure \ref{fig:nqcaustic20mestmswi}.}
+  using the data shown in \ref{fig:dcowd20}, and starting at the MSWI
+  result shown in Figure \ref{fig:mnqcaustic20mestmswi}.}
 
-\multiplot{2}{nqcaustic20resimmestmswifwi,nqcaustic20residmestmswifwi}{width=0.45\textwidth}{a: Data
+\multiplot{2}{mnqcaustic20resimmestmswifwi,mnqcaustic20residmestmswifwi}{width=0.45\textwidth}{a: Data
   corresponding to FWI + MSWI result shown in Figure
-  \ref{fig:nqcaustic20mestmswifwi}. b: Residual or data error - difference between
-  data shown in Figure \ref{fig:nqcaustic20resimmestmswifwi} and target data
-  \ref{fig:cowd20}.}
+  \ref{fig:mnqcaustic20mestmswifwi}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:mnqcaustic20resimmestmswifwi} and target data
+  \ref{fig:dcowd20}.}
 
-\plot{nqcaustic20uest1}{width=\textwidth}{Adaptive filter to match
-  data for FWI + MSWI model (Figure \ref{fig:nqcaustic20resimmestmswi}) to oblate lens
-  data (Figure \ref{fig:cowd20}). Dispersion away from 
+\plot{mnqcaustic20uest1}{width=\textwidth}{Adaptive filter to match
+  data for FWI + MSWI model (Figure \ref{fig:mnqcaustic20resimmestmswi}) to oblate lens
+  data (Figure \ref{fig:dcowd20}). Dispersion away from 
   away from $t=0$ has almost disappeared, showing that the resimulated
@@
 
-The second example based on the target model in Figure \ref{fig:cwm}
+The second example based on the target model in Figure \ref{fig:dcwm}
 differs from the first {\em only} in that the sources and receivers
@@
 offset has been doubled. Otherwise, all aspects of data generation are
-the same, and produce the data shown in Figure \ref{fig:fcwd20}.
+the same, and produce the data shown in Figure \ref{fig:dcwd20}.
 
-\plot{fcwd20}{width=\textwidth}{Data generated using the same inputs
-  as in Figure \ref{fig:cowd20}, but with sources moved to $x=2000$
+\plot{dcwd20}{width=\textwidth}{Data generated using the same inputs
+  as in Figure \ref{fig:dcowd20}, but with sources moved to $x=2000$
   and receivers to $x=6000$.}
@@
 As before, we compute the adaptive filter required to produce the data
-of Figure \ref{fig:fcwd20} from the corresponding homogeneous medium data (Figure
-\ref{fig:chwd20}). This filter, displayed in \ref{fig:caustic20uest0},
-shows that for most traces, two or more energy peaks appear in the
+of Figure \ref{fig:dcwd20} from the corresponding homogeneous medium data (Figure
+\ref{fig:chwd20}). This filter, displayed in \ref{fig:mcaustic20uest0},
+shows that for most traces, two or more major energy peaks appear in the
 adaptive filter, forced by the need to match multiple energy peaks in
@@
 Using the same parameters (tolerance of 0.01 for the normal residual
-in the CG iteration, 12 steps of steepest descent) produces the model
-estimate depicted in Figure \ref{fig:caustic20mestmswi}. The overall
-shape suggests decrease in bulk modulus in more or less the right
-place, but the minimum bulk modulus attained (3.15 GPa) is considerably larger
-than the target's, suggesting that the amount of refraction is
+in the CG iteration, same smoothing operator as inverse weight
+defining norm in model space) 12 steps of steepest descent applied to
+the MSWI objective produces the model
+estimate depicted in Figure \ref{fig:mcaustic20mestmswi}. The overall
+decrease in bulk modulus is correct, but the minimum bulk modulus attained (2.8 GPa) is considerably larger
+than the target's, and moreover is concentrated around the source
+line, suggesting that the amount of refraction produced by this model is
 considerably less. That guess is verified by resimulating the data
-(Figure \ref{fig:caustic20resimmestmswi}): the predicted data does not
+(Figure \ref{fig:mcaustic20resimmestmswi}): the predicted data does not
 exhibit triplication. The computed adaptive filter (Figure 
-\ref{fig:caustic20uestmswi}) appears to attempt a compromise between
+\ref{fig:mcaustic20uestmswi}) appears to attempt a compromise between
 earlier and later arrivals. Most importantly, the final gradient norm
-is approximately 5\% of the initial, suggesting that the minimizer has
+is approximately 18\% of the initial, suggesting that the minimizer has
 not been approximated as well as was the case in the previous
 examples. FWI initiated at the MSWI estimate gives an evidently
-cycle-skipped model estimate (Figure \ref{fig:caustic20mestmswifwi}),
+cycle-skipped model estimate (Figure \ref{fig:mcaustic20mestmswifwi}),
 with unsatisfactory resimulation and residual (Figures
-\ref{fig:caustic20resimmestmswifwi},
-\ref{fig:caustic20residmestmswifwi}), and considerable energy spread
-ion the estimated adaptive filter (Figure \ref{fig:caustic20uest1}.
+\ref{fig:mcaustic20resimmestmswifwi},
+\ref{fig:mcaustic20residmestmswifwi}), and considerable energy spread
+ion the estimated adaptive filter (Figure \ref{fig:mcaustic20uest1}.
 
-\plot{caustic20uest0}{width=\textwidth}{Adaptive filter required to produce the data
-of Figure \ref{fig:fcwd20} from corresponding homogeneous medium data (Figure
+\plot{mcaustic20uest0}{width=\textwidth}{Adaptive filter required to produce the data
+of Figure \ref{fig:dcwd20} from corresponding homogeneous medium data (Figure
 \ref{fig:chwd20}).}
 
-\plot{caustic20mestmswi}{width=\textwidth}{Result of 12 MSWI
-  iterations applied to the data in Figure \ref{fig:fcwd20}, beginning
+\plot{mcaustic20mestmswi}{width=\textwidth}{Result of 12 MSWI
+  iterations applied to the data in Figure \ref{fig:dcwd20}, beginning
   at the homogeneous model.}
 
-\multiplot{2}{caustic20resimmestmswi,caustic20residmestmswi}{width=0.45\textwidth}{a: Data
+\multiplot{2}{mcaustic20resimmestmswi,mcaustic20residmestmswi}{width=0.45\textwidth}{a: Data
   corresponding to MSWI result shown in Figure
-  \ref{fig:caustic20mestmswi}. b: Residual or data error - difference between
-  data shown in Figure \ref{fig:caustic20resimmestmswi} and target data
-  \ref{fig:fcwd20}.}
+  \ref{fig:mcaustic20mestmswi}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:mcaustic20resimmestmswi} and target data
+  \ref{fig:dcwd20}.}
 
-\plot{caustic20uestmswi}{width=\textwidth}{Adaptive filter estimate for MSWI
-  inversion shown in Figure \ref{fig:caustic20mestmswi}.}
+\plot{mcaustic20uestmswi}{width=\textwidth}{Adaptive filter estimate for MSWI
+  inversion shown in Figure \ref{fig:mcaustic20mestmswi}.}
 
-\plot{caustic20mestmswifwi}{width=\textwidth}{Bulk modulus produced by 12
+\plot{mcaustic20mestmswifwi}{width=\textwidth}{Bulk modulus produced by 12
   steepest descent steps to minimize the FWI objective \ref{eqn:fwi},
-  using the data shown in \ref{fig:fcwd20}, and starting at the MSWI
-  result shown in Figure \ref{fig:caustic20mestmswi}.}
+  using the data shown in \ref{fig:dcwd20}, and starting at the MSWI
+  result shown in Figure \ref{fig:mcaustic20mestmswi}.}
 
-\multiplot{2}{caustic20resimmestmswifwi,caustic20residmestmswifwi}{width=0.45\textwidth}{a: Data
+\multiplot{2}{mcaustic20resimmestmswifwi,mcaustic20residmestmswifwi}{width=0.45\textwidth}{a: Data
   corresponding to FWI + MSWI result shown in Figure
-  \ref{fig:caustic20mestmswifwi}. b: Residual or data error - difference between
-  data shown in Figure \ref{fig:caustic20resimmestmswifwi} and target data
-  \ref{fig:fcwd20}.}
+  \ref{fig:mcaustic20mestmswifwi}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:mcaustic20resimmestmswifwi} and target data
+  \ref{fig:dcwd20}.}
 
-\plot{caustic20uest1}{width=\textwidth}{Adaptive filter to match
-  data for FWI + MSWI model (Figure \ref{fig:caustic20resimmestmswi}) to oblate lens
-  data (Figure \ref{fig:fcwd20}). Considerable energy dispersion away
+\plot{mcaustic20uest1}{width=\textwidth}{Adaptive filter to match
+  data for FWI + MSWI model (Figure \ref{fig:mcaustic20resimmestmswi}) to oblate lens
+  data (Figure \ref{fig:dcwd20}). Considerable energy dispersion away
   from $t=0$ remains, showing that the inverted model does not match
@@
 
-As noted above, the reduction in gradient norm achieved in 12 steepest
+As noted above, the reduction in MSWI gradient norm achieved in 12 steepest
 descent iterations for this example is not as great as was the case in
@@
 optimization. Accordingly, we perform 12 additional iterations of
-steepest descent. The final gradient norm is $\approx 1.1 \times
-10^{-11}$, an order of magnitude less than the final norm achieved in
-the first 12 iterations and well over two orders of magnitude less
-than the initial gradient norm of $\approx 5.24 \times 10^{-9}$. The
-depression in the center of the estimated bulk modulus has deepened
-(Figure \ref{fig:caustic20mestmswicont}) (2.53 GPa, vs 3.15 GPa at the
-end of the first 12 iterations). While it is stiil insufficiently
-refractive to produce triplication, the resulting resimulation has
-increased in magnitude for the central tracess, indicating approach to a focus, and delayed by
-roughly a half-wavelength compared to the corresponding result at the
-end of the first MSWI iteration. As the remaining figures show, FWI
-starting from this result is much more satisfactory. The latest
-arrival for the central traces still does not have quite correct
-amplitude, but all data phases are correct.
+steepest descent applied to the MSWI objective, starting withe the
+result (Figure \ref{fig:mcaustic20mestmswi}) of the first 12
+iterations. The objective decreases by roughly 0.8, the gradient by
+0.5. Inspection of the model so obtained (Figure
+\ref{fig:mcaustic20mestmswicont}) shows that the minimum bulk modulus
+has decreased (2.09 GPa, vs 2.8  GPa at the
+end of the first 12 iterations), however the smallest values are even
+more concentrated around the source locations, and the resimulated
+data with this model is if anything less similar to the data than was
+the case at the end of the first 12 iterations (Figures
+\ref{fig:mcaustic20resimmestmswicont},
+\ref{fig:mcaustic20residmestmswicont}). The adaptive filter is no more
+focused at $t=0$ than was the case at the end of the first 12
+iterations: it appears to have adjusted the compromise between shifts
+to fit early and late arrivals, nothing more (Figure \ref{fig:mcaustic20uestmswicont}).
 
-\plot{caustic20mestmswicont}{width=\textwidth}{Result of 24 MSWI
-  iterations applied to the data in Figure \ref{fig:fcwd20}, beginning
+\plot{mcaustic20mestmswicont}{width=\textwidth}{Result of 24 MSWI
+  iterations applied to the data in Figure \ref{fig:dcwd20}, beginning
   at the homogeneous model.}
 
-\multiplot{2}{caustic20resimmestmswicont,caustic20residmestmswicont}{width=0.45\textwidth}{a: Data
+\multiplot{2}{mcaustic20resimmestmswicont,mcaustic20residmestmswicont}{width=0.45\textwidth}{a: Data
   corresponding to MSWI result shown in Figure
-  \ref{fig:caustic20mestmswicont}. b: Residual or data error - difference between
-  data shown in Figure \ref{fig:caustic20resimmestmswicont} and target data
-  \ref{fig:fcwd20}.}
+  \ref{fig:mcaustic20mestmswicont}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:mcaustic20resimmestmswicont} and target data
+  \ref{fig:dcwd20}.}
 
-\plot{caustic20uestmswicont}{width=\textwidth}{Adaptive filter estimate for MSWI
-  inversion shown in Figure \ref{fig:caustic20mestmswicont}.}
+\plot{mcaustic20uestmswicont}{width=\textwidth}{Adaptive filter estimate for MSWI
+  inversion shown in Figure \ref{fig:mcaustic20mestmswicont}.}
 
-\plot{caustic20mestmswifwicont}{width=\textwidth}{Bulk modulus produced by 25
+This result (Figure \ref{fig:mcaustic20mestmswicont}) is actually a
+{\em worse} initial estimate for FWI than was the model obtained after
+12 steepet descent iterations. We
+performed 25 steps of steepest descent applied to the FWI objective
+with the model in Figure \ref{fig:mcaustic20mestmswicont} as initial
+estimate, resulting in the obviously cycle-skipped model displayed in
+Figure \ref{fig:mcaustic20mestmswifwicont}. The residual (Figure
+\ref{fig:mcaustic20residmestmswifwicont} and adaptive filter (Figure
+\ref{fig:mcaustic20uest2}) have actually deteriorated relative to
+those obtained before.
+
+\plot{mcaustic20mestmswifwicont}{width=\textwidth}{Bulk modulus produced by 25
   steepest descent steps to minimize the FWI objective \ref{eqn:fwi},
-  using the data shown in \ref{fig:fcwd20}, and starting at the MSWI
-  result shown in Figure \ref{fig:caustic20mestmswicont}.}
+  using the data shown in \ref{fig:dcwd20}, and starting at the MSWI
+  result shown in Figure \ref{fig:mcaustic20mestmswicont}.}
 
-\multiplot{2}{caustic20resimmestmswifwicont,caustic20residmestmswifwicont}{width=0.45\textwidth}{a: Data
+\multiplot{2}{mcaustic20resimmestmswifwicont,mcaustic20residmestmswifwicont}{width=0.45\textwidth}{a: Data
   corresponding to FWI + MSWI result shown in Figure
-  \ref{fig:caustic20mestmswifwicont}. b: Residual or data error - difference between
-  data shown in Figure \ref{fig:caustic20resimmestmswifwicont} and target data
-  \ref{fig:fcwd20}.}
+  \ref{fig:mcaustic20mestmswifwicont}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:mcaustic20resimmestmswifwicont} and target data
+  \ref{fig:dcwd20}.}
 
-\plot{caustic20uest2}{width=\textwidth}{Adaptive filter to match
-  data for FWI + MSWI model (Figure \ref{fig:caustic20resimmestmswicont}) to oblate lens
-  data (Figure \ref{fig:fcwd20}). Dispersion away from 
+\plot{mcaustic20uest2}{width=\textwidth}{Adaptive filter to match
+  data for FWI + MSWI model (Figure \ref{fig:mcaustic20resimmestmswicont}) to oblate lens
+  data (Figure \ref{fig:dcwd20}). Dispersion away from 
   away from $t=0$ has almost disappeared, showing that the resimulated
   data is nearly a perfect kinematic match with the target data.}
+
+The meaning of this result will be discussed further below. For now
+merely note that the first example based on the oblate lens model in
+Figure \ref{fig:dcwm} appeared to require more computational effort
+than the cirular lens example to achieve a successful inversion,
+whereas the second was an inversion failure even with considerably
+more effort.
 

### 2024-06-02 repro/papers/awinum/paper.tex
added cam

@@
 
-\subsection{Non-smooth models: the Camembert}
+\subsection{Recovering a Camembert: effect of non-smoothness}
+
+\plot{cambulk}{width=\textwidth}{A model resembling the ``camembert''
+  of \cite{GauTarVir:86}. Diameter of circular inclusion = 2500 m,
+  approximately 8 wavelengths at median frequency. Interior value 20\%
+above exterior (4.8 GPa vs 4.0 GPa).}
+
+\plot{camcwd20}{width=\textwidth}{Data generated using ``camembert''
+  model (Figure \ref{fig:cambulk}), same source-receiver geometry as
+  in Figure \ref{fig:dcwd20}: 20 sources located at $x=2000$ m, spaced
+  150 m apart, starting at $z=500$ m; 201 receivers located at $x=6000$ m,
+  spaced 20 m apart, starting at $z=200$ m.}
+
+\plot{cam20uest0}{width=\textwidth}{Adaptive filter required to produce the data
+of Figure \ref{fig:camcwd20} from corresponding homogeneous medium data (Figure
+\ref{fig:chwd20}).}
+
+\plot{cam20mestmswi}{width=\textwidth}{Result of 12 MSWI
+  iterations applied to the data in Figure \ref{fig:camcwd20}, beginning
+  at the homogeneous model.}
+
+\multiplot{2}{cam20resimmestmswi,cam20residmestmswi}{width=0.45\textwidth}
+{a: Data  corresponding to MSWI result shown in Figure
+  \ref{fig:cam20mestmswi}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:cam20resimmestmswi} and target data
+  \ref{fig:camcwd20}.}
+
+\plot{cam20uestmswi}{width=\textwidth}{Adaptive filter estimate for MSWI
+  inversion shown in Figure \ref{fig:cam20mestmswi}.}
+
+\plot{cam20mestmswifwi}{width=\textwidth}{Bulk modulus produced by 12
+  steepest descent steps to minimize the FWI objective \ref{eqn:fwi},
+  using the data shown in \ref{fig:camcwd20}, and starting at the MSWI
+  result shown in Figure \ref{fig:cam20mestmswi}.}
+
+\multiplot{2}{cam20resimmestmswifwi,cam20residmestmswifwi}{width=0.45\textwidth}{a: Data
+  corresponding to FWI + MSWI result shown in Figure
+  \ref{fig:cam20mestmswifwi}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:cam20resimmestmswifwi} and target data
+  \ref{fig:camcwd20}.}
+
+\plot{cam20uest1}{width=\textwidth}{Adaptive filter to match
+  data for FWI + MSWI model (Figure \ref{fig:cam20resimmestmswi}) to ``camembert'' 
+  data (Figure \ref{fig:camcwd20}).}
+
+\plot{cam20mestmswifwifine}{width=\textwidth}{``Hi-res'' bulk modulus produced by
+  25 additional steepest descent steps, with inverse weight operator =
+  2 point moving average in both directions, repeated once, and
+  initial bulk modulus = output of first FWI + MSWI cycle (Figure
+  \ref{fig:cam20mestmswifwi}). Note that top and bottom edges of disk
+  shown in Figure \ref{fig:cambulk} are clearly visible.}
+
+\multiplot{2}{cam20resimmestmswifwifine,cam20residmestmswifwifine}{width=0.45\textwidth}{a: Data
+  corresponding to hi-res FWI + MSWI result shown in Figure
+  \ref{fig:cam20mestmswifwifine}. b: Residual or data error - difference between
+  data shown in Figure \ref{fig:cam20resimmestmswifwifine} and target data
+  \ref{fig:camcwd20}.}
+
+\plot{cam20uest2}{width=\textwidth}{Adaptive filter to match
+  data for hi-res FWI + MSWI model (Figure \ref{fig:cam20resimmestmswi}) to ``camembert'' 
+  data (Figure \ref{fig:camcwd20}).}
 

### 2024-07-23 repro/papers/awi/paper.tex
penalty stuff added

@@
 at its surface. Over the last several decades, this task has come to
-be carried out by minimizing the difference between computationa
-prediction of wave measurement versus actual observed
-measurements. The mechanical parameters to be inferred appear as
+be carried out by minimizing a measure of the difference between computational
+prediction of wave data versus actual observed
+data. The mechanical parameters to be inferred appear as
 coefficients in systems of partial differential equations governing
@@
 
-While heuristic justifications for AWI (and
-many details of implementation) are discussed in these and other
-works, a clear explanation for its effectiveness is 
+While these and other works discuss heuristic justifications for AWI and
+many details of implementation, a clear explanation for its effectiveness is 
 so far missing from the literature. In this short note, I show that
@@
 (that is, unique rays of geometric asymptotics between sources and
-receivers), the AWI objective function is an approximation to the
+receivers), the AWI objective function is an asymptotic (in time
+frequency) approximation to the
 mean-square time-of-travel misfit. Thus AWI and travel time fitting
@@
 note that a normalization component of the AWI objective function
-definition is very
+definition is 
 important in reaching this conclusion. The connection between
-objective normalization and the close relation to travel time
-inversion has nothing to do with
-the reasons for normalizing the objective advanced in the
-literature \cite[]{Warner:16,LiAlkhalifah:21}.
+objective normalization and the asymptotic equivalence to travel time
+inversion for transmission data has nothing to do with
+the reasons for normalizing the objective previoiusly advanced in the
+literature \cite[]{Warner:16,LiAlkhalifah:21}. Another modification of
+FWI based on similar ideas, Matched Source Waveform Inversion (MSWI)
+(\cite{HuangSymes2015SEG,HuangSymes:Geo17}) lacks the AWI normalization,
+and has a looser relation to travel time inversion. The arguments
+presented here for AWI (and MSWI) are refinements of those given by
+\cite{HuangSymes:Geo17}. which were in turn based on ideas presented in Song's thesis
+\cite[]{Song:94c}.
 
-After establishing
-the class of problems to be studied and notations for their
-description in the following section, I present the asymptotic
-analysis leading to the main conclusion in the next. The close relation
-between AWI objective and travel time misfit no longer mecessarily
-holds if multiple rays connect source and receiver. I discuss these
-and other aspects of AWI and related approaches in a brief final section.
+After establishing the class of problems to be studied and notations
+for their description in the following section, I present the
+asymptotic analysis leading to the main conclusion in the next. In
+effect, AWI regularizes or relaxes the FWI problem. For many inverse
+problems, regularization has been accomplished via additive penalty
+functions. In section four, I explain the relation between AWI and a
+penalty formulation of the wave inverse problem. The
+close relation between AWI objective and travel time misfit is limited
+in scope: in the fifth section, I show that it does not hold in
+general for transmitted wave data in which multiple rays connect
+source and receiver.
+
+The analysis explained here sheds no light on the effectiveness of AWI
+for reflected wave data, and the approach has so far been limited
+inversion of data modeled by acoustic wave propagation. I discuss
+these and other aspects of AWI and related approaches in a brief final
+section.
 
@@
 adaptive filter $u[m;d]$ with $\bar{w}=w*u[m;d]$, for which
-$S[m]u[m;d]= d$. [This proposition is dubious, but is easy to work
+$S[m]u[m;d]= d$. This proposition is dubious, but is easy to work
 around, as will be seen below. Also, \cite{Warner:16} also formulate
@@
 $u*d = F[m]w$, and in fact recommend that definition. The present
-discussion uses only the forward filter.] If
-$\kappa = \kappa^*, \beta = \beta^*$, then $\bar{w}$ should be $= w$,
+discussion uses only the forward filter.
+
+If $\kappa = \kappa^*, \beta = \beta^*$, then $\bar{w}$ should be $= w$,
 so the adaptive kernel $u[m;d]$ should be $=\delta(t)$ for all
@@
 derivative of another wavelet $v$, also transient ($v(t;{\bf x}_s)=0$
-for $|t| \gg 0$). Set $\lambda(w) = \|v\|/\|w\|$. $\lambda$ is a proxy for wavelength: it is the reciprocal of RMS frequency.
+for $|t| \gg 0$). Set $\lambda(w) = \|v\|/\|w\|$. $\lambda$ is a proxy
+for wavelength: it is the reciprocal of RMS frequency of $v$.
 
@@
 \end{equation}
-There is immediately a big problem: this $u[m;d]$ is not square-integrable, and the AWI objective $J_{\rm AWI}$ is undefined.
-To fix this, it's necessary to re-define $u[m;d]$ as the solution of a regularized least squares problem: define
+There is immediately a big problem: this $u[m;d]$ is not
+square-integrable, and the AWI objective $J_{\rm AWI}$ is
+undefined. To fix this, it's necessary to replace $u[m;d]$ by
+$u_{0,\sigma}[m;d]$,  the solution of a regularized least squares problem: 
 \begin{equation}
   \label{eqn:reg}
-  u[m;d] = \mbox{argmin}_u (\|S[m]u-d\|^2 + \sigma^2\|u\|^2)
+  u_{0,\sigma}[m;d] = \mbox{argmin}_u (\|S[m]u-d\|^2 + \sigma^2\|u\|^2)
 \end{equation}
@@
 
-With these conventions, re-write equation \ref{eqn:reg} in terms of the Fourier transforms $\hat{u}[m;d],\hat{w}$ of $u[m;d]$ and $w$:
+With these conventions, re-write equation \ref{eqn:reg} in terms of the Fourier transforms $\hat{u}[m;d],\hat{w}$ of $u_{0,\sigma}[m;d]$ and $w$:
 \begin{equation}
@@
 The normal equation for the least squares problem \ref{eqn:freqreg} is
-$$
+\[
 (a^2 |\hat{w}|^2 +\sigma^2)\hat{u} = a a^*|\hat{w}|^2e^{i\omega(\tau^*-\tau)}
-$$
+\]
 the solution of which is
-$$
-\hat{u}[m;d] \approx \frac{a^*}{a}\hat{g}_{\frac{a^*}{a}\sigma} e^{i\omega(\tau^*-\tau)}
-$$
+\[
+\hat{u}[m;d] \approx \frac{a^*}{a}\hat{g}_{\frac{\sigma}{a}} e^{i\omega(\tau^*-\tau)}
+\]
 where
 $$ 
-\hat{g}_{\sigma} = \frac{(a^*)^2|\hat{w}|^2}{(a^*)^2|\hat{w}|^2 + \sigma^2}.
+\hat{g}_{\sigma} = \frac{|\hat{w}|^2}{|\hat{w}|^2 + \sigma^2}.
 $$
-depends only the ``true'' model $m^*$.
-Transient acoustic sources are effectively bandlimited: $\hat{w}(\omega) \rightarrow 0$ faster than any negative power of $\omega$ as $|\omega| \rightarrow \infty$ and is analytic, whence $\hat{g}_{\sigma}:$ tends to $1$ almost everywhere as $\sigma \rightarrow 0$.
+Transient acoustic sources are effectively bandlimited:
+$\hat{w}(\omega) \rightarrow 0$ faster than any negative power of
+$\omega$ as $|\omega| \rightarrow \infty$ and is entire analytic, whence $\hat{g}_{\sigma}:$ tends to $1$ almost everywhere as $\sigma \rightarrow 0$.
 Hence the inverse Fourier transform $g_{\sigma}$ tends to $\delta$ in the sense of distributions, and
 $$
-u[m;d](t) \approx \frac{a^*}{a}g_{\frac{a^*}{a}\sigma}(t-(\tau^*-\tau))
+u_{0,\sigma}[m;d](t) \approx \frac{a^*}{a}g_{\frac{\sigma}{a}}(t-(\tau^*-\tau))
 $$
 to a multiple of a shifted $\delta$, in fact exactly the distribution
-solution of $\bar{S}[m,w]u=d$ given in equation \ref{eqn:notl2}. However for $\sigma>0$, $u[m;d]$ is square integrable.
+solution of $\bar{S}[m,w]u=d$ given in equation \ref{eqn:notl2}. However for $\sigma>0$, $u_{0,\sigma}[m;d]$ is square integrable.
 
@@
 \cite[]{Song:94c}. MSWI deals with the extended source $\bar{w}$, but
-the same construction applies to the adptive filter $u$. With $u$ replaced by $u[m;d]$,  this MSWI objective is
+the same construction applies to the adaptive filter $u$. With $u$ replaced by $u_{0,\sigma}[m;d]$,  the MSWI objective is
 \[
-J_{\rm MSWI}[m;d] = \int dt |tu[m;d]|^2 \approx \frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\sigma}(t-(\tau^*-\tau))|^2
+J_{\rm MSWI}[m;d] = \int dt |tu_{0,\sigma}[m;d]|^2 \approx \frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\frac{\sigma}{a}}(t-(\tau^*-\tau))|^2
 \]
 \[
-=\frac{(a^*)^2}{a^2} \int dt\, (t+(\tau^*-\tau))^2|g_{\frac{a^*}{a}\sigma}(t)|^2
+=\frac{(a^*)^2}{a^2} \int dt\, (t+(\tau^*-\tau))^2|g_{\frac{\sigma}{a}}(t)|^2
 \]
@@
   =\frac{(a^*)^2}{a^2} \int dt\, (t^2 + 2t
-  (\tau^*-\tau)+(\tau^*-\tau)^2)|g_{\frac{a^*}{a}\sigma}(t)|^2
+  (\tau^*-\tau)+(\tau^*-\tau)^2)|g_{\frac{\sigma}{a}}(t)|^2
 \end{equation}
@@
   J_{\rm MSWI}[m;d] \approx
-\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\sigma}(t)|^2
-+(\tau^*-\tau)^2\frac{(a^*)^2}{a^2}\|g_{\frac{a^*}{a}\sigma}\|^2
+\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\frac{\sigma}{a}}(t)|^2
++(\tau^*-\tau)^2\frac{(a^*)^2}{a^2}\|g_{\frac{\sigma}{a}}\|^2
 \end{equation}
@@
 
-The relation is obscured by the presence of of the amplitude ratio
-$a^*/a$ both as a multiplier and in the scaling of $\sigma$. Since the
+The relation is obscured by the presence of of the amplitudes
+$a^*, a$ both in the multiplier and in the scaling of $\sigma$. Since the
 amplitudes depend on $m$, their presence may influence the location
@@
 \ref{eqn:MSWI} of $J_{\rm MSWI}$, ones sees
-that the two functions differ only by the factor $\|u[m;d]\|^{-2}$ in
+that the two functions differ only by the factor $\|u_{0,\sigma}[m;d]\|^{-2}$ in
 the former. Using the geometric asymptotics approximation, 
 $$
-\|u[m;d]\| \approx \frac{a^*}{a}\|g_{\frac{a^*}{a}\sigma}\| + O(\sigma^2)
+\|u_{0,\sigma}[m;d]\| \approx \frac{a^*}{a}\|g_{\frac{\sigma}{a}}\| + O(\sigma^2)
 $$
-Therefore, multiplication of the MSWI objective by $\|u[m;d]\|^{-2}$
+Therefore, multiplication of the MSWI objective by $\|u_{0,\sigma}[m;d]\|^{-2}$
 precisely removes the amplitude factor in the right-hand side of the
@@
 given by equation \ref{eqn:hadamard} is an accurate approximation to the Green's function.
+
+\section{Relation to Penalty Formulation}
+Penalty functions provide a means to convert naively posed
+inverse problems into well-posed problems, with solutions that exist
+and depend stably on proper data. The ``pre-whitening'' construction
+explained in the last section is a classic example, adding a scaled
+mean-square of the adaptive filter to the residual mean-square to
+produce an objective with a square-integrable (finite-energy)
+minimizer. This penalty construction if often called {\em Tihonov
+  regularization}. The MSWI and AWI objective functions have a deeper
+relation to a penalty construction, however: they are
+suitably defined limits of penalty functions.
+
+This relation is abstract, and does not just pertain to the problems
+discussed in this paper. To derive this abstract relation, suppose for
+the moment that $S:U \rightarrow D$ is a bounded and coercive from a
+domain Hilbert space $U$ to a range Hilbert space $D$. That is, for
+real numbers $0 < C_* \le C^*$,
+\[
+  C_*\|u\|_U \le \|Su\|_D \le C^*\|u\|
+\]
+Suppose that $d \in D$ (``data''), and $T$ is another bounded operator $U \rightarrow E$, $E$
+another Hilbert space. Define
+\begin{equation}
+  \label{eqn:eq1}
+  J_{\alpha}(u) = \frac{1}{2}(\|Su-d\|_D + \alpha^2\|Tu\|_E^2).
+\end{equation}
+and
+\begin{equation}
+  \label{eqn:eq0}
+  \tJa= \min_u J_{\alpha}(u) = J_{\alpha}(u_{\alpha})
+\end{equation}
+Since $S$ is coercive, the minimum is well-defined for any
+$\alpha \ge 0$, and the minimizer $\ua \in U$ satisfies the normal
+equation
+\begin{equation}
+  \label{eqn:eq0n}
+  (S^TS + \alpha^2 T^TT) \ua = S^Td
+\end{equation}
+
+The claim to be established is that
+\begin{equation}
+  \label{eqn:eq4}
+  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tJa-\tJz).= \frac{1}{2}\|T\uz\|^2
+\end{equation}
+
+To see this, use the normal equation to write
+\[
+  \ua = (S^TS + \alpha^2 T^TT)^{-1}T^Td = (S^TS + \alpha^2 T^TT)^{-1}S^TS \uz
+\]
+\begin{equation}
+  \label{eqn:eq2}
+  = \uz - \alpha^2 (S^TS + \alpha^2 T^TT)^{-1}T^TT\uz = \uz-\alpha^2 \va
+\end{equation}
+Note that $\va = (S^TS + \alpha^2 T^TT)^{-1}T^TT\uz$ is uniformly bounded in $\alpha \ge 0$.
+
+Substitute the RHS of equation \ref{eqn:eq2} into the definition \ref{eqn:eq1} to obtain
+\begin{equation}
+  \label{eqn:eq3}
+  \tJa = \frac{1}{2}(\|S\uz-d\|_d^2 - 2 \alpha^2\langle S\uz-d, S\va\rangle_d + \alpha^2\|T\uz\|_m^2 + O(\alpha^4))
+\end{equation}
+The second term on the RHS of equation \ref{eqn:eq3} vanishes thanks to the normal equation, and the first term is precisely $\tJz$. The conclusion \ref{eqn:eq4} follows.
+
+To apply the assertion \ref{eqn:eq4} to the MSWI objective, make the
+following choices, for each $m$ in the domain of $F$: $S$ is the regularized
+modeling operator $(S[m],\sigma I)^T$.  Its domain is $U=L^2(\bR)$, of
+which $u$ (the adaptive filter) is a member, and 
+its range is $D = L^2(\bR) \oplus
+L^2(\bR)$. The role of $d$ is played  the
+augmented data $(d,0)^T$. $T$ is a modified multiply-by-$t$ operator
+$Tu(t)=\max(|t|, t_{\rm max}) \mbox{sgn}(t)u(t)$, with domain $U$ and
+range $E=L^2(\bR)$. The truncation by $t_{\rm max} > 0$ is necessary
+to make $T$ bounded, but can be chosen large enough relative to the
+support of the filter (necessarily bounded in practical calculation)
+to have no effect. Then $\Ja$ becomes $J_{{\rm MSWI},\alpha,\sigma}[m,u;d]$, where
+\begin{equation}
+  \label{eqn:eq5}
+   J_{{\rm MSWI},\alpha,\sigma}[m,u;d] = \frac{1}{2}\left(\|S[m]u-d\|^2 +
+   \sigma^2\|u\|^2 + \alpha^2\|Tu\|^2\right)
+ \end{equation}
+The minimizer $\ua$ over $u$ of $ J_{{\rm MSWI}, \alpha,\sigma}[m,u;d]$ is
+$u_{{\rm MSWI},\alpha,\sigma}[m;d]$, and $\tJa$ becomes
+\begin{equation}
+  \label{eqn:eq6}
+  \tilde{J}_{{\rm MSWI},\alpha,\sigma}[m;d] =
+  J_{{\rm MSWI},\alpha,\sigma}[m,u_{{\rm MSWI},\alpha,\sigma}[m;d];d].
+\end{equation}
+Note that $u_0$ becomes $u_{0,\sigma}[m;d]$, and $\tilde{J}_0$ becomes
+$\tilde{J}_{0,\sigma}[m;d]$ which means exactly what
+it did in the previous section. The subscript ${\rm MSWI}$ is left
+off for $\alpha=0$, for reasons that will become apparent shortly.
+
+With these correspondences, equation \ref{eqn:eq4} becomes
+\begin{equation}
+  \label{eq7}
+  \lim_{\alpha \rightarrow 0}
+  \frac{1}{\alpha^2}\left(\tilde{J}_{{\rm MSWI},\alpha,\sigma}[m;d]-\tilde{J}_{0,\sigma}[m;d]\right)=
+  J_{\rm MSWI}[m;d]
+\end{equation}
+
+To derive the analogous relation for AWI only requires re-definition
+of the operator $T$: it becomes (taking explicitly into account
+dependence on source and receiver coordinates)
+\begin{equation}
+  \label{eqn:eq8}
+  T[m;d] u(\bx_r,t;\bx_s) = \frac{\max(|t|, t_{\rm max})
+    \mbox{sgn}(t)}{\|u_{0,\sigma}[m;d](\bx_r,\cdot;\bx_s)\|}
+  u(\bx_r,t;\bx_s).
+\end{equation}
+That is, for AWI, $T$ depends on both model and data. Define
+\begin{equation}
+  \label{eqn:eq5a}
+   J_{{\rm AWI},\alpha,\sigma}[m,u;d] = \frac{1}{2}\left(\|S[m]u-d\|^2 +
+   \sigma^2\|u\|^2 + \alpha^2\|T[m;d]u\|^2\right)
+ \end{equation}
+The minimizer $\ua$ over $u$ of $ J_{{\rm AWI}, \alpha,\sigma}[m,u;d]$ is
+$u_{{\rm AWI},\alpha,\sigma}[m;d]$, and $\tJa$ becomes
+\begin{equation}
+  \label{eqn:eq6a}
+  \tilde{J}_{{\rm AWI},\alpha,\sigma}[m;d] =
+  J_{{\rm AWI};\alpha,\sigma}[m,u_{{\rm AWI};\alpha,\sigma}[m;d];d].
+\end{equation}
+Note that the definitions \ref{eqn:eq5} and \ref{eqn:eq6} are exactly
+the same as the corresponding definition \ref{eqn:eq5a} and
+\ref{eqn:eq6a} when $\sigma=0$, which is why the subscripts ${\rm
+  MSWI}$ and ${\rm AWI}$ are left off in these cases.
+
+Precisely the
+same reasoning then leads to
+\begin{equation}
+  \label{eq9}
+  \lim_{\alpha \rightarrow 0}
+  \frac{1}{\alpha^2}\left(\tilde{J}_{{\rm AWI},\alpha,\sigma}[m;d]-\tilde{J}_{0,\sigma}[m;d]\right)=
+  J_{\rm AWI}[m;d]
+\end{equation}
 
@@
 connected by unique rays. The first condition may be ignored to some
-extend - the conclusion of this paper likely holds if the data is
-dominated by ballistic transmitted waves. However the second condition
-is essential: in the presence of multiple arrivals, the connection
-with travel time inversion breaks down, and AWI and similar algorithms
+extent - the conclusion of this paper likely holds if the data is
+dominated by ballistic transmitted waves. The second condition
+is essential: even for purely transmitted waves, the presence of
+multiple arrivals breaks the connection
+with travel time inversion, and AWI and similar algorithms
 become essentially as prone to cycle-skipping as is FWI.
+
+The analysis presented here has nothing to say about AWI applied to
+reflected wave data. \cite{Warner:16} and others have
+suggested that AWI is effective in that context. If so, the underlying
+mechanism remains to be explained.
 

### 2024-07-23 repro/papers/awinum/paper.tex
LBFGS and other updates

@@
 iterations. Methods more closely related to Newton iteration require
-more involved computations (see for instance \cite{Kaufman:75}.
+more involved computations (see for instance \cite{Kaufman:75}).
 
@@
 
-\plot{chwd20}{width=\textwidth}{Data for homogeneous model $m_0$: 20
-  sources at $x=3000$ m, 201 receivers at $x=5000$ m.}
+\begin{figure}[H]   
+\begin{center}
+\includegraphics[width=\textwidth]{project/Fig/chwd20}
+\caption{Data from the homogeneous model $m_0$ acquired using 20 sources located along a vertical line at $x=3000$ m and 181 receivers at $x=5000$ m.}
+\label{fig:chwd20}
+\end{center}
+\end{figure}
+
+%\plot{chwd20}{width=\textwidth}{Data for homogeneous model $m_0$: 20
+%sources at $x=3000$ m, 201 receivers at $x=5000$ m.}
 

### 2024-08-30 repro/papers/awi/paper.tex
edits

@@
 changes in velocity leads to oscillations in the difference measures,
-especially for the most commonly used mean-square measure. Local
+especially for the most commonly used square-integral measure. Local
 optimization algorithms (relatives of Newton;'s method) are the only practical
@@
 (that is, unique rays of geometric asymptotics between sources and
-receivers), the AWI objective function is an asymptotic (in time
+receivers), and known source wavelet, the AWI objective function is an asymptotic (in time
 frequency) approximation to the
-mean-square time-of-travel misfit. Thus AWI and travel time fitting
+mean-square time-of-travel misfit, a commonly used objective for
+travel time tomography. The error in this approximation is
+proportional to the source mean-square wavelength. Thus AWI and travel time fitting
 lead to similar material parameter estimates. Since the latter is
 widely believed to be a near-optimal method for wave velocity
-estimation, this demonstration explains roughly ``why AWI works'', at
+estimation, this relation explains roughly ``why AWI works'', at
 least within this restricted scope.
@@
 note that a normalization component of the AWI objective function
-definition is 
-important in reaching this conclusion. The connection between
-objective normalization and the asymptotic equivalence to travel time
-inversion for transmission data has nothing to do with
-the reasons for normalizing the objective previoiusly advanced in the
+definition is important in reaching this conclusion. The connection
+between objective normalization and the asymptotic equivalence to
+travel time inversion for transmission data has nothing to do with the
+reasons for normalizing the objective previoiusly advanced in the
 literature \cite[]{Warner:16,LiAlkhalifah:21}. Another modification of
 FWI based on similar ideas, Matched Source Waveform Inversion (MSWI)
-(\cite{HuangSymes2015SEG,HuangSymes:Geo17}) lacks the AWI normalization,
-and has a looser relation to travel time inversion. The arguments
-presented here for AWI (and MSWI) are refinements of those given by
-\cite{HuangSymes:Geo17}. which were in turn based on ideas presented in Song's thesis
-\cite[]{Song:94c}.
+(\cite{HuangSymes2015SEG,HuangSymes:Geo17}) lacks the AWI
+normalization, and has a looser relation to travel time inversion. The
+arguments presented here for AWI (and MSWI) are refinements of those
+given by \cite{HuangSymes:Geo17}. which were in turn based on ideas
+presented in Song's thesis \cite[]{Song:94c}.
 
@@
 The version of AWI introduced by \cite{Warner:16} is based on an
-active source acoustic model of data generation. Each acoustic source
-is modeled an isotropic point radiator with known location $\bx_s$ and
-wave form (``wavelet'') $w(t;\bx_s)$. Similarly, receivers are modeled
-as isotropic point sensors, recording pressure at points ${\bf x}_r$.
-The pressure and velocity fields $p({\bf x},t;{\bf x}_s)$, ${\bf v}({\bf x},t;{\bf x}_s)$ for the source location ${\bf x}_s$ depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf x})$ (reciprocal of the density $\rho({\bf x})$), and wavelet $w(t;{\bf x}_s)$ through the acoustic system
+active source acoustic model of data generation in Euclidean space
+$\bR^3$. In this section, I will review a version of this model, place
+some useful constraints on energy sources appearing in it, and define
+AWI as an optimization problem.
+
+\subsection{Acoustic Modeling and Waveform Inversion}
+
+Each acoustic source is modeled an isotropic point radiator
+with known location $\bx_s\in \bR^3$ and wave form (``wavelet'')
+$w(t)$ (idealized as independent of $\bx_s$ for simplicity). These
+sources generate acoustic waves traveling through Euclidean space. The
+data are samples of these waves at receivers are modeled as isotropic
+point sensors, recording (excess) pressure at points
+${\bf x}_r \in \bR^3$. The recording takes place over a time interval
+$[0,t_{\rm max}]$ (also idealized as independent of $\bx_s$ and
+$\bx_r$) . I will refer to the recording interval and the
+source-receiver position pairs collectively as the {\em acquisition
+  geometry}.
+
+The pressure and velocity fields $p({\bf x},t;{\bf x}_s)$,
+${\bf v}({\bf x},t;{\bf x}_s)$ for the source location ${\bf x}_s$
+depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf
+  x})$ (reciprocal of the density $\rho({\bf x})$), and source wavelet $w(t)$ through the acoustic system
 \begin{eqnarray}
@@
  \frac{\partial p}{\partial t} & = &- \kappa \nabla \cdot {\bf v} +
-w(t;{\bf x}_s) \delta({\bf x}-{\bf x}_s); \nonumber \\
+                                    \left(\int^t w \right)
+ \delta({\bf x}-{\bf x}_s); \nonumber \\
 \frac{\partial {\bf v}}{\partial t} & = & - \beta \nabla p; \\ 
@@
 \end{eqnarray}
-For sake of brevity, define the model vector $m=(\kappa,\rho)$.
-The forward map or {\em modeling operator} is $F[m]w = \{p({\bf x}_r,t;{\bf x}_s)\}$, in which source and receiver positions ${\bf x}_s, {\bf x}_r$ define the acquisition geometry.
+The representation of the point source time-dependence as an
+indefinite integral of the source wavelet will prove convenient
+in expressing the asymptotic approximation of $p$.
 
-Acoustic inversion, for present purposes, means: given source and receiver locations $(\bx_s,\bx_r)$, source wavelet
-$w(t;\bx_s)$ and data traces $d(\bx_r,\cdot;\bx_s)$, find a model $m$
-so that $F[m]w \approx d$. The
-simplest version of FWI concretizes this task by asking for a model
-$m$ minimizing the mean square error
-\begin{equation}
-  \label{eqn:fwi}
-  J_{\rm FWI}[m;d] =\frac{1}{2} \sum_{{\bf x}_r,{\bf x}_s} \int\,dt\, |(F[m]w)({\bf x}_r,t;{\bf x}_s) - d({\bf x}_r,t;{\bf x}_s)|^2
-\end{equation}
-It is convenient to introduce the $L^2$ norm $\| \cdot \|$ as
-shorthand for the right-hand side of the definition \ref{eqn:fwi}:
-\[
-  J_{\rm FWI}[m;d]= \frac{1}{2}\|F[m]w-d\|^2.
-\]
-In the sequel I will use the $L^2$ norm notation to stand for
-mean-square over whatever variables make sense.
+For simplicity, I assume that the system \ref{eqn:awe} holds over all of Euclidean
+space-time $(\bx,t) \in \bR^3 \times \bR$, and ignore physical
+boundaries such as the Earth's surface.
+For sake of brevity, define the model vector $m=(\kappa,\rho, w)$. The
+{\em modeling operator} $F$ maps the model $m$ to the corresponding data $\{p({\bf x}_r,t;{\bf
+  x}_s): 0 \le t \le t_{\rm max}\}$.
 
-Note that in actual data acquisition, recording occurs over finite
-time intervals, so the integrals appearing in the definition
-\ref{eqn:fwi} have finite upper and lower limits, as opposed to the
-$\pm \infty$ limits implicit in the notation. It will be assumed that
-wavefield energy is causal (zero before an onset time) and decays
-rapidly enough as $t \rightarrow \infty$ that the difference between
-finite and infinite range of integration can be ignored throughout
-this discussion.
-
-The wavelet $w$ is assumed known in this formulation of the
-FWI problem. In fact, even if actual acoustic sources are
+For the purposes of this paper, the source wavelet $w$ and the density
+$\rho$ will be assumed known. In fact, even if actual acoustic sources are
 well-approximated by isotropic point radiators, their wave forms are
@@
 included as  parameters to be estimated by inversion of the
-data. However for the present discussion we will regard $w$ as known.
+data. However for the present discussion, $w$ is regarded as
+known.
+
+Acoustic inversion means: given the set of
+source-receiver position pairs $\{(\bx_s,\bx_r)\}$, source wavelet
+$w(t)$, density $\rho(\bx)$,  and data traces
+$\{d(\bx_r,t;\bx_s): 0 \le t \le t_{\rm max}\}$, find a bulk modulus $\kappa$ so that the model
+$m=(\kappa,\rho,w)$ satisfies $F[m] \approx d$ in some sense. The
+simplest version of FWI concretizes this task by asking for a model
+$m$ (that is, a bulk modulus $\kappa$) minimizing the square integral/sum
+error over the recording time interval $[0,t_{\rm max}]$ and
+source-receiver position pairs:
+\begin{equation}
+  \label{eqn:fwi}
+  J_{\rm FWI}[m;d] =\frac{1}{2} \sum_{{\bf x}_r,{\bf x}_s}
+  \int_0^{t_{\rm max}}\,dt\, |F[m]({\bf x}_r,t;{\bf x}_s) - d({\bf x}_r,t;{\bf x}_s)|^2
+\end{equation}
+The definition \ref{eqn:fwi} makes sense only if the predicted and
+observed data traces ($F[m]({\bf x}_r\cdot;{\bf x}_s)$ and $d({\bf
+  x}_r,\cdot;{\bf x}_s)$) are square-integrable functions on $[0,t_{\rm max}]$, for each
+source-receiver position pair $(\bx_r,\bx_s)$. This property is
+assumed for $d$, and follows from additional assumptions on $m$ and the
+acqquisition geometry, to be explained below.
+
+It is convenient to introduce the $L^2$ norm symbol $\| \cdot \|$ as
+shorthand for the right-hand side of the definition \ref{eqn:fwi}:
+\[
+  J_{\rm FWI}[m;d]= \frac{1}{2}\|F[m]-d\|^2.
+\]
+In this notation, the interval $[0,t_{\rm max}]$ and the
+source-receiver position pair set $\{(\bx_r,\bx_s)\}$ are implicit. In
+the sequel, I will use the $L^2$ norm notation to stand for square
+integral/sum over whatever variables make sense in context.
 
 Practical versions of FWI include various types of regularization and
-imposition of constraints on $m$, see the sources cited in the
+imposition of constraints on $m$, see the references cited in the
 Introduction. As mentioned there, local iterative optimization methods
@@
 collection and processing.
-%For now, assume that the data $d({\bf x}_r,t;{\bf x}_s)$ is the output of the modeling operator for "true" model $m^*$, that is, the "true" bulk modulus, buoyancy, and wavelet $\kappa^*, \beta^*, w^*$: that is, $d = F[m^*]w^*$.
 
-AWI is based on extending the modeling operator $F$, that is, enlarging its domain.
-The extended modeling operator ${\bar F}$ maps extended sources
+\subsection{Source Characteristics}
+
+In applications to seismology, medical imaging, and other wave imaging
+technologies, the wavelet $w$ is actually the product of signal
+processing aimed to shape the data $F[m]$ into an approximation of
+data that would have been predicted using a known wavelet. That is, the wavelet is to some extent {\em
+  chosen}. This circumstance justifies several constraints on the
+wavelet, consistent with the properties of observed signals: the
+wavelet is assumed to be of class $H^1_0(\bR)$, that is, square integrable with
+square inttegrable derivative and vanishing for sufficiently large
+$t$, and even in $t$.
+Introduce asymptotics in this setting by assuming that $w = \wl$ is chosen
+from a family $\{\wl:\}$ indexed by the dimensionless parameter
+$0 <\lambda <1\}$, and defined by
+\begin{equation}
+  \label{eqn:wfam}
+  \wl(t) = \lambda^{-1/2}w_1(t/\lambda).
+\end{equation}
+The  ``mother wavelet'' $w_1$ is even and of class $H^1_0(\bR)$, and
+all wavelets $\wl$ in this family
+inherit these properties from $w_1$, along with its dimensions (energy/time$^2$).
+
+It follows from these assumptions that the Fourier transform $\fwl$ is
+real-valued, even, of class $H^1(\bR)$. and its product with frequency $|\omega|$ is also
+square-integrable.
+
+The parameter $\lambda$ is connected to several frequency-related
+characteristics of $\wl$. Suppose that $f \in L^2(\bR)$ is also square-integrable after
+multiplication by $t$ (true if  for instance $f$ has compact
+support). The {\em pulse width} (or {\em RMS width} ) of $f$ is
+\begin{equation}
+  \label{eqn:pw}
+  l(f) = \frac{\|tf\|}{\|f\|}.
+\end{equation}
+The pulse width of $\wl$ is proportional to $\lambda$:
+\begin{equation}
+  \label{eqn:wpw}
+  l(\wl) = \left(\frac{\int dt (t \wl(t))^2}{\int dt (\wl(t))^2}
+  \right)^{1/2} = \lambda l(w_1).
+\end{equation}
+The {\em RMS frequency} of $\wl$ is
+\begin{equation}
+  \label{eqn:freq}
+  k(\wl) = \left(\frac{\int d\omega (\omega \fwl(\omega))^2}{\int dt (w(t))^2} \right)^{1/2} = \lambda^{-1}k(w_1).
+\end{equation}
+Thus the {\em RMS wavelength} $1/k(\wl)$ is proportional to $\lambda$.
+
+The Heisenberg inequality (for example, \cite{Folland:07}, p. 255) bounds the RMS
+wavelength, the reciprocal of the RMS frequency, in terms of the pulse
+width:
+\begin{equation}
+  \label{eqn:heis}
+  l(\wl) k(\wl) = l(w_1)k(w_1) \ge \frac{1}{4\pi}.
+\end{equation}
+Equality is attained for Gaussian $w_1$, and approximated for many other
+commonly used wavelets (Ricker = second derivative of Gaussian,
+trapezoidal bandpass,...).
+
+\subsection{AWI via Optimization}
+
+AWI is based an {\em extension} of the modeling operator $F$, that is,
+an operator ${\bar F}$ with larger domain, coinciding with $F$ on its domain..
+The extended modeling operator ${\bar F}$ maps extended models
+$\bar{m}=(\kappa,\rho,\bar{w})$ incorporating extended sources
 $\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
@@
   x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver
-position, then ${\bar F}[m]\bar{w} = F[m]w$. That is, $F$ is a special
+position, then ${\bar F}[\bar{m}] = F[m]$. That is, $F$ is a special
 case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
@@
 
-AWI assumes that the extended sources are time convolutions of the
+\cite{Warner:16} represent the extended sources $\bar{w}$ as time convolutions of the
 (known) source $w$ with an {\em adaptive filter} $u({\bf
@@
 \]
-The asterisk denotes convolution in time. That is,
-$u*w$ denotes the result of convolving each trace in $u$ by $w$. I
-will also use the asterisk to denote the tracae-by-trace convolution
+Here, $u*w$ denotes the result of convolving each trace in $u$ by $w$. I
+will also use the asterisk to denote the trace-by-trace convolution
 of two data sets with the same set of source and receiver coordinates,
@@
 \[
- \bar{F}[m]\bar{w} = \bar{F}[m](u* w) = (F[m]w)*u
+ \bar{F}[\bar{m}] = \bar{F}[(\kappa,\rho,,u*w)] = F[m]*u
 \]
-Since the known source $w$ is not adjusted during the inversion, 
-introduce the notation $S[m]$ for the convolution operator with kernel $F[m]w$:
+Introduce the notation $S[m]$ for the convolution operator with kernel $F[m]$:
 \begin{equation}
   \label{eqn:sdef}
-  (F[m]w)*u \equiv S[m]u
+  F[m]*u \equiv S[m]u
 \end{equation}
@@
 $S[m]u[m;d]= d$. This proposition is dubious, but is easy to work
-around, as will be seen below. Also, \cite{Warner:16} also formulate
+around, as will be seen below. Also, \cite{Warner:16} formulate
 AWI in terms of the inverse filter, that is, $u$ for which
-$u*d = F[m]w$, and in fact recommend that definition. The present
+$u*d = F[m]$, and in fact recommend that definition. The present
 discussion uses only the forward filter.
 
-If $\kappa = \kappa^*, \beta = \beta^*$, then $\bar{w}$ should be $= w$,
+If $\kappa = \kappa^*, \beta = \beta^*$, then the extended source
+$\bar{w}$ that fits the data should be $= w$,
 so the adaptive kernel $u[m;d]$ should be $=\delta(t)$ for all
@@
 multiplication by $t$.
-
 This observation leads to the definition of the AWI objective function $J_{\rm AWI}$:
@@
 Adaptive waveform inversion consists in iterative minimization of
-$J_{\rm AWI}$: choose an initial estimate $m_0=(\kappa_0,\beta_0)$ of the
-model, then update $m=(\kappa,\beta)$ repeatedly using a gradient
+$J_{\rm AWI}$: choose an initial estimate $m_0=(\kappa_0,\beta, w)$ of the
+model, then update $m$ (that is, $\kappa$) repeatedly using a gradient
 descent method. \cite{Warner:16} describes the computation of $J_{\rm AWI}$
@@
 
-It may not be obvious at first glance why this approach to estimating
-$m$ should be any more effective than FWI. To see why AWI works, in at
-least some settings, we assume that the spatial dimension is three,
+It may not be obvious at first glance why the AWI approach to estimating
+$m$, described in the last section, should be any more effective than
+FWI. To see why AWI works when FWI does not, in at
+least some settings, assume that the spatial dimension is three,
 and make use of the geometric asymptotics (ray theory) approximation
@@
   local}: it applies when sources and receivers are sufficiently
-close. We will discuss its global failure and the consequences for AWI
-in the final section.
+close. I will discuss its global failure and the consequences for AWI
+in the fifth section of this paper.
 
-Transient acoustic sources have no DC (zero frequency) component and
-are relatively small at low frequencies. One way to enforce these
-characteristics is to assume that the wavelet $w$ is the time
-derivative of another wavelet $v$, also transient ($v(t;{\bf x}_s)=0$
-for $|t| \gg 0$). Set $\lambda(w) = \|v\|/\|w\|$. $\lambda$ is a proxy
-for wavelength: it is the reciprocal of RMS frequency of $v$.
-
-The geometric asymptotics approximation for solution of the point
-radiator problem \ref{eqn:awe} expresses the pressure field in terms
+Since the fields $\kappa$ and $\rho$ have been assumed smooth,
+the geometric asymptotics approximation is applicable for solution of the point
+radiator problem \ref{eqn:awe}. This approximation expresses the pressure field in terms
 of $m$-dependent geometric amplitude $a[m]({\bf x}_s,{\bf x}_r)$ and
@@
   \label{eqn:hadamard}
-  (F[m]w)({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) w(t-\tau[m]({\bf x}_s,{\bf x}_r)) 
+  F[m_{\lambda}]({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) \wl(t-\tau[m]({\bf x}_s,{\bf x}_r)). 
 \end{equation}
+Here and in the notation to follow, $m_{\lambda} =
+(\kappa,\rho,\wl)$.
+
 In this approximation, and in other uses of the symbol $\approx$, the
-error is $O(\lambda(w))$.
+error is dominated by data wavelength, that is, RMS wavelength of the
+source wavelet $\wl$, which is proportional to $\lambda$ as mentioned above. In other words, ``$\approx$'' means ``in error by $O(\lambda)$''.
 
@@
 \cite[]{Friedlander:75,Qian:JCP24}. For smooth (slowly varying)
-$m$, it holds for $\bx_r$ sufficiently close to $\bx_s$ that a unique ray of
+$m$, it holds {\em locally}, that is, for $\bx_r$ sufficiently close to $\bx_s$ that a unique ray of
 geometric acoustics connect $\bx_s$ to $\bx_r$. This ``single
@@
 
-Applying the approximation \ref{eqn:hadamard}, obtain
+Applying the approximation \ref{eqn:hadamard} to the definition
+\ref{eqn:sdef}, obtain
 \begin{equation}
   \label{eqn:spwe}
-S[m]u({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) (w*u)(t-\tau[m]({\bf x}_s,{\bf x}_r))
+S[m_{\lambda}]u({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) (\wl*u)(t-\tau[m]({\bf x}_s,{\bf x}_r))
 \end{equation}
@@
 $$
-d({\bf x}_r,t;{\bf x}_s)=(F[m^*]w) ({\bf x}_r,t;{\bf x}_s) \approx a[m^*]({\bf x}_s,{\bf x}_r) w(t-\tau[m^*]({\bf x}_s,{\bf x}_r))
+d({\bf x}_r,t;{\bf x}_s)=(F[m^*]) ({\bf x}_r,t;{\bf x}_s) \approx a[m^*]({\bf x}_s,{\bf x}_r) \wl(t-\tau[m^*]({\bf x}_s,{\bf x}_r))
 $$
-The (approximate) solution of $S[m]u[m;d]=d$ is obvious by inspection:
+The (approximate) solution of $S[m_{\lambda}]u=d$ is obvious by inspection:
 \begin{equation}
   \label{eqn:notl2}
-  u[m;d]({\bf x}_r,t;{\bf x}_s) = \frac{a[m^*]({\bf x}_s,{\bf x}_r)}{a[m]({\bf x}_s,{\bf x}_r)}\delta(t-\tau[m^*]({\bf x}_s,{\bf x}_r)+\tau[m]({\bf x}_s,{\bf x}_r)).
+  u({\bf x}_r,t;{\bf x}_s) = \frac{a[m^*]({\bf x}_s,{\bf x}_r)}{a[m]({\bf x}_s,{\bf x}_r)}\delta(t-\tau[m^*]({\bf x}_s,{\bf x}_r)+\tau[m]({\bf x}_s,{\bf x}_r)).
 \end{equation}
-There is immediately a big problem: this $u[m;d]$ is not
+There is immediately a big problem: this $u$ is not
 square-integrable, and the AWI objective $J_{\rm AWI}$ is
-undefined. To fix this, it's necessary to replace $u[m;d]$ by
-$u_{0,\sigma}[m;d]$,  the solution of a regularized least squares problem: 
+undefined. It's not surprising that mathematical ill-behaviour arises
+in this context. Since $S[m_{\lambda}]$ is a convolution operator, solving
+$S[m_{\lambda}]u=d$ for $u$ amounts to deconovolving $d$ by the kernel $F[m]$ of
+$S[m_{\lambda}]$. Under the assumed conditions on the wavelet $\wl$, $S[m_{\lambda}]$ does
+not have a continuous (stable) convolution inverse.
+
+It is conventional to stabilize deconvolution problems such as the
+equation $S[m_{\lambda}]u=d$ via so-called ``pre-whitening'', as
+mentioned by \cite{Warner:16} ``Pre-whitening'' is a version of
+Tihonov regularization. It  amounts to replacement of the equation
+$S[m_{\lambda}]u=d$ by the regularized least squares problem: 
 \begin{equation}
   \label{eqn:reg}
-  u_{0,\sigma}[m;d] = \mbox{argmin}_u (\|S[m]u-d\|^2 + \sigma^2\|u\|^2)
+  u_0[m_{\lambda};d] = \mbox{argmin}_u (\|S[m_{\lambda}]u-d\|^2 + \lambda\|u\|^2)
 \end{equation}
+
+The choice of $\lambda$ as the  the regularization weight might seem
+natural, as $\lambda$ is more-or-less the shortest wavelength
+controlled by solution of the least-squares deconvolution problem
+\ref{eqn:reg}. It will be rigorously justified {\em post-hoc} by the
+calculations to follow.
+
+Note that in these expressions, the source and receiver coordinates
+${\bf x}_s$ and ${\bf x}_r$ act as either passive parameters or
+summation indices. The right-hand side of equation \ref{eqn:reg} is
+summed over ${\bf x}_s$ and ${\bf x}_r$, for instance. However, the
+quantities on the right-hand side are not coupled: the minimization
+could be regarded as taking place independently for each choice of
+${\bf x}_s$ and ${\bf x}_r$. Therefore, to simplify the presentation,
+drop ${\bf x}_s$ and ${\bf x}_r$ from the notation temporarily: each
+calculation to follow pertains to a single trace, i.e. a single choice
+of ${\bf x}_s$ and ${\bf x}_r$.  Also, use the
+temporary abbreviations
+$a^*=a[m^*], a=a[m], \tau^*=\tau[m^*], \tau=\tau[m]$.
+
+Using the Hadamard progressing wave approximation \ref{eqn:hadamard},
+the right-hand side of equation \ref{eqn:reg} is 
 $$
-\approx \mbox{argmin}_u (\|a[m](w*u)(\cdot-\tau[m])-a[m^*]w(\cdot-\tau[m^*])\|^2 + \sigma^2\|u\|^2)
+\approx \mbox{argmin}_u (\|a[m](\wl*u)(\cdot-\tau[m])-a[m^*]\wl(\cdot-\tau[m^*])\|^2 + \lambda\|u\|^2)
 $$
-
-Note that in these expressions, the source and receiver
-coordinates ${\bf x}_s$ and ${\bf x}_r$ act as either passive parameters
-or summation indices. Therefore, to simplify the presentation, drop them from the notation
-temporarily. Also,  use the temporary abbreviations $a^*=a[m^*], a=a[m], \tau^*=\tau[m^*],
-\tau=\tau[m]$.
-
-With these conventions, re-write equation \ref{eqn:reg} in terms of the Fourier transforms $\hat{u}[m;d],\hat{w}$ of $u_{0,\sigma}[m;d]$ and $w$:
+Re-write the approximation \ref{eqn:reg} in terms of the Fourier transforms $\hat{u_0}[m_{\lambda};d],\fwl$ of $u_0[m_{\lambda};d]$ and $\wl$:
 \begin{equation}
   \label{eqn:freqreg}
-\hat{u}[m;d] \approx \mbox{argmin}_{\hat{u}}\left(\int d\omega |a\hat{w}\hat{u}e^{i\omega \tau}-a^*\hat{w}e^{i\omega\tau^*}|^2 + \sigma^2 |\hat{u}|^2\right)
+\hat{u}_{0,\sigma}[m;d] \approx \mbox{argmin}_{\hat{u}}\frac{1}{2}\int d\omega \left(|a\hat{w}\hat{u}e^{i\omega \tau}-a^*\hat{w}e^{i\omega\tau^*}|^2 + \lambda |\hat{u}|^2\right)
 \end{equation}
@@
 \[
-(a^2 |\hat{w}|^2 +\sigma^2)\hat{u} = a a^*|\hat{w}|^2e^{i\omega(\tau^*-\tau)}
+(a^2 |\fwl|^2 +\lambda)\hat{u} = a a^*|\fwl|^2e^{i\omega(\tau^*-\tau)}
 \]
 the solution of which is
-\[
-\hat{u}[m;d] \approx \frac{a^*}{a}\hat{g}_{\frac{\sigma}{a}} e^{i\omega(\tau^*-\tau)}
-\]
+\begin{equation}
+  \label{eqn:hatuf}
+\hat{u}_0[m_{\lambda};d] \approx \frac{a^*}{a}\hat{g}_{\lambda,\frac{\lambda}{a^2}} e^{i\omega(\tau^*-\tau)}
+\end{equation}
 where
+\begin{equation}
+  \label{eqn:hatgf}
+\hat{g}_{\lambda,\sigma} = \frac{|\fwl|^2}{|\fwl|^2 + \sigma}.
+\end{equation}
+Since $\wl$ vanishes for large $|t|$, $\fwl$ is entire analytic,
+hence vanishes on a set of measure zero, whence $\hat{g}_{\lambda,\sigma}:$ tends to $1$ almost everywhere as $\sigma \rightarrow 0$.
+Hence the inverse Fourier transform $g_{\lambda,\sigma}$ tends to $\delta$ in the sense of distributions, and
 $$
-\hat{g}_{\sigma} = \frac{|\hat{w}|^2}{|\hat{w}|^2 + \sigma^2}.
-$$
-Transient acoustic sources are effectively bandlimited:
-$\hat{w}(\omega) \rightarrow 0$ faster than any negative power of
-$\omega$ as $|\omega| \rightarrow \infty$ and is entire analytic, whence $\hat{g}_{\sigma}:$ tends to $1$ almost everywhere as $\sigma \rightarrow 0$.
-Hence the inverse Fourier transform $g_{\sigma}$ tends to $\delta$ in the sense of distributions, and
-$$
-u_{0,\sigma}[m;d](t) \approx \frac{a^*}{a}g_{\frac{\sigma}{a}}(t-(\tau^*-\tau))
+u_0[m_{\lambda};d](t) \approx \frac{a^*}{a}g_{\lambda,\frac{\lambda}{a^2}}(t-(\tau^*-\tau))
 $$
 to a multiple of a shifted $\delta$, in fact exactly the distribution
-solution of $\bar{S}[m,w]u=d$ given in equation \ref{eqn:notl2}. However for $\sigma>0$, $u_{0,\sigma}[m;d]$ is square integrable.
+solution of $\bar{S}[m_{\lambda}]u=d$ given in equation \ref{eqn:notl2}. However for $\lambda>0$, $u_0[m_{\lambda};d]$ is square integrable.
 
 \cite{HuangSymes2015SEG} presented an analysis of the Matched Source
-Waveform Inversion (MSWI) objective function, which is essentially the
-numerator of the right hand side in the definition
-\ref{eqn:tJz} of the AWI objective. This analysis was based in turn on ideas presented in Song's thesis
-\cite[]{Song:94c}. MSWI deals with the extended source $\bar{w}$, but
-the same construction applies to the adaptive filter $u$. With $u$ replaced by $u_{0,\sigma}[m;d]$,  the MSWI objective is
+Waveform Inversion (MSWI) objective function, which is the numerator
+of the right hand side in the definition \ref{eqn:tJz} of the AWI
+objective. Note that this definition requires summation over sources and receivers:
 \[
-J_{\rm MSWI}[m;d] = \int dt |tu_{0,\sigma}[m;d]|^2 \approx \frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\frac{\sigma}{a}}(t-(\tau^*-\tau))|^2
+J_{\rm MSWI}[m_{\lambda};d] = \sum_{\bx_s,\bx_r}\int dt |tu_0[m_{\lambda};d]|^2 \approx \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{\lambda}{a^2}}(t-(\tau^*-\tau))|^2
 \]
 \[
-=\frac{(a^*)^2}{a^2} \int dt\, (t+(\tau^*-\tau))^2|g_{\frac{\sigma}{a}}(t)|^2
+=\sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, (t+(\tau^*-\tau))^2|g_{\lambda,\frac{\lambda}{a^2}}(t)|^2
 \]
@@
   \label{eqn:MSWI}
-  =\frac{(a^*)^2}{a^2} \int dt\, (t^2 + 2t
-  (\tau^*-\tau)+(\tau^*-\tau)^2)|g_{\frac{\sigma}{a}}(t)|^2
+  =\sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, (t^2 + 2t
+  (\tau^*-\tau)+(\tau^*-\tau)^2)|g_{\lambda,\frac{\lambda}{a^2}}(t)|^2
 \end{equation}
-As noted by \cite{HuangSymes2015SEG}, the linear term (in t) on the
-right-hand side of \ref{eqn:MSWI} vanishes,
-since $g$ is necessarily even in time (since its Fourier transform is
-real). So
+
+Since the Fourier transform $\fwl$ is even and real, so is
+$\hat{g}_{\lambda,\sigma}$, therefore $g_{\lambda,\sigma}$ is even, and the linear term (in t) on the
+right-hand side of \ref{eqn:MSWI} vanishes, So
 \begin{equation}
@@
   J_{\rm MSWI}[m;d] \approx
-\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\frac{\sigma}{a}}(t)|^2
-+(\tau^*-\tau)^2\frac{(a^*)^2}{a^2}\|g_{\frac{\sigma}{a}}\|^2
+\sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{\lambda}{a^2}}(t)|^2
++(\tau^*-\tau)^2\frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{\lambda}{a^2}}\|^2
 \end{equation}
-The first term tends to zero as $\sigma \rightarrow 0$, since
-$g_{\sigma}$ is a delta-family, so you can make it negligible with
-proper choice of $\sigma$. The second term suggests that this function
+
+Recalling the definition \ref{eqn:pw} of pulse width, the
+approximation \ref{eqn:MSWI} can be re-written as
+\begin{equation}
+  J_{\rm MSWI}[m_{\lambda};d] \approx
+  \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{\lambda}{a^2}}\|^2\left(l(g_{\lambda,\frac{\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
+\end{equation}
+
+The second term suggests that this function
 is related to the travel-time error $\tau^*-\tau$.
-
 The relation is obscured by the presence of of the amplitudes
-$a^*, a$ both in the multiplier and in the scaling of $\sigma$. Since the
+$a^*, a$ both in the multiplier and in the scaling of $\lambda$. Since the
 amplitudes depend on $m$, their presence may influence the location
@@
 \ref{eqn:MSWI} of $J_{\rm MSWI}$, ones sees
-that the two functions differ only by the factor $\|u_{0,\sigma}[m;d]\|^{-2}$ in
-the former. Using the geometric asymptotics approximation, 
+that the two functions differ only by the factor $\|u_0[m_{\lambda};d]\|^{-2}$ in
+the former, applied trace-by-trace. Using the geometric asymptotics approximation, 
 $$
-\|u_{0,\sigma}[m;d]\| \approx \frac{a^*}{a}\|g_{\frac{\sigma}{a}}\| + O(\sigma^2)
+\|u_0[m_{\lambda};d]\|^2 \approx \frac{(a^*)^2}{a^2}\|g_{\frac{\lambda}{a^2}}\|^2 + O(\lambda)
 $$
-Therefore, multiplication of the MSWI objective by $\|u_{0,\sigma}[m;d]\|^{-2}$
-precisely removes the amplitude factor in the right-hand side of the
-relation \ref{eqn:MSWIasym}. Restoring the source and receiver
-coordinates to the notation,
-$$
+Therefore, multiplication of the MSWI objective by $\|u_0[m_{\lambda};d]\|^{-2}$
+precisely removes the amplitude factor and norm-squared of $g$ in the right-hand side of the
+relation \ref{eqn:MSWIasym}. That is,
+\begin{equation}
+  \label{eqn:preawitt}
+J_{\rm AWI}[m;d] \approx \sum_{\bx_s,\bx_r} (\left(l(g_{\lambda,\frac{\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
+ \end{equation}
+
+Note that
+\[
+  \int dt\, t^2|g_{\lambda,\sigma}(t)|^2 = \int d\omega
+  \left|\frac{d\hat{g}_{\lambda,\sigma}}{d\omega}\right|^2
+  \approx \int d\omega  \left(\frac{2 \mbox{Re }\bar{\hat{w}}_{\lambda}\fwl' 
+    \sigma}{(|\fwl|^2 + \sigma)^2}\right)^2
+\]
+\[
+  =\int d\omega \frac{(2 \lambda^{1/2}\hat{w}_1(\lambda
+    \omega)\lambda^{3/2}\hat{w}_1'(\lambda \omega) 
+    \sigma)^2}{(\lambda|\hat{w}_1(\lambda
+    \omega)|^2 + \sigma)^4}
+\]
+\[
+  =\frac{1}{\lambda}\int d\omega_1 \frac{(2 \lambda^2\hat{w}_1(\omega_1)\hat{w}_1'( \omega_1) \sigma)^2}{(\lambda|\hat{w}_1(
+    \omega_1)|^2 + \sigma)^4}
+  =\frac{\sigma^2}{\lambda}\int d\omega_1
+  \frac{(2\hat{w}_1(\omega_1)\hat{w}_1'( \omega_1)
+    )^2}{\left(|\hat{w}_1(\omega_1)|^2 + \frac{\sigma}{\lambda}\right)^4}
+\]
+Substitute $\sigma = \lambda/(a^2)$ to obtain
+\[
+ \int dt\,
+  t^2|g_{\lambda,\frac{\lambda}{a^2}}(t)|^2
+  =\frac{\lambda}{a^6}\int d\omega_1
+  \frac{(2\hat{w}_1(\omega_1)\hat{w}_1'( \omega_1)
+    )^2}{\left(|\hat{w}_1(\omega_1)|^2 + \frac{1}{a^2}\right)^4}
+\]
+\begin{equation}
+  \label{eqn:g1}
+ = \frac{\lambda}{a^2} \int d\omega
+  \frac{(2a\hat{w}_1a\hat{w}_1'
+    )^2}{\left(|a\hat{w}_1|^2 + 1\right)^4}
+\end{equation}
+Also
+\[
+  \|g_{\lambda,\sigma}\|^2 = \int d\omega\,
+  \left(\frac{|\fwl|^2(\omega)}{|\fwl|^2(\omega) + \sigma}\right)^2
+= \int d\omega\,
+  \left(\frac{|\lambda^{1/2}\hat{w}_1(\lambda
+      \omega)|^2(\omega)}{|\lambda^{1/2}\hat{w}_1(\lambda
+      \omega)|^2 + \sigma}\right)^2
+\]
+\[
+=\int d\omega\,
+  \left(\frac{|\hat{w}_1(\lambda
+      \omega)|^2}{|\hat{w}_1(\lambda
+      \omega)|^2 + \frac{\sigma}{\lambda}}\right)^2
+=  \frac{1}{\lambda} \int d\omega_1\,
+  \left(\frac{|\hat{w}_1(\omega_1)|^2}{|\hat{w}_1(\omega_1)|^2 + \frac{\sigma}{\lambda}}\right)^2
+\]
+Substitute $\sigma = \lambda/(a^2)$  and rearrange to obtain
+\begin{equation}
+  \label{eqn:g0}
+  \|g_{\lambda,\frac{\lambda}{a^2}}\|^2 \approx \frac{1}{\lambda} \int d\omega\,
+  \left(\frac{|a\hat{w}_1|^2}{|a\hat{w}_1|^2 + 1}\right)^2
+\end{equation}
+Combining the relations \ref{eqn:g0} and \ref{eqn:g1}, obtain
+\begin{equation}
+  \label{eqn:pwg}
+  l(g_{\lambda,\frac{\lambda}{a^2}}) =\lambda W(m,w_1)
+\end{equation}
+in which $W(m,w_1)$ is bounded by the variation of the amplitude
+$a[m](\bx_r,\bx_s)$, and the choice of ``mother wavelet'' $w_1$, but
+independent of $\lambda$.
+
+It follows that
+\begin{equation}
+  \label{eqn:awitt}
 J_{\rm AWI}[m;d] \approx \sum_{{\bf x}_s,{\bf x}_r} (\tau[m^*]({\bf
-  x}_s,{\bf x}_r)-\tau[m]({\bf x}_s,{\bf x}_r))^2 + O(\sigma^2)
-$$
+  x}_s,{\bf x}_r)-\tau[m]({\bf x}_s,{\bf x}_r))^2 + O(\lambda^2)
+\end{equation}
 In other words, the AWI objective is {\em exactly} an approximation to
@@
 ``wavelength'' $\lambda(w)$ and the regularization parameter
-$\sigma^2$.
+$\lambda$.
 
@@
 explained in the last section is a classic example, adding a scaled
-mean-square of the adaptive filter to the residual mean-square to
+square-integral of the adaptive filter to the residual square-integral to
 produce an objective with a square-integrable (finite-energy)
@@
 following choices, for each $m$ in the domain of $F$: $S$ is the regularized
-modeling operator $(S[m],\sigma I)^T$.  Its domain is $U=L^2(\bR)$, of
+modeling operator $(S[m_{\lambda}],\sigma I)^T$.  Its domain is $U=L^2(\bR)$, of
 which $u$ (the adaptive filter) is a member, and 
@@
   \label{eqn:eq5}
-   J_{{\rm MSWI},\alpha,\sigma}[m,u;d] = \frac{1}{2}\left(\|S[m]u-d\|^2 +
-   \sigma^2\|u\|^2 + \alpha^2\|Tu\|^2\right)
+   J_{{\rm MSWI},\alpha,\sigma}[m,u;d] = \frac{1}{2}\left(\|S[m_{\lambda}]u-d\|^2 +
+   \lambda\|u\|^2 + \alpha^2\|Tu\|^2\right)
  \end{equation}
@@
 \end{equation}
-Note that $u_0$ becomes $u_{0,\sigma}[m;d]$, and $\tilde{J}_0$ becomes
+Note that $u_0$ becomes $u_0[m_{\lambda};d]$, and $\tilde{J}_0$ becomes
 $\tilde{J}_{0,\sigma}[m;d]$ which means exactly what
@@
 \begin{equation}
-  \label{eq7}
+  \label{eqn:eq7}
   \lim_{\alpha \rightarrow 0}
@@
   T[m;d] u(\bx_r,t;\bx_s) = \frac{\max(|t|, t_{\rm max})
-    \mbox{sgn}(t)}{\|u_{0,\sigma}[m;d](\bx_r,\cdot;\bx_s)\|}
+    \mbox{sgn}(t)}{\|u_0[m_{\lambda};d](\bx_r,\cdot;\bx_s)\|}
   u(\bx_r,t;\bx_s).
@@
   \label{eqn:eq5a}
-   J_{{\rm AWI},\alpha,\sigma}[m,u;d] = \frac{1}{2}\left(\|S[m]u-d\|^2 +
-   \sigma^2\|u\|^2 + \alpha^2\|T[m;d]u\|^2\right)
+   J_{{\rm AWI},\alpha,\sigma}[m,u;d] = \frac{1}{2}\left(\|S[m_{\lambda}]u-d\|^2 +
+   \lambda\|u\|^2 + \alpha^2\|T[m;d]u\|^2\right)
  \end{equation}
@@
 
-Precisely the
-same reasoning then leads to
+With these choices, equation \ref{eqn:eq4} becomes 
 \begin{equation}
-  \label{eq9}
+  \label{eqn:eq9}
   \lim_{\alpha \rightarrow 0}
@@
 \end{equation}
+
+The intervention of the $\alpha=0$ limit $\tilde{J}_{0,\sigma}$ in the
+relations \ref{eqn:eq7} and \ref{eqn:eq9} obscures the relation
+between the MSWI and AWI objective functions and their penalty
+counterparts. In fact, under the hypotheses used in deriving he
+relations 
+Recall that the wavelet $w$ is assumed essentially bandlimited, that is
+$|\hat{w}(\omega)| \rightarrow 0$ as $\omega \rightarrow \infty$
+faster than any negative power of $\omega$. It follows that $\hat{w}
+\in L^p(\bR)$ for any $p \ge 1$ (and in fact the same is true of
+$\hat{w}$ multiplied by any polynomial in $\omega$).
+
+Assume as in the last section, that is, that solutions
+of the acoustic system \ref{eqn:awe} is well-approximated by the
+geometric asymptotics as in equation \ref{eqn:hadamard}, and that the
+data is noise-free ($d=F[m^*]$). Together with the properties of $w$
+just listed, these assumptions imply that
+\begin{equation}
+  \label{eqn:eq11}
+  J_{0,\sigma}[m;d] = O(\sigma) + O(\lambda(w)).
+\end{equation}
+To see this, begin with the definition:
+\[
+  \tilde{J}_{0,\sigma}[m;d] = \frac{1}{2}
+  (\|S[m_{\lambda}]u_0[m_{\lambda};d]-d\|^2 + \lambda \|u_0[m_{\lambda};d]\|^2.
+\]
+Using the approximate identities \ref{eqn:hatuf} and \ref{eqn:hatgf} and
+rearranging,
+\[
+\|S[m_{\lambda}]u_0[m_{\lambda};d]-d\|^2 
+  \approx \int d\omega
+  |a\hat{w}\hat{u_0[m_{\lambda};d]}e^{i\omega\tau}-a^*\hat{w}e^{i\omega\tau^*}|^2
+\]
+\[
+  = \int d\omega
+  (a^*)^2|\hat{w}|^2\left(\frac{\lambda}{a^2|\hat{w}|^2 +
+      \lambda}\right)^2   
+\le \int d\omega (a^*)^2|\hat{w}|^2
+  \left(\frac{\lambda}{a^2|\hat{w}|^2 + \lambda}\right)
+\]
+\begin{equation}
+  \label{eqn:eq12}
+  = \lim_{\Omega \rightarrow \infty} \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2\lambda
+  \left(\frac{a^2|\hat{w}|^2}{a^2|\hat{w}|^2 + \lambda}\right)
+\end{equation}
+For any $\Omega > 0$,
+\[
+\int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2\lambda\left(\frac{a^2|\hat{w}|^2}{a^2|\hat{w}|^2 + \lambda}\right)
+\le \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2 \lambda
+  \left(\frac{ a|\hat{w}|}{(a^2|\hat{w}|^2 +
+      \lambda)^{\frac{1}{2}}}\right)
+\]
+\[
+  = \sigma \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2 a|\hat{w}|
+  \left(\frac{\sigma }{(a^2|\hat{w}|^2 +\lambda)^{\frac{1}{2}}}\right)
+  \le \sigma \int_{-\Omega}^{\Omega} d\omega \frac{(a^*)^2}{a}
+  |\hat{w}|
+  \le \sigma \frac{(a^*)^2}{a} \|\hat{w}\|_{L^1(\bR)}.
+\]
+Together with the approximate inequality \ref{eqn:eq12}, this
+inequality implies
+that
+\[
+  \|S[m_{\lambda}]u_0[m_{\lambda};d] - d\|^2 = O(\sigma) + O(\lambda(w)).
+\]
+A similar argument leads to the conclusion that
+\[
+  \lambda\|u_0[m_{\lambda};d]\|^2 = O(\sigma) + O(\lambda(w)).
+\]
+Together, these conclusions imply the assertion \ref{eqn:eq11}.
 
@@
 adaptive filter or extended source to data (represented by the opertor
-$S[m]$ in the discussion above) may not have a well-behaved inverse in
+$S[m_{\lambda}]$ in the discussion above) may not have a well-behaved inverse in
 the presence of multiple wavefront arrivals. \cite{Yongetal:GJI23}
@@
   \label{eqn:multi}
-  F[m]w(t) \approx \sum_{i \ge 0} a_i H^{p_i}w(t-\tau_i).
+  F[m,w](t) \approx \sum_{i \ge 0} a_i H^{p_i}w(t-\tau_i).
 \end{equation}
@@
   \label{eqn:unwrap}
-  (\delta +T_1) * F[m]w \approx  a_0w(t-\tau_0)
+  (\delta +T_1) * F[m,w] \approx  a_0w(t-\tau_0)
 \end{equation}
@@
 
-Once again, it is possible to write the adaptive filter u for which $S[m]u = (F[m]w)*u = d =
-F[m^*]w$ by inspection. Use the notations $a^*_i, t^*_i, p^*_i$ for
+Once again, it is possible to write the adaptive filter u for which $S[m_{\lambda}]u = F[m]*u = d =
+F[m^*]$ by inspection. Use the notations $a^*_i, t^*_i, p^*_i$ for
 the analogous quantities computed with the ``true'' model $m^*$, and
@@
 \[
-  d = F[m_*]w = S[m]u
+  d = F[m_*]= S[m_{\lambda}]u
 \]
@@
 Of courese, the adaptive filter just computed is not square
-integrable, and must be regularized for use in a mean-square
+integrable, and must be regularized for use in a square-integral
 construction like that of AWI or MSWI. That amounts essentially to
@@
 
+%%%%%%%%%%%%%%%%%%
+Note that this main result effectively bounds the optimal mean-{\em
+  square} traveltime error achieved by AWI by the pulse width,
+hence the RMS traveltime error by the square-root of the pulse
+width. Pulse width is related to RMS wavelength by the Heisenberg
+inequality, and in many examples these differ by a modest factor.
+Thus error estimate is suboptimal from the point of view
+of FWI, which requires something like an initial model estimate
+predicting a half-wavelength traveltime error. This suboptimality
+arises from the use of pre-whitening (Tihonov regularization) in the
+definition of the AWI objecive function \cite[]{Warner:16}. It is
+possible that a different regularization strategy or assumptions on
+the wavelet could yield a more optimal error estimate.
+
 The analysis presented here has nothing to say about AWI applied to

### 2024-09-12 repro/papers/awi/paper.tex
interim

@@
 parameter distribution within a material object  (the earth, a human
-body, a manufactured artifact,...) from the measurement of wave motion
-at its surface. Over the last several decades, this task has come to
+body, a manufactured artifact,...) from remote measurements of waves
+propagating through it. Over the last several decades, this task has come to
 be carried out by minimizing a measure of the difference between computational
@@
 parameters. However the oscillations in the difference measures lead
-these sequences to become entrapped at near parameter distributions
+these sequences to stagnate at parameter distributions
 far from the optimum, and thus do not extract the maximum information
@@
 
-While these and other works discuss heuristic justifications for AWI and
-many details of implementation, a clear explanation for its effectiveness is 
-so far missing from the literature. In this short note, I show that
-for active sensing of acoustic materials with unique arrival times of
-transmitted waves
-(that is, unique rays of geometric asymptotics between sources and
-receivers), and known source wavelet, the AWI objective function is an asymptotic (in time
-frequency) approximation to the
-mean-square time-of-travel misfit, a commonly used objective for
-travel time tomography. The error in this approximation is
-proportional to the source mean-square wavelength. Thus AWI and travel time fitting
-lead to similar material parameter estimates. Since the latter is
-widely believed to be a near-optimal method for wave velocity
-estimation, this relation explains roughly ``why AWI works'', at
-least within this restricted scope.
+While these and other works discuss heuristic justifications for AWI
+and many details of implementation, a clear explanation for its
+effectiveness is so far missing from the literature. In this short
+note, I treat a special case of FWI, the recovery of smooth acoustic
+medium parameters from data generated at localized energy sources,
+propagated through a fluid, and recorded at receiver locations remote
+from the sources. I show that the AWI objective function is an
+asymptotic (in time frequency) approximation to the mean-square
+time-of-travel misfit, provided that arrival times of transmitted
+waves are single, that is, unique rays of geometric asymptotics
+connect sources and receivers. The error in this approximation is
+proportional to the source mean-square wavelength. Thus AWI and
+minimization of mean-square travel time error (travel time tomography)
+lead to similar material parameter estimates. Travel time tomography
+appears to be much less sensitive to choice of initial model than is
+FWI, and in fact is commonly used to construct an initial model for
+FWI \cite[]{Bordingetal:87,SirguePratt:04,VirieuxOperto:09}. At least
+in the special case treated here, AWI implicitly combines travel time
+tomography with FWI.
 
@@
 
-\subsection{Acoustic Modeling and Waveform Inversion}
+\subsection{Acoustic Modeling of Transmitted Waves}
 
 Each acoustic source is modeled an isotropic point radiator
-with known location $\bx_s\in \bR^3$ and wave form (``wavelet'')
-$w(t)$ (idealized as independent of $\bx_s$ for simplicity). These
+with known location $\bx_s\in \bR^3$. These
 sources generate acoustic waves traveling through Euclidean space. The
-data are samples of these waves at receivers are modeled as isotropic
-point sensors, recording (excess) pressure at points
+data are samples of these waves at receivers, modeled as isotropic
+point sensors, recording pressure at points
 ${\bf x}_r \in \bR^3$. The recording takes place over a time interval
-$[0,t_{\rm max}]$ (also idealized as independent of $\bx_s$ and
-$\bx_r$) . I will refer to the recording interval and the
-source-receiver position pairs collectively as the {\em acquisition
-  geometry}.
+$[0,t_d]$ (also idealized as independent of $\bx_s$ and
+$\bx_r$) . The collection $X_{sr}$ of source-receiver pairs
+$(\bx_s,\bx_r)$ for which data is recorded should not intersect the
+diagonal - that is, no receiver should be co-located with a source for
+which it is active. The set of source locations $X_s = \{\bx_s: (\bx_s,\bx_r) \in X_{sr}
+\mbox{ for some } \bx_r \in \bR^3\}$ will also be useful.
+
+In practice, recording takes place over a finite time interval
+$[0,t_d]$. For convenience, assume that this interval is the same for
+all receivers. Let $\phi \in C_0^{\infty}(0,t_d)$ be a window function
+supported in this interval. Additional constraints on $t_d$ and $\phi$
+will be be specified below.
 
@@
 depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf
-  x})$ (reciprocal of the density $\rho({\bf x})$), and source wavelet $w(t)$ through the acoustic system
+  x})$ (reciprocal of the density $\rho({\bf x})$), and source wave
+form $f(t)$ through the acoustic system
 \begin{eqnarray}
@@
  \frac{\partial p}{\partial t} & = &- \kappa \nabla \cdot {\bf v} +
-                                    \left(\int^t w \right)
- \delta({\bf x}-{\bf x}_s); \nonumber \\
+                                    f(t) \delta({\bf x}-{\bf x}_s); \nonumber \\
 \frac{\partial {\bf v}}{\partial t} & = & - \beta \nabla p; \\ 
@@
 \end{eqnarray}
-The representation of the point source time-dependence as an
-indefinite integral of the source wavelet will prove convenient
-in expressing the asymptotic approximation of $p$.
-
 For simplicity, I assume that the system \ref{eqn:awe} holds over all of Euclidean
@@
 boundaries such as the Earth's surface.
-For sake of brevity, define the model vector $m=(\kappa,\rho, w)$. The
-{\em modeling operator} $F$ maps the model $m$ to the corresponding data $\{p({\bf x}_r,t;{\bf
-  x}_s): 0 \le t \le t_{\rm max}\}$.
-
-For the purposes of this paper, the source wavelet $w$ and the density
-$\rho$ will be assumed known. In fact, even if actual acoustic sources are
-well-approximated by isotropic point radiators, their wave forms are
-usually poorly determined and may reasonably be
-included as  parameters to be estimated by inversion of the
-data. However for the present discussion, $w$ is regarded as
-known.
-
-Acoustic inversion means: given the set of
-source-receiver position pairs $\{(\bx_s,\bx_r)\}$, source wavelet
-$w(t)$, density $\rho(\bx)$,  and data traces
-$\{d(\bx_r,t;\bx_s): 0 \le t \le t_{\rm max}\}$, find a bulk modulus $\kappa$ so that the model
-$m=(\kappa,\rho,w)$ satisfies $F[m] \approx d$ in some sense. The
-simplest version of FWI concretizes this task by asking for a model
-$m$ (that is, a bulk modulus $\kappa$) minimizing the square integral/sum
-error over the recording time interval $[0,t_{\rm max}]$ and
-source-receiver position pairs:
+
+As mentioned in the introduction, I will assume that the coefficients
+$\rho$ and $\kappa$ are bounded and smooth:
+\begin{equation}
+  \label{eqn:coeff}
+  \log \rho, \log \kappa \in C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)
+\end{equation}
+Then the symmetric hyperbolic system \ref{eqn:awe} has a distribution
+solution if $f \in \cal{E}'(\bR)$ (for example \cite{Lax:PDENotes}). The Propagation of
+Singularities Theorem (for example, \cite{Tay:81}, Theorem VI.2.1)
+implies that if $f \in \cal{E}'(\bR)$ and $(\bx_s,\bx_r) \in X_{sr}$,
+then the trace $p(\bx_r, \cdot;\bx_s)$ on $\{(\bx_r,t): t \in \bR\}$ is well-defined,
+as a distribution vanishing for sufficiently large
+negative $t$. Define the {\em modeling operator}
+\begin{equation}
+  \label{eqn:modop}
+  F: (C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3))^2 
+  \rightarrow X_{sr} \times {\cal D}'(\bR)
+\end{equation}
+by
+\begin{equation}
+  \label{eqn:modopdef}
+  F[\kappa,\rho](\bx_r,\bx_s) = \phi p(\bx_r,\cdot;\bx_s).
+\end{equation}
+For sake of brevity, define the model vector $m=(\kappa,\rho)$.
+
+For the purposes of this paper, the source wavelet $f$ will be assumed
+known, and to satisfy
+\begin{itemize}
+\item[F1. ] $f \in H^1_0(\bR)$,
+\item[F2. ] $f$ is odd, $f(t) = -f(-t)$.
+\end{itemize}
+In fact, even if actual acoustic sources are well-approximated by
+isotropic point radiators, their wave forms are only roughly known
+{\em a priori} and may reasonably be included as parameters to be
+estimated by inversion of the data. On the other hand, in applications
+to seismology, medical imaging, and other wave imaging technologies,
+the source wave form $f$ is actually the product of signal processing
+aimed to shape the data $F[m]$ into an approximation of data that
+would have been predicted using a prescribed $f$. That is, the wavelet
+is to some extent {\em chosen}. This circumstance justifies the
+constraints F1, F2 placed on $f$.
+
+The density $\rho$ will be presumed spatially homogeneous, or at least
+to have the same value at all of the source positions $\bx_s\in X_s$:
+introduce the {\em source density} $\rho_s > 0$ for which
+\begin{equation}
+  \label{eqn:rho}
+  \rho(\bx_s) = \rho_s\mbox{ for any } \bx_s \in X_s.
+\end{equation}
+In some settings, eg. the earth's
+upper crust, density exhibits substantially less variance than the
+bulk modulus, Also, for some acquisition designs, such as marine towed
+streamers and some medical ultrasonic technologies, sources are
+immersed in water and thus in positions with approximately the same
+densities.
+
+In any event, for the present discussion, $f$ is regarded
+as known and subject to F1 and F2, and $\rho$ is subject to the constraint \ref{eqn:rho}.
+
+In this setting, acoustic FWI means: given the set of
+source-receiver position pairs $X_{sr}$, source wavelet
+$f$,  source density $\rho_s$, and data traces
+$d:  X_{sr} \rightarrow \cal{D}(\bR)$, find $\kappa$ and $\rho$
+(subject to the constraint \ref{eqn:rho}) so that the model
+$m=(\kappa,\rho)$ satisfies $F[m] \approx d$ in some sense. If
+$d$ and $F[m]$ lie in a normed subspace of $\cal{D}(\bR)$ for each
+$(\bx_s,\bx_r) \in X_{sr}$, then the
+norm provides a straightforward candidate for this sense of
+approximation.
+
+\subsection{Hadamard's Decomposition}
+The Hadamard decomposition of the acoustic Green's function provides a
+basis for a norm-based formulation of FWI \cite[]{Friedlander:75,Qian:JCP24}. As presented by
+\cite{Friedlander:75}, for instance, this construction applies to
+second-order wave equations, but the system \ref{eqn:awe} is
+equivalent fo a second-order problem: differentiate the first equation
+with respect to $t$, eliminate $\bv$ via the second equation, and
+rearrange to obtain
+\begin{equation}
+  \label{eqn:awe2}
+  \frac{\rho}{\kappa}\frac{\partial^2 p}{\partial t^2} - \nabla^2 p -
+  \nabla \log \rho \cdot \nabla p = \rho \frac{\partial f}{\partial t}
+  \delta(\bx-\bx_s)
+\end{equation}
+The pair $(p,\bv)$ is a distribution solutions of the system
+\ref{eqn:awe} if and only if $p$ is a solution of equation
+\ref{eqn:awe2} and vis-versa (with $\bv$ constructed from $p$ in the
+obvious way, preserving causality).
+
+Hadamard's construction is local, in both space-time and in the set of
+acoustic models. A {\em normal neighborhood} of
+$\bx_s \in \bR^3$ is an open subset of $\bR^3$ 
+within which every point is connected to $\bx_s$ by  a unique ray of
+geometric acoustics (a geodesic for the Riemannian metric associated
+to the equation \ref{eqn:awe2}) (\cite{Friedlander:75}, p. 16). In the
+following, 
+$M \subset (C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3))^2$ and $\Omega
+\subset \bR^3$ are bounded open sets satisfying 
+\begin{itemize}
+  \item[G1. ] for every $(\bx_s,\bx_r) \in X_{sr}$, $\bx_s, \bx_r \in \Omega$;
+  \item[G2. ] for every source point $\bx_s\in X_{s}$ and $(\kappa,\rho) \in M$, $\Omega$ is a
+    normal neighborhood of $\bx_s$.
+  \item[G3. ] for every source point $\bx_s \in X_{s};$
+    $(\kappa,\rho) \in M$, ray ${\bf X}: [0,t_d] \rightarrow \bR^3$
+    with ${\bf X}(0)=\bx_s$, and
+    $t \in [0,t_d]$, either $\{{\bf X}(s): 0\le s \le t\} \subset \Omega$
+    or $\{{\bf X}(s): t < s < t_d\} \subset \bR^3 \setminus \Omega$.
+  \end{itemize}
+The last condition means that once a ray leaves $\Omega$ within the
+recording time interval $[0,t_d]$, it does not return to $\Omega$
+for $t<t_d$.
+
+Note that for any $X_{sr}$ there exist
+non-empty open sets $M$ and $\Omega$ satisfying these conditions: any convex open set
+$\Omega$ satisfying condition G1 satisfies conditions G2 and G3 for constant
+($\bx$-independent) $\kappa$ and $\rho$, and conditions G2 and G3 are open in
+the sense of the $C^{\infty}$ and $L^{\infty}$ topologies.
+
+Under the conditions just described, the construction described by
+\cite{Friedlander:75} produces for each $m \in M$ functions $\tau[m]$ (travel time) and $a[m]$
+(geometric amplitude) on $X_s \times \Omega$ and $b[m]$ (remainder kernel) on $X_s \times \Omega
+\times \bR$, satisfying
+\begin{itemize}
+\item[H1. ] $\tau[m], a[m],$ and $b[m]$ are smooth off the diagonal,
+that is, $\tau[m](\bx_s,\cdot) \in C^{\infty}(\Omega \setminus \{\bx_s\})$
+and similarly for $a[m],b[m]$;
+\item[H2. ] there exists $A>0$ so that for any $(\kappa,\rho) \in M$
+  and $(\bx_s,\bx_r) \in X_{sr}$,
+  $|\log a[m](\bx_s,\bx_r)| < A$;
+\item[H3. ] for any $t>0$, there is $B\ge 0$ so that for
+$(\bx_s,\bx_r) \in X_{sr}, -\infty < s \le t$, and $(\kappa,\rho) \in
+M$, $|b[m](\bx_s,\bx_r,s)| +
+|\partial_tb[m](\bx_s,\bx_r,s)| \le B$.
+\end{itemize}
+
+The Hadamard
+decomposition of the Green's function $G$, i.e. the solution of the
+equation \ref{eqn:awe2} for $f = H/\rho_s$, is (suppressing the
+dependence on $m \in M$ from the notation)::
+\begin{equation}
+  \label{eqn:green}
+  G(\bx_s,\bx,t) = a(\bx_s,\bx)\delta(t - \tau(\bx_s,\bx)) +
+  b(\bx_s,\bx,t)H(t - \tau(\bx_s,\bx))
+\end{equation}
+Under conditions G1 - G3 above, this decomposition is the restriction of
+the Greeen's function to $\Omega \times (-\infty,t_d]$. Accordingly,
+\[
+   F[m](\bx_s,\bx_r)(t) =
+   \phi(t)a(\bx_s,\bx_r)\rho_s\frac{df}{dt}(t-\tau(\bx_s,\bx_r))
+ \]
+\begin{equation}
+  \label{eqn:fwdgreen}
+ + \phi(t)\rho_s\int
+  \,ds \, \frac{df}{dt}(t-s) b(\bx_s,\bx_r,s)H(s-\tau(\bx_s,\bx_r))
+\end{equation}
+
+\subsection{FWI via Least Squares}
+It follows from the Hadamard decomposition 
+(equation \ref{eqn:fwdgreen}) that $F[m]$ is
+square-integrable. Indeed, for each $(\bx_s,\bx_r) \in X_{sr}$, the first term on the RHS of equation \ref{eqn:fwdgreen} is clearly
+square-integrable, since $f \in H^1_0(\bR)$.
+An $L^2$ bound on the second (remainder) term follows from the
+pointwise bounds on $b$ described in the last section and Young's
+inequality. In fact, there exists $C>0$ so that for $(\kappa,\rho) \in
+M$,
+\begin{equation}
+  \label{eqn:fwdnorm}
+  \|F[m]\|_d \le C
+  \left\|\frac{df}{dt}\right\|_{L^2(\bR)}.
+\end{equation}
+in which the left-hand side uses the abbreviation
+\[
+  \| \cdot \|_d = \| \cdot \|_{L^2(X_{sr} \times \bR)}
+\]
+
+This observation justfies posing FWI as a least-squares problem: given $d \in
+L^2(X_{sr}\times \bR)$, $f \in H^1_0(\bR)$, and $\rho_s>0$, find $m = (\kappa,\rho)$ to minimize
 \begin{equation}
   \label{eqn:fwi}
-  J_{\rm FWI}[m;d] =\frac{1}{2} \sum_{{\bf x}_r,{\bf x}_s}
-  \int_0^{t_{\rm max}}\,dt\, |F[m]({\bf x}_r,t;{\bf x}_s) - d({\bf x}_r,t;{\bf x}_s)|^2
-\end{equation}
-The definition \ref{eqn:fwi} makes sense only if the predicted and
-observed data traces ($F[m]({\bf x}_r\cdot;{\bf x}_s)$ and $d({\bf
-  x}_r,\cdot;{\bf x}_s)$) are square-integrable functions on $[0,t_{\rm max}]$, for each
-source-receiver position pair $(\bx_r,\bx_s)$. This property is
-assumed for $d$, and follows from additional assumptions on $m$ and the
-acqquisition geometry, to be explained below.
-
-It is convenient to introduce the $L^2$ norm symbol $\| \cdot \|$ as
-shorthand for the right-hand side of the definition \ref{eqn:fwi}:
-\[
-  J_{\rm FWI}[m;d]= \frac{1}{2}\|F[m]-d\|^2.
-\]
-In this notation, the interval $[0,t_{\rm max}]$ and the
-source-receiver position pair set $\{(\bx_r,\bx_s)\}$ are implicit. In
-the sequel, I will use the $L^2$ norm notation to stand for square
-integral/sum over whatever variables make sense in context.
-
-Practical versions of FWI include various types of regularization and
+  J_{\rm FWI}[m;d]= \frac{1}{2}\|F[m]-d\|_d^2.
+\end{equation}
+Practical versions of least-squares FWI include various types of regularization and
 imposition of constraints on $m$, see the references cited in the
-Introduction. As mentioned there, local iterative optimization methods
-are the only practical approach to minimizing $J_{\rm FWI}$ due to the
-computational cost of approximating the forward map $F$ and related
-operators, but these tend to stagnate at highly sub-optimal
-(``cycle-skipped'') estimates of $m$ unless the initial estimate is of
-quite high quality, often available only at considerable cost in data
-collection and processing.
-
-\subsection{Source Characteristics}
-
-In applications to seismology, medical imaging, and other wave imaging
-technologies, the wavelet $w$ is actually the product of signal
-processing aimed to shape the data $F[m]$ into an approximation of
-data that would have been predicted using a known wavelet. That is, the wavelet is to some extent {\em
-  chosen}. This circumstance justifies several constraints on the
-wavelet, consistent with the properties of observed signals: the
-wavelet is assumed to be of class $H^1_0(\bR)$, that is, square integrable with
-square inttegrable derivative and vanishing for sufficiently large
-$t$, and even in $t$.
+Introduction.
+%As mentioned there, local iterative optimization methods
+%are the only practical approach to minimizing $J_{\rm FWI}$ due to the
+%computational cost of approximating the forward map $F$ and related
+%operators, but these tend to stagnate at highly sub-optimal
+%(``cycle-skipped'') estimates of $m$ unless the initial estimate is of
+%quite high quality, often available only at considerable cost in data
+%collection and processing.
+
+\subsection{Source Characteristics and Asymptotics}
+
+Note that in the representation \ref{eqn:fwdgreen}, only the
+derivative $df/dt$ appears. For convenience, introduce
+\begin{equation}
+  \label{eqn:wavelet}
+  w = \rho_s \frac{df}{dt}
+\end{equation}
+From the assumed
+properties of $f$, $w \in L^2(\bR)$ and has compact support, has zero
+mean, and is
+even in $t$. Since $f$ with the properties F1 - F3 corresponds
+biuniquely with $w$ with the properties just mentioned, I will also
+refer to $w$ as ``the wavelet'', and view $F$ as parametrized by $w$.
+
 Introduce asymptotics in this setting by assuming that $w = \wl$ is chosen
 from a family $\{\wl:\}$ indexed by the dimensionless parameter
-$0 <\lambda <1\}$, and defined by
+$0 <\lambda \le 1\}$, and defined by
 \begin{equation}
@@
 \end{equation}
-The  ``mother wavelet'' $w_1$ is even and of class $H^1_0(\bR)$, and
-all wavelets $\wl$ in this family
-inherit these properties from $w_1$, along with its dimensions (energy/time$^2$).
-
-It follows from these assumptions that the Fourier transform $\fwl$ is
-real-valued, even, of class $H^1(\bR)$. and its product with frequency $|\omega|$ is also
-square-integrable.
+The  ``mother wavelet'' $w_1$ is square-integrable of compact support
+even, and mean-zero. Define $t_w$ to be the diameter of the support:
+\begin{equation}
+  \label{eqn:wdiam}
+  \mbox{supp }w_1 = [-t_w,t_w].
+\end{equation}
+All wavelets $\wl$ in this family
+inherit these properties from $w_1$, along with its dimensions
+(energy/time$^2$). It follows from these assumptions that the Fourier transform $\fwl$ is
+real-valued, even, of class $H^1(\bR)$. The corresponding source waveform $f_{\lambda}$ is
+given by
+\begin{equation}
+  \label{eqn:flambda}
+  f_{\lambda}(t) = \int_{-\infty}^t\wl =
+  \lambda^{1/2}f_1\left(\frac{t}{\lambda}\right).
+\end{equation}
+Note that
+\begin{equation}
+  \label{eqn:lambdanorms}
+  \|\wl\|_{L^2(\bR)} = \|w_1\|_{L^2(\bR)},\,\|f_{\lambda}\|_{L^2(\bR)}
+  = \lambda \|f_1\|_{L^2(\bR)}.
+\end{equation}
+
+%and its product with frequency $|\omega|$ is also
+%square-integrable.
 
@@
 
+The modeling operator with wavelet $\wl$ is denoted $\Fl$. From the
+decomposition \ref{eqn:fwdgreen}, $\Fl$ is the sum of two terms, the second of which is
+\[
+   \phi(t)\rho_s\int
+   \,ds \, \frac{df_{\lambda}}{dt}(t-s) b(\bx_s,\bx_r,s)H(s-\tau(\bx_s,\bx_r))
+\]
+\[
+  =\phi(t)\rho_s
+  f_{\lambda}(t-\tau(\bx_s,\bx_r))b(\bx_s,\bx_r,\tau(\bx_s,\bx_r))
+\]
+\[
+    -
+    \phi(t)\rho_s\int
+   \,ds \, f_{\lambda}(t-s) (\partial_t
+   b(\bx_s,\bx_r,s))H(s-\tau(\bx_s,\bx_r))
+\]
+The $L^2$ norms of the summands are bounded by a multiple of
+$\lambda$, uniform over $(\bx_s,\bx_r) \in X_{sr}, m \in M$, thanks to
+the norm scaling \ref{eqn:lambdanorms}, Young's inequality, and conclusion H3.
+
+I will write ``$+ \Ol$'' to mean an  error uniformly bounded in $L^2$ by
+a multiple of $\lambda$ over $m \in M$. Thus
+\begin{equation}
+  \label{eqn:fwdapprox}
+  \Fl[m](\bx_s,\bx_r) = \phi a(\bx_s,\bx_r)\wl(t-\tau(\bx_s,\bx_r)) +
+  \Ol
+\end{equation}
 \subsection{AWI via Optimization}
@@
 
-Since linear acoustics is time-translation invariant, its solution commutes with time convolution, that is,
-\[
- \bar{F}[\bar{m}] = \bar{F}[(\kappa,\rho,,u*w)] = F[m]*u
+The adaptive filters $u$ are presumed to vanish for $|t| > t_u$, with
+the same maximum time $t_u>0$ for all $(\bx_s,\bx_r) \in X_{sr}$, and
+to be square integrable: $u \in L^2(X_{sr}\times [-t_u,t_u])$.
+
+Since linear acoustics is time-translation invariant, its solution commutes with time convolution.
+%\[
+% \bar{F}[\bar{m}] = \bar{F}[(\kappa,\rho,,u*w)] = F[m]*u
+% \]
+Denote by $\bar{p}(\bx,t;\bx_s,\bx_r), \bar{\bv}(\bx,t;\bx_s,\bx_r)$
+the solution of the acoustic system \ref{eqn:awe} with $w(t)$ in the
+right-hand side of the first equation replaced by
+$\bar{w}(\bx_r,\cdot;\bx_s) = u({\bf  x}_r,\cdot;{\bf
+  x}_s)*w$. Then
+\[
+  \bar{p}(\bx,\cdot;\bx_s,\bx_r) = u({\bf  x}_r,\cdot;{\bf
+    x}_s)*p(\bx,\cdot;\bx_s)
 \]
@@
   \label{eqn:sdef}
-  F[m]*u \equiv S[m]u
-\end{equation}
+  S[m]u(\bx_r,t;\bx_s) \equiv \phi(t)\bar{p}(\bx_r,t;\bx_s,\bx_r).
+\end{equation}
+
 \cite{Warner:16} assume (implicitly) that there is always an exact
@@
 $S[m]u[m;d]= d$. This proposition is dubious, but is easy to work
-around, as will be seen below. Also, \cite{Warner:16} formulate
+around, as will be seen below - so for the moment accept it.
+
+Also, \cite{Warner:16} formulate
 AWI in terms of the inverse filter, that is, $u$ for which
-$u*d = F[m]$, and in fact recommend that definition. The present
+$\phi u*d = F[m]$, and in fact recommend that definition. The present
 discussion uses only the forward filter.
@@
 $m$, described in the last section, should be any more effective than
-FWI. To see why AWI works when FWI does not, in at
-least some settings, assume that the spatial dimension is three,
-and make use of the geometric asymptotics (ray theory) approximation
-to the solution of the wave equation. This approximation is {\em
+FWI. Under the smoothness and ray geometry assumptions made above (equation
+\ref{eqn:coeff}, conditions G1 - G3) hold, an asymptotic relation
+between AWI and travel time tomography emerges, justifying the notion
+that AWI is more likely to succeed than is FWI. This relation is based
+on the Hadamard decomposition, and therefore is {\em
   local}: it applies when sources and receivers are sufficiently
-close. I will discuss its global failure and the consequences for AWI
-in the fifth section of this paper.
-
-Since the fields $\kappa$ and $\rho$ have been assumed smooth,
-the geometric asymptotics approximation is applicable for solution of the point
-radiator problem \ref{eqn:awe}. This approximation expresses the pressure field in terms
-of $m$-dependent geometric amplitude $a[m]({\bf x}_s,{\bf x}_r)$ and
-travel time $\tau[m]({\bf x}_s,{\bf x}_r)$, and implies that
-\begin{equation}
-  \label{eqn:hadamard}
-  F[m_{\lambda}]({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) \wl(t-\tau[m]({\bf x}_s,{\bf x}_r)). 
-\end{equation}
-Here and in the notation to follow, $m_{\lambda} =
-(\kappa,\rho,\wl)$.
-
-In this approximation, and in other uses of the symbol $\approx$, the
-error is dominated by data wavelength, that is, RMS wavelength of the
-source wavelet $\wl$, which is proportional to $\lambda$ as mentioned above. In other words, ``$\approx$'' means ``in error by $O(\lambda)$''.
-
-Approximation \ref{eqn:hadamard} follows from a
-``progressing wave''  expansion of the fundamental
-solution (Green's function) of the wave equation due to Hadamard
-\cite[]{Friedlander:75,Qian:JCP24}. For smooth (slowly varying)
-$m$, it holds {\em locally}, that is, for $\bx_r$ sufficiently close to $\bx_s$ that a unique ray of
-geometric acoustics connect $\bx_s$ to $\bx_r$. This ``single
-arrival'' condition fails (in
-general) at larger distance. The precise region of validity depends on
-$m$. For the remainder of this section, its validity is assumed.
-
-Applying the approximation \ref{eqn:hadamard} to the definition
-\ref{eqn:sdef}, obtain
-\begin{equation}
-  \label{eqn:spwe}
-S[m_{\lambda}]u({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) (\wl*u)(t-\tau[m]({\bf x}_s,{\bf x}_r))
-\end{equation}
-Assume that the data is noise-free:
-$$
-d({\bf x}_r,t;{\bf x}_s)=(F[m^*]) ({\bf x}_r,t;{\bf x}_s) \approx a[m^*]({\bf x}_s,{\bf x}_r) \wl(t-\tau[m^*]({\bf x}_s,{\bf x}_r))
-$$
-The (approximate) solution of $S[m_{\lambda}]u=d$ is obvious by inspection:
+close to lie in a common normal neighborhood of the source locations,
+for all coefficient pairs $(\kappa,\rho)$ in the search set $M$ of the
+inverse problem. The relation to travel time tomography is asymptotic
+in wavelength/frequency, and conveniently expressed via the wavelet
+family $\{\wl\}$ and the associated family of modeling operators
+$\Fl$. With these assumptions, the AWI objective function $J_{\rm AWI}[m;d_{\lambda}$
+for noise-free data
+\begin{equation}
+  \label{eqn:noisefree}
+  d_{\lambda}=\Fl[m^*]
+\end{equation}
+is $\lambda$-asymptotic to the mean-square difference between the
+source-receiver travel times for $m$ and $m*$.
+
+This relation breaks down if the sources and receivers fail to belong
+to a normal neighborhood (that is, if multiple ray paths exist between
+sources and receivers), as explained in the fifth section of this paper.
+
+With these assumptions,
+the decomposition \ref{eqn:fwdgreen} is applicable for solution of the point
+radiator problem \ref{eqn:awe} with extended source
+$u(\bx_r,\cdot;\bx_s)*\wl$, for each source-receiver pair
+$(\bx_s,\bx_r)\in X_{sr}$. Write $\bar{p}_{\lambda}$ for the pressure
+component of the solution, and specialize the definition
+\ref{eqn:sdef} to
+\begin{equation}
+  \label{eqn:sdeflam}
+  S_{\lambda}[m]u(\bx_r,t;\bx_s) \equiv \phi(t)\bar{p}_{\lambda}(\bx_r,t;\bx_s,\bx_r).
+\end{equation}
+Following the reasoning leading to
+equation\ref{eqn:fwdapprox}, obtain
+\begin{equation}
+  \label{eqn:slamapprox}
+  S_{\lambda}[m]u({\bf x}_r,t;{\bf x}_s) = \phi(t)a[m]({\bf x}_s,{\bf
+    x}_r) (u(\bx_r,\cdot;\bx_s)*\wl)(t-\tau[m]({\bf x}_s,{\bf x}_r)) + O(\lambda).
+\end{equation}
+
+I promised additional constraints on the data time window $[0,t_d]$
+and cutoff function $\phi$. These are posed in terms of the extreme
+traveltimes over $m \in M$. Since all derivatives of the log wave velocity $\log c =
+(\log \kappa - \log \rho )/2$ are bounded in the closure of $\Omega$,
+uniformly over $m \in M$, 
+the ray trace construction of travel time implies that
+$0 < \tau_{\rm min} < \tau_{\rm max}$ exist for which $\tau_{\rm min}
+\le \tau[m](\bx_s,\bx_r) \le \tau_{\rm max}$ for $(\bx_s,\bx_r) \in
+X_{sr}$, $m \in M$.
+The additional constraints on $t_d$ and $\phi$,
+\begin{itemize}
+\item[T1. ] $t_d >  \tau_{\rm max} + t_w + t_u $,
+\item[T2. ] $0 < \tau_{\rm min} - t_w - t_u$,
+\item[T3. ] $\phi \equiv 1$ on $[\tau_{\rm min} - t_w - t_u,
+\tau_{\rm max} + t_w + t_u]$.
+\end{itemize}
+That is, the support of the leading term in \ref{eqn:slamapprox} is
+contained in $\{t: \phi(t) = 1\}$. These assumptions are physically natural, as
+they imply that the recording time interval $[0,t_d]$ contains the
+support of the extended source data, up to an error vanishing as
+$\lambda \rightarrow 0$.
+
+From T3, equation \ref{eqn:slamapprox} can be re-written as
+\begin{equation}
+  \label{eqn:slambis}
+  S_{\lambda}[m]u({\bf x}_r,t;{\bf x}_s) = a[m]({\bf x}_s,{\bf
+    x}_r) (u(\bx_r,\cdot;\bx_s)*\wl)(t-\tau[m]({\bf x}_s,{\bf x}_r)) + O(\lambda).
+\end{equation}
+
+Similarly, from the definition \ref{eqn:noisefree} and equation
+\ref{eqn:fwdgreen},
+\begin{equation}
+  \label{eqn:noisefreeapprox}
+  d_{\lambda}(\bx_r,t;\bx_s) = a[m^*](\bx_s,\bx_r) \wl(t-\tau[m^*](\bx_s,\bx_r)) + O(\lambda).
+  \end{equation}
+
+Inspection of equations \ref{eqn:slamapprox} and
+\ref{eqn:noisefreeapprox} shows that an approximate solution of $S_{\lambda}[m]u=d_{\lambda}$ is 
 \begin{equation}
   \label{eqn:notl2}
-  u({\bf x}_r,t;{\bf x}_s) = \frac{a[m^*]({\bf x}_s,{\bf x}_r)}{a[m]({\bf x}_s,{\bf x}_r)}\delta(t-\tau[m^*]({\bf x}_s,{\bf x}_r)+\tau[m]({\bf x}_s,{\bf x}_r)).
+  u({\bf x}_r,t;{\bf x}_s) = \frac{a[m^*]({\bf x}_s,{\bf
+      x}_r)}{a[m]({\bf x}_s,{\bf x}_r)}\delta(t-\tau[m^*]({\bf
+    x}_s,{\bf x}_r)+\tau[m]({\bf x}_s,{\bf x}_r)) + O(\lambda).
 \end{equation}
@@
 undefined. It's not surprising that mathematical ill-behaviour arises
-in this context. Since $S[m_{\lambda}]$ is a convolution operator, solving
-$S[m_{\lambda}]u=d$ for $u$ amounts to deconovolving $d$ by the kernel $F[m]$ of
-$S[m_{\lambda}]$. Under the assumed conditions on the wavelet $\wl$, $S[m_{\lambda}]$ does
+in this context. Since $S_{\lambda}[m]$ is a convolution operator, solving
+$S_{\lambda}[m]u=d$ for $u$ amounts to deconovolving $d$ by the kernel $F[m]$ of
+$S_{\lambda}[m]$. Under the assumed conditions on the wavelet $\wl$, $S_{\lambda}[m]$ does
 not have a continuous (stable) convolution inverse.
@@
 It is conventional to stabilize deconvolution problems such as the
-equation $S[m_{\lambda}]u=d$ via so-called ``pre-whitening'', as
+equation $S_{\lambda}[m]u=d$ via so-called ``pre-whitening'', as
 mentioned by \cite{Warner:16} ``Pre-whitening'' is a version of
 Tihonov regularization. It  amounts to replacement of the equation
-$S[m_{\lambda}]u=d$ by the regularized least squares problem: 
+$S_{\lambda}[m]u=d$ by the regularized least squares problem: 
 \begin{equation}
   \label{eqn:reg}
-  u_0[m_{\lambda};d] = \mbox{argmin}_u (\|S[m_{\lambda}]u-d\|^2 + \lambda\|u\|^2)
+  u_0[m_{\lambda};d] = \mbox{argmin}_u (\|S_{\lambda}[m]u-d\|^2 + \lambda\|u\|^2)
 \end{equation}
@@
 following choices, for each $m$ in the domain of $F$: $S$ is the regularized
-modeling operator $(S[m_{\lambda}],\sigma I)^T$.  Its domain is $U=L^2(\bR)$, of
+modeling operator $(S_{\lambda}[m],\sigma I)^T$.  Its domain is $U=L^2(\bR)$, of
 which $u$ (the adaptive filter) is a member, and 
@@
 augmented data $(d,0)^T$. $T$ is a modified multiply-by-$t$ operator
-$Tu(t)=\max(|t|, t_{\rm max}) \mbox{sgn}(t)u(t)$, with domain $U$ and
-range $E=L^2(\bR)$. The truncation by $t_{\rm max} > 0$ is necessary
+$Tu(t)=\max(|t|, t_d) \mbox{sgn}(t)u(t)$, with domain $U$ and
+range $E=L^2(\bR)$. The truncation by $t_d > 0$ is necessary
 to make $T$ bounded, but can be chosen large enough relative to the
@@
   \label{eqn:eq5}
-   J_{{\rm MSWI},\alpha,\sigma}[m,u;d] = \frac{1}{2}\left(\|S[m_{\lambda}]u-d\|^2 +
+   J_{{\rm MSWI},\alpha,\sigma}[m,u;d] = \frac{1}{2}\left(\|S_{\lambda}[m]u-d\|^2 +
    \lambda\|u\|^2 + \alpha^2\|Tu\|^2\right)
@@
   \label{eqn:eq8}
-  T[m;d] u(\bx_r,t;\bx_s) = \frac{\max(|t|, t_{\rm max})
+  T[m;d] u(\bx_r,t;\bx_s) = \frac{\max(|t|, t_d)
     \mbox{sgn}(t)}{\|u_0[m_{\lambda};d](\bx_r,\cdot;\bx_s)\|}
@@
   \label{eqn:eq5a}
-   J_{{\rm AWI},\alpha,\sigma}[m,u;d] = \frac{1}{2}\left(\|S[m_{\lambda}]u-d\|^2 +
+   J_{{\rm AWI},\alpha,\sigma}[m,u;d] = \frac{1}{2}\left(\|S_{\lambda}[m]u-d\|^2 +
    \lambda\|u\|^2 + \alpha^2\|T[m;d]u\|^2\right)
@@
   \tilde{J}_{0,\sigma}[m;d] = \frac{1}{2}
-  (\|S[m_{\lambda}]u_0[m_{\lambda};d]-d\|^2 + \lambda \|u_0[m_{\lambda};d]\|^2.
+  (\|S_{\lambda}[m]u_0[m_{\lambda};d]-d\|^2 + \lambda \|u_0[m_{\lambda};d]\|^2.
 \]
@@
 \[
-\|S[m_{\lambda}]u_0[m_{\lambda};d]-d\|^2 
+\|S_{\lambda}[m]u_0[m_{\lambda};d]-d\|^2 
   \approx \int d\omega
@@
 \[
-  \|S[m_{\lambda}]u_0[m_{\lambda};d] - d\|^2 = O(\sigma) + O(\lambda(w)).
+  \|S_{\lambda}[m]u_0[m_{\lambda};d] - d\|^2 = O(\sigma) + O(\lambda(w)).
 \]
@@
 adaptive filter or extended source to data (represented by the opertor
-$S[m_{\lambda}]$ in the discussion above) may not have a well-behaved inverse in
+$S_{\lambda}[m]$ in the discussion above) may not have a well-behaved inverse in
 the presence of multiple wavefront arrivals. \cite{Yongetal:GJI23}
@@
 
-Once again, it is possible to write the adaptive filter u for which $S[m_{\lambda}]u = F[m]*u = d =
+Once again, it is possible to write the adaptive filter u for which $S_{\lambda}[m]u = F[m]*u = d =
 F[m^*]$ by inspection. Use the notations $a^*_i, t^*_i, p^*_i$ for
@@
 \[
-  d = F[m_*]= S[m_{\lambda}]u
+  d = F[m_*]= S_{\lambda}[m]u
 \]

### 2024-10-05 repro/papers/awi/paper.tex
more

@@
 
+The conclusions of this study depend on a number of limitations on the
+acoustic model of wave propagation and the formulation of the related
+inverse problem. The principal constraints are:
+\begin{itemize}
+  \item energy sources are modeled as isotropic point radiators, with
+    known time dependence vanishing outside a finite time window;
+  \item mechanical parameter fields (coefficients in the acoustic wave
+    equation) are smooth, that is, infinitely differentiable as
+    functions of position, and the material density is known and
+    position-independent: only the bulk modulus (or wave velocity) is
+    to be determined;
+  \item waves propagate throughout Euclidean space-time - that is, no
+    physical boundaries are modeled;
+  \item the data of the inverse problem are the restrictions of the
+    pressure field to a finite number of points in space (receiver
+    positions), for a finite number of source positions and all time;
+  \item local energy decays exponentially - that is, the pressure and
+    all of its derivatives tend to zero at an exponential rate as $t
+    \rightarrow \infty$,
+    uniformly in any bounded spatial domain.
+  \end{itemize}
+  These constraints are discussed in detail in the next section. Some
+  could certainly be relaxed. For instance, exponential decay of local
+  energy implies that the finite time duration of actual measured data
+  can be treated as a perturbation of the infinite time data setting
+  posed here. However these hypotheses enable a simple and clear
+  analysis of AWI, and seem a reasonable starting point for further
+  investigation.
+
 After establishing the class of problems to be studied and notations
@@
 point sensors, recording pressure at points
-${\bf x}_r \in \bR^3$. The recording takes place over a time interval
-$[0,t_d]$ (also idealized as independent of $\bx_s$ and
-$\bx_r$) . The collection $X_{sr}$ of source-receiver pairs
+${\bf x}_r \in \bR^3$. The collection $X_{sr}$ of source-receiver pairs
 $(\bx_s,\bx_r)$ for which data is recorded should not intersect the
@@
 
-In practice, recording takes place over a finite time interval
-$[0,t_d]$. For convenience, assume that this interval is the same for
-all receivers. Let $\phi \in C_0^{\infty}(0,t_d)$ be a window function
-supported in this interval. Additional constraints on $t_d$ and $\phi$
-will be be specified below.
-
 The pressure and velocity fields $p({\bf x},t;{\bf x}_s)$,
 ${\bf v}({\bf x},t;{\bf x}_s)$ for the source location ${\bf x}_s$
-depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf
-  x})$ (reciprocal of the density $\rho({\bf x})$), and source wave
+depend on the bulk modulus $\kappa$, buoyancy $\beta$
+(reciprocal of the density $\rho$), and source wave
 form $f(t)$ through the acoustic system
@@
 space-time $(\bx,t) \in \bR^3 \times \bR$, and ignore physical
-boundaries such as the Earth's surface.
-
-As mentioned in the introduction, I will assume that the coefficients
-$\rho$ and $\kappa$ are bounded and smooth:
+boundaries such as the Earth's surface. Likewise for simplicity, I
+assume in this study that the material density $\rho$ is spatially
+homogeneous and known. In some settings, eg. the earth's
+upper crust, the material density $\rho$ exhibits substantially less variance than the
+bulk modulus, Density is however spatially variable in most geological
+and biomedical materials, and could be included amongst the parameters
+to be estimated.
+
+The parameter to be determined in the inverse problem studied here is
+the bulk modulus $\kappa$. It is presumed is bounded and smooth:
 \begin{equation}
   \label{eqn:coeff}
-  \log \rho, \log \kappa \in C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)
-\end{equation}
-Then the symmetric hyperbolic system \ref{eqn:awe} has a distribution
-solution if $f \in \cal{E}'(\bR)$ (for example \cite{Lax:PDENotes}). The Propagation of
+  \log \kappa \in C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)
+\end{equation}
+Under these hypotheses, the symmetric hyperbolic system \ref{eqn:awe}
+has a unique distribution
+solution if $f \in \cal{E}'(\bR)$ (for example \cite{Lax:PDENotes},
+Theorem 7.4). The Propagation of
 Singularities Theorem (for example, \cite{Tay:81}, Theorem VI.2.1)
@@
   \label{eqn:modop}
-  F: (C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3))^2 
+  F: C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)
   \rightarrow X_{sr} \times {\cal D}'(\bR)
@@
   \label{eqn:modopdef}
-  F[\kappa,\rho](\bx_r,\bx_s) = \phi p(\bx_r,\cdot;\bx_s).
-\end{equation}
-For sake of brevity, define the model vector $m=(\kappa,\rho)$.
-
-For the purposes of this paper, the source wavelet $f$ will be assumed
-known, and to satisfy
-\begin{itemize}
-\item[F1. ] $f \in H^1_0(\bR)$,
-\item[F2. ] $f$ is odd, $f(t) = -f(-t)$.
-\end{itemize}
-In fact, even if actual acoustic sources are well-approximated by
-isotropic point radiators, their wave forms are only roughly known
-{\em a priori} and may reasonably be included as parameters to be
-estimated by inversion of the data. On the other hand, in applications
-to seismology, medical imaging, and other wave imaging technologies,
-the source wave form $f$ is actually the product of signal processing
-aimed to shape the data $F[m]$ into an approximation of data that
-would have been predicted using a prescribed $f$. That is, the wavelet
-is to some extent {\em chosen}. This circumstance justifies the
-constraints F1, F2 placed on $f$.
-
-The density $\rho$ will be presumed spatially homogeneous, or at least
-to have the same value at all of the source positions $\bx_s\in X_s$:
-introduce the {\em source density} $\rho_s > 0$ for which
-\begin{equation}
-  \label{eqn:rho}
-  \rho(\bx_s) = \rho_s\mbox{ for any } \bx_s \in X_s.
-\end{equation}
-In some settings, eg. the earth's
-upper crust, density exhibits substantially less variance than the
-bulk modulus, Also, for some acquisition designs, such as marine towed
-streamers and some medical ultrasonic technologies, sources are
-immersed in water and thus in positions with approximately the same
-densities.
-
-In any event, for the present discussion, $f$ is regarded
-as known and subject to F1 and F2, and $\rho$ is subject to the constraint \ref{eqn:rho}.
+  F[\kappa]= p|_{X_{sr} \times \bR}
+\end{equation}
 
@@
 source-receiver position pairs $X_{sr}$, source wavelet
-$f$,  source density $\rho_s$, and data traces
-$d:  X_{sr} \rightarrow \cal{D}(\bR)$, find $\kappa$ and $\rho$
-(subject to the constraint \ref{eqn:rho}) so that the model
-$m=(\kappa,\rho)$ satisfies $F[m] \approx d$ in some sense. If
-$d$ and $F[m]$ lie in a normed subspace of $\cal{D}(\bR)$ for each
+$f$ subject to constraints F1 and F2,  density $\rho$, and data traces
+$d:  X_{sr} \rightarrow \cal{D}(\bR)$, find $\kappa$ (or equivalently
+so that $F[\kappa] \approx d$ in some sense. If
+$d$ and $F[\kappa]$ lie in a normed subspace of $\cal{D}(\bR)$ for each
 $(\bx_s,\bx_r) \in X_{sr}$, then the
@@
 approximation.
+
+%In practice, recording takes place over a finite time interval
+%$[0,t_d]$. For convenience, assume that this interval is the same for
+%all receivers. Let $\phi \in C_0^{\infty}(0,t_d)$ be a window function
+%supported in this interval. Additional constraints on $t_d$ and $\phi$
+%will be be specified below.
 
@@
   \nabla \log \rho \cdot \nabla p = \rho \frac{\partial f}{\partial t}
-  \delta(\bx-\bx_s)
-\end{equation}
-The pair $(p,\bv)$ is a distribution solutions of the system
-\ref{eqn:awe} if and only if $p$ is a solution of equation
-\ref{eqn:awe2} and vis-versa (with $\bv$ constructed from $p$ in the
-obvious way, preserving causality).
+  \delta(\bx-\bx_s), \nonumber \\
+  p=0, \, t \ll 0.
+\end{equation}
+If $(p,\bv)$ is a distribution solutions of the system
+\ref{eqn:awe} then $p$ is a solution of equation
+\ref{eqn:awe2}, as follows by differentiating the first equation of
+\ref{eqn:awe} with respect to $t$ and eliminating $\bv$ via the second
+equation. Conversely, if $p$ is a solution of the second-order initial
+value problem\ref{eqn:awe2}, it follows from the Hadamard
+decomposition detailed below and condition F1
+that 
+\[
+  \bv = \beta \int_{-\infty}^t p
+\]
+has an
+integrable singularity at $\bx=\bx_s$, hence 
+defines a distribution, and that $(p,\bv)$ solves the system \ref{eqn:awe}.
 
@@
 geometric acoustics (a geodesic for the Riemannian metric associated
-to the equation \ref{eqn:awe2}) (\cite{Friedlander:75}, p. 16). In the
-following, 
-$M \subset (C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3))^2$ and $\Omega
-\subset \bR^3$ are bounded open sets satisfying 
+to the equation \ref{eqn:awe2}) (\cite{Friedlander:75},
+p. 16). Hadamard's decomposition holds in a space-time cylinder
+$\Omega \times \bR$,  and $\Omega\subset \bR^3$ 
+and $M \subset (C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3))^2$ are bounded open sets satisfying 
 \begin{itemize}
   \item[G1. ] for every $(\bx_s,\bx_r) \in X_{sr}$, $\bx_s, \bx_r \in \Omega$;
-  \item[G2. ] for every source point $\bx_s\in X_{s}$ and $(\kappa,\rho) \in M$, $\Omega$ is a
+  \item[G2. ] for every source point $\bx_s\in X_{s}$ and $\kappa \in M$, $\Omega$ is a
     normal neighborhood of $\bx_s$.
   \item[G3. ] for every source point $\bx_s \in X_{s};$
-    $(\kappa,\rho) \in M$, ray ${\bf X}: [0,t_d] \rightarrow \bR^3$
-    with ${\bf X}(0)=\bx_s$, and
-    $t \in [0,t_d]$, either $\{{\bf X}(s): 0\le s \le t\} \subset \Omega$
-    or $\{{\bf X}(s): t < s < t_d\} \subset \bR^3 \setminus \Omega$.
+    $\kappa \in M$, ray ${\bf X}:\bR^+\rightarrow \bR^3$
+    with ${\bf X}(0)=\bx_s$ and any $t>0$, either $\{{\bf X}(s): 0\le s \le t\} \subset \Omega$
+    or $\{{\bf X}(s): t < s < \infty\} \subset \bR^3 \setminus \Omega$.
   \end{itemize}
-The last condition means that once a ray leaves $\Omega$ within the
-recording time interval $[0,t_d]$, it does not return to $\Omega$
-for $t<t_d$.
-
-Note that for any $X_{sr}$ there exist
-non-empty open sets $M$ and $\Omega$ satisfying these conditions: any convex open set
-$\Omega$ satisfying condition G1 satisfies conditions G2 and G3 for constant
-($\bx$-independent) $\kappa$ and $\rho$, and conditions G2 and G3 are open in
-the sense of the $C^{\infty}$ and $L^{\infty}$ topologies.
-
-Under the conditions just described, the construction described by
-\cite{Friedlander:75} produces for each $m \in M$ functions $\tau[m]$ (travel time) and $a[m]$
-(geometric amplitude) on $X_s \times \Omega$ and $b[m]$ (remainder kernel) on $X_s \times \Omega
-\times \bR$, satisfying
+Condition G3 (the ``no return'' condition) means that once a ray
+leaves $\Omega$, it does not return to $\Omega$. Conditions G1 - G3
+hold if $\kappa$ is constant, and for small perturbations and some
+large perturbations of constant $\kappa$: these are open conditions in
+the topology of $M$.
+
+The arguments to follow depend an additional property of solutions of
+the wave equation \ref{eqn:awe2}:
 \begin{itemize}
-\item[H1. ] $\tau[m], a[m],$ and $b[m]$ are smooth off the diagonal,
-that is, $\tau[m](\bx_s,\cdot) \in C^{\infty}(\Omega \setminus \{\bx_s\})$
-and similarly for $a[m],b[m]$;
-\item[H2. ] there exists $A>0$ so that for any $(\kappa,\rho) \in M$
+\item[G4. ] For any $n > 0$, $f \in {\cal E}'(\bR)$, there exists $K, T$ and $\delta>0$ so
+  that for any 4-multiindex ${\bf m}$ of order $\le n$, $(\bx_s,\bx_r) \in
+ X_{sr}, t \in \bR$,
+  and any $\kappa \in M$, the solution $p$ of \ref{eqn:awe2} is smooth
+  for $t>T$ and satisfies
+  \[
+    |D^{\bf m}p(\bx_r,t;\bx_s)| \le K e^{-\delta t}.
+  \]
+\end{itemize}
+Condition G4 is a special case of {\em local energy decay}, a much
+studied topic. Properties which imply G4 are in turn implied by the
+{\em non-trapping} hypothesis on the model $\kappa \in M$: all rays of
+geometric asymptotics escape to infinity (\cite{Hristova:09}, Theorem
+2, \cite{EgorovShubin}, Theorem 2.104, p. 192).
+
+Under conditions G1 - G4, the construction described by
+\cite{Friedlander:75} produces for each $m \in M$ functions $\tau[\kappa]$ (travel time), $a[\kappa]$
+(geometric amplitude) on $X_s \times \Omega$ and $b[\kappa]$
+(remainder kernel) on $X_s \times \Omega \times \bR$, satisfying
+\begin{itemize}
+\item[H1. ] $\tau[\kappa], a[\kappa],$ and $b[\kappa]$ are smooth off the diagonal,
+that is, $\tau[\kappa](\bx_s,\cdot) \in C^{\infty}(\Omega \setminus \{\bx_s\})$
+and similarly for $a[\kappa],b[\kappa]$;
+\item[H2. ] there exists $A>0$ so that for any $\kappa \in M$
   and $(\bx_s,\bx_r) \in X_{sr}$,
-  $|\log a[m](\bx_s,\bx_r)| < A$;
-\item[H3. ] for any $t>0$, there is $B\ge 0$ so that for
-$(\bx_s,\bx_r) \in X_{sr}, -\infty < s \le t$, and $(\kappa,\rho) \in
-M$, $|b[m](\bx_s,\bx_r,s)| +
-|\partial_tb[m](\bx_s,\bx_r,s)| \le B$.
+  $|\log a[\kappa](\bx_s,\bx_r)| < A$;
+\item[H3. ] there exist $B \ge 0,\delta > 0$ so that for
+$(\bx_s,\bx_r) \in X_{sr}$, and $\kappa \in
+M$,
+\begin{eqnarray}
+  |b[\kappa](\bx_r,t;\bx_r,s)| &\le& B e^{-\delta t} \nonumber \\
+  |\partial_t b[\kappa](\bx_r,t;\bx_r,s)| &\le& B e^{-\delta t}
+  \label{eqn:best}
+\end{eqnarray}
 \end{itemize}
+(The exponential decay estimate in item H3 follows from hypothesis G4.)
 
 The Hadamard
-decomposition of the Green's function $G$, i.e. the solution of the
-equation \ref{eqn:awe2} for $f = H/\rho_s$, is (suppressing the
-dependence on $m \in M$ from the notation)::
+decomposition of the Green's function $G[\kappa]$, i.e. the solution of the
+equation \ref{eqn:awe2} for $f = \beta H$, is::
 \begin{equation}
   \label{eqn:green}
-  G(\bx_s,\bx,t) = a(\bx_s,\bx)\delta(t - \tau(\bx_s,\bx)) +
-  b(\bx_s,\bx,t)H(t - \tau(\bx_s,\bx))
-\end{equation}
-Under conditions G1 - G3 above, this decomposition is the restriction of
-the Greeen's function to $\Omega \times (-\infty,t_d]$. Accordingly,
-\[
-   F[m](\bx_s,\bx_r)(t) =
-   \phi(t)a(\bx_s,\bx_r)\rho_s\frac{df}{dt}(t-\tau(\bx_s,\bx_r))
- \]
+  G[\kappa](\bx_s,\bx,t) = a[\kappa](\bx_s,\bx)\delta(t - \tau[\kappa](\bx_s,\bx)) +
+  b[\kappa](\bx_s,\bx,t-\tau[\kappa])H(t - \tau[\kappa](\bx_s,\bx))
+\end{equation}
+%Under conditions G1 - G3 above, this decomposition is the restriction of
+%the Greeen's function to $\Omega \times (-\infty,t_h]$. This fact
+%implies a decomposition of $F[\kappa]$, provided that the convolution $G*f$
+%restricted to $[0,t_d]$ is determined by the restriction of $G$ to
+%$\Omega \times [0,t_h]$. That is the case if $t_h > t_d + |\mbox{supp
+%|f/2$ (the factor of $1/2$ because $f$ is odd).
+%Making this assumption (which is weaker than assumptions on $t_h$ to be made below),
+%\[
+%   F[\kappa](\bx_s,\bx_r)(t) =
+%   \phi(t)a(\bx_s,\bx_r)\rho_s\frac{df}{dt}(t-\tau[\kappa](\bx_s,\bx_r))
+% \]
+%\begin{equation}
+%  \label{eqn:fwdgreen}
+% + \phi(t)\rho_s\int
+%  \,ds \, \frac{df}{dt}(t-s) b[\kappa](\bx_s,\bx_r,s-\tau[\kappa](\bx_s,\bx_r))H(s-\tau[\kappa](\bx_s,\bx_%r))
+%\end{equation}
+
+\subsection{FWI via Least Squares}
+Suppose that the source wavelet $f$ satisfies
+\begin{itemize}
+\item[F1. ] $f \in H^1_0(\bR)$, $\mbox{supp } \subset [-t_w,t_w]$;
+\item[F2. ] $f$ is odd, $f(t) = -f(-t)$.
+\end{itemize}
+Then the Hadamard decomposition 
+(equation \ref{eqn:green}) together with local
+energy decay (assumption G4) and Young's inequality  implies that
 \begin{equation}
   \label{eqn:fwdgreen}
- + \phi(t)\rho_s\int
-  \,ds \, \frac{df}{dt}(t-s) b(\bx_s,\bx_r,s)H(s-\tau(\bx_s,\bx_r))
-\end{equation}
-
-\subsection{FWI via Least Squares}
-It follows from the Hadamard decomposition 
-(equation \ref{eqn:fwdgreen}) that $F[m]$ is
-square-integrable. Indeed, for each $(\bx_s,\bx_r) \in X_{sr}$, the first term on the RHS of equation \ref{eqn:fwdgreen} is clearly
-square-integrable, since $f \in H^1_0(\bR)$.
-An $L^2$ bound on the second (remainder) term follows from the
-pointwise bounds on $b$ described in the last section and Young's
-inequality. In fact, there exists $C>0$ so that for $(\kappa,\rho) \in
-M$,
+  F[\kappa] = G[\kappa] * \rho \partial_t f \in L^p(X_{sr} \times \bR)
+\end{equation}
+for any $p \ge 1$, and that
+the corresponding estimates are uniform in $\kappa \in M$:
+there exists $C>0$ so that for $\kappa\in M$, $p \ge 1$,
 \begin{equation}
   \label{eqn:fwdnorm}
-  \|F[m]\|_d \le C
-  \left\|\frac{df}{dt}\right\|_{L^2(\bR)}.
-\end{equation}
-in which the left-hand side uses the abbreviation
-\[
-  \| \cdot \|_d = \| \cdot \|_{L^2(X_{sr} \times \bR)}
-\]
+  \|F[\kappa]\|_{L^p(X_{sr} \times \bR)} \le C \left\|\frac{df}{dt}\right\|_{L^p(\bR)}.
+\end{equation}
+%\[
+%  \| \cdot \|_d = \| \cdot \|_{L^2(X_{sr} \times \bR)}
+%\]
 
 This observation justfies posing FWI as a least-squares problem: given $d \in
-L^2(X_{sr}\times \bR)$, $f \in H^1_0(\bR)$, and $\rho_s>0$, find $m = (\kappa,\rho)$ to minimize
+L^2(X_{sr}\times \bR)$, $f \in H^1_0(\bR)$, and $\rho_s>0$, find $\kappa$ to minimize
 \begin{equation}
   \label{eqn:fwi}
-  J_{\rm FWI}[m;d]= \frac{1}{2}\|F[m]-d\|_d^2.
-\end{equation}
+  J_{\rm FWI}[m;d]= \frac{1}{2}\|F[\kappa]-d\|_{L^2(X_{sr} \times \bR)},
+\end{equation}
+%in which we have introduced the abbreviation
+%\[
+%  \| \cdot \|_d = \| \cdot \|_{L^2(X_{sr} \times \bR)}
+%\]
 Practical versions of least-squares FWI include various types of regularization and
@@
 Note that in the representation \ref{eqn:fwdgreen}, only the
-derivative $df/dt$ appears. For convenience, introduce
+derivative $\partial_tf$ appears. For convenience, introduce
 \begin{equation}
   \label{eqn:wavelet}
-  w = \rho_s \frac{df}{dt}
-\end{equation}
-From the assumed
-properties of $f$, $w \in L^2(\bR)$ and has compact support, has zero
-mean, and is
-even in $t$. Since $f$ with the properties F1 - F3 corresponds
-biuniquely with $w$ with the properties just mentioned, I will also
+  w = \rho \partial_t f
+\end{equation}
+The properties F1 and F2 assumed for $f$ are equivalent to these
+properties of $w$:
+\begin{itemize}
+\item[W1. ] $w \in L^2_0(\bR)$, $\mbox{supp }w \subset [-t_w,t_w]$;
+\item[W2. ] $w$ is even;
+\item[W3. ] $\int \, w = 0$.
+\end{itemize}
+I will also
 refer to $w$ as ``the wavelet'', and view $F$ as parametrized by $w$.
 
-Introduce asymptotics in this setting by assuming that $w = \wl$ is chosen
+Introduce asymptotics in this setting by choosing $w = \wl$
 from a family $\{\wl:\}$ indexed by the dimensionless parameter
@@
 The  ``mother wavelet'' $w_1$ is square-integrable of compact support
-even, and mean-zero. Define $t_w$ to be the diameter of the support:
+even, and mean-zero. So that property W1 holds for the entire family, 
 \begin{equation}
   \label{eqn:wdiam}
-  \mbox{supp }w_1 = [-t_w,t_w].
+  \mbox{supp }w_1 \subset [-t_w,t_w].
 \end{equation}
@@
 \end{equation}
-Note that
+Note that for $p \ge 1$,
 \begin{equation}
   \label{eqn:lambdanorms}
-  \|\wl\|_{L^2(\bR)} = \|w_1\|_{L^2(\bR)},\,\|f_{\lambda}\|_{L^2(\bR)}
-  = \lambda \|f_1\|_{L^2(\bR)}.
+  \|\wl\|_{L^p(\bR)} = \lambda ^{\frac{1}{p}-\frac{1}{2}}\|w_1\|_{L^p(\bR)},\,\|f_{\lambda}\|_{L^p(\bR)}
+  = \lambda ^{\frac{1}{p}+\frac{1}{2} \|f_1\|_{L^p(\bR)}}.
 \end{equation}
@@
 The modeling operator with wavelet $\wl$ is denoted $\Fl$. From the
-decomposition \ref{eqn:fwdgreen}, $\Fl$ is the sum of two terms, the second of which is
-\[
-   \phi(t)\rho_s\int
-   \,ds \, \frac{df_{\lambda}}{dt}(t-s) b(\bx_s,\bx_r,s)H(s-\tau(\bx_s,\bx_r))
-\]
-\[
-  =\phi(t)\rho_s
-  f_{\lambda}(t-\tau(\bx_s,\bx_r))b(\bx_s,\bx_r,\tau(\bx_s,\bx_r))
-\]
-\[
+decomposition \ref{eqn:fwdgreen}, $\Fl$ is the sum of two terms:
+\begin{equation}
+  \label{eqn:Flamdecomp}
+  \Fl = \Fla + \Flb.
+\end{equation}
+where
+\begin{equation}
+  \label{eqn:Fl0}
+  \Fla[\kappa](\bx_r,t;\bx_s) =
+  a[\kappa](\bx_s,\bx_r)\wl(t-\tau[\kappa](\bx_s,\bx_r))
+\end{equation}
+and
+\[
+\Flb[\kappa] = \rho_s\int
+   \,ds \, \frac{df_{\lambda}}{dt}(t-s) b[\kappa](\bx_s,\bx_r,s-\tau[\kappa](\bx_s,\bx_r))H(s-\tau[\kappa](\bx_s,\bx_r))
+\]
+\[
+  =\rho_s
+  f_{\lambda}(t-\tau(\bx_s,\bx_r))b[\kappa](\bx_s,\bx_r,0)
+\]
+\begin{equation}
+  \label{eqn:Fl1}
     -
-    \phi(t)\rho_s\int
+    \rho_s\int
    \,ds \, f_{\lambda}(t-s) (\partial_t
-   b(\bx_s,\bx_r,s))H(s-\tau(\bx_s,\bx_r))
-\]
-The $L^2$ norms of the summands are bounded by a multiple of
-$\lambda$, uniform over $(\bx_s,\bx_r) \in X_{sr}, m \in M$, thanks to
-the norm scaling \ref{eqn:lambdanorms}, Young's inequality, and conclusion H3.
-
+   b[\kappa](\bx_s,\bx_r,s-\tau[\kappa](\bx_s,\bx_r)))H(s-\tau[\kappa](\bx_s,\bx_r))
+\end{equation}
+From hypothesis G4, the bound \ref{eqn:lambdanorms} , and Young's inequality, the $\|\Fl1[\kappa]\|_{ L^p(X_{sr} \times \bR)}$ is bounded by a multiple of
+$\lambda^{\frac{1}{p}+\frac{1}{2}}$, uniformly over $\kappa \in M$.
 I will write ``$+ \Ol$'' to mean an  error uniformly bounded in $L^2$ by
@@
   \label{eqn:fwdapprox}
-  \Fl[m](\bx_s,\bx_r) = \phi a(\bx_s,\bx_r)\wl(t-\tau(\bx_s,\bx_r)) +
-  \Ol
-\end{equation}
+  \Fl[\kappa](\bx_s,\bx_r) = \Fla[\kappa] +   \Ol
+\end{equation}
+
 \subsection{AWI via Optimization}
@@
 The extended modeling operator ${\bar F}$ maps extended models
-$\bar{m}=(\kappa,\rho,\bar{w})$ incorporating extended sources
+$\bar{m}=(\kappa,\bar{w})$ incorporating extended sources
 $\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
-field. That is, the extended source depends on the receiver location
-as well as the source location - so there is one acoustic system for
-each source {\em and} receiver position - a lot of wave equations! If all
-of the wavelets for each source are the same, that is, ${\bar w}({\bf
-  x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver
-position, then ${\bar F}[\bar{m}] = F[m]$. That is, $F$ is a special
+field. Denote by $\bar{p}(\bx,t;\bx_s,\bx_r), \bar{\bv}(\bx,t;\bx_s,\bx_r)$
+the solution of the acoustic system \ref{eqn:awe} with $w(t)$ in the
+right-hand side of the first equation replaced by
+$\bar{w}(\bx_r,\cdot;\bx_s)$. Then
+$\bar{F}[\kappa,\bar{w}](\bx_r,t;\bx_s) =
+\bar{p}(\bx_r,t;\bx_s,\bx_r)$. As was the case for $F$, the Hadamard
+decomposition \ref{eqn:green} and conditions G1-G4, W1-W3 imply that
+$\bar{F}[\kappa, \cdot]$ is a bounded map: $L^2(X_{sr} \times \bR)
+\rightarrow L^2(X_{sr} \times \bR)$. If all  
+of the wavelets are the same, that is, ${\bar w}({\bf  
+  x}_r,t;{\bf x}_s) = w(t)$ is independent of source and receiver  
+position, then ${\bar F}[\kappa,\bar{w}] = F[\kappa]$. That is, $F$ is a special
 case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
@@
 \cite{Warner:16} represent the extended sources $\bar{w}$ as time convolutions of the
-(known) source $w$ with an {\em adaptive filter} $u({\bf
-  x}_r,t;{\bf x}_s)$: that is,
-\[
-  \bar{w}({\bf x}_r,t;{\bf x}_s) = (u * w)({\bf  x}_r,t;{\bf x}_s)
-  = (u({\bf  x}_r,\cdot;{\bf x}_s)*w)(t).
-\]
-Here, $u*w$ denotes the result of convolving each trace in $u$ by $w$. I
+(known) source $w$ with an {\em adaptive filter} $u \in L^2(X_{sr}
+\times \bR)$, that is, $\bar{w}=u * w$.
+Here, $u*w$ denotes the result of convolving each trace in $u$ by $w
+\in L^2(\bR)$. I
 will also use the asterisk to denote the trace-by-trace convolution
@@
 
-The adaptive filters $u$ are presumed to vanish for $|t| > t_u$, with
-the same maximum time $t_u>0$ for all $(\bx_s,\bx_r) \in X_{sr}$, and
-to be square integrable: $u \in L^2(X_{sr}\times [-t_u,t_u])$.
-
-Since linear acoustics is time-translation invariant, its solution commutes with time convolution.
-%\[
-% \bar{F}[\bar{m}] = \bar{F}[(\kappa,\rho,,u*w)] = F[m]*u
-% \]
-Denote by $\bar{p}(\bx,t;\bx_s,\bx_r), \bar{\bv}(\bx,t;\bx_s,\bx_r)$
-the solution of the acoustic system \ref{eqn:awe} with $w(t)$ in the
-right-hand side of the first equation replaced by
-$\bar{w}(\bx_r,\cdot;\bx_s) = u({\bf  x}_r,\cdot;{\bf
-  x}_s)*w$. Then
-\[
-  \bar{p}(\bx,\cdot;\bx_s,\bx_r) = u({\bf  x}_r,\cdot;{\bf
-    x}_s)*p(\bx,\cdot;\bx_s)
-\]
-Introduce the notation $S[m]$ for the convolution operator with kernel $F[m]$:
+For such extended sources,
+\[
+  \bar{F}[\kappa,u*w] = u*F[\kappa]
+\]
+Introduce the notation $S[\kappa]$ for the operator of convolution
+with $F[\kappa]$, that is,
 \begin{equation}
   \label{eqn:sdef}
-  S[m]u(\bx_r,t;\bx_s) \equiv \phi(t)\bar{p}(\bx_r,t;\bx_s,\bx_r).
-\end{equation}
-
-\cite{Warner:16} assume (implicitly) that there is always an exact
-solution of the extended inversion problem. That is, they assume that
-for any $m$, there is an extended source $\bar{w}$, or equivalently an
-adaptive filter $u[m;d]$ with $\bar{w}=w*u[m;d]$, for which
-$S[m]u[m;d]= d$. This proposition is dubious, but is easy to work
-around, as will be seen below - so for the moment accept it.
-
-Also, \cite{Warner:16} formulate
+  S[\kappa]u \equiv u*F[\kappa]
+\end{equation}
+Then $S[\kappa]$ is bounded on $L^2(X_{sr} \times \bR)$, uniformly in
+$\kappa \in M$, thanks to
+the bound \ref{eqn:fwdnorm} for $p=1$. $S$ also represents an
+extension of $F$, in the sense that with the choice
+$u(\bx_r,t;\bx_s)=\delta(t)$, the output of $F$ is recovered:
+\begin{equation}
+  \label{eqn:sconsist}
+  S[\kappa]\delta = F[\kappa].
+\end{equation}
+
+%Since $S[\kappa]$ involves solution of the wave equation with point source
+%wavelet $u*w$, application of the Hadamard decomposition \ref{eqn:green} requires a furnther
+%constraint on ray geometry, expressed by condition G3. Specifically,
+%\begin{itemize}
+%\item[G4. ]$t_h > t_d + t_w + t_u$.
+%\end{itemize}
+%Then \ref{eqn:green} implies that
+%\begin{equation}
+%  \label{eqn:sop}
+%  S[\kappa]u = \phi w* a[\kappa] T[\kappa] ((\delta + b[\kappa]H)*u)
+%\end{equation}
+%in which $T[\kappa]$ is the time translation operator by $\tau[\kappa]$:
+%\begin{equation}
+%  \label{eqn:Tdef}
+%  T[\kappa]u(\bx_r,t;\bx_s) = u(\bx_r,t-\tau[\kappa](\bx_s,\bx_r);\bx_s),
+%\end{equation}
+%Similarly,
+%\begin{equation}
+ % \label{eqn:fwdgreenop}
+%  F[\kappa] = \phi a[\kappa] T[\kappa]((\delta + b[\kappa]H)*w) 
+%\end{equation}
+
+%Note that values of $b$ for $t > t_d + t_w + t_u$ do not affect the
+%definitions of $S[\kappa]u$ or $F[\kappa]$. Therefore alter $b$ by
+%multiplication by a suitable cutoff function so that its support lies in
+%$(-\infty, t_h]$.
+
+\cite{Warner:16} assume (implicitly) that there is always a unique
+solution of the extended inversion problem, formulated in terms of
+$S$: that is, that
+for any $\kappa$ and any data $d$, there is an
+adaptive filter $u[\kappa;d]$ for which
+$S[\kappa]u[\kappa;d]= d$. If $d$ is noise-free, that is, $d=F[\kappa]$, then the consistency relation \ref{eqn:sconsist}
+suggests that $u[\kappa;d](\bx_r,t;\bx_s) = \delta(t)$. This ideal adaptive filter is in the null
+space of multiplication by $t$, that is, the operator
+$u(\bx_r,t;\bx_s) \mapsto t u(\bx_r,t;\bx_s)$. This
+observation motivates the definition of the AWI objective function
+(\cite{Warner:16}, equation XXXX)::
+\begin{equation}
+  \label{eqn:awidef}
+  J_{\rm AWI}[m;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu[\kappa;d](\bx_r,t;\bx_s)|^2}{\int dt |u[\kappa;d](\bx_r,t;\bx_s)|^2}\right) 
+\end{equation}
+in which $u[\kappa;d]$ is as defined above, that is, the solution of $S[\kappa]u[\kappa;d]
+= d$. \cite{Warner:16} also formulate
 AWI in terms of the inverse filter, that is, $u$ for which
-$\phi u*d = F[m]$, and in fact recommend that definition. The present
+$\phi u*d = F[\kappa]$, and in fact recommend that definition. The present
 discussion uses only the forward filter.
-
-If $\kappa = \kappa^*, \beta = \beta^*$, then the extended source
-$\bar{w}$ that fits the data should be $= w$,
-so the adaptive kernel $u[m;d]$ should be $=\delta(t)$ for all
-${\bf x}_s,{\bf x}_r$. Such a kernel $u[m;d]$ is in the null space of
-multiplication by $t$.
-This observation leads to the definition of the AWI objective function $J_{\rm AWI}$:
-\begin{equation}
-  \label{eqn:tJz}
-  J_{\rm AWI}[m;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu[m;d](\bx_r,t;\bx_s)|^2}{\int dt |u[m;d](\bx_r,t;\bx_s)|^2}\right)
-\end{equation}
-in which $u[m;d]$ is as defined above, that is, the solution of $S[m]u[m;d]
-= d$.
-
 Adaptive waveform inversion consists in iterative minimization of
@@
 
-The trace-by-trace normalization in definition \ref{eqn:tJz} of each
-trace in $tu[m;d]$ by the
-$L^2$ norm of the corresponding trace in $u[m;d]$ is quite important, as will be explained shortly.
+The same observation motivates the definition of the {\em Matched
+  Source Waveform Inversion} (MSWI) objective studied by
+\cite{HuangSymes2015SEG,HuangSymes:Geo17}:
+\begin{equation}
+  \label{eqn:mswidef}
+  J_{\rm MSWI}[m;d] = \sum_{\bx_s,\bx_r} \int dt |tu[\kappa;d](\bx_r,t;\bx_s)|^2
+\end{equation}
+That is, the two objectives differ only by the trace-by-trace
+normalization featured in the definition of AWI (equation
+\ref{eqn:awidef}). Much of the analysis of $J_{\rm AWI}$ pertains to
+$J_{\rm MSWI}$, as will be explained below.
+The trace-by-trace normalization in the definition of $J_{\rm AWI}$
+will prove quite consequential.
 
@@
 
-It may not be obvious at first glance why the AWI approach to estimating
-$m$, described in the last section, should be any more effective than
-FWI. Under the smoothness and ray geometry assumptions made above (equation
-\ref{eqn:coeff}, conditions G1 - G3) hold, an asymptotic relation
+It may not be obvious at first glance why AWI (local optimization of
+$J_{\rm AWI}$) should be should be any more effective at estimating
+$\kappa$ from data $d$ than is FWI (local
+optimization of $J_{\rm FWI}$). Under the smoothness and ray geometry assumptions made above (equation
+\ref{eqn:coeff}, conditions G1 - G4), an asymptotic relation
 between AWI and travel time tomography emerges, justifying the notion
-that AWI is more likely to succeed than is FWI. This relation is based
-on the Hadamard decomposition, and therefore is {\em
-  local}: it applies when sources and receivers are sufficiently
-close to lie in a common normal neighborhood of the source locations,
-for all coefficient pairs $(\kappa,\rho)$ in the search set $M$ of the
-inverse problem. The relation to travel time tomography is asymptotic
+that AWI is more likely to succeed than is FWI.
+
+\subsection{TIhonov Regularization}
+Before establishing the relation with travel time tomography, a pair of defects in the definition
+of $J_{\rm AWI}$ must be remedied. First, the objective function is
+not well-defined for the ``ideal'' case, in which $d = F[\kappa]$ and
+$u[\kappa;d]=\delta$: this adaptive filter is not square-integrable, so the
+denominator does not make sense. Second, the equation $S[\kappa]u=d$ does
+not (in general) have a solution $u \in L^2(X_{sr} \times \bR)$. Since $S[\kappa]$ is a convolution operator, solving
+$S[\kappa]u=d$ for $u$ amounts to deconovolving $d$ by $F[\kappa]$..
+Under the conditions W1-W3 on the wavelet $w$, $F[\kappa]$ does
+not have a continuous (stable) convolution inverse, and
+$S[\kappa]$ is not surjective.
+
+In fact, these two defects are really symptoms of the instability of
+deconvolution for kernels with the properties of $F[\kappa]$.  The
+conventional remedy for this instability is so-called ``pre-whitening'', as
+mentioned by \cite{Warner:16} ``Pre-whitening'' is a version of
+Tihonov regularization. It  amounts to replacement of the equation
+$S[\kappa]u=d$ by the regularized least squares problem: 
+\begin{equation}
+  \label{eqn:reg}
+  u_{0,\sigma}[\kappa;d] = \mbox{argmin}_{u \in L^2(X_{sr} \times \bR)} (\|S[\kappa]u-d\|^2 + \sigma\|u\|^2)
+\end{equation}
+For $\sigma > 0$, the Lax-Milgram Theorem (\cite{Yosida}, section
+III.7) implies that this least-squares problem has a unique solution
+$u_{0,\sigma}[\kappa;d] \in L^2(X_{sr} \times \bR)$. Define the modified AWI objective $J_{\lambda}$ by
+replacing $u$ in the definition \ref{eqn:awidef} with $u_{0,\sigma}$:
+\begin{equation}
+  \label{eqn:awidef}
+  J_{\sigma}[\kappa;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu_{0,\sigma}[\kappa;d](\bx_r,t;\bx_s)|^2}{\int dt |u_{0,\sigma}[\kappa;d](\bx_r,t;\bx_s)|^2}\right). 
+\end{equation}
+$J_{\sigma}$ is well-defined on $M \times L^2(X_{sr} \times \bR)$.
+
+A natural choice for the regularization weight $\sigma$ will emerge
+from the calculations to follow.
+
+%The second defect is insufficient constrain on the recording time
+%$t_d$: so far, $t_d$ can be arbitrarily small, and the recorded data
+%can contain arbitrarily little information about the material through%
+%which it propagates. The Hadamard decomposition motivates additional
+%constraints on $t_d$ which (in some cases) imply that the recorded
+%data embody strong constraints on the material model $m$.
+%These conditions on $t_d$  are posed in terms of the extreme
+%traveltimes over $m \in M$. Since all derivatives of the log wave velocity $\log c =
+%(\log \kappa - \log \rho )/2$ are bounded in the closure of $\Omega$,
+%uniformly over $m \in M$, 
+%the ray trace construction of travel time implies that
+%$0 < \tau_{\rm min} < \tau_{\rm max}$ exist for which $\tau_{\rm min}
+%\le \tau[\kappa](\bx_s,\bx_r) \le \tau_{\rm max}$ for $(\bx_s,\bx_r) \in
+%X_{sr}$, $m \in M$.
+%The additional constraints on $t_d$ and $\phi$ are:
+%begin{itemize}
+%\item[T1. ] $t_d >  \tau_{\rm max} + t_w + t_u $ (recording time is
+% sufficient for first arrival generated from any source with
+%waveform $u*w$ to be recorded at all receivers);
+%\item[T2. ] $0 < \tau_{\rm min} - t_w - t_u$ (signal from earliest
+%  part of wavelet $u*w$ arrives in positive time, for any
+%  source-receiver pair);
+%\item[T3. ] $\phi \equiv 1$ on $[\tau_{\rm min} - t_w - t_u,
+%\tau_{\rm max} + t_w + t_u]$ (time cutoff $\phi$ does not modify
+%leading term in Hadamard decomposition \ref{eqn:fwdgreen}, for any
+%source-receiver pair).
+%\end{itemize}
+%These assumptions are physically natural, as
+%they imply that the recording time interval $[0,t_d]$ contains the
+%support of the extended source data, up to an error vanishing with
+%source wavelength.
+
+%The choice of $\lambda$ as the  the regularization weight might seem
+%natural, as $\lambda$ is more-or-less the shortest wavelength
+%controlled by solution of the least-squares deconvolution problem
+%\ref{eqn:reg}. It will be rigorously justified {\em post-hoc} by the
+%calculations to follow. The same conclusions can be reached if
+%$\sigma$ in the definition \ref{eqn:reg}  is replaced by a function $\sigma(\lambda)$ with
+%the constraint that $C_{\rm min} \lambda \le \sigma(\lambda)
+%\le C_{\rm max}\lambda$, for suitable $C_{\rm min}, C_{\rm max}$ with
+%$0 < C_{\rm min} \le C_{\rm max}$. For simplicity, I will use
+%$\sigma(\lambda) = \lambda$.
+
+\subsection{Asymptotics: the leading term}
+The relation to travel time tomography is asymptotic
 in wavelength/frequency, and conveniently expressed via the wavelet
 family $\{\wl\}$ and the associated family of modeling operators
-$\Fl$. With these assumptions, the AWI objective function $J_{\rm AWI}[m;d_{\lambda}$
+$\Fl$.
+
+With these assumptions, I will show that the AWI objective function $J_{\rm AWI}[\kappa;d_{\lambda}]$
 for noise-free data
@@
   \label{eqn:noisefree}
-  d_{\lambda}=\Fl[m^*]
+  d_{\lambda}=\Fl[\kappa^*]
 \end{equation}
@@
 
-This relation breaks down if the sources and receivers fail to belong
-to a normal neighborhood (that is, if multiple ray paths exist between
-sources and receivers), as explained in the fifth section of this paper.
-
-With these assumptions,
-the decomposition \ref{eqn:fwdgreen} is applicable for solution of the point
+The decomposition \ref{eqn:fwdgreen} is applicable for solution of the point
 radiator problem \ref{eqn:awe} with extended source
@@
   \label{eqn:sdeflam}
-  S_{\lambda}[m]u(\bx_r,t;\bx_s) \equiv \phi(t)\bar{p}_{\lambda}(\bx_r,t;\bx_s,\bx_r).
+  S_{\lambda}[\kappa]u(\bx_r,t;\bx_s) \equiv \phi(t)\bar{p}_{\lambda}(\bx_r,t;\bx_s,\bx_r).
 \end{equation}
@@
   \label{eqn:slamapprox}
-  S_{\lambda}[m]u({\bf x}_r,t;{\bf x}_s) = \phi(t)a[m]({\bf x}_s,{\bf
-    x}_r) (u(\bx_r,\cdot;\bx_s)*\wl)(t-\tau[m]({\bf x}_s,{\bf x}_r)) + O(\lambda).
-\end{equation}
-
-I promised additional constraints on the data time window $[0,t_d]$
-and cutoff function $\phi$. These are posed in terms of the extreme
-traveltimes over $m \in M$. Since all derivatives of the log wave velocity $\log c =
-(\log \kappa - \log \rho )/2$ are bounded in the closure of $\Omega$,
-uniformly over $m \in M$, 
-the ray trace construction of travel time implies that
-$0 < \tau_{\rm min} < \tau_{\rm max}$ exist for which $\tau_{\rm min}
-\le \tau[m](\bx_s,\bx_r) \le \tau_{\rm max}$ for $(\bx_s,\bx_r) \in
-X_{sr}$, $m \in M$.
-The additional constraints on $t_d$ and $\phi$,
-\begin{itemize}
-\item[T1. ] $t_d >  \tau_{\rm max} + t_w + t_u $,
-\item[T2. ] $0 < \tau_{\rm min} - t_w - t_u$,
-\item[T3. ] $\phi \equiv 1$ on $[\tau_{\rm min} - t_w - t_u,
-\tau_{\rm max} + t_w + t_u]$.
-\end{itemize}
-That is, the support of the leading term in \ref{eqn:slamapprox} is
-contained in $\{t: \phi(t) = 1\}$. These assumptions are physically natural, as
-they imply that the recording time interval $[0,t_d]$ contains the
-support of the extended source data, up to an error vanishing as
-$\lambda \rightarrow 0$.
+  S_{\lambda}[\kappa]u({\bf x}_r,t;{\bf x}_s) = \phi(t)a[\kappa]({\bf x}_s,{\bf
+    x}_r) (u(\bx_r,\cdot;\bx_s)*\wl)(t-\tau[\kappa]({\bf x}_s,{\bf x}_r)) + O(\lambda).
+\end{equation}
 
@@
   \label{eqn:slambis}
-  S_{\lambda}[m]u({\bf x}_r,t;{\bf x}_s) = a[m]({\bf x}_s,{\bf
-    x}_r) (u(\bx_r,\cdot;\bx_s)*\wl)(t-\tau[m]({\bf x}_s,{\bf x}_r)) + O(\lambda).
-\end{equation}
-
+  S_{\lambda}[\kappa]u({\bf x}_r,t;{\bf x}_s) = a[\kappa]({\bf x}_s,{\bf
+    x}_r) (u(\bx_r,\cdot;\bx_s)*\wl)(t-\tau[\kappa]({\bf x}_s,{\bf x}_r)) + O(\lambda).
+\end{equation}
 Similarly, from the definition \ref{eqn:noisefree} and equation
@@
   \label{eqn:noisefreeapprox}
-  d_{\lambda}(\bx_r,t;\bx_s) = a[m^*](\bx_s,\bx_r) \wl(t-\tau[m^*](\bx_s,\bx_r)) + O(\lambda).
-  \end{equation}
-
-Inspection of equations \ref{eqn:slamapprox} and
-\ref{eqn:noisefreeapprox} shows that an approximate solution of $S_{\lambda}[m]u=d_{\lambda}$ is 
-\begin{equation}
-  \label{eqn:notl2}
-  u({\bf x}_r,t;{\bf x}_s) = \frac{a[m^*]({\bf x}_s,{\bf
-      x}_r)}{a[m]({\bf x}_s,{\bf x}_r)}\delta(t-\tau[m^*]({\bf
-    x}_s,{\bf x}_r)+\tau[m]({\bf x}_s,{\bf x}_r)) + O(\lambda).
-\end{equation}
-There is immediately a big problem: this $u$ is not
-square-integrable, and the AWI objective $J_{\rm AWI}$ is
-undefined. It's not surprising that mathematical ill-behaviour arises
-in this context. Since $S_{\lambda}[m]$ is a convolution operator, solving
-$S_{\lambda}[m]u=d$ for $u$ amounts to deconovolving $d$ by the kernel $F[m]$ of
-$S_{\lambda}[m]$. Under the assumed conditions on the wavelet $\wl$, $S_{\lambda}[m]$ does
-not have a continuous (stable) convolution inverse.
-
-It is conventional to stabilize deconvolution problems such as the
-equation $S_{\lambda}[m]u=d$ via so-called ``pre-whitening'', as
-mentioned by \cite{Warner:16} ``Pre-whitening'' is a version of
-Tihonov regularization. It  amounts to replacement of the equation
-$S_{\lambda}[m]u=d$ by the regularized least squares problem: 
-\begin{equation}
-  \label{eqn:reg}
-  u_0[m_{\lambda};d] = \mbox{argmin}_u (\|S_{\lambda}[m]u-d\|^2 + \lambda\|u\|^2)
-\end{equation}
-
-The choice of $\lambda$ as the  the regularization weight might seem
-natural, as $\lambda$ is more-or-less the shortest wavelength
-controlled by solution of the least-squares deconvolution problem
-\ref{eqn:reg}. It will be rigorously justified {\em post-hoc} by the
-calculations to follow.
+  d_{\lambda}(\bx_r,t;\bx_s) = a[\kappa^*](\bx_s,\bx_r) \wl(t-\tau[\kappa^*](\bx_s,\bx_r)) + O(\lambda).
+\end{equation}
 
@@
 temporary abbreviations
-$a^*=a[m^*], a=a[m], \tau^*=\tau[m^*], \tau=\tau[m]$.
+$a^*=a[\kappa^*], a=a[\kappa], \tau^*=\tau[\kappa^*], \tau=\tau[\kappa]$.
 
@@
 $$
-\approx \mbox{argmin}_u (\|a[m](\wl*u)(\cdot-\tau[m])-a[m^*]\wl(\cdot-\tau[m^*])\|^2 + \lambda\|u\|^2)
+\approx \mbox{argmin}_u (\|a[\kappa](\wl*u)(\cdot-\tau[\kappa])-a[\kappa^*]\wl(\cdot-\tau[\kappa^*])\|^2 + \lambda\|u\|^2)
 $$
-Re-write the approximation \ref{eqn:reg} in terms of the Fourier transforms $\hat{u_0}[m_{\lambda};d],\fwl$ of $u_0[m_{\lambda};d]$ and $\wl$:
+Re-write the approximation \ref{eqn:reg} in terms of the Fourier transforms $\hat{u_0}[\kappa_{\lambda};d],\fwl$ of $u_0[\kappa_{\lambda};d]$ and $\wl$:
 \begin{equation}
   \label{eqn:freqreg}
-\hat{u}_{0,\sigma}[m;d] \approx \mbox{argmin}_{\hat{u}}\frac{1}{2}\int d\omega \left(|a\hat{w}\hat{u}e^{i\omega \tau}-a^*\hat{w}e^{i\omega\tau^*}|^2 + \lambda |\hat{u}|^2\right)
+\hat{u}_{0,\sigma}[\kappa;d] \approx \mbox{argmin}_{\hat{u}}\frac{1}{2}\int d\omega \left(|a\hat{w}\hat{u}e^{i\omega \tau}-a^*\hat{w}e^{i\omega\tau^*}|^2 + \lambda |\hat{u}|^2\right)
 \end{equation}
@@
   \label{eqn:hatuf}
-\hat{u}_0[m_{\lambda};d] \approx \frac{a^*}{a}\hat{g}_{\lambda,\frac{\lambda}{a^2}} e^{i\omega(\tau^*-\tau)}
+\hat{u}_0[\kappa_{\lambda};d] \approx \frac{a^*}{a}\hat{g}_{\lambda,\frac{\lambda}{a^2}} e^{i\omega(\tau^*-\tau)}
 \end{equation}
@@
 $$
-u_0[m_{\lambda};d](t) \approx \frac{a^*}{a}g_{\lambda,\frac{\lambda}{a^2}}(t-(\tau^*-\tau))
+u_0[\kappa_{\lambda};d](t) \approx \frac{a^*}{a}g_{\lambda,\frac{\lambda}{a^2}}(t-(\tau^*-\tau))
 $$
 to a multiple of a shifted $\delta$, in fact exactly the distribution
-solution of $\bar{S}[m_{\lambda}]u=d$ given in equation \ref{eqn:notl2}. However for $\lambda>0$, $u_0[m_{\lambda};d]$ is square integrable.
+solution of $\bar{S}[\kappa_{\lambda}]u=d$ given in equation \ref{eqn:notl2}. However for $\lambda>0$, $u_0[\kappa_{\lambda};d]$ is square integrable.
 
@@
 \[
-J_{\rm MSWI}[m_{\lambda};d] = \sum_{\bx_s,\bx_r}\int dt |tu_0[m_{\lambda};d]|^2 \approx \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{\lambda}{a^2}}(t-(\tau^*-\tau))|^2
+J_{\rm MSWI}[\kappa_{\lambda};d] = \sum_{\bx_s,\bx_r}\int dt |tu_0[\kappa_{\lambda};d]|^2 \approx \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{\lambda}{a^2}}(t-(\tau^*-\tau))|^2
 \]
@@
   \label{eqn:MSWIasym}
-  J_{\rm MSWI}[m;d] \approx
+  J_{\rm MSWI}[\kappa;d] \approx
 \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{\lambda}{a^2}}(t)|^2
@@
 \begin{equation}
-  J_{\rm MSWI}[m_{\lambda};d] \approx
+  J_{\rm MSWI}[\kappa_{\lambda};d] \approx
   \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{\lambda}{a^2}}\|^2\left(l(g_{\lambda,\frac{\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
@@
 \ref{eqn:MSWI} of $J_{\rm MSWI}$, ones sees
-that the two functions differ only by the factor $\|u_0[m_{\lambda};d]\|^{-2}$ in
+that the two functions differ only by the factor $\|u_0[\kappa_{\lambda};d]\|^{-2}$ in
 the former, applied trace-by-trace. Using the geometric asymptotics approximation, 
 $$
-\|u_0[m_{\lambda};d]\|^2 \approx \frac{(a^*)^2}{a^2}\|g_{\frac{\lambda}{a^2}}\|^2 + O(\lambda)
+\|u_0[\kappa_{\lambda};d]\|^2 \approx \frac{(a^*)^2}{a^2}\|g_{\frac{\lambda}{a^2}}\|^2 + O(\lambda)
 $$
-Therefore, multiplication of the MSWI objective by $\|u_0[m_{\lambda};d]\|^{-2}$
+Therefore, multiplication of the MSWI objective by $\|u_0[\kappa_{\lambda};d]\|^{-2}$
 precisely removes the amplitude factor and norm-squared of $g$ in the right-hand side of the
@@
   \label{eqn:preawitt}
-J_{\rm AWI}[m;d] \approx \sum_{\bx_s,\bx_r} (\left(l(g_{\lambda,\frac{\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
+J_{\rm AWI}[\kappa;d] \approx \sum_{\bx_s,\bx_r} (\left(l(g_{\lambda,\frac{\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
  \end{equation}
@@
 in which $W(m,w_1)$ is bounded by the variation of the amplitude
-$a[m](\bx_r,\bx_s)$, and the choice of ``mother wavelet'' $w_1$, but
+$a[\kappa](\bx_r,\bx_s)$, and the choice of ``mother wavelet'' $w_1$, but
 independent of $\lambda$.
@@
   \label{eqn:awitt}
-J_{\rm AWI}[m;d] \approx \sum_{{\bf x}_s,{\bf x}_r} (\tau[m^*]({\bf
-  x}_s,{\bf x}_r)-\tau[m]({\bf x}_s,{\bf x}_r))^2 + O(\lambda^2)
+J_{\rm AWI}[\kappa;d] \approx \sum_{{\bf x}_s,{\bf x}_r} (\tau[\kappa^*]({\bf
+  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 + O(\lambda^2)
 \end{equation}
@@
 following choices, for each $m$ in the domain of $F$: $S$ is the regularized
-modeling operator $(S_{\lambda}[m],\sigma I)^T$.  Its domain is $U=L^2(\bR)$, of
+modeling operator $(S_{\lambda}[\kappa],\sigma I)^T$.  Its domain is $U=L^2(\bR)$, of
 which $u$ (the adaptive filter) is a member, and 
@@
 support of the filter (necessarily bounded in practical calculation)
-to have no effect. Then $\Ja$ becomes $J_{{\rm MSWI},\alpha,\sigma}[m,u;d]$, where
+to have no effect. Then $\Ja$ becomes $J_{{\rm MSWI},\alpha,\sigma}[\kappa,u;d]$, where
 \begin{equation}
   \label{eqn:eq5}
-   J_{{\rm MSWI},\alpha,\sigma}[m,u;d] = \frac{1}{2}\left(\|S_{\lambda}[m]u-d\|^2 +
+   J_{{\rm MSWI},\alpha,\sigma}[\kappa,u;d] = \frac{1}{2}\left(\|S_{\lambda}[\kappa]u-d\|^2 +
    \lambda\|u\|^2 + \alpha^2\|Tu\|^2\right)
  \end{equation}
-The minimizer $\ua$ over $u$ of $ J_{{\rm MSWI}, \alpha,\sigma}[m,u;d]$ is
-$u_{{\rm MSWI},\alpha,\sigma}[m;d]$, and $\tJa$ becomes
+The minimizer $\ua$ over $u$ of $ J_{{\rm MSWI}, \alpha,\sigma}[\kappa,u;d]$ is
+$u_{{\rm MSWI},\alpha,\sigma}[\kappa;d]$, and $\tJa$ becomes
 \begin{equation}
   \label{eqn:eq6}
-  \tilde{J}_{{\rm MSWI},\alpha,\sigma}[m;d] =
-  J_{{\rm MSWI},\alpha,\sigma}[m,u_{{\rm MSWI},\alpha,\sigma}[m;d];d].
-\end{equation}
-Note that $u_0$ becomes $u_0[m_{\lambda};d]$, and $\tilde{J}_0$ becomes
-$\tilde{J}_{0,\sigma}[m;d]$ which means exactly what
+  \tilde{J}_{{\rm MSWI},\alpha,\sigma}[\kappa;d] =
+  J_{{\rm MSWI},\alpha,\sigma}[\kappa,u_{{\rm MSWI},\alpha,\sigma}[\kappa;d];d].
+\end{equation}
+Note that $u_0$ becomes $u_0[\kappa_{\lambda};d]$, and $\tilde{J}_0$ becomes
+$\tilde{J}_{0,\sigma}[\kappa;d]$ which means exactly what
 it did in the previous section. The subscript ${\rm MSWI}$ is left
@@
   \lim_{\alpha \rightarrow 0}
-  \frac{1}{\alpha^2}\left(\tilde{J}_{{\rm MSWI},\alpha,\sigma}[m;d]-\tilde{J}_{0,\sigma}[m;d]\right)=
-  J_{\rm MSWI}[m;d]
+  \frac{1}{\alpha^2}\left(\tilde{J}_{{\rm MSWI},\alpha,\sigma}[\kappa;d]-\tilde{J}_{0,\sigma}[\kappa;d]\right)=
+  J_{\rm MSWI}[\kappa;d]
 \end{equation}
@@
   \label{eqn:eq8}
-  T[m;d] u(\bx_r,t;\bx_s) = \frac{\max(|t|, t_d)
-    \mbox{sgn}(t)}{\|u_0[m_{\lambda};d](\bx_r,\cdot;\bx_s)\|}
+  T[\kappa;d] u(\bx_r,t;\bx_s) = \frac{\max(|t|, t_d)
+    \mbox{sgn}(t)}{\|u_0[\kappa_{\lambda};d](\bx_r,\cdot;\bx_s)\|}
   u(\bx_r,t;\bx_s).
@@
   \label{eqn:eq5a}
-   J_{{\rm AWI},\alpha,\sigma}[m,u;d] = \frac{1}{2}\left(\|S_{\lambda}[m]u-d\|^2 +
-   \lambda\|u\|^2 + \alpha^2\|T[m;d]u\|^2\right)
- \end{equation}
-The minimizer $\ua$ over $u$ of $ J_{{\rm AWI}, \alpha,\sigma}[m,u;d]$ is
-$u_{{\rm AWI},\alpha,\sigma}[m;d]$, and $\tJa$ becomes
+   J_{{\rm AWI},\alpha,\sigma}[\kappa,u;d] = \frac{1}{2}\left(\|S_{\lambda}[\kappa]u-d\|^2 +
+   \lambda\|u\|^2 + \alpha^2\|T[\kappa;d]u\|^2\right)
+ \end{equation}
+The minimizer $\ua$ over $u$ of $ J_{{\rm AWI}, \alpha,\sigma}[\kappa,u;d]$ is
+$u_{{\rm AWI},\alpha,\sigma}[\kappa;d]$, and $\tJa$ becomes
 \begin{equation}
   \label{eqn:eq6a}
-  \tilde{J}_{{\rm AWI},\alpha,\sigma}[m;d] =
-  J_{{\rm AWI};\alpha,\sigma}[m,u_{{\rm AWI};\alpha,\sigma}[m;d];d].
+  \tilde{J}_{{\rm AWI},\alpha,\sigma}[\kappa;d] =
+  J_{{\rm AWI};\alpha,\sigma}[\kappa,u_{{\rm AWI};\alpha,\sigma}[\kappa;d];d].
 \end{equation}
@@
   \lim_{\alpha \rightarrow 0}
-  \frac{1}{\alpha^2}\left(\tilde{J}_{{\rm AWI},\alpha,\sigma}[m;d]-\tilde{J}_{0,\sigma}[m;d]\right)=
-  J_{\rm AWI}[m;d]
+  \frac{1}{\alpha^2}\left(\tilde{J}_{{\rm AWI},\alpha,\sigma}[\kappa;d]-\tilde{J}_{0,\sigma}[\kappa;d]\right)=
+  J_{\rm AWI}[\kappa;d]
 \end{equation}
@@
 geometric asymptotics as in equation \ref{eqn:hadamard}, and that the
-data is noise-free ($d=F[m^*]$). Together with the properties of $w$
+data is noise-free ($d=F[\kappa^*]$). Together with the properties of $w$
 just listed, these assumptions imply that
@@
   \label{eqn:eq11}
-  J_{0,\sigma}[m;d] = O(\sigma) + O(\lambda(w)).
+  J_{0,\sigma}[\kappa;d] = O(\sigma) + O(\lambda(w)).
 \end{equation}
@@
 \[
-  \tilde{J}_{0,\sigma}[m;d] = \frac{1}{2}
-  (\|S_{\lambda}[m]u_0[m_{\lambda};d]-d\|^2 + \lambda \|u_0[m_{\lambda};d]\|^2.
+  \tilde{J}_{0,\sigma}[\kappa;d] = \frac{1}{2}
+  (\|S_{\lambda}[\kappa]u_0[\kappa_{\lambda};d]-d\|^2 + \lambda \|u_0[\kappa_{\lambda};d]\|^2.
 \]
@@
 \[
-\|S_{\lambda}[m]u_0[m_{\lambda};d]-d\|^2 
+\|S_{\lambda}[\kappa]u_0[\kappa_{\lambda};d]-d\|^2 
   \approx \int d\omega
-  |a\hat{w}\hat{u_0[m_{\lambda};d]}e^{i\omega\tau}-a^*\hat{w}e^{i\omega\tau^*}|^2
+  |a\hat{w}\hat{u_0[\kappa_{\lambda};d]}e^{i\omega\tau}-a^*\hat{w}e^{i\omega\tau^*}|^2
 \]
@@
 \[
-  \|S_{\lambda}[m]u_0[m_{\lambda};d] - d\|^2 = O(\sigma) + O(\lambda(w)).
+  \|S_{\lambda}[\kappa]u_0[\kappa_{\lambda};d] - d\|^2 = O(\sigma) + O(\lambda(w)).
 \]
@@
 \[
-  \lambda\|u_0[m_{\lambda};d]\|^2 = O(\sigma) + O(\lambda(w)).
+  \lambda\|u_0[\kappa_{\lambda};d]\|^2 = O(\sigma) + O(\lambda(w)).
 \]
@@
 adaptive filter or extended source to data (represented by the opertor
-$S_{\lambda}[m]$ in the discussion above) may not have a well-behaved inverse in
+$S_{\lambda}[\kappa]$ in the discussion above) may not have a well-behaved inverse in
 the presence of multiple wavefront arrivals. \cite{Yongetal:GJI23}
@@
   \label{eqn:multi}
-  F[m,w](t) \approx \sum_{i \ge 0} a_i H^{p_i}w(t-\tau_i).
+  F[\kappa,w](t) \approx \sum_{i \ge 0} a_i H^{p_i}w(t-\tau_i).
 \end{equation}
@@
   \label{eqn:unwrap}
-  (\delta +T_1) * F[m,w] \approx  a_0w(t-\tau_0)
+  (\delta +T_1) * F[\kappa,w] \approx  a_0w(t-\tau_0)
 \end{equation}
@@
 
-Once again, it is possible to write the adaptive filter u for which $S_{\lambda}[m]u = F[m]*u = d =
-F[m^*]$ by inspection. Use the notations $a^*_i, t^*_i, p^*_i$ for
-the analogous quantities computed with the ``true'' model $m^*$, and
+Once again, it is possible to write the adaptive filter u for which $S_{\lambda}[\kappa]u = F[\kappa]*u = d =
+F[\kappa^*]$ by inspection. Use the notations $a^*_i, t^*_i, p^*_i$ for
+the analogous quantities computed with the ``true'' model $\kappa^*$, and
 $T^*, T_1^*$ defined as above but with $a^*_i$ in place of $a_i$ and so
@@
 \[
-  d = F[m_*]= S_{\lambda}[m]u
+  d = F[\kappa_*]= S_{\lambda}[\kappa] u
 \]
@@
 $\tau_1-\tau_0^*$ do not tend to zero, though they do tend to the same
-limit as $m \rightarrow m^*$ (as do the amplitude quotients), and the
+limit as $m \rightarrow \kappa^*$ (as do the amplitude quotients), and the
 Hilbert transform powers $p_1, p_1^*$ are the same as soon as $m$ is
-sufficiently close to $m^*$.
+sufficiently close to $\kappa^*$.
 
@@
 to the total. They will eventually begin to cancel as $m$ approaches
-$m^*$, however this cancellation happens only when the times are
+$\kappa^*$, however this cancellation happens only when the times are
 within the width of the approximate 

### 2024-10-16 repro/papers/awi/paper.tex
close

@@
 Theorem 7.4). The Propagation of
-Singularities Theorem (for example, \cite{Tay:81}, Theorem VI.2.1)
-implies that if $f \in \cal{E}'(\bR)$ and $(\bx_s,\bx_r) \in X_{sr}$,
+Singularities Theorem (\cite{Tay:81}, Theorem VI.2.1) and
+the pull-back property of distributions \cite{Dui:95}, Proposition 1.3.3)
+imply that if $f \in \cal{E}'(\bR)$ and $(\bx_s,\bx_r) \in X_{sr}$,
 then the trace $p(\bx_r, \cdot;\bx_s)$ on $\{(\bx_r,t): t \in \bR\}$ is well-defined,
@@
 source-receiver position pairs $X_{sr}$, source wavelet
-$f$ subject to constraints F1 and F2,  density $\rho$, and data traces
-$d:  X_{sr} \rightarrow \cal{D}(\bR)$, find $\kappa$ (or equivalently
+$f \in {\cal E}'(\bR)$,  density $\rho$, and data traces
+$d:  X_{sr} \rightarrow \cal{D}(\bR)$, find a bulk modulus $\kappa$
+(or equivalently wave velocity $c=\sqrt{\kappa/\rho}$)
 so that $F[\kappa] \approx d$ in some sense. If
-$d$ and $F[\kappa]$ lie in a normed subspace of $\cal{D}(\bR)$ for each
+$d$ and $F[\kappa]$ lie in a normed subspace of ${\cal D}'(\bR)$ for each
 $(\bx_s,\bx_r) \in X_{sr}$, then the
@@
 \subsection{Hadamard's Decomposition}
-The Hadamard decomposition of the acoustic Green's function provides a
-basis for a norm-based formulation of FWI \cite[]{Friedlander:75,Qian:JCP24}. As presented by
+The Hadamard decomposition of the acoustic Green's function \cite[]{Friedlander:75,Qian:JCP24} provides a
+basis for a norm-based formulation of FWI. As presented by
 \cite{Friedlander:75}, for instance, this construction applies to
@@
 equation. Conversely, if $p$ is a solution of the second-order initial
-value problem\ref{eqn:awe2}, it follows from the Hadamard
-decomposition detailed below and condition F1
-that 
+value problem\ref{eqn:awe2}, then under suitable conditions (including
+those to be imposed below),
 \[
@@
 \]
-has an
-integrable singularity at $\bx=\bx_s$, hence 
-defines a distribution, and that $(p,\bv)$ solves the system \ref{eqn:awe}.
+defines a distribution, and $(p,\bv)$ solves the system \ref{eqn:awe}.
 
@@
 $\Omega \times \bR$,  and $\Omega\subset \bR^3$ 
-and $M \subset (C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3))^2$ are bounded open sets satisfying 
+and $M \subset C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)$ are bounded open sets satisfying 
 \begin{itemize}
-  \item[G1. ] for every $(\bx_s,\bx_r) \in X_{sr}$, $\bx_s, \bx_r \in \Omega$;
-  \item[G2. ] for every source point $\bx_s\in X_{s}$ and $\kappa \in M$, $\Omega$ is a
+  \item[G1. ] $X_{sr} \subset \Omega \times \Omega$;
+  \item[G2. ] for every source point $\bx_s\in X_{s}$ and $\log \kappa \in M$, $\Omega$ is a
     normal neighborhood of $\bx_s$.
   \item[G3. ] for every source point $\bx_s \in X_{s};$
-    $\kappa \in M$, ray ${\bf X}:\bR^+\rightarrow \bR^3$
+    $\log \kappa \in M$, ray ${\bf X}:\bR^+\rightarrow \bR^3$
     with ${\bf X}(0)=\bx_s$ and any $t>0$, either $\{{\bf X}(s): 0\le s \le t\} \subset \Omega$
@@
 leaves $\Omega$, it does not return to $\Omega$. Conditions G1 - G3
-hold if $\kappa$ is constant, and for small perturbations and some
+hold for any convex open set $\Omega$ if $\kappa$ is constant, and for small perturbations and some
 large perturbations of constant $\kappa$: these are open conditions in
@@
 
-The arguments to follow depend an additional property of solutions of
-the wave equation \ref{eqn:awe2}:
+The arguments to follow depend an additional assumption about solutions of
+the wave equation \ref{eqn:awe2}, restricted to $X_{sr}\times \bR$:
 \begin{itemize}
-\item[G4. ] For any $n > 0$, $f \in {\cal E}'(\bR)$, there exists $K, T$ and $\delta>0$ so
-  that for any 4-multiindex ${\bf m}$ of order $\le n$, $(\bx_s,\bx_r) \in
- X_{sr}, t \in \bR$,
-  and any $\kappa \in M$, the solution $p$ of \ref{eqn:awe2} is smooth
+\item[G4. ] For any $n > 0$, $f \in {\cal E}'(\bR)$, there exist $K >
+  0, T \in \bR,$ and $\delta>0$ so
+  that for any $k$ with $0 \le k \le n$, $(\bx_s,\bx_r) \in
+ X_{sr},$
+  and any $\log \kappa \in M$, the solution $p$ of \ref{eqn:awe2} is smooth
   for $t>T$ and satisfies
   \[
-    |D^{\bf m}p(\bx_r,t;\bx_s)| \le K e^{-\delta t}.
+    |\partial_t^k p(\bx_r,t;\bx_s)| \le K e^{-\delta t}, \, t > T.
   \]
@@
 studied topic. Properties which imply G4 are in turn implied by the
-{\em non-trapping} hypothesis on the model $\kappa \in M$: all rays of
+{\em non-trapping} hypothesis on the model $\log \kappa \in M$: all rays of
 geometric asymptotics escape to infinity (\cite{Hristova:09}, Theorem
@@
 and similarly for $a[\kappa],b[\kappa]$;
-\item[H2. ] there exists $A>0$ so that for any $\kappa \in M$
+\item[H2. ] there exists $A>0$ so that for any $\log \kappa \in M$
   and $(\bx_s,\bx_r) \in X_{sr}$,
@@
 \end{equation}
+Define $\delta_{\tau} = \delta(t-\tau)$ to be the shifted delta, that
+is, $\langle \delta_{\tau},\phi\rangle = \phi(\tau)$. A useful
+restatement of the definition \ref{eqn:green}, suppressing the
+space-time coordinates, is
+\begin{equation}
+  \label{eqn:regreen}
+  G[\kappa] = (a[\kappa]\delta +  (b[\kappa]H))*\delta_{\tau[\kappa]}.
+\end{equation}
+
 %Under conditions G1 - G3 above, this decomposition is the restriction of
@@
 \subsection{FWI via Least Squares}
+For $f \in {\cal E}'(\bR)$, the Hadamard decomposition 
+(equation \ref{eqn:regreen}) gives:
+\begin{equation}
+  \label{eqn:fwdgreen}
+  F[\kappa] = G[\kappa] * \rho \partial_t f  = \rho( a[\kappa]\delta
+  +  (b[\kappa]H)) *\delta_{\tau[\kappa]}*\partial_t f.
+\end{equation}
 Suppose that the source wavelet $f$ satisfies
 \begin{itemize}
-\item[F1. ] $f \in H^1_0(\bR)$, $\mbox{supp } \subset [-t_w,t_w]$;
+\item[F1. ] $f \in C^{\infty}_0(\bR)$;
 \item[F2. ] $f$ is odd, $f(t) = -f(-t)$.
 \end{itemize}
-Then the Hadamard decomposition 
-(equation \ref{eqn:green}) together with local
-energy decay (assumption G4) and Young's inequality  implies that
-\begin{equation}
-  \label{eqn:fwdgreen}
-  F[\kappa] = G[\kappa] * \rho \partial_t f \in L^p(X_{sr} \times \bR)
-\end{equation}
-for any $p \ge 1$, and that
-the corresponding estimates are uniform in $\kappa \in M$:
-there exists $C>0$ so that for $\kappa\in M$, $p \ge 1$,
+Then equation \ref{eqn:fwdgreen}, assumptions F1, F2, bounds
+\ref{eqn:best} and Young's inequality implies that $F[\kappa] \in
+L^p(X_{sr} \times \bR)$ for any $p \ge 1$. Moreover there exists
+$C>0$ so that for $\log \kappa\in M$, $p \ge 1$,
 \begin{equation}
   \label{eqn:fwdnorm}
-  \|F[\kappa]\|_{L^p(X_{sr} \times \bR)} \le C \left\|\frac{df}{dt}\right\|_{L^p(\bR)}.
-\end{equation}
-%\[
-%  \| \cdot \|_d = \| \cdot \|_{L^2(X_{sr} \times \bR)}
-%\]
-
+  \|F[\kappa]\|_{L^p(X_{sr} \times \bR)} \le C \left\|\partial_t f\right\|_{L^p(\bR)}.
+\end{equation}
 This observation justfies posing FWI as a least-squares problem: given $d \in
@@
 \end{equation}
-%in which we have introduced the abbreviation
-%\[
-%  \| \cdot \|_d = \| \cdot \|_{L^2(X_{sr} \times \bR)}
-%\]
+
 Practical versions of least-squares FWI include various types of regularization and
@@
 %collection and processing.
-
-\subsection{Source Characteristics and Asymptotics}
-
-Note that in the representation \ref{eqn:fwdgreen}, only the
-derivative $\partial_tf$ appears. For convenience, introduce
-\begin{equation}
-  \label{eqn:wavelet}
-  w = \rho \partial_t f
-\end{equation}
-The properties F1 and F2 assumed for $f$ are equivalent to these
-properties of $w$:
-\begin{itemize}
-\item[W1. ] $w \in L^2_0(\bR)$, $\mbox{supp }w \subset [-t_w,t_w]$;
-\item[W2. ] $w$ is even;
-\item[W3. ] $\int \, w = 0$.
-\end{itemize}
-I will also
-refer to $w$ as ``the wavelet'', and view $F$ as parametrized by $w$.
-
-Introduce asymptotics in this setting by choosing $w = \wl$
-from a family $\{\wl:\}$ indexed by the dimensionless parameter
-$0 <\lambda \le 1\}$, and defined by
-\begin{equation}
-  \label{eqn:wfam}
-  \wl(t) = \lambda^{-1/2}w_1(t/\lambda).
-\end{equation}
-The  ``mother wavelet'' $w_1$ is square-integrable of compact support
-even, and mean-zero. So that property W1 holds for the entire family, 
-\begin{equation}
-  \label{eqn:wdiam}
-  \mbox{supp }w_1 \subset [-t_w,t_w].
-\end{equation}
-All wavelets $\wl$ in this family
-inherit these properties from $w_1$, along with its dimensions
-(energy/time$^2$). It follows from these assumptions that the Fourier transform $\fwl$ is
-real-valued, even, of class $H^1(\bR)$. The corresponding source waveform $f_{\lambda}$ is
-given by
-\begin{equation}
-  \label{eqn:flambda}
-  f_{\lambda}(t) = \int_{-\infty}^t\wl =
-  \lambda^{1/2}f_1\left(\frac{t}{\lambda}\right).
-\end{equation}
-Note that for $p \ge 1$,
-\begin{equation}
-  \label{eqn:lambdanorms}
-  \|\wl\|_{L^p(\bR)} = \lambda ^{\frac{1}{p}-\frac{1}{2}}\|w_1\|_{L^p(\bR)},\,\|f_{\lambda}\|_{L^p(\bR)}
-  = \lambda ^{\frac{1}{p}+\frac{1}{2} \|f_1\|_{L^p(\bR)}}.
-\end{equation}
-
-%and its product with frequency $|\omega|$ is also
-%square-integrable.
-
-The parameter $\lambda$ is connected to several frequency-related
-characteristics of $\wl$. Suppose that $f \in L^2(\bR)$ is also square-integrable after
-multiplication by $t$ (true if  for instance $f$ has compact
-support). The {\em pulse width} (or {\em RMS width} ) of $f$ is
-\begin{equation}
-  \label{eqn:pw}
-  l(f) = \frac{\|tf\|}{\|f\|}.
-\end{equation}
-The pulse width of $\wl$ is proportional to $\lambda$:
-\begin{equation}
-  \label{eqn:wpw}
-  l(\wl) = \left(\frac{\int dt (t \wl(t))^2}{\int dt (\wl(t))^2}
-  \right)^{1/2} = \lambda l(w_1).
-\end{equation}
-The {\em RMS frequency} of $\wl$ is
-\begin{equation}
-  \label{eqn:freq}
-  k(\wl) = \left(\frac{\int d\omega (\omega \fwl(\omega))^2}{\int dt (w(t))^2} \right)^{1/2} = \lambda^{-1}k(w_1).
-\end{equation}
-Thus the {\em RMS wavelength} $1/k(\wl)$ is proportional to $\lambda$.
-
-The Heisenberg inequality (for example, \cite{Folland:07}, p. 255) bounds the RMS
-wavelength, the reciprocal of the RMS frequency, in terms of the pulse
-width:
-\begin{equation}
-  \label{eqn:heis}
-  l(\wl) k(\wl) = l(w_1)k(w_1) \ge \frac{1}{4\pi}.
-\end{equation}
-Equality is attained for Gaussian $w_1$, and approximated for many other
-commonly used wavelets (Ricker = second derivative of Gaussian,
-trapezoidal bandpass,...).
-
-The modeling operator with wavelet $\wl$ is denoted $\Fl$. From the
-decomposition \ref{eqn:fwdgreen}, $\Fl$ is the sum of two terms:
-\begin{equation}
-  \label{eqn:Flamdecomp}
-  \Fl = \Fla + \Flb.
-\end{equation}
-where
-\begin{equation}
-  \label{eqn:Fl0}
-  \Fla[\kappa](\bx_r,t;\bx_s) =
-  a[\kappa](\bx_s,\bx_r)\wl(t-\tau[\kappa](\bx_s,\bx_r))
-\end{equation}
-and
-\[
-\Flb[\kappa] = \rho_s\int
-   \,ds \, \frac{df_{\lambda}}{dt}(t-s) b[\kappa](\bx_s,\bx_r,s-\tau[\kappa](\bx_s,\bx_r))H(s-\tau[\kappa](\bx_s,\bx_r))
-\]
-\[
-  =\rho_s
-  f_{\lambda}(t-\tau(\bx_s,\bx_r))b[\kappa](\bx_s,\bx_r,0)
-\]
-\begin{equation}
-  \label{eqn:Fl1}
-    -
-    \rho_s\int
-   \,ds \, f_{\lambda}(t-s) (\partial_t
-   b[\kappa](\bx_s,\bx_r,s-\tau[\kappa](\bx_s,\bx_r)))H(s-\tau[\kappa](\bx_s,\bx_r))
-\end{equation}
-From hypothesis G4, the bound \ref{eqn:lambdanorms} , and Young's inequality, the $\|\Fl1[\kappa]\|_{ L^p(X_{sr} \times \bR)}$ is bounded by a multiple of
-$\lambda^{\frac{1}{p}+\frac{1}{2}}$, uniformly over $\kappa \in M$.
-I will write ``$+ \Ol$'' to mean an  error uniformly bounded in $L^2$ by
-a multiple of $\lambda$ over $m \in M$. Thus
-\begin{equation}
-  \label{eqn:fwdapprox}
-  \Fl[\kappa](\bx_s,\bx_r) = \Fla[\kappa] +   \Ol
-\end{equation}
 
@@
 Then $S[\kappa]$ is bounded on $L^2(X_{sr} \times \bR)$, uniformly in
-$\kappa \in M$, thanks to
+$\log \kappa \in M$, thanks to
 the bound \ref{eqn:fwdnorm} for $p=1$. $S$ also represents an
@@
 %  \label{eqn:sop}
-%  S[\kappa]u = \phi w* a[\kappa] T[\kappa] ((\delta + b[\kappa]H)*u)
+%  S[\kappa]u = w* a[\kappa] T[\kappa] ((\delta + b[\kappa]H)*u)
 %\end{equation}
@@
 \begin{equation}
-  \label{eqn:awidef}
+  \label{eqn:awideforig}
   J_{\rm AWI}[m;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu[\kappa;d](\bx_r,t;\bx_s)|^2}{\int dt |u[\kappa;d](\bx_r,t;\bx_s)|^2}\right) 
@@
 AWI in terms of the inverse filter, that is, $u$ for which
-$\phi u*d = F[\kappa]$, and in fact recommend that definition. The present
+$u*d = F[\kappa]$, and in fact recommend that definition. The present
 discussion uses only the forward filter.
 Adaptive waveform inversion consists in iterative minimization of
-$J_{\rm AWI}$: choose an initial estimate $m_0=(\kappa_0,\beta, w)$ of the
-model, then update $m$ (that is, $\kappa$) repeatedly using a gradient
+$J_{\rm AWI}$: choose an initial estimate $\kappa_0$ of the
+model, then update $\kappa$ repeatedly using a gradient
 descent method. \cite{Warner:16} describes the computation of $J_{\rm AWI}$
@@
 \begin{equation}
-  \label{eqn:mswidef}
+  \label{eqn:mswideforig}
   J_{\rm MSWI}[m;d] = \sum_{\bx_s,\bx_r} \int dt |tu[\kappa;d](\bx_r,t;\bx_s)|^2
@@
 normalization featured in the definition of AWI (equation
-\ref{eqn:awidef}). Much of the analysis of $J_{\rm AWI}$ pertains to
+\ref{eqn:awideforig}). Much of the analysis of $J_{\rm AWI}$ pertains to
 $J_{\rm MSWI}$, as will be explained below.
@@
 that AWI is more likely to succeed than is FWI.
-
-\subsection{TIhonov Regularization}
-Before establishing the relation with travel time tomography, a pair of defects in the definition
-of $J_{\rm AWI}$ must be remedied. First, the objective function is
-not well-defined for the ``ideal'' case, in which $d = F[\kappa]$ and
-$u[\kappa;d]=\delta$: this adaptive filter is not square-integrable, so the
-denominator does not make sense. Second, the equation $S[\kappa]u=d$ does
-not (in general) have a solution $u \in L^2(X_{sr} \times \bR)$. Since $S[\kappa]$ is a convolution operator, solving
-$S[\kappa]u=d$ for $u$ amounts to deconovolving $d$ by $F[\kappa]$..
-Under the conditions W1-W3 on the wavelet $w$, $F[\kappa]$ does
-not have a continuous (stable) convolution inverse, and
-$S[\kappa]$ is not surjective.
+%In addition all of the
+%previously mentioned assumptions, I will assume that the data $d$ is
+%{\em noise-free}, that is, $d = F[\kappa^*]$ for a target bulk modulus
+%$\kappa^* \in M$.
+
+\subsection{Tihonov Regularization}
+Before establishing the relation with travel time tomography, several
+defects in the definition \ref{eqn:awideforig} of $J_{\rm AWI}$ must be
+remedied. First, the objective function is not well-defined for the
+``ideal'' case, in which $d = F[\kappa]$ and $u[\kappa;d]=\delta$:
+this adaptive filter is not square-integrable, so the denominator does
+not make sense. Second, the equation $S[\kappa]u=d$ does not (in
+general) have a solution $u \in L^2(X_{sr} \times \bR)$. Since
+$S[\kappa]$ is a convolution operator, solving $S[\kappa]u=d$ for $u$
+amounts to deconovolving $d$ by $F[\kappa]$..  Under the conditions
+W1-W3 on the wavelet $w$, $F[\kappa]$ does not have a continuous
+(stable) convolution inverse, and $S[\kappa]$ is not surjective.
 
 In fact, these two defects are really symptoms of the instability of
-deconvolution for kernels with the properties of $F[\kappa]$.  The
-conventional remedy for this instability is so-called ``pre-whitening'', as
+deconvolution for kernels with the properties of $F[\kappa]$.  To
+begin with, $S[\kappa] = F[\kappa]*$ defines a bounded map on
+$L^2(X_{sr}\times \bR)$, since from \ref{eqn:fwdgreen} and the definition of
+  $S[\kappa]$,
+\[
+  S[\kappa]u = \rho( a[\kappa]\delta    +  (b[\kappa]H)) *\delta_{\tau[\kappa]}*\partial_t f. *u,
+\]
+The bounds \ref{eqn:best} yield $bH \in
+  L^1(X_{sr}\times \bR)$, and property F1, $\partial_t f \in
+  L^1(\bR)$.
+
+The conventional remedy for instability in a deconvolution problem of
+this type is so-called ``pre-whitening'', as
 mentioned by \cite{Warner:16} ``Pre-whitening'' is a version of
@@
   \label{eqn:reg}
-  u_{0,\sigma}[\kappa;d] = \mbox{argmin}_{u \in L^2(X_{sr} \times \bR)} (\|S[\kappa]u-d\|^2 + \sigma\|u\|^2)
+  u_{\sigma}[\kappa;d] = \mbox{argmin}_{u \in L^2(X_{sr} \times \bR)} (\|S[\kappa]u-d\|^2 + \sigma\|u\|^2)
 \end{equation}
@@
 III.7) implies that this least-squares problem has a unique solution
-$u_{0,\sigma}[\kappa;d] \in L^2(X_{sr} \times \bR)$. Define the modified AWI objective $J_{\lambda}$ by
-replacing $u$ in the definition \ref{eqn:awidef} with $u_{0,\sigma}$:
+$u_{\sigma}[\kappa;d] \in L^2(X_{sr} \times \bR)$. Define the modified
+AWI objective $J_{\rm AWI,\sigma}$ by
+replacing $u$ in the definition \ref{eqn:awideforig} with $u_{\sigma}$:
 \begin{equation}
   \label{eqn:awidef}
-  J_{\sigma}[\kappa;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu_{0,\sigma}[\kappa;d](\bx_r,t;\bx_s)|^2}{\int dt |u_{0,\sigma}[\kappa;d](\bx_r,t;\bx_s)|^2}\right). 
-\end{equation}
-$J_{\sigma}$ is well-defined on $M \times L^2(X_{sr} \times \bR)$.
-
-A natural choice for the regularization weight $\sigma$ will emerge
-from the calculations to follow.
+  J_{\rm AWI,\sigma}[\kappa;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu_{\sigma}[\kappa;d](\bx_r,t;\bx_s)|^2}{\int dt |u_{\sigma}[\kappa;d](\bx_r,t;\bx_s)|^2}\right). 
+\end{equation}
+To make this definition a bit more concise, introduce the operator $T$ on
+${\cal D}'(X_{sr} \times \bR)$ by
+\[
+  Tp(\bx_r,t;\bx_s) = tp(\bx_r,t;\bx_s).
+\]
+Then
+\begin{equation}
+  \label{eqn:awidefbis}
+  J_{\rm AWI,\sigma}[\kappa;d] = \sum_{(\bx_s,\bx_r) \in X_{sr}}
+  \frac{\|Tu_{\sigma}[\kappa;d](\bx_r,\cdot;\bx_s)\|^2}{\|u_{\sigma}[\kappa;d](\bx_r,\cdot;\bx_s)\|^2}.
+\end{equation}
+Here $\| \cdot \|$ is the norm in $L^2(\bR)$.
+
+Define a regularized version of the MSWI objective function $J_{\rm
+  MSWI}$ (definition \ref{eqn:mswideforig}) similarly:
+\begin{equation}
+  \label{eqn:mswidef}
+  J_{\rm MSWI,\sigma}[\kappa;d] = \|Tu_{\sigma}[\kappa;d]\|^2.
+\end{equation}
+
+This definition has introduced a new defect: while the denominator is
+well-defined for $\log \kappa \in M$, $d \times L^2(X_{sr} \times
+\bR)$ for $\sigma > 0$, the numerator may not be, since there is no
+guarantee that $Tu_{\sigma}[\kappa;d] \in L^2(X_{sr} \times
+\bR)$.
+
+\begin{prop}
+  \label{thm:tl2}
+  Suppose that $d, Td \in L^2(X_{sr} \times \bR)$. Then
+  $Tu_{\sigma}[\kappa;d]  \in L^2(X_{sr} \times \bR)$.
+\end{prop}
+
+\begin{proof}
+  Since $\kappa$ is the argument of $S$ and $u_{\sigma}$ throughout, drop it
+  temporarily, also from the ingredients $\tau, a, $ and $b$ of the
+  Hadamard decomposition.
+
+  Thus
+  \[
+    S^TS =
+    \rho^2(a\delta  +  (\check{b}\check{H})) *\delta_{-\tau}*\partial_t
+    f * (a\delta  +  (bH)) *\delta_{\tau}*\partial_t f*
+  \]
+  \[
+    = \rho^2((a\delta  +  (\check{b}\check{H}))*(a\delta  +
+    (bH))*\partial_t f *\partial_t f*
+  \]
+  \[
+    [T,S^TS] = \rho^2((T\check{b}\check{H}) *(a\delta  +
+    (bH))*+ (a\delta  +  (\check{b}\check{H}))*(TbH))*\partial_t f
+    *\partial_t f*
+  \]
+  \begin{equation}
+    \label{eqn:tsscomm}
+    +2\rho^2((a\delta  +  (\check{b}\check{H}))*(a\delta  +
+    (bH)))* \partial_t f  *T\partial_t f*.
+  \end{equation}
+  Referring to the exponential decay assumption \ref{eqn:best}
+  and properties F1, F2 of $f$, 
+  observe that all summands on the RHS of identity \ref{eqn:tsscomm} are in
+  $L^1(X_{sr} \times \bR)$, so $[T,S^TS]$ restricts to a bounded
+  operator on $L^2(X_{sr}\times \bR)$. A similar computation shows the
+  same property for $[T,S^T]$. Since $u_{\sigma} \in L^2(X_{sr}\times
+  \bR)$,
+  \[
+    T(S^TS+\sigma I)u_{\sigma} =  [T,(S^TS+\sigma I)]u_{\sigma} +
+    (S^TS+\sigma I)Tu_{\sigma}
+  \]
+  \[
+    = [T,S^T]d + S^TTd,
+  \]
+  that is,
+  \[
+    (S^TS+\sigma I)Tu_{\sigma} = e \in L^2(X_{sr} \times \bR).
+  \]
+  A theorem of von Neumann (\cite{Yosida}, VII.3, Theorem 2) implies
+  that $S^TS+\sigma I$ has a bounded inverse, so the distribution
+  $Tu_{\sigma} \in L^2(X_{sr} \times \bR)$.
+\end{proof}
+
+The following fact will be useful later on:
+
+\begin{prop}
+  \label{thm:noisefree}
+  Suppose that $\kappa \in M$ and $d = F[\kappa]$. Then $Td
+  \in L^2(X_{sr} \times \bR)$.
+\end{prop}
+
+\begin{proof}
+  A simple estimate shows that $F[\kappa]$ satisfies an exponential
+  decay estimate of the form G4, whence the conclusion follows.
+\end{proof}
+
+{\bf Remark:} It is conventional to call data in the range of the
+modeling operator ``noise-free'', and I will adopt this usage.
 
@@
 
-\subsection{Asymptotics: the leading term}
-The relation to travel time tomography is asymptotic
-in wavelength/frequency, and conveniently expressed via the wavelet
-family $\{\wl\}$ and the associated family of modeling operators
-$\Fl$.
-
-With these assumptions, I will show that the AWI objective function $J_{\rm AWI}[\kappa;d_{\lambda}]$
-for noise-free data
-\begin{equation}
-  \label{eqn:noisefree}
-  d_{\lambda}=\Fl[\kappa^*]
-\end{equation}
-is $\lambda$-asymptotic to the mean-square difference between the
-source-receiver travel times for $m$ and $m*$.
-
-The decomposition \ref{eqn:fwdgreen} is applicable for solution of the point
-radiator problem \ref{eqn:awe} with extended source
-$u(\bx_r,\cdot;\bx_s)*\wl$, for each source-receiver pair
-$(\bx_s,\bx_r)\in X_{sr}$. Write $\bar{p}_{\lambda}$ for the pressure
-component of the solution, and specialize the definition
-\ref{eqn:sdef} to
-\begin{equation}
+\subsection{Source Characteristics and Asymptotics}
+
+Note that in the representation \ref{eqn:fwdgreen}, only the
+derivative $\partial_tf$ appears. For convenience, introduce
+\begin{equation}
+  \label{eqn:wavelet}
+  w = \rho \partial_t f
+\end{equation}
+The properties F1 and F2 assumed for $f$ are equivalent to these
+properties of $w$:
+\begin{itemize}
+\item[W1. ] $w \in C^{\infty}_0(\bR)$;;
+\item[W2. ] $w$ is even;
+\item[W3. ] $\int \, w = 0$.
+\end{itemize}
+I will also
+refer to $w$ as ``the wavelet'', and view $F$ as parametrized by $w$.
+
+Introduce asymptotics in this setting by choosing $w = \wl$
+from a family $\{\wl:0 <\lambda \le 1\}$ indexed by the dimensionless parameter
+$\lambda$, and defined by
+\begin{equation}
+  \label{eqn:wfam}
+  \wl(t) = \lambda^{-1/2}w_1(t/\lambda).
+\end{equation}
+The  ``mother wavelet'' $w_1$ is assumed to have properties W1 - W3, and
+all wavelets $\wl$ in this family
+inherit these properties from $w_1$, along with its dimensions
+(energy/time$^2$). It follows from these assumptions that the Fourier transform $\fwl$ is
+real-valued, even, and real-analytic. The corresponding source waveform $f_{\lambda}$ is
+given by
+\begin{equation}
+  \label{eqn:flambda}
+  f_{\lambda}(t) =\frac{1}{\rho} \int_{-\infty}^t\wl =
+  \lambda^{1/2}f_1\left(\frac{t}{\lambda}\right).
+\end{equation}
+Note that for $p \ge 1$,
+\begin{equation}
+  \label{eqn:lambdanorms}
+  \|\wl\|_{L^p(\bR)} = \lambda ^{\frac{1}{p}-\frac{1}{2}}\|w_1\|_{L^p(\bR)},\,\|f_{\lambda}\|_{L^p(\bR)}
+  = \lambda ^{\frac{1}{p}+\frac{1}{2}} \|f_1\|_{L^p(\bR)}.
+\end{equation}
+
+%and its product with frequency $|\omega|$ is also
+%square-integrable.
+
+The parameter $\lambda$ is connected to several frequency-related
+characteristics of $\wl$. Suppose that $f \in L^2(\bR)$ is also square-integrable after
+multiplication by $t$ (application of $T$, in the notation of the last
+section - true if for instance $f$ has compact
+support). The {\em pulse width} (or {\em RMS width} ) of $f$ is
+\begin{equation}
+  \label{eqn:pw}
+  l(f) = \frac{\|Tf\|}{\|f\|}.
+\end{equation}
+The pulse width of $\wl$ is proportional to $\lambda$:
+\begin{equation}
+  \label{eqn:wpw}
+  l(\wl) = \left(\frac{\int dt (t \wl(t))^2}{\int dt (\wl(t))^2}
+  \right)^{1/2} = \lambda l(w_1).
+\end{equation}
+
+Note that the AWI objective defined in \ref{eqn:awideforig} and its
+modification \ref{eqn:awidef} are actually
+squared pulse widths of the adaptive filters $u[\kappa;d]$ and
+$u_{\sigma}[\kappa;d]$ respectively.
+
+The {\em RMS frequency} of $\wl$ is
+\begin{equation}
+  \label{eqn:freq}
+  k(\wl) = \left(\frac{\int d\omega (\omega \fwl(\omega))^2}{\int dt (w(t))^2} \right)^{1/2} = \lambda^{-1}k(w_1).
+\end{equation}
+Thus the {\em RMS wavelength} $1/k(\wl)$ is proportional to $\lambda$.
+
+The Heisenberg inequality (for example, \cite{Folland:07}, p. 255) bounds the RMS
+wavelength, the reciprocal of the RMS frequency, in terms of the pulse
+width:
+\begin{equation}
+  \label{eqn:heis}
+  l(\wl) k(\wl) = l(w_1)k(w_1) \ge \frac{1}{4\pi}.
+\end{equation}
+Equality is attained for Gaussian $w_1$, and approximated for many other
+commonly used wavelets (Ricker = second derivative of Gaussian,
+trapezoidal bandpass,...).
+
+The modeling operator with wavelet $\wl$ is denoted $\Fl$. From the
+decomposition \ref{eqn:fwdgreen}, $\Fl$ is the sum of two terms:
+\begin{equation}
+  \label{eqn:Flamdecomp}
+  \Fl = \Fla + \Flb.
+\end{equation}
+where
+\begin{equation}
+  \label{eqn:Fl0}
+  \Fla[\kappa]= a[\kappa]\delta_{\tau[\kappa]}*\wl
+\end{equation}
+and
+\[
+\Flb[\kappa] = (b[\kappa]H) *\delta_{\tau[\kappa]}*\wl.
+\]
+Write $b_0[\kappa](\bx_s,\bx_r) = b[\kappa](\bx_r,0;\bx_s)$. Then
+$\Flb$ can be rearranged as
+\begin{equation}
+  \label{eqn:Fl1}
+  =\rho (b_0[\kappa]\delta + (\partial_t b H) * f_{\lambda} )*\delta_{\tau[\kappa]}.
+\end{equation}
+From the bounds \ref{eqn:best} and \ref{eqn:lambdanorms} and Young's
+inequality, there exists $C \ge 0$ so that for all $\log \kappa \in M$, $p
+\ge 1$,
+\begin{equation}
+  \label{eqn:Frembd}
+ \|\Fla[\kappa]\|_{ L^p(X_{sr} \times \bR)} \le C\lambda^{\frac{1}{p}-\frac{1}{2}},\,\|\Flb[\kappa]\|_{ L^p(X_{sr} \times \bR)} \le C\lambda^{\frac{1}{p}+\frac{1}{2}}
+\end{equation}
+
+Similarly, define
+\begin{eqnarray}
   \label{eqn:sdeflam}
-  S_{\lambda}[\kappa]u(\bx_r,t;\bx_s) \equiv \phi(t)\bar{p}_{\lambda}(\bx_r,t;\bx_s,\bx_r).
-\end{equation}
-Following the reasoning leading to
-equation\ref{eqn:fwdapprox}, obtain
-\begin{equation}
-  \label{eqn:slamapprox}
-  S_{\lambda}[\kappa]u({\bf x}_r,t;{\bf x}_s) = \phi(t)a[\kappa]({\bf x}_s,{\bf
-    x}_r) (u(\bx_r,\cdot;\bx_s)*\wl)(t-\tau[\kappa]({\bf x}_s,{\bf x}_r)) + O(\lambda).
-\end{equation}
-
-From T3, equation \ref{eqn:slamapprox} can be re-written as
-\begin{equation}
-  \label{eqn:slambis}
-  S_{\lambda}[\kappa]u({\bf x}_r,t;{\bf x}_s) = a[\kappa]({\bf x}_s,{\bf
-    x}_r) (u(\bx_r,\cdot;\bx_s)*\wl)(t-\tau[\kappa]({\bf x}_s,{\bf x}_r)) + O(\lambda).
-\end{equation}
-Similarly, from the definition \ref{eqn:noisefree} and equation
-\ref{eqn:fwdgreen},
-\begin{equation}
-  \label{eqn:noisefreeapprox}
-  d_{\lambda}(\bx_r,t;\bx_s) = a[\kappa^*](\bx_s,\bx_r) \wl(t-\tau[\kappa^*](\bx_s,\bx_r)) + O(\lambda).
-\end{equation}
-
-Note that in these expressions, the source and receiver coordinates
-${\bf x}_s$ and ${\bf x}_r$ act as either passive parameters or
-summation indices. The right-hand side of equation \ref{eqn:reg} is
-summed over ${\bf x}_s$ and ${\bf x}_r$, for instance. However, the
-quantities on the right-hand side are not coupled: the minimization
-could be regarded as taking place independently for each choice of
-${\bf x}_s$ and ${\bf x}_r$. Therefore, to simplify the presentation,
-drop ${\bf x}_s$ and ${\bf x}_r$ from the notation temporarily: each
-calculation to follow pertains to a single trace, i.e. a single choice
-of ${\bf x}_s$ and ${\bf x}_r$.  Also, use the
-temporary abbreviations
+  \Sl[\kappa]u &=& u *\Fl[\kappa],  \nonumber \\
+  \Sla[\kappa]u &=& u *\Fla[\kappa], \nonumber\\
+  \Slb[\kappa]u &=& u *\Flb[\kappa], 
+\end{eqnarray}
+so that
+\[
+  \Sl[\kappa] = \Sla[\kappa] + \Slb[\kappa].
+\]
+From the bound \ref{eqn:Frembd} for $p=1$, there is $C \ge 0$ so that
+for all $\log \kappa \in M$, $u \in L^2(X_{sr} \times \bR)$,
+\begin{eqnarray}
+  \label{eqn:Srembd}
+  \|\Sl[\kappa]u\| &\le&  C\lambda^{\frac{1}{2}}\|u\|,\nonumber \\
+  \|\Slb[\kappa]u\| & \le& C\lambda^{\frac{3}{2}}\|u\|.
+\end{eqnarray}
+
+\subsection{Linking Regularization and Wavelength}
+In the remainder of this section, I will examine the $\lambda$-parametrized families of
+objective functions obtained by
+\begin{itemize}
+\item[1. ]replacing $S$ and $d$ in the definitions \ref{eqn:awidefbis}
+  and \ref{eqn:mswidef} with $\Sl$ and $\dl =\Fl[\kappa^*]$, that is,
+  noise-free data generated by a choice of $\kappa^* \in M$ and
+  $\lambda \in (0,1]$, and
+\item[2. ] choosing the Tihonov regularization weight
+  $\sigma = \lambda$.
+\end{itemize}
+That is,
+\begin{equation}
+  \label{eqn:mswideflam}
+  J_{\rm MSWI,\lambda}[\kappa;\dl] =  \|T\ul[\kappa;d]\|^2
+\end{equation}
+and 
+\begin{equation}
+  \label{eqn:awideflam}
+  J_{\rm AWI,\lambda}[\kappa;\dl] = \sum_{(\bx_s,\bx_r) \in X_{sr}} 
+  \frac{\|T\ul[\kappa;d](\bx_r,\cdot;\bx_s)\|^2}{\|\ul[\kappa;d](\bx_r,\cdot;\bx_s)\|^2}.
+\end{equation}
+In both definitions, $\ul$ solves the least-squares problem \ref{eqn:reg} with
+$\sigma = \lambda$.
+
+Justification for choosing the regularization weight $\sigma$ to be the same as
+the wavelength proxy $\lambda$ will emerge from the analysis relating these
+objective functions to the conventional objective function of travel
+time tomography. It will be clear that choosing $\sigma$ a function of
+$\lambda$ roughly proportional to $\lambda$, that is, $C_*\lambda \le
+\sigma(\lambda) \le C^*\lambda$ for $C_*,C^*$ satisfying $0 < C_* \le
+C^*$, would lead to the same asymptotic estimates for the relation
+between $J_{\rm MSWI,\lambda}$,$ J_{\rm AWI,\lambda} $ and travel time
+error. The argument is slightly cleaner if simply $\sigma=\lambda$.
+
+\subsection{The leading term and travel time}
+
+The relation between MSWI, AWI, and travel time emerges most clearly
+through a further simplification:
+
+\begin{prop} 
+In the definitions of $J_{\rm MSWI,\sigma}$ and $J_{\rm AWI,\sigma}$, replace the data $d$ with
+$\dla = \Fla[\kappa^*]$, $S[\kappa]$ with $\Sla[\kappa]$. That is,
+\begin{eqnarray}
+  \label{eqn:leading}
+  \dla & = & \Fla[\kappa^*] = a[\kappa^*]\delta_{\tau[\kappa^*]}*\wl
+             \nonumber \\
+  \Sla[\kappa]u &=& a[\kappa]\delta_{\tau[\kappa^*]}*\wl *u\nonumber \\
+  \uzl &=& \mbox{argmin}_{u \in L^2(X_{sr} \times \bR)} (\|\Sla[\kappa]u-\dla\|^2 + \sigma\|u\|^2)
+\end{eqnarray}
+and
+\begin{eqnarray}
+  \label{eqn:awileading}
+  J^0_{\rm MSWI,\lambda}[\kappa;\dla] & = &
+                                         \|T\uzl[\kappa;\dla]\|^2\nonumber\\
+  J^0_{\rm AWI,\lambda}[\kappa;\dla] & = &
+                                       \sum_{(\bx_s,\bx_r) \in X_{sr}} \frac{\|T\uzl[\kappa;\dla](\bx_r,\cdot;\bx_s)\|^2}{\|\uzl[\kappa;\dla](\bx_r,\cdot;\bx_s)\|^2}
+\end{eqnarray}                 
+Then there exists $C>0$ so that for
+$\kappa \in M$, $0 < \lambda \le 1$, 
+\begin{equation}
+  \label{eqn:mswitt}
+ | J^0_{\rm MSWI,\lambda}[\kappa;\dla] - 
+  \frac{1}{\lambda} \sum_{{\bf x}_s,{\bf x}_r} \|u^0_1[\kappa](\bx_r,\cdot;\bx_s)\|^2(\tau[\kappa^*]({\bf
+  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 | \le C\lambda,
+\end{equation}
+and
+\begin{equation}
+  \label{eqn:awitt}
+|J^0_{\rm AWI,\lambda}[\kappa;\dla] - \sum_{{\bf x}_s,{\bf x}_r} (\tau[\kappa^*]({\bf
+  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2|  \le C\lambda^2.
+\end{equation}
+\end{prop}
+
+\begin{proof}
+The abbreviations
 $a^*=a[\kappa^*], a=a[\kappa], \tau^*=\tau[\kappa^*], \tau=\tau[\kappa]$.
-
-Using the Hadamard progressing wave approximation \ref{eqn:hadamard},
-the right-hand side of equation \ref{eqn:reg} is 
-$$
-\approx \mbox{argmin}_u (\|a[\kappa](\wl*u)(\cdot-\tau[\kappa])-a[\kappa^*]\wl(\cdot-\tau[\kappa^*])\|^2 + \lambda\|u\|^2)
-$$
-Re-write the approximation \ref{eqn:reg} in terms of the Fourier transforms $\hat{u_0}[\kappa_{\lambda};d],\fwl$ of $u_0[\kappa_{\lambda};d]$ and $\wl$:
-\begin{equation}
-  \label{eqn:freqreg}
-\hat{u}_{0,\sigma}[\kappa;d] \approx \mbox{argmin}_{\hat{u}}\frac{1}{2}\int d\omega \left(|a\hat{w}\hat{u}e^{i\omega \tau}-a^*\hat{w}e^{i\omega\tau^*}|^2 + \lambda |\hat{u}|^2\right)
-\end{equation}
-The normal equation for the least squares problem \ref{eqn:freqreg} is
+will be convenient in the computations to follow.
+
+The right-hand side of the third equation in \ref{eqn:leading} is then
+\[
+= \mbox{argmin}_u (\|a \wl*u*\delta_{\tau}-a^*\wl*\delta_{\tau^*}\|^2 + \lambda\|u\|^2)
+\]
+which in terms of the Fourier transforms $\fwl$, $\fu$ is
+\begin{equation}
+  \label{eqn:leadingft}
+= \mbox{argmin}_u \frac{1}{2\pi}(\|a \fwl \fu e^{-i\omega \tau}-a^*\fwl e^{-i\omega \tau^*}\|^2 + \lambda\|\fu\|^2).
+\end{equation}
+The normal equation for the least squares problem \ref{eqn:leadingft} is
 \[
@@
   \label{eqn:hatuf}
-\hat{u}_0[\kappa_{\lambda};d] \approx \frac{a^*}{a}\hat{g}_{\lambda,\frac{\lambda}{a^2}} e^{i\omega(\tau^*-\tau)}
+\hat{u}_{0,\lambda}[\kappa;\dla] = \frac{a^*}{a}\hat{g}_{\lambda,\frac{\lambda}{a^2}} e^{i\omega(\tau^*-\tau)}
 \end{equation}
@@
   \label{eqn:hatgf}
-\hat{g}_{\lambda,\sigma} = \frac{|\fwl|^2}{|\fwl|^2 + \sigma}.
-\end{equation}
-Since $\wl$ vanishes for large $|t|$, $\fwl$ is entire analytic,
-hence vanishes on a set of measure zero, whence $\hat{g}_{\lambda,\sigma}:$ tends to $1$ almost everywhere as $\sigma \rightarrow 0$.
-Hence the inverse Fourier transform $g_{\lambda,\sigma}$ tends to $\delta$ in the sense of distributions, and
-$$
-u_0[\kappa_{\lambda};d](t) \approx \frac{a^*}{a}g_{\lambda,\frac{\lambda}{a^2}}(t-(\tau^*-\tau))
-$$
-to a multiple of a shifted $\delta$, in fact exactly the distribution
-solution of $\bar{S}[\kappa_{\lambda}]u=d$ given in equation \ref{eqn:notl2}. However for $\lambda>0$, $u_0[\kappa_{\lambda};d]$ is square integrable.
-
-\cite{HuangSymes2015SEG} presented an analysis of the Matched Source
-Waveform Inversion (MSWI) objective function, which is the numerator
-of the right hand side in the definition \ref{eqn:tJz} of the AWI
-objective. Note that this definition requires summation over sources and receivers:
-\[
-J_{\rm MSWI}[\kappa_{\lambda};d] = \sum_{\bx_s,\bx_r}\int dt |tu_0[\kappa_{\lambda};d]|^2 \approx \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{\lambda}{a^2}}(t-(\tau^*-\tau))|^2
+\hat{g}_{\lambda,\lambda} = \frac{|\fwl|^2}{|\fwl|^2 + \lambda}.
+\end{equation}
+Since $\wl$ vanishes for large $|t|$, $\fwl$ is entire analytic hence
+vanishes on a set of measure zero, whence $\hat{g}_{\lambda,\lambda}:$
+tends to $1$ almost everywhere as $\lambda \rightarrow 0$. Hence the
+inverse Fourier transform $g_{\lambda,\lambda}$ tends to $\delta$ in
+the sense of distributions, and
+\begin{equation}
+  \label{eqn:u0s}
+\uzl[\kappa,\dla](t) = \frac{a^*}{a}g_{\lambda,\frac{\lambda}{a^2}}(t-(\tau^*-\tau))
+\end{equation}
+to a multiple of a shifted $\delta$, in fact exactly the
+non-square-integrable distribution
+solution of $\bar{\Sla}[\kappa]u=\dla$. However for $\lambda>0$, $\uzl[\kappa;\dla]$ is square integrable.
+
+Since $\fwl$ along with all of its derivatives
+tends to zero faster than any polynomial at $\infty$, $\fwl' \in
+L^2(\bR)$.
+Therefore $\hat{g}_{\lambda,\lambda}$
+and $\hat{g}_{\lambda,\lambda}'$ are also square-integrable, 
+from equation \ref{eqn:hatuf} $\hat{u}_{0,\lambda}'$ is also square
+integrable. Consequently $t u_{0\lambda}$ is square integrableTherefore
+$J_{\rm MSWI,\lambda,\lambda}$, the result of substituting in the
+right-hand side of the definition \ref{eqn:mswidef} $\uzl$ for
+$u_{\lambda}$, is well-defined. \cite{HuangSymes2015SEG} presented an analysis of
+$J_{\rm MSWI,\lambda,\lambda}$, as follows:
+\[
+J_{\rm MSWI,\lambda,\lambda}[\kappa;\dla] = \sum_{\bx_s,\bx_r}\int dt
+|t\uzl[\kappa;\dla]|^2 =  \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{\lambda}{a^2}}(t-(\tau^*-\tau))|^2
 \]
@@
 Since the Fourier transform $\fwl$ is even and real, so is
-$\hat{g}_{\lambda,\sigma}$, therefore $g_{\lambda,\sigma}$ is even, and the linear term (in t) on the
+$\hat{g}_{\lambda,\lambda}$, therefore $g_{\lambda,\lambda}$ is even, and the linear term (in t) on the
 right-hand side of \ref{eqn:MSWI} vanishes, So
@@
   \label{eqn:MSWIasym}
-  J_{\rm MSWI}[\kappa;d] \approx
+  J_{\rm MSWI,\lambda,\lambda}[\kappa;\dla] =
 \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{\lambda}{a^2}}(t)|^2
@@
 
-Recalling the definition \ref{eqn:pw} of pulse width, the
-approximation \ref{eqn:MSWI} can be re-written as
-\begin{equation}
-  J_{\rm MSWI}[\kappa_{\lambda};d] \approx
+Recalling the definition \ref{eqn:pw} of pulse width, \ref{eqn:MSWIasym} can be re-written as
+\begin{equation}
+  \label{eqn:MSWIasym2}
+  J_{\rm MSWI,\lambda,\lambda}[\kappa;\dla] = 
   \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{\lambda}{a^2}}\|^2\left(l(g_{\lambda,\frac{\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
@@
 $a^*, a$ both in the multiplier and in the scaling of $\lambda$. Since the
-amplitudes depend on $m$, their presence may influence the location
-(or even the presence) of stationary points of $J_{\rm MSWI}$ for
+amplitudes depend on $\kappa$, their presence may influence the location
+(or even the presence) of stationary points of $J_{\rm
+  MSWI,\lambda,\lambda}$ (or $J_{\rm MSWI}$, to which it is an approximation)  for
 models differing in their travel time predictions. So stationary
@@
 
-Comparing the definitions \ref{eqn:tJz} of $J_{\rm AWI}$ and
-\ref{eqn:MSWI} of $J_{\rm MSWI}$, ones sees
-that the two functions differ only by the factor $\|u_0[\kappa_{\lambda};d]\|^{-2}$ in
-the former, applied trace-by-trace. Using the geometric asymptotics approximation, 
-$$
-\|u_0[\kappa_{\lambda};d]\|^2 \approx \frac{(a^*)^2}{a^2}\|g_{\frac{\lambda}{a^2}}\|^2 + O(\lambda)
-$$
-Therefore, multiplication of the MSWI objective by $\|u_0[\kappa_{\lambda};d]\|^{-2}$
-precisely removes the amplitude factor and norm-squared of $g$ in the right-hand side of the
-relation \ref{eqn:MSWIasym}. That is,
-\begin{equation}
-  \label{eqn:preawitt}
-J_{\rm AWI}[\kappa;d] \approx \sum_{\bx_s,\bx_r} (\left(l(g_{\lambda,\frac{\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
- \end{equation}
-
-Note that
-\[
-  \int dt\, t^2|g_{\lambda,\sigma}(t)|^2 = \int d\omega
-  \left|\frac{d\hat{g}_{\lambda,\sigma}}{d\omega}\right|^2
-  \approx \int d\omega  \left(\frac{2 \mbox{Re }\bar{\hat{w}}_{\lambda}\fwl' 
-    \sigma}{(|\fwl|^2 + \sigma)^2}\right)^2
-\]
-\[
-  =\int d\omega \frac{(2 \lambda^{1/2}\hat{w}_1(\lambda
-    \omega)\lambda^{3/2}\hat{w}_1'(\lambda \omega) 
-    \sigma)^2}{(\lambda|\hat{w}_1(\lambda
-    \omega)|^2 + \sigma)^4}
-\]
-\[
-  =\frac{1}{\lambda}\int d\omega_1 \frac{(2 \lambda^2\hat{w}_1(\omega_1)\hat{w}_1'( \omega_1) \sigma)^2}{(\lambda|\hat{w}_1(
-    \omega_1)|^2 + \sigma)^4}
-  =\frac{\sigma^2}{\lambda}\int d\omega_1
-  \frac{(2\hat{w}_1(\omega_1)\hat{w}_1'( \omega_1)
-    )^2}{\left(|\hat{w}_1(\omega_1)|^2 + \frac{\sigma}{\lambda}\right)^4}
-\]
-Substitute $\sigma = \lambda/(a^2)$ to obtain
-\[
- \int dt\,
-  t^2|g_{\lambda,\frac{\lambda}{a^2}}(t)|^2
-  =\frac{\lambda}{a^6}\int d\omega_1
-  \frac{(2\hat{w}_1(\omega_1)\hat{w}_1'( \omega_1)
-    )^2}{\left(|\hat{w}_1(\omega_1)|^2 + \frac{1}{a^2}\right)^4}
-\]
-\begin{equation}
-  \label{eqn:g1}
- = \frac{\lambda}{a^2} \int d\omega
-  \frac{(2a\hat{w}_1a\hat{w}_1'
-    )^2}{\left(|a\hat{w}_1|^2 + 1\right)^4}
-\end{equation}
-Also
-\[
-  \|g_{\lambda,\sigma}\|^2 = \int d\omega\,
-  \left(\frac{|\fwl|^2(\omega)}{|\fwl|^2(\omega) + \sigma}\right)^2
-= \int d\omega\,
+The trace-by-trace normalization in the definition of
+the AWI objective removes the model-dependent factors. Denote by
+$J_{\rm AWI, \lambda,\lambda}$ the restult of subtituting
+$\uzl$ for $u_{\lambda}$ in the definition \ref{eqn:awidef} of
+$J_{\rm AWI}$:
+\begin{equation}
+  \label{eqn:AWIasym}
+J_{\rm AWI,\lambda,\lambda}[\kappa;\dla] = \sum_{\bx_s,\bx_r}\frac{\int dt
+(|t\uzl[\kappa;\dla])^2}{\|\uzl[\kappa;\dla]\|^2} 
+\end{equation}
+We have already analyzed the numerator, arriving at
+equation \ref{eqn:MSWIasym2}. From the result \ref{eqn:u0s}, for each
+$(\bx_s,\bx_r) \in X_{sr}$,
+\begin{equation}
+  \label{eqn:u0snorm}
+  \|\uzl[\kappa;\dla]\|^2=
+  \frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{\lambda}{a^2}}\|^2.
+\end{equation}
+Combine equations \ref{eqn:AWIasym} and \ref{eqn:u0snorm} to obtain
+\begin{equation}
+  \label{eqn:AWIasym2}
+  J_{\rm AWI,\lambda,\lambda}[\kappa;\dla] = \sum_{\bx_s,\bx_r} \left(l(g_{\lambda,\frac{\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
+ \end{equation}
+
+So the leading-term modification of $J_{\rm AWI}$ differs from mean square
+travel time error by the pulse width of
+$g_{\lambda,\frac{\lambda}{a^2}}$. The analogous relation for $_{\rm
+  MSWI}$ involves the norm-squared as well. To assess these
+quantities, note first that
+\[
+  \|g_{\lambda,\frac{\lambda}{a^2}}\|^2 =\frac{1}{2\pi} \int d\omega\,
+  \left(\frac{|\fwl|^2(\omega)}{|\fwl|^2(\omega) + \frac{\lambda}{a^2}}\right)^2
+=\frac{1}{2\pi} \int d\omega\,
   \left(\frac{|\lambda^{1/2}\hat{w}_1(\lambda
       \omega)|^2(\omega)}{|\lambda^{1/2}\hat{w}_1(\lambda
-      \omega)|^2 + \sigma}\right)^2
-\]
-\[
-=\int d\omega\,
+      \omega)|^2 + \frac{\lambda}{a^2}}\right)^2
+\]
+\[
+=\frac{1}{2\pi}\int d\omega\,
   \left(\frac{|\hat{w}_1(\lambda
       \omega)|^2}{|\hat{w}_1(\lambda
-      \omega)|^2 + \frac{\sigma}{\lambda}}\right)^2
-=  \frac{1}{\lambda} \int d\omega_1\,
-  \left(\frac{|\hat{w}_1(\omega_1)|^2}{|\hat{w}_1(\omega_1)|^2 + \frac{\sigma}{\lambda}}\right)^2
-\]
-Substitute $\sigma = \lambda/(a^2)$  and rearrange to obtain
+      \omega)|^2 + \frac{\lambda}{a^2\lambda}}\right)^2
+=  \frac{1}{2\pi \lambda} \int d\omega_1\,
+  \left(\frac{|\hat{w}_1(\omega_1)|^2}{|\hat{w}_1(\omega_1)|^2 + \frac{\lambda}{a^2\lambda}}\right)^2
+\]
+At this point, it is possible to identify an advantageous choice of
+the Tihonov penalty weight $\lambda$: choose
+$\lambda = \lambda$  and rearrange to obtain
 \begin{equation}
@@
 \end{equation}
+That is, with this choice $\lambda = \lambda$, the wavelength proxy
+$\lambda$ appears only in the factor in front of the integral sign.
+
+Similarly,
+\[
+  \int dt\, t^2|g_{\lambda,\frac{\lambda}{a^2}}(t)|^2 = \frac{1}{2\pi} \int d\omega
+  \left|\frac{d\hat{g}_{\lambda,\frac{\lambda}{a^2}}}{d\omega}\right|^2
+  =  \int d\omega  \left(\frac{2 \mbox{Re }\bar{\hat{w}}_{\lambda}\fwl' 
+    \frac{\lambda}{a^2}}{(|\fwl|^2 + \frac{\lambda}{a^2})^2}\right)^2
+\]
+\[
+  =\frac{1}{2\pi}\left(\frac{\lambda}{a^2}\right)^2 \int d\omega \frac{(2 \lambda^{1/2}\hat{w}_1(\lambda
+    \omega)\lambda^{3/2}\hat{w}_1'(\lambda \omega))^2}{(\lambda|\hat{w}_1(\lambda
+    \omega)|^2 + \frac{\lambda}{a^2})^4}
+\]
+\[
+  =\frac{1}{2\pi\lambda}\left(\frac{\lambda}{a^2}\right)^2\int d\omega_1 \frac{(2 \lambda^2\hat{w}_1(\omega_1)\hat{w}_1'( \omega_1))^2}{(\lambda|\hat{w}_1(
+    \omega_1)|^2 + \frac{\lambda}{a^2})^4}
+  =\frac{1}{2 \pi \lambda}\left(\frac{\lambda}{a^2}\right)^2\int d\omega_1
+  \frac{(2\hat{w}_1(\omega_1)\hat{w}_1'( \omega_1)
+    )^2}{\left(|\hat{w}_1(\omega_1)|^2 + \frac{\lambda}{a^2\lambda}\right)^4}
+\]
+With the choice $\lambda = \lambda$, this becomes
+\[
+  =\frac{\lambda}{2\pi a^4}\int d\omega_1
+  \frac{(2\hat{w}_1(\omega_1)\hat{w}_1'( \omega_1)
+    )^2}{\left(|\hat{w}_1(\omega_1)|^2 + \frac{1}{a^2}\right)^4}
+\]
+\begin{equation}
+  \label{eqn:g1}
+ = \frac{\lambda}{2\pi} \int d\omega
+  \frac{(2a\hat{w}_1a\hat{w}_1'
+    )^2}{\left(|a\hat{w}_1|^2 + 1\right)^4}
+\end{equation}
+
 Combining the relations \ref{eqn:g0} and \ref{eqn:g1}, obtain
@@
 in which $W(m,w_1)$ is bounded by the variation of the amplitude
-$a[\kappa](\bx_r,\bx_s)$, and the choice of ``mother wavelet'' $w_1$, but
+$a[\kappa](\bx_r,\bx_s)$, hence uniformly for $\log \kappa \in M$ (property
+H2), and the choice of ``mother wavelet'' $w_1$, but
 independent of $\lambda$.
@@
 \begin{equation}
+  \label{eqn:mswitt}
+  J_{\rm MSWI,\lambda,\lambda}[\kappa;\dla] =
+  \frac{1}{\lambda} \sum_{{\bf x}_s,{\bf x}_r} C[\kappa](\bx_s,\bx_r)(\tau[\kappa^*]({\bf
+  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 + O(\lambda)
+\end{equation}
+
+\begin{equation}
   \label{eqn:awitt}
-J_{\rm AWI}[\kappa;d] \approx \sum_{{\bf x}_s,{\bf x}_r} (\tau[\kappa^*]({\bf
+J_{\rm AWI,\lambda,\lambda}[\kappa;\dla] = \sum_{{\bf x}_s,{\bf x}_r} (\tau[\kappa^*]({\bf
   x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 + O(\lambda^2)
 \end{equation}
-In other words, the AWI objective is {\em exactly} an approximation to
-the travel-time tomography objective, with errors proportional to the
-``wavelength'' $\lambda(w)$ and the regularization parameter
-$\lambda$.
-
-And that's ``why it works'' - when geometric asymptotics in the form
-given by equation \ref{eqn:hadamard} is an accurate approximation to the Green's function.
+
+\end{proof}
+
+In particular, the modified AWI objective $J_{\rm AWI,\lambda}$ is {\em exactly} an approximation to
+the travel-time tomography objective, with errors proportional to the squared
+wavelength proxy $\lambda$.
+
+\subsection{Estimating the Remainder}
+
+%And that's ``why it works'' - when geometric asymptotics in the form
+%given by equation \ref{eqn:hadamard} is an accurate approximation to
+%the Green's function.
+
+To evaluate the difference between $J_{\rm AWI}$ and
+$J_{\rm AWI,\lambda}$, first estimate the norms of differences
+$\|u_{\lambda} - \uzl\|$ and $\|T(u_{\lambda} -
+\uzl)\|$.
+
+Suppressing $ \bx_s,\bx_r$ and using the notation of the last
+subsection, definitions \ref{eqn:Fl0}, \ref{eqn:Fl1} and
+\ref{eqn:sdeflam} can be re-written
+\[
+  \Sl u = (a \wl*u + (bH) * \wl *u)(t-\tau)
+\]
+where $bH$ is an abbreviation for
+$b[\kappa](\bx_r,t-\tau[\kappa];\bx_s)H(t-\tau[\kappa])$. Note that
+$b$ decays exponetially, is smooth, and all derivatives have limits
+at $t=0$. Similarly,
+\[
+  d = \Fl[\kappa^*] = (a^*\wl + (bH)  * \wl) (t-\tau).
+\]
+Recall (equations \ref{eqn:Fl0}, \ref{eqn:Fl1}, re-written as
+\[
+\Fla = a \wl * \delta_{\tau}, \,   \Flb  = (f_{\lambda}b(0) -
+f_{\lambda}*(bH))*\delta_{\tau}.
+\]
+Since $S_{\lambda,i} u = F_{\lambda,i} * u, \,i=0,1,$ $\wl$ is even,
+and $f_{\lambda}$ is odd,
+\[
+ ((\Sla)^T \Slb + (\Slb)^T\Sla)u = [\rho a b(0)( \check{\wl} *\delta_{-\tau}*f_{\lambda}*\delta_{\tau} + \wl
+ *\delta_{\tau}*\check{f}_{\lambda}*\delta_{-\tau}) 
+\]
+\[
+-
+ \rho a(\check{\wl}*\delta_{-\tau} *f_{\lambda}*(bH)*\delta_{\tau} + \wl*\delta_{\tau}*\check{f}_{\lambda}*\check{b}\check{H)}*\delta_{-\tau}) ]*u
+\]
+\[
+  = \rho a(\wl*f_{\lambda}*(\check{b}\check{H}-(bH)) *u
+\]
+Since $\wl = \rho \partial_t f_{\lambda}$ and $(\partial_t
+(\check{b}\check{H}-bH))*u = (-2\partial_tb(0)u + (\partial_t\check{b}\check{H} -\partial_tb H))*u$, this is
+\begin{equation}
+  \label{eqn:symmprods0s1}
+  = \rho^2 a(-2\partial_tb(0) \delta + (\partial_t\check{b}\check{H} -
+  \partial_tb H)) * f_{\lambda}*f_{\lambda}*u_{\lambda}
+\end{equation}
+The bounds \ref{eqn:best}, \ref{eqn:lambdanorms} imply that
+\begin{equation}
+  \label{eqn:s0s1norm}
+  \|(\Sla)^T\Slb + (\Slb)^T\Sla\| \le C \lambda^3,
+\end{equation}
+with $C$ uniform in $\log \kappa \in M$. From \ref{eqn:Srembd},
+\begin{equation}
+  \label{eqn:noperr}
+  \|\Sl^T\Sl - (\Sla)^T\Sla\| \le C \lambda^3.
+\end{equation}
+A little algebra yields
+\begin{equation}
+  \label{eqn:step1}
+  ((\Sla)^T\Sla + \lambda I) (u_{\lambda}-\uzl) = (\Slb)^T d_{\lambda} + (\Sla)^T(d_{\lambda}-d_{0,\lambda})- 
+  ((\Sla)^T\Slb + (\Slb)^T\Sla + (\Slb)^T\Slb)u_{\lambda}
+\end{equation}
+Note that the bounds \ref{eqn:Srembd} imply
+\[
+  \|\Sl^T d\| \le C \lambda^{\frac{1}{2}}\|d\|,
+\]
+\[
+  \|(\Sla)^T (d-d_0)\| = \|(\Sla)^T(f_{\lambda}b^*(0) -
+f_{\lambda}*(b^*H)*\delta_{\tau^*}\| \le C \lambda^2,
+\]
+and
+\[
+ \| (\Slb)^Td \| \le C \lambda^{3/2}.
+\]
+Hence
+\[
+  \|u_{\lambda}\| \le C\lambda^{-\frac{1}{2}}\|d\|.
+\]
+Combine the preceding three bounds with identity \ref{eqn:step1} and
+bound \ref{eqn:noperr} to obtain
+\begin{equation}
+  \label{eqn:step2}
+  \|u_{\lambda}-\uzl\| \le C
+  \lambda^{\frac{1}{2}}\|d_{\lambda}\|
+\end{equation}
+
+From identity \ref{eqn:step1},
+\[
+   ((\Sla)^T\Sla + \lambda I) T(\ul-\uzl) = -[T,
+   (\Sla)^T\Sla](\ul-\uzl) - T \Slb)^T \dl 
+ \]
+\begin{equation}
+  \label{eqn:step3}
+ + T (\Sla)^T(\dl-\dzl)- 
+  T((\Sla)^T\Slb + (\Slb)^T\Sla + (\Slb)^T\Slb)\ul)
+\end{equation}
+
+To estimate the first term on the right-hand side, note that
+\[
+  [T,\Sla] = a \tau \delta_{\tau} * \wl* + a \delta_{\tau} * (T\wl)*
+\]
+and
+\[
+  [T,(\Sla)^T] = -a \tau \delta_{-\tau} * \wl* + a 
+  \delta_{-\tau}*(T\wl)*,
+\]
+so
+\[
+  [T,  (\Sla)^T\Sla] = 2 a^2 \wl*T\wl*.
+\]
+Since
+\begin{equation}
+  \label{eqn:twl}
+  \|T\wl\|_{L^1(X_{sr}\times \bR)} =
+  \|\partial_t(T\fl)-\fl\|_{L^1(\bR)} =\lambda^{\frac{3}{2}}\int dt_1
+  |t_1||\partial f_1(t_1)| \le C \lambda^{\frac{3}{2}}
+\end{equation}
+the estimate \ref{eqn:step2} implies
+\begin{equation}
+  \label{step5}
+  \|[T,  (\Sla)^T\Sla](\ul-\uzl)\| \le C \lambda^{2}\|\dl\|,
+\end{equation}
+
+For the second term, compute
+\[
+  [T, \Slb] = [T,bH*\delta_{\tau}*\wl*] = ((T+\tau)bH*\wl
+  + bH*T\wl)*\delta_{\tau}
+\]
+so
+\[
+  [T,(\Slb)^T]  =  ((T - \tau) \check{b}\check{H})*\wl
+  + \check{b}\check{H}*T\wl)*\delta_{-\tau}
+\]
+\[
+= \check{b}\check{H}*\fl
+-\tau)b(0)\delta *\fl + (T-\tau)
+\partial_t\check{b}\check{h}*\fl +
+\check{b}\check{H}*(\partial_t(T\fl)-\fl)
+\]
+Using \ref{eqn:twl}, conclude that
+\begin{equation}
+  \label{eqn:step6}
+  [T, (\Slb)^T] \le C \lambda^{\frac{3}{2}}
+\end{equation}
+
+Observe that
+\[
+  T\dl=T(a^*\delta + b^*H)*\delta_{\tau^*}*\wl
+\]
+\[
+  =Tb^*H *\delta_{\tau^*}*\wl+(a^*\delta+b^*H)\tau^*\delta_{\tau^*}\wl
+  +(a^*\delta + b^*H)*\delta_{\tau^*}*T\wl,
+\]
+so the exponential decay of $b^*$ implies that $\|T\dl\| \le
+C\|\dl\|$. Thus the estimate \ref{eqn:step6} implies
+\begin{equation}
+  \label{eqn:step7}
+  \|T(\Slb)^T\dl\| \le C \lambda^{3/2}\|\dl\|.
+\end{equation}
+
+%%%%%%%%%%%%%%%
+
+$(Tu)(t) = tu(t)$
+
+$T^T = T$
+
+$[T,\delta_{\tau}*] = \tau \delta_{\tau}*$
+
+$[T,k *] = (Tk)*$ for any $k \in {\cal S}'(\bR)$
+
+\[
+  [T,\Sla] = a \tau \delta_{\tau} * \wl* + a \delta_{\tau} * (T\wl)*
+\]
+\[
+  [T,(\Sla)^T] = -a \tau \delta_{-\tau} * \wl* + a \delta_{-\tau}*(T\wl) *
+\]
+\[
+  [T,\Sl] = a (\tau  \wl +  (T\wl) ) \delta_{\tau} + [T,(bH)*\delta_{\tau}*\wl]
+\]
+\[
+  =a (\tau  \wl +  (T\wl) ) \delta_{\tau}* + (Tb)H *\delta_{\tau}*\wl* +
+  (bH)*(\tau \wl + T\wl)*\delta_{\tau}*
+\]
+\[
+  = (a\delta + (bH))*(\tau \wl +
+  T\wl)*\delta_{\tau}* + (Tb)H *\delta_{\tau}*\wl*
+\]
+\[
+[T,\Sl^T] =   = (a\delta + (\check{b}\check{H}))*(-\tau \wl +
+  T\wl)*\delta_{-\tau}* + (T\check{b})\check{H} *\delta_{-\tau}*\wl*
+\]
+
+\[
+  [T,(\Sla)^T\Sla] = [T,(\Sla)^T]\Sla + (\Sla)^T [T,\Sla]
+\]
+\[
+= (-a \tau \delta_{-\tau} * \wl + a \delta_{-\tau}*(T\wl)) *
+a\delta_{\tau} * \wl
+\]
+\[
+  + a \delta_{-\tau} * \wl *  ( a \tau \delta_{\tau} * \wl + a
+  \delta_{\tau} * (T\wl))
+\]
+\[
+  = 2 a^2 \wl * T\wl *
+\]
+
+\[
+  [T,\Sl^T\Sl] = [T,\Sl^T]\Sl + \Sl^T [T,\Sl]
+\]
+\[
+  =((a\delta + (\check{b}\check{H}))*(-\tau \wl +
+  T\wl)*\delta_{-\tau}* + (T\check{b})\check{H}
+  *\delta_{-\tau}*\wl*)(a \delta + (bH))*\wl*\delta_{\tau}
+\]
+\[
+  +(a \delta + (\check{b}\check{H}))*\wl*\delta_{-\tau}*
+  ((a\delta + (bH))*(\tau \wl +
+  T\wl)*\delta_{\tau}* + (Tb)H *\delta_{\tau}*\wl*)
+\]
+\[
+=(a\delta + (\check{b}\check{H}))*(-\tau \wl +
+T\wl)*\delta_{-\tau}*(a \delta + (bH))*\wl*\delta_{\tau} +
+\]
+\[
+  (a \delta + (\check{b}\check{H}))*\wl*\delta_{-\tau}*(a\delta + (bH))*(\tau \wl +
+  T\wl)*\delta_{\tau}*
+\]
+\[
+  ...
+\]
+\[
+  = 2(a \delta + (\check{b}\check{H}))*(a\delta + (bH)*\wl*T\wl +
+\]
+
+From identity \ref{eqn:symmprods0s1},
+\[
+  [T,(\Sla)^T\Slb+(\Slb)^T\Sla] = \rho^2 a[T,(-2\partial_t b(0) \delta +
+  (\partial_t\check{b}\check{H}-\partial_tb
+  H))]*f_{\lambda}*f_{\lambda}*
+\]
+\[
+  + 2\rho^2 a (-2\partial_t b(0) \delta +
+  (\partial_t\check{b}\check{H}-\partial_tb
+  H))]*f_{\lambda}*Tf_{\lambda}*
+\]
+\[
+  =
+  \rho^2 a[ T (\partial_t\check{b}\check{H}-\partial_tb
+  H))]*f_{\lambda}*f_{\lambda}* +  2\rho^4 a (-2\partial_t b(0) \delta +
+  (\partial_t\check{b}\check{H}-\partial_tb
+  H))]*f_{\lambda}*Tf_{\lambda}*
+\]
+%\[
+%  [T,\Slb] = ((Tb)H) *\wl * \delta_{\tau}* + (bH) * T\wl *
+%  \delta{\tau} * + \tau (bH)* \wl * \delta_{\tau}
+%\]
+%\[
+%[T,(\Slb)^T] = - [T,\Slb]^T =-(\checi((Tb)\check{H}*\wl*\delta{-\tau} +
+%(\check{b}\check{H}) * (-Twl)*\delta{-\tau}^* + \tau
+%(\check{b}\check{H))*\wl*\delta_{\-tau}]
+%\]
+%\begin{equation}
+%  \label{eqn:step3}
+%  ((\Sla)^T\Sla + \lambda I) (u_{\lambda}-\uzl) = (\Slb)^T d_{\lambda} + (\Sla)^T(d_{\lambda}-d_{0,\lambda})- 
+%  ((\Sla)^T\Slb + (\Slb)^T\Sla + (\Slb)^T\Slb)u_{\lambda}
+%\end{equation}
 
@@
 following choices, for each $m$ in the domain of $F$: $S$ is the regularized
-modeling operator $(S_{\lambda}[\kappa],\sigma I)^T$.  Its domain is $U=L^2(\bR)$, of
+modeling operator $(S_{\lambda}[\kappa],\lambda I)^T$.  Its domain is $U=L^2(\bR)$, of
 which $u$ (the adaptive filter) is a member, and 

### 2024-10-18 repro/papers/awi/paper.tex
more

@@
 to the equation \ref{eqn:awe2}) (\cite{Friedlander:75},
-p. 16). Hadamard's decomposition holds in a space-time cylinder
-$\Omega \times \bR$,  and $\Omega\subset \bR^3$ 
+p. 16).
+
+Throughtout the following discussion, $\Omega\subset \bR^3$ 
 and $M \subset C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)$ are bounded open sets satisfying 
@@
   $|\log a[\kappa](\bx_s,\bx_r)| < A$;
-\item[H3. ] there exist $B \ge 0,\delta > 0$ so that for
-$(\bx_s,\bx_r) \in X_{sr}$, and $\kappa \in
-M$,
-\begin{eqnarray}
-  |b[\kappa](\bx_r,t;\bx_r,s)| &\le& B e^{-\delta t} \nonumber \\
-  |\partial_t b[\kappa](\bx_r,t;\bx_r,s)| &\le& B e^{-\delta t}
+\item[H3. ] for any $k \ge 0$, there exists $B \ge 0,\delta > 0$ so
+  that for any $(\bx_s,\bx_r) \in X_{sr}$, $\kappa \in M$, 
+\begin{equation}
+  |\partial^k_t b[\kappa](\bx_r,t;\bx_r,s)| \le B e^{-\delta t}
   \label{eqn:best}
-\end{eqnarray}
+\end{equation}
 \end{itemize}
@@
 %\end{equation}
+
+\subsection{Notational Conventions for Norms and Bounds}
+The Banach spaces
+$l^p(X_{sr},L^p(\bR))$ will be abbreviated as $L^p(X_{sr} \times
+\bR)$. The norm in a Banach space $B$ will be denoted
+$\|\cdot\|_{B}$. The norm symbol
+$\|\cdot\|$ without subscript will be understood to mean either
+$\|\cdot\|_{L^2(\bR)}$ or $\|\cdot\|_{L^2(X_{sr}\times
+  \bR)}$, as context indicates.
+
+In a statement pertaining to all
+$\kappa \in M$, the letter $C$, or sometimes $C_k,
+k=0,1,2,...$ or
+$C_*,C^*$, will designate a constant (or constants) independent of
+$\kappa$.
 
@@
   \label{eqn:sdef}
-  S[\kappa]u \equiv u*F[\kappa]
+  S[\kappa]u = u*F[\kappa]
 \end{equation}
@@
 begin with, $S[\kappa] = F[\kappa]*$ defines a bounded map on
-$L^2(X_{sr}\times \bR)$, since from \ref{eqn:fwdgreen} and the definition of
-  $S[\kappa]$,
-\[
-  S[\kappa]u = \rho( a[\kappa]\delta    +  (b[\kappa]H)) *\delta_{\tau[\kappa]}*\partial_t f. *u,
-\]
-The bounds \ref{eqn:best} yield $bH \in
-  L^1(X_{sr}\times \bR)$, and property F1, $\partial_t f \in
-  L^1(\bR)$.
+$L^2(X_{sr}\times \bR)$, since from \ref{eqn:fwdgreen} and the
+definition \ref{eqn:sdef} of  $S[\kappa]$,
+\[
+  S[\kappa]u = \rho( a[\kappa]\delta    +  (b[\kappa]H)) *\delta_{\tau[\kappa]}*\partial_t f *u,
+\]
+\begin{equation}
+  \label{eqn:sgood}
+  = \rho(a[\kappa]\delta_{\tau[\kappa]}*\partial_t f +
+  (b[\kappa](0)\delta + (\partial_t b[\kappa])H)
+  *\delta_{\tau[\kappa]}*f) * u
+\end{equation}
+The key properties of $S$ follow directly from Young's inequality and
+the representation \ref{eqn:fwdgreen}:
+\begin{lemma}
+  \label{thm:snorm}
+  Under the assumptions leading to \ref{eqn:fwdgreen} and
+  the properties F1 and F2, $S[\kappa]$ is a bounded operator on
+  $L^2(X_{sr} \times \bR)$, and there exists $C>0$ so
+  that for any $\kappa \in M$,
+  \begin{equation}
+    \label{eqn:snorm}
+    \|S[\kappa]\| \le C\|\partial_t f\|_{L^1(\bR)}.
+  \end{equation}
+\end{lemma}
 
@@
 \end{equation}
+
+\begin{prop}
+  \label{thm:unorm}
+  Under the assumptions G1 - G4 and F1, F2 justifying the
+  representation \ref{eqn:fwdgreen}, the least squares problem
+  \ref{eqn:reg} with $\sigma > 0$ has a unique solution
+  $u_{\sigma}[\kappa;d] \in L^2(X_{sr} \times \bR)$ satisfying
+  \begin{equation}
+    \label{eqn:unorm}
+    \|u_{\sigma}[\kappa;d]\| \le \frac{C}{\sigma}\|\partial_t f\|_{L^1(\bR)}\|d\|.
+  \end{equation}
+\end{prop}
+\begin{proof}
 For $\sigma > 0$, the Lax-Milgram Theorem (\cite{Yosida}, section
-III.7) implies that this least-squares problem has a unique solution
-$u_{\sigma}[\kappa;d] \in L^2(X_{sr} \times \bR)$. Define the modified
+III.7) assures the existence of unique solution. 
+\end{proof}
+
+Define the modified
 AWI objective $J_{\rm AWI,\sigma}$ by
@@
   Suppose that $d, Td \in L^2(X_{sr} \times \bR)$. Then
-  $Tu_{\sigma}[\kappa;d]  \in L^2(X_{sr} \times \bR)$.
+  $Tu_{\sigma}[\kappa;d]  \in L^2(X_{sr} \times \bR)$, and
+  \[
+    \|Tu_{\sigma}[\kappa;d]\| \le
+  \]
+  \begin{equation}
+    \label{eqn:tunorm}
+    \frac{C}{\sigma}\left(\left(
+        \frac{\|\partial_t f\|^2_{L^1(\bR)}}{\sigma}(\|f\|_{L^1(\bR)}+\|T\partial_tf\|_{L^1(\bR)})
+      \right)\|d\| \\ 
+    + \|\partial_t  f\|_{L^1(\bR)}(\|d\| +\|Td\|)\right).
+  \end{equation}
 \end{prop}
@@
   \]
+  \[
+    +2\rho^2((a\delta  +  (\check{b}\check{H}))*(a\delta  +
+    (bH)))* \partial_t f  *T\partial_t f*
+  \]
+  \[
+    =\rho^2((T\partial_t\check{b}\check{H}) *(a\delta  +
+    (bH))*+ (a\delta  +  (\check{b}\check{H}))*(T\partial bH))* f
+    *\partial_t f*
+   \]
   \begin{equation}
@@
   \end{equation}
+
   Referring to the exponential decay assumption \ref{eqn:best}
@@
   $L^1(X_{sr} \times \bR)$, so $[T,S^TS]$ restricts to a bounded
-  operator on $L^2(X_{sr}\times \bR)$. A similar computation shows the
-  same property for $[T,S^T]$. Since $u_{\sigma} \in L^2(X_{sr}\times
-  \bR)$,
-  \[
-    T(S^TS+\sigma I)u_{\sigma} =  [T,(S^TS+\sigma I)]u_{\sigma} +
-    (S^TS+\sigma I)Tu_{\sigma}
-  \]
-  \[
-    = [T,S^T]d + S^TTd,
-  \]
-  that is,
-  \[
-    (S^TS+\sigma I)Tu_{\sigma} = e \in L^2(X_{sr} \times \bR).
+  operator on $L^2(X_{sr}\times \bR)$, with
+  \begin{equation}
+    \label{eqn:tsscommnorm}
+    \|[T,S^TS]\| \le C\|\partial_t f\|_{L^1(\bR)}(\|f\|_{L^1(\bR)}+\|T\partial_tf\|_{L^1(\bR)})
+  \end{equation}
+
+  Similarly,
+  \[
+    [T,S]=T(\partial_t b)H*\delta_{\tau}*f + \tau(a\delta +
+    bH)*\delta_{\tau}*\partial_t f + (a\delta +
+    bH)*\delta_{\tau}*T\partial_t f.
+  \]
+  Young's inequality shows that the RHS lies in $L^1(X_{sr}\times
+  \bR)$, and that there exists $C$ independent of $\kappa$ so that
+  $[T,S^T]$ obeys the same estimate as does $S$ (Lemma
+  \ref{thm:snorm}), that is,
+  \begin{equation}
+    \label{eqn:tcommst}  
+    \|[T,S^T]\| = \|[T,S]\| \le C\|\partial_t f\|_{L^1(\bR)}
+  \end{equation}
+
+  Multiplying the normal equation for $u_{\sigma}$Rearranging,
+  \[
+    (S^TS+\sigma I)Tu_{\sigma} = -[T,S^TS]u_{\sigma} + [T,S^T]d + S^TTd.
   \]
@@
   that $S^TS+\sigma I$ has a bounded inverse, so the distribution
-  $Tu_{\sigma} \in L^2(X_{sr} \times \bR)$.
+  $Tu_{\sigma} \in L^2(X_{sr} \times \bR)$. The inverse is bounded by
+  $1/\sigma$, so the estimates \ref{eqn:snorm}, \ref{eqn:unorm},
+  \ref{eqn:tcommst}, and \ref{eqn:tsscommnorm} combine to yield the
+  bound \ref{eqn:tunorm}.
+
 \end{proof}
@@
   \label{thm:noisefree}
-  Suppose that $\kappa \in M$ and $d = F[\kappa]$. Then $Td
-  \in L^2(X_{sr} \times \bR)$.
+  Suppose that $\kappa \in M$. Then for $p \ge 1$, $F[\kappa], TF[\kappa]
+  \in L^p(X_{sr} \times \bR)$, and there exist $C_0, C_1, C_2 >0$ so that
+  \begin{eqnarray}
+    \label{eqn:dnorms}
+    C_1\|\partial_tf\|_{L^p(\bR)} - C_2\|f\|_{L^p(\bR)} \le \|F[\kappa]\|_{L^p(\bR)} & \le & C_0\|\partial_tf\|_{L^p(\bR)}, \nonumber \\
+    \|TF[\kappa]\|_{L^p(\bR)} & \le & C_0\|\partial_tf\|_{L^p(\bR)}.
+  \end{eqnarray}
 \end{prop}
@@
   A simple estimate shows that $F[\kappa]$ satisfies an exponential
-  decay estimate of the form G4, whence the conclusion follows.
+  decay estimate of the form G4, whence the first estimate
+  follows. Since $TbH \in W^{1,1}(\bR)$,
+  \[
+    T\rho(a\delta + bH)*\delta_{\tau}*\partial_t f = \partial_t TbH *
+    \delta_{\tau} * f + \tau \rho (a\delta +
+    bH)*\delta_{\tau}*\partial_t f  + (a\delta +
+    bH)*\delta_{\tau}*T\partial_t f.
+  \]
+  All terms involve convolution of $f$ or $\partial_t f$ with kernels
+  uniformly $L^1$-bounded in $\kappa \in M$, so the second estimate follows. 
 \end{proof}
@@
   \label{eqn:Srembd}
-  \|\Sl[\kappa]u\| &\le&  C\lambda^{\frac{1}{2}}\|u\|,\nonumber \\
-  \|\Slb[\kappa]u\| & \le& C\lambda^{\frac{3}{2}}\|u\|.
+  \|\Sl[\kappa]\| &\le&  C\lambda^{\frac{1}{2}},\nonumber \\
+  \|\Sla[\kappa]\|& \le& C\lambda^{\frac{1}{2}}. \nonumber \\
+  \|\Slb[\kappa]\| & \le& C\lambda^{\frac{3}{2}}.
 \end{eqnarray}
@@
   \label{eqn:mswideflam}
-  J_{\rm MSWI,\lambda}[\kappa;\dl] =  \|T\ul[\kappa;d]\|^2
+  J_{\rm MSWI,\lambda}[\kappa;\dl] =  \|T\ul[\kappa;\dl]\|^2
 \end{equation}
@@
   J_{\rm AWI,\lambda}[\kappa;\dl] = \sum_{(\bx_s,\bx_r) \in X_{sr}} 
-  \frac{\|T\ul[\kappa;d](\bx_r,\cdot;\bx_s)\|^2}{\|\ul[\kappa;d](\bx_r,\cdot;\bx_s)\|^2}.
+  \frac{\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|^2}{\|\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|^2}.
 \end{equation}
@@
 
+\begin{prop} Under the hypotheses of Proposition \ref{thm:leading},
+  \begin{eqnarray}
+    \label{eqn:diffests}
+    \|\ul-\uzl\| &\le& C \nonumber \\
+    \|T(\ul-\uzl)\| &\le& C
+  \end{eqnarray}
+\end{prop}
+
+\begin{proof}
+  A little algebra yields
+  \begin{equation}
+    \label{eqn:step1}
+    ((\Sla)^T\Sla + \lambda I) (u_{\lambda}-\uzl) = (\Slb)^T d_{\lambda} + (\Sla)^T(d_{\lambda}-d_{0,\lambda})- 
+    ((\Sla)^T\Slb + (\Slb)^T\Sla + (\Slb)^T\Slb)u_{\lambda}
+  \end{equation}
+  Note that the bounds \ref{eqn:Srembd} imply
+\[
+  \|(\Sla)^T (\dl-\dzl)\| = \|(\Sla)^T(\rho \fl b^*(0)\delta +
+  \rho \fl*(\partial_tb^*H)*\delta_{\tau^*}\|
+\]
+\[
+  = \rho \|a \wl * \fl * (b^*(0)\delta - \partial_tb^*H)\|
+  \le C \lambda^{\frac{3}{2}}\|\dl\|
+\]
+and similarly
+\[
+ \| (\Slb)^T\dl \| \le C \lambda^{3/2}\|\dl\|.
+\]
+A bit more algebra gives
+\[
+  (\Sla)^T\Slb + (\Slb)^T\Sla + (\Slb)^T\Slb = \rho^2 (c \delta + g) *
+  \fl * \fl *,
+\]
+in which
+\[
+  c = 2a \partial t b(0) - b(0)^2)
+\]
+and
+\[
+  g = a(\partial^2_t b H + \partial^2_t \check{b}\check{H}) -
+  b((0)((\partial_t \check{b}) \check{H} + (\partial_t b) H +
+  ((\partial_t \check{b}) \check{H})*((\partial_t b)H)) 
+\]
+According to Young's inequality and the bounds \ref{eqn:best}, $g$ is integrable with a 
+$L^1(X_{sr}\times \bR)$ bound independent of $\log \kappa \in
+M$. Therefore from the bounds \ref{eqn:lambdanorms}, obtain
+\begin{equation}
+  \label{eqn:step2}
+\|(\Sla)^T\Slb + (\Slb)^T\Sla + (\Slb)^T\Slb\} \le C \lambda^3
+\end{equation}
+From the bounds \ref{eqn:unorm} and \ref{eqn:lambdanorms},
+\[
+  \|\ul\| \le C \lambda^{-\frac{1}{2}}\|\dl\|
+\]
+Combining the preceding results in the first estimate
+\ref{eqn:diffests}.
+
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%
+
+over in the expression for $g$ is exponentially decaying, hence
 %And that's ``why it works'' - when geometric asymptotics in the form
@@
   \|T(\Slb)^T\dl\| \le C \lambda^{3/2}\|\dl\|.
+\end{equation}
+
+On the other hand,
+\[
+  T(\dl - \dzl) = T(b^*H*\delta_{\tau^*}*\wl = T(b^*(0)\delta +
+  \partial_t b^*H)*\delta_{\tau^*} *\fl
+\]
+\[
+  (T\partial_tb^*H)*\delta_{\tau^*}*\fl+(b^*(0)\delta +
+  \partial_t b^*H)\tau^*\delta_{\tau^*}*\fl +( b^*(0)\delta +
+  \partial_t b^*H)*\delta_{\tau^*}T\fl
+\]
+From the second relation \ref{eqn:lambdanorms},
+\begin{equation}
+  \label{eqn:step7.5}
+  \|T(\dl-\dzl)\| \le C \lambda
+\end{equation}
+in which $C$ depends on $M$ and on the support of $f_1$ (which
+contains the support of $\fl$). The second bound \ref{eqn:rembd}
+gives
+\begin{equation}
+  \label{step8}
+  \Slb T(\dl-\dzl) \le C\lambda^{\frac{3}{2}}
+\end{equation}
+Also,
+\[
+  [T,\Slb] = ((Tb)H) *\wl * \delta_{\tau}* + (bH) * T\wl *
+delta_{\tau} * + \tau (bH)* \wl * \delta_{\tau}*
+\]
+\[
+  = ((\partial_t Tb) H * \delta_{\tau}*\fl* +  \tau(b(0)\delta + \partial_t bH)*\fl*\delta_{\tau}*
+\]
+\[
+  +( b(0)\delta + \partial_t bH)*T\fl*\delta_{\tau} -
+  Bh*\delta_{\tau}*\fl*
+\]
+SInce $\|\fl\|_{L^1(\bR)} =\lambda^{\frac{3}{2}} \|f_1\|_{L^1(\bR)}$ and
+$\| T\fl\|_{L^1(\bR)} = \lambda^{\frac{5}{2}} \|f_1\|_{L^1(\bR)}$, every term in
+the preceding expression is $O(\lambda^{\frac{3}{2}})$. Since $T$ is symmetric,
+\begin{equation}
+  \label{eqn:step9}
+  \|[T,\Slb]\| = \|[T,(\Slb)^T]\| \le C\lambda^{\frac{3}{2}}.
+\end{equation}
+Note that by reasoning almost identical to that leading to the bound
+\ref{eqn:step7.5},
+\begin{equation}
+  \label{eqn:step9.5}
+  \|\dl-\dzl\| \le C\lambda.
+\end{equation}
+so
+\begin{equation}
+  \label{eqn:step10}
+  \|T (\Slb)^T(\dl-\dzl)\| \le \|[ T, (\Slb)^T](\dl-\dzl)\| +
+  \|(Slb)^TT (\dl-\dzl)\| \le C\lambda^{\frac{5}{2}}\|\dl\|
 \end{equation}

### 2024-10-29 repro/papers/awi/paper.tex
near final

@@
 Various modifications of straighforward best-fit FWI have been
-suggested to overcome this ``cycle-skipping'' pathology. One of these
-is Adaptive Waveform Inversion \cite[]{Warner:16} (AWI). This
-technique has been successfully applied to industry-scale seismic data
+suggested to overcome this ``cycle-skipping'' pathology. Many of these
+modifications employ {\em model extension}, that is, the introduction
+of a larger space of model parameters than derives from basic
+physics. An example is the {\em source-receiver} extension \cite[]{HuangSymes:17}, in which
+source waveforms are permitted to depend (unphysically) on receiver
+locations. The focus of this paper is Adaptive Waveform Inversion
+\cite[]{Warner:16} (AWI), a modification of FWI using the
+source-receiver extension. AWI has been successfully applied to industry-scale seismic data
 \cite[]{GuaschWarnerRavaut:GEO19,Warneretal:SEG21} and shows promise for
@@
 literature \cite[]{Warner:16,LiAlkhalifah:21}. Another modification of
-FWI based on similar ideas, Matched Source Waveform Inversion (MSWI)
-(\cite{HuangSymes2015SEG,HuangSymes:Geo17}) lacks the AWI
+FWI based on source-receiver extension, Matched Source Waveform Inversion (MSWI)
+\cite[]{HuangSymes2015SEG,HuangSymes:Geo17} lacks the AWI
 normalization, and has a looser relation to travel time inversion. The
@@
   uniformly in any bounded spatial domain.
+\item only one ray of geometric
+  acoustics connects each source-receiver location pair (single arrivals)
 \end{itemize}
+
 These constraints are discussed in detail in the next section. Some
@@
 can be treated as a perturbation of the infinite time data setting
-  posed here. However these hypotheses enable a simple and clear
+posed here. The exception is the single-arrival hypothesis: it is
+essential, as without it, the link between the travel time and
+extended waveform inversion is lost. This is a feature of the
+source-receiver extension, not of a particular objective function
+(MSWI, AWI,...) formulated in terms of it. This extension treats each receiver signal
+independently. If only a single identifiable waveform arrives at
+each receiver, AWI (and similar approaches) can implicitly identify
+its arrival time, and the link to travel time inversion is
+established. If distinct waveforms are present at multiple times,
+the single receiver signal does not determine which waveform in the
+predicted signal should align with which waveform in the data to be
+fitted. This ``crosstalk'' ambiguity is analyzed in the fifth
+section, and afflicts all approaches based on the source-receivre
+extension.
+
+In any case, these hypotheses enable a simple and clear
 analysis of AWI, and seem a reasonable starting point for further
@@
 functions. In section four, I explain the relation between AWI and a
-penalty formulation of the wave inverse problem. The
-close relation between AWI objective and travel time misfit is limited
-in scope: in the fifth section, I show that it does not hold in
-general for transmitted wave data in which multiple rays connect
-source and receiver.
+penalty formulation of the wave inverse problem. The close relation
+between AWI objective and travel time misfit is limited in scope: in
+the fifth section, I show that it does not hold in general for
+transmitted wave data in which multiple rays connect source and
+receiver.
 
@@
 Singularities Theorem (\cite{Tay:81}, Theorem VI.2.1) and
-the pull-back property of distributions \cite{Dui:95}, Proposition 1.3.3)
+the pull-back property of distributions (\cite{Dui:95}, Proposition 1.3.3)
 imply that if $f \in \cal{E}'(\bR)$ and $(\bx_s,\bx_r) \in X_{sr}$,
@@
 \item[H1. ] $\tau[\kappa], a[\kappa],$ and $b[\kappa]$ are smooth off the diagonal,
-that is, $\tau[\kappa](\bx_s,\cdot) \in C^{\infty}(\Omega \setminus \{\bx_s\})$
-and similarly for $a[\kappa],b[\kappa]$;
+  that is,
+  (i) for each $\bx_s \in X_s$, $\tau[\kappa](\bx_s,\cdot) \in C^{\infty}(\Omega \setminus
+  \{\bx_s\})$, and similarly for $a[\kappa],b[\kappa]$; $\tau$
+  satisfies the eikonal equation
+  \[
+    \sqrt{\frac{\kappa}{\rho}} |\nabla \tau(\bx_s,\cdot)| = 1 \mbox{ in } \Omega \setminus
+    \{\bx_s\}, \, \lim_{\bx \rightarrow \bx_s} \tau(\bx_s,\bx) = 0.
+  \]
 \item[H2. ] there exists $A>0$ so that for any $\log \kappa \in M$
@@
 
+Generally, constants denoted $C$ etc. will absorb any dimensional
+scale factors necessary to make sense of the expressions in which they
+appear. I will call out those few cases in which explicit use of
+dimensinoal scale factors is necessary to avoid irretrievable
+dimensional confusion.
+
 \subsection{FWI via Least Squares}
@@
 The extended modeling operator ${\bar F}$ maps extended models
-$\bar{m}=(\kappa,\bar{w})$ incorporating extended sources
-$\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
+$\bar{m}=(\kappa,\bar{f})$ incorporating extended sources
+$\bar{f}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
 field. Denote by $\bar{p}(\bx,t;\bx_s,\bx_r), \bar{\bv}(\bx,t;\bx_s,\bx_r)$
-the solution of the acoustic system \ref{eqn:awe} with $w(t)$ in the
+the solution of the acoustic system \ref{eqn:awe} with $f(t)$ in the
 right-hand side of the first equation replaced by
-$\bar{w}(\bx_r,\cdot;\bx_s)$. Then
-$\bar{F}[\kappa,\bar{w}](\bx_r,t;\bx_s) =
+$\bar{f}(\bx_r,\cdot;\bx_s)$. Then
+$\bar{F}[\kappa,\bar{f}](\bx_r,t;\bx_s) =
 \bar{p}(\bx_r,t;\bx_s,\bx_r)$. As was the case for $F$, the Hadamard
@@
 \rightarrow L^2(X_{sr} \times \bR)$. If all  
-of the wavelets are the same, that is, ${\bar w}({\bf  
-  x}_r,t;{\bf x}_s) = w(t)$ is independent of source and receiver  
-position, then ${\bar F}[\kappa,\bar{w}] = F[\kappa]$. That is, $F$ is a special
+of the wavelets are the same, that is, ${\bar f}({\bf  
+  x}_r,t;{\bf x}_s) = f(t)$ is independent of source and receiver  
+position, then ${\bar F}[\kappa,\bar{f}] = F[\kappa]$. That is, $F$ is a special
 case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
@@
 
-\cite{Warner:16} represent the extended sources $\bar{w}$ as time convolutions of the
-(known) source $w$ with an {\em adaptive filter} $u \in L^2(X_{sr}
-\times \bR)$, that is, $\bar{w}=u * w$.
-Here, $u*w$ denotes the result of convolving each trace in $u$ by $w
-\in L^2(\bR)$. I
+\cite{Warner:16} represent the extended sources $\bar{f}$ as time convolutions of the
+(known) source $f$ with an {\em adaptive filter} $u \in L^2(X_{sr}
+\times \bR)$, that is, $\bar{f}=u * f$.
+Here, $u*f$ denotes the result of convolving each trace in $u$ by $f$. I
 will also use the asterisk to denote the trace-by-trace convolution
@@
 \[
-  \bar{F}[\kappa,u*w] = u*F[\kappa]
+  \bar{F}[\kappa,u*f] = u*F[\kappa]
 \]
@@
 \end{equation}
-Then $S[\kappa]$ is bounded on $L^2(X_{sr} \times \bR)$, uniformly in
-$\log \kappa \in M$, thanks to
-the bound \ref{eqn:fwdnorm} for $p=1$. $S$ also represents an
+$S$ represents an
 extension of $F$, in the sense that with the choice
@@
 observation motivates the definition of the AWI objective function
-(\cite{Warner:16}, equation XXXX)::
+(\cite{Warner:16}, equation 4)::
 \begin{equation}
   \label{eqn:awideforig}
-  J_{\rm AWI}[m;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu[\kappa;d](\bx_r,t;\bx_s)|^2}{\int dt |u[\kappa;d](\bx_r,t;\bx_s)|^2}\right) 
+  J_{\rm AWI}[\kappa;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu[\kappa;d](\bx_r,t;\bx_s)|^2}{\int dt |u[\kappa;d](\bx_r,t;\bx_s)|^2}\right) 
 \end{equation}
@@
   \label{eqn:mswideforig}
-  J_{\rm MSWI}[m;d] = \sum_{\bx_s,\bx_r} \int dt |tu[\kappa;d](\bx_r,t;\bx_s)|^2
+  J_{\rm MSWI}[\kappa;d] = \sum_{\bx_s,\bx_r} \int dt |tu[\kappa;d](\bx_r,t;\bx_s)|^2
 \end{equation}
@@
 In fact, these two defects are really symptoms of the instability of
-deconvolution for kernels with the properties of $F[\kappa]$.  To
-begin with, $S[\kappa] = F[\kappa]*$ defines a bounded map on
-$L^2(X_{sr}\times \bR)$, since from \ref{eqn:fwdgreen} and the
+deconvolution for kernels with the properties of $F[\kappa]$.  From \ref{eqn:fwdgreen} and the
 definition \ref{eqn:sdef} of  $S[\kappa]$,
@@
   (b[\kappa](0)\delta + (\partial_t b[\kappa])H)
-  *\delta_{\tau[\kappa]}*f) * u
-\end{equation}
+  *\delta_{\tau[\kappa]}*f) * u.
+\end{equation}
+Since $\partial_t f$ is even, the formal adjoint $S[\kappa]^T$ is given by
+\begin{equation}
+  \label{eqn:sadjgood}
+  S[\kappa]^Tu = \rho( a[\kappa]\delta    +  (\check{b}[\kappa]\check{H}))
+  *\delta_{-\tau[\kappa]}*\partial_t f *u,
+\end{equation}
+in which $\check{u}(t) = u(-t)$.
 The key properties of $S$ follow directly from Young's inequality and
-the representation \ref{eqn:fwdgreen}:
+the representation \ref{eqn:sgood}:
 \begin{lemma}
   \label{thm:snorm}
-  Under the assumptions leading to \ref{eqn:fwdgreen} and
-  the properties F1 and F2, $S[\kappa]$ is a bounded operator on
+  Under the assumptions G1-G4 leading to \ref{eqn:fwdgreen} and
+  the properties F1 and F2, the expressions \ref{eqn:sgood} and \ref{eqn:sadjgood} define 
+  bounded operators on
+  $\Pi_{X_{sr}} {\cal S}(\bR)$ and on
   $L^2(X_{sr} \times \bR)$, and there exists $C>0$ so
@@
 \end{lemma}
+\begin{proof} From the estimates H3, $b[\kappa](\bx_r,\cdot;\bx_s)H *
+  \partial_tf \in {\cal S}(\bR)$ for all $(\bx_s,\bx_r) \in X_{sr}$,
+  whence the first conclusion. Bounded extension to $L^2(X_{sr} \times
+  \bR)$ follows from Young's inequality.
+\end{proof}
 
@@
 \end{equation}
+For future reference, examine the
+implications of definition \ref{eqn:reg} for the units of the regularization weight $\sigma$,
+which  is
+necessarily dimensional. Since the data $d$ has dimensions of
+pressure, so must $S[m]u = F[m]*u$. Of course $F[m]$ is also sampled
+pressure, and convolution is a $t$ integration, so for $S[m]u$ to be a
+pressure, $u$ must have dimensions of 1/time. Therefore
+$\|S[m]u-d\|^2$ has dimensions of pressure$^2 \times $time, $\|u\|^2$ has
+dimensions of 1/time, whence $\sigma$ must be dimensioned as
+pressure$^2 \times$  time$^2$.
 
@@
 
-This definition has introduced a new defect: while the denominator is
-well-defined for $\log \kappa \in M$, $d \times L^2(X_{sr} \times
-\bR)$ for $\sigma > 0$, the numerator may not be, since there is no
-guarantee that $Tu_{\sigma}[\kappa;d] \in L^2(X_{sr} \times
-\bR)$.
+The definitions \ref{eqn:awidefbis} and \ref{eqn:mswidef} have introduced a new defect: there is no
+guarantee that $Tu_{\sigma}[\kappa;d] \in L^2(X_{sr} \times\bR)$.
 
@@
     = \rho^2((a\delta  +  (\check{b}\check{H}))*(a\delta  +
-    (bH))*\partial_t f *\partial_t f*
-  \]
+    (bH))*\partial_t f *\partial_t f*.
+  \]
+  $T$ is also a bounded map on $\Pi_{X_{sr}} {\cal S}(\bR)$, as is
+  its commutator with $S^TS$, and 
   \[
@@
   observe that all summands on the RHS of identity \ref{eqn:tsscomm} are in
-  $L^1(X_{sr} \times \bR)$, so $[T,S^TS]$ restricts to a bounded
+  $L^1(X_{sr} \times \bR)$, so $[T,S^TS]$ extends to a bounded
   operator on $L^2(X_{sr}\times \bR)$, with
@@
 
-  Multiplying the normal equation for $u_{\sigma}$Rearranging,
+  Multiplying the normal equation for $u_{\sigma}$ by $T$ and earranging,
   \[
@@
   \label{eqn:Fl1}
-  =\rho (b_0[\kappa]\delta + (\partial_t b H) * f_{\lambda} )*\delta_{\tau[\kappa]}.
+  =\rho (b_0[\kappa]\delta + (\partial_t b H)) * f_{\lambda} )*\delta_{\tau[\kappa]}.
 \end{equation}
@@
 
-\subsection{Linking Regularization and Wavelength}
+\subsection{Linking Regularization and Wavelength; Statement of the
+  Main Result}
 In the remainder of this section, I will examine the $\lambda$-parametrized families of
@@
 \begin{itemize}
-\item[1. ]replacing $S$ and $d$ in the definitions \ref{eqn:awidefbis}
-  and \ref{eqn:mswidef} with $\Sl$ and $\dl =\Fl[\kappa^*]$, that is,
-  noise-free data generated by a choice of $\kappa^* \in M$ and
-  $\lambda \in (0,1]$, and
-\item[2. ] choosing the Tihonov regularization weight
-  $\sigma = \lambda$.
+\item[1. ] choosing the Tihonov regularization weight in the least
+  square problem \ref{eqn:reg} as $\sigma = r\lambda$, in which $r$ is
+  a positive constant with dimensions pressure$^2 \times $ time$^2$.
+\item[2. ]replacing $S$, $d$ with $\Sl$, $\dl =\Fl[\kappa^*]$
+  (noise-free data) in the definition of $\ul$ (equation \ref{eqn:reg}).
 \end{itemize}
-That is,
-\begin{equation}
-  \label{eqn:mswideflam}
-  J_{\rm MSWI,\lambda}[\kappa;\dl] =  \|T\ul[\kappa;\dl]\|^2
-\end{equation}
-and 
-\begin{equation}
-  \label{eqn:awideflam}
-  J_{\rm AWI,\lambda}[\kappa;\dl] = \sum_{(\bx_s,\bx_r) \in X_{sr}} 
-  \frac{\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|^2}{\|\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|^2}.
-\end{equation}
-In both definitions, $\ul$ solves the least-squares problem \ref{eqn:reg} with
-$\sigma = \lambda$.
-
-Justification for choosing the regularization weight $\sigma$ to be the same as
-the wavelength proxy $\lambda$ will emerge from the analysis relating these
-objective functions to the conventional objective function of travel
-time tomography. It will be clear that choosing $\sigma$ a function of
+
+Justification for choosing the regularization weight $\sigma$ to be
+proportional to
+the wavelength proxy $\lambda$ will emerge from the analysis relating
+$\|T\ul[\kappa;\dl]\|$ to the travel
+time error $|\tau[\kappa]-\tau[\kappa^*]$. The necessary units for the
+proportionality constant $r$ follow from the discussion after equation
+\ref{eqn:reg}. It will be clear that choosing $\sigma$ a function of
 $\lambda$ roughly proportional to $\lambda$, that is, $C_*\lambda \le
@@
 C^*$, would lead to the same asymptotic estimates for the relation
-between $J_{\rm MSWI,\lambda}$,$ J_{\rm AWI,\lambda} $ and travel time
-error. The argument is slightly cleaner if simply $\sigma=\lambda$.
+between the adaptive filter $\ul$ and travel time
+error. The argument is slightly cleaner if $\sigma$ is simply
+proportional to (non-dimensional) $\lambda$ by a necessarily
+dimensional constant..
+
+The main result of this investigation is the
+
+\begin{theorem}
+  \label{thm:main}
+  Assume assertions G1 - G4 on $M$ and $X_{sr}$, W1 - W3 on the
+  ``mother wavelet'' $w_1$. Select $r >0$ with units of
+  pressure$^2 \times$time$^2$ and $\log \kappa^* \in M$. For $0 <
+  \lambda \le 1$, set
+  $\dl=\Fl[\kappa^*]$ (noise-free data), and  define $\ul[\kappa,\dl]$ by
+  equation \ref{eqn:reg} with  $\sigma=r\lambda$, $S=\Sl$ and $d=\dl$..
+  There exist $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
+  \rightarrow \bR^+$ with $C_* \le
+  W \le C^*$ so that for
+  $\log \kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$,, 
+  \begin{equation}
+    \label{eqn:mswitt}
+    |\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|- \lambda^{-\frac{1}{2}} (W[\kappa])^{\frac{1}{2}}|\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|| \le C
+  \lambda^{\frac{1}{2}}
+\end{equation}
+and for sufficiently small $\lambda$, 
+\begin{equation}
+  \label{eqn:awitt}
+  \left|\frac{\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|}{\|\ul[\kappa,\dl](\bx_r,\cdot;\bx_s)\|}-
+    |\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|\right| \le C \lambda
+  \end{equation}
+\end{theorem}
+
+Sum over the source-receiver positions in $X_{sr}$ to obtain
+\begin{cor}
+  \label{thm:cor1}
+  In addition to the assumptions of Theorem \ref{thm:main}, define $J_{\rm
+    AWI,\sigma}$ and $J_{\rm MSWI,\sigma}$ by equations
+  \ref{eqn:awidefbis}, \ref{eqn:mswidef} respectively. Then
+\begin{equation}
+\label{eqn:cor1mswi}   
+\lim_{\lambda \rightarrow 0} \lambda J_{\rm MSWI,r\lambda}[\kappa;\dl]
+= \sum_{(\bx_s,\bx_r) \in X_{sr}} M[m,m^*,w_1]^2(\bx_s,\bx_r) |\tau[m](\bx_s,\bx_r) - \tau[m^*](\bx_s,\bx_r)|^2.
+\end{equation}
+and
+\begin{equation}
+\label{eqn:cor1awi}   
+\lim_{\lambda \rightarrow 0} J_{\rm AWI,r\lambda}[\kappa;\dl]
+= \sum_{(\bx_s,\bx_r) \in X_{sr}}  |\tau[m](\bx_s,\bx_r) - \tau[m^*](\bx_s,\bx_r)|^2.
+\end{equation}
+\end{cor}
 
@@
 \begin{prop}
-In the definitions of $J_{\rm MSWI,\sigma}$ and $J_{\rm AWI,\sigma}$, replace the data $d$ with
-$\dla = \Fla[\kappa^*]$, $S[\kappa]$ with $\Sla[\kappa]$. That is,
+  \label{thm:leading}
+In the definitions \ref{eqn:mswidef} of $J_{\rm MSWI,\sigma}$ and
+\ref{eqn:awidefbis} of $J_{\rm AWI,\sigma}$, set $\sigma = r\lambda$
+and replace the data $\dl$ with
+$\dla = \Fla[\kappa^*]$, $S[\kappa]$ with $\Sla[\kappa]$, and $\ul$
+with the solution $\uzl$ of the corresponding least squares problem. That is,
 \begin{eqnarray}
@@
              \nonumber \\
-  \Sla[\kappa]u &=& a[\kappa]\delta_{\tau[\kappa^*]}*\wl *u\nonumber \\
-  \uzl &=& \mbox{argmin}_{u \in L^2(X_{sr} \times \bR)} (\|\Sla[\kappa]u-\dla\|^2 + \sigma\|u\|^2)
+  \Sla[\kappa]u &=& a[\kappa]\delta_{\tau[\kappa^*]}*\wl *u\nonumber
+  \\
+  \uzl &=& \mbox{argmin}_{u \in L^2(X_{sr} \times \bR)} (\|\Sla[\kappa]u-\dla\|^2 + r\lambda\|u\|^2)
 \end{eqnarray}
+%and
+%\begin{eqnarray}
+%  \label{eqn:awileading}
+%  J^0_{\rm MSWI,\lambda}[\kappa;\dla] & = &
+%                                         \|T\uzl[\kappa;\dla]\|^2\nonumber\\
+%  J^0_{\rm AWI,\lambda}[\kappa;\dla] & = &
+%                                       \sum_{(\bx_s,\bx_r) \in X_{sr}} \frac{\|T\uzl[\kappa;\dla](\bx_r,\cdot;\bx_s)\|^2}{\|\uzl[%\kappa;\dla](\bx_r,\cdot;\bx_s)\|^2}
+%\end{eqnarray}                 
+Then there exists $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
+\rightarrow \bR^+$ so that for
+$\kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$, $C_* \le
+W[\kappa](\bx_s,\bx_r) \le C^*$, and $\uzl$ defined in display
+\ref{eqn:leading} for $0 < \lambda \le 1$, 
+\begin{equation}
+  \label{eqn:mswittz}
+ | \|T\uzl[\kappa;\dla] (\bx_r,\cdot;\bx_r)\|^2- 
+  \frac{1}{\lambda} W[\kappa](\bx_s,\bx_r) (\tau[\kappa^*]({\bf
+  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 | \le C\lambda,
+\end{equation}
 and
-\begin{eqnarray}
-  \label{eqn:awileading}
-  J^0_{\rm MSWI,\lambda}[\kappa;\dla] & = &
-                                         \|T\uzl[\kappa;\dla]\|^2\nonumber\\
-  J^0_{\rm AWI,\lambda}[\kappa;\dla] & = &
-                                       \sum_{(\bx_s,\bx_r) \in X_{sr}} \frac{\|T\uzl[\kappa;\dla](\bx_r,\cdot;\bx_s)\|^2}{\|\uzl[\kappa;\dla](\bx_r,\cdot;\bx_s)\|^2}
-\end{eqnarray}                 
-Then there exists $C>0$ so that for
-$\kappa \in M$, $0 < \lambda \le 1$, 
-\begin{equation}
-  \label{eqn:mswitt}
- | J^0_{\rm MSWI,\lambda}[\kappa;\dla] - 
-  \frac{1}{\lambda} \sum_{{\bf x}_s,{\bf x}_r} \|u^0_1[\kappa](\bx_r,\cdot;\bx_s)\|^2(\tau[\kappa^*]({\bf
-  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 | \le C\lambda,
-\end{equation}
-and
-\begin{equation}
-  \label{eqn:awitt}
-|J^0_{\rm AWI,\lambda}[\kappa;\dla] - \sum_{{\bf x}_s,{\bf x}_r} (\tau[\kappa^*]({\bf
-  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2|  \le C\lambda^2.
-\end{equation}
-\end{prop}
+\begin{equation}
+  \label{eqn:awittz}
+  \left| \frac{\|T\uzl[\kappa;\dla] (\bx_r,\cdot;\bx_r)\|^2}{
+        \|\uzl[\kappa;\dla] (\bx_r,\cdot;\bx_r)\|^2} - (\tau[\kappa^*]({\bf
+  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2\right|  \le C\lambda^2.
+\end{equation}
+m\end{prop}
 
@@
   The abbreviations
-$a^*=a[\kappa^*], a=a[\kappa], \tau^*=\tau[\kappa^*], \tau=\tau[\kappa]$.
-will be convenient in the computations to follow.
-
-The right-hand side of the third equation in \ref{eqn:leading} is then
-\[
-= \mbox{argmin}_u (\|a \wl*u*\delta_{\tau}-a^*\wl*\delta_{\tau^*}\|^2 + \lambda\|u\|^2)
+  $a^*=a[\kappa^*](\bx_s,\bx_r), a=a[\kappa](\bx_s,\bx_r),
+  \tau^*=\tau[\kappa^*](\bx_s,\bx_r), \tau=\tau[\kappa](\bx_s,\bx_r)$.
+  will be convenient in the computations to follow. In particular, the
+  source/receiver coordinates $\bx_r,\bx_r$ will be suppressed from
+  the notation. Each identity and estimate to follow refers to every
+  pair $(\bx_s,\bx_r) \in X_{sr}$, that is, trace-by-trace, unless
+  expllicitly indiccated otherwise.,
+
+The right-hand side of the third equation in \ref{eqn:leading} is 
+\[
+= \mbox{argmin}_u (\|a \wl*u*\delta_{\tau}-a^*\wl*\delta_{\tau^*}\|^2 + r\lambda\|u\|^2)
 \]
@@
   \label{eqn:leadingft}
-= \mbox{argmin}_u \frac{1}{2\pi}(\|a \fwl \fu e^{-i\omega \tau}-a^*\fwl e^{-i\omega \tau^*}\|^2 + \lambda\|\fu\|^2).
+= \mbox{argmin}_u \frac{1}{2\pi}(\|a \fwl \fu e^{-i\omega \tau}-a^*\fwl e^{-i\omega \tau^*}\|^2 + r\lambda\|\fu\|^2).
 \end{equation}
@@
 \[
-(a^2 |\fwl|^2 +\lambda)\hat{u} = a a^*|\fwl|^2e^{i\omega(\tau^*-\tau)}
+(a^2 |\fwl|^2 +r\lambda)\hat{u} = a a^*|\fwl|^2e^{i\omega(\tau^*-\tau)}
 \]
@@
   \label{eqn:hatuf}
-\hat{u}_{0,\lambda}[\kappa;\dla] = \frac{a^*}{a}\hat{g}_{\lambda,\frac{\lambda}{a^2}} e^{i\omega(\tau^*-\tau)}
+\hat{u}_{0,\lambda}[\kappa;\dla] = \frac{a^*}{a}\hat{g}_{\lambda,\frac{r\lambda}{a^2}} e^{i\omega(\tau^*-\tau)}
 \end{equation}
@@
   \label{eqn:hatgf}
-\hat{g}_{\lambda,\lambda} = \frac{|\fwl|^2}{|\fwl|^2 + \lambda}.
-\end{equation}
-Since $\wl$ vanishes for large $|t|$, $\fwl$ is entire analytic hence
-vanishes on a set of measure zero, whence $\hat{g}_{\lambda,\lambda}:$
+\hat{g}_{\lambda,\mu} = \frac{|\fwl|^2}{|\fwl|^2 + \mu}.
+\end{equation}
+for $0 < \lambda \le 1, 0 < \mu$. Since $\wl$ vanishes for large $|t|$, $\fwl$ is entire analytic hence
+vanishes on a set of measure zero, whence $\hat{g}_{\lambda,\frac{r\lambda}{a^2}}:$
 tends to $1$ almost everywhere as $\lambda \rightarrow 0$. Hence the
-inverse Fourier transform $g_{\lambda,\lambda}$ tends to $\delta$ in
+inverse Fourier transform $g_{\lambda,\frac{r\lambda}{a^2}}$ tends to $\delta$ in
 the sense of distributions, and
@@
   \label{eqn:u0s}
-\uzl[\kappa,\dla](t) = \frac{a^*}{a}g_{\lambda,\frac{\lambda}{a^2}}(t-(\tau^*-\tau))
+\uzl[\kappa,\dla](t) = \frac{a^*}{a}g_{\lambda,\frac{r\lambda}{a^2}}(t-(\tau^*-\tau))
 \end{equation}
@@
 L^2(\bR)$.
-Therefore $\hat{g}_{\lambda,\lambda}$
-and $\hat{g}_{\lambda,\lambda}'$ are also square-integrable, 
-from equation \ref{eqn:hatuf} $\hat{u}_{0,\lambda}'$ is also square
-integrable. Consequently $t u_{0\lambda}$ is square integrableTherefore
-$J_{\rm MSWI,\lambda,\lambda}$, the result of substituting in the
+Therefore $\hat{g}_{\lambda,\frac{r\lambda}{a^2}}$
+and $\hat{g}_{\lambda,\frac{r\lambda}{a^2}}'$ are also square-integrable, 
+from equation \ref{eqn:hatuf} $(\hat{u}^0_{\lambda})'$ is also square
+integrable. Consequently $t \uzl$ is square integrable. Therefore
+$J_{\rm MSWI,r\lambda}$, the result of substituting in the
 right-hand side of the definition \ref{eqn:mswidef} $\uzl$ for
-$u_{\lambda}$, is well-defined. \cite{HuangSymes2015SEG} presented an analysis of
-$J_{\rm MSWI,\lambda,\lambda}$, as follows:
-\[
-J_{\rm MSWI,\lambda,\lambda}[\kappa;\dla] = \sum_{\bx_s,\bx_r}\int dt
-|t\uzl[\kappa;\dla]|^2 =  \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{\lambda}{a^2}}(t-(\tau^*-\tau))|^2
-\]
-\[
-=\sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, (t+(\tau^*-\tau))^2|g_{\lambda,\frac{\lambda}{a^2}}(t)|^2
+$\ul$, is well-defined. \cite{HuangSymes2015SEG} presented an analysis of
+$J_{\rm MSWI,r\lambda}$, as follows:
+\[
+J_{\rm MSWI,r\lambda}[\kappa;\dla] = \sum_{\bx_s,\bx_r}\int dt
+|t\uzl[\kappa;\dla]|^2 =  \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{r\lambda}{a^2}}(t-(\tau^*-\tau))|^2
+\]
+\[
+=\sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, (t+(\tau^*-\tau))^2|g_{\lambda,\frac{r\lambda}{a^2}}(t)|^2
 \]
@@
   =\sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, (t^2 + 2t
-  (\tau^*-\tau)+(\tau^*-\tau)^2)|g_{\lambda,\frac{\lambda}{a^2}}(t)|^2
+  (\tau^*-\tau)+(\tau^*-\tau)^2)|g_{\lambda,\frac{r\lambda}{a^2}}(t)|^2
 \end{equation}
@@
 Since the Fourier transform $\fwl$ is even and real, so is
-$\hat{g}_{\lambda,\lambda}$, therefore $g_{\lambda,\lambda}$ is even, and the linear term (in t) on the
+$\hat{g}_{\lambda,\frac{r\lambda}{a^2}}$, therefore $g_{\lambda,\frac{r\lambda}{a^2}}$ is even, and the linear term (in t) on the
 right-hand side of \ref{eqn:MSWI} vanishes, So
@@
   \label{eqn:MSWIasym}
-  J_{\rm MSWI,\lambda,\lambda}[\kappa;\dla] =
-\sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{\lambda}{a^2}}(t)|^2
-+(\tau^*-\tau)^2\frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{\lambda}{a^2}}\|^2
+  J^0_{\rm MSWI,r\lambda}[\kappa;\dla] =
+\sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{r\lambda}{a^2}}(t)|^2
++(\tau^*-\tau)^2\frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{r\lambda}{a^2}}\|^2
 \end{equation}
@@
   \label{eqn:MSWIasym2}
-  J_{\rm MSWI,\lambda,\lambda}[\kappa;\dla] = 
-  \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{\lambda}{a^2}}\|^2\left(l(g_{\lambda,\frac{\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
-\end{equation}
-
-The second term suggests that this function
-is related to the travel-time error $\tau^*-\tau$. 
-The relation is obscured by the presence of of the amplitudes
-$a^*, a$ both in the multiplier and in the scaling of $\lambda$. Since the
-amplitudes depend on $\kappa$, their presence may influence the location
-(or even the presence) of stationary points of $J_{\rm
-  MSWI,\lambda,\lambda}$ (or $J_{\rm MSWI}$, to which it is an approximation)  for
-models differing in their travel time predictions. So stationary
-points of the MSWI objective may not yield slowness models that match
-travel-times with the data, even in the noise-free case (though
-the examples reported by \cite{HuangSymes2015SEG, HuangSymes:Geo17}, and other works
-cited there, suggest that sometimes this is the case).
-
-The trace-by-trace normalization in the definition of
-the AWI objective removes the model-dependent factors. Denote by
-$J_{\rm AWI, \lambda,\lambda}$ the restult of subtituting
-$\uzl$ for $u_{\lambda}$ in the definition \ref{eqn:awidef} of
-$J_{\rm AWI}$:
-\begin{equation}
-  \label{eqn:AWIasym}
-J_{\rm AWI,\lambda,\lambda}[\kappa;\dla] = \sum_{\bx_s,\bx_r}\frac{\int dt
-(|t\uzl[\kappa;\dla])^2}{\|\uzl[\kappa;\dla]\|^2} 
-\end{equation}
-We have already analyzed the numerator, arriving at
-equation \ref{eqn:MSWIasym2}. From the result \ref{eqn:u0s}, for each
+  J^0_{\rm MSWI,\lambda}[\kappa;\dla] = 
+  \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{r\lambda}{a^2}}\|^2\left(l(g_{\lambda,\frac{r\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
+\end{equation}
+
+%The second term suggests that this function
+%5is related to the travel-time error $\tau^*-\tau$. 
+%The relation is obscured by the presence of of the amplitudes
+%$a^*, a$ both in the multiplier and in the scaling of $\lambda$. Since the
+%amplitudes depend on $\kappa$, their presence may influence the location
+%(or even the presence) of stationary points of $J_{\rm
+% MSWI,\lambda,\lambda}$ (or $J_{\rm MSWI}$, to which it is an approximation)  for
+%models differing in their travel time predictions. So stationary
+%points of the MSWI objective may not yield slowness models that match
+%travel-times with the data, even in the noise-free case (though
+%the examples reported by \cite{HuangSymes2015SEG, HuangSymes:Geo17}, and other works
+%cited there, suggest that sometimes this is the case).
+
+%The trace-by-trace normalization in the definition of
+%the AWI objective removes the model-dependent factors. Denote by
+%$J_{\rm AWI, \lambda,\lambda}$ the restult of subtituting
+%$\uzl$ for $u_{\lambda}$ in the definition \ref{eqn:awidef} of
+%$J_{\rm AWI}$:
+%\begin{equation}
+%  \label{eqn:AWIasym}
+%J^0_{\rm AWI,\lambda}[\kappa;\dla] = \sum_{\bx_s,\bx_r}\frac{\int dt
+%(|t\uzl[\kappa;\dla])^2}{\|\uzl[\kappa;\dla]\|^2} 
+%\end{equation}
+%We have already analyzed the numerator, arriving at
+%equation \ref{eqn:MSWIasym2}.
+From the result \ref{eqn:u0s}, for each
 $(\bx_s,\bx_r) \in X_{sr}$,
@@
   \|\uzl[\kappa;\dla]\|^2=
-  \frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{\lambda}{a^2}}\|^2.
-\end{equation}
-Combine equations \ref{eqn:AWIasym} and \ref{eqn:u0snorm} to obtain
+  \frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{r\lambda}{a^2}}\|^2.
+\end{equation}
+Combine equations \ref{eqn:awittz} and \ref{eqn:u0snorm} to obtain
 \begin{equation}
   \label{eqn:AWIasym2}
-  J_{\rm AWI,\lambda,\lambda}[\kappa;\dla] = \sum_{\bx_s,\bx_r} \left(l(g_{\lambda,\frac{\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
+  \frac{\|T\uzl\|^2}{\|\uzl\|^2}= \sum_{\bx_s,\bx_r} \left(l(g_{\lambda,\frac{r\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
  \end{equation}
@@
 travel time error by the pulse width of
-$g_{\lambda,\frac{\lambda}{a^2}}$. The analogous relation for $_{\rm
+$g_{\lambda,\frac{r\lambda}{a^2}}$. The analogous relation for $_{\rm
   MSWI}$ involves the norm-squared as well. To assess these
@@
 \[
-  \|g_{\lambda,\frac{\lambda}{a^2}}\|^2 =\frac{1}{2\pi} \int d\omega\,
-  \left(\frac{|\fwl|^2(\omega)}{|\fwl|^2(\omega) + \frac{\lambda}{a^2}}\right)^2
+  \|g_{\lambda,\frac{r\lambda}{a^2}}\|^2 =\frac{1}{2\pi} \int d\omega\,
+  \left(\frac{|\fwl|^2(\omega)}{|\fwl|^2(\omega) + \frac{r\lambda}{a^2}}\right)^2
 =\frac{1}{2\pi} \int d\omega\,
@@
       \omega)|^2(\omega)}{|\lambda^{1/2}\hat{w}_1(\lambda
-      \omega)|^2 + \frac{\lambda}{a^2}}\right)^2
+      \omega)|^2 + \frac{r\lambda}{a^2}}\right)^2
 \]
@@
       \omega)|^2}{|\hat{w}_1(\lambda
-      \omega)|^2 + \frac{\lambda}{a^2\lambda}}\right)^2
+      \omega)|^2 + \frac{r}{a^2}}\right)^2
 =  \frac{1}{2\pi \lambda} \int d\omega_1\,
-  \left(\frac{|\hat{w}_1(\omega_1)|^2}{|\hat{w}_1(\omega_1)|^2 + \frac{\lambda}{a^2\lambda}}\right)^2
-\]
-At this point, it is possible to identify an advantageous choice of
-the Tihonov penalty weight $\lambda$: choose
-$\lambda = \lambda$  and rearrange to obtain
+  \left(\frac{|\hat{w}_1(\omega_1)|^2}{|\hat{w}_1(\omega_1)|^2 +
+      \frac{r}{a^2}}\right)^2 
+\]
+%At this point, it is possible to identify an advantageous choice of
+%the Tihonov penalty weight $\lambda$: choose
+%$\lambda = \lambda$  and rearrange to obtain
 \begin{equation}
   \label{eqn:g0}
-  \|g_{\lambda,\frac{\lambda}{a^2}}\|^2 \approx \frac{1}{\lambda} \int d\omega\,
-  \left(\frac{|a\hat{w}_1|^2}{|a\hat{w}_1|^2 + 1}\right)^2
-\end{equation}
-That is, with this choice $\lambda = \lambda$, the wavelength proxy
-$\lambda$ appears only in the factor in front of the integral sign.
+  = \frac{1}{2\pi \lambda} \int d\omega\,
+  \left(\frac{|a\hat{w}_1|^2}{|a\hat{w}_1|^2 + r}\right)^2 =
+  \frac{1}{\lambda} \|g_{1,r}\|^2
+\end{equation}
+For future use, note that this identity together with \ref{eqn:u0snorm}: implies that
+\begin{equation}
+    \label{eqn:u0snormlam}
+  \|\uzl\| = C\lambda^{-\frac{1}{2}}.
+\end{equation}
+Note that according to the standing convention, $C$in this identity
+depends on $M$ and $w_1$ but neither on $\kappa$ nor $\lambda$.
 
@@
 \[
-  \int dt\, t^2|g_{\lambda,\frac{\lambda}{a^2}}(t)|^2 = \frac{1}{2\pi} \int d\omega
-  \left|\frac{d\hat{g}_{\lambda,\frac{\lambda}{a^2}}}{d\omega}\right|^2
-  =  \int d\omega  \left(\frac{2 \mbox{Re }\bar{\hat{w}}_{\lambda}\fwl' 
-    \frac{\lambda}{a^2}}{(|\fwl|^2 + \frac{\lambda}{a^2})^2}\right)^2
-\]
-\[
-  =\frac{1}{2\pi}\left(\frac{\lambda}{a^2}\right)^2 \int d\omega \frac{(2 \lambda^{1/2}\hat{w}_1(\lambda
+  \int dt\, t^2|g_{\lambda,\frac{r\lambda}{a^2}}(t)|^2 = \frac{1}{2\pi} \int d\omega
+  \left|\frac{d\hat{g}_{\lambda,\frac{r\lambda}{a^2}}}{d\omega}\right|^2
+  =  \frac{1}{2\pi} \int d\omega  \left(\frac{2 \mbox{Re }\bar{\hat{w}}_{\lambda}\fwl' 
+    \frac{r\lambda}{a^2}}{(|\fwl|^2 + \frac{r\lambda}{a^2})^2}\right)^2
+\]
+\[
+  =\frac{1}{2\pi}\left(\frac{r\lambda}{a^2}\right)^2 \int d\omega \frac{(2 \lambda^{1/2}\hat{w}_1(\lambda
     \omega)\lambda^{3/2}\hat{w}_1'(\lambda \omega))^2}{(\lambda|\hat{w}_1(\lambda
-    \omega)|^2 + \frac{\lambda}{a^2})^4}
-\]
-\[
-  =\frac{1}{2\pi\lambda}\left(\frac{\lambda}{a^2}\right)^2\int d\omega_1 \frac{(2 \lambda^2\hat{w}_1(\omega_1)\hat{w}_1'( \omega_1))^2}{(\lambda|\hat{w}_1(
-    \omega_1)|^2 + \frac{\lambda}{a^2})^4}
-  =\frac{1}{2 \pi \lambda}\left(\frac{\lambda}{a^2}\right)^2\int d\omega_1
+    \omega)|^2 + \frac{r\lambda}{a^2})^4}
+\]
+\[
+  =\frac{1}{2\pi\lambda}\left(\frac{r\lambda}{a^2}\right)^2\int d\omega_1 \frac{(2 \lambda^2\hat{w}_1(\omega_1)\hat{w}_1'( \omega_1))^2}{(\lambda|\hat{w}_1(
+    \omega_1)|^2 + \frac{r\lambda}{a^2})^4}
+  =\frac{1}{2 \pi \lambda}\left(\frac{r\lambda}{a^2}\right)^2\int d\omega_1
   \frac{(2\hat{w}_1(\omega_1)\hat{w}_1'( \omega_1)
-    )^2}{\left(|\hat{w}_1(\omega_1)|^2 + \frac{\lambda}{a^2\lambda}\right)^4}
-\]
-With the choice $\lambda = \lambda$, this becomes
-\[
-  =\frac{\lambda}{2\pi a^4}\int d\omega_1
+    )^2}{\left(|\hat{w}_1(\omega_1)|^2 + \frac{r}{a^2}\right)^4}
+\]
+\[
+  =\frac{r\lambda}{2\pi a^4}\int d\omega_1
   \frac{(2\hat{w}_1(\omega_1)\hat{w}_1'( \omega_1)
-    )^2}{\left(|\hat{w}_1(\omega_1)|^2 + \frac{1}{a^2}\right)^4}
+    )^2}{\left(|\hat{w}_1(\omega_1)|^2 + \frac{r}{a^2}\right)^4}
 \]
@@
   \label{eqn:g1}
- = \frac{\lambda}{2\pi} \int d\omega
+ = \frac{r\lambda}{2\pi} \int d\omega
   \frac{(2a\hat{w}_1a\hat{w}_1'
-    )^2}{\left(|a\hat{w}_1|^2 + 1\right)^4}
-\end{equation}
+    )^2}{\left(|a\hat{w}_1|^2 + r\right)^4}
+\end{equation}
+
+Define
+\begin{equation}
+  \label{eqn:Wdef}
+  W[\kappa] = \frac{(a^*)^2}{a^2}\|g_{1,r}\|^2
+\end{equation}
+Since $W$ depends on $\kappa$ through $a$, and on $a^*$, it satisfies
+bounds of the form stated in the Proposition.
 
@@
   \label{eqn:pwg}
-  l(g_{\lambda,\frac{\lambda}{a^2}}) =\lambda W(m,w_1)
-\end{equation}
-in which $W(m,w_1)$ is bounded by the variation of the amplitude
-$a[\kappa](\bx_r,\bx_s)$, hence uniformly for $\log \kappa \in M$ (property
-H2), and the choice of ``mother wavelet'' $w_1$, but
-independent of $\lambda$.
-
-It follows that
-\begin{equation}
-  \label{eqn:mswitt}
-  J_{\rm MSWI,\lambda,\lambda}[\kappa;\dla] =
-  \frac{1}{\lambda} \sum_{{\bf x}_s,{\bf x}_r} C[\kappa](\bx_s,\bx_r)(\tau[\kappa^*]({\bf
-  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 + O(\lambda)
-\end{equation}
-
-\begin{equation}
-  \label{eqn:awitt}
-J_{\rm AWI,\lambda,\lambda}[\kappa;\dla] = \sum_{{\bf x}_s,{\bf x}_r} (\tau[\kappa^*]({\bf
-  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 + O(\lambda^2)
-\end{equation}
+  |l(g_{\lambda,\frac{r\lambda}{a^2}})| \le C \lambda
+\end{equation}
+
+The conclusions \ref{eqn:mswittz} and \ref{eqn:awittz} follow directly
+from these observations and equations \ref{eqn:MSWIasym2} and
+\ref{eqn:AWIasym2}.
+
+%It follows that
+%\begin{equation}
+%  \label{eqn:mswitt}
+%  J_{\rm MSWI,\lambda,\lambda}[\kappa;\dla] =
+%  \frac{1}{\lambda} \sum_{{\bf x}_s,{\bf x}_r} C[\kappa](\bx_s,\bx_r)(\tau[\kappa^*]({\bf
+%  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 + O(\lambda)
+%\end{equation}
+%  
+%\begin{equation}
+%  \label{eqn:awitt}
+%J_{\rm AWI,\lambda,\lambda}[\kappa;\dla] = \sum_{{\bf x}_s,{\bf x}_r} (\tau[\kappa^*]({\bf
+%  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 + O(\lambda^2)
+%\end{equation}
 
@@
 
-In particular, the modified AWI objective $J_{\rm AWI,\lambda}$ is {\em exactly} an approximation to
-the travel-time tomography objective, with errors proportional to the squared
-wavelength proxy $\lambda$.
+\noindent {\bf Remark:} Note that the homogeneity in $\lambda$ of the
+expressions in equations \ref{eqn:g0} and \ref{eqn:g1} results from
+the proportionality between $\lambda$ and the regularization weight in
+the definitions \ref{eqn:reg} and \ref{eqn:leading}. Otherwise, such
+simple asymptotic relations as the conclusions
+of Proposition \ref{thm:leading} would not obtain.
 
@@
 
-\begin{prop} Under the hypotheses of Proposition \ref{thm:leading},
+\begin{prop}
+  \label{thm:remest}
+  Under the hypotheses of Theorem \ref{thm:main},
   \begin{eqnarray}
-    \label{eqn:diffests}
-    \|\ul-\uzl\| &\le& C \nonumber \\
-    \|T(\ul-\uzl)\| &\le& C
+    \|\ul-\uzl\| &\le& C \lambda^{\frac{1}{2}}\|\dl\|\label{eqn:diffu} \\
+    \|T(\ul-\uzl)\| &\le& C \lambda^{\frac{1}{2}}\|\dl\|\label{eqn:difftu}
   \end{eqnarray}
@@
     \label{eqn:step1}
-    ((\Sla)^T\Sla + \lambda I) (u_{\lambda}-\uzl) = (\Slb)^T d_{\lambda} + (\Sla)^T(d_{\lambda}-d_{0,\lambda})- 
+    ((\Sla)^T\Sla + r\lambda I) (u_{\lambda}-\uzl) = (\Slb)^T d_{\lambda} + (\Sla)^T(d_{\lambda}-d_{0,\lambda})- 
     ((\Sla)^T\Slb + (\Slb)^T\Sla + (\Slb)^T\Slb)u_{\lambda}
@@
 \]
-Combining the preceding results in the first estimate
-\ref{eqn:diffests}.
-
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%
-
-over in the expression for $g$ is exponentially decaying, hence
-%And that's ``why it works'' - when geometric asymptotics in the form
-%given by equation \ref{eqn:hadamard} is an accurate approximation to
-%the Green's function.
-
-To evaluate the difference between $J_{\rm AWI}$ and
-$J_{\rm AWI,\lambda}$, first estimate the norms of differences
-$\|u_{\lambda} - \uzl\|$ and $\|T(u_{\lambda} -
-\uzl)\|$.
-
-Suppressing $ \bx_s,\bx_r$ and using the notation of the last
-subsection, definitions \ref{eqn:Fl0}, \ref{eqn:Fl1} and
-\ref{eqn:sdeflam} can be re-written
-\[
-  \Sl u = (a \wl*u + (bH) * \wl *u)(t-\tau)
-\]
-where $bH$ is an abbreviation for
-$b[\kappa](\bx_r,t-\tau[\kappa];\bx_s)H(t-\tau[\kappa])$. Note that
-$b$ decays exponetially, is smooth, and all derivatives have limits
-at $t=0$. Similarly,
-\[
-  d = \Fl[\kappa^*] = (a^*\wl + (bH)  * \wl) (t-\tau).
-\]
-Recall (equations \ref{eqn:Fl0}, \ref{eqn:Fl1}, re-written as
-\[
-\Fla = a \wl * \delta_{\tau}, \,   \Flb  = (f_{\lambda}b(0) -
-f_{\lambda}*(bH))*\delta_{\tau}.
-\]
-Since $S_{\lambda,i} u = F_{\lambda,i} * u, \,i=0,1,$ $\wl$ is even,
-and $f_{\lambda}$ is odd,
-\[
- ((\Sla)^T \Slb + (\Slb)^T\Sla)u = [\rho a b(0)( \check{\wl} *\delta_{-\tau}*f_{\lambda}*\delta_{\tau} + \wl
- *\delta_{\tau}*\check{f}_{\lambda}*\delta_{-\tau}) 
-\]
-\[
--
- \rho a(\check{\wl}*\delta_{-\tau} *f_{\lambda}*(bH)*\delta_{\tau} + \wl*\delta_{\tau}*\check{f}_{\lambda}*\check{b}\check{H)}*\delta_{-\tau}) ]*u
-\]
-\[
-  = \rho a(\wl*f_{\lambda}*(\check{b}\check{H}-(bH)) *u
-\]
-Since $\wl = \rho \partial_t f_{\lambda}$ and $(\partial_t
-(\check{b}\check{H}-bH))*u = (-2\partial_tb(0)u + (\partial_t\check{b}\check{H} -\partial_tb H))*u$, this is
-\begin{equation}
-  \label{eqn:symmprods0s1}
-  = \rho^2 a(-2\partial_tb(0) \delta + (\partial_t\check{b}\check{H} -
-  \partial_tb H)) * f_{\lambda}*f_{\lambda}*u_{\lambda}
-\end{equation}
-The bounds \ref{eqn:best}, \ref{eqn:lambdanorms} imply that
-\begin{equation}
-  \label{eqn:s0s1norm}
-  \|(\Sla)^T\Slb + (\Slb)^T\Sla\| \le C \lambda^3,
-\end{equation}
-with $C$ uniform in $\log \kappa \in M$. From \ref{eqn:Srembd},
-\begin{equation}
-  \label{eqn:noperr}
-  \|\Sl^T\Sl - (\Sla)^T\Sla\| \le C \lambda^3.
-\end{equation}
-A little algebra yields
-\begin{equation}
-  \label{eqn:step1}
-  ((\Sla)^T\Sla + \lambda I) (u_{\lambda}-\uzl) = (\Slb)^T d_{\lambda} + (\Sla)^T(d_{\lambda}-d_{0,\lambda})- 
-  ((\Sla)^T\Slb + (\Slb)^T\Sla + (\Slb)^T\Slb)u_{\lambda}
-\end{equation}
-Note that the bounds \ref{eqn:Srembd} imply
-\[
-  \|\Sl^T d\| \le C \lambda^{\frac{1}{2}}\|d\|,
-\]
-\[
-  \|(\Sla)^T (d-d_0)\| = \|(\Sla)^T(f_{\lambda}b^*(0) -
-f_{\lambda}*(b^*H)*\delta_{\tau^*}\| \le C \lambda^2,
-\]
-and
-\[
- \| (\Slb)^Td \| \le C \lambda^{3/2}.
-\]
-Hence
-\[
-  \|u_{\lambda}\| \le C\lambda^{-\frac{1}{2}}\|d\|.
-\]
-Combine the preceding three bounds with identity \ref{eqn:step1} and
-bound \ref{eqn:noperr} to obtain
-\begin{equation}
-  \label{eqn:step2}
-  \|u_{\lambda}-\uzl\| \le C
-  \lambda^{\frac{1}{2}}\|d_{\lambda}\|
-\end{equation}
-
-From identity \ref{eqn:step1},
-\[
-   ((\Sla)^T\Sla + \lambda I) T(\ul-\uzl) = -[T,
+Combining the preceding results in thet estimate in display
+\ref{eqn:diffu}.
+
+Substituting $\fl$ for $f$, $\dl$ for $d$ and $r\lambda$ for $\sigma$ in the estimate \ref{eqn:tunorm} and using
+the bounds \ref{eqn:lambdanorms} gives
+\begin{equation}
+  \label{eqn:step3}
+  \|T\ul\| \le C \lambda^{-\frac{1}{2}}\|\dl\|
+\end{equation}
+From identity \ref{eqn:step1}, obtain
+\[
+   ((\Sla)^T\Sla + r\lambda I) T(\ul-\uzl) = -[T,
    (\Sla)^T\Sla](\ul-\uzl) - T \Slb)^T \dl 
@@
 \begin{equation}
-  \label{eqn:step3}
+  \label{eqn:step4}
  + T (\Sla)^T(\dl-\dzl)- 
@@
 \[
-  [T,\Sla] = a \tau \delta_{\tau} * \wl* + a \delta_{\tau} * (T\wl)*
-\]
-and
-\[
-  [T,(\Sla)^T] = -a \tau \delta_{-\tau} * \wl* + a 
-  \delta_{-\tau}*(T\wl)*,
+  (\Sla)^T\Sla = a^2 \wl * \wl*
 \]
@@
   \label{eqn:twl}
-  \|T\wl\|_{L^1(X_{sr}\times \bR)} =
-  \|\partial_t(T\fl)-\fl\|_{L^1(\bR)} =\lambda^{\frac{3}{2}}\int dt_1
-  |t_1||\partial f_1(t_1)| \le C \lambda^{\frac{3}{2}}
+  \|T\wl\|_{L^1(X_{sr}\times \bR)} \le C\lambda^{\frac{3}{2}}
 \end{equation}
@@
   \label{step5}
-  \|[T,  (\Sla)^T\Sla](\ul-\uzl)\| \le C \lambda^{2}\|\dl\|,
+  \|[T,  (\Sla)^T\Sla](\ul-\uzl)\| \le C \lambda^{\frac{5}{2}}\|\dl\|,
 \end{equation}
@@
 \[
-  [T, \Slb] = [T,bH*\delta_{\tau}*\wl*] = ((T+\tau)bH*\wl
-  + bH*T\wl)*\delta_{\tau}
+  [T, \Slb] = [T,bH*\delta_{\tau}*\wl*] = ((T+\tau)bH*\wl*
+  + bH*T\wl)*\delta_{\tau}*
+\]
+\[
+  = (((\partial_t (T+\tau)b) + \tau
+  b(0)\delta)* \fl * + bH*(T\wl))*\delta_{\tau}
 \]
@@
 \[
-  [T,(\Slb)^T]  =  ((T - \tau) \check{b}\check{H})*\wl
-  + \check{b}\check{H}*T\wl)*\delta_{-\tau}
-\]
-\[
-= \check{b}\check{H}*\fl
--\tau)b(0)\delta *\fl + (T-\tau)
-\partial_t\check{b}\check{h}*\fl +
-\check{b}\check{H}*(\partial_t(T\fl)-\fl)
-\]
-Using \ref{eqn:twl}, conclude that
-\begin{equation}
-  \label{eqn:step6}
-  [T, (\Slb)^T] \le C \lambda^{\frac{3}{2}}
-\end{equation}
-
-Observe that
-\[
-  T\dl=T(a^*\delta + b^*H)*\delta_{\tau^*}*\wl
-\]
-\[
-  =Tb^*H *\delta_{\tau^*}*\wl+(a^*\delta+b^*H)\tau^*\delta_{\tau^*}\wl
-  +(a^*\delta + b^*H)*\delta_{\tau^*}*T\wl,
-\]
-so the exponential decay of $b^*$ implies that $\|T\dl\| \le
-C\|\dl\|$. Thus the estimate \ref{eqn:step6} implies
-\begin{equation}
-  \label{eqn:step7}
-  \|T(\Slb)^T\dl\| \le C \lambda^{3/2}\|\dl\|.
-\end{equation}
-
-On the other hand,
-\[
-  T(\dl - \dzl) = T(b^*H*\delta_{\tau^*}*\wl = T(b^*(0)\delta +
-  \partial_t b^*H)*\delta_{\tau^*} *\fl
-\]
-\[
-  (T\partial_tb^*H)*\delta_{\tau^*}*\fl+(b^*(0)\delta +
-  \partial_t b^*H)\tau^*\delta_{\tau^*}*\fl +( b^*(0)\delta +
-  \partial_t b^*H)*\delta_{\tau^*}T\fl
-\]
-From the second relation \ref{eqn:lambdanorms},
-\begin{equation}
-  \label{eqn:step7.5}
-  \|T(\dl-\dzl)\| \le C \lambda
-\end{equation}
-in which $C$ depends on $M$ and on the support of $f_1$ (which
-contains the support of $\fl$). The second bound \ref{eqn:rembd}
-gives
-\begin{equation}
-  \label{step8}
-  \Slb T(\dl-\dzl) \le C\lambda^{\frac{3}{2}}
-\end{equation}
-Also,
-\[
-  [T,\Slb] = ((Tb)H) *\wl * \delta_{\tau}* + (bH) * T\wl *
-delta_{\tau} * + \tau (bH)* \wl * \delta_{\tau}*
-\]
-\[
-  = ((\partial_t Tb) H * \delta_{\tau}*\fl* +  \tau(b(0)\delta + \partial_t bH)*\fl*\delta_{\tau}*
-\]
-\[
-  +( b(0)\delta + \partial_t bH)*T\fl*\delta_{\tau} -
-  Bh*\delta_{\tau}*\fl*
-\]
-SInce $\|\fl\|_{L^1(\bR)} =\lambda^{\frac{3}{2}} \|f_1\|_{L^1(\bR)}$ and
-$\| T\fl\|_{L^1(\bR)} = \lambda^{\frac{5}{2}} \|f_1\|_{L^1(\bR)}$, every term in
-the preceding expression is $O(\lambda^{\frac{3}{2}})$. Since $T$ is symmetric,
-\begin{equation}
-  \label{eqn:step9}
-  \|[T,\Slb]\| = \|[T,(\Slb)^T]\| \le C\lambda^{\frac{3}{2}}.
-\end{equation}
-Note that by reasoning almost identical to that leading to the bound
-\ref{eqn:step7.5},
-\begin{equation}
-  \label{eqn:step9.5}
-  \|\dl-\dzl\| \le C\lambda.
-\end{equation}
-so
-\begin{equation}
-  \label{eqn:step10}
-  \|T (\Slb)^T(\dl-\dzl)\| \le \|[ T, (\Slb)^T](\dl-\dzl)\| +
-  \|(Slb)^TT (\dl-\dzl)\| \le C\lambda^{\frac{5}{2}}\|\dl\|
-\end{equation}
-
-%%%%%%%%%%%%%%%
-
-$(Tu)(t) = tu(t)$
-
-$T^T = T$
-
-$[T,\delta_{\tau}*] = \tau \delta_{\tau}*$
-
-$[T,k *] = (Tk)*$ for any $k \in {\cal S}'(\bR)$
-
-\[
-  [T,\Sla] = a \tau \delta_{\tau} * \wl* + a \delta_{\tau} * (T\wl)*
-\]
-\[
-  [T,(\Sla)^T] = -a \tau \delta_{-\tau} * \wl* + a \delta_{-\tau}*(T\wl) *
-\]
-\[
-  [T,\Sl] = a (\tau  \wl +  (T\wl) ) \delta_{\tau} + [T,(bH)*\delta_{\tau}*\wl]
-\]
-\[
-  =a (\tau  \wl +  (T\wl) ) \delta_{\tau}* + (Tb)H *\delta_{\tau}*\wl* +
-  (bH)*(\tau \wl + T\wl)*\delta_{\tau}*
-\]
-\[
-  = (a\delta + (bH))*(\tau \wl +
-  T\wl)*\delta_{\tau}* + (Tb)H *\delta_{\tau}*\wl*
-\]
-\[
-[T,\Sl^T] =   = (a\delta + (\check{b}\check{H}))*(-\tau \wl +
-  T\wl)*\delta_{-\tau}* + (T\check{b})\check{H} *\delta_{-\tau}*\wl*
-\]
-
-\[
-  [T,(\Sla)^T\Sla] = [T,(\Sla)^T]\Sla + (\Sla)^T [T,\Sla]
-\]
-\[
-= (-a \tau \delta_{-\tau} * \wl + a \delta_{-\tau}*(T\wl)) *
-a\delta_{\tau} * \wl
-\]
-\[
-  + a \delta_{-\tau} * \wl *  ( a \tau \delta_{\tau} * \wl + a
-  \delta_{\tau} * (T\wl))
-\]
-\[
-  = 2 a^2 \wl * T\wl *
-\]
-
-\[
-  [T,\Sl^T\Sl] = [T,\Sl^T]\Sl + \Sl^T [T,\Sl]
-\]
-\[
-  =((a\delta + (\check{b}\check{H}))*(-\tau \wl +
-  T\wl)*\delta_{-\tau}* + (T\check{b})\check{H}
-  *\delta_{-\tau}*\wl*)(a \delta + (bH))*\wl*\delta_{\tau}
-\]
-\[
-  +(a \delta + (\check{b}\check{H}))*\wl*\delta_{-\tau}*
-  ((a\delta + (bH))*(\tau \wl +
-  T\wl)*\delta_{\tau}* + (Tb)H *\delta_{\tau}*\wl*)
-\]
-\[
-=(a\delta + (\check{b}\check{H}))*(-\tau \wl +
-T\wl)*\delta_{-\tau}*(a \delta + (bH))*\wl*\delta_{\tau} +
-\]
-\[
-  (a \delta + (\check{b}\check{H}))*\wl*\delta_{-\tau}*(a\delta + (bH))*(\tau \wl +
-  T\wl)*\delta_{\tau}*
-\]
-\[
-  ...
-\]
-\[
-  = 2(a \delta + (\check{b}\check{H}))*(a\delta + (bH)*\wl*T\wl +
-\]
-
-From identity \ref{eqn:symmprods0s1},
-\[
-  [T,(\Sla)^T\Slb+(\Slb)^T\Sla] = \rho^2 a[T,(-2\partial_t b(0) \delta +
-  (\partial_t\check{b}\check{H}-\partial_tb
-  H))]*f_{\lambda}*f_{\lambda}*
-\]
-\[
-  + 2\rho^2 a (-2\partial_t b(0) \delta +
-  (\partial_t\check{b}\check{H}-\partial_tb
-  H))]*f_{\lambda}*Tf_{\lambda}*
-\]
-\[
-  =
-  \rho^2 a[ T (\partial_t\check{b}\check{H}-\partial_tb
-  H))]*f_{\lambda}*f_{\lambda}* +  2\rho^4 a (-2\partial_t b(0) \delta +
-  (\partial_t\check{b}\check{H}-\partial_tb
-  H))]*f_{\lambda}*Tf_{\lambda}*
-\]
-%\[
-%  [T,\Slb] = ((Tb)H) *\wl * \delta_{\tau}* + (bH) * T\wl *
-%  \delta{\tau} * + \tau (bH)* \wl * \delta_{\tau}
-%\]
-%\[
-%[T,(\Slb)^T] = - [T,\Slb]^T =-(\checi((Tb)\check{H}*\wl*\delta{-\tau} +
-%(\check{b}\check{H}) * (-Twl)*\delta{-\tau}^* + \tau
-%(\check{b}\check{H))*\wl*\delta_{\-tau}]
-%\]
-%\begin{equation}
-%  \label{eqn:step3}
-%  ((\Sla)^T\Sla + \lambda I) (u_{\lambda}-\uzl) = (\Slb)^T d_{\lambda} + (\Sla)^T(d_{\lambda}-d_{0,\lambda})- 
-%  ((\Sla)^T\Slb + (\Slb)^T\Sla + (\Slb)^T\Slb)u_{\lambda}
-%\end{equation}
+  \|[T,\Slb]\| \le C\lambda^{\frac{3}{2}}
+\]
+and so using \ref{eqn:dnorms} and \ref{eqn:lambdanorms}
+\[
+\|T\Slb \dl \|=  \|[T,(\Slb)^T]\dl \|+ \Slb T\dl\| \le C\lambda^{\frac{3}{2}}\|\dl\|
+\]
+Similarlly,
+\[
+\|T\Sla (\dl-\dzl) \le C \lambda^{\frac{3}{2}}\|\dl\|
+\]
+Finally,
+\[
+  [T, (\Sla)^T\Slb + (\Slb)^T\Sla + (\Slb)^T\Slb] = [T,g*\fl*\fl*] =
+  Tg*\fl*\fl* + 2 g*\fl*T\fl*
+\]
+in the notation introduced above. Since $g$ is a linear combination of
+convolutions of truncated exponentially decreasing kernels, it remains
+integrable after multiplication by $t$, and
+\[
+\|T\fl\|_{L^1(\bR)} = C\lambda^{\frac{5}{2}}. 
+\]
+Since $\|T\dl\| \le C\|\dl\|$, the estimates \ref{eqn:step2} and \ref{eqn:step3} implies
+\[
+  \|T ((\Sla)^T\Slb + (\Slb)^T\Sla + (\Slb)^T\Slb)\ul\| \le
+  C\lambda^{3}\|\ul\| \le C\lambda^{\frac{5}{2}} \|dl\|
+\]
+Adding up, the right-hand side of equation \ref{eqn:step4} is bounded
+by
+\[
+  C \lambda^{\frac{3}{2}}\|\dl\|
+\]
+which together with preceding inequalities establishes the second estimate \ref{eqn:difftu}.
+
+\end{proof}
+
+\begin{proof} {\em of Theorem \ref{thm:main}}: From the inequality \ref{eqn:mswittz}
+  (suppressing the arguments $\bx_s,\bx_r$),
+  \[
+   | \|T\uzl[\kappa;\dla] \|- 
+   \lambda^{-\frac{1}{2}}
+   W[\kappa]^{\frac{1}{2}}|\tau[\kappa^*]-\tau[\kappa]| | \le
+   C\lambda^{\frac{1}{2}}
+ \]
+ Combine with inequality \ref{eqn:difftu} to establish the first
+ result \ref{eqn:mswitt}.
+
+ The leading-term result \ref{eqn:awittz} implies that
+ \begin{equation}
+   \label{eqn:awittzsr}
+   \left| \frac{\|T\uzl[\kappa;\dla] \|}{
+       \|\uzl[\kappa;\dla] \|} - |\tau-\tau^*|\right|  \le C\lambda.
+ \end{equation}
+ A little algebra gives
+ \[
+   \frac{\|T\ul\|}{\|\ul\|} \le \left(\frac{\|T\uzl\|}{\|\uzl\|}+\frac{\|T\ul
+     -T\uzl\|}{\|\uzl}\right)\left(1-\frac{\|\ul-\uzl\|}{\|\uzl\|}\right)^{-1}
+ \]
+ \[
+   \le \left(\frac{\|T\uzl\|}{\|\uzl\|} + C\lambda \|\dl\|\right) (1-C\lambda)
+ \]
+ Using the result \ref{eqn:u0snormlam}, and assuming $\lambda \le
+ 1/C$, find
+ \[
+ \frac{\|T\ul\|}{\|\ul\|}-\frac{\|T\uzl\|}{\|\uzl\|} \le C\lambda
+ \left(1+2 \frac{\|T\uzl\|}{\|\uzl\|} \right) + 2 C^2 \lambda^2
+\]
+\[
+  \le \le C\lambda  (1+2|\tau - \tau^*| + 2 (C^2+1) \lambda^2
+\]
+Since the maximum difference of traveltimes is a function of $M$, this
+can be written as
+\[
+  \le C\lambda
+\]
+Similarly,
+\[
+  \frac{\|T\uzl\|}{\|\uzl\|}-\frac{\|T\ul\|}{\|\ul\|} \le C\lambda
+\]
+Combine these inequalities with the leading-term result
+\ref{eqn:awittzsr} to yield the estimate \ref{eqn:awitt}.
+
+\end{proof}
 
@@
 
-This relation is abstract, and does not just pertain to the problems
+This relation is abstract, and does not only pertain to the problems
 discussed in this paper. To derive this abstract relation, suppose for
-the moment that $S:U \rightarrow D$ is a bounded and coercive from a
+the moment that $S:U \rightarrow D$ is a bounded and coercive map from a
 domain Hilbert space $U$ to a range Hilbert space $D$. That is, for
@@
 \]
-Suppose that $d \in D$ (``data''), and $T$ is another bounded operator $U \rightarrow E$, $E$
+Suppose that $d \in D$ (``data''), and $T$ is another bounded operator $U \rightarrow V$, $V$
 another Hilbert space. Define
@@
   \label{eqn:eq1}
-  J_{\alpha}(u) = \frac{1}{2}(\|Su-d\|_D + \alpha^2\|Tu\|_E^2).
+  J_{\alpha}(u) = \frac{1}{2}(\|Su-d\|_D + \alpha^2\|Tu\|_V^2).
 \end{equation}
@@
 
-The claim to be established is that
+The claim to be established is
+\begin{prop}
+  \label{thm:alphalim}
+  Suppose that $U$, $V$ and $D$ are Hilbert spaces,
+  $S:U\rightarrow D$ is bounded and coercive, $T:U\rightarrow V$ is
+  bounded, $\ua$ is the solution of
+  the normal equation \ref{eqn:eq0n}, and $J_{\alpha}$,
+  and $\tJa$ are defined by \ref{eqn:eq1} and \ref{eqn:eq0}
+  respecively for $\alpha \ge 0$. Then
 \begin{equation}
@@
 \end{equation}
-
-To see this, use the normal equation to write
-\[
-  \ua = (S^TS + \alpha^2 T^TT)^{-1}T^Td = (S^TS + \alpha^2 T^TT)^{-1}S^TS \uz
+\end{prop}
+
+\begin{proof}
+Use the normal equation to write
+\[
+  \ua = (S^TS + \alpha^2 T^TT)^{-1}S^Td = (S^TS + \alpha^2 T^TT)^{-1}S^TS \uz
 \]
@@
   \label{eqn:eq3}
-  \tJa = \frac{1}{2}(\|S\uz-d\|_d^2 - 2 \alpha^2\langle S\uz-d, S\va\rangle_d + \alpha^2\|T\uz\|_m^2 + O(\alpha^4))
+  \tJa = \frac{1}{2}(\|S\uz-d\|_D^2 - 2 \alpha^2\langle S\uz-d, S\va\rangle_D + \alpha^2\|T\uz\|_V^2 + O(\alpha^4))
 \end{equation}
 The second term on the RHS of equation \ref{eqn:eq3} vanishes thanks to the normal equation, and the first term is precisely $\tJz$. The conclusion \ref{eqn:eq4} follows.
+\end{proof}
 
 To apply the assertion \ref{eqn:eq4} to the MSWI objective, make the
-following choices, for each $m$ in the domain of $F$: $S$ is the regularized
-modeling operator $(S_{\lambda}[\kappa],\lambda I)^T$.  Its domain is $U=L^2(\bR)$, of
-which $u$ (the adaptive filter) is a member, and 
-its range is $D = L^2(\bR) \oplus
-L^2(\bR)$. The role of $d$ is played  the
-augmented data $(d,0)^T$. $T$ is a modified multiply-by-$t$ operator
-$Tu(t)=\max(|t|, t_d) \mbox{sgn}(t)u(t)$, with domain $U$ and
-range $E=L^2(\bR)$. The truncation by $t_d > 0$ is necessary
-to make $T$ bounded, but can be chosen large enough relative to the
-support of the filter (necessarily bounded in practical calculation)
-to have no effect. Then $\Ja$ becomes $J_{{\rm MSWI},\alpha,\sigma}[\kappa,u;d]$, where
-\begin{equation}
-  \label{eqn:eq5}
-   J_{{\rm MSWI},\alpha,\sigma}[\kappa,u;d] = \frac{1}{2}\left(\|S_{\lambda}[\kappa]u-d\|^2 +
-   \lambda\|u\|^2 + \alpha^2\|Tu\|^2\right)
- \end{equation}
-The minimizer $\ua$ over $u$ of $ J_{{\rm MSWI}, \alpha,\sigma}[\kappa,u;d]$ is
-$u_{{\rm MSWI},\alpha,\sigma}[\kappa;d]$, and $\tJa$ becomes
-\begin{equation}
-  \label{eqn:eq6}
-  \tilde{J}_{{\rm MSWI},\alpha,\sigma}[\kappa;d] =
-  J_{{\rm MSWI},\alpha,\sigma}[\kappa,u_{{\rm MSWI},\alpha,\sigma}[\kappa;d];d].
-\end{equation}
-Note that $u_0$ becomes $u_0[\kappa_{\lambda};d]$, and $\tilde{J}_0$ becomes
-$\tilde{J}_{0,\sigma}[\kappa;d]$ which means exactly what
-it did in the previous section. The subscript ${\rm MSWI}$ is left
-off for $\alpha=0$, for reasons that will become apparent shortly.
-
-With these correspondences, equation \ref{eqn:eq4} becomes
+following choices, using the notation developed in the last section.
+In particular, $\sigma$ and $\lambda$ are related throughout by
+$\sigma=r\lambda$, where $r$ is a fixed positive constant with unit
+pressure$^2 \times$time$^2$.
+For each $\kappa \in M$, $\lambda \in (0,1]$, $S$ is the regularized
+modeling operator $[\Sl,\sqrt{r\lambda} I]^T$, with domain is
+$U=L^2(X_{sr}\times\bR)$ and range $D =  L^2(X_{sr} \times \bR) \oplus
+L^2(\bR \times \bR)$. The role of $d$ is played by the
+augmented data $(d,0)^T$, $d \in D$. $T$ is a modified multiply-by-$t$ operator
+$\Te$, defined by 
+\begin{equation}
+  \label{eqn:teps}
+  \Te u(t) = \frac{t}{\sqrt{1+\epsilon^2t^2}}.
+\end{equation}
+with domain and range $U ( = V)$. $\Te$ is bounded for $\epsilon>0$,
+and $=T$ on any natural dense domain as in the last section for $\epsilon=0$. The final result below will
+concern the limit $\epsilon \rightarrow 0$.
+
+Then $\Ja$ becomes $\Jase[\kappa,u;d]$:
+\begin{equation}
+  \label{eqn:mswipen}
+   \Jase[\kappa,u;d] = \frac{1}{2}\left(\|\Sl[\kappa]u-d\|^2 +
+   r\lambda\|u\|^2 + \alpha^2\|\Te u\|^2\right)
+ \end{equation}
+The minimizer $\ua$ over $u$ of $\Jase[\kappa,u;d]$ is
+$\uase[\kappa;d]$, and $\tJa$ becomes
+\begin{equation}
+  \label{eqn:mswipenred}
+  \tJase[\kappa;d] =
+  \Jase[\kappa,\uase[\kappa;d];d].
+\end{equation}
+
+Define
+\begin{equation}
+  \label{eqn:tj0def}
+  \tilde{J}_{\sigma}[\kappa;d] = \|S[\kappa]\ul[\kappa;d] - d\|^2 + \sigma\|\ul[\kappa;d\|^2.
+\end{equation}
+in which $\ul$ is the solution of the regularized least squares
+problem \ref{eqn:reg}.
+
+\begin{prop}
+  \label{thm:MSWIepsalphalim}
+Suppose that $\kappa^* \in M$ and $\dl=\Fl[\kappa^*]$.
 \begin{equation}
   \label{eqn:eq7}
+  \lim_{\epsilon \rightarrow 0}\lim_{\alpha \rightarrow 0}
+  \frac{1}{\alpha^2}\left(\tJase[\kappa;\dl]-\tilde{J}_{\sigma}[\kappa;d]\right)=
+ \|T\ul\|^2 =  J_{\rm MSWI,\sigma}[\kappa;d]
+\end{equation}
+in which $\ul$ is the solution of the least squares problem
+\ref{eqn:reg} with $S=\Sl$, $d=\dl$, and $J_{\rm MSWI,\sigma}$ is as
+defined in equation \ref{eqn:mswidef}.
+\end{prop}
+
+\begin{proof}
+  It follows immediately from Proposition \ref{thm:alphalim} that
+  \[
     \lim_{\alpha \rightarrow 0}
-  \frac{1}{\alpha^2}\left(\tilde{J}_{{\rm MSWI},\alpha,\sigma}[\kappa;d]-\tilde{J}_{0,\sigma}[\kappa;d]\right)=
-  J_{\rm MSWI}[\kappa;d]
-\end{equation}
-
-To derive the analogous relation for AWI only requires re-definition
-of the operator $T$: it becomes (taking explicitly into account
-dependence on source and receiver coordinates)
-\begin{equation}
-  \label{eqn:eq8}
-  T[\kappa;d] u(\bx_r,t;\bx_s) = \frac{\max(|t|, t_d)
-    \mbox{sgn}(t)}{\|u_0[\kappa_{\lambda};d](\bx_r,\cdot;\bx_s)\|}
-  u(\bx_r,t;\bx_s).
-\end{equation}
-That is, for AWI, $T$ depends on both model and data. Define
-\begin{equation}
-  \label{eqn:eq5a}
-   J_{{\rm AWI},\alpha,\sigma}[\kappa,u;d] = \frac{1}{2}\left(\|S_{\lambda}[\kappa]u-d\|^2 +
-   \lambda\|u\|^2 + \alpha^2\|T[\kappa;d]u\|^2\right)
- \end{equation}
-The minimizer $\ua$ over $u$ of $ J_{{\rm AWI}, \alpha,\sigma}[\kappa,u;d]$ is
-$u_{{\rm AWI},\alpha,\sigma}[\kappa;d]$, and $\tJa$ becomes
-\begin{equation}
-  \label{eqn:eq6a}
-  \tilde{J}_{{\rm AWI},\alpha,\sigma}[\kappa;d] =
-  J_{{\rm AWI};\alpha,\sigma}[\kappa,u_{{\rm AWI};\alpha,\sigma}[\kappa;d];d].
-\end{equation}
-Note that the definitions \ref{eqn:eq5} and \ref{eqn:eq6} are exactly
-the same as the corresponding definition \ref{eqn:eq5a} and
-\ref{eqn:eq6a} when $\sigma=0$, which is why the subscripts ${\rm
-  MSWI}$ and ${\rm AWI}$ are left off in these cases.
-
-With these choices, equation \ref{eqn:eq4} becomes 
-\begin{equation}
-  \label{eqn:eq9}
-  \lim_{\alpha \rightarrow 0}
-  \frac{1}{\alpha^2}\left(\tilde{J}_{{\rm AWI},\alpha,\sigma}[\kappa;d]-\tilde{J}_{0,\sigma}[\kappa;d]\right)=
-  J_{\rm AWI}[\kappa;d]
-\end{equation}
-
-The intervention of the $\alpha=0$ limit $\tilde{J}_{0,\sigma}$ in the
-relations \ref{eqn:eq7} and \ref{eqn:eq9} obscures the relation
-between the MSWI and AWI objective functions and their penalty
-counterparts. In fact, under the hypotheses used in deriving he
-relations 
-Recall that the wavelet $w$ is assumed essentially bandlimited, that is
-$|\hat{w}(\omega)| \rightarrow 0$ as $\omega \rightarrow \infty$
-faster than any negative power of $\omega$. It follows that $\hat{w}
-\in L^p(\bR)$ for any $p \ge 1$ (and in fact the same is true of
-$\hat{w}$ multiplied by any polynomial in $\omega$).
-
-Assume as in the last section, that is, that solutions
-of the acoustic system \ref{eqn:awe} is well-approximated by the
-geometric asymptotics as in equation \ref{eqn:hadamard}, and that the
-data is noise-free ($d=F[\kappa^*]$). Together with the properties of $w$
-just listed, these assumptions imply that
-\begin{equation}
-  \label{eqn:eq11}
-  J_{0,\sigma}[\kappa;d] = O(\sigma) + O(\lambda(w)).
-\end{equation}
-To see this, begin with the definition:
-\[
-  \tilde{J}_{0,\sigma}[\kappa;d] = \frac{1}{2}
-  (\|S_{\lambda}[\kappa]u_0[\kappa_{\lambda};d]-d\|^2 + \lambda \|u_0[\kappa_{\lambda};d]\|^2.
-\]
-Using the approximate identities \ref{eqn:hatuf} and \ref{eqn:hatgf} and
-rearranging,
-\[
-\|S_{\lambda}[\kappa]u_0[\kappa_{\lambda};d]-d\|^2 
-  \approx \int d\omega
-  |a\hat{w}\hat{u_0[\kappa_{\lambda};d]}e^{i\omega\tau}-a^*\hat{w}e^{i\omega\tau^*}|^2
-\]
-\[
-  = \int d\omega
-  (a^*)^2|\hat{w}|^2\left(\frac{\lambda}{a^2|\hat{w}|^2 +
-      \lambda}\right)^2   
-\le \int d\omega (a^*)^2|\hat{w}|^2
-  \left(\frac{\lambda}{a^2|\hat{w}|^2 + \lambda}\right)
-\]
-\begin{equation}
-  \label{eqn:eq12}
-  = \lim_{\Omega \rightarrow \infty} \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2\lambda
-  \left(\frac{a^2|\hat{w}|^2}{a^2|\hat{w}|^2 + \lambda}\right)
-\end{equation}
-For any $\Omega > 0$,
-\[
-\int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2\lambda\left(\frac{a^2|\hat{w}|^2}{a^2|\hat{w}|^2 + \lambda}\right)
-\le \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2 \lambda
-  \left(\frac{ a|\hat{w}|}{(a^2|\hat{w}|^2 +
-      \lambda)^{\frac{1}{2}}}\right)
-\]
-\[
-  = \sigma \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2 a|\hat{w}|
-  \left(\frac{\sigma }{(a^2|\hat{w}|^2 +\lambda)^{\frac{1}{2}}}\right)
-  \le \sigma \int_{-\Omega}^{\Omega} d\omega \frac{(a^*)^2}{a}
-  |\hat{w}|
-  \le \sigma \frac{(a^*)^2}{a} \|\hat{w}\|_{L^1(\bR)}.
-\]
-Together with the approximate inequality \ref{eqn:eq12}, this
-inequality implies
+    \frac{1}{\alpha^2}\left(\tJase[\kappa;\dl]-\tilde{J}_{\rm MSWI,0,\sigma,\epsilon}[\kappa;\dl]\right)=
+    \|\Te u_{\rm MSWI,0,\sigma,\epsilon}\|^2
+  \]
+  Comparing the normal equation \ref{eqn:eq2} for $\alpha=0$
+  to the definition \ref{eqn:reg}, one sees that  $u_{\rm
+    MSWI,0,\sigma,\epsilon} = u_{\rm MSWI,0,\sigma,0} = \ul$ is
+  independent of $\epsilon$. From
+  Propositions \ref{thm:tl2} and \ref{thm:noisefree}, $T\ul \in
+  L^2(X_{sr} \times \bR)$. For $\delta>0$, choose $t_{\delta} > 0$ so
   that
   \[
-  \|S_{\lambda}[\kappa]u_0[\kappa_{\lambda};d] - d\|^2 = O(\sigma) + O(\lambda(w)).
-\]
-A similar argument leads to the conclusion that
-\[
-  \lambda\|u_0[\kappa_{\lambda};d]\|^2 = O(\sigma) + O(\lambda(w)).
-\]
-Together, these conclusions imply the assertion \ref{eqn:eq11}.
+    \int_{|t| \ge t_{\delta}} \,dt\, |\Te \ul|^2 \le  \int_{|t| \ge
+      t_{\delta}} \,dt\, |T \ul|^2 \le \delta^2
+  \]
+  and $\epsilon_{\delta}>0$ so that
+  \[
+    \int_{|t| \le t_{\delta}} \,dt\, |(T-\Te) \ul|^2 \le \delta^2
+  \]
+  for $0<\epsilon<\epsilon_{\delta}$. Adding up, obtain $\|T\ul
+  -\Te\ul\|^2 \le 5\delta^2$. Since $\delta>0$ was arbitrary, it
+  follos that $\|(\Te -T)\ul\| \rightarrow 0$ as $\epsilon \rightarrow
+  0$..
+\end{proof}
+
+\begin{cor}
+  \label{thm:cor2}
+  With the hypotheses and notation of Theorem \ref{thm:main},
+  \[
+    \lim_{\lambda \rightarrow 0}\lim_{\epsilon \rightarrow 0}\lim_{\alpha \rightarrow 0}
+  \frac{\lambda}{\alpha^2}\left(\tilde{J}_{\rm MSWI,\alpha,r\lambda,\epsilon}[\kappa;\dl]-\tilde{J}_{r\lambda}[\kappa;\dl]\right)=
+  \]
+  \begin{equation}
+    \label{eqn:main1}
+    = \sum_{(\bx_s,\bx_r) \in X_{sr}}
+    M[m,m^*,w_1]^2(\bx_s,\bx_r) |\tau[\kappa](\bx_s,\bx_r) -
+    \tau[\kappa^*](\bx_s,\bx_r)|^2.
+  \end{equation}
+\end{cor}
+
+To derive the analogous relation for AWI, define a family of norms on
+$ L^2(X_{sr} \times \bR)$, depending on $\log \kappa \in M, d \in
+L^2(X_{sr} \times \bR)$, and $\sigma >0$:
+\begin{equation}
+  \label{eqn:awinormdef}
+  \|u\|^2_{\kappa,d,\sigma} = \sum_{(\bx_s,\bx_r) \in X_{sr}}\frac{\|u(\bx_r,\cdot;\bx_s)\|^2}{\|\ul[\kappa;d](\bx_r,\cdot;\bx_s)\|^2}
+\end{equation}
+Define $V_{\kappa,d,\sigma} = L^2(X_{sr} \times \bR)$, equipped with
+the norm $\|\cdot\|_{\kappa,d,\sigma}$, and define $\Tekds$ to be
+$\Te$, with domain $U$ and range $V_{\kappa,d,\sigma}$.
+
+The AWI penalty objective is then
+\begin{equation}
+  \label{eqn:awipen}
+  \JAase[\kappa,u;d] = \|S[\kappa]u-d\|^2 + \sigma\|u\|^2 + \alpha^2
+  \|\Tekds u\|_{\kappa,d,\sigma}^2
+\end{equation}
+As before, define $\uaase[\kappa;d]$ to be the minimizer of
+$\JAase[\kappa,u;d]$ over $u \in U$, and
+\begin{equation}
+  \label{eqn:awipenred}
+  \tJAase[\kappa;d] =\JAase[\kappa,\uaase[\kappa;d];d].
+\end{equation}
+An argument precisely analogous to the proof of Corollary
+\ref{thm:cor2} leads to the conclusion of
+\begin{cor}
+  \label{thm:cor3}
+  With the hypotheses and notations of Theorem \ref{thm:main}, and the
+  definitions \ref{eqn:tj0def} and \ref{eqn:awipenred} with $S=\Sl$,
+  $d=\dl$, and $\sigma=r\lambda$,
+  \[
+  \lim_{\lambda \rightarrow 0}\lim_{\epsilon \rightarrow 0}\lim_{\alpha \rightarrow 0}
+  \frac{1}{\alpha^2}\left(\tilde{J}_{\rm AWI,\alpha,r\lambda,\epsilon}[\kappa;\dl]-\tilde{J}_{r\lambda}[\kappa;\dl]\right)=
+  \]
+  \begin{equation}
+    \label{eqn:cor3}
+    = \sum_{(\bx_s,\bx_r) \in X_{sr}} |\tau[\kappa](\bx_s,\bx_r) - \tau[\kappa^*](\bx_s,\bx_r)|^2.
+  \end{equation}
+\end{cor}
 
 \section{Multiple Arrivals and Cycle-Skipping}
-The geometric asymptotics approximation \ref{eqn:hadamard} holds only
+The geometric asymptotics approximation \ref{eqn:fwdgreen} holds only
 in the ``single arrival'' case, that is, when sources and receivers
@@
 distances, multiple raypaths typically exist, corresponding to
-multiple idenfiable wavefront arrivals. Accordingly, in such settings
+multiple idenfiable wavefront arrivals \cite[]{Whi:82}. Accordingly, in such settings
 the analysis presented in the last section no longer applies, and the
@@
 receiver depths, thus leaving only data that tended to be dominated by
-single arrivals. \cite{HuangSymes:Geo17} illustrate the failure of the
+single arrivals. \cite{HuangSymes:Geo17} and
+\cite{SymesChenMinkoff:24} illustrate the failure of the
 version of MSWI reported here to avoid cycle-skipping in inversion of
-strongly refracting models, and ameliorate the this failure to some
-extent by Tihonov regularization. They observe that the mapping from
-adaptive filter or extended source to data (represented by the opertor
-$S_{\lambda}[\kappa]$ in the discussion above) may not have a well-behaved inverse in
-the presence of multiple wavefront arrivals. \cite{Yongetal:GJI23}
+strongly refracting models, using frequency-domain implementation of
+the basic definition \ref{eqn:mswidef} and time-domain implementation
+of the reduced penalty function \ref{eqn:mswipenred} respectively. \cite{Yongetal:GJI23}
 have shown that precisely the same issue afflicts AWI, and have
@@
 A simple calculation reveals the phenomenon discussed in all of these
-works. The analogue of the geometric asymptotics for a source-receiver
+works. I will argue formally, via the leading term in the Hadamard
+approximation of the Green's function \ref{eqn:green}, generalized to
+multiple arrival times. The analogue of the geometric asymptotics for a source-receiver
 pair with connected by multiple rays is (once again suppressing the
@@
   \label{eqn:multi}
-  F[\kappa,w](t) \approx \sum_{i \ge 0} a_i H^{p_i}w(t-\tau_i).
-\end{equation}
-The amplitudes $\{a_i\}$ and arrival times $\{\tau_i\}$ depend on $m$, as
+  F[\kappa](t) \approx \sum_{i \ge 0} a_i H^{p_i}w(t-\tau_i).
+\end{equation}
+The amplitudes $\{a_i\}$ and arrival times $\{\tau_i\}$ depend on $\kappa$, as
 before, through ray-tracing. $H$ is the Hilbert transform, and $p_i$
@@
 $T^*, T_1^*$ defined as above but with $a^*_i$ in place of $a_i$ and so
-on. From the observation \ref{eqn:notl2},
-\[
-  d = F[\kappa_*]= S_{\lambda}[\kappa] u
+on:
+\[
+  d = F[\kappa_*]= S[\kappa] u
 \]
@@
 $\tau_1-\tau_0^*$ do not tend to zero, though they do tend to the same
-limit as $m \rightarrow \kappa^*$ (as do the amplitude quotients), and the
-Hilbert transform powers $p_1, p_1^*$ are the same as soon as $m$ is
+limit as $\kappa \rightarrow \kappa^*$ (as do the amplitude quotients), and the
+Hilbert transform powers $p_1, p_1^*$ are the same as soon as $\kappa$ is
 sufficiently close to $\kappa^*$.
@@
 preceding section, this approximate Dirac was $g$, so use the same
-notation here. The regularized version of $u(t)$ in
-\ref{eqn:notl2explicit} contains summands with multiples of $g(t-(\tau^*_1-\tau_0))$
+notation here. The regularized version of $u(t)$ (solution of the
+least squares problem \ref{eqn:reg}) contains summands with multiples of $g(t-(\tau^*_1-\tau_0))$
 and $g(t-(\tau_1-\tau^*_0))$. These are approximately scaled by
 $\tau^*_1-\tau_0$ and $\tau_1-\tau^*_0$ respectively in the
-definitions \ref{eqn:MSWI} and \ref{eqn:tJz} of the MSWI and AWI
+definitions \ref{eqn:mswidef} and \ref{eqn:awidefbis} of the MSWI and AWI
 objectives. For other choices of objective, they contribute similarly
@@
 %%%%%%%%%%%%%%%%%%
-Note that this main result effectively bounds the optimal mean-{\em
-  square} traveltime error achieved by AWI by the pulse width,
-hence the RMS traveltime error by the square-root of the pulse
-width. Pulse width is related to RMS wavelength by the Heisenberg
-inequality, and in many examples these differ by a modest factor.
-Thus error estimate is suboptimal from the point of view
-of FWI, which requires something like an initial model estimate
-predicting a half-wavelength traveltime error. This suboptimality
-arises from the use of pre-whitening (Tihonov regularization) in the
-definition of the AWI objecive function \cite[]{Warner:16}. It is
-possible that a different regularization strategy or assumptions on
-the wavelet could yield a more optimal error estimate.
 

### 2024-11-03 repro/papers/awi/paper.tex
rearrange

@@
-\title{Adaptive Waveform Inversion for Transmitted Wavefields}
+\title{Adaptive Waveform Inversion for Transmitted Wave Data}
 \author{William W. Symes}
@@
 body, a manufactured artifact,...) from remote measurements of waves
-propagating through it. Over the last several decades, this task has come to
-be carried out by minimizing a measure of the difference between computational
-prediction of wave data versus actual observed
-data. The mechanical parameters to be inferred appear as
+propagating through it. These mechanical parameters appear as
 coefficients in systems of partial differential equations governing
-the wave motion, and discrete representations of these parameters form
-the unknown vectors in the optimization problems thus generated.
-
-While widely used in recent years, FWI exhibits a pathology when wave
-velocities or their parametric equivalents are amongst the parameters
-to be estimated: the excessive sensitivity of predicted waveforms to
-changes in velocity leads to oscillations in the difference measures,
-especially for the most commonly used square-integral measure. Local
-optimization algorithms (relatives of Newton;'s method) are the only practical
-methods to generate sequences of approximate optimal
-parameters. However the oscillations in the difference measures lead
-these sequences to stagnate at parameter distributions
-far from the optimum, and thus do not extract the maximum information
-from the data \cite[]{VirieuxOperto:09}.
-
-Various modifications of straighforward best-fit FWI have been
-suggested to overcome this ``cycle-skipping'' pathology. Many of these
-modifications employ {\em model extension}, that is, the introduction
-of a larger space of model parameters than derives from basic
-physics. An example is the {\em source-receiver} extension \cite[]{HuangSymes:17}, in which
-source waveforms are permitted to depend (unphysically) on receiver
-locations. The focus of this paper is Adaptive Waveform Inversion
-\cite[]{Warner:16} (AWI), a modification of FWI using the
-source-receiver extension. AWI has been successfully applied to industry-scale seismic data
-\cite[]{GuaschWarnerRavaut:GEO19,Warneretal:SEG21} and shows promise for
-ultrasonic imaging of the the human brain
-\cite[]{Guaschetal:NPJDM20}.
-
-While these and other works discuss heuristic justifications for AWI
-and many details of implementation, a clear explanation for its
-effectiveness is so far missing from the literature. In this short
-note, I treat a special case of FWI, the recovery of smooth acoustic
+wave propagation, and numerical solutions of these wave equations can
+be used to predict experimental data {\em ab initio}. FWI consists in
+updating the parameters to minimize a numerical measure of  the difference between computationally
+predicted wave data and actual observed
+data. If wave velocities or their parametric equivalents are amongst the parameters
+to be estimated, such numerical measures tend to oscillate, due to:
+the excessive sensitivity of predicted waveforms to changes in wave velocity.
+%especially for the most commonly used square-integral measure ($L^2$
+%norm).
+These oscillations cause iterative local optimization algorithms (relatives of
+Newton;'s method, the only practical algorithmic choices) to stagnate
+far from good data fit \cite[]{VirieuxOperto:09}, unless the initial
+parameter estimates are of very high quality.
+
+Adaptive Waveform
+Inversion \cite[]{Warner:16} (AWI) is a modification of FWI, designed to overcome this
+``cycle-skipping'' pathology. AWI has been successfully applied to
+industry-scale seismic data
+\cite[]{GuaschWarnerRavaut:GEO19,Warneretal:SEG21} and shows promise
+for ultrasonic imaging of the the human brain
+\cite[]{Guaschetal:NPJDM20}. These and other works discuss heuristic justifications for AWI
+and many details of implementation. However, a clear explanation for its
+effectiveness is so far missing from the literature.
+
+This paper supplies such an explanation for a special case of AWI, the recovery of smooth acoustic
 medium parameters from data generated at localized energy sources,
@@
 reasons for normalizing the objective previoiusly advanced in the
-literature \cite[]{Warner:16,LiAlkhalifah:21}. Another modification of
-FWI based on source-receiver extension, Matched Source Waveform Inversion (MSWI)
+literature \cite[]{Warner:16,LiAlkhalifah:21}. Another similar modification of
+FWI, Matched Source Waveform Inversion (MSWI)
 \cite[]{HuangSymes2015SEG,HuangSymes:Geo17} lacks the AWI
@@
 
-The conclusions of this study depend on a number of limitations on the
-acoustic model of wave propagation and the formulation of the related
-inverse problem. The principal constraints are:
-\begin{itemize}
-\item energy sources are modeled as isotropic point radiators, with
-  known time dependence vanishing outside a finite time window;
-\item mechanical parameter fields (coefficients in the acoustic wave
-  equation) are smooth, that is, infinitely differentiable as
-  functions of position, and the material density is known and
-  position-independent: only the bulk modulus (or wave velocity) is
-  to be determined;
-\item waves propagate throughout Euclidean space-time - that is, no
-  physical boundaries are modeled;
-\item the data of the inverse problem are the restrictions of the
-  pressure field to a finite number of points in space (receiver
-  positions), for a finite number of source positions and all time;
-\item local energy decays exponentially - that is, the pressure and
-  all of its derivatives tend to zero at an exponential rate as $t
-  \rightarrow \infty$,
-  uniformly in any bounded spatial domain.
-\item only one ray of geometric
-  acoustics connects each source-receiver location pair (single arrivals)
-\end{itemize}
-
-These constraints are discussed in detail in the next section. Some
-could certainly be relaxed. For instance, exponential decay of local
-energy implies that the finite time duration of actual measured data
-can be treated as a perturbation of the infinite time data setting
-posed here. The exception is the single-arrival hypothesis: it is
-essential, as without it, the link between the travel time and
-extended waveform inversion is lost. This is a feature of the
-source-receiver extension, not of a particular objective function
-(MSWI, AWI,...) formulated in terms of it. This extension treats each receiver signal
-independently. If only a single identifiable waveform arrives at
-each receiver, AWI (and similar approaches) can implicitly identify
-its arrival time, and the link to travel time inversion is
-established. If distinct waveforms are present at multiple times,
-the single receiver signal does not determine which waveform in the
-predicted signal should align with which waveform in the data to be
-fitted. This ``crosstalk'' ambiguity is analyzed in the fifth
-section, and afflicts all approaches based on the source-receivre
-extension.
-
-In any case, these hypotheses enable a simple and clear
-analysis of AWI, and seem a reasonable starting point for further
-investigation.
-
-After establishing the class of problems to be studied and notations
-for their description in the following section, I present the
-asymptotic analysis leading to the main conclusion in the next. In
+The next section establishes the class of problems to be studied and notations
+for their description, defines the AWI and MSWI objective functions,
+and states the main results of this paper. in the following section, I present the
+asymptotic analysis leading to the main results. In
 effect, AWI regularizes or relaxes the FWI problem. For many inverse
@@
 
-\section{Adaptive Waveform Inversion}
+\section{Main Results}
 
@@
 active source acoustic model of data generation in Euclidean space
-$\bR^3$. In this section, I will review a version of this model, place
-some useful constraints on energy sources appearing in it, and define
-AWI as an optimization problem.
-
-\subsection{Acoustic Modeling of Transmitted Waves}
+$\bR^3$.
 
@@
 
+Throughtout the following discussion, it is assumed that
+$X_{sr}$ is finite.
+
 The pressure and velocity fields $p({\bf x},t;{\bf x}_s)$,
@@
 (reciprocal of the density $\rho$), and source wave
-form $f(t)$ through the acoustic system
+form $f \in C_0^{\infty}(\bR)$ through the acoustic system
 \begin{eqnarray}
@@
 assume in this study that the material density $\rho$ is spatially
-homogeneous and known. In some settings, eg. the earth's
-upper crust, the material density $\rho$ exhibits substantially less variance than the
-bulk modulus, Density is however spatially variable in most geological
-and biomedical materials, and could be included amongst the parameters
-to be estimated.
+homogeneous and known.
+%In some settings, eg. the earth's
+%upper crust, the material density $\rho$ exhibits substantially less variance than the
+%bulk modulus, Density is however spatially variable in most geological
+%and biomedical materials, and could reasonabliy be included amongst the parameters
+%to be estimated.
 
@@
 Under these hypotheses, the symmetric hyperbolic system \ref{eqn:awe}
-has a unique distribution
-solution if $f \in \cal{E}'(\bR)$ (for example \cite{Lax:PDENotes},
+has a unique distribution solution (for example \cite{Lax:PDENotes},
 Theorem 7.4). The Propagation of
-Singularities Theorem (\cite{Tay:81}, Theorem VI.2.1) and
-the pull-back property of distributions (\cite{Dui:95}, Proposition 1.3.3)
-imply that if $f \in \cal{E}'(\bR)$ and $(\bx_s,\bx_r) \in X_{sr}$,
-then the trace $p(\bx_r, \cdot;\bx_s)$ on $\{(\bx_r,t): t \in \bR\}$ is well-defined,
-as a distribution vanishing for sufficiently large
-negative $t$. Define the {\em modeling operator}
+Singularities Theorem (\cite{Tay:81}, Theorem VI.2.1) %and
+%the pull-back property of distributions (\cite{Dui:95}, Proposition
+%1.3.3)
+imply that $p(\cdot,\cdot;\bx_s)$ is smooth away from $\{\bx_s\}
+\times \bR$, so that the traces $p(\bx_r, \cdot;\bx_s)$ on
+$\{(\bx_r,t): t \in \bR\}$ are well-defined.
+
+Define the {\em modeling operator}
 \begin{equation}
   \label{eqn:modop}
-  F: C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)
-  \rightarrow X_{sr} \times {\cal D}'(\bR)
+  F: exp(M)  \rightarrow \Pi_{X_{sr}} {\cal D}'(\bR)
 \end{equation}
@@
 source-receiver position pairs $X_{sr}$, source wavelet
-$f \in {\cal E}'(\bR)$,  density $\rho$, and data traces
-$d:  X_{sr} \rightarrow \cal{D}(\bR)$, find a bulk modulus $\kappa$
+$f \in C_0^{\infty}(\bR)$,  density $\rho$, and data traces
+$d \in \Pi_{X_{sr}} C^{\infty}(\bR)$, find a bulk modulus $\log
+\kappa$
 (or equivalently wave velocity $c=\sqrt{\kappa/\rho}$)
 so that $F[\kappa] \approx d$ in some sense. If
-$d$ and $F[\kappa]$ lie in a normed subspace of ${\cal D}'(\bR)$ for each
+$d(\bx_r,\cdot;\bx_s)$ and $F[\kappa](\bx_r,\cdot;\bx_s)$ lie in a normed subspace of ${\cal D}'(\bR)$ for each
 $(\bx_s,\bx_r) \in X_{sr}$, then the
@@
 
-%In practice, recording takes place over a finite time interval
-%$[0,t_d]$. For convenience, assume that this interval is the same for
-%all receivers. Let $\phi \in C_0^{\infty}(0,t_d)$ be a window function
-%supported in this interval. Additional constraints on $t_d$ and $\phi$
-%will be be specified below.
+The analysis to follow requires two further constraints on the domain
+$M \subset C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)$ of $F$. The first is
+the {\em Exponential Decay} assumption:
+\begin{quote}
+  For any $n > 0$, $f \in {\cal E}'(\bR)$,
+  there exist $K >
+  0, T \in \bR,$ and $\delta>0$ so
+  that for any $k$ with $0 \le k \le n$, $(\bx_s,\bx_r) \in
+ X_{sr},$ and any $\log \kappa \in M$, the solution $(p,\bv)$ of \ref{eqn:awe} is smooth
+  for $t>T$ and satisfies
+  \[
+    |\partial_t^k p(\bx_r,t;\bx_s)| \le K e^{-\delta t}, \, t > T.
+  \]
+\end{quote}
+SInce $X_{sr}$ is bounded, the property described in this assumption
+is a special case of {\em local energy decay}, a much studied topic. The {\em
+  non-trapping} hypothesis is sufficient to ensure local energy decay:
+all rays of geometric asymptotics (that is, geodesics for the Riemannian metric associated
+to the system \ref{eqn:awe}) escape to infinity (\cite{Hristova:09},
+Theorem 2, \cite{EgorovShubin}, Theorem 2.104, p. 192). That is,
+exponential decay follows from a characteristic of $\kappa$.
+
+The second assumption explicitly involves a constraint on ray geometry
+(therefore on $\kappa$). A {\em normal neighborhood} of
+$\bx_s \in \bR^3$ is an open subset $\Omega$ of $\bR^3$ 
+within which every point is connected to $\bx_s$ by  a unique ray  (\cite{Friedlander:75},
+p. 16). Therefore, to every point $\bx \in \Omega$ is associated a
+unique time of travel (geodesic distance) $\tau(\bx_s,\bx)$ from $\bx_s$ to $\bx$.
+The {\em Single Arrival} assumption is
+\begin{quote}
+  $\Omega\subset \bR^3$ 
+and $M \subset C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)$ are bounded open sets satisfying 
+\begin{itemize}
+  \item[G1. ] $X_{sr} \subset \Omega \times \Omega$;
+  \item[G2. ] for every source point $\bx_s\in X_{s}$ and $\log \kappa \in M$, $\Omega$ is a
+    normal neighborhood of $\bx_s$.
+  \item[G3. ] for every source point $\bx_s \in X_{s};$
+    $\log \kappa \in M$, ray ${\bf X}:\bR^+\rightarrow \bR^3$
+    with ${\bf X}(0)=\bx_s$ and any $t>0$, either $\{{\bf X}(s): 0\le s \le t\} \subset \Omega$
+    or $\{{\bf X}(s): t < s < \infty\} \subset \bR^3 \setminus \Omega$.
+  \end{itemize}
+\end{quote}
+Assertion G3 is the {\em no-return} assumption: once a ray leaves
+$\Omega$, its subsequent intersection with $\Omega$ is null.
+
+Thanks to exponential decay, each trace $p(\bx_r,\cdot;\bx_s)$ is in $L^{p}(\bR)$ for $p
+\ge 1$.
+Denote by $L^p(X_{sr} \times \bR)$ the product space $l^p(X_{sr},L^p(\bR))$
+with the usuall product norm.
+
+Accordingly, FWI may be posed as a least-squares
+problem: given data $d \in
+L^2(X_{sr}\times \bR)$, find $\kappa$ to minimize
+\begin{equation}
+  \label{eqn:fwi}
+  J_{\rm FWI}[\kappa;d]= \frac{1}{2}\|F[\kappa]-d\|^2_{L^2(X_{sr} \times \bR)},
+\end{equation}
+As mentioned earlier, optimization algorithms related to Newton's method, applied to
+$J_{\rm FWI}$, tend to stagnate far from global minimizers if
+the initial choice of $\kappa$ does not predict the arrival times of
+waves correctly within a half-wavelength or so (the meaning of
+``wavelength'' in this context will be explained below)
+\cite[]{GauTarVir:86,VirieuxOperto:09}.
+
+AWI modifies the least-squares FWI problem by interposing a convolution operator or
+{\em filter} $u \in L^2(X_{sr} \times \bR)$. That is, the residual
+data $F[\kappa]-d$ is replaced by the filtered residual data $F[m]*u-d$.
+Since $F[\kappa] \in L^1(X_{sr}\times \bR)$, the filtererd residual
+makes sense.
+The filter $u$ is intended to shift localized wave energy in $F[m]$ to
+match that present in $d$. If on the other hand $u(\bx_r,t;\bx_s) =
+\delta(t)$ for each $(\bx_x,\bx_r) \in X_{sr}$, then $tu = 0$. This
+observation motivates the definition of the AWI objective function
+\begin{equation}
+  \label{eqn:awidefmain}
+  J_{\rm AWI,\sigma}[\kappa;d] = \sum_{(\bx_s,\bx_r) \in X_{sr}}
+  \frac{\|Tu_{\sigma}[\kappa;d](\bx_r,\cdot;\bx_s)\|^2}{\|u_{\sigma}[\kappa;d](\bx_r,\cdot;\bx_s)\|^2}.
+\end{equation}
+in which $Tu(\bx_r,t;\bx_s) = tu(\bx_r,t;\bx_s)$ and
+\begin{equation}
+  \label{eqn:reg}
+  u_{\sigma}[\kappa;d] = \mbox{argmin}_{u \in L^2(X_{sr} \times \bR)} (\|F[\kappa]*u-d\|^2 + \sigma\|u\|^2)
+\end{equation}
+The objective function for the closely related MSWI approach,
+mentioned in the introduction, is  
+\begin{equation}
+  \label{eqn:mswidefmain}
+  J_{\rm MSWI,\sigma}[\kappa;d] = \sum_{(\bx_s,\bx_r) \in X_{sr}}
+  \|Tu_{\sigma}[\kappa;d](\bx_r,\cdot;\bx_s)\|^2.
+\end{equation}
+Note that while the Lax-Milgram Theorem (\cite{Yosida}, section
+III.7) ensures that $u_{\sigma}[\kappa;d] \in L^2(X_{sr}\times \bR)$,
+it is not immediiately obvious that $Tu_{\sigma}[\kappa;d] \in L^2(X_{sr}\times \bR)$.
+
+Asymptotic analysis of AWI and MSWI is based on a family of source
+wavelet: for $\lambda > 0$,
+\begin{equation}
+  \fl(t)= \lambda^{\frac{1}{2}}f_1\left(\frac{t}{\lambda}\right).
+\end{equation}
+The relation between the dimensionless scale parameter $\lambda$ and
+RMS wavelength will be explained below. Correspondingly, denote the
+modeling operator with $f=\fl$ by $\Fl$.
+
+The main result of this analysis is
+
+\begin{theorem}
+  \label{thm:main}
+  Assume that the Exponential Decay and Single Arrival assumptions
+  hold for $X_{sr} \subset \Omega \times \Omega, M \subset C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)$.
+  Select $r >0$ with units of
+  pressure$^2 \times$time$^2$ and $\log \kappa^* \in M$. For $0 <
+  \lambda $, set
+  $\dl=\Fl[\kappa^*]$ (``noise-free data''), and  define $\ul[\kappa,\dl]$
+  to be the solution of 
+  equation \ref{eqn:reg} with  $\sigma=r\lambda$, $F=\Fl$ and $d=\dl$..
+  There exist $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
+  \rightarrow \bR^+$ with $C_* \le
+  W \le C^*$ so that for
+  $\log \kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$, $Tu_{\sigma}[\kappa;\dl](\bx_r,\cdot;\bx_s) \in L^2(\bR)$, and 
+  \begin{equation}
+    \label{eqn:mswitt}
+    |\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|- \lambda^{-\frac{1}{2}} (W[\kappa^*,\kappa](\bx_s,\bx_r))^{\frac{1}{2}}|\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|| \le C
+  \lambda^{\frac{1}{2}}
+\end{equation}
+and for sufficiently small $\lambda$, 
+\begin{equation}
+  \label{eqn:awitt}
+  \left|\frac{\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|}{\|\ul[\kappa,\dl](\bx_r,\cdot;\bx_s)\|}-
+    |\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|\right| \le C \lambda
+  \end{equation}
+\end{theorem}
+
+Summation over the source-receiver positions in $X_{sr}$ yields
+\begin{cor}
+  \label{thm:cor1}
+  In addition to the assumptions of Theorem \ref{thm:main}, define $J_{\rm
+    AWI,\sigma}$ and $J_{\rm MSWI,\sigma}$ by equations
+  \ref{eqn:awidefmain}, \ref{eqn:mswidefmain} respectively. Then
+\begin{equation}
+\label{eqn:cor1mswi}   
+\lim_{\lambda \rightarrow 0} \lambda J_{\rm MSWI,r\lambda}[\kappa;\dl]
+= \sum_{(\bx_s,\bx_r) \in X_{sr}} W[\kappa^*,\kappa^*]^2(\bx_s,\bx_r) |\tau[m](\bx_s,\bx_r) - \tau[m^*](\bx_s,\bx_r)|^2.
+\end{equation}
+and
+\begin{equation}
+\label{eqn:cor1awi}   
+\lim_{\lambda \rightarrow 0} J_{\rm AWI,r\lambda}[\kappa;\dl]
+= \sum_{(\bx_s,\bx_r) \in X_{sr}}  |\tau[m](\bx_s,\bx_r) - \tau[m^*](\bx_s,\bx_r)|^2.
+\end{equation}
+\end{cor}
 
@@
 
-The arguments to follow depend an additional assumption about solutions of
-the wave equation \ref{eqn:awe2}, restricted to $X_{sr}\times \bR$:
-\begin{itemize}
-\item[G4. ] For any $n > 0$, $f \in {\cal E}'(\bR)$, there exist $K >
-  0, T \in \bR,$ and $\delta>0$ so
-  that for any $k$ with $0 \le k \le n$, $(\bx_s,\bx_r) \in
- X_{sr},$
-  and any $\log \kappa \in M$, the solution $p$ of \ref{eqn:awe2} is smooth
-  for $t>T$ and satisfies
-  \[
-    |\partial_t^k p(\bx_r,t;\bx_s)| \le K e^{-\delta t}, \, t > T.
-  \]
-\end{itemize}
-Condition G4 is a special case of {\em local energy decay}, a much
-studied topic. Properties which imply G4 are in turn implied by the
-{\em non-trapping} hypothesis on the model $\log \kappa \in M$: all rays of
-geometric asymptotics escape to infinity (\cite{Hristova:09}, Theorem
-2, \cite{EgorovShubin}, Theorem 2.104, p. 192).
-
 Under conditions G1 - G4, the construction described by
@@
 
-\subsection{Notational Conventions for Norms and Bounds}
-The Banach spaces
-$l^p(X_{sr},L^p(\bR))$ will be abbreviated as $L^p(X_{sr} \times
-\bR)$. The norm in a Banach space $B$ will be denoted
-$\|\cdot\|_{B}$. The norm symbol
-$\|\cdot\|$ without subscript will be understood to mean either
-$\|\cdot\|_{L^2(\bR)}$ or $\|\cdot\|_{L^2(X_{sr}\times
-  \bR)}$, as context indicates.
-
-In a statement pertaining to all
-$\kappa \in M$, the letter $C$, or sometimes $C_k,
-k=0,1,2,...$ or
-$C_*,C^*$, will designate a constant (or constants) independent of
-$\kappa$.
-
-Generally, constants denoted $C$ etc. will absorb any dimensional
-scale factors necessary to make sense of the expressions in which they
-appear. I will call out those few cases in which explicit use of
-dimensinoal scale factors is necessary to avoid irretrievable
-dimensional confusion.
-
 \subsection{FWI via Least Squares}
@@
 \end{equation}
-Suppose that the source wavelet $f$ satisfies
-\begin{itemize}
-\item[F1. ] $f \in C^{\infty}_0(\bR)$;
-\item[F2. ] $f$ is odd, $f(t) = -f(-t)$.
-\end{itemize}
-Then equation \ref{eqn:fwdgreen}, assumptions F1, F2, bounds
+
+For the remainder of this paper, source wavelet $f$ will be assumed
+smooth, of compact support: $f \in C^{\infty}_0(\bR)$. Less stringent constraints on $f$ would
+support the results below, but this is the simplest choice. Also, with
+a few notable exceptions \cite[]{SondRes:83}, source wavelets that
+adequately model physical experimentss may be take to be smooth of
+compact support, because the physical phenomena that inject energy
+into the acoustic field are transient, and the data is sampled at a
+finite rate and either analog
+or digitally filtered (or both) to well below Nyquist frequency.
+
+Then equation \ref{eqn:fwdgreen}, bounds
 \ref{eqn:best} and Young's inequality implies that $F[\kappa] \in
@@
 This observation justfies posing FWI as a least-squares problem: given $d \in
-L^2(X_{sr}\times \bR)$, $f \in H^1_0(\bR)$, and $\rho_s>0$, find $\kappa$ to minimize
+L^2(X_{sr}\times \bR)$, $f \in H^1_0(\bR)$, find $\kappa$ to minimize
 \begin{equation}
@@
 
+\subsection{A Note on Notations  for Norms and Bounds}
+The Banach spaces
+$l^p(X_{sr},L^p(\bR))$ will be abbreviated as $L^p(X_{sr} \times
+\bR)$. The norm in a Banach space $B$ will be denoted
+$\|\cdot\|_{B}$. The norm symbol
+$\|\cdot\|$ without subscript will be understood to mean either
+$\|\cdot\|_{L^2(\bR)}$ or $\|\cdot\|_{L^2(X_{sr}\times
+  \bR)}$, as context indicates.
+
+In a statement pertaining to all
+$\kappa \in M$, the letter $C$, or sometimes $C_k,
+k=0,1,2,...$ or
+$C_*,C^*$, will designate a constant (or constants) independent of
+$\kappa$.
+
+Generally, constants denoted $C$ etc. will absorb any dimensional
+scale factors necessary to make sense of the expressions in which they
+appear. I will call out those few cases in which explicit use of
+dimensinoal scale factors is necessary to avoid irretrievable
+dimensional confusion.
+
 \subsection{Tihonov Regularization}
@@
 $S[\kappa]$ is a convolution operator, solving $S[\kappa]u=d$ for $u$
-amounts to deconovolving $d$ by $F[\kappa]$..  Under the conditions
-W1-W3 on the wavelet $w$, $F[\kappa]$ does not have a continuous
+amounts to deconovolving $d$ by $F[\kappa]$..  Under the standing
+assumption that $f \in C^{\infty}_0(\bR)$, $F[\kappa]$ does not have a continuous
 (stable) convolution inverse, and $S[\kappa]$ is not surjective.
@@
   \label{thm:snorm}
-  Under the assumptions G1-G4 leading to \ref{eqn:fwdgreen} and
-  the properties F1 and F2, the expressions \ref{eqn:sgood} and \ref{eqn:sadjgood} define 
+  Under the assumptions G1-G4 leading to \ref{eqn:fwdgreen}, the
+  expressions \ref{eqn:sgood} and \ref{eqn:sadjgood} define
   bounded operators on
@@
   \label{thm:unorm}
-  Under the assumptions G1 - G4 and F1, F2 justifying the
+  Under the assumptions G1 - G4 justifying the
   representation \ref{eqn:fwdgreen}, the least squares problem
@@
   Suppose that $d, Td \in L^2(X_{sr} \times \bR)$. Then
-  $Tu_{\sigma}[\kappa;d]  \in L^2(X_{sr} \times \bR)$, and
-  \[
-    \|Tu_{\sigma}[\kappa;d]\| \le
+  $Tu_{\sigma}[\kappa;d]  \in L^2(X_{sr} \times \bR)$, and for each
+  $(\bx_s,\bx_r) \in X_{sr}$,
+  \[
+    \|Tu_{\sigma}[\kappa;d](\bx_r,\cdot;\bx_s)\| \le
   \]
@@
       \right)\|d\| \\ 
-    + \|\partial_t  f\|_{L^1(\bR)}(\|d\| +\|Td\|)\right).
+    + \|\partial_t  f\|_{L^1(\bR)}(\|d (\bx_r,\cdot;\bx_s)\| +\|Td (\bx_r,\cdot;\bx_s)\|)\right).
   \end{equation}
@@
   temporarily, also from the ingredients $\tau, a, $ and $b$ of the
-  Hadamard decomposition.
+  Hadamard decomposition. The expressions to follow pertain to
+  each trace, that is, for each $(\bx_s,\bx_r) \in X_{sr}$, so also drop these
+  from the notation also.
 
@@
   \[
-    S^TS =
+    S^TS u=
     \rho^2(a\delta  +  (\check{b}\check{H})) *\delta_{-\tau}*\partial_t
-    f * (a\delta  +  (bH)) *\delta_{\tau}*\partial_t f*
+    f * (a\delta  +  (bH)) *\delta_{\tau}*\partial_t f*u
   \]
@@
   \]
-  $T$ is also a bounded map on $\Pi_{X_{sr}} {\cal S}(\bR)$, as is
+  $T$ is also a bounded map on $ {\cal S}(\bR)$, as is
   its commutator with $S^TS$, and 
@@
   Referring to the exponential decay assumption \ref{eqn:best}
-  and properties F1, F2 of $f$, 
+  and the standing assumption $f\in C_0^{\infty}(\bR)$, 
   observe that all summands on the RHS of identity \ref{eqn:tsscomm} are in
-  $L^1(X_{sr} \times \bR)$, so $[T,S^TS]$ extends to a bounded
-  operator on $L^2(X_{sr}\times \bR)$, with
+  $L^1(\bR)$, so $[T,S^TS]$ extends to a bounded
+  operator on $L^2(\bR)$, with
   \begin{equation}
@@
   \]
-  Young's inequality shows that the RHS lies in $L^1(X_{sr}\times
-  \bR)$, and that there exists $C$ independent of $\kappa$ so that
+  Young's inequality shows that the RHS lies in $L^1(\bR)$, and that there exists $C$ independent of $\kappa$ so that
   $[T,S^T]$ obeys the same estimate as does $S$ (Lemma
@@
   Suppose that $\kappa \in M$. Then for $p \ge 1$, $F[\kappa], TF[\kappa]
-  \in L^p(X_{sr} \times \bR)$, and there exist $C_0, C_1, C_2 >0$ so that
+  \in L^p(X_{sr} \times \bR)$, and there exist $C_0, C_1, C_2 >0$ so
+  that for each $(\bx_s,\bx_r) \in X_{sr}$,
   \begin{eqnarray}
     \label{eqn:dnorms}
-    C_1\|\partial_tf\|_{L^p(\bR)} - C_2\|f\|_{L^p(\bR)} \le \|F[\kappa]\|_{L^p(\bR)} & \le & C_0\|\partial_tf\|_{L^p(\bR)}, \nonumber \\
-    \|TF[\kappa]\|_{L^p(\bR)} & \le & C_0\|\partial_tf\|_{L^p(\bR)}.
+    C_1\|\partial_tf\|_{L^p(\bR)} - C_2\|f\|_{L^p(\bR)} \le \|F[\kappa](\bx_r,\cdot;\bx_s)\|_{L^p(\bR)} & \le & C_0\|\partial_tf\|_{L^p(\bR)}, \nonumber \\
+    \|TF[\kappa](\bx_r,\cdot;\bx_s)\|_{L^p(\bR)} & \le & C_0\|\partial_tf\|_{L^p(\bR)}.
   \end{eqnarray}
@@
 \end{equation}
-The properties F1 and F2 assumed for $f$ are equivalent to these
-properties of $w$:
+Then
 \begin{itemize}
 \item[W1. ] $w \in C^{\infty}_0(\bR)$;;
-\item[W2. ] $w$ is even;
-\item[W3. ] $\int \, w = 0$.
+\item[W2. ] $\int \, w = 0$.
 \end{itemize}
@@
 \end{equation}
-The  ``mother wavelet'' $w_1$ is assumed to have properties W1 - W3, and
+The  ``mother wavelet'' $w_1$ is assumed to have properties W1, W2, and
 all wavelets $\wl$ in this family
@@
 
-The main result of this investigation is the
-
-\begin{theorem}
-  \label{thm:main}
-  Assume assertions G1 - G4 on $M$ and $X_{sr}$, W1 - W3 on the
-  ``mother wavelet'' $w_1$. Select $r >0$ with units of
-  pressure$^2 \times$time$^2$ and $\log \kappa^* \in M$. For $0 <
-  \lambda \le 1$, set
-  $\dl=\Fl[\kappa^*]$ (noise-free data), and  define $\ul[\kappa,\dl]$ by
-  equation \ref{eqn:reg} with  $\sigma=r\lambda$, $S=\Sl$ and $d=\dl$..
-  There exist $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
-  \rightarrow \bR^+$ with $C_* \le
-  W \le C^*$ so that for
-  $\log \kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$,, 
-  \begin{equation}
-    \label{eqn:mswitt}
-    |\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|- \lambda^{-\frac{1}{2}} (W[\kappa])^{\frac{1}{2}}|\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|| \le C
-  \lambda^{\frac{1}{2}}
-\end{equation}
-and for sufficiently small $\lambda$, 
-\begin{equation}
-  \label{eqn:awitt}
-  \left|\frac{\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|}{\|\ul[\kappa,\dl](\bx_r,\cdot;\bx_s)\|}-
-    |\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|\right| \le C \lambda
-  \end{equation}
-\end{theorem}
-
-Sum over the source-receiver positions in $X_{sr}$ to obtain
-\begin{cor}
-  \label{thm:cor1}
-  In addition to the assumptions of Theorem \ref{thm:main}, define $J_{\rm
-    AWI,\sigma}$ and $J_{\rm MSWI,\sigma}$ by equations
-  \ref{eqn:awidefbis}, \ref{eqn:mswidef} respectively. Then
-\begin{equation}
-\label{eqn:cor1mswi}   
-\lim_{\lambda \rightarrow 0} \lambda J_{\rm MSWI,r\lambda}[\kappa;\dl]
-= \sum_{(\bx_s,\bx_r) \in X_{sr}} M[m,m^*,w_1]^2(\bx_s,\bx_r) |\tau[m](\bx_s,\bx_r) - \tau[m^*](\bx_s,\bx_r)|^2.
-\end{equation}
-and
-\begin{equation}
-\label{eqn:cor1awi}   
-\lim_{\lambda \rightarrow 0} J_{\rm AWI,r\lambda}[\kappa;\dl]
-= \sum_{(\bx_s,\bx_r) \in X_{sr}}  |\tau[m](\bx_s,\bx_r) - \tau[m^*](\bx_s,\bx_r)|^2.
-\end{equation}
-\end{cor}
-
 \subsection{The leading term and travel time}
@@
 \ref{eqn:awidefbis} of $J_{\rm AWI,\sigma}$, set $\sigma = r\lambda$
-and replace the data $\dl$ with
+and replace the data $d$ with
 $\dla = \Fla[\kappa^*]$, $S[\kappa]$ with $\Sla[\kappa]$, and $\ul$
@@
 \]
-the solution of which is
+the solution of which is (omitting $\kappa, \dl$ from the arguments,
+as well as $\bx_s,\bx_r$):
 \begin{equation}
   \label{eqn:hatuf}
-\hat{u}_{0,\lambda}[\kappa;\dla] = \frac{a^*}{a}\hat{g}_{\lambda,\frac{r\lambda}{a^2}} e^{i\omega(\tau^*-\tau)}
+\hat{u}_{0,\lambda} = \frac{a^*}{a}\hat{g}_{\lambda,\frac{r\lambda}{a^2}} e^{i\omega(\tau^*-\tau)}
 \end{equation}
@@
   \label{eqn:u0s}
-\uzl[\kappa,\dla](t) = \frac{a^*}{a}g_{\lambda,\frac{r\lambda}{a^2}}(t-(\tau^*-\tau))
+\uzl(t) = \frac{a^*}{a}g_{\lambda,\frac{r\lambda}{a^2}}(t-(\tau^*-\tau))
 \end{equation}
@@
 
-Since $\fwl$ along with all of its derivatives
-tends to zero faster than any polynomial at $\infty$, $\fwl' \in
-L^2(\bR)$.
-Therefore $\hat{g}_{\lambda,\frac{r\lambda}{a^2}}$
-and $\hat{g}_{\lambda,\frac{r\lambda}{a^2}}'$ are also square-integrable, 
-from equation \ref{eqn:hatuf} $(\hat{u}^0_{\lambda})'$ is also square
-integrable. Consequently $t \uzl$ is square integrable. Therefore
-$J_{\rm MSWI,r\lambda}$, the result of substituting in the
-right-hand side of the definition \ref{eqn:mswidef} $\uzl$ for
-$\ul$, is well-defined. \cite{HuangSymes2015SEG} presented an analysis of
-$J_{\rm MSWI,r\lambda}$, as follows:
-\[
-J_{\rm MSWI,r\lambda}[\kappa;\dla] = \sum_{\bx_s,\bx_r}\int dt
-|t\uzl[\kappa;\dla]|^2 =  \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{r\lambda}{a^2}}(t-(\tau^*-\tau))|^2
-\]
-\[
-=\sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, (t+(\tau^*-\tau))^2|g_{\lambda,\frac{r\lambda}{a^2}}(t)|^2
+Since $\wl \in C_0^{\infty}(\bR)$, $\hat{g}_{\lambda,\frac{r\lambda}{a^2}}$
+and $\hat{g}_{\lambda,\frac{r\lambda}{a^2}}'$ are square-integrable, whence
+from equation \ref{eqn:hatuf} $(\hat{u}^0_{r\lambda})'$ is also square
+integrable. Consequently $t \uzl$ is square integrable.
+
+The following argument is a variant of one presented
+by\cite{HuangSymes2015SEG}, using ideas from Song's thesis \cite[]{Song:94}. Observe that
+\[
+\|T\uzl[\kappa;\dla]\|^2 =  \frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{r\lambda}{a^2}}(t-(\tau^*-\tau))|^2
+\]
+\[
+=\frac{(a^*)^2}{a^2} \int dt\, (t+(\tau^*-\tau))^2|g_{\lambda,\frac{r\lambda}{a^2}}(t)|^2
 \]
@@
   \label{eqn:MSWI}
-  =\sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, (t^2 + 2t
+  =\frac{(a^*)^2}{a^2} \int dt\, (t^2 + 2t
   (\tau^*-\tau)+(\tau^*-\tau)^2)|g_{\lambda,\frac{r\lambda}{a^2}}(t)|^2
 \end{equation}
-
-Since the Fourier transform $\fwl$ is even and real, so is
+Since $\wl$ is real, $|\fwl|$ is even, therefore
+so is
 $\hat{g}_{\lambda,\frac{r\lambda}{a^2}}$, therefore $g_{\lambda,\frac{r\lambda}{a^2}}$ is even, and the linear term (in t) on the
@@
   \label{eqn:MSWIasym}
-  J^0_{\rm MSWI,r\lambda}[\kappa;\dla] =
-\sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{r\lambda}{a^2}}(t)|^2
+  \|T\uzl\|^2 =
+\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\lambda,\frac{r\lambda}{a^2}}(t)|^2
 +(\tau^*-\tau)^2\frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{r\lambda}{a^2}}\|^2
 \end{equation}
-
 Recalling the definition \ref{eqn:pw} of pulse width, \ref{eqn:MSWIasym} can be re-written as
@@
   \label{eqn:MSWIasym2}
-  J^0_{\rm MSWI,\lambda}[\kappa;\dla] = 
-  \sum_{\bx_s,\bx_r}\frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{r\lambda}{a^2}}\|^2\left(l(g_{\lambda,\frac{r\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
+ \|T\uzl\|^2 =
+  \frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{r\lambda}{a^2}}\|^2\left(l(g_{\lambda,\frac{r\lambda}{a^2}})^2  + (\tau^*-\tau)^2\right)
 \end{equation}
@@
   \label{eqn:u0snorm}
-  \|\uzl[\kappa;\dla]\|^2=
+  \|\uzl\|^2=
   \frac{(a^*)^2}{a^2}\|g_{\lambda,\frac{r\lambda}{a^2}}\|^2.
@@
 
-So the leading-term modification of $J_{\rm AWI}$ differs from mean square
+So the leading-term modification of each summand in the definition of $J_{\rm AWI}$ differs from mean square
 travel time error by the pulse width of
 $g_{\lambda,\frac{r\lambda}{a^2}}$. The analogous relation for $_{\rm
-  MSWI}$ involves the norm-squared as well. To assess these
+  MSWI}$ involves the norm-squared as well.
+
+To assess these
 quantities, note first that
@@
 From the bounds \ref{eqn:unorm} and \ref{eqn:lambdanorms},
-\[
+\begin{equation}
+  \label{eqn:ulnorm}
   \|\ul\| \le C \lambda^{-\frac{1}{2}}\|\dl\|
-\]
+\end{equation}
 Combining the preceding results in thet estimate in display
@@
 \end{cor}
+
+\noindent {\bf Remark:} From inequalities \ref{eqn:ulnorm},
+\ref{eqn:Srembd}, $\tilde{J}_{r\lambda}[\kappa;\dl] \le C$. That
+suggests that the contribution of the second term on the
+left-hand-side of the limit \ref{eqn:cor2} is insignificant for small
+$\lambda$, whereas it is comparable to the first term in
+\ref{eqn:cor3}.
+
+%%%%%%%%
 
@@
 \section{Conclusion}
+
+The conclusions of this study depend on a number of limitations on the
+acoustic model of wave propagation and the formulation of the related
+inverse problem. The principal constraints are:
+\begin{itemize}
+\item energy sources are modeled as isotropic point radiators, with
+  known time dependence vanishing outside a finite time window;
+\item mechanical parameter fields (coefficients in the acoustic wave
+  equation) are smooth, that is, infinitely differentiable as
+  functions of position, and the material density is known and
+  position-independent: only the bulk modulus (or wave velocity) is
+  to be determined;
+\item waves propagate throughout Euclidean space-time - that is, no
+  physical boundaries are modeled;
+\item the data of the inverse problem are the restrictions of the
+  pressure field to a finite number of points in space (receiver
+  positions), for a finite number of source positions and all time;
+\item local energy decays exponentially - that is, the pressure and
+  all of its derivatives tend to zero at an exponential rate as $t
+  \rightarrow \infty$,
+  uniformly in any bounded spatial domain.
+\item only one ray of geometric
+  acoustics connects each source-receiver location pair (single arrivals)
+\end{itemize}
+
+These constraints are discussed in detail in the next section. Some
+could certainly be relaxed. For instance, exponential decay of local
+energy implies that the finite time duration of actual measured data
+can be treated as a perturbation of the infinite time data setting
+posed here. The exception is the single-arrival hypothesis: it is
+essential, as without it, the link between the travel time and
+extended waveform inversion is lost. This is a feature of the
+source-receiver extension, not of a particular objective function
+(MSWI, AWI,...) formulated in terms of it. This extension treats each receiver signal
+independently. If only a single identifiable waveform arrives at
+each receiver, AWI (and similar approaches) can implicitly identify
+its arrival time, and the link to travel time inversion is
+established. If distinct waveforms are present at multiple times,
+the single receiver signal does not determine which waveform in the
+predicted signal should align with which waveform in the data to be
+fitted. This ``crosstalk'' ambiguity is analyzed in the fifth
+section, and afflicts all approaches based on the source-receivre
+extension.
+
+In any case, these hypotheses enable a simple and clear
+analysis of AWI, and seem a reasonable starting point for further
+investigation.
+
 In this paper, I have shown that AWI is a good approximation to least

### 2024-11-05 repro/papers/awi/paper.tex
interim

@@
 predicted wave data and actual observed
-data. If wave velocities or their parametric equivalents are amongst the parameters
-to be estimated, such numerical measures tend to oscillate, due to:
-the excessive sensitivity of predicted waveforms to changes in wave velocity.
-%especially for the most commonly used square-integral measure ($L^2$
-%norm).
-These oscillations cause iterative local optimization algorithms (relatives of
+data. This data is typicallyl oscillatory. If wave velocities or their parametric equivalents are amongst the parameters
+to be estimated, then updates cause shifts in the arrival times of
+data oscillations, which in turn causes commonly used numerical error measures
+to oscillate. This ``cycle skipping'' phenomenon may cause iterative local optimization algorithms (relatives of
 Newton;'s method, the only practical algorithmic choices) to stagnate
-far from good data fit \cite[]{VirieuxOperto:09}, unless the initial
-parameter estimates are of very high quality.
+far from good data fit, unless the initial
+parameter estimates accurately predict the arrival times of waves \cite[]{GauTarVir:86,VirieuxOperto:09,HuetalIEEE:18}.
 
 Adaptive Waveform
-Inversion \cite[]{Warner:16} (AWI) is a modification of FWI, designed to overcome this
-``cycle-skipping'' pathology. AWI has been successfully applied to
+Inversion \cite[]{Warner:16} (AWI) is a modification of FWI, designed to overcome the
+cycle-skipping pathology. AWI has been successfully applied to
 industry-scale seismic data
@@
 
-This paper supplies such an explanation for a special case of AWI, the recovery of smooth acoustic
-medium parameters from data generated at localized energy sources,
-propagated through a fluid, and recorded at receiver locations remote
-from the sources. I show that the AWI objective function is an
-asymptotic (in time frequency) approximation to the mean-square
-time-of-travel misfit, provided that arrival times of transmitted
-waves are single, that is, unique rays of geometric asymptotics
-connect sources and receivers. The error in this approximation is
-proportional to the source mean-square wavelength. Thus AWI and
-minimization of mean-square travel time error (travel time tomography)
-lead to similar material parameter estimates. Travel time tomography
-appears to be much less sensitive to choice of initial model than is
-FWI, and in fact is commonly used to construct an initial model for
-FWI \cite[]{Bordingetal:87,SirguePratt:04,VirieuxOperto:09}. At least
-in the special case treated here, AWI implicitly combines travel time
-tomography with FWI.
-
-The reader will
-note that a normalization component of the AWI objective function
-definition is important in reaching this conclusion. The connection
-between objective normalization and the asymptotic equivalence to
-travel time inversion for transmission data has nothing to do with the
-reasons for normalizing the objective previoiusly advanced in the
-literature \cite[]{Warner:16,LiAlkhalifah:21}. Another similar modification of
-FWI, Matched Source Waveform Inversion (MSWI)
-\cite[]{HuangSymes2015SEG,HuangSymes:Geo17} lacks the AWI
-normalization, and has a looser relation to travel time inversion. The
-arguments presented here for AWI (and MSWI) are refinements of those
-given by \cite{HuangSymes:Geo17}. which were in turn based on ideas
-presented in Song's thesis \cite[]{Song:94c}.
+This paper supplies such an explanation for a special case of AWI, the
+recovery of smooth acoustic medium parameters from data generated at
+localized energy sources, propagated through a fluid, and recorded at
+receiver locations remote from the sources (transmission data). I show that the AWI
+objective function is an asymptotic (in time frequency) approximation
+to the mean-square time-of-travel misfit, provided that arrival times
+of transmitted waves are single. That is, a unique ray of geometric
+asymptotics connects each sources-receiver pair, as is the case if the associated
+Riemannian metric is simple in a domain containing the sources and
+receivers \cite[]{PestovUhlmann:05,StefanovUhlmann:05}. The error in
+this approximation is proportional to the source mean-square
+wavelength. I also show that this relation fails if arrival times have
+higher multiplicity. Travel time tomography (minimization of
+mean-square travel time error) appears to be much less sensitive to
+choice of initial model than is FWI. In fact, it is commonly used to
+construct an initial model for FWI
+\cite[]{Bordingetal:87,SirguePratt:04,VirieuxOperto:09}. At least for
+acoustic transmission data with single arrivals, AWI implicitly
+combines travel time tomography with FWI.
+
+The arguments presented here for AWI are refinements of those given by
+\cite{HuangSymes:Geo17} for Matched Source Waveform Inversion (MSWI),
+a similar modification of FWI
+\cite[]{HuangSymes2015SEG,HuangSymes:Geo17}. which were in turn based
+on ideas presented in Song's thesis \cite[]{Song:94c}. Since it
+involves no additional effort, I state results for MSWI in parallel
+with those for AWI.
 
@@
 problems, regularization has been accomplished via additive penalty
-functions. In section four, I explain the relation between AWI and a
-penalty formulation of the wave inverse problem. The close relation
+functions. In section four, I explain the relation between AWI, MSWI, and 
+penalty formulations of the wave inverse problem. The close relation
 between AWI objective and travel time misfit is limited in scope: in
@@
 transmitted wave data in which multiple rays connect source and
-receiver.
+receiver (multiiple wave arrivals, non-simple metric).
 
@@
 %1.3.3)
-imply that $p(\cdot,\cdot;\bx_s)$ is smooth away from $\{\bx_s\}
+imples that $p(\cdot,\cdot;\bx_s)$ is smooth away from $\{\bx_s\}
 \times \bR$, so that the traces $p(\bx_r, \cdot;\bx_s)$ on
@@
   \label{eqn:modop}
-  F: exp(M)  \rightarrow \Pi_{X_{sr}} {\cal D}'(\bR)
+  F: exp(M)  \rightarrow \Pi_{X_{sr}} C^{\infty}(\bR)
 \end{equation}
@@
 
-The analysis to follow requires two further constraints on the domain
+The analysis to follow requires three further constraints on the set
+of source-receiver location pairs $X_{sr}$ and the domain
 $M \subset C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)$ of $F$. The first is
@@
 \begin{quote}
-  For any $n > 0$, $f \in {\cal E}'(\bR)$,
+  For any $n > 0$, $f \in C_0^{\infty}(\bR)$,
   there exist $K >
@@
 
-The second assumption explicitly involves a constraint on ray geometry
+The second and third assumptions explicitly involves a constraint on ray geometry
 (therefore on $\kappa$). A {\em normal neighborhood} of
@@
 \begin{quote}
-  $\Omega\subset \bR^3$ 
-and $M \subset C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)$ are bounded open sets satisfying 
-\begin{itemize}
-  \item[G1. ] $X_{sr} \subset \Omega \times \Omega$;
-  \item[G2. ] for every source point $\bx_s\in X_{s}$ and $\log \kappa \in M$, $\Omega$ is a
-    normal neighborhood of $\bx_s$.
-  \item[G3. ] for every source point $\bx_s \in X_{s};$
-    $\log \kappa \in M$, ray ${\bf X}:\bR^+\rightarrow \bR^3$
+  $\Omega\subset \bR^3$ and $M \subset C^{\infty}(\bR^3) \cap
+  L^{\infty}(\bR^3)$ are bounded open sets for which $X_{sr} \subset
+  \Omega \times \Omega$, and for every source point $\bx_s\in X_{s}$
+  and $\log \kappa \in M$, $\Omega$ is a normal neighborhood of
+  $\bx_s$.
+\end{quote}
+
+The {\em No Return} assumption is
+\begin{quote}
+  for every source point $\bx_s \in X_{s}$, $\log \kappa \in M$, ray ${\bf X}:\bR^+\rightarrow \bR^3$
   with ${\bf X}(0)=\bx_s$ and any $t>0$, either $\{{\bf X}(s): 0\le s \le t\} \subset \Omega$
   or $\{{\bf X}(s): t < s < \infty\} \subset \bR^3 \setminus \Omega$.
-  \end{itemize}
 \end{quote}
-Assertion G3 is the {\em no-return} assumption: once a ray leaves
-$\Omega$, its subsequent intersection with $\Omega$ is null.
-
-Thanks to exponential decay, each trace $p(\bx_r,\cdot;\bx_s)$ is in $L^{p}(\bR)$ for $p
+This assumption states that once a ray leaves
+$\Omega$, its subsequent intersection with $\Omega$ is null (hence the name).
+
+Thanks to Exponential Decay, each trace $p(\bx_r,\cdot;\bx_s)$ is in $L^{p}(\bR)$ for $p
 \ge 1$.
@@
 Note that while the Lax-Milgram Theorem (\cite{Yosida}, section
-III.7) ensures that $u_{\sigma}[\kappa;d] \in L^2(X_{sr}\times \bR)$,
+III.7) ensures the existence of a unique solution
+$u_{\sigma}[\kappa;d] \in L^2(X_{sr}\times \bR)$ of the least squares
+problem \ref{eqn:reg},
 it is not immediiately obvious that $Tu_{\sigma}[\kappa;d] \in L^2(X_{sr}\times \bR)$.
@@
 Asymptotic analysis of AWI and MSWI is based on a family of source
-wavelet: for $\lambda > 0$,
+wavelets: for $\lambda > 0$,
 \begin{equation}
@@
   \label{thm:main}
-  Assume that the Exponential Decay and Single Arrival assumptions
+  Assume that Exponential Decay, Single Arrival, and No Return
   hold for $X_{sr} \subset \Omega \times \Omega, M \subset C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)$.
@@
   W \le C^*$ so that for
-  $\log \kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$, $Tu_{\sigma}[\kappa;\dl](\bx_r,\cdot;\bx_s) \in L^2(\bR)$, and 
+  $\log \kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$,
+  \begin{itemize}
+  \item[1. ] $Tu_{\sigma}[\kappa;\dl](\bx_r,\cdot;\bx_s) \in
+    L^2(\bR)$;
+  \item[2. ] for $\lambda > 0$,
     \begin{equation}
@@
       |\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|- \lambda^{-\frac{1}{2}} (W[\kappa^*,\kappa](\bx_s,\bx_r))^{\frac{1}{2}}|\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|| \le C
-  \lambda^{\frac{1}{2}}
-\end{equation}
-and for sufficiently small $\lambda$, 
+      \lambda^{\frac{1}{2}};
+    \end{equation}
+  \item[3. ]for sufficiently small $\lambda$, 
     \begin{equation}
@@
     \end{equation}
+  \end{itemize}
 \end{theorem}
@@
 \lim_{\lambda \rightarrow 0} \lambda J_{\rm MSWI,r\lambda}[\kappa;\dl]
-= \sum_{(\bx_s,\bx_r) \in X_{sr}} W[\kappa^*,\kappa^*]^2(\bx_s,\bx_r) |\tau[m](\bx_s,\bx_r) - \tau[m^*](\bx_s,\bx_r)|^2.
+= \sum_{(\bx_s,\bx_r) \in X_{sr}} W[\kappa^*,\kappa]^2(\bx_s,\bx_r) |\tau[\kappa](\bx_s,\bx_r) - \tau[\kappa^*](\bx_s,\bx_r)|^2.
 \end{equation}
@@
 \lim_{\lambda \rightarrow 0} J_{\rm AWI,r\lambda}[\kappa;\dl]
-= \sum_{(\bx_s,\bx_r) \in X_{sr}}  |\tau[m](\bx_s,\bx_r) - \tau[m^*](\bx_s,\bx_r)|^2.
+= \sum_{(\bx_s,\bx_r) \in X_{sr}}  |\tau[\kappa](\bx_s,\bx_r) - \tau[\kappa^*](\bx_s,\bx_r)|^2.
 \end{equation}
 \end{cor}
+
+\section{Analysis of AWI and MSWI}
+THis section is devoted to the proofs of the main result, Theorem
+\ref{thm:main}.  The principal tool in the analysis is Hadamard's
+decomposition of the acoustic Green's function, described in the
+second
+subsection below, along with some of its consequences, the chief being the
+square-integrability of the time-scaled regularized filter
+$tu_{\sigma}$. The next subsection reveals the link to travel time via
+an approximate solution of the regularized least squares problem
+\ref{eqn:reg}, based on the leading term in Hadamard[s decomposition.
+The final subsection estimates the remainder and finishes the proof of
+the main result.
+
+\subsection{A Note on Notations  for Norms and Bounds}
+The Banach spaces
+$l^p(X_{sr},L^p(\bR))$ will be abbreviated as $L^p(X_{sr} \times
+\bR)$. The norm in a Banach space $B$ will be denoted
+$\|\cdot\|_{B}$. The norm symbol
+$\|\cdot\|$ without subscript will be understood to mean either
+$\|\cdot\|_{L^2(\bR)}$ or $\|\cdot\|_{L^2(X_{sr}\times
+  \bR)}$, as context indicates.
+
+In a statement pertaining to all
+$\kappa \in M$, the letter $C$, or sometimes $C_k,
+k=0,1,2,...$ or
+$C_*,C^*$, will designate a constant (or constants) independent of
+$\kappa$.
+
+Generally, constants denoted $C$ etc. will absorb any dimensional
+scale factors necessary to make sense of the expressions in which they
+appear. I will call out those few cases in which explicit use of
+dimensinoal scale factors is necessary to avoid irretrievable
+dimensional confusion.
 
@@
 The Hadamard decomposition of the acoustic Green's function \cite[]{Friedlander:75,Qian:JCP24} provides a
-basis for a norm-based formulation of FWI. As presented by
+basis for analysis of AWI and MSWI. As presented by
 \cite{Friedlander:75}, for instance, this construction applies to
@@
 equation. Conversely, if $p$ is a solution of the second-order initial
-value problem\ref{eqn:awe2}, then under suitable conditions (including
-those to be imposed below),
+value problem\ref{eqn:awe2}, then under the conditions imposed in the
+previous section, those to be imposed below),
 \[
@@
 \]
-defines a distribution, and $(p,\bv)$ solves the system \ref{eqn:awe}.
+defines a distribution, and $(p,\bv)$ is a weak solutino of the system \ref{eqn:awe}.
 
 Hadamard's construction is local, in both space-time and in the set of
-acoustic models. A {\em normal neighborhood} of
-$\bx_s \in \bR^3$ is an open subset of $\bR^3$ 
-within which every point is connected to $\bx_s$ by  a unique ray of
-geometric acoustics (a geodesic for the Riemannian metric associated
-to the equation \ref{eqn:awe2}) (\cite{Friedlander:75},
-p. 16).
-
-Throughtout the following discussion, $\Omega\subset \bR^3$ 
-and $M \subset C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)$ are bounded open sets satisfying 
-\begin{itemize}
-  \item[G1. ] $X_{sr} \subset \Omega \times \Omega$;
-  \item[G2. ] for every source point $\bx_s\in X_{s}$ and $\log \kappa \in M$, $\Omega$ is a
-    normal neighborhood of $\bx_s$.
-  \item[G3. ] for every source point $\bx_s \in X_{s};$
-    $\log \kappa \in M$, ray ${\bf X}:\bR^+\rightarrow \bR^3$
-    with ${\bf X}(0)=\bx_s$ and any $t>0$, either $\{{\bf X}(s): 0\le s \le t\} \subset \Omega$
-    or $\{{\bf X}(s): t < s < \infty\} \subset \bR^3 \setminus \Omega$.
-  \end{itemize}
-Condition G3 (the ``no return'' condition) means that once a ray
-leaves $\Omega$, it does not return to $\Omega$. Conditions G1 - G3
-hold for any convex open set $\Omega$ if $\kappa$ is constant, and for small perturbations and some
-large perturbations of constant $\kappa$: these are open conditions in
-the topology of $M$.
-
-Under conditions G1 - G4, the construction described by
-\cite{Friedlander:75} produces for each $m \in M$ functions $\tau[\kappa]$ (travel time), $a[\kappa]$
+acoustic models. In the present context, it requires the Single
+Arrival hypothesis, and for global time validity the No Return
+assumption.
+
+\cite{Friedlander:75} produces for each $\log \kappa \in M$ functions $\tau[\kappa]$ (travel time), $a[\kappa]$
 (geometric amplitude) on $X_s \times \Omega$ and $b[\kappa]$
-(remainder kernel) on $X_s \times \Omega \times \bR$, satisfying
-\begin{itemize}
-\item[H1. ] $\tau[\kappa], a[\kappa],$ and $b[\kappa]$ are smooth off the diagonal,
-  that is,
-  (i) for each $\bx_s \in X_s$, $\tau[\kappa](\bx_s,\cdot) \in C^{\infty}(\Omega \setminus
-  \{\bx_s\})$, and similarly for $a[\kappa],b[\kappa]$; $\tau$
+(remainder kernel) on $X_s \times \Omega \times \bR$, smooth off the
+diagonal - that is, for each $\bx_s \in X_s$, $\tau[\kappa](\bx_s,\cdot) \in C^{\infty}(\Omega \setminus
+  \{\bx_s\})$, and similarly for $a[\kappa],b[\kappa]$; The traveltime $\tau$
   satisfies the eikonal equation
@@
   \]
-\item[H2. ] there exists $A>0$ so that for any $\log \kappa \in M$
-  and $(\bx_s,\bx_r) \in X_{sr}$,
-  $|\log a[\kappa](\bx_s,\bx_r)| < A$;
-\item[H3. ] for any $k \ge 0$, there exists $B \ge 0,\delta > 0$ so
-  that for any $(\bx_s,\bx_r) \in X_{sr}$, $\kappa \in M$, 
+A by-product of the construction is a bound $A>0$ so that for any $\log \kappa \in M$
+  and $(\bx_s,\bx_r) \in X_{sr}$,  $|\log a[\kappa](\bx_s,\bx_r)| <
+  A$;  With no further assumptionss, there is also a uniform bound on $b(\bx_r,t;\bx_s)$ and
+  its derivatives over any finite $t$-interval. However, with the
+  addition of the Exponetial Decay assumption, the bound becomes
+  global in time: for any $k \ge 0$, there exists $B \ge 0,\delta > 0$ so
+  that for any $(\bx_s,\bx_r) \in X_{sr}$, $\log \kappa \in M$, 
 \begin{equation}
@@
 \end{equation}
-\end{itemize}
-(The exponential decay estimate in item H3 follows from hypothesis G4.)
 
@@
 
-%Under conditions G1 - G3 above, this decomposition is the restriction of
-%the Greeen's function to $\Omega \times (-\infty,t_h]$. This fact
-%implies a decomposition of $F[\kappa]$, provided that the convolution $G*f$
-%restricted to $[0,t_d]$ is determined by the restriction of $G$ to
-%$\Omega \times [0,t_h]$. That is the case if $t_h > t_d + |\mbox{supp
-%|f/2$ (the factor of $1/2$ because $f$ is odd).
-%Making this assumption (which is weaker than assumptions on $t_h$ to be made below),
-%\[
-%   F[\kappa](\bx_s,\bx_r)(t) =
-%   \phi(t)a(\bx_s,\bx_r)\rho_s\frac{df}{dt}(t-\tau[\kappa](\bx_s,\bx_r))
-% \]
-%\begin{equation}
-%  \label{eqn:fwdgreen}
-% + \phi(t)\rho_s\int
-%  \,ds \, \frac{df}{dt}(t-s) b[\kappa](\bx_s,\bx_r,s-\tau[\kappa](\bx_s,\bx_r))H(s-\tau[\kappa](\bx_s,\bx_%r))
-%\end{equation}
-
-\subsection{FWI via Least Squares}
-For $f \in {\cal E}'(\bR)$, the Hadamard decomposition 
-(equation \ref{eqn:regreen}) gives:
-\begin{equation}
-  \label{eqn:fwdgreen}
-  F[\kappa] = G[\kappa] * \rho \partial_t f  = \rho( a[\kappa]\delta
-  +  (b[\kappa]H)) *\delta_{\tau[\kappa]}*\partial_t f.
-\end{equation}
-
-For the remainder of this paper, source wavelet $f$ will be assumed
-smooth, of compact support: $f \in C^{\infty}_0(\bR)$. Less stringent constraints on $f$ would
-support the results below, but this is the simplest choice. Also, with
-a few notable exceptions \cite[]{SondRes:83}, source wavelets that
-adequately model physical experimentss may be take to be smooth of
-compact support, because the physical phenomena that inject energy
-into the acoustic field are transient, and the data is sampled at a
-finite rate and either analog
-or digitally filtered (or both) to well below Nyquist frequency.
-
-Then equation \ref{eqn:fwdgreen}, bounds
-\ref{eqn:best} and Young's inequality implies that $F[\kappa] \in
-L^p(X_{sr} \times \bR)$ for any $p \ge 1$. Moreover there exists
-$C>0$ so that for $\log \kappa\in M$, $p \ge 1$,
-\begin{equation}
-  \label{eqn:fwdnorm}
-  \|F[\kappa]\|_{L^p(X_{sr} \times \bR)} \le C \left\|\partial_t f\right\|_{L^p(\bR)}.
-\end{equation}
-This observation justfies posing FWI as a least-squares problem: given $d \in
-L^2(X_{sr}\times \bR)$, $f \in H^1_0(\bR)$, find $\kappa$ to minimize
-\begin{equation}
-  \label{eqn:fwi}
-  J_{\rm FWI}[m;d]= \frac{1}{2}\|F[\kappa]-d\|_{L^2(X_{sr} \times \bR)},
-\end{equation}
-
-Practical versions of least-squares FWI include various types of regularization and
-imposition of constraints on $m$, see the references cited in the
-Introduction.
-%As mentioned there, local iterative optimization methods
-%are the only practical approach to minimizing $J_{\rm FWI}$ due to the
-%computational cost of approximating the forward map $F$ and related
-%operators, but these tend to stagnate at highly sub-optimal
-%(``cycle-skipped'') estimates of $m$ unless the initial estimate is of
-%quite high quality, often available only at considerable cost in data
-%collection and processing.
-
-\subsection{AWI via Optimization}
-
-AWI is based an {\em extension} of the modeling operator $F$, that is,
-an operator ${\bar F}$ with larger domain, coinciding with $F$ on its domain..
-The extended modeling operator ${\bar F}$ maps extended models
-$\bar{m}=(\kappa,\bar{f})$ incorporating extended sources
-$\bar{f}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
-field. Denote by $\bar{p}(\bx,t;\bx_s,\bx_r), \bar{\bv}(\bx,t;\bx_s,\bx_r)$
-the solution of the acoustic system \ref{eqn:awe} with $f(t)$ in the
-right-hand side of the first equation replaced by
-$\bar{f}(\bx_r,\cdot;\bx_s)$. Then
-$\bar{F}[\kappa,\bar{f}](\bx_r,t;\bx_s) =
-\bar{p}(\bx_r,t;\bx_s,\bx_r)$. As was the case for $F$, the Hadamard
-decomposition \ref{eqn:green} and conditions G1-G4, W1-W3 imply that
-$\bar{F}[\kappa, \cdot]$ is a bounded map: $L^2(X_{sr} \times \bR)
-\rightarrow L^2(X_{sr} \times \bR)$. If all  
-of the wavelets are the same, that is, ${\bar f}({\bf  
-  x}_r,t;{\bf x}_s) = f(t)$ is independent of source and receiver  
-position, then ${\bar F}[\kappa,\bar{f}] = F[\kappa]$. That is, $F$ is a special
-case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
-$F$. This is the {\em source-receiver} extension, in the terminology of \cite{HuangSymes2015SEG}.
-
-\cite{Warner:16} represent the extended sources $\bar{f}$ as time convolutions of the
-(known) source $f$ with an {\em adaptive filter} $u \in L^2(X_{sr}
-\times \bR)$, that is, $\bar{f}=u * f$.
 Here, $u*f$ denotes the result of convolving each trace in $u$ by $f$. I
@@
 \end{equation}
-
-%Since $S[\kappa]$ involves solution of the wave equation with point source
-%wavelet $u*w$, application of the Hadamard decomposition \ref{eqn:green} requires a furnther
-%constraint on ray geometry, expressed by condition G3. Specifically,
-%\begin{itemize}
-%\item[G4. ]$t_h > t_d + t_w + t_u$.
-%\end{itemize}
-%Then \ref{eqn:green} implies that
-%\begin{equation}
-%  \label{eqn:sop}
-%  S[\kappa]u = w* a[\kappa] T[\kappa] ((\delta + b[\kappa]H)*u)
-%\end{equation}
-%in which $T[\kappa]$ is the time translation operator by $\tau[\kappa]$:
-%\begin{equation}
-%  \label{eqn:Tdef}
-%  T[\kappa]u(\bx_r,t;\bx_s) = u(\bx_r,t-\tau[\kappa](\bx_s,\bx_r);\bx_s),
-%\end{equation}
-%Similarly,
-%\begin{equation}
- % \label{eqn:fwdgreenop}
-%  F[\kappa] = \phi a[\kappa] T[\kappa]((\delta + b[\kappa]H)*w) 
-%\end{equation}
-
-%Note that values of $b$ for $t > t_d + t_w + t_u$ do not affect the
-%definitions of $S[\kappa]u$ or $F[\kappa]$. Therefore alter $b$ by
-%multiplication by a suitable cutoff function so that its support lies in
-%$(-\infty, t_h]$.
 
@@
 %$\kappa^* \in M$.
-
-\subsection{A Note on Notations  for Norms and Bounds}
-The Banach spaces
-$l^p(X_{sr},L^p(\bR))$ will be abbreviated as $L^p(X_{sr} \times
-\bR)$. The norm in a Banach space $B$ will be denoted
-$\|\cdot\|_{B}$. The norm symbol
-$\|\cdot\|$ without subscript will be understood to mean either
-$\|\cdot\|_{L^2(\bR)}$ or $\|\cdot\|_{L^2(X_{sr}\times
-  \bR)}$, as context indicates.
-
-In a statement pertaining to all
-$\kappa \in M$, the letter $C$, or sometimes $C_k,
-k=0,1,2,...$ or
-$C_*,C^*$, will designate a constant (or constants) independent of
-$\kappa$.
-
-Generally, constants denoted $C$ etc. will absorb any dimensional
-scale factors necessary to make sense of the expressions in which they
-appear. I will call out those few cases in which explicit use of
-dimensinoal scale factors is necessary to avoid irretrievable
-dimensional confusion.
 
@@
 
+The reader will
+note that a normalization component of the AWI objective function
+definition is important in reaching this conclusion. The connection
+between objective normalization and the asymptotic equivalence to
+travel time inversion for transmission data has nothing to do with the
+reasons for normalizing the objective previoiusly advanced in the
+literature \cite[]{Warner:16,LiAlkhalifah:21}.
+
 The conclusions of this study depend on a number of limitations on the
@@
 
+%%%%%%%%%%%%%%%%%%%%
+
+\subsection{AWI via Optimization}
+
+AWI is based an {\em extension} of the modeling operator $F$, that is,
+an operator ${\bar F}$ with larger domain, coinciding with $F$ on its domain..
+The extended modeling operator ${\bar F}$ maps extended models
+$\bar{m}=(\kappa,\bar{f})$ incorporating extended sources
+$\bar{f}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
+field. Denote by $\bar{p}(\bx,t;\bx_s,\bx_r), \bar{\bv}(\bx,t;\bx_s,\bx_r)$
+the solution of the acoustic system \ref{eqn:awe} with $f(t)$ in the
+right-hand side of the first equation replaced by
+$\bar{f}(\bx_r,\cdot;\bx_s)$. Then
+$\bar{F}[\kappa,\bar{f}](\bx_r,t;\bx_s) =
+\bar{p}(\bx_r,t;\bx_s,\bx_r)$. As was the case for $F$, the Hadamard
+decomposition \ref{eqn:green} and conditions G1-G4, W1-W3 imply that
+$\bar{F}[\kappa, \cdot]$ is a bounded map: $L^2(X_{sr} \times \bR)
+\rightarrow L^2(X_{sr} \times \bR)$. If all  
+of the wavelets are the same, that is, ${\bar f}({\bf  
+  x}_r,t;{\bf x}_s) = f(t)$ is independent of source and receiver  
+position, then ${\bar F}[\kappa,\bar{f}] = F[\kappa]$. That is, $F$ is a special
+case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
+$F$. This is the {\em source-receiver} extension, in the terminology of \cite{HuangSymes2015SEG}.
+
+\cite{Warner:16} represent the extended sources $\bar{f}$ as time convolutions of the
+(known) source $f$ with an {\em adaptive filter} $u \in L^2(X_{sr}
+\times \bR)$, that is, $\bar{f}=u * f$.
+
 \bibliographystyle{seg}

### 2024-11-07 repro/papers/awi/paper.tex
near final

@@
 
-Here, $u*f$ denotes the result of convolving each trace in $u$ by $f$. I
-will also use the asterisk to denote the trace-by-trace convolution
-of two data sets with the same set of source and receiver coordinates,
-that is, the data set consisting of time convolutions of traces with the same source
-and receiver coordinates. The meaning will be clear from context.
-
-For such extended sources,
-\[
-  \bar{F}[\kappa,u*f] = u*F[\kappa]
-\]
+\subsection{Tihonov Regularization}
 Introduce the notation $S[\kappa]$ for the operator of convolution
@@
 \end{equation}
-
-\cite{Warner:16} assume (implicitly) that there is always a unique
-solution of the extended inversion problem, formulated in terms of
-$S$: that is, that
-for any $\kappa$ and any data $d$, there is an
-adaptive filter $u[\kappa;d]$ for which
-$S[\kappa]u[\kappa;d]= d$. If $d$ is noise-free, that is, $d=F[\kappa]$, then the consistency relation \ref{eqn:sconsist}
-suggests that $u[\kappa;d](\bx_r,t;\bx_s) = \delta(t)$. This ideal adaptive filter is in the null
-space of multiplication by $t$, that is, the operator
-$u(\bx_r,t;\bx_s) \mapsto t u(\bx_r,t;\bx_s)$. This
-observation motivates the definition of the AWI objective function
-(\cite{Warner:16}, equation 4)::
-\begin{equation}
-  \label{eqn:awideforig}
-  J_{\rm AWI}[\kappa;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu[\kappa;d](\bx_r,t;\bx_s)|^2}{\int dt |u[\kappa;d](\bx_r,t;\bx_s)|^2}\right) 
-\end{equation}
-in which $u[\kappa;d]$ is as defined above, that is, the solution of $S[\kappa]u[\kappa;d]
-= d$. \cite{Warner:16} also formulate
-AWI in terms of the inverse filter, that is, $u$ for which
-$u*d = F[\kappa]$, and in fact recommend that definition. The present
-discussion uses only the forward filter.
-Adaptive waveform inversion consists in iterative minimization of
-$J_{\rm AWI}$: choose an initial estimate $\kappa_0$ of the
-model, then update $\kappa$ repeatedly using a gradient
-descent method. \cite{Warner:16} describes the computation of $J_{\rm AWI}$
-and its gradient in detail, and their use in a quasi-Newton
-optimization algorithm.
-
-The same observation motivates the definition of the {\em Matched
-  Source Waveform Inversion} (MSWI) objective studied by
-\cite{HuangSymes2015SEG,HuangSymes:Geo17}:
-\begin{equation}
-  \label{eqn:mswideforig}
-  J_{\rm MSWI}[\kappa;d] = \sum_{\bx_s,\bx_r} \int dt |tu[\kappa;d](\bx_r,t;\bx_s)|^2
-\end{equation}
-That is, the two objectives differ only by the trace-by-trace
-normalization featured in the definition of AWI (equation
-\ref{eqn:awideforig}). Much of the analysis of $J_{\rm AWI}$ pertains to
-$J_{\rm MSWI}$, as will be explained below.
-The trace-by-trace normalization in the definition of $J_{\rm AWI}$
-will prove quite consequential.
-
-\section{Why it works}
-
-It may not be obvious at first glance why AWI (local optimization of
-$J_{\rm AWI}$) should be should be any more effective at estimating
-$\kappa$ from data $d$ than is FWI (local
-optimization of $J_{\rm FWI}$). Under the smoothness and ray geometry assumptions made above (equation
-\ref{eqn:coeff}, conditions G1 - G4), an asymptotic relation
-between AWI and travel time tomography emerges, justifying the notion
-that AWI is more likely to succeed than is FWI.
-%In addition all of the
-%previously mentioned assumptions, I will assume that the data $d$ is
-%{\em noise-free}, that is, $d = F[\kappa^*]$ for a target bulk modulus
-%$\kappa^* \in M$.
-
-\subsection{Tihonov Regularization}
-Before establishing the relation with travel time tomography, several
-defects in the definition \ref{eqn:awideforig} of $J_{\rm AWI}$ must be
-remedied. First, the objective function is not well-defined for the
-``ideal'' case, in which $d = F[\kappa]$ and $u[\kappa;d]=\delta$:
-this adaptive filter is not square-integrable, so the denominator does
-not make sense. Second, the equation $S[\kappa]u=d$ does not (in
-general) have a solution $u \in L^2(X_{sr} \times \bR)$. Since
-$S[\kappa]$ is a convolution operator, solving $S[\kappa]u=d$ for $u$
-amounts to deconovolving $d$ by $F[\kappa]$..  Under the standing
-assumption that $f \in C^{\infty}_0(\bR)$, $F[\kappa]$ does not have a continuous
-(stable) convolution inverse, and $S[\kappa]$ is not surjective.
-
-In fact, these two defects are really symptoms of the instability of
-deconvolution for kernels with the properties of $F[\kappa]$.  From \ref{eqn:fwdgreen} and the
+The regularized least-squares problem \ref{eqn:reg} can be re-written
+\begin{equation}
+  \label{eqn:regs}
+    u_{\sigma}[\kappa;d] = \mbox{argmin}_{u \in L^2(X_{sr} \times \bR)} (\|S[\kappa]u-d\|^2 + \sigma\|u\|^2)
+  \end{equation}
+\cite{Warner:16} present the definition of the AWI objective function
+\ref{eqn:awidefmain} initially as if $\sigma=0$, in fact presuming a
+zero-residual solution (that is, $S[\kappa]u=d$) but later note that
+``pre-whitening'' is necessary for a well-behaved filter construction.
+``Pre-whitening'' is a synonym for Tihonov regularization, that is,
+the least-squares minimization \ref{eqn:regs} with $\sigma>0$.
+
+Regularization has another benefit, in that it makes the objective
+well-defined. Indeed, if the data is noise-free, that is,
+$d=F[\kappa]=S[\kappa]\delta$, then $u_0[\kappa,d]=\delta$, but the
+denominator in the definition \ref{eqn:awidefmain} is not defined.
+Of course, the numerator vanishes, motivating the definition of AWI,
+but the objective function is nonsense.
+
+For future reference, examine the implications of definition
+\ref{eqn:regs} for the units of the regularization weight $\sigma$,
+which is necessarily dimensional. Since the data $d$ has dimensions of
+pressure, so must $S[m]u = F[m]*u$. Of course $F[m]$ is also sampled
+pressure, and convolution is a $t$ integration, so for $S[m]u$ to be a
+pressure, $u$ must have dimensions of 1/time. Therefore
+$\|S[m]u-d\|^2$ has dimensions of pressure$^2 \times $time, $\|u\|^2$
+has dimensions of 1/time, whence $\sigma$ must be dimensioned as
+pressure$^2 \times$ time$^2$.
+
+From \ref{eqn:green} and the
 definition \ref{eqn:sdef} of  $S[\kappa]$,
@@
 \end{equation}
-Since $\partial_t f$ is even, the formal adjoint $S[\kappa]^T$ is given by
+The formal adjoint $S[\kappa]^T$ is given by
 \begin{equation}
   \label{eqn:sadjgood}
-  S[\kappa]^Tu = \rho( a[\kappa]\delta    +  (\check{b}[\kappa]\check{H}))
-  *\delta_{-\tau[\kappa]}*\partial_t f *u,
+  S[\kappa]^Tu = \rho( a[\kappa]\delta    -  (\check{b}[\kappa]\check{H}))
+ *\delta_{-\tau[\kappa]}*\partial_t \check{f} *u,
 \end{equation}
@@
   \label{thm:snorm}
-  Under the assumptions G1-G4 leading to \ref{eqn:fwdgreen}, the
-  expressions \ref{eqn:sgood} and \ref{eqn:sadjgood} define
+  Under the Exponential Decay, Single Arrival, and No Return assumptions, the
+  expression \ref{eqn:sgood} define
   bounded operators on
@@
 \end{lemma}
-\begin{proof} From the estimates H3, $b[\kappa](\bx_r,\cdot;\bx_s)H *
+\begin{proof} From the assumptions about $f$ and bounds
+  \ref{eqn:best},  $b[\kappa](\bx_r,\cdot;\bx_s)H *
   \partial_tf \in {\cal S}(\bR)$ for all $(\bx_s,\bx_r) \in X_{sr}$,
@@
 
-The conventional remedy for instability in a deconvolution problem of
-this type is so-called ``pre-whitening'', as
-mentioned by \cite{Warner:16} ``Pre-whitening'' is a version of
-Tihonov regularization. It  amounts to replacement of the equation
-$S[\kappa]u=d$ by the regularized least squares problem: 
-\begin{equation}
-  \label{eqn:reg}
-  u_{\sigma}[\kappa;d] = \mbox{argmin}_{u \in L^2(X_{sr} \times \bR)} (\|S[\kappa]u-d\|^2 + \sigma\|u\|^2)
-\end{equation}
-For future reference, examine the
-implications of definition \ref{eqn:reg} for the units of the regularization weight $\sigma$,
-which  is
-necessarily dimensional. Since the data $d$ has dimensions of
-pressure, so must $S[m]u = F[m]*u$. Of course $F[m]$ is also sampled
-pressure, and convolution is a $t$ integration, so for $S[m]u$ to be a
-pressure, $u$ must have dimensions of 1/time. Therefore
-$\|S[m]u-d\|^2$ has dimensions of pressure$^2 \times $time, $\|u\|^2$ has
-dimensions of 1/time, whence $\sigma$ must be dimensioned as
-pressure$^2 \times$  time$^2$.
-
 \begin{prop}
   \label{thm:unorm}
-  Under the assumptions G1 - G4 justifying the
-  representation \ref{eqn:fwdgreen}, the least squares problem
-  \ref{eqn:reg} with $\sigma > 0$ has a unique solution
+  Under the assumptions of Lemma \ref{thm:snorm}, the least squares problem
+  \ref{eqn:regs} with $\sigma > 0$ and $d\in L^2(X_{sr} \times \bR)$ has a unique solution
   $u_{\sigma}[\kappa;d] \in L^2(X_{sr} \times \bR)$ satisfying
@@
 For $\sigma > 0$, the Lax-Milgram Theorem (\cite{Yosida}, section
-III.7) assures the existence of unique solution. 
+III.7) assures the existence of unique minimizer
+$u_{\sigma}[\kappa;d]$ satisfying the  normal
+equation
+\begin{equation}
+  \label{eqn:normal}
+  (S[\kappa]^TS[\kappa] + \sigma I) u_{\sigma}[\kappa;d] = S[\kappa]^Td,
+\end{equation}
+and the bound follows from the inequality \ref{eqn:snorm}.
 \end{proof}
-
-Define the modified
-AWI objective $J_{\rm AWI,\sigma}$ by
-replacing $u$ in the definition \ref{eqn:awideforig} with $u_{\sigma}$:
-\begin{equation}
-  \label{eqn:awidef}
-  J_{\rm AWI,\sigma}[\kappa;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu_{\sigma}[\kappa;d](\bx_r,t;\bx_s)|^2}{\int dt |u_{\sigma}[\kappa;d](\bx_r,t;\bx_s)|^2}\right). 
-\end{equation}
-To make this definition a bit more concise, introduce the operator $T$ on
-${\cal D}'(X_{sr} \times \bR)$ by
-\[
-  Tp(\bx_r,t;\bx_s) = tp(\bx_r,t;\bx_s).
-\]
-Then
-\begin{equation}
-  \label{eqn:awidefbis}
-  J_{\rm AWI,\sigma}[\kappa;d] = \sum_{(\bx_s,\bx_r) \in X_{sr}}
-  \frac{\|Tu_{\sigma}[\kappa;d](\bx_r,\cdot;\bx_s)\|^2}{\|u_{\sigma}[\kappa;d](\bx_r,\cdot;\bx_s)\|^2}.
-\end{equation}
-Here $\| \cdot \|$ is the norm in $L^2(\bR)$.
-
-Define a regularized version of the MSWI objective function $J_{\rm
-  MSWI}$ (definition \ref{eqn:mswideforig}) similarly:
-\begin{equation}
-  \label{eqn:mswidef}
-  J_{\rm MSWI,\sigma}[\kappa;d] = \|Tu_{\sigma}[\kappa;d]\|^2.
-\end{equation}
-
-The definitions \ref{eqn:awidefbis} and \ref{eqn:mswidef} have introduced a new defect: there is no
-guarantee that $Tu_{\sigma}[\kappa;d] \in L^2(X_{sr} \times\bR)$.
 
@@
   \label{thm:tl2}
-  Suppose that $d, Td \in L^2(X_{sr} \times \bR)$. Then
+  In addition to the Exponential Decay, Single Arrival, and No Return assumptions, uppose that $ Td \in L^2(X_{sr} \times \bR)$. Then
   $Tu_{\sigma}[\kappa;d]  \in L^2(X_{sr} \times \bR)$, and for each
@@
   \]
-  \begin{equation}
-    \label{eqn:tunorm}
+  \[
     \frac{C}{\sigma}\left(\left(
         \frac{\|\partial_t f\|^2_{L^1(\bR)}}{\sigma}(\|f\|_{L^1(\bR)}+\|T\partial_tf\|_{L^1(\bR)})
-      \right)\|d\| \\ 
+      \right)\|d\|\right.
+  \]
+  \begin{equation}
+    \label{eqn:tunorm}
+    \left.
     + \|\partial_t  f\|_{L^1(\bR)}(\|d (\bx_r,\cdot;\bx_s)\| +\|Td (\bx_r,\cdot;\bx_s)\|)\right).
@@
   \label{thm:noisefree}
-  Suppose that $\kappa \in M$. Then for $p \ge 1$, $F[\kappa], TF[\kappa]
+  Assume Exponential Decay, Single Arrival, and No Return. Then for
+  $\log \kappa \in M$,r $p \ge 1$, $F[\kappa], TF[\kappa]
   \in L^p(X_{sr} \times \bR)$, and there exist $C_0, C_1, C_2 >0$ so
@@
 
-Note that in the representation \ref{eqn:fwdgreen}, only the
+Note that in the first of the two representations in display \ref{eqn:sgood}, only the
 derivative $\partial_tf$ appears. For convenience, introduce
@@
 \end{equation}
-Then
-\begin{itemize}
-\item[W1. ] $w \in C^{\infty}_0(\bR)$;;
-\item[W2. ] $\int \, w = 0$.
-\end{itemize}
-I will also
+Then  $w \in C^{\infty}_0(\bR)$ and $\int \, w = 0$. I will also
 refer to $w$ as ``the wavelet'', and view $F$ as parametrized by $w$.
@@
 
-The modeling operator with wavelet $\wl$ is denoted $\Fl$. From the
-decomposition \ref{eqn:fwdgreen}, $\Fl$ is the sum of two terms:
+The modeling operator with wavelet $\wl$ is denoted $\Fl$
+(consistently with the prior definition using $\fl$).. From the
+decomposition \ref{eqn:green}, $\Fl$ is the sum of two terms:
 \begin{equation}

### 2024-11-07 repro/papers/awi/paper.tex
more

@@
 {\em filter} $u \in L^2(X_{sr} \times \bR)$. That is, the residual
-data $F[\kappa]-d$ is replaced by the filtered residual data $F[m]*u-d$.
+data $F[\kappa]-d$ is replaced by the filtered residual data $F[\kappa]*u-d$.
 Since $F[\kappa] \in L^1(X_{sr}\times \bR)$, the filtererd residual
@@
   Assume Exponential Decay, Single Arrival, and No Return. Then for
-  $\log \kappa \in M$,r $p \ge 1$, $F[\kappa], TF[\kappa]
+  $\log \kappa \in M$, $p \ge 1$, $F[\kappa], TF[\kappa]
   \in L^p(X_{sr} \times \bR)$, and there exist $C_0, C_1, C_2 >0$ so
@@
   A simple estimate shows that $F[\kappa]$ satisfies an exponential
-  decay estimate of the form G4, whence the first estimate
+  decay estimate of the form \ref{eqn:best} whence the first estimate
   follows. Since $TbH \in W^{1,1}(\bR)$,
@@
 
-\subsection{Linking Regularization and Wavelength; Statement of the
-  Main Result}
-In the remainder of this section, I will examine the $\lambda$-parametrized families of
-objective functions obtained by
-\begin{itemize}
-\item[1. ] choosing the Tihonov regularization weight in the least
-  square problem \ref{eqn:reg} as $\sigma = r\lambda$, in which $r$ is
-  a positive constant with dimensions pressure$^2 \times $ time$^2$.
-\item[2. ]replacing $S$, $d$ with $\Sl$, $\dl =\Fl[\kappa^*]$
-  (noise-free data) in the definition of $\ul$ (equation \ref{eqn:reg}).
-\end{itemize}
-
-Justification for choosing the regularization weight $\sigma$ to be
-proportional to
-the wavelength proxy $\lambda$ will emerge from the analysis relating
-$\|T\ul[\kappa;\dl]\|$ to the travel
-time error $|\tau[\kappa]-\tau[\kappa^*]$. The necessary units for the
-proportionality constant $r$ follow from the discussion after equation
-\ref{eqn:reg}. It will be clear that choosing $\sigma$ a function of
+\subsection{Linking Regularization and Wavelength}
+Note that noise-free data $\dl = \Fl[\kappa]$ is bounded in $L^2(X_{sr}
+\times \bR)$ uniformly in $\kappa$ and $\lambda$  (Proposition
+\ref{eqn:noisefree}, estimates \ref{eqn:Frembd}). 
+The minimum value of the right-hand side in equation \ref{eqn:regs} is
+the length of the projection of $(\dl,0)^T$ onto the range of
+$(\Sl[\kappa[],\sigma I)^T$, so is less than $\|\dl^2\|$.
+
+Suppose conversely that $\Sl[\kappa]\ul[\kappa;d]\| \ge
+\frac{1}{2}\|d\|$. Then
+\[
+  \|d\| \le 2\|\Fl[\kappa]*\ul[\kappa;d] \le
+  2\lambda^{\frac{1}{2}}\|\ul[\kappa;d]\|
+\]
+from Young's inequality, and from the bound \ref{eqn:unorm} this is in
+turn
+\[
+  \le C \frac{\lambda}{\sigma} \|d\|
+\]
+whence
+\[
+  \sigma \le C \lambda.
+\]
+This inequality suggests that in order to expect filtered predicted data
+$\Sl[\kappa]\ul[\kappa;d\|$ to remian uniformly (in $\lambda$) close to the data to be
+inverted, as the heuristic justification for AWI suggests, the
+regularization weight $\sigma$ must be coupled to the wavelength proxy
+$\lambda$.  It will be clear from the discussion to follow that choosing $\sigma$ a function of
 $\lambda$ roughly proportional to $\lambda$, that is, $C_*\lambda \le
 \sigma(\lambda) \le C^*\lambda$ for $C_*,C^*$ satisfying $0 < C_* \le
-C^*$, would lead to the same asymptotic estimates for the relation
-between the adaptive filter $\ul$ and travel time
-error. The argument is slightly cleaner if $\sigma$ is simply
+C^*$, would suffice. The argument is slightly cleaner if $\sigma$ is simply
 proportional to (non-dimensional) $\lambda$ by a necessarily
-dimensional constant..
+dimensional constant.. Therefore we assume from now on that
+\begin{equation}
+  \label{eqn:lamsig}
+  \sigma = r \lambda
+\end{equation}
+in which $r$ is a positive constant with dimensions pressure$^2 \times
+$ time$^2$.
+
+This choice will eventually be justified by a converse estimate, that
+is, showing that under the condition \ref{eqn:lamsig}, the filtered
+predicted data $\Sl[\kappa]\ul[\kappa;d]$ has norm at least half that
+of $\|\dl\|$ for all $\lambda \in (0,1]$, for properly chosen $r$.
 
@@
 
+\section{BLAH BLAH}
+
+Assume as in the last section, that is, that solutions
+of the acoustic system \ref{eqn:awe} is well-approximated by the
+geometric asymptotics as in equation \ref{eqn:hadamard}, and that the
+data is noise-free ($d=F[\kappa^*]$). Together with the properties of $w$
+just listed, these assumptions imply that
+\begin{equation}
+  \label{eqn:eq11}
+  J_{0,\sigma}[\kappa;d] = O(\sigma) + O(\lambda(w)).
+\end{equation}
+
+\begin{proof}
+  Since the arguments of $\Sl$,... are just along for the ride
+  suppress them, as before.
+  First analyse the leading order approximation: from equations
+  \ref{eqn:hatuf} and \ref{eqn:hatgf},
+  \[
+  \|\Sla \uzl -\dzl\|^2 = \frac{(a^*)^2}{2\pi} \int d\omega \left| \fwl(\omega) \frac{|\fwl(\omega)|^2}{|\fwl(\omega)|^2 + \frac{r\lambda}{a^2}}
+    - \fwl(\omega)\right|^2
+\]
+\[
+  =r^2\lambda^2\left(\frac{(a^*)^2}{2\pi a^4}\right) \int d\omega \frac{| \fwl(\omega)|^2}
+  {(|\fwl(\omega)|^2 + \frac{r\lambda}{a^2})^2}
+\]
+\[
+  =r^2\lambda^2\left(\frac{(a^*)^2}{2\pi a^4}\right) \int d\omega
+  \frac{| \lambda^{\frac{1}{2}}w_1(\lambda \omega)|^2}
+  {(|\lambda^{\frac{1}{2}}w_1(\lambda \omega)|^2 + \frac{r\lambda}{a^2})^2}
+\]
+\[
+  = \frac{1}{2\pi}\left(\frac{ra^*}{a}\right)^2 \int d\omega
+  \frac{|aw_1(\omega)|^2}
+  {(|aw_1(\omega)|^2 + r)^2}
+\]
+which depends on $M$ (and $r$) but not on $\lambda$. From the result
+\ref{eqn:u0snormlam}, the same is true of the second term in the
+objective function defined in the third equation  of the display
+\ref{eqn:leading}: that is,
+\begin{equation}
+  \label{eqn:leadingconst}
+  \|\Sla \uzl -\dzl\|^2 + r\lambda \|\uzl\|^2 = C
+\end{equation}
+
+\end{proof}
+To see this, begin with the definition:
+\[
+  \tilde{J}_{0,\sigma}[\kappa;d] = \frac{1}{2}
+  (\|S_{\lambda}[\kappa]u_0[\kappa_{\lambda};d]-d\|^2 + \lambda \|u_0[\kappa_{\lambda};d]\|^2.
+\]
+Using the approximate identities \ref{eqn:hatuf} and \ref{eqn:hatgf} and
+rearranging,
+\[
+\|\Sl[\kappa]u_0[\kappa_;d]-d\|^2 
+  \approx \int d\omega
+  |a\hat{w}\hat{u_0[\kappa_{\lambda};d]}e^{i\omega\tau}-a^*\hat{w}e^{i\omega\tau^*}|^2
+\]
+\[
+  = \int d\omega
+  (a^*)^2|\hat{w}|^2\left(\frac{\lambda}{a^2|\hat{w}|^2 +
+      \lambda}\right)^2   
+\le \int d\omega (a^*)^2|\hat{w}|^2
+  \left(\frac{\lambda}{a^2|\hat{w}|^2 + \lambda}\right)
+\]
+\begin{equation}
+  \label{eqn:eq12}
+  = \lim_{\Omega \rightarrow \infty} \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2\lambda
+  \left(\frac{a^2|\hat{w}|^2}{a^2|\hat{w}|^2 + \lambda}\right)
+\end{equation}
+For any $\Omega > 0$,
+\[
+\int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2\lambda\left(\frac{a^2|\hat{w}|^2}{a^2|\hat{w}|^2 + \lambda}\right)
+\le \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2 \lambda
+  \left(\frac{ a|\hat{w}|}{(a^2|\hat{w}|^2 +
+      \lambda)^{\frac{1}{2}}}\right)
+\]
+\[
+  = \sigma \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2 a|\hat{w}|
+  \left(\frac{\sigma }{(a^2|\hat{w}|^2 +\lambda)^{\frac{1}{2}}}\right)
+  \le \sigma \int_{-\Omega}^{\Omega} d\omega \frac{(a^*)^2}{a}
+  |\hat{w}|
+  \le \sigma \frac{(a^*)^2}{a} \|\hat{w}\|_{L^1(\bR)}.
+\]
+Together with the approximate inequality \ref{eqn:eq12}, this
+inequality implies
+that
+\[
+  \|S_{\lambda}[\kappa]u_0[\kappa_{\lambda};d] - d\|^2 = O(\sigma) + O(\lambda(w)).
+\]
+A similar argument leads to the conclusion that
+\[
+  \lambda\|u_0[\kappa_{\lambda};d]\|^2 = O(\sigma) + O(\lambda(w)).
+\]
+Together, these conclusions imply the assertion \ref{eqn:eq11}.
+
+%%%%%%%%%%%%%%%%%%%%
+
 \bibliographystyle{seg}

### 2024-11-08 repro/papers/awi/paper.tex
almost there

@@
 The relation between the dimensionless scale parameter $\lambda$ and
-RMS wavelength will be explained below. Correspondingly, denote the
+RMS wavelength will be explained below: $\lambda$ may be regarded as a
+proxy for wavelength. Correspondingly, denote the
 modeling operator with $f=\fl$ by $\Fl$.
@@
   hold for $X_{sr} \subset \Omega \times \Omega, M \subset C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)$.
-  Select $r >0$ with units of
-  pressure$^2 \times$time$^2$ and $\log \kappa^* \in M$. For $0 <
-  \lambda $, set
+  Select $\log \kappa^* \in M$. For $0 <
+  \lambda \le 1$, set
   $\dl=\Fl[\kappa^*]$ (``noise-free data''), and  define $\ul[\kappa,\dl]$
   to be the solution of 
-  equation \ref{eqn:reg} with  $\sigma=r\lambda$, $F=\Fl$ and $d=\dl$..
-  There exist $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
+  equation \ref{eqn:reg} with $F=\Fl$ and $d=\dl$..
+
+  \begin{itemize}
+  \item[1. ] $Tu_{\sigma}[\kappa;\dl](\bx_r,\cdot;\bx_s) \in
+    L^2(\bR)$;
+  \item[2. ] There exist $R_*,R^*,\lambda_*$ with $0<R_*\le R^*, 0 <
+    \lambda_*\le 1$ depending on
+    $M, X_{sr}$ so that
+    \begin{itemize}
+    \item[a. ] $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
+      \frac{1}{2}\|\dl\|$ implies that $\sigma \le R^*\lambda$;
+    \item[b. ] $\sigma \le R_*\lambda$ and $\lambda \le \lambda_*$
+      implies that $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
+      \frac{1}{2}\|\dl\|$.
+    \end{itemize}
+  \end{itemize}
+  Choose $r > 0$ and assume that
+  \begin{equation}
+    \label{eqn:lamsig}
+    \sigma = r \lambda.
+  \end{equation}.
+  Then there exist $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
   \rightarrow \bR^+$ with $C_* \le
@@
   \begin{itemize}
-  \item[1. ] $Tu_{\sigma}[\kappa;\dl](\bx_r,\cdot;\bx_s) \in
-    L^2(\bR)$;
-  \item[2. ] for $\lambda > 0$,
+  \item[2. ] for $0 <  \lambda \le 1$, 
     \begin{equation}
@@
 \end{theorem}
+\noindent{\bf Remark:} The heuristic justification for AWI is the possibility of keeping the
+filtered predicted data $\Fl[\kappa]*\ul[\kappa;d]$ close to the data
+$d$ while updating the bulk modulus $\kappa$. The conclusions 2a and
+2b show that for this to be the case, the regularization parameter $\sigma$
+must be dominated by the wavelength proxy $\lambda$. This observation underlies the choice
+$\sigma = r\lambda$ used in conclusions 3 and 4. A looser relation
+would clearly suffice, however the argument lis slightly cleaner if
+$\sigma$ is simply held proportional to $\lambda$ as $\lambda
+\rightarrow 0$. Note that the units of $r$ and $\sigma$ are pressure$^2 \times$time$^2$.
 
@@
 
-Hadamard's construction is local, in both space-time and in the set of
+Hadamard's decomposition is local, in both space-time and in the set of
 acoustic models. In the present context, it requires the Single
@@
 
-\cite{Friedlander:75} produces for each $\log \kappa \in M$ functions $\tau[\kappa]$ (travel time), $a[\kappa]$
+The construction described by \cite{Friedlander:75} produces for each $\log \kappa \in M$ functions $\tau[\kappa]$ (travel time), $a[\kappa]$
 (geometric amplitude) on $X_s \times \Omega$ and $b[\kappa]$
@@
 A by-product of the construction is a bound $A>0$ so that for any $\log \kappa \in M$
-  and $(\bx_s,\bx_r) \in X_{sr}$,  $|\log a[\kappa](\bx_s,\bx_r)| <
-  A$;  With no further assumptionss, there is also a uniform bound on $b(\bx_r,t;\bx_s)$ and
+and $(\bx_s,\bx_r) \in X_{sr}$,
+\begin{equation}
+  \label{eqn:aest}
+  |\log a[\kappa](\bx_s,\bx_r)| <  A.
+\end{equation}
+With no further assumptionss, there is also a uniform bound on
+$b(\bx_r,t;\bx_s)$ and
 its derivatives over any finite $t$-interval. However, with the
 addition of the Exponetial Decay assumption, the bound becomes
-  global in time: for any $k \ge 0$, there exists $B \ge 0,\delta > 0$ so
-  that for any $(\bx_s,\bx_r) \in X_{sr}$, $\log \kappa \in M$, 
+global in time: for any $k \ge 0$, there exists $B \ge 0,\delta > 0$
+so that for any $(\bx_s,\bx_r) \in X_{sr}$, $\log \kappa \in M$,
 \begin{equation}
@@
   \label{eqn:wfam}
-  \wl(t) = \lambda^{-1/2}w_1(t/\lambda).
+  \wl(t) = \lambda^{-1/2}w_1\left(\frac{t}{\lambda}\right).
 \end{equation}
@@
 
-Note that the AWI objective defined in \ref{eqn:awideforig} and its
-modification \ref{eqn:awidef} are actually
-squared pulse widths of the adaptive filters $u[\kappa;d]$ and
+Note that the AWI objective defined in \ref{eqn:awidefmain} is actually
+squared pulse width of the adaptive filter 
 $u_{\sigma}[\kappa;d]$ respectively.
@@
 \times \bR)$ uniformly in $\kappa$ and $\lambda$  (Proposition
-\ref{eqn:noisefree}, estimates \ref{eqn:Frembd}). 
+\ref{thm:noisefree}, estimates \ref{eqn:Frembd}). 
 The minimum value of the right-hand side in equation \ref{eqn:regs} is
@@
 
-Suppose conversely that $\Sl[\kappa]\ul[\kappa;d]\| \ge
-\frac{1}{2}\|d\|$. Then
+Suppose conversely that $\|\Sl[\kappa]\ul[\kappa;d] - \dl\| \le
+\frac{1}{2}\|d\|$. Then $\|\Sl[\kappa]\ul[\kappa;d] \| \ge
+\frac{1}{2}\|d\|$, so
 \[
@@
 \]
-from Young's inequality, and from the bound \ref{eqn:unorm} this is in
+from Young's inequality, Proposition \ref{thm:noisefree}, and
+inequalities \ref{\eqn:lambdanorms}. From the bound \ref{eqn:unorm} this is in
 turn
 \[
-  \le C \frac{\lambda}{\sigma} \|d\|
+  \le R^* \frac{\lambda}{\sigma} \|d\|
 \]
@@
 \[
-  \sigma \le C \lambda.
-\]
-This inequality suggests that in order to expect filtered predicted data
+  \sigma \le R^* \lambda.
+\]
+for a suitable $R^*>0$ independent of $\log \kappa \in M$ and $\lambda
+\in (0,1]$. This establishes conclusion 2a in Theorem \ref{thm:main}, and
+suggests that in order to expect filtered predicted data
 $\Sl[\kappa]\ul[\kappa;d\|$ to remian uniformly (in $\lambda$) close to the data to be
@@
 regularization weight $\sigma$ must be coupled to the wavelength proxy
-$\lambda$.  It will be clear from the discussion to follow that choosing $\sigma$ a function of
-$\lambda$ roughly proportional to $\lambda$, that is, $C_*\lambda \le
-\sigma(\lambda) \le C^*\lambda$ for $C_*,C^*$ satisfying $0 < C_* \le
-C^*$, would suffice. The argument is slightly cleaner if $\sigma$ is simply
-proportional to (non-dimensional) $\lambda$ by a necessarily
-dimensional constant.. Therefore we assume from now on that
-\begin{equation}
-  \label{eqn:lamsig}
-  \sigma = r \lambda
-\end{equation}
+$\lambda$. Therefore we assume from now on that equation
+\ref{eqn:lamsig} holds,  
 in which $r$ is a positive constant with dimensions pressure$^2 \times 
@@
 
-This choice will eventually be justified by a converse estimate, that
-is, showing that under the condition \ref{eqn:lamsig}, the filtered
-predicted data $\Sl[\kappa]\ul[\kappa;d]$ has norm at least half that
-of $\|\dl\|$ for all $\lambda \in (0,1]$, for properly chosen $r$.
-
 \subsection{The leading term and travel time}
@@
   \label{thm:leading}
-In the definitions \ref{eqn:mswidef} of $J_{\rm MSWI,\sigma}$ and
-\ref{eqn:awidefbis} of $J_{\rm AWI,\sigma}$, set $\sigma = r\lambda$
+In the definitions \ref{eqn:mswidefmain} of $J_{\rm MSWI,\sigma}$ and
+\ref{eqn:awidefmain} of $J_{\rm AWI,\sigma}$, set $\sigma = r\lambda$
+(equation \ref{eqn:lamsig}
 and replace the data $d$ with
@@
   \left| \frac{\|T\uzl[\kappa;\dla] (\bx_r,\cdot;\bx_r)\|^2}{
-        \|\uzl[\kappa;\dla] (\bx_r,\cdot;\bx_r)\|^2} - (\tau[\kappa^*]({\bf
-  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2\right|  \le C\lambda^2.
-\end{equation}
-m\end{prop}
+        \|\uzl[\kappa;\dla] (\bx_r,\cdot;\bx_r)\|^2} - (\tau[\kappa^*](\bx_s,\bx_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2\right|  \le C\lambda^2.
+\end{equation}
+Finally, for $\sigma = r\lambda$ with $r$ sufficiently small,
+\begin{equation}
+  \label{eqn:Sutz}
+  \Sla[\kappa]\uzl[\kappa;\dzl] \ge \frac{1}{2}\|\dzl\|
+\end{equation}
+\end{prop}
 
@@
 
-%It follows that
-%\begin{equation}
-%  \label{eqn:mswitt}
-%  J_{\rm MSWI,\lambda,\lambda}[\kappa;\dla] =
-%  \frac{1}{\lambda} \sum_{{\bf x}_s,{\bf x}_r} C[\kappa](\bx_s,\bx_r)(\tau[\kappa^*]({\bf
-%  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 + O(\lambda)
-%\end{equation}
-%  
-%\begin{equation}
-%  \label{eqn:awitt}
-%J_{\rm AWI,\lambda,\lambda}[\kappa;\dla] = \sum_{{\bf x}_s,{\bf x}_r} (\tau[\kappa^*]({\bf
-%  x}_s,{\bf x}_r)-\tau[\kappa]({\bf x}_s,{\bf x}_r))^2 + O(\lambda^2)
-%\end{equation}
-
+Similar reasoning establishes the inequality \ref{eqn:Sutz}:
+\[
+  \Sla[\kappa]\uzl[\kappa;\dla](\omega) = a^* \wl(\omega) \left(\frac{|\wl(\omega)|^2}{|\wl(\omega)|^2
+      + \frac{\sigma}{a^2}}\right)
+\]
+\[
+  = a^* \lambda^{\frac{1}{2}}w_1(\lambda \omega) \left(\frac{a^2|w_1(\lambda\omega)|^2}{a^2|w_1(\lambda\omega)|^2
+      + \frac{\sigma}{\lambda}}\right),
+\]
+so
+\[
+  \|\Sla[\kappa]\uzl[\kappa;\dla]\|^2 = \left(\frac{a*}{a}\right)^2
+  \int a^2|w_1|^2\left(\frac{a^2|w_1|^2}{a^2|w_1|^2 +
+      \frac{\sigma}{\lambda}}\right)^2 \ge \frac{\lambda}{\sigma}
+  (a^*)^2a^2 \int |w_1|^4 
+\]
+By virtue of the bound \ref{eqn:aest}, $(a^*)^2a^2 \int |w_1|^4$ is bounded below
+by a positive constant independent of $\log \kappa \in M,
+(\bx_s,\bx_r) \in X_{sr}$. Since $\dl$ is bounded above by a similar
+constant, conclude if $r = \sigma/\lambda$ is sufficiently small, the
+inequality \ref{eqn:Sutz} holds.
 \end{proof}
 
-\noindent {\bf Remark:} Note that the homogeneity in $\lambda$ of the
-expressions in equations \ref{eqn:g0} and \ref{eqn:g1} results from
-the proportionality between $\lambda$ and the regularization weight in
-the definitions \ref{eqn:reg} and \ref{eqn:leading}. Otherwise, such
-simple asymptotic relations as the conclusions
-of Proposition \ref{thm:leading} would not obtain.
+\noindent {\bf Remark:} Inequality \ref{eqn:Sutz} is the converse
+promised earlier to the observation that requiring the filtered
+predicted data $\Sl[\kappa]u_{\sigma}[\kappa;\dl]$, hence the filtered
+data residual, to be comparable in norm to the data $d$ implied that
+$\sigma$ should be dominated by $\lambda$, as the relation
+$\sigma=r\lambda$ ensures. Of course, this is an analogue, with $\Sl$
+replaced by its leading approximation $\Sla$, and $\dl$ replaced by $\dla$.
 
@@
 
-\begin{proof} {\em of Theorem \ref{thm:main}}: From the inequality \ref{eqn:mswittz}
+\begin{proof} {\em of Theorem \ref{thm:main}}:
+Conclusions 1 and 2a have already been established. Conclusion 2b
+follows from equation \ref{eqn:Sutz} in Proposition \ref{thm:leading},
+combined with the first estimate in \ref{eqn:remest} and the bounds
+\ref{eqn:Srembd}.
+
+  From the inequality \ref{eqn:mswittz}
   (suppressing the arguments $\bx_s,\bx_r$),

### 2024-11-08 repro/papers/awi/paper.tex
more cleanup

@@
 from Young's inequality, Proposition \ref{thm:noisefree}, and
-inequalities \ref{\eqn:lambdanorms}. From the bound \ref{eqn:unorm} this is in
+inequalities \ref{eqn:lambdanorms}. From the bound \ref{eqn:unorm} this is in
 turn
@@
 follows from equation \ref{eqn:Sutz} in Proposition \ref{thm:leading},
-combined with the first estimate in \ref{eqn:remest} and the bounds
+combined with the first estimate in \ref{thm:remest} and the bounds
 \ref{eqn:Srembd}.
@@
 \ref{eqn:reg} with $S=\Sl$, $d=\dl$, and $J_{\rm MSWI,\sigma}$ is as
-defined in equation \ref{eqn:mswidef}.
+defined in equation \ref{eqn:mswidefmain}.
 \end{prop}
@@
 suggests that the contribution of the second term on the
-left-hand-side of the limit \ref{eqn:cor2} is insignificant for small
+left-hand-side of the limit \ref{eqn:cor1mswi} is insignificant for small
 $\lambda$, whereas it is comparable to the first term in
@@
 \section{Multiple Arrivals and Cycle-Skipping}
-The geometric asymptotics approximation \ref{eqn:fwdgreen} holds only
+The geometric asymptotics approximation \ref{eqn:green} holds only
 in the ``single arrival'' case, that is, when sources and receivers
@@
 strongly refracting models, using frequency-domain implementation of
-the basic definition \ref{eqn:mswidef} and time-domain implementation
+the basic definition \ref{eqn:mswidefmain} and time-domain implementation
 of the reduced penalty function \ref{eqn:mswipenred} respectively. \cite{Yongetal:GJI23}
@@
 $\tau^*_1-\tau_0$ and $\tau_1-\tau^*_0$ respectively in the
-definitions \ref{eqn:mswidef} and \ref{eqn:awidefbis} of the MSWI and AWI
+definitions \ref{eqn:mswidefmain} and \ref{eqn:awidefmain} of the MSWI and AWI
 objectives. For other choices of objective, they contribute similarly
@@
 
-These constraints are discussed in detail in the next section. Some
+Some of these constraints
 could certainly be relaxed. For instance, exponential decay of local
@@
 can be treated as a perturbation of the infinite time data setting
-posed here. The exception is the single-arrival hypothesis: it is
+posed here. The conclusions of this paper likely hold in some form
+even for non-smooth sets of models, provided that the data for such models is
+dominated by ballistic transmitted waves. The exception is the single-arrival hypothesis: it is
 essential, as without it, the link between the travel time and
@@
 
-In any case, these hypotheses enable a simple and clear
-analysis of AWI, and seem a reasonable starting point for further
-investigation.
-
-In this paper, I have shown that AWI is a good approximation to least
-squares travel time inversion of transmitted wave data, up to errors proportional to a typical
-data wavelength and to a regularization (``pre-whitening'')
-weight. However, the argument leading to this conclusion works only if
-the material model is assumed slowly varying so that geometric
-acoustics accurately describes the solution of the wave equation, and if
-sources and receivers are sufficiently close to one another to be
-connected by unique rays. The first condition may be ignored to some
-extent - the conclusion of this paper likely holds if the data is
-dominated by ballistic transmitted waves. The second condition
-is essential: even for purely transmitted waves, the presence of
-multiple arrivals breaks the connection
-with travel time inversion, and AWI and similar algorithms
-become essentially as prone to cycle-skipping as is FWI.
-
-%%%%%%%%%%%%%%%%%%
-
 The analysis presented here has nothing to say about AWI applied to
@@
 
-%%%%%%%%%%%%%%%%%%%%
-
-\subsection{AWI via Optimization}
-
-AWI is based an {\em extension} of the modeling operator $F$, that is,
-an operator ${\bar F}$ with larger domain, coinciding with $F$ on its domain..
-The extended modeling operator ${\bar F}$ maps extended models
-$\bar{m}=(\kappa,\bar{f})$ incorporating extended sources
-$\bar{f}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
-field. Denote by $\bar{p}(\bx,t;\bx_s,\bx_r), \bar{\bv}(\bx,t;\bx_s,\bx_r)$
-the solution of the acoustic system \ref{eqn:awe} with $f(t)$ in the
-right-hand side of the first equation replaced by
-$\bar{f}(\bx_r,\cdot;\bx_s)$. Then
-$\bar{F}[\kappa,\bar{f}](\bx_r,t;\bx_s) =
-\bar{p}(\bx_r,t;\bx_s,\bx_r)$. As was the case for $F$, the Hadamard
-decomposition \ref{eqn:green} and conditions G1-G4, W1-W3 imply that
-$\bar{F}[\kappa, \cdot]$ is a bounded map: $L^2(X_{sr} \times \bR)
-\rightarrow L^2(X_{sr} \times \bR)$. If all  
-of the wavelets are the same, that is, ${\bar f}({\bf  
-  x}_r,t;{\bf x}_s) = f(t)$ is independent of source and receiver  
-position, then ${\bar F}[\kappa,\bar{f}] = F[\kappa]$. That is, $F$ is a special
-case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
-$F$. This is the {\em source-receiver} extension, in the terminology of \cite{HuangSymes2015SEG}.
-
-\cite{Warner:16} represent the extended sources $\bar{f}$ as time convolutions of the
-(known) source $f$ with an {\em adaptive filter} $u \in L^2(X_{sr}
-\times \bR)$, that is, $\bar{f}=u * f$.
-
-\section{BLAH BLAH}
-
-Assume as in the last section, that is, that solutions
-of the acoustic system \ref{eqn:awe} is well-approximated by the
-geometric asymptotics as in equation \ref{eqn:hadamard}, and that the
-data is noise-free ($d=F[\kappa^*]$). Together with the properties of $w$
-just listed, these assumptions imply that
-\begin{equation}
-  \label{eqn:eq11}
-  J_{0,\sigma}[\kappa;d] = O(\sigma) + O(\lambda(w)).
-\end{equation}
-
-\begin{proof}
-  Since the arguments of $\Sl$,... are just along for the ride
-  suppress them, as before.
-  First analyse the leading order approximation: from equations
-  \ref{eqn:hatuf} and \ref{eqn:hatgf},
-  \[
-  \|\Sla \uzl -\dzl\|^2 = \frac{(a^*)^2}{2\pi} \int d\omega \left| \fwl(\omega) \frac{|\fwl(\omega)|^2}{|\fwl(\omega)|^2 + \frac{r\lambda}{a^2}}
-    - \fwl(\omega)\right|^2
-\]
-\[
-  =r^2\lambda^2\left(\frac{(a^*)^2}{2\pi a^4}\right) \int d\omega \frac{| \fwl(\omega)|^2}
-  {(|\fwl(\omega)|^2 + \frac{r\lambda}{a^2})^2}
-\]
-\[
-  =r^2\lambda^2\left(\frac{(a^*)^2}{2\pi a^4}\right) \int d\omega
-  \frac{| \lambda^{\frac{1}{2}}w_1(\lambda \omega)|^2}
-  {(|\lambda^{\frac{1}{2}}w_1(\lambda \omega)|^2 + \frac{r\lambda}{a^2})^2}
-\]
-\[
-  = \frac{1}{2\pi}\left(\frac{ra^*}{a}\right)^2 \int d\omega
-  \frac{|aw_1(\omega)|^2}
-  {(|aw_1(\omega)|^2 + r)^2}
-\]
-which depends on $M$ (and $r$) but not on $\lambda$. From the result
-\ref{eqn:u0snormlam}, the same is true of the second term in the
-objective function defined in the third equation  of the display
-\ref{eqn:leading}: that is,
-\begin{equation}
-  \label{eqn:leadingconst}
-  \|\Sla \uzl -\dzl\|^2 + r\lambda \|\uzl\|^2 = C
-\end{equation}
-
-\end{proof}
-To see this, begin with the definition:
-\[
-  \tilde{J}_{0,\sigma}[\kappa;d] = \frac{1}{2}
-  (\|S_{\lambda}[\kappa]u_0[\kappa_{\lambda};d]-d\|^2 + \lambda \|u_0[\kappa_{\lambda};d]\|^2.
-\]
-Using the approximate identities \ref{eqn:hatuf} and \ref{eqn:hatgf} and
-rearranging,
-\[
-\|\Sl[\kappa]u_0[\kappa_;d]-d\|^2 
-  \approx \int d\omega
-  |a\hat{w}\hat{u_0[\kappa_{\lambda};d]}e^{i\omega\tau}-a^*\hat{w}e^{i\omega\tau^*}|^2
-\]
-\[
-  = \int d\omega
-  (a^*)^2|\hat{w}|^2\left(\frac{\lambda}{a^2|\hat{w}|^2 +
-      \lambda}\right)^2   
-\le \int d\omega (a^*)^2|\hat{w}|^2
-  \left(\frac{\lambda}{a^2|\hat{w}|^2 + \lambda}\right)
-\]
-\begin{equation}
-  \label{eqn:eq12}
-  = \lim_{\Omega \rightarrow \infty} \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2\lambda
-  \left(\frac{a^2|\hat{w}|^2}{a^2|\hat{w}|^2 + \lambda}\right)
-\end{equation}
-For any $\Omega > 0$,
-\[
-\int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2\lambda\left(\frac{a^2|\hat{w}|^2}{a^2|\hat{w}|^2 + \lambda}\right)
-\le \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2 \lambda
-  \left(\frac{ a|\hat{w}|}{(a^2|\hat{w}|^2 +
-      \lambda)^{\frac{1}{2}}}\right)
-\]
-\[
-  = \sigma \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2 a|\hat{w}|
-  \left(\frac{\sigma }{(a^2|\hat{w}|^2 +\lambda)^{\frac{1}{2}}}\right)
-  \le \sigma \int_{-\Omega}^{\Omega} d\omega \frac{(a^*)^2}{a}
-  |\hat{w}|
-  \le \sigma \frac{(a^*)^2}{a} \|\hat{w}\|_{L^1(\bR)}.
-\]
-Together with the approximate inequality \ref{eqn:eq12}, this
-inequality implies
-that
-\[
-  \|S_{\lambda}[\kappa]u_0[\kappa_{\lambda};d] - d\|^2 = O(\sigma) + O(\lambda(w)).
-\]
-A similar argument leads to the conclusion that
-\[
-  \lambda\|u_0[\kappa_{\lambda};d]\|^2 = O(\sigma) + O(\lambda(w)).
-\]
-Together, these conclusions imply the assertion \ref{eqn:eq11}.
-
-%%%%%%%%%%%%%%%%%%%%
-
 \bibliographystyle{seg}

### 2024-11-11 repro/papers/awi/paper.tex
more cleanup

@@
 
-\begin{abstract}
-Adaptive Waveform Inversion
-applied to transient transmitted wave data yields estimates of index of
-refraction (or wave velocity) similar to those obtained by travel time
-inversion, provided that the data contain a single smooth wavefront.
-\end{abstract}
+\begin{abstract} Adaptive Waveform Inversion (AWI) applied to
+  transient transmitted wave data can yield estimates of index of
+  refraction (or wave velocity) similar to those obtained by travel
+  time inversion. The AWI objective function measures normalized
+  mean-square dispersion about zero time of a family of filters, one
+  filter for each source-reeciver pair, designed to match predicted
+  data to observed data. Provided that the data contain a single
+  smooth wavefront, this function approaches the mean square
+  traveltime error as data wavelength tends to zero. The scaling of
+  each filter to have unit norm is responsible for the simple relation
+  with travel time tomography. A similar approach, Matched Source
+  Waveform Inversion (MSWI), does not normalize the filters and has a
+  looser relation with mean-square travel time error. If substantial
+  energy is spread amongst multiple arrivals, on the other hand,
+  neither AWI nor MSWI objectives approximate the travel time mean
+  square erorr, and attempts to minimize them by local
+  (Newton-related) optimization are as likely to stagnate at models
+  predicting erroneous travel times as is least-square Full Waveform
+  Inversion (FWI). The matching filter at the heart of this approach
+  must be ``pre-whitened'', that is, computed by solving a regularized
+  least squares problem for filtered data misfit. In order for the
+  relation to traveltime tomography to hold, and for the filtered data
+  misfit to stay relatively small, the regularization weight must be
+  coupled to data wavelength. These objective functions can also be
+  used as penalty terms in a homotopy of objectives connecting AWI or
+  MSWI with FWI. \end{abstract}
 
@@
 predicted wave data and actual observed
-data. This data is typicallyl oscillatory. If wave velocities or their parametric equivalents are amongst the parameters
+data. Such data is typically oscillatory. If wave velocities or their parametric equivalents are amongst the parameters
 to be estimated, then updates cause shifts in the arrival times of
@@
 This paper supplies such an explanation for a special case of AWI, the
-recovery of smooth acoustic medium parameters from data generated at
+recovery of smooth acoustic medium parameters from waves generated at
 localized energy sources, propagated through a fluid, and recorded at
@@
       \label{eqn:mswitt}
-      |\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|- \lambda^{-\frac{1}{2}} (W[\kappa^*,\kappa](\bx_s,\bx_r))^{\frac{1}{2}}|\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|| \le C
+      |\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|- \lambda^{-\frac{1}{2}} W[\kappa^*,\kappa]^{\frac{1}{2}}(\bx_s,\bx_r)|\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|| \le C
       \lambda^{\frac{1}{2}};
@@
 \lim_{\lambda \rightarrow 0} \lambda J_{\rm MSWI,r\lambda}[\kappa;\dl]
-= \sum_{(\bx_s,\bx_r) \in X_{sr}} W[\kappa^*,\kappa]^2(\bx_s,\bx_r) |\tau[\kappa](\bx_s,\bx_r) - \tau[\kappa^*](\bx_s,\bx_r)|^2.
+= \sum_{(\bx_s,\bx_r) \in X_{sr}} W[\kappa^*,\kappa](\bx_s,\bx_r) |\tau[\kappa](\bx_s,\bx_r) - \tau[\kappa^*](\bx_s,\bx_r)|^2.
 \end{equation}
@@
 
-In fact, the behaviour inherited from travel-time inversion in the
-single arrival case, namely the absence of spurious stationary points
-(``cycle-skipping''), generally fails when multiple arrivals with
-significant energy are present in the data. \cite{Symes:94c}
-illustrated this phenomenon for a version of MSWI. Crosswell
-source-receiver geometry often generates multiple arrivals,
-particularly near low-velocity sedimentary layers that are natural
-targets of seismic exploration and act as wave
-guides. \cite{Plessix:00} successfully applied a version of MSWI to
-crosswell waveform inversion by removing traces with close source and
-receiver depths, thus leaving only data that tended to be dominated by
-single arrivals. \cite{HuangSymes:Geo17} and
-\cite{SymesChenMinkoff:24} illustrate the failure of the
-version of MSWI reported here to avoid cycle-skipping in inversion of
-strongly refracting models, using frequency-domain implementation of
-the basic definition \ref{eqn:mswidefmain} and time-domain implementation
-of the reduced penalty function \ref{eqn:mswipenred} respectively. \cite{Yongetal:GJI23}
-have shown that precisely the same issue afflicts AWI, and have
-proposed a remedy by localizing the definition of the objective to
-short time intervals, within which multiple wavefront arrrivals are
-unlikely. \cite{Symes:23} constructs an extension based on surface
-sources, as opposed to the source-receiver extension which is the
-basis of MSWI and AWI. This surface source extension generates an
-invertible map from extended source to data.
-
 A simple calculation reveals the phenomenon discussed in all of these
@@
 objectives. For other choices of objective, they contribute similarly
-to the total. They will eventually begin to cancel as $m$ approaches
+to the total. They will eventually begin to cancel as $\kappa$ approaches
 $\kappa^*$, however this cancellation happens only when the times are
@@
 
-This argument shows that when multiple arrivals carry significant energy, any
-inversion approach based on the source-receiver extension, similar to
-AWI or MSWI, will cycle-skip, that is, change rapidly when the
-estimated model approaches its optimal value.
-
-It should be emphasized that this phenomenon is primarily a feature of the
-source-receiver extension, rather than of the specific objective
-function defined in terms of it. In that respect, the source-receiver
-extension is similar to the common surface parameter extensions that
-have been studied for inversion of reflection data. Inversion methods
-based on these extensions also fail to avoid cycle-skipping in the presence of
-geometrical ray complexity \cite[]{geoprosp:2008}.
-
 \section{Conclusion}
 
-The reader will
-note that a normalization component of the AWI objective function
-definition is important in reaching this conclusion. The connection
-between objective normalization and the asymptotic equivalence to
-travel time inversion for transmission data has nothing to do with the
-reasons for normalizing the objective previoiusly advanced in the
-literature \cite[]{Warner:16,LiAlkhalifah:21}.
+%The reader will
+%note that a normalization component of the AWI objective function
+%definition is important in reaching this conclusion. The connection
+%between objective normalization and the asymptotic equivalence to
+%travel time inversion for transmission data has nothing to do with the
+%reasons for normalizing the objective previoiusly advanced in the
+%literature \cite[]{Warner:16,LiAlkhalifah:21}.
+
+This paper has explained the behaviour of AWI (and MSWI) applied to
+acoustic transmission data. For data exhibiting a single smooth
+wavefront, these formulatios produce objective function that converge
+to versions of travel time mean-square error, in the limit of hgh
+data frequency. For
 
@@
 essential, as without it, the link between the travel time and
-extended waveform inversion is lost. This is a feature of the
-source-receiver extension, not of a particular objective function
-(MSWI, AWI,...) formulated in terms of it. This extension treats each receiver signal
-independently. If only a single identifiable waveform arrives at
-each receiver, AWI (and similar approaches) can implicitly identify
-its arrival time, and the link to travel time inversion is
-established. If distinct waveforms are present at multiple times,
-the single receiver signal does not determine which waveform in the
-predicted signal should align with which waveform in the data to be
-fitted. This ``crosstalk'' ambiguity is analyzed in the fifth
-section, and afflicts all approaches based on the source-receivre
+extended waveform inversion is lost.
+
+This disconnect is a fearure of the adaptive filter construction, not of a particular objective function
+(MSWI, AWI,...) formulated in terms of it. The introduction of the
+adaptive filter $u$ in the filtered data prediction $F[\kappa]*u$ is equivalent
+to permitting the source wavelet to depend on the receiver location,
+which in the basic physical model of point source acoustics it does
+not. Therefore this construction {\em extends} basic physical modeling
+of acoustic wave transmission. \cite{HuangSymes:SEG15} call this is the {\em source-receiver
+  extension}.  This ``crosstalk'' ambiguity analyzed in the previous
+section afflicts all approaches based on the source-receiver
 extension.
 
-The analysis presented here has nothing to say about AWI applied to
+A number of authors have presented numerical illustrations of the
+divergence between source-receiver extension methods and traveltime
+tomography, and some have proposed remedies. \cite{Symes:94c}
+illustrated the crosstalk phenomenon for a version of MSWI. Crosswell
+source-receiver geometry often generates multiple arrivals,
+particularly near low-velocity sedimentary layers that are natural
+targets of seismic exploration and act as wave guides.
+\cite{Plessix:00} successfully applied a version of MSWI to crosswell
+waveform inversion by removing traces with close source and receiver
+depths, thus leaving only data that tended to be dominated by single
+arrivals. \cite{HuangSymes:Geo17} and \cite{SymesChenMinkoff:24}
+illustrate the failure of the version of MSWI reported here to avoid
+cycle-skipping in inversion of strongly refracting models, using
+frequency-domain implementation of the basic definition
+\ref{eqn:mswidefmain} and time-domain implementation of the reduced
+penalty function \ref{eqn:mswipenred}, respectively.
+\cite{Yongetal:GJI23} have shown that precisely the same issue arises
+in AWI, and have proposed a remedy by localizing the definition of the
+objective to short time intervals, within which multiple wavefront
+arrrivals are unlikely to appear. \cite{Symes:23} constructs an
+extension based on surface sources, as opposed to the point source -
+point receiver extension which is the basis of MSWI and AWI. This
+surface source extension separates events by infinitesimal slope,
+information not available to the source-receiver extension.
+
+Many open questions remain about MSWI, AWI, and other extension-based
+approaches to transmitted wave data inversion. In particular, 
+the analysis presented here has nothing to say about AWI applied to
 reflected wave data. \cite{Warner:16} and others have
 suggested that AWI is effective in that context. If so, the underlying
-mechanism remains to be explained.
+mechanism remains to be explained. The application of extended model
+inversion to of elastodynamic transmission data remains to be investigated.
 

### 2024-11-11 repro/papers/awi/paper.tex
final, goes to arxiv

@@
 \cite[]{HuangSymes2015SEG,HuangSymes:Geo17}. which were in turn based
-on ideas presented in Song's thesis \cite[]{Song:94c}. Since it
+on ideas presented in Song's thesis \cite[]{Song:94}. Since it
 involves no additional effort, I state results for MSWI in parallel
@@
   \begin{itemize}
-  \item[1. ] $Tu_{\sigma}[\kappa;\dl](\bx_r,\cdot;\bx_s) \in
+  \item[1. ] for $(\bx_s,\bx_r) \in X_{sr}$, $\log \kappa \in M$, and
+    $\sigma, \lambda >0$, $Tu_{\sigma}[\kappa;\dl](\bx_r,\cdot;\bx_s) \in
     L^2(\bR)$;
@@
 This paper has explained the behaviour of AWI (and MSWI) applied to
-acoustic transmission data. For data exhibiting a single smooth
-wavefront, these formulatios produce objective function that converge
+acoustic transmission through smoothly varying material models. For
+isotropic point source/receiver data exhibiting a single smooth
+wavefront, these formulations produce objective functions that converge
 to versions of travel time mean-square error, in the limit of hgh
-data frequency. For
-
-The conclusions of this study depend on a number of limitations on the
-acoustic model of wave propagation and the formulation of the related
-inverse problem. The principal constraints are:
-\begin{itemize}
-\item energy sources are modeled as isotropic point radiators, with
-  known time dependence vanishing outside a finite time window;
-\item mechanical parameter fields (coefficients in the acoustic wave
-  equation) are smooth, that is, infinitely differentiable as
-  functions of position, and the material density is known and
-  position-independent: only the bulk modulus (or wave velocity) is
-  to be determined;
-\item waves propagate throughout Euclidean space-time - that is, no
-  physical boundaries are modeled;
-\item the data of the inverse problem are the restrictions of the
-  pressure field to a finite number of points in space (receiver
-  positions), for a finite number of source positions and all time;
-\item local energy decays exponentially - that is, the pressure and
-  all of its derivatives tend to zero at an exponential rate as $t
-  \rightarrow \infty$,
-  uniformly in any bounded spatial domain.
-\item only one ray of geometric
-  acoustics connects each source-receiver location pair (single arrivals)
-\end{itemize}
-
-Some of these constraints
-could certainly be relaxed. For instance, exponential decay of local
-energy implies that the finite time duration of actual measured data
-can be treated as a perturbation of the infinite time data setting
-posed here. The conclusions of this paper likely hold in some form
-even for non-smooth sets of models, provided that the data for such models is
-dominated by ballistic transmitted waves. The exception is the single-arrival hypothesis: it is
+data frequency.
+
+%The conclusions of this study depend on a number of limitations on the
+%acoustic model of wave propagation and the formulation of the related
+%inverse problem. The principal constraints are:
+%\begin{itemize}
+%\item energy sources are modeled as isotropic point radiators, with
+%  known time dependence vanishing outside a finite time window;
+%\item mechanical parameter fields (coefficients in the acoustic wave
+%  equation) are smooth, that is, infinitely differentiable as
+%  functions of position, and the material density is known and
+%  position-independent: only the bulk modulus (or wave velocity) is
+%  to be determined;
+%\item waves propagate throughout Euclidean space-time - that is, no
+%  physical boundaries are modeled;
+%\item the data of the inverse problem are the restrictions of the
+%  pressure field to a finite number of points in space (receiver
+%  positions), for a finite number of source positions and all time;
+%\item local energy decays exponentially - that is, the pressure and
+%  all of its derivatives tend to zero at an exponential rate as $t
+%  \rightarrow \infty$,
+%  uniformly in any bounded spatial domain.
+%\item only one ray of geometric
+%  acoustics connects each source-receiver location pair (single arrivals)
+%\end{itemize}
+
+Some of the assumptions leading to this conclusion 
+could certainly be relaxed. For instance, neither absence of boundary effects
+nor exponential decay of local energy are essential. Instead
+the finite time duration of actual measured data, together with more
+elaborate ray geometry assumptions, can replace the infinite time data setting
+posed here,. The conclusions of this paper likely hold in some form
+even for non-smooth material parameter variation, provided that the data for such models is
+dominated by ballistic transmitted waves. Similar conclusions should
+hold for multipole point sources and
+receivers, modeling non-isotropic experimental devices that are small on the
+wavelength scale.
+
+The exception is the single-arrival hypothesis: it is
 essential, as without it, the link between the travel time and
 extended waveform inversion is lost.
-
-This disconnect is a fearure of the adaptive filter construction, not of a particular objective function
-(MSWI, AWI,...) formulated in terms of it. The introduction of the
-adaptive filter $u$ in the filtered data prediction $F[\kappa]*u$ is equivalent
-to permitting the source wavelet to depend on the receiver location,
-which in the basic physical model of point source acoustics it does
-not. Therefore this construction {\em extends} basic physical modeling
-of acoustic wave transmission. \cite{HuangSymes:SEG15} call this is the {\em source-receiver
-  extension}.  This ``crosstalk'' ambiguity analyzed in the previous
-section afflicts all approaches based on the source-receiver
-extension.
-
-A number of authors have presented numerical illustrations of the
-divergence between source-receiver extension methods and traveltime
-tomography, and some have proposed remedies. \cite{Symes:94c}
-illustrated the crosstalk phenomenon for a version of MSWI. Crosswell
-source-receiver geometry often generates multiple arrivals,
-particularly near low-velocity sedimentary layers that are natural
-targets of seismic exploration and act as wave guides.
-\cite{Plessix:00} successfully applied a version of MSWI to crosswell
-waveform inversion by removing traces with close source and receiver
-depths, thus leaving only data that tended to be dominated by single
-arrivals. \cite{HuangSymes:Geo17} and \cite{SymesChenMinkoff:24}
-illustrate the failure of the version of MSWI reported here to avoid
-cycle-skipping in inversion of strongly refracting models, using
-frequency-domain implementation of the basic definition
-\ref{eqn:mswidefmain} and time-domain implementation of the reduced
-penalty function \ref{eqn:mswipenred}, respectively.
-\cite{Yongetal:GJI23} have shown that precisely the same issue arises
-in AWI, and have proposed a remedy by localizing the definition of the
-objective to short time intervals, within which multiple wavefront
-arrrivals are unlikely to appear. \cite{Symes:23} constructs an
-extension based on surface sources, as opposed to the point source -
-point receiver extension which is the basis of MSWI and AWI. This
-surface source extension separates events by infinitesimal slope,
-information not available to the source-receiver extension.
-
-Many open questions remain about MSWI, AWI, and other extension-based
-approaches to transmitted wave data inversion. In particular, 
-the analysis presented here has nothing to say about AWI applied to
+%This disconnect is a fearure of the adaptive filter construction, not of a particular objective function
+%(MSWI, AWI,...) formulated in terms of it. The introduction of the
+%adaptive filter $u$ in the filtered data prediction $F[\kappa]*u$ is equivalent
+%to permitting the source wavelet to depend on the receiver location,
+%which in the basic physical model of point source acoustics it does
+%not. Therefore this construction {\em extends} basic physical modeling
+%of acoustic wave transmission. \cite{HuangSymes:SEG15} call this is the {\em source-receiver
+%  extension}.  This ``crosstalk'' ambiguity analyzed in the previous
+%section afflicts all approaches based on the source-receiver
+%extension.
+A number of authors have illustrated this disconnect via numerical
+example. The first such examples for a version of MSWI appeared in
+\cite{Symes:94c}. Crosswell seismic surveys often generate multiple
+arrivals near low-velocity sedimentary layers that are natural targets
+of seismic exploration and act as wave guides. \cite{Plessix:00}
+successfully applied a version of MSWI to crosswell waveform inversion
+by removing traces with close source and receiver depths, thus leaving
+only data that tended to be dominated by single arrivals.
+\cite{HuangSymes:Geo17} and \cite{SymesChenMinkoff:24} illustrate the
+failure of the version of MSWI reported here to avoid cycle-skipping
+in inversion of strongly refracting models, using frequency-domain
+implementation of the basic definition \ref{eqn:mswidefmain} and
+time-domain implementation of the reduced penalty function
+\ref{eqn:mswipenred}, respectively. \cite{Yongetal:GJI23} show that
+precisely the same shortcoming afflicts AWI.
+
+Overcoming this limitation requires substantial change in the
+extension of basic acoustic modeling underlying AWI. For example,
+\cite{Yongetal:GJI23} have proposed localizing the definition of the
+AWI objective to short time intervals, within which multiple wavefront
+arrrivals are unlikely to appear, and illustrated the effectiveness of
+this approach by numerical example. \cite{Symes:23} constructs an
+extension based on surface sources, as opposed to the trace-by-trace
+extension which is the basis of MSWI and AWI. This surface source
+extension separates events by infinitesimal slope, information not
+available in the trace-by-trace construction..
+
+The analysis presented here has nothing to say about AWI applied to
 reflected wave data. \cite{Warner:16} and others have
 suggested that AWI is effective in that context. If so, the underlying
-mechanism remains to be explained. The application of extended model
-inversion to of elastodynamic transmission data remains to be investigated.
+mechanism remains to be explained.
+
+Linear elasticity is a much better model for mechanical waves
+propagating in some human tissue (notably bone) and the solid earth
+than is acoustodynamics. Least-squares FWI of elastodynamic
+transmission data also tends to stagnate due to cycle-skipping, and it
+is tempting to think that extension approaches such as AWI should
+ameliorate that obstacle. It will be necessary extract wave speed
+informaiton in the presence of potential interference of cross-talk
+between wave modes, similar to cross-talk between multiple arrivals in
+the acoustic case.
+
+Finally, virtually all materials supporting wave propagation also
+attenuate energy through various physical mechanisms, which act per
+cycle and therefore suppress high frequency components. Analysis such
+as that presented here, relying on the infinite-frequency limit
+(geometric optics), is {\em a priori} impossible for such attenuative
+wave physics without fundamental
+reformulation..
 

### 2024-11-24 repro/papers/awi/paper.tex
minor

@@
 wave propagation, and numerical solutions of these wave equations can
-be used to predict experimental data {\em ab initio}. FWI consists in
+be used to predict experimental data. FWI consists in
 updating the parameters to minimize a numerical measure of  the difference between computationally
@@
 for their description, defines the AWI and MSWI objective functions,
-and states the main results of this paper. in the following section, I present the
+and states the main results of this paper. In the following section, I present the
 asymptotic analysis leading to the main results. In
@@
 The analysis explained here sheds no light on the effectiveness of AWI
-for reflected wave data, and the approach has so far been limited
+for reflected wave data, and the approach has so far been limited to
 inversion of data modeled by acoustic wave propagation. I discuss
@@
                                     f(t) \delta({\bf x}-{\bf x}_s); \nonumber \\
-\frac{\partial {\bf v}}{\partial t} & = & - \beta \nabla p; \\ 
+\frac{\partial {\bf v}}{\partial t} & = & - \beta \nabla p, \nonumber \\ 
 p, {\bf v} & = & 0 \mbox{ for }  t \ll 0.
@@
 
-The second and third assumptions explicitly involves a constraint on ray geometry
+The second and third assumptions explicitly involve constraints on ray geometry
 (therefore on $\kappa$). A {\em normal neighborhood} of
@@
 \section{Analysis of AWI and MSWI}
-THis section is devoted to the proofs of the main result, Theorem
+This section is devoted to the proofs of the main result, Theorem
 \ref{thm:main}.  The principal tool in the analysis is Hadamard's
@@
 its derivatives over any finite $t$-interval. However, with the
-addition of the Exponetial Decay assumption, the bound becomes
+addition of the Exponential Decay assumption, the bound becomes
 global in time: for any $k \ge 0$, there exists $B \ge 0,\delta > 0$
@@
 \end{equation}
-The  ``mother wavelet'' $w_1$ is assumed to have properties W1, W2, and
+The  ``mother wavelet'' $w_1$ is assumed to have zero mean and
 all wavelets $\wl$ in this family
-inherit these properties from $w_1$, along with its dimensions
-(energy/time$^2$). It follows from these assumptions that the Fourier transform $\fwl$ is
-real-valued, even, and real-analytic. The corresponding source waveform $f_{\lambda}$ is
+inherit this property.
+It follows from these assumptions that the Fourier transform $\fwl$ is
+even and real-analytic. The corresponding source waveform $f_{\lambda}$ is
 given by
@@
 \[
-  \|d\| \le 2\|\Fl[\kappa]*\ul[\kappa;d] \le
+  \|d\| \le 2\|\Fl[\kappa]*\ul[\kappa;d] \| \le
   2\lambda^{\frac{1}{2}}\|\ul[\kappa;d]\|
@@
 \ref{eqn:awidefmain} of $J_{\rm AWI,\sigma}$, set $\sigma = r\lambda$
-(equation \ref{eqn:lamsig}
+(equation \ref{eqn:lamsig})
 and replace the data $d$ with
@@
   \|T ((\Sla)^T\Slb + (\Slb)^T\Sla + (\Slb)^T\Slb)\ul\| \le
-  C\lambda^{3}\|\ul\| \le C\lambda^{\frac{5}{2}} \|dl\|
+  C\lambda^{3}\|\ul\| \le C\lambda^{\frac{5}{2}} \|\dl\|
 \]

### 2025-02-16 repro/papers/ssefound/paper.tex
interim

\title{Mapping Properties of Surface Source Extension}
\author{William W. Symes}

\begin{abstract}
  Abstract goes here.
\end{abstract}

\section{Symmetric Hyperbolic Systems}
$k \times  k$ symmetric hyperbolic system: in $\bR^{n+1}$. Suppose that $u \in L^2_{\rm loc}(\bR^{n+1},\bR^k)$ is a weak solution of
\begin{equation}
  \label{eqn:shs}
  (\sum_{i=0}^n A_i\partial_i ) u \in L^2_{\rm loc}(\bR^{n+1},\bR^k)
\end{equation}
The coefficieent matrices $A_i \in C^{\infty}(\bR^n, \bR^{k \times k}), i=0,...,n$ are symmetric.

If $A_j$ is positive definite, then $x_j$ can play the role of time, and the Cauchy problem with initial data $u_0 \in L^2(\bR^n,\bR^k)$  on $x_j=$ constant has a unique weak solution $u \in L^2(\bR^{n+1})^k$ with well-defined $L^2$ traces on any $x_j$ level set (Lax, Taylor). Want to understand traces on other hypersurfaces, to begin with coordinate hypersurfaces, wlog $x_0=0$.

Notation:$x_0=t$, $\partial_t=\partial_0$, $\bx = (x_1,...,x_n)$, $D=(\partial_1,...\partial_n)$
\[
  K(t,\bx,D)= \sum_{i=1}^n A_i(t,\bx)\partial_i, \, M(t,\bx) = A_0(t,\bx).
\]
If $M$ is positive definite, trace properties coverd by standard theory (above). Let $\Pi_0(t,\bx)$ = projection on null space $V_0$ of $M(t,\bx)$, $\Pi_1=I-\Pi_0$ = projection on range $V_1$. Denote the rank of $M$ by $l={\rm dim} V_1$ Then
\[
  M\partial_t u + Ku = \Pi_1 M \partial_t \Pi_1 u + \sum_{\alpha,\beta=0,1} \Pi_{\alpha}K\pi_{\beta}u + \mbox{ l.o.t. }
\]
Since the ranges are orthogonal,
\begin{eqnarray}
  \Pi_1 M \partial_t \Pi_1 u + \Pi_1 K \Pi_1 u + \Pi_1 K \Pi_0 u & \in & L^2_{\rm loc}(\bR^{n+1},V_1),
  \label{eqn:decomp1} \\
  \Pi_0 K \Pi_0 u + \Pi_0 K \Pi_1 u & \in & L^2_{\rm loc}(\bR^{n+1},V_0)
  \label{eqn:decomp0}
\end{eqnarray}
Set $K_{\alpha,\beta} = \Pi_{\alpha}K\Pi_{\beta}$.

Main assumption: for all $t \in \bR, (\bx,\xi) \in T^*(\bR^n)\setminus \{0\}$, $K_{0,0}(t,\bx,\xi)$ is invertible on $V_0$.

% Suppose that $E \in OPS^1(\bR^{n+1})$ is scalar, $Eu \in H^1_{\rm loc}(\bR^{n+1})$. Denote by $i(t):\{t\} \times \bR^n \rightarrow \bR^{n+1}$ the embedding. contained in $i(0)^*(ES(E))$, and that and

% Suppose that $\Gamma_0, \Gamma \subset T^*(\bR^n)$ are conic open sets with $\overbar{\Gamma_0 \subset \Gamma$ and $\delta t >0$. Assume that  that for $|t|\le \delta t$ and $(\bx,\xi) \in Gamma$, $K_{00}(0,\bx,\xi)$ is invertible as a map $V_0 \rightarrow V_0$.
Let  $L_{00}(t,\bx,\xi) \in C^{\infty}(\bR, S^{-1}(\bR^n))$ be inverse to $K_{0,0}(t,\bx,\xi)$ on $V_0$: that is,
\[
  K_{0,0}(t,\bx,\xi)L_{0,0}(t,\bx,\xi) v=v=L_{0,0}(t,\bx,\xi)K_{0,0}(t,\bx,\xi)v
\]
for all $v \in V_0$. Construct $L_{0,0}$ to be homogeneous of degree -1 outside of a neighborhood of $\xi=0$, with $V_1$ an invariant subspace for $L_{0,0}(t,\bx,\xi)$ on which it vanishes. It follows that $L_{0,0}$ is symmetric, and that
\[
  \Pi_0 u - L_{0,0}K_{0,0}\Pi_u = \Pi_0 u - L_{0,0}K_{0,0}u = \Pi_0 u + L_{0,0}K_{0,1}u \in H^{-1}_{\rm loc}(\bR^{n+1})^k
\]
so
\[
  K_{1,0}u + K_{1,0}L_{0,0}K_{0,1}u \in L^2 _{\rm loc}(\bR^{n+1})^k
\]
Define
\[
  \tilde{K} = K_{11} - K_{10}L_{00}K_{01}.
\]
Then \ref{eqn:decomp1} can be rewritten
\begin{equation}
  \label{eqn:shs1}
  M\partial_t u_1 + \tilde{K}u_1 \in L^2_{\rm loc}(\bR^{n+1},V_1)
\end{equation}
Note that for each $t \in \bR, (\bx,\xi) \in T^*(\bR^n)$, both $V_0$ and $V_1$ are invariant subspaces of $\tilde{K}(t,\bx,\xi)$, and $V_0 \subset {\rm ker}(\tilde{K}(t,\bx,\xi)$. Also, $M$ is invertible when restricted to $V_1 = {\rm Rng} M = {\rm Rng } \Pi_1$ , but not (necessarily) positive definite.

Diagonalization: for $(t,\bx,\tau,\xi) \in T^*(\bR^n)$, define the restricted characteristic polynomial
\[
  {\cal C}(t,\bx,\tau,\xi) = \det (M(t,\bx)\tau + \tilde{K}(t,\bx,\xi) )= 0
\]
Since $V_0$ and $V_1$ are invariant subspaces of the matrix in parentheses, and $V_0$ is (contained in) its null space, there are at most $l={\rm dim}V_1$ non-zero roots in $\tau$ for $\xi \ne 0$.

Choose a conic submanifold $\gamma \subset T^*(\bR^n)$, so that for $|t| \le \delta t, (\bx,\xi) \in \gamma$ there are exactly $m \le l$ distinct real roots $\tau$ of ${\cal C}(t,\bx,\tau,\xi)$. Denote these by $\tau_i(t,\bx,\xi), i=1,...,m$.

Choose a conic $\gamma_1 \subset T^*(\bR^n)$ so that $\bar{\gamma_1} \subset \gamma$, and $Q \in OPS^0(\bR^n)$ with $ES(Q) \subset \gamma$, $Q(\bx,\xi) =1$ for $(\bx,\xi)\in \gamma_1$.

Let ${\cal N} = $ normal bundle of $x_0=0$ and $\Gamma \subset T^*(\bR^n)$ for which $\Gamma \cap {\cal N} = \emptyset$.

Assume that $u \in L^2(\bR^{n+1}) \cap H^1_{\rm loc}(\Gamma \cap \Pi_{\cal N}^{-1}(\gamma_1^C))$.

For each $\tau_i, i =1,...m$, choose $U_i(t,\bx,\xi) \in C^{\infty}(\bR \times T^*(\bR^n),V_1(t,\bx))$,  so that
\[
  \tau_i M(\bx) U_i(t,\bx,\xi) + \tilde{K}(t,\bx,\xi) U_i(t,\bx,\xi)=0, \, i=1,..,m.
\]
$U_1,...,U_m$ span a subspace $\tilde{V}_1(t,\bx,\xi)$. Scale $U_i$ so that $|U_i(t,\bx,\xi)| = |\xi|$.

Note that $\tau_i(t,\bx,\xi), i=1,...,m$ are homogeneous of order 1 in $\xi$ and smooth in $\bx$, $U_i(t,\bx,\xi), i=1,...,m$ are homogeneous of order 0 in $\xi$, smooth in $\bx$. So both are symbol-valued functions, of orders 1 and 0 respectively. Since the $U$'s must be orthogonal, $U_i^TU_j = |\xi|^2\delta_{ij}$.

Suppose first that $m=l$, that is, that $\tilde{V}_1=V_1$. Then $\{U_i(t,\bx,\xi):i=1,..,l\}$ is a basis for $V_1(t,\bx)$ for each $(t,\bx,\xi \in \Gamma$. Set $u_{1,i}(t,\bx) = U_i(t,\bx,D_x)^Tu_1$. Then for any $w \in L^2(\bR^{n+1},V_1)$
\[
  \sum_{i=1}^l U_i(t,\bx,D)U_i(t,\bx,D)^Tw(t,\bx) \approx
\]
\[
  \int \, d\xi \, e^{i \bx \cdot \xi} U_i(t,\bx,\xi)U_i(t,\bx,\xi)^T\hat{w}(t,\xi) \approx
\]
\[
  \int \, d\xi \, e^{i \bx \cdot \xi}
\]
\[
  \tilde{K}(t,\bx,D)u_{1,i} = \tilde{K}(t,\bx,D)U_i(t,\bx,D)^T
\]
\[
  M(\bx)\partial_t u_{1,i}(t,\bx) = -\tilde{K}(t,\bx,D)
\]

\section{2D Acoustics}

\begin{eqnarray}
  \frac{1}{\kappa}\partial_t p + \nabla \cdot \bv & = & 0, \nonumber \\
  \rho \partial_t \bv + \nabla p & = & 0.
  \label{eqn:awe}                                     
\end{eqnarray}

Set $u=(p,v_x,v_z), \bx=(t,x), \xi=(\omega,\xi_x), z \mapsto t$. Then \ref{eqn:awe} takes the form \ref{eqn:shs} with
\[
  M =
  \left(
    \begin{array}{ccc}
      0 & 0 & 1\\
      0 & 0 & 0\\
      1 & 0 & 0
    \end{array}
  \right),
  A_1 ={\rm diag}\left(\frac{1}{\kappa},\rho,\rho\right),
  A_2 =
 \left(
    \begin{array}{ccc}
      0 & 1 & 0\\
      1 & 0 & 0\\
      0 & 0 & 0
    \end{array}
  \right),
\]
So $V_0={\rm span}(\{e2\}),\, V_1={\rm span}(\{e1,e3\}), \Pi_0 = {\rm diag}(0,1,0)$ 
\[
  K(t,\bx,\xi) =  A_1(\bx) \xi_1 + A_2(\bx) \xi_2 =
  \left(
    \begin{array}{ccc}
      \frac{1}{\kappa(\bx)}\omega & \xi_x & 0 \\
      \xi_x & \rho(\bx) \omega & 0 \\
      0 & 0 & \rho(\bx) \omega
    \end{array}
  \right)
\]
whence $K_{0,0} = {\rm diag}(0, \rho(\bx) \omega, 0)$.
\[
  K_{1,1}(t,\bx,\xi) =
  \left(
    \begin{array}{ccc}
      \frac{1}{\kappa(\bx)}\omega & 0 & 0 \\
      0 & 0 & 0 \\
      0 & 0 & \rho(\bx) \omega
    \end{array}
  \right),
\]
\[
  K_{0,1}(t,\bx,\xi) =
  \left(
    \begin{array}{ccc}
      0 & \xi_x & 0 \\
      0 & 0 & 0 \\
      0 & 0 & 0
    \end{array}
  \right)
\]
and $K_{1,0} = K_{0,1}^T$.

Choose $\omega_0>0$, $\phi \in C^{\infty}(\bR)$ so that $\phi(\omega) = 1$ for $\omega> 2\omega_0$, $=0$ for $\omega < \omega_0$. Set
\[
  L_{0,0}= {\rm diag}\left(0, \frac{\phi(|\omega|)}{\rho(\bx) \omega}, 0\right)
\]
Then $L_{0,0}K_{0,0} = I$ restricted to $V_0$ and away from zero frequency, and
\[
  \tilde{K} = K_{1,1} - K_{1,0}L_{0,0}K_{0,1} =
  \left(
    \begin{array}{ccc}
      \frac{\omega}{\kappa} - \frac{\xi_x^2\phi(|\omega|)}{\rho \omega} & 0 & 0\\
      0 & 0  & 0 \\
      0 & 0 & \rho \omega
    \end{array}
  \right).
\]
Since $V_1={\rm span}(e_1,e_3)$, the determinant of the restricted characteristic polynomial is obtained by dropping the second row and column:
\[
  \det
  \left(
    \begin{array}{cc}
      \frac{\omega}{\kappa} - \frac{\xi_x^2\phi(|\omega|)}{\rho \omega} &  \xi_z\\
      \xi_z &  \rho \omega
    \end{array}
  \right).
\]
\[
  = \frac{\rho \omega^2 }{\kappa} - \xi_x^2\phi(|\omega|) - \xi_z^2
\]
\[
  = \frac{\rho \omega^2 }{\kappa} - \xi_x^2 - \xi_z^2 \mbox{ mod }  S^{-\infty}
\]  
as expected. So a maximal choice for $\Gamma$ would be
\[
  \Gamma = \{(z,(t,x),(\omega,\xi_x)): c(z,x)|\xi_x| < |\omega|\}.
\]
\section{2D Elasticity}

Velocity, stress, state vector:
\[
  \bv = \left(
    \begin{array}{c}
      v_1 \\
      v_2
    \end{array}
  \right), \,
  \sigma = \left(
    \begin{array}{c}
      \sigma_{11} \\
      \sigma_{12} \\
      \sigma_{21} \\
      \sigma_{22}
    \end{array}
  \right), \, 
  u = \left(
    \begin{array}{c}
      \sigma\\
      \bv
    \end{array}
  \right)
\]
Symmetry of stress: $\sigma_{12}=\sigma_{21}$. Compliance tensor (inverse of Hooke tensor):
\[
  C =\left(
    \begin{array}{cccc}
      c_{1111} &  c_{1112} &  c_{1121} &  c_{1122} \\
      c_{1211} &  c_{1212} &  c_{1221} &  c_{1222} \\
      c_{2111} &  c_{2112} &  c_{2121} &  c_{2122} \\
      c_{2211} &  c_{2212} &  c_{2221} &  c_{2222} \\
    \end{array}
  \right)
\]
Symmetries: $C_{ijkl}=C_{klij}, C_{ijkl}=C_{jikl}=C_{ijlk}$. Density = $\rho$

Elastodynamics:
\begin{eqnarray}
  \label{eqn:const}
  C \partial_ t \sigma & = & \frac{1}{2}(\nabla \bv + \nabla \bv^T) \\
  \label{eqn:mombal}
  \rho \partial_t \bv & = & \nabla \cdot \sigma
\end{eqnarray}

Matrix representation: set
\[
  B_1=\left(
    \begin{array}{cccc}
      1 & 0 & 0 & 0 \\
      0 & \frac{1}{2} & \frac{1}{2} & 0
    \end{array}
  \right),\,
  B_2=\left(
    \begin{array}{cccc}
      0 & \frac{1}{2} & \frac{1}{2} & 0 \\
      0 & 0 & 0 & 1
    \end{array}
  \right)
\]
and
\[
  A_0 = \left(
    \begin{array}{cc}
      C & 0 \\
      0 & \rho I
    \end{array}
  \right), \,
  A_j = -\left(
    \begin{array}{cc}
      0 & B_j^T \\
      B_j & 0
    \end{array}
  \right), \, j=1,2
\]

Then the elastodynamics equations \ref{eqn:const}, \ref{eqn:mombal} are equivalent to
\[
  (A_0\partial_0 + A_1 \partial_1 + A_2 \partial_2)u =0.
\]

$u = (\sigma, \bv)^T \in \bR^6$ is a null vector for $A_2$ iff $B_2\bv = 0$ and $B_2^T \sigma=$, that is, $\bv=0$ and $\sigma_{12}=\sigma_{21}=\sigma_{22} = 0$. Thus $U_0= \mbox{ker }A_2 = \mbox{span } \{\be_1\}$ and  $U_1 = \mbox{rng }A_2 = \mbox{span }\{\be_j: 2\le j \le 6\}$, and $\Pi_0 = \mbox{diag}(1,0,0,0,0,0)$, $\Pi_1 = I-\Pi_0$.

Eliminating zero rows and columns, equation \ref{eqn:shsnull1} for $u_1 = (\sigma_{12}, \sigma_{21}, \sigma_{22}, v_1, v_2)^T$ reads
\begin{eqnarray}
  \frac{1}{2}\partial_2v_1 &=& -((C \partial_0 \sigma)_{21} + \frac{1}{2}\partial_1v_2)\\
  \partial_2 v_2 &=& -(C \partial_0 \sigma)_{41}\\
  \partial_2 \sigma_{12} &=& -(\rho \partial_0v_1)
\end{eqnarray}

Thus
\begin{eqnarray}
  A_{000} &=& \Pi_0 A_0 \Pi_0 = \mbox{ diag}(c_{1111},0,0,0,0,0) \nonumber \\
  A_{100} &=& \Pi_0 A_1 \Pi_0 = 0 \nonumber \\
  A_{001} &=& \Pi_0 A_0 \Pi_1 =
              \left(
              \begin{array}{cccccc}
                0 & c_{1112} & c_{1121} & c_{1122} & 0 & 0 \\
                0 & 0 & 0 & 0 & 0 & 0 \\
                0 & 0 & 0 & 0 & 0 & 0 \\                
                0 & 0 & 0 & 0 & 0 & 0 \\
                0 & 0 & 0 & 0 & 0 & 0 \\
                0 & 0 & 0 & 0 & 0 & 0
              \end{array}
                                    \right) \nonumber \\
  A_{101} &=& \Pi_0 A_1 \Pi_1 =
                            \left(
              \begin{array}{cccccc}
                0 & 0 & 0 & 0 & 1 & 0 \\
                0 & 0 & 0 & 0 & 0 & 0 \\
                0 & 0 & 0 & 0 & 0 & 0 \\                
                0 & 0 & 0 & 0 & 0 & 0 \\
                0 & 0 & 0 & 0 & 0 & 0 \\
                0 & 0 & 0 & 0 & 0 & 0
              \end{array}
                                    \right)
                                    \label{eqn:ela0}
\end{eqnarray}
So equation \ref{eqn:rel01} reduces to
\[
  c_{1111}\sigma_{11} = -(c_{1112}\sigma_{12} + c_{1121}\sigma_21 + c_{1122}\sigma_{22} + \partial_1 \int_{-\infty}^t v_1)
\]
This is just the first row of equation \ref{eqn:const}, integrated in time! Now apply $\partial_2$ to both sides to obtain the missing equation.

\bibliographystyle{seg}
\bibliography{../../bib/masterref}

### 2025-02-19 repro/papers/ssefound/paper.tex
interim

@@
 
+\append{G\r{a}rding Interlude}
+
+In the following $U$ is an open set in $\bR^n$.
+
+\begin{definition}
+  A $k \times k$ valued symbol $M(\bx,\xi) \in OPS^m(U)$ is {\em elliptic} iff $|\xi|^{-m}M(\bx,\xi)$ is uniformly invertible for $(\bx,\xi) \in T^*(U) \setminus \{0\}$.
+That is, there exists $C>0$ so that for all  $(\bx,\xi) \in T^*(U) \setminus \{0\}$, $v \in \bR^k$,
+  \[
+    |M(\bx,\xi)v| \ge C|\xi|^m|v|.
+  \]
+\end{definition}
+
+Connection to defintion V.1.1 in \cite{Tay:81}: concerns the first order system
+\[
+  (\frac{\partial}{\partial y} - K(y,x,D_x))u = 0
+\]
+where $u=u(y,x) \in \bR^k$, $K(y,\cdot,\cdot) \in C^{\infty}(\bR,OPS^1(U))$.
+According to the cited definition, this system is elliptic iff the principal symbol $K_1(y,x,\xi)$ has no purely imaginary eigenvalues.
+The symbol of the operator on the LHS above is $i\eta I - K(y,x,\xi)$, which is not the symbol of a $\Psi$DO unless $K$ is differential.
+However $K$ having no purely imaginary eigenvalues iz equivalent to  $i\eta I - K(y,x,\xi)$ being invertible.
+
+\begin{lemma} \label{thm:lema1}
+``Interpolation'' inequality: for any $m, N \in {\bf Z}_+, 0 < \epsilon < 1$,
+\[
+  \lxr^{\frac{m}{2}-1} \le \epsilon^2 \lxr^{\frac{m}{2}} + C^2_{\epsilon,N,m}\lxr^{-N}. 
+\]
+\end{lemma}
+
+\begin{proof}
+ Equivalent to
+\[
+  \lxr^{-1} \le \epsilon^2 + C^2_{\epsilon,N,m}\lxr^{-N-\frac{m}{2}}. 
+\]
+Holds trivially if $\lxr \ge \epsilon^{-2}$. Otherwise $\lxr^{N+\frac{m}{2}-1} \le \epsilon^{-2(N+\frac{m}{2}-1)} \equiv C_{\epsilon,N,m}$.
+\end{proof}
+
+\begin{proposition}
+\label{thm:propa1}(clone of V.6.2 in \cite{Tay:81}): Suppose that $p \in OPS^0(U,\bR^k)$ has a SPD principal symbol $p_0$: there is $C > 0$ so that for sufficiently large $|\xi|$,
+  \[
+    p_0(x,\xi) = p_0(x,\xi)^T, p_0(x,\xi) \ge C
+  \]
+  Then there existss $B \in OPS^0(U,\bR^k)$ so that $P = B^*B \mbox{ mod } OPS^{-\infty}$.
+\end{proposition}
+
+\begin{proof}
+  From the spectral theorem, for each $(x,\xi) \in T^*(U)$, can construct SPD $b_0(x,\xi)$ for which $p_0(x,\xi) = b_0(x,\xi)^2$.
+Suppose that we have built $b_0,...,b_{j-1}$ with $b_i \in S^{-i}$ homogeneous of degree $-i$ for large $|\xi|$ so that
+  $r_{j}(x,D) = p(x,D) - (b_0(x,D) +...+ b_{j-1}(x,D))^2 \in OPS^{-j}$. Want $b_j \in S^{-j}$ homogeneous of degree $j$ so that
+  \[
+    p(x,D)-(b_0 +...+ b_{j-1}+b_j)^2 = r_{j-1} -  (b_0 +...+ b_{j-1})b_j - b_j((b_0 +...+ b_{j-1}) - b_j^2
+  \]
+  \[
+    = r_{j-1} - (b_0 b_j+b_j b_0) \mbox{ mod } OPS^{-j-1}
+  \]
+  is in $OPS^{j-1}$. This will be the case if the principal symbols match: denoting the principal symbol of $r_{j-1}$ by $r_{j-1,0}$, this requires that
+  \[
+    r_{j-1,0}(x,\xi) = b_0(x,\xi)b_j(x,\xi) + b_j(x,\xi)b_0(x,\xi).
+  \]
+  Since $b_0$ is SPD, the map $a \mapsto b_0a+ab_0$ on $k \times k$ symmetric matrices $a$ is invertible, as its eigenvalues are sums of the eigenvalues of $b_0$.
+So the above equation has a unique symmetric solution, and the inductive step in the construction of $b \in OPS^0$ is finished.
+\end{proof}
+
+\begin{theorem}
+  \label{thm:garding}
+  ``G\r{a}rding for systems'': Suppose that $P \in OPS^m(U,\bR^k)$ is elliptic, that is, its principal symbol $p_0$ satisfies 
+\[
+  p_0^*(x,\xi)p_0(x,\xi) \ge C|\xi|^{2m}, \, (x,\xi) \in T^*(U).
+\]
+Then for any $K \subset \subset U$, $N \ge 1$ there exists $C_{K,N} > 0$ so that for $u \in C^{\infty}_0(K)$, 
+\[
+  \|Pu\|^2_0 \ge \frac{C}{2}\|u\|^2_m - C_{K,N}\|u\|^2_{-N}.
+\]
+\end{theorem}
+
+\begin{proof}
+  For any $alpha \in [0,1)$, $\ldr^{-m}P^*P\ldr^{-m} - \frac{C}{2} \in OPS^0$ is self-adjoint  and its principal symbol $\lxr^{-,m}p_0^Tp_0\lxr^{-m} - \frac{C}{2} \in S^0$ is positive semi-definite.
+Therefore from Proposition \ref{thm:propa1} there exists $B \in OPS^0$ for which
+  \[
+    \ldr^{-m}P^*P\ldr^{-1} - \frac{C}{2} = B^*B \mbox{ mod } OPS^{-\infty}
+  \]
+  or
+  \[
+    P^*P -\frac{C}{2} \ldr^{2m} = \ldr B^*B \ldr \mbox{ mod } OPS^{-\infty}
+  \]
+  So for every $N  \in {\bf Z}_+$ there is $C_{K,N} > 0$ so that for $u \in C_0^{\infty}(K)$,  
+  \[
+    \langle u,P^*P u \rangle \ge \frac{C}{2}\|u\|_m^2 -   C_{K,N}\|u\|^2_{-N}
+  \]
+
+\end{proof}
+
+\begin{definition} $P \in OPS^m(U,\bR^k)$ is (microlocally) elliptic in the closed conic set $\Gamma \subset T^*(U)$ iff for some $C>0$ its principal symbol $p_0$ satisfies
+  \[
+    p(x,\xi)^*p(x,\xi) \ge C|\xi|^m
+  \]
+\end{definition}
+
+\begin{theorem} Suppose that $P \in OPS^m(U,\bR^k)$ is elliptic in the closed conic $\Gamma \subset T^*(U)$.
+Then for any $Q \in OPS^m(U,\bR)$ with $ES(Q) \subset \Gamma$, $N \in {\bf Z}_+$, and compact $K \subset \subset U$, there exists $C_Q > 0$ and $C_{Q,K,N} \ge 0$ so that for any $u \in C_0^{\infty}(K)$,
+  \[
+    \|P(x,D)u\|^2_0  \ge C_Q\|Q(x,D)u\|_0^2 - C_{Q,K,N}\|u\|^2_{-N}.
+  \]
+\end{theorem}
+
 \bibliographystyle{seg}

### 2025-03-13 repro/papers/ssefound/paper.tex
interim

@@
 \begin{proof}
-  For any $alpha \in [0,1)$, $\ldr^{-m}P^*P\ldr^{-m} - \frac{C}{2} \in OPS^0$ is self-adjoint  and its principal symbol $\lxr^{-,m}p_0^Tp_0\lxr^{-m} - \frac{C}{2} \in S^0$ is positive semi-definite.
+$\ldr^{-m}P^*P\ldr^{-m} - \frac{C}{2} \in OPS^0$ is self-adjoint  and its principal symbol $\lxr^{-m}p_0^Tp_0\lxr^{-m} - \frac{C}{2} \in S^0$ is positive semi-definite.
 Therefore from Proposition \ref{thm:propa1} there exists $B \in OPS^0$ for which
   \[
-    \ldr^{-m}P^*P\ldr^{-1} - \frac{C}{2} = B^*B \mbox{ mod } OPS^{-\infty}
+    \ldr^{-m}P^*P\ldr^{-m} - \frac{C}{2} = B^*B \mbox{ mod } OPS^{-\infty}
   \]
@@
   \[
-    P^*P -\frac{C}{2} \ldr^{2m} = \ldr B^*B \ldr \mbox{ mod } OPS^{-\infty}
+    P^*P -\frac{C}{2} \ldr^{2m} = \ldr^m B^*B \ldr^m \mbox{ mod } OPS^{-\infty}
   \]
@@
 
-\begin{definition} $P \in OPS^m(U,\bR^k)$ is (microlocally) elliptic in the closed conic set $\Gamma \subset T^*(U)$ iff for some $C>0$ its principal symbol $p_0$ satisfies
+\begin{definition} $P \in OPS^m(U,\bR^k)$ is (microlocally) elliptic in the open conic set $\Gamma \subset T^*(U)$ iff for some $C>0$ and conic neighborhood $\Gamma_1 \supset \bar{\Gamma}$, its principal symbol $p_0$ satisfies
   \[
-    p(x,\xi)^*p(x,\xi) \ge C|\xi|^m
+    p(x,\xi)^*p(x,\xi) \ge C|\xi|^m, \, \xi \in \Gamma_1, \, |\xi| \ge 1.
   \]
@@
 
-\begin{theorem} Suppose that $P \in OPS^m(U,\bR^k)$ is elliptic in the closed conic $\Gamma \subset T^*(U)$.
-Then for any $Q \in OPS^m(U,\bR)$ with $ES(Q) \subset \Gamma$, $N \in {\bf Z}_+$, and compact $K \subset \subset U$, there exists $C_Q > 0$ and $C_{Q,K,N} \ge 0$ so that for any $u \in C_0^{\infty}(K)$,
+Some results from \cite{BaoSy:91b}.
+Here $0 \le m \le n$, ${\cal N}$ is the normal bundle of the foliation $\bR^n = \bR^{n-m} \times \bR^m$, typically $x \in \bR^{n-m}$ is the parameter, $y \in \bR^m$ is the trace coordinate.
+
+\begin{lemma}
+\label{thm:prebao}
+Suppose that $k \in {\bf Z}_+$, $n > 0$. Then there exists $C_{k} \ge 0$ so that for $\xi, \eta \in \bR^n$,
+\begin{equation}
+  \label{eqn:pre}
+  (1 + |\xi + \eta|)^{k} \le (1 + |\xi|)^{k} (1 + |\eta|)^{k}
+\end{equation}
+\end{lemma}
+\begin{proof}
+  Obvious for $k = 0$. If equation \ref{eqn:pre} holds for $k \in {\bf Z}_+$, then
   \[
-    \|P(x,D)u\|^2_0  \ge C_Q\|Q(x,D)u\|_0^2 - C_{Q,K,N}\|u\|^2_{-N}.
+    (1 + |\xi + \eta|)^{k+1} \le (1 + |\xi)^{k}(1+ \eta|)^{k}(1+|\xi| + |\eta|)
   \]
-\end{theorem}
+  Since $1+|\xi| + |\eta| \le (1+|\xi|)(1 + |\eta|)$, the induction step is done.
+\end{proof}
+
+\begin{proposition}
+  \label{thm:bao1}
+  Suppose that $B \in \Ci(\bR^{n-m},OPS^0(\bR^m)), H \in OPS^0(\bR^n)$, $H$ is a convolution operator, $\phi \in \Ciz(\bR^n)$,
+  \[
+    ES(H) \cap {\cal N} = \emptyset.
+  \]
+  Then
+  \[
+    B\phi H \in OPS^0(\bR^n)
+  \]
+  and
+  \[
+    ES(B\phi H) \subset \Pi_2^{-1}(ES(B)) \cap ES(H)
+  \]
+\end{proposition}
+
+\begin{proof}
+  For $u \in \Ciz(\bR^n)$,
+  \[
+    {\cal F}_x(Hu)(\xi',y) = \int  d\eta H(\xi',\eta')\hat{u}(\xi',\eta)e^{ i y \eta}
+  \]
+  So
+  \[
+    (P\phi H)u(x,y) = \int d \xi P(x,y,\xi) {\cal F}_x(\phi Hu)(\xi,y)e^{i x\xi}
+  \]
+  \[
+    =\int d \xi' \int d \xi P(x,y,\xi) {\cal F}_x\phi(\xi-\xi',y) {\cal F}_x(Hu)(\xi',y)e^{i x\xi}
+  \]
+  \[
+   =\int d \xi' \int d \xi \int d\eta P(x,y,\xi) {\cal F}_x\phi(\xi-\xi',y) H(\xi',\eta)\hat{u}(\xi',\eta)e^{i x\xi + i y\eta }
+  \]
+  \[
+    =\int d \xi' \int d \xi \int d\eta P(x,y,\xi) {\cal F}_x\phi(\xi-\xi',y) e^{ix(\xi-\xi')}H(\xi',\eta)\hat{u}(\xi',\eta)e^{i x\xi' + i y\eta }
+  \]
+  So the symbol of $Q = P \phi H$ is
+  \[
+    Q(x,y,\xi,\eta) = \int d\xi' P(x,y,\xi') {\cal F}_x\phi(\xi'-\xi,y) e^{i x(\xi'-\xi)}H(\xi,\eta)
+  \]
+  \[
+    = \left(\int d\xi' P(x,y,\xi+\xi') {\cal F}_x\phi(\xi',y) e^{i x\xi'}\right) H(\xi,\eta)
+  \]
+  From the definition of $S^0_{1,0}$, for $(x,y)$ ranging over a compact $K \subset \bR^n$ , and any $\alpha \in {\bf Z}_+^m$, there is $C_{K,|\alpha|} > 0$ so that
+  \[
+    |\partial^{\alpha}_{\xi} P(x,y,\xi)| \le C_{K,|\alpha|}(1 + |\xi|)^{-|\alpha|}
+  \]
+  so for $(x,y) \in K, \alpha=(\alpha_1,\alpha_2) \in {\bf Z}_+^m \times {\bf Z}_+^{n-m}$,
+  \[
+    |\partial_{\xi,\eta}^{\alpha} Q(x,y,\xi,\eta)|
+  \]
+  \[
+    \le \sum_{\alpha_1' \le \alpha_1} \left|\left(\partial_{\xi}^{\alpha_1'}  \int d\xi' P(x,y,\xi+\xi') {\cal F}_x\phi(\xi',y) e^{i x\xi'}\right)\partial_{\xi}^{\alpha_1-\alpha_1'} \partial_{\eta}^{\alpha_2}H(\xi,\eta)       \right|
+  \]
+  \[
+    \le C \sum_{\alpha_1' \le \alpha_1} \left|\left(  \int d\xi' (1 + |\xi+\xi'|)^{-|\alpha_1'|} {\cal F}_x\phi(\xi',y) e^{i x\xi'}\right)\partial_{\xi}^{\alpha_1-\alpha_1'} \partial_{\eta}^{\alpha_2}H(\xi,\eta)       \right|
+  \]
+  Rearranging the conclusion of Lemma \ref{thm:prebao},
+  \[
+    (1 + |\xi + \xi'|)^{-k} \le (1+|\xi|)^k(1+|\xi'|)^{-k},
+  \]
+  so
+  \[
+    |\partial_{\xi,\eta}^{\alpha} Q(x,y,\xi,\eta)| \le C \sum_{\alpha_1' \le \alpha_1} (1 + |\xi|)^{-|\alpha_1'|}|\partial_{\xi}^{\alpha_1-\alpha_1'}\partial_{\eta}^{\alpha_2}H(\xi,\eta) |
+  \]
+  \[
+    \le C \sum_{\alpha_1' \le \alpha_1} (1 + |\xi|)^{-|\alpha_1'|}(1 + |(\xi,\eta)|)^{-|\alpha_1|+|\alpha_1'|-\alpha_2}
+  \]
+  since $H \in S^0_{1,0}(\bR^n)$. From norm equivalence, can replace $|(\xi,\eta)|$ with $|\xi| + |\eta|$   for the price of changing $C$.  So the above is
+  \[
+    \le C \sum_{\alpha_1' \le \alpha_1} (1 + |\xi|)^{-|\alpha_1'|}(1 + |\xi| + |\eta|)^{-|\alpha_1|+|\alpha_1'|-|\alpha_2|}
+  \]
+  \[
+    \le C \sum_{\alpha_1' \le \alpha_1} (1 + |\xi|+ |\eta|)^{-|\alpha_1'|}(1 + |\xi| + |\eta|)^{-|\alpha_1|+|\alpha_1'|-|\alpha_2|}
+  \]
+  \[
+    \le C (1 + |\xi| + |\eta|)^{-|\alpha_1|-|\alpha_2|} \le C(1 + |(\xi,\eta)|)^{-|\alpha|}
+  \]
+  whence $Q \in S^{0}_{1,0}(\bR^n)$.
+
+\end{proof}
 

### 2025-03-22 repro/papers/ssefound/paper.tex
more

@@
 
-\section{Symmetric Hyperbolic Systems}
+\section{Local Analysis of Surface Source Field}
 $k \times  k$ symmetric hyperbolic system: in $\bR^{n+1}$. Suppose that $u \in L^2_{\rm loc}(\bR^{n+1},\bR^k)$ is a weak solution of
@@
 \end{equation}
-The coefficieent matrices $A_i \in C^{\infty}(\bR^n, \bR^{k \times k}), i=0,...,n$ are symmetric.
+The coefficieent matrices $A_i \in C^{\infty}(\bR^n, \bR^{k \times
+  k}), i=0,...,n$ are symmetric, $A_0$ is positive definite.
+
+Notation:$x_0=t$, $\partial_t=\partial_0$, $\bx = (x_1,...,x_{n-1})$,
+$\partial_x=(\partial_1,...\partial_{n-1})$, $x_n=z$, $\partial_z =
+\partial_{n}$, $M=A_0(\bx)$, $A = (A_1,...,A_{n-1}$, $L=A_{n}$.
+
+Asymptotic solution 
+\[
+u(t,\bx,z) \approx e^{i\omega(t + \psi(\bx,z))} U(\bx,z)\sum_{n=0}^{\infty}
+a_n(t,\bx,z)
+\]
+Substitute first term in equation \ref{eqn:shs}, obtain
+\[
+(M\partial_t + A\cdot \partial_x + L \partial_z)u(t,\bx,z) =
+\]
+\[
+ e^{i\omega(t + \psi(\bx,z))}\left[ i\omega (M + A
+\cdot \partial_x \psi(\bx,z) + L \partial_z
+\psi(t,\bx,z))U(\bx,z)a_0(t,\bx,z) \right.
+\]
+\[
+\left. + M(\bx,z) U(\bx,z) \partial_t a_0(t,\bx,z) + \partial_x
+(a_0(t,\bx,z) U(\bx,z))\cdot A + L\partial_z (a_0(t,\bx,z)U(\bx,z))\right]= 0
+\]
+Assume $\Gamma$ is an open neighborhood of $0$ in $\bR^{n-1}$, $Z>0$ so that for each
+$\xi \in \Gamma, |\xi|=1$ and $|z| < Z$, there are $l$ independent
+solutions
+$(\zeta_1(\bx,z,\xi),U_1(\bx,z,\xi)),...,(\zeta_l(\bx,z,\xi), U_l(\bx,z,\xi))$ of
+\[
+ (M(\bx,z) + A\cdot \xi + L \zeta) U=0.
+\]
+The $U$s are orthogonal, and might as well normalize by
+$|U|=1$. For each $\xi \in \Gamma$, construct by method of characteristics solutions
+$\psi_1,...,\psi_l$ of 
+\[
+\partial_z \psi_i(\bx,z,\xi) = \zeta_i(\bx,z,\partial_x\psi(\bx,z,\xi))
+\]
+with $\psi_l(\bx,0,\xi) =\xi \cdot x$. The corresponding Hamiltonian
+system for $(\bx_i(z;\bx_0,\xi_0),\xi_i(z;\bx_0,\xi_0))$ with
+$(\bx_0,\xi_0) \in \Gamma$ is
+\begin{equation}
+  \label{eqn:zetahe1}
+  \frac{d\bx_i}{dz} = - \partial_{\xi}\zeta_i(\bx_i,z,\xi_i),\,
+  \bx_i(0;\bx_0,\xi_0) = \bx_0;
+\end{equation}
+\begin{equation}
+  \label{eqn:zetahe2}
+  \frac{d\xi_i}{dz}
+  = \partial_x\zeta_i(\bx_i,z,\xi_i),\, \xi_i(0;\bx_0,\xi_0) = \xi_0.
+\end{equation}
+and
+\[
+  \psi_l(\bx_i(z;\bx_0,\xi_0),z,\xi_i(z;\bx_0,\xi_0))= \bx_0\cdot\xi_0
+\]
+\begin{equation}
+  \label{eqn:psihe}
+- \int_0^z ds \xi(s;\bx_0,\xi_0)\cdot \partial_{\xi}\zeta_i(\bx_i(s;\bx_0,\xi_0),s,\xi_i(s;\bx_0,\xi_0))
+\end{equation}
+
+$\Pi_0$ = projection on null space $V_0$ of $M(t,\bx)$, $\Pi_1=I-\Pi_0$ = projection on range $V_1$. Denote the rank of $M$ by $l={\rm dim} V_1$ Then
+
+\section{Symmetric Hyperbolic Systems}
 
@@
 \[
-  M\partial_t u + Ku = \Pi_1 M \partial_t \Pi_1 u + \sum_{\alpha,\beta=0,1} \Pi_{\alpha}K\pi_{\beta}u + \mbox{ l.o.t. }
+  M\partial_t u + Ku = \Pi_1 M \partial_t \Pi_1 u + \sum_{\alpha,\beta=0,1} \Pi_{\alpha}K\Pi_{\beta}u + \mbox{ l.o.t. }
 \]
@@
 \[
-  \Pi_0 u - L_{0,0}K_{0,0}\Pi_u = \Pi_0 u - L_{0,0}K_{0,0}u = \Pi_0 u + L_{0,0}K_{0,1}u \in H^{-1}_{\rm loc}(\bR^{n+1})^k
+  \Pi_0 u - L_{0,0}K_{0,0}\Pi_u = \Pi_0 u - L_{0,0}K_{0,0}u = \Pi_0 u + L_{0,0}K_{0,1}u \in H^{1}_{\rm loc}(\bR^{n+1})^k
 \]
@@
 \]
-Then \ref{eqn:decomp1} can be rewritten
+Then equation \ref{eqn:decomp1} can be rewritten
 \begin{equation}
@@
 \end{equation}
-Note that for each $t \in \bR, (\bx,\xi) \in T^*(\bR^n)$, both $V_0$ and $V_1$ are invariant subspaces of $\tilde{K}(t,\bx,\xi)$, and $V_0 \subset {\rm ker}(\tilde{K}(t,\bx,\xi)$.
+Note that for each $t \in \bR, (\bx,\xi) \in T^*(\bR^n)$, both $V_0$ and $V_1$ are invariant subspaces of $\tilde{K}(t,\bx,\xi)$, and $V_0 \subset {\rm ker}\tilde{K}(t,\bx,\xi)$.
 Also, $M$ is invertible when restricted to $V_1 = {\rm Rng} M = {\rm Rng } \Pi_1$ , but not (necessarily) positive definite.
 
-Diagonalization: for $(t,\bx,\tau,\xi) \in T^*(\bR^n)$, define the restricted characteristic polynomial
+Restricted determinant: if $V$ is an invariant subspace of $A$, then ${\rm det}_{V} A$ is the determinant of the matrix of $A$ in any orthonormal basis of $V$.
+For $(t,\bx,\tau,\xi) \in T^*(\bR^n)$, define the restricted characteristic polynomial
 \[
-  {\cal C}(t,\bx,\tau,\xi) = \det (M(t,\bx)\tau + \tilde{K}(t,\bx,\xi) )= 0
+  {\cal C}(t,\bx,\tau,\xi) = {\rm det}_{V_1} (M(t,\bx)\tau + \tilde{K}(t,\bx,\xi) )= 0
 \]
@@
 \]
-$U_1,...,U_m$ span a subspace $\tilde{V}_1(t,\bx,\xi)$. Scale $U_i$ so that $|U_i(t,\bx,\xi)| = |\xi|$.
+$U_1,...,U_m$ span a subspace $\tilde{V}_1(t,\bx,\xi)$. Scale $U_i$ so that $|U_i(t,\bx,\xi)| = 1$.
 
@@
 So both are symbol-valued functions, of orders 1 and 0 respectively.
-Since the $U$'s must be orthogonal, $U_i^TU_j = |\xi|^2\delta_{ij}$.
+Since the $U$'s must be orthogonal, $U_i^TU_j = \delta_{ij}$.
 
@@
 
-Set $u=(p,v_x,v_z), \bx=(t,x), \xi=(\omega,\xi_x), z \mapsto t$. Then \ref{eqn:awe} takes the form \ref{eqn:shs} with
+Set $u=(p,v_x,v_z), \bx=x, x_2=z. \xi=\xi_x$. Then \ref{eqn:awe} takes the form \ref{eqn:shs} with
 \[
-  M =
+  L =
   \left(
@@
   \right),
-  A_1 ={\rm diag}\left(\frac{1}{\kappa},\rho,\rho\right),
-  A_2 =
+  M ={\rm diag}\left(\frac{1}{\kappa},\rho,\rho\right),
+  A =
  \left(
@@
 \]
+\[
+  M + A\xi + L\zeta =
+  \left(
+    \begin{array}{ccc}
+      \frac{1}{\kappa} & \xi & \zeta \\
+      \xi & \rho & 0 \\
+      \zeta & 0 & \rho
+    \end{array}
+  \right)
+\]
+Determinant is
+\[
+  {\rm det} = \rho\left(\frac{\rho}{\kappa} - \xi^2 - \zeta^2\right)
+\]
+Therefore take $\Gamma = \{\xi: |\xi| \lt \mbox{inf
+}\{\sqrt{\rho(\bx,z)/kappa(\bx,z)}, \mbox{ all } |z| < Z\}$. For $\xi
+\in \Gamma$, there are two roots $\zeta = \pm \sqrt{\rho/\kappa - \xi^2}.
+
 So $V_0={\rm span}(\{e2\}),\, V_1={\rm span}(\{e1,e3\}), \Pi_0 = {\rm diag}(0,1,0)$ 
@@
   \[
-    (1 + |\xi + \xi'|)^{-k} \le (1+|\xi|)^k(1+|\xi'|)^{-k},
+    (1 + |\xi + \xi'|)^{-k} \le (1+|\xi'|)^k(1+|\xi|)^{-k},
   \]
   so
+  \begin{equation}
+    \label{eqn:qbound}
+    |\partial_{\xi,\eta}^{\alpha} Q(x,y,\xi,\eta)| \le C \sum_{\alpha_1' \le \alpha_1} (1 + |\xi|)^{-|\alpha_1'|}|\partial_{\xi}^{\alpha_1-\alpha_1'}\partial_{\eta}^{\alpha_2}H(\xi,\eta) |
+  \end{equation}
+  Recall that $ES(H) \cap {\cal N} = \emptyset$. That is, there is $s>0$ so that for
+  $k,N \in {\bf Z}_+, \beta \in {\bf Z_+}^n$ with $|\beta| \le k$, there are $C_{k,N} \ge 0$ so that for $|\xi| \le s|\eta|$,
   \[
-    |\partial_{\xi,\eta}^{\alpha} Q(x,y,\xi,\eta)| \le C \sum_{\alpha_1' \le \alpha_1} (1 + |\xi|)^{-|\alpha_1'|}|\partial_{\xi}^{\alpha_1-\alpha_1'}\partial_{\eta}^{\alpha_2}H(\xi,\eta) |
+    |\partial_{\xi,\eta}^{\beta}H(\xi,\eta)| \le C_{k,N} |(\xi,\eta)|^{-N}
+  \]
+  In particular, if $k, N = |\alpha|$, then the integrand in each summand on the RHS of inequality \ref{eqn:qbound} is bounded by an $|\alpha|$-dependent multiple of $(1+|(\xi,\eta)|)^{-|\alpha|}$ in the conic neighborhood of ${\cal N}$ defined by $|\xi| \le s|\eta|$.
+
+  On the other hand, note that
+  \[
+    s|\eta| \le |\xi| \Rightarrow 1 + |(\xi,\eta)| \le 1 + |\xi| + |\eta| 
+    \le \frac{1+s}{s} (1+ |\xi|) 
+  \]
+  hence
+  \[
+    (1 + |\xi|)^{-|\alpha_1'|} \le \left(\frac{s}{1+s}(1+|(\xi,\eta)|)\right)^{-|\alpha_1'|}   
+  \]
+  In the complement defined by $|\xi|\ge s|\eta|$, the $\alpha_1'$  summand on the RHS of inequality \ref{eqn:qbound} is bounded by a multiple of
+  \[
+    \left(\frac{s}{1+s}(1 + |(\xi,\eta)|)\right)^{-|\alpha_1'|}|\partial_{\xi}^{\alpha_1-\alpha_1'}\partial_{\eta}^{\alpha_2}H(\xi,\eta) |
+  \]
+  which is in turn bounded by a multiple of
+  \[
+    \left(\frac{s}{1+s}\right)^{-|\alpha|} (1+|(\xi,\eta)|)^{-|\alpha|}
+  \]
+  because $H \in S^0(\bR^n)$.
+
+  These bounds together cover all of $T^*(\bR^n)$, so imply the bounds necessary to conclude that $Q \in S^0(\bR^n)$.
+
+  The assertion about the essential support is equivalent to existence of $C>0$ for any $\beta \in {\bf Z}_+^n, N \in {\bf Z}_+$ so that
+  \[
+    |(\partial_{\xi,\eta}^{\beta} Q)(x,y,\xi, \eta)| \le C(1+|(\xi,\eta)|)^{-N}
+  \]
+  for $\xi \notin ES(P)$ or $(\xi,\eta) \notin ES(H)$.
+\end{proof}
+
+\section{Computing Q}
+
+\subsection{Cauchy problem}
+\[
+  \partial_t S(t) = -K S(t), S(0)=I
 \]
 \[
-    \le C \sum_{\alpha_1' \le \alpha_1} (1 + |\xi|)^{-|\alpha_1'|}(1 + |(\xi,\eta)|)^{-|\alpha_1|+|\alpha_1'|-\alpha_2}
-  \]
-  since $H \in S^0_{1,0}(\bR^n)$. From norm equivalence, can replace $|(\xi,\eta)|$ with $|\xi| + |\eta|$   for the price of changing $C$.  So the above is
-  \[
-    \le C \sum_{\alpha_1' \le \alpha_1} (1 + |\xi|)^{-|\alpha_1'|}(1 + |\xi| + |\eta|)^{-|\alpha_1|+|\alpha_1'|-|\alpha_2|}
+  \partial_t S(-t) = -S(-t) \partial_t S(t) S(-t) = S(-t)K
 \]
 \[
-    \le C \sum_{\alpha_1' \le \alpha_1} (1 + |\xi|+ |\eta|)^{-|\alpha_1'|}(1 + |\xi| + |\eta|)^{-|\alpha_1|+|\alpha_1'|-|\alpha_2|}
+  \partial_t \delta S(t) = - \delta K S(t) - K \delta S(t)
 \]
 \[
-    \le C (1 + |\xi| + |\eta|)^{-|\alpha_1|-|\alpha_2|} \le C(1 + |(\xi,\eta)|)^{-|\alpha|}
+  \partial_t S(-t)\delta S(t) = S(-t)K\delta S(t) - S(-t) \delta K S(t) - S(-t)K \delta S(t)
 \]
-  whence $Q \in S^{0}_{1,0}(\bR^n)$.
+\begin{equation}
+  \label{eqn:qevol1}
+  =  -S(-t)\delta K S(t)
+\end{equation}
 
-\end{proof}
+\subsection{PsiDO RHS}
+Suppose that $K \in OPS^1$ is skew-adjoint, at least in principal part: $K_1^*=-K_1$.
+Then there is unitary $U \in OPS^0$ so that $U(x,\xi)^*K_1(x,\xi)U(x,\xi) = i\Lambda(x,\xi) = i\mbox{diag}(\lambda_1(x,\xi),...\lambda_k(x,\xi))$, with scalar $\lambda_j \in OPS^1, j=1,...,k$ (in fact, homogeneous of degree 1).
+Hence
+\[
+  \partial_t S(t) = -iU \Lambda U^*S(t) + RS(t), R \in OPS^0 
+\]
+Write $\tilde{S} = U^*SU$. Then 
+\[
+  \partial_t \tilde{S}(t) = -i\Lambda \tilde{S}(t) + \mbox{ mod } OPS^0 
+\]
+If $u^0 = (u^0_1,...,u^0_k)$, then $\tilde{u}(t) = \tilde{S}(t)u^0$ satisfies
+\[
+  \partial_t \tilde{u}(t) = -i\Lambda \tilde{u}(t) + ...
+\]
+so $\tilde{u}_i$ solves
+\[
+  \partial_t \tilde{u}_i(t) = -i\lambda_i \tilde{u}_i(t), \, i=1,...,k
+\]
+Let $\tilde{S}_i$ solve
+\[
+  \partial_t \tilde{S}_i(t) = - i\lambda_i \tilde{S}_i(t)
+\]
+Then
+\[
+  \tilde{u}_i(t) = \tilde{S}_i(t) u^0_i = (\tilde{S}(t)u^0)_i 
+\]
+whence
+\[
+  \tilde{S} = \mbox{diag}(\tilde{S}_1,....,\tilde{S}_k).
+\]
+
+From equation \ref{eqn:qevol1},
+\[
+  \tilde{Q}_i(t) = \tilde{S}_i(-t)\delta \tilde{S}_i(t)
+\]
+satisfies
+\[
+  \partial \tilde{Q}_i(t) = -i\tilde{S}_i(-t)\delta \lambda_i \tilde{S}_i(t), \, i=1,,,.k.
+\]
+and $\tilde{Q} = \mbox{diag}(\tilde{Q}_1,...,\tilde{Q}_k)$.
+From \cite{Tay:81}, Egorov's Theorem (Section VIII.1),
+\[
+  \tilde{Q}_i(t; x,\xi) = -i \int_0^t ds \,\delta \lambda_i(y_i(s;x,\xi),\eta_i(s;x,\xi))
+\]
+in which $(y_i(s;x,\xi),\eta_i(s;x,\xi))$ is the integral curve of the Hamiltonian vector field
+\[
+  H_i(x,\xi) = (\partial_{\xi}\lambda_i(x,\xi), - \partial_x\lambda_i(x,\xi))
+\]
+with $y_i(0;x,\xi)=x, \eta_i(0;x,\xi)=\xi$.
+
+\subsection{Generic H-J}
+Generic 1st order PDE:
+\begin{equation}
+  \label{eqn:gen}
+  F(x, \partial_x u) = 0.
+\end{equation}
+Suppose that $(x,p)$
+is curve in phase space, and
+\begin{equation}
+  \label{eqn:defp}
+  p(s) = \partial_x u(x(s)).
+\end{equation}
+Then
+\[
+  \frac{d}{ds}u(x(s))=p(s) \cdot \frac{dx}{ds}(s)
+\]
+Also
+\[
+  \frac{dp}{ds}(s) = \partial_x \partial_x u(x(s))\cdot
+  \frac{dx}{ds}(s).
+\]
+Note that
+\[
+  0 = \partial_x( F(x, \partial_x u(x))) = (\partial_x F)(x,
+  \partial_x u(x)) + 
+  (\partial_p F)(x, \partial_x u(x)) \partial_x \partial_x u(x)
+\]
+So if
+\begin{equation}
+  \label{eqn:H1}
+  \frac{dx}{ds} = \partial_p F(x,p)
+\end{equation}
+then
+\begin{equation}
+  \label{eqn:H2}
+  \frac{dp}{ds} = -\partial_x F(x,p)
+\end{equation}
+That is: if $u$ is a solution of $F(x,\partial_x u)=0$ and
+$(x(s),p(s))$ satisfies
+\[
+  p(s)=\partial_x u(x),\, \frac{dx}{ds} = (\partial_p
+  F)(x(s),\partial_x u(x(s)),
+\]
+then $(x(s),p(s))$ is a trajectory of the system \ref{eqn:H1},
+\ref{eqn:H2}, and
+\begin{equation}
+  \label{eqn:sol}
+  u(x(s))=u(x(0)) + \int_0^t ds p(s)\cdot (\partial_p F)(x(s),p(s)).
+\end{equation}
+Because the system the system \ref{eqn:H1},
+\ref{eqn:H2} is uniquely solvable, the result \ref{eqn:sol} holds for
+any solution $u$ of \ref{eqn:gen}, for $x(0)$ in the domain of
+definition of $u$, $p(0)=\partial_xu(x(0))$, and a trajectory $(x,p)$
+solving the system \ref{eqn:H1}, \ref{eqn:H2}, so long as $x(s)$
+remains in the domain of $u$.
+
+This observation implies a similar result for the H-J system
+\begin{equation}
+  \label{eqn:hjdef}
+  0 = \partial_t u + H(t,\bx,\partial u).
+\end{equation}
+Replace $x$ by $(t,x)$, $F(x,\partial_x u)$ by the RHS of equation
+\ref{eqn:hjdef}.  Then precisely the same results obtain, with $F$ is
+replaced by $H$.
 

### 2025-04-11 repro/papers/ssefound/shs/paper.tex
various

--- 2025-02-19 repro/papers/ssefound/paper.tex
@@
 \section{Symmetric Hyperbolic Systems}
-$k \times  k$ symmetric hyperbolic system: in $\bR^{n+1}$. Suppose that $u \in L^2_{\rm loc}(\bR^{n+1},\bR^k)$ is a weak solution of
-\begin{equation}
-  \label{eqn:shs}
-  (\sum_{i=0}^n A_i\partial_i ) u \in L^2_{\rm loc}(\bR^{n+1},\bR^k)
-\end{equation}
-The coefficieent matrices $A_i \in C^{\infty}(\bR^n, \bR^{k \times k}), i=0,...,n$ are symmetric.
 
@@
 \end{equation}
-Note that for each $t \in \bR, (\bx,\xi) \in T^*(\bR^n)$, both $V_0$ and $V_1$ are invariant subspaces of $\tilde{K}(t,\bx,\xi)$, and $V_0 \subset {\rm ker}(\tilde{K}(t,\bx,\xi)$.
+
+Note that for each $t \in \bR, (\bx,\xi) \in T^*(\bR^n)$, both $V_0$ and $V_1$ are invariant subspaces of $\tilde{K}(t,\bx,\xi)$, and $V_0 \subset {\rm ker}\tilde{K}(t,\bx,\xi)$.
 Also, $M$ is invertible when restricted to $V_1 = {\rm Rng} M = {\rm Rng } \Pi_1$ , but not (necessarily) positive definite.
 
-Diagonalization: for $(t,\bx,\tau,\xi) \in T^*(\bR^n)$, define the restricted characteristic polynomial
+Restricted determinant: if $V$ is an invariant subspace of $A$, then ${\rm det}_{V} A$ is the determinant of the matrix of $A$ in any orthonormal basis of $V$.
+For $(t,\bx,\tau,\xi) \in T^*(\bR^n)$, define the restricted characteristic polynomial
 \[
-  {\cal C}(t,\bx,\tau,\xi) = \det (M(t,\bx)\tau + \tilde{K}(t,\bx,\xi) )= 0
+  {\cal C}(t,\bx,\tau,\xi) = {\rm det}_{V_1} (M(t,\bx)\tau + \tilde{K}(t,\bx,\xi) )= 0
 \]
@@
 
-Choose a conic submanifold $\gamma \subset T^*(\bR^n)$, so that for $|t| \le \delta t, (\bx,\xi) \in \gamma$ there are exactly $m \le l$ distinct real roots $\tau$ of ${\cal C}(t,\bx,\tau,\xi)$.
-Denote these by $\tau_i(t,\bx,\xi), i=1,...,m$.
+Choose a conic submanifold $\gamma \subset T^*(\bR^n)$, so that for
+$|t| \le \delta t, (\bx,\xi) \in \gamma$ there are exactly $m \le l$
+distinct pure imaginary roots $i\tau$ of ${\cal C}(t,\bx,i\tau,\xi)$. Denote these by $i\tau_i(t,\bx,\xi), i=1,...,m$.
 
-Choose a conic $\gamma_1 \subset T^*(\bR^n)$ so that $\bar{\gamma_1} \subset \gamma$, and $Q \in OPS^0(\bR^n)$ with $ES(Q) \subset \gamma$, $Q(\bx,\xi) =1$ for $(\bx,\xi)\in \gamma_1$.
+%Choose a conic $\gamma_1 \subset T^*(\bR^n)$ so that $\bar{\gamma_1} \subset \gamma$, and $Q \in OPS^0(\bR^n)$ with $ES(Q) \subset \gamma$, $Q(\bx,\xi) =1$ for $(\bx,\xi)\in \gamma_1$.
 
-Let ${\cal N} = $ normal bundle of $x_0=0$ and $\Gamma \subset T^*(\bR^n)$ for which $\Gamma \cap {\cal N} = \emptyset$.
+%Let ${\cal N} = $ normal bundle of $x_0=0$ and $\Gamma \subset T^*(\bR^n)$ for which $\Gamma \cap {\cal N} = \emptyset$.
 
-Assume that $u \in L^2(\bR^{n+1}) \cap H^1_{\rm loc}(\Gamma \cap \Pi_{\cal N}^{-1}(\gamma_1^C))$.
+%Assume that $u \in L^2(\bR^{n+1}) \cap H^1_{\rm loc}(\Gamma \cap \Pi_{\cal N}^{-1}(\gamma_1^C))$.
 
@@
 \[
-  \tau_i M(\bx) U_i(t,\bx,\xi) + \tilde{K}(t,\bx,\xi) U_i(t,\bx,\xi)=0, \, i=1,..,m.
+  i\tau_i(t\bx,\xi) M(\bx) U_i(t,\bx,\xi) + \tilde{K}(t,\bx,\xi) U_i(t,\bx,\xi)=0, \, i=1,..,m.
 \]
-$U_1,...,U_m$ span a subspace $\tilde{V}_1(t,\bx,\xi)$. Scale $U_i$ so that $|U_i(t,\bx,\xi)| = |\xi|$.
+$U_1,...,U_m$ span a subspace $\tilde{V}_1(t,\bx,\xi)$.
 
-Note that $\tau_i(t,\bx,\xi), i=1,...,m$ are homogeneous of order 1 in $\xi$ and smooth in $\bx$, $U_i(t,\bx,\xi), i=1,...,m$ are homogeneous of order 0 in $\xi$, smooth in $\bx$.
-So both are symbol-valued functions, of orders 1 and 0 respectively.
-Since the $U$'s must be orthogonal, $U_i^TU_j = |\xi|^2\delta_{ij}$.
+Note that $\tau_i(t,\bx,\xi), i=1,...,m$ are homogeneous of order 1 in
+$\xi$ and smooth in $\bx$, $U_i(t,\bx,\xi), i=1,...,m$ are homogeneous
+of order 0 in $\xi$, smooth in $\bx$. So both are symbol-valued
+functions, of orders 1 and 0 respectively. The $U$'s are
+$M$-orthogonal, $U_i^TMU_j = \delta_{ij}$.
+
+The corresponding operators satisfy
+\begin{equation}
+  \label{eqn:ortho}
+  U(x,D)^*M(x) U(x,D) = U(x,d)M(x)U(x,D)^* = I \mbox{ mod } OPS^{-1}
+\end{equation}
+
+Extend the $\tau$s and $U$s to $t \in \bR$ smoothly, so that the above 
+properties are preserved for all $t$.
+
+The evolution equation
+\[
+  \partial_t \tilde{u}(t,\bx) =
+  i\tau_i(\t,\bx,\partial_x)\\tilde{u}(t,\bx)
+\]
+has a solution in $C^0(\bR,L^2(\bR^n))$, uniquely determined
+by $\tilde{u}(0,\bx)$. Let $\tilde{S}_i(t)$ denote the one-parameter
+group defined by this evolution equation. Then (weakly)
+\[
+  \partial_t \tilde{S}_i(t,\bx) =
+  i\tau_i(\t,\bx,\partial_x)tilde{S}_i(t,\bx),\,S(0) = I.
+\]
+Note that $\tilde{S}_i(t)$ is invertible and $\tilde{S}_i(t)^{-1} =
+\tilde{S}_i(-t)$.
+
+From \cite{Tay:81}, Egorov's Theorem (Section VIII.1), for $p \in OPS^s(\bR^n)$,
+\begin{equation}
+  \label{eqn:egorov}
+   \tilde{S}_i(-t) p \tilde{S}_i(t)=  p \circ C_i(t),\, p\tilde{S}_i(t) =
+   \tilde{S}_i(t)p \circ C_i(t)
+\end{equation}
+in which $C_i$ is the one-paraimeter family of symplectic maps given by
+\begin{equation}
+  \label{eqn:canon}
+  C_i(t)(x,\xi)=(y_i(t;x,\xi),\eta_i(t;x,\xi))
+\end{equation}
+Here $t \rightarrow y_i(t;x,\xi),\eta_i(t;x,\xi) $is the integral curve of the
+Hamiltonian vector field
+\[
+  H_i(t,x,\xi) = (\partial_{\xi}\tau_i(t,x,\xi), - \partial_x\tau_i(t,x,\xi))
+\]
+with $y_i(0;x,\xi)=x, \eta_i(0;x,\xi)=\xi$.
+
+For $p \in OPS^{\infty}$, define
+\[
+  \Gamma_i[p] = \int_0^T dt \, p \circ C_i(t)
+\]
 
@@
 
-\section{2D Acoustics}
-
-\begin{eqnarray}
-  \frac{1}{\kappa}\partial_t p + \nabla \cdot \bv & = & 0, \nonumber \\
-  \rho \partial_t \bv + \nabla p & = & 0.
-  \label{eqn:awe}                                     
-\end{eqnarray}
-
-Set $u=(p,v_x,v_z), \bx=(t,x), \xi=(\omega,\xi_x), z \mapsto t$. Then \ref{eqn:awe} takes the form \ref{eqn:shs} with
-\[
-  M =
-  \left(
-    \begin{array}{ccc}
-      0 & 0 & 1\\
-      0 & 0 & 0\\
-      1 & 0 & 0
-    \end{array}
-  \right),
-  A_1 ={\rm diag}\left(\frac{1}{\kappa},\rho,\rho\right),
-  A_2 =
- \left(
-    \begin{array}{ccc}
-      0 & 1 & 0\\
-      1 & 0 & 0\\
-      0 & 0 & 0
-    \end{array}
-  \right),
-\]
+\subsection{2D Acoustics}
+%###################
 So $V_0={\rm span}(\{e2\}),\, V_1={\rm span}(\{e1,e3\}), \Pi_0 = {\rm diag}(0,1,0)$ 
@@
 \]
-\section{2D Elasticity}
-
-Velocity, stress, state vector:
-\[
-  \bv = \left(
-    \begin{array}{c}
-      v_1 \\
-      v_2
-    \end{array}
-  \right), \,
-  \sigma = \left(
-    \begin{array}{c}
-      \sigma_{11} \\
-      \sigma_{12} \\
-      \sigma_{21} \\
-      \sigma_{22}
-    \end{array}
-  \right), \, 
-  u = \left(
-    \begin{array}{c}
-      \sigma\\
-      \bv
-    \end{array}
-  \right)
-\]
-Symmetry of stress: $\sigma_{12}=\sigma_{21}$. Compliance tensor (inverse of Hooke tensor):
-\[
-  C =\left(
-    \begin{array}{cccc}
-      c_{1111} &  c_{1112} &  c_{1121} &  c_{1122} \\
-      c_{1211} &  c_{1212} &  c_{1221} &  c_{1222} \\
-      c_{2111} &  c_{2112} &  c_{2121} &  c_{2122} \\
-      c_{2211} &  c_{2212} &  c_{2221} &  c_{2222} \\
-    \end{array}
-  \right)
-\]
-Symmetries: $C_{ijkl}=C_{klij}, C_{ijkl}=C_{jikl}=C_{ijlk}$. Density = $\rho$
-
-Elastodynamics:
-\begin{eqnarray}
-  \label{eqn:const}
-  C \partial_ t \sigma & = & \frac{1}{2}(\nabla \bv + \nabla \bv^T) \\
-  \label{eqn:mombal}
-  \rho \partial_t \bv & = & \nabla \cdot \sigma
-\end{eqnarray}
-
-Matrix representation: set
-\[
-  B_1=\left(
-    \begin{array}{cccc}
-      1 & 0 & 0 & 0 \\
-      0 & \frac{1}{2} & \frac{1}{2} & 0
-    \end{array}
-  \right),\,
-  B_2=\left(
-    \begin{array}{cccc}
-      0 & \frac{1}{2} & \frac{1}{2} & 0 \\
-      0 & 0 & 0 & 1
-    \end{array}
-  \right)
-\]
-and
-\[
-  A_0 = \left(
-    \begin{array}{cc}
-      C & 0 \\
-      0 & \rho I
-    \end{array}
-  \right), \,
-  A_j = -\left(
-    \begin{array}{cc}
-      0 & B_j^T \\
-      B_j & 0
-    \end{array}
-  \right), \, j=1,2
-\]
-
-Then the elastodynamics equations \ref{eqn:const}, \ref{eqn:mombal} are equivalent to
-\[
-  (A_0\partial_0 + A_1 \partial_1 + A_2 \partial_2)u =0.
-\]
-
-$u = (\sigma, \bv)^T \in \bR^6$ is a null vector for $A_2$ iff $B_2\bv = 0$ and $B_2^T \sigma=$, that is, $\bv=0$ and $\sigma_{12}=\sigma_{21}=\sigma_{22} = 0$.
-Thus $U_0= \mbox{ker }A_2 = \mbox{span } \{\be_1\}$ and  $U_1 = \mbox{rng }A_2 = \mbox{span }\{\be_j: 2\le j \le 6\}$, and $\Pi_0 = \mbox{diag}(1,0,0,0,0,0)$, $\Pi_1 = I-\Pi_0$.
-
-Eliminating zero rows and columns, equation \ref{eqn:shsnull1} for $u_1 = (\sigma_{12}, \sigma_{21}, \sigma_{22}, v_1, v_2)^T$ reads
-\begin{eqnarray}
-  \frac{1}{2}\partial_2v_1 &=& -((C \partial_0 \sigma)_{21} + \frac{1}{2}\partial_1v_2)\\
-  \partial_2 v_2 &=& -(C \partial_0 \sigma)_{41}\\
-  \partial_2 \sigma_{12} &=& -(\rho \partial_0v_1)
-\end{eqnarray}
-
-Thus
-\begin{eqnarray}
-  A_{000} &=& \Pi_0 A_0 \Pi_0 = \mbox{ diag}(c_{1111},0,0,0,0,0) \nonumber \\
-  A_{100} &=& \Pi_0 A_1 \Pi_0 = 0 \nonumber \\
-  A_{001} &=& \Pi_0 A_0 \Pi_1 =
-              \left(
-              \begin{array}{cccccc}
-                0 & c_{1112} & c_{1121} & c_{1122} & 0 & 0 \\
-                0 & 0 & 0 & 0 & 0 & 0 \\
-                0 & 0 & 0 & 0 & 0 & 0 \\                
-                0 & 0 & 0 & 0 & 0 & 0 \\
-                0 & 0 & 0 & 0 & 0 & 0 \\
-                0 & 0 & 0 & 0 & 0 & 0
-              \end{array}
-                                    \right) \nonumber \\
-  A_{101} &=& \Pi_0 A_1 \Pi_1 =
-                            \left(
-              \begin{array}{cccccc}
-                0 & 0 & 0 & 0 & 1 & 0 \\
-                0 & 0 & 0 & 0 & 0 & 0 \\
-                0 & 0 & 0 & 0 & 0 & 0 \\                
-                0 & 0 & 0 & 0 & 0 & 0 \\
-                0 & 0 & 0 & 0 & 0 & 0 \\
-                0 & 0 & 0 & 0 & 0 & 0
-              \end{array}
-                                    \right)
-                                    \label{eqn:ela0}
-\end{eqnarray}
-So equation \ref{eqn:rel01} reduces to
-\[
-  c_{1111}\sigma_{11} = -(c_{1112}\sigma_{12} + c_{1121}\sigma_21 + c_{1122}\sigma_{22} + \partial_1 \int_{-\infty}^t v_1)
-\]
-This is just the first row of equation \ref{eqn:const}, integrated in time! Now apply $\partial_2$ to both sides to obtain the missing equation.
-
-\append{G\r{a}rding Interlude}
-
-In the following $U$ is an open set in $\bR^n$.
-
-\begin{definition}
-  A $k \times k$ valued symbol $M(\bx,\xi) \in OPS^m(U)$ is {\em elliptic} iff $|\xi|^{-m}M(\bx,\xi)$ is uniformly invertible for $(\bx,\xi) \in T^*(U) \setminus \{0\}$.
-That is, there exists $C>0$ so that for all  $(\bx,\xi) \in T^*(U) \setminus \{0\}$, $v \in \bR^k$,
-  \[
-    |M(\bx,\xi)v| \ge C|\xi|^m|v|.
-  \]
-\end{definition}
-
-Connection to defintion V.1.1 in \cite{Tay:81}: concerns the first order system
-\[
-  (\frac{\partial}{\partial y} - K(y,x,D_x))u = 0
-\]
-where $u=u(y,x) \in \bR^k$, $K(y,\cdot,\cdot) \in C^{\infty}(\bR,OPS^1(U))$.
-According to the cited definition, this system is elliptic iff the principal symbol $K_1(y,x,\xi)$ has no purely imaginary eigenvalues.
-The symbol of the operator on the LHS above is $i\eta I - K(y,x,\xi)$, which is not the symbol of a $\Psi$DO unless $K$ is differential.
-However $K$ having no purely imaginary eigenvalues iz equivalent to  $i\eta I - K(y,x,\xi)$ being invertible.
-
-\begin{lemma} \label{thm:lema1}
-``Interpolation'' inequality: for any $m, N \in {\bf Z}_+, 0 < \epsilon < 1$,
-\[
-  \lxr^{\frac{m}{2}-1} \le \epsilon^2 \lxr^{\frac{m}{2}} + C^2_{\epsilon,N,m}\lxr^{-N}. 
-\]
-\end{lemma}
-
-\begin{proof}
- Equivalent to
-\[
-  \lxr^{-1} \le \epsilon^2 + C^2_{\epsilon,N,m}\lxr^{-N-\frac{m}{2}}. 
-\]
-Holds trivially if $\lxr \ge \epsilon^{-2}$. Otherwise $\lxr^{N+\frac{m}{2}-1} \le \epsilon^{-2(N+\frac{m}{2}-1)} \equiv C_{\epsilon,N,m}$.
-\end{proof}
-
-\begin{proposition}
-\label{thm:propa1}(clone of V.6.2 in \cite{Tay:81}): Suppose that $p \in OPS^0(U,\bR^k)$ has a SPD principal symbol $p_0$: there is $C > 0$ so that for sufficiently large $|\xi|$,
-  \[
-    p_0(x,\xi) = p_0(x,\xi)^T, p_0(x,\xi) \ge C
-  \]
-  Then there existss $B \in OPS^0(U,\bR^k)$ so that $P = B^*B \mbox{ mod } OPS^{-\infty}$.
-\end{proposition}
-
-\begin{proof}
-  From the spectral theorem, for each $(x,\xi) \in T^*(U)$, can construct SPD $b_0(x,\xi)$ for which $p_0(x,\xi) = b_0(x,\xi)^2$.
-Suppose that we have built $b_0,...,b_{j-1}$ with $b_i \in S^{-i}$ homogeneous of degree $-i$ for large $|\xi|$ so that
-  $r_{j}(x,D) = p(x,D) - (b_0(x,D) +...+ b_{j-1}(x,D))^2 \in OPS^{-j}$. Want $b_j \in S^{-j}$ homogeneous of degree $j$ so that
-  \[
-    p(x,D)-(b_0 +...+ b_{j-1}+b_j)^2 = r_{j-1} -  (b_0 +...+ b_{j-1})b_j - b_j((b_0 +...+ b_{j-1}) - b_j^2
-  \]
-  \[
-    = r_{j-1} - (b_0 b_j+b_j b_0) \mbox{ mod } OPS^{-j-1}
-  \]
-  is in $OPS^{j-1}$. This will be the case if the principal symbols match: denoting the principal symbol of $r_{j-1}$ by $r_{j-1,0}$, this requires that
-  \[
-    r_{j-1,0}(x,\xi) = b_0(x,\xi)b_j(x,\xi) + b_j(x,\xi)b_0(x,\xi).
-  \]
-  Since $b_0$ is SPD, the map $a \mapsto b_0a+ab_0$ on $k \times k$ symmetric matrices $a$ is invertible, as its eigenvalues are sums of the eigenvalues of $b_0$.
-So the above equation has a unique symmetric solution, and the inductive step in the construction of $b \in OPS^0$ is finished.
-\end{proof}
-
-\begin{theorem}
-  \label{thm:garding}
-  ``G\r{a}rding for systems'': Suppose that $P \in OPS^m(U,\bR^k)$ is elliptic, that is, its principal symbol $p_0$ satisfies 
-\[
-  p_0^*(x,\xi)p_0(x,\xi) \ge C|\xi|^{2m}, \, (x,\xi) \in T^*(U).
-\]
-Then for any $K \subset \subset U$, $N \ge 1$ there exists $C_{K,N} > 0$ so that for $u \in C^{\infty}_0(K)$, 
-\[
-  \|Pu\|^2_0 \ge \frac{C}{2}\|u\|^2_m - C_{K,N}\|u\|^2_{-N}.
-\]
-\end{theorem}
-
-\begin{proof}
-  For any $alpha \in [0,1)$, $\ldr^{-m}P^*P\ldr^{-m} - \frac{C}{2} \in OPS^0$ is self-adjoint  and its principal symbol $\lxr^{-,m}p_0^Tp_0\lxr^{-m} - \frac{C}{2} \in S^0$ is positive semi-definite.
-Therefore from Proposition \ref{thm:propa1} there exists $B \in OPS^0$ for which
-  \[
-    \ldr^{-m}P^*P\ldr^{-1} - \frac{C}{2} = B^*B \mbox{ mod } OPS^{-\infty}
-  \]
-  or
-  \[
-    P^*P -\frac{C}{2} \ldr^{2m} = \ldr B^*B \ldr \mbox{ mod } OPS^{-\infty}
-  \]
-  So for every $N  \in {\bf Z}_+$ there is $C_{K,N} > 0$ so that for $u \in C_0^{\infty}(K)$,  
-  \[
-    \langle u,P^*P u \rangle \ge \frac{C}{2}\|u\|_m^2 -   C_{K,N}\|u\|^2_{-N}
-  \]
-
-\end{proof}
-
-\begin{definition} $P \in OPS^m(U,\bR^k)$ is (microlocally) elliptic in the closed conic set $\Gamma \subset T^*(U)$ iff for some $C>0$ its principal symbol $p_0$ satisfies
-  \[
-    p(x,\xi)^*p(x,\xi) \ge C|\xi|^m
-  \]
-\end{definition}
-
-\begin{theorem} Suppose that $P \in OPS^m(U,\bR^k)$ is elliptic in the closed conic $\Gamma \subset T^*(U)$.
-Then for any $Q \in OPS^m(U,\bR)$ with $ES(Q) \subset \Gamma$, $N \in {\bf Z}_+$, and compact $K \subset \subset U$, there exists $C_Q > 0$ and $C_{Q,K,N} \ge 0$ so that for any $u \in C_0^{\infty}(K)$,
-  \[
-    \|P(x,D)u\|^2_0  \ge C_Q\|Q(x,D)u\|_0^2 - C_{Q,K,N}\|u\|^2_{-N}.
-  \]
-\end{theorem}
 

### 2025-04-11 repro/papers/ssefound/paper.tex
various

@@
 \section{Local Analysis of Surface Source Field}
-$k \times  k$ symmetric hyperbolic system: in $\bR^{n+1}$. Suppose that $u \in L^2_{\rm loc}(\bR^{n+1},\bR^k)$ is a weak solution of
+$k \times  k$ symmetric hyperbolic system: in $\bR^{n+1}$, $\Omega
+\subset \bR^n, \Gamma_r, \Gamma_s \subset \partial \Omega$. $\Gamma_r$
+is receiver locus, $\Gamma_s$ is sourcee locus. Assume that $\Gamma_s$
+is a subset of a coordinate neighborhood in  which coordinates
+$(z,\bx) \in \bR^{n-1} \times \bR$ are
+chosen so that $\Gamma_i\subset \{(z,\bx): z=z_i\}, i=s,r$ with $z_s <
+z_r$.
+
+Assume $M, A_1,...,A_{n-1}, L \in C^{\infty}(\bR^n, \bR^{k\times k})$
+are symmetric and $M$ is positive-definite.
+
+Let $\chi \in C^{\infty}(\Gamma_s \times \bR, \bR^k)$.
+Develop an asymptotic solution of
 \begin{equation}
   \label{eqn:shs}
-  (\sum_{i=0}^n A_i\partial_i ) u \in L^2_{\rm loc}(\bR^{n+1},\bR^k)
+  (M\partial_t + \sum_{i=1}^{n-1}A_i \partial_{x_i} + L \partial_z)
+  u = 0,\, u(t,z,\bx_s) = \chi(t,\bx)e^{i\omega(t + \bx \cdot \xi_0)}
 \end{equation}
-The coefficieent matrices $A_i \in C^{\infty}(\bR^n, \bR^{k \times
-  k}), i=0,...,n$ are symmetric, $A_0$ is positive definite.
+of the form
+\begin{equation}
+  \label{eqn:hfa}
+  u(t,z,\bx) \approx e^{i\omega(t + \psi(z,\bx))} \sum_{n=0}^{\infty}
+  (i\omega)^{-n}a^n(t,z,\bx) U^n(z,\bx)
+\end{equation}
+in which $a^n, n=0,...$ is scalar and $U^n, n=0,...$ is in $\bR^k$.
 
-Notation:$x_0=t$, $\partial_t=\partial_0$, $\bx = (x_1,...,x_{n-1})$,
-$\partial_x=(\partial_1,...\partial_{n-1})$, $x_n=z$, $\partial_z =
-\partial_{n}$, $M=A_0(\bx)$, $A = (A_1,...,A_{n-1}$, $L=A_{n}$.
-
-Asymptotic solution 
-\[
-u(t,\bx,z) \approx e^{i\omega(t + \psi(\bx,z))} U(\bx,z)\sum_{n=0}^{\infty}
-a_n(t,\bx,z)
-\]
 Substitute first term in equation \ref{eqn:shs}, obtain
 \[
-(M\partial_t + A\cdot \partial_x + L \partial_z)u(t,\bx,z) =
+(M\partial_t + A\cdot \partial_x + L \partial_z)u(t,z,\bx) =
 \]
 \[
- e^{i\omega(t + \psi(\bx,z))}\left[ i\omega (M + A
-\cdot \partial_x \psi(\bx,z) + L \partial_z
-\psi(t,\bx,z))U(\bx,z)a_0(t,\bx,z) \right.
+ e^{i\omega(t + \psi(z,\bx))}\left[ i\omega (M + A
+\cdot \partial_x \psi(z,\bx) + L \partial_z
+\psi(z,\bx))U(z,\bx)a^0(t,z,\bx) \right.
 \]
 \[
-\left. + M(\bx,z) U(\bx,z) \partial_t a_0(t,\bx,z) + \partial_x
-(a_0(t,\bx,z) U(\bx,z))\cdot A + L\partial_z (a_0(t,\bx,z)U(\bx,z))\right]= 0
++ M(z,\bx) U^0(z,\bx) \partial_t a^0(t,z,\bx) + A(z,\bx)\cdot \partial_x
+(a^0(t,z,\bx) U^0(z,\bx))
 \]
-Assume $\Gamma$ is an open neighborhood of $0$ in $\bR^{n-1}$, $Z>0$ so that for each
-$\xi \in \Gamma, |\xi|=1$ and $|z| < Z$, there are $l$ independent
-solutions
-$(\zeta_1(\bx,z,\xi),U_1(\bx,z,\xi)),...,(\zeta_l(\bx,z,\xi), U_l(\bx,z,\xi))$ of
 \[
- (M(\bx,z) + A\cdot \xi + L \zeta) U=0.
+ \left.   + L(z,\bx)\partial_z (a^0(t,z,\bx)U^0(z,\bx))\right]
 \]
-The $U$s are orthogonal, and might as well normalize by
-$|U|=1$. For each $\xi \in \Gamma$, construct by method of characteristics solutions
+\[
++  e^{i\omega(t + \psi(z,\bx))}\left[ (M + A
+\cdot \partial_x \psi(z,\bx) + L \partial_z
+\psi(z,\bx))U^1(z,\bx)a^1(t,z,\bx) \right.
+\]
+\[
+  + (i\omega)^{-1} (M(z,\bx) U^1(z,\bx) \partial_t a^1(t,z,\bx)
+    + A(z,\bx)\cdot \partial_x  (a^1(t,z,\bx) U^1(z,\bx))
+\]
+\begin{equation}
+  \label{eqn:series}
+  \left.   + L(z,\bx)\partial_z (a^1(t,z,\bx)U^1(z,\bx)))\right]
++...
+\end{equation}
+
+Assume $\Sigma$ is an open neighborhood of the zero section in
+$T^*\Gamma_s$, so that for each $(\bx,\xi) \in \Sigma$, $|\xi|$ is
+sufficiently small that $M + A\cdot \xi$ is positive
+definite. Further, for $z$ sufficiently small,
+there are $l$ independent solutions of 
+$(\zeta_1(z,\bx,\xi),U^0_1(z,\bx,\xi)),...,(\zeta_l(z,\bx,\xi), U^0_l(z,\bx,\xi))$ of
+\begin{equation}
+  \label{eqn:nullsp}
+ (M + A\cdot \xi + L \zeta) U^0=0, \, \zeta\ne 0.
+\end{equation}
+The $U^0$s are orthogonal, and might as well normalize by
+$|U^0|=1$. For each $(\bx,\xi) \in \Gamma$, construct by method of characteristics solutions
 $\psi_1,...,\psi_l$ of 
-\[
-\partial_z \psi_i(\bx,z,\xi) = \zeta_i(\bx,z,\partial_x\psi(\bx,z,\xi))
-\]
+\begin{equation}
+  \label{eqn:eik}
+\partial_z \psi_i(z,\bx,\xi) = \zeta_i(z,\bx,\partial_x\psi(z,\bx,\xi))
+\end{equation}
 with $\psi_l(\bx,0,\xi) =\xi \cdot x$. The corresponding Hamiltonian
@@
   \label{eqn:zetahe1}
-  \frac{d\bx_i}{dz} = - \partial_{\xi}\zeta_i(\bx_i,z,\xi_i),\,
+  \frac{d\bx_i}{dz} = - \partial_{\xi}\zeta_i(z,\bx_i,\xi_i),\,
   \bx_i(0;\bx_0,\xi_0) = \bx_0;
 \end{equation}
+
 \begin{equation}
@@
   \frac{d\xi_i}{dz}
-  = \partial_x\zeta_i(\bx_i,z,\xi_i),\, \xi_i(0;\bx_0,\xi_0) = \xi_0.
+ = \partial_x\zeta_i(z,\bx_i,\xi_i),\, \xi_i(0;\bx_0,\xi_0) = \xi_0.
 \end{equation}
-and
+
+From the relation \ref{eqn:hsol}, 
 \[
@@
   \label{eqn:psihe}
-- \int_0^z ds \xi(s;\bx_0,\xi_0)\cdot \partial_{\xi}\zeta_i(\bx_i(s;\bx_0,\xi_0),s,\xi_i(s;\bx_0,\xi_0))
++ \int_{z_s}^z ds [\zeta(\bx_i(s;\bx_0,\xi_0),s, \xi(s;\bx_0,\xi_0)) - \xi(s;\bx_0,\xi_0)\cdot \partial_{\xi}\zeta_i(\bx_i(s;\bx_0,\xi_0),s,\xi_i(s;\bx_0,\xi_0))]
 \end{equation}
 
-$\Pi_0$ = projection on null space $V_0$ of $M(t,\bx)$, $\Pi_1=I-\Pi_0$ = projection on range $V_1$. Denote the rank of $M$ by $l={\rm dim} V_1$ Then
+As previously noted, $\zeta_i \ne 0$ if $\Sigma$ is chosen
+sufficiently small. Enumerate the $\zeta$s so that $zeta_1,.../\zeta_m
+< 0, \zeta_{m+1},...,\zeta_l > 0$. Then from the eikonal equation
+\ref{eqn:eik}, the time $t$ along equal phase surfaces $t +
+\psi_i(z,\bx_i(z,\bx_0,\xi_0),\xi_i(z,\bx_0,\xi_0)) = \mbox{ const}$
+increases for $i=1,...,m$, decreases for $i=m+1,...,l$.
 
-\section{Symmetric Hyperbolic Systems}
-
-If $A_j$ is positive definite, then $x_j$ can play the role of time, and the Cauchy problem with initial data $u_0 \in L^2(\bR^n,\bR^k)$  on $x_j=$ constant has a unique weak solution $u \in L^2(\bR^{n+1})^k$ with well-defined $L^2$ traces on any $x_j$ level set (Lax, Taylor).
-Want to understand traces on other hypersurfaces, to begin with coordinate hypersurfaces, wlog $x_0=0$.
-
-Notation:$x_0=t$, $\partial_t=\partial_0$, $\bx = (x_1,...,x_n)$, $D=(\partial_1,...\partial_n)$
+% Suppose that $supp \chi \subset \{(\bx,t): t > 0\}$.
+So much for the first term in the RHS of equation
+\ref{eqn:series}. The second and third terms are of order zero in
+$i\omega$, hence should cancel. However the matrix in the third term
+is not invertible in view of the relation \ref{eqn:nullsp}. Therefore
+the second term must be orthogonal to the kernel of this matrix, which
+since the matrix is symmetric simply means in its orthogonal to $U^0$.
+That is,
 \[
-  K(t,\bx,D)= \sum_{i=1}^n A_i(t,\bx)\partial_i, \, M(t,\bx) = A_0(t,\bx).
+U^0(z,\bx) \cdot (M(z,\bx) U^0(z,\bx) \partial_t a^0(t,z,\bx) + A(z,\bx)\cdot \partial_x
+(a^0(t,z,\bx) U^0(z,\bx))
 \]
-If $M$ is positive definite, trace properties coverd by standard theory (above).
-Let $\Pi_0(t,\bx)$ = projection on null space $V_0$ of $M(t,\bx)$, $\Pi_1=I-\Pi_0$ = projection on range $V_1$.
-Denote the rank of $M$ by $l={\rm dim} V_1$ Then
+\[+ L(z,\bx)\partial_z
+  (a^0(t,z,\bx)U^0(z,\bx))) = 0,
+\]
+or
 \[
-  M\partial_t u + Ku = \Pi_1 M \partial_t \Pi_1 u + \sum_{\alpha,\beta=0,1} \Pi_{\alpha}K\Pi_{\beta}u + \mbox{ l.o.t. }
+ (b_t \partial_t + b_x \cdot \partial_x + b_z \partial_z)
+ a^0 + c a^0 = 0
 \]
-Since the ranges are orthogonal,
-\begin{eqnarray}
-  \Pi_1 M \partial_t \Pi_1 u + \Pi_1 K \Pi_1 u + \Pi_1 K \Pi_0 u & \in & L^2_{\rm loc}(\bR^{n+1},V_1),
-  \label{eqn:decomp1} \\
-  \Pi_0 K \Pi_0 u + \Pi_0 K \Pi_1 u & \in & L^2_{\rm loc}(\bR^{n+1},V_0)
-  \label{eqn:decomp0}
-\end{eqnarray}
-Set $K_{\alpha,\beta} = \Pi_{\alpha}K\Pi_{\beta}$.
-
-Main assumption: for all $t \in \bR, (\bx,\xi) \in T^*(\bR^n)\setminus \{0\}$, $K_{0,0}(t,\bx,\xi)$ is invertible on $V_0$.
-
-% Suppose that $E \in OPS^1(\bR^{n+1})$ is scalar, $Eu \in H^1_{\rm loc}(\bR^{n+1})$.
-Denote by $i(t):\{t\} \times \bR^n \rightarrow \bR^{n+1}$ the embedding.
-contained in $i(0)^*(ES(E))$, and that and
-
-% Suppose that $\Gamma_0, \Gamma \subset T^*(\bR^n)$ are conic open sets with $\overbar{\Gamma_0 \subset \Gamma$ and $\delta t >0$.
-Assume that  that for $|t|\le \delta t$ and $(\bx,\xi) \in Gamma$, $K_{00}(0,\bx,\xi)$ is invertible as a map $V_0 \rightarrow V_0$.
-Let  $L_{00}(t,\bx,\xi) \in C^{\infty}(\bR, S^{-1}(\bR^n))$ be inverse to $K_{0,0}(t,\bx,\xi)$ on $V_0$: that is,
+where
 \[
-  K_{0,0}(t,\bx,\xi)L_{0,0}(t,\bx,\xi) v=v=L_{0,0}(t,\bx,\xi)K_{0,0}(t,\bx,\xi)v
+  b_t = U^0 \cdot M U^0, b_{x,i} = U^0 \cdot A_i U^0, b_z = U^0 \cdot
+  L U_0
 \]
-for all $v \in V_0$.
-Construct $L_{0,0}$ to be homogeneous of degree -1 outside of a neighborhood of $\xi=0$, with $V_1$ an invariant subspace for $L_{0,0}(t,\bx,\xi)$ on which it vanishes.
-It follows that $L_{0,0}$ is symmetric, and that
-\[
-  \Pi_0 u - L_{0,0}K_{0,0}\Pi_u = \Pi_0 u - L_{0,0}K_{0,0}u = \Pi_0 u + L_{0,0}K_{0,1}u \in H^{1}_{\rm loc}(\bR^{n+1})^k
-\]
-so
-\[
-  K_{1,0}u + K_{1,0}L_{0,0}K_{0,1}u \in L^2 _{\rm loc}(\bR^{n+1})^k
-\]
-Define
-\[
-  \tilde{K} = K_{11} - K_{10}L_{00}K_{01}.
-\]
-Then equation \ref{eqn:decomp1} can be rewritten
-\begin{equation}
-  \label{eqn:shs1}
-  M\partial_t u_1 + \tilde{K}u_1 \in L^2_{\rm loc}(\bR^{n+1},V_1)
-\end{equation}
-Note that for each $t \in \bR, (\bx,\xi) \in T^*(\bR^n)$, both $V_0$ and $V_1$ are invariant subspaces of $\tilde{K}(t,\bx,\xi)$, and $V_0 \subset {\rm ker}\tilde{K}(t,\bx,\xi)$.
-Also, $M$ is invertible when restricted to $V_1 = {\rm Rng} M = {\rm Rng } \Pi_1$ , but not (necessarily) positive definite.
-
-Restricted determinant: if $V$ is an invariant subspace of $A$, then ${\rm det}_{V} A$ is the determinant of the matrix of $A$ in any orthonormal basis of $V$.
-For $(t,\bx,\tau,\xi) \in T^*(\bR^n)$, define the restricted characteristic polynomial
-\[
-  {\cal C}(t,\bx,\tau,\xi) = {\rm det}_{V_1} (M(t,\bx)\tau + \tilde{K}(t,\bx,\xi) )= 0
-\]
-Since $V_0$ and $V_1$ are invariant subspaces of the matrix in parentheses, and $V_0$ is (contained in) its null space, there are at most $l={\rm dim}V_1$ non-zero roots in $\tau$ for $\xi \ne 0$.
-
-Choose a conic submanifold $\gamma \subset T^*(\bR^n)$, so that for $|t| \le \delta t, (\bx,\xi) \in \gamma$ there are exactly $m \le l$ distinct real roots $\tau$ of ${\cal C}(t,\bx,\tau,\xi)$.
-Denote these by $\tau_i(t,\bx,\xi), i=1,...,m$.
-
-Choose a conic $\gamma_1 \subset T^*(\bR^n)$ so that $\bar{\gamma_1} \subset \gamma$, and $Q \in OPS^0(\bR^n)$ with $ES(Q) \subset \gamma$, $Q(\bx,\xi) =1$ for $(\bx,\xi)\in \gamma_1$.
-
-Let ${\cal N} = $ normal bundle of $x_0=0$ and $\Gamma \subset T^*(\bR^n)$ for which $\Gamma \cap {\cal N} = \emptyset$.
-
-Assume that $u \in L^2(\bR^{n+1}) \cap H^1_{\rm loc}(\Gamma \cap \Pi_{\cal N}^{-1}(\gamma_1^C))$.
-
-For each $\tau_i, i =1,...m$, choose $U_i(t,\bx,\xi) \in C^{\infty}(\bR \times T^*(\bR^n),V_1(t,\bx))$,  so that
-\[
-  \tau_i M(\bx) U_i(t,\bx,\xi) + \tilde{K}(t,\bx,\xi) U_i(t,\bx,\xi)=0, \, i=1,..,m.
-\]
-$U_1,...,U_m$ span a subspace $\tilde{V}_1(t,\bx,\xi)$. Scale $U_i$ so that $|U_i(t,\bx,\xi)| = 1$.
-
-Note that $\tau_i(t,\bx,\xi), i=1,...,m$ are homogeneous of order 1 in $\xi$ and smooth in $\bx$, $U_i(t,\bx,\xi), i=1,...,m$ are homogeneous of order 0 in $\xi$, smooth in $\bx$.
-So both are symbol-valued functions, of orders 1 and 0 respectively.
-Since the $U$'s must be orthogonal, $U_i^TU_j = \delta_{ij}$.
-
-Suppose first that $m=l$, that is, that $\tilde{V}_1=V_1$.
-Then $\{U_i(t,\bx,\xi):i=1,..,l\}$ is a basis for $V_1(t,\bx)$ for each $(t,\bx,\xi \in \Gamma$.
-Set $u_{1,i}(t,\bx) = U_i(t,\bx,D_x)^Tu_1$.
-Then for any $w \in L^2(\bR^{n+1},V_1)$
-\[
-  \sum_{i=1}^l U_i(t,\bx,D)U_i(t,\bx,D)^Tw(t,\bx) \approx
-\]
-\[
-  \int \, d\xi \, e^{i \bx \cdot \xi} U_i(t,\bx,\xi)U_i(t,\bx,\xi)^T\hat{w}(t,\xi) \approx
-\]
-\[
-  \int \, d\xi \, e^{i \bx \cdot \xi}
-\]
-\[
-  \tilde{K}(t,\bx,D)u_{1,i} = \tilde{K}(t,\bx,D)U_i(t,\bx,D)^T
-\]
-\[
-  M(\bx)\partial_t u_{1,i}(t,\bx) = -\tilde{K}(t,\bx,D)
-\]
+and $c$ is the rest, not involving derivatives of $a^0$.
 
@@
 \]
-Therefore take $\Gamma = \{\xi: |\xi| \lt \mbox{inf
-}\{\sqrt{\rho(\bx,z)/kappa(\bx,z)}, \mbox{ all } |z| < Z\}$. For $\xi
-\in \Gamma$, there are two roots $\zeta = \pm \sqrt{\rho/\kappa - \xi^2}.
-
-So $V_0={\rm span}(\{e2\}),\, V_1={\rm span}(\{e1,e3\}), \Pi_0 = {\rm diag}(0,1,0)$ 
+Therefore choose $\epsilon > 0$ and define $\Gamma = \{(\bx,\xi): |\xi| < \mbox{ inf}\{\sqrt{\rho(\bx,0)/\kappa(\bx,0)}, \mbox{ all } |z| < Z\}$. For $\xi
+\in \Gamma$, there are two roots $\zeta_{\pm} = \pm \sqrt{\rho/\kappa - \xi^2}.$
+Accordingly there are two solutions of equation \ref{eqn:eik},
 \[
-  K(t,\bx,\xi) =  A_1(\bx) \xi_1 + A_2(\bx) \xi_2 =
-  \left(
-    \begin{array}{ccc}
-      \frac{1}{\kappa(\bx)}\omega & \xi_x & 0 \\
-      \xi_x & \rho(\bx) \omega & 0 \\
-      0 & 0 & \rho(\bx) \omega
-    \end{array}
-  \right)
-\]
-whence $K_{0,0} = {\rm diag}(0, \rho(\bx) \omega, 0)$.
-\[
-  K_{1,1}(t,\bx,\xi) =
-  \left(
-    \begin{array}{ccc}
-      \frac{1}{\kappa(\bx)}\omega & 0 & 0 \\
-      0 & 0 & 0 \\
-      0 & 0 & \rho(\bx) \omega
-    \end{array}
-  \right),
+  \psi_{\pm}(\bx_{\pm}(z;\bx_0,\xi_0),z,\xi_{\pm}(z;\bx_0,\xi_0))= \bx_0\cdot\xi_0
 \]
 \[
-  K_{0,1}(t,\bx,\xi) =
-  \left(
-    \begin{array}{ccc}
-      0 & \xi_x & 0 \\
-      0 & 0 & 0 \\
-      0 & 0 & 0
-    \end{array}
-  \right)
-\]
-and $K_{1,0} = K_{0,1}^T$.
-
-Choose $\omega_0>0$, $\phi \in C^{\infty}(\bR)$ so that $\phi(\omega) = 1$ for $\omega> 2\omega_0$, $=0$ for $\omega < \omega_0$. Set
-\[
-  L_{0,0}= {\rm diag}\left(0, \frac{\phi(|\omega|)}{\rho(\bx) \omega}, 0\right)
-\]
-Then $L_{0,0}K_{0,0} = I$ restricted to $V_0$ and away from zero frequency, and
-\[
-  \tilde{K} = K_{1,1} - K_{1,0}L_{0,0}K_{0,1} =
-  \left(
-    \begin{array}{ccc}
-      \frac{\omega}{\kappa} - \frac{\xi_x^2\phi(|\omega|)}{\rho \omega} & 0 & 0\\
-      0 & 0  & 0 \\
-      0 & 0 & \rho \omega
-    \end{array}
-  \right).
-\]
-Since $V_1={\rm span}(e_1,e_3)$, the determinant of the restricted characteristic polynomial is obtained by dropping the second row and column:
-\[
-  \det
-  \left(
-    \begin{array}{cc}
-      \frac{\omega}{\kappa} - \frac{\xi_x^2\phi(|\omega|)}{\rho \omega} &  \xi_z\\
-      \xi_z &  \rho \omega
-    \end{array}
-  \right).
+ +\int_0^z ds \zeta_{\pm}(\bx_{\pm}(s;\bx_0,\xi_0),s,\xi_{\pm}(s;\bx_0,\xi_0))-\xi_{\pm}(s;\bx_0,\xi_0)\cdot \partial_{\xi}\zeta_{\pm}(\bx_{\pm}(s;\bx_0,\xi_0),s,\xi_{\pm}(s;\bx_0,\xi_0))
 \]
 \[
-  = \frac{\rho \omega^2 }{\kappa} - \xi_x^2\phi(|\omega|) - \xi_z^2
+  =\bx_0 \cdot \xi_0 \pm \int_0^z
+  \frac{ds}{c(\bx_{\pm}(s;\bx_0,\xi_0),z)
+    \sqrt{(1-\xi_{\pm}(s;\bx_0,\xi_0),z)^2
+    c(\bx_{\pm}(s;\bx_0,\xi_0),z)^2}}
 \]
-\[
-  = \frac{\rho \omega^2 }{\kappa} - \xi_x^2 - \xi_z^2 \mbox{ mod }  S^{-\infty}
-\]  
-as expected. So a maximal choice for $\Gamma$ would be
-\[
-  \Gamma = \{(z,(t,x),(\omega,\xi_x)): c(z,x)|\xi_x| < |\omega|\}.
-\]
+where $c = \sqrt{\kappa/\rho}$ is the sound velocity.
+
 \section{2D Elasticity}
@@
 \subsection{PsiDO RHS}
-Suppose that $K \in OPS^1$ is skew-adjoint, at least in principal part: $K_1^*=-K_1$.
-Then there is unitary $U \in OPS^0$ so that $U(x,\xi)^*K_1(x,\xi)U(x,\xi) = i\Lambda(x,\xi) = i\mbox{diag}(\lambda_1(x,\xi),...\lambda_k(x,\xi))$, with scalar $\lambda_j \in OPS^1, j=1,...,k$ (in fact, homogeneous of degree 1).
+Suppose that $K \in OPS^1$ is skew-adjoint, at least in principal
+part: $K_1^*=-K_1$. Then there is unitary $U \in OPS^0$ so that
+$U(x,\xi)^*K_1(x,\xi)U(x,\xi) = i\Lambda(x,\xi) =
+i\mbox{diag}(\lambda_1(x,\xi),...\lambda_k(x,\xi))$, with scalar
+$\lambda_j \in OPS^1, j=1,...,k$ (in fact, homogeneous of degree
+1). Also,
+\begin{equation}
+  \label{eqn:ortho}
+  U(x,D)^*U(x,D) = U(x,d)U(x,D)^* = I \mbox{ mod } OPS^{-1}
+\end{equation}
+Define $S(t)$ to be the one-parameter group satisfying
+\[
+  \partial_t S(t) = -K S(t), S(0)=I
+\]
+Note that $S(-t)$ is inverse to $S(t)$, and 
+\[
+  \partial_t S(-t) = -S(-t) \partial_t S(t) S(-t) = S(-t)K
+\]
 Hence
@@
 Let $\tilde{S}_i$ solve
-\[
+\begin{equation}
+  \label{eqn:stilde}
   \partial_t \tilde{S}_i(t) = - i\lambda_i \tilde{S}_i(t)
-\]
+\end{equation}
 Then
@@
 \]
+and
+\begin{equation}
+  \label{eqn:spec}
+  S(t) = \sum_{i=1}^k U_i \tilde{S}_i(t)U_i^*
+\end{equation}
 
-From equation \ref{eqn:qevol1},
+Define
 \[
-  \tilde{Q}_i(t) = \tilde{S}_i(-t)\delta \tilde{S}_i(t)
+  F = \phi_r S(T) \phi_s
 \]
-satisfies
+in which $\phi_,r \in C_0^{\infty}(\bR^{n+1})$ and $T>0$. Suppose that $K$
+is perturbed by $\delta K \in OPS^1$, also essentially skew. Then
+$\Lambda$, $U$, $S$, $\tilde{S}$ and $F$ undergo perturbations
+$\delta \Lambda$, $\delta U$, $\delta S$, $\delta \tilde{S}$, and
+$\delta F$. Our object is to understand the structure of $\delta F$, under the
+assumption that $\phi_r \delta K \in OPS^{-\infty}$.
+
+This assumption on $\delta K$ implies that $\phi_r \delta U \in
+OPS^{-\infty}$. 
 \[
-  \partial \tilde{Q}_i(t) = -i\tilde{S}_i(-t)\delta \lambda_i \tilde{S}_i(t), \, i=1,,,.k.
+  \delta F =  \phi_r \delta S(T)  = \phi_r \left(\sum _{j=1}^k \delta U_j \tilde{S}_j(T) U_j^*
+\right.
 \]
-and $\tilde{Q} = \mbox{diag}(\tilde{Q}_1,...,\tilde{Q}_k)$.
-From \cite{Tay:81}, Egorov's Theorem (Section VIII.1),
 \[
-  \tilde{Q}_i(t; x,\xi) = -i \int_0^t ds \,\delta \lambda_i(y_i(s;x,\xi),\eta_i(s;x,\xi))
+ +  \left. \sum_{j=1}^k U_j \delta\tilde{S}_j(T)U_j^* + \sum_{J=1}^k U_j
+    \tilde{S}_j(T)\delta U_j^*\right) 
 \]
-in which $(y_i(s;x,\xi),\eta_i(s;x,\xi))$ is the integral curve of the Hamiltonian vector field
+Every term in the first sum on the RHS is bounded $H^s \rightarrow H^{s-1}$ since
+$\phi_r \delta U_i \in OPS^{-\infty}$.  Since $[\phi_rU_i], \in OPS^{-1}$, 
+\[
+  \phi_r  U_j \delta \tilde{S}_j(T)  U_j^*  = U_j \phi_r
+  \tilde{S}_j(T) U_j^*
+  +...
+\]
+where the elided terms are operators mapping $H^s$ to $H^{s-1}$, any
+$s$. So,
+\[
+ \delta F = \sum_{j=1}^k U_j \phi_r \left( \delta\tilde{S}_j(T) U_i^*
+   + \tilde{S}_j(T) \delta U_j^*\right)
+  + ...
+\]
+and
+\[
+ S(-T)\delta F =\left(\sum_{i=1}^k U_i \tilde{S}_i(T)
+    U_i^*\right) \left(\sum_{j=1}^k U_j \phi_r ( \delta\tilde{S}_j(T) U_i^*
+    + \tilde{S}_j(T) \delta U_j^*)\right)
+\]  
+Using the relation \ref{eqn:ortho},
+The first order perturbation of the definition \ref{eqn:stilde} is
+\begin{equation}
+  \label{eqn:dstilde}
+  \partial_t \delta \tilde{S}_i(t) = - i\lambda_i \delta
+  \tilde{S}_i(t) - i\delta \lambda_i \tilde{S}_i(t)
+\end{equation}
+Since $\tilde{S_i}(-t)$ is inverse to $\tilde{S}_i(t)$,
+\[
+  \partial_t \tilde{S}_i(-t)= -\tilde{S}_i(-t) (\partial_t \tilde{S}_i(t)) \tilde{S}_i(-t)
+\]
+\begin{equation}
+  \label{eqn:stildeinv}
+= -\tilde{S}_i(-t)(-i\lambda_i \tilde{S}_i(t)) \tilde{S}_i(-t) = i\tilde{S}_i(-t)\lambda_i 
+\end{equation}
+So
+%\[
+% \partial_t (\tilde{S}_i(-t) \phi_r \delta \tilde{S}_i(t) \phi_s) 
+%\]
+%\[
+%  = i\tilde{S}_i(-t)\lambda_i \phi_r \delta \tilde{S}_i(t) \phi_s +
+%  \tilde{S}_i(-t) \phi_r (- i\lambda_i \delta
+%  \tilde{S}_i(t) - i\delta \lambda_i \tilde{S}_i(t)) \phi_s
+%\]
+%\[
+%  = -i\tilde{S}_i(-t)[\lambda_i,\phi_r] \delta \tilde{S}(t) \phi_s -i
+%  \tilde{S}_i(-t)\phi_r\delta\lambda_i\tilde{S}_i(t) \phi_s
+%\]
+%\begin{equation}
+%  \label{eqn:qevol1}
+%  = -i\tilde{S}_i(-t)[\lambda_i,\phi_r]\tilde{S}_i(t) \tilde{S}_i(-t)\delta \tilde{S}(t) \phi_s %-i
+%  \tilde{S}_i(-t)\phi_r\delta\lambda_i\tilde{S}_i(t) \phi_s.
+%\end{equation}
+%Similarly,
+\begin{equation}
+  \label{eqn:dstildeconj}
+  \partial_t\tilde{S}_i(-t)\delta \tilde{S}_i(t) = -i\tilde{S}_i(-t)\delta \lambda_i \tilde{S}_i(t).
+\end{equation}
+From \cite{Tay:81}, Egorov's Theorem (Section VIII.1), for $p \in OPS^s(\bR^n)$,
+\begin{equation}
+  \label{eqn:egorov}
+   \tilde{S}_i(-t) p \tilde{S}_i(t)=  p \circ C_i(t),\, p\tilde{S}_i(t) =
+   \tilde{S}_i(t)p \circ C_i(t)
+\end{equation}
+in which $C_i$ is the one-paraimeter family of symplectic maps given by
+\begin{equation}
+  \label{eqn:canon}
+  C_i(t)(x,\xi)=(y_i(t;x,\xi),\eta_i(t;x,\xi))
+\end{equation}
+Here $t \rightarrow y_i(t;x,\xi),\eta_i(t;x,\xi) $is the integral curve of the
+Hamiltonian vector field
 \[
@@
 with $y_i(0;x,\xi)=x, \eta_i(0;x,\xi)=\xi$.
+
+For $p \in OPS^{\infty}$, define
+\[
+  \Gamma_i[p] = \int_0^T dt \, p \circ C_i(t)
+\]
+Then from the identity \ref{eqn:dstildeconj} follows
+\[
+  \delta\tilde{S}_i(T) = \tilde{S}_i(T) \Gamma_i[ -i \delta \lambda_i]
+\] 
+so
+\[
+  \delta F = \sum_{i=1}^k U_i \phi_r \delta \tilde{S}_i \phi_s U_i^*
+\]
+\[
+  =\sum_{i=1}^k U_i \phi_r \tilde{S}_i(T) \Gamma_i[ -i \delta \lambda_i]\phi_s U_i^*
+\]
+\[
+ Q = -i\sum_{i=1}^k U_i \tilde{S}_i(-T) \phi_r \tilde{S}_i(T) \Gamma_i[\delta \lambda_i] U_i^*\phi_s
+  + ...
+\]
+\[
+  = -i \sum_{i=1}^k U_i (\Gamma_i[\phi_r] \Gamma_i[\delta \lambda_i] )
+  U_i^*
+\]
+
+The right hand side of equation \ref{eqn:qevol1} contains three
+expressions of the form appearing on the LHS of \ref{eqn:egorov}.
 
@@
 \end{equation}
-Replace $x$ by $(t,x)$, $F(x,\partial_x u)$ by the RHS of equation
-\ref{eqn:hjdef}.  Then precisely the same results obtain, with $F$ is
-replaced by $H$.
+Set $X=(t,x)$, $P=(q,p)$ (with $q$ dual to $t$, $F(X,\partial_X u) =
+\partial_t u + H(t,x,\partial_x u)$, so the Hamiltonian is $F(X,P) =
+q+H(t,x.p)$.  Thus
+\begin{eqnarray*}
+  \frac{dt}{ds} & = & \partial_q F = 1 \\
+  \frac{dx}{ds} & = & \partial_p F = \partial_p H \\
+  \frac{dq}{ds} & = & -\partial_t F = -\partial_t H \\
+  \frac{dp}{ds} & = & -\partial_x F = -\partial_x H
+\end{eqnarray*}
+From the first equation, $t=s$ up to a constant shift. From the third,
+\[
+  q(t) = -\int_0^t ds \partial_tH(s,x(s),p(s))
+\]
+\[
+  = -\int_0^t ds \frac{d}{ds}H(s,x(s),p(s)) - \frac{dx}{ds}(s)\cdot
+  \partial_x H(s,x(s),p(s)) -\frac{dp}{ds}(s)\cdot \partial_p
+  H(s,x(s),p(s))
+\]
+\[
+  = -H(t,x(t),p(t)) + H(0,x(0),p(0))
+\]
+in view of the 2nd and 4th equations.
+Note that these are the Hamiltonian system with Hamiltonian $H$.
+
+The result \ref{eqn:sol} becomes
+\begin{equation}
+  \label{eqn:hsol}
+  u(t,x(t))=u(0,x(0)) +tH(0,x(0),p(0))+ \int_0^t ds (-H(s,x(s),p(s)) + p(s)\cdot (\partial_p H)(s,x(s),p(s))).
+\end{equation}
+
+\section{Symmetric Hyperbolic Systems}
+
+If $A_j$ is positive definite, then $x_j$ can play the role of time, and the Cauchy problem with initial data $u_0 \in L^2(\bR^n,\bR^k)$  on $x_j=$ constant has a unique weak solution $u \in L^2(\bR^{n+1})^k$ with well-defined $L^2$ traces on any $x_j$ level set (Lax, Taylor).
+Want to understand traces on other hypersurfaces, to begin with coordinate hypersurfaces, wlog $x_0=0$.
+
+Notation:$x_0=t$, $\partial_t=\partial_0$, $\bx = (x_1,...,x_n)$, $D=(\partial_1,...\partial_n)$
+\[
+  K(t,\bx,D)= \sum_{i=1}^n A_i(t,\bx)\partial_i, \, M(t,\bx) = A_0(t,\bx).
+\]
+If $M$ is positive definite, trace properties coverd by standard theory (above).
+Let $\Pi_0(t,\bx)$ = projection on null space $V_0$ of $M(t,\bx)$, $\Pi_1=I-\Pi_0$ = projection on range $V_1$.
+Denote the rank of $M$ by $l={\rm dim} V_1$ Then
+\[
+  M\partial_t u + Ku = \Pi_1 M \partial_t \Pi_1 u + \sum_{\alpha,\beta=0,1} \Pi_{\alpha}K\Pi_{\beta}u + \mbox{ l.o.t. }
+\]
+Since the ranges are orthogonal,
+\begin{eqnarray}
+  \Pi_1 M \partial_t \Pi_1 u + \Pi_1 K \Pi_1 u + \Pi_1 K \Pi_0 u & \in & L^2_{\rm loc}(\bR^{n+1},V_1),
+  \label{eqn:decomp1} \\
+  \Pi_0 K \Pi_0 u + \Pi_0 K \Pi_1 u & \in & L^2_{\rm loc}(\bR^{n+1},V_0)
+  \label{eqn:decomp0}
+\end{eqnarray}
+Set $K_{\alpha,\beta} = \Pi_{\alpha}K\Pi_{\beta}$.
+
+Main assumption: for all $t \in \bR, (\bx,\xi) \in T^*(\bR^n)\setminus \{0\}$, $K_{0,0}(t,\bx,\xi)$ is invertible on $V_0$.
+
+% Suppose that $E \in OPS^1(\bR^{n+1})$ is scalar, $Eu \in H^1_{\rm loc}(\bR^{n+1})$.
+Denote by $i(t):\{t\} \times \bR^n \rightarrow \bR^{n+1}$ the embedding.
+contained in $i(0)^*(ES(E))$, and that and
+
+% Suppose that $\Gamma_0, \Gamma \subset T^*(\bR^n)$ are conic open sets with $\overbar{\Gamma_0 \subset \Gamma$ and $\delta t >0$.
+Assume that  that for $|t|\le \delta t$ and $(\bx,\xi) \in Gamma$, $K_{00}(0,\bx,\xi)$ is invertible as a map $V_0 \rightarrow V_0$.
+Let  $L_{00}(t,\bx,\xi) \in C^{\infty}(\bR, S^{-1}(\bR^n))$ be inverse to $K_{0,0}(t,\bx,\xi)$ on $V_0$: that is,
+\[
+  K_{0,0}(t,\bx,\xi)L_{0,0}(t,\bx,\xi) v=v=L_{0,0}(t,\bx,\xi)K_{0,0}(t,\bx,\xi)v
+\]
+for all $v \in V_0$.
+Construct $L_{0,0}$ to be homogeneous of degree -1 outside of a neighborhood of $\xi=0$, with $V_1$ an invariant subspace for $L_{0,0}(t,\bx,\xi)$ on which it vanishes.
+It follows that $L_{0,0}$ is symmetric, and that
+\[
+  \Pi_0 u - L_{0,0}K_{0,0}\Pi_u = \Pi_0 u - L_{0,0}K_{0,0}u = \Pi_0 u + L_{0,0}K_{0,1}u \in H^{1}_{\rm loc}(\bR^{n+1})^k
+\]
+so
+\[
+  K_{1,0}u + K_{1,0}L_{0,0}K_{0,1}u \in L^2 _{\rm loc}(\bR^{n+1})^k
+\]
+Define
+\[
+  \tilde{K} = K_{11} - K_{10}L_{00}K_{01}.
+\]
+Then equation \ref{eqn:decomp1} can be rewritten
+\begin{equation}
+  \label{eqn:shs1}
+  M\partial_t u_1 + \tilde{K}u_1 \in L^2_{\rm loc}(\bR^{n+1},V_1)
+\end{equation}
+
+Note that for each $t \in \bR, (\bx,\xi) \in T^*(\bR^n)$, both $V_0$ and $V_1$ are invariant subspaces of $\tilde{K}(t,\bx,\xi)$, and $V_0 \subset {\rm ker}\tilde{K}(t,\bx,\xi)$.
+Also, $M$ is invertible when restricted to $V_1 = {\rm Rng} M = {\rm Rng } \Pi_1$ , but not (necessarily) positive definite.
+
+Restricted determinant: if $V$ is an invariant subspace of $A$, then ${\rm det}_{V} A$ is the determinant of the matrix of $A$ in any orthonormal basis of $V$.
+For $(t,\bx,\tau,\xi) \in T^*(\bR^n)$, define the restricted characteristic polynomial
+\[
+  {\cal C}(t,\bx,\tau,\xi) = {\rm det}_{V_1} (M(t,\bx)\tau + \tilde{K}(t,\bx,\xi) )= 0
+\]
+Since $V_0$ and $V_1$ are invariant subspaces of the matrix in parentheses, and $V_0$ is (contained in) its null space, there are at most $l={\rm dim}V_1$ non-zero roots in $\tau$ for $\xi \ne 0$.
+
+Choose a conic submanifold $\gamma \subset T^*(\bR^n)$, so that for $|t| \le \delta t, (\bx,\xi) \in \gamma$ there are exactly $m \le l$ distinct real roots $\tau$ of ${\cal C}(t,\bx,\tau,\xi)$.
+Denote these by $\tau_i(t,\bx,\xi), i=1,...,m$.
+
+Choose a conic $\gamma_1 \subset T^*(\bR^n)$ so that $\bar{\gamma_1} \subset \gamma$, and $Q \in OPS^0(\bR^n)$ with $ES(Q) \subset \gamma$, $Q(\bx,\xi) =1$ for $(\bx,\xi)\in \gamma_1$.
+
+Let ${\cal N} = $ normal bundle of $x_0=0$ and $\Gamma \subset T^*(\bR^n)$ for which $\Gamma \cap {\cal N} = \emptyset$.
+
+Assume that $u \in L^2(\bR^{n+1}) \cap H^1_{\rm loc}(\Gamma \cap \Pi_{\cal N}^{-1}(\gamma_1^C))$.
+
+For each $\tau_i, i =1,...m$, choose $U_i(t,\bx,\xi) \in C^{\infty}(\bR \times T^*(\bR^n),V_1(t,\bx))$,  so that
+\[
+  \tau_i M(\bx) U_i(t,\bx,\xi) + \tilde{K}(t,\bx,\xi) U_i(t,\bx,\xi)=0, \, i=1,..,m.
+\]
+$U_1,...,U_m$ span a subspace $\tilde{V}_1(t,\bx,\xi)$. Scale $U_i$ so that $|U_i(t,\bx,\xi)| = 1$.
+
+Note that $\tau_i(t,\bx,\xi), i=1,...,m$ are homogeneous of order 1 in $\xi$ and smooth in $\bx$, $U_i(t,\bx,\xi), i=1,...,m$ are homogeneous of order 0 in $\xi$, smooth in $\bx$.
+So both are symbol-valued functions, of orders 1 and 0 respectively.
+Since the $U$'s must be orthogonal, $U_i^TU_j = \delta_{ij}$.
+
+Suppose first that $m=l$, that is, that $\tilde{V}_1=V_1$.
+Then $\{U_i(t,\bx,\xi):i=1,..,l\}$ is a basis for $V_1(t,\bx)$ for each $(t,\bx,\xi \in \Gamma$.
+Set $u_{1,i}(t,\bx) = U_i(t,\bx,D_x)^Tu_1$.
+Then for any $w \in L^2(\bR^{n+1},V_1)$
+\[
+  \sum_{i=1}^l U_i(t,\bx,D)U_i(t,\bx,D)^Tw(t,\bx) \approx
+\]
+\[
+  \int \, d\xi \, e^{i \bx \cdot \xi} U_i(t,\bx,\xi)U_i(t,\bx,\xi)^T\hat{w}(t,\xi) \approx
+\]
+\[
+  \int \, d\xi \, e^{i \bx \cdot \xi}
+\]
+\[
+  \tilde{K}(t,\bx,D)u_{1,i} = \tilde{K}(t,\bx,D)U_i(t,\bx,D)^T
+\]
+\[
+  M(\bx)\partial_t u_{1,i}(t,\bx) = -\tilde{K}(t,\bx,D)
+\]
+
+\subsection{2D Acoustics}
+%###################
+So $V_0={\rm span}(\{e2\}),\, V_1={\rm span}(\{e1,e3\}), \Pi_0 = {\rm diag}(0,1,0)$ 
+\[
+  K(t,\bx,\xi) =  A_1(\bx) \xi_1 + A_2(\bx) \xi_2 =
+  \left(
+    \begin{array}{ccc}
+      \frac{1}{\kappa(\bx)}\omega & \xi_x & 0 \\
+      \xi_x & \rho(\bx) \omega & 0 \\
+      0 & 0 & \rho(\bx) \omega
+    \end{array}
+  \right)
+\]
+whence $K_{0,0} = {\rm diag}(0, \rho(\bx) \omega, 0)$.
+\[
+  K_{1,1}(t,\bx,\xi) =
+  \left(
+    \begin{array}{ccc}
+      \frac{1}{\kappa(\bx)}\omega & 0 & 0 \\
+      0 & 0 & 0 \\
+      0 & 0 & \rho(\bx) \omega
+    \end{array}
+  \right),
+\]
+\[
+  K_{0,1}(t,\bx,\xi) =
+  \left(
+    \begin{array}{ccc}
+      0 & \xi_x & 0 \\
+      0 & 0 & 0 \\
+      0 & 0 & 0
+    \end{array}
+  \right)
+\]
+and $K_{1,0} = K_{0,1}^T$.
+
+Choose $\omega_0>0$, $\phi \in C^{\infty}(\bR)$ so that $\phi(\omega) = 1$ for $\omega> 2\omega_0$, $=0$ for $\omega < \omega_0$. Set
+\[
+  L_{0,0}= {\rm diag}\left(0, \frac{\phi(|\omega|)}{\rho(\bx) \omega}, 0\right)
+\]
+Then $L_{0,0}K_{0,0} = I$ restricted to $V_0$ and away from zero frequency, and
+\[
+  \tilde{K} = K_{1,1} - K_{1,0}L_{0,0}K_{0,1} =
+  \left(
+    \begin{array}{ccc}
+      \frac{\omega}{\kappa} - \frac{\xi_x^2\phi(|\omega|)}{\rho \omega} & 0 & 0\\
+      0 & 0  & 0 \\
+      0 & 0 & \rho \omega
+    \end{array}
+  \right).
+\]
+Since $V_1={\rm span}(e_1,e_3)$, the determinant of the restricted characteristic polynomial is obtained by dropping the second row and column:
+\[
+  \det
+  \left(
+    \begin{array}{cc}
+      \frac{\omega}{\kappa} - \frac{\xi_x^2\phi(|\omega|)}{\rho \omega} &  \xi_z\\
+      \xi_z &  \rho \omega
+    \end{array}
+  \right).
+\]
+\[
+  = \frac{\rho \omega^2 }{\kappa} - \xi_x^2\phi(|\omega|) - \xi_z^2
+\]
+\[
+  = \frac{\rho \omega^2 }{\kappa} - \xi_x^2 - \xi_z^2 \mbox{ mod }  S^{-\infty}
+\]  
+as expected. So a maximal choice for $\Gamma$ would be
+\[
+  \Gamma = \{(z,(t,x),(\omega,\xi_x)): c(z,x)|\xi_x| < |\omega|\}.
+\]
 

### 2025-05-29 repro/papers/ssevelan/slides/slides.tex
prep for JV talk

\documentclass[xcolor=dvipsnames,12pt,aspectratio=169]{beamer}
\usecolortheme[named=violet]{structure}

%\usetheme{umbc2} 
\usepackage{pgf}
%\usepackage[pdftex]{graphicx}
\usepackage{color}
\usepackage{amssymb,amsmath}
\usepackage[english]{babel}
\mode<presentation>

 \usepackage{multimedia}
\usefonttheme{structurebold}

%\useinnertheme{umbcboxes}

%\setbeamercolor{umbcboxes}{bg=violet!15,fg=black}

\setbeamercovered{transparent}
\setbeamertemplate{navigation symbols}{}
\input{./math-macros}
\setlength{\parskip}{2ex}   % place a blank line between paragraphs
\def\Logo{\@ifnextchar(\beamer@Logo\logo}
\def\beamer@Logo(#1,#2){\logo}

\AtBeginSection[]{
{\setbeamertemplate{logo}{}
 \begin{frame}<beamer>
    \frametitle{Agenda}
    \tableofcontents[currentsection]
  \end{frame}
}
}
% Images
% Macros
\newcommand{\peq}{\,+\hspace{-0.15cm}=}
\newcommand{\bm}{\bar{m}}
\newcommand{\dm}{\delta m}
\newcommand{\dom}{\delta \bar{m}}
\newcommand{\doma}{\delta \bar{m}_{\alpha}}
\newcommand{\om}{\bar{m}}
\newcommand{\bH}{{\bf H}}
\newcommand{\bR}{{\bf R}}
\newcommand{\bcF}{{\bar{\cal F}}}
\newcommand{\cF}{{\cal F}}
\newcommand{\oF}{\bar{F}}
\newcommand{\of}{\bar{f}}
\newcommand{\ou}{\bar{u}}
\newcommand{\odF}{\bar{F}^{\dagger}}
\newcommand{\oS}{\bar{S}}
\newcommand{\oM}{\bar{M}}
\newcommand{\oJa}{\bar{J}_{\alpha}}
\newcommand{\Na}{N_{\alpha}}
\newcommand{\dd}{\delta d}
\newcommand{\ddl}{\delta d_{\alpha}}
\title[]{Waveform Inversion via Source Extension}
\author[]{William W. Symes \inst{1}}
\institute[]{\inst{1} Rice University}
\date{Workshop in honor of Jean Virieux \\June 2025}

\begin{document}
\frame{
\titlepage
}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

%\begin{frame}\frametitle{The Plan}
%\begin{itemize}
%\item What is ``extended modeling", and why bother
%\item FWI via surface source extension - examples
%\item Theory: why \& when it works
%\item Zoology of source extensions
%\end{itemize}
%\end{frame}

\section{Inversion}

\begin{frame}%\frametitle{Conventional FWI}
FWI: adjust model parameters to make predicted data close to
observed data.

Predicted data = $Ru$, $u$ = modeled wavefield in space-time, $R$ =
sampling operator. Depends on $m$ = model parameters through wave equation
\[
L[m]u = s 
\]
$L[m]$ = wave operator, $s$ = energy source wavefield, Predicted data: $= F[m]s$, $F[m] = RL[m]^{-1}$ =
modeling operator

Least squares (``FWI''): given observed data $d$, source $s$, find $m$ to minimize
\[
J_{\rm FWI}[m] = \sum (F[m]s - d)^2
\]
$\sum$ = sum over all sources, receivers, times
\end{frame}

\begin{frame}
How to do it?

\begin{itemize}
\item Modeling (computing $u$ etc.): many methods - e.g. time-domain
  staggered grid finite differernce (\textcolor{blue}{Virieux}, 84)
\vspace{0,5cm}
\item Iterative descent update using gradient $\nabla J_{\rm FWI}$ - efficient
  computation via {\em adjoint state method} (Chavent-Lemmonier
  74,...), first multi-D demo by Gauthier-\textcolor{blue}{Virieux}-Tarantola 86
\end{itemize}

So what happens?
\end{frame}

\begin{frame}
A simple example: acoustic lens with isotropic point source @ $\bx_s$:
$f = w(t)\delta(\bx-\bx_s)$, $\bx_s$ = source position\\
%\[
%\frac{1}{\kappa}\frac{\partial p}{\partial t} = -\nabla \cdot {\bf v} + w(t)\delta(\bx-\bx_s), \,\rho \frac{\partial {\bf v}}{\partial t} = - \nabla p; \,p,{\bf v}=0, t \ll 0
%\]
%\vspace{-1.25in}
\vspace{-0.25in}
%\hspace{-0.5in}\includegraphics[height=4.0in]{Fig/bml0c.pdf}\\
%\includegraphics[height=2.0in}{Fig/bml0.pdf}\\
\begin{center}
\vspace{0.25in}
\includegraphics[height=1.25in]{Fig/bml0.png}\includegraphics[height=1.25in]{Fig/ptpwindl0.pdf}\\
\vspace{0.5cm} Left: target bulk modulus $\kappa$, source at
$x_s=3500$ m, $z_s=3000$ m, receivers at $x_r=2000-6000$ m, $z_r=1000$
m. Right: synthetic data, source pulse = [1.0, 2.0, 7.5, 12.5] Hz trapzoidal bandpass filter]
\end{center}
\end{frame}

%\begin{frame}
%Modeling operator $(S[\kappa]f)(t,x_r) = \phi(x_r) p(t,x_r,z_r)$ - linear in $f$, nonlinear in $\kappa$
%Model $m=[\kappa]$, wavefield $u=[p,\bv]$, $L[m]u=$ acoustic wave system, $R$ = sample $p$ at receivers, 
%\begin{center}
%\includegraphics[height=1.5in]{Fig/pulse00.pdf} \includegraphics[height=1.5in]{Fig/ptpwindl0.pdf}\\
%\vspace{0.25in}
%Left: [1.0, 2.0, 7.5, 12.5] Hz trapzoidal bandpass filter pulse $w$; Right: pressure traces $Ru$ at $z_r=1000 m$ for source at $x_s=3500 m, z_s=3000 m$. 
%\end{center}
%\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=1.25in]{Fig/covmestfwi2.png}\includegraphics[height=1.25in]{Fig/residcovmestfwi2wind.pdf}\\
\vspace{0.5cm}
Left: target model; 
Right: FWI estimate of $m$ = bulk modulus: initial model $\kappa_0
\equiv 4.0$ GPa, 450 steps steepest descent with smoothing
preconditioner. Right: data residual ($Ru - d$) - energy (norm)
$\approx 25$\% of data norm.
\end{center}

Failure mechanism: ``cycle skipping'', first documented in Gauthier-
\textcolor{blue}{Virieux}-Tarantola 86 - half-wavelength
criterion, (\textcolor{blue}{Virieux}-Operto 09)
\end{frame}

%\begin{frame}
%\begin{center}
%\includegraphics[height=1.2in]{Fig/ptpwindl0.pdf}\includegraphics[height=1.2in]{Fig/fwitrreslh0wind.pdf}\\
%\vspace{0.5cm}
%Left: target data; 
%Right: data residual after 20 KGN steps - final RMS residual reduction = 0.26
%\|d\|=9.72 \|S-d\|=2.51
%\end{center}
%\end{frame}

%\begin{frame}
%Big problem: "cycle skipping"
%\begin{itemize}
%\item $J[m]$ appears to have many (local) minima
%\item to avoid getting stuck at wrong estimates, initial $m$ (and $f$) must predict eve%nt times in $u$ ``correct within a wavelength''
%\end{itemize}

%OK with sufficiently low frequency data with high S/N, but not always available
%\end{frame}

\section{Extension}
\begin{frame}
  Many proposed cures for cycle skipping:
  \begin{itemize}
  \item acquire lower frequency data (Dellinger et al SEG 2016,...) 
  \item combine with traveltime tomography (Luo-Schuster 91,
    Prieux-\ldots-\textcolor{blue}{Virieux} 13)
  \item use a different objective, e.g. transport metric
    (Yang-Engquist 18, M\'{e}tivier-...-\textcolor{blue}{Virieux} 18)
  \item use a different model - {\em enlarge search space, easier data
    fit} (``model extension'')
  \end{itemize}

Many extension methods - Operto et al. 23, Farshad-Chauris 23, Huang
et al 19, S. 08. Common theme: maintain data fit while suppressing
additional degrees of freedom
\end{frame}

\begin{frame}
  \vspace{0.25in}
Example: Adaptive Waveform Inversion (AWI, Warner-Guasch 14): acoustic
model, {\em known} isotropic point source $s = w(t)\delta(\bx-\bx_s)$,

Idea: if events in predicted data $F[m]s$ are in wrong place, use a 
filter $f$ to put them in the right place: extended modeling op $\bar{F}[m,f] = f*F[m]$.
\vspace{-0.25in}
\begin{center}
  \includegraphics[height=1in]{Fig/ptph0half.pdf}\includegraphics[height=1in]{Fig/uest0half.pdf}\includegraphics[height=1in]{Fig/ptpl0half.pdf}\\
\vspace{0.5cm}
Left: Predicted data $F[m_0]s$, $m_0=$ homogenous model; Center: adaptive filter $f$;
Right: adapted data $f*F[m_0]s \approx d =$ lens data
\end{center}

\end{frame}
\begin{frame}
$f$ = $d$ deconvolved (trace-by-trace) by $F[m]s$ (``easy''), so can
make $\bar{F}[m,f]s$ fit data {\em for any m} (common feature of extension methods)

but $\bar{F}[m,f] \ne F[m]$ unless $f(\bx_s,\bx_r,t)=\delta(t)$ - to
test, find an operator vanishing on $\delta(t)$ (``annihilator'' -
another common feature), update $m$ to reduce its output

$t\delta(t)=0$, so minimize
\[
J_{\rm AWI}[m] = \sum_{\bx_s,\bx_r}\frac{{\sum_t|\color{blue}t}
  f[m](\bx_s,\bx_r,t)|^2}{\sum_{t}| f[m](\bx_s,\bx_r,t)|^2}
\]
\end{frame}

%\begin{frame}
%\vspace{-1in}
%\begin{center}
%\hspace{-0.5in}\includegraphics[height=2.5in]{Fig/uest0gz1800wind.pdf}\hspace{-0.5in}\includegraphics[height=2.5in]{Fig/uestmswigz1800wind.pdf}\\
%\vspace{-0.75in}
%Left: $f[m]$, $m$=initial model. Right: $f[m]$, $m$ updated to reduce
%$J_{\rm AWI}$
%\end{center}
%\end{frame}

\begin{frame}
  Good News: suppose
  \begin{itemize}
  \item transmission data dominated by single arrivals - unique
    travel time $\tau$ between source and receiver,
  \item $d \approx F[m_*]s$
  \item RMS wavelength of $w \rightarrow 0$.
  \end{itemize}
  Then
  \[
    J_{\rm AWI}[m] \rightarrow \sum_{\bx_s,\bx_r} |\tau[m]-\tau[m_*]|^2
  \]

  \textcolor{blue}{For single-arrival data, AWI is asymptotic to
    travel time tomography} $\Rightarrow$ no cycle-skipping

  \end{frame}

  \section{Triplication}

\begin{frame}
  Conclusion does not apply to acoustic lens example - multiple
  arrivals.

  Asymptotic analysis, numerical examples $\Rightarrow$

  Significant energy in $d$ at multiple arrival times $\Rightarrow$
  approach based on source-receiver extension (e. g. AWI) behaves like
  FWI (S. 94, Plessix et al. 00, Huang et al. 17, Yong et al. 23.)

\begin{center}
\hspace{-0.5in}\includegraphics[height=0.8in]{Fig/uest0wind.pdf}\includegraphics[height=0.8in]{Fig/uestmswiwind.pdf}\\

Adaptive filters for lens data. Left: $f[m]$, $m$=initial model. Right: $f[m]$, $m$ updated to reduce
$J_{\rm MSWI}$ (similar to $J_{\rm AWI}$).
\end{center}
\end{frame}

\begin{frame}
  Remedy: localize inversion around individual events

  \begin{itemize}
  \item localize in time - Yong-Brossier-Métivier-\textcolor{blue}{Virieux} 23: Local
    Adaptive Waveform Inversion (LAWI) via Gabor filter
  \item localize in phase space - surface source extension (SSE) -
    Huang et al. 19
  \end{itemize}

  SSE extended source: $\bar{s} = \bar{w}(\bx,t;\bx_s)\delta_{{\cal
      S}}(\bx)$, ${\cal S} $= surface containing source locations -
  {\em spread source energy over a surface}, acts as antennae,
  controls full 2D phase spectrum

For any $m$, determine $\bar{w}=\bar{w}[m]$ so that $F\bar{s} \approx d$. Update $m$ to drive
$\bar{w}(\bx,t;\bx_s) \rightarrow w(t)\delta(\bx-\bx_s)$ by minimizing
\[
  J_{\rm SSE}[m] = \sum_{\bx_s}\sum_{\bx,t} ||(\bx-\bx_s)|
  \bar{w}[m](\bx,t;\bx_s)|^2
 \]
\end{frame}

%\begin{frame}
%Detailed Formulation:

%At solution, must recover known constraints on model (``get closer to Vancouver's latitude as you get close to the sea''):

%$\Rightarrow$ introduce {\color{blue} annihilator} $A$, vanishing on models obeying these constraints - minimize output jointly with data residual

%Many possibilities - MS difference w/ known source (WRI), mean time lag (AWI)

%Many choices. For SSE: scale by offset 
%\[
%A\of(\bx,t;x_s,y_s) = \sqrt{(x-x_s)^2+(y-y_s)^2}w(x,y,t;x_s,y_s)
%\]

%{\color {blue} Extension Principle:} $m,\of$ so  that $R\bar{u} \approx d$, $A\of \approx 0$ $\Rightarrow$ approx FWI solution
%\end{frame}

%\begin{frame}
%Extended objective function: data misfit + annihilator output,

%\[
%J_{\rm SSE}[m] = \min_{\of} \,\frac{1}{2} \sum (R\ou - d)^2 + \alpha^2 \sum (A\of)^2
%\]

%Reduced: ``pre-minimized'' over source

%FWI via surface source extension: given $d$ and initial $m_0$, use numerical optimization method to find minimizer $m$ of
%$J_{\rm SSE}[m]$.
%\end{frame}

%\begin{frame}
%Examples:

%(1) diving wave data (sampling op may include window in defn)

%(2) reflection data

%In both cases, method is [??? VPM, CG for inner problem, steepest descent for outer???]
%\end{frame}

%\begin{frame}
%Mohamed's examples 
%\begin{itemize}
%\item include straight FWI for both
%\item also some info about objective values, residuals, convergence, etc.
%\item plots of models, but also sources, residuals - especially want to show extended source from early iteration
%\end{itemize}
%\end{frame}

\begin{frame}
Penalty method: minimize
\[
J_{\alpha}[m,\bar{w}] = \frac{1}{2} \sum |F[m]\bar{w} - d|^2 + \alpha^2||\bx-\bx_s|\bar{w}|^2
\]
over $m$

\begin{itemize}
\item time-domain FD, gradient via adjoint state
\item Variable Projection:
  \begin{itemize}
  \item update $\bar{w}$ using Conjugate Gradient Iteration (inner loop)
\item update $m$ using steepest descent, line search (outer loop) plus
  smoothing regularization
  \end{itemize}
\item update $\alpha$ using Discrepancy Principle (keep data residual $\in (e_-, e_+)$) - {\color{blue} can always choose $\bar{w}$ to fit data} $\Rightarrow$ can start with $\alpha=0$ (Fu \& S. 17)
\end{itemize}
\end{frame}

\begin{frame}
\begin{center}
SSE for 1 shot lens data, source surface ${\cal S} = \{z = 3000 m\}$
\end{center}
\begin{center}
\includegraphics[height=1.25in]{Fig/cgw1_bulkupdlh0.png}\includegraphics[height=1.25in]{Fig/cgw41_bulkupdlh0.png}\\
\vspace{0.25in}
Start at $\kappa_0=4.0$ GPa, use $e_-=0.05\|d\|, e_+=0.1\|d\|$. \\
Left: Iteration 1 ($\alpha=2.9 e-6$)\\
Right: Iteration 41 ($\alpha = 1.99 e-5$)
\end{center}
\end{frame}

\begin{frame}
$m=[\kappa]$ converges $\leftrightarrow$ $\alpha$ increases, extended source $\of$ (initially spread over $z=z_s=3000$ m) {\color{blue} focuses} at $x=x_s=3500$ m.
\begin{center}
\hspace{-1cm}\includegraphics[height=1.25in]{Fig/cgw0_est_source_plh0.pdf}\includegraphics[height=1.25in]{Fig/cgw8_est_source_plh0.pdf}\includegraphics[height=1.25in]{Fig/cgw18_est_source_plh0.pdf}\includegraphics[height=1.25in]{Fig/cgw40_est_source_plh0.pdf}\\
\vspace{0.25in}
Left to right: $\of$ at Iterations 1 ($\alpha=2.9 e-6$), 9
($\alpha= 5.2 e-6$ m) 19 ($\alpha=1.32 e-5$) and 41 ($\alpha=1.99e-5$).
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=1.25in]{Fig/covmestfwicgw41.png}\includegraphics[height=1.25in]{Fig/residcovmestfwicgw41wind.pdf}\\
\vspace{0.25in}
Left: ``Jump to $\alpha=\infty$": bulk modulus estimated by FWI, 200
steepest descent steps starting with iteration 41 of SSE., final RMS residual reduction = 0.05;
Right: FWI data residual 
\end{center}
\end{frame}

\begin{frame}
  Why it works - a hint:

provided that
\begin{itemize}
\item $d = F[m_*]s$,
\item RMS wavelength of $w$ (hence $d$) $\rightarrow 0$,
\end{itemize}
\[
  \nabla \nabla  J_{\rm SSE}[m_*] \rightarrow \sum_{\bx,\bp}
\]
\end{frame}

\end{document}

\begin{frame}
\begin{center}
\vspace{-0.7in}
\includegraphics[height=4.5in]{HMC1.pdf}
\end{center}
\end{frame}

\begin{frame}
HMC Salt/Carbonate Acoustic Model

Streamer acquisition $z_s = 7$ m, $\Delta x_s=37.5$ m, 315 shots, $z_r=15$ m, $\Delta x_r=12.5$ m, 475 recvs, max offset = $6$ km

Pulse for data generation = [2, 3, 6, 10] Hz trapezoidal bandpass filter

Free surface @ $z=0$, other boundaries absorbing, 2-8 staggered FD scheme

Regularization: penalize MS velocity gradient, weight (also $\alpha$) via several trials

Data contains diving waves, pre- \& post-critical reflections, multiple reflections, refracted multiple reflections, head waves, free surface multiples, ....

\end{frame}

\begin{frame}
\begin{center}
\vspace{-0.7in}
\includegraphics[height=4.5in]{HMC2.pdf}
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\vspace{-0.7in}
\includegraphics[height=4.5in]{HMC3.pdf}
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\vspace{-0.7in}
\includegraphics[height=4.5in]{HMC4.pdf}
\end{center}
\end{frame}

\section{Theory: why \& when it works}

\begin{frame}
FWI vs. SSE for transmission (diving wave, crosswell) data:

$J_{\rm FWI}[m]$ and $J_{\rm SSE}[m]$ = regular functions of $m$, locally close to quadratic approximations.

$d$ consistent: $d=Ru[m,f]$, $\of = f = w(t)\delta(\bx-\bx_s)$, $A\of=0$ $\Rightarrow$

\begin{itemize}
\item  $J_{\rm FWI}[m]=J_{\rm SSE}[m]=0$
\vspace{0.5cm}
\item $J_{\rm FWI}[m + \delta m] \approx \mbox{const. }\omega^2 \|\delta m\|^2$ for some $\delta m$ (kinematic perturbations), $\omega$ = center frequency in source pulse $w(t)$
\vspace{0.5cm}
\item $J_{\rm SSE}[m+\delta m] \le \mbox{const. }\|\delta m\|^2$ independent of $\omega, \delta m$
\end{itemize}
\end{frame}

\begin{frame}
$\Rightarrow$ radius of convexity region:
\begin{itemize}
\item $J_{\rm FWI}: O(1/\omega)$, must initialize ``within a wavelength''
\item $J_{\rm SSE}: O(1)$, converges even with (rel.) poor initial guess, HF data
\end{itemize}

Method of proof: ray theory (``mocrolocal analysis'')

More detailed analysis: minimizing $J_{\rm SSE}$ equivalent to a form of transmission travel time tomography: {\color{blue} tomography without picking}

NB: for reflection data... ???
\end{frame}

\begin{frame}\frametitle{Late Breaking News}
Theory for special reflection case: layered ($\kappa(z)$) medium, dip filter to eliminate postcritical/diving wave energy

Explicit computation of $\nabla J_{\rm SSE}$ for 
\begin{itemize}
\item homogenous background 
\item data from single reflector in otherwise slowly varying $\kappa$, wavelet with central frequency $\omega$
\end{itemize}
$\Rightarrow$ $\nabla J_{\rm SSE} \rightarrow 0$ as $\omega \rightarrow \infty$

Same behavior as $\nabla J_{\rm FWI}$!!!

Upshot: for pure reflection data, SSE may not provide constructive velocity updates
\end{frame}

\begin{frame}
{\color{blue} {\bf Summary:}} Surface Source Extension

\begin{itemize}
\item overcomes cycle-skipping for transmission data, solid theoretical foundation
\item same computational complexity as FWI
\item open question: theory for reflections {\color{blue} (initial result not positive)}
\end{itemize}
%\end{frame}

%\begin{frame}
Other source extension methods (WRI, AWI, ...)

\begin{itemize}
\item Many promising examples 
\item Little known theoretically - connection to travel time tomography???
\end{itemize}
\end{frame}

\begin{frame}\frametitle{Model Extension}
Add degrees of freedom to $v$, $\kappa$, instead of source

Annihilator: flatten/focus gathers

{\color{blue}Extracts velocity updates from reflection and refraction events} (S. 86, Kern \& S. 94, S. GP 08, Shen \& S. GEO 08, Shen EAGE 12, Biondi \& Almomin 12, 14, Chauris \& Cocher GEO 17, Hou \& S. GEO 18, Barnier \& Biondi SEG 19)

Chief challenge: expense(model extension) $\gg$ expense(FWI) (Note: expense(source extension) $\approx$ expense(FWI))
\end{frame}

\begin{frame}\frametitle{Acknowledgements}
\begin{itemize}
\item Total E\&P USA
\item Sponsors of The Rice Inversion Project
\item Organizers of SS 1 (Recent Advances and Road Ahead) at SEG 2019
\end{itemize}
\end{frame}

\end{document}

\begin{frame}
Matched Source Waveform Inversion (Hua \& Symes 1993, 4, Huang \& Symes 2017)

\begin{itemize}
\item similar to AWI (source wavelet for each trace), w/o normalization
\item if
\begin{itemize}
\item transmission data (crosswell, diving wave)
\item unique ray paths
\end{itemize}
then mean square moment $\approx$ mean square travel time misfit: {\color{blue} tomography without picking}

\item relation with tomography fails ($\approx$ FWI) if multiple rays connect sources and receivers - contrast SSE
\end{itemize}
\end{frame}

\begin{frame}
\begin{center}
Example 2: Deep Water Marmousi
\end{center}
\begin{center}
\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmVt.pdf}\\
Modified Marmousi Model: 2 km water layer added to minimize non-reflected energy, also box-snaped target
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot1}\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot81}\\
Data: isotropic point source with 8 Hz Ricker wavelet, $z_r=0.04$ km, $x_r=$ 0.05 km to 9.15 km, $\Delta x_r=$ 0.05 km, $z_s=0.08$ km, $x_s=0.6$ km to $x_s=8.6$, $\Delta x_s$ = 0.1 km. Shots at $x_s=0.6$ km and $x_s=8.6$ km displayed.
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmV0.pdf}\\
Initial $v(z)$ model for both FWI and SSE
\end{center}
\end{frame}

\begin{frame}
\vspace{-0.1cm}
\begin{center}
\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmFWI.pdf}\\
FWI estimate of $v$: 200 iterations LBFGS, frequency domain FD 5-8 Hz
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmSurf.pdf}\\
SSE estimate of $v$: 150 iterations LBFGS, frequency domain FD 5-8 Hz, fixed $\alpha$ set by trial
\end{center}
\end{frame}

\begin{frame}
%\vspace{0.07cm}
\begin{center}
\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmVt.pdf}\\
Modified Marmousi Model: 2 km water layer added to minimize non-reflected energy, also box-snaped target
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot1Inv}\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot81Inv}\\
Data predicted from SSE inverted model: shots at $x_s=0.6$ km and $x_s=8.6$ km.
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot1}\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot81}\\
Target data:\\ shots at $x_s=0.6$ km and $x_s=8.6$ km.
\end{center}
\end{frame}

\begin{frame}
Wavefield Reconstruction Inversion: $\of(\bx,t;\bx_s)$ distributed throughout space-time

van Leeuwen and Herrmann 2013, 2016, Wang \& Yingst 2016

similar: space-time extension Huang \& S. 2018

\begin{itemize}
\item[+ ]many good examples, convincing numerical evidence
\item[+ ]3D field data - Wang \& Yingst 2016
\item[- ]basic form: store complete space-time (or space-frequency) fields (6D)
\end{itemize}
\end{frame}

\begin{frame}
Adaptive Waveform Inversion (Warner \& Guatsch, 2014, 2016): $\of(\bx,t;\bx_r,\bx_s) = w(t;\bx_r)\delta(\bx-\bx_s)$ localized but depends on receiver coordinate $\bx_r$ - one wavelet per data trace

similar: source-receiver extension Hua \& S. 1994, Huang \& S. 2017

\begin{itemize}
\item[+ ]additional storage: $\approx$ data volume (5D)
\item[+ ]minor mod of standard FWI workflow
\item[- ]source-receiver: loses advantage over FWI when sources, receivers connected by multiple ray paths (fails for lens model, above)
\end{itemize}
\end{frame}

\begin{frame}
Intermediate: source distributed over space-time surface
\begin{itemize}
\item const. t surface = volume source extension - Huang \& S. 2018
\item const z surface = {\color{blue} surface source extension (SSE):} 
\[
\of(\bx,t;x_s,y_s) = w(x,y; x_s,y_s)\delta(z-z_s)
\]
\end{itemize}
\begin{itemize}
\item[+ ]additional storage = O(data volume) (5D)
\item[+ ]no problem with multiple ray paths
\item[+ ]theory for transmission data (eg. diving wave): {\color{blue} tomography w/o picking}
\end{itemize} 
\end{frame}

### 2025-05-29 repro/papers/ssevelan/slides/slidesSEG19.tex
prep for JV talk

--- 2025-05-29 repro/papers/ssevelan/slides/slides.tex
@@
 \title[]{Waveform Inversion via Source Extension}
-\author[]{William W. Symes \inst{1}}
-\institute[]{\inst{1} Rice University}
-\date{Workshop in honor of Jean Virieux \\June 2025}
+\author[]{Guanghui Huang \inst{1} \and Rami Nammour \inst{2} \and \\William W. Symes \inst{3} \and Mohamed Dolliazal \inst{2}}
+\institute[]{\inst{1} Michigan State University \and \inst{2} Total $E\&P$ $R\&T$ USA \and \inst{3} Rice University}
+\date{September 2019}
 
@@
 
-\section{Inversion}
+\section{What is ``extended modeling'', and why bother}
 
 \begin{frame}%\frametitle{Conventional FWI}
-FWI: adjust model parameters to make predicted data close to
-observed data.
+FWI: adjust model to minimize data misfit (observed - modeled)
 
-Predicted data = $Ru$, $u$ = modeled wavefield in space-time, $R$ =
-sampling operator. Depends on $m$ = model parameters through wave equation
+Wave equation:
 \[
-L[m]u = s 
+L[m]u = f 
 \]
-$L[m]$ = wave operator, $s$ = energy source wavefield, Predicted data: $= F[m]s$, $F[m] = RL[m]^{-1}$ =
-modeling operator
+$m$ = model parameters, $L$ = wave operator, $u$ = modeled wavefield(s), $f$ = energy source model
 
-Least squares (``FWI''): given observed data $d$, source $s$, find $m$ to minimize
+FWI objective: 
 \[
-J_{\rm FWI}[m] = \sum (F[m]s - d)^2
+J_{\rm FWI}[m] = \sum (R u - d)^2
 \]
-$\sum$ = sum over all sources, receivers, times
+$R$ = sampling operator, $Ru$ = modeled data, $d$ = observed data, $\sum$ = sum over all components
+
+Note: Wave equation $\Rightarrow$ $u = u[m,f]$
 \end{frame}
@@
 \begin{frame}
-How to do it?
-
-\begin{itemize}
-\item Modeling (computing $u$ etc.): many methods - e.g. time-domain
-  staggered grid finite differernce (\textcolor{blue}{Virieux}, 84)
-\vspace{0,5cm}
-\item Iterative descent update using gradient $\nabla J_{\rm FWI}$ - efficient
-  computation via {\em adjoint state method} (Chavent-Lemmonier
-  74,...), first multi-D demo by Gauthier-\textcolor{blue}{Virieux}-Tarantola 86
-\end{itemize}
-
-So what happens?
-\end{frame}
-
-\begin{frame}
-A simple example: acoustic lens with isotropic point source @ $\bx_s$:
-$f = w(t)\delta(\bx-\bx_s)$, $\bx_s$ = source position\\
+A simple example: acoustic transmission inversion w/ isotropic point source @ $\bx_s$: $f = w(t)\delta(\bx-\bx_s)$
 %\[
@@
 %\]
-%\vspace{-1.25in}
-\vspace{-0.25in}
-%\hspace{-0.5in}\includegraphics[height=4.0in]{Fig/bml0c.pdf}\\
+\vspace{-1.25in}
+%\vspace{-0.5in}
+\begin{center}
+\hspace{-0.5in}\includegraphics[height=4.0in]{Fig/bml0c.pdf}\\
 %\includegraphics[height=2.0in}{Fig/bml0.pdf}\\
-\begin{center}
-\vspace{0.25in}
-\includegraphics[height=1.25in]{Fig/bml0.png}\includegraphics[height=1.25in]{Fig/ptpwindl0.pdf}\\
-\vspace{0.5cm} Left: target bulk modulus $\kappa$, source at
-$x_s=3500$ m, $z_s=3000$ m, receivers at $x_r=2000-6000$ m, $z_r=1000$
-m. Right: synthetic data, source pulse = [1.0, 2.0, 7.5, 12.5] Hz trapzoidal bandpass filter]
-\end{center}
-\end{frame}
-
-%\begin{frame}
-%Modeling operator $(S[\kappa]f)(t,x_r) = \phi(x_r) p(t,x_r,z_r)$ - linear in $f$, nonlinear in $\kappa$
-%Model $m=[\kappa]$, wavefield $u=[p,\bv]$, $L[m]u=$ acoustic wave system, $R$ = sample $p$ at receivers, 
-%\begin{center}
-%\includegraphics[height=1.5in]{Fig/pulse00.pdf} \includegraphics[height=1.5in]{Fig/ptpwindl0.pdf}\\
-%\vspace{0.25in}
-%Left: [1.0, 2.0, 7.5, 12.5] Hz trapzoidal bandpass filter pulse $w$; Right: pressure traces $Ru$ at $z_r=1000 m$ for source at $x_s=3500 m, z_s=3000 m$. 
-%\end{center}
-%\end{frame}
-
-\begin{frame}
-\begin{center}
-\includegraphics[height=1.25in]{Fig/covmestfwi2.png}\includegraphics[height=1.25in]{Fig/residcovmestfwi2wind.pdf}\\
-\vspace{0.5cm}
-Left: target model; 
-Right: FWI estimate of $m$ = bulk modulus: initial model $\kappa_0
-\equiv 4.0$ GPa, 450 steps steepest descent with smoothing
-preconditioner. Right: data residual ($Ru - d$) - energy (norm)
-$\approx 25$\% of data norm.
-\end{center}
-
-Failure mechanism: ``cycle skipping'', first documented in Gauthier-
-\textcolor{blue}{Virieux}-Tarantola 86 - half-wavelength
-criterion, (\textcolor{blue}{Virieux}-Operto 09)
-\end{frame}
-
-%\begin{frame}
-%\begin{center}
-%\includegraphics[height=1.2in]{Fig/ptpwindl0.pdf}\includegraphics[height=1.2in]{Fig/fwitrreslh0wind.pdf}\\
-%\vspace{0.5cm}
-%Left: target data; 
-%Right: data residual after 20 KGN steps - final RMS residual reduction = 0.26
-%\|d\|=9.72 \|S-d\|=2.51
-%\end{center}
-%\end{frame}
-
-%\begin{frame}
-%Big problem: "cycle skipping"
-%\begin{itemize}
-%\item $J[m]$ appears to have many (local) minima
-%\item to avoid getting stuck at wrong estimates, initial $m$ (and $f$) must predict eve%nt times in $u$ ``correct within a wavelength''
-%\end{itemize}
-
-%OK with sufficiently low frequency data with high S/N, but not always available
-%\end{frame}
-
-\section{Extension}
-\begin{frame}
-  Many proposed cures for cycle skipping:
-  \begin{itemize}
-  \item acquire lower frequency data (Dellinger et al SEG 2016,...) 
-  \item combine with traveltime tomography (Luo-Schuster 91,
-    Prieux-\ldots-\textcolor{blue}{Virieux} 13)
-  \item use a different objective, e.g. transport metric
-    (Yang-Engquist 18, M\'{e}tivier-...-\textcolor{blue}{Virieux} 18)
-  \item use a different model - {\em enlarge search space, easier data
-    fit} (``model extension'')
-  \end{itemize}
-
-Many extension methods - Operto et al. 23, Farshad-Chauris 23, Huang
-et al 19, S. 08. Common theme: maintain data fit while suppressing
-additional degrees of freedom
-\end{frame}
-
-\begin{frame}
-  \vspace{0.25in}
-Example: Adaptive Waveform Inversion (AWI, Warner-Guasch 14): acoustic
-model, {\em known} isotropic point source $s = w(t)\delta(\bx-\bx_s)$,
-
-Idea: if events in predicted data $F[m]s$ are in wrong place, use a 
-filter $f$ to put them in the right place: extended modeling op $\bar{F}[m,f] = f*F[m]$.
-\vspace{-0.25in}
-\begin{center}
-  \includegraphics[height=1in]{Fig/ptph0half.pdf}\includegraphics[height=1in]{Fig/uest0half.pdf}\includegraphics[height=1in]{Fig/ptpl0half.pdf}\\
-\vspace{0.5cm}
-Left: Predicted data $F[m_0]s$, $m_0=$ homogenous model; Center: adaptive filter $f$;
-Right: adapted data $f*F[m_0]s \approx d =$ lens data
-\end{center}
-
-\end{frame}
-\begin{frame}
-$f$ = $d$ deconvolved (trace-by-trace) by $F[m]s$ (``easy''), so can
-make $\bar{F}[m,f]s$ fit data {\em for any m} (common feature of extension methods)
-
-but $\bar{F}[m,f] \ne F[m]$ unless $f(\bx_s,\bx_r,t)=\delta(t)$ - to
-test, find an operator vanishing on $\delta(t)$ (``annihilator'' -
-another common feature), update $m$ to reduce its output
-
-$t\delta(t)=0$, so minimize
-\[
-J_{\rm AWI}[m] = \sum_{\bx_s,\bx_r}\frac{{\sum_t|\color{blue}t}
-  f[m](\bx_s,\bx_r,t)|^2}{\sum_{t}| f[m](\bx_s,\bx_r,t)|^2}
-\]
-\end{frame}
-
-%\begin{frame}
-%\vspace{-1in}
-%\begin{center}
-%\hspace{-0.5in}\includegraphics[height=2.5in]{Fig/uest0gz1800wind.pdf}\hspace{-0.5in}\includegraphics[height=2.5in]{Fig/uestmswigz1800wind.pdf}\\
-%\vspace{-0.75in}
-%Left: $f[m]$, $m$=initial model. Right: $f[m]$, $m$ updated to reduce
-%$J_{\rm AWI}$
-%\end{center}
-%\end{frame}
-
-\begin{frame}
-  Good News: suppose
-  \begin{itemize}
-  \item transmission data dominated by single arrivals - unique
-    travel time $\tau$ between source and receiver,
-  \item $d \approx F[m_*]s$
-  \item RMS wavelength of $w \rightarrow 0$.
-  \end{itemize}
-  Then
-  \[
-    J_{\rm AWI}[m] \rightarrow \sum_{\bx_s,\bx_r} |\tau[m]-\tau[m_*]|^2
-  \]
-
-  \textcolor{blue}{For single-arrival data, AWI is asymptotic to
-    travel time tomography} $\Rightarrow$ no cycle-skipping
-
-  \end{frame}
-
-  \section{Triplication}
-
-\begin{frame}
-  Conclusion does not apply to acoustic lens example - multiple
-  arrivals.
-
-  Asymptotic analysis, numerical examples $\Rightarrow$
-
-  Significant energy in $d$ at multiple arrival times $\Rightarrow$
-  approach based on source-receiver extension (e. g. AWI) behaves like
-  FWI (S. 94, Plessix et al. 00, Huang et al. 17, Yong et al. 23.)
-
-\begin{center}
-\hspace{-0.5in}\includegraphics[height=0.8in]{Fig/uest0wind.pdf}\includegraphics[height=0.8in]{Fig/uestmswiwind.pdf}\\
-
-Adaptive filters for lens data. Left: $f[m]$, $m$=initial model. Right: $f[m]$, $m$ updated to reduce
-$J_{\rm MSWI}$ (similar to $J_{\rm AWI}$).
+\vspace{-1in}
+target bulk modulus $\kappa$, source at $x_s=3500$ m, $z_s=3000$ m, receivers at $x_r=2000-6000$ m, $z_r=1000$ m
 \end{center}
@@
 \begin{frame}
-  Remedy: localize inversion around individual events
+%Modeling operator $(S[\kappa]f)(t,x_r) = \phi(x_r) p(t,x_r,z_r)$ - linear in $f$, nonlinear in $\kappa$
+Model $m=[\kappa]$, wavefield $u=[p,\bv]$, $L[m]u=$ acoustic wave system, $R$ = sample $p$ at receivers, 
+\begin{center}
+\includegraphics[height=1.5in]{Fig/pulse00.pdf} \includegraphics[height=1.5in]{Fig/ptpwindl0.pdf}\\
+\vspace{0.25in}
+Left: [1.0, 2.0, 7.5, 12.5] Hz trapzoidal bandpass filter pulse $w$; Right: pressure traces $Ru$ at $z_r=1000 m$ for source at $x_s=3500 m, z_s=3000 m$. 
+\end{center}
+\end{frame}
+
+\begin{frame}
+Minimizing $J_{\rm FWI}[m] = \sum (R u - d)^2$:
 
 \begin{itemize}
-  \item localize in time - Yong-Brossier-Métivier-\textcolor{blue}{Virieux} 23: Local
-    Adaptive Waveform Inversion (LAWI) via Gabor filter
-  \item localize in phase space - surface source extension (SSE) -
-    Huang et al. 19
+\item Modeling (computing $u$ etc.): time-domain Finite Difference
+\vspace{0,5cm}
+\item Efficient comp'n of $\nabla J_{\rm FWI}$ = ``adjoint state method'' (Chavent-Lemmonier 74,...) - $\sim$ RTM
+\vspace{0.5cm}
+\item Iteratively update $m$ to reduce $J_{\rm FWI}$ - Newton's or related gradient/descent method, plus smoothing regularization
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\begin{center}
+\includegraphics[height=1.25in]{Fig/bml0.pdf}\includegraphics[height=1.25in]{Fig/fwitrbulkmodinvlh0.pdf}\\
+\vspace{0.5cm}
+Left: target model; 
+Right: FWI estimate of $m$ = bulk modulus: initial model $\kappa_0 \equiv 4.0$ GPa, 20 iterations trust region Krylov-Gauss-Newton with smoothing preconditioner.
+Each KGN step =  up to ``inner'' 20 CG steps $\approx$ 400 fwd models
+\end{center}
+\end{frame}
+
+\begin{frame}
+\begin{center}
+\includegraphics[height=1.2in]{Fig/ptpwindl0.pdf}\includegraphics[height=1.2in]{Fig/fwitrreslh0wind.pdf}\\
+\vspace{0.5cm}
+Left: target data; 
+Right: data residual after 20 KGN steps - final RMS residual reduction = 0.26
+%\|d\|=9.72 \|S-d\|=2.51
+\end{center}
+\end{frame}
+
+\begin{frame}
+Big problem: "cycle skipping"
+\begin{itemize}
+\item $J[m]$ appears to have many (local) minima
+\item to avoid getting stuck at wrong estimates, initial $m$ (and $f$) must predict event times in $u$ ``correct within a wavelength''
 \end{itemize}
 
-  SSE extended source: $\bar{s} = \bar{w}(\bx,t;\bx_s)\delta_{{\cal
-      S}}(\bx)$, ${\cal S} $= surface containing source locations -
-  {\em spread source energy over a surface}, acts as antennae,
-  controls full 2D phase spectrum
-
-For any $m$, determine $\bar{w}=\bar{w}[m]$ so that $F\bar{s} \approx d$. Update $m$ to drive
-$\bar{w}(\bx,t;\bx_s) \rightarrow w(t)\delta(\bx-\bx_s)$ by minimizing
-\[
-  J_{\rm SSE}[m] = \sum_{\bx_s}\sum_{\bx,t} ||(\bx-\bx_s)|
-  \bar{w}[m](\bx,t;\bx_s)|^2
- \]
+OK with sufficiently low frequency data with high S/N, but not always available
 \end{frame}
 
-%\begin{frame}
-%Detailed Formulation:
+\begin{frame}\frametitle{An Analogy}
+%\vspace{0.5cm}
+Task: Given latitude of Vancouver ($49^{\circ} 15' N$), find its longitude by traveling there from somewhere in Canadian Rockies ($\approx 121^{\circ} W$)
 
-%At solution, must recover known constraints on model (``get closer to Vancouver's latitude as you get close to the sea''):
+Algorithm: start at correct latitude (eg. Sunshine Valley, $49^{\circ} 16' N$, $121^{\circ} 11' W$) and go {\color{blue} downhill and due west} until altitude minimized 
+\vspace{-3cm}
+\begin{center}
+\includegraphics[width=0.6\textwidth,angle=270]{Fig/bc.pdf}
+\end{center}
+\vspace{-3cm}
+\footnotetext{\tiny{https://www.google.com/maps/@49.2332511,-122.8181263,9z, 2019.08.18}}
+\end{frame}
 
-%$\Rightarrow$ introduce {\color{blue} annihilator} $A$, vanishing on models obeying these constraints - minimize output jointly with data residual
+\begin{frame}
+You, looking west. Now, {\color{blue} go downhill and due west}... oops.
+\vspace{-1.5cm}
+\begin{center}
+\includegraphics[width=0.6\textwidth,angle=270]{Fig/mtns.pdf}
+\end{center}
+\vspace{-1cm}
+%\tiny{{https://www.dreamingandwandering.com/2016/01/10/25-photos-of-the-canadian-rockies-that-will-make-you-pack-your-bags-and-go/}, 2019.08.21}
+\footnotetext{\tiny{\noindent https://www.dreamingandwandering.com/2016/01/10/25-photos-of-the-canadian-rockies-that-will-make-you-pack-your-bags-and-go/}, 2019.08.21}
+\end{frame}
+
+\begin{frame}
+Analogy w/ FWI: altitude $\sim$ $J_{\rm FWI}[m] =$ mean square error
+
+Can't reduce to desired min = 0 while going downhill ($\sim$ using descent method) {\color{blue} and} maintaining latitude of Vancouver ($\sim$ known physical properties of solution)
+
+Descent methods - mandatory because of computational size, can't drop
+
+Alternative: drop ``physical constraint'' (latitude), search in larger ``unphysical'' space 
+\end{frame}
+
+\begin{frame}
+With full 2D search available, can go to local creek, then float down to local river, join Fraser River near Hope ($49^{\circ} 23' N$) ... float your boat to Vancouver, downhill all the way!
+\vspace{-2.8cm}
+\begin{center}
+\includegraphics[width=0.5\textwidth]{Fig/fraser.pdf}
+\end{center}
+\vspace{-3cm}
+Moral: dropping some known properties (physical constraints) of solution {\color{blue} during search} may make descent search feasible
+
+\footnotetext{\tiny{https://images.search.yahoo.com/yhs/search?p=fraser+river}, 2019.08.21}
+
+\end{frame}
+
+\section{FWI via source extension}
+\begin{frame}
+
+{\color{blue} Extended modeling:} allow interim models to violate known ``physical'' constraints on true solution
+
+Active source data (usually) involves {\em localized} source model with known location, eg. ($\bx=(x,y,z)$)
+\[
+f(\bx,t;x_s,y_s) = w(t) \delta(\bx-\bx_s) = \mbox{ isotropic point source at }\bx=\bx_s
+\]
+{\color{blue} Source Extension}: replace $f$ with {\em non-localized} source model $\of$
+
+Extended wavefield $\ou$  solves $L[m]\ou = \of$, data prediction = $R\ou$ - computational complexity similar to FWI
+
+\end{frame}
+
+\begin{frame}
+Source extension methods:
+\begin{itemize}
+\item
+Wavefield Reconstruction Inversion (WRI):
+van Leeuwen and Herrmann GJI 2013, Inv. Probs. 2016, Wang \& Yingst SEG 2016, Aghamiry et al GEO 2019 
+\item Adaptive Waveform Inversion (AWI): Warner \& Guatsch, GEO 2014, 2016
+\item ... (see abstract)
+\item {\color{blue} surface source extension (SSE):} $\of(\bx,t;x_s,y_s) = w(x,y,t; x_s,y_s)\delta(z-z_s)$
+\end{itemize}
+
+Common feature: additional source degrees of freedom permit good fit to data {\em for large set of $m$}
+\end{frame}
+
+\begin{frame}
+Detailed Formulation:
+
+At solution, must recover known constraints on model (``get closer to Vancouver's latitude as you get close to the sea''):
+
+$\Rightarrow$ introduce {\color{blue} annihilator} $A$, vanishing on models obeying these constraints - minimize output jointly with data residual
 
@@
 
-%Many choices. For SSE: scale by offset 
-%\[
-%A\of(\bx,t;x_s,y_s) = \sqrt{(x-x_s)^2+(y-y_s)^2}w(x,y,t;x_s,y_s)
-%\]
+Many choices. For SSE: scale by offset 
+\[
+A\of(\bx,t;x_s,y_s) = \sqrt{(x-x_s)^2+(y-y_s)^2}w(x,y,t;x_s,y_s)
+\]
 
-%{\color {blue} Extension Principle:} $m,\of$ so  that $R\bar{u} \approx d$, $A\of \approx 0$ $\Rightarrow$ approx FWI solution
-%\end{frame}
+{\color {blue} Extension Principle:} $m,\of$ so  that $R\bar{u} \approx d$, $A\of \approx 0$ $\Rightarrow$ approx FWI solution
+\end{frame}
 
@@
 \begin{frame}
-Penalty method: minimize
+FWI via SSE: given $d$ and initial $m_0$, use numerical optimization method to find minimizer $m$ of
 \[
-J_{\alpha}[m,\bar{w}] = \frac{1}{2} \sum |F[m]\bar{w} - d|^2 + \alpha^2||\bx-\bx_s|\bar{w}|^2
+J_{\rm SSE}[m] = \mbox{ min}_{\of}\frac{1}{2} \sum (R\ou - d)^2 + \alpha^2 \sum (A\of)^2
 \]
@@
 \item time-domain FD, gradient via adjoint state
-\item Variable Projection:
-  \begin{itemize}
-  \item update $\bar{w}$ using Conjugate Gradient Iteration (inner loop)
-\item update $m$ using steepest descent, line search (outer loop) plus
-  smoothing regularization
-  \end{itemize}
-\item update $\alpha$ using Discrepancy Principle (keep data residual $\in (e_-, e_+)$) - {\color{blue} can always choose $\bar{w}$ to fit data} $\Rightarrow$ can start with $\alpha=0$ (Fu \& S. 17)
+\item update $\of$ using Conjugate Gradient Iteration (inner loop)
+\item update $m$ using steepest descent, line search (outer loop) plus smoothing regularization
+\item update $\alpha$ using Discrepancy Principle (keep data residual $\in (e_-, e_+)$) - {\color{blue} can always choose $\of$ to fit data} $\Rightarrow$ can start with $\alpha=0$ (Fu \& S. GEO 2017)
 \end{itemize}
@@
 \begin{center}
-SSE for 1 shot lens data, source surface ${\cal S} = \{z = 3000 m\}$
+Example 1: Lens model
 \end{center}
 \begin{center}
-\includegraphics[height=1.25in]{Fig/cgw1_bulkupdlh0.png}\includegraphics[height=1.25in]{Fig/cgw41_bulkupdlh0.png}\\
+\includegraphics[height=1.25in]{Fig/cgw1bulkupdlh0.pdf}\includegraphics[height=1.25in]{Fig/cgw30bulkupdlh0.pdf}\\
 \vspace{0.25in}
-Start at $\kappa_0=4.0$ GPa, use $e_-=0.05\|d\|, e_+=0.1\|d\|$. \\
-Left: Iteration 1 ($\alpha=2.9 e-6$)\\
-Right: Iteration 41 ($\alpha = 1.99 e-5$)
+Start at $\kappa_0=4.0$ GPa, use $e_-=0.1\|d\|, e_+=0.2\|d\|$. \\
+Left: Iteration 1 ($\alpha=4.83 \times 10^{-6}$)\\
+Right: Iteration 30 ($\alpha = 9.1\times 10^{-5}$)
 \end{center}
@@
 \begin{frame}
-$m=[\kappa]$ converges $\leftrightarrow$ $\alpha$ increases, extended source $\of$ (initially spread over $z=z_s=3000$ m) {\color{blue} focuses} at $x=x_s=3500$ m.
+$m=[\kappa]$ converges $\leftrightarrow$ $\alpha$ increases, extended source $\of$ (initially spread over $z=z_s=1000$ m) {\color{blue} focuses} at $x=x_s=3500$ m.
 \begin{center}
-\hspace{-1cm}\includegraphics[height=1.25in]{Fig/cgw0_est_source_plh0.pdf}\includegraphics[height=1.25in]{Fig/cgw8_est_source_plh0.pdf}\includegraphics[height=1.25in]{Fig/cgw18_est_source_plh0.pdf}\includegraphics[height=1.25in]{Fig/cgw40_est_source_plh0.pdf}\\
+\hspace{-1cm}\includegraphics[height=1.25in]{Fig/cgw1estsourceplh0.pdf}\includegraphics[height=1.25in]{Fig/cgw7estsourceplh0.pdf}\includegraphics[height=1.25in]{Fig/cgw29estsourceplh0.pdf}\\
 \vspace{0.25in}
-Left to right: $\of$ at Iterations 1 ($\alpha=2.9 e-6$), 9
-($\alpha= 5.2 e-6$ m) 19 ($\alpha=1.32 e-5$) and 41 ($\alpha=1.99e-5$).
+Left to right: $\of$ at Iterations 1 ($\alpha=4.83 \times 10^{-6}, r=2000$ m), 8 ($\alpha= 1.32 \times 10^{-5}, r=1000$ m) and 30 ($\alpha=9.1 \times 10^{-5}, r=250$ m).
 \end{center}
@@
 \begin{center}
-\includegraphics[height=1.25in]{Fig/covmestfwicgw41.png}\includegraphics[height=1.25in]{Fig/residcovmestfwicgw41wind.pdf}\\
+\includegraphics[height=1.25in]{Fig/fwitrsse1bulkmodinvlh0.pdf}\includegraphics[height=1.25in]{Fig/bml0.pdf}\\
 \vspace{0.25in}
-Left: ``Jump to $\alpha=\infty$": bulk modulus estimated by FWI, 200
-steepest descent steps starting with iteration 41 of SSE., final RMS residual reduction = 0.05;
-Right: FWI data residual 
+Left: ``Jump to $\alpha=\infty$": bulk modulus estimated by FWI, starting with iteration 30 of SSE. 18 steps KGN, each $\le$ inner 20 CG steps, final RMS residual reduction = 0.012;
+Right: Target bulk modulus field
 \end{center}
 \end{frame}
-
-\begin{frame}
-  Why it works - a hint:
-
-provided that
-\begin{itemize}
-\item $d = F[m_*]s$,
-\item RMS wavelength of $w$ (hence $d$) $\rightarrow 0$,
-\end{itemize}
-\[
-  \nabla \nabla  J_{\rm SSE}[m_*] \rightarrow \sum_{\bx,\bp}
-\]
-\end{frame}
-
-\end{document}
 

### 2025-05-29 repro/papers/ssevelan/slides/Fig/gv5ref/PlotGv5Ref.m

@@ -1,0 +1,32 @@
+
+
+%% Plot 2D Data;
+
+xrec = 0:0.01:3;
+xtk = 0:1:3; ytk=0:0.5:1.5; fontsize=15;
+xi = x/1000; zi = z/1000;
+
+%% Plot velocity    
+Position = [500, 300,560, 290]; modscale = [1.4 2.5];
+Plot2DImageModel(xi,zi,reshape(v*1e-3,n),xtk,ytk,fontsize,Position, modscale);
+% hold on;
+% plot(model.xsrc/1000,model.zsrc/1000,'g*','MarkerSize',10)
+% plot(model.xrec/1000,model.zrec/1000,'bv','MarkerSize',5)
+
+
+%% Plot Data 
+xtkd = [0.5 1 1.5 2 2.5];ytkd = [0 0.5 1 1.5];
+TdPos = [500, 200,400, 400]; tdscale = [-30 40]; xlabelname = 'Receiver (km)';
+subidx = 1:451;
+for pp = 1:nsrc
+    rcv = nd(1)*(pp-1)+[1:nd(1)];
+    Plot2DTimeGather(xrec,t(subidx),-Dpt(rcv,subidx)',xtkd,ytkd,fontsize,TdPos,tdscale,xlabelname)
+    Plot2DTimeGather(xrec,t(subidx),-Sft(rcv,subidx)',xtkd,ytkd,fontsize,TdPos,tdscale,xlabelname)
+end
+
+xtkf = [0.5 1 1.5 2 2.5];ytkf = [-1.5 -1 -0.5 0 0.5 1 1.5];
+
+delta_t  = t-2;
+DeltaSub = find(abs(delta_t)<=1.5);
+FdPos = [500, 200,400, 400]; fdscale = [-30 40]*1e-2; xlabelname = 'Distance (km)';
+ Plot2DTimeGather(xrec,delta_t (DeltaSub),F0t(:,DeltaSub)',xtkf,ytkf,fontsize,FdPos,fdscale,xlabelname)

### 2025-05-29 repro/papers/ssevelan/slides/Fig/gv5ref/Test_LineSrc_Gv5Ref.m

@@ -1,0 +1,225 @@
+%% Model
+clear all;
+startup;
+%matlabpool open local 12;
+% grid
+x = 0:10:3000;
+z = 0:10:1500;
+
+[o,d,n] = grid2odn(z,x);
+
+% velocity
+[xx,zz]=meshgrid(x,z);
+
+vel = 2000*ones(n);
+[xx,zz]=meshgrid(x,z);
+w1 = zeros(n(1),1);   i0=find(abs(z-600)<=400);     w1(i0) = tukeywin(length(i0),0.4);
+w2 = zeros(n(2),1);   j0=find(abs(x-1500)<=600);    w2(j0) = tukeywin(length(j0),0.2);
+
+gv =  -600*exp(-((xx-1500).^2/300^2+(zz-600).^2/300^2));
+
+
+vref = zeros(n);
+vref(21,:) = 500;
+vref(41,:) = 500;
+vref(61,:) = 500;
+vref(81,:) = 500;
+vref(101,:) = 500;
+
+S = opKron(opSmooth(n(2),2),opSmooth(n(1),2));
+
+v = vel + reshape(vref(:),n) + (w1*w2').*gv;
+v0 = vel;
+
+%% Smooth operator
+
+Bx = opSpline1D([100:100:2900],x,[0 0]); %% dirichlet bc
+Bz = opSpline1D([100:100:1600],z,[0 0]); %% dirichlet bc;
+B  = opKron(Bx, Bz);
+
+beta = 0;
+mul  = 1;
+z1=z/1000;x1=x/1000;
+Lap = opFunction(prod(n),prod(n),{@(g)lapsmoother2(g, z1, x1, beta,mul),@(g)lapsmoother2(g, z1, x1, beta,mul)},0,1);
+
+
+% parameters
+
+% model grid
+model.o = o;
+model.d = d;
+model.n = n;
+
+% absorbing boundary
+model.nb = [40 40]; %% the number of sampling in the ABCs domain.
+
+% source/receiver grid
+model.xsrc = [500:100:2500];
+model.xsrc = 1500;
+model.zsrc = 40;
+model.xrec = 20:20:2980;
+model.zrec = 20;
+
+%% time sample and frequency sampling rate via Nyquist law.
+% time 
+t  = 0:.004:4;   %% t=0:dt:T;
[…]

### 2025-05-31 repro/papers/ssevelan/slides/slides.tex
near final

@@
 \newcommand{\ddl}{\delta d_{\alpha}}
+
 \title[]{Waveform Inversion via Source Extension}
-\author[]{William W. Symes \inst{1}}
-\institute[]{\inst{1} Rice University}
-\date{Workshop in honor of Jean Virieux \\June 2025}
+\author[]{William W. Symes}
+\institute[]{\inst{} Rice University}
+\date{To Jean, for his many contributions and his leadership\\June 2025}
 
@@
 \renewcommand{\thefootnote}{\fnsymbol{footnote}}
-
-%\begin{frame}\frametitle{The Plan}
-%\begin{itemize}
-%\item What is ``extended modeling", and why bother
-%\item FWI via surface source extension - examples
-%\item Theory: why \& when it works
-%\item Zoology of source extensions
-%\end{itemize}
-%\end{frame}
 
@@
 
-For any $m$, determine $\bar{w}=\bar{w}[m]$ so that $F\bar{s} \approx d$. Update $m$ to drive
+For any $m$, determine $\bar{w}=\bar{w}[m]$ so that $F[m]\bar{s} \approx d$. Update $m$ to drive
 $\bar{w}(\bx,t;\bx_s) \rightarrow w(t)\delta(\bx-\bx_s)$ by minimizing
@@
 \item $d = F[m_*]s$,
+\item $\Gamma \subset$ phase space over source surface, initial data
+  for rays crossing receiver surface
+\item $\tau[m](\bx_s,\bp)$ = time to receiver surface along ray taking
+  off at $(\bx_s,\bp)$
 \item RMS wavelength of $w$ (hence $d$) $\rightarrow 0$,
@@
 \[
-  \nabla \nabla  J_{\rm SSE}[m_*] \rightarrow \sum_{\bx,\bp}
+  \mbox{Hess }  J_{\rm SSE}[m_*] \rightarrow \mbox{Hess } J_{\rm TT}[m_*]
 \]
-\end{frame}
+where $J_{\rm TT}[m] = \sum_{(\bx_s,p) \in \Gamma} (\nabla_p(\tau[m](\bx_s,\bp) -
+\tau[m_*](\bx_s,\bp)))^2$
 
-\end{document}
-
-\begin{frame}
-\begin{center}
-\vspace{-0.7in}
-\includegraphics[height=4.5in]{HMC1.pdf}
-\end{center}
-\end{frame}
-
-\begin{frame}
-HMC Salt/Carbonate Acoustic Model
-
-Streamer acquisition $z_s = 7$ m, $\Delta x_s=37.5$ m, 315 shots, $z_r=15$ m, $\Delta x_r=12.5$ m, 475 recvs, max offset = $6$ km
-
-Pulse for data generation = [2, 3, 6, 10] Hz trapezoidal bandpass filter
-
-Free surface @ $z=0$, other boundaries absorbing, 2-8 staggered FD scheme
-
-Regularization: penalize MS velocity gradient, weight (also $\alpha$) via several trials
-
-Data contains diving waves, pre- \& post-critical reflections, multiple reflections, refracted multiple reflections, head waves, free surface multiples, ....
+{\color{blue} SSE asymptotic to tomography-like limit}
 
@@
 \begin{frame}
-\begin{center}
-\vspace{-0.7in}
-\includegraphics[height=4.5in]{HMC2.pdf}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\begin{center}
-\vspace{-0.7in}
-\includegraphics[height=4.5in]{HMC3.pdf}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\begin{center}
-\vspace{-0.7in}
-\includegraphics[height=4.5in]{HMC4.pdf}
-\end{center}
-\end{frame}
-
-\section{Theory: why \& when it works}
-
-\begin{frame}
-FWI vs. SSE for transmission (diving wave, crosswell) data:
-
-$J_{\rm FWI}[m]$ and $J_{\rm SSE}[m]$ = regular functions of $m$, locally close to quadratic approximations.
-
-$d$ consistent: $d=Ru[m,f]$, $\of = f = w(t)\delta(\bx-\bx_s)$, $A\of=0$ $\Rightarrow$
+Summary:
 
 \begin{itemize}
-\item  $J_{\rm FWI}[m]=J_{\rm SSE}[m]=0$
-\vspace{0.5cm}
-\item $J_{\rm FWI}[m + \delta m] \approx \mbox{const. }\omega^2 \|\delta m\|^2$ for some $\delta m$ (kinematic perturbations), $\omega$ = center frequency in source pulse $w(t)$
-\vspace{0.5cm}
-\item $J_{\rm SSE}[m+\delta m] \le \mbox{const. }\|\delta m\|^2$ independent of $\omega, \delta m$
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-$\Rightarrow$ radius of convexity region:
-\begin{itemize}
-\item $J_{\rm FWI}: O(1/\omega)$, must initialize ``within a wavelength''
-\item $J_{\rm SSE}: O(1)$, converges even with (rel.) poor initial guess, HF data
+\item for acoustic transmission data, several source extension methods
+  perform ``tomography without picking'', avoid cycle-skipping,
+  various constraints
+\item no theoretical foundation for source extension with
+  reflection inversion
+\item beyond acoustics - elasticity? attenuation?
 \end{itemize}
 
-Method of proof: ray theory (``mocrolocal analysis'')
-
-More detailed analysis: minimizing $J_{\rm SSE}$ equivalent to a form of transmission travel time tomography: {\color{blue} tomography without picking}
-
-NB: for reflection data... ???
-\end{frame}
-
-\begin{frame}\frametitle{Late Breaking News}
-Theory for special reflection case: layered ($\kappa(z)$) medium, dip filter to eliminate postcritical/diving wave energy
-
-Explicit computation of $\nabla J_{\rm SSE}$ for 
-\begin{itemize}
-\item homogenous background 
-\item data from single reflector in otherwise slowly varying $\kappa$, wavelet with central frequency $\omega$
-\end{itemize}
-$\Rightarrow$ $\nabla J_{\rm SSE} \rightarrow 0$ as $\omega \rightarrow \infty$
-
-Same behavior as $\nabla J_{\rm FWI}$!!!
-
-Upshot: for pure reflection data, SSE may not provide constructive velocity updates
-\end{frame}
-
-\begin{frame}
-{\color{blue} {\bf Summary:}} Surface Source Extension
-
-\begin{itemize}
-\item overcomes cycle-skipping for transmission data, solid theoretical foundation
-\item same computational complexity as FWI
-\item open question: theory for reflections {\color{blue} (initial result not positive)}
-\end{itemize}
-%\end{frame}
-
-%\begin{frame}
-Other source extension methods (WRI, AWI, ...)
-
-\begin{itemize}
-\item Many promising examples 
-\item Little known theoretically - connection to travel time tomography???
-\end{itemize}
-\end{frame}
-
-\begin{frame}\frametitle{Model Extension}
-Add degrees of freedom to $v$, $\kappa$, instead of source
-
-Annihilator: flatten/focus gathers
-
-{\color{blue}Extracts velocity updates from reflection and refraction events} (S.
-86, Kern \& S.
-94, S.
-GP 08, Shen \& S.
-GEO 08, Shen EAGE 12, Biondi \& Almomin 12, 14, Chauris \& Cocher GEO 17, Hou \& S.
-GEO 18, Barnier \& Biondi SEG 19)
-
-Chief challenge: expense(model extension) $\gg$ expense(FWI) (Note: expense(source extension) $\approx$ expense(FWI))
-\end{frame}
-
-\begin{frame}\frametitle{Acknowledgements}
-\begin{itemize}
-\item Total E\&P USA
-\item Sponsors of The Rice Inversion Project
-\item Organizers of SS 1 (Recent Advances and Road Ahead) at SEG 2019
-\end{itemize}
+No shortage of open questions...
 \end{frame}
@@
 \end{document}
-
-\begin{frame}
-Matched Source Waveform Inversion (Hua \& Symes 1993, 4, Huang \& Symes 2017)
-
-\begin{itemize}
-\item similar to AWI (source wavelet for each trace), w/o normalization
-\item if
-\begin{itemize}
-\item transmission data (crosswell, diving wave)
-\item unique ray paths
-\end{itemize}
-then mean square moment $\approx$ mean square travel time misfit: {\color{blue} tomography without picking}
-
-\item relation with tomography fails ($\approx$ FWI) if multiple rays connect sources and receivers - contrast SSE
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\begin{center}
-Example 2: Deep Water Marmousi
-\end{center}
-\begin{center}
-\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmVt.pdf}\\
-Modified Marmousi Model: 2 km water layer added to minimize non-reflected energy, also box-snaped target
-\end{center}
-\end{frame}
-
-\begin{frame}
-\begin{center}
-\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot1}\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot81}\\
-Data: isotropic point source with 8 Hz Ricker wavelet, $z_r=0.04$ km, $x_r=$ 0.05 km to 9.15 km, $\Delta x_r=$ 0.05 km, $z_s=0.08$ km, $x_s=0.6$ km to $x_s=8.6$, $\Delta x_s$ = 0.1 km.
-Shots at $x_s=0.6$ km and $x_s=8.6$ km displayed.
-\end{center}
-\end{frame}
-
-\begin{frame}
-\begin{center}
-\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmV0.pdf}\\
-Initial $v(z)$ model for both FWI and SSE
-\end{center}
-\end{frame}
-
-\begin{frame}
-\vspace{-0.1cm}
-\begin{center}
-\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmFWI.pdf}\\
-FWI estimate of $v$: 200 iterations LBFGS, frequency domain FD 5-8 Hz
-\end{center}
-\end{frame}
-
-\begin{frame}
-\begin{center}
-\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmSurf.pdf}\\
-SSE estimate of $v$: 150 iterations LBFGS, frequency domain FD 5-8 Hz, fixed $\alpha$ set by trial
-\end{center}
-\end{frame}
-
-\begin{frame}
-%\vspace{0.07cm}
-\begin{center}
-\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmVt.pdf}\\
-Modified Marmousi Model: 2 km water layer added to minimize non-reflected energy, also box-snaped target
-\end{center}
-\end{frame}
-
-\begin{frame}
-\begin{center}
-\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot1Inv}\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot81Inv}\\
-Data predicted from SSE inverted model: shots at $x_s=0.6$ km and $x_s=8.6$ km.
-\end{center}
-\end{frame}
-
-\begin{frame}
-\begin{center}
-\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot1}\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot81}\\
-Target data:\\ shots at $x_s=0.6$ km and $x_s=8.6$ km.
-\end{center}
-\end{frame}
-
-\begin{frame}
-Wavefield Reconstruction Inversion: $\of(\bx,t;\bx_s)$ distributed throughout space-time
-
-van Leeuwen and Herrmann 2013, 2016, Wang \& Yingst 2016
-
-similar: space-time extension Huang \& S. 2018
-
-\begin{itemize}
-\item[+ ]many good examples, convincing numerical evidence
-\item[+ ]3D field data - Wang \& Yingst 2016
-\item[- ]basic form: store complete space-time (or space-frequency) fields (6D)
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-Adaptive Waveform Inversion (Warner \& Guatsch, 2014, 2016): $\of(\bx,t;\bx_r,\bx_s) = w(t;\bx_r)\delta(\bx-\bx_s)$ localized but depends on receiver coordinate $\bx_r$ - one wavelet per data trace
-
-similar: source-receiver extension Hua \& S. 1994, Huang \& S. 2017
-
-\begin{itemize}
-\item[+ ]additional storage: $\approx$ data volume (5D)
-\item[+ ]minor mod of standard FWI workflow
-\item[- ]source-receiver: loses advantage over FWI when sources, receivers connected by multiple ray paths (fails for lens model, above)
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-Intermediate: source distributed over space-time surface
-\begin{itemize}
-\item const. t surface = volume source extension - Huang \& S. 2018
-\item const z surface = {\color{blue} surface source extension (SSE):} 
-\[
-\of(\bx,t;x_s,y_s) = w(x,y; x_s,y_s)\delta(z-z_s)
-\]
-\end{itemize}
-\begin{itemize}
-\item[+ ]additional storage = O(data volume) (5D)
-\item[+ ]no problem with multiple ray paths
-\item[+ ]theory for transmission data (eg. diving wave): {\color{blue} tomography w/o picking}
-\end{itemize} 
-\end{frame}

### 2025-08-06 repro/papers/ssefound/paper.tex
minor

@@
 \end{abstract}
+
+\section{Scalar $\Psi$DO evolution}
+
+3D case: $\lambda(x,z,\xi,\eta) \in S^1(\bR^2) \subset
+C^{\infty}(\bR,S^1(\bR^2))$ not a $\Psi$DO on $\bR^3$ however can make
+one:
+
+Suppose $H \in OPS^0(\bR^3)$ is convolution operator, $\phi \in
+C^{\infty}_0(\bR^3)$. Let ${\cal N}_t = \{(t,x,z,\omega,\xi,\eta):
+\xi=\eta=0\}$ be the normal bundle of the constant-$t$ foliation, and
+suppose that $ES(H) \cap {\cal N}_t = \emptyset$. Then $p = \lambda \phi H
+\in OPS^1(\bR^3)$,
+\[
+p(x,z,t,\xi,\zeta,\omega) = \left(\int d\xi' d\zeta'
+\lambda(x,z,\xi+\xi',\zeta+\zeta'){\cal
+  F}_{(x,z)}\phi(t,\xi',\zeta')\right)H(\xi,\zeta,\omega)
+\]
+Suppose $u \in H^s_{\rm loc}(\bR^3)$ satisfies $(\partial_t + i\lambda(x,z,\partial_x,\partial_z))u \in
+H^s_{\rm loc}(\bR^3)$, and $(I-H)u \in H^{s+1}_{\rm loc}(\bR^3)$.
+%Also
+%choose $\phi_1 \in C^{\infty}_0(\bR^3)$ for which $\phi_1=1$ on
+%$\mbox{supp }\phi$.
+Then
+\[
+  p(x,z,t,\partial_x,\partial_z,\partial_t)u = \lambda \phi u \mbox{
+    mod } H^{s}_{\rm loc}
+\]
+\[
+  = \partial_t \phi u \mbox{ mod } H^{s}_{\rm loc}.
+\]
 
@@
 \]
-in which $\phi_,r \in C_0^{\infty}(\bR^{n+1})$ and $T>0$. Suppose that $K$
+in which $\phi_{s,r} \in C_0^{\infty}(\bR^{n+1})$ and $T>0$. Suppose that $K$
 is perturbed by $\delta K \in OPS^1$, also essentially skew. Then
@@
 $\delta F$. Our object is to understand the structure of $\delta F$, under the
-assumption that $\phi_r \delta K \in OPS^{-\infty}$.
-
-This assumption on $\delta K$ implies that $\phi_r \delta U \in
-OPS^{-\infty}$. 
-\[
-  \delta F =  \phi_r \delta S(T)  = \phi_r \left(\sum _{j=1}^k \delta U_j \tilde{S}_j(T) U_j^*
+assumption that $\phi_{s,r} \delta K \in OPS^{-\infty}$, i.e. that
+$\delta K$ essentially vanishes on the supports of $\phi_{s,r}$.
+
+This assumption on $\delta K$ implies that $\phi_{s,r} \delta U_i \in
+OPS^{-\infty}, i=1,...,k$. 
+\[
+  \delta F =  \phi_r \delta S(T) \phi_s  = \phi_r \left(\sum _{j=1}^k \delta U_j \tilde{S}_j(T) U_j^*
 \right.
@@
  +  \left. \sum_{j=1}^k U_j \delta\tilde{S}_j(T)U_j^* + \sum_{J=1}^k U_j
-    \tilde{S}_j(T)\delta U_j^*\right) 
-\]
-Every term in the first sum on the RHS is bounded $H^s \rightarrow H^{s-1}$ since
-$\phi_r \delta U_i \in OPS^{-\infty}$.  Since $[\phi_rU_i], \in OPS^{-1}$, 
-\[
-  \phi_r  U_j \delta \tilde{S}_j(T)  U_j^*  = U_j \phi_r
-  \tilde{S}_j(T) U_j^*
-  +...
-\]
-where the elided terms are operators mapping $H^s$ to $H^{s-1}$, any
-$s$. So,
-\[
- \delta F = \sum_{j=1}^k U_j \phi_r \left( \delta\tilde{S}_j(T) U_i^*
-   + \tilde{S}_j(T) \delta U_j^*\right)
-  + ...
+    \tilde{S}_j(T)\delta U_j^*\right) \phi_s
+\]
+Every term in the first and last sums on the RHS is smoothing, since
+$\phi_r \delta U_i, \delta U_i^* \phi_s\in OPS^{-\infty}$:
+\[
+  \delta F = \phi_r \left(\sum_{j=1}^k U_j
+    \delta\tilde{S}_j(T)U_j^*\right)\phi_s \mbox{ mod }OPS^{-\infty}
+\]
+Denote by ${\cal B}$ operators on $H^{\infty}$ that are bounded on
+each $H^s$, all $s \in \bR$ (so $OPS^0 \subset {\cal B}$
+
+Since $[\phi_{s,r},U_i] \in OPS^{-1}$, this is
+\[
+  = \sum_{j=1}^k U_j \phi_r \delta \tilde{S}_j(T) U_J^*\phi_s 
+  \mbox{ mod }{\cal B}
 \]
@@
 \[
- S(-T)\delta F =\left(\sum_{i=1}^k U_i \tilde{S}_i(T)
-    U_i^*\right) \left(\sum_{j=1}^k U_j \phi_r ( \delta\tilde{S}_j(T) U_i^*
-    + \tilde{S}_j(T) \delta U_j^*)\right)
-\]  
-Using the relation \ref{eqn:ortho},
+ S(-T)\delta F =\left(\sum_{i=1}^k U_i \tilde{S}_i(-T)
+    U_i^*\right) \left(\sum_{j=1}^k U_j \phi_r (\delta\tilde{S}_j(T) U_j^*\phi_s\right)
+\]
+Using the relation \ref{eqn:ortho} and $\tilde{S}_i(-t)\tilde{S}_i(t)=I$, this is
+\[
+  = \left(\sum_{i=1}^k U_i \tilde{S}_i(-T) \phi_r \delta\tilde{S}_i(T) U_i^*\phi_s
+)
+\]
 The first order perturbation of the definition \ref{eqn:stilde} is

### 2025-10-28 repro/papers/awi/paper.tex
minor

@@
 
-The version of AWI introduced by \cite{Warner:16} is based on an
+As introduced by \cite{Warner:16}, AWI is based on an
 active source acoustic model of data generation in Euclidean space
@@
 The parameter to be determined in the inverse problem studied here is
-the bulk modulus $\kappa$. It is presumed is bounded and smooth:
-\begin{equation}
-  \label{eqn:coeff}
-  \log \kappa \in C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3)
-\end{equation}
+the bulk modulus $\kappa$. It is presumed is bounded and smooth: $\log
+\kappa \in M$, where $M$ is a bounded subset of $C^{\infty}(\bR^3) \cap
+L^{\infty}(\bR^3)$, on which further constraints will be placed by and
+by.
+
 Under these hypotheses, the symmetric hyperbolic system \ref{eqn:awe}
@@
 to the system \ref{eqn:awe}) escape to infinity (\cite{Hristova:09},
-Theorem 2, \cite{EgorovShubin}, Theorem 2.104, p. 192). That is,
-exponential decay follows from a characteristic of $\kappa$.
+Theorem 2, \cite{EgorovShubin}, Theorem 2.104, p. 192). In particular,
+the Exponential Decay assumption is satisfied if for any $R > r>0$ and $\log
+\kappa \in M$, there exists $T>0$ so that any ray ${\bf X}:\bR^+\rightarrow \bR^3$
+  with ${\bf X}(0) \in B_r(0)$ satisfies $|{\bf X}(t)|>R$ for $t > T$. That is,
+exponential decay may follow from a propery of $M$.
 
@@
   \begin{itemize}
-  \item[2. ] for $0 <  \lambda \le 1$, 
+  \item[3. ] for $0 <  \lambda \le 1$, 
     \begin{equation}
@@
     \end{equation}
-  \item[3. ]for sufficiently small $\lambda$, 
+  \item[4. ]for sufficiently small $\lambda$, 
     \begin{equation}
@@
 value problem\ref{eqn:awe2}, then under the conditions imposed in the
-previous section, those to be imposed below),
-\[
-  \bv = \beta \int_{-\infty}^t p
+previous section,
+\[
+  \bv = \beta \int_{-\infty}^t \nabla p
 \]
@@
   \label{eqn:Sutz}
-  \Sla[\kappa]\uzl[\kappa;\dzl] \ge \frac{1}{2}\|\dzl\|
+  \|\Sla[\kappa]\uzl[\kappa;\dzl]\| \ge \frac{1}{2}\|\dzl\|
 \end{equation}
